################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
# Visualization
library(ggplot2)
library(colorspace)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/_habitats/")
```

```{r Load clean_names_lowercase function}
clean_names_lowercase <- function(x) {

  x <- iconv(x, from = "", to = "UTF-8", sub = "")
  x <- stringi::stri_trans_general(x, "Latin-ASCII")

  x %>%
    tolower() %>%
    trimws()
}
```

# Strategy: track changes in landings of major functional groups relative to mean landings  ------------------------------

# STEP 0: Load in Region_Key
```{r Region key}
Region_Key0<- read.csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Keys/Spatial_Keys/RegionKey_FINAL_May262026.csv")

colnames(Region_Key0)

Region_Key<- Region_Key0 %>%
  select(OHI_region_est, OHI_region_name) %>%
  distinct()
```

# STEP 1: Load in and summarize GOC_MarineLandings0

```{r Load in GOC_MarineLandings0 as necessary}
GOC_MarineLandings0<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Marine.Landings_Sector.Sorted.Georef_PATCHED_Jun032026.rds"))
# 3246583 obs., 58 var.

# NOTE TO SELF: This source file may change in later Summer / Fall run. - EMF, June 4 2026
# See the patch file here: https://docs.google.com/document/d/1L6w4Gd8SOjp14f_u46vyn64rEpv4qBRaN_V2IyPjP5w/edit?tab=t.0
```


# PATCH 1 - Assign groups that are unassigned to a Major_Functional_Group:
NOTE TO SELF: This patch should not be necessary when we re-run this in later Summer / Fall run. - EMF, June 15 2026
See the patch file here: https://docs.google.com/document/d/1L6w4Gd8SOjp14f_u46vyn64rEpv4qBRaN_V2IyPjP5w/edit?tab=t.0

```{r Load in GenusFamilyGroup_Key as needed}
GenusFamilyGroup_Key<- read.csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Keys/Taxonomic_Keys/GenusFamilyGroupKey_COMPLETE_Apr042026.csv")

# NOTE TO SELF: This patch should not be necessary when we re-run this in later Summer / Fall run. - EMF, June 15 2026
# See the patch file here: https://docs.google.com/document/d/1L6w4Gd8SOjp14f_u46vyn64rEpv4qBRaN_V2IyPjP5w/edit?tab=t.0

MajorFunctionalGroup_Key<- GenusFamilyGroup_Key %>%
  select(estimated_genus_name, Major_Functional_Group)
```

```{r Major_Functional_Group patch for GOC_MarineLandings}
# Fill in missing values based on MajorFunctionalGroup_Key:
GOC_MarineLandings <- GOC_MarineLandings0 %>%
  left_join(
    MajorFunctionalGroup_Key %>%
      select(estimated_genus_name, Major_Functional_Group),
    by = "estimated_genus_name",
    suffix = c("", ".lookup")
  )

GOC_MarineLandings <- GOC_MarineLandings %>%
  mutate(
    Major_Functional_Group = coalesce(
      Major_Functional_Group,
      Major_Functional_Group.lookup
    )
  ) %>%
  select(-Major_Functional_Group.lookup)

# 3246583 obs., 58 var.

# SANITY CHECK:
sum(is.na(GOC_MarineLandings$Major_Functional_Group))
# Only 26 records unassigned. Looks good!
```

# PATCH 2 - Re-assign litopenaeus as paneus:
NOTE TO SELF: This patch should not be necessary when we re-run this in later Summer / Fall run. - EMF, June 15 2026
See the patch file here: https://docs.google.com/document/d/1L6w4Gd8SOjp14f_u46vyn64rEpv4qBRaN_V2IyPjP5w/edit?tab=t.0

```{r Re-assign paneus as litopenaeus}
GOC_MarineLandings <- GOC_MarineLandings %>%
  mutate(
    estimated_genus_name = case_when(
      estimated_genus_name == "Litopenaeus" ~ "Penaeus",
      TRUE ~ estimated_genus_name
    )
  )

unique(GOC_MarineLandings$estimated_genus_name)
```


# **DECISION POINT: Selecting which sectors to include**
```{r Select Sectors_of_Interest}
table(GOC_MarineLandings$sector_estimate_draft)

MarineFishery_Sectors<- c("marine: fishery sector yet to be classified",
                          "marine: small-scale capture",
                          "marine: large-scale capture")
  
PossibleMarineFishery_Sectors<- c("freshwater or marine: large-scale capture",
                          "freshwater or marine: small-scale capture",
                          "freshwater or marine: fishery sector yet to be classified")

GOC_MarineFisheryLandings0<- GOC_MarineLandings %>%
  filter(sector_estimate_draft %in% c(MarineFishery_Sectors, PossibleMarineFishery_Sectors))

# 3137044 obs., 58 var.
```

```{r Re-level sector_estimate_draft and Major_Functional_Group}
# Re-level sectors (mainly for plotting purposes):
GOC_MarineFisheryLandings <- GOC_MarineFisheryLandings0 %>%
  filter(
    sector_estimate_draft %in% c(
      MarineFishery_Sectors,
      PossibleMarineFishery_Sectors
    )
  ) %>%
  mutate(
    sector_estimate_draft = factor(
      sector_estimate_draft,
      levels = c(
        "marine: large-scale capture",
        "marine: small-scale capture",
        "marine: fishery sector yet to be classified",
        "freshwater or marine: large-scale capture",
        "freshwater or marine: small-scale capture",
        "freshwater or marine: fishery sector yet to be classified"
      )
    )
  )

# SANITY CHECK: How many Major_Functional_Group unassigned?
sum(is.na(GOC_MarineFisheryLandings$Major_Functional_Group))
# Only 18 records unassigned. Looks good!

table(GOC_MarineFisheryLandings$Major_Functional_Group)

GOC_MarineFisheryLandings <- GOC_MarineFisheryLandings %>%
  mutate(
    Major_Functional_Group = factor(
      Major_Functional_Group,
      levels= c(
        
        # Crustaceans
        "Shrimps, prawns",
        "Crustaceans",
        
        # Molluscs
        "Cephalopods",
        "Molluscs excl. Cephalopods",
        
        # Other invertebrates
        "Echinoderms",
        "Jellies",
        "Other marine invertebrates",
        
        # Fishes
        "Anchovy & Sardine (small pelagics)",
        "Pelagic marine fish (excl. anchovy & sardine)",
        "Demersal marine fishes",
        "Marine fishes NEI",
        "Marine or freshwater fishes NEI",
        "Diadromous fishes",
        
        # Elasmobranchs
        "Sharks",
        
        # Miscellaneous / unresolved
        "Marine algae",
        "Marine species NEI",
        "unknown",
        NA_character_
      )
    )
  )

# SANITY CHECK:
sum(is.na(GOC_MarineFisheryLandings$Major_Functional_Group))
# Only 18 records unassigned. Looks good!
```

# STEP 2: Isolate Open Ocean Taxa (OCE)

```{r Isolate Open-Ocean taxa and create OCE_Landings0}
unique(GOC_MarineFisheryLandings$Major_Functional_Group)

Open_Ocean_Taxa<- c("Pelagic marine fish (excl. anchovy & sardine)",
"Anchovy & Sardine (small pelagics)",
"Sharks",
"Cephalopods"
)

colnames(GOC_MarineFisheryLandings)

OpenOcean_Landings0<- GOC_MarineFisheryLandings %>%
  filter(Major_Functional_Group %in% Open_Ocean_Taxa) %>%
    select(state, OHI_region_est,
         office_name_clean, landing_site_clean, 
         year,
         est_aquaculture_production, reported_fleet, 
         sector_estimate_reported, sector_estimate_draft,
         recorded_taxa_name, recorded_taxa_name_clean, common_name_clean, 
         recorded_taxa_group, estimated_taxa_group, Major_Functional_Group,
         estimated_species_name, estimated_genus_name, estimated_family_name,
         landed_weight, live_weight, weight_kg_corrected, value
         )

# 653887 obs., 22 vars.
```


# METHOD 1 - STEP 3: Calculate mean condition values for small pelagics, large pelagics, sharks, and cephalopods across all OHI regions

```{r Summarize OCE landings by region in OCE_RegionalLandings_summary}
OCE_RegionalLandings_summary<- OpenOcean_Landings0 %>%
  group_by(year, OHI_region_est, Major_Functional_Group) %>%
  summarize(
    total_landings_mt = sum(weight_kg_corrected, na.rm = TRUE)/1000,
    .groups = "drop"
  ) %>%
  mutate(
    Major_Functional_Group= clean_names_lowercase(Major_Functional_Group),
    OHI_region_est= clean_names_lowercase(OHI_region_est)
  )
# 853 obs.
```


## PLOT: regional open-ocean landings over time

```{r Join in region names and re-level for plotting purposes}
OCE_RegionalLandings_summary<- OCE_RegionalLandings_summary %>%
  left_join(Region_Key, 
            by = "OHI_region_est")

# Make sure to order the regions:
Region_order <- c(
  "Western Upper Gulf",
  "Western Midriff Islands",
  "Western Central Gulf",
  "Western Southern Gulf",
  "Eastern Upper Gulf",
  "Eastern Midriff Islands",
  "Southern Sonora Coast",
  "Sinaloa Coast",
  "Southeastern Gulf"
)

OCE_RegionalLandings_summary <- OCE_RegionalLandings_summary %>%
  mutate(
    OHI_region_name = factor(
      OHI_region_name,
      levels = Region_order
    )
  )
```

```{r Plot regional open-ocean landings over time}
colnames(OCE_RegionalLandings_summary)
unique(OCE_RegionalLandings_summary$OHI_region_name)

# Plot 1, panel per region:

OCE_Landings_perRegion_plot<- OCE_RegionalLandings_summary %>%
  filter(year >= 2014, year <= 2024) %>%
  ggplot(
    aes(
      x = year,
      y = log10(total_landings_mt),
      color = Major_Functional_Group,
      group = Major_Functional_Group
    )
  ) +
  geom_line() +
  
  theme_bw() +
  
  facet_wrap(~ OHI_region_name)

OCE_Landings_perRegion_plot


# Plot 2, panel per group:

Region_colors<- c(
  "Western Upper Gulf" = "#C03729",
  "Western Midriff Islands" = "#E68C7C",
  "Western Central Gulf" = "#FC8F24",
  "Western Southern Gulf" = "#dcaa38",
  "Eastern Upper Gulf" = "#cbd2a0",
  "Eastern Midriff Islands" = "#98a25a",
  "Southern Sonora Coast" = "#747428",
  "Sinaloa Coast" = "#627391",
  "Southeastern Gulf" = "#8d61a9")

OCE_Landings_perGroup_plot<- OCE_RegionalLandings_summary %>%
  filter(year >= 2014, year <= 2024) %>%
  ggplot(
    aes(
      x = year,
      y = total_landings_mt,
      color = OHI_region_name,
      group = OHI_region_name
    )
  ) +
  geom_line() +
  
  scale_color_manual(values = Region_colors) +
  
  theme_bw() +
  
  facet_wrap(~ Major_Functional_Group)

OCE_Landings_perGroup_plot
```


## PLOT: mean condition values for small pelagics, large pelagics, sharks, and cephalopods across all OHI regions

```{r First calculate OCE_taxa_regional_conditions}
# Calculate pre- and post-2018 means:
OCE_taxa_regional_conditions <- OCE_RegionalLandings_summary %>%
  mutate(
    period = if_else(year < 2018, "Pre-2018", "Post-2018")
  ) %>%
  group_by(
    Major_Functional_Group,
    OHI_region_est,
    OHI_region_name,
    period
  ) %>%
  summarise(
    mean_catch = mean(total_landings_mt, na.rm = TRUE),
    n_years = n_distinct(year),
    period_start = min(year, na.rm = TRUE),
    period_end = max(year, na.rm = TRUE),
    .groups = "drop"
  )

# 80 obs., 8 vars.
```

```{r Evaluate data quality}
colnames(OCE_taxa_regional_conditions)

OCE_taxa_regional_conditions <- OCE_taxa_regional_conditions %>%
  mutate(
    n_years_data = case_when(
      n_years < 3 ~ "less_than_3",
      n_years < 5 ~ "between_3_and_5",
      n_years >= 5 ~ "more_than_5"
    ),
    data_quality = case_when(
      n_years < 3 ~ "poor",
      n_years < 5 ~ "moderate",
      n_years >= 5 ~ "high"
    )
  )
# 80 obs., 10 vars.
```

```{r Create separate plots for each functional group per region}
# Create separate plots for each functional group per region:
plots <- OCE_RegionalLandings_summary %>%
  split(.$Major_Functional_Group) %>%
  lapply(function(dat) {
    
    group_name <- unique(dat$Major_Functional_Group)
    
    means <- OCE_taxa_regional_conditions %>%
      filter(Major_Functional_Group == group_name)
    
    ggplot(dat, aes(x = year, y = total_landings_mt)) +
      
      # Annual time series...
      geom_line(
        color = "grey40",
        linewidth = 0.7
      ) +
      
      # ... with points:
      geom_point(
        color = "grey40",
        size = 0.7
      ) +
      
      # Pre/post mean lines
      geom_segment(
        data = means,
        aes(
          x = xmin,
          xend = xmax,
          y = mean_catch,
          yend = mean_catch,
          color = period
        ),
        linewidth = 1,
        alpha= 0.8
      ) +
      
      # 2018 reference line
      geom_vline(
        xintercept = 2018,
        linetype = "dashed",
        color = "black"
      ) +
      
      # Region facets
      facet_wrap(
        ~ OHI_region_name,
        scales = "free_y"
      ) +
      
      scale_x_continuous(
        limits = range(OCE_RegionalLandings_summary$year),
        breaks = seq(
          min(OCE_RegionalLandings_summary$year),
          max(OCE_RegionalLandings_summary$year),
          by = 5
        )
      ) +
      
      scale_color_manual(
        values = c(
          "Pre-2018" = "steelblue",
          "Post-2018" = "firebrick"
        )
      ) +
      
      labs(
        title = group_name,
        x = "Year",
        y = "Total landings (mt)",
        color = "Mean period"
      ) +
      
      theme_bw()
  })


SmallPelagics_byRegion_plot<- plots$`anchovy & sardine (small pelagics)`
LargePelagics_byRegion_plot<- plots$`pelagic marine fish (excl. anchovy & sardine)`
Sharks_byRegion_plot<- plots$sharks
Cephalopods_byRegion_plot<- plots$cephalopods
```

```{r Save plots as PNGs}
# Small Pelagics
ggsave(
  plot = SmallPelagics_byRegion_plot,
  filename = sprintf("/home/ferrer/ohi_goc_prep/_habitats/HAB_open_ocean/SmallPelagics_byRegion_plot_Aug312026.png"),
  width = 12,
  height = 12,
  dpi = 300
)

# Large Pelagics
ggsave(
  plot = LargePelagics_byRegion_plot,
  filename = sprintf("/home/ferrer/ohi_goc_prep/_habitats/HAB_open_ocean/LargePelagics_byRegion_plot_Aug312026.png"),
  width = 12,
  height = 12,
  dpi = 300
)

# Sharks
ggsave(
  plot = Sharks_byRegion_plot,
  filename = sprintf("/home/ferrer/ohi_goc_prep/_habitats/HAB_open_ocean/Sharks_byRegion_plot_Aug312026.png"),
  width = 12,
  height = 12,
  dpi = 300
)

# Cephalopods
ggsave(
  plot = Cephalopods_byRegion_plot,
  filename = sprintf("/home/ferrer/ohi_goc_prep/_habitats/HAB_open_ocean/Cephalopods_byRegion_plot_Aug312026.png"),
  width = 12,
  height = 12,
  dpi = 300
)
```

# STEP 4 - METHOD 1 - STEP 4: calculate OCE_Scores per region

```{r Calculate OCE_Scores per region}
colnames(OCE_taxa_regional_conditions)
table(OCE_taxa_regional_conditions$data_quality)

OCE_Scores0 <- OCE_taxa_regional_conditions %>%
  select(OHI_region_est, OHI_region_name, Major_Functional_Group, period, data_quality, mean_catch) %>%
  group_by(
    OHI_region_est,
    OHI_region_name,
    Major_Functional_Group
  ) %>%
  summarize(
    OCE_Score_raw = mean_catch[period == "Post-2018"] /
                mean_catch[period == "Pre-2018"] * 100,
    OCE_Score_capped = pmin(OCE_Score_raw, 100),
    data_quality = paste(unique(data_quality), collapse = ", "),
    .groups = "drop"
  )

OCE_Scores_regional<- OCE_Scores0 %>%
  filter(!is.na(OHI_region_est)) %>%
  rename(OHI_region= OHI_region_est)
# 36 obs., 6 vars.
```

# SAVE OCE_Scores_regional ------------------------------

```{r Save OCE_Scores_regional as csv}
write.csv(OCE_Scores_regional, 
          "/home/ferrer/ohi_goc_prep/_habitats/HAB_open_ocean/OCE_outputs/OCE_Scores_regional_Aug312026.csv", 
          row.names = FALSE)
# 36 obs., 6 vars.
```

# METHOD 2 - STEP 3: Calculate mean condition values for small pelagics, large pelagics, sharks, and cephalopods across the entire GOC:

```{r Summarize OCE landings gulf-wide in OCE_GOCLandings_summary}
OCE_GOCLandings_summary<- OpenOcean_Landings0 %>%
  group_by(year, Major_Functional_Group) %>%
  summarize(
    total_landings_mt = sum(weight_kg_corrected, na.rm = TRUE)/1000,
    .groups = "drop"
  ) %>%
  mutate(
    Major_Functional_Group= clean_names_lowercase(Major_Functional_Group)
  )
# 104 obs., 3 vars.
```

## PLOT: gulf-wide open-ocean landings over time

```{r Plot gulf-wide open-ocean landings over time}
colnames(OCE_GOCLandings_summary)
unique(OCE_GOCLandings_summary$OHI_region_name)

# Plot 1, panel per region:

OCE_Landings_gulfwide_plot <- OCE_GOCLandings_summary %>%
  filter(year >= 2004, year <= 2024) %>%
  
  ggplot(
    aes(
      x = year,
      y = total_landings_mt,
      group = 1
    )
  ) +
  
  geom_line(
    color = "#2C7FB8",
    linewidth = 0.8
  ) +
  
  geom_point(
    shape = 16,
    size = 2,
    stroke = 1.2,
    color = "#2C7FB8"
  ) +
  
  geom_vline(
    xintercept = 2018,
    linetype = "dashed",
    color = "grey50",
    linewidth = 0.7
  ) +
  
  facet_wrap(
    ~ Major_Functional_Group,
    scales = "free_y"
  ) +
  
  labs(
    title = "Open-Ocean Landings by Functional Group",
    subtitle = "Gulf-wide annual landings, 2004–2024",
    y = "Total landings (metric tonnes)",
    x = "Year"
  ) +
  
  theme_bw(base_size = 14) +
  theme(
    strip.text = element_text(
      size = 14,
      face = "bold"
    ),
    plot.title = element_text(
      size = 16,
      face = "bold"
    ),
    plot.subtitle = element_text(
      size = 12
    )
  )

OCE_Landings_gulfwide_plot

```

```{r Save plot as PNG}
ggsave(
  plot = OCE_Landings_gulfwide_plot,
  filename = sprintf("/home/ferrer/ohi_goc_prep/_habitats/HAB_open_ocean/OCE_Landings_gulfwide_plot_Aug312026.png"),
  width = 12,
  height = 12,
  dpi = 300
)
```


```{r Calculate OCE_taxa_gulfwide_conditions}
# Calculate pre- and post-2018 means gulf-wide:

colnames(OCE_GOCLandings_summary)

OCE_taxa_gulfwide_conditions <- OCE_GOCLandings_summary %>%

# NOTE TO SELF: region has already been filtered to the GOC-level so there is no need to exclude values where region = NA.
  # filter(!is.na(OHI_region_est)) %>% 
  
  mutate(
    period = if_else(year < 2018, "Pre-2018", "Post-2018")
  ) %>%
  group_by(
    Major_Functional_Group,
    period
  ) %>%
  summarise(
    mean_catch = mean(total_landings_mt, na.rm = TRUE),
    n_years = n_distinct(year),
    period_start = min(year, na.rm = TRUE),
    period_end = max(year, na.rm = TRUE),
    .groups = "drop"
  )
# 8 obs., 6 vars.
```

```{r Evaluate data quality}
colnames(OCE_taxa_gulfwide_conditions)

OCE_taxa_gulfwide_conditions <- OCE_taxa_gulfwide_conditions %>%
  mutate(
    n_years_data = case_when(
      n_years < 3 ~ "less_than_3",
      n_years < 5 ~ "between_3_and_5",
      n_years >= 5 ~ "more_than_5"
    ),
    data_quality = case_when(
      n_years < 3 ~ "poor",
      n_years < 5 ~ "moderate",
      n_years >= 5 ~ "high"
    )
  )
# 8 obs., 8 vars.
```

## PLOT: mean condition values for small pelagics, large pelagics, sharks, and cephalopods gulf-wide

```{r Plot pre- and post-2018 by functional group per region}
mean_lines <- OCE_taxa_gulfwide_conditions %>%
  filter(period %in% c("Pre-2018", "Post-2018")) %>%
  mutate(
    period_start_plot = if_else(period == "Pre-2018", 2004, period_start),
    period_end_plot   = if_else(period == "Pre-2018", 2017, period_end)
  )


AllGroups_gulfwide_plot <- OCE_GOCLandings_summary %>%
  filter(year >= 2004, year <= 2025) %>%
  
  ggplot(
    aes(
      x = year,
      y = total_landings_mt
    )
  ) +
  
  geom_line(
    aes(group = Major_Functional_Group),
    color = "grey40",
    linewidth = 0.7
  ) +
  
  geom_point(
    color = "grey40",
    size = 0.7
  ) +
  
  geom_segment(
    data = mean_lines,
    aes(
      x = period_start_plot,
      xend = period_end_plot,
      y = mean_catch,
      yend = mean_catch,
      color = period,
      group = period
    ),
    linewidth = 1.2,
    alpha = 0.8
  ) +
  
  geom_vline(
    xintercept = 2018,
    linetype = "dashed",
    color = "black"
  ) +
  
  facet_wrap(
    ~ Major_Functional_Group,
    scales = "free_y"
  ) +
  
  scale_x_continuous(
    limits = c(2004, 2025),
    breaks = seq(2005, 2025, by = 5)
  ) +
  
  scale_color_manual(
    values = c(
      "Pre-2018" = "steelblue",
      "Post-2018" = "firebrick"
    )
  ) +
  
  labs(
    title = "Open-Ocean Landings by Functional Group",
    subtitle = "Gulf-wide annual landings and pre-/post-2018 mean catch",
    x = "Year",
    y = "Total landings (mt)",
    color = "Mean period"
  ) +
  
  theme_bw(base_size = 14) +
  theme(
    strip.text = element_text(
      size = 14,
      face = "bold"
    ),
    plot.title = element_text(
      size = 16,
      face = "bold"
    )
  )

AllGroups_gulfwide_plot

```

```{r Save plot as PNG}
# All Groups
ggsave(
  plot = AllGroups_gulfwide_plot,
  filename = sprintf("/home/ferrer/ohi_goc_prep/_habitats/HAB_open_ocean/AllGroups_gulfwide_plot_Aug312026.png"),
  width = 12,
  height = 12,
  dpi = 300
)
```


# METHOD 2 - STEP 4: calculate OCE_Scores based on gulf-wide calculations:

```{r Calculate OCE_Scores_gulfwide}
colnames(OCE_taxa_gulfwide_conditions)
table(OCE_taxa_gulfwide_conditions$data_quality)

OCE_Scores_gulfwide <- OCE_taxa_gulfwide_conditions %>%
  select(Major_Functional_Group, period, data_quality, mean_catch) %>%
  group_by(
    Major_Functional_Group
  ) %>%
  summarize(
    OCE_Score_raw = mean_catch[period == "Post-2018"] /
                mean_catch[period == "Pre-2018"] * 100,
    OCE_Score_capped = pmin(OCE_Score_raw, 100),
    data_quality = paste(unique(data_quality), collapse = ", "),
    .groups = "drop"
  )
# 4 obs., 4 vars.
```

# SAVE OCE_Scores_gulfwide ------------------------------

```{r Save OCE_Scores_gulfwide as csv}
write.csv(OCE_Scores_gulfwide, 
          "/home/ferrer/ohi_goc_prep/_habitats/HAB_open_ocean/OCE_outputs/OCE_Scores_gulfwide_Aug312026.csv", 
          row.names = FALSE)
# 4 obs., 4 vars.
```


# METHOD 3 (RECOMMENDED) - STEP 4: calculate OCE_Scores based on gulf-wide calculations time-varying:

```{r Calculate OCE_Scores_gulfwide_timevarying}
# Calculate gulf-wide condition per taxonomic group in all years 2018-2024 compared to pre-2018 reference point:

colnames(OCE_GOCLandings_summary)

OCE_taxa_gulfwide_references <- OCE_GOCLandings_summary %>%
  filter(year < 2018) %>%
  group_by(
    Major_Functional_Group
  ) %>%
  summarise(
    mean_catch_pre2018 = mean(total_landings_mt, na.rm = TRUE),
    n_years_pre2018 = n_distinct(year[!is.na(total_landings_mt)]),
    period_start = min(year[!is.na(total_landings_mt)], na.rm = TRUE),
    period_end = max(year[!is.na(total_landings_mt)], na.rm = TRUE),
    .groups = "drop"
  )

# 4 obs., 5 vars.

OCE_Scores_gulfwide_timevarying <- OCE_GOCLandings_summary %>%
  filter(year >= 2018, year <= 2024) %>%
  left_join(
    OCE_taxa_gulfwide_references,
    by = "Major_Functional_Group"
  ) %>%
  mutate(
    OCE_Score_raw = (
      total_landings_mt / mean_catch_pre2018
    ) * 100,
    OCE_Score_capped = pmin(OCE_Score_raw, 100),
  )
# 28 obs., 8 vars.
```

# SAVE OCE_Scores_gulfwide_timevarying ------------------------------

```{r Save OCE_Scores_gulfwide_timevarying as csv}
write.csv(OCE_Scores_gulfwide_timevarying, 
          "/home/ferrer/ohi_goc_prep/_habitats/HAB_open_ocean/OCE_outputs/OCE_Scores_gulfwide_timevarying_Aug312026.csv", 
          row.names = FALSE)
# 28 obs., 8 vars.
```


# PLOT OCE_Scores ??? PENDING ??? ------------------------------

```{r Level region_order for plotting purposes}
# Make sure to order the regions:
region_order <- c(
  "Western Upper Gulf",
  "Western Midriff Islands",
  "Western Central Gulf",
  "Western Southern Gulf",
  "Eastern Upper Gulf",
  "Eastern Midriff Islands",
  "Southern Sonora Coast",
  "Sinaloa Coast",
  "Southeastern Gulf"
)

OCE_Scores <- OCE_Scores %>%
  mutate(
    OHI_region_name = factor(
      OHI_region_name,
      levels = region_order
    )
  )
```

```{r Plot OCE_Scores}
Region_colors<- c(
  "Western Upper Gulf" = "#C03729",
  "Western Midriff Islands" = "#E68C7C",
  "Western Central Gulf" = "#FC8F24",
  "Western Southern Gulf" = "#dcaa38",
  "Eastern Upper Gulf" = "#cbd2a0",
  "Eastern Midriff Islands" = "#98a25a",
  "Southern Sonora Coast" = "#747428",
  "Sinaloa Coast" = "#627391",
  "Southeastern Gulf" = "#8d61a9")

OCE_Scores_overTime<- OCE_Scores %>%
  
  filter(!is.na(OHI_region_name)) %>%
  
  ggplot(aes(x = year, y = OCE_Score, color = OHI_region_name)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  
  facet_wrap(~ OHI_region_name, scales = "fixed") +
  
  labs(
    title = "OCE Scores by Region",
    subtitle= "updated: Jul 22 2026",
    y = "FIS Score (from 0 to 100)",
    x = "Year"
  ) +
  
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey50") +
  
  ylim(0, 101) +
  
  scale_color_manual(values = Region_colors) +
  
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 16, face = "bold")
  )

OCE_Scores_overTime
```

```{r Save as PNG}
  ggsave(
    plot = OCE_Scores_overTime,
    filename = sprintf("/home/ferrer/ohi_goc_prep/BD/BD_Scores/OCE_Scores/OCE_Scores_plot_Jul222026.png"),
    width = 12,
    height = 12,
    dpi = 300
  )
```