################################################################################ 

For questions contact: Erica Ferrer, Ph.D. at emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################ 

# PRE-WORK ------------------------------

```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/NP/v2025/")
```

```{r Outdated natural products key}
NPKey_pending<- read.csv('/home/ferrer/ohi_goc_prep/NP/v2025/NP_Keys/NP_MasterList_UNDER CONSTRUCTION_Nov132025.csv')
```

```{r Load GOC_Landings_sector if needed}
GOC_Landings_sector<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Landings_SectorSorted_2000.present_EMF.Apr032026.rds"))
# 3598085 obs., 62 vars.
```

```{r Filter for marine sectors}
table(GOC_Landings_sector$sector_estimate_draft)

Possible_Sectors<- c("marine: aquaculture", "marine: fishery sector yet to be classified", 
                        "marine: large-scale capture", "marine: small-scale capture",
                        "freshwater or marine: large-scale capture", "freshwater or marine: fishery sector yet to be classified",
                        "freshwater or marine: small-scale capture", "freshwater or marine: aquaculture")

GOC_MarineLandings<- GOC_Landings_sector %>% 
  filter(sector_estimate_draft %in% Possible_Sectors)
# 3312602 obs., 62 vars.

table(GOC_MarineLandings$keywords_1)

# Filter out things from freshwater:
GOC_MarineLandings <- GOC_MarineLandings %>%
  filter(!if_any(starts_with("keywords"), ~ !is.na(.) & . == "DE AGUA DULCE"))
# 3303634 obs., 62 vars.
```



# Flag all things that are very likely to be natural products based on key terms...
```{r Estimate likelihood of ornamentals based on name of landing_site and eu_name}
#### FLAG - Flag things that are very likely to be natural products based on landing_site, keywords, key groups, and key taxa (2 levels: YES, NO, MAYBE): ####

# Clues from keywords:
table(GOC_MarineLandings$keywords_1)

# Keywords often linked to natural products:
Keyword_strong.clues<- c("VIVO", "VIVA", "VIVAS", "VIVITA", # living / live
                         
                         # Used as bait:
                         "CARNADA", "P/CARNADA",

                         # Shells and pearls:
                         "CONCHA DE", "CONCHA", "CONCHA NACAR",
                         "PERLAS", "MEDIAS PERLAS",
                         # Fin:
                         "ALETA", "ALETA ECA DE", "ALETA FCA DE",
                         
                         # Animal parts:
                         "MANTO", "MANTO DE",
                         "PULPA DE", "PULPA FRESCA", "PULPO DE", 
                         "POLVO DE",
                         "TENTACULOS DE", "TUBO DE", "TUBO",
                         "CABEZA DE", "BUCHE", "BUCHE DE", "TRIPA DE", "COLAS DE", "MANOS DE", "MANOS",
                         "ANCAS DE", "ANCAS", "ANCA",
                         "PIEL DE ECA", "PIEL DE SECA", "PIEL DE", "PIEL",
                         
                         # Dried things:
                         "SECO DE", "SECA", "SECA DE", "SECO", "SECO DE"
)

# Keywords sometimes linked to natural products:
Keyword_weak.clues<- c("DE CULTIVO", "DE CULTIVO VIVO",
                       "CULTIVADA", "CULTIVADO", "CULTIVADAS", "CULTIVADOS", # A few new adds as of May 04
                       "ABASTO RANCHO", "RANCHO",
                       # Eggs:
                       "HUEVA DE")

# Keywords typically linked to FOOD, not natural products
Keyword_opposite.clues<- c(
  # Processing state or product form (e.g., frozen, full fish, filleted, beheaded, dried, etc.):
  "S.C.", "S/C", "S",
  "FRESCO", "FRESCA", "FCA", "FCO", "FC",
  "ECO",
  "CONG", "CONGELADO",
  "COCINERO", "COCIDO", "COCIDA", "COCIDAS", "ASADO",
  "ENTERO", "ENTERA", "IND", "ENT",
  "FILETE", "FILETE DE", "ECA", "FILETE DE ECO", "DESHUESADA", "CARNE", "CARNE DE",
  "DESCABEZADO", "DES", "DSC", "DESC", "DESV", "DESVICERADO"
)

# Clues from landing_site and eu_name:
LandingSite_clues<- c("CULTIVADORES", "ESTANQUE", "VIVEROS", "ACUACULTURA") 

# Clues from recorded_taxa_group:
table(GOC_MarineLandings$recorded_taxa_group)
TaxaGroup_strong.clues<- c("PECES DE ORNATO", "ORNATO") 

# OTRAS that could be ornamental fishes:
TaxaName_strong.clues<- c("PAYASO", "PECES ORNATO", "CABALLITO MAR", 
                            "ESTRELLA MAR", "ESTRELLA LIMPIADORA", "GALLETA MAR", "CARACOL",
                            "HACHA")

colnames(GOC_MarineLandings)

# Build regex patterns dynamically
Keyword_strong.pattern <- paste0("\\b(", paste(Keyword_strong.clues, collapse = "|"), ")\\b")
Keyword_weak.pattern <- paste0("\\b(", paste(Keyword_weak.clues, collapse = "|"), ")\\b")
Keyword_opposite.pattern <- paste0("\\b(", paste(Keyword_opposite.clues, collapse = "|"), ")\\b")
LandingSite_pattern <- paste0("\\b(", paste(LandingSite_clues, collapse = "|"), ")\\b")
TaxaGroup_pattern <- paste0("\\b(", paste(TaxaGroup_strong.clues, collapse = "|"), ")\\b")
TaxaName_pattern <- paste0("\\b(", paste(TaxaName_strong.clues, collapse = "|"), ")\\b")


GOC_Possible.NPLandings <- GOC_MarineLandings %>%
  
  mutate(
    NP.Keyword_FLAG = case_when(
      if_any(keywords_1:keywords_5, ~ str_detect(.x, regex(Keyword_strong.pattern, ignore_case = TRUE))) ~ "YES",
      if_any(keywords_1:keywords_5, ~ str_detect(.x, regex(Keyword_weak.pattern, ignore_case = TRUE))) ~ "MAYBE",
      if_any(keywords_1:keywords_5, ~ str_detect(.x, regex(Keyword_opposite.pattern, ignore_case = TRUE))) ~ "NO",
      TRUE ~ "NO"
    ),
    NP.Site_FLAG = case_when(
      str_detect(landing_site, regex(LandingSite_pattern, ignore_case = TRUE)) ~ "YES",
      TRUE ~ "NO"
    ),
    NP.Group_FLAG = case_when(
      str_detect(recorded_taxa_group, regex(TaxaGroup_pattern, ignore_case = TRUE)) ~ "YES",
      TRUE ~ "NO"
    ), 
    NP.TaxaName_FLAG = case_when(
      str_detect(recorded_taxa_name_clean, regex(TaxaName_pattern, ignore_case = TRUE)) ~ "YES",
      TRUE ~ "NO"
    ),
    # Previously made ornamental flag:
      NP.PreviousOrnamental_FLAG = case_when(
        ornamentalFLAG_1 == "YES" ~ "YES",
        ornamentalFLAG_1 == "POSSIBLY" ~ "MAYBE",
        TRUE ~ "NO"
        )
  )

# Now exclude any entries that don't have at least one YES flag.
GOC_Possible.NPLandings <- GOC_Possible.NPLandings %>%
  filter(
    if_any(
      starts_with("NP"),
      ~ . == "YES"
    )
  )
# 107256 obs. of 66 vars.
```

```{r Make an alive or dead column}
test<- GOC_Possible.NPLandings %>%
  filter(recorded_taxa_name_clean=="PAYASO")

Dead_keywords<- c(
  
  # Frozen, filleted, beheaded, dried, fresh, etc.
  "S.C.", "S/C", "S",
  "ECO",
  "CONG", "CONGELADO",
  "COCINERO", "COCIDO", "COCIDA", "COCIDAS", "ASADO",
  "FILETE", "FILETE DE", "ECA", "FILETE DE ECO", "DESHUESADA", "CARNE", "CARNE DE",
  "DESCABEZADO", "DES", "DSC", "DESC", "DESV", "DESVICERADO",
  
  # Dried:
  "SECO DE", "SECA", "SECA DE", "SECO", "SECO DE",
  
  # Shells and pearls:
  "CONCHA DE", "CONCHA", "CONCHA NACAR",
  "PERLAS", "MEDIAS PERLAS",
  
  # Fin:
  "ALETA", "ALETA ECA DE", "ALETA FCA DE",
  
  # Animal parts:
  "MANTO", "MANTO DE",
  "PULPA DE", "PULPA FRESCA", "PULPO DE", 
  "POLVO DE",
  "TENTACULOS DE", "TUBO DE", "TUBO",
  "CABEZA DE", "BUCHE", "BUCHE DE", "TRIPA DE", "COLAS DE", "MANOS DE", "MANOS",
  "ANCAS DE", "ANCAS", "ANCA",
  "PIEL DE ECA", "PIEL DE SECA", "PIEL DE", "PIEL"
)

Alive_keywords<- c("VIVO", "VIVA", "VIVAS", "VIVITA") # living / live

# Assign alive, dead, unknown based on keywords:
GOC_Possible.NPLandings <- GOC_Possible.NPLandings %>%
  mutate(
    Alive.Dead = case_when(
      
      # DEAD FIRST (safer precedence)
      if_any(keywords_1:keywords_5,
             ~ str_detect(.x, str_c(Dead_keywords, collapse = "|"))) ~ "dead",
      
      # ALIVE
      if_any(keywords_1:keywords_5,
             ~ str_detect(.x, str_c(Alive_keywords, collapse = "|"))) ~ "alive",
      
      # DEFAULT
      TRUE ~ "unknown"
    )
  )

# Assign bait column:
  
  # Used as bait:
  "CARNADA", "P/CARNADA",
  
  # Whole fish? Not sure...
    "ENTERO", "ENTERA", "IND", "ENT",
    "FRESCO", "FRESCA", "FCA", "FCO", "FC",
```

# AQUARIUM TRADE SPECIES: LIVE ORNAMENTAL FISH AND INVERTS ------------------------------
Run list of species through AI and flag any species that are almost ALWAYS ornamentals.

```{r Check ornamentals flag}
table(GOC_Possible.NPLandings$ornamentalFLAG_1)
      # NO POSSIBLY      YES 
      #  1    19726     1602

table(GOC_Possible.NPLandings$NP.PreviousOrnamental_FLAG)
# MAYBE    NO   YES 
# 19726 85928  160
# Looks good!
```

```{r Pull out possible ornamentals}
colnames(GOC_Possible.NPLandings)
table(GOC_Possible.NPLandings$Alive.Dead)

GOC_Possible.OrnamentalLandings<- GOC_Possible.NPLandings %>%
  filter(
    (NP.PreviousOrnamental_FLAG %in% c("YES", "MAYBE") |
       NP.Group_FLAG %in% c("YES", "MAYBE") |
       NP.TaxaName_FLAG %in% c("YES", "MAYBE")) &
      Alive.Dead %in% c("alive", "unknown")
  )
           
# 20107 obs. of 67 vars.

table(GOC_Possible.OrnamentalLandings$Alive.Dead)
  # alive unknown 
  #   430   19677 
```

```{r Look into flagged ornamental taxa}
# Taking the information I've gleaned here to refine my sector estimates above.
table(GOC_Possible.OrnamentalLandings$estimated_genus_name)
table(GOC_Possible.OrnamentalLandings$Major_Functional_Group)
table(GOC_Possible.OrnamentalLandings$recorded_taxa_name_clean)

# Investigate Major_Taxonomic_Group unknown:
subset <- GOC_Possible.OrnamentalLandings %>% 
  filter(Major_Functional_Group %in% c("unknown"))
# 1168 obs., 62 vars.

table(subset$recorded_taxa_group)
table(subset$common_name_clean)
# Makes sense; all ornamental species.
```

```{r Pull out the taxa most likely to be sold alive for the aquarium trade}
# Very likely: 
# CABALLITO MAR → seahorse (very commonly traded live in marine aquariums)
# PAYASO → clownfish (Amphiprion spp.), one of the most common aquarium fish globally
# PECES ORNATO → literally “ornamental fish,” a general aquarium trade category

# Somewhat likely:
# CARACOL CHINO / CHINO NEGRO / CHINO ROSA / CHINO ROSADO → likely nerite-type snails (widely sold for algae control)
# CARACOL TURBO FLUCTUOSO → turbo snails (classic reef aquarium cleaners)
# CARACOL OLIVO → olive snails, sometimes kept in marine tanks
# CARACOL MARGARITA → margarita snails (cold/temperate marine aquariums)
# CARACOL UÑA / CARACOL UÑA → uncertain ID, but likely small ornamental snail types
# CARACOL CHILE / CARACOL BURRO / CARACOL PANOCHA → less standard names, but still plausibly marine snails traded as live clean-up organisms depending on region
# ESTRELLA LIMPIADORA → “cleaner star” / sand-sifting or detritus-feeding sea stars (commonly sold in marine aquariums)

# Not likely:
# ESTRELLA MAR → sea stars (some species are aquarium-kept but not common)
# GALLETA MAR → sand dollar / sea biscuit (sometimes sold live but difficult to maintain)
# HACHA CHINA / HACHA LARGA / CALLO HACHA → scallops or bivalves; occasionally live in holding systems but not typical aquarium pets
# MADRE PERLA → pearl oyster; generally not an aquarium species

AquariumTrade_LikelyTaxa<- c("CABALLITO MAR", "PAYASO", "PECES ORNATO")
AquariumTrade_PossibleTaxa<- c("CARACOL CHINO", "CARACOL CHINO NEGRO", "CARACOL CHINO ROSA", "CARACOL CHINO ROSADO",
                               "CARACOL TURBO FLUCTUOSO", "CARACOL OLIVO", "CARACOL MARGARITA",
                               "CARACOL UÑA", "CARACOL CHILE", "CARACOL BURRO", "CARACOL PANOCHA",
                               "ESTRELLA LIMPIADORA")
AquariumTrade_UnlikelyTaxa<- c("ESTRELLA MAR", "GALLETA MAR",
                               "HACHA CHINA", "HACHA LARGA", "CALLO HACHA",
                               "MADRE PERLA")

# Assign ornamental likely or unlikely:
GOC_Possible.OrnamentalLandings <- GOC_Possible.OrnamentalLandings %>%
  mutate(
    AquariumTradeLikely = case_when(
      recorded_taxa_name_clean %in% AquariumTrade_LikelyTaxa ~ "YES",
      recorded_taxa_name_clean %in% AquariumTrade_PossibleTaxa ~ "MAYBE",
      recorded_taxa_name_clean %in% AquariumTrade_UnlikelyTaxa ~ "NO",
      TRUE ~ "unknown"
    )
  )
# 20107 obs., 69 vars.

GOC_OrnamentalLandings<- GOC_Possible.OrnamentalLandings %>%
  filter(AquariumTradeLikely %in% c("YES", "MAYBE"))
# 15292 obs., 69 vars.
```

```{r Plot ornamentals}
# Summary table by region and year:
Ornamentals_summary <- GOC_OrnamentalLandings %>%
  group_by(OHI_rgn_number, year, recorded_taxa_name_clean) %>%
  summarize(annual_prod_kg = sum(weight_kg_corrected))
# 462 obs., 4 vars.

# Plot over time:
Ornamentals_summary %>%
  filter(recorded_taxa_name_clean %in% c("PECES ORNATO", "PAYASO")) %>%
  filter(year >= 2018 & year < 2025) %>%
  
  ggplot(aes(x = year, y = annual_prod_kg, color= recorded_taxa_name_clean)) +
  geom_line() +
  
  facet_wrap(~ OHI_rgn_number, scales = "free_y") +
  
  labs(x = "Year", y = "Production (kg)",
       title = "Aquarium Trade / Production of Live Ornamental Taxa [DRAFT]") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
# Residues
```{r Residues}
#### FLAG - Flag residues keywords (3 levels: YES, NO): ####

Residues_keywords <- c(
  # Animal parts:
  "MANTO", "MANTO DE",
  "PULPA DE", "PULPA FRESCA", "PULPO DE", 
  "POLVO DE",
  "TENTACULOS DE", "TUBO DE", "TUBO",
  "CABEZA DE", "BUCHE", "BUCHE DE", "TRIPA DE", "COLAS DE", "MANOS DE", "MANOS",
  "ANCAS DE", "ANCAS", "ANCA",
  "PIEL DE ECA", "PIEL DE SECA", "PIEL DE", "PIEL")

GOC_Possible.NPLandings <- GOC_Possible.NPLandings %>%
  mutate(
    ResiduesLikely = case_when(
      if_any(keywords_1:keywords_5, ~ .x %in% Residues_keywords) ~ "YES",
      TRUE ~ "NO"
    )
  )

GOC_Possible.Residues<- GOC_Possible.NPLandings %>%
  filter(ResiduesLikely %in% c("YES"))

table(GOC_Possible.Residues$Major_Functional_Group)
# Lots of cephalopods, demersal fish, and crustaceans
# NOTE TO SELF: Don't forget Aleta though!! This DOES NOT include Aleta ... yet - Apr 30 2026
```

```{r Plot residues}
# Summary table by region and year:
Residues_summary <- GOC_Possible.Residues %>%
  group_by(OHI_rgn_number, year, Major_Functional_Group) %>%
  summarize(annual_prod_kg = sum(weight_kg_corrected))
# 342 obs., 4 vars.

# Plot over time:
Residues_summary %>%
  filter(year >= 2018 & year < 2025) %>%
  
  ggplot(aes(x = year, y = annual_prod_kg, color= Major_Functional_Group)) +
  geom_line() +
  
  facet_wrap(~ OHI_rgn_number, scales = "free_y") +
  
  labs(x = "Year", y = "Production (kg)",
       title = "Residues Production over Time [DRAFT]") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

# BAIT ------------------------------
```{r Bait}
Bait_keywords <- c(
# Used as bait:
  "CARNADA", "P/CARNADA"
)

GOC_Possible.NPLandings <- GOC_Possible.NPLandings %>%
  mutate(
    BaitLikely = case_when(
      if_any(keywords_1:keywords_5, ~ .x %in% Bait_keywords) ~ "YES",
      TRUE ~ "NO"
    )
  )

GOC_BaitLandings<- GOC_Possible.NPLandings %>%
  filter(BaitLikely %in% c("YES"))
# 15292 obs., 70 vars.

table(GOC_BaitLandings$Major_Functional_Group)
           # Anchovy & Sardine (small pelagics)                                   Crustaceans 
           #                               1874                                           110 
           #             Demersal marine fishes Pelagic marine fish (excl. anchovy & sardine) 
           #                                117                                         13191 
```

```{r Plot bait production}
# Summary table by region and year:
Bait_summary <- GOC_BaitLandings %>%
  group_by(OHI_rgn_number, year, Major_Functional_Group) %>%
  summarize(annual_prod_kg = sum(weight_kg_corrected))
# 108 obs., 4 vars.

# Plot over time:
Bait_summary %>%
  filter(year >= 2018 & year < 2025) %>%
  
  ggplot(aes(x = year, y = annual_prod_kg, color= Major_Functional_Group)) +
  geom_line() +
  
  facet_wrap(~ OHI_rgn_number, scales = "free_y") +
  
  labs(x = "Year", y = "Production (kg)",
       title = "Bait Production over Time [DRAFT]") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

# PEARLS, NACRE, and SEASHELLS ------------------------------
```{r Take 1 Shells}
Shells_keywords <- c(
# Shells:
  "CONCHA DE", "CONCHA")

GOC_ShellProduction <- GOC_MarineLandings %>%
  filter(if_any(keywords_1:keywords_5, ~ . %in% Shells_keywords))

table(GOC_ShellProduction$estimated_genus_name)

#    Amiantis  Argopecten     Dosinia    Haliotis    Hexaplex     Lobatus Megapitaria         NEI 
#           1          33         993        3008         252          14          45         600 
# Triplofusus 
#           5 

# Investigating whether or not the shell is separate from the animal:
test<- GOC_MarineLandings %>% 
  filter(eu_rnpa== "0301000089" & year== 2000)
# kinda seems like catch for shell is logged separately than catch for flesh, which is good. Means I should filter out catches of shell from final fisheries estimates.

Pearls_keywords <- c(
  # Pearls
  "PERLAS", "MEDIAS PERLAS", "CONCHA NACAR")

GOC_PearlProduction <- GOC_MarineLandings %>%
  filter(if_any(keywords_1:keywords_5, ~ . %in% Pearls_keywords))
```

```{r Take 2 Shells and pearls}
# Apr 30 2026: 

Shells_keywords <- c(
# Shells:
  "CONCHA DE", "CONCHA")

Pearls_keywords <- c(
  # Pearls
  "PERLAS", "MEDIAS PERLAS", "CONCHA NACAR")

GOC_Possible.NPLandings <- GOC_Possible.NPLandings %>%
  mutate(
    ShellProduction = case_when(
      if_any(keywords_1:keywords_5, ~ .x %in% Shells_keywords) ~ "YES",
      TRUE ~ "NO"
      ),
    PearlNacreProduction = case_when(
      if_any(keywords_1:keywords_5, ~ .x %in% Pearls_keywords) ~ "YES",
      TRUE ~ "NO"
    )
  )
# 107256 obs., 72 vars.

GOC_ShellProduction<- GOC_Possible.NPLandings %>%
  filter(ShellProduction %in% c("YES"))
# 5859 obs., 70 vars.
table(GOC_ShellProduction$Major_Functional_Group)

GOC_PearlProduction<- GOC_Possible.NPLandings %>%
  filter(PearlNacreProduction %in% c("YES"))
# 18 obs., 70 vars.
# Note, registered pearl production is too small to be meaningful...
table(GOC_PearlProduction$Major_Functional_Group)
```

```{r Plot shell production}
# Summary table by region and year:
ShellProduction_summary <- GOC_ShellProduction %>%
  group_by(OHI_rgn_number, year, Major_Functional_Group) %>%
  summarize(annual_prod_kg = sum(weight_kg_corrected))
# 142 obs., 4 vars.

# Plot over time:
ShellProduction_summary %>%
  filter(year >= 2018 & year < 2025) %>%
  
  ggplot(aes(x = year, y = annual_prod_kg, color= Major_Functional_Group)) +
  geom_line() +
  
  facet_wrap(~ OHI_rgn_number, scales = "free_y") +
  
  labs(x = "Year", y = "Production (kg)",
       title = "Shell Production over Time [DRAFT]") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```


# SHARK FINS ------------------------------
```{r Shark fins}
Sharkfin_keywords <- c(
# Fins
  "ALETA", "ALETA DE", "ALETA ECA DE", "ALETA FCA DE")

GOC_SharkFinProduction <- GOC_MarineLandings %>%
  filter(if_any(keywords_1:keywords_5, ~ . %in% Sharkfin_keywords))
# No shark fin?

GOC_SharkFinProduction <- GOC_MarineLandings %>%
  filter(Major_Functional_Group == "Sharks")

table(GOC_SharkFinProduction$keywords_2)

# Shark fin, "ALETA" didn't get pulled out correctly. Pull out ALETA in CONAPESCA_CleaningWorkflow.
```


# FISH RESIDUES: fish heads, stomachs, bladder, etc. ------------------------------
Use FAO percentages to calculate whole body mass versus part mass.
Mel can look into whether or not we have edible portion of fish data from FAO. Simple multiplier, 1-edible proportion.

# FISH OIL, FISH MEAL ------------------------------
Assume that ~70% forage fish is used for fish oil fish meal

```{r Pull out forage fish}

```

```{r Apply 70% conversion}

```

# ALGAE ------------------------------
Sargassum