################################################################################ 

For questions contact: Erica Ferrer, Ph.D. at emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################ 

# PRE-WORK ------------------------------

```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/NP/v2025/")
```

```{r Outdated natural products key}
NPKey_pending<- read.csv('/home/ferrer/ohi_goc_prep/NP/v2025/NP_Keys/NP_MasterList_UNDER CONSTRUCTION_Nov132025.csv')
```

```{r Load GOC_MarineLandings if needed}
GOC_MarineLandings0<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Marine.Landings_2000.present_May042026.rds"))
# 3303634 obs., 61 vars.

# This file was generated using the CONAPESCA_CleaningWorkflow_May2026.rmd file.
```

```{r Trim GOC_MarineLandings so it's easier to work with}
colnames(GOC_MarineLandings0)

GOC_MarineLandings<- GOC_MarineLandings0 %>% 
  select(-c(keywords_5, IHH_functional_major_groups_sm,
            Assignment_for_Major_Functional_Group,
            NOTaquacultureFLAG_specific.groups, aquacultureFLAG_site.details,
            aquacultureFLAG_keywords, aquacultureFLAG_specific.taxa, 
            state_clean, office_clean, landing_site_na, state_patch, office_name_patch
            ))
# 3303634 obs., 49 vars.
```

# FLAG RESIDUES: fish heads, stomachs, bladder, shells, etc. ------------------------------
Use FAO percentages to calculate whole body mass versus part mass.
Mel can look into whether or not we have edible portion of fish data from FAO. Simple multiplier, 1-edible proportion.

```{r Remove entries where no weight whatsoever is logged: GOC_MarineLandings_corrected}
colnames(GOC_MarineLandings)

GOC_MarineLandings_corrected <- GOC_MarineLandings %>%
  filter(!(is.na(weight_kg_corrected) | weight_kg_corrected == 0)) %>%
  select(-recorded_taxa_name, -recorded_taxa_name_clean,
         -live_weight, -landed_weight, 
         -keywords_1, -keywords_2, -keywords_3, -keywords_4, 
         everything(),
         recorded_taxa_name, recorded_taxa_name_clean,
         keywords_1, keywords_2, keywords_3, keywords_4,
         live_weight, landed_weight) %>%
  select(-c(weight_kg_corrected))

colnames(GOC_MarineLandings_corrected)
dim(GOC_MarineLandings_corrected)
# 3303473 obs., 48 vars.
```

```{r First check if there is a discrepancy between wet weight and landed weight}
colnames(GOC_MarineLandings_corrected)

GOC_MarineLandings_corrected <- GOC_MarineLandings_corrected %>%
  mutate(live.landed_discrepancy = case_when(
    !is.na(live_weight) &
    !is.na(landed_weight) &
    live_weight > landed_weight ~
      "live weight is greater than landed weight",

    !is.na(live_weight) &
    !is.na(landed_weight) &
    live_weight < landed_weight ~
      "landed weight is greater than live weight; likely error?",

    !is.na(live_weight) &
    !is.na(landed_weight) &
    live_weight == landed_weight ~
      "live weight equals landed weight",

    TRUE ~ NA_character_
  ))

table(GOC_MarineLandings_corrected$live.landed_discrepancy)
#                       landed weight is greater than live weight; likely error? 
#                                                                          41562 
#                                               live weight equals landed weight 
#                                                                        1779765 
#                                     live weight is greater than landed weight 
#                                                                        1482146
```


```{r Make an alive or dead column}
Dead_keywords<- c(
  
  # Frozen, filleted, beheaded, dried, fresh, etc.
  "S.C.", "S/C", "S", # I am pretty sure this means "without shell" - May 04 2026
  "ECO",
  "CONG", "CONGELADO",
  "COCINERO", "COCIDO", "COCIDA", "COCIDAS", "ASADO",
  "FILETE", "FILETE DE", "ECA", "FILETE DE ECO", "DESHUESADA", "CARNE", "CARNE DE",
  "DESCABEZADO", "DES", "DSC", "DESC", "DESV", "DESVICERADO",
  
  # Dried:
  "SECO DE", "SECA", "SECA DE", "SECO", "SECO DE",
  
  # Shells and pearls:
  "CONCHA DE", "CONCHA", "CONCHA NACAR",
  "PERLAS", "MEDIAS PERLAS",
  
  # Fin:
  "ALETA", "ALETA ECA DE", "ALETA FCA DE",
  
  # Animal parts:
  "MANTO", "MANTO DE",
  "PULPA DE", "PULPA FRESCA", "PULPO DE", 
  "POLVO DE",
  "TENTACULOS DE", "TUBO DE", "TUBO",
  "CABEZA DE", "BUCHE", "BUCHE DE", "TRIPA DE", "COLAS DE", "MANOS DE", "MANOS",
  "ANCAS DE", "ANCAS", "ANCA",
  "PIEL DE ECA", "PIEL DE SECA", "PIEL DE", "PIEL"
)

Alive_keywords<- c("VIVO", "VIVA", "VIVAS", "VIVITA") # living / live

# Assign alive, dead, unknown based on keywords:
GOC_MarineLandings_corrected <- GOC_MarineLandings_corrected %>%
  mutate(
    Alive.Dead = case_when(
      
      # DEAD FIRST (safer precedence)
      if_any(keywords_1:keywords_4,
             ~ str_detect(.x, str_c(Dead_keywords, collapse = "|"))) ~ "dead",
      
      # ALIVE
      if_any(keywords_1:keywords_4,
             ~ str_detect(.x, str_c(Alive_keywords, collapse = "|"))) ~ "alive",
      
      # DEFAULT
      TRUE ~ "unknown, assumed dead"
    )
  )

dim(GOC_MarineLandings_corrected)
# 3303473 obs., 50 vars.
```

```{r Add flag columns}
#### FLAG 1 - Flag keywords that indicate animal was sold whole (2 levels: YES, NO): ####

SoldWhole_keywords <- c(
  "ENTERO", "ENTERA", "IND", "ENT"
)

Fillet_keywords <- c(

  # Processing state or product form (e.g., frozen, full fish, filleted, beheaded, dried, etc.):
  "FILETE", "FILETE DE", "ECA", "FILETE DE ECO", "CARNE", "CARNE DE",
  "DESCABEZADO", "DES", "DSC", "DESC", "DESV", "DESVICERADO", "DESHUESADA",
  "S.C.", "S/C", "S") # I am pretty sure this means "without shell" - May 04 2026

Residues_keywords <- c(
  # Fins:
  "ALETA DE", "ALETA",
    
  # Other animal parts:
  "MANTO", "MANTO DE",
  "PULPA DE", "PULPA FRESCA", "PULPO DE", 
  "POLVO DE",
  "TENTACULOS DE", "TUBO DE", "TUBO",
  "CABEZA DE", "BUCHE", "BUCHE DE", "TRIPA DE", "COLAS DE", "MANOS DE", "MANOS",
  "ANCAS DE", "ANCAS", "ANCA",
  "PIEL DE ECA", "PIEL DE SECA", "PIEL DE", "PIEL",
  
  # Shells:
  "CONCHA DE", "CONCHA")

GOC_MarineLandings_corrected <- GOC_MarineLandings_corrected %>%
  mutate(
    SoldWhole_FLAG = case_when(
      if_any(keywords_1:keywords_4, ~ .x %in% SoldWhole_keywords) ~ "YES",
      
      if_any(keywords_1:keywords_4, 
             ~ .x %in% Fillet_keywords | .x %in% Residues_keywords) ~ "NO",
      
      TRUE ~ "unknown"
    )
  )

#### FLAG 2 - Flag keywords that indicate animal was filleted (2 levels: YES, NO): ####

GOC_MarineLandings_corrected <- GOC_MarineLandings_corrected %>%
  mutate(
    Fillet_FLAG = case_when(
      if_any(keywords_1:keywords_4, ~ .x %in% Fillet_keywords) ~ "YES",
      
      if_any(keywords_1:keywords_4, 
             ~ .x %in% SoldWhole_keywords | .x %in% Residues_keywords) ~ "NO",
      
      TRUE ~ "unknown"
    )
  )

#### FLAG 3 - Flag keywords that indicate animal was sold as a residue (2 levels: YES, NO): ####

GOC_MarineLandings_corrected <- GOC_MarineLandings_corrected %>%
  mutate(
    Residue_FLAG = case_when(
      if_any(keywords_1:keywords_4, ~ .x %in% Residues_keywords) ~ "YES",
      
      if_any(keywords_1:keywords_4, 
             ~ .x %in% SoldWhole_keywords | .x %in% Fillet_keywords) ~ "NO",
      
      TRUE ~ "unknown"
    )
  )

dim(GOC_MarineLandings_corrected)
# 3303473 obs., 53 vars.
```

```{r SANITY CHECK: Check the flags worked as expected}
subset<- GOC_MarineLandings_corrected %>%
  filter(SoldWhole_FLAG == "NO" &
           Fillet_FLAG == "NO" &
           Residue_FLAG == "NO")
# should be 0 length. It is, good!

subset<- GOC_MarineLandings_corrected %>%
  filter(SoldWhole_FLAG == "unknown" &
           Fillet_FLAG == "unknown" &
           Residue_FLAG == "unknown")
# 82360 records that aren't assigned at all (2%)... I can live with this...

subset<- GOC_MarineLandings_corrected %>%
  filter(SoldWhole_FLAG == "YES" &
           Fillet_FLAG == "YES" &
           Residue_FLAG == "YES")
# should be 0 length. It is, good!

subset<- GOC_MarineLandings_corrected %>%
  filter(SoldWhole_FLAG == "YES" &
           Fillet_FLAG == "YES" &
           Residue_FLAG == "NO")

table(subset$recorded_taxa_name_clean)
# should be 0 length. It is, good!

subset<- GOC_MarineLandings_corrected %>%
  filter(SoldWhole_FLAG == "YES" &
           Fillet_FLAG == "NO" &
           Residue_FLAG == "YES")

table(subset$recorded_taxa_name_clean)
# should be 0 length. It's not, there are 94 obs.
# NOTE TO SELF: Here are some special edge cases, akin to blue fin tuna...

subset<- GOC_MarineLandings_corrected %>%
  filter(SoldWhole_FLAG == "NO" &
           Fillet_FLAG == "YES" &
           Residue_FLAG == "YES")

table(subset$recorded_taxa_name_clean)
# should be 0 length. It's not, there are 198 obs.
# NOTE TO SELF: Here are some special edge cases, akin to blue fin tuna...

# This flag system mostly works...

# Next let's look into entries where landed weight is greater than live weight; likely error?
table(GOC_MarineLandings_corrected$live.landed_discrepancy)
#                       landed weight is greater than live weight; likely error? 
#                                                           41562 
#                               live weight equals landed weight 
#                                                         1779765 
# live weight is greater than landed weight; suggests catch prepared before sold 
#                                                         1482146

subset<- GOC_MarineLandings_corrected %>%
  filter(live.landed_discrepancy== "landed weight is greater than live weight; likely error?")

table(subset$Alive.Dead) # Nearly all (41196, 99%) are dead.
table(subset$Residue_FLAG) # Nearly all records (41177, 99%) are flagged as most likely residues...

subset2 <- subset %>%
  filter(Residue_FLAG == "unknown")

table(subset2$recorded_taxa_name_clean)
# These entries make sense and are few.
# NOTE TO SELF: Going to need a patch for "CALAMAR GIGANTE CABEZA" (May 04 2026), fine to leave for now since few entries

subset3<- GOC_MarineLandings_corrected %>%
  filter(Residue_FLAG == "YES")

table(subset3$live.landed_discrepancy)
```

What this tells me is that for most entries where landed weight is greater than live weight; likely error?, it's an animal part that is being sold.
BUT, sometimes an animal part sold and live weight is still greater than landed weight...

# Make a recorded_weightOFanimal_kg and recorded_weightOFfillet_kg columns:
If SoldWhole_FLAG == "YES" & live.landed_discrepancy== "landed weight is greater than live weight; likely error?" then recorded_weightOFanimal_kg = landed_weight & recorded_weightOFfillet_kg = live_weight
If SoldWhole_FLAG == "YES" & live.landed_discrepancy== "live weight equals landed weight" then recorded_weightOFanimal_kg = live_weight & recorded_weightOFfillet_kg = NA
If SoldWhole_FLAG == "YES" & live.landed_discrepancy== "live weight is greater than landed weight; suggests catch prepared before sold" then recorded_weightOFanimal_kg = live_weight & recorded_weightOFfillet_kg = landed_weight

If (Fillet_FLAG == "YES" | Residue_FLAG == "YES")  & live.landed_discrepancy== "landed weight is greater than live weight; likely error?" then recorded_weightOFanimal_kg = landed_weight & recorded_weightOFfillet_kg = live_weight
If (Fillet_FLAG == "YES" | Residue_FLAG == "YES") & live.landed_discrepancy== "live weight equals landed weight" then recorded_weightOFanimal_kg = NA & recorded_weightOFfillet_kg = landed_weight
If (Fillet_FLAG == "YES" | Residue_FLAG == "YES") & live.landed_discrepancy== "live weight is greater than landed weight; suggests catch prepared before sold" then recorded_weightOFanimal_kg = live_weight & recorded_weightOFfillet_kg = landed_weight

```{r Make a recorded_weightOFanimal_kg recorded_weightOFfillet_kg and recorded_weightOFresidue_kg column based on the flags}

GOC_MarineLandings_corrected <- GOC_MarineLandings_corrected %>%
  mutate(
    recorded_weightOFanimal_kg = case_when(
      
      (Residue_FLAG == "YES") & (live_weight == 0) ~ NA_real_,
      
      SoldWhole_FLAG == "YES" &
        live.landed_discrepancy == "landed weight is greater than live weight; likely error?" ~ landed_weight,
      
      SoldWhole_FLAG == "YES" &
        live.landed_discrepancy == "live weight equals landed weight" ~ live_weight,
      
      SoldWhole_FLAG == "YES" &
        live.landed_discrepancy == "live weight is greater than landed weight" ~ live_weight,
      
      (Fillet_FLAG == "YES" | Residue_FLAG == "YES") &
        live.landed_discrepancy == "landed weight is greater than live weight; likely error?" ~ landed_weight,
      
      (Fillet_FLAG == "YES" | Residue_FLAG == "YES") &
        live.landed_discrepancy == "live weight equals landed weight" ~ NA_real_,
      
      (Fillet_FLAG == "YES" | Residue_FLAG == "YES") &
        live.landed_discrepancy == "live weight is greater than landed weight" ~ live_weight,
      
      TRUE ~ NA_real_
    ),
    
    recorded_weightOFfillet_kg = case_when(
      
      SoldWhole_FLAG == "YES" ~ NA_real_,
      Residue_FLAG == "YES" ~ NA_real_,
      
      (Fillet_FLAG == "YES") &
        live.landed_discrepancy == "landed weight is greater than live weight; likely error?" ~ live_weight,
      
      (Fillet_FLAG == "YES") &
        live.landed_discrepancy == "live weight equals landed weight" ~ landed_weight,
      
      (Fillet_FLAG == "YES") &
        live.landed_discrepancy == "live weight is greater than landed weight" ~ landed_weight,
      
      TRUE ~ NA_real_
    ),
    
    recorded_weightOFresidue_kg = case_when(
      
      SoldWhole_FLAG == "YES" ~ NA_real_,
      Fillet_FLAG == "YES" ~ NA_real_,
      
      (Residue_FLAG == "YES") & (live_weight == 0) ~ landed_weight,
      (Residue_FLAG == "YES") & (landed_weight == 0) ~ live_weight,
      
      (Residue_FLAG == "YES") & live.landed_discrepancy == "landed weight is greater than live weight; likely error?" ~ live_weight,
      
      (Residue_FLAG == "YES") &
        live.landed_discrepancy == "live weight equals landed weight" ~ landed_weight,
      
      (Residue_FLAG == "YES") &
        live.landed_discrepancy == "live weight is greater than landed weight" ~ landed_weight,
      
      (Residue_FLAG == "YES") &
        is.na(live.landed_discrepancy) ~ landed_weight,
      
      TRUE ~ NA_real_
    )
  )

# Quick look:
subset<- GOC_MarineLandings_corrected %>%
  slice_sample(n = 200)
```

```{r Make a discard column}
GOC_MarineLandings_sum<- GOC_MarineLandings_corrected %>%
  mutate(est_weightNOTused_kg = recorded_weightOFanimal_kg - recorded_weightOFfillet_kg)

dim(GOC_MarineLandings_sum)
# 3303473 obs, 57 vars.

# Quick look:
subset<- GOC_MarineLandings_sum %>%
  slice_sample(n = 200)
```

```{r SANITY CHECKS}
subset<- GOC_MarineLandings_sum %>%
  filter(Residue_FLAG == "YES")

subset<- GOC_MarineLandings_sum %>%
  filter(Residue_FLAG == "NO")

subset<- GOC_MarineLandings_sum %>%
  filter(Fillet_FLAG == "YES") %>%
  filter(is.na(recorded_weightOFanimal_kg))

test <- GOC_MarineLandings_sum %>%
  filter(recorded_taxa_group== "CAMARON") %>%
  filter(year == 2024) %>%
  filter(SoldWhole_FLAG == "YES")

# NOTE TO SELF: Shrimp that were sold whole but have a lower landed weight than live weight are cooked shrimp...
```

```{r Make type column}
GOC_MarineLandings_sum <- GOC_MarineLandings_sum %>%
  mutate(
    record_type = case_when(
      
      SoldWhole_FLAG == "YES" ~ "sold whole",
      Fillet_FLAG == "YES" ~ "fillet",
      Residue_FLAG == "YES" ~ "other residue"
    )
  )
```

```{r Summary table for shrimp}
# Quick look:
subset<- GOC_MarineLandings_sum %>%
  filter(recorded_taxa_group== "CAMARON")
  # filter(year== 2010) %>%
  # slice_sample(n = 500)

table(subset$keywords_1)
# Very few entries of CABEZA DE

colnames(GOC_MarineLandings_sum)

# Of those that were not sold whole, how much product was wasted?

ShrimpResidues_summary <- GOC_MarineLandings_sum %>%
#  filter(SoldWhole_FLAG== "NO") %>%
  filter(recorded_taxa_group== "CAMARON") %>%
  filter(!is.na(record_type)) %>%
  group_by(year, record_type) %>%
  summarise(
    recorded_weightOFanimal_kg = sum(recorded_weightOFanimal_kg, na.rm = TRUE),
    recorded_weightOFfillet_kg = sum(recorded_weightOFfillet_kg, na.rm = TRUE),
    est_weightNOTused_kg = sum(est_weightNOTused_kg, na.rm = TRUE),
    ratio_fillet.wholeanimal = recorded_weightOFfillet_kg/recorded_weightOFanimal_kg,
    ratio_NOTused.wholeanimal = est_weightNOTused_kg/recorded_weightOFanimal_kg,
    recorded_weightOFresidue_kg = sum(recorded_weightOFresidue_kg, na.rm = TRUE),
    .groups = "drop"
  )
```


```{r Let's take a deep dive into CALAMAR}
SquidResidues_summary <- GOC_MarineLandings_sum %>%
#  filter(SoldWhole_FLAG== "NO") %>%
  filter(recorded_taxa_group== "CALAMAR") %>%
  filter(!is.na(record_type)) %>%
  group_by(year, record_type) %>%
  summarise(
    recorded_weightOFanimal_kg = sum(recorded_weightOFanimal_kg, na.rm = TRUE),
    recorded_weightOFfillet_kg = sum(recorded_weightOFfillet_kg, na.rm = TRUE),
    est_weightNOTused_kg = sum(est_weightNOTused_kg, na.rm = TRUE),
    ratio_fillet.wholeanimal = recorded_weightOFfillet_kg/recorded_weightOFanimal_kg,
    ratio_NOTused.wholeanimal = est_weightNOTused_kg/recorded_weightOFanimal_kg,
    recorded_weightOFresidue_kg = sum(recorded_weightOFresidue_kg, na.rm = TRUE),
    .groups = "drop"
  )

SquidResidues_summary<- SquidResidues_summary %>%
  mutate(weight_product= case_when(
      
      record_type == "sold whole" ~ recorded_weightOFanimal_kg,
      record_type == "fillet" ~ recorded_weightOFfillet_kg,
      record_type == "other residue" ~ recorded_weightOFresidue_kg
    )
    
  )

# Plot the data: this shows us catches for items that were butchered versus not... It does not show us residues versus non-residues

scale_factor <- max(SquidResidues_summary$recorded_weightOFanimal_kg, na.rm = TRUE) /
                max(SquidResidues_summary$recorded_weightOFfillet_kg, na.rm = TRUE)

#NOTE TO SELF: It's just not totally clear that mass of residues + mass of fillets should actually adds up to mass of total whole animal catch...


SquidResidues_summary %>%
  ggplot(aes(x = year)) +
  
  geom_line(aes(y = weight_product, color = record_type), linewidth = 1) +
  
  labs(
    title = "Squid: Landings by Type",
    x = "Year",
    color = "Type"
  ) +
  theme_minimal()

```

```{r Let's take a deep dive into ABULON}
Abalone_Landings<- GOC_MarineLandings_corrected %>%
  filter(recorded_taxa_group=="ABULON")

colnames(Abalone_Landings)

Abalone_Landings_summary <- Abalone_Landings %>%
  group_by(year, Residue_FLAG) %>%
  summarise(
    annual_landed_weight_kg = sum(landed_weight, na.rm = TRUE),
    annual_total_weight_kg  = sum(weight_kg_corrected, na.rm = TRUE),
    .groups = "drop"
  )

# Plot the data:
ggplot(Abalone_Landings_summary, 
       aes(x = year, y = annual_total_weight_kg, color = Residue_FLAG, group = Residue_FLAG)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Abalone",
    x = "Year",
    y = "Annual total weight (kg)",
    color = "Residues likely"
  ) +
  theme_minimal()
```

```{r Let's take a deep dive into CARACOL}
Caracol_Landings<- GOC_MarineLandings_corrected %>%
  filter(recorded_taxa_group=="CARACOL")

Caracol_Landings_summary <- Caracol_Landings %>%
  group_by(year, Residue_FLAG) %>%
  summarise(
    annual_landed_weight_kg = sum(landed_weight, na.rm = TRUE),
    annual_total_weight_kg  = sum(weight_kg_corrected, na.rm = TRUE),
    .groups = "drop"
  )

# Plot the data:
ggplot(Caracol_Landings_summary, 
       aes(x = year, y = annual_landed_weight_kg, color = Residue_FLAG, group = Residue_FLAG)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title= "Caracol",
    x = "Year",
    y = "Annual landed weight (kg)",
    color = "Residues likely"
  ) +
  theme_minimal()
```


-------- WORKING ON RESIDUES, OTHER NPs PENDING - May 04 2026

```{r Filter out residues from GOC_MarineLandings_corrected}
GOC_Possible.Residues<- GOC_MarineLandings_corrected %>%
  filter(Residue_FLAG %in% c("YES"))

dim(GOC_Possible.Residues)
# 69548 obs., 56 vars.

table(GOC_Possible.Residues$Major_Functional_Group)
# Lots of cephalopods, demersal fish, crustaceans, and sharks. Makes sense!
# NOTE TO SELF: Don't forget Aleta though!! This DOES NOT include Aleta ... yet - Apr 30 2026
# NOTE TO SELF: Fixed the aleta issue, May 4 2026.
```

```{r Plot residues}
# Summary table by region and year:
Residues_summary <- GOC_MarineLandings_corrected %>%
  filter(Residue_FLAG %in% c("YES")) %>%
  group_by(state, year, Major_Functional_Group) %>%
  summarize(annual_prod_kg = sum(weight_kg_corrected))
# 342 obs., 4 vars.

# Plot over time:
Residues_summary %>%
  filter(year >= 2018 & year < 2025) %>%
  
  ggplot(aes(x = year, y = annual_prod_kg, color= Major_Functional_Group)) +
  geom_line() +
  
  facet_wrap(~ state, scales = "free_y") +
  
  labs(x = "Year", y = "Production (kg)",
       title = "Residues Production over Time [DRAFT]") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```


# Flag other things that are very likely to be natural products based on key terms...
```{r Estimate likelihood of ornamentals based on name of landing_site and eu_name}
#### FLAG 4 - Flag things that are very likely to be natural products based on landing_site, keywords, key groups, and key taxa (2 levels: YES, NO, MAYBE): ####

# Clues from keywords:
table(GOC_MarineLandings_corrected$keywords_1)
unique(GOC_MarineLandings_corrected$keywords_1)

# Keywords often linked to natural products:
Keyword_strong.clues<- c("VIVO", "VIVA", "VIVAS", "VIVITA", # living / live
                         
                         # Used as bait:
                         "CARNADA", "P/CARNADA",

                         # Shells and pearls:
                         "CONCHA DE", "CONCHA", "CONCHA NACAR",
                         "PERLAS", "MEDIAS PERLAS",
                         # Fin:
                         "ALETA", "ALETA DE",
                         
                         # Animal parts:
                         "MANTO", "MANTO DE",
                         "PULPA DE", "PULPA FRESCA", "PULPO DE", 
                         "POLVO DE",
                         "TENTACULOS DE", "TUBO DE", "TUBO",
                         "CABEZA DE", "BUCHE", "BUCHE DE", "TRIPA DE", "COLAS DE", "MANOS DE", "MANOS",
                         "ANCAS DE", "ANCAS", "ANCA",
                         "PIEL DE ECA", "PIEL DE SECA", "PIEL DE", "PIEL",
                         
                         # Dried things:
                         "SECO DE", "SECA", "SECA DE", "SECO", "SECO DE"
)

# Keywords sometimes linked to natural products:
Keyword_weak.clues<- c("DE CULTIVO", "DE CULTIVO VIVO",
                       "CULTIVADA", "CULTIVADO", "CULTIVADAS", "CULTIVADOS", # A few new adds as of May 04
                       "ABASTO RANCHO", "RANCHO",
                       # Eggs:
                       "HUEVA DE")

# Keywords typically linked to FOOD, not natural products
Keyword_opposite.clues<- c(
  # Processing state or product form (e.g., frozen, full fish, filleted, beheaded, dried, etc.):
  "S.C.", "S/C", "S",
  "FRESCO", "FRESCA", "FCA", "FCO", "FC",
  "ECO",
  "CONG", "CONGELADO",
  "COCINERO", "COCIDO", "COCIDA", "COCIDAS", "ASADO",
  "ENTERO", "ENTERA", "IND", "ENT",
  "FILETE", "FILETE DE", "ECA", "FILETE DE ECO", "DESHUESADA", "CARNE", "CARNE DE",
  "DESCABEZADO", "DES", "DSC", "DESC", "DESV", "DESVICERADO"
)

# Clues from landing_site and eu_name:
LandingSite_clues<- c("CULTIVADORES", "ESTANQUE", "VIVEROS", "ACUACULTURA") 

# Clues from recorded_taxa_group:
table(GOC_MarineLandings_corrected$recorded_taxa_group)
TaxaGroup_strong.clues<- c("PECES DE ORNATO", "ORNATO") 

# OTRAS that could be ornamental fishes:
otras<- GOC_MarineLandings_corrected %>%
  select(recorded_taxa_group, recorded_taxa_name_clean) %>%
  filter(recorded_taxa_group == "OTRAS") %>%
  unique()
  
TaxaName_strong.clues<- c("PAYASO", "PECES ORNATO", "ABANDERADO", "ALA ANGEL", "CABALLITO MAR", "PEZ PIPA", 
                          "ESTRELLA MAR", "ESTRELLA LIMPIADORA", "GALLETA MAR", 
                          "CARACOL", "HACHA", "CALLO", "CALLO HACHA", "CALLO MEDIA LUNA",
                          "MADRE PERLA",
                          "CORAL", "CORAL BLANCO",
                          "CANGREJO ERMITAÑO",
                          "TOTOABA"
                          )

TaxaName_weak.clues<- c("PEZ CORNETA",
                        "PEZ GLOBO", "BOTETE",
                        "MURENA", "MORENA",
                        "SOLDADITO",
                        "MEDUSA BOLA CAÑÓN"
                        )


colnames(GOC_MarineLandings_corrected)

# Build regex patterns dynamically
Keyword_strong.pattern <- paste0("\\b(", paste(Keyword_strong.clues, collapse = "|"), ")\\b")
Keyword_weak.pattern <- paste0("\\b(", paste(Keyword_weak.clues, collapse = "|"), ")\\b")
Keyword_opposite.pattern <- paste0("\\b(", paste(Keyword_opposite.clues, collapse = "|"), ")\\b")
LandingSite_pattern <- paste0("\\b(", paste(LandingSite_clues, collapse = "|"), ")\\b")
TaxaGroup_pattern <- paste0("\\b(", paste(TaxaGroup_strong.clues, collapse = "|"), ")\\b")
TaxaName_strong.pattern <- paste0("\\b(", paste(TaxaName_strong.clues, collapse = "|"), ")\\b")
TaxaName_weak.pattern <- paste0("\\b(", paste(TaxaName_weak.clues, collapse = "|"), ")\\b")


# Implement regex key:
GOC_Possible.NPLandings <- GOC_MarineLandings_corrected %>%
  
  mutate(
    NP.Keyword_FLAG = case_when(
      if_any(keywords_1:keywords_4, ~ str_detect(.x, regex(Keyword_strong.pattern, ignore_case = TRUE))) ~ "YES",
      if_any(keywords_1:keywords_4, ~ str_detect(.x, regex(Keyword_weak.pattern, ignore_case = TRUE))) ~ "MAYBE",
      if_any(keywords_1:keywords_4, ~ str_detect(.x, regex(Keyword_opposite.pattern, ignore_case = TRUE))) ~ "NO",
      TRUE ~ "NO"
    ),
    NP.Site_FLAG = case_when(
      str_detect(landing_site, regex(LandingSite_pattern, ignore_case = TRUE)) ~ "YES",
      TRUE ~ "NO"
    ),
    NP.Group_FLAG = case_when(
      str_detect(recorded_taxa_group, regex(TaxaGroup_pattern, ignore_case = TRUE)) ~ "YES",
      TRUE ~ "NO"
    ), 
    NP.TaxaName_FLAG = case_when(
      str_detect(recorded_taxa_name_clean, regex(TaxaName_strong.pattern, ignore_case = TRUE)) ~ "YES",
      str_detect(recorded_taxa_name_clean, regex(TaxaName_weak.pattern, ignore_case = TRUE)) ~ "MAYBE",
      TRUE ~ "NO"
    ),
    # Previously made ornamental flag:
      NP.PreviousOrnamental_FLAG = case_when(
        ornamentalFLAG_1 == "YES" ~ "YES",
        ornamentalFLAG_1 == "POSSIBLY" ~ "MAYBE",
        TRUE ~ "NO"
        )
  )

# Now exclude any entries that don't have at least one YES flag.
GOC_Possible.NPLandings <- GOC_Possible.NPLandings %>%
  filter(
    if_any(
      starts_with("NP"),
      ~ . == "YES"
    )
  )

dim(GOC_Possible.NPLandings)
# 142423 obs. of 53 vars.
```



# AQUARIUM TRADE SPECIES: LIVE ORNAMENTAL FISH AND INVERTS ------------------------------
Run list of species through AI and flag any species that are almost ALWAYS ornamentals.

```{r Check ornamentals flag}
table(GOC_Possible.NPLandings$ornamentalFLAG_1)
      # NO POSSIBLY      YES 
      #  1    19726     1602

table(GOC_Possible.NPLandings$NP.PreviousOrnamental_FLAG)
# MAYBE    NO   YES 
# 19726 85928  160
# Looks good!
```

```{r Pull out possible ornamentals}
colnames(GOC_Possible.NPLandings)
table(GOC_Possible.NPLandings$Alive.Dead)

GOC_Possible.OrnamentalLandings<- GOC_Possible.NPLandings %>%
  filter(
    (NP.PreviousOrnamental_FLAG %in% c("YES", "MAYBE") |
       NP.Group_FLAG %in% c("YES", "MAYBE") |
       NP.TaxaName_FLAG %in% c("YES", "MAYBE")) &
      Alive.Dead %in% c("alive", "unknown")
  )
           
# 20107 obs. of 67 vars.

table(GOC_Possible.OrnamentalLandings$Alive.Dead)
  # alive unknown 
  #   430   19677 
```

```{r Look into flagged ornamental taxa}
# Taking the information I've gleaned here to refine my sector estimates above.
table(GOC_Possible.OrnamentalLandings$estimated_genus_name)
table(GOC_Possible.OrnamentalLandings$Major_Functional_Group)
table(GOC_Possible.OrnamentalLandings$recorded_taxa_name_clean)

# Investigate Major_Taxonomic_Group unknown:
subset <- GOC_Possible.OrnamentalLandings %>% 
  filter(Major_Functional_Group %in% c("unknown"))
# 1168 obs., 62 vars.

table(subset$recorded_taxa_group)
table(subset$common_name_clean)
# Makes sense; all ornamental species.
```

```{r Pull out the taxa most likely to be sold alive for the aquarium trade}
# Very likely: 
# CABALLITO MAR → seahorse (very commonly traded live in marine aquariums)
# PAYASO → clownfish (Amphiprion spp.), one of the most common aquarium fish globally
# PECES ORNATO → literally “ornamental fish,” a general aquarium trade category

# Somewhat likely:
# CARACOL CHINO / CHINO NEGRO / CHINO ROSA / CHINO ROSADO → likely nerite-type snails (widely sold for algae control)
# CARACOL TURBO FLUCTUOSO → turbo snails (classic reef aquarium cleaners)
# CARACOL OLIVO → olive snails, sometimes kept in marine tanks
# CARACOL MARGARITA → margarita snails (cold/temperate marine aquariums)
# CARACOL UÑA / CARACOL UÑA → uncertain ID, but likely small ornamental snail types
# CARACOL CHILE / CARACOL BURRO / CARACOL PANOCHA → less standard names, but still plausibly marine snails traded as live clean-up organisms depending on region
# ESTRELLA LIMPIADORA → “cleaner star” / sand-sifting or detritus-feeding sea stars (commonly sold in marine aquariums)

# Not likely:
# ESTRELLA MAR → sea stars (some species are aquarium-kept but not common)
# GALLETA MAR → sand dollar / sea biscuit (sometimes sold live but difficult to maintain)
# HACHA CHINA / HACHA LARGA / CALLO HACHA → scallops or bivalves; occasionally live in holding systems but not typical aquarium pets
# MADRE PERLA → pearl oyster; generally not an aquarium species

AquariumTrade_LikelyTaxa<- c("CABALLITO MAR", "PAYASO", "PECES ORNATO")
AquariumTrade_PossibleTaxa<- c("CARACOL CHINO", "CARACOL CHINO NEGRO", "CARACOL CHINO ROSA", "CARACOL CHINO ROSADO",
                               "CARACOL TURBO FLUCTUOSO", "CARACOL OLIVO", "CARACOL MARGARITA",
                               "CARACOL UÑA", "CARACOL CHILE", "CARACOL BURRO", "CARACOL PANOCHA",
                               "ESTRELLA LIMPIADORA")
AquariumTrade_UnlikelyTaxa<- c("ESTRELLA MAR", "GALLETA MAR",
                               "HACHA CHINA", "HACHA LARGA", "CALLO HACHA",
                               "MADRE PERLA")

# Assign ornamental likely or unlikely:
GOC_Possible.OrnamentalLandings <- GOC_Possible.OrnamentalLandings %>%
  mutate(
    AquariumTradeLikely = case_when(
      recorded_taxa_name_clean %in% AquariumTrade_LikelyTaxa ~ "YES",
      recorded_taxa_name_clean %in% AquariumTrade_PossibleTaxa ~ "MAYBE",
      recorded_taxa_name_clean %in% AquariumTrade_UnlikelyTaxa ~ "NO",
      TRUE ~ "unknown"
    )
  )
# 20107 obs., 69 vars.

GOC_OrnamentalLandings<- GOC_Possible.OrnamentalLandings %>%
  filter(AquariumTradeLikely %in% c("YES", "MAYBE"))
# 15292 obs., 69 vars.
```

```{r Plot ornamentals}
# Summary table by region and year:
Ornamentals_summary <- GOC_OrnamentalLandings %>%
  group_by(OHI_rgn_number, year, recorded_taxa_name_clean) %>%
  summarize(annual_prod_kg = sum(weight_kg_corrected))
# 462 obs., 4 vars.

# Plot over time:
Ornamentals_summary %>%
  filter(recorded_taxa_name_clean %in% c("PECES ORNATO", "PAYASO")) %>%
  filter(year >= 2018 & year < 2025) %>%
  
  ggplot(aes(x = year, y = annual_prod_kg, color= recorded_taxa_name_clean)) +
  geom_line() +
  
  facet_wrap(~ OHI_rgn_number, scales = "free_y") +
  
  labs(x = "Year", y = "Production (kg)",
       title = "Aquarium Trade / Production of Live Ornamental Taxa [DRAFT]") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

# BAIT ------------------------------
```{r Bait}
Bait_keywords <- c(
# Used as bait:
  "CARNADA", "P/CARNADA"
)

GOC_Possible.NPLandings <- GOC_Possible.NPLandings %>%
  mutate(
    BaitLikely = case_when(
      if_any(keywords_1:keywords_4, ~ .x %in% Bait_keywords) ~ "YES",
      TRUE ~ "NO"
    )
  )

GOC_BaitLandings<- GOC_Possible.NPLandings %>%
  filter(BaitLikely %in% c("YES"))
# 15292 obs., 70 vars.

table(GOC_BaitLandings$Major_Functional_Group)
           # Anchovy & Sardine (small pelagics)                                   Crustaceans 
           #                               1874                                           110 
           #             Demersal marine fishes Pelagic marine fish (excl. anchovy & sardine) 
           #                                117                                         13191 
```

```{r Plot bait production}
# Summary table by region and year:
Bait_summary <- GOC_BaitLandings %>%
  group_by(OHI_rgn_number, year, Major_Functional_Group) %>%
  summarize(annual_prod_kg = sum(weight_kg_corrected))
# 108 obs., 4 vars.

# Plot over time:
Bait_summary %>%
  filter(year >= 2018 & year < 2025) %>%
  
  ggplot(aes(x = year, y = annual_prod_kg, color= Major_Functional_Group)) +
  geom_line() +
  
  facet_wrap(~ OHI_rgn_number, scales = "free_y") +
  
  labs(x = "Year", y = "Production (kg)",
       title = "Bait Production over Time [DRAFT]") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

# PEARLS, NACRE, and SEASHELLS ------------------------------
```{r Take 1 Shells}
Shells_keywords <- c(
# Shells:
  "CONCHA DE", "CONCHA")

GOC_ShellProduction <- GOC_MarineLandings_corrected %>%
  filter(if_any(keywords_1:keywords_4, ~ . %in% Shells_keywords))

table(GOC_ShellProduction$estimated_genus_name)

#    Amiantis  Argopecten     Dosinia    Haliotis    Hexaplex     Lobatus Megapitaria         NEI 
#           1          33         993        3008         252          14          45         600 
# Triplofusus 
#           5 

# Investigating whether or not the shell is separate from the animal:
test<- GOC_MarineLandings_corrected %>% 
  filter(eu_rnpa== "0301000089" & year== 2000)
# kinda seems like catch for shell is logged separately than catch for flesh, which is good. Means I should filter out catches of shell from final fisheries estimates.

Pearls_keywords <- c(
  # Pearls
  "PERLAS", "MEDIAS PERLAS", "CONCHA NACAR")

GOC_PearlProduction <- GOC_MarineLandings_corrected %>%
  filter(if_any(keywords_1:keywords_4, ~ . %in% Pearls_keywords))
```

```{r Take 2 Shells and pearls}
# Apr 30 2026: 

Shells_keywords <- c(
# Shells:
  "CONCHA DE", "CONCHA")

Pearls_keywords <- c(
  # Pearls
  "PERLAS", "MEDIAS PERLAS", "CONCHA NACAR")

GOC_Possible.NPLandings <- GOC_Possible.NPLandings %>%
  mutate(
    ShellProduction = case_when(
      if_any(keywords_1:keywords_4, ~ .x %in% Shells_keywords) ~ "YES",
      TRUE ~ "NO"
      ),
    PearlNacreProduction = case_when(
      if_any(keywords_1:keywords_4, ~ .x %in% Pearls_keywords) ~ "YES",
      TRUE ~ "NO"
    )
  )
# 107256 obs., 72 vars.

GOC_ShellProduction<- GOC_Possible.NPLandings %>%
  filter(ShellProduction %in% c("YES"))
# 5859 obs., 70 vars.
table(GOC_ShellProduction$Major_Functional_Group)

GOC_PearlProduction<- GOC_Possible.NPLandings %>%
  filter(PearlNacreProduction %in% c("YES"))
# 18 obs., 70 vars.
# Note, registered pearl production is too small to be meaningful...
table(GOC_PearlProduction$Major_Functional_Group)
```

```{r Plot shell production}
# Summary table by region and year:
ShellProduction_summary <- GOC_ShellProduction %>%
  group_by(OHI_rgn_number, year, Major_Functional_Group) %>%
  summarize(annual_prod_kg = sum(weight_kg_corrected))
# 142 obs., 4 vars.

# Plot over time:
ShellProduction_summary %>%
  filter(year >= 2018 & year < 2025) %>%
  
  ggplot(aes(x = year, y = annual_prod_kg, color= Major_Functional_Group)) +
  geom_line() +
  
  facet_wrap(~ OHI_rgn_number, scales = "free_y") +
  
  labs(x = "Year", y = "Production (kg)",
       title = "Shell Production over Time [DRAFT]") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```


# SHARK FINS ------------------------------
```{r Shark fins}
Sharkfin_keywords <- c(
# Fins
  "ALETA", "ALETA DE", "ALETA ECA DE", "ALETA FCA DE")

GOC_SharkFinProduction <- GOC_MarineLandings_corrected %>%
  filter(if_any(keywords_1:keywords_4, ~ . %in% Sharkfin_keywords))
# No shark fin?

GOC_SharkFinProduction <- GOC_MarineLandings_corrected %>%
  filter(Major_Functional_Group == "Sharks")

table(GOC_SharkFinProduction$keywords_2)

# Shark fin, "ALETA" didn't get pulled out correctly. Pull out ALETA in CONAPESCA_CleaningWorkflow.
```



# FISH OIL, FISH MEAL ------------------------------
Assume that ~70% forage fish is used for fish oil fish meal

```{r Pull out forage fish}

```

```{r Apply 70% conversion}

```

# ALGAE ------------------------------
Sargassum