---
title: "NP CONAPESCA Data Exploration_Oct thru Nov 2025_EMF"
output: html_document
date: "2025-10-29"
---

# Natural Products Inquiry (dataMares data) (October 2025) ----------------
## Understanding OTRAS
```{r CONAPESCA OTRAS explore}
CONAPESCA_data_dataMares <- read_csv("/Users/ferrer/Desktop/OHI/FP/dataMares_CONAPESCA_Data_Oct6_2025.csv")
names(CONAPESCA_data_dataMares) <- c("Index", "Year", "State", "Office", "Taxa", "General_Sp_Name", "Sp_Scientific_Name", "Fishery_Type", "Landing_Site", "Fishing_Site", "Days_Spent", "Date", "Revenue_mxn", "Biomass_kg", "Price_mxn")

table(CONAPESCA_data_dataMares$General_Sp_Name)
table(CONAPESCA_data_dataMares$Sp_Scientific_Name[which(CONAPESCA_data_dataMares$General_Sp_Name=="CORVINA")])
# OTRAS = 616392 obs.

States_of_Interest<- c("BAJA CALIFORNIA", "BAJA CALIFORNIA SUR", "SONORA", "SINALOA", "NAYARIT", "JALISCO")

Landings_filtered<- CONAPESCA_data_dataMares %>% filter(State %in% States_of_Interest)
# n obs.= 4,755,692

OTRAS_dataMares<- Landings_filtered %>% filter(General_Sp_Name=="OTRAS")
# n obs.= 616,392

CORVINA_dataMares<- Landings_filtered %>% filter(General_Sp_Name=="CORVINA")
# n obs.= 178,051
```

```{r Understanding OTRAS}
table(OTRAS_dataMares$Sp_Scientific_Name)
table(OTRAS_dataMares$Year) # 24 years of coverage, 2000-2023
names(OTRAS_dataMares)

OTRAS_summary <- OTRAS_dataMares %>%
  select(Year, State, Sp_Scientific_Name, Biomass_kg, Revenue_mxn) %>%
  group_by(Year, State, Sp_Scientific_Name) %>%
  summarise(
    Total_Biomass = sum(Biomass_kg, na.rm = TRUE),
    Total_Revenue = sum(Revenue_mxn, na.rm = TRUE),
    n_Records = n(),
    .groups = "drop"  # optional: ungroup after summarising
  )

# Identify species with ≥10 records in MOST years
# Step 1: get the number of unique years in your dataset
n_years <- OTRAS_summary %>% pull(Year) %>% unique() %>% length()

# Step 2: filter and summarize
species_with_10plus_most_years <- OTRAS_summary %>%
  group_by(Year, Sp_Scientific_Name, State) %>%
  summarise(yearly_records = sum(n_Records), .groups = "drop") %>%
  filter(yearly_records >= 10) %>%
  distinct(Sp_Scientific_Name, Year) %>%
  count(Sp_Scientific_Name) %>%
  filter(n > n_years / 2) %>%
  pull(Sp_Scientific_Name)

OTRAS_filtered<- OTRAS_dataMares %>% filter(Sp_Scientific_Name %in% species_with_10plus_most_years)

OTRAS_filtered_summary <- OTRAS_filtered %>%
  select(Year, State, Sp_Scientific_Name, Biomass_kg, Revenue_mxn) %>%
  group_by(Year, State, Sp_Scientific_Name) %>%
  summarise(
    Total_Biomass = sum(Biomass_kg, na.rm = TRUE),
    Total_Revenue = sum(Revenue_mxn, na.rm = TRUE),
    n_Records = n(),
    .groups = "drop"  # optional: ungroup after summarising
  )
```

```{r Plot OTRAS}
OTRAS_plot<- OTRAS_filtered_summary %>%
  filter(!Sp_Scientific_Name== "ND") %>%
  ggplot(aes(x = Year, y = Total_Biomass, color = Sp_Scientific_Name)) +
  geom_line(size = 1) +
  facet_wrap(~State, scales = "free_y") +
  labs(
    title = "Biomass Landed Over Time by Fishery and State",
    x = "Year",
    y = "Total Biomass (kg)",
    color = "Species"
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold")
  )

OTRAS_plot
```

```{Save plot}
ggplotly(OTRAS_plot)

ggsave(
  filename = "Desktop/OHI/FP/Otras_Biomass_Plot_Oct8_2025.png",  # file name
  plot = OTRAS_plot,            # the plot object
  width = 10,                        # width in inches
  height = 10,                        # height in inches
  dpi = 300                         # resolution (dots per inch)
)
```

## Understanding CORVINA:
```{r Understanding CORVINA}
CORVINA_summary <- CORVINA_dataMares %>%
  select(Year, State, Sp_Scientific_Name, Biomass_kg, Revenue_mxn) %>%
  group_by(Year, State, Sp_Scientific_Name) %>%
  summarise(
    Total_Biomass = sum(Biomass_kg, na.rm = TRUE),
    Total_Revenue = sum(Revenue_mxn, na.rm = TRUE),
    n_Records = n(),
    .groups = "drop"  # optional: ungroup after summarising
  )
```

```{r Plot CORVINA}
CORVINA_plot<- CORVINA_summary %>%
  ggplot(aes(x = Year, y = Total_Biomass, color = Sp_Scientific_Name)) +
  geom_line(size = 1) +
  facet_wrap(~State, scales = "free_y") +
  labs(
    title = "Corvina Sp. Biomass Landed Over Time by and State",
    subtitle = "Totoaba is Totoaba macdonaldi; technically illegal to fish since the 1970s",
    x = "Year",
    y = "Total Biomass (kg)",
    color = "Species"
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold")
  )

CORVINA_plot
```

## Understanding SEASHELLS:
```{r Seashells}
Species_data <- read_csv("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/CONAPESCA_Sp_MasterFile_EMF.csv")

Species_data <- Species_data %>%
  mutate(across(everything(), ~ if (is.character(.)) str_to_lower(.) else .))

CONAPESCA_data_naturalproducts0 <- CONAPESCA_data_filtered %>%
  mutate(across(everything(), ~ if (is.character(.)) str_to_lower(.) else .))

names(CONAPESCA_data_naturalproducts0) <- c("index", "consulta", "state", "office", "taxa", "general_sp_name", "sp_scientific_name", "fishery_type", "landing_site", "fishing_site", "days_spent", "fecha", "revenue_mxn", "biomass_kg", "price_mxn", "date", "year", "month", "day", "CPUE_est")

CONAPESCA_data_naturalproducts0 <- CONAPESCA_data_naturalproducts0 %>%
  left_join(Species_data, by = c("sp_scientific_name" = "ScientificName"))

table(CONAPESCA_data_naturalproducts0$NewClass)
```


# Natural Products Inquiry (J.C. CONAPESCA data) (November 2025) ----------------

## PRE-WORK
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(tidyr)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
# Taxonomy
library(taxize)
library(worrms)
library(rgbif)
# Visualization
library(ggplot2)
library(plotly)
library(htmlwidgets)


setwd("/home/ferrer/ohi_goc_prep/NP/")
```

## NP: TOP LANDINGS
```{r Natural Products Production Associated with Top Landings}
Landings_GoC_TopGroupings_wSp<- readRDS("/home/shares/ohi/OHI_GOC/goal_prep/fis/v2025/int/juan_carlos/TopGroupings_Landings_2000.present_wSp_Nov102025.rds")
LandingReceipts_Keywords<- read_csv("/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/LandingReceipts_KeywordsKey_Nov102025.csv")

NP_Keywords <- LandingReceipts_Keywords %>%
  filter(NP_flag %in% c("yes", "maybe")) %>%
  pull(Keyword)

NP_TopGroupings_Landings <- Landings_GoC_TopGroupings_wSp %>%
  filter(keyword_1 %in% NP_Keywords |
           keyword_2 %in% NP_Keywords |
           keyword_3 %in% NP_Keywords |
           keyword_4 %in% NP_Keywords |
           keyword_5 %in% NP_Keywords)

unique(NP_TopGroupings_Landings$common_name_english_estimated)
# Previous: 3031 obs. over 17 taxa (common names)
# WITH SHARK FIN: 18495 obs. over 26 taxa (common names)
# NOTE: adding additional key word "BUCHE", "CARNADA", AND "S.C." did not change total NP observations.

unique(NP_TopGroupings_Landings$species_name)
# Previous: And 21 styles of preparation
# And 65 styles of preparation
```

```{r NP Summary Table: take 1}
# How many species (log-book entries) are there per grouping?
NP_summary_table <- NP_TopGroupings_Landings %>%
  select(common_name_english_estimated, species_name) %>%
  group_by(common_name_english_estimated) %>%
  summarize(species_list = paste(unique(species_name), collapse = ", "))
```

## All CONAPESCA landings data
Let's look in the CONAPESCA database for very specific types of shellfish
Can't find any landings with Pinctada or Pteria in the name...
Can, however, find entries for "payaso" and "ornato" as suggested by Liliana

```{r Load ALL CONAPESCA data}
AllLandings_GoC<- readRDS("/home/shares/ohi/OHI_GOC/goal_prep/fis/v2025/int/juan_carlos/GoC_Landings_2000.present_EMF_created.rds")
# 3716415 obs.

AllLandings_GoC <- AllLandings_GoC %>%
  mutate(species_name = stri_trans_general(species_name, "latin-ascii"))
# 3716415 obs.
# Pinctada mazatlanica (Mother of pearl)
# Pteria sterna (nacre) 
# Pinctada imbricata (Atlantic mother of pearl)
# Pteria colymbus (Atlantic winged pearl oyster)
```

## NP: ORNAMENTALS
```{r Look into the ORNAMENTAL fishes Liliana mentions}
Ornamental_Landings <- AllLandings_GoC %>%
  filter(grepl("payaso|ornato", species_name, ignore.case = TRUE))
# 5977 obs.

print(unique(Ornamental_Landings$species_name))
unique(Ornamental_Landings$main_species_group)
unique(Ornamental_Landings$state)
colnames(Ornamental_Landings)

# Adjust fisheries 'live weight' versus 'landed weight'
# Kind of looks like sometimes 'landed weight' is 'live weight' and vice versa.
Ornamental_Landings$weight_kg_corrected<- c(NA)

Ornamental_Landings$weight_kg_corrected <- ifelse(
  Ornamental_Landings$live_weight >= Ornamental_Landings$landed_weight,
  Ornamental_Landings$live_weight,
  Ornamental_Landings$landed_weight
)

# How many species (log-book entries) are there per grouping?
Ornamental_summary_table1 <- Ornamental_Landings %>%
  select(year, state, species_name, weight_kg_corrected, value) %>%
  group_by(year, state, species_name,) %>%
  summarize(annual_weight_kg= sum(weight_kg_corrected, na.rm= T),
            annual_value_pesos= sum(value, na.rm= T))

Ornamental_summary_table2 <- Ornamental_Landings %>%
  select(year, state, species_name, weight_kg_corrected, value) %>%
  group_by(year, species_name,) %>%
  summarize(annual_weight_kg= sum(weight_kg_corrected, na.rm= T),
            annual_value_pesos= sum(value, na.rm= T), 
            produced_in = paste(unique(state), collapse = ", ")
            )

# Write CSV to send to Liliana
# write_csv(Ornamental_summary_table2, "/home/ferrer/ohi_goc_prep/NP/v2025/Ornatos_Summary_Nov132025.csv")
```

```{r}
# List of all key terms:
keyword_list <- c(
  "DESC", "IND", "ENT", "FCO", "S", "DESV", "FCA", "DES", "DSC", 
  "ECO", "CONG", "CONGELADO", "DE CULTIVO", "DE CULTIVO VIVO", 
  "FILETE", "FILETE DE", "ECA", "CABEZA DE", "FILETE DE ECO", 
  "PIEL DE ECA", "PIEL DE SECA", "ALETA ECA DE", "DESCABEZADO", "FRESCO", "SECO DE", "ALETA", "SECA", "SECA DE", "SECO", "SECO DE", "DE ALTA MAR", "ENTERO",
  "POLVO DE", "CONCHA DE", "COCINERO",
  "CARNADA", "BUCHE", "S.C.", "CONCHA" # These last 4 words are words added as of Nov. 10
)

# NOTE TO SELF: THIS LIST IS UNDER CONSTRUCTION (as of Nov 10 2025)

# Sort by descending length to match multi-word keywords first
keyword_list <- keyword_list[order(nchar(keyword_list), decreasing = TRUE)]

Seashell_Sp_List <- Seashell_Landings %>%
  select(main_species_group, recorded_species_name = species_name) %>%
  distinct() %>%
  
  # Extract keywords
  mutate(
    keywords = str_extract_all(
      recorded_species_name,
      paste0("\\b(", paste(keyword_list, collapse="|"), ")\\b[.,]?")
    )
  ) %>%
  
  # Pad for unnesting
  mutate(keywords = lapply(keywords, function(x) { length(x) <- 5; x })) %>%
  unnest_wider(keywords, names_sep = "_") %>%
  rename(
    keyword_1 = keywords_1,
    keyword_2 = keywords_2,
    keyword_3 = keywords_3,
    keyword_4 = keywords_4,
    keyword_5 = keywords_5
  ) %>%
  
  # Remove keywords from species name, except for "ALMEJA CONCHA ESPINA" (spiny oyster)
  rowwise() %>%
  mutate(
    recorded_species_name_clean = if_else(
      recorded_species_name == "ALMEJA CONCHA ESPINA",
      recorded_species_name,  # leave as-is
      str_remove_all(
        recorded_species_name,
        paste0("\\b(", paste(keyword_list, collapse="|"), ")\\b[.,]?")
      )
    ),
    recorded_species_name_clean = str_remove_all(recorded_species_name_clean, "\\bY\\b"),
    recorded_species_name_clean = str_squish(recorded_species_name_clean)
  ) %>%
  arrange(main_species_group, recorded_species_name_clean)
```


## NP: SEA SHELLS
```{r Let's look into just the ALMEJA and CARACOL}
##### ALMEJA and CARACOL
Seashell_Landings<- AllLandings_GoC %>% filter(main_species_group==c("ALMEJA", "CARACOL"))
# 50336 obs.

unique(Seashell_Landings$species_name)
```

```{r Filter out keywords from recorded_species_name}
# List of all key terms:
keyword_list <- c(
  "DESC", "IND", "ENT", "FCO", "S", "DESV", "FCA", "DES", "DSC", 
  "ECO", "CONG", "CONGELADO", "DE CULTIVO", "DE CULTIVO VIVO", 
  "FILETE", "FILETE DE", "ECA", "CABEZA DE", "FILETE DE ECO", 
  "PIEL DE ECA", "PIEL DE SECA", "ALETA ECA DE", "DESCABEZADO", "FRESCO", "SECO DE", "ALETA", "SECA", "SECA DE", "SECO", "SECO DE", "DE ALTA MAR", "ENTERO",
  "POLVO DE", "CONCHA DE", "COCINERO",
  "CARNADA", "BUCHE", "S.C.", "CONCHA" # These last 4 words are words added as of Nov. 10
)

# NOTE TO SELF: THIS LIST IS UNDER CONSTRUCTION (as of Nov 10 2025)

# Sort by descending length to match multi-word keywords first
keyword_list <- keyword_list[order(nchar(keyword_list), decreasing = TRUE)]

Seashell_Sp_List <- Seashell_Landings %>%
  select(main_species_group, recorded_species_name = species_name) %>%
  distinct() %>%
  
  # Extract keywords
  mutate(
    keywords = str_extract_all(
      recorded_species_name,
      paste0("\\b(", paste(keyword_list, collapse="|"), ")\\b[.,]?")
    )
  ) %>%
  
  # Pad for unnesting
  mutate(keywords = lapply(keywords, function(x) { length(x) <- 5; x })) %>%
  unnest_wider(keywords, names_sep = "_") %>%
  rename(
    keyword_1 = keywords_1,
    keyword_2 = keywords_2,
    keyword_3 = keywords_3,
    keyword_4 = keywords_4,
    keyword_5 = keywords_5
  ) %>%
  
  # Remove keywords from species name, except for "ALMEJA CONCHA ESPINA" (spiny oyster)
  rowwise() %>%
  mutate(
    recorded_species_name_clean = if_else(
      recorded_species_name == "ALMEJA CONCHA ESPINA",
      recorded_species_name,  # leave as-is
      str_remove_all(
        recorded_species_name,
        paste0("\\b(", paste(keyword_list, collapse="|"), ")\\b[.,]?")
      )
    ),
    recorded_species_name_clean = str_remove_all(recorded_species_name_clean, "\\bY\\b"),
    recorded_species_name_clean = str_squish(recorded_species_name_clean)
  ) %>%
  arrange(main_species_group, recorded_species_name_clean)

unique(Seashell_Sp_List$keyword_1)
unique(Seashell_Sp_List$recorded_species_name_clean)

# Only 82 names to sort through here!!
```

```{r Use taxize to determine scientific names where possible}
# Step 1: Extract scientific names from parentheses
Seashell_Sp_List$scientific_name <- sub(".*\\(([^)]+)\\).*", "\\1", Seashell_Sp_List$recorded_species_name_clean)
Seashell_Sp_List$scientific_name[!grepl("\\(", Seashell_Sp_List$recorded_species_name_clean)] <- NA
Seashell_Sp_List$scientific_name <- trimws(Seashell_Sp_List$scientific_name)

# Convert to proper binomial capitalization
capitalize_binomial <- function(x){
  x <- tolower(x)                    # all lowercase
  words <- strsplit(x, " ")[[1]]     # split words
  if(length(words) >= 2){
    words[1] <- paste0(toupper(substring(words[1],1,1)), substring(words[1],2))  # Genus capitalized
    words[2] <- words[2]  # species lowercase
    return(paste(words[1], words[2]))
  } else {
    return(x)
  }
}

Seashell_Sp_List$scientific_name_clean <- sapply(Seashell_Sp_List$scientific_name, function(x){
  if(!is.na(x)) capitalize_binomial(x) else NA
})

# Step 2: Unique names to query
taxa_list <- unique(na.omit(Seashell_Sp_List$scientific_name_clean))

# Step 3: Get GBIF IDs safely
# function to get classification from GBIF
get_gbif_classification <- function(name){
  # search for the name
  res <- name_backbone(name = name, strict = FALSE)  # fuzzy search
  if(is.null(res$species)) res$species <- NA
  if(is.null(res$genus)) res$genus <- NA
  if(is.null(res$family)) res$family <- NA
  data.frame(
    scientific_name = name,
    species = res$species,
    genus = res$genus,
    family = res$family,
    stringsAsFactors = FALSE
  )
}

gbif_results <- do.call(rbind, lapply(taxa_list, get_gbif_classification))

# Step 4: Merge
# Merge GBIF taxonomy back to the original dataframe
Seashell_Sp_List <- merge(
  Seashell_Sp_List,
  gbif_results,
  by.x = "scientific_name_clean",   # column in your original df
  by.y = "scientific_name",         # column in gbif_results
  all.x = TRUE                      # keep all rows from Seashell_Sp_List
)
```

```{r save this as a CSV}
#write.csv(Seashell_Sp_List, "/home/ferrer/ohi_goc_prep/NP/v2025/Seashell_SpKey_UNDER CONSTRUCTION_Nov102025.csv", row.names = FALSE)
```

Added information about each top-landed species manually, offline.
```{r Manually research and add species info (Nov 10 2025)}
dataMares_CONAPESCA_data<- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/CONAPESCA/d2025/SharedByDataMares/", "dataMares_CONAPESCA_Data_Oct6_2025.csv"))
# 4755692 obs.

#### Manually identify species info given their common names (Nov 10)
SpeciesInquiry<- dataMares_CONAPESCA_data %>% filter(`Especie nombre general`=="ALMEJA")
unique(SpeciesInquiry$`Especie nombre`)
unique(SpeciesInquiry$Especie_nombre_cientifico)
unique(SpeciesInquiry$Especie_nombre_cientifico[which(SpeciesInquiry$`Especie nombre`=="RONOSA NEGRA")])

get_gbifid_("Megathura", method= "backbone")

# rm(SpeciesInquiry)
# rm(dataMares_CONAPESCA_data)

#### TOP-LANDED SPECIES LIST - Nov 5 2025
#Seashell_Sp_List<- read_csv(here("/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/PrelimLook_CONAPESCA_SpeciesNames_Oct302025/Seashell_Sp_List_Nov052025.csv"))
# 600 obs., 17 columns
unique(Seashell_Sp_List$common_name_english_estimated)
```

```{r ALMEJA and CARACOL species key}
Seashell_SpKey<- read_csv("/home/ferrer/ohi_goc_prep/NP/v2025/Seashell_SpKey_UNDER CONSTRUCTION_Nov102025.csv")
```

```{r Append Seashell_SpKey to Seashell_Landings}
Seashell_Landings_wSp<- left_join(Seashell_Landings, Seashell_SpKey, 
          by = c("main_species_group", "species_name"="recorded_species_name"))
# 50336 obs., 40 vars.

Seashell_Landings_wSp<- Seashell_Landings_wSp[, -c(25:26)]
# 50336 obs., 38 vars.

unique(Seashell_Landings_wSp$keyword_1)
```

```{r Seashell Summary Table: take 2}
# How many species (log-book entries) are there per grouping?
Seashell_summary_table <- Seashell_Landings_wSp %>%
  select(genus_estimated, species_estimated) %>%
  group_by(genus_estimated) %>%
  summarize(species_list = paste(unique(species_estimated), collapse = ", "))
```

```{r NP Production Associated with Top Landings}
Keywords_Key<- read_csv("/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/LandingReceipts_KeywordsKey_Nov102025.csv")

NP_Keywords <- Keywords_Key %>%
  filter(NP_flag %in% c("yes", "maybe")) %>%
  pull(Keyword)

NP_Seashell_Landings <- Seashell_Landings_wSp %>%
  filter(acuaculture_production == "yes" |
           keyword_1 %in% NP_Keywords |
           keyword_2 %in% NP_Keywords |
           keyword_3 %in% NP_Keywords |
           keyword_4 %in% NP_Keywords |
           keyword_5 %in% NP_Keywords)

unique(NP_Seashell_Landings$common_name_english_estimated)
unique(NP_Seashell_Landings$main_species_group)
# 1428 obs. over 22 taxa (common names) and 2 groupings

unique(NP_Seashell_Landings$species_name)
# And 32 styles of preparation
```

```{r Join NP dataframes together}
NP_Fishery_Landings <- bind_rows(NP_Seashell_Landings, NP_TopGroupings_Landings)
# 19923 obs., 38 vars.

unique(NP_Fishery_Landings$main_species_group)
unique(NP_Fishery_Landings$species_name)
```

```{r Adjust fisheries 'live weight' versus 'landed weight'}
# Kind of looks like sometimes 'landed weight' is 'live weight' and vice versa.
NP_Fishery_Landings$weight_kg_corrected<- c(NA)

NP_Fishery_Landings$weight_kg_corrected <- ifelse(
  NP_Fishery_Landings$live_weight >= NP_Fishery_Landings$landed_weight,
  NP_Fishery_Landings$live_weight,
  NP_Fishery_Landings$landed_weight
)
```

```{r Join NP dataframes together}
NP_Fishery_Landings <- bind_rows(NP_Seashell_Landings, NP_TopGroupings_Landings)
# 19923 obs., 38 vars.

unique(NP_Fishery_Landings$main_species_group)
unique(NP_Fishery_Landings$species_name)
```

# Natural Products List for NP group meeting #4 - UNDER CONSTRUCTION
```{r Natural Products: products by name}
# How many species (log-book entries) are there per grouping?
NP_List_for_Review <- NP_Fishery_Landings %>%
  select(main_species_group, species_name, common_name_english_estimated,
    species_estimated, genus_estimated, family_estimated,
    keyword_1, keyword_2, keyword_3, keyword_4, keyword_5,
    year, weight_kg_corrected
  ) %>%
  group_by(main_species_group, species_name, common_name_english_estimated,
           species_estimated, genus_estimated, family_estimated) %>%
  summarize(
    mean_annual_biomass = sum(weight_kg_corrected, na.rm = TRUE) / n_distinct(year),
    products = paste(
      unique(na.omit(c(keyword_1, keyword_2, keyword_3, keyword_4, keyword_5))),
      collapse = ", "
    ),
 .groups = "drop"
  )

NP_List_for_Review <- NP_List_for_Review %>%
  relocate(products, .after = common_name_english_estimated)

NP_List_for_Review$added_by<- c("Erica_Nov_2025")
NP_List_for_Review$NP_SI.NO<- c("maybe")

# Add Liliana's additions:
NP_Additions_by_Liliana<- read_csv("/home/ferrer/ohi_goc_prep/NP/v2025/Avisos_NP_fromLiliana_Nov122025.csv")

# Reduce the dataframe down to unique records so that I can add them to the NP_List_for_Review
NP_Additions_by_Liliana<- NP_Additions_by_Liliana[,c(28,30)]
NP_Additions_by_Liliana<- unique(NP_Additions_by_Liliana)

# Re-name the columns:
colnames(NP_Additions_by_Liliana) <- c("main_species_group", "species_name")

# Add a column:
NP_Additions_by_Liliana$added_by<- c("Liliana_Nov_2025")
NP_Additions_by_Liliana$NP_SI.NO<- c("yes")

# Join both lists together into master list
NP_List_for_Review <- bind_rows(NP_Additions_by_Liliana, NP_List_for_Review)
```

```{r Save this dataframe NP_List_for_Review}
# write.csv(NP_List_for_Review, "/home/ferrer/ohi_goc_prep/NP/v2025/NP_MasterList_UNDER CONSTRUCTION_Nov122025.csv", row.names = FALSE)
```


# NP work continued...
```{r Natural Products: Fisheries Residues by State}
# How many species (log-book entries) are there per grouping?
NP_Genus_summary_table <- NP_Fishery_Landings %>%
  select(year, state, genus_estimated, weight_kg_corrected, keyword_1, keyword_2, keyword_3, keyword_4, keyword_5) %>%
  group_by(year, state, genus_estimated) %>%
  summarize(
    TotalBiomass_annual = sum(weight_kg_corrected, na.rm = TRUE),
    products = paste(
      unique(na.omit(c(keyword_1, keyword_2, keyword_3, keyword_4, keyword_5))),
      collapse = ", "
    ),
    .groups = "drop"
  )
```

```{r Plot products through time by state}
# Plot seashells, shark fins, and dried fish
NP_Fisheries_Biomass_plot<- NP_Genus_summary_table %>%
  filter(genus_estimated != "Thunnus") %>%
  filter(year > 2014 & year < 2025) %>%
  ggplot(
       aes(x = year,
           y = log10(TotalBiomass_annual),        # match your new column name
           color = genus_estimated,
           group = genus_estimated)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  facet_wrap(~ state, scales = "free_y") +
  labs(
    title = "Natural Products Total Biomass (2015–2024)",
    subtitle = "Products include seashells, heads, and shark fins [under construction]",
    x = "Year",
    y = "Total Biomass (kg)",
    color = "Common Name"
  ) +
  theme_bw(base_size = 13) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

NP_Fisheries_Biomass_plot

# Save the plot as PNG
# ggsave(filename = "/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/PrelimLook_CONAPESCA_SpeciesNames_Oct302025/Corvina.Biomass_plot_Nov032025.png",
#        plot = Corvina_Biomass_plot,
#        width = 10,        # width in inches
#        height = 6,        # height in inches
#        dpi = 300)         # resolution
```

```{r Summary table: how the products are prepared}
NP_Prep_summary_table <- NP_Fishery_Landings %>%
  select(year, state, genus_estimated, weight_kg_corrected, keyword_1, keyword_2, keyword_3, keyword_4, keyword_5) %>%
  pivot_longer(
    cols = starts_with("keyword_"),
    names_to = "keyword_type",
    values_to = "keyword"
  ) %>%
  filter(!is.na(keyword) & keyword != "") %>%  # remove blanks and NAs
  group_by(year, state, keyword) %>%
  summarize(
    TotalBiomass_annual = sum(weight_kg_corrected, na.rm = TRUE),
    products = paste(
      unique(genus_estimated),
      collapse = ", "
    ),
    .groups = "drop"
  )
```

```{r Plot products through time by state}
# Plot seashells, shark fins, and dried fish
NP_Biomass_by_Prep_plot<- NP_Prep_summary_table %>%
  filter(year > 2014 & year < 2025) %>%
  ggplot(
       aes(x = year,
           y = log10(TotalBiomass_annual),        # match your new column name
           color = keyword,
           group = keyword)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  facet_wrap(~ state, scales = "free_y") +
  labs(
    title = "Natural Products Total Biomass (2015–2024)",
    subtitle = "Products include seashells, heads, and shark fins [under construction]",
    x = "Year",
    y = "Total Biomass (kg)",
    color = "Common Name"
  ) +
  theme_bw(base_size = 13) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

NP_Biomass_by_Prep_plot

# Save the plot as PNG
# ggsave(filename = "/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/PrelimLook_CONAPESCA_SpeciesNames_Oct302025/Corvina.Biomass_plot_Nov032025.png",
#        plot = Corvina_Biomass_plot,
#        width = 10,        # width in inches
#        height = 6,        # height in inches
#        dpi = 300)         # resolution
```