---
title: 'OHI GoC 2025: Muddy Coasts'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: inline
editor_options: 
  chunk_output_type: console
---

# Summary

## Global distribution and dynamics of muddy coasts

This dataset is associated with the scientific article "Global distribution and dynamics of muddy coasts"  (https://www.nature.com/articles/s41467-023-43819-6)

The dataset contains the following items:

VisualizeResults.qgz: QGIS project with the formatted datasets ready for viewing (tested in QGIS v3.28.2.)
PredictedClasses_MuddyBeaches.shp (and associated files): Shapefile containing predicted classes of coastal transects differentiating the following types:
0: Sandy coastal transect
1: Muddy coastal transect
2: Rocky coastal transect
3: Vegetated coastal transect
4: Other undefined type
MuddyZonesFleming.shp (and associated files): Shapefile containing hand-drawn polygons based on the description of muddy coastline concentrations by (Flemming, B. W. Geographic distribution of muddy coasts, 2002).
TransectTrainingDataframe.csv : comma separated value file containing 1867 hand-labelled coastal type transects randomly distributed around the world. Each transect contains the following entries:
transect_id: individual transect ID
box_id: tiled ID
country_name
continent
New_Start_lon: longitude transect start
New_Start_lat: latitude transect start
New_End_lon: longitude transect end
New_End_lat: latitude transect end
Center_lon
Center_lat
Length: transect length
label: hand-labeled coastal type, Str (Sandy, Muddy, Rocky, Vegetated, Other)
abs lat: latitude
flag_sandy: bool, classified as a sandy area (Luiijendijk et al, 2018)
sand: % of pixels classified as sand from Sentinel 2 Multispectral land-type classification
mud: % of pixels classified as muddy from Sentinel 2 Multispectral land-type classification
water: % of pixels classified as water from Sentinel 2 Multispectral land-type classification
vegetation: % of pixels classified as vegetation from Sentinel 2 Multispectral land-type classification
dry: % of pixels classified as dry-vegetation from Sentinel 2 Multispectral land-type classification
other: % of pixels classified as other from Sentinel 2 Multispectral land-type classification
turbid % of pixels classified as turbid-water from Sentinel 2 Multispectral land-type classification
height max: max height in transect (MERIT-DEM)
height var: variance height in transect (MERIT-DEM)
mangrove: bool, mangrove presence (Global Mangrove Forest Distributions)
intertidal: intertidal range Murray et al 2019.
gsw: Global Surface Water dataset, Pekel et al 2016
maxtemp: max temperature WorldClim V1 Bioclim dataset
mintemp: min temperature WorldClim V1 Bioclim dataset
mhhw: Tide MHHW Deltares GTSM model
mllw: Tide MLLW Deltares GTSM model
tidal range: Tidal range Deltares GTSM model
PixelTrainingDataframe.csv : comma separated value file containing 3240 hand-labelled land-type pixel randomly distributed around the world. These are derived from a Sentinel 2 TOA images as a 2020 composite. Bands (BX) refer to the Sentinel-2 MSI Multispectral Instrument, level-2A spectral bands names. Each entry contains:
Aerosols: B1
Blue: B2
Green: B3
Red: B4
Red Edge 1: B5
Red Edge 2: B6
Red Edge 3: B7
NIR: B8
Red Edge 4: B8A
SWIR 1: B11
SWIR 2: B12
NDWI: Normalized Differential Water Index ((B3-B8)/(B3+B8)
NDVI ((B8-B4)/(B8+B4)
num: hand-label type differentiating the following types (1) sandy beaches, (2) mudflats, (3) clear water, (4) 111 turbid (brown) water, (5) green vegetation, (6) dry vegetation, and (7) other (containing clouds, snow, buildings, and all other indefinable points)


# Setup

```{r}
library(raster)
library(terra)
library(readr)
library(magrittr)
library(mregions2)
library(mapview)
library(sf)
library(tidyverse)
library(lwgeom)
library(here)
library(leaflet.extras2)
library(fasterize)
library(tictoc)
library(kableExtra)
library(plotly)
library(viridis)
library(readr)
library(purrr)
library(doParallel)
library(httr2) # to handle timeouts (error HTTP 524 NA)
library(data.table)
library(leaflet)
library(leaflet.extras)
library(RColorBrewer)

# ---- sources! ----
# source(here("workflow", "R", "common.R")) # file creates objects to process data
## set the mazu and neptune data_edit share based on operating system
dir_M             <- c('Windows' = '//mazu.nceas.ucsb.edu/ohi',
                       'Darwin'  = '/Volumes/ohi',    ### connect (cmd-K) to smb://mazu/ohi
                       'Linux'   = '/home/shares/ohi')[[ Sys.info()[['sysname']] ]]

# ---- set year and file path info ----
current_year <- 2025 # Update this in the future!!
version_year <- paste0("v",current_year)
data_dir_version_year <- paste0("d", current_year)

# ---- data directories ----

# raw data directory (on Mazu)
raw_data_dir <- here::here(dir_M, "OHI_GOC", "_raw_data")

# for muddy coasts raw data dir
mc_raw_dir <- here::here(dir_M, "OHI_GOC", "_raw_data", "hab_muddy_coasts", data_dir_version_year)

# for coastline characteristics intermediates
int_dir <- here::here("hab", "hab_muddy_coasts", version_year, "int")

# final output dir
output_dir <- here("hab", "hab_muddy_coasts", version_year, "output")

# spatial data for GoC
goc_spatial <- here("spatial")

# this CRS might be better for visualization, explore.
gulf_crs <- st_crs("+proj=aea +lat_1=23 +lat_2=30 +lat_0=25 +lon_0=-110 +datum=WGS84 +units=m +no_defs")
```


# Data exploration

```{r}
# read in the shapefile for 2020, examine data
mc_raw <- st_read(here(mc_raw_dir, "MuddyZonesFleming.shp")) 
View(mc_raw) # inspected object, no mexico polygons from Fleming. Would have to use the predicted classes.

predicted_classes_mb_raw <- st_read(here(mc_raw_dir, "PredictedClasses_MuddyBeaches.shp")) # WGS 84, point data
head(predicted_classes_mb_raw) 
# prediction: 
# 0: Sandy coastal transect
# 1: Muddy coastal transect
# 2: Rocky coastal transect
# 3: Vegetated coastal transect
# 4: Other undefined type
```

## Limit to GoC Ecoregion

```{r}
# read in ecoregion shapefiles in 
eco_hr_latlon <- st_read(here(goc_spatial, "GoC_polygon_ecoregion_latlon.shp"))
# mapview(eco_hr_latlon)
head(eco_hr_latlon)

eco_hr_eqarea <- st_read(here(goc_spatial, "GoC_polygon_ecoregion_eqarea.shp"))
mapview(eco_hr_eqarea)

# check CRS of ecoregion boundary
st_crs(eco_hr_latlon) # WGS84, can be used to crop global data

# transform to gulf_CRS
# eco_hr_albers <- st_transform(eco_hr_eqarea, crs = "+proj=aea +lat_1=23 +lat_2=30 +lat_0=25 +lon_0=-110 +datum=WGS84 +units=m +no_defs")
# st_crs(eco_hr_albers) # double check it worked well
# mapview(eco_hr_albers) # looks good!
```

```{r}
# crop the global data to the gulf ecoregion boundary
pc_mb_cropped <- st_intersection(predicted_classes_mb_raw, eco_hr_latlon)
head(pc_mb_cropped) # attributes are conserved
unique(pc_mb_cropped$prediction) # 0 - 4, good

mapview(pc_mb_cropped)

# write_sf(pc_mb_cropped, here::here("hab", "hab_muddy_coasts", "v2025"," ","www", "pc_mb_cropped_wgs84.shp"))
write_sf(pc_mb_cropped, here::here(int_dir, "pc_mb_cropped_wgs84.shp"))
```


# Interactive leaflet of coastline types

```{r}
library(leaflet)
library(sf)

viridis_colors <- viridis::viridis(6)

# match colors to class meanings for clarity
# 0: Sandy coastal transect
# 1: Muddy coastal transect
# 2: Rocky coastal transect
# 3: Vegetated coastal transect
# 4: Other undefined type
color_mapping <- c(
  "0" = viridis_colors[1],  # sandy
  "1" = viridis_colors[2],  # muddy
  "2" = viridis_colors[3],  # rocky
  "3" = viridis_colors[4],  # vegetated
  "4" = viridis_colors[5]   # other
)

# use color palette function
pal <- colorFactor(
  palette = color_mapping,
  domain = pc_mb_cropped$prediction
)

# class descriptions for popup
class_descriptions <- c(
  "0" = "Sandy coastal transect",
  "1" = "Muddy coastal transect",
  "2" = "Rocky coastal transect",
  "3" = "Vegetated coastal transect",
  "4" = "Other undefined type"
)

leaflet(pc_mb_cropped) %>%
  addTiles() %>%
  setView(lng = -112, lat = 28, zoom = 6) %>% 
  addCircleMarkers(
    color = ~pal(prediction),
    radius = -2,
    popup = ~paste(
      "<b>Prediciton:</b>", prediction, "<br>",
      "<b>Description:</b>", class_descriptions[as.character(prediction)] # map values to descriptions
    ),
    layerId = ~prediction  # unique layer IDs based on class
  ) %>%
  addLegend(
    pal = pal,
    values = ~prediction,
    title = "Predicted Class",
    opacity = 0.7,
    labFormat = labelFormat(
      transform = function(value) {
        class_descriptions[as.character(value)] # display description
      }
    )
  )

```














