################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(stringr)
library(stringi)
library(purrr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(fuzzyjoin)
# Visualization
library(ggplot2)
library(plotly)
library(htmlwidgets)


setwd("/home/ferrer/ohi_goc_prep/")
```


# PART 1: Filter landings data and create skeleton for species key ------------------------------
First we need to correctly clean and filter landings data so that we can append species info and extract species info for a key.

## Filtering out key words from landing receipts:
```{r Load in GoC landings data}
Landings_GoC<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fis/v2025/int/juan_carlos/GoC_Landings_2000.present_EMF_created.rds"))
# n = 3716415 obs.
```

```{r Is aleta de aleta an issue? Short answer, no.}
Landings_GoC %>%
  rename(recorded_taxa_name = species_name) %>%
  mutate(
    aleta_count = stringi::stri_count_fixed(recorded_taxa_name, "ALETA")
  ) %>%
  count(aleta_count)

# No, double aleta does not seem to be an issue...
```

```{r Filter out keywords from recorded_taxa_name to get recorded_taxa_name_clean}
# 1. Keyword list (excluding ALETA, handled separately) -----------------
keyword_list <- c(
  "S.C.", "S/C", "S",
  "ENTERO", "ENTERA", "IND", "ENT",
  "FRESCO", "FRESCA", "FCA", "FCO", "FC",
  "ECO",
  "CONG", "CONGELADO",
  "VIVO", "VIVA", "VIVAS", "VIVITA",
  "DE CULTIVO", "DE CULTIVO VIVO", "ABASTO RANCHO", "RANCHO",
  "HUEVA DE",
  "FILETE", "FILETE DE", "ECA", "FILETE DE ECO", "DESHUESADA", "CARNE", "CARNE DE",
  "DESCABEZADO", "DES", "DSC", "DESC", "DESV", "DESVICERADO",
  "CABEZA DE", "BUCHE", "BUCHE DE", "TRIPA DE", "COLAS DE", "MANOS DE", "MANOS",
  "ANCAS DE", "ANCAS", "ANCA",
  "PIEL DE ECA", "PIEL DE SECA", "PIEL DE", "PIEL",
  "SECO DE", "SECA", "SECA DE", "SECO", "SECO DE",
  "COCINERO", "COCIDO", "COCIDA", "COCIDAS", "ASADO",
  # Remove ALETA from main keywords; handled separately
  #"ALETA", "ALETA ECA DE", "ALETA FCA DE",
  "DE ALTA MAR",
  "CONCHA DE", "CONCHA",
  "POLVO DE",
  "CARNADA", "P/CARNADA",
  "DE AGUA DULCE",
  "MANTO", "MANTO DE", "PULPA DE", "PULPA FRESCA",
  "PULPO DE", "TENTACULOS DE", "TUBO DE", "TUBO",
  "NO APTO PARA EMPAQUE"
)

# Longest-first ordering
keyword_list <- keyword_list[order(nchar(keyword_list), decreasing = TRUE)]

# Escape regex meta-characters
keyword_list_escaped <- stringi::stri_replace_all_regex(
  keyword_list, "([.|(){}+*?^$\\[\\]\\\\])", "\\\\$1"
)

# Final regex
keyword_pattern <- paste0(
  "(?<![A-Za-z0-9])(",
  paste(keyword_list_escaped, collapse = "|"),
  ")(?![A-Za-z0-9])"
)


# 2. Special ALETA phrases (handled separately from other key words) -----------------

special_aleta_adj <- c("BLANCA", "AMARILLA", "CORTA", "AZUL")

remove_aleta_except_special <- function(x) {
  stringi::stri_replace_all_regex(
    x,
    pattern = paste0("\\bALETA\\b(?!\\s+(?:", paste(special_aleta_adj, collapse = "|"), "))"),
    replacement = "",
    vectorize_all = FALSE
  )
}


# 3. MAIN PIPELINE (optimized) -----------------

Landings_GoC_Cleaned <- Landings_GoC %>%
  
  # Rename columns
  rename(
    recorded_taxa_name = species_name,
    recorded_taxa_group = main_species_group
  ) %>%
  
  # Extract keywords
  mutate(
    keywords_list = stringi::stri_extract_all_regex(recorded_taxa_name, keyword_pattern)
  ) %>%
  mutate(keywords_list = map(keywords_list, ~ { length(.x) <- 5; .x })) %>%
  unnest_wider(keywords_list, names_sep = "_") %>%
  rename_with(~ paste0("keywords_", 1:5), starts_with("keywords_list_")) %>%
  
  # Clean recorded_taxa_name
  mutate(
    recorded_taxa_name_clean = recorded_taxa_name %>%
      remove_aleta_except_special() %>%                      # Remove ALETA unless special
      stringi::stri_replace_all_regex(keyword_pattern, "") %>% # Remove other keywords
      stringi::stri_replace_all_regex("\\b(Y|DE|P/)\\b", "") %>% # Remove standalone Y/DE/P/
      stringi::stri_replace_all_regex("[.,/\\\\]+", " ") %>%      # Remove punctuation
      str_squish()
  ) %>%
  
  # Extract scientific name
  mutate(
    recorded_scientific_name = stringi::stri_extract_first_regex(recorded_taxa_name_clean, "\\(([^)]+)\\)") %>%
      stringi::stri_replace_all_regex("\\(|\\)", ""),
    scientific_name_recorded_T.F = !is.na(recorded_scientific_name),
    recorded_scientific_name = if_else(
      scientific_name_recorded_T.F,
      stringi::stri_trans_totitle(recorded_scientific_name),
      NA_character_
    )
  ) %>%
  
  # Extract common name
  mutate(
    common_name_clean = stringi::stri_trim_both(
      stringi::stri_replace_first_regex(recorded_taxa_name_clean, "\\s*\\([^)]*\\)", "")
    ),
    common_name_clean = tolower(common_name_clean)
  )


# 4. REORDER COLUMNS -----------------

Landings_GoC_Cleaned <- Landings_GoC_Cleaned %>%

  select(
    # General info
    state, office_name, landing_site, landing_site_key, landing_site_primary, GoC, lat, long,
    year, month,
    receipt_date_fixed, period_start_fixed, period_end_fixed,
    fleet, acuaculture_production, eu_rnpa, eu_name, vessel_rnpa, vessel_name,
    
    # Species info
    recorded_taxa_group, recorded_taxa_name,
    
    # Cleaned / derived names
    recorded_taxa_name_clean, recorded_scientific_name,
    scientific_name_recorded_T.F, common_name_clean,
    
    # Keywords
    keywords_1, keywords_2, keywords_3, keywords_4, keywords_5,
    
    # Landing details
    landed_weight, live_weight, value
 
  )

# How many common names are there in this dataset?
length(unique(Landings_GoC_Cleaned$common_name_clean))
length(unique(Landings_GoC_Cleaned$recorded_taxa_group))
# 734 common names (versus 973 with the less comprehensive keywords_list), over 57 taxonomic groups
```

```{r Check that this worked}
TEST <- Landings_GoC_Cleaned %>%
  filter(str_detect(recorded_taxa_name, "ATUN ALETA AMARILLA"))
```
# Yes, it seems to have worked!

## Save Landings_GoC_Cleaned rds file
```{r Save this dataframe - landings keywords & species for top groupings: Nov 10 2025}
saveRDS(Landings_GoC_Cleaned, file="/home/shares/ohi/OHI_GOC/goal_prep/fis/v2025/int/juan_carlos/GoC_Landings_wKeyWords_2000.present_Dec012025.rds")
```
# PART 1 COMPLETE.


# PART 2: Create MASTER SPECIES KEY: by merging the skeleton with keys from Mar, & Hem ------------------------------
```{r Skeleton of master key from Landings_GoC_Cleaned dataframe}
Master_SpKey_Skeleton<- Landings_GoC_Cleaned %>% select(recorded_taxa_group, recorded_taxa_name_clean, common_name_clean, scientific_name_recorded_T.F, recorded_scientific_name) %>% distinct
# 802 observations, 5 variables.
```

```{r Mar and Hem Species Keys}
# AllGoC_SpKey <- read_csv("/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/CONAPESCA_SpeciesNames_OctNov2025/Outdated_SpKeys/AllGoC_SpKey_Oct312025.csv")
# 1246 obs of 8 variables

Erica_SpKey <- read_csv("/home/ferrer/ohi_goc_prep/general/Species_Keys/Erica_CONAPESCA_SpKey_IMPERFECT_Nov052025.csv")
# # 600 obs of 17 variables

Hem_SpKey <- read_csv("/home/ferrer/ohi_goc_prep/general/CONAPESCA_Keys/Species_Keys/Hem_CONAPESCA_SpKey_2018.csv")
 # 8232 obs of 3 variables

Mar_SpKey <- read_csv("/home/ferrer/ohi_goc_prep/general/CONAPESCA_Keys/Species_Keys/Mar_CONAPESCA_SpKey.csv")
# 443 obs of 30 variables
```

```{r Prepare species keys for merging}
# MAR's KEY / THE IHH KEY
Mar_SpKey <- Mar_SpKey %>%
  mutate(common_name = tolower(common_name))

Mar_SpKey_filtered <- Mar_SpKey %>% select(common_name, fishery, highest_taxonomic_id, highest_taxonomic_resolution, species, genus, family, order, class, functional_major_groups_sm= IHH_functional_major_groups_sm)

Mar_SpKey_filtered <- Mar_SpKey_filtered %>%
  select(common_name, fishery, highest_taxonomic_id, highest_taxonomic_resolution,
         species, genus, family, order, class, functional_major_groups_sm) %>%
  rename_with(~ paste0("IHH_", .x))

# HEM's KEY
Hem_SpKey <- Hem_SpKey %>%
  mutate(common_name = tolower(common_name))

# Fix the sharks...
Hem_SpKey_filtered <- Hem_SpKey %>%
  mutate(
    common_name = if_else(
      common_name == "tiburon",   # exact match only
      "tiburon se",
      common_name
    )
  )

# Fix the rockfishes...
Hem_SpKey_filtered <- Hem_SpKey_filtered %>%
  mutate(
    common_name = if_else(
      common_name == "rocote",   # exact match only
      "rocote se",
      common_name
    )
  )

# Fix the grunts...
Hem_SpKey_filtered <- Hem_SpKey_filtered %>%
  mutate(
    common_name = if_else(
      common_name == "ronco",   # exact match only
      "ronco se",
      common_name
    )
  )

Hem_SpKey_filtered<- Hem_SpKey_filtered %>%
  select(Hem_common_name= common_name, species) %>% distinct

Hem_SpKey_filtered <- Hem_SpKey_filtered %>%
  mutate(
    species = if_else(
      str_detect(Hem_common_name, "\\bse\\b"),  # \\b ensures "se" is a whole word
      "undetermined",
      species
    )
  )

# Fix the species de ornato...
Hem_SpKey_filtered <- Hem_SpKey_filtered %>%
  mutate(
    species = if_else(
      Hem_common_name == "especies de ornato",   # exact match only
      "undetermined",
      species
    )
  )

Hem_SpKey_filtered<- Hem_SpKey_filtered %>%
  select(Hem_common_name, species) %>%
  distinct() %>%
  group_by(Hem_common_name) %>%
  mutate(species_num = row_number()) %>%
  ungroup()

# 1235 obs.

Hem_SpKey_filtered <- Hem_SpKey_filtered %>%
  pivot_wider(
    names_from = species_num,
    values_from = species,
    names_prefix = "Hem_species_"
  )
```

```{r Merge}
# merged_df1 <- stringdist_join(
#   Master_SpKey_Skeleton,
#   Mar_SpKey_filtered,
#   by = c("common_name_clean" = "common_name_IHH"),
#   method = "jw",          # Jaro-Winkler works well for name matching
#   max_dist = 0.10,        # adjust as needed
#   distance_col = "dist"   # keep the distance so you can inspect match quality
# )

merged_SpKey_0 <- left_join(
  Master_SpKey_Skeleton,
  Mar_SpKey_filtered,
  by = c("common_name_clean" = "IHH_common_name")
)

Master_SpKey <- left_join(
  merged_SpKey_0,
  Hem_SpKey_filtered,
  by = c("common_name_clean" = "Hem_common_name")
)
```

```{r MASTER SPECIES KEY (not fully cleaned)}
# write.csv(Master_SpKey, "/home/ferrer/ohi_goc_prep/general/Species_Keys/Master_SpKey_Dec012025.csv", row.names = FALSE)
```

# PART 2 COMPLETE.

# PART 3: Clean up and simplify the MASTER SPECIES KEY  ------------------------------
```{r Simplify the master species key}
# WHAT IS THE LOGIC?

# If scientific_name_recorded_T.F == TRUE, then estimated_species_name = scientific_name_recorded and species_est_certainty = "virtually certain" and species_est_reasoning = "name was recorded in CONAPESCA".

# If scientific_name_recorded_T.F == FALSE, then compare IHH_species to Hem_species_.

# > If Hem_species_1 is not NA, IHH_species is not NA, and IHH_species is a fuzzy match to any of Hem_species_, then estimated_species_name = IHH_species, species_est_certainty = "high" and species_est_reasoning = "logical match in both keys".

# > If Hem_species_1 is not NA, IHH_species is not NA, and IHH_species is NOT a fuzzy match to any of Hem_species_, then estimated_species_name = IHH_species, species_est_certainty = "low", and species_est_reasoning = "mismatch between keys".

# > If Hem_species_1 is NA but IHH_species is not NA, then estimated_species_name = IHH_species, species_est_certainty = "high", and  = "match to IHH key".

# > If Hem_species_1 is not NA but IHH_species is NA, then consider the species dictated by the rest of Hem's key: if Hem_species_2 is NA, then estimated_species_name = Hem_species_1, species_est_certainty = "high", and species_est_reasoning = "match to Hem's key"; but if Hem_species_2 is not NA, then estimated_species_name = "unknown", species_est_certainty = "medium", and species_est_reasoning = "conflicting matches in Hem's key and no data from IHH" 


# APPLY THE LOGIC:

# Hem species columns
hem_cols <- paste0("Hem_species_", 1:20)

# Compute estimated_species_name, species_est_certainty, and species_est_reasoning
Master_SpKey_Cleaned <- Master_SpKey %>%
  rowwise() %>%
  mutate(
    hem_values = list(na.omit(c_across(all_of(hem_cols)))),  # all non-NA Hem species
    
    # Determine if IHH_species matches any Hem_species_ fuzzily
    ihh_matches_hem = if(!is.na(IHH_species) & length(hem_values) > 0) {
      any(stringdist::stringdist(IHH_species, hem_values, method = "jw") < 0.15)
    } else {
      FALSE
    },
    
    # Main estimation logic
    estimated_species_name = case_when(
      
      scientific_name_recorded_T.F ~ recorded_scientific_name,  # Case 1
      
      !is.na(IHH_species) & length(hem_values) > 0 & ihh_matches_hem ~ IHH_species,  # Case 2
      
      !is.na(IHH_species) & length(hem_values) > 0 & !ihh_matches_hem ~ IHH_species, # Case 3
      
      !is.na(IHH_species) & length(hem_values) == 0 ~ IHH_species,                   # Case 4
      
      is.na(IHH_species) & length(hem_values) == 1 ~ hem_values[1],                  # Case 5a
      
      is.na(IHH_species) & length(hem_values) > 1 ~ "unknown",                       # Case 5b
      
      TRUE ~ "unknown"
    ),
    
    species_est_certainty = case_when(
      scientific_name_recorded_T.F ~ "virtually certain",       # Case 1
      !is.na(IHH_species) & length(hem_values) > 0 & ihh_matches_hem ~ "high confidence",    # Case 2
      !is.na(IHH_species) & length(hem_values) > 0 & !ihh_matches_hem ~ "low confidence",    # Case 3
      !is.na(IHH_species) & length(hem_values) == 0 ~ "high confidence",                     # Case 4
      is.na(IHH_species) & length(hem_values) == 1 ~ "high confidence",                       # Case 5a
      is.na(IHH_species) & length(hem_values) > 1 ~ "medium confidence",                       # Case 5b
      TRUE ~ "not applicable"
    ),
    
    species_est_reasoning = case_when(
      scientific_name_recorded_T.F ~ "name was recorded in CONAPESCA",
      !is.na(IHH_species) & length(hem_values) > 0 & ihh_matches_hem ~ "logical match in both keys",
      !is.na(IHH_species) & length(hem_values) > 0 & !ihh_matches_hem ~ "mismatch between keys",
      !is.na(IHH_species) & length(hem_values) == 0 ~ "match to IHH key",
      is.na(IHH_species) & length(hem_values) == 1 ~ "match to Hem's key",
      is.na(IHH_species) & length(hem_values) > 1 ~ "conflicting matches in Hem's key and no data from IHH",
      TRUE ~ "no data in keys"
    )
  ) %>%
  ungroup() %>%
  select(-hem_values, -ihh_matches_hem)  # remove temporary helper columns
```

```{r A few summary stats}
sum(is.na(Master_SpKey_Cleaned$estimated_species_name)) # 0 rows without a species estimate
sum(Master_SpKey_Cleaned$estimated_species_name=="unknown") # 241 species unknown

sum(is.na(Master_SpKey_Cleaned$species_est_certainty)) # 0 rows without a species certainty estimate
table(Master_SpKey_Cleaned$species_est_certainty) # 212 where the estimate is not applicable, 168 estimates of low confidence

sum(is.na(Master_SpKey_Cleaned$species_est_reasoning)) # 0 rows without a reasoning
table(Master_SpKey_Cleaned$species_est_reasoning) # 212 where we have no data to draw conclusions from, 168 estimates where there was a mismatch between keys
```

```{r Let's review those that are uncertain and clarify them if possible}
Master_SpKey_Unknowns<- Master_SpKey_Cleaned %>% filter(species_est_certainty== "not applicable")

# Okay, a few of these unknowns can probably be resolved with my key, which I would think is slightly less reliable than Mar's and Hem's. Use my key to resolve these unknowns and figure out the rest later. (NOTE TO SELF)

# Prepare my key for use:
Erica_SpKey_filtered<-  Erica_SpKey %>%
  filter(is.na(recorded_scientific_name)) %>%
  mutate(recorded_species_name_clean = tolower(recorded_species_name_clean)) %>%
  rename(Erica_common_name = recorded_species_name_clean, recorded_taxa_group = main_species_group)

Erica_SpKey_filtered<- Erica_SpKey_filtered %>% select(recorded_taxa_group, Erica_common_name, species_estimated, genus_estimated, family_estimated, source_of_estimate) %>% distinct
# 222 obs., 6 vars.

# Control against many-to-many matches by getting rid of rows in my key that are nearly identical
Erica_SpKey_filtered <- Erica_SpKey_filtered %>%
  group_by(recorded_taxa_group, Erica_common_name) %>%
  slice(1) %>% # keep first match
  ungroup()

# Only add in my key info when IHH and Hem key info is missing: 
Master_SpKey_Cleaned <- Master_SpKey_Cleaned %>%
  # Separate unknowns from knowns
  mutate(lookup_needed = estimated_species_name == "unknown") %>%
  
  # Left join only where lookup_needed is TRUE
  left_join(
    Erica_SpKey_filtered,
    by = c("recorded_taxa_group", "common_name_clean" = "Erica_common_name")
  ) %>%
  
  # Replace estimated_species_name only where it was unknown but a match exists
  mutate(
    estimated_species_name = if_else(
      lookup_needed & !is.na(species_estimated),
      species_estimated,
      estimated_species_name
    ),
    # Optionally, update the certainty / reasoning
    species_est_certainty = if_else(
      lookup_needed & !is.na(species_estimated),
      "medium confidence",
      species_est_certainty
    ),
    species_est_reasoning = if_else(
      lookup_needed & !is.na(species_estimated),
      "match to Erica's key",
      species_est_reasoning
    )
  ) %>%
  
  # Remove temporary and unnecessary columns
  select(-lookup_needed, -species_estimated)

# 802 obs. as expected.
```

```{r New summary stats}
sum(Master_SpKey_Cleaned$estimated_species_name=="unknown") # 241 -> 228 species unknown

table(Master_SpKey_Cleaned$species_est_certainty) # 212 -> 202 where the estimate is not applicable, 168 estimates of low confidence

table(Master_SpKey_Cleaned$species_est_reasoning) # 202 where we have no data to draw conclusions from, 168 estimates where there was a mismatch between keys
```

```{r Now extract information about genus and family, where possible}
# WHAT IS THE LOGIC?

# If species is known:
# → estimated_genus_name = first word of species

# If species unknown but IHH_genus available:
# → estimated_genus_name = IHH_genus

# If both unavailable:
# → estimated_genus_name = genus unknown

# If IHH_family exists:
# → estimated_family_name = IHH_family

# Else if Erica_family_estimated exists:
# → estimated_family_name = Erica_family_estimated

# Else if Erica_SpKey gives exactly one family per larger taxonomic grouping
# → estimated_family_name = Erica_family_estimated
# → reasoning: “unique family match from Erica's key”

# Else
# → estimated_family_name = family unknown


# APPLY THE LOGIC:
# --- Build the Erica single-family lookup -------------------------------------

Erica_SpKey_simple <- Erica_SpKey %>% 
  select(
    recorded_taxa_group = main_species_group, 
    Erica_family_estimated = family_estimated
  ) %>%
  distinct() %>%
  group_by(recorded_taxa_group) %>%
  filter(n_distinct(Erica_family_estimated) == 1) %>%
  ungroup()


Master_SpKey_Cleaned <- Master_SpKey_Cleaned %>%
  
  # ----------------------- GENUS LOGIC ----------------------------------------
  mutate(
    estimated_genus_name = case_when(
      !is.na(estimated_species_name) & estimated_species_name != "unknown" ~ 
        word(estimated_species_name, 1),             # species known → genus = first word
      !is.na(IHH_genus) ~ IHH_genus,                 # fallback: IHH genus
      TRUE ~ NA_character_                           # final fallback: genus unknown
    ),
    
    genus_est_reasoning = case_when(
      !is.na(estimated_species_name) & estimated_species_name != "unknown" ~
        "genus derived from known species",
      !is.na(IHH_genus) ~
        "genus derived from IHH key",
      TRUE ~
        "genus unknown"
    )
  ) %>%
  
  
  # ----------------------- FAMILY LOGIC ---------------------------------------
  left_join(Erica_SpKey_simple, by = "recorded_taxa_group") %>%
  mutate(
    estimated_family_name = case_when(
      !is.na(IHH_family) ~ IHH_family,                         # 1. direct from IHH
      !is.na(Erica_family_estimated) ~ Erica_family_estimated, # 2. unique Erica match
      !is.na(estimated_genus_name) ~ 
        "family missing but genus available",                 # 3. genus known but no family
      TRUE ~ NA_character_                                     # 4. completely unknown
    ),
    
    family_est_reasoning = case_when(
      !is.na(IHH_family) ~ "family derived from IHH key",
      !is.na(Erica_family_estimated) ~ "unique family match from Erica's key",
      !is.na(estimated_genus_name) ~ "family unknown but genus available",
      TRUE ~ "family unknown"
    )
  ) %>%
  
  # remove columns that aren't necessary
  select(-Erica_family_estimated, -genus_estimated, -family_estimated, -source_of_estimate)
```

```{r final summary stats}
table(TEST2$species_est_reasoning) # 202 no data
table(TEST2$genus_est_reasoning) # 228 genus unknown
table(TEST2$family_est_reasoning) # 213 family unknown
```

```{r Clean MASTER SPECIES KEY}
write.csv(Master_SpKey_Cleaned, "/home/ferrer/ohi_goc_prep/general/Species_Keys/CleanedMaster_SpKey_Dec012025.csv", row.names = FALSE)
```

# PART 3 COMPLETE (for now).

# PART 4: Merge the cleaned master sp key with landings data ------------------------------
```{r Load data frames if needed}
Master_SpKey_Cleaned <- read_csv("/home/ferrer/ohi_goc_prep/general/Species_Keys/CleanedMaster_SpKey_Dec012025.csv")
# 802 obs of 41 variables

Landings_GoC_Cleaned<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fis/v2025/int/juan_carlos/GoC_Landings_wKeyWords_2000.present_Dec012025.rds"))
# n = 3716415 obs.
```

```{r Shave down the size of Master_SpKey_filtered}
Master_SpKey_filtered<- Master_SpKey_Cleaned %>% 
  select(recorded_taxa_group, recorded_taxa_name_clean, common_name_clean,
         estimated_species_name, species_est_certainty, estimated_genus_name, estimated_family_name,
         IHH_order, IHH_class, IHH_functional_major_groups_sm, IHH_fishery)
# 802 obs., 11 vars.
```

```{r Merge landings with filtered sp key}
Landings_GoC_Cleaned_wSp <- Landings_GoC_Cleaned %>%
  left_join(
    Master_SpKey_filtered,
    by = c(
      "recorded_taxa_group",
      "recorded_taxa_name_clean",
      "common_name_clean"
    )
  )
# 3716415 obs. as expected, 41 vars.
```

## Save Landings_GoC_Cleaned_wSp rds file
```{r Save this dataframe - landings keywords & species for top groupings: Nov 10 2025}
saveRDS(Landings_GoC_Cleaned_wSp, file="/home/shares/ohi/OHI_GOC/goal_prep/fis/v2025/int/juan_carlos/GoC_Landings_wSp_2000.present_Dec012025.rds")
```