---
title: 'OHI GoC 2025: Economies'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# ECO: Revenue from Marine-Related Sectors in the Censos Economicos (CE)

```{r}
## for plotting:
rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

region_levels <- paste0("Region_", 1:9)
```

## Save subset of *revenue* metrics

Reminder:
* The number of jobs, *H000A*;
* The employer contributions to social security schemes *J300A*,
* Other social benefits *J400A*;
* Total revenue from the supply of goods and services (millions of pesos) *M000A*; and
* Total revenue (millions of pesos) *A800A*

I decided to use M000A (total revenue from the supply of goods and services) because it most directly captures the market value of production generated by each economic activity, excluding financial income, extraordinary gains, and other non-operating revenues that can introduce volatility and reduce comparability across census waves. For the purpose of measuring the size and performance of marine-related economic sectors using data from the Instituto Nacional de Estadística y Geografía Censos Económicos (INEGI), operating revenue is a cleaner proxy for gross output and better reflects real economic activity within the production boundary. I adjust these values to constant 2023 pesos to remove the effects of inflation and ensure that differences across 2003, 2008, 2013, 2018, and 2023 represent real changes in economic activity rather than changes in the general price level. Using constant pesos preserves internal consistency for regional comparisons within Mexico without introducing unnecessary distortions from exchange rates or purchasing power parity conversions, which are only required for cross-country analysis.

Operating revenue, which captures the value of goods and services produced and sold, is conceptually aligned with gross output measures used in sectoral economic analysis (see U.S. Bureau of Labor Statistics on output choice). To make meaningful comparisons over time, nominal financial values must be converted to constant prices using price indices; this reflects common practice in national accounts and applied economic measurement (see World Bank constant price methodology and Office for National Statistics on deflators).

U.S. Bureau of Labor Statistics:<https://www.bls.gov/opub/mlr/2023/article/the-importance-of-output-choice.htm>
World Bank constant price methodology:<https://datahelpdesk.worldbank.org/knowledgebase/articles/114968-how-do-you-derive-your-constant-price-series-for-t>
Office for National Statistics:<https://www.ons.gov.uk/economy/inflationandpriceindices/methodologies/deflatorsandhowweusethemineconomicestimates>

```{r}
ce_cleaned_totals_per_yr_municipio_activity <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/ce_int","ce_cleaned_totals_per_yr_municipio_activity.csv")) %>% 
  filter(!(municipio == "MEXICALI" & ano_censal == "2023"))

ce_revenue_totals <- ce_cleaned_totals_per_yr_municipio_activity %>% 
  dplyr::select(-c("j400a","j300a")) # leave in number of jobs (h000a) so that we can scale to the number of jobs.
```

## Adjust for inflation

This will be done using the package `priceR`.  Documentation is located here: <https://cran.r-project.org/web/packages/priceR/readme/README.html> 

"adjust_for_inflation() automatically converts between nominal and real dollars, or in/deflates prices from one year’s prices to another’s.  It works for 304 countries / areas (see them with all by running show_countries())."

```{r}
library(priceR)

ce_revenue_totals_adj <- ce_revenue_totals %>%
  mutate(m000a_pesos = m000a * 1e6) %>% 
  mutate(
    revenue_real_pesos_2023 = adjust_for_inflation(
      price = m000a_pesos,
      from_date = ano_censal,
      country = "Mexico",
      to_date = 2023
      ),
    revenue_real_milpesos_2023 = adjust_for_inflation(
      price = m000a,
      from_date = ano_censal,
      country = "Mexico",
      to_date = 2023
      )
  ) %>% 
  dplyr::select(ano_censal, rgn_id, estado, e03, municipio, e04, codigo, general_sector, specific_sector, m000a, m000a_pesos, revenue_real_pesos_2023, revenue_real_milpesos_2023, h000a, ue)
```

Aggregating instead by region and general sector:
```{r}
total_rev_per_rgn_and_activity <- ce_revenue_totals_adj %>% 
  group_by(ano_censal, rgn_id, general_sector, specific_sector) %>% # not using codigo, notas_breve, because those are too specific
  summarize(rev_real2023_per_act = sum(revenue_real_pesos_2023, na.rm = TRUE),
            jobs_per_act = sum(h000a, na.rm = TRUE), 
            ue_per_act = sum(ue, na.rm = TRUE))
```


### Add in adjustment for municipalities that cover both coastal and pacific sides, and sectors that are not strictly coastal

```{r}
coastal_adjustment_multipliers <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/le","coastal_adjustment_multiplier_2020.csv"))
```

```{r}
coastal_total_rev_per_rgn_and_activity <- total_rev_per_rgn_and_activity %>% 
  full_join(coastal_population_multipliers, by = "rgn_id") %>% 
  mutate(
    
      # fix San Felipe
    coastal_pop_adjustment = ifelse(rgn_id %in% "Region_1" & ano_censal %in% "2023", 1, coastal_pop_adjustment),
    
    # rev
    first_adj_coastal_rev_per_act = ifelse(rgn_id %in% c("Region_3","Region_4"), rev_real2023_per_act * coastal_pop_adjustment, rev_real2023_per_act),
    coastal_rev_per_act = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), rev_real2023_per_act * coastal_pop_adjustment, first_adj_coastal_rev_per_act),
    
    # ue
    first_adj_coastal_ue_per_act = ifelse(rgn_id %in% c("Region_3","Region_4"), ue_per_act * coastal_pop_adjustment, ue_per_act),
    coastal_ue_per_act = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), ue_per_act * coastal_pop_adjustment, first_adj_coastal_ue_per_act)
  ) %>%  # if in any of these sectors, adjust for 50km inland
  
  dplyr::select(-c(jobs_per_act, rev_real2023_per_act, ue_per_act, first_adj_coastal_rev_per_act, first_adj_coastal_ue_per_act))
```

## Aggregate real 2023 revenue by region

Aggregating the revenue for each region:
```{r}
total_rev_per_rgn <- coastal_total_rev_per_rgn_and_activity %>% 
  group_by(ano_censal, rgn_id) %>% 
  summarize(total_rev = sum(coastal_rev_per_act, na.rm = TRUE),
            total_ue = sum(coastal_ue_per_act, na.rm = TRUE)) %>% 
  mutate(total_rev = round(total_rev),
         total_ue = round(total_ue)) %>% 
  ungroup
```

```{r}
## plot num jobs current status -- NOT CAPPED
rev_raw <- plotly::plot_ly(
  total_rev_per_rgn,
  x = ~ano_censal,
  y = ~total_rev,
  color = ~factor(rgn_id),
  colors = rgn_colors,
  type = "scatter",
  mode = "lines"
) %>%
  layout(
    title = "Total marine-related revenue per region from 2003-2013 (Constant 2023 Pesos)",
    xaxis = list(title = "Census Year"),
    yaxis = list(
      title = "Marine-related Revenue 2003-2013 "
    ),
    legend = list(title = list(text = "Region"))
  )
rev_raw
```


## Reference point (average value 2003-2013)

By region:
```{r}
historical_rev_ref <- total_rev_per_rgn %>% 
  filter(ano_censal %in% c("2003","2008","2013")) %>% 
  group_by(rgn_id) %>%  
  summarize(avg_value_2003_2013 = mean(total_rev))
```


## Current status

First, join back the reference point, then divide the current value of marine revenue by the historical average from 2003 - 2013 as a reference.

If it is equal to the historical average, or above, the region will receive the current status of 100.  Otherwise, it will reflect the reduction.

### By region

```{r}
ce_rev_current_status <- total_rev_per_rgn %>% 
  dplyr::select(rgn_id, ano_censal, total_rev) %>% 
  left_join(historical_rev_ref) %>% 
  mutate(current_status_raw = total_rev/avg_value_2003_2013 * 100,
         current_status = ifelse(current_status_raw >= 100, 100, current_status_raw)) %>% 
  ungroup()

write_csv(ce_rev_current_status, here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/eco","ce_rev_current_status.csv"))
```

#### plot 

```{r}
library(plotly)

## plot num jobs current status -- NOT CAPPED
rev_cs_plotly_noCAP <- plotly::plot_ly(
  ce_rev_current_status,
  x = ~ano_censal,
  y = ~current_status_raw,
  color = ~factor(rgn_id),
  colors = rgn_colors,
  type = "scatter",
  mode = "lines"
) %>%
  layout(
    title = "RAW CURRENT STATUS of marine-related revenue, reference of average value 2003-2013 (Constant 2023 Pesos)",
    xaxis = list(title = "Census Year"),
    yaxis = list(
      title = "Revenue / 2003-2013 Average Revenue (%)"
    ),
    legend = list(title = list(text = "Region"))
  ) %>%
  add_lines(
    x = range(ce_rev_current_status$ano_censal), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
rev_cs_plotly_noCAP
```


```{r}
## revenue current status by region
rev_cs_plotly <- plotly::plot_ly(
  ce_rev_current_status,
  x = ~ano_censal,
  y = ~current_status,
  color = ~factor(rgn_id),
  colors = rgn_colors,
  type = "scatter",
  mode = "lines"
) %>%
  layout(
    title = "CURRENT STATUS of marine-related revenue, reference of average value 2003-2013 (Constant 2023 Pesos)",
    xaxis = list(title = "Census Year"),
    yaxis = list(
      title = "Revenue / 2003-2013 Average Revenue (%)",
      range = c(0, 110)
    ),
    legend = list(title = list(text = "Region"))
  ) %>%
  add_lines(
    x = range(ce_rev_current_status$ano_censal), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )

rev_cs_plotly
```

