---
title: 'OHI GoC 2025: Economies'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# ECO: Revenue from Marine-Related Sectors in the Censos Economicos (CE)

## Save subset of *revenue* metrics

Reminder:
* The number of jobs, *H000A*;
* The employer contributions to social security schemes *J300A*,
* Other social benefits *J400A*;
* Total revenue from the supply of goods and services (millions of pesos) *M000A*; and
* Total revenue (millions of pesos) *A800A*

I decided to use M000A (total revenue from the supply of goods and services) because it most directly captures the market value of production generated by each economic activity, excluding financial income, extraordinary gains, and other non-operating revenues that can introduce volatility and reduce comparability across census waves. For the purpose of measuring the size and performance of marine-related economic sectors using data from the Instituto Nacional de Estadística y Geografía Censos Económicos (INEGI), operating revenue is a cleaner proxy for gross output and better reflects real economic activity within the production boundary. I adjust these values to constant 2024 pesos to remove the effects of inflation and ensure that differences across 2004, 2009, 2014, 2019, and 2024 represent real changes in economic activity rather than changes in the general price level. Using constant pesos preserves internal consistency for regional comparisons within Mexico without introducing unnecessary distortions from exchange rates or purchasing power parity conversions, which are only required for cross-country analysis.

Operating revenue, which captures the value of goods and services produced and sold, is conceptually aligned with gross output measures used in sectoral economic analysis (see U.S. Bureau of Labor Statistics on output choice). To make meaningful comparisons over time, nominal financial values must be converted to constant prices using price indices; this reflects common practice in national accounts and applied economic measurement (see World Bank constant price methodology and Office for National Statistics on deflators).

U.S. Bureau of Labor Statistics:<https://www.bls.gov/opub/mlr/2023/article/the-importance-of-output-choice.htm>
World Bank constant price methodology:<https://datahelpdesk.worldbank.org/knowledgebase/articles/114968-how-do-you-derive-your-constant-price-series-for-t>
Office for National Statistics:<https://www.ons.gov.uk/economy/inflationandpriceindices/methodologies/deflatorsandhowweusethemineconomicestimates>

```{r}
ce_cleaned_totals_per_yr_municipio_activity <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/ce_int","ce_cleaned_totals_per_yr_municipio_activity.csv")) %>% 
  filter(!(municipio == "ENSENADA" & ano_censal == "2024"))

ce_revenue_totals <- ce_cleaned_totals_per_yr_municipio_activity %>% 
  select(-c("j400a","j300a")) # leave in number of jobs (h000a) so that we can scale to the number of jobs.
```

## Adjust for inflation

This will be done using the package `priceR`.  Documentation is located here: <https://cran.r-project.org/web/packages/priceR/readme/README.html> 

"adjust_for_inflation() automatically converts between nominal and real dollars, or in/deflates prices from one year’s prices to another’s.  It works for 304 countries / areas (see them with all by running show_countries())."

```{r}
library(priceR)

ce_revenue_totals_adj <- ce_revenue_totals %>%
  mutate(m000a_pesos = m000a * 1e6) %>% 
  mutate(
    revenue_real_pesos_2024 = adjust_for_inflation(
      price = m000a_pesos,
      from_date = ano_censal,
      country = "Mexico",
      to_date = 2024
      ),
    revenue_real_milpesos_2024 = adjust_for_inflation(
      price = m000a,
      from_date = ano_censal,
      country = "Mexico",
      to_date = 2024
      )
  ) %>% 
  select(ano_censal, rgn_id, estado, e03, municipio, e04, codigo, sector_marino_propuesto, notas_breve, m000a, m000a_pesos, revenue_real_pesos_2024, revenue_real_milpesos_2024, h000a, ue)
```

## Aggregate by region and activity

### By region

```{r}
total_revenue_byrgn <- ce_revenue_totals_adj %>% 
  group_by(ano_censal, rgn_id) %>% 
  summarize(total_rev_adj = sum(revenue_real_pesos_2024, na.rm = TRUE),
            total_jobs = sum(h000a, na.rm = TRUE),
            total_ue = sum(ue, na.rm = TRUE))
```


### By sector

```{r}
total_revenue_byrgn_and_activity <- ce_revenue_totals_adj %>% 
  group_by(ano_censal, rgn_id, sector_marino_propuesto) %>% # not using codigo, notas_breve, because those are too specific
  summarize(total_rev_adj = sum(revenue_real_pesos_2024, na.rm = TRUE),
            total_jobs = sum(h000a, na.rm = TRUE),
            total_ue = sum(ue, na.rm = TRUE))
```


## Scale to the number of jobs

By region:
```{r}
ce_revenue_adj_perjob <- total_revenue_byrgn %>% 
  mutate(rev_per_job = total_rev_adj/total_jobs)
```

By region and sector:
```{r}
ce_revenue_sector_adj_perjob <- total_revenue_byrgn_and_activity %>% 
  mutate(rev_per_job = total_rev_adj/total_jobs)
```


## Reference point (average value 2004-2014)

By region:
```{r}
historical_rev_scaled_to_jobs <- ce_revenue_adj_perjob %>% 
  filter(ano_censal %in% c("2004","2009","2014")) %>% 
  group_by(rgn_id) %>%  
  summarize(avg_value_2004_2014 = mean(rev_per_job))
```

By region and sector:
```{r}
historical_rev_sector_scaled_to_jobs <- ce_revenue_sector_adj_perjob %>% 
  filter(ano_censal %in% c("2004","2009","2014")) %>% 
  group_by(rgn_id, sector_marino_propuesto) %>%  
  summarize(avg_value_2004_2014 = mean(rev_per_job))
```

## Current status

First, join back the reference point, then divide the current value of marine jobs/pop by the historical average from 2004 - 2014 as a reference.

If it is equal to the historical average, or above, the region will receive the current status of 100.  Otherwise, it will reflect the reduction.

By region:
```{r}
ce_rev_current_status <- ce_revenue_adj_perjob %>% 
  select(rgn_id, ano_censal, rev_per_job) %>% 
  left_join(historical_rev_scaled_to_jobs) %>% 
  mutate(current_status_raw = rev_per_job/avg_value_2004_2014 * 100,
         current_status = ifelse(current_status_raw >= 100, 100, current_status_raw)) %>% 
  ungroup()
```

By region and sector:
```{r}
ce_rev_sector_current_status <- ce_revenue_sector_adj_perjob %>% 
  select(rgn_id, ano_censal, sector_marino_propuesto, rev_per_job) %>% 
  left_join(historical_rev_sector_scaled_to_jobs) %>% 
  mutate(current_status_raw = rev_per_job/avg_value_2004_2014 * 100,
         current_status = ifelse(current_status_raw >= 100, 100, current_status_raw)) %>% 
  ungroup()

## now, average into one score per region:
ce_rev_sector_current_status_final <- ce_rev_sector_current_status %>% 
  group_by(ano_censal, rgn_id) %>% 
  summarize(avg_cs = mean(current_status, na.rm = TRUE)) %>% 
  ungroup
```

### Plotlys 

#### By region

```{r}
library(plotly)

## Plot revenue current status by region
rev_cs_plotly <- plotly::plot_ly(
  ce_rev_current_status,
  x = ~ano_censal,
  y = ~current_status,
  color = ~factor(rgn_id),
  type = "scatter",
  mode = "lines+markers"
) %>%
  layout(
    title = "Current Status of Marine Revenue per Job by Region (Constant 2024 Pesos)",
    xaxis = list(title = "Census Year"),
    yaxis = list(
      title = "Revenue per Job / 2004-2014 Average (%)",
      range = c(0, 110)
    ),
    legend = list(title = list(text = "Region"))
  )

rev_cs_plotly

```


#### By region and sector

```{r}
library(plotly)

## Plot revenue current status by region
rev_cs_sector_plotly <- plotly::plot_ly(
  ce_rev_sector_current_status_final,
  x = ~ano_censal,
  y = ~avg_cs,
  color = ~factor(rgn_id),
  type = "scatter",
  mode = "lines+markers"
) %>%
  layout(
    title = "Current Status of Marine Revenue per Job by Region and Averaged Across Sector (Constant 2024 Pesos)",
    xaxis = list(title = "Census Year"),
    yaxis = list(
      title = "Revenue per Job / 2004-2014 Average (%)",
      range = c(0, 110)
    ),
    legend = list(title = list(text = "Region"))
  )

rev_cs_sector_plotly
```


```{r}
##### for FACETED, before avg across sectors:
library(plotly)
library(dplyr)

# Make sure year is numeric for proper ordering
ce_rev_sector_current_status <- ce_rev_sector_current_status %>%
  mutate(ano_censal = as.numeric(ano_censal))

# Split data by region
region_list <- split(ce_rev_sector_current_status, 
                     ce_rev_sector_current_status$rgn_id)

# Create one plot per region
plots <- lapply(names(region_list), function(rgn) {
  
  plot_ly(
    region_list[[rgn]],
    x = ~ano_censal,
    y = ~current_status,
    color = ~sector_marino_propuesto,
    type = "scatter",
    mode = "lines+markers"
  ) %>%
    layout(
      title = paste("Region", rgn),
      xaxis = list(title = "Census Year"),
      yaxis = list(range = c(0, 110))
    )
})

# Combine into subplot
rev_sector_cs_plotly <- subplot(
  plots,
  nrows = length(plots),
  shareX = TRUE,
  titleY = TRUE
)

rev_sector_cs_plotly

```