---
title: "Habitat Services Scores - OHI GOC"
format: 
  html:
    code-fold: show
execute:
  echo: true
  message: false
  warning: false
---


################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

Author of original draft: Austin Martinez

################################################################################


```{r Load in Libraries}
# Load in Libraries
library(sf)
library(dplyr)
library(here)
library(terra)
library(tidyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(tmap)
library(stringr)
library(readr)
```

```{r Read in OHI region boundaries}
# Read in OHI region boundaries
ohi_regions <- st_read(here("spatial", "ohi_regions", "goc_ohi_rgns.shp")) %>% st_make_valid()
```
## Read in all habitat data
```{r Read in habitat services data}
# Read in habitat services data
habitat_services_full <- read_csv(here::here("/home/ferrer/ohi_goc_prep/HS/v2026/HS_Scores_1.1/HS_components_df_Sep112026.csv"))
```

# Calculating the HS_current_status:

```{r Calculate HS_current_status sub-scores}
colnames(habitat_services_full)

# Calculate carbon, filtration, and protection score
HS_current_status <- habitat_services_full %>%
  group_by(rgn_id) %>%
  summarise(
    PRO_score = sum(extent * coastal_protection * current_status, na.rm = TRUE) /
                    sum(extent * coastal_protection, na.rm = TRUE),

    FIL_score = sum(extent * filtration * current_status, na.rm = TRUE) /
                       sum(extent * filtration, na.rm = TRUE),

    SEQ_score = sum(extent * carbon_sequestration_gC.m2.yr * current_status, na.rm = TRUE) /
                    sum(extent * carbon_sequestration_gC.m2.yr, na.rm = TRUE)
  )

HS_current_status
```

```{r Calculate HS_current_status based on sub-scores}
# Average the 3 scores to get final OHI score
HS_current_status <- HS_current_status %>%
  mutate(HS_score = (SEQ_score + FIL_score + PRO_score) / 3)

HS_current_status
```

```{r Fill out the HS_current_status for all years 2020 through 2025 which we have data}
HS_current_status_1.1 <- HS_current_status %>%
  crossing(year = 2020:2025) %>%
  arrange(rgn_id, year)

# NOTE TO SELF: I made this dataframe (Sep 11 2026). This will change once we settle the debate on trend. - EMF

```


```{r Save HS_current_status}
write.csv(HS_current_status_1.1,
          "/home/ferrer/ohi_goc_prep/HS/v2026/HS_Scores_1.1/HS_Scores_df_Sep112026.csv",
          row.names = FALSE)
```

# PLOT HS SCORES ------------------------------
```{r Load in HS_current_status_1.1 as needed}
HS_current_status_1.1<- read.csv("/home/ferrer/ohi_goc_prep/HS/v2026/HS_Scores_1.1/HS_Scores_df_Sep112026.csv")
```

```{r Region key}
Region_Key0<- read.csv("/home/ferrer/ohi_goc_prep/FP/v2026/CONAPESCA_Keys/Spatial_Keys/RegionKey_FINAL_May262026.csv")

colnames(Region_Key0)

Region_Key <- Region_Key0 %>%
  filter(GOC_Include_Exclude == "include") %>%
  select(OHI_region_est, OHI_region_name) %>%
  mutate(
    OHI_region_est = str_to_lower(OHI_region_est)
  ) %>%
  distinct()
```

```{r Merge Region_Key to HS_current_status_1.1}
# Make rgn_id lowercase for merging purposes:
HS_current_status_1.1 <- HS_current_status_1.1 %>%
  mutate(
    rgn_id = str_to_lower(rgn_id)
  )

# Merge:
HS_current_status_1.1 <- HS_current_status_1.1 %>%
  left_join(
    Region_Key,
    by = c("rgn_id" = "OHI_region_est")
  )

# Make sure to order the regions:
Region_order <- c(
  "Western Upper Gulf",
  "Western Midriff Islands",
  "Western Central Gulf",
  "Western Southern Gulf",
  "Eastern Upper Gulf",
  "Eastern Midriff Islands",
  "Southern Sonora Coast",
  "Sinaloa Coast",
  "Southeastern Gulf"
)

HS_current_status_1.1 <- HS_current_status_1.1 %>%
  mutate(
    OHI_region_name = factor(
      OHI_region_name,
      levels = Region_order
    )
  )
```

```{r Plot HS_current_status}
# Region colors
Region_colors<- c(
  "Western Upper Gulf" = "#C03729",
  "Western Midriff Islands" = "#E68C7C",
  "Western Central Gulf" = "#FC8F24",
  "Western Southern Gulf" = "#dcaa38",
  "Eastern Upper Gulf" = "#cbd2a0",
  "Eastern Midriff Islands" = "#98a25a",
  "Southern Sonora Coast" = "#747428",
  "Sinaloa Coast" = "#627391",
  "Southeastern Gulf" = "#8d61a9")

HS_Scores_OverTime<- HS_current_status_1.1 %>%
  filter(!is.na(OHI_region_name)) %>%
  
  ggplot(aes(
    x = year,
    y = HS_score,
    color = OHI_region_name
  )) +
  
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  
  facet_wrap(~ OHI_region_name, scales = "fixed") +
  
  labs(
    title = "HS Scores by Region",
    subtitle = "Composite of PRO, FIL, and SEQ sub-scores",
    caption = "updated: Sep 30 2026",
    y = "HS Score (from 0 to 100)",
    x = "Year"
  ) +
  
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey50") +
  
  ylim(0, 101) +
  
  scale_color_manual(values = Region_colors) +
  
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 16, face = "bold")
  )

HS_Scores_OverTime

```

```{r Save as PNG}
ggsave(
  filename = sprintf("/home/ferrer/ohi_goc_prep/HS/v2026/HS_Scores_1.1/HS_Scores_1.1_plot_Sep302026.png"),
  plot = HS_Scores_OverTime,
  width = 12,
  height = 12,
  dpi = 300
)
```

# PLOT HS SUB-SCORES (PRO, FIL, SEQ) ------------------------------

```{r Plot PRO_Scores_OverTime}
# Region colors
Region_colors<- c(
  "Western Upper Gulf" = "#C03729",
  "Western Midriff Islands" = "#E68C7C",
  "Western Central Gulf" = "#FC8F24",
  "Western Southern Gulf" = "#dcaa38",
  "Eastern Upper Gulf" = "#cbd2a0",
  "Eastern Midriff Islands" = "#98a25a",
  "Southern Sonora Coast" = "#747428",
  "Sinaloa Coast" = "#627391",
  "Southeastern Gulf" = "#8d61a9")

PRO_Scores_OverTime<- HS_current_status_1.1 %>%
  filter(!is.na(OHI_region_name)) %>%
  
  ggplot(aes(
    x = year,
    y = PRO_score,
    color = OHI_region_name
  )) +
  
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  
  facet_wrap(~ OHI_region_name, scales = "fixed") +
  
  labs(
    title = "Coastal Protection (PRO) Scores by Region",
    caption = "updated: Sep 30 2026",
    y = "PRO Score (from 0 to 100)",
    x = "Year"
  ) +
  
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey50") +
  
  ylim(0, 101) +
  
  scale_color_manual(values = Region_colors) +
  
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 16, face = "bold")
  )

PRO_Scores_OverTime

```

```{r Plot FIL_Scores_OverTime}
# Region colors
Region_colors<- c(
  "Western Upper Gulf" = "#C03729",
  "Western Midriff Islands" = "#E68C7C",
  "Western Central Gulf" = "#FC8F24",
  "Western Southern Gulf" = "#dcaa38",
  "Eastern Upper Gulf" = "#cbd2a0",
  "Eastern Midriff Islands" = "#98a25a",
  "Southern Sonora Coast" = "#747428",
  "Sinaloa Coast" = "#627391",
  "Southeastern Gulf" = "#8d61a9")

FIL_Scores_OverTime<- HS_current_status_1.1 %>%
  filter(!is.na(OHI_region_name)) %>%
  
  ggplot(aes(
    x = year,
    y = FIL_score,
    color = OHI_region_name
  )) +
  
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  
  facet_wrap(~ OHI_region_name, scales = "fixed") +
  
  labs(
    title = "Pollution Filtration (FIL) Scores by Region",
    caption = "updated: Sep 30 2026",
    y = "FIL Score (from 0 to 100)",
    x = "Year"
  ) +
  
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey50") +
  
  ylim(0, 101) +
  
  scale_color_manual(values = Region_colors) +
  
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 16, face = "bold")
  )

FIL_Scores_OverTime

```

```{r Plot SEQ_Scores_OverTime}
# Region colors
Region_colors<- c(
  "Western Upper Gulf" = "#C03729",
  "Western Midriff Islands" = "#E68C7C",
  "Western Central Gulf" = "#FC8F24",
  "Western Southern Gulf" = "#dcaa38",
  "Eastern Upper Gulf" = "#cbd2a0",
  "Eastern Midriff Islands" = "#98a25a",
  "Southern Sonora Coast" = "#747428",
  "Sinaloa Coast" = "#627391",
  "Southeastern Gulf" = "#8d61a9")

SEQ_Scores_OverTime<- HS_current_status_1.1 %>%
  filter(!is.na(OHI_region_name)) %>%
  
  ggplot(aes(
    x = year,
    y = SEQ_score,
    color = OHI_region_name
  )) +
  
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  
  facet_wrap(~ OHI_region_name, scales = "fixed") +
  
  labs(
    title = "Carbon Sequestration (SEQ) Scores by Region",
    caption = "updated: Sep 30 2026",
    y = "SEQ Score (from 0 to 100)",
    x = "Year"
  ) +
  
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey50") +
  
  ylim(0, 101) +
  
  scale_color_manual(values = Region_colors) +
  
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 16, face = "bold")
  )

SEQ_Scores_OverTime

```

```{r Save as PNG}
# Coastal Protection (PRO)
ggsave(
  filename = sprintf("/home/ferrer/ohi_goc_prep/HS/v2026/HS_Scores_1.1/PRO_Scores_1.1_plot_Sep302026.png"),
  plot = PRO_Scores_OverTime,
  width = 12,
  height = 12,
  dpi = 300
)

# Pollution Filtration (FIL)
ggsave(
  filename = sprintf("/home/ferrer/ohi_goc_prep/HS/v2026/HS_Scores_1.1/FIL_Scores_1.1_plot_Sep302026.png"),
  plot = FIL_Scores_OverTime,
  width = 12,
  height = 12,
  dpi = 300
)

# Carbon Sequestration (SEQ)
ggsave(
  filename = sprintf("/home/ferrer/ohi_goc_prep/HS/v2026/HS_Scores_1.1/SEQ_Scores_1.1_plot_Sep302026.png"),
  plot = SEQ_Scores_OverTime,
  width = 12,
  height = 12,
  dpi = 300
)
```

## TABLE:
```{r HS_current_status_1.1 for 2025}

# Make a nice table for 2025:
HS_Scores.2025_table<- HS_current_status_1.1 %>%
  filter(year == 2025) %>%
  select(
    rgn_id,
    OHI_region_name,
    SEQ_score,
    FIL_score,
    PRO_score,
    HS_score
  ) %>%
  mutate(
    rgn_id = gsub("region_", "region ", rgn_id),
    SEQ_score = round(SEQ_score, 0),
    FIL_score = round(FIL_score, 0),
    PRO_score = round(PRO_score, 0),
    HS_score = round(HS_score, 0)
  ) %>%
  kable(
    col.names = c(
      "Region",
      "Region Name",
      "Sequestration Score",
      "Filtration Score",
      "Protection Score",
      "HS Score"
    )
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "bordered"),
    full_width = FALSE
  ) %>%
  add_header_above(c("OHI Scores – Current Status (2025)" = 6))
```

```{r Save HS_Scores.2025_table}
write.csv(
  HS_Scores.2025_table,
  "/home/ferrer/ohi_goc_prep/HS/v2026/HS_Scores_1.1/HS_Scores.2025_df_Sep3026.csv",
  row.names = FALSE
)
```

## MAPS:

```{r Map HS_current_status_1.1}
# Format the rgn_id column for proper merging:
HS_current_status_clean <- HS_current_status_1.1 %>%
  mutate(
    rgn_id = gsub("region ", "Region_", rgn_id)
  )

ohi_habitat_current_status <- ohi_regions %>%
  left_join(
    HS_current_status_1.1 %>%
      filter(year == 2025) %>% # Filter scores for 2025
      mutate(rgn_id = gsub("region ", "Region_", rgn_id)),
    by = "rgn_id"
  )

tmap_mode("plot")

HS_Score.2025_map <- 
  tm_shape(ohi_habitat_current_status %>% filter(location == "land")) +
  tm_polygons(
    col = "#f5e6c8",
    border.col = "gray40",
    lwd = 1,
    alpha = 0.5,
    legend.show = FALSE
  ) +
  
  tm_shape(ohi_habitat_current_status %>% filter(location == "marine")) +
  tm_polygons(
    col = "HS_score",
    palette = "RdYlGn",
    alpha = 0.8,
    title = "HS Score (2025)",
    border.col = "black",
    lwd = 1.2
  )
  
  # tm_layout(
  #   title = "OHI Habitat Services – Current Status (2025)",
  #   title.size = 2,
  #   legend.outside = TRUE,
  #   legend.title.size = 1,
  #   legend.text.size = 1,
  #   frame = FALSE
  # )

HS_Score.2025_map

```


```{r Map Sub-Scores in 2025}
# Pull score from ohi_habitat_current_status, same as above.

tmap_mode("plot")

# Protection
PRO_Scores.2025_map <- 
  tm_shape(ohi_habitat_current_status %>% filter(location == "land")) +
  tm_polygons(
    col = "#f5e6c8",
    border.col = "gray40",
    lwd = 1,
    alpha = 0.5,
    legend.show = FALSE
  ) +
  tm_shape(ohi_habitat_current_status %>% filter(location == "marine")) +
  tm_polygons(
    col = "PRO_score",
    palette = "RdYlGn",
    alpha = 0.8,
    title = "PRO Sub-Score (2025)",
    border.col = "black",
    lwd = 1.2
  )

# Filtration
FIL_Scores.2025_map <- 
  tm_shape(ohi_habitat_current_status %>% filter(location == "land")) +
  tm_polygons(
    col = "#f5e6c8",
    border.col = "gray40",
    lwd = 1,
    alpha = 0.5,
    legend.show = FALSE
  ) +
  tm_shape(ohi_habitat_current_status %>% filter(location == "marine")) +
  tm_polygons(
    col = "FIL_score",
    palette = "RdYlGn",
    alpha = 0.8,
    title = "FIL Sub-Score (2025)",
    border.col = "black",
    lwd = 1.2
  )


# Sequestration
SEQ_Scores.2025_map <- 
  tm_shape(ohi_habitat_current_status %>% filter(location == "land")) +
  tm_polygons(
    col = "#f5e6c8",
    border.col = "gray40",
    lwd = 1,
    alpha = 0.5,
    legend.show = FALSE
  ) +
  tm_shape(ohi_habitat_current_status %>% filter(location == "marine")) +
  tm_polygons(
    col = "SEQ_score",
    palette = "RdYlGn",
    alpha = 0.8,
    title = "SEQ Sub-Score (2025)",
    border.col = "black",
    lwd = 1.2
  )

PRO_Scores.2025_map
FIL_Scores.2025_map
SEQ_Scores.2025_map

```

```{r Save maps as PNGs}
# Main map (HS_Score.2025_map):
tmap_save(
  HS_Score.2025_map,
  file.path("/home/ferrer/ohi_goc_prep/HS/v2026/HS_Scores_1.1/HS_Scores.2025_map_Sep302025.png"),
  width = 12,
  height = 12,
  dpi = 300
)

# Coastal Protection (2025):
tmap_save(
  PRO_Scores.2025_map,
  file.path("/home/ferrer/ohi_goc_prep/HS/v2026/HS_Scores_1.1/SubScores_1.1/PRO_Scores.2025_map.png"),
  width = 12,
  height = 12,
  dpi = 300
)

# Pollution Filtration (2025):
tmap_save(
  FIL_Scores.2025_map,
  file.path("/home/ferrer/ohi_goc_prep/HS/v2026/HS_Scores_1.1/SubScores_1.1/FIL_Scores.2025_map.png"),
  width = 12,
  height = 12,
  dpi = 300
)

# Cabron Sequestration (2025):
tmap_save(
  SEQ_Scores.2025_map,
  file.path("/home/ferrer/ohi_goc_prep/HS/v2026/HS_Scores_1.1/SubScores_1.1/SEQ_Scores.2025_map.png"),
  width = 12,
  height = 12,
  dpi = 300
)

```


# CALCULATE HS TREND ------------------------------
Calculate the final HS_trend from PRO, FIL, and SEQ sub-scores:

# NOTE TO SELF (Sep 11 2026): I am deeply skeptical about this calculation of trend. For starters, I think the trend column includes NAs that MIGHT need to be zeros; I also think -- critically -- trend should be calculated per habitat type and then averaged.

```{r Calculate HS_trend}

colnames(habitat_services_full)

# Calculate carbon, filtration, and protection score
HS_trend <- habitat_services_full %>%
  group_by(rgn_id) %>%
  summarise(
    SEQ_score = sum(extent * carbon_sequestration_gC.m2.yr * trend, na.rm = TRUE) /
                    sum(extent * carbon_sequestration_gC.m2.yr, na.rm = TRUE) * 100,

    FIL_score = sum(extent * filtration * trend, na.rm = TRUE) /
                       sum(extent * filtration, na.rm = TRUE)* 100,

    PRO_score = sum(extent * coastal_protection * trend, na.rm = TRUE) /
                    sum(extent * coastal_protection, na.rm = TRUE) * 100
  )

HS_trend
```

```{r HS_final_trend}
# Average the 3 scores to get final OHI score
HS_final_trend <- HS_trend %>%
  mutate(HS_score = (SEQ_score + FIL_score + PRO_score) / 3)

HS_final_trend
```

```{r Make a table for HS_trend}
# Ignore NA values
HS_trend <- HS_trend %>%
  mutate(HS_score = rowMeans(
    cbind(SEQ_score, FIL_score, PRO_score),
    na.rm = TRUE
  ))

# Make a nice table
HS_trend %>%
  kable(
    col.names = c(
      "Region",
      "Carbon Trend",
      "Filtration Trend",
      "Coastal Trend",
      "Habitat Services Trend"
    )
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "bordered"),
    full_width = FALSE
  ) %>%
  add_header_above(c("OHI Scores – Trend" = 5))

```

```{r Save HS_trend as csv}
write.csv(HS_trend,
          "/home/ferrer/ohi_goc_prep/HS/v2026/HS_Scores_1.1/HS_trend_Sep112026_DRAFT.csv",
          row.names = FALSE)
```

