---
title: 'OHI GoC 2025: Clean Waters: Hazardous Waste Exploration'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: inline
editor_options: 
  chunk_output_type: inline
---

# Hazardous Waste Exploration

NOM-001-SEMARNAT-2021 : Permissible limits of pollutants in discharges to receiving bodies
<https://www.dof.gob.mx/nota_detalle.php?codigo=5645374&fecha=11/03/2022#gsc.tab=0>

^ This could be a good reference for COFEPRIS/water quality...

```{r}
library(tidyverse)
library(here)
library(janitor)
library(sf)
library(tidygeocoder)
library(mapview)
```


## Inventory of sites contaminated with hazardous waste from the Hydrocarbon Sector by municipality.

<https://datos.gob.mx/dataset/sitios_contaminados-por-tipo-de-contaminante>

Number of contaminated sites by entity, municipality, and year. Includes the type and name of the contaminant.

Category: Environment

Formats: CSV

Institution: Agency for Safety, Energy and Environment (ASEA)

```{r}
contaminated_hydrocarbon_sites <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/ASEA/d2025","ASEA_2_Cantidad_de_sitios_contaminados_por_tipo_de_contaminante_2015-2023.csv"))

contaminated_hydrocarbon_sites_geocode <- contaminated_hydrocarbon_sites %>% 
  filter(entidad_federativa %in% c("Baja California","Baja California Sur","Jalisco","Nayarit","Sinaloa","Sonora")) %>% 
  mutate(country = "Mexico") %>% 
  tidygeocoder::geocode(country = country, state = entidad_federativa, city = municipio)  # to give each city a lat, long, if possible 

# General Plutarco Elías Calles
# Sonora, Mexico
# 31.713806, -112.966550

# Cabo Corrientes
# Jalisco, Mexico
# 20.418270, -105.552399

contaminated_hydrocarbon_sites_sf <- contaminated_hydrocarbon_sites_geocode %>% 
  mutate(lat = ifelse(municipio == "Cabo Corrientes", 20.418270, lat), # for cabo corrientes
         long = ifelse(municipio == "Cabo Corrientes", -105.552399, long)) %>% 
  mutate(lat = ifelse(municipio == "General Plutarco Elías Calles", 31.713806, lat), # for cabo corrientes
         long = ifelse(municipio == "General Plutarco Elías Calles", -112.966550, long)) %>% 
  st_as_sf(coords = c("long", "lat"),
           crs = 4326, # WGS84
           remove = FALSE)

mapview(contaminated_hydrocarbon_sites_sf, zcol = "anio")
```


## Inventory of contaminated and remediated sites

Record of contaminated and remediated sites. Includes details of the responsible party, type of contaminant, affected area, and treatment technique or process.

<https://datos.gob.mx/dataset/inventario_sitios_contaminados_remediados>

### National inventory of contaminated sites
Data on contaminated sites, responsible party, type of contaminant and affected area.

Category: Environment

Formats: CSV

Institution: Secretariat of Environment and Natural Resources (SEMARNAT)

```{r}
# "/home/shares/ohi/OHI_GOC/_raw_data/SEMARNAT/d2025"
# "sitios_contaminados_geoportal.csv"

contaminated_sites <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SEMARNAT/d2025","sitios_contaminados_geoportal.csv"))

contaminated_sites_goc <- contaminated_sites %>% 
  filter(estado %in% c("Baja_California","Baja_California_Sur","Jalisco","Nayarit","Sinaloa","Sonora")) # does not need to be geocoded, already has lat and longs!!!


contaminated_sites_sf <- contaminated_sites_goc %>% 
  st_as_sf(coords = c("longitud", "latitud"),
           crs = 4326, # WGS84
           remove = FALSE)

mapview(contaminated_sites_sf, zcol = "cuenta_programa_remediacion_aprobado") # account_remediation_program_approved

```

```{r}
colnames(contaminated_sites_sf)
#  [1] "numero_registro"                           
#  [2] "anio_identificacion"                       
#  [3] "modalidad_sitio_contaminado"               
#  [4] "responsable_contaminacion"                 
#  [5] "estado"                                    
#  [6] "municipio"                                 
#  [7] "ubicacion"                                 
#  [8] "tipo_evento"                               
#  [9] "actividad_responsable_scian_2023_clase"    
# [10] "actividad_responsable_scian_2023_subsector"
# [11] "contaminante_especifico"                   
# [12] "contaminante_generico"                     
# [13] "area_suelo_contaminado_m"                  
# [14] "volumen_suelo_contaminado_m"               
# [15] "cuenta_programa_remediacion_aprobado"      
# [16] "x"                                         
# [17] "y"                                         
# [18] "fecha_identificacion"                      
# [19] "latitud"                                   
# [20] "longitud"                                  
# [21] "fn_area_suelo_contaminado_m"               
# [22] "fn_volumen_suelo_contaminado_m"            
# [23] "estado_etq"                                
# [24] "geometry"  

summary(contaminated_sites_sf$fn_area_suelo_contaminado_m)
# Show in New Window
#      Min.   1st Qu.    Median      Mean   3rd Qu.      Max.      NA's 
#      0.54    138.22    326.22   3309.59    672.19 183375.70         6 

summary(contaminated_sites_sf$fn_volumen_suelo_contaminado_m)
     # Min.   1st Qu.    Median      Mean   3rd Qu.      Max.      NA's 
     # 0.27    119.81    384.50   5208.68   1337.87 252992.27         7 

library(gtsummary)

contaminated_summary_tbl <- contaminated_sites_sf %>% 
  st_drop_geometry() %>% 
  mutate(anio_identificacion = as.character(anio_identificacion),
         fn_area_suelo_contaminado_m = as.numeric(fn_area_suelo_contaminado_m),
         fn_volumen_suelo_contaminado_m = as.numeric(fn_volumen_suelo_contaminado_m)) %>% 
  tbl_summary(include = c(anio_identificacion, modalidad_sitio_contaminado, contaminante_generico, cuenta_programa_remediacion_aprobado, fn_area_suelo_contaminado_m, fn_volumen_suelo_contaminado_m))

print(contaminated_summary_tbl)
```

#### WRONG: Reference Point

100 = all identified contaminated sites are remediated or under remediation
0 = none of the identified sites have been remediated

Or, maybe we could do:
Status = % of sites with remediation approved
Trend = change in that % over time
Pressure = number or total volume of new contaminated sites identified

```{r}
table(contaminated_sites_sf$cuenta_programa_remediacion_aprobado, useNA = "ifany")
# No Si 
# 55 87 
```

#### WRONG: Score by year?

Join with OHI Regions:
```{r}
goc_poly <- st_read(here::here("spatial/ohi_regions","goc_ohi_rgns.shp"))

goc_ohi_rgn_id <- goc_poly %>% 
  st_drop_geometry()

contaminated_sites_sf_prepjoin <- contaminated_sites_sf %>% 
  mutate(estado = ifelse(estado == "Jalisco","Nayarit",estado))

contaminated_sites_sf_ohi <- left_join(contaminated_sites_sf_prepjoin, goc_ohi_rgn_id, by = "estado")

mapview(contaminated_sites_sf_ohi, zcol = "rgn_id")
```

Crop to 50km:
```{r}
inland_50 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer","inland_50km_goc.shp"))

contaminated_sites_sf_ohi_trans <- st_transform(contaminated_sites_sf_ohi, st_crs(inland_50))
contaminated_sites_50km <- st_intersection(contaminated_sites_sf_ohi_trans, inland_50)

# contaminated_sites_50km <- st_join(contaminated_sites_sf_ohi_trans, inland_50, join = st_intersects, left = FALSE)


# st_write(contaminated_sites_50km, here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/int","hazardous_contaminated_sites_50km.gpkg")) # gpkg because of long names that would be too hard to understand otherwise!!

mapview(contaminated_sites_50km, zcol = "cuenta_programa_remediacion_aprobado") + mapview(inland_50)
```


```{r}
## by number of sites
score_by_year <- contaminated_sites_50km %>%
  st_drop_geometry() %>%
  group_by(anio_identificacion, rgn_id) %>%
  summarise(
    n_sites = n(),
    n_remediated = sum(cuenta_programa_remediacion_aprobado == "Si", na.rm = TRUE),
    pct_remediated = 100 * n_remediated / n_sites
  ) %>%
  mutate(score = pct_remediated)  # current status...

## making sure each year accounts for the previous years' unremedied sites:
score_cumulative <- contaminated_sites_50km %>%
  st_drop_geometry() %>%
  mutate(
    anio_identificacion = as.numeric(anio_identificacion),
    remediated = ifelse(cuenta_programa_remediacion_aprobado == "Si", 1, 0) # since we do not have any NAs
  ) %>%
  # sort by year to enable cumulative logic
  arrange(rgn_id, anio_identificacion) %>%
  group_by(rgn_id) %>%
  # compute cumulative totals for each year
  mutate(
    n_sites_total = cumsum(!is.na(anio_identificacion)),
    n_remediated_cum = cumsum(remediated)
  ) %>%
  group_by(rgn_id, anio_identificacion) %>%
  summarise(
    n_sites_total = max(n_sites_total),
    n_remediated = max(n_remediated_cum),
    pct_remediated = 100 * n_remediated / n_sites_total,
    .groups = "drop"
  ) %>%
  mutate(score = pct_remediated)
```


```{r}
## by volume of site -- NEEDS WORK!!!!
# score_by_year <- contaminated_sites_sf %>%
#   st_drop_geometry() %>%
#   group_by(anio_identificacion) %>%
#   summarise(
#     total_vol = sum(fn_volumen_suelo_contaminado_m, na.rm = TRUE),
#     vol_remediated = sum(ifelse(cuenta_programa_remediacion_aprobado == 1, 
#                                 fn_volumen_suelo_contaminado_m, 0), na.rm = TRUE) # fix the sum. something going wrong here!!
#   ) %>%
#   mutate(score = 100 * vol_remediated / total_vol)

```

```{r}
library(plotly)

## plot the score over time
hazardous_sites_plotly <- plotly::plot_ly(
  score_cumulative, 
  x = ~anio_identificacion, 
  y = ~score, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "CURRENT STATUS: Percent of hazardous waste sites remediated",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(score_cumulative$anio_identificacion), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
hazardous_sites_plotly

# htmlwidgets::saveWidget(hazardous_sites_plotly, here::here("CW/v2025/int","hazardous_sites_pct_remediation_score_102425.html"), selfcontained = TRUE)
```



### National inventory of remediated sites
Data on remediated sites, type of contaminant, affected area, and treatment technique or process.

Category: Environment

Formats: CSV

Institution: Secretariat of Environment and Natural Resources (SEMARNAT)

```{r}
# "/home/shares/ohi/OHI_GOC/_raw_data/SEMARNAT/d2025"
# "sitios_remediados_geoportal.csv"

remediated_sites <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SEMARNAT/d2025","sitios_remediados_geoportal.csv"))

remediated_sites_goc <- remediated_sites %>% 
  filter(estado %in% c("Baja_California","Baja_California_Sur","Jalisco","Nayarit","Sinaloa","Sonora")) # does not need to be geocoded, already has lat and longs!!!


remediated_sites_sf <- remediated_sites_goc %>% 
  st_as_sf(coords = c("longitud", "latitud"), # no NAs!! yay
           crs = 4326, # WGS84
           remove = FALSE)

colnames(remediated_sites_sf)
#  [1] "numero_registro"                            "clasificacion_tecnica_proceso_tratamiento" 
#  [3] "anio_identificacion_sitio_contaminado"      "modalidad_sitio_contaminado"               
#  [5] "responsable_contaminacion"                  "responsable_tecnico_remediacion"           
#  [7] "estado"                                     "municipio"                                 
#  [9] "ubicacion"                                  "tipo_evento"                               
# [11] "actividad_responsable_scian_2023_clase"     "actividad_responsable_scian_2023_subsector"
# [13] "contaminante_especifico"                    "contaminante_generico"                     
# [15] "tecnica_proceso_tratamiento"                "anio_conclusion_remediacion"               
# [17] "area_suelo_remediado_m"                     "volumen_suelo_remediado_m"                 
# [19] "x"                                          "y"                                         
# [21] "fecha_identificacion"                       "fecha_conclusion_remediacion"              
# [23] "latitud"                                    "longitud"                                  
# [25] "fn_area_suelo_remediado_m"                  "fn_volumen_suelo_remediado_m"              
# [27] "estado_etq"                                 "geometry"   
# does not have account_remediation_program_approved

mapview(remediated_sites_sf, zcol = "anio_conclusion_remediacion")
mapview(remediated_sites_sf, zcol = "tipo_evento")

```

### Bind contamindated and remediated 

They should not share their `numero_registro` id, they should either be contaminated or remediated.

```{r}
# Combine contaminated + remediated info conceptually
# Example weights:
#   0 = contaminated, no plan
#   0.5 = contaminated, plan approved
#   1 = remediated

# prep each for join (we will join with rgn_id and limit to 50km once they are joined)
cont <- contaminated_sites_sf %>%
  st_drop_geometry() %>%
  select(numero_registro, estado, municipio, anio_identificacion,cuenta_programa_remediacion_aprobado, fn_volumen_suelo_contaminado_m, latitud, longitud) %>%
  mutate(
    status_code = case_when(
      cuenta_programa_remediacion_aprobado == "Si" ~ 0.5,
      TRUE ~ 0
    ),
    anio = as.numeric(anio_identificacion),
    from = "cont"
  )

rem <- remediated_sites_sf %>%
  st_drop_geometry() %>%
  select(numero_registro, estado, municipio, anio_conclusion_remediacion, fn_volumen_suelo_remediado_m, latitud, longitud) %>%
  mutate(
    status_code = 1,
    anio = as.numeric(anio_conclusion_remediacion),
    from = "rem"
  )

# merge them
sites_all <- bind_rows(cont, rem)
```

```{r}
# incorrect workflow!! IGNORE THIS CHUNK

################# join with rgn_id information #################

# goc_poly <- st_read(here::here("spatial/ohi_regions","goc_ohi_rgns.shp"))
# 
# goc_ohi_rgn_id <- goc_poly %>% 
#   st_drop_geometry()
# 
# sites_all_prepjoin <- sites_all %>% 
#   mutate(estado = ifelse(estado == "Jalisco","Nayarit",estado))
# 
# sites_all_ohi_sf <- left_join(sites_all_prepjoin, goc_ohi_rgn_id, by = "estado") %>% 
#   st_as_sf(coords = c("longitud", "latitud"),
#            crs = 4326, # WGS84
#            remove = FALSE)
# 
# mapview(sites_all_ohi_sf, zcol = "rgn_id")
# 
# ########### limit to 50km from coast: this should later be replaced by watersheds???? #################
# 
# inland_50 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer","inland_50km_goc.shp"))
# sites_all_ohi_sf_trans <- st_transform(sites_all_ohi_sf, st_crs(inland_50))
# all_sites_50km <- st_intersection(sites_all_ohi_sf_trans, inland_50)
# 
# # st_write(all_sites_50km, here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/int","remediated_and_contaminated_sites_50km.gpkg")) # gpkg because of long names that would be too hard to understand otherwise!!
# 
# mapview(all_sites_50km, zcol = "status_code") + mapview(inland_50)
# 
# 
# ### fix duplicates
# inland_50_nonoverlap <- inland_50 %>%
#   st_make_valid() %>%
#   st_union() %>% ## loses region ids but that is fine in this case since we already joined with them
#   st_collection_extract("POLYGON") %>%
#   st_sf() %>%
#   mutate(rgn_id = paste0("Region_", seq_len(n())))
# 
# st_overlaps(inland_50_nonoverlap, sparse = FALSE) %>% rowSums() %>% table()
# mapview(inland_50_nonoverlap)
# all_sites_50km <- st_intersection(sites_all_ohi_sf_trans, inland_50_nonoverlap)
```


```{r}
############ to avoid duplicates and make it so much easier: ######### 
sites_all_sf <- sites_all %>% 
  st_as_sf(coords = c("longitud", "latitud"),
           crs = 4326, # WGS84
           remove = FALSE)

mapview(sites_all_sf, zcol = "status_code")

inland_50 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer","inland_50km_goc.shp"))

sites_all_sf_trans <- st_transform(sites_all_sf, st_crs(inland_50))
all_sites_50km <- st_intersection(sites_all_sf_trans, inland_50)

mapview(all_sites_50km, zcol = "status_code") + mapview(inland_50)
```

## Score 1: Are there sites in a region?

If there are no sites in a region, they should get a 1.  If they have some sites in the region, then the concentration of sites should be considered. That will be sites per unit area.

```{r}
summary(all_sites_50km) # no NAs in anio!

# summarize per year and region -- incorrect bc it is not cumulative 
status_by_year <- all_sites_50km %>%
  st_drop_geometry() %>% 
  group_by(rgn_id, anio) %>%
  summarise(
    n_sites = n()
  ) # this is not a concentration of sites by area.  we need to do that!!
```

### Cumulative hazardous waste site density by region:

```{r}
#--- 1. calculate region area (km^2) ---
## this will be changed to be WATERSHED data in the future!!!!
rgn_area <- inland_50 %>%
  mutate(area_km2 = as.numeric(st_area(.) / 1e6)) %>%
  st_drop_geometry() %>%
  select(rgn_id, area_km2)

# rgn_area_50km <- inland_50 %>%
#   #group_by(rgn_id) %>% 
#   mutate(area_km2 = as.numeric(st_area(.) / 1e6)) %>%
#   group_by(rgn_id) %>% 
#   st_drop_geometry() %>%
#   summarize(rgn_area_km2 = sum(area_km2)) %>% 
#   ungroup() 
```


```{r}
#--- 2. count sites per region and year ---
sites_per_region <- all_sites_50km %>%
  st_drop_geometry() %>%
  mutate(anio = as.numeric(anio)) %>%
  group_by(rgn_id, anio) %>%
  summarise(n_sites = n(), .groups = "drop") %>%
  group_by(rgn_id) %>% 
  complete(anio = c(2008:2023)) %>% 
  mutate(n_sites = ifelse(is.na(n_sites), 0, n_sites)) %>% 
  left_join(rgn_area, by = "rgn_id") 
```


```{r}
#--- 3. expand to full year sequence per region, at least as much as possible ---
sites_expanded <- sites_per_region %>%
  group_by(rgn_id) %>%
  #complete(anio = seq(min(anio, na.rm = TRUE), max(anio, na.rm = TRUE))) %>% # note that since this is grouped by region that you might see certain regions only having one year of data, because their min and max are that year.  even after using complete()
  #replace_na(list(n_sites = 0)) %>%
  #fill(area_km2, .direction = "downup") %>% # filling in the area for that region for all years
 # arrange(anio) %>% # arrange so that the year is ascending per region id.
  
  #--- 4. cumulative sum of sites per region ---
  mutate(
    cum_sites = cumsum(n_sites),
    density_cum = cum_sites / area_km2
  ) %>%
  ungroup()
```


```{r}
#--- 5. rescale density to 0 to 1 per year (1 = best, 0 = worst) ---
score1_cumulative <- sites_expanded %>%
  mutate(
    score1 = 1 - (density_cum / max(density_cum, na.rm = TRUE)), # essentially, across all years, if the density_cum of that row divided by the maximum density_cum across all observations. 
    # if a score1 is = 0, then that is because that observation was the max overall. Oof!
    score1_100 = score1 * 100,
    rgn_id = as.factor(rgn_id),
    anio = as.numeric(anio),
    score1_100 = as.numeric(score1_100)
  )

```

```{r}
library(plotly)

score1_cumulative_plotly <- plotly::plot_ly(
  score1_cumulative, 
  x = ~anio, 
  y = ~score1_100, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "Score 1: Cumulative site density rescaled against the maxmimum density across all years",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(score1_cumulative$anio), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
score1_cumulative_plotly

# htmlwidgets::saveWidget(score1_cumulative_plotly, here::here("CW/v2025/int","hazardous_sites_score1_110325.html"), selfcontained = TRUE)
```


## Score 2: do those sites have plans to be remediated, and to what extent are they?

We actually need to make this cumulative!!!! An example of how that would look:

| Year | New sites that year           | Cumulative sites | Cumulative average status     |
| ---- | ----------------------------- | ---------------- | ----------------------------- |
| 2017 | 5 sites, all contaminated (0) | 5                | 0.0                           |
| 2018 | 2 new, both with plans (0.5)  | 7                | (5×0 + 2×0.5)/7 = 0.14        |
| 2019 | 3 new, remediated (1)         | 10               | (5×0 + 2×0.5 + 3×1)/10 = 0.35 |

The goal would be to have all sites remediated, 100/100.

```{r}
### cumulative version: HAS DUPLICATES
# score2_cumulative <- all_sites_50km %>%
#   st_drop_geometry() %>%
#   mutate(anio = as.numeric(anio)) %>% # make sure it is numeric
#   group_by(rgn_id) %>%
#   arrange(rgn_id, anio) %>% # make sure years are in order for cumulative logic
#   group_modify(~ { # expand all combinations of region and year
#     df <- .
#     all_years <- seq(min(df$anio, na.rm = TRUE), max(df$anio, na.rm = TRUE))
#     df_expanded <- tidyr::expand_grid(anio = all_years) %>%
#       left_join(df, by = "anio")
#     
#     # cumulative calculation
#     df_expanded %>%
#       mutate(
#         n_sites_total = cumsum(!is.na(status_code)),
#         sum_status = cumsum(replace_na(status_code, 0)),
#         mean_status = if_else(n_sites_total > 0, sum_status / n_sites_total, NA_real_),
#         score = mean_status * 100
#       )
#   }) %>%
#   ungroup() %>%
#   select(rgn_id, anio, n_sites_total, mean_status, score)

## fixing duplicates
score2_cumulative <- all_sites_50km %>%
  st_drop_geometry() %>%
  mutate(anio = as.numeric(anio)) %>%
  group_by(rgn_id) %>%
  arrange(anio) %>%
  group_modify(~ {
    df <- .
    all_years <- seq(min(df$anio, na.rm = TRUE), max(df$anio, na.rm = TRUE))
    
    # summarise once per year to avoid duplicates
    df_year <- df %>%
      group_by(anio) %>%
      summarise(mean_status = mean(status_code, na.rm = TRUE), .groups = "drop")
    
    # df_expanded <- tidyr::expand_grid(anio = all_years) %>%
    #   left_join(df_year, by = "anio") %>%
    #   arrange(anio)
    
    df_expanded <- tidyr::expand_grid(anio = 2008:2023) %>%
      left_join(df_year, by = "anio") %>%
      arrange(anio)
    
    
    df_expanded %>%
      mutate(
        n_sites_total = cumsum(!is.na(mean_status)),
        sum_status = cumsum(replace_na(mean_status, 0)),
        mean_status = if_else(n_sites_total > 0, sum_status / n_sites_total, 1),
        score2_100 = mean_status * 100
      )
  }) %>%
  ungroup() %>%
  select(rgn_id, anio, n_sites_total, mean_status, score2_100)
```

##### Plotly

```{r}
library(plotly)

## plot the score2 over time
score2_cumulative_plotly <- plotly::plot_ly(
  score2_cumulative, 
  x = ~anio, 
  y = ~score2_100, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "Score 2: Cumulative percent remediation for all hazardous sites within a region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(score2_cumulative$anio), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
score2_cumulative_plotly

htmlwidgets::saveWidget(score2_cumulative_plotly, here::here("CW/v2025/int","hazardous_sites_score2_110325.html"), selfcontained = TRUE)
```

Gut check with a plotly of cumulative sites over time:

```{r}
library(plotly)

## plot the score2 over time
sites_plotly <- plotly::plot_ly(
  score2_cumulative, 
  x = ~anio, 
  y = ~n_sites_total, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "Cumulative number of sites per year",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Number of sites (cumulative)")
  ) 
sites_plotly
# 
# htmlwidgets::saveWidget(all_hazardous_sites_plotly, here::here("CW/v2025/int","hazardous_sites_pct_remediation_score_102425.html"), selfcontained = TRUE)
```


<!-- #### Gapfilled options -->

<!-- ```{r} -->
<!-- #### cumulative version of score2 gapfilled between years -->

<!-- status_all_cumulative_gf <- all_sites_50km %>% -->
<!--   st_drop_geometry() %>% -->
<!--   mutate(anio = as.numeric(anio)) %>% -->
<!--   group_by(rgn_id) %>% -->
<!--   arrange(anio, .by_group = TRUE) %>% -->
<!--   group_modify(~ { -->
<!--     df <- . -->

<!--     # collapse to one record per year (if multiple sites in same year) -->
<!--     df_year <- df %>% -->
<!--       group_by(anio) %>% -->
<!--       summarise( -->
<!--         mean_status = mean(status_code, na.rm = TRUE), -->
<!--         .groups = "drop" -->
<!--       ) -->

<!--     # ensure all years between min and max exist -->
<!--     all_years <- seq(min(df_year$anio, na.rm = TRUE), max(df_year$anio, na.rm = TRUE)) -->
<!--     df_expanded <- tidyr::expand_grid(anio = all_years) %>% -->
<!--       left_join(df_year, by = "anio") %>% -->
<!--       arrange(anio) -->

<!--     # cumulative + gapfilled calculation -->
<!--     df_expanded %>% -->
<!--       mutate( -->
<!--         n_sites_total = cumsum(!is.na(mean_status)), -->
<!--         sum_status = cumsum(replace_na(mean_status, 0)), -->
<!--         mean_status = if_else(n_sites_total > 0, sum_status / n_sites_total, NA_real_), -->
<!--         score = mean_status * 100 -->
<!--       ) %>% -->
<!--       tidyr::fill(mean_status, score, .direction = "down") # gapfill! -->
<!--   }) %>% -->
<!--   ungroup() %>% -->
<!--   select(rgn_id, anio, n_sites_total, mean_status, score) -->
<!-- ``` -->


<!-- ```{r} -->
<!-- ### updates cumulative with full gapfilling -->
<!-- # to find the full range of years across all regions -->
<!-- all_years <- seq(min(all_sites_50km$anio, na.rm = TRUE), -->
<!--                  max(all_sites_50km$anio, na.rm = TRUE)) -->

<!-- status_all_cumulative_gf_max <- all_sites_50km %>% -->
<!--   st_drop_geometry() %>% -->
<!--   mutate(anio = as.numeric(anio)) %>% -->
<!--   group_by(rgn_id) %>% -->
<!--   arrange(anio, .by_group = TRUE) %>% -->
<!--   group_modify(~ { -->
<!--     df <- . -->

<!--     # collapse to one record per year -->
<!--     df_year <- df %>% -->
<!--       group_by(anio) %>% -->
<!--       summarise(mean_status = mean(status_code, na.rm = TRUE), .groups = "drop") -->

<!--     # expand to full year range -->
<!--     df_expanded <- tibble(anio = all_years) %>% -->
<!--       left_join(df_year, by = "anio") %>% -->
<!--       arrange(anio) -->

<!--     # fill initial missing years with 100 -->
<!--     first_year_with_data <- min(df_year$anio, na.rm = TRUE) -->
<!--     df_expanded <- df_expanded %>% -->
<!--       mutate( -->
<!--         mean_status = case_when( -->
<!--           anio < first_year_with_data ~ 1, # 100% for years before region starts -->
<!--           TRUE ~ mean_status -->
<!--         ), -->
<!--         score = mean_status * 100 -->
<!--       ) -->

<!--     # cumulative/gapfill for following years -->
<!--     df_expanded <- df_expanded %>% -->
<!--       tidyr::fill(mean_status, score, .direction = "down") %>% -->
<!--       mutate( -->
<!--         n_sites_total = cumsum(!is.na(mean_status)), -->
<!--         sum_status = cumsum(replace_na(mean_status, 0)), -->
<!--         mean_status = if_else(n_sites_total > 0, sum_status / n_sites_total, NA_real_), -->
<!--         score = mean_status * 100 -->
<!--       ) -->

<!--     df_expanded -->
<!--   }) %>% -->
<!--   ungroup() %>% -->
<!--   select(rgn_id, anio, n_sites_total, mean_status, score) -->

<!-- ``` -->

#### Volume weighted? 

```{r}
# to reflect volume and use a weighted mean: need to fix to make accurate!!!! To clarify, this is incorrect because we actually need to account for the fact that there are two columns: fn_volumen_suelo_remediado_m and fn_volumen_suelo_contaminado_m: Right now it is not accounted for.

# status_all_cumulative_weighted <- all_sites_50km %>%
#   st_drop_geometry() %>%
#   mutate(anio = as.numeric(anio)) %>% # make sure it is numeric
#   group_by(rgn_id) %>%
#   arrange(rgn_id, anio) %>% # make sure years are in order for cumulative logic
#   group_modify(~ { # expand all combinations of region and year
#     df <- .
#     all_years <- seq(min(df$anio, na.rm = TRUE), max(df$anio, na.rm = TRUE))
#     df_expanded <- tidyr::expand_grid(anio = all_years) %>%
#       left_join(df, by = "anio")
# 
#     # cumulative calculation
#     df_expanded %>%
#       mutate(
#         n_sites_total = cumsum(!is.na(status_code)),
#         sum_status = cumsum(replace_na(status_code, 0)),
#         cum_vol = cumsum(replace_na(fn_volumen_suelo_contaminado_m, 0)),
#         weighted_sum = cumsum(replace_na(status_code * fn_volumen_suelo_contaminado_m, 0)),
#         mean_status = if_else(cum_vol > 0, weighted_sum / cum_vol, NA_real_),
#         score = mean_status * 100
#       )
#   }) %>%
#   ungroup() %>%
#   select(rgn_id, anio, n_sites_total, mean_status, score)

```


## Average Score1+2

Making the full current status by joining score1_cumulative and score2_cumulative:

```{r}
final_hazardous_indicator <- score1_cumulative %>%
  select(rgn_id, anio, score1 = score1_100) %>%
  left_join(score2_cumulative %>%
              select(rgn_id, anio, score2 = score2_100), 
            by = c("rgn_id", "anio")) %>%
  mutate(
    final_score = rowMeans(select(., score1, score2), na.rm = TRUE)
  ) 

# save to mazu
write_csv(final_hazardous_indicator, here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/hazardous_waste/output","current_status_hazardous_waste.csv"))
```

```{r}
library(plotly)

## plot final_score over time
final_hazardous_indicator_plotly <- plotly::plot_ly(
  final_hazardous_indicator, 
  x = ~anio, 
  y = ~final_score, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "CURRENT STATUS: Coastal Hazardous Waste Sites as Point Sources",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(final_hazardous_indicator$anio), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
final_hazardous_indicator_plotly


## plot
status_all_cumulative_gf_max_plotly <- plotly::plot_ly(
  status_all_cumulative_gf, 
  x = ~anio, 
  y = ~score, 
  color = ~rgn_id, 
  colors = ~rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: Coastal Hazardous Waste Sites as Point Sources",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(status_all_cumulative_gf$anio), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
status_all_cumulative_gf_max_plotly

htmlwidgets::saveWidget(final_hazardous_indicator_plotly, here::here("CW/v2025/int","hazardous_sites_current_status_110325.html"), selfcontained = TRUE)
```


# Estimated generation of hazardous waste by type of waste

Estimated annual generation of hazardous waste by type of waste under ASEA's jurisdiction, from 2017 to 2023. State and municipal data with information on waste type, sector, and estimated generation.

Category: Environment
Formats: CSV
Institution: Agency for Safety, Energy and Environment (ASEA)

```{r}
ASEA_3_Generacio_n_estimada_de_residuos_peligrosos_por_tipo_de_residuo_2017 <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/ASEA/d2025", "ASEA_3_Generacio_n_estimada_de_residuos_peligrosos_por_tipo_de_residuo_2017-2023.csv"))
```


# Possible resiliences 

<https://datos.gob.mx/dataset/inspeccion_vigilancia_industrial>
<https://datos.gob.mx/dataset/riesgo_ambiental>
