---
title: 'OHI GoC 2025: Clean waters pesticides'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# CW: Pesticides

## Estimated pesticide discharge

```{r setup}
library(sf)
library(utils)
library(tidyverse)
library(here)
library(raster)
library(mapview)
library(purrr)
library(stringr) # fuzzy string matching
library(stringi)
library(janitor)
library(tibble)
```

### Watersheds and pour points

```{r}
## lets use the full state for now and then summarize by the full watersheds.
goc_ohi_rgns <- st_read(here::here("spatial/ohi_regions","goc_ohi_rgns.shp"))

union_rgns <- goc_ohi_rgns %>% 
  st_union() %>% 
  st_make_valid()

# plot(goc_ohi_rgns)
# plot(union_rgns)

watershed_all <- st_read(file.path("/home/shares/ohi/stressors_2021/_dataprep/nutrients/watersheds_pourpoints/watersheds_all_4326.shp"))
st_crs(watershed_all)
# mapview(watershed_all)

watershed_goc <- st_intersection(st_transform(watershed_all, st_crs(union_rgns)), union_rgns)
mapview(watershed_goc)

# --- if we plan to change the buffer for some reason ---
# inland_50 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer","inland_50km_goc.shp"))
# watershed_goc_50km <- st_intersection(st_transform(watershed_all, st_crs(inland_50)), inland_50)
# mapview(watershed_goc_50km)
```


```{r}
# here's the associated pourpoints that should match with the watersheds:
pourpoints <- st_read("/home/shares/ohi/stressors_2021/_dataprep/nutrients/watersheds_pourpoints/pourpoints_mol.shp")

pourpoints_goc <- st_intersection(st_transform(pourpoints, st_crs(union_rgns)), union_rgns)
mapview(pourpoints_goc)

mapview(pourpoints_goc) + mapview(watershed_goc)
```

### Calculate aquacultural area per watershed

```{r}
# bring in the aquaculture land use over time output:
usv_aq <- st_read(here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/pesticides/int/usv_aquaculture_goc_allyears.gpkg"))
mapview(usv_aq)

unique(usv_aq$year)

usv_aq_2005 <- usv_aq %>% 
  filter(year %in% "2005")
usv_aq_2009 <- usv_aq %>% 
  filter(year %in% "2009")
usv_aq_2013 <- usv_aq %>% 
  filter(year %in% "2013")
usv_aq_2016 <- usv_aq %>% 
  filter(year %in% "2016")
usv_aq_2021 <- usv_aq %>% 
  filter(year %in% "2021")

mapview(usv_aq_2005, col.regions = "red") + mapview(usv_aq_2009, col.regions = "darkorange") + mapview(usv_aq_2013, col.regions = "orange") + mapview(usv_aq_2016, col.regions = "yellow") + mapview(usv_aq_2021, col.regions = "green")

mapview(usv_aq_2021, col.regions = "green") + mapview(watershed_goc) + mapview(pourpoints_goc)

```

```{r}
# reproj usv to match watersheds CRS
usv_aq_reproj <- st_transform(usv_aq, st_crs(watershed_goc)) # so that it is in meters not degrees

# Coordinate Reference System:
#   User input: Mexico ITRF2008 / LCC 
#   wkt:
# PROJCRS["Mexico ITRF2008 / LCC",
#     BASEGEOGCRS["Mexico ITRF2008",
#         DATUM["Mexico ITRF2008",
#             ELLIPSOID["GRS 1980",6378137,298.257222101,
#                 LENGTHUNIT["metre",1]]],

# intersect aq polygons with watersheds
usv_aq_watershed <- st_intersection(usv_aq_reproj, watershed_goc) %>%
  mutate(area_ha = as.numeric(st_area(.)) / 10000)  # there are 0.0001 hectares in 1 square meter, to convert from square meters to hectares, multiply by 0.0001 or divide by 10000
```

Check out 2021:
```{r}
usv_aq_watershed_2021 <- usv_aq_watershed %>% 
  filter(year %in% 2021)

mapview(usv_aq_watershed_2021, zcol = "area_ha") 

mapview(usv_aq_watershed_2021, zcol = "basin_id") + mapview(watershed_goc, zcol = "basin_id")
```


```{r}
# summarize hectares per watershed per year
aq_area_by_watershed <- usv_aq_watershed %>%
  st_drop_geometry() %>% # maybe take out...
  group_by(basin_id, year) %>% 
  summarize(aq_area_ha = sum(area_ha, na.rm = TRUE)) %>% 
  ungroup() %>% 
  mutate(year = as.numeric(year))
```

### Estimating relative risk of chemicals (both pesticides and pharmaceuticals, general) using production per hectare

González-Rivas et al. 2023, [Estimating the shrimp farm’s production and their future growth prediction by remote sensing: Case study Gulf of California](https://www.frontiersin.org/journals/marine-science/articles/10.3389/fmars.2023.1130125/full), ran a "linear regression model between the shrimp farms area, calculated using remote sensing, and statistical data on shrimp production" which was "strongly positive and significant (r2 = 0.874, p =2.209e-11)."

**Useful figures from this paper:**

- Figure 1, "Total estimated area of shrimp ponds (ha, blue line) based on remote sensing and shrimp production (tons, dotted orange line) from 1993 to 2021, for the east part of the Gulf of California corresponding to the Mexican states of Sinaloa and Sonora."

- Figure 3, "Relationship between the total shrimp pond area estimated using remote sensing (ha/year) and (A) the corresponding production (ton/year) and (B) time (years)."

Based on this linear relationship with a strong positive and significant R^2, we can use the slope between shrimp pond area (hectare/year) and corresponding production (tons/year), which is 1.5 tons/ha, to estimate the total production from each OHI region over time as land use changes.  Note that because this is a linear relationship, the slope is not changing over time. This slope was determined based on years 1993-2021.


```{r}
aq_production_by_watershed <- aq_area_by_watershed %>% 
  mutate(
    shrimp_prod_tons = aq_area_ha * 1.5 # ha * tons/ha = tons
  )
```


### Pour points and OHI regions

```{r pourpoints}
pourpoints_shrimp_prod <- pourpoints_goc %>%
  dplyr::select(basin_id) %>%
  left_join(aq_production_by_watershed, by = "basin_id")
```


```{r ohi_marine_and_inland}
inland_50 <- st_read(here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer", "inland_50km_goc.shp"))
inland_10 <- st_read(here("/home/shares/ohi/OHI_GOC/spatial/inland_10km_buffer", "inland_10km_goc.shp"))
regions_marine <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp")) %>%
  filter(location == "marine")

marine_and_inland50 <- bind_rows(regions_marine, inland_50) %>%
  dplyr::select(rgn_id, geometry) %>%
  group_by(rgn_id) %>%
  summarize(geometry = st_union(geometry), .groups = "drop")

marine_and_inland10 <- bind_rows(regions_marine, inland_10) %>%
  dplyr::select(rgn_id, geometry) %>%
  group_by(rgn_id) %>%
  summarize(geometry = st_union(geometry), .groups = "drop")
```


```{r pourpoints}
pourpoints_shrimp_prod_final <- pourpoints_shrimp_prod %>%
  filter(!is.na(shrimp_prod_tons), shrimp_prod_tons > 0) %>% # note that the watershed with the id 73730 wraps around, and that the pour point is only on the left side...not sure that is correct.
  st_intersection(marine_and_inland50) %>% # keep those within 50 km...
  mutate(is_within_10km = lengths(st_within(., marine_and_inland10)) > 0) %>% # but make sure not to keep the ones on the pacific side of the peninsula.
  filter(!(is_within_10km == FALSE & rgn_id == "Region_4"))

mapview(pourpoints_shrimp_prod_final, zcol = "shrimp_prod_tons") + mapview(usv_aq_2021, col.regions = "red") + mapview(watershed_goc, col.regions = "pink")

pourpoints_shrimp_prod_rgnid <- st_intersection(pourpoints_shrimp_prod_final, goc_ohi_rgns) %>%
  st_drop_geometry() %>%
  group_by(rgn_id, year) %>%
  summarize(
    aq_area_ha = sum(aq_area_ha, na.rm = TRUE),
    shrimp_tons = sum(shrimp_prod_tons, na.rm = TRUE),
    .groups = "drop"
  )

pourpoints_shrimp_prod_rgnid
```

# Save to mazu

```{r save}
write_csv(aq_production_by_watershed, here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/shrimp_production_chemicals/int/aq_production_tons_shrimp_by_watershed.csv"))

write_csv(pourpoints_shrimp_prod_rgnid, here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/shrimp_production_chemicals/int/shrimp_tons_pourpoints_prod_rgnid.csv"))

st_write(pourpoints_shrimp_prod_final, here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/shrimp_production_chemicals/output/shrimp_tons_pourpoints_prod_final.gpkg"), delete_dsn = TRUE # so it can overwrite if changes are made.
)
```

