---
title: 'OHI GoC 2025: Clean waters pesticides'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# CW: Pesticides

## Current status

```{r setup}
library(sf)
library(utils)
library(tidyverse)
library(here)
library(raster)
library(mapview)
library(purrr)
library(stringr) # fuzzy string matching
library(stringi)
library(janitor)
library(tibble)
```

### Read in agriculture estimates


```{r}
total_ag_pesticides_kg <- st_read(here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/pesticides/int", "total_agriculture_pesticides_kg_pourpoints_rgnid_allyears.gpkg")) %>% 
  rename(ag_pesticide_kg = total_pesticide_kg)
```

## Standardize by coastline length

To allow for comparison of pesticide pressure across regions of different sizes, total estimated kg of pesticides were standardized by the length of each region's coastline in kilometers. Coastline data were obtained from Natural Earth (version 4.1.0, 1:10m scale; naturalearthdata.com), which derives primarily from World Data Bank 2. The global coastline shapefile was clipped to the Gulf of California study area and reprojected to match the Mexico ITRF2008 Lambert Conformal Conic coordinate reference system. Island coastlines were removed by erasing a 1,500 m buffer around the Islas del Golfo de California protected area polygon prior to calculating lengths, so that only mainland coastline contributed to each region's total. Coastline segments were then intersected with each OHI region boundary and summed using st_length in the sf package in R, which returned lengths in meters that were subsequently converted to kilometers. Total estimated kg of pesticides per region and year were then divided by the corresponding coastline length to produce a standardized pressure metric of kg of pesticides per kilometer of coastline, which will subsequently be made into a current status score.

```{r}
world_coastline <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/ne_10m_coastline/ne_10m_coastline.shp"))

st_crs(marine_and_inland10) == st_crs(world_coastline) # FALSE

goc_coastline <- st_intersection(st_transform(world_coastline, st_crs(marine_and_inland10)), marine_and_inland10)
mapview(goc_coastline)

## I want to get rid of the island coastlines!!

islands_polygons <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/from_hem_POEM_ANP/232_ANP-ITRF08_16062025/232_ANP-ITRF08_16062025.shp")) %>% 
  janitor::clean_names() %>% 
  filter(nombre %in% "Islas del Golfo de California") %>% 
  st_transform(st_crs(goc_coastline))
mapview(islands_polygons)

# erase the island polygons from the municipality polygons, using st_difference, which removes the island area from any municipality that overlaps it
goc_coastline_no_islands <- goc_coastline %>%
  st_difference(st_union(st_buffer(islands_polygons, dist = 1500))) %>% # crop out island areas
  st_make_valid()
mapview(goc_coastline_no_islands)
```


```{r}
## find length of the coastline for each region using `st_length`
st_crs(goc_coastline_no_islands) # meters, Mexico ITRF2008 / LCC 

## calculate coastline length per region
coastline_by_rgn <- goc_coastline_no_islands %>%
  st_intersection(marine_and_inland10 %>% select(rgn_id)) %>%  # assign rgn_id to each coastline segment
  mutate(segment_length_m = as.numeric(st_length(.))) %>%  # length of each segment in meters
  st_drop_geometry() %>%
  group_by(rgn_id) %>%
  summarize(coastline_length_m = sum(segment_length_m, na.rm = TRUE), .groups = "drop") %>%
  mutate(coastline_length_km = coastline_length_m / 1000)
```

```{r}
## join coastline length to total kg estimates and divide
total_pesticides_standardized <- total_ag_pesticides_kg %>%
  left_join(coastline_by_rgn %>% select(rgn_id, coastline_length_km), by = "rgn_id") %>%
  mutate(kg_per_km_coastline = ag_pesticide_kg / coastline_length_km)

summary(total_pesticides_standardized)
```

```{r}
## quick plot to see standardized values over time
ggplot(total_pesticides_standardized, aes(x = year, y = kg_per_km_coastline, color = rgn_id)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  scale_color_manual(values = rgn_colors) +
  scale_x_continuous(breaks = seq(2001, 2024, 2)) +
  labs(
    title = "Total pesticides (kg) standardized by coastline length",
    x = "year",
    y = "kg pesticides per km of coastline"
    #color = "region"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

## versus un-standardized:
ggplot(total_ag_pesticides_kg, aes(x = year, y = ag_pesticide_kg, color = rgn_id)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  scale_color_manual(values = rgn_colors) +
  scale_x_continuous(breaks = seq(2001, 2024, 2)) +
  labs(
    title = "Total pesticides (kg) by region",
    x = "year",
    y = "kg pesticides"
    #color = "region"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

## save to mazu
write_csv(total_pesticides_standardized, here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/pesticides/output","total_kg_pesticides_standardized_coastline.csv"))
```




## Current status

$$current\ status\ _{pesticides} = 1 - \frac{pesticides\ _{current} - pesticides\ _{ideal}}{pesticides\ _{maximum} - pesticides\ _{ideal}}$$

Where the ideal is zero, and the maximum is the global maximum of kg estimated across all years and regions.

```{r}
### using the maximum, not one value but 5
top5_avg_kg <- total_pesticides_standardized %>%
  slice_max(kg_per_km_coastline, n = 5) %>%
  pull(kg_per_km_coastline) %>%
  mean()

current_status_pesticides <- total_pesticides_standardized %>% 
  mutate(
    ideal = 0,
    maximum = top5_avg_kg,
    current_status = (1 - ((kg_per_km_coastline - ideal) / (maximum - ideal)))* 100,
    current_status_final = ifelse(current_status < 0, 0, current_status)
  )

## save to mazu
write_csv(current_status_pesticides, here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/pesticides/output","cw_chemical_pesticide_current_status.csv"))
```


### plotly

```{r}
library(plotly)

rgn_colors <- c(
  Region_1 = "#C03729",
  Region_2 = "#E68C7C",
  Region_3 = "#FC8F24",
  Region_4 = "#dcaa38",
  Region_5 = "#cbd2a0",
  Region_6 = "#98a25a",
  Region_7 = "#747428",
  Region_8 = "#627391",
  Region_9 = "#8d61a9"
)

## plot the pesticide cs over time
pesticide_plotly <- plotly::plot_ly(
  current_status_pesticides, 
  x = ~year, 
  y = ~current_status_final, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Clean Waters: Pesticide Current Status",
    xaxis = list(title = "ano // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines( 
    x = range(current_status_pesticides$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
pesticide_plotly
```

