# CMSY Stock Assesments
# NOTE: Must run CMSY on personal machine. Can generate the main files here though. 

################################################################################ 

For questions contact: Erica Ferrer, Ph.D. at emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################ 

# STEP 0: PRE-WORK ------------------------------

```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```

## Review Instructional Files:
```{r Review the CMSY protocol developed by Ferrer et al. (2021)}
# Absolute path to your PDF
pdf_path <- normalizePath(
  "~/ohi_goc_prep/FP/CMSY_Analysis/CMSY_InstructionalFiles/Copy of Ferrer et al. (2021) CMSY Protocol.pdf",
  winslash = "/", mustWork = TRUE
)

# Convert to file URL
file_url <- paste0("file:///", URLencode(pdf_path, reserved = TRUE))

# Open in default browser
browseURL(file_url)
```

# STEP 1: CREATE FILE 1 - Landings Timeseries  ------------------------------
- ABOUT FILE 1: File with annual landings in kg (and n_count observations) grouped by year per "stock". 
- Here, stock is defined as the total population of GENUS X fished within the Gulf of California from Baja to northern Jalisco, not including landings from the Pacific.
- I’m going to run CMSY at the genus level. I actually think it’s a safer more reliable bet than species-level info and because – in many cases – fishermen seem to target taxa of the same genus interchangeably.
- This file was created in /home/ferrer/ohi_goc_prep/FP/CONAPESCA_CleanUp/CONAPESCA_CleaningWorkflow_May2026.Rmd

## Re-generate time series (from 2004-2024) per genus to use in CMSY [FINAL DRAFT: May 26 2026]

```{r Load GOC_MarineLandings dataframe as needed}
GOC_MarineLandings0<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Marine.Landings_Sector.Sorted.Georef_2000.present_May262026.rds"))
# 3246583 obs., 61 vars.
```

```{r Filter out marine sectors of interest}
table(GOC_MarineLandings0$sector_estimate_draft)

MarineFishery_Sectors<- c("marine: fishery sector yet to be classified",
                          "marine: small-scale capture",
                          "marine: large-scale capture")

GOC_Landings_mar.fisheries<- GOC_MarineLandings %>% 
  filter(sector_estimate_draft %in% MarineFishery_Sectors)
# 3107603 obs., 62 vars.

range(GOC_Landings_mar.fisheries$year)
```

```{r Sort by genera}
table(GOC_MarineLandings0$estimated_genus_name)
# Some of these genera don't appear very often. I want to filter genera for which we have more than 10 years worth of data and > 10 records per year.

# Summarize per genus and year
GOC_Genera_summary <- GOC_MarineLandings0 %>%
  filter(!is.na(estimated_genus_name)) %>%
  select(estimated_genus_name, Major_Functional_Group, year) %>%
  group_by(estimated_genus_name, Major_Functional_Group, year) %>%
  summarise(
    n_records = n(),
    .groups = "drop"
  )

table(GOC_Genera_summary$Major_Functional_Group)
table(GOC_Genera_summary$estimated_genus_name)
unique(GOC_Genera_summary$estimated_genus_name)
# Great! Only 2 freshwater taxa in here, out of 197 taxa... Let's see if they drop out below:

# - Restrict to years ≥ 2010
# - For each genus × functional group × year and check if that year meets the threshold (>10 records)
# - Count how many qualifying years each genus has
# - Keep genera with ≥10 such years


#### DETERMINE "STOCKS OF INTEREST" ####
GOC_Genera_filtered <- GOC_Genera_summary %>%
  filter(year >= 2010) %>%
  filter(!estimated_genus_name %in% c("unknown", "NEI")) %>%
  
  # count qualifying years per genus
  group_by(estimated_genus_name, Major_Functional_Group) %>%
  filter(n_records > 10) %>%   # only keep years that meet threshold
  summarise(n_qualifying_years = n_distinct(year), .groups = "drop") %>%
  
  # keep genera with >=10 qualifying years
  filter(n_qualifying_years >= 10)
# 93 obs. / genera and 3 vars.

Stocks_of_Interest<- GOC_Genera_filtered$estimated_genus_name
# Yes, the freshwater genera (i.e. Mylopharyngodon, Chirostoma) drop out.
# Some of the borderline genera (e.g. Anguilla, Gillichthys, Morone, Oncorhynchus) also drop out.
# 93 stocks of interest!
```

## Filter records for Stocks_of_Interest from GOC_MarineLandings dataframe:

```{r Filter dataset to generate time series for Stocks_of_Interest}
# Keep only genera that meet the criteria in the main dataset
CMSY_Landings <- GOC_MarineLandings %>%
  filter(year < 2025) %>%
  filter(estimated_genus_name %in% Stocks_of_Interest)
# 2256949 obs., 57 vars.
```

## Generate time series for Stocks_of_Interest:

```{r Generate time series}
colnames(CMSY_Landings)

CMSY_Landings_file1 <- CMSY_Landings %>%
  filter(!is.na(estimated_genus_name)) %>%
  # Summarize by year + genus
  group_by(year, estimated_genus_name) %>%
  summarise(
    total_biomass_kg       = sum(weight_kg_corrected, na.rm = TRUE),
    total_revenue          = sum(value, na.rm = TRUE)

  ) %>%
  arrange(estimated_genus_name, year)

dim(CMSY_Landings_file1)
# 2169 obs., 4 obs.

table(CMSY_Landings_file1$estimated_genus_name)
# This looks great! 93 genera or "stocks" that are fairly well sampled from 2000-2024.
```

## Add family and major group info to Stocks_of_Interest:
```{r Load in helpful new FAOFamilyKey}
FAO_Family.Key<- read.csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/FAO_FamilyGenusKey_COMPLETE_Apr042026.csv")

# Trim key:
colnames(FAO_Family.Key)

FAO_Family.Key<- FAO_Family.Key %>%
  select(estimated_genus_name, Major_Functional_Group, estimated_taxa_group, new_estimated_family_name)
```

```{r Merge key to CMSY_Landings_file1 and rename columns as needed}
CMSY_Landings_file1 <- left_join(CMSY_Landings_file1, FAO_Family.Key, 
                           by = "estimated_genus_name")
# 2169 obs., 7 obs. (as expected)

# Rename estimated_genus_name to "Stock"
colnames(CMSY_Landings_file1)

CMSY_Landings_file1 <- CMSY_Landings_file1 %>%
  rename(Stock = estimated_genus_name)
```


# SAVE FILE 1:

```{r Save these time series CMSY_Landings_file1}
write.csv(CMSY_Landings_file1, row.names= FALSE, file= "/home/ferrer/ohi_goc_prep/FP/CMSY_Analysis/CMSY_Final_May2026/Model_Input_Files/File1_CMSY_LandingsPerStock_May262026.csv")
# 2169 obs., 7 obs.
```

# Okay, timeseries are ready for CMSY tomorrow (May 27 2026)!


# STEP 2A: CREATE FILE 2 - Fill in metadata from priors  ------------------------------
FILE 2: Create a meta-data file that includes all the information about stock. This file can probably be trimmed down but for now I recommend formatting it exactly as it appears from before.

```{r Create File 2 based on File 1}
# STEP 1: CREATE CMSY_Metadata_file2
# Which taxa (stocks) do we want to account for in our File 2?
Stock_names<- CMSY_Landings_file1 %>%
  select(Stock, Major_Functional_Group, estimated_taxa_group, new_estimated_family_name) %>%
  distinct()
# 95 unique genera / stocks.

## Read in the column names from blueprint of File 2.
File2_blueprint <- read.csv("/home/ferrer/ohi_goc_prep/FP/CMSY_Analysis/CMSY_InstructionalFiles/CMSY_File2_BluePrint_Apr032026.csv")
File2_headers <- names(File2_blueprint)

## Create an empty dataframe with the same File2_headers
CMSY_Metadata_file2 <- data.frame(matrix(ncol = length(File2_headers), nrow = length(Stock_names$Stock)))
colnames(CMSY_Metadata_file2) <- File2_headers

# STEP 2: FILL CMSY_Metadata_file2
## Populate CMSY_Metadata_file2 with info from Stock_names and File2_blueprint
CMSY_Metadata_file2$Stock <- Stock_names$Stock
CMSY_Metadata_file2$Major_Functional_Group <- Stock_names$Major_Functional_Group
CMSY_Metadata_file2$estimated_taxa_group <- Stock_names$estimated_taxa_group
CMSY_Metadata_file2$estimated_family_name <- Stock_names$new_estimated_family_name

# Fill in some of the other columns:
CMSY_Metadata_file2$Region <- unique(File2_blueprint$Region)
CMSY_Metadata_file2$Source <- unique(File2_blueprint$Source)
CMSY_Metadata_file2$Comment <- unique(File2_blueprint$Comment)

CMSY_Metadata_file2$Sufficient_No_Years <- TRUE
CMSY_Metadata_file2$Yes_Run <- c("Yes")
CMSY_Metadata_file2$btype <- c("None")
CMSY_Metadata_file2$force.cmsy <- c("TRUE")
```

```{r Save this file as UNDER CONSTRUCTION so I can work with it offline and use it to generate priors}
write.csv(CMSY_Metadata_file2, row.names= FALSE,
          "/home/ferrer/ohi_goc_prep/FP/CMSY_Analysis/CMSY_Final_May2026/Model_Input_Files/File2_CMSY_StockMetadata_UNDERCONSTRUCTION_May262026.csv")
```


# STEP 2B: Generate Model Priors using other .R files ------------------------------
ExpertOpinion_CollatingPriors_May272026.Rmd
1_StartingBiomass_EstimatePriors_Apr032026.R
2_IntermediateYear_EstimatePriors_Apr032026.R
3_CurrentRelativeBiomass_EstimatePriors_Apr032026.R


# STEP 2C: COMPLETE FILE 2 with infro from new generated priors.   ------------------------------

```{r lowercase_function}
# Turns all entries lower-case
lowercase_function<- function(x){
  if(is.character(x)){
    tolower(x)
  } else if (is.factor(x)){
    tolower(as.character(x))
  } else {
    x
  }
}
```

```{r Read in additional files to complete file 2}
# Load File 2: CMSY_StockMetadata_file2
Stock_Metadata = read.csv("/home/ferrer/ohi_goc_prep/FP/CMSY_Analysis/CMSY_Revised_April2026/Model_Input_Files/File2_CMSY_StockMetadata_UNDERCONSTRUCTION_Apr032026.csv")
Stock_Metadata<- as.data.frame(lapply(Stock_Metadata, lowercase_function))
# 95 obs., 39 vars.

# Load in generated priors:

Priors_StartingBiomass <- read.csv("~/ohi_goc_prep/FP/CMSY_Analysis/CMSY_Revised_April2026/Generating_Model_Priors/Generated_Priors/1_StartingBiomassPriors_fromFAO_Apr042026.csv")
Priors_StartingBiomass<- as.data.frame(lapply(Priors_StartingBiomass, lowercase_function))
colnames(Priors_StartingBiomass)[1]<- "Stock_family"
colnames(Priors_StartingBiomass)[4]<- "FisheryStatus_Circa2000"
# 106 obs., 4 vars.

Priors_IntermediateYear <- read.csv("~/ohi_goc_prep/FP/CMSY_Analysis/CMSY_Revised_April2026/Generating_Model_Priors/Generated_Priors/2_IntermediateYearsPriors_fromLandings_Apr042026.csv")
Priors_IntermediateYear<- as.data.frame(lapply(Priors_IntermediateYear, lowercase_function))
# 95 obs., 5 vars.

Priors_IntEnd.Conditions <- read.csv("~/ohi_goc_prep/FP/CMSY_Analysis/CMSY_Revised_April2026/Generating_Model_Priors/Generated_Priors/3_IntermediateEndConditionsPriors_Apr042026.csv")
Priors_IntEnd.Conditions<- as.data.frame(lapply(Priors_IntEnd.Conditions, lowercase_function))
# 95 obs., 5 vars.


# UNDER CONSTRUCITON: RETURN TO THIS LATER - ERICA Apr 4 2026
Priors_Resilience <- read.csv("~/ohi_goc_prep/FP/CMSY_Analysis/CMSY_Revised_April2026/Generating_Model_Priors/Priors_Data/ResilienceEstimates_fromEMF_Dec2025.csv")
Priors_Resilience<- as.data.frame(lapply(Priors_Resilience, lowercase_function))
# 98 obs., 3 vars.
```

```{r Stitch in newly generated priors}
# Add Starting Biomass Priors
colnames(Priors_StartingBiomass)

Stock_Metadata <- Stock_Metadata %>%
  left_join(Priors_StartingBiomass,
            by = c("estimated_family_name" = "Stock_family"),
            suffix = c("", ".new")) %>%
  mutate(
    st_b.low = coalesce(st_b.low.new, st_b.low),
    st_b.hi  = coalesce(st_b.hi.new, st_b.hi)
  ) %>%
  select(-st_b.low.new, -st_b.hi.new)
# 95 obs., 40 vars.


# Add Intermediate Year Priors
colnames(Priors_IntermediateYear)

Stock_Metadata <- Stock_Metadata %>%
  left_join(Priors_IntermediateYear, by = "Stock", suffix = c("", "_overwrite")) %>%
  mutate(
    int_yr = int_yr_overwrite,
    StartYear = StartYear_overwrite,
    EndYear = EndYear_overwrite,
    Sufficient_No_Years = Sufficient_No_Years_overwrite
  ) %>%
  select(-ends_with("_overwrite"))


# Add Intermediate & End Conditions Priors
colnames(Priors_IntEnd.Conditions)

Stock_Metadata <- Stock_Metadata %>%
  left_join(Priors_IntEnd.Conditions, by = "Stock", suffix = c("", "_overwrite")) %>%
  mutate(
    int_b.low = int_b.low_overwrite,
    int_b.hi = int_b.hi_overwrite,
    end_b.low = end_b.low_overwrite,
    end_b.hi = end_b.hi_overwrite
  ) %>%
  select(-ends_with("_overwrite"))


# Add Resilience Priors - UNDER CONSTRUCTION, Apr 4 2026
colnames(Priors_Resilience)

Stock_Metadata <- Stock_Metadata %>%
  left_join(Priors_Resilience, by = "Stock", suffix = c("", "_overwrite")) %>%
  mutate(
    Resilience = Resilience_overwrite,
  ) %>%
  select(-ends_with("_overwrite"))
```

```{r Finally lets clean up File 2 before saving}
colnames(Stock_Metadata)

CMSY_StockMetadata_file2<- Stock_Metadata %>% 
  select(Region, Stock, estimated_family_name, Major_Functional_Group, estimated_taxa_group,
         Landings_Datasource= Source, 
         Sufficient_No_Years,
         StartYear, EndYear, int_yr,
         btype,
         Flim, Fpa, Blim, Bpa, Bmsy, B40,
         FMSY, MSY, MSYBtrigger, M,
         Fofl, last_F,
         q.start, q.end,
         FisheryStatus_Circa2000, st_b.low, st_b.hi, 
         int_b.low, int_b.hi, end_b.low, end_b.hi,
         r.low, r.hi, 
         Resilience, Resilience_Source,
         Yes_Run, Comment)
# 95 obs., 38 vars.
```

# SAVE FILE 2:
```{r Save these time series CMSY_Landings_file1}
write.csv(CMSY_StockMetadata_file2, row.names= FALSE, file= "/home/ferrer/ohi_goc_prep/FP/CMSY_Analysis/CMSY_Revised_April2026/Model_Input_Files/File2_CMSY_StockMetadata_Apr042026.csv")
# 95 obs., 38 vars.
```

# Stopped here April 4 2026.




# OTHER: Misc. work.

```{r How much of FP will these CMSY stocks account for?}
Landings_CMSY_TEST <- Landings_GoC_NoFresh.NoMariculture %>%
  mutate(estimated_genus_name = tolower(estimated_genus_name))

CMSY_Stocks <- CMSY_Metadata_file2_merged %>%
  filter(Yes_Run == 1) %>%
  pull(Stock) %>%
  unique()

CMSY_Stocks_hypothetical <- c(
  CMSY_Stocks, 
  "sardinops", 
  "opisthonema", 
  "sarda",
  "scomber"
)

# Calculate total from genera landings identified in the CMSY exercise:
# NOTE TO SELF: Come back to this. Find priors for these four taxa! They represent a huge component of total landings.
Landings_CMSY_summary <- Landings_CMSY_TEST %>%
  filter(year %in% 2015:2024) %>%           # only last 10 years
  mutate(is_CMSY = estimated_genus_name %in% CMSY_Stocks_hypothetical) %>%  # flag CMSY genera
  summarise(
    total_landings = sum(weight_kg_corrected, na.rm = TRUE),                # total landings
    CMSY_landings  = sum(weight_kg_corrected[is_CMSY], na.rm = TRUE)       # landings from CMSY genera
  ) %>%
  mutate(percent_CMSY = 100 * CMSY_landings / total_landings)    # percentage
  
Landings_CMSY_summary
# If I can sort out priors for these four extra txa, CMSY could account for ~50% of all GoC landings in the last five years.
```

```{r How much of FP will these CMSY stocks account for?}
Landings_CMSY_TEST <- Landings_GoC_NoFresh.NoMariculture %>%
  mutate(estimated_genus_name = tolower(estimated_genus_name))

CMSY_Stocks <- CMSY_Metadata_file2_merged %>%
  filter(Yes_Run == 1) %>%
  pull(Stock) %>%
  unique()

CMSY_Stocks_hypothetical <- c(
  CMSY_Stocks, 
  "sardinops", 
  "opisthonema", 
  "sarda",
  "scomber"
)

# Calculate total from genera landings identified in the CMSY exercise:
# NOTE TO SELF: Come back to this. Find priors for these four taxa! They represent a huge component of total landings.
Landings_CMSY_summary <- Landings_CMSY_TEST %>%
  filter(year %in% 2015:2024) %>%           # only last 10 years
  mutate(is_CMSY = estimated_genus_name %in% CMSY_Stocks_hypothetical) %>%  # flag CMSY genera
  summarise(
    total_landings = sum(weight_kg_corrected, na.rm = TRUE),                # total landings
    CMSY_landings  = sum(weight_kg_corrected[is_CMSY], na.rm = TRUE)       # landings from CMSY genera
  ) %>%
  mutate(percent_CMSY = 100 * CMSY_landings / total_landings)    # percentage
  
Landings_CMSY_summary
# If I can sort out priors for these four extra txa, CMSY could account for ~50% of all GoC landings in the last five years.
```


# RUN CMSY: Ran on personal machine. Here are the outputs from that analysis:

```{r Load in CMSY asessment results}
# Actual B/Bmsy output from CMSY:
CMSY_Timeseries_original<- read.csv("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_StockStatus_Scores/CMSY_Scores/Output_Files/CMSY_Assessment_Timeseries_Dec102025.csv")
# 775 obs., 13 vars.
```