# >> PICK 2: CMSY Stock Assesments <<
Must run CMSY on personal machine. Can generate the main files here though. 

```{r Review the CMSY protocol developed by Ferrer et al. (2021)}
# Absolute path to your PDF
pdf_path <- normalizePath(
  "~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Copy of Ferrer et al. (2021) CMSY Protocol.pdf",
  winslash = "/", mustWork = TRUE
)

# Convert to file URL
file_url <- paste0("file:///", URLencode(pdf_path, reserved = TRUE))

# Open in default browser
browseURL(file_url)
```

# GENERATING FILES 1 & 2 FROM SCRATCH

NOTES TO SELF - Dec 09 2025:

FILE 1: Create a file with annual landings in kg (and n_count observations) grouped by year per "stock". Here, stock is defined as the total population of GENUS X fished within the Gulf of California from Baja to northern Jalisco, not including landings from the Pacific.

- At present, we have had some issues with keying out species and landing_sites. Thus, whatever timeseries we have here is likely to be an UNDERESTIMATE of total landings for all genera in the GoC. I plan to re-make this file (File 1) in the new year once we have more key info.

- For now, I’m going to run CMSY at the genus level for now, since I actually think it’s a safer more reliable bet than species-level info and because – in many cases – fishermen seem to target taxa of the same genus interchangeably.

FILE 2: Create a file with OHI_Region, Stock, Group = Name, English Name, Scientific Name, Source, and a whole bunch of year columns that you populate using the various R files shared by Alfredo. This file can probably be trimmed down but for now I recommend formatting it exactly as it appears from before.

# Create FILE 1 - Timeseries
```{r File 1 - Stock Timeseries (JC data)}
# Load file, only if needed:
GoC_Landings_for_CMSY<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/fis/v2025/int/GOC_FIS.landings_CLEAN_as.of.Dec092025.rds"))

GoC_Landings_for_CMSY_original<- GoC_Landings_for_CMSY

# GoC_Landings_for_CMSY<- Landings_GoC_NoFresh.NoMariculture
# 3163120 obs., 50 vars.
```

```{r Basic checks ahead of CMSY}
# Some basic checks:
table(GoC_Landings_for_CMSY$fleet)
# large_scale small_scale 
#      142404      975060 

table(GoC_Landings_for_CMSY$acuaculture_production) 
# no aquaculture. As expected!

table(GoC_Landings_for_CMSY$freshwater_flag) 
# no freshwater flags. As expected!

table(GoC_Landings_for_CMSY$GoC) 
table(GoC_Landings_for_CMSY$GulfCal) 
# Everything in Gulf of California, that we know of!

table(GoC_Landings_for_CMSY$year)
# Plenty of data years 2000-2025

unique(GoC_Landings_for_CMSY$estimated_genus_name) 
# Okay. There are a few genera we're going to toss...
```

```{r Basic filters ahead of CMSY}
#### REMOVE FRESHWATER GENERA WE ARE NOT INTERESTED ####
genus_exclude<- c(
  "Cyprinus", # common carp; strictly freshwater.
  "Macrobrachium", # freshwater prawns/shrimps; mostly freshwater.
  "Hypophthalmichthys", # Asian carp; freshwater.
  "Chirostoma", # “charales,” endemic to freshwater lakes in Mexico.
  "Carassius", # goldfish/crucian carp; freshwater.
  "Micropterus", # black bass; freshwater.
  "Lepomis", # sunfish; freshwater.
  "Cyprinodon", # pupfish; freshwater.
  "Rana", # frogs
  "NEI", # not classed yet
  "Especies", # not specific enough to be anything
  NA # not classed
  )

# Flag genera to exclude:
GoC_Landings_for_CMSY <- GoC_Landings_for_CMSY %>%
  mutate(
    exclude_flag = if_else(
      !is.na(estimated_genus_name) & estimated_genus_name %in% genus_exclude,
      TRUE,
      FALSE
    )
  )

subset<- GoC_Landings_for_CMSY %>% filter(exclude_flag== TRUE)
# Dropping 710533 out of 3163120 total records (~22%) for now. A lot of these are NEI that will be classed in the new year!

GoC_Landings_for_CMSY<- GoC_Landings_for_CMSY %>% filter(exclude_flag== FALSE)
# 2452587 obs., 51 vars.


#### REMOVE RARE GENERA ####
table(GoC_Landings_for_CMSY$estimated_genus_name)
# Some of these genera don't appear very often. I want to filter genera for which we have more than 10 years worth of data and > 10 records per year.

# Summarize per genus and year
Genus_summary <- GoC_Landings_for_CMSY %>%
  filter(!is.na(estimated_genus_name)) %>%
  group_by(estimated_genus_name, year) %>%
  summarise(
    n_records = n(),
    .groups = "drop"
  )

# Count how many years have more than 10 records, and remove those genera with fewer than 10 years of 10 records:
Genus_filtered <- Genus_summary %>%
  filter(n_records > 10) %>%
  group_by(estimated_genus_name) %>%
  summarise(n_years = n(), .groups = "drop") %>%
  filter(n_years > 10)

Genera_of_Interest<- Genus_filtered$estimated_genus_name
# 98 remaining genera

# Keep only genera that meet the criteria in the main dataset
GoC_Landings_for_CMSY <- GoC_Landings_for_CMSY %>%
  filter(estimated_genus_name %in% Genera_of_Interest)
# 2225315 obs., 51 vars.

#### Filter out OHI Region = NA ####
sum(is.na(GoC_Landings_for_CMSY$OHI_Region_No))
# Not applicable, all entries have an OHI Region number.
```

```{r Mitigate genera key issues}
# NOTE TO SELF: Clean this up in the new year. This is a band aid solution to larger key issues.

#### GENUS KEY ####
# NOTE TO SELF: Return to this when needed (Dec 09 2025). Returned the next day, Dec 10 2025.
Genus_MajorGroup_Key <- GoC_Landings_for_CMSY %>%
  select(recorded_taxa_group, estimated_genus_name, IHH_functional_major_groups_sm) %>%
  distinct(recorded_taxa_group, estimated_genus_name, IHH_functional_major_groups_sm, .keep_all = TRUE)
# 161 obs.

# See the issue here? One genus could be associated with multiple recorded_taxa_groups. We need the unique combination of estimated_genus_name & recorded_taxa_group to key out the larger dataset. But, in the end, we'll want to be able to group by ESTIMATED_taxa_group to avoid issues of splitting the catch from a single genus across two larger groups.

# ---- Create Major_Functional_Group values based IHH major functional groups assigned to recorded_taxa_group ----

# Step 1: Assign one Major_Functional_Group per recorded_taxa_group, respecting the "OTRAS" category as inherently different.
Genus_MajorGroup_Key <- Genus_MajorGroup_Key %>%
  group_by(recorded_taxa_group) %>%
  mutate(
    # Step 1: assign group-level functional group only for non-OTRAS
    group_major = {
      tbl <- table(IHH_functional_major_groups_sm)
      tbl <- tbl[!is.na(names(tbl)) & names(tbl) != ""]
      if(length(tbl) == 0) NA_character_ else names(tbl)[which.max(tbl)]
    },
    Major_Functional_Group = if_else(
      recorded_taxa_group != "OTRAS",
      group_major,
      NA_character_
    )
  ) %>%
  ungroup() %>%
  select(-group_major) %>%
  
  # Step 2: assign genus-level functional group (including OTRAS)
  group_by(estimated_genus_name) %>%
  mutate(
    genus_major = {
      # Use either already assigned Major_Functional_Group or fallback to IHH_functional_major_groups_sm
      tbl <- table(
        c(
          Major_Functional_Group[!is.na(Major_Functional_Group)],
          IHH_functional_major_groups_sm[!is.na(IHH_functional_major_groups_sm)]
        )
      )
      if(length(tbl) == 0) NA_character_ else names(tbl)[which.max(tbl)]
    },
    Major_Functional_Group = if_else(
      is.na(Major_Functional_Group),
      genus_major,
      Major_Functional_Group
    )
  ) %>%
  ungroup() %>%
  select(-genus_major)
  
# Step 3: Enforce genus-level consistency for Major_Functional_Group
Genus_MajorGroup_Key <- Genus_MajorGroup_Key %>%
  group_by(estimated_genus_name) %>%
  mutate(
    # Only apply to genera that appear more than once
    Major_Functional_Group = if(n() > 1) {
      # Choose the most frequent functional group within the genus
      tbl <- table(Major_Functional_Group)
      tbl <- tbl[names(tbl) != "" & names(tbl) != "NA"]
      if(length(tbl) == 0) NA_character_ else names(tbl)[which.max(tbl)]
    } else {
      Major_Functional_Group
    }
  ) %>%
  ungroup()
  
# Step 4: Manual patches for specific taxonomic groups
Genus_MajorGroup_Key <- Genus_MajorGroup_Key %>%
  mutate(
    Major_Functional_Group = case_when(
      recorded_taxa_group == "ALGAS" ~ "Marine algae",
      recorded_taxa_group %in% c("PETO", "TIBURON", "CAZON") ~ "Sharks",
      TRUE ~ Major_Functional_Group
    )
  )

# Step 5: Assign estimated_taxa_group by looking at the most common recorded_taxa_group for each genus
Genus_MajorGroup_Key <- Genus_MajorGroup_Key %>%
  group_by(estimated_genus_name) %>%
  mutate(
    # Count the most frequent recorded_taxa_group for this genus
    most_common_group = {
      tbl <- table(recorded_taxa_group)
      names(tbl)[which.max(tbl)]
    },
    estimated_taxa_group = most_common_group
  ) %>%
  ungroup() %>%
  select(-most_common_group)
    
# 161 obs., 4 vars.

# TEST<- Genus_MajorGroup_Key %>%
#   select(estimated_genus_name,  Major_Functional_Group) %>%
#   distinct()
# 98 obs. = 98 genera, as expected.

Genus_MajorGroup_Key<- Genus_MajorGroup_Key %>% 
  select(recorded_taxa_group, estimated_genus_name, estimated_taxa_group, Major_Functional_Group) %>% 
  distinct()
# 134 obs., 3 vars.

# MERGE this key to the larger dataset
GoC_Landings_for_CMSY <- left_join(
  GoC_Landings_for_CMSY,
  Genus_MajorGroup_Key,
  by = c("recorded_taxa_group", "estimated_genus_name")
)
# 2225315 obs., 53 vars. as expected!

# A few quick sanity checks:
subset<- GoC_Landings_for_CMSY %>% filter(is.na(Major_Functional_Group))
sum(is.na(GoC_Landings_for_CMSY$Major_Functional_Group))
table(GoC_Landings_for_CMSY$estimated_taxa_group)
sum(is.na(GoC_Landings_for_CMSY$estimated_taxa_group))
```

```{r Save GoC_Landings_for_CMSY to be used for weighting scores later}
# I want to maintain this exact dataframe since it was filtered, etc.
# saveRDS(GoC_Landings_for_CMSY, "/home/shares/ohi/OHI_GOC/goal_prep/fp/fis/v2025/int/GOC_Landings_for_CMSY_Dec102025.rds")
```

```{r Summary table}
# Turn off scientific notation:
options(scipen = 999)

# Create timeseries FILE 1:
colnames(GoC_Landings_for_CMSY)
unique(GoC_Landings_for_CMSY$estimated_genus_name)

TotalLandings_AnnualSummary_byGenus <- GoC_Landings_for_CMSY %>%
  select(year, estimated_genus_name, weight_kg_corrected) %>%
  group_by(year, estimated_genus_name) %>%
  summarise(
    Total_Landed_Biomass = sum(weight_kg_corrected, na.rm = TRUE),
    n_Records = n(),
    .groups = "drop"  # optional: ungroup after summarising
  )
# 2401 obs., 4 vars.

TotalLandings_AnnualSummary_byGenus <- GoC_Landings_for_CMSY %>%
  select(year, Major_Functional_Group, estimated_taxa_group, estimated_genus_name, weight_kg_corrected) %>%
  group_by(year, Major_Functional_Group, estimated_taxa_group, estimated_genus_name) %>%
  summarise(
    Total_Landed_Biomass = sum(weight_kg_corrected, na.rm = TRUE),
    n_Records = n(),
    .groups = "drop"  # optional: ungroup after summarising
  )
# 2401 obs., 6 vars. GOOD!!!

table(TotalLandings_AnnualSummary_byGenus$Major_Functional_Group)
#                   Cephalopods                   Crustaceans        Demersal marine fishes             Diadromous fishes                  Marine algae 
#                            26                            41                          1153                            26                            23 
# Miscellaneous aquatic animals    Molluscs excl. Cephalopods         Pelagic marine fishes                        Sharks               Shrimps, prawns 
#                            44                           222                           472                           269                           125 
# Lots of demersal marine fishes!

File1_df<- TotalLandings_AnnualSummary_byGenus

# Re-name column
colnames(File1_df)[colnames(File1_df) == "estimated_genus_name"] <- "Stock"
colnames(File1_df)[colnames(File1_df) == "year"] <- "Year"

# Finally, and just for now, filter out data from 2025
# NOTE TO SELF: Come back to this in subsequent updates, once we have full data for 2025.
File1_df<- File1_df %>% filter(Year<2025)

CMSY_GenusKey<- File1_df %>% 
  select(Major_Functional_Group, estimated_taxa_group, Stock) %>%
  distinct()
```

## SAVE FILE 1
```{r SAVE FILE 1 & CMSY Genus Key}
# 
write.csv(File1_df, "~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Input_Files/File1_Dec092025.csv", row.names = FALSE)
# write.csv(CMSY_GenusKey, "~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Instructional_Files/CMSY_GenusKey_Dec102025.csv", row.names = FALSE)
```

# Create FILE 2 - Metadata & Model Priors
```{r Create File 2 based on File 1}
# STEP 1:
# Which taxa (stocks) do we want to account for in our File 2?
Stock_names<- unique(File1_df$Stock)
# 98 unique genera / stocks

## Read in the column names of File 2 from Ferrer et al. (2022) CSV file
headers <- names(read.csv("FIS/v2025/OHI_CMSY_Scores/Instructional_Files/CMSY_File2_BluePrint_Dec092025.csv", nrows = 1))

## Create an empty dataframe with the same headers
File2_df <- data.frame(matrix(ncol = length(headers), nrow = 0))
colnames(File2_df) <- headers

## Add Stock_names to the 'Stock' column
File2_df <- File2_df[1:length(Stock_names), ]  # Expand rows
File2_df$Stock <- Stock_names

## Reset rownames / the index column
rownames(File2_df) <- NULL


# STEP 2: TRIME AND RENAME AS NECESSARY
# Trim some of the columns off this file: only if necessary.

# Re-name columns: only if necessary.
# colnames(File2_df)[colnames(File2_df) == "Keep"] <- "Yes_Run"
# colnames(File2_df)[colnames(File2_df) == "stb.low"] <- "st_b.low"
# colnames(File2_df)[colnames(File2_df) == "stb.hi"] <- "st_b.hi"


# STEP 3: FILL AS NECESSARY
# Fill in some of the columns off this file:
File2_df$Region <- c("GOC")
File2_df$Class <- c("TBD")
File2_df$Source <- c("JC_data")
File2_df$btype <- c("None")
File2_df$force.cmsy <- c("TRUE")


# STEP 4:
# For species with less than 10 years of data, do not keep (Sufficient_No_Years = 0)
StockYears_per_Taxa <- File1_df %>%
  group_by(Stock) %>%
  summarise(n_Stock_Years = n_distinct(Year), .groups = "drop")

StockYears_per_Taxa <- StockYears_per_Taxa %>%
  mutate(Sufficient_No_Years = ifelse(n_Stock_Years < 10, 0, 1))

# How many stocks will we be keeping based on number of years?
table(StockYears_per_Taxa$Sufficient_No_Years)
# All of them, which makes sense since we already applied this filter above.

##  Add this information to the File 2 dataframe: join StockYears_per_Taxa to File2_df by "Stock", bringing in the filled Sufficient_No_Years column
File2_df <- File2_df %>%
  select(-Sufficient_No_Years) %>%  # Remove empty column to avoid duplicates
  left_join(
    StockYears_per_Taxa %>% select(Stock, Sufficient_No_Years),
    by = "Stock"
  )

# STEP 5: ADD IN THE GENUS KEY INFO
CMSY_GenusKey_distinct<- Genus_MajorGroup_Key %>% 
  select(estimated_genus_name, estimated_taxa_group) %>%
  distinct()

File2_df <- File2_df %>%
  left_join(
    CMSY_GenusKey_distinct,
    by = c("Stock"="estimated_genus_name")
  )

# STEP 6:
# Re-order columns: only if necessary.
```

## SAVE FILE 2 BLUE PRINT
```{r SAVE FILE 2 UNDER CONSTRUCTION}
write.csv(File2_df, "~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Input_Files/File2_UNDER_CONSTRUCTION_Dec092025.csv", row.names = FALSE)
```
NEXT STEPS: Fill in the rest of file 2, namely the model priors, using the ancillary files shared with me by Alfredo (Ferrer et al., 2022)
NOTE TO SELF: These model priors are likely to change in the new year but will do for now.

Did some work offline on my local machine (not available for viewing here... yet). This work will help fill out File 2.

```{r Read in additional files to complete file 2}
# Read in file if necessary:
#File2_df <- read.csv("~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Input_Files/File2_UNDER_CONSTRUCTION_Dec092025.csv")

## Additional files needed to file in the metadata
Priors_Pristine.Biomass <- read.csv("~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Model_Priors/Priors_StartingBiomass_fromFAOdata_DRAFT_Dec092025.csv")
Priors_Int.Year <- read.csv("~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Model_Priors/Priors_IntermediateYears_DRAFT_Dec092025.csv")
Priors_IntEnd.Conditions <- read.csv("~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Model_Priors/Priors_IntEndConditions_DRAFT_Dec092025.csv")
Priors_Resilience <- read.csv("~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Model_Priors/Priors_ResilienceValues_DRAFT_Dec092025.csv")
Priors_Resilience <- Priors_Resilience %>% select(-ScientificName)
# Common_Names <- read.csv("Stock and Common Names_Oct 23 2025.csv")
```

```{r lowercase_function}
to_lower_df <- function(df) {
  df %>% mutate(across(where(is.character), tolower))
}
```

```{r Stitch these dataframes together to create complete File2_df}
File2_df        <- to_lower_df(File2_df)
Priors_Pristine.Biomass  <- to_lower_df(Priors_Pristine.Biomass)
Priors_Int.Year <- to_lower_df(Priors_Int.Year)
Priors_Resilience <- to_lower_df(Priors_Resilience)

## MERGE: PRISTINE BIOMASS PRIORS
File2_df_merged <- File2_df %>% 
  left_join(
    Priors_Pristine.Biomass %>% select(estimated_taxa_group, st_b.low, st_b.hi),
    by = "estimated_taxa_group",
    suffix = c(".file2", ".priors")
  ) %>%
  mutate(
    st_b.low = coalesce(st_b.low.file2, st_b.low.priors),
    st_b.hi  = coalesce(st_b.hi.file2,  st_b.hi.priors)
  ) %>% 
  select(-ends_with(".file2"), -ends_with(".priors"))

# NOTE TO SELF: Our priors for pristine biomass, and the FAO timeseries that underly them, are fairly sparse as in missing several major taxonomic groups (like sardine). Return to this in the new year.

## MERGE: INT YEAR PRIORS
File2_df_merged <- File2_df_merged %>%
  left_join(
    Priors_Int.Year %>% 
      select(estimated_taxa_group, int_yr, StartYear, EndYear),
    by = "estimated_taxa_group",
    suffix = c(".file2", ".priors")
  ) %>%
  mutate(
    int_yr               = coalesce(int_yr.file2,               int_yr.priors),
    StartYear            = coalesce(StartYear.file2,            StartYear.priors),
    EndYear              = coalesce(EndYear.file2,              EndYear.priors)) %>%
  select(-ends_with(".file2"), -ends_with(".priors"))

## MERGE: INT END CONDITIONS PRIORS
# NOTE TO SELF: Based on the naming conventions within this file I actually think these are the priors for intermediate and ending biomass. I'll circle back to this and iron out the true meaning and utility of these particular priors in the new year.
# Join and fill missing values
File2_df_merged <- File2_df_merged %>%
  left_join(
    Priors_IntEnd.Conditions,
    by = "estimated_taxa_group",
    suffix = c("", ".priors")
  ) %>%
  mutate(
    int_b.low  = coalesce(int_b.low, int_b.low.priors),
    int_b.hi   = coalesce(int_b.hi, int_b.hi.priors),
    end_b.low  = coalesce(end_b.low, end_b.low.priors),
    end_b.hi = coalesce(end_b.hi, end_b.hi.priors)
  ) %>%
  select(-ends_with(".priors"))


## MERGE: RESILIENCE PRIORS
File2_df_merged <- File2_df_merged %>%
  left_join(
    Priors_Resilience %>% select(estimated_taxa_group, Resilience, Resilience_Source),
    by = "estimated_taxa_group",
    suffix = c(".file2", ".priors")
  ) %>%
  mutate(
    Resilience = coalesce(Resilience.file2, Resilience.priors)
  ) %>%
  select(-Resilience.file2, -Resilience.priors)
```

```{r Finally lets clean up File 2 before saving and running CMSY}
File2_df_merged<- File2_df_merged %>% 
  select(Region, Stock, Landings_Datasource=Source, 
         StartYear, EndYear, int_yr,
         btype,
         Flim, Fpa, Blim, Bpa, Bmsy, B40,
         FMSY, MSY, MSYBtrigger, M,
         Fofl, last_F,
         q.start, q.end,
         st_b.low, st_b.hi, 
         int_b.low, int_b.hi, end_b.low, end_b.hi,
         r.low, r.hi, 
         Resilience, Resilience_Source,
         Yes_Run, Comment)

# Fill in the Yes_Run column based on model priors:
File2_df_merged <- File2_df_merged %>%
  mutate(
    Yes_Run = if_else(
      is.na(Resilience) | is.na(int_b.low) | is.na(st_b.low) | is.na(end_b.low),
      0,
      1
    ),
    Comment = if_else(
      Yes_Run == 0,
      "Incomplete model priors",
      NA_character_
    )
  )

sum(File2_df_merged$Yes_Run==1)
# 38 stocks to run!

# NOTES ON NAMING, Dec 10 2025:
# Need a btype column = None
# Need a force.cmsy column = TRUE
# StartYear and EndYear must actually align with file one time series.
```

```{r How much of FP will these CMSY stocks account for?}
Landings_CMSY_TEST <- Landings_GoC_NoFresh.NoMariculture %>%
  mutate(estimated_genus_name = tolower(estimated_genus_name))

CMSY_Stocks <- File2_df_merged %>%
  filter(Yes_Run == 1) %>%
  pull(Stock) %>%
  unique()

CMSY_Stocks_hypothetical <- c(
  CMSY_Stocks, 
  "sardinops", 
  "opisthonema", 
  "sarda",
  "scomber"
)

# Calculate total from genera landings identified in the CMSY exercise:
# NOTE TO SELF: Come back to this. Find priors for these four taxa! They represent a huge component of total landings.
Landings_CMSY_summary <- Landings_CMSY_TEST %>%
  filter(year %in% 2015:2024) %>%           # only last 10 years
  mutate(is_CMSY = estimated_genus_name %in% CMSY_Stocks_hypothetical) %>%  # flag CMSY genera
  summarise(
    total_landings = sum(weight_kg_corrected, na.rm = TRUE),                # total landings
    CMSY_landings  = sum(weight_kg_corrected[is_CMSY], na.rm = TRUE)       # landings from CMSY genera
  ) %>%
  mutate(percent_CMSY = 100 * CMSY_landings / total_landings)    # percentage
  
Landings_CMSY_summary
# If I can sort out priors for these four extra txa, CMSY could account for ~50% of all GoC landings in the last five years.
```

## SAVE FILE 2 COMPLETE
```{r SAVE FILE 2 COMPLETE}
write.csv(File2_df_merged, "~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Input_Files/File2_Dec092025.csv", row.names = FALSE)
```

```{r How much of FP will these CMSY stocks account for?}
Landings_CMSY_TEST <- Landings_GoC_NoFresh.NoMariculture %>%
  mutate(estimated_genus_name = tolower(estimated_genus_name))

CMSY_Stocks <- File2_df_merged %>%
  filter(Yes_Run == 1) %>%
  pull(Stock) %>%
  unique()

CMSY_Stocks_hypothetical <- c(
  CMSY_Stocks, 
  "sardinops", 
  "opisthonema", 
  "sarda",
  "scomber"
)

# Calculate total from genera landings identified in the CMSY exercise:
# NOTE TO SELF: Come back to this. Find priors for these four taxa! They represent a huge component of total landings.
Landings_CMSY_summary <- Landings_CMSY_TEST %>%
  filter(year %in% 2015:2024) %>%           # only last 10 years
  mutate(is_CMSY = estimated_genus_name %in% CMSY_Stocks_hypothetical) %>%  # flag CMSY genera
  summarise(
    total_landings = sum(weight_kg_corrected, na.rm = TRUE),                # total landings
    CMSY_landings  = sum(weight_kg_corrected[is_CMSY], na.rm = TRUE)       # landings from CMSY genera
  ) %>%
  mutate(percent_CMSY = 100 * CMSY_landings / total_landings)    # percentage
  
Landings_CMSY_summary
# If I can sort out priors for these four extra txa, CMSY could account for ~50% of all GoC landings in the last five years.
```

# RAN CMSY ON PERSONAL MACHINE. Here are the outputs from that analysis:

```{r Load in CMSY asessment results}
# Actual B/Bmsy output from CMSY:
CMSY_Timeseries_original<- read.csv("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_StockStatus_Scores/CMSY_Scores/Output_Files/CMSY_Assessment_Timeseries_Dec102025.csv")
# 775 obs., 13 vars.
```