# CMSY Stock Assesments
# NOTE: Must run CMSY on personal machine. Can generate the main files here though. 

################################################################################ 

For questions contact: Erica Ferrer, Ph.D. at emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################ 

# PRE-WORK ------------------------------

```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```

## Review Instructional Files:
```{r Review the CMSY protocol developed by Ferrer et al. (2021)}
# Absolute path to your PDF
pdf_path <- normalizePath(
  "~/ohi_goc_prep/FP/CMSY_Analysis/CMSY_InstructionalFiles/Copy of Ferrer et al. (2021) CMSY Protocol.pdf",
  winslash = "/", mustWork = TRUE
)

# Convert to file URL
file_url <- paste0("file:///", URLencode(pdf_path, reserved = TRUE))

# Open in default browser
browseURL(file_url)
```

# LOAD FILE 1 - Landings Timeseries  ------------------------------
- ABOUT FILE 1: File with annual landings in kg (and n_count observations) grouped by year per "stock". 
- Here, stock is defined as the total population of GENUS X fished within the Gulf of California from Baja to northern Jalisco, not including landings from the Pacific.
- I’m going to run CMSY at the genus level. I actually think it’s a safer more reliable bet than species-level info and because – in many cases – fishermen seem to target taxa of the same genus interchangeably.
- This file was created in /home/ferrer/ohi_goc_prep/FP/CONAPESCA_CleanUp/CONAPESCA_CleaningWorkflow_Mar2026.Rmd

```{r File 1 - Stock Timeseries (JC data)}
CMSY_Landings_file1<- read_csv("/home/ferrer/ohi_goc_prep/FP/CMSY_Analysis/CMSY_LandingsPerGenera_Apr032026.csv")
# 2346 obs., 6 vars.

# Name estimated_genus_name to be "Stock
colnames(CMSY_Landings_file1)
CMSY_Landings_file1 <- CMSY_Landings_file1 %>%
  rename(Stock = estimated_genus_name)
```

```{r Save this file as UNDER CONSTRUCTION so I can work with it offline and use it to generate priors}
write.csv(CMSY_Landings_file1, row.names= FALSE,
          "/home/ferrer/ohi_goc_prep/FP/CMSY_Analysis/CMSY_Revised_April2026/Generating_Model_Priors/Priors_Data/File1_Landings_UNDERCONSTRUCTION_Apr032026.csv")
```

# Create FILE 2 - Metadata & Model Priors  ------------------------------
FILE 2: Create a file with OHI_Region, Stock, Group = Name, English Name, Scientific Name, Source, and a whole bunch of year columns that you populate using the various R files shared by Alfredo. This file can probably be trimmed down but for now I recommend formatting it exactly as it appears from before.

```{r Create File 2 based on File 1}
# STEP 1: CREATE CMSY_Metadata_file2
# Which taxa (stocks) do we want to account for in our File 2?
Stock_names<- CMSY_Landings_file1 %>%
  select(Stock, Major_Functional_Group, estimated_taxa_group) %>%
  distinct()
# 99 unique genera / stocks.

## Read in the column names from blueprint of File 2.
file2_blueprint <- read.csv("/home/ferrer/ohi_goc_prep/FP/CMSY_Analysis/CMSY_InstructionalFiles/CMSY_File2_BluePrint_Apr032026.csv")
file2_headers <- names(file2_blueprint)

## Create an empty dataframe with the same file2_headers
CMSY_Metadata_file2 <- data.frame(matrix(ncol = length(file2_headers), nrow = length(Stock_names$Stock)))
colnames(CMSY_Metadata_file2) <- file2_headers

# STEP 2: FILL CMSY_Metadata_file2
## Populate CMSY_Metadata_file2 with info from Stock_names and file2_blueprint
CMSY_Metadata_file2$Stock <- Stock_names$Stock
CMSY_Metadata_file2$Major_Functional_Group <- Stock_names$Major_Functional_Group
CMSY_Metadata_file2$estimated_taxa_group <- Stock_names$estimated_taxa_group

# Fill in some of the columns off this file:
CMSY_Metadata_file2$Region <- file2_blueprint$Region
CMSY_Metadata_file2$Source <- file2_blueprint$Source
CMSY_Metadata_file2$Comment <- file2_blueprint$Comment

CMSY_Metadata_file2$Sufficient_No_Years <- TRUE
CMSY_Metadata_file2$Yes_Run <- c("Yes")
CMSY_Metadata_file2$btype <- c("None")
CMSY_Metadata_file2$force.cmsy <- c("TRUE")
```

```{r Save this file as UNDER CONSTRUCTION so I can work with it offline and use it to generate priors}
write.csv(CMSY_Metadata_file2, row.names= FALSE,
          "/home/ferrer/ohi_goc_prep/FP/CMSY_Analysis/CMSY_Revised_April2026/Generating_Model_Priors/Priors_Data/File2_Metadata_UNDERCONSTRUCTION_Apr032026.csv")
```


## SAVE FILE 2 BLUE PRINT
```{r SAVE FILE 2 UNDER CONSTRUCTION}
write.csv(CMSY_Metadata_file2, "~/ohi_goc_prep/FP/CMSY_Analysis/CMSY_Revised_April2026/Input_Files/File2_UNDER_CONSTRUCTION_Apr032026.csv", row.names = FALSE)
```
NEXT STEPS: Fill in the rest of file 2, namely the model priors, using the ancillary files shared with me by Alfredo (Ferrer et al., 2022)

Did some work offline on my local machine (not available for viewing here... yet). This work will help fill out File 2.

```{r Read in additional files to complete file 2}
# Read in file if necessary:
#CMSY_Metadata_file2 <- read.csv("~/ohi_goc_prep/FP/CMSY_Analysis/CMSY_Revised_April2026/Input_Files/File2_UNDER_CONSTRUCTION_Apr032026.csv")

## Additional files needed to file in the metadata
Priors_Pristine.Biomass <- read.csv("~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Model_Priors/Priors_StartingBiomass_fromFAOdata_DRAFT_Dec092025.csv")
Priors_Int.Year <- read.csv("~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Model_Priors/Priors_IntermediateYears_DRAFT_Dec092025.csv")
Priors_IntEnd.Conditions <- read.csv("~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Model_Priors/Priors_IntEndConditions_DRAFT_Dec092025.csv")
Priors_Resilience <- read.csv("~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Model_Priors/Priors_ResilienceValues_DRAFT_Dec092025.csv")
Priors_Resilience <- Priors_Resilience %>% select(-ScientificName)
# Common_Names <- read.csv("Stock and Common Names_Oct 23 2025.csv")
```

```{r lowercase_function}
to_lower_df <- function(df) {
  df %>% mutate(across(where(is.character), tolower))
}
```

```{r Stitch these dataframes together to create complete CMSY_Metadata_file2}
CMSY_Metadata_file2        <- to_lower_df(CMSY_Metadata_file2)
Priors_Pristine.Biomass  <- to_lower_df(Priors_Pristine.Biomass)
Priors_Int.Year <- to_lower_df(Priors_Int.Year)
Priors_Resilience <- to_lower_df(Priors_Resilience)

## MERGE: PRISTINE BIOMASS PRIORS
CMSY_Metadata_file2_merged <- CMSY_Metadata_file2 %>% 
  left_join(
    Priors_Pristine.Biomass %>% select(estimated_taxa_group, st_b.low, st_b.hi),
    by = "estimated_taxa_group",
    suffix = c(".file2", ".priors")
  ) %>%
  mutate(
    st_b.low = coalesce(st_b.low.file2, st_b.low.priors),
    st_b.hi  = coalesce(st_b.hi.file2,  st_b.hi.priors)
  ) %>% 
  select(-ends_with(".file2"), -ends_with(".priors"))

# NOTE TO SELF: Our priors for pristine biomass, and the FAO timeseries that underly them, are fairly sparse as in missing several major taxonomic groups (like sardine). Return to this in the new year.

## MERGE: INT YEAR PRIORS
CMSY_Metadata_file2_merged <- CMSY_Metadata_file2_merged %>%
  left_join(
    Priors_Int.Year %>% 
      select(estimated_taxa_group, int_yr, StartYear, EndYear),
    by = "estimated_taxa_group",
    suffix = c(".file2", ".priors")
  ) %>%
  mutate(
    int_yr               = coalesce(int_yr.file2,               int_yr.priors),
    StartYear            = coalesce(StartYear.file2,            StartYear.priors),
    EndYear              = coalesce(EndYear.file2,              EndYear.priors)) %>%
  select(-ends_with(".file2"), -ends_with(".priors"))

## MERGE: INT END CONDITIONS PRIORS
# NOTE TO SELF: Based on the naming conventions within this file I actually think these are the priors for intermediate and ending biomass. I'll circle back to this and iron out the true meaning and utility of these particular priors in the new year.
# Join and fill missing values
CMSY_Metadata_file2_merged <- CMSY_Metadata_file2_merged %>%
  left_join(
    Priors_IntEnd.Conditions,
    by = "estimated_taxa_group",
    suffix = c("", ".priors")
  ) %>%
  mutate(
    int_b.low  = coalesce(int_b.low, int_b.low.priors),
    int_b.hi   = coalesce(int_b.hi, int_b.hi.priors),
    end_b.low  = coalesce(end_b.low, end_b.low.priors),
    end_b.hi = coalesce(end_b.hi, end_b.hi.priors)
  ) %>%
  select(-ends_with(".priors"))


## MERGE: RESILIENCE PRIORS
CMSY_Metadata_file2_merged <- CMSY_Metadata_file2_merged %>%
  left_join(
    Priors_Resilience %>% select(estimated_taxa_group, Resilience, Resilience_Source),
    by = "estimated_taxa_group",
    suffix = c(".file2", ".priors")
  ) %>%
  mutate(
    Resilience = coalesce(Resilience.file2, Resilience.priors)
  ) %>%
  select(-Resilience.file2, -Resilience.priors)
```

```{r Finally lets clean up File 2 before saving and running CMSY}
CMSY_Metadata_file2_merged<- CMSY_Metadata_file2_merged %>% 
  select(Region, Stock, Landings_Datasource=Source, 
         StartYear, EndYear, int_yr,
         btype,
         Flim, Fpa, Blim, Bpa, Bmsy, B40,
         FMSY, MSY, MSYBtrigger, M,
         Fofl, last_F,
         q.start, q.end,
         st_b.low, st_b.hi, 
         int_b.low, int_b.hi, end_b.low, end_b.hi,
         r.low, r.hi, 
         Resilience, Resilience_Source,
         Yes_Run, Comment)

# Fill in the Yes_Run column based on model priors:
CMSY_Metadata_file2_merged <- CMSY_Metadata_file2_merged %>%
  mutate(
    Yes_Run = if_else(
      is.na(Resilience) | is.na(int_b.low) | is.na(st_b.low) | is.na(end_b.low),
      0,
      1
    ),
    Comment = if_else(
      Yes_Run == 0,
      "Incomplete model priors",
      NA_character_
    )
  )

sum(CMSY_Metadata_file2_merged$Yes_Run==1)
# 38 stocks to run!

# NOTES ON NAMING, Dec 10 2025:
# Need a btype column = None
# Need a force.cmsy column = TRUE
# StartYear and EndYear must actually align with file one time series.
```

```{r How much of FP will these CMSY stocks account for?}
Landings_CMSY_TEST <- Landings_GoC_NoFresh.NoMariculture %>%
  mutate(estimated_genus_name = tolower(estimated_genus_name))

CMSY_Stocks <- CMSY_Metadata_file2_merged %>%
  filter(Yes_Run == 1) %>%
  pull(Stock) %>%
  unique()

CMSY_Stocks_hypothetical <- c(
  CMSY_Stocks, 
  "sardinops", 
  "opisthonema", 
  "sarda",
  "scomber"
)

# Calculate total from genera landings identified in the CMSY exercise:
# NOTE TO SELF: Come back to this. Find priors for these four taxa! They represent a huge component of total landings.
Landings_CMSY_summary <- Landings_CMSY_TEST %>%
  filter(year %in% 2015:2024) %>%           # only last 10 years
  mutate(is_CMSY = estimated_genus_name %in% CMSY_Stocks_hypothetical) %>%  # flag CMSY genera
  summarise(
    total_landings = sum(weight_kg_corrected, na.rm = TRUE),                # total landings
    CMSY_landings  = sum(weight_kg_corrected[is_CMSY], na.rm = TRUE)       # landings from CMSY genera
  ) %>%
  mutate(percent_CMSY = 100 * CMSY_landings / total_landings)    # percentage
  
Landings_CMSY_summary
# If I can sort out priors for these four extra txa, CMSY could account for ~50% of all GoC landings in the last five years.
```

## SAVE FILE 2 COMPLETE
```{r SAVE FILE 2 COMPLETE}
write.csv(CMSY_Metadata_file2_merged, "~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Input_Files/File2_Dec092025.csv", row.names = FALSE)
```

```{r How much of FP will these CMSY stocks account for?}
Landings_CMSY_TEST <- Landings_GoC_NoFresh.NoMariculture %>%
  mutate(estimated_genus_name = tolower(estimated_genus_name))

CMSY_Stocks <- CMSY_Metadata_file2_merged %>%
  filter(Yes_Run == 1) %>%
  pull(Stock) %>%
  unique()

CMSY_Stocks_hypothetical <- c(
  CMSY_Stocks, 
  "sardinops", 
  "opisthonema", 
  "sarda",
  "scomber"
)

# Calculate total from genera landings identified in the CMSY exercise:
# NOTE TO SELF: Come back to this. Find priors for these four taxa! They represent a huge component of total landings.
Landings_CMSY_summary <- Landings_CMSY_TEST %>%
  filter(year %in% 2015:2024) %>%           # only last 10 years
  mutate(is_CMSY = estimated_genus_name %in% CMSY_Stocks_hypothetical) %>%  # flag CMSY genera
  summarise(
    total_landings = sum(weight_kg_corrected, na.rm = TRUE),                # total landings
    CMSY_landings  = sum(weight_kg_corrected[is_CMSY], na.rm = TRUE)       # landings from CMSY genera
  ) %>%
  mutate(percent_CMSY = 100 * CMSY_landings / total_landings)    # percentage
  
Landings_CMSY_summary
# If I can sort out priors for these four extra txa, CMSY could account for ~50% of all GoC landings in the last five years.
```

# RAN CMSY ON PERSONAL MACHINE. Here are the outputs from that analysis:

```{r Load in CMSY asessment results}
# Actual B/Bmsy output from CMSY:
CMSY_Timeseries_original<- read.csv("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_StockStatus_Scores/CMSY_Scores/Output_Files/CMSY_Assessment_Timeseries_Dec102025.csv")
# 775 obs., 13 vars.
```