---
title: 'OHI GoC 2025: TR Current Status Visualizations'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor:
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options:
  chunk_output_type: console
---

# TR visualizations

I run this script after `tr_current_status.qmd`. It reads the saved current status outputs and builds charts with consistent region colors and labels across quantity, quality, sustainability, and the overall TR score.

```{r}
library(tidyverse)
library(ggplot2)
library(plotly)
library(sf)
library(mapview)
library(here)
```

```{r}
rgn_labels <- c(
  "Region_1" = "Western Upper Gulf",
  "Region_2" = "Western Midriff Islands",
  "Region_3" = "Western Central Gulf",
  "Region_4" = "Western Southern Gulf",
  "Region_5" = "Eastern Upper Gulf",
  "Region_6" = "Eastern Midriff Islands",
  "Region_7" = "Southern Sonora Coast",
  "Region_8" = "Sinaloa Coast",
  "Region_9" = "Southeastern Gulf"
)

rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

region_levels <- paste0("Region_", 1:9)
rgn_label_colors <- setNames(unname(rgn_colors[region_levels]), unname(rgn_labels[region_levels]))
component_levels <- c("Quantity", "Quality", "Sustainability", "Overall TR")

add_rgn_labels <- function(df) {
  df %>%
    mutate(
      rgn_id = factor(as.character(rgn_id), levels = region_levels),
      rgn_label = factor(
        rgn_labels[as.character(rgn_id)],
        levels = unname(rgn_labels[region_levels])
      )
    )
}

tr_plotly_lines <- function(df, y_var, title, y_range = NULL) {
  yr <- range(df$year, na.rm = TRUE)

  plot_ly(
    df,
    x = ~year,
    y = as.formula(paste0("~", y_var)),
    color = ~rgn_label,
    colors = rgn_label_colors,
    type = "scatter",
    mode = "lines+markers"
  ) %>%
    layout(
      title = title,
      xaxis = list(title = "Año // Year"),
      yaxis = list(title = "Puntaje // Percent (out of 100)", range = y_range),
      legend = list(title = list(text = "Region"))
    ) %>%
    add_lines(
      x = yr,
      y = c(100, 100),
      line = list(color = "black", width = 2, dash = "dash"),
      name = "Meta // Goal 2030 (100/100)",
      inherit = FALSE
    )
}
```

## Read outputs

```{r}
cs_quantity_clean <- read_csv(
  here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quantity", "tr_quantity_current_status_121025.csv"),
  show_col_types = FALSE
) %>%
  rename(final_score = score) %>%
  dplyr::select(rgn_id, year, final_score)

cs_quality_clean <- read_csv(
  here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quality", "cs_quality_all_indicators_032326.csv"),
  show_col_types = FALSE
)

cs_sust_clean <- read_csv(
  here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability", "cs_sust_all_indicators_070126.csv"),
  show_col_types = FALSE
)

all_cs_indicators <- read_csv(
  here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output", "tr_current_status_final_070126.csv"),
  show_col_types = FALSE
)

tr_components_long <- bind_rows(
  add_rgn_labels(cs_quantity_clean) %>%
    transmute(rgn_id, rgn_label, year, score = final_score, component = "Quantity"),
  add_rgn_labels(cs_quality_clean) %>%
    transmute(rgn_id, rgn_label, year, score = final_score, component = "Quality"),
  add_rgn_labels(cs_sust_clean) %>%
    transmute(rgn_id, rgn_label, year, score = final_score, component = "Sustainability"),
  add_rgn_labels(all_cs_indicators) %>%
    transmute(rgn_id, rgn_label, year, score = tr_current_status, component = "Overall TR")
) %>%
  filter(year %in% 2019:2024) %>%
  mutate(component = factor(component, levels = component_levels))

tr_2024_long <- tr_components_long %>%
  filter(year == 2024)

all_cs_indicators_2024 <- all_cs_indicators %>%
  filter(year == 2024)
```

## Interactive line charts (plotly)

```{r}
tr_score_plotly <- tr_plotly_lines(
  add_rgn_labels(all_cs_indicators),
  y_var = "tr_current_status",
  title = "CURRENT STATUS: Tourism & Recreation"
)

quantity_plotly <- tr_plotly_lines(
  add_rgn_labels(cs_quantity_clean),
  y_var = "final_score",
  title = "TR Quantity"
)

quality_plotly <- tr_plotly_lines(
  add_rgn_labels(cs_quality_clean),
  y_var = "final_score",
  title = "TR Quality"
)

sustainability_plotly <- tr_plotly_lines(
  add_rgn_labels(cs_sust_clean),
  y_var = "final_score",
  title = "TR Sustainability"
)

tr_score_plotly
quantity_plotly
quality_plotly
sustainability_plotly
```

## All components over time (ggplot)

```{r}
gg_tr_components_lines <- ggplot(
  tr_components_long,
  aes(x = year, y = score, color = rgn_label, group = rgn_label)
) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.8) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey45", linewidth = 0.4) +
  facet_wrap(~ component, ncol = 2) +
  scale_color_manual(values = rgn_label_colors, name = "Region") +
  scale_x_continuous(breaks = 2019:2024) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 25)) +
  labs(
    title = "TR current status by component (2019 to 2024)",
    x = "Year",
    y = "Score (out of 100)"
  ) +
  theme_minimal(base_family = "Calibri") +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 11),
    legend.position = "right"
  )

gg_tr_components_lines
```

## 2024 snapshot by region (ggplot)

```{r}
gg_tr_2024_bars <- ggplot(
  tr_2024_long,
  aes(x = rgn_label, y = score, fill = rgn_label)
) +
  geom_col(width = 0.82, color = "white", linewidth = 0.3) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey45", linewidth = 0.4) +
  facet_wrap(~ component, ncol = 2) +
  scale_fill_manual(values = rgn_label_colors, guide = "none") +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 25)) +
  labs(
    title = "TR current status by region (2024)",
    x = NULL,
    y = "Score (out of 100)"
  ) +
  theme_minimal(base_family = "Calibri") +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 11),
    axis.text.x = element_text(angle = 35, hjust = 1, size = 8)
  )

gg_tr_2024_bars
```

## Overall TR polar chart (2024)

```{r}
gg_tr_2024_polar <- ggplot(
  tr_2024_long %>% filter(component == "Overall TR"),
  aes(x = rgn_label, y = score, fill = rgn_label)
) +
  geom_col(width = 1, color = "white", linewidth = 0.35) +
  geom_hline(yintercept = 100, color = "grey55", linewidth = 0.5) +
  coord_polar(start = -pi / 2, clip = "off") +
  scale_fill_manual(values = rgn_label_colors, guide = "none") +
  scale_y_continuous(limits = c(0, 100), breaks = c(25, 50, 75, 100)) +
  labs(
    title = "Overall TR current status by region (2024)",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_family = "Calibri") +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    panel.grid.major.x = element_blank()
  )

gg_tr_2024_polar
```

## Map visualization (2024 overall TR)

```{r}
ohi <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp"), quiet = TRUE) %>%
  filter(location == "marine") %>%
  dplyr::select(rgn_id)

all_cs_indicators_2024_ohigns <- full_join(all_cs_indicators_2024, ohi, by = "rgn_id") %>%
  st_as_sf()
```

```{r}
mapview(
  all_cs_indicators_2024_ohigns,
  zcol = "tr_current_status",
  col.regions = colorRampPalette(c("#FEE090", "#ABD9E9", "#4575B4", "#313695")),
  layer.name = "Tourism & Recreation Current Status Scores (2024)",
  legend = TRUE
)

tr_at <- seq(0, 100, by = 10)

tr_cols <- colorRampPalette(c(
  "#8a0032", "#A50026", "#A50026", "#D73027", "#F46D43",
  "#FDAE61", "#FEE090", "#FEE090", "#ABD9E9", "#4575B4", "#313695"
))(length(tr_at) - 1)

mapview(
  all_cs_indicators_2024_ohigns,
  zcol = "tr_current_status",
  col.regions = tr_cols,
  at = tr_at,
  layer.name = "Tourism & Recreation Current Status Scores (2024)",
  legend = TRUE
)
```

## Choropleth map (ggplot)

I drew the Gulf states background first, then the marine regions on top.

```{r}
tr_palette <- c(
  "#8a0032", "#A50026", "#A50026", "#D73027", "#F46D43",
  "#FDAE61", "#FEE090", "#FEE090", "#ABD9E9", "#4575B4", "#313695"
)

states <- st_read(here::here("spatial/ohi_regions/states.shp"), quiet = TRUE) %>%
  st_transform(st_crs(all_cs_indicators_2024_ohigns))

gg_tr <- ggplot() +
  geom_sf(
    data = states,
    fill = "#d6cac0",
    color = "grey10",
    linewidth = 0.1
  ) +
  geom_sf(
    data = all_cs_indicators_2024_ohigns,
    aes(fill = tr_current_status),
    color = "black",
    linewidth = 0.2
  ) +
  scale_fill_gradientn(
    colours = tr_palette,
    limits = c(0, 100),
    na.value = "grey85",
    name = "Tourism & Recreation\nCurrent Status 2024"
  ) +
  coord_sf(expand = FALSE) +
  theme_void() +
  theme(
    text = element_text(family = "Calibri"),
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    legend.position = "right",
    legend.key.height = unit(0.8, "cm"),
    legend.key.width = unit(0.5, "cm"),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    plot.margin = margin(10, 10, 10, 10)
  )

gg_tr
```

