---
title: 'OHI GoC 2024: Catch Per Unit Effort GoC'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: inline
editor_options: 
  chunk_output_type: inline
---

# Summary

## Global Datasets of AIS-based Fishing Effort and Vessel Presence

This dataset contains the Global Fishing Watch AIS-based fishing effort and vessel presence datasets.
Data is based on fishing detections of \>114,000 unique AIS devices on fishing vessels, of which \~70,000 are active each year.
Fishing vessels are identified via a neural network classifier, vessel registry databases, and manual review by GFW and regional experts.
Data are binned into grid cells 0.01 (or 0.1) degrees on a side and measured in units of hours.
The time is calculated by assigning an amount of time to each AIS detection (which is the time to the previous position), and then summing all positions in each grid cell.

**Current version**

2.0: Fishing effort data for 2012-2020 using our latest AIS algorithms, neural network models, and vessel registry database.

**File structure**

Fishing effort and vessel presence data are available as .csv files in daily formats.
Files for each year are stored in separate .zip files.
A README.txt file is provided for each dataset version and contains the table schema and additional information.
There is also a README-known-issues-v2.txt file outlining some of the known issues with the version 2 release.

Files are names according to the following convention:

Daily file format: \[fleet/mmsi\]-daily-csvs-\[100/10\]-\[version\]-\[year\].zip

Fishing vessel format: fishing-vessels-\[version\].csv

README file format: README-\[fleet/mmsi/fishing-vessels\]-\[version\].txt File identifiers:

-   \[fleet/mmsi\]: Data by fleet (flag and geartype) or by MMSI \[100/10\]: 100th or 10th degree resolution

-   \[version\]: Fishing effort version (v2, v1-5,v1)

-   \[year\]: Year of data included in .zip or .csv file

Examples: fleet-daily-csvs-100-v2-2020.zip; mmsi-daily-csvs-10-v2-2020.csv; fishing-vessels-v2.csv; README-fleet-v2.txt

**Reference & Code**

For additional information about these results, see the associated journal article: D.A.
Kroodsma, J. Mayorga, T. Hochberg, N.A. Miller, K. Boerder, F. Ferretti, A. Wilson, B. Bergman, T.D. White, B.A. Block, P. Woods, B. Sullivan, C. Costello, and B.
Worm.
"Tracking the global footprint of fisheries." Science 361.6378 (2018).

GitHub repository for Tracking the global footprint of fisheries: https://github.com/GlobalFishingWatch/paper-global-footprint-of-fisheries License: Unless otherwise stated, Global Fishing Watch data is licensed under a Creative Commons Attribution-ShareAlike 4.0 International license and code under an Apache 2.0 license.

-----------------------------------------------------------

# Setup 

```{r}
library(raster)
library(terra)
library(readr)
library(magrittr)
library(mregions2)
library(mapview)
library(sf)
library(tidyverse)
library(lwgeom)
library(here)
library(leaflet.extras2)
library(fasterize)
library(tictoc)
library(kableExtra)
library(plotly)
library(viridis)
library(readr)
library(purrr)
library(doParallel)


# for Global Fishing Watch API (using package gfwr)
# install.packages("gfwr", repos = c("https://globalfishingwatch.r-universe.dev", "https://cran.r-project.org"))
library(gfwr)

# obtaining API key -- follow the instructions using .Renviron here <https://github.com/GlobalFishingWatch/gfwr>
# usethis::edit_r_environ()
key <- gfw_auth()

# ---- sources! ----
# source(here("workflow", "R", "common.R")) # file creates objects to process data
## set the mazu and neptune data_edit share based on operating system
dir_M             <- c('Windows' = '//mazu.nceas.ucsb.edu/ohi',
                       'Darwin'  = '/Volumes/ohi',    ### connect (cmd-K) to smb://mazu/ohi
                       'Linux'   = '/home/shares/ohi')[[ Sys.info()[['sysname']] ]]

# ---- set year and file path info ----
current_year <- 2025 # Update this in the future!!
version_year <- paste0("v",current_year)
data_dir_version_year <- paste0("d", current_year)

# ---- data directories ----

# raw data directory (on Mazu)
raw_data_dir <- here::here(dir_M, "OHI_GOC", "_raw_data")

# global_fishing_watch intermediate dir
int_dir <- here::here(dir_M, "OHI_GOC", "_raw_data", "global_fishing_watch", data_dir_version_year, "int")

# final output dir
output_dir <- here("FIS", version_year, "output")

# spatial data for GoC
goc_spatial <- here("spatial")

# this CRS might be better for visualization, explore.
gulf_crs <- st_crs("+proj=aea +lat_1=23 +lat_2=30 +lat_0=25 +lon_0=-110 +datum=WGS84 +units=m +no_defs")
```

## GoC Ecoregion shapefile

```{r}
eco_hr_eqarea <- st_read(here(goc_spatial, "GoC_polygon_ecoregion_eqarea.shp"))
mapview(eco_hr_eqarea)

# check CRS of ecoregion boundary
st_crs(eco_hr_eqarea) # NAD83 / Conus Albers, needs to be transformed

# transform to gulf_CRS
eco_hr_albers <- st_transform(eco_hr_eqarea, crs = "+proj=aea +lat_1=23 +lat_2=30 +lat_0=25 +lon_0=-110 +datum=WGS84 +units=m +no_defs")
st_crs(eco_hr_albers) # double check it worked well
mapview(eco_hr_albers) # looks good!

# to make computation faster
eco_hr_nad83 <- st_transform(eco_hr_albers, 3310)
st_crs(eco_hr_nad83)
```




# Download GFW data

```{r}
# download yearly fishing effort data for 2012 - 2021
years <- 2016:2021

# use a loop to make it all easier! parallelize too to make it much faster

# set up parallel processing for faster download
cl <- 20
registerDoParallel(cl) # creates a set of copies of R running in parallel and communicating over sockets

download_all_years_gfw <- function(year){
  
  raw_gfw <- gfwr::get_raster( # the name does not specify correctly.  There is no function for downloading a raster, it is a data frame.
    spatial_resolution = 'HIGH',
    temporal_resolution = 'YEARLY',
    group_by = 'FLAGANDGEARTYPE',
    start_date = paste0(year, '-01-01'),
    end_date = paste0(year, '-12-31'),
    region = eco_hr_albers,
    region_source = 'USER_SHAPEFILE',
    key = key
  )
  
  # convert the tabular data to sf object
  gfw_sf <- st_as_sf(raw_gfw, coords = c("Lon", "Lat"), crs = 4326) # the raw data (Lon, Lat) from gfwr::get_raster() is probably in WGS84 coordinates which is why it has to be transformed to the gulf_crs separately!  The `get_raster` function is not working as it should.  It does not limit data to the extent of eco_hr_albers, and instead provides the global data.
  
  # ensure the crs match
  gfw_sf_transformed <- st_transform(gfw_sf, gulf_crs)
  eco_hr_albers <- st_transform(eco_hr_albers, gulf_crs)
  
  # crop the effort data to the GoC ecoregion
  gfw_cropped <- st_intersection(gfw_sf_transformed, eco_hr_albers)
  
  return(write_csv(gfw_cropped, here::here(int_dir, paste0("gfw_data_", year, ".csv"))))
}

# run the function to download
lapply(years, download_all_years_gfw)

```



```{r}
# download yearly data for 2012 - 2021
raw_gfw_2012 <- gfwr::get_raster( # the name does not specify correctly.  There is no function for downloading a raster, it is a data frame.
  spatial_resolution = 'HIGH',
  temporal_resolution = 'YEARLY',
  group_by = 'FLAGANDGEARTYPE',
  start_date = '2012-01-01',
  end_date = '2012-12-31',
  region = eco_hr_albers,
  region_source = 'USER_SHAPEFILE',
  key = key
)

# convert the tabular data to sf object
gfw_sf_2012 <- st_as_sf(raw_gfw_2012, coords = c("Lon", "Lat"), crs = 4326) # the raw data (Lon, Lat) from gfwr::get_raster() is probably in WGS84 coordinates which is why it has to be transformed to the gulf_crs separately!  The `get_raster` function is not working as it should.  It does not limit data to the extent of eco_hr_albers, and instead provides the global data.

# ensure the crs match
gfw_sf_transformed_2012 <- st_transform(gfw_sf_2012, gulf_crs)
eco_hr_albers <- st_transform(eco_hr_albers, gulf_crs)

# crop the effort data to the GoC ecoregion
gfw_cropped_2012 <- st_intersection(gfw_sf_transformed_2012, eco_hr_albers)

# plot to ensure they overlay
ggplot() +
  geom_sf(data = eco_hr_albers, fill = "lightblue", color = "black") +
  geom_sf(data = gfw_cropped_2012) +
  theme_minimal() +
  labs(title = "Fishing Effort Overlaid on GoC Ecoregion for 2012")

```
```{r}
# for 2016
raw_gfw_2015 <- gfwr::get_raster( # the name does not specify correctly.  There is no function for downloading a raster, it is a data frame.
  spatial_resolution = 'HIGH',
  temporal_resolution = 'YEARLY',
  group_by = 'FLAGANDGEARTYPE',
  start_date = '2015-01-01',
  end_date = '2015-12-31',
  region = eco_hr_albers,
  region_source = 'USER_SHAPEFILE',
  key = key
)

# convert the tabular data to sf object
gfw_sf_2015 <- st_as_sf(raw_gfw_2015, coords = c("Lon", "Lat"), crs = 4326) # the raw data (Lon, Lat) from gfwr::get_raster() is probably in WGS84 coordinates which is why it has to be transformed to the gulf_crs separately!  The `get_raster` function is not working as it should.  It does not limit data to the extent of eco_hr_albers, and instead provides the global data.

# ensure the crs match
gfw_sf_transformed_2015 <- st_transform(gfw_sf_2015, gulf_crs)
eco_hr_albers <- st_transform(eco_hr_albers, gulf_crs)

# crop the effort data to the GoC ecoregion
gfw_cropped_2015 <- st_intersection(gfw_sf_transformed_2015, eco_hr_albers)

# plot to ensure they overlay
ggplot() +
  geom_sf(data = eco_hr_albers, fill = "lightblue", color = "black") +
  geom_sf(data = gfw_cropped_2015) +
  theme_minimal() +
  labs(title = "Fishing Effort Overlaid on GoC Ecoregion for 2015")
```




