---
title: 'OHI GoC 2024: Catch Per Unit Effort GoC'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: inline
editor_options: 
  chunk_output_type: console
---

# Summary

## Global Datasets of AIS-based Fishing Effort and Vessel Presence

This dataset contains the Global Fishing Watch AIS-based fishing effort and vessel presence datasets.
Data is based on fishing detections of \>114,000 unique AIS devices on fishing vessels, of which \~70,000 are active each year.
Fishing vessels are identified via a neural network classifier, vessel registry databases, and manual review by GFW and regional experts.
Data are binned into grid cells 0.01 (or 0.1) degrees on a side and measured in units of hours.
The time is calculated by assigning an amount of time to each AIS detection (which is the time to the previous position), and then summing all positions in each grid cell.

Data download: <https://globalfishingwatch.org/data-download/datasets/public-fishing-effort>
May 18 2021, MRF
Description: Fleet daily v2
Resolution: daily, 100th degree
Date Downloaded: March 4, 2025

Geartypes, defined by <https://github.com/OHI-Science/global_food_pressures/blob/main/fisheries/marine/disturbance/STEP2a_GFWdata_prep.Rmd>:
- fishing: a combination of vessels of unknown fishing gear
 - drifting_longlines: drifting longlines
 - seiners: vessels using seine nets, including potential purse seine vessels
   targeting tuna and other species, as well as danish and other seines
     - purse_seines: purse seines, both pelagic and demersal
        - tuna_purse_seines: large purse seines primarily fishing for tuna.
        - other_purse_seines: purse seiners fishing for mackerel, anchovies, etc, often smaller and operating nearer the coast than tuna purse seines.
    - other_seines: danish seines and other seiners not using purse seines.
 - trawlers: trawlers, all types
 - pole_and_line: vessel from which people fish with pole and line.
 - trollers: vessel that tows multiple fishing lines.
 - fixed_gear: a category that includes potential set longlines, set gillnets,  and pots and traps
     - pots_and_traps: vessel that deploys pots (small, portable traps) or traps to
       catch fish
     - set_longlines: vessel that fishes by setting longlines anchored to the
       seafloor. These lines have shorter hooked, typically baited, lines hanging
       from them
     - set_gillnets: vessel that fishes by setting gillnets anchored to the seafloor.
 - dredge_fishing: vessel that tows a dredge the scrapes up edible bottom
   dwellers such as scallops or oysters.
 - squid_jigger: squid jiggers, mostly large industrial pelagic operating vessels


**Current version**

2.0: Fishing effort data for 2012-2020 using our latest AIS algorithms, neural network models, and vessel registry database.

**File structure**

Fishing effort and vessel presence data are available as .csv files in daily formats.
Files for each year are stored in separate .zip files.
A README.txt file is provided for each dataset version and contains the table schema and additional information.
There is also a README-known-issues-v2.txt file outlining some of the known issues with the version 2 release.

Files are names according to the following convention:

Daily file format: \[fleet/mmsi\]-daily-csvs-\[100/10\]-\[version\]-\[year\].zip

Fishing vessel format: fishing-vessels-\[version\].csv

README file format: README-\[fleet/mmsi/fishing-vessels\]-\[version\].txt File identifiers:

-   \[fleet/mmsi\]: Data by fleet (flag and geartype) or by MMSI \[100/10\]: 100th or 10th degree resolution

-   \[version\]: Fishing effort version (v2, v1-5,v1)

-   \[year\]: Year of data included in .zip or .csv file

Examples: fleet-daily-csvs-100-v2-2020.zip; mmsi-daily-csvs-10-v2-2020.csv; fishing-vessels-v2.csv; README-fleet-v2.txt

**Reference & Code**

For additional information about these results, see the associated journal article: D.A.
Kroodsma, J. Mayorga, T. Hochberg, N.A. Miller, K. Boerder, F. Ferretti, A. Wilson, B. Bergman, T.D. White, B.A. Block, P. Woods, B. Sullivan, C. Costello, and B.
Worm.
"Tracking the global footprint of fisheries." Science 361.6378 (2018).

GitHub repository for Tracking the global footprint of fisheries: https://github.com/GlobalFishingWatch/paper-global-footprint-of-fisheries License: Unless otherwise stated, Global Fishing Watch data is licensed under a Creative Commons Attribution-ShareAlike 4.0 International license and code under an Apache 2.0 license.

-----------------------------------------------------------

# Setup 

```{r}
library(raster)
library(terra)
library(readr)
library(magrittr)
library(mregions2)
library(mapview)
library(sf)
library(tidyverse)
library(lwgeom)
library(here)
library(leaflet.extras2)
library(fasterize)
library(tictoc)
library(kableExtra)
library(plotly)
library(viridis)
library(readr)
library(purrr)
library(doParallel)
library(httr2) # to handle timeouts (error HTTP 524 NA)
library(data.table)
library(leaflet)
library(leaflet.extras)
library(RColorBrewer)


# for Global Fishing Watch API (using package gfwr)
# install.packages("gfwr", repos = c("https://globalfishingwatch.r-universe.dev", "https://cran.r-project.org"))
library(gfwr)

# obtaining API key -- follow the instructions using .Renviron here <https://github.com/GlobalFishingWatch/gfwr>
# usethis::edit_r_environ()
key <- gfw_auth()

# ---- sources! ----
# source(here("workflow", "R", "common.R")) # file creates objects to process data
## set the mazu and neptune data_edit share based on operating system
dir_M             <- c('Windows' = '//mazu.nceas.ucsb.edu/ohi',
                       'Darwin'  = '/Volumes/ohi',    ### connect (cmd-K) to smb://mazu/ohi
                       'Linux'   = '/home/shares/ohi')[[ Sys.info()[['sysname']] ]]

# ---- set year and file path info ----
current_year <- 2025 # Update this in the future!!
version_year <- paste0("v",current_year)
data_dir_version_year <- paste0("d", current_year)

# ---- data directories ----

# raw data directory (on Mazu)
raw_data_dir <- here::here(dir_M, "OHI_GOC", "_raw_data")

# global_fishing_watch intermediate dir
int_dir <- here::here(dir_M, "OHI_GOC", "_raw_data", "global_fishing_watch", data_dir_version_year, "int")

# FIS intermediate dir
fis_int_dir <-  here::here(dir_M, "OHI_GOC", "goal_prep", "fis", "v2025", "int")

# final output dir
output_dir <- here("FIS", version_year, "output")

# spatial data for GoC
goc_spatial <- here("spatial")

# this CRS might be better for visualization, explore.
gulf_crs <- st_crs("+proj=aea +lat_1=23 +lat_2=30 +lat_0=25 +lon_0=-110 +datum=WGS84 +units=m +no_defs")
```

## GoC Ecoregion shapefile

```{r}
eco_hr_eqarea <- st_read(here(goc_spatial, "GoC_polygon_ecoregion_eqarea.shp"))
# mapview(eco_hr_eqarea)

# check CRS of ecoregion boundary
# st_crs(eco_hr_eqarea) # NAD83 / Conus Albers, needs to be transformed

# transform to gulf_CRS
eco_hr_albers <- st_transform(eco_hr_eqarea, crs = "+proj=aea +lat_1=23 +lat_2=30 +lat_0=25 +lon_0=-110 +datum=WGS84 +units=m +no_defs")
# st_crs(eco_hr_albers) # double check it worked well
# mapview(eco_hr_albers) # looks good!
```




# Automated Download GFW data

```{r}
# download yearly fishing effort data for 2012 - 2021
years <- 2017:2021

# set a global timeout for all httr2 requests
# httr2::set_timeout(seconds = 600)  # ~ 10 minutes timeout

# use a loop to make it all easier! parallelize to to make it much faster

# set up parallel processing for faster download
cl <- 10
registerDoParallel(cl) # creates a set of copies of R running in parallel and communicating over sockets

download_all_years_gfw <- function(year){ #year=2017
  
  raw_gfw <- gfwr::get_raster( # the name does not specify correctly.  There is no function for downloading a raster, it is a data frame.
    spatial_resolution = 'HIGH',
    temporal_resolution = 'YEARLY',
    group_by = 'FLAGANDGEARTYPE',
    start_date = paste0(year, '-01-01'),
    end_date = paste0(year, '-12-31'),
    region = eco_hr_albers,
    region_source = 'USER_SHAPEFILE',
    key = key
  )
    # httr2::req_timeout(seconds = 600)  # ~ 10 minutes before timeout
  
  # convert the tabular data to sf object
  gfw_sf <- st_as_sf(raw_gfw, coords = c("Lon", "Lat"), crs = 4326) # the raw data (Lon, Lat) from gfwr::get_raster() is probably in WGS84 coordinates which is why it has to be transformed to the gulf_crs separately!  The `get_raster` function is not working as it should.  It does not limit data to the extent of eco_hr_albers, and instead provides the global data.
  
  # ensure the crs match
  gfw_sf_transformed <- st_transform(gfw_sf, gulf_crs)
  eco_hr_albers <- st_transform(eco_hr_albers, gulf_crs)
  
  # crop the effort data to the GoC ecoregion
  gfw_cropped <- st_intersection(gfw_sf_transformed, eco_hr_albers)
  
  return(write_csv(gfw_cropped, here::here(int_dir, paste0("gfw_data_", year, ".csv"))))
}

# run the function to download
lapply(years, download_all_years_gfw)

```



```{r}
# if the function is not working due to server-side time out, download and process each year separately. 
raw_gfw_2019 <- gfwr::get_raster( # the name does not specify correctly.  There is no function for downloading a raster, it is a data frame.
  spatial_resolution = 'HIGH',
  temporal_resolution = 'YEARLY',
  group_by = 'FLAGANDGEARTYPE',
  start_date = '2019-01-01',
  end_date = '2019-12-31',
  region = eco_hr_albers,
  region_source = 'USER_SHAPEFILE',
  key = key
  ) # %>% httr2::req_timeout(seconds = 300) # HTTP 429 Too Many Requests and HTTP 524 NA (timeout)

##### changing the way the data range is called (even though on GitHub documentation it says it can only do a year at a time) because the global `hab_prs_soft_bottom_data_prep.Rmd` does it this way and it worked for them...
raw_gfw_2019 <- gfwr::get_raster( # the name does not specify correctly.  There is no function for downloading a raster, it is a data frame.
  spatial_resolution = 'HIGH',
  temporal_resolution = 'YEARLY',
  group_by = 'FLAGANDGEARTYPE',
  date_range = '2019-01-01, 2019-12-31',
  region = eco_hr_albers,
  region_source = 'USER_SHAPEFILE',
  key = key
  ) ### NOPE--- does not work anymore! Error in gfwr::get_raster(spatial_resolution = "HIGH", temporal_resolution = "YEARLY", unused argument (date_range = "2019-01-01, 2019-12-31")

# convert the tabular data to sf object
gfw_sf_2019 <- st_as_sf(raw_gfw_2019, coords = c("Lon", "Lat"), crs = 4326) # the raw data (Lon, Lat) from gfwr::get_raster() is probably in WGS84 coordinates which is why it has to be transformed to the gulf_crs separately!  The `get_raster` function is not working as it should.  It does not limit data to the extent of eco_hr_albers, and instead provides the global data.

# ensure the crs match
gfw_sf_transformed_2019 <- st_transform(gfw_sf_2019, gulf_crs)
eco_hr_albers <- st_transform(eco_hr_albers, gulf_crs)

# crop the effort data to the GoC ecoregion
gfw_cropped_2019 <- st_intersection(gfw_sf_transformed_2019, eco_hr_albers)

# save to mazu
write_csv(gfw_cropped_2019, here::here(int_dir, "gfw_data_2019.csv"))

# plot to ensure they overlay
ggplot() +
  geom_sf(data = eco_hr_albers, fill = "lightblue", color = "black") +
  geom_sf(data = gfw_cropped_2019) +
  theme_minimal() +
  labs(title = "Fishing Effort Overlaid on GoC Ecoregion for 2019")

```

# Manual Download of GRW effort data

If the API from the package `gfwr` is presenting an HTTP 524 or 429 error, which is a timeout/server-side error, download the data manually from <https://globalfishingwatch.org/data-download/datasets/public-fishing-effort>.

Processing the daily effort data into annual effort data was originally written within fisheries/marine/disturbance/STEP2a_GFWdata_prep.Rmd, and is modified here to fit the needs of the GoC project.

## Filepath for daily files

```{r}
# create filepath for all daily CSVs for 2012:2020
all_years_list_func <- function(year){
  
  gfw_filepath <- "/home/shares/ohi/OHI_GOC/_raw_data/global_fishing_watch/d2025/global_fishing_watch_manual_data_fleet_v2/" # set baseline filepath for all folders of all years in mazu
  
  folder_path <- paste0(gfw_filepath, "fleet-daily-csvs-100-v2-", year) # dynamically replace the year
  
  filepaths <- list.files(folder_path, pattern = ".csv", full = TRUE) # full filepath for the .csvs for that year
  
  return(filepaths) # give the list of file paths, which later needs to be combined into one list for all .csvs to be processed at once
  
}

# define the range of years
years <- 2012:2020

# lapply to generate file paths for all years and combine them into one list
tic()
all_filepaths <- unlist(lapply(years, all_years_list_func)) # length of 3288 is great! 9 years * 365, + 3 days for leap years in 2012, 2016, and 2020.
toc() # v2025: 0.378 sec elapsed
```

NOT NECESSARY: code for summarizing each daily file by cell and gear, if needed:
```{r}
# function for processing files within `all_filepaths`
process_fishing_data <- function(filepath) {

  # get all file paths for each year
  filepaths <- all_years_list_func(year)

  # read and combine all files for a year
  yearly_data <- rbindlist(lapply(filepaths, fread), fill = TRUE)

  # summarize fishing hours by latitude, longitude, and gear type
  summarized_data <- yearly_data %>%
    group_by(lat = cell_ll_lat, lon = cell_ll_lon, gear = geartype) %>%
    summarize(total_fishing_hours = sum(fishing_hours), .groups = "drop") %>%
    as.data.table() # convert back to data.table

  # save summarized data to mazu
  fwrite(summarized_data,
         sprintf("/home/shares/ohi/OHI_GOC/goal_prep/fis/v2025/int/daily_summary_cell_gear/yearly_summaries/yearly_summary_%d", year)) # fwrite is optimized for large datasets, handles millions of rows efficiently... uses package `data.table` vs `readr` for write_csv. Also its default writes the file as a .csv with a "," delim

}

# apply the function to all files in parallel
cl <- makeCluster(20)
registerDoParallel(cl)

# process all daily files in parallel
foreach(filepath = all_filepaths, .packages = c("data.table", "dplyr")) %dopar% {
  process_fishing_data(filepath)
}

# stop cluster after processing is finished
stopCluster(cl)
```

## Yearly files from daily files

Similar code, but taking the next step to combine all daily files by year: 

```{r}
# create filepath for all daily CSVs for 2012:2020
all_years_list_func <- function(year){
  
  gfw_filepath <- "/home/shares/ohi/OHI_GOC/_raw_data/global_fishing_watch/d2025/global_fishing_watch_manual_data_fleet_v2/" # set baseline filepath for all folders of all years in mazu
  
  folder_path <- paste0(gfw_filepath, "fleet-daily-csvs-100-v2-", year) # dynamically replace the year
  
  filepaths <- list.files(folder_path, pattern = ".csv", full = TRUE) # full filepath for the .csvs for that year
  
  return(filepaths) # give the list of file paths, which later needs to be combined into one list for all .csvs to be processed at once
  
}

# define the range of years (again in case you only want to run this chunk)
years <- 2012:2020

# function to combine all daily files for a given year into one data frame
combine_yearly_data <- function(year) {
  
  # compile all file paths for the given year
  filepaths <- all_years_list_func(year)
  
  # read and combine all daily files for the year into one data.frame
  yearly_data <- rbindlist(lapply(filepaths, fread), fill = TRUE) # combines all daily files. fread is faster and better for larger datasets than read_csv
  
  # save the combined yearly data to mazu as a single .csv file
  fwrite(yearly_data, sprintf("/home/shares/ohi/OHI_GOC/goal_prep/fis/v2025/int/yearly_fleet_data/yearly_combined_%d.csv", year))
  
  return(yearly_data)
}

# use lapply to apply the function to all years.  May take a while, especially after 2016 since there are more AIS tracked vessels
lapply(years, combine_yearly_data)

# check_yearly_test_2018 <- read_csv("/home/shares/ohi/OHI_GOC/goal_prep/fis/v2025/int/yearly_fleet_data/yearly_combined_2018.csv")
```
## Summarize by cell and gear type

```{r}
# files within the yearly_fleet_data
yearly_combined_filepath <- here(fis_int_dir, "yearly_fleet_data")

# list yearly files
all_years_files <- list.files(path = yearly_combined_filepath, pattern = "*.csv", full.names = TRUE)

# read in the raw file
tic()
all_years_data <- rbindlist(lapply(all_years_files, fread), fill = TRUE) # fread() is faster than read_csv()
toc()
# v2025: 689.004 sec elapsed

# summarize fishing hours by latitude, longitude, and gear type
tic()
summarized_data <- all_years_data %>%
  group_by(lat = cell_ll_lat, lon = cell_ll_lon, gear = geartype) %>%
  summarize(total_fishing_hours = sum(fishing_hours), .groups = "drop") %>%
  as.data.table() # convert back to data.table
toc()

# save summarized data to mazu
fwrite(summarized_data,
       sprintf("/home/shares/ohi/OHI_GOC/goal_prep/fis/v2025/int/summary_%s", saveName)) # fwrite is optimized for large datasets, handles millions of rows efficiently... uses package `data.table` vs `readr` for write_csv. Also it automatically writes as a .csv with a "," delim

```


```{r}
# process and summarize fishing data for a single year
process_yearly_fishing_data <- function(year) {
  
  # file path for each given year
  filepaths <- paste0(yearly_combined_filepath, "/yearly_combined_", year, ".csv") # dynamically replace the year, full filepath for the .csv for that year
  
  # read the year's data in
  yearly_combined <- read_csv(filepaths) # read in the .csv for that year
  
  # summarize fishing hours by latitude, longitude, and gear type
  summarized_data <- yearly_combined %>%
    group_by(lat = cell_ll_lat, lon = cell_ll_lon, gear = geartype) %>%
    summarize(total_fishing_hours = sum(fishing_hours, na.rm = TRUE), .groups = "drop") 
  
  # output file path
  output_file <- here::here(fis_int_dir, "yearly_summary_cell_gear", sprintf("yearly_summary_%d.csv", year))
  
  # save summarized data as a csv
  write_csv(summarized_data, output_file)
  
  message(sprintf("Year %d processed and saved to %s", year, output_file)) # as a check to see where it is within the process
  
}

# # process all years in parallel
# process_all_in_parallel <- function(years, yearly_combined_filepath, fis_int_dir) {
#  
#   cl <- makeCluster(20)
#   registerDoParallel(cl)
#   
#   # process each year in parallel
#   foreach(year = years, .packages = c("data.table", "tidyverse","readr","dplyr","here"), .export = c("process_yearly_fishing_data")) %dopar% {
#     process_yearly_fishing_data(year, yearly_combined_filepath, fis_int_dir)
#   }
#   
#   # stop the cluster when done
#   stopCluster(cl)
# }

# =============== use functions to get yearly summarized data ================

# files within the yearly_fleet_data
fis_int_dir <- here::here(dir_M, "OHI_GOC", "goal_prep", "fis", "v2025", "int")
yearly_combined_filepath <- here(fis_int_dir, "yearly_fleet_data")

# define the range of years to process
years <- 2012:2020

# loop through each year and process data
for (year in years) {
  process_yearly_fishing_data(year)
}
```

## Limit to GoC Ecoregion

```{r}
# read in ecoregion shapefiles in 
eco_hr_latlon <- st_read(here(goc_spatial, "GoC_polygon_ecoregion_latlon.shp"))
# mapview(eco_hr_latlon)

eco_hr_eqarea <- st_read(here(goc_spatial, "GoC_polygon_ecoregion_eqarea.shp"))
mapview(eco_hr_eqarea)

# check CRS of ecoregion boundary
st_crs(eco_hr_eqarea) # NAD83 / Conus Albers, needs to be transformed

# transform to gulf_CRS
eco_hr_albers <- st_transform(eco_hr_eqarea, crs = "+proj=aea +lat_1=23 +lat_2=30 +lat_0=25 +lon_0=-110 +datum=WGS84 +units=m +no_defs")
st_crs(eco_hr_albers) # double check it worked well
mapview(eco_hr_albers) # looks good!
```

Convert yearly cell lat/longs into spatial geometry.

```{r}
# cl <- makeCluster(10) # CANNOT parallelize!!! Spatial objects (sf objects specifically) can be problematic in parallel loops. The sf package relies on external libraries (like GEOS, GDAL, and PROJ), and these libraries might not be thread-safe, especially if the versions you are using are not built with thread-safety in mind. When multiple cores try to access or modify these objects simultaneously, it can lead to errors, crashes, or unpredictable behavior......... I experienced this first hand.  Took 2 hours.  I should have figured that it would not work if my prints did not start right away....
# registerDoParallel(cl)


years <- 2014:2019

# double check the crs
eco_hr_albers <- st_transform(eco_hr_albers, gulf_crs)

# files within the yearly_fleet_data
yearly_summary_dir <- here(fis_int_dir, "yearly_summary_cell_gear")
yearly_summary_sf_dir <- here(fis_int_dir, "yearly_summary_cell_gear_sf")

st_as_sf_yearly_function <- function(year){
  
  print(paste("Starting year", year))
  
  # read in the yearly file for each year
  yearly_gfw_st <- read_csv(here::here(yearly_summary_dir, sprintf("yearly_summary_%s.csv", year)))
  print(paste("Read yearly data for", year))
  
  
  # convert the tabular data to sf object
  yearly_gfw_sf <- st_as_sf(yearly_gfw_st, coords = c("lon", "lat"), crs = 4326) # the raw data (Lon, Lat) from gfwr::get_raster() is probably in WGS84 coordinates which is why it has to be transformed to the gulf_crs separately!  The `get_raster` function is not working as it should.  It does not limit data to the extent of eco_hr_albers, and instead provides the global data.
  print(paste("Converted to sf object for", year))
  
  # ensure the crs match
  gfw_sf_transformed <- st_transform(yearly_gfw_sf, gulf_crs) # defined in the setup chunk
  print(paste("Transformed CRS for", year))
  
  # crop the effort data to the GoC ecoregion
  yearly_gfw_cropped <- st_intersection(gfw_sf_transformed, eco_hr_albers)
  print(paste("Cropped for", year))
  
  # save in a spatial format using st_write
  st_write(yearly_gfw_cropped, here::here(yearly_summary_sf_dir, sprintf("yearly_summary_sf_%s.shp", year)))
  print(paste("Wrote shapefile for", year))
  
  return(NULL) # do not need anything to return...
  
}

# stopCluster(cl)

# run the function to download
lapply(years, st_as_sf_yearly_function)

# inspect
# gfw_cropped_2019 <- read_csv(here::here(fis_int_dir, "yearly_summary_cell_gear_sf", "yearly_summary_sf_2019.csv"))
# gfw_cropped_2019_sf <- st_as_sf(gfw_cropped_2019, coords = geometry)
```

```{r}
# # read in the yearly file for each year
# yearly_gfw_st <- read_csv(here::here(yearly_summary_dir, "yearly_summary_2019.csv"))
# 
# # convert the tabular data to sf object
# yearly_gfw_sf <- st_as_sf(yearly_gfw_st, coords = c("lon", "lat"), crs = 4326) # the raw data (Lon, Lat) from gfwr::get_raster() is probably in WGS84 coordinates which is why it has to be transformed to the gulf_crs separately!  The `get_raster` function is not working as it should.  It does not limit data to the extent of eco_hr_albers, and instead provides the global data.
# 
# # ensure the crs match
# gfw_sf_transformed <- st_transform(yearly_gfw_sf, gulf_crs) # defined in the setup chunk
# eco_hr_albers <- st_transform(eco_hr_albers, gulf_crs)
# 
# # crop the effort data to the GoC ecoregion
# yearly_gfw_cropped <- st_intersection(gfw_sf_transformed, eco_hr_albers)
# 
# st_write(yearly_gfw_cropped, here::here(fis_int_dir, "yearly_summary_cell_gear_sf", "yearly_summary_sf_2019.tif"))
```





## Food Systems Code


Now combine the daily csv files.
```{r}

fish_effort_df <- list.files("/home/shares/food-systems/Food_footprint/all_food_systems/dataprep/fisheries/marine/disturbance/effort_summarized_by_cell_gear/", pattern = "summary", full=TRUE) %>%
  lapply(data.table::fread) %>%
  bind_rows

fish_effort_annual <- fish_effort_df %>%
  group_by(cell_ll_lat, cell_ll_lon, geartype) %>%
   summarize(fishing_hours = sum(fishing_hours)) %>%
    data.frame()

write_csv(fish_effort_annual, "/home/shares/food-systems/Food_footprint/all_food_systems/dataprep/fisheries/marine/disturbance/global_fishing_watch/FishingWatch_annual_effort_2017.csv")

```

### Determine proportion of fisheries with unidentified geartypes
```{r}

fishing_effort <- read_csv("/home/shares/food-systems/Food_footprint/all_food_systems/dataprep/fisheries/marine/disturbance/global_fishing_watch/FishingWatch_annual_effort_2017.csv")

## quick peak
totals <- fishing_effort %>%
  group_by(geartype) %>%
  summarize(fishing_hours = sum(fishing_hours))
totals$fishing_hours[totals$geartype == "fishing"]/sum(totals$fishing_hours)

#Map all fishing effort
total_map <- fishing_effort %>%
  group_by(cell_ll_lat, cell_ll_lon) %>%
  summarize(fishing_hours = sum(fishing_hours))


total_map$lat <- total_map$cell_ll_lat
total_map$lon <- total_map$cell_ll_lon
total_map <- total_map[,c("lon", "lat","fishing_hours")]
coordinates(total_map) <- ~lon+lat
proj4string(total_map) = CRS("+init=epsg:4326")
total_map_raster <- rasterFromXYZ(total_map, crs = "+init=epsg:4326")
proj4string(total_map_raster) = CRS("+init=epsg:4326")
plot(log(total_map_raster + 1))

#Map only unknowns
unk_map <- fishing_effort %>%
  filter(geartype %in% c("fishing")) %>%
  group_by(cell_ll_lat, cell_ll_lon) %>%
  summarize(fishing_hours = sum(fishing_hours, na.rm=TRUE))


unk_map$lat <- unk_map$cell_ll_lat
unk_map$lon <- unk_map$cell_ll_lon
unk_map <- unk_map[,c("lon", "lat","fishing_hours")]
coordinates(unk_map) <- ~lon+lat
proj4string(unk_map)=CRS("+init=epsg:4326")
unk_map_raster<-rasterFromXYZ(unk_map, crs = "+init=epsg:4326")
proj4string(unk_map_raster)=CRS("+init=epsg:4326")
plot(unk_map_raster)
unk_map_raster[is.na(unk_map_raster)] <- 0

prop_unk <- unk_map_raster/total_map_raster
colr <- rev(RColorBrewer::brewer.pal(4, 'RdYlBu'))
plot(prop_unk, col=colr, breaks=c(-0.01, 0.10, 0.30, 0.66, 1.01))

library(sf)
pts = rasterToPoints(prop_unk, spatial = TRUE)
pts_sf <- st_as_sf(pts)
plot(pts_sf,  pal = colr, pch=16, cex=0.005, breaks=c(-0.01, 0.10, 0.30, 0.66, 1.01))

```

### Create maps of destructive fishing
Create a map including only trawl effort and a map of dredge effort:
```{r}

fishing_effort <- read_csv("/home/shares/food-systems/Food_footprint/all_food_systems/dataprep/fisheries/marine/disturbance/global_fishing_watch/FishingWatch_annual_effort_2017.csv")


## trawl
destructive_map <- fishing_effort %>%
  filter(geartype %in% c("trawlers")) %>%
  group_by(cell_ll_lat, cell_ll_lon) %>%
  summarize(fishing_hours = sum(fishing_hours))


destructive_map$lat <- destructive_map$cell_ll_lat
destructive_map$lon <- destructive_map$cell_ll_lon
destructive_map <- destructive_map[,c("lon", "lat","fishing_hours")]
coordinates(destructive_map) <- ~lon+lat
proj4string(destructive_map)=CRS("+init=epsg:4326")
destructive_map_raster<-rasterFromXYZ(destructive_map, crs = "+init=epsg:4326")
proj4string(destructive_map_raster)=CRS("+init=epsg:4326")
plot(destructive_map_raster)

writeRaster(destructive_map_raster, "/home/shares/food-systems/Food_footprint/all_food_systems/dataprep/fisheries/marine/disturbance/global_fishing_watch/FishingWatch_annual_effort_trawlers_2017.tif", overwrite=TRUE)


## dredge
destructive_map <- fishing_effort %>%
  filter(geartype %in% c("dredge_fishing")) %>%
  group_by(cell_ll_lat, cell_ll_lon) %>%
  summarize(fishing_hours = sum(fishing_hours))


destructive_map$lat <- destructive_map$cell_ll_lat
destructive_map$lon <- destructive_map$cell_ll_lon
destructive_map <- destructive_map[,c("lon", "lat","fishing_hours")]
coordinates(destructive_map) <- ~lon+lat
proj4string(destructive_map)=CRS("+init=epsg:4326")
destructive_map_raster<-rasterFromXYZ(destructive_map, crs = "+init=epsg:4326")
proj4string(destructive_map_raster)=CRS("+init=epsg:4326")
plot(destructive_map_raster)

writeRaster(destructive_map_raster, "/home/shares/food-systems/Food_footprint/all_food_systems/dataprep/fisheries/marine/disturbance/global_fishing_watch/FishingWatch_annual_effort_dredge_2017.tif", overwrite=TRUE)


```


### Maps of fishing effort by gear

```{r}
fishing_effort <- st_read("/home/shares/ohi/OHI_GOC/goal_prep/fis/v2025/int/yearly_summary_cell_gear_sf/yearly_summary_sf_2020.shp")

# explore the .shp to ensure we understand the attributes before plotting
colnames(fishing_effort)
str(fishing_effort)
fishing_effort[1:5,]
mapview(fishing_effort)

# DID NOT LOOK GOOD...
# ggplot() +
#   # borders("world", fill = "gray95", color = "gray70") +
#   # plot points with size based on count
#   geom_sf(data = fishing_effort_count, 
#           aes(size = point_count, color = point_count),
#           alpha = 0.7) +
#   # Use a better color scale
#   scale_color_viridis_c(option = "plasma", name = "Fishing\nEffort") +
#   scale_size_continuous(name = "Point\nDensity", range = c(1, 8)) +
#   # Improve theme elements
#   theme_minimal() +
#   labs(title = "Fishing Effort Distribution (2020)",
#        subtitle = "Circle size indicates concentration of fishing activity") +
#   theme(
#     plot.title = element_text(face = "bold"),
#     legend.position = "right"
#   )

# Assuming fishing_effort is already loaded and is in a projected coordinate system
# First, check the current coordinate system
st_crs(fishing_effort)

# Transform to WGS84 (EPSG:4326) which is what leaflet expects
fishing_effort_wgs84 <- st_transform(fishing_effort, 4326)

# Now create your count summary with the transformed data
colnames(fishing_effort_wgs84)

# Then create your summary with the effort variable
fishing_effort_summary <- fishing_effort_wgs84 %>%
  group_by(geometry, gear) %>% 
  summarize(
    point_count = n(),
    total_effort = sum(ttl_fs_)  # Replace 'effort' with your actual column name
  )

# Create a color palette for the different gear types
# Get unique gear types
gear_types <- unique(fishing_effort_summary$gear)
# Create a qualitative palette with enough colors
gear_pal <- colorFactor(
  palette = colorRampPalette(brewer.pal(8, "Set1"))(length(gear_types)),
  domain = gear_types
)

# Create interactive map
leaflet(fishing_effort_summary) %>%
  addTiles() %>%
  addCircleMarkers(
    radius = ~sqrt(total_effort) * 3,  # Size based on fishing effort
    color = ~gear_pal(gear),           # Color based on gear type
    fillOpacity = 0.7,
    stroke = TRUE,
    weight = 1,
    popup = ~paste(
      "Gear Type: ", gear, "<br>",
      "Fishing Effort: ", round(total_effort, 2), "<br>",
      "Points: ", point_count
    )
  ) %>%
  # Add legend for gear types
  addLegend(
    position = "bottomright",
    pal = gear_pal,
    values = ~gear,
    title = "Gear Type"
  ) %>%
  # Add legend for effort (optional, makes a second legend)
  addLegend(
    position = "bottomleft",
    colors = viridisLite::viridis(5),
    labels = quantile(fishing_effort_summary$total_effort, probs = seq(0, 1, 0.25)) %>% round(1), # shows fishing effort values at 0% (minimum value), 25% (first quartile), 50% (median), 75% (third quartile), 100% (maximum value).  shows the data distribution... most are 0 fishing effort.  That is fishy to me (pun intended).
    title = "Fishing Effort<br>(Circle Size)",
    opacity = 0.7
  ) %>%
  # Set the view to focus on the Gulf of California
  setView(lng = -112, lat = 28, zoom = 6)
```








