---
title: 'OHI GoC 2025: Clean waters Untreated Wastewater'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# CW: Chemicals - Untreated Wastewater feeding into the ocean

Data came from the National Census of Municipal Governments and Territorial Demarcations of Mexico City (CNGMD).

> The National Census of Municipal Governments and Territorial Demarcations of Mexico City 2023, has the objective to generate statistical and geographic information on the integration and functioning of city councils and boroughs, as well as the management and performance of the institutions that make up the Public Administration of each municipality or territorial demarcation, specifically in the functions of government, public safety, civil protection, civic justice, potable water and sanitation, and urban solid residues. Its purpose is to link the data with the design, implementation, monitoring and evaluation of public policies of national scope in these functions.

"Puntos de descarga de aguas residuales municipales vertidas sin tratamiento que existían en el municipio o demarcación territorial."

The data also notes where the untreated wastewater emptied into:

res_trec	descrip
0	No aplica/Sin respuesta
1	Río o arroyo
2	Lago o laguna
3	Presa
4	Suelo o barranca
5	Canal o dren
6	Mar
7	Playa
8	Estero, marisma
9	Gran colector
10	Otro

**We are interested in 6 Mar, 7 Playa, 8 Estero/marisma** (2021/2023 coding).

Earlier census years use different receptor code schemes; each year is mapped to this 2023 standard before filtering. See `cursor_generated/README.md` for a full year-by-year comparison.

A continuación, se presenta la descripción de caracteres (siglas) para el manejo de valores que indican o representan algún estatus dentro de la información de los conjuntos de datos:

-	NSS: No cuenta con elementos para responder

	Se utiliza para identificar las variables en las que el informante, por falta de información, respondió no saber la respuesta; así como en aquellas variables que, por su carácter condicionado a otras respuestas, mantienen el mismo estatus.

-	NA: No aplica

	Se utiliza para identificar las variables que, por la naturaleza de la pregunta correspondiente, no aplica proporcionar información; así como en aquellas variables que, por su carácter condicionado a otras respuestas, no deben ser respondidas.

-	ND: Información no disponible

	Se utiliza para identificar las variables en las que, por diversas razones, el informante no respondió el cuestionario.

-	NP: No publicable por el principio de confidencialidad o por reserva de información

	Se utiliza para identificar las variables en las que el informante proporcionó información, pero que, por confidencialidad o reserva, no es susceptible de ser publicada.

```{r setup}
# load packages for spatial and tabular work
# we need sf/mapview for future mapping, tidyverse for cleaning, and stringi for accent-insensitive municipality name matching across catalogs
library(sf)
library(utils)
library(tidyverse)
library(here)
library(raster)
library(mapview)
library(purrr)
library(stringr)
library(stringi)
library(janitor)
library(tibble)
```

## Explore each year and determine if we can automate cleaning...

Data downloaded and placed at: `~/home/shares/ohi/OHI_GOC/_raw_data/CNGMD_censacional_gobiernos_municipales_demarcaciones_territoriales/agua_sin_tratamiento`.

When running locally, raw data lives in the project root (one level above this file in `cursor_generated/`):

```{r paths}
# set data directory
# here() resolves to the project root, so paths work whether this .Rmd lives in
# cursor_generated/ or at the top level
data_dir <- here::here(
  "CNGMD_censacional_gobiernos_municipales_demarcaciones_territoriales",
  "agua_sin_tratamiento"
)

# uncomment for ohi server data path
# data_dir <- "/home/shares/ohi/OHI_GOC/_raw_data/CNGMD_censacional_gobiernos_municipales_demarcaciones_territoriales/agua_sin_tratamiento"

# set municipality to region lookup path
# this file maps normalized municipality names to OHI Gulf of California region ids
municipality_to_rgnid_path <- here::here(
  "spatial", "municipality_to_rgnid", "municipality_to_rgnid.csv"
)

# uncomment for ohi server rgnid path
# municipality_to_rgnid_path <- "/home/shares/ohi/OHI_GOC/spatial/municipality_to_rgnid/municipality_to_rgnid.csv"

# define gulf of california states
# we only retain discharge points in these six coastal states
GOC_STATES <- c(
  "Baja California", "Baja California Sur", "Jalisco",
  "Nayarit", "Sinaloa", "Sonora"
)

# define ocean receptor codes in 2023 standard
# after harmonization, codes 6/7/8 correspond to Mar/Playa/Estero-marisma
OCEAN_RES_TREC_STD <- c(6L, 7L, 8L)

# build reference catalog for harmonized receptor codes
# all years are mapped onto this 2023 coding scheme before filtering
res_trec_catalog_std <- tibble::tribble(
  ~res_trec_std, ~descrip_std,
  0L,  "No aplica/Sin respuesta",
  1L,  "Río o arroyo",
  2L,  "Lago o laguna",
  3L,  "Presa",
  4L,  "Suelo o barranca",
  5L,  "Canal o dren",
  6L,  "Mar",
  7L,  "Playa",
  8L,  "Estero, marisma",
  9L,  "Gran colector",
  10L, "Otro"
)
```

Data dictionary for 2023:

```{r}
# load 2023 data dictionary for reference
diccionario_de_datos_2023 <- read_csv(
  file.path(
    data_dir,
    "aguas_residuales_sin_tratamiento_cngmd2023_csv/diccionario_de_datos",
    "diccionario_de_datos_aguaresd_cngmd2023.csv"
  )
)
```

### Helper functions

Each census year uses different column names and receptor-code schemes. The functions below harmonize municipality identifiers and map receptor codes to the 2023 standard before filtering to GoC municipalities and ocean discharge points.

```{r helpers}
# normalize mun name
# lower-case, trim, and strip accents so municipality names match across years
# and between census catalogs and the municipality_to_rgnid lookup table
normalize_mun_name <- function(x) {
  x %>%
    str_trim() %>%
    str_to_lower() %>%
    stri_trans_general("Latin-ASCII")
}

# normalize mnpio
# coerce municipality/folio ids to a 5-digit integer (e.g. "01001" and 1001 -> 1001)
# this is required because some years store folio as character with leading zeros
# while others store it as a bare integer
normalize_mnpio <- function(x) {
  as.integer(str_pad(as.character(suppressWarnings(as.integer(x))), 5, pad = "0"))
}

# read goc municipality codes
# read a year's municipality catalog, attach entity names, and filter to GoC states
# mun_id_col / mun_name_col / ent_id_col differ across census years (see README)
read_goc_municipality_codes <- function(municipio_codes, entidad_codes,
                                        mun_id_col, mun_name_col, ent_id_col) {
  municipio_codes %>%
    rename(
      mnpio = !!sym(mun_id_col),
      descrip_mun_raw = !!sym(mun_name_col),
      entidad = !!sym(ent_id_col)
    ) %>%
    mutate(
      mnpio = normalize_mnpio(mnpio),
      descrip_mun = normalize_mun_name(descrip_mun_raw),
      entidad = as.integer(entidad)
    ) %>%
    select(mnpio, descrip_mun, entidad, descrip_mun_raw) %>%
    left_join(
      entidad_codes %>%
        rename(descrip_ent = descrip),
      by = "entidad"
    ) %>%
    filter(descrip_ent %in% GOC_STATES)
}

# load municipality to rgnid
# prepare the OHI spatial lookup so it can join on normalized municipality names
load_municipality_to_rgnid <- function(path) {
  read_csv(path, show_col_types = FALSE) %>%
    mutate(
      descrip_mun = normalize_mun_name(municipio),
      rgn_id = as.integer(rgn_id)
    ) %>%
    select(descrip_mun, rgn_id)
}

# attach rgn id
# join region ids and drop municipalities with no OHI region assignment
# (e.g. Comondu is intentionally excluded when it would span both coasts)
attach_rgn_id <- function(goc_codes, municipality_to_rgnid) {
  goc_codes %>%
    left_join(municipality_to_rgnid, by = "descrip_mun") %>%
    filter(!is.na(rgn_id))
}

# map res trec to std
# translate year-specific receptor codes to the 2023 standard (res_trec_std)
# 2021/2023 already use 6=Mar, 7=Playa, 8=Estero/marisma
# 2013-2019 only distinguish Mar (code 5) and have no Playa/Estero categories
map_res_trec_to_std <- function(res_trec, census_year) {
  code <- suppressWarnings(as.integer(res_trec))
  dplyr::case_when(
    is.na(code) ~ NA_integer_,
    census_year %in% c(2021L, 2023L) ~ code,
    census_year %in% c(2013L, 2015L, 2017L, 2019L) & code == 5L ~ 6L,
    TRUE ~ NA_integer_
  )
}

# is ocean discharge
# flag rows whose harmonized receptor code indicates direct ocean discharge
is_ocean_discharge <- function(res_trec_std) {
  res_trec_std %in% OCEAN_RES_TREC_STD
}

# clean special values
# replace census missing-value codes (NSS, NA, ND, NP, NULL) with proper NA
clean_special_values <- function(x) {
  x <- str_trim(as.character(x))
  special <- c("NSS", "NA", "ND", "NP", "NULL", "")
  ifelse(toupper(x) %in% special, NA_character_, x)
}
```

### Municipality lookup for GoC

```{r municipality-lookup}
# load 2023 municipality and entity catalogs
# we use the 2023 catalogs as the master municipality list because they are the
# most recent and already include demarcaciones territoriales de CDMX
municipio_codes_2023 <- read_csv(
  file.path(data_dir, "aguas_residuales_sin_tratamiento_cngmd2023_csv/catalogos/mnpio_cngmd2023.csv"),
  show_col_types = FALSE
)

entidad_codes_2023 <- read_csv(
  file.path(data_dir, "aguas_residuales_sin_tratamiento_cngmd2023_csv/catalogos/entidad_cngmd2023.csv"),
  show_col_types = FALSE
)

# build goc municipality code table
goc_municipio_and_ent_codes <- read_goc_municipality_codes(
  municipio_codes = municipio_codes_2023,
  entidad_codes = entidad_codes_2023,
  mun_id_col = "mnpio",
  mun_name_col = "descrip",
  ent_id_col = "entidad"
)

# load and attach region ids
municipality_to_rgnid <- load_municipality_to_rgnid(municipality_to_rgnid_path)

# check municipality name mismatches between census and rgnid lookup
# character(0) means all GoC municipalities matched successfully
setdiff(
  municipality_to_rgnid$descrip_mun,
  goc_municipio_and_ent_codes$descrip_mun
)

# finalize goc codes with region ids
goc_codes <- attach_rgn_id(goc_municipio_and_ent_codes, municipality_to_rgnid)
```

Combine with the lookup table of municipalities to region id.

All years previous to 2023...

```{r read-raw-years}
# read raw discharge tables one per year
# each file has a different name, path, and column structure (see README)
aguaresd_cngmd2023 <- read_csv(
  file.path(data_dir, "aguas_residuales_sin_tratamiento_cngmd2023_csv/conjunto_de_datos/aguaresd_cngmd2023.csv"),
  show_col_types = FALSE
)

aguaresd_cngmd2021 <- read_csv(
  file.path(data_dir, "aguas_residuales_sin_tratamiento_cngmd2021_csv/conjunto_de_datos/aguaresd_cngmd2021.csv"),
  show_col_types = FALSE
)

### each year after is SOOO different!!
aguaresd_cngmd2019 <- read_csv(
  file.path(data_dir, "aguas_sintrat_cngmd2019_csv/conjunto_de_datos/secc_vii_td_m5_2018_agua_resd.csv"),
  show_col_types = FALSE
)

aguaresd_cngmd2017 <- read_csv(
  file.path(data_dir, "aguas_sintrat_cngmd2017_csv/conjunto_de_datos/secc_vii_td_m5_2016_agua_resd.csv"),
  show_col_types = FALSE
)

aguaresd_cngmd2015 <- read_csv(
  file.path(data_dir, "Aguas_sintrat_cngmd2015_csv/conjunto_de_datos/secc_vii_tr_cngmd15_m5_p9_1.csv"),
  show_col_types = FALSE
)

aguaresd_cngmd2013 <- read_csv(
  file.path(data_dir, "m5_vii_Aguas_residuales_cngmd2013_csv/conjunto_de_datos/tr_cengob13_5_vii_ar.csv"),
  show_col_types = FALSE
)
```

Limit all 2023 WWTPs to GoC municipalities

```{r explore-2023}
# exploratory join for 2023 only
# quick check of receptor codes present in GoC before running the full pipeline
goc_aguaresd_cngmd2023 <- left_join(goc_codes, aguaresd_cngmd2023, by = "mnpio")

unique(goc_aguaresd_cngmd2023$res_trec)
# **We are interested in 6 Mar, 7 Playa, 8 Estero/marisma.**
```

## Bring in and clean each year.

Year-specific receptor-code schemes:

| Census year | Receptor variable | Ocean code(s) in raw data | Mapped to 2023 std |
|-------------|-------------------|---------------------------|--------------------|
| 2023        | `res_trec`        | 6, 7, 8                   | unchanged          |
| 2021        | `cuerprec`        | 6, 7, 8                   | unchanged          |
| 2019        | `res_trec`        | 5 (Mar)                   | 5 ? 6              |
| 2017        | `agua_resd_tip_crp_recp` | 5 (Mar)              | 5 ? 6              |
| 2015        | `P9_1_2`          | 5 (Mar)                   | 5 ? 6              |
| 2013        | `P11_CPO_RP`      | 5 (Mar)                   | 5 ? 6              |

```{r clean-each-year}
# clean cngmd year
# read one census year, rename columns to a common schema, harmonize codes,
# join GoC municipality metadata, and return a standardized tibble
clean_cngmd_year <- function(census_year, goc_codes) {

  if (census_year == 2023L) {
    # clean 2023
    # most recent schema: mnpio, res_trec, res_caud (flow in L/s)
    raw <- read_csv(
      file.path(data_dir, "aguas_residuales_sin_tratamiento_cngmd2023_csv/conjunto_de_datos/aguaresd_cngmd2023.csv"),
      show_col_types = FALSE
    ) %>%
      mutate(
        census_year = census_year,
        mnpio = normalize_mnpio(mnpio),
        res_trec_raw = suppressWarnings(as.integer(res_trec)),
        res_caud = clean_special_values(res_caud)
      )
  }

  if (census_year == 2021L) {
    # clean 2021
    # same structure as 2023 but municipality id is called folio and receptor
    # code is called cuerprec instead of res_trec
    raw <- read_csv(
      file.path(data_dir, "aguas_residuales_sin_tratamiento_cngmd2021_csv/conjunto_de_datos/aguaresd_cngmd2021.csv"),
      show_col_types = FALSE
    ) %>%
      mutate(
        census_year = census_year,
        mnpio = normalize_mnpio(folio),
        res_trec_raw = suppressWarnings(as.integer(cuerprec)),
        res_caud = clean_special_values(res_caud)
      )
  }

  if (census_year == 2019L) {
    # clean 2019
    # flow reported as volume (res_v_m3 / res_v_es) rather than res_caud;
    # no cve_geo; folio stored as zero-padded character
    raw <- read_csv(
      file.path(data_dir, "aguas_sintrat_cngmd2019_csv/conjunto_de_datos/secc_vii_td_m5_2018_agua_resd.csv"),
      show_col_types = FALSE
    ) %>%
      mutate(
        census_year = census_year,
        mnpio = normalize_mnpio(folio),
        res_nom = str_squish(res_nom),
        res_trec_raw = suppressWarnings(as.integer(res_trec)),
        res_caud = NA_character_,
        res_v_m3 = clean_special_values(res_v_m3),
        res_v_es = clean_special_values(res_v_es),
        cve_geo = NA_character_
      )
  }

  if (census_year == 2017L) {
    # clean 2017
    # minimal table: only id, folio, and receptor type; no name or flow columns
    raw <- read_csv(
      file.path(data_dir, "aguas_sintrat_cngmd2017_csv/conjunto_de_datos/secc_vii_td_m5_2016_agua_resd.csv"),
      show_col_types = FALSE
    ) %>%
      rename(id_resd = agua_resd_id_serial) %>%
      mutate(
        census_year = census_year,
        mnpio = normalize_mnpio(agua_resd_folio),
        res_nom = NA_character_,
        res_trec_raw = suppressWarnings(as.integer(agua_resd_tip_crp_recp)),
        res_caud = NA_character_,
        cve_geo = NA_character_
      )
  }

  if (census_year == 2015L) {
    # clean 2015
    # P9_1_1 = discharge point name, P9_1_2 = receptor type code
    # P9_1_2_1 holds free-text when "other" is selected (not used for ocean filter)
    raw <- read_csv(
      file.path(data_dir, "Aguas_sintrat_cngmd2015_csv/conjunto_de_datos/secc_vii_tr_cngmd15_m5_p9_1.csv"),
      show_col_types = FALSE
    ) %>%
      rename(id_resd = ID_GR111224807) %>%
      mutate(
        census_year = census_year,
        mnpio = normalize_mnpio(FOLIO),
        res_nom = str_squish(P9_1_1),
        res_trec_raw = suppressWarnings(as.integer(P9_1_2)),
        res_caud = NA_character_,
        cve_geo = NA_character_
      )
  }

  if (census_year == 2013L) {
    # clean 2013
    # sparsest table: only folio and receptor code (P11_CPO_RP); no id, name, or flow
    raw <- read_csv(
      file.path(data_dir, "m5_vii_Aguas_residuales_cngmd2013_csv/conjunto_de_datos/tr_cengob13_5_vii_ar.csv"),
      show_col_types = FALSE
    ) %>%
      mutate(
        census_year = census_year,
        id_resd = NA_integer_,
        mnpio = normalize_mnpio(FOLIO),
        res_nom = NA_character_,
        res_trec_raw = suppressWarnings(as.integer(P11_CPO_RP)),
        res_caud = NA_character_,
        cve_geo = NA_character_
      )
  }

  # harmonize receptor codes and join goc metadata
  raw %>%
    mutate(
      res_trec_std = map_res_trec_to_std(res_trec_raw, census_year),
      ocean_discharge = is_ocean_discharge(res_trec_std)
    ) %>%
    left_join(
      goc_codes %>% select(mnpio, entidad, descrip_ent, descrip_mun, rgn_id),
      by = "mnpio"
    ) %>%
    left_join(res_trec_catalog_std, by = c("res_trec_std" = "res_trec_std")) %>%
    filter(!is.na(rgn_id)) %>%
    select(
      census_year,
      any_of(c("id_resd", "mnpio", "cve_geo", "res_nom", "res_caud", "res_v_m3", "res_v_es")),
      res_trec_raw,
      res_trec_std,
      descrip_std,
      ocean_discharge,
      entidad,
      descrip_ent,
      descrip_mun,
      rgn_id
    )
}

# run cleaning pipeline for all census years
cngmd_years <- c(2013L, 2015L, 2017L, 2019L, 2021L, 2023L)

goc_aguaresd_all_years <- map_dfr(cngmd_years, clean_cngmd_year, goc_codes = goc_codes)

# subset to ocean discharge points only
# these are the rows used for the clean waters untreated wastewater indicator
goc_aguaresd_ocean <- goc_aguaresd_all_years %>%
  filter(ocean_discharge)
```

### Summary by year

```{r summary-by-year}
# count all goc discharge points by year
goc_aguaresd_all_years %>%
  count(census_year, name = "n_goc_discharge_points") %>%
  arrange(census_year)

# count ocean discharge points by year and receptor type
goc_aguaresd_ocean %>%
  count(census_year, res_trec_std, descrip_std, name = "n_ocean_discharge_points") %>%
  arrange(census_year, res_trec_std)
```

### Inspect ocean discharge points

```{r inspect-ocean}
# review ocean discharge records
goc_aguaresd_ocean %>%
  arrange(census_year, descrip_ent, descrip_mun, res_nom) %>%
  select(
    census_year, mnpio, descrip_ent, descrip_mun, rgn_id,
    res_nom, res_caud, res_trec_raw, res_trec_std, descrip_std
  )
```

Unique receptor codes observed in GoC municipalities (all discharge points, not just ocean):

```{r unique-codes-by-year}
# audit raw and harmonized receptor codes by year
goc_aguaresd_all_years %>%
  group_by(census_year) %>%
  summarise(
    res_trec_raw = list(sort(unique(res_trec_raw))),
    res_trec_std = list(sort(unique(res_trec_std))),
    .groups = "drop"
  )
```
