################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```

## clean_names_lowercase function
```{r Load clean_names_lowercase function}
# Takes text and standardizes it, which we will need for merging later on:

clean_names_lowercase <- function(x) {

  x <- iconv(x, from = "", to = "UTF-8", sub = "")
  x <- stringi::stri_trans_general(x, "Latin-ASCII")

  x %>%
    tolower() %>%
    trimws()
}
```

## lowercase_function function
```{r lowercase_function}
# Turns all entries lower-case
lowercase_function<- function(x){
  if(is.character(x)){
    tolower(x)
  } else if (is.factor(x)){
    tolower(as.character(x))
  } else {
    x
  }
}
```

# STEP 0: Set Global Parameters

# **DECISION POINT: Carta Scoring Years, 2018-2024 **
```{r Carta Scoring Years}
SCORING_YEARS <- 2018:2024
```


# PICK 2: Catch vs CNP Baseline - species level ------------------------------


# STEP 1: Load in raw data, original recommendations for the Carta Nacional Pesquera
```{r Load in Carta data}
Carta_data0<- read.csv("/home/shares/ohi/OHI_GOC/goal_prep/fp/fis/v2025/int/carta/CartaNacional_Data_AllYears_Mar2026.csv")

dim(Carta_data0)
# 2699 obs., 12 vars.
```

```{r Clean Carta_data0 as needed}
# Separate Genus from scientific_name
Carta_data <- Carta_data0 %>%
  separate(scientific_name, into = c("estimated_genus_name", "species"), sep = " ", remove = FALSE)

table(Carta_data$estimated_genus_name)
# Okay, looks good! Except for Pseudobatos (currently labeled Rhinobatos and R.)

Carta_data <- Carta_data %>%
  mutate(
    notes = if_else(
      estimated_genus_name %in% c("Rhinobatos", "R."),
      "Assigned new estimated_genus_name, Pseudobatos",
      notes
    ),
    
    estimated_genus_name = if_else(
      estimated_genus_name %in% c("Rhinobatos", "R."),
      "Pseudobatos",
      estimated_genus_name
    )
  )

# Make columns clean and lowercase for safe merging:
Carta_data<- Carta_data %>%
  mutate(
    estimated_species_name= clean_names_lowercase(scientific_name),
    estimated_genus_name= clean_names_lowercase(estimated_genus_name)
  )
```


# STEP 2: Assign major groups to the Carta_data using the GenusFamilyGroup.Key
```{r Load in GenusFamilyGroup.Key and clean as needed}
GenusFamilyGroup.Key0<- read.csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Keys/CONAPESCA_Keys/Taxonomic_Keys/GenusFamilyGroupKey_COMPLETE_Apr042026.csv")
# 193 obs., 7 vars.
```

```{r Merge family details to Carta_data}
GenusFamilyGroup.Key<- GenusFamilyGroup.Key0 %>% 
  select(estimated_genus_name, estimated_taxa_group, Major_Functional_Group) %>%
  distinct()
# 193 obs., 3 vars.

# Make columns clean and lowercase for safe merging:
GenusFamilyGroup.Key<- GenusFamilyGroup.Key %>%
  mutate(
    estimated_genus_name= clean_names_lowercase(estimated_genus_name)
  )

# Merge family details to Carta_data:
Carta_data <- Carta_data %>%
  left_join(
    GenusFamilyGroup.Key,
    by = c("estimated_genus_name")
  )
# 2699 obs., 17 vars. (Good, as expected!)
```


# STEP 3: Create the Carta_Recommendations_clean based on the data

```{r Make a Carta_Recommendations0}
colnames(Carta_data)

Carta_Recommendations0<- Carta_data %>% 
  select(Year_of_Recommendation= carta_year, Major_Functional_Group, 
         estimated_species_name, estimated_genus_name, common_name, 
         OHI_region, 
         status_raw= status_adjusted, control_raw= control_adjusted, notes)
# 2699 obs., 9 vars.
```

```{r Create Carta_Recommendations - Part 1}
colnames(Carta_Recommendations0)

# Filter this down to unique number of species (disregarding common_name)
Carta_Recommendations<- Carta_Recommendations0 %>% 
  select(Year_of_Recommendation, Major_Functional_Group, 
         estimated_species_name, estimated_genus_name, 
         OHI_region, 
         status_raw, control_raw, notes) %>%
  distinct()
# 2692 obs., 8 vars.
# Suggests there were 7 duplicates.

# SANITY CHECK: Why is this key shorter from Carta_Recommendations0?
colnames(Carta_Recommendations)

common_cols <- intersect(names(Carta_Recommendations0),
                         names(Carta_Recommendations))

TEST<- anti_join(
  Carta_Recommendations0,
  Carta_Recommendations,
  by = names(Carta_Recommendations)
)
# Multiple recommendations for distinct species under different common names: common_name
```

```{r Create Carta_Recommendations - Part 2}
colnames(Carta_Recommendations)

table(Carta_Recommendations$status_raw)
  #                                decreasing         deteriorating development potential 
  #                1134                     4                    44                    26 
  # maximally harvested               unclear 
  #                 936                   112 
table(Carta_Recommendations$control_raw)
       #         decrease    increase  increasing no increase 
       # 1134          44          40           4         918


# Generate outputs regarding status_adjusted and control_adjusted at the species level:
Carta_Recommendations <- Carta_Recommendations %>%
  
  group_by(
    estimated_species_name,
    OHI_region,
    Year_of_Recommendation
  ) %>%
  
  summarise(

# STATUS:
    
    status_raw = {

      valid_status <- status_raw[!is.na(status_raw)]

      if (length(valid_status) == 0) {
        
        NA_character_

      } else if ("deteriorating" %in% valid_status) {

        "deteriorating"

      } else if ("improving" %in% valid_status) {

        "improving"

      } else {

        valid_status[1]
      }
    },


# CONTROL:

    control_raw = {

      valid_control <- control_raw[!is.na(control_raw)]

      if (length(valid_control) == 0) {

        NA_character_

      } else {

        valid_control[1]
      }
    },


# NOTES:

    notes = {

      valid_notes <- notes[
        !is.na(notes) & notes != ""
      ]

      if (length(valid_notes) == 0) {

        NA_character_

      } else {

        valid_notes[1]
      }
    },

    .groups = "drop"
  )
# 2692 obs.,  6 vars.
```

```{r Remove any rows where the control_adjusted and recommendation are both NA or "unknown"}
# Remove any rows where the control_adjusted and recommendation are both NA or "unknown"
table(Carta_Recommendations$status_raw) # 1134 NAs
table(Carta_Recommendations$control_raw) # 1134 NAs

Carta_Recommendations <- Carta_Recommendations %>%
  filter(
    !(
      (is.na(status_raw) | status_raw == "unclear" | status_raw== "") &
      (is.na(control_raw) | control_raw == "unclear" | control_raw == "")
    )
  )
# 1010 obs., 6 vars.
```

```{r Clean up rows with status_raw and control_raw that are in conflict}
table(Carta_Recommendations$status_raw)
           # decreasing         deteriorating development potential   maximally harvested 
           #          4                    44                    26                   936 
table(Carta_Recommendations$control_raw)
  # decrease    increase  increasing no increase 
  #        44          40           4         918 

# Scenarios where stock status_adjusted is "decreasing" or "maximally harvested" but control_adjusted is "increase" or "increasing"
TEST<- Carta_Recommendations %>%
  filter(status_raw %in% c("decreasing", "deteriorating", "maximally harvested") &
           control_raw %in% c("increase", "increasing"))
# 22 obs., 6 vars.

Carta_Recommendations_clean0 <- Carta_Recommendations %>%
  mutate(
    status_adjusted = case_when(
      status_raw %in% c("decreasing", "deteriorating") ~ "decreasing",
      TRUE ~ status_raw
    ),

    control_adjusted = case_when(
      status_raw %in% c("decreasing", "deteriorating") &
        (is.na(control_raw) | control_raw %in% c("increase", "increasing")) ~ "decrease",

      status_raw == "maximally harvested" &
        (is.na(control_raw) | control_raw %in% c("increase", "increasing")) ~ "no increase",
      
      status_raw == "development potential" &
        (is.na(control_raw)) ~ "increase",

      TRUE ~ control_raw
    ),
    
    control_adjusted= case_when(
      control_adjusted %in% c("increase", "increasing") ~ "increase",
      TRUE ~ control_adjusted
    )
  )

# SANITY CHECK: were the entries reassigned as expected? Yes, I believe so.
table(Carta_Recommendations_clean0$status_adjusted)
   # decreasing development potential   maximally harvested 
   #         48                    26                   936 
table(Carta_Recommendations_clean0$control_adjusted)
   # decrease    increase no increase 
   #       48          24         938 

Carta_Recommendations_clean0 %>%
  filter(!is.na(status_adjusted)) %>%
  count(status_adjusted, control_adjusted, name = "n")
# Looks good!
```


## STEP 4: SAVE the Carta_Recommendations_clean
```{r SAVE the Carta_Recommendations_clean as necessary}
write.csv(Carta_Recommendations_clean0,
          file="/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_Carta_Scores/Carta_RawRecommendations_Cleaned_Jun032026.csv",
          row.names = FALSE)
# 1010 obs., 8 vars.
```


# > SKIP TO HERE:

# STEP 5A: Create Carta_SpeciesRec_Lookup
```{r Load in Carta_Recommendations_clean as needed}
Carta_Recommendations_clean0<- read.csv("/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_Carta_Scores/Carta_RawRecommendations_Cleaned_Jun032026.csv")
# 1010 obs., 8 vars.
```

```{r Clean as needed}
colnames(Carta_Recommendations_clean0)

Carta_Recommendations_clean<- Carta_Recommendations_clean0 %>%
  select(
    estimated_species_name, OHI_region, Year_of_Recommendation,
    status_adjusted, control_adjusted, notes
  )
```

```{r Create Carta_SpeciesRec_Lookup - Part 1}
colnames(Carta_Recommendations_clean)

# Input is already one row per species × region (no duplicate keys).
Carta_Recommendations_clean <- Carta_Recommendations_clean %>%
  distinct()
# 1010 obs., 6 vars.

# How many species-region combinations are missing from the input key?
OHI_regions_all <- paste0("region_", 1:9)

Carta_Recommendations_grid <- expand.grid(
  estimated_species_name = unique(Carta_Recommendations_clean$estimated_species_name),
  OHI_region = OHI_regions_all,
  stringsAsFactors = FALSE
)
# 1359 obs., 2 obs. (1359 = 9 x 151)

Carta_Recommendations_missing <- anti_join(
  Carta_Recommendations_grid,
  Carta_Recommendations_clean %>%
    select(estimated_species_name, OHI_region),
  by = c("estimated_species_name", "OHI_region")
)
# 349 missing species × region combinations to gapfill.
```

```{r Create Carta_SpeciesRec_Lookup - Part 2 species-level modes}
colnames(Carta_Recommendations_clean)

# For each species, take the mode of status_adjusted and control_adjusted across regions where values exist.
species_modes <- Carta_Recommendations_clean %>%
  group_by(estimated_species_name) %>%
  summarise(
    Year_of_Recommendation = Year_of_Recommendation[
      !is.na(Year_of_Recommendation)
    ][1],

    status_mode = {
      valid_status <- status_adjusted[!is.na(status_adjusted) & status_adjusted != ""]
      if (length(valid_status) == 0) {
        NA_character_
      } else {
        names(sort(table(valid_status), decreasing = TRUE))[1]
      }
    },

    control_mode = {
      valid_control <- control_adjusted[!is.na(control_adjusted) & control_adjusted != ""]
      if (length(valid_control) == 0) {
        NA_character_
      } else {
        names(sort(table(valid_control), decreasing = TRUE))[1]
      }
    },

    .groups = "drop"
  )
# 151 obs., 4 vars.
```

```{r Create Carta_SpeciesRec_Lookup - Part 3 gapfill to full species × region grid}
colnames(Carta_Recommendations_clean)

Carta_SpeciesRec_Lookup <- Carta_Recommendations_grid %>%
  left_join(
    Carta_Recommendations_clean,
    by = c("estimated_species_name", "OHI_region")
  ) %>%
  left_join(
    species_modes %>%
      select(
        estimated_species_name,
        Year_species = Year_of_Recommendation,
        status_mode,
        control_mode
      ),
    by = "estimated_species_name"
  ) %>%
  mutate(
    status_original = status_adjusted,
    control_original = control_adjusted,
    original_row_present = !is.na(Year_of_Recommendation),

    carta_provenance = case_when(
      original_row_present &
        !is.na(status_original) & status_original != "" &
        !is.na(control_original) & control_original != "" ~
        "original estimate; Carta recommendation available for this species in this region",
      TRUE ~ "gapfilled estimate; no Carta recommendation for this species in this specific region"
    ),

    Year_of_Recommendation = coalesce(
      Year_of_Recommendation,
      Year_species
    ),

    status_adjusted = if_else(
      is.na(status_adjusted) | status_adjusted == "",
      status_mode,
      status_adjusted
    ),

    control_adjusted = if_else(
      is.na(control_adjusted) | control_adjusted == "",
      control_mode,
      control_adjusted
    )
  ) %>%
  select(
    estimated_species_name,
    Year_of_Recommendation,
    OHI_region,
    status_adjusted,
    control_adjusted,
    notes,
    carta_provenance
  )
# 1359 obs., 7 vars. (151 species × 9 regions)
```

```{r Parse estimated_genus_name and species from estimated_species_name}
colnames(Carta_SpeciesRec_Lookup)

# First whitespace-separated token = genus; remainder = species.
name_parts <- strsplit(
  trimws(Carta_SpeciesRec_Lookup$estimated_species_name),
  " +"
)

Carta_SpeciesRec_Lookup <- Carta_SpeciesRec_Lookup %>%
  mutate(
    estimated_genus_name = vapply(name_parts, `[`, character(1), 1),
    species = if_else(
      lengths(name_parts) >= 2,
      vapply(
        name_parts,
        function(p) paste(p[-1], collapse = " "),
        character(1)
      ),
      NA_character_
    )
  )
# 1359 obs., 9 vars.

# Names with genus only (no species epithet after the first token):
Carta_SpeciesRec_Lookup %>%
  filter(is.na(species)) %>%
  distinct(estimated_species_name)
# 0. Good!
```

```{r Sanity checks on Carta_SpeciesRec_Lookup}
colnames(Carta_SpeciesRec_Lookup)

table(Carta_SpeciesRec_Lookup$carta_provenance)
#       gapfilled estimate; no Carta recommendation for this species in this specific region 
#                                                                               349 
# original estimate; Carta recommendation available for this species in this region 
#                                                                              1010 

table(Carta_SpeciesRec_Lookup$status_adjusted)
   # decreasing development potential   maximally harvested 
   #         99                    68                  1192 
table(Carta_SpeciesRec_Lookup$control_adjusted)
   # decrease    increase no increase 
   #       99          59        1201 

# Confirm every species appears in every region:
Carta_SpeciesRec_Lookup %>%
  count(estimated_species_name, name = "n_regions") %>%
  summary()
# All species should have n_regions == 9.
```

# STEP 5B: Create Carta_GenusRec_Lookup

Stock = genus. For each genus × OHI region, take the mode status_adjusted and control_adjusted
across all species-level recommendations assigned to that genus (original and
gapfilled). Ties in status_adjusted go to the most conservative category; ties in
control_adjusted are resolved alphabetically. Genus × region combos still missing a
recommendation after species aggregation are gapfilled from the genus-wide mode.

```{r Build genus-level input from all species-level recommendations}
colnames(Carta_SpeciesRec_Lookup)

# Roll species-level recommendations (original + gapfilled) up to genus.
Carta_SpeciesRec_for_Genus <- Carta_SpeciesRec_Lookup %>%
  mutate(
    estimated_genus_name = if_else(
      is.na(estimated_genus_name),
      "unparsed",
      estimated_genus_name
    )
  )
# 1359 obs., 9 vars.

table(Carta_SpeciesRec_for_Genus$estimated_genus_name)
```

```{r Create Carta_GenusRec_Lookup - Part 1 regional modes and genus-wide fallback}
colnames(Carta_SpeciesRec_for_Genus)
table(Carta_SpeciesRec_for_Genus$status_adjusted)
           # decreasing development potential   maximally harvested 
           #         99                    68                  1192

status_priority <- c(
  "decreasing",
  "maximally harvested",
  "development potential"
)

# Match strings from Carta_SpeciesRec_Lookup$carta_provenance (STEP 5A):
carta_provenance_ORIGINAL <-
  "original estimate; Carta recommendation available for this species in this region"
carta_provenance_GAPFILLED <-
  "gapfilled estimate; no Carta recommendation for this species in this specific region"

# Region-specific mode (genus × region) — retained where available:
Carta_GenusRec_Lookup0 <- Carta_SpeciesRec_for_Genus %>%
  group_by(estimated_genus_name, OHI_region) %>%
  summarise(
    n_species = n(),
    n_species_original = sum(carta_provenance == carta_provenance_ORIGINAL),
    n_species_gapfilled = sum(carta_provenance == carta_provenance_GAPFILLED),

    status_adjusted = {
      valid_status <- status_adjusted[!is.na(status_adjusted) & status_adjusted != ""]
      if (length(valid_status) == 0) {
        NA_character_
      } else {
        tab <- table(valid_status)
        tied <- names(tab)[tab == max(tab)]
        matched <- status_priority[status_priority %in% tied]
        if (length(matched) > 0) {
          matched[1]
        } else {
          sort(tied)[1]
        }
      }
    },

    control_adjusted = {
      valid_control <- control_adjusted[!is.na(control_adjusted) & control_adjusted != ""]
      if (length(valid_control) == 0) {
        NA_character_
      } else {
        tab <- table(valid_control)
        tied <- names(tab)[tab == max(tab)]
        sort(tied)[1]
      }
    },

    .groups = "drop"
  )

# Genus-wide mode — pooled only from regions with >=1 original species-level Carta rec.
# (Species-level gapfill fills every species x region, so pooling all regions would
# incorrectly treat gapfilled-only regions as "present".)
Carta_Genus_Global <- Carta_GenusRec_Lookup0 %>%
  filter(n_species_original > 0) %>%
  group_by(estimated_genus_name) %>%
  summarise(
    n_regions_represented = n_distinct(OHI_region),
    n_species_original_global = sum(n_species_original, na.rm = TRUE),
    n_species_gapfilled_global = sum(n_species_gapfilled, na.rm = TRUE),

    status_adjusted_global = {
      valid_status <- status_adjusted[!is.na(status_adjusted) & status_adjusted != ""]
      if (length(valid_status) == 0) {
        NA_character_
      } else {
        tab <- table(valid_status)
        tied <- names(tab)[tab == max(tab)]
        matched <- status_priority[status_priority %in% tied]
        if (length(matched) > 0) {
          matched[1]
        } else {
          sort(tied)[1]
        }
      }
    },

    control_adjusted_global = {
      valid_control <- control_adjusted[!is.na(control_adjusted) & control_adjusted != ""]
      if (length(valid_control) == 0) {
        NA_character_
      } else {
        tab <- table(valid_control)
        tied <- names(tab)[tab == max(tab)]
        sort(tied)[1]
      }
    },

    .groups = "drop"
  )
# 68 obs., 7 vars.
```

```{r Create Carta_GenusRec_Lookup - Part 2 join grid; gapfill missing regions only}
OHI_regions_all <- paste0("region_", 1:9)

Carta_GenusRec_Lookup_grid <- expand.grid(
  estimated_genus_name = unique(
    Carta_SpeciesRec_for_Genus$estimated_genus_name
  ),
  OHI_region = OHI_regions_all,
  year = SCORING_YEARS,
  stringsAsFactors = FALSE
)

Carta_GenusRec_Lookup <- Carta_GenusRec_Lookup_grid %>%
  left_join(
    Carta_GenusRec_Lookup0,
    by = c("estimated_genus_name", "OHI_region")
  ) %>%
  left_join(
    Carta_Genus_Global,
    by = "estimated_genus_name"
  ) %>%
  mutate(
    # TRUE only when this genus x region has >=1 original (non-gapfilled) species rec.
    genus_region_present = coalesce(n_species_original, 0L) > 0,

    # Regional values take precedence; genus-wide fills missing regions only (per year):
    status_adjusted = if_else(
      genus_region_present,
      status_adjusted,
      status_adjusted_global
    ),

    control_adjusted = if_else(
      genus_region_present,
      control_adjusted,
      control_adjusted_global
    ),

    n_species = if_else(
      genus_region_present,
      n_species,
      0L
    ),

    n_species_original = if_else(
      genus_region_present,
      n_species_original,
      0L
    ),

    n_species_gapfilled = if_else(
      genus_region_present,
      n_species_gapfilled,
      0L
    ),

    n_regions_represented = if_else(
      genus_region_present,
      NA_integer_,
      n_regions_represented
    ),

    carta_provenance = case_when(
      genus_region_present ~
        paste0(
          "genus aggregation; mode across ",
          n_species_original,
          " original and ",
          n_species_gapfilled,
          " gapfilled species-level recommendations in this region for year ",
          year,
          " (ties: most conservative status_adjusted and control_adjusted)"
        ),
      !is.na(status_adjusted_global) & !is.na(control_adjusted_global) ~
        paste0(
          "genus-wide gapfill; carta_status and carta_control for year ",
          year,
          " assigned from mode across ",
          n_species_original_global,
          " original and ",
          n_species_gapfilled_global,
          " gapfilled species-level recommendations pooled across ",
          n_regions_represented,
          " OHI regions with original Carta data; same values applied to every missing region"
        ),
      TRUE ~ NA_character_
    )
  ) %>%
  select(
    estimated_genus_name,
    year,
    OHI_region,
    status_adjusted,
    control_adjusted,
    n_species,
    n_species_original,
    n_species_gapfilled,
    n_regions_represented,
    genus_region_present,
    carta_provenance
  )
# genus × region × scoring year (2018–2024).

# Now 4284 obs., 11 vars.
```

```{r Sanity checks on Carta_GenusRec_Lookup}
colnames(Carta_GenusRec_Lookup)

table(Carta_GenusRec_Lookup$status_adjusted)
# decreasing development potential   maximally harvested 
           #        448                   315                  3521 

table(Carta_GenusRec_Lookup$control_adjusted)
   # decrease    increase no increase 
   #      448         301        3535  

# SANITY CHECK:
Carta_GenusRec_Lookup %>%
  filter(!is.na(status_adjusted)) %>%
  count(status_adjusted, control_adjusted, name = "n")
# ALERT: the aggregation seems to have created some weird matches...

# Species-level provenance retained in counts:
summary(Carta_GenusRec_Lookup$n_species_original)
summary(Carta_GenusRec_Lookup$n_species_gapfilled)

# Genus-region rows with at least one original species-level estimate:
Carta_GenusRec_Lookup %>%
  filter(n_species_original > 0) %>%
  nrow()
# 3339

# Genus-region rows entirely gapfilled:
Carta_GenusRec_Lookup %>%
  filter(n_species_original == 0, n_species_gapfilled == 0) %>%
  nrow()
# 945 estimates are gapfilled at the genus-region.

summary(Carta_GenusRec_Lookup$n_species)
# Most genus-level estimates were made using 1 species.

dim(Carta_GenusRec_Lookup)
```

```{r Clean up rows with status_adjusted and control_adjusted that are in conflict}
Carta_GenusRec_Lookup %>%
  filter(!is.na(status_adjusted)) %>%
  count(status_adjusted, control_adjusted, name = "n")
# ALERT: the aggregation seems to have created some weird matches...

Carta_GenusRec_Lookup <- Carta_GenusRec_Lookup %>%
  mutate(
    control_adjusted = case_when(
  
      status_adjusted == "maximally harvested" ~ "no increase",
      TRUE ~ control_adjusted
    )
  )

Carta_GenusRec_Lookup %>%
  filter(!is.na(status_adjusted)) %>%
  count(status_adjusted, control_adjusted, name = "n")
# Looks good now!

# SANITY CHECK: Let's take a closer look at genera where there is development potential but carta advises no increase...
TEST<- Carta_GenusRec_Lookup %>%
  filter(status_adjusted== "development potential" & control_adjusted == "no increase")
# 63 obs.

# SANITY CHECK: Does the length of this table seem right? Yup!
length(Carta_GenusRec_Lookup$estimated_genus_name) # 4284 rows.
unique(Carta_GenusRec_Lookup$estimated_genus_name) # 68 unique genera / stocks.
unique(Carta_GenusRec_Lookup$OHI_region) # 9 unique genera / stocks.
unique(Carta_GenusRec_Lookup$year) # 7 unique genera / stocks.
# 4284 rows = 68 unique stocks * 9 regions * 7 years worth of data (2018-2024)
```


## STEP 6: SAVE Carta_SpeciesRec_Lookup and Carta_GenusRec_Lookup
```{r Save these dataframes!}
write.csv(Carta_SpeciesRec_Lookup,
          file="/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_Carta_Scores/Carta_SpeciesRec_Lookup_Jun032026.csv",
          row.names = FALSE)
# 1359 obs., 9 vars.

write.csv(Carta_GenusRec_Lookup,
          file="/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_Carta_Scores/Carta_GenusRec_Lookup_Jun052026.csv",
          row.names = FALSE)
# Now 4284 obs., 11 vars.
```


# > SKIP TO HERE:

# STEP 6: Prepare to score annual catch by aggregating catch data

```{r Load in GOC_MarineLandings0 as needed}
GOC_MarineLandings0 <- readRDS(here(
    "/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Marine.Landings_Sector.Sorted.Georef_PATCHED_Jun032026.rds"))
# 3246583 obs., 58 vars.
```

# **DECISION POINT: Selecting which sectors to include**
```{r Select Sectors_of_Interest}
table(GOC_MarineLandings0$sector_estimate_draft)

# SANITY CHECK: Which species are included in the freshwater or marine sector?
TEST<- GOC_MarineLandings0 %>%
  filter(sector_estimate_draft== "freshwater or marine: fishery sector yet to be classified")

table(TEST$recorded_taxa_group)
# Species like Mojarra (the overwhelming majority of catch), trout, and catfish, which can move from freshwater to marine systems and are a bit ambiguous in terms of where they can be caught.


MarineFishery_Sectors<- c("marine: fishery sector yet to be classified",
                          "marine: small-scale capture",
                          "marine: large-scale capture")
  
PossibleMarineFishery_Sectors<- c("freshwater or marine: large-scale capture",
                          "freshwater or marine: small-scale capture",
                          "freshwater or marine: fishery sector yet to be classified")

GOC_MarineFisheryLandings0<- GOC_MarineLandings0 %>%
  filter(sector_estimate_draft %in% c(MarineFishery_Sectors, PossibleMarineFishery_Sectors))

# 3137044 obs., 58 var.
```

```{r Summarize landings by stock (genus), year, and region}
# Sum all catch assigned to each genus in each region and year.
GOC_MarineLandings_stock <- GOC_MarineFisheryLandings0 %>%
  mutate(
    OHI_region_est = clean_names_lowercase(OHI_region_est),
    estimated_genus_name = clean_names_lowercase(estimated_genus_name)
  ) %>%
  filter(
    !is.na(estimated_genus_name),
    estimated_genus_name != ""
  )
# 2976491 obs.

GOC_MarineStocks_summary <- GOC_MarineLandings_stock %>%
  group_by(
    year,
    OHI_region_est,
    estimated_genus_name
  ) %>%
  summarize(
    annual_Catch_kg = sum(weight_kg_corrected, na.rm = TRUE),
    n_landings_records = n(),
    .groups = "drop"
  ) %>%
  group_by(OHI_region_est, estimated_genus_name) %>%
  mutate(
    mean_Catch_2018.22_kg = mean(
      annual_Catch_kg[year %in% 2018:2022],
      na.rm = TRUE
    ),
    n_years_Catch_2018.22 = sum(
      year %in% 2018:2022 & !is.na(annual_Catch_kg)
    )
  ) %>%
  ungroup() %>%
  arrange(OHI_region_est, estimated_genus_name, year)
# 12477 obs.
# One row per stock (genus) × region × year.
```

# STEP 7: Score annual catch (2018–2024) vs 2018–2022 reference using Carta genus recommendations

Stock = genus (estimated_genus_name) × OHI region.
All landings for a genus in a region are summed, regardless of species resolution.

Scoring uses cleaned status_adjusted × control_adjusted pairs (see STEP 4), with control_adjusted-only fallback.
Catch below the reference mean is never penalized (no underfishing penalty).
Status / control_adjusted values used here: decreasing, development potential, maximally harvested;
decrease, increase (development potential only), no increase.

```{r Load Carta_GenusRec_Lookup as needed}
Carta_GenusRec_Lookup <- read.csv(
  "/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_Carta_Scores/Carta_GenusRec_Lookup_Jun052026.csv"
)
# Now 4284 obs., 11 vars.
```

```{r Merge genus-level Carta recommendations}
Carta_GenusRec_Lookup <- Carta_GenusRec_Lookup %>%
  mutate(
    OHI_region_est = clean_names_lowercase(OHI_region),
    estimated_genus_name = clean_names_lowercase(estimated_genus_name)
  )

colnames(Carta_GenusRec_Lookup)

Carta_Stock_Scores0 <- GOC_MarineStocks_summary %>%
  left_join(
    Carta_GenusRec_Lookup %>%
      select(
        year,
        estimated_genus_name,
        OHI_region_est,
        status_adjusted,
        control_adjusted,
        carta_provenance,
        n_species,
        n_species_original
      ),
    by = c("year", "estimated_genus_name", "OHI_region_est")
  ) %>%
  mutate(
    carta_recommendation_source = carta_provenance,

    catch_vs_reference_expectation = case_when(
      status_adjusted == "development potential" ~
        "development potential (score = 1; no underfishing penalty)",
      status_adjusted == "maximally harvested" & control_adjusted == "no increase" ~
        "penalized if catch exceeds 2018-2022 mean; not penalized below mean",
      status_adjusted == "decreasing" & control_adjusted %in% c("decrease", "no increase") ~
        "decreasing stock penalized at/above reference; not penalized below mean",
      control_adjusted == "no increase" ~
        "fallback: penalized if catch exceeds 2018-2022 mean",
      control_adjusted == "decrease" ~
        "fallback: decreasing-style penalty at/above reference",
      TRUE ~ NA_character_
    )
  )
# 12477 obs.
```

```{r How common is it that a stock is landed only once in a given year?}
Carta_Stock_Scores0 %>%
  count(n_landings_records) %>%
  filter(n > 1) %>%
  ggplot(aes(x = n_landings_records, y = n)) +
  geom_col() +
  theme_bw() +
  coord_cartesian(
    xlim = c(1, 1000),
    ylim = c(1, 400)
  )

# Fairly common.
```

```{r Check status_adjusted and control_adjusted combinations before scoring}
Carta_Stock_Scores0 %>%
  filter(!is.na(status_adjusted)) %>%
  count(status_adjusted, control_adjusted, name = "n")
```

# **DECISION POINT: EDIT THESE SCORING PARAMETERS AS NEEDED ───────────────────────────────────────────
```{r EDIT THESE SCORING PARAMETERS AS NEEDED}
MIN_ReferenceCatch_kg <- 100   # stocks below this mean 2018-2022 catch are not scored
ReferencePoint_Buffer <- 1.05   # 5%: catch within this band of C_ref is "functionally equal"
```

```{r Calculate Stock_Status from C_current vs C_ref}
# C_ref = mean_Catch_2018.22_kg; C_current = annual_Catch_kg for each year.
# Uses MIN_ReferenceCatch_kg and ReferencePoint_Buffer from STEP 7a.
#
# Pair-based rules (then control_adjusted fallback). Catch below C_ref is never penalized.
# increase applies only via development potential (always scores 1).

Carta_Stock_Scores <- Carta_Stock_Scores0 %>%
  mutate(
    # approx_equal = abs(annual_Catch_kg - mean_Catch_2018.22_kg) <=
    #   ReferencePoint_Buffer * mean_Catch_2018.22_kg,
    
    C_ref = mean_Catch_2018.22_kg,

    Stock_Status = case_when(

      is.na(status_adjusted) | is.na(control_adjusted) | is.na(C_ref) ~
        NA_real_,

      mean_Catch_2018.22_kg < MIN_ReferenceCatch_kg ~
        NA_real_,

      # # Scenario 1: development potential (increase control_adjusted only applies here)
      # status_adjusted == "development potential" ~
      #   1,
      # 
      # # Scenario 2: maximally harvested + no increase
      # status_adjusted == "maximally harvested" &
      #   control_adjusted == "no increase" &
      #   approx_equal ~
      #   1,
      # 
      # status_adjusted == "maximally harvested" &
      #   control_adjusted == "no increase" &
      #   annual_Catch_kg > mean_Catch_2018.22_kg ~
      #   mean_Catch_2018.22_kg / annual_Catch_kg,
      # 
      # status_adjusted == "maximally harvested" &
      #   control_adjusted == "no increase" &
      #   annual_Catch_kg < mean_Catch_2018.22_kg ~
      #   1,
      # 
      # # Scenario 3: decreasing stock (penalized at/above reference)
      # status_adjusted == "decreasing" &
      #   control_adjusted %in% c("decrease", "no increase") &
      #   approx_equal ~
      #   0.5,
      # 
      # status_adjusted == "decreasing" &
      #   control_adjusted %in% c("decrease", "no increase") &
      #   annual_Catch_kg > mean_Catch_2018.22_kg ~
      #   (mean_Catch_2018.22_kg / annual_Catch_kg) * 0.5,
      # 
      # status_adjusted == "decreasing" &
      #   control_adjusted %in% c("decrease", "no increase") &
      #   annual_Catch_kg < mean_Catch_2018.22_kg ~
      #   1,

      # Let's use status_adjusted only to dictate scores
      
       # Scenario 1: development potential
      status_adjusted == "development potential" ~ 1,

      # Scenario 2: maximally harvested
      
      status_adjusted == "maximally harvested" & annual_Catch_kg <= (C_ref * ReferencePoint_Buffer) ~ 1,
    
      status_adjusted == "maximally harvested" & annual_Catch_kg > (C_ref * ReferencePoint_Buffer) ~
        1 * (C_ref / annual_Catch_kg),
    
      # Scenario 3: stock is decreasing
      status_adjusted == "decreasing" & annual_Catch_kg <= (C_ref * ReferencePoint_Buffer) ~ 0.5,
      
      status_adjusted == "decreasing" & annual_Catch_kg > (C_ref * ReferencePoint_Buffer) ~ 0.5 * (C_ref / annual_Catch_kg),

      # unpaired combos
      TRUE ~ NA_real_
    )
  )
# 12477 obs.

# kind of looks like this is working!
```

```{r NA out stock status scores in regions with very few reference years}
colnames(Carta_Stock_Scores)

Carta_Stock_Scores <- Carta_Stock_Scores %>%
  mutate(
    Stock_Status = case_when(
      n_years_Catch_2018.22 < 4 ~ NA,
      TRUE ~ Stock_Status
    )
  )
# 12477 obs.

# SANITY CHECK Did this work as expected?
TEST<- Carta_Stock_Scores %>%
  filter(n_years_Catch_2018.22 >= 4,
         is.na(Stock_Status))
# Yes. All of these rows are missing carta recommendations, which is why Stock_Status is NA.
```

```{r Clean Carta_FIS_StockStatus_scores0}
colnames(Carta_FIS_StockStatus_scores0)

Carta_FIS_StockStatus_scores<- Carta_Stock_Scores %>%
  filter(!is.na(carta_provenance),
         year >= 2018 & year <= 2024,
         mean_Catch_2018.22_kg > MIN_ReferenceCatch_kg)
# 1480 obs., 16 vars.

table(Carta_FIS_StockStatus_scores0$carta_provenance)
table(Carta_FIS_StockStatus_scores$carta_provenance)
# Makes sense.

unique(Carta_FIS_StockStatus_scores$estimated_genus_name)
# 48 unique genera / stocks for which we have reliable Carta data.

dim(Carta_FIS_StockStatus_scores)
# 1480   16, genus (stock) × region × year, 2018–2024.
```

```{r Gapfill scores we don't have enough reference data for}
# Step 1: find the mean Stock_Status for that genus x year & then coalesce
Carta_FIS_StockStatus_scores <- Carta_FIS_StockStatus_scores %>%
  group_by(year, estimated_genus_name) %>%
  mutate(
    Stock_Status_year.genus_mean = mean(Stock_Status, na.rm = TRUE)
  ) %>%
  rename(
    Stock_Status_year.genus.region = Stock_Status
  ) %>%
  ungroup() %>%
  mutate(
    # Gapfill missing values
    Stock_Status_final = coalesce(
      Stock_Status_year.genus.region,
      Stock_Status_year.genus_mean
    )
  )

dim(Carta_FIS_StockStatus_scores)
colnames(Carta_FIS_StockStatus_scores)

# Reassign carta provenance in those specific cases:
Carta_FIS_StockStatus_scores <- Carta_FIS_StockStatus_scores %>%
  mutate(
    carta_provenance = case_when(
      is.na(Stock_Status_year.genus.region) & !is.na(Stock_Status_year.genus_mean) ~
        "genus-wide gapfill; carta Stock_Status score derived from year x genus mean",
      TRUE ~ carta_provenance
    )
  )
```


```{r SANITY CHECKS on Carta_FIS_StockStatus_scores0}
colnames(Carta_FIS_StockStatus_scores)

# Rows with catch but no Carta recommendation (Stock_Status should be NA):
Carta_FIS_StockStatus_scores %>%
  filter(is.na(Stock_Status_final)) %>%
  count(estimated_genus_name, OHI_region_est, name = "n_years") %>%
  arrange(desc(n_years)) %>%
  head(10)

# Rows excluded by minimum reference catch threshold:
Carta_FIS_StockStatus_scores %>%
  filter(
    year >= 2018 & year <= 2024,
    !is.na(status_adjusted),
    mean_Catch_2018.22_kg < MIN_ReferenceCatch_kg
  ) %>%
  nrow()

summary(Carta_FIS_StockStatus_scores$Stock_Status_final)
# Most stocks doing pretty well according to the Carta...

# Before genus gapfill:
#    Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
# 0.01168 0.69029 1.00000 0.83034 1.00000 1.00000      44 

# After genus gapfill:
#    Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
# 0.01168 0.69029 1.00000 0.83034 1.00000 1.00000      44 

# Example: maximally harvested, no increase — catch above reference should score < 1:
Carta_FIS_StockStatus_scores %>%
  filter(
    control_adjusted == "no increase",
    annual_Catch_kg > mean_Catch_2018.22_kg * 1.05
  ) %>%
  select(
    year,
    estimated_genus_name,
    OHI_region_est,
    annual_Catch_kg,
    mean_Catch_2018.22_kg,
    control_adjusted,
    Stock_Status_final
  ) %>%
  head(10)
# Seems to have worked!

# Confirm one row per stock × region × year in the scored window:
Carta_FIS_StockStatus_scores %>%
  count(estimated_genus_name, OHI_region_est, year, name = "n") %>%
  filter(n > 1)
# Yup! Zero rows that don't fit this criteria.
```

## >> FINAL STEP: SAVE the Carta_FIS_StockStatus_scores0 dataframe (uncomment when ready):
```{r Write this as a CSV, eval=FALSE}
write.csv(
  Carta_FIS_StockStatus_scores,
  file = "/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_Carta_Scores/FIS_CartaScores_Jun052026.csv",
  row.names = FALSE
)
```


```{r UNDER CONSTRUCTION - Make key wide so we can plot data availability over time}
# I want to take this dataframe and make a new one where the recommendation per taxa is spread out by years so that I can easily how recommendations change over time per taxa...

Carta_ControlKey_wide <- Carta_Recommendations0 %>%
  select(
    Year_of_Recommendation,
    estimated_genus_name,
    common_name,
    OHI_region,
    status_adjusted,
    control_adjusted
  ) %>%
  pivot_wider(
    names_from = Year_of_Recommendation,
    values_from = c(status_adjusted, control_adjusted),
    names_glue = "{.value}_{Year_of_Recommendation}"
  )

# Plot...
```