################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```

## clean_names_lowercase function
```{r Load clean_names_lowercase function}
# Takes text and standardizes it, which we will need for merging later on:

clean_names_lowercase <- function(x) {

  x <- iconv(x, from = "", to = "UTF-8", sub = "")
  x <- stringi::stri_trans_general(x, "Latin-ASCII")

  x %>%
    tolower() %>%
    trimws()
}
```

## lowercase_function function
```{r lowercase_function}
# Turns all entries lower-case
lowercase_function<- function(x){
  if(is.character(x)){
    tolower(x)
  } else if (is.factor(x)){
    tolower(as.character(x))
  } else {
    x
  }
}
```


# PICK 2: Catch vs CNP Baseline - species level ------------------------------

# STEP 1: Load in raw data, original recommendations for the Carta Nacional Pesquera
```{r Load in Carta data}
Carta_data0<- read.csv("/home/shares/ohi/OHI_GOC/goal_prep/fp/fis/v2025/int/ohi_core_team/CartaNacional_Data_AllYears_Mar2026.csv")

dim(Carta_data0)
# 2699 obs., 12 vars.
```

```{r Clean Carta_data0 as needed}
# Separate Genus from scientific_name
Carta_data <- Carta_data0 %>%
  separate(scientific_name, into = c("estimated_genus_name", "species"), sep = " ", remove = FALSE)

table(Carta_data$estimated_genus_name)
# Okay, looks good! Except for Pseudobatos (currently labeled Rhinobatos and R.)

Carta_data <- Carta_data %>%
  mutate(
    notes = if_else(
      estimated_genus_name %in% c("Rhinobatos", "R."),
      "Assigned new estimated_genus_name, Pseudobatos",
      notes
    ),
    
    estimated_genus_name = if_else(
      estimated_genus_name %in% c("Rhinobatos", "R."),
      "Pseudobatos",
      estimated_genus_name
    )
  )

# Make columns clean and lowercase for safe merging:
Carta_data<- Carta_data %>%
  mutate(
    estimated_species_name= clean_names_lowercase(scientific_name),
    estimated_genus_name= clean_names_lowercase(estimated_genus_name)
  )
```
# STEP 2: Assign major groups to the Carta_data using the GenusFamilyGroup.Key
```{r Load in GenusFamilyGroup.Key and clean as needed}
GenusFamilyGroup.Key0<- read.csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Keys/CONAPESCA_Keys/Taxonomic_Keys/GenusFamilyGroupKey_COMPLETE_Apr042026.csv")
# 193 obs., 7 vars.
```

```{r Merge family details to Carta_data}
GenusFamilyGroup.Key<- GenusFamilyGroup.Key0 %>% 
  select(estimated_genus_name, estimated_taxa_group, Major_Functional_Group) %>%
  distinct()
# 193 obs., 3 vars.

# Make columns clean and lowercase for safe merging:
GenusFamilyGroup.Key<- GenusFamilyGroup.Key %>%
  mutate(
    estimated_genus_name= clean_names_lowercase(estimated_genus_name)
  )

# Merge family details to Carta_data:
Carta_data <- Carta_data %>%
  left_join(
    GenusFamilyGroup.Key,
    by = c("estimated_genus_name")
  )
# 2699 obs., 17 vars. (Good, as expected!)
```


# STEP 3: Create the Carta_Recommendations_clean based on the data

```{r Make a Carta_Recommendations0}
colnames(Carta_data)

Carta_Recommendations0<- Carta_data %>% 
  select(Year_of_Recommendation= carta_year, Major_Functional_Group, 
         estimated_species_name, estimated_genus_name, common_name, 
         OHI_region, 
         status, control, notes)
# 2699 obs., 9 vars.
```

```{r Create Carta_Recommendations - Part 1}
colnames(Carta_Recommendations0)

# Filter this down to unique number of species (disregarding common_name)
Carta_Recommendations<- Carta_Recommendations0 %>% 
  select(Year_of_Recommendation, Major_Functional_Group, 
         estimated_species_name, estimated_genus_name, 
         OHI_region, 
         status, control, notes) %>%
  distinct()
# 2692 obs., 8 vars.
# Suggests there were 7 duplicates.

# SANITY CHECK: Why is this key shorter from Carta_Recommendations0?
colnames(Carta_Recommendations)

common_cols <- intersect(names(Carta_Recommendations0),
                         names(Carta_Recommendations))

TEST<- anti_join(
  Carta_Recommendations0,
  Carta_Recommendations,
  by = names(Carta_Recommendations)
)
# Multiple recommendations for distinct species under different common names: common_name
```

```{r Create Carta_Recommendations - Part 2}
colnames(Carta_Recommendations)

table(Carta_Recommendations$status)
  #                                decreasing         deteriorating development potential 
  #                1134                     4                    44                    26 
  # maximally harvested               unclear 
  #                 936                   112 
table(Carta_Recommendations$control)
       #         decrease    increase  increasing no increase 
       # 1134          44          40           4         918


# Generate outputs regarding status and control at the species level:
Carta_Recommendations <- Carta_Recommendations %>%
  
  group_by(
    estimated_species_name,
    OHI_region,
    Year_of_Recommendation
  ) %>%
  
  summarise(

# STATUS:
    
    status = {

      valid_status <- status[!is.na(status)]

      if (length(valid_status) == 0) {
        
        NA_character_

      } else if ("deteriorating" %in% valid_status) {

        "deteriorating"

      } else if ("improving" %in% valid_status) {

        "improving"

      } else {

        valid_status[1]
      }
    },


# CONTROL:

    control = {

      valid_control <- control[!is.na(control)]

      if (length(valid_control) == 0) {

        NA_character_

      } else {

        valid_control[1]
      }
    },


# NOTES:

    notes = {

      valid_notes <- notes[
        !is.na(notes) & notes != ""
      ]

      if (length(valid_notes) == 0) {

        NA_character_

      } else {

        valid_notes[1]
      }
    },

    .groups = "drop"
  )
# 2692 obs.,  6 vars.
```

```{r Remove any rows where the control and recommendation are both NA or "unknown"}
# Remove any rows where the control and recommendation are both NA or "unknown"
table(Carta_Recommendations$status) # 1134 NAs
table(Carta_Recommendations$control) # 1134 NAs

Carta_Recommendations <- Carta_Recommendations %>%
  filter(
    !(
      (is.na(status) | status == "unclear" | status== "") &
      (is.na(control) | control == "unclear" | control == "")
    )
  )
# 1010 obs., 6 vars.
```

```{r Clean up rows with status and controls that are in conflict}
table(Carta_Recommendations$status)
           # decreasing         deteriorating development potential   maximally harvested 
           #          4                    44                    26                   936 
table(Carta_Recommendations$control)
  # decrease    increase  increasing no increase 
  #        44          40           4         918 

# Scenarios where stock status is "decreasing" or "maximally harvested" but control is "increase" or "increasing"
TEST<- Carta_Recommendations %>%
  filter(status %in% c("decreasing", "deteriorating", "maximally harvested") &
           control %in% c("increase", "increasing"))
# 22 obs., 6 vars.

Carta_Recommendations_clean0 <- Carta_Recommendations %>%
  mutate(
    status_adjusted = case_when(
      status %in% c("decreasing", "deteriorating") ~ "decreasing",
      TRUE ~ status
    ),

    control_adjusted = case_when(
      status %in% c("decreasing", "deteriorating") &
        (is.na(control) | control %in% c("increase", "increasing")) ~ "decrease",

      status == "maximally harvested" &
        (is.na(control) | control %in% c("increase", "increasing")) ~ "no increase",
      
      status == "development potential" &
        (is.na(control)) ~ "increase",

      TRUE ~ control
    ),
    
    control_adjusted= case_when(
      control_adjusted %in% c("increase", "increasing") ~ "increase",
      TRUE ~ control_adjusted
    )
  )

# SANITY CHECK: were the entries reassigned as expected? Yes, I believe so.
table(Carta_Recommendations_clean0$status_adjusted)
   # decreasing development potential   maximally harvested 
   #         48                    26                   936 
table(Carta_Recommendations_clean0$control_adjusted)
   # decrease    increase no increase 
   #       48          24         938 
```


## STEP 4: SAVE the Carta_Recommendations_clean
```{r SAVE the Carta_Recommendations_clean as necessary}
write.csv(Carta_Recommendations_clean0,
          file="/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_Carta_Scores/Carta_RawRecommendations_Cleaned_Jun032026.csv",
          row.names = FALSE)
# 1010 obs., 8 vars.
```


# STEP 5: Create Carta_SpeciesRec_Lookup
```{r Load in Carta_Recommendations_clean as needed}
Carta_Recommendations_clean0<- read.csv("/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_Carta_Scores//Carta_Recommendations_Cleaned_Jun012026.csv")
# 1010 obs., 8 vars.
```

```{r Clean as needed}
colnames(Carta_Recommendations_clean0)

Carta_Recommendations_clean<- Carta_Recommendations_clean0 %>%
  select(
    estimated_species_name, OHI_region, Year_of_Recommendation,
    status= status_adjusted, control= control_adjusted, notes
  )
```

```{r Create Carta_SpeciesRec_Lookup - Part 1}
colnames(Carta_Recommendations_clean)

# Input is already one row per species × region (no duplicate keys).
Carta_Recommendations_clean <- Carta_Recommendations_clean %>%
  distinct()
# 1010 obs., 6 vars.

# How many species-region combinations are missing from the input key?
OHI_regions_all <- paste0("region_", 1:9)

Carta_Recommendations_grid <- expand.grid(
  estimated_species_name = unique(Carta_Recommendations_clean$estimated_species_name),
  OHI_region = OHI_regions_all,
  stringsAsFactors = FALSE
)
# 1359 obs., 2 obs. (1359 = 9 x 151)

Carta_Recommendations_missing <- anti_join(
  Carta_Recommendations_grid,
  Carta_Recommendations_clean %>%
    select(estimated_species_name, OHI_region),
  by = c("estimated_species_name", "OHI_region")
)
# 349 missing species × region combinations to gapfill.
```

```{r Create Carta_SpeciesRec_Lookup - Part 2 species-level modes}
colnames(Carta_Recommendations_clean)

# For each species, take the mode of status and control across regions where values exist.
species_modes <- Carta_Recommendations_clean %>%
  group_by(estimated_species_name) %>%
  summarise(
    Year_of_Recommendation = Year_of_Recommendation[
      !is.na(Year_of_Recommendation)
    ][1],

    status_mode = {
      valid_status <- status[!is.na(status) & status != ""]
      if (length(valid_status) == 0) {
        NA_character_
      } else {
        names(sort(table(valid_status), decreasing = TRUE))[1]
      }
    },

    control_mode = {
      valid_control <- control[!is.na(control) & control != ""]
      if (length(valid_control) == 0) {
        NA_character_
      } else {
        names(sort(table(valid_control), decreasing = TRUE))[1]
      }
    },

    .groups = "drop"
  )
# 151 obs., 4 vars.
```

```{r Create Carta_SpeciesRec_Lookup - Part 3 gapfill to full species × region grid}
colnames(Carta_Recommendations_clean)

Carta_SpeciesRec_Lookup <- Carta_Recommendations_grid %>%
  left_join(
    Carta_Recommendations_clean,
    by = c("estimated_species_name", "OHI_region")
  ) %>%
  left_join(
    species_modes %>%
      select(
        estimated_species_name,
        Year_species = Year_of_Recommendation,
        status_mode,
        control_mode
      ),
    by = "estimated_species_name"
  ) %>%
  mutate(
    status_original = status,
    control_original = control,
    original_row_present = !is.na(Year_of_Recommendation),

    carta_source = case_when(
      original_row_present &
        !is.na(status_original) & status_original != "" &
        !is.na(control_original) & control_original != "" ~
        "original estimate; Carta recommendation available for this species in this region",
      TRUE ~ "gapfilled estimate; no Carta recommendation for this species in this specific region"
    ),

    Year_of_Recommendation = coalesce(
      Year_of_Recommendation,
      Year_species
    ),

    status = if_else(
      is.na(status) | status == "",
      status_mode,
      status
    ),

    control = if_else(
      is.na(control) | control == "",
      control_mode,
      control
    )
  ) %>%
  select(
    estimated_species_name,
    Year_of_Recommendation,
    OHI_region,
    status,
    control,
    notes,
    carta_source
  )
# 1359 obs., 7 vars. (151 species × 9 regions)
```

```{r Parse estimated_genus_name and species from estimated_species_name}
colnames(Carta_SpeciesRec_Lookup)

# First whitespace-separated token = genus; remainder = species.
name_parts <- strsplit(
  trimws(Carta_SpeciesRec_Lookup$estimated_species_name),
  " +"
)

Carta_SpeciesRec_Lookup <- Carta_SpeciesRec_Lookup %>%
  mutate(
    estimated_genus_name = vapply(name_parts, `[`, character(1), 1),
    species = if_else(
      lengths(name_parts) >= 2,
      vapply(
        name_parts,
        function(p) paste(p[-1], collapse = " "),
        character(1)
      ),
      NA_character_
    )
  )
# 1359 obs., 9 vars.

# Names with genus only (no species epithet after the first token):
Carta_SpeciesRec_Lookup %>%
  filter(is.na(species)) %>%
  distinct(estimated_species_name)
```

```{r Sanity checks on Carta_SpeciesRec_Lookup}
colnames(Carta_SpeciesRec_Lookup)

table(Carta_SpeciesRec_Lookup$carta_source)
#       gapfilled estimate; no Carta recommendation for this species in this specific region 
#                                                                               349 
# original estimate; Carta recommendation available for this species in this region 
#                                                                              1010 

table(Carta_SpeciesRec_Lookup$status)
   # decreasing development potential   maximally harvested 
   #         99                    68                  1192 
table(Carta_SpeciesRec_Lookup$control)
   # decrease    increase no increase 
   #       99          59        1201 

# Confirm every species appears in every region:
Carta_SpeciesRec_Lookup %>%
  count(estimated_species_name, name = "n_regions") %>%
  summary()
# All species should have n_regions == 9.
```

# STEP 5B: Create Carta_GenusRec_Lookup
```{r Build genus-level input from original Carta estimates only}
colnames(Carta_SpeciesRec_Lookup)

# Genus key (Carta_SpeciesRec_for_Genus) uses only species-level original estimates (won't use estimates that are gapfilled).
Carta_SpeciesRec_for_Genus <- Carta_SpeciesRec_Lookup %>%
  filter(carta_source == "original estimate; Carta recommendation available for this species in this region") %>%
  mutate(
    estimated_genus_name = if_else(
      is.na(estimated_genus_name),
      "unparsed",
      estimated_genus_name
    )
  )
# 1010 obs., 9 vars.

table(Carta_SpeciesRec_for_Genus$estimated_genus_name)
```

```{r Create Carta_GenusRec_Lookup - Part 1 genus x region modes}
colnames(Carta_SpeciesRec_for_Genus)
table(Carta_SpeciesRec_for_Genus$status)

status_priority <- c(
  "decreasing",
  "maximally harvested",
  "development potential"
)

Carta_GenusRec_Lookup0 <- Carta_SpeciesRec_for_Genus %>%
  group_by(estimated_genus_name, OHI_region) %>%
  summarise(
    Year_of_Recommendation = Year_of_Recommendation[
      !is.na(Year_of_Recommendation)
    ][1],

    n_species = n(),

    status = {
      valid_status <- status[!is.na(status) & status != ""]
      if (length(valid_status) == 0) {
        NA_character_
      } else {
        tab <- table(valid_status)
        tied <- names(tab)[tab == max(tab)]
        matched <- status_priority[status_priority %in% tied]
        if (length(matched) > 0) {
          matched[1]
        } else {
          sort(tied)[1]
        }
      }
    },

    control = {
      valid_control <- control[!is.na(control) & control != ""]
      if (length(valid_control) == 0) {
        NA_character_
      } else {
        tab <- table(valid_control)
        tied <- names(tab)[tab == max(tab)]
        sort(tied)[1]
      }
    },

    .groups = "drop"
  )
# 477 obs., 6 vars.

# One row per genus × region with at least one original species estimate.
```

```{r Create Carta_GenusRec_Lookup - Part 2 gapfill full genus x region grid}
colnames(Carta_GenusRec_Lookup0)

OHI_regions_all <- paste0("region_", 1:9)

Carta_GenusRec_Lookup_grid <- expand.grid(
  estimated_genus_name = unique(
    Carta_SpeciesRec_for_Genus$estimated_genus_name
  ),
  OHI_region = OHI_regions_all,
  stringsAsFactors = FALSE
)
# 612 obs.

genus_modes <- Carta_GenusRec_Lookup0 %>%
  group_by(estimated_genus_name) %>%
  summarise(
    Year_of_Recommendation = Year_of_Recommendation[
      !is.na(Year_of_Recommendation)
    ][1],

    status_mode = {
      valid_status <- status[!is.na(status) & status != ""]
      if (length(valid_status) == 0) {
        NA_character_
      } else {
        tab <- table(valid_status)
        tied <- names(tab)[tab == max(tab)]
        matched <- status_priority[status_priority %in% tied]
        if (length(matched) > 0) {
          matched[1]
        } else {
          sort(tied)[1]
        }
      }
    },

    control_mode = {
      valid_control <- control[!is.na(control) & control != ""]
      if (length(valid_control) == 0) {
        NA_character_
      } else {
        tab <- table(valid_control)
        tied <- names(tab)[tab == max(tab)]
        sort(tied)[1]
      }
    },

    .groups = "drop"
  )

Carta_GenusRec_Lookup <- Carta_GenusRec_Lookup_grid %>%
  left_join(
    Carta_GenusRec_Lookup0,
    by = c("estimated_genus_name", "OHI_region")
  ) %>%
  left_join(
    genus_modes %>%
      select(
        estimated_genus_name,
        Year_genus = Year_of_Recommendation,
        status_mode,
        control_mode
      ),
    by = "estimated_genus_name"
  ) %>%
  mutate(
    n_species_original = n_species,
    genus_region_present = !is.na(Year_of_Recommendation),

    carta_provenance = case_when(
      genus_region_present ~
        "genus aggregation; took the mode across all Carta recommendations supplied for this genus in this region (in the case of ties, I chose the most conservative status and control estimate)",
      TRUE ~
        "genus gapfill; data missing for this genus in this region, took the mode across all Carta recommendations supplied for this genus elsewhere"
    ),

    n_species = if_else(
      is.na(n_species),
      0,
      n_species
    ),

    Year_of_Recommendation = coalesce(
      Year_of_Recommendation,
      Year_genus
    ),

    status = if_else(
      is.na(status) | status == "",
      status_mode,
      status
    ),

    control = if_else(
      is.na(control) | control == "",
      control_mode,
      control
    )
  ) %>%
  select(
    estimated_genus_name,
    Year_of_Recommendation,
    OHI_region,
    status,
    control,
    n_species,
    n_species_original,
    carta_provenance
  )
# Now 612 obs.
```

```{r Sanity checks on Carta_GenusRec_Lookup}
colnames(Carta_GenusRec_Lookup)

table(Carta_GenusRec_Lookup$status)
  # decreasing development potential   maximally harvested 
  #         64                    45                   503 
table(Carta_GenusRec_Lookup$control)
   # decrease    increase no increase 
   #       64          39         509 

table(Carta_GenusRec_Lookup$carta_provenance)
# genus aggregation, n= 477
# genus gap-fill, n= 135

# Genus-region combinations built from original species estimates in that region (not gapfilled):
Carta_GenusRec_Lookup %>%
  filter(!is.na(n_species_original)) %>%
  nrow()

# Genus-region combinations gapfilled across regions at the genus level:
Carta_GenusRec_Lookup %>%
  filter(is.na(n_species_original)) %>%
  nrow()

summary(Carta_GenusRec_Lookup$n_species)
# Most genus-level estimates were made using 1 species.

dim(Carta_GenusRec_Lookup)
```


# STEP 5C: Create Carta_FamilyRec_Lookup

```{r Build group-level input from original Carta estimates only}
colnames(Carta_SpeciesRec_Lookup)

# Genus key (Carta_SpeciesRec_for_Genus) uses only species-level original estimates (won't use estimates that are gapfilled).
Carta_SpeciesRec_for_Group <- Carta_SpeciesRec_Lookup %>%
  filter(carta_source == "original estimate; Carta recommendation available for this species in this region") %>%
  mutate(
    estimated_genus_name = if_else(
      is.na(estimated_genus_name),
      "unparsed",
      estimated_genus_name
    )
  )
# 1010 obs., 9 vars.

table(Carta_SpeciesRec_for_Genus$estimated_genus_name)
```

```{r Create Carta_GenusRec_Lookup - Part 1 genus x region modes}
colnames(Carta_SpeciesRec_for_Genus)
table(Carta_SpeciesRec_for_Genus$status)

status_priority <- c(
  "decreasing",
  "maximally harvested",
  "development potential"
)

Carta_GenusRec_Lookup0 <- Carta_SpeciesRec_for_Genus %>%
  group_by(estimated_genus_name, OHI_region) %>%
  summarise(
    Year_of_Recommendation = Year_of_Recommendation[
      !is.na(Year_of_Recommendation)
    ][1],

    n_species = n(),

    status = {
      valid_status <- status[!is.na(status) & status != ""]
      if (length(valid_status) == 0) {
        NA_character_
      } else {
        tab <- table(valid_status)
        tied <- names(tab)[tab == max(tab)]
        matched <- status_priority[status_priority %in% tied]
        if (length(matched) > 0) {
          matched[1]
        } else {
          sort(tied)[1]
        }
      }
    },

    control = {
      valid_control <- control[!is.na(control) & control != ""]
      if (length(valid_control) == 0) {
        NA_character_
      } else {
        tab <- table(valid_control)
        tied <- names(tab)[tab == max(tab)]
        sort(tied)[1]
      }
    },

    .groups = "drop"
  )
# 477 obs., 6 vars.

# One row per genus × region with at least one original species estimate.
```

```{r Create Carta_GenusRec_Lookup - Part 2 gapfill full genus x region grid}
colnames(Carta_GenusRec_Lookup0)

OHI_regions_all <- paste0("region_", 1:9)

Carta_GenusRec_Lookup_grid <- expand.grid(
  estimated_genus_name = unique(
    Carta_SpeciesRec_for_Genus$estimated_genus_name
  ),
  OHI_region = OHI_regions_all,
  stringsAsFactors = FALSE
)
# 612 obs.

genus_modes <- Carta_GenusRec_Lookup0 %>%
  group_by(estimated_genus_name) %>%
  summarise(
    Year_of_Recommendation = Year_of_Recommendation[
      !is.na(Year_of_Recommendation)
    ][1],

    status_mode = {
      valid_status <- status[!is.na(status) & status != ""]
      if (length(valid_status) == 0) {
        NA_character_
      } else {
        tab <- table(valid_status)
        tied <- names(tab)[tab == max(tab)]
        matched <- status_priority[status_priority %in% tied]
        if (length(matched) > 0) {
          matched[1]
        } else {
          sort(tied)[1]
        }
      }
    },

    control_mode = {
      valid_control <- control[!is.na(control) & control != ""]
      if (length(valid_control) == 0) {
        NA_character_
      } else {
        tab <- table(valid_control)
        tied <- names(tab)[tab == max(tab)]
        sort(tied)[1]
      }
    },

    .groups = "drop"
  )

Carta_GenusRec_Lookup <- Carta_GenusRec_Lookup_grid %>%
  left_join(
    Carta_GenusRec_Lookup0,
    by = c("estimated_genus_name", "OHI_region")
  ) %>%
  left_join(
    genus_modes %>%
      select(
        estimated_genus_name,
        Year_genus = Year_of_Recommendation,
        status_mode,
        control_mode
      ),
    by = "estimated_genus_name"
  ) %>%
  mutate(
    n_species_original = n_species,
    genus_region_present = !is.na(Year_of_Recommendation),

    carta_provenance = case_when(
      genus_region_present ~
        "genus aggregation; took the mode across all Carta recommendations supplied for this genus in this region (in the case of ties, I chose the most conservative status and control estimate)",
      TRUE ~
        "genus gapfill; data missing for this genus in this region, took the mode across all Carta recommendations supplied for this genus elsewhere"
    ),

    n_species = if_else(
      is.na(n_species),
      0,
      n_species
    ),

    Year_of_Recommendation = coalesce(
      Year_of_Recommendation,
      Year_genus
    ),

    status = if_else(
      is.na(status) | status == "",
      status_mode,
      status
    ),

    control = if_else(
      is.na(control) | control == "",
      control_mode,
      control
    )
  ) %>%
  select(
    estimated_genus_name,
    Year_of_Recommendation,
    OHI_region,
    status,
    control,
    n_species,
    n_species_original,
    carta_provenance
  )
# Now 612 obs.
```


## STEP 6: SAVE Carta_SpeciesRec_Lookup and Carta_GenusRec_Lookup
```{r Save these dataframes!}
write.csv(Carta_SpeciesRec_Lookup,
          file="/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_Carta_Scores/Carta_SpeciesRec_Lookup_Jun032026.csv",
          row.names = FALSE)
# 1359 obs., 9 vars.

write.csv(Carta_GenusRec_Lookup,
          file="/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_Carta_Scores/Carta_GenusRec_Lookup_Jun032026.csv",
          row.names = FALSE)
# 612 obs., 8 vars.
```


# STEP 7: Scoring based on the Carta: compare landings over time (GOC_MarineLandings) to recommendations in the Carta_Recommendations Lookups

```{r Load in Carta_SpeciesRec_Lookup and Carta_GenusRec_Lookup as needed}
Carta_SpeciesRec_Lookup<- read.csv("/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_Carta_Scores/Carta_SpeciesRec_Lookup_Jun032026.csv")
# 1359 obs., 9 vars.

Carta_GenusRec_Lookup<- read.csv("/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_Carta_Scores/Carta_GenusRec_Lookup_Jun032026.csv")
# 612 obs., 8 vars.
```

```{r Load in GOC_MarineLandings_carta as needed}
GOC_MarineLandings0<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Marine.Landings_Sector.Sorted.Georef_PATCHED_Jun032026.rds"))
# 3246583 obs., 58 vars.
```

```{r STEP 7b - Summarize landings by taxon, year, and region}
GOC_MarineLandings <- GOC_MarineLandings0 %>%
  mutate(
    OHI_region_est = clean_names_lowercase(OHI_region_est),
    estimated_species_name = clean_names_lowercase(estimated_species_name),
    estimated_genus_name = clean_names_lowercase(estimated_genus_name),

    taxon_key = if_else(
      is.na(estimated_species_name) | estimated_species_name == "",
      estimated_genus_name,
      estimated_species_name
    ),

    taxon_level = if_else(
      is.na(estimated_species_name) | estimated_species_name == "",
      "genus",
      "species"
    )
  )
# 3246583 obs., 59 vars.

# Summarize landings:
GOC_MarineLandings_summary <- GOC_MarineLandings %>%
  group_by(
    year,
    OHI_region_est,
    taxon_key,
    taxon_level,
    estimated_genus_name, 
    estimated_taxa_group
  ) %>%
  summarize(
    annual_Catch_kg = sum(weight_kg_corrected, na.rm = TRUE),
    n_landings_records = n(),
    .groups = "drop"
  ) %>%
  group_by(OHI_region_est, taxon_key, taxon_level) %>%
  mutate(
    mean_Catch_2018.22_kg = mean(
      annual_Catch_kg[year %in% 2018:2022],
      na.rm = TRUE
    ),
    n.years_Catch_2018.22 = sum(year %in% 2018:2022)
  ) %>%
  ungroup() %>%
  arrange(taxon_level, OHI_region_est, taxon_key, year)
# 16858 obs., 9 vars.
```

```{r STEP 7c - Merge species lookup, then genus lookup fallback}
# Merge species lookup:
Carta_SpeciesRec_Lookup <- Carta_SpeciesRec_Lookup %>%
  mutate(
    OHI_region_est = clean_names_lowercase(OHI_region),
    estimated_species_name = clean_names_lowercase(estimated_species_name),
    estimated_genus_name = clean_names_lowercase(estimated_genus_name)
  )

Carta_Scores0 <- GOC_MarineLandings_summary %>%
  left_join(
    Carta_SpeciesRec_Lookup %>%
      select(
        estimated_species_name,
        OHI_region_est,
        status,
        control,
        carta_source,
        Year_of_Recommendation,
        notes
      ),
    by = c(
      "taxon_key" = "estimated_species_name",
      "OHI_region_est"
    )
  )
# 16858 obs., 14 vars.

# Then merge genus lookup as a fallback:
Carta_GenusRec_Lookup <- Carta_GenusRec_Lookup %>%
  mutate(
    OHI_region_est = clean_names_lowercase(OHI_region),
    estimated_genus_name = clean_names_lowercase(estimated_genus_name)
  )

Carta_Scores0 <- Carta_Scores0 %>%
  left_join(
    Carta_GenusRec_Lookup %>%
      select(
        estimated_genus_name,
        OHI_region_est,
        status_genus = status,
        control_genus = control,
        carta_provenance
      ),
    by = c("estimated_genus_name", "OHI_region_est")
  )


  mutate(
    carta_recommendation_source = case_when(
      taxon_level == "species" & !is.na(status) ~
        carta_source,
      taxon_level == "genus" & !is.na(status_genus) ~
        carta_provenance,
      taxon_level == "species" & !is.na(status_genus) ~
        paste0("genus fallback; ", carta_provenance),
      TRUE ~ NA_character_
    ),
    status = coalesce(status, status_genus),
    control = coalesce(control, control_genus),

    catch_vs_reference_expectation = case_when(
      status == "development potential" ~ "development potential (score = 1)",
      control == "increase" ~ "catch may be at or above 2018-2022 mean",
      control == "no increase" ~ "catch should not exceed 2018-2022 mean",
      control == "decrease" ~ "catch should be below 2018-2022 mean",
      TRUE ~ NA_character_
    )
  ) %>%
  select(-status_genus, -control_genus, -carta_provenance)
# 16858 obs., 16 vars.
```

```{r How common is it that something is landed only once?}
Carta_Scores0 %>%
  count(n_landings_records) %>%
  filter(n > 1) %>%
  ggplot(aes(x = n_landings_records, y = n)) +
  geom_col() +
  theme_bw() +
  coord_cartesian(xlim = c(1, 1000),
                  ylim = c(1, 400))
```


```{r STEP 7d - Check status and control combinations before scoring}
Carta_Scores0 %>%
  filter(!is.na(status)) %>%
  count(status, control, name = "n")

# decreasing	decrease	374		
# development potential	increase	125		
# development potential	no increase	108		
# maximally harvested	increase	49		
# maximally harvested	no increase	6348
```

```{r STEP 7e - Calculate Stock_Status from C_current vs C_ref}
# C_ref = mean_Catch_2018.22_kg; C_current = annual_Catch_kg for each year.
# 5% tolerance: within band counts as functionally equal to reference.
tolerance <- 0.05

Carta_Scores <- Carta_Scores0 %>%
  mutate(
    approx_equal = abs(annual_Catch_kg - mean_Catch_2018.22_kg) <=
      tolerance * mean_Catch_2018.22_kg,

    Stock_Status = case_when(

      is.na(status) | is.na(control) | is.na(mean_Catch_2018.22_kg) ~
        NA_real_,

      status == "development potential" ~
        1,

      control == "increase" & approx_equal ~
        1,

      control == "increase" & annual_Catch_kg >= mean_Catch_2018.22_kg ~
        1,

      control == "increase" & annual_Catch_kg < mean_Catch_2018.22_kg ~
        annual_Catch_kg / mean_Catch_2018.22_kg,

      control == "no increase" & approx_equal ~
        1,

      control == "no increase" & annual_Catch_kg > mean_Catch_2018.22_kg ~
        mean_Catch_2018.22_kg / annual_Catch_kg,

      control == "no increase" & annual_Catch_kg < mean_Catch_2018.22_kg ~
        1,

      control == "decrease" & approx_equal ~
        0.5,

      control == "decrease" & annual_Catch_kg > mean_Catch_2018.22_kg ~
        (mean_Catch_2018.22_kg / annual_Catch_kg) * 0.5,

      control == "decrease" & annual_Catch_kg < mean_Catch_2018.22_kg ~
        0.5 + 0.5 * (
          (mean_Catch_2018.22_kg - annual_Catch_kg) /
          mean_Catch_2018.22_kg
        ),

      TRUE ~ NA_real_
    )
  ) %>%
  select(-approx_equal)
# 16858 obs., 17 vars.
```

```{r STEP 7f - Filter empty stocks and stocks for which the number of reference years is less than 4}
Carta_FIS_scores <- Carta_Scores %>%
  filter(year >= 2018 & year <= 2024) %>%
  filter(!is.na(Stock_Status)) %>%
  filter(n.years_Catch_2018.22 >= 4)

dim(Carta_FIS_scores)
# 1873 obs., 17 vars.
```

```{r STEP 7g - Sanity checks on Carta_FIS_scores}
colnames(Carta_FIS_scores)

table(Carta_FIS_scores$catch_vs_reference_expectation, useNA = "ifany")
# catch may be at or above 2018-2022 mean    catch should be below 2018-2022 mean  catch should not exceed 2018-2022 mean 
#                                      12                                      70                                    1733 
#       development potential (score = 1) 
#                                      58

Carta_FIS_scores %>%
  count(taxon_level, carta_recommendation_source, name = "n")

summary(Carta_FIS_scores$Stock_Status)
#    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
# 0.00247 0.72777 1.00000 0.84179 1.00000 1.00000 

# Example: maximally harvested, no increase — catch above reference should score < 1:
Carta_FIS_scores %>%
  filter(
    control == "no increase",
    annual_Catch_kg > mean_Catch_2018.22_kg * 1.05
  ) %>%
  select(
    year,
    taxon_key,
    OHI_region_est,
    annual_Catch_kg,
    mean_Catch_2018.22_kg,
    control,
    Stock_Status
  ) %>%
  head(10)
```

## >> FINAL STEP: SAVE the Carta_FIS_scores dataframe:
```{r Write this as a CSV}
write.csv(Carta_FIS_scores,
          file= "/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_StockStatus_Scores/FIS_CartaScores_Jun012026.csv", 
          row.names = FALSE)
# 1873 obs., 17 vars.
```


```{r UNDER CONSTRUCTION - Make key wide so we can plot these data}
# I want to take this dataframe and make a new one where the recommendation per taxa is spread out by years so that I can easily how recommendations change over time per taxa...

Carta_ControlKey_wide <- Carta_Recommendations0 %>%
  select(
    Year_of_Recommendation,
    estimated_genus_name,
    common_name,
    OHI_region,
    status,
    control
  ) %>%
  pivot_wider(
    names_from = Year_of_Recommendation,
    values_from = c(status, control),
    names_glue = "{.value}_{Year_of_Recommendation}"
  )

# Plot...
```
