################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(stringr)
library(stringi)
library(purrr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(fuzzyjoin)
# Visualization
library(ggplot2)
library(plotly)
library(htmlwidgets)


setwd("/home/ferrer/ohi_goc_prep/")
```


# PART 1: Filter landings data and create skeleton for species key ------------------------------
First we need to correctly clean and filter landings data so that we can append species info and extract species info for a key.

## Filtering out key words from landing receipts:
```{r Load in GoC landings data}
Landings_GoC<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fis/v2025/int/juan_carlos/GoC_Landings_2000.present_EMF_created.rds"))
# n = 3716415 obs.
```

```{r Is aleta de aleta an issue? Short answer, no.}
Landings_GoC %>%
  rename(recorded_taxa_name = species_name) %>%
  mutate(
    aleta_count = stringi::stri_count_fixed(recorded_taxa_name, "ALETA")
  ) %>%
  count(aleta_count)

# No, double aleta does not seem to be an issue...
```

```{r Filter out keywords from recorded_taxa_name to get recorded_taxa_name_clean}
# 1. Keyword list (excluding ALETA, handled separately) -----------------
keyword_list <- c(
  "S.C.", "S/C", "S",
  "ENTERO", "ENTERA", "IND", "ENT",
  "FRESCO", "FRESCA", "FCA", "FCO", "FC",
  "ECO",
  "CONG", "CONGELADO",
  "VIVO", "VIVA", "VIVAS", "VIVITA",
  "DE CULTIVO", "DE CULTIVO VIVO", "ABASTO RANCHO", "RANCHO",
  "HUEVA DE",
  "FILETE", "FILETE DE", "ECA", "FILETE DE ECO", "DESHUESADA", "CARNE", "CARNE DE",
  "DESCABEZADO", "DES", "DSC", "DESC", "DESV", "DESVICERADO",
  "CABEZA DE", "BUCHE", "BUCHE DE", "TRIPA DE", "COLAS DE", "MANOS DE", "MANOS",
  "ANCAS DE", "ANCAS", "ANCA",
  "PIEL DE ECA", "PIEL DE SECA", "PIEL DE", "PIEL",
  "SECO DE", "SECA", "SECA DE", "SECO", "SECO DE",
  "COCINERO", "COCIDO", "COCIDA", "COCIDAS", "ASADO",
  # Remove ALETA from main keywords; handled separately
  #"ALETA", "ALETA ECA DE", "ALETA FCA DE",
  "DE ALTA MAR",
  "CONCHA DE", "CONCHA",
  "POLVO DE",
  "CARNADA", "P/CARNADA",
  "DE AGUA DULCE",
  "MANTO", "MANTO DE", "PULPA DE", "PULPA FRESCA",
  "PULPO DE", "TENTACULOS DE", "TUBO DE", "TUBO",
  "NO APTO PARA EMPAQUE"
)

# Longest-first ordering
keyword_list <- keyword_list[order(nchar(keyword_list), decreasing = TRUE)]

# Escape regex meta-characters
keyword_list_escaped <- stringi::stri_replace_all_regex(
  keyword_list, "([.|(){}+*?^$\\[\\]\\\\])", "\\\\$1"
)

# Final regex
keyword_pattern <- paste0(
  "(?<![A-Za-z0-9])(",
  paste(keyword_list_escaped, collapse = "|"),
  ")(?![A-Za-z0-9])"
)


# 2. Special ALETA phrases (handled separately from other key words) -----------------

special_aleta_adj <- c("BLANCA", "AMARILLA", "CORTA", "AZUL")

remove_aleta_except_special <- function(x) {
  stringi::stri_replace_all_regex(
    x,
    pattern = paste0("\\bALETA\\b(?!\\s+(?:", paste(special_aleta_adj, collapse = "|"), "))"),
    replacement = "",
    vectorize_all = FALSE
  )
}


# 3. MAIN PIPELINE (optimized) -----------------

Landings_GoC_Cleaned <- Landings_GoC %>%
  
  # Rename columns
  rename(
    recorded_taxa_name = species_name,
    recorded_taxa_group = main_species_group
  ) %>%
  
  # Extract keywords
  mutate(
    keywords_list = stringi::stri_extract_all_regex(recorded_taxa_name, keyword_pattern)
  ) %>%
  mutate(keywords_list = map(keywords_list, ~ { length(.x) <- 5; .x })) %>%
  unnest_wider(keywords_list, names_sep = "_") %>%
  rename_with(~ paste0("keywords_", 1:5), starts_with("keywords_list_")) %>%
  
  # Clean recorded_taxa_name
  mutate(
    recorded_taxa_name_clean = recorded_taxa_name %>%
      remove_aleta_except_special() %>%                      # Remove ALETA unless special
      stringi::stri_replace_all_regex(keyword_pattern, "") %>% # Remove other keywords
      stringi::stri_replace_all_regex("\\b(Y|DE|P/)\\b", "") %>% # Remove standalone Y/DE/P/
      stringi::stri_replace_all_regex("[.,/\\\\]+", " ") %>%      # Remove punctuation
      str_squish()
  ) %>%
  
  # Extract scientific name
  mutate(
    recorded_scientific_name = stringi::stri_extract_first_regex(recorded_taxa_name_clean, "\\(([^)]+)\\)") %>%
      stringi::stri_replace_all_regex("\\(|\\)", ""),
    scientific_name_recorded_T.F = !is.na(recorded_scientific_name),
    recorded_scientific_name = if_else(
      scientific_name_recorded_T.F,
      stringi::stri_trans_totitle(recorded_scientific_name),
      NA_character_
    )
  ) %>%
  
  # Extract common name
  mutate(
    common_name_clean = stringi::stri_trim_both(
      stringi::stri_replace_first_regex(recorded_taxa_name_clean, "\\s*\\([^)]*\\)", "")
    ),
    common_name_clean = tolower(common_name_clean)
  )


# 4. REORDER COLUMNS -----------------

Landings_GoC_Cleaned <- Landings_GoC_Cleaned %>%

  select(
    # General info
    state, office_name, landing_site, landing_site_key, landing_site_primary, GoC, lat, long,
    year, month,
    receipt_date_fixed, period_start_fixed, period_end_fixed,
    fleet, acuaculture_production, eu_rnpa, eu_name, vessel_rnpa, vessel_name,
    
    # Species info
    recorded_taxa_group, recorded_taxa_name,
    
    # Cleaned / derived names
    recorded_taxa_name_clean, recorded_scientific_name,
    scientific_name_recorded_T.F, common_name_clean,
    
    # Keywords
    keywords_1, keywords_2, keywords_3, keywords_4, keywords_5,
    
    # Landing details
    landed_weight, live_weight, value
 
  )

# How many common names are there in this dataset?
length(unique(Landings_GoC_Cleaned$common_name_clean))
length(unique(Landings_GoC_Cleaned$recorded_taxa_group))
# 734 common names (versus 973 with the less comprehensive keywords_list), over 57 taxonomic groups
```

```{r Check that this worked}
TEST <- Landings_GoC_Cleaned %>%
  filter(str_detect(recorded_taxa_name, "ATUN ALETA AMARILLA"))
```
# Yes, it seems to have worked!

## Save Landings_GoC_Cleaned rds file
```{r Save this dataframe - landings keywords & species for top groupings: Nov 10 2025}
saveRDS(Landings_GoC_Cleaned, file="/home/shares/ohi/OHI_GOC/goal_prep/fis/v2025/int/juan_carlos/GoC_Landings_wKeyWords_2000.present_Dec012025.rds")
```

```{r Skeleton of master key from Landings_GoC_Cleaned dataframe}
Master_SpKey_Skeleton<- Landings_GoC_Cleaned %>% select(recorded_taxa_group, recorded_taxa_name_clean, common_name_clean, scientific_name_recorded_T.F, recorded_scientific_name) %>% distinct
# 802 observations, 5 variables.
```

# PART 1 COMPLETE.

# PART 2: Create MASTER SPECIES KEY: by merging the skeleton with keys from Mar, & Hem ------------------------------

```{r Other Species Keys}
# Erica_SpKey <- read_csv("/home/ferrer/ohi_goc_prep/general/Species_Keys/Erica_CONAPESCA_SpKey_Nov052025.csv")
# # 600 obs of 17 variables

# AllGoC_SpKey <- read_csv("/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/CONAPESCA_SpeciesNames_OctNov2025/Outdated_SpKeys/AllGoC_SpKey_Oct312025.csv")
# 1246 obs of 8 variables

Hem_SpKey <- read_csv("/home/ferrer/ohi_goc_prep/general/Species_Keys/Hem_CONAPESCA_SpKey_2018.csv")
# 8232 obs of 3 variables

Mar_SpKey <- read_csv("/home/ferrer/ohi_goc_prep/general/Species_Keys/Mar_CONAPESCA_SpKey.csv")
# 443 obs of 30 variables
```

```{r Prepare to merge}
Mar_SpKey <- Mar_SpKey %>%
  mutate(common_name = tolower(common_name))

Mar_SpKey_filtered <- Mar_SpKey %>% select(common_name, fishery, highest_taxonomic_id, highest_taxonomic_resolution, species, genus, family, order, class, functional_major_groups_sm= IHH_functional_major_groups_sm)

Mar_SpKey_filtered <- Mar_SpKey_filtered %>%
  select(common_name, fishery, highest_taxonomic_id, highest_taxonomic_resolution,
         species, genus, family, order, class, functional_major_groups_sm) %>%
  rename_with(~ paste0("IHH_", .x))

Hem_SpKey <- Hem_SpKey %>%
  mutate(common_name = tolower(common_name))

Hem_SpKey_filtered <- Hem_SpKey %>%
  select(Hem_common_name= common_name, species)

Hem_SpKey_filtered<- Hem_SpKey_filtered %>% 
  distinct() %>%
  group_by(Hem_common_name) %>%
  mutate(species_num = row_number()) %>%
  ungroup()

Hem_SpKey_filtered <- Hem_SpKey_filtered %>%
  pivot_wider(
    names_from = species_num,
    values_from = species,
    names_prefix = "Hem_species_"
  )
```

```{r Merge}
# merged_df1 <- stringdist_join(
#   Master_SpKey_Skeleton,
#   Mar_SpKey_filtered,
#   by = c("common_name_clean" = "common_name_IHH"),
#   method = "jw",          # Jaro-Winkler works well for name matching
#   max_dist = 0.10,        # adjust as needed
#   distance_col = "dist"   # keep the distance so you can inspect match quality
# )

merged_SpKey_0 <- left_join(
  Master_SpKey_Skeleton,
  Mar_SpKey_filtered,
  by = c("common_name_clean" = "IHH_common_name")
)

Master_SpKey <- left_join(
  merged_SpKey_0,
  Hem_SpKey_filtered,
  by = c("common_name_clean" = "Hem_common_name")
)
```

```{r MASTER SPECIES KEY}
write.csv(Master_SpKey, "/home/ferrer/ohi_goc_prep/general/Species_Keys/Master_SpKey_Dec012025.csv", row.names = FALSE)
```

# PART 2 COMPLETE.

# PART 3: Merge sp key with landings data ------------------------------
```{r Merge landings with sp key}
Landings_GoC_Cleaned_wSp <- Landings_GoC_Cleaned %>%
  left_join(
    Master_SpKey,
    by = c(
      "recorded_taxa_group",
      "recorded_taxa_name_clean",
      "recorded_scientific_name",
      "scientific_name_recorded_T.F",
      "common_name_clean"
    )
  )
```

## Save Landings_GoC_Cleaned_wSp rds file
```{r Save this dataframe - landings keywords & species for top groupings: Nov 10 2025}
saveRDS(Landings_GoC_Cleaned_wSp, file="/home/shares/ohi/OHI_GOC/goal_prep/fis/v2025/int/juan_carlos/GoC_Landings_wSp_2000.present_UNDERCONSTRUCTION_Dec012025.rds")
```