---
title: 'Script 1 — Download Copernicus Primary Production (PP)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
editor_options:
  chunk_output_type: console
---

# Purpose

This script downloads Copernicus GlobColour **Primary Production (PP)** for the
Gulf of California — a modeled satellite estimate of phytoplankton carbon
fixation (closer to **NPP** than raw Chl-a alone).

PP is estimated from ocean-color Chl plus light (PAR), temperature, and related
inputs (Antoine & Morel–type algorithm in the GlobColour / CMEMS product).

**This is for a climate / declining-productivity pressure**, not eutrophication.

| Role | Path |
|------|------|
| This script | `_pressures/prs_npp/script1_download_pp.qmd` |
| Raw NetCDF (Mazu) | `/home/shares/ohi/OHI_GOC/_raw_data/npp_copernicus/` |
| Monthly GeoTIFFs | `/home/shares/ohi/OHI_GOC/_raw_data/npp_copernicus/monthly_rasters/` |
| Related Chl download | `_pressures/prs_npp/script1_download_chla.qmd` |

---

# Product information

Same GlobColour BGC family as the Chl-a download, but the **primary production**
dataset:

*DOI (product):* https://doi.org/10.48670/moi-00281

*Product identifier:* `OCEANCOLOUR_GLO_BGC_L4_MY_009_104`

*Dataset:* Primary production, multi-sensor (4 km), monthly  
`cmems_obs-oc_glo_bgc-pp_my_l4-multi-4km_P1M`

*Typical units:* mg C m⁻² day⁻¹

Catalogue:  
<https://data.marine.copernicus.eu/product/OCEANCOLOUR_GLO_BGC_L4_MY_009_104/description>

---

# Setup

```{r}
if (!require(librarian)) {
  install.packages("librarian")
  library(librarian)
}

librarian::shelf(
  here,
  terra,
  tictoc,
  RColorBrewer,
  tidyverse,
  reticulate
)

current_year <- 2025L
data_dir_version_year <- paste0("d", current_year)

# Raw download home (user-requested location on Mazu)
raw_npp_dir <- "/home/shares/ohi/OHI_GOC/_raw_data/npp_copernicus"
raw_nc_dir  <- file.path(raw_npp_dir, data_dir_version_year)
monthly_dir <- file.path(raw_npp_dir, "monthly_rasters")

dir.create(raw_nc_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(monthly_dir, showWarnings = FALSE, recursive = TRUE)

nc_filename <- "goc_globcolour_pp_monthly.nc"

message("NetCDF directory: ", raw_nc_dir)
message("Monthly GeoTIFF directory: ", monthly_dir)

cols <- rev(colorRampPalette(brewer.pal(9, "Spectral"))(255))
```

## Install / check `copernicusmarine` (Python)

```{python}
import sys
import subprocess

def install(package):
    subprocess.check_call([sys.executable, "-m", "pip", "install", package])

try:
    import copernicusmarine
except ImportError:
    install("copernicusmarine")
    import copernicusmarine

print(f"copernicusmarine version: {copernicusmarine.__version__}")
```

If the import fails from R/Quarto, restart R and try:

```{r}
# library(reticulate)
# use_virtualenv("~/.virtualenvs/r-reticulate", required = TRUE)
# py_install("copernicusmarine", pip = TRUE, upgrade = TRUE)
# py_run_string("import copernicusmarine; print(copernicusmarine.__version__)")
```

---

# Download Primary Production (GoC subset)

Same Gulf bounding box as the Chl-a Script 1 download. Update the end date when
refreshing in future years.

**Credentials:** use `copernicusmarine.login()` once on this machine (do **not**
hard-code passwords in the script). Saved credentials are reused afterward.

```{python}
import os
import copernicusmarine

# Uses saved credentials if present; otherwise prompts / opens login flow
copernicusmarine.login()

raw_nc_dir = "/home/shares/ohi/OHI_GOC/_raw_data/npp_copernicus/d2025"
os.makedirs(raw_nc_dir, exist_ok=True)
out_nc = os.path.join(raw_nc_dir, "goc_globcolour_pp_monthly.nc")

# ============== Update end_datetime when refreshing !!!! ==============
copernicusmarine.subset(
  dataset_id="cmems_obs-oc_glo_bgc-pp_my_l4-multi-4km_P1M",
  variables=["PP"],
  minimum_longitude=-118.52665401785711,
  maximum_longitude=-98.83917589285713,
  minimum_latitude=16.492029666573647,
  maximum_latitude=34.17058145228793,
  start_datetime="1997-09-01T00:00:00",
  end_datetime="2025-09-01T00:00:00",
  output_filename=out_nc
)

print(f"Data downloaded to: {out_nc}")
print(f"File size MB: {os.path.getsize(out_nc) / 1e6:.1f}")
```

---

# Open as SpatRaster and check

```{r}
nc_path <- file.path(raw_nc_dir, nc_filename)
pp_raster <- terra::rast(nc_path)

print(pp_raster)
cat("CRS:\n")
print(crs(pp_raster))

num_layers <- terra::nlyr(pp_raster)
cat("Number of layers:", num_layers, "\n")

time_attribute <- terra::time(pp_raster)
cat("First time:", as.character(time_attribute[1]), "\n")
cat("Last time: ", as.character(time_attribute[num_layers]), "\n")
```

```{r}
# Quick look at one month (log10 helps visualization)
i_ex <- which(format(time_attribute, "%Y-%m") == "2020-06")[1]
if (is.na(i_ex)) i_ex <- 1
plot(
  log10(pp_raster[[i_ex]] + 1),
  main = paste("log10(PP + 1) —", format(time_attribute[i_ex], "%Y-%m")),
  col = cols
)
```

---

# Save individual monthly GeoTIFFs

One file per month: `YYYY-MM-DD.tif` under `monthly_rasters/`.

```{r}
tic()
for (i in seq_len(num_layers)) {
  layer <- pp_raster[[i]]
  date  <- time_attribute[i]
  layer_name <- format(date, "%Y-%m-%d")
  output_file <- file.path(monthly_dir, paste0(layer_name, ".tif"))
  writeRaster(layer, filename = output_file, overwrite = TRUE)
  if (i %% 50 == 0 || i == num_layers) {
    message("Saved ", i, "/", num_layers, ": ", basename(output_file))
  }
}
toc()

cat("Monthly GeoTIFFs written to:\n ", monthly_dir, "\n")
cat("Count:", length(list.files(monthly_dir, pattern = "\\.tif$")), "\n")
```

```{r}
ex <- rast(file.path(monthly_dir, "2001-10-01.tif"))
plot(
  log10(ex + 1),
  main = "Primary production — October 2001 (log10)",
  xlab = "Longitude",
  ylab = "Latitude",
  col = cols
)
print(ex)
```

---

# Done

Raw NetCDF and monthly GeoTIFFs are ready for a PP version of the regional
monthly exploration (analogous to `script2_chla_regional_monthly_explore.Rmd`).
