---
title: "OHI GOC final scores first draft (Sep 2026)"
author: "Erica Ferrer"
date: "`r Sys.Date()`"
output: html_document
---

# PRE-WORK ------------------------------

```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load libraries & set working directories}
library(tidyverse)
library(here)
library(plotly)
library(zoo)

# No scientific notation:
options(scipen = 999)

final_output_dir <- "/home/ferrer/ohi_goc_prep/_index_scores_1.0/"
```

```{r Load clean_names_lowercase function}
clean_names_lowercase <- function(x) {

  x <- iconv(x, from = "", to = "UTF-8", sub = "")
  x <- stringi::stri_trans_general(x, "Latin-ASCII")

  x %>%
    tolower() %>%
    trimws()
}
```


# LOAD AND PREP INDIVIDUAL GOAL SCORES ------------------------------

# AO (access_opportunities)

```{r access_opportunities AKA artisanal_fishing_opporunities}
# Load original:
AFO_Scores0 <- read_csv(here::here("/home/ferrer/ohi_goc_prep/AO/v2026/AO_Scores/AFO_Scores/AFO_Scores_1.0_df_Jun152026.csv"))
# 70 obs., 3 vars.

colnames(AFO_Scores0)

# Clean:
AO_Scores<- AFO_Scores0 %>%
  select(year, OHI_region_no = OHI_region_est, AO_score= AFO_Score) %>% 
  mutate(goal = "access_opporunities")
# 70 obs., 4 vars.


# Save:
write_csv(AO_Scores, here::here(final_output_dir, "_goal_scores_1.0/AO_Scores_1.0_df_Sep222026.csv"))
```


# BD (biodiversity)

```{r biodiversity}
# Load original:
BD_Scores0 <- read_csv(here::here("/home/ferrer/ohi_goc_prep/BD/v2026/BD_Scores_1.0/BD_Scores_1.0_df_Sep082026.csv"))
# 63 obs., 4 vars.

colnames(BD_Scores0)

# Clean:
BD_Scores<- BD_Scores0 %>%
  select(year, OHI_region_no = OHI_region, BD_score) %>% 
  mutate(goal = "biodiversity")
# 63 obs., 4 vars.


# Save:
write_csv(BD_Scores, here::here(final_output_dir, "_goal_scores_1.0/BD_Scores_1.0_df_Sep222026.csv"))
```


# CW (clean_waters)

```{r clean_waters}
# Load original:
CW_Scores0 <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/cw_current_status_arithmetic.csv"))
# 81 obs., 7 vars.

colnames(CW_Scores0)

# Clean:
CW_Scores<- CW_Scores0 %>%
  select(year, OHI_region_no= rgn_id, CW_score = CW) %>% 
  mutate(goal = "clean_waters",
         OHI_region_no = clean_names_lowercase(OHI_region_no))
# 81 obs., 4 vars.


# Save:
write_csv(CW_Scores, here::here(final_output_dir, "_goal_scores_1.0/CW_Scores_1.0_df_Sep222026.csv"))
```


# FP (food_provisioning)

```{r food_provisioning}
# Load original:
FP_Scores0 <- read_csv(here::here("/home/ferrer/ohi_goc_prep/FP/v2026/FP_Scores_1.1/FP_Scores_1.1_df_Jul282026.csv"))
# NOTE TO SELF: Including 1.1 score since these do not include CPUE.
# 63 obs., 10 vars.

colnames(FP_Scores0)

# Clean:
FP_Scores<- FP_Scores0 %>%
  select(year, OHI_region_no = OHI_region_est, FP_score) %>% 
  mutate(goal = "food_provisioning")
# 63 obs., 4 vars.


# Save:
write_csv(FP_Scores, here::here(final_output_dir, "_goal_scores_1.0/FP_Scores_1.0_df_Sep222026.csv"))
```


# HS (habitat_services)

```{r habitat_services}
# Load original:
HS_Scores0 <- read_csv(here::here("/home/ferrer/ohi_goc_prep/HS/v2026/HS_Scores_1.1/HS_current_status_Sep112026.csv"))
# 54 obs., 6 vars.

colnames(HS_Scores0)

# Clean:
HS_Scores<- HS_Scores0 %>%
  select(year, OHI_region_no = rgn_id, HS_score) %>% 
  mutate(goal = "habitat_services",
         OHI_region_no= clean_names_lowercase(OHI_region_no))
# 54 obs., 4 vars.


# Save:
write_csv(HS_Scores, here::here(final_output_dir, "_goal_scores_1.0/HS_Scores_1.0_df_Sep222026.csv"))
```


# LE (livelihoods_economies)

```{r livelihoods_economies}
LE_Scores0 <- read_csv(here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/le",
    "final_le_current_status_2019ref_informality.csv"
  ))
# 41 obs., 4 vars.

colnames(LE_Scores0)

# Clean:
LE_Scores<- LE_Scores0 %>%
  select(year = ano_censal, OHI_region_no= rgn_id, LE_score = final_le_current_status_rgn) %>% 
  mutate(goal = "livelihoods_economies",
         OHI_region_no = clean_names_lowercase(OHI_region_no))
# 41 obs., 4 vars.

# Gapfill missing years (NOTE TO SELF: this will be updated)
LE_Scores_gf <- LE_Scores %>% 
  complete(year = full_seq(2003:2025, 1), OHI_region_no) %>%  # expand to every year 2003-2025
  arrange(year, OHI_region_no) %>% 
  group_by(OHI_region_no) %>% 
  mutate(LE_score = na.approx(LE_score, x = year, na.rm = FALSE, rule = 2)) %>% 
  ungroup() %>% 
  mutate(goal = "livelihoods_economies")
# 207 obs., 4 vars.


# Save:
write_csv(LE_Scores_gf, here::here(final_output_dir, "_goal_scores_1.0/LE_Scores_1.0_df_Sep222026.csv"))
```

# NP (natural_products)

```{r natural_products}
# Load original:
NP_Scores0 <- read_csv(here::here("/home/ferrer/ohi_goc_prep/NP/v2026/NP_Scores_1.0/NP_Scores_1.0_df_Sep082026.csv"))
# NOTE TO SELF: Including 1.1 score since these do not include CPUE.
# 63 obs., 10 vars.

colnames(NP_Scores0)

# Clean:
NP_Scores<- NP_Scores0 %>%
  select(year, OHI_region_no = OHI_region, NP_Score) %>% 
  mutate(goal = "natural_products")
# 63 obs., 4 vars.


# Save:
write_csv(NP_Scores, here::here(final_output_dir, "_goal_scores_1.0/NP_Scores_1.0_df_Sep222026.csv"))
```


# SP (sense_of_place)

```{r sense_of_place}
# Load original:
SP_Scores0 <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/sp_current_status_noCCE.csv"))
# 63 obs., 3 vars.

colnames(SP_Scores0)

# Clean:
SP_Scores<- SP_Scores0 %>%
  select(year, OHI_region_no= rgn_id, SP_score= current_status) %>%
  mutate(goal = "sense_of_place",
         OHI_region_no = clean_names_lowercase(OHI_region_no))
# 63 obs., 4 vars.


# Save:
write_csv(SP_Scores, here::here(final_output_dir, "_goal_scores_1.0/SP_Scores_1.0_df_Sep222026.csv"))
```


# TR (tourism_recreation) 

```{r tourism_recreation}
# Load original:
TR_Scores0 <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output/tr_current_status_final_083126.csv"))
# 63 obs., 3 vars.

colnames(TR_Scores0)

# Clean:
TR_Scores<- TR_Scores0 %>%
  rename(TR_score = tr_current_status, OHI_region_no= rgn_id) %>% 
  mutate(goal = "tourism_recreation",
         OHI_region_no = clean_names_lowercase(OHI_region_no))
# 63 obs., 4 vars.
  

# Save:
write_csv(TR_Scores, here::here(final_output_dir, "_goal_scores_1.0/TR_Scores_1.0_df_Sep222026.csv"))
```


# Join all goal dataframes:

## Long:
```{r}

Index_Scores_long0 <- bind_rows(
  AO_Scores,
  BD_Scores,
  CW_Scores,
  HS_Scores,
  FP_Scores,
  LE_Scores,
  NP_Scores,
  SP_Scores,
  TR_Scores
) %>%
  arrange(year, OHI_region_no, goal)

# Coalesce score values:
Index_Scores_long <- Index_Scores_long0 %>%
  mutate(
    goal_score = coalesce(
      AO_score,
      BD_score,
      CW_score,
      HS_score,
      FP_score,
      LE_score,
      NP_Score,
      SP_score,
      TR_score
    )
  ) %>%
  select(
    year,
    OHI_region_no,
    goal,
    goal_score
  ) %>%
  arrange(year, OHI_region_no, goal)

```

## Wide

```{r Index_Scores_wide0}
Index_Scores_wide0 <- Index_Scores_long %>%
  pivot_wider(
    id_cols = c(year, OHI_region_no),
    names_from = goal,
    values_from = goal_score
  ) %>%
  rename(
    AO_score= access_opporunities,   
    BD_score=biodiversity,
    CW_score= clean_waters,
    FP_score= food_provisioning,
    HS_score= habitat_services,
    LE_score= livelihoods_economies,
    NP_score=natural_products,
    SP_score=sense_of_place,
    TR_score=tourism_recreation
  )


Index_Scores_wide <- Index_Scores_wide0 %>%
  mutate(
    Index_score = rowMeans(
      select(., 
             AO_score,  
             BD_score,
             CW_score,
             FP_score,
             HS_score,
             LE_score,
             NP_score,
             SP_score,
             TR_score
      ),
      na.rm = TRUE
    )
  )

Index_Scores_wide_2023<- Index_Scores_wide %>%
  filter(year == 2023,
         !is.na(OHI_region_no))
```


# SAVE:

```{r Save Index_Scores_wide}
write.csv(Index_Scores_wide,
          file= "/home/ferrer/ohi_goc_prep/_index_scores_1.0/Index_Scores_1.0_df_Sep222026.csv",
          row.names = FALSE)
```


# MAP

```{r Read in OHI region boundaries}
# Read in OHI region boundaries
ohi_regions <- st_read(here("spatial", "ohi_regions", "goc_ohi_rgns.shp")) |> st_make_valid()

colnames(ohi_regions)

ohi_regions<- ohi_regions %>%
  rename(OHI_region_no= rgn_id) %>%
  mutate(
    OHI_region_no= clean_names_lowercase(OHI_region_no)
  )
```

```{r Map Index_Scores_wide_2023}

# 2. Join onto OHI polygons
Index_Scores_wide_2023 <- ohi_regions |>
  left_join(Index_Scores_wide_2023, by = "OHI_region_no")

# 3. Split marine vs land
ohi_marine <- Index_Scores_wide_2023 %>% filter(location == "marine")
ohi_land   <- Index_Scores_wide_2023 %>% filter(location == "land")

# 4. Map
tmap_mode("plot")

Index_Scores_2023_map<- tm_shape(ohi_land) +
  tm_polygons(
    col = "#f5e6c8",
    border.col = "gray40",
    lwd = 1,
    alpha = 0.5,
    legend.show = FALSE
  ) +

tm_shape(ohi_marine) +
  tm_polygons(
    col = "Index_score",
    palette = "RdYlGn",
    alpha = 0.8,
    title = "Index Scores (version 1.0)",
    border.col = "black",
    lwd = 1.2
  ) +

tm_layout(
  title = "Final Index Scores per Region (2023)",
  frame = FALSE,
  legend.outside = TRUE
)

Index_Scores_2023_map
```

```{r Save map}
tmap::tmap_save(
  tm = Index_Scores_2023_map,
  filename = here::here(
    final_output_dir,
    "Index_Scores_1.0_map_Sep222026.png"
  ),
  width = 8,
  height = 6,
  units = "in",
  dpi = 300
)
```

# GAP FILL:
Rules:
(1) If data (per region) is missing between two years where there is data available, interpolate the value. 
(2) If data (per region) is missing and is not surrounding by data before and after, take the most recent (previous) datapoint.

I.e.,
Internal gaps: linearly interpolate between the nearest available values before and after the missing year.
Leading gaps: if there is no later value and no earlier value surrounding the gap, carry the most recent previous value forward.
Don't use future values to fill a missing value when there is no previous observation.

```{r Gapfill the Index_scores_wide}
score_cols <- c(
  "LE_score", "CW_score", "AO_score", "FP_score", "NP_score",
  "BD_score", "SP_score", "TR_score", "HS_score"
)

Index_Scores_gapfilled <- Index_Scores_wide %>%
  filter(year >=2018, year <=2025) %>%
  filter(!is.na(OHI_region_no)) %>%
  arrange(OHI_region_no, year) %>%
  group_by(OHI_region_no) %>%
  mutate(
    across(
      all_of(score_cols),
      ~ {
        x <- .x
        interp <- na.approx(x, x = year, na.rm = FALSE)
        na.locf(interp, na.rm = FALSE)
      },
      .names = "{.col}_Score"
    ),
    across(
      all_of(score_cols),
      ~ {
        x <- .x
        interp <- na.approx(x, x = year, na.rm = FALSE)
        filled <- na.locf(interp, na.rm = FALSE)

        case_when(
          !is.na(x) ~ "observed",
          !is.na(interp) ~ "interpolated",
          !is.na(filled) ~ "carried_forward",
          TRUE ~ NA_character_
        )
      },
      .names = "{.col}_provenance"
    )
  ) %>%
  ungroup() %>%
  select(
    OHI_region_no,
    year,
    unlist(
      lapply(
        score_cols,
        function(x) c(
          x,
          paste0(x, "_provenance")
        )
      )
    )
  )
# 72 obs., 20 vars.
```

```{r Calculate final (gapfilled) score}
colnames(Index_Scores_gapfilled)

Index_Scores_gapfilled <- Index_Scores_gapfilled %>%
  mutate(
    Index_score_gapfilled = rowMeans(
      select(., 
             AO_score,  
             BD_score,
             CW_score,
             FP_score,
             HS_score,
             LE_score,
             NP_score,
             SP_score,
             TR_score
      ),
      na.rm = TRUE
    )
  )
```


# GAPFILLED PLOTS OVER TIME:

```{r Plot Index_Scores over time}
unique(Index_Scores_gapfilled$OHI_region_no)
colnames(Index_Scores_gapfilled)

Region_colors<- c(
  "region_1" = "#C03729",
  "region_2" = "#E68C7C",
  "region_3" = "#FC8F24",
  "region_4" = "#dcaa38",
  "region_5" = "#cbd2a0",
  "region_6" = "#98a25a",
  "region_7" = "#747428",
  "region_8" = "#627391",
  "region_9" = "#8d61a9")


Index_Scores_gf_over_time<- ggplot(Index_Scores_gapfilled, 
                                aes(x = year,
                                    y = Index_score_gapfilled,
                                    color = OHI_region_no)) +
  
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  
  facet_wrap(~ OHI_region_no) +
  scale_x_continuous(breaks = sort(unique(Index_Scores_gapfilled$year))) +
  
  labs(
    title = "Index Scores by Region [DRAFT]",
    subtitle= "updated: Sep 22 2026",
    y = "Goal Score (from 0 to 100)",
    x = "Year"
  ) +
  
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey50") +
  
  ylim(0, 101) +
  
  scale_color_manual(values = Region_colors) +
  
  theme_bw() +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank()
  )

Index_Scores_gf_over_time
```

```{r Save Index_Scores_gf_over_time as PNG}
  ggsave(
    filename = sprintf("/home/ferrer/ohi_goc_prep/_index_scores_1.0/Index_Scores_DRAFT_plot_Sep222026.png"),
    plot = Index_Scores_gf_over_time,
    width = 12,
    height = 12,
    dpi = 300
  )
```