################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

original author: "Austin Martinez"

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Load in Libraries
library(sf)
library(dplyr)
library(here)
library(terra)
library(tidyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(tmap)

```


# STEP 1: LOAD IN OHI REGIONS & CREATE 20KM BUFFER ------------------------------
 
```{r Load in the OHI Region_Key0}
Region_Key0<- read.csv("/home/ferrer/ohi_goc_prep/FP/v2026/CONAPESCA_Keys/Spatial_Keys/RegionKey_FINAL_May262026.csv")

colnames(Region_Key0)

Region_Key<- Region_Key0 %>%
  select(OHI_region= OHI_region_est, OHI_region_name) %>%
  distinct()
```

```{r Read in OHI region boundaries}
# Read in OHI region boundaries
ohi_regions <- st_read(here("spatial", "ohi_regions", "goc_ohi_rgns.shp")) |> st_make_valid()

```

```{r Create 20km coastal buffer around OHI regions}
# DECISION POINT: We clipped coverage to a 20-km coastal buffer to avoid capturing coverage located on the west side of the Baja California peninsula.

COASTAL_BUFFER = 20000 # (20,000m or 20km)

# Create 20km buffer (run once)
sf::sf_use_s2(FALSE)

ohi_regions_marine <- ohi_regions |>
  filter(location == "marine") |>
  st_union()

# Make a buffer ring
buffer_outer <- st_buffer(ohi_regions_marine, 
                          dist = COASTAL_BUFFER)

buffer_ring  <- st_difference(buffer_outer, ohi_regions_marine) |> st_make_valid()

# Union it to the marine OHI layer
ohi_marine_plus_20km <- st_union(ohi_regions_marine, buffer_ring)

```


# STEP 2: CALCULATE MANGROVE "CURRENT STATUS", i.e. condition over time (source: Global Mangrove Watch, 1996-2020)  ------------------------------

```{r Read in Global Mangrove Watch data}
# Read in mangrove data
GMW_mangrove.data_1996_raw <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/zenodo_mangrove/gmw_v3_1996_vec/gmw_v3_1996_vec.shp")) |> 
  st_make_valid()
GMW_mangrove.data_2007_raw <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/zenodo_mangrove/gmw_v3_2007_vec/gmw_v3_2007_vec.shp")) |> 
  st_make_valid()
GMW_mangrove.data_2010_raw <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/zenodo_mangrove/gmw_v3_2010_vec/gmw_v3_2010_vec.shp")) |> 
  st_make_valid()
GMW_mangrove.data_2015_raw <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/zenodo_mangrove/gmw_v3_2015_vec/gmw_v3_2015_vec.shp")) |> 
  st_make_valid()
GMW_mangrove.data_2020_raw <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/zenodo_mangrove/gmw_v3_2020_vec/gmw_v3_2020_vec.shp")) |> 
  st_make_valid()
GMW_mangrove.data_2017_raw <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/zenodo_mangrove/gmw_v3_2017_vec/gmw_v3_2017_vec.shp")) |> 
  st_make_valid()

```

```{r Match the land-use CRS to the ohi_regions CRS}
GMW_mangrove.data_1996 <- st_transform(GMW_mangrove.data_1996_raw, st_crs(ohi_regions))
GMW_mangrove.data_2007 <- st_transform(GMW_mangrove.data_2007_raw, st_crs(ohi_regions))
GMW_mangrove.data_2010 <- st_transform(GMW_mangrove.data_2010_raw, st_crs(ohi_regions))
GMW_mangrove.data_2015 <- st_transform(GMW_mangrove.data_2015_raw, st_crs(ohi_regions))
GMW_mangrove.data_2017 <- st_transform(GMW_mangrove.data_2017_raw, st_crs(ohi_regions))
GMW_mangrove.data_2020 <- st_transform(GMW_mangrove.data_2020_raw, st_crs(ohi_regions))

```

## Clip mangrove data to the 20km buffer & estimate total area per OHI region:
```{r Clip mangrove data 1996 through 2020 to the 20km buffer extent}
# Clip mangrove data to the 20km buffer extent (defined above)
GMW_mangrove.data_1996 <- st_intersection(GMW_mangrove.data_1996, ohi_marine_plus_20km)
GMW_mangrove.data_2007 <- st_intersection(GMW_mangrove.data_2007, ohi_marine_plus_20km)
GMW_mangrove.data_2010 <- st_intersection(GMW_mangrove.data_2010, ohi_marine_plus_20km)
GMW_mangrove.data_2015 <- st_intersection(GMW_mangrove.data_2015, ohi_marine_plus_20km)
GMW_mangrove.data_2017 <- st_intersection(GMW_mangrove.data_2017, ohi_marine_plus_20km)
GMW_mangrove.data_2020 <- st_intersection(GMW_mangrove.data_2020, ohi_marine_plus_20km)

```

## Assign the mangrove area to OHI regions:
```{r Assign the mangrove area to OHI regions}
# Assign the mangrove area to OHI regions (per year) in:
# 1996
MANG_area_gmw_1996 <- st_intersection(GMW_mangrove.data_1996, ohi_regions) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 1996
  ) |>
  st_drop_geometry()

# 2007
MANG_area_gmw_2007 <- st_intersection(GMW_mangrove.data_2007, ohi_regions) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2007
  ) |>
  st_drop_geometry()

# 2010
MANG_area_gmw_2010 <- st_intersection(GMW_mangrove.data_2010, ohi_regions) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2010
  ) |>
  st_drop_geometry()

# 2015
MANG_area_gmw_2015 <- st_intersection(GMW_mangrove.data_2015, ohi_regions) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2015
  ) |>
  st_drop_geometry()

# 2017
MANG_area_gmw_2017 <- st_intersection(GMW_mangrove.data_2017, ohi_regions) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2017
  ) |>
  st_drop_geometry()

# 2020
MANG_area_gmw_2020 <- st_intersection(GMW_mangrove.data_2020, ohi_regions) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2020
  ) |>
  st_drop_geometry()
```

```{r Combine all years' worth of data into one dataframe (MANG_area_gmw0)}
# Merge:
MANG_area_gmw0 <- bind_rows(MANG_area_gmw_1996, MANG_area_gmw_2007, MANG_area_gmw_2010, MANG_area_gmw_2015, MANG_area_gmw_2017, MANG_area_gmw_2020) |>
  st_drop_geometry() |>
  group_by(rgn_id, year) |>
  summarise(area_km2 = sum(area_km2), .groups = "drop")

dim(MANG_area_gmw0)
# 42 obs.,  3 vars.
```


## GAPFILL BETWEEN OBSERVED YEARS:
```{r Gapfill MANG_area_gmw0}

# First, expand the NRSI dataframe to include one row per year per region, even if that region has no data: 
MANG_area_gmw <- MANG_area_gmw0 %>%
  complete(
    rgn_id = c(paste0("Region_", 1:9)),
    year = full_seq(year, 1)
  )
# 225 obs., 3 vars.

# Keep track of what was originally observed:
MANG_area_gmw<- MANG_area_gmw %>%
  rename(observed_area_km2= area_km2) %>%
  mutate(
    estimated_area_km2 = observed_area_km2
  )

# Gapfill area_km2 in missing years using the zoo package:
MANG_area_gmw <- MANG_area_gmw %>%
  group_by(rgn_id) %>%
  fill(rgn_id, .direction = "downup") %>%
  mutate(

    # Gap-fill missing values
    estimated_area_km2 = zoo::na.approx(
      estimated_area_km2,
      x = year,
      na.rm = FALSE
    ),

    # Classify provenance
    provenance = case_when(
      !is.na(observed_area_km2) ~ "observed",
      is.na(observed_area_km2) & !is.na(estimated_area_km2) ~ "gapfilled",
      is.na(estimated_area_km2) ~ NA_character_
    )
  ) %>%
  ungroup()

# 225 obs., 5 vars.
```


# STEP 3: CALCULATE MANG SCORES (over time) based on condition from Global Mangrove Watch data ------------------------------

```{r Calculate MANG_Scores0}
colnames(MANG_area_gmw)

# Calculate MANG_Scores0 relative to 1996 baseline, where scores range from 0-100:
MANG_Scores0 <- MANG_area_gmw %>%
  group_by(rgn_id) %>%
  mutate(
    MANG_current_status = (estimated_area_km2 / estimated_area_km2[year == 1996])*100,
    MANG_current_status = pmin(MANG_current_status, 100),
    MANG_current_status = round(MANG_current_status, 3)
  ) %>%
  select(rgn_id, year, provenance, estimated_area_km2, MANG_current_status)
```

# STEP 4: CALCULATE MANGROVE "CURRENT EXTENT" (source: Hem's Data, est. 2025) ------------------------------

```{r Load in Hem's mangrove data}
# Read in Hem data
hem_mangroves <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/mangrove_Hem/USV_AGRUPADO/USV_AGRUPADO.shp")) |> 
  st_make_valid()

# Match hem data to OHI CRS
hem_mangroves <- st_transform(hem_mangroves, st_crs(ohi_regions))

```

```{r Filter Hem's data to only mangrove layers}
# Filter:

hem_mangroves_only <- hem_mangroves |>
  filter(DESCRIP_AG %in% c("MANGLAR", 
                           "VEGETACIÓN SECUNDARIA ARBÓREA DE MANGLAR", 
                           "VEGETACIÓN SECUNDARIA ARBUSTIVA DE MANGLAR",
                           "VEGETACIÓN SECUNDARIA HERBÁCEA DE MANGLAR")) |>
  st_union() |>
  st_as_sf() |>
  st_make_valid()

# Clip mangrove data to the 20km buffer extent
hem_mangroves_only <- st_intersection(hem_mangroves_only, ohi_marine_plus_20km)
```

```{r Assign the mangrove extent to each OHI region}

# Assign the mangrove extent to each OHI region
MANG_Current_Extent <- st_intersection(hem_mangroves_only, ohi_regions) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(x), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    area_km2 = round(area_km2, 1),
    year = 2025
  ) |>
  st_drop_geometry()

# Ensure all regions 1–9 are present; regions with no mangrove get area = 0
all_rgns <- tibble(
  rgn_id = paste0("Region_", 1:9),
  year = 2025
)

MANG_Current_Extent <- all_rgns |>
  left_join(MANG_Current_Extent, by = c("rgn_id", "year")) |>
  mutate(area_km2 = replace_na(area_km2, 0)) |>
  select(rgn_id, area_km2, year)

# Make a nice Ktable
MANG_Current_Extent |>
  kable(digits = 4, col.names = c("Region", "Extent (km²)", "Year")) |>
  kable_styling(bootstrap_options = c("striped", "bordered")) |>
  add_header_above(
    c('Mangrove Extent' = 3))
```

## Save MANG_Current_Extent

# NOTE TO SELF: This actually didn't get saved as I would have expted. Re-run when possible.
```{r Save MANG_Current_Extent}
write.csv(MANG_Current_Extent,
          here::here("_habitats/HAB_mangroves/MAN_outputs/MANG_Current_Extent_Sep302026.csv"),
          row.names = FALSE)
```


# STEP 5: FINALIZE MANG_Scores DATAFRAME ------------------------------

```{r Merge Hem's current extent to MANG_Scores0}
colnames(MANG_Scores0)
colnames(MANG_Current_Extent)

# Merge:
MANG_Scores <- MANG_Scores0 %>%
  left_join(select(MANG_Current_Extent,
                   c("rgn_id", "area_km2")),
            by="rgn_id") %>%
  rename(
    area_est_byCEDO.2025_km2 = area_km2,
    area_est_byGMW_km2 = estimated_area_km2
  )
  
# Make an include / exclude column based on the information reported by Hem / CEDO:
MANG_Scores<- MANG_Scores %>%
  mutate(MANG_include_exclude = ifelse(area_est_byCEDO.2025_km2 == 0, "EXCLUDE", 
                                      "INCLUDE"))

```

```{r Make a nice Ktable}
MANG_Scores |>
  select(rgn_id, year, MANG_current_status) |>
  kable(digits = 4, col.names = c("Region", "Year", "Current Status (vs. 1996)")) |>
  kable_styling(bootstrap_options = c("striped", "bordered")) |>
  add_header_above(
    c('Mangrove Current Status' = 3))


```

## Save MANG_Scores
```{r Save MANG_Scores}
# Write CSVs to output_data directory
write.csv(MANG_Scores,
          here::here("_habitats/HAB_mangroves/MAN_outputs/MANG_Scores_1.1_Sep302026.csv"),
          row.names = FALSE)
```


# STEP 6: ESTIMATE MANG_Pressure, the inverse of MANG_current_status ------------------------------
```{r Estimate pressure as 100 minus current status}
MANG_Pressure <- MANG_Scores %>%
  mutate(pressure = (100 - MANG_current_status)/100) %>%
  select(rgn_id, year, pressure)

```

## Save MANG_Pressure
```{r Save MANG_Pressure}
write.csv(MANG_Pressure,
          here::here("_habitats/HAB_mangroves/MAN_outputs/MANG_Pressure_Sep302026.csv"),
          row.names = FALSE)
```

# STEP 7: ESTIMATE MANGROVE TREND (source: GMW, 2010-2020) ------------------------------
# NOTE TO SELF: Might want to adjust how trend is calculated.

```{r Calculate MANG_Trend}
# Calculate trend in each region
MANG_Trend <- MANG_area_gmw |>
  pivot_wider(names_from = year, values_from = area_km2) |>
  mutate(trend = (`2020` / `2010`)^(1/10) - 1,
         trend = round(trend, 4) * 100,
         year = 2020) |>
  select(rgn_id, year, trend)

MANG_Trend <- left_join(select(MANG_Current_Extent, c("rgn_id", "area_km2")), MANG_Trend, by="rgn_id") %>%
   mutate(year = max(MANG_Trend$year)) %>%
  mutate(trend = ifelse(area_km2 == 0, NA, trend)) %>%
  select(-area_km2)

```

```{r Make a nice Ktable for MANG_Trend}
MANG_Trend |>
  select(rgn_id, year, trend) |>
  kable(digits = 4, col.names = c("Region", "Year", "Trend")) |>
  kable_styling(bootstrap_options = c("striped", "bordered")) |>
  add_header_above(
    c('Mangrove Trend' = 3))
```

## Save MANG_Trend
```{r Save MANG_Trend}
write.csv(MANG_Trend,
          here::here("_habitats/HAB_mangroves/MAN_outputs/MANG_Trend_Sep302026.csv"),
          row.names = FALSE)
```


# STEP 8: PLOT DATA ------------------------------

```{r Map of MANG_Scores}
MANG_Scores_clean <- MANG_Scores |>
  mutate(
    MANG_current_status_int = round(MANG_current_status * 100, 1),
    MANG_current_status_int = ifelse(rgn_id %in% c("Region_2", "Region_3"), NA, MANG_current_status_int)
  )

# Join onto OHI polygons
ohi_MANG_Scores <- ohi_regions |>
  left_join(MANG_Scores_clean, by = "rgn_id")

# Split marine vs land
ohi_marine <- ohi_MANG_Scores |> filter(location == "marine")
ohi_land   <- ohi_MANG_Scores |> filter(location == "land")

# Plot
tmap_mode("plot")

tm_shape(ohi_land) +
  tm_polygons(
    col = "#f5e6c8",
    border.col = "gray40",
    lwd = 1,
    alpha = 0.5,
    legend.show = FALSE
  ) +

tm_shape(ohi_marine) +
  tm_polygons(
    col = "MANG_current_status_int",
    palette = "RdYlGn",
    alpha = 0.7,
    title = "MANG_current_status",
    border.col = "black",
    lwd = 1.2
  ) +

tm_layout(
  title = "Mangrove MANG_current_status (2020)",
  frame = FALSE,
  legend.outside = TRUE
)

```

```{r Table of MANG_Scores}
MANG_Scores_table <- MANG_Scores_clean |>
  left_join(
    ohi_regions |> 
      filter(location == "marine") |> 
      select(rgn_id, ecorgn),
    by = "rgn_id"
  ) |>
  left_join(
    MANG_Current_Extent |> 
      select(rgn_id, area_km2),
    by = "rgn_id"
  ) |>
  mutate(
    # Clean region labels
    rgn_id = gsub("_", " ", rgn_id),
    ecorgn = gsub("_", " ", ecorgn),

    # Replace ecorgn with your custom region names
    ecorgn = case_when(
      rgn_id == "Region 1" ~ "Western Upper Gulf",
      rgn_id == "Region 2" ~ "Western Midriff Islands",
      rgn_id == "Region 3" ~ "Western Central Gulf",
      rgn_id == "Region 4" ~ "Western Southern Gulf",
      rgn_id == "Region 5" ~ "Eastern Upper Gulf",
      rgn_id == "Region 6" ~ "Eastern Midriff Islands",
      rgn_id == "Region 7" ~ "Eastern Sonora Coast",
      rgn_id == "Region 8" ~ "Sinaloa Coast",
      rgn_id == "Region 9" ~ "Southeastern Gulf",
      TRUE ~ ecorgn
    ),

    area_km2 = round(area_km2, 2),
    MANG_current_status_int = round(MANG_current_status_int, 1)
  ) |>
  select(
    rgn_id,
    ecorgn,
    area_km2,
    MANG_current_status_int
  )

MANG_Scores_table |>
  filter(!rgn_id %in% c("Region 2", "Region 3")) |>   # <-- remove 2 & 3 here
  kable(
    digits = 1,
    col.names = c("Region ID", "Region Name", "Extent (km²)", "MANG_current_status (%)")
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "bordered", "condensed"),
    full_width = FALSE
  ) |>
  add_header_above(c("Mangrove MANG_current_status" = 4))


```

```{r Map Hem's mangroves current extent}
# Add a dummy column for legend labeling
hem_mangroves_only$habitat <- "Mangrove Habitat"

tmap_mode("plot")

# 1. OHI regions (no legend)
tm_shape(ohi_regions) +
  tm_polygons(
    col = "location",
    palette = c("marine" = "#cceeff",
                "land"   = "#f5e6c8"),
    alpha = 0.5,
    border.col = "gray",
    border.lwd = 1,
    legend.show = FALSE
  ) +

# 2. Mangrove habitat polygons
tm_shape(hem_mangroves_only) +
  tm_polygons(
    col = "habitat",
    palette = c("Mangrove Habitat" = "#228B22"),   # dark green
    alpha = 1,
    border.col = "#228B22",
    border.lwd = 2,
    title = "Habitat"
  ) +

# 3. Layout + scale bar (no title)
tm_layout(
    frame = FALSE,
    legend.outside = TRUE
  ) +
tm_scale_bar(position = c("center", "bottom"))

```

```{r Mangrove area over time}
rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

# Clean color names to match cleaned labels
rgn_colors_clean <- setNames(
  rgn_colors,
  gsub("_", " ", names(rgn_colors))
)

mangrove_all_years <- MANG_area_gmw |>
  mutate(
    region_clean = gsub("_", " ", rgn_id)
  )

ggplot(
  mangrove_all_years,
  aes(
    x = year,
    y = area_km2,
    group = region_clean,
    color = region_clean
  )
) +
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = rgn_colors_clean) +
  labs(
    title = "Mangrove Extent Over Time (GMW 1996–2020)",
    x = "Year",
    y = "Mangrove Extent (km²)",
    color = "Region ID"
  ) +
  theme_minimal(base_size = 14)

```



