---
title: 'OHI GoC 2025: MPI/MPPI Exploration'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: inline
editor_options: 
  chunk_output_type: inline
---

# Summary

"The Mexico Peace Index (MPI) provides a comprehensive measure of peacefulness in Mexico, including trends, analysis and estimates of the economic impact of violence on the country. The MPI is based on the Global Peace Index, the world’s leading measure of global peacefulness, produced by Institute for Economics and Peace (IEP) every year since 2007.

This report also includes examples of the practical application of Positive Peace in Mexico at the national, state and local level.

Improving peacefulness in Mexico requires broader strategies that include addressing corruption and building effective institutions that are trusted by the public. In order to address elevated levels of violence, a holistic, integrated public security and peacebuilding framework is needed

The MPI report provides evidence for policy makers, business leaders and civil society organizations to help develop new and broader peacebuilding solutions for Mexico."

<https://www.visionofhumanity.org/maps/mexico-peace-index/#/>


```{r}
library(tidyverse)
library(here)
library(mapview)
library(sf)
library(readxl)
```


```{r}
mpi_raw <- readxl::read_xlsx(here::here("/home/shares/ohi/OHI_GOC/_raw_data/MPI/MPI_2025","MPI_PublicReleaseData_2025.xlsx"), sheet = 2, skip = 3, col_names = TRUE)
```

## Limit to GoC states

```{r}
# see the states
unique(mpi_raw$state)
#  [1] "Aguascalientes"      "Baja California"     "Baja California Sur" "Campeche"            "Chihuahua"          
#  [6] "Chiapas"             "Mexico City"         "Coahuila"            "Colima"              "Durango"            
# [11] "Guerrero"            "Guanajuato"          "Hidalgo"             "Jalisco"             "Mexico State"       
# [16] "Michoacán"           "Morelos"             NA                    "Nayarit"             "Nuevo León"         
# [21] "Oaxaca"              "Puebla"              "Querétaro"           "Quintana Roo"        "Sinaloa"            
# [26] "San Luis Potosí"     "Sonora"              "Tabasco"             "Tamaulipas"          "Tlaxcala"           
# [31] "Veracruz"            "Yucatán"             "Zacatecas"  

## read in OHI GoC regions
ohi_rgns <- st_read(here("spatial/ohi_regions","goc_ohi_rgns.shp"))
st_crs(ohi_rgns) # not WGS84: Mexico ITRF2008 / LCC
mapview(ohi_rgns)

mpi_goc <- mpi_raw %>% 
  filter(state %in% c("Baja California","Baja California Sur","Jalisco","Nayarit","Sinaloa","Sonora")) %>% 
  mutate(estado = case_when(
    state == "Baja California" ~ "Baja_California",
    state == "Baja California Sur" ~ "Baja_California_Sur",
    state == "Jalisco" ~ "Nayarit", ## small weirdness might have to address 
    state == "Nayarit" ~ "Nayarit",
    state == "Sinaloa" ~ "Sinaloa",
    state == "Sonora" ~ "Sonora",
    TRUE ~ state
  ))

# now join: expect a many-to-many relationship
mpi_goc_rgns <- full_join(ohi_rgns, mpi_goc, by = "estado")
```


```{r}
## let's see the indicator from the MPI: "fear of violence", for 2024.
fov_ohi_2024 <- mpi_goc_rgns %>% 
  filter(indicator %in% "fear of violence") %>% 
  filter(year %in% "2024") %>% 
  filter(type %in% "rate")

mapview(fov_ohi_2024, zcol = "value")
```


```{r}
## let's see organized crime.
oc_ohi_2024 <- mpi_goc_rgns %>% 
  filter(indicator %in% "organized crime") %>% 
  filter(year %in% "2024") 

mapview(oc_ohi_2024, zcol = "value")
```



