---
title: 'OHI GoC 2026: SS Safety + Security Current Status'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor:
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options:
  chunk_output_type: console
---

# Overview

In this script, we combine the **Safety** and **Security** current status scores into a single Safety and Security (SS) pressure score.

1. **Prep (loading packages)** We load the tidyverse and libraries used throughout the script.
2. **Read Safety current status** from SESNSP goal-level crime rates, scored with the **global 90th quantile** reference for now.
3. **Read Security current status** from ENSU March perception (`100 − pct_unsafe`).
4. **Simple average** the two statuses by region, year, and OHI goal.
5. **Summarize and plot** the combined SS status.

Safety status is goal-specific (SESNSP crimes mapped to each OHI goal).
Security status is not goal-specific (one ENSU city-safety perception score per region and year), so that same security score is reused for every goal when averaging.

\[
\text{SS current status} = \frac{\text{safety status (global 90th quantile)} + \text{security status}}{2}
\]

Scores are only computed for region-year (and goal) combinations where **both** components are available.
Security (ENSU March) currently starts in 2018, so the combined series is limited by that overlap with Safety (2015 through 2025).

Intermediate outputs are saved to `/home/shares/ohi/OHI_GOC/goal_prep/ss/ss_int`.
Final outputs are saved to `/home/shares/ohi/OHI_GOC/goal_prep/ss/ss_output`.

Upstream scripts:

- Safety goal-level rates and status alternatives: `ss_total/safety/ss_safety_goal_specific.qmd`
- Security ENSU current status: `ss_total/security/ensu_security.qmd`

# Background

## Safety (SESNSP)

Safety uses municipal crime incidence from the **Secretariado Ejecutivo del Sistema Nacional de Seguridad Publica** (SESNSP), standardized by population and scored by OHI goal.  Here we use the **global 90th quantile** scoring method explored in the safety script (section 6c):

- Reference crime rate = 90th percentile of `crime_rate` across **all** goals, regions, and years
- `safety_status = pmax(0, (1 - crime_rate / ref_global_q90) * 100)`
- Observations at or above the reference score 0 (scores are not allowed to go negative)

## Security (ENSU)

Security uses March rounds of the **Encuesta Nacional de Seguridad Pública Urbana** (ENSU). Region-level current status is:

\[
\text{security status} = 100 - \%\ \text{unsafe}
\]

where percent unsafe is the weighted share of adults who say living in their named city is *inseguro* (`BP1_1` = 2). Ideal is 100 (everyone feels safe).

## 1. Load packages

```{r}
#| label: step_01_load_packages

library(tidyverse) # data manipulation and visualization
library(magrittr) # pipe operators
library(here) # project relative paths
library(janitor) # clean column names
library(readr) # read csv files
library(scales) # axis helpers for plots

# intermediate and final folders for the combined SS score
ss_int <- "/home/shares/ohi/OHI_GOC/goal_prep/ss/ss_int"
ss_output <- "/home/shares/ohi/OHI_GOC/goal_prep/ss/ss_output"

# upstream safety and security products
safety_output <- "/home/shares/ohi/OHI_GOC/goal_prep/ss/safety_output"
security_output <- "/home/shares/ohi/OHI_GOC/goal_prep/ss/security_output"

# safety: crime rates by region, year, and OHI goal (from ss_safety_goal_specific.qmd)
rate_of_incidence_path <- file.path(
  safety_output,
  "rate_of_incidence_per_goal_2015_2025.csv"
)

# optional: precomputed alternatives table (if already written by safety script)
safety_status_alt_path <- file.path(
  safety_output,
  "current_status_alternatives_2015_2025.csv"
)

# security: ENSU March current status by region and year (from ensu_security.qmd)
cs_security_path <- file.path(security_output, "cs_security.csv")

# create output folders if missing (e.g. first run on a machine with the share mounted)
if (!dir.exists(ss_int)) dir.create(ss_int, recursive = TRUE)
if (!dir.exists(ss_output)) dir.create(ss_output, recursive = TRUE)
```

## 2. Read Safety data and score with global 90th quantile

We prefer recomputing `status_global_q90` from the crime-rate table so the method is explicit in this script.
If only the alternatives table is available, we can fall back to that column.

```{r}
#| label: step_02_safety_global_q90

if (file.exists(rate_of_incidence_path)) {
  rate_of_incidence_per_goal <- read_csv(here::here(rate_of_incidence_path)) %>%
    janitor::clean_names()

  # one reference across all goals, regions, and years
  ref_global_q90 <- quantile(
    rate_of_incidence_per_goal$crime_rate,
    probs = 0.90,
    na.rm = TRUE
  )

  ref_global_q90

  safety_status <- rate_of_incidence_per_goal %>%
    mutate(
      safety_status = pmax(0, (1 - (crime_rate / as.numeric(ref_global_q90))) * 100),
      method_safety = "Global 90th quantile"
    ) %>%
    dplyr::select(
      rgn_id, ano, ohi_goal,
      total_crimes_reported, population_interpolated, crime_rate,
      safety_status, method_safety
    )
} else if (file.exists(safety_status_alt_path)) {
  # fallback if the rate table is not on this machine but alternatives were saved
  message("rate table missing; using status_global_q90 from alternatives CSV")

  safety_status <- read_csv(here::here(safety_status_alt_path)) %>%
    janitor::clean_names() %>%
    mutate(
      safety_status = status_global_q90,
      method_safety = "Global 90th quantile"
    ) %>%
    dplyr::select(
      rgn_id, ano, ohi_goal,
      dplyr::any_of(c(
        "total_crimes_reported", "population_interpolated", "crime_rate"
      )),
      safety_status, method_safety
    )
} else {
  stop(
    "Could not find Safety inputs.\n",
    "Expected either:\n  ", rate_of_incidence_path, "\n",
    "or:\n  ", safety_status_alt_path
  )
}

summary(safety_status$safety_status)
unique(safety_status$ano)
unique(safety_status$rgn_id)
sort(unique(safety_status$ohi_goal))

write_csv(
  safety_status,
  here::here(ss_int, "safety_status_global_q90_2015_2025.csv")
)
```

## 3. Read Security current status (ENSU)

```{r}
#| label: step_03_read_security

if (!file.exists(cs_security_path)) {
  stop(
    "Could not find Security current status at:\n  ", cs_security_path, "\n",
    "Run ss_total/security/ensu_security.qmd first."
  )
}

cs_security <- read_csv(here::here(cs_security_path)) %>%
  janitor::clean_names()

# harmonize year column name for joining with safety (ano)
security_status <- cs_security %>%
  mutate(
    ano = as.integer(year_march),
    security_status = current_status,
    method_security = "100 - pct_unsafe (ENSU March)"
  ) %>%
  dplyr::select(
    rgn_id, ano, rgn_label,
    dplyr::any_of(c("n_cities", "n_substantive", "pct_unsafe")),
    security_status, method_security
  )

summary(security_status$security_status)
unique(security_status$ano)
unique(security_status$rgn_id)

write_csv(
  security_status,
  here::here(ss_int, "security_status_ensu_march.csv")
)
```

## 4. Simple average of Safety and Security status

Join on `rgn_id` and `ano`. Security has no `ohi_goal` dimension, so its region-year score is joined onto every goal. The combined status is the unweighted mean of the two components where both are non-missing.

```{r}
#| label: step_04_average_ss

ss_current_status <- safety_status %>%
  left_join(security_status, by = c("rgn_id", "ano")) %>%
  mutate(
    ss_current_status = dplyr::if_else(
      !is.na(safety_status) & !is.na(security_status),
      (safety_status + security_status) / 2,
      NA_real_
    ),
    has_both = !is.na(ss_current_status)
  )

# # how many rows have both components vs safety-only
# ss_coverage <- ss_current_status %>%
#   summarise(
#     n_rows = dplyr::n(),
#     n_with_both = sum(has_both, na.rm = TRUE),
#     n_safety_only = sum(!is.na(safety_status) & is.na(security_status)),
#     n_missing_safety = sum(is.na(safety_status)),
#     years_with_both = paste(sort(unique(ano[has_both])), collapse = ", "),
#     years_safety = paste(sort(unique(ano[!is.na(safety_status)])), collapse = ", "),
#     years_security = paste(sort(unique(security_status$ano)), collapse = ", ")
#   )
# 
# ss_coverage

# final ss: keep only complete safety + security nonNAs for the SS score
ss_current_status_complete <- ss_current_status %>%
  filter(has_both) %>%
  dplyr::select(
    rgn_id, ano, ohi_goal,
    dplyr::any_of(c(
      "rgn_label", "total_crimes_reported", "population_interpolated",
      "crime_rate", "pct_unsafe", "n_cities", "n_substantive"
    )),
    safety_status, security_status, ss_current_status,
    method_safety, method_security
  ) %>%
  arrange(ohi_goal, rgn_id, ano)

summary(ss_current_status_complete$ss_current_status)
unique(ss_current_status_complete$ano)
unique(ss_current_status_complete$rgn_id) # rgns 2, 3, and 5 are missing
sort(unique(ss_current_status_complete$ohi_goal))

write_csv(
  ss_current_status,
  here::here(ss_output, "ss_current_status_full_join_2015_2025.csv")
)

write_csv(
  ss_current_status_complete,
  here::here(ss_output, "ss_current_status_safety_security_avg.csv")
)
```

### 4a. Summaries by goal and overall

```{r}
#| label: step_04a_summaries

# summary of combined SS status by goal
ss_summary_by_goal <- ss_current_status_complete %>%
  group_by(ohi_goal, ano) %>%
  summarise(
    n = dplyr::n(),
    min = min(ss_current_status, na.rm = TRUE),
    q1 = quantile(ss_current_status, 0.25, na.rm = TRUE),
    median = median(ss_current_status, na.rm = TRUE),
    mean = mean(ss_current_status, na.rm = TRUE),
    q3 = quantile(ss_current_status, 0.75, na.rm = TRUE),
    max = max(ss_current_status, na.rm = TRUE),
    mean_safety = mean(safety_status, na.rm = TRUE),
    mean_security = mean(security_status, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(ohi_goal)

ss_summary_by_goal
kableExtra::kable(ss_summary_by_goal)

# overall summary comparing each layer
ss_summary_overall <- ss_current_status_complete %>%
  tidyr::pivot_longer(
    cols = c(safety_status, security_status, ss_current_status),
    names_to = "component",
    values_to = "status"
  ) %>%
  mutate(
    component = dplyr::recode(
      component,
      safety_status = "Safety (global q90)",
      security_status = "Security (ENSU)",
      ss_current_status = "SS average"
    ),
    component = factor(
      component,
      levels = c("Safety (global q90)", "Security (ENSU)", "SS average")
    )
  ) %>%
  group_by(component) %>%
  summarise(
    n = dplyr::n(),
    min = min(status, na.rm = TRUE),
    q1 = quantile(status, 0.25, na.rm = TRUE),
    median = median(status, na.rm = TRUE),
    mean = mean(status, na.rm = TRUE),
    q3 = quantile(status, 0.75, na.rm = TRUE),
    max = max(status, na.rm = TRUE),
    .groups = "drop"
  )

ss_summary_overall

write_csv(
  ss_summary_by_goal,
  here::here(ss_output, "ss_current_status_summary_by_goal.csv")
)

write_csv(
  ss_summary_overall,
  here::here(ss_output, "ss_current_status_summary_overall.csv")
)
```

## 5. Plots

Region labels and colors match the safety script so maps and line plots stay consistent.

```{r}
#| label: step_05_plot_setup
#| fig-width: 8
#| fig-height: 4.5
#| fig-dpi: 96

if (requireNamespace("showtext", quietly = TRUE)) {
  showtext::showtext_auto(FALSE)
}

rgn_labels <- c(
  "Region_1" = "Western Upper Gulf",
  "Region_2" = "Western Midriff Islands",
  "Region_3" = "Western Central Gulf",
  "Region_4" = "Western Southern Gulf",
  "Region_5" = "Eastern Upper Gulf",
  "Region_6" = "Eastern Midriff Islands",
  "Region_7" = "Southern Sonora Coast",
  "Region_8" = "Sinaloa Coast",
  "Region_9" = "Southeastern Gulf"
)

rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

theme_status <- theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11, color = "grey40"),
    axis.title = element_text(size = 11),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10),
    legend.position = "bottom",
    legend.key.size = unit(0.5, "cm"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(6, 8, 4, 6)
  )

ss_plot_data <- ss_current_status_complete %>%
  mutate(
    ano = as.integer(ano),
    rgn_id = factor(as.character(rgn_id), levels = names(rgn_labels)),
    ohi_goal = as.character(ohi_goal)
  )

goal_list <- sort(unique(ss_plot_data$ohi_goal))
```

### 5a. Component comparison 

```{r}
#| label: step_05a_component_box
#| fig-width: 9
#| fig-height: 5
#| fig-dpi: 96

status_long_components <- ss_plot_data %>%
  tidyr::pivot_longer(
    cols = c(safety_status, security_status, ss_current_status),
    names_to = "component",
    values_to = "status"
  ) %>%
  mutate(
    component = dplyr::recode(
      component,
      safety_status = "Safety (global q90)",
      security_status = "Security (ENSU)",
      ss_current_status = "SS average"
    ),
    component = factor(
      component,
      levels = c("Safety (global q90)", "Security (ENSU)", "SS average")
    )
  )

p_box_components <- ggplot(
  status_long_components,
  aes(x = component, y = status, fill = component)
) +
  geom_boxplot(outlier.size = 0.8, alpha = 0.8) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 25)) +
  labs(
    title = "Status distributions: Safety, Security, and SS average",
    subtitle = "Region-year-goal rows where both Safety and Security are available",
    x = NULL,
    y = "Status (100 = best)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 15, hjust = 1),
    plot.title = element_text(face = "bold")
  )

p_box_components

# ggsave(
#   here::here(ss_output, "plot_ss_component_boxplot.png"),
#   p_box_components, width = 9, height = 5, dpi = 300
# )
```

### 5b. SS current status by goal

One plot per OHI goal, with region trajectories over years.
Plots are stored in `p_ss_status_list` so any goal can be called individually.

```{r}
#| label: step_05b_status_plots
#| fig-width: 8
#| fig-height: 4.5
#| fig-dpi: 96

p_ss_status_list <- list()

for (g in goal_list) {
  goal_title <- stringr::str_replace_all(g, "_", " ") %>%
    stringr::str_to_title()

  p_ss_status_list[[g]] <- ss_plot_data %>%
    filter(ohi_goal == g) %>%
    ggplot(aes(x = ano, y = ss_current_status, color = rgn_id, group = rgn_id)) +
    geom_line(linewidth = 1.1, alpha = 0.9) +
    geom_point(size = 2) +
    scale_color_manual(values = rgn_colors, labels = rgn_labels) +
    scale_x_continuous(breaks = pretty_breaks(n = 6)) +
    scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 25)) +
    labs(
      title = goal_title,
      subtitle = "SS status = mean(Safety global q90, Security ENSU); 100 = best",
      x = NULL,
      y = "SS status",
      color = NULL
    ) +
    guides(color = guide_legend(nrow = 3, byrow = TRUE)) +
    theme_status

  # ggsave(
  #   here::here(ss_output, paste0("plot_ss_status_", g, ".png")),
  #   p_ss_status_list[[g]], width = 7, height = 4, dpi = 300
  # )
}

names(p_ss_status_list)
```

#### Call individual SS status plots by goal

```{r}
#| label: step_05b_print_status_by_goal
#| fig-width: 8
#| fig-height: 4.5
#| fig-dpi: 96

p_ss_status_list[["biodiversity_habitats"]]
p_ss_status_list[["biodiversity_species"]]
p_ss_status_list[["clean_waters"]]
p_ss_status_list[["coastal_opportunities_artisanal_fishing_opportunities"]]
p_ss_status_list[["coastal_opportunities_coastal_access"]]
p_ss_status_list[["food_provisioning_fisheries"]]
p_ss_status_list[["food_provisioning_mariculture"]]
p_ss_status_list[["habitat_services_carbon_storage"]]
p_ss_status_list[["habitat_services_coastal_protection"]]
p_ss_status_list[["habitat_services_filtration"]]
p_ss_status_list[["livelihoods_economies_economies"]]
p_ss_status_list[["livelihoods_economies_livelihoods"]]
p_ss_status_list[["natural_products"]]
p_ss_status_list[["research_ocean_literacy"]]
p_ss_status_list[["sentido_de_pertenencia"]]
p_ss_status_list[["tourism_recreation"]]
```

### 5c. Safety vs Security vs SS average by goal

Faceted comparison: the two inputs and the average, for one goal at a time.

```{r}
#| label: step_05c_compare_by_goal
#| fig-width: 10
#| fig-height: 7
#| fig-dpi: 96

p_ss_compare_list <- list()

for (g in goal_list) {
  goal_title <- stringr::str_replace_all(g, "_", " ") %>%
    stringr::str_to_title()

  p_ss_compare_list[[g]] <- status_long_components %>%
    filter(ohi_goal == g) %>%
    mutate(rgn_id = factor(as.character(rgn_id), levels = names(rgn_labels))) %>%
    ggplot(aes(x = ano, y = status, color = rgn_id, group = rgn_id)) +
    geom_line(linewidth = 0.9, alpha = 0.9) +
    geom_point(size = 1.5) +
    facet_wrap(~ component, ncol = 3) +
    scale_color_manual(values = rgn_colors, labels = rgn_labels) +
    scale_x_continuous(breaks = pretty_breaks(n = 5)) +
    scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 25)) +
    labs(
      title = paste0(goal_title, ": Safety, Security, and SS average"),
      x = NULL,
      y = "Status",
      color = NULL
    ) +
    guides(color = guide_legend(nrow = 3, byrow = TRUE)) +
    theme_minimal(base_size = 11) +
    theme(
      legend.position = "bottom",
      plot.title = element_text(face = "bold"),
      strip.text = element_text(face = "bold", size = 9)
    )

  # ggsave(
  #   here::here(ss_output, paste0("plot_ss_compare_", g, ".png")),
  #   p_ss_compare_list[[g]], width = 10, height = 7, dpi = 300
  # )
}

names(p_ss_compare_list)
```

#### Call individual comparison plots by goal

```{r}
#| label: step_05c_print_compare_by_goal
#| fig-width: 10
#| fig-height: 7
#| fig-dpi: 96

p_ss_compare_list[["biodiversity_habitats"]]
p_ss_compare_list[["biodiversity_species"]]
p_ss_compare_list[["clean_waters"]]
p_ss_compare_list[["coastal_opportunities_artisanal_fishing_opportunities"]]
p_ss_compare_list[["coastal_opportunities_coastal_access"]]
p_ss_compare_list[["food_provisioning_fisheries"]]
p_ss_compare_list[["food_provisioning_mariculture"]]
p_ss_compare_list[["habitat_services_carbon_storage"]]
p_ss_compare_list[["habitat_services_coastal_protection"]]
p_ss_compare_list[["habitat_services_filtration"]]
p_ss_compare_list[["livelihoods_economies_economies"]]
p_ss_compare_list[["livelihoods_economies_livelihoods"]]
p_ss_compare_list[["natural_products"]]
p_ss_compare_list[["research_ocean_literacy"]]
p_ss_compare_list[["sentido_de_pertenencia"]]
p_ss_compare_list[["tourism_recreation"]]
```

# Outputs

| File | Description |
|:---|:---|
| `ss_int/safety_status_global_q90_2015_2025.csv` | Safety status using global 90th quantile of crime rate |
| `ss_int/security_status_ensu_march.csv` | Security status from ENSU March (`100 − pct_unsafe`) |
| `ss_output/ss_current_status_full_join_2015_2025.csv` | Full left join of safety onto security (includes safety-only years) |
| `ss_output/ss_current_status_safety_security_avg.csv` | **Main product:** simple average where both components exist |
| `ss_output/ss_current_status_summary_by_goal.csv` | Summary stats of SS status by OHI goal |
| `ss_output/ss_current_status_summary_overall.csv` | Summary stats for Safety, Security, and SS average |

Main score:

```text
ss_current_status = (safety_status_global_q90 + security_status) / 2
```
