################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```


# STEP 1: Load input data ------------------------------

```{r Load in FIS_StockStatus scores as necessary}
StockStatus_Scores0<- read.csv("/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_StockStatusScores_GeoMean_Jun012026.csv")

# Species x year x region; Stock_Status is used as-is.
```

```{r Load sector-sorted landings}
GOC_MarineLandings0<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Marine.Landings_Sector.Sorted.Georef_PATCHED_Jun032026.rds"))

dim(GOC_MarineLandings0)
# 3246583 obs., 57 vars.

# NOTE TO SELF: This file path will change when I create the "fisheries only" dataframe. - June 1 2026
```


# STEP 2: Determine which fisheries are primarily small scale ------------------------------

```{r Identify stocks with >= 15% small-scale capture}
SSF_CATCH_MIN_SHARE = 0.15
```

```{r Let's look into Mar's key}
Mars_key<- read.csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Keys/CONAPESCA_Keys/Taxonomic_Keys/Mar_CONAPESCA_SpKey.csv")
```

# **DECISION POINT: Selecting which sectors to include**
```{r Select Sectors_of_Interest}
table(GOC_MarineLandings0$sector_estimate_draft)

MarineFishery_Sectors<- c("marine: fishery sector yet to be classified",
                          "marine: small-scale capture",
                          "marine: large-scale capture")
  
PossibleMarineFishery_Sectors<- c("freshwater or marine: large-scale capture",
                          "freshwater or marine: small-scale capture",
                          "freshwater or marine: fishery sector yet to be classified")

GOC_MarineFisheryLandings0<- GOC_MarineLandings0 %>%
  filter(sector_estimate_draft %in% c(MarineFishery_Sectors, PossibleMarineFishery_Sectors))

# 3137044 obs., 58 var.
```

```{r Generate GOC_MarineSector_summary}
colnames(GOC_MarineFisheryLandings0)

GOC_MarineSector_summary<- GOC_MarineFisheryLandings0 %>%
  select(year, 
         state, office_name_clean, landing_site_clean,
         sector_estimate_draft,
         GOC_est, GOC_Include_Exclude, OHI_region_est, OHI_region_name,
         eu_rnpa, eu_name, sector_estimate_draft, 
         Major_Functional_Group, recorded_taxa_group, estimated_taxa_group,
         recorded_taxa_name, recorded_taxa_name_clean,
         estimated_species_name, estimated_genus_name,
         landed_weight, live_weight, weight_kg_corrected, value
         )

GOC_MarineSector_summary<- GOC_MarineSector_summary %>%
  group_by(year, sector_estimate_draft) %>%
  summarize(
    total_weight_mt= sum(weight_kg_corrected, na.rm = TRUE) / 1000,
    total_value_peso= sum(value, na.rm = TRUE),
    n_records= n()
  )
```


## Plotting sector estimates:
```{r Plot each sector of interest over time}
# SECTORS OF INTEREST:
Sectors_of_Interest<- c("marine: fishery sector yet to be classified" , "marine: small-scale capture", "marine: large-scale capture" , "marine: aquaculture", "freshwater: aquaculture")
  
# Define colors for each sector (fleet distinction = lighter for small-scale)
sector_colors <- c(
  # Marine (blues)
  "marine: fishery sector yet to be classified"       = "#cce0ff",  # very light blue for unknown
  "marine: small-scale capture"                       = "#6699cc",  # medium blue
  "marine: large-scale capture"                       = "#336699",  # darker blue
  "marine: aquaculture"                               = "#33cccc",  # teal/aquaculture cue
  
  # Freshwater (warm shades)
  "freshwater: aquaculture"                          = "#ff9933"  # bright orange
  # "freshwater: fishery sector yet to be classified" = "#ffe0cc",  # light orange for unknown - IGNORE THIS SECTOR FOR NOW.
  # "freshwater: small-scale capture"                  = "#ff6666",  # strong red/orange - IGNORE THIS SECTOR FOR NOW.
  # "freshwater: large-scale capture"                  = "#ffb366",  # lighter orange/yellow - IGNORE THIS SECTOR FOR NOW.
  
  # Freshwater or Marine (grey and purple shades) - IGNORE FOR NOW.
  # "freshwater or marine: fishery sector yet to be classified"  = "#e0e0e0",  # light grey
  # "freshwater or marine: small-scale capture"                    = "#b399cc",  # light purple
  # "freshwater or marine: large-scale capture"                    = "#8040a0",  # medium purple
  # "freshwater or marine: aquaculture"                             = "#999999"   # neutral grey
)

# Plot
Landings_byEstimatedSector_plot<- GOC_MarineSector_summary %>%
 # filter(year >= 2018 & year <= 2024) %>%
  filter(sector_estimate_draft %in% Sectors_of_Interest) %>%
  ggplot(aes(x = year, y = total_weight_mt, fill = sector_estimate_draft)) +
  
  # Area plot with slight border for separation
 # geom_area(alpha = 0.9, color = "white", linewidth = 0.2) +
  
    geom_col(
    position = position_dodge(width = 0.8),
    width = 0.7,
    color = "white"
  ) +
  
  # Manual fill palette
  scale_fill_manual(
    values = sector_colors,
    name = "Sector"
  ) +
  
  # Labels
  labs(
    x = "Year",
    y = "Est Annual Catch (metric tons)",
    title = "Catch by Sector (GOC)",
    subtitle = "Small-scale vs. large-scale estimates are based directly on report details"
  ) +
  
  # Minimal theme and sleek styling
  theme_bw(base_size = 14) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey85"),
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 13, color = "grey30")
  ) +
  scale_x_continuous(
  breaks = 2000:2025) +
  
  # Optional: better legend layout
  guides(
    fill = guide_legend(nrow = 3, byrow = TRUE, title.position = "top")
  )

Landings_byEstimatedSector_plot
```

```{r Save plot}
ggsave(
  "/home/ferrer/ohi_goc_prep/AO/FisheryLandings_bySector_Jun042026.png",
  plot = Landings_byEstimatedSector_plot,
  width = 10,
  height = 6,
  dpi = 300
)
```


```{r Lets only look at landings post-2018}
Landings_Post.2018<- GOC_MarineFisheryLandings0 %>%
  filter(year>= 2018)

table(Landings_Post.2018$sector_reported)
# sector details were not reported                      aquaculture            small-scale fisheries 
#                             4914                                0                          1010533 
#            large-scale fisheries 
#                           142210 

table(Landings_Post.2018$sector_estimate_draft)
              # marine: fishery sector yet to be classified 
              #                                        4884 
              #                 marine: large-scale capture 
              #                                      141145 
              #                 marine: small-scale capture 
              #                                      977477
#                 freshwater or marine: small-scale capture 
#                                                     33056 
#                 freshwater or marine: large-scale capture 
#                                                      1065 
# freshwater or marine: fishery sector yet to be classified 
#                                                        30
```



```{r Small-scale fishery sector helper}
# Marine and freshwater-or-marine small-scale capture only.

is_small_scale_fishery_sector <- function(sector) {
  grepl("marine | freshwater or marine", sector, ignore.case = TRUE) &
    grepl("small-scale capture", sector, ignore.case = TRUE)
}
```


```{r Filter landings and assign stock key}
GOC_Landings <- GOC_Landings0 %>%
  filter(
    GOC_Include_Exclude == "include",
    year >= 2018,
    year <= 2024,
    !is.na(OHI_region_est),
    OHI_region_est != ""
  ) %>%
  mutate(
    OHI_region_est = clean_names_lowercase(OHI_region_est),
    estimated_species_name = clean_names_lowercase(estimated_species_name),
    estimated_genus_name = clean_names_lowercase(estimated_genus_name),
    sector_estimate_draft = as.character(sector_estimate_draft),
    small_scale_capture = is_small_scale_fishery_sector(sector_estimate_draft),
    
    taxon_key = if_else(
      is.na(estimated_species_name) | estimated_species_name == "",
      estimated_genus_name,
      estimated_species_name
    ),
    taxon_level = if_else(
      is.na(estimated_species_name) | estimated_species_name == "",
      "genus",
      "species"
    )
  ) %>%
  filter(!is.na(taxon_key), taxon_key != "")
```

```{r Summarize total and small-scale catch per stock}
stock_catch_summary <- GOC_Landings %>%
  group_by(year, OHI_region_est, taxon_key, taxon_level, estimated_genus_name) %>%
  summarize(
    total_landings_kg = sum(weight_kg_corrected, na.rm = TRUE),
    small_scale_landings_kg = sum(
      weight_kg_corrected[small_scale_capture],
      na.rm = TRUE
    ),
    n_small_scale_records = sum(small_scale_capture, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    small_scale_catch_share = small_scale_landings_kg / total_landings_kg
  )
```

```{r Keep stocks with at least 15% small-scale catch}
AFO_stocks <- stock_catch_summary %>%
  filter(
    small_scale_landings_kg > 0,
    small_scale_catch_share >= SSF_CATCH_MIN_SHARE
  )
# Stocks meeting the small-scale catch threshold in this region x year.
```

```{r Count qualifying stocks}
n_stock_region_years <- nrow(AFO_stocks)

n_unique_stocks <- AFO_stocks %>%
  distinct(taxon_key, taxon_level) %>%
  nrow()

n_unique_species <- AFO_stocks %>%
  filter(taxon_level == "species") %>%
  distinct(taxon_key) %>%
  nrow()

n_unique_genus_stocks <- AFO_stocks %>%
  filter(taxon_level == "genus") %>%
  distinct(taxon_key) %>%
  nrow()

cat(
  "Stocks (stock x region x year) with >= 15% small-scale catch:",
  n_stock_region_years,
  "\n"
)
cat("Unique stocks (taxon_key) meeting threshold:", n_unique_stocks, "\n")
cat("  species-level:", n_unique_species, "\n")
cat("  genus-level:", n_unique_genus_stocks, "\n")
```


# STEP 3: Pull matching FIS scores (no recalculation) ------------------------------

```{r Prep FIS scores}
StockStatus_Scores <- StockStatus_Scores0 %>%
  filter(year >= 2018, year <= 2024) %>%
  mutate(
    OHI_region_est = clean_names_lowercase(OHI_region_est),
    estimated_species_name = clean_names_lowercase(estimated_species_name),
    estimated_genus_name = clean_names_lowercase(estimated_genus_name)
  )
```

```{r Genus-level stocks: pull all FIS stocks in that genus}
afo_genus_keys <- AFO_stocks %>%
  filter(taxon_level == "genus") %>%
  select(year, OHI_region_est, taxon_key, taxon_level, estimated_genus_name) %>%
  distinct()

FIS_scores_afo_genus <- StockStatus_Scores %>%
  inner_join(
    afo_genus_keys,
    by = c(
      "year",
      "OHI_region_est",
      "estimated_genus_name" = "taxon_key"
    )
  ) %>%
  left_join(
    AFO_stocks %>%
      filter(taxon_level == "genus") %>%
      select(
        year,
        OHI_region_est,
        taxon_key,
        total_landings_kg,
        small_scale_landings_kg,
        small_scale_catch_share,
        n_small_scale_records
      ),
    by = c(
      "year",
      "OHI_region_est",
      "estimated_genus_name" = "taxon_key"
    )
  ) %>%
  mutate(taxon_level = "genus")
```

```{r Sanity check AFO FIS scores}
colnames(GOC_AFO_Scores)

GOC_AFO_Scores %>%
  count(taxon_level)

nrow(GOC_AFO_Scores)

summary(GOC_AFO_Scores$small_scale_catch_share)
```


# STEP 4: Stock list by OHI region ------------------------------

```{r Region totals for share calculations}
region_catch_totals <- GOC_AFO_Scores %>%
  group_by(OHI_region_est) %>%
  summarize(
    region_total_landings_kg = sum(total_landings_kg, na.rm = TRUE),
    region_small_scale_landings_kg = sum(small_scale_landings_kg, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r Build stock list organized by region}
# One row per stock x region (2018-2024 summed). Sorted region_1..region_9,
# then by small-scale catch share within region.

GOC_AFO_Stock_List <- GOC_AFO_Scores %>%
  group_by(
    OHI_region_est,
    afo_stock_key,
    taxon_level,
    estimated_species_name,
    estimated_genus_name
  ) %>%
  summarize(
    years_present = paste(sort(unique(year)), collapse = ", "),
    n_years = n_distinct(year),
    mean_Stock_Status = mean(Stock_Status, na.rm = TRUE),
    total_landings_kg = sum(total_landings_kg, na.rm = TRUE),
    small_scale_landings_kg = sum(small_scale_landings_kg, na.rm = TRUE),
    small_scale_catch_share = small_scale_landings_kg / total_landings_kg,
    .groups = "drop"
  ) %>%
  left_join(region_catch_totals, by = "OHI_region_est") %>%
  mutate(
    pct_of_region_total_catch = 100 * total_landings_kg / region_total_landings_kg,
    pct_of_region_small_scale_catch = 100 * small_scale_landings_kg /
      region_small_scale_landings_kg,
    region_num = as.integer(sub("region_", "", OHI_region_est))
  ) %>%
  arrange(region_num, desc(small_scale_landings_kg), estimated_species_name) %>%
  select(
    OHI_region_est,
    region_num,
    afo_stock_key,
    taxon_level,
    estimated_species_name,
    estimated_genus_name,
    years_present,
    n_years,
    mean_Stock_Status,
    total_landings_kg,
    small_scale_landings_kg,
    small_scale_catch_share,
    pct_of_region_total_catch,
    pct_of_region_small_scale_catch,
    region_total_landings_kg,
    region_small_scale_landings_kg
  )
```

```{r Sanity check AFO stock list}
nrow(GOC_AFO_Stock_List)

GOC_AFO_Stock_List %>%
  count(OHI_region_est)

GOC_AFO_Stock_List %>%
  group_by(OHI_region_est) %>%
  summarize(
    check_total_pct = sum(pct_of_region_total_catch),
    check_ssf_pct = sum(pct_of_region_small_scale_catch),
    .groups = "drop"
  )
```


# STEP 5: Small-scale catch-weighted AFO_score per OHI region ------------------------------

```{r Small-scale catch-weighted AFO_score per region and year}
# AFO_score = sum(Stock_Status * small_scale_landings_kg) /
#             sum(small_scale_landings_kg)
# Weights are small-scale fishery landings only.

GOC_AFO_Regional_Scores <- GOC_AFO_Scores %>%
  filter(
    !is.na(Stock_Status),
    small_scale_landings_kg > 0
  ) %>%
  group_by(year, OHI_region_est) %>%
  summarize(
    AFO_score = sum(Stock_Status * small_scale_landings_kg) /
      sum(small_scale_landings_kg),
    n_stocks = n(),
    small_scale_catch_kg = sum(small_scale_landings_kg),
    total_catch_kg = sum(total_landings_kg),
    .groups = "drop"
  ) %>%
  arrange(year, OHI_region_est)
```

```{r Sanity check AFO regional scores}
colnames(GOC_AFO_Regional_Scores)

summary(GOC_AFO_Regional_Scores$AFO_score)

GOC_AFO_Regional_Scores %>%
  arrange(year, OHI_region_est)
```

# STEP 6: Generate relevant time series including total_catch per region, total_SSF_catch per region, prop_SSF_catch per region

```{r}
Help from cursor
```

# STEP 7: Plot regional AFO scores over time ------------------------------

```{r Region colors for plot}
region_colors <- c(
  "region_1" = "royalblue",
  "region_2" = "skyblue",
  "region_3" = "orange",
  "region_4" = "sienna1",
  "region_5" = "plum",
  "region_6" = "purple4",
  "region_7" = "aquamarine3",
  "region_8" = "gold2",
  "region_9" = "tomato"
)
```

```{r Plot AFO_score over time by region}
GOC_AFO_Regional_Scores_overTime <- GOC_AFO_Regional_Scores %>%
  ggplot(aes(
    x = year,
    y = AFO_score,
    color = OHI_region_est
  )) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ OHI_region_est, scales = "fixed") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = region_colors) +
  labs(
    title = "GOC AFO: Regional catch-weighted Stock_Status scores",
    subtitle = paste0(
      ">= ",
      100 * SSF_CATCH_MIN_SHARE,
      "% small-scale capture stocks; weighted by small-scale landings (kg)"
    ),
    y = "AFO_score (0 to 1)",
    x = "Year"
  ) +
  ylim(0, 1.05) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 12, face = "bold")
  )

GOC_AFO_Regional_Scores_overTime
```


# FINAL STEP: SAVE ------------------------------

```{r SAVE GOC_AFO_Scores and stock-level detail}
# Stock-level detail (species x year x region) for downstream use.
GOC_AFO_Stocks <- GOC_AFO_Scores

write.csv(
  GOC_AFO_Scores,
  file = file.path(WORK_DIR, "GOC_AFO_Scores_Jun012026.csv"),
  row.names = FALSE
)

write.csv(
  GOC_AFO_Stocks,
  file = file.path(WORK_DIR, "GOC_AFO_Stocks_Jun012026.csv"),
  row.names = FALSE
)
```

```{r SAVE stock list and regional AFO scores}
write.csv(
  GOC_AFO_Stock_List,
  file = file.path(WORK_DIR, "GOC_AFO_Stock_List_Jun012026.csv"),
  row.names = FALSE
)

write.csv(
  GOC_AFO_Regional_Scores,
  file = file.path(WORK_DIR, "GOC_AFO_Regional_Scores_Jun012026.csv"),
  row.names = FALSE
)
```

```{r SAVE regional AFO score plot}
ggsave(
  filename = file.path(WORK_DIR, "GOC_AFO_Regional_Scores_Jun012026.png"),
  plot = GOC_AFO_Regional_Scores_overTime,
  width = 12,
  height = 8,
  dpi = 300
)
```

