################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```

# Region Key (December 2025) ------------------------------
```{r Region Key from Mel}
BajaSonoraKey<- read_csv(here("/home/frazier/ohi_goc_prep/FP/FIS/v2025/data/Baja_Sonora_states_key.csv"))

BajaSonoraKey <- BajaSonoraKey %>%
  mutate(
    long = as.numeric(long),
    lat  = as.numeric(lat)
  )

JaliscoKey<- read_csv(here("/home/frazier/ohi_goc_prep/FP/FIS/v2025/data/landing_sites_jalisco.csv"))

JaliscoKey <- JaliscoKey %>%
  rename(long= lon) %>% 
  mutate(
    long = as.numeric(long),
    lat  = as.numeric(lat)
  )

JaliscoKey_clean <- JaliscoKey %>%
  # If you want to normalize state names first
  mutate(state = toupper(state)) %>%
  
  # Remove any rows where landing_site is NA (can cause matching issues later)
  filter(!is.na(landing_site)) %>%
  
  # Keep only one row per landing_site_key; if duplicate keys exist, keep first
  group_by(landing_site_key) %>%
  slice(1) %>%  # keeps the first occurrence per key
  ungroup() %>%
  
  # Optional: also deduplicate by landing_site if there are rows with NA landing_site_key
  group_by(landing_site) %>%
  slice(1) %>%
  ungroup()

# NOTE TO SELF: Come back to this! The Jalisco key is imperfect, but we'll have to fix this in the new year.
# Identify rows that were removed in the deduplication of JaliscoKey
removed_rows <- anti_join(JaliscoKey, JaliscoKey_clean, 
                          by = c("landing_site_key", "landing_site"))


#### MERGE THESE WORKING KEYS FOR NOW: ####
RegionKey_working <- bind_rows(BajaSonoraKey, JaliscoKey_clean) %>%
  filter(!is.na(landing_site)) %>%  # remove the NA from this dataframe as this will create a many-to-many matching issue later on
  mutate(long = as.numeric(long))
# 1176 obs., 9 vars.

RegionKey_working <- RegionKey_working %>% mutate(long = as.numeric(long))

# My regional key - NOTE SELF
RegionKey_temporary<- read_csv(here("/home/ferrer/ohi_goc_prep/general/CONAPESCA_Region_Keys/RegionKey_UNDERCONSTRUCTION_Dec02024.csv"))

rm(JaliscoKey, removed_rows)
```

# Load: Landings data with species info (December 2025) ------------------------------
```{r Load in landings data with species info}
Landings_GoC_Cleaned_wSp<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fis/v2025/int/juan_carlos/GoC_Landings_wSp_2000.present_Dec012025.rds"))
# 3716415 obs, 41 vars.

Landings_GoC_Cleaned_wSp <- Landings_GoC_Cleaned_wSp %>% 
  select(-lat, -long) # remove lat, long for now to do the merge

Landings_GoC <- Landings_GoC_Cleaned_wSp %>%
  left_join(
    RegionKey_working %>%
      select(-landing_site_key),
    by = c("state", "landing_site")
  )
# 3716415 obs, 45 vars.

# A few quick checks:
table(Landings_GoC$GoC)
# All in the GoC or not otherwise noted.
table(Landings_GoC$state)
table(Landings_GoC_Cleaned_wSp$state)
table(Landings_GoC$OHI_rgn)
# Excellent, completely keyed for now. Erica - Dec 9 2025
```

```{r Add OHI region to landings w species}
Landings_GoC <- Landings_GoC %>%
  mutate(Region_pick = if_else(!is.na(OHI_rgn), OHI_rgn, state))
# 3716415 obs, 46 vars. as expected!

Landings_GoC <- Landings_GoC %>%
  left_join(
    RegionKey_temporary %>%
      select(-Region_Name_Proposed),  # remove this column for now
    by = c("Region_pick" = "Region_Mel_or_CONAPESCA")
  )
# 3716415 obs, 47 vars. as expected!

subset<- Landings_GoC %>% filter(is.na(OHI_Region_No))
table(subset$state)
# 156952 obs., 47 vars.
# Only Jalisco entries are not keyed. This makes sense. These are parts of Jalisco that either aren't in Region 9 or we can't be totally sure they are in Region 9.

Landings_GoC<- Landings_GoC %>% filter(!is.na(OHI_Region_No)) 
# 3559463 obs., 47 vars.
```

```{r Landed weight correction}
# Kind of looks like sometimes 'landed weight' is 'live weight' and vice versa. Fix this:
Landings_GoC$weight_kg_corrected <- ifelse(
  Landings_GoC$live_weight >= Landings_GoC$landed_weight,
  Landings_GoC$live_weight,
  Landings_GoC$landed_weight
)
```

```{r Filtering out things we don't care about}
colnames(Landings_GoC)
table(Landings_GoC$acuaculture_production)

#### Filter out aquaculture landings ####
aquaculture_keywords<- c(
  "DE CULTIVO", "DE CULTIVO VIVO", "ABASTO RANCHO", "RANCHO"
)

# Flag aquaculture rows:
Landings_GoC <- Landings_GoC %>%
  mutate(
    aquaculture_flag = (
      coalesce(tolower(acuaculture_production) == "yes", FALSE) |
      (rowSums(across(keywords_1:keywords_5, ~ .x %in% aquaculture_keywords), na.rm = TRUE) > 0)
    )
  )

table(Landings_GoC$aquaculture_flag)
# 115988 observations explicitly flagged as aquaculture

#### Filter out freshwater landings ####
freshwater_keywords<- c(
  "DE AGUA DULCE"
)

freshwater_taxa<- c(
  "tilapia",
  "rana", "rana toro",
  "trucha",
  "bagre canal"
)

# Flag freshwater rows:
Landings_GoC <- Landings_GoC %>%
  mutate(
    freshwater_flag = coalesce(
      (common_name_clean %in% freshwater_taxa) |
      (rowSums(across(keywords_1:keywords_5, ~ .x %in% freshwater_keywords), na.rm = TRUE) > 0) |
      (habitat == "freshwater") |
      (IHH_functional_major_groups_sm =="Freshwater fishes"),
      FALSE
    )
  )

table(Landings_GoC$freshwater_flag)
# 286013 observations explicitly flagged as freshwater landings (either based on the taxa, keyword, or landing site)

#### Now filter out freshwater & aquaculture landings from main landings ####
Landings_GoC_Fresh.Mariculture <- Landings_GoC %>%
  filter(aquaculture_flag | freshwater_flag)
# filter out the records that are  definitely aquaculture or freshwater
# 396343 obs., 50 vars

Landings_GoC_NoFresh.NoMariculture <- Landings_GoC %>%
  filter(!(aquaculture_flag | freshwater_flag))
# 3163120 obs., 50 vars

sum(is.na(Landings_GoC_NoFresh.NoMariculture$OHI_Region_No))
table(Landings_GoC_NoFresh.NoMariculture$OHI_Region_No)
# Seems like all entries have an OHI region number...
```

```{r CREATE A MARICULTURE DATAFRAM}
#### CREATE A MARICULTURE DATAFRAME: ####
Landings_GoC_Mariculture <- Landings_GoC %>%
  filter(aquaculture_flag & !freshwater_flag)
# filter out the records that are  definitely aquaculture or freshwater
# 110330 obs., 50 vars
```

# Save the FIS and MAR dataframes as of Dec 09 2025
```{r Save the FIS and MAR dataframes as of Dec 09 2025}
saveRDS(Landings_GoC_NoFresh.NoMariculture, "/home/shares/ohi/OHI_GOC/goal_prep/fp/fis/v2025/int/GOC_FIS.landings_CLEAN_as.of.Dec092025.rds")
saveRDS(Landings_GoC_Mariculture, "/home/shares/ohi/OHI_GOC/goal_prep/fp/mar/v2025/int/GOC_MAR.landings_CLEAN_as.of.Dec092025.rds")
```

```{r Top Mariculture Sp.}
table(Landings_GoC_Mariculture$common_name_clean)

Landings_GoC_Mariculture_Summary<- Landings_GoC_Mariculture %>% 
  filter(year >= 2014, year <= 2024) %>%
  select(state, recorded_taxa_group, recorded_taxa_name_clean, weight_kg_corrected, value) %>%
  group_by(state, recorded_taxa_group, recorded_taxa_name_clean) %>%
  summarize(total_catch_2014.2024= sum(weight_kg_corrected, na.rm = TRUE),
            total_value_2014.2024= sum(value, na.rm = TRUE))

unique(Landings_GoC_Mariculture_Summary$recorded_taxa_name_clean)

GoC_Top_MaricultureSp <- Landings_GoC_Mariculture_Summary %>%
  filter(total_catch_2014.2024 >= 10000) %>%
  pull(recorded_taxa_name_clean) %>%
  unique()

GoC_Top_MaricultureSp # NOTE TO SELF: 29 top species and some of them are fairly general or could be freshwater...

Landings_GoC_TopMariculture_Summary<- Landings_GoC_Mariculture %>% 
  filter(year >= 2014, year <= 2024) %>%
  filter(recorded_taxa_name_clean %in% GoC_Top_MaricultureSp) %>%
  select(recorded_taxa_group, recorded_taxa_name_clean, weight_kg_corrected, value) %>%
  group_by(recorded_taxa_group, recorded_taxa_name_clean) %>%
  summarize(total_catch_2014.2024= sum(weight_kg_corrected, na.rm = TRUE),
            total_value_2014.2024= sum(value, na.rm = TRUE))
```

```{r Save Landings_GoC_TopMariculture_Summary to find MAR sustainability scores}
# write.csv(Landings_GoC_TopMariculture_Summary, "/home/ferrer/ohi_goc_prep/FP/MAR/v2025/GOC_Mariculture_TopCatches_2014.2024_Dec082025.csv")
```

# SUMMARY STATS ABOUT OUR DATA: return to this later.
```{r How much of biomass includes information about species, genus, family?}
Landings_GoC %>%
  pull(estimated_species_name) %>%
  {c(n_na = sum(is.na(.)),
      n_unknown = sum(. == "unknown", na.rm = TRUE),
      n_NEI = sum(. == "NEI", na.rm = TRUE)
    ) }
# 1008085 (1839 + 205110 + 765582) out of 3559463 records (27%) for species are NA, "unknown", or "NEI"
# Meaning we have species info for 73% of records.

Landings_GoC %>%
  pull(estimated_genus_name) %>%
  {c(n_na = sum(is.na(.)),
     n_unknown = sum(. == "unknown", na.rm = TRUE),
     n_NEI = sum(. == "NEI", na.rm = TRUE)
    ) }
# 1008085 (206949 + 0 + 765582) out of 3559463 records (27%) for genus are NA, "unknown", or "NEI"
# Meaning we have genus info for 73% of records.

Landings_GoC %>%
  pull(estimated_family_name) %>%
  {c(n_na = sum(is.na(.)),
     n_unknown = sum(. == "unknown", na.rm = TRUE),
     n_NEI = sum(. == "NEI", na.rm = TRUE),
     n_family_info_pending = sum(. == "family missing but genus available", na.rm = TRUE)
    ) }
# 336348 (185825 + 0 + 150523) out of 3559463 records (9%) for family are NA, "unknown", or "NEI". 29438 (~1%) records have information about genus but family info pending.
# Meaning we already have or can get family info for 91% of records.
```

# SCORE A: CPUE Estimates (per major marine group) ------------------------------
```{r CPUE calculations (rowwise)}
# Load file, only if needed:
GoC_Landings_for_CPUE<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/fis/v2025/int/GOC_FIS.landings_CLEAN_as.of.Dec092025.rds"))
# GoC_Landings_for_CPUE<- Landings_GoC_NoFresh.NoMariculture
# 3163120 obs., 50 vars.

# Calculate the difference in days between period start and end:
GoC_Landings_for_CPUE$days_fishing <- as.numeric(GoC_Landings_for_CPUE$period_end_fixed - GoC_Landings_for_CPUE$period_start_fixed) + 1

GoC_Landings_for_CPUE<- GoC_Landings_for_CPUE %>%
  filter(!is.na(days_fishing) & days_fishing > 0 & live_weight > 0) %>%
  mutate(CPUE_kg.day = weight_kg_corrected / days_fishing)
# 2931534 obs., 52 vars.

max(GoC_Landings_for_CPUE$CPUE_kg.day) #288000
min(GoC_Landings_for_CPUE$CPUE_kg.day) #0.00125

# In general, the receipts from early trips are a bit funky. Let's filter those out for now. Filter >= 2003
GoC_Landings_for_CPUE<- GoC_Landings_for_CPUE %>%
  filter(year >= 2003)
# 2736299 obs.

max(GoC_Landings_for_CPUE$CPUE_kg.day) #1,400,000
min(GoC_Landings_for_CPUE$CPUE_kg.day) #0.00125

##### NOTE TO SELF:
# - There is some funky stuff going on with date end and start. Come back to this.
# - I am not sure what to do about the days with no record of live weight. Exclude these for now.
```

```{r Let's explore the landings for which IHH_major_group equals NA}
TEST<- GoC_Landings_for_CPUE %>% filter(is.na(IHH_functional_major_groups_sm))
# 180324 out of 2787363 total records contain no IHH_functional_major_groups_sm (this equate to about ~6-7% of records).

TEST2<- TEST %>% 
  select(recorded_taxa_group, recorded_taxa_name_clean, common_name_clean, keywords_1:keywords_5) %>% 
  distinct()

# write.csv(TEST2, "/home/ferrer/ohi_goc_prep/general/Species_Keys/NoIHHClass_SpKey_UNDERCONSTRUCTION_Dec022025.csv", row.names = FALSE)

rm(TEST, TEST2)
```
NOTE TO SELF: Interesting. Most things could probably be grouped to the IHH group level with a little bit of extra leg work... COME BACK TO THIS IN THE NEW YEAR.

```{r Estimate CPUE for each taxonomic group per region per year versus "historic" baseline}
# STEP 1: ESTABLISH CPUE BASELINE VALUES, 2003-2013 -----
# where baseline equals CPUE average from 2003-2013
CPUE_baseline <- GoC_Landings_for_CPUE %>%
  filter(year >= 2003, year <= 2013) %>%
  filter(!is.na(OHI_Region_No)) %>%
  group_by(OHI_Region_No, IHH_functional_major_groups_sm) %>%
  summarise(
    CPUE_baseline = mean(CPUE_kg.day, na.rm = TRUE),
    .groups = "drop"
  )

# STEP 2:CALCULATE CPUE VALUES FOR RECENT YEARS, 2014-2024 -----
CPUE_recent <- GoC_Landings_for_CPUE %>%
  filter(year >= 2014, year <= 2024) %>%
  filter(!is.na(OHI_Region_No)) %>%
  group_by(year, OHI_Region_No, IHH_functional_major_groups_sm) %>%
  summarise(
    Total_Biomass_kg = sum(weight_kg_corrected, na.rm = TRUE),
    Total_Value_mxn = sum(value, na.rm = TRUE),
    Total_Days = sum(days_fishing, na.rm = TRUE),
    Avg_CPUE = mean(CPUE_kg.day, na.rm = TRUE),
    n_records = n(),
    .groups = "drop"
  )

# Quick aside: how much of percent catch do the NAs account for?
subset<- CPUE_recent %>% filter(is.na(IHH_functional_major_groups_sm))
# Again, alot of these NAs can be accounted for with further species key work.

# STEP 3: COMPUTE PERCENT BIOMASS FOR RECENT YEARS -----
CPUE_recent <- CPUE_recent %>%
  group_by(year, OHI_Region_No) %>%
  mutate(
    total_biomass_state_year = sum(Total_Biomass_kg, na.rm = TRUE),
    pct_of_total_biomass = (Total_Biomass_kg / total_biomass_state_year) * 100
  ) %>%
  ungroup()

# STEP 4: MERGE BASELINE TO PRESENT DATA -----
CPUE_Table <- CPUE_recent %>%
  left_join(CPUE_baseline,
            by = c("OHI_Region_No", "IHH_functional_major_groups_sm")) %>%
  mutate(
    CPUE_ratio = Avg_CPUE / CPUE_baseline  # Score = CPUEpresent / CPUE_baseline
  )

# STEP 5: DESIRED TABLE -----
CPUE_Table <- CPUE_Table %>%
  transmute(
    OHI_Region = OHI_Region_No,
    Year = year,
    IHH_Taxonomic_Group = IHH_functional_major_groups_sm,
    CPUEpresent_vs_CPUEbaseline = round(CPUE_ratio, 3),
    Total_Landings_kg = round(Total_Biomass_kg, 1),
    Taxa_per_Total_Catch = round(pct_of_total_biomass, 2)
  ) %>%
  arrange(OHI_Region, Year, IHH_Taxonomic_Group)

# STEP 5B: CAP CPUE AT A MAX OF 1 per GROUP
CPUE_Table <- CPUE_Table %>%
  mutate(
    CPUEpres_vs_CPUEbase_CORRECTED = if_else(
      CPUEpresent_vs_CPUEbaseline > 1,
      1,
      CPUEpresent_vs_CPUEbaseline
    )
  )

# STEP 6: OHI SCORES PER REGION PER YEAR -----
OHI_CPUE_Scores <- CPUE_Table %>%
  mutate(TaxaProp = Taxa_per_Total_Catch/100) %>%     # IF your % values are 0–100; if 0–1 then skip this line
  group_by(Year, OHI_Region) %>%
  summarise(
    Mean_relative.CPUE_AllGroups = sum(CPUEpres_vs_CPUEbase_CORRECTED * TaxaProp, na.rm = TRUE)
  ) %>%
  ungroup()
```

```{r Save this CPUE data table}
write.csv(OHI_CPUE_Scores, , row.names = FALSE)
```

```{r Plot CPUE Scores over time}
#### PLOT CPUE OVER TIME ####

#### State Colors ####
state_colors <- c(
  "BAJA CALIFORNIA" = "royalblue",
  "BAJA CALIFORNIA SUR" = "skyblue",
  "SONORA" = "orange",
  "SINALOA" = "sienna1",
  "NAYARIT" = "plum", 
  "JALISCO" = "purple4"
)

Region_colors <- c(
  "royalblue",
  "skyblue",
  "orange",
  "sienna1",
  "plum",
  "purple4",
  "aquamarine3",
  "gold2",
  "tomato"
)

Regions_of_Interest <- c(
  "Region_1", "Region_2", "Region_3",
  "Region_4", "Region_5", "Region_6",
  "Region_7", "Region_8", "Region_9"
)

OHI_CPUEscores_overTime <- OHI_CPUE_Scores %>%
  filter(OHI_Region %in% Regions_of_Interest) %>%
  ggplot(aes(
    x = Year,
    y = Mean_relative.CPUE_AllGroups,
    color = OHI_Region
  )) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ OHI_Region, scales = "fixed") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
  
  scale_color_manual(values = Region_colors) +
  
  labs(
    title = "FIS - Regional CPUE Scores (2014–2024) relative to Baseline (2003–2013)",
    y = "CPUE Scores (Ratio of CPUE_present to CPUE_baseline capped at 1)",
    x = "Year"
  ) +
  ylim(0, 1.25) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 18, face = "bold")
  )

OHI_CPUEscores_overTime
  
  # # Save as PNG
  # ggsave(
  #   filename = sprintf("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_CPUE_Scores/OHI_CPUEScores_overTime_Dec092025.png"),
  #   plot = OHI_CPUEscores_overTime,
  #   width = 12,
  #   height = 12,
  #   dpi = 300
  # )

```

## Take a closer look at CPUE per Region and per Taxa:
```{r Plot CPUE over all years 2003-2024}
CPUE_AllYears <- GoC_Landings_for_CPUE %>%
  filter(year >= 2003, year <= 2024) %>%
  filter(!is.na(OHI_Region_No)) %>%
  group_by(OHI_Region_No, year, IHH_functional_major_groups_sm) %>%
  summarise(
    Total_Biomass_kg = sum(weight_kg_corrected, na.rm = TRUE),
    Total_Value_mxn = sum(value, na.rm = TRUE),
    Total_Days = sum(days_fishing, na.rm = TRUE),
    Avg_CPUE = mean(CPUE_kg.day, na.rm = TRUE),
    n_records = n(),
    .groups = "drop"
  )

Taxa_colors <- c("darkblue", "royalblue", "skyblue", "orange", "sienna1", "plum", "purple4", "aquamarine3", "gold2", "tomato", "grey")

Region_colors <- c("royalblue", "skyblue", "orange", "sienna1", "plum", "purple4", "aquamarine3", "gold2", "tomato")

Regions_of_Interest <- c(
  "Region_1", "Region_2", "Region_3",
  "Region_4", "Region_5", "Region_6",
  "Region_7", "Region_8", "Region_9"
)


# PLOT 1: CPUE overTime per Region, facet by Taxa
CPUE_overTime_byRegion_perTaxa<- CPUE_AllYears %>%
  filter(OHI_Region_No %in% Regions_of_Interest) %>%
  ggplot(
       aes(x = year, 
           y = Avg_CPUE, 
           group = OHI_Region_No, 
           color = OHI_Region_No)) +
  geom_line(size = 1) +                # lines connecting the points
  facet_wrap(~ IHH_functional_major_groups_sm, scales = "free_y") +  # facet by region
  labs(
    title = "Average CPUE Over Time by Functional Group",
    x = "Year",
    y = "Avg CPUE (kg/day)",
    color = "Functional Group"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_color_manual(values = Taxa_colors) +  # nice palette
  geom_vline(xintercept = 2013, linetype= "solid", color = "grey5")

CPUE_overTime_byRegion_perTaxa

  # # Save as PNG
  # ggsave(
  #   filename = sprintf("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_CPUE_Scores/CPUE_overTime_byRegion_perTaxa_Dec092025.png"),
  #   plot = CPUE_overTime_byRegion_perTaxa,
  #   width = 12,
  #   height = 12,
  #   dpi = 300
  # )


# PLOT 2: CPUE overTime per Taxa, facet by Region
CPUE_overTime_byRegion_perTaxa<- CPUE_AllYears %>%
  filter(OHI_Region_No %in% Regions_of_Interest) %>%
  ggplot(
       aes(x = year, 
           y = Avg_CPUE, 
           group = IHH_functional_major_groups_sm, 
           color = IHH_functional_major_groups_sm)) +
  geom_line(size = 1) +                # lines connecting the points
  facet_wrap(~ OHI_Region_No, scales = "free_y") +  # facet by region
  labs(
    title = "Average CPUE Over Time by Functional Group",
    x = "Year",
    y = "Avg CPUE (kg/day)",
    color = "Functional Group"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_color_manual(values = Taxa_colors) +  # nice palette
  geom_vline(xintercept = 2013, linetype= "solid", color = "grey5")

CPUE_overTime_byTaxa_perRegion

  # # Save as PNG
  # ggsave(
  #   filename = sprintf("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_CPUE_Scores/CPUE_overTime_byTaxa_perRegion_Dec092025.png"),
  #   plot = CPUE_overTime_byTaxa,
  #   width = 12,
  #   height = 12,
  #   dpi = 300
  # )


# PLOT 3: CPUE overTime per Taxa, facet by Region - normalized 0 to 1
CPUE_overTime_byTaxa_perTaxa_normalized<- CPUE_AllYears %>%
  filter(OHI_Region_No %in% Regions_of_Interest) %>%
  # Normalize Avg_CPUE per functional group across all years
  group_by(IHH_functional_major_groups_sm) %>%
  mutate(Avg_CPUE_norm = (Avg_CPUE - min(Avg_CPUE, na.rm = TRUE)) /
                            (max(Avg_CPUE, na.rm = TRUE) - min(Avg_CPUE, na.rm = TRUE))) %>%
  ungroup() %>%
  ggplot(aes(
    x = year, 
    y = Avg_CPUE_norm, 
    group = IHH_functional_major_groups_sm, 
    color = IHH_functional_major_groups_sm
  )) +
  geom_line(size = 1) +
  facet_wrap(~ OHI_Region_No, scales = "free_y") +
  labs(
    title = "Normalized CPUE Over Time by Functional Group",
    x = "Year",
    y = "Normalized CPUE (0-1)",
    color = "Functional Group"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_color_manual(values = Taxa_colors) +
  geom_vline(xintercept = 2013, linetype = "solid", color = "grey5")

CPUE_overTime_byTaxa_perRegion_normalized

  # Save as PNG
  ggsave(
    filename = sprintf("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_CPUE_Scores/CPUE_overTime_byTaxa_perRegion_normalized_Dec092025.png"),
    plot = CPUE_overTime_byTaxa_normalized,
    width = 12,
    height = 12,
    dpi = 300
  )
```

```{r Take a closer look at CPUE per Region and per Taxa}
# Region 5
subset<- GoC_Landings_for_CPUE %>% 
  filter(OHI_Region_No == "Region_5" & is.na(IHH_functional_major_groups_sm))

table(subset$recorded_taxa_name)
table(subset$recorded_taxa_group)
# Again, a lot of these NAs can be accounted for with further species key work.

# Region 7
subset<- GoC_Landings_for_CPUE %>% 
  filter(OHI_Region_No == "Region_7" & is.na(IHH_functional_major_groups_sm))

table(subset$recorded_taxa_name)
table(subset$recorded_taxa_group)
```


# SCORE B, PICK 1: RAM Stock Assessment - UNDER CONSTRUCTION ------------------------------

# SCORE B, PICK 2: CMSY Stock Assesments - UNDER CONSTRUCTION ------------------------------
Must run CMSY on personal machine. Can generate the main files here though. 

```{r Review the CMSY protocol developed by Ferrer et al. (2021)}
# Absolute path to your PDF
pdf_path <- normalizePath(
  "~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Copy of Ferrer et al. (2021) CMSY Protocol.pdf",
  winslash = "/", mustWork = TRUE
)

# Convert to file URL
file_url <- paste0("file:///", URLencode(pdf_path, reserved = TRUE))

# Open in default browser
browseURL(file_url)
```

# GENERATING FILES 1 & 2 FROM SCRATCH

NOTES TO SELF - Dec 09 2025:

FILE 1: Create a file with annual landings in kg (and n_count observations) grouped by year per "stock". Here, stock is defined as the total population of GENUS X fished within the Gulf of California from Baja to northern Jalisco, not including landings from the Pacific.

- At present, we have had some issues with keying out species and landing_sites. Thus, whatever timeseries we have here is likely to be an UNDERESTIMATE of total landings for all genera in the GoC. I plan to re-make this file (File 1) in the new year once we have more key info.

- For now, I’m going to run CMSY at the genus level for now, since I actually think it’s a safer more reliable bet than species-level info and because – in many cases – fishermen seem to target taxa of the same genus interchangeably.

FILE 2: Create a file with OHI_Region, Stock, Group = Name, English Name, Scientific Name, Source, and a whole bunch of year columns that you populate using the various R files shared by Alfredo. This file can probably be trimmed down but for now I recommend formatting it exactly as it appears from before.

# Create FILE 1 - Timeseries
```{r File 1 - Stock Timeseries (JC data)}
# Load file, only if needed:
GoC_Landings_for_CMSY<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/fis/v2025/int/GOC_FIS.landings_CLEAN_as.of.Dec092025.rds"))
# GoC_Landings_for_CMSY<- Landings_GoC_NoFresh.NoMariculture
# 3163120 obs., 50 vars.
```

```{r Basic checks ahead of CMSY}
# Some basic checks:
table(GoC_Landings_for_CMSY$fleet)
# large_scale small_scale 
#      142404      975060 

table(GoC_Landings_for_CMSY$acuaculture_production) 
# no aquaculture. As expected!

table(GoC_Landings_for_CMSY$freshwater_flag) 
# no freshwater flags. As expected!

table(GoC_Landings_for_CMSY$GoC) 
table(GoC_Landings_for_CMSY$GulfCal) 
# Everything in Gulf of California, that we know of!

table(GoC_Landings_for_CMSY$year)
# Plenty of data years 2000-2025

unique(GoC_Landings_for_CMSY$estimated_genus_name) 
# Okay. There are a few genera we're going to toss...
```

```{r Basic filters ahead of CMSY}
#### REMOVE FRESHWATER GENERA WE ARE NOT INTERESTED ####
genus_exclude<- c(
  "Cyprinus", # common carp; strictly freshwater.
  "Macrobrachium", # freshwater prawns/shrimps; mostly freshwater.
  "Hypophthalmichthys", # Asian carp; freshwater.
  "Chirostoma", # “charales,” endemic to freshwater lakes in Mexico.
  "Carassius", # goldfish/crucian carp; freshwater.
  "Micropterus", # black bass; freshwater.
  "Lepomis", # sunfish; freshwater.
  "Cyprinodon", # pupfish; freshwater.
  "Rana", # frogs
  "NEI", # not classed yet
  "Especies", # not specific enough to be anything
  NA # not classed
  )

# Flag genera to exclude:
GoC_Landings_for_CMSY <- GoC_Landings_for_CMSY %>%
  mutate(
    exclude_flag = if_else(
      !is.na(estimated_genus_name) & estimated_genus_name %in% genus_exclude,
      TRUE,
      FALSE
    )
  )

subset<- GoC_Landings_for_CMSY %>% filter(exclude_flag== TRUE)
# Dropping 710533 out of 3163120 total records (~22%) for now. A lot of these are NEI that will be classed in the new year!

GoC_Landings_for_CMSY<- GoC_Landings_for_CMSY %>% filter(exclude_flag== FALSE)
# 2452587 obs., 51 vars.


#### REMOVE RARE GENERA ####
table(GoC_Landings_for_CMSY$estimated_genus_name)
# Some of these genera don't appear very often. I want to filter genera for which we have more than 10 years worth of data and > 10 records per year.

# Summarize per genus and year
Genus_summary <- GoC_Landings_for_CMSY %>%
  filter(!is.na(estimated_genus_name)) %>%
  group_by(estimated_genus_name, year) %>%
  summarise(
    n_records = n(),
    .groups = "drop"
  )

# Count how many years have more than 10 records
Genus_filtered <- Genus_summary %>%
  filter(n_records > 10) %>%
  group_by(estimated_genus_name) %>%
  summarise(n_years = n(), .groups = "drop") %>%
  filter(n_years > 10)

Genera_of_Interest<- Genus_filtered$estimated_genus_name
# 98 remaining genera

# Keep only genera that meet the criteria in the main dataset
GoC_Landings_for_CMSY <- GoC_Landings_for_CMSY %>%
  filter(estimated_genus_name %in% Genera_of_Interest)
# 2225315 obs., 51 vars.

#### Filter out OHI Region = NA ####
sum(is.na(GoC_Landings_for_CMSY$OHI_Region_No))
# Not applicable, all entries have an OHI Region number.
```

```{r Mitigate genera key issues}
# NOTE TO SELF: Clean this up in the new year. This is a band aid solution to larger key issues.

#### GENUS KEY ####
# NOTE TO SELF: Return to this when needed (Dec 09 2025). Returned the next day, Dec 10 2025.
GenusKey <- GoC_Landings_for_CMSY %>%
  select(recorded_taxa_group, estimated_genus_name, IHH_functional_major_groups_sm) %>%
  distinct(recorded_taxa_group, estimated_genus_name, IHH_functional_major_groups_sm, .keep_all = TRUE)
# 161 obs.

# See the issue here? One genus could be associated with multiple recorded_taxa_groups. We need the unique combination of estimated_genus_name & recorded_taxa_group to key out the larger dataset. But, in the end, we'll want to be able to group by ESTIMATED_taxa_group to avoid issues of splitting the catch from a single genus across two larger groups.

# ---- Stage 1: Assign within-genus functional groups ----
GenusKey <- GenusKey %>%
  group_by(estimated_genus_name) %>%
  mutate(
    # Helper values
    n_func = n_distinct(IHH_functional_major_groups_sm, na.rm = TRUE),
    first_non_na_func = IHH_functional_major_groups_sm[!is.na(IHH_functional_major_groups_sm)][1],
    first_func = IHH_functional_major_groups_sm[1],

    # Main assignment
    Major_Functional_Group = case_when(
      n_func == 1 ~ IHH_functional_major_groups_sm,
      !is.na(first_non_na_func) ~ first_non_na_func,
      TRUE ~ first_func            # fallback if all NA
    )
  ) %>%
  ungroup() %>%
  select(-n_func, -first_non_na_func, -first_func)

# ---- Stage 2: Patch NA values based on shared recorded_taxa_group ----
GenusKey <- GenusKey %>%
  group_by(recorded_taxa_group) %>%
  mutate(
    # find any non-NA functional groups from same recorded_taxa_group
    group_level_func = Major_Functional_Group[!is.na(Major_Functional_Group)][1],

    # apply rule
    Major_Functional_Group = if_else(
      is.na(Major_Functional_Group) & !is.na(group_level_func),
      group_level_func,
      Major_Functional_Group
    )
  ) %>%
  ungroup() %>%
  select(-group_level_func)

unique(GenusKey$estimated_genus_name)
# 98 unique genera.

# Final patches for Algae, Peto, and "sharks":
GenusKey <- GenusKey %>%
  mutate(
    Major_Functional_Group = case_when(
      
      recorded_taxa_group == "ALGAS"   ~ "Marine algae",
      recorded_taxa_group == "PETO"    ~ "Sharks",
      recorded_taxa_group == "TIBURON" ~ "Sharks",
      
      TRUE ~ Major_Functional_Group
    )
  )

# MERGE this key to the larger dataset
GoC_Landings_for_CMSY <- left_join(
  GoC_Landings_for_CMSY,
  GenusKey %>%,
  by = c("recorded_taxa_group", "estimated_genus_name")
)
# 2225315 obs., 53 vars. as expected!

# A few quick sanity checks:
subset<- GoC_Landings_for_CMSY %>% filter(is.na(Major_Functional_Group))
sum(is.na(GoC_Landings_for_CMSY$Major_Functional_Group))
table(GoC_Landings_for_CMSY$estimated_taxa_group)
sum(is.na(GoC_Landings_for_CMSY$estimated_taxa_group))
```

```{r Summary table}
# Turn off scientific notation:
options(scipen = 999)

# Create timeseries FILE 1:
colnames(GoC_Landings_for_CMSY)
unique(GoC_Landings_for_CMSY$estimated_genus_name)

TotalLandings_AnnualSummary_byGenus <- GoC_Landings_for_CMSY %>%
  select(year, estimated_genus_name, weight_kg_corrected) %>%
  group_by(year, estimated_genus_name) %>%
  summarise(
    Total_Landed_Biomass = sum(weight_kg_corrected, na.rm = TRUE),
    n_Records = n(),
    .groups = "drop"  # optional: ungroup after summarising
  )
# 2401 obs., 4 vars.

TotalLandings_AnnualSummary_byGenus <- GoC_Landings_for_CMSY %>%
  select(year, Major_Functional_Group = IHH_functional_major_groups_sm, estimated_taxa_group, estimated_genus_name, weight_kg_corrected) %>%
  group_by(year, Major_Functional_Group, estimated_taxa_group, estimated_genus_name) %>%
  summarise(
    Total_Landed_Biomass = sum(weight_kg_corrected, na.rm = TRUE),
    n_Records = n(),
    .groups = "drop"  # optional: ungroup after summarising
  )
# 2401 obs., 6 vars. GOOD!!!

table(TotalLandings_AnnualSummary_byGenus$Major_Functional_Group)
        #           Cephalopods                   Crustaceans        Demersal marine fishes             Diadromous fishes 
        #                    26                            41                          1370                            26 
        #          Marine algae             Marine fishes NEI Miscellaneous aquatic animals    Molluscs excl. Cephalopods 
        #                    23                            26                            44                           222 
        # Pelagic marine fishes               Shrimps, prawns 
        #                   498                           125 
# Lots of demersal marine fishes!

File1_df<- TotalLandings_AnnualSummary_byGenus

# Re-name column
colnames(File1_df)[colnames(File1_df) == "estimated_genus_name"] <- "Stock"
colnames(File1_df)[colnames(File1_df) == "year"] <- "Year"

# Finally, and just for now, filter out data from 2025
# NOTE TO SELF: Come back to this in subsequent updates, once we have full data for 2025.
File1_df<- File1_df %>% filter(Year<2025)
```

## SAVE FILE 1
```{r SAVE FILE 1}
write.csv(File1_df, "~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Input_Files/File1_Dec092025.csv", row.names = FALSE)
```

# Create FILE 2 - Metadata & Model Priors
```{r Create File 2 based on File 1}
# STEP 1:
# Which taxa (stocks) do we want to account for in our File 2?
Stock_names<- unique(File1_df$Stock)
# 98 unique genera / stocks

## Read in the column names of File 2 from Ferrer et al. (2022) CSV file
headers <- names(read.csv("FIS/v2025/OHI_CMSY_Scores/Instructional_Files/CMSY_File2_BluePrint_Dec092025.csv", nrows = 1))

## Create an empty dataframe with the same headers
File2_df <- data.frame(matrix(ncol = length(headers), nrow = 0))
colnames(File2_df) <- headers

## Add Stock_names to the 'Stock' column
File2_df <- File2_df[1:length(Stock_names), ]  # Expand rows
File2_df$Stock <- Stock_names

## Reset rownames / the index column
rownames(File2_df) <- NULL


# STEP 2: TRIME AND RENAME AS NECESSARY
# Trim some of the columns off this file: only if necessary.

# Re-name columns: only if necessary.
# colnames(File2_df)[colnames(File2_df) == "Keep"] <- "Yes_Run"
# colnames(File2_df)[colnames(File2_df) == "stb.low"] <- "st_b.low"
# colnames(File2_df)[colnames(File2_df) == "stb.hi"] <- "st_b.hi"


# STEP 3: FILL AS NECESSARY
# Fill in some of the columns off this file:
File2_df$Region <- c("GOC")
File2_df$Class <- c("TBD")
File2_df$Source <- c("JC_data")
File2_df$btype <- c("None")
File2_df$force.cmsy <- c("TRUE")


# STEP 4:
# For species with less than 10 years of data, do not keep (Sufficient_No_Years = 0)
StockYears_per_Taxa <- File1_df %>%
  group_by(Stock) %>%
  summarise(n_Stock_Years = n_distinct(Year), .groups = "drop")

StockYears_per_Taxa <- StockYears_per_Taxa %>%
  mutate(Sufficient_No_Years = ifelse(n_Stock_Years < 10, 0, 1))

# How many stocks will we be keeping based on number of years?
table(StockYears_per_Taxa$Sufficient_No_Years)
# All of them, which makes sense since we already applied this filter above.

##  Add this information to the File 2 dataframe: join StockYears_per_Taxa to File2_df by "Stock", bringing in the filled Sufficient_No_Years column
File2_df <- File2_df %>%
  select(-Sufficient_No_Years) %>%  # Remove empty column to avoid duplicates
  left_join(
    StockYears_per_Taxa %>% select(Stock, Sufficient_No_Years),
    by = "Stock"
  )

# STEP 5: ADD IN THE GENUS KEY INFO
GenusKey_distinct<- GenusKey %>% 
  select(estimated_genus_name, estimated_taxa_group) %>%
  distinct()

File2_df <- File2_df %>%
  left_join(
    GenusKey_distinct,
    by = c("Stock"="estimated_genus_name")
  )

# STEP 6:
# Re-order columns: only if necessary.
```

## SAVE FILE 2 BLUE PRINT
```{r SAVE FILE 2 UNDER CONSTRUCTION}
write.csv(File2_df, "~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Input_Files/File2_UNDER_CONSTRUCTION_Dec092025.csv", row.names = FALSE)
```
NEXT STEPS: Fill in the rest of file 2, namely the model priors, using the ancillary files shared with me by Alfredo (Ferrer et al., 2022)
NOTE TO SELF: These model priors are likely to change in the new year but will do for now.

Did some work offline on my local machine (not available for viewing here... yet). This work will help fill out File 2.

```{r Read in additional files to complete file 2}
# Read in file if necessary:
#File2_df <- read.csv("~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Input_Files/File2_UNDER_CONSTRUCTION_Dec092025.csv")

## Additional files needed to file in the metadata
Priors_Pristine.Biomass <- read.csv("~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Model_Priors/Priors_StartingBiomass_fromFAOdata_DRAFT_Dec092025.csv")
Priors_Int.Year <- read.csv("~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Model_Priors/Priors_IntermediateYears_DRAFT_Dec092025.csv")
Priors_IntEnd.Conditions <- read.csv("~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Model_Priors/Priors_IntEndConditions_DRAFT_Dec092025.csv")
Priors_Resilience <- read.csv("~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Model_Priors/Priors_ResilienceValues_DRAFT_Dec092025.csv")
Priors_Resilience <- Priors_Resilience %>% select(-ScientificName)
# Common_Names <- read.csv("Stock and Common Names_Oct 23 2025.csv")
```

```{r lowercase_function}
to_lower_df <- function(df) {
  df %>% mutate(across(where(is.character), tolower))
}
```

```{r Stitch these dataframes together to create complete File2_df}
File2_df        <- to_lower_df(File2_df)
Priors_Pristine.Biomass  <- to_lower_df(Priors_Pristine.Biomass)
Priors_Int.Year <- to_lower_df(Priors_Int.Year)
Priors_Resilience <- to_lower_df(Priors_Resilience)

## MERGE: PRISTINE BIOMASS PRIORS
File2_df_merged <- File2_df %>% 
  left_join(
    Priors_Pristine.Biomass %>% select(estimated_taxa_group, st_b.low, st_b.hi),
    by = "estimated_taxa_group",
    suffix = c(".file2", ".priors")
  ) %>%
  mutate(
    st_b.low = coalesce(st_b.low.file2, st_b.low.priors),
    st_b.hi  = coalesce(st_b.hi.file2,  st_b.hi.priors)
  ) %>% 
  select(-ends_with(".file2"), -ends_with(".priors"))

# NOTE TO SELF: Our priors for pristine biomass, and the FAO timeseries that underly them, are fairly sparse as in missing several major taxonomic groups (like sardine). Return to this in the new year.

## MERGE: INT YEAR PRIORS
File2_df_merged <- File2_df_merged %>%
  left_join(
    Priors_Int.Year %>% 
      select(estimated_taxa_group, int_yr, StartYear, EndYear),
    by = "estimated_taxa_group",
    suffix = c(".file2", ".priors")
  ) %>%
  mutate(
    int_yr               = coalesce(int_yr.file2,               int_yr.priors),
    StartYear            = coalesce(StartYear.file2,            StartYear.priors),
    EndYear              = coalesce(EndYear.file2,              EndYear.priors)) %>%
  select(-ends_with(".file2"), -ends_with(".priors"))

## MERGE: INT END CONDITIONS PRIORS
# NOTE TO SELF: Based on the naming conventions within this file I actually think these are the priors for intermediate and ending biomass. I'll circle back to this and iron out the true meaning and utility of these particular priors in the new year.
# Join and fill missing values
File2_df_merged <- File2_df_merged %>%
  left_join(
    Priors_IntEnd.Conditions,
    by = "estimated_taxa_group",
    suffix = c("", ".priors")
  ) %>%
  mutate(
    int_b.low  = coalesce(int_b.low, int_b.low.priors),
    int_b.hi   = coalesce(int_b.hi, int_b.hi.priors),
    end_b.low  = coalesce(end_b.low, end_b.low.priors),
    end_b.hi = coalesce(end_b.hi, end_b.hi.priors)
  ) %>%
  select(-ends_with(".priors"))


## MERGE: RESILIENCE PRIORS
File2_df_merged <- File2_df_merged %>%
  left_join(
    Priors_Resilience %>% select(estimated_taxa_group, Resilience, Resilience_Source),
    by = "estimated_taxa_group",
    suffix = c(".file2", ".priors")
  ) %>%
  mutate(
    Resilience = coalesce(Resilience.file2, Resilience.priors)
  ) %>%
  select(-Resilience.file2, -Resilience.priors)
```

```{r Finally lets clean up File 2 before saving and running CMSY}
File2_df_merged<- File2_df_merged %>% 
  select(Region, Stock, Landings_Datasource=Source, 
         StartYear, EndYear, int_yr,
         btype,
         Flim, Fpa, Blim, Bpa, Bmsy, B40,
         FMSY, MSY, MSYBtrigger, M,
         Fofl, last_F,
         q.start, q.end,
         st_b.low, st_b.hi, 
         int_b.low, int_b.hi, end_b.low, end_b.hi,
         r.low, r.hi, 
         Resilience, Resilience_Source,
         Yes_Run, Comment)

# Fill in the Yes_Run column based on model priors:
File2_df_merged <- File2_df_merged %>%
  mutate(
    Yes_Run = if_else(
      is.na(Resilience) | is.na(int_b.low) | is.na(st_b.low) | is.na(end_b.low),
      0,
      1
    ),
    Comment = if_else(
      Yes_Run == 0,
      "Incomplete model priors",
      NA_character_
    )
  )

sum(File2_df_merged$Yes_Run==1)
# 38 stocks to run!

# NOTES ON NAMING, Dec 10 2025:
# Need a btype column = None
# Need a force.cmsy column = TRUE
# StartYear and EndYear must actually align with file one time series.
```

```{r How much of FP will these CMSY stocks account for?}
Landings_CMSY_TEST <- Landings_GoC_NoFresh.NoMariculture %>%
  mutate(estimated_genus_name = tolower(estimated_genus_name))

CMSY_Stocks <- File2_df_merged %>%
  filter(Yes_Run == 1) %>%
  pull(Stock) %>%
  unique()

CMSY_Stocks_hypothetical <- c(
  CMSY_Stocks, 
  "sardinops", 
  "opisthonema", 
  "sarda",
  "scomber"
)

# Calculate total from genera landings identified in the CMSY exercise:
# NOTE TO SELF: Come back to this. Find priors for these four taxa! They represent a huge component of total landings.
Landings_CMSY_summary <- Landings_CMSY_TEST %>%
  filter(year %in% 2015:2024) %>%           # only last 10 years
  mutate(is_CMSY = estimated_genus_name %in% CMSY_Stocks_hypothetical) %>%  # flag CMSY genera
  summarise(
    total_landings = sum(weight_kg_corrected, na.rm = TRUE),                # total landings
    CMSY_landings  = sum(weight_kg_corrected[is_CMSY], na.rm = TRUE)       # landings from CMSY genera
  ) %>%
  mutate(percent_CMSY = 100 * CMSY_landings / total_landings)    # percentage
  
Landings_CMSY_summary
# If I can sort out priors for these four extra txa, CMSY could account for ~50% of all GoC landings in the last five years.
```

## SAVE FILE 2 COMPLETE
```{r SAVE FILE 2 COMPLETE}
write.csv(File2_df_merged, "~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Input_Files/File2_Dec092025.csv", row.names = FALSE)
```

```{r How much of FP will these CMSY stocks account for?}
Landings_CMSY_TEST <- Landings_GoC_NoFresh.NoMariculture %>%
  mutate(estimated_genus_name = tolower(estimated_genus_name))

CMSY_Stocks <- File2_df_merged %>%
  filter(Yes_Run == 1) %>%
  pull(Stock) %>%
  unique()

CMSY_Stocks_hypothetical <- c(
  CMSY_Stocks, 
  "sardinops", 
  "opisthonema", 
  "sarda",
  "scomber"
)

# Calculate total from genera landings identified in the CMSY exercise:
# NOTE TO SELF: Come back to this. Find priors for these four taxa! They represent a huge component of total landings.
Landings_CMSY_summary <- Landings_CMSY_TEST %>%
  filter(year %in% 2015:2024) %>%           # only last 10 years
  mutate(is_CMSY = estimated_genus_name %in% CMSY_Stocks_hypothetical) %>%  # flag CMSY genera
  summarise(
    total_landings = sum(weight_kg_corrected, na.rm = TRUE),                # total landings
    CMSY_landings  = sum(weight_kg_corrected[is_CMSY], na.rm = TRUE)       # landings from CMSY genera
  ) %>%
  mutate(percent_CMSY = 100 * CMSY_landings / total_landings)    # percentage
  
Landings_CMSY_summary
# If I can sort out priors for these four extra txa, CMSY could account for ~50% of all GoC landings in the last five years.
```
# RAN CMSY ON PERSONAL MACHINE. Here are the finals from that analysis:
```{r CMSY Outputs}
CMSY_Timeseries<- read.csv("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Output_Files/CMSY_Assessment_Timeseries_Dec102025.csv")
 
OHI_CMSY_Scores<- CMSY_Timeseries %>% select(stock, yr, 
                                             B, B.Bmsy, lcl.B.Bmsy, ucl.B.Bmsy,
                                             F.Fmsy, lcl.F.Fmsy, ucl.F.Fmsy)

OHI_CMSY_Scores <- Landings_GoC_Cleaned_wSp %>%
  left_join(
    RegionKey_working %>%
      select(-landing_site_key),
    by = c("state", "landing_site")
  )
```


# SCORE B, PICK 3: Catch vs Carta Nacional Pesquera Baseline - UNDER CONSTRUCTION ------------------------------
## Load in recommendations for the Carta Nacional Pesquera
```{r Load in CNP data}
CNP_data<- read.csv("/home/frazier/ohi_goc_prep/FP/FIS/v2025/data/CNP_2023_recommendations_clean.csv")
~/ohi_goc_prep/FP/
```

```{r Catch compare to CNP}
options(scipen = 999) # Force same numeric format; don't want scientific notation

# Setting up the dataframe to compare to the C.N.P.
CPUE_Landings_summarized <- CPUE_Landings_filtered %>%
  filter(year %in% 2020:2024) %>%
  select(state, main_species_group, year, CPUE_kg.day, weight_kg_corrected) %>%
  group_by(state, main_species_group) %>%
  summarize(
    # First sum per year, then average across years 2020–2022
    Baseline_Total_Catch_2020_2022 = mean(
      tapply(weight_kg_corrected[year %in% 2020:2022],
             year[year %in% 2020:2022],
             sum, na.rm = TRUE)
    ),
    Total_Catch_2023 = sum(weight_kg_corrected[year == 2023], na.rm = TRUE),
    Total_Catch_2024 = sum(weight_kg_corrected[year == 2024], na.rm = TRUE),
    Avg_CPUE_2020_2022 = mean(CPUE_kg.day[year %in% 2020:2022], na.rm = TRUE),
    Avg_CPUE_2023 = mean(CPUE_kg.day[year == 2023], na.rm = TRUE),
    Avg_CPUE_2024 = mean(CPUE_kg.day[year == 2024], na.rm = TRUE)
  ) %>%
  ungroup() %>%
  mutate(
    # Define a threshold for suspiciously high values
    qc_flag = Baseline_Total_Catch_2020_2022 * 3,
    # Define a 30% margin of error around the baseline mean - SUBSTANTIALLY different
    lowest_bound = Baseline_Total_Catch_2020_2022 * 0.70, # most extreme cases must come first
    highest_bound = Baseline_Total_Catch_2020_2022 * 1.30,
    # Define a 5% margin of error around the baseline mean
    low_bound = Baseline_Total_Catch_2020_2022 * 0.95,
    high_bound = Baseline_Total_Catch_2020_2022 * 1.05,

    # 2023 comparison with margin of error
    CatchCompare_2023_vs_Baseline = case_when(
      Total_Catch_2023 > qc_flag ~ "qc flag: > 3x higher",
      Total_Catch_2023 > highest_bound ~ "substantially higher",
      Total_Catch_2023 < lowest_bound ~ "substantially lower",
      Total_Catch_2023 > high_bound ~ "higher",
      Total_Catch_2023 < low_bound ~ "lower",
      TRUE ~ "similar"
    ),

    # 2024 comparison with margin of error
    CatchCompare_2024_vs_Baseline = case_when(
      Total_Catch_2024 > qc_flag ~ "qc flag: > 3x higher",
      Total_Catch_2024 > highest_bound ~ "substantially higher",
      Total_Catch_2024 < lowest_bound ~ "substantially lower",
      Total_Catch_2024 > high_bound ~ "higher",
      Total_Catch_2024 < low_bound ~ "lower",
      TRUE ~ "similar"
    )
  ) %>%
  select(-low_bound, -high_bound, -lowest_bound, -highest_bound, -qc_flag)   # remove helper columns if not needed

# Add & re-arrange columns:
CPUE_Landings_summarized$CNP_Recommendation<- NA

CPUE_Landings_summarized <- CPUE_Landings_summarized %>%
  select(state,
         main_species_group,
         CNP_Recommendation,
         Baseline_Total_Catch_2020_2022,
         Total_Catch_2023,
         CatchCompare_2023_vs_Baseline,
         Total_Catch_2024,
         CatchCompare_2024_vs_Baseline,
         Avg_CPUE_2020_2022,
         Avg_CPUE_2023,
         Avg_CPUE_2024,
         everything())  # keeps any other columns after
```

```{r Write this as a CSV}

```

#### Merge the Indicator dataframes

```{r Load the indicator dataframes as necessary}
# Read in datafiles as necessary:
OHI_CPUE_Scores <- read.csv("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_CPUE_Scores/OHI_CPUEScores_FIRSTDRAFT_Dec092025.csv")

OHI_CMSY_Scores

Master_SpKey<- read.csv("/home/ferrer/ohi_goc_prep/general/CONAPESCA_Keys/Species_Keys/CleanedMaster_SpKey_Dec012025.csv")

CMSY_SpKey<- Master_SpKey
CMSY_SpKey$estimated_genus_name<- tolower(CMSY_SpKey$estimated_genus_name)

CMSY_SpKey<- CMSY_SpKey %>%
  select(IHH_functional_major_groups_sm, recorded_taxa_group, estimated_genus_name) %>% 
  distinct()
```

```{r Merge the Indicator dataframes}
# Merge:
Landings_GoC <- Landings_GoC_Cleaned_wSp %>%
  left_join(
    RegionKey_working %>%
      select(-landing_site_key),
    by = c("state", "landing_site")
  )
```