################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```

# Region Key (December 2025) ------------------------------
```{r Region Key from Mel}
BajaSonoraKey<- read_csv(here("/home/frazier/ohi_goc_prep/FP/FIS/v2025/data/Baja_Sonora_states_key.csv"))

BajaSonoraKey <- BajaSonoraKey %>%
  mutate(
    long = as.numeric(long),
    lat  = as.numeric(lat)
  )

JaliscoKey<- read_csv(here("/home/frazier/ohi_goc_prep/FP/FIS/v2025/data/landing_sites_jalisco.csv"))

JaliscoKey <- JaliscoKey %>%
  rename(long= lon) %>% 
  mutate(
    long = as.numeric(long),
    lat  = as.numeric(lat)
  )

# MERGE THESE WORKING KEYS FOR NOW:
RegionKey_working <- bind_rows(BajaSonoraKey, JaliscoKey)
# 1199 obs., 9 vars.

RegionKey_working <- RegionKey_working %>%
  mutate(
    state_key = if_else(
      is.na(state),
      "NAYARIT|JALISCO",   # permitted states when unknown
      state
    )
  )

RegionKey_expanded <- RegionKey_working %>%
  separate_rows(state_key, sep = "\\|")

RegionKey_temporary<- read_csv(here("/home/ferrer/ohi_goc_prep/general/RegionKey_UNDERCONSTRUCTION_Dec02024.csv"))

rm(BajaSonoraKey, JaliscoKey)
```
# Load: Landings data with species info (December 2025) ------------------------------
```{r Load in landings data with species info}
Landings_GoC_Cleaned_wSp<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fis/v2025/int/juan_carlos/GoC_Landings_wSp_2000.present_Dec012025.rds"))
# 3716415 obs, 41 vars.

Landings_GoC_Cleaned_wSp <- Landings_GoC_Cleaned_wSp %>%
  mutate(
    join_key = paste(state, landing_site, sep = "_")
  )
# 3716415 obs, 42 vars.

table(Landings_GoC_Cleaned_wSp$GoC)
# All in the GoC or not otherwise noted.
```

```{r Add OHI region to landings w species}
Landings_GoC_Cleaned_wSp <- Landings_GoC_Cleaned_wSp %>%
  left_join(
    RegionKey_expanded %>%
      select(-lat, -long, -GulfCal, -landing_site_key, -state),
    by = c("state" = "state_key", "landing_site")
  )
# 3716415 obs, 44 vars. as expected! (Many-to-many join issue, 3737864 observations...)

Landings_GoC_Cleaned_wSp <- Landings_GoC_Cleaned_wSp %>%
  mutate(Region_pick = if_else(!is.na(OHI_rgn), OHI_rgn, state)) %>%
  select(-OHI_rgn)
# 3716415 obs, 44 vars. as expected!

Landings_GoC_Cleaned_wSp <- Landings_GoC_Cleaned_wSp %>%
  left_join(
    RegionKey_temporary %>%
      select(-Region_Name_Proposed),  # remove this column for now
    by = c("Region_pick" = "Region_Mel_or_CONAPESCA")
  )
# 3716415 obs, 45 vars. as expected!

subset<- Landings_GoC_Cleaned_wSp %>% filter(is.na(OHI_Region_No))
# 170587 obs., 45 vars.
# Only Jalisco entries are not keyed. This makes sense. These are parts of Jalisco that either aren't in Region 9 or we can't be totally sure they are in Region 9.

Landings_GoC_Cleaned_wSp<- Landings_GoC_Cleaned_wSp %>% filter(!is.na(OHI_Region_No)) 
```

```{r Landed weight correction}
# Kind of looks like sometimes 'landed weight' is 'live weight' and vice versa. Fix this.

Landings_GoC_Cleaned_wSp$weight_kg_corrected <- ifelse(
  Landings_GoC_Cleaned_wSp$live_weight >= Landings_GoC_Cleaned_wSp$landed_weight,
  Landings_GoC_Cleaned_wSp$live_weight,
  Landings_GoC_Cleaned_wSp$landed_weight
)
```

```{r Filtering out things we don't care about}
colnames(Landings_GoC_Cleaned_wSp)
table(Landings_GoC_Cleaned_wSp$acuaculture_production)

#### Filter out aquaculture landings ####
aquaculture_keywords<- c(
  "DE CULTIVO", "DE CULTIVO VIVO", "ABASTO RANCHO", "RANCHO"
)

# Flag aquaculture rows:
Landings_GoC_Cleaned_wSp <- Landings_GoC_Cleaned_wSp %>%
  mutate(
    aquaculture_flag = (
      coalesce(tolower(acuaculture_production) == "yes", FALSE) |
      (rowSums(across(keywords_1:keywords_5, ~ .x %in% aquaculture_keywords), na.rm = TRUE) > 0)
    )
  )

table(Landings_GoC_Cleaned_wSp$aquaculture_flag)
# 116059 observations explicitly flagged as aquaculture

#### Filter out freshwater landings ####
freshwater_keywords<- c(
  "DE AGUA DULCE"
)

freshwater_taxa<- c(
  "tilapia",
  "rana", "rana toro",
  "trucha",
  "bagre canal"
)

# Flag freshwater rows:
Landings_GoC_Cleaned_wSp <- Landings_GoC_Cleaned_wSp %>%
  mutate(
    freshwater_flag = coalesce(
      (common_name_clean %in% freshwater_taxa) |
      (rowSums(across(keywords_1:keywords_5, ~ .x %in% freshwater_keywords), na.rm = TRUE) > 0) |
      (habitat == "freshwater"),
      FALSE
    )
  )

table(Landings_GoC_Cleaned_wSp$freshwater_flag)
# 232960 observations explicitly flagged as freshwater landings (either based on the taxa, keyword, or landing site)

#### Now filter out freshwater & aquaculture landings from main landings ####
Landings_GoC_Fresh.Aquaculture <- Landings_GoC_Cleaned_wSp %>%
  filter(aquaculture_flag | freshwater_flag)
# filter out the records that are  definitely aquaculture or freshwater
# 343842 obs., 49 vars

Landings_GoC_NoAquaculture <- Landings_GoC_Cleaned_wSp %>%
  filter(!(aquaculture_flag | freshwater_flag))
# 3223435 obs., 49 vars
```

```{r Top Mariculture Sp.}
#### CREATE A MARICULTURE DATAFRAME: ####
Landings_GoC_Mariculture <- Landings_GoC_Cleaned_wSp %>%
  filter(aquaculture_flag & !freshwater_flag)
# filter out the records that are  definitely aquaculture or freshwater
# 110882 obs., 49 vars

table(Landings_GoC_Mariculture$common_name_clean)

Landings_GoC_Mariculture_Summary<- Landings_GoC_Mariculture %>% 
  filter(year >= 2014, year <= 2024) %>%
  select(state, recorded_taxa_group, recorded_taxa_name_clean, weight_kg_corrected, value) %>%
  group_by(state, recorded_taxa_group, recorded_taxa_name_clean) %>%
  summarize(total_catch_2014.2024= sum(weight_kg_corrected, na.rm = TRUE),
            total_value_2014.2024= sum(value, na.rm = TRUE))

unique(Landings_GoC_Mariculture_Summary$recorded_taxa_name_clean)

GoC_Top_MaricultureSp <- Landings_GoC_Mariculture_Summary %>%
  filter(total_catch_2014.2024 >= 10000) %>%
  pull(recorded_taxa_name_clean) %>%
  unique()

GoC_Top_MaricultureSp # NOTE TO SELF: 29 top species and some of them are fairly general or could be freshwater...

Landings_GoC_TopMariculture_Summary<- Landings_GoC_Mariculture %>% 
  filter(year >= 2014, year <= 2024) %>%
  filter(recorded_taxa_name_clean %in% GoC_Top_MaricultureSp) %>%
  select(recorded_taxa_group, recorded_taxa_name_clean, weight_kg_corrected, value) %>%
  group_by(recorded_taxa_group, recorded_taxa_name_clean) %>%
  summarize(total_catch_2014.2024= sum(weight_kg_corrected, na.rm = TRUE),
            total_value_2014.2024= sum(value, na.rm = TRUE))
```

```{r Save Landings_GoC_TopMariculture_Summary to find MAR sustainability scores}
write.csv(Landings_GoC_TopMariculture_Summary, "/home/ferrer/ohi_goc_prep/FP/MAR/v2025/GOC_Mariculture_TopCatches_2014.2024_Dec082025.csv")
```

# SUMMARY STATS ABOUT OUR DATA: return to this later.
```{r How much of biomass includes information about species, genus, family?}
Landings_GoC_Cleaned_wSp %>%
  pull(estimated_species_name) %>%
  {c(n_na = sum(is.na(.)),
      n_unknown = sum(. == "unknown", na.rm = TRUE),
      n_NEI = sum(. == "NEI", na.rm = TRUE)
    ) }
# 1008085 (4186 + 210708 + 793191) out of 3716415 records (27%) for species are NA, "unknown", or "NEI"
# Meaning we have species info for 73% of records.

Landings_GoC_Cleaned_wSp %>%
  pull(estimated_genus_name) %>%
  {c(n_na = sum(is.na(.)),
     n_unknown = sum(. == "unknown", na.rm = TRUE),
     n_NEI = sum(. == "NEI", na.rm = TRUE)
    ) }
# 1008085 (214894 + 0 + 793191) out of 3716415 records (27%) for genus are NA, "unknown", or "NEI"
# Meaning we have genus info for 73% of records.

Landings_GoC_Cleaned_wSp %>%
  pull(estimated_family_name) %>%
  {c(n_na = sum(is.na(.)),
     n_unknown = sum(. == "unknown", na.rm = TRUE),
     n_NEI = sum(. == "NEI", na.rm = TRUE),
     n_family_info_pending = sum(. == "family missing but genus available", na.rm = TRUE)
    ) }
# 352911 (192402 + 0 + 160509) out of 3716415 records (9%) for family are NA, "unknown", or "NEI". 29659 records have information about genus but family info pending.
# Meaning we already have or can get family info for 91% of records.
```

# SCORE A: CPUE Estimates (per major marine group) ------------------------------
```{r CPUE calculations (rowwise)}
GoC_Landings_for_CPUE<- Landings_GoC_NoAquaculture
# 3223435 obs., 49 vars.

# Calculate the difference in days between period start and end:
GoC_Landings_for_CPUE$days_fishing <- as.numeric(GoC_Landings_for_CPUE$period_end_fixed - GoC_Landings_for_CPUE$period_start_fixed) + 1

GoC_Landings_for_CPUE<- GoC_Landings_for_CPUE %>%
  filter(!is.na(days_fishing) & days_fishing > 0 & live_weight > 0) %>%
  mutate(CPUE_kg.day = weight_kg_corrected / days_fishing)
# 2988373 obs., 50 vars.

max(GoC_Landings_for_CPUE$CPUE_kg.day) #288000
min(GoC_Landings_for_CPUE$CPUE_kg.day) #0.00125

# In general, the receipts from early trips are a bit funky. Let's filter those out for now.
GoC_Landings_for_CPUE<- GoC_Landings_for_CPUE %>%
  filter(year >= 2003)
# 2790070 obs.

max(GoC_Landings_for_CPUE$CPUE_kg.day) #1,400,000
min(GoC_Landings_for_CPUE$CPUE_kg.day) #0.00125

##### NOTE TO SELF:
# - There is some funky stuff going on with date end and start. Come back to this.
# - I am not sure what to do about the days with no record of live weight. Exclude these for now.
```

```{r Let's explore the landings for which IHH_major_group equals NA}
TEST<- GoC_Landings_for_CPUE %>% filter(is.na(IHH_functional_major_groups_sm))
# 187583 out of 2862286 records contain no IHH_functional_major_groups_sm = ~6-7% of landing records.

TEST2<- TEST %>% 
  select(recorded_taxa_group, recorded_taxa_name_clean, common_name_clean, keywords_1:keywords_5) %>% 
  distinct()

# write.csv(TEST2, "/home/ferrer/ohi_goc_prep/general/Species_Keys/NoIHHClass_SpKey_UNDERCONSTRUCTION_Dec022025.csv", row.names = FALSE)

rm(TEST, TEST2)
```
NOTE TO SELF: Interesting. Most things could probably be grouped to the IHH group level with a little bit of extra leg work... COME BACK TO THIS IN THE NEW YEAR.

```{r Estimate CPUE for each taxonomic group per region per year versus "historic" baseline}
# STEP 1: ESTABLISH CPUE BASELINE VALUES, 2003-2013 -----
# where baseline equals CPUE average from 2003-2013
CPUE_baseline <- GoC_Landings_for_CPUE %>%
  filter(year >= 2003, year <= 2013) %>%
  filter(!is.na(OHI_Region_No)) %>%
  group_by(OHI_Region_No, IHH_functional_major_groups_sm) %>%
  summarise(
    CPUE_baseline = mean(CPUE_kg.day, na.rm = TRUE),
    .groups = "drop"
  )

# STEP 2:CALCULATE CPUE VALUES FOR RECENT YEARS, 2014-2024 -----
CPUE_recent <- GoC_Landings_for_CPUE %>%
  filter(year >= 2014, year <= 2024) %>%
  filter(!is.na(OHI_Region_No)) %>%
  group_by(year, OHI_Region_No, IHH_functional_major_groups_sm) %>%
  summarise(
    Total_Biomass_kg = sum(weight_kg_corrected, na.rm = TRUE),
    Total_Value_mxn = sum(value, na.rm = TRUE),
    Total_Days = sum(days_fishing, na.rm = TRUE),
    Avg_CPUE = mean(CPUE_kg.day, na.rm = TRUE),
    n_records = n(),
    .groups = "drop"
  )

# STEP 3: COMPUTE PERCENT BIOMASS FOR RECENT YEARS -----
CPUE_recent <- CPUE_recent %>%
  group_by(year, OHI_Region_No) %>%
  mutate(
    total_biomass_state_year = sum(Total_Biomass_kg, na.rm = TRUE),
    pct_of_total_biomass = (Total_Biomass_kg / total_biomass_state_year) * 100
  ) %>%
  ungroup()

# STEP 4: MERGE BASELINE TO PRESENT DATA -----
CPUE_Table <- CPUE_recent %>%
  left_join(CPUE_baseline,
            by = c("OHI_Region_No", "IHH_functional_major_groups_sm")) %>%
  mutate(
    CPUE_ratio = Avg_CPUE / CPUE_baseline  # Score = CPUEpresent / CPUE_baseline
  )

# STEP 5: DESIRED TABLE -----
CPUE_Table <- CPUE_Table %>%
  transmute(
    OHI_Region = OHI_Region_No,
    Year = year,
    IHH_Taxonomic_Group = IHH_functional_major_groups_sm,
    CPUEpresent_vs_CPUEbaseline = round(CPUE_ratio, 3),
    Total_Landings_kg = round(Total_Biomass_kg, 1),
    Taxa_per_Total_Catch = round(pct_of_total_biomass, 2)
  ) %>%
  arrange(OHI_Region, Year, IHH_Taxonomic_Group)

# STEP 5B: CAP CPUE AT A MAX OF 1 per GROUP
CPUE_Table <- CPUE_Table %>%
  mutate(
    CPUEpres_vs_CPUEbase_CORRECTED = if_else(
      CPUEpresent_vs_CPUEbaseline > 1,
      1,
      CPUEpresent_vs_CPUEbaseline
    )
  )

# STEP 6: OHI SCORES PER REGION PER YEAR -----
OHI_CPUE_Scores <- CPUE_Table %>%
  mutate(TaxaProp = Taxa_per_Total_Catch/100) %>%     # IF your % values are 0–100; if 0–1 then skip this line
  group_by(Year, OHI_Region) %>%
  summarise(
    Mean_relative.CPUE_AllGroups = sum(CPUEpres_vs_CPUEbase_CORRECTED * TaxaProp, na.rm = TRUE)
  ) %>%
  ungroup()
```

```{r Save this CPUE data table}
write.csv(OHI_CPUE_Scores, "/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_CPUE_Scores/OHI_CPUEScores_FIRSTDRAFT_Dec082025.csv", row.names = FALSE)
```

```{r Plot CPUE over time}
#### PLOT CPUE OVER TIME ####

#### State Colors ####
state_colors <- c(
  "BAJA CALIFORNIA" = "royalblue",
  "BAJA CALIFORNIA SUR" = "skyblue",
  "SONORA" = "orange",
  "SINALOA" = "sienna1",
  "NAYARIT" = "plum", 
  "JALISCO" = "purple4"
)

Region_colors <- c(
  "royalblue",
  "skyblue",
  "orange",
  "sienna1",
  "plum",
  "purple4",
  "forestgreen",
  "gold2",
  "firebrick2"
)

Regions_of_Interest <- c(
  "Region_1", "Region_2", "Region_3",
  "Region_4", "Region_5", "Region_6",
  "Region_7", "Region_8", "Region_9"
)

OHI_avgCPUE_overTime <- OHI_CPUE_Scores %>%
  filter(OHI_Region %in% Regions_of_Interest) %>%
  ggplot(aes(
    x = Year,
    y = Mean_relative.CPUE_AllGroups,
    color = OHI_Region
  )) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ OHI_Region, scales = "fixed") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = Region_colors) +
  labs(
    title = "FIS - Regional CPUE Scores (2014–2024) relative to Baseline (2003–2013)",
    y = "CPUE Scores (Ratio of CPUE_present to CPUE_baseline capped at 1)",
    x = "Year"
  ) +
  ylim(0, 1.25) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 18, face = "bold")
  )

  
  # Save as PNG
  ggsave(
    filename = sprintf("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_CPUE_Scores/OHI_avgCPUE_overTime.png"),
    plot = OHI_avgCPUE_overTime,
    width = 12,
    height = 12,
    dpi = 300
  )

```


# SCORE B, PICK 1: RAM Stock Assessment - UNDER CONSTRUCTION ------------------------------

# SCORE B, PICK 2: CMSY Stock Assesments - UNDER CONSTRUCTION ------------------------------

# SCORE B, PICK 3: Catch vs Carta Nacional Pesquera Baseline - UNDER CONSTRUCTION ------------------------------
## Load in recommendations for the Carta Nacional Pesquera
```{r Load in CNP data}

```

```{r Catch compare to CNP}
options(scipen = 999) # Force same numeric format; don't want scientific notation

# Setting up the dataframe to compare to the C.N.P.
CPUE_Landings_summarized <- CPUE_Landings_filtered %>%
  filter(year %in% 2020:2024) %>%
  select(state, main_species_group, year, CPUE_kg.day, weight_kg_corrected) %>%
  group_by(state, main_species_group) %>%
  summarize(
    # First sum per year, then average across years 2020–2022
    Baseline_Total_Catch_2020_2022 = mean(
      tapply(weight_kg_corrected[year %in% 2020:2022],
             year[year %in% 2020:2022],
             sum, na.rm = TRUE)
    ),
    Total_Catch_2023 = sum(weight_kg_corrected[year == 2023], na.rm = TRUE),
    Total_Catch_2024 = sum(weight_kg_corrected[year == 2024], na.rm = TRUE),
    Avg_CPUE_2020_2022 = mean(CPUE_kg.day[year %in% 2020:2022], na.rm = TRUE),
    Avg_CPUE_2023 = mean(CPUE_kg.day[year == 2023], na.rm = TRUE),
    Avg_CPUE_2024 = mean(CPUE_kg.day[year == 2024], na.rm = TRUE)
  ) %>%
  ungroup() %>%
  mutate(
    # Define a threshold for suspiciously high values
    qc_flag = Baseline_Total_Catch_2020_2022 * 3,
    # Define a 30% margin of error around the baseline mean - SUBSTANTIALLY different
    lowest_bound = Baseline_Total_Catch_2020_2022 * 0.70, # most extreme cases must come first
    highest_bound = Baseline_Total_Catch_2020_2022 * 1.30,
    # Define a 5% margin of error around the baseline mean
    low_bound = Baseline_Total_Catch_2020_2022 * 0.95,
    high_bound = Baseline_Total_Catch_2020_2022 * 1.05,

    # 2023 comparison with margin of error
    CatchCompare_2023_vs_Baseline = case_when(
      Total_Catch_2023 > qc_flag ~ "qc flag: > 3x higher",
      Total_Catch_2023 > highest_bound ~ "substantially higher",
      Total_Catch_2023 < lowest_bound ~ "substantially lower",
      Total_Catch_2023 > high_bound ~ "higher",
      Total_Catch_2023 < low_bound ~ "lower",
      TRUE ~ "similar"
    ),

    # 2024 comparison with margin of error
    CatchCompare_2024_vs_Baseline = case_when(
      Total_Catch_2024 > qc_flag ~ "qc flag: > 3x higher",
      Total_Catch_2024 > highest_bound ~ "substantially higher",
      Total_Catch_2024 < lowest_bound ~ "substantially lower",
      Total_Catch_2024 > high_bound ~ "higher",
      Total_Catch_2024 < low_bound ~ "lower",
      TRUE ~ "similar"
    )
  ) %>%
  select(-low_bound, -high_bound, -lowest_bound, -highest_bound, -qc_flag)   # remove helper columns if not needed

# Add & re-arrange columns:
CPUE_Landings_summarized$CNP_Recommendation<- NA

CPUE_Landings_summarized <- CPUE_Landings_summarized %>%
  select(state,
         main_species_group,
         CNP_Recommendation,
         Baseline_Total_Catch_2020_2022,
         Total_Catch_2023,
         CatchCompare_2023_vs_Baseline,
         Total_Catch_2024,
         CatchCompare_2024_vs_Baseline,
         Avg_CPUE_2020_2022,
         Avg_CPUE_2023,
         Avg_CPUE_2024,
         everything())  # keeps any other columns after
```

```{r Write this as a CSV}

```
