################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```

# Region Key (December 2025) ------------------------------
```{r Region Key from Mel}
BajaSonoraKey<- read_csv(here("/home/frazier/ohi_goc_prep/FP/FIS/v2025/data/Baja_Sonora_states_key.csv"))

BajaSonoraKey <- BajaSonoraKey %>%
  mutate(
    long = as.numeric(long),
    lat  = as.numeric(lat)
  )

JaliscoKey<- read_csv(here("/home/frazier/ohi_goc_prep/FP/FIS/v2025/data/landing_sites_jalisco.csv"))

JaliscoKey <- JaliscoKey %>%
  rename(long= lon) %>% 
  mutate(
    long = as.numeric(long),
    lat  = as.numeric(lat)
  )

JaliscoKey_clean <- JaliscoKey %>%
  # If you want to normalize state names first
  mutate(state = toupper(state)) %>%
  
  # Remove any rows where landing_site is NA (can cause matching issues later)
  filter(!is.na(landing_site)) %>%
  
  # Keep only one row per landing_site_key; if duplicate keys exist, keep first
  group_by(landing_site_key) %>%
  slice(1) %>%  # keeps the first occurrence per key
  ungroup() %>%
  
  # Optional: also deduplicate by landing_site if there are rows with NA landing_site_key
  group_by(landing_site) %>%
  slice(1) %>%
  ungroup()

# NOTE TO SELF: Come back to this! The Jalisco key is imperfect, but we'll have to fix this in the new year.
# Identify rows that were removed in the deduplication of JaliscoKey
removed_rows <- anti_join(JaliscoKey, JaliscoKey_clean, 
                          by = c("landing_site_key", "landing_site"))


#### MERGE THESE WORKING KEYS FOR NOW: ####
RegionKey_working <- bind_rows(BajaSonoraKey, JaliscoKey_clean) %>%
  filter(!is.na(landing_site)) %>%  # remove the NA from this dataframe as this will create a many-to-many matching issue later on
  mutate(long = as.numeric(long))
# 1176 obs., 9 vars.

RegionKey_working <- RegionKey_working %>% mutate(long = as.numeric(long))

# My regional key - NOTE SELF
RegionKey_temporary<- read_csv(here("/home/ferrer/ohi_goc_prep/general/CONAPESCA_Keys/RegionKey_UNDERCONSTRUCTION_Dec02024.csv"))

rm(JaliscoKey, removed_rows)
```

# Load: Landings data with species info (December 2025) ------------------------------
```{r Load in landings data with species info}
Landings_GoC_Cleaned_wSp<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/juan_carlos/GoC_Landings_wSp_2000.present_Dec012025.rds"))
# 3716415 obs, 41 vars.

Landings_GoC_Cleaned_wSp <- Landings_GoC_Cleaned_wSp %>% 
  select(-lat, -long) # remove lat, long for now to do the merge

Landings_GoC <- Landings_GoC_Cleaned_wSp %>%
  left_join(
    RegionKey_working %>%
      select(-landing_site_key),
    by = c("state", "landing_site")
  )
# 3716415 obs, 45 vars.

# A few quick checks:
table(Landings_GoC$GoC)
# All in the GoC or not otherwise noted.
table(Landings_GoC$state)
table(Landings_GoC_Cleaned_wSp$state)
table(Landings_GoC$OHI_rgn)
# Excellent, completely keyed for now. Erica - Dec 9 2025
```

```{r Add OHI region to landings w species}
Landings_GoC <- Landings_GoC %>%
  mutate(Region_pick = if_else(!is.na(OHI_rgn), OHI_rgn, state))
# 3716415 obs, 46 vars. as expected!

Landings_GoC <- Landings_GoC %>%
  left_join(
    RegionKey_temporary %>%
      select(-Region_Name_Proposed),  # remove this column for now
    by = c("Region_pick" = "Region_Mel_or_CONAPESCA")
  )
# 3716415 obs, 47 vars. as expected!

subset<- Landings_GoC %>% filter(is.na(OHI_Region_No))
table(subset$state)
# 156952 obs., 47 vars.
# Only Jalisco entries are not keyed. This makes sense. These are parts of Jalisco that either aren't in Region 9 or we can't be totally sure they are in Region 9.

Landings_GoC<- Landings_GoC %>% filter(!is.na(OHI_Region_No)) 
# 3559463 obs., 47 vars.
```

```{r Landed weight correction}
# Kind of looks like sometimes 'landed weight' is 'live weight' and vice versa. Fix this:
Landings_GoC$weight_kg_corrected <- ifelse(
  Landings_GoC$live_weight >= Landings_GoC$landed_weight,
  Landings_GoC$live_weight,
  Landings_GoC$landed_weight
)
```

```{r Filtering out things we don't care about}
colnames(Landings_GoC)
table(Landings_GoC$acuaculture_production)

#### Filter out aquaculture landings ####
aquaculture_keywords<- c(
  "DE CULTIVO", "DE CULTIVO VIVO", "ABASTO RANCHO", "RANCHO"
)

# Flag aquaculture rows:
Landings_GoC <- Landings_GoC %>%
  mutate(
    aquaculture_flag = (
      coalesce(tolower(acuaculture_production) == "yes", FALSE) |
      (rowSums(across(keywords_1:keywords_5, ~ .x %in% aquaculture_keywords), na.rm = TRUE) > 0)
    )
  )

table(Landings_GoC$aquaculture_flag)
# 115988 observations explicitly flagged as aquaculture

#### Filter out freshwater landings ####
freshwater_keywords<- c(
  "DE AGUA DULCE"
)

freshwater_taxa<- c(
  "tilapia",
  "rana", "rana toro",
  "trucha",
  "bagre canal"
)

# Flag freshwater rows:
Landings_GoC <- Landings_GoC %>%
  mutate(
    freshwater_flag = coalesce(
      (common_name_clean %in% freshwater_taxa) |
      (rowSums(across(keywords_1:keywords_5, ~ .x %in% freshwater_keywords), na.rm = TRUE) > 0) |
      (habitat == "freshwater") |
      (IHH_functional_major_groups_sm =="Freshwater fishes"),
      FALSE
    )
  )

table(Landings_GoC$freshwater_flag)
# 286013 observations explicitly flagged as freshwater landings (either based on the taxa, keyword, or landing site)

#### Now filter out freshwater & aquaculture landings from main landings ####
Landings_GoC_Fresh.Mariculture <- Landings_GoC %>%
  filter(aquaculture_flag | freshwater_flag)
# filter out the records that are  definitely aquaculture or freshwater
# 396343 obs., 50 vars

Landings_GoC_NoFresh.NoMariculture <- Landings_GoC %>%
  filter(!(aquaculture_flag | freshwater_flag))
# 3163120 obs., 50 vars

sum(is.na(Landings_GoC_NoFresh.NoMariculture$OHI_Region_No))
table(Landings_GoC_NoFresh.NoMariculture$OHI_Region_No)
# Seems like all entries have an OHI region number...
```
NOTE TO SELF: There could still be some freshwater species hiding in here. For example, the "MOJARRA" general category is mixed freshwater and marine demersal fish.


```{r FIXING MAJOR FUNCTIONAL GROUPS}
# NOTE TO SELF: Return to this when needed (Dec 09 2025). Returned the next day, Dec 10 2025.
Genus_MajorGroup_Key_ALL <- Landings_GoC_NoFresh.NoMariculture %>%
  select(recorded_taxa_group, estimated_genus_name, IHH_functional_major_groups_sm) %>%
  distinct(recorded_taxa_group, estimated_genus_name, IHH_functional_major_groups_sm, .keep_all = TRUE)
# 367 obs., 3 vars.

# See the issue here? One genus could be associated with multiple recorded_taxa_groups. We need the unique combination of estimated_genus_name & recorded_taxa_group to key out the larger dataset. But, in the end, we'll want to be able to group by ESTIMATED_taxa_group to avoid issues of splitting the catch from a single genus across two larger functional groups.

# ---- Create Major_Functional_Group values based IHH major functional groups assigned to recorded_taxa_group ----

# Step 1: Assign one Major_Functional_Group per recorded_taxa_group, respecting that a few of the categories (like "OTRAS") are inherently different.
skip_groups <- c("OTRAS", "ESPECIE GENERICA", "FAUNA", "ORNATO", "NEI")

Genus_MajorGroup_Key_ALL$Assignment_for_Major_Functional_Group = NA_character_  # initialize flag column

Genus_MajorGroup_Key_ALL <- Genus_MajorGroup_Key_ALL %>%
  group_by(recorded_taxa_group) %>%
  mutate(
    
    # Most frequent non-NA value for IHH_functional_major_groups within the recorded_taxa_group
    group_major = {
      x <- IHH_functional_major_groups_sm
      x <- x[!is.na(x) & x != ""]
      if (length(x) == 0) NA_character_ else {
        tbl <- table(x)
        names(tbl)[which.max(tbl)]
      }
    },
    
    # Assign only to non-skipped groups
    Major_Functional_Group = if_else(
      !(recorded_taxa_group %in% skip_groups),
      group_major,
      NA_character_
    ),
      Assignment_for_Major_Functional_Group = if_else(
      !(recorded_taxa_group %in% skip_groups) & is.na(Assignment_for_Major_Functional_Group),
      "Assigned_Step_1: based on recorded_taxa_group and IHH_functional_major_groups",
      Assignment_for_Major_Functional_Group
    )
  ) %>%
  ungroup() %>%
  select(-group_major)

# Put Assignment_for_Major_Functional_Group column at the end:
Genus_MajorGroup_Key_ALL <- Genus_MajorGroup_Key_ALL %>%
  relocate(Assignment_for_Major_Functional_Group, .after = last_col())


# Step 2: Genus-level assignment (ignores cases where genus = NA or NEI)
Genus_MajorGroup_Key_ALL <- Genus_MajorGroup_Key_ALL %>%
  group_by(estimated_genus_name) %>%
  mutate(

    # ---- Determine genus_major ONLY IF genus is real (not NA, not NEI) ----
    genus_major = {
      genus_value <- cur_group()$estimated_genus_name   # single value

      if (!is.na(genus_value) && genus_value != "NEI") {

        x <- c(Major_Functional_Group, IHH_functional_major_groups_sm)
        x <- x[!is.na(x) & x != ""]
        if (length(x) == 0) NA_character_ else {
          tbl <- table(x)
          names(tbl)[which.max(tbl)]
        }

      } else {
        # genus is NA or NEI → cannot derive functional group
        NA_character_
      }
    },

    Major_Functional_Group = case_when(

      # ---- CASE 1: genus is NA or NEI → leave unchanged ----
      (is.na(estimated_genus_name) | estimated_genus_name == "NEI") ~ Major_Functional_Group,

      # ---- CASE 2: non-skip groups with missing FG → fill from genus_major ----
      !(recorded_taxa_group %in% skip_groups) &
        is.na(Major_Functional_Group) ~ genus_major,

      # ---- CASE 3: skip groups → fill *only* from genus_major ----
      recorded_taxa_group %in% skip_groups ~ genus_major,

      # ---- CASE 4: keep original ----
      TRUE ~ Major_Functional_Group
    ),
      Assignment_for_Major_Functional_Group = case_when(
      ((!(recorded_taxa_group %in% skip_groups) & is.na(Assignment_for_Major_Functional_Group) & !is.na(Major_Functional_Group)) |
       (recorded_taxa_group %in% skip_groups & is.na(Assignment_for_Major_Functional_Group) & !is.na(Major_Functional_Group))) ~ "Assigned_Step_2: based on genus (that has been keyed elsewhere)",
      TRUE ~ Assignment_for_Major_Functional_Group
    )
  ) %>%
  ungroup() %>%
  select(-genus_major)
  

# Step 3: Enforce genus-level consistency for Major_Functional_Group (ignores cases where genus = NA or NEI; assign "OTRAS" that have a verified genus)
Genus_MajorGroup_Key_ALL <- Genus_MajorGroup_Key_ALL %>%
  group_by(estimated_genus_name) %>%
  mutate(
    new_value = if (
      !is.na(cur_group()$estimated_genus_name) &&
      cur_group()$estimated_genus_name != "NEI" &&
      n() > 1
    ) {
      x <- Major_Functional_Group[!is.na(Major_Functional_Group) & Major_Functional_Group != ""]
      if(length(x) == 0) Major_Functional_Group else names(which.max(table(x))) 
    } else {
      Major_Functional_Group
    },
    Assignment_for_Major_Functional_Group = if_else(
      Major_Functional_Group != new_value & !is.na(new_value),
      "Assigned_Step_3: forced consistency across the genus",
      Assignment_for_Major_Functional_Group
    ),
    Major_Functional_Group = new_value
  ) %>%
  ungroup() %>%
  select(-new_value)
  

# Step 4: Manual patches for specific taxonomic groups (and two typos)
Genus_MajorGroup_Key_ALL <- Genus_MajorGroup_Key_ALL %>%
  mutate(
    Major_Functional_Group = case_when(
      estimated_genus_name == "Sphoeriodes" ~ "Marine fishes NEI", # Pressumably, is a typo of "Sphoeroides"
      estimated_genus_name == "Lujtanus" ~ "Demersal marine fishes", # Pressumably, is a typo of "Lutjanus"
      recorded_taxa_group == "MOJARRA" ~ "Marine fishes NEI",
      recorded_taxa_group == "ANCHOVETA" ~ "Pelagic marine fishes",
      recorded_taxa_group == "PECES ORNATO" ~ "Marine fishes NEI",
      recorded_taxa_group %in% c("ALGAS", "SARGAZO") ~ "Marine algae",
      recorded_taxa_group %in% c("PEPINO DE MAR", "ERIZO") ~ "Other marine invertebrates",
      recorded_taxa_group %in% c("PETO", "TIBURON", "CAZON") ~ "Sharks",
      TRUE ~ Major_Functional_Group
    ),
     Assignment_for_Major_Functional_Group = case_when(
      (estimated_genus_name == "Sphoeriodes" |
       estimated_genus_name == "Lujtanus" |
       recorded_taxa_group %in% c("MOJARRA", "ANCHOVETA", "PECES ORNATO", "ALGAS", "SARGAZO",
                                  "PEPINO DE MAR", "ERIZO", "PETO", "TIBURON", "CAZON")) ~ "Assigned_Step_4: manual patch",
      TRUE ~ Assignment_for_Major_Functional_Group
    )
  )


# Step 5: If Major_Functional_Group is NA then Assignment_for_Major_Functional_Group= "could not be assigned"
Genus_MajorGroup_Key_ALL <- Genus_MajorGroup_Key_ALL %>%
  mutate(
    Assignment_for_Major_Functional_Group = if_else(
      is.na(Major_Functional_Group),
      "Alert: Major_Functional_Group could not be assigned (at present)",
      Assignment_for_Major_Functional_Group
    )
  )


# Step 6: Assign estimated_taxa_group by looking at the most common recorded_taxa_group for each genus
Genus_MajorGroup_Key_ALL <- Genus_MajorGroup_Key_ALL %>%
  group_by(estimated_genus_name) %>%
  mutate(
    # Step 1: calculate mode of recorded_taxa_group within the genus
    genus_mode_group = if (!is.na(cur_group()$estimated_genus_name[1]) &&
                           cur_group()$estimated_genus_name[1] != "NEI") {
      tbl <- table(recorded_taxa_group)
      names(tbl)[which.max(tbl)]
    } else {
      NA_character_
    },
    
    # Step 2: assign estimated_taxa_group
    estimated_taxa_group = if_else(
      !is.na(genus_mode_group) & recorded_taxa_group != genus_mode_group,
      genus_mode_group,          # use mode for genus when recorded_taxa_group differs
      recorded_taxa_group        # otherwise keep recorded_taxa_group
    )
  ) %>%
  ungroup() %>%
  select(-genus_mode_group)


# Step 7: A couple of manual patches for specific taxa where we actually lost info due to "OTRAS" (e.g., Lutjanus and "Lujtanus, & Oligoplites)
Genus_MajorGroup_Key_ALL <- Genus_MajorGroup_Key_ALL %>%
  mutate(
    estimated_taxa_group = case_when(
      estimated_genus_name %in% c("Lutjanus", "Lujtanus") ~ "GUACHINANGO",
      estimated_genus_name %in% c("Oligoplites") ~ "SARDINA",
      estimated_genus_name %in% c("Pseudobatos") ~ "RAYA Y SIMILARES",
      estimated_genus_name %in% c("Scomberomorus") ~ "SIERRA",
      estimated_genus_name %in% c("Selene") ~ "RONCO",
      estimated_genus_name %in% c("Trachinotus") ~ "PAMPANO",
      TRUE ~ estimated_taxa_group
    )
  )


# Step 8: Rearrange for ease of reading:
Genus_MajorGroup_Key_ALL <- Genus_MajorGroup_Key_ALL %>%
  select(
    recorded_taxa_group,
    estimated_taxa_group,
    estimated_genus_name,
    IHH_functional_major_groups_sm,
    Major_Functional_Group,
    Assignment_for_Major_Functional_Group
  )
    
# 367 obs., 6 vars.
```

#### SAVE GENUS TO MAJOR_FUNCTIONAL_GROUP KEY
```{r Save this key!}
Genus_MajorGroup_Key<- Genus_MajorGroup_Key_ALL %>% 
  select(recorded_taxa_group, estimated_genus_name, estimated_taxa_group, Major_Functional_Group) %>%
  distinct()

write.csv(Genus_MajorGroup_Key, "~/ohi_goc_prep/general/CONAPESCA_Keys/Genus_to_MajorGroup_Key_Dec112025.csv", row.names = FALSE)
```


```{r CREATE A MARICULTURE DATAFRAM}
#### CREATE A MARICULTURE DATAFRAME: ####
Landings_GoC_Mariculture <- Landings_GoC %>%
  filter(aquaculture_flag & !freshwater_flag)
# filter out the records that are  definitely aquaculture or freshwater
# 110330 obs., 50 vars
```

# Save the FIS and MAR dataframes as of Dec 09 2025
```{r Save the FIS and MAR dataframes as of Dec 09 2025}
saveRDS(Landings_GoC_NoFresh.NoMariculture, "/home/shares/ohi/OHI_GOC/goal_prep/fp/fis/v2025/int/GOC_FIS.landings_CLEAN_as.of.Dec092025.rds")
saveRDS(Landings_GoC_Mariculture, "/home/shares/ohi/OHI_GOC/goal_prep/fp/mar/v2025/int/GOC_MAR.landings_CLEAN_as.of.Dec092025.rds")
```

```{r Top Mariculture Sp.}
table(Landings_GoC_Mariculture$common_name_clean)

Landings_GoC_Mariculture_Summary<- Landings_GoC_Mariculture %>% 
  filter(year >= 2014, year <= 2024) %>%
  select(state, recorded_taxa_group, recorded_taxa_name_clean, weight_kg_corrected, value) %>%
  group_by(state, recorded_taxa_group, recorded_taxa_name_clean) %>%
  summarize(total_catch_2014.2024= sum(weight_kg_corrected, na.rm = TRUE),
            total_value_2014.2024= sum(value, na.rm = TRUE))

unique(Landings_GoC_Mariculture_Summary$recorded_taxa_name_clean)

GoC_Top_MaricultureSp <- Landings_GoC_Mariculture_Summary %>%
  filter(total_catch_2014.2024 >= 10000) %>%
  pull(recorded_taxa_name_clean) %>%
  unique()

GoC_Top_MaricultureSp # NOTE TO SELF: 29 top species and some of them are fairly general or could be freshwater...

Landings_GoC_TopMariculture_Summary<- Landings_GoC_Mariculture %>% 
  filter(year >= 2014, year <= 2024) %>%
  filter(recorded_taxa_name_clean %in% GoC_Top_MaricultureSp) %>%
  select(recorded_taxa_group, recorded_taxa_name_clean, weight_kg_corrected, value) %>%
  group_by(recorded_taxa_group, recorded_taxa_name_clean) %>%
  summarize(total_catch_2014.2024= sum(weight_kg_corrected, na.rm = TRUE),
            total_value_2014.2024= sum(value, na.rm = TRUE))
```

```{r Save Landings_GoC_TopMariculture_Summary to find MAR sustainability scores}
# write.csv(Landings_GoC_TopMariculture_Summary, "/home/ferrer/ohi_goc_prep/FP/MAR/v2025/GOC_Mariculture_TopCatches_2014.2024_Dec082025.csv", row.names = FALSE)
```

# SUMMARY STATS ABOUT OUR DATA: return to this later.
```{r How much of biomass includes information about species, genus, family?}
Landings_GoC %>%
  pull(estimated_species_name) %>%
  {c(n_na = sum(is.na(.)),
      n_unknown = sum(. == "unknown", na.rm = TRUE),
      n_NEI = sum(. == "NEI", na.rm = TRUE)
    ) }
# 1008085 (1839 + 205110 + 765582) out of 3559463 records (27%) for species are NA, "unknown", or "NEI"
# Meaning we have species info for 73% of records.

Landings_GoC %>%
  pull(estimated_genus_name) %>%
  {c(n_na = sum(is.na(.)),
     n_unknown = sum(. == "unknown", na.rm = TRUE),
     n_NEI = sum(. == "NEI", na.rm = TRUE)
    ) }
# 1008085 (206949 + 0 + 765582) out of 3559463 records (27%) for genus are NA, "unknown", or "NEI"
# Meaning we have genus info for 73% of records.

Landings_GoC %>%
  pull(estimated_family_name) %>%
  {c(n_na = sum(is.na(.)),
     n_unknown = sum(. == "unknown", na.rm = TRUE),
     n_NEI = sum(. == "NEI", na.rm = TRUE),
     n_family_info_pending = sum(. == "family missing but genus available", na.rm = TRUE)
    ) }
# 336348 (185825 + 0 + 150523) out of 3559463 records (9%) for family are NA, "unknown", or "NEI". 29438 (~1%) records have information about genus but family info pending.
# Meaning we already have or can get family info for 91% of records.
```

# SCORE A: CPUE Estimates (per major marine group) ------------------------------
```{r Load data as necessary}
# Load file, only if needed:
GoC_Landings_for_CPUE_original<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/fis/v2025/int/GOC_FIS.landings_CLEAN_as.of.Dec092025.rds"))
GoC_Landings_for_CPUE<- GoC_Landings_for_CPUE_original
# GoC_Landings_for_CPUE<- Landings_GoC_NoFresh.NoMariculture
# 3163120 obs., 50 vars.
```

```{r CPUE calculations}
# Calculate the difference in days between period start and end:
GoC_Landings_for_CPUE$days_fishing <- as.numeric(GoC_Landings_for_CPUE$period_end_fixed - GoC_Landings_for_CPUE$period_start_fixed) + 1
# 3163120 obs., 51 vars.

# Estimate CPUE per trip:
GoC_Landings_for_CPUE<- GoC_Landings_for_CPUE %>%
  mutate(CPUE_kg.day = weight_kg_corrected / days_fishing)
# 3163120 obs., 52 vars.

# General filtering:
# In general, the receipts from early trips are a bit funky. Let's filter out trips before 2003 (n= 267072 records).
GoC_Landings_for_CPUE<- GoC_Landings_for_CPUE %>%
  filter(year >= 2003)
# 2896048 obs., 52 vars.

# Let's also filter out any records that did not have a receipt start and end date, and records where live weight was < or = 0 (n= 126955 records).
GoC_Landings_for_CPUE<- GoC_Landings_for_CPUE %>%
  filter(!is.na(days_fishing) & days_fishing > 0 & weight_kg_corrected > 0)
# 2769093 obs., 52 vars.

max(GoC_Landings_for_CPUE$CPUE_kg.day) # 1400000 (Oysters in Baja, no landing site), followed by tuna in Sinaloa
min(GoC_Landings_for_CPUE$CPUE_kg.day) #0.00125

subset<- GoC_Landings_for_CPUE %>% filter(CPUE_kg.day >= 288000)
unique(subset$estimated_genus_name)

##### NOTE TO SELF:
# - There is some funky stuff going on with SOME of the end and start dates. Come back to this in the new year.
# - I am not sure what to do about the days with no record of live weight. Exclude these for now.
```

```{r Exploratory: Plot CPUE per trip}
# PLOT CPUE PER TRIP:

CPUE_PerTrip_Distribution<- GoC_Landings_for_CPUE %>% 
  filter(year >= 2003, year <= 2013) %>%
 
  ggplot(aes(x = seq_along(CPUE_kg.day), y = CPUE_kg.day)) +
  geom_point(color = "steelblue", alpha = 0.6, size = 2) +
  geom_hline(yintercept = mean(GoC_Landings_for_CPUE$CPUE_kg.day, na.rm = TRUE),
             color = "red", linetype = "dashed", linewidth = 1) +
  labs(
    x = "# of observations",
    y = "CPUE (kg/day)",
    title = "CPUE Distribution"
  ) +
  
  theme_minimal(base_size = 14)

CPUE_PerTrip_Distribution

  # Save as PNG
  ggsave(
    filename = sprintf("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_CPUE_Scores/Distribution_of_CPUEperTrip_2003.2013_Dec112025.png"),
    plot = CPUE_PerTrip_Distribution,
    width = 12,
    height = 12,
    dpi = 300
  )
```

```{r Let's explore the landings for which IHH_major_group equals NA}
# TEST<- GoC_Landings_for_CPUE %>% filter(is.na(IHH_functional_major_groups_sm))
# # 180324 out of 2787363 total records contain no IHH_functional_major_groups_sm (this equate to about ~6-7% of records).
# 
# TEST2<- TEST %>% 
#   select(recorded_taxa_group, recorded_taxa_name_clean, common_name_clean, keywords_1:keywords_5) %>% 
#   distinct()
# 
# # write.csv(TEST2, "/home/ferrer/ohi_goc_prep/general/Species_Keys/NoIHHClass_SpKey_UNDERCONSTRUCTION_Dec022025.csv", row.names = FALSE)
# 
# rm(TEST, TEST2)
```
NOTE TO SELF: Interesting. Most things could probably be grouped to the IHH group level with a little bit of extra leg work... COME BACK TO THIS IN THE NEW YEAR.

```{r Re-assigning CPUE major groups based on new key}
Genus_MajorGroup_Key<- read.csv("~/ohi_goc_prep/general/CONAPESCA_Keys/Genus_to_MajorGroup_Key_Dec112025.csv")
# 326 obs., 4 vars.

GoC_Landings_for_CPUE <- GoC_Landings_for_CPUE %>%
  left_join(Genus_MajorGroup_Key,
            by = c("recorded_taxa_group", "estimated_genus_name"))
# 2769093 obs., 54 vars. as expected!
```

```{r Estimate CPUE for each Major Functional Group per region per year versus "historic" baseline}
# Turn off scientific notation:
options(scipen = 999)

# STEP 1: ESTABLISH CPUE BASELINE VALUES, 2003-2013 -----
# Where "baseline" equals the average Group-level CPUE from 2003-2013
CPUE_baseline <- GoC_Landings_for_CPUE %>%
  filter(year >= 2003, year <= 2013) %>%
  filter(!is.na(OHI_Region_No)) %>%
  group_by(OHI_Region_No, Major_Functional_Group) %>%
  summarise(
    OLD_CPUE_2003.2013 = mean(CPUE_kg.day, na.rm = TRUE),
    NEW_CPUE_2003.2013 = sum(weight_kg_corrected) / sum(days_fishing),
    .groups = "drop"
  )

# CPUE_baseline <- CPUE_baseline %>%
#   mutate(
#     Percent_Diff = (NEW_CPUE_2003.2013 - OLD_CPUE_2003.2013) / OLD_CPUE_2003.2013
#   )
# 
# # NOTE TO SELF: Save this table to show to Mel, can delete later -
# write.csv(CPUE_baseline, "/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_CPUE_Scores/CPUE_MethodsIllustration_PURG_Dec112025.csv", row.names = FALSE)


# STEP 2: CALCULATE CPUE VALUES FOR RECENT YEARS, 2014-2024 -----
CPUE_recent <- GoC_Landings_for_CPUE %>%
  filter(year >= 2014, year <= 2024) %>%
  filter(!is.na(OHI_Region_No)) %>%
  group_by(year, OHI_Region_No, Major_Functional_Group) %>%
  summarise(
    Total_Biomass_kg = sum(weight_kg_corrected),
    Total_Days = sum(days_fishing),
    OLD_mean_CPUE.recent = mean(CPUE_kg.day, na.rm = TRUE),
    NEW_mean_CPUE.recent = ifelse(Total_Days > 0, Total_Biomass_kg / Total_Days, NA_real_),
    Total_Value_mxn = sum(value, na.rm = TRUE),  # keep na.rm just in case value has NAs
    n_records = n(),
    .groups = "drop"
  )

# # NOTE TO SELF: Save this table to show to Mel, can delete later -
# write.csv(CPUE_recent, "/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_CPUE_Scores/CPUE_MethodsIllustration2_PURG_Dec112025.csv", row.names = FALSE)


# STEP 3: COMPUTE PERCENT BIOMASS FOR RECENT YEARS -----
CPUE_recent <- CPUE_recent %>%
  group_by(year, OHI_Region_No) %>%
  mutate(
    total_biomass_state_year = sum(Total_Biomass_kg, na.rm = TRUE),
    pct_of_total_biomass = (Total_Biomass_kg / total_biomass_state_year) * 100
  ) %>%
  ungroup()


# Quick aside: how much of percent catch do the NAs account for?
subset<- CPUE_recent %>% filter(is.na(Major_Functional_Group))
# In only a handful of cases do NAs account for the *majority* of annual catch - primarily in Region 5.
# And again, a lot of these NAs can be accounted for with further species key work.


# STEP 4: MERGE BASELINE TO PRESENT DATA -----
CPUE_Table <- CPUE_recent %>%
  left_join(CPUE_baseline,
            by = c("OHI_Region_No", "Major_Functional_Group")) %>%
  mutate(
    RatioCPUE_Recent.Baseline = NEW_mean_CPUE.recent / NEW_CPUE_2003.2013  # Score = CPUEpresent / CPUE_2003.2013
  )

# STEP 5: DESIRED TABLE -----
CPUE_Table <- CPUE_Table %>%
  transmute(
    OHI_Region = OHI_Region_No,
    Year = year,
    Major_Functional_Group = Major_Functional_Group,
    RatioCPUE_Recent.Baseline = round(RatioCPUE_Recent.Baseline, 3),
    Total_Landings_kg = round(Total_Biomass_kg, 1),
    Taxa_per_Total_Catch = round(pct_of_total_biomass, 2)
  ) %>%
  arrange(OHI_Region, Year, Major_Functional_Group)


# STEP 6: CAP CPUE AT A MAX OF 1 per GROUP
CPUE_Table <- CPUE_Table %>%
  mutate(
    RatioCPUE_Recent.Baseline_CAPPED = if_else(
      RatioCPUE_Recent.Baseline > 1,
      1,
      RatioCPUE_Recent.Baseline
    )
  )

# STEP 7: OHI CPUE SCORES PER REGION PER YEAR -----
OHI_CPUE_Scores <- CPUE_Table %>%
  mutate(TaxaProp = Taxa_per_Total_Catch/100) %>%     # IF % values are 0–100; if 0–1 then skip this line
  group_by(Year, OHI_Region) %>%
  summarise(
    weighted_CPUE_acrossAllGroups = sum(RatioCPUE_Recent.Baseline_CAPPED * TaxaProp, na.rm = TRUE)
  ) %>%
  ungroup()
```

# SAVE: OHI CPUE SCORES PER REGION PER YEAR
```{r Save this CPUE data table}
# write.csv(OHI_CPUE_Scores, "/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_CPUE_Scores/OHI_CPUEScores_SECONDDRAFT_Dec12025.csv", row.names = FALSE)
```

```{r Plot CPUE Scores over time}
#### PLOT CPUE OVER TIME ####

#### State Colors ####
state_colors <- c(
  "BAJA CALIFORNIA" = "royalblue",
  "BAJA CALIFORNIA SUR" = "skyblue",
  "SONORA" = "orange",
  "SINALOA" = "sienna1",
  "NAYARIT" = "plum", 
  "JALISCO" = "purple4"
)

Region_colors <- c(
  "royalblue",
  "skyblue",
  "orange",
  "sienna1",
  "plum",
  "purple4",
  "aquamarine3",
  "gold2",
  "tomato"
)

Regions_of_Interest <- c(
  "Region_1", "Region_2", "Region_3",
  "Region_4", "Region_5", "Region_6",
  "Region_7", "Region_8", "Region_9"
)

OHI_CPUEscores_overTime <- OHI_CPUE_Scores %>%
  filter(OHI_Region %in% Regions_of_Interest) %>%
  ggplot(aes(
    x = Year,
    y = weighted_CPUE_acrossAllGroups,
    color = OHI_Region
  )) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ OHI_Region, scales = "fixed") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
  
  scale_color_manual(values = Region_colors) +
  
  labs(
    title = "FIS - Regional CPUE Scores (2014–2024) relative to Baseline (2003–2013)",
    y = "CPUE Scores (Ratio of CPUE_present to CPUE_2003.2013 capped at 1)",
    x = "Year"
  ) +
  ylim(0, 1.25) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 18, face = "bold")
  )

OHI_CPUEscores_overTime
  
  # # Save as PNG
  # ggsave(
  #   filename = sprintf("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_CPUE_Scores/OHI_CPUEScores_overTime_Dec112025.png"),
  #   plot = OHI_CPUEscores_overTime,
  #   width = 12,
  #   height = 12,
  #   dpi = 300
  # )

```

## Take a closer look at CPUE per Region and per Major_Functional_Group:
```{r Plot CPUE over all years 2003-2024}
CPUE_AllYears <- GoC_Landings_for_CPUE %>%
  filter(year >= 2003, year <= 2024) %>%
  filter(!is.na(OHI_Region_No)) %>%
  group_by(OHI_Region_No, year, Major_Functional_Group) %>%
  summarise(
    Total_Biomass_kg = sum(weight_kg_corrected, na.rm = TRUE),
    Total_Value_mxn = sum(value, na.rm = TRUE),
    Total_Days = sum(days_fishing, na.rm = TRUE),
    Avg_CPUE = mean(CPUE_kg.day, na.rm = TRUE),
    n_records = n(),
    .groups = "drop"
  )

Taxa_colors <- c("darkblue", "royalblue", "skyblue", "orange", "sienna1", "plum", "purple4", "aquamarine3", "gold2", "tomato", "grey")

Region_colors <- c("royalblue", "skyblue", "orange", "sienna1", "plum", "purple4", "aquamarine3", "gold2", "tomato")

Regions_of_Interest <- c(
  "Region_1", "Region_2", "Region_3",
  "Region_4", "Region_5", "Region_6",
  "Region_7", "Region_8", "Region_9"
)


# PLOT 1: CPUE overTime per Region, facet by Major_Functional_Group
CPUE_overTime_byRegion_perTaxa<- CPUE_AllYears %>%
  filter(OHI_Region_No %in% Regions_of_Interest) %>%
  ggplot(
       aes(x = year, 
           y = Avg_CPUE, 
           group = OHI_Region_No, 
           color = OHI_Region_No)) +
  geom_line(size = 1) +                # lines connecting the points
  facet_wrap(~ Major_Functional_Group, scales = "free_y") +  # facet by region
  labs(
    title = "Average CPUE Over Time by Functional Group",
    x = "Year",
    y = "Avg CPUE (kg/day)",
    color = "Functional Group"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_color_manual(values = Taxa_colors) +  # nice palette
  geom_vline(xintercept = 2013, linetype= "solid", color = "grey5")

CPUE_overTime_byRegion_perTaxa

  # # Save as PNG
  # ggsave(
  #   filename = sprintf("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_CPUE_Scores/CPUE_overTime_byRegion_perTaxa_Dec092025.png"),
  #   plot = CPUE_overTime_byRegion_perTaxa,
  #   width = 12,
  #   height = 12,
  #   dpi = 300
  # )


# PLOT 2: CPUE overTime per Major_Functional_Group, facet by Region
CPUE_overTime_byRegion_perTaxa<- CPUE_AllYears %>%
  filter(OHI_Region_No %in% Regions_of_Interest) %>%
  ggplot(
       aes(x = year, 
           y = Avg_CPUE, 
           group = Major_Functional_Group, 
           color = Major_Functional_Group)) +
  geom_line(size = 1) +                # lines connecting the points
  facet_wrap(~ OHI_Region_No, scales = "free_y") +  # facet by region
  labs(
    title = "Average CPUE Over Time by Functional Group",
    x = "Year",
    y = "Avg CPUE (kg/day)",
    color = "Functional Group"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_color_manual(values = Taxa_colors) +  # nice palette
  geom_vline(xintercept = 2013, linetype= "solid", color = "grey5")

CPUE_overTime_byRegion_perTaxa

  # # Save as PNG
  # ggsave(
  #   filename = sprintf("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_CPUE_Scores/CPUE_overTime_byTaxa_perRegion_Dec092025.png"),
  #   plot = CPUE_overTime_byTaxa,
  #   width = 12,
  #   height = 12,
  #   dpi = 300
  # )


# PLOT 3: CPUE overTime per Major_Functional_Group, facet by Region - normalized 0 to 1
CPUE_overTime_byRegion_perTaxa_normalized<- CPUE_AllYears %>%
  filter(OHI_Region_No %in% Regions_of_Interest) %>%
  # Normalize Avg_CPUE per functional group across all years
  group_by(Major_Functional_Group) %>%
  mutate(Avg_CPUE_norm = (Avg_CPUE - min(Avg_CPUE, na.rm = TRUE)) /
                            (max(Avg_CPUE, na.rm = TRUE) - min(Avg_CPUE, na.rm = TRUE))) %>%
  ungroup() %>%
  ggplot(aes(
    x = year, 
    y = Avg_CPUE_norm, 
    group = Major_Functional_Group, 
    color = Major_Functional_Group
  )) +
  geom_line(size = 1) +
  facet_wrap(~ OHI_Region_No, scales = "free_y") +
  labs(
    title = "Normalized CPUE Over Time by Functional Group",
    x = "Year",
    y = "Normalized CPUE (0-1)",
    color = "Functional Group"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_color_manual(values = Taxa_colors) +
  geom_vline(xintercept = 2013, linetype = "solid", color = "grey5")

CPUE_overTime_byRegion_perTaxa_normalized

  # Save as PNG
  # ggsave(
  #   filename = sprintf("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_CPUE_Scores/CPUE_overTime_byTaxa_perRegion_normalized_Dec092025.png"),
  #   plot = CPUE_overTime_byTaxa_normalized,
  #   width = 12,
  #   height = 12,
  #   dpi = 300
  # )
```

```{r Take a closer look at CPUE per Region and per Taxa}
# Region 5
subset<- GoC_Landings_for_CPUE %>% 
  filter(OHI_Region_No == "Region_5" & is.na(IHH_functional_major_groups_sm))

table(subset$recorded_taxa_name)
table(subset$recorded_taxa_group)
# Again, a lot of these NAs can be accounted for with further species key work.

# Region 7
subset<- GoC_Landings_for_CPUE %>% 
  filter(OHI_Region_No == "Region_7" & is.na(IHH_functional_major_groups_sm))

table(subset$recorded_taxa_name)
table(subset$recorded_taxa_group)
```

# SCORE B ------------------------------
# >> PICK 1: RAM Stock Assessment <<
```{r Load in RAM and formal assessment data}
RAM_data_original<- read.csv("/home/frazier/ohi_goc_prep/FP/FIS/v2025/data/formal_assessments.csv")
table(RAM_data_original$year)
# 195 obs., 17 vars.
```
```{r Separate Genus from scientific_name}
RAM_data_cleaned<- RAM_data_original

RAM_data_cleaned <- RAM_data_cleaned %>%
  separate(scientificname, into = c("Genus", "Species"), sep = " ", remove = FALSE)
# Okay, looks good!
# 195 obs., 19 vars.
```

```{r Basic filtering & correcting}
# Keep Stock (genus-species in this case) only if it has post-2019 data
RAM_data_cleaned <- RAM_data_cleaned %>%
  group_by(Genus, Species) %>%
  filter(any(year >= 2019)) %>%
  select(year, assessyear, areaname, 
         commonname, scientificname, Genus, Species, 
         ram_bmsy, source) %>%
  ungroup()
# 181 obs., 9 vars.

# Correct the non_numeric values:
RAM_data_cleaned$RAM_BBmsy_numeric <- as.numeric(gsub("[^0-9.]", "", RAM_data_cleaned$ram_bmsy))
```

```{r Gapfill RAM data using the zoo package}
RAM_data_gapfilled <- RAM_data_cleaned %>%
  group_by(Genus, Species, areaname) %>%
  complete(year = 2019:2024) %>%
  ungroup()
# 256 obs., 10 vars.

# Gapfilling - INTERPOLATION between years with data (using the zoo package):
RAM_data_gapfilled <- RAM_data_gapfilled %>%
  # --- identify each stock correctly ---
  group_by(Genus, Species, areaname) %>%
  
  # --- add missing years WITHOUT removing any original rows ---
  complete(year = full_seq(c(year, 2019:2024), 1)) %>%
  # ^ This safely adds 2019–2024 and any intermediate years, 
  #   never removes rows
  
  # --- fill descriptors (stock metadata) ---
  fill(commonname, scientificname, source, assessyear, .direction = "downup") %>%
  
  arrange(year, .by_group = TRUE) %>%
  
  # --- interpolate only RAM_BBmsy_numeric ---
  mutate(
    RAM_BBmsy_gapfilled = zoo::na.approx(RAM_BBmsy_numeric, x = year, na.rm = FALSE)
  ) %>%
  
  ungroup()
# 347 obs., 11 vars.

# Gapfilling - EXTRAPOLATION beyond years with data (using the zoo package):
RAM_data_gapfilled <- RAM_data_gapfilled %>%
  group_by(Genus, Species, areaname) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(
    RAM_BBmsy_gapfilled = zoo::na.approx(
      RAM_BBmsy_numeric,
      x = year,
      na.rm = FALSE,
      rule = 2   # allow linear extrapolation outside observed years
    )
  ) %>%
  ungroup()
# 347 obs., 11 vars.

# Sanity check: do gapfilled values always match numeric values when they exist?
# Example vectors
numeric_BBmsy_values<- RAM_data_gapfilled$RAM_BBmsy_numeric
gapfilled_BBmsy_values <- RAM_data_gapfilled$RAM_BBmsy_gapfilled

# Logical check: TRUE wherever BBmsy_numeric is NA or matches RAM_BBmsy_gapfilled
check <- ifelse(!is.na(numeric_BBmsy_values), gapfilled_BBmsy_values == numeric_BBmsy_values, TRUE)

# Are all original values preserved?
all(check)
# Yes!

# Finally, a "RAM_BBmsy_Pick" column and "Pick_Source" column that tell which BBmsy value to work with and where that datapoint came from, a formal assessment value or a gapfilled value:
RAM_data_gapfilled <- RAM_data_gapfilled %>%
  mutate(
    RAM_BBmsy_Pick = ifelse(!is.na(RAM_BBmsy_numeric), RAM_BBmsy_numeric, RAM_BBmsy_gapfilled),
    Pick_Source    = ifelse(!is.na(RAM_BBmsy_numeric), "Direct from formal assessment (RAM or other)", "gapfilled from formal assessment")
  )
# 347 obs., 13 vars.
```

```{r BBmsy cap at 1}
# Caps values for RAM_BBmsy_Pick at 1 & filters out 2025 data:
RAM_data_gapfilled<- RAM_data_gapfilled %>%
  filter(year <= 2024) %>%
  mutate(Formal_BBmsy_capped = pmin(RAM_BBmsy_Pick, 1)) %>%
  select(Year= year, Area= areaname, CommonName= commonname,
         ScientificName= scientificname, Genus, Species,
         FormalAssessment_Source = source, FormalAssessment_Years= assessyear,
         Formal_BBmsy= ram_bmsy, Formal_BBmsy_numeric= RAM_BBmsy_numeric,
         Formal_BBmsy_gapfilled = RAM_BBmsy_gapfilled,
         Formal_BBmsy_Pick = RAM_BBmsy_Pick, Formal_BBmsy_capped,
         Formal_BBmsy_Pick_Source = Pick_Source)
# 346 obs., 14 vars.
```

```{r Save this dataframe}
write.csv(RAM_data_gapfilled,"~/ohi_goc_prep/FP/FIS/v2025/OHI_StockStatus_Scores/OHI_RAM_Scores/Gapfilled_RAMdata_Dec122025.csv", row.names = FALSE)
```


# >> PICK 2: CMSY Stock Assesments <<
Must run CMSY on personal machine. Can generate the main files here though. 

```{r Review the CMSY protocol developed by Ferrer et al. (2021)}
# Absolute path to your PDF
pdf_path <- normalizePath(
  "~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Copy of Ferrer et al. (2021) CMSY Protocol.pdf",
  winslash = "/", mustWork = TRUE
)

# Convert to file URL
file_url <- paste0("file:///", URLencode(pdf_path, reserved = TRUE))

# Open in default browser
browseURL(file_url)
```

# GENERATING FILES 1 & 2 FROM SCRATCH

NOTES TO SELF - Dec 09 2025:

FILE 1: Create a file with annual landings in kg (and n_count observations) grouped by year per "stock". Here, stock is defined as the total population of GENUS X fished within the Gulf of California from Baja to northern Jalisco, not including landings from the Pacific.

- At present, we have had some issues with keying out species and landing_sites. Thus, whatever timeseries we have here is likely to be an UNDERESTIMATE of total landings for all genera in the GoC. I plan to re-make this file (File 1) in the new year once we have more key info.

- For now, I’m going to run CMSY at the genus level for now, since I actually think it’s a safer more reliable bet than species-level info and because – in many cases – fishermen seem to target taxa of the same genus interchangeably.

FILE 2: Create a file with OHI_Region, Stock, Group = Name, English Name, Scientific Name, Source, and a whole bunch of year columns that you populate using the various R files shared by Alfredo. This file can probably be trimmed down but for now I recommend formatting it exactly as it appears from before.

# Create FILE 1 - Timeseries
```{r File 1 - Stock Timeseries (JC data)}
# Load file, only if needed:
GoC_Landings_for_CMSY<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/fis/v2025/int/GOC_FIS.landings_CLEAN_as.of.Dec092025.rds"))

GoC_Landings_for_CMSY_original<- GoC_Landings_for_CMSY

# GoC_Landings_for_CMSY<- Landings_GoC_NoFresh.NoMariculture
# 3163120 obs., 50 vars.
```

```{r Basic checks ahead of CMSY}
# Some basic checks:
table(GoC_Landings_for_CMSY$fleet)
# large_scale small_scale 
#      142404      975060 

table(GoC_Landings_for_CMSY$acuaculture_production) 
# no aquaculture. As expected!

table(GoC_Landings_for_CMSY$freshwater_flag) 
# no freshwater flags. As expected!

table(GoC_Landings_for_CMSY$GoC) 
table(GoC_Landings_for_CMSY$GulfCal) 
# Everything in Gulf of California, that we know of!

table(GoC_Landings_for_CMSY$year)
# Plenty of data years 2000-2025

unique(GoC_Landings_for_CMSY$estimated_genus_name) 
# Okay. There are a few genera we're going to toss...
```

```{r Basic filters ahead of CMSY}
#### REMOVE FRESHWATER GENERA WE ARE NOT INTERESTED ####
genus_exclude<- c(
  "Cyprinus", # common carp; strictly freshwater.
  "Macrobrachium", # freshwater prawns/shrimps; mostly freshwater.
  "Hypophthalmichthys", # Asian carp; freshwater.
  "Chirostoma", # “charales,” endemic to freshwater lakes in Mexico.
  "Carassius", # goldfish/crucian carp; freshwater.
  "Micropterus", # black bass; freshwater.
  "Lepomis", # sunfish; freshwater.
  "Cyprinodon", # pupfish; freshwater.
  "Rana", # frogs
  "NEI", # not classed yet
  "Especies", # not specific enough to be anything
  NA # not classed
  )

# Flag genera to exclude:
GoC_Landings_for_CMSY <- GoC_Landings_for_CMSY %>%
  mutate(
    exclude_flag = if_else(
      !is.na(estimated_genus_name) & estimated_genus_name %in% genus_exclude,
      TRUE,
      FALSE
    )
  )

subset<- GoC_Landings_for_CMSY %>% filter(exclude_flag== TRUE)
# Dropping 710533 out of 3163120 total records (~22%) for now. A lot of these are NEI that will be classed in the new year!

GoC_Landings_for_CMSY<- GoC_Landings_for_CMSY %>% filter(exclude_flag== FALSE)
# 2452587 obs., 51 vars.


#### REMOVE RARE GENERA ####
table(GoC_Landings_for_CMSY$estimated_genus_name)
# Some of these genera don't appear very often. I want to filter genera for which we have more than 10 years worth of data and > 10 records per year.

# Summarize per genus and year
Genus_summary <- GoC_Landings_for_CMSY %>%
  filter(!is.na(estimated_genus_name)) %>%
  group_by(estimated_genus_name, year) %>%
  summarise(
    n_records = n(),
    .groups = "drop"
  )

# Count how many years have more than 10 records, and remove those genera with fewer than 10 years of 10 records:
Genus_filtered <- Genus_summary %>%
  filter(n_records > 10) %>%
  group_by(estimated_genus_name) %>%
  summarise(n_years = n(), .groups = "drop") %>%
  filter(n_years > 10)

Genera_of_Interest<- Genus_filtered$estimated_genus_name
# 98 remaining genera

# Keep only genera that meet the criteria in the main dataset
GoC_Landings_for_CMSY <- GoC_Landings_for_CMSY %>%
  filter(estimated_genus_name %in% Genera_of_Interest)
# 2225315 obs., 51 vars.

#### Filter out OHI Region = NA ####
sum(is.na(GoC_Landings_for_CMSY$OHI_Region_No))
# Not applicable, all entries have an OHI Region number.
```

```{r Mitigate genera key issues}
# NOTE TO SELF: Clean this up in the new year. This is a band aid solution to larger key issues.

#### GENUS KEY ####
# NOTE TO SELF: Return to this when needed (Dec 09 2025). Returned the next day, Dec 10 2025.
Genus_MajorGroup_Key <- GoC_Landings_for_CMSY %>%
  select(recorded_taxa_group, estimated_genus_name, IHH_functional_major_groups_sm) %>%
  distinct(recorded_taxa_group, estimated_genus_name, IHH_functional_major_groups_sm, .keep_all = TRUE)
# 161 obs.

# See the issue here? One genus could be associated with multiple recorded_taxa_groups. We need the unique combination of estimated_genus_name & recorded_taxa_group to key out the larger dataset. But, in the end, we'll want to be able to group by ESTIMATED_taxa_group to avoid issues of splitting the catch from a single genus across two larger groups.

# ---- Create Major_Functional_Group values based IHH major functional groups assigned to recorded_taxa_group ----

# Step 1: Assign one Major_Functional_Group per recorded_taxa_group, respecting the "OTRAS" category as inherently different.
Genus_MajorGroup_Key <- Genus_MajorGroup_Key %>%
  group_by(recorded_taxa_group) %>%
  mutate(
    # Step 1: assign group-level functional group only for non-OTRAS
    group_major = {
      tbl <- table(IHH_functional_major_groups_sm)
      tbl <- tbl[!is.na(names(tbl)) & names(tbl) != ""]
      if(length(tbl) == 0) NA_character_ else names(tbl)[which.max(tbl)]
    },
    Major_Functional_Group = if_else(
      recorded_taxa_group != "OTRAS",
      group_major,
      NA_character_
    )
  ) %>%
  ungroup() %>%
  select(-group_major) %>%
  
  # Step 2: assign genus-level functional group (including OTRAS)
  group_by(estimated_genus_name) %>%
  mutate(
    genus_major = {
      # Use either already assigned Major_Functional_Group or fallback to IHH_functional_major_groups_sm
      tbl <- table(
        c(
          Major_Functional_Group[!is.na(Major_Functional_Group)],
          IHH_functional_major_groups_sm[!is.na(IHH_functional_major_groups_sm)]
        )
      )
      if(length(tbl) == 0) NA_character_ else names(tbl)[which.max(tbl)]
    },
    Major_Functional_Group = if_else(
      is.na(Major_Functional_Group),
      genus_major,
      Major_Functional_Group
    )
  ) %>%
  ungroup() %>%
  select(-genus_major)
  
# Step 3: Enforce genus-level consistency for Major_Functional_Group
Genus_MajorGroup_Key <- Genus_MajorGroup_Key %>%
  group_by(estimated_genus_name) %>%
  mutate(
    # Only apply to genera that appear more than once
    Major_Functional_Group = if(n() > 1) {
      # Choose the most frequent functional group within the genus
      tbl <- table(Major_Functional_Group)
      tbl <- tbl[names(tbl) != "" & names(tbl) != "NA"]
      if(length(tbl) == 0) NA_character_ else names(tbl)[which.max(tbl)]
    } else {
      Major_Functional_Group
    }
  ) %>%
  ungroup()
  
# Step 4: Manual patches for specific taxonomic groups
Genus_MajorGroup_Key <- Genus_MajorGroup_Key %>%
  mutate(
    Major_Functional_Group = case_when(
      recorded_taxa_group == "ALGAS" ~ "Marine algae",
      recorded_taxa_group %in% c("PETO", "TIBURON", "CAZON") ~ "Sharks",
      TRUE ~ Major_Functional_Group
    )
  )

# Step 5: Assign estimated_taxa_group by looking at the most common recorded_taxa_group for each genus
Genus_MajorGroup_Key <- Genus_MajorGroup_Key %>%
  group_by(estimated_genus_name) %>%
  mutate(
    # Count the most frequent recorded_taxa_group for this genus
    most_common_group = {
      tbl <- table(recorded_taxa_group)
      names(tbl)[which.max(tbl)]
    },
    estimated_taxa_group = most_common_group
  ) %>%
  ungroup() %>%
  select(-most_common_group)
    
# 161 obs., 4 vars.

# TEST<- Genus_MajorGroup_Key %>%
#   select(estimated_genus_name,  Major_Functional_Group) %>%
#   distinct()
# 98 obs. = 98 genera, as expected.

Genus_MajorGroup_Key<- Genus_MajorGroup_Key %>% 
  select(recorded_taxa_group, estimated_genus_name, estimated_taxa_group, Major_Functional_Group) %>% 
  distinct()
# 134 obs., 3 vars.

# MERGE this key to the larger dataset
GoC_Landings_for_CMSY <- left_join(
  GoC_Landings_for_CMSY,
  Genus_MajorGroup_Key,
  by = c("recorded_taxa_group", "estimated_genus_name")
)
# 2225315 obs., 53 vars. as expected!

# A few quick sanity checks:
subset<- GoC_Landings_for_CMSY %>% filter(is.na(Major_Functional_Group))
sum(is.na(GoC_Landings_for_CMSY$Major_Functional_Group))
table(GoC_Landings_for_CMSY$estimated_taxa_group)
sum(is.na(GoC_Landings_for_CMSY$estimated_taxa_group))
```

```{r Save GoC_Landings_for_CMSY to be used for weighting scores later}
# I want to maintain this exact dataframe since it was filtered, etc.
# saveRDS(GoC_Landings_for_CMSY, "/home/shares/ohi/OHI_GOC/goal_prep/fp/fis/v2025/int/GOC_Landings_for_CMSY_Dec102025.rds")
```

```{r Summary table}
# Turn off scientific notation:
options(scipen = 999)

# Create timeseries FILE 1:
colnames(GoC_Landings_for_CMSY)
unique(GoC_Landings_for_CMSY$estimated_genus_name)

TotalLandings_AnnualSummary_byGenus <- GoC_Landings_for_CMSY %>%
  select(year, estimated_genus_name, weight_kg_corrected) %>%
  group_by(year, estimated_genus_name) %>%
  summarise(
    Total_Landed_Biomass = sum(weight_kg_corrected, na.rm = TRUE),
    n_Records = n(),
    .groups = "drop"  # optional: ungroup after summarising
  )
# 2401 obs., 4 vars.

TotalLandings_AnnualSummary_byGenus <- GoC_Landings_for_CMSY %>%
  select(year, Major_Functional_Group, estimated_taxa_group, estimated_genus_name, weight_kg_corrected) %>%
  group_by(year, Major_Functional_Group, estimated_taxa_group, estimated_genus_name) %>%
  summarise(
    Total_Landed_Biomass = sum(weight_kg_corrected, na.rm = TRUE),
    n_Records = n(),
    .groups = "drop"  # optional: ungroup after summarising
  )
# 2401 obs., 6 vars. GOOD!!!

table(TotalLandings_AnnualSummary_byGenus$Major_Functional_Group)
#                   Cephalopods                   Crustaceans        Demersal marine fishes             Diadromous fishes                  Marine algae 
#                            26                            41                          1153                            26                            23 
# Miscellaneous aquatic animals    Molluscs excl. Cephalopods         Pelagic marine fishes                        Sharks               Shrimps, prawns 
#                            44                           222                           472                           269                           125 
# Lots of demersal marine fishes!

File1_df<- TotalLandings_AnnualSummary_byGenus

# Re-name column
colnames(File1_df)[colnames(File1_df) == "estimated_genus_name"] <- "Stock"
colnames(File1_df)[colnames(File1_df) == "year"] <- "Year"

# Finally, and just for now, filter out data from 2025
# NOTE TO SELF: Come back to this in subsequent updates, once we have full data for 2025.
File1_df<- File1_df %>% filter(Year<2025)

CMSY_GenusKey<- File1_df %>% 
  select(Major_Functional_Group, estimated_taxa_group, Stock) %>%
  distinct()
```

## SAVE FILE 1
```{r SAVE FILE 1 & CMSY Genus Key}
# 
write.csv(File1_df, "~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Input_Files/File1_Dec092025.csv", row.names = FALSE)
# write.csv(CMSY_GenusKey, "~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Instructional_Files/CMSY_GenusKey_Dec102025.csv", row.names = FALSE)
```

# Create FILE 2 - Metadata & Model Priors
```{r Create File 2 based on File 1}
# STEP 1:
# Which taxa (stocks) do we want to account for in our File 2?
Stock_names<- unique(File1_df$Stock)
# 98 unique genera / stocks

## Read in the column names of File 2 from Ferrer et al. (2022) CSV file
headers <- names(read.csv("FIS/v2025/OHI_CMSY_Scores/Instructional_Files/CMSY_File2_BluePrint_Dec092025.csv", nrows = 1))

## Create an empty dataframe with the same headers
File2_df <- data.frame(matrix(ncol = length(headers), nrow = 0))
colnames(File2_df) <- headers

## Add Stock_names to the 'Stock' column
File2_df <- File2_df[1:length(Stock_names), ]  # Expand rows
File2_df$Stock <- Stock_names

## Reset rownames / the index column
rownames(File2_df) <- NULL


# STEP 2: TRIME AND RENAME AS NECESSARY
# Trim some of the columns off this file: only if necessary.

# Re-name columns: only if necessary.
# colnames(File2_df)[colnames(File2_df) == "Keep"] <- "Yes_Run"
# colnames(File2_df)[colnames(File2_df) == "stb.low"] <- "st_b.low"
# colnames(File2_df)[colnames(File2_df) == "stb.hi"] <- "st_b.hi"


# STEP 3: FILL AS NECESSARY
# Fill in some of the columns off this file:
File2_df$Region <- c("GOC")
File2_df$Class <- c("TBD")
File2_df$Source <- c("JC_data")
File2_df$btype <- c("None")
File2_df$force.cmsy <- c("TRUE")


# STEP 4:
# For species with less than 10 years of data, do not keep (Sufficient_No_Years = 0)
StockYears_per_Taxa <- File1_df %>%
  group_by(Stock) %>%
  summarise(n_Stock_Years = n_distinct(Year), .groups = "drop")

StockYears_per_Taxa <- StockYears_per_Taxa %>%
  mutate(Sufficient_No_Years = ifelse(n_Stock_Years < 10, 0, 1))

# How many stocks will we be keeping based on number of years?
table(StockYears_per_Taxa$Sufficient_No_Years)
# All of them, which makes sense since we already applied this filter above.

##  Add this information to the File 2 dataframe: join StockYears_per_Taxa to File2_df by "Stock", bringing in the filled Sufficient_No_Years column
File2_df <- File2_df %>%
  select(-Sufficient_No_Years) %>%  # Remove empty column to avoid duplicates
  left_join(
    StockYears_per_Taxa %>% select(Stock, Sufficient_No_Years),
    by = "Stock"
  )

# STEP 5: ADD IN THE GENUS KEY INFO
CMSY_GenusKey_distinct<- Genus_MajorGroup_Key %>% 
  select(estimated_genus_name, estimated_taxa_group) %>%
  distinct()

File2_df <- File2_df %>%
  left_join(
    CMSY_GenusKey_distinct,
    by = c("Stock"="estimated_genus_name")
  )

# STEP 6:
# Re-order columns: only if necessary.
```

## SAVE FILE 2 BLUE PRINT
```{r SAVE FILE 2 UNDER CONSTRUCTION}
write.csv(File2_df, "~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Input_Files/File2_UNDER_CONSTRUCTION_Dec092025.csv", row.names = FALSE)
```
NEXT STEPS: Fill in the rest of file 2, namely the model priors, using the ancillary files shared with me by Alfredo (Ferrer et al., 2022)
NOTE TO SELF: These model priors are likely to change in the new year but will do for now.

Did some work offline on my local machine (not available for viewing here... yet). This work will help fill out File 2.

```{r Read in additional files to complete file 2}
# Read in file if necessary:
#File2_df <- read.csv("~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Input_Files/File2_UNDER_CONSTRUCTION_Dec092025.csv")

## Additional files needed to file in the metadata
Priors_Pristine.Biomass <- read.csv("~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Model_Priors/Priors_StartingBiomass_fromFAOdata_DRAFT_Dec092025.csv")
Priors_Int.Year <- read.csv("~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Model_Priors/Priors_IntermediateYears_DRAFT_Dec092025.csv")
Priors_IntEnd.Conditions <- read.csv("~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Model_Priors/Priors_IntEndConditions_DRAFT_Dec092025.csv")
Priors_Resilience <- read.csv("~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Model_Priors/Priors_ResilienceValues_DRAFT_Dec092025.csv")
Priors_Resilience <- Priors_Resilience %>% select(-ScientificName)
# Common_Names <- read.csv("Stock and Common Names_Oct 23 2025.csv")
```

```{r lowercase_function}
to_lower_df <- function(df) {
  df %>% mutate(across(where(is.character), tolower))
}
```

```{r Stitch these dataframes together to create complete File2_df}
File2_df        <- to_lower_df(File2_df)
Priors_Pristine.Biomass  <- to_lower_df(Priors_Pristine.Biomass)
Priors_Int.Year <- to_lower_df(Priors_Int.Year)
Priors_Resilience <- to_lower_df(Priors_Resilience)

## MERGE: PRISTINE BIOMASS PRIORS
File2_df_merged <- File2_df %>% 
  left_join(
    Priors_Pristine.Biomass %>% select(estimated_taxa_group, st_b.low, st_b.hi),
    by = "estimated_taxa_group",
    suffix = c(".file2", ".priors")
  ) %>%
  mutate(
    st_b.low = coalesce(st_b.low.file2, st_b.low.priors),
    st_b.hi  = coalesce(st_b.hi.file2,  st_b.hi.priors)
  ) %>% 
  select(-ends_with(".file2"), -ends_with(".priors"))

# NOTE TO SELF: Our priors for pristine biomass, and the FAO timeseries that underly them, are fairly sparse as in missing several major taxonomic groups (like sardine). Return to this in the new year.

## MERGE: INT YEAR PRIORS
File2_df_merged <- File2_df_merged %>%
  left_join(
    Priors_Int.Year %>% 
      select(estimated_taxa_group, int_yr, StartYear, EndYear),
    by = "estimated_taxa_group",
    suffix = c(".file2", ".priors")
  ) %>%
  mutate(
    int_yr               = coalesce(int_yr.file2,               int_yr.priors),
    StartYear            = coalesce(StartYear.file2,            StartYear.priors),
    EndYear              = coalesce(EndYear.file2,              EndYear.priors)) %>%
  select(-ends_with(".file2"), -ends_with(".priors"))

## MERGE: INT END CONDITIONS PRIORS
# NOTE TO SELF: Based on the naming conventions within this file I actually think these are the priors for intermediate and ending biomass. I'll circle back to this and iron out the true meaning and utility of these particular priors in the new year.
# Join and fill missing values
File2_df_merged <- File2_df_merged %>%
  left_join(
    Priors_IntEnd.Conditions,
    by = "estimated_taxa_group",
    suffix = c("", ".priors")
  ) %>%
  mutate(
    int_b.low  = coalesce(int_b.low, int_b.low.priors),
    int_b.hi   = coalesce(int_b.hi, int_b.hi.priors),
    end_b.low  = coalesce(end_b.low, end_b.low.priors),
    end_b.hi = coalesce(end_b.hi, end_b.hi.priors)
  ) %>%
  select(-ends_with(".priors"))


## MERGE: RESILIENCE PRIORS
File2_df_merged <- File2_df_merged %>%
  left_join(
    Priors_Resilience %>% select(estimated_taxa_group, Resilience, Resilience_Source),
    by = "estimated_taxa_group",
    suffix = c(".file2", ".priors")
  ) %>%
  mutate(
    Resilience = coalesce(Resilience.file2, Resilience.priors)
  ) %>%
  select(-Resilience.file2, -Resilience.priors)
```

```{r Finally lets clean up File 2 before saving and running CMSY}
File2_df_merged<- File2_df_merged %>% 
  select(Region, Stock, Landings_Datasource=Source, 
         StartYear, EndYear, int_yr,
         btype,
         Flim, Fpa, Blim, Bpa, Bmsy, B40,
         FMSY, MSY, MSYBtrigger, M,
         Fofl, last_F,
         q.start, q.end,
         st_b.low, st_b.hi, 
         int_b.low, int_b.hi, end_b.low, end_b.hi,
         r.low, r.hi, 
         Resilience, Resilience_Source,
         Yes_Run, Comment)

# Fill in the Yes_Run column based on model priors:
File2_df_merged <- File2_df_merged %>%
  mutate(
    Yes_Run = if_else(
      is.na(Resilience) | is.na(int_b.low) | is.na(st_b.low) | is.na(end_b.low),
      0,
      1
    ),
    Comment = if_else(
      Yes_Run == 0,
      "Incomplete model priors",
      NA_character_
    )
  )

sum(File2_df_merged$Yes_Run==1)
# 38 stocks to run!

# NOTES ON NAMING, Dec 10 2025:
# Need a btype column = None
# Need a force.cmsy column = TRUE
# StartYear and EndYear must actually align with file one time series.
```

```{r How much of FP will these CMSY stocks account for?}
Landings_CMSY_TEST <- Landings_GoC_NoFresh.NoMariculture %>%
  mutate(estimated_genus_name = tolower(estimated_genus_name))

CMSY_Stocks <- File2_df_merged %>%
  filter(Yes_Run == 1) %>%
  pull(Stock) %>%
  unique()

CMSY_Stocks_hypothetical <- c(
  CMSY_Stocks, 
  "sardinops", 
  "opisthonema", 
  "sarda",
  "scomber"
)

# Calculate total from genera landings identified in the CMSY exercise:
# NOTE TO SELF: Come back to this. Find priors for these four taxa! They represent a huge component of total landings.
Landings_CMSY_summary <- Landings_CMSY_TEST %>%
  filter(year %in% 2015:2024) %>%           # only last 10 years
  mutate(is_CMSY = estimated_genus_name %in% CMSY_Stocks_hypothetical) %>%  # flag CMSY genera
  summarise(
    total_landings = sum(weight_kg_corrected, na.rm = TRUE),                # total landings
    CMSY_landings  = sum(weight_kg_corrected[is_CMSY], na.rm = TRUE)       # landings from CMSY genera
  ) %>%
  mutate(percent_CMSY = 100 * CMSY_landings / total_landings)    # percentage
  
Landings_CMSY_summary
# If I can sort out priors for these four extra txa, CMSY could account for ~50% of all GoC landings in the last five years.
```

## SAVE FILE 2 COMPLETE
```{r SAVE FILE 2 COMPLETE}
write.csv(File2_df_merged, "~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Input_Files/File2_Dec092025.csv", row.names = FALSE)
```

```{r How much of FP will these CMSY stocks account for?}
Landings_CMSY_TEST <- Landings_GoC_NoFresh.NoMariculture %>%
  mutate(estimated_genus_name = tolower(estimated_genus_name))

CMSY_Stocks <- File2_df_merged %>%
  filter(Yes_Run == 1) %>%
  pull(Stock) %>%
  unique()

CMSY_Stocks_hypothetical <- c(
  CMSY_Stocks, 
  "sardinops", 
  "opisthonema", 
  "sarda",
  "scomber"
)

# Calculate total from genera landings identified in the CMSY exercise:
# NOTE TO SELF: Come back to this. Find priors for these four taxa! They represent a huge component of total landings.
Landings_CMSY_summary <- Landings_CMSY_TEST %>%
  filter(year %in% 2015:2024) %>%           # only last 10 years
  mutate(is_CMSY = estimated_genus_name %in% CMSY_Stocks_hypothetical) %>%  # flag CMSY genera
  summarise(
    total_landings = sum(weight_kg_corrected, na.rm = TRUE),                # total landings
    CMSY_landings  = sum(weight_kg_corrected[is_CMSY], na.rm = TRUE)       # landings from CMSY genera
  ) %>%
  mutate(percent_CMSY = 100 * CMSY_landings / total_landings)    # percentage
  
Landings_CMSY_summary
# If I can sort out priors for these four extra txa, CMSY could account for ~50% of all GoC landings in the last five years.
```

# RAN CMSY ON PERSONAL MACHINE. Here are the outputs from that analysis:

```{r Load in CMSY asessment results}
# Actual B/Bmsy output from CMSY:
CMSY_Timeseries_original<- read.csv("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Output_Files/CMSY_Assessment_Timeseries_Dec102025.csv")
# 775 obs., 13 vars.
```

```{r Load other useful dataframes: CMSY timeseries and key data}
# Useful genus key:
CMSY_GenusKey<- read.csv("~/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/Instructional_Files/CMSY_GenusKey_Dec102025.csv")
# 98 obs., 3 vars.

# Landings timeseries that informed the CMSY:
CMSY_LandingsData_original<- readRDS("/home/shares/ohi/OHI_GOC/goal_prep/fp/fis/v2025/int/GOC_Landings_for_CMSY_Dec102025.rds")
# 2225315 obs., 53 vars.
```

```{r Append info about Major_Functional_Group}
CMSY_Timeseries<- CMSY_Timeseries_original %>% select(stock, yr, 
                                             B, B.Bmsy, lcl.B.Bmsy, ucl.B.Bmsy,
                                             F.Fmsy, lcl.F.Fmsy, ucl.F.Fmsy)

# Convert the relevant columns to lower case
CMSY_Timeseries <- CMSY_Timeseries %>%
  mutate(stock = tolower(stock))

CMSY_GenusKey <- CMSY_GenusKey %>%
  mutate(Stock = tolower(Stock))

# Merge in Major_Functional_Group data:
CMSY_Timeseries <- CMSY_Timeseries %>%
  left_join(
    CMSY_GenusKey,
    by = c("stock" = "Stock")
  )
# 775 obs., 11 vars.
```

```{r Summarize landings by genus / stock}
Genera_in_CMSY<- unique(CMSY_Timeseries$stock)
# 31 assessed genera

CMSY_LandingsData <- CMSY_LandingsData_original %>%
  mutate(estimated_genus_name = tolower(estimated_genus_name))

CMSY_LandingsData <- CMSY_LandingsData %>% filter(estimated_genus_name %in% Genera_in_CMSY)
# 1126147 obs., 53 vars
unique(CMSY_LandingsData$estimated_genus_name)
# Yup! Looks good!

# Summarize by year and OHI region:
CMSYstocks_RegionalLandings_Summary <- CMSY_LandingsData %>% 
  filter(year >= 2014, year <= 2024) %>%
  select(year, OHI_Region_No, estimated_genus_name, weight_kg_corrected, value) %>%
  group_by(stock = estimated_genus_name, OHI_Region_No, yr = year) %>%
  summarize(
    total_catch_kg = sum(weight_kg_corrected, na.rm = TRUE),
    total_value_pesos = sum(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(stock, OHI_Region_No, yr)
# 2171 obs., 5 vars.

CMSYstocks_RegionalLandings_Summary <- CMSYstocks_RegionalLandings_Summary %>%
  left_join(
    CMSY_Timeseries,
    by = c("yr", "stock")
  )
# 2171 obs., 14 obs.
```

```{r Now filter to create the final dataframe}
colnames(CMSYstocks_RegionalLandings_Summary)

CMSY_Scores<- CMSYstocks_RegionalLandings_Summary %>% select(yr, OHI_Region_No, stock,
                                                             total_catch_kg, B, B.Bmsy, lcl.B.Bmsy,
                                                             ucl.B.Bmsy,
                                                             F.Fmsy, lcl.F.Fmsy, ucl.F.Fmsy)
```

# SAVE: SAVE B/BMSY SCORES PER GENUS PER REGION PER YEAR
```{r Save this dataframe}
# write.csv(CMSY_Scores, "/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/OHI_CMSY.BBmsyScores_SECONDDRAFT_Dec112025.csv")
```


# >> PICK 3: Catch vs Carta Nacional Pesquera Baseline <<

## Load in recommendations for the Carta Nacional Pesquera
```{r Load in Carta data}
Carta_data_original<- read.csv("/home/frazier/ohi_goc_prep/FP/FIS/v2025/data/CNP_2023_recommendations_clean.csv")
Carta_data<- Carta_data_original
```

```{r Separate Genus from scientific_name}
Carta_data <- Carta_data %>%
  separate(scientific_name, into = c("Genus", "Species"), sep = " ", remove = FALSE)
# Okay, looks good!
# Now we can try to patch in other info...
```

```{r Assigning Carta major groups based on new key}
Genus_MajorGroup_Key<- read.csv("~/ohi_goc_prep/general/CONAPESCA_Keys/Genus_to_MajorGroup_Key_Dec112025.csv")
# 326 obs., 4 vars.

Genus_MajorGroup_Key<- Genus_MajorGroup_Key %>% 
  select(estimated_genus_name, estimated_taxa_group, Major_Functional_Group) %>%
  distinct()
# 288 obs., 3 vars.

Genus_MajorGroup_Key <- Genus_MajorGroup_Key %>%
  filter(!is.na(estimated_genus_name) & estimated_genus_name != "NEI")

Carta_data <- Carta_data %>%
  left_join(
    Genus_MajorGroup_Key,
    by = c("Genus" = "estimated_genus_name")
  )
# 1008 obs., 13 vars. As expected!
```

```{r Make a CARTA CONTROL KEY}
Carta_Key<- Carta_data %>% 
  select(OHI_region, Genus, status, control, notes)
# 1008 obs., 5 vars.

Carta_Key <- Carta_Key %>%
  select(OHI_region, Genus, status, control, notes) %>%
  distinct()
# 544 obs., 5 vars.

# Make some assumptions at the genus level:
Carta_Key <- Carta_data %>%
  group_by(OHI_region, Genus) %>%
  summarise(
    ## ---- CARTA STATUS ----
    status = {
      s <- status[!is.na(status)]
      if (length(s) == 0) NA_character_
      else if ("deteriorating" %in% s) "deteriorating"
      else if ("improving" %in% s) "improving"
      else s[1]
    },

    alternative_status = {
      s <- status[!is.na(status)]
      primary <- {
        if ("deteriorating" %in% s) "deteriorating"
        else if ("improving" %in% s) "improving"
        else if (length(s) > 0) s[1]
        else NA_character_
      }
      alts <- s[s != primary]
      if (length(alts) == 0) NA_character_ else paste(unique(alts), collapse = "; ")
    },


    ## ---- CARTA CONTROL ----
    control = {
      c <- control[!is.na(control)]
      if (length(c) == 0) NA_character_ else c[1]
    },

    alternative_control = {
      c <- control[!is.na(control)]
      primary <- if (length(c) > 0) c[1] else NA_character_
      alts <- c[c != primary]
      if (length(alts) == 0) NA_character_ else paste(unique(alts), collapse = "; ")
    },


    ## ---- CARTA NOTES ----
    notes = {
      n <- notes[!is.na(notes) & notes != ""]
      if (length(n) == 0) NA_character_ else n[1]
    },

    alternative_notes = {
      n <- notes[!is.na(notes) & notes != ""]
      primary <- if (length(n) > 0) n[1] else NA_character_
      alts <- n[n != primary]
      if (length(alts) == 0) NA_character_ else paste(unique(alts), collapse = " | ")
    },

    .groups = "drop"
  ) %>%
  
  ## ---- REORDER COLUMNS (put all "alternative_*" at the end) ----
  select(
    OHI_region,
    Genus,
    status,
    control,
    notes,
    alternative_status,
    alternative_control,
    alternative_notes
  )
# 504 obs., 5 vars.
# Importantly, there don't seem to be any competing management suggestions when we aggregate at the genus level (versus the species level)

Carta_Key <- Carta_Key %>%
  select(Genus, OHI_region, status, control, notes) %>% 
  arrange(Genus, OHI_region)

Carta_Key <- Carta_Key %>%
  mutate(
    Genus      = tolower(Genus),
    OHI_region = tolower(OHI_region)
  )
```

## SAVE the CARTA KEY
```{r SAVE the Carta Key}
# write.csv(Carta_Key, "/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_Carta_Scores/OHI_CartaKey_Dec112025.csv", row.names = FALSE)
```

```{r Load in data as necessary}
# Load landings file, if needed:
Carta_Key<- read.csv("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_StockStatus_Scores/Carta_Scores/OHI_CartaKey_Dec112025.csv")
# 504 obs., 5 vars.

# GoC_Landings_for_Carta<- Landings_GoC_NoFresh.NoMariculture
GoC_Landings_for_Carta_original<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/fis/v2025/int/GOC_FIS.landings_CLEAN_as.of.Dec092025.rds"))
GoC_Landings_for_Carta<- GoC_Landings_for_Carta_original
# 3163120 obs., 50 vars.
```

```{r General filtering}
# In general, the receipts from early trips are a bit funky. Let's filter out trips before 2003 (n= 267072 records).
GoC_Landings_for_Carta<- GoC_Landings_for_Carta %>%
  filter(year >= 2003)
# 2896048 obs., 50 vars.

# Let's also filter out any records where weight_kg_corrected is < or = 0 (n= 156 records).
GoC_Landings_for_Carta<- GoC_Landings_for_Carta %>%
  filter(weight_kg_corrected > 0)
# 2895892 obs., 50 vars.

sum(is.na(GoC_Landings_for_Carta$weight_kg_corrected))
```

```{r Summarize these data by Year Region and Genus with a 2018-2022 reference period}
Carta_Landings_summary <- GoC_Landings_for_Carta %>%
  group_by(year, OHI_Region_No, estimated_genus_name) %>%
  summarize(
    Annual_Catch_kg = sum(weight_kg_corrected, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(OHI_Region_No, estimated_genus_name) %>%
  mutate(
    Avg_Catch_Ref2018.22 = mean(
      Annual_Catch_kg[year %in% 2018:2022],
      na.rm = TRUE
    ),
    n_years_Ref2018.22 = sum(year %in% 2018:2022)
  ) %>%
  ungroup() %>%
  arrange(OHI_Region_No, estimated_genus_name, year)

# Make lowercase:
Carta_Landings_summary <- Carta_Landings_summary %>%
  mutate(
    estimated_genus_name = tolower(estimated_genus_name),
    OHI_Region_No       = tolower(OHI_Region_No)
  )
# 12241 obs., 4 vars.
```

```{r Merge in Key Data}
# Merge:
Carta_Scores <- Carta_Landings_summary %>%
  left_join(
    Carta_Key,
    by = c("estimated_genus_name" = "Genus",
           "OHI_Region_No"       = "OHI_region")
  ) %>%
  arrange(OHI_Region_No, estimated_genus_name, year)
# 12241 obs., 9 vars. as expected! Good!
```

```{r Calculate Stock_Status scores based on C_current versus C_ref}
# But first, check what combinations there are in the Carta_Key between stock "status" and "control" recommendation. We're looking for weird logical mismatches such as stock is doing poorly but control says keep fishing:
Carta_Scores %>%
  count(status, control, name = "n")
# There are 6 possible combinations: (NA, NA) (maximally harvested, no increase) (unclear, NA) (deteriorating, decrease)	(deteriorating	no increase) (development potential	increase)

tolerance <- 0.10 # How similar is C_current versus C_ref? If C_current is within 10% of C_ref, then functionally equal.

Carta_Scores <- Carta_Scores %>%
  mutate(
    approx_equal = abs(Annual_Catch_kg - Avg_Catch_Ref2018.22) <=
      tolerance * Avg_Catch_Ref2018.22,

    Stock_Status = case_when(

      # ── Scenario 1: Maximally harvested ─────────────────────────
      status == "maximally harvested" &
        control == "no increase" &
        approx_equal ~
        1,

      status == "maximally harvested" &
        control == "no increase" &
        Annual_Catch_kg > Avg_Catch_Ref2018.22 ~
        Avg_Catch_Ref2018.22 / Annual_Catch_kg,

      status == "maximally harvested" &
        control == "no increase" &
        Annual_Catch_kg < Avg_Catch_Ref2018.22 ~
        1,

      # ── Scenario 2: Stock deteriorating ─────────────────────────
      status == "deteriorating" &
        control %in% c("decrease", "no increase") &
        approx_equal ~
        0.5,

      status == "deteriorating" &
        control %in% c("decrease", "no increase") &
        Annual_Catch_kg > Avg_Catch_Ref2018.22 ~
        (Avg_Catch_Ref2018.22 / Annual_Catch_kg) * 0.5,

      status == "deteriorating" &
        control %in% c("decrease", "no increase") &
        Annual_Catch_kg < Avg_Catch_Ref2018.22 ~
        0.5 + 0.5 * (
          (Avg_Catch_Ref2018.22 - Annual_Catch_kg) /
          Avg_Catch_Ref2018.22
        ),

      # ── Scenario 3: Development potential ───────────────────────
      status == "development potential" ~
        1,

      TRUE ~ NA_real_
    )
  ) %>%
  select(-approx_equal)
# 12241 obs., 10 vars.

# Filter out years before 2018 and 2025
Carta_Scores<- Carta_Scores %>% filter(year >= 2018 & year <= 2024)
# 4310 obs., 9 vars.

# Now filter only those genera that have some actual scores:
Carta_Scores<- Carta_Scores %>%
  group_by(estimated_genus_name) %>%
  filter(any(!is.na(Stock_Status))) %>%
  ungroup()
# 1267 obs., 10 vars.

unique(Carta_Scores$estimated_genus_name)
# 33 genera

mean(Carta_Scores$Stock_Status, na.rm=TRUE)
```


#### SAVE THE CARTA SCORES DATAFRAME
```{r Write this as a CSV}
write.csv(Carta_Scores, "/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_StockStatus_Scores/Carta_Scores/OHI_CartaScores_overTime_Dec162025.csv", row.names = FALSE)
```


# OHI SCORES: Merge the indicator dataframes  ------------------------------

```{r Load the indicator dataframes as necessary}
# Read in datafiles as necessary:
OHI_CPUE_Scores<- read.csv("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_CPUE_Scores/OHI_CPUEScores_SECONDDRAFT_Dec12025.csv")
OHI_CPUE_Scores<- OHI_CPUE_Scores %>% select(-X)
# 131 obs., 3 vars.

CMSY_Scores<- read.csv("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_CMSY_Scores/OHI_CMSY.BBmsyScores_SECONDDRAFT_Dec112025.csv")
CMSY_Scores<- CMSY_Scores %>% select(-X)
# 2171 obs., 11 vars.

Carta_Scores<- read.csv("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_Carta_Scores/OHI_CartaScores_overTime_Dec112025.csv")
# 1093 obs., 9 vars.
```

```{r Create Stock Status Dataframe}
# Merge CMSY_Scores and Carta_Scores by genus, year, and region:
# lowercase to make sure matching works
CMSY_Scores <- CMSY_Scores %>%
  mutate(OHI_Region_No = tolower(OHI_Region_No))
        
StockStatus_Scores <- full_join(
  CMSY_Scores,
  Carta_Scores,
  by = c("yr" = "year",
         "OHI_Region_No",
         "stock" = "estimated_genus_name")
)
# 2636 obs., 17 vars.
```

```{r Out of curiosity are there genera for which there is no CMSY score but there is a CNP score?}
# Filter for rows where B/Bmsy is missing but CNP_score exists
genera_with_CNP_only <- StockStatus_Scores %>%
  filter(is.na(B.Bmsy) & !is.na(CNP_score)) %>%
  arrange(OHI_Region_No, stock, yr)

unique(genera_with_CNP_only$stock)
unique(Carta_Scores$estimated_genus_name)
# Neat! Information about 17 additional stocks!
#  [1] "argopecten"         "chione"             "galeorhinus"        "haliotis"          
#  [5] "hoplopagrus"        "panopea"            "rhizoprionodon"     "strongylocentrotus"
#  [9] "megapitaria"        "negaprion"          "octopus"            "thunnus"           
# [13] "sphoeroides"        "callinectes"        "dosinia"            "panulirus"         
# [17] "macrobrachium"      

unique(CMSY_Scores$stock)
unique(Carta_Scores$estimated_genus_name)
```

```{r Assign Stock Status 'Pick' column}
StockStatus_Scores$RAM_BBmsy<- c(NA)

StockStatus_Scores<- StockStatus_Scores %>%
  select(OHI_Region = OHI_Region_No, Year = yr, stock_genus = stock, 
         Annual_Catch_kg = total_catch_kg, B_est = B,
         BBmsy_low = lcl.B.Bmsy, BBmsy_hi = ucl.B.Bmsy,
         FFmsy = F.Fmsy, FFmsy_low = lcl.F.Fmsy, FFmsy_hi = ucl.F.Fmsy,
         CatchTrend = Catch_t2_vs_Catch_t1, Carta_Status = status, Carta_Control = control,
         RAM_BBmsy, BBmsy = B.Bmsy, CNP_score)

# A hierarchy of which score to pick!
StockStatus_Scores <- StockStatus_Scores %>%
  mutate(
    StockStatus_Pick = case_when(
      !is.na(RAM_BBmsy) ~ RAM_BBmsy,
      !is.na(BBmsy)    ~ BBmsy,
      !is.na(CNP_score)     ~ CNP_score,
      TRUE              ~ NA_real_
    ),
    StockStatus_capped = pmin(StockStatus_Pick, 1)  # caps values at 1
  )
```


## GEOMETRIC MEAN!!

```{r Weights and geometric mean}
# Calculate the proportion of catch each stock contributed to its regional total during that 20-year reference period.

weights_data <- status_data %>%
filter(year > 1990 & year < 2010) %>%
group_by(stock_id, region_id) %>%
mutate(StockCatch = sum(catch)) %>% # total tonnes across all years for each stock and region
ungroup() %>%
group_by(region_id) %>%
mutate(TotalCatch = sum(catch)) %>% # total tonnes catch across all years and stocks for each region
ungroup() %>%
mutate(wprop = StockCatch / TotalCatch) %>% # proportion of catch each stock contributed to the total catch
distinct(stock_id, region_id, wprop)


status_data <- left_join(status_data, weights_data, by = region_id) %>%
group_by(region_id, year) %>%
summarize(status = prod(score ^ wprop)) %>% # <-- This is the geometric mean part.
ungroup()
```











## WEIGHTED CMSY-B/BMSY SCORES PER REGION PER YEAR
```{r Weighted B/Bmsy}
StockStatus_Scores_weighted <- StockStatus_Scores %>%
  group_by(Year, OHI_Region) %>%
#  filter(Stock_Status_score is not NA)
  summarize(
    
  #   # Weighted average for Stock_Status_score using total_catch_kg as weights
  #   weighted_StockStatus = sum(StockStatus_capped * Annual_Catch_kg, na.rm = TRUE) / sum(Annual_Catch_kg, na.rm = TRUE),
  #   
  #   # Total catch per region-year (optional)
  #   Annual_Catch_kg = sum(Annual_Catch_kg, na.rm = TRUE),
  #   
  #   # Number of unique stocks (genera) that went into the calculation:
  #   n_stocks = n_distinct(stock_genus),
  #   
  #   .groups = "drop"
  # ) %>%
  # 
  # # Cap the weighted_StockStatus at 1
  # mutate(
  #   weighted_StockStatus_capped = pmin(weighted_StockStatus, 1)
  # ) %>%
    # Slot in Mel's code here for weighted catches (over "historic" period) ...
  
  select(Year, OHI_Region, Annual_Catch_kg, n_stocks, weighted_StockStatus, weighted_StockStatus_capped)
```
```{r Save this dataframe}
write.csv(StockStatus_Scores_weighted, "/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_FIS_Scores/FIS_StockStatus_WeightedScores_UNDERCONSTRUCTION_Dec112025.csv", row.names = FALSE)
```

```{r Plot weighted_StockStatus over time}
#### PLOT weighted_StockStatus OVER TIME ####

#### State Colors ####
state_colors <- c(
  "BAJA CALIFORNIA" = "royalblue",
  "BAJA CALIFORNIA SUR" = "skyblue",
  "SONORA" = "orange",
  "SINALOA" = "sienna1",
  "NAYARIT" = "plum", 
  "JALISCO" = "purple4"
)

Region_colors <- c(
  "royalblue",
  "skyblue",
  "orange",
  "sienna1",
  "plum",
  "purple4",
  "aquamarine3",
  "gold2",
  "tomato"
)

Regions_of_Interest <- c(
  "Region_1", "Region_2", "Region_3",
  "Region_4", "Region_5", "Region_6",
  "Region_7", "Region_8", "Region_9"
)

Regions_of_Interest<- tolower(Regions_of_Interest)

weighted_StockStatus_overTime <- StockStatus_Scores_weighted %>%
  filter(OHI_Region %in% Regions_of_Interest) %>%
  ggplot(aes(
    x = Year,
    y = weighted_StockStatus_capped,
    color = OHI_Region
  )) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ OHI_Region, scales = "fixed") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = Region_colors) +
  labs(
    title = "FIS - Regional Stock_Status Scores, capped at 1 (2014–2024)",
    y = "Stock_Status Scores (0-1, weighted by total landings)",
    x = "Year"
  ) +
  ylim(0, 1.25) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 18, face = "bold")
  )

weighted_StockStatus_overTime
  
  # Save as PNG
  # ggsave(
  #   filename = sprintf("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_FIS_Scores/OHI_StockScores_overTime_Dec112025.png"),
  #   plot = weighted_StockStatus_overTime,
  #   width = 12,
  #   height = 12,
  #   dpi = 300
  # )

```

```{r Merge ALL dataframes for final OHI Score}
# Merge:
# lowercase to make sure matching works
OHI_CPUE_Scores <- OHI_CPUE_Scores %>%
  mutate(OHI_Region = tolower(OHI_Region))

OHI_Scores_df <- OHI_CPUE_Scores %>%
  full_join(
    StockStatus_Scores_weighted,
    by = c("Year", "OHI_Region")
  )

OHI_Scores_df <- OHI_Scores_df %>%
  mutate(FIS_Score = rowMeans(cbind(weighted_CPUE_acrossAllGroups, weighted_StockStatus_capped), na.rm = TRUE))

# SAVE THIS DATAFRAME FOR NOW AS AN EXAMPLE - Dec 11 2025:
# write.csv(OHI_Scores_df, "/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_FIS_Scores/OHI_FISscores_UNDERCONSTRUCTION_Dec112025.csv", row.names = FALSE)
```

```{r Graph FIS scores per region over time}
weighted_StockStatus_overTime <- StockStatus_Scores_weighted %>%
  filter(OHI_Region %in% Regions_of_Interest) %>%
  ggplot(aes(
    x = Year,
    y = weighted_StockStatus_capped,
    color = OHI_Region
  )) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ OHI_Region, scales = "fixed") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = Region_colors) +
  labs(
    title = "FIS - Regional Stock_Status Scores, capped at 1 (2014–2024)",
    y = "Stock_Status Scores (0-1, weighted by total landings)",
    x = "Year"
  ) +
  ylim(0, 1.25) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 18, face = "bold")
  )

weighted_StockStatus_overTime


FIS_Scores_overTime<- OHI_Scores_df %>%
  filter(OHI_Region %in% Regions_of_Interest) %>%
  
  ggplot(aes(x = Year, y = FIS_Score, color = OHI_Region)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ OHI_Region, scales = "fixed") +
  
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
  
  labs(
    title = "FIS Scores per Region Over Time",
    x = "Year",
    y = "FIS Score (0–1)"
  ) +
  
  ylim(0.50, 1.25) +
  
  scale_color_manual(values = Region_colors) +
  
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 16, face = "bold")
  )

FIS_Scores_overTime
```

