################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
# Visualization
library(ggplot2)
library(plotly)
library(htmlwidgets)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```


# Stock Assessments from RAM ------------------------------

```{r Mexico FAO Data}
## FAO data, some of the zeros were originally reported as NA (so exercise caution there)
harvest <- read_csv("https://raw.githubusercontent.com/OHI-Science/ohiprep_v2025/refs/heads/main/globalprep/mar/v2025/output/mar_harvest_tonnes.csv")
mex_harvest <- filter(harvest, rgn_id == 135)
mex_harvest <- filter(mex_harvest, year == 2017) %>% data.frame()

# data from Seafood Watch (Monterey Bay)
sustainability <- read_csv("https://raw.githubusercontent.com/OHI-Science/ohiprep_v2025/refs/heads/main/globalprep/mar/v2025/output/mar_sustainability.csv")
mex_sust <- filter(sustainability, taxa_code %in% mex_harvest$taxa_code) %>%
 filter(year==2025) %>% data.frame()

# determine whether sustainability was gapfilled (if so, fairly low confidence, depending on method used and taxa)
gf <- read_csv("https://raw.githubusercontent.com/OHI-Science/ohiprep_v2025/refs/heads/main/globalprep/mar/v2025/output/mar_sustainability_gf.csv")
mex_gf <- filter(gf, taxa_code %in% mex_harvest$taxa_code) %>% data.frame()
```


# CONAPESCA data ------------------------------

```{r CONAPESCA data shared by JC}
# These data were shared with us by Juan Carlos Villaseñor and originally accessed on September 26, 2025 from this GitHub repository: https://github.com/mex-fisheries

MexLandings_2000.present_latlong <- read_csv("/home/shares/ohi/OHI_GOC/_raw_data/CONAPESCA/d2025/SharedByJuanCarlos/mex_landings/data/clean/mex_landings_2000_present_with_lat_long.csv")
table(MexLandings_2000.present_latlong$state)
# n = 2,316,370 records from Baja, B.C.S., and Sonora from January 2000 through January 2025

MexLandings_2000.present<- readRDS(here("/home/shares/ohi/OHI_GOC/_raw_data/CONAPESCA/d2025/SharedByJuanCarlos/mex_landings/data/clean/mex_landings_2000_present.rds"))
# n = 12,153,879 records from all states in Mexico from January 2000 through January 2025
table(MexLandings_2000.present$state)

MexLandings_2018.present<- readRDS(here("/home/shares/ohi/OHI_GOC/_raw_data/CONAPESCA/d2025/SharedByJuanCarlos/mex_landings/data/clean/mex_conapesca_apertura_2018_present.rds"))

# Skimming through the dat. What does it look like?
head(MexLandings_2000.present)

table(MexLandings_2000.present$acuaculture_production) # Yes aquaculture production

max(MexLandings_2000.present$year)
max(MexLandings_2000.present$month[MexLandings_2000.present$year])
# Records up to January, 2025

```

```{r Basic filtering}
States_of_Interest<- c("BAJA CALIFORNIA", "BAJA CALIFORNIA SUR", "SONORA", "SINALOA", "NAYARIT", "JALISCO")

MexLandings_filtered<- MexLandings_2000.present %>% filter(state %in% States_of_Interest)
# n obs.= 4,713,144 (-7,440,735 obs)
# PRODUCTION NOTE: this operation filtered out 61% of all records and suggests that 39% of all fisheries records in Mexico (from 2000-present) come from these 6 states.
table(MexLandings_filtered$fleet)
# large_scale small_scale 
#      181823     1438462 
table(MexLandings_filtered$acuaculture_production) #there are also NAs in this column...

# WC.Fisheries_filtered<- MexLandings_filtered %>% filter(acuaculture_production== "NO") # n obs.= 3,701,563 (-1,011,581 obs)
# PRODUCTION NOTE: this operation filtered out 21% of all records from states of interest, suggesting that wild-catch fisheries account for 79% of all records.
WC.Fisheries_filtered<- MexLandings_filtered %>% filter(acuaculture_production== "NO" | is.na(acuaculture_production)) # 4684000 obs.
table(WC.Fisheries_filtered$year)

Mariculture_filtered<- MexLandings_filtered %>% filter(acuaculture_production== "SÍ")
# n obs.= 29,144 (-4,684,000 obs)
# PRODUCTION NOTE: this operation filtered out 99% of all records from states of interest and suggests that mariculture operations account for 1% of all records. This may or may not track with Mexico's total aquaculture production. Preliminary figure suggests that ~13.5% of Mexico's total fishery production is sources from aquaculture in 2024: https://mexicobusiness.news/agribusiness/news/mexicos-fisheries-and-aquaculture-sectors-see-growth-2024?utm_source=chatgpt.com

OTRAS_filtered<- MexLandings_filtered %>% filter(main_species_group== "OTRAS")
# n obs.= 715.547 (-1,011,581 obs)
# PRODUCTION NOTE: this operation filtered out 21% of all records from states of interest, suggesting that wild-catch fisheries account for 79% of all records.

# NOTES ON PRELIMINARY ANALYSIS: 
# SOME RECORDS ARE NOT APPENDED WITH INFORMATION ABOUT RECORD TYPE (i.e., no information on wild-catch versus aquaculture). These records are excluded from this preliminary analyses.
```

```{r Vessel summary table}
Vessel_Summary1 <- MexLandings_filtered %>%
  filter(year >= 2019 & year <=2024) %>%
  select(year, state, landing_site_key, eu_rnpa, eu_name, vessel_name) %>%
  group_by(year, state, landing_site_key) %>%
  summarise(
    Count_n_rnpa = n_distinct(eu_rnpa, na.rm = TRUE),
    Count_n_eu_name = n_distinct(eu_name, na.rm = TRUE),
    Count_n_unique_vessels = n_distinct(vessel_name, na.rm = TRUE),
    .groups = "drop"
  ) 

Vessel_Summary2 <- MexLandings_filtered %>%
  filter(year >= 2019 & year <=2024) %>%
  select(year, state, eu_rnpa, eu_name, vessel_name) %>%
  group_by(year, state) %>%
  summarise(
    Count_n_rnpa = n_distinct(eu_rnpa, na.rm = TRUE),
    Count_n_eu_name = n_distinct(eu_name, na.rm = TRUE),
    Count_n_unique_vessels = n_distinct(vessel_name, na.rm = TRUE),
    .groups = "drop"
  ) 
```

```{r W.C. cumulative percentage table}
# Summary table sorted by cumulative contribution of each fishery, per state, over the last 5 years:
WC.Fisheries_cumulative.contributions <- WC.Fisheries_filtered %>%
  filter(year >= 2019) %>%
  select(state, main_species_group, live_weight) %>%
  group_by(state, main_species_group) %>%
  summarise(
    Total_Live_Biomass = sum(live_weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(state) %>%
  mutate(
    State_Total = sum(Total_Live_Biomass, na.rm = TRUE),
    Percent_of_State = 100 * Total_Live_Biomass / State_Total
  ) %>%
  arrange(state, desc(Percent_of_State)) %>%
  mutate(
    Cumulative_Percent = cumsum(Percent_of_State)
  ) %>%
  ungroup() %>%
  mutate(
    Percent_of_State = sprintf("%.1f", Percent_of_State),
    Cumulative_Percent = sprintf("%.1f", Cumulative_Percent)
  )
```

```{r Plot cumulative percent}
df <- WC.Fisheries_cumulative.contributions %>%
  mutate(Percent_of_State = as.numeric(Percent_of_State)) %>%
  arrange(state, desc(Percent_of_State)) %>%
  mutate(
    # reorder within state for plotting
    main_species_group_ordered = reorder_within(main_species_group, Percent_of_State, state, fun = max, order = TRUE)
  )


W.C.Fisheries_cumulative_plot<- ggplot(df, aes(x = main_species_group_ordered, y = Percent_of_State)) +
  geom_segment(aes(xend = main_species_group_ordered, y = 0, yend = Percent_of_State), color = "pink") +
  geom_point(color = "tomato", size = 3) +
  facet_wrap(~state, scales = "free_y") +
  coord_flip() +
  scale_x_reordered() +   # <-- important for correct label display
  labs(
    title = "Species Group Contribution per State (Ordered by Descending Percent)",
    x = "Main Species Group",
    y = "Percentage of State Total (%)"
  ) +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))

W.C.Fisheries_cumulative_plot
W.C.Fisheries_cumulative_interactive <- ggplotly(W.C.Fisheries_cumulative_plot)
```

```{r Save plot}
ggsave(
  filename = "FIS/v2025/W.C. Fisheries Cumulative Plot_Sep 29 2025.png",  # file name
  plot = W.C.Fisheries_cumulative_plot,            # the plot object
  width = 10,                        # width in inches
  height = 10,                        # height in inches
  dpi = 300                         # resolution (dots per inch)
)

# Save your plotly object as HTML
htmlwidgets::saveWidget(W.C.Fisheries_cumulative_interactive, "FIS/v2025/W.C. Fisheries Cumulative Plotly_Sep 29 2025.html")
```

```{r MAR cumulative percentage table}
# Create a summary table of biomass landed from mariculture in the last 20 years... no data since 2019...
Mariculture_summary <- Mariculture_filtered %>%
  # filter(year >= 2019) %>%
  select(state, main_species_group, live_weight, landed_weight, value) %>%
  group_by(state, main_species_group) %>%
  summarise(
    Total_Live_Biomass = sum(live_weight, na.rm = TRUE),
    Total_Landed_Biomass = sum(landed_weight, na.rm = TRUE),
    Total_Value = sum(value, na.rm = TRUE),
    n_Records = n(),
    .groups = "drop"  # optional: ungroup after summarising
  )

# Summary table sorted by cumulative contribution of each mariculture stock, per state, over the last 20 years (no explicitly logged data since 2019)
Mariculture_cumulative.contributions <- Mariculture_filtered %>%
  #filter(year >= 2019) %>%
  select(state, main_species_group, live_weight) %>%
  group_by(state, main_species_group) %>%
  summarise(
    Total_Live_Biomass = sum(live_weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(state) %>%
  mutate(
    State_Total = sum(Total_Live_Biomass, na.rm = TRUE),
    Percent_of_State = 100 * Total_Live_Biomass / State_Total
  ) %>%
  arrange(state, desc(Percent_of_State)) %>%
  mutate(
    Cumulative_Percent = cumsum(Percent_of_State)
  ) %>%
  ungroup() %>%
  mutate(
    Percent_of_State = sprintf("%.1f", Percent_of_State),
    Cumulative_Percent = sprintf("%.1f", Cumulative_Percent)
  )
```

```{r Plot cumulative percent}
df <- Mariculture_cumulative.contributions %>%
  mutate(Percent_of_State = as.numeric(Percent_of_State)) %>%
  arrange(state, desc(Percent_of_State)) %>%
  mutate(
    # reorder within state for plotting
    main_species_group_ordered = reorder_within(main_species_group, Percent_of_State, state, fun = max, order = TRUE)
  )


Mariculture_cumulative_plot<- ggplot(df, aes(x = main_species_group_ordered, y = Percent_of_State)) +
  geom_segment(aes(xend = main_species_group_ordered, y = 0, yend = Percent_of_State), color = "lightblue") +
  geom_point(color = "blue", size = 3) +
  facet_wrap(~state, scales = "free_y") +
  coord_flip() +
  scale_x_reordered() +   # <-- important for correct label display
  labs(
    title = "Species Group Contribution per State (Ordered by Descending Percent)",
    x = "Main Species Group",
    y = "Percentage of State Total (%)"
  ) +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))

Mariculture_cumulative_plot
Mariculture_cumulative_interactive <- ggplotly(Mariculture_cumulative_plot)
```

```{r Save plot}
ggsave(
  filename = "MAR/v2025/Mariculture Stocks Cumulative Plot_Oct 1 2025.png",  # file name
  plot = Mariculture_cumulative_plot,            # the plot object
  width = 10,                        # width in inches
  height = 10,                        # height in inches
  dpi = 300                         # resolution (dots per inch)
)

# Save your plotly object as HTML
htmlwidgets::saveWidget(Mariculture_cumulative_interactive, "MAR/v2025/Mariculture Cumulative Plotly_Oct 1 2025.html")
```

```{r Summary table 2021}
# Create a summary table of biomass landed from w.c. fisheries in the last 5 years:
JCdata_Fisheries_summary <- WC.Fisheries_filtered %>%
  filter(year == 2021) %>%
  select(state, main_species_group, live_weight, landed_weight, value) %>%
  group_by(state, main_species_group) %>%
  summarise(
    Total_Live_Biomass_JCdata = sum(live_weight, na.rm = TRUE),
    Total_Landed_Biomass_JCdata = sum(landed_weight, na.rm = TRUE),
    Total_Value_JCdata = sum(value, na.rm = TRUE),
    n_Records_JCdata = n(),
    .groups = "drop"  # optional: ungroup after summarising
  )
```


# Regenerate: J.C. CONAPESCA data with species names (October 2025) ------------------------------
```{r Regenerating J.C. CONAPESCA data including species name from the landing receipts}
# How many species are fished in the GoC?
Species_data_dissertation <- read_csv("/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/CONAPESCA_SpMasterFile_EMF.csv")
length(unique(Species_data_dissertation$ScientificName))
# At least 496 unique species, according to a study I did in my dissertation (focused on a smaller sub-set of fisheries and area).

MexLandings_2000.present_wSpecies<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fis/v2025/int/juan_carlos/mex_landings_2000_present_EMF_edited.rds"))
# n = 12153879 obs.

length(unique(MexLandings_2000.present_wSpecies$species_name)) # 2052 unique species in all of Mexico.
# But remember... We don't need species from all of Mexico! Only the GoC region. Let's filter by state.

States_of_Interest<- c("BAJA CALIFORNIA", "BAJA CALIFORNIA SUR", "SONORA", "SINALOA", "NAYARIT", "JALISCO")

Landings_NWMexico <- MexLandings_2000.present_wSpecies %>% filter(state %in% States_of_Interest)
# n obs.= 4713144
length(unique(Landings_NWMexico$species_name)) 
# 1827 unique species in States of Interest.

# Filter OUT sites from Baja that are on the Pacific side:
# Lat-Long file (from Mel's repository):
Lat_Long_MasterFile<- read_csv("/home/frazier/ohi_goc_prep/FP/FIS/v2025/data/Baja_Sonora_states_key.csv")
colnames(Lat_Long_MasterFile) <- c("state", "landing_site", "landing_site_primary", "lat", "long", "GoC")

# Append lat-long info to landings data
temp_df <- Landings_NWMexico %>%
  left_join(
    Lat_Long_MasterFile,
    by = c("state", "landing_site")  # columns to match
  )

table(temp_df$GoC)
# 4713144 (total) - 996729 (no) = 3716415 obs. that are explicitly as "yes" in the GoC or NA. I have removed entries that are definitely "no" not in the Gulf.

# Generate filtered landings data for the GoC:
Landings_GoC <- temp_df %>%
  filter(GoC != "no" | is.na(GoC))
# 3716415 obs.

# Delete temp_df, not needed anymore.
rm(temp_df)

length(unique(Landings_GoC$species_name))
# 1693 unique species in the Gulf of California according to Juan Carlos' version of the CONAPESCA data base, keeping in mind that these are not really SPECIES (per se) but rather log-book entries for species.
length(unique(Landings_GoC$main_species_group))
# From 57 main species groups.
```
1693 unique species in the Gulf of California according to Juan Carlos' version of the CONAPESCA data base, keeping in mind that these are not really SPECIES (per se) but rather log-book entries for species.

## Save the Landings_GoC rds file
```{r Save the Landings_GoC dataframe: Nov 03 2025}
# saveRDS(Landings_GoC, file="/home/shares/ohi/OHI_GOC/goal_prep/fis/v2025/int/juan_carlos/GoC_Landings_2000.present_EMF_created.rds")
```


# Explore: Top species groupings in CONAPESCA database (November 2025) ------------------------------
```{r Load in the Landings_GoC dataframe}
# Skip the full file prep (above) by loading in the dataset directly:
Landings_GoC<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fis/v2025/int/juan_carlos/GoC_Landings_2000.present_EMF_created.rds"))
# n = 3716415 obs.
```

```{r Plot ALGAE by state}
Algae_Landings_summary <- Landings_GoC %>%
  filter(year > 2014 & year < 2025) %>%
  filter(main_species_group=="ALGAS") %>%
  select(year, state, main_species_group, live_weight, value) %>%
  group_by(year, state, main_species_group) %>%
  summarise(Annual_Live_Biomass = sum(live_weight, na.rm = TRUE),
            Annual_Revenue = sum(value, na.rm = TRUE), .groups = "drop")

state_colors <- c(
  "BAJA CALIFORNIA" = "royalblue",
  "BAJA CALIFORNIA SUR" = "skyblue",
  "SONORA" = "orange",
  "SINALOA" = "sienna1",
  "NAYARIT" = "plum", 
  "JALISCO" = "purple4"
)


Algae_State_plot <- ggplot(Algae_Landings_summary,
       aes(x = year,
           y = Annual_Live_Biomass,
           color = state,
           group = state)) +
  geom_line(linewidth = 1) +                     # Line plot
  geom_point(size = 2) +                         # Points on top for clarity
  facet_wrap(~ state, scales = "free_y") +
  scale_color_manual(values = state_colors) +
  labs(
    title = "Algae Landings Over Time",
    subtitle = "Annual live biomass by state",
    x = "Year",
    y = "Landings (kg)",
    color = "State"
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 9),
    legend.position = "none"                    # Legend redundant w/ facets
  )

Algae_State_plot

# Save the plot as PNG
ggsave(filename = "/home/ferrer/ohi_goc_prep/NP/v2025/Prelim_Algae_Plot_Nov202025.png",
       plot = Algae_State_plot,
       width = 10,        # width in inches
       height = 6,        # height in inches
       dpi = 300)         # resolution
```


```{r Plot taxa by total landings}
state_colors <- c(
  "BAJA CALIFORNIA" = "royalblue",
  "BAJA CALIFORNIA SUR" = "skyblue",
  "SONORA" = "orange",
  "SINALOA" = "sienna1",
  "NAYARIT" = "plum", 
  "JALISCO" = "purple4"
)

table(Landings_GoC$fleet)

Landings_GoC_summary <- Landings_GoC %>%
  filter(year > 2014 & year < 2025) %>%
  select(state, main_species_group, live_weight) %>%
  group_by(state, main_species_group) %>%
  summarise(Total_Live_Biomass = sum(live_weight, na.rm = TRUE), .groups = "drop") %>%
  group_by(state) %>%
  mutate(
    State_Total = sum(Total_Live_Biomass, na.rm = TRUE),
    Percent_of_State = 100 * Total_Live_Biomass / State_Total
  )

# %>%
#   filter(Percent_of_State >= 8) %>%
#   ungroup()

Landings_GoC_plot_filtered <- Landings_GoC_summary %>%
  filter(Percent_of_State >= 2) %>%         # <---- NEW
  filter()
  mutate(
    main_species_group_ordered =
      tidytext::reorder_within(main_species_group,
                               Percent_of_State,
                               interaction(state))
  )
  

Sp.State_Biomass_plot<- ggplot(Landings_GoC_plot_filtered,
       aes(x = main_species_group_ordered,
           y = Percent_of_State,
           color = state)) +                     # <--- color by state
  geom_segment(aes(xend = main_species_group_ordered,
                   y = 0,
                   yend = Percent_of_State),
               linewidth = 1) +
  geom_point(size = 3) +
  facet_wrap(~ state, scales = "free_y") +
  coord_flip() +
  scale_x_reordered() +
  scale_color_manual(values = state_colors) +    # <--- apply your colors
  labs(
    title = "Species Groups Contributing ≥ 2% of State Catch",
    x = "Main Species Group",
    y = "Percentage of State Total (%)",
    color = "State"
  ) +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))

Sp.State_Biomass_plot

# Save the plot as PNG
# ggsave(filename = "/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/CONAPESCA_Biomass&CPUE_Nov2025/Sp.State_Biomass_plot_Nov182025.png",
#        plot = Sp.State_Biomass_plot,
#        width = 10,        # width in inches
#        height = 6,        # height in inches
#        dpi = 300)         # resolution
```

```{r Top fisheries for SSF}
Landings_GoC_summary_SSF <- Landings_GoC %>%
  filter(year > 2014 & year < 2025) %>%
  filter(fleet=="small_scale") %>%
  select(state, main_species_group, live_weight) %>%
  group_by(state, main_species_group) %>%
  summarise(Total_Live_Biomass = sum(live_weight, na.rm = TRUE), .groups = "drop") %>%
  group_by(state) %>%
  mutate(
    State_Total = sum(Total_Live_Biomass, na.rm = TRUE),
    Percent_of_State = 100 * Total_Live_Biomass / State_Total
  )
  
Landings_GoC_SSF_filtered <- Landings_GoC_summary_SSF %>%
  filter(Percent_of_State >= 2) %>%         # <---- NEW
  mutate(
    main_species_group_ordered =
      tidytext::reorder_within(main_species_group,
                               Percent_of_State,
                               interaction(state))
  )
  
Sp.State_SSF.Biomass_plot<- ggplot(Landings_GoC_SSF_filtered,
       aes(x = main_species_group_ordered,
           y = Percent_of_State,
           color = state)) +                     # <--- color by state
  geom_segment(aes(xend = main_species_group_ordered,
                   y = 0,
                   yend = Percent_of_State),
               linewidth = 1) +
  geom_point(size = 3) +
  facet_wrap(~ state, scales = "free_y") +
  coord_flip() +
  scale_x_reordered() +
  scale_color_manual(values = state_colors) +    # <--- apply your colors
  labs(
    title = "Species Groups Contributing ≥ 2% of [SSF] State Catch",
    x = "Main Species Group",
    y = "Percentage of State Total (%)",
    color = "State"
  ) +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))

Sp.State_SSF.Biomass_plot

# Save the plot as PNG
ggsave(filename = "/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/CONAPESCA_Biomass&CPUE_Nov2025/Sp.State_SSF.Biomass_plot_Nov182025.png",
       plot = Sp.State_SSF.Biomass_plot,
       width = 10,        # width in inches
       height = 6,        # height in inches
       dpi = 300)         # resolution
```


# CMSY Stock Assesments ------------------------------


# Landings and CPUE Trends ------------------------------
```{r Estimate CPUE for GoC landings}
Landings_GoC<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fis/v2025/int/juan_carlos/GoC_Landings_2000.present_EMF_created.rds"))
# n = 3716415 obs.

CPUE_Landings_GoC<- Landings_GoC

# Calculate the difference in days between period start and end:
CPUE_Landings_GoC$days_fishing <- as.numeric(CPUE_Landings_GoC$period_end_fixed - CPUE_Landings_GoC$period_start_fixed) + 1

# Kind of looks like sometimes 'landed weight' is 'live weight' and vice versa.
CPUE_Landings_GoC$weight_kg_corrected<- c(NA)
ii<- length(CPUE_Landings_GoC$weight_kg_corrected)

CPUE_Landings_GoC$weight_kg_corrected <- ifelse(
  CPUE_Landings_GoC$live_weight >= CPUE_Landings_GoC$landed_weight,
  CPUE_Landings_GoC$live_weight,
  CPUE_Landings_GoC$landed_weight
)

CPUE_Landings_filtered<- CPUE_Landings_GoC %>%
  filter(!is.na(days_fishing) & days_fishing > 0 & live_weight > 0) %>%
  mutate(CPUE_kg.day = weight_kg_corrected / days_fishing)
# 3397596 obs.

max(CPUE_Landings_filtered$CPUE_kg.day) #4500070
min(CPUE_Landings_filtered$CPUE_kg.day) #0

# In general, the receipts from early trips are a bit funky. Let's filter those out for now.
CPUE_Landings_filtered<- CPUE_Landings_filtered %>%
  filter(year >= 2005)
# 2957910 obs.

max(CPUE_Landings_filtered$CPUE_kg.day) #4500070
min(CPUE_Landings_filtered$CPUE_kg.day) #0

##### NOTES TO SELF:
# - There is some funky stuff going on with date end and start. Come back to this.
# - I am not sure what to do about the days with no recorded live weight. Exclude these for now.
```

```{r Summarize Biommas and CPUE in the GoC for major groupings}
unique(CPUE_Landings_filtered$main_species_group)
colnames(CPUE_Landings_filtered)

CPUE_summary <- CPUE_Landings_filtered %>%
  filter(year >= 2014 & year < 2025) %>% # Optional: filter for only the last 10 years
  select(year, state, main_species_group, days_fishing, weight_kg_corrected, value, CPUE_kg.day) %>%
  group_by(year, state, main_species_group) %>%
  summarise(
    Total_Biomass_kg = sum(weight_kg_corrected, na.rm = TRUE),
    Total_Days_Spent_Fishing = sum(days_fishing, na.rm= TRUE),
    Avg_CPUE_kg.days = round(mean(CPUE_kg.day, na.rm= TRUE)),
    Total_Value_mxn = sum(value, na.rm = TRUE),
    n_Records = n(),
    .groups = "drop"  # optional: ungroup after summarising
  )  %>%
  arrange(state, main_species_group, year)
```

```{r Plot Biomass and CPUE}
#### PLOT BIOMASS & CPUE OVER TIME ####
library(ggplot2)
library(ggforce)

#### State Colors ####
state_colors <- c(
  "BAJA CALIFORNIA" = "royalblue",
  "BAJA CALIFORNIA SUR" = "skyblue",
  "SONORA" = "orange",
  "SINALOA" = "sienna1",
  "NAYARIT" = "plum", 
  "JALISCO" = "purple4"
)

# Count total number of pages
num_pages <- n_pages(
  ggplot(CPUE_summary %>% filter(year > 2010),
         aes(x = year, y = Total_Biomass_kg, color = state)) +
    geom_point(size = 2) +
    geom_line(size = 0.5, alpha = 0.5) +
    facet_wrap_paginate(~ main_species_group, scales = "free_y", ncol = 3, nrow = 3)
)

# BIOMASS PLOTS & SAVE:
for (i in 1:num_pages) {
  
  plot_i <- CPUE_summary %>%
    filter(year > 2010) %>%
    ggplot(aes(x = year, y = Total_Biomass_kg, color = state)) +
    geom_point(size = 2) +
    geom_line(size = 0.5, alpha = 0.5) +
    facet_wrap_paginate(~ main_species_group, scales = "free_y",
                        ncol = 3, nrow = 3, page = i) +
    labs(
      title = "Biomass Over Time [per Taxonomic Grouping]",
      subtitle = "Source: CONAPESCA data from J.C., cleaned & filtered (as of Nov 7)",
      x = "Year",
      y = "Biomass (kg)",
      color = "State"
    ) +
    theme_bw() +
    theme(
      legend.position = "bottom",
      plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
      plot.subtitle = element_text(hjust = 0.5, size = 14)
    ) +
    scale_color_manual(values = state_colors, name = "State")
  
  # Save as PNG
  # ggsave(
  #   filename = sprintf("/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/CONAPESCA_Biomass&CPUE_Nov2025/Biomass_Over_Time_Page_%02d.png", i),
  #   plot = plot_i,
  #   width = 12,
  #   height = 12,
  #   dpi = 300
  # )
}

# CPUE PLOTS & SAVE:
for (i in 1:num_pages) {
  
  plot_i <- CPUE_summary %>%
    filter(year > 2010) %>%
    ggplot(aes(x = year, y = Avg_CPUE_kg.days, color = state)) +
    geom_point(size = 2) +
    geom_line(size = 0.5, alpha = 0.5) +
    facet_wrap_paginate(~ main_species_group, scales = "free_y",
                        ncol = 3, nrow = 3, page = i) +
    
    labs(
      title = "CPUE Over Time [per Taxonomic Grouping]",
      subtitle = "Source: CONAPESCA data from J.C., cleaned & filtered (as of Nov 7)",
      x = "Year",
      y = "Average CPUE (kg/day)",
      color = "State"
    ) +
    
    theme_bw() +
    theme(
      legend.position = "bottom",
      plot.title = element_text(hjust = 0.5, face = "bold", size = 18),       # Center and bold title
      plot.subtitle = element_text(hjust = 0.5, size = 12)                    # Center subtitle
    ) +
    scale_color_manual(values = state_colors, name = "State")
  
  # Save as PNG
  # ggsave(
  #   filename = sprintf("/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/CONAPESCA_Biomass&CPUE_Nov2025/CPUE_Over_Time_Page_%02d.png", i),
  #   plot = plot_i,
  #   width = 12,
  #   height = 12,
  #   dpi = 300
  # )
}
```

```{r testing testing 1 2 3}
# Pre-allocate list for storing plots
plots <- vector("list", num_pages)

for (i in 1:num_pages) {

  # Create the plot for this page
  plot_i <- CPUE_summary %>%
    filter(year > 2010) %>%
    ggplot(aes(x = year, y = Avg_CPUE_kg.days, color = state)) +
    geom_point(size = 2) +
    geom_line(size = 0.5, alpha = 0.5) +
    facet_wrap_paginate(~ main_species_group,
                        scales = "free_y",
                        ncol = 3, nrow = 3, page = i) +
    labs(
      title = "CPUE Over Time [per Taxonomic Grouping]",
      subtitle = "Source: CONAPESCA data from J.C., cleaned & filtered (as of Nov 7)",
      x = "Year",
      y = "Average CPUE (kg/day)",
      color = "State"
    ) +
    theme_bw() +
    theme(
      legend.position = "bottom",
      plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
      plot.subtitle = element_text(hjust = 0.5, size = 12)
    ) +
    scale_color_manual(values = state_colors)

  # Store plot in list for interactive viewing later
  plots[[i]] <- plot_i

  # Save PNG
  # ggsave(
  #   filename = sprintf(
  #     "/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/CONAPESCA_Biomass&CPUE_Nov2025/CPUE_Over_Time_Page_%02d.png",
  #     i
  #   ),
  #   plot = plot_i,
  #   width = 12,
  #   height = 12,
  #   dpi = 300
  # )
}

print(plots[[1]])   # page 1
print(plots[[3]])   # page 3
```

```{r Summarize Biomass and CPUE for all groupings}
hist(CPUE_summary$Avg_CPUE_kg.days)

CPUE_all_summary <- CPUE_Landings_filtered %>%
  filter(year >= 2014 & year < 2025) %>% # Optional: filter for only the last 10 years
  # RELEVANT FILTERS:
  # filter(main_species_group != "SARDINA") %>%
  # filter(main_species_group != "ANCHOVETA") %>%
  filter(CPUE_kg.day < 20000) %>%
  
  select(year, state, days_fishing, weight_kg_corrected, value, CPUE_kg.day) %>%
  group_by(year, state) %>%
  summarise(
    Total_Biomass_kg = sum(weight_kg_corrected, na.rm = TRUE),
    Total_Days_Spent_Fishing = sum(days_fishing, na.rm= TRUE),
    Avg_CPUE_kg.days = round(mean(CPUE_kg.day, na.rm= TRUE)),
    Total_Value_mxn = sum(value, na.rm = TRUE),
    n_Records = n(),
    .groups = "drop"  # optional: ungroup after summarising
  )  %>%
  arrange(year, state)
```

```{r Plot CPUE and Biomass across all taxonomic groupings}
## CPUE
AllGroups_CPUE_plot <- CPUE_all_summary %>%
  filter(year > 2010) %>%
  ggplot(aes(x = year, y = Avg_CPUE_kg.days, color = state)) +
  geom_point(size = 2) +
  geom_line(size = 0.5, alpha = 0.5) +
  
  labs(
    title = "CPUE Over Time [across all taxonomic groups]",
    subtitle = "Source: CONAPESCA data from J.C., cleaned & filtered (as of Nov 13)",
    x = "Year",
    y = "Average CPUE (kg/day)",
    color = "State"
  ) +
  theme_bw() +
  geom_vline(xintercept = 2018, linetype = "dashed", color = "black") +
  geom_vline(xintercept = 2019, linetype = "dashed", color = "black") +
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 12)
  ) +
  scale_color_manual(values = state_colors)

AllGroups_CPUE_plot

# Save this plot
# ggsave(AllGroups_CPUE_plot, filename = "/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/CONAPESCA_Biomass&CPUE_Nov2025/AllGroups_CPUE_Over_Time_Nov132025.png",
#   width = 12,
#   height = 12,
#   dpi = 300
# )

## BIOMASS
AllGroups_Biomass_plot <- CPUE_all_summary %>%
  filter(year > 2010) %>%
  ggplot(aes(x = year, y = Total_Biomass_kg, color = state)) +
  geom_point(size = 2) +
  geom_line(size = 0.5, alpha = 0.5) +
  
  labs(
    title = "Landings Over Time [across all taxonomic groups]",
    subtitle = "Source: CONAPESCA data from J.C., cleaned & filtered (as of Nov 13)",
    x = "Year",
    y = "Total Biomass (kg)",
    color = "State"
  ) +
  theme_bw() +
  geom_vline(xintercept = 2018, linetype = "dashed", color = "black") +
  geom_vline(xintercept = 2019, linetype = "dashed", color = "black") +
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 12)
  ) +
  scale_color_manual(values = state_colors)

AllGroups_Biomass_plot

# Save this plot
# ggsave(AllGroups_Biomass_plot, filename = "/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/CONAPESCA_Biomass&CPUE_Nov2025/AllGroups_Biomass_Over_Time_Nov132025.png",
#   width = 12,
#   height = 12,
#   dpi = 300
# )
```

## Testing the effect of subsidy removal on Biomass & CPUE
```{r Testing the effect of subsidy removal on Biomass & CPUE}
CPUE_summary <- CPUE_summary %>%
  mutate(
    subsidy_period = case_when(
      year < 2018 ~ "pre",
      year > 2019 ~ "post",
      TRUE ~ "in between"
    )
  )

CPUE_Landings_filtered2 <- CPUE_Landings_filtered %>%
  filter(year >= 2014 & year < 2025) %>%
  mutate(
    subsidy_period = case_when(
      year < 2018 ~ "pre",
      year > 2019 ~ "post",
      TRUE ~ "in between"
    )
  )

CPUE_Landings_filtered2 %>%
  distinct(year, subsidy_period)
# looks good, four years on either side

## Mixed-effects linear model
library(Matrix)
library(lattice)
# library(lme4) # does not work
library(nlme)
library(sjPlot)

CPUE_Landings_filtered2$main_species_group <- factor(CPUE_Landings_filtered2$main_species_group)

Biomass.SubsidyTest_mixedLM <- lme(
  log10(weight_kg_corrected) ~ subsidy_period,       # fixed effect
  random = ~1 | main_species_group/state,  # nested random intercept
  data = CPUE_Landings_filtered2
)

summary(Biomass.SubsidyTest_mixedLM)

CPUE.SubsidyTest_mixedLM <- lme(
  # fixed effects of subsidy period and time
  CPUE_kg.day ~ subsidy_period + year,
  data = CPUE_Landings_filtered2,
  random = list(
    state = pdSymm(~ 1),   # random intercept + slope for year by state
    main_species_group  = pdDiag(~ 1)       # random intercept for taxa
  )
)

summary(CPUE.SubsidyTest_mixedLM)
```


```{r Biomass & CPUE of major taxa continued...}
options(scipen = 999) # Force same numeric format; don't want scientific notation

# Setting up the dataframe to compare to the C.N.P.
CPUE_Landings_summarized <- CPUE_Landings_filtered %>%
  filter(year %in% 2020:2024) %>%
  select(state, main_species_group, year, CPUE_kg.day, weight_kg_corrected) %>%
  group_by(state, main_species_group) %>%
  summarize(
    # First sum per year, then average across years 2020–2022
    Baseline_Total_Catch_2020_2022 = mean(
      tapply(weight_kg_corrected[year %in% 2020:2022],
             year[year %in% 2020:2022],
             sum, na.rm = TRUE)
    ),
    Total_Catch_2023 = sum(weight_kg_corrected[year == 2023], na.rm = TRUE),
    Total_Catch_2024 = sum(weight_kg_corrected[year == 2024], na.rm = TRUE),
    Avg_CPUE_2020_2022 = mean(CPUE_kg.day[year %in% 2020:2022], na.rm = TRUE),
    Avg_CPUE_2023 = mean(CPUE_kg.day[year == 2023], na.rm = TRUE),
    Avg_CPUE_2024 = mean(CPUE_kg.day[year == 2024], na.rm = TRUE)
  ) %>%
  ungroup() %>%
  mutate(
    # Define a threshold for suspiciously high values
    qc_flag = Baseline_Total_Catch_2020_2022 * 3,
    # Define a 30% margin of error around the baseline mean - SUBSTANTIALLY different
    lowest_bound = Baseline_Total_Catch_2020_2022 * 0.70, # most extreme cases must come first
    highest_bound = Baseline_Total_Catch_2020_2022 * 1.30,
    # Define a 5% margin of error around the baseline mean
    low_bound = Baseline_Total_Catch_2020_2022 * 0.95,
    high_bound = Baseline_Total_Catch_2020_2022 * 1.05,

    # 2023 comparison with margin of error
    CatchCompare_2023_vs_Baseline = case_when(
      Total_Catch_2023 > qc_flag ~ "qc flag: > 3x higher",
      Total_Catch_2023 > highest_bound ~ "substantially higher",
      Total_Catch_2023 < lowest_bound ~ "substantially lower",
      Total_Catch_2023 > high_bound ~ "higher",
      Total_Catch_2023 < low_bound ~ "lower",
      TRUE ~ "similar"
    ),

    # 2024 comparison with margin of error
    CatchCompare_2024_vs_Baseline = case_when(
      Total_Catch_2024 > qc_flag ~ "qc flag: > 3x higher",
      Total_Catch_2024 > highest_bound ~ "substantially higher",
      Total_Catch_2024 < lowest_bound ~ "substantially lower",
      Total_Catch_2024 > high_bound ~ "higher",
      Total_Catch_2024 < low_bound ~ "lower",
      TRUE ~ "similar"
    )
  ) %>%
  select(-low_bound, -high_bound, -lowest_bound, -highest_bound, -qc_flag)   # remove helper columns if not needed

# Add & re-arrange columns:
CPUE_Landings_summarized$CNP_Recommendation<- NA

CPUE_Landings_summarized <- CPUE_Landings_summarized %>%
  select(state,
         main_species_group,
         CNP_Recommendation,
         Baseline_Total_Catch_2020_2022,
         Total_Catch_2023,
         CatchCompare_2023_vs_Baseline,
         Total_Catch_2024,
         CatchCompare_2024_vs_Baseline,
         Avg_CPUE_2020_2022,
         Avg_CPUE_2023,
         Avg_CPUE_2024,
         everything())  # keeps any other columns after
```

```{r Write this as a CSV}
write.csv(CPUE_Landings_summarized, "/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/TaxonLvl_Biomass&CPUE_vs_Baseline_Nov072025.csv", row.names = FALSE)
```

## Load in recommendations for the Carta Nacional Pesquera - UNDER CONSTRUCTION
```{r Load in CNP data}

```
