################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
# Visualization
library(ggplot2)
library(plotly)
library(htmlwidgets)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```


# Stock Assessments from RAM ------------------------------

```{r Mexico FAO Data}
## FAO data, some of the zeros were originally reported as NA (so exercise caution there)
harvest <- read_csv("https://raw.githubusercontent.com/OHI-Science/ohiprep_v2025/refs/heads/main/globalprep/mar/v2025/output/mar_harvest_tonnes.csv")
mex_harvest <- filter(harvest, rgn_id == 135)
mex_harvest <- filter(mex_harvest, year == 2017) %>% data.frame()

# data from Seafood Watch (Monterey Bay)
sustainability <- read_csv("https://raw.githubusercontent.com/OHI-Science/ohiprep_v2025/refs/heads/main/globalprep/mar/v2025/output/mar_sustainability.csv")
mex_sust <- filter(sustainability, taxa_code %in% mex_harvest$taxa_code) %>%
 filter(year==2025) %>% data.frame()

# determine whether sustainability was gapfilled (if so, fairly low confidence, depending on method used and taxa)
gf <- read_csv("https://raw.githubusercontent.com/OHI-Science/ohiprep_v2025/refs/heads/main/globalprep/mar/v2025/output/mar_sustainability_gf.csv")
mex_gf <- filter(gf, taxa_code %in% mex_harvest$taxa_code) %>% data.frame()
```


# Load: Landings data with species info (December 2025) ------------------------------


# Estimating: CPUE over Time for General Groups (November 2025) ------------------------------
```{r CPUE}
CPUE_Landings_GoC<- Landings_GoC

# Calculate the difference in days between period start and end:
CPUE_Landings_GoC$days_fishing <- as.numeric(CPUE_Landings_GoC$period_end_fixed - CPUE_Landings_GoC$period_start_fixed) + 1

# Kind of looks like sometimes 'landed weight' is 'live weight' and vice versa.
CPUE_Landings_GoC$weight_kg_corrected<- c(NA)
ii<- length(CPUE_Landings_GoC$weight_kg_corrected)

CPUE_Landings_GoC$weight_kg_corrected <- ifelse(
  CPUE_Landings_GoC$live_weight >= CPUE_Landings_GoC$landed_weight,
  CPUE_Landings_GoC$live_weight,
  CPUE_Landings_GoC$landed_weight
)

CPUE_Landings_filtered<- CPUE_Landings_GoC %>%
  filter(!is.na(days_fishing) & days_fishing > 0 & live_weight > 0) %>%
  mutate(CPUE_kg.day = weight_kg_corrected / days_fishing)
# 3397596 obs.

max(CPUE_Landings_filtered$CPUE_kg.day) #4500070
min(CPUE_Landings_filtered$CPUE_kg.day) #0

# In general, the receipts from early trips are a bit funky. Let's filter those out for now.
CPUE_Landings_filtered<- CPUE_Landings_filtered %>%
  filter(year >= 2005)
# 2957910 obs.

max(CPUE_Landings_filtered$CPUE_kg.day) #4500070
min(CPUE_Landings_filtered$CPUE_kg.day) #0

##### NOTES TO SELF:
# - There is some funky stuff going on with date end and start. Come back to this.
# - I am not sure what to do about the days with no recorded live weight. Exclude these for now.
```



# CMSY Stock Assesments ------------------------------


# Landings and CPUE Trends ------------------------------
```{r Estimate CPUE for GoC landings}
Landings_GoC<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fis/v2025/int/juan_carlos/GoC_Landings_2000.present_EMF_created.rds"))
# n = 3716415 obs.

CPUE_Landings_GoC<- Landings_GoC

# Calculate the difference in days between period start and end:
CPUE_Landings_GoC$days_fishing <- as.numeric(CPUE_Landings_GoC$period_end_fixed - CPUE_Landings_GoC$period_start_fixed) + 1

# Kind of looks like sometimes 'landed weight' is 'live weight' and vice versa.
CPUE_Landings_GoC$weight_kg_corrected<- c(NA)
ii<- length(CPUE_Landings_GoC$weight_kg_corrected)

CPUE_Landings_GoC$weight_kg_corrected <- ifelse(
  CPUE_Landings_GoC$live_weight >= CPUE_Landings_GoC$landed_weight,
  CPUE_Landings_GoC$live_weight,
  CPUE_Landings_GoC$landed_weight
)

CPUE_Landings_filtered<- CPUE_Landings_GoC %>%
  filter(!is.na(days_fishing) & days_fishing > 0 & live_weight > 0) %>%
  mutate(CPUE_kg.day = weight_kg_corrected / days_fishing)
# 3397596 obs.

max(CPUE_Landings_filtered$CPUE_kg.day) #4500070
min(CPUE_Landings_filtered$CPUE_kg.day) #0

# In general, the receipts from early trips are a bit funky. Let's filter those out for now.
CPUE_Landings_filtered<- CPUE_Landings_filtered %>%
  filter(year >= 2005)
# 2957910 obs.

max(CPUE_Landings_filtered$CPUE_kg.day) #4500070
min(CPUE_Landings_filtered$CPUE_kg.day) #0

##### NOTES TO SELF:
# - There is some funky stuff going on with date end and start. Come back to this.
# - I am not sure what to do about the days with no recorded live weight. Exclude these for now.
```

```{r Summarize Biommas and CPUE in the GoC for major groupings}
unique(CPUE_Landings_filtered$main_species_group)
colnames(CPUE_Landings_filtered)

CPUE_summary <- CPUE_Landings_filtered %>%
  filter(year >= 2014 & year < 2025) %>% # Optional: filter for only the last 10 years
  select(year, state, main_species_group, days_fishing, weight_kg_corrected, value, CPUE_kg.day) %>%
  group_by(year, state, main_species_group) %>%
  summarise(
    Total_Biomass_kg = sum(weight_kg_corrected, na.rm = TRUE),
    Total_Days_Spent_Fishing = sum(days_fishing, na.rm= TRUE),
    Avg_CPUE_kg.days = round(mean(CPUE_kg.day, na.rm= TRUE)),
    Total_Value_mxn = sum(value, na.rm = TRUE),
    n_Records = n(),
    .groups = "drop"  # optional: ungroup after summarising
  )  %>%
  arrange(state, main_species_group, year)
```

```{r Plot Biomass and CPUE}
#### PLOT BIOMASS & CPUE OVER TIME ####
library(ggplot2)
library(ggforce)

#### State Colors ####
state_colors <- c(
  "BAJA CALIFORNIA" = "royalblue",
  "BAJA CALIFORNIA SUR" = "skyblue",
  "SONORA" = "orange",
  "SINALOA" = "sienna1",
  "NAYARIT" = "plum", 
  "JALISCO" = "purple4"
)

# Count total number of pages
num_pages <- n_pages(
  ggplot(CPUE_summary %>% filter(year > 2010),
         aes(x = year, y = Total_Biomass_kg, color = state)) +
    geom_point(size = 2) +
    geom_line(size = 0.5, alpha = 0.5) +
    facet_wrap_paginate(~ main_species_group, scales = "free_y", ncol = 3, nrow = 3)
)

# BIOMASS PLOTS & SAVE:
for (i in 1:num_pages) {
  
  plot_i <- CPUE_summary %>%
    filter(year > 2010) %>%
    ggplot(aes(x = year, y = Total_Biomass_kg, color = state)) +
    geom_point(size = 2) +
    geom_line(size = 0.5, alpha = 0.5) +
    facet_wrap_paginate(~ main_species_group, scales = "free_y",
                        ncol = 3, nrow = 3, page = i) +
    labs(
      title = "Biomass Over Time [per Taxonomic Grouping]",
      subtitle = "Source: CONAPESCA data from J.C., cleaned & filtered (as of Nov 7)",
      x = "Year",
      y = "Biomass (kg)",
      color = "State"
    ) +
    theme_bw() +
    theme(
      legend.position = "bottom",
      plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
      plot.subtitle = element_text(hjust = 0.5, size = 14)
    ) +
    scale_color_manual(values = state_colors, name = "State")
  
  # Save as PNG
  # ggsave(
  #   filename = sprintf("/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/CONAPESCA_Biomass&CPUE_Nov2025/Biomass_Over_Time_Page_%02d.png", i),
  #   plot = plot_i,
  #   width = 12,
  #   height = 12,
  #   dpi = 300
  # )
}

# CPUE PLOTS & SAVE:
for (i in 1:num_pages) {
  
  plot_i <- CPUE_summary %>%
    filter(year > 2010) %>%
    ggplot(aes(x = year, y = Avg_CPUE_kg.days, color = state)) +
    geom_point(size = 2) +
    geom_line(size = 0.5, alpha = 0.5) +
    facet_wrap_paginate(~ main_species_group, scales = "free_y",
                        ncol = 3, nrow = 3, page = i) +
    
    labs(
      title = "CPUE Over Time [per Taxonomic Grouping]",
      subtitle = "Source: CONAPESCA data from J.C., cleaned & filtered (as of Nov 7)",
      x = "Year",
      y = "Average CPUE (kg/day)",
      color = "State"
    ) +
    
    theme_bw() +
    theme(
      legend.position = "bottom",
      plot.title = element_text(hjust = 0.5, face = "bold", size = 18),       # Center and bold title
      plot.subtitle = element_text(hjust = 0.5, size = 12)                    # Center subtitle
    ) +
    scale_color_manual(values = state_colors, name = "State")
  
  # Save as PNG
  # ggsave(
  #   filename = sprintf("/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/CONAPESCA_Biomass&CPUE_Nov2025/CPUE_Over_Time_Page_%02d.png", i),
  #   plot = plot_i,
  #   width = 12,
  #   height = 12,
  #   dpi = 300
  # )
}
```

```{r testing testing 1 2 3}
# Pre-allocate list for storing plots
plots <- vector("list", num_pages)

for (i in 1:num_pages) {

  # Create the plot for this page
  plot_i <- CPUE_summary %>%
    filter(year > 2010) %>%
    ggplot(aes(x = year, y = Avg_CPUE_kg.days, color = state)) +
    geom_point(size = 2) +
    geom_line(size = 0.5, alpha = 0.5) +
    facet_wrap_paginate(~ main_species_group,
                        scales = "free_y",
                        ncol = 3, nrow = 3, page = i) +
    labs(
      title = "CPUE Over Time [per Taxonomic Grouping]",
      subtitle = "Source: CONAPESCA data from J.C., cleaned & filtered (as of Nov 7)",
      x = "Year",
      y = "Average CPUE (kg/day)",
      color = "State"
    ) +
    theme_bw() +
    theme(
      legend.position = "bottom",
      plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
      plot.subtitle = element_text(hjust = 0.5, size = 12)
    ) +
    scale_color_manual(values = state_colors)

  # Store plot in list for interactive viewing later
  plots[[i]] <- plot_i

  # Save PNG
  # ggsave(
  #   filename = sprintf(
  #     "/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/CONAPESCA_Biomass&CPUE_Nov2025/CPUE_Over_Time_Page_%02d.png",
  #     i
  #   ),
  #   plot = plot_i,
  #   width = 12,
  #   height = 12,
  #   dpi = 300
  # )
}

print(plots[[1]])   # page 1
print(plots[[3]])   # page 3
```

```{r Summarize Biomass and CPUE for all groupings}
hist(CPUE_summary$Avg_CPUE_kg.days)

CPUE_all_summary <- CPUE_Landings_filtered %>%
  filter(year >= 2014 & year < 2025) %>% # Optional: filter for only the last 10 years
  # RELEVANT FILTERS:
  # filter(main_species_group != "SARDINA") %>%
  # filter(main_species_group != "ANCHOVETA") %>%
  filter(CPUE_kg.day < 20000) %>%
  
  select(year, state, days_fishing, weight_kg_corrected, value, CPUE_kg.day) %>%
  group_by(year, state) %>%
  summarise(
    Total_Biomass_kg = sum(weight_kg_corrected, na.rm = TRUE),
    Total_Days_Spent_Fishing = sum(days_fishing, na.rm= TRUE),
    Avg_CPUE_kg.days = round(mean(CPUE_kg.day, na.rm= TRUE)),
    Total_Value_mxn = sum(value, na.rm = TRUE),
    n_Records = n(),
    .groups = "drop"  # optional: ungroup after summarising
  )  %>%
  arrange(year, state)
```

```{r Plot CPUE and Biomass across all taxonomic groupings}
## CPUE
AllGroups_CPUE_plot <- CPUE_all_summary %>%
  filter(year > 2010) %>%
  ggplot(aes(x = year, y = Avg_CPUE_kg.days, color = state)) +
  geom_point(size = 2) +
  geom_line(size = 0.5, alpha = 0.5) +
  
  labs(
    title = "CPUE Over Time [across all taxonomic groups]",
    subtitle = "Source: CONAPESCA data from J.C., cleaned & filtered (as of Nov 13)",
    x = "Year",
    y = "Average CPUE (kg/day)",
    color = "State"
  ) +
  theme_bw() +
  geom_vline(xintercept = 2018, linetype = "dashed", color = "black") +
  geom_vline(xintercept = 2019, linetype = "dashed", color = "black") +
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 12)
  ) +
  scale_color_manual(values = state_colors)

AllGroups_CPUE_plot

# Save this plot
# ggsave(AllGroups_CPUE_plot, filename = "/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/CONAPESCA_Biomass&CPUE_Nov2025/AllGroups_CPUE_Over_Time_Nov132025.png",
#   width = 12,
#   height = 12,
#   dpi = 300
# )

## BIOMASS
AllGroups_Biomass_plot <- CPUE_all_summary %>%
  filter(year > 2010) %>%
  ggplot(aes(x = year, y = Total_Biomass_kg, color = state)) +
  geom_point(size = 2) +
  geom_line(size = 0.5, alpha = 0.5) +
  
  labs(
    title = "Landings Over Time [across all taxonomic groups]",
    subtitle = "Source: CONAPESCA data from J.C., cleaned & filtered (as of Nov 13)",
    x = "Year",
    y = "Total Biomass (kg)",
    color = "State"
  ) +
  theme_bw() +
  geom_vline(xintercept = 2018, linetype = "dashed", color = "black") +
  geom_vline(xintercept = 2019, linetype = "dashed", color = "black") +
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 12)
  ) +
  scale_color_manual(values = state_colors)

AllGroups_Biomass_plot

# Save this plot
# ggsave(AllGroups_Biomass_plot, filename = "/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/CONAPESCA_Biomass&CPUE_Nov2025/AllGroups_Biomass_Over_Time_Nov132025.png",
#   width = 12,
#   height = 12,
#   dpi = 300
# )
```

## Testing the effect of subsidy removal on Biomass & CPUE
```{r Testing the effect of subsidy removal on Biomass & CPUE}
CPUE_summary <- CPUE_summary %>%
  mutate(
    subsidy_period = case_when(
      year < 2018 ~ "pre",
      year > 2019 ~ "post",
      TRUE ~ "in between"
    )
  )

CPUE_Landings_filtered2 <- CPUE_Landings_filtered %>%
  filter(year >= 2014 & year < 2025) %>%
  mutate(
    subsidy_period = case_when(
      year < 2018 ~ "pre",
      year > 2019 ~ "post",
      TRUE ~ "in between"
    )
  )

CPUE_Landings_filtered2 %>%
  distinct(year, subsidy_period)
# looks good, four years on either side

## Mixed-effects linear model
library(Matrix)
library(lattice)
# library(lme4) # does not work
library(nlme)
library(sjPlot)

CPUE_Landings_filtered2$main_species_group <- factor(CPUE_Landings_filtered2$main_species_group)

Biomass.SubsidyTest_mixedLM <- lme(
  log10(weight_kg_corrected) ~ subsidy_period,       # fixed effect
  random = ~1 | main_species_group/state,  # nested random intercept
  data = CPUE_Landings_filtered2
)

summary(Biomass.SubsidyTest_mixedLM)

CPUE.SubsidyTest_mixedLM <- lme(
  # fixed effects of subsidy period and time
  CPUE_kg.day ~ subsidy_period + year,
  data = CPUE_Landings_filtered2,
  random = list(
    state = pdSymm(~ 1),   # random intercept + slope for year by state
    main_species_group  = pdDiag(~ 1)       # random intercept for taxa
  )
)

summary(CPUE.SubsidyTest_mixedLM)
```


```{r Biomass & CPUE of major taxa continued...}
options(scipen = 999) # Force same numeric format; don't want scientific notation

# Setting up the dataframe to compare to the C.N.P.
CPUE_Landings_summarized <- CPUE_Landings_filtered %>%
  filter(year %in% 2020:2024) %>%
  select(state, main_species_group, year, CPUE_kg.day, weight_kg_corrected) %>%
  group_by(state, main_species_group) %>%
  summarize(
    # First sum per year, then average across years 2020–2022
    Baseline_Total_Catch_2020_2022 = mean(
      tapply(weight_kg_corrected[year %in% 2020:2022],
             year[year %in% 2020:2022],
             sum, na.rm = TRUE)
    ),
    Total_Catch_2023 = sum(weight_kg_corrected[year == 2023], na.rm = TRUE),
    Total_Catch_2024 = sum(weight_kg_corrected[year == 2024], na.rm = TRUE),
    Avg_CPUE_2020_2022 = mean(CPUE_kg.day[year %in% 2020:2022], na.rm = TRUE),
    Avg_CPUE_2023 = mean(CPUE_kg.day[year == 2023], na.rm = TRUE),
    Avg_CPUE_2024 = mean(CPUE_kg.day[year == 2024], na.rm = TRUE)
  ) %>%
  ungroup() %>%
  mutate(
    # Define a threshold for suspiciously high values
    qc_flag = Baseline_Total_Catch_2020_2022 * 3,
    # Define a 30% margin of error around the baseline mean - SUBSTANTIALLY different
    lowest_bound = Baseline_Total_Catch_2020_2022 * 0.70, # most extreme cases must come first
    highest_bound = Baseline_Total_Catch_2020_2022 * 1.30,
    # Define a 5% margin of error around the baseline mean
    low_bound = Baseline_Total_Catch_2020_2022 * 0.95,
    high_bound = Baseline_Total_Catch_2020_2022 * 1.05,

    # 2023 comparison with margin of error
    CatchCompare_2023_vs_Baseline = case_when(
      Total_Catch_2023 > qc_flag ~ "qc flag: > 3x higher",
      Total_Catch_2023 > highest_bound ~ "substantially higher",
      Total_Catch_2023 < lowest_bound ~ "substantially lower",
      Total_Catch_2023 > high_bound ~ "higher",
      Total_Catch_2023 < low_bound ~ "lower",
      TRUE ~ "similar"
    ),

    # 2024 comparison with margin of error
    CatchCompare_2024_vs_Baseline = case_when(
      Total_Catch_2024 > qc_flag ~ "qc flag: > 3x higher",
      Total_Catch_2024 > highest_bound ~ "substantially higher",
      Total_Catch_2024 < lowest_bound ~ "substantially lower",
      Total_Catch_2024 > high_bound ~ "higher",
      Total_Catch_2024 < low_bound ~ "lower",
      TRUE ~ "similar"
    )
  ) %>%
  select(-low_bound, -high_bound, -lowest_bound, -highest_bound, -qc_flag)   # remove helper columns if not needed

# Add & re-arrange columns:
CPUE_Landings_summarized$CNP_Recommendation<- NA

CPUE_Landings_summarized <- CPUE_Landings_summarized %>%
  select(state,
         main_species_group,
         CNP_Recommendation,
         Baseline_Total_Catch_2020_2022,
         Total_Catch_2023,
         CatchCompare_2023_vs_Baseline,
         Total_Catch_2024,
         CatchCompare_2024_vs_Baseline,
         Avg_CPUE_2020_2022,
         Avg_CPUE_2023,
         Avg_CPUE_2024,
         everything())  # keeps any other columns after
```

```{r Write this as a CSV}
write.csv(CPUE_Landings_summarized, "/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/TaxonLvl_Biomass&CPUE_vs_Baseline_Nov072025.csv", row.names = FALSE)
```

## Load in recommendations for the Carta Nacional Pesquera - UNDER CONSTRUCTION
```{r Load in CNP data}

```
