---
title: "Simple CMSY_Oct 22 2025"
output: html_document
date: "2025-10-22"
---

# PRE-WORK
```{r Loading Libraries}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(tidytext)
library(lubridate)
library(plotly)
library(htmlwidgets)
library(here)
library(janitor)
```

```{r Review PDF file for Information: NAME}
"CMSY Data-Poor Stock Analysis Protocol_v1.pdf"
```


# GENERATING FILES 1 & 2 FROM SCRATCH
Preparing Files 1 and 2
NOTES TO SELF - Oct 22 2025

FILE 1: Create a file with annual landings in kg (and n_count observations) grouped by year per "stock". Here, stock is defined as the total population of taxa X fished within the Gulf of California not including not the Pacific.

- At present, we do not have the municipality lat-longs so I will group stocks by the six state of interest first (Baja, BCS, Son, Sin, Nay, Jal) and then plan to re-make the file (File 1) once we can filter out specific municipalities.
- This means that this timeseries is likely to be an OVER-ESTIMATE of landings.

File 2: Create a file with Region, [Subregion], Stock, Group = Name, English Name, Scientific Name, Source, and a whole bunch of year columns that you populate using the various R files shared by Alfredo. This file can probably be trimmed down but for now I recommend formatting it exactly as it appears from before.

```{r File 1 - Stock Timeseries (JC data)}
# CONAPESCA data shared by JC
# These data were shared with us by Juan Carlos Villaseñor and originally accessed on September 26, 2025 from this GitHub repository: https://github.com/mex-fisheries

MexLandings_2000.present<- readRDS(here("/home/shares/ohi/OHI_GOC/_raw_data/CONAPESCA/d2025/SharedByJuanCarlos/mex_landings/data/clean/mex_landings_2000_present.rds"))
# n = 12,153,879 records from all states in Mexico from January 2000 through January 2025


# Skimming through the dat. What does it look like?
# Yes aquaculture production
# Records up to January, 2025

head(MexLandings_2000.present)

table(MexLandings_2000.present$acuaculture_production)

max(MexLandings_2000.present$year)
max(MexLandings_2000.present$month[MexLandings_2000.present$year])
```

```{r Basic filtering}
States_of_Interest<- c("BAJA CALIFORNIA", "BAJA CALIFORNIA SUR", "SONORA", "SINALOA", "NAYARIT", "JALISCO")

MexLandings_filtered<- MexLandings_2000.present %>% filter(state %in% States_of_Interest)
# n obs.= 4,713,144 (-7,440,735 obs)

table(MexLandings_filtered$fleet)
# large_scale small_scale 
#      181823     1438462 
table(MexLandings_filtered$acuaculture_production) 
#there are also mariculture landings in this dataset and NAs in this column...

FisheryLandings_filtered<- MexLandings_filtered %>% filter(acuaculture_production== "NO" | is.na(acuaculture_production)) # 4684000 obs.

MaricultureLandings_filtered<- MexLandings_filtered %>% filter(acuaculture_production== "SÍ")
# n obs.= 29,144 (-4,684,000 obs)

# NOTE: SOME RECORDS ARE NOT APPENDED WITH INFORMATION ABOUT RECORD TYPE: no information on wild-catch versus aquaculture. For this preliminary analyses, these records are assumed to be landings from fishery operations and are included.

table(FisheryLandings_filtered$year)

# OTRAS_filtered<- MexLandings_filtered %>% filter(main_species_group== "OTRAS")
# n obs.= 715.547 (-1,011,581 obs)
```

```{r Summary table}
# Turn off scientific notation:
options(scipen = 999)

# Create timeseries file 1
head(FisheryLandings_filtered)


StateLandings_AnnualSummary_byTaxa <- FisheryLandings_filtered %>%
  select(year, state, main_species_group, live_weight, landed_weight, value) %>%
  group_by(main_species_group, year, state) %>%
  summarise(
    Total_Live_Biomass = sum(live_weight, na.rm = TRUE),
    Total_Landed_Biomass = sum(landed_weight, na.rm = TRUE),
    Total_Value = sum(value, na.rm = TRUE),
    n_Records = n(),
    .groups = "drop"  # optional: ungroup after summarising
  )

TotalLandings_AnnualSummary_byTaxa <- FisheryLandings_filtered %>%
  select(year, main_species_group, live_weight, landed_weight, value) %>%
  group_by(main_species_group, year) %>%
  summarise(
    Total_Live_Biomass = sum(live_weight, na.rm = TRUE),
    Total_Landed_Biomass = sum(landed_weight, na.rm = TRUE),
    Total_Value = sum(value, na.rm = TRUE),
    n_Records = n(),
    .groups = "drop"  # optional: ungroup after summarising
  )
```


```{r Create File 2 (Metadata File) based on File 1}
# Which taxa (stocks) do we want to account for in our File 2?
Taxa_names<- unique(TotalLandings_AnnualSummary_byTaxa$main_species_group)
# don’t 57 unique taxa

# Read in the column names of File 2 from Ferrer et al. (2022) CSV file
headers <- names(read.csv("File 2_Stocks Metadata.csv", nrows = 1))

# Create an empty dataframe with the same headers
File2_df <- data.frame(matrix(ncol = length(headers), nrow = 0))
colnames(File2_df) <- headers

# Add Taxa_names to the 'Stock' column
File2_df <- File2_df[1:length(Taxa_names), ]  # Expand rows
File2_df$Stock <- Taxa_names

# Reset rownames / the index column numbered 1-57
rownames(File2_df) <- NULL

# Trim some of the columns off this file:
File2_df <- File2_df[ , !(names(File2_df) %in% c("ID", "Subregion", "Group", "Name", "Pick")) ]

# Fill in some of the columns off this file:
File2_df$Region <- c("GOC")
File2_df$Class <- c("TBD")
File2_df$Source <- c("JC_data")
File2_df$Keep <- c(1)

# Re-order columns:
File2_df <- File2_df[c(1, 3, 2, 4:5, 6, 37, 38, 7:36)]

# NEXT STEPS: Fill in the rest of column 2 using the ancillary files shared by Alfredo in Ferrer et al. (2022)
write.csv(File2_df, "File2_UNDER_CONSTRUCTION.csv", row.names = FALSE)
write.csv(File2_df, "~/ohi_goc_prep/FP/File2_UNDER_CONSTRUCTION.csv", row.names = FALSE)
```

# RUNNING CMSY WITH FILES 1 & 2