---
title: 'OHI GoC 2025: Livelihoods & Economies'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# LE: Censos Economicos (CE) 

For number of jobs, quality of jobs, and revenue of certain sectors.  


```{r}
library(tidyverse)
library(mapview)
library(sf)
```

## Isolate municipos

Using this pdf of the codes for each municipality: <https://www.gob.mx/cms/uploads/attachment/file/179869/Municipios_por_Regiones__atendidos_.pdf>

NOTE THAT SAN FELIPE was recently added in 2022 as a municipality.  Updated municipalities will need to be used. 

```{r}
library(DT)

raw_text <- paste(
  "02 BAJA_CALIFORNIA 001 ENSENADA 2 2.59",
  "02 BAJA_CALIFORNIA 002 MEXICALI 1 2.02",
  "02 BAJA_CALIFORNIA 003 TECATE 6 4.28",
  "02 BAJA_CALIFORNIA 004 TIJUANA 1 2.02",
  "02 BAJA_CALIFORNIA 005 PLAYAS_DE_ROSARITO 2 2.59",
  "02 BAJA_CALIFORNIA 006 SAN QUINTÍN 2 NA", ## added manually... because of 2022 changes: https://en.wikipedia.org/wiki/Municipalities_of_Baja_California
  "02 BAJA_CALIFORNIA 007 SAN FELIPE 2 NA", ## added manually... because of 2022 changes: https://en.wikipedia.org/wiki/Municipalities_of_Baja_California
  "03 BAJA_CALIFORNIA_SUR 001 COMONDU 3 3.03",
  "03 BAJA_CALIFORNIA_SUR 002 MULEGE 3 3.03",
  "03 BAJA_CALIFORNIA_SUR 003 LA_PAZ 3 3.03",
  "03 BAJA_CALIFORNIA_SUR 008 LOS_CABOS 3 3.03",
  "03 BAJA_CALIFORNIA_SUR 009 LORETO 4 3.53",
  "14 JALISCO 067 PUERTO_VALLARTA 6 4.28",
  "14 JALISCO 020 CABO_CORRIENTES 1 2.02",
  "18 NAYARIT 001 ACAPONETA 2 2.59",
  "18 NAYARIT 002 AHUACATLAN 1 2.02",
  "18 NAYARIT 003 AMATLAN_DE_CAÑAS 1 2.02",
  "18 NAYARIT 004 COMPOSTELA 2 2.59",
  "18 NAYARIT 006 IXTLAN_DEL_RIO 1 2.02",
  "18 NAYARIT 007 JALA 6 4.28",
  "18 NAYARIT 008 XALISCO 2 2.59",
  "18 NAYARIT 010 ROSAMORADA 2 2.59",
  "18 NAYARIT 011 RUIZ 2 2.59",
  "18 NAYARIT 012 SAN_BLAS 3 3.03",
  "18 NAYARIT 013 SAN_PEDRO_LAGUNILLAS 1 2.02",
  "18 NAYARIT 014 SANTA_MARIA_DEL_ORO 3 3.03",
  "18 NAYARIT 015 SANTIAGO_IXCUINTLA 2 2.59",
  "18 NAYARIT 016 TECUALA 1 2.02",
  "18 NAYARIT 017 TEPIC 6 4.28",
  "18 NAYARIT 018 TUXPAN 3 3.03",
  "18 NAYARIT 020 BAHIA_DE_BANDERAS 3 3.03",
  "25 SINALOA 001 AHOME 4 3.53",
  "25 SINALOA 002 ANGOSTURA 2 2.59",
  "25 SINALOA 003 BADIRAGUATO 6 4.28",
  "25 SINALOA 004 CONCORDIA 2 2.59",
  "25 SINALOA 005 COSALA 3 3.03",
  "25 SINALOA 006 CULIACAN 5 3.90",
  "25 SINALOA 007 CHOIX 5 3.90",
  "25 SINALOA 008 ELOTA 5 NA", # added in when found missing 
  "25 SINALOA 009 ESCUINAPA 2 2.59",
  "25 SINALOA 010 EL_FUERTE 3 3.03",
  "25 SINALOA 011 GUASAVE 6 4.28",
  "25 SINALOA 012 MAZATLAN 5 3.90",
  "25 SINALOA 013 MOCORITO 2 2.59",
  "25 SINALOA 014 ROSARIO 2 2.59",
  "25 SINALOA 015 SALVADOR_ALVARADO 2 2.59",
  "25 SINALOA 016 SAN_IGNACIO 2 NA", # added in when found missing 
  "25 SINALOA 017 SINALOA 3 3.03",
  "25 SINALOA 018 NAVOLATO 2 2.59",
  "26 SONORA 002 AGUA_PRIETA 6 4.28",
  "26 SONORA 003 ALAMOS 3 3.03",
  "26 SONORA 004 ALTAR 2 2.59",
  "26 SONORA 007 ATIL 6 4.28",
  "26 SONORA 012 BACUM 5 3.90",
  "26 SONORA 014 BAVIACORA 5 3.90",
  "26 SONORA 016 BENJAMIN_HILL 6 4.28",
  "26 SONORA 017 CABORCA 2 2.59",
  "26 SONORA 018 CAJEME 6 4.28",
  "26 SONORA 025 EMPALME 6 4.28",
  "26 SONORA 026 ETCHOJOA 6 4.28",
  "26 SONORA 029 GUAYMAS 5 3.90",
  "26 SONORA 030 HERMOSILLO 6 4.28",
  "26 SONORA 033 HUATABAMPO 3 3.03",
  "26 SONORA 035 IMURIS 6 4.28",
  "26 SONORA 036 MAGDALENA 5 3.90",
  "26 SONORA 037 MAZATAN 2 2.59",
  "26 SONORA 042 NAVOJOA 6 4.28",
  "26 SONORA 043 NOGALES 3 3.03",
  "26 SONORA 045 OPODEPE 4 3.53",
  "26 SONORA 047 PITIQUITO 3 3.03",
  "26 SONORA 048 PUERTO_PEÑASCO 3 3.03",
  "26 SONORA 054 SAN_JAVIER 3 3.03",
  "26 SONORA 055 SAN_LUIS_RIO_COLORADO 4 3.53",
  "26 SONORA 056 SAN_MIGUEL_DE_HORCASITAS 4 3.53",
  "26 SONORA 058 SANTA_ANA 5 3.90",
  "26 SONORA 065 TUBUTAMA 6 4.28",
  "26 SONORA 070 GENERAL_PLUTARCO_ELIAS_CALLES 1 2.02",
  "26 SONORA 071 BENITO_JUAREZ 3 3.03",
  "26 SONORA 072 SAN_IGNACIO_RIO_MUERTO 5 3.90",
  sep = "\n"
)

library(purrr)

lines <- str_split(raw_text, "\n")[[1]]

municipio_codigo_raw_df <- map_dfr(lines, function(x) {
  parts <- str_split(x, " +")[[1]]   # split on spaces
  tibble(
    clave_estado   = parts[1],
    estado         = parts[2] |> str_replace_all("_", " "),
    clave_municipio = parts[3],
    region         = as.integer(parts[length(parts) - 1]),
    tasa_regional  = as.numeric(parts[length(parts)]),
    municipio      = paste(parts[4:(length(parts) - 2)], collapse = " ") |> 
                      str_replace_all("_", " ")
  )
}) 

municipio_codigo_df <- municipio_codigo_raw_df %>% 
  select(clave_estado, estado, clave_municipio, municipio, tasa_regional)

write_csv(municipio_codigo_df, here::here("/home/shares/ohi/OHI_GOC/spatial/land_admin_boundaries", "municipio_codigos_secreteria_de_economia.csv"))
```

Combine with shapefile for each municipality.

```{r}
municipios <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/land_admin_boundaries/mex_admin_boundaries.shp","mex_admin2.shp")) %>% 
  mutate(nom_mun = str_to_upper(adm2_ref_n),
         nom_ent = str_to_upper(adm1_name)) %>% 
  select(nom_ent, nom_mun) %>% 
  filter(nom_ent %in% c("BAJA CALIFORNIA SUR", "BAJA CALIFORNIA", "SINALOA", "SONORA", "NAYARIT", "JALISCO")) %>% 
  mutate(
    nom_mun = str_replace(nom_mun, "PUERTO PENASCO", "PUERTO PEÑASCO"),
    nom_mun = str_replace(nom_mun, "AMATLAN DE CANAS", "AMATLAN DE CAÑAS")
  )

mapview(municipios)

# see if they match
setdiff(municipio_codigo_sf$municipio, municipios$nom_mun) # character(0)
# setdiff(municipios$nom_mun, municipio_codigo_sf$municipio)

municipio_codigo_sf <- inner_join(municipio_codigo_df, municipios, by = c("municipio" = "nom_mun", "estado" = "nom_ent")) %>% 
  st_as_sf()

mapview(municipio_codigo_sf)

# st_write(municipio_codigo_sf, here::here("/home/shares/ohi/OHI_GOC/spatial/land_admin_boundaries", "mex_admin2_joined_w_municipio_codigos.shp"), append = FALSE)
```


## Read in data

### 2024
```{r}
# baja california
bc_2024_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2024/conjunto_de_datos_ce_bc_2024_csv/conjunto_de_datos","tr_ce_bc_2024.csv")) 
unique(bc_2024_raw$E04) # [1] NA    "001" "002" "003" "004" "005" "006" "007", now has San Felipe!! 

bc_diccionario_2024_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2024/conjunto_de_datos_ce_bc_2024_csv/diccionario_de_datos","diccionario_de_datos_ce2024.csv"))

bc_actividad_2024_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2024/conjunto_de_datos_ce_bc_2024_csv/catalogos","tc_codigo_actividad.csv"))

# baja california sur
bcs_2024_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2024/conjunto_de_datos_ce_bcs_2024_csv/conjunto_de_datos","tr_ce_bcs_2024.csv"))

# jalisco
jal_2024_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2024/conjunto_de_datos_ce_jal_2024_csv/conjunto_de_datos","tr_ce_jal_2024.csv"))

# nayarit
nay_2024_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2024/conjunto_de_datos_ce_nay_2024_csv/conjunto_de_datos","tr_ce_nay_2024.csv"))

# sinaloa
sin_2024_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2024/conjunto_de_datos_ce_sin_2024_csv/conjunto_de_datos","tr_ce_sin_2024.csv"))

# sonora
son_2024_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2024/conjunto_de_datos_ce_son_2024_csv/conjunto_de_datos","tr_ce_son_2024.csv"))
```

Merge them into one df:
```{r}
fix_types <- \(df) df |>
  mutate(
    E03 = as.numeric(E03),
    E04 = as.numeric(E04),
    CODIGO = as.character(CODIGO)
  )

ce_bind_2024 <- bind_rows(
  fix_types(bc_2024_raw),
  fix_types(bcs_2024_raw),
  fix_types(jal_2024_raw),
  fix_types(nay_2024_raw),
  fix_types(sin_2024_raw),
  fix_types(son_2024_raw)
)
```



### 2019
```{r}
# baja california
bc_2019_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2019/ce2019_bc_csv/conjunto_de_datos","ce2019_bc.csv"))
unique(bc_2019_raw$MUNICIPIO) # [1] NA    "001" "002" "003" "004" "005"

bc_actividad_2019_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2019/ce2019_bc_csv/catalogos","tc_codigo_actividad.csv"))

# baja california sur
bcs_2019_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2019/ce2019_bcs_csv/conjunto_de_datos","ce2019_bcs.csv"))

# jalisco
jal_2019_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2019/ce2019_jal_csv/conjunto_de_datos","ce2019_jal.csv"))

# nayarit
nay_2019_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2019/ce2019_nay_csv/conjunto_de_datos","ce2019_nay.csv"))

# sinaloa
sin_2019_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2019/ce2019_sin_csv/conjunto_de_datos","ce2019_sin.csv"))

# sonora
son_2019_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2019/ce2019_son_csv/conjunto_de_datos","ce2019_son.csv"))
```

Merge them:
```{r}
spec(bc_2019_raw)
spec(jal_2019_raw)

fix_types <- \(df) df |>
  mutate(
    ENTIDAD = as.numeric(ENTIDAD),
    MUNICIPIO = as.numeric(MUNICIPIO),
    CODIGO = as.character(CODIGO)
  )

ce_bind_2019 <- bind_rows(
  fix_types(bc_2019_raw),
  fix_types(bcs_2019_raw),
  fix_types(jal_2019_raw),
  fix_types(nay_2019_raw),
  fix_types(sin_2019_raw),
  fix_types(son_2019_raw)
)
```


### 2014, 2009, 2004
```{r}
# baja california
bc_2014_2009_2004_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2014_2009_2004/ce_02_bc_csv/conjunto_de_datos","ce_inegi_02.csv"))
unique(bc_2014_2009_2004_raw$clave_municipio) # [1] NA    "001" "002" "003" "004" "005"

# baja california sur
bcs_2014_2009_2004_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2014_2009_2004/ce_03_bcs_csv/conjunto_de_datos","ce_inegi_03.csv"))

# jalisco
jal_2014_2009_2004_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2014_2009_2004/ce_14_jal_csv/conjunto_de_datos","ce_inegi_14.csv"))

# nayarit
nay_2014_2009_2004_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2014_2009_2004/ce_18_nay_csv/conjunto_de_datos","ce_inegi_18.csv"))

# sinaloa
sin_2014_2009_2004_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2014_2009_2004/ce_25_sin_csv/conjunto_de_datos","ce_inegi_25.csv"))

# sonora
son_2014_2009_2004_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2014_2009_2004/ce_26_son_csv/conjunto_de_datos","ce_inegi_26.csv"))
```

Merge them:
```{r}
spec(bc_2014_2009_2004_raw)
spec(bcs_2014_2009_2004_raw)
spec(nay_2014_2009_2004_raw)
spec(jal_2014_2009_2004_raw)
spec(sin_2014_2009_2004_raw)
spec(son_2014_2009_2004_raw)


fix_types <- \(df) df %>% 
  mutate(
    clave_entidad = as.numeric(clave_entidad),  # bc has char, others have numeric
    clave_municipio = as.numeric(clave_municipio),
    across(c(J400A, J500A, J600A, K010A, K020A, K030A, K060A, K070A, 
             K096A, K311A, K620A, K810A, K910A, M010A, M030A, 
             P000A, P000B, P000C, P030A, P030B, P030C, P100A, P100B,
             Q020A, Q030A), 
           ~(as.numeric(.x))),
    anio_censal = as.character(anio_censal)
  )

ce_bind_2014_2009_2004 <- bind_rows(
  fix_types(bc_2014_2009_2004_raw),
  fix_types(bcs_2014_2009_2004_raw),
  fix_types(jal_2014_2009_2004_raw),
  fix_types(nay_2014_2009_2004_raw),
  fix_types(sin_2014_2009_2004_raw),
  fix_types(son_2014_2009_2004_raw)
)

# pesca_lapaz <- ce_bind_2014_2009_2004 %>% 
  # filter(clave_actividad_economica %in% "1141") %>% 
  # filter(clave_entidad %in% "3") %>% 
  # filter(clave_municipio %in% "003")
```


## Merge

First do some explorations of the differences between the years.

```{r}
setdiff(colnames(ce_bind_2014_2009_2004), colnames(ce_bind_2019))
# [1] "clave_entidad"             "entidad_federativa"        "clave_municipio"           "municipio"                
# [5] "clave_actividad_economica" "actividad_economica"       "anio_censal"               "K040A"                    
# [9] "K041A"
setdiff(colnames(ce_bind_2014_2009_2004), colnames(ce_bind_2024))
# [1] "clave_entidad"             "entidad_federativa"        "clave_municipio"           "municipio"                
# [5] "clave_actividad_economica" "actividad_economica"       "anio_censal"               "K040A"                    
# [9] "K041A" 

setdiff(colnames(ce_bind_2019), colnames(ce_bind_2024)) # what is in 2019 that is not in 2024
#  [1] "ENTIDAD"   "MUNICIPIO" "A112A"     "A113A"     "A114A"     "A115A"     "A116A"     "A117A"     "A118A"    
# [10] "A119A"     "A141A"     "A143A"     "A144A"     "A145A"     "A146A"     "A147A"     "A148A"     "A149A"    
# [19] "A151A"     "A152A"     "A153A"     "A154A"     "A155A"     "A156A"     "A157A"     "A158A"     "A159A"    
# [28] "A171A"     "A173A"     "A174A"     "A175A"     "A176A"     "A177A"     "A179A"     "A181A"     "A182A"    
# [37] "A192A"     "A193A"     "A194A"     "A195A"     "A197A"     "A201A"     "A202A"     "A204A"     "A206A"    
# [46] "A208A"     "A209A"     "A212A"     "A213A"     "A214A"     "A215A"     "A216A"     "A217A"     "A218A"    
# [55] "A219A"     "A222A"     "A223A"     "A224A"     "A225A"     "A226A"     "A227A"     "A228A"     "A229A"    
# [64] "A231A"     "A235A"     "A236A"     "A237A"     "A238A"     "A242A"     "A243A"     "A244A"     "A245A"    
# [73] "A247A"     "A248A"     "A252A"     "A253A"     "A256A"     "A433A"     "A434A"     "A436A"     "A529A"    
# [82] "A534A"     "A747A"     "A748A"     "A752A"     "A764A"    

setdiff(colnames(ce_bind_2024), colnames(ce_bind_2019)) # what is in 2024 that is not in 2019
# [1] "E03"       "E04"       "SECTOR"    "SUBSECTOR" "RAMA"      "SUBRAMA"   "CLASE"
```

### Isolate metrics

What we really need is:

* The number of jobs, *H000A*;
* The employer contributions to social security schemes *J300A*,
* Other social benefits *J400A*;
* Total revenue from the supply of goods and services (millions of pesos) *M000A*; and
* Total revenue (millions of pesos) *A800A*

For now.

```{r}
## first, standardize the names of the columns for binding.

ce_rename_2014_2009_2004 <- ce_bind_2014_2009_2004 %>% 
  select(E03 = clave_entidad, E04 = clave_municipio, CODIGO = clave_actividad_economica, ano_censal = anio_censal, UE, H000A, J300A, J400A, M000A, A800A)

ce_rename_2019 <- ce_bind_2019 %>% 
  mutate(ano_censal = "2019") %>% 
  select(E03 = ENTIDAD, E04 = MUNICIPIO, CODIGO, ano_censal, ID_ESTRATO, UE, H000A, J300A, J400A, M000A, A800A)  

ce_rename_2024 <- ce_bind_2024 %>% 
  mutate(ano_censal = "2024") %>% 
  select(E03, E04, CODIGO, ano_censal, ID_ESTRATO, UE, H000A, J300A, J400A, M000A, A800A)
```


```{r}
# next, bind them together, fixing any differing types.

str(ce_rename_2014_2009_2004)
str(ce_rename_2019)
str(ce_rename_2024)

ce_all_years <- bind_rows(ce_rename_2014_2009_2004, ce_rename_2019, ce_rename_2024)

# write_csv(ce_all_years, here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/ce_int","ce_all_years.csv"))

# pesca_lapaz <- ce_all_years %>% 
#   filter(CODIGO %in% "1141") %>% 
#   filter(E03 %in% "3") %>% 
#   filter(E04 %in% "3")
```

## Identify marine activities and Ramas and Subramas (sectors) we can obtain data from


```{r}
all_sectors <- ce_bind_2014_2009_2004 %>% 
  select(clave_actividad_economica, actividad_economica) %>% 
  group_by(clave_actividad_economica, actividad_economica) %>% 
  distinct()
# write_csv(all_sectors, here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/ce_int","all_sectors.csv"))
```

```{r}
library(tibble)

ce_actividades_marino <- tribble(
  ~codigo,   ~nivel,    ~sector_marino_propuesto,                                           ~notas_breve,
  # VIAJES Y TOURS
  "56151",   "Subrama", "Coastal and marine tourism / Ecotourism",
             "Agencias de viajes; usar 56151 como subrama general, las clases 561510 están en el catálogo.",
  "56152",   "Subrama", "Coastal and marine tourism; Sports fishing tours / recreational fishing",
             "Organización de excursiones y paquetes turísticos para agencias de viajes",
  "56159",   "Subrama", "Coastal and marine tourism",
             "Otros servicios de reservaciones",

  # MARINAS Y RECREACIÓN COSTERA
  # "7139",    "Rama",   "Coastal and marine tourism?",
  #            "Otros servicios recreativos",
  "7121",    "Subrama","Ecotourism",
             "Museos, sitios históricos, zoológicos y similares; usar cuando están en zonas costeras/protegidas.",

  # ALOJAMIENTO TURÍSTICO
  "72111",   "Subrama","Coastal and marine tourism / Ecotourism",
             "Hoteles y moteles, excepto hoteles con casino",
  "72119",   "Subrama","Coastal and marine tourism / Ecotourism",
             "Cabañas, villas y similares",
  "72121",   "Subrama","Ecotourism / Coastal and marine tourism",
             "Campamentos y albergues recreativos",
  "72131",   "Subrama","Coastal and marine tourism / Ecotourism",
             "Pensiones y casas de huéspedes, y departamentos y casas amueblados con servicios de hotelería",

  # SERVICIOS DE ALIMENTOS
  "72251",   "Subrama","Coastal and marine tourism",
             "Servicios de preparación de alimentos y bebidas alcohólicas y no alcohólicas; might only use this one",
  # "722511",  "Clase",  "Coastal and marine tourism / Seafood processing and trade",
  #            "Restaurantes con servicio de preparación de alimentos a la carta o de comida corrida",
  # "722512",  "Clase",  "Seafood processing and trade; Coastal and marine tourism",
  #            "Restaurantes con servicio de preparación de pescados y mariscos",
  # "722513",  "Clase",  "Coastal and marine tourism",
  #            "Restaurantes con servicio de preparación de antojitos",
  # "722514",  "Clase",  "Coastal and marine tourism",
  #            "Restaurantes con servicio de preparación de tacos y tortas",
  # "722515",  "Clase",  "Coastal and marine tourism",
  #            "Cafeterías, fuentes de sodas, neverías, refresquerías y similares",
  # "722516",  "Clase",  "Coastal and marine tourism",
  #            "Restaurantes de autoservicio",
  # "722517",  "Clase",  "Coastal and marine tourism",
  #            "Restaurantes con servicio de preparación de pizzas, hamburguesas, hot dogs y pollos rostizados para llevar",
  # "722518",  "Clase",  "Coastal and marine tourism",
  #            "Restaurantes que preparan otro tipo de alimentos para llevar",
  # "722519",  "Clase",  "Coastal and marine tourism",
  #            "Servicios de preparación de otros alimentos para consumo inmediato",

  # MANEJO DE RESIDUOS (ENERGY AND INFRASTRUCTURE)
  "5621",    "Subsector/Rama agrupada","Energy and infrastructure",
             "Manejo de desechos y servicios de remediación; incluye 56211, 562111, 562112.",

  # AQUACULTURE
  "1125",    "Subsector", "Aquaculture",
             "Acuicultura; agregación para 11251 y clases.",
  # "11251",   "Subrama",   "Aquaculture",
  #            "Acuicultura; en tu catálogo se desagrega en 112511, 112512, etc.",
  # "112511",  "Clase",     "Aquaculture",
  #            "Camaronicultura.",
  # "112512",  "Clase",     "Aquaculture",
  #            "Piscicultura y otra acuicultura, excepto camaronicultura.",

  # FISHING
  "1141",    "Rama",   "Fishing",
             "Pesca; incluye subrama 11411 y clases específicas.",
  # "11411",   "Subrama","Fishing",
  #            "Pesca; subrama general.",
  # "114111",  "Clase", "Fishing",
  #            "Pesca de camarón.",
  # "114112",  "Clase", "Fishing",
  #            "Pesca de túnidos.",
  # "114113",  "Clase", "Fishing",
  #            "Pesca de sardina y anchoveta.",
  # "114119",  "Clase", "Fishing",
  #            "Pesca y captura de peces, crustáceos, moluscos y otras especies.",

  # SEAFOOD PROCESSING AND TRADE
  "31171",   "Subrama","Seafood processing and trade",
             "Preparación y envasado de pescados y mariscos",
  # "311710",  "Clase", "Seafood processing and trade",
  #            "Preparación y envasado de pescados y mariscos.", ## double counting
  "431123",  "Clase", "Seafood processing and trade",
             "Comercio al por mayor de pescados y mariscos; Wholesale trade of fish and seafood",
  "461123",  "Clase", "Seafood processing and trade",
             "Comercio al por menor de pescados y mariscos; Retail sale of fish and seafood",

  # SHIPBUILDING AND MARINE EQUIPMENT
  "33661",   "Subrama","Shipbuilding and marine equipment",
             "Fabricación de embarcaciones",
  # "336610",  "Clase", "Shipbuilding and marine equipment",
  #            "Fabricación de embarcaciones.", # subset but same thing

  # MARITIME TRANSPORT AND PORT SERVICES
  "4832",    "Subsector/Rama","Maritime transport and port services",
             "Transporte por aguas interiores; incluye 48321, 483210.",
  # "48321",   "Subrama", "Maritime transport and port services",
  #            "Transporte por aguas interiores.",
  # "483210",  "Clase",   "Maritime transport and port services",
  #            "Transporte por aguas interiores.",
  "4872",    "Subrama", "Maritime transport and port services; Coastal and marine tourism",
             "Transporte turístico por agua, includes 48721 and 487210",
  # "487210",  "Clase",   "Maritime transport and port services; Coastal and marine tourism",
  #            "Transporte turístico por agua.",
  "4883",    "Subrama", "Maritime transport and port services",
             "Servicios relacionados con el transporte por agua: incluye 48839, 488390, 48831 Administración de puertos y muelles, 488310, 48833 Servicios para la navegación por agua, 488330.",
  # "48839",   "Subrama", "Maritime transport and port services",
  #            "Otros servicios relacionados con el transporte por agua.",
  # "488390",  "Clase",   "Maritime transport and port services",
  #            "Otros servicios relacionados con el transporte por agua.",

  # MARINE RESEARCH AND CONSULTING
  "54162",   "Clase",   "Marine research and consulting services",
             "Servicios de consultoría en medio ambiente",
  # "541620",  "Clase",   "Marine research and consulting services",
  #            "Servicios de consultoría en medio ambiente",
  "54171",   "Subrama", "Marine research and consulting services",
             "Servicios de investigación científica y desarrollo en ciencias naturales y exactas, ingeniería, y ciencias de la vida"
  # "541711",  "Clase",   "Marine research and consulting services",
  #            "I+D en ciencias naturales y exactas, ingeniería y ciencias de la vida."
)

ce_actividades_marino_valid <- ce_actividades_marino %>%
  semi_join(all_sectors,
            by = c("codigo" = "clave_actividad_economica")) ## just checking that they are the same codes and I did not make a mistake!!

setdiff(unique(ce_actividades_marino_valid$codigo), unique(all_sectors$clave_actividad_economica)) # character(0) yay!!!

# write_csv(ce_actividades_marino_valid, here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/ce_int","all_marine_related_sectors.csv"))

```


## Limit to marine activities in coastal municipios and assign OHI rgn id

```{r}
# first, make sure the column names match.
municipio_codigo_rename <- municipio_codigo_df %>% 
  mutate(clave_estado = as.numeric(clave_estado),
         clave_municipio = as.numeric(clave_municipio)) %>% 
  rename(E03 = clave_estado, 
         E04 = clave_municipio) %>% 
  select(-tasa_regional)

# as sf if needed
# municipio_codigo_rename_sf <- municipio_codigo_sf %>%
#   mutate(clave_estado = as.numeric(clave_estado),
#          clave_municipio = as.numeric(clave_municipio)) %>% 
#   rename(E03 = clave_estado,
#          E04 = clave_municipio) %>%
#   select(-tasa_regional)
```


```{r}
# now filter the ce to only those municipalities. 
ce_all_years_goc <- ce_all_years %>% 
  inner_join(municipio_codigo_rename, by = c("E03", "E04")) %>% 
  janitor::clean_names()


## check with la paz numbers
# pesca_lapaz <- ce_all_years_goc_fix %>% 
#   filter(codigo %in% "1141") %>% 
#   filter(e03 %in% "3") %>% 
#   filter(e04 %in% "3")

# ce_all_years_goc_sf <- ce_all_years %>% 
#   inner_join(municipio_codigo_rename_sf, by = c("E03", "E04")) %>% 
#   janitor::clean_names() %>% 
#   st_as_sf()

# mapview(ce_all_years_goc_sf)

unique(ce_all_years_goc$municipio) 
#  [1] "ENSENADA"                      "MEXICALI"                      "TECATE"                       
#  [4] "TIJUANA"                       "PLAYAS DE ROSARITO"            "COMONDU"                      
#  [7] "MULEGE"                        "LA PAZ"                        "LOS CABOS"                    
# [10] "LORETO"                        "CABO CORRIENTES"               "PUERTO VALLARTA"              
# [13] "ACAPONETA"                     "AHUACATLAN"                    "AMATLAN DE CAÑAS"             
# [16] "COMPOSTELA"                    "IXTLAN DEL RIO"                "JALA"                         
# [19] "XALISCO"                       "ROSAMORADA"                    "RUIZ"                         
# [22] "SAN BLAS"                      "SAN PEDRO LAGUNILLAS"          "SANTA MARIA DEL ORO"          
# [25] "SANTIAGO IXCUINTLA"            "TECUALA"                       "TEPIC"                        
# [28] "TUXPAN"                        "BAHIA DE BANDERAS"             "AHOME"                        
# [31] "ANGOSTURA"                     "BADIRAGUATO"                   "CONCORDIA"                    
# [34] "COSALA"                        "CULIACAN"                      "CHOIX"                        
# [37] "ELOTA"                         "ESCUINAPA"                     "EL FUERTE"                    
# [40] "GUASAVE"                       "MAZATLAN"                      "MOCORITO"                     
# [43] "ROSARIO"                       "SALVADOR ALVARADO"             "SAN IGNACIO"                  
# [46] "SINALOA"                       "NAVOLATO"                      "AGUA PRIETA"                  
# [49] "ALAMOS"                        "ALTAR"                         "ATIL"                         
# [52] "BACUM"                         "BAVIACORA"                     "BENJAMIN HILL"                
# [55] "CABORCA"                       "CAJEME"                        "EMPALME"                      
# [58] "ETCHOJOA"                      "GUAYMAS"                       "HERMOSILLO"                   
# [61] "HUATABAMPO"                    "IMURIS"                        "MAGDALENA"                    
# [64] "MAZATAN"                       "NAVOJOA"                       "NOGALES"                      
# [67] "OPODEPE"                       "PITIQUITO"                     "PUERTO PEÑASCO"               
# [70] "SAN JAVIER"                    "SAN LUIS RIO COLORADO"         "SAN MIGUEL DE HORCASITAS"     
# [73] "SANTA ANA"                     "TUBUTAMA"                      "GENERAL PLUTARCO ELIAS CALLES"
# [76] "BENITO JUAREZ"                 "SAN IGNACIO RIO MUERTO"        "SAN QUINTÍN"                  
# [79] "SAN FELIPE"  

```


### Marine activities

Now limit to the actividades we care about:
```{r}
# limit to marine-related economic activities (df)
ce_all_years_goc_marine <- inner_join(ce_actividades_marino_valid, ce_all_years_goc, by = "codigo")
write_csv(ce_all_years_goc_marine, here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/ce_int","ce_all_years_goc_marine.csv"))
  
# limit to marine-related economic activities (sf)
# ce_all_years_goc_marine_sf <- inner_join(ce_actividades_marino_valid, ce_all_years_goc_sf, by = "codigo") %>% 
#   st_as_sf()
# st_write(ce_all_years_goc_marine_sf, here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/ce_int","ce_all_years_goc_marine_sf.shp"))
# mapview(ce_all_years_goc_marine_sf)
```


### Assign OHI rgn_id:

```{r}
# inland_50 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer","inland_50km_goc.shp"))
# mapview(inland_50, zcol = "rgn_id")
# mapview(municipio_codigo_sf, zcol = "estado")

## bring in the municipality to rgn_id code CSV I made...
municipality_to_rgnid <- read_csv(here::here("/home/shares/ohi/OHI_GOC/spatial/municipality_to_rgnid","municipality_to_rgnid.csv")) %>% 
  select(-c("notes","municipio_raw")) %>% 
  janitor::clean_names() %>% 
  mutate(municipio = str_trim(municipio, side = "both"))

setdiff(unique(municipality_to_rgnid$municipio), unique(ce_all_years_goc_marine$municipio)) #character(0)
```

```{r}
ce_all_years_goc_marine_rgnid  <- left_join(ce_all_years_goc_marine, municipality_to_rgnid, by = c("e03","e04", "estado", "municipio")) ## Joining with `by = join_by(e03, e04, estado, municipio)`
```

Check it out:
```{r}
# ce_goc_marine_activities_rgn_2024 <- ce_all_years_goc_marine_rgnid_sf %>%
#   filter(ano_censal %in% 2024) %>%
#   filter(codigo %in% "56151")

ce_goc_marine_activities_rgn_2024 <- ce_all_years_goc_marine_rgnid %>%
  filter(ano_censal %in% 2024) %>%
  filter(codigo %in% "1141")
unique(ce_goc_marine_activities_rgn_2024$municipio)


ce_all_years_goc_marine_sf_2024 <- ce_all_years_goc_marine %>% 
  filter(ano_censal %in% 2024) %>%
  filter(codigo %in% "1141")
unique(ce_all_years_goc_marine_sf_2024$municipio)

setdiff(unique(ce_goc_marine_activities_rgn_2024$municipio), unique(ce_all_years_goc_marine_sf_2024$municipio))
setdiff(unique(ce_all_years_goc_marine_sf_2024$h000a), unique(ce_goc_marine_activities_rgn_2024$h000a))

ce_goc_marine_activities_rgn_pesca_new <- ce_all_years_goc_marine_rgnid %>%
  filter(codigo %in% "1141")
```

Remove any municipalities that are *not* coastal municipios (aka, that have NA for rgn_id):
```{r}
isna_rgnid <- ce_all_years_goc_marine_rgnid %>% 
  filter(is.na(rgn_id))
unique(isna_rgnid$municipio)
#  [1] "TECATE"                   "TIJUANA"                 
#  [3] "COMONDU"                  "PLAYAS DE ROSARITO"      
#  [5] "AGUA PRIETA"              "NOGALES"                 
#  [7] "ALTAR"                    "BAVIACORA"               
#  [9] "MAGDALENA"                "BENJAMIN HILL"           
# [11] "IMURIS"                   "MAZATAN"                 
# [13] "SAN MIGUEL DE HORCASITAS" "SANTA ANA"               
# [15] "OPODEPE"                  "ATIL"                    
# [17] "SAN JAVIER"               "TUBUTAMA"      

ce_actividades_rgnid <- ce_all_years_goc_marine_rgnid %>% 
  filter(!is.na(rgn_id))

unique(ce_all_years_goc_marine_rgnid$municipio) # 79
unique(ce_actividades_rgnid$municipio) # 61 municipios, makes sense: 79-18 = 61

## save the filtered dataframe of all years marine activities with region ids
# write_csv(ce_actividades_rgnid, here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/ce_int","ce_all_years_goc_marine_w_rgnid.csv"))
```



## Final cleaning (ID ESTRATO)

Exploration of ID_ESTRATO:
```{r}
## now here, I am looking through to understand the id_estrato column better.

## when I look here, I see that Baja California does not have data from every municipio code.  It is only from 12567, not 3 and 4. 
pesca_ce_all_years_goc_marine <- ce_all_years_goc_marine %>% 
  filter(codigo %in% "1141") %>% 
  filter(ano_censal %in% "2024")

## when I go back to the raw data, I find that was the case originally...
pesca_ce_bind_2024 <- ce_bind_2024 %>%  filter(CODIGO %in% "1141")
### and also, that when the ID_ESTRATO is NA, then all the UEs for Baja California for this economic activity add up to the state total.  So to me that means that there are no registered economic units in municipalities 02003 (TECATE) or 02004 (TIJUANA).  WHICH, when we look at a map, those are the inland areas at the top of BC!!  Tijuana has a small area of beach, but it makes sense to me that there are no registered fishers there because it is so close to the border.  And likely why 02002 and 02005 are low numbers, considering that in 2024, San Felipe is split from Mexicali.  And San Felipe, 02007 has a high number of fishers.

## NOTE that 02001, which previously encompassed a large portion of the northern pacific-side *and* gulf-side of BC, is now only pacific-side!!! <https://en.wikipedia.org/wiki/Municipalities_of_Baja_California>.  

## so for 2024, we should only consider 02002, 02007, and 02006 to be GoC.
```

```{r}
ce_byrgn <- ce_actividades_rgnid %>% 
  filter(is.na(id_estrato)) %>% 
  group_by(rgn_id, codigo, municipio) %>% 
  summarize(years_available = n()) # 5 is the most: 2004, 2009, 2014, 2019, 2024

ce_all_years_goc_marine_rgnid_check <- ce_all_years_goc_marine_rgnid %>% 
  filter(codigo %in% "1125") %>% 
  filter(municipio %in% "MEXICALI") ## 99 is the same as NA here....

ce_all_years_goc_marine_rgnid_check <- ce_all_years_goc_marine_rgnid %>% 
  filter(codigo %in% "1141") %>% 
  filter(municipio %in% "LOS CABOS") ## 99 is the same as NA here....

colnames(ce_actividades_rgnid)
#  [1] "codigo"                  "nivel"                  
#  [3] "sector_marino_propuesto" "notas_breve"            
#  [5] "e03"                     "e04"                    
#  [7] "ano_censal"              "ue"                     
#  [9] "h000a"                   "j300a"                  
# [11] "j400a"                   "m000a"                  
# [13] "a800a"                   "id_estrato"             
# [15] "estado"                  "municipio"              
# [17] "rgn_id"                  "directly_coastal"     

unique(ce_actividades_rgnid$id_estrato)
# [1] NA 99  1  2  3  4

ce_all_years_goc_marine_rgnid_check <- ce_all_years_goc_marine_rgnid %>% 
  filter(id_estrato %in% "99")
```

Okay.  

| id_estrato | Meaning in practice                                                                            |
| ---------- | ---------------------------------------------------------------------------------------------- |
| 1–4        | Establishment size strata                                                                      |
| NA         | “Suma de Estratos” (explicit total)                                                        |
| 99         | “Agrupados por confidencialidad” (sometimes behaves like total, sometimes partial aggregation) |

So, for each group, we need to determine whether a reliable total exists....

If yes, we can just use that row.
If not, we need to reconstruct the total from strata 1-4.

```{r}
check_structure <- ce_actividades_rgnid %>%
  group_by(ano_censal, municipio, codigo) %>%
  summarize(
    n_rows = n(),
    has_total_na = any(is.na(id_estrato)),
    has_99 = any(id_estrato == 99),
    has_strata = any(id_estrato %in% 1:4),
    .groups = "drop"
  )
```

*Okay, so we can use this decision hierarchy:*

If NA total exists, use that row.
Else if only one row and it's 99, treat as total.
Else if strata 1-4 exists, sum them.
Else if only 99 rows exist, treat 99 as total.
Else... flag as having an issue.

So we only have one row per group!

*NOTE: Why 99 Sometimes Equals NA*

In INEGI Economic Census, if there is no confidentiality issue, it seems that they publish strata and the total. But, if there is a confidentiality risk, they suppress strata and replace with:

* Either 99 only
* Or 99 & partial aggregation, it seems like!

So 99 is not the “total”, but rather the “confidential grouping” they use. In our case, it behaves like a total only when strata are absent.

*Important: Do NOT Mix 99 With 1-4....* if 1-4 exist, 99 is almost always a collapsed group of suppressed strata, and is therefore NOT additive with the others.

SUMMARY: Use the total if provided, otherwise add up the strata, and never mix confidentiality strata with true strata!!

```{r}
cols_to_sum <- c("ue", "h000a", "j300a", "j400a", "m000a", "a800a")

ce_totals_clean <- ce_actividades_rgnid %>%
  group_by(ano_censal, municipio, codigo) %>% # group by census year, municipality, and activity code.this ensures we resolve strata into a single total per group
  
  group_modify(~{

    # .x contains the current group variation.  So for each group (one year x municipio x codigo)...
    df <- .x 

    # -------------------------------------------------------------
    # case 1: explicit total row exists (id_estrato is NA)
    # this is the official "suma de estratos" that is the total. if we have this, it is the safest and most complete total, so we return only that row
    # -------------------------------------------------------------
    if(any(is.na(df$id_estrato))) {
      return(df %>% filter(is.na(id_estrato)))
    }

    # -------------------------------------------------------------
    # case 2: only one row exists and it is coded as 99
    # based on the tc_estrato_ce2024 and all methodology documents for the other years, this means strata were suppressed for confidentiality. And in this case, since there are no other strata rows, we can treat 99 as the total.
    # -------------------------------------------------------------
    if(nrow(df) == 1 & df$id_estrato == 99) {
      return(df)
    }

    # -------------------------------------------------------------
    # case 3: establishment size strata (1 to 4) are present
    # this means no explicit total was provided, so we have to estimate the total by summing across strata 1 - 4
    # NOTE we do NOT include 99 here to avoid double counting!!!
    # -------------------------------------------------------------
    if(any(df$id_estrato %in% 1:4)) {

      # sum numeric variables across strata rows
      summed <- df %>%
        filter(id_estrato %in% 1:4) %>%
        summarize(across(all_of(cols_to_sum), ~sum(.x, na.rm = TRUE))) 

      # keep identifying metadata from the first row (these are constant within the group)
      metadata <- df %>% 
        slice(1) %>%
        select(ano_censal, municipio, codigo, estado, rgn_id, directly_coastal)

      # combine metadata with summed totals
      return(bind_cols(metadata, summed))
    }

    # -------------------------------------------------------------
    # case 4: multiple rows exist but *ALL* are coded 99
    # this indicates confidentiality grouping only, and since no strata exist, treat one 99 row as the total.  Very similar to case 2 but we have to take one row out of the many for that (year x municipio x codigo) grouping...
    # -------------------------------------------------------------
    if(all(df$id_estrato == 99)) {
      return(df %>% slice(1))
    }

    # -------------------------------------------------------------
    # else/case 5: if the grouping does not fit into any of these cases, then return the first row to ensure one row per group.... these NEED to be investigated.
    # -------------------------------------------------------------
    return(df %>% slice(1))
  }) %>%
  ungroup()


### check that each group has only one row!! should equal 1:
check_clean <- ce_totals_clean %>%
  group_by(ano_censal, municipio, codigo) %>%
  tally()
unique(check_clean$n) # 1 !! good!!

check_structure_2 <- ce_totals_clean %>%
  group_by(ano_censal, municipio, codigo) %>%
  summarize(
    n_rows = n(),
    has_total_na = any(is.na(id_estrato)),
    has_99 = any(id_estrato == 99),
    has_strata = any(id_estrato %in% 1:4),
    .groups = "drop"
  ) ## now we do not have any 99s in there.... need to come back and check on this.

missing_by_year <- ce_totals_clean %>%
  group_by(ano_censal) %>%
  summarize(
    n_obs = n(),
    pct_missing_jobs = mean(is.na(h000a)) * 100,
    pct_missing_ue = mean(is.na(ue)) * 100,
    pct_missing_revenue = mean(is.na(a800a)) * 100
  )

missing_by_region <- ce_totals_clean %>%
  group_by(rgn_id) %>%
  summarize(
    n_obs = n(),
    pct_missing_jobs = mean(is.na(h000a)) * 100
  ) %>%
  arrange(desc(pct_missing_jobs))

missing_rgn_year_act <- ce_totals_clean %>%
  group_by(rgn_id, ano_censal, codigo) %>%
  summarize(
    n_municipios = n(),
    n_missing_jobs = sum(is.na(h000a)),
    pct_missing_jobs = mean(is.na(h000a)) * 100,
    n_missing_ue = sum(is.na(ue)),
    pct_missing_ue = mean(is.na(ue)) * 100,
    .groups = "drop"
  ) %>%
  arrange(desc(pct_missing_jobs)) ### note that 2019 and 2024, since they are newer and did not have the summarized table of all previous years (as we had for 2004 - 2014), have many more suppressed observations... we will need to return to this and ensure what we are doing is correct.
```


## Save to Mazu

```{r}
## save the filtered dataframe of all years marine activities with region ids that only have one row per group.
# write_csv(ce_totals_clean, here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/ce_int","ce_cleaned_totals_per_yr_municipio_activity.csv"))
```

#### Exploration

```{r}
library(plotly)

ce_totals_clean_pesca <- ce_totals_clean %>% 
  filter(codigo %in% "1141")

## plot 
num_jobs_pesca_plotly <- plotly::plot_ly(
  ce_totals_clean_pesca, 
  x = ~ano_censal, 
  y = ~h000a, 
  color = ~municipio, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "Number of fishers",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "H000A")
  ) 
num_jobs_pesca_plotly
```

Totals if aggregated by region:

```{r}
ce_totals_all_activities_by_rgn <- ce_totals_clean %>% 
  group_by(rgn_id, ano_censal) %>% 
  summarize(total_jobs = sum(h000a, na.rm = TRUE),
            total_ue = sum(ue, na.rm = TRUE),
            total_revenue = sum(a800a, na.rm = TRUE)) %>% 
  ungroup()

# num jobs
num_jobs_all_plotly <- plotly::plot_ly(
  ce_totals_all_activities_by_rgn, 
  x = ~ano_censal, 
  y = ~total_jobs, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "Number of jobs",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "H000A")
  ) 
num_jobs_all_plotly

# ue total
ue_all_plotly <- plotly::plot_ly(
  ce_totals_all_activities_by_rgn, 
  x = ~ano_censal, 
  y = ~total_ue, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "Number of economic units",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = " ")
  ) 
ue_all_plotly
```


