---
title: 'OHI GoC 2025: Livelihoods & Economies'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# LE: Censos Economicos (CE) 

For number of jobs, quality of jobs, and revenue of certain sectors.  


```{r}
library(tidyverse)
library(mapview)
library(sf)
```

## Isolate municipos

Using this pdf of the codes for each municipality: <https://www.gob.mx/cms/uploads/attachment/file/179869/Municipios_por_Regiones__atendidos_.pdf>

NOTE THAT SAN FELIPE was recently added in 2022 as a municipality.  Updated municipalities will need to be used. 

```{r}
library(DT)

raw_text <- paste(
  "02 BAJA_CALIFORNIA 001 ENSENADA 2 2.59",
  "02 BAJA_CALIFORNIA 002 MEXICALI 1 2.02",
  "02 BAJA_CALIFORNIA 003 TECATE 6 4.28",
  "02 BAJA_CALIFORNIA 004 TIJUANA 1 2.02",
  "02 BAJA_CALIFORNIA 005 PLAYAS_DE_ROSARITO 2 2.59",
  "02 BAJA_CALIFORNIA 006 SAN QUINTÍN 2 NA", ## added manually... because of 2022 changes: https://en.wikipedia.org/wiki/Municipalities_of_Baja_California
  "02 BAJA_CALIFORNIA 007 SAN FELIPE 2 NA", ## added manually... because of 2022 changes: https://en.wikipedia.org/wiki/Municipalities_of_Baja_California
  "03 BAJA_CALIFORNIA_SUR 001 COMONDU 3 3.03",
  "03 BAJA_CALIFORNIA_SUR 002 MULEGE 3 3.03",
  "03 BAJA_CALIFORNIA_SUR 003 LA_PAZ 3 3.03",
  "03 BAJA_CALIFORNIA_SUR 008 LOS_CABOS 3 3.03",
  "03 BAJA_CALIFORNIA_SUR 009 LORETO 4 3.53",
  "14 JALISCO 067 PUERTO_VALLARTA 6 4.28",
  "14 JALISCO 020 CABO_CORRIENTES 1 2.02",
  "18 NAYARIT 001 ACAPONETA 2 2.59",
  "18 NAYARIT 002 AHUACATLAN 1 2.02",
  "18 NAYARIT 003 AMATLAN_DE_CAÑAS 1 2.02",
  "18 NAYARIT 004 COMPOSTELA 2 2.59",
  "18 NAYARIT 006 IXTLAN_DEL_RIO 1 2.02",
  "18 NAYARIT 007 JALA 6 4.28",
  "18 NAYARIT 008 XALISCO 2 2.59",
  "18 NAYARIT 010 ROSAMORADA 2 2.59",
  "18 NAYARIT 011 RUIZ 2 2.59",
  "18 NAYARIT 012 SAN_BLAS 3 3.03",
  "18 NAYARIT 013 SAN_PEDRO_LAGUNILLAS 1 2.02",
  "18 NAYARIT 014 SANTA_MARIA_DEL_ORO 3 3.03",
  "18 NAYARIT 015 SANTIAGO_IXCUINTLA 2 2.59",
  "18 NAYARIT 016 TECUALA 1 2.02",
  "18 NAYARIT 017 TEPIC 6 4.28",
  "18 NAYARIT 018 TUXPAN 3 3.03",
  "18 NAYARIT 020 BAHIA_DE_BANDERAS 3 3.03",
  "25 SINALOA 001 AHOME 4 3.53",
  "25 SINALOA 002 ANGOSTURA 2 2.59",
  "25 SINALOA 003 BADIRAGUATO 6 4.28",
  "25 SINALOA 004 CONCORDIA 2 2.59",
  "25 SINALOA 005 COSALA 3 3.03",
  "25 SINALOA 006 CULIACAN 5 3.90",
  "25 SINALOA 007 CHOIX 5 3.90",
  "25 SINALOA 008 ELOTA 5 NA", # added in when found missing 
  "25 SINALOA 009 ESCUINAPA 2 2.59",
  "25 SINALOA 010 EL_FUERTE 3 3.03",
  "25 SINALOA 011 GUASAVE 6 4.28",
  "25 SINALOA 012 MAZATLAN 5 3.90",
  "25 SINALOA 013 MOCORITO 2 2.59",
  "25 SINALOA 014 ROSARIO 2 2.59",
  "25 SINALOA 015 SALVADOR_ALVARADO 2 2.59",
  "25 SINALOA 016 SAN_IGNACIO 2 NA", # added in when found missing 
  "25 SINALOA 017 SINALOA 3 3.03",
  "25 SINALOA 018 NAVOLATO 2 2.59",
  "26 SONORA 002 AGUA_PRIETA 6 4.28",
  "26 SONORA 003 ALAMOS 3 3.03",
  "26 SONORA 004 ALTAR 2 2.59",
  "26 SONORA 007 ATIL 6 4.28",
  "26 SONORA 012 BACUM 5 3.90",
  "26 SONORA 014 BAVIACORA 5 3.90",
  "26 SONORA 016 BENJAMIN_HILL 6 4.28",
  "26 SONORA 017 CABORCA 2 2.59",
  "26 SONORA 018 CAJEME 6 4.28",
  "26 SONORA 025 EMPALME 6 4.28",
  "26 SONORA 026 ETCHOJOA 6 4.28",
  "26 SONORA 029 GUAYMAS 5 3.90",
  "26 SONORA 030 HERMOSILLO 6 4.28",
  "26 SONORA 033 HUATABAMPO 3 3.03",
  "26 SONORA 035 IMURIS 6 4.28",
  "26 SONORA 036 MAGDALENA 5 3.90",
  "26 SONORA 037 MAZATAN 2 2.59",
  "26 SONORA 042 NAVOJOA 6 4.28",
  "26 SONORA 043 NOGALES 3 3.03",
  "26 SONORA 045 OPODEPE 4 3.53",
  "26 SONORA 047 PITIQUITO 3 3.03",
  "26 SONORA 048 PUERTO_PEÑASCO 3 3.03",
  "26 SONORA 054 SAN_JAVIER 3 3.03",
  "26 SONORA 055 SAN_LUIS_RIO_COLORADO 4 3.53",
  "26 SONORA 056 SAN_MIGUEL_DE_HORCASITAS 4 3.53",
  "26 SONORA 058 SANTA_ANA 5 3.90",
  "26 SONORA 065 TUBUTAMA 6 4.28",
  "26 SONORA 070 GENERAL_PLUTARCO_ELIAS_CALLES 1 2.02",
  "26 SONORA 071 BENITO_JUAREZ 3 3.03",
  "26 SONORA 072 SAN_IGNACIO_RIO_MUERTO 5 3.90",
  sep = "\n"
)

library(purrr)

lines <- str_split(raw_text, "\n")[[1]]

municipio_codigo_raw_df <- map_dfr(lines, function(x) {
  parts <- str_split(x, " +")[[1]]   # split on spaces
  tibble(
    clave_estado   = parts[1],
    estado         = parts[2] |> str_replace_all("_", " "),
    clave_municipio = parts[3],
    region         = as.integer(parts[length(parts) - 1]),
    tasa_regional  = as.numeric(parts[length(parts)]),
    municipio      = paste(parts[4:(length(parts) - 2)], collapse = " ") |> 
                      str_replace_all("_", " ")
  )
}) 

municipio_codigo_df <- municipio_codigo_raw_df %>% 
  select(clave_estado, estado, clave_municipio, municipio, tasa_regional)

write_csv(municipio_codigo_df, here::here("/home/shares/ohi/OHI_GOC/spatial/land_admin_boundaries", "municipio_codigos_secreteria_de_economia.csv"))
```

Combine with shapefile for each municipality.

```{r}
municipios <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/land_admin_boundaries/mex_admin_boundaries.shp","mex_admin2.shp")) %>% 
  mutate(nom_mun = str_to_upper(adm2_ref_n),
         nom_ent = str_to_upper(adm1_name)) %>% 
  select(nom_ent, nom_mun) %>% 
  filter(nom_ent %in% c("BAJA CALIFORNIA SUR", "BAJA CALIFORNIA", "SINALOA", "SONORA", "NAYARIT", "JALISCO")) %>% 
  mutate(
    nom_mun = str_replace(nom_mun, "PUERTO PENASCO", "PUERTO PEÑASCO"),
    nom_mun = str_replace(nom_mun, "AMATLAN DE CANAS", "AMATLAN DE CAÑAS")
  )

mapview(municipios)

# see if they match
setdiff(municipio_codigo_sf$municipio, municipios$nom_mun) # character(0)
# setdiff(municipios$nom_mun, municipio_codigo_sf$municipio)

municipio_codigo_sf <- inner_join(municipio_codigo_df, municipios, by = c("municipio" = "nom_mun", "estado" = "nom_ent")) %>% 
  st_as_sf()

mapview(municipio_codigo_sf)

# st_write(municipio_codigo_sf, here::here("/home/shares/ohi/OHI_GOC/spatial/land_admin_boundaries", "mex_admin2_joined_w_municipio_codigos.shp"), append = FALSE)
```


## Read in data

### 2024
```{r}
# baja california
bc_2024_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2024/conjunto_de_datos_ce_bc_2024_csv/conjunto_de_datos","tr_ce_bc_2024.csv")) 
unique(bc_2024_raw$E04) # [1] NA    "001" "002" "003" "004" "005" "006" "007", now has San Felipe!! 

bc_diccionario_2024_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2024/conjunto_de_datos_ce_bc_2024_csv/diccionario_de_datos","diccionario_de_datos_ce2024.csv"))

bc_actividad_2024_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2024/conjunto_de_datos_ce_bc_2024_csv/catalogos","tc_codigo_actividad.csv"))

# baja california sur
bcs_2024_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2024/conjunto_de_datos_ce_bcs_2024_csv/conjunto_de_datos","tr_ce_bcs_2024.csv"))

# jalisco
jal_2024_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2024/conjunto_de_datos_ce_jal_2024_csv/conjunto_de_datos","tr_ce_jal_2024.csv"))

# nayarit
nay_2024_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2024/conjunto_de_datos_ce_nay_2024_csv/conjunto_de_datos","tr_ce_nay_2024.csv"))

# sinaloa
sin_2024_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2024/conjunto_de_datos_ce_sin_2024_csv/conjunto_de_datos","tr_ce_sin_2024.csv"))

# sonora
son_2024_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2024/conjunto_de_datos_ce_son_2024_csv/conjunto_de_datos","tr_ce_son_2024.csv"))
```

Merge them into one df:
```{r}
fix_types <- \(df) df |>
  mutate(
    E03 = as.numeric(E03),
    E04 = as.numeric(E04),
    CODIGO = as.character(CODIGO)
  )

ce_bind_2024 <- bind_rows(
  fix_types(bc_2024_raw),
  fix_types(bcs_2024_raw),
  fix_types(jal_2024_raw),
  fix_types(nay_2024_raw),
  fix_types(sin_2024_raw),
  fix_types(son_2024_raw)
)
```



### 2019
```{r}
# baja california
bc_2019_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2019/ce2019_bc_csv/conjunto_de_datos","ce2019_bc.csv"))
unique(bc_2019_raw$MUNICIPIO) # [1] NA    "001" "002" "003" "004" "005"

bc_actividad_2019_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2019/ce2019_bc_csv/catalogos","tc_codigo_actividad.csv"))

# baja california sur
bcs_2019_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2019/ce2019_bcs_csv/conjunto_de_datos","ce2019_bcs.csv"))

# jalisco
jal_2019_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2019/ce2019_jal_csv/conjunto_de_datos","ce2019_jal.csv"))

# nayarit
nay_2019_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2019/ce2019_nay_csv/conjunto_de_datos","ce2019_nay.csv"))

# sinaloa
sin_2019_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2019/ce2019_sin_csv/conjunto_de_datos","ce2019_sin.csv"))

# sonora
son_2019_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2019/ce2019_son_csv/conjunto_de_datos","ce2019_son.csv"))
```

Merge them:
```{r}
spec(bc_2019_raw)
spec(jal_2019_raw)

fix_types <- \(df) df |>
  mutate(
    ENTIDAD = as.numeric(ENTIDAD),
    MUNICIPIO = as.numeric(MUNICIPIO),
    CODIGO = as.character(CODIGO)
  )

ce_bind_2019 <- bind_rows(
  fix_types(bc_2019_raw),
  fix_types(bcs_2019_raw),
  fix_types(jal_2019_raw),
  fix_types(nay_2019_raw),
  fix_types(sin_2019_raw),
  fix_types(son_2019_raw)
)
```


### 2014, 2009, 2004
```{r}
# baja california
bc_2014_2009_2004_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2014_2009_2004/ce_02_bc_csv/conjunto_de_datos","ce_inegi_02.csv"))
unique(bc_2014_2009_2004_raw$clave_municipio) # [1] NA    "001" "002" "003" "004" "005"

# baja california sur
bcs_2014_2009_2004_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2014_2009_2004/ce_03_bcs_csv/conjunto_de_datos","ce_inegi_03.csv"))

# jalisco
jal_2014_2009_2004_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2014_2009_2004/ce_14_jal_csv/conjunto_de_datos","ce_inegi_14.csv"))

# nayarit
nay_2014_2009_2004_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2014_2009_2004/ce_18_nay_csv/conjunto_de_datos","ce_inegi_18.csv"))

# sinaloa
sin_2014_2009_2004_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2014_2009_2004/ce_25_sin_csv/conjunto_de_datos","ce_inegi_25.csv"))

# sonora
son_2014_2009_2004_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2014_2009_2004/ce_26_son_csv/conjunto_de_datos","ce_inegi_26.csv"))
```

Merge them:
```{r}
spec(bc_2014_2009_2004_raw)
spec(bcs_2014_2009_2004_raw)
spec(nay_2014_2009_2004_raw)
spec(jal_2014_2009_2004_raw)
spec(sin_2014_2009_2004_raw)
spec(son_2014_2009_2004_raw)


fix_types <- \(df) df %>% 
  mutate(
    clave_entidad = as.numeric(clave_entidad),  # bc has char, others have numeric
    clave_municipio = as.numeric(clave_municipio),
    across(c(J400A, J500A, J600A, K010A, K020A, K030A, K060A, K070A, 
             K096A, K311A, K620A, K810A, K910A, M010A, M030A, 
             P000A, P000B, P000C, P030A, P030B, P030C, P100A, P100B,
             Q020A, Q030A), 
           ~(as.numeric(.x))),
    anio_censal = as.character(anio_censal)
  )

ce_bind_2014_2009_2004 <- bind_rows(
  fix_types(bc_2014_2009_2004_raw),
  fix_types(bcs_2014_2009_2004_raw),
  fix_types(jal_2014_2009_2004_raw),
  fix_types(nay_2014_2009_2004_raw),
  fix_types(sin_2014_2009_2004_raw),
  fix_types(son_2014_2009_2004_raw)
)

# pesca_lapaz <- ce_bind_2014_2009_2004 %>% 
  # filter(clave_actividad_economica %in% "1141") %>% 
  # filter(clave_entidad %in% "3") %>% 
  # filter(clave_municipio %in% "003")
```


## Merge

First do some explorations of the differences between the years.

```{r}
setdiff(colnames(ce_bind_2014_2009_2004), colnames(ce_bind_2019))
# [1] "clave_entidad"             "entidad_federativa"        "clave_municipio"           "municipio"                
# [5] "clave_actividad_economica" "actividad_economica"       "anio_censal"               "K040A"                    
# [9] "K041A"
setdiff(colnames(ce_bind_2014_2009_2004), colnames(ce_bind_2024))
# [1] "clave_entidad"             "entidad_federativa"        "clave_municipio"           "municipio"                
# [5] "clave_actividad_economica" "actividad_economica"       "anio_censal"               "K040A"                    
# [9] "K041A" 

setdiff(colnames(ce_bind_2019), colnames(ce_bind_2024)) # what is in 2019 that is not in 2024
#  [1] "ENTIDAD"   "MUNICIPIO" "A112A"     "A113A"     "A114A"     "A115A"     "A116A"     "A117A"     "A118A"    
# [10] "A119A"     "A141A"     "A143A"     "A144A"     "A145A"     "A146A"     "A147A"     "A148A"     "A149A"    
# [19] "A151A"     "A152A"     "A153A"     "A154A"     "A155A"     "A156A"     "A157A"     "A158A"     "A159A"    
# [28] "A171A"     "A173A"     "A174A"     "A175A"     "A176A"     "A177A"     "A179A"     "A181A"     "A182A"    
# [37] "A192A"     "A193A"     "A194A"     "A195A"     "A197A"     "A201A"     "A202A"     "A204A"     "A206A"    
# [46] "A208A"     "A209A"     "A212A"     "A213A"     "A214A"     "A215A"     "A216A"     "A217A"     "A218A"    
# [55] "A219A"     "A222A"     "A223A"     "A224A"     "A225A"     "A226A"     "A227A"     "A228A"     "A229A"    
# [64] "A231A"     "A235A"     "A236A"     "A237A"     "A238A"     "A242A"     "A243A"     "A244A"     "A245A"    
# [73] "A247A"     "A248A"     "A252A"     "A253A"     "A256A"     "A433A"     "A434A"     "A436A"     "A529A"    
# [82] "A534A"     "A747A"     "A748A"     "A752A"     "A764A"    

setdiff(colnames(ce_bind_2024), colnames(ce_bind_2019)) # what is in 2024 that is not in 2019
# [1] "E03"       "E04"       "SECTOR"    "SUBSECTOR" "RAMA"      "SUBRAMA"   "CLASE"
```

What we really need is:

* The number of jobs, *H000A*;
* The employer contributions to social security schemes *J300A*,
* Other social benefits *J400A*;
* Total revenue from the supply of goods and services (millions of pesos) *M000A*; and
* Total revenue (millions of pesos) *A800A*

For now.

```{r}
## first, standardize the names of the columns for binding.

ce_rename_2014_2009_2004 <- ce_bind_2014_2009_2004 %>% 
  select(E03 = clave_entidad, E04 = clave_municipio, CODIGO = clave_actividad_economica, ano_censal = anio_censal, UE, H000A, J300A, J400A, M000A, A800A)

ce_rename_2019 <- ce_bind_2019 %>% 
  mutate(ano_censal = "2019") %>% 
  select(E03 = ENTIDAD, E04 = MUNICIPIO, CODIGO, ano_censal, ID_ESTRATO, UE, H000A, J300A, J400A, M000A, A800A)  

ce_rename_2024 <- ce_bind_2024 %>% 
  mutate(ano_censal = "2024") %>% 
  select(E03, E04, CODIGO, ano_censal, ID_ESTRATO, UE, H000A, J300A, J400A, M000A, A800A)
```


```{r}
# next, bind them together, fixing any differing types.

str(ce_rename_2014_2009_2004)
str(ce_rename_2019)
str(ce_rename_2024)

ce_all_years <- bind_rows(ce_rename_2014_2009_2004, ce_rename_2019, ce_rename_2024)

# write_csv(ce_all_years, here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/ce_int","ce_all_years.csv"))

# pesca_lapaz <- ce_all_years %>% 
#   filter(CODIGO %in% "1141") %>% 
#   filter(E03 %in% "3") %>% 
#   filter(E04 %in% "3")
```


## Clean and isolate to GoC municipalities

```{r}
# first, make sure the column names match.

municipio_codigo_rename <- municipio_codigo_df %>% 
  mutate(clave_estado = as.numeric(clave_estado),
         clave_municipio = as.numeric(clave_municipio)) %>% 
  rename(E03 = clave_estado, 
         E04 = clave_municipio) %>% 
  select(-tasa_regional)

# as sf if needed
municipio_codigo_rename_sf <- municipio_codigo_sf %>%
  mutate(clave_estado = as.numeric(clave_estado),
         clave_municipio = as.numeric(clave_municipio)) %>% 
  rename(E03 = clave_estado,
         E04 = clave_municipio) %>%
  select(-tasa_regional)
```


```{r}
# now filter the ce to only those municipalities. 
ce_all_years_goc <- ce_all_years %>% 
  inner_join(municipio_codigo_rename, by = c("E03", "E04")) %>% 
  janitor::clean_names()

# pesca_lapaz <- ce_all_years_goc_fix %>% 
#   filter(codigo %in% "1141") %>% 
#   filter(e03 %in% "3") %>% 
#   filter(e04 %in% "3")

ce_all_years_goc_sf <- ce_all_years %>% 
  inner_join(municipio_codigo_rename_sf, by = c("E03", "E04")) %>% 
  janitor::clean_names() %>% 
  st_as_sf()

# mapview(ce_all_years_goc_sf)

unique(ce_all_years_goc$municipio) 
#  [1] "ENSENADA"                      "MEXICALI"                      "TECATE"                       
#  [4] "TIJUANA"                       "PLAYAS DE ROSARITO"            "COMONDU"                      
#  [7] "MULEGE"                        "LA PAZ"                        "LOS CABOS"                    
# [10] "LORETO"                        "CABO CORRIENTES"               "PUERTO VALLARTA"              
# [13] "ACAPONETA"                     "AHUACATLAN"                    "AMATLAN DE CAÑAS"             
# [16] "COMPOSTELA"                    "IXTLAN DEL RIO"                "JALA"                         
# [19] "XALISCO"                       "ROSAMORADA"                    "RUIZ"                         
# [22] "SAN BLAS"                      "SAN PEDRO LAGUNILLAS"          "SANTA MARIA DEL ORO"          
# [25] "SANTIAGO IXCUINTLA"            "TECUALA"                       "TEPIC"                        
# [28] "TUXPAN"                        "BAHIA DE BANDERAS"             "AHOME"                        
# [31] "ANGOSTURA"                     "BADIRAGUATO"                   "CONCORDIA"                    
# [34] "COSALA"                        "CULIACAN"                      "CHOIX"                        
# [37] "ELOTA"                         "ESCUINAPA"                     "EL FUERTE"                    
# [40] "GUASAVE"                       "MAZATLAN"                      "MOCORITO"                     
# [43] "ROSARIO"                       "SALVADOR ALVARADO"             "SAN IGNACIO"                  
# [46] "SINALOA"                       "NAVOLATO"                      "AGUA PRIETA"                  
# [49] "ALAMOS"                        "ALTAR"                         "ATIL"                         
# [52] "BACUM"                         "BAVIACORA"                     "BENJAMIN HILL"                
# [55] "CABORCA"                       "CAJEME"                        "EMPALME"                      
# [58] "ETCHOJOA"                      "GUAYMAS"                       "HERMOSILLO"                   
# [61] "HUATABAMPO"                    "IMURIS"                        "MAGDALENA"                    
# [64] "MAZATAN"                       "NAVOJOA"                       "NOGALES"                      
# [67] "OPODEPE"                       "PITIQUITO"                     "PUERTO PEÑASCO"               
# [70] "SAN JAVIER"                    "SAN LUIS RIO COLORADO"         "SAN MIGUEL DE HORCASITAS"     
# [73] "SANTA ANA"                     "TUBUTAMA"                      "GENERAL PLUTARCO ELIAS CALLES"
# [76] "BENITO JUAREZ"                 "SAN IGNACIO RIO MUERTO"        "SAN QUINTÍN"                  
# [79] "SAN FELIPE"  

```

Now limit to the actividades we care about:
```{r}
# limit to marine-related economic activities (df)
ce_all_years_goc_marine <- inner_join(ce_actividades_marino, ce_all_years_goc, by = "codigo")
write_csv(ce_all_years_goc_marine, here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/ce_int","ce_all_years_goc_marine.csv"))
  
# limit to marine-related economic activities (sf)
ce_all_years_goc_marine_sf <- inner_join(ce_actividades_marino, ce_all_years_goc_sf, by = "codigo") %>% 
  st_as_sf()
st_write(ce_all_years_goc_marine_sf, here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/ce_int","ce_all_years_goc_marine_sf.shp"))

mapview(ce_all_years_goc_marine_sf)
```

Exploration of ID_ESTRATO:
```{r}
## now here, I am looking through to understand the id_estrato column better.

## when I look here, I see that Baja California does not have data from every municipio code.  It is only from 12567, not 3 and 4. 
pesca_ce_all_years_goc_marine <- ce_all_years_goc_marine %>% 
  filter(codigo %in% "1141") %>% 
  filter(ano_censal %in% "2024")

## when I go back to the raw data, I find that was the case originally...
pesca_ce_bind_2024 <- ce_bind_2024 %>%  filter(CODIGO %in% "1141")
### and also, that when the ID_ESTRATO is NA, then all the UEs for Baja California for this economic activity add up to the state total.  So to me that means that there are no registered economic units in municipalities 02003 (TECATE) or 02004 (TIJUANA).  WHICH, when we look at a map, those are the inland areas at the top of BC!!  Tijuana has a small area of beach, but it makes sense to me that there are no registered fishers there because it is so close to the border.  And likely why 02002 and 02005 are low numbers, considering that in 2024, San Felipe is split from Mexicali.  And San Felipe, 02007 has a high number of fishers.

## NOTE that 02001, which previously encompassed a large portion of the northern pacific-side *and* gulf-side of BC, is now only pacific-side!!! <https://en.wikipedia.org/wiki/Municipalities_of_Baja_California>.  

## so for 2024, we should only consider 02002, 02007, and 02006 to be GoC.
```

Now make sure a region is assigned:
```{r}
inland_50 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer","inland_50km_goc.shp"))
ce_all_years_goc_marine_sf_trans <- st_transform(ce_all_years_goc_marine_sf, st_crs(inland_50))
ce_goc_marine_activities_rgn <- st_intersection(ce_all_years_goc_marine_sf_trans, inland_50)

ce_goc_marine_activities_rgn_2024 <- ce_goc_marine_activities_rgn %>% 
  filter(ano_censal %in% 2024) %>% 
  filter(codigo %in% "1141")

mapview(ce_goc_marine_activities_rgn_2024)
```

```{r}
ce_goc_marine_activities_rgn <- st_join(
  ce_all_years_goc_marine_sf_trans,
  inland_50,
  join = st_intersects,
  left = FALSE   # keep only inside buffer
)

ce_goc_marine_activities_rgn_2024 <- ce_goc_marine_activities_rgn %>% 
  filter(ano_censal %in% 2024) %>% 
  filter(codigo %in% "56151")

mapview(ce_goc_marine_activities_rgn_2024, zcol = "rgn_id") 


```

# COME BACK HERE TMRW ^ Region 5 has a Region 1 or 2 municipio!!!

```{r}
tsdem_goc_rgn_byrgn <- tsdem_goc_rgn %>% 
  group_by(rgn_id) %>% 
  summarize(num_individ = n())

mapview(tsdem_goc_rgn_byrgn, zcol = "num_individ") + mapview(inland_50, zcol = "rgn_id")
```


## Identify the Ramas and Subramas (sectors) we can obtain data from

```{r}
library(tibble)

ce_actividades_marino <- tribble(
  ~codigo,  ~descripcion,                                                        ~nivel,   ~sector_marino_propuesto,                                           ~notas_breve,
  # --- VIAJES Y TOURS ---
  "56151",  "Agencias de viajes",                                                 "Subrama",  "Coastal and marine tourism / Ecotourism",                      "Depende de localización/mercado, clave para turismo costero/marino.",
  "56152",  "Organización de excursiones y paquetes turísticos (Operadores tours)","Subrama", "Coastal and marine tourism; Sports fishing tours / recreational fishing", "Operadores de tours, incluye pesca deportiva si los paquetes son marinos.",
  "56159",  "Otros servicios de reservaciones",                                   "Subrama",  "Coastal and marine tourism",                                   "Reservaciones vinculadas a turismo; costero si la oferta es de playa/mar.",

  # --- MARINAS Y RECREACIÓN COSTERA ---
  "71393",  "Marinas turísticas",                                                 "Subrama",  "Coastal and marine tourism; Maritime transport and port services; Sports fishing tours / recreational fishing","Infraestructura de atraque, servicios a embarcaciones de recreo.",
  "71219",  "Grutas, parques naturales y otros sitios del patrimonio cultural",   "Subrama",  "Ecotourism",                                                   "Aplica si se ubican en zonas costeras/marinas protegidas.",

  # --- ALOJAMIENTO TURÍSTICO ---
  "72111",  "Hoteles y moteles, excepto hoteles con casino",                      "Subrama",  "Coastal and marine tourism / Ecotourism",                      "Hospedaje en zonas costeras o cercanas al mar.",
  "72119",  "Cabañas, villas y similares",                                        "Subrama",  "Coastal and marine tourism / Ecotourism",                      "Cabañas frente al mar, eco-lodges costeros.",
  "72121",  "Campamentos y albergues recreativos",                                "Subrama",  "Ecotourism / Coastal and marine tourism",                      "Campamentos en litorales o áreas marinas protegidas.",
  "72131",  "Pensiones y casas de huéspedes, y deptos/casas amueblados con hotelería","Subrama","Coastal and marine tourism / Ecotourism",                "Hospedaje pequeño en destinos costeros.",

  # --- SERVICIOS DE ALIMENTOS RELACIONADOS CON TURISMO MARINO ---
  # Aquí solo con clases porque SCIAN no define una subrama única para ‘restaurantes’:
  "722512","Restaurantes de pescados y mariscos",                                 "Clase",    "Seafood processing and trade; Coastal and marine tourism",     "Consumo de productos pesqueros; turismo gastronómico marino.",
  "722511","Restaurantes a la carta o comida corrida",                            "Clase",    "Coastal and marine tourism / Seafood processing and trade",    "Restaurantes turísticos; si especializados en mariscos, ver 722512.",
  "722513","Restaurantes de antojitos",                                           "Clase",    "Coastal and marine tourism",                                   "En zonas turísticas costeras.",
  "722514","Restaurantes de tacos y tortas",                                      "Clase",    "Coastal and marine tourism",                                   "Idem.",
  "722515","Cafeterías, neverías, refresquerías y similares",                     "Clase",    "Coastal and marine tourism",                                   "Servicios de alimentos/bebidas en destinos de playa.",
  "722516","Restaurantes de autoservicio",                                        "Clase",    "Coastal and marine tourism",                                   "Fast food en zonas costeras.",
  "722517","Restaurantes de pizzas, hamburguesas, hot dogs y pollos para llevar", "Clase",    "Coastal and marine tourism",                                   "Idem.",
  "722518","Restaurantes que preparan otro tipo de alimentos para llevar",        "Clase",    "Coastal and marine tourism",                                   "Idem.",
  "722519","Otros servicios de preparación de alimentos para consumo inmediato",  "Clase",    "Coastal and marine tourism",                                   "Idem.",

  # --- MANEJO DE RESIDUOS (ENERGY AND INFRASTRUCTURE, APOYO AMBIENTAL) ---
  "5621",   "Recolección de residuos",                                            "Rama",     "Energy and infrastructure",                                    "Si opera en puertos/zonas costeras.",
  "56211",  "Recolección de residuos peligrosos",                                 "Subrama",  "Energy and infrastructure",                                    "Manejo de residuos industriales/portuarios.",
  "56212",  "Recolección de residuos no peligrosos",                              "Subrama",  "Energy and infrastructure",                                    "Residuos urbanos/costeros.",
  "5622",   "Tratamiento y disposición final de residuos",                        "Rama",     "Energy and infrastructure",                                    "Incluye rellenos sanitarios costeros.",
  "56221",  "Tratamiento y disposición de residuos peligrosos",                   "Subrama",  "Energy and infrastructure",                                    "Emisiones de puertos/industrias costeras.",
  "56222",  "Tratamiento y disposición de residuos no peligrosos",                "Subrama",  "Energy and infrastructure",                                    "Idem.",
  "5629",   "Servicios de remediación, recuperación y otros servicios de manejo de residuos","Rama","Energy and infrastructure",                      "Remediación ambiental en zonas costeras/marinas.",
  "56291",  "Servicios de remediación",                                           "Subrama",  "Energy and infrastructure",                                    "Limpieza de sitios contaminados.",
  "56292",  "Recuperación de residuos",                                           "Subrama",  "Energy and infrastructure",                                    "Reciclaje asociado a cadenas portuarias/turísticas.",
  "56299",  "Otros servicios de manejo de residuos",                              "Subrama",  "Energy and infrastructure",                                    "Servicios especializados.",

  # --- AQUACULTURE (VERSIÓN 2019–2024 CON SUBRAMA GENERAL) ---
  "11251",  "Acuicultura",                                                        "Subrama",  "Aquaculture",                                                  "Subrama general; usar para agregar clases 112511–112519.",
  "112511", "Camaronicultura",                                                    "Clase",    "Aquaculture",                                                  "Producción de camarón en sistemas controlados (frecuentemente marinos/costeros).",
  "112512", "Piscicultura y otra acuicultura, excepto camaronicultura",           "Clase",    "Aquaculture",                                                  "Incluye peces y otras especies; filtrar luego las marinas si lo necesitas.",
  "112513", "Camaronicultura y acuicultura de otros crustáceos en agua salada",   "Clase",    "Aquaculture",                                                  "Acuicultura marina/costera de camarón y otros crustáceos.",
  "112514", "Camaronicultura y acuicultura de otros crustáceos en agua dulce",    "Clase",    "Aquaculture",                                                  "Acuicultura continental; separable espacialmente.",
  "112515", "Piscicultura en agua dulce",                                         "Clase",    "Aquaculture",                                                  "Acuicultura continental.",
  "112516", "Piscicultura en agua salada",                                        "Clase",    "Aquaculture",                                                  "Acuicultura marina de peces.",
  "112519", "Otra acuicultura",                                                   "Clase",    "Aquaculture",                                                  "Otras especies acuícolas.",

  # --- FISHING (RAMA + SUBRAMA GENERAL + CLASES) ---
  "1141",   "Pesca",                                                              "Rama",     "Fishing",                                                      "Explotación de recursos pesqueros silvestres.",
  "11411",  "Pesca",                                                              "Subrama",  "Fishing",                                                      "Subrama general de pesca.",
  "114111", "Pesca de camarón",                                                   "Clase",    "Fishing",                                                      "Pesca de camarón silvestre.",
  "114112", "Pesca de túnidos",                                                   "Clase",    "Fishing",                                                      "Pesca de atunes y afines.",
  "114113", "Pesca de sardina y anchoveta",                                       "Clase",    "Fishing",                                                      "Pesca de pelágicos menores.",
  "114119", "Otras pescas",                                                       "Clase",    "Fishing",                                                      "Otras especies de peces, crustáceos y moluscos.",

  # --- SEAFOOD PROCESSING AND TRADE (CLASES CLAVE) ---
  "311711", "Congelación de pescados, mariscos y otros productos marinos",        "Clase",    "Seafood processing and trade",                                 "Procesamiento primario (congelación) de productos marinos.",
  "311712", "Secado, salado y ahumado de pescados y mariscos",                    "Clase",    "Seafood processing and trade",                                 "Procesado por conservación tradicional.",
  "311713", "Preparación y envasado de pescados y mariscos",                      "Clase",    "Seafood processing and trade",                                 "Elaboración de productos de valor agregado de origen marino.",

  # --- SHIPBUILDING AND MARINE EQUIPMENT (EJEMPLOS, AJUSTA A TU SCIAN OFICIAL) ---
  "336611", "Construcción de embarcaciones y estructuras flotantes",              "Clase",    "Shipbuilding and marine equipment",                             "Astilleros, construcción de barcos.",
  "336612", "Reparación y mantenimiento de embarcaciones",                        "Clase",    "Shipbuilding and marine equipment",                             "Talleres navales, mantenimiento marino.",

  # --- MARITIME TRANSPORT AND PORT SERVICES (EJEMPLOS) ---
  "48311",  "Transporte por vía acuática de corta distancia",                     "Clase",    "Maritime transport and port services",                          "Transporte marítimo/cabotaje.",
  "48321",  "Transporte por vía acuática de larga distancia",                     "Clase",    "Maritime transport and port services",                          "Transporte marítimo de altura.",
  "48831",  "Servicios relacionados con el transporte por vía acuática",          "Clase",    "Maritime transport and port services",                          "Servicios portuarios, maniobras, etc.",

  # --- MARINE RESEARCH AND CONSULTING SERVICES (GENÉRICOS; MARINO POR PROYECTO) ---
  "54162",  "Servicios de consultoría en medio ambiente",                         "Clase",    "Marine research and consulting services",                       "Consultoría ambiental; identificar luego proyectos marinos/costeros.",
  "54171",  "Servicios de investigación científica y desarrollo",                 "Clase",    "Marine research and consulting services",                       "I+D; identificar unidades dedicadas a temas marinos/costeros."
)

```

## Save subset of job *number* metrics

```{r}

```

## Save subset of job *quality* metrics

```{r}

```

## Save subset of *revenue* metrics

```{r}

```

