---
title: 'OHI GoC 2025: Livelihoods & Economies'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# LE: Censos Economicos (CE) 

This script brings in data from the Censos Economicos (CE) and cleans it, prepping it for analysis and use in the LIV and ECO subgoals.  The CE contains data for the number of jobs, quality of jobs, and revenue of certain marine-related sectors.  

This analysis uses only the Economic Census (CE)  conducted by Instituto Nacional de Estadística y Geografía (INEGI) rather than the National Survey of Occupation and Employment (ENOE) because the Census provides a complete count of businesses, not just a sample of households. That matters because this project evaluates marine-sector employment at the municipal level, and the Economic Census covers every formally registered establishment in each municipality. In contrast, ENOE is designed to be representative at the state and national levels, which means municipal estimates would be less precise and subject to sampling variability.

The Economic Census also allows employment, wages, benefits, and revenue to be measured using the same establishments and the same reporting system. This ensures internal consistency between the Livelihoods (number of jobs and job quality) and Economies (revenue) components of the analysis. Using a single census-based data source reduces complexity in the methodology, improves comparability across years, and avoids introducing additional assumptions that would be needed to merge survey-based labor estimates with economic data at the establishment level.

Note that while the Labor Indicators for the Municipalities of Mexico (ILMM) data source from INEGI has taken the ENOE and adjusted for municipality estimations instead of state-level, it does not disaggregate by sector. Therefore, we would not be able to isolate marine-related sectors and industries.  Additionally, it provides the percentage of the municipal labor force that is employed vs informal employment, but it neglects any job quality metrics.  

However, if we were desperate for an estimate of the number of jobs for a certain sector and municipality, we could make an assumption of sectoral homogeneity.  That would allow us to take the ENOE state-level data and multiply by the proportion of occupied individuals in that municipality within the state.  However, that is a big assumption to make, and unfortunately the locality data was suppressed for confidentiality.  

Therefore, we chose to use the Economic Census in lieu of the ENOE or ILMM.

*Outline of script:*

1. Read in data for each year and state.
2. Bind the data all into one dataframe, harmonizing the columns/variables
3. Isolate only variables we will be using in future LE analyses
4. Identify marine-related sectors out of the available sectors, also called "activities"
5. Limit to marine-related sectors
6. Limit to GoC municipalities and assign Region IDs.
7. Ensure that there is only the total annual amount for each year-region-sector (if there are no totals and data is only represented in straums, sum those).
8. Adjust the census years to reflect the year the data was collected, not released.
9. Save to Mazu
10. Explore using `plotly`.


```{r}
library(tidyverse)
library(mapview)
library(sf)
```

## Read in data

For downloading years 2024 and 2019: 

* Go to <https://www.inegi.org.mx/programas/ce/2024/> and click on "Datos Abiertos" -> "Datos de los Censos Economicos 2024" -> Entidad Federativa.


For downloading 2014 - 2004, go to "Descarga Masiva" at <https://www.inegi.org.mx/app/descarga/?p=1664#denue>, and dplyr::select the state you would like to download for on the left hand side of the page.  Then, once you have dplyr::selected the state, you will click "Indicadores" -> "Programas" -> "Censos Economicos (CE)" -> "2014" -> "Indicadores".  You will then have to do that for every one of the GoC states.

Since we have raw data for each one of the states for every year, we will need to read each one in and bind them.

### 2024
```{r}
# baja california
bc_2024_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2024/conjunto_de_datos_ce_bc_2024_csv/conjunto_de_datos","tr_ce_bc_2024.csv")) 
unique(bc_2024_raw$E04) # [1] NA    "001" "002" "003" "004" "005" "006" "007", now has San Felipe!! 

bc_diccionario_2024_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2024/conjunto_de_datos_ce_bc_2024_csv/diccionario_de_datos","diccionario_de_datos_ce2024.csv"))

bc_actividad_2024_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2024/conjunto_de_datos_ce_bc_2024_csv/catalogos","tc_codigo_actividad.csv"))

# baja california sur
bcs_2024_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2024/conjunto_de_datos_ce_bcs_2024_csv/conjunto_de_datos","tr_ce_bcs_2024.csv"))

# jalisco
jal_2024_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2024/conjunto_de_datos_ce_jal_2024_csv/conjunto_de_datos","tr_ce_jal_2024.csv"))

# nayarit
nay_2024_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2024/conjunto_de_datos_ce_nay_2024_csv/conjunto_de_datos","tr_ce_nay_2024.csv"))

# sinaloa
sin_2024_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2024/conjunto_de_datos_ce_sin_2024_csv/conjunto_de_datos","tr_ce_sin_2024.csv"))

# sonora
son_2024_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2024/conjunto_de_datos_ce_son_2024_csv/conjunto_de_datos","tr_ce_son_2024.csv"))
```

Merge them into one df:
```{r}
fix_types <- \(df) df |>
  mutate(
    E03 = as.numeric(E03),
    E04 = as.numeric(E04),
    CODIGO = as.character(CODIGO)
  )

ce_bind_2024 <- bind_rows(
  fix_types(bc_2024_raw),
  fix_types(bcs_2024_raw),
  fix_types(jal_2024_raw),
  fix_types(nay_2024_raw),
  fix_types(sin_2024_raw),
  fix_types(son_2024_raw)
)
```



### 2019
```{r}
# baja california
bc_2019_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2019/ce2019_bc_csv/conjunto_de_datos","ce2019_bc.csv"))
unique(bc_2019_raw$MUNICIPIO) # [1] NA    "001" "002" "003" "004" "005"

bc_actividad_2019_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2019/ce2019_bc_csv/catalogos","tc_codigo_actividad.csv"))

# baja california sur
bcs_2019_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2019/ce2019_bcs_csv/conjunto_de_datos","ce2019_bcs.csv"))

# jalisco
jal_2019_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2019/ce2019_jal_csv/conjunto_de_datos","ce2019_jal.csv"))

# nayarit
nay_2019_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2019/ce2019_nay_csv/conjunto_de_datos","ce2019_nay.csv"))

# sinaloa
sin_2019_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2019/ce2019_sin_csv/conjunto_de_datos","ce2019_sin.csv"))

# sonora
son_2019_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2019/ce2019_son_csv/conjunto_de_datos","ce2019_son.csv"))
```

Merge them:
```{r}
spec(bc_2019_raw)
spec(jal_2019_raw)

fix_types <- \(df) df |>
  mutate(
    ENTIDAD = as.numeric(ENTIDAD),
    MUNICIPIO = as.numeric(MUNICIPIO),
    CODIGO = as.character(CODIGO) # fixing the types so that they can bind together
  )

ce_bind_2019 <- bind_rows(
  fix_types(bc_2019_raw),
  fix_types(bcs_2019_raw),
  fix_types(jal_2019_raw),
  fix_types(nay_2019_raw),
  fix_types(sin_2019_raw),
  fix_types(son_2019_raw)
)
```


### 2014, 2009, 2004
```{r}
# baja california
bc_2014_2009_2004_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2014_2009_2004/ce_02_bc_csv/conjunto_de_datos","ce_inegi_02.csv"))
unique(bc_2014_2009_2004_raw$clave_municipio) # [1] NA    "001" "002" "003" "004" "005"

# baja california sur
bcs_2014_2009_2004_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2014_2009_2004/ce_03_bcs_csv/conjunto_de_datos","ce_inegi_03.csv"))

# jalisco
jal_2014_2009_2004_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2014_2009_2004/ce_14_jal_csv/conjunto_de_datos","ce_inegi_14.csv"))

# nayarit
nay_2014_2009_2004_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2014_2009_2004/ce_18_nay_csv/conjunto_de_datos","ce_inegi_18.csv"))

# sinaloa
sin_2014_2009_2004_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2014_2009_2004/ce_25_sin_csv/conjunto_de_datos","ce_inegi_25.csv"))

# sonora
son_2014_2009_2004_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_censos_economicos/d2026/2014_2009_2004/ce_26_son_csv/conjunto_de_datos","ce_inegi_26.csv"))
```

Merge them:
```{r}
spec(bc_2014_2009_2004_raw)
spec(bcs_2014_2009_2004_raw)
spec(nay_2014_2009_2004_raw)
spec(jal_2014_2009_2004_raw)
spec(sin_2014_2009_2004_raw)
spec(son_2014_2009_2004_raw)


fix_types <- \(df) df %>% # needed to do a lot more here because the data is presented differently, it is more cleaned up.
  mutate(
    clave_entidad = as.numeric(clave_entidad),  # bc has char, others have numeric
    clave_municipio = as.numeric(clave_municipio),
    across(c(J400A, J500A, J600A, K010A, K020A, K030A, K060A, K070A, 
             K096A, K311A, K620A, K810A, K910A, M010A, M030A, 
             P000A, P000B, P000C, P030A, P030B, P030C, P100A, P100B,
             Q020A, Q030A), 
           ~(as.numeric(.x))),
    anio_censal = as.character(anio_censal)
  )

ce_bind_2014_2009_2004 <- bind_rows(
  fix_types(bc_2014_2009_2004_raw),
  fix_types(bcs_2014_2009_2004_raw),
  fix_types(jal_2014_2009_2004_raw),
  fix_types(nay_2014_2009_2004_raw),
  fix_types(sin_2014_2009_2004_raw),
  fix_types(son_2014_2009_2004_raw)
)

## out of curiousity, looking at fishing in La Paz:
# pesca_lapaz <- ce_bind_2014_2009_2004 %>% 
  # filter(clave_actividad_economica %in% "1141") %>% 
  # filter(clave_entidad %in% "3") %>% 
  # filter(clave_municipio %in% "003")
```


## Merge all years

Now that we have separate bound datasets for *2004-2014*, *2019*, and *2024*, which really refer to data collected in *2003-2013*, *2018*, and *2023*, respectively, we can now merge them all into one dataframe. 

Before we do that, let's first do some explorations of the differences between the years.

```{r}
setdiff(colnames(ce_bind_2014_2009_2004), colnames(ce_bind_2019))
# [1] "clave_entidad"             "entidad_federativa"        "clave_municipio"           "municipio"                
# [5] "clave_actividad_economica" "actividad_economica"       "anio_censal"               "K040A"                    
# [9] "K041A"
setdiff(colnames(ce_bind_2014_2009_2004), colnames(ce_bind_2024))
# [1] "clave_entidad"             "entidad_federativa"        "clave_municipio"           "municipio"                
# [5] "clave_actividad_economica" "actividad_economica"       "anio_censal"               "K040A"                    
# [9] "K041A" 

setdiff(colnames(ce_bind_2019), colnames(ce_bind_2024)) # what is in 2019 that is not in 2024
#  [1] "ENTIDAD"   "MUNICIPIO" "A112A"     "A113A"     "A114A"     "A115A"     "A116A"     "A117A"     "A118A"    
# [10] "A119A"     "A141A"     "A143A"     "A144A"     "A145A"     "A146A"     "A147A"     "A148A"     "A149A"    
# [19] "A151A"     "A152A"     "A153A"     "A154A"     "A155A"     "A156A"     "A157A"     "A158A"     "A159A"    
# [28] "A171A"     "A173A"     "A174A"     "A175A"     "A176A"     "A177A"     "A179A"     "A181A"     "A182A"    
# [37] "A192A"     "A193A"     "A194A"     "A195A"     "A197A"     "A201A"     "A202A"     "A204A"     "A206A"    
# [46] "A208A"     "A209A"     "A212A"     "A213A"     "A214A"     "A215A"     "A216A"     "A217A"     "A218A"    
# [55] "A219A"     "A222A"     "A223A"     "A224A"     "A225A"     "A226A"     "A227A"     "A228A"     "A229A"    
# [64] "A231A"     "A235A"     "A236A"     "A237A"     "A238A"     "A242A"     "A243A"     "A244A"     "A245A"    
# [73] "A247A"     "A248A"     "A252A"     "A253A"     "A256A"     "A433A"     "A434A"     "A436A"     "A529A"    
# [82] "A534A"     "A747A"     "A748A"     "A752A"     "A764A"    

setdiff(colnames(ce_bind_2024), colnames(ce_bind_2019)) # what is in 2024 that is not in 2019
# [1] "E03"       "E04"       "SECTOR"    "SUBSECTOR" "RAMA"      "SUBRAMA"   "CLASE"
```

### & isolate metrics

What we really need is:

* The number of jobs, *H000A*;
* The employer contributions to social security schemes *J300A*,
* Other social benefits *J400A*;
* Total revenue from the supply of goods and services (millions of pesos) *M000A*; and
* Total revenue (millions of pesos) *A800A*

So, let's isolate those variables (columns).

```{r}
## first, standardize the names of the columns for binding.

ce_rename_2014_2009_2004 <- ce_bind_2014_2009_2004 %>% 
  dplyr::select(E03 = clave_entidad, E04 = clave_municipio, CODIGO = clave_actividad_economica, ano_censal = anio_censal, UE, H000A, H000B, H000C, H010A, H010B, H010C, J000A, J600A, J300A, J400A, M000A, A800A)

ce_rename_2019 <- ce_bind_2019 %>% 
  mutate(ano_censal = "2019") %>% 
  dplyr::select(E03 = ENTIDAD, E04 = MUNICIPIO, CODIGO, ano_censal, ID_ESTRATO, UE, H000A, H000B, H000C, H010A, H010B, H010C, J000A, J600A, J300A, J400A, M000A, A800A)

ce_rename_2024 <- ce_bind_2024 %>% 
  mutate(ano_censal = "2024") %>% 
  dplyr::select(E03, E04, CODIGO, ano_censal, ID_ESTRATO, UE, H000A, H000B, H000C, H010A, H010B, H010C, J000A, J600A, J300A, J400A, M000A, A800A)
```


```{r}
# next, bind them together, fixing any differing types.

str(ce_rename_2014_2009_2004)
str(ce_rename_2019)
str(ce_rename_2024)

ce_all_years <- bind_rows(ce_rename_2014_2009_2004, ce_rename_2019, ce_rename_2024)

## again, we can check La Paz:
# pesca_lapaz <- ce_all_years %>% 
#   filter(CODIGO %in% "1141") %>% 
#   filter(E03 %in% "3") %>% 
#   filter(E04 %in% "3")
```


Save the fully bound df:
```{r}
# write_csv(ce_all_years, here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/ce_int","ce_all_years.csv"))
```



## Identify marine activities and CE sector codes 

Now, we want to limit our analysis to MARINE-RELATED sectors, not all sectors in the CE.  First, we should see what sectors are available in the 2004-2014 data, to ensure we have reference points for those sectors:

```{r}
all_sectors <- ce_bind_2014_2009_2004 %>% 
  dplyr::select(clave_actividad_economica, actividad_economica) %>% 
  group_by(clave_actividad_economica, actividad_economica) %>% 
  distinct()

# write_csv(all_sectors, here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/ce_int","ce_all_sectors_2004_2014.csv"))
```

Then I went through each one and identified those that are marine-related.  See this spreadsheet for more information on marine-related sectors that were chosen: https://docs.google.com/spreadsheets/d/1X-Y8haoH3p59tzyAAGBKNggXJ10LWm1Ck15rjOc_nk8/edit?gid=265681868#gid=265681868 and [methods](https://docs.google.com/document/d/17YoomySVRz6IHWR6BYlDN11uF26A7qEjAfkDzssq1ZU/edit?tab=t.0).

We can now make a tibble of these sectors that we are interested in:
```{r}
library(tibble)

ce_actividades_marino <- tribble(
  ~codigo,   ~general_sector,           ~specific_sector,                  ~esp_sector_desc,                                                                                                                        ~eng_sector_desc,                                                                                                                       ~assumption,

  # FOOD PROVISIONING ----
  "1141",    "food_provisioning",       "fishing",
             "Pesca; incluye subrama 11411 y clases específicas.",
             "Fishing; includes sub-branch 11411 and specific classes.",
             "Assume all marine.",

  "1125",    "food_provisioning",       "mariculture",
             "Acuicultura; agregación para 11251 y clases. Las subramas y clases cambian según el año, por lo que no tiene sentido separar por tipo de acuicultura.",
             "Aquaculture; aggregation for 11251 and classes. The sub-branches and classes change depending on the year.",
             "Assume all marine for now; later include marine corrector from Erica's data.",

  "31171",   "food_provisioning",       "seafood_processing_trade",
             "Preparación y envasado de pescados y mariscos.",
             "Preparation and packaging of fish and seafood.",
             "Assume all marine.",
  "431123",  "food_provisioning",       "seafood_processing_trade",
             "Comercio al por mayor de pescados y mariscos.",
             "Wholesale trade of fish and seafood.",
             "Assume all marine.",
  "461123",  "food_provisioning",       "seafood_processing_trade",
             "Comercio al por menor de pescados y mariscos.",
             "Retail sale of fish and seafood.",
             "Assume all marine.",

  # INFRASTRUCTURE ----
  "237992",  "infrastructure",          "port_dock_construction_management",
             "Construcción de obras marítimas, fluviales y subacuáticas (muelles, rompeolas, dragados).",
             "Construction of maritime, river, and underwater works (piers, breakwaters, dredging).",
             "Assume all marine.",
  "4883",    "infrastructure",          "port_dock_construction_management",
             "Servicios relacionados con el transporte por agua: incluye 48839, 488390, 48831 Administración de puertos y muelles, 488310, 48833 Servicios para la navegación por agua, 488330.",
             "Services related to water transport: includes 48839, 488390, 48831 Port and dock administration, 488310, 48833 Water navigation services, 488330.",
             "Assume all marine.",

  "33661",   "infrastructure",          "boat_building",
             "Fabricación de embarcaciones.",
             "Manufacture of boats.",
             "Assume all marine.",

  # TIDAL ENERGY ----
  "221112",  "tidal_energy",            "tidal_energy",
             "Generación de electricidad a partir de energía hidráulica.",
             "Electricity generation from hydraulic energy.",
             "Assume all marine.",

  # MARINE AND COASTAL SHIPPING ----
  "483113",  "marine_coastal_shipping", "oil_gas_maritime_transport",
             "Transporte marítimo de petróleo y gas natural.",
             "Maritime transport of oil and natural gas.",
             "Assume all marine.",

  "4831",    "marine_coastal_shipping", "cargo_goods_maritime_transport",
             "Transporte marítimo, que incluye 483111 Transporte marítimo de altura, excepto de petróleo y gas natural, y 483112 Transporte marítimo de cabotaje, excepto de petróleo y gas natural.",
             "Maritime transport, which includes 483111 Deep-sea maritime transport, except for oil and natural gas, and 483112 Coastal maritime transport, except for oil and natural gas.",
             "Assume all marine.",

  # MARINE TOURISM ----
  "71393",   "marine_tourism",          "marine_tourist_transport",
             "Marinas turísticas.",
             "Tourist marinas.",
             "Assume all marine.",
  "4872",    "marine_tourism",          "marine_tourist_transport",
             "Transporte turístico por agua, incluye 48721 y 487210.",
             "Tourist water transport, includes 48721 and 487210.",
             "Assume all marine.",

  "7121",    "marine_tourism",          "coastal_recreation",
             "Museos, sitios históricos, zoológicos y similares.",
             "Museums, historical sites, zoos, and similar.",
             "Correct by population.",
  "72121",   "marine_tourism",          "coastal_recreation",
             "Campamentos y albergues recreativos.",
             "Campgrounds and recreational lodges.",
             "Correct by population.",
  "56152",   "marine_tourism",          "coastal_recreation",
             "Organización de excursiones y paquetes turísticos para agencias de viajes (Operadores de tours).",
             "Organization of excursions and tourist packages for travel agencies (tour operators).",
             "Correct by population.",

  "56151",   "marine_tourism",          "travel_planners",
             "Agencias de viajes.",
             "Travel agencies.",
             "Correct by population.",
  "56159",   "marine_tourism",          "travel_planners",
             "Otros servicios de reservaciones.",
             "Other reservation services.",
             "Correct by population.",

  "7211",   "marine_tourism",          "accommodations_food_drink",
             "Hoteles y moteles, y similares",
             "Hotels and motels, and similar",
             "Correct by population.",
  
  "72131",   "marine_tourism",          "accommodations_food_drink",
             "Pensiones y casas de huéspedes, y departamentos y casas amueblados con servicios de hotelería.",
             "Boarding houses, guest houses, and furnished apartments and houses with hotel services.",
             "Correct by population.",
  "72251",   "marine_tourism",          "accommodations_food_drink",
             "Servicios de preparación de alimentos y bebidas alcohólicas y no alcohólicas; incluye restaurantes y similares.",
             "Food and alcoholic and non-alcoholic beverage preparation services; includes restaurants and similar.",
             "Correct by population.",

  # NA,        "marine_tourism",          "sports_fishing",
             # NA,
             # NA,
             # "Tier 2; too data deficient to isolate. May be within tourist water transport (4872).",

  # MARINE RESEARCH AND CONSULTING ----
  "54162",   "marine_research_consulting", "environmental_research_consulting",
             "Servicios de consultoría en medio ambiente.",
             "Environmental consulting services.",
             "Correct by population.",
  "54171",   "marine_research_consulting", "environmental_research_consulting",
             "Servicios de investigación científica y desarrollo en ciencias naturales y exactas, ingeniería, y ciencias de la vida.",
             "Scientific research and development services in natural and exact sciences, engineering, and life sciences.",
             "Correct by population."
)

ce_actividades_marino
```

Quick kable to see it all in a pretty way:
```{r}
library(dplyr)
library(kableExtra)

ce_actividades_marino %>%
  arrange(specific_sector, codigo) %>%
  dplyr::select(
    codigo,
    specific_sector,
    esp_sector_desc,
    eng_sector_desc,
    assumption
  ) %>%
  kable(
    format   = "html",
    caption  = "Marine-related Economic Activities (Censos Económicos)",
    col.names = c(
      "Código",
      "Proposed Marine Sector",
      "Descripción (español)",
      "Description (English)",
      "Assumption / Notes"
    ),
    align    = c("c", "c", "l", "l", "l"),
    escape   = FALSE
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE,
    font_size         = 12
  ) %>%
  column_spec(1, bold = TRUE) %>%
  collapse_rows(
    columns = 2,  # collapse on specific_sector
    valign  = "top"
  )

```

See which sectors are present in 2019 and 2024, but not in 2004, 2009, & 2014.
```{r}
setdiff(unique(ce_actividades_marino$codigo), unique(all_sectors$clave_actividad_economica)) # [1] "221112" "483113" note that these two sectors, oil and gas maritime transport and hydraulic energy, did not exist as sectors in 2004 - 2014.

# write_csv(ce_actividades_marino, here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/ce_int","ce_identified_marine_sectors.csv"))
```


## Limit to marine activities in coastal municipios

First, bring in all municipality names and codes.
```{r}
## made in spatial/municipio/isolate_goc_municipios.qmd
municipio_codigo_df <- read_csv(here::here("/home/shares/ohi/OHI_GOC/spatial/land_admin_boundaries", "municipio_codigos_secreteria_de_economia.csv"))

# first, make sure the column names match.
municipio_codigo_rename <- municipio_codigo_df %>% 
  mutate(clave_estado = as.numeric(clave_estado),
         clave_municipio = as.numeric(clave_municipio)) %>% 
  rename(E03 = clave_estado, 
         E04 = clave_municipio) %>% 
  dplyr::select(-tasa_regional)
```

Filter the CE municipalities to those identified as coastal GoC.
```{r}
# now filter the ce to only those municipalities. 
ce_all_years_goc <- ce_all_years %>% 
  inner_join(municipio_codigo_rename, by = c("E03", "E04")) %>% 
  janitor::clean_names()

unique(ce_all_years_goc$municipio) 
#  [1] "ENSENADA"                      "MEXICALI"                      "TECATE"                       
#  [4] "TIJUANA"                       "PLAYAS DE ROSARITO"            "COMONDU"                      
#  [7] "MULEGE"                        "LA PAZ"                        "LOS CABOS"                    
# [10] "LORETO"                        "CABO CORRIENTES"               "PUERTO VALLARTA"              
# [13] "ACAPONETA"                     "AHUACATLAN"                    "AMATLAN DE CAÑAS"             
# [16] "COMPOSTELA"                    "IXTLAN DEL RIO"                "JALA"                         
# [19] "XALISCO"                       "ROSAMORADA"                    "RUIZ"                         
# [22] "SAN BLAS"                      "SAN PEDRO LAGUNILLAS"          "SANTA MARIA DEL ORO"          
# [25] "SANTIAGO IXCUINTLA"            "TECUALA"                       "TEPIC"                        
# [28] "TUXPAN"                        "BAHIA DE BANDERAS"             "AHOME"                        
# [31] "ANGOSTURA"                     "BADIRAGUATO"                   "CONCORDIA"                    
# [34] "COSALA"                        "CULIACAN"                      "CHOIX"                        
# [37] "ELOTA"                         "ESCUINAPA"                     "EL FUERTE"                    
# [40] "GUASAVE"                       "MAZATLAN"                      "MOCORITO"                     
# [43] "ROSARIO"                       "SALVADOR ALVARADO"             "SAN IGNACIO"                  
# [46] "SINALOA"                       "NAVOLATO"                      "AGUA PRIETA"                  
# [49] "ALAMOS"                        "ALTAR"                         "ATIL"                         
# [52] "BACUM"                         "BAVIACORA"                     "BENJAMIN HILL"                
# [55] "CABORCA"                       "CAJEME"                        "EMPALME"                      
# [58] "ETCHOJOA"                      "GUAYMAS"                       "HERMOSILLO"                   
# [61] "HUATABAMPO"                    "IMURIS"                        "MAGDALENA"                    
# [64] "MAZATAN"                       "NAVOJOA"                       "NOGALES"                      
# [67] "OPODEPE"                       "PITIQUITO"                     "PUERTO PEÑASCO"               
# [70] "SAN JAVIER"                    "SAN LUIS RIO COLORADO"         "SAN MIGUEL DE HORCASITAS"     
# [73] "SANTA ANA"                     "TUBUTAMA"                      "GENERAL PLUTARCO ELIAS CALLES"
# [76] "BENITO JUAREZ"                 "SAN IGNACIO RIO MUERTO"        "SAN QUINTÍN"                  
# [79] "SAN FELIPE"  

```

Now limit to the actividades we care about:
```{r}
# limit to marine-related economic activities (df)
ce_all_years_goc_marine <- inner_join(ce_actividades_marino, ce_all_years_goc, by = "codigo")
# write_csv(ce_all_years_goc_marine, here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/ce_int","ce_all_years_goc_marine.csv"))
```

Read in the municipality to rgn_id classifier df:
```{r}
## bring in the municipality to rgn_id code CSV I made...isolate_goc_municipios.qmd led to the manually made .CSV: `~/home/shares/ohi/OHI_GOC/spatial/municipality_to_rgnid/municipality_to_rgnid.csv`

municipality_to_rgnid <- read_csv(here::here("/home/shares/ohi/OHI_GOC/spatial/municipality_to_rgnid","municipality_to_rgnid.csv")) %>% 
  dplyr::select(-c("notes","municipio_raw")) %>% 
  janitor::clean_names() %>% 
  mutate(municipio = str_trim(municipio, side = "both"))

setdiff(unique(municipality_to_rgnid$municipio), unique(ce_all_years_goc_marine$municipio)) #character(0)
```

Join to obtain rgn_id:
```{r}
ce_all_years_goc_marine_rgnid  <- left_join(ce_all_years_goc_marine, municipality_to_rgnid, by = c("e03","e04", "estado", "municipio")) ## Joining with `by = join_by(e03, e04, estado, municipio)`
```

Check it out:
```{r}
# ce_goc_marine_activities_rgn_2024 <- ce_all_years_goc_marine_rgnid_sf %>%
#   filter(ano_censal %in% 2024) %>%
#   filter(codigo %in% "56151")

ce_goc_marine_activities_rgn_2024 <- ce_all_years_goc_marine_rgnid %>%
  filter(ano_censal %in% 2024) %>%
  filter(codigo %in% "1141")
unique(ce_goc_marine_activities_rgn_2024$municipio)

ce_all_years_goc_marine_sf_2024 <- ce_all_years_goc_marine %>% 
  filter(ano_censal %in% 2024) %>%
  filter(codigo %in% "1141")
unique(ce_all_years_goc_marine_sf_2024$municipio)

setdiff(unique(ce_goc_marine_activities_rgn_2024$municipio), unique(ce_all_years_goc_marine_sf_2024$municipio))
setdiff(unique(ce_all_years_goc_marine_sf_2024$h000a), unique(ce_goc_marine_activities_rgn_2024$h000a))

ce_goc_marine_activities_rgn_pesca_new <- ce_all_years_goc_marine_rgnid %>%
  filter(codigo %in% "1141")
```

Remove any municipalities that are *not* coastal municipios (aka, that have NA for rgn_id):
```{r}
isna_rgnid <- ce_all_years_goc_marine_rgnid %>% 
  filter(is.na(rgn_id))

unique(isna_rgnid$municipio)
#  [1] "ENSENADA"                 "PLAYAS DE ROSARITO"      
#  [3] "COMONDU"                  "TIJUANA"                 
#  [5] "TUBUTAMA"                 "TECATE"                  
#  [7] "NOGALES"                  "AGUA PRIETA"             
#  [9] "SANTA ANA"                "ALTAR"                   
# [11] "BAVIACORA"                "BENJAMIN HILL"           
# [13] "IMURIS"                   "MAGDALENA"               
# [15] "SAN JAVIER"               "MAZATAN"                 
# [17] "SAN MIGUEL DE HORCASITAS" "OPODEPE"                 
# [19] "ATIL"       

ce_actividades_rgnid <- ce_all_years_goc_marine_rgnid %>% 
  filter(!is.na(rgn_id))

unique(ce_all_years_goc_marine_rgnid$municipio) # 79
unique(ce_actividades_rgnid$municipio) # 60 municipios, makes sense: 79-19 = 60
```

Save the filtered dataframe of all years marine activities with region ids
```{r}
## save the filtered dataframe of all years marine activities with region ids
# write_csv(ce_actividades_rgnid, here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/ce_int","ce_all_years_goc_marine_w_rgnid.csv"))
```


## Final cleaning (Years and ID ESTRATO)

### ID ESTRATO

Exploration of ID_ESTRATO:
```{r}
## now here, I am looking through to understand the id_estrato column better.

## when I look here, I see that Baja California does not have data from every municipio code.  It is only from 12567, not 3 and 4. 
pesca_ce_all_years_goc_marine <- ce_all_years_goc_marine %>% 
  filter(codigo %in% "1141") %>% 
  filter(ano_censal %in% "2024")

## when I go back to the raw data, I find that was the case originally...
pesca_ce_bind_2024 <- ce_bind_2024 %>%  filter(CODIGO %in% "1141")
### and also, that when the ID_ESTRATO is NA, then all the UEs for Baja California for this economic activity add up to the state total.  So to me that means that there are no registered economic units in municipalities 02003 (TECATE) or 02004 (TIJUANA).  WHICH, when we look at a map, those are the inland areas at the top of BC!!  Tijuana has a small area of beach, but it makes sense to me that there are no registered fishers there because it is so close to the border.  And likely why 02002 and 02005 are low numbers, considering that in 2024, San Felipe is split from Mexicali.  And San Felipe, 02007 has a high number of fishers.

## NOTE that 02001, which previously encompassed a large portion of the northern pacific-side *and* gulf-side of BC, is now only pacific-side!!! <https://en.wikipedia.org/wiki/Municipalities_of_Baja_California>.  

## so for 2024, we should only consider 02007 and 02006 to be GoC.
```

```{r}
ce_byrgn <- ce_actividades_rgnid %>% 
  filter(is.na(id_estrato)) %>% 
  group_by(rgn_id, codigo, municipio) %>% 
  summarize(years_available = n()) # 5 is the most: 2004, 2009, 2014, 2019, 2024

ce_all_years_goc_marine_rgnid_check <- ce_all_years_goc_marine_rgnid %>% 
  filter(codigo %in% "1125") %>% 
  filter(municipio %in% "MEXICALI") ## 99 is the same as NA here....

ce_all_years_goc_marine_rgnid_check <- ce_all_years_goc_marine_rgnid %>% 
  filter(codigo %in% "1141") %>% 
  filter(municipio %in% "LOS CABOS") ## 99 is the same as NA here....

colnames(ce_actividades_rgnid)
#  [1] "codigo"                  "nivel"                  
#  [3] "specific_sector" "notas_breve"            
#  [5] "e03"                     "e04"                    
#  [7] "ano_censal"              "ue"                     
#  [9] "h000a"                   "j300a"                  
# [11] "j400a"                   "m000a"                  
# [13] "a800a"                   "id_estrato"             
# [15] "estado"                  "municipio"              
# [17] "rgn_id"                  "directly_coastal"     

unique(ce_actividades_rgnid$id_estrato)
# [1] NA 99  1  2  3  4

ce_all_years_goc_marine_rgnid_check <- ce_all_years_goc_marine_rgnid %>% 
  filter(id_estrato %in% "99")
```

Okay.  

| id_estrato | Meaning in practice                                                                            |
| ---------- | ---------------------------------------------------------------------------------------------- |
| 1-4        | Establishment size strata                                                                      |
| NA         | “Suma de Estratos” (explicit total)                                                        |
| 99         | “Agrupados por confidencialidad” (sometimes behaves like total, sometimes partial aggregation) |

So, for each group, we need to determine whether a reliable total exists....

If yes, we can just use that row.
If not, we need to reconstruct the total from strata 1-4.

```{r}
check_structure <- ce_actividades_rgnid %>%
  group_by(ano_censal, municipio, codigo) %>%
  summarize(
    n_rows = n(),
    has_total_na = any(is.na(id_estrato)),
    has_99 = any(id_estrato == 99),
    has_strata = any(id_estrato %in% 1:4),
    .groups = "drop"
  )
```

*Okay, so we can use this decision hierarchy:*

If NA total exists, use that row.
Else if only one row and it's 99, treat as total.
Else if strata 1-4 exists, sum them.
Else if only 99 rows (multiple this time) exist, treat 99 as total.
Else... flag as having an issue.

So we only have one row per group!

*NOTE: Why 99 Sometimes Equals NA*

In INEGI Economic Census, if there is no confidentiality issue, it seems that they publish strata and the total. But, if there is a confidentiality risk, they suppress strata and replace with:

* Either 99 only
* Or 99 & partial aggregation, it seems like!

So 99 is not the “total”, but rather the “confidential grouping” they use. In our case, it behaves like a total only when strata are absent.

*Important: Do NOT Mix 99 With 1-4....* if 1-4 exist, 99 is almost always a collapsed group of suppressed strata, and is therefore NOT additive with the others.

SUMMARY: Use the total if provided, otherwise add up the strata, and never mix confidentiality strata with true strata!!

```{r}
cols_to_sum <- c("ue", "h000a", "j300a", "j400a", "m000a", "a800a")

ce_totals_clean <- ce_actividades_rgnid %>%
  group_by(ano_censal, municipio, codigo) %>% # group by census year, municipality, and activity code.this ensures we resolve strata into a single total per group
  
  group_modify(~{

    # .x contains the current group variation.  So for each group (one year x municipio x codigo)...
    df <- .x 

    # -------------------------------------------------------------
    # case 1: explicit total row exists (id_estrato is NA)
    # this is the official "suma de estratos" that is the total. if we have this, it is the safest and most complete total, so we return only that row
    # -------------------------------------------------------------
    if(any(is.na(df$id_estrato))) {
      return(df %>% filter(is.na(id_estrato)))
    }

    # -------------------------------------------------------------
    # case 2: only one row exists and it is coded as 99
    # based on the tc_estrato_ce2024 and all methodology documents for the other years, this means strata were suppressed for confidentiality. And in this case, since there are no other strata rows, we can treat 99 as the total.
    # -------------------------------------------------------------
    if(nrow(df) == 1 & df$id_estrato == 99) {
      return(df)
    }

    # -------------------------------------------------------------
    # case 3: establishment size strata (1 to 4) are present
    # this means no explicit total was provided, so we have to estimate the total by summing across strata 1 - 4
    # NOTE we do NOT include 99 here to avoid double counting!!!
    # -------------------------------------------------------------
    if(any(df$id_estrato %in% 1:4)) {

      # sum numeric variables across strata rows
      summed <- df %>%
        filter(id_estrato %in% 1:4) %>%
        summarize(across(all_of(cols_to_sum), ~sum(.x, na.rm = TRUE))) 

      # keep identifying metadata from the first row (these are constant within the group)
      metadata <- df %>% 
        slice(1) %>%
        dplyr::select(ano_censal, municipio, codigo, estado, rgn_id, directly_coastal)

      # combine metadata with summed totals
      return(bind_cols(metadata, summed))
    }

    # -------------------------------------------------------------
    # case 4: multiple rows exist but *ALL* are coded 99
    # this indicates confidentiality grouping only, and since no strata exist, treat one 99 row as the total.  Very similar to case 2 but we have to take one row out of the many for that (year x municipio x codigo) grouping...
    # -------------------------------------------------------------
    if(all(df$id_estrato == 99)) {
      return(df %>% slice(1))
    }

    # -------------------------------------------------------------
    # else/case 5: if the grouping does not fit into any of these cases, then return the first row to ensure one row per group.... these NEED to be investigated.
    # -------------------------------------------------------------
    return(df %>% slice(1))
  }) %>%
  ungroup()


### check that each group has only one row!! should equal 1:
check_clean <- ce_totals_clean %>%
  group_by(ano_censal, municipio, codigo) %>%
  tally()
unique(check_clean$n) # 1 !! good!!

check_structure_2 <- ce_totals_clean %>%
  group_by(ano_censal, municipio, codigo) %>%
  summarize(
    n_rows = n(),
    has_total_na = any(is.na(id_estrato)),
    has_99 = any(id_estrato == 99),
    has_strata = any(id_estrato %in% 1:4),
    .groups = "drop"
  ) ## now we do not have any 99s in there.... 

missing_by_year <- ce_totals_clean %>%
  group_by(ano_censal) %>%
  summarize(
    n_obs = n(),
    pct_missing_jobs = mean(is.na(h000a)) * 100,
    pct_missing_ue = mean(is.na(ue)) * 100,
    pct_missing_revenue = mean(is.na(a800a)) * 100
  )

missing_by_region <- ce_totals_clean %>%
  group_by(rgn_id) %>%
  summarize(
    n_obs = n(),
    pct_missing_jobs = mean(is.na(h000a)) * 100
  ) %>%
  arrange(desc(pct_missing_jobs))

missing_rgn_year_act <- ce_totals_clean %>%
  group_by(rgn_id, ano_censal, codigo) %>%
  summarize(
    n_municipios = n(),
    n_missing_jobs = sum(is.na(h000a)),
    pct_missing_jobs = mean(is.na(h000a)) * 100,
    n_missing_ue = sum(is.na(ue)),
    pct_missing_ue = mean(is.na(ue)) * 100,
    .groups = "drop"
  ) %>%
  arrange(desc(pct_missing_jobs)) ### note that 2019 and 2024, since they are newer and did not have the summarized table of all previous years (as we had for 2004 - 2014), have many more suppressed observations... 
```

### Fixing the years

In the Economic Census year 2004, it represents data from 2003.  For example, the ano_censal = 2019 actually is referring to the year 2018.  From here on out, I will adjust the years to represent the true data.

```{r}
ce_totals_clean_years_adjusted <- ce_totals_clean %>% 
  mutate(ano_censal = case_when(
    ano_censal == 2004 ~ 2003,
    ano_censal == 2009 ~ 2008,
    ano_censal == 2014 ~ 2013,
    ano_censal == 2019 ~ 2018,
    ano_censal == 2024 ~ 2023,
    TRUE ~ NA
  ))
```



## Save to Mazu

```{r}
## save the filtered dataframe of all years marine activities with region ids that only have one row per group.
write_csv(ce_totals_clean_years_adjusted, here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/ce_int","ce_cleaned_totals_per_yr_municipio_activity.csv"))
```


We can now use this cleaned df for LIV_number_jobs, LIV_quality_jobs, and ECO_revenue.  Continue on to script 2-5.



#### Exploration

Looking at fishing over time:

```{r}
library(plotly)

ce_totals_clean_pesca <- ce_totals_clean_years_adjusted %>% 
  filter(codigo %in% "1141")

## plot 
num_jobs_pesca_plotly <- plotly::plot_ly(
  ce_totals_clean_pesca, 
  x = ~ano_censal, 
  y = ~h000a, 
  color = ~municipio, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "Number of fishers",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "H000A")
  ) 
num_jobs_pesca_plotly
```

Totals if aggregated by region:

```{r}
ce_totals_all_activities_by_rgn <- ce_totals_clean_years_adjusted %>% 
  group_by(rgn_id, ano_censal) %>% 
  summarize(total_jobs = sum(h000a, na.rm = TRUE),
            total_ue = sum(ue, na.rm = TRUE),
            total_revenue = sum(a800a, na.rm = TRUE)) %>% 
  ungroup()

# num jobs
num_jobs_all_plotly <- plotly::plot_ly(
  ce_totals_all_activities_by_rgn, 
  x = ~ano_censal, 
  y = ~total_jobs, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "Number of marine-related jobs per region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "H000A")
  ) 
num_jobs_all_plotly

# ue total
ue_all_plotly <- plotly::plot_ly(
  ce_totals_all_activities_by_rgn, 
  x = ~ano_censal, 
  y = ~total_ue, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "Number of economic units",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = " ")
  ) 
ue_all_plotly

# rev total
rev_all_plotly <- plotly::plot_ly(
  ce_totals_all_activities_by_rgn, 
  x = ~ano_censal, 
  y = ~total_revenue, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "Sum of revenue over time",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = " ")
  ) 
rev_all_plotly
```

```{r}
library(plotly)
library(dplyr)

# Aggregate jobs by general_sector (or specific_sector) and year
ce_by_sector_year <- ce_totals_clean_years_adjusted %>%
  filter(!is.na(h000a)) %>%  # exclude suppressed jobs
  group_by(general_sector, ano_censal) %>%
  summarise(
    total_jobs = sum(h000a, na.rm = TRUE),
    n_ue = sum(ue, na.rm = TRUE),
    total_revenue = sum(a800a, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(general_sector, ano_censal)

# plot 1: Jobs by general_sector over time
plot_jobs_by_sector <- plot_ly(
  ce_by_sector_year,
  x = ~ano_censal,
  y = ~total_jobs,
  color = ~general_sector,
  type = "scatter",
  mode = "lines+markers",
  text = ~paste("Jobs:", scales::comma(total_jobs), "<br>UE:", n_ue),
  hovertemplate = "%{text}<extra></extra>"
) %>%
  layout(
    title = "Marine Jobs by Sector Over Time",
    xaxis = list(title = "Census Year"),
    yaxis = list(title = "Total Jobs (H000A)", type = "log"),  # log scale to see small sectors
    legend = list(title = list(text = "General Sector"))
  )

plot_jobs_by_sector

# plot 2: UE by general_sector over time  
plot_ue_by_sector <- plot_ly(
  ce_by_sector_year,
  x = ~ano_censal,
  y = ~n_ue,
  color = ~general_sector,
  type = "scatter",
  mode = "lines+markers",
  text = ~paste("UE:", scales::comma(n_ue), "<br>Jobs:", scales::comma(total_jobs)),
  hovertemplate = "%{text}<extra></extra>"
) %>%
  layout(
    title = "Marine Economic Units by Sector Over Time",
    xaxis = list(title = "Census Year"),
    yaxis = list(title = "Economic Units"),
    legend = list(title = list(text = "General Sector"))
  )

plot_ue_by_sector

# plot 3: Revenue by general_sector over time
plot_revenue_by_sector <- plot_ly(
  ce_by_sector_year,
  x = ~ano_censal,
  y = ~total_revenue,
  color = ~general_sector,
  type = "scatter",
  mode = "lines+markers",
  text = ~paste("Revenue:", scales::comma(total_revenue), "<br>Jobs:", scales::comma(total_jobs)),
  hovertemplate = "%{text}<extra></extra>"
) %>%
  layout(
    title = "Marine Revenue by Sector Over Time (millions MXN)",
    xaxis = list(title = "Census Year"),
    yaxis = list(title = "Total Revenue (A800A)", type = "log"),
    legend = list(title = list(text = "General Sector"))
  )

plot_revenue_by_sector

```

```{r}

# agg jobs by general_sector (or specific_sector) and year
ce_by_sector_year <- ce_totals_clean_years_adjusted %>%
  group_by(specific_sector, ano_censal) %>%
  summarise(
    total_jobs = sum(h000a, na.rm = TRUE),
    n_ue = sum(ue, na.rm = TRUE),
    total_revenue = sum(m000a, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(specific_sector, ano_censal)

# plot 1: Jobs by general_sector over time
plot_jobs_by_sector <- plot_ly(
  ce_by_sector_year,
  x = ~ano_censal,
  y = ~total_jobs,
  color = ~specific_sector,
  type = "scatter",
  mode = "lines+markers",
  text = ~paste("Jobs:", scales::comma(total_jobs), "<br>UE:", n_ue),
  hovertemplate = "%{text}<extra></extra>"
) %>%
  layout(
    title = "Marine Jobs by Sector Over Time",
    xaxis = list(title = "Census Year"),
    yaxis = list(title = "Total Jobs (H000A)", type = "log"),  # log scale to see small sectors
    legend = list(title = list(text = "Specific Sector"))
  )

plot_jobs_by_sector

# plot 3: Revenue by general_sector over time
plot_revenue_by_sector <- plot_ly(
  ce_by_sector_year,
  x = ~ano_censal,
  y = ~total_revenue,
  color = ~specific_sector,
  type = "scatter",
  mode = "lines+markers",
  text = ~paste("Revenue:", scales::comma(total_revenue), "<br>Jobs:", scales::comma(total_jobs)),
  hovertemplate = "%{text}<extra></extra>"
) %>%
  layout(
    title = "Marine Revenue by Sector Over Time (millions MXN)",
    xaxis = list(title = "Census Year"),
    yaxis = list(title = "Total Revenue (M000A)", type = "log"),
    legend = list(title = list(text = "Specific Sector"))
  )

plot_revenue_by_sector
```

Looking at specific sectors over time:
```{r}
sectors_of_interest <- c(
  "fishing",
  "mariculture",
  "seafood_processing_trade",
  "port_dock_construction_management",
  "boat_building",
  "tidal_energy",
  "oil_gas_maritime_transport",
  "cargo_goods_maritime_transport",
  "marine_tourist_transport",
  "coastal_recreation",
  "travel_planners",
  "accommodations_food_drink",
  "environmental_research_consulting"
)

# filter the summarized data for only these sectors
ce_by_sector_year_filtered <- ce_totals_clean_years_adjusted %>%
  group_by(specific_sector, ano_censal, rgn_id) %>%
  summarise(
    total_jobs = sum(h000a, na.rm = TRUE),
    n_ue = sum(ue, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(specific_sector %in% sectors_of_interest)

library(plotly)
library(dplyr)
library(scales)

# lists to store plots
plot_jobs_list  <- list()

for (sector in sectors_of_interest) {
  df_sector <- ce_by_sector_year_filtered %>%
    filter(specific_sector == sector)

  if (nrow(df_sector) == 0) next  # skip if sector is missing

  # plot 1: jobs for this sector
  p_jobs <- plot_ly(
    df_sector,
    x = ~ano_censal,
    y = ~total_jobs,
    color = ~rgn_id,
    colors = rgn_colors,
    type = "scatter",
    mode = "lines+markers",
    text = ~paste("Jobs:", comma(total_jobs), "<br>UE:", n_ue),
    hovertemplate = "%{text}<extra></extra>"
  ) %>%
    layout(
      title = paste("Marine-related JOBS:", sector),
      xaxis = list(title = "Census Year"),
      yaxis = list(title = "Total Jobs (H000A)")
    )

  # store plots
  plot_jobs_list[[sector]]   <- p_jobs
}

plot_jobs_list[["fishing"]]
plot_jobs_list[["mariculture"]]
plot_jobs_list[["seafood_processing_trade"]]
plot_jobs_list[["port_dock_construction_management"]]
plot_jobs_list[["boat_building"]]
plot_jobs_list[["oil_gas_maritime_transport"]]
plot_jobs_list[["cargo_goods_maritime_transport"]]
plot_jobs_list[["marine_tourist_transport"]]
plot_jobs_list[["coastal_recreation"]]
plot_jobs_list[["travel_planners"]]
plot_jobs_list[["accommodations_food_drink"]]
plot_jobs_list[["environmental_research_consulting"]]

```

