---
title: 'OHI GoC 2025: LE Current Status Visualizations'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor:
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options:
  chunk_output_type: console
---

# LE visualizations

I ran this script after `script5_merge_liv_eco.qmd` and any LE sustainability prep script. I read the saved current status outputs and built charts with consistent region colors and labels across the merged LIV subgoal, LIV number of jobs, LIV quality of jobs, ECO revenue, sustainability, and the overall LE score.

```{r}
library(tidyverse)
library(ggplot2)
library(plotly)
library(sf)
library(mapview)
library(here)
library(rnaturalearth)
```

```{r}
rgn_labels <- c(
  "Region_1" = "Western Upper Gulf",
  "Region_2" = "Western Midriff Islands",
  "Region_3" = "Western Central Gulf",
  "Region_4" = "Western Southern Gulf",
  "Region_5" = "Eastern Upper Gulf",
  "Region_6" = "Eastern Midriff Islands",
  "Region_7" = "Southern Sonora Coast",
  "Region_8" = "Sinaloa Coast",
  "Region_9" = "Southeastern Gulf"
)

rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

region_levels <- paste0("Region_", 1:9)
rgn_label_colors <- setNames(unname(rgn_colors[region_levels]), unname(rgn_labels[region_levels]))
census_years <- c(2003, 2008, 2013, 2018, 2023)
component_levels <- c(
  "LE:LIV: Number of Jobs",
  "LE:LIV: Quality of Jobs",
  "LE:LIV",
  "LE:ECO",
  "Overall LE"
)

add_rgn_labels <- function(df) {
  df %>%
    mutate(
      rgn_id = factor(as.character(rgn_id), levels = region_levels),
      rgn_label = factor(
        rgn_labels[as.character(rgn_id)],
        levels = unname(rgn_labels[region_levels])
      )
    )
}

le_plotly_lines <- function(df, y_var, title, y_range = NULL) {
  yr <- range(df$year, na.rm = TRUE)

  plot_ly(
    df,
    x = ~year,
    y = as.formula(paste0("~", y_var)),
    color = ~rgn_label,
    colors = rgn_label_colors,
    type = "scatter",
    mode = "lines+markers"
  ) %>%
    layout(
      title = title,
      xaxis = list(title = "Año censal // Census Year"),
      yaxis = list(title = "Puntaje // Percent (out of 100)", range = y_range),
      legend = list(title = list(text = "Region"))
    ) %>%
    add_lines(
      x = yr,
      y = c(100, 100),
      line = list(color = "black", width = 2, dash = "dash"),
      name = "Meta // Goal (100/100)",
      inherit = FALSE
    )
}

le_plotly_goc <- function(df, y_var, title, y_range = NULL) {
  yr <- range(df$year, na.rm = TRUE)

  plot_ly(
    df,
    x = ~year,
    y = as.formula(paste0("~", y_var)),
    type = "scatter",
    mode = "lines+markers",
    name = "GOC Average"
  ) %>%
    layout(
      title = title,
      xaxis = list(title = "Año censal // Census Year"),
      yaxis = list(title = "Puntaje // Percent (out of 100)", range = y_range)
    ) %>%
    add_lines(
      x = yr,
      y = c(100, 100),
      line = list(color = "black", width = 2, dash = "dash"),
      name = "Meta // Goal (100/100)",
      inherit = FALSE
    )
}
```

## Read outputs

```{r}
cs_jobs_raw <- read_csv(
  here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/liv", "ce_jobs_current_status.csv"),
  show_col_types = FALSE
) %>%
  rename(year = ce_ano_censal) %>%
  mutate(year = as.numeric(year))

cs_qual_raw <- read_csv(
  here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/liv", "ce_qual_current_status.csv"),
  show_col_types = FALSE
) %>%
  rename(year = ano_censal) %>%
  mutate(year = as.numeric(year))

cs_rev_raw <- read_csv(
  here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/eco", "ce_rev_current_status.csv"),
  show_col_types = FALSE
) %>%
  rename(year = ano_censal) %>%
  mutate(year = as.numeric(year))

all_cs_indicators <- read_csv(
  here::here("/home/shares/ohi/OHI_GOC/goal_prep/le", "final_le_current_status.csv"),
  show_col_types = FALSE
) %>%
  rename(year = ano_censal, le_current_status = final_le_current_status_rgn) %>%
  mutate(year = as.numeric(year))

cs_jobs_clean <- cs_jobs_raw %>%
  rename(score = current_status) %>%
  dplyr::select(rgn_id, year, score)

cs_qual_clean <- cs_qual_raw %>%
  rename(score = current_status) %>%
  dplyr::select(rgn_id, year, score)

cs_eco_clean <- cs_rev_raw %>%
  rename(score = current_status) %>%
  dplyr::select(rgn_id, year, score)

cs_liv_clean <- full_join(
  cs_jobs_raw %>%
    dplyr::select(rgn_id, year, job_current_status = current_status),
  cs_qual_raw %>%
    dplyr::select(rgn_id, year, qual_current_status = current_status)
) %>%
  mutate(
    score = (job_current_status + qual_current_status) / 2,
    score = ifelse(rgn_id == "Region_2", qual_current_status, score)
  ) %>%
  dplyr::select(rgn_id, year, score)

le_components_long <- bind_rows(
  add_rgn_labels(cs_jobs_clean) %>%
    transmute(rgn_id, rgn_label, year, score, component = "LE:LIV: Number of Jobs"),
  add_rgn_labels(cs_qual_clean) %>%
    transmute(rgn_id, rgn_label, year, score, component = "LE:LIV: Quality of Jobs"),
  add_rgn_labels(cs_liv_clean) %>%
    transmute(rgn_id, rgn_label, year, score, component = "LE:LIV"),
  add_rgn_labels(cs_eco_clean) %>%
    transmute(rgn_id, rgn_label, year, score, component = "LE:ECO"),
  add_rgn_labels(all_cs_indicators) %>%
    transmute(rgn_id, rgn_label, year, score = le_current_status, component = "Overall LE")
) %>%
  filter(year %in% census_years) %>%
  mutate(component = factor(component, levels = component_levels))

le_2023_long <- le_components_long %>%
  filter(year == 2023)

all_cs_indicators_2023 <- all_cs_indicators %>%
  filter(year == 2023)

le_goc_avg <- all_cs_indicators %>%
  dplyr::select(year, le_goc_current_status = final_le_current_status_goc) %>%
  distinct()
```

## Interactive line charts (plotly)

```{r}
liv_plotly <- le_plotly_lines(
  add_rgn_labels(cs_liv_clean),
  y_var = "score",
  title = "LE:LIV (Number of Jobs & Quality of Jobs merged)"
)

jobs_plotly <- le_plotly_lines(
  add_rgn_labels(cs_jobs_clean),
  y_var = "score",
  title = "LE:LIV: Number of Jobs"
)

quality_plotly <- le_plotly_lines(
  add_rgn_labels(cs_qual_clean),
  y_var = "score",
  title = "LE:LIV: Quality of Jobs"
)

eco_plotly <- le_plotly_lines(
  add_rgn_labels(cs_eco_clean),
  y_var = "score",
  title = "LE:ECO"
)

le_score_plotly <- le_plotly_lines(
  add_rgn_labels(all_cs_indicators),
  y_var = "le_current_status",
  title = "CURRENT STATUS: Livelihoods & Economies"
)

liv_plotly
jobs_plotly
quality_plotly
eco_plotly
le_score_plotly
```

## Overall LE line charts (plotly)

I plotted the overall LE current status by region and the Gulf wide average across census years.

```{r}
le_overall_rgn_plotly <- le_plotly_lines(
  add_rgn_labels(all_cs_indicators),
  y_var = "le_current_status",
  title = "Overall LE current status by region",
  y_range = c(0, 110)
)

le_overall_goc_plotly <- le_plotly_goc(
  le_goc_avg,
  y_var = "le_goc_current_status",
  title = "Overall LE current status for the Gulf of California",
  y_range = c(0, 110)
)

le_overall_rgn_plotly
le_overall_goc_plotly
```

## All components over time (ggplot)

```{r}
gg_le_components_lines <- ggplot(
  le_components_long,
  aes(x = year, y = score, color = rgn_label, group = rgn_label)
) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.8) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey45", linewidth = 0.4) +
  facet_wrap(~ component, ncol = 2) +
  scale_color_manual(values = rgn_label_colors, name = "Region") +
  scale_x_continuous(breaks = census_years) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 25)) +
  labs(
    title = "LE current status by component (2003 to 2023)",
    x = "Census Year",
    y = "Score (out of 100)"
  ) +
  theme_minimal(base_family = "Calibri") +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 11),
    legend.position = "right"
  )

gg_le_components_lines
```

## 2023 snapshot by region (ggplot)

```{r}
gg_le_2023_bars <- ggplot(
  le_2023_long,
  aes(x = rgn_label, y = score, fill = rgn_label)
) +
  geom_col(width = 0.82, color = "white", linewidth = 0.3) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey45", linewidth = 0.4) +
  facet_wrap(~ component, ncol = 2) +
  scale_fill_manual(values = rgn_label_colors, guide = "none") +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 25)) +
  labs(
    title = "LE current status by region (2023)",
    x = NULL,
    y = "Score (out of 100)"
  ) +
  theme_minimal(base_family = "Calibri") +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 11),
    axis.text.x = element_text(angle = 35, hjust = 1, size = 8)
  )

gg_le_2023_bars
```

## Overall LE polar chart (2023)

```{r}
gg_le_2023_polar <- ggplot(
  le_2023_long %>% filter(component == "Overall LE"),
  aes(x = rgn_label, y = score, fill = rgn_label)
) +
  geom_col(width = 1, color = "white", linewidth = 0.35) +
  geom_hline(yintercept = 100, color = "grey55", linewidth = 0.5) +
  coord_polar(start = -pi / 2, clip = "off") +
  scale_fill_manual(values = rgn_label_colors, guide = "none") +
  scale_y_continuous(limits = c(0, 100), breaks = c(25, 50, 75, 100)) +
  labs(
    title = "Overall LE current status by region (2023)",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_family = "Calibri") +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    panel.grid.major.x = element_blank()
  )

gg_le_2023_polar
```

## Map visualization (2023 overall LE)

```{r}
ohi <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp"), quiet = TRUE) %>%
  filter(location == "marine") %>%
  dplyr::select(rgn_id) %>% 
  mutate(rgn_labels = case_when(
    rgn_id %in% "Region_1" ~ "Western Upper Gulf",
    rgn_id %in% "Region_2" ~ "Western Midriff Islands",
    rgn_id %in% "Region_3" ~ "Western Central Gulf",
    rgn_id %in% "Region_4" ~ "Western Southern Gulf",
    rgn_id %in% "Region_5" ~ "Eastern Upper Gulf",
    rgn_id %in% "Region_6" ~ "Eastern Midriff Islands",
    rgn_id %in% "Region_7" ~ "Southern Sonora Coast",
    rgn_id %in% "Region_8" ~ "Sinaloa Coast",
    rgn_id %in% "Region_9" ~ "Southeastern Gulf",
    TRUE ~ NA
  ))

all_cs_indicators_2023_ohigns <- full_join(all_cs_indicators_2023, ohi, by = "rgn_id") %>%
  st_as_sf()
```

```{r}
rgn_labels <- c(
  "Region_1" = "Western Upper Gulf",
  "Region_2" = "Western Midriff Islands",
  "Region_3" = "Western Central Gulf",
  "Region_4" = "Western Southern Gulf",
  "Region_5" = "Eastern Upper Gulf",
  "Region_6" = "Eastern Midriff Islands",
  "Region_7" = "Southern Sonora Coast",
  "Region_8" = "Sinaloa Coast",
  "Region_9" = "Southeastern Gulf"
)

rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

mapview(ohi, zcol = "rgn_labels", col.regions = rgn_colors, layer.name = "OHI GoC Regions")

```


```{r}
mapview(
  all_cs_indicators_2023_ohigns,
  zcol = "le_current_status",
  col.regions = colorRampPalette(c("#FEE090", "#ABD9E9", "#4575B4", "#313695")),
  layer.name = "Livelihoods & Economies Current Status Scores (2023)",
  legend = TRUE
)

le_at <- seq(0, 100, by = 10)

le_cols <- colorRampPalette(c(
  "#8a0032", "#A50026", "#A50026", "#D73027", "#F46D43",
  "#FDAE61", "#FEE090", "#FEE090", "#ABD9E9", "#4575B4", "#313695"
))(length(le_at) - 1)

mapview(
  all_cs_indicators_2023_ohigns,
  zcol = "le_current_status",
  col.regions = le_cols,
  at = le_at,
  layer.name = "Livelihoods & Economies Current Status Scores (2023)",
  legend = TRUE
)
```

## Choropleth map (ggplot)

I drew the Gulf states background first, then the marine regions on top.

```{r}
le_palette <- c(
  "#8a0032", "#A50026", "#A50026", "#D73027", "#F46D43",
  "#FDAE61", "#FEE090", "#FEE090", "#ABD9E9", "#4575B4", "#313695"
)

states <- st_read(here::here("spatial/ohi_regions/states.shp"), quiet = TRUE) %>%
  st_transform(st_crs(all_cs_indicators_2023_ohigns))

gg_le <- ggplot() +
  geom_sf(
    data = states,
    fill = "#d6cac0",
    color = "grey10",
    linewidth = 0.1
  ) +
  geom_sf(
    data = all_cs_indicators_2023_ohigns,
    aes(fill = le_current_status),
    color = "black",
    linewidth = 0.2
  ) +
  scale_fill_gradientn(
    colours = le_palette,
    limits = c(0, 100),
    na.value = "grey85",
    name = "Livelihoods & Economies\nCurrent Status 2023"
  ) +
  coord_sf(expand = FALSE) +
  theme_void() +
  theme(
    text = element_text(family = "Calibri"),
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    legend.position = "right",
    legend.key.height = unit(0.8, "cm"),
    legend.key.width = unit(0.5, "cm"),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    plot.margin = margin(10, 10, 10, 10)
  )

gg_le
```
