################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

original author: "Austin Martinez"

################################################################################

# PRE-WORK ------------------------------

```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Load in Libraries
library(sf)
library(dplyr)
library(here)
library(terra)
library(tidyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(tmap)

```


```{r Load in the OHI Region_Key0}
Region_Key0<- read.csv("/home/ferrer/ohi_goc_prep/FP/v2026/CONAPESCA_Keys/Spatial_Keys/RegionKey_FINAL_May262026.csv")

colnames(Region_Key0)

Region_Key<- Region_Key0 %>%
  select(rgn_id= OHI_region_est, rgn_name= OHI_region_name) %>%
  distinct()

# Make uppercase:
Region_Key$rgn_id <- sub("^r", "R", Region_Key$rgn_id)
```

```{r Read in OHI region boundaries}
# Read in OHI region boundaries
ohi_regions <- st_read(here("spatial", "ohi_regions", "goc_ohi_rgns.shp")) %>% 
  st_make_valid()

```

# STEP 1: CALCULATE CORAL REEF "CURRENT EXTENT" (sources: Allen Coral Atlas [2020], dataMares [2016], Hem / CEDO coral data [2025]) ------------------------------
Extent = Total area from dataMares - area from Allen Coral Atlas, with remainder split across all points identified in CEDO data.

```{r Read in Coral Reef extent data sets}
# Read in Allen polygon data (2018-2020):
AllenAtlas_coral_raw <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/coral_datasets/ALLEN_coral_Atlas/Mexican-Exclusive-Economic-Zone-20230309211010/Reef-Extent/reefextent.gpkg") %>% 
  st_make_valid()

# Read in CEDO point data (2025):
CEDO_coral_raw <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/Hem_data/goc_habitat_data/coral/Arrecifes_de_coral_(Golfo_de_California).shp") %>% 
  st_make_valid()

# Total area of corals in the GOC according to dataMares (2016): https://datamares.org/stories/marine-habitat-distributions-in-the-gulf-of-california-2/
CORL_area_dataMares_ha <- 657
```

```{r Match CRS to OHI regions}
# Match CRS to OHI regions
AllenAtlas_coral.data <- st_transform(AllenAtlas_coral_raw, st_crs(ohi_regions))
CEDO_coral.data <- st_transform(CEDO_coral_raw, st_crs(ohi_regions))
```

```{r Clip data to the OHI boundies extent}
# Clip data to the OHI boundies extent
AllenAtlas_coral.data <- st_intersection(AllenAtlas_coral.data, ohi_regions)
CEDO_coral.data   <- st_intersection(CEDO_coral.data, ohi_regions)
```

```{r Identify points from the CEDO data set that do / do not intersect the Allen Atlas polygons}
# Identify points from the CEDO data set that intersect the Allen Atlas polygons
corals_that_intersect <- st_intersects(CEDO_coral.data, AllenAtlas_coral.data)

# Keep only the CEDO points that do NOT intersect Allen Atlas polygons
corals_that_do_not_intersect <- CEDO_coral.data[lengths(corals_that_intersect) == 0,]

# Find how many points don't intersect a polygon:
non_intersect_coral_points <- as.numeric(nrow(corals_that_do_not_intersect))
non_intersect_coral_points
# n = 47 points

```

```{r Convert from m^2 to hectares}
CORL_area_AllenAtlas_m2 <- sum(st_area(AllenAtlas_coral.data))
CORL_area_AllenAtlas_ha <- as.numeric(CORL_area_AllenAtlas_m2) / 10000
CORL_area_AllenAtlas_ha
```

```{r Find remainder and assign area to non intersecting points}
# Subtract total coral area of CORL_area_AllenAtlas_ha from CORL_area_dataMares_ha
remaining_coral_area_ha <- CORL_area_dataMares_ha - CORL_area_AllenAtlas_ha
remaining_coral_area_ha

# Find how much area each remaning points should be assigned
area_per_point_ha = remaining_coral_area_ha / non_intersect_coral_points
area_per_point_ha

# NOTE TO SELF: 2.6, slightly different than Austin's 2.7 because of a CRS matching change. - Oct 7 2026
```

```{r Construct COMBINED coral reef dataset}

# Rename AllenAtlas_coral.data$geom → geometry
AllenAtlas_coral.data <- AllenAtlas_coral.data %>%
  rename(geometry = geom)

# Drop the unwanted columns to concat corals_that_do_not_intersect and AllenAtlas_coral.data
AllenAtlas_coral.data <- AllenAtlas_coral.data %>%
  select(-class)

corals_that_do_not_intersect <- corals_that_do_not_intersect %>%
  select(-Habitat)

# Add an area_ha column For polygons (AllenAtlas_coral.data) and For points (corals_that_do_not_intersect)
AllenAtlas_coral.data <- AllenAtlas_coral.data %>%
  mutate(area_ha = as.numeric(st_area(geometry)) / 10000)

corals_that_do_not_intersect <- corals_that_do_not_intersect %>%
  mutate(area_ha = area_per_point_ha)

# Bind the two datasets together
COMBINED_coral.data_area <- bind_rows(AllenAtlas_coral.data, corals_that_do_not_intersect)
```

```{r SANITY CHECK on total area}
# Check if it worked: All areas should sum to 657 ha (dataMares estimate)
sum(COMBINED_coral.data_area$area_ha)
```

```{r Assign CORAL extent to each OHI region}
# Assign the coral extent to each OHI region:
CORL_Current_Extent <- COMBINED_coral.data_area %>%
  st_drop_geometry() %>%
  dplyr::group_by(rgn_id) %>%
  dplyr::summarise(
    area_ha = sum(area_ha, na.rm = TRUE),
    year = 2025
  )

all_rgns <- tibble(
  rgn_id = paste0("Region_", 1:9),
  year = 2025
)

CORL_Current_Extent <- all_rgns %>%
  left_join(CORL_Current_Extent, by = c("rgn_id", "year")) %>%
  mutate(area_ha = replace_na(area_ha, 0)) %>%
  select(rgn_id, 
         area_est_2025_ha = area_ha, 
         year)

CORL_Current_Extent

```

```{r Ktable of Current Extent}
CORL_Current_Extent %>%
  kable(digits = 4, col.names = c("Region", "Extent (ha) (Allen Atlas data, 2020; CEDO data, 2025; dataMares, 2017)", "Year")) %>%
  kable_styling(bootstrap_options = c("striped", "bordered")) %>%
  add_header_above(
    c('Coral Reef Extent' = 3))

```

## Save CORL_Current_Extent
```{r Save CORL_Current_Extent}
write.csv(
  CORL_Current_Extent,
  here::here("_habitats/HAB_corals/CRL_outputs/CORL_Current_Extent_Oct072026.csv"),
  row.names = FALSE
)
```

```{r Map of CORL_Current_Extent}
tmap_mode("plot")

# Add dummy habitat columns for legend control
AllenAtlas_coral.data$habitat <- "Coral Reef"
CEDO_coral.data$habitat   <- "Coral Reef"

tm_shape(ohi_regions) +
  tm_polygons(
    col = "location",
    palette = c(
      "marine" = "#cceeff",   # light blue
      "land"   = "#f5e6c8"    # light tan
    ),
    alpha = 0.5,
    border.col = "gray",
    border.lwd = 0.6,
    legend.show = FALSE
  ) +

# Allen Coral polygons
tm_shape(AllenAtlas_coral.data) +
  tm_polygons(
    col = "habitat",
    palette = c("Allen Coral Polygons" = "#FF4500"),  # orange-red
    alpha = 1,
    border.col = "#FF4500",
    border.lwd = 2.5,
    title = "Habitat"
  ) +

# Hem Coral points
tm_shape(CEDO_coral.data) +
  tm_dots(
    col = "habitat",
    palette = c("Hem Coral Points" = "#FF4500"),       # blue
    size = 0.05,
    alpha = 1,
    title = "Habitat"
  ) +

tm_layout(
    title.size = 1.3,
    title.fontface = "bold",
    title.position = c(0.02, 0.98),
    frame = FALSE
  ) +

tm_scale_bar(position = c("left", "bottom"))

```


# STEP 2: CALCULATE CORAL REEF "CURRENT STATUS", i.e. condition over time (source: literature review from 1980s-2024)  ------------------------------

```{r Load in extent data as needed}
# Extent data from 3 data sources:
CORL_Current_Extent <- read_csv(here::here("_habitats/HAB_corals/CRL_outputs/CORL_Current_Extent_Oct072026.csv"))
# 9 obs., 3 vars.
```

```{r Load in Coral Percent Cover data}
# Percent cover data from literature review:
CORL_percent.cover_data <- read_csv("/home/shares/ohi/OHI_GOC/_raw_data/coral_datasets/Coral_cover_lit_review/coral_reef_timeseries_averaged.csv")
```

## PLOT Region 4 Time series (all sites)

```{r Filter by region to create figures}
# Filter for region 4 data
PercentCover_rgn4 <- CORL_percent.cover_data %>% 
  filter(rgn_id == "Region_4")

# Filter for region 9 data
PercentCover_rgn9 <- CORL_percent.cover_data %>% 
  filter(rgn_id == "Region_9")
```

```{r}
# Make time series for region 4 data
ggplot(PercentCover_rgn4, aes(x = year, y = percent_coral_cover, color = reef_name)) +
  geom_line() +
  geom_point() +
  theme_bw() +
  labs(
    x = "Year",
    y = "% Coral Cover",
    color = "Reef",
    title = "Time series of coral cover at common sites (Region 4)"
  )

```

```{r}
# make sure year is numeric and factors are set
PercentCover_rgn4 <- PercentCover_rgn4 %>%
  mutate(
    year = as.numeric(year),
    reef_name = factor(reef_name),
    rgn_id = factor(rgn_id)
  )
```

```{r}
# Filter Region 4 for Cabo Pulmo only
cabo <- PercentCover_rgn4 %>% 
  filter(reef_name == "Cabo Pulmo")

# Split into two time periods
cabo_1991_2012 <- cabo %>% filter(year >= 1991, year <= 2012)
cabo_2013_2023 <- cabo %>% filter(year >= 2013, year <= 2023)

# Calculate averages for each period
avg_1991_2012 <- mean(cabo_1991_2012$percent_coral_cover, na.rm = TRUE)
avg_2013_2023 <- mean(cabo_2013_2023$percent_coral_cover, na.rm = TRUE)

```

```{r}
# Plot the cabo pulmo time series with 2 periods

# Create a data frame for mean lines so they appear in the legend
mean_df <- data.frame(
  period = c("Mean 1991–2012", "Mean 2013–2023"),
  mean_value = c(avg_1991_2012, avg_2013_2023)
)

my_colors <- c(
  "1991–2012" = "blue",
  "2013–2023" = "orange",
  "Mean 1991–2012" = "blue",
  "Mean 2013–2023" = "orange"
)


ggplot() +
  # Time series 1991–2012
  geom_line(data = cabo_1991_2012,
            aes(x = year, y = percent_coral_cover, color = "1991–2012"),
            size = 1) +
  geom_point(data = cabo_1991_2012,
             aes(x = year, y = percent_coral_cover, color = "1991–2012"),
             size = 2) +
  
  # Time series 2013–2023
  geom_line(data = cabo_2013_2023,
            aes(x = year, y = percent_coral_cover, color = "2013–2023"),
            size = 1) +
  geom_point(data = cabo_2013_2023,
             aes(x = year, y = percent_coral_cover, color = "2013–2023"),
             size = 2) +
  
  # Mean lines
  geom_hline(data = mean_df,
             aes(yintercept = mean_value, color = period),
             linetype = "dashed", size = 0.8, alpha = 0.55) +
  
  # Mean value labels (left side)
  geom_text(data = mean_df,
            aes(x = min(cabo_1991_2012$year), y = mean_value,
                label = paste0(round(mean_value, 1), "%"), color = period),
            hjust = 1.8, size = 3.5, show.legend = FALSE) +
  
  scale_color_manual(values = my_colors) +
  
  coord_cartesian(clip = "off") +
  
  labs(
    x = "Year",
    y = "% Coral Cover",
    color = "Time Period",
    title = "Cabo Pulmo Coral Cover Trends: 1991–2012 vs. 2013–2023",
    caption = "Figure 1"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 11),
    plot.margin = margin(5.5, 5.5, 5.5, 40)
  )
```



# Region 9
```{r}
# Remove sites that only have 1 year of data
PercentCover_rgn9_clean <- PercentCover_rgn9 %>%
  filter(!reef_name %in% c(
    "Las Iglesias (Bahía Banderas)",
    "Yelapa (Bahía Banderas)",
    "El Paredón (Bahía Banderas)",
    "Caletitas (Bahía Banderas)"
  ))
```


```{r}
# Make time series for region 9 data
ggplot(PercentCover_rgn9_clean, aes(x = year, y = percent_coral_cover, color = reef_name)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept = 2013, linetype = "dashed", color = "grey50") +
  theme_bw() +
  labs(
    x = "Year",
    y = "% Coral Cover",
    color = "Reef",
    title = "Time Series of Percent Coral Cover in (Region 9)",
    caption = "Figure 2"
  )
```

## Curent Status Calculation
```{r}
# Region 4

# Filter reef data for Region 4 (Cabo Pulmo only)
PercentCover_rgn4_cabo <- PercentCover_rgn4 %>%
  filter(reef_name == "Cabo Pulmo")

# Split periods
reef4_pre  <- PercentCover_rgn4_cabo %>% filter(year < 2013)
reef4_post <- PercentCover_rgn4_cabo %>% filter(year >= 2013)

# Means
avg_1991_2012 <- mean(reef4_pre$percent_coral_cover, na.rm = TRUE)
avg_2013_2023 <- mean(reef4_post$percent_coral_cover, na.rm = TRUE)

# Condition score
region4_condition_score <- (avg_2013_2023 / avg_1991_2012)

# Build Region 4 row (NO hard‑coding area)
region4_condition <- tibble(
  rgn_id = "Region_4",
  year = 2025,
  current_status = region4_condition_score
) %>%
  left_join(corals_extent, by = c("rgn_id", "year")) %>%
  select(rgn_id, area_km2, year, current_status)
```



```{r}
# Region 9

PercentCover_rgn9_avg <- PercentCover_rgn9_clean %>%
  group_by(year) %>%
  summarise(
    percent_coral_cover = mean(percent_coral_cover, na.rm = TRUE),
    rgn_id = first(rgn_id)
  ) %>%
  arrange(year)

# Only include sites that have multiple observations
PercentCover_rgn9_two <- PercentCover_rgn9_clean %>%
  filter(reef_name %in% c(
    "Punta Mita (Bahía Banderas)",
    "Isla María Cleofas (Nayarit offshore)",
    "Islas Marietas"
  ))

# Make Region 9 reefs into a single site by averaging percent cover for years with multiple observations.
PercentCover_rgn9_avg <- PercentCover_rgn9_two %>%
  group_by(year) %>%
  summarise(
    percent_coral_cover = mean(percent_coral_cover, na.rm = TRUE),
    rgn_id = first(rgn_id)
  ) %>%
  arrange(year)


reef9_pre  <- PercentCover_rgn9_avg %>% filter(year < 2013)
reef9_post <- PercentCover_rgn9_avg %>% filter(year >= 2013)

mean_pre  <- mean(reef9_pre$percent_coral_cover, na.rm = TRUE)
mean_post <- mean(reef9_post$percent_coral_cover, na.rm = TRUE)

region9_condition_score <- (mean_post / mean_pre)

region9_condition <- tibble(
  rgn_id = "Region_9",
  year = 2025,
  current_status = region9_condition_score
) %>%
  left_join(corals_extent, by = c("rgn_id", "year")) %>%
  select(rgn_id, area_km2, year, current_status)
```

```{r}
# Combine region 4 and 9 data
coral_condition_combined <- bind_rows(
  region4_condition,
  region9_condition
)
```

```{r Make a table}

coral_condition_combined %>%
  select(rgn_id, area_km2, year, current_status) %>%
  kable(
    digits = 2,
    col.names = c(
      "Region ID",
      "Area (km²)",
      "Year",
      "Current Status (%)"
    ),
    align = c("c", "r", "c", "r")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "bordered", "hover"),
    full_width = FALSE
  ) %>%
  add_header_above(
    c("Coral Current Status (2025)" = 4)
  )

```

```{r}
# Select necessary columns
coral_condition_combined <- coral_condition_combined %>%
  select(rgn_id, year, current_status)

#coral_condition_combined <- read_csv(here::here("_habitats/HAB_corals/CRL_outputs/coral_current_status.csv"))

all_rgns <- tibble(
  rgn_id = paste0("Region_", 1:9),
  year = 2025
)

coral_condition <- all_rgns %>% 
  left_join(coral_condition_combined, by = c("rgn_id", "year")) %>%
  mutate(current_status = round(100*current_status, 0)) %>%
  select(rgn_id, year, current_status)

coral_condition

```

```{r}
# Current Status 
write.csv(
  coral_condition,
  here::here("_habitats/HAB_corals/CRL_outputs/coral_current_status.csv"),
  row.names = FALSE
)

## pressure
pressure <- coral_condition %>% 
  mutate(pressure = (100 - current_status)/100) %>%
  select(rgn_id, year, pressure)
  
write.csv(
  pressure,
  here::here("_habitats/HAB_corals/CRL_outputs/coral_pressure.csv"),
  row.names = FALSE
)

```






```{r}
# Split into two time periods
PercentCover_rgn9_two <- PercentCover_rgn9_clean %>%
  filter(reef_name %in% c(
    "Punta Mita (Bahía Banderas)",
    "Isla María Cleofas (Nayarit offshore)",
    "Islas Marietas"
  ))

reef_pre2013 <- PercentCover_rgn9_two %>% filter(year < 2013)
reef_post2013 <- PercentCover_rgn9_two %>% filter(year >= 2013)

# Calculate averages
mean_pre <- mean(reef_pre2013$percent_coral_cover, na.rm = TRUE)
mean_post <- mean(reef_post2013$percent_coral_cover, na.rm = TRUE)

# Data frame for mean lines
mean_df_9 <- data.frame(
  period = c("Mean 1991–2012", "Mean 2013–2024"),
  mean_value = c(mean_pre, mean_post)
)

# Colors
my_colors_9 <- c(
  "1991–2012" = "blue",
  "2013–2024" = "orange",
  "Mean 1991–2012" = "blue",
  "Mean 2013–2024" = "orange"
)

# Plot
ggplot() +
  # Time series 1991–2012
  geom_line(data = reef_pre2013,
            aes(x = year, y = percent_coral_cover, color = "1991–2012"),
            size = 1) +
  geom_point(data = reef_pre2013,
             aes(x = year, y = percent_coral_cover, color = "1991–2012"),
             size = 2) +
  
  # Time series 2013–2023
  geom_line(data = reef_post2013,
            aes(x = year, y = percent_coral_cover, color = "2013–2024"),
            size = 1) +
  geom_point(data = reef_post2013,
             aes(x = year, y = percent_coral_cover, color = "2013–2024"),
             size = 2) +
  
  # Mean lines
  geom_hline(data = mean_df_9,
             aes(yintercept = mean_value, color = period),
             linetype = "dashed", size = 0.8, alpha = 0.55) +
  
  # Mean value labels
  geom_text(data = mean_df_9,
            aes(x = min(reef_pre2013$year), y = mean_value,
                label = paste0(round(mean_value, 1), "%"), color = period),
            hjust = 2.7 , size = 3.5, show.legend = FALSE) +
  
  scale_color_manual(values = my_colors_9) +
  
  coord_cartesian(clip = "off") +
  
  labs(
    x = "Year",
    y = "% Coral Cover",
    color = "Time Period",
    title = "Region 9 Coral Cover Trends: 1991–2012 vs. 2013–2024",
    caption = "Figure 2"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 11),
    plot.margin = margin(5.5, 5.5, 5.5, 40)
  )

```

```{r}
# Filter to the two reefs
PercentCover_rgn9_two <- PercentCover_rgn9_clean %>%
  filter(reef_name %in% c(
    "Punta Mita (Bahía Banderas)",
    "Isla María Cleofas (Nayarit offshore)",
    "Islas Marietas"
  ))

# Extract original colors from your previous plot
orig_colors <- c(
  "Punta Mita (Bahía Banderas)" = "hotpink",      # pink from your plot
  "Isla María Cleofas (Nayarit offshore)" = "#7CAE00",   # olive green from your plot,
  "Islas Marietas" = "skyblue"
)

# Plot with same colors
ggplot(PercentCover_rgn9_two, aes(x = year, y = percent_coral_cover, color = reef_name)) +
  geom_line() +
  geom_point() +
  scale_color_manual(values = orig_colors) +
  geom_vline(xintercept = 2013, linetype = "dashed", color = "grey50") +
  theme_bw() +
  labs(
    x = "Year",
    y = "% Coral Cover",
    color = "Reef",
    title = "Time series of coral cover at common sites (Region 9)",
    caption = "figure 1"
  )

```

```{r}
# Prepare coral condition data for join
coral_condition_map <- coral_condition_combined %>%
  mutate(
    rgn_id = gsub(" ", "_", rgn_id),
    current_status_int = round(current_status, 0),

    # Categorize into meaningful bins
    status_cat = case_when(
      current_status_int >= 80 ~ "Good (80–100)",
      current_status_int >= 60 ~ "Moderate (60–79)",
      current_status_int >= 40 ~ "Fair (40–59)",
      TRUE ~ "Poor (0–39)"
    )
  )

# Join onto OHI polygons
ohi_coral_condition <- ohi_regions %>%
  left_join(coral_condition_map, by = "rgn_id")

# Split marine vs land
ohi_marine <- ohi_coral_condition %>% filter(location == "marine")
ohi_land   <- ohi_coral_condition %>% filter(location == "land")

# Custom colors
status_colors <- c(
  "Good (80–100)" = "#1a9850",      # green
  "Moderate (60–79)" = "#fee08b",  # yellow
  "Fair (40–59)" = "#fdae61",      # orange
  "Poor (0–39)" = "#d73027"        # red
)

# Plot
tmap_mode("plot")

tm_shape(ohi_land) +
  tm_polygons(
    col = "#f5e6c8",
    border.col = "gray40",
    lwd = 1,
    alpha = 0.5,
    legend.show = FALSE
  ) +

tm_shape(ohi_marine) +
  tm_polygons(
    col = "status_cat",
    palette = status_colors,
    alpha = 0.85,
    title = "Coral Current Status (2025)",
    border.col = "black",
    lwd = 1.2
  ) +

tm_layout(
  title = "Coral Current Status (2025)",
  frame = FALSE,
  legend.outside = TRUE
)


```



