################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

original author: "Austin Martinez"

################################################################################

# PRE-WORK ------------------------------

```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Load in Libraries
library(sf)
library(dplyr)
library(here)
library(terra)
library(tidyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(tmap)

```


```{r Load in the OHI Region_Key0}
Region_Key0<- read.csv("/home/ferrer/ohi_goc_prep/FP/v2026/CONAPESCA_Keys/Spatial_Keys/RegionKey_FINAL_May262026.csv")

colnames(Region_Key0)

Region_Key<- Region_Key0 %>%
  select(rgn_id= OHI_region_est, rgn_name= OHI_region_name) %>%
  distinct()

# Make uppercase:
Region_Key$rgn_id <- sub("^r", "R", Region_Key$rgn_id)
```

```{r Read in OHI region boundaries}
# Read in OHI region boundaries
ohi_regions <- st_read(here("spatial", "ohi_regions", "goc_ohi_rgns.shp")) %>% 
  st_make_valid()

```

# STEP 1: CALCULATE CORAL REEF "CURRENT EXTENT" (sources: Allen Coral Atlas [2020], dataMares [2016], Hem / CEDO coral data [2025]) ------------------------------
Extent = Total area from dataMares - area from Allen Coral Atlas, with remainder split across all points identified in CEDO data.

```{r Read in Coral Reef extent data sets}
# Read in Allen polygon data (2018-2020):
AllenAtlas_coral_raw <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/coral_datasets/ALLEN_coral_Atlas/Mexican-Exclusive-Economic-Zone-20230309211010/Reef-Extent/reefextent.gpkg") %>% 
  st_make_valid()

# Read in CEDO point data (2025):
CEDO_coral_raw <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/Hem_data/goc_habitat_data/coral/Arrecifes_de_coral_(Golfo_de_California).shp") %>% 
  st_make_valid()

# Total area of corals in the GOC according to dataMares (2016): https://datamares.org/stories/marine-habitat-distributions-in-the-gulf-of-california-2/
CORL_area_dataMares_ha <- 657
```

```{r Match CRS to OHI regions}
# Match CRS to OHI regions
AllenAtlas_coral.data <- st_transform(AllenAtlas_coral_raw, st_crs(ohi_regions))
CEDO_coral.data <- st_transform(CEDO_coral_raw, st_crs(ohi_regions))
```

```{r Clip data to the OHI boundies extent}
# Clip data to the OHI boundies extent
AllenAtlas_coral.data <- st_intersection(AllenAtlas_coral.data, ohi_regions)
CEDO_coral.data   <- st_intersection(CEDO_coral.data, ohi_regions)
```

```{r Identify points from the CEDO data set that do / do not intersect the Allen Atlas polygons}
# Identify points from the CEDO data set that intersect the Allen Atlas polygons
corals_that_intersect <- st_intersects(CEDO_coral.data, AllenAtlas_coral.data)

# Keep only the CEDO points that do NOT intersect Allen Atlas polygons
corals_that_do_not_intersect <- CEDO_coral.data[lengths(corals_that_intersect) == 0,]

# Find how many points don't intersect a polygon:
non_intersect_coral_points <- as.numeric(nrow(corals_that_do_not_intersect))
non_intersect_coral_points
# n = 47 points

```

```{r Convert from m^2 to hectares}
CORL_area_AllenAtlas_m2 <- sum(st_area(AllenAtlas_coral.data))
CORL_area_AllenAtlas_ha <- as.numeric(CORL_area_AllenAtlas_m2) / 10000
CORL_area_AllenAtlas_ha
```

```{r Find remainder and assign area to non intersecting points}
# Subtract total coral area of CORL_area_AllenAtlas_ha from CORL_area_dataMares_ha
remaining_coral_area_ha <- CORL_area_dataMares_ha - CORL_area_AllenAtlas_ha
remaining_coral_area_ha

# Find how much area each remaning points should be assigned
area_per_point_ha = remaining_coral_area_ha / non_intersect_coral_points
area_per_point_ha

# NOTE TO SELF: 2.6, slightly different than Austin's 2.7 because of a CRS matching change. - Oct 7 2026
```

```{r Construct COMBINED coral reef dataset}

# Rename AllenAtlas_coral.data$geom → geometry
AllenAtlas_coral.data <- AllenAtlas_coral.data %>%
  rename(geometry = geom)

# Drop the unwanted columns to concat corals_that_do_not_intersect and AllenAtlas_coral.data
AllenAtlas_coral.data <- AllenAtlas_coral.data %>%
  select(-class)

corals_that_do_not_intersect <- corals_that_do_not_intersect %>%
  select(-Habitat)

# Add an area_ha column For polygons (AllenAtlas_coral.data) and For points (corals_that_do_not_intersect)
AllenAtlas_coral.data <- AllenAtlas_coral.data %>%
  mutate(area_ha = as.numeric(st_area(geometry)) / 10000)

corals_that_do_not_intersect <- corals_that_do_not_intersect %>%
  mutate(area_ha = area_per_point_ha)

# Bind the two datasets together
COMBINED_coral.data_area <- bind_rows(AllenAtlas_coral.data, corals_that_do_not_intersect)
```

```{r SANITY CHECK on total area}
# Check if it worked: All areas should sum to 657 ha (dataMares estimate)
sum(COMBINED_coral.data_area$area_ha)
```

```{r Assign CORAL extent to each OHI region}
# Assign the coral extent to each OHI region:
CORL_Current_Extent <- COMBINED_coral.data_area %>%
  st_drop_geometry() %>%
  dplyr::group_by(rgn_id) %>%
  dplyr::summarise(
    area_ha = sum(area_ha, na.rm = TRUE),
    year = 2025
  )

all_rgns <- tibble(
  rgn_id = paste0("Region_", 1:9),
  year = 2025
)

CORL_Current_Extent <- all_rgns %>%
  left_join(CORL_Current_Extent, by = c("rgn_id", "year")) %>%
  mutate(area_ha = replace_na(area_ha, 0)) %>%
  select(rgn_id, 
         area_est_2025_ha = area_ha, 
         year)

CORL_Current_Extent

```

```{r Ktable of Current Extent}
CORL_Current_Extent %>%
  kable(digits = 4, col.names = c("Region", "Extent (ha) (Allen Atlas data, 2020; CEDO data, 2025; dataMares, 2017)", "Year")) %>%
  kable_styling(bootstrap_options = c("striped", "bordered")) %>%
  add_header_above(
    c('Coral Reef Extent' = 3))

```

## Save CORL_Current_Extent
```{r Save CORL_Current_Extent}
write.csv(
  CORL_Current_Extent,
  here::here("_habitats/HAB_corals/CRL_outputs/CORL_Current_Extent_1.1_Oct072026.csv"),
  row.names = FALSE
)
```

```{r Map of CORL_Current_Extent}
tmap_mode("plot")

# Add dummy habitat columns for legend control
AllenAtlas_coral.data$habitat <- "Coral Reef"
CEDO_coral.data$habitat   <- "Coral Reef"

tm_shape(ohi_regions) +
  tm_polygons(
    col = "location",
    palette = c(
      "marine" = "#cceeff",   # light blue
      "land"   = "#f5e6c8"    # light tan
    ),
    alpha = 0.5,
    border.col = "gray",
    border.lwd = 0.6,
    legend.show = FALSE
  ) +

# Allen Coral polygons
tm_shape(AllenAtlas_coral.data) +
  tm_polygons(
    col = "habitat",
    palette = c("Allen Coral Polygons" = "#FF4500"),  # orange-red
    alpha = 1,
    border.col = "#FF4500",
    border.lwd = 2.5,
    title = "Habitat"
  ) +

# Hem Coral points
tm_shape(CEDO_coral.data) +
  tm_dots(
    col = "habitat",
    palette = c("Hem Coral Points" = "#FF4500"),       # blue
    size = 0.05,
    alpha = 1,
    title = "Habitat"
  ) +

tm_layout(
    title.size = 1.3,
    title.fontface = "bold",
    title.position = c(0.02, 0.98),
    frame = FALSE
  ) +

tm_scale_bar(position = c("left", "bottom"))

```


# STEP 2: CALCULATE CORAL REEF "CURRENT STATUS", i.e. condition over time (source: literature review from 1980s-2024)  ------------------------------

```{r Load in extent data as needed}
# Extent data from 3 data sources:
CORL_Current_Extent <- read_csv(here::here("_habitats/HAB_corals/CRL_outputs/CORL_Current_Extent_1.1_Oct072026.csv"))
# 9 obs., 3 vars.
```

```{r Load in Coral Percent Cover data}
# Percent cover data from literature review:
CORL_percent.cover_data_raw <- read_csv("/home/shares/ohi/OHI_GOC/_raw_data/coral_datasets/Coral_cover_lit_review/Coral_PercentCover_Data_revised_Oct072026.csv")
# 103 obs.,  11 vars.
```

```{r Prep CORL_percent.cover_data}
colnames(CORL_percent.cover_data_raw)

# First ggregate at the site level:
CORL_percent.cover_data <- CORL_percent.cover_data_raw %>%
  filter(include_exclude == "INCLUDE") %>%
  
  select(rgn_id, area, specific_site, year, percent_coral_cover_est) %>%
  group_by(rgn_id, area, year) %>%
  
  summarize(
    mean_LCC = mean(percent_coral_cover_est, na.rm = TRUE),
    n_obs = sum(!is.na(percent_coral_cover_est)),
    .groups = "drop"
  )
# 59 obs.,  5 vars.
```

```{r Plot live coral cover over commonly sampled areas in Region 4 and 9}
CORL_percent.cover_data %>%
ggplot(aes(x = year, y = mean_LCC, color = area, shape= rgn_id)) +
  geom_line() +
  geom_point() +
  theme_bw() +
  labs(
    x = "Year",
    y = "% Live Coral Cover",
    color = "General Area",
    title = "Live Coral Cover Over Time at General Areas (Regions 4 & 9)"
  ) +
  theme_bw(base_size = 14) +
  theme(
    strip.text = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 16, face = "bold")
  )
```



```{r Aggregate at the Region_ level for CORL_Scores0}
# Next aggregate at the Region level:
CORL_cover_data_rgn<- CORL_percent.cover_data %>%
  group_by(rgn_id, year) %>%
  summarize(
    mean_LCC = mean(mean_LCC, na.rm = TRUE),
    n_obs = sum(n_obs),
    .groups = "drop"
  )
# 44 obs., 4 vars.

```


```{r Plot Coral Cover over time in Region_4 and 9}

Region_colors<- c(
  "Region_4" = "#dcaa38",
  "Region_9" = "#8d61a9")

CORL_cover_data_rgn %>%
ggplot(aes(x = year, y = mean_LCC, color = rgn_id)) +
  geom_line() +
  geom_point() +
  
  scale_color_manual(values = Region_colors) +
  
  labs(
    x = "Year",
    y = "% Live Coral Cover",
    color = "Region",
    title = "Live Coral Cover Over Time in Regions 4 & 9"
  ) +
  
  theme_bw(base_size = 14) +
  theme(
    strip.text = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 16, face = "bold")
  )
```

## GAPFILL BETWEEN OBSERVED YEARS:
```{r Gapfill CORL_cover_data_rgn}
colnames(CORL_cover_data_rgn)

# First, expand the WETL dataframe to include one row per year per region, even if that region has no data: 
CORL_cover_data_rgn <- CORL_cover_data_rgn %>%
  complete(
    rgn_id = c(paste0("Region_", 1:9)),
    year = full_seq(year, 1)
  )
# 297 obs., 4 vars.

# Keep track of what was originally observed:
CORL_cover_data_rgn<- CORL_cover_data_rgn %>%
  rename(observed_LCC= mean_LCC) %>%
  mutate(
    estimated_LCC = observed_LCC
  )

# Gapfill LCC in missing years using the zoo package:
CORL_cover_data_rgn <- CORL_cover_data_rgn %>%
  group_by(rgn_id) %>%
  fill(rgn_id, .direction = "downup") %>%
  mutate(

    # Gap-fill missing values
    estimated_LCC = zoo::na.approx(
      estimated_LCC,
      x = year,
      na.rm = FALSE
    ),

    # Classify provenance
    provenance = case_when(
      !is.na(observed_LCC) ~ "observed",
      is.na(observed_LCC) & !is.na(estimated_LCC) ~ "gapfilled",
      is.na(estimated_LCC) ~ NA_character_
    )
  ) %>%
  ungroup()

# 297 obs., 6 vars.
```

```{r Compare cover post- and pre- HISTORIC_CUTOFF}
# DECISION POINT: Historic cut off = pre and post 2010
HISTORIC_CUTOFF = 2010

# Distinguish between pre- and post- historic baseline
CORL_Scores0 <- CORL_cover_data_rgn %>%
  mutate(
    period = if_else(year < HISTORIC_CUTOFF, "historic", "contemporary")
  )

CORL_Scores0 <- CORL_Scores0 %>%
  group_by(rgn_id) %>%
  mutate(
    historic_mean = mean(observed_LCC[year < HISTORIC_CUTOFF], na.rm = TRUE),
    contemporary_mean = mean(observed_LCC[year >= HISTORIC_CUTOFF], na.rm = TRUE),
    CORL_current_status = pmax(0, pmin((estimated_LCC / historic_mean) * 100, 100))
  ) %>%
  ungroup() %>%
  select(rgn_id,
         year,
         period,
         historic_mean,
         contemporary_mean,
         observed_LCC,
         estimated_LCC,
         provenance,
         n_obs,
         CORL_current_status)

# Relevel:
CORL_Scores0 <- CORL_Scores0 %>%
  mutate(
    period = factor(
      period,
      levels = c("historic", "contemporary")
    ),
    provenance = factor(
      provenance,
      levels = c("observed", "gapfilled")
    )
  )
```

```{r Plot coral trends}

# Prep for plot:
Regions_of_Interest<- c("Region_4", "Region_9")

historic_mean <- unique(CORL_Scores0$historic_mean)
contemporary_mean <- unique(CORL_Scores0$contemporary_mean)

periods <- CORL_Scores0 %>%
  filter(rgn_id %in% Regions_of_Interest) %>%
  group_by(rgn_id) %>%
  summarise(
    historic_mean = first(historic_mean),
    contemporary_mean = first(contemporary_mean),
    pre_start = min(year[year < HISTORIC_CUTOFF]),
    post_end = max(year[year >= HISTORIC_CUTOFF]),
    .groups = "drop"
  )

# Plot:
CORL_Cover_OverTime<- CORL_Scores0 %>%
  filter(rgn_id %in% Regions_of_Interest) %>%
  
  ggplot(aes(year, observed_LCC, color = period)) +
  
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  
  facet_wrap(~rgn_id) +
  
  # Period means
  geom_segment(
    data = periods,
    aes(x = pre_start, xend = 2010, y = historic_mean, yend = historic_mean),
    color = "blue", linetype = "dotted", linewidth = 1
  ) +
  geom_segment(
    data = periods,
    aes(x = 2010, xend = post_end, y = contemporary_mean, yend = contemporary_mean),
    color = "orange", linetype = "dotted", linewidth = 1
  ) +
  
  geom_vline(
    xintercept = 2010,
    color = "gray40",
    linewidth = 1
  ) +
  
  scale_color_manual(
    values = c("historic" = "blue", "contemporary" = "orange")
  ) +
  
  scale_y_continuous(limits = c(0, 40)) +
  
  labs(
    title = "Historic versus Contemporary Live Coral Cover",
    subtitle = "LCC estimates taken from multiple data sets across sites that were consistently sampled",
    caption = "updated: Oct 7 2026",
    x = "Year",
    y = "% Live Coral Cover (LCC)",
    color = "Period"
  ) +
  
  theme_bw(base_size = 14) +
  theme(
    strip.text = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 16, face = "bold")
  )

CORL_Cover_OverTime

```

## Save CORL_Cover_OverTime

```{r Save as PNG}
ggsave(
  filename = sprintf("/home/ferrer/ohi_goc_prep/_habitats/HAB_corals/CRL_outputs/CORL_OverTime_plot_Oct072026.png"),
  plot = CORL_Cover_OverTime,
  width = 12,
  height = 12,
  dpi = 300
)
```

# STEP 3: FINALIZE CORL_Scores DATAFRAME ------------------------------

```{r Merge current extent estimates to CORL_Scores0}
colnames(CORL_Scores0)
colnames(CORL_Current_Extent)

# Merge:
CORL_Scores <- CORL_Scores0 %>%
  left_join(select(CORL_Current_Extent,
                   c("rgn_id", "area_est_2025_ha")),
            by="rgn_id") %>%
  rename(
    area_est_byMultiScources_ha = area_est_2025_ha
  )

# 297 obs., 11 vars.
  
# Make an include / exclude column based on the information reported by Hem / CEDO:
CORL_Scores<- CORL_Scores %>%
  mutate(CORL_include_exclude = ifelse(area_est_byMultiScources_ha == 0, "EXCLUDE", 
                                      "INCLUDE"))
# 297 obs., 12 vars.

```

```{r Make a nice Ktable}
CORL_Scores %>%
  filter(year == 2024) %>%
  select(rgn_id, year, CORL_current_status) %>%
  kable(digits = 4, col.names = c("Region", "Year", "Current Status (vs. historic pre-2010)")) %>%
  kable_styling(bootstrap_options = c("striped", "bordered")) %>%
  add_header_above(
    c('Coral Reef Current Status' = 3))


```

```{r Join rgn_name to scores}
colnames(Region_Key)
colnames(CORL_Scores)

CORL_Scores <- CORL_Scores %>%
  left_join(Region_Key, by = "rgn_id")
  
CORL_Scores<- CORL_Scores %>%
  relocate(rgn_name, .before = rgn_id)
```

## Save CORL_Scores

```{r Save CORL_Scores}
write.csv(CORL_Scores,
          here::here("_habitats/HAB_corals/CRL_outputs/CORL_Scores_1.1_df_Oct072026.csv"),
          row.names = FALSE)
```

## Plot CORL_Scores

```{r Plot CORL_Scores0}
colnames(CORL_Scores)
unique(CORL_Scores$provenance)

Region_colors<- c(
  "Western Southern Gulf" = "#dcaa38",
  "Southeastern Gulf" = "#8d61a9")

# Plot:
CORL_Scores_OverTime<- CORL_Scores %>%
  ggplot(aes(year, CORL_current_status, color = rgn_name)) +
  
  geom_line(linewidth = 1) +
  
  geom_point(
    aes(shape = provenance, fill = provenance),
    size = 2
  ) +
  
  scale_shape_manual(
    values = c(
      "gapfilled" = 21,
      "observed" = 16
    )
  ) +
  
  scale_fill_manual(
    values = c(
      "gapfilled" = "white",
      "observed" = NA
    )
  ) +
  
  geom_vline(
    xintercept = 2010,
    color = "gray40",
    linewidth = 1
  ) +
  
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey50") +
  
  ylim(0, 101) +
  
  scale_color_manual(values = Region_colors) +
  
  labs(
    title = "CORL Component Scores (1.1) by Region",
    subtitle = "Source: multiple data sources, incl. Reyes-Bonilla et al. and Cupul-Magana et al. unplished data",
    caption = "updated: Oct 7 2026",
    x = "Year",
    y = "CORL Score (from 0 to 100)",
    color = "Region",
  ) +
  
  theme_bw(base_size = 14) +
  theme(
    strip.text = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 16, face = "bold")
  )

CORL_Scores_OverTime

```

```{r Save as PNG}
ggsave(
  filename = sprintf("/home/ferrer/ohi_goc_prep/_habitats/HAB_corals/CRL_outputs/CORL_Scores_1.1_plot_Oct072026.png"),
  plot = CORL_Scores_OverTime,
  width = 12,
  height = 12,
  dpi = 300
)
```

```{r}
# Prepare coral condition data for join
coral_condition_map <- CORL_Scores %>%
  filter(year == 2023) %>%
  mutate(
    rgn_id = gsub(" ", "_", rgn_id),
    current_status_int = round(CORL_current_status, 0),

    # Categorize into meaningful bins
    status_cat = case_when(
      current_status_int >= 80 ~ "Good (80–100)",
      current_status_int >= 60 ~ "Moderate (60–79)",
      current_status_int >= 40 ~ "Fair (40–59)",
      is.na(current_status_int) ~ "No data",
      TRUE ~ "Poor (0–39)"
    )
  )

# Join onto OHI polygons
ohi_coral_condition <- ohi_regions %>%
  left_join(coral_condition_map, by = "rgn_id")

# Split marine vs land
ohi_marine <- ohi_coral_condition %>% filter(location == "marine")
ohi_land   <- ohi_coral_condition %>% filter(location == "land")

# Custom colors
status_colors <- c(
  "Good (80–100)" = "#1a9850",      # green
  "Moderate (60–79)" = "#fee08b",  # yellow
  "Fair (40–59)" = "#fdae61",      # orange
  "Poor (0–39)" = "#d73027",        # red
  "No data" = "darkgrey" # grey
)

# Plot
tmap_mode("plot")

tm_shape(ohi_land) +
  tm_polygons(
    col = "#f5e6c8",
    border.col = "gray40",
    lwd = 1,
    alpha = 0.5,
    legend.show = FALSE
  ) +

tm_shape(ohi_marine) +
  tm_polygons(
    col = "status_cat",
    palette = status_colors,
    alpha = 0.85,
    title = "Coral Current Status (2023)",
    border.col = "black",
    lwd = 1.2
  ) +

tm_layout(
  title = "Coral Current Status (2023)",
  frame = FALSE,
  legend.outside = TRUE
)

```


# STEP 6: ESTIMATE CORL_Pressure, the inverse of CORL_current_status ------------------------------

```{r Estimate pressure as 100 minus current status}
CORL_Pressure <- CORL_Scores %>%
  mutate(pressure = (100 - CORL_current_status)/100) %>%
  select(rgn_name, rgn_id, year, pressure)

```

## Save CORL_Pressure
```{r Save CORL_Pressure}
write.csv(CORL_Pressure,
          here::here("_habitats/HAB_corals/CRL_outputs/CORL_Pressure_1.1_Oct072026.csv"),
          row.names = FALSE)
```


# STEP 7: ESTIMATE CORL TREND (source: multiple data sources, 1997-present) ------------------------------

## METHOD A

```{r Calculate CORL_Trend using METHOD A}
colnames(CORL_Scores)

# METHOD A: 
# sSee GitHub issue, Oct 7 2026, for more details: https://github.com/OHI-GOC/OHI_GOC_issues/issues/127#issuecomment-6041025924
# slope <- (condition_end - condition_start) / n_years
# trend_pct_yr <- (slope / condition_start) * 100

# Calculate the trend as the linear annualized percent change between the most recent year (i.e., 2021) and 5-10 years prior (i.e., 2013)
CORL_Trend0 <- CORL_Scores %>%
  filter(CORL_include_exclude == "INCLUDE") %>%
  filter(provenance == "observed") %>%
  group_by(rgn_name, rgn_id) %>%
  mutate(
    recent_year = max(year, na.rm = TRUE),
    comparison_year = min(
      year[year < recent_year &
           year >= recent_year - 10 &
           year <= recent_year - 5]
    ),
    recent_LCC = observed_LCC[year == recent_year][1],
    comparison_LCC = observed_LCC[year == comparison_year][1],
    n_years = recent_year - comparison_year,
    slope = (recent_LCC - comparison_LCC) / n_years,
    trend = round((slope / comparison_LCC) * 100, 2)
  ) %>%
  filter(year == recent_year) %>%
  ungroup() %>%
  select(
    rgn_name,
    rgn_id,
    comparison_year,
    comparison_LCC,
    most_recent_year= recent_year,
    most_recent_LCC= recent_LCC,
    trend_linear_pct_year_opA = trend
  )

CORL_Trend0

```

## METHOD B

```{r Calculate CORL_Trend using METHOD B}
# METHOD B: 
# See GitHub issue, Oct 7 2026, for more details: https://github.com/OHI-GOC/OHI_GOC_issues/issues/127#issuecomment-6041025924
# slope = fit a linear model: lm(area ~ year), grab the slope estimate
# trend_pct_year <- (1 + slope/earliest_area_in_lm)

# Pull the most recent year and the comparison year:
most_recent_year = max(CORL_Scores$year, na.rm = TRUE)

comparison_year <- min(
  CORL_Scores$year[
    CORL_Scores$provenance == "observed" &
    CORL_Scores$year < most_recent_year &
    CORL_Scores$year >= most_recent_year - 10 &
    CORL_Scores$year <= most_recent_year - 5
  ],
  na.rm = TRUE
)

CORL_cover_lm <- CORL_Scores %>%
  filter(
    year >= comparison_year,
    year <= most_recent_year,
    provenance == "observed"
  ) %>%
  group_by(rgn_id) %>%
  summarise(
    model = list(
      lm(observed_LCC ~ year, data = pick(everything()))
    ),
    comparison_LCC = observed_LCC[year == comparison_year][1],
    .groups = "drop"
  ) %>%
  mutate(
    slope = sapply(model, function(x) coef(x)[["year"]]),
    p_value = sapply(
      model,
      function(x) summary(x)$coefficients["year", "Pr(>|t|)"]
    ),
    trend_pct_year = round((slope / comparison_LCC) * 100, 2)
  ) %>%
  select(rgn_id, 
         trend_linear_pct_year_opB = trend_pct_year,
         lm_slope= slope, 
         p_value)

```

```{r Merge options A and B}
CORL_Trend<- CORL_Trend0 %>%
  left_join(
    CORL_cover_lm, 
    by= "rgn_id"
  )
```


```{r Make a nice Ktable for CORL_Trend}
colnames(CORL_Trend)

CORL_Trend %>%
  select(rgn_id, trend_linear_pct_year_opA, trend_linear_pct_year_opB) %>%
  kable(digits = 4, col.names = c("Region", "Trend (method A)", "Trend (method B)")) %>%
  kable_styling(bootstrap_options = c("striped", "bordered")) %>%
  add_header_above(
    c('Coral Reef Trend' = 3))
```

## Save CORL_Trend
```{r Save CORL_Trend}
write.csv(CORL_Trend,
          here::here("_habitats/HAB_corals/CRL_outputs/CORL_Trend_1.1_Oct072026.csv"),
          row.names = FALSE)
```
