---
title: 'OHI GoC 2024: Lasting Special Places (LSP)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor:
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options:
  chunk_output_type: console
---

# Sentido de Pertenencia: Lasting Special Places

## Script 8: merge region scores into current status for 2020 and 2023

Scripts 2 through 7 each write a region-level condition table with a `year` column (2020 and 2023). Alrighty, now I will read those tables, stack them, and take a simple average within each year so each OHI region has one **current status** score for Lasting Special Places in 2020 and in 2023.

**What I do**

1. Read each `condition_region_*.csv` from scripts 2 through 7.
2. Make sure `rgn_id` uses underscores.
3. Average `condition` by region and year, then write `lsp_goal_current_status_by_region.csv`.

```{r}
library(tidyverse)
library(here)
```

## Read region tables from each LSP type

First I will load fishing, archaeology, protected areas, dive sites, special habitats, and all other marine places. Each table should already include `year`.

```{r}
clean_rgn <- function(path) {
  read_csv(path, show_col_types = FALSE) %>%
    mutate(
      rgn_id = str_replace_all(rgn_id, " ", "_"),
      year = as.integer(year)
    )
}

rgn_fishing_site_condition <- clean_rgn(
  here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2026/output", "condition_region_fishing_sites.csv")
)

rgn_archaeological_site_condition <- clean_rgn(
  here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2026/output", "condition_region_archaeological_sites.csv")
)

rgn_protected_area_condition <- clean_rgn(
  here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2026/output", "condition_region_protected_areas.csv")
)

rgn_dive_sites_condition <- clean_rgn(
  here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2026/output", "condition_region_dive_sites.csv")
)

rgn_special_habitats_condition <- clean_rgn(
  here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2026/output", "condition_region_special_habitats.csv")
) %>%
  dplyr::select(-any_of("n_features"))

other_sites_condition <- clean_rgn(
  here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2026/output", "condition_region_other_marine_places.csv")
)
```

## Average across LSP types within each year and write

Now that I have all six region tables, I will take an equal average within each year and write the result. If we decide one type should weigh more later, I can change this step.

```{r}
all_rgn_lsp_conditions <- bind_rows(
  rgn_fishing_site_condition,
  rgn_archaeological_site_condition,
  rgn_protected_area_condition,
  rgn_dive_sites_condition,
  rgn_special_habitats_condition,
  other_sites_condition
)

lsp_current_status <- all_rgn_lsp_conditions %>%
  group_by(year, rgn_id) %>%
  summarize(
    current_status = mean(condition, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(year, rgn_id)

write_csv(
  lsp_current_status,
  here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2026/output", "lsp_goal_current_status_by_region.csv")
)

print(lsp_current_status)

# quick look at change from 2020 to 2023
lsp_current_status %>%
  pivot_wider(names_from = year, values_from = current_status, names_prefix = "status_") %>%
  mutate(change_2020_to_2023 = status_2023 - status_2020)
```

