---
title: 'OHI GoC 2024: Lasting Special Places (LSP)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor:
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options:
  chunk_output_type: console
---

# Sentido de Pertenencia: Lasting Special Places

## Script 9: choropleth maps for 2020 and 2023

Alrighty, now that script 8 has written region status for both years, I will make static choropleth maps for GitHub and the update note. First I will map overall LSP current status for **2020** and **2023**. Then I will map each LSP type the same way so we can see which pathways pull a region up or down. White means that pathway has no score for that region.

The map style follows the earlier Sense of Place choropleths: Gulf states underlay, OHI marine regions filled 0 to 100, white for missing values.

**What I do**

1. Read `lsp_goal_current_status_by_region.csv` and each `condition_region_*.csv` from the v2026 output folder.
2. Join scores to the OHI marine region polygons.
3. Save PNG (and PDF) choropleths for overall status and for each LSP type, for 2020 and 2023.

```{r}
library(tidyverse)
library(sf)
library(here)
library(ggplot2)
```

## Paths and map styling

```{r}
out_dir <- here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2026/output")
fig_dir <- here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2026/figures_choropleth")


normalize_rgn_id <- function(x) {
  x %>%
    as.character() %>%
    str_squish() %>%
    str_replace_all(" ", "_")
}

rgn_labels <- c(
  "Region_1" = "Western Upper Gulf",
  "Region_2" = "Western Midriff Islands",
  "Region_3" = "Western Central Gulf",
  "Region_4" = "Western Southern Gulf",
  "Region_5" = "Eastern Upper Gulf",
  "Region_6" = "Eastern Midriff Islands",
  "Region_7" = "Eastern Central Gulf",
  "Region_8" = "Eastern Southern Gulf",
  "Region_9" = "Southeast Gulf / Nayarit"
)

# tr / access style ramp: low scores dark red, high scores dark blue
lsp_palette <- c(
  "#8a0032", "#A50026", "#A50026", "#D73027", "#F46D43",
  "#FDAE61", "#FEE090", "#FEE090", "#ABD9E9", "#4575B4", "#313695"
)
lsp_na_fill <- "white"
```

## Read regions and status tables

I only keep marine OHI polygons for the choropleth fill. The region shapefile also has inland buffer rows, and plotting both was stacking outlines and duplicating score labels. I also transform the states layer to the same CRS so the `#d6cac0` underlay lines up under the Gulf regions.

```{r}
ohi_marine <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp"), quiet = TRUE) %>%
  filter(location == "marine") %>%
  mutate(rgn_id = normalize_rgn_id(rgn_id)) %>%
  dplyr::select(rgn_id, geometry) %>%
  # one polygon per region so each score is labeled once
  group_by(rgn_id) %>%
  summarize(geometry = st_union(geometry), .groups = "drop") %>%
  st_make_valid()

states <- st_read(here("spatial/ohi_regions/states.shp"), quiet = TRUE) %>%
  st_transform(st_crs(ohi_marine)) %>%
  st_make_valid()

table(ohi_marine$rgn_id)

clean_rgn <- function(path, value_col = "condition") {
  read_csv(path, show_col_types = FALSE) %>%
    mutate(
      rgn_id = normalize_rgn_id(rgn_id),
      year = as.integer(year)
    ) %>%
    dplyr::select(year, rgn_id, score = all_of(value_col))
}

lsp_current_status <- clean_rgn(
  here(out_dir, "lsp_goal_current_status_by_region.csv"),
  value_col = "current_status"
)

score_layers <- list(
  overall = lsp_current_status,
  protected_areas = clean_rgn(here(out_dir, "condition_region_protected_areas.csv")),
  dive_sites = clean_rgn(here(out_dir, "condition_region_dive_sites.csv")),
  archaeological_sites = clean_rgn(here(out_dir, "condition_region_archaeological_sites.csv")),
  fishing_sites = clean_rgn(here(out_dir, "condition_region_fishing_sites.csv")),
  special_habitats = clean_rgn(here(out_dir, "condition_region_special_habitats.csv")),
  other_marine_places = clean_rgn(here(out_dir, "condition_region_other_marine_places.csv"))
)

lapply(score_layers, function(x) table(x$year, useNA = "ifany"))
```

## Choropleth helpers

```{r}
join_scores_to_ohi <- function(score_df) {
  ohi_marine %>%
    left_join(score_df, by = "rgn_id") %>%
    mutate(
      rgn_label = unname(rgn_labels[rgn_id]),
      score = ifelse(is.nan(score), NA_real_, score),
      label = ifelse(is.na(score), "NA", as.character(round(score)))
    )
}

# place one label toward the outer gulf edge so west and east scores do not stack in the middle.
# western regions (1 to 4) sit near the west side of each polygon; eastern regions (5 to 9) sit near the east side.
edge_label_points <- function(dat, west_frac = 0.22, east_frac = 0.78) {
  crs_dat <- st_crs(dat)
  west_ids <- paste0("Region_", 1:4)

  pts <- lapply(seq_len(nrow(dat)), function(i) {
    g <- st_geometry(dat)[[i]]
    bb <- st_bbox(g)
    x_frac <- if (dat$rgn_id[i] %in% west_ids) west_frac else east_frac
    x <- bb["xmin"] + x_frac * (bb["xmax"] - bb["xmin"])
    y <- bb["ymin"] + 0.50 * (bb["ymax"] - bb["ymin"])
    st_point(c(x, y))
  })

  st_sf(
    dat %>% st_drop_geometry(),
    geometry = st_sfc(pts, crs = crs_dat)
  )
}

make_lsp_choropleth <- function(score_df, plot_title = NULL, label_scores = TRUE) {
  dat <- join_scores_to_ohi(score_df)
  label_pts <- edge_label_points(dat)

  p <- ggplot() +
    # states underlay first so #d6cac0 shows on land behind the marine regions
    geom_sf(
      data = states,
      fill = "#d6cac0",
      color = "grey40",
      linewidth = 0.15
    ) +
    geom_sf(
      data = dat,
      aes(fill = score),
      color = "grey20",
      linewidth = 0.25
    ) +
    scale_fill_gradientn(
      colours = lsp_palette,
      limits = c(0, 100),
      na.value = lsp_na_fill,
      name = "Score",
      breaks = seq(0, 100, by = 25)
    ) +
    labs(title = plot_title) +
    coord_sf(expand = TRUE) +
    theme_void(base_size = 12) +
    theme(
      panel.background = element_rect(fill = "white", color = NA),
      plot.background = element_rect(fill = "white", color = NA),
      plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
      legend.position = "bottom",
      legend.direction = "horizontal",
      legend.key.height = unit(0.35, "cm"),
      legend.key.width = unit(1.2, "cm"),
      legend.title = element_text(size = 10),
      legend.text = element_text(size = 9),
      plot.margin = margin(10, 14, 10, 14)
    )

  if (isTRUE(label_scores)) {
    p <- p +
      geom_label(
        data = label_pts,
        aes(label = label, geometry = geometry),
        stat = "sf_coordinates",
        size = 3.2,
        fontface = "bold",
        color = "grey10",
        fill = "white",
        label.size = 0.15,
        label.padding = unit(0.12, "lines"),
        alpha = 0.92
      )
  }

  p
}

save_lsp_map <- function(gg, stem, width = 7.5, height = 7.5, dpi = 200) {
  png_path <- here(fig_dir, paste0(stem, ".png"))
  pdf_path <- here(fig_dir, paste0(stem, ".pdf"))
  ggsave(png_path, gg, width = width, height = height, dpi = dpi, bg = "white")
  tryCatch(
    ggsave(pdf_path, gg, width = width, height = height, device = cairo_pdf, bg = "white"),
    error = function(e) {
      ggsave(pdf_path, gg, width = width, height = height, device = "pdf", bg = "white")
    }
  )
  message("wrote ", png_path)
  invisible(png_path)
}
```

## Overall LSP status: 2020 and 2023

Now I will map the merged current status for each year. These two PNGs are the ones I would attach first in the GitHub update comment.

```{r fig.width=7.5, fig.height=7.5}
map_overall_2020 <- make_lsp_choropleth(
  score_layers$overall %>% filter(year == 2020L),
  plot_title = "LSP current status by region (2020)"
)
map_overall_2023 <- make_lsp_choropleth(
  score_layers$overall %>% filter(year == 2023L),
  plot_title = "LSP current status by region (2023)"
)

print(map_overall_2020)
print(map_overall_2023)

save_lsp_map(map_overall_2020, "lsp_current_status_choropleth_2020")
save_lsp_map(map_overall_2023, "lsp_current_status_choropleth_2023")
```

### Side by side overall comparison

```{r fig.width=12, fig.height=6.5}
overall_facet_df <- score_layers$overall %>%
  filter(year %in% c(2020L, 2023L)) %>%
  mutate(year_label = paste0("LSP current status (", year, ")"))

overall_facet_sf <- ohi_marine %>%
  left_join(overall_facet_df, by = "rgn_id") %>%
  mutate(
    score = ifelse(is.nan(score), NA_real_, score),
    label = ifelse(is.na(score), "NA", as.character(round(score)))
  )

overall_facet_labels <- edge_label_points(overall_facet_sf)

map_overall_facet <- ggplot() +
  geom_sf(data = states, fill = "#d6cac0", color = "grey40", linewidth = 0.15) +
  geom_sf(data = overall_facet_sf, aes(fill = score), color = "grey20", linewidth = 0.25) +
  geom_label(
    data = overall_facet_labels,
    aes(label = label, geometry = geometry),
    stat = "sf_coordinates",
    size = 2.9,
    fontface = "bold",
    color = "grey10",
    fill = "white",
    label.size = 0.15,
    label.padding = unit(0.12, "lines"),
    alpha = 0.92
  ) +
  scale_fill_gradientn(
    colours = lsp_palette,
    limits = c(0, 100),
    na.value = lsp_na_fill,
    name = "Score",
    breaks = seq(0, 100, by = 25)
  ) +
  facet_wrap(~year_label, ncol = 2) +
  coord_sf(expand = TRUE) +
  theme_void(base_size = 12) +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    strip.text = element_text(size = 12, face = "bold"),
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.key.height = unit(0.35, "cm"),
    legend.key.width = unit(1.2, "cm"),
    plot.margin = margin(10, 14, 10, 14)
  )

print(map_overall_facet)
save_lsp_map(map_overall_facet, "lsp_current_status_choropleth_2020_2023_side_by_side", width = 12, height = 6.5)
```

### Change from 2020 to 2023

```{r fig.width=7.5, fig.height=7.5}
status_change <- score_layers$overall %>%
  pivot_wider(names_from = year, values_from = score, names_prefix = "y") %>%
  mutate(score = y2023 - y2020) %>%
  dplyr::select(rgn_id, score)

change_sf <- join_scores_to_ohi(status_change) %>%
  mutate(label = ifelse(is.na(score), "NA", as.character(round(score, 1))))

change_labels <- edge_label_points(change_sf)

map_change <- ggplot() +
  geom_sf(data = states, fill = "#d6cac0", color = "grey40", linewidth = 0.15) +
  geom_sf(data = change_sf, aes(fill = score), color = "grey20", linewidth = 0.25) +
  geom_label(
    data = change_labels,
    aes(label = label, geometry = geometry),
    stat = "sf_coordinates",
    size = 3.2,
    fontface = "bold",
    color = "grey10",
    fill = "white",
    label.size = 0.15,
    label.padding = unit(0.12, "lines"),
    alpha = 0.92
  ) +
  scale_fill_gradient2(
    low = "#8a0032",
    mid = "white",
    high = "#313695",
    midpoint = 0,
    limits = c(-15, 15),
    na.value = lsp_na_fill,
    name = "Change",
    breaks = seq(-15, 15, by = 5)
  ) +
  labs(title = "Change in LSP current status (2023 minus 2020)") +
  coord_sf(expand = TRUE) +
  theme_void(base_size = 12) +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.key.height = unit(0.35, "cm"),
    legend.key.width = unit(1.2, "cm"),
    plot.margin = margin(10, 14, 10, 14)
  )

print(map_change)
save_lsp_map(map_change, "lsp_current_status_change_2020_to_2023")
```

## Choropleths by LSP type

Now I will make the same 2020 and 2023 maps for each pathway, side by side like the overall status figure. First I will build and `print()` every map so I can review them. Saving happens in the next chunk. Regions with no sites of that type stay white.

```{r fig.width=12, fig.height=6.5}
type_titles <- c(
  protected_areas = "Protected areas",
  dive_sites = "Dive sites",
  archaeological_sites = "Archaeological sites",
  fishing_sites = "Fishing sites",
  special_habitats = "Special habitats",
  other_marine_places = "Other marine places"
)

make_type_side_by_side <- function(layer, title_base) {
  facet_df <- layer %>%
    filter(year %in% c(2020L, 2023L)) %>%
    mutate(year_label = paste0(title_base, " (", year, ")"))

  facet_sf <- ohi_marine %>%
    left_join(facet_df, by = "rgn_id") %>%
    mutate(
      score = ifelse(is.nan(score), NA_real_, score),
      label = ifelse(is.na(score), "NA", as.character(round(score)))
    )

  facet_labels <- edge_label_points(facet_sf)

  ggplot() +
    geom_sf(data = states, fill = "#d6cac0", color = "grey40", linewidth = 0.15) +
    geom_sf(data = facet_sf, aes(fill = score), color = "grey20", linewidth = 0.25) +
    geom_label(
      data = facet_labels,
      aes(label = label, geometry = geometry),
      stat = "sf_coordinates",
      size = 2.9,
      fontface = "bold",
      color = "grey10",
      fill = "white",
      label.size = 0.15,
      label.padding = unit(0.12, "lines"),
      alpha = 0.92
    ) +
    scale_fill_gradientn(
      colours = lsp_palette,
      limits = c(0, 100),
      na.value = lsp_na_fill,
      name = "Score",
      breaks = seq(0, 100, by = 25)
    ) +
    facet_wrap(~year_label, ncol = 2) +
    coord_sf(expand = TRUE) +
    theme_void(base_size = 12) +
    theme(
      panel.background = element_rect(fill = "white", color = NA),
      plot.background = element_rect(fill = "white", color = NA),
      strip.text = element_text(size = 12, face = "bold"),
      legend.position = "bottom",
      legend.direction = "horizontal",
      legend.key.height = unit(0.35, "cm"),
      legend.key.width = unit(1.2, "cm"),
      plot.margin = margin(10, 14, 10, 14)
    )
}

# build all type maps and print them for review
type_maps <- list()

for (stem in names(type_titles)) {
  layer <- score_layers[[stem]]
  title_base <- type_titles[[stem]]

  type_maps[[paste0(stem, "_2020")]] <- make_lsp_choropleth(
    layer %>% filter(year == 2020L),
    plot_title = paste0(title_base, " condition by region (2020)")
  )
  type_maps[[paste0(stem, "_2023")]] <- make_lsp_choropleth(
    layer %>% filter(year == 2023L),
    plot_title = paste0(title_base, " condition by region (2023)")
  )
  type_maps[[paste0(stem, "_side_by_side")]] <- make_type_side_by_side(layer, title_base)
}

print(type_maps$protected_areas_2020)
print(type_maps$protected_areas_2023)
print(type_maps$protected_areas_side_by_side)

print(type_maps$dive_sites_2020)
print(type_maps$dive_sites_2023)
print(type_maps$dive_sites_side_by_side)

print(type_maps$archaeological_sites_2020)
print(type_maps$archaeological_sites_2023)
print(type_maps$archaeological_sites_side_by_side)

print(type_maps$fishing_sites_2020)
print(type_maps$fishing_sites_2023)
print(type_maps$fishing_sites_side_by_side)

print(type_maps$special_habitats_2020)
print(type_maps$special_habitats_2023)
print(type_maps$special_habitats_side_by_side)

print(type_maps$other_marine_places_2020)
print(type_maps$other_marine_places_2023)
print(type_maps$other_marine_places_side_by_side)
```

### Save the type choropleths

After I am happy with the printed maps above, I will save each one.

```{r}
save_lsp_map(type_maps$protected_areas_2020, "lsp_protected_areas_choropleth_2020")
save_lsp_map(type_maps$protected_areas_2023, "lsp_protected_areas_choropleth_2023")
save_lsp_map(
  type_maps$protected_areas_side_by_side,
  "lsp_protected_areas_choropleth_2020_2023_side_by_side",
  width = 12,
  height = 6.5
)

save_lsp_map(type_maps$dive_sites_2020, "lsp_dive_sites_choropleth_2020")
save_lsp_map(type_maps$dive_sites_2023, "lsp_dive_sites_choropleth_2023")
save_lsp_map(
  type_maps$dive_sites_side_by_side,
  "lsp_dive_sites_choropleth_2020_2023_side_by_side",
  width = 12,
  height = 6.5
)

save_lsp_map(type_maps$archaeological_sites_2020, "lsp_archaeological_sites_choropleth_2020")
save_lsp_map(type_maps$archaeological_sites_2023, "lsp_archaeological_sites_choropleth_2023")
save_lsp_map(
  type_maps$archaeological_sites_side_by_side,
  "lsp_archaeological_sites_choropleth_2020_2023_side_by_side",
  width = 12,
  height = 6.5
)

save_lsp_map(type_maps$fishing_sites_2020, "lsp_fishing_sites_choropleth_2020")
save_lsp_map(type_maps$fishing_sites_2023, "lsp_fishing_sites_choropleth_2023")
save_lsp_map(
  type_maps$fishing_sites_side_by_side,
  "lsp_fishing_sites_choropleth_2020_2023_side_by_side",
  width = 12,
  height = 6.5
)

save_lsp_map(type_maps$special_habitats_2020, "lsp_special_habitats_choropleth_2020")
save_lsp_map(type_maps$special_habitats_2023, "lsp_special_habitats_choropleth_2023")
save_lsp_map(
  type_maps$special_habitats_side_by_side,
  "lsp_special_habitats_choropleth_2020_2023_side_by_side",
  width = 12,
  height = 6.5
)

save_lsp_map(type_maps$other_marine_places_2020, "lsp_other_marine_places_choropleth_2020")
save_lsp_map(type_maps$other_marine_places_2023, "lsp_other_marine_places_choropleth_2023")
save_lsp_map(
  type_maps$other_marine_places_side_by_side,
  "lsp_other_marine_places_choropleth_2020_2023_side_by_side",
  width = 12,
  height = 6.5
)
```

## Where the figures were written

```{r}
list.files(fig_dir, pattern = "choropleth|change", full.names = FALSE) %>%
  sort()
```
