---
title: 'OHI GoC 2024: Lasting Special Places (LSP)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor:
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options:
  chunk_output_type: console
---

# Sentido de Pertenencia: Lasting Special Places

## Script 7: other marine places of importance

Scripts 2 through 6 cover named protected areas, dive sites, archaeology, fishing grounds, and special habitats. Some EWG rows are still left over as "Marine location of importance". Alrighty, in this script I will score those leftovers. First I will try protected area and SIMEC. If that gives a zero, I will fill in curated legal and literature scores from a Google Sheet.

**What I do**

1. Read the full EWG list and the condition tables from scripts 2 through 5.
2. Keep marine importance rows that were not scored yet (and skip habitat names handled in script 6).
3. Buffer, join the v2 protected area layer, fuzzy match SIMEC, apply Ramsar rules when needed.
4. Fill remaining zeros from `lsps_codified_protections_estimated_condition.csv`, average by region, write CSVs.

```{r}
library(tidyverse)
library(here)
library(sf)
library(mapview)
library(stringdist)
library(stringi)
```

## Read the inventory and earlier scores

```{r}
lsp_list_ewg <- read_csv(
  here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/google_sheets", "ewg_lsp_inventory_prepared.csv"),
  show_col_types = FALSE
) %>%
  janitor::clean_names()

fishing_site_condition <- read_csv(
  here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2026/output", "condition_lsp_fishing_sites.csv"),
  show_col_types = FALSE
)
archaeological_site_condition <- read_csv(
  here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2026/output", "condition_lsp_archaeological_sites.csv"),
  show_col_types = FALSE
)
protected_area_condition <- read_csv(
  here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2026/output", "condition_lsp_protected_areas.csv"),
  show_col_types = FALSE
)
dive_sites_condition <- read_csv(
  here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2026/output", "condition_lsp_dive_sites.csv"),
  show_col_types = FALSE
)
```

## Names already covered in other scripts

Now I will build a simple name list so I can find what is still missing.

```{r}
normalize_name <- function(x) {
  x %>%
    stringi::stri_trans_general("Latin-ASCII") %>%
    str_to_upper() %>%
    str_remove_all(
      "\\bPARQUE NACIONAL\\b|\\bPARQUE\\b|\\bNACIONAL\\b|\\bZONA MARINA\\b|\\bARCHIPIELAGO\\b"
    ) %>%
    str_squish()
}

fishing_site_condition_join <- fishing_site_condition %>%
  dplyr::select(rgn_id, lsp_name = lasting_special_place_lsp, condition = overall_condition)

archaeological_site_condition_join <- archaeological_site_condition %>%
  dplyr::select(rgn_id, lsp_name = lasting_special_place_lsp, condition = overall_condition)

protected_area_condition_join <- protected_area_condition %>%
  mutate(
    lsp_name = case_when(
      !is.na(wdpa_orig_name) ~ wdpa_orig_name,
      is.na(wdpa_orig_name) & !is.na(i_efectividad_name) ~ i_efectividad_name,
      TRUE ~ lasting_special_place_description
    ),
    lsp_name = normalize_name(lsp_name),
    lsp_name = str_replace_all(lsp_name, "LA TOVARA", "La Tovara")
  ) %>%
  dplyr::select(rgn_id, lsp_name, condition = overall_condition)

dive_sites_condition_join <- dive_sites_condition %>%
  mutate(lsp_name = paste0("Dive Site: ", nombre_sitio_de_buceo)) %>%
  dplyr::select(rgn_id, lsp_name, condition = overall_condition)

all_categorical_lsps <- bind_rows(
  fishing_site_condition_join,
  archaeological_site_condition_join,
  protected_area_condition_join,
  dive_sites_condition_join
)

lsps_not_in_categories <- setdiff(
  lsp_list_ewg$lasting_special_place_lsp,
  all_categorical_lsps$lsp_name
)
```

## Missing marine sites to score here

Now that I know which names are already covered, I will drop habitat rows that script 6 already covers with polygons. Also note: `rgn_id` should already live on the inventory when exported from the sheets.

```{r}
habitat_names_to_skip <- c(
  "Coquina Reefs in Libertad, Bahia Lobos, and Peñasco - AHW",
  "Only coral reefs in Gulf (Cabo Pulmo/Los Frailes & Bahía San Gabriel on Is. Espíritu Santo). RCB, BW",
  "Massive seagrass beds (Zostera marina), RCB, BW\n\nThe largest and most stable annual seagrass beds (Zostera marina) are in the eastern Pacific. The eelgrass in the Gulf of California is 100% annual, no perennial, and is considered a subspecies, JTC, BW",
  "Puerto Lobos N most mangrove (black mangrove) population, BW",
  "Mangrove coastal lagoons, JTC."
)

lsps_missing_condition <- lsp_list_ewg %>%
  filter(lasting_special_place_lsp %in% lsps_not_in_categories) %>%
  filter(type_of_lsp %in% "Marine location of importance") %>%
  mutate(
    lat = as.numeric(lat),
    long = as.numeric(long)
  ) %>%
  filter(!is.na(lat), !is.na(long)) %>%
  st_as_sf(coords = c("long", "lat"), crs = 4326, remove = FALSE) %>%
  filter(!lasting_special_place_lsp %in% habitat_names_to_skip) %>%
  dplyr::select(rgn_id, lsp_name = lasting_special_place_lsp) %>%
  mutate(rgn_id = str_replace_all(rgn_id, " ", "_"))

mapview(lsps_missing_condition)
```

## Protected areas and spatial join

Alrighty, now I will buffer the leftover sites and join them to the combined protected area layer so I know which parks they touch.

```{r}
all_protected <- st_read(
  here(
    "/home/shares/ohi/OHI_GOC/spatial/protected_areas_wdpa_conanp_and_hem",
    "all_goc_protected_areas_v2_2025.shp"
  ),
  quiet = TRUE
)

lsps_missing_buffered <- st_buffer(
  st_transform(lsps_missing_condition, st_crs(all_protected)),
  dist = 100
)

lsps_missing_protected <- st_join(
  lsps_missing_buffered,
  all_protected %>% dplyr::select(name, geometry),
  join = st_intersects,
  left = TRUE
) %>%
  mutate(in_protected_area = !is.na(name)) %>%
  rename(protected_area_name = name)
```

## Match overlapping parks to SIMEC

Now that I know which parks these leftover sites touch, I will prepare SIMEC for **2020 and 2023** the same way as in script 0 (averaging 2020 island-by-state rows).

```{r}
# status years i want lsp scores for
status_years <- c(2020L, 2023L)

# prepare simec for 2020 and 2023 (same logic as script 0)
# older 2020 reports split islas del golfo by state/region; i average those rows
i_efectividad_by_year <- read_csv(
  here(
    "/home/shares/ohi/OHI_GOC/_raw_data/SIMEC",
    "simec_i_efectividad_anp_2026.csv"
  ),
  show_col_types = FALSE,
  na = c("", "NA", "N/A")
) %>%
  janitor::clean_names() %>%
  mutate(
    nombre = str_squish(as.character(nombre)),
    ano_reporte = suppressWarnings(as.integer(ano_reporte)),
    i_efectividad = suppressWarnings(as.numeric(i_efectividad)),
    nombre = if_else(
      str_to_upper(nombre) == "ISLAS DEL GOLFO DE CALIFORNIA",
      "ISLAS DEL GOLFO DE CALIFORNIA",
      nombre
    )
  ) %>%
  filter(
    !is.na(ano_reporte),
    !is.na(i_efectividad),
    !is.na(nombre),
    nombre != "",
    ano_reporte %in% status_years
  ) %>%
  group_by(ano_reporte, nombre) %>%
  summarize(
    i_efectividad = mean(i_efectividad, na.rm = TRUE),
    .groups = "drop"
  )

# year-specific simec min and max (for ramsar rules / nom floors)
simec_bounds_by_year <- i_efectividad_by_year %>%
  group_by(ano_reporte) %>%
  summarize(
    simec_max = max(i_efectividad, na.rm = TRUE),
    simec_min = min(i_efectividad, na.rm = TRUE),
    .groups = "drop"
  )

i_efectividad_norm <- i_efectividad_by_year %>%
  distinct(nombre) %>%
  mutate(nombre_norm = normalize_name(nombre))

lsp_missing_sites_norm <- lsps_missing_protected %>%
  mutate(protected_area_norm = normalize_name(protected_area_name))

matches <- stringdist::amatch(
  lsp_missing_sites_norm$protected_area_norm,
  i_efectividad_norm$nombre_norm,
  method = "jw",
  maxDist = 0.2
)

lsp_missing_base <- lsp_missing_sites_norm %>%
  mutate(
    nombre_matched = i_efectividad_norm$nombre[matches],
    nombre_norm_matched = i_efectividad_norm$nombre_norm[matches]
  ) %>%
  st_drop_geometry()

# expand to both status years, then join that year's simec score
lsp_missing_w_SIMEC <- bind_rows(
  lsp_missing_base %>% mutate(ano_reporte = 2020L),
  lsp_missing_base %>% mutate(ano_reporte = 2023L)
) %>%
  left_join(
    i_efectividad_by_year,
    by = c("nombre_matched" = "nombre", "ano_reporte")
  ) %>%
  left_join(simec_bounds_by_year, by = "ano_reporte") %>%
  mutate(
    year = ano_reporte,
    protected_area_condition = ifelse(!is.na(i_efectividad), i_efectividad, NA_real_)
  )

ramsar_management_plans <- read_csv(
  here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/google_sheets", "ramsar_site_management_plan_presence.csv"),
  show_col_types = FALSE
)

# ramsar only when i have a park name but no simec match
lsp_missing_RAMSAR <- lsp_missing_w_SIMEC %>%
  filter(!is.na(protected_area_name) & is.na(nombre_norm_matched)) %>%
  left_join(
    ramsar_management_plans,
    by = c("protected_area_name" = "wdpa_orig_name")
  ) %>%
  mutate(
    protected_area_condition = case_when(
      if_ramsar_has_management_plan == "y" ~ simec_max,
      if_ramsar_has_management_plan == "in_progress" ~ 60,
      if_ramsar_has_management_plan == "n" ~ simec_min,
      TRUE ~ NA_real_
    )
  ) %>%
  dplyr::select(
    year, rgn_id, lsp_name, protected_area_condition,
    if_ramsar_has_management_plan, protected_area_name,
    in_protected_area, protected_area_norm
  )

lsp_missing_only_conanp_ANPs <- lsp_missing_w_SIMEC %>%
  filter(!protected_area_name %in% lsp_missing_RAMSAR$protected_area_name | is.na(protected_area_name)) %>%
  dplyr::select(
    year, rgn_id, lsp_name, protected_area_condition,
    protected_area_name, in_protected_area, protected_area_norm
  )

final_lsp_missing_condition <- bind_rows(
  lsp_missing_only_conanp_ANPs,
  lsp_missing_RAMSAR
) %>%
  group_by(year, rgn_id, lsp_name) %>%
  summarize(
    protected_area_condition = max(protected_area_condition, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    protected_area_condition = ifelse(
      is.infinite(protected_area_condition) | is.nan(protected_area_condition),
      0,
      protected_area_condition
    ),
    lsp_name_join = stringi::stri_trim_both(lsp_name)
  )
```

## Codified scores when still at zero

Google sheet (estimated condition of protection):  
<https://docs.google.com/spreadsheets/d/1RwNTzRZdBIDtzi7d1ceFxhNmVghN47NxTgrhZV8O2SM/edit?gid=193789975#gid=193789975>

Codified scores do not change by year; SIMEC / Ramsar scores do. Now that I have the park-based scores, I will fill any remaining zeros from the curated sheet, then average by region and year and write the tables to Mazu.

```{r}
other_protections_condition_score <- read_csv(
  here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/google_sheets", "lsps_codified_protections_estimated_condition.csv"),
  show_col_types = FALSE
) %>%
  mutate(lsp_name_join = stringi::stri_trim_both(lsp_name)) %>%
  dplyr::select(-lsp_name)

lsps_other_protections_condition <- left_join(
  final_lsp_missing_condition,
  other_protections_condition_score,
  by = "lsp_name_join"
)

final_other_protections_condition <- lsps_other_protections_condition %>%
  mutate(
    overall_condition = ifelse(
      protected_area_condition == 0,
      estimated_condition_of_protection,
      protected_area_condition
    ),
    overall_condition = as.numeric(overall_condition)
  )

rgn_final_other_protections_condition <- final_other_protections_condition %>%
  group_by(year, rgn_id) %>%
  summarize(condition = mean(overall_condition, na.rm = TRUE), .groups = "drop")


write_csv(
  final_other_protections_condition,
  here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2026/output", "condition_lsp_other_marine_places.csv")
)
write_csv(
  rgn_final_other_protections_condition,
  here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2026/output", "condition_region_other_marine_places.csv")
)
```


