---
title: 'OHI GoC 2024: Lasting Special Places (LSP)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor:
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options:
  chunk_output_type: console
---

# Script 0: Prepare multi-year SIMEC i-efectividad

In this script I prepare CONANP's SIMEC i-efectividad scores for **two status years**: **2020** and **2023**. Scripts 2 through 7 join those year-specific scores onto lasting special places, and script 8 averages them into LSP current status for each year.

**What I do**

1. Read the multi-year SIMEC extract (2020 regional reports and 2023 park reports).
2. Average **Islas del Golfo de California** when the 2020 reports split that park by state / region.
3. Keep both 2020 and 2023 as clean one-row-per-park tables.
4. Write the timeseries and the year-specific tables to Mazu.

```{r}
library(tidyverse)
library(here)
```

## Why this prep is needed

CONANP published i-efectividad scores in **2020 regional reports** and again in **2023 park reports**. That gives us two time points for management effectiveness.

The awkward case is **Islas del Golfo de California**:

- In 2020, the islands are scored separately for different states / OHI regions.
- In 2023, there is one park-wide score.

So if I join the raw 2020 rows as-is, I get three island scores for one park name. I average those island rows within 2020 into one park-level score, the same way the 2023 report treats the park.

## Read the multi-year extract

I start from the SIMEC csv. `janitor::clean_names()` makes the headers easier to work with in tidyverse.

```{r}
# status years i want lsp scores for
status_years <- c(2020L, 2023L)

i_efectividad_multi_year <- read_csv(
  here(
    "/home/shares/ohi/OHI_GOC/_raw_data/SIMEC",
    "simec_i_efectividad_anp_2026.csv"
  ),
  show_col_types = FALSE,
  na = c("", "NA", "N/A")
) %>%
  janitor::clean_names() %>%
  mutate(
    nombre = str_squish(as.character(nombre)),
    ano_reporte = suppressWarnings(as.integer(ano_reporte)),
    i_efectividad = suppressWarnings(as.numeric(i_efectividad)),
    mett = suppressWarnings(as.numeric(mett)),
    # unify island spelling (some rows have a trailing space)
    nombre = if_else(
      str_to_upper(nombre) == "ISLAS DEL GOLFO DE CALIFORNIA",
      "ISLAS DEL GOLFO DE CALIFORNIA",
      nombre
    )
  ) %>%
  # drop rows with no usable score (for example constitucion de 1857 in 2020)
  filter(
    !is.na(ano_reporte),
    !is.na(nombre),
    nombre != "",
    !is.na(i_efectividad),
    ano_reporte %in% status_years
  )

summary(i_efectividad_multi_year)

# quick look at the island rows before i average them
i_efectividad_multi_year %>%
  filter(nombre == "ISLAS DEL GOLFO DE CALIFORNIA") %>%
  select(ano_reporte, nombre, rgn_id, i_efectividad, vinculo_reporte)
```

## Average parks that have more than one row in the same year

For most parks this does nothing. For **Islas del Golfo de California** in 2020, this turns the Region_2 / Region_4 / Region_8 scores into one park-level mean. I also set `rgn_id` to `"all"` when more than one region contributed, matching the 2023 reporting style.

```{r}
i_efectividad_by_year <- i_efectividad_multi_year %>%
  group_by(ano_reporte, nombre) %>%
  summarize(
    i_efectividad = mean(i_efectividad, na.rm = TRUE),
    mett = mean(mett, na.rm = TRUE),
    ano_decreto = dplyr::first(ano_decreto),
    vinculo_reporte = dplyr::first(vinculo_reporte),
    rgn_id = if_else(
      n_distinct(rgn_id) > 1 | any(rgn_id == "all", na.rm = TRUE),
      "all",
      as.character(dplyr::first(rgn_id))
    ),
    n_rows_averaged = n(),
    .groups = "drop"
  ) %>%
  arrange(nombre, ano_reporte)

# check that 2020 islands are now one averaged row, and 2023 is still one row
i_efectividad_by_year %>%
  filter(nombre == "ISLAS DEL GOLFO DE CALIFORNIA") %>%
  select(ano_reporte, nombre, rgn_id, i_efectividad, n_rows_averaged)

i_efectividad_by_year %>%
  count(ano_reporte)
```

## Split into year-specific tables

Scripts 2 through 7 can either join the full `i_efectividad_by_year` table (two rows per park) or read the year-specific csvs below.

```{r}
i_efectividad_2020 <- i_efectividad_by_year %>%
  filter(ano_reporte == 2020L) %>%
  arrange(nombre)

i_efectividad_2023 <- i_efectividad_by_year %>%
  filter(ano_reporte == 2023L) %>%
  arrange(nombre)

i_efectividad_2020 %>%
  select(nombre, ano_reporte, i_efectividad) %>%
  print(n = Inf)

i_efectividad_2023 %>%
  select(nombre, ano_reporte, i_efectividad) %>%
  print(n = Inf)

# parks present in 2020 but missing in 2023 (expect none)
setdiff(unique(i_efectividad_2020$nombre), unique(i_efectividad_2023$nombre)) # character(0)
```

## Write prepared tables to Mazu

```{r}
write_csv(
  i_efectividad_by_year,
  here(
    "/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2026/int",
    "simec_i_efectividad_timeseries.csv"
  )
)

write_csv(
  i_efectividad_2020,
  here(
    "/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2026/int",
    "simec_i_efectividad_2020.csv"
  )
)

write_csv(
  i_efectividad_2023,
  here(
    "/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2026/int",
    "simec_i_efectividad_2023.csv"
  )
)
```

Next I run scripts 2 through 7. Each of those scripts prepares the same 2020 and 2023 SIMEC tables inline, scores lasting special places for both years, and writes a `year` column so script 8 can build LSP status for 2020 and 2023.
