---
title: 'OHI GoC 2024: Lasting Special Places (LSP)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor:
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options:
  chunk_output_type: console
---

# Sentido de Pertenencia: Lasting Special Places

## Script 2: special protected areas on the EWG list

Alrighty, in this script I will score the protected areas the EWG named explicitly. When CONANP reports a SIMEC i-efectividad score, that becomes the condition. For Ramsar-only sites I will follow simple rules based on whether a management plan is present. I will do this for **both 2020 and 2023**, then write site and region tables with a `year` column so script 8 can merge both years.

**What I do**

1. Read the combined protected area layer (`all_goc_protected_areas_v2_2025.shp`).
2. Read the EWG special protected areas table.
3. Prepare SIMEC for 2020 and 2023 (average 2020 island-by-state rows).
4. Join SIMEC onto the EWG list for each year, apply Ramsar rules, average by region and year, write CSVs.

```{r}
library(tidyverse)
library(here)
library(sf)
library(mapview)
```

## Protected areas (ANPs)

Protected areas on the list are already protected by design. I still need a condition score that reflects management quality, and for that I will rely mainly on CONANP's SIMEC i-efectividad.

### Read the combined protected area layer

Now I will load the v2 layer from script 1 (WDPA + CONANP + HEM). This is the same layer scripts 3 through 7 should use.

```{r}
all_protected <- st_read(
  here(
    "/home/shares/ohi/OHI_GOC/spatial/protected_areas_wdpa_conanp_and_hem",
    "all_goc_protected_areas_v2_2025.shp"
  ),
  quiet = TRUE
)

mapview(all_protected)
```

Optional: older CBMC GCMP MPAs for context (sites designated before 2020).

```{r}
cbmc_gcmp_2020_mpas <- st_read(
  here(
    "/home/shares/ohi/OHI_GOC/_raw_data/CBMC-GCMP-prioritization-of-diving-sites-main/shp",
    "final_mpa_layer.shp"
  ),
  quiet = TRUE
) %>%
  janitor::clean_names() %>%
  st_zm(drop = TRUE, what = "ZM")

table(st_geometry_type(cbmc_gcmp_2020_mpas))
mapview(cbmc_gcmp_2020_mpas)
```

## Read the EWG list of protected areas

Next I will pull in the EWG list so I have the expert names, WDPA labels, SIMEC names, Ramsar flags, and region ids.

```{r}
ewg_mentioned_protected_areas <- read_csv(
  here(
    "/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/google_sheets",
    "ewg_special_protected_areas_on_list.csv"
  ),
  show_col_types = FALSE
) %>%
  janitor::clean_names()
```

## Prepare SIMEC i-efectividad for 2020 and 2023

Now that I have the EWG list, I will prepare the multi-year SIMEC extract. Older 2020 reports split **Islas del Golfo de California** by state / region; the 2023 report gives one park-wide score. So first I will average any park that has more than one row in the same year, then keep both status years. (Script 0 walks through this prep in more detail.)

```{r}
# status years i want lsp scores for
status_years <- c(2020L, 2023L)

i_efectividad_by_year <- read_csv(
  here(
    "/home/shares/ohi/OHI_GOC/_raw_data/SIMEC",
    "simec_i_efectividad_anp_2026.csv"
  ),
  show_col_types = FALSE,
  na = c("", "NA", "N/A")
) %>%
  janitor::clean_names() %>%
  mutate(
    nombre = str_squish(as.character(nombre)),
    ano_reporte = suppressWarnings(as.integer(ano_reporte)),
    i_efectividad = suppressWarnings(as.numeric(i_efectividad)),
    # unify island spelling (some 2020 rows have a trailing space)
    nombre = if_else(
      str_to_upper(nombre) == "ISLAS DEL GOLFO DE CALIFORNIA",
      "ISLAS DEL GOLFO DE CALIFORNIA",
      nombre
    )
  ) %>%
  filter(
    !is.na(ano_reporte),
    !is.na(i_efectividad),
    !is.na(nombre),
    nombre != "",
    ano_reporte %in% status_years
  ) %>%
  # average multi-row parks within a year (mainly 2020 islands)
  group_by(ano_reporte, nombre) %>%
  summarize(
    i_efectividad = mean(i_efectividad, na.rm = TRUE),
    .groups = "drop"
  )

# year-specific simec min and max (for ramsar rules)
simec_bounds_by_year <- i_efectividad_by_year %>%
  group_by(ano_reporte) %>%
  summarize(
    simec_max = max(i_efectividad, na.rm = TRUE),
    simec_min = min(i_efectividad, na.rm = TRUE),
    .groups = "drop"
  )
```

## Assign condition for each year (SIMEC and Ramsar rules)

Alrighty, now that SIMEC is ready, I will duplicate the EWG list once per status year, join that year's SIMEC score, then apply Ramsar management plan rules and a few special cases (whale shark refuge, vaquita refuge, Loreto II).

```{r}
protected_area_condition <- bind_rows(
  ewg_mentioned_protected_areas %>% mutate(ano_reporte = 2020L),
  ewg_mentioned_protected_areas %>% mutate(ano_reporte = 2023L)
) %>%
  left_join(
    i_efectividad_by_year,
    by = c("i_efectividad_name" = "nombre", "ano_reporte")
  ) %>%
  left_join(simec_bounds_by_year, by = "ano_reporte") %>%
  mutate(
    overall_condition = ifelse(is.na(i_efectividad), 0, i_efectividad),
    overall_condition = case_when(
      # ramsar plan in progress: take the better of simec and 60
      if_ramsar_has_management_plan == "in_progress" ~
        pmax(i_efectividad, 60, na.rm = TRUE),
      # ramsar plan yes: use at least the highest simec in that year
      if_ramsar_has_management_plan == "y" ~
        pmax(i_efectividad, simec_max, na.rm = TRUE),
      # ramsar plan no: cap at the lowest simec in that year
      if_ramsar_has_management_plan == "n" ~
        pmin(i_efectividad, simec_min, na.rm = TRUE),
      lasting_special_place_description == "Whale shark Area de Refugio in La Paz - AHW , BW" ~ 75,
      # loreto ii is new; leave na until simec is published
      lasting_special_place_description == "Loreto II" ~ NA_real_,
      lasting_special_place_description == "Área de Refúgio para la Protección de la Vaquita" ~ 75,
      TRUE ~ overall_condition
    ),
    rgn_id = str_replace_all(rgn_id, " ", "_"),
    year = ano_reporte
  ) %>%
  dplyr::select(
    year,
    rgn_id,
    lasting_special_place_description,
    overall_condition,
    i_efectividad_name,
    wdpa_orig_name,
    if_ramsar_has_management_plan,
    ramsar_link
  )
```

Now that each site has a condition, I will attach geometry from `all_protected` by WDPA name so I can map the latest year.

```{r}
protected_condition_w_geometry <- protected_area_condition %>%
  filter(year == 2023L) %>%
  left_join(
    all_protected,
    by = c("wdpa_orig_name" = "name")
  ) %>%
  st_as_sf()

mapview(protected_condition_w_geometry, zcol = "overall_condition")
```

## Condition per region and year

Now that I have a condition for every protected area site, I will average those site conditions within each OHI region for each status year.

```{r}
rgn_protected_area_condition <- protected_area_condition %>%
  group_by(year, rgn_id) %>%
  summarize(
    condition = mean(overall_condition, na.rm = TRUE),
    .groups = "drop"
  )
```

## Write to Mazu

Finally, I will save the site-level and region-level tables for scripts 7 and 8. Both include a `year` column (2020 and 2023).

```{r}
write_csv(
  protected_area_condition,
  here(
    "/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2026/output",
    "condition_lsp_protected_areas.csv"
  )
)

write_csv(
  rgn_protected_area_condition,
  here(
    "/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2026/output",
    "condition_region_protected_areas.csv"
  )
)
```


