---
title: 'OHI GoC 2024: Lasting Special Places (LSP)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor:
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options:
  chunk_output_type: console
---

# Sentido de Pertenencia: Lasting Special Places

## Script 5: fishing sites

In this script, I will score fishing grounds the EWG flagged as iconic. I will check whether each site sits in a fishing refuge (ZRP), a cooperative TURF, or a federal protected area. Those institutional layers tell me the place has some form of stewardship beyond open access.

**What I do**

1. Filter the EWG list to fishing area rows and assign each site an OHI region.
2. Spatial join TNC ZRP polygons (with a 1 km buffer).
3. Flag the one site I confirmed sits inside a TURF using the Aceves-Bueno Shiny map.
4. Join protected areas for SIMEC scores, merge the best protection, average by region, write CSVs.

```{r}
library(tidyverse)
library(here)
library(sf)
library(mapview)
library(stringdist)
library(stringi)
```

## Why these sites matter

Fishing can be subsistence, recreation, or tradition. For lasting special places I care about whether the ground itself is managed so it can remain usable into the future. The EWG named each site.

## Read fishing sites and assign regions

```{r}
lsp_list_ewg <- read_csv(
  here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/google_sheets", "ewg_lsp_inventory_prepared.csv"),
  show_col_types = FALSE
) %>%
  janitor::clean_names()

ewg_fishing_site_list <- lsp_list_ewg %>%
  filter(
    type_of_lsp %in% c(
      '"Important fishing areas, could have fishing refuges (zonas de refugio pesquero or ZRP)"'
    )
  ) %>%
  mutate(
    long = as.numeric(long),
    lat = as.numeric(lat)
  ) %>%
  filter(!is.na(long), !is.na(lat)) %>%
  st_as_sf(coords = c("long", "lat"), crs = 4326, remove = FALSE)

mapview(ewg_fishing_site_list)

ohi_rgns <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp"), quiet = TRUE)

ewg_fishing_site_list_goc_rgnid <- st_intersection(
  st_transform(ewg_fishing_site_list, st_crs(ohi_rgns)),
  ohi_rgns
)

mapview(ewg_fishing_site_list_goc_rgnid, zcol = "rgn_id")
```

## ZRPs (fishing refuges)

Now I will pull TNC's geoportal GeoJSON and buffer sites by 1 km so points near a refuge boundary still count.

```{r}
tnc_zrps_raw <- st_read(
  here(
    "/home/shares/ohi/OHI_GOC/_raw_data/TNC_geoportal_fishing_refuge_zones/d2026",
    "Zonas_Refugios_Pesqueros_-8805549763229037385.geojson"
  ),
  quiet = TRUE
) %>%
  janitor::clean_names()

mapview(tnc_zrps_raw, col.regions = "blue") +
  mapview(ewg_fishing_site_list, col.regions = "black")

inland_50 <- st_read(
  here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer", "inland_50km_goc.shp"),
  quiet = TRUE
)

ewg_fishing_site_buffered <- st_buffer(
  st_transform(ewg_fishing_site_list_goc_rgnid, st_crs(inland_50)),
  dist = 1000
)

ewg_fishing_site_condition_zrp <- st_join(
  st_transform(ewg_fishing_site_buffered, st_crs(tnc_zrps_raw)),
  st_make_valid(tnc_zrps_raw),
  join = st_intersects,
  left = TRUE
) %>%
  mutate(in_zrp = !is.na(name)) %>%
  rename(zrp_name = name) %>%
  mutate(zrp_condition = ifelse(in_zrp, 100, 0)) %>%
  dplyr::select(rgn_id, lasting_special_place_lsp, in_zrp, zrp_condition) %>%
  distinct()
```

## TURFs (cooperative tenure)

National TURF polygons are not easy to download. I used the Shiny map from Aceves-Bueno et al. 2023 and confirmed one EWG site sits inside a TURF: the Yaqui and Mayo fishing camps at the Sonora Sinaloa border. Now I will set that site to 100. If more sites turn up later, I can extend this `ifelse`.

Papers and map:  
<https://journals.plos.org/plosone/article/figure?id=10.1371/journal.pone.0286739.g003>  
<https://ereaceves.shinyapps.io/Tapp>

```{r}
ewg_fishing_site_condition_turf <- ewg_fishing_site_condition_zrp %>%
  mutate(
    in_turf = lasting_special_place_lsp ==
      "Yaqui and Mayo People have historic (and current) important fishing camps at Sonora-Sinala border. RCB",
    overall_condition = ifelse(in_turf, 100, zrp_condition)
  )
```

## Protected area (SIMEC) scores

If ZRP and TURF both score 0, a site can still sit inside an ANP and inherit SIMEC. Now I will join the v2 protected area layer and fuzzy match SIMEC names the same way as in scripts 3 and 4.

```{r}
all_protected <- st_read(
  here(
    "/home/shares/ohi/OHI_GOC/spatial/protected_areas_wdpa_conanp_and_hem",
    "all_goc_protected_areas_v2_2025.shp"
  ),
  quiet = TRUE
)

ewg_fishing_site_condition_ANP <- st_join(
  st_transform(ewg_fishing_site_condition_turf, st_crs(all_protected)),
  all_protected %>% dplyr::select(name, geometry),
  join = st_intersects,
  left = TRUE
) %>%
  mutate(in_protected_area = !is.na(name)) %>%
  rename(protected_area_name = name)

mapview(ewg_fishing_site_condition_ANP, zcol = "in_protected_area") +
  mapview(all_protected, col.regions = "pink")
```

## Match overlapping parks to SIMEC

Alrighty, now that I know which parks each fishing site touches, I will prepare SIMEC for **2020 and 2023** the same way as in script 0 (averaging 2020 island-by-state rows). Then I will fuzzy-match park names and assign scores for each year.

```{r}
# status years i want lsp scores for
status_years <- c(2020L, 2023L)

# prepare simec for 2020 and 2023 (same logic as script 0)
# older 2020 reports split islas del golfo by state/region; i average those rows
i_efectividad_by_year <- read_csv(
  here(
    "/home/shares/ohi/OHI_GOC/_raw_data/SIMEC",
    "simec_i_efectividad_anp_2026.csv"
  ),
  show_col_types = FALSE,
  na = c("", "NA", "N/A")
) %>%
  janitor::clean_names() %>%
  mutate(
    nombre = str_squish(as.character(nombre)),
    ano_reporte = suppressWarnings(as.integer(ano_reporte)),
    i_efectividad = suppressWarnings(as.numeric(i_efectividad)),
    nombre = if_else(
      str_to_upper(nombre) == "ISLAS DEL GOLFO DE CALIFORNIA",
      "ISLAS DEL GOLFO DE CALIFORNIA",
      nombre
    )
  ) %>%
  filter(
    !is.na(ano_reporte),
    !is.na(i_efectividad),
    !is.na(nombre),
    nombre != "",
    ano_reporte %in% status_years
  ) %>%
  group_by(ano_reporte, nombre) %>%
  summarize(
    i_efectividad = mean(i_efectividad, na.rm = TRUE),
    .groups = "drop"
  )

# year-specific simec min and max (for ramsar rules / nom floors)
simec_bounds_by_year <- i_efectividad_by_year %>%
  group_by(ano_reporte) %>%
  summarize(
    simec_max = max(i_efectividad, na.rm = TRUE),
    simec_min = min(i_efectividad, na.rm = TRUE),
    .groups = "drop"
  )

normalize_name <- function(x) {
  x %>%
    stringi::stri_trans_general("Latin-ASCII") %>%
    str_to_upper() %>%
    str_remove_all(
      "\\bPARQUE NACIONAL\\b|\\bPARQUE\\b|\\bNACIONAL\\b|\\bZONA MARINA\\b|\\bARCHIPIELAGO\\b"
    ) %>%
    str_squish()
}

# fuzzy match on unique park names
i_efectividad_norm <- i_efectividad_by_year %>%
  distinct(nombre) %>%
  mutate(nombre_norm = normalize_name(nombre))

fishing_sites_norm <- ewg_fishing_site_condition_ANP %>%
  mutate(protected_area_norm = normalize_name(protected_area_name))

matches <- stringdist::amatch(
  fishing_sites_norm$protected_area_norm,
  i_efectividad_norm$nombre_norm,
  method = "jw",
  maxDist = 0.2
)

fishing_sites_base <- fishing_sites_norm %>%
  mutate(
    nombre_matched = i_efectividad_norm$nombre[matches],
    nombre_norm_matched = i_efectividad_norm$nombre_norm[matches]
  ) %>%
  st_drop_geometry()

# expand to both status years, then join that year's simec score
fishing_sites_w_SIMEC <- bind_rows(
  fishing_sites_base %>% mutate(ano_reporte = 2020L),
  fishing_sites_base %>% mutate(ano_reporte = 2023L)
) %>%
  left_join(
    i_efectividad_by_year,
    by = c("nombre_matched" = "nombre", "ano_reporte")
  ) %>%
  left_join(simec_bounds_by_year, by = "ano_reporte") %>%
  mutate(
    protected_area_condition = ifelse(!is.na(i_efectividad), i_efectividad, NA_real_),
    # ramsar with a management plan: high score for that year
    protected_area_condition = ifelse(
      protected_area_name == "Humedales del Delta del Río Colorado",
      simec_max,
      protected_area_condition
    ),
    # cabo pulmo is no take; leave it for careful manual handling
    protected_area_condition = ifelse(
      nombre_matched == "CABO PULMO",
      0,
      protected_area_condition
    ),
    year = ano_reporte
  ) %>%
  group_by(year, lasting_special_place_lsp) %>%
  summarize(
    protected_area_condition = max(protected_area_condition, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    protected_area_condition = ifelse(
      is.infinite(protected_area_condition),
      NA_real_,
      protected_area_condition
    )
  )
```

## Final site condition and region average

Now that I have ZRP, TURF, and SIMEC scores, I will start from ZRP or TURF (same in both years). When that is still 0, I will fill in that year's SIMEC. Today all sites have some of these layers, so I do not need a separate codified floor (though NOM links for clams and squid are useful if empty sites appear later). After that I will average site conditions within each OHI region for each status year and write the tables to Mazu.

```{r}
# zrp / turf scores do not change by year, so i expand them to 2020 and 2023
fishing_base_by_year <- bind_rows(
  ewg_fishing_site_condition_turf %>% st_drop_geometry() %>% mutate(year = 2020L),
  ewg_fishing_site_condition_turf %>% st_drop_geometry() %>% mutate(year = 2023L)
)

final_ewg_fishing_site_condition <- fishing_base_by_year %>%
  left_join(
    fishing_sites_w_SIMEC,
    by = c("lasting_special_place_lsp", "year")
  ) %>%
  mutate(
    overall_condition = ifelse(
      overall_condition == 0,
      protected_area_condition,
      overall_condition
    )
  )

rgn_final_ewg_fishing_site_condition <- final_ewg_fishing_site_condition %>%
  group_by(year, rgn_id) %>%
  summarize(condition = mean(overall_condition, na.rm = TRUE), .groups = "drop")
```


```{r}
write_csv(
  final_ewg_fishing_site_condition,
  here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2026/output", "condition_lsp_fishing_sites.csv")
)
write_csv(
  rgn_final_ewg_fishing_site_condition,
  here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2026/output", "condition_region_fishing_sites.csv")
)
```
