---
title: "Soft bottom - Gulf of California OHI"
author: "Austin Martinez"
date: 07/10/2026
format: 
  html:
    code-fold: show
execute:
  echo: true
  message: false
  warning: false
---


```{r}
# Load in Libraries
library(sf)
library(dplyr)
library(here)
library(terra)
library(tidyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(tmap)
library(readr)
```

```{r}
# Target CRS for all vector data (matches OHI regions + fishing rasters)
target_crs <- "EPSG:6372"  # Mexico ITRF2008 / LCC

# Master analysis grid = fishing effort (already EPSG:6372, 1 km)
years <- 2007:2025
fishing_rasters <- rast(
  paste0(
    "/home/shares/ohi/OHI_GOC/goal_prep/pressures/vessels/benthic_destructive_fishing/destructive_fishing_hours_",
    years,
    "_1km.tif"
  )
)
nlyr(fishing_rasters)
```

```{r}
# Read OHI regions and transform to Mexico ITRF2008 / LCC before any analysis
ohi_regions <- st_read(here("spatial", "ohi_regions", "goc_ohi_rgns.shp")) |>
  st_make_valid() |>
  st_transform(target_crs)

ohi_land   <- ohi_regions |> dplyr::filter(location == "land")
ohi_marine <- ohi_regions |> dplyr::filter(location == "marine")

```

```{r}
# Read habitat shapefiles and transform to EPSG:6372 before rasterizing
rocky_reef <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/Hem_data/goc_habitat_data/Arrecife rocoso/Arrecife rocoso.shp")) |>
  st_make_valid() |>
  st_transform(target_crs)

seamounts <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/Hem_data/goc_habitat_data/seamount/seamount_merge.shp")) |>
  st_make_valid() |>
  st_transform(target_crs)

```

```{r}
# Clip seamounts to OHI regions (same CRS)
seamounts <- st_intersection(seamounts, ohi_regions)
```

## rocky_reef work:
```{r}
# Find proportion of polygon that is rocky reef (areas in EPSG:6372 meters)
rocky_reef <- rocky_reef |>
  mutate(
    rocky_km2 = rocky_ha / 100, # Convert rocky_ha from hectares to km2
    polygon_km2 = as.numeric(st_area(geometry)) / 1e6, # Find area of polygons
    prop_rocky_reef = rocky_km2 / polygon_km2 # Find proportion of polygon that is rocky reef
  )
```


```{r}
# 100 m template nested under the fishing 1 km grid (same CRS, origin, extent)
# so aggregate(fact = 10) lands exactly on the fishing cells
template_100m <- disagg(rast(fishing_rasters[[1]]), fact = 10)
```

# Rocky reef rasterization:
```{r}
# Rasterize rocky reef polygon proportions
rocky_reef_prop_r <- rasterize(vect(rocky_reef), template_100m, field = "prop_rocky_reef", background = 0)

```

```{r}
# Check to see if the min and max are between 0-1
max(rocky_reef_prop_r)
```


```{r}
# Rasterize the vector seamounts data
seamounts_r  <- rasterize(vect(seamounts),  template_100m, field = 1, background = 0)
```


## Quick validation
```{r}
# Check if values are between 0-1
max(seamounts_r)
```

```{r}
# Both layers already rasterized onto template_100m — confirm alignment
compareGeom(rocky_reef_prop_r, template_100m, stopOnError = TRUE)
compareGeom(seamounts_r, template_100m, stopOnError = TRUE)
```

```{r}
# Make raster where 0 = no hard‑bottom and 1+ = hard‑bottom present
hard_bottom_100m <- rocky_reef_prop_r +
                    seamounts_r
```

```{r}
# Check if values are between 0-1
max(hard_bottom_100m)
```

```{r}
# Cap values at 1 to make a proportion
hard_bottom_100m[hard_bottom_100m > 1] <- 1
```


```{r}
# Groups every 10×10 block of 100‑m cells into one 1‑km cell, and computes the mean of values inside that block
hard_bottom_fraction <- aggregate(hard_bottom_100m, fact = 10, fun = "mean")

# flips the fraction so more soft bottom = 1
soft_bottom_fraction <- 1 - hard_bottom_fraction
```

```{r}
# Check if values are between 0-1
max(soft_bottom_fraction)
```



```{r}
# Mask to the OHI marine region extent
marine_mask <- rasterize(
  vect(ohi_marine),
  soft_bottom_fraction,
  field = 1,
  background = NA
)

soft_bottom_fraction <- mask(soft_bottom_fraction, marine_mask)

```

Make plot of soft-bottom area
```{r}
# Land first (gray), then soft-bottom raster, then marine borders
plot(vect(ohi_land), col = "gray85", border = "gray60", lwd = 0.5,
     main = "Soft-bottom fraction", axes = FALSE, box=FALSE)
plot(
  soft_bottom_fraction,
  add = TRUE,
  col = hcl.colors(100, "viridis"),
  range = c(0, 1),
  legend = TRUE,
  axes = FALSE, 
  box = FALSE
)
plot(vect(ohi_marine), add = TRUE, border = "gray70", lwd = 0.7, col = NA, box=FALSE, axes=FALSE)
```


## Redundancy Check
-   Is the soft bottom layer resolution 1km
-   Are all values between 0-1
```{r}
soft_res <- res(soft_bottom_fraction)
print(paste("Resolution (m):", soft_res))
max(soft_bottom_fraction, na.rm = TRUE)
```

## Destructive fishing data

Fishing rasters were loaded above (master grid in EPSG:6372). Soft-bottom habitat was rasterized onto that same nested grid after transforming all shapefiles to Mexico ITRF2008 / LCC.

Average head rope length: 0.030 km
Average vessel trawl speed: 3.7 km/h
```{r}
# Convert fishing hours to area km² of disturbance for each raster cell
# (already on fishing grid; no spatial transform)
disturbance_raster_km2 <- fishing_rasters * 3.7 * 0.030
compareGeom(disturbance_raster_km2, fishing_rasters[[1]], stopOnError = TRUE)
compareGeom(soft_bottom_fraction, fishing_rasters[[1]], stopOnError = TRUE)
```


## Take sum of pressures for each region and take the sum of soft bottom for each region 
## Then divide total pressures by total soft bottom in the region

disturbance_raster_km2 = the area covered by trawlers
```{r}
# now get pressure
prop_disturbed <- disturbance_raster_km2/soft_bottom_fraction
# Cap pressure at 1 - assumes that going over the same spot multiple times introduces minimal extra impact.
prop_disturbed[prop_disturbed > 1] <- 1
prop_disturbed[is.infinite(prop_disturbed)] <- 0

## save for rhodoliths
# Intermediate: soft-bottom fishing pressure (0–1) for rhodolith use
int_dir <- here("_habitats/HAB_soft_bottom/int")
# One multi-layer GeoTIFF (layers = years 2007–2025)
names(prop_disturbed) <- paste0("prop_disturbed_", years)
writeRaster(
  prop_disturbed,
  filename = file.path(int_dir, "prop_disturbed_soft_bottom_1km.tif"),
  overwrite = TRUE
)

area_soft_bottom_disturbed = prop_disturbed * soft_bottom_fraction
```

Make plot of destructive fishing pressure.
```{r}
# Land first (gray), then soft-bottom raster, then marine borders
plot(vect(ohi_land), col = "gray85", border = "gray60", lwd = 0.5,
     main = "Fishing pressure", axes = FALSE, box=FALSE)
plot(
  prop_disturbed,
  add = TRUE,
  col = hcl.colors(100, "viridis"),
  range = c(0, 1),
  legend = FALSE,
  axes = FALSE, 
  box = FALSE
)
plot(vect(ohi_marine), add = TRUE, border = "gray70", lwd = 0.7, col = NA, box=FALSE, axes=FALSE)

```


soft_bottom_fraction = soft bottom area (fraction ≈ km² on 1 km cells)
```{r}
# Rasterize OHI region IDs onto the fishing grid (vectors already in EPSG:6372)
template <- rasterize(
  vect(ohi_marine),
  fishing_rasters[[1]],
  field = "id",
  background = NA
)
compareGeom(template, fishing_rasters[[1]], stopOnError = TRUE)

# Extract region_ids
region_ids <- as.vector(values(template))

# Number of cells in the fishing-aligned region raster
n_cells <- ncell(template)

# Number of years in the disturbance stack
n_years <- nlyr(area_soft_bottom_disturbed)

# Build a tibble where each row = one pixel × one year
df_regions <- tibble(
  region_id        = rep(region_ids, times = n_years), # Region ID for each pixel, repeated for every year
  year             = rep(years, each = n_cells), # Year column: each year repeated for all n_cells pixels
  disturbed_area   = as.vector(values(area_soft_bottom_disturbed)), # Disturbed soft-bottom area (km²) for each pixel × year
  soft_bottom_area = rep(as.vector(values(soft_bottom_fraction)), times = n_years)  # Soft-bottom area (km²) for each pixel — same every year
)

# Remove rows with missing region IDs or invalid numeric values
df_regions <- df_regions |>
  filter(!is.na(region_id), !is.nan(disturbed_area), !is.nan(soft_bottom_area))

df_regions
```

```{r}
# Calculate Pressure and SOF score for each region for each year
pressure_df <- df_regions |>
  group_by(region_id, year) |>
  summarise(
    disturbed_sum   = sum(disturbed_area, na.rm = TRUE),
    soft_bottom_sum = sum(soft_bottom_area, na.rm = TRUE),
    Pressure        = disturbed_sum / soft_bottom_sum,
    SOF_score       = 1 - Pressure,
    Pressure        = Pressure * 100,
    SOF_score       = SOF_score * 100,  
    .groups = "drop"
  )

pressure_df
# SOF score per region per year
write_csv(
  pressure_df,
  here("_habitats/HAB_soft_bottom/SOF_outputs/sof_region_scores_per_year.csv"))

# pressure_df <- read_csv(here::here("_habitats/HAB_soft_bottom/SOF_outputs/sof_region_scores_per_year.csv"))

```

## Current Sof
```{r}
# Select a single year
year_to_show <- 2025   # <-- change this to any year you want

# Filter for selected year
pressure_year <- pressure_df |>
  filter(year == year_to_show) |>
  arrange(region_id)

# Make a kable table
pressure_year |>
  select(region_id, Pressure, SOF_score) |>
  kable(
    digits = 4,
    col.names = c("Region ID", "Pressure", "SOF_score"),
    align = c("c", "r", "r")
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "bordered", "hover"),
    full_width = FALSE
  ) |>
  add_header_above(
    setNames(3, paste0("Soft-bottom Pressure & SOF — Year ", year_to_show))
  )

```

```{r}
ohi_sof <- ohi_regions |>
  left_join(
    pressure_year |> transmute(id = as.integer(region_id), SOF_score),
    by = "id"
  )
ohi_land   <- ohi_sof |> filter(location == "land")
ohi_marine <- ohi_sof |> filter(location == "marine")
ggplot() +
  geom_sf(data = ohi_land, fill = "#f5e6c8", color = "gray40", linewidth = 0.3) +
  geom_sf(data = ohi_marine, aes(fill = SOF_score), color = "gray20", linewidth = 0.4) +
  scale_fill_distiller(
    palette = "RdYlGn",
    direction = 1,
    name = "SOF score",
    limits = c(0, 100),
    breaks = seq(0, 100, by = 20)
  ) +
  labs(title = "Soft-bottom habitat status (2025)") +
  theme_void() +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

```

```{r, fig.width=12, fig.height=7}
# Define region colors
rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

# Clean region names in pressure_df to match color names
pressure_df <- pressure_df |>
  mutate(
    region_clean = paste0("Region_", region_id),   # numeric → Region_#
    region_clean = gsub("_", " ", region_clean)    # Region_1 → Region 1
  )

# Clean color names to match cleaned labels ("Region 1", etc.)
rgn_colors_clean <- setNames(
  rgn_colors,
  gsub("_", " ", names(rgn_colors))                # Region_1 → Region 1
)

# SOF time series plot with custom colors
ggplot(
  pressure_df,
  aes(
    x = year,
    y = SOF_score,
    group = region_clean,
    color = region_clean
  )
) +
  geom_point(size = 2) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = rgn_colors_clean) +
  labs(
    title = "SOF Score Over Time",
    x = "Year",
    y = "SOF Score",
    color = "Region ID"
  ) +
  theme_minimal(base_size = 14)

```



## Format and save data

```{r}
# Calculate condition score
condition_df <- pressure_df |>
  select(region_id, year, condition = SOF_score) |>                  # keep needed columns
  mutate(
    condition = pmin(condition, 100)         # cap at 100
  )
condition_df

write_csv(condition_df, here::here("_habitats/HAB_soft_bottom/SOF_outputs/sof_status.csv"))

condition_df <- condition_df |>
  mutate(region_id = factor(region_id, levels = 1:9))
ggplot(condition_df, aes(x = year, y = condition)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  facet_wrap(~ region_id, ncol = 3, labeller = labeller(region_id = function(x) paste("Region", x))) +
  scale_x_continuous(breaks = pretty(condition_df$year)) +
  labs(
    x = "Year",
    y = "Condition",
    title = "Condition by region and year"
  ) +
  theme_bw()

```

## Trend Calculation

```{r}
# Trend Calculation: lm for most recent 5 years of data
year_end <- max(condition_df$year, na.rm = TRUE)
year_start <- year_end - 4
calc_trends <- function(d) {
  ys <- min(d$year)
  ye <- max(d$year)
  cs <- d$condition[d$year == ys][1]
  fit_lin <- lm(condition ~ year, data = d)
  fit_log <- lm(log(condition) ~ year, data = d)
  slope_lin <- unname(coef(fit_lin)["year"])
  slope_log <- unname(coef(fit_log)["year"])
  tibble(
    year = ye,
    year_start = ys,
    year_end = ye,
    condition_start = cs,
    trend_pct_linear = (slope_lin / cs) * 100,              # slope / start * 100
    trend_pct_cagr   = (exp(slope_log) - 1) * 100           # slope from log model
  )
}
trend_df <- condition_df |>
  filter(year >= year_start, year <= year_end) |>
  group_by(region_id) |>
  group_modify(~ calc_trends(.x)) |>
  ungroup()

trend_df

write_csv(trend_df, here::here("_habitats/HAB_soft_bottom/SOF_outputs/sof_trend.csv"))
```


## Pressure

```{r}
pressure_df_score <- pressure_df %>%
  mutate(pressure = Pressure/100) %>%
  select(region_id, year, pressure)

write_csv(pressure_df_score, here::here("_habitats/HAB_soft_bottom/SOF_outputs/sof_pressure.csv"))
```

