---
title: "Soft bottom - Gulf of California OHI"
author: "Austin Martinez"
date: 07/10/2026
format: 
  html:
    code-fold: show
execute:
  echo: true
  message: false
  warning: false
---


```{r}
# Load in Libraries
library(sf)
library(dplyr)
library(here)
library(terra)
library(tidyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(tmap)
```

```{r}
# Read in OHI region boundaries
ohi_regions <- st_read(here("spatial", "ohi_regions", "goc_ohi_rgns.shp")) |> st_make_valid()

# Filter OHI regions to marine polygons
ohi_regions_marine <- ohi_regions |>
  filter(location == "marine")
```

```{r}
# Read in Arrecife rocoso (rocky reef) data
rocky_reef <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/Hem_data/goc_habitat_data/Arrecife rocoso/Arrecife rocoso.shp")) |> 
  st_make_valid() |> 
  st_transform(st_crs(ohi_regions))

# Read in Seamounts data
seamounts <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/Hem_data/goc_habitat_data/seamount/seamount_merge.shp")) |> 
  st_make_valid() |> 
  st_transform(st_crs(ohi_regions))

# Check if CRSs are the same
print(st_crs(seamounts) == st_crs(ohi_regions))
```

```{r}
# Make data match extent of the ohi_regions layer
seamounts <- st_intersection(seamounts, ohi_regions)
```

## rocky_reef work:
```{r}
# Find proportion of pixel that is rocky reef
rocky_reef <- rocky_reef |>
  mutate(
    rocky_km2 = rocky_ha / 100, # Convert rocky_ha from hectares to km2
    polygon_km2 = as.numeric(st_area(geometry)) / 1e6, # Find area of polygons
    prop_rocky_reef = rocky_km2 / polygon_km2 # Find proportion of polygon that is rocky reef
  )
```


```{r}
# Make a template raster to use in rasterization to make sure cell grids line up

template_100m <- rast(
  ext(ohi_regions),
  resolution = 100,   # 100 m res
  crs = st_crs(ohi_regions)$wkt
)
```

# Rocky reef rasterization:
```{r}
# Rasterize rocky reef polygon proportions
rocky_reef_prop_r <- rasterize(vect(rocky_reef), template_100m, field = "prop_rocky_reef", background = 0)

# Crop raster to OHI regions
rocky_reef_prop_crop <- crop(rocky_reef_prop_r, ohi_regions)

# Mask raster to OHI regions
rocky_reef_prop_mask <- mask(rocky_reef_prop_crop, ohi_regions)
```

```{r}
# Check to see if the min and max are between 0-1
max(rocky_reef_prop_r)
```


```{r}
tmap_mode("view")

tm_shape(rocky_reef_prop_mask) +
  tm_raster(
    palette = "-viridis",
    title = "Proportion Rocky Reef",
    alpha = 0.8,
    style = "fixed",
    breaks = c(0, 0.0092, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1.0),
    legend.format = list(digits = 4)
  ) +
tm_shape(ohi_regions_marine) +
  tm_borders(
    col = "white",
    lwd = 2,
    alpha = 0.9
  ) +
tm_layout(
    legend.outside = TRUE,
    frame = FALSE
)
```


```{r}
# Rasterize the vector seamounts data
seamounts_r  <- rasterize(vect(seamounts),  template_100m, field = 1, background = 0)
```


## Quick validation
```{r}
# Check if values are between 0-1
max(seamounts_r)
```

```{r}
# Check alignment (resolution, origin, extent, CRS, rotation, cell alignment)
compareGeom(rocky_reef_prop_mask, template_100m)
compareGeom(seamounts_r, template_100m)
```

```{r}
# Make raster where 0 = no hard‑bottom and 1+ = hard‑bottom present
hard_bottom_100m <- rocky_reef_prop_mask +
                    seamounts_r
```

```{r}
# Check if values are between 0-1
max(hard_bottom_100m)
```

```{r}
# Cap values at 1 to make a proportion
hard_bottom_100m[hard_bottom_100m > 1] <- 1
```

```{r}
# Check if values are between 0-1
max(hard_bottom_100m)
```


```{r}
# Groups every 10×10 block of 100‑m cells into one 1‑km cell, and computes the mean of values inside that block
hard_bottom_fraction <- aggregate(hard_bottom_100m, fact = 10, fun = "mean")

# flips the fraction so more soft bottom = 1
soft_bottom_fraction <- 1 - hard_bottom_fraction
```

```{r}
# Check if values are between 0-1
max(soft_bottom_fraction)
```


```{r}
tmap_mode("view")

tm_shape(soft_bottom_fraction) +
  tm_raster(
    palette = "-viridis",
    title = "Proportion Soft bottom",
    alpha = 0.8,
    breaks = seq(0, 1, by = 0.1),   # FORCE scale 0 → 1
    legend.format = list(digits = 2)
  ) +
tm_shape(ohi_regions_marine) +
  tm_borders(
    col = "gray",
    lwd = 2,
    alpha = 0.9
  ) +
tm_layout(
    legend.outside = TRUE,
    frame = FALSE
)

```

```{r}
# Mask to the OHI marine region extent
marine_mask <- rasterize(
  vect(ohi_regions_marine),
  soft_bottom_fraction,
  field = 1,
  background = NA
)

soft_bottom_fraction_marine <- mask(soft_bottom_fraction, marine_mask)
```


```{r}
tmap_mode("view")

tm_shape(soft_bottom_fraction_marine) +
  tm_raster(
    palette = "-viridis",
    title = "Proportion Soft Bottom",
    alpha = 0.8,
    breaks = seq(0, 1, by = 0.1),   # FORCE scale 0 → 1
    legend.format = list(digits = 2)
  ) +
tm_shape(ohi_regions_marine) +
  tm_borders(
    col = "white",
    lwd = 2,
    alpha = 0.9
  ) +
tm_layout(
    legend.outside = TRUE,
    frame = FALSE
)
```


```{r}
# View the 
tmap_mode("view")
tmap_options(check.and.fix = TRUE)

tm_shape(ohi_regions_marine) +
  tm_borders(col = "black", lwd = 1) +

tm_shape(seamounts) +
  tm_fill(col = "red", alpha = 0.5) +
  tm_borders(col = "red") +

tm_layout(
  frame = FALSE,
  legend.outside = TRUE,
  legend.outside.position = "right"
)
```

## Redundancy Check
-   Is the soft bottom layer resolution 1km
-   Are all values between 0-1
```{r}
soft_res <- res(soft_bottom_fraction_marine)
print(paste("Resolution (m):", soft_res))
max(soft_bottom_fraction_marine, na.rm = TRUE)
```

## Destructive fishing data

```{r}
# Years of destructive fishing data available
years <- 2007:2025

# Build full file paths for each year's raster and load them into a multi-layer SpatRaster
fishing_rasters <- rast(
  paste0(
    "/home/shares/ohi/OHI_GOC/goal_prep/pressures/vessels/benthic_destructive_fishing/destructive_fishing_hours_",
    years,
    "_1km.tif"
  )
)

# Print the total rasters in the stack
nlyr(fishing_rasters)
```

```{r}
# Reproject soft-bottom habitat raster to match the fishing raster grid
soft_bottom_aligned <- project(soft_bottom_fraction_marine, fishing_rasters[[1]])
# Check if geometries of the aligned raster and fishing rasters match
compareGeom(soft_bottom_aligned, fishing_rasters)
```

Average head rope length: 0.030 km
Average vessel trawl speed: 3.704 km/h
```{r}
# Convert fishing hours to area km² of disturbance for each raster cell
disturbance_raster_km2 <- fishing_rasters * 3.704 * 0.030
```

```{r}
# Check if values are between 0-1
max(disturbance_raster_km2[[7]])
```

```{r}
# Cap disturbance at 1 km² since there can't be more than 1 km² disturbance in a 1km² cell
disturbance_raster_km2[disturbance_raster_km2 > 1] <- 1
```


## Take sum of pressures for each region and take the sum of soft bottom for each region 
## Then divide total pressures by total soft bottom in the region

disturbance_raster_km2 = actual km² of seafloor disturbed in each pixel
disturbance_km_per_km_soft_bottom = ratio of disturbance to soft-bottom habitat

We use disturbance_raster_km2 because this requires area-weighted pressure:
```{r}
# Here we use disturbance_raster_km2 and not disturbance_km_per_km_soft_bottom because ...
# Find the km² of soft‑bottom habitat disturbed
disturbed_soft_bottom_area <- disturbance_raster_km2 * soft_bottom_aligned
summary(values(disturbed_soft_bottom_area[[7]]))
```
soft_bottom_aligned = soft bottom area
```{r}
# Load region raster template (1 km resolution)
template <- rast(here("spatial", "ohi_regions", "ohi_marine_raster_1km.tif"))

# Extract region_ids
region_ids <- as.vector(values(template))

# Correct alignment: project your rasters TO the template grid
disturbed_soft_bottom_area_aligned <- project(disturbed_soft_bottom_area, template)
soft_bottom_area_aligned           <- project(soft_bottom_aligned, template)

# Number of cells in the 1‑km OHI template raster
n_cells <- ncell(template)

# Number of years (layers) in the disturbed_soft_bottom_area raster stack
n_years <- nlyr(disturbed_soft_bottom_area_aligned)

# Build a tibble where each row = one pixel × one year
df_regions <- tibble(
  region_id        = rep(region_ids, times = n_years), # Region ID for each pixel, repeated for every year
  year             = rep(years, each = n_cells), # Year column: each year repeated for all n_cells pixels
  disturbed_area   = as.vector(values(disturbed_soft_bottom_area_aligned)), # Disturbed soft-bottom area (km²) for each pixel × year
  soft_bottom_area = rep(as.vector(values(soft_bottom_area_aligned)), times = n_years)  # Soft-bottom area (km²) for each pixel — same every year
)

# Remove rows with missing region IDs or invalid numeric values
df_regions <- df_regions |>
  filter(!is.na(region_id), !is.nan(disturbed_area), !is.nan(soft_bottom_area))

df_regions
```
```{r}
# Calculate Pressure and SOF score for each region for each year
pressure_df <- df_regions |>
  group_by(region_id, year) |>
  summarise(
    disturbed_sum   = sum(disturbed_area, na.rm = TRUE),
    soft_bottom_sum = sum(soft_bottom_area, na.rm = TRUE),
    Pressure        = disturbed_sum / soft_bottom_sum,
    SOF_score       = 1 - Pressure,
    Pressure        = Pressure * 100,
    SOF_score       = SOF_score * 100,  
    .groups = "drop"
  )

pressure_df
# SOF score per region per year
write_csv(
  pressure_df,
  "SOF_outputs/sof_region_scores_per_year.csv",
  row.names = FALSE
)

# pressure_df <- read_csv(here::here("_habitats/HAB_soft_bottom/SOF_outputs/sof_region_scores_per_year.csv"))

```

## Current Sof
```{r}
# Select a single year
year_to_show <- 2025   # <-- change this to any year you want

# Filter for selected year
pressure_year <- pressure_df |>
  filter(year == year_to_show) |>
  arrange(region_id)

# Make a kable table
pressure_year |>
  select(region_id, Pressure, SOF_score) |>
  kable(
    digits = 4,
    col.names = c("Region ID", "Pressure", "SOF_score"),
    align = c("c", "r", "r")
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "bordered", "hover"),
    full_width = FALSE
  ) |>
  add_header_above(
    setNames(3, paste0("Soft-bottom Pressure & SOF — Year ", year_to_show))
  )

```

```{r}
# Match region_id type to polygon ID
pressure_year <- pressure_year |>
  mutate(region_id = as.integer(region_id))

# Join SOF scores to polygons using region ID column
ohi_regions_map <- ohi_regions_marine |>
  left_join(
    pressure_year |> select(region_id, SOF_score),
    by = c("id" = "region_id")
  )


tmap_mode("view")
tm_shape(ohi_regions_map) +
  tm_polygons(
    col = "SOF_score",
    palette = "-viridis",
    title = paste0("SOF_score (Year ", year_to_show, ")"),
    alpha = 0.9,
    border.col = "white",
    border.lwd = 1
  ) +
tm_layout(
    legend.outside = TRUE,
    frame = FALSE
)

```

```{r, fig.width=12, fig.height=7}
# Define region colors
rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

# Clean region names in pressure_df to match color names
pressure_df <- pressure_df |>
  mutate(
    region_clean = paste0("Region_", region_id),   # numeric → Region_#
    region_clean = gsub("_", " ", region_clean)    # Region_1 → Region 1
  )

# Clean color names to match cleaned labels ("Region 1", etc.)
rgn_colors_clean <- setNames(
  rgn_colors,
  gsub("_", " ", names(rgn_colors))                # Region_1 → Region 1
)

# SOF time series plot with custom colors
ggplot(
  pressure_df,
  aes(
    x = year,
    y = SOF_score,
    group = region_clean,
    color = region_clean
  )
) +
  geom_point(size = 2) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = rgn_colors_clean) +
  labs(
    title = "SOF Score Over Time",
    x = "Year",
    y = "SOF Score",
    color = "Region ID"
  ) +
  theme_minimal(base_size = 14)

```



## Current Status Calculation

```{r}
# Calculate condition score: SOF_2025 / SOF_2007 for each region
condition_df <- pressure_df |>
  select(region_id, year, condition = SOF_score) |>                  # keep needed columns
  mutate(
    condition = pmin(condition, 100)         # cap at 100
  )
condition_df

write_csv(condition_df, here::here("_habitats/HAB_soft_bottom/SOF_outputs/sof_status.csv"))

condition_df <- condition_df |>
  mutate(region_id = factor(region_id, levels = 1:9))
ggplot(condition_df, aes(x = year, y = condition)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  facet_wrap(~ region_id, ncol = 3, labeller = labeller(region_id = function(x) paste("Region", x))) +
  scale_x_continuous(breaks = pretty(condition_df$year)) +
  labs(
    x = "Year",
    y = "Condition",
    title = "Condition by region and year"
  ) +
  theme_bw()

```

## Trend Calculation

```{r}
# Trend Calculation: lm for most recent 5 years of data
year_end <- max(condition_df$year, na.rm = TRUE)
year_start <- year_end - 4
calc_trends <- function(d) {
  ys <- min(d$year)
  ye <- max(d$year)
  cs <- d$condition[d$year == ys][1]
  fit_lin <- lm(condition ~ year, data = d)
  fit_log <- lm(log(condition) ~ year, data = d)
  slope_lin <- unname(coef(fit_lin)["year"])
  slope_log <- unname(coef(fit_log)["year"])
  tibble(
    year = ye,
    year_start = ys,
    year_end = ye,
    condition_start = cs,
    trend_pct_linear = (slope_lin / cs) * 100,              # slope / start * 100
    trend_pct_cagr   = (exp(slope_log) - 1) * 100           # slope from log model
  )
}
trend_df <- condition_df |>
  filter(year >= year_start, year <= year_end) |>
  group_by(region_id) |>
  group_modify(~ calc_trends(.x)) |>
  ungroup()

trend_df

write_csv(trend_df, here::here("_habitats/HAB_soft_bottom/SOF_outputs/sof_trend.csv"))
```


## Pressure

```{r}

```

