################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
# Visualization
library(ggplot2)
library(colorspace)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/BD/")
```

```{r Load clean_names_lowercase function}
clean_names_lowercase <- function(x) {

  x <- iconv(x, from = "", to = "UTF-8", sub = "")
  x <- stringi::stri_trans_general(x, "Latin-ASCII")

  x %>%
    tolower() %>%
    trimws()
}
```


# STEP 0: Load in Region_Key
```{r Region key}
Region_Key0<- read.csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Keys/Spatial_Keys/RegionKey_FINAL_May262026.csv")

colnames(Region_Key0)

Region_Key<- Region_Key0 %>%
  select(OHI_region_est, OHI_region_name) %>%
  distinct()
```

# STEP 1: Load in and summarize GOC_MarineLandings0

```{r Load in GOC_MarineLandings0 as necessary}
GOC_MarineLandings0<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Marine.Landings_Sector.Sorted.Georef_PATCHED_Jun032026.rds"))
# 3246583 obs., 58 var.

# NOTE TO SELF: This source file may change in later Summer / Fall run. - EMF, June 4 2026
# See the patch file here: https://docs.google.com/document/d/1L6w4Gd8SOjp14f_u46vyn64rEpv4qBRaN_V2IyPjP5w/edit?tab=t.0
```


# PATCH 1 - Assign groups that are unassigned to a Major_Functional_Group:
NOTE TO SELF: This patch should not be necessary when we re-run this in later Summer / Fall run. - EMF, June 15 2026
See the patch file here: https://docs.google.com/document/d/1L6w4Gd8SOjp14f_u46vyn64rEpv4qBRaN_V2IyPjP5w/edit?tab=t.0

```{r Load in GenusFamilyGroup_Key as needed}
GenusFamilyGroup_Key<- read.csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Keys/Taxonomic_Keys/GenusFamilyGroupKey_COMPLETE_Apr042026.csv")

# NOTE TO SELF: This patch should not be necessary when we re-run this in later Summer / Fall run. - EMF, June 15 2026
# See the patch file here: https://docs.google.com/document/d/1L6w4Gd8SOjp14f_u46vyn64rEpv4qBRaN_V2IyPjP5w/edit?tab=t.0

MajorFunctionalGroup_Key<- GenusFamilyGroup_Key %>%
  select(estimated_genus_name, Major_Functional_Group)
```

```{r Major_Functional_Group patch for GOC_MarineLandings}
# Fill in missing values based on MajorFunctionalGroup_Key:
GOC_MarineLandings <- GOC_MarineLandings0 %>%
  left_join(
    MajorFunctionalGroup_Key %>%
      select(estimated_genus_name, Major_Functional_Group),
    by = "estimated_genus_name",
    suffix = c("", ".lookup")
  )

GOC_MarineLandings <- GOC_MarineLandings %>%
  mutate(
    Major_Functional_Group = coalesce(
      Major_Functional_Group,
      Major_Functional_Group.lookup
    )
  ) %>%
  select(-Major_Functional_Group.lookup)

# 3246583 obs., 58 var.

# SANITY CHECK:
sum(is.na(GOC_MarineLandings$Major_Functional_Group))
# Only 26 records unassigned. Looks good!
```

# PATCH 2 - Re-assign litopenaeus as paneus:
NOTE TO SELF: This patch should not be necessary when we re-run this in later Summer / Fall run. - EMF, June 15 2026
See the patch file here: https://docs.google.com/document/d/1L6w4Gd8SOjp14f_u46vyn64rEpv4qBRaN_V2IyPjP5w/edit?tab=t.0

```{r Re-assign paneus as litopenaeus}
GOC_MarineLandings <- GOC_MarineLandings %>%
  mutate(
    estimated_genus_name = case_when(
      estimated_genus_name == "Litopenaeus" ~ "Penaeus",
      TRUE ~ estimated_genus_name
    )
  )

unique(GOC_MarineLandings$estimated_genus_name)
```


# **DECISION POINT: Selecting which sectors to include**
```{r Select Sectors_of_Interest}
table(GOC_MarineLandings$sector_estimate_draft)

MarineFishery_Sectors<- c("marine: fishery sector yet to be classified",
                          "marine: small-scale capture",
                          "marine: large-scale capture")
  
PossibleMarineFishery_Sectors<- c("freshwater or marine: large-scale capture",
                          "freshwater or marine: small-scale capture",
                          "freshwater or marine: fishery sector yet to be classified")

GOC_MarineFisheryLandings0<- GOC_MarineLandings %>%
  filter(sector_estimate_draft %in% c(MarineFishery_Sectors, PossibleMarineFishery_Sectors))

# 3137044 obs., 58 var.
```

```{r Re-level sector_estimate_draft and Major_Functional_Group}
# Re-level sectors (mainly for plotting purposes):
GOC_MarineFisheryLandings <- GOC_MarineFisheryLandings0 %>%
  filter(
    sector_estimate_draft %in% c(
      MarineFishery_Sectors,
      PossibleMarineFishery_Sectors
    )
  ) %>%
  mutate(
    sector_estimate_draft = factor(
      sector_estimate_draft,
      levels = c(
        "marine: large-scale capture",
        "marine: small-scale capture",
        "marine: fishery sector yet to be classified",
        "freshwater or marine: large-scale capture",
        "freshwater or marine: small-scale capture",
        "freshwater or marine: fishery sector yet to be classified"
      )
    )
  )

# SANITY CHECK: How many Major_Functional_Group unassigned?
sum(is.na(GOC_MarineFisheryLandings$Major_Functional_Group))
# Only 18 records unassigned. Looks good!

table(GOC_MarineFisheryLandings$Major_Functional_Group)

GOC_MarineFisheryLandings <- GOC_MarineFisheryLandings %>%
  mutate(
    Major_Functional_Group = factor(
      Major_Functional_Group,
      levels= c(
        
        # Crustaceans
        "Shrimps, prawns",
        "Crustaceans",
        
        # Molluscs
        "Cephalopods",
        "Molluscs excl. Cephalopods",
        
        # Other invertebrates
        "Echinoderms",
        "Jellies",
        "Other marine invertebrates",
        
        # Fishes
        "Anchovy & Sardine (small pelagics)",
        "Pelagic marine fish (excl. anchovy & sardine)",
        "Demersal marine fishes",
        "Marine fishes NEI",
        "Marine or freshwater fishes NEI",
        "Diadromous fishes",
        
        # Elasmobranchs
        "Sharks",
        
        # Miscellaneous / unresolved
        "Marine algae",
        "Marine species NEI",
        "unknown",
        NA_character_
      )
    )
  )

# SANITY CHECK:
sum(is.na(GOC_MarineFisheryLandings$Major_Functional_Group))
# Only 18 records unassigned. Looks good!
```

# Strategy 1 - Track changes in landings of major functional groups relative to mean landings  ------------------------------

## Open Ocean Taxa (OCE)

```{r Isolate Open-Ocean taxa and create OCE_Landings0}
unique(GOC_MarineFisheryLandings$Major_Functional_Group)

Open_Ocean_Taxa<- c("Pelagic marine fish (excl. anchovy & sardine)",
"Anchovy & Sardine (small pelagics)",
"Sharks",
"Cephalopods"
)

colnames(GOC_MarineFisheryLandings)

OpenOcean_Landings0<- GOC_MarineFisheryLandings %>%
  filter(Major_Functional_Group %in% Open_Ocean_Taxa) %>%
    select(state, OHI_region_est,
         office_name_clean, landing_site_clean, 
         year,
         est_aquaculture_production, reported_fleet, 
         sector_estimate_reported, sector_estimate_draft,
         recorded_taxa_name, recorded_taxa_name_clean, common_name_clean, 
         recorded_taxa_group, estimated_taxa_group, Major_Functional_Group,
         estimated_species_name, estimated_genus_name, estimated_family_name,
         landed_weight, live_weight, weight_kg_corrected, value
         )

# 653887 obs., 22 vars.
```

```{r Summarize OCE landings}
OpenOcean_Landings_summary<- OpenOcean_Landings0 %>%
  group_by(year, OHI_region_est, Major_Functional_Group) %>%
  summarize(
    total_landings_mt = sum(weight_kg_corrected, na.rm = TRUE)/1000,
    .groups = "drop"
  ) %>%
  mutate(
    Major_Functional_Group= clean_names_lowercase(Major_Functional_Group),
    OHI_region_est= clean_names_lowercase(OHI_region_est)
  )
# 853 obs.
```

# Join in region names:

```{r Merge Region_Key to OCE_Scores}
OpenOcean_Landings_summary<- OpenOcean_Landings_summary %>%
  left_join(Region_Key, 
            by = "OHI_region_est")

# Make sure to order the regions:
Region_order <- c(
  "Western Upper Gulf",
  "Western Midriff Islands",
  "Western Central Gulf",
  "Western Southern Gulf",
  "Eastern Upper Gulf",
  "Eastern Midriff Islands",
  "Southern Sonora Coast",
  "Sinaloa Coast",
  "Southeastern Gulf"
)

OpenOcean_Landings_summary <- OpenOcean_Landings_summary %>%
  mutate(
    OHI_region_name = factor(
      OHI_region_name,
      levels = Region_order
    )
  )
```

```{r Plot OCE landings over time}
colnames(OpenOcean_Landings_summary)
unique(OpenOcean_Landings_summary$OHI_region_name)

# Plot 1, panel per region:

OCE_Landings_perRegion_plot<- OpenOcean_Landings_summary %>%
  filter(year >= 2014, year <= 2024) %>%
  ggplot(
    aes(
      x = year,
      y = log10(total_landings_mt),
      color = Major_Functional_Group,
      group = Major_Functional_Group
    )
  ) +
  geom_line() +
  
  theme_bw() +
  
  facet_wrap(~ OHI_region_name)

OCE_Landings_perRegion_plot


# Plot 2, panel per group:

Region_colors<- c(
  "Western Upper Gulf" = "#C03729",
  "Western Midriff Islands" = "#E68C7C",
  "Western Central Gulf" = "#FC8F24",
  "Western Southern Gulf" = "#dcaa38",
  "Eastern Upper Gulf" = "#cbd2a0",
  "Eastern Midriff Islands" = "#98a25a",
  "Southern Sonora Coast" = "#747428",
  "Sinaloa Coast" = "#627391",
  "Southeastern Gulf" = "#8d61a9")

OCE_Landings_perGroup_plot<- OpenOcean_Landings_summary %>%
  filter(year >= 2014, year <= 2024) %>%
  ggplot(
    aes(
      x = year,
      y = log10(total_landings_mt),
      color = OHI_region_name,
      group = OHI_region_name
    )
  ) +
  geom_line() +
  
  scale_color_manual(values = Region_colors) +
  
  theme_bw() +
  
  facet_wrap(~ Major_Functional_Group)

OCE_Landings_perGroup_plot
```


# STEP 3: Strategy 2 - Use FIS Stock Status scores for open-ocean groups ------------------------------

```{r Load in FIS_StockStatus scores as necessary}
StockStatus_Scores0<- read.csv("/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_Stock_Status_SubScores/Stock_Statuses_Compiled_df_Jun072026.csv")

# Species x year x region; Stock_Status is used as-is.

# CSV was previously titled: StockStatus_Estimates_Jun072026.csv
```

```{r Prep StockStatus_Scores}
StockStatus_Scores <- StockStatus_Scores0 %>%
  filter(year >= 2018, year <= 2024) %>%
  mutate(
    OHI_region_est = clean_names_lowercase(OHI_region_est),
    estimated_genus_name = clean_names_lowercase(estimated_genus_name)
  )
# 4130 obs.

unique(StockStatus_Scores$Major_Functional_Group)
unique(StockStatus_Scores$estimated_genus_name)
```

# PATCH 3 - Patch sardinops, penaeus, and litopenaeus 
```{r Patch penaeus major functional group}
StockStatus_Scores <- StockStatus_Scores %>%
  mutate(
    estimated_genus_name = case_when(
      estimated_genus_name == "litopenaeus" ~ "penaeus",
      TRUE ~ estimated_genus_name
    ),
    Major_Functional_Group = case_when(
      estimated_genus_name %in% c("penaeus", "litopenaeus") ~ "Shrimps, prawns",
      estimated_genus_name %in% c("sardinops") ~ "Anchovy & Sardine (small pelagics)",
      TRUE ~ Major_Functional_Group
    )
  )
```

# PATCH 4 - Patch loligo, dosidicus, octopus to "Cephalopods"
NOTE TO SELF: Hopefully won't need this patch eventually - Jul 22 2026
```{r Patch penaeus major functional group}
StockStatus_Scores <- StockStatus_Scores %>%
  mutate(
    Major_Functional_Group = case_when(
      estimated_genus_name == "loligo" ~ "Cephalopods",
      estimated_genus_name == "dosidicus" ~ "Cephalopods",
      estimated_genus_name == "octopus" ~ "Cephalopods",
      TRUE ~ Major_Functional_Group
    )
  )

unique(StockStatus_Scores$Major_Functional_Group)
```

## Stock_Status: Open Ocean Communities

```{r Create OCE_StockStatus_Scores}

Open_Ocean_Taxa<- c("Pelagic marine fish (excl. anchovy & sardine)",
"Anchovy & Sardine (small pelagics)",
"Sharks",
"Cephalopods"
)

colnames(StockStatus_Scores)

OCE_StockStatus_Scores<- StockStatus_Scores %>%
  filter(Major_Functional_Group %in% Open_Ocean_Taxa) %>%
    select(year, Major_Functional_Group,
           estimated_genus_name, estimated_family_name,
           OHI_region_est,
           total_weight_mt,
           StockStatus_final,
           StockStatus_source
         )

# 853 obs.

OCE_StockStatus_Scores<- OCE_StockStatus_Scores %>%
  left_join(
    Region_Key,
    by= "OHI_region_est"
  )
```

```{r Plot OCE Stock Status over time}
colnames(OCE_StockStatus_Scores)

# Plot 1, panel per region:

OCE_StockStatus_perRegion_plot<- OCE_StockStatus_Scores %>%
  filter(year >= 2014, year <= 2024) %>%
  filter(!is.na(OHI_region_name)) %>%
  
  ggplot(
    aes(
      x = year,
      y = StockStatus_final,
      color = Major_Functional_Group,
      group = estimated_genus_name
    )
  ) +
  geom_line() +
  
  facet_wrap(~ OHI_region_name) +

  labs(
    title = "Status of Open-Ocean Stocks per Region",
    subtitle= "updated: Jul 22 2026",
    y = "Stock Status (0 to 1)",
    x = "Year"
  ) +
  
  theme_bw(base_size = 14) +
  theme(
    strip.text = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 16, face = "bold")
  )

OCE_StockStatus_perRegion_plot

# Plot 2, panel per group:

Region_colors<- c(
  "Western Upper Gulf" = "#C03729",
  "Western Midriff Islands" = "#E68C7C",
  "Western Central Gulf" = "#FC8F24",
  "Western Southern Gulf" = "#dcaa38",
  "Eastern Upper Gulf" = "#cbd2a0",
  "Eastern Midriff Islands" = "#98a25a",
  "Southern Sonora Coast" = "#747428",
  "Sinaloa Coast" = "#627391",
  "Southeastern Gulf" = "#8d61a9")


OCE_StockStatus_perGroup_plot<- OCE_StockStatus_Scores %>%
  filter(year >= 2014, year <= 2024) %>%
  filter(!is.na(OHI_region_name)) %>%
  
  ggplot(
    aes(
      x = year,
      y = StockStatus_final,
      color = OHI_region_name,
      group = interaction(OHI_region_name, estimated_genus_name)
    )
  ) +
  
  facet_wrap(~ Major_Functional_Group) +
  
  geom_line() +
  
  scale_color_manual(values = Region_colors) +
  
  labs(
    title = "Status of Open-Ocean Stocks per Group",
    subtitle= "updated: Jul 22 2026",
    y = "Stock Status (0 to 1)",
    x = "Year"
  ) +
  
  theme_bw(base_size = 14) +
  theme(
    strip.text = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 16, face = "bold")
  )

OCE_StockStatus_perGroup_plot
```

# Summarize by Major_Functional_Group before calculating scores?

```{r Summarize stock status - PURG?}
colnames(OCE_StockStatus_Scores)

OCE_StockStatus_summarized<- OCE_StockStatus_Scores %>%
  group_by(year, OHI_region_est, Major_Functional_Group) %>%
  summarize(
    mean_StockStatus = mean(StockStatus_final, na.rm = TRUE),
    total_weight_mt = sum(total_weight_mt, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Major_Functional_Group= clean_names_lowercase(Major_Functional_Group),
    OHI_region_est= clean_names_lowercase(OHI_region_est)
  )
# 201 obs.

OCE_StockStatus_summarized<- OCE_StockStatus_summarized %>%
  left_join(
    Region_Key,
    by= "OHI_region_est"
  )
```

```{r Plot OCE Stock Status over time}
colnames(OCE_StockStatus_summarized)

# Plot 1, panel per region:

OCE_SummaryStockStatus_perRegion_plot<- OCE_StockStatus_summarized %>%
  filter(year >= 2014, year <= 2024) %>%
  filter(!is.na(OHI_region_name)) %>%
  
  ggplot(
    aes(
      x = year,
      y = mean_StockStatus,
      color = Major_Functional_Group,
      group = Major_Functional_Group
    )
  ) +
  geom_line() +
  
  facet_wrap(~ OHI_region_name) +

  labs(
    title = "Status of Open-Ocean Groups per Region",
    subtitle= "updated: Jul 22 2026",
    y = "Stock Status (0 to 1)",
    x = "Year"
  ) +
  
  theme_bw(base_size = 14) +
  theme(
    strip.text = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 16, face = "bold")
  )

OCE_SummaryStockStatus_perRegion_plot

# Plot 2, panel per group:

Region_colors<- c(
  "Western Upper Gulf" = "#C03729",
  "Western Midriff Islands" = "#E68C7C",
  "Western Central Gulf" = "#FC8F24",
  "Western Southern Gulf" = "#dcaa38",
  "Eastern Upper Gulf" = "#cbd2a0",
  "Eastern Midriff Islands" = "#98a25a",
  "Southern Sonora Coast" = "#747428",
  "Sinaloa Coast" = "#627391",
  "Southeastern Gulf" = "#8d61a9")

OCE_SummaryStockStatus_perRegion_plot

OCE_SummaryStockStatus_perGroup_plot<- OCE_StockStatus_summarized %>%
  filter(year >= 2014, year <= 2024) %>%
  filter(!is.na(OHI_region_name)) %>%
  
  ggplot(
    aes(
      x = year,
      y = mean_StockStatus,
      color = OHI_region_name,
      group = OHI_region_name
    )
  ) +
  
  facet_wrap(~ Major_Functional_Group) +
  
  geom_line() +
  
  scale_color_manual(values = Region_colors) +
  
  labs(
    title = "Status of Open-Ocean Groups",
    subtitle= "updated: Jul 22 2026",
    y = "Stock Status (0 to 1)",
    x = "Year"
  ) +
  
  theme_bw(base_size = 14) +
  theme(
    strip.text = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 16, face = "bold")
  )

OCE_SummaryStockStatus_perGroup_plot
```

```{r Save plots as PNGs}
ggsave(
  plot = OCE_SummaryStockStatus_perRegion_plot,
  filename = sprintf("/home/ferrer/ohi_goc_prep/BD/BD_Scores/OCE_Scores/OCE_SummaryStockStatus_perRegion_plot_Jul222026.png"),
  width = 12,
  height = 12,
  dpi = 300
)

ggsave(
  plot = OCE_SummaryStockStatus_perGroup_plot,
  filename = sprintf("/home/ferrer/ohi_goc_prep/BD/BD_Scores/OCE_Scores/OCE_SummaryStockStatus_perGroup_plot_Jul222026.png"),
  width = 12,
  height = 12,
  dpi = 300
)
```

# Calculate final OCE_Scores:

```{r Calculate OCE_Scores}
colnames(OCE_StockStatus_summarized)

OCE_Scores0 <- OCE_StockStatus_summarized %>%
  select(year, OHI_region_est, OHI_region_name, mean_StockStatus) %>%
  group_by(year, OHI_region_est, OHI_region_name) %>%
  summarize(
    OCE_Score = mean(mean_StockStatus, na.rm = TRUE) * 100,
    n_groups = n(),
    .groups = "drop"
  )
# 69 obs., 5 vars.

OCE_Scores<- OCE_Scores0 %>%
  filter(!is.na(OHI_region_est))
# 63 obs., 5 vars.
```

```{r Sanity check final OCE_Scores}
colnames(OCE_Scores)

summary(OCE_Scores$OCE_Score)
  #  Min. 1st Qu.  Median    Mean 3rd Qu.    Max.
  # 49.22   71.61   77.69   76.10   82.26   94.57

# Year 2024 mean for OCE scores:
mean(OCE_Scores$OCE_Score[OCE_Scores$year == 2024], na.rm = TRUE)
# 66
```

# SAVE OCE_Scores ------------------------------

```{r Save OCE_Scores as csv}
write.csv(OCE_Scores, 
          "/home/ferrer/ohi_goc_prep/BD/BD_Scores/OCE_Scores/OCE_Scores_df_Jul222026.csv", 
          row.names = FALSE)
# 63 obs., 5 vars.
```


# PLOT OCE_Scores  ------------------------------

```{r Merge Region_Key to FP_Scores}
# Make sure to order the regions:
region_order <- c(
  "Western Upper Gulf",
  "Western Midriff Islands",
  "Western Central Gulf",
  "Western Southern Gulf",
  "Eastern Upper Gulf",
  "Eastern Midriff Islands",
  "Southern Sonora Coast",
  "Sinaloa Coast",
  "Southeastern Gulf"
)

OCE_Scores <- OCE_Scores %>%
  mutate(
    OHI_region_name = factor(
      OHI_region_name,
      levels = region_order
    )
  )
```

```{r Plot OCE_Scores}
Region_colors<- c(
  "Western Upper Gulf" = "#C03729",
  "Western Midriff Islands" = "#E68C7C",
  "Western Central Gulf" = "#FC8F24",
  "Western Southern Gulf" = "#dcaa38",
  "Eastern Upper Gulf" = "#cbd2a0",
  "Eastern Midriff Islands" = "#98a25a",
  "Southern Sonora Coast" = "#747428",
  "Sinaloa Coast" = "#627391",
  "Southeastern Gulf" = "#8d61a9")

OCE_Scores_overTime<- OCE_Scores %>%
  
  filter(!is.na(OHI_region_name)) %>%
  
  ggplot(aes(x = year, y = OCE_Score, color = OHI_region_name)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  
  facet_wrap(~ OHI_region_name, scales = "fixed") +
  
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey50") +
  
  labs(
    title = "OCE Scores by Region",
    subtitle= "updated: Jul 22 2026",
    y = "FIS Score (from 0 to 100)",
    x = "Year"
  ) +
  
  ylim(0, 101) +
  
  scale_color_manual(values = Region_colors) +
  
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 16, face = "bold")
  )

OCE_Scores_overTime
```

```{r Save as PNG}
  ggsave(
    plot = OCE_Scores_overTime,
    filename = sprintf("/home/ferrer/ohi_goc_prep/BD/BD_Scores/OCE_Scores/OCE_Scores_plot_Jul222026.png"),
    width = 12,
    height = 12,
    dpi = 300
  )
```