---
title: "Coastal Dune Habitat Condition and Trend - Gulf of California OHI with a 1km and 5km buffer"
author: "Austin Martinez"
date: 06/05/2026
format: 
  html:
    code-fold: show
execute:
  echo: true
  message: false
  warning: false
---

```{r}
# Load in Libraries
library(sf)
library(dplyr)
library(here)
library(terra)
library(tidyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(tmap)
library(stringr)
```

```{r}
# Read in OHI region boundaries
ohi_regions <- st_read(here("spatial", "ohi_regions", "goc_ohi_rgns.shp")) |> st_make_valid()
```

# CANABIO Land Use and Vegitation all years

```{r}
dune_canabio_landuse_1980_91 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250kcs1agw/usv250kcs1agw.shp") |> st_make_valid()

dune_canabio_landuse_2001 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250ks2gw/usv250ks2gw.shp") |> st_make_valid()

dune_canabio_landuse_2005 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250ks3gw/usv250ks3gw.shp") |> st_make_valid()

dune_canabio_landuse_2009 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250ks4gw/usv250ks4gw.shp") |> st_make_valid()

dune_canabio_landuse_2013 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250s5ugw/usv250s5ugw.shp") |> st_make_valid()

dune_canabio_landuse_2016 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250s6gw/usv250s6gw.shp") |> st_make_valid()

dune_canabio_landuse_2021 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250s7gw/usv250s7gw.shp") |> st_make_valid()
```


```{r}
# Create 5km buffer (run once)
sf::sf_use_s2(FALSE)

ohi_regions_marine <- ohi_regions |>
  filter(location == "marine") |>
  st_union()

# Make a 5km buffer ring
buffer_outer <- st_buffer(ohi_regions_marine, dist = 5000)
buffer_ring  <- st_difference(buffer_outer, ohi_regions_marine) |> st_make_valid()

# Union it to the marine OHI layer
ohi_marine_plus_5km <- st_union(ohi_regions_marine, buffer_ring)

# Make a 1km buffer ring
buffer_outer <- st_buffer(ohi_regions_marine, dist = 1000)
buffer_ring  <- st_difference(buffer_outer, ohi_regions_marine) |> st_make_valid()

# Union it to the marine OHI layer
ohi_marine_plus_1km <- st_union(ohi_regions_marine, buffer_ring)
```

```{r}
tmap_mode("view")

# Plot buffers
tm_shape(ohi_marine_plus_5km) +
  tm_polygons(col = "lightgreen", border.col = "forestgreen", alpha = 0.2, lwd = 0.5) +
tm_shape(ohi_marine_plus_1km) +
  tm_polygons(col = "lightblue", border.col = "blue", alpha = 0.2, lwd = 0.5)
```


We are using these layers to represent Coastal dunes:
-     VEGETACIÓN DE DUNAS COSTERAS with a 5km buffer
-     VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA XERÓFILA with a 1km buffer
-     VEGETACIÓN HALÓFILA XERÓFILA with a 1km buffer

```{r}
sf::sf_use_s2(FALSE)

# 1980_91
dune_canabio_landuse_1980_91_sublayer_5km <- dune_canabio_landuse_1980_91 |>
  dplyr::filter(
    descripcio %in% c(
      "VEGETACIÓN DE DUNAS COSTERAS"
    )
  )

dune_canabio_landuse_1980_91_sublayer_1km <- dune_canabio_landuse_1980_91 |>
  dplyr::filter(
    descripcio %in% c(
      "VEGETACIÓN HALÓFILA XERÓFILA",
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA XERÓFILA"
    )
  )

# Match CRS of OHI and Buffer OHI regions
ohi_region <- st_transform(ohi_regions, st_crs(dune_canabio_landuse_1980_91))


ohi_marine_plus_1km <- st_transform(ohi_marine_plus_1km, st_crs(dune_canabio_landuse_1980_91))
ohi_marine_plus_5km <- st_transform(ohi_marine_plus_5km, st_crs(dune_canabio_landuse_1980_91))


#dune_canabio_landuse_1980_91_sublayer_1km <- st_intersection(dune_canabio_landuse_1980_91_sublayer_1km, ohi_marine_plus_1km)
dune_canabio_landuse_1980_91_sublayer_5km <- st_intersection(dune_canabio_landuse_1980_91_sublayer_5km, ohi_marine_plus_5km)

#dune_canabio_landuse_1980_91_sublayer <- st_union(dune_canabio_landuse_1980_91_sublayer_1km, dune_canabio_landuse_1980_91_sublayer_5km)

dune_canabio_landuse_1980_91_sublayer_area <- st_intersection(dune_canabio_landuse_1980_91_sublayer_5km, ohi_region) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 1991
  ) |>
  st_drop_geometry()
```
1991 has no observations for layers:
      "VEGETACIÓN HALÓFILA XERÓFILA",
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA XERÓFILA"
      
So, only the 5km buffer was used in the analysis.

```{r}
tmap_mode("view")

# Plot buffers
tm_shape(dune_canabio_2001_sublayer_5km) +
  tm_polygons(col = "red", border.col = "red", alpha = 0.2, lwd = 0.5) +
tm_shape(dune_canabio_2001_sublayer_1km) +
  tm_polygons(col = "blue", border.col = "blue", alpha = 0.2, lwd = 0.5) +
tm_shape(ohi_marine_plus_1km) +
  tm_polygons(col = "green", border.col = "darkgreen", alpha = 0.1, lwd = 0.5)
```




```{r}
# 2001
dune_canabio_2001_sublayer_5km <- dune_canabio_landuse_2001 |>
  dplyr::filter(
    descripcio %in% c(
      "VEGETACIÓN DE DUNAS COSTERAS"
    )
  )

dune_canabio_2001_sublayer_1km <- dune_canabio_landuse_2001 |>
  dplyr::filter(
    descripcio %in% c(
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA XERÓFILA",
      "VEGETACIÓN HALÓFILA XERÓFILA"
    )
  )

# Match CRS of OHI and Buffer OHI regions
ohi_region <- st_transform(ohi_regions, st_crs(dune_canabio_landuse_2001))
ohi_marine_plus_5km <- st_transform(ohi_marine_plus_5km, st_crs(dune_canabio_landuse_2001))
ohi_marine_plus_1km <- st_transform(ohi_marine_plus_1km, st_crs(dune_canabio_landuse_2001))

dune_canabio_2001_sublayer_5km <- st_intersection(dune_canabio_2001_sublayer_5km, ohi_marine_plus_5km)
dune_canabio_2001_sublayer_1km <- st_intersection(dune_canabio_2001_sublayer_1km, ohi_marine_plus_1km)

dune_canabio_2001_sublayer <- st_union(dune_canabio_2001_sublayer_5km, dune_canabio_2001_sublayer_1km)

dune_canabio_2001_sublayer_area <- st_intersection(dune_canabio_2001_sublayer, ohi_region) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2001
  ) |>
  st_drop_geometry()
```



```{r}
# 2005
dune_canabio_2005_sublayer_5km <- dune_canabio_landuse_2005 |>
  dplyr::filter(
    descripcio %in% c(
      "VEGETACIÓN DE DUNAS COSTERAS"
    )
  )

dune_canabio_2005_sublayer_1km <- dune_canabio_landuse_2005 |>
  dplyr::filter(
    descripcio %in% c(
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA XERÓFILA",
      "VEGETACIÓN HALÓFILA XERÓFILA"
    )
  )

# Match CRS of OHI and Buffer OHI regions
ohi_region <- st_transform(ohi_regions, st_crs(dune_canabio_landuse_2005))
ohi_marine_plus_5km <- st_transform(ohi_marine_plus_5km, st_crs(dune_canabio_landuse_2005))
ohi_marine_plus_1km <- st_transform(ohi_marine_plus_1km, st_crs(dune_canabio_landuse_2005))

dune_canabio_2005_sublayer_5km <- st_intersection(dune_canabio_2005_sublayer_5km, ohi_marine_plus_5km)
dune_canabio_2005_sublayer_1km <- st_intersection(dune_canabio_2005_sublayer_1km, ohi_marine_plus_1km)

dune_canabio_2005_sublayer <- st_union(dune_canabio_2005_sublayer_5km, dune_canabio_2005_sublayer_1km)

dune_canabio_2005_sublayer_area <- st_intersection(dune_canabio_2005_sublayer, ohi_region) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2005
  ) |>
  st_drop_geometry()
```









```{r}
# 2009
dune_canabio_2009_sublayer_5km <- dune_canabio_landuse_2009 |>
  dplyr::filter(
    descripcio %in% c(
      "VEGETACIÓN DE DUNAS COSTERAS"
    )
  )

dune_canabio_2009_sublayer_1km <- dune_canabio_landuse_2009 |>
  dplyr::filter(
    descripcio %in% c(
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA XERÓFILA",
      "VEGETACIÓN HALÓFILA XERÓFILA"
    )
  )

# Match CRS of OHI and Buffer OHI regions
ohi_region <- st_transform(ohi_regions, st_crs(dune_canabio_landuse_2009))
ohi_marine_plus_5km <- st_transform(ohi_marine_plus_5km, st_crs(dune_canabio_landuse_2009))
ohi_marine_plus_1km <- st_transform(ohi_marine_plus_1km, st_crs(dune_canabio_landuse_2009))

dune_canabio_2009_sublayer_5km <- st_intersection(dune_canabio_2009_sublayer_5km, ohi_marine_plus_5km)
dune_canabio_2009_sublayer_1km <- st_intersection(dune_canabio_2009_sublayer_1km, ohi_marine_plus_1km)

dune_canabio_2009_sublayer <- st_union(dune_canabio_2009_sublayer_5km, dune_canabio_2009_sublayer_1km)

dune_canabio_2009_sublayer_area <- st_intersection(dune_canabio_2009_sublayer, ohi_region) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2009
  ) |>
  st_drop_geometry()
```




```{r}
# 2013
dune_canabio_2013_sublayer_5km <- dune_canabio_landuse_2013 |>
  dplyr::filter(
    descripcio %in% c(
      "VEGETACIÓN DE DUNAS COSTERAS"
    )
  )

dune_canabio_2013_sublayer_1km <- dune_canabio_landuse_2013 |>
  dplyr::filter(
    descripcio %in% c(
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA XERÓFILA",
      "VEGETACIÓN HALÓFILA XERÓFILA"
    )
  )

# Match CRS of OHI and Buffer OHI regions
ohi_region <- st_transform(ohi_regions, st_crs(dune_canabio_landuse_2013))
ohi_marine_plus_5km <- st_transform(ohi_marine_plus_5km, st_crs(dune_canabio_landuse_2013))
ohi_marine_plus_1km <- st_transform(ohi_marine_plus_1km, st_crs(dune_canabio_landuse_2013))

dune_canabio_2013_sublayer_5km <- st_intersection(dune_canabio_2013_sublayer_5km, ohi_marine_plus_5km)
dune_canabio_2013_sublayer_1km <- st_intersection(dune_canabio_2013_sublayer_1km, ohi_marine_plus_1km)

dune_canabio_2013_sublayer <- st_union(dune_canabio_2013_sublayer_5km, dune_canabio_2013_sublayer_1km)

dune_canabio_2013_sublayer_area <- st_intersection(dune_canabio_2013_sublayer, ohi_region) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2013
  ) |>
  st_drop_geometry()
```




```{r}
# 2016
dune_canabio_2016_sublayer_5km <- dune_canabio_landuse_2016 |>
  dplyr::filter(
    DESCRIPCIO %in% c(
      "VEGETACIÓN DE DUNAS COSTERAS"
    )
  )

dune_canabio_2016_sublayer_1km <- dune_canabio_landuse_2016 |>
  dplyr::filter(
    DESCRIPCIO %in% c(
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA XERÓFILA",
      "VEGETACIÓN HALÓFILA XERÓFILA"
    )
  )

# Match CRS of OHI and Buffer OHI regions
ohi_region <- st_transform(ohi_regions, st_crs(dune_canabio_landuse_2016))
ohi_marine_plus_5km <- st_transform(ohi_marine_plus_5km, st_crs(dune_canabio_landuse_2016))
ohi_marine_plus_1km <- st_transform(ohi_marine_plus_1km, st_crs(dune_canabio_landuse_2016))

dune_canabio_2016_sublayer_5km <- st_intersection(dune_canabio_2016_sublayer_5km, ohi_marine_plus_5km)
dune_canabio_2016_sublayer_1km <- st_intersection(dune_canabio_2016_sublayer_1km, ohi_marine_plus_1km)

dune_canabio_2016_sublayer <- st_union(dune_canabio_2016_sublayer_5km, dune_canabio_2016_sublayer_1km)

dune_canabio_2016_sublayer_area <- st_intersection(dune_canabio_2016_sublayer, ohi_region) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2016
  ) |>
  st_drop_geometry()
```




```{r}
# 2021
dune_canabio_2021_sublayer_5km <- dune_canabio_landuse_2021 |>
  dplyr::filter(
    DESCRIPCIO %in% c(
      "VEGETACIÓN DE DUNAS COSTERAS"
    )
  )

dune_canabio_2021_sublayer_1km <- dune_canabio_landuse_2021 |>
  dplyr::filter(
    DESCRIPCIO %in% c(
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA XERÓFILA",
      "VEGETACIÓN HALÓFILA XERÓFILA"
    )
  )

# Match CRS of OHI and Buffer OHI regions
ohi_region <- st_transform(ohi_regions, st_crs(dune_canabio_landuse_2021))
ohi_marine_plus_5km <- st_transform(ohi_marine_plus_5km, st_crs(dune_canabio_landuse_2021))
ohi_marine_plus_1km <- st_transform(ohi_marine_plus_1km, st_crs(dune_canabio_landuse_2021))

dune_canabio_2021_sublayer_5km <- st_intersection(dune_canabio_2021_sublayer_5km, ohi_marine_plus_5km)
dune_canabio_2021_sublayer_1km <- st_intersection(dune_canabio_2021_sublayer_1km, ohi_marine_plus_1km)

dune_canabio_2021_sublayer <- st_union(dune_canabio_2021_sublayer_5km, dune_canabio_2021_sublayer_1km)

dune_canabio_2021_sublayer_area <- st_intersection(dune_canabio_2021_sublayer, ohi_region) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2021
  ) |>
  st_drop_geometry()
```

```{r}
# Combine all years into one long table
dune_all_years <- dplyr::bind_rows(
  dune_canabio_landuse_1980_91_sublayer_area,
  dune_canabio_2001_sublayer_area,
  dune_canabio_2005_sublayer_area,
  dune_canabio_2009_sublayer_area,
  dune_canabio_2013_sublayer_area,
  dune_canabio_2016_sublayer_area,
  dune_canabio_2021_sublayer_area
)
```


```{r}
# Pivot wider: regions on rows, years on columns
saltmarsh_wide <- dune_all_years |>
  tidyr::pivot_wider(
    names_from = year,
    values_from = area_km2) |>
  dplyr::arrange(rgn_id)

# Make the kable
saltmarsh_wide |>
  kable( digits = 4,
    col.names = c(
      "Region ID", "1990","2001", "2005", "2009", "2013", "2016", "2021")) |>
  kable_styling(
    bootstrap_options = c("striped", "bordered", "hover")) |>
  add_header_above(
    c("Coastal Dune Extent (km2) CONABIO Land Use (All Years)" = 8))

```

Exclude 1990 because it's numbers don't make sense. It also is missing layers:
-      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA XERÓFILA"
-      "VEGETACIÓN HALÓFILA XERÓFILA"

```{r}
# Combine all years into one long table
dune_all_years <- dplyr::bind_rows(
  dune_canabio_2001_sublayer_area,
  dune_canabio_2005_sublayer_area,
  dune_canabio_2009_sublayer_area,
  dune_canabio_2013_sublayer_area,
  dune_canabio_2016_sublayer_area,
  dune_canabio_2021_sublayer_area
)
```

```{r}
library(ggplot2)

ggplot(dune_all_years, aes(x = year, y = area_km2, group = rgn_id, color = factor(rgn_id))) +
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  labs(
    title = "Coastal Dune Extent Over Time (CONABIO Land Use)",
    x = "Year",
    y = "Coastal Dune Extent (km²)",
    color = "Region ID"
  ) +
  theme_minimal(base_size = 14)

```

```{r}
dune_total <- dune_all_years |>
  group_by(year) |>
  summarise(total_km2 = sum(area_km2, na.rm = TRUE))

ggplot(dune_total, aes(x = year, y = total_km2)) +
  geom_point(size = 3) +
  geom_line(linewidth = 1.2) +
  labs(
    title = "Total Coastal Dune Extent Over Time",
    x = "Year",
    y = "Total Extent (km²)"
  ) +
  theme_minimal(base_size = 14)

```


# Extent: 2021
# This is the CONABIO Land use and vegetation data : http://www.conabio.gob.mx/informacion/gis/?vns=gis_root/usv/inegi/usv250s7gw

```{r}
# 2021
dune_canabio_2021_sublayer_5km <- dune_canabio_landuse_2021 |>
  dplyr::filter(
    DESCRIPCIO %in% c(
      "VEGETACIÓN DE DUNAS COSTERAS"
    )
  )

dune_canabio_2021_sublayer_1km <- dune_canabio_landuse_2021 |>
  dplyr::filter(
    DESCRIPCIO %in% c(
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA XERÓFILA",
      "VEGETACIÓN HALÓFILA XERÓFILA"
    )
  )

# Match CRS of OHI and Buffer OHI regions
ohi_region <- st_transform(ohi_regions, st_crs(dune_canabio_landuse_2021))
ohi_marine_plus_5km <- st_transform(ohi_marine_plus_5km, st_crs(dune_canabio_landuse_2021))
ohi_marine_plus_1km <- st_transform(ohi_marine_plus_1km, st_crs(dune_canabio_landuse_2021))

dune_canabio_2021_sublayer_5km <- st_intersection(dune_canabio_2021_sublayer_5km, ohi_marine_plus_5km)
dune_canabio_2021_sublayer_1km <- st_intersection(dune_canabio_2021_sublayer_1km, ohi_marine_plus_1km)

dune_canabio_2021_sublayer <- st_union(dune_canabio_2021_sublayer_5km, dune_canabio_2021_sublayer_1km)

dune_canabio_2021_sublayer_area <- st_intersection(dune_canabio_2021_sublayer, ohi_region) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2021
  ) |>
  st_drop_geometry()

# Make a nice kable table
dune_canabio_2021_sublayer_area |>
select(rgn_id, year, area_km2) |>
  kable(
    digits = 4,
    col.names = c('Region ID', 'Year', 'Extent (km²)'),
    align = c('c', 'c', 'r')
  ) |>
  kable_styling(
    bootstrap_options = c('striped', 'bordered', 'hover')
  ) |>
  add_header_above(
    c('Coastal Dune Extent - CONABIO Land Use and Vegetation (2021)' = 3)
  )
```




# Trend: 2021/2013

```{r}
# File 3: coastal_dune_trend.csv using CANABIO LU data
coastal_dune_trend <- bind_rows(dune_canabio_2021_sublayer_area, dune_canabio_2013_sublayer_area) |>
  pivot_wider(names_from = year, values_from = area_km2) |>
  mutate(
    trend = (`2021` / `2013`)^(1/8) - 1,
    year = 2021
  ) |>
  select(rgn_id, year, trend)

coastal_dune_trend |>
  kable(digits = 4, col.names = c('Region ID', 'Year', 'Trend')) |>
  kable_styling(bootstrap_options = c('striped', 'bordered', 'hover')) |>
  add_header_above(c('Coastal Dune Trend CONABIO 2021/2013' = 3))
```





# Condition

```{r}
# File 1: coastal_dune_condition using CANABIO LU data
coastal_dune_condition <- bind_rows(dune_canabio_2001_sublayer_area, dune_canabio_2021_sublayer_area) |>
  pivot_wider(names_from = year, values_from = area_km2) |>
  mutate(condition = `2021` / `2001`, 
         condition = pmin(condition, 1),   # cap values > 1 at 1
         year = 2021) |>
  select(rgn_id, year, condition)

coastal_dune_condition |>
  kable(digits = 4, col.names = c('Region ID', 'Year', 
                                   'Condition (2021/2001)')) |>
  kable_styling(bootstrap_options = c('striped', 'bordered', 'hover')) |>
  add_header_above(c('Coastal Dune Condition' = 3)) # make above 1 values 1, and 2 decimal places 
```

```{r}
# Write CSVs to output_data directory
write.csv(dune_canabio_2021_sublayer_area,
          "habitat_services_output/coastal_dune_extent_2021.csv",
          row.names = FALSE)

write.csv(coastal_dune_trend,
          "habitat_services_output/coastal_dune_trend.csv",
          row.names = FALSE)

write.csv(coastal_dune_condition,
          "habitat_services_output/coastal_dune_condition.csv",
          row.names = FALSE)
```

