---
title: 'OHI GoC 2025: SP ICO Gulf of California'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# Summary

The Iconic Species (ICO) subgoal of Sense of Place captures the condition of iconic species that are relevant to local cultural identity.  We must (1) create a list of iconic species and which OHI regions they belong to, (2) determine their condition by their risk status, such as from NOM-059-SEMARNAT-2010, or IUCN, and (3) then, for each region, the average of the status for all iconic species can be calculated.

(1) = script1_iconic_species_list.qmd
(2) = script2_norma059.qmd, script3_cites_prep.qmd, script4_iucn_prep.qmd
(3) = script5_ico_current_status.qmd

In this fifth script, we will be assigning the condition weights to each species in this order:

NOM -> literature -> IUCN -> CITES


To so do, lets bring in the cleaned intermediate weights for each species from each source.


## ICO List

```{r}
ico_spp_rgnid <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/ico/int","ico_species_rgnid_040626.csv"))

colnames(ico_spp_rgnid)

ico_spp_rgnid_complete <- ico_spp_rgnid %>%  # so that every species and region has the years 2010:2025
  mutate(year = NA) %>% 
  arrange(taxon_scientific_name, rgn_id) %>%
  group_by(taxon_scientific_name, rgn_id) %>% 
  complete(year = 2010:2025) %>% 
  filter(!is.na(year)) # because I introduced year as NA, I need to remove it now.

length(unique(ico_spp_rgnid$taxon_scientific_name)) # 117
```


## NOM-059-SEMARNAT-2010

```{r}
nom059_condition_ico_spp <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/ico/int","nom059_condition_ico_spp_040726.csv"))

colnames(nom059_condition_ico_spp)
# [1] "taxon_scientific_name" "year"                  "categoria_de_riesgo"  
# [4] "w_cat_nom059"          "rgn_id"                "nombres_comunes" 

```


## IUCN

```{r}
iucn_condition_ico_spp <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/ico/int","iucn_condition_ico_spp_040726_only_ico.csv"))

colnames(iucn_condition_ico_spp)
# [1] "taxon_scientific_name"  "rgn_id"                 "year"                  
# [4] "red_list_category_code" "w_cat_iucn"       
```


## CITES

```{r}
cites_condition_ico_spp <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/ico/int","cites_condition_ico_spp_040726.csv"))

colnames(cites_condition_ico_spp)
# [1] "taxon_scientific_name" "appendix_listing"      "year"                 
# [4] "rgn_id"                "w_cat_cites"  
```

## Merge

```{r}
ico_spp_all_conditions <- full_join(ico_spp_rgnid_complete, nom059_condition_ico_spp) %>% # should have the same # of obs since NOM gives a 1 to all iconic species not on the list (v2026: 9648)
  full_join(cites_condition_ico_spp) %>%  
  full_join(iucn_condition_ico_spp) #by = c("taxon_scientific_name","rgn_id","year"))

# species on the red list try to get reassessed every 5-10 years, but for this case we can say if the year is after the assessment and no changes have been made, give that species in that region and subsequent years the same red list category.
ico_spp_all_conditions_iucnfill <- ico_spp_all_conditions %>% 
  arrange(taxon_scientific_name, rgn_id, year) %>%
  group_by(taxon_scientific_name, rgn_id) %>% 
  fill(w_cat_iucn, .direction = "down") %>% # leaving the iucn `red_list_category_code` unfilled to represent which years did not have a new IUCN assessment.
  ungroup()
```

## Current status

```{r}
# first, we need to assign only one risk category weight (w_cat) for each species, rgn it occurs, and year.
ico_spp_finalize_risk <- ico_spp_all_conditions_iucnfill %>% 
  mutate(w_cat = case_when(
    
    w_cat_nom059 == 1.0 & !is.na(w_cat_iucn) ~ w_cat_iucn, 
    w_cat_nom059 == 1.0 & is.na(w_cat_iucn) & !is.na(w_cat_cites) ~ w_cat_cites,
    TRUE ~ w_cat_nom059
    
    ## need to identify for which species not being on the norma is significant and accurate. Endemic vs not endemic. 
    ## need to double check the ones that are not on the norma, but are listed as at risk IUCN or CITEs, do a quick run through.  Especially because IUCN includes Atlantic and Pacific.  Need to account for regional differences.
    
  ))

# check the logic for everything that has w_cat_nom059 = 1. 
check_no_risk <- ico_spp_finalize_risk %>% 
  filter(w_cat_nom059 == 1.0) 

check_no_risk_w_iucn1 <- check_no_risk %>% 
  filter(w_cat_iucn < 1) # go through these, read through IUCN and how the Pacific vs Atlantic stock are doing. And if Pacific is doing well, we can defer to the norma!!
unique(check_no_risk_w_iucn1$taxon_scientific_name)
#  [1] "Cynoscion othonopterus"   "Epinephelus itajara"     
#  [3] "Hyporthodus acanthistius" "Hyporthodus nigritus"    
#  [5] "Hyporthodus niveatus"     "Istiophorus platypterus" 
#  [7] "Kajikia audax"            "Leuresthes sardina"      
#  [9] "Makaira nigricans"        "Mobula tarapacana"       
# [11] "Mobula thurstoni"         "Mycteroperca jordani"    
# [13] "Pavona gigantea"          "Pocillopora damicornis"  
# [15] "Pocillopora meandrina"    "Pocillopora verrucosa"   
# [17] "Porites panamensis"       "Sphyrna corona"          
# [19] "Sphyrna lewini"           "Sphyrna media"           
# [21] "Sphyrna mokarran"         "Sphyrna tiburo"          
# [23] "Sphyrna zygaena"          "Thalasseus elegans"      
# [25] "Xiphias gladius"  
# <https://docs.google.com/spreadsheets/d/1D6VWMVoEhhkJf7aqx0DaqKFaP6XKi8OW1fAR9l_XfhA/edit?gid=2131308524#gid=2131308524> 

check_no_risk_w_cites1 <- check_no_risk %>% 
  filter(w_cat_cites < 1) 
#%>% 
 # filter(is.na(w_cat_iucn))
unique(check_no_risk_w_cites1$taxon_scientific_name)
# [1] "Pachycereus pringlei" "Sphyrna lewini"       "Sphyrna mokarran"    
# [4] "Sphyrna zygaena" 
# <https://docs.google.com/spreadsheets/d/1D6VWMVoEhhkJf7aqx0DaqKFaP6XKi8OW1fAR9l_XfhA/edit?gid=2124372570#gid=2124372570> 

```

Now that the logic is sound, lets average the values across all species in each region. 

```{r}
# first, I am just curious at how many species there are in each region. Based on the occurrence data, just merely for me to see.
ico_spp_per_rgn <- ico_spp_finalize_risk %>% 
  group_by(rgn_id) %>% 
  summarise(
    n_species = n_distinct(taxon_scientific_name)
  )

# now, let us determine the value for each region and year.
ico_mean_value_per_rgn <- ico_spp_finalize_risk %>% 
  summarise(
    mean_value = mean(w_cat),
    .by = c(year, rgn_id)
  ) %>% 
  mutate(current_status = mean_value * 100) %>% 
  ungroup()


```

### Write to Mazu

```{r}
write_csv(ico_mean_value_per_rgn, here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/ico/output","ico_mean_risk_status_per_region_040726.csv"))
```



```{r}
library(plotly)

## for plotting:
rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

ico_plotly <- plotly::plot_ly(
  ico_mean_value_per_rgn, 
  x = ~year, 
  y = ~current_status, 
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Risk Status: average condition (risk status score) for all iconic species occurring in each region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(ico_mean_value_per_rgn$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
ico_plotly
```

## Explore each region

Let's look at Region 1, which is doing the worst, and see why.

```{r}
ico_clean_species_occurrence_sf <- st_read(here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/ico/int","ico_clean_species_occurrence_sf.shp"))
ico_clean_species_occurrence_goc_rgnid <- st_read(here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/ico/int","ico_clean_species_occurrence_goc_rgnid.shp"))

mapview(ico_clean_species_occurrence_sf)

# hm, I am seeing a lot of totoaba in region 1, which has a w_cat = 0.25 for all years... 
ico_clean_species_occurrence_sf_totoaba <- ico_clean_species_occurrence_goc_rgnid %>% 
  filter(old_name %in% "Totoaba macdonaldi")
mapview(ico_clean_species_occurrence_sf_totoaba, zcol = "rgn_id")
```

```{r}
## lets take a look at the frequency of occurrence of species in Region 1 (even though that does not have an impact on the current status)
ico_species_occurrence_total_per_rgn <- ico_clean_species_occurrence_goc_rgnid %>% 
  group_by(rgn_id, old_name) %>% 
  summarize(
    n_occurrence = n()
  ) %>% 
  ungroup()

## WOW! So many vaquita, which also across all years have a risk status of 0.25! 
ico_clean_species_occurrence_sf_vaquita <- ico_clean_species_occurrence_goc_rgnid %>% 
  filter(old_name %in% "Phocoena sinus")
mapview(ico_clean_species_occurrence_sf_vaquita, zcol = "rgn_id")

## lets look at the gulf corvina, which was the second most occurring in Region 1, to see if other regions also contain it:
ico_clean_species_occurrence_sf_gulfcorvina <- ico_clean_species_occurrence_goc_rgnid %>% 
  filter(old_name %in% "Cynoscion othonopterus")
mapview(ico_clean_species_occurrence_sf_gulfcorvina, zcol = "rgn_id")
# they do, but it is the other two lowest scores as well: Region 2 and Region 5.
```

It appears that Region 1 contains many endemic species that are also at a high risk of extinction.

