---
title: 'OHI GoC 2025: SP ICO Gulf of California'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# Summary

The Iconic Species (ICO) subgoal of Sense of Place captures the condition of iconic species that are relevant to local cultural identity.  We must (1) create a list of iconic species and which OHI regions they belong to, (2) determine their condition by their risk status, such as from NOM-059-SEMARNAT-2010, or IUCN, and (3) then, for each region, the average of the status for all iconic species can be calculated.

(1) = script1_iconic_species_list.qmd
(2) = script2_norma059.qmd, script3_cites_prep.qmd, script4_iucn_prep.qmd
(3) = script5_ico_current_status.qmd

In this first script, we will be bringing in the list developed by the Expert Working Group and assigning each species to a OHI GoC Region (1-9) based on a species occurence (presence and absence) database from Dr. Hem Nalini Morzaria-Luna.


## Read in list

```{r}
especies_iconicas <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/ico","especies_iconicas_040326.csv"))
```


## Clean to prep for joining with risk condition 


```{r}
especies_iconicas_scinames <- especies_iconicas %>% 
  mutate(sci_nombre = str_replace_all(sci_nombre, "spp.",""))
```


## Determine species occurence throughout the Gulf


```{r}

```



## Assign OHI GoC Region ID

```{r}

```





















