---
title: 'OHI GoC 2025: SP ICO Gulf of California'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# Summary

The Iconic Species (ICO) subgoal of Sense of Place captures the condition of iconic species that are relevant to local cultural identity.  We must (1) create a list of iconic species and which OHI regions they belong to, (2) determine their condition by their risk status, such as from NOM-059-SEMARNAT-2010, or IUCN, and (3) then, for each region, the average of the status for all iconic species can be calculated.

(1) = script1_iconic_species_list.qmd
(2) = script2_norma059.qmd, script3_cites_prep.qmd, script4_iucn_prep.qmd
(3) = script5_ico_current_status.qmd

In this second script, we will be cleaning the two NORMA-059-SEMARNAT releases, 2010 and 2019.  In 2021, they re-released the same list as 2019, deciding there were no changes.

The NOM-059-SEMARNAT-2010 is divided into amphibians, birds, fungi, invertebrates, mammals, fish, plants and reptiles and consider four risk categories:

- Subject to special protection (Pr)
- Threatened (A)
- Endangered (P)
- Probably extinct in the wild (E)

[NOM-059-SEMARNAT-2010](https://www.dof.gob.mx/normasOficiales/4254/semarnat/semarnat.htm)
[2019 Modification](https://www.dof.gob.mx/nota_detalle.php?codigo=5578808&fecha=14/11/2019#gsc.tab=0)
[Conoce las categorías de riesgo de la NOM 059 SEMARNAT-2010 para especies de flora y fauna](https://www.gob.mx/semarnat/articulos/conoce-las-categorias-de-riesgo-de-la-nom-059-semarnat-2010-para-especies-de-flora-y-fauna)

So, if none of the iconic species located within an OHI Region are listed on NOM-059-SEMARNAT-2010, nor CITES/IUCN, the subgoal would have a current status of 100 for that region and year.

*We plan to rank the status of the species as such:*

Not listed = 1
Subject to special protection (Pr) = 0.75
Threatened (A) = 0.5
Endangered (P) = 0.25
Probably extinct in the wild (E) = 0

To use this, we can assign approximate rankings for the status of species based on their listing.  So for example, if an iconic species is not listed, it would receive a 1.  If the species is subject to special protection it would receive a 0.75, threatened receive a 0.5, endangered a 0.25, and extinct a zero.  

After that, for each region, the average of the status for all iconic species that are found within that region can be calculated.

## Specifics on each risk category

*Subject to special protection (Pr)*

Those that could be threatened by factors that negatively affect their viability, thus determining the need to promote their recovery and conservation or the recovery and conservation of populations of associated species.

Examples:

Sonoran green toad
Great blue heron, great blue heron, Holy Spirit heron
Sea cucumber


*Threatened (A)*

Those that could be in danger of disappearing in the short or medium term, if the factors that negatively affect their viability continue to operate, by causing the deterioration or modification of their habitat or directly decreasing the size of their populations.

Examples:

Cozumel Badger
Elephant seal
Tuza de Querétaro


*Endangered (P)*

Species whose distribution areas or population sizes in the national territory have drastically decreased, putting their biological viability at risk throughout their natural habitat, due to factors such as habitat destruction or drastic modification, unsustainable exploitation, diseases or predation, among others.

Examples:

Poblano frog
Fine wolf of Guadalupe
Little sardine fish Cuatrociénegas


*Probably extinct in the wild (E)*

That species native to Mexico whose specimens in the wild within the national territory have disappeared, as far as the documentation and studies carried out prove, and of which the existence of live specimens is known, in confinement or outside the Mexican territory.

Examples:

Socorro Island Dove
Grizzly bear, brown bear
Sea of Cortez finch


## 2010

```{r}
norma_2010 <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/NOM-059-SEMARNAT-2010","NOM-059-SEMARNAT-2010.csv")) %>% 
  janitor::clean_names() %>% 
  mutate(ano = 2010)
```


## 2019

```{r}
norma_2019 <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/NOM-059-SEMARNAT-2010","MODIFIED_2019_NOM-059-SEMARNAT-2010.csv")) %>% 
  janitor::clean_names() %>% 
  mutate(ano = 2019)
```


## 2021

```{r}
norma_2021 <- norma_2019 %>% 
  mutate(ano = 2021) # no changes since 2019
```


## 2025

```{r}
norma_2025 <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/NOM-059-SEMARNAT-2010","MODIFIED_2025_NOM-059-SEMARNAT-2010.csv")) %>% 
  janitor::clean_names() %>% 
  mutate(ano = 2025)
```



# Merge all years

```{r}
colnames(norma_2010)
colnames(norma_2019)
colnames(norma_2025)

setdiff(colnames(norma_2025), colnames(norma_2019)) # the same, lets change 2010 to reflect the same name

norma_2010_merge <- norma_2010 %>% 
  mutate(categoria_infraespecifica = ifelse(!is.na(subespecie), "subsp.", NA_character_)) %>% 
  select(familia, genero, especie, categoria_infraespecifica, nombre_infraespecifico = subespecie, autor = metodo, sinonimia, nombres_comunes = nombre, distribucion, categoria_de_riesgo = categoria, ano)
```


```{r}
norma_all_years <- bind_rows(norma_2010_merge, norma_2019, norma_2021, norma_2025)
```







