---
title: 'OHI GoC 2025: SP ICO Gulf of California'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# Summary

The Iconic Species (ICO) subgoal of Sense of Place captures the condition of iconic species that are relevant to local cultural identity.  We must (1) create a list of iconic species and which OHI regions they belong to, (2) determine their condition by their risk status, such as from NOM-059-SEMARNAT-2010, or IUCN, and (3) then, for each region, the average of the status for all iconic species can be calculated.

(1) = script1_iconic_species_list.qmd
(2) = script2_norma059.qmd, script3_cites_prep.qmd, script4_iucn_prep.qmd
(3) = script5_ico_current_status.qmd

In this fourth script, we will be cleaning up IUCN information and assessing iconic species condition based on the IUCN Red List Categories and Criteria.

Note that if none of the iconic species are listed on NOM-059-SEMARNAT-2010, we will check to see if they are listed as *at risk* by CITES/IUCN.  If not, the subgoal would have a current status of 100 for that region and year.

The IUCN Red List divides species into nine categories: *Not Evaluated, Data Deficient, Least Concern, Near Threatened, Vulnerable, Endangered, Critically Endangered, Extinct in the Wild and Extinct.*

-   **Reference**: IUCN 2024. IUCN Red List of Threatened Species. Version 2025-1 \<www.iucnredlist.org\>
-   **Accessed**: July 15, 2025
-   **Download**: Download is done through iucnredlist API
-   **Native data resolution**: Country level (by country name)
-   **Time range**: 1965-2024 (discrete past assessments by species)
-   **Format**: JSON

Using the IUCN API, we accessed the full IUCN species list. With some minor formatting, this list contains the following variables:

iucn_sid \| kingdom \| phylum \| class \| order \| family \| genus \| sciname \| subpopulation \| category

## Download Species List from IUCN

Get total species list, using the IUCN API.

To access the API v4, you will need to sign up for a key on the [IUCN website](https://api.iucnredlist.org/). This key can be added to the local .renviron on your machine and retrieved using the code below.


### Setup

```{r}
if (!require(librarian)){install.packages("librarian")}
if (!require(ohicore)){devtools::install_github('ohi-science/ohicore@dev')}
librarian::shelf(
  tidyverse,
  here,
  janitor,
  jsonlite,
  plotly,
  parallel,
  devtools,
  ohicore,
  kableExtra,
  magrittr,
  rredlist
)  
```


```{r set up API}
library(rredlist)

remotes::install_github("iucn-uk/iucnredlist")
library(iucnredlist)

if (!requireNamespace("pak", quietly = TRUE)) {
  install.packages("pak", dependencies = TRUE)
}
pak::pkg_install("stangandaho/redlist")

library(redlist)

rl_check_api()
```

```{r}
rl_check_api()

rl_countries()

# 158 Mexico MX
mexico_data <- rl_countries(code = "MX", latest = TRUE)

mexico_2023 <- rl_countries(code = "MX", year_published = 2023)
mexico_2024 <- rl_countries(code = "MX", year_published = 2024)

```

```{r}
str(mexico_data)
names(mexico_data)

mexico_data_df <- bind_rows(mexico_data$assessments) %>% 
  dplyr::select(-scopes) # since all are Global, 1

library(dplyr)
library(stringr)

target_species <- c(
  "Tivela planulata", "Megapitaria aurantiaca", "Megapitaria squalida",
  "Penaeus vannamei", "Penaeus stylirostris", "Penaeus californiensis",
  "Penaeus brevirostris", "Pinna rugosa", "Atrina tuberculosa",
  "Atrina maura", "Pteria sterna", "Pinctada mazatlanica",
  "Chione californiensis", "Chione undatella", "Ilochione subrugosa",
  "Coryphaena hippurus", "Kajikia audax", "Tetrapturus audax",
  "Makaira nigricans", "Makaira ampla", "Istiompax indica",
  "Makaira indica", "Tetrapturus albidus", "Kajikia albida",
  "Istiophorus platypterus", "Istiophorus albicans", "Xiphias gladius",
  "Tetrapturus angustirostris", "Rhincodon typus", "Sphyrna",
  "Mycteroperca", "Hyporthodus", "Stereolepis", "Epinephelus",
  "Lutjanus peru", "Balistes polylepis", "Cynoscion othonopterus",
  "Leuresthes sardina", "Zostera marina", "Distichlis palmeri",
  "Pachycereus pringlei", "Rhizophora mangle", "Totoaba macdonaldi",
  "Zalophus californianus", "Phocoena sinus", "Tursiops truncatus",
  "Stenella attenuata", "Stenella longirostris", "Stenella coeruleoalba",
  "Delphinus delphis", "Delphinus capensis", "Lagenorhynchus obliquidens",
  "Grampus griseus", "Caretta caretta", "Chelonia agassizii",
  "Chelonia mydas", "Eretmochelys imbricata", "Lepidochelys olivacea",
  "Dermochelys coriacea", "Orcinus orca", "Physeter macrocephalus",
  "Balaenoptera musculus", "Balaenoptera physalus", "Megaptera novaeangliae",
  "Eschrichtius robustus", "Larus livens", "Thalasseus elegans",
  "Larus heermanni", "Endomychura craveri", "Sula leucogaster",
  "Sula nebouxii", "Fregata magnificens", "Pandion haliaetus",
  "Pelecanus occidentalis", "Mobula", "Dosidicus gigas",
  "Crocodylus acutus", "Myotis vivesi", "Pavona gigantea",
  "Pavona clavus", "Porites panamensis", "Pocillopora verrucosa",
  "Pocillopora damicornis", "Pocillopora meandrina"
)

# Build a regex pattern from the genus/species names
pattern <- paste(target_species, collapse = "|")

mexico_focal <- mexico_data_df %>%
  filter(str_detect(taxon_scientific_name, regex(pattern, ignore_case = TRUE)))

# See which of your targets were found
found <- target_species[sapply(target_species, function(sp) 
  any(str_detect(mexico_focal$taxon_scientific_name, regex(sp, ignore_case = TRUE))))]

# See which were NOT found
missing <- setdiff(target_species, found)
missing

#  [1] "Tivela planulata"       "Megapitaria aurantiaca" "Megapitaria squalida"   "Penaeus vannamei"      
#  [5] "Penaeus stylirostris"   "Penaeus californiensis" "Penaeus brevirostris"   "Pinna rugosa"          
#  [9] "Atrina tuberculosa"     "Atrina maura"           "Pteria sterna"          "Pinctada mazatlanica"  
# [13] "Chione californiensis"  "Chione undatella"       "Ilochione subrugosa"    "Tetrapturus audax"     
# [17] "Makaira ampla"          "Makaira indica"         "Tetrapturus albidus"    "Istiophorus albicans"  
# [21] "Distichlis palmeri"     "Delphinus capensis"     "Chelonia agassizii"     "Endomychura craveri" 
```


I want to see change over time, so...

```{r}
years <- 2000:2025

mexico_all_years <- purrr::map(years, function(yr) {
  Sys.sleep(0.5)
  rl_countries(code = "MX", year_published = yr)$assessments %>%
    filter(str_detect(taxon_scientific_name, regex(pattern, ignore_case = TRUE)))
}) %>% 
  bind_rows()

unique(mexico_all_years$taxon_scientific_name)
```


















