################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```

# CPUE Estimates per Major_Functional_Group ------------------------------

# STEP 1: Interrogate the economic unit (eu_rnpa) and vessel (vessel_name) columns. ------------------------------

```{r Load in dataframe as needed}
GOC_MarineLandings0<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Marine.Landings_Sector.Sorted.Georef_PATCHED_Jun032026.rds"))
# 3246583 obs., 58 vars.

dim(GOC_MarineLandings0)
# 3246583 obs., 58 vars.

# NOTE TO SELF: This file path will change when I create the "fisheries only" dataframe. - June 1 2026
# And when I patch any remaining species, genus, family, or sector adjustments - June 3 2026

table(GOC_MarineLandings0$GOC_Include_Exclude)
```

```{r Filter out sectors of interest}
MarineFishery_Sectors<- c("marine: fishery sector yet to be classified" , 
                          "marine: small-scale capture", 
                          "marine: large-scale capture")

GOC_MarineFishery_Landings<- GOC_MarineLandings0 %>% 
  filter(sector_estimate_draft %in% MarineFishery_Sectors)

dim(GOC_MarineFishery_Landings)
# 3039221 obs., 58 vars.

range(GOC_MarineFishery_Landings$year)
```

```{r First look}
length(unique(GOC_MarineFishery_Landings$eu_name))
# 4774 unique economic units in the dataframe.

length(unique(GOC_MarineFishery_Landings$eu_rnpa))
# 4435 unique permits given to economic units in the dataframe.

length(unique(GOC_MarineFishery_Landings$vessel_name))
# 1981 unique vessel names in the dataframe.

length(unique(GOC_MarineFishery_Landings$vessel_rnpa))
# 1974 unique vessel permit numbers in the dataframe.

# So, for the most part, each vessel is registered to a unique vessel permit.

# How many of the 3229063 total observations have no economic unit name associated with them? Only ~2% of all records.
GOC_MarineFishery_Landings %>%
  count(eu_name == "NO DISPONIBLE")
# 62343 / 3039221

# How many of all observations have no vessel name associated with them? About 50% of all records.
GOC_MarineFishery_Landings %>%
  count(vessel_rnpa == "00000000")
# 1530584 / 3039221

# This tells us that about half of all records have no information about the fishers' vessel, but nearly all (98%) have permit info.
```

```{r Testing some ideas}
table(GOC_MarineFishery_Landings$reported_aquaculture_production)
table(GOC_MarineFishery_Landings$est_aquaculture_production)
# Potentially a small number of aquaculture records hiding in here (>0.03%)...

# Look into a specific unit:
subset<- GOC_MarineFishery_Landings %>%
  filter(eu_name == "SCPP DE SERV LA CAIMANERA SC DE RL")

table(subset$reported_aquaculture_production)
# Notice that some of these economic units say nothing about aquaculture, whereas others do. 

# Look into another specific unit:
subset2<- GOC_MarineFishery_Landings %>%
  filter(eu_name == "SCPP PUNTA TIBURON, S DE RL DE CV")

table(subset2$reported_aquaculture_production)
# It looks like this particular economic unit (possibly a cooperative?) does not work in aquaculture. 

# Keep workspace clean:
rm(subset, subset2)
```

```{r Let's plot missing data over time}
# Summarize missing vessel IDs over time
vessel_missing_over_time <- GOC_MarineFishery_Landings %>%
  mutate(missing_vessel = is.na(vessel_rnpa) | vessel_rnpa == "00000000" | vessel_rnpa == "0000000_") %>%
  group_by(year) %>%   # Replace 'year' with your time column
  summarise(
    total = n(),
    missing_count = sum(missing_vessel),
    missing_percent = 100 * missing_count / total
  )

# Plot missing vessel prevalence over time
ggplot(vessel_missing_over_time, aes(x = year, y = missing_percent)) +
  geom_line(color = "orange", size = 1.2) +
  geom_point(color = "tomato", size = 2) +
  labs(
    title = "Prevalence of Missing Vessel IDs Over Time",
    x = "Year",
    y = "Percentage of Records Missing Vessel ID"
  ) +
  scale_y_continuous(limits = c(0, 100)) +
  theme_bw()

# Vessel data is missing throughout the time series.

# Plot the opposite: what's not missing?
vessel_present_over_time <- GOC_MarineFishery_Landings %>%
  mutate(missing_vessel = is.na(vessel_rnpa) | vessel_rnpa == "00000000" | vessel_rnpa == "0000000_") %>%
  group_by(year) %>%
  summarise(
    total = n(),
    present_count = sum(!missing_vessel),
    present_percent = 100 * present_count / total
  )

ggplot(vessel_present_over_time, aes(x = year, y = present_percent)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_point(color = "navy", size = 2) +
  
  labs(
    title = "Proportion of Records WITH Vessel IDs Over Time",
    x = "Year",
    y = "Percentage of Records with Vessel ID"
  ) +
  scale_y_continuous(limits = c(0, 100)) +
  theme_bw()

# Plot the opposite: what's not missing?
eu_present_over_time <- GOC_MarineFishery_Landings %>%
  mutate(missing_eu = eu_name == "NO DISPONIBLE") %>%
  group_by(year) %>%
  summarise(
    total = n(),
    present_count = sum(!missing_eu),
    present_percent = 100 * present_count / total
  )

ggplot(eu_present_over_time, aes(x = year, y = present_percent)) +
  geom_line(color = "lavender", size = 1.2) +
  geom_point(color = "orchid", size = 2) +
  
  labs(
    title = "Proportion of Records WITH EU Details Over Time",
    x = "Year",
    y = "Percentage of Records with EU details"
  ) +
  scale_y_continuous(limits = c(0, 100)) +
  theme_bw()
```

```{r Investigate vessel names since 2018}
GOC_CPUE_Landings<- GOC_MarineFishery_Landings %>%
  filter(year >= 2018)
# 1123506 obs., 57 vars.

length(unique(GOC_CPUE_Landings$vessel_name))
# 1167 unique vessel names in the dataframe.

length(unique(GOC_CPUE_Landings$vessel_rnpa))
# 1148 unique vessel permit numbers in the dataframe.
# So, for the most part, each vessel is registered to a unique vessel permit.

# How many of the 1104876 total observations have no economic unit name associated with them? Essentially 0% of records.
GOC_CPUE_Landings %>%
  count(eu_name == "NO DISPONIBLE")
# 4 / 1123506

# How many of all observations have no vessel PERMIT associated with them? 87% of all records.
GOC_CPUE_Landings %>%
  count(vessel_rnpa == "00000000" | vessel_rnpa == "0000000_" | is.na(vessel_rnpa))
# 977629 / 1123506

# How many of all observations have no vessel NAME associated with them? About 87% of all records.
# Okay, so we know vessel name alone won't be a good estimator for CPUE...
GOC_CPUE_Landings %>%
  count(is.na(vessel_name))
# 977629 / 1123506

# Take a look at a random assortment of rnpas versus names to understand what's missing and what's not:
names_missing_slice <- GOC_CPUE_Landings %>%
  filter(is.na(vessel_name)) %>%
  select(vessel_rnpa, vessel_name, eu_name, eu_rnpa) %>%
  distinct() %>%
  slice_sample(n = 200)
# This shows that records missing vessel details do still have unique eu names.
```

```{r Testing some ideas}
colnames(GOC_CPUE_Landings)

# Case studies (eu_rnpa from names_missing_slice)
subset <- GOC_CPUE_Landings %>%
  filter(eu_rnpa == "2607603921")

length(unique(subset$vessel_name))
# 1 vessel registered to this economic unit permit

subset2<- GOC_CPUE_Landings %>%
  filter(eu_rnpa =="2611000700")

length(unique(subset2$vessel_name))
# 1 vessel name registered to this economic unit permit

length(unique(subset2$eu_name))
# 1 name for this economic unit

unique(subset2$vessel_rnpa)
# 1 vessel permit for this economic unit, 0000000_

range(subset2$year)
# From 2018 to 2025

unique(subset2$recorded_taxa_name_clean)
# Across 34 taxa

table(subset2$sector_estimate_reported)
# All small-scale fisheries, so at least the same eu doesn't appear to be switching between sector types.
```


# STEP 2: Esitmate CPUE per economic unit 2018-2024 
```{r Estimate CPUE per economic unit 2018-2024}
# Let's look through unit Economic Units. All records have a non-NA economic unit (at least since 2018)
EconomicUnits <- GOC_CPUE_Landings %>%
  distinct(eu_rnpa)    # Keeps unique values as a one-column tibble
# 2773 obs.

# Estimate Catch per economic unit 2018-2024: summarize weight_kg_corrected by eu_rnpa per year
CatchPerEconomicUnit_summary <- GOC_CPUE_Landings %>%
  filter(year >= 2018 & year <= 2024) %>%   # filter the years
  group_by(year, eu_rnpa) %>%              # group by year and economic unit
  summarize(
    total_catch_kg = sum(weight_kg_corrected, na.rm = TRUE)  # sum catch, ignore NAs
  ) %>%
  ungroup()  # remove grouping after summarizing
# 13163 obs., 3 vars.
```

```{r Pull out the eu_rnpas that are consistently present in all years 2018-2025}
# Using these eu_rnpas only, estimate the total catch of Major_Taxa_Group divided by the total eu_rnpa per year

# Step 1: Identify eu_rnpa present in all years
ConsistentEconomicUnits <- GOC_CPUE_Landings %>%
  filter(year >= 2018 & year <= 2025) %>%
  group_by(eu_rnpa) %>%
  summarize(years_present = n_distinct(year), .groups = "drop") %>%
  filter(years_present == 8) %>%    # 2018–2025 = 8 years
  pull(eu_rnpa)                     # get vector of eu_rnpa
# 1003 eu.rnpas are consistently present in all years 2018-2025

# Step 2: Filter data to only include these eu_rnpa
GOC_CPUE_Landings <- GOC_CPUE_Landings %>%
  filter(year >= 2018 & year <= 2025) %>%
  filter(eu_rnpa %in% ConsistentEconomicUnits)
# 860336 obs., 57 vars. (down -263170 obs.)

# SANITY CHECK: Do all landings have a Major_Functional_Group assigned? Yes!
sum(is.na(GOC_CPUE_Landings$Major_Functional_Group))
```

```{r Summarize catch per EU by Major_Functional_Group per year}
colnames(GOC_CPUE_Landings)

CPUE_summary <- GOC_CPUE_Landings %>%
  group_by(year, Major_Functional_Group) %>%
  summarize(
    total_catch_mt = sum(weight_kg_corrected, na.rm = TRUE)/1000,
    n_eu = n_distinct(eu_rnpa),                    # number of economic units that year
    catch.mt_per_eu = total_catch_mt / n_eu          # divide total catch by number of units
  ) %>%
  ungroup()

# Looking promising! Let's plot it...
```

```{r Plot CPUE for each Major_Functional_Group}
# PLOT FOR ALL GROUPS, excluding Small Pelagics:
# Remove NAs or 'unknown' if you don’t want them cluttering the plot
# First compute overall yearly mean (across all groups)

# First we need mean CPUE over years per group (2018-2024)
CPUE_mean_df <- CPUE_summary %>%
  filter(year > 2017 & year < 2025) %>%
  group_by(Major_Functional_Group) %>%
  summarise(mean_cpue = mean(catch.mt_per_eu), 
            .groups = "drop")

CPUE_perEU_plot <- 
  
  CPUE_summary %>%
  filter(year < 2025) %>%
  
  ggplot(aes(x = year, y = catch.mt_per_eu)) +
  
  # Group-specific lines
  geom_line(aes(color = Major_Functional_Group), size = 1.1) +
  geom_point(aes(color = Major_Functional_Group), size = 2) +
  
  facet_wrap(~ Major_Functional_Group, scales = "free_y") +
  
  geom_hline(
  data = CPUE_mean_df,
  aes(yintercept = mean_cpue, group = Major_Functional_Group),
  color = "grey40",
  linetype = "dashed",
  linewidth = 0.8) +
  
  labs(
    title = "CPUE per Economic Unit by Functional Group (2018-2024)",
    x = "Year",
    y = "Catch per Economic Unit (mt)"
  ) +
  
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

CPUE_perEU_plot
```

```{r Save CPUE_perEU_plot}
# Save the plot as a high-resolution PNG
ggsave(
  filename = "/home/ferrer/ohi_goc_prep/FP/CONAPESCA_CleanUp/CPUE_perEU_plot_Apr022026.png",
  plot = CPUE_perEU_plot,
  width = 10,       # width in inches
  height = 6,       # height in inches
  dpi = 300         # resolution
)
```

# STEP 2: Estimate CPUE now and over a "historic" baseline
```{r Establish what is the "historic" baseline?}
HistoricBaseline_low= 2018
HistoricBaseline_high= 2022

# It was previously 2003 to 2013 (December 2025), but now I will make it 2018-2022 to be consistent with the Carta baseline.
```

```{r Estimate CPUE for each Major Functional Group per region per year versus "historic" baseline}
# Turn off scientific notation:
options(scipen = 999)

colnames(CPUE_summary)

# Calculate CPUE values (at the group-level) over the "historic" baseline period defined above:
CPUE_baseline <- CPUE_summary %>%
  filter(year >= HistoricBaseline_low, year <= HistoricBaseline_high) %>%
  group_by(Major_Functional_Group) %>%
  summarise(
    CPUE_baseline_2018.2024 = mean(catch.mt_per_eu, na.rm = TRUE),
    n_eu_baseline = sum(n_eu, na.rm = TRUE),
    n_years_baseline = n_distinct(year),
    .groups = "drop"
  )


# Calculate CPUE values (at the group-level) from 2018 to 2024
# CPUE_ <- CPUE_summary %>%
#   filter(year >= 2018 & year <= 2024) %>%
#   group_by(year, Major_Functional_Group) %>%
#   summarise(
#     CPUE_current = catch.mt_per_eu,
#     n_eu = sum(n_eu, na.rm = TRUE),
#     .groups = "drop"
#   )
```

```{r Merge baseline to present}
options(scipen = 999)

# Merge CPUE_baseline to CPUE_recent:
colnames(CPUE_summary)

CPUE_FIS_scores <- CPUE_summary %>%
  left_join(CPUE_baseline,
            by = c("Major_Functional_Group")) %>%
  mutate(
    CPUE_score = catch.mt_per_eu / CPUE_baseline_2018.2024
  ) %>%
  filter(year<=2024)

# Cap CPUE_score at a max of 1:
CPUE_FIS_scores <- CPUE_FIS_scores %>%
  mutate(
    CPUE_score_capped = if_else(
      CPUE_score > 1,
      1,
      CPUE_score
    )
  )
```

```{r Add a GOC region column}
CPUE_FIS_scores$region<- c("GOC")

CPUE_FIS_scores<- CPUE_FIS_scores %>%
  relocate(region, .before= year)
```


# FINAL STEP: SAVE the CPUE_FIS_scores dataframe:
```{r Save this CPUE data table}
write.csv(CPUE_FIS_scores, 
"/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_CPUE_Scores/FIS_CPUE_Scores_Jun032026.csv", 
row.names = FALSE)
```


# UNDER CONSTRUCTION / PURG: Other Inquiries - June 1 2026

Estimate how much biomass per group?
```{r COMPUTE PERCENT BIOMASS FOR RECENT YEARS}
CPUE_recent <- CPUE_recent %>%
  group_by(year) %>%
  mutate(
    total_biomass_state_year = sum(Total_Biomass_kg, na.rm = TRUE),
    pct_of_total_biomass = (Total_Biomass_kg / total_biomass_state_year) * 100
  ) %>%
  ungroup()
```

Take a closer look at CPUE per Region and per Major_Functional_Group:
```{r Plot CPUE over all years HistoricBaseline_low-2024}
CPUE_AllYears <- GOC_CPUE_Landings %>%
  filter(year >= HistoricBaseline_low, year <= 2024) %>%
  filter(!is.na(OHI_region_est)) %>%
  group_by(OHI_region_est, year, Major_Functional_Group) %>%
  summarise(
    Total_Biomass_kg = sum(weight_kg_corrected, na.rm = TRUE),
    Total_Value_mxn = sum(value, na.rm = TRUE),
    Total_Days = sum(days_fishing, na.rm = TRUE),
    Avg_CPUE = mean(CPUE_kg.day, na.rm = TRUE),
    n_records = n(),
    .groups = "drop"
  )

Taxa_colors <- c("darkblue", "royalblue", "skyblue", "orange", "sienna1", "plum", "purple4", "aquamarine3", "gold2", "tomato", "grey")

Region_colors <- c("royalblue", "skyblue", "orange", "sienna1", "plum", "purple4", "aquamarine3", "gold2", "tomato")

Regions_of_Interest <- c(
  "Region_1", "Region_2", "Region_3",
  "Region_4", "Region_5", "Region_6",
  "Region_7", "Region_8", "Region_9"
)


# PLOT 1: CPUE overTime per Region, facet by Major_Functional_Group
CPUE_overTime_byRegion_perTaxa<- CPUE_AllYears %>%
  filter(OHI_region_est %in% Regions_of_Interest) %>%
  ggplot(
       aes(x = year, 
           y = Avg_CPUE, 
           group = OHI_region_est, 
           color = OHI_region_est)) +
  geom_line(size = 1) +                # lines connecting the points
  facet_wrap(~ Major_Functional_Group, scales = "free_y") +  # facet by region
  labs(
    title = "Average CPUE Over Time by Functional Group",
    x = "Year",
    y = "Avg CPUE (kg/day)",
    color = "Functional Group"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_color_manual(values = Taxa_colors) +  # nice palette
  geom_vline(xintercept = HistoricBaseline_high, linetype= "solid", color = "grey5")

CPUE_overTime_byRegion_perTaxa

  # # Save as PNG
  # ggsave(
  #   filename = sprintf("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_CPUE_FIS_scores/CPUE_PlotOverTime_byRegion_perTaxa_Dec092025.png"),
  #   plot = CPUE_overTime_byRegion_perTaxa,
  #   width = 12,
  #   height = 12,
  #   dpi = 300
  # )


# PLOT 2: CPUE overTime per Major_Functional_Group, facet by Region
CPUE_overTime_byRegion_perTaxa<- CPUE_AllYears %>%
  filter(OHI_region_est %in% Regions_of_Interest) %>%
  ggplot(
       aes(x = year, 
           y = Avg_CPUE, 
           group = Major_Functional_Group, 
           color = Major_Functional_Group)) +
  geom_line(size = 1) +                # lines connecting the points
  facet_wrap(~ OHI_region_est, scales = "free_y") +  # facet by region
  labs(
    title = "Average CPUE Over Time by Functional Group",
    x = "Year",
    y = "Avg CPUE (kg/day)",
    color = "Functional Group"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_color_manual(values = Taxa_colors) +  # nice palette
  geom_vline(xintercept = HistoricBaseline_high, linetype= "solid", color = "grey5")

CPUE_overTime_byRegion_perTaxa

  # # Save as PNG
  # ggsave(
  #   filename = sprintf("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_CPUE_FIS_scores/CPUE_PlotOverTime_byTaxa_perRegion_Dec092025.png"),
  #   plot = CPUE_overTime_byTaxa,
  #   width = 12,
  #   height = 12,
  #   dpi = 300
  # )


# PLOT 3: CPUE overTime per Major_Functional_Group, facet by Region - normalized 0 to 1
CPUE_overTime_byRegion_perTaxa_normalized<- CPUE_AllYears %>%
  filter(OHI_region_est %in% Regions_of_Interest) %>%
  # Normalize Avg_CPUE per functional group across all years
  group_by(Major_Functional_Group) %>%
  mutate(Avg_CPUE_norm = (Avg_CPUE - min(Avg_CPUE, na.rm = TRUE)) /
                            (max(Avg_CPUE, na.rm = TRUE) - min(Avg_CPUE, na.rm = TRUE))) %>%
  ungroup() %>%
  ggplot(aes(
    x = year, 
    y = Avg_CPUE_norm, 
    group = Major_Functional_Group, 
    color = Major_Functional_Group
  )) +
  geom_line(size = 1) +
  facet_wrap(~ OHI_region_est, scales = "free_y") +
  labs(
    title = "Normalized CPUE Over Time by Functional Group",
    x = "Year",
    y = "Normalized CPUE (0-1)",
    color = "Functional Group"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_color_manual(values = Taxa_colors) +
  geom_vline(xintercept = HistoricBaseline_high, linetype = "solid", color = "grey5")

CPUE_overTime_byRegion_perTaxa_normalized

  # Save as PNG
  # ggsave(
  #   filename = sprintf("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_CPUE_FIS_scores/CPUE_PlotOverTime_byTaxa_perRegion_normalized_Dec092025.png"),
  #   plot = CPUE_overTime_byTaxa_normalized,
  #   width = 12,
  #   height = 12,
  #   dpi = 300
  # )
```