---
title: 'OHI GoC 2025: Coastline Classification'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: inline
editor_options: 
  chunk_output_type: console
---

# Summary

## GCL_FCS30: a global coastline dataset with 30-m resolution and a fine classification system from 2010 to 2020

The GCL_FCS30 dataset provides a high-resolution global coastline classification system with six distinct categories, developed using advanced remote sensing techniques and machine learning. This dataset spans 2010, 2015, and 2020, offering continuous 30-meter resolution vectors that capture both natural and human-modified coastal features.

Downloaded: April 3, 2025.

### Dataset Classification Codes

The six coastline categories are numerically encoded as follows:\
- **0: Artificial coastlines** – Human-built structures like seawalls, ports, piers, and reclaimed land.\
- **1: Biogenic coastlines** – Mangroves, salt marshes, coral reefs, and other biologically formed features.\
- **2: Sandy coastlines** – Beaches, dunes, and other sand-dominated shores.\
- **3: Muddy coastlines** – Sediment-rich zones like tidal flats and mudflats.\
- **4: Rocky coastlines** – Cliffs, bedrock platforms, and other rock-dominated shores.\
- **5: Estuary coastlines** – Transitional zones where rivers meet the sea, including deltas and brackish wetlands.

### Methodology Overview

**Coastline Extraction Process**:\
1.
**MNDWI Calculation**: Enhanced water/land separation using Modified Normalized Difference Water Index ($$ \text{MNDWI} = \frac{\text{Green} - \text{SWIR1}}{\text{Green} + \text{SWIR1}} $$).\
2.
**OTSU Adaptive Thresholding**: Automated sea-land segmentation via maximum between-class variance optimization.\
3.
**Canny Edge Detection**: Precise coastline vectorization from raster data.

**Classification Approach**:\
- A **hybrid transect classifier** combining:\
- **Random Forest algorithm** trained on multi-source geophysical data (e.g., elevation, mangrove maps, land cover).\
- **Stable training samples** derived from cross-validated global datasets.\
- Validation achieved **\>85% overall accuracy** using 964 stratified reference points verified by expert consensus.

See [Zuo et al. 2025](https://www.nature.com/articles/s41597-025-04430-0#Sec1) for more details.

### Citation

Zuo, J., Zhang, L., Xiao, J. et al. GCL_FCS30: a global coastline dataset with 30-m resolution and a fine classification system from 2010 to 2020. Sci Data 12, 129 (2025). https://doi.org/10.1038/s41597-025-04430-0

```{r}
library(raster)
library(terra)
library(readr)
library(magrittr)
library(mregions2)
library(mapview)
library(sf)
library(tidyverse)
library(lwgeom)
library(here)
library(leaflet.extras2)
library(fasterize)
library(tictoc)
library(kableExtra)
library(plotly)
library(viridis)
library(readr)
library(purrr)
library(doParallel)
library(httr2) # to handle timeouts (error HTTP 524 NA)
library(data.table)
library(leaflet)
library(leaflet.extras)
library(RColorBrewer)

# ---- sources! ----
# source(here("workflow", "R", "common.R")) # file creates objects to process data
## set the mazu and neptune data_edit share based on operating system
dir_M             <- c('Windows' = '//mazu.nceas.ucsb.edu/ohi',
                       'Darwin'  = '/Volumes/ohi',    ### connect (cmd-K) to smb://mazu/ohi
                       'Linux'   = '/home/shares/ohi')[[ Sys.info()[['sysname']] ]]

# ---- set year and file path info ----
current_year <- 2025 # Update this in the future!!
version_year <- paste0("v",current_year)
data_dir_version_year <- paste0("d", current_year)

# ---- data directories ----

# raw data directory (on Mazu)
raw_data_dir <- here::here(dir_M, "OHI_GOC", "_raw_data")

# for GCL_FCS30 raw data dir
gcl_raw_dir <- here::here(dir_M, "OHI_GOC", "_raw_data", "GCL_FCS30", data_dir_version_year)

# for coastline characteristics intermediates
int_dir <- here::here("hab", "hab_coastlines", version_year, "int")

# final output dir
output_dir <- here("hab", "hab_coastlines", version_year, "output")

# spatial data for GoC
goc_spatial <- here("spatial")

# this CRS might be better for visualization, explore.
gulf_crs <- st_crs("+proj=aea +lat_1=23 +lat_2=30 +lat_0=25 +lon_0=-110 +datum=WGS84 +units=m +no_defs")
```

# Data exploration

```{r}
# read in the shapefile for 2020, examine data
gcl_raw_2020 <- st_read(here(gcl_raw_dir, "GCL2020.shp"))

mapview(gcl_raw_2020)
head(gcl_raw_2020)
unique(gcl_raw_2020$class) # yes, so 0 - 5 are present

st_crs(gcl_raw_2020) # WGS84
```
## Limit to GoC Ecoregion

```{r}
# read in ecoregion shapefiles in 
eco_hr_latlon <- st_read(here(goc_spatial, "GoC_polygon_ecoregion_latlon.shp"))
# mapview(eco_hr_latlon)
head(eco_hr_latlon)

eco_hr_eqarea <- st_read(here(goc_spatial, "GoC_polygon_ecoregion_eqarea.shp"))
mapview(eco_hr_eqarea)

# check CRS of ecoregion boundary
st_crs(eco_hr_latlon) # WGS84, can be used to crop global data

# transform to gulf_CRS
# eco_hr_albers <- st_transform(eco_hr_eqarea, crs = "+proj=aea +lat_1=23 +lat_2=30 +lat_0=25 +lon_0=-110 +datum=WGS84 +units=m +no_defs")
# st_crs(eco_hr_albers) # double check it worked well
# mapview(eco_hr_albers) # looks good!
```

```{r}
# crop the global data to the gulf ecoregion boundary
gcl_wgs84 <- st_transform(gcl_raw_2020, crs = 4326) # ensure it is wgs84
eco_hr_wgs84 <- st_transform(eco_hr_latlon, crs = 4326) # ensure it is wgs84

gcl_2020_cropped <- st_intersection(gcl_wgs84, eco_hr_latlon)
head(gcl_2020_cropped) # attributes are conserved

mapview(gcl_2020_cropped)

write_sf(gcl_2020_cropped, here::here("hab", "hab_coastlines", "v2025","GCL_FCS30_app","www", "gcl_2020_cropped.shp"))
write_sf(gcl_2020_cropped, here::here(int_dir, "gcl_2020_cropped.shp"))
```

```{r}
# crop the global data to the gulf ecoregion boundary
# gcl_wgs84 <- st_transform(gcl_raw_2020, crs = 3857) # ensure it is wgs84
eco_hr_wgs84_pm <- st_transform(eco_hr_latlon, crs = 3857) # ensure it is wgs84

gcl_2020_cropped_pm <- st_intersection(gcl_raw_2020, eco_hr_wgs84_pm)
head(gcl_2020_cropped_pm) # attributes are conserved

mapview(gcl_2020_cropped_pm)
```
# Visualize interactively

```{r}
library(leaflet)
library(sf)

# Colorblind-friendly palette (Viridis)
viridis_colors <- viridis::viridis(6)

# Match colors to class meanings for clarity
color_mapping <- c(
  "0" = viridis_colors[1],  # Artificial
  "1" = viridis_colors[2],  # Biogenic
  "2" = viridis_colors[3],  # Sandy
  "3" = viridis_colors[4],  # Muddy
  "4" = viridis_colors[5],  # Rocky
  "5" = viridis_colors[6]   # Estuary
)

# Create color palette function
pal <- colorFactor(
  palette = color_mapping,
  domain = gcl_2020_cropped$class
)

# Class descriptions for popup
class_descriptions <- c(
  "0" = "Artificial: Human-built structures",
  "1" = "Biogenic: Mangroves, marshes, reefs",
  "2" = "Sandy: Beaches, dunes",
  "3" = "Muddy: Tidal flats, mudflats",
  "4" = "Rocky: Cliffs, bedrock",
  "5" = "Estuary: River-sea transition zones"
)

# Create leaflet map with layers control
leaflet(gcl_2020_cropped) %>%
  addTiles() %>%
  setView(lng = -112, lat = 28, zoom = 6) %>%  # Or use fitBounds()
  addPolylines(
    color = ~pal(class),
    weight = 6,
    opacity = 0.7,
    popup = ~paste(
      "<b>Class:</b>", class, "<br>",
      "<b>Description:</b>", class_descriptions[as.character(class)] # Map values to descriptions
    ),
    layerId = ~class  # Assign unique layer IDs based on class
  ) %>%
  addLegend(
    pal = pal,
    values = ~class,
    title = "Coastline Type",
    opacity = 1,
    labFormat = labelFormat(
      transform = function(value) {
        class_descriptions[as.character(value)] # display description
      }
    )
  ) %>%
  addLayersControl(
    overlayGroups = unique(gcl_2020_cropped$class), # Create layer groups for each class
    options = layersControlOptions(collapsed = FALSE)  # Keep the layer control expanded
  )

```



