---
title: "PANGAS_biomass data"
author: "Erica Ferrer"
date: "`r Sys.Date()`"
output: html_document
---


# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
library(pangaear)

# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
library(sf) # for work on shape fines

# Visualization
library(ggplot2)
library(ggforce)
library(ggpubr) # Plotting lm statistics on ggplot.
library(colorspace)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/_habitats/")
```

# PANGAS DATA ------------------------------


```{r Load in PANGAS_biomass data}
PANGAS_biomass <- pg_data("10.1594/PANGAEA.963907")

PANGAS_biomass_Data0 <- PANGAS_biomass[[1]]$data
```

```{r Clean PANGAS_biomass}
PANGAS_biomass <- PANGAS_biomass_Data0 %>%
  rename(
    event = Event,
    id = ID,
    area = Area,
    latitude = Latitude,
    longitude = Longitude,
    activity = Activity,
    zone = Zone,
    institution = Institution,
    year_initial = `Year obs [a AD] (Initial year)`,
    year_final = `Year obs [a AD] (Final year)`,
    years = `Years [a]`,
    family = Family,
    species = `Species UID`,
    mean_biomass = `Biom [g/m**2] (Estimated based on the abunda...)`,
    sd_biomass = `Std dev [±] (Of mean biomass values, Calcu...)`,
    samples = `Samples [#]`,
    slope = `Slope (Calculated)`,
    p_value = `Prob (Calculated)`,
    slope_se = `Prob std e [±] (Calculated)`,
    biomass_min = `Biom [g/m**2] (Minimum of biomass. Estimated...)`,
    biomass_max = `Biom [g/m**2] (Maximum of biomass. Estimated...)`
  ) %>%
  mutate(
    year_mid = (year_initial + year_final) / 2
  )

```

```{r Load in PANGAS_density data}
PANGAS_density <- pg_data("10.1594/PANGAEA.963910")

PANGAS_density_Data0 <- PANGAS_density[[1]]$data
```

```{r Clean PANGAS_density}
colnames(PANGAS_density_Data0)

PANGAS_density <- PANGAS_density_Data0 %>%
  rename(
    event = Event,
    id = ID,
    area = Area,
    latitude = Latitude,
    longitude = Longitude,
    activity = Activity,
    zone = Zone,
    institution = Institution,
    year_initial = `Year obs [a AD] (Initial year)`,
    year_final = `Year obs [a AD] (Final year)`,
    years = `Years [a]`,
    family = Family,
    species = `Species UID`,
    mean_density = `Ind a [#/m**2] (Mean density. Estimated based...)`,
    sd_density = `Std dev [±] (Of mean density values, Calcu...)`,
    samples = `Samples [#]`,
    slope = `Slope (Calculated)`,
    p_value = `Prob (Calculated)`,
    slope_se = `Prob std e [±] (Calculated)`,
    density_min = `Ind a [#/m**2] (Minimum of density. Estimated...)`,
    density_max = `Ind a [#/m**2] (Maximum of density. Estimated...)`
  ) %>%
  mutate(
    year_mid = (year_initial + year_final) / 2
  )

```

# Add OHI region details to the PANGAS density data

```{r Loading in OHI regions and buffer shape files}
# OHI regions:
OHI_regions_metadata <- readLines("/home/shares/ohi/OHI_GOC/spatial/OHI_regions/marine/ohi_goc_final_071025.prj")
# CRS: EPSG:4326 in practice, poorly labeled in meta-data.

OHI_regions_shape0 <- sf::st_read(
  "/home/shares/ohi/OHI_GOC/spatial/OHI_regions/marine/ohi_goc_final_071025.shp",
  options = "ENCODING=LATIN1"
)

st_crs(OHI_regions_shape0)

OHI_regions_shape0 <- st_make_valid(OHI_regions_shape0)

# Quick plot:
OHI_Region_plot<- ggplot(OHI_regions_shape0) +
  geom_sf(aes(fill = rgn_id)) +
  theme_bw()

OHI_Region_plot

# 50km buffer:
OHI_buffer_metadata <- readLines("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer/inland_50km_goc.prj")
# Projected CRS name: Mexico_ITRF2008_LCC

OHI_buffer_shape0 <- sf::st_read(
  "/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer/inland_50km_goc.shp",
  options = "ENCODING=LATIN1"
)

st_crs(OHI_buffer_shape0)

OHI_buffer_shape0 <- st_make_valid(OHI_buffer_shape0)

# Quick plot:
OHI_Buffer_plot<- ggplot(OHI_buffer_shape0) +
  geom_sf(aes(fill = rgn_id)) +
  theme_bw()

OHI_Buffer_plot
```

```{r Some quick cleaning}
# Let's use CRS 4326 for all projections now.
crs_use = 4326

# Change CRS on OHI_buffer_shape to match CRS on OHI_regions_shape
OHI_regions_shape <- st_transform(OHI_regions_shape0, crs_use)

# Change CRS on OHI_buffer_shape to match CRS on OHI_regions_shape
OHI_buffer_shape <- st_transform(OHI_buffer_shape0, crs_use)

# Remove columns and extract useful info:
OHI_buffer_shape<- OHI_buffer_shape0 %>%
  select(-c(estado))

additional_info<- OHI_buffer_shape0 %>%
  select(rgn_id, ecorgn) %>%
  st_drop_geometry()

# Add new columns to OHI_regions_shape
OHI_regions_shape <- OHI_regions_shape %>%
  left_join(
    additional_info,
    by = "rgn_id"
  ) %>%
  mutate(
    location= c("sea")
  ) %>%
  relocate(location, .after= rgn_id)

rm(additional_info)
```

```{r Join OHI geometric key to PANGAS_density}
# Step 0 — define a Coordinate Reference System (CRS) once and stick to it:
crs_use <- 4326
# Alternative could be: CRS 5070 is the USA Contiguous Albers Equal Area (good proxy for northern Mexico too). EPSG:5070 is centered on the U.S., so Baja / Sonora are near the edge of its optimal zone.

# Step 1 — Convert master table to spatial points
# Turn PANGAS_density into an sf object
colnames(PANGAS_density)

master_sf <- PANGAS_density %>%
  filter(!is.na(longitude), !is.na(latitude)) %>%
  st_as_sf(coords = c("longitude", "latitude"), 
           crs = crs_use,
           remove = FALSE) %>%
  st_transform(crs_use)

# Step 2 — Transform polygons ONCE
buffer_sf <- OHI_buffer_shape %>%
  st_transform(crs_use)

sea_sf <- OHI_regions_shape %>%
  st_transform(crs_use)

# Check CRS:
st_crs(master_sf)
st_crs(buffer_sf)
st_crs(sea_sf)

# Check spatial extents
st_bbox(master_sf)
st_bbox(buffer_sf)
st_bbox(sea_sf)

# Check number of polygon matches for each PANGAS point:
st_intersects(master_sf, buffer_sf)
st_intersects(master_sf, sea_sf)
# Only intersects with the sea boundaries and none of the land boundaries, which makes sense.

# Step 3 — Spatial join to land buffer polygons
# Join points → OHI_buffer_shape
master_land <- st_join(
  master_sf,
  buffer_sf,
  join = st_within,
  left = TRUE
)

# Re-name columns to be safe:
master_land <- master_land %>%
  rename(
    rgn_id_land = rgn_id,
    ecorgn_land = ecorgn,
    location_land = location
  )

# Step 5 — Spatial join to sea regions
# Join same (or filtered) points → OHI_regions_shape
master_all <- st_join(
  master_land,
  sea_sf,
  join = st_within,
  left = TRUE
)

# Step 6 — Remove duplicate column names cleanly
# master_all <- master_all %>%
#   rename(
#     rgn_id_sea = rgn_id,
#     ecorgn_sea = ecorgn,
#     location_sea = location
#   )

# Clean
colnames(master_all)

PANGAS_density<- master_all %>%
  select(area, zone, latitude, longitude,
         rgn_id, ecorgn,
         activity, institution, samples,
         year_initial, year_final, year_mid, years,
         family, species,
         density_min, density_max, sd_density,
         mean_density, slope,
         p_value)
```

## QC: Does sampling effort impact mean density estimates? No, at least not at this level.

```{r Does sampling effort impact mean density estimates?}

plot_df <- PANGAS_density |>
  distinct(species, samples, mean_density) |>
  filter(!is.na(samples), !is.na(mean_density))

# Calculate p-value for each species
pvals <- plot_df |>
  group_by(species) |>
  summarise(
    p_value = summary(lm(mean_density ~ samples))$coefficients[
      "samples", "Pr(>|t|)"
    ],
    .groups = "drop"
  ) |>
  mutate(
    label = paste0(
      "p = ",
      format.pval(p_value, digits = 3, eps = 0.001)
    )
  )

# Plot
ggplot(plot_df, aes(x = samples, y = mean_density)) +
  geom_point(size = 2.5, alpha = 0.7) +
  
  geom_smooth(
    method = "lm",
    se = TRUE,
    color = "steelblue"
  ) +
  
  geom_text(
    data = pvals,
    aes(
      x = Inf,
      y = Inf,
      label = label
    ),
    hjust = 1.1,
    vjust = 1.5,
    inherit.aes = FALSE
  ) +
  
  facet_wrap(~ species, scales = "free") +
  
  labs(
    x = "Number of samples",
    y = "Mean density",
    title = "Sampling effort vs. mean density"
  ) +
  
  theme_bw()

```

## Visualize density and biomass data:

```{r Simple slope plots}
plot(PANGAS_biomass$slope)

plot(PANGAS_density$slope)
```

```{r Summary stats}
summary(PANGAS_biomass$biomass, na.rm= TRUE)
summary(PANGAS_biomass$slope, na.rm= TRUE) # Slope seems to be generally increasing
summary(PANGAS_biomass$slope_se)

```

```{r Plot PANGAS biomass slopes per region}
ggplot(
  PANGAS_biomass %>% filter(!is.na(slope)),
  aes(x = zone, y = slope)
) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = "grey40"
  ) +
  geom_jitter(
    width = 0.15,
    alpha = 0.3
  ) +
  geom_boxplot(
    width = 0.4,
    alpha = 0.5,
    outlier.shape = NA
  ) +
  labs(
    x = "Gulf region",
    y = "Estimated annual change in abundance"
  ) +
  theme_bw()

```

```{r Plot PANGAS density slopes per region}
ggplot(
  PANGAS_density %>% filter(!is.na(slope)),
  aes(x = family, y = slope)
) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = "grey40"
  ) +
  geom_jitter(
    width = 0.15,
    alpha = 0.25
  ) +
  geom_boxplot(
    width = 0.5,
    alpha = 0.5,
    outlier.shape = NA
  ) +
  facet_wrap(~ zone) +
  labs(
    x = "Family",
    y = "Estimated annual change in density"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

## Slope means:
```{r Summarize biomass slopes by species and zone}
colnames(PANGAS_biomass)

species_zone_biomass_slope <- PANGAS_biomass %>%
  group_by(zone, species) %>%
  summarise(
    mean_slope = mean(slope, na.rm = TRUE),
    .groups = "drop"
  )

species_zone_density_slope <- PANGAS_density %>%
  group_by(zone, species) %>%
  summarise(
    mean_slope = mean(slope, na.rm = TRUE),
    .groups = "drop"
  )
```


```{r Plot mean biomass per species / zone}
ggplot(
  species_zone_biomass_slope,
  aes(x = species, y = mean_slope, fill = species)
) +
  geom_col() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~ zone) +
  labs(
    x = "Species",
    y = "Mean biomass slope (g/m²/year)",
    fill = "Species"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )


ggplot(
  species_zone_density_slope,
  aes(x = species, y = mean_slope, fill = species)
) +
  geom_col() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~ zone) +
  labs(
    x = "Species",
    y = "Mean density slope (g/m²/year)",
    fill = "Species"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )

```

```{r What if we filter by years available?}
unique(PANGAS_biomass$year_final)
unique(PANGAS_density$year_final)
# 2010 2016 2017 2020 2019 2011 2018


unique(PANGAS_biomass$year_mid)
# 2008.5 2013.0 2012.0 2015.0 2011.5 2014.5 2009.0 2012.5 2013.5 2015.5 2014.0 2016.5

is.numeric(PANGAS_biomass$year_mid)
```

```{r Summarize slope by species and zone but filter for more recent data sets}
species_zone_biomass_slope <- PANGAS_biomass %>%
  filter(year_final>= 2015) %>%
  group_by(zone, species) %>%
  summarise(
    mean_slope = mean(slope, na.rm = TRUE),
    .groups = "drop"
  )

species_zone_density_slope <- PANGAS_density %>%
  filter(year_final>= 2015) %>%
  group_by(zone, species) %>%
  summarise(
    mean_slope = mean(slope, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r Simple plots}
plot(species_zone_biomass_slope$mean_slope)

plot(species_zone_density_slope$mean_slope)
```

## Slopes pre and post 2015:
```{r Pre and post 2015}

dat <- PANGAS_biomass %>%
  mutate(
    assessment_period = case_when(
      year_mid < 2015 ~ "Pre-2015",
      year_mid >= 2015 ~ "Post-2015",
      TRUE ~ NA_character_
    ),
    assessment_period = factor(
      assessment_period,
      levels = c("Pre-2015", "Post-2015")
    )
  )

table(dat$assessment_period)
 # Pre-2015 Post-2015 
 #      325       117

table(dat$assessment_period, dat$zone)
table(dat$assessment_period, dat$family)

```


```{r Visualization pre and post 2015}

ggplot(
  dat %>% filter(!is.na(slope), !is.na(assessment_period)),
  aes(x = assessment_period, y = slope)
) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_jitter(
    width = 0.15,
    alpha = 0.35,
    color = "grey30"
  ) +
  geom_boxplot(
    width = 0.4,
    alpha = 0.5,
    outlier.shape = NA
  ) +
  labs(
    x = NULL,
    y = "Estimated biomass slope (g/m²/year)"
  ) +
  theme_bw()

```

```{r Model pre and post 2015}
dat_test <- dat %>%
  filter(
    !is.na(slope),
    !is.na(assessment_period)
  )

mod <- lm(
  slope ~ assessment_period + family + zone,
  data = dat_test
)

summary(mod)
#                            Estimate Std. Error t value  Pr(>|t|)    
# (Intercept)                 2.64525    0.59201   4.468 0.0000101 ***
# assessment_periodPost-2015 -0.32220    0.60092  -0.536   0.59211    
# familyGirellidae            0.90343    1.05669   0.855   0.39304    
# familyKyphosidae            0.08168    1.05669   0.077   0.93842    
# familyPomacentridae        -1.32344    0.72186  -1.833   0.06743 .  
# familyScaridae             -0.40456    0.66831  -0.605   0.54527    
# zoneNorth                  -0.93533    0.61354  -1.524   0.12812    
# zoneSouth                  -1.92226    0.67104  -2.865   0.00438 ** 

# No compelling evidence of an effect of period...

```


```{r Is there evidence density has declined pre and post 2015?}
library(dplyr)
library(ggplot2)

slope_summary <- PANGAS_density %>%
  filter(!is.na(slope)) %>%
  summarise(
    n = n(),
    mean_slope = mean(slope),
    median_slope = median(slope),
    sd_slope = sd(slope),
    n_negative = sum(slope < 0),
    n_positive = sum(slope > 0),
    n_zero = sum(slope == 0)
  )

slope_summary


ggplot(
  PANGAS_density %>% filter(!is.na(slope)),
  aes(x = slope)
) +
  geom_histogram(
    bins = 40,
    fill = "steelblue",
    color = "white"
  ) +
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    linewidth = 1
  ) +
  labs(
    x = "Estimated annual change in density",
    y = "Number of site × species observations"
  ) +
  theme_bw()

```




## Revised approach (Sep 29 2026): testing two scoring options

```{r Estimate annual density based on estimated mean_density and slope}
# Use the mean_density and slope parameters to estimate annual density. Bound by density_max and density_min.

colnames(PANGAS_density)

PANGAS_density_yearly0 <- PANGAS_density %>%
  rowwise() %>%
  mutate(
    year = list(seq(year_initial, year_final))
  ) %>%
  tidyr::unnest(year)

# 4719 obs., 23 vars.
```

```{r filter out p is NA}
colnames(PANGAS_density_yearly0)

PANGAS_density_yearly <- PANGAS_density_yearly0 %>%
  filter(!is.na(p_value)) %>%
  mutate(
    slope_classification = case_when(
      slope < 0 & p_value < 0.05  ~ "significant negative slope",
      slope < 0 & p_value >= 0.05 ~ "non-significant negative slope",
      slope >= 0 & p_value < 0.05 ~ "significant positive slope",
      slope >= 0 & p_value >= 0.05 ~ "non-significant positive slope"
    )
  ) %>%
  select(area, zone, latitude, longitude,
         rgn_id, ecorgn,
         activity, institution, samples,
         year_initial, year_final, year_mid,
         family, species,
         density_min, density_max, sd_density,
         year, mean_density, slope,
         p_value, slope_classification)

table(PANGAS_density_yearly$slope_classification)

# non-significant negative slope non-significant positive slope     significant negative slope     significant positive slope 
#                            797                           1630                            265                           1033 
```

# Option 1: proportion of statistically negative slopes
```{r Summarize slopes per year zone species}
colnames(PANGAS_density_yearly)

PANGAS_scores_option1 <- PANGAS_density_yearly %>%
  select(rgn_id, species, year, samples, institution, slope_classification) %>%
  group_by(rgn_id, species, year) %>%
  summarise(
    n_samples = sum(samples),
    n_datasets = n_distinct(institution),
    prop_sig_negative =
      mean(slope_classification == "significant negative slope"),
    prop_nonsig_negative =
      mean(slope_classification == "non-significant negative slope"),
    prop_sig_positive =
      mean(slope_classification == "significant positive slope"),
    prop_nonsig_positive =
      mean(slope_classification == "non-significant positive slope"),
    .groups = "drop"
  )

PANGAS_scores_option1<- PANGAS_scores_option1 %>%
  mutate(
    score_option1 = (1-prop_sig_negative)*100
  )
```

```{r Simple plot}
ggplot(PANGAS_scores_option1, 
       aes(x = year, y = prop_sig_negative, color = species)) +
  
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5) +
  
  facet_wrap(~ rgn_id) +
  
  scale_y_continuous(
    limits = c(0, 1),
    labels = scales::percent
  ) +

  labs(
    x = "Year",
    y = "Proportion of significant negative slopes",
    color = "Species"
  ) +
  
  scale_color_viridis_d()+
  theme_bw()
```

```{r Summarize by rgn_id and year}
colnames(PANGAS_scores_option1)

PANGAS_scores_option11<- PANGAS_scores_option1 %>%
  group_by(rgn_id, year) %>%
  summarize(
    score_option1= mean(score_option1)
  )
```

```{r Simple plot of Option 1}
ggplot(PANGAS_scores_option11, 
       aes(x = year, y = score_option1, color = rgn_id)) +
  
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5) +
  
  facet_wrap(~ rgn_id) +
  
  scale_y_continuous(
    limits = c(0, 100),
  ) +
  labs(
    title = "PANGAS Rocky Reef Data - Scoring Option 1",
    caption = "Option 1 represents one minus the proportion of significant negative slopes (p < 0.05) \nmeaned across species & data sets at the annual x zone level",
    x = "Year",
    y = "Rocky Reef Fish Scores [Option 1]",
    color = "Zone"
  ) +
#  scale_color_viridis_d()+
  theme_bw()
```

## Option 2: score equals estimated fish density (that year at that site) relative to mean density
```{r PANGAS_scores_option2}
colnames(PANGAS_density_yearly)

# Estimate density per species per year per survey:
PANGAS_scores_option2 <-PANGAS_density_yearly %>%
  mutate(
    estimated_density_raw = mean_density +
      slope * (year - year_mid),

    estimated_density = pmax(0, estimated_density_raw),

    density_was_bounded = estimated_density_raw < 0
  )

# For slopes that are not significant, estimated_density should equal mean_density:
PANGAS_scores_option2<- PANGAS_scores_option2 %>%
  mutate(
    estimated_density = case_when(
      slope_classification == "non-significant negative slope" ~ mean_density,
      slope_classification == "non-significant positive slope" ~ mean_density,
      slope_classification == "significant positive slope" ~ estimated_density,
      slope_classification == "significant negative slope" ~ estimated_density,
    )
  )
  

colnames(PANGAS_scores_option2)

PANGAS_scores_option22<- PANGAS_scores_option2 %>%
  select(year, rgn_id, species, samples, institution, mean_density, estimated_density) %>%
  group_by(rgn_id, species, year) %>%
  summarise(
    n_samples = sum(samples),
    n_datasets = n_distinct(institution),
    mean_density = mean(mean_density, na.rm= TRUE),
    mean_est_annual_density= mean(estimated_density, na.rm= TRUE)
  )
```

```{r How does sampling effort impact mean density estimates at this level?}
plot_df <- PANGAS_scores_option22 |>
  distinct(species, n_samples, mean_density) |>
  filter(!is.na(n_samples), !is.na(mean_density))

# Calculate p-value for each species
pvals <- plot_df |>
  group_by(species) |>
  summarise(
    p_value = summary(lm(mean_density ~ n_samples))$coefficients[
      "n_samples", "Pr(>|t|)"
    ],
    .groups = "drop"
  ) |>
  mutate(
    label = paste0(
      "p = ",
      format.pval(p_value, digits = 3, eps = 0.001)
    )
  )

# Plot
ggplot(plot_df, aes(x = n_samples, y = mean_density)) +
  geom_point(size = 2.5, alpha = 0.7) +
  
  geom_smooth(
    method = "lm",
    se = TRUE,
    color = "steelblue"
  ) +
  
  geom_text(
    data = pvals,
    aes(
      x = Inf,
      y = Inf,
      label = label
    ),
    hjust = 1.1,
    vjust = 1.5,
    inherit.aes = FALSE
  ) +
  
  facet_wrap(~ species, scales = "free") +
  
  labs(
    x = "Number of samples",
    y = "Mean density",
    title = "Sampling effort vs. mean density"
  ) +
  
  theme_bw()

# At this level, the number of samples does bias the mean density outcome.
```

This exercise leads me to believe that we should not use option 2 since it is biased by sample size.

```{r Summarize by rgn_id and year}
colnames(PANGAS_scores_option2)

TEST<- PANGAS_scores_option22 %>%
  group_by(rgn_id, year) %>%
  summarize(
    mean_density= mean(mean_density, na.rm= TRUE)
  )
```

```{r Simple plot mean densities by rgn_id}
ggplot(TEST, 
       aes(x = year, y = mean_density, color = rgn_id)) +
  
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5) +
  
  facet_wrap(~ rgn_id) +

  labs(
    title = "PANGAS Rocky Reef Data: mean fish densities per zone",
    caption = "These mean density estimates are biased by the postive relationship between mean density and # samples.",
    x = "Year",
    y = "Mean fish density",
    color = "OHI Region"
  ) +
  theme_bw()
```





# COBI DATA ------------------------------

## COBI_Invert_Data0

```{r Look into COBI inverts data}
COBI_Invert_Data0<- read.csv("/home/shares/ohi/OHI_GOC/_raw_data/COBI/COBI_Invertebrados_Reserva_Biosfera_IslaSanPedroMartir_2007_2022.csv")
```

```{r Clean df}
colnames(COBI_Invert_Data0)

COBI_Invert_Data<- COBI_Invert_Data0 %>%
  select(
    Site = SITIO,
    Day = DIA,
    Month = MES,
    Year = ANIO,
    State = ESTADO,
    Lat = LATITUD,
    Lon = LONGITUD,
    Depth_start = PROFUNDIDAD_INICIAL,
    Depth_end = PROFUNDIDAD_FINAL,
    Species = ESPECIE,
    Abudance = ABUNDANCIA
  )
```

```{r Summarize}
COBI_Invert_Data_summarized<- COBI_Invert_Data %>%
  select(Year, Species, Abudance) %>%
  group_by(Year, Species) %>%
  summarize(
    mean_Abudance = mean(Abudance, na.rm = TRUE)
  )
```

```{r Plot summary data}
ggplot(
  COBI_Invert_Data_summarized,
  aes(x = Year, y = mean_Abudance)
) +
  geom_point(alpha = 0.5) +
  geom_smooth(
    method = "lm",
    se = TRUE
  ) +
  facet_wrap(~ Species, scales = "free_y") +
  geom_vline(
    xintercept = 2015,
    linetype = "dashed"
  ) +
  labs(
    x = "Year",
    y = "Mean abundance"
  ) +
  theme_bw()

```

```{r Enhance this plot}
library(dplyr)
library(ggplot2)

# Calculate LM p-value separately for each species
species_lm <- COBI_Invert_Data_summarized %>%
  group_by(Species) %>%
  summarise(
    n_years = n_distinct(Year[!is.na(Year) & !is.na(mean_Abudance)]),
    slope = if (n_years >= 2) {
      coef(lm(mean_Abudance ~ Year))["Year"]
    } else {
      NA_real_
    },
    p_value = if (n_years >= 2) {
      summary(lm(mean_Abudance ~ Year))$coefficients["Year", "Pr(>|t|)"]
    } else {
      NA_real_
    },
    .groups = "drop"
  ) %>%
  mutate(
    trend = case_when(
      is.na(p_value) ~ "Insufficient data",
      p_value < 0.10 & slope < 0 ~ "Significant decline",
      p_value < 0.10 & slope > 0 ~ "Significant increase",
      TRUE ~ "No significant trend"
    )
  )


# Add significance information back to the data
COBI_plot <- COBI_Invert_Data_summarized %>%
  left_join(species_lm, by = "Species")

# Plot
ggplot(
  COBI_plot,
  aes(x = Year, y = mean_Abudance)
) +
  geom_point(
    aes(color = trend),
    alpha = 0.5
  ) +
  geom_smooth(
    aes(color = trend),
    method = "lm",
    se = TRUE
  ) +
  facet_wrap(~ Species, scales = "free_y") +
  geom_vline(
    xintercept = 2015,
    linetype = "dashed"
  ) +
scale_color_manual(
  values = c(
    "Significant decline" = "#D73027",
    "Significant increase" = "#1A9850",
    "No significant trend" = "grey70",
    "Insufficient data" = "grey90"
  )
) + 

  labs(
    x = "Year",
    y = "Mean abundance"
  ) +
  theme_bw()

```


## COBI_Fish_Data0
```{r Look into COBI inverts data}
COBI_Fish_Data0<- read.csv("/home/shares/ohi/OHI_GOC/_raw_data/COBI/COBI_Peces_Reserva_Biosfera_IslaSanPedroMartir_2007_2022.csv")
```

```{r Clean df}
colnames(COBI_Fish_Data0)

COBI_Fish_Data<- COBI_Fish_Data0 %>%
  select(
    Site = SITIO,
    Day = DIA,
    Month = MES,
    Year = ANIO,
    State = ESTADO,
    Lat = LATITUD,
    Lon = LONGITUD,
    Depth_start = PROFUNDIDAD_INICIAL,
    Depth_end = PROFUNDIDAD_FINDL,
    Species = ESPECIE,
    Abudance = ABUNDANCIA
  )
```

```{r Summarize}
COBI_Fish_Data_summarized<- COBI_Fish_Data %>%
  select(Year, Species, Abudance) %>%
  group_by(Year, Species) %>%
  summarize(
    mean_Abudance = mean(Abudance, na.rm = TRUE)
  )
```

```{r Enhance this plot}
library(dplyr)
library(ggplot2)

# Calculate LM p-value separately for each species
species_lm <- COBI_Fish_Data_summarized %>%
  group_by(Species) %>%
  summarise(
    n_years = n_distinct(Year[!is.na(Year) & !is.na(mean_Abudance)]),
    slope = if (n_years >= 2) {
      coef(lm(mean_Abudance ~ Year))["Year"]
    } else {
      NA_real_
    },
    p_value = if (n_years >= 2) {
      summary(lm(mean_Abudance ~ Year))$coefficients["Year", "Pr(>|t|)"]
    } else {
      NA_real_
    },
    .groups = "drop"
  ) %>%
  mutate(
    trend = case_when(
      is.na(p_value) ~ "Insufficient data",
      p_value < 0.10 & slope < 0 ~ "Significant decline",
      p_value < 0.10 & slope > 0 ~ "Significant increase",
      TRUE ~ "No significant trend"
    )
  )


# Add significance information back to the data
COBI_plot_fish <- COBI_Fish_Data_summarized %>%
  left_join(species_lm, by = "Species")

# Plot
ggplot(
  COBI_plot_fish,
  aes(x = Year, y = mean_Abudance)
) +
  geom_point(
    aes(color = trend),
    alpha = 0.5
  ) +
  geom_smooth(
    aes(color = trend),
    method = "lm",
    se = TRUE
  ) +
  facet_wrap(~ Species, scales = "free_y") +
  geom_vline(
    xintercept = 2015,
    linetype = "dashed"
  ) +
scale_color_manual(
  values = c(
    "Significant decline" = "#D73027",
    "Significant increase" = "#1A9850",
    "No significant trend" = "grey70",
    "Insufficient data" = "grey90"
  )
) + 

  labs(
    x = "Year",
    y = "Mean abundance"
  ) +
  theme_bw()

```

