################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```

# PENDING - Must merge with new techniques (May 28 2026)
# SCORE A: CPUE Estimates (per major marine group) ------------------------------
```{r Load data as necessary}
# Load file, only if needed:
GoC_Landings_for_CPUE_original<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/fis/v2025/int/GOC_FIS.landings_CLEAN_as.of.Dec092025.rds"))
GoC_Landings_for_CPUE<- GoC_Landings_for_CPUE_original
# GoC_Landings_for_CPUE<- Landings_GoC_NoFresh.NoMariculture
# 3163120 obs., 50 vars.
```

```{r TEMPORARY - Graph of Top Fisheries Per Region}
GoC_FIS.Landings_CLEAN<- GoC_Landings_for_CPUE_original

colnames(GoC_FIS.Landings_CLEAN)

top_fisheries <- GoC_FIS.Landings_CLEAN %>%
  group_by(OHI_Region_No, estimated_genus_name) %>%
  summarise(
    total_weight = sum(weight_kg_corrected, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(OHI_Region_No) %>%
  slice_max(total_weight, n = 10, with_ties = FALSE)


library(dplyr)
library(ggplot2)
library(tidytext)

regions_to_drop <- c(
  "Region_4_5_6_Sonora_TBD",
  "Region_3_4_BCS_TBD",
  "Region_1_2_Baja_TBD"
)

TOPFisheries_Jan202026<- ggplot(
  top_fisheries %>%
    filter(!OHI_Region_No %in% regions_to_drop),
  aes(
    x = reorder_within(estimated_genus_name, total_weight, OHI_Region_No),
    y = total_weight,
    fill = OHI_Region_No
  )
) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~ OHI_Region_No, scales = "free_y") +
  scale_x_reordered() +
  scale_y_log10(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    x = "Fishery",
    y = "Total landed weight (kg, log scale)",
    title = "Top 10 fisheries by landed weight (per region)"
  ) +
  theme_minimal()


  # Save as PNG
  ggsave(
    filename = sprintf("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/TOPFisheries_Jan202026.png"),
    plot = TOPFisheries_Jan202026,
    width = 12,
    height = 12,
    dpi = 300
  )
  
top_Artisanal_fisheries <- GoC_FIS.Landings_CLEAN %>%
  filter(year >= 2005) %>%
  filter(fleet=="small_scale") %>%
  group_by(OHI_Region_No, estimated_genus_name) %>%
  summarise(
    total_weight = sum(weight_kg_corrected, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(OHI_Region_No) %>%
  slice_max(total_weight, n = 10, with_ties = FALSE)
  
TOP_SSF_Plot_Genus <- ggplot(
  top_Artisanal_fisheries %>%
    filter(!OHI_Region_No %in% regions_to_drop),
  aes(
    x = reorder_within(estimated_genus_name, total_weight, OHI_Region_No),
    y = log10(total_weight),
    fill = OHI_Region_No
  )
) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~ OHI_Region_No, scales = "free_y") +
  scale_x_reordered() +
  scale_fill_viridis_d(option = "C", end = 0.9) +
  labs(
    x = "Fishery",
    y = "Total landed weight (kg, log scale)",
    title = "Top 10 fisheries by landed weight (per region) BY GENUS"
  ) +
  theme_bw() +
  theme(
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    axis.text.x  = element_text(size = 14),
    axis.text.y  = element_text(size = 14),
    plot.title   = element_text(size = 22, face = "bold", hjust = 0.5)
  )

TOP_SSF_Plot_Genus
  
# Save as PNG
  ggsave(
    filename = sprintf("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/TOPSSF_Genus_Jan212026.png"),
    plot = TOP_SSF_Plot_Genus,
    width = 12,
    height = 12,
    dpi = 300
  )

```

```{r}
subset<- GoC_FIS.Landings_CLEAN"PESQUERA PUNTA PEÑASCO SA DE CV"
```


```{r CPUE calculations}
# Calculate the difference in days between period start and end:
GoC_Landings_for_CPUE$days_fishing <- as.numeric(GoC_Landings_for_CPUE$period_end_fixed - GoC_Landings_for_CPUE$period_start_fixed) + 1
# 3163120 obs., 51 vars.

# Estimate CPUE per trip:
GoC_Landings_for_CPUE<- GoC_Landings_for_CPUE %>%
  mutate(CPUE_kg.day = weight_kg_corrected / days_fishing)
# 3163120 obs., 52 vars.

# General filtering:
# In general, the receipts from early trips are a bit funky. Let's filter out trips before 2003 (n= 267072 records).
GoC_Landings_for_CPUE<- GoC_Landings_for_CPUE %>%
  filter(year >= 2003)
# 2896048 obs., 52 vars.

# Let's also filter out any records that did not have a receipt start and end date, and records where live weight was < or = 0 (n= 126955 records).
GoC_Landings_for_CPUE<- GoC_Landings_for_CPUE %>%
  filter(!is.na(days_fishing) & days_fishing > 0 & weight_kg_corrected > 0)
# 2769093 obs., 52 vars.

max(GoC_Landings_for_CPUE$CPUE_kg.day) # 1400000 (Oysters in Baja, no landing site), followed by tuna in Sinaloa
min(GoC_Landings_for_CPUE$CPUE_kg.day) #0.00125

subset<- GoC_Landings_for_CPUE %>% filter(CPUE_kg.day >= 288000)
unique(subset$estimated_genus_name)

##### NOTE TO SELF:
# - There is some funky stuff going on with SOME of the end and start dates. Come back to this in the new year.
# - I am not sure what to do about the days with no record of live weight. Exclude these for now.
```

```{r Exploratory: Plot CPUE per trip}
# PLOT CPUE PER TRIP:

CPUE_PerTrip_Distribution<- GoC_Landings_for_CPUE %>% 
  filter(year >= 2003, year <= 2013) %>%
 
  ggplot(aes(x = seq_along(CPUE_kg.day), y = CPUE_kg.day)) +
  geom_point(color = "steelblue", alpha = 0.6, size = 2) +
  geom_hline(yintercept = mean(GoC_Landings_for_CPUE$CPUE_kg.day, na.rm = TRUE),
             color = "red", linetype = "dashed", linewidth = 1) +
  labs(
    x = "# of observations",
    y = "CPUE (kg/day)",
    title = "CPUE Distribution"
  ) +
  
  theme_minimal(base_size = 14)

CPUE_PerTrip_Distribution

  # Save as PNG
  ggsave(
    filename = sprintf("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_CPUE_Scores/Distribution_of_CPUEperTrip_2003.2013_Dec112025.png"),
    plot = CPUE_PerTrip_Distribution,
    width = 12,
    height = 12,
    dpi = 300
  )
```

```{r Let's explore the landings for which IHH_major_group equals NA}
# TEST<- GoC_Landings_for_CPUE %>% filter(is.na(IHH_functional_major_groups_sm))
# # 180324 out of 2787363 total records contain no IHH_functional_major_groups_sm (this equate to about ~6-7% of records).
# 
# TEST2<- TEST %>% 
#   select(recorded_taxa_group, recorded_taxa_name_clean, common_name_clean, keywords_1:keywords_5) %>% 
#   distinct()
# 
# # write.csv(TEST2, "/home/ferrer/ohi_goc_prep/general/Species_Keys/NoIHHClass_SpKey_UNDERCONSTRUCTION_Dec022025.csv", row.names = FALSE)
# 
# rm(TEST, TEST2)
```
NOTE TO SELF: Interesting. Most things could probably be grouped to the IHH group level with a little bit of extra leg work... COME BACK TO THIS IN THE NEW YEAR.

```{r Re-assigning CPUE major groups based on new key}
Genus_MajorGroup_Key<- read.csv("~/ohi_goc_prep/general/CONAPESCA_Keys/Genus_to_MajorGroup_Key_Dec112025.csv")
# 326 obs., 4 vars.

GoC_Landings_for_CPUE <- GoC_Landings_for_CPUE %>%
  left_join(Genus_MajorGroup_Key,
            by = c("recorded_taxa_group", "estimated_genus_name"))
# 2769093 obs., 54 vars. as expected!
```

```{r Estimate CPUE for each Major Functional Group per region per year versus "historic" baseline}
# Turn off scientific notation:
options(scipen = 999)

# STEP 1: ESTABLISH CPUE BASELINE VALUES, 2003-2013 -----
# Where "baseline" equals the average Group-level CPUE from 2003-2013
CPUE_baseline <- GoC_Landings_for_CPUE %>%
  filter(year >= 2003, year <= 2013) %>%
  filter(!is.na(OHI_Region_No)) %>%
  group_by(OHI_Region_No, Major_Functional_Group) %>%
  summarise(
    OLD_CPUE_2003.2013 = mean(CPUE_kg.day, na.rm = TRUE),
    NEW_CPUE_2003.2013 = sum(weight_kg_corrected) / sum(days_fishing),
    .groups = "drop"
  )

# CPUE_baseline <- CPUE_baseline %>%
#   mutate(
#     Percent_Diff = (NEW_CPUE_2003.2013 - OLD_CPUE_2003.2013) / OLD_CPUE_2003.2013
#   )
# 
# # NOTE TO SELF: Save this table to show to Mel, can delete later -
# write.csv(CPUE_baseline, "/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_CPUE_Scores/CPUE_MethodsIllustration_PURG_Dec112025.csv", row.names = FALSE)


# STEP 2: CALCULATE CPUE VALUES FOR RECENT YEARS, 2014-2024 -----
CPUE_recent <- GoC_Landings_for_CPUE %>%
  filter(year >= 2014, year <= 2024) %>%
  filter(!is.na(OHI_Region_No)) %>%
  group_by(year, OHI_Region_No, Major_Functional_Group) %>%
  summarise(
    Total_Biomass_kg = sum(weight_kg_corrected),
    Total_Days = sum(days_fishing),
    OLD_mean_CPUE.recent = mean(CPUE_kg.day, na.rm = TRUE),
    NEW_mean_CPUE.recent = ifelse(Total_Days > 0, Total_Biomass_kg / Total_Days, NA_real_),
    Total_Value_mxn = sum(value, na.rm = TRUE),  # keep na.rm just in case value has NAs
    n_records = n(),
    .groups = "drop"
  )

# # NOTE TO SELF: Save this table to show to Mel, can delete later -
# write.csv(CPUE_recent, "/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_CPUE_Scores/CPUE_MethodsIllustration2_PURG_Dec112025.csv", row.names = FALSE)


# STEP 3: COMPUTE PERCENT BIOMASS FOR RECENT YEARS -----
CPUE_recent <- CPUE_recent %>%
  group_by(year, OHI_Region_No) %>%
  mutate(
    total_biomass_state_year = sum(Total_Biomass_kg, na.rm = TRUE),
    pct_of_total_biomass = (Total_Biomass_kg / total_biomass_state_year) * 100
  ) %>%
  ungroup()


# Quick aside: how much of percent catch do the NAs account for?
subset<- CPUE_recent %>% filter(is.na(Major_Functional_Group))
# In only a handful of cases do NAs account for the *majority* of annual catch - primarily in Region 5.
# And again, a lot of these NAs can be accounted for with further species key work.


# STEP 4: MERGE BASELINE TO PRESENT DATA -----
CPUE_Table <- CPUE_recent %>%
  left_join(CPUE_baseline,
            by = c("OHI_Region_No", "Major_Functional_Group")) %>%
  mutate(
    RatioCPUE_Recent.Baseline = NEW_mean_CPUE.recent / NEW_CPUE_2003.2013  # Score = CPUEpresent / CPUE_2003.2013
  )

# STEP 5: DESIRED TABLE -----
CPUE_Table <- CPUE_Table %>%
  transmute(
    OHI_Region = OHI_Region_No,
    Year = year,
    Major_Functional_Group = Major_Functional_Group,
    RatioCPUE_Recent.Baseline = round(RatioCPUE_Recent.Baseline, 3),
    Total_Landings_kg = round(Total_Biomass_kg, 1),
    Taxa_per_Total_Catch = round(pct_of_total_biomass, 2)
  ) %>%
  arrange(OHI_Region, Year, Major_Functional_Group)


# STEP 6: CAP CPUE AT A MAX OF 1 per GROUP
CPUE_Table <- CPUE_Table %>%
  mutate(
    RatioCPUE_Recent.Baseline_CAPPED = if_else(
      RatioCPUE_Recent.Baseline > 1,
      1,
      RatioCPUE_Recent.Baseline
    )
  )

# STEP 7: OHI CPUE SCORES PER REGION PER YEAR -----
CPUE_Scores <- CPUE_Table %>%
  mutate(TaxaProp = Taxa_per_Total_Catch/100) %>%     # IF % values are 0–100; if 0–1 then skip this line
  group_by(Year, OHI_Region) %>%
  summarise(
    weighted_CPUE_acrossAllGroups = sum(RatioCPUE_Recent.Baseline_CAPPED * TaxaProp, na.rm = TRUE)
  ) %>%
  ungroup()
```

# SAVE: OHI CPUE SCORES PER REGION PER YEAR
```{r Save this CPUE data table}
# write.csv(CPUE_Scores, "/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_CPUE_Scores/FIS_CPUEScores_WeightedMean_Dec12025.csv", row.names = FALSE)
```

## Take a closer look at CPUE per Region and per Major_Functional_Group:
```{r Plot CPUE over all years 2003-2024}
CPUE_AllYears <- GoC_Landings_for_CPUE %>%
  filter(year >= 2003, year <= 2024) %>%
  filter(!is.na(OHI_Region_No)) %>%
  group_by(OHI_Region_No, year, Major_Functional_Group) %>%
  summarise(
    Total_Biomass_kg = sum(weight_kg_corrected, na.rm = TRUE),
    Total_Value_mxn = sum(value, na.rm = TRUE),
    Total_Days = sum(days_fishing, na.rm = TRUE),
    Avg_CPUE = mean(CPUE_kg.day, na.rm = TRUE),
    n_records = n(),
    .groups = "drop"
  )

Taxa_colors <- c("darkblue", "royalblue", "skyblue", "orange", "sienna1", "plum", "purple4", "aquamarine3", "gold2", "tomato", "grey")

Region_colors <- c("royalblue", "skyblue", "orange", "sienna1", "plum", "purple4", "aquamarine3", "gold2", "tomato")

Regions_of_Interest <- c(
  "Region_1", "Region_2", "Region_3",
  "Region_4", "Region_5", "Region_6",
  "Region_7", "Region_8", "Region_9"
)


# PLOT 1: CPUE overTime per Region, facet by Major_Functional_Group
CPUE_overTime_byRegion_perTaxa<- CPUE_AllYears %>%
  filter(OHI_Region_No %in% Regions_of_Interest) %>%
  ggplot(
       aes(x = year, 
           y = Avg_CPUE, 
           group = OHI_Region_No, 
           color = OHI_Region_No)) +
  geom_line(size = 1) +                # lines connecting the points
  facet_wrap(~ Major_Functional_Group, scales = "free_y") +  # facet by region
  labs(
    title = "Average CPUE Over Time by Functional Group",
    x = "Year",
    y = "Avg CPUE (kg/day)",
    color = "Functional Group"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_color_manual(values = Taxa_colors) +  # nice palette
  geom_vline(xintercept = 2013, linetype= "solid", color = "grey5")

CPUE_overTime_byRegion_perTaxa

  # # Save as PNG
  # ggsave(
  #   filename = sprintf("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_CPUE_Scores/CPUE_PlotOverTime_byRegion_perTaxa_Dec092025.png"),
  #   plot = CPUE_overTime_byRegion_perTaxa,
  #   width = 12,
  #   height = 12,
  #   dpi = 300
  # )


# PLOT 2: CPUE overTime per Major_Functional_Group, facet by Region
CPUE_overTime_byRegion_perTaxa<- CPUE_AllYears %>%
  filter(OHI_Region_No %in% Regions_of_Interest) %>%
  ggplot(
       aes(x = year, 
           y = Avg_CPUE, 
           group = Major_Functional_Group, 
           color = Major_Functional_Group)) +
  geom_line(size = 1) +                # lines connecting the points
  facet_wrap(~ OHI_Region_No, scales = "free_y") +  # facet by region
  labs(
    title = "Average CPUE Over Time by Functional Group",
    x = "Year",
    y = "Avg CPUE (kg/day)",
    color = "Functional Group"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_color_manual(values = Taxa_colors) +  # nice palette
  geom_vline(xintercept = 2013, linetype= "solid", color = "grey5")

CPUE_overTime_byRegion_perTaxa

  # # Save as PNG
  # ggsave(
  #   filename = sprintf("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_CPUE_Scores/CPUE_PlotOverTime_byTaxa_perRegion_Dec092025.png"),
  #   plot = CPUE_overTime_byTaxa,
  #   width = 12,
  #   height = 12,
  #   dpi = 300
  # )


# PLOT 3: CPUE overTime per Major_Functional_Group, facet by Region - normalized 0 to 1
CPUE_overTime_byRegion_perTaxa_normalized<- CPUE_AllYears %>%
  filter(OHI_Region_No %in% Regions_of_Interest) %>%
  # Normalize Avg_CPUE per functional group across all years
  group_by(Major_Functional_Group) %>%
  mutate(Avg_CPUE_norm = (Avg_CPUE - min(Avg_CPUE, na.rm = TRUE)) /
                            (max(Avg_CPUE, na.rm = TRUE) - min(Avg_CPUE, na.rm = TRUE))) %>%
  ungroup() %>%
  ggplot(aes(
    x = year, 
    y = Avg_CPUE_norm, 
    group = Major_Functional_Group, 
    color = Major_Functional_Group
  )) +
  geom_line(size = 1) +
  facet_wrap(~ OHI_Region_No, scales = "free_y") +
  labs(
    title = "Normalized CPUE Over Time by Functional Group",
    x = "Year",
    y = "Normalized CPUE (0-1)",
    color = "Functional Group"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_color_manual(values = Taxa_colors) +
  geom_vline(xintercept = 2013, linetype = "solid", color = "grey5")

CPUE_overTime_byRegion_perTaxa_normalized

  # Save as PNG
  # ggsave(
  #   filename = sprintf("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_CPUE_Scores/CPUE_PlotOverTime_byTaxa_perRegion_normalized_Dec092025.png"),
  #   plot = CPUE_overTime_byTaxa_normalized,
  #   width = 12,
  #   height = 12,
  #   dpi = 300
  # )
```

```{r Take a closer look at CPUE per Region and per Taxa}
# Region 5
subset<- GoC_Landings_for_CPUE %>% 
  filter(OHI_Region_No == "Region_5" & is.na(IHH_functional_major_groups_sm))

table(subset$recorded_taxa_name)
table(subset$recorded_taxa_group)
# Again, a lot of these NAs can be accounted for with further species key work.

# Region 7
subset<- GoC_Landings_for_CPUE %>% 
  filter(OHI_Region_No == "Region_7" & is.na(IHH_functional_major_groups_sm))

table(subset$recorded_taxa_name)
table(subset$recorded_taxa_group)
```