# > PENDING STEP 14: Interrogate the economic unit (eu_rnpa) and vessel (vessel_name) columns. Move to CPUE? ------------------------------

```{r Load in dataframe if needed}
GOC_MarineLandings0<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Marine.Landings_Sector.Sorted.Georef_2000.present_May262026.rds"))
# 3322933 obs., 61 vars.
```


```{r Filter for marine sectors}
table(GOC_MarineLandings0$sector_estimate_draft)

Possible_Marine_Sectors<- c("marine: aquaculture", "marine: fishery sector yet to be classified", 
                        "marine: large-scale capture", "marine: small-scale capture",
                        "freshwater or marine: large-scale capture", "freshwater or marine: fishery sector yet to be classified",
                        "freshwater or marine: small-scale capture", "freshwater or marine: aquaculture")

GOC_MarineLandings<- GOC_MarineLandings0 %>% 
  filter(sector_estimate_draft %in% Possible_Marine_Sectors)
# 3312602 obs., 62 vars.

table(GOC_MarineLandings$keywords_1)

# Filter out things from freshwater:
GOC_MarineLandings <- GOC_MarineLandings %>%
  filter(!if_any(starts_with("keywords"), ~ !is.na(.) & . == "DE AGUA DULCE"))
# 3303634 obs., 62 vars.
```

```{r FILTER OUT MARINE SECTORS OF INTEREST:}

MarineFishery_Sectors<- c("marine: fishery sector yet to be classified" , "marine: small-scale capture", "marine: large-scale capture")

GOC_Landings_mar.fisheries<- GOC_MarineLandings0 %>% 
  filter(sector_estimate_draft %in% MarineFishery_Sectors)
# 3105164 obs., 61 vars.

range(GOC_Landings_mar.fisheries$year)
```

```{r First look}
length(unique(GOC_Landings_mar.fisheries$eu_name))
# 5002 unique economic units in the dataframe.

length(unique(GOC_Landings_mar.fisheries$eu_rnpa))
# 4647 unique permits given to economic units in the dataframe.

length(unique(GOC_Landings_mar.fisheries$vessel_name))
# 2141 unique vessel names in the dataframe.

length(unique(GOC_Landings_mar.fisheries$vessel_rnpa))
# 2134 unique vessel permit numbers in the dataframe.

# So, for the most part, each vessel is registered to a unique vessel permit.'

# How many of the 3229063 total observations have no economic unit name associated with them? Only ~2% of all records.
GOC_Landings_mar.fisheries %>%
  count(eu_name == "NO DISPONIBLE")
# 68006 / 3105164

# How many of all observations have no vessel name associated with them? About 51% of all records.
GOC_Landings_mar.fisheries %>%
  count(vessel_rnpa == "00000000")
# 1611882 / 3105164

# This tells us that about half of all records have no information about the fishers' vessel, but nearly all have permit info.
```

```{r Testing some ideas}
table(GOC_Landings_mar.fisheries$reported_aquaculture_production)
# No aquaculture hiding in here...

# Look into a specific unit:
subset2<- GOC_Landings_mar.fisheries %>%
  filter(eu_name == "SCPP PUNTA TIBURON, S DE RL DE CV")
table(subset2$reported_aquaculture_production)
# It looks like this particular economic unit (possibly a cooperative?) does not work in aquaculture. 

# Look into another specific unit:
subset3<- GOC_Landings_mar.fisheries %>%
  filter(eu_name == "SCPP DE SERV LA CAIMANERA SC DE RL")
table(subset3$reported_aquaculture_production)
# Notice that some of these economic units say nothing about aquaculture, whereas others do. 

# Keep workspace clean:
rm(subset2, subset3)
```
Let's leave this section under construction. I'm not totally sure what to do with this yet. But it will impact how and if we can calculate CPUE.

Waiting to here back from Juan Carlos is if there is a key or roster for economic units...

```{r Let's plot missing data over time}
# Summarize missing vessel IDs over time
vessel_missing_over_time <- GOC_Landings_mar.fisheries %>%
  mutate(missing_vessel = is.na(vessel_rnpa) | vessel_rnpa == "00000000" | vessel_rnpa == "0000000_") %>%
  group_by(year) %>%   # Replace 'year' with your time column
  summarise(
    total = n(),
    missing_count = sum(missing_vessel),
    missing_percent = 100 * missing_count / total
  )

# Plot missing vessel prevalence over time
ggplot(vessel_missing_over_time, aes(x = year, y = missing_percent)) +
  geom_line(color = "orange", size = 1.2) +
  geom_point(color = "tomato", size = 2) +
  labs(
    title = "Prevalence of Missing Vessel IDs Over Time",
    x = "Year",
    y = "Percentage of Records Missing Vessel ID"
  ) +
  scale_y_continuous(limits = c(0, 100)) +
  theme_bw()

# Vessel data is missing throughout the time series.

# Plot the opposite: what's not missing?
vessel_present_over_time <- GOC_Landings_mar.fisheries %>%
  mutate(missing_vessel = is.na(vessel_rnpa) | vessel_rnpa == "00000000" | vessel_rnpa == "0000000_") %>%
  group_by(year) %>%
  summarise(
    total = n(),
    present_count = sum(!missing_vessel),
    present_percent = 100 * present_count / total
  )

ggplot(vessel_present_over_time, aes(x = year, y = present_percent)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_point(color = "navy", size = 2) +
  
  labs(
    title = "Proportion of Records WITH Vessel IDs Over Time",
    x = "Year",
    y = "Percentage of Records with Vessel ID"
  ) +
  scale_y_continuous(limits = c(0, 100)) +
  theme_bw()

# Plot the opposite: what's not missing?
eu_present_over_time <- GOC_Landings_mar.fisheries %>%
  mutate(missing_eu = eu_name == "NO DISPONIBLE") %>%
  group_by(year) %>%
  summarise(
    total = n(),
    present_count = sum(!missing_eu),
    present_percent = 100 * present_count / total
  )

ggplot(eu_present_over_time, aes(x = year, y = present_percent)) +
  geom_line(color = "lavender", size = 1.2) +
  geom_point(color = "orchid", size = 2) +
  
  labs(
    title = "Proportion of Records WITH EU Details Over Time",
    x = "Year",
    y = "Percentage of Records with EU details"
  ) +
  scale_y_continuous(limits = c(0, 100)) +
  theme_bw()
```

```{r Investigate vessel names since 2018}
GOC_Landings_vessels<- GOC_Landings_mar.fisheries %>%
  filter(year >= 2018)
# 1104230 obs., 61 vars.

length(unique(GOC_Landings_vessels$vessel_name))
# 1297 unique vessel names in the dataframe.
length(unique(GOC_Landings_vessels$vessel_rnpa))
# 1279 unique vessel permit numbers in the dataframe.

# So, for the most part, each vessel is registered to a unique vessel permit.'

# How many of the 1104876 total observations have no economic unit name associated with them? Essentially 0% of records.
GOC_Landings_vessels %>%
  count(eu_name == "NO DISPONIBLE")
# 4 / 1104230

# How many of all observations have no vessel permit associated with them? 86% of all records.
GOC_Landings_vessels %>%
  count(vessel_rnpa == "00000000" | vessel_rnpa == "0000000_" | is.na(vessel_rnpa))
# 946953 / 1104230

# How many of all observations have no vessel NAME associated with them? About 86% of all records.
# Okay, so we know vessel name alone won't be a good estimator for CPUE...
GOC_Landings_vessels %>%
  count(is.na(vessel_name))
# 946953 / 1104230

# Take a look at a random assortment of rnpas versus names to understnad what's missing and what's not:
subset <- GOC_Landings_vessels %>%
  filter(is.na(vessel_name)) %>%
  select(vessel_rnpa, vessel_name, eu_name, eu_rnpa) %>%
  distinct() %>%
  slice_sample(n = 200)
# This shows they have real eu names, just no vessel details.
```

```{r Testing some ideas}
colnames(GOC_Landings_vessels)

# Case studies:
subset <- GOC_Landings_vessels %>%
  filter(eu_rnpa == "2508005234")

length(unique(subset$vessel_name))
# 9 vessels registered to this economic unit permit

subset2<- GOC_Landings_vessels %>%
  filter(eu_rnpa =="0203010103")

length(unique(subset2$vessel_name))
# 1 vessel name registered to this economic unit permit
length(unique(subset2$eu_name))
# 1 name for this economic unit
unique(subset2$vessel_rnpa)
# 1 vessel permit for this economic unit, 0000000_

range(subset2$year)
# From 2018 to 2025
unique(subset2$recorded_taxa_name_clean)
# Across several taxa
unique(subset2$sector_estimate_reported)
# And all small-scale (or unclassified)
```


# PENDING STEP 15: Esitmate CPUE per economic unit 2018-2024 
```{r Estimate CPUE per economic unit 2018-2024}
# Let's look through unit Economic Units. All records have a non-NA economic unit (at least since 2018)
EconomicUnits_df <- GOC_Landings_vessels %>%
  distinct(eu_rnpa)    # Keeps unique values as a one-column tibble
# 2849 obs.

GOC_Landings_vessels %>%
  count(is.na(eu_rnpa))


# Estimate Catch per economic unit 2018-2024: summarize weight_kg_corrected by eu_rnpa per year

CatchPerEU_summary <- GOC_Landings_vessels %>%
  filter(year >= 2018 & year <= 2024) %>%   # filter the years
  group_by(year, eu_rnpa) %>%              # group by year and economic unit
  summarize(
    total_catch_kg = sum(weight_kg_corrected, na.rm = TRUE)  # sum catch, ignore NAs
  ) %>%
  ungroup()  # remove grouping after summarizing

# Next, pull out the eu_rnpas that are present in all years 2018-2025. Using these eu_rnpas only, estimate the total catch of Major_Taxa_Group divided by the total eu_rnpa per year

# Step 1: Identify eu_rnpa present in all years
EU_all_years <- GOC_Landings_vessels %>%
  filter(year >= 2018 & year <= 2025) %>%
  group_by(eu_rnpa) %>%
  summarize(years_present = n_distinct(year), .groups = "drop") %>%
  filter(years_present == 8) %>%    # 2018–2025 = 8 years
  pull(eu_rnpa)                     # get vector of eu_rnpa
# 1003 eu.rnpas are present in all years 2018-2025

# Step 2: Filter data to only include these eu_rnpa
filtered_data <- GOC_Landings_vessels %>%
  filter(eu_rnpa %in% EU_all_years, year >= 2018 & year <= 2025)

# Step 3: Summarize catch by Major_Functional_Group per year
CPUE_summary <- filtered_data %>%
  group_by(year, Major_Functional_Group) %>%
  summarize(
    total_catch_mt = sum(weight_kg_corrected, na.rm = TRUE)/1000,
    n_eu = n_distinct(eu_rnpa),                    # number of economic units that year
    catch.mt_per_eu = total_catch_mt / n_eu          # divide total catch by number of units
  ) %>%
  ungroup()

# Looking promising! Let's plot it...
```

```{r Plot CPUE for Small Pelagics}
# PLOT FOR ALL GROUPS, excluding Small Pelagics:
# Remove NAs or 'unknown' if you don’t want them cluttering the plot
# First compute overall yearly mean (across all groups)

# First we need mean CPUE over years per group (2018-2024)
CPUE_mean_df <- CPUE_summary %>%
  filter(year > 2017 & year < 2025) %>%
  group_by(Major_Functional_Group) %>%
  summarise(mean_cpue = mean(catch.mt_per_eu), .groups = "drop")

CPUE_perEU_plot <- 
  
  CPUE_summary %>%
  filter(year < 2025) %>%
  
  ggplot(aes(x = year, y = catch.mt_per_eu)) +
  
  # Group-specific lines
  geom_line(aes(color = Major_Functional_Group), size = 1.1) +
  geom_point(aes(color = Major_Functional_Group), size = 2) +
  
  facet_wrap(~ Major_Functional_Group, scales = "free_y") +
  
  geom_hline(
  data = CPUE_mean_df,
  aes(yintercept = mean_cpue, group = Major_Functional_Group),
  color = "grey40",
  linetype = "dashed",
  linewidth = 0.8) +
  
  labs(
    title = "CPUE per Economic Unit by Functional Group (2018-2024)",
    x = "Year",
    y = "Catch per Economic Unit (mt)"
  ) +
  
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

CPUE_perEU_plot
```

```{r Save CPUE_perEU_plot}
# Save the plot as a high-resolution PNG
ggsave(
  filename = "/home/ferrer/ohi_goc_prep/FP/CONAPESCA_CleanUp/CPUE_perEU_plot_Apr022026.png",
  plot = CPUE_perEU_plot,
  width = 10,       # width in inches
  height = 6,       # height in inches
  dpi = 300         # resolution
)
```