---
title: 'OHI GoC 2024: Lasting Special Places (LSP)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# Sentido de Pertenencia: Lasting Special Places


## Make one protected area layer

```{r}
voluntariamente_conanp <- st_read(here::here("/home/shares/ohi/OHI_GOC/_raw_data/CONANP/619_ADVC_VIGENTES_MARZO_2026/619_ADVC_VIGENTES_MARZO_2026.shp")) %>% 
  st_zm(voluntariamente_conanp, drop = TRUE, what = "ZM") %>% 
  janitor::clean_names() %>% 
  filter(estado %in% c("Baja California", "Baja California Sur","Sonora","Sinaloa","Nayarit","Jalisco"))

# verify the geometry types are now consistent
table(st_geometry_type(voluntariamente_conanp))

mapview(voluntariamente_conanp)
```

```{r}
zonification_primaria <- st_read(here::here("/home/shares/ohi/OHI_GOC/_raw_data/CONANP/SHAPE_ANPS_ZN/ZP_ANPS_22042024.shp")) %>% 
  janitor::clean_names() 
  filter(estado %in% c("Baja California", "Baja California Sur","Sonora","Sinaloa","Nayarit","Jalisco"))

mapview(zonification_primaria)
```

```{r}
conanp_anps_julio_2025 <- st_read(here::here("/home/shares/ohi/OHI_GOC/_raw_data/CONANP/232_ANP-ITRF08_04072025/232_ANP-ITRF08_04072025.shp")) %>% 
  st_zm(conanp_anps_julio_2025, drop = TRUE, what = "ZM") %>% 
  janitor::clean_names()

# verify the geometry types are now consistent
table(st_geometry_type(conanp_anps_julio_2025))

mapview(conanp_anps_julio_2025) + mapview(all_protected, col.regions = "pink")
```

```{r}
regions_CONANP <- st_read(here::here("/home/shares/ohi/OHI_GOC/_raw_data/CONANP/Regionalizacion_22052017/Regionalizacion_22052017.shp")) %>% 
  janitor::clean_names()
```

##### Dissolve protected areas into one layer

```{r}
# check CRS
st_crs(wdpa_poly_all) == st_crs(hem_protected_anps) # FALSE

# transform hem to match wdpa (or vice versa)
hem_protected_anps_reproj <- st_transform(hem_protected_anps, st_crs(wdpa_poly_all))
```


```{r}
st_crs(conanp_anps_julio_2025) == st_crs(hem_protected_anps_reproj) ## FALSE

# transform conanp to match wdpa 
conanp_anps_julio_2025_reproj <- st_transform(conanp_anps_julio_2025, st_crs(wdpa_poly_all))
```


```{r}
# harmonize geometry types (cast both to MULTIPOLYGON to be safe)
wdpa_clean <- st_cast(wdpa_poly_all, "MULTIPOLYGON") %>% 
  mutate(source = "wdpa_2025")
hem_clean  <- st_cast(hem_protected_anps_reproj, "MULTIPOLYGON") %>% 
  mutate(source = "hem")
conanp_clean <- st_cast(conanp_anps_julio_2025_reproj, "MULTIPOLYGON") %>% 
  mutate(source = "conanp_2025")

# select only shared/needed columns before binding (bind_rows will fill NA for mismatched columns — fine, but tidy to be explicit)
all_protected <- bind_rows(
  wdpa_clean %>% dplyr::select(name, source, geometry),
  hem_clean %>% dplyr::select(name = nombre, source, geometry),
  conanp_clean %>% dplyr::select(name = nombre, source, geometry)
) 
mapview(all_protected)

# dissolve overlaps for a single coverage layer
all_protected_dissolved <- st_union(all_protected)
mapview(all_protected_dissolved)

# write out to mazu
st_write(all_protected, here::here("/home/shares/ohi/OHI_GOC/spatial/protected_areas_wdpa_conanp_and_hem", "all_goc_protected_areas_v2_2025.shp"))
st_write(all_protected_dissolved, here::here("/home/shares/ohi/OHI_GOC/spatial/protected_areas_wdpa_conanp_and_hem", "all_goc_protected_areas_dissolved_v2_2025.shp"))
```


## List of coordinates from the Expert Working Group (EWG)

This subgoal's current status is based on whether there is a protection on the Lasting Special Place (LSP), and the condition is then determined by the effectiveness of that protection on the LSP.  For example, if the LSP is a specific island such as Isla Tiburón, so the island falls within a protected area, then its condition will be based on the management effectiveness determined by CONANP's SIMEC index of effectiveness (i-efectividad).

If they do not fall within a protected area, then the condition of protection will be determined accordingly.  Here is a table of the types of LSPs and their associated protections: <https://docs.google.com/spreadsheets/d/1RwNTzRZdBIDtzi7d1ceFxhNmVghN47NxTgrhZV8O2SM/edit?gid=1758421523#gid=1758421523>

So, first, we need to go through the list and separate into the type of LSP...


## Read in EWG list of LSP sites

```{r}
lsp_google_sheets_dir <- "/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/google_sheets"
lsp_list_ewg <- read_csv(file.path(lsp_google_sheets_dir, "ewg_lsp_inventory_prepared.csv")) %>% 
  janitor::clean_names()

lsp_list_ewg_pts <- lsp_list_ewg %>%
  mutate(lat = as.numeric(lat), long = as.numeric(long)) %>%
  filter(!is.na(lat), !is.na(long)) %>%
  sf::st_as_sf(coords = c("long", "lat"), crs = 4326, remove = FALSE)

mapview(lsp_list_ewg_pts, col.regions = "orange") + mapview(all_protected, col.regions = "blue")
```




