---
title: "Saltmarsh Habitat data exploration"
author: "Austin Martinez"
date: 08/014/2026
format: 
  html:
    code-fold: show
execute:
  echo: true
  message: false
  warning: false
---

```{r}
# Load in Libraries
library(sf)
library(dplyr)
library(here)
library(terra)
library(tidyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(tmap)
library(stringr)
```

```{r}
# Read in OHI region boundaries
ohi_regions <- st_read(here("spatial", "ohi_regions", "goc_ohi_rgns.shp")) |> st_make_valid()
```

```{r}
# Read in Hem's satmarsh layer
hem_saltmarsh <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/Hem_data/goc_habitat_data/wetlands/wetlands_merge.shp") |> st_make_valid()
```

## CANABIO Land Use and Vegitation all years
#### Dunes data
```{r}
dune_canabio_landuse_1980_91 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250kcs1agw/usv250kcs1agw.shp") |> st_make_valid()

dune_canabio_landuse_2001 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250ks2gw/usv250ks2gw.shp") |> st_make_valid()

dune_canabio_landuse_2005 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250ks3gw/usv250ks3gw.shp") |> st_make_valid()

dune_canabio_landuse_2009 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250ks4gw/usv250ks4gw.shp") |> st_make_valid()

dune_canabio_landuse_2013 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250s5ugw/usv250s5ugw.shp") |> st_make_valid()

dune_canabio_landuse_2016 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250s6gw/usv250s6gw.shp") |> st_make_valid()

dune_canabio_landuse_2021 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250s7gw/usv250s7gw.shp") |> st_make_valid()
```

### Saltmarsh
```{r}
saltmarsh_canabio_landuse_1980_91 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250kcs1agw/usv250kcs1agw.shp") |> st_make_valid()

saltmarsh_canabio_landuse_2001 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250ks2gw/usv250ks2gw.shp") |> st_make_valid()

saltmarsh_canabio_landuse_2005 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250ks3gw/usv250ks3gw.shp") |> st_make_valid()

saltmarsh_canabio_landuse_2009 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250ks4gw/usv250ks4gw.shp") |> st_make_valid()

saltmarsh_canabio_landuse_2013 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250s5ugw/usv250s5ugw.shp") |> st_make_valid()

saltmarsh_canabio_landuse_2016 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250s6gw/usv250s6gw.shp") |> st_make_valid()

saltmarsh_canabio_landuse_2021 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250s7gw/usv250s7gw.shp") |> st_make_valid()
```


## Habitat buffers:
### Dunes
```{r}
# Create 5km buffer (run once)
sf::sf_use_s2(FALSE)

ohi_regions_marine <- ohi_regions |>
  filter(location == "marine") |>
  st_union()

# Make a 5km buffer ring
buffer_outer <- st_buffer(ohi_regions_marine, dist = 5000)
buffer_ring  <- st_difference(buffer_outer, ohi_regions_marine) |> st_make_valid()

# Union it to the marine OHI layer
ohi_marine_plus_5km <- st_union(ohi_regions_marine, buffer_ring)

# Make a 1km buffer ring
buffer_outer <- st_buffer(ohi_regions_marine, dist = 1000)
buffer_ring  <- st_difference(buffer_outer, ohi_regions_marine) |> st_make_valid()

# Union it to the marine OHI layer
ohi_marine_plus_1km <- st_union(ohi_regions_marine, buffer_ring)
```

### Saltmarsh
```{r}
ohi_regions_marine <- ohi_regions |>
  filter(location == "marine") |>
  st_union()

# Make a buffer ring
buffer_outer <- st_buffer(ohi_regions_marine, dist = 5000)
buffer_ring  <- st_difference(buffer_outer, ohi_regions_marine) |> st_make_valid()

# Union it to the marine OHI layer
ohi_marine_plus_5km <- st_union(ohi_regions_marine, buffer_ring)
```


### Process Saltmarsh data:
We are using these layers to represent Saltmarshes:
-     VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA 
-     VEGETACIÓN HALÓFILA


```{r}
# 1980-91
saltmarsh_canabio_1980_91_sublayer <- saltmarsh_canabio_landuse_1980_91 |>
  dplyr::filter(
    descripcio %in% c(
      "VEGETACIÓN HALÓFILA",
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA"
    )
  )

# Match CRS of OHI and Buffer OHI regions
ohi_region <- st_transform(ohi_regions, st_crs(saltmarsh_canabio_1980_91_sublayer))
ohi_marine_plus_5km <- st_transform(ohi_marine_plus_5km, st_crs(saltmarsh_canabio_1980_91_sublayer))

saltmarsh_canabio_1980_91_sublayer <- st_intersection(saltmarsh_canabio_1980_91_sublayer, ohi_marine_plus_5km)

saltmarsh_canabio_1980_91_sublayer_area <- st_intersection(saltmarsh_canabio_1980_91_sublayer, ohi_region) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 1980
  ) |>
  st_drop_geometry()
```

```{r}
# Check for double counting
any(lengths(st_overlaps(saltmarsh_canabio_1980_91_sublayer)) > 0)
```


```{r}
# 2001
saltmarsh_canabio_2001_sublayer <- saltmarsh_canabio_landuse_2001 |>
  dplyr::filter(
    descripcio %in% c(
      "VEGETACIÓN HALÓFILA HIDRÓFILA",
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA HIDRÓFILA"
    )
  )

# Match CRS of OHI and Buffer OHI regions
ohi_region <- st_transform(ohi_regions, st_crs(saltmarsh_canabio_2001_sublayer))
ohi_marine_plus_5km <- st_transform(ohi_marine_plus_5km, st_crs(saltmarsh_canabio_2001_sublayer))

saltmarsh_canabio_2001_sublayer <- st_intersection(saltmarsh_canabio_2001_sublayer, ohi_marine_plus_5km)

saltmarsh_canabio_2001_sublayer_area <- st_intersection(saltmarsh_canabio_2001_sublayer, ohi_region) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2001
  ) |>
  st_drop_geometry()
```


```{r}
# 2005
saltmarsh_canabio_2005_sublayer <- saltmarsh_canabio_landuse_2005 |>
  dplyr::filter(
    descripcio %in% c(
      "VEGETACIÓN HALÓFILA HIDRÓFILA",
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA HIDRÓFILA"
    )
  )

# Match CRS of OHI and Buffer OHI regions
ohi_region <- st_transform(ohi_regions, st_crs(saltmarsh_canabio_2005_sublayer))
ohi_marine_plus_5km <- st_transform(ohi_marine_plus_5km, st_crs(saltmarsh_canabio_2005_sublayer))

saltmarsh_canabio_2005_sublayer <- st_intersection(saltmarsh_canabio_2005_sublayer, ohi_marine_plus_5km)

saltmarsh_canabio_2005_sublayer_area <- st_intersection(saltmarsh_canabio_2005_sublayer, ohi_region) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2005
  ) |>
  st_drop_geometry()
```


```{r}
# 2009
saltmarsh_canabio_2009_sublayer <- saltmarsh_canabio_landuse_2009 |>
  dplyr::filter(
    descripcio %in% c(
      "VEGETACIÓN HALÓFILA HIDRÓFILA",
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA HIDRÓFILA"
    )
  )

# Match CRS of OHI and Buffer OHI regions
ohi_region <- st_transform(ohi_regions, st_crs(saltmarsh_canabio_2009_sublayer))
ohi_marine_plus_5km <- st_transform(ohi_marine_plus_5km, st_crs(saltmarsh_canabio_2009_sublayer))

saltmarsh_canabio_2009_sublayer <- st_intersection(saltmarsh_canabio_2009_sublayer, ohi_marine_plus_5km)

saltmarsh_canabio_2009_sublayer_area <- st_intersection(saltmarsh_canabio_2009_sublayer, ohi_region) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2009
  ) |>
  st_drop_geometry()
```


```{r}
# 2013
saltmarsh_canabio_2013_sublayer <- saltmarsh_canabio_landuse_2013 |>
  dplyr::filter(
    descripcio %in% c(
      "VEGETACIÓN HALÓFILA HIDRÓFILA",
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA HIDRÓFILA"
    )
  )

# Match CRS of OHI and Buffer OHI regions
ohi_region <- st_transform(ohi_regions, st_crs(saltmarsh_canabio_2013_sublayer))
ohi_marine_plus_5km <- st_transform(ohi_marine_plus_5km, st_crs(saltmarsh_canabio_2013_sublayer))

saltmarsh_canabio_2013_sublayer <- st_intersection(saltmarsh_canabio_2013_sublayer, ohi_marine_plus_5km)

saltmarsh_canabio_2013_sublayer_area <- st_intersection(saltmarsh_canabio_2013_sublayer, ohi_region) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2013
  ) |>
  st_drop_geometry()
```


```{r}
# 2016
saltmarsh_canabio_2016_sublayer <- saltmarsh_canabio_landuse_2016 |>
  dplyr::filter(
    DESCRIPCIO %in% c(
      "VEGETACIÓN HALÓFILA HIDRÓFILA",
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA HIDRÓFILA"
    )
  )

# Match CRS of OHI and Buffer OHI regions
ohi_region <- st_transform(ohi_regions, st_crs(saltmarsh_canabio_2016_sublayer))
ohi_marine_plus_5km <- st_transform(ohi_marine_plus_5km, st_crs(saltmarsh_canabio_2016_sublayer))

saltmarsh_canabio_2016_sublayer <- st_intersection(saltmarsh_canabio_2016_sublayer, ohi_marine_plus_5km)

saltmarsh_canabio_2016_sublayer_area <- st_intersection(saltmarsh_canabio_2016_sublayer, ohi_region) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2016
  ) |>
  st_drop_geometry()
```


```{r}
# 2021
saltmarsh_canabio_2021_sublayer <- saltmarsh_canabio_landuse_2021 |>
  dplyr::filter(
    DESCRIPCIO %in% c(
      "VEGETACIÓN HALÓFILA HIDRÓFILA",
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA HIDRÓFILA"
    )
  )

# Match CRS of OHI and Buffer OHI regions
ohi_region <- st_transform(ohi_regions, st_crs(saltmarsh_canabio_2021_sublayer))
ohi_marine_plus_5km <- st_transform(ohi_marine_plus_5km, st_crs(saltmarsh_canabio_2021_sublayer))

saltmarsh_canabio_2021_sublayer <- st_intersection(saltmarsh_canabio_2021_sublayer, ohi_marine_plus_5km)

saltmarsh_canabio_2021_sublayer_area <- st_intersection(saltmarsh_canabio_2021_sublayer, ohi_region) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2021
  ) |>
  st_drop_geometry()
```

```{r}
# Combine all years into one long table
saltmarsh_all_years <- dplyr::bind_rows(
  saltmarsh_canabio_1980_91_sublayer_area,
  saltmarsh_canabio_2001_sublayer_area,
  saltmarsh_canabio_2005_sublayer_area,
  saltmarsh_canabio_2009_sublayer_area,
  saltmarsh_canabio_2013_sublayer_area,
  saltmarsh_canabio_2016_sublayer_area,
  saltmarsh_canabio_2021_sublayer_area
)
```

### Process dune data:
We are using these layers to represent Coastal dunes:
-     VEGETACIÓN DE DUNAS COSTERAS with a 5km buffer
-     VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA XERÓFILA with a 1km buffer
-     VEGETACIÓN HALÓFILA XERÓFILA with a 1km buffer


1991 has no observations for layers:
      "VEGETACIÓN HALÓFILA XERÓFILA",
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA XERÓFILA"
      
So, only the 5km buffer was used in the analysis.
```{r}
sf::sf_use_s2(FALSE)

# 1980_91
dune_canabio_landuse_1980_91_sublayer_5km <- dune_canabio_landuse_1980_91 |>
  dplyr::filter(
    descripcio %in% c(
      "VEGETACIÓN DE DUNAS COSTERAS"
    )
  )

dune_canabio_landuse_1980_91_sublayer_1km <- dune_canabio_landuse_1980_91 |>
  dplyr::filter(
    descripcio %in% c(
      "VEGETACIÓN HALÓFILA XERÓFILA",
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA XERÓFILA"
    )
  )

# Match CRS of OHI and Buffer OHI regions
ohi_region <- st_transform(ohi_regions, st_crs(dune_canabio_landuse_1980_91))


ohi_marine_plus_1km <- st_transform(ohi_marine_plus_1km, st_crs(dune_canabio_landuse_1980_91))
ohi_marine_plus_5km <- st_transform(ohi_marine_plus_5km, st_crs(dune_canabio_landuse_1980_91))


#dune_canabio_landuse_1980_91_sublayer_1km <- st_intersection(dune_canabio_landuse_1980_91_sublayer_1km, ohi_marine_plus_1km)
dune_canabio_landuse_1980_91_sublayer_5km <- st_intersection(dune_canabio_landuse_1980_91_sublayer_5km, ohi_marine_plus_5km)

#dune_canabio_landuse_1980_91_sublayer <- st_union(dune_canabio_landuse_1980_91_sublayer_1km, dune_canabio_landuse_1980_91_sublayer_5km)

dune_canabio_landuse_1980_91_sublayer_area <- st_intersection(dune_canabio_landuse_1980_91_sublayer_5km, ohi_region) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 1991
  ) |>
  st_drop_geometry()
```


```{r}
# 2001
dune_canabio_2001_sublayer_5km <- dune_canabio_landuse_2001 |>
  dplyr::filter(
    descripcio %in% c(
      "VEGETACIÓN DE DUNAS COSTERAS"
    )
  )

dune_canabio_2001_sublayer_1km <- dune_canabio_landuse_2001 |>
  dplyr::filter(
    descripcio %in% c(
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA XERÓFILA",
      "VEGETACIÓN HALÓFILA XERÓFILA"
    )
  )

# Match CRS of OHI and Buffer OHI regions
ohi_region <- st_transform(ohi_regions, st_crs(dune_canabio_landuse_2001))
ohi_marine_plus_5km <- st_transform(ohi_marine_plus_5km, st_crs(dune_canabio_landuse_2001))
ohi_marine_plus_1km <- st_transform(ohi_marine_plus_1km, st_crs(dune_canabio_landuse_2001))

dune_canabio_2001_sublayer_5km <- st_intersection(dune_canabio_2001_sublayer_5km, ohi_marine_plus_5km)
dune_canabio_2001_sublayer_1km <- st_intersection(dune_canabio_2001_sublayer_1km, ohi_marine_plus_1km)

dune_canabio_2001_sublayer <- st_union(dune_canabio_2001_sublayer_5km, dune_canabio_2001_sublayer_1km)

dune_canabio_2001_sublayer_area <- st_intersection(dune_canabio_2001_sublayer, ohi_region) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2001
  ) |>
  st_drop_geometry()
```

```{r}
# 2005
dune_canabio_2005_sublayer_5km <- dune_canabio_landuse_2005 |>
  dplyr::filter(
    descripcio %in% c(
      "VEGETACIÓN DE DUNAS COSTERAS"
    )
  )

dune_canabio_2005_sublayer_1km <- dune_canabio_landuse_2005 |>
  dplyr::filter(
    descripcio %in% c(
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA XERÓFILA",
      "VEGETACIÓN HALÓFILA XERÓFILA"
    )
  )

# Match CRS of OHI and Buffer OHI regions
ohi_region <- st_transform(ohi_regions, st_crs(dune_canabio_landuse_2005))
ohi_marine_plus_5km <- st_transform(ohi_marine_plus_5km, st_crs(dune_canabio_landuse_2005))
ohi_marine_plus_1km <- st_transform(ohi_marine_plus_1km, st_crs(dune_canabio_landuse_2005))

dune_canabio_2005_sublayer_5km <- st_intersection(dune_canabio_2005_sublayer_5km, ohi_marine_plus_5km)
dune_canabio_2005_sublayer_1km <- st_intersection(dune_canabio_2005_sublayer_1km, ohi_marine_plus_1km)

dune_canabio_2005_sublayer <- st_union(dune_canabio_2005_sublayer_5km, dune_canabio_2005_sublayer_1km)

dune_canabio_2005_sublayer_area <- st_intersection(dune_canabio_2005_sublayer, ohi_region) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2005
  ) |>
  st_drop_geometry()
```

```{r}
# 2009
dune_canabio_2009_sublayer_5km <- dune_canabio_landuse_2009 |>
  dplyr::filter(
    descripcio %in% c(
      "VEGETACIÓN DE DUNAS COSTERAS"
    )
  )

dune_canabio_2009_sublayer_1km <- dune_canabio_landuse_2009 |>
  dplyr::filter(
    descripcio %in% c(
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA XERÓFILA",
      "VEGETACIÓN HALÓFILA XERÓFILA"
    )
  )

# Match CRS of OHI and Buffer OHI regions
ohi_region <- st_transform(ohi_regions, st_crs(dune_canabio_landuse_2009))
ohi_marine_plus_5km <- st_transform(ohi_marine_plus_5km, st_crs(dune_canabio_landuse_2009))
ohi_marine_plus_1km <- st_transform(ohi_marine_plus_1km, st_crs(dune_canabio_landuse_2009))

dune_canabio_2009_sublayer_5km <- st_intersection(dune_canabio_2009_sublayer_5km, ohi_marine_plus_5km)
dune_canabio_2009_sublayer_1km <- st_intersection(dune_canabio_2009_sublayer_1km, ohi_marine_plus_1km)

dune_canabio_2009_sublayer <- st_union(dune_canabio_2009_sublayer_5km, dune_canabio_2009_sublayer_1km)

dune_canabio_2009_sublayer_area <- st_intersection(dune_canabio_2009_sublayer, ohi_region) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2009
  ) |>
  st_drop_geometry()
```

```{r}
# 2013
dune_canabio_2013_sublayer_5km <- dune_canabio_landuse_2013 |>
  dplyr::filter(
    descripcio %in% c(
      "VEGETACIÓN DE DUNAS COSTERAS"
    )
  )

dune_canabio_2013_sublayer_1km <- dune_canabio_landuse_2013 |>
  dplyr::filter(
    descripcio %in% c(
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA XERÓFILA",
      "VEGETACIÓN HALÓFILA XERÓFILA"
    )
  )

# Match CRS of OHI and Buffer OHI regions
ohi_region <- st_transform(ohi_regions, st_crs(dune_canabio_landuse_2013))
ohi_marine_plus_5km <- st_transform(ohi_marine_plus_5km, st_crs(dune_canabio_landuse_2013))
ohi_marine_plus_1km <- st_transform(ohi_marine_plus_1km, st_crs(dune_canabio_landuse_2013))

dune_canabio_2013_sublayer_5km <- st_intersection(dune_canabio_2013_sublayer_5km, ohi_marine_plus_5km)
dune_canabio_2013_sublayer_1km <- st_intersection(dune_canabio_2013_sublayer_1km, ohi_marine_plus_1km)

dune_canabio_2013_sublayer <- st_union(dune_canabio_2013_sublayer_5km, dune_canabio_2013_sublayer_1km)

dune_canabio_2013_sublayer_area <- st_intersection(dune_canabio_2013_sublayer, ohi_region) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2013
  ) |>
  st_drop_geometry()
```

```{r}
# 2016
dune_canabio_2016_sublayer_5km <- dune_canabio_landuse_2016 |>
  dplyr::filter(
    DESCRIPCIO %in% c(
      "VEGETACIÓN DE DUNAS COSTERAS"
    )
  )

dune_canabio_2016_sublayer_1km <- dune_canabio_landuse_2016 |>
  dplyr::filter(
    DESCRIPCIO %in% c(
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA XERÓFILA",
      "VEGETACIÓN HALÓFILA XERÓFILA"
    )
  )

# Match CRS of OHI and Buffer OHI regions
ohi_region <- st_transform(ohi_regions, st_crs(dune_canabio_landuse_2016))
ohi_marine_plus_5km <- st_transform(ohi_marine_plus_5km, st_crs(dune_canabio_landuse_2016))
ohi_marine_plus_1km <- st_transform(ohi_marine_plus_1km, st_crs(dune_canabio_landuse_2016))

dune_canabio_2016_sublayer_5km <- st_intersection(dune_canabio_2016_sublayer_5km, ohi_marine_plus_5km)
dune_canabio_2016_sublayer_1km <- st_intersection(dune_canabio_2016_sublayer_1km, ohi_marine_plus_1km)

dune_canabio_2016_sublayer <- st_union(dune_canabio_2016_sublayer_5km, dune_canabio_2016_sublayer_1km)

dune_canabio_2016_sublayer_area <- st_intersection(dune_canabio_2016_sublayer, ohi_region) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2016
  ) |>
  st_drop_geometry()
```

```{r}
# 2021
dune_canabio_2021_sublayer_5km <- dune_canabio_landuse_2021 |>
  dplyr::filter(
    DESCRIPCIO %in% c(
      "VEGETACIÓN DE DUNAS COSTERAS"
    )
  )

dune_canabio_2021_sublayer_1km <- dune_canabio_landuse_2021 |>
  dplyr::filter(
    DESCRIPCIO %in% c(
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA XERÓFILA",
      "VEGETACIÓN HALÓFILA XERÓFILA"
    )
  )

# Match CRS of OHI and Buffer OHI regions
ohi_region <- st_transform(ohi_regions, st_crs(dune_canabio_landuse_2021))
ohi_marine_plus_5km <- st_transform(ohi_marine_plus_5km, st_crs(dune_canabio_landuse_2021))
ohi_marine_plus_1km <- st_transform(ohi_marine_plus_1km, st_crs(dune_canabio_landuse_2021))

dune_canabio_2021_sublayer_5km <- st_intersection(dune_canabio_2021_sublayer_5km, ohi_marine_plus_5km)
dune_canabio_2021_sublayer_1km <- st_intersection(dune_canabio_2021_sublayer_1km, ohi_marine_plus_1km)

dune_canabio_2021_sublayer <- st_union(dune_canabio_2021_sublayer_5km, dune_canabio_2021_sublayer_1km)

dune_canabio_2021_sublayer_area <- st_intersection(dune_canabio_2021_sublayer, ohi_region) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2021
  ) |>
  st_drop_geometry()
```

```{r}
# Combine all years into one long table
dune_all_years <- dplyr::bind_rows(
  dune_canabio_landuse_1980_91_sublayer_area,
  dune_canabio_2001_sublayer_area,
  dune_canabio_2005_sublayer_area,
  dune_canabio_2009_sublayer_area,
  dune_canabio_2013_sublayer_area,
  dune_canabio_2016_sublayer_area,
  dune_canabio_2021_sublayer_area
)
```

# Hem's data
```{r}
hem_saltmarsh <- st_transform(hem_saltmarsh, st_crs(ohi_regions))
```

```{r}
hem_saltmarsh_no_mangrove <- hem_saltmarsh |>
  dplyr::filter(VEG_TYPES != "Manglar - Vegetacion Primaria")
```


# Other wetland Hem data
```{r}
tmap_mode("view")

tm_basemap("Esri.WorldImagery") +
tm_shape(hem_saltmarsh_no_mangrove) +
  tm_polygons(
    col = "VEG_TYPES",
    border.col = "gray",   # outline color
    lwd = 0.01,              # thicker outline
    alpha = 0.35,
    palette = c(
      "Agua" = "darkorange",                     
      "Popal-tular - Vegetacion Primaria" = "yellow",
      "Vegetacion de dunas costeras - Vegetacion Primaria" = "#7ED321",
      "Vegetacion halofila y gipsofila - Vegetacion Primaria" = "#BD10E0"
    )
  ) +

tm_shape(saltmarsh_canabio_2021_sublayer) +
  tm_polygons(
    col = "red",
    border.col = "red",
    lwd = .9,
    alpha = 0.2,
    group = "CONABIO Saltmarsh"
  )

```

```{r}
tmap_mode("view")

# Agua
tm_shape(hem_saltmarsh_no_mangrove |> filter(VEG_TYPES == "Agua")) +
  tm_polygons(
    col = "#4A90E2",
    border.col = "#4A90E2",
    lwd = 0.4,
    alpha = 0.4,
    group = "Agua"
  ) +

# Popal-tular
tm_shape(hem_saltmarsh_no_mangrove |> filter(VEG_TYPES == "Popal-tular - Vegetacion Primaria")) +
  tm_polygons(
    col = "#7ED321",
    border.col = "#7ED321",
    lwd = 0.4,
    alpha = 0.4,
    group = "Popal-tular"
  ) +

# Dune vegetation
tm_shape(hem_saltmarsh_no_mangrove |> filter(VEG_TYPES == "Vegetacion de dunas costeras - Vegetacion Primaria")) +
  tm_polygons(
    col = "#F5A623",
    border.col = "#F5A623",
    lwd = 0.4,
    alpha = 0.4,
    group = "Dunas"
  ) +

# Halophilic/gypseous vegetation
tm_shape(hem_saltmarsh_no_mangrove |> filter(VEG_TYPES == "Vegetacion halofila y gipsofila - Vegetacion Primaria")) +
  tm_polygons(
    col = "#BD10E0",
    border.col = "#BD10E0",
    lwd = 0.4,
    alpha = 0.4,
    group = "Halofila/Gipsofila"
  ) +

# Marine regions
tm_shape(ohi_regions |> filter(location == "marine")) +
  tm_borders(col = "gray60", lwd = 0.2)

```



```{r}
tmap_mode("view")

tm_shape(hem_saltmarsh_no_mangrove) +
  tm_polygons(
    col = "VEG_TYPES",
    border.col = "gray",
    lwd = 0.01,
    alpha = 0.3,
    palette = c(
      "Agua" = "#4A90E2",
      "Popal-tular - Vegetacion Primaria" = "#7ED321",
      "Vegetacion de dunas costeras - Vegetacion Primaria" = "#F5A623",
      "Vegetacion halofila y gipsofila - Vegetacion Primaria" = "#BD10E0"
    ),
    group = "Hem Vegetation"
  ) +

tm_shape(saltmarsh_canabio_2021_sublayer) +
  tm_polygons(
    col = "red",
    border.col = "red",
    lwd = 0.7,
    alpha = 0.1,
    group = "CONABIO Saltmarsh"
  )

```

## Identify intersections with any Hem saltmarsh polygon
```{r}
# Match CRS
saltmarsh_canabio_2021_sublayer <- st_transform(
  saltmarsh_canabio_2021_sublayer,
  st_crs(hem_saltmarsh_no_mangrove)
)

# Add ID + area
conabio <- saltmarsh_canabio_2021_sublayer
conabio$conabio_id <- seq_len(nrow(conabio))
conabio$conabio_area <- as.numeric(st_area(conabio))

# Intersection with ANY Hem saltmarsh
ints_any <- st_intersection(conabio, hem_saltmarsh_no_mangrove)

# Compute intersection area BEFORE dropping geometry
ints_any$intersection_area <- as.numeric(st_area(ints_any))

# Summarize by ID
ints_any_df <- ints_any |>
  st_drop_geometry() |>
  group_by(conabio_id) |>
  summarise(intersection_area = sum(intersection_area))

# Join + percent overlap
overlap_any <- conabio |>
  st_drop_geometry() |>
  left_join(ints_any_df, by = "conabio_id") |>
  mutate(
    intersection_area = replace_na(intersection_area, 0),
    pct_inside_any = 100 * intersection_area / conabio_area
  )

# ONE NUMBER: percent of total CONABIO area inside ANY Hem saltmarsh
pct_any <- sum(overlap_any$intersection_area) / sum(overlap_any$conabio_area) * 100
pct_any

```

## Identify intersections with any Hem saltmarsh polygon "Vegetacion halofila y gipsofila - Vegetacion Primaria" only
```{r}
# Filter Hem halophyte/gypseous
hem_halo <- hem_saltmarsh_no_mangrove |>
  filter(VEG_TYPES == "Vegetacion halofila y gipsofila - Vegetacion Primaria")

# Intersection
ints_halo <- st_intersection(conabio, hem_halo)

# Compute intersection area BEFORE dropping geometry
ints_halo$intersection_area_halo <- as.numeric(st_area(ints_halo))

# Summarize by ID
ints_halo_df <- ints_halo |>
  st_drop_geometry() |>
  group_by(conabio_id) |>
  summarise(intersection_area_halo = sum(intersection_area_halo))

# Join + percent overlap
overlap_halo <- conabio |>
  st_drop_geometry() |>
  left_join(ints_halo_df, by = "conabio_id") |>
  mutate(
    intersection_area_halo = replace_na(intersection_area_halo, 0),
    pct_inside_halo = 100 * intersection_area_halo / conabio_area
  )

# ONE NUMBER: percent of total CONABIO area inside Hem halophyte/gypseous vegetation
pct_halo <- sum(overlap_halo$intersection_area_halo) / sum(overlap_halo$conabio_area) * 100
pct_halo

```


```{r}
# Intersect Hem saltmarsh with OHI regions
hem_region_intersection <- st_intersection(hem_saltmarsh_no_mangrove, ohi_region)

# Compute area BEFORE summarise
hem_region_intersection$area_km2 <- as.numeric(st_area(hem_region_intersection)) / 1e6

# Summarize Hem area per region
hem_region_area <- hem_region_intersection |>
  st_drop_geometry() |>
  group_by(rgn_id) |>
  summarise(hem_area_km2 = sum(area_km2))

# Join Hem + CONABIO region areas
area_compare <- hem_region_area |>
  left_join(saltmarsh_canabio_2021_sublayer_area, by = "rgn_id") |>
  rename(conabio_area_km2 = area_km2) |>
  mutate(
    diff_km2 = hem_area_km2 - conabio_area_km2,
    pct_difference = (hem_area_km2 / conabio_area_km2) * 100
  )

# Display comparison table
area_compare |>
  select(rgn_id, hem_area_km2, conabio_area_km2, diff_km2) |>
  kable(digits = 2, caption = "Hem non-mangove vs CONABIO Saltmarsh Area per OHI Region") |>
  kable_styling( bootstrap_options = c("striped", "bordered", "condensed"),
    full_width = FALSE)


```



```{r}
# Filter Hem halophyte/gypseous vegetation
hem_halo <- hem_saltmarsh_no_mangrove |>
  filter(VEG_TYPES == "Vegetacion halofila y gipsofila - Vegetacion Primaria")

# Intersect Hem halophyte/gypseous with OHI regions
hem_halo_region_intersection <- st_intersection(hem_halo, ohi_region)

# Compute area BEFORE summarise
hem_halo_region_intersection$area_km2 <- as.numeric(st_area(hem_halo_region_intersection)) / 1e6

# Summarize Hem halophyte/gypseous area per region
hem_halo_region_area <- hem_halo_region_intersection |>
  st_drop_geometry() |>
  group_by(rgn_id) |>
  summarise(hem_halo_area_km2 = sum(area_km2))

# Join Hem halophyte/gypseous + CONABIO region areas
area_compare_halo <- hem_halo_region_area |>
  left_join(saltmarsh_canabio_2021_sublayer_area, by = "rgn_id") |>
  rename(conabio_area_km2 = area_km2) |>
  mutate(
    diff_km2 = hem_halo_area_km2 - conabio_area_km2
  )

# Display comparison table (no year, no pct_difference)
area_compare_halo |>
  select(rgn_id, hem_halo_area_km2, conabio_area_km2, diff_km2) |>
  kable(
    digits = 2,
    caption = "Hem halophyte/gypseous vs CONABIO Saltmarsh Area per OHI Region"
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "bordered", "condensed"),
    full_width = FALSE
  )

```

# Aquaculture
```{r}
# 2021
saltmarsh_canabio_2021_aquaculture <- saltmarsh_canabio_landuse_2021 |>
  dplyr::filter(
    DESCRIPCIO %in% c(
      "ACUÍCOLA"
    )
  )
```


```{r}
tmap_mode("view")

tm_basemap("Esri.WorldImagery") +
tm_shape(saltmarsh_canabio_2021_aquaculture) +
  tm_polygons(
    col = "red",
    border.col = "red",
    lwd = .9,
    alpha = 0.2,
    group = "Aquaculture"
  )

```

# Gypsophilous vegetation
```{r}
# 2021
# 2021 Gypsophilous subclasses
saltmarsh_canabio_2021_gyp1 <- saltmarsh_canabio_landuse_2021 %>%
  filter(DESCRIPCIO == "VEGETACIÓN SECUNDARIA ARBUSTIVA DE PASTIZAL GIPSÓFILO")

saltmarsh_canabio_2021_gyp2 <- saltmarsh_canabio_landuse_2021 %>%
  filter(DESCRIPCIO == "VEGETACIÓN GIPSÓFILA")

saltmarsh_canabio_2021_gyp3 <- saltmarsh_canabio_landuse_2021 %>%
  filter(DESCRIPCIO == "PASTIZAL GIPSÓFILO")

```

```{r}
library(sf)

gyp_all <- rbind(
  saltmarsh_canabio_2021_gyp1,
  saltmarsh_canabio_2021_gyp2,
  saltmarsh_canabio_2021_gyp3
)

# Compute convex hull
gyp_hull <- st_convex_hull(st_union(gyp_all))

# Compute minimum bounding circle
gyp_circle <- lwgeom::st_minimum_bounding_circle(gyp_hull)


```

```{r}
tmap_mode("view")

# Layer 1
tm_shape(saltmarsh_canabio_2021_gyp1) +
  tm_polygons(
    col = "#e41a1c",        # red
    border.col = "#e41a1c",
    lwd = 1,
    alpha = 0.35,
    group = "Secundaria Arbustiva Gipsófila"
  ) +

# Layer 2
tm_shape(saltmarsh_canabio_2021_gyp2) +
  tm_polygons(
    col = "#377eb8",        # blue
    border.col = "#377eb8",
    lwd = 1,
    alpha = 0.35,
    group = "Vegetación Gipsófila"
  ) +

# Layer 3
tm_shape(saltmarsh_canabio_2021_gyp3) +
  tm_polygons(
    col = "#4daf4a",        # green
    border.col = "#4daf4a",
    lwd = 1,
    alpha = 0.35,
    group = "Pastizal Gipsófilo"
  ) +

# ⭐ Extent circle
tm_shape(gyp_circle) +
  tm_borders(
    col = "yellow",
    lwd = 3,
    alpha = 0.9
  ) +
  tm_fill(
    col = "yellow",
    alpha = 0.15
  )

```

dune_canabio_2021_sublayer, saltmarsh_canabio_2021_sublayer, saltmarsh_canabio_2021_aquaculture
```{r}
target_crs <- st_crs(ohi_region)

dune_canabio_2021_sublayer        <- st_transform(dune_canabio_2021_sublayer, target_crs)
saltmarsh_canabio_2021_sublayer   <- st_transform(saltmarsh_canabio_2021_sublayer, target_crs)
saltmarsh_canabio_2021_aquaculture <- st_transform(saltmarsh_canabio_2021_aquaculture, target_crs)

# Union all three 2021 layers
dune_2021_std        <- dune_canabio_2021_sublayer        %>% select(geometry)
saltmarsh_2021_std   <- saltmarsh_canabio_2021_sublayer   %>% select(geometry)
aquaculture_2021_std <- saltmarsh_canabio_2021_aquaculture %>% select(geometry)

saltmarsh_2021_all <- rbind(
  dune_2021_std,
  saltmarsh_2021_std,
  aquaculture_2021_std
)

saltmarsh_2021_union <- st_union(saltmarsh_2021_all)

# Convert union geometry into an sf object
saltmarsh_2021_union_sf <- st_as_sf(data.frame(geometry = saltmarsh_2021_union))

saltmarsh_2021_union_area <- st_intersection(saltmarsh_2021_union_sf, ohi_region) %>%
  group_by(rgn_id) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2021
  ) %>%
  st_drop_geometry()
```

```{r}
saltmarsh_vs_hem_5km <- saltmarsh_2021_union_area %>%
  left_join(
    hem_saltmarsh_5km_area %>% 
      rename(hem_area_km2 = area_km2),
    by = "rgn_id"
  ) %>%
  mutate(
    diff_km2 = area_km2 - hem_area_km2
  )

saltmarsh_vs_hem_5km %>%
  select(
    rgn_id,
    conabio_area_km2 = area_km2,
    hem_area_km2,
    diff_km2
  ) %>%
  kable(
    digits = 2,
    caption = "CONABIO 2021 Union vs Hem Saltmarsh (5 km Buffer) Area per OHI Region"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "bordered", "condensed"),
    full_width = FALSE
  )


```

```{r}
tmap_mode("view")

tm_basemap("Esri.WorldImagery") +

# CONABIO 2021 union layer
tm_shape(saltmarsh_2021_union_sf) +
  tm_polygons(
    col = "#e41a1c",        # red
    border.col = "#e41a1c",
    alpha = 0.4,
    lwd = 1.2,
    group = "CONABIO 2021 Union"
  ) +

# Hem saltmarsh (no mangrove)
tm_shape(hem_saltmarsh_no_mangrove) +
  tm_polygons(
    col = "#377eb8",        # blue
    border.col = "#377eb8",
    alpha = 0.4,
    lwd = 1.2,
    group = "Hem Saltmarsh (No Mangrove)"
  )
```


```{r}
# Match CRS of Hem and OHI + 5 km buffer
target_crs <- st_crs(hem_saltmarsh_no_mangrove)

ohi_region <- st_transform(ohi_regions, target_crs)
ohi_marine_plus_5km <- st_transform(ohi_marine_plus_5km, target_crs)

hem_saltmarsh_no_mangrove <- st_transform(hem_saltmarsh_no_mangrove, target_crs)

hem_saltmarsh_5km <- st_intersection(hem_saltmarsh_no_mangrove, ohi_marine_plus_5km)

hem_saltmarsh_5km_area <- st_intersection(hem_saltmarsh_5km, ohi_region) %>%
  group_by(rgn_id) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2021
  ) %>%
  st_drop_geometry()
hem_saltmarsh_5km_area
```

