################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```

## clean_names_lowercase function
```{r Load clean_names_lowercase function}
# Takes text and standardizes it, which we will need for merging later on:

clean_names_lowercase <- function(x) {

  x <- iconv(x, from = "", to = "UTF-8", sub = "")
  x <- stringi::stri_trans_general(x, "Latin-ASCII")

  x %>%
    tolower() %>%
    trimws()
}
```

## lowercase_function function
```{r lowercase_function}
# Turns all entries lower-case
lowercase_function<- function(x){
  if(is.character(x)){
    tolower(x)
  } else if (is.factor(x)){
    tolower(as.character(x))
  } else {
    x
  }
}
```


# PICK 1: RAM Stock Assessment - species level ------------------------------

# >> SKIP: I have already downloaded and gapfilled the RAM data through 2024.
```{r Load in RAM and formal assessment data}
RAM_data0<- read.csv("/home/frazier/ohi_goc_prep/FP/FIS/v2025/data/formal_assessments.csv")
table(RAM_data0$year)
# 195 obs., 17 vars.
```

```{r Separate Genus from scientific_name}
RAM_data<- RAM_data0

RAM_data <- RAM_data %>%
  separate(scientificname, into = c("Genus", "Species"), sep = " ", remove = FALSE)
# Okay, looks good!
# 195 obs., 19 vars.
```

```{r Basic filtering & correcting}
# Keep Stock (genus-species in this case) only if it has post-2019 data
RAM_data <- RAM_data %>%
  group_by(Genus, Species) %>%
  filter(any(year >= 2019)) %>%
  select(year, assessyear, areaname, 
         commonname, scientificname, Genus, Species, 
         ram_bmsy, source) %>%
  ungroup()
# 181 obs., 9 vars.

# Correct the non_numeric values:
RAM_data$RAM_BBmsy_numeric <- as.numeric(gsub("[^0-9.]", "", RAM_data$ram_bmsy))
```

```{r Gapfill RAM data using the zoo package}
RAM_gapfilled <- RAM_data %>%
  group_by(Genus, Species, areaname) %>%
  complete(year = 2019:2024) %>%
  ungroup()
# 256 obs., 10 vars.

# Gapfilling - INTERPOLATION between years with data (using the zoo package):
RAM_gapfilled <- RAM_gapfilled %>%
  # --- identify each stock correctly ---
  group_by(Genus, Species, areaname) %>%
  
  # --- add missing years WITHOUT removing any original rows ---
  complete(year = full_seq(c(year, 2019:2024), 1)) %>%
  # ^ This safely adds 2019–2024 and any intermediate years, 
  #   never removes rows
  
  # --- fill descriptors (stock metadata) ---
  fill(commonname, scientificname, source, assessyear, .direction = "downup") %>%
  
  arrange(year, .by_group = TRUE) %>%
  
  # --- interpolate only RAM_BBmsy_numeric ---
  mutate(
    RAM_gapfilled = zoo::na.approx(RAM_BBmsy_numeric, x = year, na.rm = FALSE)
  ) %>%
  
  ungroup()
# 347 obs., 11 vars.

# Gapfilling - EXTRAPOLATION beyond years with data (using the zoo package):
RAM_gapfilled <- RAM_gapfilled %>%
  group_by(Genus, Species, areaname) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(
    RAM_gapfilled = zoo::na.approx(
      RAM_BBmsy_numeric,
      x = year,
      na.rm = FALSE,
      rule = 2   # allow linear extrapolation outside observed years
    )
  ) %>%
  ungroup()
# 347 obs., 11 vars.

# Sanity check: do gapfilled values always match numeric values when they exist?
# Example vectors
numeric_BBmsy_values<- RAM_gapfilled$RAM_BBmsy_numeric
gapfilled_BBmsy_values <- RAM_gapfilled$RAM_gapfilled

# Logical check: TRUE wherever BBmsy_numeric is NA or matches RAM_gapfilled
check <- ifelse(!is.na(numeric_BBmsy_values), gapfilled_BBmsy_values == numeric_BBmsy_values, TRUE)

# Are all original values preserved?
all(check)
# Yes!
```

```{r Annotate RAM_gapfilled dataframe}
# Finally, a "RAM_BBmsy_Pick" column and "Pick_Source" column that tell which BBmsy value to work with and where that data point came from, a formal assessment value or a gapfilled value:
RAM_gapfilled <- RAM_gapfilled %>%
  mutate(
    RAM_BBmsy_Pick = ifelse(!is.na(RAM_BBmsy_numeric), RAM_BBmsy_numeric, RAM_gapfilled),
    Pick_Source    = ifelse(!is.na(RAM_BBmsy_numeric), "Direct from formal assessment (RAM or other)", "gapfilled from formal assessment")
  )
# 347 obs., 13 vars.
```

```{r Save the RAM_gapfilled file}
# PENDING... cannot recreate as of June 1. Must find missing formal_assessments file.
```

```{r BBmsy cap at 1}
colnames(RAM_gapfilled)

# Caps values for RAM_BBmsy_Pick at 1 & filters out 2025 data:
RAM_FIS_scores<- RAM_gapfilled %>%
  filter(Year <= 2024) %>%
  mutate(Formal_BBmsy_capped = pmin(Formal_BBmsy_Pick, 1)) %>%
  select(Year, Area, CommonName,
         ScientificName= scientificname, Genus, Species,
         FormalAssessment_Source = source, FormalAssessment_Years= assessyear,
         Formal_BBmsy= ram_bmsy, Formal_BBmsy_numeric= RAM_BBmsy_numeric,
         Formal_BBmsy_gapfilled = RAM_gapfilled,
         Formal_BBmsy_Pick = RAM_BBmsy_Pick, Formal_BBmsy_capped,
         Formal_BBmsy_Pick_Source = Pick_Source)
# 346 obs., 14 vars.
```

# >> SKIP TO HERE: Convert RAM data into a usable BBMSY score.
```{r Load in RAM_gapfilled0 as needed}
RAM_gapfilled0<- read.csv("/home/frazier/ohi_goc_prep/FP/FIS_Scores/FIS_Scores_Dec2026/FIS_StockStatus_Scores_Dec2025/RAM_Scores/Gapfilled_RAMdata_Dec122025.csv")
# 346 obs., 14 vars.
```

```{r Convert RAM_gapfilled0 to RAM_FIS_scores}
RAM_FIS_scores<- RAM_gapfilled0
# 346 obs., 14 vars.
```

# >> FINAL STEP: SAVE the RAM_FIS_scores dataframe:
```{r Save this dataframe}
write.csv(RAM_FIS_scores,"~/ohi_goc_prep/FP/FIS_Scores/FIS_RevisedScores_May2026/FIS_StockStatus_Scores_May2026/FIS_RAMScores_June012026.csv", row.names = FALSE)
```


# PICK 2: Catch vs CNP Baseline - species level ------------------------------

## STEP A: Load in recommendations for the Carta Nacional Pesquera
```{r Load in Carta data}
Carta_data0<- read.csv("/home/shares/ohi/OHI_GOC/goal_prep/fp/fis/v2025/int/ohi_core_team/CartaNacional_Data_AllYears_Mar2026.csv")

dim(Carta_data0)
# 2699 obs., 12 vars.
```

```{r Separate Genus from scientific_name}
Carta_data <- Carta_data0 %>%
  separate(scientific_name, into = c("estimated_genus_name", "species"), sep = " ", remove = FALSE)
```

```{r Clean Carta_data as needed}
table(Carta_data$estimated_genus_name)
# Okay, looks good! Except for Pseudobatos (currently labeled Rhinobatos and R.)

Carta_data <- Carta_data %>%
  mutate(
    notes = if_else(
      estimated_genus_name %in% c("Rhinobatos", "R."),
      "Assigned new estimated_genus_name, Pseudobatos",
      notes
    ),
    
    estimated_genus_name = if_else(
      estimated_genus_name %in% c("Rhinobatos", "R."),
      "Pseudobatos",
      estimated_genus_name
    )
  )

# Make columns clean and lowercase for safe merging:
Carta_data<- Carta_data %>%
  mutate(
    estimated_species_name= clean_names_lowercase(scientific_name),
    estimated_genus_name= clean_names_lowercase(estimated_genus_name)
  )
```

```{r Assigning Carta major groups based on new key}
FAO_FamilyGenus.Key<- read.csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/FAO_FamilyGenusKey_COMPLETE_Apr042026.csv")
# 326 obs., 4 vars.

FAO_FamilyGenus.Key<- FAO_FamilyGenus.Key %>% 
  select(estimated_genus_name, estimated_taxa_group, Major_Functional_Group) %>%
  distinct()
# 193 obs., 3 vars.

# Make columns clean and lowercase for safe merging:
FAO_FamilyGenus.Key<- FAO_FamilyGenus.Key %>%
  mutate(
    estimated_genus_name= clean_names_lowercase(estimated_genus_name)
  )

# Merge family details to Carta_data:
Carta_data <- Carta_data %>%
  left_join(
    FAO_FamilyGenus.Key,
    by = c("estimated_genus_name")
  )
# 2699 obs., 17 vars. (Good, as expected!)
```

## STEP B: Create the Carta_RecommendationsKey based on the data.

```{r Make a Carta_RecommendationsKey0}
colnames(Carta_data)

Carta_RecommendationsKey0<- Carta_data %>% 
  select(Year_of_Recommendation= carta_year, Major_Functional_Group, 
         estimated_species_name, estimated_genus_name, common_name, 
         OHI_region, 
         status, control, notes)
# 2699 obs., 9 vars.
```

```{r Make key wide so we can plot these data}
# I want to take this dataframe and make a new one where the recommendation per taxa is spread out by years so that I can easily how recommendations change over time per taxa...

Carta_ControlKey_wide <- Carta_RecommendationsKey0 %>%
  select(
    Year_of_Recommendation,
    estimated_genus_name,
    common_name,
    OHI_region,
    status,
    control
  ) %>%
  pivot_wider(
    names_from = Year_of_Recommendation,
    values_from = c(status, control),
    names_glue = "{.value}_{Year_of_Recommendation}"
  )

# Plot...
```

```{r Create Carta_RecommendationsKey - Part 1}
colnames(Carta_RecommendationsKey0)

# Filter this down to unique number of species (disregarding common_name)
Carta_RecommendationsKey<- Carta_RecommendationsKey0 %>% 
  select(Year_of_Recommendation, Major_Functional_Group, 
         estimated_species_name, estimated_genus_name, 
         OHI_region, 
         status, control, notes) %>%
  distinct()
# 2692 obs., 8 vars.
# Suggests there were 7 duplicates.

# SANITY CHECK: Why is this key shorter from Carta_RecommendationsKey0?
colnames(Carta_RecommendationsKey)

common_cols <- intersect(names(Carta_RecommendationsKey0),
                         names(Carta_RecommendationsKey))

TEST<- anti_join(
  Carta_RecommendationsKey0 %>% select(-common_name),
  Carta_RecommendationsKey,
  by = names(Carta_RecommendationsKey)
)
# Multiple recommendations for distinct species under different common names: common_name
```

```{r Create Carta_RecommendationsKey - Part 2}
colnames(Carta_RecommendationsKey)

table(Carta_RecommendationsKey$status)
  #                                decreasing         deteriorating development potential 
  #                1134                     4                    44                    26 
  # maximally harvested               unclear 
  #                 936                   112 
table(Carta_RecommendationsKey$control)
       #         decrease    increase  increasing no increase 
       # 1134          44          40           4         918


# Generate outputs regarding status and control at the species level:
Carta_RecommendationsKey <- Carta_RecommendationsKey %>%
  
  group_by(
    estimated_species_name,
    OHI_region,
    Year_of_Recommendation
  ) %>%
  
  summarise(

# STATUS:
    
    status = {

      valid_status <- status[!is.na(status)]

      if (length(valid_status) == 0) {
        
        NA_character_

      } else if ("deteriorating" %in% valid_status) {

        "deteriorating"

      } else if ("improving" %in% valid_status) {

        "improving"

      } else {

        valid_status[1]
      }
    },


# CONTROL:

    control = {

      valid_control <- control[!is.na(control)]

      if (length(valid_control) == 0) {

        NA_character_

      } else {

        valid_control[1]
      }
    },


# NOTES:

    notes = {

      valid_notes <- notes[
        !is.na(notes) & notes != ""
      ]

      if (length(valid_notes) == 0) {

        NA_character_

      } else {

        valid_notes[1]
      }
    },

    .groups = "drop"
  )
# 2692 obs.,  6 vars.
```

```{r Remove any rows where the control and recommendation are both NA or "unknown"}
# Remove any rows where the control and recommendation are both NA or "unknown"
table(Carta_RecommendationsKey$status) # 1134 NAs
table(Carta_RecommendationsKey$control) # 1134 NAs

Carta_RecommendationsKey <- Carta_RecommendationsKey %>%
  filter(
    !(
      (is.na(status) | status == "unclear") &
      (is.na(control) | control == "unclear")
    )
  )
# 2144 obs., 6 vars.
```

## STEP C: SAVE the Carta_RecommendationsKey.
```{r SAVE the Carta_RecommendationsKey as necessary}
write.csv(Carta_RecommendationsKey,
          file="/home/ferrer/ohi_goc_prep/FP/FIS_Scores/FIS_RevisedScores_May2026/FIS_StockStatus_Scores_May2026/CartaRecommendationsKey_May282026.csv",
          row.names = FALSE)
# 2144 obs., 6 vars.
```

## STEP D: Scoring based on the Carta: compare landings over time (GOC_MarineLandings) to recommendations in the Carta_RecommendationsKey
```{r Load in Carta_RecommendationsKey as needed}
Carta_RecommendationsKey<- read.csv("/home/ferrer/ohi_goc_prep/FP/FIS_Scores/FIS_RevisedScores_May2026/FIS_StockStatus_Scores_May2026/CartaRecommendationsKey_May282026.csv")
# 2144 obs., 6 vars.
```

```{r Load in GOC_MarineLandings_carta as needed}
GOC_MarineLandings0<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Marine.Landings_Sector.Sorted.Georef_2000.present_May262026.rds"))
# 3246583 obs., 57 vars.
```

```{r Filter and clean GOC_MarineLandings_carta}
GOC_MarineLandings_carta<- GOC_MarineLandings0
# 3246583 obs., 57 vars.

table(GOC_MarineLandings_carta$sector_estimate_draft)

GOC_MarineLandings_carta<- GOC_MarineLandings_carta %>%
  select(state, office_name_clean, landing_site_clean,
         year, month,
         Assignment_for_Major_Functional_Group, Major_Functional_Group,
         recorded_taxa_group, estimated_taxa_group, recorded_taxa_name, recorded_taxa_name_clean,
         estimated_species_name, estimated_genus_name, estimated_family_name,
         landed_weight, live_weight, weight_kg_corrected, value,
         sector_estimate_draft,
         GOC_est, GOC_Include_Exclude, OHI_region_est, OHI_region_name
         )
# 3246583 obs., 23 vars.
```

```{r Summarize these data by Year Region and estimated_species_name with a 2018-2022 reference period}
colnames(GOC_MarineLandings_carta)

GOC_MarineLandings_carta.summary <- GOC_MarineLandings_carta %>%
  group_by(year, OHI_region_est, estimated_species_name) %>%
  summarize(
    annual_Catch_kg = sum(weight_kg_corrected, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(OHI_region_est, estimated_species_name) %>%
  mutate(
    mean_Catch_2018.22_kg = mean(
      annual_Catch_kg[year %in% 2018:2022],
      na.rm = TRUE
    ),
    n.years_Catch_2018.22 = sum(year %in% 2018:2022)
  ) %>%
  ungroup() %>%
  arrange(OHI_region_est, estimated_species_name, year)
# 16802 obs., 6 vars.
```

```{r Prep GOC_MarineLandings_carta.summary and Carta_RecommendationsKey for merge}
# Make everything lowercase in preparation for merge:
colnames(GOC_MarineLandings_carta.summary)

GOC_MarineLandings_carta.summary <- GOC_MarineLandings_carta.summary %>%
  mutate(
    OHI_region_est = clean_names_lowercase(OHI_region_est),
    estimated_species_name = clean_names_lowercase(estimated_species_name)
  )
# 16802 obs., 6 vars.

colnames(Carta_RecommendationsKey)

Carta_RecommendationsKey <- Carta_RecommendationsKey %>%
  mutate(
    OHI_region_est = clean_names_lowercase(OHI_region),
    estimated_species_name = clean_names_lowercase(estimated_species_name)
  ) %>%
  select(-c(OHI_region))
# 16802 obs., 6 vars.
```

```{r Merge in Carta_RecommendationsKey data}
# Merge Carta_RecommendationsKey to GOC_MarineLandings_carta.summary:
Carta_Scores0 <- GOC_MarineLandings_carta.summary %>%
  left_join(
    Carta_RecommendationsKey,
    by = c("estimated_species_name", "OHI_region_est")
  ) %>%
  arrange(OHI_region_est, estimated_species_name, year)
# 16802 obs., 10 vars. (As expected, good!)
```

```{r Calculate Stock_Status scores based on C_current versus C_ref - Part 1}
# But first, check what combinations there are in the Carta_RecommendationsKey between stock "status" and "control" recommendation. We're looking for weird logical mismatches such as stock is doing poorly but control says keep fishing:
Carta_Scores0 %>%
  count(status, control, name = "n")
# There are 6 possible combinations:
# - decreasing,	increase	4		
# - deteriorating,	decrease	171		
# - development potential,	increase	50		
# - maximally harvested,	increase	100		
# - maximally harvested,	no increase	4114		
# - NA	NA	9567	

# Important adjustment:
# - In scenarios where the fishery status was “maximally harvested”, we assumed that the control was “no increase” even when the original Carta data file said otherwise. Double check this with Austin. 
# - In scenarios where the fishery status was “decreasing”, we assumed that the control was “decrease [pressure]” even when the original Carta data file said otherwise. Double check this with Austin.

# Quick re-assignment:
Carta_Scores <- Carta_Scores0 %>%
  mutate(
    control = case_when(
      status == "maximally harvested" ~ "no increase",
      status == "decreasing" ~ "decrease",
      TRUE ~ control
    )
  )
```

```{r Calculate Stock_Status scores based on C_current versus C_ref - Part 2}
# How similar is C_current versus C_ref? If C_current is within 10% of C_ref, then functionally equal.
tolerance <- 0.10 

Carta_Scores <- Carta_Scores %>%
  mutate(
    approx_equal = abs(annual_Catch_kg - mean_Catch_2018.22_kg) <=
      tolerance * mean_Catch_2018.22_kg,

    Stock_Status = case_when(

      # ── Scenario 1: Maximally harvested ─────────────────────────
      status == "maximally harvested" &
        control == "no increase" &
        approx_equal ~
        1,

      status == "maximally harvested" &
        control == "no increase" &
        annual_Catch_kg > mean_Catch_2018.22_kg ~
        mean_Catch_2018.22_kg / annual_Catch_kg,

      status == "maximally harvested" &
        control == "no increase" &
        annual_Catch_kg < mean_Catch_2018.22_kg ~
        1,

      # ── Scenario 2: Stock deteriorating ─────────────────────────
      status == "deteriorating" &
        control %in% c("decrease", "no increase") &
        approx_equal ~
        0.5,

      status == "deteriorating" &
        control %in% c("decrease", "no increase") &
        annual_Catch_kg > mean_Catch_2018.22_kg ~
        (mean_Catch_2018.22_kg / annual_Catch_kg) * 0.5,

      status == "deteriorating" &
        control %in% c("decrease", "no increase") &
        annual_Catch_kg < mean_Catch_2018.22_kg ~
        0.5 + 0.5 * (
          (mean_Catch_2018.22_kg - annual_Catch_kg) /
          mean_Catch_2018.22_kg
        ),

      # ── Scenario 3: Development potential ───────────────────────
      status == "development potential" ~
        1,

      TRUE ~ NA_real_
    )
  ) %>%
  select(-approx_equal)
# 16802 obs., 11 vars.
```

```{r Filter Carta_FIS_scores}
# Filter out years before 2018 and 2025
Carta_FIS_scores<- Carta_Scores %>% 
  filter(year >= 2018 & year <= 2024)
# 5800 obs., 11 vars.

# Now filter only those genera that have some actual scores:
Carta_FIS_scores<- Carta_FIS_scores %>%
  filter(!is.na(Stock_Status))
# 1507 obs., 11 vars.

unique(Carta_FIS_scores$estimated_species_name)
# 66 species

mean(Carta_FIS_scores$Stock_Status, na.rm=TRUE)
# 0.8289082

median(Carta_FIS_scores$Stock_Status, na.rm=TRUE)
# 1
```

```{r Filter stocks for which the number of reference years is less than 4}
# Finally, filter stocks for which the number of reference years is less than 4:
colnames(Carta_FIS_scores)

Carta_FIS_scores<- Carta_FIS_scores %>%
  filter(n.years_Catch_2018.22 >= 4)
# 1142 obs. (-365 obs.), 11 vars.

mean(Carta_FIS_scores$Stock_Status, na.rm=TRUE)
# 0.8440899

median(Carta_FIS_scores$Stock_Status, na.rm=TRUE)
# 1
```


## >> FINAL STEP: SAVE the Carta_FIS_scores dataframe:
```{r Write this as a CSV}
write.csv(Carta_FIS_scores,
          file= "/home/ferrer/ohi_goc_prep/FP/FIS_Scores/FIS_RevisedScores_May2026/FIS_StockStatus_Scores_May2026/FIS_CartaScores_May282026.csv", 
          row.names = FALSE)
```


# PICK 3: CMSY Stock Assesments - genus level ------------------------------
- This task represents a considerable amount of work that was not completed in this file.
- To learn how the CMSY scores were generated, review this file:
CMSY_FinalAnalysis_May2026.Rmd.
- I used this file to generate the model priors.
- However, the actual CMSY script (RunCMSY_May282026.R) was run on my personal machine.

Important Notes: 
- I ran CMSY at genus level, since I actually think it’s a safer more reliable bet than species-level info and because – in many cases – fishermen seem to target taxa of the same genus interchangeably.

# >> SKIP: I have already run the CMSY through 2024.

FILE 1: Create a file with annual landings in kg (and n_count observations) grouped by year per "stock". Here, stock is defined as the total population of GENUS X fished within the Gulf of California from Baja to northern Jalisco, not including landings from the Pacific.

FILE 2: Create a file with OHI_Region, Stock, Group = Name, English Name, Scientific Name, Source, and a whole bunch of year columns that you populate using the various R files shared by Alfredo. This file can probably be trimmed down but for now I recommend formatting it exactly as it appears from before.


```{r Load in CMSY asessment results}
# Actual B/Bmsy output from CMSY:
CMSY_Timeseries0<- read.csv("/home/ferrer/ohi_goc_prep/FP/CMSY_Analysis/CMSY_FinalLook_May2026/Model_Output_Files/")
# 900 obs., 14 vars.
```

```{r Load other useful dataframes: CMSY timeseries and key data}
# Useful genus key:
CMSY_GenusKey<- read.csv("~/ohi_goc_prep/FP/FIS/v2025/OHI_StockStatus_Scores/CMSY_FIS_scores/Instructional_Files/CMSY_GenusKey_Dec102025.csv")
# 98 obs., 3 vars.
```

```{r Append info about Major_Functional_Group}
CMSY_Timeseries<- CMSY_Timeseries0 %>% select(stock, yr, 
                                             B, B.Bmsy, lcl.B.Bmsy, ucl.B.Bmsy,
                                             F.Fmsy, lcl.F.Fmsy, ucl.F.Fmsy)

# Convert the relevant columns to lower case
CMSY_Timeseries <- CMSY_Timeseries %>%
  mutate(stock = tolower(stock))

CMSY_GenusKey <- CMSY_GenusKey %>%
  mutate(Stock = tolower(Stock))

# Merge in Major_Functional_Group data:
CMSY_Timeseries <- CMSY_Timeseries %>%
  left_join(
    CMSY_GenusKey,
    by = c("stock" = "Stock")
  )
# 775 obs., 11 vars.
```

```{r Summarize landings by genus / stock}
Genera_in_CMSY<- unique(CMSY_Timeseries$stock)
# 31 assessed genera

CMSY_LandingsData <- CMSY_LandingsData_original %>%
  mutate(estimated_genus_name = tolower(estimated_genus_name))

CMSY_LandingsData <- CMSY_LandingsData %>% filter(estimated_genus_name %in% Genera_in_CMSY)
# 1126147 obs., 53 vars
unique(CMSY_LandingsData$estimated_genus_name)
# Yup! Looks good!

# Summarize by year and OHI region:
CMSYstocks_RegionalLandings_Summary <- CMSY_LandingsData %>% 
  filter(year >= 2014, year <= 2024) %>%
  select(year, OHI_Region_No, estimated_genus_name, weight_kg_corrected, value) %>%
  group_by(stock = estimated_genus_name, OHI_Region_No, yr = year) %>%
  summarize(
    total_catch_kg = sum(weight_kg_corrected, na.rm = TRUE),
    total_value_pesos = sum(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(stock, OHI_Region_No, yr)
# 2171 obs., 5 vars.

CMSYstocks_RegionalLandings_Summary <- CMSYstocks_RegionalLandings_Summary %>%
  left_join(
    CMSY_Timeseries,
    by = c("yr", "stock")
  )
# 2171 obs., 14 obs.
```

```{r Now filter to create the final dataframe}
colnames(CMSYstocks_RegionalLandings_Summary)

CMSY_FIS_scores<- CMSYstocks_RegionalLandings_Summary %>% select(yr, OHI_Region_No, stock,
                                                             total_catch_kg, B, B.Bmsy, lcl.B.Bmsy,
                                                             ucl.B.Bmsy,
                                                             F.Fmsy, lcl.F.Fmsy, ucl.F.Fmsy)
```

# FINAL STEP: SAVE the CMSY_FIS_scores dataframe:
```{r Save this dataframe}
write.csv(CMSY_FIS_scores,
          file= "/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_StockStatus_Scores/CMSY_FIS_scores/FIS_CMSY.BBmsyScores_Dec162025.csv"
          row.names = FALSE)
```