################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```

## clean_names_lowercase function
```{r Load clean_names_lowercase function}
# Takes text and standardizes it, which we will need for merging later on:

clean_names_lowercase <- function(x) {

  x <- iconv(x, from = "", to = "UTF-8", sub = "")
  x <- stringi::stri_trans_general(x, "Latin-ASCII")

  x %>%
    tolower() %>%
    trimws()
}
```

## lowercase_function function
```{r lowercase_function}
# Turns all entries lower-case
lowercase_function<- function(x){
  if(is.character(x)){
    tolower(x)
  } else if (is.factor(x)){
    tolower(as.character(x))
  } else {
    x
  }
}
```


# PICK 2: Catch vs CNP Baseline - species level ------------------------------

## STEP A: Load in recommendations for the Carta Nacional Pesquera
```{r Load in Carta data}
Carta_data0<- read.csv("/home/shares/ohi/OHI_GOC/goal_prep/fp/fis/v2025/int/ohi_core_team/CartaNacional_Data_AllYears_Mar2026.csv")

dim(Carta_data0)
# 2699 obs., 12 vars.
```

```{r Separate Genus from scientific_name}
Carta_data <- Carta_data0 %>%
  separate(scientific_name, into = c("estimated_genus_name", "species"), sep = " ", remove = FALSE)
```

```{r Clean Carta_data as needed}
table(Carta_data$estimated_genus_name)
# Okay, looks good! Except for Pseudobatos (currently labeled Rhinobatos and R.)

Carta_data <- Carta_data %>%
  mutate(
    notes = if_else(
      estimated_genus_name %in% c("Rhinobatos", "R."),
      "Assigned new estimated_genus_name, Pseudobatos",
      notes
    ),
    
    estimated_genus_name = if_else(
      estimated_genus_name %in% c("Rhinobatos", "R."),
      "Pseudobatos",
      estimated_genus_name
    )
  )

# Make columns clean and lowercase for safe merging:
Carta_data<- Carta_data %>%
  mutate(
    estimated_species_name= clean_names_lowercase(scientific_name),
    estimated_genus_name= clean_names_lowercase(estimated_genus_name)
  )
```

```{r Assigning Carta major groups based on new key}
FAO_FamilyGenus.Key<- read.csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/FAO_FamilyGenusKey_COMPLETE_Apr042026.csv")
# 326 obs., 4 vars.

FAO_FamilyGenus.Key<- FAO_FamilyGenus.Key %>% 
  select(estimated_genus_name, estimated_taxa_group, Major_Functional_Group) %>%
  distinct()
# 193 obs., 3 vars.

# Make columns clean and lowercase for safe merging:
FAO_FamilyGenus.Key<- FAO_FamilyGenus.Key %>%
  mutate(
    estimated_genus_name= clean_names_lowercase(estimated_genus_name)
  )

# Merge family details to Carta_data:
Carta_data <- Carta_data %>%
  left_join(
    FAO_FamilyGenus.Key,
    by = c("estimated_genus_name")
  )
# 2699 obs., 17 vars. (Good, as expected!)
```

## STEP B: Create the Carta_RecommendationsKey based on the data.

```{r Make a Carta_RecommendationsKey0}
colnames(Carta_data)

Carta_RecommendationsKey0<- Carta_data %>% 
  select(Year_of_Recommendation= carta_year, Major_Functional_Group, 
         estimated_species_name, estimated_genus_name, common_name, 
         OHI_region, 
         status, control, notes)
# 2699 obs., 9 vars.
```

```{r Create Carta_RecommendationsKey - Part 1}
colnames(Carta_RecommendationsKey0)

# Filter this down to unique number of species (disregarding common_name)
Carta_RecommendationsKey<- Carta_RecommendationsKey0 %>% 
  select(Year_of_Recommendation, Major_Functional_Group, 
         estimated_species_name, estimated_genus_name, 
         OHI_region, 
         status, control, notes) %>%
  distinct()
# 2692 obs., 8 vars.
# Suggests there were 7 duplicates.

# SANITY CHECK: Why is this key shorter from Carta_RecommendationsKey0?
colnames(Carta_RecommendationsKey)

common_cols <- intersect(names(Carta_RecommendationsKey0),
                         names(Carta_RecommendationsKey))

TEST<- anti_join(
  Carta_RecommendationsKey0 %>% select(-common_name),
  Carta_RecommendationsKey,
  by = names(Carta_RecommendationsKey)
)
# Multiple recommendations for distinct species under different common names: common_name
```

```{r Create Carta_RecommendationsKey - Part 2}
colnames(Carta_RecommendationsKey)

table(Carta_RecommendationsKey$status)
  #                                decreasing         deteriorating development potential 
  #                1134                     4                    44                    26 
  # maximally harvested               unclear 
  #                 936                   112 
table(Carta_RecommendationsKey$control)
       #         decrease    increase  increasing no increase 
       # 1134          44          40           4         918


# Generate outputs regarding status and control at the species level:
Carta_RecommendationsKey <- Carta_RecommendationsKey %>%
  
  group_by(
    estimated_species_name,
    OHI_region,
    Year_of_Recommendation
  ) %>%
  
  summarise(

# STATUS:
    
    status = {

      valid_status <- status[!is.na(status)]

      if (length(valid_status) == 0) {
        
        NA_character_

      } else if ("deteriorating" %in% valid_status) {

        "deteriorating"

      } else if ("improving" %in% valid_status) {

        "improving"

      } else {

        valid_status[1]
      }
    },


# CONTROL:

    control = {

      valid_control <- control[!is.na(control)]

      if (length(valid_control) == 0) {

        NA_character_

      } else {

        valid_control[1]
      }
    },


# NOTES:

    notes = {

      valid_notes <- notes[
        !is.na(notes) & notes != ""
      ]

      if (length(valid_notes) == 0) {

        NA_character_

      } else {

        valid_notes[1]
      }
    },

    .groups = "drop"
  )
# 2692 obs.,  6 vars.
```

```{r Remove any rows where the control and recommendation are both NA or "unknown"}
# Remove any rows where the control and recommendation are both NA or "unknown"
table(Carta_RecommendationsKey$status) # 1134 NAs
table(Carta_RecommendationsKey$control) # 1134 NAs

Carta_RecommendationsKey <- Carta_RecommendationsKey %>%
  filter(
    !(
      (is.na(status) | status == "unclear" | status== "") &
      (is.na(control) | control == "unclear" | control == "")
    )
  )
# 1010 obs., 6 vars.
```

```{r Make sure there is a recommendation for every region even if assumed}
# Asking Cursor for help with this chunk of code.
# 1010 obs., 6 vars.
```

## STEP C: SAVE the Carta_RecommendationsKey.
```{r SAVE the Carta_RecommendationsKey as necessary}
write.csv(Carta_RecommendationsKey,
          file="/home/ferrer/ohi_goc_prep/FP/FIS_Scores/FIS_RevisedScores_May2026/FIS_StockStatus_Scores_May2026/CartaRecommendationsKey_May282026.csv",
          row.names = FALSE)
# 2144 obs., 6 vars.
```

```{r Make key wide so we can plot these data}
# I want to take this dataframe and make a new one where the recommendation per taxa is spread out by years so that I can easily how recommendations change over time per taxa...

Carta_ControlKey_wide <- Carta_RecommendationsKey0 %>%
  select(
    Year_of_Recommendation,
    estimated_genus_name,
    common_name,
    OHI_region,
    status,
    control
  ) %>%
  pivot_wider(
    names_from = Year_of_Recommendation,
    values_from = c(status, control),
    names_glue = "{.value}_{Year_of_Recommendation}"
  )

# Plot...
```

## STEP D: Scoring based on the Carta: compare landings over time (GOC_MarineLandings) to recommendations in the Carta_RecommendationsKey
```{r Load in Carta_RecommendationsKey as needed}
Carta_RecommendationsKey<- read.csv("/home/ferrer/ohi_goc_prep/FP/FIS_Scores/FIS_RevisedScores_May2026/FIS_StockStatus_Scores_May2026/CartaRecommendationsKey_May282026.csv")
# 2144 obs., 6 vars.
```

```{r Load in GOC_MarineLandings_carta as needed}
GOC_MarineLandings0<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Marine.Landings_Sector.Sorted.Georef_2000.present_May262026.rds"))
# 3246583 obs., 57 vars.
```

```{r Filter and clean GOC_MarineLandings_carta}
GOC_MarineLandings_carta<- GOC_MarineLandings0
# 3246583 obs., 57 vars.

table(GOC_MarineLandings_carta$sector_estimate_draft)

GOC_MarineLandings_carta<- GOC_MarineLandings_carta %>%
  select(state, office_name_clean, landing_site_clean,
         year, month,
         Assignment_for_Major_Functional_Group, Major_Functional_Group,
         recorded_taxa_group, estimated_taxa_group, recorded_taxa_name, recorded_taxa_name_clean,
         estimated_species_name, estimated_genus_name, estimated_family_name,
         landed_weight, live_weight, weight_kg_corrected, value,
         sector_estimate_draft,
         GOC_est, GOC_Include_Exclude, OHI_region_est, OHI_region_name
         )
# 3246583 obs., 23 vars.
```

```{r Summarize these data by Year Region and estimated_species_name with a 2018-2022 reference period}
colnames(GOC_MarineLandings_carta)

GOC_MarineLandings_carta.summary <- GOC_MarineLandings_carta %>%
  group_by(year, OHI_region_est, estimated_species_name) %>%
  summarize(
    annual_Catch_kg = sum(weight_kg_corrected, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(OHI_region_est, estimated_species_name) %>%
  mutate(
    mean_Catch_2018.22_kg = mean(
      annual_Catch_kg[year %in% 2018:2022],
      na.rm = TRUE
    ),
    n.years_Catch_2018.22 = sum(year %in% 2018:2022)
  ) %>%
  ungroup() %>%
  arrange(OHI_region_est, estimated_species_name, year)
# 16802 obs., 6 vars.
```

```{r Prep GOC_MarineLandings_carta.summary and Carta_RecommendationsKey for merge}
# Make everything lowercase in preparation for merge:
colnames(GOC_MarineLandings_carta.summary)

GOC_MarineLandings_carta.summary <- GOC_MarineLandings_carta.summary %>%
  mutate(
    OHI_region_est = clean_names_lowercase(OHI_region_est),
    estimated_species_name = clean_names_lowercase(estimated_species_name)
  )
# 16802 obs., 6 vars.

colnames(Carta_RecommendationsKey)

Carta_RecommendationsKey <- Carta_RecommendationsKey %>%
  mutate(
    OHI_region_est = clean_names_lowercase(OHI_region),
    estimated_species_name = clean_names_lowercase(estimated_species_name)
  ) %>%
  select(-c(OHI_region))
# 16802 obs., 6 vars.
```

```{r Merge in Carta_RecommendationsKey data}
# Merge Carta_RecommendationsKey to GOC_MarineLandings_carta.summary:
Carta_Scores0 <- GOC_MarineLandings_carta.summary %>%
  left_join(
    Carta_RecommendationsKey,
    by = c("estimated_species_name", "OHI_region_est")
  ) %>%
  arrange(OHI_region_est, estimated_species_name, year)
# 16802 obs., 10 vars. (As expected, good!)
```

```{r Calculate Stock_Status scores based on C_current versus C_ref - Part 1}
# But first, check what combinations there are in the Carta_RecommendationsKey between stock "status" and "control" recommendation. We're looking for weird logical mismatches such as stock is doing poorly but control says keep fishing:
Carta_Scores0 %>%
  count(status, control, name = "n")
# There are 6 possible combinations:
# - decreasing,	increase	4		
# - deteriorating,	decrease	171		
# - development potential,	increase	50		
# - maximally harvested,	increase	100		
# - maximally harvested,	no increase	4114		
# - NA	NA	9567	

# Important adjustment:
# - In scenarios where the fishery status was “maximally harvested”, we assumed that the control was “no increase” even when the original Carta data file said otherwise. Double check this with Austin. 
# - In scenarios where the fishery status was “decreasing”, we assumed that the control was “decrease [pressure]” even when the original Carta data file said otherwise. Double check this with Austin.

# Quick re-assignment:
Carta_Scores <- Carta_Scores0 %>%
  mutate(
    control = case_when(
      status == "maximally harvested" ~ "no increase",
      status == "decreasing" ~ "decrease",
      TRUE ~ control
    )
  )
```

```{r Calculate Stock_Status scores based on C_current versus C_ref - Part 2}
# How similar is C_current versus C_ref? If C_current is within 10% of C_ref, then functionally equal.
tolerance <- 0.10 

Carta_Scores <- Carta_Scores %>%
  mutate(
    approx_equal = abs(annual_Catch_kg - mean_Catch_2018.22_kg) <=
      tolerance * mean_Catch_2018.22_kg,

    Stock_Status = case_when(

      # ── Scenario 1: Maximally harvested ─────────────────────────
      status == "maximally harvested" &
        control == "no increase" &
        approx_equal ~
        1,

      status == "maximally harvested" &
        control == "no increase" &
        annual_Catch_kg > mean_Catch_2018.22_kg ~
        mean_Catch_2018.22_kg / annual_Catch_kg,

      status == "maximally harvested" &
        control == "no increase" &
        annual_Catch_kg < mean_Catch_2018.22_kg ~
        1,

      # ── Scenario 2: Stock deteriorating ─────────────────────────
      status == "deteriorating" &
        control %in% c("decrease", "no increase") &
        approx_equal ~
        0.5,

      status == "deteriorating" &
        control %in% c("decrease", "no increase") &
        annual_Catch_kg > mean_Catch_2018.22_kg ~
        (mean_Catch_2018.22_kg / annual_Catch_kg) * 0.5,

      status == "deteriorating" &
        control %in% c("decrease", "no increase") &
        annual_Catch_kg < mean_Catch_2018.22_kg ~
        0.5 + 0.5 * (
          (mean_Catch_2018.22_kg - annual_Catch_kg) /
          mean_Catch_2018.22_kg
        ),

      # ── Scenario 3: Development potential ───────────────────────
      status == "development potential" ~
        1,

      TRUE ~ NA_real_
    )
  ) %>%
  select(-approx_equal)
# 16802 obs., 11 vars.
```

```{r Filter Carta_FIS_scores}
# Filter out years before 2018 and 2025
Carta_FIS_scores<- Carta_Scores %>% 
  filter(year >= 2018 & year <= 2024)
# 5800 obs., 11 vars.

# Now filter only those genera that have some actual scores:
Carta_FIS_scores<- Carta_FIS_scores %>%
  filter(!is.na(Stock_Status))
# 1507 obs., 11 vars.

unique(Carta_FIS_scores$estimated_species_name)
# 66 species

mean(Carta_FIS_scores$Stock_Status, na.rm=TRUE)
# 0.8289082

median(Carta_FIS_scores$Stock_Status, na.rm=TRUE)
# 1
```

```{r Filter stocks for which the number of reference years is less than 4}
# Finally, filter stocks for which the number of reference years is less than 4:
colnames(Carta_FIS_scores)

Carta_FIS_scores<- Carta_FIS_scores %>%
  filter(n.years_Catch_2018.22 >= 4)
# 1142 obs. (-365 obs.), 11 vars.

mean(Carta_FIS_scores$Stock_Status, na.rm=TRUE)
# 0.8440899

median(Carta_FIS_scores$Stock_Status, na.rm=TRUE)
# 1
```


## >> FINAL STEP: SAVE the Carta_FIS_scores dataframe:
```{r Write this as a CSV}
write.csv(Carta_FIS_scores,
          file= "/home/ferrer/ohi_goc_prep/FP/FIS_Scores/FIS_RevisedScores_May2026/FIS_StockStatus_Scores_May2026/FIS_CartaScores_May282026.csv", 
          row.names = FALSE)
```