---
title: 'OHI GoC 2025: CW Chemicals'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# CW: Chemicals

```{r}
library(tidyverse)
library(here)
library(plotly)

rgn_colors <- c(
  Region_1 = "#C03729",
  Region_2 = "#E68C7C",
  Region_3 = "#FC8F24",
  Region_4 = "#dcaa38",
  Region_5 = "#cbd2a0",
  Region_6 = "#98a25a",
  Region_7 = "#747428",
  Region_8 = "#627391",
  Region_9 = "#8d61a9"
)
```

We now have status scores for each of these 5 subcomponents of Chemicals (a subcomponent of the Clean Waters goal):

- Pesticides (PEST) 
- Pharmaceuticals (PHRM) 
- Hazardous waste sites (HAZW) 
- Untreated wastewater discharge point sites (WWPS) 
- Vessel hydrocarbon estimation (VESS) 

::: {.callout-note collapse="true"}
## Click here to see a table of each subcomponent with more information on each.

![](https://github.com/user-attachments/assets/093565f3-cd5e-4fe8-a765-7a4a31fc5f0d){width=991 height=558}
:::

We discussed: (1) Using [this goalkeeper group activity](https://docs.google.com/document/d/1mNmaxgHCGd57YzAA0afcbNObEByE9meEl_LUy-7m4so/edit?tab=t.0#heading=h.uev2w5dkrcdm) to inform our weighting, (2) a geometric mean across all CS scores, so that the lower scores have a larger impact, and (3) computing overall Chemical status scores using an arithmetic mean across all subcomponents.

Ultimately, we decided that it would make the most sense to weight these status scores based on the strength of their consequences on both human and ocean health.  If we do not have enough evidence to support those weights, then we will use a simple arithmetic mean (no weighting). 

## Bring in all status scores

### Pesticides
```{r}
pesticides_cs <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/pesticides/output/cw_chemical_pesticide_current_status.csv")) %>% 
dplyr::select(rgn_id, year, current_status = current_status_final)

## expand
library(tidyverse)
library(zoo)

pesticides_cs_complete <- pesticides_cs %>%
  group_by(rgn_id) %>%
  complete(year = seq(min(year), 2023, by = 1)) %>% # expand all years to 2023
  mutate(current_status = zoo::na.approx(current_status, na.rm = FALSE)) %>% # interpolate linearly between known values
  mutate(current_status = zoo::na.locf(current_status, na.rm = FALSE)) %>% # carry the last known value forward (for any trailing NAs)
  ungroup()
```

```{r}
pest_plotly <- plotly::plot_ly(
  pesticides_cs_complete, 
  x = ~year, 
  y = ~current_status, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Current Status: CW Chemicals Pesticides",
    xaxis = list(title = "ano // year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines( 
    x = range(pesticides_cs_complete$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
pest_plotly
```


### Pharmaceuticals
```{r}
pharmaceuticals_cs <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/pharmaceuticals","current_status_pharmaceuticals.csv")) %>% 
dplyr::select(rgn_id, year, current_status = current_status_final)

## expand
library(tidyverse)
library(zoo)

pharmaceuticals_cs_complete <- pharmaceuticals_cs %>%
  group_by(rgn_id) %>%
  complete(year = seq(min(year), 2023, by = 1)) %>% # expand all years to 2023
  mutate(current_status = zoo::na.approx(current_status, na.rm = FALSE)) %>% # interpolate linearly between known values
  mutate(current_status = zoo::na.locf(current_status, na.rm = FALSE)) %>% # carry the last known value forward (for any trailing NAs)
  ungroup()
```

```{r}
pharm_plotly <- plotly::plot_ly(
  pharmaceuticals_cs_complete, 
  x = ~year, 
  y = ~current_status, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Current Status: CW Chemicals Pharmaceuticals",
    xaxis = list(title = "ano // year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines( 
    x = range(pharmaceuticals_cs_complete$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
pharm_plotly
```


### Untreated wastewater discharge points
```{r}
untreated_ww_cs <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/agua_sin_tratamiento/output/untreated_ww_current_status.csv")) %>% 
dplyr::select(rgn_id, year = census_year, current_status = current_status_final)

## expand
library(tidyverse)
library(zoo)

untreated_ww_cs_complete <- untreated_ww_cs %>%
  group_by(rgn_id) %>%
  complete(year = seq(min(year), 2023, by = 1)) %>% # expand all years to 2023
  mutate(current_status = zoo::na.approx(current_status, na.rm = FALSE)) %>% # interpolate linearly between known values
  mutate(current_status = zoo::na.locf(current_status, na.rm = FALSE)) %>% # carry the last known value forward (for any trailing NAs)
  ungroup()
```

```{r}
untreated_ww_plotly <- plotly::plot_ly(
  untreated_ww_cs_complete, 
  x = ~year, 
  y = ~current_status, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Current Status: CW Chemicals Untreated Wastewater",
    xaxis = list(title = "ano // year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines( 
    x = range(untreated_ww_cs_complete$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
untreated_ww_plotly
```


### Hazardous waste sites
```{r}
hazardous_waste_cs <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/hazardous_waste/output","current_status_hazardous_waste.csv")) %>% 
dplyr::select(rgn_id, year = anio, current_status = final_score)
```

```{r}
hazardous_waste_plotly <- plotly::plot_ly(
  hazardous_waste_cs, 
  x = ~year, 
  y = ~current_status, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Current Status: CW Chemicals Hazardous Waste",
    xaxis = list(title = "ano // year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines( 
    x = range(hazardous_waste_cs$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
hazardous_waste_plotly
```


### Vessel hydrocarbon pollution
```{r}
vessel_cs <- read_csv(here::here("_pressures/vessels/output/hydrocarbon_pollution_risk.csv")) %>% 
mutate(current_status = (1 - pressure)*100) %>% 
dplyr::select(rgn_id, year, current_status)
```

```{r}
vessel_plotly <- plotly::plot_ly(
  vessel_cs, 
  x = ~year, 
  y = ~current_status, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Current Status: CW Chemicals vessel hydrocarbon pollution",
    xaxis = list(title = "ano // year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines( 
    x = range(vessel_cs$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
vessel_plotly
```



## Explore weighting

### 1. Equal weighting, simple average across current_status scores:

```{r}
chemical_cs_equal <- pharmaceuticals_cs_complete %>%
  rename(PHRM = current_status) %>%
  left_join(pesticides_cs_complete %>% rename(PEST = current_status), by = c("rgn_id", "year")) %>%
  left_join(untreated_ww_cs_complete %>% rename(WWPS = current_status), by = c("rgn_id", "year")) %>%
  left_join(hazardous_waste_cs %>% rename(HAZW = current_status), by = c("rgn_id", "year")) %>%
  left_join(vessel_cs %>% rename(VESS = current_status), by = c("rgn_id", "year")) %>%
  mutate(WWPS = ifelse(rgn_id %in% "Region_2", 100, WWPS),
         HAZW = ifelse(rgn_id %in% "Region_1", 100, HAZW),
         PEST = ifelse(rgn_id %in% "Region_1", 100, PEST)) %>% 
  rowwise() %>%
  mutate(
    chemical_cs = mean(c(PEST, PHRM, WWPS, HAZW, VESS), na.rm = TRUE)
  ) %>%
  ungroup()
```

```{r}
chemical_cs_equal_plotly <- plotly::plot_ly(
  chemical_cs_equal, 
  x = ~year, 
  y = ~chemical_cs, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Current Status: CW Chemicals (arithmetic mean of PEST, HAZW, WWPS, PHRM and VESS)",
    xaxis = list(title = "ano // year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines( 
    x = range(chemical_cs_equal$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
chemical_cs_equal_plotly
```


### 2. Assign PEST, HAZW, and WWPS 1/4, 1/4, 1/4 based on the idea that these types of pollutants directly prevent swimming even when levels are low and split the last 1/4 between PHRM and VESS (1/8 + 1/8) because these type of pollutants only prevent swimming when levels are high. 

```{r}
chemical_cs_weighted <- pharmaceuticals_cs_complete %>%
  rename(PHRM = current_status) %>%
  left_join(pesticides_cs_complete %>% rename(PEST = current_status), by = c("rgn_id", "year")) %>%
  left_join(untreated_ww_cs_complete %>% rename(WWPS = current_status), by = c("rgn_id", "year")) %>%
  mutate(WWPS = ifelse(rgn_id %in% "Region_2", 100, WWPS)) %>% 
  left_join(hazardous_waste_cs %>% rename(HAZW = current_status), by = c("rgn_id", "year")) %>%
  mutate(HAZW = ifelse(rgn_id %in% "Region_1", 100, HAZW),
         PEST = ifelse(rgn_id %in% "Region_1", 100, PEST)) %>% 
  left_join(vessel_cs %>% rename(VESS = current_status), by = c("rgn_id", "year")) %>%
  mutate(
    chemical_cs = (PEST * (1/4)) + (HAZW * (1/4)) + (WWPS * (1/4)) + (PHRM * (1/8)) + (VESS * (1/8))
  )
```

```{r}
chemical_cs_weighted_plotly <- plotly::plot_ly(
  chemical_cs_weighted, 
  x = ~year, 
  y = ~chemical_cs, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Current Status: CW Chemicals [weighted PEST, HAZW, and WWPS (1/4, 1/4, 1/4), PHRM and VESS (1/8 + 1/8)]",
    xaxis = list(title = "ano // year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines( 
    x = range(chemical_cs_weighted$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
chemical_cs_weighted_plotly
```



## Viz

Before combining subcomponents:
```{r}
library(ggplot2)
library(patchwork)
library(cowplot)

# ── shared theme ──────────────────────────────────────────────────────────────
cw_theme <- theme_bw() +
  theme(
    legend.position  = "none",
    plot.title       = element_text(size = 11, face = "bold"),
    axis.title       = element_text(size = 9),
    axis.text        = element_text(size = 8)
  )

fixed_y <- scale_y_continuous(limits = c(0, 100), name = "cs score (0-100)")

# ── helper function ───────────────────────────────────────────────────────────
make_panel <- function(df, title) {
  yr    <- sort(unique(df$year))
  n_yr  <- length(yr)
  step  <- if (n_yr > 10) 5 else if (n_yr > 6) 2 else 1
  brks  <- seq(floor(min(yr) / step) * step, max(yr), by = step)

  ggplot(df, aes(x = year, y = current_status, color = factor(rgn_id), group = factor(rgn_id))) +
    geom_line(linewidth = 1.25) +
    geom_point(size = 1.15) +
    geom_hline(yintercept = 100, linetype = "dashed", color = "black", linewidth = 0.7) +
    fixed_y +
    scale_x_continuous(
      name   = "año // year",
      breaks = brks,
      labels = brks
    ) +
    scale_color_manual(values = rgn_colors) +
    labs(title = title) +
    cw_theme +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

# ── combine: 2x3 layout ───────────────────────────────────────────────────────
panel <- (p1 | p2) / (p3 | p4) / (p5 | plot_spacer())

final_fig <- cowplot::plot_grid(
  panel, shared_legend,
  rel_widths = c(1, 0.15)
)

final_fig

# ── build panels ──────────────────────────────────────────────────────────────
p1 <- make_panel(pesticides_cs_complete,      "Pesticides")
p2 <- make_panel(pharmaceuticals_cs_complete,  "Pharmaceuticals")
p3 <- make_panel(untreated_ww_cs_complete,     "Untreated Wastewater")
p4 <- make_panel(hazardous_waste_cs,           "Hazardous Waste")
p5 <- make_panel(vessel_cs,                    "Vessel Hydrocarbon Pollution")

# ── get all unique region IDs across all datasets ─────────────────────────────
all_rgns <- sort(unique(c(
  pesticides_cs_complete$rgn_id,
  pharmaceuticals_cs_complete$rgn_id,
  untreated_ww_cs_complete$rgn_id,
  hazardous_waste_cs$rgn_id,
  vessel_cs$rgn_id
)))

# ── dummy data to drive a complete legend ─────────────────────────────────────
dummy_df <- data.frame(
  year           = 2000,
  current_status = 50,
  rgn_id         = all_rgns
)

p_legend <- ggplot(dummy_df, aes(x = year, y = current_status, color = factor(rgn_id))) +
  geom_line(linewidth = 1.5) +
  scale_color_manual(values = rgn_colors, name = "Region") +
  theme(legend.position = "right")

shared_legend <- cowplot::get_legend(p_legend)

# ── combine ───────────────────────────────────────────────────────────────────
panel <- (p1 | p2) / (p3 | p4) / (p5 | plot_spacer())

final_fig <- cowplot::plot_grid(
  cowplot::plot_grid(panel, shared_legend, rel_widths = c(1, 0.15)),
  ncol = 1,
  rel_heights = c(1, 0.05)
)

final_fig
```












