---
title: 'OHI GoC 2025: CW Chemicals'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# CW: Chemicals

```{r}
library(tidyverse)
library(here)
library(plotly)

rgn_colors <- c(
  Region_1 = "#C03729",
  Region_2 = "#E68C7C",
  Region_3 = "#FC8F24",
  Region_4 = "#dcaa38",
  Region_5 = "#cbd2a0",
  Region_6 = "#98a25a",
  Region_7 = "#747428",
  Region_8 = "#627391",
  Region_9 = "#8d61a9"
)
```

We now have status scores for each of these 5 subcomponents of Chemicals (a subcomponent of the Clean Waters goal):

- Pesticides (PEST) 
- Pharmaceuticals (PHRM) 
- Hazardous waste sites (HAZW) 
- Untreated wastewater discharge point sites (WWPS) 
- Vessel hydrocarbon estimation (VESS) 

::: {.callout-note collapse="true"}
## Click here to see a table of each subcomponent with more information on each.

![](https://github.com/user-attachments/assets/093565f3-cd5e-4fe8-a765-7a4a31fc5f0d){width=991 height=558}
:::

We discussed: (1) Using [this goalkeeper group activity](https://docs.google.com/document/d/1mNmaxgHCGd57YzAA0afcbNObEByE9meEl_LUy-7m4so/edit?tab=t.0#heading=h.uev2w5dkrcdm) to inform our weighting, (2) a geometric mean across all CS scores, so that the lower scores have a larger impact, and (3) computing overall Chemical status scores using an arithmetic mean across all subcomponents.

Ultimately, we decided that it would make the most sense to weight these status scores based on the strength of their consequences on both human and ocean health.  If we do not have enough evidence to support those weights, then we will use a simple arithmetic mean (no weighting). 

## Bring in all status scores

### Pesticides
```{r}
pesticides_cs <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/pesticides/output/cw_chemical_pesticide_current_status.csv")) %>% 
dplyr::select(rgn_id, year, current_status = current_status_final)
```

```{r}
pest_plotly <- plotly::plot_ly(
  pesticides_cs, 
  x = ~year, 
  y = ~current_status, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Current Status: CW Pesticides",
    xaxis = list(title = "ano // year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines( 
    x = range(pesticides_cs$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
pest_plotly
```


### Pharmaceuticals
```{r}
pharmaceuticals_cs <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/pharmaceuticals","current_status_pharmaceuticals.csv")) %>% 
dplyr::select(rgn_id, year, current_status = current_status_final)
```

```{r}
pharm_plotly <- plotly::plot_ly(
  pharmaceuticals_cs, 
  x = ~year, 
  y = ~current_status, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Current Status: CW Pharmaceuticals",
    xaxis = list(title = "ano // year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines( 
    x = range(pharmaceuticals_cs$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
pharm_plotly
```


### Untreated wastewater discharge points
```{r}
untreated_ww_cs <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/agua_sin_tratamiento/output/untreated_ww_current_status.csv")) %>% 
dplyr::select(rgn_id, year = census_year, current_status = current_status_final)
```

```{r}
untreated_ww_plotly <- plotly::plot_ly(
  untreated_ww_cs, 
  x = ~year, 
  y = ~current_status, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Current Status: CW Untreated Wastewater",
    xaxis = list(title = "ano // year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines( 
    x = range(untreated_ww_cs$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
untreated_ww_plotly
```


### Hazardous waste sites
```{r}
hazardous_waste_cs <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/hazardous_waste/output","current_status_hazardous_waste.csv")) %>% 
dplyr::select(rgn_id, year = anio, current_status = final_score)
```

```{r}
hazardous_waste_plotly <- plotly::plot_ly(
  untreated_ww_cs, 
  x = ~year, 
  y = ~current_status, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Current Status: CW Untreated Wastewater",
    xaxis = list(title = "ano // year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines( 
    x = range(hazardous_waste_plotly$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
untreated_ww_plotly
```


### Vessel hydrocarbon pollution
```{r}

```

```{r}

```



## Explore weighting

```{r}

```



## Viz












