---
title: "Beach access — workflow (split scripts)"
output: html_document
date: "2026-01-06"
---

Determining access for each beach based on presence of:

1. Parking  
2. Road access  
3. Bus access  
4. Beach entry: paths and no blockages  

The analysis is split into four R scripts (run in order from the repo root, or set working directory accordingly):

| Step | Script | Purpose |
|------|--------|---------|
| 1 | `01_collect_access_support_data.R` | Download/read beaches, OSM (paths, parking), INEGI roads and sites, CONANP MPAs, urban localities |
| 2 | `02_identify_beaches_urban_area.R` | For each city in `data/GofC_urban_areas.csv`, buffer included urban locality polygons, intersect beaches, and write one per-city CSV (with expert-opinion fields) |
| 3 | `03_map_city_support_layers.R` | One Mapview QA map for all urban areas in `GofC_urban_areas.csv` (union of buffers; support layers clipped to that region) |
| 4 | `04_visualize_beach_access.R` | Join filled expert CSV to beach geometries; save Leaflet HTML map |

```{r run-step-1, eval=FALSE}
source(here::here("_pressures/access/01_collect_access_support_data.R"))
```

```{r run-step-2, eval=FALSE}
source(here::here("_pressures/access/02_identify_beaches_urban_area.R"))
```

```{r run-step-3, eval=FALSE}
source(here::here("_pressures/access/03_map_city_support_layers.R"))
```

```{r run-step-4, eval=FALSE}
source(here::here("_pressures/access/04_visualize_beach_access.R"))
```

## INEGI field reference (roads / POI / localities)

The following notes are unchanged from the original notebook.

inegi road data  
3.2.1 Road Network (Red Vial)  
A terrestrial transport route intended for vehicular and/or pedestrian traffic. It is a linear element designed to represent vehicular circulation and features a junction-type element at each end. Road Network elements may contain a series of attributes that distinguish them from others, such as type of roadway, official name, traffic direction, etc. (see 3.3 Attribute Tables).

The network is composed of highways, dirt roads (unpaved), paved roads and trails, streets, avenues, and general roadways in urban and rural localities, as well as vehicular transition elements such as U-turns, interchanges, and roundabouts. Any addition to the attributes of a Road Network element—such as a restriction or an alternative name—may require the line to be cut or segmented into two or more parts.

Field Name,Data Type,Description (English),Domain / Constraints,Length  
Id_Red,Integer,Road network identifier,,  
Tipo_Vial,String,Type of roadway,Catalogo_Tipo_Vial,16  
Nombre,String,Roadway name,,254  
Codigo,String,Highway number,,5  
Cond_Pav,String,Type of driving surface,Catalogo_Cond_Pav,14  
Recubri,String,Driving surface material,Catalogo_Recubri,10  
Carriles,String,Number of vehicle lanes,"1..n, N/A",3  
Estatus,String,Routing service status,Catalogo_Estatus,13  
Condicion,String,Roadway condition,Catalogo_Condicion,26  
Nivel,Integer,Data to distinguish elevation levels,-3...5,  
Peaje,String,Payment for transit rights (Toll),Catalogo_Peaje,2  
Administra,String,Administration,Catalogo_Administra,11  
Jurisdi,String,Jurisdiction,Catalogo_Jurisdi,13  
Circula,String,Direction of vehicle circulation,Catalogo_Sentido,25  
Escala_Vis,Integer,Visualization scale,1..5,  
Velocidad,String,Speed for routing,"10..110, N/A",3  
Union_Ini,Integer,Initial junction identifier,,  
Union_Fin,Integer,Final junction identifier,,  
Longitud,Double,Length in meters,,  
Ancho,Double,Highway or roadway width,,  
Fecha_Act,String,Update date,,19  
Calirepr,String,Position qualifier,Catalogo_Calirepr,10  

3.2.6 Point of Interest (Sitio de interés)  
A location where services and places of interest are situated. Many services are particularly relevant to the highway theme, such as recreation, maintenance, emergency services, customs, and various minor retail services. Services may be related to a Road Network element or to junctions to describe information necessary for vehicular access. Services can be further characterized by attributes.

For this version of the product, relevant services along the highway network that serve as connections to other modes of transport—such as airports and seaports—are considered, as well as toll plazas (caseta de cobro) on toll roads to estimate travel costs, fuel stations, and others such as phytosanitary and zoosanitary inspection stations, in addition to tourist attractions.

Note: These sites are placed based on network coverage. Therefore, in urban areas where only main thoroughfares are included, the recording of these points is conditioned upon their proximity to the network lines. Once all urban roadways are incorporated, the remaining sites will be included.  
Field Name,Data Type,Description (English),Domain / Constraints,Length  
Id_Loc,Integer,Unique locality identifier,,  
Tipo,String,Locality type,Catalogo_Tipo_Localidad,6  
Nombre,String,Official designation,,254  
Cve_Geo,String,Geostatistical code,,15  
Fecha_Act,String,Update date,,19  
Calirepr,String,Position qualifier,Catalogo_Calirepr,10  

3.2.9 Locality (Localidad)  
A place occupied by one or more buildings used as dwellings, which may or may not be inhabited; this place is recognized by a name given by legal provision or custom. Localities are represented by a point feature, which can spatially connect with one or more Road Network elements to identify origins or destinations for route determination. The types of localities in this project are classified as urban and rural.

Field Name,Data Type,Description (English),Domain / Constraints,Length  
Id_Loc,Integer,Unique locality identifier,,  
Tipo,String,Locality type,Catalogo_Tipo_Localidad,6  
Nombre,String,Official designation,,254  
Cve_Geo,String,Geostatistical code,,15  
Fecha_Act,String,Update date,,19  
Calirepr,String,Position qualifier,Catalogo_Calirepr,10  

## Example: state bounding box (Mazatlán / Sinaloa)

```{r example-bbox, eval=FALSE}
library(osmdata)
library(sf)

bb <- getbb("Sinaloa, Mexico")
bb_poly <- st_as_sfc(st_bbox(c(xmin = bb[1, 1], ymin = bb[2, 1], xmax = bb[1, 2], ymax = bb[2, 2]), crs = 4326))
# roads <- st_read(".../red_vial.shp")
# bb_poly_tr <- st_transform(bb_poly, st_crs(roads))
# mapview::mapview(bb_poly_tr)
```
