################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```

# >> Report Statistics <<  ------------------------------
```{r Load in AFO_Scores as needed}
AFO_Scores0<- read.csv("/home/ferrer/ohi_goc_prep/AO/v2025/AO_Scores/AFO_Scores/AFO_Scores_1.0_df_Jun152026.csv")

AFO_Scores<- AFO_Scores0
```

```{r Average AFO scores over time}
AFO_averages<- AFO_Scores %>%
  group_by(OHI_region_est) %>%
  summarize(
    mean_AFO_score = mean(AFO_Score, na.rm= TRUE)
  )
```



```{r How much seafood is produced in this region in 2024?}
AllMex_Landings<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/juan_carlos/AllMex_landings_2000_present_EMF_edited.rds"))
# 12153879 obs., 20 vars.

# Subset records from 2024:
AllMex_Landings_2024<- AllMex_Landings %>% filter(year==2024)
# 483775 obs., 20 vars.

# Landed weight correction (kind of looks like sometimes 'landed weight' is 'live weight' and vice versa. Fix this):
AllMex_Landings_2024$weight_kg_corrected <- ifelse(
  AllMex_Landings_2024$live_weight >= AllMex_Landings_2024$landed_weight,
  AllMex_Landings_2024$live_weight,
  AllMex_Landings_2024$landed_weight
)
# 483775 obs., 21 vars.

# Let's filter out any records where weight_kg_corrected is < or = 0.
AllMex_Landings_2024<- AllMex_Landings_2024 %>%
  filter(weight_kg_corrected > 0)
# 483775 obs., 21 vars.

sum(is.na(AllMex_Landings_2024$weight_kg_corrected))
# 0. Good!

# Finally, let's apply a coarse filter to filter out landings from aquaculture. This won't get everything.
table(AllMex_Landings_2024$acuaculture_production) #  NO: 421421 and the rest are NAs.


## VERSUS - [MARINE] LANDINGS IN ONLY THE GoC
# This is an imperfect calculation because we can't be certain that all of these sites are inside the GoC.


GoC_Landings_cleaned<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/fis/v2025/int/GOC_FIS.landings_CLEAN_as.of.Dec092025.rds"))
# GoC_Landings<- Landings_GoC_NoFresh.NoMariculture
# 3163120 obs., 50 vars.

# Subset records from 2024:
GoC_Landings_2024<- GoC_Landings_cleaned %>% filter(year== 2024)
# 143479 obs., 50 vars.

# Let's filter out any records where weight_kg_corrected is < or = 0.
GoC_Landings_2024<- GoC_Landings_2024 %>%
  filter(weight_kg_corrected > 0)
# 143479 obs., 50 vars.

sum(is.na(GoC_Landings_2024$weight_kg_corrected))
# 0. Good!

table(GoC_Landings_2024$OHI_Region_No)
unique(GoC_Landings_2024$OHI_Region_No)

Definitely_GoC_Regions<- c("Region_1", "Region_2", "Region_3",
                           "Region_4", "Region_5", "Region_6",
                           "Region_7", "Region_8", "Region_9",
                           "Region_4_5_6_Sonora_TBD")
# We cannot be certain about the landings from "Region_1_2_Baja_TBD" or "Region_3_4_BCS_TBD"

GoC_Landings_2024<- GoC_Landings_2024 %>% filter(OHI_Region_No %in% Definitely_GoC_Regions)
# 142533 obs., 50 vars.

#### CALCULATION: How much seafood produced in the region? ####
# ANSWER: ~50% in 2024

sum(AllMex_Landings_2024$weight_kg_corrected)
# 1993632680 kg

sum(GoC_Landings_2024$weight_kg_corrected)
# 995487742 kg


#### CALCULATION: How much seafood VALUE is produced in the region? ####
# ANSWER: ~23% in 2024

sum(AllMex_Landings_2024$value)
# 44196283211 pesos

sum(GoC_Landings_2024$value)
# 9978483456 pesos


#### CALCULATION: How much seafood in the region is produced by SSF? ####
# ANSWER: 20% total in 2024; anywhere from 25-48% when exclude large-scale anchovy and sardine

table(GoC_Landings_2024$fleet)
# large_scale small_scale NA
#       14546      127261 

GoC_Landings_2024_All<- GoC_Landings_2024 %>% filter(!is.na(fleet))
# 141807 obs.

GoC_Landings_2024_SSF<- GoC_Landings_2024 %>% filter(fleet=="small_scale")
# 127261 obs.

sum(GoC_Landings_2024_All$weight_kg_corrected) # 992204985 kg
sum(GoC_Landings_2024_SSF$weight_kg_corrected) # 199330559 kg

# What about for direct consumption?
GoC_Landings_2024_Large<- GoC_Landings_2024 %>% filter(fleet=="large_scale")
# 14546 obs.

table(GoC_Landings_2024_Large$recorded_taxa_group)

GoC_Landings_2024_withoutLargeAnchovy <- GoC_Landings_2024 %>% 
  filter(
    !(fleet == "large_scale" &
      recorded_taxa_group %in% c("ANCHOVETA", "SARDINA"))
  )
# 137815 obs.

sum(GoC_Landings_2024_withoutLargeAnchovy$weight_kg_corrected) # 417828233 kg
sum(GoC_Landings_2024_SSF$weight_kg_corrected) # 199330559 kg
```


```{r How much of the seafood can we account for in our analysis?}
# Load file, only if needed:
GoC_Landings_TEST<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/fis/v2025/int/GOC_FIS.landings_CLEAN_as.of.Dec092025.rds"))
# 3163120 obs., 50 vars.

# General filtering:
# In general, the receipts from early trips are a bit funky. Let's filter out trips before 2003.
GoC_Landings_TEST<- GoC_Landings_TEST %>%
  filter(year >= 2003)
# 2896048 obs., 50 vars.

# Let's also filter out any records where weight was < or = 0 (n= 126955 records).
GoC_Landings_TEST<- GoC_Landings_TEST %>%
  filter(weight_kg_corrected > 0 & !is.na(weight_kg_corrected))
# 2895892 obs., 50 vars.

# Re-assigning CPUE major groups based on new key:
Genus_MajorGroup_Key<- read.csv("~/ohi_goc_prep/general/CONAPESCA_Keys/Genus_to_MajorGroup_Key_Dec112025.csv")
# 326 obs., 4 vars.

GoC_Landings_TEST <- GoC_Landings_TEST %>%
  left_join(Genus_MajorGroup_Key,
            by = c("recorded_taxa_group", "estimated_genus_name"))
# 2895892 obs., 52 vars. as expected!


#### CALCULATION: How much of the seafood can we account for in our analysis? ####
# ANSWER: Anywhere from 92-99% in any given year.

table(GoC_Landings_TEST$Major_Functional_Group)
# We can account for 2778209 out of 2895892 records (96%).

# How much of the landings can we account for?
Yearly_weight_summary <- GoC_Landings_TEST %>%
  group_by(year) %>%
  summarise(
    # Total weight across ALL records
    total_weight_all = sum(weight_kg_corrected, na.rm = TRUE),

    # Weight where Major_Functional_Group is present and not NEI
    total_weight_tax_grouped = sum(
      weight_kg_corrected[
        !is.na(Major_Functional_Group) &
        Major_Functional_Group != "NEI"
      ],
      na.rm = TRUE
    ),

    # Optional: proportion of total that is taxonomically grouped
    prop_tax_grouped = total_weight_tax_grouped / total_weight_all
  ) %>%
  ungroup()
```

