---
title: "Maintaining beach access — PROFEPA capacity & municipal ZOFEMAT coverage"
author: "OHI Gulf of California / beach access pressure"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9.5,
  fig.height = 5.0,
  dpi = 120
)
```

# Why this indicator

Equitable beach access in Mexico rests on federal law: public transit on beaches and the Zona Federal Marítimo Terrestre (ZOFEMAT) must not be impeded. When hotels, condos, or gates block that access, **opening it again depends on enforcement capacity**, not only on municipal fee collection.

This report tracks three capacity signals that can be built from public data:

1. **PROFEPA authorized plazas** (whole agency) — raw and per capita  
2. **PROFEPA inspection-program PEF** — raw, inflation-adjusted, and inflation-adjusted per capita  
3. **Share of study municipalities with an active municipal ZOFEMAT office**, by **OHI marine region**

> **Core caveat:** PROFEPA covers *all* federal environmental inspection (industry, wildlife, ANPs, forests, ports, ZOFEMAT…). PEF plazas cannot be split into an “inspection program” series; inspection **budget** can (PP 005→014). Neither series is ZOFEMAT-only. Treat both as **upper-bound envelopes of capacity that could be applied to blocked beaches**.

```{r packages-paths}
library(tidyverse)
library(scales)
library(knitr)

ref_dir <- {
  cand <- c(
    getwd(),
    file.path(getwd(), "references"),
    file.path(getwd(), "_pressures", "access", "references")
  )
  hit <- cand[file.exists(file.path(cand, "output", "pef_national_series_2018_2026.csv"))]
  if (!length(hit)) stop("Cannot find references/output PEF CSVs")
  normalizePath(hit[[1]])
}
out_dir <- file.path(ref_dir, "output")
years <- 2018:2026

rgn_labels <- c(
  Region_1 = "Western Upper Gulf",
  Region_2 = "Western Midriff Islands",
  Region_3 = "Western Central Gulf",
  Region_4 = "Western Southern Gulf",
  Region_5 = "Eastern Upper Gulf",
  Region_6 = "Eastern Midriff Islands",
  Region_7 = "Southern Sonora Coast",
  Region_8 = "Sinaloa Coast",
  Region_9 = "Southeastern Gulf"
)
rgn_colors <- c(
  Region_1 = "#C03729", Region_2 = "#E68C7C", Region_3 = "#FC8F24",
  Region_4 = "#dcaa38", Region_5 = "#cbd2a0", Region_6 = "#98a25a",
  Region_7 = "#747428", Region_8 = "#627391", Region_9 = "#8d61a9"
)
```

```{r load-data}
nat <- read_csv(file.path(out_dir, "pef_national_series_2018_2026.csv"), show_col_types = FALSE)
plazas <- read_csv(file.path(out_dir, "pef_plazas_dgz_profepa_2018_2026.csv"), show_col_types = FALSE)
denoms <- read_csv(file.path(out_dir, "indicator_denominators.csv"), show_col_types = FALSE)
inpc <- read_csv(file.path(out_dir, "indicator_inpc_annual.csv"), show_col_types = FALSE)
zof_rgn <- read_csv(file.path(out_dir, "indicator_zofemat_by_ohi_region.csv"), show_col_types = FALSE)

inpc_2018 <- inpc$inpc_annual_avg[inpc$year == 2018]

plazas_p <- plazas %>%
  filter(unit == "PROFEPA") %>%
  left_join(denoms %>% select(year, mexico_pop_millions), by = "year") %>%
  mutate(
    plazas_per_million_people = plazas / mexico_pop_millions,
    index_plazas_2018 = 100 * plazas / plazas[year == 2018],
    index_plazas_pop_2018 = 100 * plazas_per_million_people /
      plazas_per_million_people[year == 2018]
  )

insp <- nat %>%
  filter(series == "PROFEPA_inspeccion") %>%
  select(year, mdp_current = mdp) %>%
  left_join(denoms %>% select(year, mexico_pop_millions), by = "year") %>%
  left_join(inpc %>% select(year, inpc_annual_avg), by = "year") %>%
  mutate(
    mdp_real_2018 = mdp_current * (inpc_2018 / inpc_annual_avg),
    mdp_real_per_million_people = mdp_real_2018 / mexico_pop_millions,
    index_current_2018 = 100 * mdp_current / mdp_current[year == 2018],
    index_real_2018 = 100 * mdp_real_2018 / mdp_real_2018[year == 2018],
    index_real_pc_2018 = 100 * mdp_real_per_million_people /
      mdp_real_per_million_people[year == 2018]
  )

zof_rgn <- zof_rgn %>%
  mutate(
    rgn_name = coalesce(rgn_labels[rgn_id], as.character(ecorgn)),
    rgn_label = paste0(rgn_id, " — ", rgn_name),
    prop_pct = 100 * prop_active
  )

write_csv(plazas_p, file.path(out_dir, "indicator_profepa_plazas_national.csv"))
write_csv(insp, file.path(out_dir, "indicator_profepa_inspection_national.csv"))
```

# Agencies and how blocked access is handled

ZOFEMAT is **federal national property**. Public use of the beach and transit to the sea must not be impeded. Occupation of the federal strip normally needs a SEMARNAT concession. Walls, gates, private guards, or “guests only” barriers on or to the beach are classic enforcement cases.

## Who does what

| Actor | Role |
|-------|------|
| **Public / NGOs / visitors** | Notice blockage; file complaints |
| **Municipal ZOFEMAT** (where it exists) | Local front desk — derechos, cleaning, anuencias, documentation, soft pressure. **Cannot** grant federal concessions or usually issue federal clausuras |
| **SEMARNAT / DGZFMTAC** | Concession authority (who may occupy; delimitation). National unit; **not** the main beach-access policeman |
| **PROFEPA** (state Oficina de Representación + national capacity) | **Main enforcement engine** — inspects; can order temporary closure, fines, removal |

## How a complaint typically moves (example: La Paz)

1. Someone finds a gated path or wall blocking beach access.  
2. They may contact **municipal ZOFEMAT** (in La Paz: Tesorería ZOFEMAT) *or* go straight to **PROFEPA BCS**.  
3. The municipality can check occupancy/derechos and urge reopening; if the actor refuses, the case needs **PROFEPA**.  
4. PROFEPA receives a **denuncia** (state office or hotline **01 800 776 3372**), inspects, and may impose measures.  
5. SEMARNAT enters if the core problem is a missing or illegal concession.

**Implication:** municipal offices help routing and soft management; **sustained libre acceso under development pressure depends on PROFEPA capacity** — measured here with plazas + inspection funding — plus whether a local ZOFEMAT desk even exists to receive complaints.

## Why both plazas and inspection budget

| Series | Isolatable as “inspection”? | Interpretation |
|--------|-----------------------------|----------------|
| **PROFEPA plazas** | **No** (agency-wide; PEF plazas lack a programa presupuestario split) | Best available **people** envelope |
| **PROFEPA inspection PEF** | **Yes** (PP 005 through 2025 / PP 014 in 2026) | Money tagged to the inspection program — closer to enforcement operations than total PROFEPA $ |

People are usually the better capacity signal, but only the budget side cleanly isolates the inspection component. Using **both** is intentional.

# Indicator 1 — PROFEPA plazas (raw and per capita)

**Source:** PEF *Analítico de Plazas* for UR E00 (PROFEPA). Counts are **authorized posts**, not necessarily filled headcount. Job titles include only a small number of explicit “Inspector Especializado…” posts; most plazas are other roles that may still support inspection.

```{r plot-plazas-raw}
ggplot(plazas_p, aes(year, plazas)) +
  geom_line(linewidth = 1.15, color = "#7c3aed") +
  geom_point(size = 2.6, color = "#7c3aed") +
  scale_x_continuous(breaks = years) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title = "PROFEPA authorized plazas (national, whole agency)",
    subtitle = "PEF plazas analitico — all-environment mandate; not ZOFEMAT-only",
    x = NULL, y = "Plazas"
  ) +
  theme_minimal(base_size = 12)
```

```{r plot-plazas-pc, fig.height=4.6}
plazas_p %>%
  select(year, index_plazas_2018, index_plazas_pop_2018) %>%
  pivot_longer(-year, names_to = "series", values_to = "index") %>%
  mutate(series = recode(series,
    index_plazas_2018 = "Plazas (raw)",
    index_plazas_pop_2018 = "Plazas per million people"
  )) %>%
  ggplot(aes(year, index, color = series)) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey60") +
  geom_line(linewidth = 1.05) +
  geom_point(size = 2.2) +
  scale_x_continuous(breaks = years) +
  labs(
    title = "PROFEPA plazas — raw vs population-adjusted",
    subtitle = "Index 2018 = 100; population from CONAPO/INEGI approximations",
    x = NULL, y = "Index (2018 = 100)", color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

```{r tbl-plazas}
plazas_p %>%
  transmute(
    Year = year,
    Plazas = round(plazas),
    `Per million people` = round(plazas_per_million_people, 2),
    `Index raw` = round(index_plazas_2018, 1),
    `Index per capita` = round(index_plazas_pop_2018, 1)
  ) %>%
  kable(caption = "National PROFEPA plazas")
```

Plazas fell around 2020 (≈2242 → ≈1942) and remained near ≈1975 through 2026 — **below 2018**, especially per capita.

# Indicator 2 — PROFEPA inspection PEF (raw, real, real per capita)

**Series:** authorized PEF for PROFEPA’s inspection program — **PP 005** (2018–2025) / **PP 014** (2026).  
**Inflation:** annual-average **INPC** (base 2Q Jul 2018 = 100); amounts expressed in **constant 2018 MDP**.  
**Per capita:** real MDP ÷ Mexico population (millions).

```{r plot-insp-raw-real, fig.height=5.0}
insp %>%
  select(year, mdp_current, mdp_real_2018) %>%
  pivot_longer(-year, names_to = "series", values_to = "mdp") %>%
  mutate(series = recode(series,
    mdp_current = "Current MDP",
    mdp_real_2018 = "Real MDP (2018 pesos)"
  )) %>%
  ggplot(aes(year, mdp, color = series)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.3) +
  scale_x_continuous(breaks = years) +
  scale_y_continuous(labels = label_comma()) +
  scale_color_manual(values = c("Current MDP" = "#0f172a", "Real MDP (2018 pesos)" = "#0369a1")) +
  labs(
    title = "PROFEPA inspection program — authorized PEF",
    subtitle = "PP 005 / 014; inflation via annual-average INPC → constant 2018 MDP",
    x = NULL, y = "MDP", color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

```{r plot-insp-indexes, fig.height=5.0}
insp %>%
  select(year, index_current_2018, index_real_2018, index_real_pc_2018) %>%
  pivot_longer(-year, names_to = "series", values_to = "index") %>%
  mutate(series = recode(series,
    index_current_2018 = "Current MDP",
    index_real_2018 = "Real MDP (2018 pesos)",
    index_real_pc_2018 = "Real MDP per million people"
  )) %>%
  ggplot(aes(year, index, color = series)) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey60") +
  geom_line(linewidth = 1.05) +
  geom_point(size = 2.2) +
  scale_x_continuous(breaks = years) +
  labs(
    title = "Inspection PEF capacity indexes",
    subtitle = "2018 = 100. Real series remove price growth; per-capita also accounts for population",
    x = NULL, y = "Index (2018 = 100)", color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

```{r tbl-insp}
insp %>%
  transmute(
    Year = year,
    `Current MDP` = round(mdp_current, 1),
    `Real MDP (2018)` = round(mdp_real_2018, 1),
    `Real MDP / M people` = round(mdp_real_per_million_people, 2),
    INPC = round(inpc_annual_avg, 1)
  ) %>%
  kable(caption = "Inspection PEF — current, real, and real per capita")
```

In current pesos, inspection PEF fell from ~296 MDP (2018) to ~162 MDP (2026). In **real** terms the decline is steeper; real **per capita** capacity is weaker still. For access maintenance, the relevant story is whether this envelope keeps up with coastal pressure — here it has declined.

# Indicator 3 — Municipal ZOFEMAT coverage by OHI region

**Universe:** the 25 municipios that contain the 41 beach-access study localities.  
**Active office:** `has_local_zofemat_office = yes` from open-web inventory (15 Jul 2026) — La Paz, Los Cabos, Puerto Peñasco, Bahía de Banderas.  
**Region assignment:** INEGI locality polygons for study localities → centroid → intersect/nearest **marine** OHI region (`spatial/ohi_regions/goc_ohi_rgns.shp`). Each study municipio maps to one region in this set.

**Metric:** for each OHI region with ≥1 study municipio,  
\(\text{proportion} = n_{\text{municipios with active ZOFEMAT}} / n_{\text{study municipios in region}}\).

```{r plot-zof-rgn, fig.height=5.2}
# ensure all Region_1..9 appear if desired — only those with study mun
zof_plot <- zof_rgn %>%
  mutate(rgn_id = factor(rgn_id, levels = paste0("Region_", 1:9)))

ggplot(zof_plot, aes(rgn_id, prop_pct, fill = rgn_id)) +
  geom_col(width = 0.75) +
  geom_text(
    aes(label = sprintf("%.0f%%\n(%d/%d)", prop_pct, n_active_zofemat, n_municipios)),
    vjust = -0.15, size = 3.1
  ) +
  scale_fill_manual(values = rgn_colors, guide = "none") +
  scale_y_continuous(limits = c(0, 100), expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "Share of study municipalities with an active municipal ZOFEMAT office",
    subtitle = "By OHI marine region; active = confirmed dedicated municipal unit (yes)",
    x = NULL, y = "Percent of study municipios"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

```{r tbl-zof-rgn}
zof_rgn %>%
  arrange(rgn_id) %>%
  transmute(
    Region = rgn_id,
    Name = rgn_name,
    `Study municipios` = n_municipios,
    `With active ZOFEMAT` = n_active_zofemat,
    `Percent` = round(prop_pct, 1),
    `Active municipios` = if_else(active_municipios == "", "—", active_municipios),
    `All study municipios` = municipios
  ) %>%
  kable(caption = "Municipal ZOFEMAT coverage by OHI region (study municipios only)")
```

```{r zof-overall}
overall_mun <- read_csv(
  file.path(ref_dir, "..", "local_zofemat", "local_zofemat_by_municipio.csv"),
  show_col_types = FALSE
)
n_yes <- sum(overall_mun$has_local_zofemat_office == "yes")
n_mun <- nrow(overall_mun)
```

Overall, **`r n_yes` of `r n_mun`** study municipios (`r round(100 * n_yes / n_mun, 0)`%) have a confirmed active municipal ZOFEMAT office. Most OHI regions with study sites show **0%** coverage; the higher shares are Western Southern Gulf (La Paz / Los Cabos), Eastern Upper Gulf (Puerto Peñasco), and Southeastern Gulf (Bahía de Banderas).

Municipal units facilitate complaints and beach admin; **they do not replace PROFEPA** for forcing open a blocked access.

Detail and contacts: [`../local_zofemat/local_zofemat_offices.html`](../local_zofemat/local_zofemat_offices.html).

# Data weaknesses

1. **Not ZOFEMAT-only.** Inspection PEF and plazas cover all PROFEPA monitoring; beaches are an unknown slice.  
2. **Plazas ≠ inspection staff.** PEF plazas have no programa presupuestario field; agency-wide counts only.  
3. **Authorized ≠ executed; plazas ≠ filled posts.**  
4. **INPC 2026 is provisional** (Jan–May mean) for comparing to PEF 2026.  
5. **Population denominators** are national approximations (CONAPO/INEGI).  
6. **Municipal “active office”** from Jul 2026 open-web research — offices may exist without public branding; coding is `yes` only when a dedicated unit was found.  
7. **OHI assignment** uses locality centroids vs marine regions; municipalities without study localities are out of scope. Region_2 has no study municipios in this set.  
8. **State PROFEPA PEF EF tags** are omitted here — they are small CDMX-dominated accounting tags and are not a good proxy for state plazas or office budgets.  
9. **DGZFMTAC** is concession admin (national), not used as an enforcement indicator.

# Putting the pieces together

| Component | Desired direction for access maintenance |
|-----------|------------------------------------------|
| PROFEPA plazas (raw + per capita) | Stable or ↑ |
| Inspection PEF (real + real per capita) | Stable or ↑ |
| % municipalities with active ZOFEMAT (by OHI region) | ↑ / gap rates ↓ |

**2018–2026 reading:** national plazas and real inspection funding are **weaker than in 2018**, while many OHI regions have **no** confirmed municipal ZOFEMAT desk among study municipios. That combination does not favor maintaining libre acceso under coastal development pressure. These remain **governance/capacity proxies**, not substitutes for field equitable-access beach scores.

# Sources & outputs

- PEF Programa Presupuestario analiticos 2018–2026 (`sources/pef_analiticos/pp/`)  
- PEF Plazas analiticos (`sources/pef_analiticos/plazas/`)  
- INPC annual means: `output/indicator_inpc_annual.csv`  
- Population: `output/indicator_denominators.csv`  
- ZOFEMAT × OHI: `output/indicator_zofemat_by_ohi_region.csv`, `output/study_municipios_ohi_region_zofemat.csv`  
- Municipal inventory: [`../local_zofemat/`](../local_zofemat/)  
- Prior narrative summary: [`zofemat_budget_enforcement_summary.html`](zofemat_budget_enforcement_summary.html)
