################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
# Visualization
library(ggplot2)
library(colorspace)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/BD/")
```

```{r Load clean_names_lowercase function}
clean_names_lowercase <- function(x) {

  x <- iconv(x, from = "", to = "UTF-8", sub = "")
  x <- stringi::stri_trans_general(x, "Latin-ASCII")

  x %>%
    tolower() %>%
    trimws()
}
```


# STEP 1: Load in and summarize GOC_MarineLandings0

```{r Load in GOC_MarineLandings0 as necessary}
GOC_MarineLandings0<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Marine.Landings_Sector.Sorted.Georef_PATCHED_Jun032026.rds"))
# 3246583 obs., 58 var.

# NOTE TO SELF: This source file may change in later Summer / Fall run. - EMF, June 4 2026
# See the patch file here: https://docs.google.com/document/d/1L6w4Gd8SOjp14f_u46vyn64rEpv4qBRaN_V2IyPjP5w/edit?tab=t.0
```


# PATCH 1 - Assign groups that are unassigned to a Major_Functional_Group:
NOTE TO SELF: This patch should not be necessary when we re-run this in later Summer / Fall run. - EMF, June 15 2026
See the patch file here: https://docs.google.com/document/d/1L6w4Gd8SOjp14f_u46vyn64rEpv4qBRaN_V2IyPjP5w/edit?tab=t.0

```{r Load in GenusFamilyGroup_Key as needed}
GenusFamilyGroup_Key<- read.csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Keys/Taxonomic_Keys/GenusFamilyGroupKey_COMPLETE_Apr042026.csv")

# NOTE TO SELF: This patch should not be necessary when we re-run this in later Summer / Fall run. - EMF, June 15 2026
# See the patch file here: https://docs.google.com/document/d/1L6w4Gd8SOjp14f_u46vyn64rEpv4qBRaN_V2IyPjP5w/edit?tab=t.0

MajorFunctionalGroup_Key<- GenusFamilyGroup_Key %>%
  select(estimated_genus_name, Major_Functional_Group)
```

```{r Major_Functional_Group patch for GOC_MarineLandings}
# Fill in missing values based on MajorFunctionalGroup_Key:
GOC_MarineLandings <- GOC_MarineLandings0 %>%
  left_join(
    MajorFunctionalGroup_Key %>%
      select(estimated_genus_name, Major_Functional_Group),
    by = "estimated_genus_name",
    suffix = c("", ".lookup")
  )

GOC_MarineLandings <- GOC_MarineLandings %>%
  mutate(
    Major_Functional_Group = coalesce(
      Major_Functional_Group,
      Major_Functional_Group.lookup
    )
  ) %>%
  select(-Major_Functional_Group.lookup)

# 3246583 obs., 58 var.

# SANITY CHECK:
sum(is.na(GOC_MarineLandings$Major_Functional_Group))
# Only 26 records unassigned. Looks good!
```

# PATCH 2 - Re-assign paneus as litopenaeus:
NOTE TO SELF: This patch should not be necessary when we re-run this in later Summer / Fall run. - EMF, June 15 2026
See the patch file here: https://docs.google.com/document/d/1L6w4Gd8SOjp14f_u46vyn64rEpv4qBRaN_V2IyPjP5w/edit?tab=t.0

```{r Re-assign paneus as litopenaeus}
GOC_MarineLandings <- GOC_MarineLandings %>%
  mutate(
    estimated_genus_name = case_when(
      estimated_genus_name == "Penaeus" ~ "Litopenaeus",
      TRUE ~ estimated_genus_name
    )
  )

unique(GOC_MarineLandings$estimated_genus_name)
```


# **DECISION POINT: Selecting which sectors to include**
```{r Select Sectors_of_Interest}
table(GOC_MarineLandings$sector_estimate_draft)

MarineFishery_Sectors<- c("marine: fishery sector yet to be classified",
                          "marine: small-scale capture",
                          "marine: large-scale capture")
  
PossibleMarineFishery_Sectors<- c("freshwater or marine: large-scale capture",
                          "freshwater or marine: small-scale capture",
                          "freshwater or marine: fishery sector yet to be classified")

GOC_MarineFisheryLandings0<- GOC_MarineLandings %>%
  filter(sector_estimate_draft %in% c(MarineFishery_Sectors, PossibleMarineFishery_Sectors))

# 3137044 obs., 58 var.
```

```{r Re-level sector_estimate_draft and Major_Functional_Group}
# Re-level sectors (mainly for plotting purposes):
GOC_MarineFisheryLandings <- GOC_MarineFisheryLandings0 %>%
  filter(
    sector_estimate_draft %in% c(
      MarineFishery_Sectors,
      PossibleMarineFishery_Sectors
    )
  ) %>%
  mutate(
    sector_estimate_draft = factor(
      sector_estimate_draft,
      levels = c(
        "marine: large-scale capture",
        "marine: small-scale capture",
        "marine: fishery sector yet to be classified",
        "freshwater or marine: large-scale capture",
        "freshwater or marine: small-scale capture",
        "freshwater or marine: fishery sector yet to be classified"
      )
    )
  )

# SANITY CHECK: How many Major_Functional_Group unassigned?
sum(is.na(GOC_MarineFisheryLandings$Major_Functional_Group))
# Only 18 records unassigned. Looks good!

table(GOC_MarineFisheryLandings$Major_Functional_Group)

GOC_MarineFisheryLandings <- GOC_MarineFisheryLandings %>%
  mutate(
    Major_Functional_Group = factor(
      Major_Functional_Group,
      levels= c(
        
        # Crustaceans
        "Shrimps, prawns",
        "Crustaceans",
        
        # Molluscs
        "Cephalopods",
        "Molluscs excl. Cephalopods",
        
        # Other invertebrates
        "Echinoderms",
        "Jellies",
        "Other marine invertebrates",
        
        # Fishes
        "Anchovy & Sardine (small pelagics)",
        "Pelagic marine fish (excl. anchovy & sardine)",
        "Demersal marine fishes",
        "Marine fishes NEI",
        "Marine or freshwater fishes NEI",
        "Diadromous fishes",
        
        # Elasmobranchs
        "Sharks",
        
        # Miscellaneous / unresolved
        "Marine algae",
        "Marine species NEI",
        "unknown",
        NA_character_
      )
    )
  )

# SANITY CHECK:
sum(is.na(GOC_MarineFisheryLandings$Major_Functional_Group))
# Only 18 records unassigned. Looks good!
```

# Strategy 1 - Track changes in major functional groups relative to mean  ------------------------------

## Rocky Reef Communities

```{r Isolate Rocky Reef taxa and create RockyReef_Landings0}
Rocky_Reef_Taxa<- c("Crustaceans", "Shrimps, prawns", "Echinoderms", "Demersal marine fishes", "Marine algae")

colnames(GOC_MarineFisheryLandings)

RockyReef_Landings0<- GOC_MarineFisheryLandings %>%
  filter(Major_Functional_Group %in% Rocky_Reef_Taxa) %>%
    select(state, OHI_region_est,
         office_name_clean, landing_site_clean, 
         year,
         est_aquaculture_production, reported_fleet, 
         sector_estimate_reported, sector_estimate_draft,
         recorded_taxa_name, recorded_taxa_name_clean, common_name_clean, 
         recorded_taxa_group, estimated_taxa_group, Major_Functional_Group,
         estimated_species_name, estimated_genus_name, estimated_family_name,
         landed_weight, live_weight, weight_kg_corrected, value
         )

# 2189051 obs., 22 vars.
```

```{r Summarize landings}
RockyReef_Landings_summary<- RockyReef_Landings0 %>%
  group_by(year, OHI_region_est, Major_Functional_Group) %>%
  summarize(
    total_landings_mt = sum(weight_kg_corrected, na.rm = TRUE)/1000,
    .groups = "drop"
  ) %>%
  mutate(
    Major_Functional_Group= clean_names_lowercase(Major_Functional_Group),
    OHI_region_est= clean_names_lowercase(OHI_region_est)
  )
# 868 obs.
```

```{r Plot over time}
colnames(RockyReef_Landings_summary)

RockyReef_Landings_summary %>%
  ggplot(
    aes(
      x = year,
      y = log10(total_landings_mt),
      color = Major_Functional_Group,
      group = Major_Functional_Group
    )
  ) +
  geom_line() +
  theme_bw() +
  facet_wrap(~ OHI_region_est)
```

## Open Ocean Communities

```{r Isolate Rocky Reef taxa and create RockyReef_Landings0}
unique(GOC_MarineFisheryLandings$Major_Functional_Group)

Open_Ocean_Taxa<- c("Pelagic marine fish (excl. anchovy & sardine)",
"Anchovy & Sardine (small pelagics)",
"Sharks",
"Cephalopods"
)

colnames(GOC_MarineFisheryLandings)

OpenOcean_Landings0<- GOC_MarineFisheryLandings %>%
  filter(Major_Functional_Group %in% Open_Ocean_Taxa) %>%
    select(state, OHI_region_est,
         office_name_clean, landing_site_clean, 
         year,
         est_aquaculture_production, reported_fleet, 
         sector_estimate_reported, sector_estimate_draft,
         recorded_taxa_name, recorded_taxa_name_clean, common_name_clean, 
         recorded_taxa_group, estimated_taxa_group, Major_Functional_Group,
         estimated_species_name, estimated_genus_name, estimated_family_name,
         landed_weight, live_weight, weight_kg_corrected, value
         )

# 653887 obs., 22 vars.
```

```{r Summarize landings}
OpenOcean_Landings_summary<- OpenOcean_Landings0 %>%
  group_by(year, OHI_region_est, Major_Functional_Group) %>%
  summarize(
    total_landings_mt = sum(weight_kg_corrected, na.rm = TRUE)/1000,
    .groups = "drop"
  ) %>%
  mutate(
    Major_Functional_Group= clean_names_lowercase(Major_Functional_Group),
    OHI_region_est= clean_names_lowercase(OHI_region_est)
  )
# 853 obs.
```

```{r Plot over time}
colnames(OpenOcean_Landings_summary)

OpenOcean_Landings_summary %>%
  filter(year >= 2014, year <= 2024) %>%
  ggplot(
    aes(
      x = year,
      y = log10(total_landings_mt),
      color = Major_Functional_Group,
      group = Major_Functional_Group
    )
  ) +
  geom_line() +
  theme_bw() +
  facet_wrap(~ OHI_region_est)
```


# STEP 3: Strategy 2 - Pull matching FIS scores ------------------------------

```{r Load in FIS_StockStatus scores as necessary}
StockStatus_Scores0<- read.csv("/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_Stock_Status_SubScores/Stock_Statuses_Compiled_df_Jun072026.csv")

# Species x year x region; Stock_Status is used as-is.

# CSV was previously titled: StockStatus_Estimates_Jun072026.csv
```

```{r Prep FIS scores}
StockStatus_Scores <- StockStatus_Scores0 %>%
  filter(year >= 2018, year <= 2024) %>%
  mutate(
    OHI_region_est = clean_names_lowercase(OHI_region_est),
    estimated_genus_name = clean_names_lowercase(estimated_genus_name)
  )
# 4130 obs.
```

# PATCH: Patch sardinops, penaeus, and litopenaeus 
```{r Patch penaeus major functional group}
StockStatus_Scores <- StockStatus_Scores %>%
  mutate(
    estimated_genus_name = case_when(
      estimated_genus_name == "penaeus" ~ "litopenaeus",
      TRUE ~ estimated_genus_name
    ),
    Major_Functional_Group = case_when(
      estimated_genus_name %in% c("penaeus", "litopenaeus") ~ "Shrimps, prawns",
      estimated_genus_name %in% c("sardinops") ~ "Anchovy & Sardine (small pelagics)",
      TRUE ~ Major_Functional_Group
    )
  )
```

## Stock_Status: Open Ocean Communities

```{r Create OpenOcean_StockStatus}
unique(StockStatus_Scores$Major_Functional_Group)

Open_Ocean_Taxa<- c("Pelagic marine fish (excl. anchovy & sardine)",
"Anchovy & Sardine (small pelagics)",
"Sharks",
"Cephalopods"
)

colnames(StockStatus_Scores)

OpenOcean_StockStatus<- StockStatus_Scores %>%
  filter(Major_Functional_Group %in% Open_Ocean_Taxa) %>%
    select(year, Major_Functional_Group,
           estimated_genus_name, estimated_family_name,
           OHI_region_est,
           total_weight_mt,
           StockStatus_final,
           StockStatus_source
         )

# 791 obs.
```

```{r Summarize landings}
colnames(OpenOcean_StockStatus)

OpenOcean_Status_summary<- OpenOcean_StockStatus %>%
  group_by(year, OHI_region_est, Major_Functional_Group) %>%
  summarize(
    mean_StockStatus = mean(StockStatus_final, na.rm = TRUE),
    total_weight_mt = sum(total_weight_mt, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Major_Functional_Group= clean_names_lowercase(Major_Functional_Group),
    OHI_region_est= clean_names_lowercase(OHI_region_est)
  )
# 150 obs.
```

```{r Plot over time}
colnames(OpenOcean_Status_summary)

OpenOcean_Status_summary %>%
  filter(year >= 2014, year <= 2024) %>%
  ggplot(
    aes(
      x = year,
      y = mean_StockStatus,
      color = Major_Functional_Group,
      group = Major_Functional_Group
    )
  ) +
  geom_line() +
  theme_bw() +
  facet_wrap(~ OHI_region_est)
```

## Stock_Status: Open Ocean Communities

```{r Create OpenOcean_StockStatus}
unique(StockStatus_Scores$Major_Functional_Group)

Open_Ocean_Taxa<- c("Pelagic marine fish (excl. anchovy & sardine)",
"Anchovy & Sardine (small pelagics)",
"Sharks",
"Cephalopods"
)

colnames(StockStatus_Scores)

OpenOcean_StockStatus<- StockStatus_Scores %>%
  filter(Major_Functional_Group %in% Open_Ocean_Taxa) %>%
    select(year, Major_Functional_Group,
           estimated_genus_name, estimated_family_name,
           OHI_region_est,
           total_weight_mt,
           StockStatus_final,
           StockStatus_source
         )

# 791 obs.
```

```{r Summarize landings}
colnames(OpenOcean_StockStatus)

OpenOcean_Status_summary<- OpenOcean_StockStatus %>%
  group_by(year, OHI_region_est, Major_Functional_Group) %>%
  summarize(
    mean_StockStatus = mean(StockStatus_final, na.rm = TRUE),
    total_weight_mt = sum(total_weight_mt, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Major_Functional_Group= clean_names_lowercase(Major_Functional_Group),
    OHI_region_est= clean_names_lowercase(OHI_region_est)
  )
# 150 obs.
```

```{r Plot over time}
colnames(OpenOcean_Status_summary)

OpenOcean_Status_summary %>%
  filter(year >= 2014, year <= 2024) %>%
  ggplot(
    aes(
      x = year,
      y = mean_StockStatus,
      color = Major_Functional_Group,
      group = Major_Functional_Group
    )
  ) +
  geom_line() +
  theme_bw() +
  facet_wrap(~ OHI_region_est)
```