---
title: "Mangrove Habitat Condition and Trend - Gulf of California OHI"
author: "Austin Martinez"
date: 03/26/2026
format: 
  html:
    code-fold: show
execute:
  echo: true
  message: false
  warning: false
---


```{r}
# Load in Libraries
library(sf)
library(dplyr)
library(here)
library(terra)
library(tidyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(tmap)
library(readr)
```

```{r}
# Read in OHI region boundaries
ohi_regions <- st_read(here("spatial", "ohi_regions", "goc_ohi_rgns.shp")) |> st_make_valid()

# Read in mangrove area shape files
mangrove_1981 <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/north_pacific_mangrove_land_use/pnoc1981gw/pnoc1981gw.shp"))
mangrove_2005 <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/north_pacific_mangrove_land_use/pnoc2005gw/pnoc2005gw.shp"))
mangrove_2010 <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/north_pacific_mangrove_land_use/pnoc2010gw/pnoc2010gw.shp"))
mangrove_2015 <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/north_pacific_mangrove_land_use/pn_oc2015gw/pn_oc2015gw.shp"))
mangrove_2020 <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/north_pacific_mangrove_land_use/pn_oc2020gw/pn_oc2020gw.shp"))

# Read in greenness data
greenness <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/mangrove_greenness/mantend_gw/mantend_gw.shp"))

```

```{r}
# Make all mangrove layers have the same CRS as OHI regions
mangrove_1981 <- st_transform(mangrove_1981, st_crs(ohi_regions))
mangrove_2005 <- st_transform(mangrove_2005, st_crs(ohi_regions))
mangrove_2010 <- st_transform(mangrove_2010, st_crs(ohi_regions))
mangrove_2015 <- st_transform(mangrove_2015, st_crs(ohi_regions))
mangrove_2020 <- st_transform(mangrove_2020, st_crs(ohi_regions))

# Make greenness layer have the same CRS as OHI regions
greenness <- st_transform(greenness, st_crs(ohi_regions))

```


## Step 1

To measure condition we compare current habitat distributions to historic distributions (1970/1980 in this case).

For each OHI region summarize the area of mangrove for each of the years of data. The spreadsheet at the end should include the variables:

```{r}
# There are not just mangroves in this data
print(unique(mangrove_1981$Descrip))

# Takes 35 mins to run
# Intersect each year with OHI regions, calculate area, and convert to km2
rgns_1981 <- st_intersection(mangrove_1981, ohi_regions) |>
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6, year = 1981)

rgns_2005 <- st_intersection(mangrove_2005, ohi_regions) |>
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6, year = 2005)

rgns_2010 <- st_intersection(mangrove_2010, ohi_regions) |>
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6, year = 2010)

rgns_2015 <- st_intersection(mangrove_2015, ohi_regions) |>
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6, year = 2015)

rgns_2020 <- st_intersection(mangrove_2020, ohi_regions) |>
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6, year = 2020)

# Filter for only 
rgns_1981 <- rgns_1981 |> filter(Descrip %in% c("Manglar", "Manglar perturbado"))
rgns_2005 <- rgns_2005 |> filter(Descrip %in% c("Manglar", "Manglar perturbado"))
rgns_2010 <- rgns_2010 |> filter(Descrip %in% c("Manglar", "Manglar perturbado"))
rgns_2015 <- rgns_2015 |> filter(Descrip %in% c("Manglar", "Manglar perturbado"))
rgns_2020 <- rgns_2020 |> filter(Descrip %in% c("Manglar", "Manglar perturbado"))
```
```{r}
tm_shape(rgns_2020) +
  tm_polygons(col = "red", border.col = "red", lwd = 0.5) +
tm_shape(buffer_ring) +
  tm_polygons(col = "lightblue", border.col = "lightblue", lwd = 0.5, alpha = 0.2)
```


```{r}
# Combine and summaries area per region per year
area_by_year <- bind_rows(rgns_1981, rgns_2005, rgns_2010, rgns_2015, rgns_2020) |>
  st_drop_geometry() |>
  group_by(rgn_id, year) |>
  summarise(area_km2 = sum(area_km2, na.rm = TRUE), .groups = "drop")
```

Then we can find condition for each region by dividing 2020 data by 1970/1980 data.

```{r}
# Dividing 2020 data by 1981 data to get condition
condition <- area_by_year |>
  filter(year %in% c(1981, 2020)) |>
  pivot_wider(names_from = year, values_from = area_km2) |>
  mutate(condition = `2020` / `1981`)
print(condition)
```

## Step 2

```{r}
# Subtract each of the year rasters from historic (1981)
change_2005 <- area_by_year |> 
  filter(year %in% c(1981, 2005)) |>
  pivot_wider(names_from = year, values_from = area_km2) |>
  mutate(area_change = `2005` - `1981`)

change_2010 <- area_by_year |> 
  filter(year %in% c(1981, 2010)) |>
  pivot_wider(names_from = year, values_from = area_km2) |>
  mutate(area_change = `2010` - `1981`)

change_2015 <- area_by_year |> 
  filter(year %in% c(1981, 2015)) |>
  pivot_wider(names_from = year, values_from = area_km2) |>
  mutate(area_change = `2015` - `1981`)

change_2020 <- area_by_year |> 
  filter(year %in% c(1981, 2020)) |>
  pivot_wider(names_from = year, values_from = area_km2) |>
  mutate(area_change = `2020` - `1981`)
```


```{r fig.width = 15, fig.height =8}
# Concatenate data frames
all_years_change <- bind_rows(
  change_2005 |> mutate(year = 2005),
  change_2010 |> mutate(year = 2010),
  change_2015 |> mutate(year = 2015),
  change_2020 |> mutate(year = 2020)
) 

# Plot the change over the years for each region
all_years_change |>
  mutate(rgn_id = gsub("_", " ", rgn_id)) |>
  ggplot(aes(x = year, y = area_change, color = rgn_id, group = rgn_id)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c(
    "Region 2" = "green",
    "Region 3" = "forestgreen",
    "Region 4" = "red",
    "Region 6" = "orange",
    "Region 7" = "purple",
    "Region 8" = "dodgerblue",
    "Region 9" = "chocolate"
  )) +
  labs(title = "Mangrove Area Change from 1981",
       y = "Change in Area (km²)", x = "Year", color = "Region") +
  theme_bw()
```

## Trend

We also need to get some idea of whether mangroves are getting better/worse in more recent years.

Use the above data to measure the yearly change in mangrove area each region across years.

```{r}
# measure the yearly change in mangrove area each region across years.
yearly_change <- area_by_year |>
  group_by(rgn_id) |>
  arrange(year) |>
  mutate(annual_change = area_km2 - lag(area_km2)) |>
  filter(!is.na(annual_change))

kable(yearly_change, col.names = c("Region", "Year", "Area (km²)", "Annual Change (km²)"), digits = 4) |>
  kable_styling(bootstrap_options = c("striped", "bordered")) |>
  row_spec(seq(0, nrow(yearly_change), by = 7), extra_css = "border-bottom: 1px solid black;")
```

But also, lets explore using the trend in greenness data (see #71). Summarize the amount of increasing/decreasing/constant in each ohi region and see if it makes sense to use this as a measure of trend.

```{r}
# Takes 30 mins to run
# Summarize the amount of increasing/decreasing/constant in each ohi region
greenness_trend <- st_intersection(greenness, ohi_regions) |>
  st_drop_geometry() |>
  group_by(rgn_id, direccion) |>
  summarise(total_ha = sum(ha, na.rm = TRUE), .groups = "drop")
```


```{r}
# Translate Decremento/Incremento to english and remove "_" from region names
greenness_trend |>
  mutate(
    rgn_id = gsub("_", " ", rgn_id),
    direccion = case_when(
      direccion == "Decremento" ~ "Decrease",
      direccion == "Incremento" ~ "Increase"
    )
  ) |>
  # Create a ktable
  kable(col.names = c("Region", "Direction", "Total Area (ha)"), digits = 4) |>
  kable_styling(bootstrap_options = c("striped", "bordered")) |>
  row_spec(seq(0, nrow(greenness_trend), by = 2), extra_css = "border-bottom: 1px solid black;")

```













# Comparing results to other data

### Using the Conabio data you have been looking at: Summarize for each region and year the area of mangroves that are disturbed vs. non-disturbed (I want to see if these can be used to assess condition and/or trend)

```{r}
# Summarize for each region and year the area of mangroves that are disturbed vs. non-disturbed
area_by_year_disturbed <- bind_rows(rgns_1981, rgns_2005, rgns_2010, rgns_2015, rgns_2020) |>
  st_drop_geometry() |>
  group_by(rgn_id, year, Descrip) |>
  summarise(area_km2 = sum(area_km2, na.rm = TRUE), .groups = "drop")

```




### Truns out there is a shapefile verion, so I am goint to use it instead of dealing with the tiles up above for a raster
.
```{r}
# Read in zenodo data for 1996
gmw_1996 <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/zenodo_mangrove/gmw_v3_1996_vec/gmw_v3_1996_vec.shp"))

# Match gmw_1996 CRS to ohi_regions
gmw_1996 <- st_transform(gmw_1996, st_crs(ohi_regions))

# Intersect gmw_1996 with OHI regions, calculate area, and convert to km2
rgns_gmw_1996 <- st_intersection(gmw_1996, ohi_regions) |>
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6, year = 1996)
```


```{r}
# Read in zenodo data for 2020
gmw_2020 <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/zenodo_mangrove/gmw_v3_2020_vec/gmw_v3_2020_vec.shp"))

# Match gmw_2020 CRS to ohi_regions
gmw_2020 <- st_transform(gmw_2020, st_crs(ohi_regions))

# Intersect gmw_2020 with OHI regions, calculate area, and convert to km2
rgns_gmw_2020 <- st_intersection(gmw_2020, ohi_regions) |>
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6, year = 2020)
```
```{r}
tm_shape(rgns_gmw_2020) +
  tm_polygons(col = "red", border.col = "red", lwd = 0.5) +
tm_shape(buffer_ring) +
  tm_polygons(col = "lightblue", border.col = "lightblue", lwd = 0.5, alpha = 0.2)
```



### Summarize Hem's mangrove data by region: /home/shares/ohi/OHI_GOC/_raw_data/mangrove_Hem/USV_AGRUPADO and compare the number of hectares to the CONABIO data for 2020.

    - "MANGLAR"
    - "VEGETACIÓN SECUNDARIA ARBÓREA DE MANGLAR",
    - "VEGETACIÓN SECUNDARIA ARBUSTIVA DE MANGLAR",
    - "VEGETACIÓN SECUNDARIA HERBÁCEA DE MANGLAR"
    
```{r}
# Read in Hem's data
hem_mangroves <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/mangrove_Hem/USV_AGRUPADO/USV_AGRUPADO.shp"))

# Match hem_mangroves CRS to ohi_regions
hem_mangroves <- st_transform(hem_mangroves, st_crs(ohi_regions))

# Filter out layers that aren't "MANGLAR" and make it valid
hem_mangroves_only <- hem_mangroves |>
  filter(DESCRIP_AG %in% c(
    "MANGLAR")) |>
  st_make_valid()

# Intersect hem_mangroves_only with OHI regions, calculate area, and convert to km2
rgns_hem <- st_intersection(hem_mangroves_only, ohi_regions) |>
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6)

# Print first 5 rows
head(rgns_hem)
```
```{r}
tm_shape(rgns_hem) +
  tm_polygons(col = "red", border.col = "red", lwd = 2) +
tm_shape(buffer_ring) +
  tm_polygons(col = "lightblue", border.col = "lightblue", lwd = 0.5, alpha = 0.3)
```

