################################################################################

Erica Ferrer, Ph.D.
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(stringr)
library(stringi)
library(purrr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(fuzzyjoin)
# Visualization
library(ggplot2)
library(plotly)
library(htmlwidgets)


setwd("/home/ferrer/ohi_goc_prep/")
```


# PART 1: Filter landings data and create skeleton for species key ------------------------------
First we need to correctly clean and filter landings data so that we can append species info and extract species info for a key.

## Filtering out key words from landing receipts:
```{r Load in GoC landings data}
GOC_Landings<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fis/v2025/int/juan_carlos/GoC_Landings_2000.present_EMF_created.rds"))
# n = 3716415 obs.
# NOTE TO SELF: THIS FILE WILL CHANGE WITH CLEAN UP!
```

```{r Is aleta de aleta an issue? Short answer, no.}
GOC_Landings %>%
  rename(recorded_taxa_name = species_name) %>%
  mutate(
    aleta_count = stringi::stri_count_fixed(recorded_taxa_name, "ALETA")
  ) %>%
  count(aleta_count)

# No, double aleta does not seem to be an issue...
```

```{r Filter out keywords from recorded_taxa_name to get recorded_taxa_name_clean}
# 1. Keyword list (excluding ALETA, handled separately) -----------------
keyword_list <- c(
  "S.C.", "S/C", "S",
  "ENTERO", "ENTERA", "IND", "ENT",
  "FRESCO", "FRESCA", "FCA", "FCO", "FC",
  "ECO",
  "CONG", "CONGELADO",
  "VIVO", "VIVA", "VIVAS", "VIVITA",
  "DE CULTIVO", "DE CULTIVO VIVO", "ABASTO RANCHO", "RANCHO",
  "HUEVA DE",
  "FILETE", "FILETE DE", "ECA", "FILETE DE ECO", "DESHUESADA", "CARNE", "CARNE DE",
  "DESCABEZADO", "DES", "DSC", "DESC", "DESV", "DESVICERADO",
  "CABEZA DE", "BUCHE", "BUCHE DE", "TRIPA DE", "COLAS DE", "MANOS DE", "MANOS",
  "ANCAS DE", "ANCAS", "ANCA",
  "PIEL DE ECA", "PIEL DE SECA", "PIEL DE", "PIEL",
  "SECO DE", "SECA", "SECA DE", "SECO", "SECO DE",
  "COCINERO", "COCIDO", "COCIDA", "COCIDAS", "ASADO",
  # Remove ALETA from main keywords; handled separately
  #"ALETA", "ALETA ECA DE", "ALETA FCA DE",
  "DE ALTA MAR",
  "CONCHA DE", "CONCHA",
  "POLVO DE",
  "CARNADA", "P/CARNADA",
  "DE AGUA DULCE",
  "MANTO", "MANTO DE", "PULPA DE", "PULPA FRESCA",
  "PULPO DE", "TENTACULOS DE", "TUBO DE", "TUBO",
  "NO APTO PARA EMPAQUE"
)

# Longest-first ordering
keyword_list <- keyword_list[order(nchar(keyword_list), decreasing = TRUE)]

# Escape regex meta-characters
keyword_list_escaped <- stringi::stri_replace_all_regex(
  keyword_list, "([.|(){}+*?^$\\[\\]\\\\])", "\\\\$1"
)

# Final regex
keyword_pattern <- paste0(
  "(?<![A-Za-z0-9])(",
  paste(keyword_list_escaped, collapse = "|"),
  ")(?![A-Za-z0-9])"
)


# 2. Special ALETA phrases (handled separately from other key words) -----------------

special_aleta_adj <- c("BLANCA", "AMARILLA", "CORTA", "AZUL")

remove_aleta_except_special <- function(x) {
  stringi::stri_replace_all_regex(
    x,
    pattern = paste0("\\bALETA\\b(?!\\s+(?:", paste(special_aleta_adj, collapse = "|"), "))"),
    replacement = "",
    vectorize_all = FALSE
  )
}


# 3. MAIN PIPELINE (optimized) -----------------

GOC_Landings_clean <- GOC_Landings %>%
  
  # Rename columns
  rename(
    recorded_taxa_name = species_name,
    recorded_taxa_group = main_species_group
  ) %>%
  
  # Extract keywords
  mutate(
    keywords_list = stringi::stri_extract_all_regex(recorded_taxa_name, keyword_pattern)
  ) %>%
  mutate(keywords_list = map(keywords_list, ~ { length(.x) <- 5; .x })) %>%
  unnest_wider(keywords_list, names_sep = "_") %>%
  rename_with(~ paste0("keywords_", 1:5), starts_with("keywords_list_")) %>%
  
  # Clean recorded_taxa_name
  mutate(
    recorded_taxa_name_clean = recorded_taxa_name %>%
      remove_aleta_except_special() %>%                      # Remove ALETA unless special
      stringi::stri_replace_all_regex(keyword_pattern, "") %>% # Remove other keywords
      stringi::stri_replace_all_regex("\\b(Y|DE|P/)\\b", "") %>% # Remove standalone Y/DE/P/
      stringi::stri_replace_all_regex("[.,/\\\\]+", " ") %>%      # Remove punctuation
      str_squish()
  ) %>%
  
  # Extract scientific name
  mutate(
    recorded_scientific_name = stringi::stri_extract_first_regex(recorded_taxa_name_clean, "\\(([^)]+)\\)") %>%
      stringi::stri_replace_all_regex("\\(|\\)", ""),
    scientific_name_recorded_T.F = !is.na(recorded_scientific_name),
    recorded_scientific_name = if_else(
      scientific_name_recorded_T.F,
      stringi::stri_trans_totitle(recorded_scientific_name),
      NA_character_
    )
  ) %>%
  
  # Extract common name
  mutate(
    common_name_clean = stringi::stri_trim_both(
      stringi::stri_replace_first_regex(recorded_taxa_name_clean, "\\s*\\([^)]*\\)", "")
    ),
    common_name_clean = tolower(common_name_clean)
  )


# 4. REORDER COLUMNS -----------------

GOC_Landings_clean <- GOC_Landings_clean %>%

  select(
    # General info
    state, office_name, landing_site, landing_site_key, landing_site_primary, GoC, lat, long,
    year, month,
    receipt_date_fixed, period_start_fixed, period_end_fixed,
    fleet, acuaculture_production, eu_rnpa, eu_name, vessel_rnpa, vessel_name,
    
    # Species info
    recorded_taxa_group, recorded_taxa_name,
    
    # Cleaned / derived names
    recorded_taxa_name_clean, recorded_scientific_name,
    scientific_name_recorded_T.F, common_name_clean,
    
    # Keywords
    keywords_1, keywords_2, keywords_3, keywords_4, keywords_5,
    
    # Landing details
    landed_weight, live_weight, value
 
  )

# How many common names are there in this dataset?
length(unique(GOC_Landings_clean$common_name_clean))
length(unique(GOC_Landings_clean$recorded_taxa_group))
# 734 common names (versus 973 with the less comprehensive keywords_list), over 57 taxonomic groups
```

```{r Check that this worked}
TEST <- GOC_Landings_clean %>%
  filter(str_detect(recorded_taxa_name, "ATUN ALETA AMARILLA"))
```
# Yes, it seems to have worked!

## Save GOC_Landings_cleaned rds file
# Revising this as of Mar 25, 2025: Fixing unicore on estimated_species_name
```{r Save this dataframe - landings keywords & species for top groupings: Nov 10 2025}
GOC_Landings_clean<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Landings_KeywordsParsed_2000.present_EMF.Mar232026.rds"))
# 3622776 obs., 34 vars.
```
# PART 1 COMPLETE.


# PART 2: Create MASTER SPECIES KEY: by merging the skeleton with keys from Mar, & Hem ------------------------------
```{r Skeleton of master key from GOC_Landings_clean dataframe}
Master_SpKey_skeleton<- GOC_Landings_clean %>% select(recorded_taxa_group, recorded_taxa_name_clean, common_name_clean, scientific_name_recorded_T.F, recorded_scientific_name) %>% distinct()
# 807 observations, 5 variables.
```

```{r Mar and Hem Species Keys}
Erica_SpKey <- read_csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/Species_Keys/Erica_CONAPESCA_SpKey_IMPERFECT_Nov052025.csv")
# # 600 obs of 17 variables

Hem_SpKey <- read_csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/Species_Keys/Hem_CONAPESCA_SpKey_2018.csv")
 # 8232 obs of 3 variables

Mar_SpKey <- read_csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/Species_Keys/Mar_CONAPESCA_SpKey.csv")
 # 443 obs of 30 variables
```

```{r Prepare species keys for merging}
# MAR's KEY - THE IHH KEY
Mar_SpKey <- Mar_SpKey %>%
  mutate(common_name = tolower(common_name))

Mar_SpKey_filtered <- Mar_SpKey %>% select(common_name, fishery, highest_taxonomic_id, highest_taxonomic_resolution, species, genus, family, order, class, functional_major_groups_sm= IHH_functional_major_groups_sm)

Mar_SpKey_filtered <- Mar_SpKey_filtered %>%
  select(common_name, fishery, highest_taxonomic_id, highest_taxonomic_resolution,
         species, genus, family, order, class, functional_major_groups_sm) %>%
  rename_with(~ paste0("IHH_", .x))
# 443 obs of 10 variables

# HEM's KEY
Hem_SpKey <- Hem_SpKey %>%
  mutate(common_name = tolower(common_name))

# Fix the sharks...
Hem_SpKey_filtered <- Hem_SpKey %>%
  mutate(
    common_name = if_else(
      common_name == "tiburon",   # exact match only
      "tiburon se",
      common_name
    )
  )

# Fix the rockfishes...
Hem_SpKey_filtered <- Hem_SpKey_filtered %>%
  mutate(
    common_name = if_else(
      common_name == "rocote",   # exact match only
      "rocote se",
      common_name
    )
  )

# Fix the grunts...
Hem_SpKey_filtered <- Hem_SpKey_filtered %>%
  mutate(
    common_name = if_else(
      common_name == "ronco",   # exact match only
      "ronco se",
      common_name
    )
  )

Hem_SpKey_filtered<- Hem_SpKey_filtered %>%
  select(Hem_common_name= common_name, species) %>% 
  distinct()

Hem_SpKey_filtered <- Hem_SpKey_filtered %>%
  mutate(
    species = if_else(
      str_detect(Hem_common_name, "\\bse\\b"),  # \\b ensures "se" is a whole word
      "undetermined",
      species
    )
  )

# Fix the species de ornato...
Hem_SpKey_filtered <- Hem_SpKey_filtered %>%
  mutate(
    species = if_else(
      Hem_common_name == "especies de ornato",   # exact match only
      "undetermined",
      species
    )
  )

Hem_SpKey_filtered<- Hem_SpKey_filtered %>%
  select(Hem_common_name, species) %>%
  distinct() %>%
  group_by(Hem_common_name) %>%
  mutate(species_num = row_number()) %>%
  ungroup()

# 1235 obs., 3 vars.

Hem_SpKey_filtered <- Hem_SpKey_filtered %>%
  pivot_wider(
    names_from = species_num,
    values_from = species,
    names_prefix = "Hem_species_"
  )
# 695 obs., 21 vars.
```

```{r Merge}
merged_SpKey_0 <- left_join(
  Master_SpKey_skeleton,
  Mar_SpKey_filtered,
  by = c("common_name_clean" = "IHH_common_name")
)

Master_SpKey <- left_join(
  merged_SpKey_0,
  Hem_SpKey_filtered,
  by = c("common_name_clean" = "Hem_common_name")
)
# 807 obs., 34 vars.
```

```{r MASTER SPECIES KEY (not fully cleaned)}
# write.csv(Master_SpKey, "/home/ferrer/ohi_goc_prep/general/Species_Keys/Master_SpKey_Dec012025.csv", row.names = FALSE)
```

# PART 2 COMPLETE.

# PART 3: Clean up and simplify the MASTER SPECIES KEY  ------------------------------
```{r Simplify the master species key}
# WHAT IS THE LOGIC?

# If scientific_name_recorded_T.F == TRUE, then estimated_species_name = scientific_name_recorded and species_est_certainty = "virtually certain" and species_est_reasoning = "name was recorded in CONAPESCA".

# If scientific_name_recorded_T.F == FALSE, then compare IHH_species to Hem_species_.

# > If Hem_species_1 is NOT NA & IHH_species is NOT NA & IHH_species is a fuzzy match to any of Hem_species_ >> then estimated_species_name = IHH_species, species_est_certainty = "high" and species_est_reasoning = "logical match in both keys".

# > If Hem_species_1 is NOT NA & IHH_species is NOT NA, but IHH_species is NOT a fuzzy match to any of Hem_species_ >> then estimated_species_name = IHH_species, species_est_certainty = "low", and species_est_reasoning = "mismatch between keys".

# > If Hem_species_1 is NA but IHH_species is NOT NA >> then estimated_species_name = IHH_species, species_est_certainty = "high", and  = "match to IHH key".

# > If Hem_species_1 is NOT NA & IHH_species is NA, then consider the species dictated by the rest of Hem's key: if Hem_species_2 is NA, then estimated_species_name = Hem_species_1, species_est_certainty = "high", and species_est_reasoning = "match to Hem's key"; but if Hem_species_2 is not NA, then estimated_species_name = "unknown", species_est_certainty = "medium", and species_est_reasoning = "conflicting matches in Hem's key and no data from IHH" 


# APPLY THE LOGIC:

# Hem species columns
hem_cols <- paste0("Hem_species_", 1:20)

# Compute estimated_species_name, species_est_certainty, and species_est_reasoning
Master_SpKey_clean <- Master_SpKey %>%
  rowwise() %>%
  mutate(
    hem_values = list(na.omit(c_across(all_of(hem_cols)))),  # all non-NA Hem species
    
    # Determine if IHH_species matches any Hem_species_ fuzzily
    ihh_matches_hem = if(!is.na(IHH_species) & length(hem_values) > 0) {
      any(stringdist::stringdist(IHH_species, hem_values, method = "jw") < 0.15)
    } else {
      FALSE
    },
    
    # Main estimation logic
    estimated_species_name = case_when(
      
      scientific_name_recorded_T.F ~ recorded_scientific_name,  # Case 1
      
      !is.na(IHH_species) & length(hem_values) > 0 & ihh_matches_hem ~ IHH_species,  # Case 2
      
      !is.na(IHH_species) & length(hem_values) > 0 & !ihh_matches_hem ~ IHH_species, # Case 3
      
      !is.na(IHH_species) & length(hem_values) == 0 ~ IHH_species,                   # Case 4
      
      is.na(IHH_species) & length(hem_values) == 1 ~ hem_values[1],                  # Case 5a
      
      is.na(IHH_species) & length(hem_values) > 1 ~ "unknown",                       # Case 5b
      
      TRUE ~ "unknown"
    ),
    
    species_est_certainty = case_when(
      scientific_name_recorded_T.F ~ "virtually certain",       # Case 1
      !is.na(IHH_species) & length(hem_values) > 0 & ihh_matches_hem ~ "high confidence",    # Case 2
      !is.na(IHH_species) & length(hem_values) > 0 & !ihh_matches_hem ~ "low confidence",    # Case 3
      !is.na(IHH_species) & length(hem_values) == 0 ~ "high confidence",                     # Case 4
      is.na(IHH_species) & length(hem_values) == 1 ~ "high confidence",                       # Case 5a
      is.na(IHH_species) & length(hem_values) > 1 ~ "medium confidence",                       # Case 5b
      TRUE ~ "not applicable"
    ),
    
    species_est_reasoning = case_when(
      scientific_name_recorded_T.F ~ "name was recorded in CONAPESCA",
      !is.na(IHH_species) & length(hem_values) > 0 & ihh_matches_hem ~ "logical match in both keys",
      !is.na(IHH_species) & length(hem_values) > 0 & !ihh_matches_hem ~ "mismatch between keys",
      !is.na(IHH_species) & length(hem_values) == 0 ~ "match to IHH key",
      is.na(IHH_species) & length(hem_values) == 1 ~ "match to Hem's key",
      is.na(IHH_species) & length(hem_values) > 1 ~ "conflicting matches in Hem's key and no data from IHH",
      TRUE ~ "no data in keys"
    )
  ) %>%
  ungroup() %>%
  select(-hem_values, -ihh_matches_hem)  # remove temporary helper columns
```

```{r NORMALIZE UNICODE and RE-ESTIMATE estimated_genus_name}
#### UNDER CONSTRUCTION: NORMALIZE UNICODE and RE-ESTIMATE estimated_genus_name ####
Master_SpKey_clean <- Master_SpKey_clean %>%
  mutate(
    estimated_species_name = stri_trans_nfkc(estimated_species_name), # normalize Unicode
    estimated_species_name = trimws(estimated_species_name)           # remove leading/trailing spaces
  )
# 807 obs.
```


```{r A few summary stats}
sum(is.na(Master_SpKey_clean$estimated_species_name)) # 0 rows without a species estimate
sum(Master_SpKey_clean$estimated_species_name=="unknown") # 246 species unknown
subset<- Master_SpKey_clean %>%
  filter(estimated_species_name=="unknown")

sum(is.na(Master_SpKey_clean$species_est_certainty)) # 0 rows without a species certainty estimate
table(Master_SpKey_clean$species_est_certainty)
  # high confidence    low confidence medium confidence    not applicable virtually certain 
  #             317               168                29               217                76 

sum(is.na(Master_SpKey_clean$species_est_reasoning)) # 0 rows without a reasoning
table(Master_SpKey_clean$species_est_reasoning)

# conflicting matches in Hem's key and no data from IHH 
#                                                    29 
#                            logical match in both keys 
#                                                   134 
#                                    match to Hem's key 
#                                                   117 
#                                      match to IHH key 
#                                                    66 
#                                 mismatch between keys 
#                                                   168 
#                        name was recorded in CONAPESCA 
#                                                    76 
#                                       no data in keys 
#                                                   217 
```

```{r Let's review those that are uncertain and clarify them if possible}
Master_SpKey_unknowns<- Master_SpKey_clean %>% filter(species_est_certainty== "not applicable")

# Okay, a few of these unknowns can probably be resolved with my key, which I would imagine is slightly less reliable than Mar's and Hem's. Use my key to resolve these unknowns and figure out the rest later. (NOTE TO SELF)

# Prepare my key for use:
Erica_SpKey_filtered<-  Erica_SpKey %>%
  filter(is.na(recorded_scientific_name)) %>%
  mutate(recorded_species_name_clean = tolower(recorded_species_name_clean)) %>%
  rename(Erica_common_name = recorded_species_name_clean, recorded_taxa_group = main_species_group)

Erica_SpKey_filtered<- Erica_SpKey_filtered %>% 
  select(recorded_taxa_group, Erica_common_name, species_estimated, genus_estimated, family_estimated, source_of_estimate) %>% 
  distinct()
# 222 obs., 6 vars.

# Control against many-to-many matches by getting rid of rows in my key that are nearly identical
Erica_SpKey_filtered <- Erica_SpKey_filtered %>%
  group_by(recorded_taxa_group, Erica_common_name) %>%
  slice(1) %>% # keep first match
  ungroup()

# Only add in my key info when IHH and Hem key info is missing: 
Master_SpKey_clean <- Master_SpKey_clean %>%
  # Separate unknowns from knowns
  mutate(lookup_needed = estimated_species_name == "unknown") %>%
  
  # Left join only where lookup_needed is TRUE
  left_join(
    Erica_SpKey_filtered,
    by = c("recorded_taxa_group", "common_name_clean" = "Erica_common_name")
  ) %>%
  
  # Replace estimated_species_name only where it was unknown but a match exists
  mutate(
    estimated_species_name = if_else(
      lookup_needed & !is.na(species_estimated),
      species_estimated,
      estimated_species_name
    ),
    # Optionally, update the certainty / reasoning
    species_est_certainty = if_else(
      lookup_needed & !is.na(species_estimated),
      "medium confidence",
      species_est_certainty
    ),
    species_est_reasoning = if_else(
      lookup_needed & !is.na(species_estimated),
      "match to Erica's key",
      species_est_reasoning
    )
  ) %>%
  
  # Remove temporary and unnecessary columns
  select(-lookup_needed, -species_estimated)

# 807 obs. (as expected.) and 40 vars.

#### UNDER CONSTRUCTION: NORMALIZE UNICODE and RE-ESTIMATE estimated_genus_name ####
Master_SpKey_clean <- Master_SpKey_clean %>%
  mutate(
    estimated_species_name = stri_trans_nfkc(estimated_species_name), # normalize Unicode
    estimated_species_name = trimws(estimated_species_name)           # remove leading/trailing spaces
  )
# 807 obs., 40 vars.
```

```{r New summary stats}
sum(Master_SpKey_clean$estimated_species_name=="unknown") # 246 -> 233 species unknown

table(Master_SpKey_clean$species_est_certainty)
  # high confidence    low confidence medium confidence    not applicable virtually certain 
  #             317               168                39               207                76 

table(Master_SpKey_clean$species_est_reasoning)

# conflicting matches in Hem's key and no data from IHH 
#                                                    26 
#                            logical match in both keys 
#                                                   134 
#                                  match to Erica's key 
#                                                    13 
#                                    match to Hem's key 
#                                                   117 
#                                      match to IHH key 
#                                                    66 
#                                 mismatch between keys 
#                                                   168 
#                        name was recorded in CONAPESCA 
#                                                    76 
#                                       no data in keys 
#                                                   207
```

```{r Now extract information about genus and family, where possible}
# WHAT IS THE LOGIC?

# If species is known:
# → estimated_genus_name = first word of species

# If species unknown but IHH_genus available:
# → estimated_genus_name = IHH_genus

# If both unavailable:
# → estimated_genus_name = genus unknown

# If IHH_family exists:
# → estimated_family_name = IHH_family

# Else if Erica_family_estimated exists:
# → estimated_family_name = Erica_family_estimated

# Else if Erica_SpKey gives exactly one family per larger taxonomic grouping
# → estimated_family_name = Erica_family_estimated
# → reasoning: “unique family match from Erica's key”

# Else
# → estimated_family_name = family unknown


# APPLY THE LOGIC:
# --- Build the Erica single-family lookup -------------------------------------

Erica_SpKey_simple <- Erica_SpKey %>% 
  select(
    recorded_taxa_group = main_species_group, 
    Erica_family_estimated = family_estimated
  ) %>%
  distinct() %>%
  group_by(recorded_taxa_group) %>%
  filter(n_distinct(Erica_family_estimated) == 1) %>%
  ungroup()


Master_SpKey_clean <- Master_SpKey_clean %>%
  
  # ----------------------- GENUS LOGIC ----------------------------------------
  mutate(
    estimated_genus_name = case_when(
      !is.na(estimated_species_name) & estimated_species_name != "unknown" ~ 
        word(estimated_species_name, 1),             # species known → genus = first word
      !is.na(IHH_genus) ~ IHH_genus,                 # fallback: IHH genus
      TRUE ~ NA_character_                           # final fallback: genus unknown
    ),
    
    genus_est_reasoning = case_when(
      !is.na(estimated_species_name) & estimated_species_name != "unknown" ~
        "genus derived from known species",
      !is.na(IHH_genus) ~
        "genus derived from IHH key",
      TRUE ~
        "genus unknown"
    )
  ) %>%
  
  
  # ----------------------- FAMILY LOGIC ---------------------------------------
  left_join(Erica_SpKey_simple, by = "recorded_taxa_group") %>%
  mutate(
    estimated_family_name = case_when(
      !is.na(IHH_family) ~ IHH_family,                         # 1. direct from IHH
      !is.na(Erica_family_estimated) ~ Erica_family_estimated, # 2. unique Erica match
      !is.na(estimated_genus_name) ~ 
        "family missing but genus available",                 # 3. genus known but no family
      TRUE ~ NA_character_                                     # 4. completely unknown
    ),
    
    family_est_reasoning = case_when(
      !is.na(IHH_family) ~ "family derived from IHH key",
      !is.na(Erica_family_estimated) ~ "unique family match from Erica's key",
      !is.na(estimated_genus_name) ~ "family unknown but genus available",
      TRUE ~ "family unknown"
    )
  ) %>%
  
  # remove columns that aren't necessary
  select(-Erica_family_estimated, -genus_estimated, -family_estimated, -source_of_estimate)
```

```{r Clean MASTER SPECIES KEY}
write.csv(Master_SpKey_clean, "/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/Species_Keys/CleanedMaster_SpKey_Mar252026.csv", row.names = FALSE)
```

# PART 3 COMPLETE. NEW SAVE AS OF Mar 25 2026.

# FIXING MAJOR FUNCTIONAL GROUPS
# NOTE TO SELF: UNDER CONSTRUCTION (Mar 25 2026)

```{r Load MasterSpKey_pending}
MasterSpKey_pending <- read_csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/Species_Keys/CleanedMaster_SpKey_Mar252026.csv")
# 807 obs of 41 variables

table(MasterSpKey_pending$recorded_taxa_group)
```

```{r Assign Major_Functional_Group computationally for all taxa exluding Special_Groups like otras nei ornatos etc.}
# Split the OTRAS from everything else. They will require additional manual work.
Special_groups <- c("OTRAS", "ESPECIE GENERICA", "FAUNA", "ORNATO", "PECES DE ORNATO", "NEI")

NONOTRAS_MajorGroupKey <- MasterSpKey_pending %>%
  filter(!recorded_taxa_group %in% Special_groups) %>%
  select(recorded_taxa_group, estimated_genus_name, IHH_functional_major_groups_sm) %>%
  distinct(recorded_taxa_group, estimated_genus_name, IHH_functional_major_groups_sm, .keep_all = TRUE)
# 246 obs., 3 vars.

NONOTRAS_MajorGroupKey$Major_Functional_Group = NA_character_  # initialize the Major_Functional_Group column
NONOTRAS_MajorGroupKey$Assignment_for_Major_Functional_Group = NA_character_  # initialize the Assignment_for_Major_Functional_Group column

# Here's the logic:
# Step 1 → Assignment based on direct IHH data (strongest signal)
# Step 2 → GENUS-LEVEL assignment (stronger signal)
# Step 3 → for genus-level consistency
# Step 4 → GROUP-LEVEL fallback (only if genus missing)

# STEP 1: Assign Major_Functional_Group values based solely on the values assigned by the IHH major functional groups ----
NONOTRAS_MajorGroupKey <- NONOTRAS_MajorGroupKey %>%
  mutate(
    Major_Functional_Group = case_when(
      !is.na(IHH_functional_major_groups_sm) ~ IHH_functional_major_groups_sm,
      TRUE ~ Major_Functional_Group
    ),
    Assignment_for_Major_Functional_Group = case_when(
      !is.na(IHH_functional_major_groups_sm) ~ "Assigned_Step_1: based directly on IHH_functional_major_group assignment for this genus in this group.",
      TRUE ~ Assignment_for_Major_Functional_Group
    )
  )

# STEP 2: Assign Major_Functional_Group values based on genus-level information (strong signal)
NONOTRAS_MajorGroupKey <- NONOTRAS_MajorGroupKey %>%
  group_by(estimated_genus_name) %>%
  mutate(
    
    # Determine the most common genus (genus_major) ONLY IF genus info is present (not NA nor NEI)
    genus_major = {
      genus_value <- first(estimated_genus_name)  # safer than cur_group()

      if (!is.na(genus_value) && genus_value != "NEI") {

        # Use IHH first (most reliable), fallback to existing Major_Functional_Group
        x <- IHH_functional_major_groups_sm
        x <- x[!is.na(x) & x != ""]

        if (length(x) == 0) {
          x <- Major_Functional_Group
          x <- x[!is.na(x) & x != ""]
        }

        if (length(x) == 0) NA_character_ else {
          names(which.max(table(x)))
        }

      } else {
        NA_character_  # cannot derive functional group
      }
    },

    # Assign Major_Functional_Group based on genus_major only if missing
    Major_Functional_Group = case_when(
      !is.na(genus_major) & is.na(Major_Functional_Group) ~ genus_major,
      TRUE ~ Major_Functional_Group
    ),

    # Update assignment label
    Assignment_for_Major_Functional_Group = case_when(
      !is.na(genus_major) & is.na(Assignment_for_Major_Functional_Group) & !is.na(Major_Functional_Group) ~
        "Assigned_Step_2: based on genus-level details; interpolated Major_Functional_Group based on records with matching genus name.",
      TRUE ~ Assignment_for_Major_Functional_Group
    )
  ) %>%
  ungroup() %>%
  select(-genus_major)

# STEP 3: Force genus-level consistency (except when genus = NA or NEI) ----
NONOTRAS_MajorGroupKey <- NONOTRAS_MajorGroupKey %>%
  # Keep original Major_Functional_Group for comparison
  mutate(old_Major_Functional_Group = Major_Functional_Group) %>%
  
  # Compute genus-level consensus for non-NA/NEI genera
  group_by(estimated_genus_name) %>%
  mutate(
    genus_consensus_group = if(!is.na(first(estimated_genus_name)) & first(estimated_genus_name) != "NEI") {
      # Get mode of Major_Functional_Group in the genus
      vals <- Major_Functional_Group[!is.na(Major_Functional_Group)]
      if(length(vals) == 0) NA_character_ else names(sort(table(vals), decreasing = TRUE))[1]
    } else {
      NA_character_
    }
  ) %>%
  ungroup() %>%
  
  # Apply the consensus to all rows in the genus (if applicable)
  mutate(
    Major_Functional_Group = if_else(
      !is.na(genus_consensus_group),
      genus_consensus_group,
      Major_Functional_Group
    ),
    
    # Update Assignment_for_Major_Functional_Group if the value changed
    Assignment_for_Major_Functional_Group = if_else(
      !is.na(genus_consensus_group) &
        (is.na(old_Major_Functional_Group) | old_Major_Functional_Group != Major_Functional_Group),
      "Assigned_Step_3: forced genus-level consistency; if genus is equal then assume Major_Functional_Group is equal too.",
      Assignment_for_Major_Functional_Group
    )
  ) %>%
  select(-old_Major_Functional_Group, -genus_consensus_group)


# STEP 4: Assign Major_Functional_Group values based on recorded_taxa_group fallback (weaker signal)
# > If we don’t know anything yet, borrow a typical value from a reference to IHH_functional_major_groups.
NONOTRAS_MajorGroupKey <- NONOTRAS_MajorGroupKey %>%
  group_by(recorded_taxa_group) %>%
  mutate(
    
    # Take the most frequent non-NA value for IHH_functional_major_groups within the group
    group_major = {
      x <- IHH_functional_major_groups_sm
      x <- x[!is.na(x) & x != ""]
      if (length(x) == 0) NA_character_ else {
        names(which.max(table(x)))
      }
    },

    # Assign only to non-special groups and only if missing
    Major_Functional_Group = if_else(
      !(recorded_taxa_group %in% Special_groups) & is.na(Major_Functional_Group),
      group_major,
      Major_Functional_Group
    ),

    # Update assignment label
    Assignment_for_Major_Functional_Group = if_else(
      !(recorded_taxa_group %in% Special_groups) & is.na(Assignment_for_Major_Functional_Group) & !is.na(Major_Functional_Group),
      "Assigned_Step_4: assigned Major_Functional_Group based on the average IHH functional group associated with this recorded_taxa_group.",
      Assignment_for_Major_Functional_Group
    )
  ) %>%
  ungroup() %>%
  select(-group_major)

# STEP 5: Enforce consistency for Major_Functional_Group at the recorded_taxa_group level in cases where genus = NA or NEI ----
# > Now that we have some real assignments, make the group internally consistent.
NONOTRAS_MajorGroupKey <- NONOTRAS_MajorGroupKey %>%
  mutate(old_Major_Functional_Group = Major_Functional_Group) %>%
  
  group_by(recorded_taxa_group) %>%
  mutate(
    taxa_group_consensus = if (any(is.na(estimated_genus_name) | estimated_genus_name == "NEI")) {
      vals <- Major_Functional_Group[!is.na(Major_Functional_Group) & Major_Functional_Group != "NEI"]
      if (length(vals) == 0) NA_character_ else names(sort(table(vals), decreasing = TRUE))[1]
    } else {
      NA_character_
    },

    # Apply only when genus or Major_Functional_Group is NA or NEI
    Major_Functional_Group = if_else(
      !is.na(taxa_group_consensus) & 
        # THIS IS THE LINE!!
        (is.na(estimated_genus_name) | estimated_genus_name == "NEI" | is.na(Major_Functional_Group) | Major_Functional_Group == "NEI"),
      taxa_group_consensus,
      Major_Functional_Group
    ),
    
    # Step 5 assignment: robust to NAs
    Assignment_for_Major_Functional_Group = if_else(
      !is.na(taxa_group_consensus) &
        (
          (is.na(old_Major_Functional_Group) & !is.na(Major_Functional_Group)) |
          (old_Major_Functional_Group == "NEI" & Major_Functional_Group != "NEI") |
          (!is.na(old_Major_Functional_Group) & !is.na(Major_Functional_Group) & old_Major_Functional_Group != Major_Functional_Group)
        ),
      "Assigned_Step_5: enforced Major_Functional_Group consistency across recorded_taxa_group.",
      Assignment_for_Major_Functional_Group
    )
  ) %>%
  ungroup() %>%
  select(-old_Major_Functional_Group, -taxa_group_consensus)


# STEP 6: Manual patches for specific taxonomic groups and two typos ----
unique(NONOTRAS_MajorGroupKey$recorded_taxa_group)
# List all recorded_taxa_group values where Major_Functional_Group is NA
unique(NONOTRAS_MajorGroupKey$recorded_taxa_group[is.na(NONOTRAS_MajorGroupKey$Major_Functional_Group)])

# Split_Groups: groups of fish for which it is not possible to assign a taxonomic group with high accuracy based on recorded_taxa_group alone...
# NOTE TO SELF: COME BACK TO THESE GROUPS WHEN / IF THERE'S TIME. Categorization across groups of fishes are still a little wrong tbh. - Mar 23 2026
# A note about PETO: I feel pretty confident that PETO is not shark. I think Hem's key for Peto carrito is incorrect...
# Because of this Lamna is incorrectly assigned as a demersal fish and not a shark.
SplitGroup_list <- c("MOJARRA", "BAGRE", "CARPA", "PECES ORNATO")

NONOTRAS_MajorGroupKey <- NONOTRAS_MajorGroupKey %>%
  mutate(
    # Assign Major_Functional_Group and Assignment_for_Major_Functional_Group in one go:
    
    Major_Functional_Group = case_when(
      
      # Split groups, cannot be fully categorized:
      recorded_taxa_group %in% SplitGroup_list & (is.na(estimated_genus_name) | estimated_genus_name == "NEI" | is.na(IHH_functional_major_groups_sm)) ~ "Marine or freshwater fishes NEI",
      
      # Specific taxa and groups for which there is no IHH data:
      recorded_taxa_group == "ANCHOVETA" ~ "Pelagic marine fishes",
      recorded_taxa_group %in% c("ALGAS", "SARGAZO") ~ "Marine algae",
      recorded_taxa_group %in% c("PEPINO DE MAR", "ERIZO") ~ "Echinoderms",
      recorded_taxa_group %in% c("TIBURON", "CAZON") ~ "Sharks",
      recorded_taxa_group %in% c("CINTILLA") ~ "Pelagic marine fishes",
      recorded_taxa_group %in% c("TRUCHA") ~ "Marine or freshwater fishes NEI",
      
      # Typos:
      # estimated_genus_name == "Sphoeriodes" ~ "Marine fishes NEI",
      estimated_genus_name == "Lujtanus" ~ "Demersal marine fishes",
      TRUE ~ Major_Functional_Group  # keep existing if no condition matches
    ),
    
    Assignment_for_Major_Functional_Group = case_when(
      # For all manual patch conditions, assign the label
      (recorded_taxa_group %in% SplitGroup_list & (is.na(estimated_genus_name) | estimated_genus_name == "NEI")) ~ "Assigned_Step_6: manual patch",
      recorded_taxa_group == "ANCHOVETA" ~ "Assigned_Step_6: manual patch",
      recorded_taxa_group %in% c("ALGAS", "SARGAZO") ~ "Assigned_Step_6: manual patch",
      recorded_taxa_group %in% c("PEPINO DE MAR", "ERIZO") ~ "Assigned_Step_6: manual patch",
      recorded_taxa_group %in% c("TIBURON", "CAZON") ~ "Assigned_Step_6: manual patch",
      recorded_taxa_group %in% c("CINTILLA") ~ "Assigned_Step_6: manual patch",
      recorded_taxa_group %in% c("TRUCHA") ~ "Assigned_Step_6: manual patch",
      # estimated_genus_name == "Sphoeriodes" ~ "Assigned_Step_6: manual patch",
      estimated_genus_name == "Lujtanus" ~ "Assigned_Step_6: manual patch or override",
      TRUE ~ Assignment_for_Major_Functional_Group  # keep existing if no condition matches
    )
  )

# STEP 3 REPEATED: Reinforce genus-level consistency (except when genus = NA or NEI) ----
NONOTRAS_MajorGroupKey <- NONOTRAS_MajorGroupKey %>%
  # Keep original Major_Functional_Group for comparison
  mutate(old_Major_Functional_Group = Major_Functional_Group) %>%
  
  # Compute genus-level consensus for non-NA/NEI genera
  group_by(estimated_genus_name) %>%
  mutate(
    genus_consensus_group = if(!is.na(first(estimated_genus_name)) & first(estimated_genus_name) != "NEI") {
      # Get mode of Major_Functional_Group in the genus
      vals <- Major_Functional_Group[!is.na(Major_Functional_Group)]
      if(length(vals) == 0) NA_character_ else names(sort(table(vals), decreasing = TRUE))[1]
    } else {
      NA_character_
    }
  ) %>%
  ungroup() %>%
  
  # Apply the consensus to all rows in the genus (if applicable)
  mutate(
    Major_Functional_Group = if_else(
      !is.na(genus_consensus_group),
      genus_consensus_group,
      Major_Functional_Group
    ),
    
    # Update Assignment_for_Major_Functional_Group if the value changed
    Assignment_for_Major_Functional_Group = if_else(
      !is.na(genus_consensus_group) &
        (is.na(old_Major_Functional_Group) | old_Major_Functional_Group != Major_Functional_Group),
      "Assigned_Step_3: forced genus-level consistency; if genus is equal then assume Major_Functional_Group is equal too.",
      Assignment_for_Major_Functional_Group
    )
  ) %>%
  select(-old_Major_Functional_Group, -genus_consensus_group)


# STEP 7: If Major_Functional_Group is NA then Assignment_for_Major_Functional_Group= "could not be assigned" ----
NONOTRAS_MajorGroupKey <- NONOTRAS_MajorGroupKey %>%
  mutate(
    Assignment_for_Major_Functional_Group = if_else(
      is.na(Major_Functional_Group),
      "Alert: Major_Functional_Group could not be assigned at present",
      Assignment_for_Major_Functional_Group
    )
  )


# STEP 8: Finally, assign estimated_taxa_group based on recorded_taxa_group and force the estimated_taxa_group to be equal for all records with the same genus (so genera like Lutjanus) ---- 

# Initialize the new column
NONOTRAS_MajorGroupKey$estimated_taxa_group = NA_character_

NONOTRAS_MajorGroupKey <- NONOTRAS_MajorGroupKey %>%
  
  # Fill in missing estimated_taxa_group using recorded_taxa_group
  mutate(
    estimated_taxa_group = if_else(
      is.na(estimated_taxa_group),
      recorded_taxa_group,
      estimated_taxa_group
    )
  ) %>%
  
  # Enforce consistency at the genus level, ignoring NA or "NEI"
  group_by(estimated_genus_name) %>%
  mutate(
    genus_consensus_taxa_group = if_else(
      !is.na(estimated_genus_name) & estimated_genus_name != "NEI",
      {
        vals <- estimated_taxa_group[!is.na(estimated_taxa_group)]
        if (length(vals) == 0) NA_character_ else names(sort(table(vals), decreasing = TRUE))[1]
      },
      NA_character_
    ),

    # Apply consensus only to valid genera
    estimated_taxa_group = if_else(
      !is.na(estimated_genus_name) & estimated_genus_name != "NEI",
      genus_consensus_taxa_group,
      estimated_taxa_group
    )
  ) %>%
  ungroup() %>%
  select(-genus_consensus_taxa_group)

# Step 9: Rearrange for ease of reading ----
NONOTRAS_MajorGroupKey <- NONOTRAS_MajorGroupKey %>%
  select(
    recorded_taxa_group,
    estimated_taxa_group,
    estimated_genus_name,
    IHH_functional_major_groups_sm,
    Major_Functional_Group,
    Assignment_for_Major_Functional_Group
  )

# STEP 10: Two manual patches for PETO and MERO
# NOTE TO SELF: Honestly it would probably be easiest to just manually create a key assignment for the ~57 major groups in the CONAPESCA database... come back to this.
NONOTRAS_MajorGroupKey <- NONOTRAS_MajorGroupKey %>%
  mutate(
    estimated_taxa_group = case_when(
      recorded_taxa_group == "TIBURON" & estimated_genus_name == "Lamna" ~ "TIBURON",
      TRUE ~ estimated_taxa_group
    ),
    Major_Functional_Group = case_when(
      recorded_taxa_group == "TIBURON" & estimated_genus_name == "Lamna" ~ "Sharks",
      TRUE ~ Major_Functional_Group
    ),
    Assignment_for_Major_Functional_Group = case_when(
      recorded_taxa_group == "TIBURON" & estimated_genus_name == "Lamna" ~ 
        "Manual patch, come back to correctly file PETO and Lamna. Issue can be traced back to Hem's species key.",
      TRUE ~ Assignment_for_Major_Functional_Group
    )
  )

NONOTRAS_MajorGroupKey <- NONOTRAS_MajorGroupKey %>%
  mutate(
    estimated_taxa_group = case_when(
      recorded_taxa_group == "BAQUETA" & estimated_genus_name == "Hyporthodus" ~ "MERO",
      TRUE ~ estimated_taxa_group
    ),
    Major_Functional_Group = case_when(
      recorded_taxa_group == "BAQUETA" & estimated_genus_name == "Hyporthodus" ~ "Demersal marine fishes",
      TRUE ~ Major_Functional_Group
    ),
    Assignment_for_Major_Functional_Group = case_when(
      recorded_taxa_group == "BAQUETA" & estimated_genus_name == "Hyporthodus" ~ 
        "Manual patch, come back to assign BAQUETA with an earlier manual key later on.",
      TRUE ~ Assignment_for_Major_Functional_Group
    )
  )
    
# 246 obs., 6 vars.
table(NONOTRAS_MajorGroupKey$Major_Functional_Group)

          #           Cephalopods                     Crustaceans          Demersal marine fishes 
          #                     6                               7                              87 
          #           Echinoderms               Freshwater fishes                    Marine algae 
          #                     5                               6                               4 
          #     Marine fishes NEI Marine or freshwater fishes NEI      Molluscs excl. Cephalopods 
          #                     4                              15                              27 
          # Pelagic marine fishes                          Sharks                 Shrimps, prawns 
          #                    40                              31                              14 
```

```{r Assign Major_Functional_Group manually for OTRAS taxa}
# Alright, we're going to need to treat the "OTRAS" with no genus as their own special thing. I'll need to assign a genus or Major_Functional_Group manually.

# Split the OTRAS from everything else. They will require additional manual work.
Special_groups <- Special_groups

OTRAS_MajorGroupKey <- MasterSpKey_pending %>% filter(recorded_taxa_group %in% Special_groups) %>%
  select(recorded_taxa_group, common_name_clean,
         estimated_species_name, species_est_certainty, species_est_reasoning,
         estimated_genus_name, genus_est_reasoning,
         IHH_functional_major_groups_sm) %>%
  distinct(recorded_taxa_group, estimated_genus_name, IHH_functional_major_groups_sm, common_name_clean, .keep_all = TRUE)
# 280 obs., 8 vars.
```

```{r Fill in what's happening with OTRAS based on NON OTRAS}
# Create a reference dataframe from NONOTRAS:
Genus_lookup <- NONOTRAS_MajorGroupKey %>%
  filter(!is.na(estimated_genus_name)) %>%
  filter(estimated_genus_name != "NEI") %>% # remove all NEI rows as they create a many to many matching error
  select(
    estimated_genus_name,
    estimated_taxa_group,
    Major_Functional_Group
  ) %>%
  distinct()
# 130 obs., 3 vars.

# Perform a left join to bring in data from the NON-OTRAS data frame:
OTRAS_MajorGroupKey <- OTRAS_MajorGroupKey %>%
  left_join(Genus_lookup, by = "estimated_genus_name")
# 280 obs., 10 vars.

# Lift value from IHH_functional_major_groups_sm to inform Major_Functional_Group wherever possible:
OTRAS_MajorGroupKey <- OTRAS_MajorGroupKey %>%
  mutate(
    Major_Functional_Group = case_when(
      is.na(Major_Functional_Group) & !is.na(IHH_functional_major_groups_sm) ~ IHH_functional_major_groups_sm,
      TRUE ~ Major_Functional_Group  # otherwise keep existing value
    )
  )
```

```{r SANITY CHECK}
# SANITY CHECK: Does Major_Functional_Group = IHH_functional_major_groups_sm in most cases? 
OTRAS_MajorGroupKey_comparison <- OTRAS_MajorGroupKey %>%
  mutate(
    MajorGroup_match = Major_Functional_Group == IHH_functional_major_groups_sm
  ) %>% 
  filter(!MajorGroup_match)
# Yes. 0 observations where these don't match.

# Delete this to keep work space clean:
rm(OTRAS_MajorGroupKey_comparison)
```

```{r Separate out the OTRAS that have info about Major_Functional_Group and those that don't}
OTRAS_MajorGroup_completed.auto<- OTRAS_MajorGroupKey %>%
  filter(!is.na(Major_Functional_Group))

OTRAS_MajorGroup_completed.auto$Assignment_for_Major_Functional_Group <- c("Assigned based on genus name and IHH functional group")
# 137 obs., 11 vars.

OTRAS_MajorGroup_incomplete<- OTRAS_MajorGroupKey %>%
  filter(is.na(Major_Functional_Group))
# 143 obs., 10 vars.

# Save this key to work on offline.
```

```{r # Save this key to work on offline}
# # Save this key to work on offline. Compare this to the key I previously worked on.
# OTRAS_MajorGroup_completed.manually<- read_csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/OTRAS_MajorGroupKey_COMPLETED.MANUALLY_Mar132026.csv")
# 
# # Shave down OTRAS_MajorGroup_incomplete for a correct merge:
# colnames(OTRAS_MajorGroup_incomplete)
# OTRAS_MajorGroup_incomplete<- OTRAS_MajorGroup_incomplete %>%
#   select(-IHH_functional_major_groups_sm, -estimated_taxa_group, -Major_Functional_Group)
# 
# # Fill in info to OTRAS_MajorGroup_incomplete from OTRAS_MajorGroup_completed.manually
# OTRAS_MajorGroup_incomplete <- OTRAS_MajorGroup_incomplete %>%
#   left_join(
#     OTRAS_MajorGroup_completed.manually %>%
#       select(recorded_taxa_group, estimated_taxa_group,
#              common_name_clean, estimated_species_name,
#              Manual_Classification_Notes,
#              ornamentalFLAG_1,
#              IHH_functional_major_groups_sm,
#              Major_Functional_Group,
#              Assignment_for_Major_Functional_Group),
#     by = c("recorded_taxa_group", "common_name_clean")
#   )
# # 145 obs., 10 vars. (should be 143 obs, but charrito creates two duplicates)
# 
# test<- anti_join(
#   OTRAS_MajorGroup_incomplete,
#   OTRAS_MajorGroup_completed.manually,
#   by = c("recorded_taxa_group", "common_name_clean")
# )
# 
# # Save this dataframe and work on it offline:
# write.csv(OTRAS_MajorGroup_incomplete,
#   "/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/OTRAS_MajorGroupKey_INCOMPLETE_Mar252026.csv",
#   row.names = FALSE
# )
# # NOTE TO SELF: Charrito is a weird case here... Come back to this at some point - Mar 13 2026
```

```{r Load back in the OTRAS completed manually}
# Work offline complete. Load in completed key.
OTRAS_MajorGroup_completed.manually<- read_csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/OTRAS_MajorGroupKey_COMPLETED.MANUALLY_Mar252026.csv")
# 143 obs., 13 vars.

rm(OTRAS_MajorGroup_incomplete)
```

#### SAVE GENUS TO MAJOR_FUNCTIONAL_GROUP KEY
```{r Merge keys}
OTRAS_MajorGroup_complete.merge <- bind_rows(
  OTRAS_MajorGroup_completed.auto,
  OTRAS_MajorGroup_completed.manually
)
# 280 obs. (as expected) and 13 vars.

# Call this OTRAS_MajorGroupKey and then...
OTRAS_MajorGroupKey<- OTRAS_MajorGroup_complete.merge

OTRAS_MajorGroupKey$estimated_taxa_group<- "OTRAS"

# ... delete unnecessary dataframes now that we have a merged frame:
rm(OTRAS_MajorGroup_completed.auto, OTRAS_MajorGroup_completed.manually, OTRAS_MajorGroup_complete.merge)

# NOTE TO SELF REGARDING NATURAL PRODUCTS: Missing a lot of ornamental flags. Come back to this key work when I switch to working on the NP goal.
```

```{r Merge the OTRAS and the NON-OTRAS keys}
# Combine the two dataframes
COMBINED_MajorGroupKey <- bind_rows(NONOTRAS_MajorGroupKey, OTRAS_MajorGroupKey)
# 526 obs. (as expected)

# Define preferred order (optional)
preferred_order <- c(
  "recorded_taxa_group",  "estimated_taxa_group",
  "common_name_clean", "estimated_species_name", "estimated_genus_name", 
  "IHH_functional_major_groups_sm",
  "Major_Functional_Group", "Assignment_for_Major_Functional_Group",
  "Manual_Classification_Notes", "species_est_certainty", 
  "species_est_reasoning", "genus_est_reasoning", "ornamentalFLAG_1"
)

# Reorder columns (only those that exist in combined dataframe)
COMBINED_MajorGroupKey <- COMBINED_MajorGroupKey[, intersect(preferred_order, colnames(COMBINED_MajorGroupKey))]

COMBINED_MajorGroupKey <- COMBINED_MajorGroupKey %>%
  mutate(
    estimated_taxa_group = ifelse(
      is.na(estimated_taxa_group) & !is.na(recorded_taxa_group),
      recorded_taxa_group,
      estimated_taxa_group
    )
  )

# test <- COMBINED_MajorGroupKey %>%
#   select(recorded_taxa_group, estimated_genus_name, common_name_clean) %>%
#   mutate(
#     distinct_key = if_else(
#       recorded_taxa_group %in% Special_groups,
#       common_name_clean,
#       NA_character_
#     )
#   ) %>%
#   distinct(recorded_taxa_group, estimated_genus_name, distinct_key, .keep_all = TRUE) %>%
#   select(-distinct_key)
# 501 obs.! Good!
```

```{r Finally, ensure that estimated_genus_name is correctly parsed from estimated_species_name}
COMBINED_MajorGroupKey <- COMBINED_MajorGroupKey %>%
  mutate(
    estimated_genus_name = ifelse(
      is.na(estimated_genus_name) & !is.na(estimated_species_name),
      str_extract(estimated_species_name, "^[^ ]+"),  # extract the first word (genus)
      estimated_genus_name  # keep existing if not NA
    )
  )

# Ensure newly filled genus names are unicode normalized:
COMBINED_MajorGroupKey$estimated_species_name <- stri_trans_nfc(COMBINED_MajorGroupKey$estimated_species_name)
COMBINED_MajorGroupKey$estimated_genus_name <- stri_trans_nfc(COMBINED_MajorGroupKey$estimated_genus_name)
```

```{r Save this key!}
# Save this key to work on offline
write.csv(
  COMBINED_MajorGroupKey,
  "/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/MajorGroupKey_Mar252026.csv", row.names = FALSE)
```
# COMPLETE ENOUGH FOR NOW! MAR 13 2026, Mar 24 2026, Mar 25 2026

# ADDITIONAL KEY WORK [Apr 4 2026] - Create genus to family key for use in CMSY
# Eventually integrate this into the CONAPESCA CleanUp main file.

```{r Load GOC_MarineLandings dataframe as needed}
GOC_MarineLandings<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_MarineLandings_2000.present_EMF.Apr032026.rds"))
# 3312602 obs., 62 vars.
```

```{r FILTER OUT MARINE SECTORS OF INTEREST:}
# FILTER OUT MARINE FISHERY SECTORS OF INTEREST: ----
MarineFishery_Sectors<- c("marine: fishery sector yet to be classified" , "marine: small-scale capture", "marine: large-scale capture")

GOC_Landings_mar.fisheries<- GOC_MarineLandings %>% 
  filter(sector_estimate_draft %in% MarineFishery_Sectors)
# 3107603 obs., 62 vars.

range(GOC_Landings_mar.fisheries$year)
```

```{r Create genus to family key for use in CMSY}
GenusFamily.Key <- GOC_Landings_mar.fisheries %>%
  filter(!is.na(estimated_genus_name) & !(estimated_genus_name== "NEI") & !(estimated_genus_name== "unknown") ) %>%
  select(estimated_genus_name, estimated_family_name, estimated_taxa_group, Major_Functional_Group) %>%
  distinct()

GenusFamily.Key <- GenusFamily.Key %>%
  mutate(estimated_family_name = ifelse(
    estimated_family_name %in% c(
      "family missing but genus available", "NEI", "unknown"
    ),
    NA,
    estimated_family_name
  ))

# Clean the key, taking the most common non-NA value for estimated_taxa_group, estimated_family_name, and Major_Functional_Group per genera:
GenusFamilyGroup.Key_clean <- GenusFamily.Key %>%
  group_by(estimated_genus_name) %>%
  summarise(
    
    # Major_Functional_Group: pick most common value for the genus, ingoring NA values
    Major_Functional_Group = {
      candidates <- Major_Functional_Group[!is.na(Major_Functional_Group)]
      if(length(candidates) == 0) NA else names(sort(table(candidates), decreasing = TRUE))[1]
    },
    
    # estimated_taxa_group: pick most common value for the genus, ingnoring NA values; allow "OTRAS" if nothing else
    estimated_taxa_group = {
      candidates <- estimated_taxa_group[!is.na(estimated_taxa_group) & estimated_taxa_group != "OTRAS"]
      if(length(candidates) == 0 && "OTRAS" %in% estimated_taxa_group) {
        "OTRAS"
      } else if(length(candidates) == 0) {
        NA
      } else {
        names(sort(table(candidates), decreasing = TRUE))[1]
      }
    },
    
    # estimated_family_name: most common non-NA value
    family_from_previous.keys = {
      candidates <- estimated_family_name[!is.na(estimated_family_name)]
      if(length(candidates) == 0) NA else names(sort(table(candidates), decreasing = TRUE))[1]
    },
    
    .groups = "drop"
  )
# 193 obs., 4 vars.

AllGenera <- unique(GenusFamilyGroup.Key_clean$estimated_genus_name)

# TAXIZE: Careful, this takes a couple minutes to run. Don't re-run unknown you need to.
# Taxize_Family.Key <- tax_name(AllGenera, get = "family", db = "ncbi")
Taxize_Family.Key <- Taxize_Family.Key %>%
  rename(family_from_taxize = family)


# Create final clean key for use in CMSY and eventual integration into Clean CONAPESCA - Apr 4 2026:
GenusFamilyGroup.Key_clean <- GenusFamilyGroup.Key_clean %>%
  # Join taxize lookup (only new families):
  left_join(Taxize_Family.Key %>% select(query, family_from_taxize), 
            by = c("estimated_genus_name" = "query")) %>%
  mutate(
    # Use original family if present, otherwise fill from taxize
    new_estimated_family_name = ifelse(is.na(family_from_previous.keys), family_from_taxize, family_from_previous.keys),
    
    # Create reasoning column
    new_family_est_reasoning = case_when(
      !is.na(family_from_previous.keys) ~ "family name was estimated from previous keys",
      is.na(family_from_previous.keys) & !is.na(family_from_taxize) ~ "family name was estimated from genus name using taxize function",
      is.na(family_from_previous.keys) & is.na(family_from_taxize) ~ "family name will need to be logged manually"
    )
  )
```

```{r Save key as needed}
# Great! Export this key (integrate later): call it FAO Family Key
write.csv(GenusFamilyGroup.Key_clean, row.names = FALSE, file= "/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/GenusFamilyGroupKey_INCOMPLETE_Apr042026.csv")
```
# Key work ends here. Filled in last 8 missing families manually offline.
See: /home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/FAOFamilyKey_COMPLETE_Apr042026.csv

# AQUACULTURE KEY UNDER CONSTRUCTION [Apr 4 2026]
```{r Load data}
GOC_MarineLandings<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Marine.Landings_Sector.Sorted.Georef_2000.present_May262026.rds"))
# 3246583 obs., 57 vars.

Mariculture_Sectors<- c("marine: aquaculture")

GOC_MaricultureProduction<- GOC_MarineLandings %>% 
  filter(sector_estimate_draft %in% Mariculture_Sectors)
# 108009 obs., 57 vars.

range(GOC_MaricultureProduction$year)
```

```{r Create genus to family key for use in AQUACULTURE}
GenusFamily.Key <- GOC_MaricultureProduction %>%
  select(estimated_genus_name, estimated_family_name, estimated_taxa_group, Major_Functional_Group) %>%
  distinct()

AllGenera <- unique(GenusFamily.Key$estimated_genus_name)

# TAXIZE: Careful, this takes a couple minutes to run. Don't re-run unknown you need to.
# Taxize_Family.Key <- tax_name(AllGenera, get = "family", db = "ncbi")
Taxize_Family.Key <- Taxize_Family.Key %>%
  rename(family_from_taxize = family)


# Clean the key, taking the most common non-NA value for estimated_taxa_group, estimated_family_name, and Major_Functional_Group per genera:

GenusFamilyGroup.Key_clean <- GenusFamily.Key %>%
  # Identify placeholder genera
  mutate(
    is_placeholder = is.na(estimated_genus_name) | 
                     estimated_genus_name %in% c("NEI", "unknown")
  ) %>%
  
  # Group by genus for aggregation, keeping placeholders separate
  group_by(estimated_genus_name, is_placeholder) %>%
  mutate(
    # Pick most common values only for real genera
    Major_Functional_Group = if(!first(is_placeholder)) {
      candidates <- Major_Functional_Group[!is.na(Major_Functional_Group)]
      if(length(candidates) == 0) NA else names(sort(table(candidates), decreasing = TRUE))[1]
    } else {
      Major_Functional_Group
    },
    
    estimated_taxa_group = if(!first(is_placeholder)) {
      candidates <- estimated_taxa_group[!is.na(estimated_taxa_group) & estimated_taxa_group != "OTRAS"]
      if(length(candidates) == 0 && "OTRAS" %in% estimated_taxa_group) "OTRAS"
      else if(length(candidates) == 0) NA
      else names(sort(table(candidates), decreasing = TRUE))[1]
    } else {
      estimated_taxa_group
    },
    
    # Use the correct column name
    family_from_previous_keys = if(!first(is_placeholder)) {
      candidates <- estimated_family_name[!is.na(estimated_family_name)]
      if(length(candidates) == 0) NA else names(sort(table(candidates), decreasing = TRUE))[1]
    } else {
      estimated_family_name
    }
  ) %>%
  ungroup() %>%
  select(-is_placeholder) %>%
  
  # Join with Taxize lookup
  left_join(
    Taxize_Family.Key %>% select(query, family_from_taxize),
    by = c("estimated_genus_name" = "query")
  ) %>%
  
  # Fill new_estimated_family_name using previous keys first, then Taxize if NA, NEI, or unknown
  mutate(
    new_estimated_family_name = if_else(
      is.na(family_from_previous_keys) | 
      family_from_previous_keys %in% c("NEI", "unknown", "family missing but genus available"),
      family_from_taxize,
      family_from_previous_keys
    ),
    
    # Create reasoning column
    new_family_est_reasoning = case_when(
      !is.na(family_from_previous_keys) & !family_from_previous_keys %in% c("NEI", "unknown") ~ 
        "family name was estimated from previous keys",
      !is.na(family_from_taxize) ~ "family name was estimated from genus name using taxize function",
      TRUE ~ "family name will need to be logged manually"
    )
  )
# 113 obs., 4 vars.

# Great! Export this key (integrate later): call it FAO Family Key
# write.csv(GenusFamilyGroup.Key_clean, row.names = FALSE, file= "/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/AquacultureFamilyKey_INCOMPLETE_Apr042026.csv")
```


# Create most-likley family if 

```{r Load data}
NW.Mex_Landings<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/NW.Mex_Landings_SectorSorted.Georef_2000.present_May262026.rds"))

dim(NW.Mex_Landings)
# 4713144 obs., 57 vars.
```

```{r Create backbone for key}
colnames(NW.Mex_Landings)

LikelyFamily.Key_backbone0<- NW.Mex_Landings %>%
  select(estimated_genus_name, estimated_family_name, estimated_taxa_group) %>%
  distinct()
```

```{r Cleaning}
LikelyFamily.Key_backbone <- LikelyFamily.Key_backbone0 %>%
  mutate(
    most_likely_family = case_when(
      estimated_family_name == "family missing but genus available" ~ NA_character_,
      TRUE ~ estimated_family_name
    )
  ) %>%
  select(estimated_taxa_group, estimated_genus_name, estimated_family_name, most_likely_family)
```

```{r How was family estimated if species and genus is NEI?}
TEST<- NW.Mex_Landings %>%
  filter(estimated_taxa_group== "BANDERA")
# Must have been estimated like this in Hem or Mars key
```

```{r Estimate most_likely_family based on the other families available in that group}
write.csv(LikelyFamily.Key_backbone, 
          file="/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Keys/CONAPESCA_Keys/Taxonomic_Keys/LikelyFamilyKey_INCOMPLETE_Jun032026.csv", 
          row.names = FALSE)
```

# Sent this to Cursor - June 3 2026. I checked it but not closely. Needs to be closely scruntinized at some point.
## STEP A: Load the incomplete Likely Family Key

```{r Load LikelyFamilyKey_INCOMPLETE}
KEY_DIR <- "/Users/ferrer/Desktop/Cursor Projects/OHI/Taxonomic Keys"

LikelyFamilyKey0 <- read.csv(
  file.path(KEY_DIR, "LikelyFamilyKey_INCOMPLETE_Jun032026.csv"),
  stringsAsFactors = FALSE
)

colnames(LikelyFamilyKey0)
# 381 obs., 4 vars.
```

```{r Clean taxonomic text fields}
invalid_genus_tokens <- c(
  "nei", "unknown", "especies", "na", ""
)
invalid_family_tokens <- c(
  "nei",
  "family missing but genus available",
  "na",
  ""
)
excluded_taxa_groups <- c("otras")

LikelyFamilyKey0 <- LikelyFamilyKey0 %>%
  mutate(
    estimated_taxa_group = str_squish(as.character(estimated_taxa_group)),
    estimated_genus_name = str_squish(as.character(estimated_genus_name)),
    estimated_family_name = str_squish(as.character(estimated_family_name)),
    most_likely_family = str_squish(as.character(most_likely_family)),

    taxa_group_clean = tolower(estimated_taxa_group),
    genus_clean = tolower(estimated_genus_name),
    family_clean = tolower(estimated_family_name),

    genus_valid = !is.na(genus_clean) &
      !genus_clean %in% invalid_genus_tokens,

    family_valid = !is.na(family_clean) &
      !family_clean %in% invalid_family_tokens,

    taxa_group_eligible_for_gapfill = !is.na(taxa_group_clean) &
      !taxa_group_clean %in% excluded_taxa_groups
  )
# 381 obs., 12 vars.
```

## STEP B: Assign most_likely_family when family information is already available

```{r Create LikelyFamilyKey - Part 1 known family}
colnames(LikelyFamilyKey0)

original_most_likely_family <- LikelyFamilyKey0$most_likely_family

LikelyFamilyKey <- LikelyFamilyKey0 %>%
  mutate(
    most_likely_family = case_when(
      # Multi-family lists (e.g. "FamilyA;;FamilyB") are kept as-is.
      family_valid ~ estimated_family_name,
      family_clean == "nei" & !genus_valid ~ "NEI",
      !is.na(most_likely_family) &
        most_likely_family != "" &
        most_likely_family != "NEI" ~
        most_likely_family,
      TRUE ~ NA_character_
    ),

    family_source = case_when(
      !is.na(original_most_likely_family) &
        original_most_likely_family != "" &
        !(original_most_likely_family == "NEI" &
          genus_valid &
          family_clean == "nei") ~
        "original estimate",
      family_valid ~ "direct family assignment",
      family_clean == "nei" & !genus_valid ~ "family reported as NEI",
      TRUE ~ NA_character_
    )
  )

sum(is.na(LikelyFamilyKey$most_likely_family) |
  LikelyFamilyKey$most_likely_family == "")
# Rows still needing genus lookup or taxa-group gapfill.
# NEI rows with a valid genus are held for taxize lookup in STEP C.
```

## STEP C: Use taxize to estimate family from genus when family is missing

```{r Identify genera needing taxize lookup}
genera_for_taxize <- LikelyFamilyKey %>%
  filter(
    genus_valid,
    is.na(most_likely_family) | most_likely_family == ""
  ) %>%
  distinct(estimated_genus_name) %>%
  pull(estimated_genus_name)

length(genera_for_taxize)
# Unique genera to query with taxize.
```

```{r Helper - extract family rank from taxize classification}
extract_family_from_classification <- function(classification_df) {
  if (is.null(classification_df) || !is.data.frame(classification_df)) {
    return(NA_character_)
  }

  rank_col <- intersect(
    c("rank", "Rank"),
    names(classification_df)
  )[1]
  name_col <- intersect(
    c("name", "Name"),
    names(classification_df)
  )[1]

  if (is.na(rank_col) || is.na(name_col)) {
    return(NA_character_)
  }

  family_match <- classification_df[[name_col]][
    tolower(classification_df[[rank_col]]) == "family"
  ]

  if (length(family_match) == 0) {
    return(NA_character_)
  }

  family_match[1]
}
```

```{r Query taxize for genus-level family estimates}
GENUS_LOOKUP_FILE <- file.path(
  KEY_DIR,
  "LikelyFamilyKey_GenusFamilyLookup_Jun032026.csv"
)

# WoRMS is preferred for marine taxa; ITIS is used as a fallback.
lookup_genus_family <- function(genus_name) {
  worms_family <- tryCatch(
    {
      worms_cls <- classification(genus_name, db = "worms")
      extract_family_from_classification(worms_cls[[1]])
    },
    error = function(e) NA_character_
  )

  if (!is.na(worms_family) && worms_family != "") {
    return(worms_family)
  }

  tryCatch(
    {
      itis_cls <- classification(genus_name, db = "itis")
      extract_family_from_classification(itis_cls[[1]])
    },
    error = function(e) NA_character_
  )
}

if (file.exists(GENUS_LOOKUP_FILE)) {
  GenusFamilyLookup <- read.csv(
    GENUS_LOOKUP_FILE,
    stringsAsFactors = FALSE
  )

  missing_genera <- setdiff(genera_for_taxize, GenusFamilyLookup$estimated_genus_name)

  if (length(missing_genera) > 0) {
    missing_lookup <- data.frame(
      estimated_genus_name = missing_genera,
      taxize_family = vapply(
        missing_genera,
        lookup_genus_family,
        character(1)
      ),
      stringsAsFactors = FALSE
    )

    GenusFamilyLookup <- bind_rows(GenusFamilyLookup, missing_lookup)
    write.csv(
      GenusFamilyLookup,
      GENUS_LOOKUP_FILE,
      row.names = FALSE
    )
  }
} else {
  GenusFamilyLookup <- data.frame(
    estimated_genus_name = genera_for_taxize,
    taxize_family = vapply(
      genera_for_taxize,
      lookup_genus_family,
      character(1)
    ),
    stringsAsFactors = FALSE
  )

  write.csv(
    GenusFamilyLookup,
    GENUS_LOOKUP_FILE,
    row.names = FALSE
  )
}
# One row per genus queried with taxize.

table(is.na(GenusFamilyLookup$taxize_family))
```

```{r Merge taxize family estimates onto the key}
colnames(LikelyFamilyKey)

LikelyFamilyKey <- LikelyFamilyKey %>%
  left_join(GenusFamilyLookup, by = "estimated_genus_name") %>%
  mutate(
    most_likely_family = case_when(
      !is.na(most_likely_family) & most_likely_family != "" ~ most_likely_family,
      !is.na(taxize_family) & taxize_family != "" ~ taxize_family,
      family_clean == "nei" & genus_valid ~ "NEI",
      TRUE ~ most_likely_family
    ),

    family_source = case_when(
      !is.na(family_source) ~ family_source,
      !is.na(taxize_family) & taxize_family != "" ~ "taxize genus lookup",
      family_clean == "nei" & genus_valid ~ "family reported as NEI",
      genus_valid &
        (is.na(most_likely_family) | most_likely_family == "") ~
        "taxize lookup failed",
      TRUE ~ family_source
    )
  ) %>%
  select(-taxize_family)
# 381 obs.
# taxize may override NEI when a valid genus is available; otherwise NEI is kept.
```

## STEP D: Gapfill remaining rows using the most likely family within each taxonomic group

```{r Create genus-level family assignments within each taxa group}
GenusGroupFamily <- LikelyFamilyKey %>%
  filter(
    genus_valid,
    !is.na(most_likely_family),
    most_likely_family != "",
    most_likely_family != "NEI",
    !is.na(estimated_taxa_group),
    estimated_taxa_group != "",
    taxa_group_eligible_for_gapfill
  ) %>%
  group_by(estimated_taxa_group, estimated_genus_name) %>%
  summarise(
    genus_family = names(sort(table(most_likely_family), decreasing = TRUE))[1],
    .groups = "drop"
  )
# One family per valid genus within each eligible taxonomic group.
```

```{r Create taxa-group family lookup}
TaxaGroupFamilyLookup <- GenusGroupFamily %>%
  group_by(estimated_taxa_group) %>%
  summarise(
    most_likely_family_by_group = names(
      sort(table(genus_family), decreasing = TRUE)
    )[1],

    n_genera_assigned_to_modal_family = sum(
      genus_family == names(sort(table(genus_family), decreasing = TRUE))[1]
    ),

    n_genera_assigned_to_other_family = sum(
      genus_family != names(sort(table(genus_family), decreasing = TRUE))[1]
    ),

    n_genera_in_group = n(),
    .groups = "drop"
  ) %>%
  mutate(
    modal_family_genus_percent = round(
      100 * n_genera_assigned_to_modal_family / n_genera_in_group,
      1
    ),

    family_estimate_likelihood = case_when(
      modal_family_genus_percent == 100 ~
        "family estimate is virtually certain",
      modal_family_genus_percent >= 80 ~
        "high confidence in family estimate",
      modal_family_genus_percent >= 60 ~
        "moderate confidence in family estimate",
      TRUE ~
        "low confidence in family estimate"
    )
  )
# One row per taxonomic group with a resolved family mode.
# OTRAS is excluded; groups with no resolved genera remain absent.
```

```{r Create LikelyFamilyKey - Part 2 taxa-group gapfill}
colnames(LikelyFamilyKey)

LikelyFamilyKey <- LikelyFamilyKey %>%
  left_join(
    TaxaGroupFamilyLookup,
    by = "estimated_taxa_group"
  ) %>%
  mutate(
    most_likely_family = if_else(
      (is.na(most_likely_family) | most_likely_family == "") &
        taxa_group_eligible_for_gapfill,
      most_likely_family_by_group,
      most_likely_family
    ),

    family_source = case_when(
      !is.na(family_source) ~ family_source,
      taxa_group_eligible_for_gapfill &
        !is.na(most_likely_family) &
        !is.na(most_likely_family_by_group) &
        most_likely_family == most_likely_family_by_group ~
        "taxa-group mode gapfill",
      TRUE ~ family_source
    )
  )
# OTRAS rows keep row-level assignments only; no group-level family is applied.
```

```{r Restrict OTRAS rows without genus or family information}
colnames(LikelyFamilyKey)

LikelyFamilyKey <- LikelyFamilyKey %>%
  mutate(
    otras_without_taxonomic_info = taxa_group_clean == "otras" &
      !genus_valid &
      !family_valid,

    most_likely_family = if_else(
      otras_without_taxonomic_info,
      NA_character_,
      most_likely_family
    ),

    family_source = if_else(
      otras_without_taxonomic_info,
      NA_character_,
      family_source
    ),

    most_likely_family_by_group = if_else(
      taxa_group_clean == "otras",
      NA_character_,
      most_likely_family_by_group
    ),

    modal_family_genus_percent = if_else(
      taxa_group_clean == "otras",
      NA_real_,
      modal_family_genus_percent
    ),

    family_estimate_likelihood = if_else(
      taxa_group_clean == "otras",
      NA_character_,
      family_estimate_likelihood
    ),

    n_genera_assigned_to_modal_family = if_else(
      taxa_group_clean == "otras",
      NA_integer_,
      n_genera_assigned_to_modal_family
    ),

    n_genera_assigned_to_other_family = if_else(
      taxa_group_clean == "otras",
      NA_integer_,
      n_genera_assigned_to_other_family
    ),

    n_genera_in_group = if_else(
      taxa_group_clean == "otras",
      NA_integer_,
      n_genera_in_group
    )
  ) %>%
  select(
    estimated_taxa_group,
    estimated_genus_name,
    estimated_family_name,
    most_likely_family,
    family_source,
    most_likely_family_by_group,
    modal_family_genus_percent,
    family_estimate_likelihood,
    n_genera_assigned_to_modal_family,
    n_genera_assigned_to_other_family,
    n_genera_in_group
  )
# 381 obs., 10 vars.
# OTRAS rows without a valid genus or family remain NA.
```

```{r Sanity checks on LikelyFamilyKey}
colnames(LikelyFamilyKey)

table(LikelyFamilyKey$family_source, useNA = "ifany")

sum(is.na(LikelyFamilyKey$most_likely_family) |
  LikelyFamilyKey$most_likely_family == "")
# Remaining unresolved rows after row-level and taxa-group gapfill.

LikelyFamilyKey %>%
  filter(
    is.na(most_likely_family) | most_likely_family == ""
  ) %>%
  count(estimated_taxa_group, sort = TRUE)

# Rows where genus and family inputs were missing but a group-level family was assigned:
LikelyFamilyKey %>%
  filter(family_source == "taxa-group mode gapfill") %>%
  nrow()

# OTRAS should never receive a taxa-group gapfill assignment:
LikelyFamilyKey %>%
  filter(
    tolower(estimated_taxa_group) == "otras",
    family_source == "taxa-group mode gapfill"
  ) %>%
  nrow()

# OTRAS rows without genus or family information should remain unresolved:
LikelyFamilyKey %>%
  filter(
    tolower(estimated_taxa_group) == "otras",
    (is.na(estimated_genus_name) |
      estimated_genus_name %in% c("NEI", "unknown", "Especies", "")) &
      (is.na(estimated_family_name) |
        estimated_family_name %in% c(
          "NEI",
          "family missing but genus available",
          ""
        )),
    !is.na(most_likely_family),
    most_likely_family != ""
  ) %>%
  nrow()
```

## STEP E: Save the completed key and taxa-group lookup

```{r Save LikelyFamilyKey outputs}
LikelyFamilyKey_out <- LikelyFamilyKey %>%
  select(
    estimated_taxa_group,
    estimated_genus_name,
    estimated_family_name,
    most_likely_family
  )

write.csv(
  LikelyFamilyKey_out,
  file = file.path(KEY_DIR, "LikelyFamilyKey_Jun032026.csv"),
  row.names = FALSE
)
# 381 obs., 4 vars.

write.csv(
  GenusFamilyLookup,
  file = file.path(KEY_DIR, "LikelyFamilyKey_GenusFamilyLookup_Jun032026.csv"),
  row.names = FALSE
)

write.csv(
  TaxaGroupFamilyLookup,
  file = file.path(KEY_DIR, "LikelyFamilyKey_TaxaGroupLookup_Jun032026.csv"),
  row.names = FALSE
)

write.csv(
  LikelyFamilyKey,
  file = file.path(KEY_DIR, "LikelyFamilyKey_with_provenance_Jun032026.csv"),
  row.names = FALSE
)
# Provenance file includes family_source and group-level lookup fields.
```