---
title: 'OHI GoC 2025: Clean Waters: COFEPRIS'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: inline
editor_options: 
  chunk_output_type: console
---

# Summary

## COFEPRIS: Federal Commission for the Protection against Health Risks

**History of sanitary closures, precautionary closures, and red tide event notes in bivalve mollusk harvesting areas 2013-2024** *Published: March 8, 2025*

COFEPRIS monitors harmful algal blooms along the country's coastlines, with the goal of implementing, where appropriate, precautionary closures and sanitary bans in affected areas and preventing the sale of bivalve mollusks that pose a health risk related to harmful algal blooms.

These documents provide information on the history of health bans, precautionary closures, and related events from 2013 to 2024.

### List of documents

1.  Cuadro_Cierres_precautorios_2024

2.  Cuadro_Vedas_Sanitarias_2023

3.  Cuadro_Vedas_Sanitarias_2022

4.  Cuadro_Cierres_precautorios_2022-2

5.  Cuadro_Cierres_precautorios_2023_1

6.  Presencia_de_Marea_Roja_en_M_xico_durante_2021

7.  Presencia_de_Florecimientos_Algales_Nocivos_2020.pdf

8.  Vedas_Sanitarias_Florecimientos_Algales_Nocivos_2019

9.  Presencia_de_Marea_Roja_en_costas_nacionales_durante_2017

10. Presencia_de_Marea_Roja_en_costas_nacionales_durante_2016

11. Presencia_de_Marea_Roja_en_costas_nacionales_durante_2015

12. Presencia_de_Marea_Roja_en_costas_nacionales_durante_2014

13. Presencia_de_Marea_Roja_en_costas_nacionales_durante_2013

14. Cuadro_Cierres_precautorios_2025

15. Cuadro_Cierres_precautorios_otros_Riesgos_Sanitarios_2024-2025

#### Using `Tabula` to tabulate pdfs (issue #39)

Unfortunately, finding a consistent delimiter was difficult with these pdfs.  I began using the package pdftools, but realized that this would not be sufficient for obtaining the appropriate data. I then tried out using AI to take a screenshot of the table and produce a .csv, but that was creating issues and inconsistencies.  Not just in how it was separating the columns, but also in the actual **content** of the table; numbers were not always exact. Perhaps it was because I was not using the pro version.

So, I researched and found the application **[Tabula](https://opencollective.com/tabulapdf)**, which is a free, open-source tool for liberating data trapped inside PDF files! Download the newest version [HERE](https://tabula.technology/#:~:text=How%20to%20Use%20Tabula,one%2Dtime%20or%20monthly%20donation.). 

> Tabula was created by [Manuel Aristarán](http://jazzido.com/), [Mike Tigas](https://mike.tig.as/) and [Jeremy B. Merrill](http://jeremybmerrill.com/) with the support of [ProPublica](https://www.propublica.org/), [La Nación DATA](http://blogs.lanacion.com.ar/data/), [Knight-Mozilla OpenNews](https://opennews.org/), [The New York Times](http://www.nytimes.com/), [Northwestern University Knight Lab](http://knightlab.northwestern.edu/), [The Knight Foundation](http://www.knightfoundation.org/), and [The Shuttleworth Foundation](https://shuttleworthfoundation.org/). Tabula was designed by [Jason Das.](http://jasondas.com/)

It has two methods for extraction: STREAM and LATTICE.  Stream looks for whitespace between columns, while Lattice looks for boundary lines between columns.  

To use Tabula, 
1. Upload a PDF file containing a data table.
2. Browse to the page you want, then select the table by clicking and dragging to draw a box around the table. You can let it auto-identify the table, but make sure that the outline of the table is entirely within the selected area.
3. Click "Preview & Export Extracted Data". Tabula will try to extract the data and display a preview. Inspect the data to make sure it looks correct. If data is missing, you can go back to adjust your selection.
4. Click the "Export" button, specifying your desired file type (for us, typically a CSV).

#### Using Tabula for COFEPRIS

In this case, I used the Lattice method, and manually ensured that the extracted .CSV was correct. Unfortunately, only the tables from 2022-2025 were correctly able to be extracted.  If we need more than 3 years of data for this goal, I will return and determine another reproducible way to extract the previous year's data.  

The files are located at: ~/home/shares/ohi/OHI_GOC/_raw_data/COFEPRIS/d2025/tabula_csv_cw_closures

### Setup 

```{r}
library(tidyverse)
library(here)
library(janitor)
library(plotly)
library(pdftools)
library(ggplot2)
```

### PDF Text Extraction (NOT used anymore)

```{r}
# bring in the 2024 pdf! we will start here, even though there are pdfs with different structures for each year...
advisory_2024_file <- "Cuadro_Cierres_precautorios_2024.pdf" # the file name
advisory_2024_raw <- pdf_text(here::here("/home/shares/ohi/OHI_GOC/_raw_data/COFEPRIS/d2025", advisory_2024_file)) # using the package `pdftools`, read in using pdf_text()

advisory_2024_table <- advisory_2024_raw %>%
  as.data.frame() # make the page as a data frame.  Be warned, the data is all there! You just can't see it yet.

columns_2024 = c(
  "estado",
  "area_de_cosecha",
  "fecha_de_implementacion",
  "origen_de_la_medida",
  "toxina_marina_o_especie_fitoplanctonica_toxica",
  "producto",
  "resultado",
  "resultado_confirmatorio_ug_kg",
  "condicion",
  "fecha_del_levantamiento"
) # mirror the pdf and create clean column names

# advisory_2024_clean <- advisory_2024_table %>% 
#   separate_longer_position(everything(.), 6) %>% 
#   rename(main_column = ".") %>%  # rename the only column as main_column instead of just the period
#   slice(5:180) %>% 
#   separate_wider_position(
#     main_column, # the issue here is that there is not a common delimiter, and the three columns blur together. So, we can use `separate_wider_position()`, which allows you to instead define column breaks by the number of characters, which includes spaces
#     widths = c(site = 44, date_of_designation = 17, region_province_state = 20, area = 10, coordinates = 15), # these numbers were found through trial and error, but are always at least the length of the longest character string
#     names_sep = NULL, # default, to avoid renaming the column names
#     names_repair = "check_unique", #the default), no name repair, but checks they are unique
#     too_many = "debug", # will add additional columns to the output to help you locate and resolve the underlying problem
#     too_few = "debug", # adds additional columns to the output to help you locate and resolve the underlying problem... these will not remain in final code
#     cols_remove = FALSE # always FALSE if too_few or too_many are set to "debug"
#   ) 
```

### Cleaning Tabulated PDFs

Now that we extracted the data into CSVs using [Tabula](https://tabula.technology/#:~:text=How%20to%20Use%20Tabula,one%2Dtime%20or%20monthly%20donation.), let's see if we can clean them up and bind years 2022 - 2025.  Unfortunately, these have very different columns/fields, varying by year. 

First, lets read them all in:

```{r}
# for 2022 files
closed_seasons_2022_1_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/COFEPRIS/d2025/tabula_csv_cw_closures", "tabula-Cuadro_Vedas_Sanitarias_2022.csv"))
precautions_2022_2_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/COFEPRIS/d2025/tabula_csv_cw_closures", "tabula-Cuadro_Cierres_precautorios_2022-2.csv"))


# for 2023 files
closed_seasons_2023_1_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/COFEPRIS/d2025/tabula_csv_cw_closures", "tabula-Cuadro_Vedas_Sanitarias_2023.csv"))
precautions_2023_2_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/COFEPRIS/d2025/tabula_csv_cw_closures", "tabula-Cuadro_Cierres_precautorios_2023__1_.csv"))


# for 2024 files
closed_seasons_2024_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/COFEPRIS/d2025/tabula_csv_cw_closures", "tabula-Cuadro_Cierres_precautorios_2024.csv"))
precautions_2024_2025_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/COFEPRIS/d2025/tabula_csv_cw_closures", "tabula-Cuadro_Cierres_precautorios_otros_Riesgos_Sanitarios_2024_-_2025.csv"))


# for the 2025 file
closed_seasons_2025_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/COFEPRIS/d2025/tabula_csv_cw_closures", "tabula-Cuadro_Cierres_precautorios_2025.csv"))
```
```{r}
# start with 2024/2025 because it is easier... and we can see what we have now and what is important to do for the less clean extracted tables in 2022 and 2023.

# clean the column names and create a year identifier
closed_seasons_2024 <- closed_seasons_2024_raw %>% 
  janitor::clean_names() %>% 
  mutate(year = "2024")

closed_seasons_2025 <- closed_seasons_2025_raw %>% 
  janitor::clean_names() %>% 
  mutate(year = "2025") %>% 
  rename(resultado_confirmatorio_mg_kg = resultado_confirmatorio_o_de_seguimiento_mg_kg)

setdiff(colnames(closed_seasons_2025), colnames(closed_seasons_2024)) # no differences
setdiff(colnames(closed_seasons_2024), colnames(closed_seasons_2025)) # no differences

closed_seasons_2024_2025_clean <- rbind(closed_seasons_2024, closed_seasons_2025) %>% 
  mutate(estado = str_to_title(estado),
         estado = str_replace_all(estado, "Aja California Sur","Baja California Sur"),
         area_de_cosecha = str_to_title(area_de_cosecha)) %>%  # cosecha means harvest.  AKA where are the mollusks harvested?
  filter(estado %in% c("Baja California","Baja California Sur","Sonora","Sinaloa","Nayarit","Jalisco")) # only GoC states
```

I am thinking about how this will be used. We have to look up all of the harvest areas to ensure they are in the Gulf of California.  Many of the BCS or BC ones are on the Pacific-side, not all are in the Gulf.  This makes it difficult.

```{r}
unique(closed_seasons_2024_2025_clean$area_de_cosecha)

closed_seasons_2024_2025_forexplore <- closed_seasons_2024_2025_clean %>% 
  select(estado, area_de_cosecha)

# We can assume all sinaloa and sonora are coastal.
# however, we need to look at all baja ones...
# and we need to make sure that the Jalisco areas do not lie outside of the OHI GoC regions...

########## BC/BCS ###########
closed_seasons_2024_2025_bc_bcs <- closed_seasons_2024_2025_clean %>% 
  filter(estado %in% c("Baja California", "Baja California Sur"))
unique(closed_seasons_2024_2025_bc_bcs$area_de_cosecha)

# NOT in the GoC:
 # [1] "Estero El Coyote Y\rEstero El Cardón"
# [2] "Zona Iv, Comundú Baja\rCalifornia"                       
#  [3] "Punta Eugenia"   
# [5] "La Bocana"
# [6] "Estero El Coyote"                                        
#  [7] "Estero El Cardón" 
# [8] "San Hipólito Bahía\rAsunción" 
# [9] "Estero La Bocana" 
# [10] "Zona Iv Complejo Lagunar\rBahía Magdalena Bahia\rAlmejas"
# [11] "San Hipólito -  Bahía\rAsunción" 
```

#### COME BACK

Found something interesting: for Zona Iv, Comundú Baja California in BCS, Comundu is inland.  However, the product alerted is grinding oysters, due to demoic acid.  If the oysters are part of the BCS Gulf-side, then this should be counted, but otherwise, not.

```{r}
########## Jalisco ###########
closed_seasons_2024_2025_jalisco <- closed_seasons_2024_2025_clean %>% 
  filter(estado == "Jalisco")

unique(closed_seasons_2024_2025_jalisco$area_de_cosecha)
# Those that are NOT in the GoC: "Barra De Navidad" "Bahía De La Manzanilla" "Punta Pérula" "Bahía Punta Perula, La\rHuerta" "Bahía De Cuastecomates" 
# Unsure: "Malecón I, Malecón Ii,\rDársenas" 
```
```{r}
# only GOC
closed_seasons_2024_2025_goc <- closed_seasons_2024_2025_clean %>% 
  filter(area_de_cosecha != c("Estero El Coyote Y Estero El Cardón",
  "Zona IV, Comondú Baja California",
  "Punta Eugenia",
  "La Bocana",
  "Estero El Coyote",
  "Estero El Cardón",
  "San Hipólito Bahía Asunción",
  "Estero La Bocana",
  "Zona IV Complejo Lagunar Bahía Magdalena Bahía Almejas",
  "San Hipólito - Bahía Asunción",
  "Barra De Navidad",
  "Bahía De La Manzanilla",
  "Punta Pérula",
  "Bahía Punta Perula, La Huerta",
  "Bahía De Cuastecomates",
  "Malecón I, Malecón Ii, Dársenas"))
```




