---
title: 'OHI GoC 2025: Clean Waters Eutrophication'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# Combining all components of the Clean Waters goal

The Clean Waters (CW) goal measures the degree to which local waters are unpolluted by anthropogenic sources.

People value marine waters that are free of pollution and debris for aesthetic and health reasons. Contamination of waters comes from various sources, such as oil spills, chemicals, nutrient loading (eutrophication), disease pathogens (e.g., fecal coliform, viruses, and parasites from sewage outflow), and floating trash. People are sensitive to these pollutants, especially in areas they access for recreation, and place a high “existence value” on simply knowing that clean waters exist.

We have included four sub-goals of pollution in our assessment clean water goal: eutrophication (nutrients), chemicals, pathogens and marine debris. The status of these components is typically the inverse of their intensity. This means that the high input results in a low status score, and this goal scores highest when the contamination level is zero.

- Eutrophication (EUT) 
- Pathogens (PAT) 
- Chemicals (CHM)
  - Pesticides (PEST) 
  - Pharmaceuticals (PHRM) 
  - Hazardous waste sites (HAZW) 
  - Untreated wastewater discharge point sites (WWPS) 
  - Vessel hydrocarbon estimation (VESS) 
- Plastics (PLA)
  - Microplastics (MCRO): not actually calculated, not enough data
  - Risk of ghost gear (RIGG) 

## Bring in status scores

```{r}
library(tidyverse)
library(here)
library(sf)
library(mapview)
library(stringdist)
library(stringi)
```


Eutrophication
```{r}
eutro_cs <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/eutrophication/output","eutrophication_cs.csv")) %>% 
  group_by(rgn_id) %>% 
  mutate(year = 2017) %>% 
  complete(year = 2017:2025) %>% 
  fill(current_status, .direction = "down") %>% 
  dplyr::select(rgn_id, year, EUT = current_status) %>% 
  ungroup()

eutro_cs_plotly <- plotly::plot_ly(
  eutro_cs,
  x = ~year,
  y = ~EUT,
  color = ~as.factor(rgn_id),
  colors = rgn_colors,
  type = "scatter",
  mode = "lines+markers"
) %>%
  layout(
    title = "Clean Waters (CW) Current Status: EUT",
    xaxis = list(title = "Year"),
    yaxis = list(
      title = "Score // Percent (out of 100)",
      range = c(0, 105)
    )
  ) %>%
  add_lines(
    x = range(eutro_cs$year),
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )

eutro_cs_plotly
```

Pathogens
```{r}
path_cs <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/pathogens/output","pathogen_cs.csv")) %>% 
  dplyr::select(rgn_id, year, PAT = current_status) %>% 
  group_by(rgn_id) %>% 
  complete(year = 2017:2025) %>% 
  fill(PAT, .direction = "up") %>% 
  ungroup()

path_cs_plotly <- plotly::plot_ly(
  path_cs,
  x = ~year,
  y = ~PAT,
  color = ~as.factor(rgn_id),
  colors = rgn_colors,
  type = "scatter",
  mode = "lines+markers"
) %>%
  layout(
    title = "Clean Waters (CW) Current Status: PAT",
    xaxis = list(title = "Year"),
    yaxis = list(
      title = "Score // Percent (out of 100)",
      range = c(0, 105)
    )
  ) %>%
  add_lines(
    x = range(path_cs$year),
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )

path_cs_plotly
```

Chemicals
```{r}
chem_cs <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chemicals/chemical_cs.csv")) %>% 
  dplyr::select(rgn_id, year, CHM = chemical_cs) %>% 
  group_by(rgn_id) %>% 
  complete(year = 2017:2025) %>% 
  fill(CHM, .direction = "down") %>% 
  ungroup()

chem_cs_plotly <- plotly::plot_ly(
  chem_cs,
  x = ~year,
  y = ~CHM,
  color = ~as.factor(rgn_id),
  colors = rgn_colors,
  type = "scatter",
  mode = "lines+markers"
) %>%
  layout(
    title = "Clean Waters (CW) Current Status: CHM",
    xaxis = list(title = "Year"),
    yaxis = list(
      title = "Score // Percent (out of 100)",
      range = c(0, 105)
    )
  ) %>%
  add_lines(
    x = range(chem_cs$year),
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )

chem_cs_plotly
```

Plastics
```{r}
ghostgear_cs <- read_csv(here::here("~/OHI_Intro/ohi_goc_prep/_pressures/vessels/output/ghost_gear_risk.csv")) %>% 
  mutate(status = 1 - pressure,
         current_status = status * 100) %>% 
  dplyr::select(rgn_id, year, PLA = current_status) 

ghostgear_cs_plotly <- plotly::plot_ly(
  ghostgear_cs,
  x = ~year,
  y = ~PLA,
  color = ~as.factor(rgn_id),
  colors = rgn_colors,
  type = "scatter",
  mode = "lines+markers"
) %>%
  layout(
    title = "Clean Waters (CW) Current Status: PLA",
    xaxis = list(title = "Year"),
    yaxis = list(
      title = "Score // Percent (out of 100)",
      range = c(0, 105)
    )
  ) %>%
  add_lines(
    x = range(ghostgear_cs$year),
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )

ghostgear_cs_plotly
```

## Geometric mean

```{r}
# join all components and calculate geometric mean
cw_status <- eutro_cs %>%
  left_join(path_cs, by = c("rgn_id", "year")) %>%
  left_join(chem_cs, by = c("rgn_id", "year")) %>%
  left_join(ghostgear_cs, by = c("rgn_id", "year")) %>%
  group_by(rgn_id, year) %>%
  mutate(
    CW = exp(mean(log(c(EUT, PAT, CHM, PLA)), na.rm = TRUE))
  ) %>%
  ungroup()
```

```{r}
cw_plotly <- plotly::plot_ly(
  cw_status,
  x = ~year,
  y = ~CW,
  color = ~as.factor(rgn_id),
  colors = rgn_colors,
  type = "scatter",
  mode = "lines+markers"
) %>%
  layout(
    title = "Clean Waters (CW) Current Status: Geometric Mean of EUT, PAT, CHM, PLA",
    xaxis = list(title = "Year"),
    yaxis = list(
      title = "Score // Percent (out of 100)",
      range = c(0, 105)
    )
  ) %>%
  add_lines(
    x = range(cw_status$year),
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )

cw_plotly
```


## Arithmetic mean

```{r}
# join all components and calculate arithmetic mean
cw_status_arithmetic <- eutro_cs %>%
  left_join(path_cs, by = c("rgn_id", "year")) %>%
  left_join(chem_cs, by = c("rgn_id", "year")) %>%
  left_join(ghostgear_cs, by = c("rgn_id", "year")) %>%
  group_by(rgn_id, year) %>%
  mutate(
    CW = mean(c(EUT, PAT, CHM, PLA), na.rm = TRUE)
  ) %>%
  ungroup()
```

```{r}
cw_status_arithmetic_plotly <- plotly::plot_ly(
  cw_status_arithmetic,
  x = ~year,
  y = ~CW,
  color = ~as.factor(rgn_id),
  colors = rgn_colors,
  type = "scatter",
  mode = "lines+markers"
) %>%
  layout(
    title = "Clean Waters (CW) Current Status: Arithmetic Mean of EUT, PAT, CHM, PLA",
    xaxis = list(title = "Year"),
    yaxis = list(
      title = "Score // Percent (out of 100)",
      range = c(0, 105)
    )
  ) %>%
  add_lines(
    x = range(cw_status_arithmetic$year),
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )

cw_status_arithmetic_plotly
```


## Weighted Mean

```{r}
# join all components and calculate weighted mean
cw_status_weighted <- eutro_cs %>%
  left_join(path_cs, by = c("rgn_id", "year")) %>%
  left_join(chem_cs, by = c("rgn_id", "year")) %>%
  left_join(ghostgear_cs, by = c("rgn_id", "year")) %>%
  group_by(rgn_id, year) %>%
  mutate(
    CW = weighted.mean(c(EUT, PAT, CHM, PLA), w = c(1/6, 2/6, 2/6, 1/6), na.rm = TRUE)
  ) %>%
  ungroup()
```

```{r}
cw_status_weighted_plotly <- plotly::plot_ly(
  cw_status_weighted,
  x = ~year,
  y = ~CW,
  color = ~as.factor(rgn_id),
  colors = rgn_colors,
  type = "scatter",
  mode = "lines+markers"
) %>%
  layout(
    title = "Clean Waters (CW) Current Status: Weighted Mean of EUT (1/6), PAT (2/6), CHM (2/6), PLA (1/6)",
    xaxis = list(title = "Year"),
    yaxis = list(
      title = "Score // Percent (out of 100)",
      range = c(0, 105)
    )
  ) %>%
  add_lines(
    x = range(cw_status_weighted$year),
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
cw_status_weighted_plotly
```




