---
title: 'OHI GoC 2025: Clean Waters'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
    include-in-header: '../../../src/templates/ohi_hdr.html'
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: inline
editor_options: 
  chunk_output_type: inline
---

# Clean Waters Goal

## RENAMECA

Resultados de la Red Nacional de Medición de Calidad del Agua (RENAMECA).
[Location of surface and underground sites of the National Water Quality Measurement Network monitored during the period 2012-2024.](https://www.gob.mx/conagua/es/articulos/resultados-de-la-red-nacional-de-medicion-de-calidad-del-agua-renameca?idiom=es)

> CONAGUA, through the National Water Quality Measurement Network (RENAMECA), carries out systematic and ongoing monitoring of the water quality of the country's main bodies of water.
> Water quality measurement includes the analysis of physicochemical and microbiological parameters, established according to the type of water body to be characterized.

> On the other hand, RENAMECA addresses specific environmental issues related to water resources at the national level, including requests from the National Human Rights Commission.

> Finally, it should be noted that water quality information contributes to decision-making at all three levels of government to improve the sustainable management of water resources, as well as to informing society about the current status of major bodies of water.

The RENAMECA results are presented according to the following categories:

1.  **By type of water body**


A.  *Lotic* (Water bodies that always move in the same direction such as rivers, streams and creeks)

B.  *Lentic* (Still or stagnant inland waters, such as dams, lakes and reservoirs)

C.  *Coastal* (Coastal waters, such as beaches, bays, marshes and estuaries)

D.  *Subterranean* (Underground waters such as cenotes, galleries and springs)


2.  **By State**

3.  **All results**

[Download RENAMECA Results](https://files.conagua.gob.mx/aguasnacionales/TODOS%20LOS%20MONITOREOS.xlsb)

### Read in data

```{r}
# first let's read in the .xslsb
library(readxl)
library(here)
library(tidyverse)
library(janitor)

# sheet 1: the site names/IDs tested for water quality and their location
sitios_raw <- read_excel(here::here("/home/shares/ohi/OHI_GOC/_raw_data/CONAGUA_RENAMECA/d2025","TODOS_LOS_MONITOREOS.xlsm"), sheet = 1, col_names = TRUE)

# sheet 2: the annual results for all sites
resultados_raw <- read_excel(here::here("/home/shares/ohi/OHI_GOC/_raw_data/CONAGUA_RENAMECA/d2025","TODOS_LOS_MONITOREOS.xlsm"), sheet = 2, col_names = TRUE)

# sheet 3: the water quality variables tested... 462 of them.
simbologia_raw <- read_excel(here::here("/home/shares/ohi/OHI_GOC/_raw_data/CONAGUA_RENAMECA/d2025","TODOS_LOS_MONITOREOS.xlsm"), sheet = 3, col_names = TRUE)
```

### Clean and join

At least for now, let's only look at coastal sites (costeros) in the GoC, and then join sheet 1 with sheet 2 to have the results for only those sites.

```{r}
# filter the sites to only be coastal
unique(sitios_raw$`TIPO DE CUERPO DE AGUA`)
colnames(sitios_costeros)

sitios_costeros <- sitios_raw %>% 
  janitor::clean_names() %>% 
  filter(tipo_de_cuerpo_de_agua %in% c("COSTERO", "COSTERO (HUMEDAL)", "LÓTICO - COSTERO", "LÓTICO - COSTERO (HUMEDAL)", "LÉNTICO - COSTERO", "LÉNTICO - COSTERO (HUMEDAL)","COSTERO Y SEDIMENTOS")) %>%  # all coastal/coastal adjacent sites
  select(-c(clave_acuifero, acuifero)) # only present in non-coastal sites


# filter sites to be in GoC
unique(sitios_costeros$estado)

sitios_costeros_goc <- sitios_costeros %>% 
  filter(estado %in% c("BAJA CALIFORNIA SUR","NAYARIT","SONORA","BAJA CALIFORNIA","SINALOA", "JALISCO"))
```

#### Filter for GoC sites

This is to see how many are not actually *in* the Gulf, but are instead on the Pacific-side of the peninsula.

```{r}
library(sf)

sitios_costeros_goc_sf <- st_as_sf(sitios_costeros_goc, coords = c("longitud", "latitud"), crs = 4326)
head(sitios_costeros_goc_sf) # looks correct
st_crs(sitios_costeros_goc_sf) # good, WGS 84 for now


library(mapview)
mapview(sitios_costeros_goc_sf)
```


```{r}
# many are outside of the GoC.  Let's see if we can narrow down using the municipio...
unique(sitios_costeros_goc_sf$municipio)

gulf_municipios <- c("LOS CABOS", "LA PAZ", "MULEGÉ", "COMONDÚ", "LORETO", 
                     "TECUALA", "SANTIAGO IXCUINTLA", "BAHÍA DE BANDERAS", 
                     "SAN BLAS", "COMPOSTELA", "ROSAMORADA", "TOMATLÁN", 
                     "CIHUATLÁN", "LA HUERTA", "PUERTO VALLARTA", "CABO CORRIENTES",
                     "HERMOSILLO", "HUATABAMPO", "SAN IGNACIO RÍO MUERTO", 
                     "GUAYMAS", "EMPALME", "PUERTO PEÑASCO", "NAVOLATO", 
                     "CULIACÁN", "GUASAVE", "AHOME", "ANGOSTURA", "ELOTA", 
                     "SAN IGNACIO", "MAZATLÁN", "ROSARIO", "ESCUINAPA","MEXICALI")

sitios_costeros_goc_sf_2 <- sitios_costeros_goc_sf %>% 
  filter(municipio %in% gulf_municipios)

mapview(sitios_costeros_goc_sf_2) # oh no... there are some inconsistencies.
```

There are some inconsistencies, in which there are sites that are labeled with incorrect municipio names.  For example, clave_sitio DLBAJ166 is on the Pacific-side, not Gulf-side, yet the municipio is "MULEGÉ".  This is incorrect, because MULEGÉ is quite small and is Gulf-side.  It also seems that I have lost some northern sites...some sites in Bahia Los Angeles are designated as "ENSENADA"...

```{r}
# let's try to use the cuerpo_de_agua as a delimiter
unique(sitios_costeros_goc_sf$cuerpo_de_agua)

gulf_water_bodies <- c(
  "GOLFO DE CALIFORNIA",
  "BAHIA CONCEPCION", 
  "BAHIA DE LA PAZ",
  "BAHIA LA VENTANA",
  "BAHIA DE MUERTOS", 
  "BAHIA LAS PALMAS",
  "BAHIA SAN JOSE DEL CABO",
  "BAHIA SAN LUCAS",
  "LAGUNA AGUA GRANDE",
  "ESTERO TEACAPAN",
  "ESTERO TEAPAQUEÑOS", 
  "RIO SAN PEDRO",
  "CUERPO DE AGUA DEL HUMEDAL",
  "ESTERO EL CONCHAL",
  "ESTERO GACHUPINES",
  "ESTERO VARADERO", 
  "RIO GRANDE DE SANTIAGO",
  "RIO AMECA",
  "ESTERO EL QUELELE",
  "LAGUNA EL PESCADERO",
  "LAGUNA AGUA BRAVA",
  "ESTERO LAGUNA LOS VERGELES",
  "ESTERO CUAUTLA",
  "SISTEMA LAGUNA AGUA BRAVA",
  "BARRA DE NOVILLERO",
  "LAGUNA LA SABANITA", 
  "LAGUNA SABANITA",
  "LAGUNA BARRA DE NAVIDAD",
  "BAHIA MELAQUE",
  "BAHIA MELAQUE 1",
  "ESTERO LA MANZANILLA",
  "RÍO PURIFICACIÓN",
  "RÍO SAN NICOLAS", 
  "LAGUNA CHALACATEPEC",
  "LAGUNA XOLA-PARAMAN",
  "ESTERO MAJAHUAS",
  "RIO TOMATLAN",
  "ESTERO EL CHORRO",
  "CUERPO DE AGUA DEL HUMEDAL - LAGUNA COSTERA EL ERMITAÑO",
  "CUERPO DE AGUA DEL HUMEDAL - LAGUNA AGUA DULCE",
  "RÍO MARABASCO",
  "ESTERO LA CRUZ",
  "HUMEDALES DE YAVAROS-MORONCARIT 1",
  "HUMEDALES DE YAVAROS-MORONCARIT 2", 
  "HUMEDALES DE YAVAROS-MORONCARIT 3",
  "HUMEDALES DE YAVAROS-MORONCARIT 4",
  "HUMEDALES DE YAVAROS-MORONCARIT 5",
  "HUMEDALES DE YAVAROS-MORONCARIT 6",
  "HUMEDALES DE YAVAROS-MORONCARIT 7",
  "LAGUNA YAVAROS",
  "RIO MAYO",
  "DREN JUPATECO",
  "COMPLEJO LAGUNAR BAHIA GUASIMAS-ESTERO LOBOS",
  "HUMEDAL - BAHIA DE LOBOS",
  "RIO YAQUI",
  "BAHIA DE GUAYMAS",
  "AFLUENTE DE DESCARGAS AGUA DE MAR Y CUERPO DE DESCARGA BAHIA EL PARAJE",
  "ESTERO EL SOLDADO",
  "BAHIA LAS GUÁSIMAS",
  "LAGUNA AGIABAMPO",
  "SISTEMA LAGUNAR AGIABAMPO-BACREHUIS-RÍO FUERTE ANTIGUO 1",
  "SISTEMA LAGUNAR AGIABAMPO-BACREHUIS-RÍO FUERTE ANTIGUO 2",
  "SISTEMA LAGUNAR AGIABAMPO-BACREHUIS-RÍO FUERTE ANTIGUO 3",
  "SISTEMA LAGUNAR AGIABAMPO-BACREHUIS-RÍO FUERTE ANTIGUO 4",
  "SISTEMA LAGUNAR AGIABAMPO-BACREHUIS-RÍO FUERTE ANTIGUO 5",
  "SISTEMA LAGUNAR AGIABAMPO",
  "SISTEMA LAGUNAR AGIABAMPO 4",
  "BAHIA DE LOS ANGELES",
  "RIO CULIACAN",
  "BAHIA DE ALTATA SEG",
  "BAHIA ALTATA SALIDA",
  "BAHÍA ENSENADA PABELLONES",
  "LAGUNA PLAYA COLORADA-SANTA MARIA LA REFORMA",
  "BAHIA DE OUHIRA TOPOLOBAMPO SEG",
  "LAGUNAS DE SANTA MARIA TOPOLOBAMPO-OHUIRA",
  "LAGUNA DE SANTA MARIA TOPOLOBAMPO OHUIRA",
  "BAHÍA EL COLORADO",
  "RIO SINALOA",
  "BAHIA DE ALTATA",
  "RIO SAN LORENZO",
  "RÍO SAN LORENZO",
  "BAHIA DE CEUTA",
  "LAGUNA HUIZACHE-CAIMANERO",
  "LAGUNA HUIZACHE",
  "SISTEMA LAGUNAR HUIZACHE - CAIMANERO",
  "RIO BALUARTE",
  "ESTERO DE URIAS",
  "ESTERO URIAS/B. MAZATLAN",
  "ESTERO EL INFIERNILLO/B. MAZATLAN"
)

sitios_costeros_goc_sf_3 <- sitios_costeros_goc_sf %>% 
  filter(cuerpo_de_agua %in% gulf_water_bodies)

mapview(sitios_costeros_goc_sf_3) # NOT working... OCEANO PACIFICO used in the northern gulf and pacific side......
```

Okay, I think by watershed is the best way. 

```{r}
# let's try by watershed.
unique(sitios_costeros_goc_sf$cuenca)

#  [1] "SAN JOSÉ DEL CABO"                       "SAN LUCAS"                              
#  [3] "TODOS SANTOS"                            "ALFREDO B. BONFIL"                      
#  [5] "SAN IGNACIO"                             "SANTO DOMINGO"                          
#  [7] "LORETO"                                  "BAHÍA CONCEPCIÓN"                       
#  [9] "MULEGÉ"                                  "SANTA AGUEDA"                           
# [11] "SANTA ROSALÍA"                           "PUNTA EUGENIA"                          
# [13] "SAN LUIS"                                "VIZCAÍNO"                               
# [15] "EL ARCO"                                 "LA PURÍSIMA"                            
# [17] "SANTA RITA"                              "LA PAZ"                                 
# [19] "EL COYOTE"                               "LOS PLANES"                             
# [21] "SAN BARTOLO"                             "SANTIAGO"                               
# [23] "CABO PULMO"                              "GRUPO DE CORRIENTES MARISMAS NACIONALES"
# [25] "RÍO SAN PEDRO-DESEMBOCADURA"             "AMECA IXTAPA B"                         
# [27] "SAN BLAS"                                "RÍO SANTIAGO 6"                         
# [29] "IXTAPA"                                  "HUICICILA"                              
# [31] "RÍO ACAPONETA 2"                         "RÍO MARÍA GARCÍA"                       
# [33] "RÍO PURIFICACIÓN"                        "RÍO SAN NICOLÁS B"                      
# [35] "RÍO SAN NICOLÁS A"                       "RÍO TOMATLÁN B"                         
# [37] "RÍO IPALA"                               "RÍO MARABASCO A"                        
# [39] "TECOMALA"                                "PITILLAL"                               
# [41] "RÍO SONORA 3"                            "RÍO MAYO 3"                             
# [43] "RÍO YAQUI 3"                             "RÍO MÁTAPE 2"                           
# [45] "GRUPO DE CORRIENTES AGIABAMPO"           "LOS VIDRIOS 1"                          
# [47] "RÍO SONOYTA 1"                           "RÍO CONCEPCIÓN"                         
# [49] "DESCANSO-LOS MÉDANOS"                    "TIJUANA"                                
# [51] "GUADALUPE"                               "ENSENADA-EL GALLO"                      
# [53] "SAN CARLOS"                              "MANEADERO-LAS ANIMAS"                   
# [55] "SANTO TOMÁS"                             "SAN VICENTE"                            
# [57] "SAN RAFAEL"                              "SAN TELMO"                              
# [59] "SAN QUINTÍN"                             "SANTA CATARINA"                         
# [61] "BAHÍA SAN FELIPE"                        "EL BORREGO"                             
# [63] "VALLE DE SAN LUIS"                       "TEPETATE"                               
# [65] "SAN PEDRO"                               "RÍO CULIACÁN"                           
# [67] "GRUPO DE CORRIENTES PABELLONES"          "GRUPO DE CORRIENTES HUYAQUI"            
# [69] "GRUPO DE CORRIENTE TPOPOLOBAMPO"         "RÍO FUERTE 2"                           
# [71] "RÍO SINALOA 2"                           "GRUPO DE CORRIENTES REFORMA"            
# [73] "RÍO PERICOS 2"                           "RÍO SAN LORENZO 2"                      
# [75] "RÍO ELOTA"                               "GRUPO DE CORRIENTES TEMPEHUAYA"         
# [77] "RÍO PIAXTLA 2"                           "GRUPO DE CORRIENTES HUIZACHE"           
# [79] "RÍO BALUARTE 2"                          "GRUPO DE CORRIENTES MAZATLAN"           
# [81] "RÍO QUELITE 2"                           "BALANDRA Y EL MERITO"   


# those in the gulf-side:
cuencas_goc <- c("SAN JOSÉ DEL CABO", "TODOS SANTOS","LORETO","MULEGÉ","SANTA ROSALÍA","BAHÍA SAN FELIPE","EL BORREGO","VALLE DE SAN LUIS","LOS VIDRIOS 1","RÍO SONOYTA 1","RÍO CONCEPCIÓN","TEPETATE","SAN PEDRO","RÍO SONORA 3","RÍO MÁTAPE 2","RÍO YAQUI 3","RÍO MAYO 3","GRUPO DE CORRIENTES AGIABAMPO","RÍO FUERTE 2","GRUPO DE CORRIENTE TPOPOLOBAMPO","RÍO SINALOA 2","GRUPO DE CORRIENTES HUYAQUI","GRUPO DE CORRIENTES REFORMA","RÍO PERICOS 2","GRUPO DE CORRIENTES PABELLONES","RÍO CULIACÁN","GRUPO DE CORRIENTES PABELLONES","RÍO SAN LORENZO 2","GRUPO DE CORRIENTES TEMPEHUAYA","RÍO ELOTA","RÍO PIAXTLA 2","RÍO QUELITE 2","GRUPO DE CORRIENTES MAZATLAN","GRUPO DE CORRIENTES HUIZACHE","RÍO BALUARTE 2","GRUPO DE CORRIENTES MARISMAS NACIONALES","RÍO ACAPONETA 2","RÍO SAN PEDRO-DESEMBOCADURA","RÍO SANTIAGO 6","SAN BLAS","IXTAPA","HUICICILA","AMECA IXTAPA B","PITILLAL","TECOMALA","SANTA ROSALÍA","SANTA AGUEDA","MULEGÉ","BAHÍA CONCEPCIÓN","ALFREDO B. BONFIL","LA PAZ","EL COYOTE","BALANDRA Y EL MERITO","LOS PLANES","SAN BARTOLO","SANTIAGO","CABO PULMO","SAN JOSÉ DEL CABO","SAN LUCAS")

sitios_costeros_goc_sf_4 <- sitios_costeros_goc_sf %>% 
  filter(cuenca %in% cuencas_goc) %>% 
  filter(clave_sitio != "DLBAJ106W2") # out of GoC

mapview(sitios_costeros_goc_sf)
mapview(sitios_costeros_goc_sf_4)
```



#### Join with results

```{r}
# clean the column names
resultados <- resultados_raw %>% 
  janitor::clean_names()

# join cleaned sheet 1 and sheet 2
sitios_resultados_goc <- left_join(sitios_costeros_goc_sf_4, resultados, by = c("clave_sitio","nombre_del_sitio")) 

# check that everything is there...
length(unique(sitios_resultados_goc$clave_sitio))
length(unique(sitios_costeros_goc_sf_4$clave_sitio)) # both 268! yay. that means there are 268 water quality testing sites for the GoC!

# check the number of years per site
years_per_site <- sitios_resultados_goc %>% 
  group_by(clave_sitio) %>% 
  summarise(num_years = length(unique(.$ano))) # yay!! They all have 12, which is from 2012-2024.

colnames(sitios_resultados_goc) # 480 cols

# out of curiosity, it looks like only 131 variables have at least one non-NA value
filtered_for_col_wo_NAs <- sitios_resultados_goc %>% 
  select(where(~ any(!is.na(.)))) %>% 
  mutate(across(where(is.character), ~ str_replace_all(.x, "<", "")))

##### let's save these subsets, which is in wide format, but contains data only from the Gulf.
write_csv(filtered_for_col_wo_NAs, here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/int", "renameca_goc_at_least_one_obs.csv"))
write_csv(sitios_resultados_goc, here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/int", "renameca_goc_wide.csv"))

wide_check <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/int", "renameca_goc_wide.csv"), col_types = cols(.default = "c"))
clean_wide_check <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/int", "renameca_goc_at_least_one_obs.csv"), col_types = cols(.default = "c")) # remember to use the col_types otherwise you will lose the observations that had "<" that have not been mutated to as.character or as.numeric -- we will do that later
```


```{r}
# for fun, what about for hg_tot?
years_per_site_hg <- sitios_resultados_goc %>% 
  group_by(clave_sitio, ano) %>%
  summarise(hg_tot_values = sum(!is.na(hg_tot)), .groups = "drop") # depends on the site.  some have multiple observations of results per site per year.

hg_tot_only <- sitios_resultados_goc %>% 
  select(clave_sitio, fecha_realizacion,ano, hg_tot)
```


### Explore the available metrics 

...and see how many years they have been monitored!

```{r}
# first, let's see the symbology:
simbologia <- simbologia_raw %>% 
  janitor::clean_names()

## now, lets pivot longer so that there is a column named "clave_parametro" like the simbologia df

# see what columns are available (aka variables monitored)
colnames(filtered_for_col_wo_NAs) # okay, so we want to pivot, but likely depth and temperature should be their own pivot. and maybe their own df???

# pivot all the water quality indicators EXCEPT temp profiles
sitios_long_wq <- filtered_for_col_wo_NAs %>% 
  select(-starts_with("temp_")) %>%  # note that profundidad = depth (meters), st = (total solids), caudal (L/s) is "flow or discharge"
  mutate(across(c(alc_fen:st), ~ as.numeric(.x))) %>% 
  pivot_longer(
    cols = c(alc_fen:st),
    names_to = "clave_parametro",
    values_to = "value"
  )

# pivot temperature profile separately
sitios_long_temp <- filtered_for_col_wo_NAs %>%
  pivot_longer(
    cols = starts_with("temp_agua"),
    names_to = "depth_raw",
    values_to = "temp_C"
  ) %>%
  select(-c(alc_fen:st)) %>% 
  mutate(
    # clean depth information from column names
    depth_m = case_when(
      depth_raw == "temp_agua_sup" ~ 0, # surface ("superficie")
      depth_raw == "temp_agua_med" ~ NA_real_, # "media" ("average")... hard for us to know, come back to this!!!
      depth_raw == "temp_agua_fon" ~ NA_real_, # "fondo" ("background..." maybe environmental?)
      TRUE ~ as.numeric(
        str_remove(depth_raw, "temp_agua_") %>%    # remove prefix
          str_replace("_", ".")   # for the 0.5 meters and 39.5
      )
    )
  )
```

```{r}
library(lubridate)

# now, lets bind the water quality goc subset with the simbologia to see what each of the variables represent

simb_clean <- simbologia %>% 
  mutate(clave_parametro = str_to_lower(clave_parametro),
         nombre_parametro = str_to_lower(nombre_parametro)) # for joining

wq_goc_w_simb <- inner_join(sitios_long_wq, simb_clean, by = "clave_parametro") %>%  # 19 variables, correct!
  relocate(geometry, .after = unidad_de_medida) %>% # so this is not made lower
  mutate(across(c(nombre_del_sitio:tipo_cuerpo_de_agua), ~ str_to_lower(.x))) %>% # for easier reading (just my aesthetic preference)
  mutate(fecha_realizacion = ymd(fecha_realizacion)) # ensure it is a lubridate

# length(unique(wq_goc_w_simb$clave_sitio)) # 268 sites! good

# mapview(wq_goc_w_simb) # looks good (took a while to load though since it is in long format)

#### SAVE for now (Water quality in GoC)
write_csv(wq_goc_w_simb, here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/int", "renameca_goc_long.csv"))
long_check <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/int", "renameca_goc_long.csv"))
```


```{r}
# lets look at the number of times each site was monitored per year.
# some years early on would sample 8 times in a year, but now it is typically 1, if any.
wq_variables_avail <- wq_goc_w_simb %>% 
  st_drop_geometry() %>% 
  group_by(clave_parametro, ano, clave_sitio) %>% 
  summarise(times_monitored = n(), .groups = "drop")
```


```{r}
# for each parameter and site, check if monitoring occurred in every year from 2018 to 2024
# group data by parameter, site, and year, count distinct years monitored per parameter/site
# filter to include only parameter/site pairs monitored at least once in all years

# summarise by parameter, year, site
coverage_per_year <- wq_goc_w_simb %>%
  st_drop_geometry() %>%             
  group_by(clave_parametro, ano, clave_sitio) %>%
  summarise(times_monitored = n(), .groups = "drop")

# identify parameter/site pairs monitored at least once per year from 2018_2024
coverage_summary <- coverage_per_year %>%
  group_by(clave_parametro, clave_sitio) %>%
  summarise(years_monitored = n_distinct(ano), .groups = "drop") %>%
  filter(years_monitored == 7)  # Only those monitored every year

# select sites with at least one parameter monitored at least once in every year from 2018 to 2024
sites_with_full_yearly_parameter <- coverage_summary %>% 
  distinct(clave_sitio)

sites_with_full_yearly_parameter <- unique(coverage_summary$clave_sitio) # v2025: 24 sites
parameters_w_yrly_coverage <- unique(coverage_summary$clave_parametro) # v2025: 74 parameters 
```


```{r}
# # 1. Calculate yearly coverage for all site-parameter combos
# coverage_per_year <- wq_goc_w_simb %>%
#   st_drop_geometry() %>%
#   group_by(clave_sitio, clave_parametro, ano) %>%
#   summarise(times_monitored = n(), .groups = 'drop')
# 
# # 2. For each site-parameter, count how many years are covered
# coverage_summary <- coverage_per_year %>%
#   group_by(clave_sitio, clave_parametro) %>%
#   summarise(years_monitored = n_distinct(ano), .groups = 'drop')
# 
# # 3. Filter to only those site-parameter combos with full coverage (7 years)
# full_coverage_combos <- coverage_summary %>%
#   filter(years_monitored == 7)
# 
# # 4. requiring parameters to be monitored fully at (e.g.) 80% or more of the 24 qualifying sites that meet full yearly coverage criteria, rather than the entire original 268 sites
# total_filtered_sites <- length(unique(full_coverage_combos$clave_sitio))  # Should be 24 here
# threshold <- 0.8  # For example, 80% of filtered sites
# 
# params_full_at_most_sites <- full_coverage_combos %>%
#   group_by(clave_parametro) %>%
#   summarise(sites_covered = n_distinct(clave_sitio), .groups = "drop") %>%
#   filter(sites_covered >= threshold * total_filtered_sites) %>%
#   pull(clave_parametro) # all 74
```

```{r}
wq_goc_comparable <- wq_goc_w_simb %>%
  filter(
    clave_sitio %in% unique(full_coverage_combos$clave_sitio),
    clave_parametro %in% params_full_at_most_sites
  )

mapview(wq_goc_comparable)
```


#### Instead of requiring at least one observation yearly from 2018_2024 for each site, now:

Lets apply a threshold like 0.7 to require that sites have data for at least 70% of the years between 2018-2024 for each parameter, instead of demanding 100% coverage.

1. Calculate the number of years with data per site and parameter (2018–2024).

2. Calculate the percentage of years with data relative to total years (7).

3. Filter to keep site-parameter pairs where percent of years with data >= 70% (i.e., at least 5 out of 7 years).

Use this filtered set to identify usable sites and parameters.

```{r}
total_years <- length(2018:2024)  # 7 years
threshold <- 0.7  # 70%

# calc years monitored per site and parameter
years_per_site_param <- wq_goc_w_simb %>%
  filter(ano %in% 2018:2024) %>% 
  group_by(clave_sitio, clave_parametro) %>%
  summarise(years_monitored = n_distinct(ano), .groups = "drop") %>%
  mutate(coverage_ratio = years_monitored / total_years) %>%
  filter(coverage_ratio >= threshold) # basically only those that had 5 years of coverage or more

# find the sites and parameters that meet the threshold
sites_thres <- unique(years_per_site_param$clave_sitio) # v2025: 23
params_thres <- unique(years_per_site_param$clave_parametro) # v2025: 74

# filter the original data to keep these sites and parameters only
wq_goc_filtered <- wq_goc_w_simb %>%
  filter(clave_sitio %in% sites_thres,
         clave_parametro %in% params_thres)

mapview(wq_goc_filtered) # weird !!! These are completely different sites than those in wq_goc_comparable...
```


```{r}
sites_w_parameters_generally_complete <- wq_goc_w_simb %>% 
  filter(clave_parametro %in% parameters_w_yrly_coverage)

mapview(sites_w_parameters_generally_complete)

sites_w_parameters_generally_complete_noNA <- sites_w_parameters_generally_complete %>% 
  filter(!is.na(value))

library(kableExtra)
sites_w_parameters_generally_complete_noNA %>%
  distinct(clave_parametro, unidad_de_medida, nombre_parametro) %>%
  arrange(clave_parametro) %>%
  kable("html", caption = "Unique Parameters with Units and Names") %>%
  kable_styling(full_width = FALSE, 
                bootstrap_options = c("striped", "hover", "condensed", "responsive"))

```

#### Values over time


##### coliformes fecales

```{r}
library(plotly)

cf <- sites_w_parameters_generally_complete_noNA %>% 
  filter(clave_parametro %in% "coli_fec") %>% 
  filter(ano %in% 2018:2024)

ct <- sites_w_parameters_generally_complete_noNA %>% 
  filter(clave_parametro %in% "coli_tot") %>% 
  filter(ano %in% 2018:2024)

### preliminary plot 
line_plot <- plotly::plot_ly(ct, x = ~fecha_realizacion, y = ~value, color = ~clave_sitio, type = "scatter", mode = "lines") %>% 
  layout(title = "coliformes fecales in MPN/100mL", 
         xaxis = list(title = "Year"),
         yaxis = list(title = "Value (MPN / 100mL"))

line_plot

## site DLNAY1893 had a big spike in Oct 2017 and july 2014 for cf.  This is in the bahia de banderas in nayarit, by puerto vallarta.
cf_DLNAY1893 <- cf %>% 
  filter(clave_sitio %in% "DLNAY1893") 
mapview(cf_DLNAY1893) # it is within a waterway called Punta Pool, that contains turtles (https://www.google.com/maps/place/Pool+punta/@20.6764982,-105.2861224,3a,75y,90t/data=!3m8!1e2!3m6!1sCIABIhADyc5UcAUH7mfJAOIABnB4!2e10!3e12!6shttps:%2F%2Flh3.googleusercontent.com%2Fgps-cs-s%2FAC9h4nqIrPnx3_pyZEV8CL_0lnWBChDs9T6LW0IqC8RSf7aQV16R442e5WymlaVFGoSqHL8pAq27VWWGTO1_1GdtDA7Ds58XDJ-bE9frltgqxpBI8KxAoMcLl-dQrz0ASLCO-OEdpzM2V3Lczt6F%3Dw203-h270-k-no!7i3000!8i4000!4m15!1m7!3m6!1s0x84214541150fa951:0x3aee618b164963f2!2sPool+punta!8m2!3d20.6764982!4d-105.2861224!16s%2Fg%2F11tcd8qxj1!3m6!1s0x84214541150fa951:0x3aee618b164963f2!8m2!3d20.6764982!4d-105.2861224!10e5!16s%2Fg%2F11tcd8qxj1?entry=ttu&g_ep=EgoyMDI1MDgxOC4wIKXMDSoASAFQAw%3D%3D).  

## plus, after Tropical Storm Lidia hit the Pacific coast at the end of Aug–early Sept 2017 and caused widespread flooding and infrastructure issues across the region; so it makes sense that water-quality impacts can persist for weeks, especially near river mouths (Ameca, Pitillal, Cuale) that drain into Bahía de Banderas.

## lets look at sites nearby during 2017: DLNAY1933, also at the mouth of the river, and DLNAY1934, outside near a harbor.
cf_DLNAY <- cf %>% 
  filter(clave_sitio %in% c("DLNAY1893", "DLNAY1933","DLNAY1934"))

line_plot_DLNAY <- plotly::plot_ly(cf_DLNAY, x = ~fecha_realizacion, y = ~value, color = ~clave_sitio, type = "scatter", mode = "lines") %>% 
  layout(title = "coliformes fecales in MPN/100mL", 
         xaxis = list(title = "Year"),
         yaxis = list(title = "Value (MPN / 100mL"))

line_plot_DLNAY # very localized effects at DLNAY18933.  

```


#### Cont. exploration

```{r}
total_years <- length(2018:2024)  # = 7 years
threshold <- 0.5  # i.e., 50%

years_per_site_param <- wq_goc_w_simb %>%
  filter(ano %in% 2018:2024) %>% 
  st_drop_geometry() %>% 
  group_by(clave_sitio, clave_parametro) %>%
  summarise(years_monitored = n_distinct(ano), .groups = "drop") %>%
  mutate(coverage_ratio = years_monitored / total_years)


site_param_kept <- years_per_site_param %>%
  filter(coverage_ratio >= threshold)

sites_thres <- unique(site_param_kept$clave_sitio)
params_thres <- unique(site_param_kept$clave_parametro)


param_coverage <- site_param_kept %>%
  group_by(clave_parametro) %>%
  summarise(sites_meeting_threshold = n_distinct(clave_sitio), .groups = "drop")

# require that parameter is present in at least 70% of sites that had any data
site_cutoff <- 0.5 * length(unique(site_param_kept$clave_sitio))

params_generalizable <- param_coverage %>%
  filter(sites_meeting_threshold >= site_cutoff) %>%
  pull(clave_parametro)


wq_goc_comparable <- wq_goc_w_simb %>%
  filter(clave_sitio %in% sites_thres,
         clave_parametro %in% params_generalizable)

mapview(wq_goc_comparable) 

```


```{r}
threshold_seq <- seq(0.5, 1, 0.1) # from 50% to 100%
diagnostic_thresholds <- purrr::map_df(threshold_seq, function(th) {
  
  site_param_temp <- years_per_site_param %>%
    filter(coverage_ratio >= th)
  
  tibble(
    threshold = th,
    n_sites = n_distinct(site_param_temp$clave_sitio),
    n_params = n_distinct(site_param_temp$clave_parametro)
  )
})

ggplot(diagnostic_thresholds, aes(x = threshold, y = n_sites)) +
  geom_line(color = "steelblue") +
  geom_point() +
  ylab("Number of sites retained") +
  xlab("Coverage threshold") +
  ggtitle("Impact of coverage threshold on site retention") +
  theme_minimal()

```


```{r}
coverage_summary <- wq_goc_w_simb %>%
  filter(ano %in% 2018:2024) %>%
  st_drop_geometry() %>%
  group_by(clave_parametro, ano, clave_sitio) %>%  # include site to avoid duplicates
  summarise(has_data = any(!is.na(value)), .groups = "drop") %>%
  group_by(clave_parametro, ano) %>%
  summarise(
    coverage_fraction = mean(has_data),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = ano, values_from = coverage_fraction, values_fill = 0) %>%
  column_to_rownames("clave_parametro") %>%
  as.matrix()


plot_ly(
  z = coverage_summary,
  x = colnames(coverage_summary),
  y = rownames(coverage_summary),
  type = "heatmap",
  colors = colorRamp(c("white", "royalblue")),
  hoverinfo = "x+y+z"
) %>%
  layout(
    title = "Fraction of sites in the GoC that have data by parameter and year",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Parameter")
  )

# ### CHECK:
# 
# data_2024 <- wq_goc_w_simb %>% 
#   filter(ano %in% 2024) 
# 
# unique(data_2024$clave_sitio) # only one...
# 
# data_2024_noNA <- wq_goc_w_simb %>% 
#   filter(ano %in% 2024) %>% 
#   filter(!is.na(value))
# unique(data_2024_noNA$clave_parametro) # 7.... instead of 74


```

##### Explore SST (Sólidos Suspendidos Totales)

```{r}
sst <- sitios_resultados_goc %>% 
  mutate(across(c(alc_fen:salinidad_469), ~ as.numeric(.x))) %>% 
  pivot_longer(
    cols = c(alc_fen:salinidad_469),
    names_to = "clave_parametro",
    values_to = "value"
  ) %>% 
  filter(clave_parametro %in% "sst") %>% 
  filter(ano %in% 2018:2024)

mapview(sst)

sst_2018 <- sst %>% 
  filter(ano %in% 2018)
mapview(sst_2018, zcol = "value")

sst_2019 <- sst %>% 
  filter(ano %in% 2019)
mapview(sst_2019, zcol = "value")

sst_2020 <- sst %>% 
  filter(ano %in% 2020)
mapview(sst_2020, zcol = "value")

sst_2021 <- sst %>% 
  filter(ano %in% 2021)
mapview(sst_2021, zcol = "value")

sst_2022 <- sst %>% 
  filter(ano %in% 2022)
mapview(sst_2022, zcol = "value")

sst_2023 <- sst %>% 
  filter(ano %in% 2023)
mapview(sst_2023, zcol = "value")

sst_2024 <- sst %>% 
  filter(ano %in% 2024)
mapview(sst_2024, zcol = "value")

glifosato <- sitios_resultados_goc %>% 
  mutate(across(c(alc_fen:salinidad_469), ~ as.numeric(.x))) %>% 
  pivot_longer(
    cols = c(alc_fen:salinidad_469),
    names_to = "clave_parametro",
    values_to = "value"
  ) %>% 
  filter(clave_parametro %in% "glifosato") %>% 
  filter(ano %in% 2018:2024)
```

###### Old

```{r}
sst <- sites_w_parameters_generally_complete_noNA %>% 
  filter(clave_parametro %in% "sst") %>% 
  filter(ano %in% 2018:2024)

unique(sst$clave_sitio) # v2025: 254 sites!!!
mapview(sst)

# summarize years monitored per site for sst
sst_site_coverage <- sst %>%
  group_by(clave_sitio) %>%
  summarise(years_monitored = n_distinct(ano)) %>%
  arrange(desc(years_monitored))

# bar plot for # of sites by years monitored
ggplot(sst_site_coverage, aes(x = years_monitored)) +
  geom_bar(fill = "steelblue") +
  scale_x_continuous(breaks = 1:7) +
  labs(
    title = "Number of Sites by Years of SST Data Coverage (2018-2024)",
    x = "Years Monitored",
    y = "Number of Sites"
  ) +
  theme_minimal()
```

```{r}
line_plot_sst <- plotly::plot_ly(sst, x = ~fecha_realizacion, y = ~value, color = ~clave_sitio, type = "scatter", mode = "lines") %>% 
  layout(title = "suspended solids", 
         xaxis = list(title = "Year"),
         yaxis = list(title = "Value"))
line_plot_sst

### look at the only site that has a value in 2024...
sst_DLNAY1915 <- sst %>% 
  filter(clave_sitio %in% "DLNAY1915")

line_plot_sstDLNAY1915 <- plotly::plot_ly(sst_DLNAY1915, x = ~fecha_realizacion, y = ~value, color = ~clave_sitio, type = "scatter", mode = "lines") %>% 
  layout(title = "suspended solids", 
         xaxis = list(title = "Year"),
         yaxis = list(title = "Value"))
line_plot_sstDLNAY1915
```

### Sampling frequency

```{r}
sampling_counts <- wq_goc_w_simb %>%
  st_drop_geometry() %>%
  group_by(clave_sitio, clave_parametro, ano) %>%
  summarise(n_samples = n(), .groups = "drop")


site_summary <- sampling_counts %>%
  group_by(clave_sitio, clave_parametro) %>%
  summarise(mean_per_year = mean(n_samples),
            sd_per_year   = sd(n_samples),
            years_sampled = n(),
            .groups = "drop")

coverage_thresholds <- sampling_counts %>%
  mutate(quarterly = n_samples >= 4,
         monthly   = n_samples >= 12)

```

### Explore available data further, in a different way...

```{r}
# count years of data per site/parameter
site_param_coverage <- wq_goc_w_simb %>%
  filter(ano %in% 2018:2024) %>%
  filter(!is.na(value)) %>% 
  st_drop_geometry() %>% 
  group_by(clave_parametro, clave_sitio) %>%
  summarise(n_years = n_distinct(ano), .groups = "drop")

# count how many sites meet threshold (>= 3 years)
param_threshold <- site_param_coverage %>%
  mutate(threshold_met = n_years >= 3)

param_threshold_sites <- param_threshold %>%  
  filter(threshold_met == TRUE) %>% 
  distinct(clave_sitio)

param_threshold_count <- param_threshold %>% 
  group_by(clave_parametro) %>% 
  summarise(
    total_sites = n(),
    sites_meeting = sum(threshold_met),
    perc_sites_meeting = sites_meeting / total_sites,
    .groups = "drop"
  )

# plot
ggplot(param_threshold_count, aes(x = reorder(clave_parametro, perc_sites_meeting),
                            y = perc_sites_meeting)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Usable parameters (≥3 years of data per site, 2018–2024)",
    x = "Parameter",
    y = "Proportion of sites meeting threshold (3 years)"
  ) +
  theme_minimal()


######################
# for 2020-2024
######################

# count years of data per site/parameter
site_param_coverage <- wq_goc_w_simb %>%
  filter(ano %in% 2020:2024) %>%
  filter(!is.na(value)) %>% 
  st_drop_geometry() %>% 
  group_by(clave_parametro, clave_sitio) %>%
  summarise(n_years = n_distinct(ano), .groups = "drop")

# count how many sites meet threshold (>= 3 years)
param_threshold <- site_param_coverage %>%
  mutate(threshold_met = n_years >= 3)

param_threshold_sites <- param_threshold %>%  
  filter(threshold_met == TRUE) %>% 
  distinct(clave_sitio)

param_threshold_count <- param_threshold %>% 
  group_by(clave_parametro) %>% 
  summarise(
    total_sites = n(),
    sites_meeting = sum(threshold_met),
    perc_sites_meeting = sites_meeting / total_sites,
    .groups = "drop"
  )

# plot
ggplot(param_threshold_count, aes(x = reorder(clave_parametro, perc_sites_meeting),
                            y = perc_sites_meeting)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Usable parameters (≥3 years of data per site, 2020–2024)",
    x = "Parameter",
    y = "Proportion of sites meeting threshold (3 years)"
  ) +
  theme_minimal()
```

```{r}
n_tot <- sites_w_parameters_generally_complete_noNA %>% 
  filter(clave_parametro %in% "n_tot") %>% 
  filter(ano %in% 2018:2024)

line_plot_n_tot <- plotly::plot_ly(n_tot, x = ~fecha_realizacion, y = ~value, color = ~clave_sitio, type = "scatter", mode = "lines") %>% 
  layout(title = "n_tot", 
         xaxis = list(title = "Year"),
         yaxis = list(title = "Value"))
line_plot_n_tot
```

#### For those that have data for 2020-2024

```{r}
# see if each parameter has data in >= 4 of 5 years (somewhere)
param_year_presence_4of5 <- spy %>%
  group_by(clave_parametro) %>%
  summarise(years_with_any_site = n_distinct(ano), .groups = "drop") %>%
  filter(years_with_any_site >= 4)

params_2020_2024_4of5 <- sp %>%
  group_by(clave_parametro) %>%
  summarise(
    sites_union_2020_2024 = n_distinct(clave_sitio),
    sites_5of5            = sum(years_monitored == 5),
    sites_4of5plus        = sum(years_monitored >= 4),
    pct_sites_4of5plus    = sites_4of5plus / sites_union_2020_2024,
    median_years_per_site = median(years_monitored),
    median_samples_per_site_year = median(median_samples_per_year),
    .groups = "drop"
  ) %>%
  inner_join(param_year_presence_4of5, by = "clave_parametro") %>%
  arrange(desc(sites_4of5plus), desc(sites_union_2020_2024))

params_2020_2024_4of5 %>%
  kable("html", caption = "Parameters with coverage in all years (2020–2024)") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed", "responsive"))

```

