---
title: 'OHI GoC 2025: Clean Waters'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
    include-in-header: '../../../src/templates/ohi_hdr.html'
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: inline
editor_options: 
  chunk_output_type: inline
---

# Clean Waters Goal

## RENAMECA

Resultados de la Red Nacional de Medición de Calidad del Agua (RENAMECA).
[Location of surface and underground sites of the National Water Quality Measurement Network monitored during the period 2012-2024.](https://www.gob.mx/conagua/es/articulos/resultados-de-la-red-nacional-de-medicion-de-calidad-del-agua-renameca?idiom=es)

> CONAGUA, through the National Water Quality Measurement Network (RENAMECA), carries out systematic and ongoing monitoring of the water quality of the country's main bodies of water.
> Water quality measurement includes the analysis of physicochemical and microbiological parameters, established according to the type of water body to be characterized.

> On the other hand, RENAMECA addresses specific environmental issues related to water resources at the national level, including requests from the National Human Rights Commission.

> Finally, it should be noted that water quality information contributes to decision-making at all three levels of government to improve the sustainable management of water resources, as well as to informing society about the current status of major bodies of water.

The RENAMECA results are presented according to the following categories:

1.  **By type of water body**


A.  *Lotic* (Water bodies that always move in the same direction such as rivers, streams and creeks)

B.  *Lentic* (Still or stagnant inland waters, such as dams, lakes and reservoirs)

C.  *Coastal* (Coastal waters, such as beaches, bays, marshes and estuaries)

D.  *Subterranean* (Underground waters such as cenotes, galleries and springs)


2.  **By State**

3.  **All results**

[Download RENAMECA Results](https://files.conagua.gob.mx/aguasnacionales/TODOS%20LOS%20MONITOREOS.xlsb)

### Read in data

```{r}
# first let's read in the .xslsb
library(readxl)
library(here)
library(tidyverse)
library(janitor)

# sheet 1: the site names/IDs tested for water quality and their location
sitios_raw <- read_excel(here::here("/home/shares/ohi/OHI_GOC/_raw_data/CONAGUA_RENAMECA/d2025","TODOS_LOS_MONITOREOS.xlsm"), sheet = 1, col_names = TRUE)

# sheet 2: the annual results for all sites
resultados_raw <- read_excel(here::here("/home/shares/ohi/OHI_GOC/_raw_data/CONAGUA_RENAMECA/d2025","TODOS_LOS_MONITOREOS.xlsm"), sheet = 2, col_names = TRUE)

# sheet 3: the water quality variables tested... 462 of them.
simbologia_raw <- read_excel(here::here("/home/shares/ohi/OHI_GOC/_raw_data/CONAGUA_RENAMECA/d2025","TODOS_LOS_MONITOREOS.xlsm"), sheet = 3, col_names = TRUE)
```

### Clean and join

At least for now, let's only look at coastal sites (costeros) in the GoC, and then join sheet 1 with sheet 2 to have the results for only those sites.

```{r}
# filter the sites to only be coastal
unique(sitios_raw$`TIPO DE CUERPO DE AGUA`)
colnames(sitios_costeros)

sitios_costeros <- sitios_raw %>% 
  janitor::clean_names() %>% 
  filter(tipo_de_cuerpo_de_agua %in% c("COSTERO", "COSTERO (HUMEDAL)", "LÓTICO - COSTERO", "LÓTICO - COSTERO (HUMEDAL)", "LÉNTICO - COSTERO", "LÉNTICO - COSTERO (HUMEDAL)","COSTERO Y SEDIMENTOS")) %>%  # all coastal/coastal adjacent sites
  select(-c(clave_acuifero, acuifero)) # only present in non-coastal sites


# filter sites to be in GoC
unique(sitios_costeros$estado)

sitios_costeros_goc <- sitios_costeros %>% 
  filter(estado %in% c("BAJA CALIFORNIA SUR","NAYARIT","SONORA","BAJA CALIFORNIA","SINALOA", "JALISCO"))
```

#### Filter for GoC sites

This is to see how many are not actually *in* the Gulf, but are instead on the Pacific-side of the peninsula.

```{r}
library(sf)

sitios_costeros_goc_sf <- st_as_sf(sitios_costeros_goc, coords = c("longitud", "latitud"), crs = 4326)
head(sitios_costeros_goc_sf) # looks correct
st_crs(sitios_costeros_goc_sf) # good, WGS 84 for now


library(mapview)
mapview(sitios_costeros_goc_sf)
```


```{r}
# many are outside of the GoC.  Let's see if we can narrow down using the municipio...
unique(sitios_costeros_goc_sf$municipio)

gulf_municipios <- c("LOS CABOS", "LA PAZ", "MULEGÉ", "COMONDÚ", "LORETO", 
                     "TECUALA", "SANTIAGO IXCUINTLA", "BAHÍA DE BANDERAS", 
                     "SAN BLAS", "COMPOSTELA", "ROSAMORADA", "TOMATLÁN", 
                     "CIHUATLÁN", "LA HUERTA", "PUERTO VALLARTA", "CABO CORRIENTES",
                     "HERMOSILLO", "HUATABAMPO", "SAN IGNACIO RÍO MUERTO", 
                     "GUAYMAS", "EMPALME", "PUERTO PEÑASCO", "NAVOLATO", 
                     "CULIACÁN", "GUASAVE", "AHOME", "ANGOSTURA", "ELOTA", 
                     "SAN IGNACIO", "MAZATLÁN", "ROSARIO", "ESCUINAPA","MEXICALI")

sitios_costeros_goc_sf_2 <- sitios_costeros_goc_sf %>% 
  filter(municipio %in% gulf_municipios)

mapview(sitios_costeros_goc_sf_2) # oh no... there are some inconsistencies.
```

There are some inconsistencies, in which there are sites that are labeled with incorrect municipio names.  For example, clave_sitio DLBAJ166 is on the Pacific-side, not Gulf-side, yet the municipio is "MULEGÉ".  This is incorrect, because MULEGÉ is quite small and is Gulf-side.  It also seems that I have lost some northern sites...some sites in Bahia Los Angeles are designated as "ENSENADA"...

```{r}
# let's try to use the cuerpo_de_agua as a delimiter
unique(sitios_costeros_goc_sf$cuerpo_de_agua)

gulf_water_bodies <- c(
  "GOLFO DE CALIFORNIA",
  "BAHIA CONCEPCION", 
  "BAHIA DE LA PAZ",
  "BAHIA LA VENTANA",
  "BAHIA DE MUERTOS", 
  "BAHIA LAS PALMAS",
  "BAHIA SAN JOSE DEL CABO",
  "BAHIA SAN LUCAS",
  "LAGUNA AGUA GRANDE",
  "ESTERO TEACAPAN",
  "ESTERO TEAPAQUEÑOS", 
  "RIO SAN PEDRO",
  "CUERPO DE AGUA DEL HUMEDAL",
  "ESTERO EL CONCHAL",
  "ESTERO GACHUPINES",
  "ESTERO VARADERO", 
  "RIO GRANDE DE SANTIAGO",
  "RIO AMECA",
  "ESTERO EL QUELELE",
  "LAGUNA EL PESCADERO",
  "LAGUNA AGUA BRAVA",
  "ESTERO LAGUNA LOS VERGELES",
  "ESTERO CUAUTLA",
  "SISTEMA LAGUNA AGUA BRAVA",
  "BARRA DE NOVILLERO",
  "LAGUNA LA SABANITA", 
  "LAGUNA SABANITA",
  "LAGUNA BARRA DE NAVIDAD",
  "BAHIA MELAQUE",
  "BAHIA MELAQUE 1",
  "ESTERO LA MANZANILLA",
  "RÍO PURIFICACIÓN",
  "RÍO SAN NICOLAS", 
  "LAGUNA CHALACATEPEC",
  "LAGUNA XOLA-PARAMAN",
  "ESTERO MAJAHUAS",
  "RIO TOMATLAN",
  "ESTERO EL CHORRO",
  "CUERPO DE AGUA DEL HUMEDAL - LAGUNA COSTERA EL ERMITAÑO",
  "CUERPO DE AGUA DEL HUMEDAL - LAGUNA AGUA DULCE",
  "RÍO MARABASCO",
  "ESTERO LA CRUZ",
  "HUMEDALES DE YAVAROS-MORONCARIT 1",
  "HUMEDALES DE YAVAROS-MORONCARIT 2", 
  "HUMEDALES DE YAVAROS-MORONCARIT 3",
  "HUMEDALES DE YAVAROS-MORONCARIT 4",
  "HUMEDALES DE YAVAROS-MORONCARIT 5",
  "HUMEDALES DE YAVAROS-MORONCARIT 6",
  "HUMEDALES DE YAVAROS-MORONCARIT 7",
  "LAGUNA YAVAROS",
  "RIO MAYO",
  "DREN JUPATECO",
  "COMPLEJO LAGUNAR BAHIA GUASIMAS-ESTERO LOBOS",
  "HUMEDAL - BAHIA DE LOBOS",
  "RIO YAQUI",
  "BAHIA DE GUAYMAS",
  "AFLUENTE DE DESCARGAS AGUA DE MAR Y CUERPO DE DESCARGA BAHIA EL PARAJE",
  "ESTERO EL SOLDADO",
  "BAHIA LAS GUÁSIMAS",
  "LAGUNA AGIABAMPO",
  "SISTEMA LAGUNAR AGIABAMPO-BACREHUIS-RÍO FUERTE ANTIGUO 1",
  "SISTEMA LAGUNAR AGIABAMPO-BACREHUIS-RÍO FUERTE ANTIGUO 2",
  "SISTEMA LAGUNAR AGIABAMPO-BACREHUIS-RÍO FUERTE ANTIGUO 3",
  "SISTEMA LAGUNAR AGIABAMPO-BACREHUIS-RÍO FUERTE ANTIGUO 4",
  "SISTEMA LAGUNAR AGIABAMPO-BACREHUIS-RÍO FUERTE ANTIGUO 5",
  "SISTEMA LAGUNAR AGIABAMPO",
  "SISTEMA LAGUNAR AGIABAMPO 4",
  "BAHIA DE LOS ANGELES",
  "RIO CULIACAN",
  "BAHIA DE ALTATA SEG",
  "BAHIA ALTATA SALIDA",
  "BAHÍA ENSENADA PABELLONES",
  "LAGUNA PLAYA COLORADA-SANTA MARIA LA REFORMA",
  "BAHIA DE OUHIRA TOPOLOBAMPO SEG",
  "LAGUNAS DE SANTA MARIA TOPOLOBAMPO-OHUIRA",
  "LAGUNA DE SANTA MARIA TOPOLOBAMPO OHUIRA",
  "BAHÍA EL COLORADO",
  "RIO SINALOA",
  "BAHIA DE ALTATA",
  "RIO SAN LORENZO",
  "RÍO SAN LORENZO",
  "BAHIA DE CEUTA",
  "LAGUNA HUIZACHE-CAIMANERO",
  "LAGUNA HUIZACHE",
  "SISTEMA LAGUNAR HUIZACHE - CAIMANERO",
  "RIO BALUARTE",
  "ESTERO DE URIAS",
  "ESTERO URIAS/B. MAZATLAN",
  "ESTERO EL INFIERNILLO/B. MAZATLAN"
)

sitios_costeros_goc_sf_3 <- sitios_costeros_goc_sf %>% 
  filter(cuerpo_de_agua %in% gulf_water_bodies)

mapview(sitios_costeros_goc_sf_3) # NOT working... OCEANO PACIFICO used in the northern gulf and pacific side......
```

Okay, I think by watershed is the best way. 

```{r}
# let's try by watershed.
unique(sitios_costeros_goc_sf$cuenca)

#  [1] "SAN JOSÉ DEL CABO"                       "SAN LUCAS"                              
#  [3] "TODOS SANTOS"                            "ALFREDO B. BONFIL"                      
#  [5] "SAN IGNACIO"                             "SANTO DOMINGO"                          
#  [7] "LORETO"                                  "BAHÍA CONCEPCIÓN"                       
#  [9] "MULEGÉ"                                  "SANTA AGUEDA"                           
# [11] "SANTA ROSALÍA"                           "PUNTA EUGENIA"                          
# [13] "SAN LUIS"                                "VIZCAÍNO"                               
# [15] "EL ARCO"                                 "LA PURÍSIMA"                            
# [17] "SANTA RITA"                              "LA PAZ"                                 
# [19] "EL COYOTE"                               "LOS PLANES"                             
# [21] "SAN BARTOLO"                             "SANTIAGO"                               
# [23] "CABO PULMO"                              "GRUPO DE CORRIENTES MARISMAS NACIONALES"
# [25] "RÍO SAN PEDRO-DESEMBOCADURA"             "AMECA IXTAPA B"                         
# [27] "SAN BLAS"                                "RÍO SANTIAGO 6"                         
# [29] "IXTAPA"                                  "HUICICILA"                              
# [31] "RÍO ACAPONETA 2"                         "RÍO MARÍA GARCÍA"                       
# [33] "RÍO PURIFICACIÓN"                        "RÍO SAN NICOLÁS B"                      
# [35] "RÍO SAN NICOLÁS A"                       "RÍO TOMATLÁN B"                         
# [37] "RÍO IPALA"                               "RÍO MARABASCO A"                        
# [39] "TECOMALA"                                "PITILLAL"                               
# [41] "RÍO SONORA 3"                            "RÍO MAYO 3"                             
# [43] "RÍO YAQUI 3"                             "RÍO MÁTAPE 2"                           
# [45] "GRUPO DE CORRIENTES AGIABAMPO"           "LOS VIDRIOS 1"                          
# [47] "RÍO SONOYTA 1"                           "RÍO CONCEPCIÓN"                         
# [49] "DESCANSO-LOS MÉDANOS"                    "TIJUANA"                                
# [51] "GUADALUPE"                               "ENSENADA-EL GALLO"                      
# [53] "SAN CARLOS"                              "MANEADERO-LAS ANIMAS"                   
# [55] "SANTO TOMÁS"                             "SAN VICENTE"                            
# [57] "SAN RAFAEL"                              "SAN TELMO"                              
# [59] "SAN QUINTÍN"                             "SANTA CATARINA"                         
# [61] "BAHÍA SAN FELIPE"                        "EL BORREGO"                             
# [63] "VALLE DE SAN LUIS"                       "TEPETATE"                               
# [65] "SAN PEDRO"                               "RÍO CULIACÁN"                           
# [67] "GRUPO DE CORRIENTES PABELLONES"          "GRUPO DE CORRIENTES HUYAQUI"            
# [69] "GRUPO DE CORRIENTE TPOPOLOBAMPO"         "RÍO FUERTE 2"                           
# [71] "RÍO SINALOA 2"                           "GRUPO DE CORRIENTES REFORMA"            
# [73] "RÍO PERICOS 2"                           "RÍO SAN LORENZO 2"                      
# [75] "RÍO ELOTA"                               "GRUPO DE CORRIENTES TEMPEHUAYA"         
# [77] "RÍO PIAXTLA 2"                           "GRUPO DE CORRIENTES HUIZACHE"           
# [79] "RÍO BALUARTE 2"                          "GRUPO DE CORRIENTES MAZATLAN"           
# [81] "RÍO QUELITE 2"                           "BALANDRA Y EL MERITO"   


# those in the gulf-side:
cuencas_goc <- c("SAN JOSÉ DEL CABO", "TODOS SANTOS","LORETO","MULEGÉ","SANTA ROSALÍA","BAHÍA SAN FELIPE","EL BORREGO","VALLE DE SAN LUIS","LOS VIDRIOS 1","RÍO SONOYTA 1","RÍO CONCEPCIÓN","TEPETATE","SAN PEDRO","RÍO SONORA 3","RÍO MÁTAPE 2","RÍO YAQUI 3","RÍO MAYO 3","GRUPO DE CORRIENTES AGIABAMPO","RÍO FUERTE 2","GRUPO DE CORRIENTE TPOPOLOBAMPO","RÍO SINALOA 2","GRUPO DE CORRIENTES HUYAQUI","GRUPO DE CORRIENTES REFORMA","RÍO PERICOS 2","GRUPO DE CORRIENTES PABELLONES","RÍO CULIACÁN","GRUPO DE CORRIENTES PABELLONES","RÍO SAN LORENZO 2","GRUPO DE CORRIENTES TEMPEHUAYA","RÍO ELOTA","RÍO PIAXTLA 2","RÍO QUELITE 2","GRUPO DE CORRIENTES MAZATLAN","GRUPO DE CORRIENTES HUIZACHE","RÍO BALUARTE 2","GRUPO DE CORRIENTES MARISMAS NACIONALES","RÍO ACAPONETA 2","RÍO SAN PEDRO-DESEMBOCADURA","RÍO SANTIAGO 6","SAN BLAS","IXTAPA","HUICICILA","AMECA IXTAPA B","PITILLAL","TECOMALA","SANTA ROSALÍA","SANTA AGUEDA","MULEGÉ","BAHÍA CONCEPCIÓN","ALFREDO B. BONFIL","LA PAZ","EL COYOTE","BALANDRA Y EL MERITO","LOS PLANES","SAN BARTOLO","SANTIAGO","CABO PULMO","SAN JOSÉ DEL CABO","SAN LUCAS")

sitios_costeros_goc_sf_4 <- sitios_costeros_goc_sf %>% 
  filter(cuenca %in% cuencas_goc) %>% 
  filter(clave_sitio != "DLBAJ106W2") # out of GoC

mapview(sitios_costeros_goc_sf)
mapview(sitios_costeros_goc_sf_4)
```



#### Join with results

```{r}
# clean the column names
resultados <- resultados_raw %>% 
  janitor::clean_names()

# join cleaned sheet 1 and sheet 2
sitios_resultados_goc <- left_join(sitios_costeros_goc_sf_4, resultados, by = c("clave_sitio","nombre_del_sitio")) 

# check that everything is there...
length(unique(sitios_resultados_goc$clave_sitio))
length(unique(sitios_costeros_goc_sf_4$clave_sitio)) # both 268! yay. that means there are 268 water quality testing sites for the GoC!

# check the number of years per site
years_per_site <- sitios_resultados_goc %>% 
  group_by(clave_sitio) %>% 
  summarise(num_years = length(unique(.$ano))) # yay!! They all have 12, which is from 2012-2024.

colnames(sitios_resultados_goc) # 480 cols

# out of curiosity, it looks like only 131 variables have at least one non-NA value
filtered_for_col_wo_NAs <- sitios_resultados_goc %>% 
  select(where(~ any(!is.na(.)))) %>% 
  mutate(across(where(is.character), ~ str_replace_all(.x, "<", "")))

##### let's save these subsets, which is in wide format, but contains data only from the Gulf.
write_csv(filtered_for_col_wo_NAs, here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/int", "renameca_goc_at_least_one_obs.csv"))
write_csv(sitios_resultados_goc, here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/int", "renameca_goc_wide.csv"))

wide_check <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/int", "renameca_goc_wide.csv"), col_types = cols(.default = "c"))
clean_wide_check <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/int", "renameca_goc_at_least_one_obs.csv"), col_types = cols(.default = "c")) # remember to use the col_types otherwise you will lose the observations that had "<" that have not been mutated to as.character or as.numeric -- we will do that later
```


```{r}
# for fun, what about for hg_tot?
years_per_site_hg <- sitios_resultados_goc %>% 
  group_by(clave_sitio, ano) %>%
  summarise(hg_tot_values = sum(!is.na(hg_tot)), .groups = "drop") # depends on the site.  some have multiple observations of results per site per year.

hg_tot_only <- sitios_resultados_goc %>% 
  select(clave_sitio, fecha_realizacion,ano, hg_tot)
```


### Explore the available metrics 

...and see how many years they have been monitored!

```{r}
# first, let's see the symbology:
simbologia <- simbologia_raw %>% 
  janitor::clean_names()

## now, lets pivot longer so that there is a column named "clave_parametro" like the simbologia df

# see what columns are available (aka variables monitored)
colnames(filtered_for_col_wo_NAs) # okay, so we want to pivot, but likely depth and temperature should be their own pivot. and maybe their own df???

# pivot all the water quality indicators EXCEPT temp profiles
sitios_long_wq <- filtered_for_col_wo_NAs %>% 
  select(-starts_with("temp_")) %>%  # note that profundidad = depth (meters), st = (total solids), caudal (L/s) is "flow or discharge"
  mutate(across(c(alc_fen:st), ~ as.numeric(.x))) %>% 
  pivot_longer(
    cols = c(alc_fen:st),
    names_to = "clave_parametro",
    values_to = "value"
  )

# pivot temperature profile separately
sitios_long_temp <- filtered_for_col_wo_NAs %>%
  pivot_longer(
    cols = starts_with("temp_agua"),
    names_to = "depth_raw",
    values_to = "temp_C"
  ) %>%
  select(-c(alc_fen:st)) %>% 
  mutate(
    # clean depth information from column names
    depth_m = case_when(
      depth_raw == "temp_agua_sup" ~ 0, # surface ("superficie")
      depth_raw == "temp_agua_med" ~ NA_real_, # "media" ("average")... hard for us to know, come back to this!!!
      depth_raw == "temp_agua_fon" ~ NA_real_, # "fondo" ("background..." maybe environmental?)
      TRUE ~ as.numeric(
        str_remove(depth_raw, "temp_agua_") %>%    # remove prefix
          str_replace("_", ".")   # for the 0.5 meters and 39.5
      )
    )
  )
```

```{r}
# now, lets bind the water quality goc subset with the simbologia to see what each of the variables represent

simb_clean <- simbologia %>% 
  mutate(clave_parametro = str_to_lower(clave_parametro),
         nombre_parametro = str_to_lower(nombre_parametro))

wq_goc_w_simb <- full_join(sitios_long_wq, simb_clean, by = "clave_parametro") # 19 variables, correct!

mapview(wq_goc_w_simb)

# some years early on would sample 8 times in a year, but now it is typically 1, if any.
wq_variables_avail <- wq_goc_w_simb %>% 
  st_drop_geometry() %>% 
  group_by(clave_parametro, ano, clave_sitio) %>% 
  summarise(times_monitored = n(), .groups = "drop")
```


```{r}
# summarise by parameter and year
param_year_counts <- wq_goc_w_simb %>%
  st_drop_geometry() %>%             
  group_by(clave_parametro, ano) %>%
  summarise(times_monitored = n(), .groups = "drop")

# filter for variables that cover all years from 2018:2024
params_full_coverage <- param_year_counts %>%
  group_by(clave_parametro) %>%
  summarise(years_monitored = n_distinct(ano), .groups = "drop") %>%
  filter(years_monitored == 7)   # 7 years: 2018-2024

# OR just first filter the year...
params_full_coverage <- param_year_counts %>%
  filter(ano %in% 2018:2024) %>% 
  group_by(clave_parametro) %>%
  summarise(years_monitored = n_distinct(ano), .groups = "drop") %>%
  filter(years_monitored == 7)   # 7 years: 2018-2024

params_full_coverage
```

```{r}
library(dplyr)
library(tidyr)

# Step 1: summarise to param-year
param_year_counts <- wq_goc_w_simb %>%
  st_drop_geometry() %>%
  group_by(clave_parametro, ano) %>%
  summarise(times_monitored = n(), .groups = "drop")

# Step 2: define full set of years
all_years <- 2018:2024

# Step 3: create complete grid param × year
diagnostic_table <- param_year_counts %>%
  complete(clave_parametro, ano = all_years, fill = list(times_monitored = 0)) %>%
  mutate(monitored = times_monitored > 0)

# Step 4: flag which parameters have full coverage
diagnostic_summary <- diagnostic_table %>%
  group_by(clave_parametro) %>%
  summarise(
    years_monitored = sum(monitored),
    missing_years = paste(all_years[!all_years %in% ano[monitored]], collapse = ", "),
    .groups = "drop"
  ) %>%
  mutate(has_full_coverage = years_monitored == length(all_years))

```






