---
title: 'OHI GoC 2025: Clean Waters: HAEDAT (HABs) Mexico Specific'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: inline
editor_options: 
  chunk_output_type: inline
---


# Mexico Specific Harmful Algal Event Database (HAEDAT) Exploration

**Using this site instead: <https://haedat.iode.org/>**

What is the Harmful Algal Event Database?
The Harmful Algal Event Database (HAEDAT) is a component of the Harmful Algal Information system (HAIS) within the "International Oceanographic Data and Information Exchange" (IODE) of the "Intergovernmental Oceanographic Commission" (IOC) of UNESCO, and in cooperation with ICES, and PICES.

The HAEDAT is a meta database containing records of harmful algal events. HAEDAT contains records from the ICES area (North Atlantic) since 1985, and from the PICES area (North Pacific) since 2000. IOC Regional networks in South America, South Pacific and Asia, and North Africa are preparing to contribute. Guidance on submission of data and questions re HAEDAT can be found here.

HAEDAT Disclaimer: The HAEDAT database contains information based on yearly national reports by ICES and PICES member states. The available information on individual events varies greatly from event to event or country to country. Monitoring intensity, number of monitoring stations, number of samplings, stations, etc. also varies greatly and therefore there is not a direct proportionality between recorded events and actual occurrences of e.g. toxicity in a given region. Furthermore, areas with numerous recorded occurrences of HAE's, but with an efficient monitoring and management programmes, may have very few problems and a low risk of intoxications, whereas rare HAE's in other areas may cause severe problems and represent significant health risks.

To download:

1. "Browse Events" tab at the top menu
2. Click on "MEXICO"
3. Click on "[Download these events as a CSV file]" at the top right.
4. Open in Excel and resave as an .xlsx to fix delimiter issue.


Other events not in data:
Short answer: yes—there’s solid evidence of at least one 2024 HAB episode in the Upper Gulf of California (Alto Golfo), plus some 2024 investigations elsewhere that did not confirm HAB.

What we can say for 2024

Upper Gulf (Baja California side): State/COFEPRIS notices show sanitary closures (veda sanitaria) in January 2024 in Alto Golfo harvest polygons (Los Amigos del Puerto, Demetrio Soberanes). This indicates a HAB/toxin risk significant enough for closures. 
El Informador

Upper Gulf (region-wide note): Early March 2024 press (citing CICESE monitoring) reported a red tide confirmed in the vaquita habitat; authorities later indicated the immediate danger had passed—still a 2024 HAB signal for the Alto Golfo. 
Excélsior
+1

Bahía de La Paz (BCS): Mid-July 2024 fish kills were investigated; COEPRIS BCS explicitly ruled out red tide after microscopy—so do not treat this as a confirmed 2024 HAB entry for La Paz. 
ssbcs.gob.mx
+2
diarioelindependiente.mx
+2

COFEPRIS HAB page: Central reference hub for state HAB emergencies/closures; useful to cross-check updates through 2024. 
Gobierno de México

Suggested entries for your dataset

You can add these with “medium” confidence (official closure + press confirmation), pending species/footprint details:

event_name	event_year	standardized_name	state	initial_date	final_date	notes
MX-24-001	2024	Alto Golfo de California (polígonos Los Amigos del Puerto / Demetrio Soberanes)	Baja California	2024-01-18	2024-01-(open)	COFEPRIS/state closure due to marea roja risk; mark date range as at least 18–19 Jan from notices; fill end date when you find the lift notice. 
El Informador

MX-24-002	2024	Alto Golfo de California (vaquita habitat)	BC / Sonora (regional)	~2024-03-04	~2024-03-10	Press citing CICESE monitoring: marea roja confirmada, “pasó el peligro”; treat as regional HAB episode; add species when available. 
Excélsior
+1

I did not add Bahía de La Paz (July 2024) because the state lab said no HAB; if you keep an “investigated events” tab, list it there with outcome = “ruled out.” 
ssbcs.gob.mx


## Exploration

```{r}
library(tidyverse)
library(here)
library(janitor)
library(sf)
library(mapview)
library(readxl)

# ISSUE WITH DELIMITERS: resave as xlsx!!!
# mex_hab_occurrence_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/HAEDAT_mex/d2025","haedat_search.csv"))

# resaved as .xlsx
mex_hab_occurrence_raw <- read_excel(here::here("/home/shares/ohi/OHI_GOC/_raw_data/HAEDAT_mex/d2025","haedat_search.xlsx"), sheet = 1, col_names = TRUE)

```



```{r}
# clean by removing NA rows!
mex_hab_occurrence <- mex_hab_occurrence_raw %>% 
  janitor::clean_names() %>% 
  filter(!is.na(event_name)) %>% 
  mutate(event_year = as.character(event_year)) 

unique(mex_hab_occurrence$region)

# [1] "Sinaloa"                                    "Quintana Roo"                               "Baja California Sur"                       
#  [4] "Golfo de California"                        "MulegÈ, Baja California Sur"                "Navolato, Sinaloa"                         
#  [7] "Colima"                                     "El Caribe"                                  "Chiapas"                                   
# [10] "SInaloa"                                    "Baja California"                            "Oaxaca"                                    
# [13] "Jalisco"                                    "Sonora"                                     "Tamaulipas, north of Gulf of Mexico"       
# [16] "Guerrero"                                   "Yucatan"                                    "Yucat·n"                                   
# [19] "Nayarit"                                    "Campeche"                                   "Peninsula de Yucat·n and Mexican Caribbean"
# [22] "Sinalo"                                     "Todos Santos Bay"                           NA                                          
# [25] "Tabasco"                                    "Tamaulipas"                                 "Tehuantepec"                               
# [28] "Veracruz"                                   "Los Mochis, Sinaloa"                        "Michoac·n"                                 
# [31] "Bja California Sur"                         "La Paz, B. C. S."                           "Todos Santos Bay, B. C. N." 

mex_hab_goc <- mex_hab_occurrence %>% 
  filter(!region %in% c("Colima","Guerrero","Tabasco","Veracruz","Quintana Roo","El Caribe","Yucatan","Campeche","Todos Santos Bay","Tamaulipas","Chiapas","Oaxaca","Tamaulipas, north of Gulf of Mexico","Yucat·n","Peninsula de Yucat·n and Mexican Caribbean","Tehuantepec","Michoac·n","Todos Santos Bay, B. C. N."))

unique(mex_hab_goc$region)
#  [1] "Sinaloa"                     "Baja California Sur"         "Golfo de California"         "MulegÈ, Baja California Sur"
#  [5] "Navolato, Sinaloa"           "SInaloa"                     "Baja California"             "Jalisco"                    
#  [9] "Sonora"                      "Nayarit"                     "Sinalo"                      NA                           
# [13] "Los Mochis, Sinaloa"         "Bja California Sur"          "La Paz, B. C. S."
```

Let's take a look at the lats/longs that are already provided in the dataset (note: most do not have lat/longs):
```{r}
 mex_hab_goc_sf <- mex_hab_goc %>% 
   filter(!is.na(latitude)) %>% 
   st_as_sf(coords = c("longitude", "latitude"),
            crs = 4326, # WGS84
            remove = FALSE)
# mapview(mex_hab_goc_sf)
```

Unfortunately, some of these look accurate, and the others look very inaccurate. For example, MX-12-022 should be in Puerto Peñasco, but it is not. 

I identified the unique locations in the data and used ChatGPT to help me get coordinates and a standardized name. This was also used to identify the sites in the Gulf of California.

```{r}

lat_lon <- read_csv(here("CW/v2025/data/Habs_locations.csv")) %>%
  rename(location_text = locs) %>%
  select(-region)

setdiff(mex_hab_goc$location_text, lat_lon$location_text)

habs_locs <- mex_hab_goc %>%
  select(-region) %>%
  left_join(lat_lon, by="location_text") %>%
  filter(GofC %in% "Gulf of California") 


```

Next I am going to associate the estimated size of the HAB with the data.
These are currently based on estimates from ChatGPT based on the description of the region and, in some cases, the length of the outbreak. Ideally we can find a better source of information for these sizes. But we are going to go with this for now.

```{r}
size <- read_csv(here("CW/v2025/data/HAB_size_estimate.csv")) %>%
  select(area, event_name)
setdiff(habs_locs$event_name, size$event_name)

habs_locs_size <- left_join(habs_locs, size, by="event_name") 

habs_locs_size[duplicated(habs_locs_size$event_name),]

# appear to be replicates, just picking the first one.
#tmp <- filter(habs_locs_size, event_name %in% c("MX-11-015", "MX-10-017"))
habs_locs_size <- habs_locs_size[!duplicated(habs_locs_size$event_name), ]

#  Events MX-17-014 and MX-17-013 seem like they might be the same event, but I am going to keep them as
# separate events for now. I am going to change the year to the longer one to 2018
# since most of its time series was in 2018
filter(habs_locs_size, location_text %in% "Alto Golfo de California") %>%
  unique()

habs_locs_size$event_year[habs_locs_size$event_name == "MX-17-014"] <- 2018

```

Looking at variables to use for rescaling.

```{r}

explore <- habs_locs_size %>%
  select(event_name, occurred_before, high_phyto, seafood_toxin, mass_mortal,
         foam_mucil, toxicity_detected, humans_affected, natural_fish_affected, aquaculture_fish_affected, planktonic_affected, benthic_affected, shellfish_affected, birds_affected, aquatic_mammals_affected) %>%
  mutate(num_affected = humans_affected + natural_fish_affected + aquaculture_fish_affected + planktonic_affected + benthic_affected + shellfish_affected + birds_affected + aquatic_mammals_affected) %>%
  select(-natural_fish_affected, -aquaculture_fish_affected, -planktonic_affected, -benthic_affected, -shellfish_affected, -birds_affected, -aquatic_mammals_affected)
plot(jitter(check$num_effected), check$toxicity_detected)
plot(jitter(check$num_effected), check$mass_mortal)



score <- explore %>%
   mutate(score = case_when(
      mass_mortal == 1 ~ 1,
      num_affected >= 3 ~ 1,
      num_affected == 2 ~ 0.75,
      num_affected == 1 ~ 0.50,
      num_affected == 0 & toxicity_detected == 1 ~ 0.25,
      TRUE ~ 0.1
    )) %>%
  select(event_name, score)


```

Create a plot by size.
```{r}
plot_data <- habs_locs_size %>%
  select(event_name, event_year, standardized_name, state, lat, lon, area) %>%
  left_join(score, by="event_name") %>%
  filter(event_year >= 2015)

pts_wgs <- st_as_sf(plot_data,
                    coords = c("lon", "lat"),
                    crs = 4326, # WGS84
                    remove = FALSE)
mapview::mapview(pts_wgs)


# 2) Project to a meter-based CRS (Mexico ITRF2008 / LCC = EPSG:6372)
pts_mx <- st_transform(pts_wgs, 6372)

# 3) Build circular polygons with target area = `area` (assumed km^2)
#    area = pi * r^2  =>  r = sqrt(area_m2 / pi)
polys_mx <- pts_mx %>%
  mutate(
    area_km2   = as.numeric(area),
    area_m2    = area_km2 * 1e6,
    radius_m   = sqrt(area_m2 / pi),
    geometry   = st_buffer(geometry, dist = radius_m)  # circle per point
  )

mapview::mapview(polys_mx)

```
Rasterize by year and save:



```{r}
mex_hab_goc_clean <- mex_hab_goc %>% 
  mutate(region = case_when( # fix all that need fixing
    
     # BCS
    region %in% c("Baja California Sur", 
                  "Bja California Sur", 
                  "MulegÈ, Baja California Sur", 
                  "La Paz, B. C. S.") ~ "Baja California Sur",

    # Sinaloa
    region %in% c("Sinaloa", 
                  "SInaloa", 
                  "Sinalo", 
                  "Navolato, Sinaloa", 
                  "Los Mochis, Sinaloa") ~ "Sinaloa",
    TRUE ~ region
  )) %>% 
  tidygeocoder::geocode(city = location_text, country = country_name, state = region)  # to give each city a lat, long, if possible 
  
mex_hab_goc_cleaner <- mex_hab_goc_clean %>% 
  mutate(longitude = ifelse(is.na(longitude), long, longitude)) %>% 
  mutate(latitude = ifelse(is.na(latitude), lat, latitude)) %>% 
  tidygeocoder::geocode(country = country_name, state = region)

mex_hab_goc_cleanest <- mex_hab_goc_cleaner %>% 
  janitor::clean_names() %>% 
  mutate(longitude = ifelse(is.na(longitude), long_88, longitude)) %>% 
  mutate(latitude = ifelse(is.na(latitude), lat_87, latitude)) %>% 
  mutate(latitude = ifelse(location_text == "Bahía de la Paz", 24.312716332893793, latitude)) %>%  # 24.312716332893793, -110.32933847561796
  mutate(latitude = ifelse(location_text == "Bahia de La Paz", 24.312716332893793, latitude)) %>% 
  mutate(longitude = ifelse(location_text == "Bahia de La Paz", -110.32933847561796, longitude)) %>% 
  filter(!location_text %in% c("Bahia de Acapulco", "BAHIA DE TODOS SANTOS")) 
  
# mex_hab_goc_cleanest_na <- mex_hab_goc_cleanest %>% 
  # filter(is.na(latitude))

mex_hab_goc_cleanest_sf <- mex_hab_goc_cleanest %>% 
  st_as_sf(coords = c("longitude", "latitude"),
           crs = 4326, # WGS84
           remove = FALSE)

mapview(mex_hab_goc_cleanest_sf, zcol = "event_year")
```


```{r}
summary(mex_hab_goc_clean)

library(gtsummary)

mex_hab_summary_tbl <- mex_hab_goc_clean %>% 
  tbl_summary(include = c(event_year, region, occurred_before, humans_affected, fish_affected, causative_species_name0))

print(mex_hab_summary_tbl)
```








