---
title: 'OHI GoC 2025: Clean Waters - Microplastics'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# Clean Waters: Microplastics


```{r}
library(tidyverse) 
library(sf)
library(janitor)
library(here)
library(mapview)
```

## Read in data from NCEI NOAA global microplastics (1972 - 2018)

[From the website:](https://www.ncei.noaa.gov/products/microplastics)

>This product is a publicly available collection of marine microplastics data from around the world. The database is a repository for multiple datasets of marine microplastics. The data are aggregated, archived, and accessible to users in a consistent and reliable manner. Contributors from various research cruises and field work provide the data. This is a growing database and data collectors are encouraged to submit findings to NCEI to be added to the Marine Microplastics application.

>Data Usage Disclaimer: Microplastics vary in origin, density, chemical properties, morphology, size and color. At present, there is no single combination of methods for sampling, extracting, analyzing, and reporting. Thus, the microplastic concentrations in the database may not always be comparable across studies. Users should consider the metadata in the archives which contain more details about the data records (such as sampling protocols and instrumental analysis) for further usability of the data.

```{r}
ncei_raw <- st_read(here::here("/home/shares/ohi/OHI_GOC/_raw_data/NCEI_NOAA_microplastics/d2026/Marine_Microplastic_Concentrations.geojson")) %>% 
  janitor::clean_names()

mapview(ncei_raw, zcol = "sampling_method")
mapview(ncei_raw, zcol = "date_m_d_yyyy")

summary(ncei_raw)
```

For PVC cylinder, which are sand cores of beaches:

```{r}
numeric_cols <- c(
  "latitude_degree", "longitude_degree",
  "ocean_bottom_depth_m", "water_sample_depth_m",
  "sediment_sample_depth_m", "mesh_size_mm",
  "transect_no", "volunteers_number",
  "collecting_time_min", "standardized_nurdle_amount",
  "microplastics_measurement"
)

ncei_pvc_sediment <- ncei_raw %>% 
  filter(short_reference %in% "Alvarez-Zeferino et al. 2020") %>% 
  mutate(across(all_of(numeric_cols), as.numeric))

summary(ncei_pvc_sediment)

mapview(ncei_pvc_sediment, zcol = "microplastics_measurement")
```

Convert to particles/kg:

```{r}
# [Based on this article: ](https://cementindustrial.com/sand-density-and-how-to-measure-sand-particle-size/) loose sand has a density of 1442 kg/m³,  and wet packed sand has a density of 2082 kg/m³.

ncei_pvc_sediment_kg <- ncei_pvc_sediment %>% 
  mutate(microplastics_measurement_mp_kg_loose = microplastics_measurement/1442,
         microplastics_measurement_mp_kg_packed = microplastics_measurement/2082)
```

Surface water:

```{r}
ncei_surface_water_neuston_nets <- ncei_raw %>% 
  filter(short_reference %in% "Law et al.2014") %>% 
  mutate(across(all_of(numeric_cols), as.numeric))

summary(ncei_surface_water_neuston_nets$microplastics_measurement)
mapview(ncei_surface_water_neuston_nets, zcol = "microplastics_measurement")
```

Grab samples:

```{r}
ncei_grab_samples <- ncei_raw %>% 
  filter(short_reference %in% "Barrows et al.2018") %>% 
  mutate(across(all_of(numeric_cols), as.numeric))

summary(ncei_grab_samples$microplastics_measurement)
mapview(ncei_grab_samples, zcol = "microplastics_measurement")
```


## Other data sources without data for the GOC from Hafiz et al. 2026 (2019 - 2026) - DUD

1.  [A global dataset of microplastics in urban stormwater runoff (2018–2024)](https://www.nature.com/articles/s41597-026-07137-y)

[Data availability:](https://osf.io/9cgu4/overview)

"Abstract: Microplastic (<5 mm) pollution is a growing environmental concern, with urban stormwater systems serving as critical transport pathways. While microplastics predominantly originate and circulate within urban environments, transported via stormwater runoff to receiving water bodies, existing databases focus almost exclusively on marine and coastal waters. This mismatch between contamination sources and data infrastructure has left urban stormwater systems without dedicated data resources to support research and management. In this research, we present the Database of Urban RUnoff Microplastics (DURUM), a global database dedicated to microplastics in urban stormwater systems. The objective of DURUM is to compile, standardize, and serve urban runoff microplastic data in a consistent format to enable data-driven approaches for developing predictive transport models. DURUM systematically compiles 180 sampling efforts from peer-reviewed studies published between 2018 and 2024, encompassing surface runoff from streets, residential catchments, commercial districts, industrial areas, and drainage infrastructure across 15 countries. This paper describes DURUM’s structure, compilation methodology, standardization protocols, and future research directions."

GitHub: [abdulmobinibnahafiz/DURUM-Urban-Runoff-Microplastics](https://github.com/abdulmobinibnahafiz/DURUM-Urban-Runoff-Microplastics)
DOI: https://doi.org/10.17605/OSF.IO/9CGU4
License (Data): CC-BY 4.0 | License (Code): MIT

**UNFORTUNATELY, NO DATA IN THE GULF OF CALIFORNIA.**


2.  [A Review of Global Microplastic (MP) Databases: A Study on the Challenges and Opportunities for Data Integration in the Context of MP Pollution, Ahamed et al 2025](https://www.mdpi.com/2673-4605/22/1/6)

"Abstract: Microplastic (MP) pollution is an escalating global environmental concern, with a growing body of research addressing diverse dimensions of this issue. Despite this progress, the field remains hindered by generating large, heterogeneous datasets that follow inconsistent reporting standards, resulting in fragmented and often incompatible databases. While various databases on MPs have been developed, they primarily operate in isolation, limiting the accessibility and cross-comparison of data. This study presents a foundational approach to aggregating and accessing existing MP pollution datasets. A comprehensive review of the currently available databases was conducted to evaluate their integration potential. It revealed key challenges such as non-standardized data formats, limited accessibility, and difficulty performing comparative analyses across sources. To address these barriers, a prototype web-based platform was developed that enables unified access to MP datasets. The architecture includes a smart standardization layer that harmonizes inputs from disparate sources. The integration of Large Language Models (LLMs) with Retrieval-Augmented Generation (RAG) techniques was proposed to facilitate natural language querying. This enables researchers to interact with the platform intuitively and extract meaningful insights more efficiently. The proposed system aims to enhance data discoverability, promote interoperability, and support robust, data-driven environmental research, paving the way toward more informed policy-making and scientific collaboration in the fight against MP pollution. With this platform, there is a potential for new discoveries and a future in which the tools to effectively combat this global issue are available, making the audience realize the potential for new discoveries."

[Has really cool application with multiple data sources on microplastics globally, but these do not have GoC data...](https://microplastics.streamlit.app/)

**UNFORTUNATELY, NO DATA IN THE GULF OF CALIFORNIA.**


## Mexico-specific papers regarding micro- and macro- plastic

1. [The state of microplastic pollution in México: a review and evolving perspectives, Caro-Martínez et al. 2025](https://www.sciencedirect.com/science/article/abs/pii/S0048969725014135)

>"Abstract: Microplastic pollution has emerged as a significant global environmental concern. These small plastic fragments, originating from the degradation of larger plastics, infiltrate marine and terrestrial ecosystems. This article reviews the current state of research on microplastics in Mexico using the PRISMA method. The information collected focused on the distribution, characterization and abundances of microplastics across different environmental matrices. A literature search was conducted in high-quality databases from 2014 to 2024, resulting in 80 scientific articles documenting studies in environmental matrices such as sediments/soil, biota, water and atmosphere across 20 states of Mexico. The most extensively studied environmental matrix is sediment -particularly beach sand-, followed by biota, water and atmosphere. The most reported microplastic shapes and color are transparent and blue fibers, primarily made of polypropylene and polyethylene. It is imperative to standardize the methods for studying microplastics to serve as a monitoring tool for assessing and comparing microplastic abundances both nationally and globally. The standardization of sampling, laboratory analysis, and abundance reporting units will facilitate the comparison of studies. Establishing a national microplastics database would support long term monitoring and inform mitigation strategies. While this initiative aligns with Sustainable Development Goal 14 -Life Below Water- specifically target 14.1.1.b aimed at reducing marine pollution, the widespread presence of microplastics in freshwater, soil and air underscores the need to integrate SDGs 6, SDG 12 and SDG 15 to address plastic pollution holistically."



2. [Inventories of plastic pollution sources, flows and hotspots as a baseline for national action plans: The experience of Mexico,  Vázquez-Morillas et al. 2024](https://www.sciencedirect.com/science/article/pii/S0048969724074953)

>"Abstract: This article explores the development of Mexico's National Inventory of Plastic Pollution Sources (NIPPS), designed to establish a baseline for the country's National Action Plan. The study employed a multifaceted approach, combining material flow analysis tools, literature reviews, and stakeholder workshops. This methodology successfully identified critical leakage points within the waste management system, such as deficiencies in collection and the prevalence of uncontrolled disposal sites. It also recognized priority geographic regions and fostered engagement with relevant stakeholders. However, a comprehensive assessment was hampered by data limitations, including a lack of information on waste composition, inconsistent recycling data, and the use of varied methodologies in field studies of plastic pollution. By highlighting the strengths and limitations of the process, this article underscores the critical role of standardized data collection and analysis in creating robust national plastic pollution inventories. Such inventories serve as a cornerstone for effective National Action Plans and contribute significantly to broader efforts tackling the global plastic pollution crisis. The valuable lessons from Mexico's experience offer insights for other countries facing similar challenges."





