---
title: 'OHI GoC 2025: Clean waters pesticides'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: inline
editor_options: 
  chunk_output_type: inline
---

# CW: pesticides

## Download agricultural land raster

```{r}
library(utils)
library(tidyverse)
library(here)
library(raster)

zip_file_path <- "/home/shares/ohi/OHI_GOC/_raw_data/inegi_landuse_vegetation/d2025/794551132210.zip"
extraction_directory <- "/home/shares/ohi/OHI_GOC/_raw_data/inegi_landuse_vegetation/d2025/" # directory to extract the files
unzip(zipfile = zip_file_path, exdir = extraction_directory)

# Warning in unzip(zipfile = zip_file_path, exdir = extraction_directory) :
#   zip file is corrupt
```

```{r}
## lets try again!! (downloaded once more)
url  <- "https://www.inegi.org.mx/contenidos/productos/prod_serv/contenidos/espanol/bvinegi/productos/geografia/tematicas/uso_suelo_veg/int_art/794551132210_t.zip"
dest <- "/home/shares/ohi/OHI_GOC/_raw_data/inegi_landuse_vegetation/d2025/USYVIA_IA2018.zip"

message("Downloading IA 2018 dataset...")

resp <- GET(
  url,
  timeout(600),             # 10 min
  write_disk(dest, overwrite = TRUE),
  progress()
)

stop_for_status(resp)
message("Download finished: ", dest)

zip_file_path <- "/home/shares/ohi/OHI_GOC/_raw_data/inegi_landuse_vegetation/d2025/USYVIA_IA2018.zip"
extraction_directory <- "/home/shares/ohi/OHI_GOC/_raw_data/inegi_landuse_vegetation/d2025/" # directory to extract the files
unzip(zipfile = zip_file_path, exdir = extraction_directory) # STILL: Warning in unzip(zipfile = zip_file_path, exdir = extraction_directory) :zip file is corrupt
```

```{r}
## keep trying...
zip_file <- "/home/shares/ohi/OHI_GOC/_raw_data/inegi_landuse_vegetation/d2025/USYVIA_IA2018.zip"
exdir <- "/home/shares/ohi/OHI_GOC/_raw_data/inegi_landuse_vegetation/d2025/USYVIA_IA2018/"


unzip(zip_file, exdir = exdir)
tif_files <- list.files(exdir, pattern = "\\.tif$", full.names = TRUE)
rasters <- lapply(tif_files, raster) # Warning in unzip(zip_file, exdir = exdir) : zip file is corrupt

```

STILL NOT WORKIING!!

### Using the terminal to finish unzipping

In the terminal, do:

"cd /home/shares/ohi/OHI_GOC/_raw_data/inegi_landuse_vegetation/d2025/

while true; do
  echo "Checking/continuing download..."
  wget -c --timeout=900 --tries=10 \
    "https://www.inegi.org.mx/contenidos/productos/prod_serv/contenidos/espanol/bvinegi/productos/geografia/tematicas/uso_suelo_veg/int_art/794551132210_t.zip" \
    -O USYVIA_IA2018.zip

  echo "Verifying ZIP integrity..."
  unzip -t USYVIA_IA2018.zip >/tmp/zipcheck.txt 2>&1

  if grep -q "No errors detected" /tmp/zipcheck.txt; then
    echo "File verified successfully!"
    break
  else
    echo "File still incomplete or corrupt — retrying in 60 s..."
    sleep 60
  fi
done"

To ensure everything was downloaded. You should see:

<!-- 🔁 Checking/continuing download... -->
<!-- --2025-10-31 19:26:27--  https://www.inegi.org.mx/contenidos/productos/prod_serv/contenidos/espanol/bvinegi/productos/geografia/tematicas/uso_suelo_veg/int_art/794551132210_t.zip -->
<!-- Resolving www.inegi.org.mx (www.inegi.org.mx)... 200.23.8.5 -->
<!-- Connecting to www.inegi.org.mx (www.inegi.org.mx)|200.23.8.5|:443... connected. -->
<!-- HTTP request sent, awaiting response... 416 Requested Range Not Satisfiable -->

<!--     The file is already fully retrieved; nothing to do. -->

<!-- Verifying ZIP integrity... -->
<!-- File verified successfully! -->

Then, do:

"cd /home/shares/ohi/OHI_GOC/_raw_data/inegi_landuse_vegetation/d2025/
unzip -t USYVIA_IA2018.zip"

to check if it is there and fully unzipped. You should see:

<!-- "Archive:  USYVIA_IA2018.zip -->
<!--     testing: map_08nov2023_meta_004003.tif   OK -->
<!--     testing: map_08nov2023_meta_005001.tif   OK -->
<!--     testing: map_08nov2023_meta_005002.tif   OK -->
<!--     testing: map_08nov2023_meta_005003.tif   OK -->
<!--     testing: map_08nov2023_meta_005004.tif   OK -->
<!--     testing: map_08nov2023_meta_006001.tif   OK -->
<!--     testing: map_08nov2023_meta_006002.tif   OK -->
<!--     testing: map_08nov2023_meta_006003.tif   OK -->
<!--     testing: map_08nov2023_meta_006004.tif   OK -->
<!--     testing: map_08nov2023_meta_006005.tif   OK -->
<!--     testing: map_08nov2023_meta_007001.tif   OK -->
<!--     testing: map_08nov2023_meta_007002.tif   OK -->
<!--     testing: map_08nov2023_meta_007003.tif   OK -->
<!--     testing: map_08nov2023_meta_007004.tif   OK -->
<!--     testing: map_08nov2023_meta_007005.tif   OK -->
<!--     testing: map_08nov2023_meta_007006.tif   OK -->
<!--     testing: map_08nov2023_meta_007007.tif   OK -->
<!--     testing: map_08nov2023_meta_008002.tif   OK -->
<!--     testing: map_08nov2023_meta_008003.tif   OK -->
<!--     testing: map_08nov2023_meta_008004.tif   OK -->
<!--     testing: map_08nov2023_meta_008005.tif   OK -->
<!--     testing: map_08nov2023_meta_008006.tif   OK -->
<!--     testing: map_08nov2023_meta_008007.tif   OK -->
<!--     testing: map_08nov2023_meta_008008.tif   OK -->
<!--     testing: map_08nov2023_meta_009002.tif   OK -->
<!--     testing: map_08nov2023_meta_009003.tif   OK -->
<!--     testing: map_08nov2023_meta_009004.tif   OK -->
<!--     testing: map_08nov2023_meta_009005.tif   OK -->
<!--     testing: map_08nov2023_meta_009006.tif   OK -->
<!--     testing: map_08nov2023_meta_009007.tif   OK -->
<!--     testing: map_08nov2023_meta_009008.tif   OK -->
<!--     testing: map_08nov2023_meta_010002.tif   OK -->
<!--     testing: map_08nov2023_meta_010003.tif   OK -->
<!--     testing: map_08nov2023_meta_010004.tif   OK -->
<!--     testing: map_08nov2023_meta_010005.tif   OK -->
<!--     testing: map_08nov2023_meta_010006.tif   OK -->
<!--     testing: map_08nov2023_meta_010007.tif   OK -->
<!--     testing: map_08nov2023_meta_011002.tif   OK -->
<!--     testing: map_08nov2023_meta_011003.tif   OK -->
<!--     testing: map_08nov2023_meta_011004.tif   OK -->
<!--     testing: map_08nov2023_meta_011005.tif   OK -->
<!--     testing: map_08nov2023_meta_011006.tif   OK -->
<!--     testing: map_08nov2023_meta_011007.tif   OK -->
<!--     testing: map_08nov2023_meta_011008.tif   OK -->
<!--     testing: map_08nov2023_meta_011009.tif   OK -->
<!--     testing: map_08nov2023_meta_011010.tif   OK -->
<!--     testing: map_08nov2023_meta_012002.tif   OK -->
<!--     testing: map_08nov2023_meta_012003.tif   OK -->
<!--     testing: map_08nov2023_meta_012004.tif   OK -->
<!--     testing: map_08nov2023_meta_012005.tif   OK -->
<!--     testing: map_08nov2023_meta_012006.tif   OK -->
<!--     testing: map_08nov2023_meta_012007.tif   OK -->
<!--     testing: map_08nov2023_meta_012008.tif   OK -->
<!--     testing: map_08nov2023_meta_012009.tif   OK -->
<!--     testing: map_08nov2023_meta_012010.tif   OK -->
<!--     testing: map_08nov2023_meta_012011.tif   OK -->
<!--     testing: map_08nov2023_meta_013003.tif   OK -->
<!--     testing: map_08nov2023_meta_013004.tif   OK -->
<!--     testing: map_08nov2023_meta_013005.tif   OK -->
<!--     testing: map_08nov2023_meta_013006.tif   OK -->
<!--     testing: map_08nov2023_meta_013007.tif   OK -->
<!--     testing: map_08nov2023_meta_013008.tif   OK -->
<!--     testing: map_08nov2023_meta_013009.tif   OK -->
<!--     testing: map_08nov2023_meta_013010.tif   OK -->
<!--     testing: map_08nov2023_meta_013011.tif   OK -->
<!--     testing: map_08nov2023_meta_013012.tif   OK -->
<!--     testing: map_08nov2023_meta_014003.tif   OK -->
<!--     testing: map_08nov2023_meta_014004.tif   OK -->
<!--     testing: map_08nov2023_meta_014005.tif   OK -->
<!--     testing: map_08nov2023_meta_014006.tif   OK -->
<!--     testing: map_08nov2023_meta_014007.tif   OK -->
<!--     testing: map_08nov2023_meta_014008.tif   OK -->
<!--     testing: map_08nov2023_meta_014009.tif   OK -->
<!--     testing: map_08nov2023_meta_014009.tif.aux.xml   OK -->
<!--     testing: map_08nov2023_meta_014009.tif.vat.cpg   OK -->
<!--     testing: map_08nov2023_meta_014009.tif.vat.dbf   OK -->
<!--     testing: map_08nov2023_meta_014010.tif   OK -->
<!--     testing: map_08nov2023_meta_014011.tif   OK -->
<!--     testing: map_08nov2023_meta_014012.tif   OK -->
<!--     testing: map_08nov2023_meta_015003.tif   OK -->
<!--     testing: map_08nov2023_meta_015004.tif   OK -->
<!--     testing: map_08nov2023_meta_015005.tif   OK -->
<!--     testing: map_08nov2023_meta_015006.tif   OK -->
<!--     testing: map_08nov2023_meta_015007.tif   OK -->
<!--     testing: map_08nov2023_meta_015008.tif   OK -->
<!--     testing: map_08nov2023_meta_015009.tif   OK -->
<!--     testing: map_08nov2023_meta_015010.tif   OK -->
<!--     testing: map_08nov2023_meta_015011.tif   OK -->
<!--     testing: map_08nov2023_meta_015012.tif   OK -->
<!--     testing: map_08nov2023_meta_015013.tif   OK -->
<!--     testing: map_08nov2023_meta_016005.tif   OK -->
<!--     testing: map_08nov2023_meta_016006.tif   OK -->
<!--     testing: map_08nov2023_meta_016007.tif   OK -->
<!--     testing: map_08nov2023_meta_016008.tif   OK -->
<!--     testing: map_08nov2023_meta_016009.tif   OK -->
<!--     testing: map_08nov2023_meta_016010.tif   OK -->
<!--     testing: map_08nov2023_meta_016011.tif   OK -->
<!--     testing: map_08nov2023_meta_016012.tif   OK -->
<!--     testing: map_08nov2023_meta_016013.tif   OK -->
<!--     testing: map_08nov2023_meta_017006.tif   OK -->
<!--     testing: map_08nov2023_meta_017007.tif   OK -->
<!--     testing: map_08nov2023_meta_017008.tif   OK -->
<!--     testing: map_08nov2023_meta_017009.tif   OK -->
<!--     testing: map_08nov2023_meta_017010.tif   OK -->
<!--     testing: map_08nov2023_meta_017011.tif   OK -->
<!--     testing: map_08nov2023_meta_017012.tif   OK -->
<!--     testing: map_08nov2023_meta_017013.tif   OK -->
<!--     testing: map_08nov2023_meta_017014.tif   OK -->
<!--     testing: map_08nov2023_meta_018010.tif   OK -->
<!--     testing: map_08nov2023_meta_018011.tif   OK -->
<!--     testing: map_08nov2023_meta_018012.tif   OK -->
<!--     testing: map_08nov2023_meta_018013.tif   OK -->
<!--     testing: map_08nov2023_meta_018014.tif   OK -->
<!--     testing: map_08nov2023_meta_019011.tif   OK -->
<!--     testing: map_08nov2023_meta_019012.tif   OK -->
<!--     testing: map_08nov2023_meta_019013.tif   OK -->
<!--     testing: map_08nov2023_meta_019014.tif   OK -->
<!--     testing: map_08nov2023_meta_020012.tif   OK -->
<!--     testing: map_08nov2023_meta_020013.tif   OK -->
<!--     testing: map_08nov2023_meta_020014.tif   OK -->
<!--     testing: map_08nov2023_meta_021009.tif   OK -->
<!--     testing: map_08nov2023_meta_021010.tif   OK -->
<!--     testing: map_08nov2023_meta_021011.tif   OK -->
<!--     testing: map_08nov2023_meta_021012.tif   OK -->
<!--     testing: map_08nov2023_meta_021013.tif   OK -->
<!--     testing: map_08nov2023_meta_021014.tif   OK -->
<!--     testing: map_08nov2023_meta_022008.tif   OK -->
<!--     testing: map_08nov2023_meta_022009.tif   OK -->
<!--     testing: map_08nov2023_meta_022010.tif   OK -->
<!--     testing: map_08nov2023_meta_022011.tif   OK -->
<!--     testing: map_08nov2023_meta_022012.tif   OK -->
<!--     testing: map_08nov2023_meta_022013.tif   OK -->
<!--     testing: map_08nov2023_meta_023008.tif   OK -->
<!--     testing: map_08nov2023_meta_023009.tif   OK -->
<!--     testing: map_08nov2023_meta_023010.tif   OK -->
<!--     testing: map_08nov2023_meta_023011.tif   OK -->
<!--     testing: map_08nov2023_meta_023012.tif   OK -->
<!--     testing: map_08nov2023_meta_024009.tif   OK -->
<!--     testing: map_08nov2023_meta_024010.tif   OK -->
<!--     testing: map_08nov2023_meta_024011.tif   OK -->
<!--     testing: map_08nov2023_meta_024012.tif   OK -->
<!--     testing: map_08nov2023_meta_025009.tif   OK -->
<!--     testing: map_08nov2023_meta_025010.tif   OK -->
<!--     testing: map_08nov2023_meta_025011.tif   OK -->
<!--     testing: map_08nov2023_meta_025011.tif.vat.cpg   OK -->
<!--     testing: map_08nov2023_meta_025011.tif.vat.dbf   OK -->
<!-- No errors detected in compressed data of USYVIA_IA2018.zip." -->

Okay!!! Now unzip in the terminal without using R's unzip() command, which does not handle large .zip archives well (especially those with many files).

"unzip USYVIA_IA2018.zip -d USYVIA_IA2018/"

### Now we can look at the individual tiles!
```{r}
map_08nov2023_meta_005001 <- terra::rast(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_landuse_vegetation/d2025/USYVIA_IA2018","map_08nov2023_meta_005001.tif"))

# check the current CRS
crs(map_08nov2023_meta_005001) # v2025: "PROJCRS[\"unnamed\",\n    BASEGEOGCRS[\"GRS 1980(IUGG, 1980)\",\n        DATUM[\"unknown\",\n            ELLIPSOID[\"GRS80\",6378137,298.257222101,\n      "
plot(map_08nov2023_meta_005001) # to get a quick visualization

# check the number of layers
num_layers <- terra::nlyr(map_08nov2023_meta_005001)
print(num_layers) # yup, just 1
```

```{r}
map_08nov2023_meta_019011 <- terra::rast(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_landuse_vegetation/d2025/USYVIA_IA2018","map_08nov2023_meta_019011.tif"))

# check the current CRS
crs(map_08nov2023_meta_019011) # v2025: "PROJCRS[\"unnamed\",\n    BASEGEOGCRS[\"GRS 1980(IUGG, 1980)\",\n        DATUM[\"unknown\",\n            ELLIPSOID[\"GRS80\",6378137,298.257222101,\n      " 
plot(map_08nov2023_meta_019011) # to get a quick viz

# check the number of layers
num_layers <- terra::nlyr(map_08nov2023_meta_005001)
print(num_layers) # yup, just 1
```
### Merge the tiles into one mosaic raster!!

```{r}
# directory path to all tiles
tiles_dir <- here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_landuse_vegetation/d2025/USYVIA_IA2018")

# lList all .tif files
tif_files <- list.files(tiles_dir, pattern = "\\.tif$", full.names = TRUE)

# check the number of them
length(tif_files) # 141
head(tif_files)
# [1] "/home/shares/ohi/OHI_GOC/_raw_data/inegi_landuse_vegetation/d2025/USYVIA_IA2018/map_08nov2023_meta_004003.tif"
# [2] "/home/shares/ohi/OHI_GOC/_raw_data/inegi_landuse_vegetation/d2025/USYVIA_IA2018/map_08nov2023_meta_005001.tif"
# [3] "/home/shares/ohi/OHI_GOC/_raw_data/inegi_landuse_vegetation/d2025/USYVIA_IA2018/map_08nov2023_meta_005002.tif"
# [4] "/home/shares/ohi/OHI_GOC/_raw_data/inegi_landuse_vegetation/d2025/USYVIA_IA2018/map_08nov2023_meta_005003.tif"
# [5] "/home/shares/ohi/OHI_GOC/_raw_data/inegi_landuse_vegetation/d2025/USYVIA_IA2018/map_08nov2023_meta_005004.tif"
# [6] "/home/shares/ohi/OHI_GOC/_raw_data/inegi_landuse_vegetation/d2025/USYVIA_IA2018/map_08nov2023_meta_006001.tif"

# load them all as SpatRasters
tiles <- lapply(tif_files, terra::rast)

# merge into one raster.... this will take time!!!!
# (mosaic takes care of overlapping borders)
mosaic_raster <- do.call(terra::mosaic, tiles)

# save it
all_tiles_path <- here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_landuse_vegetation/d2025/USYVIA_IA2018/IA2018_mosaic.tif")
terra::writeRaster(mosaic_raster, all_tiles_path, overwrite = TRUE)

# does it look right??
mosaic_raster <- terra::rast(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_landuse_vegetation/d2025/USYVIA_IA2018/IA2018_mosaic.tif"))
plot(mosaic_raster, main = "Uso del Suelo y Vegetación IA 2018 (Mosaico INEGI)") ## yay!!

tab <- terra::freq(mosaic_raster)  # returns a data.frame of pixel counts per raster value
colnames(tab)
# [1] "layer" "value" "count"
```

### Classification of land

```{r}
library(readxl)

classification <- readxl::read_xlsx(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_landuse_vegetation/d2025/sist_clas_peaseuv_ia", "claves_PEUSUEV_IA.xlsx"), skip = 1) %>% janitor::clean_names()
colnames(classification)
# [1] "num"         "clave"       "descripcion" "nota"

```

```{r}
tab_joined <- left_join(tab, classification, by = c("value" = "num"))
```
```{r}
# classification_unique <- classification %>%
#   distinct(num, .keep_all = TRUE)
# 
# raster_vals <- sort(unique(terra::values(mosaic_raster)))
# 
# class_labels <- classification_unique %>%
#   filter(num %in% raster_vals) %>%
#   arrange(num) %>%
#   pull(descripcion)
```

