---
title: 'OHI GoC 2025: Clean Waters: Chl-A'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: inline
editor_options: 
  chunk_output_type: inline
---

# Global Ocean Colour (Copernicus-GlobColour), Bio-Geo-Chemical, L4 (monthly and interpolated) from Satellite Observations (1997-ongoing): Chlorophyll-A concentration

For the Global Ocean Satellite Observations, ACRI-ST company (Sophia Antipolis, France) is providing Bio-Geo-Chemical (BGC) products based on the Copernicus-GlobColour processor.

*Upstreams:* SeaWiFS, MODIS, MERIS, VIIRS-SNPP & JPSS1, OLCI-S3A & S3B for the ""multi"" products, and S3A & S3B only for the ""olci"" products.

*Variables:* Chlorophyll-a (CHL), Phytoplankton Functional types and sizes (PFT), Primary Production (PP), Suspended Matter (SPM), Secchi Transparency Depth (ZSD), Diffuse Attenuation (KD490), Particulate Backscattering (BBP), Absorption Coef. (CDM) and Reflectance (RRS).

*Temporal resolutions:* monthly plus, for some variables, daily gap-free based on a space-time interpolation to provide a ""cloud free"" product.

*Spatial resolutions:* 4 km and a finer resolution based on olci 300 meters inputs.

Recent products are organized in datasets called Near Real Time (NRT) and long time-series (from 1997) in datasets called Multi-Years (MY).

To find the Copernicus-GlobColour products in the catalogue, use the search keyword ""GlobColour""."

*DOI (product):*
https://doi.org/10.48670/moi-00281

*Product identifier*
OCEANCOLOUR_GLO_BGC_L4_MY_009_104

*Product name*
Global Ocean Colour (Copernicus-GlobColour), Bio-Geo-Chemical, L4 (monthly and interpolated) from Satellite Observations (1997-ongoing)

*Dataset*
Plankton, multi-sensor (4 km), monthly   ·   cmems_obs-oc_glo_bgc-plankton_my_l4-multi-4km_P1M

<https://data.marine.copernicus.eu/product/OCEANCOLOUR_GLO_BGC_L4_MY_009_104/download?dataset=cmems_obs-oc_glo_bgc-plankton_my_l4-multi-4km_P1M_202411> 

## Setup

```{r}
# load packages
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}

librarian::shelf(
  here,
  CopernicusMarine,
  janitor,
  raster,
  terra,
  maps,
  httr,
  jsonlite,
  purrr,
  tictoc,
  sf, 
  googleVis,
  RColorBrewer,
  foreach,
  doParallel, # for using multiple cores, if needed
  tidyverse, 
  ohicore,
  ggplot2,
  reticulate, # for python coding
  ncmeta,
  ncdf4,
  plotly
)


# monthly chl-a rasters
int_monthly_dir <- here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chla_int/monthly_rasters")
# dir.create(int_dir) # to create the path on Mazu if it has not already been done

chl_files <- list.files(int_monthly_dir, full=TRUE)
plot(log(rast(chl_files[1])+1))
pos_colors <- colorRampPalette(c("#75A68F", "#C0C19E", "#FBC8A4", "#FD928F", "#FE3A5A"))(100)

month <- "02"
rast_tmp <- rast(sprintf("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chla_int/monthly_rasters/2024-%s-01.tif", month))
tmp <- ifel(rast_tmp>7, 7, rast_tmp)
plot(tmp, col=pos_colors, main=month)

# 1) linear scale]
scaled <- (rast_tmp - 5) / 5

# 2) apply thresholds
rescaled <-  ifel(rast_tmp < 5, 0,
              ifel(rast_tmp >= 10, 1,
                   scaled))

pos_colors <- c("#87B1DE", colorRampPalette(c("#DEF2BB", "#ECE46A", "#F8CD21", "#FA8433", "#FE3A5A"))(100))
plot(rescaled, col=pos_colors, zlim=c(0, 1))

# 1) linear scale
scaled <- (rast_tmp - 2) / (5-2)

# 2) apply thresholds
rescaled <-  ifel(rast_tmp < 2, 0,
              ifel(rast_tmp >= 5, 1,
                   scaled))

plot(rescaled, col=pos_colors,  zlim=c(0, 1))


## Trend
slopes <- list.files("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chla_int/trend/", full=TRUE, pattern="slope")
tmp <- rast(slopes)
slope_avg <- mean(tmp)
quants <- quantile(values(slope_avg), c(0.01, 0.99), na.rm=TRUE)
slope_avg <- ifel(slope_avg<quants[1], quants[1], slope_avg)
slope_avg <- ifel(slope_avg>quants[2], quants[2], slope_avg)

# rescale:
# 1) linear scale
scaled <- (slope_avg - 0) / (0.1-0)

# 2) apply thresholds
rescaled_trend <-  ifel(slope_avg < 0, 0,
              ifel(slope_avg >= 0.1, 1,
                   scaled))

plot(rescaled_trend, col=pos_colors,  zlim=c(0, 1))



pos_colors <- colorRampPalette(c("#87B1DE", "#DEF2BB", "#ECE46A", "#F8CD21", "#FA8433", "#FE3A5A"))(100)

plot(slope_avg, col=pos_colors, zlim=c(-0.05, 0.15))


jan <- rast("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chla_int/trend//month01_chl_trend_slope.tif")
quants <- quantile(values(jan), c(0.01, 0.99), na.rm=TRUE)
jan <- ifel(jan<quants[1], quants[1], jan)
jan <- ifel(jan>quants[2], quants[2], jan)
  
pos_colors <- colorRampPalette(c("#87B1DE", "#87B1DE", "#87B1DE", "#DEF2BB", "#ECE46A", "#F8CD21", "#FA8433", "#FE3A5A"))(100)  
plot(jan, col=pos_colors, zlim=c(-0.05, 0.15))

## combine
pressure <- mean(rescaled_trend, rescaled)
plot(pressure, col=pos_colors,  zlim=c(0, 1))

regions <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp"))
regions_ll <- st_transform(regions, crs(pressure)) %>%
  filter(location == "marine")

plot(pressure, col = pos_colors, zlim = c(0, 1))
plot(st_geometry(regions_ll), add = TRUE, border = "black", lwd = 0.75)

scores <- terra::zonal(pressure, vect(regions_ll), fun="mean", na.rm=TRUE)
```


## Checking NAs, since this can be a problem with these data.
```{r}

# this seems fine! I don't see any systematic problem here.
na_count_data <- data.frame()
years <- 1998:2024
for(year in years){ #year = years[27]
  tmp_files <- grep(chl_files, pattern=year, value=TRUE)
  tmp_stack <- rast(tmp_files)
  na_count <- app(tmp_stack, function(x) sum(is.na(x)))
  plot(na_count)
  new_data <- data.frame(table(values(na_count)))
  new_data$year <- year
  na_count_data <- rbind(na_count_data, new_data)
  writeRaster(na_count, sprintf("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chla_int/missing_data_rasters/na_count_%s.tif", year), overwrite = TRUE)
}
write_csv(na_count_data, here("CW/v2025/data/chla_missing_data.csv"))

na_rasts <- list.files("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chla_int/missing_data_rasters/", full=TRUE)
for(year in years){ # year <- years[1]
  rast_file <- grep(na_rasts, pattern = sprintf("na_count_%s.tif", year), value=TRUE)
  plot(raster(rast_file), main=year)
}


```

Do a trend analysis for each month.

```{r}

trend_fun <- function(y, t) {
  # y = cell values
  if (all(is.na(y))) return(c(NA, NA))  # skip cells with no data

  ok <- !is.na(y) & !is.na(t)
  y_ok <- y[ok]
  t_ok <- t[ok]

  # need at least 3 points for a slope
  if (length(y_ok) < 3) return(c(NA, NA))

  fit <- try(lm(y_ok ~ t_ok), silent = TRUE)
  if (inherits(fit, "try-error")) return(c(NA, NA))

  sm <- summary(fit)$coefficients

  slope <- sm["t_ok", "Estimate"]
  pval  <- sm["t_ok", "Pr(>|t|)"]

  c(slope, pval)
}

months <- c("01", "02", "03", "04", "05", "06", "07", "08", "09", "10", "11", "12")
trend_years <- 2015:2024
for(month in months){ # month = months[1]
tmp_month <- grep(chl_files, pattern= paste(trend_years, collapse="|"), value=TRUE)
  tmp_month <- grep(tmp_month, pattern=sprintf("-%s-01.tif", month), value=TRUE)
#tmp_month <- grep(tmp_month, pattern = "/2025-", invert=TRUE, value=TRUE)
month_stack <- rast(tmp_month)
yrs <- as.numeric(format(time(month_stack), "%Y"))

sum_raster <- sum(month_stack, na.rm=TRUE)
writeRaster(sum_raster,  
            sprintf("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chla_int/trend/month%s_chl_mean.tif", month),  overwrite = TRUE)

trend_raster <- app(month_stack, trend_fun, t = yrs)
names(trend_raster) <- c("slope", "p_value")
writeRaster(trend_raster$slope,  
            sprintf("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chla_int/trend/month%s_chl_trend_slope.tif", month),  overwrite = TRUE)
writeRaster(trend_raster$p_value, 
            sprintf("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chla_int/trend/month%s_chl_trend_pvalue.tif", month), overwrite = TRUE)
}


for(i in 1:12){ # i = 1
slopes <- list.files("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chla_int/trend/", full=TRUE, pattern="slope")
tmp_s <- rast(slopes[i])
tmp_s <- ifel(tmp_s>0, 1, -1)

pvalue <- list.files("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chla_int/trend/", full=TRUE, pattern="pvalue")
tmp_p <- rast(pvalue[i])
tmp_p <- ifel(tmp_p<0.05, 1, 0)

out_rasts <- gsub("_chl_trend_slope.tif", "", basename(slopes[i]))

writeRaster(tmp_s,  
            sprintf("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chla_int/trend2/%s_chl_slope.tif", out_rasts),  overwrite = TRUE)
writeRaster(tmp_p, 
            sprintf("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chla_int/trend2/%s_chl_trend.tif", out_rasts), overwrite = TRUE)

out_png <- sprintf("trend_%s.png", gsub("_chl_trend_slope.tif", "", basename(slopes[i])))
png(sprintf(here("CW/v2025/chl_int/monthly_slopes/%s"), out_png), width = 2400, height = 1200, res = 200)

par(mfrow = c(1, 2))      # 1 row, 2 columns
par(mar = c(4, 4, 4, 5))  # give room for legends/titles if needed

plot(tmp_s, main=gsub(".tif", "", basename(slopes[i])), col=c("#008080", "red"))
plot(tmp_p, main=gsub(".tif", "", basename(pvalue[i])), col=c("#ffefb3", "orange"))
dev.off()

}

## get overview

neg_colors <- colorRampPalette(c("#045a8d", "#74a9cf"))(12)   # -12 to -1
zero_color <- "white"                                         # 0
pos_colors <- colorRampPalette(c("#ffefb3", "#ffb400"))(12)   # 1 to 12

pal <- c(neg_colors, zero_color, pos_colors)

slope <- list.files("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chla_int/trend2", pattern= "slope", full=TRUE)
tmp <- rast(slope)
slope_total <- sum(tmp)
plot(slope_total, col=pal, zlim=c(-12, 12))


pal <- c(colorRampPalette(c("#fff8db", "red"))(13)) 
pval <- list.files("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chla_int/trend2", pattern="trend", full=TRUE)
tmp <- rast(pval)
slope_total <- sum(tmp)
plot(slope_total,
     col = pal)

```


The previous analysis provides some indication of change over time. Now I want to assess the magnitude of the change over time. I also rescale the values as one indicator of future pressure (sustainability).

```{r}

int_monthly_dir <- here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chla_int/monthly_rasters")
list.files(int_monthly_dir, full=TRUE)

for(i in 1:12){ # i = 1
slopes <- list.files("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chla_int/trend/", full=TRUE, pattern="slope")
tmp_s <- rast(slopes[i])
quants <- quantile(values(tmp_s), c(0.01, 0.99), na.rm=TRUE)
tmp_s <- ifel(tmp_s<quants[1], quants[1], tmp_s)
tmp_s <- ifel(tmp_s>quants[2], quants[2], tmp_s)
plot(tmp_s)

out_rasts <- gsub("_chl_trend_slope.tif", "", basename(slopes[i]))

writeRaster(tmp_s,  
            sprintf("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chla_int/slope_capped/%s_chl_slope.tif", out_rasts),  overwrite = TRUE)

out_png <- sprintf("trend_%s.png", gsub("_chl_trend_slope.tif", "", basename(slopes[i])))
png(sprintf(here("CW/v2025/chl_int/monthly_slopes_capped/%s"), out_png), width = 2400, height = 1200, res = 200)

par(mfrow = c(1, 2))      # 1 row, 2 columns
par(mar = c(4, 4, 4, 5))  # give room for legends/titles if needed

plot(tmp_s, main=gsub(".tif", "", basename(slopes[i])))
dev.off()

## now rescale between 0 and 1
r <- tmp_s
minv <- 0
maxv <- 0.1

# linear rescale
scaled <- (r - minv) / (maxv - minv)

# clamp
rescaled <- ifel(r <= minv, 0,
             ifel(r >= maxv, 1,
                  scaled))

rescaled
plot(rescaled)

writeRaster(rescaled, sprintf("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chla_int/rescaled_monthly_trend/%s.tif", out_rasts), overwrite=TRUE)
}


## average the monthly data
months <- list.files("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chla_int/rescaled_monthly_trend", full=TRUE)

rast_stack <- rast(months)
avg_raster <- mean(rast_stack, na.rm=TRUE)
plot(avg_raster)
writeRaster(avg_raster,  
            sprintf("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chla_int/current_pressure/current_pressure_%s.tif", year),  overwrite = TRUE)


```


## get exceedences for each year
```{r}

# get mean chlorphyll for a year
int_monthly_dir <- here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chla_int/monthly_rasters")
list.files(int_monthly_dir, full=TRUE)

years <- 1998:2024
months <- c("01", "02", "03", "04", "05", "06", "07", "08", "09", "10", "11", "12")

for(year in years){ # year <- 2024 
for(month in months) { # month <- "01"  
r <- rast(sprintf("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chla_int/monthly_rasters/%s-%s-01.tif", year, month))

# 1) linear scale
min <- 5
max <- 10
scaled <- (r - min) / (max -  min)

# 2) apply thresholds
rescaled <-  ifel(r < min, 0,
              ifel(r >= max, 1,
                   scaled))

writeRaster(rescaled, sprintf("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chla_int/rescaled/%s_%s_rescaled.tif", year, month), overwrite=TRUE)
}}

```

Average monthly exceedences for each year to get current pressure estimate for each raster cell

```{r}
## now average per year
years <- 1998:2024
for(year in years){ #year = 2024
rasts_dir <- list.files("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chla_int/rescaled", full=TRUE, pattern = sprintf("%s_", year))
cat(sprintf("number of rasts for year %s = ", year), length(rasts_dir))

rast_stack <- rast(rasts_dir)
avg_raster <- mean(rast_stack, na.rm=TRUE)

writeRaster(avg_raster,  
            sprintf("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chla_int/current_pressure/current_pressure_%s.tif", year),  overwrite = TRUE)
}


current_pressure <- list.files("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chla_int/current_pressure/", full=TRUE)

pos_colors <- c("#87B1DE", colorRampPalette(c("#DEF2BB", "#ECE46A", "#F8CD21", "#FA8433", "#FE3A5A"))(100))

plot(rast("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chla_int/current_pressure//current_pressure_2024.tif"), col=pos_colors, zlim=c(0,1))
     


```
