---
title: 'OHI GoC 2025: Safety and Security (ENSU)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
    include-in-header: '../../../src/templates/ohi_hdr.html'
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: inline
editor_options: 
  chunk_output_type: inline
---

# Safety and Security Goal: ENSU city perception

## Summary

### Background

This script explores feelings of safety using the National Survey of Urban Public Security (ENSU) for cities in the six states that border or fringe the Gulf of California: Baja California, Baja California Sur, Sonora, Sinaloa, Nayarit, and Jalisco.
Following discussions from the OHI GoC Expert Working Group Meeting (June 2025), the aim is to inform a Safety and Security goal with perception indicators that can be assigned to designed urban domains rather than forced onto municipios that the survey never powered.

The analysis focuses on March rounds across all years available in the ENSU folder.
March is treated as a comparable seasonal snapshot so that year to year change is not mixed with within year seasonality from June, September, or December rounds.

**Important data gap:** there is no March microdata package for 2021 in the INEGI release set used here (2021 folders contain June and later rounds only).

*Note: This analysis maintains sensitivity to regional contexts and focuses on constructive data interpretation as it relates to the Gulf's health.*

### Why ENSU for this goal

ENSU is built to produce city level estimates of perceived public safety for named áreas urbanas de interés.
That matches the OHI need more closely than either ENVIPE or the Mexico Peace Index (MPI) and Mexico Positive Peace Index (MPPI) for the following reasons:

* ENVIPE is designed for national, state, and a short list of urban areas of interest.
  Municipal means from ENVIPE microdata are exploratory only.
* MPI fear of violence is already an ENVIPE based state score.
  MPPI is a state level Positive Peace construct.
  Painting those state scores onto Gulf side municipios imports border and Pacific dynamics and cannot isolate Sea of Cortez localities.
* ENSU asks, by design, whether living in the named city feels safe or unsafe, and whether people changed habits from fear of crime.

### Data source and folder layout

Program page: [INEGI ENSU](https://www.inegi.org.mx/programas/ensu/)

Primary table:

* `ENSU_CB_*`: selected person questionnaire (perception and habit change).

Raw data are read from the first existing path among:

1. `/home/shares/ohi/OHI_GOC/_raw_data/inegi_crime/d2025/ENSU` (shared server)
2. the project local folder `ENSU/` (for example `.../SS_security/ENSU`)

Intermediate products go to `/home/shares/ohi/OHI_GOC/goal_prep/ss/security_int` when that tree exists, otherwise to `goal_prep/ss/security_int` under the project.
Outputs go to `/home/shares/ohi/OHI_GOC/goal_prep/ss/security_output` or `goal_prep/ss/security_output` locally.

### Year to year file and schema changes handled here

| Years | Format | Notes handled by the loader |
|---|---|---|
| 2014 to 2020 | DBF (`.dbf` / `.DBF`) | Read with `foreign::read.dbf` |
| 2022 to present | CSV | Read with `readr::read_csv` |
| 2021 | (no March file) | Explicitly absent; not an error |
| 2014 to 2015 | Legacy questionnaire names | `P1` becomes city safety; `P4_1` to `P4_5` become habit change; `FACTOR` becomes the person weight |
| 2016 onward | Modern `BP1_*` names | `BP1_1` and `BP1_5_*` with `FAC_SEL` |
| through 2016 | `ENT` / `MUN` | Renamed to `CVE_ENT` / `CVE_MUN` |
| 2017 onward | `CVE_ENT`, `NOM_ENT` | State names usually present |
| through 2020 | city code `CD` only | City names filled from a crosswalk built from the latest CSV year that has `NOM_CD` |
| 2022 onward | `NOM_CD` present | Preferred city labels, still keyed by `CD` |
| file names | mixed | Accepts `ENSU_CB_03YY` and `ENSU_CB03YY` |

City codes (`CD`) are the stable join key across years.
City display names can shift slightly (for example `CULIACAN` versus `CULIACAN ROSALES`), so labels are standardized from the newest catalog when possible.

### Methodology notes

ENSU is a probabilistic urban household survey.
Representativeness is for each área urbana de interés in the sample, not for every municipio.

Harmonized analysis items:

* City safety: modern `BP1_1`, or legacy `P1` in 2014 to 2015.
  Codes: `1` safe, `2` unsafe, `9` does not know or does not answer.
* Habit change: modern `BP1_5_1` to `BP1_5_5`, or legacy `P4_1` to `P4_5` in 2014 to 2015.
  Codes: `1` yes, `2` no, `3` not applicable, `9` does not know or does not answer.
* Person weight: modern `FAC_SEL`, or legacy `FACTOR` in 2014 to 2015.

Estimates are weighted shares among substantive responses.
Cities that enter ENSU later will show missing early years; that is coverage change, not a zero rate.

# Setup

```{r}
# install librarian if it is missing
if (!require(librarian)) {
  # install librarian from cran
  install.packages("librarian")
  # attach librarian after install
  library(librarian)
  # end of librarian install branch
}

# load project packages through librarian
librarian::shelf(
  tidyverse, # data wrangling and plotting utilities
  here, # project relative path helpers
  ggplot2, # grammar of graphics plotting
  readr, # fast csv reading and writing
  stringr, # string cleaning helpers
  stringi, # accent stripping for city name joins
  lubridate, # date helpers if needed later
  janitor, # clean_names for ensu columns
  scales, # axis formatting helpers
  purrr, # map over march files by year
  foreign # read dbf microdata for 2020 and earlier
)

# candidate raw ensu roots (server first, then local project)
path_raw_candidates <- c(
  "/home/shares/ohi/OHI_GOC/_raw_data/inegi_crime/d2025/ENSU", # shared ohi server path
  here::here("ENSU") # local project ensu folder
)

# keep only candidate roots that currently exist
path_raw_existing <- path_raw_candidates[dir.exists(path_raw_candidates)]

# stop if no ensu root can be found
if (length(path_raw_existing) == 0) {
  # explain both expected locations
  stop("no ensu raw folder found. expected server path or local project ENSU/ folder.")
}

# use the first existing raw root
path_raw_ensu <- path_raw_existing[[1]]

# candidate intermediate folders
path_int_candidates <- c(
  "/home/shares/ohi/OHI_GOC/goal_prep/ss/security_int", # shared server intermediate path
  here::here("goal_prep", "ss", "security_int") # local intermediate fallback
)

# candidate output folders
path_out_candidates <- c(
  "/home/shares/ohi/OHI_GOC/goal_prep/ss/security_output", # shared server output path
  here::here("goal_prep", "ss", "security_output") # local output fallback
)

# default to local intermediate path
path_int <- path_int_candidates[[2]]

# prefer server intermediate path when the ohi shares tree exists
if (dir.exists("/home/shares/ohi/OHI_GOC") || dir.exists(path_int_candidates[[1]])) {
  # use server intermediate path
  path_int <- path_int_candidates[[1]]
}

# default to local output path
path_out <- path_out_candidates[[2]]

# prefer server output path when the ohi shares tree exists
if (dir.exists("/home/shares/ohi/OHI_GOC") || dir.exists(path_out_candidates[[1]])) {
  # use server output path
  path_out <- path_out_candidates[[1]]
}

# create the intermediate folder if it does not exist
dir.create(path_int, recursive = TRUE, showWarnings = FALSE)

# create the output folder if it does not exist
dir.create(path_out, recursive = TRUE, showWarnings = FALSE)

# print which paths are in use
message("using raw ensu path: ", path_raw_ensu)
# print intermediate path
message("using intermediate path: ", path_int)
# print output path
message("using output path: ", path_out)

# state codes for the six gulf adjacent states
goc_state_codes <- c(
  "02", # baja california
  "03", # baja california sur
  "14", # jalisco
  "18", # nayarit
  "25", # sinaloa
  "26" # sonora
)

# list ensu cities in those states keyed by stable city code cd
goc_city_meta <- tibble::tribble(
  ~cd, ~nom_cd, ~gulf_relevance, # column headers for city metadata
  "04", "LA PAZ", "strong goc candidate", # bcs gulf side city
  "55", "LOS CABOS", "pacific baja", # bcs pacific side city
  "42", "LOS MOCHIS", "northern sinaloa goc adjacent", # ahome urban area
  "94", "CIUDAD OBREGON", "southern sonora coast adjacent", # cajeme urban area
  "43", "HERMOSILLO", "inland sonora metro", # inland sonora capital
  "40", "CULIACAN ROSALES", "inland sinaloa metro", # inland sinaloa capital
  "41", "MAZATLAN", "open pacific sinaloa", # pacific sinaloa city
  "25", "PUERTO VALLARTA", "pacific bahia de banderas fringe", # jalisco pacific fringe
  "33", "TEPIC", "inland nayarit", # nayarit capital
  "59", "GUADALAJARA", "inland jalisco metro", # jalisco core city
  "63", "ZAPOPAN", "inland jalisco metro", # guadalajara metro member
  "60", "TONALA", "inland jalisco metro", # guadalajara metro member
  "62", "SAN PEDRO TLAQUEPAQUE", "inland jalisco metro", # guadalajara metro member
  "61", "TLAJOMULCO DE ZUNIGA", "inland jalisco metro", # guadalajara metro member
  "03", "TIJUANA", "border pacific not goc", # border pacific city
  "02", "MEXICALI", "border not goc", # border city
  "44", "NOGALES", "border sonora not goc" # sonora border city
)
```

The chunk above sets paths (server with local fallback), packages, and city metadata keyed by `CD`.

# Discover March microdata (CSV and DBF)

March files are discovered recursively.
Accepted names include `ENSU_CB_0325.csv` and older spellings such as `ENSU_CB0316.dbf`.
Year 2021 is expected to be missing.

```{r}
# find march cb files in csv or dbf form under the ensu root
march_cb_files <- list.files(
  path = path_raw_ensu,
  pattern = "^ENSU_CB_?03[0-9]{2}\\.(csv|dbf)$",
  full.names = TRUE,
  recursive = TRUE,
  ignore.case = TRUE
)

# stop if nothing was found
if (length(march_cb_files) == 0) {
  # provide a clear recovery message
  stop("no march ensu_cb csv/dbf files found under ", path_raw_ensu)
}

# build an inventory tibble from discovered files
march_inventory <- tibble::tibble(
  path = march_cb_files
) %>%
  # parse yy from either ensu_cb_03yy or ensu_cb03yy
  mutate(
    yy = stringr::str_match(basename(path), "(?i)ENSU_CB_?03([0-9]{2})")[, 2],
    year = dplyr::if_else(as.integer(yy) >= 90L, 1900L + as.integer(yy), 2000L + as.integer(yy)),
    file = basename(path),
    format = tolower(tools::file_ext(path))
  ) %>%
  # prefer csv over dbf if both somehow exist for one year
  arrange(year, dplyr::desc(format == "csv"), path) %>%
  # keep one file per year
  distinct(year, .keep_all = TRUE) %>%
  # flag known missing march year for documentation
  mutate(note = dplyr::if_else(year == 2021L, "unexpected if present; 2021 usually has no march", NA_character_))

# explicitly record that 2021 march is unavailable when absent
missing_2021 <- !2021L %in% march_inventory$year

# message about the 2021 gap when relevant
if (missing_2021) {
  # confirm the expected gap
  message("note: no march 2021 file found (expected). that year will be absent from the series.")
}

# print inventory
print(march_inventory)

# save inventory
readr::write_csv(march_inventory, file.path(path_int, "ensu_march_file_inventory.csv"))
```

# Load and harmonize March CB tables across schema generations

The loader:

1. Reads CSV or DBF
2. Maps legacy 2014 to 2015 names onto modern `BP1_*` / `FAC_SEL` fields
3. Renames `ENT` / `MUN` to `CVE_ENT` / `CVE_MUN`
4. Zero pads codes
5. Attaches standardized city names from a `CD` catalog built after all years are scanned

```{r}
# helper: read one ensu table from csv or dbf as character columns
read_ensu_file <- function(path) {
  # branch on file extension
  ext <- tolower(tools::file_ext(path))
  # read csv microdata
  if (ext == "csv") {
    # read all columns as character
    out <- readr::read_csv(path, col_types = readr::cols(.default = "c"), show_col_types = FALSE)
    # return csv tibble
    return(out)
  }
  # read dbf microdata for 2020 and earlier
  if (ext == "dbf") {
    # foreign::read.dbf returns a data.frame
    out <- foreign::read.dbf(path, as.is = TRUE)
    # convert every column to character for binding safety
    out[] <- lapply(out, function(col) as.character(col))
    # return dbf table as tibble
    return(tibble::as_tibble(out))
  }
  # unsupported extension
  stop("unsupported ensu file type: ", path)
}

# helper: harmonize one march cb file into a common schema
harmonize_ensu_march_cb <- function(path, year) {
  # read raw file
  raw <- read_ensu_file(path)
  # clean names to snake case
  raw <- janitor::clean_names(raw)
  # map legacy 2014 to 2015 safety item p1 onto bp1_1 when needed
  if (!("bp1_1" %in% names(raw)) && ("p1" %in% names(raw))) {
    # copy legacy p1 into bp1_1
    raw$bp1_1 <- raw$p1
  }
  # map legacy habit items p4_1 to p4_5 onto bp1_5_* when needed
  for (i in 1:5) {
    # build modern and legacy names for this habit item
    modern <- paste0("bp1_5_", i)
    # build legacy name
    legacy <- paste0("p4_", i)
    # copy legacy habit item when modern is missing
    if (!(modern %in% names(raw)) && (legacy %in% names(raw))) {
      # assign legacy responses into the modern column name
      raw[[modern]] <- raw[[legacy]]
    }
  }
  # map legacy factor weight onto fac_sel when needed
  if (!("fac_sel" %in% names(raw)) && ("factor" %in% names(raw))) {
    # copy factor into fac_sel
    raw$fac_sel <- raw$factor
  }
  # rename legacy state code ent to cve_ent
  if ("ent" %in% names(raw) && !("cve_ent" %in% names(raw))) {
    # rename ent
    raw <- dplyr::rename(raw, cve_ent = ent)
  }
  # rename legacy municipality code mun to cve_mun
  if ("mun" %in% names(raw) && !("cve_mun" %in% names(raw))) {
    # rename mun
    raw <- dplyr::rename(raw, cve_mun = mun)
  }
  # ensure cve_ent exists even if empty
  if (!("cve_ent" %in% names(raw))) {
    # create empty state code column
    raw$cve_ent <- NA_character_
  }
  # ensure cd exists
  if (!("cd" %in% names(raw))) {
    # create empty city code column
    raw$cd <- NA_character_
  }
  # ensure nom_ent exists
  if (!("nom_ent" %in% names(raw))) {
    # create empty state name column
    raw$nom_ent <- NA_character_
  }
  # ensure nom_cd exists
  if (!("nom_cd" %in% names(raw))) {
    # create empty city name column
    raw$nom_cd <- NA_character_
  }
  # ensure fac_sel exists
  if (!("fac_sel" %in% names(raw))) {
    # create empty weight column
    raw$fac_sel <- NA_character_
  }
  # ensure bp1_1 exists
  if (!("bp1_1" %in% names(raw))) {
    # create empty safety column
    raw$bp1_1 <- NA_character_
  }
  # ensure habit columns exist
  for (i in 1:5) {
    # modern habit column name
    modern <- paste0("bp1_5_", i)
    # add empty habit column when absent
    if (!(modern %in% names(raw))) {
      # create empty habit column
      raw[[modern]] <- NA_character_
    }
  }
  # columns to retain for the multi year bind
  keep_cols <- c(
    "cve_ent", "nom_ent", "cve_mun", "nom_mun",
    "cd", "nom_cd", "per",
    "sexo", "edad",
    "bp1_1",
    "bp1_5_1", "bp1_5_2", "bp1_5_3", "bp1_5_4", "bp1_5_5",
    "fac_sel", "fac_viv",
    "upm", "upm_dis", "est_dis", "dominio"
  )
  # keep only columns that exist
  keep_cols <- intersect(keep_cols, names(raw))
  # select harmonized columns
  out <- dplyr::select(raw, dplyr::all_of(keep_cols))
  # add march year
  out$year_march <- year
  # add source file name
  out$source_file <- basename(path)
  # zero pad state codes to two digits
  out$cve_ent <- stringr::str_pad(stringr::str_trim(out$cve_ent), width = 2, pad = "0")
  # zero pad city codes to two digits
  out$cd <- stringr::str_pad(stringr::str_trim(out$cd), width = 2, pad = "0")
  # uppercase and strip accents from state names when present
  out$nom_ent <- stringr::str_to_upper(stringr::str_squish(out$nom_ent))
  # transliterate state names
  out$nom_ent <- stringi::stri_trans_general(out$nom_ent, "Latin-ASCII")
  # uppercase and strip accents from city names when present
  out$nom_cd <- stringr::str_to_upper(stringr::str_squish(out$nom_cd))
  # transliterate city names
  out$nom_cd <- stringi::stri_trans_general(out$nom_cd, "Latin-ASCII")
  # standardize a common culican label variant early when present
  out$nom_cd <- dplyr::if_else(out$nom_cd == "CULIACAN", "CULIACAN ROSALES", out$nom_cd)
  # return harmonized tibble
  out
}

# read and harmonize every march year
march_list <- purrr::pmap(
  list(march_inventory$path, march_inventory$year),
  function(path, year) {
    # announce the year being read
    message("reading march ensu cb for ", year, " (", basename(path), ")")
    # harmonize this file
    harmonize_ensu_march_cb(path, year)
  }
)

# bind all years
ensu_march_all <- dplyr::bind_rows(march_list)

# build a cd to city name catalog from rows that already have nom_cd
cd_catalog <- ensu_march_all %>%
  # keep rows with both code and name
  dplyr::filter(!is.na(cd), cd != "", !is.na(nom_cd), nom_cd != "") %>%
  # count support for each label by year recency
  dplyr::count(cd, nom_cd, year_march, name = "n") %>%
  # prefer the newest year label for each cd
  dplyr::group_by(cd) %>%
  dplyr::arrange(dplyr::desc(year_march), dplyr::desc(n), .by_group = TRUE) %>%
  dplyr::slice(1) %>%
  dplyr::ungroup() %>%
  # keep catalog columns
  dplyr::select(cd, nom_cd_std = nom_cd)

# also fold in the hand curated goc labels when they exist
cd_catalog <- cd_catalog %>%
  # join curated names
  dplyr::left_join(goc_city_meta %>% dplyr::select(cd, nom_cd_meta = nom_cd), by = "cd") %>%
  # prefer curated goc label when available
  dplyr::mutate(nom_cd_std = dplyr::coalesce(nom_cd_meta, nom_cd_std)) %>%
  # drop helper column
  dplyr::select(cd, nom_cd_std)

# attach standardized city names to all rows by cd
ensu_march_all <- ensu_march_all %>%
  # join catalog
  dplyr::left_join(cd_catalog, by = "cd") %>%
  # fill nom_cd from catalog when missing or blank
  dplyr::mutate(nom_cd = dplyr::if_else(is.na(nom_cd) | nom_cd == "", nom_cd_std, nom_cd)) %>%
  # drop helper
  dplyr::select(-nom_cd_std)

# build a simple state name catalog from known codes for early years without nom_ent
state_catalog <- tibble::tribble(
  ~cve_ent, ~nom_ent_std,
  "02", "BAJA CALIFORNIA",
  "03", "BAJA CALIFORNIA SUR",
  "14", "JALISCO",
  "18", "NAYARIT",
  "25", "SINALOA",
  "26", "SONORA"
)

# fill missing state names for goc states
ensu_march_all <- ensu_march_all %>%
  # join state catalog
  dplyr::left_join(state_catalog, by = "cve_ent") %>%
  # coalesce state names
  dplyr::mutate(nom_ent = dplyr::coalesce(dplyr::na_if(nom_ent, ""), nom_ent_std)) %>%
  # drop helper
  dplyr::select(-nom_ent_std)

# filter to the six gulf adjacent states using codes (robust when names are messy)
ensu_march_goc <- ensu_march_all %>%
  # keep goc state codes
  dplyr::filter(cve_ent %in% goc_state_codes)

# save person level intermediate extract
readr::write_csv(ensu_march_goc, file.path(path_int, "ensu_march_goc_persons.csv"))

# show counts by year after filter
ensu_march_goc %>%
  # count selected adults by march year
  dplyr::count(year_march, name = "n_selected_adults") %>%
  # print counts
  print()
```

# City coverage by March year

Because ENSU added cities over time, coverage is uneven.
`CD` identifies cities; names come from the standardized catalog.

```{r}
# city by year presence for goc states
city_year_coverage <- ensu_march_goc %>%
  # drop rows without a city code
  dplyr::filter(!is.na(cd), cd != "") %>%
  # count respondents
  dplyr::count(year_march, cve_ent, nom_ent, cd, nom_cd, name = "n") %>%
  # join gulf relevance by cd
  dplyr::left_join(goc_city_meta %>% dplyr::select(cd, gulf_relevance), by = "cd") %>%
  # arrange
  dplyr::arrange(nom_ent, cd, year_march)

# save coverage intermediate
readr::write_csv(city_year_coverage, file.path(path_int, "ensu_march_goc_city_year_coverage.csv"))

# wide presence matrix
city_year_wide <- city_year_coverage %>%
  # keep identity fields and counts
  dplyr::select(nom_ent, cd, nom_cd, gulf_relevance, year_march, n) %>%
  # spread years
  tidyr::pivot_wider(names_from = year_march, values_from = n)

# save wide presence to output
readr::write_csv(city_year_wide, file.path(path_out, "ensu_march_goc_city_year_presence.csv"))

# print wide table
print(city_year_wide)
```

# Indicator 1: share who feel living in their city is unsafe

Uses harmonized `bp1_1` (including legacy `P1` for 2014 to 2015) and `fac_sel` (including legacy `FACTOR`).

```{r}
# weighted percent unsafe by city and march year
bp1_1_city_year <- ensu_march_goc %>%
  # parse codes and weights
  dplyr::mutate(
    bp1_1 = as.integer(bp1_1),
    fac_sel = as.numeric(fac_sel)
  ) %>%
  # keep substantive safe or unsafe answers with positive weights
  dplyr::filter(bp1_1 %in% c(1L, 2L), !is.na(fac_sel), fac_sel > 0) %>%
  # group
  dplyr::group_by(year_march, cve_ent, nom_ent, cd, nom_cd) %>%
  # summarize
  dplyr::summarise(
    n_substantive = dplyr::n(),
    wt_total = sum(fac_sel, na.rm = TRUE),
    wt_unsafe = sum(fac_sel[bp1_1 == 2L], na.rm = TRUE),
    pct_unsafe = 100 * wt_unsafe / wt_total,
    .groups = "drop"
  ) %>%
  # join relevance
  dplyr::left_join(goc_city_meta %>% dplyr::select(cd, gulf_relevance), by = "cd") %>%
  # arrange
  dplyr::arrange(nom_ent, cd, year_march)

# save intermediate and output copies
readr::write_csv(bp1_1_city_year, file.path(path_int, "ensu_march_bp1_1_city_year.csv"))
# save output table
readr::write_csv(bp1_1_city_year, file.path(path_out, "ensu_march_pct_unsafe_city_year.csv"))

# preview
print(bp1_1_city_year)
```

# Plot March unsafe shares over years

```{r}
# line plot of percent unsafe
p_unsafe <- ggplot(
  data = bp1_1_city_year,
  mapping = aes(x = year_march, y = pct_unsafe, color = nom_cd, group = cd)
) +
  # lines by city code
  geom_line(linewidth = 0.6, alpha = 0.9) +
  # points by city year
  geom_point(size = 1.8) +
  # facet by state
  facet_wrap(~ nom_ent, scales = "free_y") +
  # labels
  labs(
    title = "ENSU March: share of adults who feel living in their city is unsafe",
    subtitle = "Weighted with FAC_SEL (or legacy FACTOR). 2021 has no March round.",
    x = "March year",
    y = "Percent unsafe",
    color = "City"
  ) +
  # theme
  theme_minimal(base_size = 11) +
  # rotate x labels
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "bottom") +
  # y limits
  scale_y_continuous(limits = c(0, 100)) +
  # show integer years on x
  scale_x_continuous(breaks = sort(unique(bp1_1_city_year$year_march)))

# print plot
print(p_unsafe)

# save figure
ggsave(
  filename = file.path(path_out, "ensu_march_pct_unsafe_by_city_year.png"),
  plot = p_unsafe,
  width = 12,
  height = 8,
  dpi = 150
)
```

# Optional focus: stronger Gulf candidates only

```{r}
# relevance labels closer to gulf side needs
gulf_focus_labels <- c(
  "strong goc candidate",
  "northern sinaloa goc adjacent",
  "southern sonora coast adjacent",
  "pacific bahia de banderas fringe"
)

# filter focus cities
bp1_1_gulf_focus <- bp1_1_city_year %>%
  # keep chosen relevance labels
  dplyr::filter(gulf_relevance %in% gulf_focus_labels)

# save focused table
readr::write_csv(bp1_1_gulf_focus, file.path(path_out, "ensu_march_pct_unsafe_gulf_focus_city_year.csv"))

# plot if rows exist
if (nrow(bp1_1_gulf_focus) > 0) {
  # focused plot
  p_focus <- ggplot(
    data = bp1_1_gulf_focus,
    mapping = aes(x = year_march, y = pct_unsafe, color = nom_cd, group = cd)
  ) +
    # lines
    geom_line(linewidth = 0.8) +
    # points
    geom_point(size = 2) +
    # labels
    labs(
      title = "ENSU March: percent unsafe in gulf focus cities",
      subtitle = "Subset of ENSU cities closer to GoC analysis needs",
      x = "March year",
      y = "Percent unsafe",
      color = "City"
    ) +
    # theme
    theme_minimal(base_size = 12) +
    # y limits
    scale_y_continuous(limits = c(0, 100)) +
    # integer year breaks
    scale_x_continuous(breaks = sort(unique(bp1_1_gulf_focus$year_march)))

  # print
  print(p_focus)

  # save
  ggsave(
    filename = file.path(path_out, "ensu_march_pct_unsafe_gulf_focus.png"),
    plot = p_focus,
    width = 10,
    height = 6,
    dpi = 150
  )
}
```

# Indicator 2: changed habits from fear of crime

Uses harmonized `bp1_5_*` (including legacy `P4_*` for 2014 to 2015).

```{r}
# habit item columns
habit_cols <- c("bp1_5_1", "bp1_5_2", "bp1_5_3", "bp1_5_4", "bp1_5_5")

# long habit table
habit_long <- ensu_march_goc %>%
  # parse weight
  dplyr::mutate(fac_sel = as.numeric(fac_sel)) %>%
  # keep positive weights
  dplyr::filter(!is.na(fac_sel), fac_sel > 0) %>%
  # reshape
  tidyr::pivot_longer(
    cols = dplyr::all_of(habit_cols),
    names_to = "habit_item",
    values_to = "resp"
  ) %>%
  # parse response
  dplyr::mutate(resp = as.integer(resp)) %>%
  # keep substantive yes or no
  dplyr::filter(resp %in% c(1L, 2L))

# weighted percent yes by item
habit_city_year <- habit_long %>%
  # group
  dplyr::group_by(year_march, cve_ent, nom_ent, cd, nom_cd, habit_item) %>%
  # summarize
  dplyr::summarise(
    n_substantive = dplyr::n(),
    wt_total = sum(fac_sel, na.rm = TRUE),
    wt_yes = sum(fac_sel[resp == 1L], na.rm = TRUE),
    pct_yes = 100 * wt_yes / wt_total,
    .groups = "drop"
  ) %>%
  # join relevance
  dplyr::left_join(goc_city_meta %>% dplyr::select(cd, gulf_relevance), by = "cd")

# save habit item rates
readr::write_csv(habit_city_year, file.path(path_int, "ensu_march_bp1_5_item_city_year.csv"))
# save output copy
readr::write_csv(habit_city_year, file.path(path_out, "ensu_march_habit_change_item_city_year.csv"))

# person level any habit changed flag
any_habit <- ensu_march_goc %>%
  # parse
  dplyr::mutate(
    fac_sel = as.numeric(fac_sel),
    dplyr::across(dplyr::all_of(habit_cols), as.integer)
  ) %>%
  # keep weights
  dplyr::filter(!is.na(fac_sel), fac_sel > 0) %>%
  # rowwise any yes among substantive answers
  dplyr::rowwise() %>%
  dplyr::mutate(
    any_yes = {
      # collect habit answers
      vals <- c_across(dplyr::all_of(habit_cols))
      # keep yes no only
      vals <- vals[vals %in% c(1L, 2L)]
      # missing if no substantive habit answers
      if (length(vals) == 0) NA else any(vals == 1L)
    }
  ) %>%
  dplyr::ungroup() %>%
  # drop empty habit rows
  dplyr::filter(!is.na(any_yes))

# weighted percent any habit change
any_habit_city_year <- any_habit %>%
  # group
  dplyr::group_by(year_march, cve_ent, nom_ent, cd, nom_cd) %>%
  # summarize
  dplyr::summarise(
    n_substantive = dplyr::n(),
    wt_total = sum(fac_sel, na.rm = TRUE),
    wt_any_yes = sum(fac_sel[any_yes], na.rm = TRUE),
    pct_any_habit_change = 100 * wt_any_yes / wt_total,
    .groups = "drop"
  ) %>%
  # join relevance
  dplyr::left_join(goc_city_meta %>% dplyr::select(cd, gulf_relevance), by = "cd")

# save composite
readr::write_csv(any_habit_city_year, file.path(path_out, "ensu_march_any_habit_change_city_year.csv"))

# preview
print(any_habit_city_year)
```

# Latest March snapshot

```{r}
# latest march year present in the unsafe series
latest_year <- max(bp1_1_city_year$year_march, na.rm = TRUE)

# snapshot joining unsafe and habit change
latest_snapshot <- bp1_1_city_year %>%
  # latest year only
  dplyr::filter(year_march == latest_year) %>%
  # select key fields
  dplyr::select(year_march, nom_ent, cd, nom_cd, gulf_relevance, n_substantive, pct_unsafe) %>%
  # join habit composite
  dplyr::left_join(
    any_habit_city_year %>%
      dplyr::select(year_march, cd, pct_any_habit_change),
    by = c("year_march", "cd")
  ) %>%
  # arrange
  dplyr::arrange(nom_ent, cd)

# save snapshot
readr::write_csv(latest_snapshot, file.path(path_out, paste0("ensu_march_", latest_year, "_city_snapshot.csv")))

# print snapshot
print(latest_snapshot)
```

# Practical notes

1. March 2021 is intentionally missing from the time series.
2. Years 2014 to 2020 are read from DBF; 2022 onward from CSV.
3. Years 2014 to 2015 use legacy `P1` / `P4_*` / `FACTOR` mappings documented above.
4. City identity is `CD`; names are standardized from the newest available catalog.
5. When copying this folder to the shared server, keep the same `ensu_bd_YYYY_*` layout so recursive discovery continues to work.
