---
title: 'OHI GoC 2025: Safety and Security (ENSU)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
    include-in-header: '../../../src/templates/ohi_hdr.html'
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# Safety and Security Goal: ENSU city perception

# Summary

### Background

This script explores feelings of safety using the National Survey of Urban Public Security (ENSU) for cities in the six states that border or fringe the Gulf of California: Baja California, Baja California Sur, Sonora, Sinaloa, Nayarit, and Jalisco. Following discussions from the OHI GoC Expert Working Group Meeting (June 2025), the aim is to inform a Safety and Security goal with perception indicators that can be assigned to designed urban domains rather than forced onto municipios that the survey never powered.

The analysis focuses on March rounds from **2018 onward**. Earlier March years (2014 to 2017) are still read when present, mainly so older files can borrow city names from newer files, but they are left out of the Gulf of California results because those early years have fewer people and fewer cities, and the city codes change over time. March is treated as a comparable seasonal snapshot so that year to year change is not mixed with within year seasonality from June, September, or December rounds.

March 2021 is included when the person-level files are available (DBF under `ensu_bd_2021_dbf/ensu_bd_marzo_2021_dbf`, including on Mazu at `/home/shares/ohi/OHI_GOC/_raw_data/inegi_crime/d2025/ENSU/ensu_bd_2021_dbf/ensu_bd_marzo_2021_dbf`). INEGI did run March 2021; older project copies sometimes lacked only that microdata package.

*Note: This analysis maintains sensitivity to regional contexts and focuses on constructive data interpretation as it relates to the Gulf's health.*

### Why ENSU for this goal

ENSU is built to produce city level estimates of perceived public safety for named áreas urbanas de interés. That matches the OHI need more closely than either ENVIPE or the Mexico Peace Index (MPI) and Mexico Positive Peace Index (MPPI) for the following reasons:

* ENVIPE is designed for national, state, and a short list of urban areas of interest. Municipal means from ENVIPE microdata are exploratory only.
* MPI fear of violence is already an ENVIPE based state score. MPPI is a state level Positive Peace construct.  Painting those state scores onto Gulf side municipios imports border and Pacific dynamics and cannot isolate GoC localities.
* ENSU asks, by design, whether living in the *named city* feels safe or unsafe, and whether people changed habits from fear of crime.

### Data source and folder layout

Program page: [INEGI ENSU](https://www.inegi.org.mx/programas/ensu/)

Primary table:

* `ENSU_CB_*`: selected person questionnaire (perception and habit change).

Raw data are read from the first existing path among:

1. `/home/shares/ohi/OHI_GOC/_raw_data/inegi_crime/d2025/ENSU` (shared server)
2. the project local folder `ENSU/` (for example `.../SS_security/ENSU`)

Intermediate outputs go to `/home/shares/ohi/OHI_GOC/goal_prep/ss/security_int` when that tree exists, otherwise to `goal_prep/ss/security_int` under the project.
Final outputs go to `/home/shares/ohi/OHI_GOC/goal_prep/ss/security_output` or `goal_prep/ss/security_output` locally.

OHI GoC regions come from a CSV that lists each municipality and which Gulf region it belongs to (`municipality_to_rgnid_update.csv` on the server under `/home/shares/ohi/OHI_GOC/spatial/municipality_to_rgnid/`, or the copy in this project folder). Each ENSU city is matched to that list through the main municipality the city sits in (for example Los Mochis uses Ahome; Ciudad Obregón uses Cajeme).

### Year to year file and schema changes handled here

| Years | Format | Notes handled by the loader |
|---|---|---|
| 2014 to 2020 | DBF (`.dbf` / `.DBF`) | Read with `foreign::read.dbf` |
| 2021 March | DBF | `ENSU_CB_0321.dbf` under `ensu_bd_2021_dbf/ensu_bd_marzo_2021_dbf` (modern `BP1_*` / `FAC_SEL` / `NOM_CD`) |
| 2022 to present | CSV | Read with `readr::read_csv` |
| 2014 to 2015 | Older question names | `P1` becomes city safety; `P4_1` to `P4_5` become habit change; `FACTOR` becomes the person weight |
| 2014 to 2015 | Older city codes `CD` | Different numbering than 2016 onward; converted to the later codes using state + city code, checked against municipality |
| 2016 onward | Modern `BP1_*` names | `BP1_1` and `BP1_5_*` with `FAC_SEL` |
| 2016 to 2018 | Guadalajara metro as `CD` 24 | Converted to later code `CD` 59 so it matches the city list used here |
| through 2016 | `ENT` / `MUN` | Renamed to `CVE_ENT` / `CVE_MUN` |
| 2017 onward | `CVE_ENT`, `NOM_ENT` | State names usually present |
| through 2020 | city code `CD` only | City names filled from newer years that already include `NOM_CD` |
| 2022 onward | `NOM_CD` present | Preferred city labels, still keyed by `CD` |
| file names | mixed | Accepts `ENSU_CB_03YY` and `ENSU_CB03YY` |

After older city codes are converted, `CD` is the stable city ID across years. City display names can shift slightly (for example `CULIACAN` versus `CULIACAN ROSALES`), so names are standardized from the newest files when possible. Do not attach city names using `CD` alone on unconverted 2014 to 2015 rows: those older codes mean different places (for example Jalisco `CD` 02 is Guadalajara, not Mexicali).

### Methodology notes

ENSU is a probabilistic urban household survey. Representativeness is for each named urban area (city) in the sample, not for every municipality.

Person weight: modern `FAC_SEL` (older files: `FACTOR`) says how many adults in that city one respondent stands for. All percents below use these weights among people who gave a clear answer.

#### Perception of safety (`pct_unsafe`) — city overall, not by place

**Question used:** `BP1_1` (older files: `P1`).

INEGI wording (section 1.1): *En términos de delincuencia, ¿considera que vivir actualmente en (CIUDAD), es…*

| Code | Meaning |
|---|---|
| `1` | seguro (safe) |
| `2` | inseguro (unsafe) |
| `9` | no sabe / no responde (dropped from the percent) |

This is one overall judgment about **living in that named city**. It is **not** built by combining answers about home, work, streets, car, and so on.

**How the percent is calculated:** keep people with answers `1` or `2` and a positive weight; then

\[
\%\ \text{unsafe} = 100 \times \frac{\text{sum of weights for people who said inseguro (2)}}{\text{sum of weights for people who said seguro or inseguro}}
\]

The same calculation is done by city × March year and again by OHI GoC region × March year (aggregating people whose city sits in that region).

#### Place-specific safety (`BP1_2_*`) — asked by ENSU, **not used in this script**

ENSU also asks (section 1.2): *En términos de delincuencia, dígame si en (LUGAR) se siente seguro(a) o inseguro(a)* for each place below.
Answers are seguro / inseguro / no aplica / no sabe.

| Code | Place |
|---|---|
| `BP1_2_01` | su casa |
| `BP1_2_02` | su trabajo |
| `BP1_2_03` | las calles que habitualmente usa |
| `BP1_2_04` | la escuela |
| `BP1_2_05` | el mercado |
| `BP1_2_06` | el centro comercial |
| `BP1_2_07` | el banco |
| `BP1_2_08` | el cajero automático en la vía pública |
| `BP1_2_09` | el transporte público |
| `BP1_2_10` | el automóvil |
| `BP1_2_11` | la carretera |
| `BP1_2_12` | el parque recreativo o centro recreativo |

**This analysis does not use `BP1_2_*` because we are interested in the general perception of safety in the city.** 
`pct_unsafe` is **not** “felt unsafe in any of these places.” It only uses the single city question `BP1_1`.

#### Habit changes from fear of crime (`BP1_5_*`)

**Question stem (section 1.5):** *En este mismo periodo de tres meses, por temor a sufrir algún delito (robo, asalto, secuestro, entre otros), ¿usted cambió sus hábitos respecto a …*

(Older files use `P4_1` to `P4_5` for the same items.)

| Code | Habit the person is asked about |
|---|---|
| `BP1_5_1` | llevar cosas de valor (joyas, dinero o tarjetas de crédito) |
| `BP1_5_2` | caminar por los alrededores de su vivienda, pasadas las ocho de la noche |
| `BP1_5_3` | visitar a parientes o amigos(as) |
| `BP1_5_4` | permitir que los (las) menores de edad que viven en el hogar salgan solos(as) |
| `BP1_5_5` | otro (any other habit change) |

| Code | Meaning |
|---|---|
| `1` | sí (yes, changed this habit) |
| `2` | no |
| `3` | no aplica (dropped from item percents) |
| `9` | no sabe / no responde (dropped from item percents) |

**Two habit products are saved:**

1. **Per item** (`pct_yes` in the habit-item tables): for each of the five habits separately, the weighted share who said sí among clear sí/no answers.
2. **Any habit change** (`pct_any_habit_change`): for each person, if **at least one** of the five items is sí (among their clear sí/no answers), that person counts as having changed habits. Then take the weighted share of such people.

So the combined measure is the share of adults who, in the last three months, changed **at least one** listed habit out of fear of crime.
It is **not** the average of the five item percents, and it does **not** use the place-safety questions.

Cities that enter ENSU later will show missing early years; that is coverage change, not a zero rate.

# Setup

```{r}
# install librarian if it is missing
if (!require(librarian)) {
  # install librarian from cran
  install.packages("librarian")
  # attach librarian after install
  library(librarian)
  # end of librarian install branch
}

# load project packages through librarian
librarian::shelf(
  tidyverse, # data wrangling and plotting utilities
  here, # project relative path helpers
  ggplot2, # grammar of graphics plotting
  readr, # fast csv reading and writing
  stringr, # string cleaning helpers
  stringi, # accent stripping for city name joins
  lubridate, # date helpers if needed later
  janitor, # clean_names for ensu columns
  scales, # axis formatting helpers
  purrr, # map over march files by year
  foreign # read dbf microdata for 2020 and earlier
)

# candidate raw ensu roots (server first, then local project)
path_raw_candidates <- c(
  "/home/shares/ohi/OHI_GOC/_raw_data/inegi_crime/d2025/ENSU", # shared ohi server path
  here::here("ENSU") # local project ensu folder
)

# keep only candidate roots that currently exist
path_raw_existing <- path_raw_candidates[dir.exists(path_raw_candidates)]

# stop if no ensu root can be found
if (length(path_raw_existing) == 0) {
  # explain both expected locations
  stop("no ensu raw folder found. expected server path or local project ENSU/ folder.")
}

# use the first existing raw root
path_raw_ensu <- path_raw_existing[[1]]

# candidate intermediate folders
path_int_candidates <- c(
  "/home/shares/ohi/OHI_GOC/goal_prep/ss/security_int", # shared server intermediate path
  here::here("goal_prep", "ss", "security_int") # local intermediate folder if server path is missing
)

# candidate output folders
path_out_candidates <- c(
  "/home/shares/ohi/OHI_GOC/goal_prep/ss/security_output", # shared server output path
  here::here("goal_prep", "ss", "security_output") # local output folder if server path is missing
)

# default to local intermediate path
path_int <- path_int_candidates[[2]]

# prefer server intermediate path when the ohi shares tree exists
if (dir.exists("/home/shares/ohi/OHI_GOC") || dir.exists(path_int_candidates[[1]])) {
  # use server intermediate path
  path_int <- path_int_candidates[[1]]
}

# default to local output path
path_out <- path_out_candidates[[2]]

# prefer server output path when the ohi shares tree exists
if (dir.exists("/home/shares/ohi/OHI_GOC") || dir.exists(path_out_candidates[[1]])) {
  # use server output path
  path_out <- path_out_candidates[[1]]
}

# create the intermediate folder if it does not exist
dir.create(path_int, recursive = TRUE, showWarnings = FALSE)

# create the output folder if it does not exist
dir.create(path_out, recursive = TRUE, showWarnings = FALSE)

# print which paths are in use
message("using raw ensu path: ", path_raw_ensu)
# print intermediate path
message("using intermediate path: ", path_int)
# print output path
message("using output path: ", path_out)

# CSV that says which municipality belongs to which OHI GoC region (server first, then local copy)
path_mun_rgn_candidates <- c(
  "/home/shares/ohi/OHI_GOC/spatial/municipality_to_rgnid/municipality_to_rgnid_update.csv",
  here::here("municipality_to_rgnid_update.csv")
)
# keep existing candidates
path_mun_rgn_existing <- path_mun_rgn_candidates[file.exists(path_mun_rgn_candidates)]
# stop if the municipality-to-region file is missing
if (length(path_mun_rgn_existing) == 0) {
  stop("municipality_to_rgnid_update.csv not found on server or in project root.")
}
# use first existing path
path_mun_rgn <- path_mun_rgn_existing[[1]]
message("using municipality-to-region file: ", path_mun_rgn)

# readable names for each OHI GoC region
rgn_labels <- c(
  "Region_1" = "Western Upper Gulf",
  "Region_2" = "Western Midriff Islands",
  "Region_3" = "Western Central Gulf",
  "Region_4" = "Western Southern Gulf",
  "Region_5" = "Eastern Upper Gulf",
  "Region_6" = "Eastern Midriff Islands",
  "Region_7" = "Southern Sonora Coast",
  "Region_8" = "Sinaloa Coast",
  "Region_9" = "Southeastern Gulf"
)

# colors for each OHI GoC region
rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

# ordered region ids
region_levels <- paste0("Region_", 1:9)

# same colors, but named by the readable region labels for plots
rgn_label_colors <- setNames(unname(rgn_colors[region_levels]), unname(rgn_labels[region_levels]))

# state codes for the six gulf adjacent states
goc_state_codes <- c(
  "02", # baja california
  "03", # baja california sur
  "14", # jalisco
  "18", # nayarit
  "25", # sinaloa
  "26" # sonora
)

# ENSU cities we care about, with the municipality code used to look up the region
goc_city_meta <- tibble::tribble(
  ~cve_ent, ~cd, ~cve_mun, ~nom_cd, ~gulf_relevance, # column headers
  "03", "04", "003", "LA PAZ", "goc", # bcs gulf side city
  "03", "55", "008", "LOS CABOS", "goc", # bcs pacific side city
  "25", "42", "001", "LOS MOCHIS", "goc", # ahome urban area
  "26", "94", "018", "CIUDAD OBREGON", "goc", # cajeme urban area
  "26", "43", "030", "HERMOSILLO", "goc", # inland sonora capital
  "25", "40", "006", "CULIACAN ROSALES", "goc", # inland sinaloa capital
  "25", "41", "012", "MAZATLAN", "goc", # pacific sinaloa city
  "14", "25", "067", "PUERTO VALLARTA", "goc", # jalisco pacific fringe
  "18", "33", "017", "TEPIC", "goc", # nayarit capital
  # "14", "59", "039", "GUADALAJARA", "inland jalisco metro", # jalisco core city
  # "14", "63", "120", "ZAPOPAN", "inland jalisco metro", # guadalajara metro member
  # "14", "60", "101", "TONALA", "inland jalisco metro", # guadalajara metro member
  # "14", "62", "098", "SAN PEDRO TLAQUEPAQUE", "inland jalisco metro", # guadalajara metro member
  # "14", "61", "097", "TLAJOMULCO DE ZUNIGA", "inland jalisco metro", # guadalajara metro member
  # "02", "03", "004", "TIJUANA", "border pacific not goc", # border pacific city
  "02", "02", "002", "MEXICALI", "border not goc", # border city
  "26", "44", "043", "NOGALES", "border sonora not goc" # sonora border city; not in goc region table
)

# read the municipality-to-region file
mun_rgn <- readr::read_csv(path_mun_rgn, show_col_types = FALSE, locale = readr::locale(encoding = "UTF-8")) %>%
  # clean column names
  janitor::clean_names() %>%
  # make state and municipality codes the same width as in ENSU
  dplyr::mutate(
    cve_ent = stringr::str_pad(as.character(e03), width = 2, pad = "0"),
    cve_mun = stringr::str_pad(as.character(e04), width = 3, pad = "0"),
    # treat the text "NA" as missing region
    rgn_id = dplyr::na_if(as.character(rgn_id), "NA")
  ) %>%
  # keep the columns needed to match cities to regions
  dplyr::select(cve_ent, cve_mun, rgn_id, both_pacific_gulf, directly_coastal) %>%
  # one row per municipality
  dplyr::distinct(cve_ent, cve_mun, .keep_all = TRUE)

# add the region id to each ENSU city using its municipality
goc_city_meta <- goc_city_meta %>%
  dplyr::left_join(mun_rgn %>% dplyr::select(cve_ent, cve_mun, rgn_id), by = c("cve_ent", "cve_mun")) %>%
  # keep regions in order 1 to 9 for plots
  dplyr::mutate(
    rgn_id = factor(rgn_id, levels = region_levels),
    rgn_label = unname(rgn_labels[as.character(rgn_id)]),
    rgn_label = factor(rgn_label, levels = unname(rgn_labels[region_levels]))
  )

# show which city was placed in which region
print(
  goc_city_meta %>%
    dplyr::select(cve_ent, cd, cve_mun, nom_cd, gulf_relevance, rgn_id, rgn_label)
)

# 2014 to 2015 used different city codes than later years.
# Checked against municipality codes (for example Jalisco CD 02 = Guadalajara, not Mexicali).
cd_legacy_2014_2015 <- tibble::tribble(
  ~cve_ent, ~cd_legacy, ~cd, ~nom_cd_legacy,
  "02", "21", "03", "TIJUANA", # baja california mun 004
  "03", "40", "04", "LA PAZ", # bcs mun 003
  "14", "02", "59", "GUADALAJARA", # jalisco metro munis; modern split starts 2019
  "18", "27", "33", "TEPIC", # nayarit mun 017
  "25", "24", "40", "CULIACAN ROSALES", # sinaloa mun 006
  "26", "25", "43", "HERMOSILLO" # sonora mun 030
)
```

This setup chunk finds the data folders, loads packages, sets region names and colors, lists the ENSU cities and their municipalities, and converts the old 2014 to 2015 city codes to the later ones.

### Explore March microdata (CSV and DBF)

March files are discovered recursively under the ENSU root (server or local `ENSU/`).
Accepted names include `ENSU_CB_0325.csv`, `ENSU_CB_0321.dbf`, and older spellings such as `ENSU_CB0316.dbf`.
March 2021 is expected as DBF when present (`ensu_bd_2021_dbf/ensu_bd_marzo_2021_dbf`).

```{r}
# find march cb files in csv or dbf form under the ensu root
march_cb_files <- list.files(
  path = path_raw_ensu,
  pattern = "^ENSU_CB_?03[0-9]{2}\\.(csv|dbf)$",
  full.names = TRUE,
  recursive = TRUE,
  ignore.case = TRUE
)

# stop if nothing was found
if (length(march_cb_files) == 0) {
  # provide a clear recovery message
  stop("no march ensu_cb csv/dbf files found under ", path_raw_ensu)
}

# build an inventory tibble from discovered files
march_inventory <- tibble::tibble(
  path = march_cb_files
) %>%
  # parse yy from either ensu_cb_03yy or ensu_cb03yy
  mutate(
    yy = stringr::str_match(basename(path), "(?i)ENSU_CB_?03([0-9]{2})")[, 2],
    year = dplyr::if_else(as.integer(yy) >= 90L, 1900L + as.integer(yy), 2000L + as.integer(yy)),
    file = basename(path),
    format = tolower(tools::file_ext(path))
  ) %>%
  # prefer csv over dbf if both somehow exist for one year
  arrange(year, dplyr::desc(format == "csv"), path) %>%
  # keep one file per year
  distinct(year, .keep_all = TRUE) %>%
  # note march 2021 source when present
  mutate(
    note = dplyr::case_when(
      year == 2021L & format == "dbf" ~ "march 2021 from dbf package",
      year == 2021L ~ "march 2021",
      TRUE ~ NA_character_
    )
  )

# message if march 2021 is still missing from this ensu root
if (!(2021L %in% march_inventory$year)) {
  message(
    "note: no march 2021 ensu_cb file under ", path_raw_ensu,
    ". expected at ensu_bd_2021_dbf/ensu_bd_marzo_2021_dbf/ENSU_CB_0321.dbf"
  )
}

# print inventory
print(march_inventory)

# save inventory
readr::write_csv(march_inventory, file.path(path_int, "ensu_march_file_inventory.csv"))
```

## Load and clean March person tables so all years share the same columns

The loader:

1. Reads CSV or DBF
2. Renames older 2014 to 2015 question names to the later `BP1_*` / `FAC_SEL` names
3. Renames `ENT` / `MUN` to `CVE_ENT` / `CVE_MUN`
4. Adds leading zeros so codes line up (for example `2` becomes `02`)
5. Converts 2014 to 2015 city codes (and 2016 to 2018 Guadalajara `CD` 24) to the later city codes
6. Fills missing city names from years that already have names

```{r}
# helper: read one ensu table from csv or dbf as character columns
read_ensu_file <- function(path) {
  # branch on file extension
  ext <- tolower(tools::file_ext(path))
  # read csv microdata
  if (ext == "csv") {
    # read all columns as character
    out <- readr::read_csv(path, col_types = readr::cols(.default = "c"), show_col_types = FALSE)
    # return csv tibble
    return(out)
  }
  # read dbf microdata for 2020 and earlier
  if (ext == "dbf") {
    # foreign::read.dbf returns a data.frame
    out <- foreign::read.dbf(path, as.is = TRUE)
    # convert every column to character for binding safety
    out[] <- lapply(out, function(col) as.character(col))
    # return dbf table as tibble
    return(tibble::as_tibble(out))
  }
  # unsupported extension
  stop("unsupported ensu file type: ", path)
}

# helper: clean one march file so columns match later years
harmonize_ensu_march_cb <- function(path, year) {
  # read raw file
  raw <- read_ensu_file(path)
  # clean names to snake case
  raw <- janitor::clean_names(raw)
  # rename older 2014 to 2015 safety question p1 to bp1_1 when needed
  if (!("bp1_1" %in% names(raw)) && ("p1" %in% names(raw))) {
    # copy older p1 into bp1_1
    raw$bp1_1 <- raw$p1
  }
  # rename older habit questions p4_1 to p4_5 to bp1_5_* when needed
  for (i in 1:5) {
    # build later and older names for this habit item
    modern <- paste0("bp1_5_", i)
    # build older name
    legacy <- paste0("p4_", i)
    # copy older habit item when later name is missing
    if (!(modern %in% names(raw)) && (legacy %in% names(raw))) {
      # put older answers under the later column name
      raw[[modern]] <- raw[[legacy]]
    }
  }
  # rename older weight column factor to fac_sel when needed
  if (!("fac_sel" %in% names(raw)) && ("factor" %in% names(raw))) {
    # copy factor into fac_sel
    raw$fac_sel <- raw$factor
  }
  # rename older state code ent to cve_ent
  if ("ent" %in% names(raw) && !("cve_ent" %in% names(raw))) {
    # rename ent
    raw <- dplyr::rename(raw, cve_ent = ent)
  }
  # rename older municipality code mun to cve_mun
  if ("mun" %in% names(raw) && !("cve_mun" %in% names(raw))) {
    # rename mun
    raw <- dplyr::rename(raw, cve_mun = mun)
  }
  # ensure cve_ent exists even if empty
  if (!("cve_ent" %in% names(raw))) {
    # create empty state code column
    raw$cve_ent <- NA_character_
  }
  # ensure cd exists
  if (!("cd" %in% names(raw))) {
    # create empty city code column
    raw$cd <- NA_character_
  }
  # ensure nom_ent exists
  if (!("nom_ent" %in% names(raw))) {
    # create empty state name column
    raw$nom_ent <- NA_character_
  }
  # ensure nom_cd exists
  if (!("nom_cd" %in% names(raw))) {
    # create empty city name column
    raw$nom_cd <- NA_character_
  }
  # ensure fac_sel exists
  if (!("fac_sel" %in% names(raw))) {
    # create empty weight column
    raw$fac_sel <- NA_character_
  }
  # ensure bp1_1 exists
  if (!("bp1_1" %in% names(raw))) {
    # create empty safety column
    raw$bp1_1 <- NA_character_
  }
  # ensure habit columns exist
  for (i in 1:5) {
    # modern habit column name
    modern <- paste0("bp1_5_", i)
    # add empty habit column when absent
    if (!(modern %in% names(raw))) {
      # create empty habit column
      raw[[modern]] <- NA_character_
    }
  }
  # columns to retain for the multi year bind
  keep_cols <- c(
    "cve_ent", "nom_ent", "cve_mun", "nom_mun",
    "cd", "nom_cd", "per",
    "sexo", "edad",
    "bp1_1",
    "bp1_5_1", "bp1_5_2", "bp1_5_3", "bp1_5_4", "bp1_5_5",
    "fac_sel", "fac_viv",
    "upm", "upm_dis", "est_dis", "dominio"
  )
  # keep only columns that exist
  keep_cols <- intersect(keep_cols, names(raw))
  # select cleaned columns
  out <- dplyr::select(raw, dplyr::all_of(keep_cols))
  # add march year
  out$year_march <- year
  # add source file name
  out$source_file <- basename(path)
  # add a leading zero so state codes are two digits
  out$cve_ent <- stringr::str_pad(stringr::str_trim(out$cve_ent), width = 2, pad = "0")
  # add a leading zero so city codes are two digits
  out$cd <- stringr::str_pad(stringr::str_trim(out$cd), width = 2, pad = "0")
  # uppercase and strip accents from state names when present
  out$nom_ent <- stringr::str_to_upper(stringr::str_squish(out$nom_ent))
  # transliterate state names
  out$nom_ent <- stringi::stri_trans_general(out$nom_ent, "Latin-ASCII")
  # uppercase and strip accents from city names when present
  out$nom_cd <- stringr::str_to_upper(stringr::str_squish(out$nom_cd))
  # transliterate city names
  out$nom_cd <- stringi::stri_trans_general(out$nom_cd, "Latin-ASCII")
  # standardize a common culican label variant early when present
  out$nom_cd <- dplyr::if_else(out$nom_cd == "CULIACAN", "CULIACAN ROSALES", out$nom_cd)
  # return cleaned table
  out
}

# read and clean every march year
march_list <- purrr::pmap(
  list(march_inventory$path, march_inventory$year),
  function(path, year) {
    # announce the year being read
    message("reading march ensu cb for ", year, " (", basename(path), ")")
    # clean this file
    harmonize_ensu_march_cb(path, year)
  }
)

# bind all years
ensu_march_all <- dplyr::bind_rows(march_list)

# convert 2014 to 2015 city codes to the later codes before attaching city names.
# if we skip this step, the same code can get the wrong city name
# (for example Jalisco CD 02 becomes Mexicali; Sonora CD 25 becomes Puerto Vallarta).
ensu_march_all <- ensu_march_all %>%
  # keep the original code for checking later if needed
  dplyr::mutate(cd_raw = cd) %>%
  # match the old-code lookup table for early years
  dplyr::left_join(
    cd_legacy_2014_2015 %>% dplyr::select(cve_ent, cd_legacy, cd_modern = cd),
    by = c("cve_ent", "cd_raw" = "cd_legacy")
  ) %>%
  # replace old codes when a later match exists
  dplyr::mutate(
    cd = dplyr::if_else(
      year_march %in% c(2014L, 2015L) & !is.na(cd_modern),
      cd_modern,
      cd
    ),
    # 2016 to 2018 used CD 24 for Guadalajara metro; use later CD 59
    cd = dplyr::if_else(cve_ent == "14" & cd == "24", "59", cd)
  ) %>%
  # drop helper
  dplyr::select(-cd_modern)

# build a simple city-code to city-name lookup from rows that already have names
cd_catalog <- ensu_march_all %>%
  # keep rows with both code and name
  dplyr::filter(!is.na(cd), cd != "", !is.na(nom_cd), nom_cd != "") %>%
  # count how often each name appears by year
  dplyr::count(cd, nom_cd, year_march, name = "n") %>%
  # prefer the newest year label for each city code
  dplyr::group_by(cd) %>%
  dplyr::arrange(dplyr::desc(year_march), dplyr::desc(n), .by_group = TRUE) %>%
  dplyr::slice(1) %>%
  dplyr::ungroup() %>%
  # keep lookup columns
  dplyr::select(cd, nom_cd_std = nom_cd)

# prefer the hand-picked GoC city names when available
cd_catalog <- cd_catalog %>%
  dplyr::left_join(goc_city_meta %>% dplyr::select(cd, nom_cd_meta = nom_cd), by = "cd") %>%
  dplyr::mutate(nom_cd_std = dplyr::coalesce(nom_cd_meta, nom_cd_std)) %>%
  dplyr::select(cd, nom_cd_std)

# fill blank city names using that lookup
ensu_march_all <- ensu_march_all %>%
  dplyr::left_join(cd_catalog, by = "cd") %>%
  dplyr::mutate(nom_cd = dplyr::if_else(is.na(nom_cd) | nom_cd == "", nom_cd_std, nom_cd)) %>%
  dplyr::select(-nom_cd_std)

# simple state-name lookup for early years that lack nom_ent
state_catalog <- tibble::tribble(
  ~cve_ent, ~nom_ent_std,
  "02", "BAJA CALIFORNIA",
  "03", "BAJA CALIFORNIA SUR",
  "14", "JALISCO",
  "18", "NAYARIT",
  "25", "SINALOA",
  "26", "SONORA"
)

# fill missing state names for goc states
ensu_march_all <- ensu_march_all %>%
  dplyr::left_join(state_catalog, by = "cve_ent") %>%
  dplyr::mutate(nom_ent = dplyr::coalesce(dplyr::na_if(nom_ent, ""), nom_ent_std)) %>%
  dplyr::select(-nom_ent_std)

# keep the six gulf-adjacent states and years 2018 onward
ensu_march_goc <- ensu_march_all %>%
  dplyr::filter(cve_ent %in% goc_state_codes) %>%
  # drop early years with few people / changing city codes
  dplyr::filter(year_march >= 2018L)

# save person level intermediate extract
readr::write_csv(ensu_march_goc, file.path(path_int, "ensu_march_goc_persons.csv"))

# show counts by year after filter
ensu_march_goc %>%
  # count selected adults by march year
  dplyr::count(year_march, name = "n_selected_adults") %>%
  # print counts
  print()
```

### City coverage by March year

ENSU added cities over time, so not every city appears in every year.
After older codes are converted, `CD` identifies cities; names come from newer files when older ones are blank.

```{r}
# city by year presence for goc states
city_year_coverage <- ensu_march_goc %>%
  # drop rows without a city code
  dplyr::filter(!is.na(cd), cd != "") %>%
  # count respondents
  dplyr::count(year_march, cve_ent, nom_ent, cd, nom_cd, name = "n") %>%
  # join gulf relevance by state and city code
  dplyr::left_join(
    goc_city_meta %>% dplyr::select(cve_ent, cd, gulf_relevance, rgn_id, rgn_label),
    by = c("cve_ent", "cd")
  ) %>%
  # arrange
  dplyr::arrange(nom_ent, cd, year_march)

# save coverage intermediate
readr::write_csv(city_year_coverage, file.path(path_int, "ensu_march_goc_city_year_coverage.csv"))

# wide presence matrix
city_year_wide <- city_year_coverage %>%
  # keep identity fields and counts
  dplyr::select(nom_ent, cd, nom_cd, gulf_relevance, rgn_id, rgn_label, year_march, n) %>%
  # spread years
  tidyr::pivot_wider(names_from = year_march, values_from = n)

# save wide presence to output
readr::write_csv(city_year_wide, file.path(path_out, "ensu_march_goc_city_year_presence.csv"))

# print wide table
print(city_year_wide)
```

# Indicator 1: share who feel living in their city is unsafe

Uses only `BP1_1` (older: `P1`): *En términos de delincuencia, ¿considera que vivir actualmente en (CIUDAD), es seguro o inseguro?*

This is **not** an aggregate of feeling unsafe at home, work, in the car, and so on (`BP1_2_*` is left unused).
Person weights are `FAC_SEL` (older: `FACTOR`).
People who answered “no sabe / no responde” (`9`) are left out of the percent.

```{r}
# weighted percent unsafe by city and march year
bp1_1_city_year <- ensu_march_goc %>%
  # parse codes and weights
  dplyr::mutate(
    bp1_1 = as.integer(bp1_1),
    fac_sel = as.numeric(fac_sel)
  ) %>%
  # keep clear safe/unsafe answers with positive weights
  dplyr::filter(bp1_1 %in% c(1L, 2L), !is.na(fac_sel), fac_sel > 0) %>%
  # group
  dplyr::group_by(year_march, cve_ent, nom_ent, cd, nom_cd) %>%
  # summarize
  dplyr::summarise(
    n_substantive = dplyr::n(),
    wt_total = sum(fac_sel, na.rm = TRUE),
    wt_unsafe = sum(fac_sel[bp1_1 == 2L], na.rm = TRUE),
    pct_unsafe = 100 * wt_unsafe / wt_total,
    .groups = "drop"
  ) %>%
  # add gulf tag and region
  dplyr::left_join(
    goc_city_meta %>% dplyr::select(cve_ent, cd, gulf_relevance, rgn_id, rgn_label),
    by = c("cve_ent", "cd")
  ) %>%
  # arrange
  dplyr::arrange(nom_ent, cd, year_march)

# save intermediate and output copies
readr::write_csv(bp1_1_city_year, file.path(path_int, "ensu_march_bp1_1_city_year.csv"))
# save output table
readr::write_csv(bp1_1_city_year, file.path(path_out, "ensu_march_pct_unsafe_city_year.csv"))

# preview
print(bp1_1_city_year)
```

### Plot March unsafe shares over years

Two views of the same city-year series: one colored by city (faceted by state), and one colored by OHI GoC region.

```{r}
# city-colored plot: one line per city, panels by state
p_unsafe <- ggplot(
  data = bp1_1_city_year,
  mapping = aes(x = year_march, y = pct_unsafe, color = nom_cd, group = cd)
) +
  # lines by city code
  geom_line(linewidth = 0.6, alpha = 0.9) +
  # points by city year
  geom_point(size = 1.8) +
  # one panel per state
  facet_wrap(~ nom_ent, scales = "free_y") +
  # labels
  labs(
    title = "ENSU (March): pct of adults who feel living in their city is unsafe",
    subtitle = "Weighted with FAC_SEL, which describes how many adults in that city that one respondent stands for",
    x = "Year",
    y = "Percent unsafe",
    color = "City"
  ) +
  # theme
  theme_minimal(base_size = 11) +
  # rotate x labels
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "bottom") +
  # y limits
  scale_y_continuous(limits = c(0, 100)) +
  # show integer years on x
  scale_x_continuous(breaks = sort(unique(bp1_1_city_year$year_march)))

# print city plot
print(p_unsafe)

# save city plot
ggsave(
  filename = file.path(path_out, "ensu_march_pct_unsafe_by_city_year.png"),
  plot = p_unsafe,
  width = 12,
  height = 8,
  dpi = 150
)

# same data, colored by OHI GoC region
p_unsafe_rgn_color <- ggplot(
  data = bp1_1_city_year,
  mapping = aes(
    x = year_march,
    y = pct_unsafe,
    color = dplyr::coalesce(as.character(rgn_label), "Unassigned"),
    group = cd
  )
) +
  geom_line(linewidth = 0.6, alpha = 0.9) +
  geom_point(size = 1.8) +
  facet_wrap(~ nom_ent, scales = "free_y") +
  labs(
    title = "ENSU (March): pct of adults who feel living in their city is unsafe",
    subtitle = "Same city series as above, colored by OHI GoC region",
    x = "Year",
    y = "Percent unsafe",
    color = "Region"
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "bottom") +
  scale_y_continuous(limits = c(0, 100)) +
  scale_x_continuous(breaks = sort(unique(bp1_1_city_year$year_march))) +
  scale_color_manual(
    values = c(rgn_label_colors, "Unassigned" = "grey50"),
    breaks = c(unname(rgn_labels[region_levels]), "Unassigned")
  )

print(p_unsafe_rgn_color)

ggsave(
  filename = file.path(path_out, "ensu_march_pct_unsafe_by_city_year_rgn_color.png"),
  plot = p_unsafe_rgn_color,
  width = 12,
  height = 8,
  dpi = 150
)
```

### GoC cities only (excludes border comparison cities)

```{r}
# keep cities tagged goc (excludes mexicali / nogales)
bp1_1_gulf_focus <- bp1_1_city_year %>%
  dplyr::filter(gulf_relevance == "goc")

# save focused table
readr::write_csv(bp1_1_gulf_focus, file.path(path_out, "ensu_march_pct_unsafe_gulf_focus_city_year.csv"))

# plot if rows exist
if (nrow(bp1_1_gulf_focus) > 0) {
  # color by region; line style separates cities that share a region
  p_focus <- ggplot(
    data = bp1_1_gulf_focus,
    mapping = aes(
      x = year_march,
      y = pct_unsafe,
      color = rgn_label,
      linetype = nom_cd,
      group = cd
    )
  ) +
    # lines
    geom_line(linewidth = 0.8) +
    # points
    geom_point(size = 2) +
    # labels
    labs(
      title = "ENSU March: percent unsafe in gulf focus cities",
      subtitle = "GoC-tagged ENSU cities colored by OHI GoC region",
      x = "March year",
      y = "Percent unsafe",
      color = "Region",
      linetype = "City"
    ) +
    # theme
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom") +
    # y limits
    scale_y_continuous(limits = c(0, 100)) +
    # integer year breaks
    scale_x_continuous(breaks = sort(unique(bp1_1_gulf_focus$year_march))) +
    # region colors
    scale_color_manual(values = rgn_label_colors, drop = FALSE)

  # print
  print(p_focus)

  # save
  ggsave(
    filename = file.path(path_out, "ensu_march_pct_unsafe_gulf_focus.png"),
    plot = p_focus,
    width = 10,
    height = 6,
    dpi = 150
  )

  # same data, one panel per region
  p_focus_rgn <- ggplot(
    data = bp1_1_gulf_focus %>% dplyr::filter(!is.na(rgn_id)),
    mapping = aes(x = year_march, y = pct_unsafe, color = rgn_label, group = cd)
  ) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 2) +
    facet_wrap(~ rgn_label, scales = "free_y") +
    labs(
      title = "ENSU March: percent unsafe by OHI GoC region",
      subtitle = "Each line is an ENSU city; color matches that region's color",
      x = "March year",
      y = "Percent unsafe",
      color = "Region"
    ) +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
    scale_y_continuous(limits = c(0, 100)) +
    scale_x_continuous(breaks = sort(unique(bp1_1_gulf_focus$year_march))) +
    scale_color_manual(values = rgn_label_colors, drop = FALSE)

  print(p_focus_rgn)

  ggsave(
    filename = file.path(path_out, "ensu_march_pct_unsafe_by_rgn.png"),
    plot = p_focus_rgn,
    width = 11,
    height = 7,
    dpi = 150
  )
}
```

# Indicator 2: changed habits from fear of crime

Uses `BP1_5_1` to `BP1_5_5` (older: `P4_1` to `P4_5`).
Stem: in the last three months, from fear of crime, did you change habits regarding…

1. `BP1_5_1` — carrying valuables (jewelry, money, credit cards)
2. `BP1_5_2` — walking near home after 8 pm
3. `BP1_5_3` — visiting relatives or friends
4. `BP1_5_4` — allowing minors in the household to go out alone
5. `BP1_5_5` — other habit change

For each item, `pct_yes` is the weighted share who said sí among clear sí/no answers.
`pct_any_habit_change` counts a person if **any** of those five is sí, then takes the weighted share of such people.
“No aplica” (`3`) and “no sabe” (`9`) are not treated as yes or no for these percents.

```{r}
# habit item columns (same order as the list above)
habit_cols <- c("bp1_5_1", "bp1_5_2", "bp1_5_3", "bp1_5_4", "bp1_5_5")

# readable labels for habit items (for saved tables)
habit_item_labels <- c(
  "bp1_5_1" = "carry_valuables",
  "bp1_5_2" = "walk_near_home_after_8pm",
  "bp1_5_3" = "visit_relatives_or_friends",
  "bp1_5_4" = "allow_minors_out_alone",
  "bp1_5_5" = "other_habit_change"
)
# long habit table
habit_long <- ensu_march_goc %>%
  # parse weight
  dplyr::mutate(fac_sel = as.numeric(fac_sel)) %>%
  # keep positive weights
  dplyr::filter(!is.na(fac_sel), fac_sel > 0) %>%
  # reshape
  tidyr::pivot_longer(
    cols = dplyr::all_of(habit_cols),
    names_to = "habit_item",
    values_to = "resp"
  ) %>%
  # parse response
  dplyr::mutate(
    resp = as.integer(resp),
    habit_item_label = unname(habit_item_labels[habit_item])
  ) %>%
  # keep clear yes or no answers
  dplyr::filter(resp %in% c(1L, 2L))

# weighted percent yes by item
habit_city_year <- habit_long %>%
  # group
  dplyr::group_by(year_march, cve_ent, nom_ent, cd, nom_cd, habit_item, habit_item_label) %>%
  # summarize
  dplyr::summarise(
    n_substantive = dplyr::n(),
    wt_total = sum(fac_sel, na.rm = TRUE),
    wt_yes = sum(fac_sel[resp == 1L], na.rm = TRUE),
    pct_yes = 100 * wt_yes / wt_total,
    .groups = "drop"
  ) %>%
  # join relevance
  dplyr::left_join(
    goc_city_meta %>% dplyr::select(cve_ent, cd, gulf_relevance, rgn_id, rgn_label),
    by = c("cve_ent", "cd")
  )

# save habit item rates
readr::write_csv(habit_city_year, file.path(path_int, "ensu_march_bp1_5_item_city_year.csv"))
# save output copy
readr::write_csv(habit_city_year, file.path(path_out, "ensu_march_habit_change_item_city_year.csv"))

# person level any habit changed flag
any_habit <- ensu_march_goc %>%
  # parse
  dplyr::mutate(
    fac_sel = as.numeric(fac_sel),
    dplyr::across(dplyr::all_of(habit_cols), as.integer)
  ) %>%
  # keep weights
  dplyr::filter(!is.na(fac_sel), fac_sel > 0) %>%
  # flag if any habit answer is yes
  dplyr::rowwise() %>%
  dplyr::mutate(
    any_yes = {
      # collect habit answers
      vals <- c_across(dplyr::all_of(habit_cols))
      # keep yes no only
      vals <- vals[vals %in% c(1L, 2L)]
      # missing if there are no clear habit answers
      if (length(vals) == 0) NA else any(vals == 1L)
    }
  ) %>%
  dplyr::ungroup() %>%
  # drop empty habit rows
  dplyr::filter(!is.na(any_yes))

# weighted percent any habit change
any_habit_city_year <- any_habit %>%
  # group
  dplyr::group_by(year_march, cve_ent, nom_ent, cd, nom_cd) %>%
  # summarize
  dplyr::summarise(
    n_substantive = dplyr::n(),
    wt_total = sum(fac_sel, na.rm = TRUE),
    wt_any_yes = sum(fac_sel[any_yes], na.rm = TRUE),
    pct_any_habit_change = 100 * wt_any_yes / wt_total,
    .groups = "drop"
  ) %>%
  # join relevance and region
  dplyr::left_join(
    goc_city_meta %>% dplyr::select(cve_ent, cd, gulf_relevance, rgn_id, rgn_label),
    by = c("cve_ent", "cd")
  )
# save composite
readr::write_csv(any_habit_city_year, file.path(path_out, "ensu_march_any_habit_change_city_year.csv"))

# preview
print(any_habit_city_year)
```

## Region totals (weighted from people, not from city averages)

City tables above stay as they are.
Region numbers pool the same people by region using survey weights (`FAC_SEL`).
This is not the average of city percentages.

```{r}
# people with a region, for the unsafe question
bp1_1_persons_rgn <- ensu_march_goc %>%
  dplyr::mutate(
    bp1_1 = as.integer(bp1_1),
    fac_sel = as.numeric(fac_sel)
  ) %>%
  dplyr::filter(bp1_1 %in% c(1L, 2L), !is.na(fac_sel), fac_sel > 0) %>%
  dplyr::left_join(
    goc_city_meta %>% dplyr::select(cve_ent, cd, rgn_id, rgn_label),
    by = c("cve_ent", "cd")
  ) %>%
  dplyr::filter(!is.na(rgn_id))

# percent unsafe by region and march year
bp1_1_rgn_year <- bp1_1_persons_rgn %>%
  dplyr::group_by(year_march, rgn_id, rgn_label) %>%
  dplyr::summarise(
    n_cities = dplyr::n_distinct(cd),
    n_substantive = dplyr::n(),
    wt_total = sum(fac_sel, na.rm = TRUE),
    wt_unsafe = sum(fac_sel[bp1_1 == 2L], na.rm = TRUE),
    pct_unsafe = 100 * wt_unsafe / wt_total,
    .groups = "drop"
  ) %>%
  dplyr::arrange(rgn_id, year_march)

# save region unsafe series
readr::write_csv(bp1_1_rgn_year, file.path(path_int, "ensu_march_bp1_1_rgn_year.csv"))
readr::write_csv(bp1_1_rgn_year, file.path(path_out, "ensu_march_pct_unsafe_rgn_year.csv"))

# people with a region, for any habit change
any_habit_persons_rgn <- any_habit %>%
  dplyr::left_join(
    goc_city_meta %>% dplyr::select(cve_ent, cd, rgn_id, rgn_label),
    by = c("cve_ent", "cd")
  ) %>%
  dplyr::filter(!is.na(rgn_id))

# percent with any habit change by region and march year
any_habit_rgn_year <- any_habit_persons_rgn %>%
  dplyr::group_by(year_march, rgn_id, rgn_label) %>%
  dplyr::summarise(
    n_cities = dplyr::n_distinct(cd),
    n_substantive = dplyr::n(),
    wt_total = sum(fac_sel, na.rm = TRUE),
    wt_any_yes = sum(fac_sel[any_yes], na.rm = TRUE),
    pct_any_habit_change = 100 * wt_any_yes / wt_total,
    .groups = "drop"
  ) %>%
  dplyr::arrange(rgn_id, year_march)

# save region habit series
readr::write_csv(any_habit_rgn_year, file.path(path_out, "ensu_march_any_habit_change_rgn_year.csv"))

# one line per region (pooled people in that region)
p_rgn_agg <- ggplot(
  data = bp1_1_rgn_year,
  mapping = aes(x = year_march, y = pct_unsafe, color = rgn_label, group = rgn_id)
) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  labs(
    title = "ENSU March: percent unsafe by OHI GoC region",
    subtitle = "Survey-weighted total across ENSU cities in each region",
    x = "March year",
    y = "Percent unsafe",
    color = "Region"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom") +
  scale_y_continuous(limits = c(0, 100)) +
  scale_x_continuous(breaks = sort(unique(bp1_1_rgn_year$year_march))) +
  scale_color_manual(values = rgn_label_colors, drop = FALSE)

print(p_rgn_agg)

ggsave(
  filename = file.path(path_out, "ensu_march_pct_unsafe_rgn_aggregate.png"),
  plot = p_rgn_agg,
  width = 10,
  height = 6,
  dpi = 150
)

# preview
print(bp1_1_rgn_year)
print(any_habit_rgn_year)
```

## Latest March snapshot (city and region)

```{r}
# latest march year present in the unsafe series
latest_year <- max(bp1_1_city_year$year_march, na.rm = TRUE)

# city snapshot joining unsafe and habit change
latest_snapshot_city <- bp1_1_city_year %>%
  dplyr::filter(year_march == latest_year) %>%
  dplyr::select(
    year_march, nom_ent, cd, nom_cd, gulf_relevance, rgn_id, rgn_label,
    n_substantive, pct_unsafe
  ) %>%
  dplyr::left_join(
    any_habit_city_year %>%
      dplyr::select(year_march, cd, pct_any_habit_change),
    by = c("year_march", "cd")
  ) %>%
  dplyr::arrange(rgn_id, nom_cd)

# region snapshot joining unsafe and habit change
latest_snapshot_rgn <- bp1_1_rgn_year %>%
  dplyr::filter(year_march == latest_year) %>%
  dplyr::select(year_march, rgn_id, rgn_label, n_cities, n_substantive, pct_unsafe) %>%
  dplyr::left_join(
    any_habit_rgn_year %>%
      dplyr::select(year_march, rgn_id, pct_any_habit_change),
    by = c("year_march", "rgn_id")
  ) %>%
  dplyr::arrange(rgn_id)

# save city and region snapshots
readr::write_csv(
  latest_snapshot_city,
  file.path(path_out, paste0("ensu_march_", latest_year, "_city_snapshot.csv"))
)
readr::write_csv(
  latest_snapshot_rgn,
  file.path(path_out, paste0("ensu_march_", latest_year, "_rgn_snapshot.csv"))
)

# combined long table for convenience (city rows + region rows)
latest_snapshot_combined <- dplyr::bind_rows(
  latest_snapshot_city %>%
    dplyr::mutate(level = "city") %>%
    dplyr::select(
      level, year_march, rgn_id, rgn_label, nom_ent, cd, nom_cd,
      n_substantive, pct_unsafe, pct_any_habit_change
    ),
  latest_snapshot_rgn %>%
    dplyr::mutate(level = "region", nom_ent = NA_character_, cd = NA_character_, nom_cd = NA_character_) %>%
    dplyr::select(
      level, year_march, rgn_id, rgn_label, nom_ent, cd, nom_cd,
      n_substantive, pct_unsafe, pct_any_habit_change
    )
) %>%
  dplyr::arrange(rgn_id, level, nom_cd)

readr::write_csv(
  latest_snapshot_combined,
  file.path(path_out, paste0("ensu_march_", latest_year, "_city_rgn_snapshot.csv"))
)

# print
print(latest_snapshot_city)
print(latest_snapshot_rgn)
```

# Current status score

Current status is the complement of feeling unsafe: **100 - percent unsafe**.
A score of 100 means everyone who answered felt that living in their city is safe (ideal).
A score of 0 means everyone felt unsafe.
This uses the region-pooled March series (`bp1_1_rgn_year`).

```{r}
# current status by region and march year
cs_security <- bp1_1_rgn_year %>%
  dplyr::mutate(current_status = 100 - pct_unsafe) %>%
  dplyr::select(
    year_march, rgn_id, rgn_label, n_cities, n_substantive,
    pct_unsafe, current_status
  ) %>%
  dplyr::arrange(rgn_id, year_march)

# save current status table
readr::write_csv(cs_security, file.path(path_out, "cs_security.csv"))

# one line per region: current status over years
p_cs_security <- ggplot(
  data = cs_security,
  mapping = aes(x = year_march, y = current_status, color = rgn_label, group = rgn_id)
) +
  # ideal reference at 100
  geom_hline(yintercept = 100, linetype = "dashed", color = "black", linewidth = 0.6) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  labs(
    title = "ENSU March: current status score by OHI GoC region",
    subtitle = "Current status = 100 - percent unsafe",
    x = "March year",
    y = "Current status score",
    color = "Region"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom") +
  scale_y_continuous(limits = c(0, 100)) +
  scale_x_continuous(breaks = sort(unique(cs_security$year_march))) +
  scale_color_manual(values = rgn_label_colors, drop = FALSE)

print(p_cs_security)

ggsave(
  filename = file.path(path_out, "cs_security_rgn_status.png"),
  plot = p_cs_security,
  width = 10,
  height = 6,
  dpi = 150
)

# preview
print(cs_security)
```

### Alternative current status score

Alternative current status averages two “good” shares at the region level:

* **percent safe** = `100 − pct_unsafe` (people who said living in their city is safe)
* **percent habits not changed** = `100 − pct_any_habit_change` (people who did not report changing any of the five habits from fear of crime)

\[
\text{alternative current status} = \frac{\text{percent safe} + \text{percent habits not changed}}{2}
\]

Ideal is still 100 (everyone safe and no fear-driven habit changes).

```{r}
# join region unsafe and any-habit series, then average the two "good" percents
cs_security_alt <- bp1_1_rgn_year %>%
  dplyr::select(year_march, rgn_id, rgn_label, n_cities, pct_unsafe) %>%
  dplyr::left_join(
    any_habit_rgn_year %>%
      dplyr::select(year_march, rgn_id, pct_any_habit_change),
    by = c("year_march", "rgn_id")
  ) %>%
  dplyr::mutate(
    pct_safe = 100 - pct_unsafe,
    pct_habits_not_changed = 100 - pct_any_habit_change,
    current_status_alt = (pct_safe + pct_habits_not_changed) / 2
  ) %>%
  dplyr::select(
    year_march, rgn_id, rgn_label, n_cities,
    pct_unsafe, pct_safe,
    pct_any_habit_change, pct_habits_not_changed,
    current_status_alt
  ) %>%
  dplyr::arrange(rgn_id, year_march)

# save alternative current status table
readr::write_csv(cs_security_alt, file.path(path_out, "cs_security_alt.csv"))

# one line per region: alternative current status over years
p_cs_security_alt <- ggplot(
  data = cs_security_alt,
  mapping = aes(x = year_march, y = current_status_alt, color = rgn_label, group = rgn_id)
) +
  # ideal reference at 100
  geom_hline(yintercept = 100, linetype = "dashed", color = "black", linewidth = 0.6) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  labs(
    title = "ENSU March: alternative current status by OHI GoC region",
    subtitle = "Avg of percent safe + percent with no fear-driven habit change",
    x = "Year",
    y = "alt current status score",
    color = "Region"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom") +
  scale_y_continuous(limits = c(0, 100)) +
  scale_x_continuous(breaks = sort(unique(cs_security_alt$year_march))) +
  scale_color_manual(values = rgn_label_colors, drop = FALSE)

print(p_cs_security_alt)

ggsave(
  filename = file.path(path_out, "cs_security_alt_rgn_status.png"),
  plot = p_cs_security_alt,
  width = 10,
  height = 6,
  dpi = 150
)

# preview
print(cs_security_alt)
```

# Saving all plots as PDFs

All plots are saved as PDFs: `/home/shares/ohi/OHI_GOC/goal_prep/ss/security_output/plot_pdfs`.

```{r}
# pdf output folder (server first, local fallback)
path_pdf <- file.path(path_out, "plot_pdfs")
dir.create(path_pdf, recursive = TRUE, showWarnings = FALSE)

# slide-friendly dimensions (16:9 widescreen)
pdf_w <- 10
pdf_h <- 5.63

# helper: save a plot as pdf with legend inside the page
save_slide_pdf <- function(plot, filename) {
  ggsave(
    filename = file.path(path_pdf, filename),
    plot = plot +
      theme(
        legend.position = "bottom",
        legend.box = "horizontal",
        legend.margin = margin(t = 2, b = 2),
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 9),
        plot.margin = margin(10, 10, 10, 10)
      ),
    width = pdf_w,
    height = pdf_h,
    device = cairo_pdf
  )
}

# city-colored unsafe by state
save_slide_pdf(p_unsafe, "ensu_march_pct_unsafe_by_city_year.pdf")

# region-colored unsafe by state
save_slide_pdf(p_unsafe_rgn_color, "ensu_march_pct_unsafe_by_city_year_rgn_color.pdf")

# gulf focus city plot (only if it exists)
if (exists("p_focus")) {
  save_slide_pdf(p_focus, "ensu_march_pct_unsafe_gulf_focus.pdf")
}

# gulf focus faceted by region
if (exists("p_focus_rgn")) {
  save_slide_pdf(p_focus_rgn, "ensu_march_pct_unsafe_by_rgn.pdf")
}

# region aggregate unsafe
save_slide_pdf(p_rgn_agg, "ensu_march_pct_unsafe_rgn_aggregate.pdf")

# current status score
save_slide_pdf(p_cs_security, "cs_security_rgn_status.pdf")

# alternative current status score
save_slide_pdf(p_cs_security_alt, "cs_security_alt_rgn_status.pdf")

message("PDFs saved to: ", path_pdf)
```

# Practical notes

1. March 2021 is included from the DBF package (`ensu_bd_2021_dbf/ensu_bd_marzo_2021_dbf`) when that folder is under the ENSU root (local or Mazu).
2. Years 2014 to 2021 March are often DBF; 2022 onward are CSV. Results keep March 2018 onward only.
3. Years 2014 to 2015 use older `P1` / `P4_*` / `FACTOR` column names (still read when present, not used in the GoC series).
4. City codes from 2014 to 2015 (and Guadalajara `CD` 24) are converted to the later codes; city names come from newer files when older ones are blank.
5. Each ENSU city gets an OHI GoC region through its main municipality in `municipality_to_rgnid_update.csv` (for example Los Mochis → Ahome → Region_8). Nogales has no GoC region and stays unassigned.
6. Region percentages pool people by region with survey weights, and are saved next to the city tables (`*_rgn_year.csv` / `*_rgn_snapshot.csv`).
7. `pct_unsafe` uses only city-level `BP1_1`. Place questions (`BP1_2_*`: casa, trabajo, automóvil, etc.) are not used. Habit change uses `BP1_5_1`–`BP1_5_5`; `pct_any_habit_change` is “yes to at least one of those five.”
8. Current status score is `100 − pct_unsafe` by region (saved as `cs_security.csv`); ideal is 100.
9. Alternative current status averages percent safe and percent habits not changed (saved as `cs_security_alt.csv`); ideal is 100.
10. All plots are also saved as PDFs (10 × 5.63 in, 16:9) in `plot_pdfs/` for slide use.
11. When copying this folder to the shared server, keep the same `ensu_bd_YYYY_*` layout so file discovery continues to work.