---
title: 'OHI GoC 2025: Safety & Security'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
    include-in-header: '../../../src/templates/ohi_hdr.html'
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: inline
editor_options: 
  chunk_output_type: inline
---

# Safety & Security Goal

## Summary

### Background

This script explores data related to safety and security across the five/six (small portion of Jalisco) states bordering the Gulf of California, with particular attention to organized crime and its regional impacts.
Following discussions from the OHI GoC Expert Working Group Meeting (June 2025), this analysis aims to inform the potential development of a dedicated goal addressing safety and security concerns in the region.

**Some key questions brought up in the meeting:**

-   How do safety and security conditions vary across the GoC states?
-   What role does organized crime play as a **pressure** affecting regional well-being and sustainability?
-   How can social justice and resilience metrics be integrated into safety assessments?
    -   How might green spaces act as components of community resilience?
-   What spatial patterns emerge when examining crime incidence and security indicators?

**Essential considerations:** - Sensitive and respectful analysis that avoids country criticism - Perhaps an integration of multiple data sources for comprehensive assessment, if possible - Consideration of both quantitative indicators and qualitative resilience factors

### Data Sources

-   **Organized Crime Index**: Comprehensive assessment of organized crime presence and impact
-   **INEGI Crime Incidence Data**: Official Mexican statistics on crime occurrence and patterns
-   **Semaforo Platform**: Real-time security and risk monitoring data
-   **UN Resilience Indices**: International frameworks for measuring community resilience and green area benefits

*Note: This analysis maintains sensitivity to regional contexts and focuses on constructive data interpretation as it relates to the Gulf's health.*

# Exploring available data

## INEGI: National Survey on Victimization and Perception of Public Safety (ENVIPE) 2024

### About the survey

Information is located [here](https://www.inegi.org.mx/programas/envipe/2024/#microdatos), and is available up to the year 2023 for victimization and 2024 for perception of public safety and performance of authorities.

Full methodology is located [here](https://www.inegi.org.mx/contenidos/programas/envipe/2024/doc/fd_envipe2024.pdf) and in home/shares/ohi/OHI_GOC/_raw_data/inegi_crime/d2025/fd_envipe2024.pdf. 

The 2024 National Survey on Victimization and Perception of Public Safety (ENVIPE) is the fourteenth installment in the statistical series generated by the National Subsystem of Government, Public Safety, and Justice Information (SNIGSPIJ), coordinated by the National Institute of Statistics and Geography (INEGI).
This project continues the topics addressed in previous editions of this survey, the results of which have been declared Information of National Interest by the INEGI Governing Board.

Currently, the implementation of victimization surveys has gained global relevance, both within the government sector, which is responsible for public safety and justice, and within academia and research.
This is due to the fact that societies aspire to a violence-free living environment, in which the physical and property integrity of citizens and institutions is not violated.

The National Survey on Victimization and Perception of Public Safety (ENVIPE) statistical series, conducted by the National Institute of Statistics and Geography (INEGI), responds to the need for statistical data that allow for the planning of public security policies. It gathers information on the incidence of crime affecting households and their members, the crime rate, the characteristics of the crime, the victims and the context of victimization, the perception of public safety, as well as information on the performance and experiences with the institutions in charge of public safety and justice. This is done so that the country's competent authorities have the tools to develop public policies in these areas.

For the design of this statistical series, best practices in victimization surveys were considered, specifically the recommendations of the United Nations (UN). Proposals from public security and justice authorities and academic experts from Mexico were also considered. INEGI's experience in conducting surveys related to security and justice, which dates back to the late 1980s, was also invaluable in its design.

Due to its importance, the ENVIPE was declared information of national interest by the INEGI Governing Board in December 2011. This means that its use is mandatory for the Federation, the states, Mexico City, and municipalities, and its publication is therefore required on a regular and periodic basis.

The 2024 ENVIPE is the fourteenth exercise in this series, which follows the annual exercises conducted uninterruptedly from 2011 to 2023.

**Specific objectives:**

-   Make estimates of the number of people who suffered criminal victimization in 2023.

-   Calculate the number of crimes that occurred during 2023.

-   Make estimates of the black figure of crimes and their causes.

-   To measure the perception of the population aged 18 and over regarding the safety of their place of residence (neighborhood or town, municipality or borough, state) and the places where they carry out their daily activities (work, school, market, public transportation, among others).

-   To measure the degree of trust in public security institutions and the perception of the performance of public security and justice authorities.

-   Identify changes in people's activities due to fear of crime.

-   Estimate household expenses to protect against crime.

-   Know the impact of the crime on the victims.

-   Gather information on the sociodemographic characteristics of the victims.

-   Gather information about the characteristics of the perpetrator and his or her relationship with the victim.

-   Identify victims' attitudes and experiences with public security and law enforcement institutions.

### Exploration

First, let's look at the 6 different CSVs that can be downloaded as "National Survey on Victimization and Perception of Public Safety 2024" [here](https://www.inegi.org.mx/programas/envipe/2024/#datos_abiertos).

There is:

`TVivienda.csv`: Características de las viviendas seleccionadas, sus integrantes y los hogares que las conforman. Contains basic information about the selected dwelling, including geographic identifiers.

`THogar.csv`: Características de los hogares que componen la vivienda seleccionada. Contains household-level data, including some geographic information (linked to TVivienda).

`TSDem.csv`: Características sociodemográficas correspondientes a los integrantes del hogar. Contains sociodemographic information for each household member.

`TPer_Vic1.csv`: Variables correspondientes a la percepción de seguridad pública y el desempeño institucional.  Contains variables related to perceptions of public security, antisocial behaviors, changes in habits due to fear of crime, and trust in authorities. 

`TPer_Vic2.csv`: Características correspondientes a la victimización en el hogar y personal. Contains information about victimization in the household and personal victimization.

`TMod_Vic.csv`: Características correspondientes a los delitos captados por el módulo sobre victimización.  Contains detailed information about crimes experienced by the selected person and their household during the reference year.


```{r}
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  tidyverse,
  here,
  ggplot2,
  RColorBrewer,
  dplyr,
  readr,
  stringr,
  lubridate,
  plotly,
  tictoc
)

# read each of them in to explore them
tvivienda_raw <- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_crime/d2025/bd_envipe_2024_csv","TVivienda.csv")) %>% 
  janitor::clean_names()

thogar_raw <- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_crime/d2025/bd_envipe_2024_csv","THogar.csv")) %>% 
  janitor::clean_names()

tsdem_raw <- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_crime/d2025/bd_envipe_2024_csv","TSDem.csv")) %>% 
  janitor::clean_names()

tper_vic1_raw <- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_crime/d2025/bd_envipe_2024_csv","TPer_Vic1.csv")) %>% 
  janitor::clean_names()

tper_vic2_raw <- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_crime/d2025/bd_envipe_2024_csv","TPer_Vic2.csv")) %>% 
  janitor::clean_names()

tmod_vic_raw <- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_crime/d2025/bd_envipe_2024_csv","TMod_Vic.csv")) %>% 
  janitor::clean_names()
```

Start with tper_vic1 to explore perceptions:

```{r}
municipios <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/land_admin_boundaries/mex_admin_boundaries.shp","mex_admin2.shp")) %>% 
  mutate(nom_mun = str_to_upper(adm2_ref_n),
         nom_ent = str_to_upper(adm1_name)) %>% 
  select(nom_ent, nom_mun)

mapview(municipios)
```

```{r}
tsdem_join_municip <- left_join(tsdem_goc_states, municipios, by = c("nom_ent", "nom_mun"))

total_indiv_municip <- tsdem_join_municip %>% 
  group_by(nom_ent, nom_mun, geometry) %>% 
  summarize(num_individ = n()) %>% 
  st_as_sf()
  

mapview(total_indiv_municip, zcol = "num_individ", layer.name = "Number of responses per municipio")
```

----

```{r}
# explore how many goc cities are present
options(max.print = 2000)

tsdem_goc_states <- tsdem_raw %>% 
  filter(nom_ent %in% c("BAJA CALIFORNIA","SONORA","SINALOA","JALISCO","BAJA CALIFORNIA SUR","NAYARIT")) %>% 
  mutate(edad = as.numeric(edad))
unique(tsdem_goc_states$nom_mun) # 104

library(tidygeocoder)

tsdem_municipios <- tsdem_goc_states %>% 
  mutate(country = "Mexico") %>% 
  tidygeocoder::geocode(city = nom_mun, country = country, state = nom_ent)

library(sf)
tsdem_municipios_sf <- tsdem_municipios %>% 
  filter(!nom_mun %in% c("ZAPOTLAN EL GRANDE")) %>% 
  mutate(lat = ifelse(nom_mun == "BAHIA DE BANDERAS", 20.8295, lat),
         long = ifelse(nom_mun == "BAHIA DE BANDERAS", -105.2706, long)) %>% 
  st_as_sf(coords = c("long", "lat"),
           crs = 4326, # WGS84
           remove = FALSE)

library(mapview)
mapview(tsdem_municipios_sf)

inland_50 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer","inland_50km_goc.shp"))
tsdem_municipios_sf_trans <- st_transform(tsdem_municipios_sf, st_crs(inland_50))
tsdem_goc_rgn <- st_intersection(tsdem_municipios_sf_trans, inland_50)

tsdem_goc_rgn_byrgn <- tsdem_goc_rgn %>% 
  group_by(rgn_id) %>% 
  summarize(num_individ = n())

mapview(tsdem_goc_rgn_byrgn, zcol = "num_individ") + mapview(inland_50, zcol = "rgn_id")
```


```{r}
tper_vic1_goc_states <- tper_vic1_raw %>% 
  filter(nom_ent %in% c("BAJA CALIFORNIA","SONORA","SINALOA","JALISCO","BAJA CALIFORNIA SUR","NAYARIT")) 
unique(tper_vic1_goc_states$nom_mun) # 104

setdiff(unique(tsdem_goc_states$nom_mun), unique(tper_vic1_goc_states$nom_mun))
```


#### The municipalities that have survey data associated with them:

**Baja California/BCS:**
LA PAZ [6]
LOS CABOS [8] (includes Cabo San Lucas, San Jose)
MULEGE [9] (just above bahia de concepcion)
LORETO [10]
COMONDU [7] (I would consider this too far inland)

**Sonora:**
CAJEME [85] (by bahia kino)
GUAYMAS [86]
HERMOSILLO [87] (I might consider this too far inland, depends on our inland bordering)
EMPALME [94] (right next to Guaymas)
HUATABAMPO [96]
NAVOJOA [88] (just inland of Huatabampo)
ETCHOJOA [102] (also just inland of Huatabampo)

**Sinaloa:** 
AHOME [70] (contains Los Mochis)

**Jalisco:** 
PUERTO VALLARTA [12]

```{r}
# if (!require("devtools")) {
#     install.packages("devtools")
# }
# devtools::install_github("diegovalle/mxmaps")
library(mxmaps)

data("df_mxmunicipio") # read in 2015 municipios

join_tper_vic1_tsdem <- left_join(tsdem_goc_states, tper_vic1_raw)

# tper_vic1_goc <- tper_vic1_raw %>% 
  # filter(nom_ent %in% c("BAJA CALIFORNIA","SONORA","SINALOA","JALISCO","BAJA CALIFORNIA SUR","NAYARIT"))

tper_vic1_goc <- tper_vic1_raw %>% 
  filter(nom_mun %in% c(
    "LA PAZ", "LOS CABOS", "MULEGE", "LORETO", "COMONDU",
    "CAJEME", "GUAYMAS", "HERMOSILLO", "EMPALME", "HUATABAMPO", "NAVOJOA", "ETCHOJOA",
    "AHOME",
    "PUERTO VALLARTA"
  ))

tsdem_goc_states_int <- tper_vic1_goc %>% 
  mutate(cve_ent = as.numeric(cve_ent),
         cve_mun = as.numeric(cve_mun),
         region = sprintf("%02d%03d", cve_ent, cve_mun)) %>% 
  rename(value = ap4_3_2) %>% 
  filter(value != "9") %>%  # 4.3 9 = No sabe / no responde (pg 19 https://www.inegi.org.mx/contenidos/programas/envipe/2024/doc/fd_envipe2024.pdf)
  group_by(region) %>%
  summarize(value = mean(value, na.rm = TRUE)) %>%
  ungroup()

tsdem_goc_states_chloro <- mxmunicipio_choropleth(
  tsdem_goc_states_int,
  title = "How safe individuals feel their municipio is (higher the number, the less safe they feel)",
  legend = "Value",
  num_colors = 1 
)


### how safe in jobs people feel 

tsdem_goc_states_job <- tper_vic1_goc %>% 
  mutate(cve_ent = as.numeric(cve_ent),
         cve_mun = as.numeric(cve_mun),
         region = sprintf("%02d%03d", cve_ent, cve_mun)) %>% 
  rename(value = ap4_4_02) %>% 
  filter(!value %in% c("9", "3")) %>%  # 4.3 9 = No sabe / no responde (pg 19 https://www.inegi.org.mx/contenidos/programas/envipe/2024/doc/fd_envipe2024.pdf)
  group_by(region) %>%
  summarize(value = mean(value, na.rm = TRUE)) %>%
  ungroup()

tsdem_goc_states_job_chloro <- mxmunicipio_choropleth(
  tsdem_goc_states_job,
  title = "In terms of crime, tell me if you feel safe or unsafe in your job (higher the number, the less safe they feel)",
  legend = "Value",
  num_colors = 1 
)
" En términos de delincuencia, dígame si se siente seguro(a) o inseguro(a) en…"

# BUT, THIS ONLY MAKES SENSE WHEN COMPARED TO THE POPULATION PER MUNICIPIO!
```
The mxmunicipio_chloropleth function is only really useful if we want to see all of Mexico.  In the future we should instead just use a leaflet.

#### How many GoC municipio responses (how complete is the ENVIPE data?)

##### TPer_Viv1

```{r}
# determine the variables for each dataset that indicates a question of interest

## for tper_viv1
tper_vic1_goc <- tper_vic1_raw %>% 
  filter(nom_mun %in% c(
    "LA PAZ", "LOS CABOS", "MULEGE", "LORETO", "COMONDU",
    "CAJEME", "GUAYMAS", "HERMOSILLO", "EMPALME", "HUATABAMPO", "NAVOJOA", "ETCHOJOA",
    "AHOME",
    "PUERTO VALLARTA"
  ))
colnames(tper_vic1_goc) # 232 columns... ap start questions

tper_vic1_goc_complete_detailed <- tper_vic1_goc %>%
  group_by(nom_mun) %>%
  summarise(
    # total observations per municipality
    total_observations = n(),
    
    # for each ap question, let's calculate both counts and rates
    across(starts_with("ap"), 
           list(
             # substantive responses 
             substantive_count = ~sum(. %in% 1:8, na.rm = TRUE),
             
             # "don't know" responses  
             dont_know_count = ~sum(. == 9, na.rm = TRUE),
             
             # true missing values
             missing_count = ~sum(is.na(.)),
             
             # prop of substantive responses 
             substantive_rate = ~mean(. %in% 1:8, na.rm = TRUE),
             
             # check the math: substantive_count / total_observations should equal substantive_rate
             manual_rate_check = ~sum(. %in% 1:8, na.rm = TRUE) / n()
           ),
           .names = "{.fn}_{.col}")
  )

## make heatmap to visualize the substantive rate for tmod_vic
heatmap_tper_vic1_goc_complete <- tper_vic1_goc_complete_detailed %>%
  select(nom_mun, starts_with("substantive_rate")) %>%
  pivot_longer(cols = starts_with("substantive_rate"), 
               names_to = "question", 
               values_to = "substantive_rate") %>%
  mutate(question = str_remove(question, "substantive_rate_"),
         substantive_rate = as.numeric(substantive_rate)) %>% 
  mutate(completion_level = case_when(
    substantive_rate < 0.5 ~ "bad",
    substantive_rate > 0.5 ~ "good",
    TRUE ~ NA
  ))

ggplot(heatmap_tper_vic1_goc_complete, aes(x = question, y = nom_mun, fill = substantive_rate)) +
  # white borders for better visualizing
  geom_tile(color = "white", size = 0.5) +
  
  # green to red, good to bad (bad meaning missing values)
  scale_fill_gradient2(low = "red",   # low substantive response rate (bad)
                       mid = "yellow", # medium substantive response rate (okay) 
                       high = "darkgreen",  # high substantive response rate (good)
                       midpoint = 0.5, # set midpoint at 50% completion
                       name = "Substantive\nResponse Rate \nNot '9' or 'NA'",
                       labels = scales::percent_format()) +  # as percentages
  
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 70, hjust = 1, size = 9),
        plot.title = element_text(size = 14),
        axis.title = element_text(size = 12)) +
  
  labs(title = "Substantive response rates by municipality and question: TPer_Viv1",
       subtitle = "Perception of safety and ability of institutions in charge of public safety and justice",
       x = "Survey questions", 
       y = "Municipio")

#########################
# the questions with good completion (usable questions)
heatmap_tper_vic1_goc_good <- heatmap_tper_vic1_goc_complete %>% 
  filter(completion_level %in% "good") %>%
  group_by(question) %>% 
  summarise(num_municipios = n()) %>% # see how many municipios have a good completion % for that question.  The maximum is 14.
  filter(num_municipios %in% 10:14) # only keep the ones that have 10 or more municipios that have good completion. If the question does not have a good substantive rate then we cannot compare across regions.
unique(heatmap_tper_vic1_goc_good$question)
```


```{r}
# ### how many responses there were per municipio/estado
# response_summary <- tper_vic1_goc %>% 
#   group_by(nom_ent, nom_mun) %>%  #group by both state and municipality
#   summarise(
#     n_responses = n(),  # count total responses per municipality
#     .groups = "drop"  # remove grouping after summarise
#   ) %>%
#   # add state-level summaries for comparison
#   group_by(nom_ent) %>%
#   mutate(
#     state_total = sum(n_responses),  # tot responses per state
#     mun_pct_of_state = (n_responses / state_total) * 100,  # each muni's share of state responses
#     n_municipios = n()  # the number of municipalities per state
#   ) %>%
#   ungroup()
# 
# # calc additional summary statistics for deeper analysis
# detailed_summary <- response_summary %>%
#   group_by(nom_ent) %>%
#   summarise(
#     n_municipios = n(),
#     total_responses = sum(n_responses),
#     avg_responses_per_mun = mean(n_responses),
#     median_responses_per_mun = median(n_responses),
#     min_responses = min(n_responses),
#     max_responses = max(n_responses),
#     # coefficient of variation shows how much municipalities vary within states
#     cv_responses = sd(n_responses) / mean(n_responses),
#     .groups = "drop"
#   )
# 
# # viz showing response distribution characteristics by state
# p2 <- ggplot(response_summary, aes(x = reorder(nom_ent, state_total))) +
#   # Box plots show distribution of municipality response counts within each state
#   geom_boxplot(aes(y = n_responses), alpha = 0.6, outlier.alpha = 0.5) +
#   # Add points for individual municipalities
#   geom_jitter(aes(y = n_responses, color = nom_ent), 
#               width = 0.2, alpha = 0.7, size = 1.5) +
#   coord_flip() +
#   scale_y_continuous(labels = scales::comma_format()) +
#   labs(
#     title = "Distribution of Municipality Response Counts by State",
#     subtitle = "Box plots show quartiles, points show individual municipalities",
#     x = "State",
#     y = "Responses per Municipality"
#   ) +
#   theme_minimal() +
#   theme(legend.position = "none")
# 
# print(p2)
# 
# tper_vic1_goc_complete <- tper_vic1_goc %>% 
#   summarise_all(~sum(is.na(.))) %>% 
#   pivot_longer(everything(), names_to = "variable", values_to = "missing_count") %>%
#   mutate(completeness = 1 - missing_count/nrow(tper_vic1_goc)) %>%
#   arrange(completeness)
# 
# ## completeness by state and question
# tper_vic1_goc_complete_state_ap4 <- tper_vic1_goc %>%
#   group_by(nom_ent) %>%
#   summarise(
#     n_responses = n(),
#     across(starts_with("ap4"), 
#            list(count = ~sum(. == "9", na.rm = TRUE), # when they answer "9", it is that they do not know or they did not respond.
#                 pct = ~mean(. == "9", na.rm = TRUE) * 100), 
#            .names = "nosabe_{.col}_{.fn}")
#   )
# 
# ## make heatmap to visualize
# heatmap_tper_vic1_goc <- tper_vic1_goc_complete_state_ap4 %>%
#   select(nom_ent, ends_with("_pct")) %>%
#   pivot_longer(cols = ends_with("_pct"), 
#                names_to = "question", 
#                values_to = "pct_nosabe") %>%
#   mutate(question = str_remove(question, "nosabe_|_pct"))
# 
# ggplot(heatmap_tper_vic1_goc, aes(x = question, y = nom_ent, fill = pct_nosabe)) +
#   geom_tile(color = "white", size = 0.5) +
#   scale_fill_gradient2(low = "white", mid = "yellow", high = "red", 
#                        midpoint = 5, name = "% 'I don't know'/No Response") +
#   theme_minimal() +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
#   labs(title = "Percentage of Missing Responses by GOC State and Question: TPer_Viv1",
#        x = "Questions", y = "State")
# 
# ## completeness by municipio and question
# tper_vic1_goc_complete_mun_ap4 <- tper_vic1_goc %>%
#   group_by(nom_mun) %>%
#   summarise(
#     n_responses = n(),
#     across(starts_with("ap4"), 
#            list(count = ~sum(. == "9", na.rm = TRUE), # when they answer "9", it is that they do not know or they did not respond.
#                 pct = ~mean(. == "9", na.rm = TRUE) * 100), 
#            .names = "nosabe_{.col}_{.fn}")
#   )
# 
# ## make heatmap to visualize
# heatmap_tper_vic1_goc_mun <- tper_vic1_goc_complete_mun_ap4 %>%
#   select(nom_mun, ends_with("_pct")) %>%
#   pivot_longer(cols = ends_with("_pct"), 
#                names_to = "question", 
#                values_to = "pct_nosabe") %>%
#   mutate(question = str_remove(question, "nosabe_|_pct"))
# 
# ggplot(heatmap_tper_vic1_goc_mun, aes(x = question, y = nom_mun, fill = pct_nosabe)) +
#   geom_tile(color = "white", size = 0.5) +
#   scale_fill_gradient2(low = "white", mid = "yellow", high = "red", 
#                        midpoint = 5, name = "% 'I don't know'/No Response") +
#   theme_minimal() +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
#   labs(title = "Percentage of Missing Responses by GOC Municipio and Question: TPer_Viv1",
#        x = "Questions", y = "Municipio")
# 
# tper_vic1_goc_complete_state_ap4 <- tper_vic1_goc %>%
#   filter(if_any(starts_with("ap4"), ~ . == "9")) %>%  # rows with at least one 9
#   group_by(nom_ent) %>%
#   summarise(across(starts_with("ap4"), ~sum(. == "9", na.rm = TRUE), .names = "nosabe_{.col}"))
# 
# ## double check that there are no missing values in ap4
# tper_vic1_goc_ap4 <- tper_vic1_goc %>% 
#   select(across(starts_with("ap4")))
# 
# tper_vic1_goc_complete_state_ap5 <- tper_vic1_goc %>%
#   group_by(nom_ent) %>%
#   summarise(across(starts_with("a57"), ~mean(!is.na(.)), .names = "complete_{.col}"))
```

##### TPer_Viv2

```{r}
#######################################
## for tper_viv2
tper_vic2_goc <- tper_vic2_raw %>% 
  filter(nom_mun %in% c(
    "LA PAZ", "LOS CABOS", "MULEGE", "LORETO", "COMONDU",
    "CAJEME", "GUAYMAS", "HERMOSILLO", "EMPALME", "HUATABAMPO", "NAVOJOA", "ETCHOJOA",
    "AHOME",
    "PUERTO VALLARTA"
  ))
colnames(tper_vic2_goc) # 150 columns... 

# tper_vic2_goc_complete <- tper_vic2_goc %>% 
#   summarise_all(~sum(is.na(.))) %>% 
#   pivot_longer(everything(), names_to = "variable", values_to = "missing_count") %>%
#   mutate(completeness = 1 - missing_count/nrow(tper_vic2_goc)) %>%
#   arrange(completeness)
# 
# ## completeness by state and question
# tper_vic2_goc_complete_state_ap6 <- tper_vic2_goc %>%
#   group_by(nom_ent) %>%
#   summarise(across(starts_with("ap6"), ~mean(!is.na(.)), .names = "complete_{.col}"))
# 
# tper_vic2_goc_complete_state_ap7 <- tper_vic2_goc %>%
#   group_by(nom_ent) %>%
#   summarise(across(starts_with("ap7"), ~mean(!is.na(.)), .names = "complete_{.col}"))

tper_vic2_goc_complete_detailed <- tper_vic2_goc %>%
  group_by(nom_mun) %>%
  summarise(
    # total observations per municipality
    total_observations = n(),
    
    # for each ap question, let's calculate both counts and rates
    across(starts_with("ap"), 
           list(
             # substantive responses 
             substantive_count = ~sum(. %in% 1:8, na.rm = TRUE),
             
             # "don't know" responses  
             dont_know_count = ~sum(. == 9, na.rm = TRUE),
             
             # true missing values
             missing_count = ~sum(is.na(.)),
             
             # prop of substantive responses 
             substantive_rate = ~mean(. %in% 1:8, na.rm = TRUE),
             
             # check the math: substantive_count / total_observations should equal substantive_rate
             manual_rate_check = ~sum(. %in% 1:8, na.rm = TRUE) / n()
           ),
           .names = "{.fn}_{.col}")
  )

## make heatmap to visualize the substantive rate for tmod_vic
heatmap_tper_vic2_goc_complete <- tper_vic2_goc_complete_detailed %>%
  select(nom_mun, starts_with("substantive_rate")) %>%
  pivot_longer(cols = starts_with("substantive_rate"), 
               names_to = "question", 
               values_to = "substantive_rate") %>%
  mutate(question = str_remove(question, "substantive_rate_"),
         substantive_rate = as.numeric(substantive_rate)) %>% 
  mutate(completion_level = case_when(
    substantive_rate < 0.5 ~ "bad",
    substantive_rate > 0.5 ~ "good",
    TRUE ~ NA
  ))

ggplot(heatmap_tper_vic2_goc_complete, aes(x = question, y = nom_mun, fill = substantive_rate)) +
  # white borders for better visualizing
  geom_tile(color = "white", size = 0.5) +
  
  # green to red, good to bad (bad meaning missing values)
  scale_fill_gradient2(low = "red",   # low substantive response rate (bad)
                       mid = "yellow", # medium substantive response rate (okay) 
                       high = "darkgreen",  # high substantive response rate (good)
                       midpoint = 0.5, # set midpoint at 50% completion
                       name = "Substantive\nResponse Rate \nNot '9' or 'NA'",
                       labels = scales::percent_format()) +  # as percentages
  
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 70, hjust = 1, size = 9),
        plot.title = element_text(size = 14),
        axis.title = element_text(size = 12)) +
  
  labs(title = "Substantive response rates by municipality and question: TPer_Viv2",
       subtitle = "Information about victimization in the household and personal victimization.",
       x = "Survey questions", 
       y = "Municipio")

#########################
# the questions with good completion (usable questions)
heatmap_tper_vic2_goc_good <- heatmap_tper_vic2_goc_complete %>% 
  filter(completion_level %in% "good") %>%
  group_by(question) %>% 
  summarise(num_municipios = n()) %>% # see how many municipios have a good completion % for that question.  The maximum is 14.
  filter(num_municipios %in% 10:14) # only keep the ones that have 10 or more municipios that have good completion. If the question does not have a good substantive rate then we cannot compare across regions.
unique(heatmap_tper_vic2_goc_good$question)
```


##### TMod_Vic 

TMod_Vic is about crimes that occurred.  Upon exploration, note that sometimes there are multiple individuals from the same household (id_viv)

```{r}
#######################################
## for tmod_vic
setdiff(tmod_vic_raw$id_viv, tvivienda_raw$id_viv)
# character(0)
setdiff(tvivienda_raw$id_viv, tmod_vic_raw$id_viv)
# [1] "0100005.01" "0100005.04" "0100005.05" "0100012.02"

# okay, so all of the id_viv observations are present in tmod_vic_raw as well!

intersect(colnames(tmod_vic_raw), colnames(tvivienda_raw))
# [1] "x1"      "id_viv"  "upm"     "viv_sel" "dominio" "est_dis"
# [7] "upm_dis"

tmod_vic_raw_0160078.15 <- tmod_vic_raw %>% 
  filter(id_viv %in% "0160078.15") # multiple of the same id_viv

tmod_vic_w_nom_ent <- left_join(tmod_vic_raw, tvivienda_raw, by = c("id_viv", "upm", "viv_sel", "dominio", "est_dis", "upm_dis")) %>% 
  select(-c("x1.x","x1.y"))

nrow(tmod_vic_w_nom_ent) == nrow(tmod_vic_raw) # TRUE, same number of rows!
c("nom_ent","nom_mun") %in% colnames(tmod_vic_w_nom_ent) # TRUE

tmod_vic_goc <- tmod_vic_w_nom_ent %>% 
  filter(nom_mun %in% c(
    "LA PAZ", "LOS CABOS", "MULEGE", "LORETO", "COMONDU",
    "CAJEME", "GUAYMAS", "HERMOSILLO", "EMPALME", "HUATABAMPO", "NAVOJOA", "ETCHOJOA",
    "AHOME",
    "PUERTO VALLARTA"
  ))
colnames(tmod_vic_goc) # 144 columns...

# determine the completeness of the responses (how many NAs) or "9", which means they left it blank
tmod_vic_goc_complete_detailed <- tmod_vic_goc %>%
  group_by(nom_mun) %>%
  summarise(
    # total observations per municipality
    total_observations = n(),
    
    # for each bp question, let's calculate both counts and rates
    across(starts_with("bp"), 
           list(
             # substantive responses 
             substantive_count = ~sum(. %in% 1:8, na.rm = TRUE),
             
             # "don't know" responses  
             dont_know_count = ~sum(. == 9, na.rm = TRUE),
             
             # true missing values
             missing_count = ~sum(is.na(.)),
             
             # prop of substantive responses 
             substantive_rate = ~mean(. %in% 1:8, na.rm = TRUE),
             
             # check the math: substantive_count / total_observations should equal substantive_rate
             manual_rate_check = ~sum(. %in% 1:8, na.rm = TRUE) / n()
           ),
           .names = "{.fn}_{.col}")
  )

## make heatmap to visualize the substantive rate for tmod_vic
heatmap_tmod_vic_goc_complete <- tmod_vic_goc_complete_detailed %>%
  select(nom_mun, starts_with("substantive_rate")) %>%
  pivot_longer(cols = starts_with("substantive_rate"), 
               names_to = "question", 
               values_to = "substantive_rate") %>%
  mutate(question = str_remove(question, "substantive_rate_"),
         substantive_rate = as.numeric(substantive_rate)) %>% 
  mutate(completion_level = case_when(
    substantive_rate < 0.5 ~ "bad",
    substantive_rate > 0.5 ~ "good",
    TRUE ~ NA
  ))

ggplot(heatmap_tmod_vic_goc_complete, aes(x = question, y = nom_mun, fill = substantive_rate)) +
  # white borders for better visualizing
  geom_tile(color = "white", size = 0.5) +
  
  # green to red, good to bad (bad meaning missing values)
  scale_fill_gradient2(low = "red",   # low substantive response rate (bad)
                       mid = "yellow", # medium substantive response rate (okay) 
                       high = "darkgreen",  # high substantive response rate (good)
                       midpoint = 0.5, # set midpoint at 50% completion
                       name = "Substantive\nResponse Rate \nNot '9' or 'NA'",
                       labels = scales::percent_format()) +  # as percentages
  
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 14),
        axis.title = element_text(size = 12)) +
  
  labs(title = "Substantive response rates by municipality and question: TMod_Vic",
       subtitle = "Information about crimes experienced by the selected person and their household during the reference year",
       x = "Survey questions", 
       y = "Municipality")

#########################
# the questions with good completion (usable questions)
heatmap_tmod_vic_goc_good <- heatmap_tmod_vic_goc_complete %>% 
  filter(completion_level %in% "good") %>%
  group_by(question) %>% 
  summarise(num_municipios = n()) %>% # see how many municipios have a good completion % for that question.  The maximum is 14.
  filter(num_municipios %in% 10:14) # only keep the ones that have 10 or more municipios that have good completion. If the question does not have a good substantive rate then we cannot compare across regions.
unique(heatmap_tmod_vic_goc_good$question)

```




