################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```

```{r Load clean_names_lowercase function}
clean_names_lowercase <- function(x) {

  x <- iconv(x, from = "", to = "UTF-8", sub = "")
  x <- stringi::stri_trans_general(x, "Latin-ASCII")

  x %>%
    tolower() %>%
    trimws()
}
```

```{r Sector helper functions}
# Fisheries: marine and freshwater-or-marine capture and fishery sectors.
# Mariculture: marine and freshwater-or-marine aquaculture only (excludes
# freshwater-only aquaculture).

is_fishery_sector <- function(sector) {
  grepl("marine|freshwater or marine", sector, ignore.case = TRUE) &
    grepl("fishery|capture", sector, ignore.case = TRUE)
}

is_mariculture_sector <- function(sector) {
  grepl("marine|freshwater or marine", sector, ignore.case = TRUE) &
    grepl("aquaculture", sector, ignore.case = TRUE)
}
```


# STEP A: Load input data ------------------------------

```{r Load FIS scores}
FIS_scores0 <- read.csv("/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_StockStatusScores_GeoMean_Jun012026.csv")
# 88 obs., 3 vars.
```

```{r Load MAR scores}
MAR_scores0 <- read.csv("/home/ferrer/ohi_goc_prep/FP/FP_Scores/MAR_Scores/MAR_Scores_Jun012026.csv")
# 55 obs., 16 vars.
```

```{r Load in GOC_MarineLandings0 as necessary}
GOC_Landings0<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Landings_SectorSorted.Georef_2000.present_May262026.rds"))
# 3490044 obs., 57 var.

table(GOC_Landings0$GOC_Include_Exclude)
```


# STEP B: Prep landings for sector weights and fishery catch ------------------------------

```{r Filter landings to GOC include and analysis years}
GOC_Landings <- GOC_Landings0 %>%
  filter(
    year >= 2018,
    year <= 2024,
    !is.na(OHI_region_est),
    OHI_region_est != ""
  ) %>%
  mutate(
    OHI_region_est = clean_names_lowercase(OHI_region_est),
    estimated_species_name = clean_names_lowercase(estimated_species_name),
    estimated_genus_name = clean_names_lowercase(estimated_genus_name),
    sector_estimate_draft = as.character(sector_estimate_draft)
  )
# Subset of full landings for 2018-2024, included GOC records only.
```

```{r Summarize sector landings weights per region and year}
sector_landings <- GOC_Landings %>%
  mutate(
    fishery_landings_kg = if_else(
      is_fishery_sector(sector_estimate_draft),
      weight_kg_corrected,
      0
    ),
    mariculture_landings_kg = if_else(
      is_mariculture_sector(sector_estimate_draft),
      weight_kg_corrected,
      0
    )
  ) %>%
  group_by(year, OHI_region_est) %>%
  summarize(
    fishery_landings_kg = sum(fishery_landings_kg, na.rm = TRUE),
    mariculture_landings_kg = sum(mariculture_landings_kg, na.rm = TRUE),
  .groups = "drop"
  ) %>%
  mutate(
    total_fp_landings_kg = fishery_landings_kg + mariculture_landings_kg,
    w_fisheries = fishery_landings_kg / total_fp_landings_kg,
    w_mariculture = mariculture_landings_kg / total_fp_landings_kg
  )
```

```{r Sanity check sector landings}
colnames(sector_landings)

sector_landings %>%
  filter(total_fp_landings_kg > 0) %>%
  summarize(
    min_w_fisheries = min(w_fisheries, na.rm = TRUE),
    max_w_fisheries = max(w_fisheries, na.rm = TRUE)
  )
```


# STEP C: Fisheries catch by stock (species or genus) ------------------------------

```{r Filter landings to fishery sectors and define stock key}
fishery_landings <- GOC_Landings %>%
  filter(is_fishery_sector(sector_estimate_draft)) %>%
  mutate(
    taxon_key = if_else(
      is.na(estimated_species_name) | estimated_species_name == "",
      estimated_genus_name,
      estimated_species_name
    ),
    taxon_level = if_else(
      is.na(estimated_species_name) | estimated_species_name == "",
      "genus",
      "species"
    )
  ) %>%
  filter(!is.na(taxon_key), taxon_key != "")
```

```{r Summarize fishery catch by stock, region, and year}
fishery_stock_catch <- fishery_landings %>%
  group_by(year, OHI_region_est, taxon_key, taxon_level, estimated_genus_name) %>%
  summarize(
    stock_catch_kg = sum(weight_kg_corrected, na.rm = TRUE),
    .groups = "drop"
  )
# Catch per stock (species when available, otherwise genus).
```


# STEP D: Attach FIS Stock_Status and compute regional fisheries score ------------------------------

```{r Prep sector_landings FIS_scores0 and MAR_scores0 for merge}
colnames(sector_landings)

sector_landings<- sector_landings %>%
  mutate(
    OHI_region_est= clean_names_lowercase(OHI_region_est)
  )

FIS_scores0<- FIS_scores0 %>%
  mutate(
    OHI_region_est= clean_names_lowercase(OHI_region_est),
    year= yr
  )

MAR_scores0<- MAR_scores0 %>%
  mutate(
    OHI_region_est= clean_names_lowercase(OHI_region_est)
  )
```

```{r Merge FIS scores onto fishery stock catch}
Final_frame <- sector_landings %>%
  left_join(
    FIS_scores0,
    by = c(
      "year",
      "OHI_region_est"
    )
  )
```

```{r Catch-weighted fisheries score per OHI region and year}
# FIS = 100 * sum(Stock_Status * catch_stock) / sum(catch_stock) over scored stocks.

FIS_region_scores <- Final_frame %>%
  filter(!is.na(GeoMean_StockStatus), total_fp_landings_kg > 0) %>%
  group_by(year, OHI_region_est) %>%
  summarize(
    FIS_score = 100 * sum(GeoMean_StockStatus * total_fp_landings_kg) / sum(total_fp_landings_kg),
    n_stocks_scored = n(),
    scored_fishery_catch_kg = sum(total_fp_landings_kg),
    .groups = "drop"
  )
```

```{r Sanity check FIS regional scores}
colnames(FIS_region_scores)

summary(FIS_region_scores$FIS_score)
  #  Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  # 42.03   63.80   78.14   76.44   91.11  100.00

FIS_region_scores %>%
  arrange(year, OHI_region_est)
```


# STEP E: Prep mariculture scores ------------------------------

```{r Prep MAR scores}
MAR_region_scores <- MAR_scores0 %>%
  filter(
    year >= 2018,
    year <= 2024,
    !is.na(OHI_region_est),
    OHI_region_est != ""
  ) %>%
  mutate(OHI_region_est = clean_names_lowercase(OHI_region_est)) %>%
  select(
    year,
    OHI_region_est,
    MAR_score = total_score,
    total_production_mt
  )
```

```{r Sanity check MAR scores}
colnames(MAR_region_scores)

summary(MAR_region_scores$MAR_score)
```


# STEP F: Combine fisheries and mariculture into food score ------------------------------

```{r Build final food provision scores}
GOC_FP_scores <- sector_landings %>%
  left_join(FIS_region_scores, by = c("year", "OHI_region_est")) %>%
  left_join(MAR_region_scores, by = c("year", "OHI_region_est")) %>%
  mutate(
  FP_score = rowSums(
    cbind(
      w_fisheries * FIS_score,
      w_mariculture * MAR_score
    ),
    na.rm = TRUE
  )
) %>%
  select(
    year,
    OHI_region_est,
    fishery_landings_kg,
    mariculture_landings_kg,
    total_fp_landings_kg,
    w_fisheries,
    w_mariculture,
    FIS_score,
    n_stocks_scored,
    scored_fishery_catch_kg,
    MAR_score,
    total_production_mt,
    FP_score
  ) %>%
  arrange(OHI_region_est, year)
```

```{r Sanity check final food scores}
colnames(GOC_FP_scores)

GOC_FP_scores %>%
  filter(!is.na(FP_score))

summary(GOC_FP_scores$FP_score)
```


```{r Plot}
GOC_FP_scores %>%
  ggplot(aes(
    x = year,
    y = FP_score,
    color = OHI_region_est
  )) +
  
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  
  facet_wrap(~ OHI_region_est, scales = "fixed") +
  
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey50") +
  
  labs(
    title = "FP Scores by Region [DRAFT - June 1 2026]",
    y = "FP Score (from 0 to 100)",
    x = "Year"
  ) +
  
  ylim(0, 101) +
  
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 14, face = "bold")
    # plot.title = element_text(size = 18, face = "bold")
  )
```


# FINAL STEP: SAVE ------------------------------

```{r SAVE GOC_FP_scores}
write.csv(
  GOC_FP_scores,
  file = file.path(WORK_DIR, "GOC_FP_scores_June012026.csv"),
  row.names = FALSE
)
```