---
title: 'OHI GoC 2024: Lasting Special Places (LSP)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# Sentido de Pertenencia: Lasting Special Places

**Script 7 — Residual “marine location of importance” rows.** After scripts 2–6 chew through thematic inventories, some EWG coordinates still float outside those pipelines. Here we diff the master spreadsheet against outputs already saved to disk, spatially test leftovers against ANPs/SIMEC/Ramsar, and—for stubborn zeros—pull curated legal narratives from `lsps_codified_protections_estimated_condition.csv`.

**Contents**

1. Load prepared inventory + previously computed category CSVs; harmonise column names for anti-join logic.
2. Build `lsps_missing_condition`, buffer, spatial join ANPs, fuzzy SIMEC + Ramsar splits similar to earlier scripts.
3. Merge codified scores for remaining zeros, average per region, write “other marine places” outputs.

## All other LSPs that should lie within protected areas, or should have codified protections in the law

Grab `ewg_lsp_inventory_prepared.csv` fresh from `google_sheets/`—everything downstream keys off this authoritative row count.

```{r}
lsp_google_sheets_dir <- "/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/google_sheets"
lsp_list_ewg <- read_csv(file.path(lsp_google_sheets_dir, "ewg_lsp_inventory_prepared.csv")) %>% 
  janitor::clean_names()
```


### Join together the large categories of LSPs

Read script outputs (`condition_lsp_*`) so we know which `lasting_special_place_lsp` strings already have scores—saves us double-counting habitat polygons handled in script 6, etc.

```{r}
# fishing sites
fishing_site_condition <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/output/condition_lsp_fishing_sites.csv"))

# archaeological sites
archaeological_site_condition <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/output/condition_lsp_archaeological_sites.csv"))

# protected area sites
protected_area_condition <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/output/condition_lsp_protected_areas.csv"))

# dive sites
dive_sites_condition <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/output/condition_lsp_dive_sites.csv"))
```

```{r}
colnames(fishing_site_condition)
colnames(archaeological_site_condition)
colnames(protected_area_condition)
colnames(dive_sites_condition)
```


```{r}
fishing_site_condition_join <- fishing_site_condition %>% 
  dplyr::select(rgn_id, lsp_name = lasting_special_place_lsp, condition = overall_condition)

archaeological_site_condition_join <- archaeological_site_condition %>% 
  dplyr::select(rgn_id, lsp_name = lasting_special_place_lsp, condition = overall_condition)

protected_area_condition_join <- protected_area_condition %>% 
  mutate(lsp_name = case_when(
    !is.na(wdpa_orig_name) ~ wdpa_orig_name,
    is.na(wdpa_orig_name) & !is.na(i_efectividad_name) ~ i_efectividad_name,
    is.na(wdpa_orig_name) & is.na(i_efectividad_name) ~ lasting_special_place_description,
    TRUE ~ NA
  )) %>% 
  mutate(lsp_name = normalize_name(lsp_name)) %>% 
  mutate(lsp_name = str_replace_all(lsp_name, "LA TOVARA","La Tovara")) %>% 
  dplyr::select(rgn_id, lsp_name, condition = overall_condition)

dive_sites_condition_join <- dive_sites_condition %>% 
  mutate(lsp_name = paste0("Dive Site:", sep = " ", nombre_sitio_de_buceo)) %>% 
  dplyr::select(rgn_id, lsp_name, condition = overall_condition)
```


# Join them all together

`bind_rows` builds the union of labelled sites with conditions; `setdiff` reveals whose coordinates still need bespoke logic.

```{r}
all_categorical_lsps <- bind_rows(fishing_site_condition_join, archaeological_site_condition_join, protected_area_condition_join, dive_sites_condition_join) 

lsps_not_in_categories <- setdiff(lsp_list_ewg$lasting_special_place_lsp, all_categorical_lsps$lsp_name)
```

## LSPs from the EWG list that are not captured by the previous categories:

Filter to marine-importance rows still absent from the stitched table, drop habitat duplicates explicitly enumerated in `filter(!lasting_special_place_lsp %in% ...)`, and promote coordinates to `sf` for buffering steps ahead.

```{r}
lsps_missing_condition <- lsp_list_ewg %>% 
  filter(lasting_special_place_lsp %in% lsps_not_in_categories) %>% 
  filter(type_of_lsp %in% "Marine location of importance") %>% 
  mutate(lat = as.numeric(lat),
         lont = as.numeric(long)) %>% 
  filter(!is.na(lat)) %>% 
  st_as_sf(
    coords = c("long", "lat"), # making sf object
    crs = 4326, # WGS84
    remove = FALSE
  ) %>% 
  filter(!lasting_special_place_lsp %in% c("Coquina Reefs in Libertad, Bahia Lobos, and Peñasco - AHW","Only coral reefs in Gulf (Cabo Pulmo/Los Frailes & Bahía San Gabriel on Is. Espíritu Santo). RCB, BW","Massive seagrass beds (Zostera marina), RCB, BW\n\nThe largest and most stable annual seagrass beds (Zostera marina) are in the eastern Pacific. The eelgrass in the Gulf of California is 100% annual, no perennial, and is considered a subspecies, JTC, BW","Only coral reefs in Gulf (Cabo Pulmo/Los Frailes & Bahía San Gabriel on Is. Espíritu Santo). RCB, BW", "Puerto Lobos N most mangrove (black mangrove) population, BW","Mangrove coastal lagoons, JTC.")) %>% 
  dplyr::select(rgn_id, lsp_name = lasting_special_place_lsp)
```

## Determine if they overlap with protected areas

Normalize `rgn_id` spacing, map against `all_protected`, buffer 100 m identical to other point workflows, run `st_join` for polygon labels.

```{r}
lsps_missing_condition_clean <- lsps_missing_condition %>% 
  mutate(rgn_id = str_replace_all(rgn_id, " ", "_"))
```

```{r}
mapview(lsps_missing_condition_clean) + mapview(all_protected)
```

```{r}
# buffer dive sites by 1000m
lsps_missing_condition_clean_buffered <- st_buffer(
  st_transform(lsps_missing_condition_clean, st_crs(all_protected)),
  dist = 100
)
mapview(lsps_missing_condition_clean_buffered) + mapview(all_protected)

# spatial join to the non-dissolved layer (all_protected) to keep protected area names
lsps_missing_protected <- st_join(
  lsps_missing_condition_clean_buffered,
  all_protected %>% dplyr::select(name, geometry),
  join = st_intersects,
  left = TRUE  # keep all dive sites, even those outside of protected areas
) %>%
  mutate(in_protected_area = !is.na(name)) %>%
  rename(protected_area_name = name) 

# quick checks for sanity
mapview(lsps_missing_condition_clean) + mapview(lsps_missing_protected, zcol = "in_protected_area") + mapview(all_protected, col.regions = "pink")
table(lsps_missing_protected$in_protected_area)
table(lsps_missing_protected$protected_area_name)


```

The only problem is that if a buffered site overlaps multiple protected areas, `st_join()` will return one row per match, giving us duplicate sites. So that is something to keep in mind.

### i-efectividad for each dive site in a protected area

(Historical heading—) reload SIMEC (`google_sheets/simec_i_efectividad_anp_2025.csv`) and rebuild normalization helpers immediately ahead of fuzzy matching chunks below.

```{r}
i_efectividad_raw <- read_csv(file.path(lsp_google_sheets_dir, "simec_i_efectividad_anp_2025.csv")) %>% 
  janitor::clean_names()

unique(i_efectividad_raw$nombre)
unique(lsps_missing_protected$protected_area_name)
```

```{r}
library(tidyverse)
library(here)
library(sf)
library(mapview)
library(stringdist)
library(stringi)

# normalize protected area names...title case, remove accents
normalize_name <- function(x) {
  x %>%
    stringi::stri_trans_general("Latin-ASCII") %>%  # remove accents
    str_to_title() %>%
    str_squish()  # remove extra whitespace left behind
}
```


### Fuzzy matching the protected area names

Rebuild uppercase normalizations, `amatch()` against SIMEC, capture nombre pairs for QA via printed distinct tables upstream.

```{r}
library(stringdist)
library(stringi)

# normalize protected area names...uppercase, remove accents, remove common filler words
normalize_name <- function(x) {
  x %>%
    stringi::stri_trans_general("Latin-ASCII") %>%  # remove accents
    str_to_upper() %>%
    str_remove_all("\\bPARQUE NACIONAL\\b|\\bPARQUE\\b|\\bNACIONAL\\b|\\bZONA MARINA\\b|\\bARCHIPIELAGO\\b") %>%
    str_squish()  # remove extra whitespace left behind
}

# make normalized versions for matching
i_efectividad_norm <- i_efectividad_raw %>%
  mutate(nombre_norm = normalize_name(nombre))

lsp_missing_sites_norm <- lsps_missing_protected %>%
  mutate(protected_area_norm = normalize_name(protected_area_name))

# fuzzy match: for each dive site protected area name, find the closest match in i_efectividad
matches <- stringdist::amatch(
  lsp_missing_sites_norm$protected_area_norm,
  i_efectividad_norm$nombre_norm,
  method = "jw", # Jaro-Winkler works well for place names
  maxDist = 0.2 # adjust threshold as needed; lower = stricter
)
```


```{r}
# attach matched nombre from i_efectividad back to dive sites
lsp_missing_matched <- lsp_missing_sites_norm %>%
  mutate(
    nombre_matched = i_efectividad_norm$nombre[matches],
    nombre_norm_matched = i_efectividad_norm$nombre_norm[matches]
  )
```


```{r}
# inspect the matches before joining
# join them!!!
lsp_missing_w_SIMEC <- lsp_missing_matched %>%
  left_join(
    i_efectividad_raw,
    by = c("nombre_matched" = "nombre")
  ) %>% 
  mutate(protected_area_condition = ifelse(!is.na(i_efectividad), i_efectividad, NA))

ramsar_management_plans <- read_csv(file.path(lsp_google_sheets_dir, "ramsar_site_management_plan_presence.csv"))

lsp_missing_RAMSAR <- lsp_missing_w_SIMEC %>% 
  filter(!is.na(protected_area_name) & is.na(nombre_norm_matched)) %>% 
  left_join(ramsar_management_plans, by = c("protected_area_name" = "wdpa_orig_name")) %>% 
  mutate(protected_area_condition = case_when(
    if_ramsar_has_management_plan == "y" ~ max(i_efectividad_raw$i_efectividad),
    if_ramsar_has_management_plan == "in_progress" ~ 60,
    if_ramsar_has_management_plan == "n" ~ min(i_efectividad_raw$i_efectividad),
    TRUE ~ NA
  )) %>% 
  dplyr::select(rgn_id, lsp_name, protected_area_condition, if_ramsar_has_management_plan, protected_area_name, in_protected_area, protected_area_norm)

lsp_missing_only_conanp_ANPs <- lsp_missing_w_SIMEC %>% 
  filter(!protected_area_name %in% lsp_missing_RAMSAR$protected_area_name)
```


```{r}
final_lsp_missing_condition <- bind_rows(lsp_missing_only_conanp_ANPs, lsp_missing_RAMSAR) %>% 
  group_by(rgn_id, lsp_name) %>% 
  summarize(protected_area_condition = mean(protected_area_condition, na.rm = TRUE)) %>% 
  ungroup() %>% 
  mutate(protected_area_condition = ifelse(protected_area_condition == "NaN", 0, protected_area_condition)) %>% 
  mutate(lsp_name_join = stri_trim_both(lsp_name)) 

length(unique(final_lsp_missing_condition$lsp_name)) # 37, only running the first line of the code chunk
length(final_lsp_missing_condition$lsp_name) # 37, great!

mapview(final_lsp_missing_condition, zcol = "protected_area_condition") + mapview(all_protected)
```

## Determine if those that are not within protected areas are still protected in some way

Anything still scoring zero after ANP/Ramsar passes joins `lsps_codified_protections_estimated_condition.csv`—those rows encode statutes, NOM citations, and literature-backed estimates keyed by trimmed `lsp_name`.

```{r}
lsps_other_protections <- final_lsp_missing_condition %>% 
  filter(protected_area_condition == 0) %>%
  mutate(lsp_name_join = stri_trim_both(lsp_name)) %>% 
  dplyr::select(-lsp_name)

mapview(lsps_other_protections)

# From: <https://docs.google.com/spreadsheets/d/1RwNTzRZdBIDtzi7d1ceFxhNmVghN47NxTgrhZV8O2SM/edit?gid=193789975#gid=193789975>
other_protections_condition_score <- read_csv(file.path(lsp_google_sheets_dir, "lsps_codified_protections_estimated_condition.csv")) %>% 
  mutate(lsp_name_join = stri_trim_both(lsp_name))  %>% 
  dplyr::select(-lsp_name)

lsps_other_protections_condition <- left_join(final_lsp_missing_condition, other_protections_condition_score)
```


# Condition per region

Collapse to means per `rgn_id`, respecting both SIMEC-derived overlaps and codified fill-ins.

```{r}
final_other_protections_condition <- lsps_other_protections_condition %>% 
  mutate(overall_condition = ifelse(protected_area_condition == 0, estimated_condition_of_protection, protected_area_condition),
         overall_condition = as.numeric(overall_condition)) %>% 
  st_drop_geometry()

rgn_final_other_protections_condition <- final_other_protections_condition %>% 
  group_by(rgn_id) %>% 
  summarize(
    condition = mean(overall_condition)
  ) %>% 
  ungroup()
```

## Write to Mazu

Emit place-level + regional CSVs consumed by script 8 when averaging categories.

```{r}
write_csv(final_other_protections_condition, here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/output/condition_lsp_other_marine_places.csv"))
write_csv(rgn_final_other_protections_condition, here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/output/condition_region_other_marine_places.csv"))
```


