---
title: 'OHI GoC 2026: SP ICO Gulf of California — expert-final iconic list'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# Summary

The Iconic Species (ICO) subgoal of Sense of Place captures the condition of iconic species that are relevant to local cultural identity.  We must (1) create a list of iconic species and which OHI regions they belong to, (2) determine their condition by their risk status, such as from NOM-059-SEMARNAT-2010, or IUCN, and (3) then, for each region, the average of the status for all iconic species can be calculated.

(1) = script1_iconic_species_list.qmd
(2) = script2_norma059.qmd, script3_cites_prep.qmd, script4_iucn_prep.qmd
(3) = script5_ico_current_status.qmd

In this fourth script, we will be cleaning up IUCN information and assessing iconic species condition based on the IUCN Red List Categories and Criteria.

Note that if none of the iconic species are listed on NOM-059-SEMARNAT-2010, we will check to see if they are listed as *at risk* by CITES/IUCN.  If not, the subgoal would have a current status of 100 for that region and year.

The IUCN Red List divides species into nine categories: *Not Evaluated, Data Deficient, Least Concern, Near Threatened, Vulnerable, Endangered, Critically Endangered, Extinct in the Wild and Extinct.*

-   **Reference**: IUCN 2024. IUCN Red List of Threatened Species. Version 2025-1 \<www.iucnredlist.org\>
-   **Accessed**: July 15, 2025
-   **Download**: Download is done through iucnredlist API
-   **Native data resolution**: Country level (by country name)
-   **Time range**: 1965-2024 (discrete past assessments by species)
-   **Format**: JSON

The Status of this sub-goal (X~ICO~) is then the % of iconic species in each threat category (as defined by the IUCN Red List), such that:

$$X_{ICO} = \frac{\displaystyle\sum_{category}S_{cat}*w_{cat}}{\displaystyle\sum_{category}S_{cat}}$$

where for each IUCN threat category:

-   *S~cat~* is the number of assessed species in the category
-   *w~cat~* is the status weight assigned for that category (note, these are the inverse of the risk value used in the SPP calculations):
-   'LC' = 1.0, 'NT' = 0.8, 'VU' = 0.6, 'EN' = 0.4, 'CR' = 0.2, 'EX' = 0.0

For this September expert-final update, I am not re-pulling the IUCN API. The April run already produced `mexico_all_years_w_condition_clean` (saved as `iucn_condition_ico_spp_041326_only_ico.csv`), and nothing material changed in how those IUCN weights were assigned. I will reuse that April table and filter it to the Keep-only species–region pairs from script1.

```{r}
library(tidyverse)
library(here)
library(janitor)
```


## Reuse April IUCN conditions for the Keep list

```{r}
ico_spp_rgnid <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/ico/int", "ico_species_rgnid_093026.csv"))

mexico_all_years_w_condition_clean <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/ico/int", "iucn_condition_ico_spp_041326_only_ico.csv"))

# keep only taxa (and region pairs) that remain on the expert Keep list
mexico_all_years_w_condition_clean_keep <- mexico_all_years_w_condition_clean %>% 
  semi_join(ico_spp_rgnid, by = c("taxon_scientific_name", "rgn_id"))

length(unique(mexico_all_years_w_condition_clean$taxon_scientific_name))
length(unique(mexico_all_years_w_condition_clean_keep$taxon_scientific_name))

setdiff(
  unique(ico_spp_rgnid$taxon_scientific_name),
  unique(mexico_all_years_w_condition_clean_keep$taxon_scientific_name)
) # keep-list species with no IUCN condition row in the april file

write_csv(
  mexico_all_years_w_condition_clean_keep,
  here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/ico/int", "iucn_condition_ico_spp_093026_only_ico.csv")
)
```
