---
title: 'OHI GoC 2026: SP ICO Gulf of California — expert-final iconic list'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# Summary

The Iconic Species (ICO) subgoal of Sense of Place captures the condition of iconic species that are relevant to local cultural identity.  We must (1) create a list of iconic species and which OHI regions they belong to, (2) determine their condition by their risk status, such as from NOM-059-SEMARNAT-2010, or IUCN, and (3) then, for each region, the average of the status for all iconic species can be calculated.

(1) = script1_iconic_species_list.qmd
(2) = script2_norma059.qmd, script3_cites_prep.qmd, script4_iucn_prep.qmd
(3) = script5_ico_current_status.qmd

This folder (`v2026_expert_final_093026`) is a copy of `ICO/v2026`. I updated it to use only the species the experts decided to keep (43 kept and 36 removed).

## Expert-final criteria (September 2026)

After several rounds of Expert Working Group discussion, and a clarifying email from the OHI Core Team asking for objections or corrections by September 30, we settled on the following criteria for what counts as iconic for the index. Iconic species must (1) have local cultural meaning (for example spiritual, religious, heritage, or deep aesthetic meaning) beyond being regularly caught and eaten as seafood, (2) be dependent on the health of the Gulf, meaning a marine species or a coastal obligate, and (3) be important to the general coastal community, not just biologists or conservationists.

We also reconsidered existence value as a criterion and decided against it, because every species has existence value and that alone does not justify an iconic designation. Applying the criteria consistently, we excluded several charismatic megafauna that are present and valued for tourism or conservation but are not clearly celebrated as local cultural icons by coastal communities. For shrimp, I kept only Pacific blue shrimp (*Penaeus stylirostris*), which seemed to have the strongest ties to regional identity, and I removed white, brown, and pink shrimp as primarily commercial or culinary species under criterion 1.

In this first script, I will be bringing in the Keep-only list and assigning each species to an OHI GoC Region (1-9) based on a species occurrence (presence and absence) database from Dr. Hem Nalini Morzaria-Luna.

```{r}
library(tidyverse)
library(here)
library(janitor)
```


## Read in list

I filtered the expert sheet to Keep only and saved that as `especies_iconicas_093026.csv` on the shared OHI drive. The full Keep/Remove sheet is archived there as `sp_ico_list_w_general_groups_093026.csv` for provenance.

```{r}
especies_iconicas <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/ico", "especies_iconicas_093026.csv")) %>%
  mutate(
    sci_nombre = str_replace_all(sci_nombre, "[\u200b\u200c\u200d\ufeff]", ""), # strip zero-width characters
    sci_nombre = str_squish(sci_nombre),
    general_nombre = str_squish(general_nombre)
  )

nrow(especies_iconicas) # expect 43
n_distinct(especies_iconicas$sci_nombre)
```


## Clean to prep for joining with risk condition 


```{r}
especies_iconicas_scinames <- especies_iconicas %>% 
  mutate(sci_nombre = str_replace_all(sci_nombre, "spp.",""),
         sci_nombre = str_trim(sci_nombre, side = "both"))

target_species <- unique(especies_iconicas_scinames$sci_nombre)

# making a regex pattern from the genus/species names
pattern <- paste(target_species, collapse = "|")
```


## Determine species occurence throughout the Gulf


```{r}
clean_species_occurrence <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/ico/clean_species_ocurrence_hem_2018.csv")) %>% 
  filter(wordnum %in% "2") %>% 
  dplyr::select(old_name, lat, lon, source) %>% 
  mutate(old_name = str_trim(old_name, side = "both"))
```


```{r}
ico_clean_species_occurrence <- clean_species_occurrence  %>%
  filter(str_detect(old_name, regex(pattern, ignore_case = TRUE)))

setdiff(unique(ico_clean_species_occurrence$old_name), target_species)

setdiff(target_species, unique(ico_clean_species_occurrence$old_name))
```

Many of the target species not in the occurrence database are the general genera, not the species name, so we can ignore those.  However, for species names, we need to look deeper at missing occurrence data.

Compared with the earlier full list, several former missing-occurrence cases (for example *Penaeus brevirostris*, *Ilochione subrugosa*, *Tetrapturus albidus*, *Endomychura craveri*, and *Lagenorynchus obliquidens*) were removed by the experts, so I no longer need occurrence fixes for those. I handle the remaining spelling and synonym fixes for kept species below.


```{r}
clean_string <- function(x) {
  x %>% 
    str_replace_all("[^[:ascii:]]", " ") %>%  # replace non-ascii with space
    str_replace_all("\\s+", " ") %>% # collapse multiple spaces
    str_trim(side = "both")
}

# re-clean both sides
target_species_clean <- clean_string(target_species)

ico_clean_species_occurrence <- clean_species_occurrence %>% 
  mutate(old_name = clean_string(old_name))

# now check again
setdiff(target_species_clean, unique(ico_clean_species_occurrence$old_name))
```


*ICO List --> old species name on occurrence df (kept species only):*

Chelonia agassizi --> missing, but could be synonymous as Chelonia mydas (since it is a subspecies of green turtle).  However, the NORMA lists them separately. But, they are listed with the same risk category on the NORMA, so for now, let's just use Chelonia mydas overall.
Larus heermani --> Larus heermanni (note the double "n"s)
Pandion haeliatus --> Pandion haliaetus (note different species name spelling)
Sphyrna / Mobula --> these are genus-level entries, and the occurrence data may use specific species within the genus

But I am realizing here that the occurrence data likely has the correct spelling, and my ICO list must be incorrect.  So let's update the iconic species list.


```{r}
clean_species_occurrence <- clean_species_occurrence %>%
  mutate(old_name = clean_string(old_name))

especies_iconicas_scinames <- especies_iconicas_scinames %>%
  mutate(sci_nombre = clean_string(sci_nombre)) %>%
  mutate(sci_nombre = case_when(
    sci_nombre == "Chelonia agassizi" ~ "Chelonia mydas",
    sci_nombre == "Larus heermani" ~ "Larus heermanni",
    sci_nombre == "Pandion haeliatus" ~ "Pandion haliaetus",
    TRUE ~ sci_nombre
  ))
```


Now, let us re-run the filtering using the pattern:

```{r}
target_species <- unique(especies_iconicas_scinames$sci_nombre)

# making a regex pattern from the genus/species names
pattern <- paste(target_species, collapse = "|")

ico_clean_species_occurrence <- clean_species_occurrence  %>%
  filter(str_detect(old_name, regex(pattern, ignore_case = TRUE)))

setdiff(unique(ico_clean_species_occurrence$old_name), target_species)

setdiff(target_species, unique(ico_clean_species_occurrence$old_name))
```

I expect any remaining misses to be mostly genus-level entries such as *Sphyrna* and *Mobula*, plus any true occurrence gaps for flora that were emphasized in the Keep list (for example *Distichlis* and *Pachycereus*).



## Assign OHI GoC Region ID

```{r}
library(sf)
library(mapview)

ico_clean_species_occurrence_sf <- ico_clean_species_occurrence %>% 
  st_as_sf(coords = c("lon", "lat"), # making sf object
           crs = 4326, # wgs84
           remove = FALSE)

st_write(
  ico_clean_species_occurrence_sf,
  here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/ico/int", "ico_clean_species_occurrence_sf_093026.shp")
)
```

First, group by region and unite the polygons for inland 10km with the marine polygons.
```{r}
# intersect with inland_10
inland_10 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_10km_buffer","inland_10km_goc.shp"))
regions_marine <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp")) %>%
  filter(location == "marine")

st_crs(inland_10) == st_crs(regions_marine) # true

marine_and_inland10 <- bind_rows(regions_marine, inland_10) %>% 
  dplyr::select(rgn_id, geometry) %>% 
  group_by(rgn_id) %>% 
  summarize(
    geometry = st_union(geometry)
  )

mapview(marine_and_inland10)
```

Now, we can intersect with the `marine_and_inland10` so that each species occurrence is assigned an OHI GoC Region ID.
```{r}
# intersect with goc
ico_clean_species_occurrence_goc_rgnid <- st_intersection(st_transform(ico_clean_species_occurrence_sf, st_crs(marine_and_inland10)), marine_and_inland10)

mapview(ico_clean_species_occurrence_goc_rgnid, zcol = "rgn_id")

st_write(
  ico_clean_species_occurrence_goc_rgnid,
  here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/ico/int", "ico_clean_species_occurrence_goc_rgnid_093026.shp")
)
```


We can group by rgn_id and determine how many times a species was found in that region.
```{r}
ico_species_occurrence_total_per_rgn <- ico_clean_species_occurrence_goc_rgnid %>% 
  group_by(rgn_id, old_name) %>% 
  summarize(
    n_occurrence = n()
  ) %>% 
  ungroup()
```

If we do not care about frequency of occurrence:
```{r}
ico_species_rgnid <- ico_species_occurrence_total_per_rgn %>% 
  st_drop_geometry() %>% 
  dplyr::select(taxon_scientific_name = old_name, rgn_id)

write_csv(
  ico_species_rgnid,
  here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/ico/int", "ico_species_rgnid_093026.csv")
)
```
