---
title: "Population data"
output: html_document
date: "2025-06-24"
---

## NOTE: I started a script that estimates population at the Manzana scale for urban populations. Rural populatiosn needed to be estimated at the Local scale.

I decided to go the Local route because it was more straightforward, and I didn't think the data was going to be worse. However, if we need higher resolution, the manzana approach is available (with a bit more work to complete the script).

Regions data downloaded from here (2020 version of regions used): https://en.www.inegi.org.mx/app/biblioteca/ficha.html?upc=889463807469
```{r setup, include=FALSE}

library(sf)
library(terra)
library(tidyverse)
library(dplyr)
library(stringr)
library(here)
library(mapview)

# Function to covert lat/long into decimal degrees

dms_to_dd <- function(x, is_lon = FALSE) {
  parse_one <- function(s) {
    if (is.na(s) || !nzchar(s)) return(NA_real_)
    s <- str_trim(s)
    # normalize quotes and strip letters (N/S/E/W) but keep digits and separators
    s <- str_replace_all(s, "[“”\"]", "\"")
    s <- str_replace_all(s, "[’`′']", "'")
    s <- str_replace_all(s, "[NnSsEeWw]", "")
    # try D M S
    m <- str_match(s, "^(\\d{1,3})\\D+(\\d{1,2})\\D+(\\d{1,2}(?:\\.\\d+)?)")
    if (!any(is.na(m))) {
      deg <- as.numeric(m[,2]); min <- as.numeric(m[,3]); sec <- as.numeric(m[,4])
    } else {
      # try D M only
      m2 <- str_match(s, "^(\\d{1,3})\\D+(\\d{1,2})")
      if (any(is.na(m2))) return(NA_real_)
      deg <- as.numeric(m2[,2]); min <- as.numeric(m2[,3]); sec <- 0
    }
    deg + min/60 + sec/3600
  }
  out <- vapply(x, parse_one, numeric(1))
  # Longitudes in Mexico are West → negative. If any already negative, keep it negative.
  if (is_lon) out <- -abs(out)
  out
}


```


# Making base raster 
(although maybe it would be better to use our other raster)
```{r}

#zip_dir <- "/home/shares/ohi/OHI_GOC/_raw_data/boundaries_inegi/794551132173_s"

# List all .zip files
#zip_files <- list.files(zip_dir, pattern = "\\.zip$", full.names = TRUE)

## Unzip each file into the same folder
#for (z in zip_files) {
#  unzip(z, exdir = zip_dir)
#}

ent02 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/02ent.shp")
ent03 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/03ent.shp")
ent26 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/26ent.shp")
ent25 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/25ent.shp")
ent18 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/18ent.shp")
ent14 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/14ent.shp")
plot(st_geometry(ent02)) 
plot(st_geometry(ent03))  
plot(st_geometry(ent26))  
plot(st_geometry(ent25)) 
plot(st_geometry(ent18)) 
plot(st_geometry(ent14)) 
states <- bind_rows(ent02, ent03, ent26, ent25, ent18, ent14)
states <- st_transform(states, crs=6372)
plot(st_geometry(states), add=TRUE)
bbox_coords_6372 <- c(
  xmin = 1050000,  # western edge (near Baja California)
  ymin = 900000,   # southern edge (mid-Jalisco)
  xmax = 2300000,  # eastern edge (into Sonora/Sinaloa)
  ymax = 2380000   # northern edge (toward US border)
)
# Convert to bbox and then to polygon
gulf_bbox <- st_bbox(bbox_coords_6372, crs = st_crs(6372)) |> st_as_sfc()
plot(gulf_bbox, add=TRUE)

gulf <- st_crop(states, gulf_bbox)
plot(gulf_bbox)
plot(st_geometry(gulf), add=TRUE)

gulf_bbox_vect <- vect(gulf_bbox)  # gulf_bbox must be in EPSG:6372

# Now use that with terra::ext()
r_template <- rast(
  extent = ext(gulf_bbox_vect),
  resolution = 1000,         # 1 km
  crs = crs(gulf_bbox_vect)  # or "EPSG:6372"
)
```

## Obtain location population data
Spatial boundary data originally downloaded from here:
https://en.www.inegi.org.mx/app/biblioteca/ficha.html?upc=889463807469 these are the 2020 data.

Population data:
https://www.inegi.org.mx/app/descarga/ficha.html?tit=326108&ag=0&f=csv

I confirmed that the data summed across locations equals the total reported for the state.

```{r}
tmp <- read_csv("/home/shares/ohi/OHI_GOC/_raw_data/census_inegi/iter_location/iter_00_cpv2020/conjunto_de_datos/conjunto_de_datos_iter_00CSV20.csv") %>%
  filter(ENTIDAD == "02") %>%
  filter(NOM_LOC != "Localidades de una vivienda") %>%
  filter(NOM_LOC != "Localidades de dos viviendas") %>%
  filter(as.numeric(as.character(LOC)) > 0)

sum(tmp$POBTOT)

```

```{r}


loc_file <- rbind(
  st_read("/home/shares/ohi/OHI_GOC/_raw_data/boundaries_inegi/census_2020/02_bajacalifornia/conjunto_de_datos/02l.shp"),
  st_read("/home/shares/ohi/OHI_GOC/_raw_data/boundaries_inegi/census_2020/03_bajacaliforniasur/conjunto_de_datos/03l.shp"),
  st_read("/home/shares/ohi/OHI_GOC/_raw_data/boundaries_inegi/census_2020/14_jalisco/conjunto_de_datos/14l.shp"),
  st_read("/home/shares/ohi/OHI_GOC/_raw_data/boundaries_inegi/census_2020/18_nayarit/conjunto_de_datos/18l.shp"),
  st_read("/home/shares/ohi/OHI_GOC/_raw_data/boundaries_inegi/census_2020/25_sinaloa/conjunto_de_datos/25l.shp"),
  st_read("/home/shares/ohi/OHI_GOC/_raw_data/boundaries_inegi/census_2020/26_sonora/conjunto_de_datos/26l.shp"))

pop_data <- "/home/shares/ohi/OHI_GOC/_raw_data/census_inegi/iter_location/iter_00_cpv2020/conjunto_de_datos/conjunto_de_datos_iter_00CSV20.csv"

loc <- loc_file %>%
  mutate(id = paste(CVE_ENT, CVE_MUN, CVE_LOC)) %>%
  arrange(CVE_MUN, CVE_LOC) 
#mapview(filter(loc, AMBITO=="Rural")) + mapview(filter(loc, AMBITO=="Urbana"), col.region="red")
#mapview(loc)

loc_02 <- loc %>% filter(CVE_ENT %in% "02")
mapview(loc_02)

pop <- read_csv(pop_data) %>%
  filter(!str_detect(NOM_LOC, "Total")) %>%
  mutate(id = paste(ENTIDAD, MUN, LOC)) %>%
  filter(ENTIDAD %in% c("02", "03", "14", "18", "25", "26")) %>%
  filter(NOM_LOC != "Localidades de una vivienda") %>%
  filter(NOM_LOC != "Localidades de dos viviendas") %>%
  filter(as.numeric(as.character(LOC)) > 0)

missing_loc <- setdiff(pop$id, loc$id)
missing_pop_loc <- filter(pop, id %in% missing_loc)
#sum(missing_pop_loc$POBTOT)/sum(pop$POBTOT) # only about 2% not accounted for in local polygons

## Take care of the polygon data
pop_loc <- left_join(loc, pop, by="id") %>%
  select(CVE_ENT, CVE_MUN, CVE_LOC, AMBITO, NOM_ENT, NOM_MUN, id, POBTOT) %>%
  filter(!is.na(POBTOT)) %>%
   mutate(
    area_km2  = st_area(geometry)/1e6) %>%
  mutate(POBTOT_km2 = POBTOT/area_km2)
## Check
#sum(pop_loc$POBTOT)
#sum(pop_loc$POBTOT_km2 * pop_loc$area_km2)
## summary(pop_loc$area_km2)
## quantile(pop_loc$area_km2, c(0.99, 0.999))

pop_loc_02 <- pop_loc %>% filter(CVE_ENT %in% "02")
mapview(pop_loc_02)

#st_write(pop_loc, "/home/shares/ohi/OHI_GOC/goal_prep/pressures/population_localities/inegi_boundaries_pop.shp")

pop_loc_vect <- vect(pop_loc)
pop_rast <- rasterize(
  x = pop_loc_vect,
  y = r_template, # 1km template raster
  field = "POBTOT_km2",
  fun = "sum",
  background = 0
)
##global(pop_rast, "sum", na.rm=TRUE) ## this drops because most of Jalisco is not included in the raster


## Take care of the data that did not align with a polygon
pop_pts <- missing_pop_loc %>%
  mutate(
    lon_dd = dms_to_dd(LONGITUD, is_lon = TRUE),
    lat_dd = dms_to_dd(LATITUD,  is_lon = FALSE)
  ) %>%
  # (optional) drop rows with missing/invalid coords
  filter(!is.na(lon_dd), !is.na(lat_dd)) %>%
  st_as_sf(coords = c("lon_dd", "lat_dd"), crs = 4326)
#sum(missing_pop_loc$POBTOT)

#mapview(pop_pts)

pop_pts <- st_transform(pop_pts, "EPSG:6372")  # MEXICO_ITRF_2008_LCC

pop_pts_vect <- vect(pop_pts)
pop_pts_rast <- rasterize(
  x = pop_pts_vect,
  y = r_template, # 1km template raster
  field = "POBTOT",
  fun = "sum",
  background = 0
)
#global(pop_pts_rast, "sum", na.rm=TRUE)
# mapview(pop_pts_rast)

final_raster <- pop_pts_rast + pop_rast
writeRaster(final_raster, "/home/shares/ohi/OHI_GOC/goal_prep/pressures/population_rasters/pop_2020.tif")
final_cap <- ifel(final_raster>10000, 10000, final_raster)
final_cap <- ifel(final_cap==0, NA, final_cap)

final_cap_sinaloa <- crop(final_cap, ent25)

#mapview(final_cap, maxpixels=1.85e+08)
# global(final_raster, "sum", na.rm=TRUE)
pts <- as.data.frame(final_cap, xy = TRUE, na.rm = TRUE)
cols <- grDevices::colorRampPalette(c("red","orange","yellow","green","cyan","blue","purple"))(256)
scale_color_gradientn(colors = cols)

ggplot(pts, aes(x = x, y = y, color = sum)) +
  geom_point(size = 0.01, alpha = 0.1) +
  geom_sf(data = gulf, fill = NA, color = "black", linewidth = 0.4,
          inherit.aes = FALSE) +      # <— key line
  coord_sf(datum = NA) +              # works well with sf overlays
  scale_color_gradientn(colors = rev(cols)) +
  labs(color = "people/km2") +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank())

```

