---
title: "Population data"
output: html_document
date: "2025-06-24"
---


Regions data downloaded from here (2020 version of regions used): https://en.www.inegi.org.mx/app/biblioteca/ficha.html?upc=889463807469
```{r setup, include=FALSE}
library(sf)
library(terra)
library(tidyverse)
library(dplyr)
library(stringr)
library(here)
library(mapview)

```

# map of Mexico states
```{r}

zip_dir <- "/home/shares/ohi/OHI_GOC/_raw_data/boundaries_inegi/794551132173_s"

# List all .zip files
zip_files <- list.files(zip_dir, pattern = "\\.zip$", full.names = TRUE)

# Unzip each file into the same folder
for (z in zip_files) {
  unzip(z, exdir = zip_dir)
}

ent02 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/02ent.shp")
ent03 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/03ent.shp")
ent26 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/26ent.shp")
ent25 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/25ent.shp")
ent18 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/18ent.shp")
ent14 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/14ent.shp")
plot(st_geometry(ent02)) 
plot(st_geometry(ent03))  
plot(st_geometry(ent26))  
plot(st_geometry(ent25)) 
plot(st_geometry(ent18)) 
plot(st_geometry(ent14)) 
states <- bind_rows(ent02, ent03, ent26, ent25, ent18, ent14)
states <- st_transform(states, crs=6372)
plot(st_geometry(states))
bbox_coords_6372 <- c(
  xmin = 1050000,  # western edge (near Baja California)
  ymin = 900000,   # southern edge (mid-Jalisco)
  xmax = 2300000,  # eastern edge (into Sonora/Sinaloa)
  ymax = 2380000   # northern edge (toward US border)
)

# Convert to bbox and then to polygon
gulf_bbox <- st_bbox(bbox_coords_6372, crs = st_crs(6372)) |> st_as_sfc()

gulf <- st_crop(states, gulf_bbox)
plot(gulf_bbox)
plot(st_geometry(gulf), add=TRUE)

gulf_bbox_vect <- vect(gulf_bbox)  # gulf_bbox must be in EPSG:6372

# Now use that with terra::ext()
r_template <- rast(
  extent = ext(gulf_bbox_vect),
  resolution = 1000,         # 1 km
  crs = crs(gulf_bbox_vect)  # or "EPSG:6372"
)

gulf_rast <- rasterize(states, r_template, field = "CVEGEO", background = 0)

coast <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_coastline/conjunto_de_datos/lineacosta_mex_2023.shp")



mapview(states) + mapview(coast)
## failed try again later
url <- "https://en.www.inegi.org.mx/contenidos/productos/prod_serv/contenidos/espanol/bvinegi/productos/geografia/cart_geo_hist_mex/794551131954_s.zip"
destfile <- "/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/geoestadistico/794551131954_s.zip" 

# Download the file
download.file(url, destfile, mode = "wb")
unzip(zipfile = destfile, exdir = "/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/geoestadistico/")

library(mxmaps)


data("df_mxstate_2020")
df_mxstate_2020$value <- df_mxstate_2020$pop
mxstate_choropleth(df_mxstate_2020,
                    title = "Total population, by state") 
st_read("data-raw/simpl/entidades.shp", "entidades")
```


Get the boundaries: Will need to get boundaries for rural areas as well.
https://en.www.inegi.org.mx/app/biblioteca/ficha.html?upc=889463807469 these are teh 2020 data
```{r}

manz_prep <- function(manz_file){ # manz_file <- "/home/shares/ohi/OHI_GOC/_raw_data/boundaries_inegi/census_2020/02_bajacalifornia/conjunto_de_datos/02m.shp"

manz <- st_read(manz_file) %>%
  mutate(id = paste(CVE_ENT, CVE_MUN, CVE_AGEB, CVE_MZA)) %>%
  mutate(id_loc = paste(CVE_ENT, CVE_MUN, CVE_LOC)) %>%
  arrange(CVE_MUN, CVE_LOC)
plot(st_geometry(filter(manz, AMBITO=="Rural")), border="red", add=TRUE)
plot(st_geometry(filter(manz, AMBITO=="Urbana")), border="blue", add=TRUE)

# check the area, 99.9% are < 0.5 km2, so reasonable to convert points to 1degree raster
manz_area <- st_area(manz) / 1e6
hist(manz_area)
summary(manz_area)

# getting the points:
manz_point <- st_centroid(manz)
manz_point}

manz_points <- c(manz_prep("/home/shares/ohi/OHI_GOC/_raw_data/boundaries_inegi/census_2020/02_bajacalifornia/conjunto_de_datos/02m.shp"),
                 manz_prep(""))


```

Get the population from 2020 census:
This only includes the urban and rural neighborhoods, so only includes about 3.5 million people.
(3.77 million is the actual total in Baja California, for example)

https://www.inegi.org.mx/programas/ccpv/2020/#microdatos

merge with points data
``` {r}

pop_manz <- read_csv("/home/shares/ohi/OHI_GOC/_raw_data/census_inegi/RESAGEBURB_02CSV20.csv") %>%
  filter(!str_detect(NOM_LOC, "Total")) %>%
  mutate(id = paste(ENTIDAD, MUN, AGEB, MZA))

## check to see that we are accounting for most of the population
mun_pop <- pop_manz %>%
  group_by(MUN) %>%
  summarize(pop = sum(POBTOT, na.rm=TRUE))

missing_pop <- setdiff(pop_manz$id, manz_point$id) # very few! which is good. losing 0.2% of people (N=7318)
missing_block <- setdiff(manz_point$id, pop$id)
missing_pop <- filter(pop, id %in% missing_pop)
sum(missing_pop$POBTOT)/
sum(pop$POBTOT) 

manz_pop_points <- left_join(manz_point, pop_manz, by="id") %>%
  select(CVE_ENT, CVE_MUN, CVE_LOC, CVE_AGEB, CVE_MZA, AMBITO, TIPOMZA, NOM_ENT, NOM_MUN,id, id_loc, POBTOT) %>%
  filter(!is.na(POBTOT))


ggplot(data = manz_pop_points) +
  geom_sf(aes(color = log(POBTOT)), size = 0.3) +
  scale_color_viridis_c(option = "plasma", trans = "sqrt") +  # good for skewed population data
  theme_minimal() +
  labs(title = "Population by Point",
       color = "Total Population")

```

Convert to a raster
```{r}
manz_points_vect <- vect(manz_pop_points)

# 2. Rasterize: sum POBTOT in each cell
manz_pop_raster <- rasterize(
  x = manz_points_vect,
  y = r_template,          # 1-km raster template
  field = "POBTOT",        # field to sum
  fun = "sum",             # sum values in each cell
  background = NA          # cells with no points = NA
)

plot(log(manz_pop_raster))
```


Explore data
```{r}

mapview(states)+ mapview(ohi) + mapview(coast) + mapview(goc_ocean)
ohi <- st_read("/home/shares/ohi/OHI_GOC/spatial/OHI_regions/marine/ohi_goc_final_071025.shp")
goc_ocean <- st_read(here("spatial/GoC_polygon_ecoregion_eqarea.shp"))

if (!require(devtools)) {
    install.packages("devtools")
}
devtools::install_github('diegovalle/mxmaps')
library(mxmaps)

data("df_mxmunicipio_2020")
knitr::kable(head(df_mxmunicipio_2020))

df_mxmunicipio_2020$value <-  df_mxmunicipio_2020$pop 
mxmunicipio_choropleth(df_mxmunicipio_2020, num_colors = 1,
                       title = "Population")


## ITER DATA

tmp <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/02ent.shp") 
plot(st_geometry(tmp))

tmp <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/02l.shp") %>% arrange(CVE_MUN, CVE_LOC)
plot(st_geometry(filter(tmp, CVE_LOC %in% c("0001", "0002", "0003", "0004", "0005", "0006"))))
tmp <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/02lpr.shp") %>% arrange(CVE_MUN, CVE_LOC) %>%
  mutate(id = paste(CVE_ENT, CVE_MUN, CVE_AGEB, CVE_MZA)) %>%
  mutate(id_loc = paste(CVE_ENT, CVE_MUN, CVE_LOC))

setdiff(tmp$id_loc, manz$id_loc)
setdiff(manz$id_loc, tmp$id_loc)
plot(st_geometry(tmp), color="orange", add=TRUE)
filter(tmp, CVE_MUN=="002")
```