---
title: "Population data"
output: html_document
date: "2025-06-24"
---

## NOTE Manzanas are rural and urban, but these data only include urban populations (with a few exceptions.)
Regions data downloaded from here (2020 version of regions used): https://en.www.inegi.org.mx/app/biblioteca/ficha.html?upc=889463807469
```{r setup, include=FALSE}

library(sf)
library(terra)
library(tidyverse)
library(dplyr)
library(stringr)
library(here)
library(mapview)

```


# Making base raster (although maybe it would be better to use our other raster)
```{r}

#zip_dir <- "/home/shares/ohi/OHI_GOC/_raw_data/boundaries_inegi/794551132173_s"

# List all .zip files
#zip_files <- list.files(zip_dir, pattern = "\\.zip$", full.names = TRUE)

## Unzip each file into the same folder
#for (z in zip_files) {
#  unzip(z, exdir = zip_dir)
#}

ent02 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/02ent.shp")
ent03 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/03ent.shp")
ent26 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/26ent.shp")
ent25 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/25ent.shp")
ent18 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/18ent.shp")
ent14 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/14ent.shp")
plot(st_geometry(ent02)) 
plot(st_geometry(ent03))  
plot(st_geometry(ent26))  
plot(st_geometry(ent25)) 
plot(st_geometry(ent18)) 
plot(st_geometry(ent14)) 
states <- bind_rows(ent02, ent03, ent26, ent25, ent18, ent14)
states <- st_transform(states, crs=6372)
plot(st_geometry(states))
bbox_coords_6372 <- c(
  xmin = 1050000,  # western edge (near Baja California)
  ymin = 900000,   # southern edge (mid-Jalisco)
  xmax = 2300000,  # eastern edge (into Sonora/Sinaloa)
  ymax = 2380000   # northern edge (toward US border)
)
# Convert to bbox and then to polygon
gulf_bbox <- st_bbox(bbox_coords_6372, crs = st_crs(6372)) |> st_as_sfc()
plot(gulf_bbox, add=TRUE)

gulf <- st_crop(states, gulf_bbox)
plot(gulf_bbox)
plot(st_geometry(gulf), add=TRUE)

gulf_bbox_vect <- vect(gulf_bbox)  # gulf_bbox must be in EPSG:6372

# Now use that with terra::ext()
r_template <- rast(
  extent = ext(gulf_bbox_vect),
  resolution = 1000,         # 1 km
  crs = crs(gulf_bbox_vect)  # or "EPSG:6372"
)
```

## Convert manz population data to raster
Spatial boundary data originally downloaded from here:
https://en.www.inegi.org.mx/app/biblioteca/ficha.html?upc=889463807469 these are teh 2020 data.

Population data:
https://www.inegi.org.mx/programas/ccpv/2020/#microdatos

This takes care of the majority of people, but not all. Based on the 2020 census data, the manzanas include about 3.5 million people in Baja, whereas the entire Baja California population is 3.77 million. We will take care of the remainder of the population in the next step.


```{r}

manz_raster <- function(manz_file, pop_data, r_template){ # 
  
#manz_file <- "/home/shares/ohi/OHI_GOC/_raw_data/boundaries_inegi/census_2020/02_bajacalifornia/conjunto_de_datos/02m.shp"
#pop_data <- "/home/shares/ohi/OHI_GOC/_raw_data/census_inegi/RESAGEBURB_02CSV20.csv"

manz <- st_read(manz_file) %>%
  mutate(id = paste(CVE_ENT, CVE_MUN, CVE_AGEB, CVE_MZA)) %>%
  mutate(id_loc = paste(CVE_ENT, CVE_MUN, CVE_LOC)) %>%
  arrange(CVE_MUN, CVE_LOC)
#mapview(filter(manz, AMBITO=="Rural")) + mapview(filter(manz, AMBITO=="Urbana"), col.region="red")

# check the area, 99.9% are < 0.5 km2, so reasonable to convert points to 1km raster
#manz_area <- st_area(manz) / 1e6
#hist(manz_area)
#summary(manz_area)

# getting the points:
manz_point <- st_centroid(manz)

pop_manz <- read_csv(pop_data) %>%
  filter(!str_detect(NOM_LOC, "Total")) %>%
  mutate(id = paste(ENTIDAD, MUN, AGEB, MZA))
#pop_ageb <- read_csv(pop_data) %>%
#  filter(str_detect(NOM_LOC, "Total")) %>%
#  filter(AGEB!="0000") %>%
#  mutate(id = paste(ENTIDAD, MUN, AGEB))

manz_pop_points <- left_join(manz_point, pop_manz, by="id") %>%
  select(CVE_ENT, CVE_MUN, CVE_LOC, CVE_AGEB, CVE_MZA, AMBITO, TIPOMZA, NOM_ENT, NOM_MUN,id, id_loc, POBTOT) 
%>%
  filter(!is.na(POBTOT))

#ggplot(data = manz_pop_points) +
#  geom_sf(aes(color = log(POBTOT)), size = 0.3) +
#  scale_color_viridis_c(option = "plasma", trans = "sqrt") +  # good for skewed #population data
#  theme_minimal() +
#  labs(title = "Population by Point",
#       color = "Total Population")

manz_points_vect <- vect(manz_pop_points)

# 2. Rasterize: sum POBTOT in each cell
manz_pop_raster <- rasterize(
  x = manz_points_vect,
  y = r_template,          # 1-km raster template
  field = "POBTOT",        # field to sum
  fun = "sum",             # sum values in each cell
  background = NA          # cells with no points = NA
)

#plot(log(manz_pop_raster+1))
#global(manz_pop_raster, "sum", na.rm=TRUE)
manz_pop_raster
}



```

## Population of rest of areas
For this, I remove the manz population data and then distribute at the remainder of AGEB polygons.

``` {r}

local_pr <- st_read(list.files("/home/shares/ohi/OHI_GOC/_raw_data/boundaries_inegi/census_2020/", pattern = sprintf("%slpr.shp", st_id), recursive=TRUE, full=TRUE))
local <- st_read(list.files("/home/shares/ohi/OHI_GOC/_raw_data/boundaries_inegi/census_2020/", pattern = sprintf("%sl.shp", st_id), recursive=TRUE, full=TRUE))
mapview(local, col.regions="red") +  mapview(manz_shp, col.regions="blue")

ageb_raster <- function(st_id="02", r_template){ # 
  
  ## Get data
manz_shp <- st_read(list.files("/home/shares/ohi/OHI_GOC/_raw_data/boundaries_inegi/census_2020/", pattern = sprintf("%sm.shp", st_id), recursive=TRUE, full=TRUE)) 

ageb_shp_rural <- st_read(list.files("/home/shares/ohi/OHI_GOC/_raw_data/boundaries_inegi/census_2020/", pattern = sprintf("%sar.shp", st_id), recursive=TRUE, full=TRUE))
ageb_shp_urban <- select(
         st_read(list.files("/home/shares/ohi/OHI_GOC/_raw_data/boundaries_inegi/census_2020/", pattern = sprintf("%sa.shp", st_id), recursive=TRUE, full=TRUE)), -"CVE_LOC")

ageb_shp <- rbind(ageb_shp_rural, ageb_shp_urban)

pop_data <- read_csv(sprintf("/home/shares/ohi/OHI_GOC/_raw_data/census_inegi/RESAGEBURB_%sCSV20.csv", st_id))
 
## POPULATION DATA:  
   
# get manz regions:  
manz <- manz_shp %>%
  mutate(id = paste(CVE_ENT, CVE_MUN, CVE_AGEB, CVE_MZA)) %>%
  arrange(CVE_MUN, CVE_LOC)

# remove the manz region population and summarize by AGEB
pop_ageb_nomanz <- pop_data %>%
  filter(!str_detect(NOM_LOC, "Total")) %>%
  mutate(id = paste(ENTIDAD, MUN, AGEB, MZA)) %>%
  filter(!(id %in% manz$id)) %>% 
  group_by(ENTIDAD, MUN, AGEB) %>%
  summarize(POBTOT_noMZA = sum(POBTOT, na.rm=TRUE)) %>%
  ungroup() %>%
  select(CVE_AGEB = AGEB, POBTOT_noMZA)

## REGIONS
ageb_fix <- st_make_valid(ageb_shp)
manz_fix <- st_make_valid(manz_shp)
manz_u <- st_union(manz_fix)
ageb_minus_manz <- st_difference(ageb_fix, manz_u)

#drop empty geometries (AGEBs fully covered by MANZ)
ageb <- ageb_minus_manz[!st_is_empty(ageb_minus_manz), ]

ageb <- ageb %>%
  left_join(pop_ageb_nomanz, by = "CVE_AGEB") %>%
  mutate(POBTOT_noMZA = ifelse(is.na(POBTOT_noMZA), 0, POBTOT_noMZA)) %>%
  mutate(
    area_km2  = drop_units(st_area(geometry))/1e6) %>%
  mutate(POBTOT_km2 = POBTOT_noMZA/area_km2)

mapview(ageb, zcol= "POBTOT_km2")
```

Convert to a raster
```{r}
manz_points_vect <- vect(manz_pop_points)

# 2. Rasterize: sum POBTOT in each cell
manz_pop_raster <- rasterize(
  x = manz_points_vect,
  y = r_template,          # 1-km raster template
  field = "POBTOT",        # field to sum
  fun = "sum",             # sum values in each cell
  background = NA          # cells with no points = NA
)

plot(log(manz_pop_raster+1))
```


Explore data
```{r}
## rural AGEB
rural <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/boundaries_inegi/census_2020/02_bajacalifornia/conjunto_de_datos/02ar.shp")
mapview(rural)
## urban AGEB
urban <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/boundaries_inegi/census_2020/02_bajacalifornia/conjunto_de_datos/02a.shp")
mapview(rural) + mapview(urban, col.region="red")


setdiff(pop_ageb$AGEB, tmp$CVE_AGEB)

tmp <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/boundaries_inegi/census_2020/02_bajacalifornia/conjunto_de_datos/02ti.shp")
mapview(tmp)

tmp <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/boundaries_inegi/census_2020/02_bajacalifornia/conjunto_de_datos/02l.shp")
mapview(tmp)

mapview(states)+ mapview(ohi) + mapview(coast) + mapview(goc_ocean)
ohi <- st_read("/home/shares/ohi/OHI_GOC/spatial/OHI_regions/marine/ohi_goc_final_071025.shp")

goc_ocean <- st_read(here("spatial/GoC_polygon_ecoregion_eqarea.shp"))

if (!require(devtools)) {
    install.packages("devtools")
}
devtools::install_github('diegovalle/mxmaps')
library(mxmaps)

data("df_mxmunicipio_2020")
knitr::kable(head(df_mxmunicipio_2020))

df_mxmunicipio_2020$value <-  df_mxmunicipio_2020$pop 
mxmunicipio_choropleth(df_mxmunicipio_2020, num_colors = 1,
                       title = "Population")


## ITER DATA

tmp <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/02mun.shp") 
mapview(tmp)
plot(st_geometry(tmp))

tmp <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/02l.shp") %>% arrange(CVE_MUN, CVE_LOC)
plot(st_geometry(filter(tmp, CVE_LOC %in% c("0001", "0002", "0003", "0004", "0005", "0006"))))
tmp <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/02lpr.shp") %>% arrange(CVE_MUN, CVE_LOC) %>%
  mutate(id = paste(CVE_ENT, CVE_MUN, CVE_AGEB, CVE_MZA)) %>%
  mutate(id_loc = paste(CVE_ENT, CVE_MUN, CVE_LOC))

setdiff(tmp$id_loc, manz$id_loc)
setdiff(manz$id_loc, tmp$id_loc)
plot(st_geometry(tmp), color="orange", add=TRUE)
filter(tmp, CVE_MUN=="002")
```



#Exploring state boundaries and data.
# Making base raster (although maybe it would be better to use our other raster)
```{r}

#zip_dir <- "/home/shares/ohi/OHI_GOC/_raw_data/boundaries_inegi/794551132173_s"

# List all .zip files
#zip_files <- list.files(zip_dir, pattern = "\\.zip$", full.names = TRUE)

## Unzip each file into the same folder
#for (z in zip_files) {
#  unzip(z, exdir = zip_dir)
#}

ent02 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/02ent.shp")
ent03 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/03ent.shp")
ent26 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/26ent.shp")
ent25 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/25ent.shp")
ent18 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/18ent.shp")
ent14 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/14ent.shp")
plot(st_geometry(ent02)) 
plot(st_geometry(ent03))  
plot(st_geometry(ent26))  
plot(st_geometry(ent25)) 
plot(st_geometry(ent18)) 
plot(st_geometry(ent14)) 
states <- bind_rows(ent02, ent03, ent26, ent25, ent18, ent14)
states <- st_transform(states, crs=6372)
plot(st_geometry(states))
bbox_coords_6372 <- c(
  xmin = 1050000,  # western edge (near Baja California)
  ymin = 900000,   # southern edge (mid-Jalisco)
  xmax = 2300000,  # eastern edge (into Sonora/Sinaloa)
  ymax = 2380000   # northern edge (toward US border)
)

# Convert to bbox and then to polygon
gulf_bbox <- st_bbox(bbox_coords_6372, crs = st_crs(6372)) |> st_as_sfc()

gulf <- st_crop(states, gulf_bbox)
plot(gulf_bbox)
plot(st_geometry(gulf), add=TRUE)

gulf_bbox_vect <- vect(gulf_bbox)  # gulf_bbox must be in EPSG:6372

# Now use that with terra::ext()
r_template <- rast(
  extent = ext(gulf_bbox_vect),
  resolution = 1000,         # 1 km
  crs = crs(gulf_bbox_vect)  # or "EPSG:6372"
)

gulf_rast <- rasterize(states, r_template, field = "CVEGEO", background = 0)

coast <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_coastline/conjunto_de_datos/lineacosta_mex_2023.shp")



mapview(states) + mapview(coast)

## failed try again later
url <- "https://en.www.inegi.org.mx/contenidos/productos/prod_serv/contenidos/espanol/bvinegi/productos/geografia/cart_geo_hist_mex/794551131954_s.zip"
destfile <- "/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/geoestadistico/794551131954_s.zip" 

# Download the file
download.file(url, destfile, mode = "wb")
unzip(zipfile = destfile, exdir = "/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/geoestadistico/")

library(mxmaps)


data("df_mxstate_2020")
df_mxstate_2020$value <- df_mxstate_2020$pop
mxstate_choropleth(df_mxstate_2020,
                    title = "Total population, by state") 
st_read("data-raw/simpl/entidades.shp", "entidades")
```

