---
title: 'OHI GoC 2025: Clean waters pharmaceuticals'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# CW: Pharmaceuticals - Livestock

Estimated pharmaceutical discharge.

**Setup: Bring in watersheds and pour points:**

```{r setup}
library(sf)
library(utils)
library(tidyverse)
library(here)
library(raster)
library(mapview)
library(purrr)
library(stringr) # fuzzy string matching
library(stringi)
library(janitor)
library(tibble)
```

```{r}
## lets use the full state for now and then summarize by the full watersheds.
goc_ohi_rgns <- st_read(here::here("spatial/ohi_regions","goc_ohi_rgns.shp"))

union_rgns <- goc_ohi_rgns %>% 
  st_union() %>% 
  st_make_valid()

# plot(goc_ohi_rgns)
# plot(union_rgns)

watershed_all <- st_read(file.path("/home/shares/ohi/stressors_2021/_dataprep/nutrients/watersheds_pourpoints/watersheds_all_4326.shp"))
st_crs(watershed_all)
# mapview(watershed_all)

watershed_goc <- st_intersection(st_transform(watershed_all, st_crs(union_rgns)), union_rgns)
mapview(watershed_goc)

# --- if we plan to change the buffer for some reason ---
# inland_50 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer","inland_50km_goc.shp"))
# watershed_goc_50km <- st_intersection(st_transform(watershed_all, st_crs(inland_50)), inland_50)
# mapview(watershed_goc_50km)
```

```{r}
# here's the associated pourpoints that should match with the watersheds:
pourpoints <- st_read("/home/shares/ohi/stressors_2021/_dataprep/nutrients/watersheds_pourpoints/pourpoints_mol.shp")

pourpoints_goc <- st_intersection(st_transform(pourpoints, st_crs(union_rgns)), union_rgns)
mapview(pourpoints_goc)

mapview(pourpoints_goc) + mapview(watershed_goc)
```

```{r ohi_marine_and_inland}
inland_50 <- st_read(here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer", "inland_50km_goc.shp"))
inland_10 <- st_read(here("/home/shares/ohi/OHI_GOC/spatial/inland_10km_buffer", "inland_10km_goc.shp"))
regions_marine <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp")) %>%
  filter(location == "marine")

marine_and_inland50 <- bind_rows(regions_marine, inland_50) %>%
  dplyr::select(rgn_id, geometry) %>%
  group_by(rgn_id) %>%
  summarize(geometry = st_union(geometry), .groups = "drop")

marine_and_inland10 <- bind_rows(regions_marine, inland_10) %>%
  dplyr::select(rgn_id, geometry) %>%
  group_by(rgn_id) %>%
  summarize(geometry = st_union(geometry), .groups = "drop")
```


# Estimating risk of pharmaceutical discharge based on the intensity of livestock farming in each OHI Region

I plan to, instead of using number of heads, use the total volume in tonnes (toneladas) slaughtered per year and region of all livestock animals (by separately determining the raster distribution of volume within each municipality for each livestock type, chickn cattle, goat, sheep and pig, and then once those are in raster form then aggregating for each year and cell so that we have the total yearly livestock slaughtered in tonnes, then convert to kg), then multiply by 11 mg/kg of antimicrobials given per kg of terrestrial livestock animal, which gives estimated mg of antimicrobials that could be present in each cell. THEN I will summarize by watershed polygons, and take that value to each pourpoint and assign them to OHI regions.

All my current code uses the number of heads from cattle to do the dasymetric mapping, and the FAO maps are livestock density in heads/km^2. We have 5 species, each with their own GLW4 raster, and SIAP provides the `nomespecie`. So the dasymetric step needs to happen **separately per species**, and then we will sum across species at the end. Here is how we can do this:

---

### Workflow

**For each species × year combination:**

1. Load the species-specific GLW4 raster (e.g. `CTL` for bovino, `CHK` for ave, etc.)
2. Crop/resample to 1km 
3. Compute the weight raster for that species (GLW cell density / municipal GLW total), which is the spatial proxy
4. Filter SIAP to that species × year, get `toneladas_volumen` per municipality
5. Rasterize the SIAP tonnes onto the municipality polygons
6. Multiply: `siap_volume_raster * weight_raster` --> tonnes of that species slaughtered in each 1km cell

**Then across species:**

7. Stack the 5 species rasters and sum cellwise, which = total tonnes slaughtered per 1km cell per year
8. Multiply by 1000 (tonnes --> kg) then by 11 mg/kg, to get mg antimicrobials per cell
9. Summarize by watershed, assign to pourpoints, assign to OHI regions
10. Calculate the risk factor by comparing across regions and years

_REMEMBER: GLW4 (FAO livestock density) = **where**, DGSIAP (Mexican Govt) = **how much**!!_


## Reading in all livestock reared from 2006-2024 (missing 2023 right now)

Data was downloaded from <https://nube.agricultura.gob.mx/datosAbiertos/Pecuario.php>: 

For each year, click on the associated link for the year, then bring it into Mazu at `~/home/shares/ohi/OHI_GOC/_raw_data/DGSIAP/d2026`.

```{r}
## write function to read in all of the files and limit them to GoC states

read_in_livestock <- function (year) {
  
  df <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/DGSIAP/d2026", paste0("cierre_", year, ".csv"))) %>% 
    
    janitor::clean_names() %>% 
    
    mutate(across(where(is.character), ~iconv(., from = "WINDOWS-1252", to = "UTF-8"))) %>%  # to address encoding issues
    
    filter(nomestado %in% c("Baja California Sur","Baja California","Jalisco","Sinaloa","Sonora","Nayarit")) 
  
  return(df)
  
}
```

Now, we can use the function to read in each year's WWTP data and limit to GoC states.

```{r}
# specify the years we want to iterate through. ideally, we would have 2023 in the future...
years_to_read_in <- c(2006:2022, 2024)

# use `lapply` and bind rows to obtain data over time.
list_of_all_dfs <- lapply(years_to_read_in, read_in_livestock)
  
# bind them across all years
bind_all_years_livestock <- bind_rows(list_of_all_dfs) %>% 
  mutate(cvempio = as.numeric(cvempio),
         cveestado = as.numeric(cveestado))

summary(bind_all_years_livestock)
unique(bind_all_years_livestock$nomestado)
unique(bind_all_years_livestock$cvempio)
colnames(bind_all_years_livestock)

unique(bind_all_years_livestock$nomespecie)
# [1] "Abeja"     "Ave"       "Bovino"    "Caprino"   "Ovino"    
# [6] "Porcino"   "Guajolote"

unique(bind_all_years_livestock$nomproducto)
# [1] "Miel"          "Carne"         "Huevo-plato"   "Ganado en pie" "Leche"        
# [6] "Cera"          "Huevo plato"   "Ganado en Pie"

unique(bind_all_years_livestock$cveestado)

```

## Aggregate all livestock slaughtered volume for each municipality and year

For each livestock type, we will have a separate gridded 10km livestock density map from FAO.

```{r}
# first, we need to isolate the product to tonnes volume of "Carne" of `nomespecies` = c("Ave","Bovino","Caprino","Ovino","Porcino").
sum_livestock_municipal <- bind_all_years_livestock %>% 
  filter(nomproducto %in% "Carne" & nomespecie %in% c("Ave","Bovino","Caprino","Ovino","Porcino")) %>% 
  group_by(anio, cveestado, nomestado, cvempio, nommunicipio, nomespecie) %>% 
  summarize(toneladas_volumen = sum(volumen, na.rm = TRUE),
            cabezas_asacrificados = sum(asacrificado, na.rm = TRUE)) %>% 
  # mutate(weight = toneladas_volumen / cabezas_asacrificados) %>% 
  ungroup()


```


## Join SIAP data with municipality polygons

```{r}
# read them in
municipio_codigo_sf <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/land_admin_boundaries", "mex_admin2_joined_w_municipio_codigos.shp")) %>% 
  mutate(cvempio = as.numeric(clv_mnc),
         cveestado = as.numeric(clv_std))

unique(municipio_codigo_sf$clv_mnc)
unique(municipio_codigo_sf$cveestado)

mapview(municipio_codigo_sf)
```

```{r}
siap_municipal <- inner_join(sum_livestock_municipal, municipio_codigo_sf) %>% 
  st_as_sf()

siap_municipal_2022 <- siap_municipal %>% 
  filter(anio %in% 2022 & nomespecie %in% "Bovino")

unique(siap_municipal_2022$nomespecie)

mapview(siap_municipal_2022, zcol = "toneladas_volumen")
```

```{r}
gridded_cattle_density_2020_10km <- terra::rast(here::here("/home/shares/ohi/OHI_GOC/_raw_data/FAO_livestock_density/d2026/GLW4-2020.D-DA.CTL.tif"))

plot(gridded_cattle_density_2020_10km)

st_crs(gridded_cattle_density_2020_10km) # WGS 84

freq(gridded_cattle_density_2020_10km)

gridded_cattle_density_2020_10km_goc <- terra::crop(terra::project(gridded_cattle_density_2020_10km, crs(marine_and_inland50)), marine_and_inland50)

crs(gridded_cattle_density_2020_10km_goc)

plot(gridded_cattle_density_2020_10km_goc)
```

### Bring in FAO livestock density rasters

This data is from FAO's gridded 10km raster of heads/km^2 of each livestock type.  The data was downloaded from <https://data.apps.fao.org/catalog//iso/9d1e149b-d63f-4213-978b-317a8eb42d02> and placed into `~/home/shares/ohi/OHI_GOC/_raw_data/FAO_livestock_density/d2026`:

chicken/ave = ~/home/shares/ohi/OHI_GOC/_raw_data/FAO_livestock_density/d2026/GLW4-2020.D-DA.CHK.tif
cattle/bovino = ~/home/shares/ohi/OHI_GOC/_raw_data/FAO_livestock_density/d2026/GLW4-2020.D-DA.CTL.tif
goat/caprino = ~/home/shares/ohi/OHI_GOC/_raw_data/FAO_livestock_density/d2026/GLW4-2020.D-DA.GTS.tif
pigs/porcino = ~/home/shares/ohi/OHI_GOC/_raw_data/FAO_livestock_density/d2026/GLW4-2020.D-DA.PGS.tif
sheep/ovino = ~/home/shares/ohi/OHI_GOC/_raw_data/FAO_livestock_density/d2026/GLW4-2020.D-DA.SHP.tif

```{r}
species_glw_map <- tribble(
  ~nomespecie, ~glw_file,
  "Bovino", "GLW4-2020.D-DA.CTL.tif", # cattle
  "Ave", "GLW4-2020.D-DA.CHK.tif", # bird, which we assume is chicken because turkeys were separate and those classified as Aves also produce eggs which are in the SIAP data
  "Caprino", "GLW4-2020.D-DA.GTS.tif", # goat
  "Porcino", "GLW4-2020.D-DA.PGS.tif", # pig
  "Ovino", "GLW4-2020.D-DA.SHP.tif" # sheep
)

glw_dir <- "/home/shares/ohi/OHI_GOC/_raw_data/FAO_livestock_density/d2026"
out_dir <- "/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/pharmaceuticals/livestock_int"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
```


### EXAMPLE Cattle density (global 10km raster)

For an example of how to do it for just cattle:

```{r}
library(terra)

# read in cattle density
gridded_cattle_density_2020_10km <- terra::rast(here::here("/home/shares/ohi/OHI_GOC/_raw_data/FAO_livestock_density/d2026/GLW4-2020.D-DA.CTL.tif"))

# temp project the 9 polygons to WGS 84 just for the fast global crop
marine_and_inland50_wgs84 <- st_transform(marine_and_inland50, terra::crs(gridded_cattle_density_2020_10km))

# crop the massive global raster down to marine_and_inland50_wgs84 rectangular box
cattle_cropped_wgs84 <- terra::crop(gridded_cattle_density_2020_10km, marine_and_inland50_wgs84)
terra::plot(cattle_cropped_wgs84)

# project only this small cropped raster into our preferred metric CRS (Mexico LCC)
cattle_projected_lcc <- terra::project(cattle_cropped_wgs84, terra::crs(marine_and_inland50))

# mask the raster to the exact boundaries of the 9 region polygons
gridded_cattle_density_2020_10km_goc <- terra::mask(cattle_projected_lcc, marine_and_inland50)

# verify the final CRS
terra::crs(gridded_cattle_density_2020_10km_goc)
terra::plot(gridded_cattle_density_2020_10km_goc)
```

#### 1. Resample to 1km

```{r}
# now resample to 1km
# make an empty template with 1000m (1km) resolution matching the extent and CRS
template_1km <- terra::rast(gridded_cattle_density_2020_10km_goc)
terra::res(template_1km) <- 1000

# resample the data using bilinear interpolation to smoothly estimate new 1km density values. original data is in cattle heads/km^2
cattle_density_1km <- terra::resample(
  x = gridded_cattle_density_2020_10km_goc, 
  y = template_1km, 
  method = "bilinear"
)

# reapply the polygon mask to clean up any edge cells bleeding past the borders
gridded_cattle_density_2020_1km_goc <- terra::mask(cattle_density_1km, marine_and_inland50)

# check and plot
print(gridded_cattle_density_2020_1km_goc)
terra::plot(gridded_cattle_density_2020_1km_goc)
```

#### 2. Dasymetric mapping of municipal numbers to spatially where estimated cattle locations are from the global raster

Because SIAP data is locked to arbitrary administrative boundaries (municipalities), we can't intersect it with watersheds, as a single municipality might span multiple watersheds. The GLW4 raster provides the "spatial proxy" to figure out where within those municipalities the livestock actually are.

###### a. prep the DGSIAP data and rasterize

Ensure municipality polygons (ex: siap_municipal_2022) are in the exact same metric projection (Mexico LCC) as the 1km raster.

```{r}
library(terra)
library(sf)
library(dplyr)

# make the 2022 SIAP sf object a terra SpatVector in Mexico LCC
mun_vector <- siap_municipal_2022 %>% 
  st_transform(st_crs(gridded_cattle_density_2020_1km_goc)) %>% 
  vect()
```


###### b. calculate the municipal weights

We need to find the total GLW raster value for each municipality and divide each cell by that total.

```{r}
# sum up the GLW 1km density values inside each municipality
mun_totals <- terra::extract(gridded_cattle_density_2020_1km_goc, mun_vector, fun = sum, na.rm = TRUE)

# append this municipal total back to the vector object (assuming the extraction order matches the vector rows)
mun_vector$glw_total_density <- mun_totals[, 2]

# rasterize the municipal totals so we can do raster algebra. we use the 1km cattle raster as a geometry template
mun_totals_raster <- terra::rasterize(mun_vector, gridded_cattle_density_2020_1km_goc, field = "glw_total_density")

# create the weight raster (Cell Weight = Cell Density / Municipal Total Density). cells will range from 0 to 1, representing the proportion of the municipality's cattle in that cell.
cattle_weight_raster <- gridded_cattle_density_2020_1km_goc / mun_totals_raster

plot(cattle_weight_raster)
```


###### c. distribute SIAP values (cattle volume) across the 1km grid

```{r}
## rasterize the actual SIAP livestock volume (tons) for 2022
siap_volume_raster <- terra::rasterize(mun_vector, gridded_cattle_density_2020_1km_goc, field = "toneladas_volumen")

# distribute the tons across the 1km grid
# result: a 1km raster where cell values = estimated SIAP tons of cattle in that specific 1km block
siap_cattle_tons_1km <- siap_volume_raster * cattle_weight_raster

# quick check: plot it to see the refined 1km distribution of Mexican govt data
plot(siap_cattle_tons_1km)

```

###### d. do for all years!

```{r}
install.packages(c("future", "furrr"))

library(future)
library(furrr)
```


```{r}
# ---- define output directory ----------------------------------------------------
out_dir <- "/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/pharmaceuticals/livestock_int"
# dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

# ---- per-year function ----------------------------------------------------
distribute_siap_year <- function(yr, siap_municipal, template_raster,
                                  weight_raster, out_dir) {
  
  message("processing year: ", yr)
  
  # 1. filter to this year and attach volume to geometry
  mun_yr_vector <- municipio_codigo_sf %>% 
    st_transform(st_crs(template_raster)) %>% 
    mutate(clv_mnc = as.numeric(clv_mnc),
           clv_std = as.numeric(clv_std)) %>%   # coerce both keys to match
    left_join(
      st_drop_geometry(siap_municipal %>% filter(anio == yr)) %>% 
        select(cvempio, cveestado, toneladas_volumen),
      by = c("clv_mnc" = "cvempio", "clv_std" = "cveestado")
    ) %>% 
    vect()
  
  # 2. rasterize SIAP volume for this year
  siap_volume_raster <- terra::rasterize(mun_yr_vector, template_raster,
                                          field = "toneladas_volumen")
  
  # 3. dasymetric distribution
  siap_cattle_tons_1km <- siap_volume_raster * weight_raster
  
  # 4. save to disk
  out_path <- file.path(out_dir, paste0("siap_cattle_tons_1km_", yr, ".tif"))
  terra::writeRaster(siap_cattle_tons_1km, out_path, overwrite = TRUE)
  
  message("  saved: ", basename(out_path))
  return(out_path)
}


# ---- run in parallel across all years ----------------------------------------------------
years_to_process <- c(2006:2022, 2024)

output_paths <- purrr::map(
  years_to_process,
  \(yr) distribute_siap_year(
    yr = yr,
    siap_municipal = siap_municipal,
    template_raster = gridded_cattle_density_2020_1km_goc,
    weight_raster = cattle_weight_raster,
    out_dir = out_dir
  )
)

# ---- quick sanity check ----------------------------------------------------
tibble(
  year = years_to_process,
  path = unlist(output_paths),
  exists = file.exists(unlist(output_paths))
)
```

#### 3. In theory, we would do the same thing for not just cattle, but also sheep, pigs, goat, and chicken (birds).

```{r}

```


#### 4. Summarize total kg of livestock production in each 1km cell, across all nomespecies.

First, we need to do dimensional analysis to convert from tonnes to kg of volume produced in each municipality and year.

```{r}

```


Now that we ideally have a 1km gridded raster of the total livestock production (kg) of all livestock types, we can estimate the mg of antimicrobial quantities for each municipality and year by applying:

*__ kg total volume terrestrial livestock slaughtered × 11 mg/kg of antimicrobials  = __ mg of antimicriobials applied that could potentially discharge into the GoC eventually.*

^ This number was determined based on Table 5 from WOAH <https://www.woah.org/app/uploads/2025/05/en-woah-amu-report-2025-final.pdf>.










### Dasymetric mapping 

Because SIAP data is locked to arbitrary administrative boundaries (municipalities), we can't intersect it with watersheds, as a single municipality might span multiple watersheds. The GLW4 (FAO livestock density) rasters provide the "spatial proxy" to figure out where within those municipalities the livestock actually are.


#### 2. Crop and resample all GLW4 rasters (for each nomespecie) to 1km

```{r}
# helper function to crop, project, and resample a single GLW4 raster to 1km
prep_glw_1km <- function(glw_path, mask_polygon) {
  
  glw_raw <- terra::rast(glw_path)
  mask_wgs84 <- st_transform(mask_polygon, terra::crs(glw_raw))
  
  glw_cropped <- terra::crop(glw_raw, mask_wgs84)
  glw_projected <- terra::project(glw_cropped, terra::crs(mask_polygon))
  glw_masked <- terra::mask(glw_projected, mask_polygon)
  
  # resample to 1km
  template_1km <- terra::rast(glw_masked)
  terra::res(template_1km) <- 1000
  glw_1km <- terra::resample(glw_masked, template_1km, method = "bilinear")
  glw_1km <- terra::mask(glw_1km, mask_polygon)
  
  return(glw_1km)
}

# apply to all species — named list for easy lookup by nomespecie
glw_1km_list <- purrr::map(
  setNames(species_glw_map$glw_file, species_glw_map$nomespecie),
  \(f) prep_glw_1km(file.path(glw_dir, f), marine_and_inland50)
)
```


#### 3. Compute the weight raster for that species (GLW cell density / municipal GLW total)

```{r}
# helper: given a 1km GLW raster and the municipality vector, return a weight raster
compute_weight_raster <- function(glw_1km, mun_vector_base, municipio_codigo_sf) {
  
  mun_vec <- municipio_codigo_sf %>% 
    st_transform(st_crs(glw_1km)) %>% 
    mutate(clv_mnc = as.numeric(clv_mnc),
           clv_std = as.numeric(clv_std)) %>% 
    vect()
  
  mun_totals <- terra::extract(glw_1km, mun_vec, fun = sum, na.rm = TRUE)
  mun_vec$glw_total <- mun_totals[, 2]
  mun_totals_raster <- terra::rasterize(mun_vec, glw_1km, field = "glw_total")
  weight_raster <- glw_1km / mun_totals_raster
  
  return(weight_raster)
}

# pre-compute weight rasters for all species (year-invariant)
weight_raster_list <- purrr::map(
  glw_1km_list,
  \(glw) compute_weight_raster(glw, mun_vector_base, municipio_codigo_sf)
)
```


#### 4-6. Filter SIAP to that species and year, get `toneladas_volumen` per municipality, rasterize the DIAP tonnes onto the municipality polygons, and multiply: siap_volume_raster * weight_raster

This is the actual dasymetric distribution step right here which will produce tonnes of that species slaughtered in each 1km cell.

```{r}
# these three steps are in one function per species and year
distribute_species_year <- function(yr, species, siap_municipal, 
                                     municipio_codigo_sf, glw_1km, 
                                     weight_raster, out_dir) {
  
  message("  processing: ", species, " - ", yr)
  
  # step 4: filter SIAP to species x year
  siap_yr_sp <- siap_municipal %>% 
    filter(anio == yr, nomespecie == species) %>% 
    st_drop_geometry() %>% 
    select(cvempio, cveestado, toneladas_volumen)
  
  # step 5: join to municipality polygons and rasterize
  mun_vec <- municipio_codigo_sf %>% 
    st_transform(st_crs(glw_1km)) %>% 
    mutate(clv_mnc = as.numeric(clv_mnc),
           clv_std = as.numeric(clv_std)) %>% 
    left_join(siap_yr_sp, by = c("clv_mnc" = "cvempio", "clv_std" = "cveestado")) %>% 
    vect()
  
  siap_volume_raster <- terra::rasterize(mun_vec, glw_1km, field = "toneladas_volumen")
  
  # step 6: dasymetric distribution
  species_tons_1km <- siap_volume_raster * weight_raster
  
  # save intermediate per-species x year raster
  out_path <- file.path(out_dir, paste0("siap_tons_1km_", species, "_", yr, ".tif"))
  terra::writeRaster(species_tons_1km, out_path, overwrite = TRUE)
  
  return(out_path)
}

# run across all species x year combinations
years_to_process <- c(2006:2022, 2024)

combos <- tidyr::crossing(
  yr  = years_to_process,
  species = species_glw_map$nomespecie
)

message("running dasymetric mapping for ", nrow(combos), " species and year combinations...")

purrr::walk2(
  combos$yr, combos$species,
  \(yr, sp) distribute_species_year(
    yr = yr,
    species = sp,
    siap_municipal = siap_municipal,
    municipio_codigo_sf = municipio_codigo_sf,
    glw_1km = glw_1km_list[[sp]],
    weight_raster = weight_raster_list[[sp]],
    out_dir = out_dir
  )
)
```


#### 7. Stack across species rasters and sum cellwise

This will produce total tonnes slaughtered per 1km cell per year.

```{r}
# for each year, read the 5 species rasters, stack, and sum to get total tonnes
sum_species_year <- function(yr, out_dir) {
  
  message("summing species for year: ", yr)
  
  species_files <- file.path(out_dir, 
    paste0("siap_tons_1km_", species_glw_map$nomespecie, "_", yr, ".tif"))
  
  # check all files exist before stacking
  missing <- species_files[!file.exists(species_files)]
  if (length(missing) > 0) {
    warning("missing files for year ", yr, ": ", paste(basename(missing), collapse = ", "))
    return(NULL)
  }
  
  stack <- terra::rast(species_files)
  total <- terra::app(stack, fun = sum, na.rm = TRUE)
  
  out_path <- file.path(out_dir, paste0("siap_total_tons_1km_", yr, ".tif"))
  terra::writeRaster(total, out_path, overwrite = TRUE)
  
  return(out_path)
}

total_tons_paths <- purrr::map(years_to_process, \(yr) sum_species_year(yr, out_dir))

# sanity check
tibble(
  year = years_to_process,
  path = unlist(total_tons_paths),
  exists = file.exists(unlist(total_tons_paths))
)
```


#### 8. Convert to mg of potential antimicrobials

```{r}
# conversion: tonnes -> kg -> mg antimicrobials
# 1 tonne = 1000 kg
# 11.87 mg antimicrobials per kg of terrestrial livestock slaughtered (WOAH 2025, Table 5)

convert_to_antimicrobials <- function(yr, out_dir) {
  
  message("Converting to antimicrobial load: ", yr)
  
  in_path  <- file.path(out_dir, paste0("siap_total_tons_1km_", yr, ".tif"))
  total_tonnes <- terra::rast(in_path)
  
  # tonnes -> kg -> mg antimicrobials
  mg_antimicrobials <- total_tonnes * 1000 * 11.87
  names(mg_antimicrobials) <- paste0("mg_antimicrobials_", yr)
  
  out_path <- file.path(out_dir, paste0("mg_antimicrobials_1km_", yr, ".tif"))
  terra::writeRaster(mg_antimicrobials, out_path, overwrite = TRUE)
  
  return(out_path)
}

antimicrobial_paths <- purrr::map(years_to_process, \(yr) convert_to_antimicrobials(yr, out_dir))

# final sanity check
tibble(
  year = years_to_process,
  path = unlist(antimicrobial_paths),
  exists = file.exists(unlist(antimicrobial_paths))
)
```


## Compare across regions

Then, we can take the *__ mg of antimicriobials applied that could potentially discharge into the GoC eventually* and determine, across all regions and time, which region is more likely to discharge pharmaceuticals from livestock relative to other regions and years.

$$current\ status\ _{livestock\_pharma} = 1 - \frac{mg\_antimicrobials\ _{current} - mg\_antimicrobials\ _{ideal}}{mg\_antimicrobials\ _{maximum} - mg\_antimicrobials\ _{ideal}}$$

Where the ideal is zero, and the maximum is the global maximum of estimated antimicrobials given to livestock across all years and regions.

#### 9a. Summarize within watersheds

```{r}
# project watersheds to match the antimicrobial rasters (Mexico LCC)
# we use the first year's raster just to grab the CRS
ref_crs <- terra::crs(terra::rast(file.path(out_dir, paste0("mg_antimicrobials_1km_", years_to_process[1], ".tif"))))

watersheds_lcc <- watershed_goc %>% 
  st_transform(ref_crs) %>% 
  vect()

# summarize mg antimicrobials within each watershed polygon, for each year
summarize_watersheds_year <- function(yr, out_dir, watersheds_lcc) {
  
  message("summarizing mg antimicrobials within watersheds for ", yr)
  
  mg_rast <- terra::rast(file.path(out_dir, paste0("mg_antimicrobials_1km_", yr, ".tif")))
  
  # sum all raster cells within each watershed polygon
  watershed_totals <- terra::extract(mg_rast, watersheds_lcc, fun = sum, na.rm = TRUE) #### need to double check the method
  
  # bind back to watershed attributes and tidy
  result <- as.data.frame(watersheds_lcc) %>% 
    mutate(
      year = yr,
      mg_antimicrobials = watershed_totals[, 2]
    )
  
  return(result)
}

watershed_antimicrobials <- purrr::map(
  years_to_process,
  \(yr) summarize_watersheds_year(yr, out_dir, watersheds_lcc)
) %>% 
  bind_rows() %>% 
  filter(!is.nan(mg_antimicrobials)) # filter out any basins without livestock pharmaceutical use

summary(watershed_antimicrobials)
```


#### 9b. Assign to pourpoints and then OHI regions

Note that in the shrimp farming pharmaceutical risk (`~/ohi_goc_prep/CW/v2025/chemicals/pesticides/script3_shrimp_farming_risk.qmd`), dropping zero/NA pourpoints makes sense there because a watershed with no aquaculture genuinely has no shrimp production. It is not a data gap, it's a true absence that shouldn't contribute to the regional score.

HOWEVER, for livestock, it is a bit more nuanced. There are two different reasons a pourpoint might have zero or NA antimicrobials:

(1) Actual absence, where the watershed drains an area with genuinely no livestock. These would be okay to drop.
(2) There is a data gap, in which the municipality exists but SIAP has no reported `toneladas_volumen` for that species and year. This is a real data gap and dropping it silently could underestimate pressure in that region.

```{r}
# project pourpoints to match
pourpoints_lcc <- pourpoints_goc %>% 
  st_transform(ref_crs)

# intersect with ohi regions
pourpoints_rgnid <- st_intersection(st_transform(pourpoints_lcc, st_crs(marine_and_inland50)), marine_and_inland50)
# mapview(pourpoints_rgnid)
# st_crs(pourpoints_rgnid)

# each pourpoint sits at the outlet of a watershed. pourpoints and watersheds share the `basin_id`
watershed_pourpoint_join <- pourpoints_rgnid %>% 
  select(basin_id, rgn_id) %>% 
  left_join(
    watershed_antimicrobials,
    by = "basin_id"
  ) %>% 
  st_as_sf() %>% 
  mutate(is_within_10km = lengths(st_within(., marine_and_inland10)) > 0) %>% # flag 10km
  filter(!(is_within_10km == FALSE & rgn_id %in% c("Region_2","Region_4","Region_9"))) # drop Pacific-side

mapview(watershed_pourpoint_join, zcol = "rgn_id")

# aggregate to OHI region: sum all watershed pourpoints that drain into each region
rgn_antimicrobials <- watershed_pourpoint_join %>% 
  group_by(rgn_id, year) %>% 
  summarize(mg_antimicrobials = sum(mg_antimicrobials, na.rm = TRUE), .groups = "drop")

summary(rgn_antimicrobials)

na_years <- rgn_antimicrobials %>% 
  filter(is.na(year))
mapview(na_years)

mapview(rgn_antimicrobials, zcol = "mg_antimicrobials")

# drop the pourpoints with zero contributions
rgn_antimicrobials_drop0 <- rgn_antimicrobials %>% 
  filter(!mg_antimicrobials == 0)
mapview(rgn_antimicrobials_drop0, zcol = "mg_antimicrobials")

summary(rgn_antimicrobials_drop0)
```

#### 10. Risk factor and current status

```{r}
# status = 1 - (mg_current - mg_ideal) / (mg_max - mg_ideal)
# ideal = 0 (no antimicrobial discharge), so: status = 1 - (mg_current / mg_max), where mg_max is the global maximum across all regions and years

mg_max <- max(rgn_antimicrobials$mg_antimicrobials, na.rm = TRUE)

current_status_livestock <- rgn_antimicrobials %>% 
  mutate(
   
     # rescaled pressure score (0 = no pressure, 1 = maximum pressure)
    pressure_score = mg_antimicrobials / mg_max,
    risk_factor = pressure_score * 100,
    
    # current status (1 = ideal, 0 = worst)
    current_status = 100 - risk_factor
  )

# check it to make sure the range is not wild
summary(current_status_livestock$current_status)
 #   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
 #   0.00   54.75   89.66   75.94   95.60   99.93

# --------------------- alt ------------------
mg_max_alt <- max(rgn_antimicrobials_drop0$mg_antimicrobials, na.rm = TRUE)

current_status_livestock_alt <- rgn_antimicrobials_drop0 %>% 
  mutate(
   
     # rescaled pressure score (0 = no pressure, 1 = maximum pressure)
    pressure_score = mg_antimicrobials / mg_max_alt,
    risk_factor = pressure_score * 100,
    
    # current status (1 = ideal, 0 = worst)
    current_status = 100 - risk_factor
  )

# check it to make sure the range is not wild
summary(current_status_livestock_alt$current_status)
   # Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
   # 0.00   52.47   86.81   73.80   95.71   99.93 
```

##### ~ plotlys

```{r}
library(plotly)

rgn_colors <- c(
  Region_1 = "#C03729",
  Region_2 = "#E68C7C",
  Region_3 = "#FC8F24",
  Region_4 = "#dcaa38",
  Region_5 = "#cbd2a0",
  Region_6 = "#98a25a",
  Region_7 = "#747428",
  Region_8 = "#627391",
  Region_9 = "#8d61a9"
)

## plot the livestock pharma cs over time
current_status_livestock_plotly <- plotly::plot_ly(
  current_status_livestock_alt, 
  x = ~year, 
  y = ~current_status, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Clean Waters: Current status (1 - risk_factor) of pharmaceutical chemicals from livestock, based on tonnes/year of production",
    xaxis = list(title = "ano // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines( 
    x = range(current_status_livestock_alt$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
current_status_livestock_plotly

## plot the livestock pharma rf over time
rf_livestock_plotly <- plotly::plot_ly(
  current_status_livestock_alt, 
  x = ~year, 
  y = ~risk_factor, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Clean Waters: Relative risk of pharmaceutical chemicals from livestock, based on tonnes/year of production",
    xaxis = list(title = "ano // Year"),
    yaxis = list(title = "Risk factor (mg_antimicrobials - ideal)/(maximum - ideal) * 100")
  ) %>%
  add_lines( 
    x = range(current_status_livestock_alt$year), 
    y = c(0, 0),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (0)",
    inherit = FALSE
  )
rf_livestock_plotly

```


### Save to Mazu

```{r}
final_out_dir <- "/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/pharmaceuticals/livestock_output"

# save full intermediate table (watershed level)
write_csv(rgn_antimicrobials_drop0, here::here(final_out_dir, "watershed_antimicrobials_livestock.csv"))

# save region-level pressure and status scores
write_csv(current_status_livestock_alt, here::here(final_out_dir, "rgn_cw_pharmaceuticals_livestock_status.csv"))
```


