---
title: 'OHI GoC 2025: Clean waters pharmaceuticals'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# CW: Pharmaceuticals - WWTP Effluent

*Alfonso-Muniozguren P, Serna-Galvis EA, Bussemaker M, Torres-Palma RA, Lee J. A review on pharmaceuticals removal from waters by single and combined biological, membrane filtration and ultrasound systems. Ultrason Sonochem. 2021 Aug;76:105656. doi: 10.1016/j.ultsonch.2021.105656. Epub 2021 Jul 1. PMID: 34274706; PMCID: PMC8319449.*

"Currently the biggest threat to the global freshwater reserve and, ecological and public health is the rising of emerging contaminants in the environment such as pharmaceutical drugs [2](https://doi.org/10.1016/j.scitotenv.2013.12.065), [3](https://doi.org/10.1007/s10661-017-6097-x). Wastewater treatment plants (WWTPs) are considered the primary source of pharmaceuticals in the environment, although a higher concentration of certain compounds can be found in hospital effluents [4](https://doi.org/10.1007/s11356-017-9662-5). Current water treatment methods implemented in WWTPs are inefficient for the complete removal of most pharmaceuticals [2](https://doi.org/10.1016/j.scitotenv.2013.12.065), [5](https://doi.org/10.1016/j.watres.2012.11.027), leading to the discharge of these compounds directly into water bodies. The concentration of pharmaceuticals in hospital effluents [4](https://doi.org/10.1007/s11356-017-9662-5), [6](https://doi.org/10.1021/es400708w), [7](https://doi.org/10.1007/s11356-016-6389-7), pharmaceutical industry effluents [8](https://pmc.ncbi.nlm.nih.gov/articles/PMC8319449/#b0040), [9](https://scholar.google.com/scholar_lookup?journal=Ind.%20Eng.%20Chem.%20Res.&title=Pharmaceutical%20Industry%20Wastewater:%20Review%20of%20the%20Technologies%20for%20Water%20Treatment%20and%20Reuse&author=C.%20Gadipelly&author=A.%20P%C3%A9rez-Gonz%C3%A1lez&author=G.D.%20Yadav&author=I.%20Ortiz&author=R.%20Ib%C3%A1%C3%B1ez&volume=53&publication_year=2014&pages=11571-11592&) and WWTPs [10](https://doi.org/10.1016/j.watres.2009.07.004), [11](https://pubmed.ncbi.nlm.nih.gov/15497842/) range from ng L−1 to µg L−1. The concentration of these compounds in the environment depends on discharge, geographical location, climate conditions and consumption [12](https://doi.org/10.1016/j.envint.2006.09.014)."

## Estimated pharmaceutical discharge

We will be doing this by estimating the total liters per year of effluent from each WWTP across the Gulf, determining which watershed the WWTP falls into, and aggregate the liters per year at the associated pourpoints.  

From there, we can classify which OHI Region each pourpoint falls into, and aggregate the total liters per year of WWTP effluent for each year and region. This will then represent pharmaceutical discharge risk for each year and region.

The regions will then be compared across time in `script4_current_status.qmd`.

### Workflow

*For each WWTP x year combination:*

1. Load the year-specific WWTP shapefile from SINA/CONAGUA (ex: m29_vwplantastarm_2021)
2. Filter to GoC states (Baja California, BCS, Jalisco, Sinaloa, Sonora, Nayarit) and fix encoding
3. Flag whether each WWTP is over capacity (caudal > capacidad)
4. Reproject WWTPs to match watershed CRS, then intersect with GoC watersheds (wwtps_watershed)
5. Convert flow rate to annual volume: caudal (L/s) * 31,536,000 -> estimated L/year per WWTP (upper bound, assumes continuous operation)
6. Summarize by basin_id x year: sum ef_l_yr, count total WWTPs, count over-capacity plants
7. Join watershed summaries to GoC pourpoints (pourpoints_goc) by basin_id
8. Spatially filter pourpoints: keep only those within the 50km inland buffer, and exclude Pacific-side peninsula pourpoints (Regions 4, 5, 9) not within the 10km buffer
9. Intersect with OHI regions and aggregate total_ef_l_yr by rgn_id x year

*Convert to mg pharmaceuticals:*

10. Multiply total_ef_l_yr by the mean pharmaceutical concentration across literature compounds (mean_ng/L from Vázquez-Tapia et al. 2020 + Garduño-Jiménez et al. 2023), then convert ng -> mg (* 1/1,000,000)

*Data quality:*

11a. Explore what happened in 2020
11b. Flag 2020 anomalies: for each WWTP, compare 2020 caudal to its non-2020 median, flag if ratio > 3x or < 0.33x, or if treatment process changed
11c. Set 2020 values to NA and linearly interpolate within each region using zoo::na.approx
12. Move to script4_current_status.qmd to aggregate the total mg from WWTPs with livestock and shrimp farming, and use the total_pharma_mg against the global maximum (or top-5 average maximum) across all regions and years to obtain a current status score.

*REMEMBER: SINA/CONAGUA WWTP data = where and how much flow, Vázquez-Tapia concentration values = how much pharmaceutical load per liter!*


### Setup

```{r setup}
library(sf)
library(utils)
library(tidyverse)
library(here)
library(raster)
library(mapview)
library(purrr)
library(stringr) # fuzzy string matching
library(stringi)
library(janitor)
library(tibble)
```


First, let's isolate the watersheds and associated pourpoints in the Gulf of California.

```{r}
## lets use the full state for now and then summarize by the full watersheds.
goc_ohi_rgns <- st_read(here::here("spatial/ohi_regions","goc_ohi_rgns.shp"))

union_rgns <- goc_ohi_rgns %>% 
  st_union() %>% 
  st_make_valid()

# plot(goc_ohi_rgns)
# plot(union_rgns)

watershed_all <- st_read(file.path("/home/shares/ohi/stressors_2021/_dataprep/nutrients/watersheds_pourpoints/watersheds_all_4326.shp"))
st_crs(watershed_all)
# mapview(watershed_all)

watershed_goc <- st_intersection(st_transform(watershed_all, st_crs(union_rgns)), union_rgns)
mapview(watershed_goc)

# --- if we plan to change the buffer for some reason ---
# inland_50 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer","inland_50km_goc.shp"))
# watershed_goc_50km <- st_intersection(st_transform(watershed_all, st_crs(inland_50)), inland_50)
# mapview(watershed_goc_50km)
```

```{r}
# here's the associated pourpoints that should match with the watersheds:
pourpoints <- st_read("/home/shares/ohi/stressors_2021/_dataprep/nutrients/watersheds_pourpoints/pourpoints_mol.shp")

pourpoints_goc <- st_intersection(st_transform(pourpoints, st_crs(union_rgns)), union_rgns)
mapview(pourpoints_goc)

mapview(pourpoints_goc) + mapview(watershed_goc)
```

### 1-3. Reading in all WWTP capacities and usages from 2017-2023

Data was downloaded from <https://sinav30.conagua.gob.mx:8080/SINA/?opcion=salud#infrah>: 

(1) click on "Infraestructura Hidráulica", 

(2) Click on "Plantas de Tratamiento de Aguas Residuales", 

(3) For each year, go through this process: select the year, click PTAR (the acronym in Spanish, it would be WWTPs in English), and then select the next year and do the same thing, which ensures that the data is shown on the interactive map,

(4) Once all layers from all years are showing, go to the bottom left-hand corner and click the button that is meant to represent "see all layers currently on the map".  It has an eye and multiple squares, looking like layered papers.  It is located just above the question mark button.  Here, you will go one by one, and download each of the years by clicking on the button representing ".SHP".  Because the .zip files are all named the same, regardless of the year, you will need to rename each folder as you go before adding it to `~/home/shares/ohi/OHI_GOC/_raw_data/SINA_3_CONAGUA/d2025/wastewater_treatment_plants`.  Once you unzip each folder, add the appropriate year in the end of the folder name, for example: "_2020".

```{r}
## write function to read in all of the shapefiles and limit them to GoC states, as well as determine whether they were over capacity for that year and WWTP
read_in_wwtp <- function (year) {
  
  df <- st_read(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SINA_3_CONAGUA/d2025/wastewater_treatment_plants", paste0("m29_vwplantastarm_", year))) %>% 
    mutate(across(where(is.character), ~iconv(., from = "WINDOWS-1252", to = "UTF-8"))) %>%  # to address encoding issues
    filter(estado %in% c("Baja California Sur","Baja California","Jalisco","Sinaloa","Sonora","Nayarit")) %>% 
    mutate(over_proc = case_when(
      caudal > capacidad ~ TRUE,
      TRUE ~ FALSE
    ))
  
  return(df)
  
}
```

Now, we can use the function to read in each year's WWTP data and limit to GoC states.

```{r}
# specify the years we want to iterate through.  Change this in the future if necessary.
years_to_read_in <- 2017:2023

# use `lapply` and bind rows to obtain data over time.
list_of_all_dfs <- lapply(years_to_read_in, read_in_wwtp)
  
# bind them across all years
bind_all_years_wwtp <- bind_rows(list_of_all_dfs)

summary(bind_all_years_wwtp)
unique(bind_all_years_wwtp$estado)
procesos <- unique(bind_all_years_wwtp$proceso)
```

Take a look...

```{r}
mapview(bind_all_years_wwtp, zcol = "over_proc")
```


#### Exploring WWTPs over time

This is just some exploratory stats to understand the landscape better.

```{r}
pct_over_processing <- bind_all_years_wwtp %>% 
  st_drop_geometry() %>% 
  group_by(anio, estado) %>% 
  summarize(
    total_wwtp = n(),
    num_over_proc = sum(over_proc, na.rm = TRUE),
    pct_over_proc = round(num_over_proc / total_wwtp * 100, 1),
    mean_caudal = mean(caudal, na.rm = TRUE),
    mean_capacidad = mean(capacidad, na.rm = TRUE),
    .groups = "drop"
  )

number_of_wwtp <- bind_all_years_wwtp %>% 
  st_drop_geometry() %>% 
  group_by(anio, estado) %>% 
  summarize(
    num_wwtp = n()
  ) %>% 
  ungroup()
```

### 4. Reproject WWTPs to match watershed CRS, then intersect with GoC watersheds (wwtps_watershed)

Now that we have data about the capacity and the actual L/s of each WWTP, let's assign each WWTP to its respective watershed.

```{r}
# reproject to match watersheds CRS
bind_all_years_wwtp_reproj <- st_transform(bind_all_years_wwtp, st_crs(watershed_goc))

# Coordinate Reference System:
#   User input: Mexico ITRF2008 / LCC 
#   wkt:
# PROJCRS["Mexico ITRF2008 / LCC",
#     BASEGEOGCRS["Mexico ITRF2008",
#         DATUM["Mexico ITRF2008",
#             ELLIPSOID["GRS 1980",6378137,298.257222101,
#                 LENGTHUNIT["metre",1]]],

# intersect wwtps with watersheds
wwtps_watershed <- st_intersection(bind_all_years_wwtp_reproj, watershed_goc)
colnames(wwtps_watershed)
```

Check out 2021:
```{r}
wwtps_watershed_2021 <- wwtps_watershed %>% 
  filter(anio %in% 2021)

mapview(wwtps_watershed_2021, zcol = "capacidad") 

mapview(wwtps_watershed_2021, zcol = "basin_id") + mapview(watershed_goc, zcol = "basin_id")
```


### 5. Convert flow rate to annual volume

To go from L/s to L/year, we need to multiply by 31536000.  Note that this is a rough assumption... it assumes that the WWTP is running constantly, but it is likely not running 24/7/365.  This is basically an upper bound for the year.

```{r}
summary(bind_all_years_wwtp$caudal) # no zeros or NAs:      Min.   1st Qu.    Median      Mean   3rd Qu.      Max. 
                                                 #  0.010     1.377     3.100    53.311    15.100 28743.230 

effluent_liters_wwtp_yearly <- bind_all_years_wwtp %>%
  mutate(
    ef_l_yr = caudal * 31536000 # assumes that the WWTP is running constantly... but it is likely not running 24/7/365.  This is basically an upper bound for the year.
  )
```

### 6. Summarize liters per watershed per year

We can now aggregate the total L/year for each watershed. Here, you can summarize by basin_id x year: sum ef_l_yr, count total WWTPs, and count over-capacity plants.

```{r}
# summarize liters per watershed per year
wwtp_liters_effluent_by_watershed <- wwtps_watershed %>%
  st_drop_geometry() %>% # maybe take out...
  group_by(basin_id, anio) %>% 
  mutate(
    ef_l_yr = caudal * 31536000 # assumes that the WWTP is running constantly... but it is likely not running 24/7/365.  This is basically an upper bound for the year
  ) %>% 
  summarize(
    total_wwtp = n(),
    num_over_proc = sum(over_proc, na.rm = TRUE),
    pct_over_proc = num_over_proc / total_wwtp * 100,
    total_ef_l_yr = sum(ef_l_yr),
    total_ef_m3_yr = total_ef_l_yr / 1000 # more readable
  ) %>% 
  ungroup() %>% 
  mutate(year = as.numeric(anio))
```

### 7-8. Join watershed summaries to GoC pourpoints (pourpoints_goc) by basin_id

And now we can take that to the pourpoints!

```{r pourpoints}
pourpoints_wwtp_liters <- pourpoints_goc %>%
  dplyr::select(basin_id) %>%
  left_join(wwtp_liters_effluent_by_watershed, by = "basin_id")

mapview(pourpoints_wwtp_liters)
```

Now, spatially filter pourpoints: keep only those within the 50km inland buffer, and exclude Pacific-side peninsula pourpoints (Regions 4, 5, 9) not within the 10km buffer.

*This is just so that we are not losing any WWTPs that may be on the coastal border:*

```{r ohi_marine_and_inland}
inland_50 <- st_read(here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer", "inland_50km_goc.shp"))
inland_10 <- st_read(here("/home/shares/ohi/OHI_GOC/spatial/inland_10km_buffer", "inland_10km_goc.shp"))
regions_marine <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp")) %>%
  filter(location == "marine")

marine_and_inland50 <- bind_rows(regions_marine, inland_50) %>%
  dplyr::select(rgn_id, geometry) %>%
  group_by(rgn_id) %>%
  summarize(geometry = st_union(geometry), .groups = "drop")

marine_and_inland10 <- bind_rows(regions_marine, inland_10) %>%
  dplyr::select(rgn_id, geometry) %>%
  group_by(rgn_id) %>%
  summarize(geometry = st_union(geometry), .groups = "drop")
```

Now we can ensure that we only keep the pourpoints in the GoC that have pharmaceutical effluent associated with them.

```{r pourpoints}
pourpoints_wwtp_liters_final <- pourpoints_wwtp_liters %>%
  filter(!is.na(total_ef_l_yr), total_ef_l_yr > 0) %>% # note that the watershed with the id 73730 wraps around, and that the pour point is only on the left side...not sure that is correct.
  st_intersection(marine_and_inland50) %>% # keep those within 50 km...
  mutate(is_within_10km = lengths(st_within(., marine_and_inland10)) > 0) %>% # but make sure *not* to keep the ones on the pacific side of the peninsula.
  filter(!(is_within_10km == "FALSE" & rgn_id %in% c("Region_4","Region_5","Region_9")))

# ensure that all of the pourpoints outside of the Gulf are not included.
mapview(pourpoints_wwtp_liters_final, zcol = "total_ef_l_yr") + mapview(wwtps_watershed, col.regions = "red") + mapview(watershed_goc, col.regions = "pink")
```


### 9. Aggregate total_ef_l_yr by rgn_id x year

And now we can aggregate by region to determine the total L/year and region.

```{r pourpoints}
pourpoints_wwtp_liters_rgnid <- st_intersection(pourpoints_wwtp_liters_final, goc_ohi_rgns) %>%
  st_drop_geometry() %>%
  group_by(rgn_id, year) %>%
  summarize(
    total_ef_l_yr = sum(total_ef_l_yr),
    total_ef_m3_yr = sum(total_ef_m3_yr),
    total_wwtp = sum(total_wwtp),
    num_over_proc = sum(num_over_proc),
    pct_over_proc = num_over_proc / total_wwtp * 100,
    .groups = "drop"
  )

pourpoints_wwtp_liters_rgnid
```

### 10. Convert to mg pharmaceuticals

#### a. WWTP effluent concentration values

```{r}
concentrations_lit <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/VazquezTapia_2022_supplemental/table_of_lit_manually_adding_tula_valley_2023/VazquezTapia_et_al_2020_S2-isolated_priority_list_env_risk_WITH_tula_valley.csv")) %>% 
  janitor::clean_names() %>% 
  mutate(across(everything(), ~ str_replace_all(.x, "ND", NA_character_)))

concentrations_tidy <- concentrations_lit %>%
  mutate(
    garduno_jimenez_et_al_2023_minimum = garduno_jimenez_et_al_2023,
    garduno_jimenez_et_al_2023_maximum = garduno_jimenez_et_al_2023
  ) %>%
  select(-garduno_jimenez_et_al_2023) %>%
  pivot_longer(
    cols = -compound_reference_ng_per_liter,
    names_to = c("paper", "bound"),
    names_pattern = "(.*)_(minimum|maximum)$",
    values_to = "value"
  ) %>%
  pivot_wider(names_from = bound, values_from = value) 

concentrations_midpoint <- concentrations_tidy %>%
  mutate(
    minimum = as.numeric(minimum),
    maximum = as.numeric(maximum)
  ) %>% 
  mutate(
    geometric_mean = sqrt(minimum * maximum),
    midpoint = (minimum + maximum) / 2, 
    end_value = case_when(
      is.na(minimum) & is.na(maximum) ~ NA_real_,
      is.na(minimum) ~ maximum,
      is.na(maximum) ~ minimum,
      TRUE ~ midpoint
    )
  )

concentractions_per_compound <- concentrations_midpoint %>% 
  group_by(compound_reference_ng_per_liter) %>% 
  summarize(
    mean_ng_per_liter = mean(end_value, na.rm = TRUE),
    .groups = "drop"
  )

## save to Mazu
write_csv(concentractions_per_compound, here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/pharmaceuticals/humans_int","concentractions_per_compound_across_lit_ng_per_liter.csv"))

concentraction_pharmaceuticals <- concentractions_per_compound %>% 
  summarize(
    mean_ng_per_liter = mean(mean_ng_per_liter),
    .groups = "drop"
  )

## save to Mazu
write_csv(concentraction_pharmaceuticals, here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/pharmaceuticals/humans_int","concentraction_all_env_risk_pharmaceuticals_ng_per_liter.csv"))
```

#### b. Use the values to convert to mg

Now that we have the ng/L estimated pharmaceutical chemicals per liter of WWTP effluent, we can take the aggregated L/year found for each region and multiply by the ng/L to obtain total ng potentially flowing into the GoC.

To harmonize units across different chemical sources, we can convert to mg from ng.

```{r}
mean_ng_per_liter_value <- unique(concentraction_pharmaceuticals$mean_ng_per_liter)

pourpoints_wwtp_mg_rgnid <- pourpoints_wwtp_liters_rgnid %>% 
  mutate(total_pharma_ng = total_ef_l_yr * mean_ng_per_liter_value,
         total_pharma_mg = total_pharma_ng * (1/1000000)) # 1000000 ng in 1 mg
```



<!-- ### Current status -->

<!-- Note that the risk factor is essentially the pressure on the system.  The current status is the inverse of that (1 - risk_factor). -->

<!-- $$current\ status\ _{human\_wastewater\_pharmaceutical\_risk} = 1 - \frac{wastewater\_in\_liters\ _{current} - wastewater\_in\_liters\ _{ideal}}{wastewater\_in\_liters\ _{maximum} - wastewater\_in\_liters\ _{ideal}}$$ -->

<!-- Where the ideal is zero, and the maximum is the global maximum of liters estimated across all years and regions. -->

<!-- ```{r} -->
<!-- current_status_wwtp_effluent <- pourpoints_wwtp_liters_rgnid %>%  -->
<!--   mutate( -->
<!--     ideal = 0, -->
<!--     maximum = max(total_ef_l_yr), -->
<!--     current_status = 1 - ((total_ef_l_yr - ideal)/(maximum - ideal)), -->
<!--     current_status_final = current_status * 100 -->
<!--   ) -->

<!-- current_status_wwtp_effluent_drop2020 <- pourpoints_wwtp_liters_rgnid %>%  -->
<!--   filter(!year %in% 2020) %>%  -->
<!--   mutate( -->
<!--     ideal = 0, -->
<!--     maximum = max(total_ef_l_yr), -->
<!--     current_status = 1 - ((total_ef_l_yr - ideal)/(maximum - ideal)), -->
<!--     current_status_final = current_status * 100 -->
<!--   ) -->

<!-- ``` -->

<!-- #### plotly -->

<!-- Plot both the current status, but also the L/year across all regions and years to understand how the current status was derived. -->

<!-- ```{r} -->
<!-- library(plotly) -->

<!-- rgn_colors <- c( -->
<!--   Region_1 = "#C03729", -->
<!--   Region_2 = "#E68C7C", -->
<!--   Region_3 = "#FC8F24", -->
<!--   Region_4 = "#dcaa38", -->
<!--   Region_5 = "#cbd2a0", -->
<!--   Region_6 = "#98a25a", -->
<!--   Region_7 = "#747428", -->
<!--   Region_8 = "#627391", -->
<!--   Region_9 = "#8d61a9" -->
<!-- ) -->

<!-- ## plot the wwtp effluent rf over time -->
<!-- wwtp_plotly <- plotly::plot_ly( -->
<!--   current_status_wwtp_effluent,  -->
<!--   x = ~year,  -->
<!--   y = ~current_status_final,  -->
<!--   color = ~rgn_id,  -->
<!--   colors = rgn_colors, -->
<!--   type = "scatter",  -->
<!--   mode = "lines+markers" -->
<!-- ) %>%  -->
<!--   layout( -->
<!--     title = "Clean Waters: current status of WWTP effluent, estimation of pharmaceutical risk from human wastewater", -->
<!--     xaxis = list(title = "ano // Year"), -->
<!--     yaxis = list(title = "Puntaje // Percent (out of 100)") -->
<!--   ) %>% -->
<!--   add_lines(  -->
<!--     x = range(current_status_wwtp_effluent$year),  -->
<!--     y = c(100, 100), -->
<!--     line = list(color = "black", width = 2, dash = "dash"), -->
<!--     name = "Meta // Goal (100/100)", -->
<!--     inherit = FALSE -->
<!--   ) -->
<!-- wwtp_plotly -->

<!-- ## plot the wwtp effluent rf over time -->
<!-- wwtp_drop2020_plotly <- plotly::plot_ly( -->
<!--   current_status_wwtp_effluent_drop2020,  -->
<!--   x = ~year,  -->
<!--   y = ~current_status_final,  -->
<!--   color = ~rgn_id,  -->
<!--   colors = rgn_colors, -->
<!--   type = "scatter",  -->
<!--   mode = "lines+markers" -->
<!-- ) %>%  -->
<!--   layout( -->
<!--     title = "Current status of WWTP effluent, estimation of pharmaceutical risk from human wastewater (dropped 2020)", -->
<!--     xaxis = list(title = "ano // Year"), -->
<!--     yaxis = list(title = "Puntaje // Percent (out of 100)") -->
<!--   ) %>% -->
<!--   add_lines(  -->
<!--     x = range(current_status_wwtp_effluent_drop2020$year),  -->
<!--     y = c(100, 100), -->
<!--     line = list(color = "black", width = 2, dash = "dash"), -->
<!--     name = "Meta // Goal (100/100)", -->
<!--     inherit = FALSE -->
<!--   ) -->
<!-- wwtp_drop2020_plotly -->

<!-- # --- -->

<!-- ## plot the liters per year of effluent for all rgns -->
<!-- rgn_eff_plotly <- plotly::plot_ly( -->
<!--   pourpoints_wwtp_liters_rgnid,  -->
<!--   x = ~year,  -->
<!--   y = ~total_ef_l_yr,  -->
<!--   color = ~rgn_id,  -->
<!--   colors = rgn_colors, -->
<!--   type = "scatter",  -->
<!--   mode = "lines+markers" -->
<!-- ) %>%  -->
<!--   layout( -->
<!--     title = "Clean Waters: liters per year effluent from WWTPs", -->
<!--     xaxis = list(title = "ano // Year"), -->
<!--     yaxis = list(title = "Liters of effluent at pourpoints within the region") -->
<!--   ) -->
<!-- rgn_eff_plotly -->
<!-- ``` -->

### 11. Deal w 2020

#### a. Examining the spike in 2020 found in Region 8 and 9

```{r}
wwtps_watershed_na_80866_2020 <- wwtps_watershed %>% 
  filter(basin_id %in% "na_80866") %>% 
  filter(anio %in% 2020)

wwtps_watershed_na_80866_2019 <- wwtps_watershed %>% 
  filter(basin_id %in% "na_80866") %>% 
  filter(anio %in% 2019)

mapview(wwtps_watershed_na_80866_2020, zcol = "total_ef_l_yr") # look at it over time, see if 2020 has any outlier or inconsistent internal results over time.  Determine what is going on there!! Is it a mistake???
mapview(wwtps_watershed_na_80866_2019, zcol = "total_ef_l_yr")
```

```{r}
## isolate region 8 and 9 basins
pourpoints_wwtp_liters_final

rgn_8_9_basins <- pourpoints_wwtp_liters_final %>% 
  filter(rgn_id %in% c("Region_8","Region_9")) %>% 
  st_drop_geometry()

## plot the liters per year of effluent for these basins
inspecting_8_9_plotly <- plotly::plot_ly(
  rgn_8_9_basins, 
  x = ~year, 
  y = ~total_ef_l_yr, 
  color = ~basin_id, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "L/y of effluent from each watershed over time in regions 8 and 9",
    xaxis = list(title = "ano // Year"),
    yaxis = list(title = "Liters of effluent at pourpoints for each basin")
  )
inspecting_8_9_plotly
```


```{r}
### so na_76895 has the spike, and na_80866 has the reduction in 2020.
basins_76895_80866 <- pourpoints_wwtp_liters_final %>% 
  filter(basin_id %in% c("na_76895","na_80866")) 

### okay, so let's see which WWTPs in the basins caused the spikes.
wwtps_watershed_only_basins_76895_80866 <- wwtps_watershed %>% 
  st_drop_geometry() %>% 
  group_by(clvplanta, anio, caudal) %>% 
  distinct() %>% 
  filter(basin_id %in% c("na_76895","na_80866")) %>% 
  mutate(
    ef_l_yr = caudal * 31536000, # assumes that the WWTP is running constantly... but it is likely not running 24/7/365.  This is basically an upper bound for the year
    year = as.numeric(anio)
  ) %>% 
  ungroup()

## plot the liters per year of effluent for these basins
inspecting_basins_76895_80866_plotly <- plotly::plot_ly(
  wwtps_watershed_only_basins_76895_80866, 
  x = ~year, 
  y = ~ef_l_yr, 
  color = ~clvplanta, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "L/y of effluent from all WWTPs in basins na_76895 and na_80866",
    xaxis = list(title = "ano // Year"),
    yaxis = list(title = "Liters of effluent for each WWTP over time")
  )
inspecting_basins_76895_80866_plotly
```


```{r}
## it appears that clvplanta p_25-4117 has the spike in 2020, while p_14-4871 and p_14-3656 had the drastic decreases in 2020.

wwtps_watershed_only_clvplanta_outliers <- wwtps_watershed_only_basins_76895_80866 %>% 
  filter(clvplanta %in% c("p_25-4177","p_14-4871","p_14-3656"))


## plot the liters per year of effluent for these plants
inspecting_basins_76895_80866_plotly <- plotly::plot_ly(
  wwtps_watershed_only_basins_76895_80866, 
  x = ~year, 
  y = ~ef_l_yr, 
  color = ~clvplanta, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "L/y of effluent from all WWTPs in basins na_76895 and na_80866",
    xaxis = list(title = "ano // Year"),
    yaxis = list(title = "Liters of effluent for each WWTP over time")
  )
inspecting_basins_76895_80866_plotly
```

#### b. Flagging anomalies

Check for 2020 anomalies across all WWTPs in the GoC.

```{r check-2020-anomalies}
# for each WWTP (clvplanta), find the median caudal across non-2020 years, then flag 2020 rows where caudal deviates by more than 3x (or is zero/NA when it shouldnt be)
wwtps_2020_anomaly_check <- wwtps_watershed %>%
  st_drop_geometry() %>%
  group_by(clvplanta, anio, caudal, proceso, capacidad, basin_id) %>%
  distinct() %>%
  mutate(year = as.numeric(anio)) %>%
  ungroup() %>%
  group_by(clvplanta) %>%
  mutate(
    median_caudal_non2020 = median(caudal[year != 2020], na.rm = TRUE), # to see how much higher/lower the L/s processing jumped or fell
    caudal_ratio_2020 = ifelse(
      year == 2020 & !is.na(median_caudal_non2020) & median_caudal_non2020 > 0,
      caudal / median_caudal_non2020,
      NA_real_
    ),
    proceso_2020 = proceso[year == 2020][1],
    proceso_modal = names(sort(table(proceso[year != 2020]), decreasing = TRUE))[1],
    proceso_changed_2020 = ifelse(proceso_2020 != proceso_modal, TRUE, FALSE)
  ) %>%
  ungroup()

## pull out just the 2020 rows that look suspicious: ex, caudal is >3x or < (1/3) x the plant's individual non-2020 median, OR treatment process changed
anomalies_2020 <- wwtps_2020_anomaly_check %>%
  filter(year == 2020) %>%
  filter(
    (caudal_ratio_2020 > 3 | caudal_ratio_2020 < 0.33) | proceso_changed_2020 %in% "TRUE" # 1/3 = 0.33
  ) %>%
  arrange(basin_id, clvplanta) %>%
  select(basin_id, clvplanta, municipio, localidad,
         proceso, proceso_modal, capacidad,
         caudal, median_caudal_non2020, caudal_ratio_2020,
         proceso_changed_2020)

anomalies_all_yrs <- wwtps_2020_anomaly_check %>%
  filter(
    (caudal_ratio_2020 > 3 | caudal_ratio_2020 < 0.33) | proceso_changed_2020 == TRUE # 1/3 = 0.33
  ) %>%
  arrange(basin_id, clvplanta) %>%
  select(basin_id, clvplanta, municipio, localidad,
         proceso, proceso_modal, capacidad,
         caudal, median_caudal_non2020, caudal_ratio_2020,
         proceso_changed_2020)


## how many unique plants are flagged?
cat("how many flagged WWTPs in 2020:", n_distinct(anomalies_2020$clvplanta), "\n") #258
cat("how many basins:", n_distinct(anomalies_2020$basin_id), "\n") # 76

## interactive plot of ALL flagged plants over time for visual confirmation!!
flagged_plants <- unique(anomalies_2020$clvplanta)

wwtps_flagged_timeseries <- wwtps_2020_anomaly_check %>%
  filter(clvplanta %in% flagged_plants) %>%
  mutate(ef_l_yr = caudal * 31536000)

wwtps_flagged_timeseries_plotly <- plotly::plot_ly(
  wwtps_flagged_timeseries,
  x = ~year,
  y = ~ef_l_yr,
  color = ~clvplanta,
  text = ~paste("Basin:", basin_id, "<br>Proceso:", proceso,
                  "<br>Capacidad:", capacidad, "<br>Caudal:", caudal),
  type = "scatter",
  mode = "lines+markers",
  hoverinfo = "text+x+y"
) %>%
  layout(
    title = "All flagged WWTPs with potential 2020 anomalies",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Estimated L/yr of effluent"),
    shapes = list(list(  # vertical line to highlight 2020
      type = "line", x0 = 2020, x1 = 2020,
      y0 = 0, y1 = 1, yref = "paper",
      line = list(color = "red", dash = "dot")
    ))
  )
wwtps_flagged_timeseries_plotly
```

#### c. Interpolate 2020

Because 2020 seems to have database entry errors, we should interpolate and drop any of the original 2020 values.

```{r}
library(zoo) # for na.approx

# set 2020 values to NA so they can be interpolated
pourpoints_wwtp_mg_rgnid_interp <- pourpoints_wwtp_mg_rgnid %>%
  mutate(
    total_ef_l_yr_raw = total_ef_l_yr,
    total_pharma_ng_raw = total_pharma_ng,
    total_pharma_mg_raw = total_pharma_mg,
    total_ef_l_yr = if_else(year == 2020, NA_real_, total_ef_l_yr),
    total_pharma_ng = if_else(year == 2020, NA_real_, total_pharma_ng),
    total_pharma_mg= if_else(year == 2020, NA_real_, total_pharma_mg)
  ) %>%
  # interpolate within each region 
  arrange(rgn_id, year) %>%
  group_by(rgn_id) %>%
  mutate(
    total_ef_l_yr = zoo::na.approx(total_ef_l_yr,   x = year, na.rm = FALSE),
    total_pharma_ng = zoo::na.approx(total_pharma_ng, x = year, na.rm = FALSE),
    total_pharma_mg = zoo::na.approx(total_pharma_mg, x = year, na.rm = FALSE),
    interpolated_2020 = if_else(year == 2020, TRUE, FALSE)
  ) %>%
  ungroup()

# sanity check, compare raw vs interpolated for 2020
pourpoints_wwtp_mg_rgnid_interp %>%
  filter(year == 2020) %>%
  select(rgn_id, year,
         total_ef_l_yr_raw, total_ef_l_yr,
         total_pharma_mg_raw, total_pharma_mg)
```


<!-- ## FIXED current status -->

<!-- ```{r} -->
<!-- current_status_wwtp_effluent <- pourpoints_wwtp_mg_rgnid_interp %>%  # use interpolated -->
<!--   mutate( -->
<!--     ideal = 0, -->
<!--     maximum = max(total_pharma_mg), -->
<!--     current_status = 1 - ((total_pharma_mg - ideal) / (maximum - ideal)), -->
<!--     current_status_final = current_status * 100 -->
<!--   ) -->

<!-- ### using the alt maximum, not one value but 5 -->
<!-- top5_avg_mg <- pourpoints_wwtp_mg_rgnid_interp %>% -->
<!--   slice_max(total_pharma_mg, n = 5) %>% -->
<!--   pull(total_pharma_mg) %>% -->
<!--   mean() -->

<!-- current_status_wwtp_effluent_5top_max <- pourpoints_wwtp_mg_rgnid_interp %>%  # use interpolated -->
<!--   mutate( -->
<!--     ideal = 0, -->
<!--     maximum = top5_avg_mg, -->
<!--     current_status = (1 - ((total_pharma_mg - ideal) / (maximum - ideal)))* 100, -->
<!--     current_status_final = ifelse(current_status < 0, 0, current_status) -->
<!--   ) -->

<!-- ## plot the wwtp effluent cs over time -->
<!-- wwtp_cs_plotly <- plotly::plot_ly( -->
<!--   current_status_wwtp_effluent,  -->
<!--   x = ~year,  -->
<!--   y = ~current_status_final,  -->
<!--   color = ~rgn_id,  -->
<!--   colors = rgn_colors, -->
<!--   type = "scatter",  -->
<!--   mode = "lines+markers" -->
<!-- ) %>%  -->
<!--   layout( -->
<!--     title = "Current status of WWTP effluent, potential pharmaceutical chemicals from human wastewater (interpolated 2020)", -->
<!--     xaxis = list(title = "ano // Year"), -->
<!--     yaxis = list(title = "Puntaje // Percent (out of 100)") -->
<!--   ) %>% -->
<!--   add_lines(  -->
<!--     x = range(current_status_wwtp_effluent$year),  -->
<!--     y = c(100, 100), -->
<!--     line = list(color = "black", width = 2, dash = "dash"), -->
<!--     name = "Meta // Goal (100/100)", -->
<!--     inherit = FALSE -->
<!--   ) -->
<!-- wwtp_cs_plotly -->

<!-- ## using the new maximum -->
<!-- wwtp_cs_5top_plotly <- plotly::plot_ly( -->
<!--   current_status_wwtp_effluent_5top_max, -->
<!--   x = ~year, -->
<!--   y = ~current_status_final, -->
<!--   color = ~rgn_id, -->
<!--   colors = rgn_colors, -->
<!--   type = "scatter", -->
<!--   mode = "lines+markers" -->
<!-- ) %>% -->
<!--   layout( -->
<!--     title = "Current status of WWTP effluent, potential pharmaceutical chemicals from human wastewater (interpolated 2020)", -->
<!--     xaxis = list(title = "ano // Year"), -->
<!--     yaxis = list(title = "Puntaje // Percent (out of 100)") -->
<!--   ) %>% -->
<!--   add_lines( -->
<!--     x = range(current_status_wwtp_effluent_5top_max$year), -->
<!--     y = c(100, 100), -->
<!--     line = list(color = "black", width = 2, dash = "dash"), -->
<!--     name = "Meta // Goal (100/100)", -->
<!--     inherit = FALSE -->
<!--   ) -->
<!-- wwtp_cs_5top_plotly -->
<!-- ``` -->

## Save to mazu

```{r save}
total_pharma_mg <- pourpoints_wwtp_mg_rgnid_interp %>% 
  select(rgn_id, year, total_pharma_mg)

write_csv(total_pharma_mg, here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/pharmaceuticals/humans_output/wwtp_mg_rgnid_pharma.csv"))
```


```{r save}
# write_csv(risk_factor_wwtp_effluent, here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/pharmaceuticals/output/risk_factor_wwtp_effluent_pharma_risk.csv"))

write_csv(wwtp_liters_effluent_by_watershed, here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/pharmaceuticals/int/wwtp_liters_effluent_by_watershed.csv"))

write_csv(pourpoints_wwtp_liters_rgnid, here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/pharmaceuticals/int/effluent_wwtp_liters_rgnid.csv"))

st_write(pourpoints_wwtp_liters_final,here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/pharmaceuticals/int/pourpoints_effluent_wwtp_liters_final.gpkg"), delete_dsn = TRUE) # so it can overwrite if changes are made.

# write_csv(current_status_wwtp_effluent, here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/pharmaceuticals/output/current_status_wwtp_effluent_pharma_risk.csv"))
```


