---
title: 'OHI GoC 2025: Clean waters pharmaceuticals'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# CW: Pharmaceuticals - Livestock

## Estimated pharmaceutical discharge

```{r setup}
library(sf)
library(utils)
library(tidyverse)
library(here)
library(raster)
library(mapview)
library(purrr)
library(stringr) # fuzzy string matching
library(stringi)
library(janitor)
library(tibble)
```

### Watersheds and pour points

```{r}
## lets use the full state for now and then summarize by the full watersheds.
goc_ohi_rgns <- st_read(here::here("spatial/ohi_regions","goc_ohi_rgns.shp"))

union_rgns <- goc_ohi_rgns %>% 
  st_union() %>% 
  st_make_valid()

# plot(goc_ohi_rgns)
# plot(union_rgns)

watershed_all <- st_read(file.path("/home/shares/ohi/stressors_2021/_dataprep/nutrients/watersheds_pourpoints/watersheds_all_4326.shp"))
st_crs(watershed_all)
# mapview(watershed_all)

watershed_goc <- st_intersection(st_transform(watershed_all, st_crs(union_rgns)), union_rgns)
mapview(watershed_goc)

# --- if we plan to change the buffer for some reason ---
# inland_50 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer","inland_50km_goc.shp"))
# watershed_goc_50km <- st_intersection(st_transform(watershed_all, st_crs(inland_50)), inland_50)
# mapview(watershed_goc_50km)
```


```{r}
# here's the associated pourpoints that should match with the watersheds:
pourpoints <- st_read("/home/shares/ohi/stressors_2021/_dataprep/nutrients/watersheds_pourpoints/pourpoints_mol.shp")

pourpoints_goc <- st_intersection(st_transform(pourpoints, st_crs(union_rgns)), union_rgns)
mapview(pourpoints_goc)

mapview(pourpoints_goc) + mapview(watershed_goc)
```


## Estimating risk of pharmaceutical discharge based on the intensity of livestock farming in each OHI Region

Here I will plan to aggregate the total number of "cabezas", aka "heads", of all livestock and then rasterize the polygon of each municipio. I will then assume that the production is homogeneously distributed across the municipio, so that every cell has the same value within the municipality. 

I will then summarize the total cabezas in each watershed, and use the associated pourpoints to assign values to OHI regions. I will then aggregate the total livestock produced in each OHI Region, where the more livestock produced means that there is a higher risk of pharmaceutical discharge in that region and year.

### Reading in all livestock reared from 2006-2024 (missing 2023 right now)

Data was downloaded from <https://nube.agricultura.gob.mx/datosAbiertos/Pecuario.php>: 

For each year, click on the associated link for the year, then bring it into Mazu at `~/home/shares/ohi/OHI_GOC/_raw_data/DGSIAP/d2026`.

```{r}
## write function to read in all of the files and limit them to GoC states

read_in_livestock <- function (year) {
  
  df <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/DGSIAP/d2026", paste0("cierre_", year, ".csv"))) %>% 
    
    janitor::clean_names() %>% 
    
    mutate(across(where(is.character), ~iconv(., from = "WINDOWS-1252", to = "UTF-8"))) %>%  # to address encoding issues
    
    filter(nomestado %in% c("Baja California Sur","Baja California","Jalisco","Sinaloa","Sonora","Nayarit")) 
  
  return(df)
  
}
```

Now, we can use the function to read in each year's WWTP data and limit to GoC states.

```{r}
# specify the years we want to iterate through. ideally, we would have 2023 in the future...
years_to_read_in <- c(2006:2022, 2024)

# use `lapply` and bind rows to obtain data over time.
list_of_all_dfs <- lapply(years_to_read_in, read_in_livestock)
  
# bind them across all years
bind_all_years_livestock <- bind_rows(list_of_all_dfs) %>% 
  mutate(cvempio = as.numeric(cvempio),
         cveestado = as.numeric(cveestado))

summary(bind_all_years_livestock)
unique(bind_all_years_livestock$nomestado)
unique(bind_all_years_livestock$cvempio)
colnames(bind_all_years_livestock)

unique(bind_all_years_livestock$nomproducto)
# [1] "Miel"          "Carne"         "Huevo-plato"   "Ganado en pie" "Leche"        
# [6] "Cera"          "Huevo plato"   "Ganado en Pie"

unique(bind_all_years_livestock$cveestado)

```

### Aggregate all livestock cabezas for each municipality and year

Later, come back and determine how to weigh the different types of livestock appropriately.

```{r}
## check that carne is always less than "ganado en pie" (live cattle), so that we know that carne is a subset of ganado en pie?

# for now, just use ganado en pie only
sum_livestock_municipal <- bind_all_years_livestock %>% 
  filter(nomproducto %in% c("Ganado en pie","Ganado en Pie")) %>% 
  group_by(anio, cveestado, nomestado, cvempio, nommunicipio) %>% 
  summarize(toneladas_volumen = sum(volumen, na.rm = TRUE)) %>% 
  ungroup()
```


#### Join with municipality polygons

```{r}
# read them in
municipio_codigo_sf <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/land_admin_boundaries", "mex_admin2_joined_w_municipio_codigos.shp")) %>% 
  mutate(cvempio = as.numeric(clv_mnc),
         cveestado = as.numeric(clv_std))

unique(municipio_codigo_sf$clv_mnc)
unique(municipio_codigo_sf$cveestado)

mapview(municipio_codigo_sf)
```

```{r}
siap_municipal <- inner_join(sum_livestock_municipal, municipio_codigo_sf) %>% 
  st_as_sf()

siap_municipal_2022 <- siap_municipal %>% 
  filter(anio %in% 2022)

mapview(siap_municipal_2022, zcol = "toneladas_volumen")
```









