---
title: 'OHI GoC 2025: Clean waters Pharmaceuticals'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# CW: Pharmaceuticals

Estimated chemical (antimicrobial) discharge from shrimp farming in the Gulf of California.

## Workflow

**For each land use survey year:**

1. Load the aquaculture land use polygons (`usv_aquaculture_goc_allyears.gpkg`) and filter to that year
2. Reproject to match watershed CRS (Mexico ITRF2008 / LCC, meters not degrees)
3. Intersect aquaculture polygons with GoC watersheds, then compute area: `st_area() / 10000` -> hectares per polygon
4. Summarize by `basin_id` x year: sum `aq_area_ha`

**Estimate shrimp production:**

5. Apply the González-Rivas et al. 2023 slope: `aq_area_ha * 1.5` -> estimated tons of shrimp produced per watershed per year (based on R^2 = 0.874 linear relationship from 1993-2021 remote sensing data for Sinaloa and Sonora)

**Spatially:**

6. Join watershed production summaries to GoC pourpoints (`pourpoints_goc`) by `basin_id`
7. Spatially filter pourpoints: keep only those within the 50km inland buffer, and exclude Pacific-side peninsula pourpoints (Region 4) not within the 10km buffer
8. Intersect with OHI regions and aggregate `aq_area_ha` and `shrimp_tons` by `rgn_id` x year

**Convert to mg antimicrobials:**

9. Convert tons -> kg: `shrimp_tons / 907.18474` (note: González-Rivas tons are short tons, not metric tonnes)
10. Multiply by 20.64 mg/kg (WOAH 2025 report, Table 5, North/Central America aquaculture AMU value) -> `mg_antimicrobials` per region per year

11. Move to `script4_current_status.qmd` to aggregate mg from shrimp farming with WWTP effluent and livestock, then calculate current status against the simple avg of the top 5 global maximum values across all regions and years.

*REMEMBER: USV land use polygons (INEGI) = where and how much area, González-Rivas et al. 2023 slope = how much production per hectare, WOAH 2025 = how much antimicrobial load per kg of production!*

## Setup 

```{r setup}
library(sf)
library(utils)
library(tidyverse)
library(here)
library(raster)
library(mapview)
library(purrr)
library(stringr) # fuzzy string matching
library(stringi)
library(janitor)
library(tibble)
```

Watersheds and pour points

```{r}
## lets use the full state for now and then summarize by the full watersheds.
goc_ohi_rgns <- st_read(here::here("spatial/ohi_regions","goc_ohi_rgns.shp"))

union_rgns <- goc_ohi_rgns %>% 
  st_union() %>% 
  st_make_valid()

# plot(goc_ohi_rgns)
# plot(union_rgns)

watershed_all <- st_read(file.path("/home/shares/ohi/stressors_2021/_dataprep/nutrients/watersheds_pourpoints/watersheds_all_4326.shp"))
st_crs(watershed_all)
# mapview(watershed_all)

watershed_goc <- st_intersection(st_transform(watershed_all, st_crs(union_rgns)), union_rgns)
mapview(watershed_goc)

# --- if we plan to change the buffer for some reason ---
# inland_50 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer","inland_50km_goc.shp"))
# watershed_goc_50km <- st_intersection(st_transform(watershed_all, st_crs(inland_50)), inland_50)
# mapview(watershed_goc_50km)
```


```{r}
# here's the associated pourpoints that should match with the watersheds:
pourpoints <- st_read("/home/shares/ohi/stressors_2021/_dataprep/nutrients/watersheds_pourpoints/pourpoints_mol.shp")

pourpoints_goc <- st_intersection(st_transform(pourpoints, st_crs(union_rgns)), union_rgns)
mapview(pourpoints_goc)

#mapview(pourpoints_goc) + mapview(watershed_goc)
```

### Calculate aquacultural area per watershed

#### 1. Load the aquaculture land use polygons (`usv_aquaculture_goc_allyears.gpkg`) and filter to that year

```{r}
# bring in the aquaculture land use over time output:
usv_aq <- st_read(here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/pesticides/int/usv_aquaculture_goc_allyears.gpkg"))
mapview(usv_aq)

unique(usv_aq$year)

usv_aq_2005 <- usv_aq %>% 
  filter(year %in% "2005")
usv_aq_2009 <- usv_aq %>% 
  filter(year %in% "2009")
usv_aq_2013 <- usv_aq %>% 
  filter(year %in% "2013")
usv_aq_2016 <- usv_aq %>% 
  filter(year %in% "2016")
usv_aq_2021 <- usv_aq %>% 
  filter(year %in% "2021")

#mapview(usv_aq_2005, col.regions = "red") + mapview(usv_aq_2009, col.regions = "darkorange") + mapview(usv_aq_2013, col.regions = "orange") + mapview(usv_aq_2016, col.regions = "yellow") + mapview(usv_aq_2021, col.regions = "green")

#mapview(usv_aq_2021, col.regions = "green") + mapview(watershed_goc) + mapview(pourpoints_goc)

```

#### 2. Reproject to match watershed CRS (Mexico ITRF2008 / LCC, meters not degrees)

```{r}
# reproj usv to match watersheds CRS
usv_aq_reproj <- st_transform(usv_aq, st_crs(watershed_goc)) # so that it is in meters not degrees

# Coordinate Reference System:
#   User input: Mexico ITRF2008 / LCC 
#   wkt:
# PROJCRS["Mexico ITRF2008 / LCC",
#     BASEGEOGCRS["Mexico ITRF2008",
#         DATUM["Mexico ITRF2008",
#             ELLIPSOID["GRS 1980",6378137,298.257222101,
#                 LENGTHUNIT["metre",1]]],
```

#### 3. Intersect aquaculture polygons with GoC watersheds, then compute area: `st_area() / 10000` -> hectares per polygon

```{r}
# intersect aq polygons with watersheds
usv_aq_watershed <- st_intersection(usv_aq_reproj, watershed_goc) %>%
  mutate(area_ha = as.numeric(st_area(.)) / 10000)  # there are 0.0001 hectares in 1 square meter, to convert from square meters to hectares, multiply by 0.0001 or divide by 10000
```

Check out 2021:
```{r}
usv_aq_watershed_2021 <- usv_aq_watershed %>% 
  filter(year %in% 2021)

mapview(usv_aq_watershed_2021, zcol = "area_ha") 

mapview(usv_aq_watershed_2021, zcol = "basin_id") + mapview(watershed_goc, zcol = "basin_id")
```

Compute the area:

```{r}
# summarize hectares per watershed per year
aq_area_by_watershed <- usv_aq_watershed %>%
  st_drop_geometry() %>% # maybe take out...
  group_by(basin_id, year) %>% 
  summarize(aq_area_ha = sum(area_ha, na.rm = TRUE)) %>% 
  ungroup() %>% 
  mutate(year = as.numeric(year))
```

### Estimating shrimp production using production per hectare

González-Rivas et al. 2023, [Estimating the shrimp farm’s production and their future growth prediction by remote sensing: Case study Gulf of California](https://www.frontiersin.org/journals/marine-science/articles/10.3389/fmars.2023.1130125/full), ran a "linear regression model between the shrimp farms area, calculated using remote sensing, and statistical data on shrimp production" which was "strongly positive and significant (r2 = 0.874, p =2.209e-11)."

**Useful figures from this paper:**

- Figure 1, "Total estimated area of shrimp ponds (ha, blue line) based on remote sensing and shrimp production (tons, dotted orange line) from 1993 to 2021, for the east part of the Gulf of California corresponding to the Mexican states of Sinaloa and Sonora."

- Figure 3, "Relationship between the total shrimp pond area estimated using remote sensing (ha/year) and (A) the corresponding production (ton/year) and (B) time (years)."

Based on this linear relationship with a strong positive and significant R^2, we can use the slope between shrimp pond area (hectare/year) and corresponding production (tons/year), which is 1.5 tons/ha, to estimate the total production from each OHI region over time as land use changes.  Note that because this is a linear relationship, the slope is not changing over time. This slope was determined based on years 1993-2021.

#### 5. Apply the González-Rivas et al. 2023 slope

Remember that `aq_area_ha * 1.5` -> estimated tons of shrimp produced per watershed per year (based on R^2 = 0.874 linear relationship from 1993-2021 remote sensing data for Sinaloa and Sonora).

```{r}
aq_production_by_watershed <- aq_area_by_watershed %>% 
  mutate(
    shrimp_prod_tons = aq_area_ha * 1.5 # ha * tons/ha = tons
  )
```


#### 6. Join watershed production summaries to GoC pourpoints (`pourpoints_goc`) by `basin_id`

```{r pourpoints}
pourpoints_shrimp_prod <- pourpoints_goc %>%
  dplyr::select(basin_id) %>%
  left_join(aq_production_by_watershed, by = "basin_id")
```

#### 7. Spatially filter pourpoints

```{r ohi_marine_and_inland}
inland_50 <- st_read(here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer", "inland_50km_goc.shp"))
inland_10 <- st_read(here("/home/shares/ohi/OHI_GOC/spatial/inland_10km_buffer", "inland_10km_goc.shp"))
regions_marine <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp")) %>%
  filter(location == "marine")

marine_and_inland50 <- bind_rows(regions_marine, inland_50) %>%
  dplyr::select(rgn_id, geometry) %>%
  group_by(rgn_id) %>%
  summarize(geometry = st_union(geometry), .groups = "drop")

marine_and_inland10 <- bind_rows(regions_marine, inland_10) %>%
  dplyr::select(rgn_id, geometry) %>%
  group_by(rgn_id) %>%
  summarize(geometry = st_union(geometry), .groups = "drop")
```


```{r pourpoints}
pourpoints_shrimp_prod_final <- pourpoints_shrimp_prod %>%
  filter(!is.na(shrimp_prod_tons), shrimp_prod_tons > 0) %>% # note that the watershed with the id 73730 wraps around, and that the pour point is only on the left side...not sure that is correct.
  st_intersection(marine_and_inland50) %>% # keep those within 50 km...
  mutate(is_within_10km = lengths(st_within(., marine_and_inland10)) > 0) %>% # but make sure not to keep the ones on the pacific side of the peninsula.
  filter(!(is_within_10km == FALSE & rgn_id == "Region_4"))

mapview(pourpoints_shrimp_prod_final, zcol = "shrimp_prod_tons") + mapview(usv_aq_2021, col.regions = "red") + mapview(watershed_goc, col.regions = "pink")
```


#### 8. Intersect with OHI regions and aggregate `aq_area_ha` and `shrimp_tons` by `rgn_id` x year

```{r pourpoints}
pourpoints_shrimp_prod_rgnid <- st_intersection(pourpoints_shrimp_prod_final, goc_ohi_rgns) %>%
  st_drop_geometry() %>%
  group_by(rgn_id, year) %>%
  summarize(
    aq_area_ha = sum(aq_area_ha, na.rm = TRUE),
    shrimp_tons = sum(shrimp_prod_tons, na.rm = TRUE),
    .groups = "drop"
  )

pourpoints_shrimp_prod_rgnid
```

#### 9-10. Convert to mg

Now that we have an estimated tons/year of production, lets convert the production to kg, then use that to estimate the mg of pharmaceuticals used in shrimp farming. 

*__ kg total volume shrimp production × 20.64 mg/kg of antimicrobials  = __ mg of antimicriobials applied that could potentially discharge into the GoC eventually.*

^ This number was determined based on Table 5 from WOAH <https://www.woah.org/app/uploads/2025/05/en-woah-amu-report-2025-final.pdf>.

Essentially, we will be converting tons -> kg: `shrimp_tons / 907.18474` (note: González-Rivas tons are short tons, not metric tonnes), then multiplying by 20.64 mg/kg (WOAH 2025 report, Table 5, North/Central America aquaculture AMU value) -> `mg_antimicrobials` per region per year.

```{r}
pourpoints_shrimp_prod_rgnid <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/pharmaceuticals/shrimp_int/shrimp_tons_pourpoints_prod_rgnid.csv"))

pourpoints_shrimp_mg_rgnid <- pourpoints_shrimp_prod_rgnid %>% 
  group_by(rgn_id, year) %>% 
  mutate(
    shrimp_kg = shrimp_tons * 907.18474, # remember that 1 ton = 907.18474 kg, the 2023 paper was not in tonnes
    mg_antimicrobials = shrimp_kg * 20.64 # see above, Table 5 from 2025 WOAH report
  )
summary(pourpoints_shrimp_mg_rgnid)
```

#### 11. Interpolate/gapfill to all years 2005-2023

```{r}
pourpoints_shrimp_mg_rgnid_interp <- pourpoints_shrimp_mg_rgnid %>%
  group_by(rgn_id) %>% # interpolate, expand each region to full year
  complete(year = 2005:2023) %>%  # creates rows for missing years
  arrange(rgn_id, year) %>%
  mutate(
    
    # linear interpolation between known survey years
    aq_area_ha = zoo::na.approx(aq_area_ha, x = year, na.rm = FALSE),
    shrimp_tons = zoo::na.approx(shrimp_tons, x = year, na.rm = FALSE),
    shrimp_kg = zoo::na.approx(shrimp_kg, x = year, na.rm = FALSE),
    mg_antimicrobials = zoo::na.approx(mg_antimicrobials, x = year, na.rm = FALSE),
    
    # "extrapolate"/gapfill 2022-2023 by carrying forward the 2021 value.....
    aq_area_ha = zoo::na.locf(aq_area_ha, na.rm = FALSE),
    shrimp_tons = zoo::na.locf(shrimp_tons, na.rm = FALSE),
    shrimp_kg = zoo::na.locf(shrimp_kg, na.rm = FALSE),
    mg_antimicrobials = zoo::na.locf(mg_antimicrobials, na.rm = FALSE)
  ) %>%
  ungroup()

# check
summary(pourpoints_shrimp_mg_rgnid_interp)
check_interp <- pourpoints_shrimp_mg_rgnid_interp %>% 
  filter(year >= 2017)
```


## Save to mazu

```{r save}
# total mg per region and year
write_csv(pourpoints_shrimp_mg_rgnid_interp, here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/pharmaceuticals/shrimp_output/shrimp_mg_pharmaceuticals_rgnid.csv"))


# all other int
st_write(pourpoints_shrimp_prod_final, here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/pharmaceuticals/shrimp_int/shrimp_tons_pourpoints_prod_final.gpkg"))
write_csv(aq_production_by_watershed, here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/pharmaceuticals/shrimp_int/aq_production_tons_shrimp_by_watershed.csv"))
write_csv(pourpoints_shrimp_prod_rgnid, here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/pharmaceuticals/shrimp_int/shrimp_tons_pourpoints_prod_rgnid.csv"))
```



