---
title: 'OHI GoC 2025: Clean waters pharmaceuticals'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# CW: Pharmaceuticals - WWTP Effluent

*Alfonso-Muniozguren P, Serna-Galvis EA, Bussemaker M, Torres-Palma RA, Lee J. A review on pharmaceuticals removal from waters by single and combined biological, membrane filtration and ultrasound systems. Ultrason Sonochem. 2021 Aug;76:105656. doi: 10.1016/j.ultsonch.2021.105656. Epub 2021 Jul 1. PMID: 34274706; PMCID: PMC8319449.*

"Currently the biggest threat to the global freshwater reserve and, ecological and public health is the rising of emerging contaminants in the environment such as pharmaceutical drugs [2](https://doi.org/10.1016/j.scitotenv.2013.12.065), [3](https://doi.org/10.1007/s10661-017-6097-x). Wastewater treatment plants (WWTPs) are considered the primary source of pharmaceuticals in the environment, although a higher concentration of certain compounds can be found in hospital effluents [4](https://doi.org/10.1007/s11356-017-9662-5). Current water treatment methods implemented in WWTPs are inefficient for the complete removal of most pharmaceuticals [2](https://doi.org/10.1016/j.scitotenv.2013.12.065), [5](https://doi.org/10.1016/j.watres.2012.11.027), leading to the discharge of these compounds directly into water bodies. The concentration of pharmaceuticals in hospital effluents [4](https://doi.org/10.1007/s11356-017-9662-5), [6](https://doi.org/10.1021/es400708w), [7](https://doi.org/10.1007/s11356-016-6389-7), pharmaceutical industry effluents [8](https://pmc.ncbi.nlm.nih.gov/articles/PMC8319449/#b0040), [9](https://scholar.google.com/scholar_lookup?journal=Ind.%20Eng.%20Chem.%20Res.&title=Pharmaceutical%20Industry%20Wastewater:%20Review%20of%20the%20Technologies%20for%20Water%20Treatment%20and%20Reuse&author=C.%20Gadipelly&author=A.%20P%C3%A9rez-Gonz%C3%A1lez&author=G.D.%20Yadav&author=I.%20Ortiz&author=R.%20Ib%C3%A1%C3%B1ez&volume=53&publication_year=2014&pages=11571-11592&) and WWTPs [10](https://doi.org/10.1016/j.watres.2009.07.004), [11](https://pubmed.ncbi.nlm.nih.gov/15497842/) range from ng L−1 to µg L−1. The concentration of these compounds in the environment depends on discharge, geographical location, climate conditions and consumption [12](https://doi.org/10.1016/j.envint.2006.09.014)."

## Estimated pharmaceutical discharge

We will be doing this by estimating the total liters per year of effluent from each WWTP across the Gulf, determining which watershed the WWTP falls into, and aggregate the liters per year at the associated pourpoints.  

From there, we can classify which OHI Region each pourpoint falls into, and aggregate the total liters per year of WWTP effluent for each year and region.

We calculate **two** pressure metrics:

1. **Volume pressure** (`total_ef_l_yr`) — effluent volume only (original approach).
2. **Pharmaceutical-weighted pressure** (`total_ef_pharma_l_yr`) — effluent volume × literature-based expected micropollutant **retention** in effluent (by SINA `proceso`) × optional overload multiplier when `caudal > capacidad`.

The regions are compared across time using the global historical maximum (across-Gulf) for each metric. The year–region at each maximum receives a risk factor of 100 and a current status of 0. Everything else is scaled accordingly, with ideal = zero effluent (volume or weighted volume). 

```{r setup}
library(sf)
library(utils)
library(tidyverse)
library(here)
library(raster)
library(mapview)
library(purrr)
library(stringr) # fuzzy string matching
library(stringi)
library(janitor)
library(tibble)
```

### Watersheds and pour points

First, let's isolate the watersheds and associated pourpoints in the Gulf of California.

```{r}
## lets use the full state for now and then summarize by the full watersheds.
goc_ohi_rgns <- st_read(here::here("spatial/ohi_regions","goc_ohi_rgns.shp"))

union_rgns <- goc_ohi_rgns %>% 
  st_union() %>% 
  st_make_valid()

# plot(goc_ohi_rgns)
# plot(union_rgns)

watershed_all <- st_read(file.path("/home/shares/ohi/stressors_2021/_dataprep/nutrients/watersheds_pourpoints/watersheds_all_4326.shp"))
st_crs(watershed_all)
# mapview(watershed_all)

watershed_goc <- st_intersection(st_transform(watershed_all, st_crs(union_rgns)), union_rgns)
# mapview(watershed_goc)

# --- if we plan to change the buffer for some reason ---
# inland_50 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer","inland_50km_goc.shp"))
# watershed_goc_50km <- st_intersection(st_transform(watershed_all, st_crs(inland_50)), inland_50)
# mapview(watershed_goc_50km)
```

```{r}
# here's the associated pourpoints that should match with the watersheds:
pourpoints <- st_read("/home/shares/ohi/stressors_2021/_dataprep/nutrients/watersheds_pourpoints/pourpoints_mol.shp")

pourpoints_goc <- st_intersection(st_transform(pourpoints, st_crs(union_rgns)), union_rgns)
mapview(pourpoints_goc)

# mapview(pourpoints_goc) + mapview(watershed_goc)
```

### Reading in all WWTP capacities and usages from 2017-2023

Data was downloaded from <https://sinav30.conagua.gob.mx:8080/SINA/?opcion=salud#infrah>: 

(1) click on "Infraestructura Hidráulica", 

(2) Click on "Plantas de Tratamiento de Aguas Residuales", 

(3) For each year, go through this process: select the year, click PTAR (the acronym in Spanish, it would be WWTPs in English), and then select the next year and do the same thing, which ensures that the data is shown on the interactive map,

(4) Once all layers from all years are showing, go to the bottom left-hand corner and click the button that is meant to represent "see all layers currently on the map".  It has an eye and multiple squares, looking like layered papers.  It is located just above the question mark button.  Here, you will go one by one, and download each of the years by clicking on the button representing ".SHP".  Because the .zip files are all named the same, regardless of the year, you will need to rename each folder as you go before adding it to `~/home/shares/ohi/OHI_GOC/_raw_data/SINA_3_CONAGUA/d2025/wastewater_treatment_plants`.  Once you unzip each folder, add the appropriate year in the end of the folder name, for example: "_2020".

```{r}
## write function to read in all of the shapefiles and limit them to GoC states, as well as determine whether they were over capacity for that year and WWTP
read_in_wwtp <- function (year) {
  
  df <- st_read(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SINA_3_CONAGUA/d2025/wastewater_treatment_plants", paste0("m29_vwplantastarm_", year))) %>% 
    mutate(across(where(is.character), ~iconv(., from = "WINDOWS-1252", to = "UTF-8"))) %>%  # to address encoding issues
    filter(estado %in% c("Baja California Sur","Baja California","Jalisco","Sinaloa","Sonora","Nayarit")) %>% 
    mutate(over_proc = case_when(
      caudal > capacidad ~ TRUE,
      TRUE ~ FALSE
    ))
  
  return(df)
  
}
```

Now, we can use the function to read in each year's WWTP data and limit to GoC states.

```{r}
# specify the years we want to iterate through.  Change this in the future if necessary.
years_to_read_in <- 2017:2023

# use `lapply` and bind rows to obtain data over time.
list_of_all_dfs <- lapply(years_to_read_in, read_in_wwtp)
  
# bind them across all years
bind_all_years_wwtp <- bind_rows(list_of_all_dfs)

summary(bind_all_years_wwtp)
unique(bind_all_years_wwtp$estado)
procesos <- unique(bind_all_years_wwtp$proceso)
```

### Literature-based retention factors (by `proceso`)

SINA reports each plant's dominant treatment process in `proceso`. We join `proceso_retention_lookup.csv` (derived from `wwtp_pharmaceutical_removal_by_process.csv`) to assign a conservative **retention** fraction — the share of influent pharmaceutical load expected to remain in effluent after treatment (not measured removal).

Retention by overall literature rating: Good = 0.30, Moderate = 0.45, Limited = 0.75, Variable/Unknown = 0.50 (with low/high bounds for sensitivity). Plants operating over capacity (`over_proc`) use `overproc_multiplier` (default 1.25).

```{r retention-lookup}
seconds_per_year <- 31536000

proceso_retention_candidates <- c(
  here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/pharmaceuticals/treatment_process_differences/proceso_retention_lookup.csv")
)

proceso_retention_lookup <- read_csv(proceso_retention_candidates)

retention_defaults <- proceso_retention_lookup %>%
  filter(treatment_process == "Other") %>%
  slice(1)

# helper: join retention + compute volume and pharma-weighted effluent (L/yr)
add_pharma_effluent <- function(df) {
  df %>%
    left_join(
      proceso_retention_lookup %>%
        dplyr::select(
          proceso,
          treatment_process_mapped = treatment_process,
          pharmaceutical_removal_rating,
          retention_pharma,
          retention_pharma_low,
          retention_pharma_high,
          overproc_multiplier
        ),
      by = "proceso"
    ) %>%
    mutate(
      proceso_mapped = !is.na(retention_pharma),
      pharmaceutical_removal_rating = coalesce(
        pharmaceutical_removal_rating,
        retention_defaults$pharmaceutical_removal_rating
      ),
      treatment_process_mapped = coalesce(
        treatment_process_mapped,
        retention_defaults$treatment_process
      ),
      retention_pharma = coalesce(retention_pharma, retention_defaults$retention_pharma),
      retention_pharma_low = coalesce(
        retention_pharma_low,
        retention_defaults$retention_pharma_low
      ),
      retention_pharma_high = coalesce(
        retention_pharma_high,
        retention_defaults$retention_pharma_high
      ),
      overproc_multiplier = coalesce(
        overproc_multiplier,
        retention_defaults$overproc_multiplier
      ),
      overproc_mult = if_else(over_proc %in% TRUE, overproc_multiplier, 1),
      ef_l_yr = caudal * seconds_per_year,
      ef_pharma_l_yr = ef_l_yr * retention_pharma * overproc_mult,
      ef_pharma_l_yr_low = ef_l_yr * retention_pharma_low * overproc_mult,
      ef_pharma_l_yr_high = ef_l_yr * retention_pharma_high * overproc_mult
    )
}

# audit: SINA proceso values not in lookup (before defaulting to Other)
proceso_not_in_lookup <- bind_all_years_wwtp %>%
  st_drop_geometry() %>%
  distinct(proceso) %>%
  anti_join(proceso_retention_lookup, by = "proceso")

if (nrow(proceso_not_in_lookup) > 0) {
  message(
    "SINA proceso values not in lookup (assigned Other retention): ",
    paste(proceso_not_in_lookup$proceso, collapse = "; ")
  )
}
```

Take a look...

```{r}
# mapview(bind_all_years_wwtp, zcol = "over_proc")
```


#### Understanding WWTPs over time

This is just some exploratory stats to understand the landscape better.

```{r}
pct_over_processing <- bind_all_years_wwtp %>% 
  st_drop_geometry() %>% 
  group_by(anio, estado) %>% 
  summarize(
    total_wwtp = n(),
    num_over_proc = sum(over_proc, na.rm = TRUE),
    pct_over_proc = round(num_over_proc / total_wwtp * 100, 1),
    mean_caudal = mean(caudal, na.rm = TRUE),
    mean_capacidad = mean(capacidad, na.rm = TRUE),
    .groups = "drop"
  )

number_of_wwtp <- bind_all_years_wwtp %>% 
  st_drop_geometry() %>% 
  group_by(anio, estado) %>% 
  summarize(
    num_wwtp = n()
  ) %>% 
  ungroup()
```

### Assign to watersheds

Now that we have data about the capacity and the actual L/s of each WWTP, let's assign each WWTP to its respective watershed.

```{r}
# reproject to match watersheds CRS
bind_all_years_wwtp_reproj <- st_transform(bind_all_years_wwtp, st_crs(watershed_goc))

# Coordinate Reference System:
#   User input: Mexico ITRF2008 / LCC 
#   wkt:
# PROJCRS["Mexico ITRF2008 / LCC",
#     BASEGEOGCRS["Mexico ITRF2008",
#         DATUM["Mexico ITRF2008",
#             ELLIPSOID["GRS 1980",6378137,298.257222101,
#                 LENGTHUNIT["metre",1]]],

# intersect wwtps with watersheds
wwtps_watershed <- st_intersection(bind_all_years_wwtp_reproj, watershed_goc)

wwtps_watershed <- wwtps_watershed %>%
  add_pharma_effluent()

colnames(wwtps_watershed)
```

Check out 2021:
```{r}
wwtps_watershed_2021 <- wwtps_watershed %>% 
  filter(anio %in% 2021)

# mapview(wwtps_watershed_2021, zcol = "capacidad") 

# mapview(wwtps_watershed_2021, zcol = "basin_id") + mapview(watershed_goc, zcol = "basin_id")
```


#### Exploring liters of effluent calculation

To go from L/s to L/year, multiply by `seconds_per_year` (31,536,000). This assumes the WWTP runs continuously — an upper bound for the year. Pharmaceutical-weighted effluent applies literature retention and an overload multiplier when `over_proc` is TRUE.

```{r}
summary(bind_all_years_wwtp$caudal) # no zeros or NAs:      Min.   1st Qu.    Median      Mean   3rd Qu.      Max. 
                                                        #  0.010     1.377     3.100    53.311    15.100     28743.230 

effluent_liters_wwtp_yearly <- wwtps_watershed %>%
  st_drop_geometry() %>%
  dplyr::select(
    clvplanta, anio, estado, proceso, treatment_process_mapped,
    pharmaceutical_removal_rating, retention_pharma, over_proc, overproc_mult,
    caudal, capacidad, ef_l_yr, ef_pharma_l_yr,
    ef_pharma_l_yr_low, ef_pharma_l_yr_high, proceso_mapped
  )
```

#### Summarize liters per watershed per year

Aggregate total effluent volume and pharmaceutical-weighted effluent per watershed.

```{r}
# summarize liters per watershed per year
wwtp_liters_effluent_by_watershed <- wwtps_watershed %>%
  st_drop_geometry() %>%
  group_by(basin_id, anio) %>%
  summarize(
    total_wwtp = n(),
    num_over_proc = sum(over_proc, na.rm = TRUE),
    pct_over_proc = num_over_proc / total_wwtp * 100,
    num_proceso_unmapped = sum(!proceso_mapped, na.rm = TRUE),
    total_ef_l_yr = sum(ef_l_yr, na.rm = TRUE),
    total_ef_pharma_l_yr = sum(ef_pharma_l_yr, na.rm = TRUE),
    total_ef_pharma_l_yr_low = sum(ef_pharma_l_yr_low, na.rm = TRUE),
    total_ef_pharma_l_yr_high = sum(ef_pharma_l_yr_high, na.rm = TRUE),
    total_ef_m3_yr = total_ef_l_yr / 1000,
    total_ef_pharma_m3_yr = total_ef_pharma_l_yr / 1000,
    .groups = "drop"
  ) %>%
  mutate(year = as.numeric(anio))
```

### Pour points and OHI regions

And now we can take that to the pourpoints!

```{r pourpoints}
pourpoints_wwtp_liters <- pourpoints_goc %>%
  dplyr::select(basin_id) %>%
  left_join(wwtp_liters_effluent_by_watershed, by = "basin_id")

# mapview(pourpoints_wwtp_liters)
```

*This is just so that we are not losing any WWTPs that may be on the coastal border:*

```{r ohi_marine_and_inland}
inland_50 <- st_read(here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer", "inland_50km_goc.shp"))
inland_10 <- st_read(here("/home/shares/ohi/OHI_GOC/spatial/inland_10km_buffer", "inland_10km_goc.shp"))
regions_marine <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp")) %>%
  filter(location == "marine")

marine_and_inland50 <- bind_rows(regions_marine, inland_50) %>%
  dplyr::select(rgn_id, geometry) %>%
  group_by(rgn_id) %>%
  summarize(geometry = st_union(geometry), .groups = "drop")

marine_and_inland10 <- bind_rows(regions_marine, inland_10) %>%
  dplyr::select(rgn_id, geometry) %>%
  group_by(rgn_id) %>%
  summarize(geometry = st_union(geometry), .groups = "drop")
```

Now we can ensure that we only keep the pourpoints in the GoC that have pharmaceutical effluent associated with them.

```{r pourpoints}
pourpoints_wwtp_liters_final <- pourpoints_wwtp_liters %>%
  filter(!is.na(total_ef_l_yr), total_ef_l_yr > 0) %>% # note that the watershed with the id 73730 wraps around, and that the pour point is only on the left side...not sure that is correct.
  st_intersection(marine_and_inland50) %>% # keep those within 50 km...
  mutate(is_within_10km = lengths(st_within(., marine_and_inland10)) > 0) %>% # but make sure *not* to keep the ones on the pacific side of the peninsula.
  filter(!(is_within_10km == "FALSE" & rgn_id %in% c("Region_4","Region_5","Region_9")))

# ensure that all of the pourpoints outside of the Gulf are not included.
# mapview(pourpoints_wwtp_liters_final, zcol = "total_ef_l_yr") + mapview(wwtps_watershed, col.regions = "red") + mapview(watershed_goc, col.regions = "pink")
```

And now we can aggregate by region to determine the total L/year and region.

```{r pourpoints}
pourpoints_wwtp_liters_rgnid <- st_intersection(pourpoints_wwtp_liters_final, goc_ohi_rgns) %>%
  st_drop_geometry() %>%
  group_by(rgn_id, year) %>%
  summarize(
    total_ef_l_yr = sum(total_ef_l_yr, na.rm = TRUE),
    total_ef_pharma_l_yr = sum(total_ef_pharma_l_yr, na.rm = TRUE),
    total_ef_pharma_l_yr_low = sum(total_ef_pharma_l_yr_low, na.rm = TRUE),
    total_ef_pharma_l_yr_high = sum(total_ef_pharma_l_yr_high, na.rm = TRUE),
    total_ef_m3_yr = sum(total_ef_m3_yr, na.rm = TRUE),
    total_ef_pharma_m3_yr = sum(total_ef_pharma_m3_yr, na.rm = TRUE),
    total_wwtp = sum(total_wwtp, na.rm = TRUE),
    num_over_proc = sum(num_over_proc, na.rm = TRUE),
    pct_over_proc = num_over_proc / total_wwtp * 100,
    .groups = "drop"
  )

pourpoints_wwtp_liters_rgnid
```

### Calculate risk factor

The risk factor is pressure on the system; current status is its inverse (higher status = lower pressure).

**Volume metric** (primary, SINA-native):

$$current\ status\ _{effluent\_volume} = 1 - \frac{total\_ef\_l\_yr - 0}{max(total\_ef\_l\_yr) - 0}$$

**Pharmaceutical-weighted metric** (secondary, literature retention by `proceso`):

$$current\ status\ _{effluent\_pharma} = 1 - \frac{total\_ef\_pharma\_l\_yr - 0}{max(total\_ef\_pharma\_l\_yr) - 0}$$

```{r}
calc_current_status <- function(df, pressure_col, prefix) {
  ideal <- 0
  maximum <- max(df[[pressure_col]], na.rm = TRUE)
  risk_col <- paste0("risk_factor_", prefix)
  status_col <- paste0("current_status_", prefix)
  status_final_col <- paste0("current_status_final_", prefix)
  maximum_col <- paste0("maximum_", prefix)

  df %>%
    mutate(
      ideal = ideal,
      !!maximum_col := maximum,
      !!risk_col := (.data[[pressure_col]] - ideal) / (maximum - ideal),
      !!status_col := 1 - .data[[risk_col]],
      !!status_final_col := .data[[status_col]] * 100
    )
}

current_status_wwtp_effluent <- pourpoints_wwtp_liters_rgnid %>%
  calc_current_status("total_ef_l_yr", "volume") %>%
  calc_current_status("total_ef_pharma_l_yr", "pharma")

# backward-compatible column names (volume = original single metric)
current_status_wwtp_effluent <- current_status_wwtp_effluent %>%
  rename(
    maximum = maximum_volume,
    current_status = current_status_volume,
    current_status_final = current_status_final_volume,
    risk_factor = risk_factor_volume
  )

risk_factor_wwtp_effluent <- current_status_wwtp_effluent
```

#### plotly

Plot both the current status, but also the L/year across all regions and years to understand how the current status was derived.

```{r}
library(plotly)

rgn_colors <- c(
  Region_1 = "#C03729",
  Region_2 = "#E68C7C",
  Region_3 = "#FC8F24",
  Region_4 = "#dcaa38",
  Region_5 = "#cbd2a0",
  Region_6 = "#98a25a",
  Region_7 = "#747428",
  Region_8 = "#627391",
  Region_9 = "#8d61a9"
)

## plot the wwtp effluent current status over time (volume and pharma-weighted)
wwtp_plotly_volume <- plotly::plot_ly(
  current_status_wwtp_effluent,
  x = ~year,
  y = ~current_status_final_volume,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines+markers"
) %>%
  layout(
    title = "Clean Waters: current status — WWTP effluent volume (primary)",
    xaxis = list(title = "ano // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(current_status_wwtp_effluent$year),
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
wwtp_plotly_volume

wwtp_plotly_pharma <- plotly::plot_ly(
  current_status_wwtp_effluent,
  x = ~year,
  y = ~current_status_final_pharma,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines+markers"
) %>%
  layout(
    title = "Clean Waters: current status — pharma-weighted effluent (literature retention by proceso)",
    xaxis = list(title = "ano // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(current_status_wwtp_effluent$year),
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
wwtp_plotly_pharma

# ---

## plot the liters per year of effluent for all rgns
rgn_eff_plotly <- plotly::plot_ly(
  pourpoints_wwtp_liters_rgnid, 
  x = ~year, 
  y = ~total_ef_l_yr, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Clean Waters: liters per year effluent from WWTPs",
    xaxis = list(title = "ano // Year"),
    yaxis = list(title = "Liters of effluent at pourpoints within the region")
  )
rgn_eff_plotly
```

#### Examining the spike in 2020 found in Region 8 and 9
```{r}
wwtps_watershed_na_80866_2020 <- wwtps_watershed %>% 
  filter(basin_id %in% "na_80866") %>% 
  filter(anio %in% 2020)

wwtps_watershed_na_80866_2019 <- wwtps_watershed %>% 
  filter(basin_id %in% "na_80866") %>% 
  filter(anio %in% 2019)

# mapview(wwtps_watershed_na_80866_2020, zcol = "total_ef_l_yr") # look at it over time, see if 2020 has any outlier or inconsistent internal results over time.  Determine what is going on there!! Is it a mistake???
# mapview(wwtps_watershed_na_80866_2019, zcol = "total_ef_l_yr")
```

```{r}
## isolate region 8 and 9 basins
pourpoints_wwtp_liters_final

rgn_8_9_basins <- pourpoints_wwtp_liters_final %>% 
  filter(rgn_id %in% c("Region_8","Region_9")) %>% 
  st_drop_geometry()

## plot the liters per year of effluent for these basins
inspecting_8_9_plotly <- plotly::plot_ly(
  rgn_8_9_basins, 
  x = ~year, 
  y = ~total_ef_l_yr, 
  color = ~basin_id, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "L/y of effluent from each watershed over time in regions 8 and 9",
    xaxis = list(title = "ano // Year"),
    yaxis = list(title = "Liters of effluent at pourpoints for each basin")
  )
inspecting_8_9_plotly
```


```{r}
### so na_76895 has the spike, and na_80866 has the reduction in 2020.
basins_76895_80866 <- pourpoints_wwtp_liters_final %>% 
  filter(basin_id %in% c("na_76895","na_80866")) 

### okay, so let's see which WWTPs in the basins caused the spikes.
wwtps_watershed_only_basins_76895_80866 <- wwtps_watershed %>% 
  st_drop_geometry() %>% 
  group_by(clvplanta, anio, caudal) %>% 
  distinct() %>% 
  filter(basin_id %in% c("na_76895","na_80866")) %>%
  mutate(year = as.numeric(anio)) %>%
  ungroup()

## plot the liters per year of effluent for these basins
inspecting_basins_76895_80866_plotly <- plotly::plot_ly(
  wwtps_watershed_only_basins_76895_80866, 
  x = ~year, 
  y = ~ef_l_yr, 
  color = ~clvplanta, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "L/y of effluent from all WWTPs in basins na_76895 and na_80866",
    xaxis = list(title = "ano // Year"),
    yaxis = list(title = "Liters of effluent for each WWTP over time")
  )
inspecting_basins_76895_80866_plotly
```


```{r}
## it appears that clvplanta p_25-4117 has the spike in 2020, while p_14-4871 and p_14-3656 had the drastic decreases in 2020.

wwtps_watershed_only_clvplanta_outliers <- wwtps_watershed_only_basins_76895_80866 %>% 
  filter(clvplanta %in% c("p_25-4177","p_14-4871","p_14-3656"))


## plot the liters per year of effluent for these plants
inspecting_basins_76895_80866_plotly <- plotly::plot_ly(
  wwtps_watershed_only_basins_76895_80866, 
  x = ~year, 
  y = ~ef_l_yr, 
  color = ~clvplanta, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "L/y of effluent from all WWTPs in basins na_76895 and na_80866",
    xaxis = list(title = "ano // Year"),
    yaxis = list(title = "Liters of effluent for each WWTP over time")
  )
inspecting_basins_76895_80866_plotly
```

#### Flagging anomalies

Check for 2020 anomalies across all WWTPs in the GoC

```{r check-2020-anomalies}
# for each WWTP (clvplanta), find the median caudal across non-2020 years, then flag 2020 rows where caudal deviates by more than 3x (or is zero/NA when it shouldnt be)
wwtps_2020_anomaly_check <- wwtps_watershed %>%
  st_drop_geometry() %>%
  group_by(clvplanta, anio, caudal, proceso, capacidad, basin_id) %>%
  distinct() %>%
  mutate(year = as.numeric(anio)) %>%
  ungroup() %>%
  group_by(clvplanta) %>%
  mutate(
    median_caudal_non2020 = median(caudal[year != 2020], na.rm = TRUE), # to see how much higher/lower the L/s processing jumped or fell
    caudal_ratio_2020 = ifelse(
      year == 2020 & !is.na(median_caudal_non2020) & median_caudal_non2020 > 0,
      caudal / median_caudal_non2020,
      NA_real_
    ),
    proceso_2020 = proceso[year == 2020][1],
    proceso_modal = names(sort(table(proceso[year != 2020]), decreasing = TRUE))[1],
    proceso_changed_2020 = ifelse(proceso_2020 != proceso_modal, TRUE, FALSE)
  ) %>%
  ungroup()

## pull out just the 2020 rows that look suspicious: ex, caudal is >3x or < (1/3) x the plant's individual non-2020 median, OR treatment process changed
anomalies_2020 <- wwtps_2020_anomaly_check %>%
  filter(year == 2020) %>%
  filter(
    (caudal_ratio_2020 > 3 | caudal_ratio_2020 < 0.33) | proceso_changed_2020 %in% "TRUE" # 1/3 = 0.33
  ) %>%
  arrange(basin_id, clvplanta) %>%
  select(basin_id, clvplanta, municipio, localidad,
         proceso, proceso_modal, capacidad,
         caudal, median_caudal_non2020, caudal_ratio_2020,
         proceso_changed_2020)

anomalies_all_yrs <- wwtps_2020_anomaly_check %>%
  filter(
    (caudal_ratio_2020 > 3 | caudal_ratio_2020 < 0.33) | proceso_changed_2020 == TRUE # 1/3 = 0.33
  ) %>%
  arrange(basin_id, clvplanta) %>%
  select(basin_id, clvplanta, municipio, localidad,
         proceso, proceso_modal, capacidad,
         caudal, median_caudal_non2020, caudal_ratio_2020,
         proceso_changed_2020)


## how many unique plants are flagged?
cat("how many flagged WWTPs in 2020:", n_distinct(anomalies_2020$clvplanta), "\n") #258
cat("how many basins:", n_distinct(anomalies_2020$basin_id), "\n") # 76

## interactive plot of ALL flagged plants over time for visual confirmation!!
flagged_plants <- unique(anomalies_2020$clvplanta)

wwtps_flagged_timeseries <- wwtps_2020_anomaly_check %>%
  filter(clvplanta %in% flagged_plants)

wwtps_flagged_timeseries_plotly <- plotly::plot_ly(
  wwtps_flagged_timeseries,
  x = ~year,
  y = ~ef_l_yr,
  color = ~clvplanta,
  text = ~paste("Basin:", basin_id, "<br>Proceso:", proceso,
                  "<br>Capacidad:", capacidad, "<br>Caudal:", caudal),
  type = "scatter",
  mode = "lines+markers",
  hoverinfo = "text+x+y"
) %>%
  layout(
    title = "All flagged WWTPs with potential 2020 anomalies",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Estimated L/yr of effluent"),
    shapes = list(list(  # vertical line to highlight 2020
      type = "line", x0 = 2020, x1 = 2020,
      y0 = 0, y1 = 1, yref = "paper",
      line = list(color = "red", dash = "dot")
    ))
  )
wwtps_flagged_timeseries_plotly
```


# Save to mazu

```{r save}
out_dir <- here::here("goal_prep/cw/v2025/pharmaceuticals")
int_dir <- file.path(out_dir, "int")
output_dir <- file.path(out_dir, "output")

dir.create(int_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

write_csv(
  risk_factor_wwtp_effluent,
  file.path(output_dir, "risk_factor_wwtp_effluent_pharma_risk.csv")
)

write_csv(
  effluent_liters_wwtp_yearly,
  file.path(int_dir, "wwtp_effluent_by_plant_yearly.csv")
)

write_csv(
  wwtp_liters_effluent_by_watershed,
  file.path(int_dir, "wwtp_liters_effluent_by_watershed.csv")
)

write_csv(
  pourpoints_wwtp_liters_rgnid,
  file.path(int_dir, "effluent_wwtp_liters_rgnid.csv")
)

write_csv(
  proceso_retention_lookup,
  file.path(int_dir, "proceso_retention_lookup.csv")
)

if (nrow(proceso_not_in_lookup) > 0) {
  write_csv(
    proceso_not_in_lookup,
    file.path(int_dir, "proceso_not_in_retention_lookup.csv")
  )
}

st_write(
  pourpoints_wwtp_liters_final,
  file.path(int_dir, "pourpoints_effluent_wwtp_liters_final.gpkg"),
  delete_dsn = TRUE
)
```


