---
title: 'OHI GoC 2025: Clean waters pharmaceuticals'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# CW: Pharmaceuticals - Livestock

Estimated pharmaceutical discharge.

```{r setup}
library(sf)
library(utils)
library(tidyverse)
library(here)
library(raster)
library(mapview)
library(purrr)
library(stringr) # fuzzy string matching
library(stringi)
library(janitor)
library(tibble)
```

### Watersheds and pour points

```{r}
## lets use the full state for now and then summarize by the full watersheds.
goc_ohi_rgns <- st_read(here::here("spatial/ohi_regions","goc_ohi_rgns.shp"))

union_rgns <- goc_ohi_rgns %>% 
  st_union() %>% 
  st_make_valid()

# plot(goc_ohi_rgns)
# plot(union_rgns)

watershed_all <- st_read(file.path("/home/shares/ohi/stressors_2021/_dataprep/nutrients/watersheds_pourpoints/watersheds_all_4326.shp"))
st_crs(watershed_all)
# mapview(watershed_all)

watershed_goc <- st_intersection(st_transform(watershed_all, st_crs(union_rgns)), union_rgns)
mapview(watershed_goc)

# --- if we plan to change the buffer for some reason ---
# inland_50 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer","inland_50km_goc.shp"))
# watershed_goc_50km <- st_intersection(st_transform(watershed_all, st_crs(inland_50)), inland_50)
# mapview(watershed_goc_50km)
```


```{r}
# here's the associated pourpoints that should match with the watersheds:
pourpoints <- st_read("/home/shares/ohi/stressors_2021/_dataprep/nutrients/watersheds_pourpoints/pourpoints_mol.shp")

pourpoints_goc <- st_intersection(st_transform(pourpoints, st_crs(union_rgns)), union_rgns)
mapview(pourpoints_goc)

mapview(pourpoints_goc) + mapview(watershed_goc)
```


```{r ohi_marine_and_inland}
inland_50 <- st_read(here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer", "inland_50km_goc.shp"))
inland_10 <- st_read(here("/home/shares/ohi/OHI_GOC/spatial/inland_10km_buffer", "inland_10km_goc.shp"))
regions_marine <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp")) %>%
  filter(location == "marine")

marine_and_inland50 <- bind_rows(regions_marine, inland_50) %>%
  dplyr::select(rgn_id, geometry) %>%
  group_by(rgn_id) %>%
  summarize(geometry = st_union(geometry), .groups = "drop")

marine_and_inland10 <- bind_rows(regions_marine, inland_10) %>%
  dplyr::select(rgn_id, geometry) %>%
  group_by(rgn_id) %>%
  summarize(geometry = st_union(geometry), .groups = "drop")
```


## Estimating risk of pharmaceutical discharge based on the intensity of livestock farming in each OHI Region

Here I will plan to aggregate the total number of "cabezas", aka "heads", of all livestock and then rasterize the polygon of each municipio. I will then assume that the production is homogeneously distributed across the municipio, so that every cell has the same value within the municipality. 

I will then summarize the total cabezas in each watershed, and use the associated pourpoints to assign values to OHI regions. I will then aggregate the total livestock produced in each OHI Region, where the more livestock produced means that there is a higher risk of pharmaceutical discharge in that region and year.

### Reading in all livestock reared from 2006-2024 (missing 2023 right now)

Data was downloaded from <https://nube.agricultura.gob.mx/datosAbiertos/Pecuario.php>: 

For each year, click on the associated link for the year, then bring it into Mazu at `~/home/shares/ohi/OHI_GOC/_raw_data/DGSIAP/d2026`.

```{r}
## write function to read in all of the files and limit them to GoC states

read_in_livestock <- function (year) {
  
  df <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/DGSIAP/d2026", paste0("cierre_", year, ".csv"))) %>% 
    
    janitor::clean_names() %>% 
    
    mutate(across(where(is.character), ~iconv(., from = "WINDOWS-1252", to = "UTF-8"))) %>%  # to address encoding issues
    
    filter(nomestado %in% c("Baja California Sur","Baja California","Jalisco","Sinaloa","Sonora","Nayarit")) 
  
  return(df)
  
}
```

Now, we can use the function to read in each year's WWTP data and limit to GoC states.

```{r}
# specify the years we want to iterate through. ideally, we would have 2023 in the future...
years_to_read_in <- c(2006:2022, 2024)

# use `lapply` and bind rows to obtain data over time.
list_of_all_dfs <- lapply(years_to_read_in, read_in_livestock)
  
# bind them across all years
bind_all_years_livestock <- bind_rows(list_of_all_dfs) %>% 
  mutate(cvempio = as.numeric(cvempio),
         cveestado = as.numeric(cveestado))

summary(bind_all_years_livestock)
unique(bind_all_years_livestock$nomestado)
unique(bind_all_years_livestock$cvempio)
colnames(bind_all_years_livestock)

unique(bind_all_years_livestock$nomproducto)
# [1] "Miel"          "Carne"         "Huevo-plato"   "Ganado en pie" "Leche"        
# [6] "Cera"          "Huevo plato"   "Ganado en Pie"

unique(bind_all_years_livestock$cveestado)

```

### Aggregate all livestock cabezas for each municipality and year

Later, come back and determine how to weigh the different types of livestock appropriately.

```{r}
## check that carne is always less than "ganado en pie" (live cattle), so that we know that carne is a subset of ganado en pie?

# for now, just use ganado en pie only
sum_livestock_municipal <- bind_all_years_livestock %>% 
  filter(nomproducto %in% c("Ganado en pie","Ganado en Pie")) %>% 
  group_by(anio, cveestado, nomestado, cvempio, nommunicipio) %>% 
  summarize(toneladas_volumen = sum(volumen, na.rm = TRUE)) %>% 
  ungroup()
```


#### 1. Join with municipality polygons

```{r}
# read them in
municipio_codigo_sf <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/land_admin_boundaries", "mex_admin2_joined_w_municipio_codigos.shp")) %>% 
  mutate(cvempio = as.numeric(clv_mnc),
         cveestado = as.numeric(clv_std))

unique(municipio_codigo_sf$clv_mnc)
unique(municipio_codigo_sf$cveestado)

mapview(municipio_codigo_sf)
```

```{r}
siap_municipal <- inner_join(sum_livestock_municipal, municipio_codigo_sf) %>% 
  st_as_sf()

siap_municipal_2022 <- siap_municipal %>% 
  filter(anio %in% 2022)

mapview(siap_municipal_2022, zcol = "toneladas_volumen")
```

```{r}
gridded_cattle_density_2020_10km <- terra::rast(here::here("/home/shares/ohi/OHI_GOC/_raw_data/FAO_livestock_density/d2026/GLW4-2020.D-DA.CTL.tif"))

plot(gridded_cattle_density_2020_10km)

st_crs(gridded_cattle_density_2020_10km) # WGS 84

freq(gridded_cattle_density_2020_10km)

gridded_cattle_density_2020_10km_goc <- terra::crop(terra::project(gridded_cattle_density_2020_10km, crs(marine_and_inland50)), marine_and_inland50)

crs(gridded_cattle_density_2020_10km_goc)

plot(gridded_cattle_density_2020_10km_goc)
```

#### 2. Cattle density (global 10km raster)

```{r}
library(terra)

# read in cattle density
gridded_cattle_density_2020_10km <- terra::rast(here::here("/home/shares/ohi/OHI_GOC/_raw_data/FAO_livestock_density/d2026/GLW4-2020.D-DA.CTL.tif"))

# temp project the 9 polygons to WGS 84 just for the fast global crop
marine_and_inland50_wgs84 <- st_transform(marine_and_inland50, terra::crs(gridded_cattle_density_2020_10km))

# crop the massive global raster down to marine_and_inland50_wgs84 rectangular box
cattle_cropped_wgs84 <- terra::crop(gridded_cattle_density_2020_10km, marine_and_inland50_wgs84)
terra::plot(cattle_cropped_wgs84)

# project only this small cropped raster into our preferred metric CRS (Mexico LCC)
cattle_projected_lcc <- terra::project(cattle_cropped_wgs84, terra::crs(marine_and_inland50))

# mask the raster to the exact boundaries of the 9 region polygons
gridded_cattle_density_2020_10km_goc <- terra::mask(cattle_projected_lcc, marine_and_inland50)

# verify the final CRS
terra::crs(gridded_cattle_density_2020_10km_goc)
terra::plot(gridded_cattle_density_2020_10km_goc)
```

#### 3. Resample to 1km

```{r}
# now resample to 1km
# make an empty template with 1000m (1km) resolution matching the extent and CRS
template_1km <- terra::rast(gridded_cattle_density_2020_10km_goc)
terra::res(template_1km) <- 1000

# resample the data using bilinear interpolation to smoothly estimate new 1km density values. original data is in cattle heads/km^2
cattle_density_1km <- terra::resample(
  x = gridded_cattle_density_2020_10km_goc, 
  y = template_1km, 
  method = "bilinear"
)

# reapply the polygon mask to clean up any edge cells bleeding past the borders
gridded_cattle_density_2020_1km_goc <- terra::mask(cattle_density_1km, marine_and_inland50)

# check and plot
print(gridded_cattle_density_2020_1km_goc)
terra::plot(gridded_cattle_density_2020_1km_goc)
```

### Dasymetric mapping of municipal numbers to the spatially where estimated cattle locations are from the global raster

Because SIAP data is locked to arbitrary administrative boundaries (municipalities), we can't intersect it with watersheds, as a single municipality might span multiple watersheds. The GLW4 raster provides the "spatial proxy" to figure out where within those municipalities the livestock actually are.


#### 1. prep the DGSIAP data and rasterize

Ensure municipality polygons (ex: siap_municipal_2022) are in the exact same metric projection (Mexico LCC) as the 1km raster.

```{r}
library(terra)
library(sf)
library(dplyr)

# make the 2022 SIAP sf object a terra SpatVector in Mexico LCC
mun_vector <- siap_municipal_2022 %>% 
  st_transform(st_crs(gridded_cattle_density_2020_1km_goc)) %>% 
  vect()
```


#### 2. calculate the municipal weights

We need to find the total GLW raster value for each municipality and divide each cell by that total.

```{r}
# sum up the GLW 1km density values inside each municipality
mun_totals <- terra::extract(gridded_cattle_density_2020_1km_goc, mun_vector, fun = sum, na.rm = TRUE)

# append this municipal total back to the vector object (assuming the extraction order matches the vector rows)
mun_vector$glw_total_density <- mun_totals[, 2]

# rasterize the municipal totals so we can do raster algebra. we use the 1km cattle raster as a geometry template
mun_totals_raster <- terra::rasterize(mun_vector, gridded_cattle_density_2020_1km_goc, field = "glw_total_density")

# create the weight raster (Cell Weight = Cell Density / Municipal Total Density). cells will range from 0 to 1, representing the proportion of the municipality's cattle in that cell.
cattle_weight_raster <- gridded_cattle_density_2020_1km_goc / mun_totals_raster

plot(cattle_weight_raster)
```


#### 3. distribute SIAP values (cattle volume) across the 1km grid

```{r}
## rasterize the actual SIAP livestock volume (tons) for 2022
siap_volume_raster <- terra::rasterize(mun_vector, gridded_cattle_density_2020_1km_goc, field = "toneladas_volumen")

# distribute the tons across the 1km grid
# result: a 1km raster where cell values = estimated SIAP tons of cattle in that specific 1km block
siap_cattle_tons_1km <- siap_volume_raster * cattle_weight_raster

# quick check: plot it to see the refined 1km distribution of Mexican govt data
plot(siap_cattle_tons_1km)

```



### Summarize within watersheds

```{r}
# load in watershed sf object, project to match, and convert to SpatVector
watersheds_lcc <- st_read(here::here("/path/to/your/watersheds.shp")) %>% 
  st_transform(st_crs(siap_cattle_tons_1km)) %>% 
  vect()

# sum the SIAP cattle tons within each individual watershed
watershed_cattle_totals <- terra::extract(siap_cattle_tons_1km, watersheds_lcc, fun = sum, na.rm = TRUE)

# bind the calculated tons back to the watershed polygons!!!!
watersheds_lcc$total_cattle_tons <- watershed_cattle_totals[, 2]

```


### Assign to pourpoints and then OHI regions

```{r}

```




