---
title: "Saltmarsh Habitat Condition and Trend - Gulf of California OHI"
author: "Austin Martinez"
date: 04/06/2026
format: 
  html:
    code-fold: show
execute:
  echo: true
  message: false
  warning: false
---

```{r}
# Load in Libraries
library(sf)
library(dplyr)
library(here)
library(terra)
library(tidyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(tmap)
library(stringr)
```

```{r}
source(here::here("BD", "HAB", "temp_processing", "setup_temp.R"))
```

```{r}
# Read in OHI region boundaries
ohi_regions <- st_read(here("spatial", "ohi_regions", "goc_ohi_rgns.shp")) |> st_make_valid()
```
## Testing to see how changes in CRS change area values
```{r}
# Read in hem saltmarsh area shape files
Saltmarsh_data_hem_raw <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/Hem_data/goc_habitat_data/wetlands/wetlands_merge.shp"), options = "ENCODING=LATIN1") |>
  st_make_valid()

Saltmarsh_data_hem <- st_transform(Saltmarsh_data_hem_raw, st_crs(ohi_regions))

paste("before transform:", sum(st_area(Saltmarsh_data_hem_raw)))
paste("after tranform:", sum(st_area(Saltmarsh_data_hem)))
```


```{r}
# check
canabio_landcover_2020_raw <- rast(here("/home/shares/ohi/OHI_GOC/_raw_data/mexico_land_cover_conabio/land_cover_MEX_2020_30m/nalcmsmx20cw.tif"))

canabio_landcover_2020_14_raw <- ifel(canabio_landcover_2020_raw == 14, 1, NA)

canabio_landcover_2020 <- project(canabio_landcover_2020_14_raw, crs(ohi_regions), method = "near")

CONABIO_before <- global(cellSize(canabio_landcover_2020_14_raw, unit = "m"), "sum", na.rm = TRUE)
CONABIO_after <- global(cellSize(canabio_landcover_2020, unit = "m"), "sum", na.rm = TRUE)

paste("before transform:", CONABIO_before)
paste("after transform:", CONABIO_after)
```

```{r}
# Read in hem CONABIO Land use and vegetation area shape file
saltmarsh_canabio_landuse_raw <-st_read("/home/shares/ohi/OHI_GOC/_raw_data/CANABIO_Saltmarsh_landuse_veg/usv250s7gw/usv250s7gw.shp") |>
  st_make_valid()

saltmarsh_canabio_landuse <- st_transform(saltmarsh_canabio_landuse_raw, st_crs(ohi_regions))

paste("before transform:", sum(st_area(saltmarsh_canabio_landuse_raw)))
paste("after tranform:", sum(st_area(saltmarsh_canabio_landuse)))
```


```{r}
# Read in Oceanplus data
saltmarsh_oceanplus2_raw <-st_read("/home/shares/ohi/OHI_GOC/_raw_data/oceanplus_WCMC027_Saltmarsh_v6_1/01_Data/WCMC027_Saltmarshes_Py_v6_1.shp") |> 
  st_make_valid()

saltmarsh_oceanplus2 <- st_transform(saltmarsh_oceanplus2_raw, st_crs(ohi_regions))

paste("before transform:", sum(st_area(saltmarsh_oceanplus2_raw)))
paste("after tranform:", sum(st_area(saltmarsh_oceanplus2)))
```

```{r}
Blue_Carbon_saltmarsh_raw <-st_read("/home/shares/ohi/OHI_GOC/_raw_data/bluecarbon_saltmarsh_shapefile/NA_BlueCarbon_Saltmarsh_2021/data/North_America_Saltmarshes_Py.shp") |>
  st_make_valid()

Blue_Carbon_saltmarsh <- st_transform(Blue_Carbon_saltmarsh_raw, st_crs(ohi_regions))

paste("before transform:", sum(st_area(Blue_Carbon_saltmarsh_raw)))
paste("after tranform:", sum(st_area(Blue_Carbon_saltmarsh)))
```

```{r}
# Path to folder containing raster tiles
folder <- "/home/shares/ohi/OHI_GOC/_raw_data/CANABIO_Saltmarsh_landuse_veg/2020"

# Get bounding box of OHI regions in geographic coordinates (lat/lon)
bbox <- st_bbox(st_transform(ohi_regions, 4326))

# Create sequence of longitudes and latitudes at 5-degree intervals (matches naming convention of raster tiles)
lons <- seq(floor(bbox$xmin / 5) * 5, floor(bbox$xmax / 5) * 5, by = 5)
lats <- seq(floor(bbox$ymin / 5) * 5, floor(bbox$ymax / 5) * 5, by = 5)

# Create all combinations of lon/lat tiles and format names
tiles_needed <- expand.grid(lon = lons, lat = lats) |>
  mutate(
    lon_str = ifelse(lon >= 0, paste0("E", abs(lon)), paste0("W", abs(lon))),
    lat_str = ifelse(lat >= 0, paste0("N", lat), paste0("S", abs(lat))),
    tile_name = paste0(lon_str, lat_str)
  )

# Extract tile names needed for filtering files
needed <- tiles_needed$tile_name

# List only raster files that match required tile names
files <- list.files(
  "/home/shares/ohi/OHI_GOC/_raw_data/CANABIO_Saltmarsh_landuse_veg/2020",
  pattern = paste(needed, collapse = "|"),
  full.names = TRUE
)

# Creates a virtual raster (VRT)
tifs <- vrt(files)

# Mask to salt marsh only (value 186)
saltmarsh_raw <- ifel(tifs == 186, 1, NA)

# Compute area BEFORE transform (m²)
area_before <- global(cellSize(saltmarsh_raw, unit = "m"), "sum", na.rm = TRUE)

# Project masked raster into OHI CRS
saltmarsh_proj <- project(saltmarsh_raw, crs(ohi_regions), method = "near")

# Compute area AFTER transform (m²) ---
area_after <- global(cellSize(saltmarsh_proj, unit = "m"), "sum", na.rm = TRUE)


paste("before transform:", area_before)
paste("after transform:",  area_after)
```



## Redo the group bys

# This is data from Hem

```{r}
# Read in hem saltmarsh area shape files
Saltmarsh_data_hem_raw <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/Hem_data/goc_habitat_data/wetlands/wetlands_merge.shp"), options = "ENCODING=LATIN1") |>
  st_make_valid()

# Transform crs of ohi_regions to Saltmarsh_data_hem_raw crs
ohi_regions <- st_transform(ohi_regions, st_crs(Saltmarsh_data_hem_raw))

# Only keep Saltmarsh_data that intersects ohi_regions
saltmarsh_ohi_hem <- st_intersection(Saltmarsh_data_hem_raw, ohi_regions)

# Intersect, dissolve within regions, calculate area
saltmarsh_area_hem_2 <- st_intersection(Saltmarsh_data_hem_raw, ohi_regions) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6) |>
  st_drop_geometry()

saltmarsh_area_hem_2

# make sure no double counting is happening
Saltmarsh_data_hem_raw <- st_union(Saltmarsh_data_hem_raw)
# Extract area of Hem's saltmarsh data by region lets do this now
saltmarsh_area_hem_2 <- saltmarsh_ohi_hem |> 
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6) |> 
  group_by(rgn_id) |> 
  summarise(area_km2 = sum(area_km2)) |>
  st_drop_geometry()
```


# This is data from CONABIO: http://www.conabio.gob.mx/informacion/gis/?vns=gis_root/usv/otras/nalcmsmx20gw for 2020

```{r}
# check
canabio_landcover_2020_raw <- rast(here("/home/shares/ohi/OHI_GOC/_raw_data/mexico_land_cover_conabio/land_cover_MEX_2020_30m/nalcmsmx20cw.tif"))

ohi_regions_vect <- vect(ohi_regions)


# Project regions to match raster CRS
ohi_regions_vect <- project(ohi_regions_vect, crs(canabio_landcover_2020_raw))
```

```{r}
# Crop and mask to OHI regions
canabio_landcover_2020_crop <- crop(canabio_landcover_2020_raw, ohi_regions_vect)
canabio_landcover_2020_mask <- mask(canabio_landcover_2020_crop, ohi_regions_vect)

# Makes wetland (14) = 1, everything else = NA
canabio_landcover_2020_14 <- ifel(canabio_landcover_2020_mask == 14, 1, NA)

# Pixel area in km²
pixel_area_2020_canabio <- prod(res(canabio_landcover_2020_14)) / 1e6
```

```{r}
# Summarize by OHI region
wetland_2020_zonal <- zonal(canabio_landcover_2020_14, ohi_regions_vect, fun = "sum", na.rm = TRUE)
total_2020_zonal <- zonal(!is.na(canabio_landcover_2020_mask), ohi_regions_vect, fun = "sum", na.rm = TRUE)



wetland_area_2020_canabio_2 <- wetland_2020_zonal |>
  as_tibble() |>
  mutate(rgn_id = ohi_regions_vect$rgn_id,
         area_km2 = OID * pixel_area_2020_canabio) |>
  select(rgn_id, area_km2)

wetland_area_2020_canabio_2 <- wetland_area_2020_canabio_2 |>
  group_by(rgn_id) |>
  summarise(area_km2 = sum(area_km2, na.rm = TRUE))

wetland_density_2020_canabio <- wetland_area_2020_canabio_2 |>
  mutate(total_area_km2 = total_2020_zonal$OID * pixel_area_2020_canabio,
         density = area_km2 / total_area_km2) |>
  select(rgn_id, density)
wetland_density_2020_canabio
```

# This is the CONABIO Land use and vegetation data : http://www.conabio.gob.mx/informacion/gis/?vns=gis_root/usv/inegi/usv250s7gw
```{r}
sf::sf_use_s2(FALSE)

# read in CONABIO Land use and vegetation data
saltmarsh_canabio_landuse <-st_read("/home/shares/ohi/OHI_GOC/_raw_data/CANABIO_Saltmarsh_landuse_veg/usv250s7gw/usv250s7gw.shp") |>
  st_make_valid()

# Filter for Saltmarsh variables
saltmarsh_canabio <- saltmarsh_canabio_landuse |>
  dplyr::filter(
    DESCRIPCIO %in% c(
      "VEGETACIÓN HALÓFILA HIDRÓFILA",
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA HIDRÓFILA"
    )
  )

# Transform crs of ohi_regions to saltmarsh_canabio crs
ohi_regions <- st_transform(ohi_regions, st_crs(saltmarsh_canabio))

# Intersect, dissolve within regions, calculate area
saltmarsh_canabio_2021_area2 <- st_intersection(saltmarsh_canabio, ohi_regions) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6, year = 2021) |>
  st_drop_geometry()
saltmarsh_canabio_2021_area2
```

## This data comes from Oceanplus: https://habitats.oceanplus.org/mexico from March 2021
```{r}
# Read in Oceanplus data
saltmarsh_oceanplus2_raw <-st_read("/home/shares/ohi/OHI_GOC/_raw_data/oceanplus_WCMC027_Saltmarsh_v6_1/01_Data/WCMC027_Saltmarshes_Py_v6_1.shp") |> 
  st_make_valid()

# Transform crs of ohi_regions to saltmarsh_canabio crs
ohi_regions <- st_transform(ohi_regions, st_crs(saltmarsh_oceanplus2_raw))

# Intersect, dissolve within regions, calculate area
saltmarsh_oceanplus2_area_2 <- st_intersection(saltmarsh_oceanplus2_raw, ohi_regions) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6) |>
  st_drop_geometry()
saltmarsh_oceanplus2_area_2
```

## This data comes from Zenodo: https://zenodo.org/records/10068479 from March 2021


```{r}
# Path to folder containing raster tiles
folder <- "/home/shares/ohi/OHI_GOC/_raw_data/zenodo_saltmarsh/2020"

# Get bounding box of OHI regions in geographic coordinates (lat/lon)
bbox <- st_bbox(st_transform(ohi_regions, 4326))

# Create sequence of longitudes and latitudes at 5-degree intervals (matches naming convention of raster tiles)
lons <- seq(floor(bbox$xmin / 5) * 5, floor(bbox$xmax / 5) * 5, by = 5)
lats <- seq(floor(bbox$ymin / 5) * 5, floor(bbox$ymax / 5) * 5, by = 5)

# Create all combinations of lon/lat tiles and format names
tiles_needed <- expand.grid(lon = lons, lat = lats) |>
  mutate(
    lon_str = ifelse(lon >= 0, paste0("E", abs(lon)), paste0("W", abs(lon))),
    lat_str = ifelse(lat >= 0, paste0("N", lat), paste0("S", abs(lat))),
    tile_name = paste0(lon_str, lat_str)
  )

# Extract tile names needed for filtering files
needed <- tiles_needed$tile_name

# List only raster files that match required tile names
files <- list.files(
  "/home/shares/ohi/OHI_GOC/_raw_data/CANABIO_Saltmarsh_landuse_veg/2020",
  pattern = paste(needed, collapse = "|"),
  full.names = TRUE
)
```

```{r}
# Load and mosaic the tiles
tifs <- vrt(files)  # virtual mosaic, no need to load into memory

# Mask to salt marsh only (value 186)
saltmarsh_raster <- ifel(tifs == 186, 1, NA)


# Reproject ohi_regions to match raster CRs
ohi_vect <- vect(st_transform(ohi_regions, crs(tifs)))

# Compute per-cell area in km²
cell_areas <- cellSize(saltmarsh_raster, unit = "km")
cell_areas_masked <- mask(cell_areas, saltmarsh_raster)

# Zonal sum by region (this already groups correctly)
# IMPORTANT: explicitly use zone field


saltmarsh_area2 <- zonal(
  cell_areas_masked,
  ohi_vect,
  fun = "sum",
  na.rm = TRUE
)

saltmarsh_area2$rgn_id <- ohi_vect$rgn_id

saltmarsh_area2_saltmarsh <- saltmarsh_area2 |>
  group_by(rgn_id) |>
  summarise(area_km2 = sum(area, na.rm = TRUE)) |>
  mutate(area_km2 = ifelse(area_km2 == 0, NA, area_km2))

saltmarsh_area2_saltmarsh
```



## This data comes from North American Blue Carbon: https://www.cec.org/north-american-environmental-atlas/north-american-blue-carbon-2021/ extent for 2021

```{r}
# Read in Blue Carbon data
Blue_Carbon_saltmarsh_raw <-st_read("/home/shares/ohi/OHI_GOC/_raw_data/bluecarbon_saltmarsh_shapefile/NA_BlueCarbon_Saltmarsh_2021/data/North_America_Saltmarshes_Py.shp")

# Make saltmarsh_oceanplus2 layer have the same CRS as OHI regions
ohi_regions <- st_transform(ohi_regions, st_crs(Blue_Carbon_saltmarsh_raw))

# Intersect, dissolve within regions, calculate area
blue_carbon_saltmarsh_area <- st_intersection(Blue_Carbon_saltmarsh_raw, ohi_regions) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6) |>
  st_drop_geometry()
blue_carbon_saltmarsh_area
```


# Create ktable
```{r}
full_join(saltmarsh_canabio_2021_area2, saltmarsh_area_hem_2, by = "rgn_id", suffix = c("_conabio", "_hem")) |>
  full_join(saltmarsh_oceanplus2_area_2, by = "rgn_id") |>
  rename(area_km2_oceanplus = area_km2) |>
  full_join(wetland_area_2020_canabio_2, by = "rgn_id") |>
  rename(area_km2_wetland_2020 = area_km2) |>
  full_join(saltmarsh_area2_saltmarsh, by = "rgn_id") |>
  rename(area_km2_zenodo = area_km2) |>
  full_join(blue_carbon_saltmarsh_area, by = "rgn_id") |>
  rename(area_km2_bluecarbon = area_km2) |>
  select(
    rgn_id,
    area_km2_hem,
    area_km2_conabio,
    area_km2_oceanplus,
    area_km2_wetland_2020,
    area_km2_zenodo,
    area_km2_bluecarbon
  ) |>
  arrange(rgn_id) |>
  kable(
    digits = 4,
    col.names = c(
      "Region",
      "Hem (km²)",
      "CONABIO LU shp 2021 (km²)",
      "Oceanplus (km²)",
      "CONABIO LC tif 2020 (km²)",
      "Zenodo 2020 (km²)",
      "Blue Carbon (km²)"
    )
  ) |>
  kable_styling(bootstrap_options = c("striped", "bordered"))

```

```{r}
sf::sf_use_s2(FALSE)

# Filter for Saltmarsh variables
saltmarsh_canabio <- saltmarsh_canabio_landuse |>
  dplyr::filter(
    DESCRIPCIO %in% c(
      "VEGETACIÓN HALÓFILA HIDRÓFILA",
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA HIDRÓFILA"
    )
  )

# Transform crs of ohi_regions to saltmarsh_canabio crs
ohi_regions <- st_transform(ohi_regions, st_crs(saltmarsh_canabio))

# Intersect, dissolve within regions, calculate area
saltmarsh_canabio_2021_area2 <- st_intersection(saltmarsh_canabio, ohi_regions) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6, year = 2021) |>
  st_drop_geometry()
```





# CANABIO Land Use and Vegitation all yeats

```{r}
saltmarsh_canabio_landuse_1980_91 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CANABIO_Saltmarsh_landuse_veg/usv250kcs1agw/usv250kcs1agw.shp") |> st_make_valid()

saltmarsh_canabio_landuse_2001 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CANABIO_Saltmarsh_landuse_veg/usv250ks2gw/usv250ks2gw.shp") |> st_make_valid()

saltmarsh_canabio_landuse_2005 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CANABIO_Saltmarsh_landuse_veg/usv250ks3gw/usv250ks3gw.shp") |> st_make_valid()

saltmarsh_canabio_landuse_2009 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CANABIO_Saltmarsh_landuse_veg/usv250ks4gw/usv250ks4gw.shp") |> st_make_valid()

saltmarsh_canabio_landuse_2013 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CANABIO_Saltmarsh_landuse_veg/usv250s5ugw/usv250s5ugw.shp") |> st_make_valid()

saltmarsh_canabio_landuse_2016 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CANABIO_Saltmarsh_landuse_veg/usv250s6gw/usv250s6gw.shp") |> st_make_valid()

saltmarsh_canabio_landuse_2021 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CANABIO_Saltmarsh_landuse_veg/usv250s7gw/usv250s7gw.shp") |> st_make_valid()
```

VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA 
VEGETACIÓN HALÓFILA

```{r}
# 1980-91
saltmarsh_canabio_1980_91 <- saltmarsh_canabio_landuse_1980_91 |>
  dplyr::filter(
    descripcio %in% c(
      "VEGETACIÓN HALÓFILA",
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA"
    )
  )

ohi_regions_1980_91 <- st_transform(ohi_regions, st_crs(saltmarsh_canabio_1980_91))

saltmarsh_canabio_1980_91_area <- st_intersection(saltmarsh_canabio_1980_91, ohi_regions_1980_91) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 1980
  ) |>
  st_drop_geometry()

saltmarsh_canabio_1980_91_area
```


```{r}
# 2001

saltmarsh_canabio_2001 <- saltmarsh_canabio_landuse_2001 |>
  dplyr::filter(
    descripcio %in% c(
      "VEGETACIÓN HALÓFILA HIDRÓFILA",
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA HIDRÓFILA"
    )
  )

ohi_regions_2001 <- st_transform(ohi_regions, st_crs(saltmarsh_canabio_2001))

saltmarsh_canabio_2001_area <- st_intersection(saltmarsh_canabio_2001, ohi_regions_2001) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2001
  ) |>
  st_drop_geometry()

saltmarsh_canabio_2001_area
```


```{r}
# 2005

saltmarsh_canabio_2005 <- saltmarsh_canabio_landuse_2005 |>
  dplyr::filter(
    descripcio %in% c(
      "VEGETACIÓN HALÓFILA HIDRÓFILA",
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA HIDRÓFILA"
    )
  )

ohi_regions_2005 <- st_transform(ohi_regions, st_crs(saltmarsh_canabio_2005))

saltmarsh_canabio_2005_area <- st_intersection(saltmarsh_canabio_2005, ohi_regions_2005) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2005
  ) |>
  st_drop_geometry()

saltmarsh_canabio_2005_area
```


```{r}
# 2009

saltmarsh_canabio_2009 <- saltmarsh_canabio_landuse_2009 |>
  dplyr::filter(
    descripcio %in% c(
      "VEGETACIÓN HALÓFILA HIDRÓFILA",
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA HIDRÓFILA"
    )
  )


ohi_regions_2009 <- st_transform(ohi_regions, st_crs(saltmarsh_canabio_2009))

saltmarsh_canabio_2009_area <- st_intersection(saltmarsh_canabio_2009, ohi_regions_2009) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2009
  ) |>
  st_drop_geometry() |>
    complete(rgn_id = unique(ohi_regions_2009$rgn_id), fill = list(area_km2 = 0, year = 2009))

saltmarsh_canabio_2009_area
```


```{r}
# 2013
saltmarsh_canabio_2013 <- saltmarsh_canabio_landuse_2013 |>
  dplyr::filter(
    descripcio %in% c(
      "VEGETACIÓN HALÓFILA HIDRÓFILA",
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA HIDRÓFILA"
    )
  )

ohi_regions_2013 <- st_transform(ohi_regions, st_crs(saltmarsh_canabio_2013))

saltmarsh_canabio_2013_area <- st_intersection(saltmarsh_canabio_2013, ohi_regions_2013) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2013
  ) |>
  st_drop_geometry()

saltmarsh_canabio_2013_area
```


```{r}
# 2016

saltmarsh_canabio_2016 <- saltmarsh_canabio_landuse_2016 |>
  dplyr::filter(
    DESCRIPCIO %in% c(
      "VEGETACIÓN HALÓFILA HIDRÓFILA",
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA HIDRÓFILA"
    )
  )

ohi_regions_2016 <- st_transform(ohi_regions, st_crs(saltmarsh_canabio_2016))

saltmarsh_canabio_2016_area <- st_intersection(saltmarsh_canabio_2016, ohi_regions_2016) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2016
  ) |>
  st_drop_geometry() |>
    complete(rgn_id = unique(ohi_regions_2016$rgn_id), fill = list(area_km2 = 0, year = 2016))

saltmarsh_canabio_2016_area
```


```{r}
# 2021

saltmarsh_canabio_2021 <- saltmarsh_canabio_landuse_2021 |>
  dplyr::filter(
    DESCRIPCIO %in% c(
      "VEGETACIÓN HALÓFILA HIDRÓFILA",
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA HIDRÓFILA"
    )
  )

ohi_regions_2021 <- st_transform(ohi_regions, st_crs(saltmarsh_canabio_2021))

saltmarsh_canabio_2021_area <- st_intersection(saltmarsh_canabio_2021, ohi_regions_2021) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2021
  ) |>
  st_drop_geometry()

saltmarsh_canabio_2021_area
```


```{r}
# Combine all years into one long table
saltmarsh_all_years <- dplyr::bind_rows(
  saltmarsh_canabio_1980_91_area,
  saltmarsh_canabio_2001_area,
  saltmarsh_canabio_2005_area,
  saltmarsh_canabio_2009_area,
  saltmarsh_canabio_2013_area,
  saltmarsh_canabio_2016_area,
  saltmarsh_canabio_2021_area
)

# Pivot wider: regions on rows, years on columns
saltmarsh_wide <- saltmarsh_all_years |>
  tidyr::pivot_wider(
    names_from = year,
    values_from = area_km2) |>
  dplyr::arrange(rgn_id)

# Make the kable
saltmarsh_wide |>
  kable( digits = 4,
    col.names = c(
      "Region ID", "1991", "2001", "2005", "2009", "2013", "2016", "2021")) |>
  kable_styling(
    bootstrap_options = c("striped", "bordered", "hover")) |>
  add_header_above(
    c("Saltmarsh Extent CONABIO Land Use (All Years)" = 8))

```

```{r}
print(paste("1980_91 unique values:", 
            length(unique(saltmarsh_canabio_landuse_1980_91$descripcio))))

print(paste("2001 unique values:", 
            length(unique(saltmarsh_canabio_landuse_2001$descripcio))))

print(paste("2005 unique values:", 
            length(unique(saltmarsh_canabio_landuse_2005$descripcio))))

print(paste("2009 unique values:", 
            length(unique(saltmarsh_canabio_landuse_2009$descripcio))))

print(paste("2013 unique values:", 
            length(unique(saltmarsh_canabio_landuse_2013$descripcio))))

print(paste("2016 unique values:", 
            length(unique(saltmarsh_canabio_landuse_2016$DESCRIPCIO))))

print(paste("2021 unique values:", 
            length(unique(saltmarsh_canabio_landuse_2021$DESCRIPCIO))))
```
```{r}
library(ggplot2)

ggplot(saltmarsh_all_years, aes(x = year, y = area_km2, group = rgn_id, color = factor(rgn_id))) +
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  labs(
    title = "Saltmarsh Extent Over Time (CONABIO Land Use)",
    x = "Year",
    y = "Saltmarsh Extent (km²)",
    color = "Region ID"
  ) +
  theme_minimal(base_size = 14)

```
```{r}
saltmarsh_total <- saltmarsh_all_years |>
  group_by(year) |>
  summarise(total_km2 = sum(area_km2, na.rm = TRUE))

ggplot(saltmarsh_total, aes(x = year, y = total_km2)) +
  geom_point(size = 3) +
  geom_line(linewidth = 1.2) +
  labs(
    title = "Total Saltmarsh Extent Over Time",
    x = "Year",
    y = "Total Extent (km²)"
  ) +
  theme_minimal(base_size = 14)

```


# Extent: 2021
# This is the CONABIO Land use and vegetation data : http://www.conabio.gob.mx/informacion/gis/?vns=gis_root/usv/inegi/usv250s7gw
```{r}
# 2021
# Filter for saltmarsh classes in 2021 CONABIO land use data
saltmarsh_canabio_2021 <- saltmarsh_canabio_landuse_2021 |>
  dplyr::filter(
    DESCRIPCIO %in% c(
      "VEGETACIÓN HALÓFILA HIDRÓFILA",
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA HIDRÓFILA"
    )
  )

# Transform OHI regions to match the CRS of the saltmarsh data
ohi_regions_2021 <- st_transform(ohi_regions, st_crs(saltmarsh_canabio_2021))

# Intersect saltmarsh polygons with OHI regions and calculate area
saltmarsh_canabio_2021_area <- st_intersection(saltmarsh_canabio_2021, ohi_regions_2021) |>
  group_by(rgn_id) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2021
  ) |>
  st_drop_geometry()

# Make a nice kable table
saltmarsh_canabio_2021_area |>
select(rgn_id, year, area_km2) |>
  kable(
    digits = 4,
    col.names = c('Region ID', 'Year', 'Extent (km²)'),
    align = c('c', 'c', 'r')
  ) |>
  kable_styling(
    bootstrap_options = c('striped', 'bordered', 'hover')
  ) |>
  add_header_above(
    c('Saltmarsh Extent - CONABIO Land Use and Vegetation (2021)' = 3)
  )
```




# Trend: 2021/2013

```{r}
# File 3: saltmarsh_trend.csv using CANABIO LU data
canabio_trend <- bind_rows(saltmarsh_canabio_2021_area, saltmarsh_canabio_2013_area) |>
  pivot_wider(names_from = year, values_from = area_km2) |>
  mutate(
    trend = (`2021` / `2013`)^(1/8) - 1,
    year = 2021
  ) |>
  select(rgn_id, year, trend)

canabio_trend |>
  kable(digits = 4, col.names = c('Region ID', 'Year', 'Trend')) |>
  kable_styling(bootstrap_options = c('striped', 'bordered', 'hover')) |>
  add_header_above(c('Saltmarsh Trend CONABIO 2021/2013' = 3))
```





# Condition

```{r}
# File 1: saltmarsh_condition using CANABIO LU data
saltmarsh_condition <- bind_rows(saltmarsh_canabio_2001_area, saltmarsh_canabio_2021_area) |>
  pivot_wider(names_from = year, values_from = area_km2) |>
  mutate(condition = `2021` / `2001`, year = 2021) |>
  select(rgn_id, year, condition)

saltmarsh_condition |>
  kable(digits = 4, col.names = c('Region ID', 'Year', 
                                   'Condition (2021/2001)')) |>
  kable_styling(bootstrap_options = c('striped', 'bordered', 'hover')) |>
  add_header_above(c('Saltmarsh Condition' = 3)) # make above 1 values 1, and 2 decimal places 
```

```{r}
# Write CSVs to output_data directory
write.csv(saltmarsh_canabio_2021_area,
          "output_data/saltmarsh_extent_2021.csv",
          row.names = FALSE)

write.csv(canabio_trend,
          "output_data/saltmarsh_trend.csv",
          row.names = FALSE)

write.csv(saltmarsh_condition,
          "output_data/saltmarsh_condition.csv",
          row.names = FALSE)
```

