---
title: "Simple CMSY_Oct 22 2025"
output: html_document
date: "2025-10-22"
---

# PRE-WORK
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Loading Libraries}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(tidytext)
library(lubridate)
library(plotly)
library(htmlwidgets)
library(here)
library(janitor)

setwd("/home/ferrer/ohi_goc_prep/FP/CMSY")
```

```{r Review the CMSY protocol developed by Ferrer et al. (2021)}
### View PDF in Notebook
pdf_path <- "~/ohi_goc_prep/FP/CMSY/Copy of Ferrer et al. (2021) CMSY Protocol.pdf"

# Get the absolute path
abs_path <- normalizePath("~/ohi_goc_prep/FP/CMSY/Copy of Ferrer et al. (2021) CMSY Protocol.pdf")

# Convert to file URL (Windows or Unix-style)
file_url <- paste0("file:///", URLencode(abs_path))

# Open the PDF
browseURL(file_url)
```


# GENERATING FILES 1 & 2 FROM SCRATCH
Preparing Files 1 and 2
NOTES TO SELF - Oct 22 2025

FILE 1: Create a file with annual landings in kg (and n_count observations) grouped by year per "stock". Here, stock is defined as the total population of taxa X fished within the Gulf of California not including not the Pacific.

- At present, we do not have the municipality lat-longs so I will group stocks by the six state of interest first (Baja, BCS, Son, Sin, Nay, Jal) and then plan to re-make the file (File 1) once we can filter out specific municipalities.
- This means that this timeseries is likely to be an OVER-ESTIMATE of landings.

File 2: Create a file with Region, [Subregion], Stock, Group = Name, English Name, Scientific Name, Source, and a whole bunch of year columns that you populate using the various R files shared by Alfredo. This file can probably be trimmed down but for now I recommend formatting it exactly as it appears from before.

# Create FILE 1 - Timeseries
```{r File 1 - Stock Timeseries (JC data)}
# CONAPESCA data shared by JC
# These data were shared with us by Juan Carlos Villaseñor and originally accessed on September 26, 2025 from this GitHub repository: https://github.com/mex-fisheries

MexLandings_2000.present<- readRDS(here("/home/shares/ohi/OHI_GOC/_raw_data/CONAPESCA/d2025/SharedByJuanCarlos/mex_landings/data/clean/mex_landings_2000_present.rds"))
# n = 12,153,879 records from all states in Mexico from January 2000 through January 2025


# Skimming through the dat. What does it look like?
# Yes aquaculture production
# Records up to January, 2025

head(MexLandings_2000.present)

table(MexLandings_2000.present$acuaculture_production)

max(MexLandings_2000.present$year)
max(MexLandings_2000.present$month[MexLandings_2000.present$year])
```

```{r Basic filtering}
States_of_Interest<- c("BAJA CALIFORNIA", "BAJA CALIFORNIA SUR", "SONORA", "SINALOA", "NAYARIT", "JALISCO")

# PENDING: HERE IS WHERE WE COULD FILTER BY MUNICIPALITY
# Landing_Sites_of_Interest<- c(...)

MexLandings_filtered<- MexLandings_2000.present %>% filter(state %in% States_of_Interest)
# n obs.= 4,713,144 (-7,440,735 obs)

table(MexLandings_filtered$fleet)
# large_scale small_scale 
#      181823     1438462 
table(MexLandings_filtered$acuaculture_production) 
#there are also mariculture landings in this dataset and NAs in this column...

FisheryLandings_filtered<- MexLandings_filtered %>% filter(acuaculture_production== "NO" | is.na(acuaculture_production)) # 4684000 obs.

MaricultureLandings_filtered<- MexLandings_filtered %>% filter(acuaculture_production== "SÍ")
# n obs.= 29,144 (-4,684,000 obs)

# NOTE: SOME RECORDS ARE NOT APPENDED WITH INFORMATION ABOUT RECORD TYPE: no information on wild-catch versus aquaculture. For this preliminary analyses, these records are assumed to be landings from fishery operations and are included.

table(FisheryLandings_filtered$year)

# OTRAS_filtered<- MexLandings_filtered %>% filter(main_species_group== "OTRAS")
# n obs.= 715.547 (-1,011,581 obs)
```

```{r Summary table}
# Turn off scientific notation:
options(scipen = 999)

# Create timeseries FILE 1
head(FisheryLandings_filtered)

StateLandings_AnnualSummary_byTaxa <- FisheryLandings_filtered %>%
  select(year, state, main_species_group, live_weight, landed_weight, value) %>%
  group_by(main_species_group, year, state) %>%
  summarise(
    Total_Live_Biomass = sum(live_weight, na.rm = TRUE),
    Total_Landed_Biomass = sum(landed_weight, na.rm = TRUE),
    Total_Value = sum(value, na.rm = TRUE),
    n_Records = n(),
    .groups = "drop"  # optional: ungroup after summarising
  )

TotalLandings_AnnualSummary_byTaxa <- FisheryLandings_filtered %>%
  select(year, main_species_group, live_weight, landed_weight, value) %>%
  group_by(main_species_group, year) %>%
  summarise(
    Total_Live_Biomass = sum(live_weight, na.rm = TRUE),
    Total_Landed_Biomass = sum(landed_weight, na.rm = TRUE),
    Total_Value = sum(value, na.rm = TRUE),
    n_Records = n(),
    .groups = "drop"  # optional: ungroup after summarising
  )

File1_df<- TotalLandings_AnnualSummary_byTaxa

# Re-name column
colnames(File1_df)[colnames(File1_df) == "main_species_group"] <- "Stock"
colnames(File1_df)[colnames(File1_df) == "year"] <- "Year"
```
## SAVE FILE 1
```{r SAVE FILE 1}
# write.csv(File1_df, "~/ohi_goc_prep/FP/CMSY/File1_UNDER_CONSTRUCTION_Oct 22 2025.csv", row.names = FALSE)
```

# Create FILE 2 - Metadata
```{r Create File 2 (Metadata File) based on File 1}
# STEP 1:
# Which taxa (stocks) do we want to account for in our File 2?
Stock_names<- unique(File1_df$Stock)
# don’t 57 unique taxa

## Read in the column names of File 2 from Ferrer et al. (2022) CSV file
headers <- names(read.csv("Ferrer et al 2021_File 2_Stocks Metadata.csv", nrows = 1))

## Create an empty dataframe with the same headers
File2_df <- data.frame(matrix(ncol = length(headers), nrow = 0))
colnames(File2_df) <- headers

## Add Stock_names to the 'Stock' column
File2_df <- File2_df[1:length(Stock_names), ]  # Expand rows
File2_df$Stock <- Stock_names

## Reset rownames / the index column numbered 1-57
rownames(File2_df) <- NULL


# STEP 2:
# Trim some of the columns off this file:
File2_df <- File2_df[ , !(names(File2_df) %in% c("ID", "Subregion", "Group", "Name","MinOfYear", "MaxOfYear")) ]

# Re-name columns
colnames(File2_df)[colnames(File2_df) == "Pick"] <- "Sufficient_No_Years"
colnames(File2_df)[colnames(File2_df) == "Keep"] <- "Yes_Run"
colnames(File2_df)[colnames(File2_df) == "stb.low"] <- "st_b.low"
colnames(File2_df)[colnames(File2_df) == "stb.hi"] <- "st_b.hi"
colnames(File2_df)[colnames(File2_df) == "int.yr"] <- "int_yr"
colnames(File2_df)[colnames(File2_df) == "intb.low"] <- "int_b.low"
colnames(File2_df)[colnames(File2_df) == "intb.hi"] <- "int_b.hi"
colnames(File2_df)[colnames(File2_df) == "endb.low"] <- "end_b.low"
colnames(File2_df)[colnames(File2_df) == "endb.high"] <- "end_b.hi"


# STEP 3:
# Fill in some of the columns off this file:
File2_df$Region <- c("GOC")
File2_df$Class <- c("TBD")
File2_df$Source <- c("JC_data")
File2_df$btype <- c("None")
File2_df$force.cmsy <- c("TRUE")

# STEP 4:
# For species with less than 10 years of data, do not keep (Sufficient_No_Years formerly "Pick" = 0)
StockYears_per_Taxa <- File1_df %>%
  group_by(Stock) %>%
  summarise(n_Stock_Years = n_distinct(Year), .groups = "drop")

StockYears_per_Taxa <- StockYears_per_Taxa %>%
  mutate(Sufficient_No_Years = ifelse(n_Stock_Years < 10, 0, 1))

## How many stocks will we be keeping based on number of years?
table(StockYears_per_Taxa$Sufficient_No_Years)
## We will keep 54 out of 57 stocks, exclude 3

##  Add this information to the File 2 dataframe: join StockYears_per_Taxa to File2_df by "Stock", bringing in the filled Sufficient_No_Years column
File2_df <- File2_df %>%
  select(-Sufficient_No_Years) %>%  # Remove empty column to avoid duplicates
  left_join(
    StockYears_per_Taxa %>% select(Stock, Sufficient_No_Years),
    by = "Stock"
  )

# STEP 5:
# Re-order columns:
#File2_df <- File2_df[c(1, 3, 2, 4:5, 6, 39, 37:38, 7:36)]
```

## SAVE FILE 2
```{r SAVE FILE 2}
# NEXT STEPS: Fill in the rest of column 2 using the ancillary files shared by Alfredo in Ferrer et al. (2022)
write.csv(File2_df, "~/ohi_goc_prep/FP/CMSY/File2_UNDER_CONSTRUCTION_Oct 22 2025.csv", row.names = FALSE)
```

Did some work offline on my local machine (not available for viewing here... yet). This work will help fill out File 2.

```{r Read in additional files to complete file 2}
#File2_df <- File2_df
#File2_df0 <- read.csv("File2_UNDER_CONSTRUCTION_Oct 22 2025.csv")

## Additional files needed to file in the metadata
Biomass_Priors <- read.csv("Priors_Starting Biomass_from FAO data_Oct 23 2025.csv")

Int_Year_Priors <- read.csv("Priors_Intermediate Years_Oct 23 2025.csv")

Common_Names <- read.csv("Stock and Common Names_Oct 23 2025.csv")
```

```{r lowercase_function}
lowercase_function<- function(x){
  if(is.character(x)){
    tolower(x)
  } else if (is.factor(x)){
    tolower(as.character(x))
  } else {
    x
  }
}
```

```{r Stitch these dataframes together}
File2_df_merged<- as.data.frame(lapply(File2_df, lowercase_function))
Biomass_Priors<- as.data.frame(lapply(Biomass_Priors, lowercase_function))
Int_Year_Priors<- as.data.frame(lapply(Int_Year_Priors, lowercase_function))
Common_Names<- as.data.frame(lapply(Common_Names, lowercase_function))

File2_df_merged <- File2_df_merged %>%
  left_join(
    Biomass_Priors %>% select(Stock, st_b.low, st_b.hi),
    by = "Stock",
    suffix = c(".file2", ".priors")
  ) %>%
  mutate(
    st_b.low = ifelse(is.na(st_b.low.file2) | st_b.low.file2 == "",
                      st_b.low.priors,
                      st_b.low.file2),
    st_b.hi = ifelse(is.na(st_b.hi.file2) | st_b.hi.file2 == "",
                     st_b.hi.priors,
                     st_b.hi.file2)
  ) %>%
  select(-st_b.low.file2, -st_b.low.priors, -st_b.hi.file2, -st_b.hi.priors)

File2_df_merged <- File2_df_merged %>%
  left_join(
    Int_Year_Priors %>% select(Stock, int_yr, StartYear, EndYear, Sufficient_No_Years),
    by = "Stock",
    suffix = c(".file2", ".priors")
  ) %>%
  mutate(
    int_yr = ifelse(is.na(int_yr.file2) | int_yr.file2 == "",
                      int_yr.priors,
                      int_yr.file2),
    StartYear = ifelse(is.na(StartYear.file2) | StartYear.file2 == "",
                     StartYear.priors,
                     StartYear.file2),
    EndYear = ifelse(is.na(EndYear.file2) | EndYear.file2 == "",
                      EndYear.priors,
                      EndYear.file2),
    Sufficient_No_Years = ifelse(is.na(Sufficient_No_Years.file2) | Sufficient_No_Years.file2 == "",
                     Sufficient_No_Years.priors,
                     Sufficient_No_Years.file2)
  ) %>%
  select(-int_yr.file2, -int_yr.priors, -StartYear.file2, -StartYear.priors, -EndYear.file2, -EndYear.priors, -Sufficient_No_Years.file2, -Sufficient_No_Years.priors)

File2_df_merged <- File2_df_merged %>%
  left_join(
    Common_Names %>% select(Stock, Class, EnglishName),
    by = "Stock",
    suffix = c(".file2", ".priors")
  ) %>%
  mutate(
    Class = ifelse(is.na(Class.file2) | Class.file2 == "",
                      Class.priors,
                      Class.file2),
    EnglishName = ifelse(is.na(EnglishName.file2) | EnglishName.file2 == "",
                     EnglishName.priors,
                     EnglishName.file2)
  ) %>%
  select(-Class.file2, -Class.priors, -EnglishName.file2, -EnglishName.priors)
```

```{r Abbreviated file under construction}
File2_df_abbreviated <- File2_df_merged %>%
  filter(!is.na(st_b.low))

File2_df_abbreviated <- File2_df_abbreviated %>%
  mutate(Yes_Run = ifelse(Sufficient_No_Years == 1, 1, 0))
```

## RE-SAVE FILE 2 - as abbreviated file under construction
```{r SAVE FILE 2 ABBREVIATED UNDER CONSTRUCTION OCT 23 2025}
write.csv(File2_df_abbreviated, "~/ohi_goc_prep/FP/CMSY/File2_Abbreviated_UNDER_CONSTRUCTION_Oct 23 2025.csv", row.names = FALSE)
```
# NEXT STEPS: Fill in the rest of the columns using the 3_Initial_Conditions file to estimate biomass file. Then add in preliminary "resilience" (that is, "r") values for each stock. Once these columns in file 2 are filled out, I can run an abbreviated CMSY with the 22 stocks for which we have complete priors. Once this is done I can get creative about filling in my priors for the other >20 stocks in our database.

Did some work offline on my local machine (not available for viewing here... yet). Filed out the intermediate and end biomass priors and resilience values. Add this to File 2 (Abbreviated) and its ready to run CMSY.

```{r Add final priors and resilience values to this dataframe}
#File2_df0 <- File2_df_abbreviated
File2_df0 <- read.csv("File2_Abbreviated_UNDER_CONSTRUCTION_Oct 23 2025.csv")
Intermediate_Biomass_Priors <- read.csv("Priors_Initial Conditions_Oct 23 2025.csv")
Resilience_Priors <- read.csv("Priors_Resilience Values_Oct 23 2025.csv")

# Apply lower-case function
File2_df_merged<- as.data.frame(lapply(File2_df0, lowercase_function))
Intermediate_Biomass_Priors<- as.data.frame(lapply(Intermediate_Biomass_Priors, lowercase_function))


File2_df_merged <- File2_df_merged %>%
  left_join(
    Intermediate_Biomass_Priors %>% select(Stock, int_b.low, int_b.hi, end_b.low, end_b.hi),
    by = "Stock",
    suffix = c(".file2", ".priors")
  ) %>%
  mutate(
    int_b.low = ifelse(is.na(int_b.low.file2) | int_b.low.file2 == "",
                      int_b.low.priors,
                      int_b.low.file2),
    int_b.hi = ifelse(is.na(int_b.hi.file2) | int_b.hi.file2 == "",
                     int_b.hi.priors,
                     int_b.hi.file2),
    end_b.low = ifelse(is.na(end_b.low.file2) | end_b.low.file2 == "",
                      end_b.low.priors,
                      end_b.low.file2),
    end_b.hi = ifelse(is.na(end_b.hi.file2) | end_b.hi.file2 == "",
                     end_b.hi.priors,
                     end_b.hi.file2)
  ) %>%
  select(-int_b.low.file2, -int_b.low.priors, 
         -int_b.hi.file2, -int_b.hi.priors, 
         -end_b.low.file2, -end_b.low.priors, 
         -end_b.hi.file2, -end_b.hi.priors)

File2_df_merged <- File2_df_merged %>%
  left_join(
    Resilience_Priors %>% select(Stock, Resilience),
    by = "Stock",
    suffix = c(".file2", ".priors")
  ) %>%
  mutate(
    Resilience = ifelse(is.na(Resilience.file2) | Resilience.file2 == "",
                      Resilience.priors,
                      Resilience.file2)
  ) %>%
  select(-Resilience.file2, -Resilience.priors)

File2_df_abbreviated <- File2_df_merged %>%
  mutate(Yes_Run = ifelse(is.na(Resilience), 0, 1))
```

## RE-RE-SAVE FILE 2 - as abbreviated file under construction
```{r SAVE FILE 2 ABBREVIATED UNDER CONSTRUCTION OCT 23 2025}
write.csv(File2_df_abbreviated, "~/ohi_goc_prep/FP/CMSY/File2_Abbreviated_UNDER_CONSTRUCTION_Oct 23 2025.csv", row.names = FALSE)
```
# NEXT STEPS: I can run an abbreviated CMSY with the 21 stocks for which we have complete priors (and marine). Once this is done I can get creative about filling in my priors for the other >20 additional stocks in our database.

# RUNNING CMSY WITH FILES 1 & 2 - Run on local machine

# TESTING CMSY on FERRER ET AL (2021) DATA
Check out testing in "SI_4_msy_Ch2_Overfishing.R"