---
title: 'OHI GoC 2025: Sense of Place CCE'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# 0. Workflow for this script

This is the Sense of Place coastal community engagement script for lasting non natural cultural places.

1. Read and clean the cultural places file from Script 1.
2. Bring in coastal population within 10 km.
3. Join culture with population.
4. Scale to population and plot rates.
5. Look at which activity types drive Region_1 and Region_3 growth from 2023 to 2026.
6. Compare current status options.

See `tr_quantity_updated_scripts/script1_denue_prep.qmd` (section 5) for how I build the `non_natural_cultural_places_2019_2026_goc_10km.shp` file.

# 1. Read and clean cultural places

I will read in the non natural cultural places and estimate employees from the worker size labels.

```{r}
library(tidyverse)
library(sf)
library(mapview)
library(here)
library(plotly)

inland_10 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_10km_buffer", "inland_10km_goc.shp"))

culture_int_dir <- "/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_esparcimiento_culturales_y_deportivos_servicios_recreativos/int"
culture_shp_new <- here::here(culture_int_dir, "non_natural_cultural_places_2019_2026_goc_10km.shp")

non_natural_culture_gulf_10km <- st_read(culture_shp_new)

# keep one row per place for later checks, then summarize by municipio for scoring
non_natural_culture_places <- non_natural_culture_gulf_10km %>% 
  mutate(nombre_act = nombre_act %>% 
           str_replace_all("<fa>", "ú") %>%
           str_replace_all("<e1>", "á") %>%
           str_replace_all("<ed>", "í") %>%
           str_replace_all("<f1>", "ñ") %>%
           str_replace_all("<e9>", "é"),
         
         municipio = municipio %>%
           str_replace_all("<fa>", "ú") %>%
           str_replace_all("<e1>", "á") %>%
           str_replace_all("<ed>", "í") %>%
           str_replace_all("<f1>", "ñ") %>%
           str_replace_all("<e9>", "é")
  ) %>% 
  mutate(municipio_clean = municipio %>% 
           str_to_lower() %>%
           str_replace_all(" ", "_"),
         municipio_clean = case_when(
           municipio == "San Ignacio Río Muerto" ~ "cajeme",
           municipio == "Bácum" ~ "cajeme",
           municipio == "Mexicali" ~ "san_felipe", # because these locations are already filtered to 10km coastal of GoC, so if in Mexicali then it was before the municipality San Felipe was made.
           municipio == "Sinaloa" ~ "guasave",
           municipio == "Santa María del Oro" ~ "san_pedro_lagunillas",
           TRUE ~ municipio_clean
           )) %>% 
  mutate(
    employees_estimated = case_when(
      per_ocu == "0 a 5 personas" ~ 3,
      per_ocu == "6 a 10 personas" ~ 8,
      per_ocu == "11 a 30 personas" ~ 20.5,
      per_ocu == "31 a 50 personas" ~ 40.5,
      per_ocu == "51 a 100 personas" ~ 75.5,
      per_ocu == "101 a 250 personas" ~ 175.5,
      per_ocu %in% c("251 y m<e1>s personas", "251 y más personas") ~ 300,
      TRUE ~ NA_real_
    )
  )

# ################################################################################
# CHANGE HERE: drop "Marinas turísticas" from the CCE non natural cultural places
# panel (nombre_act filter).
# need to discuss this further, but these should not be within the CCE non natural
# cultural places subcomponent, as that category is already separated from cultural
# recreation economic units in script1_denue_prep.qmd (marine tourism / nautical).
# Tourist marinas are commercial waterfront infrastructure more than community
# cultural venues. This also avoids treating San Felipe marina re registration as
# a real new cultural place.
# ################################################################################
non_natural_culture_places <- non_natural_culture_places %>%
  filter(nombre_act != "Marinas turísticas")

non_natural_culture_gulf_10km_clean <- non_natural_culture_places %>% 
  group_by(municipio_clean, data_year, rgn_id) %>% 
  summarize(
    agg_employees = sum(employees_estimated),
    n_units = n(), # number of economic units (establishments)
    .groups = "drop"
  )

rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)
```

# 2. Bring in coastal population

Now bring in population data by municipality, limited to the 10 km coastal strip.

```{r}
## main cities/towns 2020 population
cities <- st_read(here::here("/home/shares/ohi/OHI_GOC/goal_prep/pressures/population_localities/inegi_boundaries_pop.shp")) %>%
  janitor::clean_names()

mapview(cities) + mapview(non_natural_culture_gulf_10km_clean, zcol = "agg_employees")

## but first, we need to clean the population data.

cities_clean <- st_intersection(st_transform(cities, st_crs(inland_10)), inland_10) %>% 
  mutate(municipio = nom_mun %>%
           str_replace_all("<fa>", "ú") %>%
           str_replace_all("<e1>", "á") %>%
           str_replace_all("<ed>", "í") %>%
           str_replace_all("<f1>", "ñ") %>%
           str_replace_all("<e9>", "é")
  ) %>%
  mutate(municipio_clean = municipio %>%
           str_to_lower() %>%
           str_replace_all(" ", "_"),
         municipio_clean = case_when(
           municipio == "San Ignacio Río Muerto" ~ "cajeme",
           municipio == "Bácum" ~ "cajeme",
           municipio == "Mexicali" ~ "san_felipe", # because these locations are already filtered to 10km coastal of GoC, so if in Mexicali then it was before the municipality San Felipe was made.
           municipio == "Sinaloa" ~ "guasave",
           municipio == "Santa María del Oro" ~ "san_pedro_lagunillas",
           TRUE ~ municipio_clean
         )) %>% 
  dplyr::select(municipio_clean, rgn_id, pobtot, area_km2, pobtot_km2)

```


<!-- ## Bring in population by year -->

```{r}
# pop_goc <- read_csv(here::here("/home/shares/ohi/OHI_GOC/spatial/population_municipal","population_goc_municipality_inegi_2000_2020.csv"))
# 
# # target municipios
# municipios_culturales_lugares <- unique(non_natural_culture_gulf_10km_clean$municipio_clean)
# 
# ## aggregate by region
# agg_pop_rgnid <- pop_goc %>% 
#   mutate(municipio = desc_municipio %>%
#            str_replace_all("<fa>", "ú") %>%
#            str_replace_all("<e1>", "á") %>%
#            str_replace_all("<ed>", "í") %>%
#            str_replace_all("<f1>", "ñ") %>%
#            str_replace_all("<e9>", "é")
#   ) %>% 
#   mutate(municipio_clean = municipio %>% 
#            str_to_lower() %>%
#            str_replace_all(" ", "_"),
#          municipio_clean = case_when(
#            municipio == "San Ignacio Río Muerto" ~ "cajeme",
#            municipio == "Bácum" ~ "cajeme",
#            municipio == "Sinaloa" ~ "guasave",
#            municipio == "Santa María del Oro" ~ "san_pedro_lagunillas",
#            TRUE ~ municipio_clean
#          )) %>% 
#   filter(municipio_clean %in% municipios_culturales_lugares) %>% 
#   group_by(rgn_id, ano_censal, unidad_medida) %>% 
#   summarize(total_rgn_pop = sum(pop_count, na.rm = TRUE)) %>% 
#   ungroup()

# unfortunately, I think we really need to wait until 2025 is released.  We need to use the population locality polygons for this, since we are limiting to 10km coastal. There are many municipalities that have a majority of their population outside of that 10km. 
mapview(cities) + mapview(inland_10)
```

# 3. Join culture with population

```{r}
non_natural_culture_gulf_10km_clean_join <- non_natural_culture_gulf_10km_clean %>% 
  st_drop_geometry()

cities_clean_join <- cities_clean %>% 
  st_drop_geometry() %>% 
  group_by(municipio_clean) %>% 
  summarize(
    pobtot = sum(pobtot),
    area_km2 = sum(area_km2),
    pobtot_km2 = sum(pobtot_km2)
  ) %>% 
  ungroup()
  

join_culture_pop_10km <- inner_join(non_natural_culture_gulf_10km_clean_join, cities_clean_join)
```

# 4. Scale to population and plot rates

I will aggregate employees and places to each region and year, then scale both to coastal population.

```{r}
## aggregate employees and economic units to each region and year, then scale both to coastal population
# rgn_colors defined in section 1

cultural_places_scaled_to_population <- join_culture_pop_10km %>% 
  group_by(rgn_id, data_year) %>% 
  summarize(
    agg_employees = sum(agg_employees),
    n_units = sum(n_units),
    pobtot = sum(pobtot),
    area_km2 = sum(area_km2),
    pobtot_km2 = sum(pobtot_km2),
    .groups = "drop"
  ) %>% 
  mutate(
    culture_emp_per_local_pop = agg_employees / pobtot,
    culture_units_per_local_pop = n_units / pobtot
  )
```


```{r}
## plot the culture_emp_per_local_pop over time
culture_gulf_emp_per_pop_plotly <- plotly::plot_ly(
  cultural_places_scaled_to_population, 
  x = ~data_year, 
  y = ~culture_emp_per_local_pop, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Non natural cultural places: employees per coastal population",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "culture_emp_per_local_pop")
  ) 
culture_gulf_emp_per_pop_plotly
```

```{r}
## plot the culture_units_per_local_pop over time
culture_gulf_units_per_pop_plotly <- plotly::plot_ly(
  cultural_places_scaled_to_population, 
  x = ~data_year, 
  y = ~culture_units_per_local_pop, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Non natural cultural places: economic units per coastal population",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "culture_units_per_local_pop")
  ) 
culture_gulf_units_per_pop_plotly
```

# 5. What is driving Region_1 and Region_3 growth from 2023 to 2026?

Region_1 and Region_3 jump a lot after 2023 in the current status plots. Coastal population is fixed (2020 localities), so the score change has to come from more places and / or more estimated employees in the DENUE file.

Here I will check:

1. raw place counts and employees by year for those regions,
2. which activity types (`nombre_act`) grew or appeared,
3. which municipios added the most places,
4. whether worker size labels shifted toward bigger places.

If growth is only a few activity types or one municipio, that is useful to know. If whole new activity labels appear that never existed before 2024, I will look closer for a cleaning or filter issue.

## 5.1 Raw counts for Region_1 and Region_3

```{r}
focus_rgns <- c("Region_1", "Region_3")

places_focus <- non_natural_culture_places %>% 
  st_drop_geometry() %>% 
  filter(rgn_id %in% focus_rgns)

# region year totals (no population scaling)
focus_year_totals <- places_focus %>% 
  group_by(rgn_id, data_year) %>% 
  summarize(
    n_units = n(),
    n_emp = sum(employees_estimated, na.rm = TRUE),
    n_municipios = n_distinct(municipio_clean),
    n_activity_types = n_distinct(nombre_act),
    .groups = "drop"
  ) %>% 
  arrange(rgn_id, data_year)

focus_year_totals

focus_units_plotly <- plot_ly(
  focus_year_totals,
  x = ~data_year,
  y = ~n_units,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines+markers"
) %>% 
  layout(
    title = "Region_1 and Region_3: number of cultural places by year",
    xaxis = list(title = "Year"),
    yaxis = list(title = "n_units")
  )
focus_units_plotly

focus_emp_plotly <- plot_ly(
  focus_year_totals,
  x = ~data_year,
  y = ~n_emp,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines+markers"
) %>% 
  layout(
    title = "Region_1 and Region_3: estimated employees by year",
    xaxis = list(title = "Year"),
    yaxis = list(title = "estimated employees")
  )
focus_emp_plotly
```

## 5.2 Activity types by year

```{r}
# counts by activity type and year
focus_act_year <- places_focus %>% 
  count(rgn_id, data_year, nombre_act, name = "n_units") %>% 
  arrange(rgn_id, data_year, desc(n_units))

focus_act_year

# wide view for 2023 vs 2026: what grew, what is new
focus_act_2023_2026 <- focus_act_year %>% 
  filter(data_year %in% c(2023, 2026)) %>% 
  tidyr::pivot_wider(
    names_from = data_year,
    values_from = n_units,
    values_fill = 0,
    names_prefix = "y"
  ) %>% 
  mutate(
    delta_2023_to_2026 = y2026 - y2023,
    status = case_when(
      y2023 == 0 & y2026 > 0 ~ "new in 2026 (not in 2023)",
      y2023 > 0 & y2026 == 0 ~ "present in 2023, gone by 2026",
      delta_2023_to_2026 > 0 ~ "grew",
      delta_2023_to_2026 < 0 ~ "shrank",
      TRUE ~ "same count"
    )
  ) %>% 
  arrange(rgn_id, desc(delta_2023_to_2026))

focus_act_2023_2026

# activity types that did not exist in that region in 2023 but do in 2026
focus_new_act_types <- focus_act_2023_2026 %>% 
  filter(status == "new in 2026 (not in 2023)")

focus_new_act_types

focus_act_r1_plotly <- plot_ly(
  focus_act_year %>% filter(rgn_id == "Region_1", data_year >= 2023),
  x = ~data_year,
  y = ~n_units,
  color = ~nombre_act,
  type = "scatter",
  mode = "lines+markers"
) %>% 
  layout(
    title = "Region_1: places by activity type (2023 to 2026)",
    xaxis = list(title = "Year"),
    yaxis = list(title = "n_units"),
    legend = list(orientation = "h", y = -0.25)
  )
focus_act_r1_plotly

focus_act_r3_plotly <- plot_ly(
  focus_act_year %>% filter(rgn_id == "Region_3", data_year >= 2023),
  x = ~data_year,
  y = ~n_units,
  color = ~nombre_act,
  type = "scatter",
  mode = "lines+markers"
) %>% 
  layout(
    title = "Region_3: places by activity type (2023 to 2026)",
    xaxis = list(title = "Year"),
    yaxis = list(title = "n_units"),
    legend = list(orientation = "h", y = -0.25)
  )
focus_act_r3_plotly
```

## 5.3 Which municipios added places?

```{r}
focus_muni_2023_2026 <- places_focus %>% 
  filter(data_year %in% c(2023, 2026)) %>% 
  count(rgn_id, municipio_clean, data_year, name = "n_units") %>% 
  tidyr::pivot_wider(
    names_from = data_year,
    values_from = n_units,
    values_fill = 0,
    names_prefix = "y"
  ) %>% 
  mutate(delta_2023_to_2026 = y2026 - y2023) %>% 
  arrange(rgn_id, desc(delta_2023_to_2026))

focus_muni_2023_2026

# also break municipio growth by activity type for the biggest gainers
top_muni_gains <- focus_muni_2023_2026 %>% 
  filter(delta_2023_to_2026 > 0) %>% 
  group_by(rgn_id) %>% 
  slice_max(delta_2023_to_2026, n = 5) %>% 
  ungroup()

focus_muni_act_detail <- places_focus %>% 
  filter(
    data_year %in% c(2023, 2026),
    municipio_clean %in% top_muni_gains$municipio_clean
  ) %>% 
  count(rgn_id, municipio_clean, nombre_act, data_year, name = "n_units") %>% 
  tidyr::pivot_wider(
    names_from = data_year,
    values_from = n_units,
    values_fill = 0,
    names_prefix = "y"
  ) %>% 
  mutate(delta_2023_to_2026 = y2026 - y2023) %>% 
  filter(delta_2023_to_2026 != 0) %>% 
  arrange(rgn_id, municipio_clean, desc(delta_2023_to_2026))

focus_muni_act_detail
```

## 5.4 Did worker size labels change?

If estimated employees rise much faster than place counts, growth may be bigger places rather than more places.

```{r}
focus_size_year <- places_focus %>% 
  filter(data_year >= 2023) %>% 
  count(rgn_id, data_year, per_ocu, name = "n_units") %>% 
  arrange(rgn_id, data_year, desc(n_units))

focus_size_year

focus_size_summary <- places_focus %>% 
  filter(data_year %in% c(2023, 2026)) %>% 
  group_by(rgn_id, data_year) %>% 
  summarize(
    n_units = n(),
    n_emp = sum(employees_estimated, na.rm = TRUE),
    emp_per_place = n_emp / n_units,
    .groups = "drop"
  ) %>% 
  arrange(rgn_id, data_year)

focus_size_summary
```

## 5.5 Quick sanity notes while I read the tables

Things that would worry me:

- brand new `nombre_act` labels that look like encoding junk,
- one municipio suddenly gaining dozens of places of one odd type,
- employee totals jumping while economic unit counts barely change (number of employees issue??).

Things that would reassure me:

- growth is mostly more of the same activity types already present in 2023,
- several municipios contribute,
- place/economic unit counts and employees move together.

# 6. Current status alternatives

I will compare several ways of turning the per capita rates into a 0 to 100 style current status.

For the quantile versions (90th, 80th, and 75th), every region and year is compared to the same cross regional, cross year reference:

`current status = value (current) / value (quantile of all regions and years) * 100`

For the 2019 reference versions, each region is compared to its own 2019 baseline so we can see whether it is increasing or decreasing relative to itself:

`current status = value (year) / value (2019, same region) * 100`

Here 100 means matching the 2019 baseline. Values above 100 mean an increase since 2019. I leave those uncapped in `current_status` so the direction of change is visible, and also store a capped `score` (max 100) for OHI style bookkeeping.

## 6.1 Employees per person, 90th quantile

```{r}
## find the 90th quantile of employees per capita
cultural_emp_90th <- cultural_places_scaled_to_population %>% 
  mutate(
    percap_90th = quantile(culture_emp_per_local_pop, probs = 0.9),
    current_status = (culture_emp_per_local_pop / percap_90th) * 100
  ) %>% 
  mutate(score = ifelse(current_status > 100, 100, current_status))

write_csv(
  cultural_emp_90th,
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026/output",
    "cultural_places_scaled_to_population_current_status_quantile_041026.csv"
  )
)
```


```{r}
culture_emp_90th_plotly <- plotly::plot_ly(
  cultural_emp_90th, 
  x = ~data_year, 
  y = ~score, 
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: employees per capita, scaled to the 90th quantile",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cultural_emp_90th$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
culture_emp_90th_plotly
```

## 6.2 Employees per person, 80th quantile

```{r}
## find the 80th quantile of employees per capita
cultural_emp_80th <- cultural_places_scaled_to_population %>% 
  mutate(
    percap_80th = quantile(culture_emp_per_local_pop, probs = 0.8),
    current_status = (culture_emp_per_local_pop / percap_80th) * 100
  ) %>% 
  mutate(score = ifelse(current_status > 100, 100, current_status))

write_csv(
  cultural_emp_80th,
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026/output",
    "cultural_places_employees_per_pop_current_status_80th_quantile.csv"
  )
)
```

```{r}
culture_emp_80th_plotly <- plotly::plot_ly(
  cultural_emp_80th, 
  x = ~data_year, 
  y = ~score, 
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: employees per capita, scaled to the 80th quantile",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cultural_emp_80th$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
culture_emp_80th_plotly
```

## 6.3 Employees per person, 75th quantile

```{r}
## find the 75th quantile of employees per capita
cultural_emp_75th <- cultural_places_scaled_to_population %>% 
  mutate(
    percap_75th = quantile(culture_emp_per_local_pop, probs = 0.75),
    current_status = (culture_emp_per_local_pop / percap_75th) * 100
  ) %>% 
  mutate(score = ifelse(current_status > 100, 100, current_status))

write_csv(
  cultural_emp_75th,
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026/output",
    "cultural_places_employees_per_pop_current_status_75th_quantile.csv"
  )
)

# temporary 2025 gapfill: denue 2025 is not available yet, so copy each region's 2024 score into 2025.
# this is saved as a separate csv so we can reverse it once 2025 denue arrives
# (switch the sp merge back to the non gapfilled csv above).
# cultural_emp_75th_gapfill2025 <- cultural_emp_75th %>%
#   filter(data_year == 2024) %>%
#   mutate(
#     data_year = 2025L,
#     gapfill_note = "copied from 2024; temporary until 2025 DENUE is available"
#   ) %>%
#   bind_rows(cultural_emp_75th) %>%
#   arrange(rgn_id, data_year)

# write_csv(
#   cultural_emp_75th_gapfill2025,
#   here::here(
#     "/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026/output",
#     "cultural_places_employees_per_pop_current_status_75th_quantile_gapfill2025.csv"
#   )
# )
```

```{r}
culture_emp_75th_plotly <- plotly::plot_ly(
  cultural_emp_75th, 
  x = ~data_year, 
  y = ~score, 
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: employees per capita, scaled to the 75th quantile",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cultural_emp_75th$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
culture_emp_75th_plotly
```

## 6.4 Employees per person, 2019 regional reference

```{r}
cultural_emp_2019ref <- cultural_places_scaled_to_population %>% 
  group_by(rgn_id) %>% 
  mutate(
    emp_percap_2019 = culture_emp_per_local_pop[data_year == 2019][1],
    current_status = (culture_emp_per_local_pop / emp_percap_2019) * 100
  ) %>% 
  ungroup() %>% 
  mutate(score = ifelse(current_status > 100, 100, current_status))

write_csv(
  cultural_emp_2019ref,
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026/output",
    "cultural_places_employees_per_pop_current_status_2019ref.csv"
  )
)
```

```{r}
culture_emp_2019ref_plotly <- plotly::plot_ly(
  cultural_emp_2019ref, 
  x = ~data_year, 
  y = ~current_status, 
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: employees per capita relative to each region's 2019 baseline",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (100 = 2019 baseline)")
  ) %>%
  add_lines(
    x = range(cultural_emp_2019ref$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "2019 baseline (100)",
    inherit = FALSE
  )
culture_emp_2019ref_plotly
```

## 6.5 Places per person, 90th quantile

```{r}
cultural_units_90th <- cultural_places_scaled_to_population %>% 
  mutate(
    units_percap_90th = quantile(culture_units_per_local_pop, probs = 0.9),
    current_status = (culture_units_per_local_pop / units_percap_90th) * 100
  ) %>% 
  mutate(score = ifelse(current_status > 100, 100, current_status))

write_csv(
  cultural_units_90th,
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026/output",
    "cultural_places_units_per_pop_current_status_quantile.csv"
  )
)
```

```{r}
culture_units_90th_plotly <- plotly::plot_ly(
  cultural_units_90th, 
  x = ~data_year, 
  y = ~score, 
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: economic units per capita, scaled to the 90th quantile",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cultural_units_90th$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
culture_units_90th_plotly
```

## 6.6 Places per person, 80th quantile

```{r}
cultural_units_80th <- cultural_places_scaled_to_population %>% 
  mutate(
    units_percap_80th = quantile(culture_units_per_local_pop, probs = 0.8),
    current_status = (culture_units_per_local_pop / units_percap_80th) * 100
  ) %>% 
  mutate(score = ifelse(current_status > 100, 100, current_status))

write_csv(
  cultural_units_80th,
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026/output",
    "cultural_places_units_per_pop_current_status_80th_quantile.csv"
  )
)
```

```{r}
culture_units_80th_plotly <- plotly::plot_ly(
  cultural_units_80th, 
  x = ~data_year, 
  y = ~score, 
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: economic units per capita, scaled to the 80th quantile",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cultural_units_80th$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
culture_units_80th_plotly
```

## 6.7 Places per person, 75th quantile

```{r}
cultural_units_75th <- cultural_places_scaled_to_population %>% 
  mutate(
    units_percap_75th = quantile(culture_units_per_local_pop, probs = 0.75),
    current_status = (culture_units_per_local_pop / units_percap_75th) * 100
  ) %>% 
  mutate(score = ifelse(current_status > 100, 100, current_status))

write_csv(
  cultural_units_75th,
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026/output",
    "cultural_places_units_per_pop_current_status_75th_quantile.csv"
  )
)
```

```{r}
culture_units_75th_plotly <- plotly::plot_ly(
  cultural_units_75th, 
  x = ~data_year, 
  y = ~score, 
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: economic units per capita, scaled to the 75th quantile",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cultural_units_75th$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
culture_units_75th_plotly
```

## 6.8 Places per person, 2019 regional reference

```{r}
cultural_units_2019ref <- cultural_places_scaled_to_population %>% 
  group_by(rgn_id) %>% 
  mutate(
    units_percap_2019 = culture_units_per_local_pop[data_year == 2019][1],
    current_status = (culture_units_per_local_pop / units_percap_2019) * 100
  ) %>% 
  ungroup() %>% 
  mutate(score = ifelse(current_status > 100, 100, current_status))

write_csv(
  cultural_units_2019ref,
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026/output",
    "cultural_places_units_per_pop_current_status_2019ref.csv"
  )
)
```

```{r}
culture_units_2019ref_plotly <- plotly::plot_ly(
  cultural_units_2019ref, 
  x = ~data_year, 
  y = ~current_status, 
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: economic units per capita relative to each region's 2019 baseline",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (100 = 2019 baseline)")
  ) %>%
  add_lines(
    x = range(cultural_units_2019ref$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "2019 baseline (100)",
    inherit = FALSE
  )
culture_units_2019ref_plotly
```
