################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```

## clean_names_lowercase function
```{r Load clean_names_lowercase function}
# Takes text and standardizes it, which we will need for merging later on:

clean_names_lowercase <- function(x) {

  x <- iconv(x, from = "", to = "UTF-8", sub = "")
  x <- stringi::stri_trans_general(x, "Latin-ASCII")

  x %>%
    tolower() %>%
    trimws()
}
```

## lowercase_function function
```{r lowercase_function}
# Turns all entries lower-case
lowercase_function<- function(x){
  if(is.character(x)){
    tolower(x)
  } else if (is.factor(x)){
    tolower(as.character(x))
  } else {
    x
  }
}
```


# PICK 1: RAM Stock Assessment - species level ------------------------------

# >> SKIP: I have already downloaded and gapfilled the RAM data through 2024.
```{r Load in RAM and formal assessment data}
RAM_data0<- read.csv("/home/ferrer/ohi_goc_prep/FP/Exploratory_Research/FIS_TeamResearch/data/formal_assessments.csv")

table(RAM_data0$year)
# 195 obs., 17 vars.
```

```{r Separate Genus from scientific_name}
RAM_data<- RAM_data0

RAM_data <- RAM_data %>%
  separate(scientificname, into = c("Genus", "Species"), sep = " ", remove = FALSE)
# Okay, looks good!
# 195 obs., 19 vars.
```

```{r Basic filtering & correcting}
# Keep Stock (genus-species in this case) only if it has post-2019 data
RAM_data <- RAM_data %>%
  group_by(Genus, Species) %>%
  filter(any(year >= 2019)) %>%
  select(year, assessyear, areaname, 
         commonname, scientificname, Genus, Species, 
         RAM_BBmsy_raw= ram_bmsy, source) %>%
  ungroup()
# 181 obs., 9 vars.

# Correct the non_numeric values:
RAM_data$RAM_BBmsy_numeric <- as.numeric(gsub("[^0-9.]", "", RAM_data$RAM_BBmsy_raw))
```

```{r Gapfill RAM data using the zoo package}
RAM_gapfilled0 <- RAM_data %>%
  group_by(Genus, Species, areaname) %>%
  complete(year = 2019:2024) %>%
  ungroup()
# 256 obs., 10 vars.

# Gapfilling - INTERPOLATION between years with data (using the zoo package):
RAM_gapfilled <- RAM_gapfilled0 %>%
  # --- identify each stock correctly ---
  group_by(Genus, Species, areaname) %>%
  
  # --- add missing years WITHOUT removing any original rows ---
  complete(year = full_seq(c(year, 2019:2024), 1)) %>%
  # ^ This safely adds 2019–2024 and any intermediate years, 
  #   never removes rows
  
  # --- fill descriptors (stock metadata) ---
  fill(commonname, scientificname, source, assessyear, .direction = "downup") %>%
  
  arrange(year, .by_group = TRUE) %>%
  
  # --- interpolate only RAM_BBmsy_numeric ---
  mutate(
    RAM_gapfilled0 = zoo::na.approx(RAM_BBmsy_numeric, x = year, na.rm = FALSE)
  ) %>%
  
  ungroup()
# 347 obs., 11 vars.

# Gapfilling - EXTRAPOLATION beyond years with data (using the zoo package):
RAM_gapfilled <- RAM_gapfilled %>%
  group_by(Genus, Species, areaname) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(
    RAM_gapfilled = zoo::na.approx(
      RAM_BBmsy_numeric,
      x = year,
      na.rm = FALSE,
      rule = 2   # allow linear extrapolation outside observed years
    )
  ) %>%
  ungroup()
# 347 obs., 11 vars.

# Sanity check: do gapfilled values always match numeric values when they exist?
# Example vectors
numeric_BBmsy_values<- RAM_gapfilled$RAM_BBmsy_numeric
gapfilled_BBmsy_values <- RAM_gapfilled$RAM_gapfilled

# Logical check: TRUE wherever BBmsy_numeric is NA or matches RAM_gapfilled
TEST <- ifelse(!is.na(numeric_BBmsy_values), gapfilled_BBmsy_values == numeric_BBmsy_values, TRUE)

# Are all original values preserved?
all(TEST)
# Yes!
```

```{r Annotate RAM_gapfilled dataframe}
# Finally, a "RAM_BBmsy_final" column and "RAM_provenance" column that tell which BBmsy value to work with and where that data point came from, a formal assessment value or a gapfilled value:
RAM_gapfilled <- RAM_gapfilled %>%
  mutate(
    RAM_BBmsy_final = ifelse(!is.na(RAM_BBmsy_numeric), RAM_BBmsy_numeric, RAM_gapfilled),
    RAM_provenance    = ifelse(!is.na(RAM_BBmsy_numeric), "Direct from formal assessment (RAM or other)", "gapfilled from formal assessment")
  )
# 347 obs., 13 vars.

table(RAM_gapfilled$RAM_provenance)
# Direct from formal assessment (RAM or other)             gapfilled from formal assessment 
#                                          181                                          166
```

```{r Cap BBmsy at 1}
colnames(RAM_gapfilled)

# Caps values for RAM_BBmsy_final at 1 & filters out 2025 data:
RAM_gapfilled_final<- RAM_gapfilled %>%
  filter(year <= 2024) %>%
  mutate(RAM_BBmsy_capped = pmin(RAM_BBmsy_final, 1)) %>%
  select(year, Area= areaname, Common_Name= commonname,
         ScientificName= scientificname, Genus, Species, FormalAssesment_Period= assessyear,
         RAM_BBmsy_raw, RAM_BBmsy_numeric, RAM_gapfilled,
         RAM_BBmsy_final, RAM_BBmsy_capped,
         RAM_provenance, FormalAssessment_Source= source)
# 346 obs., 14 vars.
```

# SAVE the RAM_gapfilled dataframe:
```{r Save RAM_gapfilled as needed}
write.csv(RAM_gapfilled_final,
          "~/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_RAM_Scores/RAM_gapfilled_assessments_June032026.csv", 
          row.names = FALSE)
```


# >> SKIP TO HERE: Convert RAM data into a usable BBMSY score.
```{r Load in RAM_gapfilled0 as needed}
RAM_gapfilled0<- read.csv("~/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_RAM_Scores/RAM_gapfilled_assessments_June032026.csv")
# 346 obs., 14 vars.
```

```{r Convert RAM_gapfilled0 to RAM_FIS_StockStatus_scores}
RAM_FIS_StockStatus_scores<- RAM_gapfilled0
# 346 obs., 14 vars.

table(RAM_FIS_StockStatus_scores$FormalAssessment_Source)

# filter out estimates from the Carta:
RAM_FIS_StockStatus_scores <- RAM_gapfilled0 %>%
  filter(!FormalAssessment_Source %in% c(
    "carta",
    "carta: robalos are regulated as one group",
    "carta: rough estimate from plot",
    "carta: snappers are regulated as one taxa"
  ))
# 216 obs., 14 vars.
```

# >> FINAL STEP: SAVE the RAM_FIS_StockStatus_scores dataframe:
```{r Save this dataframe}
write.csv(RAM_FIS_StockStatus_scores,
          "~/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_RAM_Scores/FIS_RAM_Scores_Jun032026.csv", 
          row.names = FALSE)
```
