---
title: 'OHI GoC 2025: SP ICO Gulf of California'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# Summary

The Iconic Species (ICO) subgoal of Sense of Place captures the condition of iconic species that are relevant to local cultural identity.  We must (1) create a list of iconic species and which OHI regions they belong to, (2) determine their condition by their risk status, such as from NOM-059-SEMARNAT-2010, or IUCN, and (3) then, for each region, the average of the status for all iconic species can be calculated.

(1) = script1_iconic_species_list.qmd
(2) = script2_norma059.qmd, script3_cites_prep.qmd, script4_iucn_prep.qmd
(3) = script5_ico_current_status.qmd

In this first script, we will be bringing in the list developed by the Expert Working Group and assigning each species to a OHI GoC Region (1-9) based on a species occurence (presence and absence) database from Dr. Hem Nalini Morzaria-Luna.

```{r}
library(tidyverse)
library(here)
library(janitor)
```


## Read in list

```{r}
especies_iconicas <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/ico","especies_iconicas_040326.csv"))
```


## Clean to prep for joining with risk condition 


```{r}
especies_iconicas_scinames <- especies_iconicas %>% 
  mutate(sci_nombre = str_replace_all(sci_nombre, "spp.",""),
         sci_nombre = str_trim(sci_nombre, side = "both")) %>% 
  left_join(key_for_general_groups, by = "sci_nombre")

target_species <- unique(especies_iconicas_scinames$sci_nombre)

# making a regex pattern from the genus/species names
pattern <- paste(target_species, collapse = "|")
```


## Determine species occurence throughout the Gulf


```{r}
clean_species_occurrence <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/ico/clean_species_ocurrence_hem_2018.csv")) %>% 
  filter(wordnum %in% "2") %>% 
  dplyr::select(old_name, lat, lon, source) %>% 
  mutate(old_name = str_trim(old_name, side = "both"))
```


```{r}
ico_clean_species_occurrence <- clean_species_occurrence  %>%
  filter(str_detect(old_name, regex(pattern, ignore_case = TRUE)))

setdiff(unique(ico_clean_species_occurrence$old_name), target_species)

setdiff(target_species, unique(ico_clean_species_occurrence$old_name))
#  [1] "Penaeus brevirostris"      "Pinctada mazatlanica​​"     
#  [3] "Ilochione subrugosa"       "Tetrapturus albidus"      
#  [5] "Sphyrna "                  "Mycteroperca "            
#  [7] "Hyporthodus "              "Stereolepis "             
#  [9] "Epinephelus "              "Lagenorynchus obliquidens"
# [11] "Chelonia agassizi"         "Larus heermani"           
# [13] "Endomychura craveri"       "Pandion haeliatus"        
# [15] "Mobula "   
```

Many of the target species not in the occurrence database are the general genera, not the species name, so we can ignore those.  However, for species names, we need to look deeper at missing occurrence data.

So, we need to look specifically at "Penaeus brevirostris", "Ilochione subrugosa", "Pinctada mazatlanica", "Tetrapturus albidus", "Lagenorynchus obliquidens", "Chelonia agassizi", "Larus heermani", "Endomychura craveri", "Pandion haeliatus".

However, I know for a fact that Pinctada mazatlanica is in both dataframes.  So I am going to see if I can fix the regex...

```{r}
clean_string <- function(x) {
  x %>% 
    str_replace_all("[^[:ascii:]]", " ") %>%  # replace non-ASCII with space
    str_replace_all("\\s+", " ") %>% # collapse multiple spaces
    str_trim(side = "both")
}

# re-clean both sides
target_species_clean <- clean_string(target_species)

ico_clean_species_occurrence <- clean_species_occurrence %>% 
  mutate(old_name = clean_string(old_name))

# now check again
setdiff(target_species_clean, unique(ico_clean_species_occurrence$old_name)) # great, no Pinctada!!!

#  [1] "Penaeus brevirostris"      "Ilochione subrugosa"      
#  [3] "Tetrapturus albidus"       "Sphyrna"                  
#  [5] "Mycteroperca"              "Hyporthodus"              
#  [7] "Stereolepis"               "Epinephelus"              
#  [9] "Lagenorynchus obliquidens" "Chelonia agassizi"        
# [11] "Larus heermani"            "Endomychura craveri"      
# [13] "Pandion haeliatus"         "Mobula" 
```


*ICO List --> old species name on occurrence df:*

Penaeus brevirostris --> Farfantepenaeus brevirostris (old genus)
Ilochione subrugosa --> missing, not even the genus name
Tetrapturus albidus --> missing, not even the genus name
Lagenorynchus obliquidens --> Lagenorhynchus obliquidens, (note "h" in occurrence dataset)
Chelonia agassizi --> missing, but could be synonymous as Chelonia mydas (since it is a subspecies of green turtle).  However, the NORMA lists them separately. But, they are listed with the same risk category on the NORMA, so for now, let's just use Chelonia mydas overall.
Larus heermani --> Larus heermanni (note the double "n"s)
Endomychura craveri --> missing, not even the genus name
Pandion haeliatus --> Pandion haliaetus (note different species name spelling)


But I am realizing here that the occurrence data likely has the correct spelling, and my ICO list must be incorrect.  So let's update the iconic species list.


```{r}
clean_species_occurrence <- clean_species_occurrence %>%
  mutate(old_name = clean_string(old_name)) %>%
  mutate(old_name = case_when(
    old_name == "Farfantepenaeus brevirostris" ~ "Penaeus brevirostris",
    TRUE ~ old_name
  ))

especies_iconicas_scinames <- especies_iconicas_scinames %>%
  mutate(sci_nombre = clean_string(sci_nombre)) %>%
  mutate(sci_nombre = case_when(
    sci_nombre == "Lagenorynchus obliquidens" ~ "Lagenorhynchus obliquidens",
    sci_nombre == "Chelonia agassizi" ~ "Chelonia mydas",
    sci_nombre == "Larus heermani" ~ "Larus heermanni",
    sci_nombre == "Pandion haeliatus" ~ "Pandion haliaetus",
    TRUE ~ sci_nombre
  ))
```


Now, let us re-run the filtering using the pattern:

```{r}
target_species <- unique(especies_iconicas_scinames$sci_nombre)
general_groupings <- especies_iconicas_scinames %>% 
  select(sci_nombre, general_group)

# making a regex pattern from the genus/species names
pattern <- paste(target_species, collapse = "|")

ico_clean_species_occurrence <- clean_species_occurrence  %>%
  filter(str_detect(old_name, regex(pattern, ignore_case = TRUE))) %>% 
  left_join(general_groupings, by = c("old_name" = "sci_nombre")) %>% 
  mutate(general_group = case_when(
    str_detect(old_name, "^Sphyrna") ~ "sharks", # ^ indicates it starts with Sphyrna.
    is.na(general_group) ~ "fish",
    TRUE ~ general_group
  ))

setdiff(unique(ico_clean_species_occurrence$old_name), target_species)

setdiff(target_species, unique(ico_clean_species_occurrence$old_name))

## now, we are only missing these, most of which are just genus names:
# [1] "Ilochione subrugosa" "Tetrapturus albidus" "Sphyrna"            
# [4] "Mycteroperca"        "Hyporthodus"         "Stereolepis"        
# [7] "Epinephelus"         "Endomychura craveri" "Mobula"
```



## Assign OHI GoC Region ID

```{r}
library(sf)
library(mapview)

ico_clean_species_occurrence_sf <- ico_clean_species_occurrence %>% 
  st_as_sf(coords = c("lon", "lat"), # making sf object
           crs = 4326, # WGS84
           remove = FALSE)

st_write(ico_clean_species_occurrence_sf, here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/ico/int","ico_clean_species_occurrence_w_groups_sf.gpkg"), append=FALSE)
```

First, group by region and unite the polygons for inland 10km with the marine polygons.
```{r}
## intersect with inland_10:
inland_10 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_10km_buffer","inland_10km_goc.shp"))
regions_marine <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp")) %>%
  filter(location == "marine")

st_crs(inland_10) == st_crs(regions_marine) # TRUE

marine_and_inland10 <- bind_rows(regions_marine, inland_10) %>% 
  dplyr::select(rgn_id, geometry) %>% 
  group_by(rgn_id) %>% 
  summarize(
    geometry = st_union(geometry)
  )

mapview(marine_and_inland10)
```

Now, we can intersect with the `marine_and_inland10` so that each species occurrence is assigned an OHI GoC Region ID.
```{r}
# intersect with GoC
ico_clean_species_occurrence_goc_rgnid <- st_intersection(st_transform(ico_clean_species_occurrence_sf, st_crs(marine_and_inland10)), marine_and_inland10)

st_write(ico_clean_species_occurrence_goc_rgnid, here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/ico/int","ico_clean_species_occurrence_goc_rgnid_w_groups.gpkg"), append=FALSE)
```


We can group by rgn_id and determine how many times a species was found in that region.
```{r}
ico_species_occurrence_total_per_rgn <- ico_clean_species_occurrence_goc_rgnid %>% 
  group_by(rgn_id, old_name) %>% 
  summarize(
    n_occurrence = n()
  ) %>% 
  ungroup()
```

If we do not care about frequency of occurrence:
```{r}
ico_species_rgnid <- ico_species_occurrence_total_per_rgn %>% 
  st_drop_geometry() %>% 
  dplyr::select(taxon_scientific_name = old_name, rgn_id)

write_csv(ico_species_rgnid, here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/ico/int","ico_species_rgnid_040626.csv"))
```





## Making keys for pressures

```{r}
ico_clean_species_occurrence_goc_rgnid <- st_read(here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/ico/int","ico_clean_species_occurrence_goc_rgnid_w_groups.gpkg"))

# key_for_general_groups <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/ico/sp_ico_list_w_general_groups.csv")) %>% 
#   mutate(sci_nombre = str_replace_all(sci_nombre, "spp.",""),
#          sci_nombre = str_trim(sci_nombre, side = "both")) %>% 
#   select(sci_nombre, general_group = "general_group\n\nfish, sharks, shellfish, coral, mammals, turtles, birds, other")

ico_group_occurrence_total_per_rgn <- ico_clean_species_occurrence_goc_rgnid %>% 
  group_by(rgn_id, general_group) %>% 
  summarize(
    n_occurrence = n()
  ) %>% 
  ungroup()

# missing_general_group <- ico_clean_species_occurrence_goc_rgnid %>% 
#   filter(is.na(general_group))
# unique(missing_general_group$old_name)

proportion_of_groups_occurrence <- ico_group_occurrence_total_per_rgn %>% 
  group_by(rgn_id) %>% 
  mutate(total_occurrence = sum(n_occurrence)) %>% 
  ungroup() %>% 
  group_by(rgn_id, general_group) %>% 
  mutate(prop_of_rgn = n_occurrence / total_occurrence,
         pct_of_rgn = prop_of_rgn * 100) %>% 
  ungroup()

# save to mazu for pressures

```













