---
title: 'OHI GoC 2025: SP ICO Gulf of California'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# Summary

The Iconic Species (ICO) subgoal of Sense of Place captures the condition of iconic species that are relevant to local cultural identity.  We must (1) create a list of iconic species and which OHI regions they belong to, (2) determine their condition by their risk status, such as from NOM-059-SEMARNAT-2010, or IUCN, and (3) then, for each region, the average of the status for all iconic species can be calculated.

(1) = script1_iconic_species_list.qmd
(2) = script2_norma059.qmd, script3_cites_prep.qmd, script4_iucn_prep.qmd
(3) = script5_ico_current_status.qmd

In this final script, I will be calculating the proportion of general group occurrence within each region, to be used for pressure calculations.

```{r}
library(tidyverse)
library(here)
library(janitor)
```


## Making keys for pressures

```{r}
ico_clean_species_occurrence_goc_rgnid <- st_read(here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/ico/int","ico_clean_species_occurrence_goc_rgnid_w_groups.gpkg"))

# key_for_general_groups <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/ico/sp_ico_list_w_general_groups.csv")) %>% 
#   mutate(sci_nombre = str_replace_all(sci_nombre, "spp.",""),
#          sci_nombre = str_trim(sci_nombre, side = "both")) %>% 
#   select(sci_nombre, general_group = "general_group\n\nfish, sharks, shellfish, coral, mammals, turtles, birds, other")

ico_group_occurrence_total_per_rgn <- ico_clean_species_occurrence_goc_rgnid %>% 
  group_by(rgn_id, general_group) %>% 
  summarize(
    n_occurrence = n()
  ) %>% 
  ungroup()

# missing_general_group <- ico_clean_species_occurrence_goc_rgnid %>% 
#   filter(is.na(general_group))
# unique(missing_general_group$old_name)

proportion_of_groups_occurrence <- ico_group_occurrence_total_per_rgn %>% 
  group_by(rgn_id) %>% 
  mutate(total_occurrence = sum(n_occurrence)) %>% 
  ungroup() %>% 
  group_by(rgn_id, general_group) %>% 
  mutate(prop_of_rgn = n_occurrence / total_occurrence,
         pct_of_rgn = prop_of_rgn * 100) %>% 
  ungroup() %>% 
  mutate(goal = "ICO") %>% 
  select(rgn_id, general_group, prop_of_rgn, pct_of_rgn)

# save to mazu for pressures
st_write(proportion_of_groups_occurrence, here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/ico/for_pressures","proportion_of_groups_occurrence.gpkg"), append=FALSE)

unique(proportion_of_groups_occurrence$general_group)
unique(ico_clean_species_occurrence_goc_rgnid$source)
```


