---
title: 'OHI GoC 2025: SP ICO Gulf of California'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# Summary

The Iconic Species (ICO) subgoal of Sense of Place captures the condition of iconic species that are relevant to local cultural identity.  We must (1) create a list of iconic species and which OHI regions they belong to, (2) determine their condition by their risk status, such as from NOM-059-SEMARNAT-2010, or IUCN, and (3) then, for each region, the average of the status for all iconic species can be calculated.

(1) = script1_iconic_species_list.qmd
(2) = script2_norma059.qmd, script3_cites_prep.qmd, script4_iucn_prep.qmd
(3) = script5_ico_current_status.qmd

In this fourth script, we will be cleaning up IUCN information and assessing iconic species condition based on the IUCN Red List Categories and Criteria.

Note that if none of the iconic species are listed on NOM-059-SEMARNAT-2010, we will check to see if they are listed as *at risk* by CITES/IUCN.  If not, the subgoal would have a current status of 100 for that region and year.

The IUCN Red List divides species into nine categories: *Not Evaluated, Data Deficient, Least Concern, Near Threatened, Vulnerable, Endangered, Critically Endangered, Extinct in the Wild and Extinct.*

-   **Reference**: IUCN 2024. IUCN Red List of Threatened Species. Version 2025-1 \<www.iucnredlist.org\>
-   **Accessed**: July 15, 2025
-   **Download**: Download is done through iucnredlist API
-   **Native data resolution**: Country level (by country name)
-   **Time range**: 1965-2024 (discrete past assessments by species)
-   **Format**: JSON

Using the IUCN API, we accessed the full IUCN species list. With some minor formatting, this list contains the following variables:

iucn_sid \| kingdom \| phylum \| class \| order \| family \| genus \| sciname \| subpopulation \| category

The Status of this sub-goal (X~ICO~) is then the % of iconic species in each threat category (as defined by the IUCN Red List), such that:

$$X_{ICO} = \frac{\displaystyle\sum_{category}S_{cat}*w_{cat}}{\displaystyle\sum_{category}S_{cat}}$$

where for each IUCN threat category:

-   *S~cat~* is the number of assessed species in the category
-   *w~cat~* is the status weight assigned for that category (note, these are the inverse of the risk value used in the SPP calculations):
-   'LC' = 1.0, 'NT' = 0.8, 'VU' = 0.6, 'EN' = 0.4, 'CR' = 0.2, 'EX' = 0.0

ICO trend is calculated in a similar manner, but weightings are assigned according to IUCN population trend: 'Decreasing' = -0.5, 'Stable' = 0.0, 'Increasing' = +0.5.

## Download Species List from IUCN

Get total species list, using the IUCN API.

To access the API v4, you will need to sign up for a key on the [IUCN website](https://api.iucnredlist.org/). This key can be added to the local .renviron on your machine and retrieved using the code below.


### Setup

```{r}
if (!require(librarian)){install.packages("librarian")}
if (!require(ohicore)){devtools::install_github('ohi-science/ohicore@dev')}
librarian::shelf(
  tidyverse,
  here,
  janitor,
  jsonlite,
  plotly,
  parallel,
  devtools,
  ohicore,
  kableExtra,
  magrittr,
  rredlist
)  
```


```{r set up API}
library(rredlist)

remotes::install_github("iucn-uk/iucnredlist")
library(iucnredlist)

if (!requireNamespace("pak", quietly = TRUE)) {
  install.packages("pak", dependencies = TRUE)
}
pak::pkg_install("stangandaho/redlist")

library(redlist)

rl_check_api()
```

```{r}
rl_check_api()

rl_countries()

# 158 Mexico MX
# mexico_data <- rl_countries(code = "MX", latest = TRUE)
# mexico_2023 <- rl_countries(code = "MX", year_published = 2023)
# mexico_2024 <- rl_countries(code = "MX", year_published = 2024)
# str(mexico_data)
# names(mexico_data)
# mexico_data_df <- bind_rows(mexico_data$assessments) %>% 
#   dplyr::select(-scopes) # since all are Global, 1
```

Use the list of Iconic Species to pull IUCN Assessments:
```{r}
ico_spp_rgnid <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/ico/int","ico_species_rgnid_040626.csv"))

target_species <- unique(ico_spp_rgnid$taxon_scientific_name)

# making a regex pattern from the genus/species names
pattern <- paste(target_species, collapse = "|")
```


```{r}
# mexico_focal <- mexico_data_df %>%
#   filter(str_detect(taxon_scientific_name, regex(pattern, ignore_case = TRUE)))
# 
# found <- target_species[sapply(target_species, function(sp) 
#   any(str_detect(mexico_focal$taxon_scientific_name, regex(sp, ignore_case = TRUE))))]
# 
# # see which were NOT found
# missing <- setdiff(target_species, found)
# missing

#  [1] "Tivela planulata"       "Megapitaria aurantiaca" "Megapitaria squalida"   "Penaeus vannamei"      
#  [5] "Penaeus stylirostris"   "Penaeus californiensis" "Penaeus brevirostris"   "Pinna rugosa"          
#  [9] "Atrina tuberculosa"     "Atrina maura"           "Pteria sterna"          "Pinctada mazatlanica"  
# [13] "Chione californiensis"  "Chione undatella"       "Ilochione subrugosa"    "Tetrapturus audax"     
# [17] "Makaira ampla"          "Makaira indica"         "Tetrapturus albidus"    "Istiophorus albicans"  
# [21] "Distichlis palmeri"     "Delphinus capensis"     "Chelonia agassizii"     "Endomychura craveri" 
```


### Obtain IUCN Assessments for each year

I want to see change over time, so...

```{r}
years <- 2010:2025

mexico_all_years <- purrr::map(years, function(yr) {
  Sys.sleep(0.5)
  rl_countries(code = "MX", year_published = yr)$assessments %>%
    filter(str_detect(taxon_scientific_name, regex(pattern, ignore_case = TRUE)))
}) %>% 
  bind_rows()

iucn <- unique(mexico_all_years$taxon_scientific_name)
ico_spp <- unique(ico_spp_rgnid$taxon_scientific_name)
setdiff(ico_spp, iucn) # 42 spp are missing...
```


### Assign condition weights

```{r}
# *w~cat~* is the status weight assigned for that category (note, these are the inverse of the risk value used in the SPP calculations):
# -   'LC' = 1.0, 'NT' = 0.8, 'VU' = 0.6, 'EN' = 0.4, 'CR' = 0.2, 'EX' = 0.0

unique(mexico_all_years$red_list_category_code) # 2010-2025: [1] "LC" "NT" "VU" "CR" "DD" "EN"

# when including 200-2025: [1] "VU"    "LR/nt" "LR/lc" "CR"    "LC"    "EN"    "NT"    "NE"    "DD"    "NR"   
# "Lower Risk" (LR) is a formerly used category in the IUCN Red List (version 2.3, 1994-2000) for species not currently threatened with extinction. It was superseded by the Near Threatened (NT) and Least Concern (LC) categories. It was historically subdivided into Conservation Dependent (LR/cd), Near Threatened (LR/nt), and Least Concern (LR/lc).
# Therefore, they should be treated as such: LR/nt -> NT, and LR/lc -> LC

#### when we only assess 2010:2025, we do not have the issue of LR. But I will leave this information in case we ever want to go earlier than 2010.

mexico_all_years_w_condition <- mexico_all_years %>% 
  dplyr::select(-scopes) %>% # since all are Global, 1
  inner_join(ico_spp_rgnid) %>%
  dplyr::select(taxon_scientific_name, year_published, rgn_id, red_list_category_code) %>% 
  distinct() %>% # needed because some years had multiple assessments with different assessment ids but for the same species, with the same risk category.  For example, 2011 Coryphaena hippurus.
  mutate(w_cat = case_when(
    
    red_list_category_code %in% c("LC","LR/lc") ~ 1.0,
    red_list_category_code %in% c("NT","LR/nt") ~ 0.75,
    red_list_category_code == "VU" ~ 0.50,
    red_list_category_code == "EN" ~ 0.25,
    red_list_category_code == "CR" ~ 0.12,
    red_list_category_code == "EX" ~ 0.0,
    # red_list_category_code == "DD" ~ 99,
    TRUE ~ NA
    
  ))

mexico_all_years_w_condition_clean <- mexico_all_years_w_condition %>% 
  dplyr::select(taxon_scientific_name, rgn_id, year = year_published, red_list_category_code, w_cat_iucn = w_cat) 

old_mexico_all_years_w_condition_clean <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/ico/int","iucn_condition_ico_spp_040626.csv"))

setdiff(unique(mexico_all_years_w_condition_clean$taxon_scientific_name), unique(old_mexico_all_years_w_condition_clean$taxon_scientific_name))
# [1] "Larus heermanni"            "Lagenorhynchus obliquidens"
# [3] "Pandion haliaetus"  


setdiff(unique(old_mexico_all_years_w_condition_clean$taxon_scientific_name), unique(mexico_all_years_w_condition_clean$taxon_scientific_name))
#  [1] "Mobula birostris"            "Epinephelus adscensionis"   
#  [3] "Hyporthodus drummondhayi"    "Epinephelus guttatus"       
#  [5] "Epinephelus morio"           "Epinephelus striatus"       
#  [7] "Hyporthodus flavolimbatus"   "Hyporthodus mystacinus"     
#  [9] "Mycteroperca acutirostris"   "Mycteroperca bonaci"        
# [11] "Mycteroperca interstitialis" "Mycteroperca microlepis"    
# [13] "Mycteroperca phenax"         "Mycteroperca tigris"        
# [15] "Mycteroperca venenosa"       "Hyporthodus cifuentesi"     
# [17] "Mobula hypostoma"            "Mobula mobular" 

# write_csv(mexico_all_years_w_condition_clean, here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/ico/int","iucn_condition_ico_spp_041326_only_ico.csv"))
```

### Deal with DD

Search for literature regarding these species.

```{r}
## see which are NAs (meaning that they were NR (Not Recognized), NE (Not Evaluated), or DD (Data Deficient))

nas_mexico_all_years_w_condition <- mexico_all_years_w_condition %>% 
  filter(is.na(w_cat)) # note that all are DD or are not even considered by IUCN, none are NR or NE

unique(nas_mexico_all_years_w_condition$taxon_scientific_name)

# DD:
#  [1] "Istiompax indica"             "Tetrapturus angustirostris"  
#  [3] "Epinephelus quinquefasciatus" "Stenella longirostris"       
#  [5] "Orcinus orca"                 "Dosidicus gigas"             
#  [7] "Hyporthodus exsul"            "Hyporthodus niphobles"       
#  [9] "Hyporthodus drummondhayi"     "Mycteroperca tigris"         
# [11] "Mycteroperca phenax"          "Mycteroperca prionura"       
# [13] "Mycteroperca xenarcha"

# All, when joining w ICO list before assigning conidtion:
#  [1] "Istiompax indica"             "Tetrapturus angustirostris"  
#  [3] "Epinephelus quinquefasciatus" "Stenella longirostris"       
#  [5] "Orcinus orca"                 "Dosidicus gigas"             
#  [7] "Hyporthodus exsul"            "Hyporthodus niphobles"       
#  [9] "Hyporthodus drummondhayi"     "Mycteroperca tigris"         
# [11] "Mycteroperca phenax"          "Mycteroperca prionura"       
# [13] "Mycteroperca xenarcha"        "Tivela planulata"            
# [15] "Megapitaria aurantiaca"       "Megapitaria squalida"        
# [17] "Penaeus vannamei"             "Penaeus stylirostris"        
# [19] "Penaeus californiensis"       "Penaeus brevirostris"        
# [21] "Pinna rugosa"                 "Atrina tuberculosa"          
# [23] "Atrina maura"                 "Pteria sterna"               
# [25] "Pinctada mazatlanica​​"         "Chione californiensis"       
# [27] "Chione undatella"             "Ilochione subrugosa"         
# [29] "Tetrapturus albidus"          "Sphyrna "                    
# [31] "Mycteroperca "                "Hyporthodus "                
# [33] "Stereolepis "                 "Epinephelus "                
# [35] "Distichlis palmeri"           "Delphinus capensis"          
# [37] "Lagenorynchus obliquidens"    "Chelonia agassizi"           
# [39] "Eretmochelys imbricata"       "Lepidochelys olivacea"       
# [41] "Larus heermani"               "Endomychura craveri"         
# [43] "Pandion haeliatus"            "Mobula "    
```

















