---
title: 'OHI GoC 2026: CCE with Security subcomponent (GAPFILL)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    theme: cerulean
    highlight-style: haddock
editor_options:
  chunk_output_type: console
---

# Coastal Community Engagement (CCE) with Security — GAPFILL

Same three-way CCE mean as `new_cce_integration.qmd`, but **security NAs are gapfilled** before averaging:

1. **Temporal** — region mean across available ENSU years (e.g. Region 7 / Ciudad Obregón).
2. **Spatial** — GoC-only donors (no Nogales / border):
   - Regions 2 & 3 ← Region 4 (BCS)
   - Region 5 ← mean of Regions 6 & 7 (Sonora)

Ama tu Playa Region 2 NA (no beaches) is still left as NA and ignored in the CCE mean.

```{r}
library(tidyverse)

rgn_labels <- c(
  "Region_1" = "Western Upper Gulf",
  "Region_2" = "Western Midriff Islands",
  "Region_3" = "Western Central Gulf",
  "Region_4" = "Western Southern Gulf",
  "Region_5" = "Eastern Upper Gulf",
  "Region_6" = "Eastern Midriff Islands",
  "Region_7" = "Southern Sonora Coast",
  "Region_8" = "Sinaloa Coast",
  "Region_9" = "Southeastern Gulf"
)

rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

region_levels <- paste0("Region_", 1:9)
rgn_label_colors <- setNames(
  unname(rgn_colors[region_levels]),
  unname(rgn_labels[region_levels])
)

path_cce_out <- "/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026/output"
path_viz <- "/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026/viz"
path_security <- "/home/shares/ohi/OHI_GOC/goal_prep/ss/security_output/cs_security.csv"
path_security_gapfill <- "/home/shares/ohi/OHI_GOC/goal_prep/ss/security_output/cs_security_gapfill.csv"

dir.create(path_viz, recursive = TRUE, showWarnings = FALSE)
dir.create(path_cce_out, recursive = TRUE, showWarnings = FALSE)
```

# Gapfill helper (same rules as security_current_status_plot_GAPFILL.qmd)

```{r}
spatial_donors <- list(
  "Region_2" = c("Region_4"),
  "Region_3" = c("Region_4"),
  "Region_5" = c("Region_6", "Region_7")
)

gapfill_security <- function(security_raw, years, regions = region_levels) {
  out <- tidyr::crossing(rgn_id = regions, year = years) %>%
    dplyr::left_join(security_raw, by = c("rgn_id", "year")) %>%
    dplyr::mutate(
      security_observed = security_status,
      gapfilled = FALSE,
      gapfill_reason = NA_character_
    )

  rgn_means <- out %>%
    dplyr::filter(!is.na(security_status)) %>%
    dplyr::group_by(rgn_id) %>%
    dplyr::summarize(rgn_mean = mean(security_status), .groups = "drop")

  regions_with_data <- rgn_means$rgn_id

  out <- out %>%
    dplyr::left_join(rgn_means, by = "rgn_id") %>%
    dplyr::mutate(
      needs_temporal = is.na(security_status) & rgn_id %in% regions_with_data,
      security_status = dplyr::if_else(needs_temporal, rgn_mean, security_status),
      gapfilled = dplyr::if_else(needs_temporal, TRUE, gapfilled),
      gapfill_reason = dplyr::if_else(
        needs_temporal,
        "temporal: mean of same region across available ENSU years",
        gapfill_reason
      )
    ) %>%
    dplyr::select(-rgn_mean, -needs_temporal)

  for (target in names(spatial_donors)) {
    donors <- spatial_donors[[target]]
    for (yr in years) {
      donor_vals <- out %>%
        dplyr::filter(rgn_id %in% donors, year == yr, !is.na(security_status)) %>%
        dplyr::pull(security_status)
      if (length(donor_vals) == 0) {
        donor_vals <- out %>%
          dplyr::filter(rgn_id %in% donors, !is.na(security_status)) %>%
          dplyr::pull(security_status)
      }
      if (length(donor_vals) == 0) next

      fill_val <- mean(donor_vals)
      idx <- which(out$rgn_id == target & out$year == yr & is.na(out$security_status))
      if (length(idx) == 0) next

      out$security_status[idx] <- fill_val
      out$gapfilled[idx] <- TRUE
      out$gapfill_reason[idx] <- paste0(
        "spatial: GoC donor mean from ",
        paste(donors, collapse = ", "),
        " (excludes border cities)"
      )
    }
  }

  out
}
```

# Read the three subcomponents

```{r}
# 1) Ama tu Playa
ama <- read_csv(
  file.path(path_cce_out, "current_status_ama_tu_playa_progress_072326.csv"),
  show_col_types = FALSE
) %>%
  select(
    rgn_id,
    year,
    ama_status = current_status,
    any_of("reason")
  )

if (!"year" %in% names(ama)) {
  ama <- ama %>% mutate(year = 2025L)
}

if (!"reason" %in% names(ama)) {
  ama <- ama %>% mutate(reason = NA_character_)
}

if (!"Region_2" %in% ama$rgn_id) {
  ama <- bind_rows(
    ama,
    crossing(rgn_id = "Region_2", year = 2019:2025) %>%
      mutate(
        ama_status = NA_real_,
        reason = "no beaches in program inventory"
      )
  )
} else {
  ama <- ama %>%
    mutate(
      reason = if_else(
        rgn_id == "Region_2" & is.na(ama_status),
        "no beaches in program inventory",
        reason
      )
    )
}

ama <- ama %>% arrange(rgn_id, year)

# 2) Non-natural cultural places
cult <- read_csv(
  file.path(
    path_cce_out,
    "cultural_places_employees_per_pop_current_status_75th_quantile.csv"
  ),
  show_col_types = FALSE
) %>%
  select(rgn_id, year = data_year, cult_status = score)

# 3) Security (gapfilled)
# Prefer table from security_current_status_plot_GAPFILL.qmd if already written
if (file.exists(path_security_gapfill)) {
  security <- read_csv(path_security_gapfill, show_col_types = FALSE) %>%
    select(
      rgn_id, year,
      security_status = current_status,
      security_gapfilled = gapfilled,
      security_gapfill_reason = gapfill_reason
    )
} else {
  message("cs_security_gapfill.csv not found; gapfilling from cs_security.csv")
  security_raw <- read_csv(path_security, show_col_types = FALSE) %>%
    transmute(rgn_id, year = year_march, security_status = current_status)

  cce_years_preview <- sort(unique(c(ama$year, cult$year, security_raw$year)))

  security <- gapfill_security(security_raw, years = cce_years_preview) %>%
    select(
      rgn_id, year, security_status,
      security_gapfilled = gapfilled,
      security_gapfill_reason = gapfill_reason
    )
}

ama
cult
security
```

# Average the three subcomponents

```{r}
cce_years <- sort(unique(c(ama$year, cult$year, security$year)))
cce_regions <- region_levels

cce_with_security <- crossing(
  rgn_id = cce_regions,
  year = cce_years
) %>%
  left_join(ama, by = c("rgn_id", "year")) %>%
  left_join(cult, by = c("rgn_id", "year")) %>%
  left_join(security, by = c("rgn_id", "year")) %>%
  arrange(rgn_id, year) %>%
  mutate(
    current_status = rowMeans(
      cbind(ama_status, cult_status, security_status),
      na.rm = TRUE
    ),
    current_status = if_else(is.nan(current_status), NA_real_, current_status)
  ) %>%
  select(
    rgn_id, year, current_status,
    ama_status, cult_status, security_status,
    security_gapfilled, security_gapfill_reason, reason
  )

cce_with_security

write_csv(
  cce_with_security,
  file.path(path_cce_out, "cce_current_status_with_security_gapfill.csv")
)
```

# Plot possible CCE current status (with gapfilled security)

```{r}
cce_plot_df <- cce_with_security %>%
  mutate(
    region_label = factor(
      unname(rgn_labels[rgn_id]),
      levels = unname(rgn_labels[region_levels])
    ),
    security_gapfilled = replace_na(security_gapfilled, FALSE)
  ) %>% 
  filter(year %in% c(2019:2025))

p_cce <- ggplot(
  cce_plot_df,
  aes(x = year, y = current_status, color = region_label, group = rgn_id)
) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "black", linewidth = 0.6) +
  geom_line(linewidth = 0.9) +
  geom_point() +
  #geom_point(aes(shape = security_gapfilled), size = 2.2) +
  scale_shape_manual(
    values = c("FALSE" = 16, "TRUE" = 1),
    #labels = c("FALSE" = "Security observed", "TRUE" = "Security gapfilled"),
    name = "Security"
  ) +
  scale_color_manual(values = rgn_label_colors, drop = FALSE) +
  scale_y_continuous(limits = c(0, 100)) +
  scale_x_continuous(breaks = sort(unique(cce_plot_df$year))) +
  labs(
    title = "CURRENT STATUS: CCE with Security (GAPFILL)",
    subtitle = "Mean of Ama tu Playa, cultural places, and security.",
    x = "Year",
    y = "Current status (0-100)",
    color = "Region"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

p_cce

ggsave(
  filename = file.path(path_viz, "cce_current_status_with_security_gapfill.pdf"),
  plot = p_cce,
  width = 6,
  height = 6,
  device = cairo_pdf
)
```
