---
title: 'OHI GoC 2026: CCE with Security subcomponent'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    theme: cerulean
    highlight-style: haddock
editor_options:
  chunk_output_type: console
---

# Coastal Community Engagement (CCE) with Security

This script rebuilds CCE as the **arithmetic mean of three subcomponents**:

1. **Ama tu Playa / Adopta una Playa** — beach adoption current status  
2. **Non-natural cultural places** — employees per coastal population, 75th quantile score  
3. **Security** — ENSU percent who feel safe in their city (`100 − pct_unsafe`)

NAs are ignored in the mean (same pattern as the existing two-layer CCE: e.g. Region 2 has no beaches in the program inventory).

```{r}
library(tidyverse)

# ---------------------------------------------------------------------------
# Region labels and colors
# ---------------------------------------------------------------------------
rgn_labels <- c(
  "Region_1" = "Western Upper Gulf",
  "Region_2" = "Western Midriff Islands",
  "Region_3" = "Western Central Gulf",
  "Region_4" = "Western Southern Gulf",
  "Region_5" = "Eastern Upper Gulf",
  "Region_6" = "Eastern Midriff Islands",
  "Region_7" = "Southern Sonora Coast",
  "Region_8" = "Sinaloa Coast",
  "Region_9" = "Southeastern Gulf"
)

rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

region_levels <- paste0("Region_", 1:9)
rgn_label_colors <- setNames(
  unname(rgn_colors[region_levels]),
  unname(rgn_labels[region_levels])
)

# ---------------------------------------------------------------------------
# Paths (Mazu)
# ---------------------------------------------------------------------------
path_cce_out <- "/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026/output"
path_viz     <- "/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026/viz"
path_security <- "/home/shares/ohi/OHI_GOC/goal_prep/ss/security_output/cs_security.csv"

dir.create(path_viz, recursive = TRUE, showWarnings = FALSE)
dir.create(path_cce_out, recursive = TRUE, showWarnings = FALSE)
```

# Read the three subcomponents

```{r}
# 1) Ama tu Playa
ama <- read_csv(
  file.path(path_cce_out, "current_status_ama_tu_playa_progress_072326.csv"),
  show_col_types = FALSE
) %>%
  select(
    rgn_id,
    year,
    ama_status = current_status,
    any_of("reason")
  )

# support older single-year files
if (!"year" %in% names(ama)) {
  ama <- ama %>% mutate(year = 2025L)
}

if (!"reason" %in% names(ama)) {
  ama <- ama %>% mutate(reason = NA_character_)
}

# keep explicit NA for Region 2 (no beaches in inventory)
if (!"Region_2" %in% ama$rgn_id) {
  ama <- bind_rows(
    ama,
    crossing(rgn_id = "Region_2", year = 2019:2025) %>%
      mutate(
        ama_status = NA_real_,
        reason = "no beaches in program inventory"
      )
  )
} else {
  ama <- ama %>%
    mutate(
      reason = if_else(
        rgn_id == "Region_2" & is.na(ama_status),
        "no beaches in program inventory",
        reason
      )
    )
}

ama <- ama %>% arrange(rgn_id, year)

# 2) Non-natural cultural places (75th quantile score)
cult <- read_csv(
  file.path(
    path_cce_out,
    "cultural_places_employees_per_pop_current_status_75th_quantile.csv"
  ),
  show_col_types = FALSE
) %>%
  select(rgn_id, year = data_year, cult_status = score)

# 3) Security (percent who feel safe)
security <- read_csv(path_security, show_col_types = FALSE) %>%
  select(
    rgn_id,
    year = year_march,
    security_status = current_status
  )

ama
cult
security
```

# Average the three subcomponents

```{r}
cce_years <- sort(unique(c(ama$year, cult$year, security$year)))
cce_regions <- sort(unique(c(ama$rgn_id, cult$rgn_id, security$rgn_id)))

cce_with_security <- crossing(
  rgn_id = cce_regions,
  year = cce_years
) %>%
  left_join(ama, by = c("rgn_id", "year")) %>%
  left_join(cult, by = c("rgn_id", "year")) %>%
  left_join(security, by = c("rgn_id", "year")) %>%
  arrange(rgn_id, year) %>%
  mutate(
    # equal-weight mean of available subcomponents
    current_status = rowMeans(
      cbind(ama_status, cult_status, security_status),
      na.rm = TRUE
    ),
    # rowMeans of all-NA rows returns NaN; treat as missing
    current_status = if_else(is.nan(current_status), NA_real_, current_status)
  ) %>%
  select(
    rgn_id, year, current_status,
    ama_status, cult_status, security_status, reason
  )

cce_with_security

write_csv(
  cce_with_security,
  file.path(path_cce_out, "cce_current_status_with_security.csv")
)
```

# Plot possible CCE current status (with security)

```{r}
cce_plot_df <- cce_with_security %>%
  mutate(
    region_label = factor(
      unname(rgn_labels[rgn_id]),
      levels = unname(rgn_labels[region_levels])
    )
  ) %>% 
  filter(year %in% c(2019:2025))

p_cce <- ggplot(
  cce_plot_df,
  aes(x = year, y = current_status, color = region_label, group = rgn_id)
) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "black", linewidth = 0.6) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  scale_color_manual(values = rgn_label_colors, drop = FALSE) +
  scale_y_continuous(limits = c(0, 100)) +
  scale_x_continuous(breaks = sort(unique(cce_plot_df$year))) +
  labs(
    title = "CURRENT STATUS: Coastal Community Engagement (CCE) with Security",
    subtitle = "Arithmetic mean of Ama tu Playa, non-natural cultural places, and security (NAs ignored)",
    x = "Year",
    y = "Current status (0–100)",
    color = "Region"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

p_cce

ggsave(
  filename = file.path(path_viz, "cce_current_status_with_security.pdf"),
  plot = p_cce,
  width = 10,
  height = 6,
  device = cairo_pdf
)
```
