---
title: 'OHI GoC 2026: Security current status plot (GAPFILL)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    theme: cerulean
    highlight-style: haddock
editor_options:
  chunk_output_type: console
---

# Security current status (with gapfill)

Same ENSU security current status plot as `security_current_status_plot.qmd`, but **gapfills NAs** before plotting:

1. **Temporal** — if a region has ENSU data in some years but not others (e.g. Region 7 / Ciudad Obregón), fill missing years with that **region’s mean across available March years** (for single-city regions this equals the city mean).
2. **Spatial** — if a region has **no** ENSU city, fill from **GoC-only** donor regions in the same state (border cities like Nogales excluded):
   - Regions 2 & 3 ← Region 4 (La Paz / Los Cabos, BCS)
   - Region 5 ← mean of Regions 6 & 7 (Hermosillo & Ciudad Obregón, Sonora)

Gapfilled rows are flagged (`gapfilled`, `gapfill_reason`) and drawn with open points on the plot.

```{r}
library(tidyverse)

rgn_labels <- c(
  "Region_1" = "Western Upper Gulf",
  "Region_2" = "Western Midriff Islands",
  "Region_3" = "Western Central Gulf",
  "Region_4" = "Western Southern Gulf",
  "Region_5" = "Eastern Upper Gulf",
  "Region_6" = "Eastern Midriff Islands",
  "Region_7" = "Southern Sonora Coast",
  "Region_8" = "Sinaloa Coast",
  "Region_9" = "Southeastern Gulf"
)

rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

region_levels <- paste0("Region_", 1:9)
rgn_label_colors <- setNames(
  unname(rgn_colors[region_levels]),
  unname(rgn_labels[region_levels])
)

path_security <- "/home/shares/ohi/OHI_GOC/goal_prep/ss/security_output/cs_security.csv"
path_security_out <- "/home/shares/ohi/OHI_GOC/goal_prep/ss/security_output"
path_viz <- "/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026/viz"

dir.create(path_viz, recursive = TRUE, showWarnings = FALSE)
dir.create(path_security_out, recursive = TRUE, showWarnings = FALSE)
```

# Gapfill helper

```{r}
# Spatial donors: GoC ENSU cities only (no Nogales / border)
spatial_donors <- list(
  "Region_2" = c("Region_4"),           # BCS → La Paz / Los Cabos
  "Region_3" = c("Region_4"),
  "Region_5" = c("Region_6", "Region_7") # Sonora → Hermosillo / Cd. Obregón
)

gapfill_security <- function(security_raw, years, regions = region_levels) {
  # security_raw: tibble with rgn_id, year, security_status

  out <- tidyr::crossing(rgn_id = regions, year = years) %>%
    dplyr::left_join(security_raw, by = c("rgn_id", "year")) %>%
    dplyr::mutate(
      security_observed = security_status,
      gapfilled = FALSE,
      gapfill_reason = NA_character_
    )

  # --- 1) Temporal: same-region mean across available years ---
  rgn_means <- out %>%
    dplyr::filter(!is.na(security_status)) %>%
    dplyr::group_by(rgn_id) %>%
    dplyr::summarize(rgn_mean = mean(security_status), .groups = "drop")

  regions_with_data <- rgn_means$rgn_id

  out <- out %>%
    dplyr::left_join(rgn_means, by = "rgn_id") %>%
    dplyr::mutate(
      needs_temporal = is.na(security_status) & rgn_id %in% regions_with_data,
      security_status = dplyr::if_else(needs_temporal, rgn_mean, security_status),
      gapfilled = dplyr::if_else(needs_temporal, TRUE, gapfilled),
      gapfill_reason = dplyr::if_else(
        needs_temporal,
        "temporal: mean of same region across available ENSU years",
        gapfill_reason
      )
    ) %>%
    dplyr::select(-rgn_mean, -needs_temporal)

  # --- 2) Spatial: GoC donor regions (excludes border cities) ---
  for (target in names(spatial_donors)) {
    donors <- spatial_donors[[target]]

    for (yr in years) {
      donor_vals <- out %>%
        dplyr::filter(rgn_id %in% donors, year == yr, !is.na(security_status)) %>%
        dplyr::pull(security_status)

      # if that year is missing for all donors, fall back to donor all-year mean
      if (length(donor_vals) == 0) {
        donor_vals <- out %>%
          dplyr::filter(rgn_id %in% donors, !is.na(security_status)) %>%
          dplyr::pull(security_status)
      }
      if (length(donor_vals) == 0) next

      fill_val <- mean(donor_vals)
      idx <- which(out$rgn_id == target & out$year == yr & is.na(out$security_status))
      if (length(idx) == 0) next

      out$security_status[idx] <- fill_val
      out$gapfilled[idx] <- TRUE
      out$gapfill_reason[idx] <- paste0(
        "spatial: GoC donor mean from ",
        paste(donors, collapse = ", "),
        " (excludes border cities)"
      )
    }
  }

  out
}
```

# Read, gapfill, and save

```{r}
cs_security_raw <- read_csv(path_security, show_col_types = FALSE) %>%
  transmute(
    rgn_id,
    year = year_march,
    security_status = current_status
  )

security_years <- sort(unique(cs_security_raw$year))

cs_security <- gapfill_security(cs_security_raw, years = security_years) %>%
  mutate(
    rgn_id = factor(rgn_id, levels = region_levels),
    region_label = factor(
      unname(rgn_labels[as.character(rgn_id)]),
      levels = unname(rgn_labels[region_levels])
    ),
    current_status = security_status
  ) %>%
  arrange(rgn_id, year)

cs_security

write_csv(
  cs_security %>%
    select(
      rgn_id, year, current_status, security_observed,
      gapfilled, gapfill_reason
    ),
  file.path(path_security_out, "cs_security_gapfill.csv")
)

# quick check of what was filled
cs_security %>%
  filter(gapfilled) %>%
  count(rgn_id, gapfill_reason)
```

# Plot

```{r}
p_security <- ggplot(
  cs_security,
  aes(x = year, y = current_status, color = region_label, group = rgn_id)
) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "black", linewidth = 0.6) +
  geom_line(linewidth = 0.9) +
  geom_point() +
  # observed = solid; gapfilled = open
  #geom_point(aes(shape = gapfilled), size = 2.2) +
  #scale_shape_manual(
    #values = c("FALSE" = 16, "TRUE" = 1),
    #labels = c("FALSE" = "Observed", "TRUE" = "Gapfilled"),
    #name = "Value"
  #) +
  scale_color_manual(values = rgn_label_colors, drop = FALSE) +
  scale_y_continuous(limits = c(0, 100)) +
  scale_x_continuous(breaks = sort(unique(cs_security$year))) +
  labs(
    title = "CURRENT STATUS: Security (GAPFILL)",
    subtitle = "ENSU March; temporal = same-region year mean; spatial = GoC donors (no border cities). Dashed line = ideal (100).",
    x = "Year",
    y = "Current status (0-100)",
    color = "Region"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

p_security

ggsave(
  filename = file.path(path_viz, "security_current_status_gapfill.pdf"),
  plot = p_security,
  width = 6,
  height = 6,
  device = cairo_pdf
)
```
