---
title: 'OHI GoC 2026: Sense of Place (SP) with Security in CCE'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    theme: cerulean
    highlight-style: haddock
editor_options:
  chunk_output_type: console
---

# Sense of Place with Security incorporated into CCE

Sense of Place (SP) remains the equal-weight arithmetic mean of three subgoals:

1. **CCE** — Coastal Community Engagement (now including Security as a third subcomponent)  
2. **LSP** — Lasting Special Places  
3. **ICO** — Iconic Species  

This script expects `cce_current_status_with_security.csv` from `new_cce_integration.qmd` (or rebuilds that three-way CCE mean if needed).

```{r}
library(tidyverse)

# ---------------------------------------------------------------------------
# Region labels and colors
# ---------------------------------------------------------------------------
rgn_labels <- c(
  "Region_1" = "Western Upper Gulf",
  "Region_2" = "Western Midriff Islands",
  "Region_3" = "Western Central Gulf",
  "Region_4" = "Western Southern Gulf",
  "Region_5" = "Eastern Upper Gulf",
  "Region_6" = "Eastern Midriff Islands",
  "Region_7" = "Southern Sonora Coast",
  "Region_8" = "Sinaloa Coast",
  "Region_9" = "Southeastern Gulf"
)

rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

region_levels <- paste0("Region_", 1:9)
rgn_label_colors <- setNames(
  unname(rgn_colors[region_levels]),
  unname(rgn_labels[region_levels])
)

# ---------------------------------------------------------------------------
# Paths (Mazu)
# ---------------------------------------------------------------------------
path_cce_out  <- "/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026/output"
path_viz      <- "/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026/viz"
path_lsp      <- "/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/output/lsp_current_status.csv"
path_ico      <- "/home/shares/ohi/OHI_GOC/goal_prep/sp/ico/output/ico_mean_risk_status_per_region_041326.csv"
path_security <- "/home/shares/ohi/OHI_GOC/goal_prep/ss/security_output/cs_security.csv"
path_sp_out   <- "/home/shares/ohi/OHI_GOC/goal_prep/sp"

dir.create(path_viz, recursive = TRUE, showWarnings = FALSE)
dir.create(path_sp_out, recursive = TRUE, showWarnings = FALSE)
```

# CCE with Security

Prefer the integrated CCE table from `new_cce_integration.qmd`. If it is missing, rebuild the three-way mean here.

```{r}
path_cce_w_sec <- file.path(path_cce_out, "cce_current_status_with_security.csv")

if (file.exists(path_cce_w_sec)) {
  cce_with_security <- read_csv(path_cce_w_sec, show_col_types = FALSE)
} else {
  message("cce_current_status_with_security.csv not found; rebuilding from subcomponents.")

  ama <- read_csv(
    file.path(path_cce_out, "current_status_ama_tu_playa_progress_072326.csv"),
    show_col_types = FALSE
  ) %>%
    select(rgn_id, year, ama_status = current_status)

  if (!"year" %in% names(ama)) {
    ama <- ama %>% mutate(year = 2025L)
  }

  cult <- read_csv(
    file.path(
      path_cce_out,
      "cultural_places_employees_per_pop_current_status_75th_quantile.csv"
    ),
    show_col_types = FALSE
  ) %>%
    select(rgn_id, year = data_year, cult_status = score)

  security <- read_csv(path_security, show_col_types = FALSE) %>%
    select(rgn_id, year = year_march, security_status = current_status)

  cce_years <- sort(unique(c(ama$year, cult$year, security$year)))
  cce_regions <- sort(unique(c(ama$rgn_id, cult$rgn_id, security$rgn_id)))

  cce_with_security <- crossing(rgn_id = cce_regions, year = cce_years) %>%
    left_join(ama, by = c("rgn_id", "year")) %>%
    left_join(cult, by = c("rgn_id", "year")) %>%
    left_join(security, by = c("rgn_id", "year")) %>%
    mutate(
      current_status = rowMeans(
        cbind(ama_status, cult_status, security_status),
        na.rm = TRUE
      ),
      current_status = if_else(is.nan(current_status), NA_real_, current_status)
    ) %>%
    select(rgn_id, year, current_status, ama_status, cult_status, security_status)
}

cce_status <- cce_with_security %>%
  select(rgn_id, year, cce_status = current_status)

cce_status
```

# LSP and ICO

```{r}
lsp <- read_csv(path_lsp, show_col_types = FALSE) %>%
  select(rgn_id, lsp_status = current_status)

ico <- read_csv(path_ico, show_col_types = FALSE) %>%
  select(rgn_id, year, ico_status = current_status)

# expand LSP snapshot across CCE years (same approach as sp_current_status.qmd)
cce_years <- sort(unique(cce_status$year))

lsp_expanded <- crossing(
  rgn_id = unique(lsp$rgn_id),
  year = cce_years
) %>%
  left_join(lsp, by = "rgn_id")

lsp_expanded
ico
```

# SP = mean of CCE (with security), LSP, and ICO

```{r}
sp_with_safety <- cce_status %>%
  full_join(lsp_expanded, by = c("rgn_id", "year")) %>%
  full_join(ico, by = c("rgn_id", "year")) %>%
  filter(year %in% cce_years) %>%
  arrange(rgn_id, year) %>%
  mutate(
    current_status = rowMeans(
      cbind(cce_status, lsp_status, ico_status),
      na.rm = TRUE
    ),
    current_status = if_else(is.nan(current_status), NA_real_, current_status)
  ) %>%
  select(rgn_id, year, current_status, cce_status, lsp_status, ico_status)

sp_with_safety

write_csv(
  sp_with_safety %>% select(rgn_id, year, current_status),
  file.path(path_sp_out, "sp_current_status_with_security.csv")
)

# optional fuller table with subgoal columns
write_csv(
  sp_with_safety,
  file.path(path_sp_out, "sp_current_status_with_security_detail.csv")
)
```

# Plot overall SP current status (with safety in CCE)

```{r}
sp_plot_df <- sp_with_safety %>%
  mutate(
    region_label = factor(
      unname(rgn_labels[rgn_id]),
      levels = unname(rgn_labels[region_levels])
    )
  ) %>% 
  filter(year %in% c(2019:2025))

p_sp <- ggplot(
  sp_plot_df,
  aes(x = year, y = current_status, color = region_label, group = rgn_id)
) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "black", linewidth = 0.6) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  scale_color_manual(values = rgn_label_colors, drop = FALSE) +
  scale_y_continuous(limits = c(0, 100)) +
  scale_x_continuous(breaks = sort(unique(sp_plot_df$year))) +
  labs(
    title = "CURRENT STATUS: Sense of Place (SP) with Security in CCE",
    subtitle = "Arithmetic mean of CCE (Ama + cultural + security), LSP, and ICO",
    x = "Year",
    y = "Current status (0–100)",
    color = "Region"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

p_sp

ggsave(
  filename = file.path(path_viz, "sp_current_status_with_security.pdf"),
  plot = p_sp,
  width = 10,
  height = 6,
  device = cairo_pdf
)
```
