---
title: 'OHI GoC 2025: LE ggplot PDF exports'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
editor_options:
  chunk_output_type: console
---

# LE visualizations (ggplot → PDF)

Companion to `LE/script6_le_visualizations.qmd` and the EWG 3.1 LE website page. Rebuilds current-status line charts as **ggplot** objects and writes **vector PDFs** (best quality for Google Slides).

For metrics that use two reference options, each chart is saved twice: **Historical reference** and **2019 reference**. Job quality uses one shared Gulf-wide 90th-quantile reference.

**Extra raw plots:** Gulf-wide marine sector trends (jobs, job quality per job, revenue) matching the website section, saved as a three-panel figure plus separate panels and a sector color key.

```{r setup}
library(tidyverse)
library(ggplot2)
library(scales)
library(here)
library(priceR)

# Vector PDF via cairo when available (sharper text for slides)
pdf_device <- if (isTRUE(capabilities("cairo"))) cairo_pdf else pdf

out_dir <- here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/output_viz/ggplots_for_slides_071326")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
png_dir <- file.path(out_dir, "png")
dir.create(png_dir, recursive = TRUE, showWarnings = FALSE)

# Slide-friendly landscape (approx. 16:9 content area)
FIG_W <- 9
FIG_H <- 5
SECTOR_H <- 8
LEGEND_H <- 1
```

```{r helpers}
rgn_labels <- c(
  "Region_1" = "Western Upper Gulf",
  "Region_2" = "Western Midriff Islands",
  "Region_3" = "Western Central Gulf",
  "Region_4" = "Western Southern Gulf",
  "Region_5" = "Eastern Upper Gulf",
  "Region_6" = "Eastern Midriff Islands",
  "Region_7" = "Southern Sonora Coast",
  "Region_8" = "Sinaloa Coast",
  "Region_9" = "Southeastern Gulf"
)

rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

region_levels <- paste0("Region_", 1:9)
rgn_label_colors <- setNames(
  unname(rgn_colors[region_levels]),
  unname(rgn_labels[region_levels])
)
census_years <- c(2003, 2008, 2013, 2018, 2023)
component_levels <- c(
  "LE:LIV: Number of Jobs",
  "LE:LIV: Quality of Jobs",
  "LE:LIV",
  "LE:ECO",
  "Overall LE"
)

add_rgn_labels <- function(df) {
  df %>%
    mutate(
      rgn_id = factor(as.character(rgn_id), levels = region_levels),
      rgn_label = factor(
        rgn_labels[as.character(rgn_id)],
        levels = unname(rgn_labels[region_levels])
      )
    )
}

first_existing <- function(...) {
  paths <- c(...)
  hit <- paths[file.exists(paths)]
  if (length(hit) == 0) {
    stop("None of these paths exist:\n", paste(" -", paths, collapse = "\n"))
  }
  hit[[1]]
}

gg_theme_le <- function() {
  theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", size = 16, hjust = 0),
      plot.subtitle = element_text(size = 11, color = "grey30"),
      axis.title = element_text(size = 12),
      axis.text = element_text(size = 11),
      legend.title = element_text(face = "bold", size = 11),
      legend.text = element_text(size = 10),
      legend.position = "right",
      panel.grid.minor = element_blank(),
      plot.margin = margin(8, 12, 8, 8)
    )
}

le_gg_score <- function(df, y_var, title, y_lab = "Puntaje // Percent (out of 100)",
                        y_range = c(0, 110), add_goal = TRUE) {
  df <- df %>%
    mutate(year = as.numeric(year)) %>%
    arrange(rgn_label, year)

  y_col <- rlang::sym(y_var)

  p <- ggplot(df, aes(x = year, y = !!y_col, color = rgn_label, group = rgn_label)) +
    geom_line(linewidth = 0.9, lineend = "round") +
    geom_point(size = 2.2) +
    scale_color_manual(values = rgn_label_colors, name = "Region", drop = FALSE) +
    scale_x_continuous(breaks = census_years) +
    labs(title = title, x = "Año censal // Census Year", y = y_lab) +
    gg_theme_le()

  if (add_goal) {
    p <- p +
      geom_hline(
        yintercept = 100,
        linetype = "dashed",
        color = "black",
        linewidth = 0.55
      )
  }

  if (!is.null(y_range)) {
    p <- p + coord_cartesian(ylim = y_range)
  }

  p
}

le_gg_goc <- function(df, y_var, title, y_lab = "Puntaje // Percent (out of 100)",
                      y_range = c(0, 110)) {
  df <- df %>%
    mutate(year = as.numeric(year)) %>%
    arrange(year)

  y_col <- rlang::sym(y_var)

  p <- ggplot(df, aes(x = year, y = !!y_col)) +
    geom_line(linewidth = 1.05, color = "#333333", lineend = "round") +
    geom_point(size = 2.6, color = "#333333") +
    geom_hline(
      yintercept = 100,
      linetype = "dashed",
      color = "black",
      linewidth = 0.55
    ) +
    scale_x_continuous(breaks = census_years) +
    labs(title = title, x = "Año censal // Census Year", y = y_lab) +
    gg_theme_le() +
    theme(legend.position = "none")

  if (!is.null(y_range)) {
    p <- p + coord_cartesian(ylim = y_range)
  }

  p
}

# Raw / scaled regional line chart (no 100 goal line)
le_gg_raw <- function(df, y_var, title, y_lab, y_scale = "comma") {
  df <- df %>%
    mutate(year = as.numeric(year)) %>%
    arrange(rgn_label, year)

  y_col <- rlang::sym(y_var)

  p <- ggplot(df, aes(x = year, y = !!y_col, color = rgn_label, group = rgn_label)) +
    geom_line(linewidth = 0.9, lineend = "round") +
    geom_point(size = 2.2) +
    scale_color_manual(values = rgn_label_colors, name = "Region", drop = FALSE) +
    scale_x_continuous(breaks = census_years) +
    labs(title = title, x = "Año censal // Census Year", y = y_lab) +
    gg_theme_le()

  if (identical(y_scale, "comma")) {
    p <- p + scale_y_continuous(labels = label_comma())
  } else if (identical(y_scale, "sci")) {
    p <- p + scale_y_continuous(labels = label_number(accuracy = 0.0001))
  }

  p
}

save_le_pdf <- function(plot, filename, width = FIG_W, height = FIG_H) {
  path <- file.path(out_dir, filename)
  ggsave(
    filename = path,
    plot = plot,
    device = pdf_device,
    width = width,
    height = height,
    units = "in",
    dpi = 300,
    bg = "white"
  )
  message("Wrote: ", path)
  invisible
}

save_le_png <- function(plot, filename, width = FIG_W, height = FIG_H) {
  ggsave(
    filename = file.path(png_dir, filename),
    plot = plot,
    width = width,
    height = height,
    units = "in",
    dpi = 300,
    bg = "white"
  )
}

build_le_bundle <- function(jobs_path, rev_path, qual_raw, final_path) {
  cs_jobs_raw <- read_csv(jobs_path, show_col_types = FALSE) %>%
    rename(year = ce_ano_censal) %>%
    mutate(year = as.numeric(year))

  cs_rev_raw <- read_csv(rev_path, show_col_types = FALSE) %>%
    rename(year = ano_censal) %>%
    mutate(year = as.numeric(year))

  all_cs_indicators <- read_csv(final_path, show_col_types = FALSE) %>%
    rename(year = ano_censal, le_current_status = final_le_current_status_rgn) %>%
    mutate(year = as.numeric(year))

  cs_jobs_clean <- cs_jobs_raw %>%
    rename(score = current_status) %>%
    mutate(score = ifelse(rgn_id == "Region_2", NA_real_, score)) %>%
    dplyr::select(rgn_id, year, score)

  cs_eco_clean <- cs_rev_raw %>%
    rename(score = current_status) %>%
    mutate(score = ifelse(rgn_id == "Region_2", NA_real_, score)) %>%
    dplyr::select(rgn_id, year, score)

  cs_liv_clean <- full_join(
    cs_jobs_raw %>%
      dplyr::select(rgn_id, year, job_current_status = current_status),
    qual_raw %>%
      dplyr::select(rgn_id, year, qual_current_status = current_status)
  ) %>%
    mutate(
      score = (job_current_status + qual_current_status) / 2,
      score = ifelse(rgn_id == "Region_2", NA_real_, score)
    ) %>%
    dplyr::select(rgn_id, year, score)

  le_goc_avg <- all_cs_indicators %>%
    dplyr::select(year, le_goc_current_status = final_le_current_status_goc) %>%
    distinct()

  list(
    jobs = cs_jobs_clean,
    eco = cs_eco_clean,
    liv = cs_liv_clean,
    le = all_cs_indicators,
    le_goc = le_goc_avg
  )
}
```

## Read current-status outputs (both references)

```{r read-cs}
path_qual <- first_existing(
  "/home/shares/ohi/OHI_GOC/goal_prep/le/liv/ce_qual_current_status.csv"
)
path_jobs_hist <- first_existing(
  "/home/shares/ohi/OHI_GOC/goal_prep/le/liv/ce_jobs_current_status.csv"
)
path_jobs_2019 <- first_existing(
  "/home/shares/ohi/OHI_GOC/goal_prep/le/liv/ce_jobs_current_status_2019ref.csv"
)
path_rev_hist <- first_existing(
  "/home/shares/ohi/OHI_GOC/goal_prep/le/eco/ce_rev_current_status.csv"
)
path_rev_2019 <- first_existing(
  "/home/shares/ohi/OHI_GOC/goal_prep/le/eco/ce_rev_current_status_2019ref.csv"
)
path_final_hist <- first_existing(
  "/home/shares/ohi/OHI_GOC/goal_prep/le/final_le_current_status.csv"
)
path_final_2019 <- first_existing(
  "/home/shares/ohi/OHI_GOC/goal_prep/le/final_le_current_status_2019ref.csv"
)
path_coastal_adj <- first_existing(
  "/home/shares/ohi/OHI_GOC/goal_prep/le/coastal_adjustment_multiplier_2020.csv"
)
path_ce_cleaned <- first_existing(
  "/home/shares/ohi/OHI_GOC/goal_prep/le/ce_int/ce_cleaned_totals_per_yr_municipio_activity.csv"
)

cs_qual_raw <- read_csv(path_qual, show_col_types = FALSE) %>%
  rename(year = ano_censal) %>%
  mutate(year = as.numeric(year))

cs_qual_clean <- cs_qual_raw %>%
  rename(score = current_status) %>%
  mutate(score = ifelse(rgn_id == "Region_2", NA_real_, score)) %>%
  dplyr::select(rgn_id, year, score)

le_hist <- build_le_bundle(
  jobs_path = path_jobs_hist,
  rev_path = path_rev_hist,
  qual_raw = cs_qual_raw,
  final_path = path_final_hist
)

le_2019 <- build_le_bundle(
  jobs_path = path_jobs_2019,
  rev_path = path_rev_2019,
  qual_raw = cs_qual_raw,
  final_path = path_final_2019
)

coastal_adj <- read_csv(path_coastal_adj, show_col_types = FALSE)

coastal_sectors <- c(
  "coastal_recreation",
  "travel_planners",
  "accommodations_food_drink",
  "environmental_research_consulting"
)

# Scaled jobs (jobs per coastal resident) from jobs current-status file
jobs_scaled_df <- read_csv(path_jobs_hist, show_col_types = FALSE) %>%
  rename(year = ce_ano_censal) %>%
  mutate(
    year = as.numeric(year),
    marine_jobs_per_local_pop = as.numeric(marine_jobs_per_local_pop)
  ) %>%
  filter(!rgn_id %in% "Region_2") %>% 
  dplyr::select(rgn_id, year, marine_jobs_per_local_pop)

# Unscaled coastal-adjusted job counts by region (before dividing by population)
jobs_unscaled_df <- read_csv(path_ce_cleaned, show_col_types = FALSE) %>%
  filter(!(municipio == "MEXICALI" & ano_censal == "2023")) %>%
  group_by(ano_censal, rgn_id, specific_sector) %>%
  summarize(jobs_per_act = sum(h000a, na.rm = TRUE), .groups = "drop") %>%
  full_join(coastal_adj, by = "rgn_id") %>%
  mutate(
    coastal_pop_adjustment = ifelse(
      rgn_id == "Region_1" & ano_censal == "2023",
      1,
      coastal_pop_adjustment
    ),
    first_adj = ifelse(
      rgn_id %in% c("Region_3", "Region_4"),
      jobs_per_act * coastal_pop_adjustment,
      jobs_per_act
    ),
    coastal_jobs_per_act = ifelse(
      specific_sector %in% coastal_sectors,
      jobs_per_act * coastal_pop_adjustment,
      first_adj
    )
  ) %>%
  group_by(ano_censal, rgn_id) %>%
  summarize(total_jobs = round(sum(coastal_jobs_per_act, na.rm = TRUE)), .groups = "drop") %>%
  mutate(year = as.numeric(ano_censal)) %>%
  dplyr::select(rgn_id, year, total_jobs)
```

## Build ggplot score objects

```{r build-score-plots}
# Overall LE
gg_le_overall_hist <- le_gg_score(
  add_rgn_labels(le_hist$le),
  y_var = "le_current_status",
  title = "CURRENT STATUS: Livelihoods & Economies (Historical Reference)"
)
gg_le_overall_2019 <- le_gg_score(
  add_rgn_labels(le_2019$le),
  y_var = "le_current_status",
  title = "CURRENT STATUS: Livelihoods & Economies (2019 Reference)"
)

gg_le_rgn_hist <- le_gg_score(
  add_rgn_labels(le_hist$le),
  y_var = "le_current_status",
  title = "Overall LE current status by region (Historical Reference)"
)
gg_le_rgn_2019 <- le_gg_score(
  add_rgn_labels(le_2019$le),
  y_var = "le_current_status",
  title = "Overall LE current status by region (2019 Reference)"
)

gg_le_goc_hist <- le_gg_goc(
  le_hist$le_goc,
  y_var = "le_goc_current_status",
  title = "Overall LE current status for the Gulf of California (Historical Reference)"
)
gg_le_goc_2019 <- le_gg_goc(
  le_2019$le_goc,
  y_var = "le_goc_current_status",
  title = "Overall LE current status for the Gulf of California (2019 Reference)"
)

# LIV
gg_liv_hist <- le_gg_score(
  add_rgn_labels(le_hist$liv),
  y_var = "score",
  title = "LE:LIV (Number of Jobs & Quality of Jobs merged) (Historical Reference)"
)
gg_liv_2019 <- le_gg_score(
  add_rgn_labels(le_2019$liv),
  y_var = "score",
  title = "LE:LIV (Number of Jobs & Quality of Jobs merged) (2019 Reference)"
)

gg_jobs_hist <- le_gg_score(
  add_rgn_labels(le_hist$jobs),
  y_var = "score",
  title = "LE:LIV: Number of Jobs (Historical Reference)"
)
gg_jobs_2019 <- le_gg_score(
  add_rgn_labels(le_2019$jobs),
  y_var = "score",
  title = "LE:LIV: Number of Jobs (2019 Reference)"
)

gg_quality <- le_gg_score(
  add_rgn_labels(cs_qual_clean),
  y_var = "score",
  title = "LE:LIV: Quality of Jobs (same reference in both scenarios)"
)

# ECO
gg_eco_hist <- le_gg_score(
  add_rgn_labels(le_hist$eco),
  y_var = "score",
  title = "LE:ECO (Historical Reference)"
)
gg_eco_2019 <- le_gg_score(
  add_rgn_labels(le_2019$eco),
  y_var = "score",
  title = "LE:ECO (2019 Reference)"
)

# Faceted components (both refs)
make_components_long <- function(bundle, qual_df) {
  bind_rows(
    add_rgn_labels(bundle$jobs) %>%
      transmute(rgn_id, rgn_label, year, score, component = "LE:LIV: Number of Jobs"),
    add_rgn_labels(qual_df) %>%
      transmute(rgn_id, rgn_label, year, score, component = "LE:LIV: Quality of Jobs"),
    add_rgn_labels(bundle$liv) %>%
      transmute(rgn_id, rgn_label, year, score, component = "LE:LIV"),
    add_rgn_labels(bundle$eco) %>%
      transmute(rgn_id, rgn_label, year, score, component = "LE:ECO"),
    add_rgn_labels(bundle$le) %>%
      transmute(
        rgn_id, rgn_label, year,
        score = le_current_status,
        component = "Overall LE"
      )
  ) %>%
    filter(year %in% census_years) %>%
    mutate(component = factor(component, levels = component_levels))
}

le_components_hist <- make_components_long(le_hist, cs_qual_clean)
le_components_2019 <- make_components_long(le_2019, cs_qual_clean)

gg_components_hist <- ggplot(
  le_components_hist,
  aes(x = year, y = score, color = rgn_label, group = rgn_label)
) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.8) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey45", linewidth = 0.4) +
  facet_wrap(~ component, ncol = 2) +
  scale_color_manual(values = rgn_label_colors, name = "Region", drop = FALSE) +
  scale_x_continuous(breaks = census_years) +
  coord_cartesian(ylim = c(0, 110)) +
  labs(
    title = "LE current status by component (Historical Reference)",
    x = "Census Year",
    y = "Score (out of 100)"
  ) +
  gg_theme_le() +
  theme(strip.text = element_text(face = "bold", size = 11))

gg_components_2019 <- ggplot(
  le_components_2019,
  aes(x = year, y = score, color = rgn_label, group = rgn_label)
) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.8) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey45", linewidth = 0.4) +
  facet_wrap(~ component, ncol = 2) +
  scale_color_manual(values = rgn_label_colors, name = "Region", drop = FALSE) +
  scale_x_continuous(breaks = census_years) +
  coord_cartesian(ylim = c(0, 110)) +
  labs(
    title = "LE current status by component (2019 Reference)",
    x = "Census Year",
    y = "Score (out of 100)"
  ) +
  gg_theme_le() +
  theme(strip.text = element_text(face = "bold", size = 11))

# Raw job counts (coastal-adjusted, unscaled) and scaled jobs per coastal resident
gg_jobs_unscaled <- le_gg_raw(
  add_rgn_labels(jobs_unscaled_df),
  y_var = "total_jobs",
  title = "Marine-related jobs by region (coastal-adjusted, unscaled counts)",
  y_lab = "Number of jobs (H000A, coastal-adjusted)"
)

gg_jobs_scaled <- le_gg_raw(
  add_rgn_labels(jobs_scaled_df),
  y_var = "marine_jobs_per_local_pop",
  title = "Marine-related jobs by region (scaled to coastal population)",
  y_lab = "Jobs per coastal resident",
  y_scale = "sci"
)
```

## Gulf-wide marine sector trends (extra raw plots)

Matches the website **Gulf-wide marine sector trends** section: coastal-adjusted H000A jobs, (J300A + J400A) per job in constant 2023 MXN, and coastal-adjusted M000A revenue, aggregated across all OHI regions and colored by marine sector.

```{r sector-trends}
sector_labels <- c(
  "fishing" = "Fishing",
  "mariculture" = "Mariculture",
  "seafood_processing_trade" = "Seafood processing & trade",
  "port_dock_construction_management" = "Port & dock construction",
  "boat_building" = "Boat building",
  "tidal_energy" = "Tidal energy",
  "oil_gas_maritime_transport" = "Oil & gas maritime transport",
  "cargo_goods_maritime_transport" = "Cargo maritime transport",
  "marine_tourist_transport" = "Marine tourist transport",
  "coastal_recreation" = "Coastal recreation",
  "travel_planners" = "Travel planners",
  "accommodations_food_drink" = "Accommodations & food/drink",
  "environmental_research_consulting" = "Environmental research"
)

sector_levels <- unname(sector_labels)

# Custom palette for slides (keyed by display label)
sector_colors <- c(
  "Fishing"                        = "#0091FF",  # vivid bright blue (prominent)
  "Mariculture"                    = "#FF9E1B",  # bright orange
  "Seafood processing & trade"     = "#00C781",  # bright emerald green
  "Port & dock construction"       = "#B98B4E",  # warm light tan
  "Boat building"                  = "#FF4FA0",  # bright pink
  "Tidal energy"                   = "#5FD4E3",  # bright aqua
  "Oil & gas maritime transport"   = "#E8590C",  # bright burnt orange/vermillion
  "Cargo maritime transport"       = "#FFD500",  # bright gold/yellow
  "Marine tourist transport"       = "#7C6FF0",  # bright periwinkle
  "Coastal recreation"             = "#9ED60B",  # bright lime green
  "Travel planners"                = "#C061E8",  # bright orchid purple
  "Accommodations & food/drink"    = "#FF2255",  # vivid crimson (prominent)
  "Environmental research"         = "#A8A8A8"   # light neutral gray
)

sector_linetypes <- setNames(
  rep("solid", length(sector_levels)),
  sector_levels
)

quiet_inflation_factor_2023 <- function(year) {
  result <- NA_real_
  suppressMessages(
    suppressWarnings({
      capture.output(
        result <- as.numeric(
          adjust_for_inflation(
            price = 1,
            from_date = year,
            country = "Mexico",
            to_date = 2023
          )
        ),
        type = "output"
      )
    })
  )
  result
}

inflation_factors <- tibble(
  year = census_years,
  infl_factor = vapply(census_years, quiet_inflation_factor_2023, numeric(1))
)

apply_coastal_sector_metrics <- function(df) {
  df %>%
    full_join(coastal_adj, by = "rgn_id") %>%
    mutate(
      coastal_pop_adjustment = ifelse(
        rgn_id == "Region_1" & ano_censal == "2023",
        1,
        coastal_pop_adjustment
      ),
      coastal_jobs = ifelse(
        rgn_id %in% c("Region_3", "Region_4"),
        jobs * coastal_pop_adjustment,
        jobs
      ),
      coastal_jobs = ifelse(
        specific_sector %in% coastal_sectors,
        jobs * coastal_pop_adjustment,
        coastal_jobs
      ),
      coastal_benefits = ifelse(
        rgn_id %in% c("Region_3", "Region_4"),
        benefits_real_2023 * coastal_pop_adjustment,
        benefits_real_2023
      ),
      coastal_benefits = ifelse(
        specific_sector %in% coastal_sectors,
        benefits_real_2023 * coastal_pop_adjustment,
        coastal_benefits
      ),
      coastal_revenue = ifelse(
        rgn_id %in% c("Region_3", "Region_4"),
        revenue_real_2023 * coastal_pop_adjustment,
        revenue_real_2023
      ),
      coastal_revenue = ifelse(
        specific_sector %in% coastal_sectors,
        revenue_real_2023 * coastal_pop_adjustment,
        coastal_revenue
      )
    ) %>%
    dplyr::select(-coastal_pop_adjustment)
}

ce_sector_raw <- read_csv(path_ce_cleaned, show_col_types = FALSE) %>%
  filter(!(municipio == "MEXICALI" & ano_censal == "2023")) %>%
  mutate(
    year = as.numeric(ano_censal),
    benefits_mxn = (coalesce(j300a, 0) + coalesce(j400a, 0)) * 1e6,
    revenue_mxn = coalesce(m000a, 0) * 1e6,
    jobs = coalesce(h000a, 0)
  ) %>%
  left_join(inflation_factors, by = "year") %>%
  mutate(
    benefits_real_2023 = benefits_mxn * infl_factor,
    revenue_real_2023 = revenue_mxn * infl_factor
  )

ce_sector_coastal <- ce_sector_raw %>%
  apply_coastal_sector_metrics() %>%
  mutate(
    sector_label = factor(
      sector_labels[specific_sector],
      levels = sector_levels
    )
  ) %>%
  filter(!is.na(sector_label), year %in% census_years)

sector_goc <- ce_sector_coastal %>%
  group_by(year, specific_sector, sector_label) %>%
  summarize(
    total_jobs = sum(coastal_jobs, na.rm = TRUE),
    total_benefits = sum(coastal_benefits, na.rm = TRUE),
    total_revenue = sum(coastal_revenue, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    quality_per_job = ifelse(total_jobs > 0, total_benefits / total_jobs, NA_real_)
  )

sector_goc_long <- sector_goc %>%
  pivot_longer(
    c(total_jobs, quality_per_job, total_revenue),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(
    metric = factor(
      metric,
      levels = c("total_jobs", "quality_per_job", "total_revenue"),
      labels = c(
        "Total jobs (coastal-adjusted, formal and informal)",
        "Job quality (pesos of social benefits per job, 2023 MXN)",
        "Total revenue (2023 MXN)"
      )
    )
  )

gg_sector_theme <- function(show_legend = FALSE) {
  theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(size = 11, color = "grey30"),
      strip.text = element_text(face = "bold", size = 11),
      legend.position = if (show_legend) "right" else "none",
      panel.grid.minor = element_blank()
    )
}

add_sector_scales <- function(p) {
  p +
    scale_color_manual(
      values = sector_colors,
      name = "Sector",
      breaks = sector_levels,
      drop = FALSE
    ) +
    scale_linetype_manual(
      values = sector_linetypes,
      name = "Sector",
      breaks = sector_levels,
      drop = FALSE
    ) +
    guides(
      color = guide_legend(override.aes = list(linewidth = 1.6)),
      linetype = guide_legend(override.aes = list(linewidth = 1.6))
    )
}

gg_sector_goc_three_metrics <- ggplot(
  sector_goc_long,
  aes(
    x = year,
    y = value,
    color = sector_label,
    linetype = sector_label,
    group = sector_label
  )
) +
  geom_line(linewidth = 0.9, lineend = "round") +
  geom_point(size = 1.8) +
  facet_wrap(~ metric, ncol = 1, scales = "free_y") +
  scale_x_continuous(breaks = census_years) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title = "Gulf-wide marine sector trends by census year",
    subtitle = "Totals aggregated across all OHI regions; coastal adjustments applied.",
    x = "Census year",
    y = NULL
  ) +
  gg_sector_theme(show_legend = FALSE)

gg_sector_goc_three_metrics <- add_sector_scales(gg_sector_goc_three_metrics)

make_sector_panel <- function(df, y_var, title, y_lab) {
  y_col <- rlang::sym(y_var)
  p <- ggplot(
    df,
    aes(
      x = year,
      y = !!y_col,
      color = sector_label,
      linetype = sector_label,
      group = sector_label
    )
  ) +
    geom_line(linewidth = 0.95, lineend = "round") +
    geom_point(size = 2) +
    scale_x_continuous(breaks = census_years) +
    scale_y_continuous(labels = label_comma()) +
    labs(
      title = title,
      subtitle = "Gulf-wide totals; coastal adjustments applied.",
      x = "Census year",
      y = y_lab
    ) +
    gg_sector_theme(show_legend = TRUE)

  add_sector_scales(p)
}

gg_sector_jobs <- make_sector_panel(
  sector_goc,
  "total_jobs",
  "Gulf-wide marine sector trends: Total jobs",
  "Coastal-adjusted jobs (formal and informal)"
)

gg_sector_quality <- make_sector_panel(
  sector_goc,
  "quality_per_job",
  "Gulf-wide marine sector trends: Job quality",
  "contributed social benefits per job (2023 MXN)"
)

gg_sector_revenue <- make_sector_panel(
  sector_goc,
  "total_revenue",
  "Gulf-wide marine sector trends: Total revenue",
  "Coastal-adjusted revenue (2023 MXN)"
)

sector_legend_lines <- tibble(
  sector_label = factor(rep(sector_levels, each = 2), levels = sector_levels),
  x = rep(c(1, 2), times = length(sector_levels)),
  y = rep(seq_along(sector_levels), each = 2)
)

gg_sector_legend <- ggplot(
  sector_legend_lines,
  aes(
    x = x,
    y = y,
    color = sector_label,
    linetype = sector_label,
    group = sector_label
  )
) +
  geom_line(linewidth = 2, alpha = 0) +
  scale_color_manual(
    values = sector_colors,
    name = "Sector",
    breaks = sector_levels,
    drop = FALSE
  ) +
  scale_linetype_manual(
    values = sector_linetypes,
    name = "Sector",
    breaks = sector_levels,
    drop = FALSE
  ) +
  guides(
    color = guide_legend(
      nrow = 4,
      byrow = TRUE,
      override.aes = list(linewidth = 2.5, alpha = 1)
    ),
    linetype = guide_legend(
      nrow = 4,
      byrow = TRUE,
      override.aes = list(linewidth = 2.5, alpha = 1)
    )
  ) +
  coord_cartesian(xlim = c(0.5, 2.5), ylim = c(0, length(sector_levels) + 1)) +
  labs(title = "Sector color key (Gulf-wide marine sector trends)") +
  theme_void(base_size = 12) +
  theme(
    legend.position = "bottom",
    legend.box = "horizontal",
    legend.background = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank(),
    legend.key.width = unit(1.6, "cm"),
    legend.key.height = unit(0.5, "cm"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 11, face = "bold"),
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.background = element_rect(fill = "white", color = NA),
    plot.margin = margin(8, 8, 8, 8)
  )
```

## Preview

```{r preview, fig.width=9, fig.height=5}
gg_le_overall_hist
gg_le_goc_hist
gg_quality
gg_jobs_unscaled
gg_jobs_scaled
gg_sector_jobs
```

```{r preview-sector-panel, fig.width=9, fig.height=8}
gg_sector_goc_three_metrics
```

## Save all PDFs (+ PNG copies)

```{r save-pdfs}
score_plots <- list(
  "le_overall_historical.pdf" = gg_le_overall_hist,
  "le_overall_2019ref.pdf" = gg_le_overall_2019,
  "le_overall_by_region_historical.pdf" = gg_le_rgn_hist,
  "le_overall_by_region_2019ref.pdf" = gg_le_rgn_2019,
  "le_goc_average_historical.pdf" = gg_le_goc_hist,
  "le_goc_average_2019ref.pdf" = gg_le_goc_2019,
  "le_liv_merged_historical.pdf" = gg_liv_hist,
  "le_liv_merged_2019ref.pdf" = gg_liv_2019,
  "le_liv_jobs_historical.pdf" = gg_jobs_hist,
  "le_liv_jobs_2019ref.pdf" = gg_jobs_2019,
  "le_liv_quality.pdf" = gg_quality,
  "le_eco_historical.pdf" = gg_eco_hist,
  "le_eco_2019ref.pdf" = gg_eco_2019,
  "le_jobs_unscaled_coastal_adjusted.pdf" = gg_jobs_unscaled,
  "le_jobs_scaled_per_coastal_resident.pdf" = gg_jobs_scaled
)

facet_plots <- list(
  "le_components_historical.pdf" = gg_components_hist,
  "le_components_2019ref.pdf" = gg_components_2019
)

iwalk(score_plots, ~ {
  save_le_pdf(.x, .y)
  save_le_png(.x, sub("\\.pdf$", ".png", .y))
})

iwalk(facet_plots, ~ {
  save_le_pdf(.x, .y, width = FIG_W, height = 7.5)
  save_le_png(.x, sub("\\.pdf$", ".png", .y), width = FIG_W, height = 7.5)
})

# Three-panel sector figure (taller) + separate metric panels + color key
save_le_pdf(gg_sector_goc_three_metrics, "le_sector_goc_three_metrics.pdf",
            width = FIG_W, height = SECTOR_H)
save_le_png(gg_sector_goc_three_metrics, "le_sector_goc_three_metrics.png",
            width = FIG_W, height = SECTOR_H)

sector_single_plots <- list(
  "le_sector_goc_jobs.pdf" = gg_sector_jobs,
  "le_sector_goc_job_quality.pdf" = gg_sector_quality,
  "le_sector_goc_revenue.pdf" = gg_sector_revenue
)

iwalk(sector_single_plots, ~ {
  save_le_pdf(.x, .y)
  save_le_png(.x, sub("\\.pdf$", ".png", .y))
})

save_le_pdf(gg_sector_legend, "le_sector_color_key.pdf",
            width = FIG_W, height = LEGEND_H)
save_le_png(gg_sector_legend, "le_sector_color_key.png",
            width = FIG_W, height = LEGEND_H)

message("All LE figure PDFs written to: ", out_dir)
message("PNG copies (300 dpi) written to: ", png_dir)
```

## Notes for Google Slides

- Prefer **PDF** files (vector) when inserting into Slides so text stays sharp.
- Dual-reference pairs are labeled `_historical` and `_2019ref`.
- Jobs raw plots: `le_jobs_unscaled_coastal_adjusted.pdf` (counts) and `le_jobs_scaled_per_coastal_resident.pdf` (jobs per coastal resident).
- Sector trends: use `le_sector_goc_three_metrics.pdf` for the stacked figure, or the separate jobs / job quality / revenue PDFs; pair with `le_sector_color_key.pdf`.
- Region colors match `script6_le_visualizations.qmd` / the LE website page.

