---
title: 'OHI GoC 2025: LE Current Status Visualizations'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor:
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options:
  chunk_output_type: console
---

# LE visualizations


> **2019-reference variant:** Reads `_2019ref` outputs produced by the scripts in `cursor_generated_scripts/LE/`.

I ran this script after `script5_merge_liv_eco_2019ref.qmd` and any LE sustainability prep script. I read the saved current status outputs and built charts with consistent region colors and labels across the merged LIV subgoal, LIV number of jobs, LIV quality of jobs, ECO revenue, sustainability, and the overall LE score. I also compare LIV number of jobs with and without the ENOE informality multiplier (`ce_jobs_current_status_2019ref_informality.csv`).

```{r}
library(tidyverse)
library(ggplot2)
library(plotly)
library(sf)
library(mapview)
library(here)
library(rnaturalearth)
```

```{r}
rgn_labels <- c(
  "Region_1" = "Western Upper Gulf",
  "Region_2" = "Western Midriff Islands",
  "Region_3" = "Western Central Gulf",
  "Region_4" = "Western Southern Gulf",
  "Region_5" = "Eastern Upper Gulf",
  "Region_6" = "Eastern Midriff Islands",
  "Region_7" = "Southern Sonora Coast",
  "Region_8" = "Sinaloa Coast",
  "Region_9" = "Southeastern Gulf"
)

rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

region_levels <- paste0("Region_", 1:9)
rgn_label_colors <- setNames(unname(rgn_colors[region_levels]), unname(rgn_labels[region_levels]))
census_years <- c(2003, 2008, 2013, 2018, 2023)
component_levels <- c(
  "LE:LIV: Number of Jobs",
  "LE:LIV: Quality of Jobs",
  "LE:LIV",
  "LE:ECO",
  "Overall LE"
)

add_rgn_labels <- function(df) {
  df %>%
    mutate(
      rgn_id = factor(as.character(rgn_id), levels = region_levels),
      rgn_label = factor(
        rgn_labels[as.character(rgn_id)],
        levels = unname(rgn_labels[region_levels])
      )
    )
}

le_plotly_lines <- function(df, y_var, title, y_range = NULL) {
  yr <- range(df$year, na.rm = TRUE)

  plot_ly(
    df,
    x = ~year,
    y = as.formula(paste0("~", y_var)),
    color = ~rgn_label,
    colors = rgn_label_colors,
    type = "scatter",
    mode = "lines+markers"
  ) %>%
    layout(
      title = title,
      xaxis = list(title = "Año censal // Census Year"),
      yaxis = list(title = "Puntaje // Percent (out of 100)", range = y_range),
      legend = list(title = list(text = "Region"))
    ) %>%
    add_lines(
      x = yr,
      y = c(100, 100),
      line = list(color = "black", width = 2, dash = "dash"),
      name = "Meta // Goal (100/100)",
      inherit = FALSE
    )
}

le_plotly_goc <- function(df, y_var, title, y_range = NULL) {
  yr <- range(df$year, na.rm = TRUE)

  plot_ly(
    df,
    x = ~year,
    y = as.formula(paste0("~", y_var)),
    type = "scatter",
    mode = "lines+markers",
    name = "GOC Average"
  ) %>%
    layout(
      title = title,
      xaxis = list(title = "Año censal // Census Year"),
      yaxis = list(title = "Puntaje // Percent (out of 100)", range = y_range)
    ) %>%
    add_lines(
      x = yr,
      y = c(100, 100),
      line = list(color = "black", width = 2, dash = "dash"),
      name = "Meta // Goal (100/100)",
      inherit = FALSE
    )
}
```

## Read outputs

```{r}
cs_jobs_raw <- read_csv(
  here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/liv", "ce_jobs_current_status_2019ref.csv"),
  show_col_types = FALSE
) %>%
  rename(year = ce_ano_censal) %>%
  mutate(year = as.numeric(year))

cs_jobs_informality_raw <- read_csv(
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/le/liv",
    "ce_jobs_current_status_2019ref_informality.csv"
  ),
  show_col_types = FALSE
) %>%
  rename(year = ce_ano_censal) %>%
  mutate(year = as.numeric(year))

cs_qual_raw <- read_csv(
  here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/liv", "ce_qual_current_status.csv"),
  show_col_types = FALSE
) %>%
  rename(year = ano_censal) %>%
  mutate(year = as.numeric(year))

cs_rev_raw <- read_csv(
  here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/eco", "ce_rev_current_status_2019ref.csv"),
  show_col_types = FALSE
) %>%
  rename(year = ano_censal) %>%
  mutate(year = as.numeric(year))

all_cs_indicators <- read_csv(
  here::here("/home/shares/ohi/OHI_GOC/goal_prep/le", "final_le_current_status_2019ref.csv"),
  show_col_types = FALSE
) %>%
  rename(year = ano_censal, le_current_status = final_le_current_status_rgn) %>%
  mutate(year = as.numeric(year))

cs_jobs_clean <- cs_jobs_raw %>%
  rename(score = current_status) %>%
  dplyr::select(rgn_id, year, score)

cs_jobs_informality_clean <- cs_jobs_informality_raw %>%
  rename(score = current_status) %>%
  dplyr::select(rgn_id, year, score)

cs_qual_clean <- cs_qual_raw %>%
  rename(score = current_status) %>%
  dplyr::select(rgn_id, year, score)

cs_eco_clean <- cs_rev_raw %>%
  rename(score = current_status) %>%
  dplyr::select(rgn_id, year, score)

cs_liv_clean <- full_join(
  cs_jobs_raw %>%
    dplyr::select(rgn_id, year, job_current_status = current_status),
  cs_qual_raw %>%
    dplyr::select(rgn_id, year, qual_current_status = current_status)
) %>%
  mutate(
    score = (job_current_status + qual_current_status) / 2,
    score = ifelse(rgn_id == "Region_2", qual_current_status, score)
  ) %>%
  dplyr::select(rgn_id, year, score)

le_components_long <- bind_rows(
  add_rgn_labels(cs_jobs_clean) %>%
    transmute(rgn_id, rgn_label, year, score, component = "LE:LIV: Number of Jobs"),
  add_rgn_labels(cs_qual_clean) %>%
    transmute(rgn_id, rgn_label, year, score, component = "LE:LIV: Quality of Jobs"),
  add_rgn_labels(cs_liv_clean) %>%
    transmute(rgn_id, rgn_label, year, score, component = "LE:LIV"),
  add_rgn_labels(cs_eco_clean) %>%
    transmute(rgn_id, rgn_label, year, score, component = "LE:ECO"),
  add_rgn_labels(all_cs_indicators) %>%
    transmute(rgn_id, rgn_label, year, score = le_current_status, component = "Overall LE")
) %>%
  filter(year %in% census_years) %>%
  mutate(component = factor(component, levels = component_levels))

le_2023_long <- le_components_long %>%
  filter(year == 2023)

all_cs_indicators_2023 <- all_cs_indicators %>%
  filter(year == 2023)

le_goc_avg <- all_cs_indicators %>%
  dplyr::select(year, le_goc_current_status = final_le_current_status_goc) %>%
  distinct()
```

## Interactive line charts (plotly)

```{r}
liv_plotly <- le_plotly_lines(
  add_rgn_labels(cs_liv_clean),
  y_var = "score",
  title = "LE:LIV (Number of Jobs & Quality of Jobs merged)"
)

jobs_plotly <- le_plotly_lines(
  add_rgn_labels(cs_jobs_clean),
  y_var = "score",
  title = "LE:LIV: Number of Jobs"
)

quality_plotly <- le_plotly_lines(
  add_rgn_labels(cs_qual_clean),
  y_var = "score",
  title = "LE:LIV: Quality of Jobs"
)

eco_plotly <- le_plotly_lines(
  add_rgn_labels(cs_eco_clean),
  y_var = "score",
  title = "LE:ECO"
)

le_score_plotly <- le_plotly_lines(
  add_rgn_labels(all_cs_indicators),
  y_var = "le_current_status",
  title = "CURRENT STATUS: Livelihoods & Economies"
)

liv_plotly
jobs_plotly
quality_plotly
eco_plotly
le_score_plotly
```

## LIV number of jobs: CE vs CE x ENOE informality

Side-by-side Plotlys of the CE-only current status (`ce_jobs_current_status_2019ref.csv`) and the informality-adjusted series (`ce_jobs_current_status_2019ref_informality.csv`). Capped scores can hide differences when both series sit at 100, so the raw (uncapped) scores are shown as well.

```{r}
jobs_informality_plotly <- le_plotly_lines(
  add_rgn_labels(cs_jobs_informality_clean),
  y_var = "score",
  title = "LE:LIV: Number of Jobs (CE x ENOE informality)"
)

jobs_compare_capped <- subplot(
  jobs_plotly %>% layout(title = "CE only (capped at 100)", showlegend = FALSE),
  jobs_informality_plotly %>% layout(title = "CE x ENOE informality (capped at 100)"),
  nrows = 1,
  shareY = TRUE,
  titleX = TRUE,
  titleY = TRUE,
  margin = 0.06
) %>%
  layout(
    title = list(
      text = "LIV number of jobs current status: CE vs CE x ENOE informality",
      y = 0.98
    ),
    legend = list(title = list(text = "Region")),
    annotations = list(
      list(
        x = 0.22, y = 1.08, xref = "paper", yref = "paper",
        text = "CE only", showarrow = FALSE, font = list(size = 13)
      ),
      list(
        x = 0.80, y = 1.08, xref = "paper", yref = "paper",
        text = "CE x ENOE informality", showarrow = FALSE, font = list(size = 13)
      )
    )
  )

cs_jobs_raw_compare <- bind_rows(
  add_rgn_labels(cs_jobs_raw) %>%
    transmute(
      rgn_id, rgn_label, year,
      score = current_status_raw,
      version = "CE only"
    ),
  add_rgn_labels(cs_jobs_informality_raw) %>%
    transmute(
      rgn_id, rgn_label, year,
      score = current_status_raw,
      version = "CE x ENOE informality"
    )
)

jobs_raw_ce_plotly <- le_plotly_lines(
  add_rgn_labels(cs_jobs_raw) %>% rename(score = current_status_raw),
  y_var = "score",
  title = "CE only (raw, not capped)"
)

jobs_raw_informality_plotly <- le_plotly_lines(
  add_rgn_labels(cs_jobs_informality_raw) %>% rename(score = current_status_raw),
  y_var = "score",
  title = "CE x ENOE informality (raw, not capped)"
)

jobs_compare_raw <- subplot(
  jobs_raw_ce_plotly %>% layout(title = "CE only (raw, not capped)", showlegend = FALSE),
  jobs_raw_informality_plotly %>% layout(title = "CE x ENOE informality (raw, not capped)"),
  nrows = 1,
  shareY = TRUE,
  titleX = TRUE,
  titleY = TRUE,
  margin = 0.06
) %>%
  layout(
    title = list(
      text = "LIV number of jobs raw current status: CE vs CE x ENOE informality",
      y = 0.98
    ),
    legend = list(title = list(text = "Region")),
    annotations = list(
      list(
        x = 0.22, y = 1.08, xref = "paper", yref = "paper",
        text = "CE only", showarrow = FALSE, font = list(size = 13)
      ),
      list(
        x = 0.80, y = 1.08, xref = "paper", yref = "paper",
        text = "CE x ENOE informality", showarrow = FALSE, font = list(size = 13)
      )
    )
  )

jobs_raw_overlay_plotly <- plot_ly()

for (rgn in region_levels) {
  dat_ce <- cs_jobs_raw_compare %>%
    filter(rgn_id == rgn, version == "CE only")
  dat_inf <- cs_jobs_raw_compare %>%
    filter(rgn_id == rgn, version == "CE x ENOE informality")

  jobs_raw_overlay_plotly <- jobs_raw_overlay_plotly %>%
    add_trace(
      x = dat_ce$year,
      y = dat_ce$score,
      type = "scatter",
      mode = "lines+markers",
      name = unname(rgn_labels[rgn]),
      legendgroup = rgn,
      line = list(color = unname(rgn_colors[rgn]), dash = "solid", width = 2),
      marker = list(color = unname(rgn_colors[rgn])),
      showlegend = TRUE
    ) %>%
    add_trace(
      x = dat_inf$year,
      y = dat_inf$score,
      type = "scatter",
      mode = "lines+markers",
      name = unname(rgn_labels[rgn]),
      legendgroup = rgn,
      line = list(color = unname(rgn_colors[rgn]), dash = "dot", width = 2),
      marker = list(color = unname(rgn_colors[rgn]), symbol = "diamond"),
      showlegend = FALSE
    )
}

jobs_raw_overlay_plotly <- jobs_raw_overlay_plotly %>%
  add_trace(
    x = range(cs_jobs_raw_compare$year, na.rm = TRUE),
    y = c(100, 100),
    type = "scatter",
    mode = "lines",
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE,
    showlegend = TRUE
  ) %>%
  layout(
    title = "LIV number of jobs raw current status overlay (solid = CE only, dotted = CE x ENOE informality)",
    xaxis = list(title = "Año censal // Census Year"),
    yaxis = list(title = "Puntaje // Percent (uncapped)"),
    legend = list(title = list(text = "Region"))
  )

jobs_score_diff <- add_rgn_labels(cs_jobs_raw) %>%
  dplyr::select(rgn_id, rgn_label, year, score_ce = current_status) %>%
  left_join(
    add_rgn_labels(cs_jobs_informality_raw) %>%
      dplyr::select(rgn_id, year, score_inf = current_status),
    by = c("rgn_id", "year")
  ) %>%
  mutate(score_diff = score_inf - score_ce)

jobs_diff_plotly <- plot_ly(
  jobs_score_diff,
  x = ~year,
  y = ~score_diff,
  color = ~rgn_label,
  colors = rgn_label_colors,
  type = "scatter",
  mode = "lines+markers"
) %>%
  layout(
    title = "Difference in capped LIV jobs current status (informality-adjusted minus CE only)",
    xaxis = list(title = "Año censal // Census Year"),
    yaxis = list(title = "Score difference (percentage points)"),
    legend = list(title = list(text = "Region")),
    shapes = list(
      list(
        type = "line",
        x0 = min(jobs_score_diff$year, na.rm = TRUE),
        x1 = max(jobs_score_diff$year, na.rm = TRUE),
        y0 = 0,
        y1 = 0,
        line = list(color = "grey45", width = 1, dash = "dash")
      )
    )
  )

jobs_informality_plotly
jobs_compare_capped
jobs_compare_raw
jobs_raw_overlay_plotly
jobs_diff_plotly
```

## Overall LE line charts (plotly)

I plotted the overall LE current status by region and the Gulf wide average across census years.

```{r}
le_overall_rgn_plotly <- le_plotly_lines(
  add_rgn_labels(all_cs_indicators),
  y_var = "le_current_status",
  title = "Overall LE current status by region",
  y_range = c(0, 110)
)

le_overall_goc_plotly <- le_plotly_goc(
  le_goc_avg,
  y_var = "le_goc_current_status",
  title = "Overall LE current status for the Gulf of California",
  y_range = c(0, 110)
)

le_overall_rgn_plotly
le_overall_goc_plotly
```

## All components over time (ggplot)

```{r}
gg_le_components_lines <- ggplot(
  le_components_long,
  aes(x = year, y = score, color = rgn_label, group = rgn_label)
) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.8) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey45", linewidth = 0.4) +
  facet_wrap(~ component, ncol = 2) +
  scale_color_manual(values = rgn_label_colors, name = "Region") +
  scale_x_continuous(breaks = census_years) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 25)) +
  labs(
    title = "LE current status by component (2003 to 2023)",
    x = "Census Year",
    y = "Score (out of 100)"
  ) +
  theme_minimal(base_family = "Calibri") +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 11),
    legend.position = "right"
  )

gg_le_components_lines
```

## Sector trends (jobs, job quality, revenue)

These plots use the cleaned CE activity file from script 1, with the same coastal population adjustments applied in scripts 2 to 4. **Jobs** are coastal-adjusted H000A totals. **Job quality** is (J300A + J400A) in constant 2023 MXN per job. **Revenue** is coastal-adjusted M000A in constant 2023 MXN. Sector labels use `specific_sector`.

```{r}
library(priceR)
library(scales)

coastal_sectors <- c(
  "coastal_recreation",
  "travel_planners",
  "accommodations_food_drink",
  "environmental_research_consulting"
)

sector_labels <- c(
  "fishing" = "Fishing",
  "mariculture" = "Mariculture",
  "seafood_processing_trade" = "Seafood processing & trade",
  "port_dock_construction_management" = "Port & dock construction",
  "boat_building" = "Boat building",
  "tidal_energy" = "Tidal energy",
  "oil_gas_maritime_transport" = "Oil & gas maritime transport",
  "cargo_goods_maritime_transport" = "Cargo maritime transport",
  "marine_tourist_transport" = "Marine tourist transport",
  "coastal_recreation" = "Coastal recreation",
  "travel_planners" = "Travel planners",
  "accommodations_food_drink" = "Accommodations & food/drink",
  "environmental_research_consulting" = "Environmental research"
)

sector_levels <- unname(sector_labels)

apply_coastal_sector_metrics <- function(df) {
  df %>%
    full_join(coastal_adj, by = "rgn_id") %>%
    mutate(
      coastal_pop_adjustment = ifelse(
        rgn_id == "Region_1" & ano_censal == "2023",
        1,
        coastal_pop_adjustment
      ),
      coastal_jobs = ifelse(
        rgn_id %in% c("Region_3", "Region_4"),
        jobs * coastal_pop_adjustment,
        jobs
      ),
      coastal_jobs = ifelse(
        specific_sector %in% coastal_sectors,
        jobs * coastal_pop_adjustment,
        coastal_jobs
      ),
      coastal_benefits = ifelse(
        rgn_id %in% c("Region_3", "Region_4"),
        benefits_real_2023 * coastal_pop_adjustment,
        benefits_real_2023
      ),
      coastal_benefits = ifelse(
        specific_sector %in% coastal_sectors,
        benefits_real_2023 * coastal_pop_adjustment,
        coastal_benefits
      ),
      coastal_revenue = ifelse(
        rgn_id %in% c("Region_3", "Region_4"),
        revenue_real_2023 * coastal_pop_adjustment,
        revenue_real_2023
      ),
      coastal_revenue = ifelse(
        specific_sector %in% coastal_sectors,
        revenue_real_2023 * coastal_pop_adjustment,
        coastal_revenue
      )
    ) %>%
    dplyr::select(-coastal_pop_adjustment)
}

coastal_adj <- read_csv(
  here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/coastal_adjustment_multiplier_2020.csv"),
  show_col_types = FALSE
)

ce_sector_raw <- read_csv(
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/le/ce_int",
    "ce_cleaned_totals_per_yr_municipio_activity.csv"
  ),
  show_col_types = FALSE
) %>%
  filter(!(municipio == "MEXICALI" & ano_censal == "2023")) %>%
  mutate(
    year = as.numeric(ano_censal),
    benefits_mxn = (coalesce(j300a, 0) + coalesce(j400a, 0)) * 1e6,
    revenue_mxn = coalesce(m000a, 0) * 1e6,
    benefits_real_2023 = adjust_for_inflation(
      price = benefits_mxn,
      from_date = year,
      country = "Mexico",
      to_date = 2023
    ),
    revenue_real_2023 = adjust_for_inflation(
      price = revenue_mxn,
      from_date = year,
      country = "Mexico",
      to_date = 2023
    ),
    jobs = coalesce(h000a, 0)
  )

ce_sector_coastal <- ce_sector_raw %>%
  apply_coastal_sector_metrics() %>%
  mutate(
    sector_label = factor(
      sector_labels[specific_sector],
      levels = sector_levels
    )
  ) %>%
  filter(!is.na(sector_label), year %in% census_years)

sector_by_region <- ce_sector_coastal %>%
  group_by(year, rgn_id, specific_sector, sector_label) %>%
  summarize(
    total_jobs = sum(coastal_jobs, na.rm = TRUE),
    total_benefits = sum(coastal_benefits, na.rm = TRUE),
    total_revenue = sum(coastal_revenue, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    quality_per_job = ifelse(total_jobs > 0, total_benefits / total_jobs, NA_real_)
  ) %>%
  add_rgn_labels()

sector_goc <- sector_by_region %>%
  group_by(year, specific_sector, sector_label) %>%
  summarize(
    total_jobs = sum(total_jobs, na.rm = TRUE),
    total_benefits = sum(total_benefits, na.rm = TRUE),
    total_revenue = sum(total_revenue, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    quality_per_job = ifelse(total_jobs > 0, total_benefits / total_jobs, NA_real_)
  )

sector_goc_long <- sector_goc %>%
  pivot_longer(
    c(total_jobs, quality_per_job, total_revenue),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(
    metric = factor(
      metric,
      levels = c("total_jobs", "quality_per_job", "total_revenue"),
      labels = c(
        "Total jobs (coastal-adjusted H000A)",
        "Job quality (J300A + J400A per job, 2023 MXN)",
        "Total revenue (M000A, 2023 MXN)"
      )
    )
  )

sector_by_region_long <- sector_by_region %>%
  pivot_longer(
    c(total_jobs, quality_per_job, total_revenue),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(
    metric = factor(
      metric,
      levels = c("total_jobs", "quality_per_job", "total_revenue"),
      labels = c(
        "Total jobs (coastal-adjusted H000A)",
        "Job quality (J300A + J400A per job, 2023 MXN)",
        "Total revenue (M000A, 2023 MXN)"
      )
    )
  )
```

### Gulf-wide totals by sector and year

One figure with three panels: total jobs, benefits per job, and total revenue, each summed across all OHI regions.

```{r}
gg_sector_goc_three_metrics <- ggplot(
  sector_goc_long,
  aes(x = year, y = value, color = sector_label, group = sector_label)
) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.4) +
  facet_wrap(~ metric, ncol = 1, scales = "free_y") +
  scale_color_viridis_d(option = "turbo", name = "Sector") +
  scale_x_continuous(breaks = census_years) +
  labs(
    title = "Marine sector trends across the Gulf of California",
    subtitle = "Totals aggregated across all regions; coastal adjustments applied",
    x = "Census year",
    y = NULL
  ) +
  theme_minimal(base_family = "Calibri") +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 10),
    legend.position = "bottom",
    legend.text = element_text(size = 7),
    legend.key.width = unit(0.45, "cm")
  ) +
  guides(color = guide_legend(nrow = 3, byrow = TRUE))

gg_sector_goc_three_metrics
```

### Jobs by sector within each region

```{r}
gg_sector_region_jobs <- ggplot(
  sector_by_region_long %>% filter(metric == "Total jobs (coastal-adjusted H000A)"),
  aes(x = year, y = value, color = sector_label, group = sector_label)
) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.1) +
  facet_wrap(~ rgn_label, ncol = 3, scales = "free_y") +
  scale_color_viridis_d(option = "turbo", name = "Sector") +
  scale_x_continuous(breaks = census_years) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Marine jobs by sector and region",
    x = "Census year",
    y = "Total jobs"
  ) +
  theme_minimal(base_family = "Calibri") +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 8),
    axis.text.x = element_text(size = 7),
    legend.position = "none"
  )

gg_sector_region_jobs
```

### Job quality by sector within each region

```{r}
gg_sector_region_quality <- ggplot(
  sector_by_region_long %>%
    filter(metric == "Job quality (J300A + J400A per job, 2023 MXN)"),
  aes(x = year, y = value, color = sector_label, group = sector_label)
) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.1) +
  facet_wrap(~ rgn_label, ncol = 3, scales = "free_y") +
  scale_color_viridis_d(option = "turbo", name = "Sector") +
  scale_x_continuous(breaks = census_years) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Marine job quality by sector and region",
    subtitle = "Employer social security contributions plus other benefits per job (constant 2023 MXN)",
    x = "Census year",
    y = "Benefits per job"
  ) +
  theme_minimal(base_family = "Calibri") +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 8),
    axis.text.x = element_text(size = 7),
    legend.position = "none"
  )

gg_sector_region_quality
```

### Revenue by sector within each region

```{r}
gg_sector_region_revenue <- ggplot(
  sector_by_region_long %>%
    filter(metric == "Total revenue (M000A, 2023 MXN)"),
  aes(x = year, y = value, color = sector_label, group = sector_label)
) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.1) +
  facet_wrap(~ rgn_label, ncol = 3, scales = "free_y") +
  scale_color_viridis_d(option = "turbo", name = "Sector") +
  scale_x_continuous(breaks = census_years) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Marine revenue by sector and region",
    x = "Census year",
    y = "Total revenue (2023 MXN)"
  ) +
  theme_minimal(base_family = "Calibri") +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 8),
    axis.text.x = element_text(size = 7),
    legend.position = "none"
  )

gg_sector_region_revenue
```

### Shared sector legend (for regional facet plots)

```{r}
gg_sector_legend <- ggplot(
  sector_by_region_long %>% filter(metric == "Total jobs (coastal-adjusted H000A)"),
  aes(x = year, y = value, color = sector_label)
) +
  geom_line(linewidth = 0.8) +
  scale_color_viridis_d(option = "turbo", name = "Sector") +
  theme_void() +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 9),
    legend.title = element_text(size = 10, face = "bold")
  ) +
  guides(color = guide_legend(nrow = 4, byrow = TRUE))

gg_sector_legend
```

Region 2 sector values are shown for exploration but should be interpreted cautiously because municipal census data cannot reliably separate Gulf-side activity for Ensenada and San Quintín.

## 2023 snapshot by region (ggplot)

```{r}
gg_le_2023_bars <- ggplot(
  le_2023_long,
  aes(x = rgn_label, y = score, fill = rgn_label)
) +
  geom_col(width = 0.82, color = "white", linewidth = 0.3) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey45", linewidth = 0.4) +
  facet_wrap(~ component, ncol = 2) +
  scale_fill_manual(values = rgn_label_colors, guide = "none") +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 25)) +
  labs(
    title = "LE current status by region (2023)",
    x = NULL,
    y = "Score (out of 100)"
  ) +
  theme_minimal(base_family = "Calibri") +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 11),
    axis.text.x = element_text(angle = 35, hjust = 1, size = 8)
  )

gg_le_2023_bars
```

## Overall LE polar chart (2023)

```{r}
gg_le_2023_polar <- ggplot(
  le_2023_long %>% filter(component == "Overall LE"),
  aes(x = rgn_label, y = score, fill = rgn_label)
) +
  geom_col(width = 1, color = "white", linewidth = 0.35) +
  geom_hline(yintercept = 100, color = "grey55", linewidth = 0.5) +
  coord_polar(start = -pi / 2, clip = "off") +
  scale_fill_manual(values = rgn_label_colors, guide = "none") +
  scale_y_continuous(limits = c(0, 100), breaks = c(25, 50, 75, 100)) +
  labs(
    title = "Overall LE current status by region (2023)",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_family = "Calibri") +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    panel.grid.major.x = element_blank()
  )

gg_le_2023_polar
```

## Map visualization (2023 overall LE)

```{r}
ohi <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp"), quiet = TRUE) %>%
  filter(location == "marine") %>%
  dplyr::select(rgn_id) %>% 
  mutate(rgn_labels = case_when(
    rgn_id %in% "Region_1" ~ "Western Upper Gulf",
    rgn_id %in% "Region_2" ~ "Western Midriff Islands",
    rgn_id %in% "Region_3" ~ "Western Central Gulf",
    rgn_id %in% "Region_4" ~ "Western Southern Gulf",
    rgn_id %in% "Region_5" ~ "Eastern Upper Gulf",
    rgn_id %in% "Region_6" ~ "Eastern Midriff Islands",
    rgn_id %in% "Region_7" ~ "Southern Sonora Coast",
    rgn_id %in% "Region_8" ~ "Sinaloa Coast",
    rgn_id %in% "Region_9" ~ "Southeastern Gulf",
    TRUE ~ NA
  ))

all_cs_indicators_2023_ohigns <- full_join(all_cs_indicators_2023, ohi, by = "rgn_id") %>%
  st_as_sf()
```

```{r}
rgn_labels <- c(
  "Region_1" = "Western Upper Gulf",
  "Region_2" = "Western Midriff Islands",
  "Region_3" = "Western Central Gulf",
  "Region_4" = "Western Southern Gulf",
  "Region_5" = "Eastern Upper Gulf",
  "Region_6" = "Eastern Midriff Islands",
  "Region_7" = "Southern Sonora Coast",
  "Region_8" = "Sinaloa Coast",
  "Region_9" = "Southeastern Gulf"
)

rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

mapview(ohi, zcol = "rgn_labels", col.regions = rgn_colors, layer.name = "OHI GoC Regions")

```


```{r}
mapview(
  all_cs_indicators_2023_ohigns,
  zcol = "le_current_status",
  col.regions = colorRampPalette(c("#FEE090", "#ABD9E9", "#4575B4", "#313695")),
  layer.name = "Livelihoods & Economies Current Status Scores (2023)",
  legend = TRUE
)

le_at <- seq(0, 100, by = 10)

le_cols <- colorRampPalette(c(
  "#8a0032", "#A50026", "#A50026", "#D73027", "#F46D43",
  "#FDAE61", "#FEE090", "#FEE090", "#ABD9E9", "#4575B4", "#313695"
))(length(le_at) - 1)

mapview(
  all_cs_indicators_2023_ohigns,
  zcol = "le_current_status",
  col.regions = le_cols,
  at = le_at,
  layer.name = "Livelihoods & Economies Current Status Scores (2023)",
  legend = TRUE
)
```

## Choropleth map (ggplot)

I drew the Gulf states background first, then the marine regions on top.

```{r}
le_palette <- c(
  "#8a0032", "#A50026", "#A50026", "#D73027", "#F46D43",
  "#FDAE61", "#FEE090", "#FEE090", "#ABD9E9", "#4575B4", "#313695"
)

states <- st_read(here::here("spatial/ohi_regions/states.shp"), quiet = TRUE) %>%
  st_transform(st_crs(all_cs_indicators_2023_ohigns))

gg_le <- ggplot() +
  geom_sf(
    data = states,
    fill = "#d6cac0",
    color = "grey10",
    linewidth = 0.1
  ) +
  geom_sf(
    data = all_cs_indicators_2023_ohigns,
    aes(fill = le_current_status),
    color = "black",
    linewidth = 0.2
  ) +
  scale_fill_gradientn(
    colours = le_palette,
    limits = c(0, 100),
    na.value = "grey85",
    name = "Livelihoods & Economies\nCurrent Status 2023"
  ) +
  coord_sf(expand = FALSE) +
  theme_void() +
  theme(
    text = element_text(family = "Calibri"),
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    legend.position = "right",
    legend.key.height = unit(0.8, "cm"),
    legend.key.width = unit(0.5, "cm"),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    plot.margin = margin(10, 10, 10, 10)
  )

gg_le
```
