---
title: 'OHI GoC 2025: Livelihoods & Economies'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# Merge LE: LIV + ECO


> **2019-reference variant:** This copy scores number of jobs and revenue against each region's 2019 Economic Census value (2018 in the adjusted year series), not the 2003-2013 average. Outputs use `_2019ref` filenames.

Remember that within this Livelihoods & Economies (LE) goal, there are two subgoals: (1) Livelihoods (LIV), which is the quantity and quality of marine-related jobs, and (2) Economies (ECO), which is the revenue for all coastal regions from marine-related sectors.  

And subsequently, experts agreed on the ideal state for each of the subgoals, in which, for livelihoods, there are stable/more jobs of the same/higher quality in the future, and for economies, that revenue is maintained or increases in the future.

This script merges the current status found for the number of jobs, quality of jobs, and revenue of certain sectors, where both LIV components are first averaged, and then LIV and ECO current status scores are averaged to yield the final LE current status scores.  

*Outline of script:*

1. Read in the current status scores for LIV_number_jobs, LIV_quality_jobs, and ECO_revenue.
2. Bind both LIV current status scores into one dataframe, and average between the two to obtain the overall LIV current status scores.
3. Explore overall LIV current status scores.
4. Merge LIV with ECO and take the average of the current status scores.  This results in the overall LE current status scores across time and space. 
5. Save to Mazu.
6. Explore using `plotly`.

## Read in all current status of components

### LIV: number of jobs

```{r}
ce_jobs_current_status <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/liv","ce_jobs_current_status_2019ref.csv")) %>% 
  dplyr::select(-c(marine_jobs_per_local_pop, ref_value_2019)) %>% 
  rename(job_current_status_raw = current_status_raw, 
         job_current_status = current_status,
         ano_censal = ce_ano_censal)
```

### LIV: quality of jobs

```{r}
ce_qual_current_status <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/liv","ce_qual_current_status.csv")) %>% 
  dplyr::select(rgn_id, ano_censal, qual_current_status_raw = current_status_raw, qual_current_status = current_status) 
```


### ECO

```{r}
ce_rev_current_status <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/eco","ce_rev_current_status_2019ref.csv")) %>% 
  dplyr::select(rgn_id, ano_censal, rev_current_status_raw = current_status_raw, rev_current_status = current_status) 
```



## Merge LIV

```{r}
liv_merge <- full_join(ce_jobs_current_status, ce_qual_current_status) %>% 
  mutate(liv_current_status_raw = ((job_current_status_raw + qual_current_status_raw) / 2),
         liv_current_status = ((job_current_status + qual_current_status) / 2)) %>% 
  mutate(liv_current_status_raw = ifelse(rgn_id == "Region_2", qual_current_status_raw, liv_current_status_raw), # for Region_2 only use job quality becasue the reference point is not strictly temporal within region
         liv_current_status = ifelse(rgn_id == "Region_2", qual_current_status, liv_current_status)) %>% 
  dplyr::select(rgn_id, ano_censal, liv_current_status_raw, liv_current_status)
```

```{r}
library(plotly)
library(dplyr)

liv_merge <- liv_merge %>%
  mutate(ano_censal = as.numeric(ano_censal))

liv_merge_plotly <- plot_ly(
  liv_merge,
  x = ~ano_censal,
  y = ~liv_current_status_raw,
  color = ~factor(rgn_id),
  colors = rgn_colors,
  type = "scatter",
  mode = "lines"
) %>%
  layout(
    title = "RAW CURRENT STATUS of LIV (LIV_number_job & LIV_quality merged) per region",
    xaxis = list(title = "Census Year"),
    yaxis = list(
      title = "LE raw score (0-100 scale)"
    ),
    legend = list(title = list(text = "GoC Region"))
  ) %>%
  add_lines(
    x = range(liv_merge$ano_censal), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )

liv_merge_plotly



```


## Merge LIV and ECO

```{r}
le_current_status <- full_join(ce_rev_current_status, liv_merge) %>% 
  mutate(le_current_status_raw = ((liv_current_status_raw + rev_current_status_raw) / 2),
         le_current_status = ((liv_current_status + rev_current_status) / 2)) %>% 
  mutate(le_current_status_raw = ifelse(rgn_id == "Region_2", NA, le_current_status_raw), # for Region_2 NA because even quality does not have reliable data
         le_current_status = ifelse(rgn_id == "Region_2", NA, le_current_status)) %>% 
  mutate(le_current_status_raw_recapped = ifelse(le_current_status_raw >= 100, 100, le_current_status_raw)) %>% 
  dplyr::select(rgn_id, ano_censal, le_current_status_raw, le_current_status_raw_recapped, le_current_status)
```


### Plot

#### le_current_status_raw

```{r}
library(plotly)
library(dplyr)

le_current_status <- le_current_status %>%
  mutate(ano_censal = as.numeric(ano_censal))

le_raw_plotly <- plot_ly(
  le_current_status,
  x = ~ano_censal,
  y = ~le_current_status_raw,
  color = ~factor(rgn_id),
  colors = rgn_colors,
  type = "scatter",
  mode = "lines"
) %>%
  layout(
    title = "RAW CURRENT STATUS of LE (LIV & ECO merged) per region",
    xaxis = list(title = "Census Year"),
    yaxis = list(
      title = "LE raw score (0-100 scale)"
    ),
    legend = list(title = list(text = "GoC Region"))
  ) %>%
  add_lines(
    x = range(le_current_status$ano_censal), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )

le_raw_plotly



```

#### le_current_status_raw_recapped

```{r}
library(plotly)
library(dplyr)

le_current_status <- le_current_status %>%
  mutate(ano_censal = as.numeric(ano_censal))

le_raw_recapped_plotly <- plot_ly(
  le_current_status,
  x = ~ano_censal,
  y = ~le_current_status_raw_recapped,
  color = ~factor(rgn_id),
  colors = rgn_colors,
  type = "scatter",
  mode = "lines+markers"
) %>%
  layout(
    title = "RAW CURRENT STATUS of LE (LIV & ECO merged), recapped at 100",
    xaxis = list(title = "Census Year"),
    yaxis = list(
      title = "LE raw score (0-100, capped at 100)",
      range = c(0, 110)
    ),
    legend = list(title = list(text = "Region"))
  ) %>%
  add_lines(
    x = range(le_current_status$ano_censal),
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )

le_raw_recapped_plotly



```

#### le_current_status

```{r}
library(plotly)
library(dplyr)

le_current_status <- le_current_status %>%
  mutate(ano_censal = as.numeric(ano_censal))

le_cs_plotly <- plot_ly(
  le_current_status,
  x = ~ano_censal,
  y = ~le_current_status,
  color = ~factor(rgn_id),
  colors = rgn_colors,
  type = "scatter",
  mode = "lines+markers"
) %>%
  layout(
    title = "CURRENT STATUS of LE (LIV & ECO merged) per region",
    xaxis = list(title = "Census Year"),
    yaxis = list(
      title = "LE current status (0-100)",
      range = c(0, 110)
    ),
    legend = list(title = list(text = "Region"))
  ) %>%
  add_lines(
    x = range(le_current_status$ano_censal),
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )

le_cs_plotly



```


# Final LE Current Status Scores 

The most accurate one to use is le_current_status_raw_recapped, since we lose information in le_current_status by averaging between capped current status metrics.  Therefore, I will rename it:

```{r}
final_le_current_status <- le_current_status %>% 
  dplyr::select(rgn_id, ano_censal, final_le_current_status_rgn = le_current_status) %>% 
  group_by(ano_censal) %>% 
  mutate(final_le_current_status_goc = mean(final_le_current_status_rgn, na.rm = TRUE)) %>% 
  ungroup() # 88.19257 overall GoC LE score for 2023.
```

## Save to Mazu

```{r}
write_csv(final_le_current_status, here::here("/home/shares/ohi/OHI_GOC/goal_prep/le","final_le_current_status_2019ref.csv"))
```

## Plotlys

```{r}
final_le_current_status <- final_le_current_status %>%
  mutate(ano_censal = as.numeric(ano_censal))

x_range <- range(final_le_current_status$ano_censal)

le_cs_plotly <- plot_ly(
  final_le_current_status,
  x = ~ano_censal,
  y = ~final_le_current_status_rgn,
  color = ~factor(rgn_id),
  colors = rgn_colors,
  type = "scatter",
  mode = "lines"
) %>%
layout(
  title = "CURRENT STATUS of LE (LIV & ECO merged) per region",
  xaxis = list(title = "Census Year"),
  yaxis = list(
    title = "LE current status (0-100)",
    range = c(0, 110)
  ),
  legend = list(title = list(text = "Region"))
) %>%
add_lines(
  data = final_le_current_status,  
  x = x_range,                     
  y = c(100, 100),
  line = list(color = "black", width = 2, dash = "dash"),
  name = "Meta // Goal (100/100)",
  inherit = FALSE
)

le_cs_plotly
```

```{r}
final_le_current_status <- final_le_current_status %>%
  mutate(ano_censal = as.numeric(ano_censal))

x_range <- range(final_le_current_status$ano_censal)

le_cs_plotly <- plot_ly(
  final_le_current_status,
  x = ~ano_censal,
  y = ~final_le_current_status_goc,
  type = "scatter",
  mode = "lines", 
  name = "GOC Average"
) %>%
layout(
  title = "CURRENT STATUS of LE (LIV & ECO merged) for the GOC per year",
  xaxis = list(title = "Census Year"),
  yaxis = list(
    title = "LE current status (0-100)",
    range = c(0, 110)
  )
) %>%
add_lines(
  data = final_le_current_status, 
  x = x_range,                     
  y = c(100, 100),
  line = list(color = "black", width = 2, dash = "dash"),
  name = "Meta // Goal (100/100)",
  inherit = FALSE
)

le_cs_plotly
```
