---
title: 'OHI GoC 2025: Livelihoods & Economies (informality-adjusted jobs)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor:
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options:
  chunk_output_type: console
---

# Merge LE: LIV + ECO, using informality-adjusted jobs

> **2019-reference variant, informality-adjusted jobs:** Same merge as `script5_merge_liv_eco_2019ref.qmd`, except LIV number of jobs comes from `ce_jobs_current_status_2019ref_informality.csv`. Job quality and ECO revenue are unchanged. The original CE-only LE file is not overwritten.

LIV is the average of number of jobs and quality of jobs (Region 2 uses quality only). LE is the average of LIV and ECO. Region 2 LE is set to NA, matching script 5.

*Outline of script:*

1. Read informality-adjusted LIV jobs, LIV job quality, and ECO revenue current status.
2. Average the two LIV components.
3. Average LIV with ECO to get overall LE current status.
4. Save to Mazu under `_informality` filenames.
5. Explore using `plotly`.

```{r}
library(tidyverse)
library(here)
library(plotly)

rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)
```

## Read in all current status of components

```{r}
ce_jobs_current_status <- read_csv(
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/le/liv",
    "ce_jobs_current_status_2019ref_informality.csv"
  ),
  show_col_types = FALSE
) %>%
  dplyr::select(
    rgn_id,
    ano_censal = ce_ano_censal,
    job_current_status_raw = current_status_raw,
    job_current_status = current_status
  )

ce_qual_current_status <- read_csv(
  here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/liv", "ce_qual_current_status.csv"),
  show_col_types = FALSE
) %>%
  dplyr::select(
    rgn_id,
    ano_censal,
    qual_current_status_raw = current_status_raw,
    qual_current_status = current_status
  )

ce_rev_current_status <- read_csv(
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/le/eco",
    "ce_rev_current_status_2019ref.csv"
  ),
  show_col_types = FALSE
) %>%
  dplyr::select(
    rgn_id,
    ano_censal,
    rev_current_status_raw = current_status_raw,
    rev_current_status = current_status
  )
```

## Merge LIV

```{r}
liv_merge <- full_join(ce_jobs_current_status, ce_qual_current_status) %>%
  mutate(
    liv_current_status_raw = (job_current_status_raw + qual_current_status_raw) / 2,
    liv_current_status = (job_current_status + qual_current_status) / 2
  ) %>%
  mutate(
    # Region_2: use job quality only because the jobs reference is not strictly temporal within region
    liv_current_status_raw = ifelse(
      rgn_id == "Region_2",
      qual_current_status_raw,
      liv_current_status_raw
    ),
    liv_current_status = ifelse(
      rgn_id == "Region_2",
      qual_current_status,
      liv_current_status
    )
  ) %>%
  dplyr::select(rgn_id, ano_censal, liv_current_status_raw, liv_current_status) %>%
  mutate(ano_censal = as.numeric(ano_censal))
```

```{r}
liv_merge_plotly <- plot_ly(
  liv_merge,
  x = ~ano_censal,
  y = ~liv_current_status,
  color = ~factor(rgn_id),
  colors = rgn_colors,
  type = "scatter",
  mode = "lines+markers"
) %>%
  layout(
    title = "CURRENT STATUS of LIV (informality-adjusted jobs & quality merged) per region",
    xaxis = list(title = "Census Year"),
    yaxis = list(title = "LIV score (0-100)", range = c(0, 110)),
    legend = list(title = list(text = "GoC Region"))
  ) %>%
  add_lines(
    x = range(liv_merge$ano_censal, na.rm = TRUE),
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )

liv_merge_plotly
```

## Merge LIV and ECO

```{r}
le_current_status <- full_join(ce_rev_current_status, liv_merge) %>%
  mutate(
    ano_censal = as.numeric(ano_censal),
    le_current_status_raw = (liv_current_status_raw + rev_current_status_raw) / 2,
    le_current_status = (liv_current_status + rev_current_status) / 2
  ) %>%
  mutate(
    # Region_2 NA because even quality does not have reliable data
    le_current_status_raw = ifelse(rgn_id == "Region_2", NA, le_current_status_raw),
    le_current_status = ifelse(rgn_id == "Region_2", NA, le_current_status)
  ) %>%
  mutate(
    le_current_status_raw_recapped = ifelse(
      le_current_status_raw >= 100,
      100,
      le_current_status_raw
    )
  ) %>%
  dplyr::select(
    rgn_id,
    ano_censal,
    le_current_status_raw,
    le_current_status_raw_recapped,
    le_current_status
  )
```

```{r}
le_cs_plotly <- plot_ly(
  le_current_status,
  x = ~ano_censal,
  y = ~le_current_status,
  color = ~factor(rgn_id),
  colors = rgn_colors,
  type = "scatter",
  mode = "lines+markers"
) %>%
  layout(
    title = "CURRENT STATUS of LE (informality-adjusted LIV & ECO merged) per region",
    xaxis = list(title = "Census Year"),
    yaxis = list(title = "LE current status (0-100)", range = c(0, 110)),
    legend = list(title = list(text = "Region"))
  ) %>%
  add_lines(
    x = range(le_current_status$ano_censal, na.rm = TRUE),
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )

le_cs_plotly
```

# Final LE current status scores

This matches script 5: the saved regional score is `le_current_status` (average of the capped LIV and ECO components).

```{r}
final_le_current_status <- le_current_status %>%
  dplyr::select(
    rgn_id,
    ano_censal,
    final_le_current_status_rgn = le_current_status
  ) %>%
  group_by(ano_censal) %>%
  mutate(final_le_current_status_goc = mean(final_le_current_status_rgn, na.rm = TRUE)) %>%
  ungroup()

final_le_current_status
```

## Save to Mazu

```{r}
write_csv(
  liv_merge,
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/le/liv",
    "liv_current_status_2019ref_informality.csv"
  )
)

write_csv(
  final_le_current_status,
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/le",
    "final_le_current_status_2019ref_informality.csv"
  )
)
```

```{r}
x_range <- range(final_le_current_status$ano_censal, na.rm = TRUE)

le_goc_plotly <- plot_ly(
  final_le_current_status,
  x = ~ano_censal,
  y = ~final_le_current_status_goc,
  type = "scatter",
  mode = "lines+markers",
  name = "GOC Average"
) %>%
  layout(
    title = "CURRENT STATUS of LE (informality-adjusted LIV & ECO merged) for the GOC per year",
    xaxis = list(title = "Census Year"),
    yaxis = list(title = "LE current status (0-100)", range = c(0, 110))
  ) %>%
  add_lines(
    x = x_range,
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )

le_goc_plotly
```
