---
title: 'OHI GoC eutrophication update: agriculture nitrogen (script 2)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    theme: cerulean
    highlight-style: haddock
editor_options:
  chunk_output_type: console
---

# Script 2: agriculture nitrogen

Here we estimate nitrogen runoff from agriculture following Romero-Gil et al. (2026), then allocate municipal totals to GoC watersheds (`basin_id`).

For scoring later, we use a **10% leaching coefficient** (JARG Supplemental Table 6). JARG's main text used 20%; Table 6 also reports 5%, 10%, and 15%.

We take crop area and crop mix from **DGSIAP cierre agricola**. Fertilizer rates come from JARG's `Fertilizantes.csv` (INIFAP / Páez-Osuna). Where crops sit spatially (for allocating municipal N into watersheds) comes from processed CONABIO **USV cropland**.

---

## What JARG did

$$
N_{\mathrm{ag}} = \sum (\mathrm{area}_c \times \mathrm{fertilizer}_c \times L_c)
$$

- Area: SIAP irrigated sown area (`Sembrada`, modalidad riego)
- Fertilizer: kg N / ha for ~18 crops that cover most irrigated area
- Leaching: we use **$L_c = 0.10$**

JARG kept municipal totals (with a Gulf-drainage area filter). We go one step further and split each municipality's nitrogen across GoC watersheds using USV cropland area shares (see the worked example in section 3).

---

## Setup

```{r setup}
library(tidyverse)
library(here)
library(sf)
library(janitor)
library(mapview)
library(plotly)
sf_use_s2(FALSE)
```

```{r paths-constants}
pesticides_int <- here::here(
  "/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chemicals/pesticides/int"
)
dgsiap_agricola <- here::here(
  "/home/shares/ohi/OHI_GOC/_raw_data/DGSIAP_produccion_agricola/d2026"
)
jarg_repo <- here::here(
  "/home/shares/ohi/OHI_GOC/_raw_data/romero_gil_2026_estimated_nitrogen_discharge_goc/d2026/OceanHealth-Sanitation-main"
)
eut_int <- here::here(
  "/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/eutrophication/jarg_analysis_update/int"
)
spatial_regions <- here::here("/home/shares/ohi/OHI_GOC/spatial/OHI_regions")
spatial_admin <- here::here("/home/shares/ohi/OHI_GOC/spatial/land_admin_boundaries")

lc_cw <- 0.10
ag_years <- 2019:2025

goc_states <- c(
  "Baja California", "Baja California Sur", "Sonora",
  "Sinaloa", "Nayarit", "Jalisco"
)

# dir.create(eut_int, recursive = TRUE, showWarnings = FALSE)
```

---

## 1. Fertilizer rates (18 crops)

```{r fertilizer-rates}
fert <- read_csv(
  here::here(jarg_repo, "Fertilizantes.csv"),
  locale = locale(encoding = "latin1"),
  show_col_types = FALSE
) %>%
  clean_names() %>%
  rename(fert_n_kg_ha = fertilizante_n_kg_ha)

fert
```

---

## 2. DGSIAP irrigated sown area (2019-2025)

Files live at:

`/home/shares/ohi/OHI_GOC/_raw_data/DGSIAP_produccion_agricola/d2026/Cierre_agricola_mun_YYYY.csv`

We keep GoC states, **riego** (`idmodalidad == 1`), and crops that have a fertilizer rate in `Fertilizantes.csv` (same 18-crop focus as JARG).

```{r read-dgsiap-agricola}
read_ag_year <- function(year) {
  read_csv(
    here::here(dgsiap_agricola, paste0("Cierre_agricola_mun_", year, ".csv")),
    locale = locale(encoding = "latin1"),
    show_col_types = FALSE
  ) %>%
    clean_names() %>%
    mutate(
      across(where(is.character), ~ iconv(.x, from = "latin1", to = "UTF-8")),
      year = as.numeric(anio),
      idestado = as.numeric(idestado),
      idmunicipio = as.numeric(idmunicipio),
      idcultivo = as.numeric(idcultivo),
      sembrada = as.numeric(sembrada)
    )
}

cultivos_dgsiap <- map_dfr(ag_years, read_ag_year) %>%
  filter(
    nomestado %in% goc_states,
    idmodalidad == 1
  )

glimpse(cultivos_dgsiap)
unique(cultivos_dgsiap$nomestado)
unique(cultivos_dgsiap$anio) # 2019-2025
```

```{r join-fert-municipal-n}
# applied N, then 10% leaching, at the crop x municipality x year row
tn_ag_crop <- cultivos_dgsiap %>%
  inner_join(fert %>% dplyr::select(idcultivo, cultivo, fert_n_kg_ha), by = "idcultivo") %>%
  mutate(
    nin_kg = sembrada * fert_n_kg_ha,
    nag_kg = nin_kg * lc_cw,
    tn_agriculture_t_yr = nag_kg / 1000
  )

# municipal totals (before watershed allocation)
tn_ag_municipal <- tn_ag_crop %>%
  group_by(year, idestado, nomestado, idmunicipio, nommunicipio) %>%
  summarize(
    sembrada_ha = sum(sembrada, na.rm = TRUE),
    nin_kg = sum(nin_kg, na.rm = TRUE),
    nag_kg = sum(nag_kg, na.rm = TRUE),
    tn_agriculture_t_yr = sum(tn_agriculture_t_yr, na.rm = TRUE),
    .groups = "drop"
  )

total_goc_annual_tn <- tn_ag_municipal %>%
  group_by(year) %>%
  summarize(
    n_mun = n_distinct(paste(idestado, idmunicipio)),
    tn_agriculture_t_yr = sum(tn_agriculture_t_yr, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(year) # interesting, it has decreased in the last three years!!

total_goc_annual_tn
```

```{r plotly-tn-ag-municipal-lines}
tn_ag_municipal_plotly <- plot_ly(
  tn_ag_municipal %>% filter(year >= 2019, year <= 2025),
  x = ~year,
  y = ~tn_agriculture_t_yr,
  color = ~nommunicipio,
  type = "scatter",
  mode = "lines+markers",
  text = ~paste0(nommunicipio, ", ", nomestado),
  hovertemplate = paste(
    "%{text}<br>",
    "Year: %{x}<br>",
    "TN agriculture: %{y:.2f} t N / yr",
    "<extra></extra>"
  )
) %>%
  layout(
    title = "Agriculture nitrogen by municipality (2019-2025)",
    xaxis = list(title = "Year", dtick = 1),
    yaxis = list(title = "t N / year"),
    showlegend = TRUE,
    legend = list(title = list(text = "Municipality"))
  )

tn_ag_municipal_plotly
```

Quick coverage check (matched fert crops vs all riego GoC area) so we can see how much sown area sits outside the 18-crop fertilizer table:

```{r fert-coverage-qa}
coverage <- cultivos_dgsiap %>%
  group_by(year) %>%
  summarize(
    sembrada_all_ha = sum(sembrada, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(
    tn_ag_crop %>%
      group_by(year) %>%
      summarize(sembrada_fert_ha = sum(sembrada, na.rm = TRUE), .groups = "drop"),
    by = "year"
  ) %>%
  mutate(frac_area_with_fert = sembrada_fert_ha / sembrada_all_ha)

coverage
```

---

## 3. Allocate municipal N to watersheds (`basin_id`)

Municipal SIAP / DGSIAP totals are one number per municipality × year. Watersheds cut across those boundaries, so one municipality can send N into several `basin_id`s, and some of its land may not drain to the GoC at all.

**USV cropland area shares** means we use mapped cropland polygons (CONABIO USV) to decide how much of that municipal N belongs in each GoC watershed.

### Worked example

If Mexicali has **100 t N / year** from our DGSIAP × fertilizer × 10% calc, and USV shows cropland in two GoC watersheds inside Mexicali:

| basin_id | USV cropland in that mun ∩ basin | Share |
|----------|----------------------------------|-------|
| A | 800 ha | 80% |
| B | 200 ha | 20% |

Then:

- basin A gets \(100 \times 0.80 = 80\) t N / year
- basin B gets \(100 \times 0.20 = 20\) t N / year

We are **not** saying fertilizer was applied at the watershed scale. We are saying: SIAP / DGSIAP tells us how much N the municipality generated; USV tells us where the cropland sits, so we split that municipal total proportionally.

### Why we use USV, not raw municipality ∩ watershed land area

Raw polygon area would spread N over desert, cities, and non-crop land inside the municipality. USV cropland is a better proxy for where agriculture actually is, which matches the idea that fertilizer N comes from cropped land.

### Steps we take in code

1. Read GoC watersheds, municipality polygons, and USV agriculture.
2. Keep cropland classes only (same filter as pesticides).
3. Intersect USV ∩ municipality ∩ watershed; get `ag_area_ha`.
4. For each DGSIAP year, use the nearest USV survey year for weights (USV years are sparse; DGSIAP is annual).
5. `tn_basin = tn_mun * (ag_ha_mun_basin / sum_ag_ha_mun_in_goc_basins)`.

Municipalities with **no** USV cropland inside any GoC watershed contribute **0** at the watershed step (even if they still have municipal TN in `tn_ag_municipal`). That is intentional for "N reaching the GoC via cropland in GoC-draining basins," but it can drop some SIAP N that never gets a spatial home under this rule. We keep the municipal table for QA.

```{r read-spatial}
watershed_goc <- st_read(here::here(spatial_regions, "watersheds_goc.gpkg")) %>%
  st_make_valid()

municipio_sf <- st_read(
  here::here(spatial_admin, "mex_admin2_joined_w_municipio_codigos.shp")
) %>%
  clean_names() %>%
  mutate(
    idestado = as.numeric(clv_std),
    idmunicipio = as.numeric(clv_mnc)
  ) %>%
  filter(idestado %in% unique(tn_ag_municipal$idestado)) %>%
  select(idestado, idmunicipio, geometry) %>%
  st_make_valid()

usv_ag <- st_read(here::here(pesticides_int, "usv_agriculture_goc_allyears.gpkg")) %>%
  st_make_valid()

types_of_agriculture <- read_csv(
  here::here(pesticides_int, "types_of_agriculture_usv_agriculture_goc_allyears.csv"),
  show_col_types = FALSE
)

keep_ag <- types_of_agriculture %>%
  filter(is_cropland == "Y")

usv_ag_clean <- usv_ag %>%
  filter(descripcio %in% keep_ag$descripcio)
```

```{r mun-basin-usv-weights}
# project everything to watershed CRS (meters)
target_crs <- st_crs(watershed_goc)

usv_r <- st_transform(usv_ag_clean, target_crs)
mun_r <- st_transform(municipio_sf, target_crs)
ws_r <- st_transform(watershed_goc %>% select(basin_id), target_crs)

# USV x municipality
usv_mun <- st_intersection(
  usv_r %>% select(year, descripcio),
  mun_r
) %>%
  mutate(area_ha = as.numeric(st_area(.)) / 10000)

# then x watershed
usv_mun_ws <- st_intersection(
  usv_mun,
  ws_r
) %>%
  mutate(area_ha = as.numeric(st_area(.)) / 10000)

mun_basin_ag <- usv_mun_ws %>%
  st_drop_geometry() %>%
  group_by(usv_year = as.numeric(year), idestado, idmunicipio, basin_id) %>%
  summarize(ag_area_ha = sum(area_ha, na.rm = TRUE), .groups = "drop")

glimpse(mun_basin_ag)
```

```{r allocate-to-watershed}
usv_years <- sort(unique(mun_basin_ag$usv_year))
usv_years # [1] 1985 2001 2005 2009 2013 2016 2021

pick_usv_year <- function(y) {
  usv_years[which.min(abs(usv_years - y))]
}

# weights within each municipality (among GoC basins only)
mun_basin_w <- mun_basin_ag %>%
  group_by(usv_year, idestado, idmunicipio) %>%
  mutate(
    ag_area_mun_goc = sum(ag_area_ha, na.rm = TRUE),
    w_basin = if_else(ag_area_mun_goc > 0, ag_area_ha / ag_area_mun_goc, NA_real_)
  ) %>%
  ungroup()

tn_ag_by_watershed <- tn_ag_municipal %>%
  mutate(usv_year = map_dbl(year, pick_usv_year)) %>%
  left_join(
    mun_basin_w,
    by = c("usv_year", "idestado", "idmunicipio")
  ) %>%
  filter(!is.na(basin_id), !is.na(w_basin), w_basin > 0) %>%
  mutate(
    tn_agriculture_t_yr = tn_agriculture_t_yr * w_basin,
    nag_kg = nag_kg * w_basin,
    nin_kg = nin_kg * w_basin
  ) %>%
  group_by(basin_id, year, usv_year) %>%
  summarize(
    tn_agriculture_t_yr = sum(tn_agriculture_t_yr, na.rm = TRUE),
    nag_kg = sum(nag_kg, na.rm = TRUE),
    nin_kg = sum(nin_kg, na.rm = TRUE),
    n_municipalities = n_distinct(paste(idestado, idmunicipio)),
    .groups = "drop"
  ) %>%
  arrange(year, basin_id)

tn_ag_by_watershed

tn_ag_by_watershed %>%
  group_by(year) %>%
  summarize(
    n_basins = n_distinct(basin_id),
    tn_agriculture_t_yr = sum(tn_agriculture_t_yr, na.rm = TRUE),
    .groups = "drop"
  )
```

Optional QA: we can compare to the existing USV-only watershed area table from pesticides.

```{r compare-ag-area}
ag_area_by_watershed <- read_csv(
  here::here(pesticides_int, "ag_area_ha_by_watershed.csv"),
  show_col_types = FALSE
)

glimpse(ag_area_by_watershed)
```

---

## 4. Write watershed output

Primary deliverable for script 5 is watershed TN (10% leaching already applied). Pourpoints and OHI regions are handled in script 5.

```{r write-outputs}
write_csv(tn_ag_municipal, here::here(eut_int, "tn_agriculture_municipal.csv"))
write_csv(mun_basin_ag, here::here(eut_int, "usv_ag_area_ha_mun_basin.csv"))
write_csv(tn_ag_by_watershed, here::here(eut_int, "tn_agriculture_by_watershed.csv"))
```
