---
title: 'OHI GoC eutrophication update: current status (script 6)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    theme: cerulean
    highlight-style: haddock
editor_options:
  chunk_output_type: console
---

# Script 6: eutrophication current status (on hold)

This is where we will turn regional total nitrogen into a pressure and a current status score.

**We are not ready to lock a reference point yet.** Keep this script as a place for notes and options until we have that discussion.

Input will come from script 5:

`jarg_analysis_update/output/tn_total_by_region.csv`

---

## What we did last time (for context only)

The old `cw_eutrophication.qmd` workflow:

1. Used JARG Table 6 with Mean / 10% / 40% / 100%
2. Divided municipal TN by coastline km
3. Used the max municipal ton/km as the reference
4. Weighted municipalities into regions by coastline
5. Gapfilled Region 9 with Region 8

That pushed Region 1 (Mexicali) very low because of a short coastline, and it treated Region 9 as a copy of Region 8. We do not want to repeat that.

---

## What we want this update to support

- Ideal = 0 nitrogen
- Use pourpoint-based regional TN from script 5 (t N / year), not coastline-standardized municipal values
- No Region 8 → Region 9 copy-paste

$$
\mathrm{Status} = (1 - \mathrm{pressure}) \times 100
$$

---

## Reference point options to discuss

These are options only. Do not implement until we agree.

1. **Top-5 mean** of regional TN across all regions and years (same pattern as Chemicals)
2. **Single historical maximum** regional TN
3. A literature / policy threshold if we find one that actually applies to annual land-based loads to the GoC
4. Something else the EWG prefers

Open questions:

- Do we keep raw t N / year, or normalize by something other than coastline (shelf area, discharge volume, etc.)?
- Single year snapshot vs multi-year panel for the maximum?
- Cap pressure at 1 when a region exceeds the reference?

---

## Placeholder code (do not treat as final)

```{r setup}
library(tidyverse)
library(here)
```

```{r paths}
eut_out <- here::here(
  "/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/eutrophication/jarg_analysis_update/output"
)
```

```{r read-when-ready}
# tn <- read_csv(here::here(eut_out, "tn_total_by_region.csv"))
```

```{r score-placeholder}
# example only, not approved:
# max_ref <- tn %>%
#   slice_max(tn_total_t_yr, n = 5) %>%
#   summarize(maximum = mean(tn_total_t_yr)) %>%
#   pull(maximum)
#
# scored <- tn %>%
#   mutate(
#     pressure = pmin(1, tn_total_t_yr / max_ref),
#     status = (1 - pressure) * 100
#   )
```

```{r write-placeholder}
# later, once reference points are agreed:
# write_csv(scored %>% select(rgn_id, year, status),
#           here::here(eut_out, "nutrient_status.csv"))
# write_csv(scored %>% select(rgn_id, year, pressure),
#           here::here(eut_out, "nutrient_pressure.csv"))
```

---

## Status

| Piece | Status |
|-------|--------|
| Needs reference-point discussion | Yes |
| Implement scoring | Not yet |
| Depends on | Script 5 totals |
