---
title: 'OHI GoC eutrophication update: total nitrogen (script 5)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    theme: cerulean
    highlight-style: haddock
editor_options:
  chunk_output_type: console
---

# Script 5: total nitrogen (watersheds → pourpoints → regions)

Here we combine the four source watershed tables from scripts 1–4, gapfill missing years so the panel lines up, then use pourpoints to assign coastal watershed nitrogen output to OHI regions.

Source coefficients are already applied upstream:

| Source | Coefficient (scripts 1–4) |
|--------|---------------------------|
| Wastewater | Mean NRE pathway |
| Agriculture | 10% leaching |
| Livestock | 40% export |
| Shrimp | 100% export |

$$
TN = N_{\mathrm{ww}} + N_{\mathrm{ag}} + N_{\mathrm{liv}} + N_{\mathrm{shrimp}}
$$

Pourpoints here are only a delivery / region-assignment step (not a second attenuation step). Coastline standardization and scoring live in scripts 6–7 (t N / km coastline, Chemicals pattern).

**Data and methods overview:** gapfill with zoo the same way as Chemicals — linear interpolate (`na.approx`) between known years (e.g. livestock 2023; shrimp between USV survey years), then carry forward (`na.locf`) after the last known year (e.g. shrimp 2021 → 2022–2025). We keep only years where all four sources are present, through 2025.

---

## Inputs (from scripts 1-4)

Under:

`/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/eutrophication/jarg_analysis_update/int/`

| File | From |
|------|------|
| `tn_wastewater_by_watershed.csv` | script 1 |
| `tn_agriculture_by_watershed.csv` | script 2 |
| `tn_livestock_by_watershed.csv` | script 3 |
| `tn_shrimp_by_watershed.csv` | script 4 |

Each should have at least: `basin_id`, `year`, and the source TN column in **t N per year**.

---

## Outline

1. Read the four watershed tables (keep missing years as NA, not 0).
2. Choose the panel years (first shared start → 2025).
3. Gapfill each source by basin (keep observed years as anchors → `na.approx` between them → `na.locf` after the last year).
4. Join, sum to `tn_total_t_yr`, print coverage.
5. Pourpoints → regions.
6. Plotlys; write outputs.

---

## Setup

```{r setup}
library(tidyverse)
library(here)
library(sf)
library(janitor)
library(plotly)
library(zoo)
sf_use_s2(FALSE)
```

```{r paths}
eut_int <- here::here(
  "/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/eutrophication/jarg_analysis_update/int"
)
eut_out <- here::here(
  "/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/eutrophication/jarg_analysis_update/output"
)
spatial_regions <- here::here("/home/shares/ohi/OHI_GOC/spatial/OHI_regions")

year_end_panel <- 2025L

dir.create(eut_out, recursive = TRUE, showWarnings = FALSE)
```

---

## 1. Read watershed sources

```{r read-sources}
ww <- read_csv(
  here::here(eut_int, "tn_wastewater_by_watershed.csv"),
  show_col_types = FALSE
) %>%
  mutate(year = as.integer(year)) %>%
  select(basin_id, year, tn_wastewater_t_yr)

ag <- read_csv(
  here::here(eut_int, "tn_agriculture_by_watershed.csv"),
  show_col_types = FALSE
) %>%
  mutate(year = as.integer(year)) %>%
  select(basin_id, year, tn_agriculture_t_yr)

liv <- read_csv(
  here::here(eut_int, "tn_livestock_glw4_by_watershed.csv"),
  show_col_types = FALSE
) %>%
  mutate(year = as.integer(year)) %>%
  select(basin_id, year, tn_livestock_t_yr)

shr <- read_csv(
  here::here(eut_int, "tn_shrimp_by_watershed.csv"),
  show_col_types = FALSE
) %>%
  mutate(year = as.integer(year)) %>%
  select(basin_id, year, tn_shrimp_t_yr)

# observed year ranges (before gapfill)
source_years <- tibble(
  source = c("wastewater", "agriculture", "livestock", "shrimp"),
  years = list(
    sort(unique(ww$year)),
    sort(unique(ag$year)),
    sort(unique(liv$year)),
    sort(unique(shr$year))
  )
) %>%
  mutate(
    year_min = map_int(years, min),
    year_max = map_int(years, max),
    n_years = map_int(years, length),
    missing_2023 = map_lgl(years, ~ !(2023L %in% .x))
  ) %>%
  select(source, year_min, year_max, n_years, missing_2023)

source_years
```

---

## 2. Gapfill by basin, then keep the complete panel

```{r gapfill-sources}
# panel starts when every source has begun (usually agriculture is the limiter)
year_start_panel <- max(
  min(ww$year),
  min(ag$year),
  min(liv$year),
  min(shr$year)
)
years_panel <- year_start_panel:year_end_panel

cat("complete panel years:", paste(years_panel, collapse = ", "), "\n")

# gapfill one tn column within each basin.
# keep ALL observed years as anchors (e.g. shrimp 2016 + 2021), expand through year_end,
# then: na.approx between known years, na.locf after the last known year.
gapfill_basin <- function(df, value_col, years_out) {
  col <- rlang::as_string(rlang::ensym(value_col))

  df %>%
    group_by(basin_id) %>%
    group_modify(function(.x, .y) {
      y_lo <- min(c(.x$year, years_out), na.rm = TRUE)
      y_hi <- max(years_out)

      out <- .x %>%
        tidyr::complete(year = y_lo:y_hi) %>%
        arrange(year)

      v <- out[[col]]
      if (sum(is.finite(v)) >= 2L) {
        v <- zoo::na.approx(v, x = out$year, na.rm = FALSE)
      }
      # carry forward after last observation (shrimp 2021 → 2022:2025)
      v <- zoo::na.locf(v, na.rm = FALSE)
      out[[col]] <- v

      out %>%
        filter(year %in% years_out) %>%
        select(year, all_of(col))
    }) %>%
    ungroup()
}

ww_g <- gapfill_basin(ww, tn_wastewater_t_yr, years_panel)
ag_g <- gapfill_basin(ag, tn_agriculture_t_yr, years_panel)
liv_g <- gapfill_basin(liv, tn_livestock_t_yr, years_panel)
shr_g <- gapfill_basin(shr, tn_shrimp_t_yr, years_panel)

# sanity check: shrimp should be non-zero across the panel after fill, not only 2016/2021
cat("\nshrimp gulf totals AFTER gapfill (should be >0 each panel year):\n")
print(
  shr_g %>%
    group_by(year) %>%
    summarize(tn_shrimp_t_yr = sum(tn_shrimp_t_yr, na.rm = TRUE), .groups = "drop")
)

# join; basins with no history for a source get 0 (not invented from other basins)
tn_by_watershed <- ww_g %>%
  full_join(ag_g, by = c("basin_id", "year")) %>%
  full_join(liv_g, by = c("basin_id", "year")) %>%
  full_join(shr_g, by = c("basin_id", "year")) %>%
  mutate(
    across(starts_with("tn_"), ~ replace_na(.x, 0)),
    tn_total_t_yr = tn_wastewater_t_yr + tn_agriculture_t_yr +
      tn_livestock_t_yr + tn_shrimp_t_yr
  ) %>%
  arrange(year, desc(tn_total_t_yr))

tn_by_watershed

# gulf-wide coverage after gapfill (all four should be present every panel year)
tn_year_coverage <- tn_by_watershed %>%
  group_by(year) %>%
  summarize(
    n_basins = n_distinct(basin_id),
    tn_wastewater_t_yr = sum(tn_wastewater_t_yr),
    tn_agriculture_t_yr = sum(tn_agriculture_t_yr),
    tn_livestock_t_yr = sum(tn_livestock_t_yr),
    tn_shrimp_t_yr = sum(tn_shrimp_t_yr),
    tn_total_t_yr = sum(tn_total_t_yr),
    has_ww = tn_wastewater_t_yr > 0,
    has_ag = tn_agriculture_t_yr > 0,
    has_liv = tn_livestock_t_yr > 0,
    has_shr = tn_shrimp_t_yr > 0,
    n_sources_present = has_ww + has_ag + has_liv + has_shr,
    .groups = "drop"
  ) %>%
  arrange(year)

tn_year_coverage

# keep only years where all four sources are present gulf-wide
years_complete <- tn_year_coverage %>%
  filter(n_sources_present == 4) %>%
  pull(year)

tn_by_watershed <- tn_by_watershed %>%
  filter(year %in% years_complete)

cat("years kept (all four sources):", paste(years_complete, collapse = ", "), "\n")
```

### QA: livestock 2023 and shrimp carry-forward

```{r qa-gapfill-plotly}
# livestock: raw vs gapfilled gulf totals (2023 should no longer be a hole)
liv_raw_gulf <- liv %>%
  group_by(year) %>%
  summarize(tn_livestock_t_yr = sum(tn_livestock_t_yr, na.rm = TRUE), .groups = "drop") %>%
  mutate(series = "raw")

liv_fill_gulf <- liv_g %>%
  group_by(year) %>%
  summarize(tn_livestock_t_yr = sum(tn_livestock_t_yr, na.rm = TRUE), .groups = "drop") %>%
  mutate(series = "gapfilled")

liv_qa <- bind_rows(liv_raw_gulf, liv_fill_gulf)

liv_gapfill_plotly <- plot_ly(
  liv_qa,
  x = ~year,
  y = ~tn_livestock_t_yr,
  color = ~series,
  type = "scatter",
  mode = "lines+markers"
) %>%
  layout(
    title = "Livestock TN (gulf total): raw vs gapfilled",
    xaxis = list(title = "Year", dtick = 1),
    yaxis = list(title = "t N / year"),
    shapes = list(list(
      type = "line", x0 = 2023, x1 = 2023,
      y0 = 0, y1 = 1, yref = "paper",
      line = list(color = "red", dash = "dot")
    ))
  )

liv_gapfill_plotly

# shrimp: raw vs gapfilled (2022-2025 should match 2021)
shr_raw_gulf <- shr %>%
  group_by(year) %>%
  summarize(tn_shrimp_t_yr = sum(tn_shrimp_t_yr, na.rm = TRUE), .groups = "drop") %>%
  mutate(series = "raw")

shr_fill_gulf <- shr_g %>%
  group_by(year) %>%
  summarize(tn_shrimp_t_yr = sum(tn_shrimp_t_yr, na.rm = TRUE), .groups = "drop") %>%
  mutate(series = "gapfilled")

shr_qa <- bind_rows(shr_raw_gulf, shr_fill_gulf)

shr_gapfill_plotly <- plot_ly(
  shr_qa,
  x = ~year,
  y = ~tn_shrimp_t_yr,
  color = ~series,
  type = "scatter",
  mode = "lines+markers"
) %>%
  layout(
    title = "Shrimp TN (gulf total): raw vs gapfilled (approx between USV years; carry 2021 → 2022-2025)",
    xaxis = list(title = "Year", dtick = 1),
    yaxis = list(title = "t N / year"),
    shapes = list(list(
      type = "line", x0 = 2021, x1 = 2021,
      y0 = 0, y1 = 1, yref = "paper",
      line = list(color = "red", dash = "dot")
    ))
  )

shr_gapfill_plotly
```

---

## 3. Pourpoints → OHI regions

Same Chemicals pattern: join by `basin_id`, keep pourpoints inside the 50 km inland buffer, drop Pacific-side peninsula pourpoints that are outside the 10 km buffer for Regions 2, 4, and 9.

```{r pourpoints-to-regions}
pourpoints_goc <- st_read(here::here(spatial_regions, "pourpoints_goc.gpkg"), quiet = TRUE)

marine_and_inland50 <- st_read(
  here::here(spatial_regions, "marine_and_inland50.gpkg"),
  quiet = TRUE
)
marine_and_inland10 <- st_read(
  here::here(spatial_regions, "marine_and_inland10.gpkg"),
  quiet = TRUE
)

pour_tn <- pourpoints_goc %>%
  select(basin_id) %>%
  left_join(tn_by_watershed, by = "basin_id") %>%
  filter(!is.na(tn_total_t_yr), tn_total_t_yr > 0) %>%
  st_transform(st_crs(marine_and_inland50)) %>%
  st_intersection(marine_and_inland50) %>%
  mutate(
    is_within_10km = lengths(
      st_within(., st_transform(marine_and_inland10, st_crs(.)))
    ) > 0
  ) %>%
  filter(!(is_within_10km == FALSE & rgn_id %in% c("Region_2", "Region_4", "Region_9")))

tn_by_region <- pour_tn %>%
  st_drop_geometry() %>%
  group_by(rgn_id, year) %>%
  summarize(
    across(starts_with("tn_"), ~ sum(.x, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  arrange(year, rgn_id)

tn_by_region

tn_by_region %>%
  group_by(year) %>%
  summarize(
    n_regions = n_distinct(rgn_id),
    across(starts_with("tn_"), ~ sum(.x, na.rm = TRUE)),
    .groups = "drop"
  )
```

---

## 4. Quick look (plotly)

```{r plotly-sources-by-region}
tn_long <- tn_by_region %>%
  select(
    rgn_id, year,
    tn_wastewater_t_yr, tn_agriculture_t_yr,
    tn_livestock_t_yr, tn_shrimp_t_yr
  ) %>%
  pivot_longer(starts_with("tn_"), names_to = "source", values_to = "t_yr") %>%
  mutate(
    source = recode(
      source,
      tn_wastewater_t_yr = "wastewater",
      tn_agriculture_t_yr = "agriculture",
      tn_livestock_t_yr = "livestock",
      tn_shrimp_t_yr = "shrimp"
    )
  )

tn_sources_plotly <- plot_ly(
  tn_long,
  x = ~rgn_id,
  y = ~t_yr,
  color = ~source,
  frame = ~year,
  type = "bar",
  text = ~paste0(source, ": ", round(t_yr, 1), " t N/yr"),
  hoverinfo = "text+x"
) %>%
  layout(
    barmode = "stack",
    title = "Nitrogen by source and region (t N / year)",
    xaxis = list(title = NULL),
    yaxis = list(title = "t N / year"),
    legend = list(title = list(text = "Source"))
  )

tn_sources_plotly

tn_total_plotly <- plot_ly(
  tn_by_region,
  x = ~year,
  y = ~tn_total_t_yr,
  color = ~rgn_id,
  type = "scatter",
  mode = "lines+markers",
  text = ~paste0(rgn_id, "<br>", year, ": ", round(tn_total_t_yr, 1), " t N/yr"),
  hoverinfo = "text"
) %>%
  layout(
    title = "Total nitrogen by region over time (gapfilled complete panel)",
    xaxis = list(title = "Year", dtick = 1),
    yaxis = list(title = "t N / year"),
    legend = list(title = list(text = "Region"))
  )

tn_total_plotly
```

---

## 5. Write outputs

```{r write-outputs}
write_csv(tn_by_watershed, here::here(eut_out, "tn_total_by_watershed.csv"))
write_csv(tn_by_region, here::here(eut_out, "tn_total_by_region.csv"))
write_csv(tn_year_coverage, here::here(eut_out, "tn_year_coverage.csv"))
write_csv(source_years, here::here(eut_out, "tn_source_years_raw.csv"))
```
