---
title: 'OHI GoC eutrophication update: wastewater nitrogen (script 1)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    theme: cerulean
    highlight-style: haddock
editor_options:
  chunk_output_type: console
---

# Script 1: wastewater nitrogen

Here we estimate nitrogen from human wastewater following Romero-Gil et al. (2026), then allocate it to GoC watersheds (`basin_id`). We then use pourpoints to assign coastal watershed nitrogen output to OHI regions in script 5.

We will use the **mean** wastewater treatment pathway from JARG's Supplemental Table 6.

**Data and methods overview:** we reuse Chemicals pharmaceuticals WWTP spatial layers for treatment removal (NRE), and build the nitrogen load from INEGI ITER 2020 locality age × sex population plus ENSANUT / JARG Table 1 protein. Other years are scaled with CONAPO municipal total population ratios. Many 2020 SINA plant flow (`caudal`) values look wrong; we flag outliers and interpolate before computing NRE.

---

## What JARG did

$$
N_{\mathrm{ww}} = \mathrm{pop} \times \mathrm{protein}_{g/day} \times 365 \times 0.16 \times (1 - \mathrm{NRE})
$$

Protein came from ENSANUT by age and sex. Population came from the INEGI 2020 locality census. Urban areas got an NRE from CONAGUA treatment plants (flow-weighted by process). Rural areas had NRE = 0 (everything reaches water).

Their study covered 29 coastal municipalities and did not fully cover Nayarit and Jalisco, which we need for the GoC.

---

## Population data

### INEGI ITER 2020 (locality age × sex)

On mazu:

`/home/shares/ohi/OHI_GOC/_raw_data/inegi_population/d2026/iter_00_cpv2020/`

This is the **2020** Censo de Población y Vivienda ITER (locality-level). It gives:

- Locality IDs + names, and DMS `LONGITUD` / `LATITUD`
- `POBTOT`, `POBFEM`, `POBMAS`
- Quinquenal age × sex columns so we can match JARG Table 1
- `TAMLOC` size class (we treat TAMLOC ≥ 5, i.e. ≥ 2,500 people, as urban)

### CONAPO municipal projections (`pobproy_ggrupos.csv`)

On mazu:

`/home/shares/ohi/OHI_GOC/_raw_data/CONAPO_demographic_indicators/d2026/pobproy_ggrupos.csv`

Annual municipal population **1990–2040** by sex and broad age groups (`POB_00_011`, `POB_012_29`, `POB_30_59`, `POB_60_mm`, `POB_TOTAL`). Those age bins do not match Table 1 exactly, so we use CONAPO mainly to **scale the 2020 ITER nitrogen baseline** through time with `POB_TOTAL` ratios (municipality × year / municipality × 2020).

**Workflow:** 2020 detailed N from ITER × Table 1 × NRE → then multiply municipal N by CONAPO total-pop scale factors for other years → push to watersheds with 2020 locality→basin shares.

---

## What we already have from CW

From `chemicals/pharmaceuticals/script1_wwtp_effluent.qmd`:

| File | What it is |
|------|------------|
| `humans_int/wwtps_watershed_plant_level.gpkg` | Plant-level WWTPs ∩ watersheds |
| `humans_int/wwtp_liters_effluent_by_watershed.csv` | Effluent liters per watershed × year |
| `humans_int/effluent_wwtp_liters_rgnid.csv` | Same idea, already at region |
| `humans_int/pourpoints_effluent_wwtp_liters_final.gpkg` | Pourpoints with liters |

---

## Setup

```{r setup}
library(tidyverse)
library(here)
library(sf)
library(stringi)
library(janitor)
library(mapview)
library(plotly)
sf_use_s2(FALSE)
```

```{r paths-constants}
pharma_humans_int <- here::here(
  "/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chemicals/pharmaceuticals/humans_int"
)
eut_int <- here::here(
  "/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/eutrophication/jarg_analysis_update/int"
)
spatial_regions <- here::here("/home/shares/ohi/OHI_GOC/spatial/OHI_regions")

iter_dir <- here::here(
  "/home/shares/ohi/OHI_GOC/_raw_data/inegi_population/d2026/iter_00_cpv2020"
)
iter_csv <- here::here(iter_dir, "conjunto_de_datos", "conjunto_de_datos_iter_00CSV20.csv")
if (!file.exists(iter_csv)) {
  # sometimes the CSV sits directly in the folder
  cand <- list.files(iter_dir, pattern = "iter_00CSV20\\.csv$|conjunto_de_datos_iter", recursive = TRUE, full.names = TRUE)
  if (length(cand) > 0) iter_csv <- cand[[1]]
}

conapo_csv <- here::here(
  "/home/shares/ohi/OHI_GOC/_raw_data/CONAPO_demographic_indicators/d2026/pobproy_ggrupos.csv"
)

proto <- 0.16  # protein to N
year_census <- 2020L
# years we will scale to with CONAPO (overlap with WWTP / other eut sources as needed)
conapo_years <- 2000:2025

goc_entidad <- c("02", "03", "14", "18", "25", "26")  # BC, BCS, Jal, Nay, Sin, Son
goc_clave_ent <- c(2, 3, 14, 18, 25, 26)

# Table 1 mean protein (g / person / day)
protein_table1 <- tribble(
  ~age_group, ~sex, ~protein_g_day,
  "0-4",   "F", 37.10,
  "0-4",   "M", 39.80,
  "5-14",  "F", 54.80,
  "5-14",  "M", 59.80,
  "15-19", "F", 59.60,
  "15-19", "M", 75.00,
  "20-29", "F", 53.60,
  "20-29", "M", 69.60,
  "30-59", "F", 57.00,
  "30-59", "M", 77.70,
  "60+",   "F", 44.00,
  "60+",   "M", 61.50
)

# map ITER quinquenal columns onto Table 1 groups
age_col_map <- tribble(
  ~age_group, ~sex, ~iter_col,
  "0-4",   "F", "P_0A4_F",
  "0-4",   "M", "P_0A4_M",
  "5-14",  "F", "P_5A9_F",
  "5-14",  "M", "P_5A9_M",
  "5-14",  "F", "P_10A14_F",
  "5-14",  "M", "P_10A14_M",
  "15-19", "F", "P_15A19_F",
  "15-19", "M", "P_15A19_M",
  "20-29", "F", "P_20A24_F",
  "20-29", "M", "P_20A24_M",
  "20-29", "F", "P_25A29_F",
  "20-29", "M", "P_25A29_M",
  "30-59", "F", "P_30A34_F",
  "30-59", "M", "P_30A34_M",
  "30-59", "F", "P_35A39_F",
  "30-59", "M", "P_35A39_M",
  "30-59", "F", "P_40A44_F",
  "30-59", "M", "P_40A44_M",
  "30-59", "F", "P_45A49_F",
  "30-59", "M", "P_45A49_M",
  "30-59", "F", "P_50A54_F",
  "30-59", "M", "P_50A54_M",
  "30-59", "F", "P_55A59_F",
  "30-59", "M", "P_55A59_M",
  "60+",   "F", "P_60YMAS_F",
  "60+",   "M", "P_60YMAS_M"
)
```

---

## 1. NRE from WWTP plants (flow-weighted by watershed)

We reuse the Chemicals plant×watershed table, but first we fix **2020 caudal**. Many plants show a 2020 flow that is far from every other year for that plant. We:

1. keep one row per plant × year
2. compare each plant's 2020 caudal to its median in non-2020 years
3. drop 2020 values that are more than 5× larger or smaller than that median
4. fill the gap by interpolating across years (or, if the plant only has one year of data, reuse that value)

Then we attach an NRE by process name and take a **flow-weighted mean NRE per watershed** in the census year.

```{r nre-lookup}
nre_by_process <- tribble(
  ~proceso_key,                ~nre,
  "lodos activados",           0.75,
  "lagunas aireadas",          0.60,
  "lagunas de estabilizacion", 0.60,
  "lagunas de estabilización", 0.60,
  "primario avanzado",         0.10,
  "tanque septico",            0.85,
  "tanque séptico",            0.85,
  "biodiscos",                 0.45,
  "filtros biologicos",        0.45,
  "filtros biológicos",        0.45,
  "humedal",                   0.45
)
```

```{r read-wwtp-plants}
wwtps_plants <- st_read(here::here(pharma_humans_int, "wwtps_watershed_plant_level.gpkg")) %>%
  st_drop_geometry() %>%
  clean_names() %>%
  mutate(
    anio = as.integer(anio),
    caudal = as.numeric(caudal),
    proceso_clean = str_to_lower(stri_trans_general(as.character(proceso), "Latin-ASCII"))
  ) %>%
  # sina gpkg currently repeats each plant-year 7 times; keep one
  distinct(clvplanta, anio, .keep_all = TRUE)
```

```{r fix-2020-caudal}
# fill missing caudal along a plant's year series.
# needs at least two years with data to interpolate; one year → copy that value; none → leave NA.
fill_caudal_series <- function(year, caudal) {
  keep <- is.finite(caudal)
  year_ok <- year[keep]
  caudal_ok <- caudal[keep]

  n_years <- n_distinct(year_ok)

  if (n_years >= 2) {
    # one value per year, then linear interpolate / extrapolate
    series <- tibble(year = year_ok, caudal = caudal_ok) %>%
      group_by(year) %>%
      summarize(caudal = mean(caudal), .groups = "drop")

    approx(series$year, series$caudal, xout = year, rule = 2)$y
  } else if (n_years == 1) {
    rep(mean(caudal_ok), length(year))
  } else {
    caudal
  }
}

wwtps_plants <- wwtps_plants %>%
  group_by(clvplanta) %>%
  mutate(
    # plant's typical flow outside 2020
    caudal_typical = median(caudal[anio != 2020], na.rm = TRUE),

    # 2020 looks broken if it is >5x or <1/5 of that typical flow
    caudal_2020_outlier = anio == 2020 &
      is.finite(caudal) &
      is.finite(caudal_typical) &
      caudal_typical > 0 &
      (caudal > 5 * caudal_typical | caudal < caudal_typical / 5),

    # set outliers to NA so the fill step can replace them
    caudal_clean = if_else(caudal_2020_outlier, NA_real_, caudal)
  ) %>%
  arrange(anio) %>%
  mutate(caudal_fix = fill_caudal_series(anio, caudal_clean)) %>%
  ungroup()
```

### QA: did the 2020 fix look right?

Same idea as Chemicals `script1_wwtp_effluent.qmd`: interactive plant time series + a red dotted line at 2020. Chemicals used a **3×** flag and plotted the **raw** series; here we use **5×**, and we add a **fixed** series plot plus a 2020 raw-vs-fixed scatter so you can confirm the interpolation.

```{r qa-2020-caudal-plotly}
flagged_plants <- wwtps_plants %>%
  filter(caudal_2020_outlier) %>%
  distinct(clvplanta) %>%
  pull(clvplanta)

cat("flagged 2020 plants:", length(flagged_plants), "\n")

caudal_qa <- wwtps_plants %>%
  filter(clvplanta %in% flagged_plants) %>%
  select(clvplanta, anio, basin_id, proceso, capacidad, caudal, caudal_fix, caudal_typical)

hover_txt <- ~paste0(
  "Plant: ", clvplanta,
  "<br>Basin: ", basin_id,
  "<br>Proceso: ", proceso,
  "<br>Capacidad: ", capacidad,
  "<br>Caudal: ", caudal_l_s, " L/s"
)

# 1) raw series (chemicals-style: find the 2020 spike/dip)
caudal_raw_flagged_plotly <- plot_ly(
  caudal_qa %>% mutate(caudal_l_s = caudal),
  x = ~anio,
  y = ~caudal_l_s,
  color = ~clvplanta,
  type = "scatter",
  mode = "lines+markers",
  text = hover_txt,
  hoverinfo = "text+x+y"
) %>%
  layout(
    title = "Flagged plants: raw caudal (look for 2020 outliers)",
    xaxis = list(title = "Year", dtick = 1),
    yaxis = list(title = "Caudal (L/s)"),
    shapes = list(list(
      type = "line", x0 = 2020, x1 = 2020,
      y0 = 0, y1 = 1, yref = "paper",
      line = list(color = "red", dash = "dot")
    ))
  )

caudal_raw_flagged_plotly

# 2) fixed series (same plants; 2020 should sit with the rest of the series)
caudal_fix_flagged_plotly <- plot_ly(
  caudal_qa %>% mutate(caudal_l_s = caudal_fix),
  x = ~anio,
  y = ~caudal_l_s,
  color = ~clvplanta,
  type = "scatter",
  mode = "lines+markers",
  text = hover_txt,
  hoverinfo = "text+x+y"
) %>%
  layout(
    title = "Flagged plants: fixed caudal (after 2020 interpolation)",
    xaxis = list(title = "Year", dtick = 1),
    yaxis = list(title = "Caudal (L/s)"),
    shapes = list(list(
      type = "line", x0 = 2020, x1 = 2020,
      y0 = 0, y1 = 1, yref = "paper",
      line = list(color = "red", dash = "dot")
    ))
  )

caudal_fix_flagged_plotly

# 3) 2020 only: raw vs fixed (points off the 1:1 line were corrected)
caudal_2020_compare <- caudal_qa %>%
  filter(anio == 2020)

one_to_one <- range(
  c(caudal_2020_compare$caudal, caudal_2020_compare$caudal_fix),
  na.rm = TRUE
)

caudal_2020_scatter_plotly <- plot_ly(
  caudal_2020_compare,
  x = ~caudal,
  y = ~caudal_fix,
  type = "scatter",
  mode = "markers",
  text = ~paste0(
    clvplanta, " | ", basin_id,
    "<br>Raw: ", caudal, " L/s",
    "<br>Fixed: ", caudal_fix, " L/s",
    "<br>Typical (non-2020 median): ", round(caudal_typical, 2)
  ),
  hoverinfo = "text",
  name = "plants"
) %>%
  add_trace(
    x = one_to_one,
    y = one_to_one,
    type = "scatter",
    mode = "lines",
    line = list(color = "gray", dash = "dash"),
    showlegend = FALSE,
    hoverinfo = "skip",
    inherit = FALSE
  ) %>%
  layout(
    title = "2020 caudal: raw (x) vs fixed (y); dashed = 1:1",
    xaxis = list(title = "Raw caudal 2020 (L/s)", type = "log"),
    yaxis = list(title = "Fixed caudal 2020 (L/s)", type = "log")
  )

caudal_2020_scatter_plotly
```

```{r nre-by-basin}
# first matching process key wins
match_nre <- function(proceso_clean) {
  hits <- nre_by_process$nre[str_detect(proceso_clean, fixed(nre_by_process$proceso_key))]
  if (length(hits) == 0 || all(is.na(hits))) return(NA_real_)
  hits[[1]]
}

wwtps_plants <- wwtps_plants %>%
  mutate(nre = map_dbl(proceso_clean, match_nre))

# flow-weighted NRE in the census year (caudal is L/s → L/year)
nre_by_basin <- wwtps_plants %>%
  filter(
    anio == year_census,
    !is.na(basin_id),
    is.finite(caudal_fix),
    caudal_fix > 0
  ) %>%
  mutate(ef_l_yr = caudal_fix * 31536000) %>%
  group_by(basin_id) %>%
  summarize(
    nre_flow_weighted = sum(nre * ef_l_yr, na.rm = TRUE) / sum(ef_l_yr, na.rm = TRUE),
    n_plants = n(),
    .groups = "drop"
  )

summary(nre_by_basin$nre_flow_weighted)
```

---

## 2. Read 2020 ITER localities and compute protein-weighted N (before NRE)

```{r parse-dms}
# ITER stores coords like 116°35'42.482" W
parse_dms <- function(x) {
  x <- str_trim(as.character(x))
  out <- rep(NA_real_, length(x))
  ok <- !is.na(x) & x != "" & x != "*"
  # degrees, minutes, seconds, hemisphere
  m <- str_match(x[ok], "([0-9]+).*?([0-9]+).*?([0-9.]+).*?([NnSsEeWw])")
  deg <- as.numeric(m[, 2])
  minu <- as.numeric(m[, 3])
  sec <- as.numeric(m[, 4])
  hemi <- toupper(m[, 5])
  val <- deg + minu / 60 + sec / 3600
  val[hemi %in% c("W", "S")] <- -val[hemi %in% c("W", "S")]
  out[ok] <- val
  out
}
```

```{r read-iter-protein-n}
iter_cols_needed <- c(
  "ENTIDAD", "NOM_ENT", "MUN", "NOM_MUN", "LOC", "NOM_LOC",
  "LONGITUD", "LATITUD", "TAMLOC", "POBTOT",
  unique(age_col_map$iter_col)
)

iter_raw <- read_csv(
  iter_csv,
  col_select = any_of(iter_cols_needed),
  locale = locale(encoding = "UTF-8"),
  show_col_types = FALSE,
  na = c("", "NA", "N/A", "*")
) %>%
  clean_names() %>%
  mutate(
    entidad = str_pad(as.character(entidad), 2, pad = "0"),
    mun = str_pad(as.character(mun), 3, pad = "0"),
    loc = str_pad(as.character(loc), 4, pad = "0")
  ) %>%
  filter(
    entidad %in% goc_entidad,
    mun != "000",
    !loc %in% c("0000", "9998", "9999")
  )

# long age × sex pops, join Table 1 protein
age_col_map_clean <- age_col_map %>%
  mutate(iter_col_clean = make_clean_names(iter_col))

iter_age_long <- iter_raw %>%
  pivot_longer(
    cols = any_of(age_col_map_clean$iter_col_clean),
    names_to = "iter_col_clean",
    values_to = "pop",
    values_transform = list(pop = as.numeric)
  ) %>%
  left_join(age_col_map_clean, by = "iter_col_clean", relationship = "many-to-many") %>%
  left_join(protein_table1, by = c("age_group", "sex")) %>%
  filter(!is.na(protein_g_day), is.finite(pop), pop > 0) %>%
  # kg N / year before NRE: protein_g_day * 0.365 * 0.16 * pop
  mutate(n_kg_yr_gross = protein_g_day * 0.365 * proto * pop)

tn_loc_gross <- iter_age_long %>%
  group_by(entidad, nom_ent, mun, nom_mun, loc, nom_loc, longitud, latitud, tamloc, pobtot) %>%
  summarize(
    n_kg_yr_gross = sum(n_kg_yr_gross, na.rm = TRUE),
    pop_age_sex = sum(pop, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    year = year_census,
    lon = parse_dms(longitud),
    lat = parse_dms(latitud),
    urban = as.numeric(tamloc) >= 5,  # >= 2500 people
    tn_gross_t_yr = n_kg_yr_gross / 1000
  ) %>%
  filter(is.finite(lon), is.finite(lat))

glimpse(tn_loc_gross)
sum(tn_loc_gross$tn_gross_t_yr, na.rm = TRUE)
```

---

## 3. Assign localities to watersheds and apply (1 - NRE)

```{r loc-to-watershed-nre}
watershed_goc <- st_read(here::here(spatial_regions, "watersheds_goc.gpkg")) %>%
  st_make_valid()

loc_sf <- tn_loc_gross %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
  st_transform(st_crs(watershed_goc))

loc_ws <- st_join(loc_sf, watershed_goc %>% select(basin_id), join = st_within, left = TRUE)

# points on edges can miss; try nearest basin if needed
missing <- loc_ws %>% filter(is.na(basin_id))
if (nrow(missing) > 0) {
  nearest_i <- st_nearest_feature(missing, watershed_goc)
  loc_ws$basin_id[is.na(loc_ws$basin_id)] <- watershed_goc$basin_id[nearest_i]
}

tn_ww_locality <- loc_ws %>%
  st_drop_geometry() %>%
  left_join(nre_by_basin, by = "basin_id") %>%
  mutate(
    nre = case_when(
      urban & is.finite(nre_flow_weighted) ~ nre_flow_weighted,
      urban & !is.finite(nre_flow_weighted) ~ 0,  # urban but no plant in basin: treat as untreated
      TRUE ~ 0                                   # rural
    ),
    n_kg_yr = n_kg_yr_gross * (1 - nre),
    tn_wastewater_t_yr = n_kg_yr / 1000
  )

tn_ww_by_watershed_2020 <- tn_ww_locality %>%
  group_by(basin_id, year) %>%
  summarize(
    tn_wastewater_t_yr = sum(tn_wastewater_t_yr, na.rm = TRUE),
    n_localities = n(),
    pop_age_sex = sum(pop_age_sex, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(tn_wastewater_t_yr))

tn_ww_by_watershed_2020
sum(tn_ww_by_watershed_2020$tn_wastewater_t_yr, na.rm = TRUE)
```

---

## 4. Scale 2020 N through time with CONAPO `POB_TOTAL`

We hold the 2020 protein × NRE structure fixed, and scale each municipality's 2020 wastewater N by:

$$
\mathrm{scale}_{m,t} = \frac{\mathrm{POB\_TOTAL}_{m,t}}{\mathrm{POB\_TOTAL}_{m,2020}}
$$

Then we redistribute each year's municipal N to basins using the **2020** locality→basin shares (so spatial pattern stays from ITER).

```{r conapo-scales}
conapo <- read_csv(conapo_csv, locale = locale(encoding = "latin1"), show_col_types = FALSE) %>%
  clean_names() %>%
  filter(clave_ent %in% goc_clave_ent, ano %in% conapo_years) %>%
  mutate(
    idestado = as.integer(clave_ent),
    idmunicipio = as.integer(clave %% 1000),
    year = as.integer(ano),
    pob_total = as.numeric(pob_total)
  ) %>%
  group_by(idestado, idmunicipio, year) %>%
  summarize(pob_total = sum(pob_total, na.rm = TRUE), .groups = "drop")  # sum HOMBRES + MUJERES

conapo_scale <- conapo %>%
  left_join(
    conapo %>%
      filter(year == year_census) %>%
      select(idestado, idmunicipio, pob_2020 = pob_total),
    by = c("idestado", "idmunicipio")
  ) %>%
  mutate(scale_pop = if_else(pob_2020 > 0, pob_total / pob_2020, NA_real_))

summary(conapo_scale$scale_pop)
```

```{r scale-mun-to-watershed-years}
# 2020 municipal TN after NRE
tn_ww_mun_2020 <- tn_ww_locality %>%
  mutate(
    idestado = as.integer(entidad),
    idmunicipio = as.integer(mun)
  ) %>%
  group_by(idestado, nom_ent, idmunicipio, nom_mun) %>%
  summarize(tn_wastewater_t_yr_2020 = sum(tn_wastewater_t_yr, na.rm = TRUE), .groups = "drop")

# 2020 locality shares of each municipality's TN within each basin
loc_mun_basin_w <- tn_ww_locality %>%
  mutate(
    idestado = as.integer(entidad),
    idmunicipio = as.integer(mun)
  ) %>%
  group_by(idestado, idmunicipio, basin_id) %>%
  summarize(tn_loc = sum(tn_wastewater_t_yr, na.rm = TRUE), .groups = "drop") %>%
  group_by(idestado, idmunicipio) %>%
  mutate(
    tn_mun = sum(tn_loc, na.rm = TRUE),
    w_basin = if_else(tn_mun > 0, tn_loc / tn_mun, NA_real_)
  ) %>%
  ungroup() %>%
  select(idestado, idmunicipio, basin_id, w_basin)

tn_ww_by_watershed <- tn_ww_mun_2020 %>%
  inner_join(
    conapo_scale %>% select(idestado, idmunicipio, year, scale_pop),
    by = c("idestado", "idmunicipio")
  ) %>%
  mutate(tn_wastewater_t_yr = tn_wastewater_t_yr_2020 * scale_pop) %>%
  inner_join(loc_mun_basin_w, by = c("idestado", "idmunicipio")) %>%
  mutate(tn_wastewater_t_yr = tn_wastewater_t_yr * w_basin) %>%
  group_by(basin_id, year) %>%
  summarize(
    tn_wastewater_t_yr = sum(tn_wastewater_t_yr, na.rm = TRUE),
    n_municipalities = n_distinct(paste(idestado, idmunicipio)),
    .groups = "drop"
  ) %>%
  arrange(year, desc(tn_wastewater_t_yr))

tn_ww_by_watershed %>%
  group_by(year) %>%
  summarize(tn_wastewater_t_yr = sum(tn_wastewater_t_yr, na.rm = TRUE), .groups = "drop")
```

Note: NRE is baked into the 2020 baseline. We do not re-estimate plant NRE for every CONAPO year here (WWTP years are shorter anyway). A later refinement can apply year-specific flow-weighted NRE where SINA years exist.

---

## 5. Write watershed output

Primary deliverable for script 5 is watershed TN (coefficients / NRE already applied). Pourpoints and OHI regions are handled in script 5.

```{r write-outputs}
# optional QA intermediates
# write_csv(tn_ww_locality, here::here(eut_int, "tn_wastewater_locality_2020.csv"))
# write_csv(tn_ww_mun_2020, here::here(eut_int, "tn_wastewater_municipal_2020.csv"))

write_csv(tn_ww_by_watershed, here::here(eut_int, "tn_wastewater_by_watershed.csv"))
```
