---
title: 'OHI GoC eutrophication update: livestock nitrogen (script 3)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    theme: cerulean
    highlight-style: haddock
editor_options:
  chunk_output_type: console
---

# Script 3: livestock nitrogen

This script estimates nitrogen from livestock following Romero-Gil et al. (2026), then routes it through watersheds and pourpoints.

For scoring later, we want a **40% export coefficient** (JARG Supplemental Table 6).

**Data decision:** use SIAP municipal **heads** from the pharmaceuticals clean file  
(`siap_livestock_municipal_clean.csv`, product Carne, column `cabezas_asacrificados`).  
That matches JARG’s “livestock heads from SIAP yearbooks” closely enough. Do not overwrite that Chemicals file.

---

## What JARG did

$$
N_{\mathrm{liv}} = \left(\sum \mathrm{heads}_s \times EF_s\right) \times E_{\mathrm{export}}
$$

| Species (`nomespecie`) | EF (kg N head⁻¹ year⁻¹) |
|------------------------|-------------------------|
| Bovino | 21.9 |
| Caprino | 6.3 |
| Ovino | 4.1 |
| Porcino | 4.1 |
| Ave | 0.79 |

---

## Setup

```{r setup}
library(tidyverse)
library(here)
library(sf)
library(janitor)
library(mapview)
```

```{r paths-constants}
export_cw <- 0.40

ef_kg_n_per_head_yr <- tribble(
  ~nomespecie, ~ef,
  "Bovino",  21.9,
  "Caprino",  6.3,
  "Ovino",    4.1,
  "Porcino",  4.1,
  "Ave",      0.79
)

livestock_int <- here::here(
  "/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chemicals/pharmaceuticals/livestock_int"
)
eut_int <- here::here(
  "/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/eutrophication/jarg_analysis_update/int"
)
```

### Read SIAP municipal heads (already written from pharmaceuticals)

```{r read-siap}
siap_municipal <- read_csv(
  here::here(livestock_int, "siap_livestock_municipal_clean.csv")
)

glimpse(siap_municipal)
```

### Apply emission factors and 40% export (municipal)

```{r livestock-n-municipal}
tn_liv_municipal <- siap_municipal %>%
  filter(nomespecie %in% ef_kg_n_per_head_yr$nomespecie) %>%
  left_join(ef_kg_n_per_head_yr, by = "nomespecie") %>%
  mutate(
    heads = cabezas_asacrificados,
    n_kg_gross = heads * ef,
    n_kg_export = n_kg_gross * export_cw,
    tn_livestock_t_yr = n_kg_export / 1000
  )

tn_liv_municipal %>%
  group_by(anio) %>%
  summarize(tn_livestock_t_yr = sum(tn_livestock_t_yr, na.rm = TRUE), .groups = "drop") %>%
  arrange(anio)
```

### Still needed: watershed → pourpoint → region

```{r spatial-todo}
# still to do in this script (same pattern as pharmaceuticals livestock):
# 1) dasymetric-allocate municipal tn with FAO GLW4
# 2) sum by basin_id
# 3) pourpoint -> ohi region (50 km / 10 km filters)
# 4) write:
#    here::here(eut_int, "tn_livestock_by_watershed.csv")
#    here::here(eut_int, "tn_livestock_by_region.csv")
```

### Write municipal intermediate (optional QA)

```{r write-todo}
# write_csv(tn_liv_municipal, here::here(eut_int, "tn_livestock_municipal.csv"))
```

---

## Status

| Piece | Status |
|-------|--------|
| SIAP heads path | Ready (`siap_livestock_municipal_clean.csv`) |
| EF × 40% at municipality | Coded |
| Watershed / pourpoint / region | **Not coded yet** |
| Region TN file for script 5 | Pending |

Next: `script4_shrimp_aquaculture.qmd`
