---
title: 'OHI GoC eutrophication update: agriculture nitrogen (script 2)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    theme: cerulean
    highlight-style: haddock
editor_options:
  chunk_output_type: console
---

# Script 2: agriculture nitrogen

This script estimates nitrogen runoff from agriculture following Romero-Gil et al. (2026), then sends it through watersheds and pourpoints.

For scoring later, we want a **10% leaching coefficient** (JARG Supplemental Table 6).

Important: we only want **processed** CONABIO USV agriculture layers for where crops sit. Crop area × fertilizer rates come from DGSIAP / SIAP.

---

## What JARG did

$$
N_{\mathrm{ag}} = \sum (\mathrm{area}_c \times \mathrm{fertilizer}_c \times L_c)
$$

- Area: SIAP irrigated sown area
- Fertilizer rates: `Fertilizantes.csv` (INIFAP / Páez-Osuna)
- Leaching: **we use 10%**

JARG's older extract stopped at 2020 (`cultivos4.csv`). We now also have DGSIAP cierre files through 2025.

---

## Data paths

```{r paths}
pesticides_int <- here::here(
  "/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chemicals/pesticides/int"
)
dgsiap_agricola <- here::here(
  "/home/shares/ohi/OHI_GOC/_raw_data/DGSIAP_produccion_agricola/d2026"
)
jarg_repo <- here::here(
  "/home/shares/ohi/OHI_GOC/_raw_data/romero_gil_2026_estimated_nitrogen_discharge_goc/d2026/OceanHealth-Sanitation-main"
)
eut_int <- here::here(
  "/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/eutrophication/jarg_analysis_update/int"
)
lc_cw <- 0.10

goc_states <- c(
  "Baja California", "Baja California Sur", "Sonora",
  "Sinaloa", "Nayarit", "Jalisco"
)
```

Already available:

| File | Role |
|------|------|
| `pesticides/int/ag_area_ha_by_watershed.csv` | USV ag area by watershed (weights) |
| `Fertilizantes.csv` (JARG repo) | kg N / ha by crop |
| `DGSIAP_produccion_agricola/d2026/Cierre_agricola_mun_YYYY.csv` | Municipal sown area 2019-2025 (extend with older years / cultivos4 as needed) |

---

## Setup

```{r setup}
library(tidyverse)
library(here)
library(sf)
library(janitor)
library(mapview)
```

### Read USV ag area by watershed

```{r read-ag-area-watershed}
ag_area_by_watershed <- read_csv(
  here::here(pesticides_int, "ag_area_ha_by_watershed.csv")
)

glimpse(ag_area_by_watershed)
```

### Read fertilizer rates

```{r fertilizer-rates}
fert <- read_csv(here::here(jarg_repo, "Fertilizantes.csv")) %>%
  clean_names()

fert
```

### Read DGSIAP cierre agricola (2019-2025 shown here)

```{r read-dgsiap-agricola}
# files currently in the folder: 2019 through 2025
ag_years <- 2019:2025

# read one year of municipal cierre
read_ag_year <- function(year) {
  read_csv(
    here::here(dgsiap_agricola, paste0("Cierre_agricola_mun_", year, ".csv"))
  ) %>%
    clean_names() %>%
    mutate(year = as.numeric(anio))
}

# bind years and keep GoC states + irrigated modality
cultivos_dgsiap <- map_dfr(ag_years, read_ag_year) %>%
  filter(
    nomestado %in% goc_states,
    idmodalidad == 1  # riego, same as jarg
  )

glimpse(cultivos_dgsiap)
unique(cultivos_dgsiap$nomestado)
```

### Optional: older years from JARG cultivos4 (2003-2020)

```{r read-cultivos4-optional}
# use if you want the long pre-2019 series
# cultivos4 <- read_csv(here::here(jarg_repo, "cultivos4.csv")) %>%
#   clean_names() %>%
#   filter(nomestado %in% goc_states, idmodalidad == 1)
```

### Sketch: applied N then 10% leaching

```{r ag-n-todo}
# 1) join cultivos to fert by idcultivo
# 2) nin_kg = sembrada * fertilizante_n_kg_ha
# 3) nag_kg = nin_kg * 0.10
# 4) allocate municipal nag to watersheds with ag_area_by_watershed shares
# 5) pourpoint -> region
```

### Write outputs (when ready)

```{r write-todo}
# write_csv(tn_ag_by_watershed, here::here(eut_int, "tn_agriculture_by_watershed.csv"))
# write_csv(tn_ag_by_region,    here::here(eut_int, "tn_agriculture_by_region.csv"))
```

---

## Status

| Piece | Status |
|-------|--------|
| DGSIAP 2019-2025 path | Wired in |
| Fertilizantes + USV watershed area | Ready |
| Full N calc + pourpoints | Still to finish |
| Pre-2019 series | Optional via cultivos4 |

Next: `script3_livestock.qmd`
