---
title: 'OHI GoC eutrophication update: total nitrogen (script 5)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    theme: cerulean
    highlight-style: haddock
editor_options:
  chunk_output_type: console
---

# Script 5: total nitrogen by region

This script just adds the four sources together so we have total nitrogen entering the Gulf by OHI region (and by watershed, for QA).

Coefficients baked into scripts 1-4:

| Source | Coefficient |
|--------|-------------|
| Wastewater | Mean |
| Agriculture | 10% leaching |
| Livestock | 40% export |
| Shrimp | 100% export |

$$
TN = N_{\mathrm{ww}} + N_{\mathrm{ag}} + N_{\mathrm{liv}} + N_{\mathrm{shrimp}}
$$

We are **not** dividing by coastline here. That was the old approach and it made Mexicali look much worse than the water-delivery story suggests. Scoring / reference points live in script 6 (and we are not ready for that yet).

---

## Inputs (from scripts 1-4)

Expected under:

`/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/eutrophication/jarg_analysis_update/int/`

| File | From |
|------|------|
| `tn_wastewater_by_region.csv` | script 1 |
| `tn_agriculture_by_region.csv` | script 2 |
| `tn_livestock_by_region.csv` | script 3 |
| `tn_shrimp_by_region.csv` | script 4 |

Same idea at watershed level is nice for maps:

- `tn_*_by_watershed.csv`

Each region table should at least have: `rgn_id`, `year` (if available), and the source TN column in **t N per year**.

---

## Outline

1. Read the four region tables.
2. Full join on `rgn_id` (+ `year` if present).
3. Replace missing sources with 0 only after checking they are true zeros, not missing files.
4. Sum to `tn_total_t_yr`.
5. Quick plots (stacked bars by source, map if watershed version exists).
6. Write `output/tn_total_by_region.csv` (keep source columns too).

---

## Setup

```{r setup}
library(tidyverse)
library(here)
library(janitor)
```

```{r paths}
eut_int <- here::here(
  "/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/eutrophication/jarg_analysis_update/int"
)
eut_out <- here::here(
  "/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/eutrophication/jarg_analysis_update/output"
)
```

### Read sources (once scripts 1-4 have written them)

```{r read-sources}
# ww  <- read_csv(here::here(eut_int, "tn_wastewater_by_region.csv"))
# ag  <- read_csv(here::here(eut_int, "tn_agriculture_by_region.csv"))
# liv <- read_csv(here::here(eut_int, "tn_livestock_by_region.csv"))
# shr <- read_csv(here::here(eut_int, "tn_shrimp_by_region.csv"))
```

### Add them up

```{r combine}
# tn_total_by_region <- ww %>%
#   full_join(ag,  by = c("rgn_id", "year")) %>%
#   full_join(liv, by = c("rgn_id", "year")) %>%
#   full_join(shr, by = c("rgn_id", "year")) %>%
#   mutate(
#     across(starts_with("tn_"), ~ replace_na(.x, 0)),
#     tn_total_t_yr = tn_wastewater_t_yr + tn_agriculture_t_yr +
#       tn_livestock_t_yr + tn_shrimp_t_yr
#   )
#
# tn_total_by_region
```

### Quick look

```{r plot-todo}
# tn_long <- tn_total_by_region %>%
#   select(rgn_id, year, starts_with("tn_")) %>%
#   select(-tn_total_t_yr) %>%
#   pivot_longer(starts_with("tn_"), names_to = "source", values_to = "t_yr")
#
# ggplot(tn_long, aes(x = rgn_id, y = t_yr, fill = source)) +
#   geom_col() +
#   labs(x = NULL, y = "t N / year", title = "Nitrogen by source and region") +
#   theme_classic() +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Write

```{r write-todo}
# write_csv(tn_total_by_region, here::here(eut_out, "tn_total_by_region.csv"))
```

---

## Status

| Piece | Status |
|-------|--------|
| Combine plan | Ready |
| Actual joins | Blocked until scripts 1-4 write region tables |
| No coastline standardization | Intentional |

Next: `script6_eut_current_status.qmd` (hold for discussion)
