---
title: 'OHI GoC eutrophication update: current status (script 6)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    theme: cerulean
    highlight-style: haddock
editor_options:
  chunk_output_type: console
---

# Script 6: eutrophication current status

Here we turn regional total nitrogen from script 5 into a pressure and a current status score.

Input:

`jarg_analysis_update/output/tn_total_by_region.csv`

We follow the same **Chemicals** status pattern (ideal = 0; maximum = mean of the top 5 observations), but we score on **raw t N / year** from pourpoints. We intentionally do **not** divide by coastline. The old coastline workflow made Region 1 (Mexicali) look much worse than the water-delivery story, and it gapfilled Region 9 from Region 8. We do not repeat that.

$$
\mathrm{pressure} = \min\left(1,\ \frac{TN_{\mathrm{current}} - TN_{\mathrm{ideal}}}{TN_{\mathrm{maximum}} - TN_{\mathrm{ideal}}}\right)
$$

$$
\mathrm{status} = (1 - \mathrm{pressure}) \times 100
$$

with $TN_{\mathrm{ideal}} = 0$ and $TN_{\mathrm{maximum}}$ = mean of the top 5 `tn_total_t_yr` values across all regions and years in the scoring panel.

This is a **working default** aligned with pharmaceuticals / pesticides / untreated wastewater. If the EWG prefers a different reference later, we only need to change the maximum step.

---

## What we did last time (context only)

The old `cw_eutrophication.qmd` workflow:

1. Used JARG Table 6 with Mean / 10% / 40% / 100%
2. Divided municipal TN by coastline km
3. Used the max municipal ton/km as the reference
4. Weighted municipalities into regions by coastline
5. Gapfilled Region 9 with Region 8

We keep that history here so we remember why this update is different.

---

## Setup

```{r setup}
library(tidyverse)
library(here)
library(plotly)

region_levels <- paste0("Region_", 1:9)

rgn_colors <- c(
  Region_1 = "#C03729",
  Region_2 = "#E68C7C",
  Region_3 = "#FC8F24",
  Region_4 = "#dcaa38",
  Region_5 = "#cbd2a0",
  Region_6 = "#98a25a",
  Region_7 = "#747428",
  Region_8 = "#627391",
  Region_9 = "#8d61a9"
)

rgn_labels <- c(
  "Region_1" = "Western Upper Gulf",
  "Region_2" = "Western Midriff Islands",
  "Region_3" = "Western Central Gulf",
  "Region_4" = "Western Southern Gulf",
  "Region_5" = "Eastern Upper Gulf",
  "Region_6" = "Eastern Midriff Islands",
  "Region_7" = "Southern Sonora Coast",
  "Region_8" = "Sinaloa Coast",
  "Region_9" = "Southeastern Gulf"
)

# plotly legend uses the color column values; map palette to display names
rgn_colors_labeled <- setNames(
  unname(rgn_colors[region_levels]),
  unname(rgn_labels[region_levels])
)
```

```{r paths}
eut_out <- here::here(
  "/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/eutrophication/jarg_analysis_update/output"
)

dir.create(eut_out, recursive = TRUE, showWarnings = FALSE)
```

---

## 1. Read script 5 totals

```{r read-tn}
tn <- read_csv(
  here::here(eut_out, "tn_total_by_region.csv"),
  show_col_types = FALSE
) %>%
  mutate(year = as.integer(year)) %>%
  arrange(year, rgn_id)

# optional: year coverage from script 5, if present
coverage_path <- here::here(eut_out, "tn_year_coverage.csv")
if (file.exists(coverage_path)) {
  tn_year_coverage <- read_csv(coverage_path, show_col_types = FALSE)
  tn_year_coverage
}

tn
```

### Scoring panel

Script 5 already gapfills and keeps only years where all four sources are present. We score on that full table.

```{r scoring-panel}
tn_score <- tn

score_years <- sort(unique(tn_score$year))

cat("scoring years (from script 5 complete panel):", paste(score_years, collapse = ", "), "\n")
cat("n region-years:", nrow(tn_score), "\n")

if (exists("tn_year_coverage")) {
  tn_year_coverage
}

tn_score
```

---

## 2. Reference point and status

```{r score}
ideal <- 0

top5 <- tn_score %>%
  slice_max(tn_total_t_yr, n = 5, with_ties = FALSE)

top5

maximum <- mean(top5$tn_total_t_yr)
maximum

nutrient_status <- tn_score %>%
  mutate(
    ideal = ideal,
    maximum = maximum,
    pressure = pmin(1, (tn_total_t_yr - ideal) / (maximum - ideal)),
    status = (1 - pressure) * 100
  ) %>%
  arrange(year, rgn_id)

nutrient_status
```

---

## 3. Quick look (plotly)

```{r plotly-status}
nutrient_status_plot <- nutrient_status %>%
  mutate(
    rgn_id = factor(rgn_id, levels = region_levels),
    rgn_label = factor(
      rgn_labels[as.character(rgn_id)],
      levels = unname(rgn_labels[region_levels])
    )
  )

status_plotly <- plot_ly(
  nutrient_status_plot,
  x = ~year,
  y = ~status,
  color = ~rgn_label,
  colors = rgn_colors_labeled,
  type = "scatter",
  mode = "lines+markers",
  text = ~paste0(
    rgn_id, " (", rgn_label, ")",
    "<br>Year: ", year,
    "<br>TN: ", round(tn_total_t_yr, 1), " t N/yr",
    "<br>Pressure: ", round(pressure, 3),
    "<br>Status: ", round(status, 1)
  ),
  hoverinfo = "text"
) %>%
  layout(
    title = "Clean Waters: eutrophication current status (top-5 mean reference)",
    xaxis = list(title = "Year", dtick = 1),
    yaxis = list(title = "Status (0-100)", range = c(0, 100)),
    legend = list(title = list(text = "Region")),
    shapes = list(list(
      type = "line",
      x0 = min(nutrient_status_plot$year),
      x1 = max(nutrient_status_plot$year),
      y0 = 100, y1 = 100,
      line = list(color = "gray", dash = "dot")
    ))
  )

status_plotly

pressure_plotly <- plot_ly(
  nutrient_status_plot,
  x = ~year,
  y = ~pressure,
  color = ~rgn_label,
  colors = rgn_colors_labeled,
  type = "scatter",
  mode = "lines+markers",
  text = ~paste0(
    rgn_id, " (", rgn_label, ")",
    "<br>", year, ": pressure ", round(pressure, 3)
  ),
  hoverinfo = "text"
) %>%
  layout(
    title = "Eutrophication pressure (1 = at or above top-5 mean reference)",
    xaxis = list(title = "Year", dtick = 1),
    yaxis = list(title = "Pressure (0-1)", range = c(0, 1.05)),
    legend = list(title = list(text = "Region"))
  )

pressure_plotly
```

---

## 4. Write outputs

```{r write-outputs}
write_csv(
  nutrient_status,
  here::here(eut_out, "nutrient_status_full.csv")
)

write_csv(
  nutrient_status %>% select(rgn_id, year, status),
  here::here(eut_out, "nutrient_status.csv")
)

write_csv(
  nutrient_status %>% select(rgn_id, year, pressure),
  here::here(eut_out, "nutrient_pressure.csv")
)

# small reference log so we remember how the score was built
ref_log <- tibble(
  ideal = ideal,
  maximum_top5_mean = maximum,
  n_top5 = nrow(top5),
  scoring_years = paste(score_years, collapse = ";"),
  top5_detail = top5 %>%
    mutate(lab = paste0(rgn_id, " ", year, "=", round(tn_total_t_yr, 1))) %>%
    pull(lab) %>%
    paste(collapse = "; ")
)

write_csv(ref_log, here::here(eut_out, "nutrient_reference_point.csv"))
```
