---
title: 'OHI GoC eutrophication update: current status ALT (script 7)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    theme: cerulean
    highlight-style: haddock
editor_options:
  chunk_output_type: console
---

# Script 7: eutrophication current status (ALT rescaling)

Alternate scoring to `script6_eut_current_status.qmd`. Same coastline-standardized TN (`tn_t_per_km_coastline`) and same top-5 mean reference. Script 6 maps the reference to pressure = 1 (status = 0). Here we **rescale** that same pressure linearly so the reference maps to pressure = 0.5 (status = 50), and the full score range is **50 to 100** (not a hard floor that flattens mid values).

$$
TN_{\mathrm{km}} = \frac{TN_{\mathrm{region}}}{L_{\mathrm{coast,\,km}}}
$$

$$
\mathrm{pressure}_{\mathrm{raw}} = \min\left(1,\ \frac{TN_{\mathrm{km,\,current}} - TN_{\mathrm{ideal}}}{TN_{\mathrm{maximum}} - TN_{\mathrm{ideal}}}\right)
$$

$$
\mathrm{pressure} = 0.5 \times \mathrm{pressure}_{\mathrm{raw}}
$$

$$
\mathrm{status} = (1 - \mathrm{pressure}) \times 100 = 50 + 50 \times (1 - \mathrm{pressure}_{\mathrm{raw}})
$$

with $TN_{\mathrm{ideal}} = 0$ and $TN_{\mathrm{maximum}}$ = mean of the top 5 `tn_t_per_km_coastline` values. So zero TN → status 100; at the top-5 mean → status 50; halfway to the reference → status 75 (not 50). Values above the top-5 mean stay at pressure 0.5 / status 50.

**Data and methods overview:** reads the same script 5 region totals as script 6; rebuilds coastline lengths the same way. Writes parallel `*_ALT.csv` outputs so the default script 6 files stay untouched.

---

## Setup

```{r setup}
library(tidyverse)
library(here)
library(sf)
library(janitor)
library(plotly)
sf_use_s2(FALSE)

region_levels <- paste0("Region_", 1:9)

rgn_colors <- c(
  Region_1 = "#C03729",
  Region_2 = "#E68C7C",
  Region_3 = "#FC8F24",
  Region_4 = "#dcaa38",
  Region_5 = "#cbd2a0",
  Region_6 = "#98a25a",
  Region_7 = "#747428",
  Region_8 = "#627391",
  Region_9 = "#8d61a9"
)

rgn_labels <- c(
  "Region_1" = "Western Upper Gulf",
  "Region_2" = "Western Midriff Islands",
  "Region_3" = "Western Central Gulf",
  "Region_4" = "Western Southern Gulf",
  "Region_5" = "Eastern Upper Gulf",
  "Region_6" = "Eastern Midriff Islands",
  "Region_7" = "Southern Sonora Coast",
  "Region_8" = "Sinaloa Coast",
  "Region_9" = "Southeastern Gulf"
)

rgn_colors_labeled <- setNames(
  unname(rgn_colors[region_levels]),
  unname(rgn_labels[region_levels])
)

# ALT: rescale script-6 pressure so the top-5 mean reference → 0.5 (status 50-100)
pressure_at_reference <- 0.5
```

```{r paths}
eut_out <- here::here(
  "/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/eutrophication/jarg_analysis_update/output"
)
spatial_regions <- here::here("/home/shares/ohi/OHI_GOC/spatial/OHI_regions")

dir.create(eut_out, recursive = TRUE, showWarnings = FALSE)
```

---

## 1. Read script 5 totals

```{r read-tn}
tn <- read_csv(
  here::here(eut_out, "tn_total_by_region.csv"),
  show_col_types = FALSE
) %>%
  mutate(year = as.integer(year)) %>%
  arrange(year, rgn_id)

tn
```

---

## 2. Standardize by coastline length (same as script 6 / Chemicals)

```{r coastline-by-region}
coast_no_islands_path <- here::here(spatial_regions, "goc_coastline_no_islands.gpkg")
marine_and_inland10 <- st_read(
  here::here(spatial_regions, "marine_and_inland10.gpkg"),
  quiet = TRUE
)

if (file.exists(coast_no_islands_path)) {
  goc_coastline_no_islands <- st_read(coast_no_islands_path, quiet = TRUE)
} else {
  world_coastline <- st_read(
    here::here("/home/shares/ohi/OHI_GOC/spatial/ne_10m_coastline/ne_10m_coastline.shp"),
    quiet = TRUE
  )
  goc_coastline <- st_intersection(
    st_transform(world_coastline, st_crs(marine_and_inland10)),
    marine_and_inland10
  )
  islands_polygons <- st_read(
    here::here(
      "/home/shares/ohi/OHI_GOC/spatial/from_hem_POEM_ANP/232_ANP-ITRF08_16062025/232_ANP-ITRF08_16062025.shp"
    ),
    quiet = TRUE
  ) %>%
    clean_names() %>%
    filter(nombre %in% "Islas del Golfo de California") %>%
    st_transform(st_crs(goc_coastline))

  goc_coastline_no_islands <- goc_coastline %>%
    st_difference(st_union(st_buffer(islands_polygons, dist = 1500))) %>%
    st_make_valid()

  st_write(
    goc_coastline_no_islands,
    coast_no_islands_path,
    delete_dsn = TRUE,
    quiet = TRUE
  )
}

coastline_by_rgn <- goc_coastline_no_islands %>%
  st_intersection(marine_and_inland10 %>% select(rgn_id)) %>%
  mutate(segment_length_m = as.numeric(st_length(.))) %>%
  st_drop_geometry() %>%
  group_by(rgn_id) %>%
  summarize(
    coastline_length_m = sum(segment_length_m, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(coastline_length_km = coastline_length_m / 1000)

coastline_by_rgn
```

```{r standardize-tn}
tn_std <- tn %>%
  left_join(coastline_by_rgn %>% select(rgn_id, coastline_length_km), by = "rgn_id") %>%
  mutate(tn_t_per_km_coastline = tn_total_t_yr / coastline_length_km)

tn_score <- tn_std
score_years <- sort(unique(tn_score$year))
cat("scoring years:", paste(score_years, collapse = ", "), "\n")
tn_score
```

---

## 3. ALT rescaling: status 50-100 (reference → pressure 0.5)

```{r score-alt}
ideal <- 0

top5 <- tn_score %>%
  slice_max(tn_t_per_km_coastline, n = 5, with_ties = FALSE)

top5

maximum <- mean(top5$tn_t_per_km_coastline)
maximum

nutrient_status_alt <- tn_score %>%
  mutate(
    ideal = ideal,
    maximum = maximum,
    # same raw pressure as script 6 (0 at ideal, 1 at/above top-5 mean)
    pressure_raw = pmin(1, (tn_t_per_km_coastline - ideal) / (maximum - ideal)),
    # rescale so reference → 0.5; full range maps linearly to status 50-100
    pressure = pressure_at_reference * pressure_raw,
    status = (1 - pressure) * 100
  ) %>%
  arrange(year, rgn_id)

# sanity: zero → 100; at/above reference → 50; mid-reference → 75
stopifnot(min(nutrient_status_alt$status) >= 50 - 1e-8)
stopifnot(max(nutrient_status_alt$pressure) <= pressure_at_reference + 1e-8)
stopifnot(
  all(
    abs(
      nutrient_status_alt$status -
        (50 + 50 * (1 - nutrient_status_alt$pressure_raw))
    ) < 1e-8
  )
)

nutrient_status_alt %>%
  summarize(
    status_min = min(status),
    status_max = max(status),
    pressure_max = max(pressure),
    .groups = "drop"
  )

nutrient_status_alt
```

---

## 4. Quick look (plotly)

```{r plotly-status-alt}
nutrient_status_plot <- nutrient_status_alt %>%
  mutate(
    rgn_id = factor(rgn_id, levels = region_levels),
    rgn_label = factor(
      rgn_labels[as.character(rgn_id)],
      levels = unname(rgn_labels[region_levels])
    )
  )

status_plotly_alt <- plot_ly(
  nutrient_status_plot,
  x = ~year,
  y = ~status,
  color = ~rgn_label,
  colors = rgn_colors_labeled,
  type = "scatter",
  mode = "lines+markers",
  text = ~paste0(
    rgn_id, " (", rgn_label, ")",
    "<br>Year: ", year,
    "<br>TN/km: ", round(tn_t_per_km_coastline, 3),
    "<br>Pressure: ", round(pressure, 3),
    "<br>Status: ", round(status, 1)
  ),
  hoverinfo = "text"
) %>%
  layout(
    title = "Eutrophication status ALT (rescaled 50-100; top-5 mean TN/km = status 50)",
    xaxis = list(title = "Year", dtick = 1),
    yaxis = list(title = "Status (0-100)", range = c(0, 100)),
    legend = list(title = list(text = "Region")),
    shapes = list(
      list(
        type = "line",
        x0 = min(nutrient_status_plot$year),
        x1 = max(nutrient_status_plot$year),
        y0 = 100, y1 = 100,
        line = list(color = "gray", dash = "dot")
      ),
      list(
        type = "line",
        x0 = min(nutrient_status_plot$year),
        x1 = max(nutrient_status_plot$year),
        y0 = 50, y1 = 50,
        line = list(color = "gray", dash = "dash")
      )
    )
  )

status_plotly_alt

pressure_plotly_alt <- plot_ly(
  nutrient_status_plot,
  x = ~year,
  y = ~pressure,
  color = ~rgn_label,
  colors = rgn_colors_labeled,
  type = "scatter",
  mode = "lines+markers",
  text = ~paste0(
    rgn_id, " (", rgn_label, ")",
    "<br>", year, ": pressure ", round(pressure, 3)
  ),
  hoverinfo = "text"
) %>%
  layout(
    title = "Eutrophication pressure ALT (rescaled; 0.5 = at or above top-5 mean TN/km)",
    xaxis = list(title = "Year", dtick = 1),
    yaxis = list(title = "Pressure (0-1)", range = c(0, 1.05)),
    legend = list(title = list(text = "Region"))
  )

pressure_plotly_alt
```

---

## 5. Write ALT outputs

```{r write-outputs-alt}
write_csv(
  nutrient_status_alt,
  here::here(eut_out, "nutrient_status_full_ALT.csv")
)

write_csv(
  nutrient_status_alt %>% select(rgn_id, year, status),
  here::here(eut_out, "nutrient_status_ALT.csv")
)

write_csv(
  nutrient_status_alt %>% select(rgn_id, year, pressure),
  here::here(eut_out, "nutrient_pressure_ALT.csv")
)

ref_log_alt <- tibble(
  ideal = ideal,
  maximum_top5_mean_tn_per_km = maximum,
  n_top5 = nrow(top5),
  scoring_years = paste(score_years, collapse = ";"),
  metric = "tn_t_per_km_coastline",
  method = "rescale: pressure = 0.5 * pressure_raw (script6); status 50-100",
  pressure_at_reference = pressure_at_reference,
  status_at_reference = (1 - pressure_at_reference) * 100,
  top5_detail = top5 %>%
    mutate(
      lab = paste0(
        rgn_id, " ", year, "=", round(tn_t_per_km_coastline, 3), " t/km"
      )
    ) %>%
    pull(lab) %>%
    paste(collapse = "; ")
)

write_csv(ref_log_alt, here::here(eut_out, "nutrient_reference_point_ALT.csv"))
```
