---
title: 'OHI GoC eutrophication update: shrimp aquaculture nitrogen (script 4)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    theme: cerulean
    highlight-style: haddock
editor_options:
  chunk_output_type: console
---

# Script 4: shrimp aquaculture nitrogen

This script estimates nitrogen from shrimp farming following Romero-Gil et al. (2026), then routes it through watersheds and pourpoints.

For scoring later, we want **100%** of estimated shrimp-farm nitrogen to reach the sea (JARG Supplemental Table 6). Table 6 also has 50% and 85%; we use 100% because ponds discharge right to nearby waters.

Good news: Chemicals already estimated shrimp production the way we want, using pond area × 1.5 tons/ha (González-Rivas et al. 2023).

---

## What JARG did

$$
N_{\mathrm{shrimp}} = P_{\mathrm{shrimp}} \times 47~\mathrm{kg~N~per~ton} \times E_{\mathrm{export}}
$$

Romero-Gil et al. (2026) take the **47 kg N per ton of shrimp** factor from Páez-Osuna (2007). From their Aquaculture methods:

> "To estimate the nitrogen load generated by shrimp farming in the study area, we adopted the methodology developed by Páez-Osuna (2007). This approach relies on empirical measurements of nitrogen discharges from semi-intensive shrimp farms operating with two production cycles per year -- practices that are typical in the state of Sinaloa. According to Páez-Osuna, an average of 47 kg of nitrogen is released per ton of shrimp produced. This nitrogen originates primarily from uneaten feed and metabolic waste, which are discharged directly into adjacent water bodies."

They then multiply that factor by municipal production:

> "Nitrogen discharge was calculated by multiplying the nitrogen release factor (47 kg N/ton) by the average annual shrimp production for each municipality. This approach assumes that the nitrogen discharge rate remains constant across the study area, consistent with semi-intensive farming practices."

For **how much of that N reaches the Gulf**, the main text uses 100% export (Supplemental Table 6 also reports 50% and 85%). We match the main-text / Table 6 100% case:

> "Since all shrimp farms in this region are located along the coastal strip and discharge their effluents directly into estuarine or marine environments, we assumed that 100% of the nitrogen produced ultimately reaches the Gulf of California. This assumption, although conservative, is intended to represent an upper-bound scenario due to the direct hydrological connectivity of shrimp ponds to marine waters and the absence of effective attenuation processes such as sedimentation ponds or wetlands."

So in practice: production tons × 47 kg N/ton × 1.0, then convert kg → t (/ 1000).

JARG's production came from the Aquaculture Production Census (2015-2020 municipal means). We instead use Chemicals watershed production (pond area × 1.5 t/ha), then apply the same 47 and 100% export.

Paper: https://doi.org/10.2166/wpt.2026.239

---

## What we already have from CW/Chemicals 

From `chemicals/pharmaceuticals/script2_shrimp_farming.qmd`:

<!-- ```r -->
<!-- aq_production_by_watershed <- aq_area_by_watershed %>% -->
<!--   mutate( -->
<!--     shrimp_prod_tons = aq_area_ha * 1.5  # ha * tons/ha = tons -->
<!--   ) -->
<!-- ``` -->

Already written on mazu:

| File | Path |
|------|------|
| Production by watershed | `chemicals/pharmaceuticals/shrimp_int/aq_production_tons_shrimp_by_watershed.csv` |
| Production by region | `chemicals/pharmaceuticals/shrimp_int/shrimp_tons_pourpoints_prod_rgnid.csv` |
| Pourpoint gpkg | `chemicals/pharmaceuticals/shrimp_int/shrimp_tons_pourpoints_prod_final.gpkg` |
| Aquaculture USV polygons | `chemicals/pesticides/int/usv_aquaculture_goc_allyears.gpkg` |

To do this analysis, we only need the tons of shrimp produced, then × 47 kg N/ton × 1.0.

**Unit assumption:** we treat González-Rivas / Chemicals `shrimp_prod_tons` (and `shrimp_tons`) as **metric tonnes** (1 t = 1000 kg), then apply Páez-Osuna / Romero-Gil’s 47 kg N per tonne directly:

`tn_shrimp_t_yr = shrimp_prod_tons * 47 / 1000`

We do **not** apply the pharmaceuticals short-ton conversion (`* 907.18474`). That step is only for antimicrobial mg/kg. Romero-Gil’s 47 factor and Table 6 nitrogen totals are in metric terms, so assuming metric tonnes keeps the nitrogen calc consistent with JARG. If QA against Table 6 looks systematically high/low by ~10%, revisit whether González-Rivas “tons” were actually short tons.

---

## Outline of what to do here

1. Read `aq_production_tons_shrimp_by_watershed.csv`.
2. Compute `tn_shrimp_t_yr = shrimp_prod_tons * 47 / 1000` (kg → t), with export = 1.
3. Join to pourpoints / regions (or start from `shrimp_tons_pourpoints_prod_rgnid.csv` if you only need region totals).
4. Write eutrophication intermediates.

---

## Setup

```{r setup}
library(tidyverse)
library(here)
library(sf)
library(janitor)
library(mapview)
```

```{r paths-constants}
shrimp_int <- here::here(
  "/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chemicals/pharmaceuticals/shrimp_int"
)
eut_int <- here::here(
  "/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/eutrophication/jarg_analysis_update/int"
)


kg_n_per_t_shrimp <- 47  # Paez-Osuna (2007) via Romero-Gil et al. (2026): kg N released per ton shrimp
export_cw <- 1.0 # Table 6 100% export to coastal waters
```

### Read in shrimp production (again, already calculated within `CW/chemicals/pharmaceuticals/script2_shrimp_farming.qmd`)

```{r read-shrimp-prod}
aq_production_by_watershed <- read_csv(
  here::here(shrimp_int, "aq_production_tons_shrimp_by_watershed.csv")
)

glimpse(aq_production_by_watershed)
```

### Convert production to nitrogen

```{r shrimp-to-nitrogen}
tn_shrimp_by_watershed <- aq_production_by_watershed %>%
  mutate(
    tn_shrimp_t_yr = shrimp_prod_tons * kg_n_per_t_shrimp / 1000 * export_cw
  )

summary(tn_shrimp_by_watershed$tn_shrimp_t_yr)
```

### Optional: region totals from existing pourpoint table

```{r region-from-existing}
shrimp_by_region_tons <- read_csv(
  here::here(shrimp_int, "shrimp_tons_pourpoints_prod_rgnid.csv")
)

tn_shrimp_by_region <- shrimp_by_region_tons %>%
  mutate(
    tn_shrimp_t_yr = shrimp_tons * kg_n_per_t_shrimp / 1000 * export_cw
  )

tn_shrimp_by_region
```

### Write outputs

```{r write-outputs}
write_csv(tn_shrimp_by_watershed, here::here(eut_int, "tn_shrimp_by_watershed.csv"))
write_csv(tn_shrimp_by_region, here::here(eut_int, "tn_shrimp_by_region.csv"))
```

