---
title: 'OHI GoC eutrophication update: livestock nitrogen (script 3)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    theme: cerulean
    highlight-style: haddock
editor_options:
  chunk_output_type: console
---

# Script 3: livestock nitrogen

Here we estimate nitrogen from livestock following Romero-Gil et al. (2026), then allocate municipal totals to GoC watersheds (`basin_id`) using FAO **GLW4** livestock density (same spatial proxy as Chemicals pharmaceuticals). Pourpoints → OHI regions happen once in script 5.

For scoring later, we want a **40% export coefficient** (JARG Supplemental Table 6).

**Data decision:** we use SIAP municipal **heads** from the pharmaceuticals clean file  
(`siap_livestock_municipal_clean.csv`, product Carne, column `cabezas_asacrificados`).  
That matches JARG's "livestock heads from SIAP yearbooks" closely enough. We do not overwrite that Chemicals file.

---

## What JARG did

$$
N_{\mathrm{liv}} = \left(\sum \mathrm{heads}_s \times EF_s\right) \times E_{\mathrm{export}}
$$

| Species (`nomespecie`) | EF (kg N head⁻¹ year⁻¹) |
|------------------------|-------------------------|
| Bovino | 21.9 |
| Caprino | 6.3 |
| Ovino | 4.1 |
| Porcino | 4.1 |
| Ave | 0.79 |

---

## How we get to watershed TN (GLW4)

SIAP gives municipal heads (how much). A municipality can drain to several watersheds, so we need a spatial proxy for where animals sit inside the municipality.

Same rule as Chemicals pharmaceuticals `script3_livestock.qmd`:

> GLW4 (FAO livestock density) = **where**, SIAP / DGSIAP = **how much**

1. Compute municipal nitrogen by species (heads × EF × 0.40).
2. Prep each species' FAO GLW4 density raster (2020, heads/km²), crop to GoC municipalities, project to Mexico LCC, resample to 1 km (Chemicals helpers).
3. Intersect municipalities with GoC watersheds.
4. For each species, sum GLW density inside each mun ∩ basin polygon; convert to weights within the municipality (among GoC basins only).
5. Allocate species nitrogen with those weights; sum to `basin_id` × year.

If a municipality has no GLW density inside any GoC basin for a species, we fall back to intersection **area** share for that mun × species only (rare edge case).

GLW files on mazu (from Chemicals):

`/home/shares/ohi/OHI_GOC/_raw_data/FAO_livestock_density/d2026/`

| Species | File |
|---------|------|
| Bovino | `GLW4-2020.D-DA.CTL.tif` |
| Ave | `GLW4-2020.D-DA.CHK.tif` |
| Caprino | `GLW4-2020.D-DA.GTS.tif` |
| Porcino | `GLW4-2020.D-DA.PGS.tif` |
| Ovino | `GLW4-2020.D-DA.SHP.tif` |

---

## Setup

```{r setup}
library(tidyverse)
library(here)
library(sf)
library(terra)
library(janitor)
library(mapview)
sf_use_s2(FALSE)
```

```{r paths-constants}
export_cw <- 0.40

ef_kg_n_per_head_yr <- tribble(
  ~nomespecie, ~ef,
  "Bovino",  21.9,
  "Caprino",  6.3,
  "Ovino",    4.1,
  "Porcino",  4.1,
  "Ave",      0.79
)

# same species → GLW file map as Chemicals pharmaceuticals script3
species_glw_map <- tribble(
  ~nomespecie, ~glw_file,
  "Bovino",  "GLW4-2020.D-DA.CTL.tif",
  "Ave",     "GLW4-2020.D-DA.CHK.tif",
  "Caprino", "GLW4-2020.D-DA.GTS.tif",
  "Porcino", "GLW4-2020.D-DA.PGS.tif",
  "Ovino",   "GLW4-2020.D-DA.SHP.tif"
)

livestock_int <- here::here(
  "/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chemicals/pharmaceuticals/livestock_int"
)
eut_int <- here::here(
  "/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/eutrophication/jarg_analysis_update/int"
)
glw_dir <- here::here(
  "/home/shares/ohi/OHI_GOC/_raw_data/FAO_livestock_density/d2026"
)
spatial_regions <- here::here("/home/shares/ohi/OHI_GOC/spatial/OHI_regions")
spatial_admin <- here::here("/home/shares/ohi/OHI_GOC/spatial/land_admin_boundaries")

# Mexico ITRF / LCC (same target CRS Chemicals uses for GLW prep)
mexico_lcc <- "+proj=lcc +lat_1=17.5 +lat_2=29.5 +lat_0=12 +lon_0=-102 +x_0=2500000 +y_0=0 +datum=WGS84 +units=m +no_defs"

dir.create(eut_int, recursive = TRUE, showWarnings = FALSE)
```

### Read SIAP municipal heads (already written from pharmaceuticals)

```{r read-siap}
siap_municipal <- read_csv(
  here::here(livestock_int, "siap_livestock_municipal_clean.csv"),
  show_col_types = FALSE
) %>%
  clean_names() %>%
  mutate(
    anio = as.integer(anio),
    cveestado = as.numeric(cveestado),
    cvempio = as.numeric(cvempio),
    cabezas_asacrificados = as.numeric(cabezas_asacrificados)
  )

glimpse(siap_municipal)
```

### Apply emission factors and 40% export (municipal × species)

Keep species rows here so we can weight each species with its own GLW4 density map.

```{r livestock-n-species}
tn_liv_species <- siap_municipal %>%
  filter(nomespecie %in% ef_kg_n_per_head_yr$nomespecie) %>%
  left_join(ef_kg_n_per_head_yr, by = "nomespecie") %>%
  mutate(
    heads = cabezas_asacrificados,
    n_kg_gross = heads * ef,
    n_kg_export = n_kg_gross * export_cw,
    tn_livestock_t_yr = n_kg_export / 1000,
    year = anio
  )

tn_liv_species %>%
  group_by(year) %>%
  summarize(
    n_mun = n_distinct(paste(cveestado, cvempio)),
    tn_livestock_t_yr = sum(tn_livestock_t_yr, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(year)
```

---

## Build GLW4 mun × basin weights (Chemicals pattern)

### Municipality polygons (erase GoC islands, same as Chemicals)

```{r mun-polygons}
municipio_sf <- st_read(
  here::here(spatial_admin, "mex_admin2_joined_w_municipio_codigos.shp"),
  quiet = TRUE
) %>%
  clean_names() %>%
  mutate(
    cveestado = as.numeric(clv_std),
    cvempio = as.numeric(clv_mnc)
  ) %>%
  filter(cveestado %in% unique(tn_liv_species$cveestado)) %>%
  select(cveestado, cvempio, geometry) %>%
  st_make_valid()

# erase GoC islands in a metric CRS (110 m buffer). buffering in lon/lat would wipe Mexico.
islands_path <- here::here(
  "/home/shares/ohi/OHI_GOC/spatial/from_hem_POEM_ANP/232_ANP-ITRF08_16062025",
  "232_ANP-ITRF08_16062025.shp"
)

if (file.exists(islands_path)) {
  islands_polygons <- st_read(islands_path, quiet = TRUE) %>%
    clean_names() %>%
    filter(str_detect(nombre, regex("Islas del Golfo de California", ignore_case = TRUE)))

  if (nrow(islands_polygons) > 0) {
    islands_lcc <- islands_polygons %>%
      st_transform(mexico_lcc) %>%
      st_make_valid()

    municipio_sf <- municipio_sf %>%
      st_transform(mexico_lcc) %>%
      st_difference(st_union(st_buffer(islands_lcc, dist = 110))) %>%
      st_make_valid() %>%
      st_collection_extract("POLYGON", warn = FALSE)
  } else {
    message("island filter returned 0 rows; keeping municipality polygons as-is")
    municipio_sf <- st_transform(municipio_sf, mexico_lcc)
  }
} else {
  message("island shapefile not found; keeping municipality polygons as-is")
  municipio_sf <- st_transform(municipio_sf, mexico_lcc)
}

stopifnot(nrow(municipio_sf) > 0)
stopifnot(!all(st_is_empty(municipio_sf)))

# single multipolygon mask for GLW crop (WGS84 to match FAO rasters)
mun_union_geom <- municipio_sf %>%
  st_union() %>%
  st_make_valid()

stopifnot(!st_is_empty(mun_union_geom))

mun_union <- st_sf(id = 1L, geometry = st_transform(mun_union_geom, 4326))

cat(
  "mun_union area (km2):",
  as.numeric(st_area(st_transform(mun_union, mexico_lcc))) / 1e6,
  "\n"
)
```

### Prep GLW4 rasters to 1 km (Chemicals `prep_glw_1km`)

```{r prep-glw-1km}
prep_glw_1km <- function(glw_path, mask_polygon) {
  glw_raw <- terra::rast(glw_path)

  mask_sf <- mask_polygon %>%
    st_make_valid() %>%
    st_transform(terra::crs(glw_raw))

  # terra needs a SpatVector / SpatExtent, not a bare sf with empty geom
  mask_v <- terra::vect(mask_sf)
  if (terra::nrow(mask_v) < 1) {
    stop("mask SpatVector is empty after transform to GLW CRS")
  }

  glw_cropped <- terra::crop(glw_raw, mask_v)

  # project to Mexico LCC, then resample to 1 km
  glw_projected <- terra::project(glw_cropped, mexico_lcc)
  mask_lcc <- terra::vect(st_transform(mask_polygon, mexico_lcc))
  glw_masked <- terra::mask(glw_projected, mask_lcc)

  template_1km <- terra::rast(glw_masked)
  terra::res(template_1km) <- 1000
  glw_1km <- terra::resample(glw_masked, template_1km, method = "near")
  glw_1km <- terra::mask(glw_1km, mask_lcc)

  glw_1km
}

glw_1km_list <- purrr::map(
  setNames(species_glw_map$glw_file, species_glw_map$nomespecie),
  function(f) {
    message("prep GLW 1km: ", f)
    prep_glw_1km(file.path(glw_dir, f), mun_union)
  }
)
```

### Mun ∩ watershed polygons + area fallback

```{r mun-basin-polygons}
watershed_goc <- st_read(
  here::here(spatial_regions, "watersheds_goc.gpkg"),
  quiet = TRUE
) %>%
  st_make_valid()

target_crs <- mexico_lcc
mun_r <- st_transform(municipio_sf, target_crs)
ws_r <- st_transform(watershed_goc %>% select(basin_id), target_crs)

mun_ws_poly <- st_intersection(mun_r, ws_r) %>%
  st_make_valid() %>%
  mutate(area_m2 = as.numeric(st_area(.))) %>%
  filter(is.finite(area_m2), area_m2 > 0)

# area shares (fallback when GLW sum is 0 for a mun × species)
mun_ws_area <- mun_ws_poly %>%
  st_drop_geometry() %>%
  group_by(cveestado, cvempio, basin_id) %>%
  summarize(area_m2 = sum(area_m2, na.rm = TRUE), .groups = "drop") %>%
  group_by(cveestado, cvempio) %>%
  mutate(
    area_mun_goc = sum(area_m2, na.rm = TRUE),
    w_area = area_m2 / area_mun_goc
  ) %>%
  ungroup() %>%
  select(cveestado, cvempio, basin_id, w_area)

glimpse(mun_ws_area)
```

### Species-specific GLW density weights

```{r mun-basin-glw-weights}
# sum GLW cells in each mun ∩ basin polygon; weight within municipality
glw_weights_one_species <- function(glw_1km, mun_ws_poly, nomespecie) {
  mun_ws_lcc <- st_transform(mun_ws_poly, terra::crs(glw_1km))
  glw_sum <- terra::extract(
    glw_1km,
    terra::vect(mun_ws_lcc),
    fun = sum,
    na.rm = TRUE
  )

  mun_ws_lcc %>%
    st_drop_geometry() %>%
    mutate(
      nomespecie = nomespecie,
      glw_sum = glw_sum[[2]]
    ) %>%
    group_by(cveestado, cvempio, basin_id, nomespecie) %>%
    summarize(glw_sum = sum(glw_sum, na.rm = TRUE), .groups = "drop") %>%
    group_by(cveestado, cvempio, nomespecie) %>%
    mutate(
      glw_mun_goc = sum(glw_sum, na.rm = TRUE),
      w_glw = if_else(glw_mun_goc > 0, glw_sum / glw_mun_goc, NA_real_)
    ) %>%
    ungroup()
}

mun_basin_glw <- purrr::imap_dfr(
  glw_1km_list,
  function(glw, sp) {
    message("GLW mun×basin weights: ", sp)
    glw_weights_one_species(glw, mun_ws_poly, sp)
  }
)

# final weight: GLW when available, else area share
mun_basin_w <- mun_basin_glw %>%
  left_join(mun_ws_area, by = c("cveestado", "cvempio", "basin_id")) %>%
  mutate(
    w_basin = coalesce(w_glw, w_area),
    weight_source = if_else(is.finite(w_glw), "glw4", "area_fallback")
  ) %>%
  filter(is.finite(w_basin), w_basin > 0)

mun_basin_w %>%
  count(nomespecie, weight_source)

glimpse(mun_basin_w)
```

---

## Allocate species TN to watersheds

```{r allocate-to-watershed}
tn_liv_by_watershed <- tn_liv_species %>%
  inner_join(
    mun_basin_w %>% select(cveestado, cvempio, basin_id, nomespecie, w_basin, weight_source),
    by = c("cveestado", "cvempio", "nomespecie")
  ) %>%
  mutate(
    tn_livestock_t_yr = tn_livestock_t_yr * w_basin,
    n_kg_export = n_kg_export * w_basin
  ) %>%
  group_by(basin_id, year) %>%
  summarize(
    tn_livestock_t_yr = sum(tn_livestock_t_yr, na.rm = TRUE),
    n_municipalities = n_distinct(paste(cveestado, cvempio)),
    n_species = n_distinct(nomespecie),
    .groups = "drop"
  ) %>%
  arrange(year, desc(tn_livestock_t_yr))

tn_liv_by_watershed

tn_liv_by_watershed %>%
  group_by(year) %>%
  summarize(
    n_basins = n_distinct(basin_id),
    tn_livestock_t_yr = sum(tn_livestock_t_yr, na.rm = TRUE),
    .groups = "drop"
  )
```

---

## Write watershed output

Primary deliverable for script 5 is watershed TN (40% export already applied; GLW4 allocation already applied). Pourpoints and OHI regions are handled in script 5.

```{r write-outputs}
write_csv(mun_basin_w, here::here(eut_int, "livestock_mun_basin_glw4_weights.csv"))
write_csv(tn_liv_by_watershed, here::here(eut_int, "tn_livestock_by_watershed.csv"))
```
