---
title: 'OHI GoC eutrophication update: shrimp aquaculture nitrogen (script 4)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    theme: cerulean
    highlight-style: haddock
editor_options:
  chunk_output_type: console
---

# Script 4: shrimp aquaculture nitrogen

Here we estimate nitrogen from shrimp farming following Romero-Gil et al. (2026) at GoC watersheds (`basin_id`). We then use pourpoints to assign coastal watershed nitrogen output to OHI regions in script 5.

We will use a **100% export coefficient**, chosen from JARG's Supplemental Table 6 (Table 6 also has 50% and 85%; we use 100% because ponds discharge directly to nearby waters).

**Data and methods overview:** production comes from Chemicals pharmaceuticals shrimp intermediates: pond area × 1.5 t/ha (González-Rivas et al. 2023), already summarized by watershed. We apply Páez-Osuna / Romero-Gil's 47 kg N per metric tonne of shrimp, then convert to t N / year.

---

## What JARG did

$$
N_{\mathrm{shrimp}} = P_{\mathrm{shrimp}} \times 47~\mathrm{kg~N~per~ton} \times E_{\mathrm{export}}
$$

Romero-Gil et al. (2026) take the **47 kg N per ton of shrimp** factor from Páez-Osuna (2007). From their Aquaculture methods:

> "To estimate the nitrogen load generated by shrimp farming in the study area, we adopted the methodology developed by Páez-Osuna (2007). This approach relies on empirical measurements of nitrogen discharges from semi-intensive shrimp farms operating with two production cycles per year -- practices that are typical in the state of Sinaloa. According to Páez-Osuna, an average of 47 kg of nitrogen is released per ton of shrimp produced. This nitrogen originates primarily from uneaten feed and metabolic waste, which are discharged directly into adjacent water bodies."

They then multiply that factor by municipal production:

> "Nitrogen discharge was calculated by multiplying the nitrogen release factor (47 kg N/ton) by the average annual shrimp production for each municipality. This approach assumes that the nitrogen discharge rate remains constant across the study area, consistent with semi-intensive farming practices."

For **how much of that N reaches the Gulf**, the main text uses 100% export (Supplemental Table 6 also reports 50% and 85%). We match the main-text / Table 6 100% case:

> "Since all shrimp farms in this region are located along the coastal strip and discharge their effluents directly into estuarine or marine environments, we assumed that 100% of the nitrogen produced ultimately reaches the Gulf of California. This assumption, although conservative, is intended to represent an upper-bound scenario due to the direct hydrological connectivity of shrimp ponds to marine waters and the absence of effective attenuation processes such as sedimentation ponds or wetlands."

So in practice: production tons × 47 kg N/ton × 1.0, then convert kg → t (/ 1000).

JARG's production came from the Aquaculture Production Census (2015-2020 municipal means). We instead use Chemicals watershed production (pond area × 1.5 t/ha), then apply the same 47 and 100% export.

Paper: https://doi.org/10.2166/wpt.2026.239

---

## What we already have from CW (use this)

From `chemicals/pharmaceuticals/script2_shrimp_farming.qmd`:

```r
aq_production_by_watershed <- aq_area_by_watershed %>%
  mutate(
    shrimp_prod_tons = aq_area_ha * 1.5  # ha * tons/ha = tons
  )
```

Already written on mazu:

| File | Path |
|------|------|
| Production by watershed | `chemicals/pharmaceuticals/shrimp_int/aq_production_tons_shrimp_by_watershed.csv` |
| Production by region | `chemicals/pharmaceuticals/shrimp_int/shrimp_tons_pourpoints_prod_rgnid.csv` |
| Pourpoint gpkg | `chemicals/pharmaceuticals/shrimp_int/shrimp_tons_pourpoints_prod_final.gpkg` |
| Aquaculture USV polygons | `chemicals/pesticides/int/usv_aquaculture_goc_allyears.gpkg` |

We do **not** need the pharmaceutical mg columns for eutrophication. We only need tons of shrimp, then × 47 kg N/ton × 1.0.

**Unit assumption:** we treat González-Rivas / Chemicals `shrimp_prod_tons` (and `shrimp_tons`) as **metric tonnes** (1 t = 1000 kg), then apply Páez-Osuna / Romero-Gil’s 47 kg N per tonne directly:

`tn_shrimp_t_yr = shrimp_prod_tons * 47 / 1000`

We do **not** apply the pharmaceuticals short-ton conversion (`* 907.18474`). That step is only for antimicrobial mg/kg. Romero-Gil’s 47 factor and Table 6 nitrogen totals are in metric terms, so assuming metric tonnes keeps the nitrogen calc consistent with JARG. If QA against Table 6 looks systematically high/low by ~10%, revisit whether González-Rivas “tons” were actually short tons.

---

## Outline of what we do here

1. Read `aq_production_tons_shrimp_by_watershed.csv`.
2. Compute `tn_shrimp_t_yr = shrimp_prod_tons * 47 / 1000` (kg → t), with export = 1.
3. Write `tn_shrimp_by_watershed.csv` for script 5 (pourpoints / regions happen there).

---

## Setup

```{r setup}
library(tidyverse)
library(here)
library(sf)
library(janitor)
library(mapview)
```

```{r paths-constants}
shrimp_int <- here::here(
  "/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chemicals/pharmaceuticals/shrimp_int"
)
eut_int <- here::here(
  "/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/eutrophication/jarg_analysis_update/int"
)

kg_n_per_t_shrimp <- 47  # Paez-Osuna (2007) via Romero-Gil et al. (2026): kg N released per ton shrimp
export_cw <- 1.0          # main-text / Table 6 100% export to coastal waters
```

### Read shrimp production (already built)

```{r read-shrimp-prod}
aq_production_by_watershed <- read_csv(
  here::here(shrimp_int, "aq_production_tons_shrimp_by_watershed.csv")
)

glimpse(aq_production_by_watershed)
```

### Convert production to nitrogen

```{r shrimp-to-nitrogen}
tn_shrimp_by_watershed <- aq_production_by_watershed %>%
  mutate(
    tn_shrimp_t_yr = shrimp_prod_tons * kg_n_per_t_shrimp / 1000 * export_cw
  )

summary(tn_shrimp_by_watershed$tn_shrimp_t_yr)
```

### Write watershed output

Primary deliverable for script 5 is watershed TN (47 kg N/t × 100% export already applied). Pourpoints and OHI regions are handled in script 5.

```{r write-outputs}
write_csv(tn_shrimp_by_watershed, here::here(eut_int, "tn_shrimp_by_watershed.csv"))
```

---

## QA: pourpoints vs pond polygons (are we losing coastal shrimp N?)

Ponds are already inside GoC watersheds (script 4 only reads the watershed production table). Any gulf-wide gap between watershed TN and pourpoint→region TN is from the **pourpoint delivery filter** (50 km inland buffer; Pacific-side peninsula pourpoints outside 10 km), not from ponds sitting outside coastal watersheds.

This chunk maps USV aquaculture polygons, GoC pourpoints, and flags basins that have shrimp TN but do not survive the Chemicals shrimp pourpoint filter (same 50 km / Region 4 rule as `script2_shrimp_farming.qmd`).

```{r qa-shrimp-pourpoint-mapview}
library(sf)
sf_use_s2(FALSE)

watershed_goc <- st_read(here::here(spatial_regions, "watersheds_goc.gpkg")) %>%
  st_make_valid()

regions_ohi <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp"))

spatial_regions <- here::here("/home/shares/ohi/OHI_GOC/spatial/OHI_regions")
pesticides_int <- here::here(
  "/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/chemicals/pesticides/int"
)

pourpoints_goc <- st_read(
  here::here(spatial_regions, "pourpoints_goc.gpkg"),
  quiet = TRUE
)
marine_and_inland50 <- st_read(
  here::here(spatial_regions, "marine_and_inland50.gpkg"),
  quiet = TRUE
)
marine_and_inland10 <- st_read(
  here::here(spatial_regions, "marine_and_inland10.gpkg"),
  quiet = TRUE
)

# USV aquaculture (same source Chemicals used for pond area)
usv_aq_path <- here::here(pesticides_int, "usv_aquaculture_goc_allyears.gpkg")
usv_aq <- st_read(usv_aq_path, quiet = TRUE) %>%
  clean_names()

# prefer a recent survey year for the map
aq_years <- sort(unique(as.integer(usv_aq$year)))
map_year <- if (2021L %in% aq_years) 2021L else max(aq_years)
usv_aq_map <- usv_aq %>% filter(as.integer(year) == map_year)

tn_year <- tn_shrimp_by_watershed %>%
  filter(year == map_year) %>%
  select(basin_id, tn_shrimp_t_yr, shrimp_prod_tons, aq_area_ha)

# join watershed TN to pourpoints; apply same Chemicals shrimp spatial filter
pour_shrimp <- pourpoints_goc %>%
  select(basin_id) %>%
  left_join(tn_year, by = "basin_id") %>%
  filter(!is.na(tn_shrimp_t_yr), tn_shrimp_t_yr > 0) %>%
  st_transform(st_crs(marine_and_inland50)) %>%
  st_intersection(marine_and_inland50) %>%
  mutate(
    is_within_10km = lengths(
      st_within(., st_transform(marine_and_inland10, st_crs(.)))
    ) > 0
  ) %>%
  # Chemicals shrimp filter: drop Pacific-side peninsula pourpoints outside 10 km for Region 4
  filter(!(is_within_10km == FALSE & rgn_id == "Region_4"))

basins_kept <- unique(pour_shrimp$basin_id)
basins_all <- unique(tn_year$basin_id)
basins_lost <- setdiff(basins_all, basins_kept)

tn_lost <- tn_year %>%
  filter(basin_id %in% basins_lost) %>%
  summarize(
    n_basins_lost = n(),
    shrimp_prod_tons = sum(shrimp_prod_tons, na.rm = TRUE),
    tn_shrimp_t_yr = sum(tn_shrimp_t_yr, na.rm = TRUE),
    .groups = "drop"
  )

tn_kept <- tn_year %>%
  filter(basin_id %in% basins_kept) %>%
  summarize(
    n_basins_kept = n(),
    tn_shrimp_t_yr = sum(tn_shrimp_t_yr, na.rm = TRUE),
    .groups = "drop"
  )

cat(
  "map year:", map_year, "\n",
  "watershed basins with shrimp:", length(basins_all), "\n",
  "basins kept after pourpoint filter:", tn_kept$n_basins_kept, "\n",
  "basins lost:", tn_lost$n_basins_lost, "\n",
  "TN kept (t N/yr):", round(tn_kept$tn_shrimp_t_yr, 2), "\n",
  "TN lost (t N/yr):", round(tn_lost$tn_shrimp_t_yr, 2), "\n",
  "% retained:", round(100 * tn_kept$tn_shrimp_t_yr /
    (tn_kept$tn_shrimp_t_yr + tn_lost$tn_shrimp_t_yr), 2), "\n"
)

tn_year %>%
  filter(basin_id %in% basins_lost) %>%
  arrange(desc(tn_shrimp_t_yr)) %>%
  print(n = 20)

# pourpoints that never get shrimp TN (no join or filtered out) vs kept
pour_all_with_tn <- pourpoints_goc %>%
  select(basin_id) %>%
  left_join(tn_year, by = "basin_id") %>%
  mutate(has_shrimp = !is.na(tn_shrimp_t_yr) & tn_shrimp_t_yr > 0)

pour_lost_basins <- pour_all_with_tn %>%
  filter(basin_id %in% basins_lost)

mapview(
  usv_aq_map,
  col.regions = "darkorange",
  alpha.regions = 0.55,
  layer.name = paste0("USV aquaculture ", map_year)
) +
  mapview(
    pour_shrimp,
    zcol = "tn_shrimp_t_yr",
    layer.name = "Pourpoints kept (TN)"
  ) +
  mapview(
    pour_lost_basins,
    col.regions = "red",
    layer.name = "Pourpoints in lost basins (filtered or no coastal pourpoint)"
  ) +
  mapview(
    watershed_goc,
    col.regions = "blue",
    layer.name = "Watersheds") +
  mapview(
    regions_ohi,
    col.regions = "pink",
    layer.name = "OHI Regions")
```
