---
title: 'OHI GoC 2025: Clean waters pesticides'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# CW: pesticides

## Land Use Prep: automating the reading in of each land use shapefile for each year and limiting to the GoC.

Data: CONABIO Land Use & Vegetation Series <http://www.conabio.gob.mx/informacion/gis/?vns=gis_root/usv/inegi/usv250s7gw>

This script filters the shapefiles to agriculture and aquaculture within GOC OHI Regions.  Note that each folder represents a different year of data.

usv250kcs1agw = 1980-91
usv250ks2gw = 2001
usv250ks3gw = 2005
usv250ks4gw = 2009
usv250s5ugw = 2013
usv250s6gw = 2016
usv250s7gw = 2021

```{r setup}
library(sf)
library(utils)
library(tidyverse)
library(here)
library(raster)
library(mapview)
library(purrr)
library(stringr) # fuzzy string matching
library(stringi)
library(janitor)
library(tibble)
```


```{r step_1}
#  ---------------- step 1: load in boundary ----------------
# goc ohi regions shapefile, used to clip all data
goc_ohi_rgns <- st_read(here("spatial/ohi_regions", "goc_ohi_rgns.shp"))
```


```{r step_2}
# ---------------- step 2: save the folder names that are each associated with a year  ---------------- 
# each folder corresponds to a different survey year
usv_years <- tibble::tribble(
  ~folder,           ~year,
  "usv250kcs1agw",   "1985", # 1980-1991 survey, using midpoint year
  "usv250ks2gw",     "2001",
  "usv250ks3gw",     "2005",
  "usv250ks4gw",     "2009",
  "usv250s5ugw",     "2013",
  "usv250s6gw",      "2016",
  "usv250s7gw",      "2021"
)

# root directory for all raw USV shapefiles on mazu
raw_dir <- "/home/shares/ohi/OHI_GOC/_raw_data/CANABIO_Saltmarsh_landuse_veg" # ask Austin to change the name to CONABIO
```


```{r step_3}
#  ---------------- step 3: using fuzzy string matching to limit to ag and aquaculture ----------------

# the fxn returns TRUE for any DESCRIPCIO value that contains "AGRICULT" (catches all agriculture observations) or "ACUICUL" (aquaculture)
# `stri_trans_general` strips accents first (ex:. Ú = U, Í = I) so "ACUÍCOLA" matches "ACUICUL", avoids missed matches due to encoding differences

is_agriculture <- function(x) {
  str_detect(
    str_to_upper( # convert to uppercase for matching
      stri_trans_general(x, "Latin-ASCII") # remove accent characters before matching
    ),
    "AGRICULT|ACUICO" # | = or; matches any agriculture type or aquaculture
  )
}
```


```{r step_4}
# ---------------- step 4: make fxn to read it in, filter to ag/aq, and clip to GOC  ----------------

# takes a folder name and year label, reads the shapefile, standardizes column names, filters to agriculture/aquaculture, reprojects and clips to the GOC boundary, and tags each row with its respective year

read_usv <- function(folder, year) {
  
  # path to each folder
  shp_path <- file.path(raw_dir, folder, paste0(folder, ".shp"))
  
  message("reading: ", folder, " (", year, ")")
  
  # read shapefile
  shp <- st_read(shp_path) %>%
    clean_names() # standardize all column names
  
  # keep only rows where descripcio matches agriculture or aquaculture patterns
  agr <- shp %>%
    filter(is_agriculture(descripcio))
  
  message("  = ", nrow(agr), " agriculture/aquaculture kept")
  
  # if it did not match with anything, skip that year...
  if (nrow(agr) == 0) return(NULL)
  
  # reproject to match GOC regions CRS before clipping
  agr_goc <- agr %>%
    st_transform(st_crs(goc_ohi_rgns)) %>%    
   
  # intersect w goc (right now, not limited to a 50km buffer from the coast, includes the entire state.)                            
    st_intersection(goc_ohi_rgns) %>%         
    mutate(
      year = year,                       
      folder = folder 
    )
  
  message("  = ", nrow(agr_goc), " clipped successfully")
  
  return(agr_goc)
}
```


```{r}
# ---------------- step 5: actually use the fxn  ----------------

# map2() applies read_usv() to each folder and year pair (in parallel!).  It returns a list of sf objects (one per year), and NULLs where files were skipped

usv_list <- map2(
  usv_years$folder,
  usv_years$year,
  read_usv
)
```


```{r}
# ---------------- step 6: keep all years and make it into one sf object ----------------

usv_all <- bind_rows(usv_list)

# before saving, check it out!!
message("Total features across all years: ", nrow(usv_all))
message("Years present: ", paste(unique(usv_all$year), collapse = ", "))
message("Land use types found:\n",
        paste(sort(unique(usv_all$descripcio)), collapse = "\n"))

usv_ag <- usv_all %>% 
  filter(!descripcio %in% "ACUÍCOLA")

usv_aq <- usv_all %>% 
  filter(descripcio %in% "ACUÍCOLA")
```


```{r}
# ---------------- step 7: save the output to mazu ----------------

# saving as .gpkg to handle long field names and special characters in case there are still some??

# save all
st_write(
  usv_all,
  here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/pesticides/int", "usv_agriculture_and_aquaculture_goc_allyears.gpkg"),
  delete_dsn = TRUE # overwrite if file already exists
)

# save only ag
st_write(
  usv_ag,
  here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/pesticides/int", "usv_agriculture_goc_allyears.gpkg"),
  delete_dsn = TRUE # overwrite if file already exists
)

# save only aquaculture
st_write(
  usv_aq,
  here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/pesticides/int", "usv_aquaculture_goc_allyears.gpkg"),
  delete_dsn = TRUE # overwrite if file already exists
)


```

