################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```

## clean_names_lowercase function
```{r Load clean_names_lowercase function}
# Takes text and standardizes it, which we will need for merging later on:

clean_names_lowercase <- function(x) {

  x <- iconv(x, from = "", to = "UTF-8", sub = "")
  x <- stringi::stri_trans_general(x, "Latin-ASCII")

  x %>%
    tolower() %>%
    trimws()
}
```


# STEP 1: Generate GOC_Stocks_summary ------------------------------

# Load in and summarize GOC_MarineLandings0
```{r Load in GOC_MarineLandings0 as necessary}
GOC_MarineLandings0<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Marine.Landings_Sector.Sorted.Georef_PATCHED_Jun032026.rds"))
# 3246583 obs., 58 var.

# NOTE TO SELF: This source file may change in later Summer / Fall run. - EMF, June 4 2026
```

# **DECISION POINT: Selecting which sectors to include**
```{r Select Sectors_of_Interest}
table(GOC_MarineLandings0$sector_estimate_draft)

# SANITY CHECK: Which species are included in the freshwater or marine sector?
TEST<- GOC_MarineLandings0 %>%
  filter(sector_estimate_draft== "freshwater or marine: fishery sector yet to be classified")

table(TEST$recorded_taxa_group)
# Species like Mojarra (the overwhelming majority of catch), trout, and catfish, which can move from freshwater to marine systems and are a bit ambiguous in terms of where they can be caught.


MarineFishery_Sectors<- c("marine: fishery sector yet to be classified",
                          "marine: small-scale capture",
                          "marine: large-scale capture")
  
PossibleMarineFishery_Sectors<- c("freshwater or marine: large-scale capture",
                          "freshwater or marine: small-scale capture",
                          "freshwater or marine: fishery sector yet to be classified")

GOC_MarineFisheryLandings0<- GOC_MarineLandings0 %>%
  filter(sector_estimate_draft %in% c(MarineFishery_Sectors, PossibleMarineFishery_Sectors))

# 3137044 obs., 58 var.
```

```{r Clean up names of GOC_MarineFisheryLandings}
GOC_MarineFisheryLandings <- GOC_MarineFisheryLandings0 %>%
  mutate(
    OHI_region_est = clean_names_lowercase(OHI_region_est),
    estimated_genus_name = clean_names_lowercase(estimated_genus_name),
    Major_Functional_Group = clean_names_lowercase(Major_Functional_Group)
  )

# 3137044 obs., 58 var.
```

```{r Generate GOC_Stocks_summary}
colnames(GOC_MarineFisheryLandings)

GOC_Stocks_summary<- GOC_MarineFisheryLandings %>%
  select(year, 
         state, office_name_clean, landing_site_clean,
         GOC_est, GOC_Include_Exclude, OHI_region_est, OHI_region_name,
         eu_rnpa, eu_name, sector_estimate_draft, 
         Major_Functional_Group, recorded_taxa_group, estimated_taxa_group,
         recorded_taxa_name, recorded_taxa_name_clean,
         estimated_species_name, estimated_genus_name,
         landed_weight, live_weight, weight_kg_corrected, value
         )

GOC_Stocks_summary<- GOC_Stocks_summary %>%
  group_by(year, OHI_region_est, estimated_genus_name) %>%
  summarize(
    total_weight_mt= sum(weight_kg_corrected, na.rm = TRUE) / 1000,
    total_value_peso= sum(value, na.rm = TRUE)
  )
# 12696 obs., 5 vars.
```


# STEP 2: Load and merge the various FIS scores ------------------------------

```{r Load the indicator dataframes as necessary}
RAM_Scores0<- read.csv("/home/ferrer/ohi_goc_prep/FP/FP_Scores_1.0/FIS_Scores_1.0/FIS_Stock_Status_SubScores/FIS_RAM_Scores/FIS_RAM_Scores_Jun032026.csv")
# 216 obs., 14 vars.

Carta_Scores0<- read.csv("/home/ferrer/ohi_goc_prep/FP/FP_Scores_1.0/FIS_Scores_1.0/FIS_Stock_Status_SubScores/FIS_Carta_Scores/FIS_CartaScores_Jun052026.csv")
# 1480 obs., 17 vars.

CMSY_Scores0<- read.csv("/home/ferrer/ohi_goc_prep/FP/FP_Scores_1.0/FIS_Scores_1.0/FIS_Stock_Status_SubScores/FIS_CMSY_Scores/FIS_CMSY_Scores_Jun012026.csv")
# 1569 obs., 11 vars.
```

## RAM Data
```{r Augment the RAM dataframe by collapsing all stock info into one score per genus per year}
colnames(RAM_Scores0)

RAM_Scores <- RAM_Scores0 %>%
  group_by(year, Genus) %>%
  summarize(
    RAM_BBmsy = mean(RAM_BBmsy_final, na.rm = TRUE),
    n_Formal_Assessments = sum(!is.na(RAM_BBmsy_numeric)),
    .groups = "drop"
  ) %>%
  mutate(
    RAM_provenance = case_when(
      n_Formal_Assessments == 0 ~ "value was interpolated based on formal assessment from other years (RAM or other*)",
      n_Formal_Assessments == 1 ~ "value was sourced from 1 formal assessment (RAM or other*)",
      n_Formal_Assessments >= 2 ~ "value was sourced from 2 or more formal assessments (RAM or other*)"
    )
  )
# 168 obs., 5 var.
# The total number of observations is reduced because we are meaning across species to the genus level.

# lowercase to make sure matching works
RAM_Scores <- RAM_Scores %>%
  mutate(estimated_genus_name = tolower(Genus),
         Stock_Status_RAM= RAM_BBmsy) %>%
  select(-c(Genus, RAM_BBmsy))
# 168 obs., 6 var. 

# Add 0 helpful columns for merging:
# RAM_Scores$StockStatus_source<- c("formal assessment (RAM or other*)")
# RAM_Scores$StockStatus_est.level<- c("genus level, across the GOC")
# RAM_Scores$OHI_region_est<- c("GOC")
```

```{r Rename the Stock_Status column in Carta_Scores}
Carta_Scores<- Carta_Scores0 %>%
  mutate(
    Stock_Status_Carta= Stock_Status_final
  ) %>%
  select(-c(Stock_Status_final))

# Add 0 helpful column for merging:
# Carta_Scores$StockStatus_source<- c("Carta")
# Carta_Scores$StockStatus_est.level<- c("genus-region level")
```

```{r Rename the Stock_Status column in Stock_Status_CMSY}
CMSY_Scores<- CMSY_Scores0 %>%
  mutate(
    Stock_Status_CMSY= B.Bmsy
  ) %>%
  rename(year= yr)

# Add 1 helpful columns for merging:
# CMSY_Scores$StockStatus_source<- c("CMSY")
# CMSY_Scores$StockStatus_est.level<- c("genus level, across the GOC")
# CMSY_Scores$OHI_region_est<- c("GOC")
CMSY_Scores$CMSY_provenance<- c("CMSY model run from May 28")
```

# Join dataframes together the long way:

```{r Create a backbone to merge to and then add}
# Join in RAM data, merge by year x genus:
Stock_Statuses0 <- GOC_Stocks_summary %>%
  left_join(RAM_Scores, by = c("year", "estimated_genus_name"))
# 12696 obs.

# Join in Carta data, merge is more specific by year x genus x region:
colnames(Carta_Scores)

Stock_Statuses00 <- Stock_Statuses0 %>%
  left_join(Carta_Scores, by = c("year", "estimated_genus_name", "OHI_region_est"))
# 12696 obs.

# Join in CMSY data, merge is more specific by year x genus x region:
colnames(CMSY_Scores)

Stock_Statuses000 <- Stock_Statuses00 %>%
  left_join(CMSY_Scores, by = c("year", "estimated_genus_name"))
# 12696 obs., 34 vars.
```

```{r Clean up this dataframe}
colnames(Stock_Statuses000)

Stock_Statuses<- Stock_Statuses000 %>%
  select(year, Major_Functional_Group, estimated_genus_name, estimated_family_name,
         OHI_region_est,
         total_weight_mt, total_value_peso,
         Stock_Status_RAM, Stock_Status_Carta, Stock_Status_CMSY,
         Carta_provenance= carta_provenance, RAM_provenance
         ) %>%
  mutate(Stock_Status_RAM_capped = pmin(Stock_Status_RAM, 1),
         Stock_Status_Carta_capped = pmin(Stock_Status_Carta, 1), 
         Stock_Status_CMSY_capped = pmin(Stock_Status_CMSY, 1))

# 12696 obs., 15 vars.
```

# **DECISION POINT: Pick 1 Stock Status**
```{r Pick one StockStatus_final}
# A hierarchy of which score to pick!
Stock_Statuses <- Stock_Statuses %>%
  mutate(
    StockStatus_final = coalesce(
      Stock_Status_RAM_capped,
      Stock_Status_Carta_capped,
      Stock_Status_CMSY_capped
    ),
    StockStatus_source = case_when(
      !is.na(Stock_Status_RAM_capped)   ~ "RAM",
      !is.na(Stock_Status_Carta_capped) ~ "Carta",
      !is.na(Stock_Status_CMSY_capped)  ~ "CMSY",
      TRUE ~ NA_character_
    ),
    StockStatus_est.level = case_when(
      !is.na(Stock_Status_RAM_capped)   ~ "genus level",
      !is.na(Stock_Status_Carta_capped) ~ "genus x region level",
      !is.na(Stock_Status_CMSY_capped)  ~ "genus level",
      TRUE ~ NA_character_
    )
  )

# Still some NAs in here. Let's proceed.
```

```{r How much is left unassigned at the stock level?}
unique(Stock_Statuses$estimated_genus_name)
# 196 unique genera in landings

Stock_Statuses %>%
  filter(year >= 2018 & year <= 2024) %>%
  mutate(
    in_filter = is.na(StockStatus_final) |
                is.na(OHI_region_est)
  ) %>%
  group_by(year) %>%
  summarise(
    total_mass_mt = sum(total_weight_mt, na.rm = TRUE),
    mass_in_filter = sum(total_weight_mt[in_filter], na.rm = TRUE),
    mass_not_in_filter = sum(total_weight_mt[!in_filter], na.rm = TRUE),
    percent_unaccounted = (mass_in_filter/total_mass_mt)*100
  )

# Less than ~32% of total biomass unaccounted for by Stock Status in any given year.
# Estimates seem poorer in 2023 and 2024 (this is because genus estimates are poorer these years)...
```

```{r Save this dataframe as needed}
# write.csv(Stock_Statuses, 
#           "/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/Stock_Statuses_Compiled_df_Jun072026.csv", 
#           row.names = FALSE)
```

# STEP 3: Calculate GEOMETRIC MEAN of stock landings * Stock_Status ------------------------------

```{r Load in this dataframe as needed}
Stock_Statuses<- read.csv("/home/ferrer/ohi_goc_prep/FP/FP_Scores_1.1/FIS_Scores_1.1/copy_Stock_Statuses_Compiled_df_Jun072026.csv")
```


```{r Weights and geometric mean}
# Calculate the proportion of catch each "Stock" (i.e. genus) contributed to its regional total during the study period:
colnames(Stock_Statuses)

StudyPeriod_low= 2018
StudyPeriod_high= 2024

Weights_data <- GOC_Stocks_summary %>%
  filter(year >= StudyPeriod_low & year <= StudyPeriod_high) %>%
  group_by(estimated_genus_name, OHI_region_est) %>%
  mutate(StockCatch_mt = sum(total_weight_mt)) %>% # total tonnes across all years for each stock and region during the study period
  ungroup() %>%
  group_by(OHI_region_est) %>%
  mutate(TotalCatch_mt = sum(total_weight_mt, na.rm = TRUE)) %>% # total tonnes catch across all years and stocks for each region
  ungroup() %>%
  mutate(wprop_2018.24 = StockCatch_mt / TotalCatch_mt) %>% # proportion of catch each stock contributed to the total catch
  distinct(estimated_genus_name, OHI_region_est, wprop_2018.24)

# NOTE TO SELF: see point about geometric weights. At present, the weights reflect the relative contribution per stock to total FOFM catch (per region) DURING the 2018–2024 period. Verify with Mel that this is correct. - EMF, Sep 5 2026.

# SANITY CHECK: Do the weights add up to 1?
TEST<- Weights_data %>%
  group_by(OHI_region_est) %>%
  summarize(sum_w = sum(wprop_2018.24, na.rm= TRUE))
# Yup, looks good!
```

```{r Add weights data to Stock Status df:}
colnames(Weights_data)

StockStatus_Scores_finalized <- left_join(Stock_Statuses, 
                                      Weights_data, 
                                      by = c("OHI_region_est", "estimated_genus_name"))


# This is the geometric mean part:
FIS_Scores <- StockStatus_Scores_finalized %>%
  group_by(OHI_region_est, year) %>%
  summarize(
    GeoMean_StockStatus = prod(StockStatus_final ^ wprop_2018.24, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(StockStatus_SubScore= GeoMean_StockStatus * 100)
```

```{r Save this dataframe}
# write.csv(FIS_Scores,
#           "/home/ferrer/ohi_goc_prep/FP/FP_Scores_1.1/FIS_Scores_1.1/copy_FIS_StockStatus_SubScores_df_Jun052026.csv",
#           row.names = FALSE)
```

# >> COMPLETE: FIS_Scores!

```{r Load in FIS_Scores0 as needed}
FIS_Scores0<- read.csv("/home/ferrer/ohi_goc_prep/FP/FP_Scores_1.1/FIS_Scores_1.1/copy_FIS_StockStatus_SubScores_df_Jun052026.csv")

FIS_Scores<- FIS_Scores0
```


# PLOT: Stock_Status Sub-Score (Geometric Mean) over time
```{r Plot Stock Status Geo Mean over time}
# Region Colors:
Region_colors <-  c(
"region_1" = "#C03729",
"region_2" = "#E68C7C",
"region_3" = "#FC8F24",
"region_4" = "#dcaa38",
"region_5" = "#cbd2a0",
"region_6" = "#98a25a",
"region_7" = "#747428",
"region_8" = "#627391",
"region_9" = "#8d61a9"
)

Regions_of_Interest <- c(
  "region_1", "region_2", "region_3",
  "region_4", "region_5", "region_6",
  "region_7", "region_8", "region_9"
)

FIS_Scores_overTime <- FIS_Scores %>%
  filter(OHI_region_est %in% Regions_of_Interest,
         year>=2018 & year <= 2024) %>%
  ggplot(aes(
    x = year,
    y = StockStatus_SubScore,
    color = OHI_region_est
  )) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ OHI_region_est, scales = "fixed") +
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = Region_colors) +
  labs(
    title = expression(bold("FIS: Regional " * bolditalic("Stock_Status") * " Sub-Scores (Geometric Mean)")),
    y = "Stock_Status Sub-Score (from 0 to 1)",
    x = "Year"
  ) +
  ylim(0, 101) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 14, face = "bold")
  )

FIS_Scores_overTime
```

```{r Save this plot}
# Save as PNG
ggsave(
    filename = sprintf("/home/ferrer/ohi_goc_prep/FP/FP_Scores_1.1/FIS_Scores_1.1/FIS_Scores_1.1_plot_Jul282026.png"),
    plot = FIS_Scores_overTime,
    width = 12,
    height = 12,
    dpi = 300
  )
```

# Closer look into the scores of Region 1:
```{r Whats happening in Region 1 lately?}
colnames(Stock_Statuses)

Region1_Look <- Stock_Statuses %>%
  filter(year>= StudyPeriod_low & year<= StudyPeriod_high, 
         OHI_region_est == "region_1") %>%
  group_by(year, Major_Functional_Group) %>%
  summarize(
    mean_StockStatus = mean(StockStatus_final, na.rm = TRUE),
    mode_StockStatus_est.level =
      names(which.max(table(na.omit(StockStatus_source)))),
    .groups = "drop"
  )
# Looks like the change in small pelagics might be driving the changes, and those changes aren't only reflected in the CMSY.

ggplot(Region1_Look,
       aes(x = year,
           y = mean_StockStatus,
           color = mode_StockStatus_est.level,
           group = Major_Functional_Group)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  
  facet_wrap(~ Major_Functional_Group, scales = "free_y") +
  
  scale_color_brewer(palette = "Dark2") +
  
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.2)
  ) +
  
  labs(
    x = "Year",
    y = "Mean Stock Status",
    color = "Average assessment level",
    title = "Region 1: Mean Stock Status through Time by Functional Group",
    subtitle = "Diagnosing decline in FIS"
  ) +
  theme_bw() +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

```{r Let's continue this close look}
Region1_CloserLook <- Stock_Statuses %>%
  filter(year>= StudyPeriod_low & year<= StudyPeriod_high, 
         OHI_region_est == "region_1",
         Major_Functional_Group == "Pelagic marine fish (excl. anchovy & sardine)") %>%
  group_by(year, estimated_genus_name) %>%
  summarize(
    mean_StockStatus = mean(StockStatus_final, na.rm = TRUE),
    mode_StockStatus_est.level =
      names(which.max(table(na.omit(StockStatus_source)))),
    .groups = "drop"
  )

ggplot(Region1_CloserLook,
       aes(x = year,
           y = mean_StockStatus,
           color = mode_StockStatus_est.level,
           group = estimated_genus_name)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  
  facet_wrap(~ estimated_genus_name, scales = "free_y") +
  
  scale_color_brewer(palette = "Dark2") +
  
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.2)
  ) +
  
  labs(
    x = "Year",
    y = "Mean Stock Status",
    color = "Average assessment level",
    title = "Region 1: Mean Stock Status through Time by Functional Group",
    subtitle = "Diagnosing decline in FIS"
  ) +
  theme_bw() +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "bottom"
  )

# Here we go. Catch for scomberomorus has been incredibly high in recent years.

Region1_CloserLook <- Stock_Statuses %>%
    filter(year>= StudyPeriod_low & year<= StudyPeriod_high, 
         OHI_region_est == "region_1",
         estimated_genus_name == "scomberomorus")

# This decline could be real, frankly. Look at the declining Carta and CMSY scores.
```
