---
title: "OHI GoC CCE: DENUE marine NGO prep (10 km and 50 km)"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    theme: cerulean
editor:
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options:
  chunk_output_type: console
---

# 0. Workflow

This script follows the same DENUE download pattern as `tr_quantity_forexample/script1_denue_prep.qmd`:

1. Read **annual DENUE CSV snapshots** (November for 2019-2024; May for 2025-2026).
2. Filter to the six GoC states.
3. Keep **nonprofit / civic organization** SCIAN classes (see below).
4. Optionally require **marine / coastal keywords** in establishment names.
5. Clip to **10 km** and **50 km** coastal buffers and write both geopackages.

**Important:** DENUE does not have a field called "nonprofit." We use SCIAN activity codes under **sector 81, subsector 813**. In Mexico the main civil-society class is **`813230` — Asociaciones y organizaciones civiles** (not NAICS `813410`, which is US/Canada only).

Political (`813220`) and religious (`813210`) organizations are excluded. Professional (`813130`) and producer-chamber (`813110`) classes are optional and flagged in the exploration tables.

**Currently downloaded:** May 2026 only, split like sector 72:

- Otros servicios excepto actividades gubernamentales (**1 de 2**)
- Otros servicios excepto actividades gubernamentales (**2 de 2**)

Path on mazu:

`/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/otros_servicios_excepto_actividades_gubernamentales/`

INEGI unzip names (no month suffix until renamed):

- `denue_00_81_1_csv/conjunto_de_datos/denue_inegi_81_1.csv`
- `denue_00_81_2_csv/conjunto_de_datos/denue_inegi_81_2.csv`

If later years are renamed the TR way (`denue_00_81_1_0526_csv`, etc.), the helpers below will find those too.

Run **Section 3** first: it lists every `nombre_act` in this sector 81 snapshot.

```{r}
library(tidyverse)
library(sf)
library(here)
library(janitor)

sf::sf_use_s2(FALSE)

# Only May 2026 sector 81 is downloaded so far.
series_years <- 2026L

denue_base_path <- "/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025"
ngo_sector <- "otros_servicios_excepto_actividades_gubernamentales"

if (!dir.exists(file.path(denue_base_path, ngo_sector))) {
  local_copy <- here::here(ngo_sector)
  if (dir.exists(local_copy)) {
    message("Shares path missing; using local copy: ", local_copy)
    denue_base_path <- here::here()
  }
}

int_dir <- here::here(
  "/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026/int/marine_ngos"
)
dir.create(int_dir, recursive = TRUE, showWarnings = FALSE)

out_10km_gpkg <- here::here(int_dir, "marine_ngos_2019_2026_goc_10km.gpkg")
out_50km_gpkg <- here::here(int_dir, "marine_ngos_2019_2026_goc_50km.gpkg")
out_explore_csv <- here::here(int_dir, "marine_ngo_denue_activity_exploration.csv")
out_nombre_act_csv <- here::here(int_dir, "denue_81_2026may_nombre_act_inventory.csv")
out_list_10km_csv <- here::here(int_dir, "marine_ngo_unit_list.csv")
out_list_50km_csv <- here::here(int_dir, "marine_ngo_unit_list_50km.csv")

# strict | medium | broad  (see filter section)
ngo_filter_tier <- "medium"
```

# 1. Shared helpers (same conventions as script1_denue_prep.qmd)

```{r}
goc_states <- c(
  "baja california", "baja california sur", "jalisco",
  "nayarit", "sinaloa", "sonora"
)

get_folder_suffix <- function(year) {
  suffixes <- list(
    "2016" = "1016",
    "2017" = "1117",
    "2018" = "1118",
    "2019" = "1119",
    "2020" = "1120",
    "2021" = "1121",
    "2022" = "1122",
    "2023" = "1123",
    "2024" = "1124",
    "2025" = "0525",
    "2026" = "0526"
  )
  suffixes[[as.character(year)]]
}

fix_denue_text <- function(x) {
  x %>%
    as.character() %>%
    str_replace_all("<fa>", "ú") %>%
    str_replace_all("<e1>", "á") %>%
    str_replace_all("<ed>", "í") %>%
    str_replace_all("<f3>", "ó") %>%
    str_replace_all("<f1>", "ñ") %>%
    str_replace_all("<e9>", "é")
}

clean_municipio <- function(municipio_raw) {
  municipio_fixed <- fix_denue_text(municipio_raw)
  municipio_lower <- municipio_fixed %>%
    str_to_lower() %>%
    str_replace_all(" ", "_")
  case_when(
    municipio_fixed == "San Ignacio Río Muerto" ~ "cajeme",
    municipio_fixed == "Bácum" ~ "cajeme",
    municipio_fixed == "Mexicali" ~ "san_felipe",
    municipio_fixed == "Sinaloa" ~ "guasave",
    municipio_fixed == "Santa María del Oro" ~ "san_pedro_lagunillas",
    TRUE ~ municipio_lower
  )
}

employees_from_per_ocu <- function(per_ocu) {
  case_when(
    per_ocu == "0 a 5 personas" ~ 3,
    per_ocu == "6 a 10 personas" ~ 8,
    per_ocu == "11 a 30 personas" ~ 20.5,
    per_ocu == "31 a 50 personas" ~ 40.5,
    per_ocu == "51 a 100 personas" ~ 75.5,
    per_ocu == "101 a 250 personas" ~ 175.5,
    per_ocu %in% c("251 y m<e1>s personas", "251 y más personas") ~ 300,
    TRUE ~ NA_real_
  )
}

normalize_codigo_act <- function(x) {
  str_pad(str_remove_all(as.character(x), "[^0-9]"), width = 6, side = "left", pad = "0")
}

# SCIAN 813 classes in Mexico (SCIAN 2018/2023 Mexico, not NAICS 813410)
ngo_scian_strict <- c("813230") # Asociaciones y organizaciones civiles

ngo_scian_medium <- c(
  "813230", # civiles
  "813130", # profesionistas (e.g. biologist / planner associations)
  "813110"  # cámaras de productores (fishing chambers possible; review in exploration)
)

ngo_scian_broad <- c(
  "813110", "813120", "813130", "813140",
  "813230"
  # exclude 813210 religious, 813220 political (also absent from DENUE)
)

ngo_scian_exclude <- c("813210", "813220")

marine_ngo_regex <- regex(
  paste(
    "marin|oc[eé]an|cost(er|a|ero)|golfo|cort[eé]|mar de cort",
    "pesca|pesquer|acu[ií]cul|manglar|tortug|vaquita|ballen|delfin",
    "conserv|ambient|ecolog|biodivers|arrecif|humedal|playa",
    "parque marino|área natural|area natural|anp|reserva",
    "lobo marino|totoaba|comunidad.*pesqu|cooperativa.*pesqu",
    "vaquita|totoaba|semarnat|conanp|conapesca",
    sep = "|"
  ),
  ignore_case = TRUE
)

pick_ngo_scian <- function(tier = ngo_filter_tier) {
  switch(
    tier,
    strict = ngo_scian_strict,
    medium = ngo_scian_medium,
    broad = ngo_scian_broad,
    stop("ngo_filter_tier must be strict, medium, or broad")
  )
}

# Sector 81 is two INEGI files (1 de 2 / 2 de 2), same pattern as sector 72.
# Accept both unzipped names (denue_00_81_1_csv) and TR-style suffixes (denue_00_81_1_0526_csv).
denue_81_part_path <- function(year, part, base_path = denue_base_path, sector = ngo_sector) {
  suffix <- get_folder_suffix(year)
  folder_candidates <- c(
    paste0("denue_00_81_", part, "_", suffix, "_csv"),
    paste0("denue_00_81_", part, "_csv")
  )
  file_candidates <- c(
    paste0("denue_inegi_81_", part, ".csv"),
    "denue_inegi_81_.csv"
  )
  for (folder in folder_candidates) {
    for (fname in file_candidates) {
      csv_path <- here::here(base_path, sector, folder, "conjunto_de_datos", fname)
      if (file.exists(csv_path)) return(csv_path)
    }
  }
  NA_character_
}

denue_81_paths <- function(year, base_path = denue_base_path, sector = ngo_sector) {
  paths <- c(
    denue_81_part_path(year, 1, base_path, sector),
    denue_81_part_path(year, 2, base_path, sector)
  )
  paths <- paths[!is.na(paths)]
  if (length(paths) == 0) {
    stop(
      "Missing DENUE sector 81 CSVs for ", year, " under ",
      here::here(base_path, sector)
    )
  }
  paths
}

read_denue_81_year <- function(year) {
  csv_paths <- denue_81_paths(year)
  map_dfr(csv_paths, function(csv_path) {
    message("Reading ", year, ": ", csv_path)
    # INEGI splits leave the same column as character in one file and numeric in the other.
    read_csv(
      csv_path,
      locale = locale(encoding = "latin1"),
      col_types = cols(.default = col_character()),
      show_col_types = FALSE
    ) %>%
      select(-any_of(c("cod_postal", "edificio_e", "telefono")))
  })
}

prepare_denue_goc <- function(dat, year) {
  dat %>%
    clean_names() %>%
    mutate(entidad = str_to_lower(entidad)) %>%
    select(-any_of(c("cod_postal", "edificio_e", "telefono"))) %>%
    filter(entidad %in% goc_states) %>%
    mutate(
      data_year = as.integer(year),
      codigo_act = normalize_codigo_act(codigo_act),
      nom_estab = fix_denue_text(nom_estab),
      raz_social = fix_denue_text(raz_social),
      nombre_act = fix_denue_text(nombre_act),
      municipio = fix_denue_text(municipio),
      id = as.character(id),
      latitud = as.numeric(latitud),
      longitud = as.numeric(longitud)
    ) %>%
    filter(!is.na(longitud), !is.na(latitud), longitud != 0, latitud != 0)
}

label_ngo_scian <- function(codigo_act) {
  case_when(
    codigo_act == "813110" ~ "chamber_producer_commerce",
    codigo_act == "813120" ~ "labor_union",
    codigo_act == "813130" ~ "professional_association",
    codigo_act == "813140" ~ "recreational_regulatory",
    codigo_act == "813210" ~ "religious",
    codigo_act == "813220" ~ "political",
    codigo_act == "813230" ~ "civic_civil_association",
    str_starts(codigo_act, "813") ~ "other_813",
    TRUE ~ "not_813"
  )
}

filter_ngo_rows <- function(df, tier = ngo_filter_tier, require_marine_text = TRUE) {
  allowed <- pick_ngo_scian(tier)
  df <- df %>%
    mutate(
      ngo_class = label_ngo_scian(codigo_act),
      text_blob = paste(nom_estab, raz_social, nombre_act, sep = " "),
      has_marine_text = str_detect(text_blob, marine_ngo_regex)
    ) %>%
    filter(
      codigo_act %in% allowed,
      !(codigo_act %in% ngo_scian_exclude)
    )
  if (require_marine_text) {
    df <- df %>% filter(has_marine_text)
  }
  df
}

clip_to_study_area <- function(points_sf, study_layer, label) {
  clipped <- st_intersection(
    st_transform(points_sf, st_crs(study_layer)),
    study_layer
  )
  message("Kept ", nrow(clipped), " rows inside ", label)
  clipped
}

save_marine_ngo_outputs <- function(marine_ngo_goc, gpkg_path, list_csv_path) {
  marine_ngo_goc %>%
    st_drop_geometry() %>%
    arrange(data_year, rgn_id, nom_estab) %>%
    write_csv(list_csv_path)
  st_write(marine_ngo_goc, gpkg_path, delete_dsn = TRUE, quiet = TRUE)
  marine_ngo_goc %>%
    st_drop_geometry() %>%
    count(data_year, rgn_id, name = "n_units") %>%
    arrange(data_year, rgn_id)
}
```

# 2. Study areas (10 km and 50 km, same buffers as script1_denue_prep.qmd)

```{r}
inland_10 <- st_read(
  here::here("/home/shares/ohi/OHI_GOC/spatial/inland_10km_buffer", "inland_10km_goc.shp"),
  quiet = TRUE
)
inland_50 <- st_read(
  here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer", "inland_50km_goc.shp"),
  quiet = TRUE
)
regions_marine <- st_read(
  here::here("spatial/ohi_regions/goc_ohi_rgns.shp"),
  quiet = TRUE
) %>%
  filter(location == "marine")

marine_and_inland10 <- bind_rows(
  regions_marine %>% dplyr::select(rgn_id, geometry),
  inland_10 %>% dplyr::select(rgn_id, geometry)
) %>%
  group_by(rgn_id) %>%
  summarize(geometry = st_union(geometry), .groups = "drop")
```

# 3. Explore every `nombre_act` in the May 2026 sector 81 snapshot

Sector 81 is not only NGOs. Most of it is auto repair, beauty salons, laundry, and churches. This chunk lists **all 43 SCIAN classes** in the two CSVs, then zooms in on subsector **813**.

```{r}
denue_81_2026 <- read_denue_81_year(2026) %>% prepare_denue_goc(2026)

denue_81_nombre_act <- denue_81_2026 %>%
  count(codigo_act, nombre_act, name = "n_goc_six_states") %>%
  arrange(desc(n_goc_six_states)) %>%
  mutate(
    scian_3digit = str_sub(codigo_act, 1, 3),
    is_813 = str_starts(codigo_act, "813"),
    ngo_class = label_ngo_scian(codigo_act)
  )

write_csv(denue_81_nombre_act, out_nombre_act_csv)
denue_81_nombre_act

denue_81_nombre_act %>%
  count(scian_3digit, wt = n_goc_six_states, name = "n_establishments") %>%
  arrange(desc(n_establishments))
```

```{r}
denue_813_explore <- denue_81_2026 %>%
  filter(str_starts(codigo_act, "813")) %>%
  mutate(ngo_class = label_ngo_scian(codigo_act))

unique(denue_813_explore$ngo_class)
# [1] "religious"                 "chamber_producer_commerce" "civic_civil_association"   "labor_union"               "professional_association" 
# [6] "recreational_regulatory"  

denue_813_activity_summary <- denue_813_explore %>%
  count(data_year, codigo_act, nombre_act, ngo_class, name = "n_goc_statewide") %>%
  arrange(desc(n_goc_statewide))

write_csv(denue_813_activity_summary, out_explore_csv)
denue_813_activity_summary

# frequent civic-association names (813230), mostly ejidos, churches excluded already, etc.!
denue_813_explore %>%
  filter(codigo_act == "813230") %>%
  count(nom_estab, nombre_act, sort = TRUE) %>%
  slice_head(n = 40)

# marine-keyword hits inside civic associations (many false positives!!!: Marina condos, Playas, Costa...check them)
marine_keywords <- denue_813_explore %>%
  filter(codigo_act == "813230") %>%
  mutate(text_blob = paste(nom_estab, raz_social, nombre_act, sep = " ")) %>%
  filter(str_detect(text_blob, marine_ngo_regex)) %>%
  select(entidad, municipio, nom_estab, raz_social) %>%
  arrange(entidad, municipio, nom_estab)
```

**How to read this**

| SCIAN | `nombre_act` (expected) | NGO type | Keep for marine CCE? |
|-------|-------------------------|----------|----------------------|
| **813230** | Asociaciones y organizaciones civiles | AC / civil society | **Yes**, core tier, plus marine keyword |
| **813130** | Asociaciones y organizaciones de profesionistas | Professional associations | Maybe, if marine keyword (medium tier) |
| **813110** | Asociaciones, organizaciones y cámaras de productores... | Chambers / producer groups | Maybe for fishing chambers only (review names) |
| **813120** | Asociaciones y organizaciones laborales y sindicales | Labor unions | Usually no (broad tier only if EWG wants) |
| **813140** | Asociaciones regulatorias de actividades recreativas | Sports/fishing leagues | Review case by case |
| **813210** | Asociaciones y organizaciones religiosas | Religious | **No** |
| **813220** | Asociaciones y organizaciones políticas | Political | **No** (also excluded from DENUE) |

DENUE also lists some **associations under sector 71** (for example sports teams). Those are not in sector 81 and are not included here.

# 4. Process all years, filter NGOs, clip to 10 km and 50 km

```{r}
process_denue_ngo_year <- function(year, tier = ngo_filter_tier) {
  dat <- read_denue_81_year(year) %>%
    prepare_denue_goc(year) %>%
    filter_ngo_rows(tier = tier, require_marine_text = TRUE) %>%
    mutate(
      municipio_clean = clean_municipio(municipio),
      employees_estimated = employees_from_per_ocu(per_ocu),
      filter_tier = tier
    )
  if (nrow(dat) == 0) return(dat)
  dat %>%
    st_as_sf(coords = c("longitud", "latitud"), crs = 4326, remove = FALSE)
}

marine_ngo_sf_list <- map(series_years, function(yr) {
  tryCatch(
    process_denue_ngo_year(yr, tier = ngo_filter_tier),
    error = function(e) {
      message("Skipping ", yr, ": ", conditionMessage(e))
      tibble()
    }
  )
})

marine_ngo_sf <- bind_rows(marine_ngo_sf_list) %>%
  distinct(id, data_year, .keep_all = TRUE)

if (nrow(marine_ngo_sf) == 0) {
  stop(
    "No marine NGO rows found. Download DENUE sector 81 for 2019-2026 ",
    "and check int/marine_ngo_denue_activity_exploration.csv."
  )
}

message(
  "Filtered to ", nrow(marine_ngo_sf),
  " marine NGO unit-years (tier = ", ngo_filter_tier, ") before spatial clip."
)

marine_ngo_goc_10km <- clip_to_study_area(marine_ngo_sf, marine_and_inland10, "10 km")
marine_ngo_goc_50km <- clip_to_study_area(marine_ngo_sf, inland_50, "50 km")
```

# 5. Save outputs

```{r}
save_marine_ngo_outputs(marine_ngo_goc_10km, out_10km_gpkg, out_list_10km_csv)
save_marine_ngo_outputs(marine_ngo_goc_50km, out_50km_gpkg, out_list_50km_csv)
```

Then run:

- `marine_ngo_per_capita.qmd` for 10 km scores
- `marine_ngo_per_capita_50km.qmd` for 50 km scores

Change `ngo_filter_tier` at the top (`strict`, `medium`, or `broad`) after reviewing `marine_ngo_denue_activity_exploration.csv`.

