################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```

# >> Report Statistics <<  ------------------------------
```{r Load in AFO_Scores as needed}
AFO_Scores0<- read.csv("/home/ferrer/ohi_goc_prep/AO/v2025/AO_Scores/AFO_Scores/AFO_Scores_1.0_df_Jun152026.csv")

AFO_Scores<- AFO_Scores0
```

```{r Average AFO scores over time}
AFO_averages<- AFO_Scores %>%
  group_by(OHI_region_est, year) %>%
  summarize(
    mean_AFO_score = mean(AFO_Score, na.rm= TRUE)
  )

# Year 2024 mean for AO scores:
mean(AFO_Scores$AFO_Score[AFO_Scores$year == 2024], na.rm = TRUE)
mean(AFO_Scores$AFO_Score, na.rm = TRUE)
```

# STEP 1: Load in and summarize GOC_MarineLandings0

```{r Load in GOC_MarineLandings0 as necessary}
GOC_MarineLandings0<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Marine.Landings_Sector.Sorted.Georef_PATCHED_Jun032026.rds"))
# 3246583 obs., 58 var.

# NOTE TO SELF: This source file may change in later Summer / Fall run. - EMF, June 4 2026
# See the patch file here: https://docs.google.com/document/d/1L6w4Gd8SOjp14f_u46vyn64rEpv4qBRaN_V2IyPjP5w/edit?tab=t.0
```


# PATCH 1 - Look into groups that are unassigned to a Major_Functional_Group:
NOTE TO SELF: This patch should not be necessary when we re-run this in later Summer / Fall run. - EMF, June 15 2026
See the patch file here: https://docs.google.com/document/d/1L6w4Gd8SOjp14f_u46vyn64rEpv4qBRaN_V2IyPjP5w/edit?tab=t.0

```{r Load in GenusFamilyGroup_Key as needed}
GenusFamilyGroup_Key<- read.csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Keys/Taxonomic_Keys/GenusFamilyGroupKey_COMPLETE_Apr042026.csv")

# NOTE TO SELF: This patch should not be necessary when we re-run this in later Summer / Fall run. - EMF, June 15 2026
# See the patch file here: https://docs.google.com/document/d/1L6w4Gd8SOjp14f_u46vyn64rEpv4qBRaN_V2IyPjP5w/edit?tab=t.0

MajorFunctionalGroup_Key<- GenusFamilyGroup_Key %>%
  select(estimated_genus_name, Major_Functional_Group)
```

```{r Major_Functional_Group patch for GOC_MarineLandings}
# Fill in missing values based on MajorFunctionalGroup_Key:
GOC_MarineLandings <- GOC_MarineLandings0 %>%
  left_join(
    MajorFunctionalGroup_Key %>%
      select(estimated_genus_name, Major_Functional_Group),
    by = "estimated_genus_name",
    suffix = c("", ".lookup")
  )

GOC_MarineLandings <- GOC_MarineLandings %>%
  mutate(
    Major_Functional_Group = coalesce(
      Major_Functional_Group,
      Major_Functional_Group.lookup
    )
  ) %>%
  select(-Major_Functional_Group.lookup)

# 3246583 obs., 58 var.

# SANITY CHECK:
sum(is.na(GOC_MarineLandings$Major_Functional_Group))
# Only 26 records unassigned. Looks good!
```

# PATCH 2 - Re-assign litopenaeus as paneus:

```{r Re-assign paneus as litopenaeus}
GOC_MarineLandings <- GOC_MarineLandings %>%
  mutate(
    estimated_genus_name = case_when(
      estimated_genus_name == "Litopenaeus" ~ "Penaeus",
      TRUE ~ estimated_genus_name
    )
  )

unique(GOC_MarineLandings$estimated_genus_name)
```

```{r Select Sectors_of_Interest}
table(GOC_MarineLandings$sector_estimate_draft)

# SANITY CHECK: Which species are included in the freshwater or marine sector?
TEST<- GOC_MarineLandings %>%
  filter(sector_estimate_draft== "freshwater or marine: fishery sector yet to be classified")

table(TEST$recorded_taxa_group)
# Species like Mojarra (the overwhelming majority of catch), trout, and catfish, which can move from freshwater to marine systems and are a bit ambiguous in terms of where they can be caught.


MarineFishery_Sectors<- c("marine: fishery sector yet to be classified",
                          "marine: small-scale capture",
                          "marine: large-scale capture")
  
PossibleMarineFishery_Sectors<- c("freshwater or marine: large-scale capture",
                          "freshwater or marine: small-scale capture",
                          "freshwater or marine: fishery sector yet to be classified")

GOC_MarineFisheryLandings0<- GOC_MarineLandings %>%
  filter(sector_estimate_draft %in% c(MarineFishery_Sectors, PossibleMarineFishery_Sectors))

# 3137044 obs., 58 var.
```

```{r How many landings are produced by SSF in the GOC?}
table(GOC_MarineFisheryLandings0$GOC_est)
# All GOC records

colnames(GOC_MarineFisheryLandings0)
table(GOC_MarineFisheryLandings0$sector_estimate_draft)

set.seed(123)

slice <- GOC_MarineFisheryLandings0 %>%
  filter(year >= 2018 & year <= 2024) %>%
  slice_sample(n = 200)

# Summarize total landings and value from SSF
GOC_SSF <- GOC_MarineFisheryLandings0 %>%
  filter(year >= 2018 & year <= 2024) %>%
  group_by(year) %>%
  summarise(
    SSF_landings = sum(
      weight_kg_corrected[grepl("small-scale capture", sector_estimate_draft)],
      na.rm = TRUE
    ),
    total_landings = sum(weight_kg_corrected, na.rm = TRUE),
    
    SSF_value = sum(
      value[grepl("small-scale capture", sector_estimate_draft)],
      na.rm = TRUE
    ),
    total_value = sum(value, na.rm = TRUE),
    
    .groups = "drop"
  ) %>%
  mutate(
    pct_catch = 100 * SSF_landings / total_landings,
    pct_value = 100 * SSF_value / total_value
  )
  

# Mean from 2018-2024
mean(GOC_SSF$pct_catch) # ~19%
mean(GOC_SSF$pct_value) # ~51%

#
GOC_boats <- GOC_MarineFisheryLandings0 %>%
  filter(
    year >= 2018 & year <= 2024,
    grepl("small-scale capture|large-scale capture", sector_estimate_draft)
  ) %>%
  mutate(
    sector = case_when(
      grepl("small-scale capture", sector_estimate_draft) ~ "SSF",
      grepl("large-scale capture", sector_estimate_draft) ~ "Large-scale"
    )
  ) %>%
  group_by(year, sector) %>%
  summarise(
    n_eu = n_distinct(eu_rnpa),
    .groups = "drop"
  ) %>%
  group_by(year) %>%
  mutate(
    pct_eu = 100 * n_eu / sum(n_eu)
  ) %>%
  ungroup()

# Mean from 2018-2024
mean(GOC_boats$pct_eu[GOC_boats$sector=="SSF"]) # ~88%
```

```{r Graph of top SSF capture}
GOC_SSF_taxa <- GOC_MarineFisheryLandings0 %>%
  filter(
    year >= 2018 & year <= 2024,
    grepl("small-scale capture", sector_estimate_draft),
    !is.na(Major_Functional_Group)
  ) %>%
  group_by(year, Major_Functional_Group) %>%
  summarise(
    total_landings = sum(weight_kg_corrected, na.rm = TRUE),
    .groups = "drop"
  )

library(ggplot2)

table(GOC_SSF_taxa$Major_Functional_Group)

# Remove marine NEI and unknown:
GOC_SSF_taxa_plot <- GOC_SSF_taxa %>%
  filter(
    !Major_Functional_Group %in% c(
      "marine species NEI",
      "unknown",
      "Other marine invertebrates",
      "Marine species NEI",
      "Diadromous fishes"
    )
  )

unique(GOC_SSF_taxa_plot$Major_Functional_Group)

# Relevel:
GOC_SSF_taxa_plot <- GOC_SSF_taxa_plot %>%
  mutate(
    Major_Functional_Group = factor(
      Major_Functional_Group,
      levels = c(
        "Anchovy & Sardine (small pelagics)",
        "Demersal marine fishes",
        "Pelagic marine fish (excl. anchovy & sardine)",
        "Sharks",
        "Marine fishes NEI",
        "Marine or freshwater fishes NEI",
        "Cephalopods",
        "Molluscs excl. Cephalopods",
        "Crustaceans",
        "Shrimps, prawns",
        "Echinoderms",
        "Jellies",
        "Marine algae"
      )
    )
  )

GOC_SSF_taxa_plot <- GOC_SSF_taxa_plot %>%
  mutate(
    Functional_group = recode(
      Major_Functional_Group,
      "Anchovy & Sardine (small pelagics)" = "Small pelagic fishes",
      "Demersal marine fishes" = "Demersal fishes",
      "Pelagic marine fish (excl. anchovy & sardine)" = "Other pelagic fishes",
      "Sharks" = "Sharks",
      "Marine fishes NEI" = "Marine fishes NEI",
      "Marine or freshwater fishes NEI" = "Marine/freshwater fishes NEI",
      "Cephalopods" = "Cephalopods",
      "Molluscs excl. Cephalopods" = "Other molluscs",
      "Crustaceans" = "Crustaceans",
      "Shrimps, prawns" = "Shrimps & prawns",
      "Echinoderms" = "Echinoderms",
      "Jellies" = "Jellies",
      "Marine algae" = "Marine algae"
    )
  )

# Then plot:
SSF_Taxa_Plot<- ggplot(
  GOC_SSF_taxa_plot,
  aes(
    x = factor(year),
    y = total_landings,
    fill = Major_Functional_Group
  )
) +
  geom_col(width = 0.75) +
  labs(
    x = "Year",
    y = "SSF landings (kg)",
    fill = "Major functional group",
    title = "SSF catch by major functional group",
    subtitle = "2018–2024"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "right",
    plot.title = element_text(face = "bold")
  )

SSF_Taxa_Plot
```

```{r SAVE regional AFO score plot}
ggsave(
  plot = SSF_Taxa_Plot,
  filename = "/home/ferrer/ohi_goc_prep/AO/v2025/AO_Scores/AFO_Scores/SSF_Catch_plot_MOVE_Aug132026.png",
  width = 12,
  height = 8,
  dpi = 300
)
```



```{r How much seafood is produced in this region in 2024?}
AllMex_Landings<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/juan_carlos/AllMex_landings_2000_present_EMF_edited.rds"))
# 12153879 obs., 20 vars.

# Subset records from 2024:
AllMex_Landings_2024<- AllMex_Landings %>% filter(year==2024)
# 483775 obs., 20 vars.

# Landed weight correction (kind of looks like sometimes 'landed weight' is 'live weight' and vice versa. Fix this):
AllMex_Landings_2024$weight_kg_corrected <- ifelse(
  AllMex_Landings_2024$live_weight >= AllMex_Landings_2024$landed_weight,
  AllMex_Landings_2024$live_weight,
  AllMex_Landings_2024$landed_weight
)
# 483775 obs., 21 vars.

# Let's filter out any records where weight_kg_corrected is < or = 0.
AllMex_Landings_2024<- AllMex_Landings_2024 %>%
  filter(weight_kg_corrected > 0)
# 483775 obs., 21 vars.

sum(is.na(AllMex_Landings_2024$weight_kg_corrected))
# 0. Good!

# Finally, let's apply a coarse filter to filter out landings from aquaculture. This won't get everything.
table(AllMex_Landings_2024$acuaculture_production) #  NO: 421421 and the rest are NAs.


## VERSUS - [MARINE] LANDINGS IN ONLY THE GoC
# This is an imperfect calculation because we can't be certain that all of these sites are inside the GoC.


GoC_Landings_cleaned<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/fis/v2025/int/GOC_FIS.landings_CLEAN_as.of.Dec092025.rds"))
# GoC_Landings<- Landings_GoC_NoFresh.NoMariculture
# 3163120 obs., 50 vars.

# Subset records from 2024:
GoC_Landings_2024<- GoC_Landings_cleaned %>% filter(year== 2024)
# 143479 obs., 50 vars.

# Let's filter out any records where weight_kg_corrected is < or = 0.
GoC_Landings_2024<- GoC_Landings_2024 %>%
  filter(weight_kg_corrected > 0)
# 143479 obs., 50 vars.

sum(is.na(GoC_Landings_2024$weight_kg_corrected))
# 0. Good!

table(GoC_Landings_2024$OHI_Region_No)
unique(GoC_Landings_2024$OHI_Region_No)

Definitely_GoC_Regions<- c("Region_1", "Region_2", "Region_3",
                           "Region_4", "Region_5", "Region_6",
                           "Region_7", "Region_8", "Region_9",
                           "Region_4_5_6_Sonora_TBD")
# We cannot be certain about the landings from "Region_1_2_Baja_TBD" or "Region_3_4_BCS_TBD"

GoC_Landings_2024<- GoC_Landings_2024 %>% filter(OHI_Region_No %in% Definitely_GoC_Regions)
# 142533 obs., 50 vars.

#### CALCULATION: How much seafood produced in the region? ####
# ANSWER: ~50% in 2024

sum(AllMex_Landings_2024$weight_kg_corrected)
# 1993632680 kg

sum(GoC_Landings_2024$weight_kg_corrected)
# 995487742 kg


#### CALCULATION: How much seafood VALUE is produced in the region? ####
# ANSWER: ~23% in 2024

sum(AllMex_Landings_2024$value)
# 44196283211 pesos

sum(GoC_Landings_2024$value)
# 9978483456 pesos


#### CALCULATION: How much seafood in the region is produced by SSF? ####
# ANSWER: 20% total in 2024; anywhere from 25-48% when exclude large-scale anchovy and sardine

table(GoC_Landings_2024$fleet)
# large_scale small_scale NA
#       14546      127261 

GoC_Landings_2024_All<- GoC_Landings_2024 %>% filter(!is.na(fleet))
# 141807 obs.

GoC_Landings_2024_SSF<- GoC_Landings_2024 %>% filter(fleet=="small_scale")
# 127261 obs.

sum(GoC_Landings_2024_All$weight_kg_corrected) # 992204985 kg
sum(GoC_Landings_2024_SSF$weight_kg_corrected) # 199330559 kg

# What about for direct consumption?
GoC_Landings_2024_Large<- GoC_Landings_2024 %>% filter(fleet=="large_scale")
# 14546 obs.

table(GoC_Landings_2024_Large$recorded_taxa_group)

GoC_Landings_2024_withoutLargeAnchovy <- GoC_Landings_2024 %>% 
  filter(
    !(fleet == "large_scale" &
      recorded_taxa_group %in% c("ANCHOVETA", "SARDINA"))
  )
# 137815 obs.

sum(GoC_Landings_2024_withoutLargeAnchovy$weight_kg_corrected) # 417828233 kg
sum(GoC_Landings_2024_SSF$weight_kg_corrected) # 199330559 kg
```


```{r How much of the seafood can we account for in our analysis?}
# Load file, only if needed:
GoC_Landings_TEST<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/fis/v2025/int/GOC_FIS.landings_CLEAN_as.of.Dec092025.rds"))
# 3163120 obs., 50 vars.

# General filtering:
# In general, the receipts from early trips are a bit funky. Let's filter out trips before 2003.
GoC_Landings_TEST<- GoC_Landings_TEST %>%
  filter(year >= 2003)
# 2896048 obs., 50 vars.

# Let's also filter out any records where weight was < or = 0 (n= 126955 records).
GoC_Landings_TEST<- GoC_Landings_TEST %>%
  filter(weight_kg_corrected > 0 & !is.na(weight_kg_corrected))
# 2895892 obs., 50 vars.

# Re-assigning CPUE major groups based on new key:
Genus_MajorGroup_Key<- read.csv("~/ohi_goc_prep/general/CONAPESCA_Keys/Genus_to_MajorGroup_Key_Dec112025.csv")
# 326 obs., 4 vars.

GoC_Landings_TEST <- GoC_Landings_TEST %>%
  left_join(Genus_MajorGroup_Key,
            by = c("recorded_taxa_group", "estimated_genus_name"))
# 2895892 obs., 52 vars. as expected!


#### CALCULATION: How much of the seafood can we account for in our analysis? ####
# ANSWER: Anywhere from 92-99% in any given year.

table(GoC_Landings_TEST$Major_Functional_Group)
# We can account for 2778209 out of 2895892 records (96%).

# How much of the landings can we account for?
Yearly_weight_summary <- GoC_Landings_TEST %>%
  group_by(year) %>%
  summarise(
    # Total weight across ALL records
    total_weight_all = sum(weight_kg_corrected, na.rm = TRUE),

    # Weight where Major_Functional_Group is present and not NEI
    total_weight_tax_grouped = sum(
      weight_kg_corrected[
        !is.na(Major_Functional_Group) &
        Major_Functional_Group != "NEI"
      ],
      na.rm = TRUE
    ),

    # Optional: proportion of total that is taxonomically grouped
    prop_tax_grouped = total_weight_tax_grouped / total_weight_all
  ) %>%
  ungroup()
```

