---
title: "Wetland Habitat Condition and Trend - Gulf of California OHI"
author: "Austin Martinez"
date: 08/20/2026
format: 
  html:
    code-fold: show
execute:
  echo: true
  message: false
  warning: false
---

```{r}
# Load in Libraries
library(sf)
library(dplyr)
library(here)
library(terra)
library(tidyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(tmap)
library(stringr)
```

```{r}
# Read in OHI region boundaries
ohi_regions <- st_read(here("spatial", "ohi_regions", "goc_ohi_rgns.shp")) |> st_make_valid()
```

```{r}
# Read in Hem's satmarsh layer
hem_saltmarsh <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/Hem_data/goc_habitat_data/wetlands/wetlands_merge.shp") |> st_make_valid()
```

## CANABIO Land Use and Vegitation all years
#### Dunes data
```{r}
dune_canabio_landuse_1980_91 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250kcs1agw/usv250kcs1agw.shp") |> st_make_valid()

dune_canabio_landuse_2001 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250ks2gw/usv250ks2gw.shp") |> st_make_valid()

dune_canabio_landuse_2005 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250ks3gw/usv250ks3gw.shp") |> st_make_valid()

dune_canabio_landuse_2009 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250ks4gw/usv250ks4gw.shp") |> st_make_valid()

dune_canabio_landuse_2013 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250s5ugw/usv250s5ugw.shp") |> st_make_valid()

dune_canabio_landuse_2016 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250s6gw/usv250s6gw.shp") |> st_make_valid()

dune_canabio_landuse_2021 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250s7gw/usv250s7gw.shp") |> st_make_valid()
```

### Saltmarsh
```{r}
saltmarsh_canabio_landuse_1980_91 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250kcs1agw/usv250kcs1agw.shp") |> st_make_valid()

saltmarsh_canabio_landuse_2001 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250ks2gw/usv250ks2gw.shp") |> st_make_valid()

saltmarsh_canabio_landuse_2005 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250ks3gw/usv250ks3gw.shp") |> st_make_valid()

saltmarsh_canabio_landuse_2009 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250ks4gw/usv250ks4gw.shp") |> st_make_valid()

saltmarsh_canabio_landuse_2013 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250s5ugw/usv250s5ugw.shp") |> st_make_valid()

saltmarsh_canabio_landuse_2016 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250s6gw/usv250s6gw.shp") |> st_make_valid()

saltmarsh_canabio_landuse_2021 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250s7gw/usv250s7gw.shp") |> st_make_valid()
```


## Habitat buffers:
### Dunes
```{r}
# Create 5km buffer (run once)
sf::sf_use_s2(FALSE)

ohi_regions_marine <- ohi_regions |>
  filter(location == "marine") |>
  st_union()

# Make a 5km buffer ring
buffer_outer <- st_buffer(ohi_regions_marine, dist = 5000)
buffer_ring  <- st_difference(buffer_outer, ohi_regions_marine) |> st_make_valid()

# Union it to the marine OHI layer
ohi_marine_plus_5km <- st_union(ohi_regions_marine, buffer_ring)

# Make a 1km buffer ring
buffer_outer <- st_buffer(ohi_regions_marine, dist = 1000)
buffer_ring  <- st_difference(buffer_outer, ohi_regions_marine) |> st_make_valid()

# Union it to the marine OHI layer
ohi_marine_plus_1km <- st_union(ohi_regions_marine, buffer_ring)
```

### Saltmarsh
```{r}
ohi_regions_marine <- ohi_regions |>
  filter(location == "marine") |>
  st_union()

# Make a buffer ring
buffer_outer <- st_buffer(ohi_regions_marine, dist = 5000)
buffer_ring  <- st_difference(buffer_outer, ohi_regions_marine) |> st_make_valid()

# Union it to the marine OHI layer
ohi_marine_plus_5km <- st_union(ohi_regions_marine, buffer_ring)
```


### Process Saltmarsh data:
We are using these layers to represent Saltmarshes:
-     VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA 
-     VEGETACIÓN HALÓFILA

### Process dune data:
We are using these layers to represent Coastal dunes:
-     VEGETACIÓN DE DUNAS COSTERAS with a 5km buffer
-     VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA XERÓFILA with a 1km buffer
-     VEGETACIÓN HALÓFILA XERÓFILA with a 1km buffer


1991 has no observations for layers:
      "VEGETACIÓN HALÓFILA XERÓFILA",
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA XERÓFILA"
      
So, only the 5km buffer was used in the analysis.

```{r}
#2001
# Extract the SAL layers
saltmarsh_2001 <- saltmarsh_canabio_landuse_2001 %>%
  filter(descripcio %in% c(
    "VEGETACIÓN HALÓFILA HIDRÓFILA",
    "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA HIDRÓFILA"
  ))

# Extract the 5km DUN layers
dunes_2001_5km <- dune_canabio_landuse_2001 %>%
  filter(descripcio %in% c("VEGETACIÓN DE DUNAS COSTERAS"))

# Extract the 1km DUN layers
dunes_2001_1km <- dune_canabio_landuse_2001 %>%
  filter(descripcio %in% c(
    "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA XERÓFILA",
    "VEGETACIÓN HALÓFILA XERÓFILA"
  ))

# Match CRSs to saltmarsh_2001
target_crs <- st_crs(saltmarsh_2001)
ohi_region <- st_transform(ohi_regions, target_crs)
ohi_marine_plus_5km <- st_transform(ohi_marine_plus_5km, target_crs)
ohi_marine_plus_1km <- st_transform(ohi_marine_plus_1km, target_crs)

# Clip SAL and DUN data to the buffer extents
saltmarsh_2001 <- st_intersection(saltmarsh_2001, ohi_marine_plus_5km)
dunes_2001_5km <- st_intersection(dunes_2001_5km, ohi_marine_plus_5km)
dunes_2001_1km <- st_intersection(dunes_2001_1km, ohi_marine_plus_1km)

# Combine data into 1 dataframe 
habitat_2001 <- rbind(saltmarsh_2001, dunes_2001_5km, dunes_2001_1km)

# Find the DUN area in each OHI region
habitat_2001_area <- st_intersection(habitat_2001, ohi_region) %>%
  group_by(rgn_id) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6,
         year = 2001) %>%
  st_drop_geometry()

```

```{r}
# 2005
# Extract the SAL layers
saltmarsh_2005 <- saltmarsh_canabio_landuse_2005 %>%
  filter(descripcio %in% c(
    "VEGETACIÓN HALÓFILA HIDRÓFILA",
    "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA HIDRÓFILA"
  ))

# Extract the 5km DUN layers
dunes_2005_5km <- dune_canabio_landuse_2005 %>%
  filter(descripcio %in% c("VEGETACIÓN DE DUNAS COSTERAS"))

# Extract the 1km DUN layers
dunes_2005_1km <- dune_canabio_landuse_2005 %>%
  filter(descripcio %in% c(
    "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA XERÓFILA",
    "VEGETACIÓN HALÓFILA XERÓFILA"
  ))

# Match CRSs to saltmarsh data
target_crs <- st_crs(saltmarsh_2005)
ohi_region <- st_transform(ohi_regions, target_crs)
ohi_marine_plus_5km <- st_transform(ohi_marine_plus_5km, target_crs)
ohi_marine_plus_1km <- st_transform(ohi_marine_plus_1km, target_crs)

# Clip SAL and DUN data to the buffer extents
saltmarsh_2005 <- st_intersection(saltmarsh_2005, ohi_marine_plus_5km)
dunes_2005_5km <- st_intersection(dunes_2005_5km, ohi_marine_plus_5km)
dunes_2005_1km <- st_intersection(dunes_2005_1km, ohi_marine_plus_1km)

# Combine data into 1 dataframe 
habitat_2005 <- rbind(saltmarsh_2005, dunes_2005_5km, dunes_2005_1km)

# Find the DUN area in each OHI region
habitat_2005_area <- st_intersection(habitat_2005, ohi_region) %>%
  group_by(rgn_id) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6,
         year = 2005) %>%
  st_drop_geometry()

```

```{r}
# 2009
# Extract the SAL layers
saltmarsh_2009 <- saltmarsh_canabio_landuse_2009 %>%
  filter(descripcio %in% c(
    "VEGETACIÓN HALÓFILA HIDRÓFILA",
    "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA HIDRÓFILA"
  ))

# Extract the 5km DUN layers
dunes_2009_5km <- dune_canabio_landuse_2009 %>%
  filter(descripcio %in% c("VEGETACIÓN DE DUNAS COSTERAS"))

# Extract the 1km DUN layers
dunes_2009_1km <- dune_canabio_landuse_2009 %>%
  filter(descripcio %in% c(
    "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA XERÓFILA",
    "VEGETACIÓN HALÓFILA XERÓFILA"
  ))

# Match CRSs to saltmarsh data
target_crs <- st_crs(saltmarsh_2009)
ohi_region <- st_transform(ohi_regions, target_crs)
ohi_marine_plus_5km <- st_transform(ohi_marine_plus_5km, target_crs)
ohi_marine_plus_1km <- st_transform(ohi_marine_plus_1km, target_crs)

# Clip SAL and DUN data to the buffer extents
saltmarsh_2009 <- st_intersection(saltmarsh_2009, ohi_marine_plus_5km)
dunes_2009_5km <- st_intersection(dunes_2009_5km, ohi_marine_plus_5km)
dunes_2009_1km <- st_intersection(dunes_2009_1km, ohi_marine_plus_1km)

# Combine data into 1 dataframe 
habitat_2009 <- rbind(saltmarsh_2009, dunes_2009_5km, dunes_2009_1km)

# Find the DUN area in each OHI region
habitat_2009_area <- st_intersection(habitat_2009, ohi_region) %>%
  group_by(rgn_id) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6,
         year = 2009) %>%
  st_drop_geometry()

```

```{r}
# 2013
# Extract the SAL layers
saltmarsh_2013 <- saltmarsh_canabio_landuse_2013 %>%
  filter(descripcio %in% c(
    "VEGETACIÓN HALÓFILA HIDRÓFILA",
    "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA HIDRÓFILA"
  ))

# Extract the 5km DUN layers
dunes_2013_5km <- dune_canabio_landuse_2013 %>%
  filter(descripcio %in% c("VEGETACIÓN DE DUNAS COSTERAS"))

# Extract the 1km DUN layers
dunes_2013_1km <- dune_canabio_landuse_2013 %>%
  filter(descripcio %in% c(
    "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA XERÓFILA",
    "VEGETACIÓN HALÓFILA XERÓFILA"
  ))

# Match CRSs to saltmarsh data
target_crs <- st_crs(saltmarsh_2013)
ohi_region <- st_transform(ohi_regions, target_crs)
ohi_marine_plus_5km <- st_transform(ohi_marine_plus_5km, target_crs)
ohi_marine_plus_1km <- st_transform(ohi_marine_plus_1km, target_crs)

# Clip SAL and DUN data to the buffer extents
saltmarsh_2013 <- st_intersection(saltmarsh_2013, ohi_marine_plus_5km)
dunes_2013_5km <- st_intersection(dunes_2013_5km, ohi_marine_plus_5km)
dunes_2013_1km <- st_intersection(dunes_2013_1km, ohi_marine_plus_1km)

# Combine data into 1 dataframe 
habitat_2013 <- rbind(saltmarsh_2013, dunes_2013_5km, dunes_2013_1km)

# Find the DUN area in each OHI region
habitat_2013_area <- st_intersection(habitat_2013, ohi_region) %>%
  group_by(rgn_id) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6,
         year = 2013) %>%
  st_drop_geometry()

```

```{r}
#2016
# Extract the SAL layers
saltmarsh_2016 <- saltmarsh_canabio_landuse_2016 %>%
  filter(DESCRIPCIO %in% c(
    "VEGETACIÓN HALÓFILA HIDRÓFILA",
    "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA HIDRÓFILA"
  ))

# Extract the 5km DUN layers
dunes_2016_5km <- dune_canabio_landuse_2016 %>%
  filter(DESCRIPCIO %in% c("VEGETACIÓN DE DUNAS COSTERAS"))

# Extract the 1km DUN layers
dunes_2016_1km <- dune_canabio_landuse_2016 %>%
  filter(DESCRIPCIO %in% c(
    "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA XERÓFILA",
    "VEGETACIÓN HALÓFILA XERÓFILA"
  ))

# Match CRSs to saltmarsh data
target_crs <- st_crs(saltmarsh_2016)
ohi_region <- st_transform(ohi_regions, target_crs)
ohi_marine_plus_5km <- st_transform(ohi_marine_plus_5km, target_crs)
ohi_marine_plus_1km <- st_transform(ohi_marine_plus_1km, target_crs)

# Clip SAL and DUN data to the buffer extents
saltmarsh_2016 <- st_intersection(saltmarsh_2016, ohi_marine_plus_5km)
dunes_2016_5km <- st_intersection(dunes_2016_5km, ohi_marine_plus_5km)
dunes_2016_1km <- st_intersection(dunes_2016_1km, ohi_marine_plus_1km)

# Combine data into 1 dataframe 
habitat_2016 <- rbind(saltmarsh_2016, dunes_2016_5km, dunes_2016_1km)

# Find the DUN area in each OHI region
habitat_2016_area <- st_intersection(habitat_2016, ohi_region) %>%
  group_by(rgn_id) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6,
         year = 2016) %>%
  st_drop_geometry()

```

```{r}
#2021
# Extract the SAL layers
saltmarsh_2021 <- saltmarsh_canabio_landuse_2021 %>%
  filter(DESCRIPCIO %in% c(
    "VEGETACIÓN HALÓFILA HIDRÓFILA",
    "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA HIDRÓFILA"
  ))

# Extract the 5km DUN layers
dunes_2021_5km <- dune_canabio_landuse_2021 %>%
  filter(DESCRIPCIO %in% c("VEGETACIÓN DE DUNAS COSTERAS"))

# Extract the 1km DUN layers
dunes_2021_1km <- dune_canabio_landuse_2021 %>%
  filter(DESCRIPCIO %in% c(
    "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA XERÓFILA",
    "VEGETACIÓN HALÓFILA XERÓFILA"
  ))

# Match CRSs to saltmarsh data
target_crs <- st_crs(saltmarsh_2021)
ohi_region <- st_transform(ohi_regions, target_crs)
ohi_marine_plus_5km <- st_transform(ohi_marine_plus_5km, target_crs)
ohi_marine_plus_1km <- st_transform(ohi_marine_plus_1km, target_crs)

# Clip SAL and DUN data to the buffer extents
saltmarsh_2021 <- st_intersection(saltmarsh_2021, ohi_marine_plus_5km)
dunes_2021_5km <- st_intersection(dunes_2021_5km, ohi_marine_plus_5km)
dunes_2021_1km <- st_intersection(dunes_2021_1km, ohi_marine_plus_1km)

# Combine data into 1 dataframe 
habitat_2021 <- rbind(saltmarsh_2021, dunes_2021_5km, dunes_2021_1km)

# Find the DUN area in each OHI region
habitat_2021_area <- st_intersection(habitat_2021, ohi_region) %>%
  group_by(rgn_id) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6,
         year = 2021) %>%
  st_drop_geometry()

```

```{r}
# Combine all years of data
habitat_all_years <- bind_rows(
  #habitat_1980_91_area,
  habitat_2001_area,
  habitat_2005_area,
  habitat_2009_area,
  habitat_2013_area,
  habitat_2016_area,
  habitat_2021_area
)
```

# Aquaculture
```{r}
# 2021
# Extract the Aquaculture layer
saltmarsh_canabio_2021_aquaculture <- saltmarsh_canabio_landuse_2021 |>
  dplyr::filter(
    DESCRIPCIO %in% c(
      "ACUÍCOLA"
    )
  )
```


## Extent
### Hem's data
```{r}
# Match Hem's data to the CRS of OHI regions
hem_saltmarsh <- st_transform(hem_saltmarsh, st_crs(ohi_regions))

# Remove the mangrove layer
hem_saltmarsh_no_mangrove <- hem_saltmarsh |>
  dplyr::filter(VEG_TYPES != "Manglar - Vegetacion Primaria")
```

```{r}
# Match CRS of Hem and OHI + 5 km buffer
target_crs <- st_crs(hem_saltmarsh_no_mangrove)

ohi_region <- st_transform(ohi_regions, target_crs)
ohi_marine_plus_5km <- st_transform(ohi_marine_plus_5km, target_crs)

hem_saltmarsh_no_mangrove <- st_transform(hem_saltmarsh_no_mangrove, target_crs)

# Clip Hem saltmarsh to 5 km buffer
hem_saltmarsh_5km <- st_intersection(hem_saltmarsh_no_mangrove, ohi_marine_plus_5km)

# Match CRS of aquaculture to Hem saltmarsh
saltmarsh_canabio_2021_aquaculture <- st_transform(
  saltmarsh_canabio_2021_aquaculture,
  st_crs(hem_saltmarsh_5km)
)

# Union aquaculture
saltmarsh_canabio_2021_aquaculture_union <- st_union(saltmarsh_canabio_2021_aquaculture)

# Make geometries valid
hem_saltmarsh_5km <- st_make_valid(hem_saltmarsh_5km)
saltmarsh_canabio_2021_aquaculture_union <- st_make_valid(saltmarsh_canabio_2021_aquaculture_union)

# Remove aquaculture from Hem saltmarsh
hem_saltmarsh_5km_minus_aqua <- st_difference(hem_saltmarsh_5km, saltmarsh_canabio_2021_aquaculture_union)

# Compute area per OHI region
hem_saltmarsh_5km_minus_aqua_area <- st_intersection(hem_saltmarsh_5km_minus_aqua, ohi_region) %>%
  group_by(rgn_id) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    year = 2021
  ) %>%
  st_drop_geometry()

hem_saltmarsh_5km_minus_aqua_area
```

## Current Status
```{r}
# Wetland condition (2021/2001) using CANABIO LU data
wetland_condition <- habitat_all_years %>%
  filter(year %in% c(2001, 2021)) %>%
  pivot_wider(names_from = year, values_from = area_km2) %>%
  mutate(
    condition = (`2021` / `2001`),   # percent change
    condition = pmin(condition, 1),      # cap at 100
    year = 2021
  ) %>%
  select(rgn_id, year, condition)
wetland_condition
```

## Trend
```{r}
# Wetland trend (annual % change) using CANABIO LU data
wetland_trend <- habitat_all_years %>%
  filter(year %in% c(2013, 2021)) %>%
  pivot_wider(names_from = year, values_from = area_km2) %>%
  mutate(
    trend = ((`2021` / `2013`)^(1/8) - 1),   # annual % change
    year = 2021,
    rgn_id = gsub("_", " ", rgn_id)
  ) %>%
  select(rgn_id, year, trend)
wetland_trend
```

```{r}
# Write CSVs to output_data directory
write.csv(hem_saltmarsh_5km_minus_aqua_area,
          "WET_outputs/wetland_extent.csv",
          row.names = FALSE)

write.csv(wetland_trend,
          "WET_outputs/wetland_trend.csv",
          row.names = FALSE)

write.csv(wetland_condition,
          "WET_outputs/wetland_condition.csv",
          row.names = FALSE)
```









```{r}
# Rename columns for clarity
hem_area <- hem_saltmarsh_5km_minus_aqua_area %>%
  rename(hem_minus_aqua_km2 = area_km2)

habitat_area <- habitat_2021_area %>%
  rename(habitat_km2 = area_km2)

# Join and compute difference
habitat_vs_hem_minus_aqua <- habitat_area %>%
  left_join(hem_area, by = "rgn_id") %>%
  mutate(
    diff_km2 = habitat_km2 - hem_minus_aqua_km2
  )

# Table with edited column names and title
habitat_vs_hem_minus_aqua %>%
  mutate(rgn_id = gsub("_", " ", rgn_id)) %>%   # remove underscores
  select(
    `Region ID` = rgn_id,
    `CONABIO Data (DUN+SAL)` = habitat_km2,
    `Hem Data (Hem - aqua)` = hem_minus_aqua_km2
  ) %>%
  kable(
    digits = 2,
    caption = "Comparison of Hem Data Minus Aquaculture vs CONABIO (DUN+SAL) data"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "bordered", "condensed"),
    full_width = FALSE,
    font_size = 14
  ) %>%
  row_spec(0, color = "black")   
```

```{r}
# Rename columns for clarity
hem_total <- hem_saltmarsh_5km_minus_aqua_area %>%
  rename(hem_minus_aqua_km2 = area_km2)

habitat_total <- habitat_2021_area %>%
  rename(habitat_km2 = area_km2)

# Compute total area values
total_comparison <- tibble(
  habitat_total_km2 = sum(habitat_total$habitat_km2, na.rm = TRUE),
  hem_minus_aqua_total_km2 = sum(hem_total$hem_minus_aqua_km2, na.rm = TRUE),
  diff_km2 = habitat_total_km2 - hem_minus_aqua_total_km2
)

# Table
total_comparison %>%
    select(
    `CONABIO Data (DUN+SAL) Total Area` = habitat_total_km2,
    `Hem Data (Hem - aqua) Total Area` = hem_minus_aqua_total_km2,
    `Difference` = diff_km2
  ) %>%
  kable(
    digits = 2,
    caption = "Total Area Comparison: CONABIO Data (DUN+SAL) vs Hem data Minus Aquaculture"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "bordered", "condensed"),
    full_width = FALSE
  )

```

```{r}
# Add habitat type column BEFORE rbind
saltmarsh_2021$habitat_type <- "Saltmarsh"
dunes_2021_5km$habitat_type <- "Dune"
dunes_2021_1km$habitat_type <- "Dune"

# Rebuild combined layer with the new column
habitat_2021 <- rbind(
  saltmarsh_2021,
  dunes_2021_5km,
  dunes_2021_1km
)

tmap_mode("view")

tm_shape(habitat_2021) +
  tm_polygons(
    col = "habitat_type",
    palette = c(
      "Saltmarsh" = "orange",
      "Dune"      = "green"
    ),
    alpha = 0.2,
    border.col = "gray",
    border.lwd = 0.1,
    title = "Habitat Type"
  )

```

```{r}
tmap_mode("view")

tm_basemap("Esri.WorldImagery") +

tm_shape(hem_saltmarsh_5km) +
  tm_polygons(
    col = "#377eb8",
    alpha = 0.5,
    border.col = "#377eb8",
    group = "Hem Saltmarsh 5km (Before Removal)"
  ) +

tm_shape(hem_saltmarsh_5km_minus_aqua) +
  tm_polygons(
    col = "#e41a1c",
    alpha = 0.7,
    border.col = "#e41a1c",
    group = "Hem Saltmarsh 5km (After Aquaculture Removal)"
  ) +

tm_shape(saltmarsh_canabio_2021_aquaculture) +
  tm_polygons(
    col = "#4daf4a",
    alpha = 0.6,
    border.col = "#4daf4a",
    group = "Aquaculture"
  )

```


`







```{r}
# Add dummy column directly to the geometry layer
hem_saltmarsh_5km_minus_aqua$layer <- "Wetlands"

tmap_mode("plot")

tm_shape(ohi_region) +
  tm_polygons(
    col = "location",
    palette = c("marine" = "#cceeff",
                "land"   = "#f5e6c8"),
    alpha = 0.5,
    border.col = "gray",
    border.lwd = 1,
    legend.show = FALSE
  ) +

tm_shape(hem_saltmarsh_5km_minus_aqua) +
  tm_polygons(
    col = "layer",
    palette = c("Hem Minus Aquaculture" = "#228B22"),
    alpha = 1,
    border.col = "#228B22",
    border.lwd = 2,
    title = "Habitat"
  ) +

tm_layout(
    title = "Hem's Wetland data",
    title.size = 1.2,
    title.fontface = "bold",
    title.position = c(0.02, 0.98),
    frame = FALSE
  ) +
tm_scale_bar(position = c("left", "bottom"))


```
```{r}
# Add dummy column directly to the geometry layer
habitat_2021$layer <- "Wetlands (DUN + SAL)"

tmap_mode("plot")

tm_shape(ohi_region) +
  tm_polygons(
    col = "location",
    palette = c("marine" = "#cceeff",
                "land"   = "#f5e6c8"),
    alpha = 0.5,
    border.col = "gray",
    border.lwd = 1,
    legend.show = FALSE
  ) +

tm_shape(habitat_2021) +
  tm_polygons(
    col = "layer",
    palette = c("CONABIO Habitat 2021" = "#228B22"),
    alpha = 1,
    border.col = "#228B22",
    border.lwd = 2,
    title = "Habitat"
  ) +

tm_layout(
    title = "CONABIO Saltmarsh + Dune Habitat (2021)",
    title.size = 1.2,
    title.fontface = "bold",
    title.position = c(0.02, 0.98),
    frame = FALSE
  ) +
tm_scale_bar(position = c("left", "bottom"))

```









