################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

original author: "Austin Martinez"

################################################################################

# PRE-WORK ------------------------------

```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Load in Libraries
library(sf)
library(dplyr)
library(here)
library(terra)
library(tidyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(tmap)

```

# STEP 1: LOAD IN OHI REGIONS & CREATE 5KM BUFFER ------------------------------

```{r Load in the OHI Region_Key0}
Region_Key0<- read.csv("/home/ferrer/ohi_goc_prep/FP/v2026/CONAPESCA_Keys/Spatial_Keys/RegionKey_FINAL_May262026.csv")

colnames(Region_Key0)

Region_Key<- Region_Key0 %>%
  select(rgn_id= OHI_region_est, rgn_name= OHI_region_name) %>%
  distinct()

# Make uppercase:
Region_Key$rgn_id <- sub("^r", "R", Region_Key$rgn_id)
```

```{r Read in OHI region boundaries}
# Read in OHI region boundaries
ohi_regions <- st_read(here("spatial", "ohi_regions", "goc_ohi_rgns.shp")) %>% 
  st_make_valid()

```

```{r Create 5km coastal buffer around OHI regions}
# DECISION POINT: We clipped coverage to a 5-km coastal buffer to avoid capturing coverage that was not coastal.

COASTAL_BUFFER = 5000 # (5,000m or 5km)

# Create 5km buffer (run once)
sf::sf_use_s2(FALSE)

ohi_regions_marine <- ohi_regions %>%
  filter(location == "marine") %>%
  st_union()

# Make a buffer ring
buffer_outer <- st_buffer(ohi_regions_marine, 
                          dist = COASTAL_BUFFER)

buffer_ring  <- st_difference(buffer_outer, ohi_regions_marine) %>% st_make_valid()

# Union it to the marine OHI layer
ohi_marine_plus_5km <- st_union(ohi_regions_marine, buffer_ring)

```


# STEP 2: CALCULATE WETLAND "CURRENT STATUS", i.e. condition over time (source: CONABIO wetland and scrubland layers, 1998/1991-2021)  ------------------------------

## CONABIO Land-Use Data (1991-2021)
```{r CONABIO Land-Use Data}

CONABIO_landuse.data_1980_91_raw <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250kcs1agw/usv250kcs1agw.shp") %>% st_make_valid()

CONABIO_landuse.data_2001_raw <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250ks2gw/usv250ks2gw.shp") %>% st_make_valid()

CONABIO_landuse.data_2005_raw <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250ks3gw/usv250ks3gw.shp") %>% st_make_valid()

CONABIO_landuse.data_2009_raw <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250ks4gw/usv250ks4gw.shp") %>% st_make_valid()

CONABIO_landuse.data_2013_raw <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250s5ugw/usv250s5ugw.shp") %>% st_make_valid()

CONABIO_landuse.data_2016_raw <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250s6gw/usv250s6gw.shp") %>% st_make_valid()

CONABIO_landuse.data_2021_raw <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/CONABIO_Saltmarsh_landuse_veg/usv250s7gw/usv250s7gw.shp") %>% st_make_valid()

```

```{r Match the land-use CRS to the ohi_regions CRS}
CONABIO_landuse.data_1980_91 <- st_transform(CONABIO_landuse.data_1980_91_raw, st_crs(ohi_regions))
CONABIO_landuse.data_2001 <- st_transform(CONABIO_landuse.data_2001_raw, st_crs(ohi_regions))
CONABIO_landuse.data_2005 <- st_transform(CONABIO_landuse.data_2005_raw, st_crs(ohi_regions))
CONABIO_landuse.data_2009 <- st_transform(CONABIO_landuse.data_2009_raw, st_crs(ohi_regions))
CONABIO_landuse.data_2013 <- st_transform(CONABIO_landuse.data_2013_raw, st_crs(ohi_regions))
CONABIO_landuse.data_2016 <- st_transform(CONABIO_landuse.data_2016_raw, st_crs(ohi_regions))
CONABIO_landuse.data_2021 <- st_transform(CONABIO_landuse.data_2021_raw, st_crs(ohi_regions))

```

## Clip coastal wetland, dune, and scrubland data to the 5km buffer & estimate total area per OHI region:
```{r Clip data 1991 through 2021 to the 20km buffer extent}
# Clip coastal wetland, dune, and scrubland data to the 5km buffer extent (defined above)
CONABIO_landuse.data_1980_91 <- st_intersection(CONABIO_landuse.data_1980_91, ohi_marine_plus_5km)
CONABIO_landuse.data_2001 <- st_intersection(CONABIO_landuse.data_2001, ohi_marine_plus_5km)
CONABIO_landuse.data_2005 <- st_intersection(CONABIO_landuse.data_2005, ohi_marine_plus_5km)
CONABIO_landuse.data_2009 <- st_intersection(CONABIO_landuse.data_2009, ohi_marine_plus_5km)
CONABIO_landuse.data_2013 <- st_intersection(CONABIO_landuse.data_2013, ohi_marine_plus_5km)
CONABIO_landuse.data_2016 <- st_intersection(CONABIO_landuse.data_2016, ohi_marine_plus_5km)
CONABIO_landuse.data_2021 <- st_intersection(CONABIO_landuse.data_2021, ohi_marine_plus_5km)

```


### Process wetland data:
We are using these layers to represent wetlandes:
-     VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA 
-     VEGETACIÓN HALÓFILA

### Process dune data:
We are using these layers to represent Coastal dunes:
-     VEGETACIÓN DE DUNAS COSTERAS with a 5km buffer
-     VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA XERÓFILA with a 5km buffer
-     VEGETACIÓN HALÓFILA XERÓFILA with a 5km buffer


1991 has no observations for layers:
      "VEGETACIÓN HALÓFILA XERÓFILA",
      "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA XERÓFILA"


```{r Select CONABIO layers for coastal wetland dune and scrubland}
# Define wetland layers:
wetland_layers <- c(
    "VEGETACIÓN HALÓFILA HIDRÓFILA",
    "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA HIDRÓFILA"
    )

# Define dune layers:
dunes_layers <- c("VEGETACIÓN DE DUNAS COSTERAS")

# Define scrubland layers:
scrubland_layers <- c(
    "VEGETACIÓN SECUNDARIA ARBUSTIVA DE VEGETACIÓN HALÓFILA XERÓFILA",
    "VEGETACIÓN HALÓFILA XERÓFILA"
    )

WETL_layers <- c(wetland_layers,
                 dunes_layers,
                 scrubland_layers)
```

```{r Extract WETL layers per year}
WETL_conabio_2001_polygons <- CONABIO_landuse.data_2001 %>%
  filter(descripcio %in% WETL_layers)

WETL_conabio_2005_polygons <- CONABIO_landuse.data_2005 %>%
  filter(descripcio %in% WETL_layers)

WETL_conabio_2009_polygons <- CONABIO_landuse.data_2009 %>%
  filter(descripcio %in% WETL_layers)

WETL_conabio_2013_polygons <- CONABIO_landuse.data_2013 %>%
  filter(descripcio %in% WETL_layers)

WETL_conabio_2016_polygons <- CONABIO_landuse.data_2016 %>%
  filter(DESCRIPCIO %in% WETL_layers)

WETL_conabio_2021_polygons <- CONABIO_landuse.data_2021 %>%
  filter(DESCRIPCIO %in% WETL_layers)
```

```{r Map coverage in 2021}

ohi_land   <- ohi_regions %>% dplyr::filter(location == "land")
ohi_marine <- ohi_regions %>% dplyr::filter(location == "marine")
ggplot() +
  geom_sf(data = ohi_land, fill = "#f5e6c8", color = "gray40", linewidth = 0.3) +
  geom_sf(data = ohi_marine, fill = "gray95", color = "gray20", linewidth = 0.4) +
  geom_sf(data = WETL_conabio_2021_polygons, fill = "#2a9d8f", color = NA, alpha = 0.85) +
  labs(title = "Wetlands (saltmarsh + dunes), 2021 — within 5 km of marine OHI") +
  theme_void() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

```

```{r Component wetland dune and scruland layers per year (if useful)}
# 2001:
wetland_area_conabio_2001 <- CONABIO_landuse.data_2001 %>%
  filter(descripcio %in% wetland_layers)

dunes_area_conabio_2001 <- CONABIO_landuse.data_2001 %>%
  filter(descripcio %in% dunes_layers)

scrubland_area_conabio_2001 <- CONABIO_landuse.data_2001 %>%
  filter(descripcio %in% scrubland_layers)

# 2005:
wetland_area_conabio_2005 <- CONABIO_landuse.data_2005 %>%
  filter(descripcio %in% wetland_layers)

dunes_area_conabio_2005 <- CONABIO_landuse.data_2005 %>%
  filter(descripcio %in% dunes_layers)

scrubland_area_conabio_2005 <- CONABIO_landuse.data_2005 %>%
  filter(descripcio %in% scrubland_layers)

# 2009:
wetland_area_conabio_2009 <- CONABIO_landuse.data_2009 %>%
  filter(descripcio %in% wetland_layers)

dunes_area_conabio_2009 <- CONABIO_landuse.data_2009 %>%
  filter(descripcio %in% dunes_layers)

scrubland_area_conabio_2009 <- CONABIO_landuse.data_2009 %>%
  filter(descripcio %in% scrubland_layers)

# 2013:
wetland_area_conabio_2013 <- CONABIO_landuse.data_2013 %>%
  filter(descripcio %in% wetland_layers)

dunes_area_conabio_2013 <- CONABIO_landuse.data_2013 %>%
  filter(descripcio %in% dunes_layers)

scrubland_area_conabio_2013 <- CONABIO_landuse.data_2013 %>%
  filter(descripcio %in% scrubland_layers)

# 2016:
wetland_area_conabio_2016 <- CONABIO_landuse.data_2016 %>%
  filter(DESCRIPCIO %in% wetland_layers)

dunes_area_conabio_2016 <- CONABIO_landuse.data_2016 %>%
  filter(DESCRIPCIO %in% dunes_layers)

scrubland_area_conabio_2016 <- CONABIO_landuse.data_2016 %>%
  filter(DESCRIPCIO %in% scrubland_layers)

# 2021
wetland_area_conabio_2021 <- CONABIO_landuse.data_2021 %>%
  filter(DESCRIPCIO %in% wetland_layers)

dunes_2021 <- CONABIO_landuse.data_2021 %>%
  filter(DESCRIPCIO %in% dunes_layers)

scrubland_area_conabio_2021 <- CONABIO_landuse.data_2021 %>%
  filter(DESCRIPCIO %in% scrubland_layers)

```

## Assign the WETL area to OHI regions:
```{r Assign the WETL area to OHI regions}
# Assign the WETL to OHI regions (per year) in:

# 2001
WETL_area_conabio_2001 <- st_intersection(WETL_conabio_2001_polygons, ohi_regions) %>%
  group_by(rgn_id) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6,
         year = 2001) %>%
  st_drop_geometry()


# 2005
WETL_area_conabio_2005 <- st_intersection(WETL_conabio_2005_polygons, ohi_regions) %>%
  group_by(rgn_id) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6,
         year = 2005) %>%
  st_drop_geometry()


# 2009
WETL_area_conabio_2009 <- st_intersection(WETL_conabio_2009_polygons, ohi_regions) %>%
  group_by(rgn_id) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6,
         year = 2009) %>%
  st_drop_geometry()

# 2013
WETL_area_conabio_2013 <- st_intersection(WETL_conabio_2013_polygons, ohi_regions) %>%
  group_by(rgn_id) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6,
         year = 2013) %>%
  st_drop_geometry()

# 2016
WETL_area_conabio_2016 <- st_intersection(WETL_conabio_2016_polygons, ohi_regions) %>%
  group_by(rgn_id) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6,
         year = 2016) %>%
  st_drop_geometry()

# 2021
WETL_area_conabio_2021 <- st_intersection(WETL_conabio_2021_polygons, ohi_regions) %>%
  group_by(rgn_id) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6,
         year = 2021) %>%
  st_drop_geometry()

```


```{r Combine all years' worth of data into one dataframe (WETL_area_conabio0)}
# Merge:
WETL_area_conabio0 <- bind_rows(
  # WETL_area_conabio_1980.91,
  WETL_area_conabio_2001,
  WETL_area_conabio_2005,
  WETL_area_conabio_2009,
  WETL_area_conabio_2013,
  WETL_area_conabio_2016,
  WETL_area_conabio_2021
)

dim(WETL_area_conabio0)
# 54 obs.,  3 vars.
```


## GAPFILL BETWEEN OBSERVED YEARS:
```{r Gapfill WETL_area_conabio0}

# First, expand the WETL dataframe to include one row per year per region, even if that region has no data: 
WETL_area_conabio <- WETL_area_conabio0 %>%
  complete(
    rgn_id = c(paste0("Region_", 1:9)),
    year = full_seq(year, 1)
  )
# 189 obs., 3 vars.

# Keep track of what was originally observed:
WETL_area_conabio<- WETL_area_conabio %>%
  rename(observed_area_km2= area_km2) %>%
  mutate(
    estimated_area_km2 = observed_area_km2
  )

# Gapfill area_km2 in missing years using the zoo package:
WETL_area_conabio <- WETL_area_conabio %>%
  group_by(rgn_id) %>%
  fill(rgn_id, .direction = "downup") %>%
  mutate(

    # Gap-fill missing values
    estimated_area_km2 = zoo::na.approx(
      estimated_area_km2,
      x = year,
      na.rm = FALSE
    ),

    # Classify provenance
    provenance = case_when(
      !is.na(observed_area_km2) ~ "observed",
      is.na(observed_area_km2) & !is.na(estimated_area_km2) ~ "gapfilled",
      is.na(estimated_area_km2) ~ NA_character_
    )
  ) %>%
  ungroup()

# 189 obs., 5 vars.
```


# STEP 3: CALCULATE WETL SCORES (over time) based on condition from CONABIO land-use data ------------------------------

```{r Calculate WETL_Scores0}
colnames(WETL_area_conabio)

# Calculate WETL_Scores0 relative to 2001 baseline, where scores range from 0-100:
WETL_Scores0 <- WETL_area_conabio %>%
  group_by(rgn_id) %>%
  mutate(
    WETL_current_status = (estimated_area_km2 / estimated_area_km2[year == 2001])*100,
    WETL_current_status = pmin(WETL_current_status, 100),
    WETL_current_status = round(WETL_current_status, 3)
  ) %>%
  select(rgn_id, year, provenance, estimated_area_km2, WETL_current_status)
```


# STEP 4: CALCULATE WETLAND & SCRUBLAND "CURRENT EXTENT" (source: Hem / CEDO Data [est. 2025] minus CONABIO aquaculture data [2021]) ------------------------------
Extent = Hem / CEDO's coastal wetland layer - aquaculture (CONABIO layer), clipped to a 5km buffer

```{r Load in Hem / CEDO wetland data}
# Read in Hem / CEDO data
CEDO_wetland_raw <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/Hem_data/goc_habitat_data/wetlands/wetlands_merge.shp")) %>% 
  st_make_valid()

```

```{r Match CRS of ohi_regions remove mangroves and clip to 5km}
# Match Hem / CEDO data to the CRS of ohi_regions
CEDO_wetland.data <- st_transform(CEDO_wetland_raw, st_crs(ohi_regions))

# Remove the mangrove layer
CEDO_wetland.data_no_mangrove <- CEDO_wetland.data %>%
  dplyr::filter(VEG_TYPES != "Manglar - Vegetacion Primaria")

# Clip Hem / CEDO wetland layer to 5 km buffer:
CEDO_wetland.data_5km <- st_intersection(CEDO_wetland.data_no_mangrove, ohi_marine_plus_5km)

# Coordinate Reference System:
#   User input: Mexico ITRF2008 / LCC 
```

## Aquaculture Data (CONABIO, 2021)
```{r Aquaculture Data}
# Extract the Aquaculture layer (2021, most recent year)
Aquaculture_area_conabio_2021 <- CONABIO_landuse.data_2021 %>%
  dplyr::filter(
    DESCRIPCIO %in% c(
      "ACUÍCOLA"
    )
  )

# Coordinate Reference System:
#   User input: Mexico ITRF2008 / LCC 
```

## Subtract aquaculture layer (CONABIO layer, 2021):
```{r Subtract aquaculture}

# Check that the CRS for Aquaculture_area_conabio_2021 and CEDO_wetland.data_5km match:
st_crs(Aquaculture_area_conabio_2021) == st_crs(CEDO_wetland.data_5km)
# TRUE

# Union the Aquaculture polygons:
Aquaculture_area_conabio_2021_union <- st_union(Aquaculture_area_conabio_2021)

# Make sure geometries are valid:
CEDO_wetland.data_5km <- st_make_valid(CEDO_wetland.data_5km)
Aquaculture_area_conabio_2021_union <- st_make_valid(Aquaculture_area_conabio_2021_union)

# Remove aquaculture from Hem / CEDO wetland layer:
CEDO_wetland.data_5km_no.aqua_area <- st_difference(CEDO_wetland.data_5km, Aquaculture_area_conabio_2021_union)
```

```{r Assign WETL extent to each OHI region}
# Assign the wetland scrubland extent to each OHI region:
CEDO_wetland.data_5km_no.aqua_area_perOHI <- st_intersection(CEDO_wetland.data_5km_no.aqua_area, ohi_regions) %>%
  group_by(rgn_id) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    area_km2 = round(area_km2, 1),
    year = 2021 # The year could be 2021 as that's when the aquaculture layer was produced. The layer could also be 2025, since Hem / CEDO layer doesn't have a year assigned.
  ) %>%
  st_drop_geometry()

# Ensure all regions 1–9 are present; regions with no WETL get area = 0
all_rgns <- tibble(
  rgn_id = paste0("Region_", 1:9),
  year = 2021
)

WETL_Current_Extent <- all_rgns %>%
  left_join(CEDO_wetland.data_5km_no.aqua_area_perOHI, by = c("rgn_id", "year")) %>%
  mutate(area_km2 = replace_na(area_km2, 0)) %>%
  select(rgn_id, 
         area_est_byCEDO.2025_km2= area_km2, 
         year)

```

```{r Ktable of Current Extent}
WETL_Current_Extent %>%
  kable(digits = 4, col.names = c("Region", "Extent (km²) (CEDO data, 2025; CONABIO Aquaculture data, 2021)", "Year")) %>%
  kable_styling(bootstrap_options = c("striped", "bordered")) %>%
  add_header_above(
    c('Wetland, Dune, & Scrubland Extent' = 3))

```

## Save WETL_Current_Extent
```{r Save WETL_Current_Extent}
write.csv(WETL_Current_Extent,
          here::here("_habitats/HAB_wetlands/WET_outputs/WETL_Current_Extent_Oct072026.csv"),
          row.names = FALSE)

```

## SANITY CHECKS on CEDO_wetland.data_5km_no.aqua_area_perOHI

```{r Compare to estimates from WETL_area_conabio}
# Compare out WETL esitmates versus Hem / CEDO wetland layer:
CEDO_wetland.data_5km_no.aqua_area_perOHI

colnames(WETL_area_conabio)

compare <- WETL_area_conabio %>%
  filter(year == 2021 ) %>%
  select(rgn_id, CONABIO_area = observed_area_km2) %>%
  left_join(CEDO_wetland.data_5km_no.aqua_area_perOHI, by="rgn_id")

```

```{r Plot Hem / CEDO wetland data minus aquaculture}
# Add dummy column directly to the geometry layer
CEDO_wetland.data_5km_no.aqua_area$layer <- "Wetlands"

tmap_mode("plot")

tm_shape(ohi_regions) +
  tm_polygons(
    col = "location",
    palette = c("marine" = "#cceeff",
                "land"   = "#f5e6c8"),
    alpha = 0.5,
    border.col = "gray",
    border.lwd = 1,
    legend.show = FALSE
  ) +

tm_shape(CEDO_wetland.data_5km_no.aqua_area) +
  tm_polygons(
    col = "layer",
    palette = c("Hem Minus Aquaculture" = "#228B22"),
    alpha = 1,
    border.col = "#228B22",
    border.lwd = 2,
    title = "Habitat"
  ) +

tm_layout(
    title = "Hem / CEDO Wetland data",
    title.size = 1.2,
    title.fontface = "bold",
    title.position = c(0.02, 0.98),
    frame = FALSE
  ) +
tm_scale_bar(position = c("left", "bottom"))


```

```{r Plot WETL_conabio_2021_polygons}
# Add dummy column directly to the geometry layer
WETL_conabio_2021_polygons$layer <- "Wetlands (DUN + SAL)"

tmap_mode("plot")

tm_shape(ohi_regions) +
  tm_polygons(
    col = "location",
    palette = c("marine" = "#cceeff",
                "land"   = "#f5e6c8"),
    alpha = 0.5,
    border.col = "gray",
    border.lwd = 1,
    legend.show = FALSE
  ) +

tm_shape(WETL_conabio_2021_polygons) +
  tm_polygons(
    col = "layer",
    palette = c("CONABIO Habitat 2021" = "#228B22"),
    alpha = 1,
    border.col = "#228B22",
    border.lwd = 2,
    title = "Habitat"
  ) +

tm_layout(
    title = "CONABIO wetland + Dune Habitat (2021)",
    title.size = 1.2,
    title.fontface = "bold",
    title.position = c(0.02, 0.98),
    frame = FALSE
  ) +
tm_scale_bar(position = c("left", "bottom"))

```

```{r Plot overlap with aquaculture layer on ESRI}
tmap_mode("view")

tm_basemap("Esri.WorldImagery") +

tm_shape(CEDO_wetland.data_5km) +
  tm_polygons(
    col = "#377eb8",
    alpha = 0.5,
    border.col = "#377eb8",
    group = "CEDO wetland, clipped to 5km (before removal)"
  ) +

tm_shape(CEDO_wetland.data_5km_no.aqua_area) +
  tm_polygons(
    col = "#e41a1c",
    alpha = 0.7,
    border.col = "#e41a1c",
    group = "CEDO wetland, clipped to 5km (after removing CONABIO Aquaculture layer)"
  ) +

tm_shape(Aquaculture_area_conabio_2021) +
  tm_polygons(
    col = "#4daf4a",
    alpha = 0.6,
    border.col = "#4daf4a",
    group = "Aquaculture"
  )

```


# STEP 5: FINALIZE WETL_Scores DATAFRAME ------------------------------

```{r Merge CEDO current extent estimates to MANG_Scores0}
colnames(WETL_Scores0)
colnames(WETL_Current_Extent)

# Merge:
WETL_Scores <- WETL_Scores0 %>%
  left_join(select(WETL_Current_Extent,
                   c("rgn_id", "area_est_byCEDO.2025_km2")),
            by="rgn_id") %>%
  rename(
    area_est_byCONABIO_km2 = estimated_area_km2
  )
  
# Make an include / exclude column based on the information reported by Hem / CEDO:
WETL_Scores<- WETL_Scores %>%
  mutate(WETL_include_exclude = ifelse(area_est_byCEDO.2025_km2 == 0, "EXCLUDE", 
                                      "INCLUDE"))

# Relevel:
WETL_Scores <- WETL_Scores %>%
  mutate(
    provenance = factor(
      provenance,
      levels = c("observed", "gapfilled")
    )
  )

```

```{r Make a nice Ktable}
WETL_Scores %>%
  select(rgn_id, year, WETL_current_status) %>%
  kable(digits = 4, col.names = c("Region", "Year", "Current Status (vs. 2001)")) %>%
  kable_styling(bootstrap_options = c("striped", "bordered")) %>%
  add_header_above(
    c('Wetland Dune & Scrubland Current Status' = 3))

```

```{r Join rgn_name to scores}
colnames(Region_Key)
colnames(WETL_Scores)

WETL_Scores <- WETL_Scores %>%
  left_join(Region_Key, by = "rgn_id")
  
WETL_Scores<- WETL_Scores %>%
  relocate(rgn_name, .before = rgn_id)
```


## Save WETL_Scores
```{r Save WETL_Scores}
write.csv(WETL_Scores,
          here::here("_habitats/HAB_wetlands/WET_outputs/WETL_Scores_1.1_df_Oct072026.csv"),
          row.names = FALSE)
```

## Plot WETL_Scores

```{r Plot WETL_Scores}
colnames(WETL_Scores)

Region_colors<- c(
  "Western Upper Gulf" = "#C03729",
  "Western Midriff Islands" = "#E68C7C",
  "Western Central Gulf" = "#FC8F24",
  "Western Southern Gulf" = "#dcaa38",
  "Eastern Upper Gulf" = "#cbd2a0",
  "Eastern Midriff Islands" = "#98a25a",
  "Southern Sonora Coast" = "#747428",
  "Sinaloa Coast" = "#627391",
  "Southeastern Gulf" = "#8d61a9")

# Plot:
WETL_Scores_OverTime<- WETL_Scores %>%
  ggplot(aes(year, WETL_current_status, color = rgn_name)) +
  
  geom_line(linewidth = 1) +
  
  geom_point(
  aes(shape = provenance, fill = provenance),
  size = 2
  ) +
  
  scale_shape_manual(
    values = c(
      "gapfilled" = 21,
      "observed" = 16
    )
  ) +
  
  scale_fill_manual(
    values = c(
      "gapfilled" = "white",
      "observed" = NA
    )
  ) +
  
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey50") +
  
  ylim(0, 101) +
  
  scale_color_manual(values = Region_colors) +
  
  labs(
    title = "WETL Component Scores (1.1) by Region",
    subtitle = "Data sources: CONABIO (2021)",
    caption = "updated: Oct 7 2026",
    x = "Year",
    y = "WETL Score (from 0 to 100)",
    color = "Region",
  ) +
  
  theme_bw(base_size = 14) +
  theme(
    strip.text = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 16, face = "bold")
  )

WETL_Scores_OverTime

```

```{r Save as PNG}
ggsave(
  filename = sprintf("/home/ferrer/ohi_goc_prep/_habitats/HAB_wetlands/WETL_Scores_1.1_plot_Oct072026.png"),
  plot = WETL_Scores_OverTime,
  width = 12,
  height = 12,
  dpi = 300
)
```


# STEP 6: ESTIMATE WETL_Pressure, the inverse of WETL_current_status ------------------------------

```{r Estimate pressure as 100 minus current status}
WETL_Pressure <- WETL_Scores %>%
  mutate(pressure = (100 - WETL_current_status)/100) %>%
  select(rgn_name, rgn_id, year, pressure)

```

## Save WETL_Pressure
```{r Save WETL_Pressure}
write.csv(WETL_Pressure,
          here::here("_habitats/HAB_wetlands/WET_outputs/WETL_Pressure_Oct072026.csv"),
          row.names = FALSE)
```


# STEP 7: ESTIMATE WETL TREND (source: CONABIO, 2010-2020) ------------------------------

```{r Plot wetland trends}
colnames(WETL_Scores)
unique(WETL_Scores$rgn_name)

Region_colors<- c(
  "Western Upper Gulf" = "#C03729",
  "Western Midriff Islands" = "#E68C7C",
  "Western Central Gulf" = "#FC8F24",
  "Western Southern Gulf" = "#dcaa38",
  "Eastern Upper Gulf" = "#cbd2a0",
  "Eastern Midriff Islands" = "#98a25a",
  "Southern Sonora Coast" = "#747428",
  "Sinaloa Coast" = "#627391",
  "Southeastern Gulf" = "#8d61a9")

WETL_Area_OverTime<- WETL_Scores %>%
  filter(year >= 2000, year <= 2025) %>%
  filter(provenance == "observed") %>%
  
  ggplot(aes(
    x = year,
    y = area_est_byCONABIO_km2,
    color = rgn_name
  )) +
  

  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  
  labs(
    title = "WETL (Wetland, Scrubland, Dune) Area by Region",
    subtitle = "Source: CONABIO land-use data 2001-2021",
    caption = "updated: Oct 7 2026",
    y = "Area (km2)",
    x = "Year"
  ) +
  
  scale_color_manual(values = Region_colors) +
  
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 16, face = "bold")
  )

WETL_Area_OverTime

```

## Save WETL_Area_OverTime

```{r Save as PNG}
ggsave(
  filename = sprintf("/home/ferrer/ohi_goc_prep/_habitats/HAB_wetlands/WET_outputs/WETL_OverTime_plot_Oct072026.png"),
  plot = WETL_Area_OverTime,
  width = 12,
  height = 12,
  dpi = 300
)
```

## METHOD A

```{r Calculate WETL_Trend using METHOD A}
colnames(WETL_Scores)

# METHOD A: 
# sSee GitHub issue, Oct 7 2026, for more details: https://github.com/OHI-GOC/OHI_GOC_issues/issues/127#issuecomment-6041025924
# slope <- (condition_end - condition_start) / n_years
# trend_pct_yr <- (slope / condition_start) * 100

# Calculate the trend as the linear annualized percent change between the most recent year (i.e., 2021) and 5-10 years prior (i.e., 2013)
WETL_Trend0 <- WETL_Scores %>%
  # filter(WETL_include_exclude == "INCLUDE") %>%
  filter(provenance == "observed") %>%
  group_by(rgn_name, rgn_id) %>%
  mutate(
    recent_year = max(year, na.rm = TRUE),
    comparison_year = min(
      year[year < recent_year &
           year >= recent_year - 10 &
           year <= recent_year - 5]
    ),
    recent_area = area_est_byCONABIO_km2[year == recent_year][1],
    comparison_area = area_est_byCONABIO_km2[year == comparison_year][1],
    n_years = recent_year - comparison_year,
    slope = (recent_area - comparison_area) / n_years,
    trend = round((slope / comparison_area) * 100, 2)
  ) %>%
  filter(year == recent_year) %>%
  ungroup() %>%
  select(
    rgn_name,
    rgn_id,
    comparison_year,
    comparison_area,
    most_recent_year= recent_year,
    most_recent_area= recent_area,
    trend_linear_pct_year_opA = trend
  )

WETL_Trend0

```

## METHOD B

```{r Calculate WETL_Trend using METHOD B}
# METHOD B: 
# See GitHub issue, Oct 7 2026, for more details: https://github.com/OHI-GOC/OHI_GOC_issues/issues/127#issuecomment-6041025924
# slope = fit a linear model: lm(area ~ year), grab the slope estimate
# trend_pct_year <- (1 + slope/earliest_area_in_lm)

# Pull the most recent year and the comparison year:
most_recent_year = max(WETL_Scores$year, na.rm = TRUE)

comparison_year <- min(
  WETL_Scores$year[
    WETL_Scores$provenance == "observed" &
    WETL_Scores$year < most_recent_year &
    WETL_Scores$year >= most_recent_year - 10 &
    WETL_Scores$year <= most_recent_year - 5
  ],
  na.rm = TRUE
)

WETL_area_lm <- WETL_Scores %>%
  # filter(WETL_include_exclude == "INCLUDE") %>%
  filter(
    year >= comparison_year,
    year <= most_recent_year,
    provenance == "observed"
  ) %>%
  group_by(rgn_id) %>%
  summarise(
    model = list(
      lm(area_est_byCONABIO_km2 ~ year, data = pick(everything()))
    ),
    comparison_area = area_est_byCONABIO_km2[year == comparison_year][1],
    .groups = "drop"
  ) %>%
  mutate(
    slope = sapply(model, function(x) coef(x)[["year"]]),
    p_value = sapply(
      model,
      function(x) summary(x)$coefficients["year", "Pr(>|t|)"]
    ),
    trend_pct_year = round((slope / comparison_area) * 100, 2)
  ) %>%
  select(rgn_id, 
         trend_linear_pct_year_opB = trend_pct_year,
         lm_slope= slope, 
         p_value)

```

```{r Merge options A and B}
WETL_Trend<- WETL_Trend0 %>%
  left_join(
    WETL_area_lm, 
    by= "rgn_id"
  )
```


```{r Make a nice Ktable for WETL_Trend}
colnames(WETL_Trend)

WETL_Trend %>%
  select(rgn_id, trend_linear_pct_year_opA, trend_linear_pct_year_opB) %>%
  kable(digits = 4, col.names = c("Region", "Trend (method A)", "Trend (method B)")) %>%
  kable_styling(bootstrap_options = c("striped", "bordered")) %>%
  add_header_above(
    c('Wetland & Scrubland Trend' = 3))
```

## Save WETL_Trend
```{r Save WETL_Trend}
write.csv(WETL_Trend,
          here::here("_habitats/HAB_wetlands/WET_outputs/WETL_Trend_Oct072026.csv"),
          row.names = FALSE)
```


