---
title: 'OHI GoC 2024: Lasting Special Places (LSP) Native Lands'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: inline
editor_options: 
  chunk_output_type: inline
---

# Native Lands in the Gulf of California

Slug present after the name. 

Example: for Pericu, it is *pericu*. `Sample Request:https://native-land.ca/api/index.php?maps=territories&name=pericu`

### Baja California + BCS

- Pericu: pericu
- Guaycura: guaycura
- Monqui: monqui
- Comom'ti-pa (Cochimi): cochimi
- Xawiƚƚ kwñchawaay (Cocopah): cocopah

### Sonora

- Hia-Ced O’odham: hia-ced-oodham
- O’odham Jeweḍ: tohono-oodham
- Comcaac (Seri): seri
- Yoeme (Yaqui): yaqui

- Tohono O’odham: papago-tohono-oodham

### Sinaloa

- Cahita: cahita
- Yoreme (Mayo): mayo

- Achires: achires
- Tahues: tahues
- Acaxee: acaxee
- Totorames: totorames
- Tepehuan: tepehuan-2

### Nayarit

- Wixárika (Huichol): huichol-wixaritari
- Náayerite (Cora): cora


```{r}
library(tidyverse)
library(here)
library(janitor)
library(httr)
library(jsonlite)

library(tidyverse)
library(httr)
library(jsonlite)
library(sf)

# --- Your API key, from Native-Land.ca ---
api_key <- "0d0mnukIMvZ-ZC9Yg-64c"  # from this dashboard: https://native-land.ca/dashboard/api

base_url <- "https://native-land.ca/api/index.php"

# Function: get territory by slug -> sf
get_territory_by_slug_sf <- function(slug) {
  res <- GET(
    base_url,
    query = list(
      maps = "territories",
      name = slug,
      key = api_key
    )
  )
  
  if (status_code(res) != 200) {
    warning(paste("API request failed for:", slug))
    return(NULL)
  }
  
  json_data <- content(res, as = "text", encoding = "UTF-8") %>%
    fromJSON(flatten = TRUE)
  
  if (length(json_data) == 0) {
    warning(paste("Empty data for:", slug))
    return(NULL)
  }
  
  # If no geometry present, skip
  if (!"geometry" %in% names(json_data)) {
    warning(paste("No geometry for:", slug))
    return(NULL)
  }
  
  # Convert to sf
  tryCatch({
    sf_obj <- st_as_sf(json_data, crs = 4326)
    sf_obj$slug <- slug
    return(sf_obj)
  }, error = function(e) {
    warning(paste("Error converting to sf for:", slug, "-", e$message))
    return(NULL)
  })
}

# Slugs
slugs <- c(
  "pericu", "guaycura", "monqui", "cochimi", "cocopah",
  "hia-ced-oodham", "tohono-oodham", "seri", "yaqui", "papago-tohono-oodham",
  "cahita", "mayo", "achires", "tahues", "acaxee", "totorames", "tepehuan-2",
  "huichol-wixaritari", "cora"
)

# Fetch all
territories_sf_list <- map(slugs, get_territory_by_slug_sf) %>%
  compact()

# Combine into single sf if any succeeded
if (length(territories_sf_list) > 0) {
  territories_sf <- do.call(rbind, territories_sf_list)
  print(territories_sf)
  plot(territories_sf["properties.Name"])
} else {
  message("No territories returned with geometry.")
}

library(mapview)
mapview(territories_sf)
```

# Trying by position: slug queries do not produce geometry.  

```{r}
library(httr)
library(jsonlite)
library(sf)
library(dplyr)
library(mapview)
library(here)
library(rmapshaper)

goc_poly <- st_read(here("spatial/ohi_regions","goc_ohi_rgns.shp"))
st_crs(goc_poly) # not WGS84: Mexico ITRF2008 / LCC
mapview(goc_poly)

# reproject to WGS84 (required by API)
goc_poly <- st_transform(goc_poly, 4326)

# in order for this to run and not get error 502, we have to reduce the OHI rgns polygon to ~5% of original vertices
goc_poly_simple <- rmapshaper::ms_simplify(goc_poly, keep = 0.05, keep_shapes = TRUE)
mapview(goc_poly_simple)

# convert simplified polygon to GeoJSON
goc_geojson <- geojsonsf::sf_geojson(goc_poly_simple)
goc_fc <- jsonlite::fromJSON(goc_geojson, simplifyVector = FALSE)


# the request
req_body <- list(
  key = "0d0mnukIMvZ-ZC9Yg-64c", # the API key
  maps = "territories", # could also be "territories,languages,treaties"
  polygon_geojson = goc_fc
)

# POST request
res <- httr::POST(
  url = "https://native-land.ca/api/index.php",
  body = jsonlite::toJSON(req_body, auto_unbox = TRUE),
  encode = "raw",
  httr::content_type_json()
)

if (status_code(res) == 200) {
  geojson_txt <- content(res, as = "text", encoding = "UTF-8")
  
  # convert GeoJSON string into an sf object directly
  territories_sf <- geojsonsf::geojson_sf(geojson_txt)
  
  print(territories_sf)
  plot(territories_sf["Name"])
} else {
  message("Request failed with status: ", status_code(res))
}


mapview(territories_sf)

```

## Make a kable

```{r}
library(kableExtra)

territories_sf %>%
  st_drop_geometry() %>%
  janitor::clean_names() %>%
  kable("html", caption = "Territories returned by Native-Land API that intersect with OHI GoC Regions") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  row_spec(0, bold = TRUE, color = "white", background = "#3C6E71")
```

