---
title: 'OHI GoC 2024: Lasting Special Places (LSP): Cultural Centers'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# Ministry of Culture: Directorio de centros culturales y casas de cultura

## Clean the datasets

Note that while many of these centers have longitude and latitude, some are at 0,0 though they have specified municipality locations.  So I will have to identify their location from their federal entity, municipality, and locality.

```{r}
library(tidyverse)
library(magrittr)
library(here)
library(janitor)
library(tmaptools)
library(sf)
library(mapview)
library(tidygeocoder)

## cc = centros culturales / casas de cultura
cc_raw <- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/ministry_of_culture","0_centro_cultural_directorio.csv"))

## mu = museos
mu_raw <- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/ministry_of_culture","museos.csv"))
  
## mon = monumentos registrados 
mon_raw <- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/ministry_of_culture","INAH_6_monumentos_registrados.csv"))
  
## uni = universidades
uni_raw <- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/ministry_of_culture","0_universidad_directorio.csv"))

## gal = galleria 
gal_raw <- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/ministry_of_culture","0_galeria_directorio.csv"))

## cdi = centro desarrollo indigena
cdi_raw <- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/ministry_of_culture","0_centro_desarrollo_indigena_directorio.csv"))

## ce = centro educacion
ce_raw <- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/ministry_of_culture","0_centro_educacion_directorio.csv"))

## ca = casa artesania
ca_raw <- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/ministry_of_culture","0_casa_artesania_directorio.csv"))

## cat = catedrals
cat_raw <- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/ministry_of_culture","0_catedral_directorio.csv"))
```

### Filter out CCs without long/lat

We will use tmaptools::geocode_OSM() to obtain coordinates for cities.

```{r}
## first, lets filter out all the CCs that do not have coordinates.
cc_no_coords <- cc_raw %>% 
  filter(longitud == "0") # if either long or lat are zero, the other one seems to be as well.
```


Let's try using tmaptools::geocode_OSM to get coordinates for localidades o municipios:

```{r}
## now lets try obtaining coordinates for these based on the localidad (more granular than municipio)
# cc_new_coords <- cc_no_coords %>% 
#   group_by(nom_mun) %>% 
#   mutate(geometry = purrr::map(nom_mun, ~ st_geometry(geocode_OSM(.x, as.sf = TRUE)))) %>%
#       unnest(geometry) %>%
#       st_as_sf()
```

Did not work, too many to geocode. Trying `tidygeocoder`:

```{r}
devtools::install_github("jessecambon/tidygeocoder")
library(dplyr, warn.conflicts = FALSE)
library(tidygeocoder)
library(mapview)

cc_new_coords <- cc_no_coords %>%
  geocode(nom_loc, method = 'osm', lat = latitude , long = longitude) # 411.9 seconds!

cc_new_coords_na <- cc_new_coords %>% 
  mutate(coord_na = ifelse(is.na(longitude) | is.na(latitude), TRUE, FALSE)) %>% 
  filter(coord_na == TRUE) # if NA in either longitude or latitude, filter

na_id <- unique(cc_new_coords_na$centro_cultural_id)

cc_new_coords_sf <- cc_new_coords %>% 
  select(-c(longitud, latitud)) %>% 
  filter(!centro_cultural_id %in% na_id) %>% 
  st_as_sf(coords = c("longitude", "latitude"),
           crs = 4326, # WGS84
           remove = FALSE) # convert to sf object

mapview(cc_new_coords_sf)
```

Okay, so there were issues using nom_loc because sometimes they did not make sense out of context.  For example, San Antonio was what the Nominatim (“osm”) geocoding service used, but it is actually in La Paz:
<!-- 1	centro_cultural_id 1635  -->
<!-- 2	centro_cultural_nombre Casa de Cultura de San Antonio  -->
<!-- 3	centro_cultural_fecha_fundacion	1972-01-01 -->
<!-- 4	centro_cultural_adscripcion	Ayuntamiento de La Paz -->
<!-- 5	estado_id 3 -->
<!-- 6	nom_ent Baja California Sur -->
<!-- 7	municipio_id 3 -->
<!-- 8	nom_mun	La Paz -->
<!-- 9	localidad_id 766 -->
<!-- 10	nom_loc San Antonio -->

```{r}
## make a combined name for the place of nom_ent, nom_mun, and nom_loc

# unite columns for one name, "nom"
cc_no_coords_nom <- cc_no_coords %>% 
  unite("nom", nom_ent, nom_mun, nom_loc, sep = ",", remove = FALSE)

# with the nom column, use tidygeocoder::geocode to obtain long/lat coords
cc_new_coords_nom <- cc_no_coords_nom %>%
  geocode(nom, method = 'osm', lat = latitude , long = longitude) # 414.3 seconds!

# remove NAs but keep track of them to add them back later
cc_new_coords_na_nom <- cc_new_coords_nom %>% 
  mutate(coord_na = ifelse(is.na(longitude) | is.na(latitude), TRUE, FALSE)) %>% 
  filter(coord_na == TRUE) # if NA in either longitude or latitude, filter

na_id_nom <- unique(cc_new_coords_na_nom$centro_cultural_id) # these are all not in GoC states, so we can drop them 

cc_new_coords_sf_nom <- cc_new_coords_nom %>% 
  select(-c(longitud, latitud)) %>% 
  filter(!centro_cultural_id %in% na_id_nom) %>% 
  st_as_sf(coords = c("longitude", "latitude"),
           crs = 4326, # WGS84
           remove = FALSE) # convert to sf object

mapview(cc_new_coords_sf_nom) # yay!! none outside of Mexico! 
```

### Combine new coordinate CCs with ones that already had lat/longs

```{r}
## first, let's make the ones that already had lat/longs into sf objects
cc_sf <- cc_raw %>% 
  filter(longitud != "0") %>% 
  filter(centro_cultural_id != "1063") %>% # drop 1063 because it is in Zacatecas, Sain Alto, and it has incorrect long/lats
  st_as_sf(coords = c("longitud", "latitud"),
           crs = 4326, # WGS84
           remove = FALSE) # convert to sf object

mapview(cc_sf) # yay!! none outside of Mexico! 
```

Now, we combine them:

```{r}
cc_all_sf <- bind_rows(cc_sf, cc_new_coords_sf_nom)  # bind them!
summary(st_coordinates(cc_all_sf)) # not all are within 180 and 90.  Specifically, this is happening in longitud/latitud, not the longitude/latitudes.

# which are outside of the bounding box?
cc_all_sf_out <- cc_all_sf %>% 
  mutate(bad_coord = case_when(
    longitud < -180 | longitud > 180 ~ TRUE,
    latitud < -90  | latitud > 90  ~ TRUE,
    TRUE ~ FALSE
  )) %>% 
  filter(bad_coord == TRUE) # these are not in GoC.  Just drop them.

# the ids for the CCs outside of the bounding box, which are not in the GoC.
out_id <- unique(cc_all_sf_out$centro_cultural_id)

# filter them out
cc_all_clean_sf <- cc_all_sf %>% 
  filter(!centro_cultural_id %in% out_id)

# map it!
mapview(cc_all_clean_sf) # yay!! no more warnings.
```

## Filter for only GoC CCs 

First, lets bring in the most recent OHI GoC regions, including inland areas.

```{r}
ohi_rgns <- st_read(here("spatial/ohi_regions","goc_ohi_rgns.shp"))
st_crs(ohi_rgns) # not WGS84: Mexico ITRF2008 / LCC

mapview(ohi_rgns)
```

Now, let's filter for only CCs in the OHI rgn area.

```{r}
## make st_crs(cc_all_clean_sf) == st_crs(ohi_rgns)
cc_all_clean_sf_ohicrs <- st_transform(cc_all_clean_sf, st_crs(ohi_rgns))

# check it
st_crs(cc_all_clean_sf_ohicrs) == st_crs(ohi_rgns) # TRUE

# intersect to only keep the points within the ohi_rgns polygon.
cc_goc_sf <- st_intersection(cc_all_clean_sf_ohicrs, ohi_rgns)

mapview(cc_goc_sf) # great!!
```

## Explore the cultural centers 

```{r}
unique(cc_goc_sf$centro_cultural_nombre) # 324, many are casas or centros

unique(cc_goc_sf$centro_cultural_fecha_fundacion) # the dates the CCs were founded

colnames(cc_goc_sf)
```
I want to make a plotly to see how many were founded per year:
```{r}
library(lubridate)
library(plotly)

# prep for time series
cc_goc_ts <- cc_goc_sf %>% 
  mutate(centro_cultural_fecha_fundacion = ymd(centro_cultural_fecha_fundacion))

# count CCs per year
cc_goc_year <- cc_goc_ts %>%
  mutate(year = year(centro_cultural_fecha_fundacion)) %>%
  filter(!is.na(year)) %>%
  count(year) %>% 
  st_drop_geometry()

# interactive plotly line chart
fig <- plot_ly(cc_goc_year,
               x = ~year,
               y = ~n,
               type = 'scatter',
               mode = 'lines+markers',
               line = list(color = 'steelblue'),
               marker = list(size = 6)) %>%
  layout(title = "Cultural centers (CCs) founded per year",
         xaxis = list(title = "Year"),
         yaxis = list(title = "Number of CCs"))
fig
```



# Doing the same for all the datasets

```{r}
## function to clean and geocode any dataset
clean_places <- function(df, id_col, lon_col, lat_col, ent_col, mun_col, loc_col, place_type) {
  
  ## 1. separate into those with and without coords
  df_no_coords <- df %>% 
    filter(.data[[lon_col]] == 0 | .data[[lat_col]] == 0)
  
  df_yes_coords <- df %>%
    filter(.data[[lon_col]] != 0 & .data[[lat_col]] != 0)
  
  ## 2. build combined location string for geocoding
  df_no_coords_nom <- df_no_coords %>%
    unite("nom", all_of(c(ent_col, mun_col, loc_col)), sep = ",", remove = FALSE)
  
  ## 3. geocode (can take a while!)
  df_new_coords_nom <- df_no_coords_nom %>%
    geocode(nom, method = 'osm', lat = "latitude", long = "longitude")
  
  ## 4. drop failed geocodes
  na_ids <- df_new_coords_nom %>%
    filter(is.na(longitude) | is.na(latitude)) %>%
    pull(!!sym(id_col))
  
  df_new_coords_sf <- df_new_coords_nom %>%
    filter(!.data[[id_col]] %in% na_ids) %>%
    st_as_sf(coords = c("longitude", "latitude"), crs = 4326, remove = FALSE)
  
  ## 5. convert original coords into sf
  df_yes_coords_sf <- df_yes_coords %>%
    st_as_sf(coords = c(lon_col, lat_col), crs = 4326, remove = FALSE)
  
  ## 6. bind together and filter out invalid coords
  df_all_sf <- bind_rows(df_yes_coords_sf, df_new_coords_sf) %>%
    mutate(bad_coord = case_when(
      .data[[lon_col]] < -180 | .data[[lon_col]] > 180 ~ TRUE,
      .data[[lat_col]] < -90  | .data[[lat_col]] > 90  ~ TRUE,
      TRUE ~ FALSE
    )) %>%
    filter(!bad_coord) %>%
    mutate(type = place_type)  ## add column for place type
  
  return(df_all_sf)
}

```


```{r}
## cultural centers -- works, ofc!!
cc_clean <- clean_places(cc_raw, 
                         id_col = "centro_cultural_id", 
                         lon_col = "longitud", 
                         lat_col = "latitud", 
                         ent_col = "nom_ent", 
                         mun_col = "nom_mun", 
                         loc_col = "nom_loc", 
                         place_type = "Centro Cultural")

## museums -- worked!!
mu_clean <- clean_places(mu_raw, 
                         id_col = "museo_id", 
                         lon_col = "longitud", 
                         lat_col = "latitud", 
                         ent_col = "nom_ent", 
                         mun_col = "nom_mun", 
                         loc_col = "nom_loc", 
                         place_type = "Museo")

## registered monuments: -- different set up
mon_int <- mon_raw %>% 
  mutate(monumento_id = row_number()) 

mon_clean <- clean_places(mon_int, 
                          id_col = "monumento_id", 
                          lon_col = "longitud", 
                          lat_col = "latitud", 
                          ent_col = "entidad_federativa", 
                          mun_col = "municipio_alcaldia", # municipality city hall
                          loc_col = "localidad_colonia", 
                          place_type = "Monumento")

## universities -- works!!
uni_clean <- clean_places(uni_raw, 
                          id_col = "universidad_id", 
                          lon_col = "longitud", 
                          lat_col = "latitud", 
                          ent_col = "nom_ent", 
                          mun_col = "nom_mun", 
                          loc_col = "nom_loc", 
                          place_type = "Universidad")

## galleries -- NO zeros in lat/long, func not needed!!
gal_clean <- gal_raw %>% 
  mutate(type = "Galería") %>% 
  st_as_sf(coords = c("longitud", "latitud"), crs = 4326, remove = FALSE)

## indigenous development centers -- worked!!
cdi_clean <- clean_places(cdi_raw, 
                          id_col = "centro_desarrollo_indigena_id", 
                          lon_col = "longitud", 
                          lat_col = "latitud", 
                          ent_col = "nom_ent", 
                          mun_col = "nom_mun", 
                          loc_col = "nom_loc", 
                          place_type = "Centro Desarrollo Indígena")

## education centers -- works!!
ce_clean <- clean_places(ce_raw, 
                         id_col = "centro_educacion_id", 
                         lon_col = "longitud", 
                         lat_col = "latitud", 
                         ent_col = "nom_ent", 
                         mun_col = "nom_mun", 
                         loc_col = "nom_loc", 
                         place_type = "Centro Educación")

## artisan houses -- workss!
ca_clean <- clean_places(ca_raw, 
                         id_col = "casa_artesania_id", 
                         lon_col = "longitud", 
                         lat_col = "latitud", 
                         ent_col = "nom_ent", 
                         mun_col = "nom_mun", 
                         loc_col = "nom_loc", 
                         place_type = "Casa Artesanía")

## cathedrals -- same error as gallerias
cat_clean <- cat_raw %>%
  mutate(type = "Cateral") %>% 
  st_as_sf(coords = c("longitud", "latitud"), crs = 4326, remove = FALSE)
```

First, group by region and unite the polygons for inland 10km with the marine polygons.
```{r}
## intersect with inland_10:
inland_10 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_10km_buffer","inland_10km_goc.shp"))
regions_marine <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp")) %>%
  filter(location == "marine")

st_crs(inland_10) == st_crs(regions_marine) # TRUE

marine_and_inland10 <- bind_rows(regions_marine, inland_10) %>% 
  dplyr::select(rgn_id, geometry) %>% 
  group_by(rgn_id) %>% 
  summarize(
    geometry = st_union(geometry)
  ) %>% 
  ungroup()

mapview(marine_and_inland10)
```

Now, we can intersect with the `marine_and_inland10` so that each cultural site is assigned an OHI GoC Region ID.
```{r}
all_places_sf <- bind_rows(cc_clean, mu_clean, uni_clean, cdi_clean, ce_clean, ca_clean,
                           gal_clean, cat_clean) %>% # mon_clean no latitud or longitud
  st_transform(., st_crs(marine_and_inland10)) %>%
  st_intersection(marine_and_inland10)

```

```{r}
mapview(all_places_sf, zcol = "rgn_id")
mapview(all_places_sf, zcol = "type") + mapview(marine_and_inland10)
```


```{r}
## add in protected areas
wdpa_prot <- st_read(here("LSP", "v2025", "output", "WDPA_Jan2025_shp_clipped_all.shp"))

mapview(all_places_sf, zcol = "type") + mapview(marine_and_inland10, col.regions = "pink") + mapview(wdpa_prot, color = "black")
```


<!-- # Quick look at ZOFEMAT for La Paz -->

<!-- ```{r} -->
<!-- el_sargento <- st_read(here("/home/shares/ohi/OHI_GOC/spatial/delimitaciones_zona_federal/03003_2024_02_EL_SARGENTO","03003_2024_02.shp")) -->
<!-- mapview(el_sargento) -->
<!-- ``` -->

