---
title: 'OHI GoC 2024: Lasting Special Places (LSP)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# Sentido de Pertenencia: Lasting Special Places

## Protected areas (ANPs)

This subgoal's current status is based on whether there is a protection on the Lasting Special Place (LSP), and the condition is then determined by the effectiveness of that protection on the LSP.  

For protected areas, they are all inherently protected, and are also all on the list from experts. The condition will be based on the management effectiveness determined by CONANP's SIMEC index of effectiveness (i-efectividad).

```{r}
library(tidyverse)
library(here)
library(sf)
library(mapview)
```


## Bring in protected areas

#### WDPA January 2025

```{r}
wdpa_2025 <- st_read(here::here("/home/lecuona/OHI_Intro/ohi_goc_prep/LSP/v2025/output/WDPA_Jan2025_shp_clipped_all.shp"))

mapview(wdpa_2025)

wdpa_2025_df <- wdpa_2025 %>% 
  st_drop_geometry()
```

`sub_loc` can help us make sure the protected area is in the correct region. 

```{r}

```




#### For more information on all PAs designated before 2020: CBMC GCMP 

```{r}
cbmc_gcmp_2020_mpas <- st_read(here::here("/home/shares/ohi/OHI_GOC/_raw_data/CBMC-GCMP-prioritization-of-diving-sites-main/shp/final_mpa_layer.shp")) %>%
  janitor::clean_names()

# get rid of Z coordinates to make all geometries 2D
cbmc_gcmp_2020_mpas <- st_zm(cbmc_gcmp_2020_mpas, drop = TRUE, what = "ZM")

# verify the geometry types are now consistent
table(st_geometry_type(cbmc_gcmp_2020_mpas))

mapview(cbmc_gcmp_2020_mpas)
```


### Assign OHI GoC Regions


## i-efectividad for each CONANP protected area

```{r}
i_efectividad_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SIMEC/identifying_lasting_special_places_SIMEC.csv")) %>% 
  janitor::clean_names()

```

```{r}
ewg_protected_areas_list <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/LSP_list_ewg_mentioned_protected_areas_v1_042426.csv")) %>% 
  janitor::clean_names()
```

```{r}
i_efectividad_w_ewg_list <- full_join(i_efectividad_raw, ewg_protected_areas_list, by = c("nombre" = "i_efectividad_name"))
```



## RAMSAR site


### Join all together

Join the i-efectividad with each protected area, and the condition for the ramsar sites.


## Avg value per region


