---
title: 'OHI GoC 2024: Lasting Special Places (LSP)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# Sentido de Pertenencia: Lasting Special Places

## Fishing sites

*Overview*
Fishing is an activity that can be done for subsistence, but it can also be done recreationally, or as part of traditional/cultural engagement. Ideally, fishing grounds are well managed so the locations to fish can be maintained.

We will determine the condition of whether the fishing site is "protected" to last into the future based on whether there is an active zona de refugio pesquero (ZRP) or fishing cooperative (TURF) in that location.  

Each fishing site was named by the Expert Working Group.


```{r}
library(tidyverse)
library(here)
library(sf)
library(mapview)
```


## Read in fishing sites

```{r}
lsp_list_ewg_v3 <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/identifying_lasting_special_places_collapsed_lsp_protections_v3.csv")) %>% 
  janitor::clean_names()

ewg_fishing_site_list <- lsp_list_ewg_v3 %>% 
  filter(protection_should_be %in% c("TURFs + refugia", "TURFs + fishing refuge (specific to squid?)","TURFs + fishing refuge (specific to clam?)")) %>%
  mutate(long = as.numeric(long),
         lat = as.numeric(lat)) %>% 
  filter(!is.na(long)) %>% 
  st_as_sf(
    coords = c("long", "lat"), # making sf object
    crs = 4326, # WGS84
    remove = FALSE
  )

mapview(ewg_fishing_site_list)
```


## Determine if priority fishing sites fall within ZRPs or TURFs

## Condition for each fishing site

# Condition per region

## Write to Mazu

```{r}
write_csv(fishing_sites_condition, here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/output/fishing_sites_condition.csv"))
write_csv(rgn_fishing_sites_condition, here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/output/rgn_fishing_sites_condition.csv"))
```


