---
title: 'OHI GoC 2024: Lasting Special Places (LSP)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# Sentido de Pertenencia: Lasting Special Places

## Cultural and archaelogical sites

*Overview*
This includes archaeological sites, such as the shell middens near Puerto Peñasco. These locations have existence, cultural, and educational value. Condition of the archaeological site is determined by whether the agency, National Institute of Anthropology and History (INAH).

*Why the INAH?*
[From their website for archaeological sites:](https://www.inah.gob.mx/zonas-arqueologicas)

> "The National Institute of Anthropology and History (INAH) oversees 194 archaeological sites and one paleontological site open to the public throughout the country, 365 days a year. The following are exempt from the entrance fee: seniors over 60, children under 13, retirees, pensioners, people with disabilities, active teachers and students, as well as interns or researchers with INAH permission to conduct studies related to the museums, monuments, and archaeological sites referred to in this article. National and foreign visitors residing in Mexico who visit museums, monuments, and archaeological sites on Sundays are also exempt from this fee. (Art. 288 of the Federal Law of Fees.)"

[From their website for historical monuments:](https://monumentoshistoricos.inah.gob.mx/index.php)

> "The need to preserve, study, and disseminate the nation's historical heritage, expressed by the Federal Government of Lázaro Cárdenas, led to the creation of the National Institute of Anthropology and History (INAH) on February 3, 1939.

> One of INAH's primary functions was the oversight, conservation, and restoration of the Republic's archaeological, historical, and artistic monuments, as well as the objects found within them. To this end, the Department of Colonial Monuments was created and divided into two parts: the Directorate of Pre-Hispanic Monuments and the Directorate of Colonial Monuments.

> Around 1973, the Department of Colonial Monuments was incorporated into the newly created Directorate of Historical Monuments, and six years later it was dissolved.

> In 1989, the Directorate of Historical Monuments underwent a restructuring and became the National Coordination of Historical Monuments, whose specific tasks include the conservation, restoration, protection, cataloging, research, and dissemination of the nation's built historical heritage. To carry them out, it has the following areas: National Coordination, Technical Support Directorate, Authorizations and Inspections Directorate, Authorizations and Inspections Subdirectorate, Support for Registration and Linkage of Historical Monuments Subdirectorate, Administrative Subdirectorate, Catalog and Zones Subdirectorate, Research Subdirectorate, Supervision of External Projects and Works Subdirectorate, Information Technology Unit and the Unit for Attention to Buildings in the Face of Natural Disasters."


```{r}
library(tidyverse)
library(here)
library(sf)
library(mapview)
```


## Read in archaeological sites the EWG identified as special in the GoC

```{r}
lsp_list_ewg_v3 <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/identifying_lasting_special_places_collapsed_lsp_protections_v3.csv")) %>% 
  janitor::clean_names()

ewg_cultural_site_list <- lsp_list_ewg_v3 %>% 
  filter(protection_should_be %in% "INAH") %>% 
  mutate(long = as.numeric(long),
         lat = as.numeric(lat)) %>% 
  st_as_sf(
    coords = c("long", "lat"), # making sf object
    crs = 4326, # WGS84
    remove = FALSE
  )

mapview(ewg_cultural_site_list)
```

### Limit to the GoC 

First, read in the marine and land polygons.
```{r}
ohi_rgns <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp")) 

mapview(ohi_rgns, zcol = "rgn_id")
```

Now, we can intersect with `ohi_rgns` so that each site is assigned an OHI GoC Region ID.
```{r}
# intersect with GoC
ewg_cultural_site_list_goc_rgnid <- st_intersection(st_transform(ewg_cultural_site_list, st_crs(ohi_rgns)), ohi_rgns)

mapview(ewg_cultural_site_list_goc_rgnid, zcol = "rgn_id") 
```


## Bring in all sites in the GoC managed by the INAH

```{r}
INAH_aechaeological_sites <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/INAH_archaeological_sites_v2026.csv")) %>% 
  janitor::clean_names() %>% 
  mutate(long = as.numeric(long),
         lat = as.numeric(lat)) %>% 
  st_as_sf(
    coords = c("long", "lat"), # making sf object
    crs = 4326, # WGS84
    remove = FALSE
  )

mapview(ewg_cultural_site_list_goc_rgnid, col.regions = "blue") + mapview(INAH_aechaeological_sites, col.regions = "brown")
```


### Determine if there is overlap between the EWG and INAH sites

I chose to do this spatially, in case the names between the sites do not match up exactly.

```{r}
# buffer dive sites by 10000m
ewg_cultural_site_list_buffered <- st_buffer(
  st_transform(ewg_cultural_site_list_goc_rgnid, st_crs(INAH_aechaeological_sites)),
  dist = 10000
)

INAH_aechaeological_sites_buffered <- st_buffer(INAH_aechaeological_sites, dist = 10000)

mapview(ewg_cultural_site_list_buffered) + mapview(INAH_aechaeological_sites_buffered, col.regions = "red")
```


```{r}
# spatial join to the INAH site layer
ewg_archaeological_sites_protected_INAH <- st_join(
  ewg_cultural_site_list_buffered,
  INAH_aechaeological_sites_buffered,
  join = st_intersects,
  left = TRUE  # keep all sites, even those outside of INAH sites
) %>%
  mutate(in_INAH_directory = !is.na(nombre)) %>% 
  mutate(overall_condition = ifelse(in_INAH_directory == "TRUE", 100, 0)) 

ewg_archaeological_sites_condition <- ewg_archaeological_sites_protected_INAH %>% 
  dplyr::select( rgn_id, lasting_special_place_lsp, in_INAH_directory, nombre_INAH = nombre, overall_condition, link, significado, relevancia_cultural, costo_pesos_nonresidences) %>% 
  st_drop_geometry()

## only 2 are currently managed by INAH.
```

### Determine if there is overlap between the EWG sites and protected areas

#### Bring in protected areas

This aggregated layer was built using Hem's layer of ANPs, and WDPA's protected area polygons downloaded in January 2025.

```{r}
all_protected <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/protected _areas_wdpa_and_hem", "all_goc_protected_areas_2025.shp"))
mapview(all_protected)
```

### Intersect with protected areas

```{r}
## NEW BUFFER that is smaller:
# buffer dive sites by 10000m
ewg_cultural_site_list_buffered <- st_buffer(
  st_transform(ewg_cultural_site_list_goc_rgnid, st_crs(INAH_aechaeological_sites)),
  dist = 500
)

# spatial join to the non-dissolved layer (all_protected) to keep protected area names
ewg_archaeological_sites_protected_ANP <- st_join(
  st_transform(ewg_cultural_site_list_buffered, st_crs(all_protected)),
  all_protected %>% dplyr::select(name, geometry),
  join = st_intersects,
  left = TRUE  # keep all sites, even those outside of protected areas
) %>%
  mutate(in_protected_area = !is.na(name)) %>%
  rename(protected_area_name = name) 

# quick checks for sanity
mapview(ewg_archaeological_sites_protected_ANP, zcol = "in_protected_area") + mapview(all_protected, col.regions = "pink")
table(ewg_archaeological_sites_protected_ANP$in_protected_area)
table(ewg_archaeological_sites_protected_ANP$protected_area_name)
table(ewg_archaeological_sites_protected_ANP$lasting_special_place_lsp) # multiple observations for each site.
```

The only problem is that if a buffered site overlaps multiple protected areas, `st_join()` will return one row per match, giving us duplicate sites. So that is something to keep in mind.

#### i-efectividad for each dive site in a protected area

```{r}
i_efectividad_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SIMEC/identifying_lasting_special_places_SIMEC.csv")) %>% 
  janitor::clean_names()

unique(i_efectividad_raw$nombre)
unique(ewg_archaeological_sites_protected_ANP$protected_area_name)
```

#### Fuzzy matching the protected area names

```{r}
library(stringdist)
library(stringi)

# normalize protected area names...uppercase, remove accents, remove common filler words
normalize_name <- function(x) {
  x %>%
    stringi::stri_trans_general("Latin-ASCII") %>%  # remove accents
    str_to_upper() %>%
    str_remove_all("\\bPARQUE NACIONAL\\b|\\bPARQUE\\b|\\bNACIONAL\\b|\\bZONA MARINA\\b|\\bARCHIPIELAGO\\b") %>%
    str_squish()  # remove extra whitespace left behind
}

# make normalized versions for matching
i_efectividad_norm <- i_efectividad_raw %>%
  mutate(nombre_norm = normalize_name(nombre))

arch_sites_norm <- ewg_archaeological_sites_protected_ANP %>%
  mutate(protected_area_norm = normalize_name(protected_area_name))

# fuzzy match: for each dive site protected area name, find the closest match in i_efectividad
matches <- stringdist::amatch(
  arch_sites_norm$protected_area_norm,
  i_efectividad_norm$nombre_norm,
  method = "jw", # Jaro-Winkler works well for place names
  maxDist = 0.2 # adjust threshold as needed; lower = stricter
)
```


```{r}
# attach matched nombre from i_efectividad back to dive sites
arch_sites_matched <- arch_sites_norm %>%
  mutate(
    nombre_matched = i_efectividad_norm$nombre[matches],
    nombre_norm_matched = i_efectividad_norm$nombre_norm[matches]
  )

# inspect the matches before joining
# join them!!!
arch_sites_w_SIMEC <- arch_sites_matched %>%
  left_join(
    i_efectividad_raw,
    by = c("nombre_matched" = "nombre")
  ) %>% 
  mutate(protected_area_condition = ifelse(!is.na(i_efectividad), i_efectividad, NA),
         protected_area_condition = ifelse(protected_area_name == "Humedales de Bahía Adaír", 30, protected_area_condition), # Humedales de Bahía Adaír does not have a managment plan in place: https://rsis.ramsar.org/?f%5B0%5D=region_en_s%3ANorth+America&f%5B1%5D=managementPlanAvailable_i%3A0&q=Mexico#map 
         protected_area_condition = ifelse(lasting_special_place_lsp == "LA PIEDRA DE LA VIRGEN, / O PIEDRA DEL CRISTO DEL MAR.  Important for the huichol people, they recognize as the point of origin of their people (https://visitnayarit.travel/en/blog-nayarit-en/tatei-haramara-the-origin-of-the-world-according-to-the-wixarika-worldview/) (JGH)", 100, protected_area_condition), # "In November 2008, Tatei Haramara, along with other Wixarika sacred sites, was protected by governments through the signing of a common agreement called the ‘Hauxa Manaka Pact for the Preservation and Development of Wixarika Culture.’ A total of 6 hectares of this site were protected, committing to ‘prevent the Wixaritari and/or individuals not of their culture from using elements of the same for purposes contrary to the ceremonial and tradition of the Wixárika People." https://visitnayarit.travel/en/blog-nayarit-en/tatei-haramara-the-origin-of-the-world-according-to-the-wixarika-worldview/ https://www.wixarika.org/pronouncement-wixarika-sacred-site-route-wirikuta-cultural-and-natural-world-heritage-unesco https://www.wixarika.org/recognition-wixarika-sacred-route-will-avoid-pillage-peyote 
         protected_area_condition = ifelse(is.na(protected_area_condition), 0, protected_area_condition)
  ) %>%
  group_by(lasting_special_place_lsp) %>% 
  mutate(protected_area_condition = mean(protected_area_condition)) %>% 
  ungroup() %>% 
  dplyr::select(rgn_id, lasting_special_place_lsp, protected_area_condition) %>% 
  distinct()

## note that "LA PIEDRA DE LA VIRGEN, / O PIEDRA DEL CRISTO DEL MAR.  Important for the huichol people, they recognize as the point of origin of their people (https://visitnayarit.travel/en/blog-nayarit-en/tatei-haramara-the-origin-of-the-world-according-to-the-wixarika-worldview/) (JGH)" is fully protected by Hauxa Manaka Pact for the Preservation and Development of Wixarika Culture (2008) and UNESCO (as of 2025).
```

### Codified protection condition

From [Guerrero-Arenas et al. 2022:](https://paleontologiamexicana.geologia.unam.mx/index.php/paleontologia/article/view/165/299)

"The paleontological heritage of Mexico consists of the fossil specimens housed in scientific collections (movable heritage), as well as those found in fossiliferous localities throughout the country’s territory (immovable heritage). Since 1986, as stated in the Federal Law on Monuments and Archaeological, Artistic and Historical Zones, the National Institute of Anthropology and History is in charge of protecting fossils discovered in Mexico. Fossils are considered cultural heritage by law in several Latin American countries, such as Argentina, Chile, El Salvador, Peru, and Mexico. *However, this classification is not the most adequate because it does not guarantee its comprehensive protection.*"

"There is no law in Mexico defining the criteria for considering fossil sites as objects of protection; however, there are various instruments that allow their registration and therefore, a proposal for their long-term management and protection. Nonetheless, paleontologists working on Mexican territory must be commited to comply with the requirements already established by the law for the registration of palaeontological movable and immovable property. With this, it is possible to improve and adapt the instruments and legal processes that allow the safeguarding of Mexico’s paleontological heritage."

Therefore, I will be giving any archaeological site that is a managed site by INAH nor does it fall into a natural protected are a management condition score of 30 (relatively weak, but partial protection).

```{r}
ewg_archaeological_sites_condition_protected <- ewg_archaeological_sites_condition %>% 
  left_join(arch_sites_w_SIMEC) %>% 
  mutate(overall_condition = ifelse(overall_condition == 0, protected_area_condition, overall_condition)) %>% 
  mutate(overall_condition = ifelse(overall_condition == 0, 30, overall_condition)) # codified
```



## Condition per region

```{r}
final_ewg_archaeological_sites_condition <- ewg_archaeological_sites_condition_protected %>% 
  st_drop_geometry() %>% 
  dplyr::select(-geometry)

rgn_final_ewg_archaeological_sites_condition <- final_ewg_archaeological_sites_condition %>% 
  group_by(rgn_id) %>% 
  summarize(
    condition = mean(overall_condition, na.rm = TRUE)
  ) %>% 
  ungroup()
```

### Write to Mazu

```{r}
write_csv(final_ewg_archaeological_sites_condition, here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/output/ewg_archaeological_sites_condition.csv"))
write_csv(rgn_final_ewg_archaeological_sites_condition, here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/output/rgn_ewg_archaeological_sites_condition.csv"))
```








