---
title: 'OHI GoC 2025: Sense of Place CCE'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# Sense of Place 

## Coastal Community Engagement: Lasting non-natural cultural places


See `TR/denue_prep.qmd` for the download of `/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_esparcimiento_culturales_y_deportivos_servicios_recreativos/int/non_natural_cultural_places_2019_2024_goc_10km.shp`.


Read in the non-natural cultural places.

```{r}
library(tidyverse)
library(sf)
library(mapview)
library(here)
```


### Culture service establishments 


```{r}
non_natural_culture_gulf_10km <- st_read(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_esparcimiento_culturales_y_deportivos_servicios_recreativos/int/non_natural_cultural_places_2019_2024_goc_10km.shp"))

non_natural_culture_gulf_10km_clean <- non_natural_culture_gulf_10km %>% 
  #st_drop_geometry() %>% 
  mutate(nombre_act = nombre_act %>% 
           str_replace_all("<fa>", "ú") %>%
           str_replace_all("<e1>", "á") %>%
           str_replace_all("<ed>", "í") %>%
           str_replace_all("<f1>", "ñ") %>%
           str_replace_all("<e9>", "é"),
         
         municipio = municipio %>%
           str_replace_all("<fa>", "ú") %>%
           str_replace_all("<e1>", "á") %>%
           str_replace_all("<ed>", "í") %>%
           str_replace_all("<f1>", "ñ") %>%
           str_replace_all("<e9>", "é")
  ) %>% 
  mutate(municipio_clean = municipio %>% 
           str_to_lower() %>%
           str_replace_all(" ", "_"),
         municipio_clean = case_when(
           municipio == "San Ignacio Río Muerto" ~ "cajeme",
           municipio == "Bácum" ~ "cajeme",
           municipio == "Mexicali" ~ "san_felipe", # because these locations are already filtered to 10km coastal of GoC, so if in Mexicali then it was before the municipality San Felipe was made.
           municipio == "Sinaloa" ~ "guasave",
           municipio == "Santa María del Oro" ~ "san_pedro_lagunillas",
           TRUE ~ municipio_clean
           )) %>% 
  mutate(
    employees_estimated = case_when(
      per_ocu == "0 a 5 personas" ~ 3,
      per_ocu == "6 a 10 personas" ~ 8,
      per_ocu == "11 a 30 personas" ~ 20.5,
      per_ocu == "31 a 50 personas" ~ 40.5,
      per_ocu == "51 a 100 personas" ~ 75.5,
      per_ocu == "101 a 250 personas" ~ 175.5,
      per_ocu %in% c("251 y m<e1>s personas", "251 y más personas") ~ 300,
      TRUE ~ NA_real_
    )
  ) %>% 
  group_by(municipio_clean, data_year, rgn_id) %>% 
  summarize(
    agg_employees = sum(employees_estimated),
    n_units = n(), # number of economic units (establishments)
    .groups = "drop"
  )
```

Now, bring in population data by municipality. 

```{r}
## main cities/towns 2020 population
cities <- st_read(here::here("/home/shares/ohi/OHI_GOC/goal_prep/pressures/population_localities/inegi_boundaries_pop.shp")) %>%
  janitor::clean_names()

mapview(cities) + mapview(non_natural_culture_gulf_10km_clean, zcol = "agg_employees")

## but first, we need to clean the population data.

cities_clean <- st_intersection(st_transform(cities, st_crs(inland_10)), inland_10) %>% 
  mutate(municipio = nom_mun %>%
           str_replace_all("<fa>", "ú") %>%
           str_replace_all("<e1>", "á") %>%
           str_replace_all("<ed>", "í") %>%
           str_replace_all("<f1>", "ñ") %>%
           str_replace_all("<e9>", "é")
  ) %>%
  mutate(municipio_clean = municipio %>%
           str_to_lower() %>%
           str_replace_all(" ", "_"),
         municipio_clean = case_when(
           municipio == "San Ignacio Río Muerto" ~ "cajeme",
           municipio == "Bácum" ~ "cajeme",
           municipio == "Mexicali" ~ "san_felipe", # because these locations are already filtered to 10km coastal of GoC, so if in Mexicali then it was before the municipality San Felipe was made.
           municipio == "Sinaloa" ~ "guasave",
           municipio == "Santa María del Oro" ~ "san_pedro_lagunillas",
           TRUE ~ municipio_clean
         )) %>% 
  dplyr::select(municipio_clean, rgn_id, pobtot, area_km2, pobtot_km2)

```


<!-- ## Bring in population by year -->

```{r}
# pop_goc <- read_csv(here::here("/home/shares/ohi/OHI_GOC/spatial/population_municipal","population_goc_municipality_inegi_2000_2020.csv"))
# 
# # target municipios
# municipios_culturales_lugares <- unique(non_natural_culture_gulf_10km_clean$municipio_clean)
# 
# ## aggregate by region
# agg_pop_rgnid <- pop_goc %>% 
#   mutate(municipio = desc_municipio %>%
#            str_replace_all("<fa>", "ú") %>%
#            str_replace_all("<e1>", "á") %>%
#            str_replace_all("<ed>", "í") %>%
#            str_replace_all("<f1>", "ñ") %>%
#            str_replace_all("<e9>", "é")
#   ) %>% 
#   mutate(municipio_clean = municipio %>% 
#            str_to_lower() %>%
#            str_replace_all(" ", "_"),
#          municipio_clean = case_when(
#            municipio == "San Ignacio Río Muerto" ~ "cajeme",
#            municipio == "Bácum" ~ "cajeme",
#            municipio == "Sinaloa" ~ "guasave",
#            municipio == "Santa María del Oro" ~ "san_pedro_lagunillas",
#            TRUE ~ municipio_clean
#          )) %>% 
#   filter(municipio_clean %in% municipios_culturales_lugares) %>% 
#   group_by(rgn_id, ano_censal, unidad_medida) %>% 
#   summarize(total_rgn_pop = sum(pop_count, na.rm = TRUE)) %>% 
#   ungroup()

# unfortunately, I think we really need to wait until 2025 is released.  We need to use the population locality polygons for this, since we are limiting to 10km coastal. There are many municipalities that have a majority of their population outside of that 10km. 
mapview(cities) + mapview(inland_10)
```

### Join with DENUE

```{r}
non_natural_culture_gulf_10km_clean_join <- non_natural_culture_gulf_10km_clean %>% 
  st_drop_geometry()

cities_clean_join <- cities_clean %>% 
  st_drop_geometry() %>% 
  group_by(municipio_clean) %>% 
  summarize(
    pobtot = sum(pobtot),
    area_km2 = sum(area_km2),
    pobtot_km2 = sum(pobtot_km2)
  ) %>% 
  ungroup()
  

join_culture_pop_10km <- inner_join(non_natural_culture_gulf_10km_clean_join, cities_clean_join)
```


## Non natural cultural areas scaled to population

Let's aggregate employees and economic units to each region and year, then scale both to coastal population.

```{r}
rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

cultural_places_scaled_to_population <- join_culture_pop_10km %>% 
  group_by(rgn_id, data_year) %>% 
  summarize(
    agg_employees = sum(agg_employees),
    n_units = sum(n_units),
    pobtot = sum(pobtot),
    area_km2 = sum(area_km2),
    pobtot_km2 = sum(pobtot_km2),
    .groups = "drop"
  ) %>% 
  mutate(
    culture_emp_per_local_pop = agg_employees / pobtot,
    culture_units_per_local_pop = n_units / pobtot
  )
```


```{r}
library(plotly)

## plot the culture_emp_per_local_pop over time
culture_gulf_emp_per_pop_plotly <- plotly::plot_ly(
  cultural_places_scaled_to_population, 
  x = ~data_year, 
  y = ~culture_emp_per_local_pop, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Non natural cultural places: employees per coastal population",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "culture_emp_per_local_pop")
  ) 
culture_gulf_emp_per_pop_plotly
```

```{r}
## plot the culture_units_per_local_pop over time
culture_gulf_units_per_pop_plotly <- plotly::plot_ly(
  cultural_places_scaled_to_population, 
  x = ~data_year, 
  y = ~culture_units_per_local_pop, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Non natural cultural places: economic units per coastal population",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "culture_units_per_local_pop")
  ) 
culture_gulf_units_per_pop_plotly
```

# Current status alternatives

I will compare several ways of turning the per capita rates into a 0 to 100 style current status.

For the quantile versions (90th, 80th, and 75th), every region and year is compared to the same cross regional, cross year reference:

`current status = value (current) / value (quantile of all regions and years) * 100`

For the 2019 reference versions, each region is compared to its own 2019 baseline so we can see whether it is increasing or decreasing relative to itself:

`current status = value (year) / value (2019, same region) * 100`

Here 100 means matching the 2019 baseline. Values above 100 mean an increase since 2019. I leave those uncapped in `current_status` so the direction of change is visible, and also store a capped `score` (max 100) for OHI style bookkeeping.

## 1. Employees per capita, 90th quantile reference

```{r}
## find the 90th quantile of employees per capita
cultural_emp_90th <- cultural_places_scaled_to_population %>% 
  mutate(
    percap_90th = quantile(culture_emp_per_local_pop, probs = 0.9),
    current_status = (culture_emp_per_local_pop / percap_90th) * 100
  ) %>% 
  mutate(score = ifelse(current_status > 100, 100, current_status))

write_csv(
  cultural_emp_90th,
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026/output",
    "cultural_places_scaled_to_population_current_status_quantile_041026.csv"
  )
)
```


```{r}
culture_emp_90th_plotly <- plotly::plot_ly(
  cultural_emp_90th, 
  x = ~data_year, 
  y = ~score, 
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: employees per capita, scaled to the 90th quantile",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cultural_emp_90th$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
culture_emp_90th_plotly
```

## 2. Employees per capita, 80th quantile reference

```{r}
## find the 80th quantile of employees per capita
cultural_emp_80th <- cultural_places_scaled_to_population %>% 
  mutate(
    percap_80th = quantile(culture_emp_per_local_pop, probs = 0.8),
    current_status = (culture_emp_per_local_pop / percap_80th) * 100
  ) %>% 
  mutate(score = ifelse(current_status > 100, 100, current_status))

write_csv(
  cultural_emp_80th,
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026/output",
    "cultural_places_employees_per_pop_current_status_80th_quantile.csv"
  )
)
```

```{r}
culture_emp_80th_plotly <- plotly::plot_ly(
  cultural_emp_80th, 
  x = ~data_year, 
  y = ~score, 
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: employees per capita, scaled to the 80th quantile",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cultural_emp_80th$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
culture_emp_80th_plotly
```

## 3. Employees per capita, 75th quantile reference

```{r}
## find the 75th quantile of employees per capita
cultural_emp_75th <- cultural_places_scaled_to_population %>% 
  mutate(
    percap_75th = quantile(culture_emp_per_local_pop, probs = 0.75),
    current_status = (culture_emp_per_local_pop / percap_75th) * 100
  ) %>% 
  mutate(score = ifelse(current_status > 100, 100, current_status))

write_csv(
  cultural_emp_75th,
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026/output",
    "cultural_places_employees_per_pop_current_status_75th_quantile.csv"
  )
)

# temporary 2025 gapfill: denue 2025 is not available yet, so copy each region's 2024 score into 2025.
# this is saved as a separate csv so we can reverse it once 2025 denue arrives
# (switch the sp merge back to the non gapfilled csv above).
cultural_emp_75th_gapfill2025 <- cultural_emp_75th %>%
  filter(data_year == 2024) %>%
  mutate(
    data_year = 2025L,
    gapfill_note = "copied from 2024; temporary until 2025 DENUE is available"
  ) %>%
  bind_rows(cultural_emp_75th) %>%
  arrange(rgn_id, data_year)

write_csv(
  cultural_emp_75th_gapfill2025,
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026/output",
    "cultural_places_employees_per_pop_current_status_75th_quantile_gapfill2025.csv"
  )
)
```

```{r}
culture_emp_75th_plotly <- plotly::plot_ly(
  cultural_emp_75th, 
  x = ~data_year, 
  y = ~score, 
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: employees per capita, scaled to the 75th quantile",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cultural_emp_75th$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
culture_emp_75th_plotly
```

## 4. Employees per capita, 2019 regional reference

```{r}
cultural_emp_2019ref <- cultural_places_scaled_to_population %>% 
  group_by(rgn_id) %>% 
  mutate(
    emp_percap_2019 = culture_emp_per_local_pop[data_year == 2019][1],
    current_status = (culture_emp_per_local_pop / emp_percap_2019) * 100
  ) %>% 
  ungroup() %>% 
  mutate(score = ifelse(current_status > 100, 100, current_status))

write_csv(
  cultural_emp_2019ref,
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026/output",
    "cultural_places_employees_per_pop_current_status_2019ref.csv"
  )
)
```

```{r}
culture_emp_2019ref_plotly <- plotly::plot_ly(
  cultural_emp_2019ref, 
  x = ~data_year, 
  y = ~current_status, 
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: employees per capita relative to each region's 2019 baseline",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (100 = 2019 baseline)")
  ) %>%
  add_lines(
    x = range(cultural_emp_2019ref$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "2019 baseline (100)",
    inherit = FALSE
  )
culture_emp_2019ref_plotly
```

## 5. Economic units per capita, 90th quantile reference

```{r}
cultural_units_90th <- cultural_places_scaled_to_population %>% 
  mutate(
    units_percap_90th = quantile(culture_units_per_local_pop, probs = 0.9),
    current_status = (culture_units_per_local_pop / units_percap_90th) * 100
  ) %>% 
  mutate(score = ifelse(current_status > 100, 100, current_status))

write_csv(
  cultural_units_90th,
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026/output",
    "cultural_places_units_per_pop_current_status_quantile.csv"
  )
)
```

```{r}
culture_units_90th_plotly <- plotly::plot_ly(
  cultural_units_90th, 
  x = ~data_year, 
  y = ~score, 
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: economic units per capita, scaled to the 90th quantile",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cultural_units_90th$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
culture_units_90th_plotly
```

## 6. Economic units per capita, 80th quantile reference

```{r}
cultural_units_80th <- cultural_places_scaled_to_population %>% 
  mutate(
    units_percap_80th = quantile(culture_units_per_local_pop, probs = 0.8),
    current_status = (culture_units_per_local_pop / units_percap_80th) * 100
  ) %>% 
  mutate(score = ifelse(current_status > 100, 100, current_status))

write_csv(
  cultural_units_80th,
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026/output",
    "cultural_places_units_per_pop_current_status_80th_quantile.csv"
  )
)
```

```{r}
culture_units_80th_plotly <- plotly::plot_ly(
  cultural_units_80th, 
  x = ~data_year, 
  y = ~score, 
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: economic units per capita, scaled to the 80th quantile",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cultural_units_80th$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
culture_units_80th_plotly
```

## 7. Economic units per capita, 75th quantile reference

```{r}
cultural_units_75th <- cultural_places_scaled_to_population %>% 
  mutate(
    units_percap_75th = quantile(culture_units_per_local_pop, probs = 0.75),
    current_status = (culture_units_per_local_pop / units_percap_75th) * 100
  ) %>% 
  mutate(score = ifelse(current_status > 100, 100, current_status))

write_csv(
  cultural_units_75th,
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026/output",
    "cultural_places_units_per_pop_current_status_75th_quantile.csv"
  )
)
```

```{r}
culture_units_75th_plotly <- plotly::plot_ly(
  cultural_units_75th, 
  x = ~data_year, 
  y = ~score, 
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: economic units per capita, scaled to the 75th quantile",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cultural_units_75th$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
culture_units_75th_plotly
```

## 8. Economic units per capita, 2019 regional reference

```{r}
cultural_units_2019ref <- cultural_places_scaled_to_population %>% 
  group_by(rgn_id) %>% 
  mutate(
    units_percap_2019 = culture_units_per_local_pop[data_year == 2019][1],
    current_status = (culture_units_per_local_pop / units_percap_2019) * 100
  ) %>% 
  ungroup() %>% 
  mutate(score = ifelse(current_status > 100, 100, current_status))

write_csv(
  cultural_units_2019ref,
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026/output",
    "cultural_places_units_per_pop_current_status_2019ref.csv"
  )
)
```

```{r}
culture_units_2019ref_plotly <- plotly::plot_ly(
  cultural_units_2019ref, 
  x = ~data_year, 
  y = ~current_status, 
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: economic units per capita relative to each region's 2019 baseline",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (100 = 2019 baseline)")
  ) %>%
  add_lines(
    x = range(cultural_units_2019ref$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "2019 baseline (100)",
    inherit = FALSE
  )
culture_units_2019ref_plotly
```




