---
title: 'OHI GoC 2025: Sense of Place CCE'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# Sense of Place 

## Coastal Community Engagement: Lasting non-natural cultural places


See `TR/denue_prep.qmd` for the download of `/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_esparcimiento_culturales_y_deportivos_servicios_recreativos/int/non_natural_cultural_places_2019_2024_goc_10km.shp`.


Read in the non-natural cultural places.

```{r}
library(tidyverse)
library(sf)
library(mapview)
library(here)
```


### Culture service establishments 


```{r}
non_natural_culture_gulf_10km <- st_read(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_esparcimiento_culturales_y_deportivos_servicios_recreativos/int/non_natural_cultural_places_2019_2024_goc_10km.shp"))

non_natural_culture_gulf_10km_clean <- non_natural_culture_gulf_10km %>% 
  #st_drop_geometry() %>% 
  mutate(nombre_act = nombre_act %>% 
           str_replace_all("<fa>", "ú") %>%
           str_replace_all("<e1>", "á") %>%
           str_replace_all("<ed>", "í") %>%
           str_replace_all("<f1>", "ñ") %>%
           str_replace_all("<e9>", "é"),
         
         municipio = municipio %>%
           str_replace_all("<fa>", "ú") %>%
           str_replace_all("<e1>", "á") %>%
           str_replace_all("<ed>", "í") %>%
           str_replace_all("<f1>", "ñ") %>%
           str_replace_all("<e9>", "é")
  ) %>% 
  mutate(municipio_clean = municipio %>% 
           str_to_lower() %>%
           str_replace_all(" ", "_"),
         municipio_clean = case_when(
           municipio == "San Ignacio Río Muerto" ~ "cajeme",
           municipio == "Bácum" ~ "cajeme",
           municipio == "Mexicali" ~ "san_felipe", # because these locations are already filtered to 10km coastal of GoC, so if in Mexicali then it was before the municipality San Felipe was made.
           municipio == "Sinaloa" ~ "guasave",
           municipio == "Santa María del Oro" ~ "san_pedro_lagunillas",
           TRUE ~ municipio_clean
           )) %>% 
  mutate(
    employees_estimated = case_when(
      per_ocu == "0 a 5 personas" ~ 3,
      per_ocu == "6 a 10 personas" ~ 8,
      per_ocu == "11 a 30 personas" ~ 20.5,
      per_ocu == "31 a 50 personas" ~ 40.5,
      per_ocu == "51 a 100 personas" ~ 75.5,
      per_ocu == "101 a 250 personas" ~ 175.5,
      per_ocu %in% c("251 y m<e1>s personas", "251 y más personas") ~ 300,
      TRUE ~ NA_real_
    )
  ) %>% 
  group_by(municipio_clean, data_year, rgn_id) %>% 
  summarize(agg_employees = sum(employees_estimated)) %>% 
  ungroup()
```

Now, bring in population data by municipality. 

```{r}
## main cities/towns 2020 population
cities <- st_read(here::here("/home/shares/ohi/OHI_GOC/goal_prep/pressures/population_localities/inegi_boundaries_pop.shp")) %>%
  janitor::clean_names()

mapview(cities) + mapview(non_natural_culture_gulf_10km_clean, zcol = "agg_employees")

## but first, we need to clean the population data.

cities_clean <- st_intersection(st_transform(cities, st_crs(inland_10)), inland_10) %>% 
  mutate(municipio = nom_mun %>%
           str_replace_all("<fa>", "ú") %>%
           str_replace_all("<e1>", "á") %>%
           str_replace_all("<ed>", "í") %>%
           str_replace_all("<f1>", "ñ") %>%
           str_replace_all("<e9>", "é")
  ) %>%
  mutate(municipio_clean = municipio %>%
           str_to_lower() %>%
           str_replace_all(" ", "_"),
         municipio_clean = case_when(
           municipio == "San Ignacio Río Muerto" ~ "cajeme",
           municipio == "Bácum" ~ "cajeme",
           municipio == "Mexicali" ~ "san_felipe", # because these locations are already filtered to 10km coastal of GoC, so if in Mexicali then it was before the municipality San Felipe was made.
           municipio == "Sinaloa" ~ "guasave",
           municipio == "Santa María del Oro" ~ "san_pedro_lagunillas",
           TRUE ~ municipio_clean
         )) %>% 
  dplyr::select(municipio_clean, rgn_id, pobtot, area_km2, pobtot_km2)

```


<!-- ## Bring in population by year -->

```{r}
# pop_goc <- read_csv(here::here("/home/shares/ohi/OHI_GOC/spatial/population_municipal","population_goc_municipality_inegi_2000_2020.csv"))
# 
# # target municipios
# municipios_culturales_lugares <- unique(non_natural_culture_gulf_10km_clean$municipio_clean)
# 
# ## aggregate by region
# agg_pop_rgnid <- pop_goc %>% 
#   mutate(municipio = desc_municipio %>%
#            str_replace_all("<fa>", "ú") %>%
#            str_replace_all("<e1>", "á") %>%
#            str_replace_all("<ed>", "í") %>%
#            str_replace_all("<f1>", "ñ") %>%
#            str_replace_all("<e9>", "é")
#   ) %>% 
#   mutate(municipio_clean = municipio %>% 
#            str_to_lower() %>%
#            str_replace_all(" ", "_"),
#          municipio_clean = case_when(
#            municipio == "San Ignacio Río Muerto" ~ "cajeme",
#            municipio == "Bácum" ~ "cajeme",
#            municipio == "Sinaloa" ~ "guasave",
#            municipio == "Santa María del Oro" ~ "san_pedro_lagunillas",
#            TRUE ~ municipio_clean
#          )) %>% 
#   filter(municipio_clean %in% municipios_culturales_lugares) %>% 
#   group_by(rgn_id, ano_censal, unidad_medida) %>% 
#   summarize(total_rgn_pop = sum(pop_count, na.rm = TRUE)) %>% 
#   ungroup()

# unfortunately, I think we really need to wait until 2025 is released.  We need to use the population locality polygons for this, since we are limiting to 10km coastal. There are many municipalities that have a majority of their population outside of that 10km. 
mapview(cities) + mapview(inland_10)
```

### Join with DENUE

```{r}
non_natural_culture_gulf_10km_clean_join <- non_natural_culture_gulf_10km_clean %>% 
  st_drop_geometry()

cities_clean_join <- cities_clean %>% 
  st_drop_geometry() %>% 
  group_by(municipio_clean) %>% 
  summarize(
    pobtot = sum(pobtot),
    area_km2 = sum(area_km2),
    pobtot_km2 = sum(pobtot_km2)
  ) %>% 
  ungroup()
  

join_culture_pop_10km <- inner_join(non_natural_culture_gulf_10km_clean_join, cities_clean_join)
```


## Non-natural cultural areas scaled to population

```{r}
cultural_places_scaled_to_population <- join_culture_pop_10km %>% 
  group_by(rgn_id, data_year) %>% 
  summarize(
    agg_employees = sum(agg_employees),
    pobtot = sum(pobtot),
    area_km2 = sum(area_km2),
    pobtot_km2 = sum(pobtot_km2)
  ) %>% 
  ungroup() %>% 
  mutate(culture_emp_per_local_pop = agg_employees/pobtot)
```


```{r}
## plot the culture_emp_per_local_pop over time
culture_gulf_emp_per_pop_plotly <- plotly::plot_ly(
  cultural_places_scaled_to_population, 
  x = ~data_year, 
  y = ~culture_emp_per_local_pop, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Non-natural lasting special place employee per capita population",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "culture_emp_per_local_pop")
  ) 
culture_gulf_emp_per_pop_plotly

```

Now let's actually calculate current status for each region and year. The reference point is the 90th quantile?

current status = culture_emp_per_local_pop (current) / culture_emp_per_local_pop (90th quantile of all regions and years)

Where every region and year is compared to the highest.

```{r}
## find the 90th quantile
cultural_places_scaled_to_population_cs <- cultural_places_scaled_to_population %>% 
  mutate(percap_90th = quantile(culture_emp_per_local_pop, probs = 0.9),
         current_status = (culture_emp_per_local_pop / percap_90th) * 100) %>% 
  mutate(score = ifelse(current_status > 100, 100, current_status))

write_csv(cultural_places_scaled_to_population_cs, here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026/output","cultural_places_scaled_to_population_current_status_quantile_041026.csv"))
```


```{r}
library(plotly)

## plot the score over time
culture_quantile_score_plotly <- plotly::plot_ly(
  cultural_places_scaled_to_population_cs, 
  x = ~data_year, 
  y = ~score, 
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: Non-natural cultural places -- employees in establishments per capita population, scaled to the 90th quantile",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cultural_places_scaled_to_population_cs$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
culture_quantile_score_plotly

```










