---
title: 'OHI GoC 2024: Lasting Special Places (LSP)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor:
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options:
  chunk_output_type: console
---

# Sentido de Pertenencia: Lasting Special Places

## Script 5: fishing sites

> **v5_pmax_streamlining note:** this copy replaces the hard-coded Humedales Delta rule with general Ramsar `pmax` vs SIMEC for overlapping parks. Outputs write to `v2026/output_v5_pmax/` so the original `v3_most_updated_scripts` results stay untouched.

In this script, I will score fishing grounds the EWG flagged as iconic. I will check whether each site sits in a fishing refuge (ZRP), a cooperative TURF, or a federal protected area. Those institutional layers tell me the place has some form of stewardship beyond open access.

**What I do**

1. Filter the EWG list to fishing area rows and assign each site an OHI region.
2. Spatial join TNC ZRP polygons (with a 1 km buffer).
3. Flag the one site I confirmed sits inside a TURF using the Aceves-Bueno Shiny map.
4. Join protected areas for SIMEC scores, merge the best protection, average by region, write CSVs.

```{r}
library(tidyverse)
library(here)
library(sf)
library(mapview)
library(stringdist)
library(stringi)
```

## Why these sites matter

Fishing can be subsistence, recreation, or tradition. For lasting special places I care about whether the ground itself is managed so it can remain usable into the future. The EWG named each site.

## Read fishing sites and assign regions

```{r}
lsp_list_ewg <- read_csv(
  here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2026/google_sheets", "ewg_lsp_inventory_prepared.csv"),
  show_col_types = FALSE
) %>%
  janitor::clean_names()

ewg_fishing_site_list <- lsp_list_ewg %>%
  filter(
    type_of_lsp %in% c(
      '"Important fishing areas, could have fishing refuges (zonas de refugio pesquero or ZRP)"'
    )
  ) %>%
  mutate(
    long = as.numeric(long),
    lat = as.numeric(lat)
  ) %>%
  filter(!is.na(long), !is.na(lat)) %>%
  st_as_sf(coords = c("long", "lat"), crs = 4326, remove = FALSE)

mapview(ewg_fishing_site_list)

ohi_rgns <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp"), quiet = TRUE)

ewg_fishing_site_list_goc_rgnid <- st_intersection(
  st_transform(ewg_fishing_site_list, st_crs(ohi_rgns)),
  ohi_rgns
)

mapview(ewg_fishing_site_list_goc_rgnid, zcol = "rgn_id")
```

## ZRPs (fishing refuges)

Now I will pull TNC's geoportal GeoJSON and buffer sites by 1 km so points near a refuge boundary still count.

```{r}
tnc_zrps_raw <- st_read(
  here(
    "/home/shares/ohi/OHI_GOC/_raw_data/TNC_geoportal_fishing_refuge_zones/d2026",
    "Zonas_Refugios_Pesqueros_-8805549763229037385.geojson"
  ),
  quiet = TRUE
) %>%
  janitor::clean_names()

mapview(tnc_zrps_raw, col.regions = "blue") +
  mapview(ewg_fishing_site_list, col.regions = "black")

inland_50 <- st_read(
  here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer", "inland_50km_goc.shp"),
  quiet = TRUE
)

ewg_fishing_site_buffered <- st_buffer(
  st_transform(ewg_fishing_site_list_goc_rgnid, st_crs(inland_50)),
  dist = 1000
)

ewg_fishing_site_condition_zrp <- st_join(
  st_transform(ewg_fishing_site_buffered, st_crs(tnc_zrps_raw)),
  st_make_valid(tnc_zrps_raw),
  join = st_intersects,
  left = TRUE
) %>%
  mutate(in_zrp = !is.na(name)) %>%
  rename(zrp_name = name) %>%
  mutate(zrp_condition = ifelse(in_zrp, 100, 0)) %>%
  dplyr::select(rgn_id, lasting_special_place_lsp, in_zrp, zrp_condition) %>%
  distinct()
```

## TURFs (cooperative tenure)

National TURF polygons are not easy to download from the Aceves Bueno et al. 2023 Shiny map. For now I curate which EWG fishing sites sit in TURF or cooperative tenure and set those sites to a condition of 100. That list started with the Yaqui and Mayo camps at the Sonora Sinaloa border. I am now also including the three Region 8 sites that MBL added: Mazatlán, Topolobampo, and Altata y Bahía de Santa María. If we later get downloadable concession polygons, I can replace this name list with a spatial join.

Papers and map:  
<https://journals.plos.org/plosone/article/figure?id=10.1371/journal.pone.0286739.g003>  
<https://ereaceves.shinyapps.io/Tapp>

```{r}
# these are curated turf or cooperative tenure matches because we do not yet have a national concession polygon file to join spatially. the yaqui and mayo site was confirmed on the aceves bueno et al. 2023 shiny map. mazatlan, topolobampo, and altata are mbl 2026 expert additions for region 8. they describe large cooperative and ribereño fleets, so i treat them as turf equivalent stewardship until we can join concession polygons spatially.
turf_lsp_names <- c(
  "Yaqui and Mayo People have historic (and current) important fishing camps at Sonora-Sinala border. RCB",
  "Mazatlán - puerto pesquero industrial más grande del estado. Aunque destaca por su flota de altamar (atún y camarón), sus zonas costeras e industriales procesan la mayor parte del volumen ribereño del sur. Concentra más del 70% de la flota de altamar del estado. -MBL",
  "Topolobampo - Importante en la captura de camarón de bahía, jaiba, callo de hacha y una enorme variedad de escama marina. Albergan una de las concentraciones de cooperativas y pangas más organizadas de México, sumando miles de pescadores ribereños -MBL",
  "Altata y Bahía de Santa María - destacan casi exclusivamente por la pesca ribereña y costera en bahías interiores. Son zonas clave para la recolección de moluscos (ostión, almeja), jaiba y camarón ribereño. Al no tener la infraestructura de muelles profundos para barcos grandes, estas bahías operan al 100% con embarcaciones menores (pangas). -MBL"
)

ewg_fishing_site_condition_turf <- ewg_fishing_site_condition_zrp %>%
  mutate(
    in_turf = lasting_special_place_lsp %in% turf_lsp_names,
    overall_condition = ifelse(in_turf, 100, zrp_condition)
  )

table(ewg_fishing_site_condition_turf$in_turf)
```

## Protected area (SIMEC) scores

If ZRP and TURF both score 0, a site can still sit inside an ANP and inherit SIMEC. Now I will join the v2 protected area layer and fuzzy match SIMEC names the same way as in scripts 3 and 4.

```{r}
all_protected <- st_read(
  here(
    "/home/shares/ohi/OHI_GOC/spatial/protected_areas_wdpa_conanp_and_hem",
    "all_goc_protected_areas_v2_2025.shp"
  ),
  quiet = TRUE
)

ewg_fishing_site_condition_ANP <- st_join(
  st_transform(ewg_fishing_site_condition_turf, st_crs(all_protected)),
  all_protected %>% dplyr::select(name, geometry),
  join = st_intersects,
  left = TRUE
) %>%
  mutate(in_protected_area = !is.na(name)) %>%
  rename(protected_area_name = name)

mapview(ewg_fishing_site_condition_ANP, zcol = "in_protected_area") +
  mapview(all_protected, col.regions = "pink")
```

## Match overlapping parks to SIMEC and Ramsar

Alrighty, now that I know which parks each fishing site touches, I will prepare SIMEC for **2020 and 2023** the same way as in script 0 (averaging 2020 island-by-state rows). Then I will fuzzy-match park names and also join Ramsar management plan presence. For each overlapping park I will take `pmax(SIMEC, Ramsar)`, then keep the highest park score per fishing site. That replaces the older hard-coded Humedales Delta rule with the same general logic used in habitats.

```{r}
# status years i want lsp scores for
status_years <- c(2020L, 2023L)

# prepare simec for 2020 and 2023 (same logic as script 0)
# older 2020 reports split islas del golfo by state/region; i average those rows
i_efectividad_by_year <- read_csv(
  here(
    "/home/shares/ohi/OHI_GOC/_raw_data/SIMEC",
    "simec_i_efectividad_anp_2026.csv"
  ),
  show_col_types = FALSE,
  na = c("", "NA", "N/A")
) %>%
  janitor::clean_names() %>%
  mutate(
    nombre = str_squish(as.character(nombre)),
    ano_reporte = suppressWarnings(as.integer(ano_reporte)),
    i_efectividad = suppressWarnings(as.numeric(i_efectividad)),
    nombre = if_else(
      str_to_upper(nombre) == "ISLAS DEL GOLFO DE CALIFORNIA",
      "ISLAS DEL GOLFO DE CALIFORNIA",
      nombre
    )
  ) %>%
  filter(
    !is.na(ano_reporte),
    !is.na(i_efectividad),
    !is.na(nombre),
    nombre != "",
    ano_reporte %in% status_years
  ) %>%
  group_by(ano_reporte, nombre) %>%
  summarize(
    i_efectividad = mean(i_efectividad, na.rm = TRUE),
    .groups = "drop"
  )

# year-specific simec min and max (for ramsar rules / nom floors)
simec_bounds_by_year <- i_efectividad_by_year %>%
  group_by(ano_reporte) %>%
  summarize(
    simec_max = max(i_efectividad, na.rm = TRUE),
    simec_min = min(i_efectividad, na.rm = TRUE),
    .groups = "drop"
  )

normalize_name <- function(x) {
  x %>%
    stringi::stri_trans_general("Latin-ASCII") %>%
    str_to_upper() %>%
    str_remove_all(
      "\\bPARQUE NACIONAL\\b|\\bPARQUE\\b|\\bNACIONAL\\b|\\bZONA MARINA\\b|\\bARCHIPIELAGO\\b"
    ) %>%
    str_squish()
}

# fuzzy match on unique park names
i_efectividad_norm <- i_efectividad_by_year %>%
  distinct(nombre) %>%
  mutate(nombre_norm = normalize_name(nombre))

fishing_sites_norm <- ewg_fishing_site_condition_ANP %>%
  mutate(protected_area_norm = normalize_name(protected_area_name))

matches <- stringdist::amatch(
  fishing_sites_norm$protected_area_norm,
  i_efectividad_norm$nombre_norm,
  method = "jw",
  maxDist = 0.2
)

fishing_sites_base <- fishing_sites_norm %>%
  mutate(
    nombre_matched = i_efectividad_norm$nombre[matches],
    nombre_norm_matched = i_efectividad_norm$nombre_norm[matches]
  ) %>%
  st_drop_geometry()

# ramsar management plan presence (same sheet as scripts 6 and 7)
ramsar_management_plans <- read_csv(
  here(
    "/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/google_sheets",
    "ramsar_site_management_plan_presence.csv"
  ),
  show_col_types = FALSE
)

# expand to both status years, then join that year's simec and ramsar scores.
# for each overlapping park, take pmax(simec, ramsar), then max across parks per site.
fishing_sites_w_SIMEC <- bind_rows(
  fishing_sites_base %>% mutate(ano_reporte = 2020L),
  fishing_sites_base %>% mutate(ano_reporte = 2023L)
) %>%
  left_join(
    i_efectividad_by_year,
    by = c("nombre_matched" = "nombre", "ano_reporte")
  ) %>%
  left_join(simec_bounds_by_year, by = "ano_reporte") %>%
  left_join(
    ramsar_management_plans,
    by = c("protected_area_name" = "wdpa_orig_name")
  ) %>%
  mutate(
    year = ano_reporte,
    simec_score = ifelse(!is.na(i_efectividad), as.numeric(i_efectividad), NA_real_),
    ramsar_score = case_when(
      if_ramsar_has_management_plan == "y" ~ simec_max,
      if_ramsar_has_management_plan == "in_progress" ~ 60,
      if_ramsar_has_management_plan == "n" ~ simec_min,
      TRUE ~ NA_real_
    ),
    protected_area_condition = case_when(
      !is.na(simec_score) | !is.na(ramsar_score) ~ pmax(
        tidyr::replace_na(simec_score, -Inf),
        tidyr::replace_na(ramsar_score, -Inf)
      ),
      TRUE ~ NA_real_
    ),
    # cabo pulmo is no take; leave it for careful manual handling
    protected_area_condition = ifelse(
      nombre_matched == "CABO PULMO",
      0,
      protected_area_condition
    )
  ) %>%
  group_by(year, lasting_special_place_lsp) %>%
  summarize(
    protected_area_condition = max(protected_area_condition, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    protected_area_condition = ifelse(
      is.infinite(protected_area_condition),
      NA_real_,
      protected_area_condition
    )
  )
```

## Final site condition and region average

Now that I have ZRP, TURF, and park scores (`pmax` of SIMEC and Ramsar when both apply), I will start from ZRP or TURF (same in both years). When that is still 0, I will fill in that year's best park score. Today all sites have some of these layers, so I do not need a separate codified floor (though NOM links for clams and squid are useful if empty sites appear later). After that I will average site conditions within each OHI region for each status year and write the tables to Mazu.

```{r}
# zrp / turf scores do not change by year, so i expand them to 2020 and 2023
fishing_base_by_year <- bind_rows(
  ewg_fishing_site_condition_turf %>% st_drop_geometry() %>% mutate(year = 2020L),
  ewg_fishing_site_condition_turf %>% st_drop_geometry() %>% mutate(year = 2023L)
)

final_ewg_fishing_site_condition <- fishing_base_by_year %>%
  left_join(
    fishing_sites_w_SIMEC,
    by = c("lasting_special_place_lsp", "year")
  ) %>%
  mutate(
    overall_condition = ifelse(
      overall_condition == 0,
      protected_area_condition,
      overall_condition
    )
  )

rgn_final_ewg_fishing_site_condition <- final_ewg_fishing_site_condition %>%
  group_by(year, rgn_id) %>%
  summarize(condition = mean(overall_condition, na.rm = TRUE), .groups = "drop")

write_csv(
  final_ewg_fishing_site_condition,
  here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2026/output_v5_pmax", "condition_lsp_fishing_sites.csv")
)
write_csv(
  rgn_final_ewg_fishing_site_condition,
  here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2026/output_v5_pmax", "condition_region_fishing_sites.csv")
)
```
