---
title: "Untitled"
output: html_document
date: "2025-07-29"
---

```{r}
library(sf)
library(terra)
library(tidyverse)
library(dplyr)
library(stringr)
library(here)
library(mapview)
library(lwgeom)
sf::sf_use_s2(FALSE)
```

# get OHI ocean regions.
```{r}

# Load in POEM regions
poem <- st_read("/home/shares/ohi/OHI_GOC/spatial/subregions_drafts/poem_w_ohi_draft2/poem_ohi_draft2_071025_final.shp") %>%
  st_make_valid() %>%
  st_transform(6372)

#st_write(poem, here("spatial/poem/POEM.shp"))

ohi_rgn_data <- data.frame(poem) %>%
  select(rgn_id, estado, ecorgn) %>%
  unique()

poem <- st_read(here("spatial/poem/POEM.shp"))
# merge into OHI regions:
poem_buffered <- poem %>%
  st_buffer(0.2)        

# Step 3: Group and dissolve
ohi <- poem_buffered %>%
  group_by(rgn_id) %>%
  summarize(geometry = st_union(geometry), .groups = "drop") 

# Remove buffer
ohi <- st_buffer(ohi, -0.2)

mapview(ohi)
# st_write(ohi, here("spatial/marine_ohi/ohirgns_nolines.geojson"))

ohi <- st_cast(ohi, "MULTIPOLYGON") %>%
  left_join(ohi_rgn_data, by="rgn_id") %>%
  select(rgn_id, estado, ecorgn, geometry)


```


## Add in the corresponding land
```{r}
ent02 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/02ent.shp")
ent03 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/03ent.shp")
ent26 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/26ent.shp")
ent25 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/25ent.shp")
ent18 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/18ent.shp")
ent14 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/14ent.shp")
states <- bind_rows(ent02, ent03, ent26, ent25, ent18, ent14)
states <- st_transform(states, crs=6372)
mapview(states) + mapview(ohi)


```

Some of the water regions overlap with land. Subtract the water regions from the land.
```{r}
ocean <- ohi

states_minus_ocean <- st_difference(states, st_union(ocean))
mapview(states_minus_ocean) + mapview(ohi)

st_area(states)-
st_area(states_minus_ocean) # area lost, which is good!
```

Some gaps between the ocean and land. I want this to be ocean. Therefore, I will buffer the ocean. Then overlay the ocean and land and subtract out ocean that overlaps with land.

```{r}
## function to buffer and clip
cut_rgn <- function(rgn, distance=200, ocean, states_minus_ocean){
#rgn = rgns[1]  
  # Subset for current region
  ocean_poly <- filter(ocean, rgn_id == rgn)
  ocean_buffer <- st_buffer(ocean_poly, dist=distance)
  ocean_minus_rgn <- filter(ocean, rgn_id != rgn)
      land_ocean <-  
      st_union(c(st_geometry(states_minus_ocean),
               st_geometry(ocean_minus_rgn)))
     
    ocean_new <- st_difference(ocean_buffer, land_ocean)
     ocean_new
}

new_ocean <- ocean

new_1 <- cut_rgn(rgn="Region_1", dist=200, ocean=new_ocean, states_minus_ocean = states_minus_ocean)
new_ocean <- rbind(filter(new_ocean, rgn_id != "Region_1"),new_1)

new_5 <- cut_rgn(rgn="Region_5", dist=200, ocean=new_ocean, states_minus_ocean = states_minus_ocean)
new_ocean <- rbind(filter(new_ocean, rgn_id != "Region_5"), new_5)

new_2 <- cut_rgn(rgn="Region_2", dist=300, ocean=new_ocean, states_minus_ocean = states_minus_ocean)
new_ocean <- rbind(filter(new_ocean, rgn_id != "Region_2"),new_2)

new_6 <- cut_rgn(rgn="Region_6", dist=300, ocean=new_ocean, states_minus_ocean = states_minus_ocean)
new_ocean <- rbind(filter(new_ocean, rgn_id != "Region_6"), new_6)

new_3 <- cut_rgn(rgn="Region_3", dist=300, ocean=new_ocean, states_minus_ocean = states_minus_ocean)
new_ocean <- rbind(filter(new_ocean, rgn_id != "Region_3"),new_3)

new_7 <- cut_rgn(rgn="Region_7", dist=300, ocean=new_ocean, states_minus_ocean = states_minus_ocean)
new_ocean <- rbind(filter(new_ocean, rgn_id != "Region_7"), new_7)

new_4 <- cut_rgn(rgn="Region_4", dist=300, ocean=new_ocean, states_minus_ocean = states_minus_ocean)
new_ocean <- rbind(filter(new_ocean, rgn_id != "Region_4"),new_4)

new_8 <- cut_rgn(rgn="Region_8", dist=300, ocean=new_ocean, states_minus_ocean = states_minus_ocean)
new_ocean <- rbind(filter(new_ocean, rgn_id != "Region_8"),new_8)

new_9 <- cut_rgn(rgn="Region_9", dist=200, ocean=new_ocean, states_minus_ocean = states_minus_ocean)
new_ocean <- rbind(filter(new_ocean, rgn_id != "Region_9"), new_9)

mapview(new_ocean) + mapview(states_minus_ocean)


## Fix gap between 1 and 5 and below polygons
rgn <- "Region_5"
distance <- 0.5
  ocean_poly <- filter(new_ocean, rgn_id == rgn)
  ocean_buffer <- st_buffer(ocean_poly, dist=distance)
  ocean_minus_rgn <- filter(new_ocean, !(rgn_id %in% c("Region_5", "Region_1")))
      land_ocean <-  
      st_union(c(st_geometry(states_minus_ocean),
               st_geometry(ocean_minus_rgn)))
     
    tmp_ocean_new <- st_difference(ocean_buffer, land_ocean)
mapview(tmp_ocean_new)

new_ocean <- rbind(filter(new_ocean, rgn_id != rgn), tmp_ocean_new)

#get rid of overlap
new_1v2 <- cut_rgn(rgn="Region_1", dist=0, ocean=new_ocean, states_minus_ocean = states_minus_ocean)
new_ocean <- rbind(filter(new_ocean, rgn_id != "Region_1"), new_1v2)

mapview(new_ocean) + mapview(states_minus_ocean)


## Fix gap between 4 and 7 
rgn <- "Region_4"
distance <- 2
  ocean_poly <- filter(new_ocean, rgn_id == rgn)
  ocean_buffer <- st_buffer(ocean_poly, dist=distance)
  ocean_minus_rgn <- filter(new_ocean, !(rgn_id %in% c("Region_4", "Region_3")))
  
  land_ocean <-  
      st_union(c(st_geometry(states_minus_ocean),
            st_geometry(ocean_minus_rgn)))
     
    tmp_ocean_new <- st_difference(ocean_buffer, land_ocean)
mapview(tmp_ocean_new)

new_ocean <- rbind(filter(new_ocean, rgn_id != rgn), tmp_ocean_new)

#get rid of overlap
new_1v2 <- cut_rgn(rgn="Region_3", dist=0, ocean=new_ocean, states_minus_ocean = states_minus_ocean)
new_ocean <- rbind(filter(new_ocean, rgn_id != "Region_3"), new_1v2)

new_4v2 <- cut_rgn(rgn="Region_4", dist=0, ocean=new_ocean, states_minus_ocean = states_minus_ocean)
new_ocean <- rbind(filter(new_ocean, rgn_id != "Region_4"), new_4v2)


mapview(new_ocean) + mapview(states_minus_ocean)


# Fix a weird gap that should be land. 
box_latlon <- st_as_sfc(st_bbox(c(
  xmin = -114.90406,
  ymin = 31.33796,
  xmax = -114.82854,
  ymax = 31.43288
), crs = 4326))
box_proj <- st_transform(box_latlon, crs = 6372)
plot(st_geometry(box_proj), col="red")  

plot(st_geometry(states_minus_ocean), add=TRUE)

land_poly <- filter(states_minus_ocean, CVE_ENT == "02")
land_minus_rgn <- filter(states_minus_ocean, CVE_ENT != "02")
      land_box <-  
      st_union(c(st_geometry(box_proj),
               st_geometry(land_poly)))
    land_new2 <- st_difference(land_box, st_union(new_ocean))
    st_geometry(land_poly) <- land_new2
    land_newfull <- rbind(land_poly, land_minus_rgn)

mapview(new_ocean) + mapview(land_newfull)

## few more fixes
# fix overlap between region 3 and 4
fix <- filter(new_ocean, rgn_id=="Region_3")
rgn3polygons <- st_cast(fix, "POLYGON")   
mapview(rgn3polygons)
rgn3polygon <- rgn3polygons[1,]
mapview(rgn3polygon)

new_ocean <- rbind(filter(new_ocean, rgn_id != "Region_3"), rgn3polygon)
mapview(new_ocean) + mapview(land_newfull)

new_4v2 <- cut_rgn(rgn="Region_4", dist=50, ocean=new_ocean, states_minus_ocean = states_minus_ocean)
new_ocean <- rbind(filter(new_ocean, rgn_id != "Region_4"), new_4v2)

mapview(new_ocean) + mapview(land_newfull)


# fix a weird jag in region 5
latlong_coords <- matrix(c(-113.52029, 31.29291, -113.51925, 31.29309), ncol = 2, byrow = TRUE)
latlong_line <- st_sfc(st_linestring(latlong_coords))  # Create line geometry
st_crs(latlong_line) <- 4326  # Assign CRS to the line
latlong_sf <- st_sf(geometry = latlong_line)  # Convert to Simple Feature
mapview(new_ocean) + mapview(latlong_sf)

cut02 <- st_transform(latlong_line, st_crs(new_ocean))

rgn02cut <- filter(new_ocean, rgn_id=="Region_5")

cut_polygons <- st_split(rgn02cut, cut02)

 split_polygons <- st_collection_extract(cut_polygons, "POLYGON")   

new_ocean <- filter(new_ocean, rgn_id != "Region_5") 
new_ocean <- rbind(new_ocean, split_polygons[1,])

# fix small gap in region 6
new_6 <- cut_rgn(rgn="Region_6", dist=200, ocean=new_ocean, states_minus_ocean = states_minus_ocean)
new_ocean <- rbind(filter(new_ocean, rgn_id != "Region_6"), new_6)
 mapview(new_ocean) + mapview(land_newfull)

st_write(new_ocean, here("spatial/marine_ohi/marine_ohi.shp"), append=FALSE)
st_write(land_newfull, here("spatial/ohi_regions/states.shp"), append=FALSE)
```

## cut the OHI regions on land
Explore watersheds. We considered using these boundaries, but it didn't really seem to work. So skip this!
```{r}

shed <- st_read("/home/shares/ohi/OHI_GOC/spatial/Cuencas_Hidrograficas_de_Mexico_2007/cue250k_07gw/cue250k_07gw.shp")

```

Hand cut the regions. 

```{r}
ocean <- st_read("spatial/marine_ohi/marine_ohi.shp")
land <- st_read("spatial/ohi_regions/states.shp")
mapview(ocean) + mapview(land) 


# cut Baja into two regions
latlong_coords <- matrix(c(-114.78779, 30.96744, -116.13537, 30.54735), ncol = 2, byrow = TRUE)
latlong_line <- st_sfc(st_linestring(latlong_coords))  # Create line geometry
st_crs(latlong_line) <- 4326  # Assign CRS to the line
latlong_sf <- st_sf(geometry = latlong_line)  # Convert to Simple Feature
mapview(ocean) + mapview(land) + mapview(latlong_sf)

cut02 <- st_transform(latlong_line, st_crs(land))

rgn02cut <- filter(land, CVE_ENT=="02")

cut_polygons <- st_split(rgn02cut, cut02)
 split_polygons <- st_collection_extract(cut_polygons, "POLYGON")   
rgn1 <- split_polygons[c(14, 17,18,19,20,21,22),] %>%
  st_union() %>%
  st_sf(geometry = .) %>%
  mutate(rgn_id = "Region_1",
         location = "land") 

rgn2 <- split_polygons[-c(14, 17, 18, 19, 20, 21, 22), ] %>%
  st_union() %>%
  st_sf(geometry = .) %>%
  mutate(rgn_id = "Region_2",
         location = "land")

# cut Baja Sur into two regions
latlong_coords <- matrix(c(-111.40100,26.34925, -112.47126, 25.95715), ncol = 2, byrow = TRUE)
latlong_line <- st_sfc(st_linestring(latlong_coords))  # Create line geometry
st_crs(latlong_line) <- 4326  # Assign CRS to the line
latlong_sf <- st_sf(geometry = latlong_line)  # Convert to Simple Feature
mapview(ocean) + mapview(land) + mapview(latlong_sf)

cut02 <- st_transform(latlong_line, st_crs(land))

rgn02cut <- filter(land, CVE_ENT=="03")

cut_polygons <- st_split(rgn02cut, cut02)
 split_polygons <- st_collection_extract(cut_polygons, "POLYGON")   
 mapview(split_polygons)
rgn3 <- split_polygons[c(42,43,44,45,46,47,48,
                         50,51,52,53,59,
                         60,82,83,84,85,86,87,88,89),] %>%
  st_union() %>%
  st_sf(geometry = .) %>%
  mutate(rgn_id = "Region_3",
         location = "land") 

rgn4 <- split_polygons[-c(42,43,44,45,46,47,48,
                          50,51,52,53,59,
                         60,82,83,84,85,86,87,88,89), ] %>%
  st_union() %>%
  st_sf(geometry = .) %>%
  mutate(rgn_id = "Region_4",
         location = "land")


# cut Sonora into three regions
latlong_coords <- matrix(c(-113.52018, 31.29295, -111.44855, 31.60423), ncol = 2, byrow = TRUE)
latlong_line <- st_sfc(st_linestring(latlong_coords))  # Create line geometry
st_crs(latlong_line) <- 4326  # Assign CRS to the line
latlong_sf <- st_sf(geometry = latlong_line)  # Convert to Simple Feature
mapview(ocean) + mapview(land) + mapview(latlong_sf)

cut02 <- st_transform(latlong_line, st_crs(land))

rgn02cut <- filter(land, CVE_ENT=="26")

cut_polygons <- st_split(rgn02cut, cut02)
 split_polygons <- st_collection_extract(cut_polygons, "POLYGON")   
 mapview(split_polygons)

 rgn5 <- split_polygons[c(25,26,74),] %>%
  st_union() %>%
  st_sf(geometry = .) %>%
  mutate(rgn_id = "Region_5",
         location = "land") 

rgn6_7 <- split_polygons[-c(25,26,74), ] %>%
  st_union() %>%
  st_sf(geometry = .) %>%
  mutate(rgn_id = "Region_6",
         location = "land")
mapview(rgn6_7)

## 2nd cut
latlong_coords <- matrix(c(-111.45738, 28.31612, -108.46474, 29.56174), ncol = 2, byrow = TRUE)
latlong_line <- st_sfc(st_linestring(latlong_coords))  # Create line geometry
st_crs(latlong_line) <- 4326  # Assign CRS to the line
latlong_sf <- st_sf(geometry = latlong_line)  # Convert to Simple Feature
mapview(ocean) + mapview(land) + mapview(latlong_sf)

cut02 <- st_transform(latlong_line, st_crs(land))

cut_polygons <- st_split(rgn6_7, cut02)
 split_polygons <- st_collection_extract(cut_polygons, "POLYGON")   
 mapview(split_polygons) + mapview(ocean)

 rgn6 <- split_polygons[c(1,19,20,21,22,23,25,27,54),] %>%
  st_union() %>%
  st_sf(geometry = .) %>%
  mutate(rgn_id = "Region_6",
         location = "land") 

rgn7 <- split_polygons[-c(1,19,20,21,22,23,25,27,54), ] %>%
  st_union() %>%
  st_sf(geometry = .) %>%
  mutate(rgn_id = "Region_7",
         location = "land")
mapview(rgn7)


# Pull out Sinaloa

rgn8 <- filter(land, CVE_ENT=="25") %>%
  st_union() %>%
  st_sf(geometry = .) %>%
  mutate(rgn_id = "Region_8",
         location = "land") 

# Pull out Nayarit and Jalisco
rgn9 <- filter(land, CVE_ENT %in% c("18", "14")) %>%
  st_union() %>%
  st_sf(geometry = .) %>%
  mutate(rgn_id = "Region_9",
         location = "land") 

mapview(rgn9)

## combine land into file
land <- rbind(rgn1, rgn2, rgn3, rgn4, rgn5, rgn6, 
              rgn7, rgn8, rgn9)

```

combine land and water
```{r}

land <- left_join(land, st_drop_geometry(ocean), by="rgn_id")

ocean <- ocean %>%
  mutate(location = "marine") 

finalOHI <- rbind(land, ocean)
mapview(finalOHI)
st_write(finalOHI, "spatial/ohi_regions/goc_ohi_rgns.shp")
ohi <- st_read("spatial/ohi_regions/goc_ohi_rgns.shp")

```


## coastal line
never figured out how to work with this.
```{r}
coast <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_coastline/conjunto_de_datos/lineacosta_mex_2023.shp") %>%
  st_transform(6372) %>%
  filter(TIPO == "Continental") 

```

Trimming to a boxed area. Might want to do this.
```{r}
bbox_coords_6372 <- c(
  xmin = 1050000,  # western edge (near Baja California)
  ymin = 900000,   # southern edge (mid-Jalisco)
  xmax = 2300000,  # eastern edge (into Sonora/Sinaloa)
  ymax = 2380000   # northern edge (toward US border)
)

# Convert to bbox and then to polygon
gulf_bbox <- st_bbox(bbox_coords_6372, crs = st_crs(6372)) |> st_as_sfc()

gulf <- st_crop(states, gulf_bbox)
plot(gulf_bbox)
plot(st_geometry(gulf), add=TRUE)

gulf_bbox_vect <- vect(gulf_bbox)  # gulf_bbox must be in EPSG:6372

# Now use that with terra::ext()
r_template <- rast(
  extent = ext(gulf_bbox_vect),
  resolution = 1000,         # 1 km
  crs = crs(gulf_bbox_vect)  # or "EPSG:6372"
)

gulf_rast <- rasterize(states, r_template, field = "CVEGEO", background = 0)

coast <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_coastline/conjunto_de_datos/lineacosta_mex_2023.shp")

```

