---
title: 'OHI GoC 2025: Clean waters Untreated Wastewater, Script 2'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# CW: Chemicals - Untreated Wastewater feeding into the ocean

Data came from the National Census of Municipal Governments and Territorial Demarcations of Mexico City (CNGMD).

> The National Census of Municipal Governments and Territorial Demarcations of Mexico City 2023, has the objective to generate statistical and geographic information on the integration and functioning of city councils and boroughs, as well as the management and performance of the institutions that make up the Public Administration of each municipality or territorial demarcation, specifically in the functions of government, public safety, civil protection, civic justice, potable water and sanitation, and urban solid residues. Its purpose is to link the data with the design, implementation, monitoring and evaluation of public policies of national scope in these functions.

"Puntos de descarga de aguas residuales municipales vertidas sin tratamiento que existían en el municipio o demarcación territorial."

This script will calculate current status, in which the reference point is that there are no untreated wastewater points that directly enter the GoC.

```{r setup}
# load packages for spatial and tabular work
# we need sf/mapview for future mapping, tidyverse for cleaning, and stringi for accent-insensitive municipality name matching across catalogs
library(sf)
library(utils)
library(tidyverse)
library(here)
library(raster)
library(mapview)
library(purrr)
library(stringr)
library(stringi)
library(janitor)
library(tibble)
```

## Read in the cleaned data

```{r}
# all years merged and harmonized (yay!)
goc_aguaresd_all_years <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/agua_sin_tratamiento/int","goc_aguaresd_all_years.csv"))

# all points that discharge into the GoC through the ocean, beach, or estuary.
goc_aguaresd_ocean <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/agua_sin_tratamiento/int","goc_aguaresd_ocean.csv"))
```

## Current status

$$current\ status\ _{wastewater\ discharge} = 1 - \frac{number\ point\ sources\ _{current} - 0\ _{point\ sources}}{historical\ maximum\ _{point\ sources} - 0\ _{point\ sources}}$$

Keep river/stream (1), ground/ravine (4), canal/drain (5), ocean (6), beach (7), estuary/marsh (8) locations:

```{r}
count_goc_aguaresd_ocean_and_river <- goc_aguaresd_all_years %>% 
  filter(res_trec_std %in% c(1L, 4L, 5L, 6L, 7L, 8L)) %>% 
  group_by(rgn_id, census_year) %>% 
  count(census_year, name = "n_goc_discharge_points") %>%
  arrange(census_year) %>% 
  ungroup()
```

### Site density by region:

```{r}
# read municipalities in
municipio_codigo_sf <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/land_admin_boundaries", "mex_admin2_joined_w_municipio_codigos.shp")) %>% 
  mutate(cvempio = as.numeric(clv_mnc),
         cveestado = as.numeric(clv_std))

unique(municipio_codigo_sf$clv_mnc)
unique(municipio_codigo_sf$cveestado)

mapview(municipio_codigo_sf)

islands_polygons <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/from_hem_POEM_ANP/232_ANP-ITRF08_16062025/232_ANP-ITRF08_16062025.shp")) %>% 
  janitor::clean_names() %>% 
  filter(nombre %in% "Islas del Golfo de California")
mapview(islands_polygons)

# erase the island polygons from the municipality polygons, using st_difference, which removes the island area from any municipality that overlaps it
municipio_codigo_sf_no_islands <- municipio_codigo_sf %>%
  st_transform(st_crs(islands_polygons)) %>%
  st_difference(st_union(st_buffer(islands_polygons, dist = 110))) %>% # crop out island areas
  st_make_valid()

# isolate to municipios that ww was measured in: 
municipio_codigo_sf_no_islands_ww <- municipio_codigo_sf_no_islands %>% 
  mutate(mnpio = as.numeric(paste0(clv_std, clv_mnc))) 

# join:
municipio_polygons_ww <- left_join(goc_codes, municipio_codigo_sf_no_islands_ww) %>% 
  st_as_sf()
```


```{r}
#--- 1. calculate region area (km^2) ---
rgn_area_50km <- inland_50 %>%
  #group_by(rgn_id) %>% 
  mutate(area_km2 = as.numeric(st_area(.) / 1e6)) %>%
  group_by(rgn_id) %>% 
  st_drop_geometry() %>%
  summarize(rgn_area_km2 = sum(area_km2)) %>% 
  ungroup() 

rgn_area_municipios <- municipio_polygons_ww %>%
  #group_by(rgn_id) %>% 
  mutate(area_km2 = as.numeric(st_area(.) / 1e6)) %>%
  group_by(rgn_id) %>% 
  st_drop_geometry() %>%
  summarize(rgn_area_km2 = sum(area_km2)) %>% 
  ungroup() %>% 
  mutate(rgn_area_km2 = ifelse(rgn_id %in% "Region_2", 21662.012, rgn_area_km2)) # taken from rgn_area_50km, though there are not any untreated ww in region 2 so it does not matter
```

```{r}
#--- 2. expand to full year sequence per region, at least as much as possible ---
sites_expanded <- count_goc_aguaresd_ocean_and_river %>%
  group_by(rgn_id) %>%
  complete(census_year = c(2013, 2015, 2017, 2019, 2021, 2023)) %>% # note that since this is grouped by region that you might see certain regions only having one year of data, because their min and max are that year.  even after using complete()
  left_join(rgn_area_municipios, by = "rgn_id") %>% 
  replace_na(list(n_goc_discharge_points = 0)) %>%  # because the census was done for that year, and those regions did not report any untreated ww point sources with receptors into the ocean, beach, estuary, etc.
  ungroup() 
  
#--- 3. find number of sites per km2 ---
ww_density_per_rgn <- sites_expanded %>% 
  mutate(
    sites_per_km2 = n_goc_discharge_points / rgn_area_km2
  ) 
```

### Current status

```{r}
ocean_river_current_status <- ww_density_per_rgn %>% 
  mutate(
    ideal = 0,
    maximum = mean(sort(sites_per_km2, decreasing = TRUE)[1:5]),
    pressure_score = ((sites_per_km2 - ideal)/(maximum - ideal)),
    current_status = 1 - ((sites_per_km2 - ideal)/(maximum - ideal)),
    current_status_final = ifelse(current_status < 0, 0, current_status * 100)
  )
```

## Write to Mazu

```{r}
write_csv(ww_density_per_rgn, here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/agua_sin_tratamiento/output","density_goc_discharge_points_per_km2.csv"))

write_csv(ocean_river_current_status, here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/agua_sin_tratamiento/output","untreated_ww_current_status.csv"))
```


## Visualizations

### plotly

```{r}
library(plotly)

rgn_colors <- c(
  Region_1 = "#C03729",
  Region_2 = "#E68C7C",
  Region_3 = "#FC8F24",
  Region_4 = "#dcaa38",
  Region_5 = "#cbd2a0",
  Region_6 = "#98a25a",
  Region_7 = "#747428",
  Region_8 = "#627391",
  Region_9 = "#8d61a9"
)
```


```{r}
goc_ww_ocean_river_plotly <- plotly::plot_ly(
  ocean_river_current_status, 
  x = ~census_year, 
  y = ~current_status_final, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Current Status: CW Untreated wastewater point source density, ocean + beaches + estuaries + rivers + ravines + canals (top 5 maximum)",
    xaxis = list(title = "ano // year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines( 
    x = range(ocean_river_current_status$census_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
goc_ww_ocean_river_plotly
```










