---
title: 'OHI GoC 2025: Clean waters Untreated Wastewater, Script 2'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# CW: Chemicals - Untreated Wastewater feeding into the ocean

Data came from the National Census of Municipal Governments and Territorial Demarcations of Mexico City (CNGMD).

> The National Census of Municipal Governments and Territorial Demarcations of Mexico City 2023, has the objective to generate statistical and geographic information on the integration and functioning of city councils and boroughs, as well as the management and performance of the institutions that make up the Public Administration of each municipality or territorial demarcation, specifically in the functions of government, public safety, civil protection, civic justice, potable water and sanitation, and urban solid residues. Its purpose is to link the data with the design, implementation, monitoring and evaluation of public policies of national scope in these functions.

"Puntos de descarga de aguas residuales municipales vertidas sin tratamiento que existían en el municipio o demarcación territorial."

This script will calculate current status, in which the reference point is that there are no untreated wastewater points that directly enter the GoC.

```{r setup}
# load packages for spatial and tabular work
# we need sf/mapview for future mapping, tidyverse for cleaning, and stringi for accent-insensitive municipality name matching across catalogs
library(sf)
library(utils)
library(tidyverse)
library(here)
library(raster)
library(mapview)
library(purrr)
library(stringr)
library(stringi)
library(janitor)
library(tibble)
```

## Read in the cleaned data

```{r}
# all years merged and harmonized (yay!)
goc_aguaresd_all_years <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/agua_sin_tratamiento/int","goc_aguaresd_all_years.csv"))

# all points that discharge into the GoC specifically.
goc_aguaresd_ocean <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/agua_sin_tratamiento/int","goc_aguaresd_ocean.csv"))
```

## Current status

$$current\ status\ _{wastewater\ discharge} = 1 - \frac{number\ point\ sources\ _{current} - 0\ _{point\ sources}}{historical\ maximum\ _{point\ sources} - 0\ _{point\ sources}}$$

All destination locations:
```{r}
count_goc_aguaresd_all_years <- goc_aguaresd_all_years %>% 
  group_by(rgn_id, census_year) %>% 
  count(census_year, name = "n_goc_discharge_points") %>%
  arrange(census_year) %>% 
  ungroup()

any_destination_current_status <- count_goc_aguaresd_all_years %>% 
  mutate(
    ideal = 0,
    maximum = max(n_goc_discharge_points),
    current_status = 1 - ((n_goc_discharge_points - ideal)/(maximum - ideal)),
    current_status_final = current_status * 100
  )

# --- region-specific reference point ---
any_destination_current_status_per_rgn <- count_goc_aguaresd_all_years %>% 
  group_by(rgn_id) %>% 
  mutate(
    ideal = 0,
    maximum = max(n_goc_discharge_points),
    current_status = 1 - ((n_goc_discharge_points - ideal)/(maximum - ideal)),
    current_status_final = current_status * 100
  ) %>% 
  ungroup()
```

Ocean, beach, estuary, *and* river/stream locations:
```{r}
count_goc_aguaresd_ocean_and_river <- goc_aguaresd_all_years %>% 
  filter(res_trec_std %in% c(1L, 6L, 7L, 8L)) %>% 
  group_by(rgn_id, census_year) %>% 
  count(census_year, name = "n_ocean_river_discharge_points") %>%
  arrange(census_year) %>% 
  ungroup()

ocean_river_current_status <- count_goc_aguaresd_ocean_and_river %>% 
  mutate(
    ideal = 0,
    maximum = max(n_ocean_river_discharge_points),
    current_status = 1 - ((n_ocean_river_discharge_points - ideal)/(maximum - ideal)),
    current_status_final = current_status * 100
  )

# --- region-specific reference point ---
ocean_river_current_status_per_rgn <- count_goc_aguaresd_ocean_and_river %>% 
  group_by(rgn_id) %>% 
  mutate(
    ideal = 0,
    maximum = max(n_ocean_river_discharge_points),
    current_status = 1 - ((n_ocean_river_discharge_points - ideal)/(maximum - ideal)),
    current_status_final = current_status * 100
  )
```


Only ocean, beach, and estuary locations:
```{r}
count_goc_aguaresd_ocean <- goc_aguaresd_ocean %>% 
  group_by(rgn_id, census_year) %>% 
  count(census_year, name = "n_ocean_discharge_points") %>%
  arrange(census_year) %>% 
  ungroup()

only_ocean_current_status <- count_goc_aguaresd_ocean %>% 
  mutate(
    ideal = 0,
    maximum = max(n_ocean_discharge_points),
    current_status = 1 - ((n_ocean_discharge_points - ideal)/(maximum - ideal)),
    current_status_final = current_status * 100
  )

# --- region-specific reference point ---
only_ocean_current_status_per_rgn <- count_goc_aguaresd_ocean %>% 
  group_by(rgn_id) %>% 
  mutate(
    ideal = 0,
    maximum = max(n_ocean_discharge_points),
    current_status = 1 - ((n_ocean_discharge_points - ideal)/(maximum - ideal)),
    current_status_final = current_status * 100
  )
  
```

## Comparative visualizations

### plotly

```{r}
library(plotly)

rgn_colors <- c(
  Region_1 = "#C03729",
  Region_2 = "#E68C7C",
  Region_3 = "#FC8F24",
  Region_4 = "#dcaa38",
  Region_5 = "#cbd2a0",
  Region_6 = "#98a25a",
  Region_7 = "#747428",
  Region_8 = "#627391",
  Region_9 = "#8d61a9"
)
```


```{r}
## plot the ww sources cs over time
goc_ww_sources_plotly <- plotly::plot_ly(
  any_destination_current_status, 
  x = ~census_year, 
  y = ~current_status_final, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Current Status: CW Untreated wastewater point sources, all destination locations (global historical maximum)",
    xaxis = list(title = "ano // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines( 
    x = range(any_destination_current_status$census_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
goc_ww_sources_plotly


### using the per region maximum in the reference point:
## plot the ww sources cs over time
goc_ww_sources_rgn_plotly <- plotly::plot_ly(
  any_destination_current_status_per_rgn, 
  x = ~census_year, 
  y = ~current_status_final, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Current Status: CW Untreated wastewater point sources, all destination locations (rgn-specific maximum)",
    xaxis = list(title = "ano // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines( 
    x = range(any_destination_current_status_per_rgn$census_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
goc_ww_sources_rgn_plotly
```


```{r}
goc_ww_ocean_river_plotly <- plotly::plot_ly(
  ocean_river_current_status, 
  x = ~census_year, 
  y = ~current_status_final, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Current Status: CW Untreated wastewater point sources, ocean + beaches + estuaries + rivers (global historical maximum)",
    xaxis = list(title = "ano // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines( 
    x = range(ocean_river_current_status$census_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
goc_ww_ocean_river_plotly

## using the per region maximum in the reference point:
## plot the ww sources cs over time
goc_ww_ocean_river_rgn_plotly <- plotly::plot_ly(
  ocean_river_current_status_per_rgn, 
  x = ~census_year, 
  y = ~current_status_final, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Current Status: CW Untreated wastewater point sources, ocean + beaches + estuaries + rivers (rgn-specific maximum)",
    xaxis = list(title = "ano // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines( 
    x = range(ocean_river_current_status_per_rgn$census_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
goc_ww_ocean_river_rgn_plotly
```


```{r}
goc_ww_ocean_only_plotly <- plotly::plot_ly(
  only_ocean_current_status, 
  x = ~census_year, 
  y = ~current_status_final, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Current Status: CW Untreated wastewater point sources, ocean + beaches + estuaries (global historical maximum)",
    xaxis = list(title = "ano // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines( 
    x = range(only_ocean_current_status$census_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
goc_ww_ocean_only_plotly

## using the per region maximum in the reference point:
## plot the ww sources cs over time
goc_ww_ocean_rgn_plotly <- plotly::plot_ly(
  only_ocean_current_status_per_rgn, 
  x = ~census_year, 
  y = ~current_status_final, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Current Status: CW Untreated wastewater point sources, ocean + beaches + estuaries (rgn-specific maximum)",
    xaxis = list(title = "ano // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines( 
    x = range(only_ocean_current_status_per_rgn$census_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
goc_ww_ocean_rgn_plotly
```

```{r}
latest_year <- max(any_destination_current_status$census_year)

comparison_df <- bind_rows(
  any_destination_current_status %>% filter(census_year == latest_year) %>% mutate(scenario = "All destinations"),
  ocean_river_current_status  %>% filter(census_year == latest_year) %>% mutate(scenario = "Ocean + River"),
  only_ocean_current_status %>% filter(census_year == latest_year) %>% mutate(scenario = "Ocean/Beach/Estuary only")
)

goc_ww_comparison_plotly <- plotly::plot_ly(
  comparison_df,
  x = ~rgn_id,
  y = ~current_status_final,
  color = ~scenario,
  type = "bar"
) %>%
  layout(
    title = paste("Clean Waters: wastewater score comparison for", latest_year),
    barmode = "group",
    xaxis = list(title = "Region // Región"),
    yaxis = list(title = "Puntaje // Percent (out of 100)", range = c(0, 100))
  )
goc_ww_comparison_plotly
```









