---
title: "OHI GoC: LSP points, protected areas, SIMEC *i*-efectividad"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
editor:
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options:
  chunk_output_type: console
execute:
  warning: true
  message: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

## Goal

Filter collapsed LSP rows (protected area + SIMEC/RAMSAR), test whether each point **intersects** WDPA or HEM land ANPs (`st_intersects`), attach SIMEC scores via the EWG list, then output **`rgn_id`**, **`lasting_special_place_lsp`**, **`condition`**: SIMEC score if present, `1` = Ramsar-only in EWG, `0` = other PA, `NA` = outside both layers.

```{r libraries}
library(tidyverse)
library(here)
library(janitor)
library(sf)
library(mapview)
mapviewOptions(fgb = FALSE)
```

## Paths (`here`)

```{r paths}
WDPA_SHP <- here::here("/home/lecuona/OHI_Intro/ohi_goc_prep/LSP/v2025/output/WDPA_Jan2025_shp_clipped_all.shp")

COLLAPSED_CSV <- here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/identifying_lasting_special_places_collapsed_lsp_protections.csv")

SIMEC_CSV <- here("/home/shares/ohi/OHI_GOC/_raw_data/SIMEC/identifying_lasting_special_places_SIMEC.csv")

EWG_LIST_CSV <- here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/LSP_list_ewg_mentioned_protected_areas_v1_042426.csv")

HEM_ANP_SHP <- here(
  "/home/shares/ohi/OHI_GOC/spatial/from_hem_POEM_ANP/232_ANP-ITRF08_16062025",
  "232_ANP-ITRF08_16062025.shp"
)
```

## Read tables

```{r read-data}
i_efectividad_raw <- read_csv(SIMEC_CSV, show_col_types = FALSE) %>%
  clean_names()

ewg_protected_areas_list <- read_csv(EWG_LIST_CSV, show_col_types = FALSE) %>%
  clean_names()

if (!"only_ramsar_site" %in% names(ewg_protected_areas_list)) {
  ewg_protected_areas_list$only_ramsar_site <- NA_character_
}

i_efectividad_w_ewg_list <- full_join(
  i_efectividad_raw,
  ewg_protected_areas_list,
  by = c("nombre" = "i_efectividad_name")
)

collapsed_raw <- read_csv(COLLAPSED_CSV, show_col_types = FALSE, na = c("", "NA")) %>%
  clean_names()

if ("ohi_go_c_region" %in% names(collapsed_raw) && !"ohi_goc_region" %in% names(collapsed_raw)) {
  collapsed_raw <- collapsed_raw %>% rename(ohi_goc_region = ohi_go_c_region)
}
```

## LSP points to test (protected area + SIMEC/RAMSAR, valid coords)

```{r lsp-points}
lsp_pa <- collapsed_raw %>%
  filter(
    str_detect(protection_should_be, regex("protected\\s*area", ignore_case = TRUE)),
    str_detect(protection_condition_determined_by, regex("SIMEC|RAMSAR", ignore_case = TRUE))
  ) %>%
  mutate(
    lat = parse_number(as.character(lat)),
    long = parse_number(as.character(long))
  ) %>%
  filter(!is.na(lat), !is.na(long))

lsp_pts <- lsp_pa %>%
  st_as_sf(coords = c("long", "lat"), crs = 4326, remove = FALSE)
```

## WDPA + HEM polygons as one layer (`pa_all`)

Polygon **geometry** decides inclusion; names are only for joining SIMEC / Ramsar flags afterward.

```{r pa-layers}
stopifnot(file.exists(WDPA_SHP), file.exists(HEM_ANP_SHP))

wdpa <- st_read(WDPA_SHP, quiet = TRUE) %>%
  st_transform(4326) %>%
  st_make_valid()

hem_protected_anps <- st_read(HEM_ANP_SHP, quiet = TRUE) %>%
  st_transform(4326) %>%
  st_make_valid() %>%
  clean_names() %>%
  filter(
    estados %in% c(
      "Baja California Sur", "Jalisco", "Baja California", "Sonora", "Nayarit", "Sinaloa",
      "Baja California y Baja California Sur", "B.C., B.C.S., Sonora y Sinaloa", "Jalisco y Nayarit"
    )
  )

# first matching name column in each layer — edit vectors if your fields differ
wdpa_col <- intersect(c("NAME", "name", "NAME_1"), names(wdpa)) %>% .[1]
hem_col <- intersect(c("nombre", "nom_anp", "NOM_N1", "name"), names(hem_protected_anps)) %>% .[1]
if (is.na(wdpa_col)) stop("Set a label column for WDPA (see wdpa names: ", paste(names(wdpa), collapse = ", "), ")")
if (is.na(hem_col)) stop("Set a label column for HEM (see hem names)")

pa_wdpa <- wdpa %>%
  transmute(
    pa_label = as.character(.data[[wdpa_col]]),
    pa_source = "wdpa"
  )
pa_hem <- hem_protected_anps %>%
  transmute(
    pa_label = as.character(.data[[hem_col]]),
    pa_source = "hem_anp"
  )
pa_all <- bind_rows(pa_wdpa, pa_hem)
```

## EWG + SIMEC lookups (and Ramsar-only flag per `wdpa_key`)

Match on a **normalised** name string (lowercase, no accents) so labels from either layer can meet `wdpa_orig_name` in the EWG file.

```{r name-key}
# one place to normalise text for joins
norm_name <- function(x) {
  x %>%
    as.character() %>%
    stringi::stri_trans_general("Latin-ASCII") %>%
    str_squish() %>%
    str_to_lower()
}
```

```{r ewg-simec-ramsar}
simec_scores <- i_efectividad_raw %>%
  transmute(
    simec_key = str_squish(str_to_upper(nombre)),
    i_efectividad = as.numeric(i_efectividad),
    ano_reporte,
    vinculo_reporte
  ) %>%
  filter(!is.na(simec_key), simec_key != "")

ewg_match <- ewg_protected_areas_list %>%
  mutate(
    wdpa_key = norm_name(wdpa_orig_name),
    simec_key = ifelse(
      is.na(i_efectividad_name) | i_efectividad_name %in% c("", "NA", "na"),
      NA_character_,
      str_squish(str_to_upper(i_efectividad_name))
    )
  ) %>%
  filter(!is.na(wdpa_key), wdpa_key != "", wdpa_key != "na") %>%
  group_by(wdpa_key) %>%
  slice(1) %>%
  ungroup() %>%
  dplyr::select(wdpa_key, simec_key)

ramsar_y <- function(x) str_to_lower(str_squish(as.character(x))) %in%
  c("y", "yes", "true", "1", "s", "sí", "si")

ewg_ramsar_keys <- ewg_protected_areas_list %>%
  mutate(
    wdpa_key = norm_name(wdpa_orig_name),
    is_ramsar_only = ramsar_y(only_ramsar_site) &
      (is.na(i_efectividad_name) | i_efectividad_name %in% c("", "NA", "na"))
  ) %>%
  filter(!is.na(wdpa_key), wdpa_key != "", wdpa_key != "na") %>%
  group_by(wdpa_key) %>%
  summarize(is_ramsar_only = any(is_ramsar_only, na.rm = TRUE)) %>%
  ungroup()
```

## Spatial join + SIMEC scores

```{r join-intersect}
joined <- st_join(lsp_pts, pa_all, join = st_intersects, left = TRUE) %>%
  mutate(wdpa_key = norm_name(pa_label))

with_simec <- joined %>%
  left_join(ewg_match, by = "wdpa_key") %>%
  left_join(ewg_ramsar_keys, by = "wdpa_key") %>%
  mutate(
    # if EWG did not match, try polygon label as SIMEC nombre
    simec_key = ifelse(
      !is.na(simec_key),
      simec_key,
      ifelse(is.na(pa_label), NA_character_, str_squish(str_to_upper(pa_label)))
    ),
    is_ramsar_only = replace_na(is_ramsar_only, FALSE)
  ) %>%
  left_join(simec_scores, by = "simec_key")
```

## Summary + `condition` + output columns

```{r summary}
lsp_summary <- with_simec %>%
  #st_drop_geometry() %>%
  group_by(ohi_goc_region, type_of_lsp, lasting_special_place_lsp, lat, long) %>%
  summarize(
    in_pa_poly = any(!is.na(pa_label)),
    pa_names = paste(unique(na.omit(pa_label)), collapse = " | "),
    pa_sources = paste(unique(na.omit(pa_source)), collapse = ", "),
    i_efectividad_best = {
      v <- i_efectividad[!is.na(i_efectividad)]
      if (length(v)) max(v) else NA_real_
    },
    any_ramsar_only = any(is_ramsar_only, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    condition = case_when(
      !in_pa_poly ~ NA_real_,
      !is.na(i_efectividad_best) ~ as.numeric(i_efectividad_best),
      any_ramsar_only ~ 1,
      in_pa_poly ~ 0,
      TRUE ~ NA_real_
    )
  )

mapview(lsp_summary, zcol = "in_pa_poly")

lsp_condition <- lsp_summary %>%
  transmute(
    rgn_id = parse_number(as.character(ohi_goc_region)),
    lasting_special_place_lsp,
    condition
  )
```

## Quick checks

```{r counts}
lsp_summary %>%
  count(in_pa_poly, name = "n_rows")
```

```{r missing-simec-not-ramsar}
lsp_summary %>%
  filter(in_pa_poly, is.na(i_efectividad_best), !any_ramsar_only) %>%
  select(lasting_special_place_lsp, pa_names, pa_sources, condition)
```

## Map

```{r map, fig.width = 8, fig.height = 6}
mapview(wdpa, layer.name = "WDPA", col.regions = "blue", alpha.regions = 0.35) +
  mapview(hem_protected_anps, layer.name = "HEM ANP", col.regions = "green") +mapview(lsp_summary, zcol = "in_pa_poly")
```

## Main result

```{r result}
lsp_condition
```

```{r objects}
result_pts_sf <- with_simec
# write_csv(lsp_condition, here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2026/output", "lsp_condition.csv"))
```

Objects: **`lsp_condition`**, **`lsp_summary`**, **`result_pts_sf`**, **`i_efectividad_w_ewg_list`**.
