---
title: 'OHI GoC 2024: Lasting Special Places (LSP)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# Sentido de Pertenencia: Lasting Special Places

## Fishing sites

*Overview*
Fishing is an activity that can be done for subsistence, but it can also be done recreationally, or as part of traditional/cultural engagement. Ideally, fishing grounds are well managed so the locations to fish can be maintained.

We will determine the condition of whether the fishing site is "protected" to last into the future based on whether there is an active zona de refugio pesquero (ZRP) or fishing cooperative (TURF) in that location.  

Each fishing site was named by the Expert Working Group.


```{r}
library(tidyverse)
library(here)
library(sf)
library(mapview)
```


## Read in fishing sites

```{r}
lsp_list_ewg_v3 <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/identifying_lasting_special_places_collapsed_lsp_protections_v3.csv")) %>% 
  janitor::clean_names()

lsp_list_ewg_v4 <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/identifying_lasting_special_places_collapsed_lsp_protections_v4.csv")) %>% 
  janitor::clean_names()

ewg_fishing_site_list <- lsp_list_ewg_v4 %>% 
  filter(type_of_lsp %in% c('"Important fishing areas, could have fishing refuges (zonas de refugio pesquero or ZRP)"')) %>%
  mutate(long = as.numeric(long),
         lat = as.numeric(lat)) %>% 
  filter(!is.na(long)) %>% 
  st_as_sf(
    coords = c("long", "lat"), # making sf object
    crs = 4326, # WGS84
    remove = FALSE
  )

mapview(ewg_fishing_site_list)
```

### Assign regions

First, read in the marine and land polygons.
```{r}
ohi_rgns <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp")) 

# mapview(ohi_rgns, zcol = "rgn_id")
```

Now, we can intersect with `ohi_rgns` so that each site is assigned an OHI GoC Region ID.
```{r}
# intersect with GoC
ewg_fishing_site_list_goc_rgnid <- st_intersection(st_transform(ewg_fishing_site_list, st_crs(ohi_rgns)), ohi_rgns)

mapview(ewg_fishing_site_list_goc_rgnid, zcol = "rgn_id") 
```


## Determine if priority fishing sites fall within ZRPs or TURFs or codified regulation

### ZRPs

```{r}
tnc_zrps_raw <- st_read(here::here("/home/shares/ohi/OHI_GOC/_raw_data/TNC_geoportal_fishing_refuge_zones/d2026/Zonas_Refugios_Pesqueros_-8805549763229037385.geojson")) %>% 
  janitor::clean_names()

mapview(tnc_zrps_raw, col.regions = "blue") + mapview(ewg_fishing_site_list, col.regions = "black")
```

```{r}
# for mexico CRS
inland_50 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer","inland_50km_goc.shp"))

# buffer fishing sites by 1000m
ewg_fishing_site_buffered <- st_buffer(
  st_transform(ewg_fishing_site_list_goc_rgnid, st_crs(inland_50)),
  dist = 1000
)
#mapview(ewg_fishing_site_buffered)

# spatial join to the non-dissolved layer (all_protected) to keep protected area names
ewg_fishing_site_condition_zrp <- st_join(
  st_transform(ewg_fishing_site_buffered, st_crs(tnc_zrps_raw)),
  st_make_valid(tnc_zrps_raw),
  join = st_intersects,
  left = TRUE  # keep all sites, even those outside of protected areas
) %>%
  mutate(in_zrp = !is.na(name)) %>%
  rename(zrp_name = name) %>% 
  mutate(zrp_condition = ifelse(in_zrp == "TRUE", 100, 0)) %>% 
  dplyr::select(rgn_id, lasting_special_place_lsp, in_zrp, zrp_condition) %>% 
  distinct()
```


### TURFs

This cooperative data is from authors Eréndira Aceves-Bueno ,Mateja Nenadovic,India Dove,Claire Atkins-Davis,Juan Salvador Aceves-Bueno,Abel Trejo-Ramirez,Cristina Rivas-Ochoa,Salvador Rodriguez-Van Dyck,Amy Hudson Weaver, from the paper *Sustaining small-scale fisheries through a nation-wide Territorial Use Rights in Fisheries system* published in 2023.

<https://journals.plos.org/plosone/article/figure?id=10.1371/journal.pone.0286739.g003>
<https://ereaceves.shinyapps.io/Tapp>

Unfortunately these are not easily downloadable, but using the interactive Shiny app, we can determine visually whether the iconic fishing site lies within a TURF. 

As of now, only the `lasting_special_place_lsp` == "Yaqui and Mayo People have historic (and current) important fishing camps at Sonora-Sinala border. RCB" lies within a TURF.  

Specifically, SON	S.C.P.P. PROGRESO DE BACOREGUIS	59.5021330485373	CAMARON DE ESTERO (Penaeus spp.), 	SON	S.C.P.P. PUNTA COLORADA	10.1437898288521	CAMARON DE ESTERO (Penaeus spp.), 	SON	S.C.P.P. BAHIA DE LA ACEITUNA	10.9384359771656	CAMARON DE ESTERO (Penaeus spp.), 	SON	S.C.P.P. SYLVIA RAMIREZ	6.25839729540253	CAMARON DE ESTERO (Penaeus spp.), SON	S.C.P.P. ENSENADA DEL COROBOCHA	18.5936130597429	CAMARON DE ESTERO (Penaeus spp.), SON	S.C.P.P. BAHIA DE SIALIBOCA	18.761539480154	CAMARON DE ESTERO (Penaeus spp.), SIN	S.C.P.P. BAHIA DE JITZAMURI	29.5269278412638	CAMARON DE ESTERO (Penaeus spp.), SON	S.C.P.P. PESCADORES AUTENTICOS DEL JITZAMURI	16.9434518590803	CAMARON DE ESTERO (Penaeus spp.), 	SON	S.C.P.P. INDIGENAS UNIDOS DE BACOREHUIS	13.0599505730419	CAMARON DE ESTERO (Penaeus spp.), 	SIN	S.C.P.P. DREN CINCO	7.5655624022361	CAMARON DE ESTERO (Penaeus spp.), SIN	S.C.P.P. PACIOTECOLA	14.0565957834506	CAMARON DE ESTERO (Penaeus spp.), SIN	S.C.P.P. ESTRELLA DEL NORTE	14.3260790098044	CAMARON DE ESTERO (Penaeus spp.), 	SIN	S.C.P.P. IGNACIO SARMIENTO	12.6098901269236	CAMARON DE ESTERO (Penaeus spp.).

```{r}
ewg_fishing_site_condition_turf <- ewg_fishing_site_condition_zrp %>% 
  mutate(in_turf = ifelse(lasting_special_place_lsp == "Yaqui and Mayo People have historic (and current) important fishing camps at Sonora-Sinala border. RCB", TRUE, FALSE)) %>% 
  mutate(overall_condition = ifelse(in_turf == "TRUE", 100, zrp_condition))
```



## Intersect with protected areas

```{r}
# buffer fishing sites by 1000m
ewg_fishing_site_buffered <- st_buffer(
  st_transform(ewg_fishing_site_list_goc_rgnid, st_crs(inland_50)),
  dist = 1000
)

ewg_fishing_site_condition_ANP <- st_join(
  st_transform(ewg_fishing_site_condition_turf, st_crs(all_protected)),
  all_protected %>% dplyr::select(name, geometry),
  join = st_intersects,
  left = TRUE  # keep all sites, even those outside of protected areas
) %>%
  mutate(in_protected_area = !is.na(name)) %>%
  rename(protected_area_name = name) 

# quick checks for sanity
mapview(ewg_fishing_site_condition_ANP, zcol = "in_protected_area") + mapview(all_protected, col.regions = "pink")
table(ewg_fishing_site_condition_ANP$in_protected_area)
table(ewg_fishing_site_condition_ANP$protected_area_name)
table(ewg_fishing_site_condition_ANP$lasting_special_place_lsp) # multiple observations for each site.
```

The only problem is that if a buffered site overlaps multiple protected areas, `st_join()` will return one row per match, giving us duplicate sites. So that is something to keep in mind.

### i-efectividad for each dive site in a protected area

```{r}
i_efectividad_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SIMEC/identifying_lasting_special_places_SIMEC.csv")) %>% 
  janitor::clean_names()

unique(i_efectividad_raw$nombre)
unique(ewg_fishing_site_condition_ANP$protected_area_name)
```

### Fuzzy matching the protected area names

```{r}
library(stringdist)
library(stringi)

# normalize protected area names...uppercase, remove accents, remove common filler words
normalize_name <- function(x) {
  x %>%
    stringi::stri_trans_general("Latin-ASCII") %>%  # remove accents
    str_to_upper() %>%
    str_remove_all("\\bPARQUE NACIONAL\\b|\\bPARQUE\\b|\\bNACIONAL\\b|\\bZONA MARINA\\b|\\bARCHIPIELAGO\\b") %>%
    str_squish()  # remove extra whitespace left behind
}

# make normalized versions for matching
i_efectividad_norm <- i_efectividad_raw %>%
  mutate(nombre_norm = normalize_name(nombre))

fishing_sites_norm <- ewg_fishing_site_condition_ANP %>%
  mutate(protected_area_norm = normalize_name(protected_area_name))

# fuzzy match: for each dive site protected area name, find the closest match in i_efectividad
matches <- stringdist::amatch(
  fishing_sites_norm$protected_area_norm,
  i_efectividad_norm$nombre_norm,
  method = "jw", # Jaro-Winkler works well for place names
  maxDist = 0.2 # adjust threshold as needed; lower = stricter
)
```


```{r}
# attach matched nombre from i_efectividad back to dive sites
fishing_sites_matched <- fishing_sites_norm %>%
  mutate(
    nombre_matched = i_efectividad_norm$nombre[matches],
    nombre_norm_matched = i_efectividad_norm$nombre_norm[matches]
  )

# inspect the matches before joining
# join them!!!
fishing_sites_w_SIMEC <- fishing_sites_matched %>%
  left_join(
    i_efectividad_raw,
    by = c("nombre_matched" = "nombre")
  ) %>% 
  mutate(protected_area_condition = ifelse(!is.na(i_efectividad), i_efectividad, NA),
         protected_area_condition = ifelse(protected_area_name == "Humedales del Delta del Río Colorado", max(i_efectividad_raw$i_efectividad), protected_area_condition), # this Ramsar site has a management plan in place: https://rsis.ramsar.org/?f%5B0%5D=region_en_s%3ANorth+America&f%5B1%5D=managementPlanAvailable_i%3A-1&q=Mexico#map  https://docs.google.com/spreadsheets/d/1RwNTzRZdBIDtzi7d1ceFxhNmVghN47NxTgrhZV8O2SM/edit?gid=1116165746#gid=1116165746
        protected_area_condition = ifelse(nombre_matched == "CABO PULMO", 0, protected_area_condition) # CABO PULMO IS NO TAKE!!!!
         ) %>%
  group_by(lasting_special_place_lsp) %>% 
  summarize(protected_area_condition = mean(protected_area_condition, na.rm = TRUE)) %>% 
  ungroup() %>% 
  st_drop_geometry()

## note that "LA PIEDRA DE LA VIRGEN, / O PIEDRA DEL CRISTO DEL MAR.  Important for the huichol people, they recognize as the point of origin of their people (https://visitnayarit.travel/en/blog-nayarit-en/tatei-haramara-the-origin-of-the-world-according-to-the-wixarika-worldview/) (JGH)" is fully protected by Hauxa Manaka Pact for the Preservation and Development of Wixarika Culture (2008) and UNESCO (as of 2025).
```



## Condition for each fishing site

```{r}
final_ewg_fishing_site_condition <- ewg_fishing_site_condition_turf %>% 
  left_join(fishing_sites_w_SIMEC) %>% 
  mutate(overall_condition = ifelse(overall_condition == 0, protected_area_condition, overall_condition)) %>% 
  st_drop_geometry()
```

### Codified regulation

None of the sites are not protected in some way by ZRPs, TURFs, or ANPs.  So we will not have to estimate a condition score based on codified protections, though some resources to do so in the future are here:

#### Clams (almejas)

[Official Mexican Standard NOM-014-SAG/PESC-2015, Specifications to regulate the use of geoduck clam ( panopea generosa and panopea globosa ) in waters of federal jurisdiction of the Pacific Ocean coast and Gulf of California.](https://dof.gob.mx/nota_detalle_popup.php?codigo=5398036)

[2022 Anuncian implementación de veda temporal para Almeja Chocolata](https://www.gob.mx/imipas/prensa/anuncian-implementacion-de-veda-temporal-para-almeja-chocolata?idiom=es)

[Catarina Clam](https://www.gob.mx/imipas/es/articulos/almeja-catarina-208871?idiom=es)

[2021 CARACTERIZACIÓN DE LA PESQUERÍA DE LA ALMEJA BLANCA (DOSINIA PONDEROSA) EN LA REGIÓN DE PUERTO PEÑASCO, SONORA, MÉXICO.](https://fisheryprogress.org/sites/default/files/documents_actions/Annex%2004%20Rep%20Final-Almeja_Blanaca%20Nov2021.pdf)

#### Squid

[Low management score from Seafood Watch.](https://www.seachoice.org/wp-content/uploads/2015/06/MBA_SeafoodWatch_JumboSquidMexicoReport.pdf)


## Condition per region

```{r}
rgn_final_ewg_fishing_site_condition <- final_ewg_fishing_site_condition %>% 
  group_by(rgn_id) %>% 
  summarize(
    condition = mean(overall_condition, na.rm = TRUE)
  ) %>% 
  ungroup()
```


## Write to Mazu

```{r}
write_csv(final_ewg_fishing_site_condition, here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/output/fishing_sites_condition.csv"))
write_csv(rgn_final_ewg_fishing_site_condition, here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/output/rgn_fishing_sites_condition.csv"))
```


