---
title: 'OHI GoC 2024: Lasting Special Places (LSP)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# Sentido de Pertenencia: Lasting Special Places

## Dive Sites

*Overview*
Dive sites are considered lasting special places because people make a concerted effort to travel to these locations, some of them rather remote, and view wildlife. It is a direct connection between humans and the ocean, and if these places are not protected, then it is possible there will not continue to be a visited dive spot because their unique features are not there anymore. If possible, we should separate out dive spots that are known for the rock formations versus the wildlife viewing.

[From Arcos-Aguilar et al. 2021](https://datamares.org/wp-content/uploads/2021/12/Arcos-etal-2021-Diving-tourism-in-Mexico-Economic-and-conservation-importance.pdf):

  "The trip price is related to the accessibility of the diving sites. Areas with low accessibility charge more per trip than areas with high accessibility (Fig. 3). The increase in trip price can be related to an increase in costs to reach the diving sites by operators (e.g. higher gasoline consumption per trip); thus, accessibility of the diving sites can represent an advantage to the operators. However, the dive sites close to shore are under more environmental pressures than remote areas and are harder to manage [54]. Whenever management measures are implemented, the support of the community is essential. The success of MPAs is linked to the involvement and support of local communities in MPA management [55].

  "Diving operators can increase revenues by supporting environmental protections that increase the chances of encounters with pristine wildlife. This is one of the top factors influencing travel decisions made by ecotourists, and such encounters have a higher probability of occurring within a No-Take Zone where biodiversity has full protection [47]. Establishing higher prices to access protected areas is not unheard of, especially when it can deliver an improved experience. Indeed, the number of No-Take Zones was found to be positively correlated with the popularity of a site, and in turn popularity drives revenues (Fig. 3). Tourists may be willing to pay more to ensure local conservation and to visit healthy habitats [23,26]."

  "There is high potential revenue and greater benefits to be had by promoting a system of small-scale diving operations that prioritize the quality of tourists’ experiences, which can then translate into higher prices, increasing net revenues and decreasing pressure on biodiversity over time."

  "Preservation of biodiversity embedded in diving sites must take into consideration long-term effects to the environment, the economy, and the livelihood of people and communities. To achieve this, diving operators should actively participate in marine spatial planning decisionmaking processes due to their economic importance and the necessity of community-based management. As of now, Mexico has the opportunity to become a beacon for community-led management through ecotourism activities that stimulate a novel and profitable way of using marine resources sustainably."

*Potential protections:* 
Protected areas/no-take zones, tourism use regulations/permits (CONANP visitor permits), Normas Oficiales Mexicanas (NOMs) for ecotourism (for diving, Norma Oficial Mexicana, NOM-012-TUR-2016).

*Planned methods:*
If it falls into a protected area, the condition will be based on the management effectiveness determined by CONANP's SIMEC index of effectiveness (i-efectividad). If not, the condition will be a 0.60 out of 1, because in [NOM-012-TUR-2016](https://dof.gob.mx/nota_to_doc.php%3Fcodnota%3D5450981), in section 6.11, it says this: "6.11 Para la práctica de buceo, tanto en las áreas naturales protegidas como en aquellas que se encuentran fuera, el prestador de servicios turísticos de buceo debe informar que las actividades de buceo deben realizarse sin dañar o extraer la flora y fauna silvestre acuática (viva o muerta), así como no intervenir, alterar o extraer monumentos y vestigios arqueológicos, históricos y artísticos, entendiéndose por éstos tanto los bienes prehistóricos y paleontológicos, así como los naufragios o pecios. En caso de realizar inmersiones en lugares que contengan patrimonio cultural subacuático, por seguridad del turista o usuario y de dicho patrimonio deberán mantener una distancia mínima de un metro de los elementos que conformen el patrimonio cultural subacuático. Los usuarios deberán verificar que todos los instrumentos que porten estén debidamente sujetos para que no arrastren y no dañen el patrimonio cultural o las formaciones coralinas. En caso de practicarse el buceo en cavidades naturales, lo anterior debe aplicarse también para no ocasionar daño a las formaciones geológicas ahí presentes." Therefore, there are some legal limitations on the actions of divers to protected the marine area, regardless of whether it falls into a protected area or not. 

https://ecojurisprudence.org/initiatives/mexico-law-on-ecological-equilibrium-and-environmental-protection//


```{r}
library(tidyverse)
library(here)
library(sf)
library(mapview)
```


## Read in dive sites


<!-- ```{r} -->
<!-- final_mpa_layer <- st_read(here::here("/home/shares/ohi/OHI_GOC/_raw_data/CBMC-GCMP-prioritization-of-diving-sites-main/shp/final_mpa_layer.shp")) %>% -->
<!--   janitor::clean_names() -->

<!-- # get rid of Z coordinates to make all geometries 2D -->
<!-- final_mpa_layer <- st_zm(final_mpa_layer, drop = TRUE, what = "ZM") -->

<!-- # verify the geometry types are now consistent -->
<!-- table(st_geometry_type(final_mpa_layer)) -->
<!-- ``` -->

### All dive sites

```{r}
dive_sites_raw <- st_read(here::here("/home/shares/ohi/OHI_GOC/_raw_data/CBMC-GCMP-prioritization-of-diving-sites-main/shp/Diving_sites_v1.0_01032021/dive_sites.shp")) %>% 
  janitor::clean_names()
  
mapview(dive_sites_raw)
```

### Dive sites with conservation priority

```{r}
dive_sites_w_priority <- readxl::read_xlsx(here::here("/home/shares/ohi/OHI_GOC/_raw_data/CBMC-GCMP-prioritization-of-diving-sites-main/data/tabla_prioridad.xlsx")) %>% 
  janitor::clean_names() %>% 
  st_as_sf(coords = c("longitud", "latitud"), # making sf object
           crs = 4326, # WGS84
           remove = FALSE)

mapview(dive_sites_w_priority)
```


### Limit to the GoC 

First, group by region and unite the polygons for inland 10km with the marine polygons.
```{r}
## intersect with inland_10:
inland_10 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_10km_buffer","inland_10km_goc.shp"))
regions_marine <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp")) %>%
  filter(location == "marine")

st_crs(inland_10) == st_crs(regions_marine) # TRUE

marine_and_inland10 <- bind_rows(regions_marine, inland_10) %>% 
  dplyr::select(rgn_id, geometry) %>% 
  group_by(rgn_id) %>% 
  summarize(
    geometry = st_union(geometry)
  )

mapview(marine_and_inland10)
```

Now, we can intersect with the `marine_and_inland10` so that each dive site is assigned an OHI GoC Region ID.
```{r}
# intersect with GoC
dive_sites_goc_rgnid <- st_intersection(st_transform(dive_sites_raw, st_crs(marine_and_inland10)), marine_and_inland10)
dive_sites_w_priority_goc_rgnid <- st_intersection(st_transform(dive_sites_w_priority, st_crs(marine_and_inland10)), marine_and_inland10)

mapview(dive_sites_goc_rgnid, zcol = "rgn_id") 
mapview(dive_sites_w_priority_goc_rgnid, zcol = "rgn_id")

# compare them
mapview(dive_sites_goc_rgnid, col.regions = "blue") + mapview(dive_sites_w_priority_goc_rgnid, col.regions = "red")

st_write(dive_sites_goc_rgnid, here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/int","dive_sites_goc_rgnid.gpkg"))
st_write(dive_sites_w_priority_goc_rgnid, here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/int","dive_sites_w_priority_goc_rgnid.gpkg"))
```


<!-- ```{r} -->
<!-- tourism_operators <- st_read(here::here("/home/shares/ohi/OHI_GOC/_raw_data/CBMC-GCMP-prioritization-of-diving-sites-main/shp/Tourism_operators_v1.0_01032021/operators.shp")) %>%  -->
<!--   janitor::clean_names() -->

<!-- mapview(tourism_operators) -->
<!-- ``` -->

## Determine if priority dive sites fall within protected areas

```{r}
mapview(dive_sites_w_priority_goc_rgnid, col.regions = "red") + mapview(wdpa_poly_all, col.regions = "blue") + mapview(hem_protected_anps, col.regions = "black")
```

### Dissolve protected areas into one layer

```{r}
# check CRS
st_crs(wdpa_poly_all) == st_crs(hem_protected_anps) # FALSE

# transform hem to match wdpa (or vice versa)
hem_protected_anps_reproj <- st_transform(hem_protected_anps, st_crs(wdpa_poly_all))

# harmonize geometry types (cast both to MULTIPOLYGON to be safe)
wdpa_clean <- st_cast(wdpa_poly_all, "MULTIPOLYGON")
hem_clean  <- st_cast(hem_protected_anps_reproj, "MULTIPOLYGON")

# select only shared/needed columns before binding (bind_rows will fill NA for mismatched columns — fine, but tidy to be explicit)
all_protected <- bind_rows(
  wdpa_clean %>% dplyr::select(name, geometry),
  hem_clean  %>% dplyr::select(name = nombre, geometry)  # adjust column names to match yours
)
mapview(all_protected)

# dissolve overlaps for a single coverage layer
all_protected_dissolved <- st_union(all_protected)
mapview(all_protected_dissolved)

# write out to mazu
st_write(all_protected, here::here("/home/shares/ohi/OHI_GOC/spatial/protected _areas_wdpa_and_hem", "all_goc_protected_areas_2025.shp"))
st_write(all_protected_dissolved, here::here("/home/shares/ohi/OHI_GOC/spatial/protected _areas_wdpa_and_hem", "all_goc_protected_areas_dissolved_2025.shp"))
```

### Intersect diving priority areas w protected areas

```{r}
# buffer dive sites by 100m
dive_sites_buffered <- st_buffer(
  st_transform(dive_sites_w_priority_goc_rgnid, st_crs(all_protected)),
  dist = 100
)
mapview(dive_sites_buffered) + mapview(dive_sites_w_priority_goc_rgnid)

# spatial join to the non-dissolved layer (all_protected) to keep protected area names
dive_sites_protected <- st_join(
  dive_sites_buffered,
  all_protected %>% dplyr::select(name, geometry),
  join = st_intersects,
  left = TRUE  # keep all dive sites, even those outside of protected areas
) %>%
  mutate(in_protected_area = !is.na(name)) %>%
  rename(protected_area_name = name) 

# quick checks for sanity
mapview(dive_sites_w_priority_goc_rgnid) + mapview(dive_sites_protected, zcol = "in_protected_area") + mapview(all_protected, col.regions = "pink")
table(dive_sites_protected$in_protected_area)
table(dive_sites_protected$protected_area_name)
table(dive_sites_protected$nombre_sitio_de_buceo) # multiple observations for each dive site.
```

The only problem is that if a buffered dive site overlaps multiple protected areas, `st_join()` will return one row per match, giving us duplicate dive sites. So that is something to keep in mind.

## i-efectividad for each dive site in a protected area

```{r}
i_efectividad_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SIMEC/identifying_lasting_special_places_SIMEC.csv")) %>% 
  janitor::clean_names()

unique(i_efectividad_raw$nombre)
unique(dive_sites_protected$protected_area_name)
```

### Fuzzy matching the protected area names

```{r}
library(stringdist)
library(stringi)

# normalize protected area names...uppercase, remove accents, remove common filler words
normalize_name <- function(x) {
  x %>%
    stringi::stri_trans_general("Latin-ASCII") %>%  # remove accents
    str_to_upper() %>%
    str_remove_all("\\bPARQUE NACIONAL\\b|\\bPARQUE\\b|\\bNACIONAL\\b|\\bZONA MARINA\\b|\\bARCHIPIELAGO\\b") %>%
    str_squish()  # remove extra whitespace left behind
}
```


```{r}
# make normalized versions for matching
i_efectividad_norm <- i_efectividad_raw %>%
  mutate(nombre_norm = normalize_name(nombre))

dive_sites_norm <- dive_sites_protected %>%
  mutate(protected_area_norm = normalize_name(protected_area_name))
```


```{r}
# fuzzy match: for each dive site protected area name, find the closest match in i_efectividad
matches <- stringdist::amatch(
  dive_sites_norm$protected_area_norm,
  i_efectividad_norm$nombre_norm,
  method = "jw", # Jaro-Winkler works well for place names
  maxDist = 0.2 # adjust threshold as needed; lower = stricter
)
```


```{r}
# attach matched nombre from i_efectividad back to dive sites
dive_sites_matched <- dive_sites_norm %>%
  mutate(
    nombre_matched = i_efectividad_norm$nombre[matches],
    nombre_norm_matched = i_efectividad_norm$nombre_norm[matches]
  )

# inspect the matches before joining, here or by opening the df ^
dive_sites_matched %>%
  st_drop_geometry() %>%
  distinct(protected_area_name, protected_area_norm, nombre_matched) %>%
  print(n = Inf)
```

```{r}
# join them!!!
dive_sites_w_SIMEC <- dive_sites_matched %>%
  left_join(
    i_efectividad_raw,
    by = c("nombre_matched" = "nombre")
  ) %>% 
  mutate(protected_area_condition = ifelse(!is.na(i_efectividad), i_efectividad, NA)) %>% 
  dplyr::select(rgn_id, nombre_sitio_de_buceo, protected_area_name, protected_area_condition, popularidad_de_los_sitios, riqueza_de_especies, amenaza_de_cambio_climatico, valor_turistico, estado, latitud, longitud)
```

## Condition for each dive site

```{r}
# now, summarize by dive site, so we have one condition for each site
dive_sites_condition <- dive_sites_w_SIMEC %>% 
  mutate(nom_condition = min(protected_area_condition, na.rm = TRUE)) %>%  # because in NOM-012-TUR-2016, in section 6.11, it says this: "6.11 Para la práctica de buceo, tanto en las áreas naturales protegidas como en aquellas que se encuentran fuera, el prestador de servicios turísticos de buceo debe informar que las actividades de buceo deben realizarse sin dañar o extraer la flora y fauna silvestre acuática (viva o muerta), así como no intervenir, alterar o extraer monumentos y vestigios arqueológicos, históricos y artísticos, entendiéndose por éstos tanto los bienes prehistóricos y paleontológicos, así como los naufragios o pecios. En caso de realizar inmersiones en lugares que contengan patrimonio cultural subacuático, por seguridad del turista o usuario y de dicho patrimonio deberán mantener una distancia mínima de un metro de los elementos que conformen el patrimonio cultural subacuático. Los usuarios deberán verificar que todos los instrumentos que porten estén debidamente sujetos para que no arrastren y no dañen el patrimonio cultural o las formaciones coralinas. En caso de practicarse el buceo en cavidades naturales, lo anterior debe aplicarse también para no ocasionar daño a las formaciones geológicas ahí presentes." Therefore, there are some legal limitations on the actions of divers to protected the marine area, regardless of whether it falls into a protected area or not.  Legal restrictions are not perfect, though: <https://insideclimatenews.org/news/02082025/mexico-marine-tourism-damaging-wildlife/>
  group_by(nombre_sitio_de_buceo, rgn_id, latitud) %>% 
  mutate(nom_condition = mean(nom_condition), # not really needed because it is all the same value, but still
         protected_area_condition = mean(protected_area_condition)) %>% 
  dplyr::select(-protected_area_name) %>% 
  distinct() %>% 
  ungroup() %>% 
  mutate(overall_condition = ifelse(!is.na(protected_area_condition), protected_area_condition, nom_condition)) %>% 
  st_drop_geometry() %>% 
  dplyr::select(rgn_id, nombre_sitio_de_buceo, overall_condition, protected_area_condition, popularidad_de_los_sitios, riqueza_de_especies, amenaza_de_cambio_climatico, valor_turistico, estado, latitud, longitud)

dive_sites_condition_sf <- dive_sites_condition %>% 
  st_as_sf(coords = c("longitud", "latitud"), # making points again, not buffered area
           crs = 4326, # WGS84
           remove = FALSE) 

mapview(dive_sites_condition_sf, zcol = "protected_area_condition") + mapview(all_protected)
mapview(dive_sites_condition_sf, zcol = "overall_condition") + mapview(all_protected)
```

# Condition per region

```{r}
rgn_dive_sites_condition <- dive_sites_condition %>% 
  group_by(rgn_id) %>% 
  summarize(
    condition = mean(overall_condition, na.rm = TRUE)
  ) %>% 
  ungroup()
```


## Write to Mazu

```{r}
write_csv(dive_sites_condition, here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/output/dive_sites_condition.csv"))
write_csv(rgn_dive_sites_condition, here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/output/rgn_dive_sites_condition.csv"))

dive_sites_condition <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/output/dive_sites_condition.csv"))
```








