################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
# Visualization
library(ggplot2)
library(colorspace)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```

```{r Load clean_names_lowercase function}
clean_names_lowercase <- function(x) {

  x <- iconv(x, from = "", to = "UTF-8", sub = "")
  x <- stringi::stri_trans_general(x, "Latin-ASCII")

  x %>%
    tolower() %>%
    trimws()
}
```


# STEP 1: Load in and summarize GOC_MarineLandings0

```{r Load in GOC_MarineLandings0 as necessary}
GOC_MarineLandings0<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Marine.Landings_Sector.Sorted.Georef_PATCHED_Jun032026.rds"))
# 3246583 obs., 58 var.

# NOTE TO SELF: This source file may change in later Summer / Fall run. - EMF, June 4 2026
# See the patch file here: https://docs.google.com/document/d/1L6w4Gd8SOjp14f_u46vyn64rEpv4qBRaN_V2IyPjP5w/edit?tab=t.0
```


# PATCH: Look into groups that are unassigned to a Major_Functional_Group:
NOTE TO SELF: This patch should not be necessary when we re-run this in later Summer / Fall run. - EMF, June 15 2026
See the patch file here: https://docs.google.com/document/d/1L6w4Gd8SOjp14f_u46vyn64rEpv4qBRaN_V2IyPjP5w/edit?tab=t.0

```{r Review generathat are not assigned to a Major_Functional_Group}
TEST<- GOC_MarineFisheryLandings0 %>%
  filter(estimated_genus_name== "Sardinops")

TEST2<- GOC_MarineFisheryLandings0 %>%
  filter(is.na(Major_Functional_Group))

table(TEST2$estimated_genus_name)
```

```{r Load in GenusFamilyGroup_Key as needed}
GenusFamilyGroup_Key<- read.csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Keys/Taxonomic_Keys/GenusFamilyGroupKey_COMPLETE_Apr042026.csv")

# NOTE TO SELF: This patch should not be necessary when we re-run this in later Summer / Fall run. - EMF, June 15 2026
# See the patch file here: https://docs.google.com/document/d/1L6w4Gd8SOjp14f_u46vyn64rEpv4qBRaN_V2IyPjP5w/edit?tab=t.0

MajorFunctionalGroup_Key<- GenusFamilyGroup_Key %>%
  select(estimated_genus_name, Major_Functional_Group)
```

```{r Major_Functional_Group patch for GOC_MarineLandings}
# Fill in missing values based on MajorFunctionalGroup_Key:
GOC_MarineLandings <- GOC_MarineLandings0 %>%
  left_join(
    MajorFunctionalGroup_Key %>%
      select(estimated_genus_name, Major_Functional_Group),
    by = "estimated_genus_name",
    suffix = c("", ".lookup")
  )

GOC_MarineLandings <- GOC_MarineLandings %>%
  mutate(
    Major_Functional_Group = coalesce(
      Major_Functional_Group,
      Major_Functional_Group.lookup
    )
  ) %>%
  select(-Major_Functional_Group.lookup)

# 3246583 obs., 58 var.

# SANITY CHECK:
sum(is.na(GOC_MarineLandings$Major_Functional_Group))
# Only 26 records unassigned. Looks good!
```


# **DECISION POINT: Selecting which sectors to include**
```{r Select Sectors_of_Interest}
table(GOC_MarineLandings$sector_estimate_draft)

# SANITY CHECK: Which species are included in the freshwater or marine sector?
TEST<- GOC_MarineLandings %>%
  filter(sector_estimate_draft== "freshwater or marine: fishery sector yet to be classified")

table(TEST$recorded_taxa_group)
# Species like Mojarra (the overwhelming majority of catch), trout, and catfish, which can move from freshwater to marine systems and are a bit ambiguous in terms of where they can be caught.


MarineFishery_Sectors<- c("marine: fishery sector yet to be classified",
                          "marine: small-scale capture",
                          "marine: large-scale capture")
  
PossibleMarineFishery_Sectors<- c("freshwater or marine: large-scale capture",
                          "freshwater or marine: small-scale capture",
                          "freshwater or marine: fishery sector yet to be classified")

GOC_MarineFisheryLandings0<- GOC_MarineLandings %>%
  filter(sector_estimate_draft %in% c(MarineFishery_Sectors, PossibleMarineFishery_Sectors))

# 3137044 obs., 58 var.
```

```{r Re-level sector_estimate_draft and Major_Functional_Group}
# Re-level sectors (mainly for plotting purposes):
GOC_MarineFisheryLandings <- GOC_MarineFisheryLandings0 %>%
  filter(
    sector_estimate_draft %in% c(
      MarineFishery_Sectors,
      PossibleMarineFishery_Sectors
    )
  ) %>%
  mutate(
    sector_estimate_draft = factor(
      sector_estimate_draft,
      levels = c(
        "marine: large-scale capture",
        "marine: small-scale capture",
        "marine: fishery sector yet to be classified",
        "freshwater or marine: large-scale capture",
        "freshwater or marine: small-scale capture",
        "freshwater or marine: fishery sector yet to be classified"
      )
    )
  )

# SANITY CHECK: How many Major_Functional_Group unassigned?
sum(is.na(GOC_MarineFisheryLandings$Major_Functional_Group))
# Only 18 records unassigned. Looks good!

table(GOC_MarineFisheryLandings$Major_Functional_Group)

GOC_MarineFisheryLandings <- GOC_MarineFisheryLandings %>%
  mutate(
    Major_Functional_Group = factor(
      Major_Functional_Group,
      levels= c(
        
        # Crustaceans
        "Shrimps, prawns",
        "Crustaceans",
        
        # Molluscs
        "Cephalopods",
        "Molluscs excl. Cephalopods",
        
        # Other invertebrates
        "Echinoderms",
        "Jellies",
        "Other marine invertebrates",
        
        # Fishes
        "Anchovy & Sardine (small pelagics)",
        "Pelagic marine fish (excl. anchovy & sardine)",
        "Demersal marine fishes",
        "Marine fishes NEI",
        "Marine or freshwater fishes NEI",
        "Diadromous fishes",
        
        # Elasmobranchs
        "Sharks",
        
        # Miscellaneous / unresolved
        "Marine algae",
        "Marine species NEI",
        "unknown",
        NA_character_
      )
    )
  )

# SANITY CHECK:
sum(is.na(GOC_MarineFisheryLandings$Major_Functional_Group))
# Only 18 records unassigned. Looks good!
```


# STEP 2: Review & Plot Sector Data ------------------------------

```{r Generate GOC_MarineSector_summary}
colnames(GOC_MarineFisheryLandings)

GOC_MarineSector_summary<- GOC_MarineFisheryLandings %>%
  select(year, 
         state, office_name_clean, landing_site_clean,
         sector_estimate_draft,
         GOC_est, GOC_Include_Exclude, OHI_region_est, OHI_region_name,
         eu_rnpa, eu_name, sector_estimate_draft, 
         Major_Functional_Group, recorded_taxa_group, estimated_taxa_group,
         recorded_taxa_name, recorded_taxa_name_clean,
         estimated_species_name, estimated_genus_name,
         landed_weight, live_weight, weight_kg_corrected, value
         )

GOC_MarineSector_summary<- GOC_MarineSector_summary %>%
  group_by(year, OHI_region_est, sector_estimate_draft) %>%
  summarize(
    total_weight_mt= sum(weight_kg_corrected, na.rm = TRUE) / 1000,
    total_value_peso= sum(value, na.rm = TRUE),
    n_records= n()
  )
# 646 obs.

# Regions 7 and 8 produce A LOT of catch relative to the other regions!
# Work (in the next two chunks) reveal that most of this catch is "Anchovy & Sardine (small pelagics)"
```

## Plotting sector estimates:
```{r Plot each sector of interest over time}
# Define colors for each sector (fleet distinction = lighter for small-scale)
sector_colors <- c(
  # Marine (blues)
  "marine: large-scale capture"                       = "#336699",  # darker blue
  "marine: small-scale capture"                       = "#33cccc",  # teal/aquaculture cue
  "marine: fishery sector yet to be classified"       = "#E78AC3"  # pink
  
  # Freshwater or Marine (grey and purple shades) - IGNORE FOR NOW.
  # "freshwater or marine: fishery sector yet to be classified"  = "#e0e0e0",  # light grey
  # "freshwater or marine: small-scale capture"                    = "#b399cc",  # light purple
  # "freshwater or marine: large-scale capture"                    = "#8040a0",  # medium purple
)

# Plot
Landings_bySector_plot <- GOC_MarineSector_summary %>%
  filter(sector_estimate_draft %in% MarineFishery_Sectors,
         year >= 2018,
         year <= 2024) %>%
  ggplot(
    aes(
      x = factor(year),
      y = total_weight_mt,
      fill = sector_estimate_draft
    )
  ) +
  geom_col(
    position = position_dodge2(width = 0.8, preserve = "single"),
    width = 0.7,
    color = "white",
    linewidth = 0.2
  ) +
  facet_wrap(~ OHI_region_est, ncol = 3) +
  scale_y_continuous(
    trans = "log10",
    labels = scales::comma
  ) +
  scale_fill_manual(
    name = "Fishery sector",
    values= sector_colors
  ) +
  labs(
    x = NULL,
    y = "Annual catch (metric tons, log-scale)",
    title = "Catch by Sector in the Gulf of California",
    subtitle = "Small-scale versus large-scale classification reported directly to CONAPESCA"
  ) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", size = 16),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  guides(
  fill = guide_legend(nrow = 2, byrow = TRUE)
)

Landings_bySector_plot
```

```{r Save Landings_bySector_plot}
ggsave(
  "/home/ferrer/ohi_goc_prep/AO/v2025/ExploratoryResearch/Landings_bySector_Jun152026.png",
  plot = Landings_bySector_plot,
  width = 10,
  height = 10,
  dpi = 300
)
```


## Look into catches from Region 7 & 8
```{r Summarize what they are catching in Region 7 and 8}
colnames(GOC_MarineFisheryLandings0)

Region_summary<- GOC_MarineFisheryLandings %>%
  select(year, OHI_region_est, Major_Functional_Group, weight_kg_corrected) %>%
  group_by(year, OHI_region_est, Major_Functional_Group) %>%
  summarize(
    total_weight_mt= sum(weight_kg_corrected, na.rm = TRUE) / 1000,
    n_records= n()
  )

Region78_summary<- GOC_MarineFisheryLandings %>%
  filter(OHI_region_est %in% c("Region_7", "Region_8")) %>%
  select(year, OHI_region_est, Major_Functional_Group, weight_kg_corrected) %>%
  group_by(year, OHI_region_est, Major_Functional_Group) %>%
  summarize(
    total_weight_mt= sum(weight_kg_corrected, na.rm = TRUE) / 1000,
    n_records= n()
  )

# Remove anything less than 2 metric tons:

Region78_summary<- Region78_summary %>%
  filter(year >= 2018 & year <= 2024) %>%
  group_by(Major_Functional_Group) %>%
  filter(max(total_weight_mt, na.rm = TRUE) >= 5) %>%
  ungroup()
```


```{r Plot what they are catching in Region 7 and 8}
unique(Region78_summary$Major_Functional_Group)

# Colors for these 13 groups:
mfg_cols <- qualitative_hcl(
  17,
  palette = "Dark 3"
)

# Pelagics<- c("Anchovy & Sardine (small pelagics)", 
#              "Pelagic marine fish (excl. anchovy & sardine)")


Region78_plot<- Region78_summary %>%
  filter(year >= 2018 & year <= 2024) %>%
  ggplot(aes(x = year, y = total_weight_mt, group= OHI_region_est)) +
  
    geom_col(
    aes(color= OHI_region_est, fill= OHI_region_est), 
    position = position_dodge(width = 0.8),
    width = 0.7
  ) +
  
  # Colors (most useful when grouping by Major_Functional_Group)
  # scale_fill_manual(values = mfg_cols) +
  # scale_color_manual(values = mfg_cols) +

  # Labels
  labs(
    x = "Year",
    y = "Annual Catch (metric tons)",
    title = "Catch by Major_Functional_Group (Regions 7 and 8)"
  ) +
  
  facet_wrap(~Major_Functional_Group) +
  
  # Minimal theme and sleek styling
  theme_bw(base_size = 14) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey85"),
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 13, color = "grey30")
  )

Region78_plot
```

```{r Save Landings_bySector_plot}
ggsave(
  "/home/ferrer/ohi_goc_prep/AO/v2025/ExploratoryResearch/Rgn78_Landings_byGroup_Jun152026.png",
  plot = Region78_plot,
  width = 10,
  height = 10,
  dpi = 300
)
```


# STEP 3: Pull matching FIS scores (no recalculation) ------------------------------

```{r Load in FIS_StockStatus scores as necessary}
StockStatus_Scores0<- read.csv("/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_Stock_Status_SubScores/Stock_Statuses_Compiled_df_Jun072026.csv")

# Species x year x region; Stock_Status is used as-is.

# CSV was previously titled: StockStatus_Estimates_Jun072026.csv
```

```{r Prep FIS scores}
StockStatus_Scores <- StockStatus_Scores0 %>%
  filter(year >= 2018, year <= 2024) %>%
  mutate(
    OHI_region_est = clean_names_lowercase(OHI_region_est),
    estimated_genus_name = clean_names_lowercase(estimated_genus_name)
  )
# 4130 obs.
```

# Prep AFO dataframe:
```{r Create GOC_ArtisanalLandings}
colnames(GOC_MarineFisheryLandings)
unique(GOC_MarineFisheryLandings$sector_estimate_draft)

Artisanal_Sectors<- c("marine: small-scale capture", "freshwater or marine: small-scale capture ")

GOC_ArtisanalLandings <- GOC_MarineFisheryLandings %>%
  select(state, OHI_region_est,
         office_name_clean, landing_site_clean, 
         year,
         est_aquaculture_production, reported_fleet, sector_estimate_reported, sector_estimate_draft,
         recorded_taxa_name, recorded_taxa_name_clean, common_name_clean, 
         recorded_taxa_group, estimated_taxa_group, Major_Functional_Group,
         estimated_species_name, estimated_genus_name, estimated_family_name,
         landed_weight, live_weight, weight_kg_corrected, value
         ) %>%
  filter(sector_estimate_draft %in% Artisanal_Sectors)
# 977477 obs., 22 vars.

ArtisanalLandings_summary <- GOC_ArtisanalLandings %>%
  group_by(year, OHI_region_est, estimated_genus_name) %>%
  summarize(
    total_landings_kg = sum(weight_kg_corrected, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    estimated_genus_name= clean_names_lowercase(estimated_genus_name),
    OHI_region_est= clean_names_lowercase(OHI_region_est)
  )
# 4295 obs., 4 vars.
```

```{r Merge StockStatus_Scores to ArtisanalLandings_summary}
colnames(StockStatus_Scores)

TEST <- ArtisanalLandings_summary %>%
  left_join(
    StockStatus_Scores %>%
      select(
        OHI_region_est,
        year,
        estimated_genus_name,
        StockStatus_final,
        StockStatus_source,
        StockStatus_est.level
      ),
    by = c(
      "year",
      "estimated_genus_name",
      "OHI_region_est"
    )
  ) %>%
  filter(year <= 2024)

# 4295 obs. (as expected!), 7 vars.
```

```{r penaeus patch}

```



```{r Genus-level stocks: pull all FIS stocks in that genus}
AFO_StockStatus_Scores <- AFO_stocks %>%
  inner_join(
    StockStatus_Scores,
    by = c(
      "year",
      "OHI_region_est",
      "estimated_genus_name"
    )
  )
# 3475 obs. (as expected), 25 varsity.
```

```{r Sanity check AFO FIS scores}
colnames(AFO_StockStatus_Scores)

nrow(AFO_StockStatus_Scores)

summary(AFO_StockStatus_Scores$small_scale_catch_share)
 #   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
 # 0.1561  1.0000  1.0000  0.9574  1.0000  1.0000 
```


# STEP 4: Stock list by OHI region ------------------------------

```{r Region totals for share calculations}
Regional_Catch <- AFO_StockStatus_Scores %>%
  group_by(OHI_region_est) %>%
  summarize(
    total_landings_kg = sum(total_landings_kg, na.rm = TRUE),
    small_scale_landings_kg = sum(small_scale_landings_kg, na.rm = TRUE),
    small_scale_catch_share = small_scale_landings_kg / total_landings_kg,
    .groups = "drop"
  )
```


# STEP 5: Small-scale catch-weighted AFO_score per OHI region ------------------------------

# AFO Scores: Option 1
```{r Small-scale catch-weighted AFO_score per region and year}
# AFO_score = sum(Stock_Status * small_scale_landings_kg) /
#             sum(small_scale_landings_kg)
# Weights are small-scale fishery landings only.

colnames(AFO_StockStatus_Scores)

AFO_Scores_Ex1 <- AFO_StockStatus_Scores %>%
  filter(
    !is.na(StockStatus_final),
    small_scale_landings_kg > 0
  ) %>%
  group_by(year, OHI_region_est) %>%
  summarize(
    AFO_score_mean_SS = mean(StockStatus_final, na.rm = TRUE)*100,
    n_stocks = n(),
    .groups = "drop"
  ) %>%
  arrange(year, OHI_region_est)
```

```{r Plot Example 1}
AFO_Scores_overTime <- AFO_Scores_Ex1 %>%
  ggplot(aes(
    x = year,
    y = AFO_score_mean_SS,
    color = OHI_region_est
  )) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ OHI_region_est, scales = "fixed") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = region_colors) +
  labs(
    title = "AFO - Option 1: mean Stock_Status scores of regional SSF stocks",
    subtitle = paste0(
      "mean Stock_Status of qualifying stocks, not weighted by landings"
    ),
    y = "AFO_score (0 to 100)",
    x = "Year"
  ) +
  ylim(0, 105) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 12, face = "bold")
  )

AFO_Scores_overTime
```


# AFO Scores: Option 2
```{r Small-scale catch-weighted AFO_score per region and year}
# AFO_score = sum(Stock_Status * small_scale_landings_kg) /
#             sum(small_scale_landings_kg)
# Weights are small-scale fishery landings only.

colnames(AFO_StockStatus_Scores)

AFO_Scores_Ex2 <- AFO_StockStatus_Scores %>%
  filter(
    !is.na(StockStatus_final),
    small_scale_landings_kg > 0
  ) %>%
  group_by(year, OHI_region_est) %>%
  summarize(
    AFO_score_SSxLandings = (sum(StockStatus_final * small_scale_landings_kg) /
      sum(small_scale_landings_kg))*100,
    n_stocks = n(),
    small_scale_catch_kg = sum(small_scale_landings_kg),
    total_catch_kg = sum(total_landings_kg),
    .groups = "drop"
  ) %>%
  arrange(year, OHI_region_est)
```

```{r Sanity check AFO regional scores}
colnames(AFO_Scores_Ex2)

summary(AFO_Scores_Ex2$AFO_score)

AFO_Scores_Ex2 %>%
  arrange(year, OHI_region_est)
```

```{r Plot Example 2}
AFO_Scores_overTime <- AFO_Scores_Ex2 %>%
  ggplot(aes(
    x = year,
    y = AFO_score_SSxLandings,
    color = OHI_region_est
  )) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ OHI_region_est, scales = "fixed") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = region_colors) +
  labs(
    title = "AFO - Option 2: mean Stock_Status of SSF stocks weighted by total SSF catch",
    subtitle = paste0(
      "weighted by landings"
    ),
    y = "AFO_score (0 to 100)",
    x = "Year"
  ) +
  ylim(0, 105) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 12, face = "bold")
  )

AFO_Scores_overTime
```


# STEP 7: Plot regional AFO scores over time ------------------------------

```{r Region colors for plot}
region_colors <- c(
  "region_1" = "royalblue",
  "region_2" = "skyblue",
  "region_3" = "orange",
  "region_4" = "sienna1",
  "region_5" = "plum",
  "region_6" = "purple4",
  "region_7" = "aquamarine3",
  "region_8" = "gold2",
  "region_9" = "tomato"
)
```

```{r Plot AFO_score over time by region}
AFO_Scores_overTime <- AFO_Scores %>%
  ggplot(aes(
    x = year,
    y = AFO_score,
    color = OHI_region_est
  )) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ OHI_region_est, scales = "fixed") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = region_colors) +
  labs(
    title = "GOC AFO: Regional catch-weighted Stock_Status scores",
    subtitle = paste0(
      ">= ",
      100 * SSF_CATCH_MIN_SHARE,
      "% small-scale capture stocks; weighted by small-scale landings (kg)"
    ),
    y = "AFO_score (0 to 1)",
    x = "Year"
  ) +
  ylim(0, 1.05) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 12, face = "bold")
  )

AFO_Scores_overTime
```


# FINAL STEP: SAVE ------------------------------

```{r SAVE AFO_StockStatus_Scores and stock-level detail}
# Stock-level detail (species x year x region) for downstream use.
GOC_AFO_Stocks <- AFO_StockStatus_Scores

write.csv(
  AFO_StockStatus_Scores,
  file = file.path(WORK_DIR, "AFO_StockStatus_Scores_Jun012026.csv"),
  row.names = FALSE
)

write.csv(
  GOC_AFO_Stocks,
  file = file.path(WORK_DIR, "GOC_AFO_Stocks_Jun012026.csv"),
  row.names = FALSE
)
```

```{r SAVE stock list and regional AFO scores}
write.csv(
  GOC_AFO_Stock_List,
  file = file.path(WORK_DIR, "GOC_AFO_Stock_List_Jun012026.csv"),
  row.names = FALSE
)

write.csv(
  AFO_Scores,
  file = file.path(WORK_DIR, "AFO_Scores_Jun012026.csv"),
  row.names = FALSE
)
```

```{r SAVE regional AFO score plot}
ggsave(
  filename = file.path(WORK_DIR, "AFO_Scores_Jun012026.png"),
  plot = AFO_Scores_overTime,
  width = 12,
  height = 8,
  dpi = 300
)
```


# PURG as of June 15 2026
# First Strategy: Pull out only SSF stocks ------------------------------

```{r Lets only look at landings post-2018}
Landings_Post.2018<- GOC_MarineFisheryLandings0 %>%
  filter(year>= 2018)

table(Landings_Post.2018$reported_fleet)
# large_scale small_scale 
#      142210     1010533 

table(Landings_Post.2018$sector_estimate_draft)
              # marine: fishery sector yet to be classified 
              #                                        4884 
              #                 marine: large-scale capture 
              #                                      141145 
              #                 marine: small-scale capture 
              #                                      977477
#                 freshwater or marine: small-scale capture 
#                                                     33056 
#                 freshwater or marine: large-scale capture 
#                                                      1065 
# freshwater or marine: fishery sector yet to be classified 
#                                                        30
```


```{r Filter landings and assign stock key}
Landings_Post.2018 <- Landings_Post.2018 %>%
  filter(
    GOC_Include_Exclude == "include",
    year >= 2018,
    year <= 2024,
    !is.na(OHI_region_est),
    OHI_region_est != ""
  ) %>%
  mutate(
    OHI_region_est = clean_names_lowercase(OHI_region_est),
    estimated_species_name = clean_names_lowercase(estimated_species_name),
    estimated_genus_name = clean_names_lowercase(estimated_genus_name),
    sector_estimate_draft = as.character(sector_estimate_draft),
    
    taxon_key = if_else(
      is.na(estimated_species_name) | estimated_species_name == "",
      estimated_genus_name,
      estimated_species_name
    ),
    taxon_level = if_else(
      is.na(estimated_species_name) | estimated_species_name == "",
      "genus",
      "species"
    )
  ) %>%
  filter(!is.na(taxon_key), taxon_key != "")

# 1092972 obs.
```

```{r Summarize total and small-scale catch per stock}
table(Landings_Post.2018$sector_estimate_draft)

SSF_sectors<- c("marine: small-scale capture", "freshwater or marine: small-scale capture")

# Summarize landings at the genus (i.e., stock) level:
GOC_Stocks_summary <- Landings_Post.2018 %>%
  group_by(
    year,
    OHI_region_est,
    taxon_level,
    estimated_genus_name
  ) %>%
  summarize(
    total_landings_kg = sum(weight_kg_corrected, na.rm = TRUE),
    
    small_scale_landings_kg = sum(
      weight_kg_corrected[sector_estimate_draft %in% SSF_sectors],
      na.rm = TRUE
    ),
    
    n_small_scale_records = sum(
      sector_estimate_draft %in% SSF_sectors,
      na.rm = TRUE
    ),
    
    industrial_landings_kg = sum(
      weight_kg_corrected[!(sector_estimate_draft %in% SSF_sectors)],
      na.rm = TRUE
    ),
    
    n_industrial_records = sum(
      !(sector_estimate_draft %in% SSF_sectors),
      na.rm = TRUE
    ),
    
    .groups = "drop"
  ) %>%
  mutate(
    small_scale_catch_share = small_scale_landings_kg / total_landings_kg
  )
# 4034
```

# ** DECISION POINT: Identify stocks with >= 15% small-scale capture

```{r Identify stocks with >= 15% small-scale capture}
SSF_CATCH_MIN_SHARE = 0.15
```

```{r Keep stocks with at least 15% small-scale catch}
# Stocks meeting the small-scale catch threshold in this region x year.
AFO_stocks0 <- GOC_Stocks_summary %>%
  filter(
    small_scale_landings_kg > 0,
    small_scale_catch_share >= SSF_CATCH_MIN_SHARE
  )
# 3581 obs.
```

```{r}
# Remove NAs and unknowns:
AFO_stocks <- AFO_stocks0 %>%
  filter(
    !is.na(estimated_genus_name) &
      estimated_genus_name != "unknown"
  )
# 3475 obs.
```

```{r Count qualifying stocks}
nrow(AFO_stocks)
# 3475 obs.

# n_unique_genus_stocks
unique(AFO_stocks$estimated_genus_name)
# 165 unique stocks / genera
```

