---
title: 'OHI GoC 2024: Lasting Special Places (LSP)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor:
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options:
  chunk_output_type: console
---

# Sentido de Pertenencia: Lasting Special Places

## Script 7: other marine places of importance

Scripts 2 through 6 cover named protected areas, dive sites, archaeology, fishing grounds, and special habitats. Some EWG rows are still left over as "Marine location of importance". In this script I score those leftovers. I try protected area and SIMEC first. If that gives a zero, I use curated legal and literature scores from a Google Sheet.

**What I do**

1. Read the full EWG list and the condition tables from scripts 2 through 5.
2. Keep marine importance rows that were not scored yet (and skip habitat names handled in script 6).
3. Buffer, join the v2 protected area layer, fuzzy match SIMEC, apply Ramsar rules when needed.
4. Fill remaining zeros from `lsps_codified_protections_estimated_condition.csv`, average by region, write CSVs.

```{r}
library(tidyverse)
library(here)
library(sf)
library(mapview)
library(stringdist)
library(stringi)
```

## Read the inventory and earlier scores

```{r}
lsp_google_sheets_dir <- "/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/google_sheets"
out_dir <- "/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/output"

lsp_list_ewg <- read_csv(
  file.path(lsp_google_sheets_dir, "ewg_lsp_inventory_prepared.csv"),
  show_col_types = FALSE
) %>%
  janitor::clean_names()

fishing_site_condition <- read_csv(
  file.path(out_dir, "condition_lsp_fishing_sites.csv"),
  show_col_types = FALSE
)
archaeological_site_condition <- read_csv(
  file.path(out_dir, "condition_lsp_archaeological_sites.csv"),
  show_col_types = FALSE
)
protected_area_condition <- read_csv(
  file.path(out_dir, "condition_lsp_protected_areas.csv"),
  show_col_types = FALSE
)
dive_sites_condition <- read_csv(
  file.path(out_dir, "condition_lsp_dive_sites.csv"),
  show_col_types = FALSE
)
```

## Names already covered in other scripts

I build a simple name list so I can find what is still missing.

```{r}
normalize_name <- function(x) {
  x %>%
    stringi::stri_trans_general("Latin-ASCII") %>%
    str_to_upper() %>%
    str_remove_all(
      "\\bPARQUE NACIONAL\\b|\\bPARQUE\\b|\\bNACIONAL\\b|\\bZONA MARINA\\b|\\bARCHIPIELAGO\\b"
    ) %>%
    str_squish()
}

fishing_site_condition_join <- fishing_site_condition %>%
  dplyr::select(rgn_id, lsp_name = lasting_special_place_lsp, condition = overall_condition)

archaeological_site_condition_join <- archaeological_site_condition %>%
  dplyr::select(rgn_id, lsp_name = lasting_special_place_lsp, condition = overall_condition)

protected_area_condition_join <- protected_area_condition %>%
  mutate(
    lsp_name = case_when(
      !is.na(wdpa_orig_name) ~ wdpa_orig_name,
      is.na(wdpa_orig_name) & !is.na(i_efectividad_name) ~ i_efectividad_name,
      TRUE ~ lasting_special_place_description
    ),
    lsp_name = normalize_name(lsp_name),
    lsp_name = str_replace_all(lsp_name, "LA TOVARA", "La Tovara")
  ) %>%
  dplyr::select(rgn_id, lsp_name, condition = overall_condition)

dive_sites_condition_join <- dive_sites_condition %>%
  mutate(lsp_name = paste0("Dive Site: ", nombre_sitio_de_buceo)) %>%
  dplyr::select(rgn_id, lsp_name, condition = overall_condition)

all_categorical_lsps <- bind_rows(
  fishing_site_condition_join,
  archaeological_site_condition_join,
  protected_area_condition_join,
  dive_sites_condition_join
)

lsps_not_in_categories <- setdiff(
  lsp_list_ewg$lasting_special_place_lsp,
  all_categorical_lsps$lsp_name
)
```

## Missing marine sites to score here

I drop habitat rows that script 6 already covers with polygons. Also note: `rgn_id` should already live on the inventory when exported from the sheets.

```{r}
habitat_names_to_skip <- c(
  "Coquina Reefs in Libertad, Bahia Lobos, and Peñasco - AHW",
  "Only coral reefs in Gulf (Cabo Pulmo/Los Frailes & Bahía San Gabriel on Is. Espíritu Santo). RCB, BW",
  "Massive seagrass beds (Zostera marina), RCB, BW\n\nThe largest and most stable annual seagrass beds (Zostera marina) are in the eastern Pacific. The eelgrass in the Gulf of California is 100% annual, no perennial, and is considered a subspecies, JTC, BW",
  "Puerto Lobos N most mangrove (black mangrove) population, BW",
  "Mangrove coastal lagoons, JTC."
)

lsps_missing_condition <- lsp_list_ewg %>%
  filter(lasting_special_place_lsp %in% lsps_not_in_categories) %>%
  filter(type_of_lsp %in% "Marine location of importance") %>%
  mutate(
    lat = as.numeric(lat),
    long = as.numeric(long)
  ) %>%
  filter(!is.na(lat), !is.na(long)) %>%
  st_as_sf(coords = c("long", "lat"), crs = 4326, remove = FALSE) %>%
  filter(!lasting_special_place_lsp %in% habitat_names_to_skip) %>%
  dplyr::select(rgn_id, lsp_name = lasting_special_place_lsp) %>%
  mutate(rgn_id = str_replace_all(rgn_id, " ", "_"))

mapview(lsps_missing_condition)
```

## Protected areas, SIMEC, and Ramsar

```{r}
all_protected <- st_read(
  here(
    "/home/shares/ohi/OHI_GOC/spatial/protected_areas_wdpa_conanp_and_hem",
    "all_goc_protected_areas_v2_2025.shp"
  ),
  quiet = TRUE
)

lsps_missing_buffered <- st_buffer(
  st_transform(lsps_missing_condition, st_crs(all_protected)),
  dist = 100
)

lsps_missing_protected <- st_join(
  lsps_missing_buffered,
  all_protected %>% dplyr::select(name, geometry),
  join = st_intersects,
  left = TRUE
) %>%
  mutate(in_protected_area = !is.na(name)) %>%
  rename(protected_area_name = name)

i_efectividad_raw <- read_csv(
  file.path(lsp_google_sheets_dir, "simec_i_efectividad_anp_2025.csv"),
  show_col_types = FALSE
) %>%
  janitor::clean_names()

i_efectividad_norm <- i_efectividad_raw %>%
  mutate(nombre_norm = normalize_name(nombre))

lsp_missing_sites_norm <- lsps_missing_protected %>%
  mutate(protected_area_norm = normalize_name(protected_area_name))

matches <- stringdist::amatch(
  lsp_missing_sites_norm$protected_area_norm,
  i_efectividad_norm$nombre_norm,
  method = "jw",
  maxDist = 0.2
)

lsp_missing_w_SIMEC <- lsp_missing_sites_norm %>%
  mutate(
    nombre_matched = i_efectividad_norm$nombre[matches],
    nombre_norm_matched = i_efectividad_norm$nombre_norm[matches]
  ) %>%
  left_join(i_efectividad_raw, by = c("nombre_matched" = "nombre")) %>%
  mutate(
    protected_area_condition = ifelse(!is.na(i_efectividad), i_efectividad, NA_real_)
  )

ramsar_management_plans <- read_csv(
  file.path(lsp_google_sheets_dir, "ramsar_site_management_plan_presence.csv"),
  show_col_types = FALSE
)

# Ramsar only when I have a park name but no SIMEC match
lsp_missing_RAMSAR <- lsp_missing_w_SIMEC %>%
  filter(!is.na(protected_area_name) & is.na(nombre_norm_matched)) %>%
  left_join(
    ramsar_management_plans,
    by = c("protected_area_name" = "wdpa_orig_name")
  ) %>%
  mutate(
    protected_area_condition = case_when(
      if_ramsar_has_management_plan == "y" ~
        max(i_efectividad_raw$i_efectividad, na.rm = TRUE),
      if_ramsar_has_management_plan == "in_progress" ~ 60,
      if_ramsar_has_management_plan == "n" ~
        min(i_efectividad_raw$i_efectividad, na.rm = TRUE),
      TRUE ~ NA_real_
    )
  ) %>%
  dplyr::select(
    rgn_id, lsp_name, protected_area_condition,
    if_ramsar_has_management_plan, protected_area_name,
    in_protected_area, protected_area_norm
  )

lsp_missing_only_conanp_ANPs <- lsp_missing_w_SIMEC %>%
  filter(!protected_area_name %in% lsp_missing_RAMSAR$protected_area_name)

final_lsp_missing_condition <- bind_rows(
  lsp_missing_only_conanp_ANPs,
  lsp_missing_RAMSAR
) %>%
  group_by(rgn_id, lsp_name) %>%
  summarize(
    protected_area_condition = max(protected_area_condition, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    protected_area_condition = ifelse(
      is.infinite(protected_area_condition) | is.nan(protected_area_condition),
      0,
      protected_area_condition
    ),
    lsp_name_join = stringi::stri_trim_both(lsp_name)
  )

mapview(final_lsp_missing_condition, zcol = "protected_area_condition") +
  mapview(all_protected)
```

## Codified scores when still at zero

Google sheet (estimated condition of protection):  
<https://docs.google.com/spreadsheets/d/1RwNTzRZdBIDtzi7d1ceFxhNmVghN47NxTgrhZV8O2SM/edit?gid=193789975#gid=193789975>

```{r}
other_protections_condition_score <- read_csv(
  file.path(lsp_google_sheets_dir, "lsps_codified_protections_estimated_condition.csv"),
  show_col_types = FALSE
) %>%
  mutate(lsp_name_join = stringi::stri_trim_both(lsp_name)) %>%
  dplyr::select(-lsp_name)

lsps_other_protections_condition <- left_join(
  final_lsp_missing_condition,
  other_protections_condition_score,
  by = "lsp_name_join"
)

final_other_protections_condition <- lsps_other_protections_condition %>%
  mutate(
    overall_condition = ifelse(
      protected_area_condition == 0,
      estimated_condition_of_protection,
      protected_area_condition
    ),
    overall_condition = as.numeric(overall_condition)
  ) %>%
  st_drop_geometry()

rgn_final_other_protections_condition <- final_other_protections_condition %>%
  group_by(rgn_id) %>%
  summarize(condition = mean(overall_condition, na.rm = TRUE), .groups = "drop")

dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

write_csv(
  final_other_protections_condition,
  file.path(out_dir, "condition_lsp_other_marine_places.csv")
)
write_csv(
  rgn_final_other_protections_condition,
  file.path(out_dir, "condition_region_other_marine_places.csv")
)
```


