---
title: 'OHI GoC 2024: Lasting Special Places (LSP)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor:
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options:
  chunk_output_type: console
---

# Sentido de Pertenencia: Lasting Special Places

## Script 2: special protected areas on the EWG list

In this script I score the protected areas the EWG named explicitly. When CONANP reports a SIMEC i-efectividad score, I use that as condition. For Ramsar-only sites I follow simple rules based on whether a management plan is present. I write both a site table and a region average so script 8 can merge everything later.

**What I do**

1. Read the combined protected area layer (`all_goc_protected_areas_v2_2025.shp`).
2. Read the EWG special protected areas table and the SIMEC spreadsheet.
3. Join SIMEC onto the EWG list, apply Ramsar rules, average by region, write CSVs.

```{r}
library(tidyverse)
library(here)
library(sf)
library(mapview)
```

## Protected areas (ANPs)

Protected areas on the list are already protected by design. I still need a condition score that reflects management quality. I rely mainly on CONANP's SIMEC i-efectividad.

### Read the combined protected area layer

Let's load the v2 layer from script 1 (WDPA + CONANP + HEM). This is the same layer scripts 3 through 7 should use.

```{r}
all_protected <- st_read(
  here(
    "/home/shares/ohi/OHI_GOC/spatial/protected_areas_wdpa_conanp_and_hem",
    "all_goc_protected_areas_v2_2025.shp"
  ),
  quiet = TRUE
)

mapview(all_protected)
```

Optional: older CBMC GCMP MPAs for context (sites designated before 2020).

```{r}
cbmc_gcmp_2020_mpas <- st_read(
  here(
    "/home/shares/ohi/OHI_GOC/_raw_data/CBMC-GCMP-prioritization-of-diving-sites-main/shp",
    "final_mpa_layer.shp"
  ),
  quiet = TRUE
) %>%
  janitor::clean_names() %>%
  st_zm(drop = TRUE, what = "ZM")

table(st_geometry_type(cbmc_gcmp_2020_mpas))
mapview(cbmc_gcmp_2020_mpas)
```

## Read the EWG list of protected areas

I need the expert names, WDPA labels, SIMEC names, Ramsar flags, and region ids.

```{r}
lsp_google_sheets_dir <- "/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/google_sheets"

ewg_mentioned_protected_areas <- read_csv(
  file.path(lsp_google_sheets_dir, "ewg_special_protected_areas_on_list.csv"),
  show_col_types = FALSE
) %>%
  janitor::clean_names()
```

## Join with i-efectividad

I join the SIMEC table on the CONANP name (`i_efectividad_name` to `nombre`).

```{r}
i_efectividad_raw <- read_csv(
  file.path(lsp_google_sheets_dir, "simec_i_efectividad_anp_2025.csv"),
  show_col_types = FALSE
) %>%
  janitor::clean_names()

i_efectividad_w_ewg_list <- left_join(
  ewg_mentioned_protected_areas,
  i_efectividad_raw,
  by = c("i_efectividad_name" = "nombre")
)
```

## Assign condition (SIMEC and Ramsar rules)

Let's start with SIMEC when it exists. Then I apply Ramsar management plan rules and a few special cases (whale shark refuge, vaquita refuge, Loreto II).

```{r}
protected_area_condition <- i_efectividad_w_ewg_list %>%
  mutate(
    overall_condition = ifelse(is.na(i_efectividad), 0, i_efectividad),
    overall_condition = case_when(
      # Ramsar plan in progress: take the better of SIMEC and 60
      if_ramsar_has_management_plan == "in_progress" ~
        max(c(i_efectividad, 60), na.rm = TRUE),
      # Ramsar plan yes: use at least the highest SIMEC in the table
      if_ramsar_has_management_plan == "y" ~
        max(c(i_efectividad, max(i_efectividad_raw$i_efectividad, na.rm = TRUE)), na.rm = TRUE),
      # Ramsar plan no: cap at the lowest SIMEC in the table
      if_ramsar_has_management_plan == "n" ~
        min(c(i_efectividad, min(i_efectividad_raw$i_efectividad, na.rm = TRUE)), na.rm = TRUE),
      lasting_special_place_description == "Whale shark Area de Refugio in La Paz - AHW , BW" ~ 75,
      # Loreto II is new; leave NA until SIMEC is published
      lasting_special_place_description == "Loreto II" ~ NA_real_,
      lasting_special_place_description == "Área de Refúgio para la Protección de la Vaquita" ~ 75,
      TRUE ~ overall_condition
    ),
    rgn_id = str_replace_all(rgn_id, " ", "_")
  ) %>%
  dplyr::select(
    rgn_id,
    lasting_special_place_description,
    overall_condition,
    i_efectividad_name,
    wdpa_orig_name,
    if_ramsar_has_management_plan,
    ramsar_link
  )
```

I attach geometry from `all_protected` by WDPA name so I can map condition.

```{r}
protected_condition_w_geometry <- left_join(
  protected_area_condition,
  all_protected,
  by = c("wdpa_orig_name" = "name")
) %>%
  st_as_sf()

mapview(protected_condition_w_geometry, zcol = "overall_condition")
```

## Condition per region

I average site conditions within each OHI region.

```{r}
rgn_protected_area_condition <- protected_area_condition %>%
  group_by(rgn_id) %>%
  summarize(
    condition = mean(overall_condition, na.rm = TRUE),
    .groups = "drop"
  )
```

## Write to Mazu

I save site level and region level tables for scripts 7 and 8.

```{r}
out_dir <- "/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/output"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

write_csv(
  protected_area_condition,
  file.path(out_dir, "condition_lsp_protected_areas.csv")
)
write_csv(
  rgn_protected_area_condition,
  file.path(out_dir, "condition_region_protected_areas.csv")
)
```
