---
title: 'OHI GoC 2024: Lasting Special Places (LSP)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor:
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options:
  chunk_output_type: console
---

# Sentido de Pertenencia: Lasting Special Places

## Script 1: combine protected area layers and read the EWG list

In this script I build one protected area layer for the Gulf of California. Let's put CONANP, WDPA, and HEM polygons into the same table, dissolve once if I need a simple footprint map, and write the result to Mazu for scripts 2 through 7. At the end I also load the prepared EWG inventory so I can plot expert points on top of those polygons.

**What I do**

1. Explore CONANP downloads for the Gulf states.
2. Align CRS and geometry types, then bind WDPA, HEM, and CONANP into `all_protected`.
3. Write `all_goc_protected_areas_v2_2025.shp` (and a dissolved version).
4. Read the EWG list and map the points.

```{r}
library(tidyverse)
library(here)
library(sf)
library(mapview)
```

## Make one protected area layer

### Explore CONANP layers

First I read voluntary conservation sites, primary zonification, July 2025 ANPs, and CONANP regions. I only keep Gulf states we score for OHI GoC.

```{r}
gulf_states <- c(
  "Baja California", "Baja California Sur",
  "Sonora", "Sinaloa", "Nayarit", "Jalisco"
)

voluntariamente_conanp <- st_read(
  here(
    "/home/shares/ohi/OHI_GOC/_raw_data/CONANP/619_ADVC_VIGENTES_MARZO_2026",
    "619_ADVC_VIGENTES_MARZO_2026.shp"
  ),
  quiet = TRUE
) %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  janitor::clean_names() %>%
  filter(estado %in% gulf_states)

table(st_geometry_type(voluntariamente_conanp))
mapview(voluntariamente_conanp)
```

```{r}
zonification_primaria <- st_read(
  here(
    "/home/shares/ohi/OHI_GOC/_raw_data/CONANP/SHAPE_ANPS_ZN",
    "ZP_ANPS_22042024.shp"
  ),
  quiet = TRUE
) %>%
  janitor::clean_names() %>%
  filter(estado %in% gulf_states)

mapview(zonification_primaria)
```

```{r}
conanp_anps_julio_2025 <- st_read(
  here(
    "/home/shares/ohi/OHI_GOC/_raw_data/CONANP/232_ANP-ITRF08_04072025",
    "232_ANP-ITRF08_04072025.shp"
  ),
  quiet = TRUE
) %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  janitor::clean_names()

table(st_geometry_type(conanp_anps_julio_2025))
mapview(conanp_anps_julio_2025)
```

```{r}
regions_CONANP <- st_read(
  here(
    "/home/shares/ohi/OHI_GOC/_raw_data/CONANP/Regionalizacion_22052017",
    "Regionalizacion_22052017.shp"
  ),
  quiet = TRUE
) %>%
  janitor::clean_names()
```

### Combine WDPA, HEM, and CONANP

I need WDPA and HEM objects in the R session (from earlier prep, or re-read them if needed). Then I reproject everything to the WDPA CRS, cast to multipolygon, and bind the rows. I keep a `source` column so I know where each polygon came from.

```{r}
# Reproject so all three layers share CRS
hem_protected_anps_reproj <- st_transform(hem_protected_anps, st_crs(wdpa_poly_all))
conanp_anps_julio_2025_reproj <- st_transform(conanp_anps_julio_2025, st_crs(wdpa_poly_all))

wdpa_clean <- st_cast(wdpa_poly_all, "MULTIPOLYGON") %>%
  mutate(source = "wdpa_2025")

hem_clean <- st_cast(hem_protected_anps_reproj, "MULTIPOLYGON") %>%
  mutate(source = "hem")

conanp_clean <- st_cast(conanp_anps_julio_2025_reproj, "MULTIPOLYGON") %>%
  mutate(source = "conanp_2025")

all_protected <- bind_rows(
  wdpa_clean %>% dplyr::select(name, source, geometry),
  hem_clean %>% dplyr::select(name = nombre, source, geometry),
  conanp_clean %>% dplyr::select(name = nombre, source, geometry)
)

mapview(all_protected)

# One multipolygon if I just need coverage for a quick map
all_protected_dissolved <- st_union(all_protected)
mapview(all_protected_dissolved)
```

### Write the combined layer to Mazu

Later scripts (2 through 7) should read this v2 file so everyone uses the same polygons, including Alto Golfo.

```{r}
out_pa_dir <- "/home/shares/ohi/OHI_GOC/spatial/protected_areas_wdpa_conanp_and_hem"
dir.create(out_pa_dir, recursive = TRUE, showWarnings = FALSE)

st_write(
  all_protected,
  file.path(out_pa_dir, "all_goc_protected_areas_v2_2025.shp"),
  delete_dsn = TRUE
)
st_write(
  all_protected_dissolved,
  file.path(out_pa_dir, "all_goc_protected_areas_dissolved_v2_2025.shp"),
  delete_dsn = TRUE
)
```

## Read in the EWG list of LSP sites

Current status for this goal depends on whether a lasting special place has a protection, and how well that protection is managed. For example, if Isla Tiburón sits inside an ANP, its condition may follow CONANP's SIMEC i-efectividad score. Sites outside ANPs get other rules in later scripts.

Types of LSPs and their intended protections:  
<https://docs.google.com/spreadsheets/d/1RwNTzRZdBIDtzi7d1ceFxhNmVghN47NxTgrhZV8O2SM/edit?gid=1758421523#gid=1758421523>

Let's load the prepared inventory and map points on the combined protected areas.

```{r}
lsp_google_sheets_dir <- "/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/google_sheets"

lsp_list_ewg <- read_csv(
  file.path(lsp_google_sheets_dir, "ewg_lsp_inventory_prepared.csv"),
  show_col_types = FALSE
) %>%
  janitor::clean_names()

lsp_list_ewg_pts <- lsp_list_ewg %>%
  mutate(
    lat = as.numeric(lat),
    long = as.numeric(long)
  ) %>%
  filter(!is.na(lat), !is.na(long)) %>%
  st_as_sf(coords = c("long", "lat"), crs = 4326, remove = FALSE)

mapview(lsp_list_ewg_pts, col.regions = "orange") +
  mapview(all_protected, col.regions = "blue")
```
