---
title: "Simple CMSY_Oct 22 2025"
output: html_document
date: "2025-10-22"
---

# PRE-WORK
```{r Loading Libraries}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(tidytext)
library(lubridate)
library(plotly)
library(htmlwidgets)
library(here)
library(janitor)
```

```{r Review the CMSY protocol developed by Ferrer et al. (2021)}
### View PDF in Notebook
pdf_path <- "~/ohi_goc_prep/FP/CMSY/Copy of Ferrer et al. (2021) CMSY Protocol.pdf"

# Get the absolute path
abs_path <- normalizePath("~/ohi_goc_prep/FP/CMSY/Copy of Ferrer et al. (2021) CMSY Protocol.pdf")

# Convert to file URL (Windows or Unix-style)
file_url <- paste0("file:///", URLencode(abs_path))

# Open the PDF
browseURL(file_url)
```


# GENERATING FILES 1 & 2 FROM SCRATCH
Preparing Files 1 and 2
NOTES TO SELF - Oct 22 2025

FILE 1: Create a file with annual landings in kg (and n_count observations) grouped by year per "stock". Here, stock is defined as the total population of taxa X fished within the Gulf of California not including not the Pacific.

- At present, we do not have the municipality lat-longs so I will group stocks by the six state of interest first (Baja, BCS, Son, Sin, Nay, Jal) and then plan to re-make the file (File 1) once we can filter out specific municipalities.
- This means that this timeseries is likely to be an OVER-ESTIMATE of landings.

File 2: Create a file with Region, [Subregion], Stock, Group = Name, English Name, Scientific Name, Source, and a whole bunch of year columns that you populate using the various R files shared by Alfredo. This file can probably be trimmed down but for now I recommend formatting it exactly as it appears from before.

```{r File 1 - Stock Timeseries (JC data)}
# CONAPESCA data shared by JC
# These data were shared with us by Juan Carlos Villaseñor and originally accessed on September 26, 2025 from this GitHub repository: https://github.com/mex-fisheries

MexLandings_2000.present<- readRDS(here("/home/shares/ohi/OHI_GOC/_raw_data/CONAPESCA/d2025/SharedByJuanCarlos/mex_landings/data/clean/mex_landings_2000_present.rds"))
# n = 12,153,879 records from all states in Mexico from January 2000 through January 2025


# Skimming through the dat. What does it look like?
# Yes aquaculture production
# Records up to January, 2025

head(MexLandings_2000.present)

table(MexLandings_2000.present$acuaculture_production)

max(MexLandings_2000.present$year)
max(MexLandings_2000.present$month[MexLandings_2000.present$year])
```

```{r Basic filtering}
States_of_Interest<- c("BAJA CALIFORNIA", "BAJA CALIFORNIA SUR", "SONORA", "SINALOA", "NAYARIT", "JALISCO")

MexLandings_filtered<- MexLandings_2000.present %>% filter(state %in% States_of_Interest)
# n obs.= 4,713,144 (-7,440,735 obs)

table(MexLandings_filtered$fleet)
# large_scale small_scale 
#      181823     1438462 
table(MexLandings_filtered$acuaculture_production) 
#there are also mariculture landings in this dataset and NAs in this column...

FisheryLandings_filtered<- MexLandings_filtered %>% filter(acuaculture_production== "NO" | is.na(acuaculture_production)) # 4684000 obs.

MaricultureLandings_filtered<- MexLandings_filtered %>% filter(acuaculture_production== "SÍ")
# n obs.= 29,144 (-4,684,000 obs)

# NOTE: SOME RECORDS ARE NOT APPENDED WITH INFORMATION ABOUT RECORD TYPE: no information on wild-catch versus aquaculture. For this preliminary analyses, these records are assumed to be landings from fishery operations and are included.

table(FisheryLandings_filtered$year)

# OTRAS_filtered<- MexLandings_filtered %>% filter(main_species_group== "OTRAS")
# n obs.= 715.547 (-1,011,581 obs)
```

```{r Summary table}
# Turn off scientific notation:
options(scipen = 999)

# Create timeseries FILE 1
head(FisheryLandings_filtered)

StateLandings_AnnualSummary_byTaxa <- FisheryLandings_filtered %>%
  select(year, state, main_species_group, live_weight, landed_weight, value) %>%
  group_by(main_species_group, year, state) %>%
  summarise(
    Total_Live_Biomass = sum(live_weight, na.rm = TRUE),
    Total_Landed_Biomass = sum(landed_weight, na.rm = TRUE),
    Total_Value = sum(value, na.rm = TRUE),
    n_Records = n(),
    .groups = "drop"  # optional: ungroup after summarising
  )

TotalLandings_AnnualSummary_byTaxa <- FisheryLandings_filtered %>%
  select(year, main_species_group, live_weight, landed_weight, value) %>%
  group_by(main_species_group, year) %>%
  summarise(
    Total_Live_Biomass = sum(live_weight, na.rm = TRUE),
    Total_Landed_Biomass = sum(landed_weight, na.rm = TRUE),
    Total_Value = sum(value, na.rm = TRUE),
    n_Records = n(),
    .groups = "drop"  # optional: ungroup after summarising
  )

File1_df<- TotalLandings_AnnualSummary_byTaxa

# Re-name column
colnames(File1_df)[colnames(File1_df) == "main_species_group"] <- "Stock"
colnames(File1_df)[colnames(File1_df) == "year"] <- "Year"
```
## SAVE FILE 1
```{r SAVE FILE 1}
write.csv(File1_df, "~/ohi_goc_prep/FP/CMSY/File1_UNDER_CONSTRUCTION_Oct 22 2025.csv", row.names = FALSE)
```

```{r Create File 2 (Metadata File) based on File 1}
# STEP 1:
# Which taxa (stocks) do we want to account for in our File 2?
Stock_names<- unique(File1_df$Stock)
# don’t 57 unique taxa

## Read in the column names of File 2 from Ferrer et al. (2022) CSV file
headers <- names(read.csv("File 2_Stocks Metadata.csv", nrows = 1))

## Create an empty dataframe with the same headers
File2_df <- data.frame(matrix(ncol = length(headers), nrow = 0))
colnames(File2_df) <- headers

## Add Stock_names to the 'Stock' column
File2_df <- File2_df[1:length(Stock_names), ]  # Expand rows
File2_df$Stock <- Stock_names

## Reset rownames / the index column numbered 1-57
rownames(File2_df) <- NULL


# STEP 2:
# Trim some of the columns off this file or re-name:
File2_df <- File2_df[ , !(names(File2_df) %in% c("ID", "Subregion", "Group", "Name")) ]
colnames(File2_df)[colnames(File2_df) == "Keep"] <- "Sufficient_No_Years"
colnames(File2_df)[colnames(File2_df) == "Pick"] <- "Yes_Run"


# STEP 3:
# Fill in some of the columns off this file:
File2_df$Region <- c("GOC")
File2_df$Class <- c("TBD")
File2_df$Source <- c("JC_data")


# STEP 4:
# For species with less than 10 years of data, do not keep (Keep = 0)
StockYears_per_Taxa <- File1_df %>%
  group_by(Stock) %>%
  summarise(n_Stock_Years = n_distinct(Year), .groups = "drop")

StockYears_per_Taxa <- StockYears_per_Taxa %>%
  mutate(Sufficient_No_Years = ifelse(n_Stock_Years < 10, 0, 1))

## How many stocks will we be keeping based on number of years?
table(StockYears_per_Taxa$Sufficient_No_Years)
## We will keep 54 out of 57 stocks, exclude 3

##  Add this information to the File 2 dataframe: join StockYears_per_Taxa to File2_df by "Stock", bringing in the filled Sufficient_No_Years column
File2_df <- File2_df %>%
  select(-Sufficient_No_Years) %>%  # Remove empty column to avoid duplicates
  left_join(
    StockYears_per_Taxa %>% select(Stock, Sufficient_No_Years),
    by = "Stock"
  )

# STEP 5:
# Re-order columns:
File2_df <- File2_df[c(1, 3, 2, 4:5, 6, 37, 39, 38, 7:36)]
```

## SAVE FILE 2
```{r SAVE FILE 2}
# NEXT STEPS: Fill in the rest of column 2 using the ancillary files shared by Alfredo in Ferrer et al. (2022)
write.csv(File2_df, "~/ohi_goc_prep/FP/CMSY/File2_UNDER_CONSTRUCTION_Oct 22 2025.csv", row.names = FALSE)
```

# RUNNING CMSY WITH FILES 1 & 2