---
title: 'OHI GoC 2025: TR Sustainability — Cruise ship pressure'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# TR Sustainability: cruise ship pressure

This script builds a **sustainability pressure** indicator from SECTUR cruise port statistics.

**Idea:** if a region gets more cruise passengers than its pre-COVID average, that is extra day-visitor pressure → lower sustainability score.

This belongs in **Sustainability**, not Quantity. Cruise visits are short port stops and are not split into international vs domestic.

## Method

1. Read annual passengers and ship arrivals (`arribos`) by port (2016–2025).
2. Keep GoC cruise ports only (10 ports); drop pacific-side stops and all other national ports.
3. Sum by OHI region and year.
4. Reference = mean passengers in **2016–2019** by region.
5. Score = `100 × min(1, ref / passengers)` — at or below baseline → 100; double baseline → 50.

Regions with no GoC cruise ports stay **NA** (not zero).

# 1. Setup

```{r}
library(tidyverse)
library(here)
library(readxl)
library(stringi)
library(plotly)

region_levels <- paste0("Region_", 1:9)

rgn_labels <- c(
  "Region_1" = "Western Upper Gulf",
  "Region_2" = "Western Midriff Islands",
  "Region_3" = "Western Central Gulf",
  "Region_4" = "Western Southern Gulf",
  "Region_5" = "Eastern Upper Gulf",
  "Region_6" = "Eastern Midriff Islands",
  "Region_7" = "Southern Sonora Coast",
  "Region_8" = "Sinaloa Coast",
  "Region_9" = "Southeastern Gulf"
)

rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

analysis_years <- 2019:2025
baseline_years <- 2016:2019

cruise_xlsx <- here::here(
  "/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_cruceros/cruceros_datos_2015_2025.xlsx"
)

output_sustainability <- here::here(
  "/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability"
)
```

# 2. Read the cruise workbook

Each sheet = one year. We read **columns B through P** only:

- Column B = port name or `Pasajeros` / `Arribos` label
- Columns D–P = monthly values + **annual total** (last number in the row)

We skip the national block at the top (`Pasajeros Totales`) and only keep port-level rows.

```{r}
excel_sheets(cruise_xlsx) #  [1] "2016" "2017" "2018" "2019" "2020" "2021" "2022" "2023" "2024" "2025"

years_to_read <- 2016:2025
cruise_rows <- list()

for (yr in years_to_read) {
  raw <- read_excel(
    cruise_xlsx,
    sheet = as.character(yr),
    range = "B1:P200",
    col_names = FALSE,
    col_types = rep("text", 15)
  )

  ## within B:P, column 1 = labels (Excel col B), data starts at column 3 (Excel col D)
  current_port <- NA_character_

  for (i in seq_len(nrow(raw))) {
    label <- str_squish(as.character(raw[[1]][i]))
    if (is.na(label) || label == "") next

    skip_row <- str_detect(
      label,
      regex(
        "Total|MOVIMIENTO|Puertos|Pasajeros Totales|movimiento de pasajeros|cifras pueden variar",
        ignore_case = TRUE
      )
    )
    if (skip_row) next

    is_passengers <- str_detect(label, regex("Pasajeros", ignore_case = TRUE)) &&
      !str_detect(label, regex("Totales", ignore_case = TRUE))

    is_arribos <- str_detect(label, regex("Arribos", ignore_case = TRUE))

    if (is_passengers || is_arribos) {
      if (is.na(current_port)) next

      row_nums <- suppressWarnings(as.numeric(unlist(raw[i, 3:ncol(raw)])))
      row_nums <- row_nums[!is.na(row_nums)]
      annual <- if (length(row_nums) > 0) tail(row_nums, 1) else 0

      cruise_rows[[length(cruise_rows) + 1]] <- tibble(
        year = yr,
        port_raw = current_port,
        metric = if (is_passengers) "passengers" else "arribos",
        annual = annual
      )
      next
    }

    current_port <- label
  }
}

cruise_long <- bind_rows(cruise_rows) # looks good!!! finally

cruise_annual <- cruise_long %>%
  mutate(
    port_norm = port_raw %>%
      str_squish() %>%
      stringi::stri_trans_general("Latin-ASCII") %>%
      str_to_lower()
  ) %>%
  pivot_wider(names_from = metric, values_from = annual) %>%
  mutate(
    passengers = replace_na(passengers, 0),
    arribos = replace_na(arribos, 0)
  )

## quick sanity check.  should be ~31 ports and 2 metrics per year (no NA port names)
cruise_annual %>%
  count(year) # good, they were all captured for every year

## looking at cabo for instance.
cruise_long %>%
  filter(str_detect(port_raw, "Cabo San Lucas")) %>%
  filter(metric == "passengers") %>%
  select(year, port_raw, annual) %>%
  arrange(year)
```

# 3. Keep GoC ports and assign regions

**Included:** Cabo San Lucas, Guaymas, La Paz, Loreto, Mazatlán, Pichilingue, Puerto Escondido (B.C.S.), Puerto Vallarta, Santa Rosalía, Topolobampo.

**Excluded (pacific-side only):** Ensenada, Rosarito, San Carlos, and all other ports in the workbook outside of the GoC.

```{r}
cruise_goc <- cruise_annual %>%
  filter(
    !str_detect(port_norm, "ensenada|rosarito|san carlos")
  ) %>%
  mutate(
    rgn_id = case_when(
      str_detect(port_norm, "cabo san lucas") ~ "Region_4",
      str_detect(port_norm, "la paz") ~ "Region_4",
      str_detect(port_norm, "pichilingue") ~ "Region_4", # right next to la paz
      str_detect(port_norm, "loreto") ~ "Region_4",
      str_detect(port_norm, "puerto escondido") & str_detect(port_norm, "b\\.c") ~ "Region_4", # just below loreto
      str_detect(port_norm, "santa rosal") ~ "Region_3",
      str_detect(port_norm, "guaymas") ~ "Region_7",
      str_detect(port_norm, "mazatlan") ~ "Region_8",
      str_detect(port_norm, "topolobampo") ~ "Region_8",
      str_detect(port_norm, "puerto vallarta") ~ "Region_9",
      TRUE ~ NA_character_
    ),
    rgn_id = factor(rgn_id, levels = region_levels)
  ) %>%
  filter(!is.na(rgn_id))

# check that I did not miss one
na_rgnid <- cruise_goc %>% 
  filter(is.na(rgn_id))
unique(na_rgnid$port_raw) # all good!

num_passengers <- cruise_goc %>%
  count(year, port_raw, rgn_id, wt = passengers, name = "passengers") %>%
  arrange(year, rgn_id, desc(passengers))
```

# 4. Sum passengers by region and year

```{r}
cruise_rgn_year <- cruise_goc %>%
  group_by(year, rgn_id) %>%
  summarise(
    passengers = sum(passengers),
    arribos = sum(arribos),
    n_ports = n_distinct(port_raw),
    .groups = "drop"
  )

## fill in zero years for regions that have GoC cruise ports
cruise_regions <- c("Region_3", "Region_4", "Region_5", "Region_7", "Region_8", "Region_9")

cruise_rgn_year <- crossing(
  year = years_to_read,
  rgn_id = factor(cruise_regions, levels = region_levels)
) %>%
  left_join(cruise_rgn_year, by = c("year", "rgn_id")) %>%
  mutate(
    passengers = replace_na(passengers, 0),
    arribos = replace_na(arribos, 0),
    n_ports = replace_na(n_ports, 0)
  ) # now we have all the regions in there.

cruise_rgn_year
```

# 5. Reference points (2016–2019 mean for each rgn)

```{r}
cruise_ref <- cruise_rgn_year %>%
  filter(year %in% baseline_years) %>%
  group_by(rgn_id) %>%
  summarise(
    ref_passengers = mean(passengers),
    ref_arribos = mean(arribos),
    .groups = "drop"
  )

cruise_ref
```

# 6. Current status scores

Higher passengers than baseline → lower score.

```{r}
cruise_scores <- cruise_rgn_year %>%
  left_join(cruise_ref, by = "rgn_id") %>%
  mutate(
    current_status = case_when(
      ref_passengers <= 0 & passengers <= 0 ~ 100,
      ref_passengers <= 0 & passengers > 0 ~ NA_real_,
      passengers <= 0 ~ 100,
      TRUE ~ 100 * pmin(1, ref_passengers / passengers)
    ),
    current_status_arribos = case_when(
      ref_arribos <= 0 & arribos <= 0 ~ 100,
      ref_arribos <= 0 & arribos > 0 ~ NA_real_,
      arribos <= 0 ~ 100,
      TRUE ~ 100 * pmin(1, ref_arribos / arribos)
    ),
    rgn_label = rgn_labels[as.character(rgn_id)],
    sust_type = "cruise_pressure"
  )

cruise_scores_cs <- cruise_scores %>%
  filter(year %in% analysis_years) %>%
  select(
    rgn_id, year, passengers, arribos,
    ref_passengers, ref_arribos,
    current_status, current_status_arribos,
    rgn_label, sust_type
  ) %>%
  arrange(rgn_id, year)

cruise_scores_cs
```

# 7. Format for merging into overall Sustainability

```{r}
## cruise regions only (for inspection)
cruise_cs_for_merge <- cruise_scores_cs %>%
  transmute(
    rgn_id = factor(rgn_id, levels = region_levels),
    year,
    score = current_status,
    sust_type = "cruise_pressure"
  )

## all 9 regions — non-cruise regions are NA
cruise_cs_all_regions <- crossing(
  year = analysis_years,
  rgn_id = factor(region_levels, levels = region_levels)
) %>%
  left_join(cruise_cs_for_merge, by = c("year", "rgn_id")) %>%
  mutate(sust_type = "cruise_pressure")

cruise_cs_all_regions
```

# 8. Plots

```{r}
plot_ly(
  cruise_scores,
  x = ~year,
  y = ~passengers,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines+markers",
  text = ~paste0(rgn_label, "<br>", passengers, " passengers"),
  hoverinfo = "text+x"
) %>%
  layout(
    title = "GoC cruise passengers by OHI region",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Annual cruise passengers")
  )

plot_ly(
  cruise_scores_cs,
  x = ~year,
  y = ~current_status,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines+markers"
) %>%
  layout(
    title = "TR Sustainability: cruise pressure (vs 2016-2019 mean)",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Score (0-100; higher = less pressure)", range = c(0, 105))
  ) %>%
  add_lines(
    x = range(analysis_years),
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "No extra pressure (100)",
    inherit = FALSE
  )
```

# 9. Save outputs

```{r}
write_csv(cruise_annual, here::here(output_sustainability, "cruise_port_annual_2016_2025.csv"))
write_csv(cruise_goc, here::here(output_sustainability, "cruise_goc_ports_mapped_2016_2025.csv"))
write_csv(cruise_scores, here::here(output_sustainability, "cruise_pressure_scores_detail_2016_2025.csv"))
write_csv(cruise_cs_for_merge, here::here(output_sustainability, "tr_sust_cruise_pressure_current_status_2019_2025.csv"))
write_csv(cruise_cs_all_regions, here::here(output_sustainability, "tr_sust_cruise_pressure_current_status_all_regions_2019_2025.csv"))
```

# 10. Passengers per coastal resident (comparison)

Same coastal population as **TR Quality** (`tr_quality.qmd`): INEGI municipal counts for GoC-bordering municipalities, summed by `rgn_id`, linearly interpolated between census years (2000–2020, held flat 2020–2025).

Source: `interpolated_population_per_rgn_coastalmunicipios_2000_2025.csv` in `population_municipal/` (not the `pop_2020.tif` raster — the CSV is already region-level and year-matched).

Per-capita pressure uses the same formula as section 6, but with **passengers / coastal residents** instead of raw passenger counts.

```{r}
pop_interp_path <- here::here(
  "/home/shares/ohi/OHI_GOC/spatial/population_municipal",
  "interpolated_population_per_rgn_coastalmunicipios_2000_2025.csv"
)
if (!file.exists(pop_interp_path)) {
  pop_interp_path <- here::here(
    "population_municipal",
    "interpolated_population_per_rgn_coastalmunicipios_2000_2025.csv"
  )
}

coastal_pop <- read_csv(pop_interp_path, show_col_types = FALSE) %>%
  select(rgn_id, year = year_interpolated, coastal_pop = population_interpolated)

cruise_percap <- cruise_rgn_year %>%
  inner_join(coastal_pop, by = c("rgn_id", "year")) %>%
  mutate(
    passengers_per_resident = passengers / coastal_pop,
    arribos_per_resident = arribos / coastal_pop,
    rgn_label = rgn_labels[as.character(rgn_id)]
  )

cruise_percap_ref <- cruise_percap %>%
  filter(year %in% baseline_years) %>%
  group_by(rgn_id) %>%
  summarise(
    ref_passengers_per_resident = mean(passengers_per_resident),
    ref_arribos_per_resident = mean(arribos_per_resident),
    .groups = "drop"
  )

cruise_percap_scores <- cruise_percap %>%
  left_join(cruise_percap_ref, by = "rgn_id") %>%
  mutate(
    current_status_percap = case_when(
      ref_passengers_per_resident <= 0 & passengers_per_resident <= 0 ~ 100,
      ref_passengers_per_resident <= 0 & passengers_per_resident > 0 ~ NA_real_,
      passengers_per_resident <= 0 ~ 100,
      TRUE ~ 100 * pmin(1, ref_passengers_per_resident / passengers_per_resident)
    )
  )

## side-by-side with raw-passenger score (section 6)
cruise_compare <- cruise_scores %>%
  select(rgn_id, year, passengers, current_status) %>%
  inner_join(
    cruise_percap_scores %>%
      select(
        rgn_id, year, coastal_pop,
        passengers_per_resident, ref_passengers_per_resident,
        current_status_percap
      ),
    by = c("rgn_id", "year")
  ) %>%
  mutate(
    score_diff_percap_minus_raw = current_status_percap - current_status,
    rgn_label = rgn_labels[as.character(rgn_id)]
  )

cruise_compare %>%
  filter(year %in% analysis_years) %>%
  select(
    rgn_id, rgn_label, year, passengers, coastal_pop,
    passengers_per_resident, current_status, current_status_percap,
    score_diff_percap_minus_raw
  ) %>%
  arrange(rgn_id, year)
```

```{r}
plot_ly(
  cruise_percap,
  x = ~year,
  y = ~passengers_per_resident,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines+markers",
  text = ~paste0(
    rgn_label, "<br>",
    round(passengers_per_resident, 2), " passengers / coastal resident"
  ),
  hoverinfo = "text+x"
) %>%
  layout(
    title = "Cruise passengers per coastal resident by OHI region",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Passengers per coastal resident")
  )
```


```{r}
cruise_percap_scores_cs <- cruise_percap_scores %>%
  filter(year %in% analysis_years)

plot_ly(
  cruise_percap_scores_cs,
  x = ~year,
  y = ~current_status_percap,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines+markers",
  text = ~paste0(
    rgn_label, "<br>",
    round(passengers_per_resident, 2), " passengers / resident<br>",
    "Score: ", round(current_status_percap, 1)
  ),
  hoverinfo = "text+x"
) %>%
  layout(
    title = "TR Sustainability: cruise pressure per coastal resident (vs 2016-2019 mean)",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Score (0-100; higher = less pressure)", range = c(0, 105))
  ) %>%
  add_lines(
    x = range(analysis_years),
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "No extra pressure (100)",
    inherit = FALSE
  )
```


```{r}
plot_ly(
  cruise_compare %>% filter(year %in% analysis_years),
  x = ~current_status,
  y = ~current_status_percap,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "markers",
  text = ~paste0(rgn_label, " ", year),
  hoverinfo = "text"
) %>%
  layout(
    title = "Cruise pressure score: raw passengers vs per coastal resident",
    xaxis = list(title = "Score: raw passengers (section 6)", range = c(0, 105)),
    yaxis = list(title = "Score: passengers / coastal resident", range = c(0, 105))
  ) %>%
  add_lines(
    x = c(0, 100),
    y = c(0, 100),
    line = list(color = "black", width = 1, dash = "dash"),
    name = "1:1 line",
    inherit = FALSE
  )
```

```{r}
write_csv(
  cruise_percap_scores,
  here::here(output_sustainability, "cruise_passengers_per_coastal_resident_2016_2025.csv")
)
write_csv(
  cruise_compare,
  here::here(output_sustainability, "cruise_pressure_score_comparison_raw_vs_percap_2016_2025.csv")
)
```

## Notes

- **COVID (2020-2021):** near-zero passengers means that scores are near 100. That means no cruise pressure, not excellent management.
- Next we can bind `tr_sust_cruise_pressure_current_status_2019_2025.csv` with other sustainability indicators in `tr_current_status.qmd` if we think this is a good indicator.
