---
title: 'OHI GoC 2025: SUSTAINABILITY Tourism & Recreation'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
    number-sections: false
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# TR Sustainability: hotel certifications

This script builds the hotel sustainability certification current status for the Gulf of California OHI Tourism & Recreation goal. The score is the share of large DENUE hotels (100+ employees within the 50 km coastal buffer) that hold at least one of the certifications below in a given year.

## Outline

1. [Setup](#1-setup)
2. [Distintivo S](#2-distintivo-s)
3. [DENUE accommodations](#3-denue-accommodations)
4. [Certification programs and DENUE IDs](#4-certification-programs-and-denue-ids)
    - 4a. Green Key
    - 4b. Green Globe
    - 4c. EarthCheck *(One&Only Mandarina still needs a DENUE ID)*
    - 4d. Preferred by Nature *(Secrets Puerto Los Cabos and One&Only Mandarina still need DENUE IDs)*
    - 4e. Calidad Ambiental Turística / PROFEPA PNAA
    - 4f. Distintivo S establishments
    - 4g. Certified hotels master table
5. [Join certifications to DENUE](#5-join-certifications-to-denue)
6. [Large hotels filter (100+ employees)](#6-large-hotels-filter-100-employees)
7. [Current status scores and plots](#7-current-status-scores-and-plots)

I prepared cleaned lists for EarthCheck, Preferred by Nature, and Calidad Ambiental Turística below. For EarthCheck, I found DENUE IDs by hand for most GoC hotels (One&Only Mandarina is still missing). Preferred by Nature and Calidad Ambiental Turística DENUE IDs are in the DENUE `int` folder (Secrets Puerto Los Cabos and One&Only Mandarina still need DENUE IDs for Preferred by Nature).

# 1. Setup

```{r}
library(tidyverse)
library(here)
library(janitor)
library(readxl)
library(sf)
library(stringi)
library(mapview)
library(plotly)
library(gt)
library(tidygeocoder)
```

# 2. Distintivo S

"https://www.gob.mx/presidencia/prensa/el-objetivo-para-turismo-en-el-plan-mexico-es-ser-el-5to-pais-mas-visitado-del-mundo-para-el-2030-presidenta-claudia-sheinbaum?idiom=es"

To achieve the goal of being the fifth most visited country in the world by 2030, it is estimated that it is necessary to increase the number of tourists by three million each year, as well as increase:
 

The economic impact of international tourists by 46 percent.
Hotel rooms by 12 percent.
The generation of tourism jobs by 27 percent.
The tourism Gross Domestic Product (GDP) by 9 percent.
Domestic tourism by 9.8 percent.
*The number of service providers certified in sustainability by 31.5 percent.*
The average expenditure per international tourist by 3.3 percent.
The arrival of international visitors by 30 percent
The increase in DNR revenue by 40.5 percent.

However, instead we will impose a more stringent and easily comprehensible reference point --> any hotel that has 100+ employees in the coastal area of the Gulf (50km) should be certified in Distintivo S.

```{r}
## read the given spreadsheet from SECTUR via PNT
dist_S_2023_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/distintivo_S/tabulated_distintivo_S", "tabula-Directorio_de_empresas_vigentes_Distintivo_S.csv"))
```

```{r}
# # from PNT request
# dist_S_raw <- read_excel(here::here("/home/shares/ohi/OHI_GOC/_raw_data/distintivo_S/requested_PNT/4-260126 Respuesta UDT 340027000021125/Anexo SST 340027000021125","Anexo SST DISTINTIVO S  2025.xls")) %>% 
#   janitor::clean_names() 
# 
# colnames(dist_S_raw)
# 
# # first, define col_types vector based on column positions
# # fecha columns: 3, 15, 16, 18, 19, 21, 22, 24, 25, 27, 28, 30, 31, 33, 34, 36, 37, 39, 40, 42, 43, 45, 46, 48, 49
# col_types_vec <- c(
#   "text",   # 1  no_consecutivos
#   "text",   # 2  no_de_folio_distintivo
#   "date",   # 3  fecha_de_inicio_del_programa_en_la_empresa_inicio_historico
#   "text",   # 4  nombre_comercial_de_la_empresa
#   "text",   # 5  direccion
#   "text",   # 6  colonia
#   "text",   # 7  c_p
#   "text",   # 8  ciudad
#   "text",   # 9  estado
#   "text",   # 10 localidad
#   "text",   # 11 lada
#   "text",   # 12 giro_general
#   "text",   # 13 distintivo_nuevo_renovacion
#   "text",   # 14 renovacion_primera_segunda_tercera_cuarta_quinta
#   "date",   # 15 fecha_emision_2013
#   "date",   # 16 fecha_de_expiracion_2014
#   "text",   # 17 tipo_de_distintivo_17
#   "date",   # 18 fecha_de_emision_2014
#   "date",   # 19 fecha_de_expiracion_2015
#   "text",   # 20 tipo_de_distintivo_20
#   "date",   # 21 fecha_de_emision_2015
#   "date",   # 22 fecha_de_expiracion_2016
#   "text",   # 23 tipo_de_distintivo_23
#   "date",   # 24 fecha_de_emision_2016
#   "date",   # 25 fecha_de_expiracion_2017
#   "text",   # 26 tipo_de_distintivo_26
#   "date",   # 27 fecha_de_emision_2017
#   "date",   # 28 fecha_de_expiracion_2018
#   "text",   # 29 tipo_de_distintivo_29
#   "date",   # 30 fecha_de_emision_2018
#   "date",   # 31 fecha_de_expiracion_2019
#   "text",   # 32 tipo_de_distintivo_32
#   "date",   # 33 fecha_de_emision_2019
#   "date",   # 34 fecha_de_expiracion_2020
#   "text",   # 35 tipo_de_distintivo_35
#   "date",   # 36 fecha_de_emision_2020
#   "date",   # 37 fecha_de_expiracion_2021
#   "text",   # 38 tipo_de_distintivo_38
#   "date",   # 39 fecha_de_emision_2021
#   "date",   # 40 fecha_de_expiracion_2022
#   "text",   # 41 tipo_de_distintivo_41
#   "date",   # 42 fecha_de_emision_2022
#   "date",   # 43 fecha_de_expiracion_2023
#   "text",   # 44 tipo_de_distintivo_44
#   "date",   # 45 fecha_de_emision_2023
#   "date",   # 46 fecha_de_expiracion_2024
#   "text",   # 47 tipo_de_distintivo_47
#   "date",   # 48 fecha_de_emision_2024
#   "date",   # 49 fecha_de_expiracion_2025
#   "text",   # 50 tipo_de_distintivo_50
#   "text",   # 51 status_vigente_vencido
#   "text"    # 52 ultimo_organismo_evaluador
# )
# 
# ## re-read it in now that we know the col_types!!
# dist_S_raw <- read_excel(
#   here::here("/home/shares/ohi/OHI_GOC/_raw_data/distintivo_S/requested_PNT/4-260126 Respuesta UDT 340027000021125/Anexo SST 340027000021125",
#              "Anexo SST DISTINTIVO S  2025.xls"),
#   col_types = col_types_vec
# ) %>% 
#   janitor::clean_names()
```

```{r}
# after manually fixing the historical dates:
col_types_vec <- c(
  "text",   # 1  no_consecutivos
  "text",   # 2  no_de_folio_distintivo
  "date",   # 2.5  ADDING IN UPDATED DATE!!!!
  "date",   # 3  fecha_de_inicio_del_programa_en_la_empresa_inicio_historico
  "text",   # 4  nombre_comercial_de_la_empresa
  "text",   # 5  direccion
  "text",   # 6  colonia
  "text",   # 7  c_p
  "text",   # 8  ciudad
  "text",   # 9  estado
  "text",   # 10 localidad
  "text",   # 11 lada
  "text",   # 12 giro_general
  "text",   # 13 distintivo_nuevo_renovacion
  "text",   # 14 renovacion_primera_segunda_tercera_cuarta_quinta
  "date",   # 15 fecha_emision_2013
  "date",   # 16 fecha_de_expiracion_2014
  "text",   # 17 tipo_de_distintivo_17
  "date",   # 18 fecha_de_emision_2014
  "date",   # 19 fecha_de_expiracion_2015
  "text",   # 20 tipo_de_distintivo_20
  "date",   # 21 fecha_de_emision_2015
  "date",   # 22 fecha_de_expiracion_2016
  "text",   # 23 tipo_de_distintivo_23
  "date",   # 24 fecha_de_emision_2016
  "date",   # 25 fecha_de_expiracion_2017
  "text",   # 26 tipo_de_distintivo_26
  "date",   # 27 fecha_de_emision_2017
  "date",   # 28 fecha_de_expiracion_2018
  "text",   # 29 tipo_de_distintivo_29
  "date",   # 30 fecha_de_emision_2018
  "date",   # 31 fecha_de_expiracion_2019
  "text",   # 32 tipo_de_distintivo_32
  "date",   # 33 fecha_de_emision_2019
  "date",   # 34 fecha_de_expiracion_2020
  "text",   # 35 tipo_de_distintivo_35
  "date",   # 36 fecha_de_emision_2020
  "date",   # 37 fecha_de_expiracion_2021
  "text",   # 38 tipo_de_distintivo_38
  "date",   # 39 fecha_de_emision_2021
  "date",   # 40 fecha_de_expiracion_2022
  "text",   # 41 tipo_de_distintivo_41
  "date",   # 42 fecha_de_emision_2022
  "date",   # 43 fecha_de_expiracion_2023
  "text",   # 44 tipo_de_distintivo_44
  "date",   # 45 fecha_de_emision_2023
  "date",   # 46 fecha_de_expiracion_2024
  "text",   # 47 tipo_de_distintivo_47
  "date",   # 48 fecha_de_emision_2024
  "date",   # 49 fecha_de_expiracion_2025
  "text",   # 50 tipo_de_distintivo_50
  "text",   # 51 status_vigente_vencido
  "text"    # 52 ultimo_organismo_evaluador
)

dist_S_raw_2 <- read_excel(
  here::here("/home/shares/ohi/OHI_GOC/_raw_data/distintivo_S/requested_PNT/4-260126 Respuesta UDT 340027000021125/Anexo SST 340027000021125",
             "updated_dates_Anexo_SST_DISTINTIVO_S_2025.xlsx"),
  col_types = col_types_vec
) %>% 
  janitor::clean_names() %>% 
  mutate(localidad = str_trim(localidad), # remove whitespaces between, before, and after words.
         localidad = str_squish(localidad),
         nombre_comercial_de_la_empresa = str_trim(nombre_comercial_de_la_empresa),
         nombre_comercial_de_la_empresa = str_squish(nombre_comercial_de_la_empresa))
```


## Geocode the hotels

```{r}
# see how estados are formatted
unique(dist_S_raw_2$estado)

#  [1] "Oaxaca"              "Guerrero"            "Quintana Roo"       
#  [4] "Nayarit"             "Jalisco"             "Sonora"             
#  [7] "Baja California Sur" "Ciudad de México"    "Nuevo León"         
# [10] "Colima"              "Yucatán"             "Veracruz"           
# [13] "Hidalgo"             "Chiapas"             "Campeche"           
# [16] "Puebla"              "Morelos"             "Sinaloa"            
# [19] "San Luis Potosí"     "Guanajuato"          "Zacatecas"          
# [22] "Chihuahua"           "Querétaro"           "Estado de México"   
# [25] "Coahuila"            "Baja California"     "Aguascalientes"     
# [28] "TAMAULIPAS"          "Tamaulipas"          "OAXACA"             
# [31] "PUEBLA"              "NAYARIT"             "NUEVO LEÓN"         
# [34] "MORELOS"             "BAJA CALIFORNIA SUR" "QUINTANA ROO" 

dist_S_goc_estados <- dist_S_raw_2 %>% 
  filter(estado %in% c("Nayarit","Jalisco","Sonora","Baja California Sur","Baja  Calfornia Sur", "BAJA CALIFORNIA SUR","Baja California","Sinaloa")) %>%
  mutate(country = "Mexico")

dist_S_goc_estados_geocode <- dist_S_goc_estados %>% 
  tidygeocoder::geocode(city = localidad, country = country, state = estado) # to give each city a lat, long

```

Fix all lat/longs that tidygeocoder was not able to do.

```{r}
isna_lat <- dist_S_goc_estados_geocode %>% 
  filter(is.na(lat))
# [1] "Dreams VillaMagna Resort & Spa"  (found manually on website <https://www.rivieranayarit.com/hotels/nuevo-nayarit-dreams-villamagna/> that address is Paseo de los Cocoteros Lotes 32 y 33, Fraccionamiento Nuevo Vallarta, Nayarit, México, C.P 63732, and then on Google Maps) = 20.712368, -105.302158    
# [2] "Sunscape Puerto Vallarta Resort & Spa" (found manually on Google Maps) = 20.642284, -105.238695
# [3] "IBEROSTAR SELECTION PLAYA MITA" (found manually on Google Maps) = 20.790514, -105.488250

## fix the lats and longs that were NA after `geocode()`
dist_S_goc_estados_geocode_fix <- dist_S_goc_estados_geocode %>% 
  mutate(lat = case_when(
    nombre_comercial_de_la_empresa == "Dreams VillaMagna Resort & Spa" ~ 20.712368,
    nombre_comercial_de_la_empresa == "Sunscape Puerto Vallarta Resort & Spa" ~ 20.642284,
    nombre_comercial_de_la_empresa == "IBEROSTAR SELECTION PLAYA MITA" ~ 20.790514,
    TRUE ~ lat
  ),
  long = case_when(
    nombre_comercial_de_la_empresa == "Dreams VillaMagna Resort & Spa" ~ -105.302158,
    nombre_comercial_de_la_empresa == "Sunscape Puerto Vallarta Resort & Spa" ~ -105.238695,
    nombre_comercial_de_la_empresa == "IBEROSTAR SELECTION PLAYA MITA" ~ -105.488250,
    TRUE ~ long
  ))
```


```{r}
dist_S_geocode_sf <- dist_S_goc_estados_geocode_fix %>% 
  st_as_sf(coords = c("long", "lat"),
           crs = 4326, # WGS84
           remove = FALSE)

library(mapview)
mapview(dist_S_geocode_sf)
plot(dist_S_geocode_sf)

# one in tijuana and one in guadalajara that should be filtered out.
```

## Standardize hotel names

```{r}
library(stringi)

# trime all whitespaces, str to lower, and remove accents
dist_S_all_clean <- dist_S_geocode_sf %>% 
  mutate(
    nombre_comercial_clean = nombre_comercial_de_la_empresa %>% 
      str_to_lower() %>% 
      str_squish() %>% 
      stri_trans_general("Latin-ASCII") %>% # remove accents
      str_replace_all("[^a-z0-9 ]", "") %>% # remove punctuation
      str_replace_all("\\s+", " ") # normalize spaces
  )

# unique(dist_S_all_clean$nombre_comercial_clean)

# go one step futher and remove parts of the names that are not unique to the hotel/business
dist_S_all_clean_2 <- dist_S_all_clean %>% 
  mutate(
    nombre_comercial_clean = nombre_comercial_clean %>% 
      str_replace_all("\\b(hotel|resort|spa|suites|suite|all inclusive|the|grupo)\\b", "") %>% 
      str_squish()
  )

# unique(dist_S_all_clean_2$nombre_comercial_clean)
```


## Classify OHI Rgn ID

```{r}
library(sf)

## first, join with OHI GoC Regions

inland_50 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer","inland_50km_goc.shp"))
dist_S_all_clean_2_trans <- st_transform(dist_S_all_clean_2, st_crs(inland_50))
dist_S_goc_rgn <- st_intersection(dist_S_all_clean_2_trans, inland_50)

mapview(dist_S_goc_rgn, zcol = "rgn_id") + mapview(inland_50, zcol = "rgn_id")  # looks good!
mapview(dist_S_goc_rgn, zcol = "rgn_id") + mapview(inland_50, zcol = "rgn_id") + mapview(dist_S_all_clean_2) # double checking with the original layered on top

## NOTE: we lose a hotel in culiacan if we do 10km, which fits into the 50km buffer.
```

## Last date certified? How many times certified?

```{r}
library(dplyr)

expiration_cols <- c(
  "fecha_de_expiracion_2014",
  "fecha_de_expiracion_2015",
  "fecha_de_expiracion_2016",
  "fecha_de_expiracion_2017",
  "fecha_de_expiracion_2018",
  "fecha_de_expiracion_2019",
  "fecha_de_expiracion_2020",
  "fecha_de_expiracion_2021",
  "fecha_de_expiracion_2022",
  "fecha_de_expiracion_2023",
  "fecha_de_expiracion_2024",
  "fecha_de_expiracion_2025"
)

dist_S_goc_rgn_recent_yr <- dist_S_goc_rgn %>%
  rowwise() %>%  # go row by row, looking across columns for each hotel
  mutate(
    
    # --- most_recent_expiration_year ---
    # we want to know the last year a hotel had an expiration date recorded/when did they last have an active certification??
    most_recent_expiration_year = {
      
      # grab all the expiration column values for this hotel
      vals <- c_across(all_of(expiration_cols))
      
      # find which columns actually have a date (are not NA) because each non-NA means the hotel had a cert that expired that year
      non_empty <- which(!is.na(vals))
      
      # if every expiration column is NA, this hotel has no certification history at all... this would NOT happen, but it is a good check for us to make sure this is working as it should.
      if (length(non_empty) == 0) {
        NA_real_  # just return NA
        
      } else {
        # otherwise, grab the value from the last non-NA column (the most recent one)
        # max(non_empty) gives us the index of the rightmost non-NA column
        last_val <- vals[[max(non_empty)]]
        
        # the columns are POSIXt datetime objects, not plain strings, so we can use format() to pull out just the year
        as.numeric(format(last_val, "%Y"))
      }
    },
    
    # --- gapfilling updated_fecha_de_inicio_del_programa_en_la_empresa_inicio_historico ---
    # in the case where there is an NA in the sheet, replace with the first year they were certified.
    updated_fecha_de_inicio_del_programa_en_la_empresa_inicio_historico = {
      if (is.na(updated_fecha_de_inicio_del_programa_en_la_empresa_inicio_historico)) {
        
        vals <- c_across(all_of(expiration_cols))
        earliest_val <- vals[[min(which(!is.na(vals)))]]
        as.numeric(format(earliest_val, "%Y"))
        
      } else {
        # otherwise, use what was given!
        as.numeric(format(updated_fecha_de_inicio_del_programa_en_la_empresa_inicio_historico, "%Y"))
      }
    },
    
    # --- times_certified ---
    # how many certification cycles has this hotel completed between 2014-2025?
    # each expiration column = one potential cycle.
    # non-NA = they did it, NA = they didn't.
    times_certified = {
      
      # all the expiration columns again for this hotel
      vals <- c_across(all_of(expiration_cols))
      
      # count the non-NAs:
      # times_certified = 1 means they certified once and never came back.
      # times_certified = 12 means they renewed every single year since 2014!!
      sum(!is.na(vals))
    },
    
    # --- years_in_program ---
    # how many years since the hotel first joined? from start year to 2025.
    years_in_program = 2025 - updated_fecha_de_inicio_del_programa_en_la_empresa_inicio_historico,
    
    # --- certification_consistency_pct ---
    # times_certified / years_in_program
    # 100% = renewed every year since joining, 25% = only 1 in every 4 years, and so on.
    certification_consistency_pct = (times_certified / years_in_program) * 100
    
  ) %>%
  ungroup() # drop the rowwise grouping

# see how the distribution looks across hotels
table(dist_S_goc_rgn_recent_yr$times_certified, useNA = "always")


### now, pivot to long format...
# define the expiration cols with their corresponding years
expiration_cols_named <- c(
  "fecha_de_expiracion_2014" = 2014,
  "fecha_de_expiracion_2015" = 2015,
  "fecha_de_expiracion_2016" = 2016,
  "fecha_de_expiracion_2017" = 2017,
  "fecha_de_expiracion_2018" = 2018,
  "fecha_de_expiracion_2019" = 2019,
  "fecha_de_expiracion_2020" = 2020,
  "fecha_de_expiracion_2021" = 2021,
  "fecha_de_expiracion_2022" = 2022,
  "fecha_de_expiracion_2023" = 2023,
  "fecha_de_expiracion_2024" = 2024,
  "fecha_de_expiracion_2025" = 2025
)

dist_S_goc_rgn_long <- dist_S_goc_rgn_recent_yr %>%
  
  # drop the wide expiration columns and pivot only the certified years
  pivot_longer(
    cols = all_of(names(expiration_cols_named)),
    names_to = "cert_year_col",
    values_to = "expiration_date"
  ) %>%
  
  # keep only rows where the hotel actually had a certification that year
  filter(!is.na(expiration_date)) %>%
  
  # extract the year as a clean integer for easy joining later
  mutate(cert_year = as.integer(format(expiration_date, "%Y"))) %>%
  
  # drop the helper column, you don't need it anymore
  dplyr::select(-cert_year_col)
```


```{r}
dist_S_goc_rgn_recent_yr_select <- dist_S_goc_rgn_long %>% 
  dplyr::select(cert_year, most_recent_expiration_year, updated_fecha_de_inicio_del_programa_en_la_empresa_inicio_historico, nombre_comercial_de_la_empresa, nombre_comercial_clean, times_certified, years_in_program, certification_consistency_pct, rgn_id, estado, localidad, direccion, ultimo_organismo_evaluador) %>% 
  st_as_sf()

st_write(dist_S_goc_rgn_recent_yr_select, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int","distintivo_S_certified_50km_030926.gpkg"), append = FALSE) 
```

## Counts by rgn

```{r}
# count the 
dist_S_counts_by_rgn <- dist_S_goc_rgn_recent_yr_select %>% 
  st_drop_geometry() %>% 
  distinct(rgn_id, nombre_comercial_clean, cert_year) %>% 
  group_by(rgn_id, cert_year) %>% 
  summarise(n_hotels = n(), .groups = "drop")
```


## Plotly over time

```{r}
dist_S_counts_by_rgn <- dist_S_counts_by_rgn %>% 
  mutate(year = as.numeric(cert_year))

library(plotly)

plot_ly(
  dist_S_counts_by_rgn,
  x = ~year,
  y = ~n_hotels,
  color = ~factor(rgn_id),
  type = 'scatter',
  mode = 'lines+markers'
) %>% 
  layout(
    title = "Distintivo S Certified Hotels by GoC Region (2013-2025)",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Number of Certified Hotels"),
    legend = list(title = list(text = "Region ID"))
  )

```


# 3. DENUE accommodations

## Compare with DENUE accommodations

First, lets read in the list of hotel economic units for each year, and make a `nombre_comercial_clean` column.

```{r}
denue_int <- here::here(
  "/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas/int"
)

alojamientos_2019_2026_50km <- st_read(file.path(denue_int, "update_alojamientos_2019_2026_goc_50km.shp"))

library(stringi)

alojamientos_2019_2026_50km_clean <- alojamientos_2019_2026_50km %>% 
  mutate(
    nombre_comercial_clean = nom_estab %>% 
      str_to_lower() %>% 
      str_squish() %>% 
      stri_trans_general("Latin-ASCII") %>% # remove accents
      str_replace_all("[^a-z0-9 ]", "") %>% # remove punctuation
      str_replace_all("\\s+", " ") %>%  # normalize spaces
      str_replace_all("\\b(hotel|resort|spa|suites|suite|all inclusive|the|grupo)\\b", "") %>% 
      str_squish()
  ) %>% 
  mutate(
    employees_estimated = case_when(
      per_ocu == "0 a 5 personas" ~ 3,
      per_ocu == "6 a 10 personas" ~ 8,
      per_ocu == "11 a 30 personas" ~ 20.5,
      per_ocu == "31 a 50 personas" ~ 40.5,
      per_ocu == "51 a 100 personas" ~ 75.5,
      per_ocu == "101 a 250 personas" ~ 175.5,
      per_ocu %in% c("251 y m<e1>s personas", "251 y más personas") ~ 300,
      TRUE ~ NA_real_
    )
  )

denue_hotels <- unique(alojamientos_2019_2026_50km_clean$nombre_comercial_clean)
```

Now, let's read in the list of Distintivo S certified hotels that I manually found the DENUE ID for by using setdiff and then walking through any that are not clear and searching their name in `alojamientos_2019_2026_50km_clean`.

Note that before I had the data from SECTUR, I was using three publicly available PDFs released in 2015, 2017, and 2023.  Since I already did work to find the DENUE ID for those hotels listed, I will use what I can to ID the full SECTUR list.

```{r}
dist_S_goc_50km_SECTUR <- st_read(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int","distintivo_S_certified_50km_030926.gpkg"))

sectur_dist_S_businesses <- dist_S_goc_50km_SECTUR %>% 
  dplyr::select(nombre_comercial_de_la_empresa, nombre_comercial_clean, estado, localidad, rgn_id) %>% 
  mutate(across(c(nombre_comercial_de_la_empresa, nombre_comercial_clean, estado, localidad), 
                ~tolower(str_squish(.)))) %>% 
  distinct()
```

```{r}
# use the old key for DENUE IDs, to save time
OLD_dist_s_denue_id <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int","DistintivoS_x_DENUE_Accommodations.csv")) %>% 
  janitor::clean_names()

dist_S_goc_50km_PDFs <- st_read(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int","distintivo_S_certified_50km.shp")) %>% 
  st_drop_geometry()
```

Some exploration before manual work:
```{r}
sectur_dist_S_businesses_names <- unique(sectur_dist_S_businesses$nombre_comercial_clean)

not_in_OLD <- setdiff(sectur_dist_S_businesses_names, unique(OLD_dist_s_denue_id$hotel_resort_name))
##those not present in y (OLD_dist_s) but are present in x (new from SECTUR):
#  [1] "barcelo los cabos palace deluxe"      "casa mexicana de la ballena gris"    
#  [3] "ecoturismo kuyima"                    "secrets marquis los cabos"           
#  [5] "fiesta americana grand los cabos"     "sandos finisterra los cabos"         
#  [7] "vista encantada residences"           "el cid marina mazatlan"              
#  [9] "estrella del mar golf y country club" "dreams villamagna"                   
# [11] "now amber puerto vallarta"  

new_dist_s_establishments_withoutDENUE_yet <- sectur_dist_S_businesses %>% 
  filter(nombre_comercial_clean %in% not_in_OLD)

# to see if any of those without DENUE ID already are easily identifiable in DENUE
clearly_in_DENUE <- setdiff(new_dist_s_establishments_withoutDENUE_yet$nombre_comercial_clean, denue_hotels)

setdiff(unique(new_dist_s_establishments_withoutDENUE_yet$nombre_comercial_clean), clearly_in_DENUE)
## only [1] "fiesta americana grand los cabos"!!!
```

# 4. Certification programs and DENUE IDs

All manual work was done in this spreadsheet: <https://docs.google.com/spreadsheets/d/1QwVcd9Zj3qSvbu0Zshmm64V6zycQ1-pCnCnwZMYhils/edit?gid=2040057616#gid=2040057616> 

The most updated Dist. S list with DENUE IDs is on the sheet tab "all_distS_w_IDs", while the Green Key hotels are under "green_key_w_IDs", and the Green Globe hotels are under "green_globe_hotels". EarthCheck DENUE IDs are in `earthcheck_w_IDs.csv` on the OHI share (One&Only Mandarina is still missing). Preferred by Nature DENUE IDs are in `preferred_by_nature_w_IDs.csv` in the DENUE `int` folder (Secrets Puerto Los Cabos and One&Only Mandarina still need DENUE IDs). Calidad Ambiental Turística DENUE IDs are in `certificados_emitidos_profepa_pnaa_w_denue_ids.csv` in the DENUE `int` folder (same place as Preferred by Nature).

## 4a. Green Key hotels

```{r}
greenkey_hotels <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/FEE_foundation_environmental_education/greenkey_mexico/greenkey_hotels_03_10_26.csv")) %>% 
  janitor::clean_names() %>% 
  mutate(
    nombre_comercial_clean = nombre %>% 
      str_to_lower() %>% 
      str_squish() %>% 
      stri_trans_general("Latin-ASCII") %>% # remove accents
      str_replace_all("[^a-z0-9 ]", "") %>% # remove punctuation
      str_replace_all("\\s+", " ") %>%  # normalize spaces
      str_replace_all("\\b(hotel|resort|spa|suites|suite|all inclusive|the|grupo)\\b", "") %>% 
      str_squish(),
    certification = "green_key"
  )  %>% 
  st_as_sf(coords = c("long", "lat"),
           crs = 4326, # WGS84
           remove = FALSE)

## after manual work in this google spreadsheet: <https://docs.google.com/spreadsheets/d/1QwVcd9Zj3qSvbu0Zshmm64V6zycQ1-pCnCnwZMYhils/edit?gid=2040057616#gid=2040057616> 
greenkey_denue_id <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int","green_key_w_IDs.csv")) %>% 
  janitor::clean_names() %>% 
  mutate(certification = "green_key") %>% 
  filter(!is.na(id_denue)) %>% 
  dplyr::select(nombre_comercial_clean, id_denue, certification) %>% 
  mutate(cert_year = 2025) # waiting on updated greenkey data that has years previous to 2025.

greenkey_goc_rgn <- greenkey_hotels %>% 
  st_transform(st_crs(inland_50)) %>% 
  st_intersection(inland_50) %>% 
  st_drop_geometry() %>% 
  dplyr::select(nombre_comercial_clean, rgn_id) %>% 
  distinct()

all_greenkey_id_w_yr <- greenkey_denue_id %>% 
  inner_join(greenkey_goc_rgn, by = "nombre_comercial_clean") %>% 
  inner_join(
    greenkey_hotels %>% 
      st_drop_geometry() %>% 
      dplyr::select(nombre_comercial_clean, nombre) %>% 
      distinct(),
    by = "nombre_comercial_clean"
  ) %>% 
  dplyr::select(cert_year, certification, id_denue, rgn_id, nombre)
```

## 4b. Green Globe hotels

I pulled the americas member list from the Green Globe website in 2026. Platinum Member status is awarded to members certified for 10 consecutive years, Gold Members for 5 years, and regular Green Globe Members are working towards certification (see `Green Globe American Members - Green Globe.pdf` in `/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int`). Because of that, I only counted gold and platinum members as certified hotels, and I backdated their certification years from the 2026 lookup year. I also assumed that all green globe hotels in this list have more than 100 employees.

```{r}
dms_to_decimal <- function(dms) {
  dms <- str_squish(dms)
  hemisphere <- str_extract(dms, "[NSEW]$")
  sign <- ifelse(hemisphere %in% c("S", "W"), -1, 1)

  numeric_parts <- dms %>%
    str_remove("[NSEW]$") %>%
    str_replace_all("[°'\"]", " ") %>%
    str_squish() %>%
    str_split(" ") %>%
    map(~ as.numeric(.x))

  map2_dbl(numeric_parts, sign, function(parts, s) {
    if (length(parts) < 3 || any(is.na(parts))) {
      return(NA_real_)
    }
    s * (parts[1] + parts[2] / 60 + parts[3] / 3600)
  })
}

green_globe_lookup_year <- 2026L

green_globe_hotels <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int","green_globe_w_IDs.csv")) %>% 
  janitor::clean_names() %>% 
  mutate(
    lat = dms_to_decimal(lat),
    long = dms_to_decimal(long),
    nombre_comercial_clean = nombre_comercial_clean %>% 
      str_to_lower() %>% 
      str_squish(),
    certification = "green_globe"
  )  %>% 
  st_as_sf(coords = c("long", "lat"),
           crs = 4326, # WGS84
           remove = FALSE)

## after i cross referenced green globe hotels in this google spreadsheet: <https://docs.google.com/spreadsheets/d/1QwVcd9Zj3qSvbu0Zshmm64V6zycQ1-pCnCnwZMYhils/edit?gid=625683625#gid=625683625> 
green_globe_denue_id <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int","green_globe_w_IDs.csv")) %>% 
  janitor::clean_names() %>% 
  mutate(
    id_denue = as.numeric(id_denue),
    certification = "green_globe",
    membership_tier = case_when(
      str_detect(str_to_upper(green_globe_certification_status), "PLATINUM") ~ "platinum",
      str_detect(str_to_upper(green_globe_certification_status), "GOLD") ~ "gold",
      TRUE ~ "member"
    ),
    cert_start_year = case_when(
      membership_tier == "platinum" ~ green_globe_lookup_year - 9L,
      membership_tier == "gold" ~ green_globe_lookup_year - 4L,
      TRUE ~ NA_integer_
    )
  ) %>% 
  filter(!is.na(id_denue)) %>% 
  filter(membership_tier %in% c("platinum", "gold")) %>% 
  rowwise() %>%
  mutate(cert_year = list(seq(cert_start_year, green_globe_lookup_year))) %>%
  unnest(cert_year) %>%
  ungroup() %>%
  filter(cert_year %in% 2019:2026) %>% 
  dplyr::select(nombre, nombre_comercial_clean, id_denue, certification, cert_year)

## i_excluded_regular_green_globe_members_because_they_are_still_working_towards_certification
green_globe_members_excluded <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int","green_globe_w_IDs.csv")) %>% 
  janitor::clean_names() %>% 
  filter(!str_detect(str_to_upper(green_globe_certification_status), "GOLD|PLATINUM")) %>% 
  dplyr::select(nombre, green_globe_certification_status)

green_globe_large_hotel_ids <- green_globe_denue_id %>% 
  distinct(id_denue) %>% 
  pull(id_denue)

green_globe_goc_rgn <- green_globe_hotels %>% 
  st_transform(st_crs(inland_50)) %>% 
  st_intersection(inland_50) %>% 
  st_drop_geometry() %>% 
  dplyr::select(nombre_comercial_clean, rgn_id) %>% 
  distinct()

all_green_globe_id_w_yr <- green_globe_denue_id %>% 
  inner_join(green_globe_goc_rgn, by = "nombre_comercial_clean") %>% 
  dplyr::select(cert_year, certification, id_denue, rgn_id, nombre)
```

## 4c. EarthCheck hotels

I read the EarthCheck members export (`EarthCheckMembers_072326.xlsx`), kept Mexico accommodation / villa / resort / hostel members, and intersected them with the 50 km GoC buffer. I backdated certification years from each hotel's `achievement_level` and `achievement_date`, using the tier rules in `EarthCheck Certified - EarthCheck.pdf` (Silver after 1 year, Gold after 5, Platinum after 10, Master after 15 consecutive years of EarthCheck certification). When `achievement_level` is NA but dates exist, I assumed the hotel was only certified for that one year (`year(achievement_date)`). When both level and dates are NA, like Villa Premiere Boutique Hotel & Romantic Getaway, I left the hotel out of the year list for now.

I found DENUE IDs by hand for most GoC hotels and saved them in `earthcheck_w_IDs.csv`. One&Only Mandarina is still missing a DENUE ID even though the hotel is listed on the resort website (<https://www.oneandonlyresorts.com/mandarina>).

```{r}
clean_hotel_name <- function(x) {
  x %>%
    str_to_lower() %>%
    str_squish() %>%
    stri_trans_general("Latin-ASCII") %>%
    str_replace_all("[^a-z0-9 ]", "") %>%
    str_replace_all("\\s+", " ") %>%
    str_replace_all("\\b(hotel|resort|spa|suites|suite|all inclusive|the|grupo)\\b", "") %>%
    str_squish()
}

goc_estados <- c(
  "Baja California", "Baja California Sur", "Sonora",
  "Sinaloa", "Nayarit", "Jalisco"
)

earthcheck_raw <- read_excel(here::here("/home/shares/ohi/OHI_GOC/_raw_data/EarthCheck/EarthCheckMembers_072326.xlsx")) %>%
  janitor::clean_names()

earthcheck_mexico_accommodation <- earthcheck_raw %>%
  filter(
    str_detect(str_to_lower(country), "mexico"),
    str_detect(
      str_to_lower(sector),
      "accommodation|resort|villa|hostel|hotel"
    )
  ) %>%
  mutate(
    nombre = coalesce(member_operation, member),
    nombre_comercial_clean = clean_hotel_name(nombre),
    achievement_date = as.Date(achievement_date),
    achievement_expiry_date = as.Date(achievement_expiry_date),
    certification = "earthcheck",
    id_denue = NA_real_
  ) %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326, remove = FALSE)

earthcheck_goc <- earthcheck_mexico_accommodation %>%
  st_transform(st_crs(inland_50)) %>%
  st_intersection(inland_50)

earthcheck_goc_rgn <- earthcheck_goc %>%
  st_drop_geometry() %>%
  dplyr::select(
    nombre,
    nombre_comercial_clean,
    sector,
    address,
    achievement_level,
    earth_check_product,
    achievement_date,
    achievement_expiry_date,
    rgn_id,
    id_denue,
    certification
  ) %>%
  distinct()

## backdate certification years from achievement_level (see EarthCheck Certified - EarthCheck.pdf)
earthcheck_years <- earthcheck_goc_rgn %>%
  mutate(
    lookup_year = year(coalesce(achievement_date, achievement_expiry_date))
  ) %>%
  rowwise() %>%
  mutate(
    cert_year = list({
      if (is.na(lookup_year)) {
        integer(0)
      } else if (is.na(achievement_level)) {
        lookup_year
      } else {
        years_back <- dplyr::case_when(
          str_detect(str_to_lower(achievement_level), "master") ~ 14L,
          str_detect(str_to_lower(achievement_level), "platinum") ~ 9L,
          str_detect(str_to_lower(achievement_level), "gold") ~ 4L,
          str_detect(str_to_lower(achievement_level), "silver") ~ 0L,
          str_detect(str_to_lower(achievement_level), "bronze") ~ 0L,
          TRUE ~ 0L
        )
        seq(lookup_year - years_back, lookup_year)
      }
    })
  ) %>%
  unnest(cert_year) %>%
  ungroup() %>%
  filter(cert_year %in% 2019:2026) %>%
  dplyr::select(
    cert_year, certification, id_denue, rgn_id, nombre,
    nombre_comercial_clean, achievement_level, achievement_date, address
  ) %>%
  distinct()

earthcheck_unresolved_years <- earthcheck_goc_rgn %>%
  filter(is.na(achievement_level), is.na(achievement_date), is.na(achievement_expiry_date)) %>%
  dplyr::select(nombre, nombre_comercial_clean, achievement_level, achievement_date, achievement_expiry_date, rgn_id, address) %>%
  distinct()

## audit file for hotels still missing DENUE IDs
earthcheck_for_denue_match <- earthcheck_goc_rgn %>%
  dplyr::select(
    nombre, nombre_comercial_clean, sector, address, rgn_id,
    achievement_level, achievement_date, achievement_expiry_date, id_denue
  ) %>%
  mutate(certification = "earthcheck") %>%
  distinct()

write_csv(
  earthcheck_for_denue_match,
  here::here("earthcheck_for_denue_match_082726.csv")
)

## after manual work in earthcheck_w_IDs.csv
earthcheck_denue_id <- read_csv(
  here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int", "earthcheck_w_IDs.csv"),
  show_col_types = FALSE
) %>%
  janitor::clean_names() %>%
  mutate(
    id_denue = as.numeric(id_denue),
    certification = "earthcheck"
  ) %>%
  filter(!is.na(id_denue)) %>%
  dplyr::select(nombre_comercial_clean, id_denue, certification) %>%
  distinct()

## Grupo Vidanta Vallarta - Hotel maps to two DENUE units (Marina Vallarta and Nuevo Vallarta)
earthcheck_missing_denue <- earthcheck_goc_rgn %>%
  anti_join(earthcheck_denue_id, by = "nombre_comercial_clean") %>%
  distinct(nombre, nombre_comercial_clean, rgn_id, address)

all_earthcheck_id_w_yr <- earthcheck_years %>%
  dplyr::select(-id_denue) %>%
  inner_join(earthcheck_denue_id, by = c("nombre_comercial_clean", "certification")) %>%
  dplyr::select(cert_year, certification, id_denue, rgn_id, nombre) %>%
  distinct()

if (nrow(all_earthcheck_id_w_yr) == 0) {
  all_earthcheck_id_w_yr <- tibble(
    cert_year = integer(),
    certification = character(),
    id_denue = numeric(),
    rgn_id = character(),
    nombre = character()
  )
}

earthcheck_goc_rgn
earthcheck_unresolved_years
earthcheck_missing_denue
all_earthcheck_id_w_yr
```

## 4d. Preferred by Nature hotels

I read Preferred by Nature's Mexico tourism certificates (`Preferred_by_Nature_Certificates-2026-08-19.xlsx`), kept GoC states (Baja California, BCS, Sonora, Sinaloa, Nayarit, Jalisco), and expanded years from first issued through expiry (Active and Terminated both count for years they were certified). I assigned DENUE IDs and `rgn_id` by hand in `preferred_by_nature_w_IDs.csv` on the OHI share.

```{r}
preferred_by_nature_raw <- read_excel(
  here::here("/home/shares/ohi/OHI_GOC/_raw_data/Preferred_by_Nature/Preferred_by_Nature_Certificates-2026-08-19.xlsx"),
  skip = 5,
  col_names = c(
    "row_num",
    "certificate_name",
    "blank",
    "certificate_registration_code",
    "status",
    "country",
    "organisation_name",
    "certificate_category",
    "certificate_issued",
    "first_issued",
    "certificate_expiry",
    "primary_address",
    "city",
    "state_province",
    "country_cert"
  )
) %>%
  filter(
    !is.na(organisation_name),
    !str_detect(str_to_lower(organisation_name), "^total$"),
    !is.na(certificate_registration_code)
  )

parse_flex_date <- function(x) {
  if (inherits(x, "POSIXt")) {
    return(as.Date(x))
  }
  if (inherits(x, "Date")) {
    return(x)
  }
  if (is.numeric(x)) {
    return(as.Date(x, origin = "1899-12-30"))
  }

  x_chr <- as.character(x)
  x_chr[is.na(x) | x_chr %in% c("", "NA", "Total", "Count")] <- NA_character_

  as.Date(
    lubridate::parse_date_time(
      x_chr,
      orders = c("mdy", "ymd", "dmy", "Ymd"),
      quiet = TRUE,
      tz = "UTC"
    )
  )
}

preferred_by_nature_goc_estados <- preferred_by_nature_raw %>%
  mutate(
    estado = str_to_title(str_squish(state_province)),
    nombre = organisation_name,
    nombre_comercial_clean = clean_hotel_name(nombre),
    certificate_issued = parse_flex_date(certificate_issued),
    first_issued = parse_flex_date(first_issued),
    certificate_expiry = parse_flex_date(certificate_expiry),
    certification = "preferred_by_nature",
    id_denue = NA_real_,
    rgn_id = NA_character_
  ) %>%
  filter(estado %in% goc_estados) %>%
  dplyr::select(
    nombre,
    nombre_comercial_clean,
    certificate_registration_code,
    status,
    estado,
    city,
    primary_address,
    certificate_issued,
    first_issued,
    certificate_expiry,
    certification,
    id_denue,
    rgn_id
  ) %>%
  distinct()

preferred_by_nature_for_denue_match <- preferred_by_nature_goc_estados

write_csv(
  preferred_by_nature_for_denue_match,
  file.path(denue_int, "preferred_by_nature_for_denue_match_082726.csv")
)

preferred_by_nature_years <- preferred_by_nature_goc_estados %>%
  filter(!is.na(first_issued) | !is.na(certificate_issued)) %>%
  mutate(
    start_year = year(coalesce(first_issued, certificate_issued)),
    end_year = year(coalesce(certificate_expiry, certificate_issued, first_issued))
  ) %>%
  rowwise() %>%
  mutate(cert_year = list(seq(start_year, end_year))) %>%
  unnest(cert_year) %>%
  ungroup() %>%
  filter(cert_year %in% 2019:2026) %>%
  dplyr::select(
    cert_year, certification, nombre, nombre_comercial_clean,
    certificate_registration_code, status, estado
  ) %>%
  distinct()

preferred_by_nature_denue_id <- read_csv(
  file.path(denue_int, "preferred_by_nature_w_IDs.csv"),
  show_col_types = FALSE
) %>%
  janitor::clean_names() %>%
  mutate(
    id_denue = as.numeric(id_denue),
    certification = "preferred_by_nature"
  ) %>%
  filter(!is.na(id_denue), !is.na(rgn_id)) %>%
  dplyr::select(nombre_comercial_clean, id_denue, rgn_id, certification) %>%
  distinct()

all_preferred_by_nature_id_w_yr <- preferred_by_nature_years %>%
  inner_join(preferred_by_nature_denue_id, by = c("nombre_comercial_clean", "certification")) %>%
  dplyr::select(cert_year, certification, id_denue, rgn_id, nombre) %>%
  distinct()

if (nrow(all_preferred_by_nature_id_w_yr) == 0) {
  all_preferred_by_nature_id_w_yr <- tibble(
    cert_year = integer(),
    certification = character(),
    id_denue = numeric(),
    rgn_id = character(),
    nombre = character()
  )
}

preferred_by_nature_goc_estados
preferred_by_nature_for_denue_match
all_preferred_by_nature_id_w_yr
```

## 4e. Calidad Ambiental Turística (PROFEPA PNAA)

I used the GoC extract of PROFEPA PNAA certificates and assigned DENUE IDs and `rgn_id` by hand in `certificados_emitidos_profepa_pnaa_w_denue_ids.csv` in the DENUE `int` folder. Each row is already a certification year (`cert_year`). When one certificate maps to more than one DENUE ID, I split it into separate rows (or the code below splits comma-separated IDs). Rows without a DENUE ID or region are left out of the score for now.

```{r}
calidad_ambiental_w_denue_ids <- read_csv(
  file.path(denue_int, "certificados_emitidos_profepa_pnaa_w_denue_ids.csv"),
  show_col_types = FALSE
) %>%
  janitor::clean_names()

calidad_ambiental_denue_id <- calidad_ambiental_w_denue_ids %>%
  mutate(
    cert_year = as.integer(coalesce(cert_year, ano)),
    id_denue = str_trim(as.character(id_denue)),
    id_denue = if_else(id_denue %in% c("", "NA"), NA_character_, id_denue),
    rgn_id = str_squish(as.character(rgn_id)),
    rgn_id = if_else(rgn_id %in% c("", "NA"), NA_character_, rgn_id),
    certification = coalesce(certification, "calidad_ambiental_turistica")
  ) %>%
  separate_rows(id_denue, sep = ",\\s*") %>%
  mutate(
    id_denue = str_trim(id_denue),
    id_denue = if_else(id_denue %in% c("", "NA"), NA_character_, id_denue),
    id_denue = as.numeric(id_denue)
  ) %>%
  filter(!is.na(id_denue), !is.na(rgn_id)) %>%
  dplyr::select(cert_year, certification, id_denue, rgn_id, nombre) %>%
  distinct()

all_calidad_ambiental_id_w_yr <- calidad_ambiental_denue_id

if (nrow(all_calidad_ambiental_id_w_yr) == 0) {
  all_calidad_ambiental_id_w_yr <- tibble(
    cert_year = integer(),
    certification = character(),
    id_denue = numeric(),
    rgn_id = character(),
    nombre = character()
  )
}

calidad_ambiental_w_denue_ids
all_calidad_ambiental_id_w_yr
```

## 4f. Distintivo S establishments

`dist_S_counts_by_rgn` is only a summary table I used for the plotly chart above (hotel counts per region per year). It does not have `id_denue`, so I could not use it to join with DENUE accommodations.

For the current status score, I needed two pieces: (1) which DENUE economic unit each hotel is (`dist_s_denue_id`, from my google spreadsheet where I found the IDs by hand), and (2) which years and region each hotel was certified (`dist_S_goc_50km_SECTUR`, from the sectur gpkg I wrote earlier). I joined those together to build `all_dist_s_id_w_yr`.

```{r}
dist_S_goc_50km_SECTUR <- st_read(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int","distintivo_S_certified_50km_030926.gpkg"))

dist_s_denue_id <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int","DistintivoS_x_DENUE_Accommodations_all_distS_w_IDs.csv")) %>%
  janitor::clean_names() %>% 
  filter(!is.na(id_denue)) %>%  # if it did not have a denue id, it is either not a hotel or it did not have enough information to know which hotel it was. This was done through manual searches *after* using setdiff first.  the DENUE ID could also be NA if I found duplicates between the new SECTUR-given data and the PDFs we originally used.
  dplyr::select(nombre_comercial_clean = hotel_resort_name, id_denue) %>% 
  mutate(certification = "distintivo_s")

all_dist_s_id_w_yr <- dist_s_denue_id %>% 
  inner_join(
    dist_S_goc_50km_SECTUR %>% 
      st_drop_geometry() %>% 
      dplyr::select(
        nombre_comercial_clean,
        nombre = nombre_comercial_de_la_empresa,
        cert_year,
        rgn_id
      ) %>% 
      distinct(),
    by = "nombre_comercial_clean"
  ) %>% 
  dplyr::select(cert_year, certification, id_denue, rgn_id, nombre)
```

## 4g. Certified hotels master table

Join them so that we have one list of all certified hotels. Before I calculated current status, I built `certified_hotels_master` so I could check that every hotel had a `nombre`, `rgn_id`, `cert_year`, and `id_denue`. EarthCheck, Preferred by Nature, and Calidad Ambiental Turística are included for rows where I found DENUE IDs and regions.

```{r}
certified_hotels_master <- bind_rows(
  all_dist_s_id_w_yr,
  all_greenkey_id_w_yr,
  all_green_globe_id_w_yr,
  all_earthcheck_id_w_yr,
  all_preferred_by_nature_id_w_yr,
  all_calidad_ambiental_id_w_yr
) %>% 
  distinct(rgn_id, cert_year, id_denue, certification, nombre) %>% 
  arrange(rgn_id, cert_year, nombre)

certified_hotels_master

write_csv(
  certified_hotels_master,
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability",
    "sust_cert_hotels_master_082726.csv"
  )
)

# check the number of hotels:
length(unique(certified_hotels_master$id_denue))
table(certified_hotels_master$certification, useNA = "always")
```

# 5. Join certifications to DENUE

```{r}
certified_hotels_master <- read_csv(
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability",
    "sust_cert_hotels_master_082726.csv"
  ),
  show_col_types = FALSE
)

# sf object must be first in the join or mapview loses geometry
join_denue_alojamientos_certified_and_not <- full_join(
  alojamientos_2019_2026_50km_clean,
  certified_hotels_master,
  by = c("id" = "id_denue", "data_year" = "cert_year", "rgn_id")
) %>%
  filter(data_year %in% 2019:2026) %>%
  mutate(certification = ifelse(is.na(certification), "no_certification", certification))

join_denue_alojamientos_certified <- alojamientos_2019_2026_50km_clean %>%
  inner_join(
    certified_hotels_master,
    by = c("id" = "id_denue", "data_year" = "cert_year", "rgn_id")
  )

mapview(join_denue_alojamientos_certified, zcol = "certification")
```

<!-- ```{r} -->
<!-- # looking at including OSM hotel locations -->
<!-- new_hotel_data <- st_read(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025","OpenStreetMap_Tourist_Attractions_for_North_America.geojson")) %>%  -->
<!--   st_zm() -->
<!--   filter(addr_country %in% "MX") -->

<!-- mapview(new_hotel_data) -->

<!-- t2 <- new_hotel_data %>%  -->
<!--   filter(tourism %in% "hotel") -->

<!-- mapview(t2) -->

<!-- t2_mex <- t2 %>%  -->
<!--   filter(addr_country %in% "MX") # not validated data. loss of MX hotels  -->
<!-- mapview(t2_mex) -->

<!-- inland_50 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer","inland_50km_goc.shp")) -->
<!-- t2_trans <- st_transform(t2, st_crs(inland_50)) -->
<!-- t2_50km <- st_intersection(t2_trans, inland_50) -->

<!-- mapview(t2_50km, zcol = "rgn_id") + mapview(alojamientos_2019_2026_50km_clean) -->

<!-- summary(t2_50km) -->
<!-- ``` -->


# 6. Large hotels filter (100+ employees)

For current status, I compared two counts for each region and year from 2019 to 2024: (1) the total number of unique DENUE accommodations with 100+ employees, plus any green globe hotels below that threshold in years they were certified (which I assumed are still large hotels), and (2) how many of those same DENUE ids had a sustainability certification that year.

```{r}
analysis_years <- 2019:2025

green_globe_large_hotel_yrs <- certified_hotels_master %>% 
  filter(certification == "green_globe") %>% 
  dplyr::select(id_denue, cert_year) %>% 
  distinct()

denue_large_hotels <- alojamientos_2019_2026_50km_clean %>% 
  st_drop_geometry() %>% 
  filter(data_year %in% analysis_years) %>% 
  mutate(cert_year = data_year) %>% 
  left_join(
    green_globe_large_hotel_yrs %>% mutate(is_green_globe_cert_year = TRUE),
    by = c("id" = "id_denue", "cert_year")
  ) %>% 
  mutate(
    is_large_hotel = employees_estimated >= 100 | is_green_globe_cert_year %in% TRUE
  ) %>% 
  filter(is_large_hotel) %>% 
  dplyr::select(rgn_id, cert_year, id_denue = id, nombre = nom_estab, employees_estimated) %>% 
  distinct()

total_large_hotels_by_rgn_yr <- denue_large_hotels %>% 
  group_by(rgn_id, cert_year) %>% 
  summarize(total_large_hotels = n_distinct(id_denue), .groups = "drop")

certified_ids_by_yr <- certified_hotels_master %>% 
  filter(cert_year %in% analysis_years) %>% 
  dplyr::select(id_denue, cert_year) %>% 
  distinct()

certified_in_scope <- denue_large_hotels %>% 
  inner_join(certified_ids_by_yr, by = c("id_denue", "cert_year")) %>% 
  group_by(rgn_id, cert_year) %>% 
  summarize(certified = n_distinct(id_denue), .groups = "drop")

large_hotels_denue <- total_large_hotels_by_rgn_yr

large_hotels_denue_w_cert <- total_large_hotels_by_rgn_yr %>% 
  left_join(certified_in_scope, by = c("rgn_id", "cert_year")) %>% 
  replace_na(list(certified = 0L)) %>% 
  mutate(
    no_certification = total_large_hotels - certified,
    current_status = (certified / total_large_hotels) * 100
  )
```


# 7. Current status scores and plots

I saved two output files to `output_sustainability`: `sust_cert_hotels_master_082726.csv` (hotel level audit with nombre, rgn_id, cert_year, and id_denue) and `sust_cert_current_status_final_2019_2026_100over_082726.csv` (region by year scores). These names reflect all hotel sustainability certifications included in the master table once DENUE IDs are available.

```{r}

### table to visualize
region_levels <- paste0("Region_", 1:9)

large_hotels_denue_w_cert %>%
  mutate(
    rgn_id = factor(rgn_id, levels = region_levels),
    current_status = current_status / 100
  ) %>%
  arrange(rgn_id) %>%
  gt() %>%
  cols_label(
    rgn_id = "Region",
    cert_year = "Year",
    certified = "Certified Hotels (all programs)",
    total_large_hotels = "Total Large Hotels (+100 employees)",
    current_status = "Certification Coverage"
  ) %>%
  fmt_percent(
    columns = current_status,
    decimals = 1
  ) %>%
  cols_align(
    align = "center",
    everything()
  ) %>%
  tab_header(
    title = "Sustainability Certification Coverage",
    subtitle = "Certified hotels (Distintivo S, Green Key, Green Globe, EarthCheck, Preferred by Nature, Calidad Ambiental Turística) as a share of total large hotels (+100 employees)"
  ) %>%
  opt_table_outline() %>%
  opt_all_caps()

```

## Compare to July 2026 run

The previous score used Distintivo S, Green Key, and Green Globe only (`sust_cert_current_status_final_2019_2024_100over_070126.csv` on the OHI share). The table below puts that run next to the August 2026 run (all six programs, 2019 to 2026). Change is in percentage points (pp).

```{r}
output_sustainability <- here::here(
  "/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability"
)

sust_cert_current_status_jul2026 <- read_csv(
  here::here(
    output_sustainability,
    "sust_cert_current_status_final_2019_2024_100over_070126.csv"
  ),
  show_col_types = FALSE
) %>%
  rename(current_status_jul2026 = current_status)

sust_cert_comparison %>%
  mutate(
    rgn_id = factor(rgn_id, levels = region_levels),
    current_status_jul2026_pct = current_status_jul2026 / 100,
    current_status_aug2026_pct = current_status_aug2026 / 100
  ) %>%
  arrange(rgn_id, cert_year) %>%
  gt() %>%
  cols_label(
    rgn_id = "Region",
    cert_year = "Year",
    certified = "Certified Hotels (Aug 2026)",
    total_large_hotels = "Total Large Hotels (+100 employees)",
    current_status_aug2026_pct = "Coverage Aug 2026 (6 programs)",
    current_status_jul2026_pct = "Coverage Jul 2026 (3 programs)",
    change_pp = "Change (pp)"
  ) %>%
  fmt_percent(
    columns = c(current_status_aug2026_pct, current_status_jul2026_pct),
    decimals = 1
  ) %>%
  fmt_number(
    columns = change_pp,
    decimals = 1,
    pattern = "{x} pp"
  ) %>%
  sub_missing(
    columns = change_pp,
    missing_text = "—"
  ) %>%
  cols_hide(c(current_status_jul2026, current_status_aug2026)) %>%
  cols_align(
    align = "center",
    everything()
  ) %>%
  tab_header(
    title = "Sustainability Certification Coverage: August 2026 vs July 2026",
    subtitle = paste(
      "Aug 2026: Distintivo S, Green Key, Green Globe, EarthCheck,",
      "Preferred by Nature, Calidad Ambiental Turística.",
      "Jul 2026: Distintivo S, Green Key, and Green Globe only."
    )
  ) %>%
  opt_table_outline() %>%
  opt_all_caps()

sust_cert_comparison
```

```{r}
sust_cert_current_status_final <- large_hotels_denue_w_cert %>%
  dplyr::select(rgn_id, cert_year, current_status) %>%
  ungroup()

write_csv(
  sust_cert_current_status_final,
  file.path(
    output_sustainability,
    "sust_cert_current_status_final_2019_2026_100over_082726.csv"
  )
)

## historical comparison files from distintivo_s_only runs
sust_cert_cs_dist_s_only <- read_csv(
  file.path(output_sustainability, "dist_S_current_status_final_75over_010726.csv"),
  show_col_types = FALSE
)
sust_cert_cs_old <- read_csv(
  file.path(output_sustainability, "dist_S_current_status_final_121225.csv"),
  show_col_types = FALSE
)
```


```{r}
## plot the score over time
sust_cert_current_status_final <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","sust_cert_current_status_final_2019_2026_100over_082726.csv"))

final_score_plotly <- plotly::plot_ly(
  sust_cert_current_status_final, 
  x = ~cert_year, 
  y = ~current_status, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "TR Sustainability: certified hotels (all programs) / total large hotels",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(sust_cert_current_status_final$cert_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
final_score_plotly

large_hotels_denue_w_cert <- large_hotels_denue_w_cert %>% 
  arrange(cert_year) %>% 
  ungroup()

## plot the score over time
denue_large_hotels_plotly <- plotly::plot_ly(
  large_hotels_denue_w_cert, 
  x = ~cert_year, 
  y = ~total_large_hotels, 
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Total DENUE Large Hotels (>100 employees) in coastal areas of the GoC (50km)",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Number of hotels")
  ) 
denue_large_hotels_plotly

```
