---
title: 'OHI GoC 2025: TR Sustainability — Investment in potential tourism management'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor:
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options:
  chunk_output_type: console
---

# TR Sustainability: investment + infrastructure for tourism management

Combines **federal CONANP investment** (budgets + returned fees) with **SIMEC i-efectividad** (C2 + C3).

## Formula

```text
score_budget = mean(V003, V005, returned cobro de derechos)
               # national; V003/V005 vs 2012–2018 mean; capped at 100

score_anp    = mean(C2 Administrativo y financiero, C3 Usos y beneficios)
score_simec  = mean(score_anp) across ANPs in the region  # by SIMEC report year

IF region has SIMEC that year:
  score = mean(score_budget, score_simec)
ELSE:
  score = score_budget
```

| Layer | What it captures | Variation |
|-------|------------------|-----------|
| Budget + fees | National investment in ANP management | Mostly **by year** |
| SIMEC C2 + C3 | Admin/finance + visitor/use benefits | **By region**, and now **2020 and 2023** |

**C2:** presupuesto, personal, equipo, infraestructura de manejo, cobro de derechos.  
**C3:** beneficios económicos, producción sustentable, uso sustentable, infraestructura para visitantes.

**SIMEC years:** `simec_i_efectividad_anp_2026.csv` has assessments for **2020** and **2023** (2023 scores were previously mislabeled as 2021/2022).  
For each analysis year we use the **most recent SIMEC report on or before that year**:

| Analysis year | SIMEC report used |
|---------------|-------------------|
| 2019–2022 | 2020 |
| 2023–2025 | 2023 |

# 1. Setup

```{r}
library(tidyverse)
library(here)
library(plotly)
library(janitor)

region_levels <- paste0("Region_", 1:9)

rgn_labels <- c(
  "Region_1" = "Western Upper Gulf",
  "Region_2" = "Western Midriff Islands",
  "Region_3" = "Western Central Gulf",
  "Region_4" = "Western Southern Gulf",
  "Region_5" = "Eastern Upper Gulf",
  "Region_6" = "Eastern Midriff Islands",
  "Region_7" = "Southern Sonora Coast",
  "Region_8" = "Sinaloa Coast",
  "Region_9" = "Southeastern Gulf"
)

rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

analysis_years <- 2019:2025
budget_hist_years <- 2012:2018

output_sustainability <- here::here(
  "/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability"
)

simec_path <- here::here(
  "/home/shares/ohi/OHI_GOC/_raw_data/SIMEC",
  "simec_i_efectividad_anp_2026.csv"
)
```

# 2. Budget: V003, V005, returned fees

National only — same value assigned to every region. **S046 / PROCODES is not included.**

## 2.1 V003 & V005 (vs 2012–2018)

```{r}
fed_budget_raw <- read_csv(
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int/conanp_int",
    "fed_budget_programas_importantes_2012_2026.csv"
  ),
  show_col_types = FALSE
)

fed_budget_programs <- fed_budget_raw %>%
  select(year, v003_anp_management, v005_protection_restoration) %>%
  pivot_longer(
    cols = c(v003_anp_management, v005_protection_restoration),
    names_to = "programas",
    values_to = "mdp_2026"
  )

hist_avg_budget <- fed_budget_programs %>%
  filter(year %in% budget_hist_years) %>%
  group_by(programas) %>%
  summarise(hist_avg_mdp_2026 = mean(mdp_2026, na.rm = TRUE), .groups = "drop")

cs_budget_programs <- fed_budget_programs %>%
  left_join(hist_avg_budget, by = "programas") %>%
  mutate(
    current_status_raw = (mdp_2026 / hist_avg_mdp_2026) * 100,
    current_status = pmin(current_status_raw, 100)
  )

cs_budget_programs %>%
  filter(year %in% analysis_years) %>%
  arrange(programas, year)
```

```{r}
plot_ly(
  cs_budget_programs,
  x = ~year,
  y = ~current_status_raw,
  color = ~programas,
  type = "scatter",
  mode = "lines+markers"
) %>%
  layout(
    title = "CONANP V003 & V005 vs 2012–2018 mean (raw, uncapped)",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Percent of historical mean")
  ) %>%
  add_lines(
    x = range(cs_budget_programs$year, na.rm = TRUE),
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Goal (100)",
    inherit = FALSE
  )
```

## 2.2 Returned cobro de derechos

```{r}
cs_fees <- read_csv(
  here::here(
    "/home/shares/ohi/OHI_GOC/_raw_data/NOSSA_cuidar_lo_que_importa/d2026",
    "CONANP_fees_returned_2007_2026.csv"
  ),
  show_col_types = FALSE
) %>%
  clean_names() %>%
  select(year, pct_returned) %>%
  mutate(
    programas = "return_cobros_from_shcp",
    current_status_raw = pct_returned * 100,
    current_status = pmin(current_status_raw, 100)
  )

cs_fees %>% filter(year %in% analysis_years)
```

## 2.3 Mean of three parts → expand to all regions

```{r}
cs_budget_parts <- bind_rows(
  cs_budget_programs %>% select(year, programas, current_status),
  cs_fees %>% select(year, programas, current_status)
) %>%
  filter(year %in% analysis_years)

cs_budget <- cs_budget_parts %>%
  group_by(year) %>%
  summarise(
    score_budget = mean(current_status, na.rm = TRUE),
    n_budget_parts = n(),
    .groups = "drop"
  ) %>%
  crossing(rgn_id = factor(region_levels, levels = region_levels)) %>%
  relocate(rgn_id, year)

cs_budget
```

# 3. SIMEC i-efectividad: C2 + C3 (2020 and 2023)

**Region rules**

1. Drop non-GoC (`rgn_id == "NA"`).
2. Split multi-region strings (`"Region_1, Region_5"` → both).
3. Expand `"all"` (Islas del Golfo in **2023**) to all 9 regions.
4. Keep **2020** Islas del Golfo rows as already assigned (Region_2 / 4 / 8).

```{r}
simec_raw <- read_csv(simec_path, show_col_types = FALSE) %>%
  clean_names() %>%
  rename(
    report_year = ano_reporte,
    c2 = componente_administrativo_y_financiero,
    c3 = componente_de_usos_y_beneficios
  )

names(simec_raw)
count(simec_raw, report_year)
```

```{r}
simec_anp <- simec_raw %>%
  mutate(
    nombre = str_squish(nombre),
    rgn_id = str_squish(as.character(rgn_id)),
    score_anp = rowMeans(cbind(c2, c3), na.rm = TRUE)
  ) %>%
  filter(
    !is.na(c2), !is.na(c3),
    !is.na(rgn_id),
    !rgn_id %in% c("NA", "na")
  ) %>%
  ## "all" (Islas del Golfo, 2023) → every region
  mutate(
    rgn_id = if_else(
      str_to_lower(rgn_id) == "all",
      paste(region_levels, collapse = ", "),
      rgn_id
    )
  ) %>%
  separate_rows(rgn_id, sep = ",\\s*") %>%
  mutate(
    rgn_id = str_squish(rgn_id),
    rgn_id = factor(rgn_id, levels = region_levels)
  ) %>%
  filter(!is.na(rgn_id)) %>%
  select(
    report_year, nombre, rgn_id,
    c2, c3, score_anp, i_efectividad, vinculo_reporte
  )

simec_anp %>% count(report_year, rgn_id) %>% arrange(report_year, rgn_id)

simec_anp %>%
  filter(str_detect(nombre, regex("ISLAS DEL GOLFO", ignore_case = TRUE))) %>%
  arrange(report_year, rgn_id) %>%
  select(report_year, nombre, rgn_id, c2, c3, score_anp)
```

## 3.1 Regional means by SIMEC report year

```{r}
cs_simec_by_report <- simec_anp %>%
  group_by(report_year, rgn_id) %>%
  summarise(
    score_simec = mean(score_anp, na.rm = TRUE),
    score_c2 = mean(c2, na.rm = TRUE),
    score_c3 = mean(c3, na.rm = TRUE),
    n_anp = n_distinct(nombre),
    anp_names = paste(sort(unique(nombre)), collapse = "; "),
    .groups = "drop"
  )

cs_simec_by_report %>% arrange(rgn_id, report_year)
```

## 3.2 Attach SIMEC to each analysis year

```{r}
simec_lookup <- tibble(year = analysis_years) %>%
  mutate(report_year = if_else(year < 2023, 2020L, 2023L))

cs_simec <- crossing(
  year = analysis_years,
  rgn_id = factor(region_levels, levels = region_levels)
) %>%
  left_join(simec_lookup, by = "year") %>%
  left_join(cs_simec_by_report, by = c("report_year", "rgn_id"))

cs_simec %>%
  filter(year %in% c(2020, 2022, 2023, 2025)) %>%
  select(rgn_id, year, report_year, score_simec, n_anp) %>%
  arrange(rgn_id, year)
```

```{r}
plot_ly(
  cs_simec_by_report,
  x = ~rgn_id,
  y = ~score_simec,
  color = ~factor(report_year),
  type = "bar",
  text = ~paste0(n_anp, " ANPs<br>", anp_names),
  hoverinfo = "text+y"
) %>%
  layout(
    barmode = "group",
    title = "SIMEC mean(C2, C3) by region: 2020 vs 2023",
    xaxis = list(title = "Region"),
    yaxis = list(title = "Score (0–100)", range = c(0, 105)),
    legend = list(title = list(text = "Report year"))
  )
```

# 4. Combine budget + SIMEC

```{r}
cs_invest_mgmt <- cs_budget %>%
  left_join(
    cs_simec %>%
      select(rgn_id, year, report_year, score_simec, score_c2, score_c3, n_anp, anp_names),
    by = c("rgn_id", "year")
  ) %>%
  mutate(
    has_simec = !is.na(score_simec),
    score = if_else(
      has_simec,
      (score_budget + score_simec) / 2,
      score_budget
    ),
    sust_type = "conanp_fee_budget",
    rgn_label = rgn_labels[as.character(rgn_id)],
    method = if_else(
      has_simec,
      "mean(budget_V003_V005_fees, simec_C2_C3)",
      "budget_V003_V005_fees_only"
    )
  ) %>%
  arrange(rgn_id, year)

cs_invest_mgmt %>%
  select(
    rgn_id, year, report_year,
    score_budget, score_simec, has_simec, score, n_anp, method
  )
```

## Merge-ready table

Same shape as before for `tr_merge.qmd` / `tr_current_status.qmd`.

```{r}
cs_for_merge <- cs_invest_mgmt %>%
  transmute(
    rgn_id = factor(rgn_id, levels = region_levels),
    year,
    score,
    sust_type,
    score_budget,
    score_simec,
    simec_report_year = report_year,
    has_simec,
    method
  )

cs_for_merge
```

# 5. Plots

```{r}
plot_ly(
  cs_budget_parts,
  x = ~year,
  y = ~current_status,
  color = ~programas,
  type = "scatter",
  mode = "lines+markers"
) %>%
  layout(
    title = "Budget parts: V003, V005, returned cobro de derechos",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Score (0–100)", range = c(0, 105))
  ) %>%
  add_lines(
    x = range(analysis_years),
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Goal (100)",
    inherit = FALSE
  )
```

```{r}
## final score over time (SIMEC steps from 2020 → 2023 at analysis year 2023)
plot_ly(
  cs_invest_mgmt,
  x = ~year,
  y = ~score,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines+markers",
  text = ~paste0(
    rgn_label, "<br>",
    "Final: ", round(score, 1), "<br>",
    "Budget: ", round(score_budget, 1), "<br>",
    "SIMEC C2+C3 (", report_year, "): ",
    ifelse(has_simec, round(score_simec, 1), "NA — budget only")
  ),
  hoverinfo = "text+x"
) %>%
  layout(
    title = "Investment + infrastructure (tourism management)",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Score (0–100)", range = c(0, 105))
  ) %>%
  add_lines(
    x = range(analysis_years),
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Goal (100)",
    inherit = FALSE
  )
```

```{r}
## budget vs SIMEC vs final — 2020 report years vs 2023 report years
compare_df <- cs_invest_mgmt %>%
  filter(year %in% c(2020, 2023)) %>%
  select(rgn_id, year, score_budget, score_simec, score) %>%
  pivot_longer(
    cols = c(score_budget, score_simec, score),
    names_to = "layer",
    values_to = "value"
  ) %>%
  mutate(
    layer = recode(
      layer,
      score_budget = "Budget (V003+V005+fees)",
      score_simec = "SIMEC mean(C2, C3)",
      score = "Final combined"
    )
  )

compare_2020 <- compare_df %>% filter(year == 2020)
compare_2023 <- compare_df %>% filter(year == 2023)

subplot(
  plot_ly(
    compare_2020, x = ~rgn_id, y = ~value, color = ~layer,
    type = "bar", showlegend = FALSE
  ) %>% layout(barmode = "group", yaxis = list(title = "Score", range = c(0, 105))),
  plot_ly(
    compare_2023, x = ~rgn_id, y = ~value, color = ~layer,
    type = "bar"
  ) %>% layout(barmode = "group", yaxis = list(title = "", range = c(0, 105))),
  nrows = 1,
  shareY = TRUE,
  titleX = TRUE
) %>%
  layout(
    title = "Budget vs SIMEC vs final — 2020 (left) vs 2023 (right)",
    annotations = list(
      list(x = 0.22, y = 1.08, text = "2020 SIMEC report", showarrow = FALSE,
           xref = "paper", yref = "paper"),
      list(x = 0.78, y = 1.08, text = "2023 SIMEC report", showarrow = FALSE,
           xref = "paper", yref = "paper")
    )
  )
```

# 6. Save outputs

```{r}
write_csv(
  cs_budget_parts,
  file.path(output_sustainability, "invest_mgmt_budget_parts_v003_v005_fees_2019_2025.csv")
)
write_csv(
  simec_anp,
  file.path(output_sustainability, "invest_mgmt_simec_anp_c2_c3_goc_2020_2023.csv")
)
write_csv(
  cs_simec_by_report,
  file.path(output_sustainability, "invest_mgmt_simec_region_means_c2_c3_by_report_year.csv")
)
write_csv(
  cs_invest_mgmt,
  file.path(output_sustainability, "invest_mgmt_detail_budget_simec_2019_2025.csv")
)
write_csv(
  cs_for_merge,
  file.path(output_sustainability, "investment_tourism_management_final_092826.csv")
)
write_csv(
  cs_for_merge %>% select(rgn_id, year, score, sust_type),
  file.path(
    output_sustainability,
    "tr_sust_invest_mgmt_tourism_current_status_2019_2025.csv"
  )
)
```

## Notes

- **Formula:** `mean( mean(V003, V005, fees), mean(C2, C3) )` when SIMEC exists; else budget only.
- **SIMEC path:** `/home/shares/ohi/OHI_GOC/_raw_data/SIMEC/simec_i_efectividad_anp_2026.csv`
- **SIMEC timing:** 2020 scores → analysis years 2019–2022; 2023 scores → 2023–2025. Final scores step at 2023 when SIMEC updates; budget still moves every year.
- **Islas del Golfo:** 2020 keeps region-specific rows (R2 / R4 / R8); 2023 `"all"` expands to all 9 regions.
- **Alto Golfo:** `"Region_1, Region_5"` counted in both.
- **Non-GoC** (`NA`) dropped.
- **Merge:** use `investment_tourism_management_final_092826.csv`; `sust_type` stays `"conanp_fee_budget"`.
