---
title: 'OHI GoC 2025: TR Sustainability — Investment in potential tourism management'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# TR Sustainability: investment + infrastructure for tourism management

This script rebuilds the **Investment in potential tourism management** sustainability subcomponent as:

**investment** (federal CONANP budgets + returned fees) **+ infrastructure / management effectiveness** (SIMEC i-efectividad C2 & C3).

## Method

1. **Budget/fees (temporal variation):**  
   `score_budget = mean(V003, V005, returned cobro de derechos)`  
   Reference for V003/V005 = 2012–2018 mean (same as before). Scores capped at 100.

2. **SIMEC i-efectividad (spatial variation):**  
   For each GoC ANP, `score_anp = mean(Componente Administrativo y Financiero, Componente de Usos y Beneficios)`.  
   Then `score_simec = mean(score_anp)` across ANPs in that region.  
   Scores are held constant across years until SIMEC is updated (reports are ~2021–2022).

3. **Combine:**  
   - If the region has ≥1 SIMEC-scored GoC ANP:  
     `score = mean(score_budget, score_simec)`  
   - If not:  
     `score = score_budget` only

In theory, once SIMEC is updated regularly it could replace the budget/fee indicators. For now, budgets give year-to-year change; SIMEC gives cross-region structure.

**C2** (Administrativo y financiero) includes presupuesto, personal, equipo, infraestructura de manejo, and cobro de derechos.  
**C3** (Usos y beneficios) includes beneficios económicos, producción sustentable, uso sustentable, and infraestructura para visitantes.

# 1. Setup

```{r}
library(tidyverse)
library(here)
library(plotly)
library(janitor)

region_levels <- paste0("Region_", 1:9)

rgn_labels <- c(
  "Region_1" = "Western Upper Gulf",
  "Region_2" = "Western Midriff Islands",
  "Region_3" = "Western Central Gulf",
  "Region_4" = "Western Southern Gulf",
  "Region_5" = "Eastern Upper Gulf",
  "Region_6" = "Eastern Midriff Islands",
  "Region_7" = "Southern Sonora Coast",
  "Region_8" = "Sinaloa Coast",
  "Region_9" = "Southeastern Gulf"
)

rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

analysis_years <- 2019:2025
budget_hist_years <- 2012:2018

output_sustainability <- here::here(
  "/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability"
)

simec_dir <- here::here("/home/shares/ohi/OHI_GOC/_raw_data/SIMEC")
if (!dir.exists(simec_dir)) {
  simec_dir <- here::here("SIMEC")
}
```

# 2. Budget component: V003, V005, returned fees

## 2.1 Federal CONANP program budgets

```{r}
fed_budget_raw <- read_csv(
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int/conanp_int",
    "fed_budget_programas_importantes_2012_2026.csv"
  ),
  show_col_types = FALSE
)

## keep V003 (ANP management) and V005 (protection/restoration) only —
## not S046 / PROCODES, per the restructured formula
fed_budget_programs <- fed_budget_raw %>%
  select(
    year,
    v003_anp_management,
    v005_protection_restoration
  ) %>%
  pivot_longer(
    cols = c(v003_anp_management, v005_protection_restoration),
    names_to = "programas",
    values_to = "mdp_2026"
  )

w_hist_avg_fed_budget <- fed_budget_programs %>%
  filter(year %in% budget_hist_years) %>%
  group_by(programas) %>%
  summarise(hist_avg_mdp_2026 = mean(mdp_2026, na.rm = TRUE), .groups = "drop")

current_status_fed_budget <- fed_budget_programs %>%
  left_join(w_hist_avg_fed_budget, by = "programas") %>%
  mutate(
    current_status_raw = (mdp_2026 / hist_avg_mdp_2026) * 100,
    current_status = pmin(current_status_raw, 100)
  )

current_status_fed_budget %>%
  filter(year %in% analysis_years) %>%
  arrange(programas, year)
```

```{r}
fed_budget_plotly <- plot_ly(
  current_status_fed_budget,
  x = ~year,
  y = ~current_status_raw,
  color = ~programas,
  type = "scatter",
  mode = "lines+markers"
) %>%
  layout(
    title = "CURRENT STATUS: CONANP V003 & V005 (millones de pesos 2026)",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(current_status_fed_budget$year, na.rm = TRUE),
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
fed_budget_plotly
```

## 2.2 Returned cobro de derechos

```{r}
pct_returned_df <- read_csv(
  here::here(
    "/home/shares/ohi/OHI_GOC/_raw_data/NOSSA_cuidar_lo_que_importa/d2026",
    "CONANP_fees_returned_2007_2026.csv"
  ),
  show_col_types = FALSE
) %>%
  clean_names() %>%
  select(year, pct_returned)

current_status_fees <- pct_returned_df %>%
  mutate(
    programas = "return_cobros_from_shcp",
    current_status_raw = pct_returned * 100,
    current_status = pmin(current_status_raw, 100)
  )

current_status_fees %>%
  filter(year %in% analysis_years)
```

## 2.3 Average budget + fees (same national value → all regions)

```{r}
cs_budget_parts <- bind_rows(
  current_status_fed_budget %>%
    select(year, programas, current_status),
  current_status_fees %>%
    select(year, programas, current_status)
) %>%
  filter(year %in% analysis_years)

## national score: mean of V003, V005, returned fees
cs_budget_national <- cs_budget_parts %>%
  group_by(year) %>%
  summarise(
    score_budget = mean(current_status, na.rm = TRUE),
    n_budget_parts = n(),
    .groups = "drop"
  )

## expand to all 9 regions (budget has no regional variation)
cs_budget <- crossing(
  year = analysis_years,
  rgn_id = factor(region_levels, levels = region_levels)
) %>%
  left_join(cs_budget_national, by = "year")

cs_budget
```

# 3. SIMEC i-efectividad: C2 + C3

Read ANP-level scores with GoC region assignments from `identifying_lasting_special_places_SIMEC_w_rgns.csv`.

**Region assignment rules:**

- Drop non-GoC ANPs (`rgn_id == "NA"`).
- Split multi-region strings (e.g. `"Region_1, Region_5"` → both regions).
- **`Islas del Golfo de California` (`rgn_id == "all"`)** is expanded to **all 9 OHI regions** for now (Gulf-wide ANP).

```{r}
simec_raw <- read_csv(
  file.path(simec_dir, "identifying_lasting_special_places_SIMEC_w_rgns.csv"),
  show_col_types = FALSE
) %>%
  clean_names()

## clean_names turns components into snake_case; confirm names
names(simec_raw)

simec_anp <- simec_raw %>%
  rename(
    c2_admin_financiero = componente_administrativo_y_financiero,
    c3_usos_beneficios = componente_de_usos_y_beneficios
  ) %>%
  mutate(
    score_anp_c2_c3 = rowMeans(
      cbind(c2_admin_financiero, c3_usos_beneficios),
      na.rm = TRUE
    )
  ) %>%
  ## drop non-GoC only
  filter(!is.na(rgn_id), !rgn_id %in% c("NA", "na")) %>%
  ## Islas del Golfo ("all") → every OHI region
  mutate(
    rgn_id = if_else(
      str_to_lower(str_squish(rgn_id)) == "all",
      paste(region_levels, collapse = ", "),
      as.character(rgn_id)
    )
  ) %>%
  ## explode "Region_1, Region_5" and the expanded "all" string
  separate_rows(rgn_id, sep = ",\\s*") %>%
  mutate(
    rgn_id = str_squish(rgn_id),
    rgn_id = factor(rgn_id, levels = region_levels)
  ) %>%
  filter(!is.na(rgn_id))

simec_anp %>%
  select(
    ano_reporte, nombre, rgn_id,
    c2_admin_financiero, c3_usos_beneficios, score_anp_c2_c3, i_efectividad
  ) %>%
  arrange(rgn_id, nombre)
```

```{r}
## mean across ANPs within each region (one static score per region)
cs_simec <- simec_anp %>%
  group_by(rgn_id) %>%
  summarise(
    score_simec = mean(score_anp_c2_c3, na.rm = TRUE),
    score_c2 = mean(c2_admin_financiero, na.rm = TRUE),
    score_c3 = mean(c3_usos_beneficios, na.rm = TRUE),
    n_anp_simec = n_distinct(nombre),
    anp_names = paste(sort(unique(nombre)), collapse = "; "),
    .groups = "drop"
  )

cs_simec
```

```{r}
plot_ly(
  cs_simec,
  x = ~rgn_id,
  y = ~score_simec,
  type = "bar",
  marker = list(color = ~rgn_colors[as.character(rgn_id)]),
  text = ~paste0(n_anp_simec, " ANPs<br>"),
  hoverinfo = "text+y"
) %>%
  layout(
    title = "SIMEC mean(C2 Administrativo, C3 Usos y beneficios) by region",
    xaxis = list(title = "Region"),
    yaxis = list(title = "Score (0 - 100)", range = c(0, 105))
  )
```

# 4. Combine investment + infrastructure

```{r}
cs_invest_mgmt <- cs_budget %>%
  left_join(cs_simec, by = "rgn_id") %>%
  mutate(
    has_simec = !is.na(score_simec),
    ## regions with SIMEC: avg(budget, simec); without: budget only
    score = if_else(
      has_simec,
      rowMeans(cbind(score_budget, score_simec), na.rm = TRUE),
      score_budget
    ),
    sust_type = "conanp_fee_budget",
    rgn_label = rgn_labels[as.character(rgn_id)],
    method = if_else(
      has_simec,
      "mean(budget_V003_V005_fees, simec_C2_C3)",
      "budget_V003_V005_fees_only"
    )
  ) %>%
  arrange(rgn_id, year)

cs_invest_mgmt %>%
  select(
    rgn_id, year, score_budget, score_simec, has_simec,
    score, method, n_anp_simec
  )
```

## Merge-ready table

Same shape as the previous `investment_tourism_management_final_*.csv` so `tr_merge.qmd` / `tr_current_status.qmd` can keep using `score` + `sust_type = "conanp_fee_budget"`.

```{r}
cs_for_merge <- cs_invest_mgmt %>%
  transmute(
    rgn_id = factor(rgn_id, levels = region_levels),
    year,
    score,
    sust_type = "conanp_fee_budget",
    score_budget,
    score_simec,
    has_simec,
    method
  )

cs_for_merge
```

# 5. Plots

```{r}
## parts that go into the budget average
budget_parts_plotly <- plot_ly(
  cs_budget_parts,
  x = ~year,
  y = ~current_status,
  color = ~programas,
  type = "scatter",
  mode = "lines+markers"
) %>%
  layout(
    title = "Budget parts: V003, V005, returned cobro de derechos",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)", range = c(0, 105))
  ) %>%
  add_lines(
    x = range(analysis_years),
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
budget_parts_plotly
```

```{r}
invest_mgmt_plotly <- plot_ly(
  cs_invest_mgmt,
  x = ~year,
  y = ~score,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines+markers",
  text = ~paste0(
    rgn_label, "<br>",
    "Final: ", round(score, 1), "<br>",
    "Budget: ", round(score_budget, 1), "<br>",
    "SIMEC C2+C3: ", ifelse(has_simec, round(score_simec, 1), "NA (budget only)")
  ),
  hoverinfo = "text+x"
) %>%
  layout(
    title = "TR Sustainability: investment + infrastructure (tourism management)",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)", range = c(0, 105))
  ) %>%
  add_lines(
    x = range(analysis_years),
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
invest_mgmt_plotly
```

```{r}
## compare budget-only vs combined for a recent year
compare_yr <- max(analysis_years)
compare_df <- cs_invest_mgmt %>%
  filter(year == compare_yr) %>%
  select(rgn_id, score_budget, score_simec, score) %>%
  pivot_longer(
    cols = c(score_budget, score_simec, score),
    names_to = "layer",
    values_to = "value"
  ) %>%
  mutate(
    layer = recode(
      layer,
      score_budget = "Budget (V003+V005+fees)",
      score_simec = "SIMEC mean(C2, C3)",
      score = "Final combined"
    )
  )

plot_ly(
  compare_df,
  x = ~rgn_id,
  y = ~value,
  color = ~layer,
  type = "bar"
) %>%
  layout(
    barmode = "group",
    title = paste0("Budget vs SIMEC vs final — ", compare_yr),
    xaxis = list(title = "Region"),
    yaxis = list(title = "Score (0–100)", range = c(0, 105))
  )
```

# 6. Save outputs

```{r}
write_csv(
  cs_budget_parts,
  file.path(output_sustainability, "invest_mgmt_budget_parts_v003_v005_fees_2019_2025.csv")
)
write_csv(
  simec_anp,
  file.path(output_sustainability, "invest_mgmt_simec_anp_c2_c3_goc.csv")
)
write_csv(
  cs_simec,
  file.path(output_sustainability, "invest_mgmt_simec_region_means_c2_c3.csv")
)
write_csv(
  cs_invest_mgmt,
  file.path(output_sustainability, "invest_mgmt_detail_budget_simec_2019_2025.csv")
)

## drop-in replacement for the previous merge file
write_csv(
  cs_for_merge,
  file.path(output_sustainability, "investment_tourism_management_final_091526.csv")
)

## also write a clearly named merge file
write_csv(
  cs_for_merge %>% select(rgn_id, year, score, sust_type),
  file.path(
    output_sustainability,
    "tr_sust_invest_mgmt_tourism_current_status_2019_2025.csv"
  )
)
```

## Notes

- **Formula:** `avg( avg(V003, V005, returned fees), avg(C2, C3) )` when SIMEC exists; else budget only.
- **S046 / PROCODES** is no longer in this average (restructure uses V003 + V005 + fees).
- **SIMEC years:** ANP reports are 2021 or 2022; the regional mean is held constant across 2019–2025 so budgets still drive temporal change.
- **Islas del Golfo:** `rgn_id == "all"` is expanded to all 9 regions, so every region (including Region_7) has at least that SIMEC ANP in the mean.
- **Merge:** point `tr_merge.qmd` / `tr_current_status.qmd` at `investment_tourism_management_final_091526.csv` (or the `tr_sust_invest_mgmt_tourism_...` file). `sust_type` stays `"conanp_fee_budget"` so the overall Sustainability mean does not need renaming.
- **Future:** if SIMEC is updated annually, it can replace the budget/fee layer entirely.
