---
title: 'OHI GoC 2025: CCE Beach Cleanups'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# Sense of Pertenencia: Coastal Community Engagement (CCE)

## Ama tu Playa and Adopta una Playa

Ama tu Playa (also branded as **Adopta una Playa**) is the beach adoption arm of Mexico's *Estrategia Nacional de Limpieza y Conservación de Playas y Costas de México 2025-2030*. It is led with SEMARNAT and partners across federal, state, and municipal government, civil society, and private organizations.

"UN COMPROMISO CON EL FUTURO DE NUESTROS OCÉANOS La Estrategia Nacional de Limpieza y Conservación de Playas y Costas de México 2025-2030 es una iniciativa ambiciosa del Gobierno de México para erradicar la contaminación por plásticos del 100% de playas y costas."

The national goal is ambitious: **100% of beaches and coasts free of plastic waste by 2030**. The program frames coastal plastic pollution as both an ecological and community problem. An estimated 60 to 90% of beach waste is plastics and microplastics. Organizations are asked to "adopt" a beach or coastal stretch and keep it clean through 2030.

Adopting a beach typically means:

- Removing waste from the adopted beach or coast
- Separating and disposing of collected waste appropriately
- Helping eliminate single use plastics
- Reducing waste generation more broadly
- Reporting cleanup activities regularly (the live platform also tracks coverage and tons collected per beach)

How to join: email <adoptatuplaya@semarnat.gob.mx> with the organization name, beach(es) of interest, and contact details. Catalog, map, and beach reports live at <https://www.adoptaunaplaya.com/playas/> (see also <https://www.adoptaunaplaya.com/playas/sumate/>).

For OHI GoC CCE, I am not scoring "tons collected" directly. More trash removed can mean either dirtier beaches or stronger cleanup effort. Instead, I use the **share of program inventory beaches that have already been adopted** in each OHI region. That is progress toward the program's own adoption and stewardship target.

Important caveat: the SEMARNAT catalog is biased toward monitored or listed beaches (often COFEPRIS, Blue Flag, or campaign sites). A region with **no beaches in the inventory** is not the same as a region that is plastic free. Those coasts may simply be less populated or not yet listed. I therefore do *not* assign a perfect score of 100 when a region is missing from the inventory.

```{r}
library(tidyverse)
library(mapview)
library(sf)
library(here)
library(janitor)
```

### Read in beach adoption data

```{r}
library(textclean)

## bring in data tabulated using `Tabula` from a pdf into a csv
adopted_beaches_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/ama_tu_playa/ama_tu_playa_040126","tabula-adoptaunaplaya_amatuplaya_lista.csv")) %>% 
  janitor::clean_names()
```

Note that the pdf itself has duplicate entries: for example, Playa El Anclote in Nayarit.  So we will clean that up.

```{r}

unique_beaches <- adopted_beaches_raw %>% 
  dplyr::select(-adopcion) %>% 
  distinct()

unique_adopted_beaches <- adopted_beaches_raw %>% 
  filter(adopcion == 1) %>% 
  distinct()

adopted_beaches_raw_nodups <- left_join(unique_beaches, unique_adopted_beaches)

```



### Restrict to GoC

```{r}
adopted_beaches_goc <- adopted_beaches_raw_nodups %>% 
  filter(entidad %in% c("Baja California","Baja California Sur","Jalisco","Nayarit","Sinaloa","Sonora")) %>% 
  mutate(municipio_clean = textclean::replace_non_ascii(municipio),
         municipio_clean = str_to_lower(municipio_clean)) %>% 
  mutate(long = str_replace_all(x, ",","."),
         lat = str_replace_all(y, ",",".")) %>% 
  st_as_sf(coords = c("long", "lat"), # making sf object
           crs = 4326, # WGS84
           remove = FALSE) 

mapview(adopted_beaches_goc)
```


### Assign to rgn_id

First, group by region and unite the polygons for inland 10km with the marine polygons.
```{r}
## intersect with inland_10:
inland_10 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_10km_buffer","inland_10km_goc.shp"))
regions_marine <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp")) %>%
  filter(location == "marine")

st_crs(inland_10) == st_crs(regions_marine) # TRUE

marine_and_inland10 <- bind_rows(regions_marine, inland_10) %>% 
  dplyr::select(rgn_id, geometry) %>% 
  group_by(rgn_id) %>% 
  summarize(
    geometry = st_union(geometry)
  )

mapview(adopted_beaches_goc) + mapview(marine_and_inland10)
```

Now, we can intersect with the `marine_and_inland10` so that each beach is assigned an OHI GoC Region ID.
```{r}
# intersect with GoC
adopted_beaches_goc_rgnid <- st_intersection(st_transform(adopted_beaches_goc, st_crs(marine_and_inland10)), marine_and_inland10) %>% 
  dplyr::select(playa, adopcion, organismo, municipio_clean, entidad, poblacion, long, lat, rgn_id, geometry) %>% 
  mutate(adopcion = ifelse(is.na(adopcion), "no","yes")) 

all_yes <- adopted_beaches_goc_rgnid %>% 
  filter(adopcion %in% "yes")
mapview(all_yes)

mapview(adopted_beaches_goc_rgnid, zcol = "adopcion")
```


### Current status

Now that we have a list of all the beaches that SEMARNAT aims to be adopted by 2030, we can determine, as of 04.01.26, how many beaches have been adopted. That is inherently representative of how close each GoC region is to the program goal of having its listed beaches adopted.

The Ama tu Playa / Adopta una Playa campaign is part of the 2025-2030 national strategy, so prior to 2025 none of these beaches were adopted under this program.

I will expand the snapshot into a 2019-2025 series: years 2019-2024 get `current_status = 0`, and 2025 gets the observed adoption rate.

Region 2 has no beaches in the program inventory after the spatial join, so I add an explicit row with `current_status = NA` and `reason = "no beaches in program inventory"` rather than treating that gap as a score of 0 or 100.

```{r}
ama_tu_playa_progress_2025 <- adopted_beaches_goc_rgnid %>% 
  st_drop_geometry() %>% 
  group_by(rgn_id) %>% 
  summarize(
    adopted_beaches = sum(adopcion == "yes"),
    total_beaches = n(),
    pct_adopted = adopted_beaches / total_beaches,
  ) %>% 
  ungroup() %>% 
  mutate(
    current_status = pct_adopted * 100,
    reason = NA_character_
  )

# explicit row for regions with no program inventory beaches
ama_tu_playa_missing <- tibble(
  rgn_id = "Region_2",
  adopted_beaches = NA_real_,
  total_beaches = 0,
  pct_adopted = NA_real_,
  current_status = NA_real_, # ama_status stays na; do not invent 0 or 100
  reason = "no beaches in program inventory"
)

ama_tu_playa_progress_2025 <- bind_rows(ama_tu_playa_progress_2025, ama_tu_playa_missing) %>% 
  arrange(rgn_id)

# expand to 2019-2025: no program adoptions before 2025
ama_tu_playa_progress <- tidyr::crossing(
  rgn_id = ama_tu_playa_progress_2025$rgn_id,
  year = 2019:2025
) %>%
  left_join(ama_tu_playa_progress_2025, by = "rgn_id") %>%
  mutate(
    adopted_beaches = case_when(
      !is.na(reason) ~ NA_real_, # region_2 stays missing
      year < 2025 ~ 0,
      TRUE ~ adopted_beaches
    ),
    pct_adopted = case_when(
      !is.na(reason) ~ NA_real_,
      year < 2025 ~ 0,
      TRUE ~ pct_adopted
    ),
    current_status = case_when(
      !is.na(reason) ~ NA_real_,
      year < 2025 ~ 0,
      TRUE ~ current_status
    )
  ) %>%
  arrange(rgn_id, year)

ama_tu_playa_progress
```


#### Write to Mazu

```{r}
write_csv(ama_tu_playa_progress, here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026/output","current_status_ama_tu_playa_progress_072326.csv"))

ama_tu_playa_progress <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026/output","current_status_ama_tu_playa_progress_040126.csv"))
ama_tu_playa_progress_new <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026/output","current_status_ama_tu_playa_progress_072326.csv"))
```


















































