---
title: "Untitled"
output: html_document
date: "2025-10-01"
---
Identifying the location of CONAPESCA fishing locations.

Exploring Juan Carlos fishing data

https://github.com/mex-Fisheries

office Names?
https://github.com/mex-fisheries/mex_ports/blob/main/data/clean/mex_ports.csv
also: mex_vessel_registry

```{r setup, include=FALSE}
library(tidyverse)
library(tidygeocoder)
library(dplyr)
library(stringr)
library(readr)
library(here)
library(sf)
library(mapview)
library(osmdata)
library(purrr)
library(stringi)
library(fuzzyjoin)


jc_data <- "/home/shares/ohi/OHI_GOC/_raw_data/CONAPESCA/d2025/SharedByJuanCarlos/mex_landings/data/clean"

list.files(jc_data)

sites <- read_csv(here("FP/FIS/v2025/data/conapesca_landing_sites_baja_sonora.csv"))
fish <- readRDS(file.path(jc_data, "mex_landings_2000_present.rds")) 

```


```{r}

baja_sites <- filter(sites, State=="Baja California") 

## Take care of regions with the same name (very rude!)
## two of these regions, one on pacific and one on gulf, but the gulf site is very small
## will just assume landings are for the pacific side
filter(baja_sites, landing_site == "SAN CARLOS PUNTA")

baja_sites <- baja_sites %>%
filter(!(landing_site =="SAN CARLOS PUNTA" & GofC == "yes"))
## average the other duplicates which are close to each other
baja_sites <- baja_sites %>%
  group_by(landing_site) %>%
  summarize(clave=mean(clave), 
            subregion= mean(subregion),
            long = mean(long),
            lat = mean(lat),
            State = unique(State))
write_csv(baja_sites, here("FP/FIS/v2025/data/fishing_site_locs_raw.csv"))

## Add some more sites that were not in the site data but were in the fisheries data.
tmp25 <- data.frame(landing_site = "FARO DE SAN JOSE", clave = NA, subregion = 1, long = -109.665, lat = 23.065, State = "Baja California")
tmp26 <- data.frame(landing_site = "LAGUNA MACUILE", clave = NA, subregion = 1, long = -109.665, lat = 23.065, State = "Baja California")
tmp27 <- data.frame(landing_site = "LOS MORROS", clave = NA, subregion = 1, long = -114.9790, lat = 29.3823, State = "Baja California")

baja_sites <- rbind(baja_sites, tmp21, tmp22, tmp23, tmp24, tmp25, tmp26, tmp27)


baja_fish <- filter(fish, state %in% c("BAJA CALIFORNIA")) 

sort(setdiff(baja_fish$landing_site, baja_sites$landing_site))
sort(setdiff(baja_sites$landing_site, baja_fish$landing_site))
intersect(baja_sites$landing_site, baja_fish$landing_site)
                      
                      
# Words that don't help matching
# ---- stoplist you can tweak ----
stop_terms <- c("PUERTO","PUNTA","ISLOTE","BAHIA","BAHÍA","LITORAL",
                "CALA","CALETA","CAMPO","PLAYA","RANCHO",
                "DEL","DE","LA","LAS","LOS","EL")

# ---- helpers (compatible with older R) ----
norm_tokens <- function(x) {
  x <- str_to_upper(x)
  x <- stringi::stri_trans_general(x, "Latin-ASCII")
  x <- str_replace_all(x, "[^A-Z0-9 ]", " ")
  x <- str_squish(x)
  str_split(x, "\\s+")
}

normalize_basic <- function(x) {
  toks <- norm_tokens(x)
  vapply(toks, function(v) paste(sort(unique(v)), collapse = " "),
         FUN.VALUE = character(1))
}

normalize_stopless <- function(x) {
  toks <- norm_tokens(x)
  toks <- lapply(toks, function(v) v[!v %in% stop_terms])
  vapply(toks, function(v) paste(sort(unique(v)), collapse = " "),
         FUN.VALUE = character(1))
}

# ---- build working copies ----
sites_k <- baja_sites %>%
  mutate(
    name_basic   = normalize_basic(landing_site),
    name_stopless = normalize_stopless(landing_site)
  ) %>%
  select(landing_site_sites = landing_site,
         name_basic, name_stopless, clave, subregion, long, lat)

fish_k <- baja_fish %>%
  mutate(
    name_basic   = normalize_basic(landing_site),
    name_stopless = normalize_stopless(landing_site)
  )

# Distinct fish names to map once
fish_names <- fish_k %>% distinct(landing_site, name_basic, name_stopless)

# ---- 1) EXACT RAW ----
# N=55
map1 <- fish_names %>%
  inner_join(sites_k %>% select(landing_site_sites, clave, subregion, long, lat),
             by = c("landing_site" = "landing_site_sites")) %>%
  transmute(landing_site,
            landing_site_sites = landing_site,
            clave, subregion, long, lat,
            dist = 0, match_type = "exact_raw")

todo1 <- anti_join(fish_names, map1, by = "landing_site")

# ---- 2) EXACT BASIC (no stop removal) ----
# N=104, do a quick check to make sure they are sensible!
map2 <- todo1 %>%
  inner_join(sites_k %>% select(name_basic, landing_site_sites, clave, subregion, long, lat),
             by = "name_basic") %>%
  transmute(landing_site,
            landing_site_sites, clave, subregion, long, lat,
            dist = 0, match_type = "exact_basic")

todo2 <- anti_join(todo1, map2 %>% select(landing_site), by = "landing_site")

# ---- 3) EXACT STOPLESS (with stop removal) ----
# N=24
map3 <- todo2 %>%
  inner_join(sites_k %>% select(name_stopless, landing_site_sites, clave, subregion, long, lat),
             by = "name_stopless") %>%
  transmute(landing_site,
            landing_site_sites, clave, subregion, long, lat,
            dist = 0, match_type = "exact_stopless") %>%
  filter(!is.na(landing_site))

todo3 <- anti_join(todo2, map3 %>% select(landing_site), by = "landing_site")



# ---- 4) FUZZY on STOPLESS ----

## N = 41: Some questionable ones, but mostly seem reasonable.
map4 <- if (nrow(todo3)) {
  stringdist_inner_join(
    todo3,
    sites_k %>% select(name_stopless, landing_site_sites, clave, subregion, long, lat),
    by = c("name_stopless" = "name_stopless"),
    method = "jw", max_dist = 0.15, distance_col = "dist"
  ) %>%
    group_by(landing_site) %>%
    slice_min(dist, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    transmute(landing_site,
              landing_site_sites, clave, subregion, long, lat,
              match_type = "fuzzy_stopless", dist)
} else dplyr::tibble()

# ---- combine + attach back ----
map_all <- bind_rows(map1, map2, map3, map4)

fish_matched <- fish_k %>%
  left_join(map_all, by = "landing_site")


fish_matched %>%
  filter(!is.na(lat)) %>%
  summarize(sum(live_weight))
fish_matched %>%
  filter(is.na(lat)) %>%
  summarize(sum(live_weight))
252954428/3520818879

missing_sites <- fish_matched %>%
  filter(is.na(lat)) %>%
  select(landing_site) %>%
  unique()

# ---- quick QA ----
fish_matched %>% count(match_type, matched = !is.na(lat))
fish_matched %>%
  filter(match_type == "fuzzy_stopless") %>%
  arrange(desc(dist)) %>%
  select(landing_site, landing_site_sites, dist) %>%
  head(20)



#, "BAJA CALIFORNIA SUR", " NAYARIT", "JALISCO", "SINALOA", "SONORA"))



tmp <- readRDS(file.path(jc_data, "mex_conapesca_apertura_2018_present.rds")) %>%
  filter(state %in% c("BAJA CALIFORNIA", "BAJA CALIFORNIA SUR", " NAYARIT", "JALISCO", "SINALOA", "SONORA")) %>%
  filter(fishing_zone_type != "AGUAS CONTINENTALES")

fishing_sites <- tmp %>%
  select(state, fishing_site_name) %>%
    mutate(fishing_site_clean = iconv(fishing_site_name, from = "latin1", to = "UTF-8")) %>%
    mutate(fishing_site_clean = str_squish(fishing_site_clean),
         fishing_site_clean = str_to_title(fishing_site_clean)) %>% # keep nice casing
  unique()

write_csv(fishing_sites, here("FIS/v2025/data/fishing_site_locs_raw.csv"))

## There is not always a match between the state and fishing location, but should be general correspondence
baja <- fishing_sites %>%
  filter(state %in% c("BAJA CALIFORNIA"))

```

```{r}
state_sites <- baja
# geocode via OSM (Nominatim). Add ", Mexico" to focus results.
geo1 <- state_sites %>%
  mutate(address = paste(fishing_site_clean, ", Mexico")) %>%
  geocode(address = address,
          method  = "osm",
          limit   = 1,
          full_results = TRUE)  # keep extras for QC

pts <- geo1 %>%
  filter(!is.na(lat)) %>%
  select(state, fishing_site_name, lat, long, osm_type, osm_id)

pts <- st_as_sf(pts, coords = c("long", "lat"),
                crs = 4326, remove = FALSE)

mapview::mapview(pts)



# OSM
# your cleaned names vector
nm <- state_sites$fishing_site_clean

# (1) escape regex specials so names like "San Isidro (Ejido …)" don't break the regex
escape_regex <- function(x) str_replace_all(x, "([\\W])", "\\\\\\1")

# (2) split into manageable batches
make_batches <- function(x, size = 30) split(x, ceiling(seq_along(x)/size))
batches <- make_batches(nm, size = 30)

# (3) helper to build one query for a batch
build_query <- function(names_batch, bbox = "Mexico", timeout = 180,
                        restrict = TRUE) {
  # optional: restrict by likely feature keys to speed up
  # (turn off with restrict = FALSE if you want all features)
  name_pat <- paste0("^(", paste(escape_regex(names_batch), collapse="|"), ")$")
  q <- opq(bbox = bbox, timeout = timeout) |>
    add_osm_feature(key = "name", value = name_pat, value_exact = FALSE, match_case = FALSE)
  if (restrict) {
    # Add a few typical keys to cut search space; each additional add_osm_feature
    # narrows the result, so we need OR logic. We'll just run separate queries per key
    # for speed in the outer function instead of stacking here.
    return(list(q = q, name_pat = name_pat))
  }
  list(q = q, name_pat = name_pat)
}

# (4) fetch one batch with retry + optional key-type splitting (faster)
fetch_batch <- function(names_batch, bbox = "Mexico", timeout = 180, pause = 2) {
  qinfo <- build_query(names_batch, bbox = bbox, timeout = timeout)
  name_pat <- qinfo$name_pat

  # try a few narrower key-based passes to reduce load (fast in Mexico):
  key_passes <- list(
    list(key="place", values=c("city","town","village","hamlet")),
    list(key="natural", values=c("beach","cape","coastline")),
    list(key="harbour", values="yes"),
    list(key="man_made", values=c("pier","breakwater","lighthouse","dam")),
    list(key="waterway", values=c("dam")),
    list(key="amenity", values=c("ferry_terminal"))
  )

  res_list <- list()

  # helper: run a single overpass call with retry
  run_q <- function(q) {
    tries <- 3
    for (i in 1:tries) {
      out <- try(osmdata_sf(q), silent = TRUE)
      if (!inherits(out, "try-error")) return(out)
      Sys.sleep(pause * i)  # backoff
    }
    NULL
  }

  # First: run several narrower queries (OR across them)
  for (kp in key_passes) {
    q <- opq(bbox = bbox, timeout = timeout) |>
      add_osm_feature(key = "name", value = name_pat, match_case = FALSE) |>
      add_osm_feature(key = kp$key, value = kp$values)
    out <- run_q(q)
    if (!is.null(out)) res_list <- append(res_list, list(out))
  }

  # Fallback: a general name-only query for any leftovers
  remaining <- names_batch
  if (length(res_list)) {
    found_names <- res_list |>
      map(~ bind_rows(.$osm_points, .$osm_polygons, .$osm_multipolygons, .$osm_lines)) |>
      bind_rows() |>
      as_tibble() |>
      distinct(name) |>
      pull(name) |>
      unique() |>
      str_to_upper()
    remaining <- setdiff(str_to_upper(names_batch), found_names)
  }

  if (length(remaining)) {
    q_rem <- opq(bbox = bbox, timeout = timeout) |>
      add_osm_feature(key = "name",
                      value = paste0("^(", paste(escape_regex(remaining), collapse="|"), ")$"),
                      match_case = FALSE)
    out_rem <- run_q(q_rem)
    if (!is.null(out_rem)) res_list <- append(res_list, list(out_rem))
  }

  if (!length(res_list)) return(NULL)

  # combine geometries and create a representative point for non-point features
  comb <- res_list |>
    map(~ bind_rows(.$osm_points, .$osm_polygons, .$osm_multipolygons, .$osm_lines)) |>
    purrr::compact() |>
    bind_rows()

  if (!nrow(comb)) return(NULL)

  sf_all <- st_as_sf(comb) |>
    mutate(name_u = str_to_upper(name)) |>
    group_by(name_u) |>
    summarize(geom = st_point_on_surface(st_combine(geometry)), .groups="drop") |>
    st_set_crs(4326)

  sf_all
}

# (5) run all batches and bind

tmp <- fetch_batch(batches[[1]], bbox = "Mexico", timeout = 180, pause = 3)

res_sf <- batches |>
  purrr::map(fetch_batch, bbox = "Mexico", timeout = 180, pause = 3) |>
  purrr::compact() |>
  (\(lst) if (length(lst)) do.call(rbind, lst) else NULL)()

# (6) join coordinates back to your table (left join by uppercase cleaned name)
coords <- res_sf |>
  mutate(lat = st_coordinates(geom)[,2],
         long = st_coordinates(geom)[,1]) |>
  st_drop_geometry() |>
  select(name_u, lat, long)

final <- state_sites %>%
  mutate(name_u = str_to_upper(fishing_site_clean)) %>%
  left_join(coords, by = "name_u")
