---
title: "Untitled"
output: html_document
date: "2025-10-01"
---

STEP 1: Ran through this script for Baja California and Baja California Sur with this data: conapesca_landing_sites_baja_sonora.csv (digitized pdf of fishing sites) and the fishing data from Juan Carlos: https://github.com/mex-Fisheries

STEP 2: Identified which of the fishing data sites could be 
identified based on different levels of matching. Then hand 
added the ones that couldn't be matched (landing_sites_BC and landing_sites_BCS). These are the files that should be corrected for errors/additions.
STEP 3: Rerun the code to attach the locations to the fishing data.




```{r setup, include=FALSE}
library(tidyverse)
library(tidygeocoder)
library(dplyr)
library(stringr)
library(readr)
library(here)
library(sf)
library(mapview)
library(osmdata)
library(purrr)
library(stringi)
library(fuzzyjoin)


jc_data <- "/home/shares/ohi/OHI_GOC/_raw_data/CONAPESCA/d2025/SharedByJuanCarlos/mex_landings/data/clean"

list.files(jc_data)

```

# 1 : Do Baja California

```{r}
## started with this file, but then added some regions that were missing from the 
## catch data
#sites <- read_csv(here("FP/FIS/v2025/data/conapesca_landing_sites_baja_sonora.csv"))
# baja_sites <- filter(sites, State=="Baja California") 
# 
# ## Take care of regions with the same name (very rude!)
# ## two of these regions, one on pacific and one on gulf, but the gulf site is very small
# ## will just assume landings are for the pacific side
# filter(baja_sites, landing_site == "SAN CARLOS PUNTA")
# 
# baja_sites <- baja_sites %>%
# filter(!(landing_site =="SAN CARLOS PUNTA" & GofC == "yes"))
# ## average the other duplicates which are close to each other
# baja_sites <- baja_sites %>%
#   group_by(landing_site) %>%
#   summarize(clave=mean(clave), 
#             subregion= mean(subregion),
#             long = mean(long),
#             lat = mean(lat),
#             State = unique(State))
# write_csv(baja_sites, here("FP/FIS/v2025/data/fishing_site_locs_raw_BC.csv"))


baja_sites <- read_csv(here("FP/FIS/v2025/data/landing_sites_BC.csv")) %>%
  mutate(OHI_rgn = ifelse(lat>= 30.96716, "Baja_California_North", "Baja_California_South"))

fish <- readRDS(file.path(jc_data, "mex_landings_2000_present.rds")) 
baja_fish <- filter(fish, state %in% c("BAJA CALIFORNIA")) 


```


```{r}


# Words that don't help matching
# ---- stoplist you can tweak ----
stop_terms <- c("PUERTO","PUNTA","ISLOTE","BAHIA","BAHÍA","LITORAL",
                "CALA","CALETA","CAMPO","PLAYA","RANCHO",
                "DEL","DE","LA","LAS","LOS","EL")

# ---- helpers (compatible with older R) ----
norm_tokens <- function(x) {
  x <- str_to_upper(x)
  x <- stringi::stri_trans_general(x, "Latin-ASCII")
  x <- str_replace_all(x, "[^A-Z0-9 ]", " ")
  x <- str_squish(x)
  str_split(x, "\\s+")
}

normalize_basic <- function(x) {
  toks <- norm_tokens(x)
  vapply(toks, function(v) paste(sort(unique(v)), collapse = " "),
         FUN.VALUE = character(1))
}

normalize_stopless <- function(x) {
  toks <- norm_tokens(x)
  toks <- lapply(toks, function(v) v[!v %in% stop_terms])
  vapply(toks, function(v) paste(sort(unique(v)), collapse = " "),
         FUN.VALUE = character(1))
}

# ---- build working copies ----
sites_k <- baja_sites %>%
  mutate(
    name_basic   = normalize_basic(landing_site),
    name_stopless = normalize_stopless(landing_site)
  ) %>%
  select(landing_site_sites = landing_site,
         name_basic, name_stopless, clave, subregion, long, lat, OHI_rgn)

fish_k <- baja_fish %>%
  mutate(
    name_basic   = normalize_basic(landing_site),
    name_stopless = normalize_stopless(landing_site)
  )

# Distinct fish names to map once
fish_names <- fish_k %>% distinct(landing_site, name_basic, name_stopless)

# ---- 1) EXACT RAW ----
# N=187
map1 <- fish_names %>%
  inner_join(sites_k %>% select(landing_site_sites, clave, subregion, OHI_rgn, long, lat),
             by = c("landing_site" = "landing_site_sites")) %>%
  transmute(landing_site,
            landing_site_sites = landing_site,
            clave, subregion, OHI_rgn, long, lat, 
            dist = 0, match_type = "exact_raw")

todo1 <- anti_join(fish_names, map1, by = "landing_site")

# ---- 2) EXACT BASIC (no stop removal) ----
# N=105, do a quick check to make sure they are sensible!
map2 <- todo1 %>%
  inner_join(sites_k %>% select(name_basic, landing_site_sites, clave, subregion, OHI_rgn, long, lat),
             by = "name_basic") %>%
  transmute(landing_site,
            landing_site_sites, clave, subregion, OHI_rgn, long, lat,
            dist = 0, match_type = "exact_basic")

todo2 <- anti_join(todo1, map2 %>% select(landing_site), by = "landing_site")

# ---- 3) EXACT STOPLESS (with stop removal) ----
# N=24
map3 <- todo2 %>%
  inner_join(sites_k %>% select(name_stopless, landing_site_sites, clave, subregion, OHI_rgn, long, lat),
             by = "name_stopless") %>%
  transmute(landing_site,
            landing_site_sites, clave, subregion, OHI_rgn, long, lat,
            dist = 0, match_type = "exact_stopless") %>%
  filter(!is.na(landing_site))

todo3 <- anti_join(todo2, map3 %>% select(landing_site), by = "landing_site")



# ---- 4) FUZZY on STOPLESS ----

## N = 35: Some questionable ones, but mostly seem reasonable.
map4 <- if (nrow(todo3)) {
  stringdist_inner_join(
    todo3,
    sites_k %>% select(name_stopless, landing_site_sites, clave, subregion, OHI_rgn, long, lat),
    by = c("name_stopless" = "name_stopless"),
    method = "jw", max_dist = 0.15, distance_col = "dist"
  ) %>%
    group_by(landing_site) %>%
    slice_min(dist, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    transmute(landing_site,
              landing_site_sites, clave, subregion, OHI_rgn, long, lat,
              match_type = "fuzzy_stopless", dist)
} else dplyr::tibble()


# ---- combine + attach back ----
map_all <- bind_rows(map1, map2, map3, map4)

fish_matched <- fish_k %>%
  left_join(map_all, by = "landing_site") %>%
  mutate(GulfCal = ifelse(subregion<4, "no", "yes"))
table(fish_matched$GulfCal)
175849/(324083+175849) 


# ---- quick QA ----

# these should be the same:
sum(fish_matched$live_weight)
sum(baja_fish$live_weight)

fish_matched %>%
  filter(!is.na(lat)) %>%
  summarize(sum(live_weight))
fish_matched %>%
  filter(is.na(lat)) %>%
  summarize(sum(live_weight))
142644549/(3631128240+142644549) #4% of catch couldn't find a match for.

fish_matched %>%
  filter(!is.na(lat) & year == 2024) %>%
  summarize(sum(live_weight))
fish_matched %>%
  filter(is.na(lat) & year == 2024) %>%
  summarize(sum(live_weight))

## for 2024 only 0.5% not accounted for. Most of mismatches are in previous years
1664744/(1664744+353647569)
missing_sites <- fish_matched %>%
  filter(is.na(lat)) %>%
  select(state, office_name, landing_site, year, landing_site_sites)
summary(missing_sites)
filter(missing_sites, is.na(landing_site_sites)) ## all sites were accounted for in the list, but
# I couldn't find the location unambiguously of all sites.
table(missing_sites$year)
table(baja_fish$year)
## nearly all the missmatches are from 2005 and earlier
bc_data <- fish_matched %>%
  select(-name_basic, -name_stopless, -clave, -dist, -match_type)
write_csv(bc_data, file.path(jc_data, "mex_BC_landings_2000_present_with_lat_long.csv"))



```



# 2 : Do Baja California SUR
```{r}
## started with this file, but then added some regions that were missing from the 
## catch data
#sites <- read_csv(here("FP/FIS/v2025/data/conapesca_landing_sites_baja_sonora.csv")) 
#bajasur_sites <- filter(sites, State=="Baja California Sur") 
#bajasur_sites <- filter(bajasur_sites, landing_site != "AGUAS LITORALES DEL SUR")

bajasur_sites <- read_csv(here("FP/FIS/v2025/data/landing_sites_BCS.csv")) %>%
    mutate(OHI_rgn = ifelse(lat>= 26.34775, "Baja_California_Sur_North", "Baja_California_Sur_South"))
sum(duplicated(bajasur_sites))



fish <- readRDS(file.path(jc_data, "mex_landings_2000_present.rds")) 
bajasur_fish <- filter(fish, state %in% c("BAJA CALIFORNIA SUR")) %>%
  filter(!is.na(landing_site))

```


```{r}

# ---- build working copies ----
sites_k <- bajasur_sites %>%
  mutate(
    name_basic   = normalize_basic(landing_site),
    name_stopless = normalize_stopless(landing_site)
  ) %>%
  select(landing_site_sites = landing_site,
         name_basic, name_stopless, clave, subregion, OHI_rgn, long, lat)

fish_k <- bajasur_fish %>%
  mutate(
    name_basic   = normalize_basic(landing_site),
    name_stopless = normalize_stopless(landing_site)
  )

# Distinct fish names to map once
fish_names <- fish_k %>% distinct(landing_site, name_basic, name_stopless)

# ---- 1) EXACT RAW ----
# N=233
map1 <- fish_names %>%
  inner_join(sites_k %>% select(landing_site_sites, clave, subregion, OHI_rgn, long, lat),
             by = c("landing_site" = "landing_site_sites")) %>%
  transmute(landing_site,
            landing_site_sites = landing_site,
            clave, subregion, OHI_rgn, long, lat,
            dist = 0, match_type = "exact_raw")

todo1 <- anti_join(fish_names, map1, by = "landing_site")

# ---- 2) EXACT BASIC (no stop removal) ----
# N=250, do a quick check to make sure they are sensible!
map2 <- todo1 %>%
  inner_join(sites_k %>% select(name_basic, landing_site_sites, clave, subregion, OHI_rgn, long, lat),
             by = "name_basic") %>%
  transmute(landing_site,
            landing_site_sites, clave, subregion, OHI_rgn, long, lat,
            dist = 0, match_type = "exact_basic")

todo2 <- anti_join(todo1, map2 %>% select(landing_site), by = "landing_site")

# ---- 3) EXACT STOPLESS (with stop removal) ----
# N=26
map3 <- todo2 %>%
  inner_join(sites_k %>% select(name_stopless, landing_site_sites, clave, subregion, OHI_rgn, long, lat),
             by = "name_stopless") %>%
  transmute(landing_site,
            landing_site_sites, clave, subregion, OHI_rgn, long, lat,
            dist = 0, match_type = "exact_stopless") %>%
  filter(!is.na(landing_site))

todo3 <- anti_join(todo2, map3 %>% select(landing_site), by = "landing_site")


# ---- 4) FUZZY on STOPLESS ----

## N = 1: Some questionable ones, but mostly seem reasonable.
map4 <- if (nrow(todo3)) {
  stringdist_inner_join(
    todo3,
    sites_k %>% select(name_stopless, landing_site_sites, clave, subregion, OHI_rgn, long, lat),
    by = c("name_stopless" = "name_stopless"),
    method = "jw", max_dist = 0.15, distance_col = "dist"
  ) %>%
    group_by(landing_site) %>%
    slice_min(dist, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    transmute(landing_site,
              landing_site_sites, clave, subregion, OHI_rgn, long, lat,
              match_type = "fuzzy_stopless", dist)
} else dplyr::tibble()


# ---- combine + attach back ----
map_all <- bind_rows(map1, map2, map3, map4)
map_all <- unique(map_all)

fish_matched <- fish_k %>%
  left_join(map_all, by = "landing_site") %>%
  mutate(GulfCal = ifelse(subregion<3, "no", "yes"))

filter(fish_matched, is.na(landing_site_sites)) # everything got matched

table(fish_matched$GulfCal)
318204/(672646 + 318204)   

# ---- quick QA ----
# these should be the same:
sum(fish_matched$live_weight)
sum(bajasur_fish$live_weight)

fish_matched %>%
  filter(!is.na(lat)) %>%
  summarize(sum(live_weight))
fish_matched %>%
  filter(is.na(lat)) %>%
  summarize(sum(live_weight))
488104769/(3149693408+488104769) #13% of catch couldn't find a match for.

fish_matched %>%
  filter(!is.na(lat) & year == 2024) %>%
  summarize(sum(live_weight))
fish_matched %>%
  filter(is.na(lat) & year == 2024) %>%
  summarize(sum(live_weight))

## for 2024 only 0.5% not accounted for. Most of mismatches are in previous years
 25651/( 25651+190704289)

missing_sites <- fish_matched %>%
  filter(is.na(lat)) %>%
  select(state, office_name, landing_site, year, landing_site_sites)
unique(missing_sites$landing_site)
summary(missing_sites)
tmp <-filter(missing_sites, is.na(landing_site_sites)) ## all sites were accounted for in the list, but
unique(tmp$landing_site)
# I couldn't find the location unambiguously of all sites.
table(missing_sites$year)
table(baja_fish$year)

## nearly all the missmatches are from 2005 and earlier
bc_data <- fish_matched %>%
  select(-name_basic, -name_stopless,  -clave, -dist, -match_type)
write_csv(bc_data, file.path(jc_data, "mex_BCS_landings_2000_present_with_lat_long.csv"))


```


# 3 : Do Sonora

```{r}
## started with this file, but then added some regions that were missing from the 
## catch data
#sites <- read_csv(here("FP/FIS/v2025/data/conapesca_landing_sites_baja_sonora.csv"))
# sonora_sites <- filter(sites, subregion>=5) %>%
#   mutate(State = "Sonora") %>%
#   mutate(GofC = "yes") %>%
#   select(-Data)
#dup_landings <- sonora_sites[duplicated(sonora_sites$landing_site),]
#filter(sonora_sites, landing_site %in% dup_landings$landing_site)

# ## Take care of regions with the same name (very rude!)
# ## average the duplicates which are close to each other

# sonora_sites <- sonora_sites %>%
#   group_by(landing_site) %>%
#   summarize(clave=mean(clave), 
#             subregion= mean(subregion),
#             long = mean(long),
#             lat = mean(lat),
#             State = unique(State))
# write_csv(sonora_sites, here("FP/FIS/v2025/data/fishing_site_locs_raw_Sonora.csv"))


sonora_sites <- read_csv(here("FP/FIS/v2025/data/landing_sites_Sonora.csv")) %>%
  mutate(OHI_rgn = ifelse(lat>= 31.29309, "Sonora_North", "Sonora_South"))



fish <- readRDS(file.path(jc_data, "mex_landings_2000_present.rds")) 
sonora_fish <- filter(fish, state %in% c("SONORA")) 

```


```{r}


# ---- build working copies ----
sites_k <- sonora_sites %>%
  mutate(
    name_basic   = normalize_basic(landing_site),
    name_stopless = normalize_stopless(landing_site)
  ) %>%
  select(landing_site_sites = landing_site,
         name_basic, name_stopless, clave, subregion, OHI_rgn, long, lat)

fish_k <- sonora_fish %>%
  mutate(
    name_basic   = normalize_basic(landing_site),
    name_stopless = normalize_stopless(landing_site)
  )

# Distinct fish names to map once
fish_names <- fish_k %>% distinct(landing_site, name_basic, name_stopless)

# ---- 1) EXACT RAW ----
# N=183
map1 <- fish_names %>%
  inner_join(sites_k %>% select(landing_site_sites, clave, subregion, OHI_rgn, long, lat),
             by = c("landing_site" = "landing_site_sites")) %>%
  transmute(landing_site,
            landing_site_sites = landing_site,
            clave, subregion, OHI_rgn, long, lat,
            dist = 0, match_type = "exact_raw")

todo1 <- anti_join(fish_names, map1, by = "landing_site")

# ---- 2) EXACT BASIC (no stop removal) ----
# N=2, do a quick check to make sure they are sensible!
map2 <- todo1 %>%
  inner_join(sites_k %>% select(name_basic, landing_site_sites, clave, subregion, OHI_rgn, long, lat),
             by = "name_basic") %>%
  transmute(landing_site,
            landing_site_sites, clave, subregion, OHI_rgn, long, lat,
            dist = 0, match_type = "exact_basic")

todo2 <- anti_join(todo1, map2 %>% select(landing_site), by = "landing_site")

# ---- 3) EXACT STOPLESS (with stop removal) ----
# N=0
map3 <- todo2 %>%
  inner_join(sites_k %>% select(name_stopless, landing_site_sites, clave, subregion, OHI_rgn, long, lat),
             by = "name_stopless") %>%
  transmute(landing_site,
            landing_site_sites, clave, subregion, OHI_rgn, long, lat,
            dist = 0, match_type = "exact_stopless") %>%
  filter(!is.na(landing_site))

todo3 <- anti_join(todo2, map3 %>% select(landing_site), by = "landing_site")



# ---- 4) FUZZY on STOPLESS ----

## N = 2: Some questionable ones, but mostly seem reasonable.
map4 <- if (nrow(todo3)) {
  stringdist_inner_join(
    todo3,
    sites_k %>% select(name_stopless, landing_site_sites, clave, subregion, OHI_rgn, long, lat),
    by = c("name_stopless" = "name_stopless"),
    method = "jw", max_dist = 0.15, distance_col = "dist"
  ) %>%
    group_by(landing_site) %>%
    slice_min(dist, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    transmute(landing_site,
              landing_site_sites, clave, subregion, OHI_rgn, long, lat,
              match_type = "fuzzy_stopless", dist)
} else dplyr::tibble()


# ---- combine + attach back ----
map_all <- bind_rows(map1, map2, map3, map4)

fish_matched <- fish_k %>%
  left_join(map_all, by = "landing_site") %>%
  mutate(GulfCal = "yes")


# ---- quick QA ----

# these should be the same:
sum(fish_matched$live_weight)
sum(sonora_fish$live_weight)

fish_matched %>%
  filter(!is.na(lat)) %>%
  summarize(sum(live_weight))
fish_matched %>%
  filter(is.na(lat)) %>%
  summarize(sum(live_weight))
4216144936/(4216144936+9428303173) #30% of catch couldn't find a match for.

fish_matched %>%
  filter(!is.na(lat) & year == 2024) %>%
  summarize(sum(live_weight))
fish_matched %>%
  filter(is.na(lat) & year == 2024) %>%
  summarize(sum(live_weight))

## for 2024 14% not accounted for. Most of mismatches are in previous years
103054906/(103054906+634228902)
missing_sites <- fish_matched %>%
  filter(is.na(lat)) %>%
  select(state, office_name, landing_site, year, landing_site_sites)
summary(missing_sites)
tmp <- unique(missing_sites$landing_site)


filter(missing_sites, is.na(landing_site_sites)) ## all sites were accounted for in the list, but


# I couldn't find the location unambiguously of all sites.
table(missing_sites$year)
table(sonora_fish$year)
## there is a lot of unidentified catch in these data, but it does drop off after 2004.
sonora_data <- fish_matched %>%
  select(-name_basic, -name_stopless, -clave, -dist, -match_type)
write_csv(sonora_data, file.path(jc_data, "mex_sonora_landings_2000_present_with_lat_long.csv"))



```



## Combine
```{r}

BC <- read_csv(file.path(jc_data, "mex_BC_landings_2000_present_with_lat_long.csv"))
BCS <- read_csv(file.path(jc_data, "mex_BCS_landings_2000_present_with_lat_long.csv"))
Sonora <- read_csv(file.path(jc_data, "mex_sonora_landings_2000_present_with_lat_long.csv"))

data <- rbind(BC, BCS, Sonora) 

site_key <- data %>%
  select(state, landing_site_catch = landing_site, landing_site_primary=landing_site_sites, lat, long, GulfCal) %>%
  unique()
write_csv(site_key, here("FP/FIS/v2025/data/Baja_Sonora_states_key.csv"))

catch_data <- data %>%
  select(-landing_site_sites)

write_csv(catch_data, file.path(jc_data, "mex_landings_2000_present_with_lat_long.csv"))

```
