---
title: "Script 3 OG-model: Gapfill with frozen model3.5 + model1.5 (train 2019–2024)"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor:
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options:
  chunk_output_type: console
---

# 0. Workflow for this script

This is an **optional parallel** to main Script 3. Run it after Scripts 1 and 2. It does **not** replace `script3_gapfilling_arrivals_by_employees.qmd`.

**Why this exists:** compare a **frozen** copy of the original production stack from `rstudio-export/gapfilling_arrivals_by_employees.qmd` against main Script 3 (which refits through available SECTUR years). Same models, clearer naming:

- **model3.5** (`arrivals ~ total_employees + municipio_clean + year_factor`) for municipios with some SECTUR history  
  (`year_factor = 1` in 2020, `2` otherwise — so every filled year uses model3.5; 2020 just gets the COVID intercept)
- **model1.5** (`arrivals ~ total_employees + year_factor`) for DENUE-only municipios

**Training freeze (reproducibility):** fit on complete SECTUR + DENUE cases for **2019–2024 only**. Hold those coefficients fixed and `predict()` missing municipio-years on **2019–2025** only (2026 SECTUR is incomplete, so we do not gapfill 2026 here). Observed SECTUR values are never overwritten — so pre-2025 *observed* rows stay put, and gapfilled levels follow the locked 2019–2024 mapping.

1. Bring in DENUE accommodations (employees) from Script 1 (file may include 2026; this script only uses employees through 2025 for fills).
2. Bring in SECTUR total arrivals shapefiles from Script 2 (complete years through 2025).
3. Harmonize municipio names, join employees and arrivals, and check coverage gaps.
4. Fit model1 / model3 / model1.5 / model3.5 on **2019–2024** (model1 / model3 kept for diagnostics / AIC only).
5. Gapfill missing rows with **model3.5 / model1.5**; censor negatives at 0.
6. Write OG-model outputs for `script4_OGmodel_tr_quantity.qmd` (filenames end in `_OGmodel`).

**Data window:** train **2019–2024**; predict / write fills for **2019–2025**.

# 1. Bring in DENUE accommodations (employees)
We will need to clean it up, because the number of employees for each business is binned.

*This is what the per_ocu variable description says in the metadata:*
"This includes personnel directly employed by the company and external personnel supplied by another company, who worked for the economic unit, under its direction and control, and who covered at least one-third of the workday. This may include permanent staff, paid temporary staff, or unpaid temporary staff.
Economic units are classified by employee size ranges, which allows for the identification of the size of Economic Units based on the number of personnel they employ; that is, according to their total workforce. For the purposes of the Directory, Economic Units are grouped into employee size ranges or strata, based on the total number of employees reported by each unit, as shown in the following table.
Number of Employees
1 = 0 to 5
2 = 6 to 10
3 = 11 to 30
4 = 31 to 50
5 = 51 to 100
6 = 101 to 250
7 = 251 and above" 

So, we will take the midpoint of each range.

```{r}
library(tidyverse)
library(magrittr)
library(here)
library(janitor)
library(tmaptools)
library(sf)
library(mapview)

alojamientos_2019_2026_50km <- st_read(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas/int","update_alojamientos_2019_2026_goc_50km.shp"))

## see what the variable per_ocu looks like
unique(alojamientos_2019_2026_50km$per_ocu)

# [1] "0 a 5 personas"        "6 a 10 personas"       "11 a 30 personas"      "31 a 50 personas"      "101 a 250 personas"    "51 a 100 personas"     "251 y m<e1>s personas" "251 y más personas" 

## use a case_when to reclassify the number of people employed based on the midpoint of the range.  NOTE that I am assuming that you can never have 0 employees, so the first midpoint is between 1 and 5.
alojamientos_est_perocu <- alojamientos_2019_2026_50km %>% 
  mutate(
    employees_estimated = case_when(
      per_ocu == "0 a 5 personas" ~ 3,
      per_ocu == "6 a 10 personas" ~ 8,
      per_ocu == "11 a 30 personas" ~ 20.5,
      per_ocu == "31 a 50 personas" ~ 40.5,
      per_ocu == "51 a 100 personas" ~ 75.5,
      per_ocu == "101 a 250 personas" ~ 175.5,
      per_ocu %in% c("251 y m<e1>s personas", "251 y más personas") ~ 300,
      TRUE ~ NA_real_
    )
  )

mapview(alojamientos_est_perocu, zcol = "employees_estimated")

unique(alojamientos_est_perocu$municipio)

# when 10km:
#  [1] "Mexicali"                 "Ensenada"                 "Muleg<e9>"                "Los Cabos"                "La Paz"                   "Loreto"                   "Puerto Vallarta"          "Cabo Corrientes"         
#  [9] "Bah<ed>a de Banderas"     "San Blas"                 "Compostela"               "Tecuala"                  "Puerto Pe<f1>asco"        "San Luis R<ed>o Colorado" "Hermosillo"               "Pitiquito"               
# [17] "Guaymas"                  "Empalme"                  "Benito Ju<e1>rez"         "Huatabampo"               "Elota"                    "Mazatl<e1>n"              "Escuinapa"                "San Ignacio"             
# [25] "Ahome"                    "Navolato"                 "Culiac<e1>n"              "Guasave"                  "Angostura"                "Rosario"                  "San Quint<ed>n"           "San Felipe"              
# [33] "San Quintín"              "Mulegé"                   "Bahía de Banderas"        "Puerto Peñasco"           "San Luis Río Colorado"    "Benito Juárez"            "Mazatlán"                 "Culiacán"                
# [41] "Caborca" 

# when 50km:
#  [1] "Mexicali"                   "Ensenada"                   "Muleg<e9>"                  "Comond<fa>"                
#  [5] "Los Cabos"                  "La Paz"                     "Loreto"                     "Puerto Vallarta"           
#  [9] "Talpa de Allende"           "Tomatl<e1>n"                "San Sebasti<e1>n del Oeste" "Cabo Corrientes"           
# [13] "Mascota"                    "Bah<ed>a de Banderas"       "Tepic"                      "San Blas"                  
# [17] "Compostela"                 "Acaponeta"                  "Ru<ed>z"                    "Xalisco"                   
# [21] "Tuxpan"                     "Tecuala"                    "Santiago Ixcuintla"         "Rosamorada"                
# [25] "Huajicori"                  "San Pedro Lagunillas"       "Puerto Pe<f1>asco"          "San Luis R<ed>o Colorado"  
# [29] "Hermosillo"                 "Caborca"                    "Pitiquito"                  "Navojoa"                   
# [33] "Cajeme"                     "Guaymas"                    "Huatabampo"                 "Etchojoa"                  
# [37] "Empalme"                    "Benito Ju<e1>rez"           "Elota"                      "Mazatl<e1>n"               
# [41] "Navolato"                   "Ahome"                      "Culiac<e1>n"                "El Fuerte"                 
# [45] "Escuinapa"                  "San Ignacio"                "Guasave"                    "Rosario"                   
# [49] "Mocorito"                   "Salvador Alvarado"          "Concordia"                  "Angostura"                 
# [53] "San Quint<ed>n"             "San Felipe"                 "San Quintín"                "Mulegé"                    
# [57] "Comondú"                    "Tomatlán"                   "San Sebastián del Oeste"    "Bahía de Banderas"         
# [61] "Ruíz"                       "Puerto Peñasco"             "San Luis Río Colorado"      "Benito Juárez"             
# [65] "Culiacán"                   "Mazatlán"    

unique(alojamientos_est_perocu$localidad)

# when 10km:
#  [1] "San Felipe"                            "Villas de Cortez"                      "Campo Beluga"                          "Campo Daggets"                         "Campo Sacrificio"                     
#   [6] "Punta Final (Villa Mar y Sol)"         "Campo Raquel y Larry"                  "Punta Bufeo [Campo Tur<ed>stico]"      "Bah<ed>a de los <c1>ngeles"            "La Ventana [Campo]"                   
#  [11] "Puerto San Francisquito"               "Heroica Muleg<e9>"                     "Santa Rosal<ed>a"                      "Cabo San Lucas"                        "Buenavista"                           
#  [16] "San Jos<e9> del Cabo"                  "La Cuadra Regina"                      "La Paz"                                "Chametla"                              "La Ribera"                            
#  [21] "El Sargento"                           "Puerto de Pichilingue"                 "Loreto"                                "La Ventana"                            "El Cardonal"                          
#  [26] "Nopol<f3>"                             "Buenos Aires"                          "Los Barriles"                          "Punta Pescadero"                       "Tripu<ed> [Trailer Park]"             
#  [31] "Boca de la Vinorama"                   "Danzante Bay"                          "Punta Prieta"                          "Ninguno"                               "El Leonero"                           
#  [36] "Puerto Vallarta"                       "Sun Resorts"                           "Mismaloya"                             "Yelapa"                                "Las Juntas y los Veranos"             
#  [41] "Le<f3>n de la Sierra"                  "La Quebrada"                           "Ixtapa"                                "Las Juntas"                            "Playa Gemelas"                        
#  [46] "Sayulita"                              "Nuevo Vallarta"                        "Guadalupe Victoria"                    "Rinc<f3>n de Guayabitos"               "Bucer<ed>as"                          
#  [51] "Aticama"                               "Cruz de Huanacaxtle"                   "El Porvenir"                           "La Pe<f1>ita de Jaltemba"              "Los Ayala"                            
#  [56] "Platanitos"                            "Lo de Marcos"                          "Chacala"                               "La Palma"                              "San Blas"                             
#  [61] "Matanch<e9>n (La Aguada)"              "San Francisco"                         "Corral del Risco (Punta de Mita)"      "Lidia [Bungalows]"                     "Playa de los Cocos"                   
#  [66] "La Joya"                               "P<e1>tzcuaro"                          "Monte<f3>n"                            "Punta Raza"                            "Las Varas"                            
#  [71] "La Manzanilla"                         "Flamingos"                             "Santa Cruz de Miramar"                 "Tondoroque"                            "Desarrollo Punta Mita"                
#  [76] "Bah<ed>a de Matanch<e9>n"              "El Crucero de Chacala"                 "Punta Negra (Careyeros)"               "San Jos<e9> del Valle"                 "Para<ed>so Sereno"                    
#  [81] "Mezcales"                              "Playa Escondida"                       "Universo [Colonia]"                    "Playa Novillero"                       "Grand Palladium Vallarta [Hotel]"     
#  [86] "San Vicente"                           "Las Jarretaderas"                      "San Clemente de Lima"                  "Arboledas [Bungalows]"                 "La Tranquila [Condominios]"           
#  [91] "Imanta"                                "Los Arroyos Verdes"                    "Dos Palmeras (Casa Pel<ed>canos)"      "El Tonino [Bloquera]"                  "Punta Monterrey"                      
#  [96] "Rancho Banderas"                       "Papelillos [Hotel]"                    "Puerto Pe<f1>asco"                     "Golfo de Santa Clara"                  "Bah<ed>a la Choya"                    
# [101] "Playa Arenosa"                         "Bah<ed>a de Kino"                      "Puerto Libertad"                       "Mayan Palace Hotel"                    "Playa Miramar"                        
# [106] "Las Conchas"                           "Heroica Guaymas"                       "San Carlos (San Carlos Nuevo Guaymas)" "Condominios Playa Blanca"              "Empalme"                              
# [111] "Villa Ju<e1>rez"                       "Agua Blanca"                           "Las Bocas"                             "Costa del Mar"                         "La Cruz"                              
# [116] "Mazatl<e1>n"                           "Color Marino"                          "Dimas (Estaci<f3>n Dimas)"             "Teacap<e1>n"                           "Topolobampo"                          
# [121] "Altata"                                "Isla del Bosque"                       "Eldorado"                              "El Saucito"                            "Celestino Gazca Villase<f1>or"        
# [126] "Las Glorias"                           "Fraccionamiento Isla Para<ed>so"       "La Reforma"                            "Gustavo D<ed>az Ordaz (El Carrizo)"    "La Escondida"                         
# [131] "Agua Verde"                            "Localidad Econ<f3>mica"                "Rancho Girolando"                      "El Patio"                              "Ejido Plan Nacional Agrario"          
# [136] "Alfonsinas"                            "Tailwind"                              "El Encuentro"                          "Los Mochis"                            "Zacualpan"                            
# [141] "Desarrollo Puerto Bah<ed>a"            "Litib<fa>"                             "Bahía de los Ángeles"                  "Punta Bufeo [Campo Turístico]"         "Heroica Mulegé"                       
# [146] "Santa Rosalía"                         "San José del Cabo"                     "Tripuí [Trailer Park]"                 "San Román"                             "Nopoló"                               
# [151] "Costa Vida Vallarta"                   "Rincón de Guayabitos"                  "Bucerías"                              "Bahía de Matanchén"                    "La Peñita de Jaltemba"                
# [156] "San José del Valle"                    "Dos Palmeras (Casa Pelícanos)"         "Pátzcuaro"                             "Desarrollo Puerto Bahía"               "Litibú"                               
# [161] "Puerto Peñasco"                        "Bahía de Kino"                         "Villa Juárez"                          "Mazatlán"                              "Campo Santa Fe"                       
# [166] "Teacapán"                              "Dimas (Estación Dimas)"                "Motel Popeye"                          "Gustavo Díaz Ordaz (El Carrizo)"       "Playa del Sol"                        
# [171] "Pap<e1> Fern<e1>ndez [Campo]"          "Los Vientos"                           "Campo Lobo"                            "Cabo Golf Cars"                        "Cuesta Blanca"                        
# [176] "Las <c1>nimas de Quimixto"             "Lienzo Charro"                         "Mezcalitos"                            "Para<ed>so Escondido [Colonia]"        "Jalcocot<e1>n"                        
# [181] "Las Tijerillas"                        "Luis Pe<f1>a"                          "Hip<f3>lito Gonz<e1>lez L<f3>pez"      "Limoncito Hills"                       "La Playa Estates"                     
# [186] "Haixa (Litib<fa>)"                     "Playa Encanto"                         "Puerto Lobos"                          "Mayan Lakes"                           "Playa Huatabampito"                   
# [191] "Coch<f3>rit" 

# when 50km:
#   [1] "San Felipe"                                      "Villas de Cortez"                               
#   [3] "Campo Beluga"                                    "Campo Daggets"                                  
#   [5] "Campo Sacrificio"                                "Punta Final (Villa Mar y Sol)"                  
#   [7] "Campo Raquel y Larry"                            "Punta Bufeo [Campo Tur<ed>stico]"               
#   [9] "Bah<ed>a de los <c1>ngeles"                      "La Ventana [Campo]"                             
#  [11] "Catavi<f1>a"                                     "Puerto San Francisquito"                        
#  [13] "Rancho Escondido"                                "Piedra Blanca"                                  
#  [15] "Heroica Muleg<e9>"                               "Santa Rosal<ed>a"                               
#  [17] "San Isidro"                                      "San Lino"                                       
#  [19] "Cabo San Lucas"                                  "Buenavista"                                     
#  [21] "San Jos<e9> del Cabo"                            "La Cuadra Regina"                               
#  [23] "La Paz"                                          "Chametla"                                       
#  [25] "El Triunfo"                                      "La Acacia"                                      
#  [27] "La Ribera"                                       "El Sargento"                                    
#  [29] "Puerto de Pichilingue"                           "Loreto"                                         
#  [31] "San Miguel de Comond<fa>"                        "La Ventana"                                     
#  [33] "El Cardonal"                                     "Nopol<f3>"                                      
#  [35] "Buenos Aires"                                    "Los Barriles"                                   
#  [37] "Punta Pescadero"                                 "Tripu<ed> [Trailer Park]"                       
#  [39] "Boca de la Vinorama"                             "Danzante Bay"                                   
#  [41] "Punta Prieta"                                    "Ninguno"                                        
#  [43] "El Leonero"                                      "Puerto Vallarta"                                
#  [45] "Talpa de Allende"                                "La Cruz de Loreto"                              
#  [47] "Sun Resorts"                                     "San Sebasti<e1>n del Oeste"                     
#  [49] "Mismaloya"                                       "Yelapa"                                         
#  [51] "Las Juntas y los Veranos"                        "Mascota"                                        
#  [53] "El Tuito"                                        "Puerta de Enmedio"                              
#  [55] "Le<f3>n de la Sierra"                            "La Quebrada"                                    
#  [57] "Ixtapa"                                          "San Nicol<e1>s (La Troje)"                      
#  [59] "La Puntilla de Maito"                            "Las Juntas"                                     
#  [61] "Playa Gemelas"                                   "Sayulita"                                       
#  [63] "Nuevo Vallarta"                                  "Tepic"                                          
#  [65] "Guadalupe Victoria"                              "Rinc<f3>n de Guayabitos"                        
#  [67] "Bucer<ed>as"                                     "Aticama"                                        
#  [69] "Cruz de Huanacaxtle"                             "El Porvenir"                                    
#  [71] "La Pe<f1>ita de Jaltemba"                        "Los Ayala"                                      
#  [73] "Platanitos"                                      "Lo de Marcos"                                   
#  [75] "Chacala"                                         "La Palma"                                       
#  [77] "San Blas"                                        "Matanch<e9>n (La Aguada)"                       
#  [79] "San Francisco"                                   "Corral del Risco (Punta de Mita)"               
#  [81] "Lidia [Bungalows]"                               "Playa de los Cocos"                             
#  [83] "La Joya"                                         "P<e1>tzcuaro"                                   
#  [85] "Monte<f3>n"                                      "Punta Raza"                                     
#  [87] "Las Varas"                                       "Acaponeta"                                      
#  [89] "La Manzanilla"                                   "Flamingos"                                      
#  [91] "Santa Cruz de Miramar"                           "Tondoroque"                                     
#  [93] "Desarrollo Punta Mita"                           "Compostela"                                     
#  [95] "Ru<ed>z"                                         "San Juan de Abajo"                              
#  [97] "Xalisco"                                         "Bah<ed>a de Matanch<e9>n"                       
#  [99] "Tuxpan"                                          "El Crucero de Chacala"                          
# [101] "Tecuala"                                         "Punta Negra (Careyeros)"                        
# [103] "Santiago Ixcuintla"                              "Rosamorada"                                     
# [105] "San Jos<e9> del Valle"                           "Para<ed>so Sereno"                              
# [107] "Huajicori"                                       "Mezcales"                                       
# [109] "Playa Escondida"                                 "Villa Hidalgo"                                  
# [111] "Universo [Colonia]"                              "Playa Novillero"                                
# [113] "Grand Palladium Vallarta [Hotel]"                "San Vicente"                                    
# [115] "Mexcaltit<e1>n de Uribe"                         "Las Jarretaderas"                               
# [117] "L<e1>zaro C<e1>rdenas [Colonia]"                 "San Clemente de Lima"                           
# [119] "Arboledas [Bungalows]"                           "La Tranquila [Condominios]"                     
# [121] "Imanta"                                          "Los Arroyos Verdes"                             
# [123] "Dos Palmeras (Casa Pel<ed>canos)"                "San Cayetano"                                   
# [125] "El Tonino [Bloquera]"                            "San Pedro Lagunillas"                           
# [127] "Punta Monterrey"                                 "Rancho Banderas"                                
# [129] "Papelillos [Hotel]"                              "Puerto Pe<f1>asco"                              
# [131] "Golfo de Santa Clara"                            "Bah<ed>a la Choya"                              
# [133] "Playa Arenosa"                                   "Bah<ed>a de Kino"                               
# [135] "Miguel Alem<e1>n (La Doce)"                      "Plutarco El<ed>as Calles (La Y Griega)"         
# [137] "Puerto Libertad"                                 "Mayan Palace Hotel"                             
# [139] "Playa Miramar"                                   "Las Conchas"                                    
# [141] "Navojoa"                                         "Ciudad Obreg<f3>n"                              
# [143] "Heroica Guaymas"                                 "San Carlos (San Carlos Nuevo Guaymas)"          
# [145] "Condominios Playa Blanca"                        "Huatabampo"                                     
# [147] "Bacobampo"                                       "Empalme"                                        
# [149] "Esperanza"                                       "Villa Ju<e1>rez"                                
# [151] "Agua Blanca"                                     "Pueblo Yaqui"                                   
# [153] "Los Bahuises"                                    "V<ed>cam (Switch)"                              
# [155] "Las Bocas"                                       "Costa del Mar"                                  
# [157] "La Cruz"                                         "Campo el Salto"                                 
# [159] "Mazatl<e1>n"                                     "San Marcos"                                     
# [161] "Navolato"                                        "Los Mochis"                                     
# [163] "Culiac<e1>n Rosales"                             "Color Marino"                                   
# [165] "Dimas (Estaci<f3>n Dimas)"                       "Adolfo Ruiz Cortines"                           
# [167] "Leopoldo S<e1>nchez Celis"                       "Escuinapa de Hidalgo"                           
# [169] "San Ignacio"                                     "La Papalota"                                    
# [171] "Teacap<e1>n"                                     "El Rosario"                                     
# [173] "Mocorito"                                        "El Fuerte"                                      
# [175] "Topolobampo"                                     "Villa Uni<f3>n"                                 
# [177] "Altata"                                          "Guam<fa>chil"                                   
# [179] "Isla del Bosque"                                 "Guasave"                                        
# [181] "Juan Jos<e9> R<ed>os"                            "Quil<e1>"                                       
# [183] "Eldorado"                                        "El Saucito"                                     
# [185] "El Batall<f3>n"                                  "Pericos"                                        
# [187] "Celestino Gazca Villase<f1>or"                   "Las Glorias"                                    
# [189] "Costa Rica"                                      "Concordia"                                      
# [191] "Fraccionamiento Isla Para<ed>so"                 "Sistema de Producci<f3>n Intensiva del Noroeste"
# [193] "La Reforma"                                      "Gustavo D<ed>az Ordaz (El Carrizo)"             
# [195] "Licenciado Benito Ju<e1>rez (Campo Gobierno)"    "La Escondida"                                   
# [197] "Santa Eduwiges"                                  "Caimanero"                                      
# [199] "Callejones de Guasavito"                         "Agua Verde"                                     
# [201] "Le<f3>n Fonseca (Estaci<f3>n Verdura)"           "Nestle Culiac<e1>n"                             
# [203] "Localidad Econ<f3>mica"                          "Rancho Girolando"                               
# [205] "Ninguno [Motel]"                                 "El Patio"                                       
# [207] "Gabriel Leyva Solano (Benito Ju<e1>rez)"         "Ejido Plan Nacional Agrario"                    
# [209] "Alfonsinas"                                      "Tailwind"                                       
# [211] "Providencia"                                     "El Encuentro"                                   
# [213] "Zacualpan"                                       "Ruiz"                                           
# [215] "Desarrollo Puerto Bah<ed>a"                      "Litib<fa>"                                      
# [217] "Heroica Ciudad Concordia"                        "Productos Enlatados Madre Tierra"               
# [219] "Ferretera y Torniller<ed>a Zapata [FETOZA]"      "Walamito"                                       
# [221] "Bahía de los Ángeles"                            "Cataviña"                                       
# [223] "Punta Bufeo [Campo Turístico]"                   "Heroica Mulegé"                                 
# [225] "Santa Rosalía"                                   "San José del Cabo"                              
# [227] "Tripuí [Trailer Park]"                           "San Román"                                      
# [229] "Nopoló"                                          "San Sebastián del Oeste"                        
# [231] "San Nicolás (La Troje)"                          "Costa Vida Vallarta"                            
# [233] "Rincón de Guayabitos"                            "Bucerías"                                       
# [235] "Bahía de Matanchén"                              "La Peñita de Jaltemba"                          
# [237] "San José del Valle"                              "Dos Palmeras (Casa Pelícanos)"                  
# [239] "Pátzcuaro"                                       "Potrero las Lomas"                              
# [241] "Desarrollo Puerto Bahía"                         "Litibú"                                         
# [243] "Mexcaltitán de Uribe"                            "Lázaro Cárdenas [Colonia]"                      
# [245] "Los Tabachines"                                  "Puerto Peñasco"                                 
# [247] "Bahía de Kino"                                   "Miguel Alemán (La Doce)"                        
# [249] "Plutarco Elías Calles (La Y Griega)"             "Villa Juárez"                                   
# [251] "Ciudad Obregón"                                  "Vícam (Switch)"                                 
# [253] "Culiacán Rosales"                                "Mazatlán"                                       
# [255] "Guamúchil"                                       "Sistema de Producción Intensiva del Noroeste"   
# [257] "Campo Santa Fe"                                  "El Batallón"                                    
# [259] "Teacapán"                                        "La Caída"                                       
# [261] "Villa Unión"                                     "Dimas (Estación Dimas)"                         
# [263] "Quilá"                                           "Juan José Ríos"                                 
# [265] "Motel Popeye"                                    "Licenciado Benito Juárez (Campo Gobierno)"      
# [267] "Gustavo Díaz Ordaz (El Carrizo)"                 "Nestle Culiacán"                                
# [269] "León Fonseca (Estación Verdura)"                 "Motel Paradise"                                 
# [271] "Playa del Sol"                                   "Pap<e1> Fern<e1>ndez [Campo]"                   
# [273] "Los Vientos"                                     "Campo Lobo"                                     
# [275] "Cabo Golf Cars"                                  "Club Campestre el Madrigal"                     
# [277] "Cuesta Blanca"                                   "San Dionisio"                                   
# [279] "Las <c1>nimas de Quimixto"                       "Lienzo Charro"                                  
# [281] "Mezcalitos"                                      "Para<ed>so Escondido [Colonia]"                 
# [283] "Jalcocot<e1>n"                                   "Las Tijerillas"                                 
# [285] "Luis Pe<f1>a"                                    "Hip<f3>lito Gonz<e1>lez L<f3>pez"               
# [287] "Limoncito Hills"                                 "La Playa Estates"                               
# [289] "La Cantera"                                      "Haixa (Litib<fa>)"                              
# [291] "Playa Encanto"                                   "Puerto Lobos"                                   
# [293] "Mayan Lakes"                                     "Playa Huatabampito"                             
# [295] "Coch<f3>rit"                                     "Anglers Inn [Bill Chapman Jr.]"                 
# [297] "La Ca<ed>da"                                     "San Pedro"                                      
# [299] "El Bordo"                                        "TUMSA [Transportes]"                            
# [301] "El Diez"                                         "Cortijo la Piedra"  
```

```{r}
# aggregate the accommodation employee data by municipality and year
employees_by_muni_yr_50 <- alojamientos_est_perocu %>% 
  mutate(municipio = case_when(
    municipio == "Mexicali" ~ "San Felipe",
    TRUE ~ municipio
  )) %>% 
  group_by(municipio, estado, data_year) %>% 
  summarize(
    total_employees = sum(employees_estimated, na.rm = TRUE),
    num_businesses = n(),
    avg_employees_per_business = mean(employees_estimated, na.rm = TRUE),
    .groups = "drop"
  )
```


# 2. Bring in SECTUR total visitors from Script 2
These shapefiles were written out from `~TR/script2_arrivals_prep.qmd`.

```{r}
## from SECTUR complete annual years (through 2025); re-written from script2_arrivals_prep.qmd
dom_goc <- st_read(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/hotel_monitoring_int","domestic_arrivals_sectur_goc.shp"))

int_goc <- st_read(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/hotel_monitoring_int","international_arrivals_sectur_goc.shp"))

## join back together to have both spatial attributes and the number of visitors
all_arrivals <- bind_rows(dom_goc, int_goc)

## Datatur Ensenada / San Quintín are Pacific tourist centers. Drop their SECTUR
## arrivals so they are gapfilled from GOC-buffer DENUE hotels (Region_2), not
## scored with Pacific observed demand.
pacific_sectur_centers <- c("ensenada", "san_quintin", "san_quintín")
n_pacific <- sum(tolower(all_arrivals$municip) %in% pacific_sectur_centers)
message(
  "Dropping Pacific SECTUR centers from GoC gapfill training (n rows = ",
  n_pacific, "): ensenada, san_quintín"
)
all_arrivals <- all_arrivals %>%
  filter(!tolower(municip) %in% pacific_sectur_centers)

total_arrivals_yr_muni <- all_arrivals %>% 
  group_by(year, municip, estado, rgn_id) %>% 
  summarize(
    arrivals = {
      x <- as.numeric(value)
      x <- x[is.finite(x)]
      if (length(x) == 0) NA_real_ else sum(x)
    },
    .groups = "drop"
  )

mapview(total_arrivals_yr_muni, zcol = "arrivals")
```


# 3. Harmonize municipio names and merge datasets
First, clean them both up:

```{r}
## look for differences

unique(employees_by_muni_yr_50$municipio)
# [1] "Ahome"                    "Angostura"                "Bah<ed>a de Banderas"     "Bahía de Banderas"        "Benito Ju<e1>rez"         "Benito Juárez"            "Cabo Corrientes"          "Caborca"                 
#  [9] "Compostela"               "Culiac<e1>n"              "Culiacán"                 "Elota"                    "Empalme"                  "Ensenada"                 "Escuinapa"                "Guasave"                 
# [17] "Guaymas"                  "Hermosillo"               "Huatabampo"               "La Paz"                   "Loreto"                   "Los Cabos"                "Mazatl<e1>n"              "Mazatlán"                
# [25] "Muleg<e9>"                "Mulegé"                   "Navolato"                 "Pitiquito"                "Puerto Pe<f1>asco"        "Puerto Peñasco"           "Puerto Vallarta"          "Rosario"                 
# [33] "San Blas"                 "San Felipe"               "San Ignacio"              "San Luis R<ed>o Colorado" "San Luis Río Colorado"    "San Quint<ed>n"           "San Quintín"              "Tecuala"               



unique(total_arrivals_yr_muni$municip)
#  [1] "cabo_san_lucas"          "culiacan"                "el_fuerte"               "el_rosario"              "escuinapa"               "guaymas"                 "hermosillo"              "la_paz"                  "loreto"                 
# [10] "los_mochis"              "mazatlan"                "puerto_vallarta"         "san_felipe"              "san_jose_del_cabo"       "zona_corredor_los_cabos"
```
```{r}
## for 50km
employees_by_muni_yr_join_50 <- employees_by_muni_yr_50 %>%
  st_drop_geometry() %>% 
  mutate(
    municipio_clean = municipio %>%
      str_replace_all("Bah<ed>a", "Bahía") %>%
      str_replace_all("Ju<e1>rez", "Juárez") %>%
      str_replace_all("Culiac<e1>n", "Culiacán") %>%
      str_replace_all("Mazatl<e1>n", "Mazatlán") %>%
      str_replace_all("Muleg<e9>", "Mulegé") %>%
      str_replace_all("Pe<f1>asco", "Peñasco") %>%
      str_replace_all("R<ed>o", "Río") %>%
      str_replace_all("Ru<ed>z", "Ruíz") %>%
      str_replace_all("Quint<ed>n", "Quintín") %>%
      str_replace_all("Comond<fa>", "Comondú") %>%
      str_replace_all("San Sebasti<e1>n del Oeste", "San Sebastián del Oeste") %>%
      str_replace_all("Tomatl<e1>n", "Tomatlán") %>%
      str_replace_all("San Luis R<ed>o Colorado", "San Luis Río Colorado") %>%
      str_replace_all("<fa>", "ú") %>%
      str_replace_all("<e1>", "á") %>%
      str_replace_all("<ed>", "í") %>%
      str_replace_all("<f1>", "ñ") %>%
      str_replace_all("<e9>", "é") %>%
      str_to_lower() %>%
      str_replace_all(" ", "_"),
    
    estado = estado %>% 
      str_to_lower() %>%
      str_replace_all(" ", "_") %>% 
      ifelse(municipio_clean == "puerto_vallarta", "jalisco", .),
    
    data_year = as.numeric(data_year)
  ) %>% 
  dplyr::select(-municipio) %>% 
  complete(nesting(municipio_clean, estado), data_year) %>%  # for ensenada (missing 2024), san quintin (missing 2019, 2020) and san pedro laguilla (missing 2019 and 2020)
  mutate(
    total_employees = replace_na(total_employees, 0),
    num_businesses = replace_na(num_businesses, 0),
    avg_employees_per_business = replace_na(avg_employees_per_business, 0)
  ) ## if there are no economic units in the denue directory for a year, then we can assume there are no formal businesses. Will need to note this.
```


```{r}
## for 50km
employees_by_muni_yr_join_50_sf <- employees_by_muni_yr_50 %>%
  mutate(
    municipio_clean = municipio %>%
      str_replace_all("Bah<ed>a", "Bahía") %>%
      str_replace_all("Ju<e1>rez", "Juárez") %>%
      str_replace_all("Culiac<e1>n", "Culiacán") %>%
      str_replace_all("Mazatl<e1>n", "Mazatlán") %>%
      str_replace_all("Muleg<e9>", "Mulegé") %>%
      str_replace_all("Pe<f1>asco", "Peñasco") %>%
      str_replace_all("R<ed>o", "Río") %>%
      str_replace_all("Ru<ed>z", "Ruíz") %>%
      str_replace_all("Quint<ed>n", "Quintín") %>%
      str_replace_all("Comond<fa>", "Comondú") %>%
      str_replace_all("San Sebasti<e1>n del Oeste", "San Sebastián del Oeste") %>%
      str_replace_all("Tomatl<e1>n", "Tomatlán") %>%
      str_replace_all("San Luis R<ed>o Colorado", "San Luis Río Colorado") %>%
      str_replace_all("<fa>", "ú") %>%
      str_replace_all("<e1>", "á") %>%
      str_replace_all("<ed>", "í") %>%
      str_replace_all("<f1>", "ñ") %>%
      str_replace_all("<e9>", "é") %>%
      str_to_lower() %>%
      str_replace_all(" ", "_"),
    
    estado = estado %>% 
      str_to_lower() %>%
      str_replace_all(" ", "_") %>% 
      ifelse(municipio_clean == "puerto_vallarta", "jalisco", .),
    
    data_year = as.numeric(data_year)
  ) %>% 
  dplyr::select(municipio_clean, geometry) %>% 
  distinct()
```


```{r}
## Reproducible region assignment:
## 1) one representative point per municipio (point on surface)
## 2) must fall inside inland_50km_goc (same coastal mask as DENUE)
## 3) assign OHI region by nearest OHI region polygon (handles boundary ties)
## No manual case_when — re-running with the same shapefiles gives the same rgn_id.
inland_50 <- st_read(here::here(
  "/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer",
  "inland_50km_goc.shp"
))
goc_poly <- st_read(here("spatial/ohi_regions", "goc_ohi_rgns.shp")) %>%
  select(rgn_id) %>%
  st_make_valid()

st_crs(inland_50) == st_crs(employees_by_muni_yr_join_50_sf) ## TRUE

muni_pts <- employees_by_muni_yr_join_50_sf %>%
  st_make_valid() %>%
  mutate(geometry = st_point_on_surface(geometry))

in_50 <- lengths(st_intersects(muni_pts, inland_50)) > 0
message(
  "DENUE municipios outside inland_50 (dropped from region lookup): ",
  paste(sort(muni_pts$municipio_clean[!in_50]), collapse = ", ")
)
muni_pts_50 <- muni_pts[in_50, ]

## nearest OHI region polygon to exactly one rgn_id per municipio (reproducible)
goc_poly_t <- st_transform(goc_poly, st_crs(muni_pts_50))
nn <- st_nearest_feature(muni_pts_50, goc_poly_t)

denue_municipios_rgn_fix <- muni_pts_50 %>%
  st_drop_geometry() %>%
  mutate(rgn_id = goc_poly_t$rgn_id[nn]) %>%
  distinct(municipio_clean, rgn_id) %>%
  arrange(municipio_clean)

## sanity: one row per municipio
denue_municipios_rgn_fix %>%
  count(municipio_clean) %>%
  filter(n > 1)

denue_municipios_rgn_fix %>%
  count(rgn_id)

mapview(muni_pts_50 %>% mutate(rgn_id = goc_poly_t$rgn_id[nn]), zcol = "rgn_id") +
  mapview(goc_poly_t, alpha.regions = 0.2)

write_csv(
  denue_municipios_rgn_fix,
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/hotel_monitoring_int",
    "ohi_rgn_assign_for_municipio_clean_OGmodel.csv"
  )
)
```

```{r}
### for 10km
## denue accommodation employees
# employees_by_muni_yr_join <- employees_by_muni_yr %>%
#   st_drop_geometry() %>% 
#   mutate(
#     municipio_clean = municipio %>%
#       str_replace_all("Bah<ed>a", "Bahía") %>%
#       str_replace_all("Ju<e1>rez", "Juárez") %>%
#       str_replace_all("Culiac<e1>n", "Culiacán") %>%
#       str_replace_all("Mazatl<e1>n", "Mazatlán") %>%
#       str_replace_all("Muleg<e9>", "Mulegé") %>%
#       str_replace_all("Pe<f1>asco", "Peñasco") %>%
#       str_replace_all("R<ed>o", "Río") %>%
#       str_replace_all("Quint<ed>n", "Quintín") %>%
#       
#       # make lower case and remove spaces to match the arrivals data
#       str_to_lower() %>%
#       str_replace_all(" ", "_"),
#     estado = estado %>% 
#       str_to_lower() %>%
#       str_replace_all(" ", "_"),
#     data_year = as.numeric(data_year)
#   )
# 
# unique(employees_by_muni_yr_join$municipio_clean)
```


```{r}
## arrivals data tweaks for join
## IMPORTANT:
## 1) Some SECTUR points land on OHI region boundaries and get *two* rgn_id values
##    with the *same* arrivals total. Do NOT sum across rgn_id (double-count).
## 2) Dom vs int geocodes can also leave the same center with different estado
##    strings (e.g. "baja_california" vs ""). Grouping by estado then kept two
##    rows/year for ensenada / san_felipe (mexicali merge) / bahía_de_banderas
##    (nuevo_nayarit merge). Collapse to one row per municipio-year.
## 3) max(..., na.rm = TRUE) on an all-NA group returns -Inf in R. That broke lm().
##    Use helpers that return NA instead.
pick_estado <- function(x) {
  x <- x[!is.na(x) & x != "" & x != "na"]
  if (length(x) == 0) NA_character_ else x[[1]]
}

max_or_na <- function(x) {
  x <- x[is.finite(x)]
  if (length(x) == 0) NA_real_ else max(x)
}

sum_or_na <- function(x) {
  x <- x[is.finite(x)]
  if (length(x) == 0) NA_real_ else sum(x)
}

total_arrivals_aggregated_join <- total_arrivals_yr_muni %>%
  st_drop_geometry() %>%
  mutate(
    estado = estado %>%
      str_to_lower() %>%
      str_replace_all(" ", "_"),
    year = as.numeric(year)
  ) %>%
  # 1) one row per original SECTUR center-year (drop duplicate rgn / blank estado)
  group_by(year, municip) %>%
  summarize(
    arrivals = max_or_na(arrivals),
    estado = pick_estado(estado),
    .groups = "drop"
  ) %>%
  mutate(
    municipio_clean = case_when(
      municip %in% c("cabo_san_lucas", "san_jose_del_cabo", "zona_corredor_los_cabos") ~ "los_cabos",
      municip == "los_mochis" ~ "ahome",
      municip == "el_rosario" ~ "rosario",
      municip == "culiacan" ~ "culiacán",
      municip == "mazatlan" ~ "mazatlán",
      municip == "san_quintin" ~ "san_quintín",
      municip == "nuevo_nayarit" ~ "bahía_de_banderas", # Nuevo Nayarit sits in Bahía de Banderas
      municip == "mexicali" ~ "san_felipe", # same rule as DENUE: Mexicali coastal tourism -> San Felipe
      TRUE ~ municip
    )
  ) %>%
  # 2) sum only after renaming (Los Cabos centers; mexicali+san_felipe; etc.)
  #    key is municipio-year ONLY (not estado)
  group_by(year, municipio_clean) %>%
  summarize(
    arrivals = sum_or_na(arrivals),
    estado = pick_estado(estado),
    .groups = "drop"
  ) %>%
  # drop center-years that never had a usable SECTUR total
  filter(is.finite(arrivals)) %>%
  ## belt-and-suspenders: never keep Pacific Datatur Ensenada / San Quintín
  filter(!municipio_clean %in% c("ensenada", "san_quintín", "san_quintin")) %>%
  # 3) attach a single OHI region from the DENUE lookup (Script 4 re-joins anyway)
  left_join(
    denue_municipios_rgn_fix %>% distinct(municipio_clean, .keep_all = TRUE),
    by = "municipio_clean"
  )

# should be empty after the renames above
setdiff(
  unique(total_arrivals_aggregated_join$municipio_clean),
  unique(employees_by_muni_yr_join_50$municipio_clean)
)

# must be one row per municipio-year (ensenada / san_felipe / bahía_de_banderas included)
total_arrivals_aggregated_join %>%
  count(municipio_clean, year) %>%
  filter(n > 1)

# no -Inf / Inf left
summary(total_arrivals_aggregated_join$arrivals)

unique(total_arrivals_aggregated_join$municipio_clean)
```

## 3.1 Check municipio coverage differences
```{r}
# what the arrivals data has but the employment data does not
setdiff(unique(total_arrivals_aggregated_join$municipio_clean), unique(employees_by_muni_yr_join_50$municipio_clean))
# expect character(0) after san_quintin / nuevo_nayarit / mexicali renames

# what DENUE data has but hotel arrivals data does not
setdiff(unique(employees_by_muni_yr_join_50$municipio_clean), unique(total_arrivals_aggregated_join$municipio_clean))
```

## 3.2 Join arrivals and accommodation employees
```{r}
## Join on municipio + year only (estado strings can disagree between SECTUR and DENUE).
## Also collapse DENUE to one row per municipio-year in case complete() left
## duplicate estado variants for the same municipio.
employees_for_join <- employees_by_muni_yr_join_50 %>%
  group_by(municipio_clean, data_year) %>%
  summarize(
    total_employees = sum(total_employees, na.rm = TRUE),
    num_businesses = sum(num_businesses, na.rm = TRUE),
    avg_employees_per_business = mean(avg_employees_per_business, na.rm = TRUE),
    estado_denue = {
      x <- estado[!is.na(estado) & estado != ""]
      if (length(x) == 0) NA_character_ else x[[1]]
    },
    .groups = "drop"
  )

# safety: one employee row per municipio-year
employees_for_join %>%
  count(municipio_clean, data_year) %>%
  filter(n > 1)

## Training panel: freeze on 2019-2024 complete cases only.
## 2025+ SECTUR is kept for scoring / observed totals elsewhere, but does NOT
## re-estimate gapfill coefficients (assessment reproducibility).
combined_arrivals_employees_data <- total_arrivals_aggregated_join %>%
  left_join(employees_for_join, by = c("municipio_clean", "year" = "data_year")) %>%
  mutate(estado = coalesce(estado, estado_denue)) %>%
  select(-estado_denue) %>%
  filter(year %in% 2019:2024, is.finite(arrivals))

message(
  "OG-model training years: ",
  paste(sort(unique(combined_arrivals_employees_data$year)), collapse = ", ")
)

## Municipios with no complete arrivals+employees rows cannot enter model3.
## If has_arrivals is FALSE, they are not in the GOC SECTUR shapefile panel
## (often dropped by st_intersection) and will use model1 / model1.5 instead.
incomplete_for_model3 <- combined_arrivals_employees_data %>%
  group_by(municipio_clean) %>%
  summarize(
    has_arrivals = sum(!is.na(arrivals)) > 0,
    has_employees = sum(!is.na(total_employees)) > 0,
    n_complete = sum(!is.na(arrivals) & !is.na(total_employees)),
    .groups = "drop"
  ) %>%
  filter(n_complete == 0)

incomplete_for_model3

# confirm whether SECTUR brought them in at all
total_arrivals_aggregated_join %>%
  filter(municipio_clean %in% incomplete_for_model3$municipio_clean)
```


# 4. Visitor and accommodation employees relationship

Models below are fit on the **frozen 2019–2024** training panel (`combined_arrivals_employees_data`), after renaming newer SECTUR centers onto DENUE municipios (`mexicali` to `san_felipe`, `nuevo_nayarit` to `bahía_de_banderas`, `san_quintin` to `san_quintín`) and fixing the arrivals/employees join. SECTUR 2025 is **not** used to estimate coefficients.

**Notes on model comparison**

1. Production fills use **model3.5 / model1.5** (same as the final merge in the rstudio-export script). model1 / model3 are fit for diagnostics and AIC/BIC only.
2. AIC/BIC comments after `#` below are from earlier runs. Printed values here reflect the **2019–2024 freeze**.
3. In model3 / model3.5, most of the fit is municipio fixed effects; employees matter more for never-observed municipios (model1 / model1.5).

**Decision for this OG-model script:** **model3.5** for tourist centers with some SECTUR history; **model1.5** for DENUE-only municipios. Coefficients from **2019–2024 only**; `predict()` onto **2019–2025**. Main Script 3 uses the same model family but refits through available SECTUR years — that is the comparison.

## 4.1 Model 1
```{r}
model1 <- lm(arrivals ~ total_employees, data = combined_arrivals_employees_data)
summary(model1)

library(ggplot2)

ggplot(combined_arrivals_employees_data, aes(x = total_employees, y = arrivals)) +
  geom_point() +
  # geom_abline(slope = 160.088, intercept = 0) +
  # geom_smooth(method = "lm") +
  theme_minimal() +
  labs(
    title = "Total arrivals (domestic + international) vs Total accommodation employees",
    x = "Total employees",
    y = "Total arrivals"
  )

```

## 4.2 Model 1.5
```{r}
# model1.5 <- lm(arrivals ~ total_employees + as.factor(year), data = combined_arrivals_employees_data)
# summary(model1.5)

combined_arrivals_employees_data_wo2020 <- combined_arrivals_employees_data %>% 
  mutate(year_factor = ifelse(year %in% 2020, 1, 2))

## 2020 has more employees than we would expect based on arrivals (large dip in arrivals)

## we need to treat 2020 as something different

model1.5 <- lm(arrivals ~ total_employees + as.factor(year_factor), data = combined_arrivals_employees_data_wo2020)
summary(model1.5)
```


<!-- ### model2 -->

<!-- ```{r} -->
<!-- ## just to check -->
<!-- model2 <- lm(log(arrivals) ~ log(total_employees), data = combined_arrivals_employees_data) -->
<!-- summary(model2) -->
<!-- ``` -->

## 4.3 Model 3
With municipio effects:
```{r}
model3 <- lm(arrivals ~ total_employees + municipio_clean, data = combined_arrivals_employees_data)
summary(model3)
```

## 4.4 Model 3.5
Accounting for 2020:
```{r}
combined_arrivals_employees_data_wo2020 <- combined_arrivals_employees_data %>% 
  mutate(year_factor = ifelse(year %in% 2020, 1, 2))

model3.5 <- lm(arrivals ~ total_employees + municipio_clean + as.factor(year_factor), data = combined_arrivals_employees_data_wo2020)
summary(model3.5)

```

## 4.5 Model 4
```{r}
model4 <- lm(arrivals ~ total_employees + estado, data = combined_arrivals_employees_data)
summary(model4)

```

## 4.6 AIC / BIC comparison

Comments after `#` track older runs (some trained through 2025). Printed values are from the **2019–2024 freeze**.

```{r}
AIC(model1) # old: 1983.7 ; 2019-2025: ~2383.6 ; after renames: ~2597.1
AIC(model3) # old: 1901.3 ; 2019-2025: ~2279.3 ; after renames: ~2473.9
AIC(model4) # old: 1981.3 ; 2019-2025: ~2380.2 ; after renames: ~2580.3
# AIC(model5) # old note only; model5 not used in this script

BIC(model1) # old: 1990.4 ; 2019-2025: ~2390.8 ; after renames: ~2604.6
BIC(model3) # old: 1934.6 ; 2019-2025: ~2317.8 ; after renames: ~2516.2
BIC(model4) # old: 1996.9 ; 2019-2025: ~2397.0 ; after renames: ~2600.2
# BIC(model5) # old note only; model5 not used in this script

BIC(model3) # still higher than model3.5
BIC(model3.5) # old: 1911.9 ; 2019-2025: ~2293.6 ; after renames: ~2486.0 (still best)

BIC(model1)
BIC(model1.5) # old: 1988.6 ; 2019-2025: ~2389.0 ; after renames: ~2602.2 (still better than model1)

AIC(model3.5) # 2019-2025: ~2252.7 ; after renames: ~2441.2
```



# 5. Gapfilling tourist arrivals (OG: model3.5 + model1.5, train 2019–2024 to predict 2019–2025)

Same production decision as the final merge in `rstudio-export/gapfilling_arrivals_by_employees.qmd`:

1. **model3.5** for tourist centers that have gaps in some years but at least some SECTUR history (employees + municipio FE + 2020 year factor).
2. **model1.5** for municipalities with DENUE employees but no SECTUR tourist-center history (employees + 2020 year factor).
3. Fit on **2019–2024** complete cases; apply with `predict()` to missing rows in **2019–2025** only. Observed arrivals stay as-is.

`year_factor` is 1 in 2020 and 2 otherwise. That means **every** filled year uses the .5 model; 2020 simply receives the COVID intercept. (Older prose in the export script sometimes still said “model3 / model1” after the year-factor decision — here the written fill is explicitly model3.5 / model1.5.)

Use `predict()`.

```{r}
## prep the full municipio-year panel for predict() fills
## same rule as 3.2: one row per municipio-year; do not join on estado
## IMPORTANT: rgn_id must come from the DENUE inland_50 lookup for ALL municipios.
## Taking rgn_id only from the SECTUR arrivals side left DENUE-only gapfills as NA
## even when those municipios clearly fall in a region.
rgn_lookup <- denue_municipios_rgn_fix %>%
  distinct(municipio_clean, .keep_all = TRUE) %>%
  select(municipio_clean, rgn_id)

combined_arrivals_employees_data_full <- full_join(
  employees_for_join,
  total_arrivals_aggregated_join %>%
    select(municipio_clean, year, arrivals, estado_sectur = estado),
  by = c("municipio_clean", "data_year" = "year")
) %>%
  mutate(estado = coalesce(estado_sectur, estado_denue)) %>%
  select(-estado_denue, -estado_sectur) %>%
  left_join(rgn_lookup, by = "municipio_clean")

# final safety check before fill/plot
combined_arrivals_employees_data_full %>%
  count(municipio_clean, data_year) %>%
  filter(n > 1)

## who still lacks a region?
combined_arrivals_employees_data_full %>%
  distinct(municipio_clean, rgn_id) %>%
  filter(is.na(rgn_id) | rgn_id == "") %>%
  arrange(municipio_clean)
```

## 5.1 Diagnostic plots / bptest

```{r}
plot(model1, which = 1) # residuals vs fitted
plot(model1, which = 2) # QQ plot
plot(model1, which = 3) # scale - location

library(lmtest)
bptest(model1)
bptest(model3)
bptest(model3.5)
bptest(model1.5)
bptest(model4)
```

## 5.2 Prep for fill (model3.5 / model1.5 need year_factor)
```{r}
## IMPORTANT: use the factor levels that were actually fit in model3 / model3.5,
## not unique(municipio_clean) from the join. Municipios with arrivals but no
## complete arrivals+employees training rows must go to model1.5.
munis_model3 <- model3$xlevels$municipio_clean

setdiff(unique(combined_arrivals_employees_data$municipio_clean), munis_model3)
# municipios in the SECTUR join but not in model3 -> filled with model1.5

## year_factor: 1 = 2020 (COVID), 2 = all other years (including 2025 fills)
df_model3 <- combined_arrivals_employees_data_full %>%
  filter(municipio_clean %in% munis_model3) %>%
  mutate(year_factor = ifelse(data_year == 2020, 1, 2))

df_model1 <- combined_arrivals_employees_data_full %>%
  filter(!municipio_clean %in% munis_model3) %>%
  mutate(year_factor = ifelse(data_year == 2020, 1, 2))
```

## 5.3 Use predict() to fill (2019–2025 only)

### 5.3.1 Model 3.5 (SECTUR-history municipios)
Also store model3 predictions for diagnostics; **production merge uses model3.5**.

```{r}
df_model3_yr <- df_model3 %>%
  filter(data_year %in% 2019:2025)

df_model3_filled <- df_model3_yr %>%
  mutate(
    arrivals_filled_model3 = ifelse(
      is.na(arrivals) & !is.na(total_employees),
      predict(model3, newdata = df_model3_yr),
      arrivals
    ),
    arrivals_filled_model3.5 = ifelse(
      is.na(arrivals) & !is.na(total_employees),
      predict(model3.5, newdata = df_model3_yr),
      arrivals
    )
  )
```

### 5.3.2 Model 1.5 (DENUE-only municipios)
Also store model1 predictions for diagnostics; **production merge uses model1.5**.

```{r}
df_model1_yr <- df_model1 %>%
  filter(data_year %in% 2019:2025)

df_model1_filled <- df_model1_yr %>%
  mutate(
    arrivals_filled_model1 = ifelse(
      is.na(arrivals) & !is.na(total_employees),
      predict(model1, newdata = df_model1_yr),
      arrivals
    ),
    arrivals_filled_model1.5 = ifelse(
      is.na(arrivals) & !is.na(total_employees),
      predict(model1.5, newdata = df_model1_yr),
      arrivals
    )
  )
```

### 5.3.3 Combine and merge (production = model3.5 then model1.5)
```{r}
filled_combined <- bind_rows(df_model3_filled, df_model1_filled) %>%
  mutate(is_filled = ifelse(is.na(arrivals), "1", "0"))

## same rule as rstudio-export final merge; names match the models actually used
arrivals_employees_filled <- filled_combined %>%
  mutate(arrivals_merged = case_when(
    is.na(arrivals) & !is.na(arrivals_filled_model3.5) ~ arrivals_filled_model3.5,
    is.na(arrivals) & is.na(arrivals_filled_model3.5) & !is.na(arrivals_filled_model1.5) ~ arrivals_filled_model1.5,
    TRUE ~ arrivals
  ))

## confirm fill years stop at 2025
stopifnot(max(arrivals_employees_filled$data_year, na.rm = TRUE) <= 2025)
```

### 5.3.4 Censor negative predictions at zero
For small municipalities, predicted arrivals can dip below zero. That cannot happen conceptually, so negatives are set to zero after predicting (same as the original script).

```{r}
arrivals_employees_filled_negfix <- arrivals_employees_filled %>% 
  mutate(arrivals_merged_no_neg = ifelse(arrivals_merged < 0, 0, arrivals_merged))
```


## 5.5 Plotly checks
```{r}
library(plotly)

plot_data <- arrivals_employees_filled %>% 
  arrange(municipio_clean, data_year)

plot_data_plotly <- plot_ly(
  data = plot_data,
  x = ~data_year,
  y = ~arrivals_merged,
  color = ~municipio_clean,
  type = 'scatter',
  mode = 'lines+markers'
) %>%
  layout(
    title = "Total arrivals after gapfilling with model3.5 + model1.5 (OG freeze 2019–2024 to predict 2019–2025)",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Arrivals")
  )
plot_data_plotly
```


```{r}
#### after fixing negative predicted employees
plot_data_noneg <- arrivals_employees_filled_negfix %>% 
  arrange(municipio_clean, data_year)

plot_data_noneg_plotly <- plot_ly(
  data = plot_data_noneg,
  x = ~data_year,
  y = ~arrivals_merged_no_neg,
  color = ~municipio_clean,
  type = 'scatter',
  mode = 'lines+markers'
) %>%
  layout(
    title = "Total arrivals after gapfilling with model3.5 + model1.5 (OG) + negatives censored at zero",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Arrivals")
  )
plot_data_noneg_plotly

```



### 5.5.1 Diagnostic checks
```{r}
## first, check that there are no negative predictions
arrivals_employees_filled %>%
  filter(arrivals_merged < 0) ## none!!
```


```{r}
# ggplot(df_model3_filled, aes(x = arrivals, y = arrivals_filled_model3)) +
#   geom_point(alpha = 0.6) +
#   geom_abline(slope = 1, intercept = 0, color = "red") +
#   theme_minimal() +
#   labs(
#     title = "Model3 Diagnostic: Actual vs Predicted Arrivals",
#     x = "Actual arrivals",
#     y = "Predicted arrivals (model3)"
#   )
```

### 5.5.2 Histogram: model 3.5 filled
```{r}
hist(arrivals_employees_filled_negfix$arrivals_filled_model3.5, breaks = 30,
     main = "Distribution of filled arrivals",
     xlab = "arrivals")
```

### 5.5.3 Histogram: model 1.5 filled
```{r}
hist(arrivals_employees_filled_negfix$arrivals_filled_model1.5, breaks = 30,
     main = "Distribution of filled arrivals",
     xlab = "arrivals")
```


```{r}
df_compare <- arrivals_employees_filled %>%
  mutate(source = ifelse(is.na(arrivals), "predicted", "actual"))

ggplot(df_compare, aes(x = arrivals_merged, fill = source)) +
  geom_histogram(alpha = 0.5, position = "identity") +
  theme_minimal()
```

```{r}
# ggplot(arrivals_employees_filled,
#        aes(x = data_year, y = arrivals_merged, color = is_filled)) +
#   geom_line() +
#   facet_wrap(~ municipio_clean, scales = "free_y") +
#   theme_minimal() +
#   labs(title = "Per-Municipio Time Series (Actual vs Filled)")

# df_model3_filled %>%
#   filter(!is.na(arrivals)) %>%
#   summarize(
#     RMSE = sqrt(mean((arrivals - predict(model3, newdata = .))^2)), # a lower RMSE indicates a better fit and more accurate predictions
#     MAE  = mean(abs(arrivals - predict(model3, newdata = .)))
#   )
# 
# df_model1_filled %>%
#   filter(!is.na(arrivals)) %>%
#   summarize(
#     RMSE = sqrt(mean((arrivals - predict(model1, newdata = .))^2)),
#     MAE  = mean(abs(arrivals - predict(model1, newdata = .)))
#   )
```

### 5.5.4 Arrivals vs employees
```{r}
ggplot(arrivals_employees_filled_negfix,
       aes(x = total_employees, y = arrivals_merged, color = is_filled)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_minimal()

ggplot(arrivals_employees_filled_negfix,
       aes(x = total_employees, y = arrivals_merged_no_neg, color = is_filled)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_minimal()
```

# 6. Write gapfilled arrivals for Script 4
```{r}
## confirm regions are populated before writing
arrivals_employees_filled_negfix %>%
  distinct(municipio_clean, rgn_id) %>%
  count(missing_rgn = is.na(rgn_id) | rgn_id == "")

agg_merged <- arrivals_employees_filled_negfix %>%
  distinct(municipio_clean, rgn_id) %>%
  count(missing_rgn = is.na(rgn_id) | rgn_id == "")

write_csv(arrivals_employees_filled, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/gapfilled_int","arrivals_filled_by_DENUE_accommodation_employees_2019_2026_OGmodel.csv"))

write_csv(arrivals_employees_filled_negfix, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/gapfilled_int","arrivals_filled_by_DENUE_accommodation_employees_2019_2026_OGmodel_UPDATED.csv"))

arrivals_employees_filled_negfix <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/gapfilled_int","arrivals_filled_by_DENUE_accommodation_employees_2019_2026_OGmodel_UPDATED.csv"))

# arrivals_employees_filled <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/gapfilled_int","arrivals_filled_by_DENUE_accommodation_employees_2019_2026_OGmodel.csv"))
```


# 7. Note on covid-scaled sensitivity

Not included in this OG-model parallel. The covid-scaled DENUE-only 2020 fill lives only in main `script3_gapfilling_arrivals_by_employees.qmd` (Section 7). This script’s purpose is a **frozen** model3.5 + model1.5 comparison (train 2019–2024, predict 2019–2025) against the refitting main Script 3.
