---
title: "Script 4: TR quantity scores (with gapfilled arrivals)"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
date: "2026-7-30"
editor_options:
  chunk_output_type: console
---

# 0. Workflow for this script

This is **Script 4** in the TR quantity update pipeline. Run it after Scripts 1-3.

1. Read cleaned SECTUR yearly arrivals from Script 2.
2. Join with DENUE-gapfilled totals from Script 3.
3. Estimate international vs domestic arrivals where SECTUR is missing.
4. Keep only municipios whose representative point lies in `inland_50km_goc` (same coastal mask as DENUE / Script 2), assign OHI regions, and compute current status scores (2024 baseline targets).
5. Make regional plots and write final quantity outputs.
6. Optional Section 7: re-score using Script 3’s covid-scaled gapfill file for comparison.

Baseline targets stay the same: domestic = 2024 arrivals x 1.098; international = 2024 arrivals x 1.30.

# 1. Combine SECTUR arrivals with DENUE-gapfilled totals (2019-2026)
In the script `script3_script3_gapfilling_arrivals_by_employees.qmd`, I predicted missing tourist arrival data using a model that used the number of employees in hotels/accommodations to predict the total number of arrivals.  To apply it here, we will need to determine the proportion of arrivals that are international vs domestic out of the total in each state.  That way, we can apply it to the regions that have missing data.

Essentially, we will determine for each state the proportion of domestic/total and international/total and just assume that same proportion for the regions we are estimating using the regression models.

I am now aligning this with the updated DENUE series (**2019-2026**). For SECTUR annual arrivals I use **2019-2025** only, because 2026 is still incomplete (Enero-Junio only as of 07/30/26).

```{r}
## first, we will read in the cleaned sectur data, and only select arrivals.  We want to find the total number of arrivals, int + dom, for every year and municipio.
all_sectur_arrivals <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/hotel_monitoring_int","hotel_data_cleaned_yearly_2013_2025.csv")) %>% 
  filter(metric %in% c("llegadas_de_turistas_residentes", "llegadas_de_turistas_no_residentes")) %>% # domestic and international arrivals
  filter(category %in% "total_categorias")  %>% # since we do not care what star designation it is
  filter(year %in% 2019:2025) %>% 
  mutate(arrival_type = case_when(
    metric == "llegadas_de_turistas_residentes" ~ "dom",
    metric == "llegadas_de_turistas_no_residentes" ~ "int",
    TRUE ~ NA
  ))

# summarize to find the total number of arrivals for every year and municipio
all_sectur_arrivals_sum <- all_sectur_arrivals %>%
  group_by(municipio, year) %>% 
  summarize(total_arrivals = sum(value))

# obtain the proportion of the total for each value.  Double check that for the same municipio, the proportion of the dom + the proportion of the int = 1. Then, assign them each to states.
all_sectur_arrivals_w_prop <- left_join(all_sectur_arrivals, all_sectur_arrivals_sum, by = c("municipio","year")) %>%
  mutate(prop_of_total = value/total_arrivals) %>% 
  mutate(country = "Mexico") %>%  
  mutate(estado = case_when(
    municipio %in% c("cabo_san_lucas", "la_paz", "loreto", "san_jose_del_cabo", "zona_corredor_los_cabos") ~ "baja_california_sur",
    municipio %in% c("san_felipe", "ensenada", "mexicali", "san_quintin") ~ "baja_california",
    municipio %in% c("guaymas", "hermosillo") ~ "sonora",
    municipio %in% c("culiacan", "el_fuerte", "los_mochis", "mazatlan", "escuinapa", "el_rosario") ~ "sinaloa",
    municipio %in% c("nuevo_nayarit", "tepic") ~ "nayarit",
    municipio %in% c("puerto_vallarta") ~ "jalisco",
    TRUE ~ ""  # default case for any unmatched cities
  ))  
  
```

For now, we will average the proportion across the years 2019:2025 for every state. This is because in some states, notably Sonora, we have very little information on the numbers of arrivals throughout all years.  So we cannot obtain a number for the proportion of dom vs. int for every year for each state. 

Note that Nayarit has little/no usable SECTUR coverage in places, so I will assume it has the same proportions as Jalisco when needed.  

I will need to check in with Mel on all of this!! -- checked, this is the best we can do (120925).

```{r}
# summarize the average proportion of int and dom for every state.
all_sectur_prop_by_state <- all_sectur_arrivals_w_prop %>% 
  group_by(arrival_type, estado) %>% 
  summarize(avg_prop = mean(prop_of_total, na.rm = TRUE))

## save it!!
write_csv(all_sectur_prop_by_state, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/gapfilled_int","sectur_avg_prop_by_state_2019_2025.csv"))
```

## 1.1 Read gapfilled arrivals from Script 3
```{r}
arrivals_filled_by_DENUE_accommodation_employees <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/gapfilled_int","arrivals_filled_by_DENUE_accommodation_employees_2019_2026_UPDATED.csv"))

gapfilled_total_arrivals <- arrivals_filled_by_DENUE_accommodation_employees %>%
  mutate(country = "Mexico") %>% 
  select(year = data_year, country, estado, municipio_clean, rgn_id, arrivals_merged = arrivals_merged_no_neg, is_filled) 

```

```{r}
## prep all sectur arrivals for a join with the gapfilled totals
all_sectur_arrivals_prep_join <- all_sectur_arrivals %>% 
  mutate(
    municipio_clean = case_when(
      municipio %in% c("cabo_san_lucas", "san_jose_del_cabo", "zona_corredor_los_cabos") ~ "los_cabos",
      municipio == "los_mochis" ~ "ahome", # los mochis is a city within the municipality of ahome!!
      municipio == "el_rosario" ~ "rosario",
      municipio == "culiacan" ~ "culiacán",
      municipio == "mazatlan" ~ "mazatlán",
      municipio == "nuevo_nayarit" ~ "bahía_de_banderas", # Nuevo Nayarit is located inside the Municipality of Bahía de Banderas
      municipio == "san_quintin" ~ "san_quintín",
      TRUE ~ municipio
    )
  ) %>% 
  # reaggregate with changes
  group_by(year, municipio_clean, arrival_type) %>%
  summarize(value = sum(value), .groups = "drop") %>% 
  select(year, municipio_clean, arrival_type, raw_arrivals = value) %>% # leave in "municipio" to double check at first
  filter(!municipio_clean %in% "mexicali")

## check!!
setdiff(unique(all_sectur_arrivals_prep_join$municipio_clean), unique(gapfilled_total_arrivals$municipio_clean)) # character(0)
setdiff(unique(gapfilled_total_arrivals$municipio_clean), unique(all_sectur_arrivals_prep_join$municipio_clean)) 

## makes sense, those were all the municipalities that were gapfilled using DENUE accommodation employees!!
```


```{r}
## municipalities missing in SECTUR arrivals
gapfilled_municipios <- setdiff(unique(gapfilled_total_arrivals$municipio_clean), unique(all_sectur_arrivals_prep_join$municipio_clean))

## expand into dataframe with years 2019-2026 and arrival types int/dom
## (SECTUR observed values stop at 2025; 2026 rows will be DENUE-gapfilled only)
gapfilled_municipios_df <- expand.grid(
  municipio_clean = gapfilled_municipios,
  year = 2019:2026,
  arrival_type = c("int", "dom"),
  KEEP.OUT.ATTRS = FALSE, # a logical indicating the "out.attrs" attribute should be computed and returned.
  stringsAsFactors = FALSE # logical specifying if character vectors are converted to factors.
)

## join with the prepped sectur arrivals df
gapfilled_municipios_and_sectur_arrivals_prep_join <- full_join(all_sectur_arrivals_prep_join, gapfilled_municipios_df, by = c("municipio_clean","year","arrival_type"))

```

Join the gapfilled totals with the raw int and dom values for arrivals:

```{r}
joined_gapfilled_and_raw_arrivals <- full_join(gapfilled_total_arrivals, gapfilled_municipios_and_sectur_arrivals_prep_join, by = c("year","municipio_clean"), relationship = "many-to-many")

## check that the raw arrivals for int + dom add up to the "arrivals_merged" when "is_filled" = 0 (it was not predicted using DENUE employees for accommodations)
## looks good!
```

Now, join this data frame with the avg proportions by state.

```{r}
all_sectur_prop_by_state <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/gapfilled_int","sectur_avg_prop_by_state_2019_2025.csv"))

joined_gapfilled_arrivals_and_prop <- left_join(joined_gapfilled_and_raw_arrivals, all_sectur_prop_by_state, by = c("estado", "arrival_type")) %>% 
  mutate(avg_prop = case_when(
    estado == "nayarit" & arrival_type == "dom" ~ 0.56415674,
    estado == "nayarit" & arrival_type == "int" ~ 0.43584326,
    TRUE ~ avg_prop
  )) # assume for now that nayarit has the save avg proportion as jalisco, though this is definitely not true.
```

Now we can actually estimate the number of international arrivals and domestic arrivals for all predicted total arrivals from DENUE data.

```{r}
estimated_int_dom_arrivals <- joined_gapfilled_arrivals_and_prop %>% 
  mutate(estimated_arrivals = ifelse(is_filled == "1", avg_prop*arrivals_merged, raw_arrivals)) # if it was gapfilled, then estimate the number of int or dom arrivals by multiplying the average proportion of int or dom arrivals by state and the predicted total arrivals from the DENUE data. If they were not gapfilled, use the origianl int and dom arrival values.
```


# 2. Plot estimated arrivals over time
```{r}
library(plotly)

### INT Arrivals over time ###
estimated_int_arrivals <- estimated_int_dom_arrivals %>% 
  filter(arrival_type %in% "int") %>% 
  arrange(year)


int_arrivals_plotly <- plotly::plot_ly(
  estimated_int_arrivals, 
  x = ~year, 
  y = ~estimated_arrivals, 
  color = ~municipio_clean, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "International arrivals by year and municipio",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Number of arrivals")
  )
int_arrivals_plotly
```


```{r}
### DOM Arrivals ###
estimated_dom_arrivals <- estimated_int_dom_arrivals %>% 
  filter(arrival_type %in% "dom") %>% 
  arrange(year)

dom_arrivals_plotly <- plotly::plot_ly(
  estimated_dom_arrivals, 
  x = ~year, 
  y = ~estimated_arrivals, 
  color = ~municipio_clean, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "Domestic arrivals by year and municipio",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Number of arrivals")
  )
dom_arrivals_plotly

```


# 3. Current status by municipio (optional / commented)
```{r}
# baseline_year <- 2024
# 
# quantity_current_status <- estimated_int_dom_arrivals %>% 
#   mutate(year = as.numeric(year)) %>%
#   
#   # group by municipio + arrival type
#   group_by(municipio_clean, arrival_type) %>%
#   
#   mutate(
#     # baseline per municipio + type
#     baseline_arrivals = estimated_arrivals[year == baseline_year][1],
# 
#     # target multiplier varies by type
#     target_multiplier = case_when(
#       arrival_type == "dom" ~ 1.098, # domestic +9.8%
#       arrival_type == "int" ~ 1.30,  # international +30%
#       TRUE ~ NA_real_
#     ),
# 
#     # calculate targets
#     target_arrivals = baseline_arrivals * target_multiplier,
# 
#     # CURRENT STATUS (0 to 100)
#     current_status = (estimated_arrivals / target_arrivals) * 100,
# 
#     # absolute and % gap
#     gap_to_target = target_arrivals - estimated_arrivals,
#     gap_percentage = ((target_arrivals - estimated_arrivals) / estimated_arrivals) * 100
#   ) 
# 
# ## save this to outputs
# write_csv(quantity_current_status, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quantity","quantity_current_status_120125.csv"))
```

## 3.1 Municipio plotlys (optional / commented)
International:

```{r}
# library(plotly)
# 
# int_quantity_current_status <- quantity_current_status %>% 
#   filter(arrival_type %in% "int") %>% 
#   arrange(year)
# 
# ## plot the score over time
# final_int_score_plotly <- plotly::plot_ly(
#   int_quantity_current_status, 
#   x = ~year, 
#   y = ~current_status, 
#   color = ~municipio_clean, 
#   type = "scatter", 
#   mode = "lines"
# ) %>% 
#   layout(
#     title = "CURRENT STATUS: Llegadas de turistas internacionales por municipio",
#     xaxis = list(title = "Año // Year"),
#     yaxis = list(title = "Puntaje // Percent (out of 100)")
#   ) %>%
#   add_lines(
#     x = range(int_quantity_current_status$year), 
#     y = c(100, 100),
#     line = list(color = "black", width = 2, dash = "dash"),
#     name = "Meta // Goal 2030 (100/100)",
#     inherit = FALSE
#   )
# final_int_score_plotly
# 
# #htmlwidgets::saveWidget(final_int_score_plotly, here::here("TR/sectur_quantity_exploration_int","int_current_status_gapfilled_by_municipio_120125.html"), selfcontained = TRUE)
```

Domestic:

```{r}
# library(plotly)
# 
# dom_quantity_current_status <- quantity_current_status %>% 
#   filter(arrival_type %in% "dom") %>% 
#   arrange(year)
# 
# ## plot the score over time
# final_dom_score_plotly <- plotly::plot_ly(
#   dom_quantity_current_status, 
#   x = ~year, 
#   y = ~current_status, 
#   color = ~municipio_clean, 
#   type = "scatter", 
#   mode = "lines"
# ) %>% 
#   layout(
#     title = "CURRENT STATUS: Llegadas de turistas nacionales por municipio",
#     xaxis = list(title = "Año // Year"),
#     yaxis = list(title = "Puntaje // Percent (out of 100)")
#   ) %>%
#   add_lines(
#     x = range(dom_quantity_current_status$year), 
#     y = c(100, 100),
#     line = list(color = "black", width = 2, dash = "dash"),
#     name = "Meta // Goal 2030 (100/100)",
#     inherit = FALSE
#   )
# final_dom_score_plotly
# 
# #htmlwidgets::saveWidget(final_dom_score_plotly, here::here("TR/sectur_quantity_exploration_int","dom_current_status_gapfilled_by_municipio_120125.html"), selfcontained = TRUE)
```

# 4. Current status by OHI region
Somewhere along the way we lost the lat/longs for each municipio.  Lets add them back in and join with ohi regions, so that we can attribute each municipality to an OHI Region.

```{r}
estimated_int_dom_arrivals_latlong <- estimated_int_dom_arrivals %>% 
  tidygeocoder::geocode(city = municipio_clean, country = country, state = estado) %>%  # to give each city a lat, long
  mutate(coordinate_na = is.na(lat))

na_municipios_latlong <- estimated_int_dom_arrivals_latlong %>% 
  filter(coordinate_na == "TRUE") %>%
  distinct(municipio_clean)

# ahome
# 2
# bahía_de_banderas
# 3
# rosario
# 4
# cabo_corrientes
# 5
# mascota
# 6
# san_sebastián_del_oeste
# 7
# talpa_de_allende
# 8
# tomatlán
  
  # manually found:
  ## tomatlán = 19.938547, -105.249363
  ## talpa_de_allende = 20.382398, -104.820536
  ## san_sebastián_del_oeste = 20.762123, -104.854319
  ## san_quintín = 23.709045, -110.396823
  ## san_pedro_lagunillas = 21.218043, -104.753181
  ## mascota = 20.527390, -104.792100
  ## ensenada = 28.443258, -112.880537 (if you look at employees_by_muni_yr_join_50, it is interchanged with san quintín for some of the hotels, because ensenada as a municipality stretches far along baja california on the interior gulf.)
  ## cabo_corrientes = 20.392035, -105.453138
  ## bahía_de_banderas = 20.759572, -105.292800

estimated_int_dom_arrivals_latlong_clean <- estimated_int_dom_arrivals_latlong %>%   
  mutate(
    lat = case_when(
      municipio_clean == "talpa_de_allende" ~ 20.382398,
      municipio_clean == "tomatlán" ~ 19.938547,
      municipio_clean == "san_sebastián_del_oeste" ~ 20.762123,
      municipio_clean == "san_quintín" ~ 23.709045,
      municipio_clean == "san_pedro_lagunillas" ~ 21.218043,
      municipio_clean == "mascota" ~ 20.527390,
      municipio_clean == "ensenada" ~ 28.443258,
      municipio_clean == "cabo_corrientes" ~ 20.392035,
      municipio_clean == "bahía_de_banderas" ~ 20.759572,
      municipio_clean == "ahome" ~ 25.918973,
      municipio_clean == "rosario" ~ 27.84484348043768,
      municipio_clean == "mascota" ~ 20.526449771988734,
      TRUE ~ lat 
    ),
    long = case_when(
      municipio_clean == "talpa_de_allende" ~ -104.820536,
      municipio_clean == "tomatlán" ~ -105.249363,
      municipio_clean == "san_sebastián_del_oeste" ~ -104.854319,
      municipio_clean == "san_quintín" ~ -110.396823,
      municipio_clean == "san_pedro_lagunillas" ~ -104.753181,
      municipio_clean == "mascota" ~ -104.792100,
      municipio_clean == "ensenada" ~ -112.880537,
      municipio_clean == "cabo_corrientes" ~ -105.453138,
      municipio_clean == "bahía_de_banderas" ~ -105.292800,
      municipio_clean == "ahome" ~ -109.174074,
      municipio_clean == "rosario" ~ -109.37340921304606,
      municipio_clean == "mascota" ~ -104.79243186410368,
      TRUE ~ long
    )
  )


estimated_int_dom_arrivals_sf <- estimated_int_dom_arrivals_latlong_clean %>% 
  st_as_sf(coords = c("long", "lat"),
           crs = 4326, # WGS84
           remove = FALSE) %>% 
  select(-rgn_id) # because many were missing and it is easier to just intersect with ohi rgns.

## Same rule as Scripts 1-2: only municipios whose point lies in inland_50km_goc
inland_50 <- st_read(here::here(
  "/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer",
  "inland_50km_goc.shp"
))
estimated_int_dom_arrivals_sf <- st_transform(estimated_int_dom_arrivals_sf, st_crs(inland_50))

munis_pts <- estimated_int_dom_arrivals_sf %>%
  distinct(municipio_clean, .keep_all = TRUE)
in_50_flag <- lengths(st_intersects(munis_pts, inland_50)) > 0
munis_in_50 <- munis_pts$municipio_clean[in_50_flag]
munis_out_50 <- munis_pts$municipio_clean[!in_50_flag]
message("Script 4 municipios KEPT (within inland_50km): ", paste(sort(munis_in_50), collapse = ", "))
message("Script 4 municipios DROPPED (outside inland_50km): ", paste(sort(munis_out_50), collapse = ", "))

estimated_int_dom_arrivals_sf <- estimated_int_dom_arrivals_sf %>%
  filter(municipio_clean %in% munis_in_50)
```

Now, assign each municipio to an OHI region from its coordinates (reproducible spatial join — not a manual table).

```{r}
goc_poly <- st_read(here("spatial/ohi_regions", "goc_ohi_rgns.shp")) %>%
  select(rgn_id) %>%
  st_make_valid()

goc_poly_t <- st_transform(goc_poly, st_crs(estimated_int_dom_arrivals_sf))

## one point per municipio for assignment, then join back to all year/type rows
muni_pts_score <- estimated_int_dom_arrivals_sf %>%
  distinct(municipio_clean, .keep_all = TRUE)

nn <- st_nearest_feature(muni_pts_score, goc_poly_t)
rgn_spatial <- muni_pts_score %>%
  st_drop_geometry() %>%
  transmute(municipio_clean, rgn_id = goc_poly_t$rgn_id[nn])

## any still missing? should be none after inland_50 filter
rgn_spatial %>% filter(is.na(rgn_id) | rgn_id == "")

## also refresh the lookup CSV so Script 3 / other readers stay in sync
write_csv(
  rgn_spatial %>% arrange(municipio_clean),
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/hotel_monitoring_int",
    "ohi_rgn_assign_for_municipio_clean.csv"
  )
)

estimated_int_dom_arrivals_rgn <- estimated_int_dom_arrivals_sf %>%
  select(-any_of("rgn_id")) %>%
  left_join(rgn_spatial, by = "municipio_clean")

estimated_int_dom_arrivals_rgn %>%
  st_drop_geometry() %>%
  distinct(municipio_clean, rgn_id) %>%
  count(missing = is.na(rgn_id))

mapview(
  muni_pts_score %>% mutate(rgn_id = goc_poly_t$rgn_id[nn]),
  zcol = "rgn_id"
) + mapview(goc_poly_t, alpha.regions = 0.2)
```

Let's aggregate all tourism arrivals by region and type.

```{r}
estimated_int_dom_arrivals_ohirgn <- estimated_int_dom_arrivals_rgn %>% 
  group_by(rgn_id, arrival_type, year) %>% 
  summarize(agg_tourist_arrivals = sum(estimated_arrivals, na.rm = TRUE)) %>% 
  filter(year %in% 2019:2025)
```

## 4.1 Domestic trend visualizations

<!-- ```{r} -->
<!-- last_3yrs <- c(2023, 2024, 2025) -->

<!-- trend_last3_slope <- estimated_int_dom_arrivals_ohirgn %>% -->
<!--   st_drop_geometry() %>%  -->
<!--   filter(year %in% last_3yrs) %>% -->
<!--   group_by(rgn_id, arrival_type) %>% -->
<!--   summarise( -->
<!--     slope_per_year = coef(lm(agg_tourist_arrivals ~ year))[2], -->
<!--     .groups = "drop" -->
<!--   ) -->

<!-- trend_last3_pct <- estimated_int_dom_arrivals_ohirgn %>% -->
<!--   st_drop_geometry() %>%  -->
<!--   filter(year %in% last_3yrs) %>% -->
<!--   group_by(rgn_id, arrival_type) %>% -->
<!--   summarise( -->
<!--     arrivals_2023 = agg_tourist_arrivals[year == 2023], -->
<!--     arrivals_2025 = agg_tourist_arrivals[year == 2025], -->
<!--     pct_change_3yr = (arrivals_2025 - arrivals_2023) / arrivals_2023 * 100, -->
<!--     .groups = "drop" -->
<!--   )  -->

<!-- dom_trend_last3yrs <- full_join(trend_last3_pct, trend_last3_slope) %>%  -->
<!--   filter(arrival_type == "dom") %>%  -->
<!--   mutate(target_arrivals = arrivals_2025 * 1.098, # illustrative vs latest complete year; official score baseline remains 2024 elsewhere -->
<!--          proj_arrivals_2030 = arrivals_2025 + (5 * slope_per_year), # 2025 -> 2030 = 5 years -->
<!--          on_track = ifelse(proj_arrivals_2030 >= target_arrivals, 1, 0)) -->
<!-- write_csv(dom_trend_last3yrs, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_viz","dom_trend_last3yrs_arrivals_2023_2025.csv")) -->

<!-- #### join with ohi rgn polygons -->
<!-- ohi <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp")) %>% -->
<!--   filter(location == "marine") %>%  -->
<!--   select(rgn_id) -->

<!-- dom_trend_last3yrs_ohigns <- full_join(dom_trend_last3yrs, ohi, by = "rgn_id") %>%  -->
<!--   st_as_sf() -->

<!-- mapview(dom_trend_last3yrs_ohigns,  -->
<!--         zcol = "pct_change_3yr", -->
<!--         col.regions = colorRampPalette(c('#8a0032', '#A50026', '#A50026', '#D73027', '#F46D43', '#FDAE61', '#FEE090', '#FEE090', '#ABD9E9', '#4575B4', '#313695')), -->
<!--         layer.name = "Percent change of domestic arrivals (2023-2025)", -->
<!--         legend = TRUE) -->

<!-- ######################################## -->

<!-- ### across the gulf: -->
<!-- trend_overall <- estimated_int_dom_arrivals_ohirgn %>% -->
<!--   st_drop_geometry() %>%  -->
<!--   filter(year %in% last_3yrs) %>% -->
<!--   group_by(arrival_type, year) %>% -->
<!--   summarise( -->
<!--     total_arrivals = sum(agg_tourist_arrivals, na.rm = TRUE), -->
<!--     .groups = "drop" -->
<!--   ) -->

<!-- trend_overall_slope <- trend_overall %>% -->
<!--   group_by(arrival_type) %>% -->
<!--   summarise( -->
<!--     slope_per_year = coef(lm(total_arrivals ~ year))[2] -->
<!--   ) -->

<!-- trend_overall_pct_change <- trend_overall %>% -->
<!--   group_by(arrival_type) %>% -->
<!--   summarise( -->
<!--     pct_change_3yr = -->
<!--       (total_arrivals[year == 2025] - total_arrivals[year == 2023]) / -->
<!--       total_arrivals[year == 2023] * 100 -->
<!--   ) -->

<!-- dom_trend_goc <- full_join(trend_overall_slope, trend_overall_pct_change) %>%  -->
<!--   filter(arrival_type == "dom") -->
<!-- write_csv(dom_trend_goc, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_viz","dom_trend_goc_arrivals_2023_2025.csv")) -->
<!-- ``` -->


<!-- ```{r} -->
<!-- #### for visualizations: see the proportion of int vs dom for each rgn -->
<!-- prop_dom_by_rgn <- estimated_int_dom_arrivals_ohirgn %>%  -->
<!--   group_by(rgn_id, year) %>%  -->
<!--   mutate(rgn_agg_arr = sum(agg_tourist_arrivals)) %>%  -->
<!--   ungroup() %>%  -->
<!--   mutate(prop_rgn_total = agg_tourist_arrivals / rgn_agg_arr) %>%  -->
<!--   filter(year == 2025) %>%  -->
<!--   filter(arrival_type == "dom") %>%  -->
<!--   st_drop_geometry() -->
<!-- write_csv(prop_dom_by_rgn, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_viz","prop_dom_by_rgn_arrivals_2025.csv")) -->

<!-- ### join with ohi rgn polygons -->
<!-- ohi <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp")) %>% -->
<!--   filter(location == "marine") %>%  -->
<!--   select(rgn_id) -->

<!-- prop_dom_arrivals_ohigns <- full_join(prop_dom_by_rgn, ohi, by = "rgn_id") %>%  -->
<!--   st_as_sf() -->

<!-- mapview(prop_dom_arrivals_ohigns,  -->
<!--         zcol = "prop_state_total", -->
<!--         col.regions = colorRampPalette(c("#096b91", "#a057ce")), -->
<!--         layer.name = "Proportion of Regionwide Domestic Arrivals (2025)", -->
<!--         legend = TRUE) -->

<!-- #### prop of total goc -->
<!-- prop_dom_all <- estimated_int_dom_arrivals_ohirgn %>%  -->
<!--   filter(year == 2025) %>%  -->
<!--   st_drop_geometry() %>%  -->
<!--   group_by(year) %>%  -->
<!--   mutate(all_agg_arr = sum(agg_tourist_arrivals)) %>%  -->
<!--   ungroup() %>%  -->
<!--   group_by(arrival_type, year) %>%  -->
<!--   mutate(total_type_by_yr = sum(agg_tourist_arrivals)) %>%  -->
<!--   ungroup() %>%  -->
<!--   select(-c(rgn_id, agg_tourist_arrivals)) %>%  -->
<!--   distinct() %>%  -->
<!--   mutate(prop_dom_total = total_type_by_yr / all_agg_arr) %>%  -->
<!--   filter(arrival_type == "dom") -->
<!-- ``` -->

## 4.2 Calculate CS
```{r}
baseline_year <- 2024

quantity_current_status_rgn <- estimated_int_dom_arrivals_ohirgn %>% 
  mutate(year = as.numeric(year)) %>%
  
  # group by ohi_rgn + arrival type
  group_by(rgn_id, arrival_type) %>%
  
  mutate(
    # baseline per municipio + type
    baseline_arrivals = agg_tourist_arrivals[year == baseline_year][1],

    # target multiplier varies by type
    target_multiplier = case_when(
      arrival_type == "dom" ~ 1.098, # domestic +9.8%
      arrival_type == "int" ~ 1.30,  # international +30%
      TRUE ~ NA_real_
    ),

    # calculate targets
    target_arrivals = baseline_arrivals * target_multiplier,

    # CURRENT STATUS (0 to 100)
    current_status = (agg_tourist_arrivals / target_arrivals) * 100,
  ) %>% 
  st_drop_geometry()

## save this to outputs
write_csv(quantity_current_status_rgn, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quantity","int_dom_quantity_current_status_2019_2026.csv"))

quantity_current_status_rgn <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quantity","int_dom_quantity_current_status_2019_2026.csv"))
```


# 5. Regional plotlys for international and domestic scores
International:

```{r}
library(plotly)

int_quantity_current_status <- quantity_current_status_rgn %>% 
  filter(arrival_type %in% "int") %>% 
  arrange(year)

## plot the score over time
final_int_score_plotly <- plotly::plot_ly(
  int_quantity_current_status, 
  x = ~year, 
  y = ~current_status, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: Llegadas de turistas internacionales por municipio",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(int_quantity_current_status$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
final_int_score_plotly

#htmlwidgets::saveWidget(final_int_score_plotly, here::here("TR/sectur_quantity_exploration_int","int_current_status_gapfilled_by_municipio_2019_2026.html"), selfcontained = TRUE)

## plot the number of arrivals over time
final_int_arrivals_plotly <- plotly::plot_ly(
  int_quantity_current_status, 
  x = ~year, 
  y = ~agg_tourist_arrivals, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Llegadas de turistas internacionales por region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Aggregated tourist arrivals // Llegadas turísticas agregadas")
  ) 
final_int_arrivals_plotly
```

Domestic:

```{r}
library(plotly)

dom_quantity_current_status <- quantity_current_status_rgn %>% 
  filter(arrival_type %in% "dom") %>% 
  arrange(year)

## plot the score over time
final_dom_score_plotly <- plotly::plot_ly(
  dom_quantity_current_status, 
  x = ~year, 
  y = ~current_status, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: Llegadas de turistas nacionales por municipio",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(dom_quantity_current_status$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
final_dom_score_plotly

#htmlwidgets::saveWidget(final_dom_score_plotly, here::here("TR/sectur_quantity_exploration_int","dom_current_status_gapfilled_by_municipio_2019_2026.html"), selfcontained = TRUE)

## plot the number of arrivals over time
final_dom_arrivals_plotly <- plotly::plot_ly(
  dom_quantity_current_status, 
  x = ~year, 
  y = ~agg_tourist_arrivals, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Llegadas de turistas nacionales por region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Aggregated tourist arrivals // Llegadas turísticas agregadas")
  ) 
final_dom_arrivals_plotly
```


# 6. Average TR quantity score by region and year
```{r}

tr_quantity_overall_cs <- quantity_current_status_rgn %>% 
  group_by(rgn_id, year) %>% 
  summarize(current_status_new = mean(current_status)) %>% 
  ungroup() %>% 
  rename(score = current_status_new) %>% 
  select(rgn_id, year, score)

write_csv(tr_quantity_overall_cs, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quantity","tr_quantity_current_status_2019_2026.csv"))


```

```{r}
library(plotly)

final_quantity_current_status <- tr_quantity_overall_cs %>% 
  arrange(year)

## plot the score over time
final_score_plotly <- plotly::plot_ly(
  final_quantity_current_status, 
  x = ~year, 
  y = ~score, 
  color = ~rgn_id, 
  colors = ~rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: TR Quantity",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(final_quantity_current_status$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
final_score_plotly

#htmlwidgets::saveWidget(final_score_plotly,here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quantity","plotly_quantity_current_status_2019_2026.html"), selfcontained = TRUE)
```


# 7. Sensitivity: covid-scaled DENUE-only 2020 fills
Uses Script 3 Section 7 output:
`arrivals_filled_by_DENUE_accommodation_employees_2019_2026_covid_scaled.csv`

Same scoring pipeline as above, but DENUE-only 2020 fills are scaled by the observed
SECTUR \(A_{2020}/A_{2019}\) ratio instead of model1.5’s raw year factor. Baseline
Section 1-6 outputs are unchanged. Requires Sections 1 and 4 objects already in memory
(`all_sectur_arrivals_prep_join`, `all_sectur_prop_by_state`, `munis_in_50`,
`rgn_spatial`, `rgn_colors`).

```{r}
gapfilled_total_arrivals_covid_scaled <- read_csv(here::here(
  "/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/gapfilled_int",
  "arrivals_filled_by_DENUE_accommodation_employees_2019_2026_covid_scaled.csv"
)) %>%
  mutate(country = "Mexico") %>%
  select(
    year = data_year, country, estado, municipio_clean, rgn_id,
    arrivals_merged = arrivals_merged_no_neg, is_filled
  )

gapfilled_municipios_covid <- setdiff(
  unique(gapfilled_total_arrivals_covid_scaled$municipio_clean),
  unique(all_sectur_arrivals_prep_join$municipio_clean)
)

gapfilled_municipios_covid_df <- expand.grid(
  municipio_clean = gapfilled_municipios_covid,
  year = 2019:2026,
  arrival_type = c("int", "dom"),
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)

gapfilled_and_sectur_covid <- full_join(
  all_sectur_arrivals_prep_join,
  gapfilled_municipios_covid_df,
  by = c("municipio_clean", "year", "arrival_type")
)

joined_covid_scaled <- full_join(
  gapfilled_total_arrivals_covid_scaled,
  gapfilled_and_sectur_covid,
  by = c("year", "municipio_clean"),
  relationship = "many-to-many"
) %>%
  left_join(all_sectur_prop_by_state, by = c("estado", "arrival_type")) %>%
  mutate(avg_prop = case_when(
    estado == "nayarit" & arrival_type == "dom" ~ 0.56415674,
    estado == "nayarit" & arrival_type == "int" ~ 0.43584326,
    TRUE ~ avg_prop
  )) %>%
  mutate(estimated_arrivals = ifelse(
    is_filled == "1" | is_filled == 1,
    avg_prop * arrivals_merged,
    raw_arrivals
  ))

## observed vs gapfilled totals by type (milder 2020 fill drop for both)
joined_covid_scaled %>%
  filter(year %in% c(2019, 2020, 2024)) %>%
  group_by(arrival_type, year, is_filled) %>%
  summarize(total = sum(estimated_arrivals, na.rm = TRUE), .groups = "drop")
```

```{r}
## reuse spatially assigned municipio -> region from Section 4 (rgn_spatial)
estimated_covid_scaled_ohirgn <- joined_covid_scaled %>%
  select(-any_of("rgn_id")) %>%
  ## same coastal mask as main Script 4 path
  filter(municipio_clean %in% munis_in_50) %>%
  left_join(rgn_spatial, by = "municipio_clean") %>%
  group_by(rgn_id, arrival_type, year) %>%
  summarize(agg_tourist_arrivals = sum(estimated_arrivals, na.rm = TRUE), .groups = "drop") %>%
  filter(year %in% 2019:2025)

baseline_year <- 2024

quantity_current_status_rgn_covid_scaled <- estimated_covid_scaled_ohirgn %>%
  mutate(year = as.numeric(year)) %>%
  group_by(rgn_id, arrival_type) %>%
  mutate(
    baseline_arrivals = agg_tourist_arrivals[year == baseline_year][1],
    target_multiplier = case_when(
      arrival_type == "dom" ~ 1.098,
      arrival_type == "int" ~ 1.30,
      TRUE ~ NA_real_
    ),
    target_arrivals = baseline_arrivals * target_multiplier,
    current_status = (agg_tourist_arrivals / target_arrivals) * 100
  ) %>%
  ungroup()

write_csv(
  quantity_current_status_rgn_covid_scaled,
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quantity",
    "int_dom_quantity_current_status_2019_2026_covid_scaled.csv"
  )
)

tr_quantity_overall_cs_covid_scaled <- quantity_current_status_rgn_covid_scaled %>%
  group_by(rgn_id, year) %>%
  summarize(score = mean(current_status), .groups = "drop")

write_csv(
  tr_quantity_overall_cs_covid_scaled,
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quantity",
    "tr_quantity_current_status_2019_2026_covid_scaled.csv"
  )
)
```

```{r}
library(plotly)

## international scores
int_cs_covid_scaled <- quantity_current_status_rgn_covid_scaled %>%
  filter(arrival_type == "int") %>%
  arrange(year)

final_int_score_plotly_covid_scaled <- plot_ly(
  int_cs_covid_scaled,
  x = ~year,
  y = ~current_status,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines+markers"
) %>%
  layout(
    title = "CURRENT STATUS (covid-scaled fills): International arrivals by region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(int_cs_covid_scaled$year),
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
final_int_score_plotly_covid_scaled

## international raw arrivals
final_int_arrivals_plotly_covid_scaled <- plot_ly(
  int_cs_covid_scaled,
  x = ~year,
  y = ~agg_tourist_arrivals,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines+markers"
) %>%
  layout(
    title = "International arrivals by region (covid-scaled DENUE-only 2020 fills)",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Aggregated tourist arrivals")
  )
final_int_arrivals_plotly_covid_scaled

## domestic scores
dom_cs_covid_scaled <- quantity_current_status_rgn_covid_scaled %>%
  filter(arrival_type == "dom") %>%
  arrange(year)

final_dom_score_plotly_covid_scaled <- plot_ly(
  dom_cs_covid_scaled,
  x = ~year,
  y = ~current_status,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines+markers"
) %>%
  layout(
    title = "CURRENT STATUS (covid-scaled fills): Domestic arrivals by region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(dom_cs_covid_scaled$year),
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
final_dom_score_plotly_covid_scaled

## domestic raw arrivals
final_dom_arrivals_plotly_covid_scaled <- plot_ly(
  dom_cs_covid_scaled,
  x = ~year,
  y = ~agg_tourist_arrivals,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines+markers"
) %>%
  layout(
    title = "Domestic arrivals by region (covid-scaled DENUE-only 2020 fills)",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Aggregated tourist arrivals")
  )
final_dom_arrivals_plotly_covid_scaled

## overall TR quantity score (mean of int + dom)
final_score_plotly_covid_scaled <- plot_ly(
  tr_quantity_overall_cs_covid_scaled %>% arrange(year),
  x = ~year,
  y = ~score,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines+markers"
) %>%
  layout(
    title = "CURRENT STATUS: TR Quantity (covid-scaled fills)",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(tr_quantity_overall_cs_covid_scaled$year),
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
final_score_plotly_covid_scaled

#htmlwidgets::saveWidget(
  final_int_score_plotly_covid_scaled,
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quantity",
    "plotly_int_current_status_2019_2026_covid_scaled.html"
  ),
  selfcontained = TRUE
)

#htmlwidgets::saveWidget(
  final_dom_score_plotly_covid_scaled,
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quantity",
    "plotly_dom_current_status_2019_2026_covid_scaled.html"
  ),
  selfcontained = TRUE
)

#htmlwidgets::saveWidget(
  final_score_plotly_covid_scaled,
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quantity",
    "plotly_quantity_current_status_2019_2026_covid_scaled.html"
  ),
  selfcontained = TRUE
)
```
