---
title: "Script 4 OG-model: TR quantity scores (frozen model3.5/1.5 + CSV regions)"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
date: "2026-7-30"
editor_options:
  chunk_output_type: console
---

# 0. Workflow for this script

This is an **optional parallel** to main Script 4. Run it after Scripts 1–2 and `script3_OGmodel_gapfilling_arrivalls.qmd`. It does **not** replace `script4_tr_quantity.qmd`.

Uses gapfills from the **OG frozen stack** (model3.5 + model1.5; coefficients frozen on **2019–2024**; fills through **2019–2025**).

1. Read cleaned SECTUR yearly arrivals from Script 2.
2. Join with DENUE-gapfilled totals from Script 3 OG-model (`*_OGmodel_UPDATED.csv`).
3. Estimate international vs domestic arrivals where SECTUR is missing.
4. Assign OHI regions from the curated GoC municipio table (`municipality_to_rgnid_update_tr.csv`) — not from geocode ∩ polygons. Flag any `municipio_clean` that does not match the table but leave it in for evaluation. Then compute current status scores (2024 baseline targets).
5. Make regional plots and write outputs with `_OGmodel` filenames (does not overwrite main Script 4 products).

Baseline targets stay the same: domestic = 2024 arrivals x 1.098; international = 2024 arrivals x 1.30.

# 1. Combine SECTUR arrivals with DENUE-gapfilled totals (2019-2026)
In the script `script3_OGmodel_gapfilling_arrivalls.qmd`, I predicted missing tourist arrival data using a model that used the number of employees in hotels/accommodations to predict the total number of arrivals.  To apply it here, we will need to determine the proportion of arrivals that are international vs domestic out of the total in each state.  That way, we can apply it to the regions that have missing data.

Essentially, we will determine for each state the proportion of domestic/total and international/total and just assume that same proportion for the regions we are estimating using the regression models.

I am now aligning this with the updated DENUE series (**2019-2026**). For SECTUR annual arrivals I use **2019-2025** only, because 2026 is still incomplete (Enero-Junio only as of 07/30/26).

```{r}
library(tidyverse)
library(here)
library(sf)
library(mapview)
library(plotly)

## first, we will read in the cleaned sectur data, and only select arrivals.  We want to find the total number of arrivals, int + dom, for every year and municipio.
all_sectur_arrivals <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/hotel_monitoring_int","hotel_data_cleaned_yearly_2013_2025.csv")) %>% 
  filter(metric %in% c("llegadas_de_turistas_residentes", "llegadas_de_turistas_no_residentes")) %>% # domestic and international arrivals
  filter(category %in% "total_categorias")  %>% # since we do not care what star designation it is
  filter(year %in% 2019:2025) %>% 
  mutate(arrival_type = case_when(
    metric == "llegadas_de_turistas_residentes" ~ "dom",
    metric == "llegadas_de_turistas_no_residentes" ~ "int",
    TRUE ~ NA
  ))

# summarize to find the total number of arrivals for every year and municipio
all_sectur_arrivals_sum <- all_sectur_arrivals %>%
  group_by(municipio, year) %>% 
  summarize(total_arrivals = sum(value))

# obtain the proportion of the total for each value.  Double check that for the same municipio, the proportion of the dom + the proportion of the int = 1. Then, assign them each to states.
all_sectur_arrivals_w_prop <- left_join(all_sectur_arrivals, all_sectur_arrivals_sum, by = c("municipio","year")) %>%
  mutate(prop_of_total = value/total_arrivals) %>% 
  mutate(country = "Mexico") %>%  
  mutate(estado = case_when(
    municipio %in% c("cabo_san_lucas", "la_paz", "loreto", "san_jose_del_cabo", "zona_corredor_los_cabos") ~ "baja_california_sur",
    municipio %in% c("san_felipe", "ensenada", "mexicali", "san_quintin") ~ "baja_california",
    municipio %in% c("guaymas", "hermosillo") ~ "sonora",
    municipio %in% c("culiacan", "el_fuerte", "los_mochis", "mazatlan", "escuinapa", "el_rosario") ~ "sinaloa",
    municipio %in% c("nuevo_nayarit", "tepic") ~ "nayarit",
    municipio %in% c("puerto_vallarta") ~ "jalisco",
    TRUE ~ ""  # default case for any unmatched cities
  ))  
  
```

For now, we will average the proportion across the years 2019:2025 for every state. This is because in some states, notably Sonora, we have very little information on the numbers of arrivals throughout all years.  So we cannot obtain a number for the proportion of dom vs. int for every year for each state. 

Note that Nayarit has little/no usable SECTUR coverage in places, so I will assume it has the same proportions as Jalisco when needed.  

I will need to check in with Mel on all of this!! -- checked, this is the best we can do (120925).

```{r}
# summarize the average proportion of int and dom for every state.
all_sectur_prop_by_state <- all_sectur_arrivals_w_prop %>% 
  group_by(arrival_type, estado) %>% 
  summarize(avg_prop = mean(prop_of_total, na.rm = TRUE))

## save it!!
write_csv(all_sectur_prop_by_state, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/gapfilled_int","sectur_avg_prop_by_state_2019_2025.csv"))
```

## 1.1 Read gapfilled arrivals from Script 3
```{r}
arrivals_filled_by_DENUE_accommodation_employees <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/gapfilled_int","arrivals_filled_by_DENUE_accommodation_employees_2019_2026_OGmodel_UPDATED.csv"))

gapfilled_total_arrivals <- arrivals_filled_by_DENUE_accommodation_employees %>%
  mutate(country = "Mexico") %>% 
  select(year = data_year, country, estado, municipio_clean, rgn_id, arrivals_merged = arrivals_merged_no_neg, is_filled) 

```

```{r}
## prep all sectur arrivals for a join with the gapfilled totals
all_sectur_arrivals_prep_join <- all_sectur_arrivals %>% 
  mutate(
    municipio_clean = case_when(
      municipio %in% c("cabo_san_lucas", "san_jose_del_cabo", "zona_corredor_los_cabos") ~ "los_cabos",
      municipio == "los_mochis" ~ "ahome", # los mochis is a city within the municipality of ahome!!
      municipio == "el_rosario" ~ "rosario",
      municipio == "culiacan" ~ "culiacán",
      municipio == "mazatlan" ~ "mazatlán",
      municipio == "nuevo_nayarit" ~ "bahía_de_banderas", # Nuevo Nayarit is located inside the Municipality of Bahía de Banderas
      municipio == "san_quintin" ~ "san_quintín",
      TRUE ~ municipio
    )
  ) %>% 
  # reaggregate with changes
  group_by(year, municipio_clean, arrival_type) %>%
  summarize(value = sum(value), .groups = "drop") %>% 
  select(year, municipio_clean, arrival_type, raw_arrivals = value) %>% # leave in "municipio" to double check at first
  ## Drop Pacific Datatur centers: Ensenada / San Quintín arrivals are Pacific-side.
  ## Also drop mexicali (merged to san_felipe for DENUE). These municipios are
  ## gapfilled from GOC-buffer DENUE hotels instead (Region_2 for ensenada / san_quintín).
  filter(!municipio_clean %in% c("mexicali", "ensenada", "san_quintín", "san_quintin"))

## check!!
setdiff(unique(all_sectur_arrivals_prep_join$municipio_clean), unique(gapfilled_total_arrivals$municipio_clean)) # character(0)
setdiff(unique(gapfilled_total_arrivals$municipio_clean), unique(all_sectur_arrivals_prep_join$municipio_clean)) 

## makes sense, those were all the municipalities that were gapfilled using DENUE accommodation employees!!
```


```{r}
## municipalities missing in SECTUR arrivals
gapfilled_municipios <- setdiff(unique(gapfilled_total_arrivals$municipio_clean), unique(all_sectur_arrivals_prep_join$municipio_clean))

## expand into dataframe with years 2019-2026 and arrival types int/dom
## (SECTUR observed values stop at 2025; 2026 rows will be DENUE-gapfilled only)
gapfilled_municipios_df <- expand.grid(
  municipio_clean = gapfilled_municipios,
  year = 2019:2026,
  arrival_type = c("int", "dom"),
  KEEP.OUT.ATTRS = FALSE, # a logical indicating the "out.attrs" attribute should be computed and returned.
  stringsAsFactors = FALSE # logical specifying if character vectors are converted to factors.
)

## join with the prepped sectur arrivals df
gapfilled_municipios_and_sectur_arrivals_prep_join <- full_join(all_sectur_arrivals_prep_join, gapfilled_municipios_df, by = c("municipio_clean","year","arrival_type"))

```

Join the gapfilled totals with the raw int and dom values for arrivals:

```{r}
joined_gapfilled_and_raw_arrivals <- full_join(gapfilled_total_arrivals, gapfilled_municipios_and_sectur_arrivals_prep_join, by = c("year","municipio_clean"), relationship = "many-to-many")

## check that the raw arrivals for int + dom add up to the "arrivals_merged" when "is_filled" = 0 (it was not predicted using DENUE employees for accommodations)
## looks good!
```

Now, join this data frame with the avg proportions by state.

```{r}
all_sectur_prop_by_state <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/gapfilled_int","sectur_avg_prop_by_state_2019_2025.csv"))

joined_gapfilled_arrivals_and_prop <- left_join(joined_gapfilled_and_raw_arrivals, all_sectur_prop_by_state, by = c("estado", "arrival_type")) %>% 
  mutate(avg_prop = case_when(
    estado == "nayarit" & arrival_type == "dom" ~ 0.56415674,
    estado == "nayarit" & arrival_type == "int" ~ 0.43584326,
    TRUE ~ avg_prop
  )) # assume for now that nayarit has the save avg proportion as jalisco, though this is definitely not true.
```

Now we can actually estimate the number of international arrivals and domestic arrivals for all predicted total arrivals from DENUE data.

```{r}
estimated_int_dom_arrivals <- joined_gapfilled_arrivals_and_prop %>% 
  mutate(estimated_arrivals = ifelse(is_filled == "1", avg_prop*arrivals_merged, raw_arrivals)) # if it was gapfilled, then estimate the number of int or dom arrivals by multiplying the average proportion of int or dom arrivals by state and the predicted total arrivals from the DENUE data. If they were not gapfilled, use the origianl int and dom arrival values.
```


# 2. Plot estimated arrivals over time
```{r}
library(plotly)

### INT Arrivals over time ###
estimated_int_arrivals <- estimated_int_dom_arrivals %>% 
  filter(arrival_type %in% "int") %>% 
  arrange(year)


int_arrivals_plotly <- plotly::plot_ly(
  estimated_int_arrivals, 
  x = ~year, 
  y = ~estimated_arrivals, 
  color = ~municipio_clean, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "International arrivals by year and municipio",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Number of arrivals")
  )
int_arrivals_plotly
```


```{r}
### DOM Arrivals ###
estimated_dom_arrivals <- estimated_int_dom_arrivals %>% 
  filter(arrival_type %in% "dom") %>% 
  arrange(year)

dom_arrivals_plotly <- plotly::plot_ly(
  estimated_dom_arrivals, 
  x = ~year, 
  y = ~estimated_arrivals, 
  color = ~municipio_clean, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "Domestic arrivals by year and municipio",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Number of arrivals")
  )
dom_arrivals_plotly

```


# 3. Current status by municipio (optional / commented)
```{r}
# baseline_year <- 2024
# 
# quantity_current_status <- estimated_int_dom_arrivals %>% 
#   mutate(year = as.numeric(year)) %>%
#   
#   # group by municipio + arrival type
#   group_by(municipio_clean, arrival_type) %>%
#   
#   mutate(
#     # baseline per municipio + type
#     baseline_arrivals = estimated_arrivals[year == baseline_year][1],
# 
#     # target multiplier varies by type
#     target_multiplier = case_when(
#       arrival_type == "dom" ~ 1.098, # domestic +9.8%
#       arrival_type == "int" ~ 1.30,  # international +30%
#       TRUE ~ NA_real_
#     ),
# 
#     # calculate targets
#     target_arrivals = baseline_arrivals * target_multiplier,
# 
#     # CURRENT STATUS (0 to 100)
#     current_status = (estimated_arrivals / target_arrivals) * 100,
# 
#     # absolute and % gap
#     gap_to_target = target_arrivals - estimated_arrivals,
#     gap_percentage = ((target_arrivals - estimated_arrivals) / estimated_arrivals) * 100
#   ) 
# 
# ## save this to outputs
# write_csv(quantity_current_status, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quantity","quantity_current_status_120125.csv"))
```

## 3.1 Municipio plotlys (optional / commented)
International:

```{r}
# library(plotly)
# 
# int_quantity_current_status <- quantity_current_status %>% 
#   filter(arrival_type %in% "int") %>% 
#   arrange(year)
# 
# ## plot the score over time
# final_int_score_plotly <- plotly::plot_ly(
#   int_quantity_current_status, 
#   x = ~year, 
#   y = ~current_status, 
#   color = ~municipio_clean, 
#   type = "scatter", 
#   mode = "lines"
# ) %>% 
#   layout(
#     title = "CURRENT STATUS: Llegadas de turistas internacionales por municipio",
#     xaxis = list(title = "Año // Year"),
#     yaxis = list(title = "Puntaje // Percent (out of 100)")
#   ) %>%
#   add_lines(
#     x = range(int_quantity_current_status$year), 
#     y = c(100, 100),
#     line = list(color = "black", width = 2, dash = "dash"),
#     name = "Meta // Goal 2030 (100/100)",
#     inherit = FALSE
#   )
# final_int_score_plotly
# 
# htmlwidgets::saveWidget(final_int_score_plotly, here::here("TR/sectur_quantity_exploration_int","int_current_status_gapfilled_by_municipio_120125.html"), selfcontained = TRUE)
```

Domestic:

```{r}
# library(plotly)
# 
# dom_quantity_current_status <- quantity_current_status %>% 
#   filter(arrival_type %in% "dom") %>% 
#   arrange(year)
# 
# ## plot the score over time
# final_dom_score_plotly <- plotly::plot_ly(
#   dom_quantity_current_status, 
#   x = ~year, 
#   y = ~current_status, 
#   color = ~municipio_clean, 
#   type = "scatter", 
#   mode = "lines"
# ) %>% 
#   layout(
#     title = "CURRENT STATUS: Llegadas de turistas nacionales por municipio",
#     xaxis = list(title = "Año // Year"),
#     yaxis = list(title = "Puntaje // Percent (out of 100)")
#   ) %>%
#   add_lines(
#     x = range(dom_quantity_current_status$year), 
#     y = c(100, 100),
#     line = list(color = "black", width = 2, dash = "dash"),
#     name = "Meta // Goal 2030 (100/100)",
#     inherit = FALSE
#   )
# final_dom_score_plotly
# 
# htmlwidgets::saveWidget(final_dom_score_plotly, here::here("TR/sectur_quantity_exploration_int","dom_current_status_gapfilled_by_municipio_120125.html"), selfcontained = TRUE)
```

# 4. Current status by OHI region

Assign each `municipio_clean` to an OHI region using the curated GoC municipality table
(`/home/shares/ohi/OHI_GOC/spatial/municipality_to_rgnid/municipality_to_rgnid_update_tr.csv`
on mazu; local copy also lives in this repo). This replaces inland_50 filters and
`st_nearest_feature` region assignment so regions do not depend on geocode jitter.

- GoC municipios in the table get their `rgn_id` from the CSV.
- `nuevo_nayarit` to `bahía_de_banderas`; `juan_josé_ríos` to `ahome`.
- Drop inland / Pacific / out-of-study names: comondú, mascota, san_sebastián_del_oeste, talpa_de_allende, tomatlán.
- Keep `huajicori` (Region_9). Ensenada / san_quintín GOC DENUE fills to Region_2.
- Any remaining unmatched `municipio_clean` is flagged for evaluation.
- CSV column `include_tr_quantity` (Y/N) drops inland / mountainous municipios from regional scores (especially Region_9 interiors like Tepic).

No municipio geocoding here: tidygeocode often misplaces Baja / Nayarit / Jalisco names, and regions come from the CSV anyway. Regional maps below join OHI polygons directly.

```{r}
## Aliases / drops before region join
drop_non_goc_munis <- c(
  "comondú", "comondu",
  "mascota",
  "san_sebastián_del_oeste", "san_sebastian_del_oeste",
  "talpa_de_allende",
  "tomatlán", "tomatlan"
)

estimated_int_dom_arrivals <- estimated_int_dom_arrivals %>%
  mutate(
    municipio_clean = case_when(
      municipio_clean %in% c("juan_josé_ríos", "juan_jose_rios") ~ "ahome",
      municipio_clean %in% c("nuevo_nayarit") ~ "bahía_de_banderas",
      TRUE ~ municipio_clean
    )
  ) %>%
  filter(!municipio_clean %in% drop_non_goc_munis) %>%
  group_by(year, municipio_clean, arrival_type) %>%
  summarize(
    estimated_arrivals = sum(estimated_arrivals, na.rm = TRUE),
    country = dplyr::first(stats::na.omit(country)),
    estado = {
      x <- estado[!is.na(estado) & estado != ""]
      if (length(x) == 0) NA_character_ else x[[1]]
    },
    is_filled = dplyr::first(as.character(is_filled)),
    .groups = "drop"
  )

message(
  "Dropped non-GoC municipios from scoring panel: ",
  paste(unique(drop_non_goc_munis), collapse = ", ")
)
message("Aliased juan_josé_ríos to ahome; kept huajicori (Region_9).")
```

```{r}
## ASCII fold + snake_case so CSV names (BAHIA DE BANDERAS, trailing nbsp, etc.)
## match pipeline municipio_clean (bahía_de_banderas, accents, underscores)
fold_muni_key <- function(x) {
  x <- as.character(x)
  x <- gsub("\u00A0|\u2007|\u202F|\uFEFF|\u200B", "", x, perl = TRUE)
  x <- trimws(x)
  x <- tolower(x)
  x <- iconv(x, from = "UTF-8", to = "ASCII//TRANSLIT")
  x <- gsub("[^a-z0-9]+", "_", x)
  x <- gsub("^_|_$", "", x)
  x
}

muni_rgn_candidates <- c(
  "/home/shares/ohi/OHI_GOC/spatial/municipality_to_rgnid/municipality_to_rgnid_update_tr.csv",
  here::here("municipality_to_rgnid_update_tr.csv"),
  here::here("../municipality_to_rgnid_update_tr.csv")
)
muni_rgn_csv_path <- muni_rgn_candidates[file.exists(muni_rgn_candidates)][1]
if (is.na(muni_rgn_csv_path)) {
  stop("Could not find municipality_to_rgnid_update_tr.csv (tried mazu path + local repo copies).")
}
message("Using municipality to region CSV: ", muni_rgn_csv_path)

goc_muni_rgn <- read_csv(muni_rgn_csv_path, show_col_types = FALSE) %>%
  mutate(
    municipio = gsub("\u00A0|\u2007|\u202F|\uFEFF|\u200B", "", municipio, perl = TRUE),
    municipio = trimws(municipio),
    municipio_key = fold_muni_key(municipio),
    ## treat literal "NA" / blank as missing region
    rgn_id = na_if(as.character(rgn_id), "NA"),
    rgn_id = na_if(rgn_id, ""),
    include_tr_quantity = toupper(trimws(as.character(include_tr_quantity))),
    include_tr_quantity = if_else(
      include_tr_quantity %in% c("Y", "YES", "TRUE", "1"),
      "Y",
      "N"
    )
  ) %>%
  distinct(municipio_key, .keep_all = TRUE) %>%
  select(
    municipio_key, rgn_id_csv = rgn_id, municipio_csv = municipio,
    both_pacific_gulf, directly_coastal, include_tr_quantity, tr_exclude_reason, notes
  )

## one row per municipio in the (already filtered/aliased) arrivals panel
muni_in_data <- estimated_int_dom_arrivals %>%
  distinct(municipio_clean) %>%
  mutate(
    municipio_for_rgn = municipio_clean,
    municipio_key = fold_muni_key(municipio_for_rgn)
  )

rgn_lookup <- muni_in_data %>%
  left_join(goc_muni_rgn, by = "municipio_key") %>%
  mutate(
    ## GOC DENUE gapfills for these municipios score as Western Midriff (Region_2).
    ## Datatur Pacific SECTUR for the same names is dropped in Section 1.
    ## Huajicori: Nayarit GoC table entry (Region_9); force if mazu CSV not yet updated.
    rgn_id_csv = case_when(
      municipio_clean %in% c("ensenada", "san_quintín", "san_quintin") ~ "Region_2",
      municipio_clean %in% c("huajicori") ~ coalesce(rgn_id_csv, "Region_9"),
      TRUE ~ rgn_id_csv
    ),
    ## default include_tr_quantity from CSV; force-include ensenada/san_quintín/huajicori
    include_tr_quantity = case_when(
      municipio_clean %in% c("ensenada", "san_quintín", "san_quintin", "huajicori") ~ "Y",
      is.na(include_tr_quantity) ~ "N",
      TRUE ~ include_tr_quantity
    ),
    rgn_assign_status = case_when(
      municipio_clean %in% c("ensenada", "san_quintín", "san_quintin", "huajicori") ~
        "matched_goc_muni_table",
      is.na(municipio_csv) ~ "FLAG_not_in_goc_muni_table",
      is.na(rgn_id_csv) ~ "FLAG_in_table_but_rgn_id_missing",
      TRUE ~ "matched_goc_muni_table"
    ),
    rgn_id = rgn_id_csv,
    in_goc_muni_table = !is.na(municipio_csv) |
      municipio_clean %in% c("ensenada", "san_quintín", "san_quintin", "huajicori")
  ) %>%
  select(
    municipio_clean, municipio_for_rgn, rgn_id, rgn_assign_status,
    in_goc_muni_table, include_tr_quantity, tr_exclude_reason,
    both_pacific_gulf, directly_coastal, notes
  )

message("=== OHI region assignment via municipality_to_rgnid_update_tr.csv ===")
message("Matched with rgn_id: ",
        sum(rgn_lookup$rgn_assign_status == "matched_goc_muni_table"))
message("In table but rgn_id missing: ",
        paste(sort(rgn_lookup$municipio_clean[rgn_lookup$rgn_assign_status == "FLAG_in_table_but_rgn_id_missing"]),
              collapse = ", "))
message("FLAG — not in GoC muni table (kept for evaluation): ",
        paste(sort(rgn_lookup$municipio_clean[rgn_lookup$rgn_assign_status == "FLAG_not_in_goc_muni_table"]),
              collapse = ", "))
message(
  "Excluded from TR quantity scores (include_tr_quantity = N): ",
  paste(sort(rgn_lookup$municipio_clean[rgn_lookup$include_tr_quantity == "N"]), collapse = ", ")
)
message(
  "Region_9 kept for scoring: ",
  paste(
    sort(rgn_lookup$municipio_clean[
      rgn_lookup$rgn_id == "Region_9" & rgn_lookup$include_tr_quantity == "Y"
    ]),
    collapse = ", "
  )
)

rgn_lookup %>%
  filter(rgn_assign_status != "matched_goc_muni_table" | include_tr_quantity == "N") %>%
  arrange(include_tr_quantity, rgn_assign_status, municipio_clean)

write_csv(
  rgn_lookup %>% arrange(rgn_assign_status, include_tr_quantity, municipio_clean),
  here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/hotel_monitoring_int",
    "ohi_rgn_assign_for_municipio_clean_OGmodel.csv"
  )
)

## keep matched GoC municipios (+ flagged only if still present); drop list already applied above
estimated_int_dom_arrivals_rgn <- estimated_int_dom_arrivals %>%
  select(-any_of("rgn_id")) %>%
  left_join(
    rgn_lookup %>% select(
      municipio_clean, rgn_id, rgn_assign_status, in_goc_muni_table,
      include_tr_quantity, tr_exclude_reason
    ),
    by = "municipio_clean"
  )

estimated_int_dom_arrivals_rgn %>%
  distinct(municipio_clean, rgn_id, rgn_assign_status, include_tr_quantity) %>%
  count(rgn_assign_status, include_tr_quantity)
```

Let's aggregate all tourism arrivals by region and type.

Matched GoC municipios with a valid `rgn_id` **and** `include_tr_quantity == "Y"` enter regional scores.
Inland / mountainous municipios (`include_tr_quantity == "N"` in the CSV) stay in
`estimated_int_dom_arrivals_rgn` for evaluation but are excluded from regional sums
(so Region_9 is not inflated by Tepic and other interior centers).

```{r}
estimated_int_dom_arrivals_ohirgn <- estimated_int_dom_arrivals_rgn %>%
  filter(
    rgn_assign_status == "matched_goc_muni_table",
    include_tr_quantity == "Y",
    !is.na(rgn_id),
    rgn_id != ""
  ) %>%
  group_by(rgn_id, arrival_type, year) %>%
  summarize(agg_tourist_arrivals = sum(estimated_arrivals, na.rm = TRUE), .groups = "drop") %>%
  filter(year %in% 2019:2025)
```

## 4.1 Domestic trend visualizations
```{r}
last_3yrs <- c(2023, 2024, 2025)

trend_last3_slope <- estimated_int_dom_arrivals_ohirgn %>% 
  filter(year %in% last_3yrs) %>%
  group_by(rgn_id, arrival_type) %>%
  summarise(
    slope_per_year = coef(lm(agg_tourist_arrivals ~ year))[2],
    .groups = "drop"
  )

trend_last3_pct <- estimated_int_dom_arrivals_ohirgn %>% 
  filter(year %in% last_3yrs) %>%
  group_by(rgn_id, arrival_type) %>%
  summarise(
    arrivals_2023 = agg_tourist_arrivals[year == 2023],
    arrivals_2025 = agg_tourist_arrivals[year == 2025],
    pct_change_3yr = (arrivals_2025 - arrivals_2023) / arrivals_2023 * 100,
    .groups = "drop"
  ) 

dom_trend_last3yrs <- full_join(trend_last3_pct, trend_last3_slope) %>% 
  filter(arrival_type == "dom") %>% 
  mutate(target_arrivals = arrivals_2025 * 1.098, # illustrative vs latest complete year; official score baseline remains 2024 elsewhere
         proj_arrivals_2030 = arrivals_2025 + (5 * slope_per_year), # 2025 -> 2030 = 5 years
         on_track = ifelse(proj_arrivals_2030 >= target_arrivals, 1, 0))
write_csv(dom_trend_last3yrs, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_viz","dom_trend_last3yrs_arrivals_2023_2025_OGmodel.csv"))

#### join with ohi rgn polygons
ohi <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp")) %>%
  filter(location == "marine") %>% 
  select(rgn_id)

dom_trend_last3yrs_ohigns <- full_join(dom_trend_last3yrs, ohi, by = "rgn_id") %>% 
  st_as_sf()

mapview(dom_trend_last3yrs_ohigns, 
        zcol = "pct_change_3yr",
        col.regions = colorRampPalette(c('#8a0032', '#A50026', '#A50026', '#D73027', '#F46D43', '#FDAE61', '#FEE090', '#FEE090', '#ABD9E9', '#4575B4', '#313695')),
        layer.name = "Percent change of domestic arrivals (2023-2025)",
        legend = TRUE)

########################################

### across the gulf:
trend_overall <- estimated_int_dom_arrivals_ohirgn %>% 
  filter(year %in% last_3yrs) %>%
  group_by(arrival_type, year) %>%
  summarise(
    total_arrivals = sum(agg_tourist_arrivals, na.rm = TRUE),
    .groups = "drop"
  )

trend_overall_slope <- trend_overall %>%
  group_by(arrival_type) %>%
  summarise(
    slope_per_year = coef(lm(total_arrivals ~ year))[2]
  )

trend_overall_pct_change <- trend_overall %>%
  group_by(arrival_type) %>%
  summarise(
    pct_change_3yr =
      (total_arrivals[year == 2025] - total_arrivals[year == 2023]) /
      total_arrivals[year == 2023] * 100
  )

dom_trend_goc <- full_join(trend_overall_slope, trend_overall_pct_change) %>% 
  filter(arrival_type == "dom")
write_csv(dom_trend_goc, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_viz","dom_trend_goc_arrivals_2023_2025_OGmodel.csv"))
```


```{r}
#### for visualizations: see the proportion of int vs dom for each rgn
prop_dom_by_rgn <- estimated_int_dom_arrivals_ohirgn %>% 
  group_by(rgn_id, year) %>% 
  mutate(rgn_agg_arr = sum(agg_tourist_arrivals)) %>% 
  ungroup() %>% 
  mutate(prop_rgn_total = agg_tourist_arrivals / rgn_agg_arr) %>% 
  filter(year == 2025) %>% 
  filter(arrival_type == "dom")
write_csv(prop_dom_by_rgn, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_viz","prop_dom_by_rgn_arrivals_2025_OGmodel.csv"))

### join with ohi rgn polygons
ohi <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp")) %>%
  filter(location == "marine") %>% 
  select(rgn_id)

prop_dom_arrivals_ohigns <- full_join(prop_dom_by_rgn, ohi, by = "rgn_id") %>% 
  st_as_sf()

mapview(prop_dom_arrivals_ohigns, 
        zcol = "prop_state_total",
        col.regions = colorRampPalette(c("#096b91", "#a057ce")),
        layer.name = "Proportion of Regionwide Domestic Arrivals (2025)",
        legend = TRUE)

#### prop of total goc
prop_dom_all <- estimated_int_dom_arrivals_ohirgn %>% 
  filter(year == 2025) %>% 
  group_by(year) %>% 
  mutate(all_agg_arr = sum(agg_tourist_arrivals)) %>% 
  ungroup() %>% 
  group_by(arrival_type, year) %>% 
  mutate(total_type_by_yr = sum(agg_tourist_arrivals)) %>% 
  ungroup() %>% 
  select(-c(rgn_id, agg_tourist_arrivals)) %>% 
  distinct() %>% 
  mutate(prop_dom_total = total_type_by_yr / all_agg_arr) %>% 
  filter(arrival_type == "dom")
```

## 4.2 Calculate CS
```{r}
baseline_year <- 2024

quantity_current_status_rgn <- estimated_int_dom_arrivals_ohirgn %>% 
  mutate(year = as.numeric(year)) %>%
  
  # group by ohi_rgn + arrival type
  group_by(rgn_id, arrival_type) %>%
  
  mutate(
    # baseline per municipio + type
    baseline_arrivals = agg_tourist_arrivals[year == baseline_year][1],

    # target multiplier varies by type
    target_multiplier = case_when(
      arrival_type == "dom" ~ 1.098, # domestic +9.8%
      arrival_type == "int" ~ 1.30,  # international +30%
      TRUE ~ NA_real_
    ),

    # calculate targets
    target_arrivals = baseline_arrivals * target_multiplier,

    # CURRENT STATUS (0 to 100)
    current_status = (agg_tourist_arrivals / target_arrivals) * 100,
  ) %>%
  ungroup()

## save this to outputs
write_csv(quantity_current_status_rgn, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quantity","int_dom_quantity_current_status_2019_2026_OGmodel.csv"))

quantity_current_status_rgn <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quantity","int_dom_quantity_current_status_2019_2026_OGmodel.csv"))
```


# 5. Regional plotlys for international and domestic scores
International:

```{r}
library(plotly)

int_quantity_current_status <- quantity_current_status_rgn %>% 
  filter(arrival_type %in% "int") %>% 
  arrange(year)

## plot the score over time
final_int_score_plotly <- plotly::plot_ly(
  int_quantity_current_status, 
  x = ~year, 
  y = ~current_status, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: Llegadas de turistas internacionales por municipio",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(int_quantity_current_status$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
final_int_score_plotly

htmlwidgets::saveWidget(final_int_score_plotly, here::here("TR/sectur_quantity_exploration_int","int_current_status_gapfilled_by_municipio_2019_2026_OGmodel.html"), selfcontained = TRUE)

## plot the number of arrivals over time
final_int_arrivals_plotly <- plotly::plot_ly(
  int_quantity_current_status, 
  x = ~year, 
  y = ~agg_tourist_arrivals, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Llegadas de turistas internacionales por region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Aggregated tourist arrivals // Llegadas turísticas agregadas")
  ) 
final_int_arrivals_plotly
```

Domestic:

```{r}
library(plotly)

dom_quantity_current_status <- quantity_current_status_rgn %>% 
  filter(arrival_type %in% "dom") %>% 
  arrange(year)

## plot the score over time
final_dom_score_plotly <- plotly::plot_ly(
  dom_quantity_current_status, 
  x = ~year, 
  y = ~current_status, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: Llegadas de turistas nacionales por municipio",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(dom_quantity_current_status$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
final_dom_score_plotly

htmlwidgets::saveWidget(final_dom_score_plotly, here::here("TR/sectur_quantity_exploration_int","dom_current_status_gapfilled_by_municipio_2019_2026_OGmodel.html"), selfcontained = TRUE)

## plot the number of arrivals over time
final_dom_arrivals_plotly <- plotly::plot_ly(
  dom_quantity_current_status, 
  x = ~year, 
  y = ~agg_tourist_arrivals, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Llegadas de turistas nacionales por region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Aggregated tourist arrivals // Llegadas turísticas agregadas")
  ) 
final_dom_arrivals_plotly
```


# 6. Average TR quantity score by region and year
```{r}

tr_quantity_overall_cs <- quantity_current_status_rgn %>% 
  group_by(rgn_id, year) %>% 
  summarize(current_status_new = mean(current_status)) %>% 
  ungroup() %>% 
  rename(score = current_status_new) %>% 
  select(rgn_id, year, score)

write_csv(tr_quantity_overall_cs, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quantity","tr_quantity_current_status_2019_2026_OGmodel.csv"))


```

```{r}
library(plotly)

final_quantity_current_status <- tr_quantity_overall_cs %>% 
  arrange(year)

## plot the score over time
final_score_plotly <- plotly::plot_ly(
  final_quantity_current_status, 
  x = ~year, 
  y = ~score, 
  color = ~rgn_id, 
  colors = ~rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: TR Quantity",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(final_quantity_current_status$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
final_score_plotly

htmlwidgets::saveWidget(final_score_plotly,here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quantity","plotly_quantity_current_status_2019_2026_OGmodel.html"), selfcontained = TRUE)
```


# 7. Note on covid-scaled sensitivity

Not included in this OG-model parallel. Compare against main `script4_tr_quantity.qmd` Section 7 if needed. This script scores only the model3 + model1 gapfills from `script3_OGmodel_gapfilling_arrivalls.qmd`.



