---
title: " "
format: html
toc: true
---

## Habitat Data for the Gulf of California

There are a variety of habitats located within the Gulf of California ecoregion. In an effort to understand the data available, various sources were explored. This includes a coastline habitat classification system, the wetlands of importance as determined by the Ramsar List, mangrove polygons from SIMAR and CONABIO, and the area of different types of habitats within a 500km\^2 area.

### Global Coastline Fine Classification System (GCL_FCS30)

This high-resolution dataset classifies coastlines globally, and is cropped to the Gulf of California's ecoregion. It splits the coastline into six distinct categories at 30-meter resolution. The interactive map shows how these different coastline types are distributed:

-   Artificial coastlines (purple): Human-built structures like ports and seawalls

-   Biogenic coastlines (blue): Natural formations including mangroves and marshes

-   Sandy coastlines (teal): Beach and dune environments

-   Muddy coastlines (green): Tidal flats and sediment-rich zones

-   Rocky coastlines (yellow-green): Cliff faces and bedrock shorelines

-   Estuary coastlines (yellow): Transition zones where rivers meet the sea

The map allows you to toggle between different coastline types to visualize their distribution throughout the Gulf. Zooming in on the interactive map will provide a better understanding of each coastline type's spatial distribution.

While this map only shows the classification for 2020, the dataset also includes data for 2010 and 2015. Therefore, it is valuable for monitoring coastal changes, development impacts, and natural habitat distribution over time.

```{r}
#| warning: false
#| message: false
#| echo: false

library(magrittr)
library(mapview)
library(sf)
library(tidyverse)
library(here)
library(plotly)
library(viridis)
library(leaflet)
library(leaflet.extras)
library(RColorBrewer)


# ---- sources! ----
# source(here("workflow", "R", "common.R")) # file creates objects to process data
## set the mazu and neptune data_edit share based on operating system
dir_M             <- c('Windows' = '//mazu.nceas.ucsb.edu/ohi',
                       'Darwin'  = '/Volumes/ohi',    ### connect (cmd-K) to smb://mazu/ohi
                       'Linux'   = '/home/shares/ohi')[[ Sys.info()[['sysname']] ]]

gcl_2020_cropped <- read_sf(here::here(dir_M, "OHI_GOC/_raw_data/GCL_FCS30/d2025/int/gcl_2020_cropped.shp"))

viridis_colors <- viridis::viridis(6)

# match colors to class meanings for clarity
color_mapping <- c(
  "0" = viridis_colors[1],  # artificial
  "1" = viridis_colors[2],  # biogenic
  "2" = viridis_colors[3],  # sandy
  "3" = viridis_colors[4],  # muddy
  "4" = viridis_colors[5],  # rocky
  "5" = viridis_colors[6]   # estuary
)

# use color palette function
pal <- colorFactor(
  palette = color_mapping,
  domain = gcl_2020_cropped$class
)

# class descriptions for popup
class_descriptions <- c(
  "0" = "Artificial: Human-built structures",
  "1" = "Biogenic: Mangroves, marshes, reefs",
  "2" = "Sandy: Beaches, dunes",
  "3" = "Muddy: Tidal flats, mudflats",
  "4" = "Rocky: Cliffs, bedrock",
  "5" = "Estuary: River-sea transition zones"
)

leaflet(gcl_2020_cropped) %>%
  addProviderTiles("CartoDB.Positron", group = "Light Map") %>% # add the base layers first
  addProviderTiles("Esri.OceanBasemap", group = "Ocean Basemap") %>%
  addProviderTiles("Esri.WorldImagery", group = "Satellite") %>%
  
  # now add the habitat layers - one group per class!! That way they can be controlled in the map by the viewer
  addPolylines(
    data = gcl_2020_cropped[gcl_2020_cropped$class == "0", ], # 0
    color = ~pal(class),
    weight = 6,
    opacity = 0.7,
    popup = ~paste("<b>Class:</b>", class, "<br>",
                  "<b>Description:</b>", class_descriptions[as.character(class)]),
    group = "Artificial"
  ) %>%
  addPolylines(
    data = gcl_2020_cropped[gcl_2020_cropped$class == "1", ], # 1
    color = ~pal(class),
    weight = 6,
    opacity = 0.7,
    popup = ~paste("<b>Class:</b>", class, "<br>",
                  "<b>Description:</b>", class_descriptions[as.character(class)]),
    group = "Biogenic"
  ) %>%
  addPolylines(
    data = gcl_2020_cropped[gcl_2020_cropped$class == "2", ], # 2
    color = ~pal(class),
    weight = 6,
    opacity = 0.7,
    popup = ~paste("<b>Class:</b>", class, "<br>",
                  "<b>Description:</b>", class_descriptions[as.character(class)]),
    group = "Sandy"
  ) %>%
  addPolylines(
    data = gcl_2020_cropped[gcl_2020_cropped$class == "3", ], # 3
    color = ~pal(class),
    weight = 6,
    opacity = 0.7,
    popup = ~paste("<b>Class:</b>", class, "<br>",
                  "<b>Description:</b>", class_descriptions[as.character(class)]),
    group = "Muddy"
  ) %>%
  addPolylines(
    data = gcl_2020_cropped[gcl_2020_cropped$class == "4", ], # 4 
    color = ~pal(class),
    weight = 6,
    opacity = 0.7,
    popup = ~paste("<b>Class:</b>", class, "<br>",
                  "<b>Description:</b>", class_descriptions[as.character(class)]),
    group = "Rocky"
  ) %>%
  addPolylines(
    data = gcl_2020_cropped[gcl_2020_cropped$class == "5", ], # 6
    color = ~pal(class),
    weight = 6,
    opacity = 0.7,
    popup = ~paste("<b>Class:</b>", class, "<br>",
                  "<b>Description:</b>", class_descriptions[as.character(class)]),
    group = "Estuary"
  ) %>%

  # group layer control
  addLayersControl(
    baseGroups = c("Light Map", "Ocean Basemap", "Satellite"),
    overlayGroups = c("Artificial", "Biogenic", "Sandy", "Muddy","Rocky","Estuary"),
    options = layersControlOptions(collapsed = FALSE)
  ) 
  # addLegend(
  #   position = "bottomright",
  #   pal = pal,
  #   values = c("Artificial", "Biogenic", "Sandy", "Muddy","Rocky","Estuary"),
  #   title = "Habitat Classification",
  #   opacity = 0.7
  # )
```

#### Citation

Zuo, J., Zhang, L., Xiao, J. et al. GCL_FCS30: a global coastline dataset with 30-m resolution and a fine classification system from 2010 to 2020. Sci Data 12, 129 (2025). <https://doi.org/10.1038/s41597-025-04430-0>

Data Download: <https://doi.org/10.5281/zenodo.13943679>

### Habitat Extent and Distribution (dataMares)

This 2017 dataset from dataMares visualizes the distribution and extent of five critical marine and coastal habitats within 500km² areas throughout the Gulf of California:

-   **Rocky reefs**: Hard-bottom submarine habitats supporting diverse marine life

-   **Mangroves**: Coastal forests that serve as crucial nursery habitats

-   **Sargassum**: Floating brown algae creating habitat for numerous species

-   **Coral reefs**: Biodiverse underwater ecosystems built by colonial cnidarians

-   **Seamounts**: Underwater mountains that create hotspots of marine biodiversity

The interactive map uses circle markers, the size of which indicates the habitat area (in hectares). You can toggle between habitat types to see their distribution patterns throughout the Gulf. Despite its low resolution, this dataset still provides a good framework for understanding habitat extents throughout the entire Gulf.

```{r}
#| warning: false
#| message: false
#| echo: false

hab_extent_500km2_sf <- read_sf(here::here(dir_M, "OHI_GOC/_raw_data/dataMares/hab_extent_500km2/int/hab_extent_500km2_wgs84.shp"))

unique_habitats <- unique(hab_extent_500km2_sf$habitat) # [1] "Rocky reef"  "Mangroves"   "Sargassum"   "Coral reefs" [5] "Seamounts"  
pal <- colorFactor( # make a color palette based on the 5 habitat types in the GoC
  palette = viridis(length(unique_habitats), option = "D"), # [1] "#440154FF" "#3B528BFF" "#21908CFF" "#5DC863FF" [5] "#FDE725FF"
  domain = unique_habitats
)

# baselayer leaflet map
habitat_map <- leaflet() %>%
  addProviderTiles("Esri.OceanBasemap", group = "Ocean Basemap") %>%
  addProviderTiles("CartoDB.Positron", group = "Light Map") %>%
  addProviderTiles("Esri.WorldImagery", group = "Satellite")

# add a layer for each habitat using a loop, later add layer controls
for (hab in unique_habitats) {
  hab_subset <- hab_extent_500km2_sf[hab_extent_500km2_sf$habitat == hab, ]
  habitat_map <- habitat_map %>%
    addCircleMarkers(
      data = hab_subset, 
      color = "white", # color of the border of each circle marker
      fillColor = pal(hab), # filled based on the color in the data for each circle marker
      weight = 1, # width of the border color
      radius = ~sqrt(area_ha)/5, # need to use the square root of that the area for the radius of the circle (bc of pi*r^2). Divided by 10 as a scaling factor for easier visualization (otherwise the circles are too big)
      fillOpacity = 0.8,
      stroke = TRUE,
      popup = ~paste(
        "<strong>ID:</strong>", id_reg, "<br>",
        "<strong>Habitat:</strong>", habitat, "<br>",
        "<strong>Area (ha):</strong>", round(area_ha, 2)
      ),
      group = hab
    )
}

habitat_map <- habitat_map %>%
  addLayersControl(
    baseGroups = c("Light Map", "Ocean Basemap", "Satellite"), # so the viewer can select whichever basemap is best for them
    overlayGroups = unique_habitats, # there is a group for each habitat
    options = layersControlOptions(collapsed = FALSE) # show it clearly 
  ) %>%
  addLegend(
    position = "bottomright",
    pal = pal,
    values = hab_extent_500km2_sf$habitat,
    title = "Habitat Types",
    opacity = 0.7
  )

habitat_map
```

#### Citation

Aburto-Oropeza, Octavio; Moreno-Baez, Marcia; Johnson, Andrew; Giron-Nava, Alfredo; Cisneros-Montemayor, Andres; Suarez, Alvin (2017). Mapping Ocean's Wealth - Habitats extent. In dataMares: Ecosystem Dynamics. UC San Diego Library Digital Collections. <https://doi.org/10.6075/J01N7Z9K>

**Contact Information:** Octavio Aburto ([maburto\@ucsd.edu](mailto:maburto@ucsd.edu))

Data Download: [**Mapping Ocean's Wealth - Habitats extent**](https://library.ucsd.edu/dc/object/bb6009652h)

### Wetlands of Importance (Ramsar List)

This map displays wetland sites within the Gulf of California region that have been recognized for their international significance under the Ramsar Convention on Wetlands. These wetlands have met strict criteria demonstrating their ecological, botanical, zoological, or hydrological importance. Each blue circle represents a designated Ramsar site, with the circle size corresponding to the wetland's area in hectares. Clicking on a site reveals:

-  The official site name

-  Date of designation as a Ramsar wetland

-  Total protected area in hectares

These sites form part of an international network of protected wetlands recognized for both their biodiversity value and their ecological functions that sustain human communities.

```{r}
#| warning: false
#| message: false
#| echo: false

ramsar_list_goc <- read_sf(here::here(dir_M, "OHI_GOC/_raw_data/ramsar_list/d2025/int/ramsar_list_goc_wgs84.shp"))


# baselayer leaflet map
wetland_map <- leaflet() %>%
  addProviderTiles("Esri.OceanBasemap", group = "Ocean Basemap") %>%
  addProviderTiles("CartoDB.Positron", group = "Light Map") %>%
  addProviderTiles("Esri.WorldImagery", group = "Satellite") %>% 
  
  addCircleMarkers(
    data = ramsar_list_goc, 
    color = "white", # color of the border of each circle marker
    fillColor = "royalblue", # filled based on the color in the data for each circle marker
    weight = 1, # width of the border color
    radius = ~sqrt(area_ha)/10, # need to use the square root of that the area for the radius of the circle (bc of pi*r^2). Divided by 10 as a scaling factor for easier visualization (otherwise the circles are too big)
    fillOpacity = 0.8,
    stroke = TRUE,
    popup = ~paste(
      "<strong>Site:</strong>", site, "<br>",
      "<strong>Date of Designation:</strong>", dt_f_ds, "<br>",
      "<strong>Area (ha):</strong>", area_ha
    )
  ) %>%
  addLayersControl(
    baseGroups = c("Light Map", "Ocean Basemap", "Satellite"), # so the viewer can select whichever basemap is best for them
    options = layersControlOptions(collapsed = FALSE) # show it clearly 
  )

wetland_map
```

#### Citation

The List of Wetlands of International Importance (the Ramsar List) download: <https://www.ramsar.org/document/list-wetlands-international-importance-ramsar-list>

Published: 14 April 2025

The Secretariat of the Convention on Wetlands (Ramsar, Iran, 1971)
Rue Mauverney 28, CH-1196 Gland, Switzerland
Tel. + 41 22 999 0170, e-mail <ramsar@ramsar.org>


### Mangrove Distribution

The below map shows the distribution of mangrove forests throughout the Gulf of California region based on data from the Sistema de Información y Análisis Marino Costero (SIMAR-Explorer) developed with CONABIO (the National Commission for the Knowledge and Use of Biodiversity).  The dark green polygons represent actual mangrove stands, showing:

-  Their geographic extent and distribution

-  Area coverage in hectares

-  Date of mapping/assessment

Mangroves serve as critical nursery habitat for fish, protect coastlines from erosion, sequester carbon, and support coastal livelihoods.

```{r}
#| warning: false
#| message: false
#| echo: false

sc_mangroves_cropped <- read_sf(here::here(dir_M, "OHI_GOC/_raw_data/SIMAR_CONABIO/d2025/hab_mangrove/int/sc_mangroves_cropped_wgs84.shp"))

# mapview::mapview(sc_mangroves_cropped, layer.name = "Mangrove Coverage")

mangrove_map <- leaflet() %>%
  addProviderTiles("Esri.OceanBasemap", group = "Ocean Basemap") %>% # add base layers
  addProviderTiles("CartoDB.Positron", group = "Light Map") %>%
  addProviderTiles("Esri.WorldImagery", group = "Satellite") %>% 
  
  # add the polygons to the map
  addPolygons(
    data = sc_mangroves_cropped,
    color = "black",
    fillColor = "darkgreen",
    fillOpacity = 0.6,
    weight = 1,
    opacity = 0.5,
    fill = TRUE,
    popup = ~paste("<b>Area (ha):</b>", HA, "<br>",
                  "<b>Date:</b>", FECHA)
    ) %>%

  # base layer control
  addLayersControl(
    baseGroups = c("Light Map", "Ocean Basemap", "Satellite"),
    options = layersControlOptions(collapsed = FALSE)
  )

mangrove_map
```

#### Citation

Sistema de Información y Análisis Marino Costero (SIMAR-Explorer) and National Commission for the Knowledge and Use of Biodiversity (CONABIO)

Links to download and explore:
<https://simar.conabio.gob.mx/explorer/>

<https://fmcn.org/en/proyectos/marine-and-coastal-monitoring>

<https://fmcn.org/es/noticia/distribucion-de-mangle-rojo-en-el-noroeste-de-mexico-y-la-peninsula-de-yucatan>

Data ➤ Bases Cartográficas ➤ Mangrove Coverage ➤ Distribution of Mangroves in México for 2020 (CONABIO)




