---
title: "Mangrove - Gulf of California OHI"
author: "Austin Martinez"
date: 08/03/2026
format: 
  html:
    code-fold: show
execute:
  echo: true
  message: false
  warning: false
---

```{r}
# Load in Libraries
library(sf)
library(dplyr)
library(here)
library(terra)
library(tidyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(tmap)
```

```{r}
here::i_am("mangrove_goc_ohi.qmd")

# Change this to any year with a completed download (1984–2023)
MANGROVE_YEAR <- 2023L
```

```{r}
# Read in OHI region boundaries
Sys.setenv(SHAPE_RESTORE_SHX = "YES")

ohi_regions <- st_read(here("data", "goc_ohi_rgns.shp")) |>
  st_make_valid()

if (is.na(st_crs(ohi_regions))) {
  ohi_regions <- st_set_crs(ohi_regions, 6372)
}

if ("id" %in% names(ohi_regions)) {
  ohi_regions <- ohi_regions |>
    mutate(region_id = as.integer(id))
} else {
  ohi_regions <- ohi_regions |>
    mutate(region_id = row_number())
}

ohi_regions_marine <- if ("location" %in% names(ohi_regions)) {
  ohi_regions |> filter(location == "marine")
} else {
  ohi_regions
}

ohi_regions_marine
```

## Read annual mangrove extent (vector)

Data source: Continuous Global Mangrove Dynamics (Zenodo 10.5281/zenodo.20617284).
Each year is an ESRI Shapefile in `projected_YYYY.zip` (Behrmann equal-area).

```{r}
mangrove_zip <- here(
  "data", "mangroves",
  paste0("projected_", MANGROVE_YEAR, ".zip")
)

stopifnot(file.exists(mangrove_zip))

shp_name <- grep(
  "\\.shp$",
  utils::unzip(mangrove_zip, list = TRUE)$Name,
  value = TRUE
)[1]

mangrove_path <- paste0("/vsizip/", normalizePath(mangrove_zip), "/", shp_name)

mangrove_global <- st_read(mangrove_path, quiet = TRUE) |>
  st_make_valid()

mangrove_global
```

```{r}
# gridcode: 1 = valid Landsat observation, 2 = gap-filled
table(mangrove_global$gridcode)
```

## Clip mangroves to OHI study area

```{r}
mangrove_goc <- mangrove_global |>
  st_transform(st_crs(ohi_regions_marine)) |>
  st_intersection(ohi_regions_marine)

mangrove_goc
```

## Mangrove area (km²) by OHI region

Use the dataset `area_km2` field (30-m polygon footprints in Behrmann projection).

```{r}
region_mangrove <- mangrove_goc |>
  st_drop_geometry() |>
  group_by(region_id) |>
  summarise(
    mangrove_km2      = sum(area_km2, na.rm = TRUE),
    valid_km2         = sum(area_km2[gridcode == 1L], na.rm = TRUE),
    gapfilled_km2     = sum(area_km2[gridcode == 2L], na.rm = TRUE),
    n_polygons        = n(),
    .groups = "drop"
  ) |>
  right_join(
    ohi_regions_marine |> st_drop_geometry() |> select(region_id),
    by = "region_id"
  ) |>
  mutate(across(where(is.numeric), ~ replace_na(.x, 0)))

region_mangrove
```

```{r}
region_mangrove |>
  kable(
    digits = 2,
    col.names = c(
      "Region ID", "Total mangrove (km²)", "Valid obs. (km²)",
      "Gap-filled (km²)", "Polygon count"
    ),
    align = c("c", "r", "r", "r", "r")
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "bordered", "hover"),
    full_width = FALSE
  ) |>
  add_header_above(
    setNames(5, paste0("Mangrove extent — ", MANGROVE_YEAR))
  )
```

## Map mangrove polygons inside OHI regions

```{r}
tmap_mode("view")

tm_shape(ohi_regions_marine) +
  tm_borders(col = "white", lwd = 2) +
  tm_shape(mangrove_goc) +
  tm_fill(
    col = "gridcode",
    palette = c("1" = "darkgreen", "2" = "yellowgreen"),
    labels = c("Valid Landsat", "Gap-filled"),
    alpha = 0.7,
    title = "Mangrove source"
  ) +
  tm_borders(col = "darkgreen", lwd = 0.3) +
  tm_layout(
    legend.outside = TRUE,
    frame = FALSE
  )
```

## Join summary to region polygons

```{r}
ohi_regions_map <- ohi_regions_marine |>
  left_join(region_mangrove, by = "region_id")

tmap_mode("view")

tm_shape(ohi_regions_map) +
  tm_polygons(
    col = "mangrove_km2",
    palette = "viridis",
    title = paste0("Mangrove area ", MANGROVE_YEAR, " (km²)"),
    alpha = 0.9,
    border.col = "white",
    border.lwd = 1
  ) +
  tm_layout(
    legend.outside = TRUE,
    frame = FALSE
  )
```

## Optional: loop all downloaded years for a time series

```{r}
available_years <- list.files(
  here("data", "mangroves"),
  pattern = "^projected_[0-9]{4}\\.zip$",
  full.names = FALSE
) |>
  sub("^projected_|\\.zip$", "", x = _) |>
  as.integer() |>
  sort()

available_years
```

```{r}
read_mangrove_year <- function(year) {
  zip <- here("data", "mangroves", paste0("projected_", year, ".zip"))
  shp <- grep("\\.shp$", utils::unzip(zip, list = TRUE)$Name, value = TRUE)[1]
  path <- paste0("/vsizip/", normalizePath(zip), "/", shp)

  st_read(path, quiet = TRUE) |>
    st_make_valid() |>
    st_transform(st_crs(ohi_regions_marine)) |>
    st_intersection(ohi_regions_marine) |>
    st_drop_geometry() |>
    mutate(year = year)
}

mangrove_ts <- lapply(available_years, function(y) {
  tryCatch(read_mangrove_year(y), error = function(e) NULL)
}) |>
  bind_rows()

mangrove_ts_summary <- mangrove_ts |>
  group_by(region_id, year) |>
  summarise(
    mangrove_km2 = sum(area_km2, na.rm = TRUE),
    .groups = "drop"
  )

mangrove_ts_summary
```

```{r, fig.width=12, fig.height=7}
ggplot(mangrove_ts_summary, aes(x = year, y = mangrove_km2, group = region_id, color = factor(region_id))) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Mangrove extent time series by OHI region",
    x = "Year",
    y = "Mangrove area (km²)",
    color = "Region ID"
  ) +
  theme_minimal(base_size = 14)
```

## Redundancy check

```{r}
cat("Mangrove year:", MANGROVE_YEAR, "\n")
cat("Total mangrove area in study area (km²):", sum(region_mangrove$mangrove_km2), "\n")
cat("Downloaded years available:", paste(available_years, collapse = ", "), "\n")
```
