---
title: "Tourist data"
output: html_document
date: "2025-3-6"
editor_options: 
  chunk_output_type: console
---
Data: Temporary accommodation and food and beverage preparation services and with the number of employees
Location: https://www.inegi.org.mx/app/descarga/?ti=6

There are two files, this just splits up the states.

Another option is the number of tourists. Data are available for this as well, but I don't think it is available for states, and it only includes foreign visits.

https://github.com/OHI-GOC/OHI_GOC_issues/issues/13

```{r setup, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}

library(mapview)
library(sf)
library(tidyverse)
library(here)
library(terra)

```

Attribute	Description
id	        Unique identification number of the DENUE record. It's a unique numeric key for each record in the National Statistical Directory of Economic Units (DENUE).

nom_estab	  Commercial name of the economic unit (business name as displayed to the public).  If none, the generic name of the activity is used.

raz_social	Legal or corporate name of the entity (for corporations). For individuals, this is omitted for privacy.

codigo_act	Economic activity code assigned based on the SCIAN 2018 classification, representing the main activity of the unit.

nombre_act	Name of the economic activity according to SCIAN 2018.

per_ocu	Number of employees, grouped into ranges: 1 = 0 to 5, 2 = 6 to 10, 3 = 11 to 30, 4 = 31 to 50, 5 = 51 to 100, 6 = 101 to 250, 7 = 251 and more.

tipo_vial	Type of road where the economic unit is located (street, avenue, walkway, etc.).

nom_vial	Name of the road.

tipo_v_e_1/2/3	Additional road types (for intersections or corners).

nom_v_e_1/2/3	Additional road names.

numero_ext	Exterior number of the building.

letra_ext	Exterior letter of the building (if applicable).

edificio	Building name.

edificio_e	Floor or level of the building.

numero_int	Interior number (unit or suite number).

letra_int	Interior letter.

tipo_asent	Type of settlement (colony, neighborhood, etc.).

nomb_asent	Name of the settlement.

tipoCenCom	Type of commercial center (mall, plaza, etc.).

nom_CenCom	Name of the commercial center.

num_local	Local unit number (within a plaza or market).

cod_postal	Postal code.

cve_ent	State code.

entidad	Name of the state.

cve_mun	Municipality code.

municipio	Name of the municipality.

cve_loc	Locality code.

localidad	Name of the locality.

ageb	Basic Geo-statistical Area (AGEB) code.

manzana	Block code (geo-statistical).

telefono	Phone number of the business.

correoelec	Email address of the business.

www	Website URL of the business.

tipoUniEco	Type of economic unit: Fixed or Semi-fixed.

latitud	Latitude of the business location.

longitud	Longitude of the business location.

fecha_alta	Date the business was added to DENUE and the National Business Statistical Registry (RENEM).

clee	Unique statistical identification key assigned by INEGI to identify and link establishments belonging to the same company.


## Hotel data
Get the hotel data and select only the GofC regions.

```{r}

hot1 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue_hotels/denue_00_72_1_shp/conjunto_de_datos/denue_inegi_72_1.shp")
hot2 <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue_hotels/denue_00_72_2_shp/conjunto_de_datos/denue_inegi_72_2.shp")

regions <- c("Baja California", "Baja California Sur", "Sonora", "Sinaloa", "Nayarit", "Jalisco")

hot1 <- hot1 %>%
  filter(entidad %in% regions)
hot2 <- hot2 %>%
  filter(entidad %in% regions)

hots <- rbind(hot1, hot2)

```

Now identify the accommodation data. Cut the food data for now. (although consider whether this could be relevant)

```{r}
accom_cats <- read_csv(here("TR/data/tourism_categories.csv")) %>%
  filter(Accommodation == 1) %>%
  pull(codigo_act)

hots <- hots %>%
  filter(codigo_act %in% accom_cats) %>%
  select(nom_estab, per_ocu, entidad, municipio, correoelec, www, latitud, longitud)

size <- data.frame(per_ocu = c("0 a 5 personas", "101 a 250 personas", "11 a 30 personas",  "251 y m\xe1s personas",
                               "31 a 50 personas", "51 a 100 personas", "6 a 10 personas"),
                   med_ocu = c(3, 175, 20, 250, 40, 75, 8),
                   rank_ocu = c(1, 6, 3, 7, 4, 5, 2))
hots <- hots %>%
  left_join(size, by="per_ocu")

```

Plot

```{r}
library(leafpop)

hots_clean <- hots |> 
  mutate(across(where(is.character), ~ iconv(.x, from = "latin1", to = "UTF-8", sub = "")))


mapview(hots_clean)

mapview(
  hots_clean,
  zcol = "rank_ocu",         # colors based on rank_ocu (optional)
  cex = hots$rank_ocu,   # scale the point size
  popup = popupTable(hots_clean, 
                     zcol = c("nom_estab", "per_ocu", "entidad", 
                               "municipio", "correoelec", "www", 
                               "med_ocu", "rank_ocu")),
  legend = TRUE
)
```
