---
title: 'OHI GoC 2025: SUSTAINABILITY Tourism & Recreation'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# TR Sustainability

## Blue Flag Beaches

This indicator will reflect the goal that all touristy beaches are managed by the Blue Flag certification. [This document](https://blueflag.us/wp-content/uploads/2021/11/USA-National-Beach-Criteria-2021_track-changes_11.21.pdf) details what it means for a beach to be compliant with the Blue Flag program, and it hits everything that I think is really important (infrastructure availability such as available trashcans, environmental education, litter/plastic cleanups on beaches, safety of beachgoers, water quality, free access/only small fees for parking etc, management of different users on the beach, toilet or restroom facilities with controlled sewage disposals, sensitive areas and species must be protected and managed, and more!).

So, current status = [the km length of touristy beaches with a blue flag] / [the total km of touristy beaches in the Gulf] * 100

The key here is to determine which beaches *should* have this certification (aka, which beaches are urban/touristy), so that we can find the *length* of beaches that need to be protected. 


### Make a list of all GoC coastal cities

- Los Cabos
- La Paz
- Loreto ? 
- Mulege ?
- San Felipe
- Puerto Peñasco
- Bahía de Kino
- Guaymas
- Huatabampo
- Topolobampo
- Mazatlán
- La Peñita de Jaltemba
- Sayulita
- Bahía de Banderas
  - Punta Mita
  - La Cruz de Huanacaxtle
  - Bucerías
  - Nuevo Nayarit/Las Jarretaderas
- Puerto Vallarta
- Boca de Tomatlan
- Boca Yelapa
- Chimo (cabo corrientes)
  


```{r}
mun_w_rgnids <- read_csv(here::here("/home/shares/ohi/OHI_GOC/spatial/municipality_to_rgnid","municipality_to_rgnid.csv"))

## main cities/towns
cities_m <- st_read( "/home/shares/ohi/OHI_GOC/goal_prep/pressures/population_localities/inegi_boundaries_pop.shp")  %>% filter(AMBITO=="Urbana") 
cities <- st_read( "/home/shares/ohi/OHI_GOC/goal_prep/pressures/population_localities/inegi_boundaries_pop.shp") 

mapview(cities_buf) + mapview(cities_m)  

muns <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/00mun.shp") %>% st_transform(crs=4326)

#mapview(muns) + mapview(cities) 

## beaches
coastline <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_coastline/conjunto_de_datos_2024/conjunto_de_datos/lc_rep_mex_v_2024.shp") 
#beach <-  coastline %>% filter(SUBTIPO %in% c("Playa") & TIPO=="Continental") %>%
#  st_transform(crs=4326)
#beach$ID_Beach <- 1:nrow(beach)
#st_write(beach, here("_pressures/access/data/beach.shp"))
beach <- st_read(here("_pressures/access/data/beach.shp"))
```

```{r}
## join mun_w_rgnids and cities
cities_w_rgnids <- cities_m %>% # all "urban" cities according to INEGI
  mutate(E03 = as.numeric(CVE_ENT),
         E04 = as.numeric(CVE_MUN)) %>% 
  inner_join(mun_w_rgnids) %>% 
  filter(!is.na(rgn_id))

mapview(cities_w_rgnids) ## still some not coastal, lets limit to inland_50


################### including rural ####################
## join mun_w_rgnids and cities
cities_w_rgnids_RUR <- cities %>% # all "urban" cities according to INEGI
  mutate(E03 = as.numeric(CVE_ENT),
         E04 = as.numeric(CVE_MUN)) %>% 
  inner_join(mun_w_rgnids) %>% 
  filter(!is.na(rgn_id))

mapview(cities_w_rgnids_RUR) ## still some not coastal, lets limit to inland_50
```


```{r}
inland_50 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer","inland_50km_goc.shp"))

cities_w_rgnids_trans <- st_transform(cities_w_rgnids, st_crs(inland_50))
cities_w_rgnids_50km <- st_intersection(cities_w_rgnids_trans, inland_50) %>% 
  st_make_valid()

mapview(cities_w_rgnids_50km) # might need to remove Caborca and others that seem unreasonable to include as touristy beaches later on.


################### including rural ####################
cities_w_rgnids_trans_RUR <- st_transform(cities_w_rgnids_RUR, st_crs(inland_50))
cities_w_rgnids_50km_RUR <- st_intersection(cities_w_rgnids_trans_RUR, inland_50) %>% 
  st_make_valid()

mapview(cities_w_rgnids_50km_RUR) 

```


### Make a buffer around those cities

```{r}
cities_buf <- st_buffer(cities_w_rgnids_50km, dist = 500) # reminder: inland_50 is in CRS = Mexico ITRF2008 / LCC , which is a metric projection, so we can interpret the dist = 500 as meters.  

mapview(cities_w_rgnids_50km) + mapview(cities_buf)

# merge / dissolve all buffers into one geometry
buf_merged <- st_union(cities_buf)
cities_buf_2 <- st_as_sf(buf_merged) %>% st_transform(crs=4326)

mapview(cities_w_rgnids_50km) + mapview(cities_buf_2)


################### including rural ####################
cities_buf_RUR <- st_buffer(cities_w_rgnids_50km_RUR, dist = 500) # reminder: inland_50 is in CRS = Mexico ITRF2008 / LCC , which is a metric projection, so we can interpret the dist = 500 as meters.  

mapview(cities_w_rgnids_50km_RUR) + mapview(cities_buf_RUR)

# merge / dissolve all buffers into one geometry
buf_merged_RUR <- st_union(cities_buf_RUR)
cities_buf_2_RUR <- st_as_sf(buf_merged_RUR) %>% st_transform(crs=4326)

mapview(cities_w_rgnids_50km_RUR) + mapview(cities_buf_2_RUR)
```


### Crop beaches to the city buffers

```{r}
beach_urban <- st_intersection(beach, cities_buf_2)

mapview(beach_urban, color = "magenta") + mapview(cities_buf_2, col.regions ="lightblue") + mapview(cities_w_rgnids_50km)

st_write(beach_urban, here::here("/home/shares/ohi/OHI_GOC/spatial/urban_beaches","urban_beaches_500m_buffer_INEGI.shp"))

################### including rural ####################
beach_urban_RUR <- st_intersection(beach, cities_buf_2_RUR)

mapview(beach_urban_RUR, color = "magenta") + mapview(cities_buf_2_RUR, col.regions ="lightblue") + mapview(cities_w_rgnids_50km_RUR)

```



### Read in Blue Flag Beaches

```{r}
library(tidyverse)
library(here)
library(janitor)
library(sf)
library(mapview)
```

```{r}
## read in the data taken from the website
blue_flag_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/FEE_foundation_environmental_education/blueflag_mexico", "blueflag_beaches_03_10_26.csv"))

blueflag_sf <- blue_flag_raw %>% 
  janitor::clean_names() %>% 
  st_as_sf(coords = c("long", "lat"), # making sf object
           crs = 4326, # WGS84
           remove = FALSE) %>% 
  filter(tipo %in% c("playa del mar", "marina del mar"))

mapview(blueflag_sf, color = "blue") + mapview(beach_urban, color = "magenta") + mapview(cities_buf_2, col.regions ="lightblue")


################### including rural ####################
mapview(blueflag_sf, col.region = "blue") + mapview(beach_urban_RUR, color = "magenta") + mapview(cities_w_rgnids_50km_RUR, zcol ="AMBITO")
```

### ** Use Urban for now

### Assign ID_Beach to Blue Flag beaches

Now that we have a registry of the urban/touristy beaches, we need to match up the BlueFlag coordinates to the `ID_Beach` in the beach_urban sf. 

That way, we can use `st_length()` to compare the length of beaches protected under the Blue Flag certification vs. total length of touristy beaches.

```{r}
mapview(blueflag_sf, color = "blue") + mapview(beach_urban, color = "magenta")


blueflag_buffer <- st_buffer(blueflag_sf, dist = 400)
mapview(blueflag_buffer) + mapview(beach_urban, color = "magenta")

# there are some that need to be done manually because there are a couple of beach segments within the buffer, and sometimes the coordinates of the BF beach is actually a little further away from the beach than it really is.  For example, "Hotel Marriott Puerto Vallarta Resort & Spa" upon further research applies to ID_Beach 1417 and not the other small segmented beaches nearby (ex: 2000) because those are in front of different hotels and Marriott's BF certification would not stretch there.

# because of this, I will walk through each one and assign the ID_Beach...
```


```{r}
library(tibble)
library(readr)
  
bf_beaches <- tibble( # bf = blue flag
  nombre = c(
    "Hotel Marriott Puerto Vallarta Resort & Spa",
    "Hotel The Westin Resort & Spa Puerto Vallarta",
    "Sheraton",
    "Playa Camarones",
    "Nuevo Nayarit",
    "Lo de Marcos",
    "Marina Riviera Nayarit",
    "Marina Riviera Nayarit",
    "Esmeralda Resort",
    "La Gaviota",
    "Playa Paraíso Leonero",
    "El Surgidero",
    "La Ribera",
    "Playa La Playa",
    "Acapulquito",
    "Palmilla",
    "Grand Velas Los Cabos",
    "Santa María",
    "El Chileno",
    "Grand Fiesta Americana Los Cabos",
    "Las Viudas",
    "Médano Riu Palace Cabo San Lucas",
    "Médano Riu Santa Fe",
    "Médano Riu Palace Baja California",
    "Médano Villa del Palmar",
    "Médano Villa La Estancia",
    "Médano Villa del Arco",
    "Médano Club Cascadas de Baja",
    "Médano Pueblo Bonito Rose",
    "Médano Pueblo Bonito Los Cabos",
    "Médano ME Cabo",
    "Médano Casa Dorada",
    "Playa Médano Corazón",
    "Hacienda",
    "Médano Breathless",
    "El Corsario"
  ),
  ID_Beach = c(
    1417, 2007, 2019, 2022, 1233, 1379, 1412, 1413,
    483, 243, 243, 243, 625, 230, 92, 228, 223,
    NA, NA, 219, 214, 214, 214, 214, 214, 214, 214,
    214, 214, 214, 214, 214, 214, 214, 214, 293
  ),
  AMBITO = c(
    "Urbana", "Urbana", "Urbana", "Urbana", "Rural", "Rural",
    "Marina", "Marina", "Urbana", "Rural", "Rural", "Rural",
    "Rural", "Urbana", "Urbana", "Urbana", "Urbana", "Urbana",
    "Urbana", "Urbana", "Urbana", "Urbana", "Urbana", "Urbana",
    "Urbana", "Urbana", "Urbana", "Urbana", "Urbana", "Urbana",
    "Urbana", "Urbana", "Urbana", "Urbana", "Urbana", "Urbana"
  )
) %>% 
  mutate(certified = "yes") %>% 
  rename(blue_flag_nombre = nombre)
```


### Join so that the urban beach dataset indicates whether it is blue flag certified or not

```{r}
## first, add any beaches that are rural but are blue flag certified already into the `beach_urban` df.
bf_rural_beaches <- beach %>% 
  filter(ID_Beach %in% c("1233","1379","243","625"))

st_crs(beach_urban) == st_crs(bf_rural_beaches) # TRUE

beach_urban_complete <- bind_rows(beach_urban, bf_rural_beaches) # join the bf rural beaches
mapview(beach_urban_complete, color = "magenta")


```


```{r}
## join with the tibble indicating whether it is a bf certified beach in 2025
bf_w_beach_urban_complete <- full_join(beach_urban_complete, bf_beaches, relationship = "many-to-many") %>% 
  janitor::clean_names() %>% 
  mutate(ambito = ifelse(is.na(ambito), "Urbana", ambito),
         certified = ifelse(is.na(certified), "no", certified)) %>% 
  dplyr::select(id_beach, ambito, certified, blue_flag_nombre, geometry) %>% 
  st_as_sf()

mapview(bf_w_beach_urban_complete, color = "magenta")
```


Since there are multiple blue flags for some beaches, we can just call a beach certified if it has at least one BF certification:
```{r}
bf_w_beach_urban_complete_nodups <- bf_w_beach_urban_complete %>% 
  group_by(id_beach, ambito, certified, geometry) %>%
  summarise(blue_flag_nombre = paste(blue_flag_nombre, collapse = "; "),
            .groups = "drop") %>%
  st_as_sf() %>% 
  st_cast("MULTILINESTRING") %>% # to ensure they are the same types
  mutate(length_m = st_length(geometry)) # find each beach's length in meters

# st_write(bf_w_beach_urban_complete_nodups, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int/blue_flag_int","all_urban_beaches_plus_bf_rural_beaches.gpkg"))
  
mapview(bf_w_beach_urban_complete_nodups, color = "magenta")

st_geometry_type(bf_w_beach_urban_complete_nodups) %>% unique()
# should return LINESTRING or MULTILINESTRING: returned MULTILINESTRING because I used st_cast

mapview(bf_w_beach_urban_complete_nodups, color = "magenta") + mapview(blueflag_sf, color = "blue") + mapview(cities_buf_2, col.regions ="lightblue")
```

```{r}
## intersect with OHI rgn polygons for rgn_id
goc_ohi_rgns <- st_read(here::here("spatial/ohi_regions","goc_ohi_rgns.shp")) %>% 
  dplyr::select(rgn_id) %>%
  group_by(rgn_id) %>%
  summarise(geometry = st_union(geometry)) %>%
  ungroup()
mapview(goc_ohi_rgns)

bf_w_beach_urban_complete_rgnid <- st_intersection(st_transform(bf_w_beach_urban_complete_nodups, st_crs(goc_ohi_rgns)), goc_ohi_rgns)

mapview(bf_w_beach_urban_complete_rgnid, zcol = "rgn_id") + mapview(blueflag_sf, color = "blue") + mapview(cities_buf_2, col.regions ="lightblue")

mapview(bf_w_beach_urban_complete_rgnid, zcol = "certified")

# st_write(bf_w_beach_urban_complete_rgnid, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int/blue_flag_int","blue_flag_w_urban_beaches_length_meters_rgn_id.gpkg"))
```

### Current status

current status = [the m length of touristy beaches with a blue flag] / [the total m of touristy beaches in the Gulf] * 100

```{r}
blue_flag_current_status <- bf_w_beach_urban_complete_rgnid %>%
  st_drop_geometry() %>%
  mutate(length_m = as.numeric(length_m)) %>%   # drop units before pivot
  group_by(rgn_id, certified) %>%
  summarize(total_length = sum(length_m, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(
    names_from = certified,
    values_from = total_length,
    values_fill = 0
  ) %>%
  mutate(total_length = no + yes,
         prop_blue_flag = yes / total_length,
         current_status = prop_blue_flag * 100) %>% 
  mutate(year = 2025)

# write_csv(blue_flag_current_status, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","blue_flag_current_status_final_031626.csv"))

```

## Questions to resolve:

(1) What should we do about the blue flag beaches Santa María and El Chileno?  INEGI does not register that part of the coastline as a beach, but it does look like one on the ESRI World Imagery. 

(2) Should we limit the beaches we consider "touristy" further by filtering out certain coastal cities that we would not consider highly visited, or needing to be blue flag beaches? EX: Loreto and other cities in Region_3?

(3) If they have zero BF beaches, should it absolutely have a current status of zero? 

(4) Some BF beaches have been BF since 2015, for example, some since 2025.  I need to go through and add the year that each one became Blue Flag.  The reason why I did not do that at first is because it would take a bit of time to walk through each one, but I will do that. 


