---
title: 'OHI GoC 2025: SUSTAINABILITY Tourism & Recreation'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# TR Sustainability

## Blue Flag Beaches

This indicator will reflect the goal that all touristy beaches are managed by the Blue Flag certification. [This document](https://blueflag.us/wp-content/uploads/2021/11/USA-National-Beach-Criteria-2021_track-changes_11.21.pdf) details what it means for a beach to be compliant with the Blue Flag program, and it hits everything that I think is really important (infrastructure availability such as available trashcans, environmental education, litter/plastic cleanups on beaches, safety of beachgoers, water quality, free access/only small fees for parking etc, management of different users on the beach, toilet or restroom facilities with controlled sewage disposals, sensitive areas and species must be protected and managed, and more!).

So, current status = [the m length of touristy beaches WITH a blue flag] / [the total m of touristy beaches in the Gulf] * 100

The key here is to determine which beaches *should* have this certification (aka, which beaches are urban/touristy), so that we can find the *length* of beaches that need to be protected. 


### Make a list of all GoC coastal cities


This list specifically is not used yet, but if we want to do manual cuts, these are the ones I could include:
- Los Cabos
- La Paz
- Loreto ? 
- Mulege ?
- San Felipe
- Puerto Peñasco
- Bahía de Kino
- Guaymas
- Huatabampo
- Topolobampo
- Mazatlán
- La Peñita de Jaltemba
- Sayulita
- Bahía de Banderas
  - Punta Mita
  - La Cruz de Huanacaxtle
  - Bucerías
  - Nuevo Nayarit/Las Jarretaderas
- Puerto Vallarta
- Boca de Tomatlan
- Boca Yelapa
- Chimo (cabo corrientes)
  


```{r}
mun_w_rgnids <- read_csv(here::here("/home/shares/ohi/OHI_GOC/spatial/municipality_to_rgnid","municipality_to_rgnid.csv"))

## main cities/towns
cities_m <- st_read( "/home/shares/ohi/OHI_GOC/goal_prep/pressures/population_localities/inegi_boundaries_pop.shp")  %>% filter(AMBITO=="Urbana") 
cities <- st_read( "/home/shares/ohi/OHI_GOC/goal_prep/pressures/population_localities/inegi_boundaries_pop.shp") 

mapview(cities_buf) + mapview(cities_m)  

muns <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_boundaries/889463807469_s/conjunto_de_datos/00mun.shp") %>% st_transform(crs=4326)

#mapview(muns) + mapview(cities) 

## beaches
coastline <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/inegi_coastline/conjunto_de_datos_2024/conjunto_de_datos/lc_rep_mex_v_2024.shp") 
#beach <-  coastline %>% filter(SUBTIPO %in% c("Playa") & TIPO=="Continental") %>%
#  st_transform(crs=4326)
#beach$ID_Beach <- 1:nrow(beach)
#st_write(beach, here("_pressures/access/data/beach.shp"))
beach <- st_read(here("_pressures/access/data/beach.shp"))

beach2 <- coastline %>% 
  filter(SUBTIPO %in% c("Playa","Acantilado") & TIPO == "Continental") %>%
  st_transform(crs=4326)
beach2$ID_Beach <- 1:nrow(beach2)
mapview(beach2)

# subtipo Construcción portuaria
ports <- coastline %>% 
  filter(SUBTIPO %in% c("Construcción portuaria") & TIPO=="Continental") %>%
  st_transform(crs=4326)


```

```{r}
## join mun_w_rgnids and cities
cities_w_rgnids <- cities_m %>% # all "urban" cities according to INEGI
  mutate(E03 = as.numeric(CVE_ENT),
         E04 = as.numeric(CVE_MUN)) %>% 
  inner_join(mun_w_rgnids) %>% 
  filter(!is.na(rgn_id))

mapview(cities_w_rgnids) ## still some not coastal, lets limit to inland_50


################### including rural ####################
## join mun_w_rgnids and cities
cities_w_rgnids_RUR <- cities %>% # all "urban" cities according to INEGI
  mutate(E03 = as.numeric(CVE_ENT),
         E04 = as.numeric(CVE_MUN)) %>% 
  inner_join(mun_w_rgnids) %>% 
  filter(!is.na(rgn_id))

mapview(cities_w_rgnids_RUR) ## still some not coastal, lets limit to inland_50
```


```{r}
inland_50 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer","inland_50km_goc.shp"))

cities_w_rgnids_trans <- st_transform(cities_w_rgnids, st_crs(inland_50))
cities_w_rgnids_50km <- st_intersection(cities_w_rgnids_trans, inland_50) %>% 
  st_make_valid()

mapview(cities_w_rgnids_50km) # might need to remove Caborca and others that seem unreasonable to include as touristy beaches later on.


################### including rural ####################
cities_w_rgnids_trans_RUR <- st_transform(cities_w_rgnids_RUR, st_crs(inland_50))
cities_w_rgnids_50km_RUR <- st_intersection(cities_w_rgnids_trans_RUR, inland_50) %>% 
  st_make_valid()

mapview(cities_w_rgnids_50km_RUR) 

```


### Make a buffer around those cities

```{r}
cities_buf <- st_buffer(cities_w_rgnids_50km, dist = 500) # reminder: inland_50 is in CRS = Mexico ITRF2008 / LCC , which is a metric projection, so we can interpret the dist = 500 as meters.  

mapview(cities_w_rgnids_50km) + mapview(cities_buf)

# merge / dissolve all buffers into one geometry
buf_merged <- st_union(cities_buf)
cities_buf_2 <- st_as_sf(buf_merged) %>% st_transform(crs=4326)

mapview(cities_w_rgnids_50km) + mapview(cities_buf_2)


################### including rural ####################
cities_buf_RUR <- st_buffer(cities_w_rgnids_50km_RUR, dist = 500) # reminder: inland_50 is in CRS = Mexico ITRF2008 / LCC , which is a metric projection, so we can interpret the dist = 500 as meters.  

mapview(cities_w_rgnids_50km_RUR) + mapview(cities_buf_RUR)

# merge / dissolve all buffers into one geometry
buf_merged_RUR <- st_union(cities_buf_RUR)
cities_buf_2_RUR <- st_as_sf(buf_merged_RUR) %>% st_transform(crs=4326)

mapview(cities_w_rgnids_50km_RUR) + mapview(cities_buf_2_RUR)
```


### Crop beaches to the city buffers

```{r}
beach_urban <- st_intersection(beach, cities_buf_2)

mapview(beach_urban, color = "magenta") + mapview(cities_buf_2, col.regions ="lightblue") + mapview(cities_w_rgnids_50km)

st_write(beach_urban, here::here("/home/shares/ohi/OHI_GOC/spatial/urban_beaches","urban_beaches_500m_buffer_INEGI.shp"))

################### including rural ####################
beach_urban_RUR <- st_intersection(beach, cities_buf_2_RUR)

mapview(beach_urban_RUR, color = "magenta") + mapview(cities_buf_2_RUR, col.regions ="lightblue") + mapview(cities_w_rgnids_50km_RUR)

```


### Read in Blue Flag Beaches

```{r}
library(tidyverse)
library(here)
library(janitor)
library(sf)
library(mapview)
```

```{r}
## read in the data taken from the website
blue_flag_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/FEE_foundation_environmental_education/blueflag_mexico", "blueflag_beaches_03_10_26.csv"))

### with the year since becoming blue flag
blue_flag_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/FEE_foundation_environmental_education/blueflag_mexico", "blueflag_beaches_03_17_26.csv"))

blueflag_sf <- blue_flag_raw %>% 
  janitor::clean_names() %>% 
  st_as_sf(coords = c("long", "lat"), # making sf object
           crs = 4326, # WGS84
           remove = FALSE) %>% 
  filter(tipo %in% c("playa del mar", "marina del mar"))

mapview(blueflag_sf, color = "blue") + mapview(beach_urban, color = "magenta") + mapview(cities_buf_2, col.regions ="lightblue")


################### including rural ####################
mapview(blueflag_sf, col.region = "blue") + mapview(beach_urban_RUR, color = "magenta") + mapview(cities_w_rgnids_50km_RUR, zcol ="AMBITO")
```

### ** Use Urban for now

### Assign ID_Beach to Blue Flag beaches

Now that we have a registry of the urban/touristy beaches, we need to match up the BlueFlag coordinates to the `ID_Beach` in the beach_urban sf. 

That way, we can use `st_length()` to compare the length of beaches protected under the Blue Flag certification vs. total length of touristy beaches.

```{r}
mapview(blueflag_sf, color = "blue") + mapview(beach_urban, color = "magenta")


blueflag_buffer <- st_buffer(blueflag_sf, dist = 400)
mapview(blueflag_buffer) + mapview(beach_urban, color = "magenta")

# there are some that need to be done manually because there are a couple of beach segments within the buffer, and sometimes the coordinates of the BF beach is actually a little further away from the beach than it really is.  For example, "Hotel Marriott Puerto Vallarta Resort & Spa" upon further research applies to ID_Beach 1417 and not the other small segmented beaches nearby (ex: 2000) because those are in front of different hotels and Marriott's BF certification would not stretch there.

# because of this, I will walk through each one and assign the ID_Beach...
```

<https://docs.google.com/spreadsheets/d/1tVsSwANpZAdqtuugBZJ50l_xPl6uVBwOE0geRvmWjXU/edit?gid=761823854#gid=761823854>

```{r}
#read in %>% 
  mutate(certified = "yes") %>% 
  left_join(blueflag_sf, by = c("blue_flag_nombre" = "nombre")) %>% 
  st_drop_geometry() %>% 
  dplyr::select(-geometry)
```


### Join so that the urban beach dataset indicates whether it is blue flag certified or not

```{r}
## first, add any beaches that are rural but are blue flag certified already into the `beach_urban` df.
bf_rural_beaches <- beach %>% 
  filter(ID_Beach %in% c("1233","1379","243","625"))

st_crs(beach_urban) == st_crs(bf_rural_beaches) # TRUE

beach_urban_complete <- bind_rows(beach_urban, bf_rural_beaches) # join the bf rural beaches
mapview(beach_urban_complete, color = "magenta")
```


```{r}
## join with the tibble indicating whether it is a bf certified beach in 2025
bf_w_beach_urban_complete <- full_join(beach_urban_complete, bf_beaches, relationship = "many-to-many") %>% 
  janitor::clean_names() %>% 
  mutate(ambito = ifelse(is.na(ambito), "Urbana", ambito),
         certified = ifelse(is.na(certified), "no", certified)) %>% 
  dplyr::select(id_beach, blue_flag_since, ambito, certified, blue_flag_nombre, geometry) %>% 
  st_as_sf()

mapview(bf_w_beach_urban_complete, color = "magenta")
```


Since there are multiple blue flags for some beaches, we can just call a beach certified if it has at least one BF certification:
```{r}
bf_w_beach_urban_complete_nodups <- bf_w_beach_urban_complete %>% 
  group_by(id_beach, ambito, geometry) %>%
  summarize(
    certified = first(certified),
    blue_flag_nombre = paste(blue_flag_nombre, collapse = "; "), # put all names of blue flag beaches associated with one INEGI beach
    blue_flag_since = suppressWarnings(min(blue_flag_since, na.rm = TRUE)), # take the earliest year that beach was certified if it has multiple blue flag beaches on one INEGI beach. Suppress warnings because any urban beach that does not have a blue flag would produce Inf but we will fix that... ->
    .groups = "drop") %>%
  mutate(blue_flag_since = ifelse(is.infinite(blue_flag_since), NA, blue_flag_since)) %>% # ...here!!
  st_as_sf() %>% 
  st_cast("MULTILINESTRING") %>% # to ensure they are the same types
  mutate(length_m = st_length(geometry)) # find each beach's length in meters

# st_write(bf_w_beach_urban_complete_nodups, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int/blue_flag_int","all_urban_beaches_plus_bf_rural_beaches_and_yr.gpkg"))
  
mapview(bf_w_beach_urban_complete_nodups, color = "magenta")

st_geometry_type(bf_w_beach_urban_complete_nodups) %>% unique()
# should return LINESTRING or MULTILINESTRING: returned MULTILINESTRING because I used st_cast

mapview(bf_w_beach_urban_complete_nodups, color = "magenta") + mapview(blueflag_sf, color = "blue") + mapview(cities_buf_2, col.regions ="lightblue") + mapview(beach, color = "yellow")
```

```{r}
## intersect with OHI rgn polygons for rgn_id
goc_ohi_rgns <- st_read(here::here("spatial/ohi_regions","goc_ohi_rgns.shp")) %>% 
  dplyr::select(rgn_id) %>%
  group_by(rgn_id) %>%
  summarise(geometry = st_union(geometry)) %>%
  ungroup()
mapview(goc_ohi_rgns)

bf_w_beach_urban_complete_rgnid <- st_intersection(st_transform(bf_w_beach_urban_complete_nodups, st_crs(goc_ohi_rgns)), goc_ohi_rgns)

mapview(bf_w_beach_urban_complete_rgnid, zcol = "rgn_id") + mapview(blueflag_sf, color = "blue") + mapview(cities_buf_2, col.regions ="lightblue")

mapview(bf_w_beach_urban_complete_rgnid, zcol = "certified") + mapview(blueflag_sf, color = "blue")

# st_write(bf_w_beach_urban_complete_rgnid, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int/blue_flag_int","blue_flag_w_urban_beaches_length_meters_rgn_id.gpkg"))
```

### Current status

current status = [the m length of touristy beaches with a blue flag] / [the total m of touristy beaches in the Gulf] * 100

First, we need to adjust for the "blue_flag_since" column, where if a beach has been blue flag certified since 2015, it should count as a certified beach for every year after that.

```{r}
### expand the blue flags into one row for each year it has been certified, earliest is 2013, to 2025
blue_flag_status_allyrs <- bf_w_beach_urban_complete_rgnid %>%
  st_drop_geometry() %>%
  mutate(length_m = as.numeric(length_m)) %>% # make numeric
  # expand certified beaches into one row per year
  mutate(year = ifelse(certified == "yes", blue_flag_since, NA)) %>%
  filter(certified == "yes" & !is.na(year)) %>%
  rowwise() %>%
  mutate(year = list(seq(year, 2025, by = 1))) %>%
  unnest(year) %>%
  ungroup()

# now get total certified length per rgn_id per year
certified_by_yr <- blue_flag_status_allyrs %>%
  group_by(rgn_id, year) %>%
  summarize(yes = sum(length_m, na.rm = TRUE), .groups = "drop")

# total beach length per region (constant across years)
total_urban_length_by_rgn <- bf_w_beach_urban_complete_rgnid %>%
  st_drop_geometry() %>%
  mutate(length_m = as.numeric(length_m)) %>%
  group_by(rgn_id) %>%
  summarize(total_length = sum(length_m, na.rm = TRUE), .groups = "drop")
```


#### Now we can calculate the current status for every region and year.

```{r}
# join and calculate status for every year
blue_flag_status_allyrs_final <- expand_grid(
    rgn_id = unique(total_urban_length_by_rgn$rgn_id),
    year = 2013:2025
  ) %>%
  left_join(total_urban_length_by_rgn, by = "rgn_id") %>%
  left_join(certified_by_yr, by = c("rgn_id", "year")) %>%
  mutate(
    yes = replace_na(yes, 0), # no certifications yet that year = 0
    prop_blue_flag = yes / total_length,
    current_status = prop_blue_flag * 100
  )

# write_csv(blue_flag_status_allyrs_final, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","blue_flag_current_status_final_by_yr_031726.csv"))
```

#### OLD current status without years since being a blue flag:

```{r}
blue_flag_current_status <- bf_w_beach_urban_complete_rgnid %>%
  st_drop_geometry() %>%
  mutate(length_m = as.numeric(length_m)) %>%   # drop units before pivot
  group_by(rgn_id, certified) %>%
  summarize(total_length = sum(length_m, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(
    names_from = certified,
    values_from = total_length,
    values_fill = 0
  ) %>%
  mutate(total_length = no + yes,
         prop_blue_flag = yes / total_length,
         current_status = prop_blue_flag * 100) %>% 
  mutate(year = 2025)

# write_csv(blue_flag_current_status, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","blue_flag_current_status_final_031626.csv"))

```

### plotly

```{r}
rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

## TR sustainability BlueFlag score over time
bf_plotly <- plotly::plot_ly(
  blue_flag_status_allyrs_final,
  x = ~year, 
  y = ~current_status, 
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>%
  layout(
    title = "CURRENT STATUS: length of urban beaches that have at least one Blue Flag site / total urban beach length",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Score (out of 100)"),
    legend = list(title = list(text = "Región"))
  ) %>%
  add_lines(
    x = range(blue_flag_status_allyrs_final$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (all beaches are blue flag = 100/100)",
    inherit = FALSE
  )
bf_plotly
```


## Questions to resolve:

(1) What should we do about the blue flag beaches Santa María and El Chileno?  INEGI does not register that part of the coastline as a beach, but it does look like one on the ESRI World Imagery. 

(2) Should we limit the beaches we consider "touristy" further by filtering out certain coastal cities that we would not consider highly visited, or needing to be blue flag beaches? EX: Loreto and other cities in Region_3?

(3) If they have zero BF beaches, should it absolutely have a current status of zero? 

## UPDATES

```{r}
## use POBTOT_km2 (population per km2) as a way to define whether a beach is touristy or not.

## use both AMBITO == urban and rural:
cities_w_rgnids_50km_RUR

# summary(cities_w_rgnids_50km_RUR$POBTOT_km2)
#    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
#     2.1   973.4  1696.4  2091.6  2642.9 44800.7 

## puerto peñasco is 1531.980210226832 POBTOT_km2, San Felipe is 1543.4270893317282, Sayulita is 2080.617258333723.

## This makes me feel good about having a threshold of greater than or equal to San Felipe POBTOT_km2 to be a touristy locality/city polygon.  Even small polygons like El Chileno/Santa Maria in Los Cabos has a POBTOT_km2 = 3281.6748449526376, so it really would limit out any that are not touristy (like puerto libertad in Sonora, 829.2587738359624)!!

## filter POBTOT_km2:
urban_cities_w_rgnids_50km <- cities_w_rgnids_50km_RUR %>% 
  filter(POBTOT_km2 >= 1500)
mapview(urban_cities_w_rgnids_50km)

pobtot_cities_buf <- st_buffer(urban_cities_w_rgnids_50km, dist = 400) # reminder: inland_50 is in CRS = Mexico ITRF2008 / LCC , which is a metric projection, so we can interpret the dist = 500 as meters.  

mapview(urban_cities_w_rgnids_50km) + mapview(pobtot_cities_buf, col.region = "lightblue") + mapview(pobtot_beach_urban, color = "magenta") + mapview(alojamientos_2019_2024_50km)


## UHOH!! ran into some places, like Bahia de Kino, that have many hotels in the locality polygon but < 1000 POBTOT_km2.  This makes me think we need to use employees per km2 instead.


alojamientos_est_perocu <- alojamientos_2019_2024_50km %>% 
  mutate(
    employees_estimated = case_when(
      per_ocu == "0 a 5 personas" ~ 3,
      per_ocu == "6 a 10 personas" ~ 8,
      per_ocu == "11 a 30 personas" ~ 20.5,
      per_ocu == "31 a 50 personas" ~ 40.5,
      per_ocu == "51 a 100 personas" ~ 75.5,
      per_ocu == "101 a 250 personas" ~ 175.5,
      per_ocu %in% c("251 y m<e1>s personas", "251 y más personas") ~ 300,
      TRUE ~ NA_real_
    )
  ) %>% 
  filter(data_year %in% 2024) # most recent hotel directory to understand where people visit

## join by cve_loc
hotel_emp_per_area_cveloc_join <- st_join(cities_w_rgnids_50km_RUR, alojamientos_est_perocu, by = c("CVE_LOC" = "cve_loc", "CVE_MUN" = "cve_mun","CVE_ENT" = "cve_ent"), relationship = "many-to-many") 

hotel_emp_per_area_cveloc <- hotel_emp_per_area_cveloc_join %>% 
  unite(col = "cve_mun_loc", c("cve_mun", "cve_loc"), sep = "_") %>% 
  group_by(cve_mun_loc, geometry) %>% 
  summarize(
    total_hotel_employees = sum(employees_estimated, na.rm = TRUE),
    total_area_km2 = sum(area_km2, na.rm = TRUE)
  ) %>% 
  ungroup() %>% 
  mutate(hotel_emp_per_km2 = total_hotel_employees / total_area_km2,
         total_hotel_employees = as.numeric(total_hotel_employees)) %>% 
  st_as_sf()

mapview(hotel_emp_per_area_cveloc, zcol = "hotel_emp_per_km2")
summary(hotel_emp_per_area_cveloc$hotel_emp_per_km2)
summary(hotel_emp_per_area_cveloc$total_hotel_employees)
# the min, qu, median are 0: more than 50% of areas have zero hotel employees, max shows large outliers.

hist(hotel_emp_per_area_cveloc$total_hotel_employees)

hotel_emp_per_area_cveloc$total_hotel_employees[hotel_emp_per_area_cveloc$total_hotel_employees == 0] <- NA
summary(hotel_emp_per_area_cveloc$total_hotel_employees) 

hotel_emp_per_area_cveloc_touristy <- hotel_emp_per_area_cveloc %>% 
  filter(total_hotel_employees >= 56) # use the lowest number of summed employees that have a blue flag certification (v2026: Lo de Marcos in Nayarit, which has 56 total hotel employees)

# hotel_emp_per_area_cveloc_touristy <- hotel_emp_per_area_cveloc %>% 
#   filter(total_hotel_employees >= 25.76) # use the mean!! Includes all places that have a blue flag as well. when we use 100+, then we lose La Ribera (BCS), Lo de Marcos (NAY), Punta Mita and La Cruz de Huanacaxtle (NAY).
# hotel_emp_per_area_cveloc_touristy <- hotel_emp_per_area_cveloc %>% 
#   filter(total_hotel_employees >= 50)
# hotel_emp_per_area_cveloc_touristy <- hotel_emp_per_area_cveloc %>% 
#   filter(total_hotel_employees >= 100)

mapview(hotel_emp_per_area_cveloc_touristy, zcol = "hotel_emp_per_km2") + mapview(alojamientos_est_perocu)
mapview(hotel_emp_per_area_cveloc_touristy, zcol = "total_hotel_employees") + mapview(hotel_emp_per_area_cveloc) + mapview(blueflag_sf) + mapview(cities)

hotel_emp_cities_buf <- st_buffer(hotel_emp_per_area_cveloc_touristy, dist = 400) # reminder: inland_50 is in CRS = Mexico ITRF2008 / LCC , which is a metric projection, so we can interpret the dist = 500 as meters.  

mapview(hotel_emp_per_area_cveloc_touristy, zcol = "total_hotel_employees") + mapview(hotel_emp_cities_buf)

# merge / dissolve all buffers into one geometry
hotel_buf_merged <- st_union(hotel_emp_cities_buf)
cities_buf_hotel <- st_as_sf(hotel_buf_merged) %>% 
  st_transform(crs=4326)

hot_emp_beach_urban <- st_intersection(st_transform(beach2, st_crs(cities_buf_hotel)), cities_buf_hotel)

# mapview(hotel_emp_per_area_cveloc_touristy, zcol = "total_hotel_employees") + mapview(hot_emp_beach_urban) + mapview(beach)

# mapview(hot_emp_beach_urban)+ mapview(hot_emp_beach_urban2)

hot_emp_beach_urban_df <- hot_emp_beach_urban %>% 
  st_drop_geometry()

# write_csv(hot_emp_beach_urban_df, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int/blue_flag_int","beaches_and_cliffs_bordering_localities_w_56_hotel_emp_or_more.csv"))

# mapview(beach2_urban_RUR, zcol = "ID_Beach") + mapview(cities_buf_2_RUR, col.regions ="lightblue") + mapview(cities_w_rgnids_50km_RUR)
```


#### Now, find where in the coastline is not "touristy" or take note of any that should be added back in (based on ID_Beach)

```{r}
mapview(beach2, zcol = "SUBTIPO") + mapview(hotel_emp_per_area_cveloc_touristy, col.region = "orange") + mapview(hotel_emp_per_area_cveloc, col.region = "red") + mapview(cities, col.region = "black") + mapview(alojamientos_est_perocu)
```


If there were no original polygon within `cities` and therefore the coastline was not caught as "touristy/urban" based on the buffer around the number of hotel employees in that polygon, then I will add them in manually.

NOTE that this does not apply to small locality polygons that were already filtered out by `total_hotel_employees >= 100`.  I will only be adding back in beaches that never had a locality in front of them, but have large hotels surrounding them.

<https://docs.google.com/spreadsheets/d/1tVsSwANpZAdqtuugBZJ50l_xPl6uVBwOE0geRvmWjXU/edit?gid=463454199#gid=463454199>

```{r}
large_hotels <- alojamientos_est_perocu %>% 
  filter(employees_estimated >= 56)

large_hotels_buffer <- st_buffer(large_hotels, dist = 400)
```

Use this mapview to walk through and identify any ID_Beach that has a large hotel beachfront to it, that is not already accounted for because it did not fall within a city/locality. 

If the hotel has 100 or more employees, then it fits the same logic as used for `hotel_emp_per_area_cveloc_touristy`. 

```{r}
mapview(large_hotels, zcol = "nom_estab") +mapview(large_hotels_buffer, col.region = "magenta") + mapview(alojamientos_est_perocu, col.region = "pink" ) + mapview(beach2, zcol = "SUBTIPO") + mapview(hotel_emp_per_area_cveloc_touristy, col.region = "orange") + mapview(cities, col.region = "black") 
```

Now that we have the list of those beaches, let's add them back in:

```{r}
beaches_w_large_hotels_not_in_locality <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int/blue_flag_int","beaches_w_large_hotels_not_in_locality.csv")) %>% 
  inner_join(beach2, by = "ID_Beach") %>% 
  st_as_sf()

total_touristy_beaches <- bind_rows(beaches_w_large_hotels_not_in_locality, hot_emp_beach_urban) %>% 
  mutate(reason = ifelse(is.na(reason), "Beach is within 400m of a locality polygon that within it has a sum of at least 100, or more, hotel employees.", reason))

mapview(total_touristy_beaches, color = "blue")
```


```{r}
mapview(large_hotels, col.region = "magenta") + mapview(alojamientos_est_perocu, col.region = "pink" ) + mapview(total_touristy_beaches, color = "blue") + mapview(beach2, zcol = "SUBTIPO") + mapview(hotel_emp_per_area_cveloc_touristy, col.region = "orange") + mapview(cities, col.region = "black") 
```


```{r}
## take original length and full beach of ID_Beach:

full_length_touristy_beaches <- st_join(coastline, st_transform(total_touristy_beaches, st_crs(coastline)), left = FALSE) # for an inner join

## to see what the differences would be if we used the full beach segments instead of the beach line cropped to the 400m buffer.
mapview(full_length_touristy_beaches, color = "magenta") + mapview(total_touristy_beaches, color = "green")
```

Clean the lines associated with ports, ensure it is in the GoC.
```{r}
# subtipo Construcción portuaria
ports <- coastline %>% 
  filter(SUBTIPO %in% c("Construcción portuaria") & TIPO =="Continental") %>%
  st_transform(crs=st_crs(goc_ohi_rgns))

## intersect with OHI rgn polygons for rgn_id
goc_ohi_rgns <- st_read(here::here("spatial/ohi_regions","goc_ohi_rgns.shp")) %>% 
  dplyr::select(rgn_id) %>%
  group_by(rgn_id) %>%
  summarise(geometry = st_union(geometry)) %>%
  ungroup() %>% 
  st_make_valid()
mapview(goc_ohi_rgns)

# bf_w_beach_urban_complete_rgnid <- st_intersection(st_transform(bf_w_beach_urban_complete_nodups, st_crs(goc_ohi_rgns)), goc_ohi_rgns)

ports_goc <- st_intersection(ports, goc_ohi_rgns)
ports_goc$ID_Port <- 1:nrow(ports_goc)
ports_goc_clean <- ports_goc %>% 
  filter(!ID_Port %in% c("1","2","3","4","6","31"))  %>% # pacific-side or outside of Gulf
  dplyr::select(c(ID_Port, rgn_id, geometry))
mapview(ports_goc_clean, zcol = "ID_Port")
```

Add in ports to each:
```{r}
total_touristy_beaches_w_ports <- bind_rows(st_transform(total_touristy_beaches, st_crs(ports_goc_clean)), ports_goc_clean) %>% 
  mutate(touristy = ifelse(is.na(touristy), "yes", touristy))

full_length_touristy_beaches_w_ports <- bind_rows(st_transform(full_length_touristy_beaches, st_crs(ports_goc_clean)), ports_goc_clean) %>% 
  mutate(touristy = ifelse(is.na(touristy), "yes", touristy))

## to see what the differences would be if we used the full beach segments instead of the beach line cropped to the 400m buffer.
mapview(full_length_touristy_beaches_w_ports, color = "magenta") + mapview(total_touristy_beaches_w_ports, color = "green")
```



### Assign ID_Beach to Blue Flag beaches

Now that we have a registry of the urban/touristy beaches, we need to match up the BlueFlag coordinates to the `ID_Beach` in the beach_urban sf. 

That way, we can use `st_length()` to compare the length of beaches protected under the Blue Flag certification vs. total length of touristy beaches.


```{r}
mapview(blueflag_sf, color = "blue") + mapview(beach2, zcol = "SUBTIPO")

blueflag_buffer <- st_buffer(blueflag_sf, dist = 400)
mapview(blueflag_sf, color = "blue") + mapview(blueflag_buffer) + mapview(beach2, zcol = "SUBTIPO")

# there are some that need to be done manually because there are a couple of beach segments within the buffer, and sometimes the coordinates of the BF beach is actually a little further away from the beach than it really is.  For example, "Hotel Marriott Puerto Vallarta Resort & Spa" upon further research applies to ID_Beach 1417 and not the other small segmented beaches nearby (ex: 2000) because those are in front of different hotels and Marriott's BF certification would not stretch there.

# because of this, I will walk through each one and assign the ID_Beach...
```

<https://docs.google.com/spreadsheets/d/1tVsSwANpZAdqtuugBZJ50l_xPl6uVBwOE0geRvmWjXU/edit?gid=761823854#gid=761823854>

```{r}
blueflag_ID_Beach <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int/blue_flag_int","blueflag_ID_Beach.csv")) %>% 
  mutate(certified = "yes") %>% 
  left_join(blueflag_sf, by = c("nombre")) %>% 
  # st_as_sf()
  st_drop_geometry() %>% 
  dplyr::select(-geometry) %>% 
  mutate(blue_flag_nombre = nombre)
```

# Add in Playas Limpias

```{r}
raw_playas_limpias <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/CONAGUA_playas_limpias/d2026","Playas_Limpias_Digitized_Data_Current_2022_2025.csv"))


## convert to decimal degrees
library(sf)
library(dplyr)
library(stringr)

# look at the raw bytes of one coordinate string
x <- raw_playas_limpias$latitud[1]
cat(x, "\n")          # how it prints
chartr("", "", x)     # won't help yet, but...
utf8::utf8_print(x)   # see exact characters

# check what quote/degree characters are actually in there
stringr::str_extract_all(x, "[^\\d\\s\\.]")  # extract all non-numeric, non-space chars

# Print first 5 lat/lon values to see what you're working with
head(raw_playas_limpias$latitud, 5)
head(raw_playas_limpias$longitud, 5)

# Check how many parsed successfully vs returned NA
playas_limpias %>%
  summarise(
    lat_na  = sum(is.na(lat_dd)),
    lon_na  = sum(is.na(lon_dd)),
    lat_ok  = sum(!is.na(lat_dd)),
    lon_ok  = sum(!is.na(lon_dd))
  )

dms_to_dd <- function(dms_string) {
  dms_string <- str_trim(dms_string)
  
  # Normalize: replace any degree-like and quote-like symbols with standard ones
  dms_clean <- dms_string %>%
    str_replace_all("[°º]", "°") %>%                  # normalize degree symbol
    str_replace_all("[\u2019\u2018\u0060\u00b4]", "'") %>%  # normalize single quotes
    str_replace_all("[\u201c\u201d]", "''")            # normalize double quotes

  # Flexible regex: tolerates spacing around symbols
  parts <- str_match(
    dms_clean,
    "(\\d+)\\s*°\\s*(\\d+)\\s*'\\s*([\\d.]+)\\s*[''\"]{1,2}\\s*(Norte|Sur|Este|Oeste)"
  )
  
  degrees   <- as.numeric(parts[, 2])
  minutes   <- as.numeric(parts[, 3])
  seconds   <- as.numeric(parts[, 4])
  direction <- parts[, 5]
  
  dd <- degrees + (minutes / 60) + (seconds / 3600)
  dd <- ifelse(direction %in% c("Sur", "Oeste"), -dd, dd)
  
  return(dd)
}

playas_limpias <- raw_playas_limpias %>%
  mutate(
    lat_dd  = dms_to_dd(latitud),
    lon_dd  = dms_to_dd(longitud)
  ) %>% 
  st_as_sf(
    coords = c("lon_dd", "lat_dd"), # making sf object
    crs = 4326, # WGS84
    remove = FALSE
  ) %>% 
  dplyr::select(fed_ent, destino, playa, sitio_de_muestro, lat_dd, lon_dd, geometry) %>% 
  distinct()

mapview(playas_limpias) + mapview(cropped_bf_w_beach_nodups_rgnid)
```

### Buffer playas limpias

```{r}
playas_limpias_metric_proj <- st_transform(playas_limpias, st_crs(inland_50))
playas_limpias_buf <- st_buffer(playas_limpias_metric_proj, dist = 500) # reminder: it is in CRS = Mexico ITRF2008 / LCC , which is a metric projection, so we can interpret the dist = 500 as meters.  

mapview(cropped_bf_w_beach_nodups_rgnid) + mapview(playas_limpias_buf) + mapview(beach2)

# merge / dissolve all buffers into one geometry
playas_limpias_buf_merged <- st_union(playas_limpias_buf)
playas_limpias_buf_wgs84 <- st_as_sf(playas_limpias_buf_merged) %>% st_transform(crs=4326)

mapview(beach2) + mapview(playas_limpias_buf_wgs84)
```

### Intersect with beach linestrings

```{r}
playas_limpias_linestring <- st_intersection(beach2, playas_limpias_buf_wgs84)

mapview(playas_limpias_linestring, color = "magenta") + mapview(playas_limpias_buf_wgs84, col.regions ="lightblue") + mapview(beach2)

playas_limpias_beachIDs <- unique(playas_limpias_linestring$ID_Beach)
```

Convert back to EPSG 6365

```{r}
playas_limpias_linestring_mex <- st_transform(playas_limpias_linestring, st_crs(inland_50))
```



```{r}
## filter beach2 for the beaches crossing the playas limpias:
beach2_playas_limpias <- beach2 %>% 
  filter(ID_Beach %in% playas_limpias_beachIDs)

##### these are far too large. would make the scores miniscule.
mapview(playas_limpias_linestring, color = "magenta") + mapview(beach2_playas_limpias, color = "green")
```


### Join with touristy beaches to make a total "highly-visited beaches" object

```{r}
total_highly_visited_beaches <- bind_rows(
  total_touristy_beaches_w_ports %>% mutate(source = "touristy"),
  playas_limpias_linestring_mex %>% mutate(source = "playas_limpias")
) %>%
  mutate(
    # make a unique key: use ID_Beach if available, otherwise flag as port
    group_key = case_when(
      !is.na(ID_Beach) ~ paste0("b-", ID_Beach),
      !is.na(ID_Port) ~ paste0("p-", ID_Port),
      TRUE ~ paste0("unknown-", row_number())
    )
  ) %>%
  group_by(group_key) %>%
  summarize(
    ID_Beach = first(ID_Beach),
    ID_Port = first(ID_Port),
    touristy = first(touristy),
    geometry = st_combine(geometry), # merge geometries from both sources
    .groups = "drop"
  ) %>%
  st_as_sf() %>%
  st_cast("MULTILINESTRING")

mapview(playas_limpias_linestring, color = "magenta") + mapview(total_touristy_beaches_w_ports, color = "green") + mapview(total_highly_visited_beaches, color = "purple")
```



# Join so that the urban beach dataset indicates whether it is blue flag certified or not


```{r}
################ cropped length #####################
cropped_bf_w_beach <- left_join(total_highly_visited_beaches, blueflag_ID_Beach, by = c("ID_Beach", "ID_Port"), relationship = "many-to-many") %>% 
  janitor::clean_names() %>% 
  mutate(ambito = ifelse(is.na(ambito), "Urbana", ambito),
         certified = ifelse(is.na(certified), "no", certified)) %>% 
  dplyr::select(id_beach, id_port, blue_flag_since, ambito, certified, blue_flag_nombre, geometry) %>%
  st_as_sf()

mapview(cropped_bf_w_beach, color = "green")
```


Since there are multiple blue flags for some beaches, we can just call a beach certified if it has at least one BF certification:
```{r}

################ cropped length #####################
cropped_bf_w_beach_nodups <- cropped_bf_w_beach %>% 
  mutate(id_beach_or_port = ifelse(is.na(id_beach), paste0("p-", id_port), as.character(id_beach))) %>%  # so we have one identifier for each beach and/or port.
  group_by(id_beach_or_port) %>%
  # group_by(id_beach, ambito, geometry) %>%
  summarize(
    certified = first(certified),
    geometry = first(geometry),
    blue_flag_nombre = paste(blue_flag_nombre, collapse = "; "), # put all names of blue flag beaches associated with one INEGI beach
    blue_flag_since = suppressWarnings(min(blue_flag_since, na.rm = TRUE)), # take the earliest year that beach was certified if it has multiple blue flag beaches on one INEGI beach. Suppress warnings because any urban beach that does not have a blue flag would produce Inf but we will fix that... ->
    .groups = "drop") %>%
  mutate(blue_flag_since = ifelse(is.infinite(blue_flag_since), NA, blue_flag_since)) %>% # ...here!!
  st_as_sf() %>% 
  st_cast("MULTILINESTRING") %>% # to ensure they are the same types
  mutate(length_m = st_length(geometry)) # find each beach's length in meters

# st_write(bf_w_beach_urban_complete_nodups, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int/blue_flag_int","all_urban_beaches_plus_bf_rural_beaches_and_yr.gpkg"))

mapview(cropped_bf_w_beach_nodups, color = "green")

st_geometry_type(cropped_bf_w_beach_nodups) %>% unique()
# should return LINESTRING or MULTILINESTRING: returned MULTILINESTRING because I used st_cast

mapview(cropped_bf_w_beach_nodups, color = "green") + mapview(blueflag_sf, color = "blue") + mapview(hotel_emp_per_area_cveloc_touristy, col.regions ="lightblue") + mapview(beach, color = "yellow")
```

```{r}
## intersect with OHI rgn polygons for rgn_id
goc_ohi_rgns <- st_read(here::here("spatial/ohi_regions","goc_ohi_rgns.shp")) %>% 
  dplyr::select(rgn_id) %>%
  group_by(rgn_id) %>%
  summarise(geometry = st_union(geometry)) %>%
  ungroup()
mapview(goc_ohi_rgns)


################ cropped length #####################

cropped_bf_w_beach_nodups_rgnid <- st_intersection(st_transform(cropped_bf_w_beach_nodups, st_crs(goc_ohi_rgns)), goc_ohi_rgns)

mapview(cropped_bf_w_beach_nodups_rgnid, zcol = "rgn_id") + mapview(blueflag_sf, color = "blue") + mapview(cities_buf_2, col.regions ="lightblue")

mapview(cropped_bf_w_beach_nodups_rgnid, zcol = "certified") + mapview(blueflag_sf, color = "blue") + mapview(large_hotels, col.region = "magenta") + mapview(hotel_emp_per_area_cveloc_touristy, col.region = "orange") + mapview(playas_limpias, col.region = "green")

# st_write(cropped_bf_w_beach_nodups_rgnid, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int/blue_flag_int","400m city_crop_bf_w_beach_length_meters_rgn_id.gpkg"))
```

# Current status

current status = [the m length of touristy beaches with a blue flag] / [the total m of touristy beaches in the Gulf] * 100

First, we need to adjust for the "blue_flag_since" column, where if a beach has been blue flag certified since 2015, it should count as a certified beach for every year after that.

```{r}
################ cropped length #####################

### expand the blue flags into one row for each year it has been certified, earliest is 2013, to 2025
blue_flag_status_allyrs <- cropped_bf_w_beach_nodups_rgnid %>%
  st_drop_geometry() %>%
  mutate(length_m = as.numeric(length_m)) %>% # make numeric
  # expand certified beaches into one row per year
  mutate(year = ifelse(certified == "yes", blue_flag_since, NA)) %>%
  filter(certified == "yes" & !is.na(year)) %>%
  rowwise() %>%
  mutate(year = list(seq(year, 2025, by = 1))) %>%
  unnest(year) %>%
  ungroup()

# now get total certified length per rgn_id per year
cropped_certified_by_yr <- blue_flag_status_allyrs %>%
  group_by(rgn_id, year) %>%
  summarize(yes = sum(length_m, na.rm = TRUE), .groups = "drop")

# total beach length per region (constant across years)
cropped_total_urban_length_by_rgn <- cropped_bf_w_beach_nodups_rgnid %>%
  st_drop_geometry() %>%
  mutate(length_m = as.numeric(length_m)) %>%
  group_by(rgn_id) %>%
  summarize(total_length = sum(length_m, na.rm = TRUE), .groups = "drop")
```



#### Now we can calculate the current status for every region and year.

```{r}
############### cropped length #####################

# join and calculate status for every year
cropped_blue_flag_status_allyrs_final <- expand_grid(
    rgn_id = unique(cropped_total_urban_length_by_rgn$rgn_id),
    year = 2013:2025
  ) %>%
  left_join(cropped_total_urban_length_by_rgn, by = "rgn_id") %>%
  left_join(cropped_certified_by_yr, by = c("rgn_id", "year")) %>%
  mutate(
    yes = replace_na(yes, 0), # no certifications yet that year = 0
    prop_blue_flag = yes / total_length,
    current_status = prop_blue_flag * 100
  )

# write_csv(cropped_blue_flag_status_allyrs_final, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","cropped_blue_flag_current_status_final_by_yr_031926.csv"))
# write_csv(cropped_blue_flag_status_allyrs_final, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","cropped_blue_flag_current_status_final_by_yr_033026.csv"))
```

##### plotly

```{r}
rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)


################ full length #####################

full_bf_plotly <- plotly::plot_ly(
  full_blue_flag_status_allyrs_final,
  x = ~year, 
  y = ~current_status, 
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>%
  layout(
    title = "CURRENT STATUS: FULL length of regularly-visited beaches that have at least one Blue Flag site / total regularly-visited beach length",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Score (out of 100)"),
    legend = list(title = list(text = "Región"))
  ) %>%
  add_lines(
    x = range(full_blue_flag_status_allyrs_final$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (all beaches are blue flag = 100/100)",
    inherit = FALSE
  )
full_bf_plotly





################ cropped length #####################

## TR sustainability BlueFlag score over time
crop_bf_plotly <- plotly::plot_ly(
  cropped_blue_flag_status_allyrs_final,
  x = ~year, 
  y = ~current_status, 
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>%
  layout(
    title = "CURRENT STATUS: length of highly-visited beaches that have at least one Blue Flag site / total highly-visited beach length", #  (buffered 400m to localities & beach coordinates)
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Score (out of 100)"),
    legend = list(title = list(text = "Región"))
  ) %>%
  add_lines(
    x = range(cropped_blue_flag_status_allyrs_final$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (all beaches are blue flag = 100/100)",
    inherit = FALSE
  )
crop_bf_plotly
```

