---
title: 'OHI GoC 2025: DENUE Establishments'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: inline
editor_options: 
  chunk_output_type: inline
---


# DENUE Establishments

This is the link to download the data: <https://www.inegi.org.mx/app/descarga/default.html>

Note that there is [an API available for INEGI DENUE](https://www.inegi.org.mx/servicios/api_denue.html). However, it would not be helpful for our purposes, because it will only call up the most recent values for the economic unit.  In order to evaluate changes over time, and not just current data on specific businesses or areas, we need to download each of the released .CSV lists of businesses per economic activity for every year.

This allows us to look at changes over time.

Additionally, each of these lists are published in May and November of each year.  I will be downloading the November lists for each year, and avoiding the May ones. 

I will be downloading the .CSVs for each economic activity from 2016 - 2025.

These are the ones that I will be downloading:

- Agricultura, cría y explotación de animales, aprovechamiento forestal, pesca y caza (`~/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/agricultura_cria_explotacion_forestal_pesca_y_caza`)
(`denue_00_11_1124_csv`)

- Minería 
(`~/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/mineria`)
(`denue_00_21_1124_csv`)

- Servicios de alojamiento temporal y de preparación de alimentos y bebidas (1 de 2)
- Servicios de alojamiento temporal y de preparación de alimentos y bebidas (2 de 2)
(`~/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas `)
(`denue_00_72_1_1124_csv`)
(`denue_00_72_2_1124_csv`)

- Servicios de apoyo a los negocios y manejo de desechos y servicios de remediación
(`~/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_apoyo_negocios_y_manejo_desechos_y_servicios_de_remediación`)
(`denue_00_56_1124_csv`)

- Servicios de esparcimiento culturales y deportivos, y otros servicios recreativos
(`~/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_esparcimiento_culturales_y_deportivos_servicios_recreativos`)
(`denue_00_71_1124_csv`)

- Servicios profesionales, científicos y técnicos
(`~/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_profesionales_científicos_técnicos`)
(`denue_00_54_1124_csv`)


Note that for the years 2025, 2017, and 2016 you will need to rename the files to have the appropriate month and date when they are unzipped.


NOTE: `fecha_alta` is "Fecha en la que la unidad económica se integró al Directorio Estadístico Nacional de Unidades Económicas y al Registro Estadístico de Negocios de México (RENEM)."

Aka, in English, it is the date on which the economic unit was added to the National Statistical Directory of Economic Units and the Statistical Registry of Businesses in Mexico (RENEM).

*Other things to note:*

1.  DENUE versions are cumulative but filtered by active status.
  Every DENUE release only includes active establishments at the time of publication.
  If an establishment was closed (temporarily or permanently) in 2022, it might be missing from the 2022 file, even if it has reopened or been reactivated by 2024.

2. Reactivated establishments keep their original `fecha_alta`.
  That is why there are fecha_altas as far back as 2010 in the 2024 file that were not present in the 2022 file: these businesses first entered RENEM in 2010, later disappeared from the active roster, and were likely revalidated by INEGI in 2024.

3. Differences between years do NOT equal purely new/closed businesses.
  Some IDs are newly added (new businesses).
  Some are reactivations.
  Some disappear because they closed, changed name or location, or failed verification.
  
SO, there can be establishments split/merged (1 --> 2 or 2 --> 1), occasional ID reassignments, etc.


## servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas

```{r}
library(tidyverse)
library(magrittr)
library(here)
library(janitor)
library(tmaptools)
library(sf)
library(mapview)

inland_10 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_10km_buffer","inland_10km_goc.shp"))
```


### Exploring 2024

```{r}
sp_1_2024 <- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas/denue_00_72_1_1124_csv/conjunto_de_datos","denue_inegi_72_1.csv"))

sp_1_2024_diccionario_de_datos <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas/denue_00_72_1_1124_csv/diccionario_de_datos","denue_diccionario_de_datos.csv"))

sp_1_2024_goc <- sp_1_2024 %>% 
  filter(entidad %in% c("Baja California","Baja California Sur","Jalisco","Nayarit","Sinaloa","Sonora"))

sf_sp_1_2024_goc <- sp_1_2024_goc %>% 
  st_as_sf(coords = c("longitud", "latitud"),
           crs = 4326, # WGS84
           remove = FALSE)

sf_sp_1_2024_goc_trans <- st_transform(sf_sp_1_2024_goc, st_crs(inland_10))
sf_sp_1_2024_goc_10km <- st_intersection(sf_sp_1_2024_goc_trans, inland_10)

sf_sp_1_2024_goc_10km[] <- lapply(sf_sp_1_2024_goc_10km, function(x) {
  if (is.character(x)) iconv(x, from = "", to = "UTF-8", sub = "byte") else x
})

mapview(sf_sp_1_2024_goc_10km)
```

```{r}
sp_2_2024 <- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas /denue_00_72_2_1124_csv/conjunto_de_datos","denue_inegi_72_2.csv"))

sp_2_2024_goc <- sp_2_2024 %>% 
  filter(entidad %in% c("Baja California","Baja California Sur","Jalisco","Nayarit","Sinaloa","Sonora"))

sf_sp_2_2024_goc <- sp_2_2024_goc %>% 
  st_as_sf(coords = c("longitud", "latitud"),
           crs = 4326, # WGS84
           remove = FALSE)

sf_sp_2_2024_goc_trans <- st_transform(sf_sp_2_2024_goc, st_crs(inland_10))
sf_sp_2_2024_goc_10km <- st_intersection(sf_sp_2_2024_goc_trans, inland_10)

sf_sp_2_2024_goc_10km[] <- lapply(sf_sp_2_2024_goc_10km, function(x) {
  if (is.character(x)) iconv(x, from = "", to = "UTF-8", sub = "byte") else x
})

mapview(sf_sp_2_2024_goc_10km)
```

Join them together:
```{r}
alojamientos_bebidas_2024_goc_10km <- rbind(sf_sp_1_2024_goc_10km, sf_sp_2_2024_goc_10km) %>% 
  mutate(data_year = 2024)
mapview(alojamientos_bebidas_2024_goc_10km)
```


What are the types of services here?
```{r}
unique(alojamientos_bebidas_2024_goc_10km$nombre_act)

#  [1] "Cafeter<ed>as, fuentes de sodas, never<ed>as, refresquer<ed>as y similares"                                   
#  [2] "Caba<f1>as, villas y similares"                                                                               
#  [3] "Restaurantes con servicio de preparaci<f3>n de alimentos a la carta o de comida corrida"                      
#  [4] "Servicios de preparaci<f3>n de otros alimentos para consumo inmediato"                                        
#  [5] "Restaurantes con servicio de preparaci<f3>n de antojitos"                                                     
#  [6] "Restaurantes con servicio de preparaci<f3>n de tacos y tortas"                                                
#  [7] "Restaurantes con servicio de preparaci<f3>n de pizzas, hamburguesas, hot dogs y pollos rostizados para llevar"
#  [8] "Hoteles con otros servicios integrados"                                                                       
#  [9] "Restaurantes con servicio de preparaci<f3>n de pescados y mariscos"                                           
# [10] "Bares, cantinas y similares"                                                                                  
# [11] "Restaurantes de autoservicio"                                                                                 
# [12] "Campamentos y albergues recreativos"                                                                          
# [13] "Hoteles sin otros servicios integrados"                                                                       
# [14] "Departamentos y casas amueblados con servicios de hoteler<ed>a"                                               
# [15] "Pensiones y casas de hu<e9>spedes"                                                                            
# [16] "Restaurantes que preparan otro tipo de alimentos para llevar"                                                 
# [17] "Moteles"                                                                                                      
# [18] "Centros nocturnos, discotecas y similares"                                                                    
# [19] "Servicios de preparaci<f3>n de alimentos en unidades m<f3>viles"                                              
# [20] "Servicios de preparaci<f3>n de alimentos para ocasiones especiales"                                           
# [21] "Servicios de comedor para empresas e instituciones"  
```

Separate out the accommodations:

```{r}
alojamientos_2024_goc_10km <- alojamientos_bebidas_2024_goc_10km %>% 
  filter(nombre_act %in% c("Moteles","Hoteles sin otros servicios integrados","Pensiones y casas de hu<e9>spedes","Departamentos y casas amueblados con servicios de hoteler<ed>a","Campamentos y albergues recreativos","Hoteles con otros servicios integrados","Caba<f1>as, villas y similares"))

mapview(alojamientos_2024_goc_10km)
```

And the food & drink services:

```{r}
comida_bebidas_2024_goc_10km <- alojamientos_bebidas_2024_goc_10km %>% 
  filter(nombre_act %in% c("Cafeter<ed>as, fuentes de sodas, never<ed>as, refresquer<ed>as y similares","Restaurantes con servicio de preparaci<f3>n de alimentos a la carta o de comida corrida","Servicios de preparaci<f3>n de otros alimentos para consumo inmediato","Restaurantes con servicio de preparaci<f3>n de antojitos","Restaurantes con servicio de preparaci<f3>n de tacos y tortas","Restaurantes con servicio de preparaci<f3>n de pizzas, hamburguesas, hot dogs y pollos rostizados para llevar","Restaurantes con servicio de preparaci<f3>n de pescados y mariscos","Bares, cantinas y similares","Restaurantes de autoservicio","Restaurantes que preparan otro tipo de alimentos para llevar","Centros nocturnos, discotecas y similares","Servicios de preparaci<f3>n de alimentos en unidades m<f3>viles","Servicios de preparaci<f3>n de alimentos para ocasiones especiales","Servicios de comedor para empresas e instituciones"))

mapview(comida_bebidas_2024_goc_10km)
```

### Function to read in and separate alojamientos from comida & bebidas 

```{r}
library(tidyverse)
library(magrittr)
library(here)
library(janitor)
library(tmaptools)
library(sf)
library(mapview)

# defining these categories based on 2024
alojamientos_categories <- c(
  "Moteles",
  "Hoteles sin otros servicios integrados",
  "Pensiones y casas de hu<e9>spedes",
  "Departamentos y casas amueblados con servicios de hoteler<ed>a",
  "Campamentos y albergues recreativos",
  "Hoteles con otros servicios integrados",
  "Caba<f1>as, villas y similares"
)

comida_bebidas_categories <- c(
  "Cafeter<ed>as, fuentes de sodas, never<ed>as, refresquer<ed>as y similares",
  "Restaurantes con servicio de preparaci<f3>n de alimentos a la carta o de comida corrida",
  "Servicios de preparaci<f3>n de otros alimentos para consumo inmediato",
  "Restaurantes con servicio de preparaci<f3>n de antojitos",
  "Restaurantes con servicio de preparaci<f3>n de tacos y tortas",
  "Restaurantes con servicio de preparaci<f3>n de pizzas, hamburguesas, hot dogs y pollos rostizados para llevar",
  "Restaurantes con servicio de preparaci<f3>n de pescados y mariscos",
  "Bares, cantinas y similares",
  "Restaurantes de autoservicio",
  "Restaurantes que preparan otro tipo de alimentos para llevar",
  "Centros nocturnos, discotecas y similares",
  "Servicios de preparaci<f3>n de alimentos en unidades m<f3>viles",
  "Servicios de preparaci<f3>n de alimentos para ocasiones especiales",
  "Servicios de comedor para empresas e instituciones"
)

# since the folders are not saved by year but instead by month and year, but are not consistent in the month for each year, lets extract the "suffix" that corresponds to each year.
get_folder_suffix <- function(year) {
  # years 2016 - 2019 would use their respective month/year codes
  # ex: 2016 is _1016_, 2017 is _1117_
  suffixes <- list(
    "2016" = "1016",
    "2017" = "1117", 
    "2018" = "1118",
    "2019" = "1119",
    "2020" = "1120",
    "2021" = "1121",
    "2022" = "1122",
    "2023" = "1123",
    "2024" = "1124",
    "2025" = "0525"
  )
  return(suffixes[[as.character(year)]])
}

###########################
# function to process data for a given year
process_denue_year <- function(year, inland_10, base_path = "/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025") {
  
  cat(paste0("Processing year: ", year, "\n"))
  
  # for each year, obtain the corresponding date and year "suffix"
  suffix <- get_folder_suffix(year)

  
  # the estados
  goc_states <- c("baja california", "baja california sur", "jalisco", 
                  "nayarit", "sinaloa", "sonora")
  
  # process dataset 1 (72_1)
  cat("  Reading 72_1 dataset...\n")
  sp_1_path <- here(base_path, 
                    "servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas",
                    paste0("denue_00_72_1_", suffix, "_csv"),
                    "conjunto_de_datos",
                    "denue_inegi_72_1.csv")
  
  sp_1 <- read_csv(sp_1_path)
  
  cat("  Mutating 72_1 dataset...\n")
  sp_1_goc <- sp_1 %>%
    mutate(entidad = stringr::str_to_lower(entidad)) %>% 
    select(-c(edificio_e, telefono)) %>% 
    filter(entidad %in% goc_states)
  
  cat("  Converting 72_1 to spatial object ...\n")
  sf_sp_1_goc <- sp_1_goc %>%
    st_as_sf(coords = c("longitud", "latitud"),
             crs = 4326,
             remove = FALSE)
  
  cat("  Transforming 72_1 dataset...\n")
  sf_sp_1_goc_trans <- st_transform(sf_sp_1_goc, st_crs(inland_10))
  
  cat("  Intersct 72_1 dataset...\n")
  sf_sp_1_goc_10km <- st_intersection(sf_sp_1_goc_trans, inland_10)
  
  cat("  Fix UTF issue for 72_1 dataset...\n")
  sf_sp_1_goc_10km[] <- lapply(sf_sp_1_goc_10km, function(x) {
    if (is.character(x)) iconv(x, from = "", to = "UTF-8", sub = "byte") else x
  })
  
  # process dataset 2 (72_2)
  cat("  Reading 72_2 dataset...\n")
  sp_2_path <- here(base_path,
                    "servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas",
                    paste0("denue_00_72_2_", suffix, "_csv"),
                    "conjunto_de_datos",
                    "denue_inegi_72_2.csv")
  
  sp_2 <- read_csv(sp_2_path, show_col_types = FALSE)
  
  cat("  Mutating 72_2 dataset...\n")
  sp_2_goc <- sp_2 %>%
    mutate(entidad = stringr::str_to_lower(entidad)) %>% 
    select(-c(edificio_e, telefono)) %>% 
    filter(entidad %in% goc_states)
  
  cat("  Converting 72_2 to spatial object ...\n")
  sf_sp_2_goc <- sp_2_goc %>%
    st_as_sf(coords = c("longitud", "latitud"),
             crs = 4326,
             remove = FALSE)
  
  cat("  Transforming 72_2 dataset...\n")
  sf_sp_2_goc_trans <- st_transform(sf_sp_2_goc, st_crs(inland_10))
  
  cat("  Intersct 72_2 dataset...\n")
  sf_sp_2_goc_10km <- st_intersection(sf_sp_2_goc_trans, inland_10)
  
  cat("  Fix UTF issue for 72_2 dataset...\n")
  sf_sp_2_goc_10km[] <- lapply(sf_sp_2_goc_10km, function(x) {
    if (is.character(x)) iconv(x, from = "", to = "UTF-8", sub = "byte") else x
  })
  
  # combine sp_1 and sp_2
  cat("  Combining datasets...\n")
  alojamientos_bebidas_goc_10km <- rbind(sf_sp_1_goc_10km, sf_sp_2_goc_10km) %>%
    mutate(data_year = year)
  
  # separate accommodations
  cat("  Filtering accommodations...\n")
  alojamientos_goc_10km <- alojamientos_bebidas_goc_10km %>%
    filter(nombre_act %in% alojamientos_categories)
  
  # separate out food & beverages
  cat("  Filtering comida y bebidas...\n")
  comida_bebidas_goc_10km <- alojamientos_bebidas_goc_10km %>%
    filter(nombre_act %in% comida_bebidas_categories)
  
  cat(paste0("  Completed year ", year, "!\n"))
  cat(paste0("    Accommodations: ", nrow(alojamientos_goc_10km), " records\n"))
  cat(paste0("    Food & Beverages: ", nrow(comida_bebidas_goc_10km), " records\n\n"))
  
  # returning them as a list
  return(list(
    alojamientos = alojamientos_goc_10km,
    comida_bebidas = comida_bebidas_goc_10km,
    combined = alojamientos_bebidas_goc_10km
  ))
}

# GoC 10km buffer
inland_10 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_10km_buffer","inland_10km_goc.shp"))

# check with 2024 that I did individually
result_2024 <- process_denue_year(2024, inland_10)
alojamientos_2024_goc_10km <- result_2024$alojamientos
comida_bebidas_2024_goc_10km <- result_2024$comida_bebidas
```

### Using function

```{r}
##########################################
# process years 2019 to 2024!
years <- 2019:2024
results <- lapply(years, function(y) process_denue_year(y, inland_10))
names(results) <- paste0("year_", years)

# joining all of these into one object for alojamientos and comida & bebidas.

all_alojamientos <- bind_rows(lapply(results, function(x) x$alojamientos))
# 1701 + 1107 + 1630 + 1630 + 1639 + 1551 = 9258
# great!!
mapview(all_alojamientos, zcol = "data_year")


all_comida_bebidas <- bind_rows(lapply(results, function(x) x$comida_bebidas))
mapview(all_comida_bebidas, zcol = "data_year")
##########################################
```

#### Checking fxn for various years

Was hitting lots of warnings prior to 2021, so I decided to take a look, and it seems like they notated the states in all caps instead of camel case.  Will need to alter the function to account for that. 

```{r}
sp_1_2020 <- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas/denue_00_72_1_1120_csv/conjunto_de_datos","denue_inegi_72_1.csv"))

sp_1_2020_diccionario_de_datos <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas/denue_00_72_1_1120_csv/diccionario_de_datos","denue_diccionario_de_datos.csv"))

sp_1_2020_goc <- sp_1_2020 %>% 
  mutate(entidad = stringr::str_to_lower(entidad)) %>% 
  filter(entidad %in% c("Baja California","Baja California Sur","Jalisco","Nayarit","Sinaloa","Sonora")) #### here is the issue

unique(sp_1_2020_goc$entidad)

#  [1] "aguascalientes"         "baja california"        "baja california sur"    "campeche"               "coahuila de zaragoza"   "colima"                
#  [7] "chiapas"                "chihuahua"              "ciudad de m\xc9xico"    "durango"                "guanajuato"             "guerrero"              
# [13] "hidalgo"                "jalisco"                "m\xc9xico"              "michoac\xc1n de ocampo" "morelos"                "nayarit"               
# [19] "nuevo le\xd3n"          "oaxaca"                 "puebla"   

na_2020 <-sp_1_2020_goc %>% 
  filter(is.na(longitud))

sf_sp_1_2020_goc <- sp_1_2020_goc %>% 
  st_as_sf(coords = c("longitud", "latitud"),
           crs = 4326, # WGS84
           remove = FALSE)
```

Exploring 2019: making sure this looks similar to later years.

```{r}
sp_1_2019 <- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas/denue_00_72_1_1119_csv/conjunto_de_datos","denue_inegi_72_1.csv"))

# sp_1_2019_goc <- sp_1_2019 %>% 
  # filter(entidad %in% c("Baja California","Baja California Sur","Jalisco","Nayarit","Sinaloa","Sonora")) #### here is the issue also.

na_2019 <-sp_1_2019 %>% 
  filter(is.na(longitud)) # none even without filtering for GoC
```


2023 looks suspiciously low for accommodations in terms of records (comparatively)

```{r}
sp_1_2023 <- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas/denue_00_72_1_1123_csv/conjunto_de_datos","denue_inegi_72_1.csv"))

sp_1_2023_goc <- sp_1_2023 %>% 
  mutate(entidad = stringr::str_to_lower(entidad)) %>% 
  filter(entidad %in% c("baja california","baja california sur","jalisco","nayarit","sinaloa","sonora")) 

na_2023 <-sp_1_2023_goc %>% 
  filter(is.na(longitud)) ## none

sf_sp_1_2023_goc <- sp_1_2023_goc %>% 
  st_as_sf(coords = c("longitud", "latitud"),
           crs = 4326, # WGS84
           remove = FALSE)

sf_sp_1_2023_goc_trans <- st_transform(sf_sp_1_2023_goc, st_crs(inland_10))
sf_sp_1_2023_goc_10km <- st_intersection(sf_sp_1_2023_goc_trans, inland_10)

sf_sp_1_2023_goc_10km[] <- lapply(sf_sp_1_2023_goc_10km, function(x) {
  if (is.character(x)) iconv(x, from = "", to = "UTF-8", sub = "byte") else x
})

mapview(sf_sp_1_2023_goc_10km)

alojamientos_2023_sp1_goc_10km <- sf_sp_1_2023_goc_10km %>% 
  filter(nombre_act %in% c("Moteles","Hoteles sin otros servicios integrados","Pensiones y casas de hu<e9>spedes","Departamentos y casas amueblados con servicios de hoteler<ed>a","Campamentos y albergues recreativos","Hoteles con otros servicios integrados","Caba<f1>as, villas y similares")) # 762 obs.

###############################
sp_2_2023 <- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas/denue_00_72_2_1123_csv/conjunto_de_datos","denue_inegi_72_2.csv"))

sp_2_2023_goc <- sp_2_2023 %>% 
  mutate(entidad = stringr::str_to_lower(entidad)) %>% 
  filter(entidad %in% c("baja california","baja california sur","jalisco","nayarit","sinaloa","sonora")) 

na_2023 <-sp_2_2023_goc %>% 
  filter(is.na(longitud)) ## none

sf_sp_2_2023_goc <- sp_2_2023_goc %>% 
  st_as_sf(coords = c("longitud", "latitud"),
           crs = 4326, # WGS84
           remove = FALSE)

sf_sp_2_2023_goc_trans <- st_transform(sf_sp_2_2023_goc, st_crs(inland_10))
sf_sp_2_2023_goc_10km <- st_intersection(sf_sp_2_2023_goc_trans, inland_10)

sf_sp_2_2023_goc_10km[] <- lapply(sf_sp_2_2023_goc_10km, function(x) {
  if (is.character(x)) iconv(x, from = "", to = "UTF-8", sub = "byte") else x
})

mapview(sf_sp_2_2023_goc_10km)

alojamientos_2023_sp2_goc_10km <- sf_sp_2_2023_goc_10km %>% 
  filter(nombre_act %in% c("Moteles","Hoteles sin otros servicios integrados","Pensiones y casas de hu<e9>spedes","Departamentos y casas amueblados con servicios de hoteler<ed>a","Campamentos y albergues recreativos","Hoteles con otros servicios integrados","Caba<f1>as, villas y similares")) # 345 obs.

##### 762 + 345 = 1107, which is the same as what the function produced!! Yay, so it works!!!
```

### Write out accommodations and comida/bebidas

```{r}
### alojamientos
# st_write(all_alojamientos, here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas/int","alojamientos_2019_2024_goc_10km.shp"))

# ### comida bebidas
# st_write(all_comida_bebidas, here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas/int","comida_bebidas_2019_2024_goc_10km.shp"))
```

## agricultura_cria_explotacion_forestal_pesca_y_caza


## mineria


## servicios_apoyo_negocios_y_manejo_desechos_y_servicios_de_remediación


## servicios_esparcimiento_culturales_y_deportivos_servicios_recreativos


## servicios_profesionales_científicos_técnicos

