---
title: 'OHI GoC 2025: DENUE Establishments'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: inline
editor_options: 
  chunk_output_type: inline
---


# DENUE Establishments

This is the link to download the data: <https://www.inegi.org.mx/app/descarga/default.html>

Note that there is [an API available for INEGI DENUE](https://www.inegi.org.mx/servicios/api_denue.html). However, it would not be helpful for our purposes, because it will only call up the most recent values for the economic unit.  In order to evaluate changes over time, and not just current data on specific businesses or areas, we need to download each of the released .CSV lists of businesses per economic activity for every year.

This allows us to look at changes over time.

Additionally, each of these lists are published in May and November of each year.  I will be downloading the November lists for each year, and avoiding the May ones. 

I will be downloading the .CSVs for each economic activity from 2016 - 2025.

These are the ones that I will be downloading:

- Agricultura, cría y explotación de animales, aprovechamiento forestal, pesca y caza (`~/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/agricultura_cria_explotacion_forestal_pesca_y_caza`)
(`denue_00_11_1124_csv`)

- Minería 
(`~/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/mineria`)
(`denue_00_21_1124_csv`)

- Servicios de alojamiento temporal y de preparación de alimentos y bebidas (1 de 2)
- Servicios de alojamiento temporal y de preparación de alimentos y bebidas (2 de 2)
(`~/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas `)
(`denue_00_72_1_1124_csv`)
(`denue_00_72_2_1124_csv`)

- Servicios de apoyo a los negocios y manejo de desechos y servicios de remediación
(`~/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_apoyo_negocios_y_manejo_desechos_y_servicios_de_remediación`)
(`denue_00_56_1124_csv`)

- Servicios de esparcimiento culturales y deportivos, y otros servicios recreativos
(`~/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_esparcimiento_culturales_y_deportivos_servicios_recreativos`)
(`denue_00_71_1124_csv`)

- Servicios profesionales, científicos y técnicos
(`~/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_profesionales_científicos_técnicos`)
(`denue_00_54_1124_csv`)


Note that for the years 2025, 2017, and 2016 you will need to rename the files to have the appropriate month and date when they are unzipped.


NOTE: `fecha_alta` is "Fecha en la que la unidad económica se integró al Directorio Estadístico Nacional de Unidades Económicas y al Registro Estadístico de Negocios de México (RENEM)."

Aka, in English, it is the date on which the economic unit was added to the National Statistical Directory of Economic Units and the Statistical Registry of Businesses in Mexico (RENEM).

*Other things to note:*

1.  DENUE versions are cumulative but filtered by active status.
  Every DENUE release only includes active establishments at the time of publication.
  If an establishment was closed (temporarily or permanently) in 2022, it might be missing from the 2022 file, even if it has reopened or been reactivated by 2024.

2. Reactivated establishments keep their original `fecha_alta`.
  That is why there are fecha_altas as far back as 2010 in the 2024 file that were not present in the 2022 file: these businesses first entered RENEM in 2010, later disappeared from the active roster, and were likely revalidated by INEGI in 2024.

3. Differences between years do NOT equal purely new/closed businesses.
  Some IDs are newly added (new businesses).
  Some are reactivations.
  Some disappear because they closed, changed name or location, or failed verification.
  
SO, there can be establishments split/merged (1 --> 2 or 2 --> 1), occasional ID reassignments, etc.


## servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas

```{r}
library(tidyverse)
library(magrittr)
library(here)
library(janitor)
library(tmaptools)
library(sf)
library(mapview)

inland_10 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_10km_buffer","inland_10km_goc.shp"))
inland_50 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer","inland_50km_goc.shp"))
```

### Function to read in and separate alojamientos from comida & bebidas 

```{r}
library(tidyverse)
library(magrittr)
library(here)
library(janitor)
library(tmaptools)
library(sf)
library(mapview)

# defining these categories based on 2024
alojamientos_categories <- c(
  "Moteles",
  "Hoteles sin otros servicios integrados",
  "Pensiones y casas de huéspedes",
  "Departamentos y casas amueblados con servicios de hotelería",
  "Campamentos y albergues recreativos",
  "Hoteles con otros servicios integrados",
  "Cabañas, villas y similares"
)

comida_bebidas_categories <- c(
  "Cafeterías, fuentes de sodas, neverías, refresquerías y similares",
  "Restaurantes con servicio de preparación de alimentos a la carta o de comida corrida",
  "Servicios de preparación de otros alimentos para consumo inmediato",
  "Restaurantes con servicio de preparación de antojitos",
  "Restaurantes con servicio de preparación de tacos y tortas",
  "Restaurantes con servicio de preparación de pizzas, hamburguesas, hot dogs y pollos rostizados para llevar",
  "Restaurantes con servicio de preparación de pescados y mariscos",
  "Bares, cantinas y similares",
  "Restaurantes de autoservicio",
  "Restaurantes que preparan otro tipo de alimentos para llevar",
  "Centros nocturnos, discotecas y similares",
  "Servicios de preparación de alimentos en unidades móviles",
  "Servicios de preparación de alimentos para ocasiones especiales",
  "Servicios de comedor para empresas e instituciones"
)

# since the folders are not saved by year but instead by month and year, but are not consistent in the month for each year, lets extract the "suffix" that corresponds to each year.
get_folder_suffix <- function(year) {
  # years 2016 - 2019 would use their respective month/year codes
  # ex: 2016 is _1016_, 2017 is _1117_
  suffixes <- list(
    "2016" = "1016",
    "2017" = "1117", 
    "2018" = "1118",
    "2019" = "1119",
    "2020" = "1120",
    "2021" = "1121",
    "2022" = "1122",
    "2023" = "1123",
    "2024" = "1124",
    "2025" = "0525"
  )
  return(suffixes[[as.character(year)]])
}

###########################
# function to process data for a given year
process_denue_year <- function(year, inland_50, base_path = "/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025") {
  
  cat(paste0("Processing year: ", year, "\n"))
  
  # for each year, obtain the corresponding date and year "suffix"
  suffix <- get_folder_suffix(year)
  
  
  # the estados
  goc_states <- c("baja california", "baja california sur", "jalisco", 
                  "nayarit", "sinaloa", "sonora")
  
  # process dataset 1 (72_1)
  cat("  Reading 72_1 dataset...\n")
  sp_1_path <- here(base_path, 
                    "servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas",
                    paste0("denue_00_72_1_", suffix, "_csv"),
                    "conjunto_de_datos",
                    "denue_inegi_72_1.csv")
  
  # sp_1 <- read_csv(sp_1_path, locale = locale(encoding = "latin1"), show_col_types = FALSE)
  if (year != 2023) {
    sp_1 <- read_csv(sp_1_path, locale = locale(encoding = "latin1"), show_col_types = FALSE)
  } else {
    sp_1 <- read_csv(sp_1_path, show_col_types = FALSE)
  }
  
  cat("  Mutating 72_1 dataset...\n")
  sp_1_goc <- sp_1 %>%
    mutate(entidad = stringr::str_to_lower(entidad)) %>% 
    select(-c(edificio_e, telefono)) %>% 
    filter(entidad %in% goc_states)
  
  cat("  Converting 72_1 to spatial object ...\n")
  sf_sp_1_goc <- sp_1_goc %>%
    st_as_sf(coords = c("longitud", "latitud"),
             crs = 4326,
             remove = FALSE)
  
  cat("  Transforming 72_1 dataset...\n")
  sf_sp_1_goc_trans <- st_transform(sf_sp_1_goc, st_crs(inland_50))
  
  cat("  Intersct 72_1 dataset...\n")
  sf_sp_1_goc_50km <- st_intersection(sf_sp_1_goc_trans, inland_50)
  
  # cat("  Fix UTF issue for 72_1 dataset...\n")
  # sf_sp_1_goc_50km[] <- lapply(sf_sp_1_goc_50km, function(x) {
  #   if (is.character(x)) iconv(x, from = "", to = "UTF-8", sub = "byte") else x
  # })
  
  # process dataset 2 (72_2)
  cat("  Reading 72_2 dataset...\n")
  sp_2_path <- here(base_path,
                    "servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas",
                    paste0("denue_00_72_2_", suffix, "_csv"),
                    "conjunto_de_datos",
                    "denue_inegi_72_2.csv")
  
  # sp_2 <- read_csv(sp_1_path, locale = locale(encoding = "latin1"), show_col_types = FALSE)
  if (year != 2023) {
    sp_2 <- read_csv(sp_2_path, locale = locale(encoding = "latin1"), show_col_types = FALSE)
  } else {
    sp_2 <- read_csv(sp_2_path, show_col_types = FALSE)
  }
  
  cat("  Mutating 72_2 dataset...\n")
  sp_2_goc <- sp_2 %>%
    mutate(entidad = stringr::str_to_lower(entidad)) %>% 
    select(-c(edificio_e, telefono)) %>% 
    filter(entidad %in% goc_states)
  
  cat("  Converting 72_2 to spatial object ...\n")
  sf_sp_2_goc <- sp_2_goc %>%
    st_as_sf(coords = c("longitud", "latitud"),
             crs = 4326,
             remove = FALSE)
  
  cat("  Transforming 72_2 dataset...\n")
  sf_sp_2_goc_trans <- st_transform(sf_sp_2_goc, st_crs(inland_50))
  
  cat("  Intersct 72_2 dataset...\n")
  sf_sp_2_goc_50km <- st_intersection(sf_sp_2_goc_trans, inland_50)
  
  # cat("  Fix UTF issue for 72_2 dataset...\n")
  # sf_sp_2_goc_50km[] <- lapply(sf_sp_2_goc_50km, function(x) {
  #   if (is.character(x)) iconv(x, from = "", to = "UTF-8", sub = "byte") else x
  # })
  
  # combine sp_1 and sp_2
  cat("  Combining datasets...\n")
  alojamientos_bebidas_goc_50km <- rbind(sf_sp_1_goc_50km, sf_sp_2_goc_50km) %>%
    mutate(data_year = year)
  
  # separate accommodations
  cat("  Filtering accommodations...\n")
  alojamientos_goc_50km <- alojamientos_bebidas_goc_50km %>%
    filter(nombre_act %in% alojamientos_categories)
  
  # separate out food & beverages
  cat("  Filtering comida y bebidas...\n")
  comida_bebidas_goc_50km <- alojamientos_bebidas_goc_50km %>%
    filter(nombre_act %in% comida_bebidas_categories)
  
  cat(paste0("  Completed year ", year, "!\n"))
  cat(paste0("    Accommodations: ", nrow(alojamientos_goc_50km), " records\n"))
  cat(paste0("    Food & Beverages: ", nrow(comida_bebidas_goc_50km), " records\n\n"))
  
  # returning them as a list
  return(list(
    alojamientos = alojamientos_goc_50km,
    comida_bebidas = comida_bebidas_goc_50km,
    combined = alojamientos_bebidas_goc_50km
  ))
}
```


```{r}
# GoC 10km buffer
inland_10 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_10km_buffer","inland_10km_goc.shp"))

# GoC 50km buffer
inland_50 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer","inland_50km_goc.shp"))

# check with 2023, 2024 that I did individually
result_2023 <- process_denue_year(2023, inland_50)
result_2024 <- process_denue_year(2024, inland_50)
alojamientos_2024_goc_10km <- result_2024$alojamientos
comida_bebidas_2024_goc_10km <- result_2024$comida_bebidas
```

### Using function

```{r}
# using 10km buffer
# process years 2019 to 2024!
# years <- 2019:2024
# results_10km <- lapply(years, function(y) process_denue_year(y, inland_10))
# names(results_10km) <- paste0("year_", years)
# 
# # joining all of these into one object for alojamientos and comida & bebidas.
# 
# all_alojamientos_10km <- bind_rows(lapply(results_10km, function(x) x$alojamientos))
# # 1701 + 1107 + 1630 + 1630 + 1639 + 1551 = 9258
# # great!!
# mapview(all_alojamientos, zcol = "data_year")
# 
# 
# all_comida_bebidas_10km <- bind_rows(lapply(results_10km, function(x) x$comida_bebidas))
# mapview(all_comida_bebidas, zcol = "data_year")


##########################################
# using 50km buffer
# process years 2019 to 2024!
inland_50 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer","inland_50km_goc.shp"))
years <- 2019:2024
results_50km <- lapply(years, function(y) process_denue_year(y, inland_50))
names(results_50km) <- paste0("year_", years)

# joining all of these into one object for alojamientos and comida & bebidas.

all_alojamientos_50km <- bind_rows(lapply(results_50km, function(x) x$alojamientos))
mapview(all_alojamientos_50km, zcol = "data_year")

all_comida_bebidas_50km <- bind_rows(lapply(results_50km, function(x) x$comida_bebidas))
mapview(all_comida_bebidas_50km, zcol = "data_year")
##########################################
```

### Write out accommodations and comida/bebidas

```{r}
############## 10km ###############
# ### alojamientos
# st_write(all_alojamientos, here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas/int","alojamientos_2019_2024_goc_10km.shp"))
# 
# # ### comida bebidas
# st_write(all_comida_bebidas, here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas/int","comida_bebidas_2019_2024_goc_10km.shp"))

############## 50km ###############
### alojamientos
st_write(all_alojamientos_50km, here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas/int","update_alojamientos_2019_2024_goc_50km.shp"))

# ### comida bebidas
st_write(all_comida_bebidas_50km, here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas/int","update_comida_bebidas_2019_2024_goc_50km.shp"))
```

#### Checking fxn for various years

Was hitting lots of warnings prior to 2021, so I decided to take a look, and it seems like they notated the states in all caps instead of camel case.  Will need to alter the function to account for that. 


##### 2020

```{r}
###############################################
### 1. SP1 – YEAR 2020
###############################################

sp_1_2020 <- read_csv(here::here(
  "/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas/denue_00_72_1_1120_csv/conjunto_de_datos",
  "denue_inegi_72_1.csv"
))

sp_1_2020_goc <- sp_1_2020 %>% 
  mutate(entidad = stringr::str_to_lower(entidad)) %>% 
  filter(entidad %in% c("baja california","baja california sur","jalisco","nayarit","sinaloa","sonora"))

# Any missing longitud?
na_2020_sp1 <- sp_1_2020_goc %>% filter(is.na(longitud))

sf_sp_1_2020_goc <- sp_1_2020_goc %>% 
  st_as_sf(coords = c("longitud", "latitud"),
           crs = 4326,
           remove = FALSE)

sf_sp_1_2020_goc_trans <- st_transform(sf_sp_1_2020_goc, st_crs(inland_50))
sf_sp_1_2020_goc_50km <- st_intersection(sf_sp_1_2020_goc_trans, inland_50)

# Fix encoding
sf_sp_1_2020_goc_50km[] <- lapply(sf_sp_1_2020_goc_50km, function(x) {
  if (is.character(x)) iconv(x, from = "", to = "UTF-8", sub = "byte") else x
})

### Subsets: alojamientos + comida/bebidas (SP1 2020)
alojamientos_2020_sp1_goc_50km <- sf_sp_1_2020_goc_50km %>% 
  filter(nombre_act %in% c(
    "Moteles",
    "Hoteles sin otros servicios integrados",
    "Pensiones y casas de hu\u00e9spedes",
    "Departamentos y casas amueblados con servicios de hoteler\u00eda",
    "Campamentos y albergues recreativos",
    "Hoteles con otros servicios integrados",
    "Caba\u00f1as, villas y similares"
  ))

comida_bebidas_2020_sp1_goc_50km <- sf_sp_1_2020_goc_50km %>% 
  filter(nombre_act %in% c(
    "Cafeter\u00edas, fuentes de sodas, never\u00edas, refresquer\u00edas y similares",
    "Restaurantes con servicio de preparaci\u00f3n de alimentos a la carta o de comida corrida",
    "Servicios de preparaci\u00f3n de otros alimentos para consumo inmediato",
    "Restaurantes con servicio de preparaci\u00f3n de antojitos",
    "Restaurantes con servicio de preparaci\u00f3n de tacos y tortas",
    "Restaurantes con servicio de preparaci\u00f3n de pizzas, hamburguesas, hot dogs y pollos rostizados para llevar",
    "Restaurantes con servicio de preparaci\u00f3n de pescados y mariscos",
    "Bares, cantinas y similares",
    "Restaurantes de autoservicio",
    "Restaurantes que preparan otro tipo de alimentos para llevar",
    "Centros nocturnos, discotecas y similares",
    "Servicios de preparaci\u00f3n de alimentos en unidades m\u00f3viles",
    "Servicios de preparaci\u00f3n de alimentos para ocasiones especiales",
    "Servicios de comedor para empresas e instituciones"
  ))


###############################################
### 2. SP2 – YEAR 2020
###############################################

sp_2_2020 <- read_csv(here::here(
  "/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas/denue_00_72_2_1120_csv/conjunto_de_datos",
  "denue_inegi_72_2.csv"
), locale = locale(encoding = "latin1"))

sp_2_2020_goc <- sp_2_2020 %>% 
  mutate(entidad = stringr::str_to_lower(entidad)) %>% 
  filter(entidad %in% c("baja california","baja california sur","jalisco","nayarit","sinaloa","sonora"))

na_2020_sp2 <- sp_2_2020_goc %>% filter(is.na(longitud))

sf_sp_2_2020_goc <- sp_2_2020_goc %>% 
  st_as_sf(coords = c("longitud", "latitud"),
           crs = 4326,
           remove = FALSE)

sf_sp_2_2020_goc_trans <- st_transform(sf_sp_2_2020_goc, st_crs(inland_50))
sf_sp_2_2020_goc_50km <- st_intersection(sf_sp_2_2020_goc_trans, inland_50)

# # Fix encoding
# sf_sp_2_2020_goc_50km[] <- lapply(sf_sp_2_2020_goc_50km, function(x) {
#   if (is.character(x)) iconv(x, from = "latin1", to = "UTF-8", sub = "byte") else x
# })

unique(sf_sp_2_2020_goc_50km$nombre_act)

### Subsets: alojamientos + comida/bebidas (SP2 2020)

alojamientos_2020_sp2_goc_50km <- sf_sp_2_2020_goc_50km %>% 
  filter(nombre_act %in% c(
    "Moteles",
    "Hoteles sin otros servicios integrados",
    "Pensiones y casas de hu\u00e9spedes",
    "Departamentos y casas amueblados con servicios de hoteler\u00eda",
    "Campamentos y albergues recreativos",
    "Hoteles con otros servicios integrados",
    "Caba\u00f1as, villas y similares"
  ))

comida_bebidas_2020_sp2_goc_50km <- sf_sp_2_2020_goc_50km %>% 
  filter(nombre_act %in% c(
    "Cafeter\u00edas, fuentes de sodas, never\u00edas, refresquer\u00edas y similares",
    "Restaurantes con servicio de preparaci\u00f3n de alimentos a la carta o de comida corrida",
    "Servicios de preparaci\u00f3n de otros alimentos para consumo inmediato",
    "Restaurantes con servicio de preparaci\u00f3n de antojitos",
    "Restaurantes con servicio de preparaci\u00f3n de tacos y tortas",
    "Restaurantes con servicio de preparaci\u00f3n de pizzas, hamburguesas, hot dogs y pollos rostizados para llevar",
    "Restaurantes con servicio de preparaci\u00f3n de pescados y mariscos",
    "Bares, cantinas y similares",
    "Restaurantes de autoservicio",
    "Restaurantes que preparan otro tipo de alimentos para llevar",
    "Centros nocturnos, discotecas y similares",
    "Servicios de preparaci\u00f3n de alimentos en unidades m\u00f3viles",
    "Servicios de preparaci\u00f3n de alimentos para ocasiones especiales",
    "Servicios de comedor para empresas e instituciones"
  ))

mapview(comida_bebidas_2020_sp2_goc_50km)

```


```{r}
sp_1_2020 <- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas/denue_00_72_1_1120_csv/conjunto_de_datos","denue_inegi_72_1.csv"))

sp_1_2020_diccionario_de_datos <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas/denue_00_72_1_1120_csv/diccionario_de_datos","denue_diccionario_de_datos.csv"))

sp_1_2020_goc <- sp_1_2020 %>% 
  mutate(entidad = stringr::str_to_lower(entidad)) %>% 
  filter(entidad %in% c("Baja California","Baja California Sur","Jalisco","Nayarit","Sinaloa","Sonora")) #### here is the issue

unique(sp_1_2020_goc$entidad)

#  [1] "aguascalientes"         "baja california"        "baja california sur"    "campeche"               "coahuila de zaragoza"   "colima"                
#  [7] "chiapas"                "chihuahua"              "ciudad de m\xc9xico"    "durango"                "guanajuato"             "guerrero"              
# [13] "hidalgo"                "jalisco"                "m\xc9xico"              "michoac\xc1n de ocampo" "morelos"                "nayarit"               
# [19] "nuevo le\xd3n"          "oaxaca"                 "puebla"   

na_2020 <-sp_1_2020_goc %>% 
  filter(is.na(longitud))

sf_sp_1_2020_goc <- sp_1_2020_goc %>% 
  st_as_sf(coords = c("longitud", "latitud"),
           crs = 4326, # WGS84
           remove = FALSE)
```

##### 2019 

Exploring 2019: making sure this looks similar to later years.

```{r}
sp_1_2019 <- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas/denue_00_72_1_1119_csv/conjunto_de_datos","denue_inegi_72_1.csv"))

# sp_1_2019_goc <- sp_1_2019 %>% 
  # filter(entidad %in% c("Baja California","Baja California Sur","Jalisco","Nayarit","Sinaloa","Sonora")) #### here is the issue also.

na_2019 <-sp_1_2019 %>% 
  filter(is.na(longitud)) # none even without filtering for GoC
```

##### 2023 

2023 looks suspiciously low for accommodations in terms of records (comparatively)

```{r}
sp_1_2023 <- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas/denue_00_72_1_1123_csv/conjunto_de_datos","denue_inegi_72_1.csv"))

sp_1_2023_goc <- sp_1_2023 %>% 
  mutate(entidad = stringr::str_to_lower(entidad)) %>% 
  filter(entidad %in% c("baja california","baja california sur","jalisco","nayarit","sinaloa","sonora")) 

na_2023 <-sp_1_2023_goc %>% 
  filter(is.na(longitud)) ## none

sf_sp_1_2023_goc <- sp_1_2023_goc %>% 
  st_as_sf(coords = c("longitud", "latitud"),
           crs = 4326, # WGS84
           remove = FALSE)

st_crs(sf_sp_1_2023_goc) == st_crs(inland_50) ## FALSE
sf_sp_1_2023_goc_trans <- st_transform(sf_sp_1_2023_goc, st_crs(inland_50))
sf_sp_1_2023_goc_50km <- st_intersection(sf_sp_1_2023_goc_trans, inland_50)

sf_sp_1_2023_goc_50km[] <- lapply(sf_sp_1_2023_goc_50km, function(x) {
  if (is.character(x)) iconv(x, from = "", to = "UTF-8", sub = "byte") else x
})

mapview(sf_sp_1_2023_goc_50km)

alojamientos_2023_sp1_goc_50km <- sf_sp_1_2023_goc_50km %>% 
  filter(nombre_act %in% c("Moteles","Hoteles sin otros servicios integrados","Pensiones y casas de hu<e9>spedes","Departamentos y casas amueblados con servicios de hoteler<ed>a","Campamentos y albergues recreativos","Hoteles con otros servicios integrados","Caba<f1>as, villas y similares")) #980 obs.

alojamientos_2023_sp1_goc_50km <- sf_sp_1_2023_goc_50km %>% 
  filter(nombre_act %in% alojamientos_categories) # 1453 obs.

comida_bebidas_2023_sp1_goc_50km <- sf_sp_1_2023_goc_50km %>% 
  filter(nombre_act %in% c("Cafeter<ed>as, fuentes de sodas, never<ed>as, refresquer<ed>as y similares",
  "Restaurantes con servicio de preparaci<f3>n de alimentos a la carta o de comida corrida",
  "Servicios de preparaci<f3>n de otros alimentos para consumo inmediato",
  "Restaurantes con servicio de preparaci<f3>n de antojitos",
  "Restaurantes con servicio de preparaci<f3>n de tacos y tortas",
  "Restaurantes con servicio de preparaci<f3>n de pizzas, hamburguesas, hot dogs y pollos rostizados para llevar",
  "Restaurantes con servicio de preparaci<f3>n de pescados y mariscos",
  "Bares, cantinas y similares",
  "Restaurantes de autoservicio",
  "Restaurantes que preparan otro tipo de alimentos para llevar",
  "Centros nocturnos, discotecas y similares",
  "Servicios de preparaci<f3>n de alimentos en unidades m<f3>viles",
  "Servicios de preparaci<f3>n de alimentos para ocasiones especiales",
  "Servicios de comedor para empresas e instituciones")) # 1843 obs

comida_bebidas_2023_sp1_goc_50km <- sf_sp_1_2023_goc_50km %>% 
  filter(nombre_act %in% comida_bebidas_categories) # 18337 obs.

###############################
sp_2_2023 <- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas/denue_00_72_2_1123_csv/conjunto_de_datos","denue_inegi_72_2.csv"))

sp_2_2023_goc <- sp_2_2023 %>% 
  mutate(entidad = stringr::str_to_lower(entidad)) %>% 
  filter(entidad %in% c("baja california","baja california sur","jalisco","nayarit","sinaloa","sonora")) 

na_2023 <-sp_2_2023_goc %>% 
  filter(is.na(longitud)) ## none

sf_sp_2_2023_goc <- sp_2_2023_goc %>% 
  st_as_sf(coords = c("longitud", "latitud"),
           crs = 4326, # WGS84
           remove = FALSE)

sf_sp_2_2023_goc_trans <- st_transform(sf_sp_2_2023_goc, st_crs(inland_50))
sf_sp_2_2023_goc_50km <- st_intersection(sf_sp_2_2023_goc_trans, inland_50)

sf_sp_2_2023_goc_50km[] <- lapply(sf_sp_2_2023_goc_50km, function(x) {
  if (is.character(x)) iconv(x, from = "", to = "UTF-8", sub = "byte") else x
})

mapview(sf_sp_2_2023_goc_50km)

unique(sf_sp_2_2023_goc_50km$nombre_act)

alojamientos_2023_sp2_goc_50km <- sf_sp_2_2023_goc_50km %>% 
  filter(nombre_act %in% c("Moteles","Hoteles sin otros servicios integrados","Pensiones y casas de hu<e9>spedes","Departamentos y casas amueblados con servicios de hoteler<ed>a","Campamentos y albergues recreativos","Hoteles con otros servicios integrados","Caba<f1>as, villas y similares")) 
unique(alojamientos_2023_sp2_goc_50km$nombre_act)

comida_bebidas_2023_sp2_goc_50km <- sf_sp_2_2023_goc_50km %>% 
  filter(nombre_act %in% c("Cafeter<ed>as, fuentes de sodas, never<ed>as, refresquer<ed>as y similares",
  "Restaurantes con servicio de preparaci<f3>n de alimentos a la carta o de comida corrida",
  "Servicios de preparaci<f3>n de otros alimentos para consumo inmediato",
  "Restaurantes con servicio de preparaci<f3>n de antojitos",
  "Restaurantes con servicio de preparaci<f3>n de tacos y tortas",
  "Restaurantes con servicio de preparaci<f3>n de pizzas, hamburguesas, hot dogs y pollos rostizados para llevar",
  "Restaurantes con servicio de preparaci<f3>n de pescados y mariscos",
  "Bares, cantinas y similares",
  "Restaurantes de autoservicio",
  "Restaurantes que preparan otro tipo de alimentos para llevar",
  "Centros nocturnos, discotecas y similares",
  "Servicios de preparaci<f3>n de alimentos en unidades m<f3>viles",
  "Servicios de preparaci<f3>n de alimentos para ocasiones especiales",
  "Servicios de comedor para empresas e instituciones"))

mapview(comida_bebidas_2023_sp2_goc_50km) + mapview(comida_bebidas_2023_sp1_goc_50km)

```

##### 2024

Exploring 2024: what I did to write the function `process_denue_year`!

```{r}
sp_1_2024 <- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas/denue_00_72_1_1124_csv/conjunto_de_datos","denue_inegi_72_1.csv"))

sp_1_2024_diccionario_de_datos <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas/denue_00_72_1_1124_csv/diccionario_de_datos","denue_diccionario_de_datos.csv"))

sp_1_2024_goc <- sp_1_2024 %>% 
  filter(entidad %in% c("Baja California","Baja California Sur","Jalisco","Nayarit","Sinaloa","Sonora"))

sf_sp_1_2024_goc <- sp_1_2024_goc %>% 
  st_as_sf(coords = c("longitud", "latitud"),
           crs = 4326, # WGS84
           remove = FALSE)

sf_sp_1_2024_goc_trans <- st_transform(sf_sp_1_2024_goc, st_crs(inland_10))
sf_sp_1_2024_goc_10km <- st_intersection(sf_sp_1_2024_goc_trans, inland_10)

sf_sp_1_2024_goc_10km[] <- lapply(sf_sp_1_2024_goc_10km, function(x) {
  if (is.character(x)) iconv(x, from = "", to = "UTF-8", sub = "byte") else x
})

mapview(sf_sp_1_2024_goc_10km)
```

```{r}
sp_2_2024 <- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas /denue_00_72_2_1124_csv/conjunto_de_datos","denue_inegi_72_2.csv"))

sp_2_2024_goc <- sp_2_2024 %>% 
  filter(entidad %in% c("Baja California","Baja California Sur","Jalisco","Nayarit","Sinaloa","Sonora"))

sf_sp_2_2024_goc <- sp_2_2024_goc %>% 
  st_as_sf(coords = c("longitud", "latitud"),
           crs = 4326, # WGS84
           remove = FALSE)

sf_sp_2_2024_goc_trans <- st_transform(sf_sp_2_2024_goc, st_crs(inland_10))
sf_sp_2_2024_goc_10km <- st_intersection(sf_sp_2_2024_goc_trans, inland_10)

sf_sp_2_2024_goc_10km[] <- lapply(sf_sp_2_2024_goc_10km, function(x) {
  if (is.character(x)) iconv(x, from = "", to = "UTF-8", sub = "byte") else x
})

mapview(sf_sp_2_2024_goc_10km)
```

Join them together:
```{r}
alojamientos_bebidas_2024_goc_10km <- rbind(sf_sp_1_2024_goc_10km, sf_sp_2_2024_goc_10km) %>% 
  mutate(data_year = 2024)
mapview(alojamientos_bebidas_2024_goc_10km)
```


What are the types of services here?
```{r}
unique(alojamientos_bebidas_2024_goc_10km$nombre_act)

#  [1] "Cafeter<ed>as, fuentes de sodas, never<ed>as, refresquer<ed>as y similares"                                   
#  [2] "Caba<f1>as, villas y similares"                                                                               
#  [3] "Restaurantes con servicio de preparaci<f3>n de alimentos a la carta o de comida corrida"                      
#  [4] "Servicios de preparaci<f3>n de otros alimentos para consumo inmediato"                                        
#  [5] "Restaurantes con servicio de preparaci<f3>n de antojitos"                                                     
#  [6] "Restaurantes con servicio de preparaci<f3>n de tacos y tortas"                                                
#  [7] "Restaurantes con servicio de preparaci<f3>n de pizzas, hamburguesas, hot dogs y pollos rostizados para llevar"
#  [8] "Hoteles con otros servicios integrados"                                                                       
#  [9] "Restaurantes con servicio de preparaci<f3>n de pescados y mariscos"                                           
# [10] "Bares, cantinas y similares"                                                                                  
# [11] "Restaurantes de autoservicio"                                                                                 
# [12] "Campamentos y albergues recreativos"                                                                          
# [13] "Hoteles sin otros servicios integrados"                                                                       
# [14] "Departamentos y casas amueblados con servicios de hoteler<ed>a"                                               
# [15] "Pensiones y casas de hu<e9>spedes"                                                                            
# [16] "Restaurantes que preparan otro tipo de alimentos para llevar"                                                 
# [17] "Moteles"                                                                                                      
# [18] "Centros nocturnos, discotecas y similares"                                                                    
# [19] "Servicios de preparaci<f3>n de alimentos en unidades m<f3>viles"                                              
# [20] "Servicios de preparaci<f3>n de alimentos para ocasiones especiales"                                           
# [21] "Servicios de comedor para empresas e instituciones"  
```

Separate out the accommodations:

```{r}
alojamientos_2024_goc_10km <- alojamientos_bebidas_2024_goc_10km %>% 
  filter(nombre_act %in% c("Moteles","Hoteles sin otros servicios integrados","Pensiones y casas de hu<e9>spedes","Departamentos y casas amueblados con servicios de hoteler<ed>a","Campamentos y albergues recreativos","Hoteles con otros servicios integrados","Caba<f1>as, villas y similares"))

mapview(alojamientos_2024_goc_10km)
```

And the food & drink services:

```{r}
comida_bebidas_2024_goc_10km <- alojamientos_bebidas_2024_goc_10km %>% 
  filter(nombre_act %in% c("Cafeter<ed>as, fuentes de sodas, never<ed>as, refresquer<ed>as y similares","Restaurantes con servicio de preparaci<f3>n de alimentos a la carta o de comida corrida","Servicios de preparaci<f3>n de otros alimentos para consumo inmediato","Restaurantes con servicio de preparaci<f3>n de antojitos","Restaurantes con servicio de preparaci<f3>n de tacos y tortas","Restaurantes con servicio de preparaci<f3>n de pizzas, hamburguesas, hot dogs y pollos rostizados para llevar","Restaurantes con servicio de preparaci<f3>n de pescados y mariscos","Bares, cantinas y similares","Restaurantes de autoservicio","Restaurantes que preparan otro tipo de alimentos para llevar","Centros nocturnos, discotecas y similares","Servicios de preparaci<f3>n de alimentos en unidades m<f3>viles","Servicios de preparaci<f3>n de alimentos para ocasiones especiales","Servicios de comedor para empresas e instituciones"))

mapview(comida_bebidas_2024_goc_10km)
```

```{r}
categories_2023_sp1 <- debug_2023$sp_1_50km %>%
  st_drop_geometry() %>%
  distinct(nombre_act)

categories_2023_sp2 <- debug_2023$sp_2_50km %>%
  st_drop_geometry() %>%
  distinct(nombre_act)

unique_categories_2023 <- bind_rows(
  categories_2023_sp1,
  categories_2023_sp2
) %>%
  distinct(nombre_act) %>%
  arrange(nombre_act)
unique_categories_2023

categories_2024_sp1 <- debug_2024$sp_1_50km %>%
  st_drop_geometry() %>%
  distinct(nombre_act)

categories_2024_sp2 <- debug_2024$sp_2_50km %>%
  st_drop_geometry() %>%
  distinct(nombre_act)

unique_categories_2024 <- bind_rows(
  categories_2024_sp1,
  categories_2024_sp2
) %>%
  distinct(nombre_act) %>%
  arrange(nombre_act)
unique_categories_2024

# Compare them
setdiff(unique_categories_2024$nombre_act, unique_categories_2023$nombre_act)
setdiff(unique_categories_2023$nombre_act, unique_categories_2024$nombre_act)
```




## agricultura_cria_explotacion_forestal_pesca_y_caza


## mineria


## servicios_apoyo_negocios_y_manejo_desechos_y_servicios_de_remediación


## servicios_esparcimiento_culturales_y_deportivos_servicios_recreativos

```{r}
library(tidyverse)
library(magrittr)
library(here)
library(janitor)
library(tmaptools)
library(sf)
library(mapview)

inland_10 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_10km_buffer","inland_10km_goc.shp"))
inland_50 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer","inland_50km_goc.shp"))
```

### Exploring 2024

```{r}
sp_71_2024 <- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_esparcimiento_culturales_y_deportivos_servicios_recreativos/denue_00_71_1124_csv/conjunto_de_datos","denue_inegi_71_.csv"))

sp_71_2024_diccionario_de_datos <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_esparcimiento_culturales_y_deportivos_servicios_recreativos/denue_00_71_1123_csv/diccionario_de_datos","denue_diccionario_de_datos.csv"))

sp_71_2024_goc <- sp_71_2024 %>% 
  filter(entidad %in% c("Baja California","Baja California Sur","Jalisco","Nayarit","Sinaloa","Sonora"))

sf_sp_71_2024_goc <- sp_71_2024_goc %>% 
  st_as_sf(coords = c("longitud", "latitud"),
           crs = 4326, # WGS84
           remove = FALSE)

sf_sp_71_2024_goc_trans <- st_transform(sf_sp_71_2024_goc, st_crs(inland_50))
sf_sp_71_2024_goc_50km <- st_intersection(sf_sp_71_2024_goc_trans, inland_50)

sf_sp_71_2024_goc_50km[] <- lapply(sf_sp_71_2024_goc_50km, function(x) {
  if (is.character(x)) iconv(x, from = "", to = "UTF-8", sub = "byte") else x
})

mapview(sf_sp_71_2024_goc_50km)
```

```{r}
### fix encoding issues... manually for now...
unique(sf_sp_71_2024_goc_50km$nombre_act)

sf_sp_71_2024_goc_50km_clean <- sf_sp_71_2024_goc_50km %>%
  mutate(
    nombre_act = case_when(
      TRUE ~ nombre_act %>%
        str_replace_all("<e1>", "á") %>%
        str_replace_all("<e9>", "é") %>%
        str_replace_all("<ed>", "í") %>%
        str_replace_all("<f3>", "ó") %>%
        str_replace_all("<fa>", "ú") %>%
        str_replace_all("<f1>", "ñ")
    )
  )

unique(sf_sp_71_2024_goc_50km_clean$nombre_act) ### make vector of those we are interested in
```



**Separate out the types of service activities:**

*1. Sports/fitness*
  
  "Private fitness centers",
  "Public fitness centers",
  "Private sports clubs",
  "Public sports clubs",
  "Professional sports teams",
  "Professional athletes",
  "Amateur clubs or leagues",
  "Golf courses"
  
  "Centros de acondicionamiento físico del sector privado",
  "Centros de acondicionamiento físico del sector público",
    "Clubes deportivos del sector privado",
    "Clubes deportivos del sector público",
    "Equipos deportivos profesionales",
    "Deportistas profesionales",
    "Clubes o ligas de aficionados",
    "Campos de golf"

*2. arts and entertainment*

“Independent artists, writers, and technicians,”
  “Private sector dance companies,”
  “Public sector theater companies,”
  “Private sector theater companies,”
  “Public sector music groups,”
  “Private sector singers and music groups,”
  “Other private sector performing arts companies and groups,”
  
  "Artistas, escritores y técnicos independientes",
    "Compañías de danza del sector privado",
    "Compañías de teatro del sector público",
    "Compañías de teatro del sector privado",
    "Grupos musicales del sector público",
    "Cantantes y grupos musicales del sector privado",
    "Otras compañías y grupos de espectáculos artísticos del sector privado"
  
*3. Museums*

  “Public sector museums,”
  “Private sector museums,”
  
  "Museos del sector público",
    "Museos del sector privado"
  

*4. Theme parks/recreation*

  "Private sector amusement and theme parks",
  "Public sector amusement and theme parks",
  "Private sector water parks and resorts",
  "Public sector water parks and resorts",
  "Billiard halls",
  "Arcade arcades",
  "Bowling alleys",
  "Other private sector recreational services",
  "Other public sector recreational services"
  
  "Parques de diversiones y temáticos del sector privado",
    "Parques de diversiones y temáticos del sector público",
    "Parques acuáticos y balnearios del sector privado",
    "Parques acuáticos y balnearios del sector público",
    "Billares",
    "Casas de juegos electrónicos",
    "Boliches",
    "Otros servicios recreativos prestados por el sector privado",
    "Otros servicios recreativos prestados por el sector público"

*5. Gambling*

“Sale of lottery tickets, sports betting, and other raffle tickets,”
    “Other games of chance,”
    “Casinos”
    
   "Venta de billetes de lotería, pronósticos deportivos y otros boletos de sorteo",
    "Otros juegos de azar",
    "Casinos"

*6. Nature, parks, heritage sites*

  “Caves, natural parks, and other sites of national cultural heritage,”
  “Private botanical gardens and zoos,”
  “Public botanical gardens and zoos”
  
  "Grutas, parques naturales y otros sitios del patrimonio cultural de la nación",
    "Jardines botánicos y zoológicos del sector privado",
    "Jardines botánicos y zoológicos del sector público"

*7. Coastal/marine tourism*

  “Tourist marinas”
  
  "Marinas turísticas"

    
*8. Agents or event production*

   “Public sector promoters of artistic, cultural, sporting, and similar events that have facilities to host them,”
    “Private sector promoters of artistic, cultural, sporting, and similar events that have facilities to host them,”
    “Promoters of artistic, cultural, sporting, and similar events that do not have facilities to host them,”
    “Agents and representatives of artists, athletes, and similar professionals”
    
  "Promotores del sector público de espectáculos artísticos, culturales, deportivos y similares que cuentan con instalaciones para presentarlos",
    "Promotores del sector privado de espectáculos artísticos, culturales, deportivos y similares que cuentan con instalaciones para presentarlos",
    "Promotores de espectáculos artísticos, culturales, deportivos y similares que no cuentan con instalaciones para presentarlos",
    "Agentes y representantes de artistas, deportistas y similares"
    
*9. Historic sites*

  “Historic sites”    
  "Sitios históricos"


For now, I am going to only use the Museums and historic sites, Nature, parks and heritage sites, and coastal/marine tourism groups.

```{r}
museums_2024_goc_10km <- sf_sp_71_2024_goc_50km_clean %>%
  filter(nombre_act %in% c(
    "Museos del sector público",
    "Museos del sector privado"
  ))
mapview(museums_2024_goc_10km)


historical_sites_2024_goc_10km <- sf_sp_71_2024_goc_50km_clean %>% 
  filter(nombre_act %in% c(
    "Sitios históricos"
  ))
mapview(historical_sites_2024_goc_10km)


nature_heritage_2024_goc_10km <- sf_sp_71_2024_goc_50km_clean %>%
  filter(nombre_act %in% c(
    "Grutas, parques naturales y otros sitios del patrimonio cultural de la nación",
    "Jardines botánicos y zoológicos del sector privado",
    "Jardines botánicos y zoológicos del sector público"
  ))
mapview(nature_heritage_2024_goc_10km)

marine_tourism_2024_goc_10km <- sf_sp_71_2024_goc_50km_clean %>%
  filter(nombre_act %in% c(
    "Marinas turísticas"
  ))

mapview(marine_tourism_2024_goc_10km)
```

### Function to read in all years of data

```{r}
library(tidyverse)
library(magrittr)
library(here)
library(janitor)
library(tmaptools)
library(sf)
library(mapview)

museums_categories <- c(
  "Museos del sector público",
  "Museos del sector privado"
)

historical_sites_categories <- c(
  "Sitios históricos"
)

nature_heritage_categories <- c(
  "Grutas, parques naturales y otros sitios del patrimonio cultural de la nación",
  "Jardines botánicos y zoológicos del sector privado",
  "Jardines botánicos y zoológicos del sector público"
)

marine_tourism_categories <- c(
  "Marinas turísticas"
)

recreation_other_categories <- c(
  "Parques de diversiones y temáticos del sector privado",
  "Parques de diversiones y temáticos del sector público",
  "Parques acuáticos y balnearios del sector privado",
  "Parques acuáticos y balnearios del sector público",
  "Billares",
  "Casas de juegos electrónicos",
  "Boliches",
  "Otros servicios recreativos prestados por el sector privado",
  "Otros servicios recreativos prestados por el sector público"
)

# since the folders are not saved by year but instead by month and year, but are not consistent in the month for each year, lets extract the "suffix" that corresponds to each year.
get_folder_suffix <- function(year) {
  # years 2016 - 2019 would use their respective month/year codes
  # ex: 2016 is _1016_, 2017 is _1117_
  suffixes <- list(
    "2016" = "1016",
    "2017" = "1117", 
    "2018" = "1118",
    "2019" = "1119",
    "2020" = "1120",
    "2021" = "1121",
    "2022" = "1122",
    "2023" = "1123",
    "2024" = "1124",
    "2025" = "0525"
  )
  return(suffixes[[as.character(year)]])
}

###########################
# function to process data for a given year
process_denue_cultural_year <- function(year, inland_50, base_path = "/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025") {
  
  cat(paste0("Processing year: ", year, "\n"))
  
  # for each year, obtain the corresponding date and year "suffix"
  suffix <- get_folder_suffix(year)
  
  
  # the estados
  goc_states <- c("baja california", "baja california sur", "jalisco", 
                  "nayarit", "sinaloa", "sonora")
  
  # read in DENUE 71
  cat("  Reading 71 dataset...\n")
  sp_71_path <- here(base_path, 
                     "servicios_esparcimiento_culturales_y_deportivos_servicios_recreativos",
                     paste0("denue_00_71_", suffix, "_csv"),
                     "conjunto_de_datos",
                     "denue_inegi_71_.csv")
  
  sp_71 <- read_csv(sp_71_path)
  
  cat("  Mutating 71 dataset...\n")
  sp_71_goc <- sp_71 %>%
    mutate(entidad = stringr::str_to_lower(entidad)) %>% 
    select(-c(edificio_e, telefono)) %>% 
    filter(entidad %in% goc_states) %>% 
    mutate(data_year = year)
  
  cat("  Converting 71 to spatial object ...\n")
  sf_sp_71_goc <- sp_71_goc %>%
    st_as_sf(coords = c("longitud", "latitud"),
             crs = 4326,
             remove = FALSE)
  
  cat("  Transforming 71 dataset...\n")
  sf_sp_71_goc_trans <- st_transform(sf_sp_71_goc, st_crs(inland_50))
  
  cat("  Intersct 71 dataset...\n")
  sf_sp_71_goc_50km <- st_intersection(sf_sp_71_goc_trans, inland_50)
  
  cat("  Fix UTF issue for 71 dataset...\n")
  sf_sp_71_goc_50km[] <- lapply(sf_sp_71_goc_50km, function(x) {
    if (is.character(x)) iconv(x, from = "", to = "UTF-8", sub = "byte") else x
  })
  
  cat("  Fixing UTF issue futher ...\n")
  sf_sp_71_goc_50km_clean <- sf_sp_71_goc_50km %>% 
    mutate(
    nombre_act = case_when(
      TRUE ~ nombre_act %>%
        str_replace_all("<e1>", "á") %>%
        str_replace_all("<e9>", "é") %>%
        str_replace_all("<ed>", "í") %>%
        str_replace_all("<f3>", "ó") %>%
        str_replace_all("<fa>", "ú") %>%
        str_replace_all("<f1>", "ñ")
    ))
  
  # --- separate out recreation types ---
  
  cat("  Filtering museums...\n")
  museums_goc_50km <- sf_sp_71_goc_50km_clean %>%
    filter(nombre_act %in% museums_categories)
  
  cat("  Filtering historical sites...\n")
  historical_sites_goc_50km <- sf_sp_71_goc_50km_clean %>%
    filter(nombre_act %in% historical_sites_categories)
  
  cat("  Filtering nature and heritage...\n")
  nature_heritage_goc_50km <- sf_sp_71_goc_50km_clean %>%
    filter(nombre_act %in% nature_heritage_categories)
  
  cat("  Filtering marine tourism...\n")
  marine_tourism_goc_50km <- sf_sp_71_goc_50km_clean %>%
    filter(nombre_act %in% marine_tourism_categories)
  
  cat("  Filtering other recreation...\n")
  recreation_other_goc_50km <- sf_sp_71_goc_50km_clean %>%
    filter(nombre_act %in% recreation_other_categories)
  
  cat(paste0("  Completed year ", year, "!\n"))
  cat(paste0("    Museums: ", nrow(museums_goc_50km), " records\n"))
  cat(paste0("    Historical sites: ", nrow(historical_sites_goc_50km), " records\n"))
  cat(paste0("    Nature heritage sites: ", nrow(nature_heritage_goc_50km), " records\n"))
  cat(paste0("    Marine tourism: ", nrow(marine_tourism_goc_50km), " records\n\n"))
  cat(paste0("    Other recreation: ", nrow(recreation_other_goc_50km), " records\n"))
  
  # returning them as a list
  return(list(
    museums = museums_goc_50km,
    historical_sites = historical_sites_goc_50km,
    nature_heritage = nature_heritage_goc_50km,
    marine_tourism = marine_tourism_goc_50km,
    recreation_other = recreation_other_goc_50km
  ))
}

# GoC 50km buffer
inland_50 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer","inland_50km_goc.shp"))

# check with 2024 that I did individually
## using 50km
result_2024 <- process_denue_cultural_year(2024, inland_50) ## yay!! same as my exploration
marine_2024 <- result_2024$marine_tourism
```

### Using function

```{r}
##########################################
# using 50km buffer
# process years 2019 to 2024!
years <- 2019:2024
culture_results_50km <- lapply(years, function(y) process_denue_cultural_year(y, inland_50))
names(culture_results_50km) <- paste0("year_", years)
```


```{r}
# joining all of these into one object for each recreation type.

all_museums_goc_50km <- bind_rows(lapply(culture_results_50km, function(x) x$museums))
mapview(all_museums_goc_50km, zcol = "data_year") ## ageb/municipio shifts in 2021, so it definitely makes sense to aggregate by region rather than by municipio.  See ID 6498864.

all_historical_sites_goc_50km <- bind_rows(lapply(culture_results_50km, function(x) x$historical_sites))
mapview(all_historical_sites_goc_50km, zcol = "data_year")

all_nature_heritage_goc_50km <- bind_rows(lapply(culture_results_50km, function(x) x$nature_heritage))
mapview(all_nature_heritage_goc_50km, zcol = "data_year")

all_marine_tourism_goc_50km <- bind_rows(lapply(culture_results_50km, function(x) x$marine_tourism))
mapview(all_marine_tourism_goc_50km, zcol = "data_year")

all_recreation_other_goc_50km <- bind_rows(lapply(culture_results_50km, function(x) x$recreation_other))
mapview(all_recreation_other_goc_50km, zcol = "data_year")
```

### Write out recreation service businesses

```{r}
############## 50km ###############
output_dir <- here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_esparcimiento_culturales_y_deportivos_servicios_recreativos/int")

recreation_types_list <- list(
  museums = all_museums_goc_50km,
  historical_sites = all_historical_sites_goc_50km,
  nature_heritage = all_nature_heritage_goc_50km,
  marine_tourism = all_marine_tourism_goc_50km,
  recreation_other = all_recreation_other_goc_50km
)

## loop to write out the recreation businesses
for (x in names(recreation_types_list)) {
  
  message("Writing out: ", x)
  
  st_write(recreation_types_list[[x]], dsn = here::here(output_dir, paste0(x, "_2019_2024_goc_50km.shp")), delete_layer = TRUE)
  
}
```



### Plotly businesses

```{r}

# all combined datasets in a named list
culture_combined_list <- list(
  museums = all_museums_goc_50km,
  historical_sites = all_historical_sites_goc_50km,
  nature_heritage = all_nature_heritage_goc_50km,
  nautical_services = all_marine_tourism_goc_50km,
  recreation_other = all_recreation_other_goc_50km
)

## count rows per year for each category
culture_counts <- map_df(
  names(culture_combined_list),
  ~ culture_combined_list[[.x]] %>% 
    count(data_year, name = "n") %>%
    mutate(category = .x),
  .id = NULL
)

culture_counts

culture_counts_df <- culture_counts %>% 
  st_drop_geometry()

culture_counts_plotly <- plot_ly(culture_counts_df,
        x = ~data_year,
        y = ~n,
        color = ~category,
        type = 'scatter',
        mode = 'lines+markers') %>%
  layout(
    title = "DENUE observations by year (50km buffer)",
    xaxis = list(title = "Data year"),
    yaxis = list(title = "Total observations"),
    legend = list(title=list(text="Recreation type"))
  )
culture_counts_plotly

htmlwidgets::saveWidget(culture_counts_plotly, here::here("TR/tr_quality_int","culture_obs_counts_2019_2024_120325.html"), selfcontained = TRUE)
```


See if there are businesses that keep the same IDs over time:

```{r}
# id_years <- all_museums_goc_50km %>%
#   count(id, data_year) %>%
#   count(id, name = "n_years") %>%
#   arrange(desc(n_years))

# museum_id_persistence <- all_museums_goc_50km %>%
#   distinct(id, data_year) %>% # remove duplicate year-ID combos
#   count(id, name = "n_years") # count years per DENUE ID
# 
# museum_persistence_summary <- museum_id_persistence %>%
#   count(n_years, name = "n_ids") %>%
#   mutate(
#     pct = 100 * n_ids / sum(n_ids)
#   )
# 
# plot_ly(
#   museum_persistence_summary,
#   x = ~n_years,
#   y = ~pct,
#   type = "bar",
#   text = ~paste0(round(pct, 1), "%"),
#   textposition = "outside"
# ) %>%
#   layout(
#     title = "Percentage of Museum DENUE IDs observed across *n* years (Museums, 50km Buffer)",
#     xaxis = list(title = "Number of Years the Business Appears in DENUE"),
#     yaxis = list(title = "Percent of Total IDs"),
#     bargap = 0.2
#   )

```

Add all onto the same one:

```{r}
recreation_types_list <- list(
  museums = all_museums_goc_50km,
  historical_sites = all_historical_sites_goc_50km,
  nature_heritage = all_nature_heritage_goc_50km,
  nautical_services = all_marine_tourism_goc_50km,
  recreation_other = all_recreation_other_goc_50km
)

# func to compute persistence for a single dataset
total_persistence <- function(df, type_name) {
  df %>%
    distinct(id, data_year) %>%
    count(id, name = "n_years") %>%
    count(n_years, name = "n_ids") %>%
    mutate(
      pct = 100 * n_ids / sum(n_ids),
      category = type_name
    )
}

# apply across list
persistence_all <- map_df(
  names(recreation_types_list),
  ~ compute_persistence(recreation_types_list[[.x]], .x)
)

## plotly bar graph!!!
business_persistence_plotly <- plot_ly(
  data = persistence_all,
  x = ~n_years,
  y = ~pct,
  color = ~category,
  type = "bar",
  text = ~paste0(round(pct, 1), "%"),
  textposition = "outside"
) %>%
  layout(
    title = "Percent of DENUE IDs observed across *n* years in 2019 - 2024 (all recreation types, 50km buffer)",
    xaxis = list(title = "Number of years the business appears in DENUE"),
    yaxis = list(title = "Percent of total IDs"),
    barmode = "group",
    bargap = 0.15
  )
business_persistence_plotly

htmlwidgets::saveWidget(business_persistence_plotly, here::here("TR/tr_quality_int","culture_id_persistence_2019_2024_120325.html"), selfcontained = TRUE)

```

### Plotly employees

```{r}
convert_employees <- function(x) {
  case_when(
    str_detect(x, "^0 a 5") ~ 3.5,
    str_detect(x, "^6 a 10") ~ 8,
    str_detect(x, "^11 a 30") ~ 20,
    str_detect(x, "^31 a 50") ~ 40,
    str_detect(x, "^51 a 100") ~ 75,
    str_detect(x, "^101 a 250") ~ 175,
    str_detect(x, "251") ~ 300,
    TRUE ~ NA_real_
  )
}

recreation_types_list <- list(
  museums = all_museums_goc_50km,
  historical_sites = all_historical_sites_goc_50km,
  nature_heritage = all_nature_heritage_goc_50km,
  marine_tourism = all_marine_tourism_goc_50km,
  recreation_other = all_recreation_other_goc_50km
)

recreation_with_emp <- lapply(
  recreation_types_list,
  function(df) df %>% mutate(emp_mid = convert_employees(per_ocu))
)

employees_time <- map_df(
  names(recreation_with_emp),
  ~ recreation_with_emp[[.x]] %>%
      group_by(data_year) %>%
      summarise(total_emp = sum(emp_mid, na.rm = TRUE)) %>%
      mutate(category = .x)
) %>% 
  st_drop_geometry()
```


```{r}
total_est_employees <- plot_ly(
  employees_time,
  x = ~data_year,
  y = ~total_emp,
  color = ~category,
  type = "scatter",
  mode = "lines+markers"
) %>%
  layout(
    title = "Total estimated employees per year by recreation type (50km buffer)",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Estimated total employees")
  )
total_est_employees

htmlwidgets::saveWidget(total_est_employees, here::here("TR/tr_quality_int","culture_total_est_employees_2019_2024_120325.html"), selfcontained = TRUE)

```


## servicios_profesionales_científicos_técnicos

