---
title: 'OHI GoC 2025: TR Quantity Gapfilling'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# Seeing if the number of employees INEGIs DENUE "Servicios de alojamiento temporal y de preparación de alimentos y bebidas" has a relationship with tourist arrivals (SECTUR Hotel Monitoring Program)

See ~TR/denue_prep.qmd for the data downloading and cleaning up to this point.

## Bring in Accommodations

We will need to clean it up, because the number of employees for each business is binned.

*This is what the per_ocu variable description says in the metadata:*
"This includes personnel directly employed by the company and external personnel supplied by another company, who worked for the economic unit, under its direction and control, and who covered at least one-third of the workday. This may include permanent staff, paid temporary staff, or unpaid temporary staff.
Economic units are classified by employee size ranges, which allows for the identification of the size of Economic Units based on the number of personnel they employ; that is, according to their total workforce. For the purposes of the Directory, Economic Units are grouped into employee size ranges or strata, based on the total number of employees reported by each unit, as shown in the following table.
Number of Employees
1 = 0 to 5
2 = 6 to 10
3 = 11 to 30
4 = 31 to 50
5 = 51 to 100
6 = 101 to 250
7 = 251 and above" 

So, we will take the midpoint of each range.

```{r}
library(tidyverse)
library(magrittr)
library(here)
library(janitor)
library(tmaptools)
library(sf)
library(mapview)

alojamientos_2019_2024_50km <- st_read(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas/int","alojamientos_2019_2024_goc_50km.shp"))

## see what the variable per_ocu looks like
unique(alojamientos_2019_2024_50km$per_ocu)

# [1] "0 a 5 personas"        "6 a 10 personas"       "11 a 30 personas"      "31 a 50 personas"      "101 a 250 personas"    "51 a 100 personas"     "251 y m<e1>s personas" "251 y más personas" 

## use a case_when to reclassify the number of people employed based on the midpoint of the range.  NOTE that I am assuming that you can never have 0 employees, so the first midpoint is between 1 and 5.
alojamientos_est_perocu <- alojamientos_2019_2024_50km %>% 
  mutate(
    employees_estimated = case_when(
      per_ocu == "0 a 5 personas" ~ 3,
      per_ocu == "6 a 10 personas" ~ 8,
      per_ocu == "11 a 30 personas" ~ 20.5,
      per_ocu == "31 a 50 personas" ~ 40.5,
      per_ocu == "51 a 100 personas" ~ 75.5,
      per_ocu == "101 a 250 personas" ~ 175.5,
      per_ocu %in% c("251 y m<e1>s personas", "251 y más personas") ~ 300,
      TRUE ~ NA_real_
    )
  )

mapview(alojamientos_est_perocu, zcol = "employees_estimated")

unique(alojamientos_est_perocu$municipio)

# when 10km:
#  [1] "Mexicali"                 "Ensenada"                 "Muleg<e9>"                "Los Cabos"                "La Paz"                   "Loreto"                   "Puerto Vallarta"          "Cabo Corrientes"         
#  [9] "Bah<ed>a de Banderas"     "San Blas"                 "Compostela"               "Tecuala"                  "Puerto Pe<f1>asco"        "San Luis R<ed>o Colorado" "Hermosillo"               "Pitiquito"               
# [17] "Guaymas"                  "Empalme"                  "Benito Ju<e1>rez"         "Huatabampo"               "Elota"                    "Mazatl<e1>n"              "Escuinapa"                "San Ignacio"             
# [25] "Ahome"                    "Navolato"                 "Culiac<e1>n"              "Guasave"                  "Angostura"                "Rosario"                  "San Quint<ed>n"           "San Felipe"              
# [33] "San Quintín"              "Mulegé"                   "Bahía de Banderas"        "Puerto Peñasco"           "San Luis Río Colorado"    "Benito Juárez"            "Mazatlán"                 "Culiacán"                
# [41] "Caborca" 

# when 50km:
#  [1] "Mexicali"                   "Ensenada"                   "Muleg<e9>"                  "Comond<fa>"                
#  [5] "Los Cabos"                  "La Paz"                     "Loreto"                     "Puerto Vallarta"           
#  [9] "Talpa de Allende"           "Tomatl<e1>n"                "San Sebasti<e1>n del Oeste" "Cabo Corrientes"           
# [13] "Mascota"                    "Bah<ed>a de Banderas"       "Tepic"                      "San Blas"                  
# [17] "Compostela"                 "Acaponeta"                  "Ru<ed>z"                    "Xalisco"                   
# [21] "Tuxpan"                     "Tecuala"                    "Santiago Ixcuintla"         "Rosamorada"                
# [25] "Huajicori"                  "San Pedro Lagunillas"       "Puerto Pe<f1>asco"          "San Luis R<ed>o Colorado"  
# [29] "Hermosillo"                 "Caborca"                    "Pitiquito"                  "Navojoa"                   
# [33] "Cajeme"                     "Guaymas"                    "Huatabampo"                 "Etchojoa"                  
# [37] "Empalme"                    "Benito Ju<e1>rez"           "Elota"                      "Mazatl<e1>n"               
# [41] "Navolato"                   "Ahome"                      "Culiac<e1>n"                "El Fuerte"                 
# [45] "Escuinapa"                  "San Ignacio"                "Guasave"                    "Rosario"                   
# [49] "Mocorito"                   "Salvador Alvarado"          "Concordia"                  "Angostura"                 
# [53] "San Quint<ed>n"             "San Felipe"                 "San Quintín"                "Mulegé"                    
# [57] "Comondú"                    "Tomatlán"                   "San Sebastián del Oeste"    "Bahía de Banderas"         
# [61] "Ruíz"                       "Puerto Peñasco"             "San Luis Río Colorado"      "Benito Juárez"             
# [65] "Culiacán"                   "Mazatlán"    

unique(alojamientos_est_perocu$localidad)

# when 10km:
#  [1] "San Felipe"                            "Villas de Cortez"                      "Campo Beluga"                          "Campo Daggets"                         "Campo Sacrificio"                     
#   [6] "Punta Final (Villa Mar y Sol)"         "Campo Raquel y Larry"                  "Punta Bufeo [Campo Tur<ed>stico]"      "Bah<ed>a de los <c1>ngeles"            "La Ventana [Campo]"                   
#  [11] "Puerto San Francisquito"               "Heroica Muleg<e9>"                     "Santa Rosal<ed>a"                      "Cabo San Lucas"                        "Buenavista"                           
#  [16] "San Jos<e9> del Cabo"                  "La Cuadra Regina"                      "La Paz"                                "Chametla"                              "La Ribera"                            
#  [21] "El Sargento"                           "Puerto de Pichilingue"                 "Loreto"                                "La Ventana"                            "El Cardonal"                          
#  [26] "Nopol<f3>"                             "Buenos Aires"                          "Los Barriles"                          "Punta Pescadero"                       "Tripu<ed> [Trailer Park]"             
#  [31] "Boca de la Vinorama"                   "Danzante Bay"                          "Punta Prieta"                          "Ninguno"                               "El Leonero"                           
#  [36] "Puerto Vallarta"                       "Sun Resorts"                           "Mismaloya"                             "Yelapa"                                "Las Juntas y los Veranos"             
#  [41] "Le<f3>n de la Sierra"                  "La Quebrada"                           "Ixtapa"                                "Las Juntas"                            "Playa Gemelas"                        
#  [46] "Sayulita"                              "Nuevo Vallarta"                        "Guadalupe Victoria"                    "Rinc<f3>n de Guayabitos"               "Bucer<ed>as"                          
#  [51] "Aticama"                               "Cruz de Huanacaxtle"                   "El Porvenir"                           "La Pe<f1>ita de Jaltemba"              "Los Ayala"                            
#  [56] "Platanitos"                            "Lo de Marcos"                          "Chacala"                               "La Palma"                              "San Blas"                             
#  [61] "Matanch<e9>n (La Aguada)"              "San Francisco"                         "Corral del Risco (Punta de Mita)"      "Lidia [Bungalows]"                     "Playa de los Cocos"                   
#  [66] "La Joya"                               "P<e1>tzcuaro"                          "Monte<f3>n"                            "Punta Raza"                            "Las Varas"                            
#  [71] "La Manzanilla"                         "Flamingos"                             "Santa Cruz de Miramar"                 "Tondoroque"                            "Desarrollo Punta Mita"                
#  [76] "Bah<ed>a de Matanch<e9>n"              "El Crucero de Chacala"                 "Punta Negra (Careyeros)"               "San Jos<e9> del Valle"                 "Para<ed>so Sereno"                    
#  [81] "Mezcales"                              "Playa Escondida"                       "Universo [Colonia]"                    "Playa Novillero"                       "Grand Palladium Vallarta [Hotel]"     
#  [86] "San Vicente"                           "Las Jarretaderas"                      "San Clemente de Lima"                  "Arboledas [Bungalows]"                 "La Tranquila [Condominios]"           
#  [91] "Imanta"                                "Los Arroyos Verdes"                    "Dos Palmeras (Casa Pel<ed>canos)"      "El Tonino [Bloquera]"                  "Punta Monterrey"                      
#  [96] "Rancho Banderas"                       "Papelillos [Hotel]"                    "Puerto Pe<f1>asco"                     "Golfo de Santa Clara"                  "Bah<ed>a la Choya"                    
# [101] "Playa Arenosa"                         "Bah<ed>a de Kino"                      "Puerto Libertad"                       "Mayan Palace Hotel"                    "Playa Miramar"                        
# [106] "Las Conchas"                           "Heroica Guaymas"                       "San Carlos (San Carlos Nuevo Guaymas)" "Condominios Playa Blanca"              "Empalme"                              
# [111] "Villa Ju<e1>rez"                       "Agua Blanca"                           "Las Bocas"                             "Costa del Mar"                         "La Cruz"                              
# [116] "Mazatl<e1>n"                           "Color Marino"                          "Dimas (Estaci<f3>n Dimas)"             "Teacap<e1>n"                           "Topolobampo"                          
# [121] "Altata"                                "Isla del Bosque"                       "Eldorado"                              "El Saucito"                            "Celestino Gazca Villase<f1>or"        
# [126] "Las Glorias"                           "Fraccionamiento Isla Para<ed>so"       "La Reforma"                            "Gustavo D<ed>az Ordaz (El Carrizo)"    "La Escondida"                         
# [131] "Agua Verde"                            "Localidad Econ<f3>mica"                "Rancho Girolando"                      "El Patio"                              "Ejido Plan Nacional Agrario"          
# [136] "Alfonsinas"                            "Tailwind"                              "El Encuentro"                          "Los Mochis"                            "Zacualpan"                            
# [141] "Desarrollo Puerto Bah<ed>a"            "Litib<fa>"                             "Bahía de los Ángeles"                  "Punta Bufeo [Campo Turístico]"         "Heroica Mulegé"                       
# [146] "Santa Rosalía"                         "San José del Cabo"                     "Tripuí [Trailer Park]"                 "San Román"                             "Nopoló"                               
# [151] "Costa Vida Vallarta"                   "Rincón de Guayabitos"                  "Bucerías"                              "Bahía de Matanchén"                    "La Peñita de Jaltemba"                
# [156] "San José del Valle"                    "Dos Palmeras (Casa Pelícanos)"         "Pátzcuaro"                             "Desarrollo Puerto Bahía"               "Litibú"                               
# [161] "Puerto Peñasco"                        "Bahía de Kino"                         "Villa Juárez"                          "Mazatlán"                              "Campo Santa Fe"                       
# [166] "Teacapán"                              "Dimas (Estación Dimas)"                "Motel Popeye"                          "Gustavo Díaz Ordaz (El Carrizo)"       "Playa del Sol"                        
# [171] "Pap<e1> Fern<e1>ndez [Campo]"          "Los Vientos"                           "Campo Lobo"                            "Cabo Golf Cars"                        "Cuesta Blanca"                        
# [176] "Las <c1>nimas de Quimixto"             "Lienzo Charro"                         "Mezcalitos"                            "Para<ed>so Escondido [Colonia]"        "Jalcocot<e1>n"                        
# [181] "Las Tijerillas"                        "Luis Pe<f1>a"                          "Hip<f3>lito Gonz<e1>lez L<f3>pez"      "Limoncito Hills"                       "La Playa Estates"                     
# [186] "Haixa (Litib<fa>)"                     "Playa Encanto"                         "Puerto Lobos"                          "Mayan Lakes"                           "Playa Huatabampito"                   
# [191] "Coch<f3>rit" 

# when 50km:
#   [1] "San Felipe"                                      "Villas de Cortez"                               
#   [3] "Campo Beluga"                                    "Campo Daggets"                                  
#   [5] "Campo Sacrificio"                                "Punta Final (Villa Mar y Sol)"                  
#   [7] "Campo Raquel y Larry"                            "Punta Bufeo [Campo Tur<ed>stico]"               
#   [9] "Bah<ed>a de los <c1>ngeles"                      "La Ventana [Campo]"                             
#  [11] "Catavi<f1>a"                                     "Puerto San Francisquito"                        
#  [13] "Rancho Escondido"                                "Piedra Blanca"                                  
#  [15] "Heroica Muleg<e9>"                               "Santa Rosal<ed>a"                               
#  [17] "San Isidro"                                      "San Lino"                                       
#  [19] "Cabo San Lucas"                                  "Buenavista"                                     
#  [21] "San Jos<e9> del Cabo"                            "La Cuadra Regina"                               
#  [23] "La Paz"                                          "Chametla"                                       
#  [25] "El Triunfo"                                      "La Acacia"                                      
#  [27] "La Ribera"                                       "El Sargento"                                    
#  [29] "Puerto de Pichilingue"                           "Loreto"                                         
#  [31] "San Miguel de Comond<fa>"                        "La Ventana"                                     
#  [33] "El Cardonal"                                     "Nopol<f3>"                                      
#  [35] "Buenos Aires"                                    "Los Barriles"                                   
#  [37] "Punta Pescadero"                                 "Tripu<ed> [Trailer Park]"                       
#  [39] "Boca de la Vinorama"                             "Danzante Bay"                                   
#  [41] "Punta Prieta"                                    "Ninguno"                                        
#  [43] "El Leonero"                                      "Puerto Vallarta"                                
#  [45] "Talpa de Allende"                                "La Cruz de Loreto"                              
#  [47] "Sun Resorts"                                     "San Sebasti<e1>n del Oeste"                     
#  [49] "Mismaloya"                                       "Yelapa"                                         
#  [51] "Las Juntas y los Veranos"                        "Mascota"                                        
#  [53] "El Tuito"                                        "Puerta de Enmedio"                              
#  [55] "Le<f3>n de la Sierra"                            "La Quebrada"                                    
#  [57] "Ixtapa"                                          "San Nicol<e1>s (La Troje)"                      
#  [59] "La Puntilla de Maito"                            "Las Juntas"                                     
#  [61] "Playa Gemelas"                                   "Sayulita"                                       
#  [63] "Nuevo Vallarta"                                  "Tepic"                                          
#  [65] "Guadalupe Victoria"                              "Rinc<f3>n de Guayabitos"                        
#  [67] "Bucer<ed>as"                                     "Aticama"                                        
#  [69] "Cruz de Huanacaxtle"                             "El Porvenir"                                    
#  [71] "La Pe<f1>ita de Jaltemba"                        "Los Ayala"                                      
#  [73] "Platanitos"                                      "Lo de Marcos"                                   
#  [75] "Chacala"                                         "La Palma"                                       
#  [77] "San Blas"                                        "Matanch<e9>n (La Aguada)"                       
#  [79] "San Francisco"                                   "Corral del Risco (Punta de Mita)"               
#  [81] "Lidia [Bungalows]"                               "Playa de los Cocos"                             
#  [83] "La Joya"                                         "P<e1>tzcuaro"                                   
#  [85] "Monte<f3>n"                                      "Punta Raza"                                     
#  [87] "Las Varas"                                       "Acaponeta"                                      
#  [89] "La Manzanilla"                                   "Flamingos"                                      
#  [91] "Santa Cruz de Miramar"                           "Tondoroque"                                     
#  [93] "Desarrollo Punta Mita"                           "Compostela"                                     
#  [95] "Ru<ed>z"                                         "San Juan de Abajo"                              
#  [97] "Xalisco"                                         "Bah<ed>a de Matanch<e9>n"                       
#  [99] "Tuxpan"                                          "El Crucero de Chacala"                          
# [101] "Tecuala"                                         "Punta Negra (Careyeros)"                        
# [103] "Santiago Ixcuintla"                              "Rosamorada"                                     
# [105] "San Jos<e9> del Valle"                           "Para<ed>so Sereno"                              
# [107] "Huajicori"                                       "Mezcales"                                       
# [109] "Playa Escondida"                                 "Villa Hidalgo"                                  
# [111] "Universo [Colonia]"                              "Playa Novillero"                                
# [113] "Grand Palladium Vallarta [Hotel]"                "San Vicente"                                    
# [115] "Mexcaltit<e1>n de Uribe"                         "Las Jarretaderas"                               
# [117] "L<e1>zaro C<e1>rdenas [Colonia]"                 "San Clemente de Lima"                           
# [119] "Arboledas [Bungalows]"                           "La Tranquila [Condominios]"                     
# [121] "Imanta"                                          "Los Arroyos Verdes"                             
# [123] "Dos Palmeras (Casa Pel<ed>canos)"                "San Cayetano"                                   
# [125] "El Tonino [Bloquera]"                            "San Pedro Lagunillas"                           
# [127] "Punta Monterrey"                                 "Rancho Banderas"                                
# [129] "Papelillos [Hotel]"                              "Puerto Pe<f1>asco"                              
# [131] "Golfo de Santa Clara"                            "Bah<ed>a la Choya"                              
# [133] "Playa Arenosa"                                   "Bah<ed>a de Kino"                               
# [135] "Miguel Alem<e1>n (La Doce)"                      "Plutarco El<ed>as Calles (La Y Griega)"         
# [137] "Puerto Libertad"                                 "Mayan Palace Hotel"                             
# [139] "Playa Miramar"                                   "Las Conchas"                                    
# [141] "Navojoa"                                         "Ciudad Obreg<f3>n"                              
# [143] "Heroica Guaymas"                                 "San Carlos (San Carlos Nuevo Guaymas)"          
# [145] "Condominios Playa Blanca"                        "Huatabampo"                                     
# [147] "Bacobampo"                                       "Empalme"                                        
# [149] "Esperanza"                                       "Villa Ju<e1>rez"                                
# [151] "Agua Blanca"                                     "Pueblo Yaqui"                                   
# [153] "Los Bahuises"                                    "V<ed>cam (Switch)"                              
# [155] "Las Bocas"                                       "Costa del Mar"                                  
# [157] "La Cruz"                                         "Campo el Salto"                                 
# [159] "Mazatl<e1>n"                                     "San Marcos"                                     
# [161] "Navolato"                                        "Los Mochis"                                     
# [163] "Culiac<e1>n Rosales"                             "Color Marino"                                   
# [165] "Dimas (Estaci<f3>n Dimas)"                       "Adolfo Ruiz Cortines"                           
# [167] "Leopoldo S<e1>nchez Celis"                       "Escuinapa de Hidalgo"                           
# [169] "San Ignacio"                                     "La Papalota"                                    
# [171] "Teacap<e1>n"                                     "El Rosario"                                     
# [173] "Mocorito"                                        "El Fuerte"                                      
# [175] "Topolobampo"                                     "Villa Uni<f3>n"                                 
# [177] "Altata"                                          "Guam<fa>chil"                                   
# [179] "Isla del Bosque"                                 "Guasave"                                        
# [181] "Juan Jos<e9> R<ed>os"                            "Quil<e1>"                                       
# [183] "Eldorado"                                        "El Saucito"                                     
# [185] "El Batall<f3>n"                                  "Pericos"                                        
# [187] "Celestino Gazca Villase<f1>or"                   "Las Glorias"                                    
# [189] "Costa Rica"                                      "Concordia"                                      
# [191] "Fraccionamiento Isla Para<ed>so"                 "Sistema de Producci<f3>n Intensiva del Noroeste"
# [193] "La Reforma"                                      "Gustavo D<ed>az Ordaz (El Carrizo)"             
# [195] "Licenciado Benito Ju<e1>rez (Campo Gobierno)"    "La Escondida"                                   
# [197] "Santa Eduwiges"                                  "Caimanero"                                      
# [199] "Callejones de Guasavito"                         "Agua Verde"                                     
# [201] "Le<f3>n Fonseca (Estaci<f3>n Verdura)"           "Nestle Culiac<e1>n"                             
# [203] "Localidad Econ<f3>mica"                          "Rancho Girolando"                               
# [205] "Ninguno [Motel]"                                 "El Patio"                                       
# [207] "Gabriel Leyva Solano (Benito Ju<e1>rez)"         "Ejido Plan Nacional Agrario"                    
# [209] "Alfonsinas"                                      "Tailwind"                                       
# [211] "Providencia"                                     "El Encuentro"                                   
# [213] "Zacualpan"                                       "Ruiz"                                           
# [215] "Desarrollo Puerto Bah<ed>a"                      "Litib<fa>"                                      
# [217] "Heroica Ciudad Concordia"                        "Productos Enlatados Madre Tierra"               
# [219] "Ferretera y Torniller<ed>a Zapata [FETOZA]"      "Walamito"                                       
# [221] "Bahía de los Ángeles"                            "Cataviña"                                       
# [223] "Punta Bufeo [Campo Turístico]"                   "Heroica Mulegé"                                 
# [225] "Santa Rosalía"                                   "San José del Cabo"                              
# [227] "Tripuí [Trailer Park]"                           "San Román"                                      
# [229] "Nopoló"                                          "San Sebastián del Oeste"                        
# [231] "San Nicolás (La Troje)"                          "Costa Vida Vallarta"                            
# [233] "Rincón de Guayabitos"                            "Bucerías"                                       
# [235] "Bahía de Matanchén"                              "La Peñita de Jaltemba"                          
# [237] "San José del Valle"                              "Dos Palmeras (Casa Pelícanos)"                  
# [239] "Pátzcuaro"                                       "Potrero las Lomas"                              
# [241] "Desarrollo Puerto Bahía"                         "Litibú"                                         
# [243] "Mexcaltitán de Uribe"                            "Lázaro Cárdenas [Colonia]"                      
# [245] "Los Tabachines"                                  "Puerto Peñasco"                                 
# [247] "Bahía de Kino"                                   "Miguel Alemán (La Doce)"                        
# [249] "Plutarco Elías Calles (La Y Griega)"             "Villa Juárez"                                   
# [251] "Ciudad Obregón"                                  "Vícam (Switch)"                                 
# [253] "Culiacán Rosales"                                "Mazatlán"                                       
# [255] "Guamúchil"                                       "Sistema de Producción Intensiva del Noroeste"   
# [257] "Campo Santa Fe"                                  "El Batallón"                                    
# [259] "Teacapán"                                        "La Caída"                                       
# [261] "Villa Unión"                                     "Dimas (Estación Dimas)"                         
# [263] "Quilá"                                           "Juan José Ríos"                                 
# [265] "Motel Popeye"                                    "Licenciado Benito Juárez (Campo Gobierno)"      
# [267] "Gustavo Díaz Ordaz (El Carrizo)"                 "Nestle Culiacán"                                
# [269] "León Fonseca (Estación Verdura)"                 "Motel Paradise"                                 
# [271] "Playa del Sol"                                   "Pap<e1> Fern<e1>ndez [Campo]"                   
# [273] "Los Vientos"                                     "Campo Lobo"                                     
# [275] "Cabo Golf Cars"                                  "Club Campestre el Madrigal"                     
# [277] "Cuesta Blanca"                                   "San Dionisio"                                   
# [279] "Las <c1>nimas de Quimixto"                       "Lienzo Charro"                                  
# [281] "Mezcalitos"                                      "Para<ed>so Escondido [Colonia]"                 
# [283] "Jalcocot<e1>n"                                   "Las Tijerillas"                                 
# [285] "Luis Pe<f1>a"                                    "Hip<f3>lito Gonz<e1>lez L<f3>pez"               
# [287] "Limoncito Hills"                                 "La Playa Estates"                               
# [289] "La Cantera"                                      "Haixa (Litib<fa>)"                              
# [291] "Playa Encanto"                                   "Puerto Lobos"                                   
# [293] "Mayan Lakes"                                     "Playa Huatabampito"                             
# [295] "Coch<f3>rit"                                     "Anglers Inn [Bill Chapman Jr.]"                 
# [297] "La Ca<ed>da"                                     "San Pedro"                                      
# [299] "El Bordo"                                        "TUMSA [Transportes]"                            
# [301] "El Diez"                                         "Cortijo la Piedra"  
```

```{r}
# aggregate the accommodation employee data by municipality and year
employees_by_muni_yr_50 <- alojamientos_est_perocu %>% 
  mutate(municipio = case_when(
    municipio == "Mexicali" ~ "San Felipe",
    TRUE ~ municipio
  )) %>% 
  group_by(municipio, estado, data_year) %>% 
  summarize(
    total_employees = sum(employees_estimated, na.rm = TRUE),
    num_businesses = n(),
    avg_employees_per_business = mean(employees_estimated, na.rm = TRUE),
    .groups = "drop"
  )
```


## Bring in Total Visitors

These shapefiles were written out from `~TR/tr_quantity.qmd`.

```{r}
## from 2018 - 2025
dom_goc <- st_read(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/hotel_monitoring_int","domestic_arrivals_sectur_goc.shp"))

int_goc <- st_read(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/hotel_monitoring_int","international_arrivals_sectur_goc.shp"))

## join back together to have both spatial attributes and the number of visitors
all_arrivals <- bind_rows(dom_goc, int_goc) 

total_arrivals_yr_muni <- all_arrivals %>% 
  group_by(year, municip, estado, rgn_id) %>% 
  summarize(arrivals = sum(value))

mapview(total_arrivals_yr_muni, zcol = "arrivals")
```


## Merge the datasets

First, clean them both up:

```{r}
## look for differences

unique(employees_by_muni_yr_50$municipio)
# [1] "Ahome"                    "Angostura"                "Bah<ed>a de Banderas"     "Bahía de Banderas"        "Benito Ju<e1>rez"         "Benito Juárez"            "Cabo Corrientes"          "Caborca"                 
#  [9] "Compostela"               "Culiac<e1>n"              "Culiacán"                 "Elota"                    "Empalme"                  "Ensenada"                 "Escuinapa"                "Guasave"                 
# [17] "Guaymas"                  "Hermosillo"               "Huatabampo"               "La Paz"                   "Loreto"                   "Los Cabos"                "Mazatl<e1>n"              "Mazatlán"                
# [25] "Muleg<e9>"                "Mulegé"                   "Navolato"                 "Pitiquito"                "Puerto Pe<f1>asco"        "Puerto Peñasco"           "Puerto Vallarta"          "Rosario"                 
# [33] "San Blas"                 "San Felipe"               "San Ignacio"              "San Luis R<ed>o Colorado" "San Luis Río Colorado"    "San Quint<ed>n"           "San Quintín"              "Tecuala"               



unique(total_arrivals_yr_muni$municip)
#  [1] "cabo_san_lucas"          "culiacan"                "el_fuerte"               "el_rosario"              "escuinapa"               "guaymas"                 "hermosillo"              "la_paz"                  "loreto"                 
# [10] "los_mochis"              "mazatlan"                "puerto_vallarta"         "san_felipe"              "san_jose_del_cabo"       "zona_corredor_los_cabos"
```
```{r}
## for 50km
employees_by_muni_yr_join_50 <- employees_by_muni_yr_50 %>%
  st_drop_geometry() %>% 
  mutate(
    municipio_clean = municipio %>%
      str_replace_all("Bah<ed>a", "Bahía") %>%
      str_replace_all("Ju<e1>rez", "Juárez") %>%
      str_replace_all("Culiac<e1>n", "Culiacán") %>%
      str_replace_all("Mazatl<e1>n", "Mazatlán") %>%
      str_replace_all("Muleg<e9>", "Mulegé") %>%
      str_replace_all("Pe<f1>asco", "Peñasco") %>%
      str_replace_all("R<ed>o", "Río") %>%
      str_replace_all("Ru<ed>z", "Ruíz") %>%
      str_replace_all("Quint<ed>n", "Quintín") %>%
      str_replace_all("Comond<fa>", "Comondú") %>%
      str_replace_all("San Sebasti<e1>n del Oeste", "San Sebastián del Oeste") %>%
      str_replace_all("Tomatl<e1>n", "Tomatlán") %>%
      str_replace_all("San Luis R<ed>o Colorado", "San Luis Río Colorado") %>%
      str_replace_all("<fa>", "ú") %>%
      str_replace_all("<e1>", "á") %>%
      str_replace_all("<ed>", "í") %>%
      str_replace_all("<f1>", "ñ") %>%
      str_replace_all("<e9>", "é") %>%
      str_to_lower() %>%
      str_replace_all(" ", "_"),
    
    estado = estado %>% 
      str_to_lower() %>%
      str_replace_all(" ", "_") %>% 
      ifelse(municipio_clean == "puerto_vallarta", "jalisco", .),
    
    data_year = as.numeric(data_year)
  ) %>% 
  dplyr::select(-municipio) %>% 
  complete(nesting(municipio_clean, estado), data_year) %>%  # for ensenada (missing 2024), san quintin (missing 2019, 2020) and san pedro laguilla (missing 2019 and 2020)
  mutate(
    total_employees = replace_na(total_employees, 0),
    num_businesses = replace_na(num_businesses, 0),
    avg_employees_per_business = replace_na(avg_employees_per_business, 0)
  ) ## if there are no economic units in the denue directory for a year, then we can assume there are no formal businesses. Will need to note this.
```


```{r}
## for 50km
employees_by_muni_yr_join_50_sf <- employees_by_muni_yr_50 %>%
  mutate(
    municipio_clean = municipio %>%
      str_replace_all("Bah<ed>a", "Bahía") %>%
      str_replace_all("Ju<e1>rez", "Juárez") %>%
      str_replace_all("Culiac<e1>n", "Culiacán") %>%
      str_replace_all("Mazatl<e1>n", "Mazatlán") %>%
      str_replace_all("Muleg<e9>", "Mulegé") %>%
      str_replace_all("Pe<f1>asco", "Peñasco") %>%
      str_replace_all("R<ed>o", "Río") %>%
      str_replace_all("Ru<ed>z", "Ruíz") %>%
      str_replace_all("Quint<ed>n", "Quintín") %>%
      str_replace_all("Comond<fa>", "Comondú") %>%
      str_replace_all("San Sebasti<e1>n del Oeste", "San Sebastián del Oeste") %>%
      str_replace_all("Tomatl<e1>n", "Tomatlán") %>%
      str_replace_all("San Luis R<ed>o Colorado", "San Luis Río Colorado") %>%
      str_replace_all("<fa>", "ú") %>%
      str_replace_all("<e1>", "á") %>%
      str_replace_all("<ed>", "í") %>%
      str_replace_all("<f1>", "ñ") %>%
      str_replace_all("<e9>", "é") %>%
      str_to_lower() %>%
      str_replace_all(" ", "_"),
    
    estado = estado %>% 
      str_to_lower() %>%
      str_replace_all(" ", "_") %>% 
      ifelse(municipio_clean == "puerto_vallarta", "jalisco", .),
    
    data_year = as.numeric(data_year)
  ) %>% 
  dplyr::select(municipio_clean, geometry) %>% 
  distinct()
```


```{r}
inland_50 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer","inland_50km_goc.shp"))
st_crs(inland_50) == st_crs(employees_by_muni_yr_join_50_sf) ## TRUE

denue_municipios_rgn_sf <- st_intersection(employees_by_muni_yr_join_50_sf, inland_50)

denue_municipios_rgn <- denue_municipios_rgn_sf %>% 
  st_drop_geometry() %>%
  dplyr::select(municipio_clean, rgn_id) %>% 
  distinct()

dups <- denue_municipios_rgn %>% 
  group_by(municipio_clean) %>% 
  summarise(dups = sum(n())) %>% 
  filter(dups > 1)
unique(dups$municipio_clean) ## [1] "comondú"        "puerto_peñasco" "san_felipe"  

### Okay, I am going to assign each one by a case_when to Regions that these municipalities typically are... for now.
denue_municipios_rgn_fix <- denue_municipios_rgn %>% 
  mutate(rgn_id = case_when(
    municipio_clean == "comondú" ~ "Region_3",
    municipio_clean == "puerto_peñasco" ~ "Region_5",
    municipio_clean == "san_felipe" ~ "Region_1",
    TRUE ~ rgn_id
  )) ## no more duplicates!!!!! Save this for later.
  # distinct() %>% 
  # group_by(municipio_clean) %>% 
  # summarise(dups = sum(n())) %>% 
  # filter(dups > 1)

write_csv(denue_municipios_rgn_fix, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/hotel_monitoring_int","ohi_rgn_assign_for_municipio_clean.csv"))
```

```{r}
### for 10km
## denue accommodation employees
# employees_by_muni_yr_join <- employees_by_muni_yr %>%
#   st_drop_geometry() %>% 
#   mutate(
#     municipio_clean = municipio %>%
#       str_replace_all("Bah<ed>a", "Bahía") %>%
#       str_replace_all("Ju<e1>rez", "Juárez") %>%
#       str_replace_all("Culiac<e1>n", "Culiacán") %>%
#       str_replace_all("Mazatl<e1>n", "Mazatlán") %>%
#       str_replace_all("Muleg<e9>", "Mulegé") %>%
#       str_replace_all("Pe<f1>asco", "Peñasco") %>%
#       str_replace_all("R<ed>o", "Río") %>%
#       str_replace_all("Quint<ed>n", "Quintín") %>%
#       
#       # make lower case and remove spaces to match the arrivals data
#       str_to_lower() %>%
#       str_replace_all(" ", "_"),
#     estado = estado %>% 
#       str_to_lower() %>%
#       str_replace_all(" ", "_"),
#     data_year = as.numeric(data_year)
#   )
# 
# unique(employees_by_muni_yr_join$municipio_clean)
```


```{r}
## arrivals data tweaks for join
total_arrivals_aggregated_join <- total_arrivals_yr_muni %>%
  st_drop_geometry() %>% 
  mutate(
    municipio_clean = case_when(
      municip %in% c("cabo_san_lucas", "san_jose_del_cabo", "zona_corredor_los_cabos") ~ "los_cabos",
      municip == "los_mochis" ~ "ahome",
      municip == "el_rosario" ~ "rosario",
      municip == "culiacan" ~ "culiacán",
      municip == "mazatlan" ~ "mazatlán",
      TRUE ~ municip
    ),
    estado = estado %>% 
      str_to_lower() %>%
      str_replace_all(" ", "_"),
    year = as.numeric(year)
  ) %>%
  # reaggregate with changes
  group_by(year, municipio_clean, estado, rgn_id) %>%
  summarize(arrivals = sum(arrivals), .groups = "drop")

unique(total_arrivals_aggregated_join$municipio_clean)
```

### See the differences now in municipios

```{r}
# what the arrivals data has but the employment data does not
setdiff(unique(total_arrivals_aggregated_join$municipio_clean), unique(employees_by_muni_yr_join_50$municipio_clean))
# none!!!

# what DENUE data has but hotel arrivals data does not
setdiff(unique(employees_by_muni_yr_join_50$municipio_clean), unique(total_arrivals_aggregated_join$municipio_clean))

#  [1] "acaponeta"               "angostura"               "bahía_de_banderas"       "benito_juárez"          
#  [5] "cabo_corrientes"         "caborca"                 "cajeme"                  "comondú"                
#  [9] "compostela"              "concordia"               "elota"                   "empalme"                
# [13] "ensenada"                "etchojoa"                "guasave"                 "huajicori"              
# [17] "huatabampo"              "mascota"                 "mocorito"                "mulegé"                 
# [21] "navojoa"                 "navolato"                "pitiquito"               "puerto_peñasco"         
# [25] "rosamorada"              "ruíz"                    "salvador_alvarado"       "san_blas"               
# [29] "san_ignacio"             "san_luis_río_colorado"   "san_pedro_lagunillas"    "san_quintín"            
# [33] "san_sebastián_del_oeste" "santiago_ixcuintla"      "talpa_de_allende"        "tecuala"                
# [37] "tepic"                   "tomatlán"                "tuxpan"                  "xalisco"            
```

### Join the arrivals and accommodation employees data

```{r}
## join the datasets
combined_arrivals_employees_data <- total_arrivals_aggregated_join %>%
  left_join(employees_by_muni_yr_join_50, by = c("municipio_clean" = "municipio_clean", "year" = "data_year", "estado" = "estado")) %>% 
  filter(year %in% 2019:2024)

# combined_arrivals_employees_data %>%
#   group_by(municipio_clean) %>%
#   summarize(
#     has_arrivals = sum(!is.na(arrivals)) > 0,
#     has_employees = sum(!is.na(total_employees)) > 0,
#     n_years = n()
#   ) %>%
#   print(n = Inf)
```


## Visitor and Accommodation Employees Relationship

### model1

```{r}
model1 <- lm(arrivals ~ total_employees, data = combined_arrivals_employees_data)
summary(model1)

library(ggplot2)

ggplot(combined_arrivals_employees_data, aes(x = total_employees, y = arrivals)) +
  geom_point() +
  # geom_abline(slope = 160.088, intercept = 0) +
  # geom_smooth(method = "lm") +
  theme_minimal() +
  labs(
    title = "Total arrivals (domestic + international) vs Total accommodation employees",
    x = "Total employees",
    y = "Total arrivals"
  )

```

### model1.5

```{r}
# model1.5 <- lm(arrivals ~ total_employees + as.factor(year), data = combined_arrivals_employees_data)
# summary(model1.5)

combined_arrivals_employees_data_wo2020 <- combined_arrivals_employees_data %>% 
  mutate(year_factor = ifelse(year %in% 2020, 1, 2))

## 2020 has more employees than we would expect based on arrivals (large dip in arrivals)

## we need to treat 2020 as something different

model1.5 <- lm(arrivals ~ total_employees + as.factor(year_factor), data = combined_arrivals_employees_data_wo2020)
summary(model1.5)
```


<!-- ### model2 -->

<!-- ```{r} -->
<!-- ## just to check -->
<!-- model2 <- lm(log(arrivals) ~ log(total_employees), data = combined_arrivals_employees_data) -->
<!-- summary(model2) -->
<!-- ``` -->

### model3

With municipio effects:
```{r}
model3 <- lm(arrivals ~ total_employees + municipio_clean, data = combined_arrivals_employees_data)
summary(model3)
```

### model3.5
Accounting for 2020:
```{r}
combined_arrivals_employees_data_wo2020 <- combined_arrivals_employees_data %>% 
  mutate(year_factor = ifelse(year %in% 2020, 1, 2))

model3.5 <- lm(arrivals ~ total_employees + municipio_clean + as.factor(year_factor), data = combined_arrivals_employees_data_wo2020)
summary(model3.5)

```

### model4

```{r}
model4 <- lm(arrivals ~ total_employees + estado, data = combined_arrivals_employees_data)
summary(model4)

```


```{r}
AIC(model1) # 1983.695 / 1993.01
AIC(model3) # 1901.315 / 1901.315
AIC(model4) # 1981.343 / 1981.343
AIC(model5) # 1979.68

BIC(model1) # 1990.354 / 1997.449
BIC(model3) # 1934.607 / 1934.607
BIC(model4) # 1996.879 / 1996.879 - overfitting
BIC(model5) # 1934.607

BIC(model3) # 1934.607
BIC(model3.5) # 1911.944 yay!

BIC(model1) # 1997.449
BIC(model1.5) # 1988.558 yay!
```



# Gapfilling tourist arrivals 

We will be using model3 for any tourist centers that have gaps in certain years, but at least have some data.  For municipalities that have DENUE hotel employee data but not any SECTUR arrivals data/tourist centers, we will use model1.

Use `predict()`.

```{r}
## we will be prepping the df `combined_arrivals_employees_data` 

combined_arrivals_employees_data_full <- full_join(employees_by_muni_yr_join_50, total_arrivals_aggregated_join, by = c("municipio_clean" = "municipio_clean", "data_year" = "year", "estado" = "estado")) 
```

```{r}
# filled_combined_arrivals_employees_data <- combined_arrivals_employees_data_full %>%
#   mutate(
#     arrivals_filled = ifelse(
#       is.na(arrivals) & !is.na(total_employees),
#       predict(model1, newdata = combined_arrivals_employees_data),
#       arrivals
#     )
#   )
```

## bptest

```{r}
plot(model1, which = 1) # residuals vs fitted
plot(model1, which = 2) # QQ plot
plot(model1, which = 3) # scale - location

library(lmtest)
bptest(model1) # Breusch Pagan test. BP = 19.161, df = 1, p-value = 1.201e-05, heteroscedastic
bptest(model3) # when employees + municipio: BP = 21.914, df = 13, p-value = 0.05672, just over 0.05... homoscedastic or hetero
bptest(model2) # when log(): BP = 13.647, df = 1, p-value = 0.0002206... hetero
bptest(model4) # employees + estado: BP = 22.289, df = 5, p-value = 0.0004613. .... hetero
```

## prep for fill

```{r}
# use predict() to fill arrivals by employees
## make a vector of all municipios that can be filled with model3
munis_model3 <- unique(combined_arrivals_employees_data$municipio_clean)

df_model3 <- combined_arrivals_employees_data_full %>%
  filter(municipio_clean %in% munis_model3)

df_model1 <- combined_arrivals_employees_data_full %>%
  filter(!municipio_clean %in% munis_model3)
```


## use predict() to fill

### model3
```{r}
## fill using model3
df_model3_filled <- df_model3 %>%
  filter(data_year %in% 2019:2024) %>% 
  mutate(
    arrivals_filled_model3 = ifelse(
      is.na(arrivals) & !is.na(total_employees), # if there are no arrivals data but there are employees for that municipality, predict using model3 !!
      predict(model3, newdata = df_model3),
      arrivals
    )
  )
```

### model1
```{r}
## fill using model 1
df_model1_filled <- df_model1 %>%
  filter(data_year %in% 2019:2024) %>% 
  mutate(
    arrivals_filled_model1 = ifelse(
      is.na(arrivals) & !is.na(total_employees),  # if, of the municipalities left, there are no arrivals data but there are employees for that municipality, predict using model1 !!
      predict(model1, newdata = df_model1),
      arrivals
    )
  )
```

```{r}
## combine the two datasets
filled_combined <- bind_rows(
  df_model3_filled,
  df_model1_filled
) %>% 
  mutate(is_filled = ifelse(is.na(arrivals), "1", "0")) # so we can see what has been gapfilled
```


```{r}
## now, lets make the column of arrivals contained of all that have been filled, and the OG data.
arrivals_employees_filled <- filled_combined %>% 
  mutate(arrivals_merged = case_when(
    is.na(arrivals) & !is.na(arrivals_filled_model3) ~ arrivals_filled_model3,
    is.na(arrivals) & is.na(arrivals_filled_model3) & !is.na(arrivals_filled_model1) ~ arrivals_filled_model1,
    TRUE ~ arrivals
  ))
```


## account for 2020

```{r}
# use predict() to fill arrivals by employees
# both need year_factor so the .5 models can use it

munis_model3 <- unique(combined_arrivals_employees_data$municipio_clean)

df_model3 <- combined_arrivals_employees_data_full %>%
  filter(municipio_clean %in% munis_model3) %>%
  mutate(year_factor = ifelse(data_year == 2020, 1, 2))

df_model1 <- combined_arrivals_employees_data_full %>%
  filter(!municipio_clean %in% munis_model3) %>%
  mutate(year_factor = ifelse(data_year == 2020, 1, 2))
```


### model3.5

```{r}
df_model3_filled <- df_model3 %>%
  filter(data_year %in% 2019:2024) %>%
  mutate(
    arrivals_filled_model3 = ifelse(
      is.na(arrivals) & !is.na(total_employees),
      predict(model3, newdata = df_model3 %>% filter(data_year %in% 2019:2024)),
      arrivals
    ),
    arrivals_filled_model3.5 = ifelse(
      is.na(arrivals) & !is.na(total_employees),
      predict(model3.5, newdata = df_model3 %>% 
                filter(data_year %in% 2019:2024) %>%
                mutate(year_factor = ifelse(data_year == 2020, 1, 2))),
      arrivals
    )
  )
```


### model1.5

```{r}
df_model1_filled <- df_model1 %>%
  filter(data_year %in% 2019:2024) %>%
  mutate(
    arrivals_filled_model1 = ifelse(
      is.na(arrivals) & !is.na(total_employees),
      predict(model1, newdata = df_model1 %>% filter(data_year %in% 2019:2024)),
      arrivals
    ),
    arrivals_filled_model1.5 = ifelse(
      is.na(arrivals) & !is.na(total_employees),
      predict(model1.5, newdata = df_model1 %>% 
                filter(data_year %in% 2019:2024) %>%
                mutate(year_factor = ifelse(data_year == 2020, 1, 2))),
      arrivals
    )
  )
```

### combine&merge

```{r}
filled_combined <- bind_rows(df_model3_filled, df_model1_filled) %>%
  mutate(is_filled = ifelse(is.na(arrivals), "1", "0"))

arrivals_employees_filled <- filled_combined %>%
  mutate(arrivals_merged = case_when(
    is.na(arrivals) & !is.na(arrivals_filled_model3.5) ~ arrivals_filled_model3.5,
    is.na(arrivals) & is.na(arrivals_filled_model3.5) & !is.na(arrivals_filled_model1.5) ~ arrivals_filled_model1.5,
    TRUE ~ arrivals
  ))
```

### censor at zero

For small municipalities, we now see that the predicted arrivals dip below zero.  However, that would obviously never happen.  

So, we will make any negative arrivals equal to zero, as there cannot be any negative arrivals conceptually.

Because it would not make sense to log transform this data, nor to constrain the model itself with a log link (poisson/GLM which would assume mean = variance which is not true here, see `bptest`....), we will simply restrain the values after predicting.  

This is fine for gapfilling.

```{r}
arrivals_employees_filled_negfix <- arrivals_employees_filled %>% 
  mutate(arrivals_merged_no_neg = ifelse(arrivals_merged < 0, 0, arrivals_merged))
```


## plotlys

```{r}
library(plotly)

plot_data <- arrivals_employees_filled %>% 
  arrange(data_year)

plot_data_plotly <- plot_ly(
  data = plot_data,
  x = ~data_year,
  y = ~arrivals_merged,
  color = ~municipio_clean,
  type = 'scatter',
  mode = 'lines+markers'
) %>%
  layout(
    title = "Total arrivals (international + domestic) by municipio after gapfilling with model3.5 and model1.5",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Arrivals")
  )
plot_data_plotly
```


```{r}
#### after fixing negative predicted employees
plot_data_noneg <- arrivals_employees_filled_negfix %>% 
  arrange(data_year)

plot_data_noneg_plotly <- plot_ly(
  data = plot_data_noneg,
  x = ~data_year,
  y = ~arrivals_merged_no_neg,
  color = ~municipio_clean,
  type = 'scatter',
  mode = 'lines+markers'
) %>%
  layout(
    title = "Total arrivals (international + domestic) by municipio after gapfilling with model3.5 and model1.5 + making negative arrivals zero",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Arrivals")
  )
plot_data_noneg_plotly

```



#### Check it diagnostically:

```{r}
## first, check that there are no negative predictions
arrivals_employees_filled %>%
  filter(arrivals_merged < 0) ## none!!
```


```{r}
# ggplot(df_model3_filled, aes(x = arrivals, y = arrivals_filled_model3)) +
#   geom_point(alpha = 0.6) +
#   geom_abline(slope = 1, intercept = 0, color = "red") +
#   theme_minimal() +
#   labs(
#     title = "Model3 Diagnostic: Actual vs Predicted Arrivals",
#     x = "Actual arrivals",
#     y = "Predicted arrivals (model3)"
#   )
```

### hist model3.5 filled
```{r}
hist(arrivals_employees_filled_negfix$arrivals_filled_model3.5, breaks = 30,
     main = "Distribution of filled arrivals",
     xlab = "arrivals")
```

### hist model1.5 filled

```{r}
hist(arrivals_employees_filled$arrivals_filled_model1.5, breaks = 30,
     main = "Distribution of filled arrivals",
     xlab = "arrivals")
```


```{r}
df_compare <- arrivals_employees_filled %>%
  mutate(source = ifelse(is.na(arrivals), "predicted", "actual"))

ggplot(df_compare, aes(x = arrivals_merged, fill = source)) +
  geom_histogram(alpha = 0.5, position = "identity") +
  theme_minimal()
```

```{r}
# ggplot(arrivals_employees_filled,
#        aes(x = data_year, y = arrivals_merged, color = is_filled)) +
#   geom_line() +
#   facet_wrap(~ municipio_clean, scales = "free_y") +
#   theme_minimal() +
#   labs(title = "Per-Municipio Time Series (Actual vs Filled)")

# df_model3_filled %>%
#   filter(!is.na(arrivals)) %>%
#   summarize(
#     RMSE = sqrt(mean((arrivals - predict(model3, newdata = .))^2)), # a lower RMSE indicates a better fit and more accurate predictions
#     MAE  = mean(abs(arrivals - predict(model3, newdata = .)))
#   )
# 
# df_model1_filled %>%
#   filter(!is.na(arrivals)) %>%
#   summarize(
#     RMSE = sqrt(mean((arrivals - predict(model1, newdata = .))^2)),
#     MAE  = mean(abs(arrivals - predict(model1, newdata = .)))
#   )
```

### arrivals vs emp

```{r}
ggplot(arrivals_employees_filled,
       aes(x = total_employees, y = arrivals_merged, color = is_filled)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_minimal()
```

## Write out

```{r}
write_csv(arrivals_employees_filled, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/gapfilled_int","arrivals_filled_by_DENUE_accommodation_employees_2019_2024.csv"))

write_csv(arrivals_employees_filled_negfix, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/gapfilled_int","arrivals_filled_by_DENUE_accommodation_employees_2019_2024_UPDATED_031026.csv"))

# arrivals_employees_filled <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/gapfilled_int","arrivals_filled_by_DENUE_accommodation_employees_2019_2024.csv"))
```
