---
title: "Tourist data"
output: html_document
date: "2025-3-6"
editor_options: 
  chunk_output_type: console
---
Data: "Ocupación en hoteles y moteles por centro turístico, corte mensual DataTur"
Go to: <https://monitoreohotelero.sectur.gob.mx/Reportes/Menu.aspx>

To download:

> 1. Go to the “Generate Report” section.
> 2. Select the “Mensual”, or monthly, time frame.
> 3. Choose the option “Hotel and motel occupancy by tourist center, monthly DataTur snapshot.”
> 4. Click “Continue” and then “Advanced criteria” (Criterios Avanzados)
> 5. Select the municipality of interest, then all the variables of interest.  Specify "Serie" for "Cobertura Temporal corte mensual", starting from 2018 January (Enero) to 2025 Junio (June).
> 6. Finally, click on “Generate Report.”

The report can be downloaded in various formats and contains the information available as of the date of consultation.  It is important to note that this information is presented at the tourist center level, not the municipal level. However, it can be useful for territorial approximations in the Gulf of California region, particularly in key tourist destinations.

```{r}
library(mapview)
library(sf)
library(tidyverse)
library(here)
library(terra)
```


## Clean data

```{r}
# read in one municipio first. start with San Felipe
san_felipe_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles","DataturReportePublico_San_Felipe.csv")) # hmm.. Error in vroom_(file, delim = delim %||% col_types$delim, col_names = col_names,  :                 
  # embedded nul in string: 'O\0c\0u\0p\0a\0c\0i\0\xf3\0n\0 \0e\0n\0 \0h\0o\0t\0e\0l\0e\0s\0 \0y\0 \0m\0o\0t\0e\0l\0e\0s\0 \0p\0o\0r\0 \0c\0e\0n\0t\0r\0o\0 \0t\0u\0r\0\xed\0s\0t\0i\0c\0o\0,\0 \0c\0o\0r\0t\0e\0 \0m\0e\0n\0s\0u\0a\0l\0.\0 \0D\0a\0t\0a\0T\0u\0r'

library(readr)
library(data.table)

# It looks like the file is encoded in UTF-16 (or UCS-2) instead of the usual UTF-8 or Latin1, so you need to use `data.table::fread()`
san_felipe_raw <- read_lines("/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/DataturReportePublico_San_Felipe.csv", 
                    locale = locale(encoding = "UTF-16LE"))

# paste lines back together into one string for read_csv
san_felipe_text <- paste(san_felipe_raw, collapse = "\n")

# read into a dataframe now
san_felipe_csv_raw <- read_csv(I(san_felipe_text)) %>% 
  janitor::clean_names()

### did not work... try downloading the report as an excel file instead... could do some cleaning from now on but it might be easier to download as an excel.  Check!!
```
Try the excel file download instead...

```{r}
library(readxl)

# read in San Felipe .xls
san_felipe_raw <- readxl::read_excel(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/antiquated_resaves","DataturReportePublico_San_Felipe.xls"))

tools::file_ext(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/antiquated_resaves/DataturReportePublico_San_Felipe.xls")) # [1] "xls"
readLines(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/antiquated_resaves/DataturReportePublico_San_Felipe.xls"))

# .... [2] "            <table id=\"tablaReportePublico\" class=\"overflow-y\">"
# likely has html code which is why it looks like this. Try resaving as .xlsx (YES, when I tried opening it in excel and resaving it, it was currently a .htm)

# read in San Felipe .xlsx
san_felipe_raw <- readxl::read_xlsx(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/antiquated_resaves","DataturReportePublico_San_Felipe.xlsx"), sheet = 1) # must still have .htm weirdness... Error in read_fun(path = path, sheet_i = sheet, limits = limits, shim = shim,  : invalid Unicode point 56564
```

Try re-saving the raw file as a .CSV again...
```{r}
# for San Felipe still
san_felipe_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/resaved_as_csv","DataturReportePublico_San_Felipe_resaved.csv")) # it works!!!

san_felipe_clean <- san_felipe_raw %>% 
  set_names(.[9, ]) %>%  # make the column names what they should be
  slice(-c(1:10)) %>% 
  janitor::clean_names()

# clean further by trimming/replacing white spaces in the concept headers, as well as making it lower case (for my aesthetics haha)
san_felipe_clean_concept <- san_felipe_clean %>%
  mutate(
    concept = str_to_lower(concepto),
    concept = str_trim(concept),
    concept = str_replace_all(concept, " ", "_")
  ) %>% 
  relocate(concept, .after = concepto) %>% 
  select(-concepto)

# identify which rows are the topic headers, and then give each row a specific name for easier pivoting.
san_felipe_clean_by_topic <- san_felipe_clean_concept %>% 
  mutate(
    is_topic = !str_detect(concept, "^[0-9]_estrellas|[1-5]_estrellas|total_categorias") & !is.na(concept) # identify, true or false, if it is not the subheader (aka it IS a topic header)
  ) %>%
  # fill down header concepts
  mutate(header_concept = if_else(is_topic, concept, NA)) %>% # or NA_character if there are problems.
  fill(header_concept) %>% 
  # make the combined concept by topic: if it’s a subcategory, combine, if not, leave as original
  mutate(
    concept_full = if_else(concept != header_concept, paste0(header_concept, ":", concept), concept
    )
  )

# now, pivot longer, where we have the columns as the header_concept!
san_felipe_long <- san_felipe_clean_by_topic %>% 
  pivot_longer(
    cols = 2:100,
    names_to = "month_year",
    values_to = "value"
  ) # check with original data to ensure it makes sense!

# great.  now let's make a column to indicate the year, making the final San Felipe dataset
san_felipe_long_final <- san_felipe_long %>% 
  mutate(
    is_mo_yr = !str_detect(month_year, "acum|total"),
    temp = if_else(is_mo_yr, month_year, NA)
  ) %>% 
  separate(temp, into = c("month", "year"), sep = "_", remove = FALSE) %>% 
  filter(is_topic %in% FALSE) %>% # we do not want the rows of just titles
  select(header_concept, concept, concept_full, month_year, month, year, value) %>% 
  mutate(municipio = "san_felipe") # before rbinding with other cities!
```

### Function to clean the files based on this file

```{r}
clean_hotel_occupations_data <- function(municipio) {
  
  # where to find the file
  file_path <- here::here(
    "/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/resaved_as_csv",
    paste0("DataturReportePublico_", municipio, "_resaved.csv")
  )
  
  # read raw data
  # raw_data <- read_csv(file_path)
  raw_data <- read_csv(file_path, locale = readr::locale(encoding = "UTF-8")) # changed because having error: "Error in `map()`: ℹ In index: 1. Caused by error in `nchar()`: ! invalid multibyte string, element 1"
  
  # clean the column names, remove top rows, and tidy
  cleaned <- raw_data %>% 
    set_names(.[9, ]) %>%  
    slice(-c(1:10)) %>% # NOTE: not all are the same length, so some of them will have the bottom text (ex: ND No disponible) as part of the output, and there will have to be extra cleaning done at the end.
    janitor::clean_names()
  
  # clean and standardize the "concepto" column
  cleaned_concept <- cleaned %>%
    mutate(
      concept = str_to_lower(concepto),
      concept = str_trim(concept),
      concept = str_replace_all(concept, " ", "_")
    ) %>% 
    relocate(concept, .after = concepto) %>% # these two lines of code are not really necessary because the func will do it all at once
    select(-concepto)
  
  # identify topic headers and combine them with subcategories
  by_topic <- cleaned_concept %>% 
    mutate(
      is_topic = !str_detect(concept, "^[1-5]_estrella[s]?|total_categorias") & !is.na(concept)
    ) %>%
    mutate(header_concept = if_else(is_topic, concept, NA_character_)) %>% # use NA_character_ for this stuff just in case
    fill(header_concept) %>% 
    mutate(
      concept_full = if_else(concept != header_concept, paste0(header_concept, ":", concept), concept)
    )
  
  # pivot longer
  long <- by_topic %>% 
    pivot_longer(
      cols = 2:100, # adjust if more columns
      names_to = "month_year",
      values_to = "value"
    )
  
  # separate year and month, filter, and tag city
  final <- long %>% 
    mutate(
      is_mo_yr = !str_detect(month_year, "acum|total"),
      temp = if_else(is_mo_yr, month_year, NA_character_)
    ) %>% 
    separate(temp, into = c("month", "year"), sep = "_", remove = FALSE) %>% 
    filter(is_topic == FALSE) %>% 
    select(header_concept, concept, concept_full, month_year, month, year, value) %>% 
    mutate(municipio = str_to_lower(municipio))
  
  return(final)
}
```

Okay! Let's try it with Cabo San Lucas.

```{r}
cabo_san_lucas_final <- clean_hotel_occupations_data("Cabo_San_Lucas") # hmmm... there are weird crosses when resaved as UTF... "†".. but it only works as a UTF-8!

la_paz_final <- clean_hotel_occupations_data("La_Paz") # okay it is happening more than once.  Need to figure that out.  Come back tomorrow!
```

Okay!! So it looks like you CANNOT save the file as a .CSV first, instead you **must** to do it in this order:

1.  On the SECTUR report site, after generating the report, click "Exportar a Excel".  

2.  Open the ".xls" (which is really a .htm) in Excel, and then "Save As..." a UTF-8 .CSV. Rename it `DataturReportePublico_municipio_resaved.csv` where you replace municipio with the municipality you just saved. 

3.  Bring into mazu in `/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/resaved_as_csv` to be read in here.  This is the only way that you do not get any "†" in whitespaces.

**These instructions will be brought to the top of the script when it is cleaned up.**

```{r}
cabo_san_lucas_final <- clean_hotel_occupations_data("Cabo_San_Lucas") # YAY, now that I followed the above procedure, it looks great!!
```

#### Apply function to all municipios

```{r}
all_cities <- c(
  "Cabo_San_Lucas", "Culiacan", "El_Fuerte", "El_Rosario", "Escuinapa",
  "Guaymas", "Hermosillo", "La_Paz", "Loreto", "Los_Mochis",
  "Mazatlan", "Puerto_Vallarta", "San_Felipe", "San_Jose_del_Cabo", 
  "Zona_Corredor_Los_Cabos"
)

library(purrr)

all_data <- all_cities %>% 
  map(clean_hotel_occupations_data) %>% 
  bind_rows() # looks good! keep checking after lunch.

all_data_clean <- all_data %>% 
  rename( 
    metric = header_concept,
    category = concept
  ) %>%  # for better clarity, since we are done cleaning and these names make more sense now
  mutate(
    concept_full = str_replace_all(concept_full, "\U{0eef2f}th>|\U{0ee832}|\U{0ee82e}", "_"),
    # metric_label = str_replace_all(metric, "_", " "),
    # metric_label = str_to_title(metric_label), # for later viz 
    metric = str_replace_all(metric, "\U{0eef2f}th>|\U{0ee832}|\U{0ee82e}", "_"),
    category = str_replace_all(category, "\U{0eef2f}th>|\U{0ee832}|\U{0ee82e}", "_"),
    value = str_replace_all(value, ",",""),
    year = str_replace_all(year, "p",""), # only in 2025, assume it means "preliminary"
    month = case_when(
      tolower(month) == "enero" ~ "01",
      tolower(month) == "febrero" ~ "02",
      tolower(month) == "marzo" ~ "03",
      tolower(month) == "abril" ~ "04",
      tolower(month) == "mayo" ~ "05",
      tolower(month) == "junio" ~ "06",
      tolower(month) == "julio" ~ "07",
      tolower(month) == "agosto" ~ "08",
      tolower(month) == "septiembre" ~ "09",
      tolower(month) == "octubre" ~ "10",
      tolower(month) == "noviembre" ~ "11",
      tolower(month) == "diciembre" ~ "12",
      TRUE ~ NA_character_  # otherwise
    ) # for lubridate later
  ) %>% 
  filter(!category %in% NA) 

unique(all_data_clean$concept_full) # check that there are no weirdnesses left
```

#### Check data completeness for all municipios

```{r}
# check the number of municipalities, there should be 15
unique(all_data$municipio) # 15

# check the unique categories
unique(all_data$concept_full) # there should be 10593 observations for each municipio, but there are 132,561 total observations, not 155,925. Why would there be less for some? A: because not all of them have 1-5 star establishment subsets. This could pose a difficulty for using this data as a way to measure "quality" of tourism, depending on the methodology of this data.  If it is representative of all the hotels in the municipio, then it could be used.  BUT, if there is just a lack of data on the "star rating" of hotels in that area, or if it is not a representative sample, then it should not be used.

# group by municipio, and see the number of unique `concpet_full` for each municipio:
concept_counts <- all_data %>%
  group_by(municipio) %>%
  summarise(unique_concepts = n_distinct(concept_full)) ## hm... not consistent. But it should be in theory...
```

Okay, so upon further look, it appears that some municipios do not have information about hotels that are not 5 stars.  For example, in Zona Corredor Los Cabos, there is only a total category and a 5 star subset, but they are always equal, presumably because there are **only** 5 star hotels in the municipio that are monitored.

Loreto does not have any 2 star subcategories, only 1, 3, 4, 5.  

Cabo San Lucas does not have any 1 star establishments as a subset for observations. Nor does San Felipe or Hermosillo.

Guaymas only has 3, 4, 5 star establishments monitored.


```{r}
# count the number of observations per municipio
observations_per_municipio <- all_data %>%
  group_by(municipio) %>%
  summarise(n_obs = n()) # guaymas, loreto, and zona corredor have much less observations than the others.
```


## Exploration of data

From this data, let us see what is available for analysis, and what each type of visitation means.

This data source translates to: "Hotel and motel occupancy by tourist center, monthly breakdown. DataTur
Provides up to 126 variables of information with final results presented no later than the 20th of each month. Consult metadata to identify availability of information by tourist destination." 

So, `concept_full_clean` contains values of the number of visitors, with monthly data for each municipio, and each variable is a combination of:

* Metrics: eg, cuartos_disponibles, cuartos_ocupados, llegadas_de_turistas, porcentaje_de_ocupación, etc.

* Type of visitor: overall, residentes, no_residentes

* Hotel category: 5_estrellas, 4_estrellas, etc, or total_categorias


Explore the metrics available:
```{r}
# check which metrics are available for analysis
count_metric <- all_data_clean %>%
  count(metric, sort = TRUE) # hm, `densidad_de_ocupaci_o_residentes` has many more observations than the others, (10890 vs 7920 for all the others)

# oh no... a category is within the metric (1 estrella).... [FIXED]
# all_data_1_est <- all_data_clean %>% 
#   filter(metric == "1_estrella") # AH, this is because of the missing "s" at the end of estrella, since there is only one star.  Will alter the cleaning function.

all_data_densidad_de_ocupaci_o_residentes <- all_data_clean %>% 
  filter(metric %in% "densidad_de_ocupaci_o_residentes")
unique(all_data_densidad_de_ocupaci_o_residentes$category) # 7: has NA as a category

all_data_densidad_de_ocupaci_esidentes <- all_data_clean %>% 
  filter(metric %in% "densidad_de_ocupaci_esidentes")
unique(all_data_densidad_de_ocupaci_esidentes$category) # 6

all_data_NA_category <- all_data_clean %>% 
  filter(category %in% NA)
length(unique(all_data_NA_category$municipio)) # yup, for all 15. Without any values associated. Let's get rid of this when creating `all_data_clean` on line 229.
# that was fixed, now they all have 7920 observations for each metric.  YAY!
```


Explore the categories available (aka star levels) each metric has:
```{r}
# okay, now lets explore categories once more
count_cat <- all_data_clean %>%
  group_by(metric) %>%
  summarise(n_categories = n_distinct(category)) %>%
  arrange(desc(n_categories)) # 6 for each
```

### Vizualizing data availability

```{r}
library(ggplot2)

# make a heatmap of the data availability for each municipio.
all_data_clean %>%
  count(municipio, metric) %>%
  ggplot(aes(x = metric, y = municipio, fill = n)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title = "Number of Observations per Municipio and Metric", x = "Metric", y = "Municipio")
```

It makes sense that Zona Corredor Los Cabos has much less data associated with it, as it is not a full municipality, but instead just a stretch of high tourism activity in an area of Los Cabos.

```{r}
library(plotly)

all_data_llegadas_de_turistas <- all_data_clean %>% 
  filter(metric %in% "llegadas_de_turistas")

plot_data <- all_data_clean %>%
  filter(metric %in% "llegadas_de_turistas", category %in% "total_categorias") %>% 
  select(-c(month, year)) %>% 
  mutate(value = as.numeric(value),
         month_year_total = str_detect(month_year, "total")) %>% 
  filter(month_year_total == "TRUE") %>% 
  separate(month_year, into = c("type", "year"), sep = "_anual_", remove = FALSE)


# preliminary plot 
line_plot <- plotly::plot_ly(plot_data, x = ~year, y = ~value, color = ~municipio, type = "scatter", mode = "lines") %>% 
  layout(title = "Llegadas de Turistas (total per year for all hotels) por Municipio", 
         xaxis = list(title = "Year"),
         yaxis = list(title = "Value"))

line_plot

# htmlwidgets::saveWidget(line_plot, file = "llegadas_de_turistas.html")
```

Oh wow, okay this brings to my attention that we are missing data after 2022 for El Rosario and Escuinapa in Sinaloa, and El Fuerte in Sinaloa after 2023.  The rest have an appropriate dip in 2020 (keep in mind these are annual totals of arrivals).  This is only looking at "Llegadas de Turistas", when there are also: 

| Metric Variable                         | English Translation                           |
|-----------------------------------------|-----------------------------------------------|
| cuartos_registrados_fin_per_o           | Rooms Registered at End of Period             |
| cuartos_disponibles                     | Available Rooms                               |
| cuartos_disponibles_promedio            | Average Available Rooms                       |
| cuartos_ocupados                        | Occupied Rooms                                |
| cuartos_ocupados_residentes             | Occupied Rooms (Residents)                    |
| cuartos_ocupados_no_residentes          | Occupied Rooms (Non-Residents)                |
| llegadas_de_turistas                    | Tourist Arrivals                              |
| llegadas_de_turistas_residentes         | Tourist Arrivals (Residents)                  |
| llegadas_de_turistas_no_residentes      | Tourist Arrivals (Non-Residents)              |
| turistas_noche                          | Tourist-Nights (Person-Nights)                |
| turistas_noche_residentes               | Tourist-Nights (Residents)                    |
| turistas_noche_no_residentes            | Tourist-Nights (Non-Residents)                |
| porcentaje_de_ocupaci_                  | Occupancy Rate                                |
| porcentaje_de_ocupaci_esidentes         | Occupancy Rate (Residents)                    |
| porcentaje_de_ocupaci_o_residentes      | Occupancy Rate (Non-Residents)                |
| estad_promedio                          | Average Length of Stay                        |
| estad_promedio_residentes               | Average Length of Stay (Residents)            |
| estad_promedio_no_residentes            | Average Length of Stay (Non-Residents)        |
| densidad_de_ocupaci_                    | Occupancy Density                             |
| densidad_de_ocupaci_esidentes           | Occupancy Density (Residents)                 |
| densidad_de_ocupaci_o_residentes        | Occupancy Density (Non-Residents)             |

Let's look only at "total_categorias" within categories:

```{r}
# total category: the total of all possible hotels (1-5 stars, if available)
all_data_total_categorias <- all_data_clean %>% 
  filter(category == "total_categorias") %>%  # all 15 municipios present
  filter(!grepl("total_anual_", month_year, ignore.case = TRUE),
         !grepl("acumulado_", month_year, ignore.case = TRUE))

# check on guaymas
all_data_total_categorias_guay <- all_data_clean %>%
  filter(category == "total_categorias") %>% 
  mutate(has_data = !is.na(value)) %>% 
  filter(municipio %in% "guaymas")

# list of metric variables
metric_vars <- c(
  "cuartos_registrados_fin_per_o",
  "cuartos_disponibles",
  "cuartos_disponibles_promedio",
  "cuartos_ocupados",
  "cuartos_ocupados_residentes",
  "cuartos_ocupados_no_residentes",
  "llegadas_de_turistas",
  "llegadas_de_turistas_residentes",
  "llegadas_de_turistas_no_residentes",
  "turistas_noche",
  "turistas_noche_residentes",
  "turistas_noche_no_residentes",
  "porcentaje_de_ocupaci_",
  "porcentaje_de_ocupaci_esidentes",
  "porcentaje_de_ocupaci_o_residentes",
  "estad_promedio",
  "estad_promedio_residentes",
  "estad_promedio_no_residentes",
  "densidad_de_ocupaci_",
  "densidad_de_ocupaci_esidentes",
  "densidad_de_ocupaci_o_residentes"
)

# func to make a heatmap for each metric variable
plot_metric_heatmap <- function(metric_var) {
  all_data_total_categorias %>%
    filter(metric == metric_var) %>%
    mutate(has_data = !is.na(value)) %>% 
    ggplot(aes(x = year, y = municipio, fill = has_data)) +
    geom_tile(color = "white") +
    scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "grey90"),
                      name = "Data Available") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    labs(
      title = paste("Data Completeness:", metric_var),
      x = "Year",
      y = "Municipio"
    )
}

heatmap_list <- map(metric_vars, plot_metric_heatmap)

# view them.. there are 21 variables
heatmap_list[[1]]

# save them all to files:
# walk2(heatmap_list, metric_vars, ~ ggsave(
#   filename = paste0("heatmap_sectur_completeness_", .y, ".png"),
#   plot = .x, width = 8, height = 5
# ))

```




