---
title: "Tourist data"
output: html_document
date: "2025-3-6"
editor_options: 
  chunk_output_type: console
---
Data: "Ocupación en hoteles y moteles por centro turístico, corte mensual DataTur"
Go to: <https://monitoreohotelero.sectur.gob.mx/Reportes/Menu.aspx>

To download:

> 1. Go to the “Generate Report” section.
> 2. Select the “Mensual”, or monthly, time frame.
> 3. Choose the option “Hotel and motel occupancy by tourist center, monthly DataTur snapshot.”
> 4. Click “Continue” and then “Advanced criteria” (Criterios Avanzados)
> 5. Select the municipality of interest, then all the variables of interest.  Specify "Serie" for "Cobertura Temporal corte mensual", starting from 2018 January (Enero) to 2025 Junio (June).
> 6. Finally, click on “Generate Report.”

The report can be downloaded in various formats and contains the information available as of the date of consultation.  It is important to note that this information is presented at the tourist center level, not the municipal level. However, it can be useful for territorial approximations in the Gulf of California region, particularly in key tourist destinations.

```{r}
library(mapview)
library(sf)
library(tidyverse)
library(here)
library(terra)
```


## Clean data

```{r}
# read in one municipio first. start with San Felipe
san_felipe_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles","DataturReportePublico_San_Felipe.csv")) # hmm.. Error in vroom_(file, delim = delim %||% col_types$delim, col_names = col_names,  :                 
  # embedded nul in string: 'O\0c\0u\0p\0a\0c\0i\0\xf3\0n\0 \0e\0n\0 \0h\0o\0t\0e\0l\0e\0s\0 \0y\0 \0m\0o\0t\0e\0l\0e\0s\0 \0p\0o\0r\0 \0c\0e\0n\0t\0r\0o\0 \0t\0u\0r\0\xed\0s\0t\0i\0c\0o\0,\0 \0c\0o\0r\0t\0e\0 \0m\0e\0n\0s\0u\0a\0l\0.\0 \0D\0a\0t\0a\0T\0u\0r'

library(readr)
library(data.table)

# It looks like the file is encoded in UTF-16 (or UCS-2) instead of the usual UTF-8 or Latin1, so you need to use `data.table::fread()`
san_felipe_raw <- read_lines("/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/DataturReportePublico_San_Felipe.csv", 
                    locale = locale(encoding = "UTF-16LE"))

# paste lines back together into one string for read_csv
san_felipe_text <- paste(san_felipe_raw, collapse = "\n")

# read into a dataframe now
san_felipe_csv_raw <- read_csv(I(san_felipe_text)) %>% 
  janitor::clean_names()

### did not work... try downloading the report as an excel file instead... could do some cleaning from now on but it might be easier to download as an excel.  Check!!
```
Try the excel file download instead...

```{r}
library(readxl)

# read in San Felipe .xls
san_felipe_raw <- readxl::read_excel(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/antiquated_resaves","DataturReportePublico_San_Felipe.xls"))

tools::file_ext(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/antiquated_resaves/DataturReportePublico_San_Felipe.xls")) # [1] "xls"
readLines(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/antiquated_resaves/DataturReportePublico_San_Felipe.xls"))

# .... [2] "            <table id=\"tablaReportePublico\" class=\"overflow-y\">"
# likely has html code which is why it looks like this. Try resaving as .xlsx (YES, when I tried opening it in excel and resaving it, it was currently a .htm)

# read in San Felipe .xlsx
san_felipe_raw <- readxl::read_xlsx(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/antiquated_resaves","DataturReportePublico_San_Felipe.xlsx"), sheet = 1) # must still have .htm weirdness... Error in read_fun(path = path, sheet_i = sheet, limits = limits, shim = shim,  : invalid Unicode point 56564
```

Try re-saving the raw file as a .CSV again...
```{r}
# for San Felipe still
san_felipe_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/resaved_as_csv","DataturReportePublico_San_Felipe_resaved.csv")) # it works!!!
```




