---
title: "Tourist data"
output: html_document
date: "2025-3-6"
editor_options: 
  chunk_output_type: console
---
Data: "Ocupación en hoteles y moteles por centro turístico, corte mensual DataTur"
Go to: <https://monitoreohotelero.sectur.gob.mx/Reportes/Menu.aspx>

To download:

> 1. Go to the “Generate Report” section.
> 2. Select the “Mensual”, or monthly, time frame.
> 3. Choose the option “Hotel and motel occupancy by tourist center, monthly DataTur snapshot.”
> 4. Click “Continue” and then “Advanced criteria” (Criterios Avanzados)
> 5. Select the municipality of interest, then all the variables of interest.  Specify "Serie" for "Cobertura Temporal corte mensual", starting from 2018 January (Enero) to 2025 Junio (June).
> 6. Finally, click on “Generate Report.”

The report can be downloaded in various formats and contains the information available as of the date of consultation.  It is important to note that this information is presented at the tourist center level, not the municipal level. However, it can be useful for territorial approximations in the Gulf of California region, particularly in key tourist destinations.

```{r}
library(mapview)
library(sf)
library(tidyverse)
library(here)
library(terra)
```


## Clean data

```{r}
# # read in one municipio first. start with San Felipe
# san_felipe_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/antiquated_resaves","DataturReportePublico_San_Felipe.csv")) # hmm.. Error in vroom_(file, delim = delim %||% col_types$delim, col_names = col_names,  :                 
#   # embedded nul in string: 'O\0c\0u\0p\0a\0c\0i\0\xf3\0n\0 \0e\0n\0 \0h\0o\0t\0e\0l\0e\0s\0 \0y\0 \0m\0o\0t\0e\0l\0e\0s\0 \0p\0o\0r\0 \0c\0e\0n\0t\0r\0o\0 \0t\0u\0r\0\xed\0s\0t\0i\0c\0o\0,\0 \0c\0o\0r\0t\0e\0 \0m\0e\0n\0s\0u\0a\0l\0.\0 \0D\0a\0t\0a\0T\0u\0r'
# 
# library(readr)
# library(data.table)
# 
# # It looks like the file is encoded in UTF-16 (or UCS-2) instead of the usual UTF-8 or Latin1, so you need to use `data.table::fread()`
# san_felipe_raw <- read_lines("/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/antiquated_resaves/DataturReportePublico_San_Felipe.csv", 
#                     locale = locale(encoding = "UTF-16LE"))
# 
# # paste lines back together into one string for read_csv
# san_felipe_text <- paste(san_felipe_raw, collapse = "\n")
# 
# # read into a dataframe now
# san_felipe_csv_raw <- read_csv(I(san_felipe_text)) %>% 
#   janitor::clean_names()
# 
# ### did not work... try downloading the report as an excel file instead... could do some cleaning from now on but it might be easier to download as an excel.  Check!!
```
Try the excel file download instead...

```{r}
# library(readxl)
# 
# # read in San Felipe .xls
# san_felipe_raw <- readxl::read_excel(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/antiquated_resaves","DataturReportePublico_San_Felipe.xls"))
# 
# tools::file_ext(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/antiquated_resaves/DataturReportePublico_San_Felipe.xls")) # [1] "xls"
# readLines(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/antiquated_resaves/DataturReportePublico_San_Felipe.xls"))
# 
# # .... [2] "            <table id=\"tablaReportePublico\" class=\"overflow-y\">"
# # likely has html code which is why it looks like this. Try resaving as .xlsx (YES, when I tried opening it in excel and resaving it, it was currently a .htm)
# 
# # read in San Felipe .xlsx
# san_felipe_raw <- readxl::read_xlsx(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/antiquated_resaves","DataturReportePublico_San_Felipe.xlsx"), sheet = 1) # must still have .htm weirdness... Error in read_fun(path = path, sheet_i = sheet, limits = limits, shim = shim,  : invalid Unicode point 56564
```

Try re-saving the raw file as a .CSV again...
```{r}
# for San Felipe still
san_felipe_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/resaved_as_csv","DataturReportePublico_San_Felipe_resaved.csv")) # it works!!!

san_felipe_clean <- san_felipe_raw %>% 
  set_names(.[9, ]) %>%  # make the column names what they should be
  slice(-c(1:10)) %>% 
  janitor::clean_names()

# clean further by trimming/replacing white spaces in the concept headers, as well as making it lower case (for my aesthetics haha)
san_felipe_clean_concept <- san_felipe_clean %>%
  mutate(
    concept = str_to_lower(concepto),
    concept = str_trim(concept),
    concept = str_replace_all(concept, " ", "_")
  ) %>% 
  relocate(concept, .after = concepto) %>% 
  select(-concepto)

# identify which rows are the topic headers, and then give each row a specific name for easier pivoting.
san_felipe_clean_by_topic <- san_felipe_clean_concept %>% 
  mutate(
    is_topic = !str_detect(concept, "^[0-9]_estrellas|[1-5]_estrellas|total_categorias") & !is.na(concept) # identify, true or false, if it is not the subheader (aka it IS a topic header)
  ) %>%
  # fill down header concepts
  mutate(header_concept = if_else(is_topic, concept, NA)) %>% # or NA_character if there are problems.
  fill(header_concept) %>% 
  # make the combined concept by topic: if it’s a subcategory, combine, if not, leave as original
  mutate(
    concept_full = if_else(concept != header_concept, paste0(header_concept, ":", concept), concept
    )
  )

# now, pivot longer, where we have the columns as the header_concept!
san_felipe_long <- san_felipe_clean_by_topic %>% 
  pivot_longer(
    cols = 2:100,
    names_to = "month_year",
    values_to = "value"
  ) # check with original data to ensure it makes sense!

# great.  now let's make a column to indicate the year, making the final San Felipe dataset
san_felipe_long_final <- san_felipe_long %>% 
  mutate(
    is_mo_yr = !str_detect(month_year, "acum|total"),
    temp = if_else(is_mo_yr, month_year, NA)
  ) %>% 
  separate(temp, into = c("month", "year"), sep = "_", remove = FALSE) %>% 
  filter(is_topic %in% FALSE) %>% # we do not want the rows of just titles
  select(header_concept, concept, concept_full, month_year, month, year, value) %>% 
  mutate(municipio = "san_felipe") # before rbinding with other cities!
```

### Function to clean the files based on San Felipe

```{r}
clean_hotel_occupations_data <- function(municipio) {
  
  # where to find the file
  file_path <- here::here(
    "/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/resaved_as_csv",
    paste0("DataturReportePublico_", municipio, "_resaved.csv")
  )
  
  # read raw data
  # raw_data <- read_csv(file_path)
  raw_data <- read_csv(file_path, locale = readr::locale(encoding = "UTF-8")) # changed because having error: "Error in `map()`: ℹ In index: 1. Caused by error in `nchar()`: ! invalid multibyte string, element 1"
  
  # clean the column names, remove top rows, and tidy
  cleaned <- raw_data %>% 
    set_names(.[9, ]) %>%  
    slice(-c(1:10)) %>% # NOTE: not all are the same length, so some of them will have the bottom text (ex: ND No disponible) as part of the output, and there will have to be extra cleaning done at the end.
    janitor::clean_names()
  
  # clean and standardize the "concepto" column
  cleaned_concept <- cleaned %>%
    mutate(
      concept = str_to_lower(concepto),
      concept = str_trim(concept),
      concept = str_replace_all(concept, " ", "_")
    ) %>% 
    relocate(concept, .after = concepto) %>% # these two lines of code are not really necessary because the func will do it all at once
    select(-concepto)
  
  # identify topic headers and combine them with subcategories
  by_topic <- cleaned_concept %>% 
    mutate(
      is_topic = !str_detect(concept, "^[1-5]_estrella[s]?|total_categorias") & !is.na(concept)
    ) %>%
    mutate(header_concept = if_else(is_topic, concept, NA_character_)) %>% # use NA_character_ for this stuff just in case
    fill(header_concept) %>% 
    mutate(
      concept_full = if_else(concept != header_concept, paste0(header_concept, ":", concept), concept)
    )
  
  # pivot longer
  long <- by_topic %>% 
    pivot_longer(
      cols = 2:100, # adjust if more columns
      names_to = "month_year",
      values_to = "value"
    )
  
  # separate year and month, filter, and tag city
  final <- long %>% 
    mutate(
      is_mo_yr = !str_detect(month_year, "acum|total"),
      temp = if_else(is_mo_yr, month_year, NA_character_)
    ) %>% 
    separate(temp, into = c("month", "year"), sep = "_", remove = FALSE) %>% 
    filter(is_topic == FALSE) %>% 
    select(header_concept, concept, concept_full, month_year, month, year, value) %>% 
    mutate(municipio = str_to_lower(municipio)) %>% 
    mutate(value = as.numeric(str_replace_all(value, ",", "")))
  
  return(final)
}
```

Okay! Let's try it with Cabo San Lucas.

```{r}
cabo_san_lucas_final <- clean_hotel_occupations_data("Cabo_San_Lucas") # hmmm... there are weird crosses when resaved as UTF... "†".. but it only works as a UTF-8!

la_paz_final <- clean_hotel_occupations_data("La_Paz") # okay it is happening more than once.  Need to figure that out.  Come back tomorrow!
```

Okay!! So it looks like you CANNOT save the file as a .CSV first, instead you **must** to do it in this order:

1.  On the SECTUR report site, after generating the report, click "Exportar a Excel".  

2.  Open the ".xls" (which is really a .htm) in Excel, and then "Save As..." a UTF-8 .CSV. Rename it `DataturReportePublico_municipio_resaved.csv` where you replace municipio with the municipality you just saved. 

3.  Bring into mazu in `/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/resaved_as_csv` to be read in here.  This is the only way that you do not get any "†" in whitespaces.

**These instructions will be brought to the top of the script when it is cleaned up.**

```{r}
cabo_san_lucas_final <- clean_hotel_occupations_data("Cabo_San_Lucas") # YAY, now that I followed the above procedure, it looks great!!
```

#### Apply function to all municipios

```{r}
all_tourist_centers <- c(
  "Cabo_San_Lucas", "Culiacan", "El_Fuerte", "El_Rosario", "Escuinapa",
  "Guaymas", "Hermosillo", "La_Paz", "Loreto", "Los_Mochis",
  "Mazatlan", "Puerto_Vallarta", "San_Felipe", "San_Jose_del_Cabo", 
  "Zona_Corredor_Los_Cabos"
)

library(purrr)

all_data <- all_tourist_centers %>% 
  map(clean_hotel_occupations_data) %>% 
  bind_rows() # looks good! keep checking after lunch.

all_data_clean <- all_data %>% 
  rename( 
    metric = header_concept,
    category = concept
  ) %>%  # for better clarity, since we are done cleaning and these names make more sense now
  mutate(
    concept_full = str_replace_all(concept_full, "\U{0eef2f}th>|\U{0ee832}|\U{0ee82e}", "_"),
    # metric_label = str_replace_all(metric, "_", " "),
    # metric_label = str_to_title(metric_label), # for later viz 
    metric = str_replace_all(metric, "\U{0eef2f}th>|\U{0ee832}|\U{0ee82e}", "_"),
    category = str_replace_all(category, "\U{0eef2f}th>|\U{0ee832}|\U{0ee82e}", "_"),
    value = str_replace_all(value, ",",""),
    year = str_replace_all(year, "p",""), # only in 2025, assume it means "preliminary"
    month = case_when(
      tolower(month) == "enero" ~ "01",
      tolower(month) == "febrero" ~ "02",
      tolower(month) == "marzo" ~ "03",
      tolower(month) == "abril" ~ "04",
      tolower(month) == "mayo" ~ "05",
      tolower(month) == "junio" ~ "06",
      tolower(month) == "julio" ~ "07",
      tolower(month) == "agosto" ~ "08",
      tolower(month) == "septiembre" ~ "09",
      tolower(month) == "octubre" ~ "10",
      tolower(month) == "noviembre" ~ "11",
      tolower(month) == "diciembre" ~ "12",
      TRUE ~ NA_character_  # otherwise
    ) # for lubridate later
  ) %>% 
  filter(!category %in% NA) 

unique(all_data_clean$concept_full) # check that there are no weirdnesses left

# data_check <- all_data_clean %>% filter(year %in% 2024) %>% filter(metric %in% "cuartos_ocupados_no_residentes") # double check with raw.  looks good!
```

#### Check data completeness for all municipios

```{r}
# check the number of municipalities, there should be 15
unique(all_data_clean$municipio) # 15

# check the unique categories
unique(all_data_clean$concept_full) # there should be 10593 observations for each municipio, but there are 132,561 total observations, not 155,925. Why would there be less for some? A: because not all of them have 1-5 star establishment subsets. This could pose a difficulty for using this data as a way to measure "quality" of tourism, depending on the methodology of this data.  If it is representative of all the hotels in the municipio, then it could be used.  BUT, if there is just a lack of data on the "star rating" of hotels in that area, or if it is not a representative sample, then it should not be used.

# group by municipio, and see the number of unique `concpet_full` for each municipio:
concept_counts <- all_data_clean %>%
  group_by(municipio) %>%
  summarise(unique_concepts = n_distinct(concept_full)) ## hm... not consistent. But it should be in theory...
```

Okay, so upon further look, it appears that some municipios do not have information about hotels that are not 5 stars.  For example, in Zona Corredor Los Cabos, there is only a total category and a 5 star subset, but they are always equal, presumably because there are **only** 5 star hotels in the municipio that are monitored.

Loreto does not have any 2 star subcategories, only 1, 3, 4, 5.  

Cabo San Lucas does not have any 1 star establishments as a subset for observations. Nor does San Felipe or Hermosillo.

Guaymas only has 3, 4, 5 star establishments monitored.


```{r}
# count the number of observations per municipio
observations_per_municipio <- all_data_clean %>%
  group_by(municipio) %>%
  summarise(n_obs = n()) # guaymas, loreto, and zona corredor have much less observations than the others.
```


## Exploration of data

From this data, let us see what is available for analysis, and what each type of visitation means.

This data source translates to: "Hotel and motel occupancy by tourist center, monthly breakdown. DataTur
Provides up to 126 variables of information with final results presented no later than the 20th of each month. Consult metadata to identify availability of information by tourist destination." 

So, `concept_full_clean` contains values of the number of visitors, with monthly data for each municipio, and each variable is a combination of:

* Metrics: eg, cuartos_disponibles, cuartos_ocupados, llegadas_de_turistas, porcentaje_de_ocupación, etc.

* Type of visitor: overall, residentes, no_residentes

* Hotel category: 5_estrellas, 4_estrellas, etc, or total_categorias


Explore the metrics available:
```{r}
# check which metrics are available for analysis
count_metric <- all_data_clean %>%
  count(metric, sort = TRUE) # hm, `densidad_de_ocupaci_o_residentes` has many more observations than the others, (10890 vs 7920 for all the others)

# oh no... a category is within the metric (1 estrella).... [FIXED]
# all_data_1_est <- all_data_clean %>% 
#   filter(metric == "1_estrella") # AH, this is because of the missing "s" at the end of estrella, since there is only one star.  Will alter the cleaning function.

all_data_densidad_de_ocupaci_o_residentes <- all_data_clean %>% 
  filter(metric %in% "densidad_de_ocupaci_o_residentes")
unique(all_data_densidad_de_ocupaci_o_residentes$category) # 7: has NA as a category

missing_data_densidad_de_ocupaci_o_residentes <- all_data_clean %>% 
  filter(metric %in% "densidad_de_ocupaci_o_residentes") %>% 
  filter(category %in% "total_categorias") %>% 
  filter(is.na(value)) # guaymas is missing data for every month of every year after 2018.
unique(missing_data_densidad_de_ocupaci_o_residentes$municipio) # "el_fuerte"  "el_rosario" "escuinapa"  "guaymas" 


all_data_densidad_de_ocupaci_esidentes <- all_data_clean %>% 
  filter(metric %in% "densidad_de_ocupaci_esidentes")
unique(all_data_densidad_de_ocupaci_esidentes$category) # 6

all_data_NA_category <- all_data_clean %>% 
  filter(category %in% NA)
length(unique(all_data_NA_category$municipio)) # yup, for all 15. Without any values associated. Let's get rid of this when creating `all_data_clean` on line 229.
# that was fixed, now they all have 7920 observations for each metric.  YAY!
```


Explore the categories available (aka star levels) each metric has:
```{r}
# okay, now lets explore categories once more
count_cat <- all_data_clean %>%
  filter(category %in% "total_categorias") %>%  # since we will likely not be using the star system
  filter(str_detect(month_year, "total_anual_")) %>%  # only keep observations with values
  group_by(metric) %>%
  summarise(n_muni = n_distinct(year)) %>%
  arrange(desc(n_muni)) 
```

### Vizualizing data availability

```{r}
library(ggplot2)

# make a heatmap of the number of observations for each municipio.
# does not actually show us the data completeness
all_data_clean %>%
  count(municipio, metric) %>% # unique counts for each municipio and metric.  like doing group_by(municipio, metric) %>% summarise(n = n())
  ggplot(aes(x = metric, y = municipio, fill = n)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title = "Number of observations per municipio and metric", x = "Metric", y = "Municipio") # but, this could vary because each municipio might have less observations if they only consider 5 star hotels for example. 
```

It makes sense that Zona Corredor Los Cabos has less observations because it only considers 5 star hotels.  This also make sense, as it is not a full municipality, but instead just a stretch of high tourism activity in an area of Los Cabos, so many of the hotels likely fit the requirements for a 5 star: "Establishments with high-end amenities such as restaurants, bars with entertainment, extensive services, bilingual staff, quality furnishings. Includes Gran Turismo and Special Class categories with additional luxury or unique historical/architectural value."

```{r}
## to see the number of tourism arrivals per year for each municipio.  Note that some are missing data after 2022.

library(plotly)

all_data_llegadas_de_turistas <- all_data_clean %>% 
  filter(metric %in% "llegadas_de_turistas")

plot_data <- all_data_clean %>%
  filter(metric %in% "llegadas_de_turistas", category %in% "total_categorias") %>% 
  select(-c(month, year)) %>% 
  mutate(value = as.numeric(value),
         month_year_total = str_detect(month_year, "total")) %>% 
  filter(month_year_total == "TRUE") %>% 
  separate(month_year, into = c("type", "year"), sep = "_anual_", remove = FALSE) # to only look at anual total for each metric (as there will only be a total yearly value if there are values for the months in that year)


# preliminary plot 
line_plot <- plotly::plot_ly(plot_data, x = ~year, y = ~value, color = ~municipio, type = "scatter", mode = "lines") %>% 
  layout(title = "Llegadas de Turistas (total per year for all hotels) por Municipio", 
         xaxis = list(title = "Year"),
         yaxis = list(title = "Value"))

line_plot

# htmlwidgets::saveWidget(line_plot, file = "llegadas_de_turistas.html")
```

Oh wow, okay this brings to my attention that we are missing data after 2019 for Guaymas, 2022 for El Rosario and Escuinapa in Sinaloa, and El Fuerte in Sinaloa after 2023.  The rest have an appropriate dip in 2020 (keep in mind these are annual totals of arrivals).  This is only looking at "Llegadas de Turistas", though there are also: 

| Metric Variable                         | English Translation                           |
|-----------------------------------------|-----------------------------------------------|
| cuartos_registrados_fin_per_o           | Rooms Registered at End of Period             |
| cuartos_disponibles                     | Available Rooms                               |
| cuartos_disponibles_promedio            | Average Available Rooms                       |
| cuartos_ocupados                        | Occupied Rooms                                |
| cuartos_ocupados_residentes             | Occupied Rooms (Residents)                    |
| cuartos_ocupados_no_residentes          | Occupied Rooms (Non-Residents)                |
| llegadas_de_turistas                    | Tourist Arrivals                              |
| llegadas_de_turistas_residentes         | Tourist Arrivals (Residents)                  |
| llegadas_de_turistas_no_residentes      | Tourist Arrivals (Non-Residents)              |
| turistas_noche                          | Tourist-Nights (Person-Nights)                |
| turistas_noche_residentes               | Tourist-Nights (Residents)                    |
| turistas_noche_no_residentes            | Tourist-Nights (Non-Residents)                |
| porcentaje_de_ocupaci_                  | Occupancy Rate                                |
| porcentaje_de_ocupaci_esidentes         | Occupancy Rate (Residents)                    |
| porcentaje_de_ocupaci_o_residentes      | Occupancy Rate (Non-Residents)                |
| estad_promedio                          | Average Length of Stay                        |
| estad_promedio_residentes               | Average Length of Stay (Residents)            |
| estad_promedio_no_residentes            | Average Length of Stay (Non-Residents)        |
| densidad_de_ocupaci_                    | Occupancy Density                             |
| densidad_de_ocupaci_esidentes           | Occupancy Density (Residents)                 |
| densidad_de_ocupaci_o_residentes        | Occupancy Density (Non-Residents)             |

Let's look only at "total_categorias" within categories:

```{r}
# total category: the total of all possible hotels (1-5 stars, if available)
all_data_total_categorias <- all_data_clean %>% 
  filter(category == "total_categorias") %>% 
  filter(!grepl("total_anual_", month_year, ignore.case = TRUE), # removing the totals, so we can evaluate monthly
         !grepl("acumulado_", month_year, ignore.case = TRUE)) # and removing the accumulated total from 2018 - 2025

# check on guaymas
all_data_total_categorias_guay <- all_data_clean %>%
  filter(category == "total_categorias") %>% 
  mutate(has_data = !is.na(value)) %>% 
  filter(municipio %in% "guaymas") # for all metrics, it seems there is missing data after mid-2019

# list of metric variables
metric_vars <- c(
  "cuartos_registrados_fin_per_o",
  "cuartos_disponibles",
  "cuartos_disponibles_promedio",
  "cuartos_ocupados",
  "cuartos_ocupados_residentes",
  "cuartos_ocupados_no_residentes",
  "llegadas_de_turistas",
  "llegadas_de_turistas_residentes",
  "llegadas_de_turistas_no_residentes",
  "turistas_noche",
  "turistas_noche_residentes",
  "turistas_noche_no_residentes",
  "porcentaje_de_ocupaci_",
  "porcentaje_de_ocupaci_esidentes",
  "porcentaje_de_ocupaci_o_residentes",
  "estad_promedio",
  "estad_promedio_residentes",
  "estad_promedio_no_residentes",
  "densidad_de_ocupaci_",
  "densidad_de_ocupaci_esidentes",
  "densidad_de_ocupaci_o_residentes"
)

# func to make a heatmap for each metric variable
plot_metric_heatmap <- function(metric_var) {
  all_data_total_categorias %>%
    filter(metric == metric_var) %>%
    mutate(has_data = !is.na(value)) %>% 
    ggplot(aes(x = year, y = municipio, fill = has_data)) +
    geom_tile(color = "white") +
    scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "grey90"),
                      name = "Data available") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    labs(
      title = paste("Data completeness:", metric_var),
      x = "Year",
      y = "Municipio"
    )
}

heatmap_list <- map(metric_vars, plot_metric_heatmap)

# view them.. there are 21 variables
heatmap_list[[1]]

# save them all to files:
# walk2(heatmap_list, metric_vars, ~ ggsave(
#   filename = paste0("heatmap_sectur_completeness_", .y, ".png"),
#   plot = .x, width = 8, height = 5
# ))

```

# Tourism Quantity

Now that we have the data source cleaned up in general, let's clean it up specifically for tourism "quantity" use.

We think that tourism arrivals, aka, the number of individuals, matter the most in terms of value. If we count not just the number of tourist that come to the area, but also the days that they stay, then we are inferring that there is a higher value of the visit by the person who stays longer.  However, many people do not have the finances to stay very long, but they still value the trip and location just as much, if not more.  

So, we want to filter to the tourism metric tourist arrivals, "llegadas_de_turistas", and have category = "total_categorias" because we value 1-star establishments the same as 5-star (just a different type of tourist, does not necessarily mean one provides the better value trip).

```{r}
# so that we can have the yearly total, not just by month, because we will have the scores per year.
all_data_yearly <- all_data_clean %>% 
  filter(str_detect(month_year, "total_anual_")) %>% 
  select(-c(month, year)) %>% 
  separate_wider_delim(., month_year, delim = "_anual_", names = c("type","year"))

## save into int
write_csv(all_data_yearly, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/int","hotel_data_cleaned_yearly.csv"))
```


```{r}
quantity_clean <- all_data_yearly %>% 
  filter(metric %in% "llegadas_de_turistas") %>% 
  filter(category %in% "total_categorias")

# let's make them an sf object
quantity_geocode <- quantity_clean %>% 
  mutate(country = "Mexico") %>% 
  mutate(estado = case_when(
    municipio %in% c("cabo_san_lucas", "la_paz", "loreto", "san_jose_del_cabo", "zona_corredor_los_cabos") ~ "Baja California Sur",
    municipio %in% c("san_felipe") ~ "Baja California",
    municipio %in% c("guaymas", "hermosillo") ~ "Sonora",
    municipio %in% c("culiacan", "el_fuerte", "los_mochis", "mazatlan", "escuinapa", "el_rosario") ~ "Sinaloa",
    municipio %in% c("puerto_vallarta") ~ "Jalisco",
    TRUE ~ ""  # default case for any unmatched cities
  )) %>% 
  tidygeocoder::geocode(city = municipio, country = country, state = estado) # to give each city a lat, long
  
quantity_sf <- quantity_geocode %>% 
  mutate(lat = ifelse(municipio %in% "zona_corredor_los_cabos", 23.1539, lat)) %>% # adding in the coordinates manually from <https://latitude.to/articles-by-country/mx/mexico/56998/los-cabos-corridor>
  mutate(long = ifelse(municipio %in% "zona_corredor_los_cabos", -109.720233, long)) %>% 
  st_as_sf(coords = c("long", "lat"),
           crs = 4326, # WGS84
           remove = FALSE)

mapview(quantity_sf)
```

Now, let us assign each tourist center to an OHI region. 

```{r}
goc_poly <- st_read(here("spatial/ohi_regions","goc_ohi_rgns.shp"))
st_crs(goc_poly) # not WGS84: Mexico ITRF2008 / LCC
mapview(goc_poly)

quantity_sf_transform <- st_transform(quantity_sf, crs = st_crs(goc_poly)) %>%  # transform the tourist centers to Mexico ITRF2008 / LCC
   mutate(value = as.numeric(value))
crs(quantity_sf_transform)

quantity_goc <- st_intersection(quantity_sf_transform, goc_poly)
mapview(quantity_goc) # now each tourist center has associated region data

```

A cool plot to see them each:

```{r}
# map of all goc tourist centers
mapview(quantity_goc, zcol = "rgn_id") + mapview(goc_poly, zcol = "rgn_id")
```

### Current status

The reference point could be a 40% increase in total tourist arrivals per year, assuming that it does not really matter if they are international or domestic: <https://tourismanalytics.com/blog-posts/how-mexicos-forty-percent-tourism-growth-plan-aims-for-fifth-global-spot-by-2030-with-twenty-billion-dollar-investment-in-infrastructure-and-market-expansion>

The most current full year of data is 2024.  Plan México was announced on January 13, 2025, which means the 40% tourism growth target by 2030 would logically use 2024 as the baseline year.

So, the current status for the quantity could be:

(Current arrivals / Target arrivals) × 100

Score of 100 = meeting the 2030 target
Score < 100 = below target
Score > 100 = exceeding target

```{r}
# identify the reference point for each tourist center using 2024 as the baseline
baseline_2024 <- quantity_goc %>% 
  filter(year == "2024") %>% 
  select(municipio, rgn_id, baseline_arrivals = value) %>% 
  mutate(baseline_arrivals = as.numeric(baseline_arrivals))
```


```{r}
## there are NAs for tourist centers without data. drop them to viz
baseline_2024_no_NAs <- drop_na(baseline_2024, baseline_arrivals)

# map of the tourist centers with available data 2024
mapview(baseline_2024_no_NAs, zcol = "rgn_id") + mapview(goc_poly, zcol = "rgn_id")
```


```{r}
# calc the target values (40% increase from 2024 baseline)
tourism_targets <- baseline_2024 %>% 
  mutate(target_2030 = baseline_arrivals * 1.40) %>%  # 40% increase means we multiply by 1.40
  st_drop_geometry()
```

Now, use the reference point to identify how close each tourist center is to the ideal (by 2030).

```{r}
tourism_quantity_score <- quantity_goc %>%
  left_join(tourism_targets, by = c("municipio", "rgn_id")) %>% 
  mutate(
    # calculate current tourism arrivals as percentage of 2030 sustainability target (baseline year 2024)
    index_score = (value / target_2030) * 100,
    
    # rename 2030 target values for easy comparison and reference in analysis
    target_arrivals = target_2030,
    
    # the difference between the # of current arrivals for that year and 2030 target in raw numbers
    gap_to_target = target_2030 - value,
    
    # calc percentage increase required from current levels to reach 2030 target (approximately 40% growth needed from 2024 baseline)
    gap_percentage = ((target_2030 - value) / value) * 100
  ) %>% 
  select(municipio, rgn_id, year, current_arrivals = value, 
         baseline_arrivals, target_arrivals, index_score, 
         gap_to_target, gap_percentage, everything())
```


```{r}
# preliminary plot of the index score with target reference line
line_plot_score <- plotly::plot_ly(tourism_quantity_score, x = ~year, y = ~index_score, color = ~municipio, type = "scatter", mode = "lines") %>% 
  layout(title = "INDEX SCORE: Llegadas de Turistas (total per year for all hotels) por Municipio", 
         xaxis = list(title = "Year"),
         yaxis = list(title = "Score (out of 100)")) %>%
  add_lines(x = range(tourism_quantity_score$year), y = c(100, 100), 
            line = list(color = "black", width = 2, dash = "dash"),
            name = "2030 Target (100/100)", 
            inherit = FALSE)

line_plot_score
```


# Domestic vs. International

Since there are different targets within Plan México for Domestic versus International visitors:

> "Para lograr la meta de ser el quinto país visitado del mundo en el 2030 se estima que es necesario aumentar tres millones más de turistas cada año, así como incrementar:

> Los cuartos de hotel en 12 por ciento.
> Del turismo nacional en 9.8 por ciento.
> La llegada de visitantes internacionales en 30 por ciento."

What we will do is determine the current status for domestic and international visitors separately, and then average those two to produce the tourism:quantity current status.

```{r}
unique(all_data_yearly$metric)

#  [1] "cuartos_registrados_fin_per_o"      "cuartos_disponibles"               
#  [3] "cuartos_disponibles_promedio"       "cuartos_ocupados"                  
#  [5] "cuartos_ocupados_residentes"        "cuartos_ocupados_no_residentes"    
#  [7] "llegadas_de_turistas"               "llegadas_de_turistas_residentes"   
#  [9] "llegadas_de_turistas_no_residentes" "turistas_noche"                    
# [11] "turistas_noche_residentes"          "turistas_noche_no_residentes"      
# [13] "porcentaje_de_ocupaci_"             "porcentaje_de_ocupaci_esidentes"   
# [15] "porcentaje_de_ocupaci_o_residentes" "estad_promedio"                    
# [17] "estad_promedio_residentes"          "estad_promedio_no_residentes"      
# [19] "densidad_de_ocupaci_"               "densidad_de_ocupaci_esidentes"     
# [21] "densidad_de_ocupaci_o_residentes" 
```


### Domestic

```{r}
## "llegadas_de_turistas_residentes" = domestic visitors
dom_arrivals <- all_data_yearly %>% 
  filter(metric %in% "llegadas_de_turistas_residentes") %>% # domestic arrivals
  filter(category %in% "total_categorias") # since we do not care what star designation it is
```

```{r}
# let's make them an sf object
dom_geocode <- dom_arrivals %>% 
  mutate(country = "Mexico") %>% 
  mutate(estado = case_when(
    municipio %in% c("cabo_san_lucas", "la_paz", "loreto", "san_jose_del_cabo", "zona_corredor_los_cabos") ~ "Baja California Sur",
    municipio %in% c("san_felipe") ~ "Baja California",
    municipio %in% c("guaymas", "hermosillo") ~ "Sonora",
    municipio %in% c("culiacan", "el_fuerte", "los_mochis", "mazatlan", "escuinapa", "el_rosario") ~ "Sinaloa",
    municipio %in% c("puerto_vallarta") ~ "Jalisco",
    TRUE ~ ""  # default case for any unmatched cities
  )) %>% 
  tidygeocoder::geocode(city = municipio, country = country, state = estado) # to give each city a lat, long
  
dom_geocode_sf <- dom_geocode %>% 
  mutate(lat = ifelse(municipio %in% "zona_corredor_los_cabos", 23.1539, lat)) %>% # adding in the coordinates manually from <https://latitude.to/articles-by-country/mx/mexico/56998/los-cabos-corridor>
  mutate(long = ifelse(municipio %in% "zona_corredor_los_cabos", -109.720233, long)) %>% 
  st_as_sf(coords = c("long", "lat"),
           crs = 4326, # WGS84
           remove = FALSE)

mapview(dom_geocode_sf)

############################################################
# Now, let us assign each tourist center to an OHI region. 
############################################################

goc_poly <- st_read(here("spatial/ohi_regions","goc_ohi_rgns.shp"))
st_crs(goc_poly) # not WGS84: Mexico ITRF2008 / LCC
mapview(goc_poly)

dom_geocode_sf_transform <- st_transform(dom_geocode_sf, crs = st_crs(goc_poly)) %>%  # transform the tourist centers to Mexico ITRF2008 / LCC
   mutate(value = as.numeric(value))
crs(dom_geocode_sf_transform)

dom_goc <- st_intersection(dom_geocode_sf_transform, goc_poly) %>% 
  mutate(arrivals_type = "domestic")
mapview(dom_goc) # now each tourist center has associated region data

##############################
#A cool plot to see them each:
##############################

# map of all goc tourist centers
mapview(dom_goc, zcol = "rgn_id") + mapview(goc_poly, zcol = "rgn_id")
```


The reference point could be a 9.8% increase in domestic tourist arrivals by 2030, which starting from 2024 is 5 years (since you would want to reach that goal at the end of 2029 to reach the target *by* 2030). <https://tourismanalytics.com/blog-posts/how-mexicos-forty-percent-tourism-growth-plan-aims-for-fifth-global-spot-by-2030-with-twenty-billion-dollar-investment-in-infrastructure-and-market-expansion>

Therefore, each year, domestic tourism arrivals should increase by 1.96%.

The most current full year of data is 2024.  Plan México was announced on January 13, 2025, which means the 9.8% tourism growth target by 2030 would logically use 2024 as the baseline year.

So, the current status for domestic quantity could be:

(Current arrivals / Target arrivals) × 100

Score of 100 = on track to meet the 2030 target
Score < 100 = below target expected level
Score > 100 = exceeding target, yay!

*We will use linear growth, even if it is not as realistic, for now.*

```{r}
growth_rate_linear_dom <- 0.0196  # 1.96% per year (9.8/5, linear growth, for now)
baseline_year <- 2024

dom_quantity_score <- dom_goc %>%
  group_by(municipio, rgn_id) %>%
  mutate(
    year = as.numeric(year),
    
    # determine the baseline value for each municipio (2024)
    dom_baseline_arrivals = value[year == baseline_year][1],

    # to know how far from 2024 we are (note: negative numbers are because of years before 2024)
    years_since_baseline = year - baseline_year,

    # target arrivals for domestic visitors
    dom_target_arrivals = dom_baseline_arrivals * (1 + (growth_rate_linear_dom * years_since_baseline)),  # the target number of arrivals should change based on how many years it is from the baseline. For example, two years from 2024 should have a growth of 1.96% * 2.  

    # make it a score
    dom_index_score = (value / dom_target_arrivals) * 100,
    dom_gap_to_target = dom_target_arrivals - value,
    dom_gap_percentage = ((dom_target_arrivals - value) / value) * 100
  ) %>%
  ungroup() %>%
  select(municipio, rgn_id, year, dom_current_arrivals = value,
         dom_baseline_arrivals, dom_target_arrivals, dom_index_score,
         dom_gap_to_target, dom_gap_percentage, everything())
```


```{r}
library(plotly)

## plot the score over time
dom_score_plotly <- plotly::plot_ly(
  dom_quantity_score, 
  x = ~year, 
  y = ~dom_index_score, 
  color = ~municipio, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "CURRENT STATUS: Llegadas de turistas nacionales por municipio",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(dom_quantity_score$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
dom_score_plotly

htmlwidgets::saveWidget(dom_score_plotly, here::here("TR/sectur_quantity_exploration_int","dom_arrivals_score_100825.html"), selfcontained = TRUE)


## compare to the patterns of domestic arrivals over time
dom_arrivals_plot <- plotly::plot_ly(
  dom_goc, 
  x = ~year, 
  y = ~value, 
  color = ~municipio, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "Llegadas de turistas nacionales por municipio (domestic arrivals)",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Llegadas // Arrivals")
  ) %>% 
  add_lines(
    x = range(dom_goc$year),
    y = c(100, 100),
    line = list(color = "white", width = 2),
    name = "",
    inherit = FALSE
  )
dom_arrivals_plot

htmlwidgets::saveWidget(dom_arrivals_plot, here::here("TR/sectur_quantity_exploration_int","dom_arrivals_plot_100825.html"), selfcontained = TRUE)
```


### International

```{r}
## "llegadas_de_turistas_no_residentes" = international visitors
int_arrivals <- all_data_yearly %>% 
  filter(metric %in% "llegadas_de_turistas_no_residentes")%>% # int arrivals
  filter(category %in% "total_categorias") # since we do not care what star designation it is
```

```{r}
# let's make them an sf object
int_geocode <- int_arrivals %>% 
  mutate(country = "Mexico") %>% 
  mutate(estado = case_when(
    municipio %in% c("cabo_san_lucas", "la_paz", "loreto", "san_jose_del_cabo", "zona_corredor_los_cabos") ~ "Baja California Sur",
    municipio %in% c("san_felipe") ~ "Baja California",
    municipio %in% c("guaymas", "hermosillo") ~ "Sonora",
    municipio %in% c("culiacan", "el_fuerte", "los_mochis", "mazatlan", "escuinapa", "el_rosario") ~ "Sinaloa",
    municipio %in% c("puerto_vallarta") ~ "Jalisco",
    TRUE ~ ""  # default case for any unmatched cities
  )) %>% 
  tidygeocoder::geocode(city = municipio, country = country, state = estado) # to give each city a lat, long
  
int_geocode_sf <- int_geocode %>% 
  mutate(lat = ifelse(municipio %in% "zona_corredor_los_cabos", 23.1539, lat)) %>% # adding in the coordinates manually from <https://latitude.to/articles-by-country/mx/mexico/56998/los-cabos-corridor>
  mutate(long = ifelse(municipio %in% "zona_corredor_los_cabos", -109.720233, long)) %>% 
  st_as_sf(coords = c("long", "lat"),
           crs = 4326, # WGS84
           remove = FALSE)

mapview(int_geocode_sf)

############################################################
# Now, let us assign each tourist center to an OHI region. 
############################################################

goc_poly <- st_read(here("spatial/ohi_regions","goc_ohi_rgns.shp"))
st_crs(goc_poly) # not WGS84: Mexico ITRF2008 / LCC
mapview(goc_poly)

int_geocode_sf_transform <- st_transform(int_geocode_sf, crs = st_crs(goc_poly)) %>%  # transform the tourist centers to Mexico ITRF2008 / LCC
   mutate(value = as.numeric(value))
crs(int_geocode_sf_transform)

int_goc <- st_intersection(int_geocode_sf_transform, goc_poly) %>% 
  mutate(arrivals_type = "international")
mapview(int_goc) # now each tourist center has associated region data

##############################
#A cool plot to see them each:
##############################

# map of all goc tourist centers
mapview(int_goc, zcol = "rgn_id") + mapview(goc_poly, zcol = "rgn_id")
```

The reference point could be a 30% increase in international tourist arrivals by 2030, which starting from 2024 is 5 years (since you would want to reach that goal at the end of 2029 to reach the target *by* 2030). <https://tourismanalytics.com/blog-posts/how-mexicos-forty-percent-tourism-growth-plan-aims-for-fifth-global-spot-by-2030-with-twenty-billion-dollar-investment-in-infrastructure-and-market-expansion>

Therefore, each year, international tourism arrivals should increase by 6%.

The most current full year of data is 2024.  Plan México was announced on January 13, 2025, which means the 30% tourism growth target by 2030 would logically use 2024 as the baseline year.

So, the current status for domestic quantity could be:

(Current arrivals / Target arrivals) × 100

Score of 100 = on track to meet the 2030 target
Score < 100 = below target expected level
Score > 100 = exceeding target, yay!

*We will use linear growth, even if it is not as realistic, for now.*

```{r}
growth_rate_linear_int <- 0.06 # 1.96% per year (30/5, linear growth, for now)
baseline_year <- 2024

int_quantity_score <- int_goc %>%
  group_by(municipio, rgn_id) %>%
  mutate(
    year = as.numeric(year),
    
    # determine the baseline value for each municipio (2024)
    int_baseline_arrivals = value[year == baseline_year][1],

    # to know how far from 2024 we are (note: negative numbers are because of years before 2024)
    years_since_baseline = year - baseline_year,

    # target arrivals for international visitors
    int_target_arrivals = int_baseline_arrivals * (1 + (growth_rate_linear_int * years_since_baseline)),  # the target number of arrivals should change based on how many years it is from the baseline. For example, two years from 2024 should have a growth of 1.96% * 2.  

    # make it a score
    int_index_score = (value / int_target_arrivals) * 100,
    int_gap_to_target = int_target_arrivals - value,
    int_gap_percentage = ((int_target_arrivals - value) / value) * 100
  ) %>%
  ungroup() %>%
  select(municipio, rgn_id, year, int_current_arrivals = value,
         int_baseline_arrivals, int_target_arrivals, int_index_score,
         int_gap_to_target, int_gap_percentage, everything())
```


```{r}
library(plotly)

## plot the score over time
int_score_plotly <- plotly::plot_ly(
  int_quantity_score, 
  x = ~year, 
  y = ~int_index_score, 
  color = ~municipio, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "CURRENT STATUS: Llegadas de turistas internacionales por municipio",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(int_quantity_score$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
int_score_plotly

htmlwidgets::saveWidget(int_score_plotly, here::here("TR/sectur_quantity_exploration_int","int_arrivals_score_100825.html"), selfcontained = TRUE)


## compare to the patterns of international arrivals over time
int_arrivals_plot <- plotly::plot_ly(
  int_goc, 
  x = ~year, 
  y = ~value, 
  color = ~municipio, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "Llegadas de turistas internacionales por municipio (international arrivals)",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Llegadas // Arrivals")
  ) %>% 
  add_lines(
    x = range(int_goc$year),
    y = c(100, 100),
    line = list(color = "white", width = 2),
    name = "",
    inherit = FALSE
  )
int_arrivals_plot

htmlwidgets::saveWidget(int_arrivals_plot, here::here("TR/sectur_quantity_exploration_int","int_arrivals_plot_100825.html"), selfcontained = TRUE)
```




### Merge

```{r}
# join int_quantity_score and dom_quantity_score

## check for duplicates
dom_dups <- dom_quantity_score %>%
  st_drop_geometry() %>%
  count(municipio, rgn_id, year) %>%
  filter(n > 1)

int_dups <- int_quantity_score %>%
  st_drop_geometry() %>%
  count(municipio, rgn_id, year) %>%
  filter(n > 1)

dom_dups    # correct: they should not have any duplicates!!
int_dups

dom_df <- dom_quantity_score %>% st_drop_geometry()
int_df <- int_quantity_score %>% st_drop_geometry()

quantity_scores <- full_join(
  dom_df,
  int_df,
  by = c("municipio", "rgn_id", "year"),
  suffix = c(".dom", ".int")
)

quantity_scores %>%
  mutate(has_dom = !is.na(dom_index_score),
         has_int = !is.na(int_index_score)) %>%
  count(has_dom, has_int)

# Missing int
quantity_scores %>% filter(is.na(int_index_score)) %>% select(municipio, rgn_id, year)

# Missing dom
quantity_scores %>% filter(is.na(dom_index_score)) %>% select(municipio, rgn_id, year)

dom_geom <- dom_quantity_score %>%
  select(municipio, rgn_id, geometry) %>%
  distinct(municipio, rgn_id, .keep_all = TRUE)
```


```{r}
quantity_scores_sf <- quantity_scores %>%
  left_join(dom_geom, by = c("municipio", "rgn_id")) %>%
  st_as_sf() %>% 
  select(municipio, rgn_id, year, dom_index_score, int_index_score, geometry) %>% 
  mutate(
    current_status_quantity = (dom_index_score + int_index_score) / 2
  )
```

```{r}
## plot the score over time
quantity_score_plotly <- plotly::plot_ly(
  quantity_scores_sf, 
  x = ~year, 
  y = ~current_status_quantity, 
  color = ~municipio, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "CURRENT STATUS: Llegadas de turistas (todos) por municipio",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(quantity_scores_sf$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
quantity_score_plotly

htmlwidgets::saveWidget(quantity_score_plotly, here::here("TR/sectur_quantity_exploration_int","all_arrivals_score_100825.html"), selfcontained = TRUE)

```

```{r}
# ## averaged score per region
# quantity_scores_rgn <- quantity_scores_sf %>% 
#   group_by(rgn_id) %>% 
#   summarize(quantity_score_rgn = mean(current_status_quantity))
# 
# mapview(quantity_scores_rgn)
```

# NEXT STEPS: I need to join with goc_poly and assign the values within the regions to the entire region. 
