---
title: "Tourist data"
output: html_document
date: "2025-3-6"
editor_options: 
  chunk_output_type: console
---
Data: "Ocupación en hoteles y moteles por centro turístico, corte mensual DataTur"
Go to: <https://monitoreohotelero.sectur.gob.mx/Reportes/Menu.aspx>

To download:

> 1. Go to the “Generate Report” section.
> 2. Select the “Mensual”, or monthly, time frame.
> 3. Choose the option “Hotel and motel occupancy by tourist center, monthly DataTur snapshot.”
> 4. Click “Continue” and then “Advanced criteria” (Criterios Avanzados)
> 5. Select the municipality of interest, then all the variables of interest.  Specify "Serie" for "Cobertura Temporal corte mensual", starting from 2018 January (Enero) to 2025 Junio (June).
> 6. Finally, click on “Generate Report.”

The report can be downloaded in various formats and contains the information available as of the date of consultation.  It is important to note that this information is presented at the tourist center level, not the municipal level. However, it can be useful for territorial approximations in the Gulf of California region, particularly in key tourist destinations.

```{r}
library(mapview)
library(sf)
library(tidyverse)
library(here)
library(terra)
```


## Clean data

```{r}
# read in one municipio first. start with San Felipe
san_felipe_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles","DataturReportePublico_San_Felipe.csv")) # hmm.. Error in vroom_(file, delim = delim %||% col_types$delim, col_names = col_names,  :                 
  # embedded nul in string: 'O\0c\0u\0p\0a\0c\0i\0\xf3\0n\0 \0e\0n\0 \0h\0o\0t\0e\0l\0e\0s\0 \0y\0 \0m\0o\0t\0e\0l\0e\0s\0 \0p\0o\0r\0 \0c\0e\0n\0t\0r\0o\0 \0t\0u\0r\0\xed\0s\0t\0i\0c\0o\0,\0 \0c\0o\0r\0t\0e\0 \0m\0e\0n\0s\0u\0a\0l\0.\0 \0D\0a\0t\0a\0T\0u\0r'

library(readr)
library(data.table)

# It looks like the file is encoded in UTF-16 (or UCS-2) instead of the usual UTF-8 or Latin1, so you need to use `data.table::fread()`
san_felipe_raw <- read_lines("/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/DataturReportePublico_San_Felipe.csv", 
                    locale = locale(encoding = "UTF-16LE"))

# paste lines back together into one string for read_csv
san_felipe_text <- paste(san_felipe_raw, collapse = "\n")

# read into a dataframe now
san_felipe_csv_raw <- read_csv(I(san_felipe_text)) %>% 
  janitor::clean_names()

### did not work... try downloading the report as an excel file instead... could do some cleaning from now on but it might be easier to download as an excel.  Check!!
```
Try the excel file download instead...

```{r}
library(readxl)

# read in San Felipe .xls
san_felipe_raw <- readxl::read_excel(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/antiquated_resaves","DataturReportePublico_San_Felipe.xls"))

tools::file_ext(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/antiquated_resaves/DataturReportePublico_San_Felipe.xls")) # [1] "xls"
readLines(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/antiquated_resaves/DataturReportePublico_San_Felipe.xls"))

# .... [2] "            <table id=\"tablaReportePublico\" class=\"overflow-y\">"
# likely has html code which is why it looks like this. Try resaving as .xlsx (YES, when I tried opening it in excel and resaving it, it was currently a .htm)

# read in San Felipe .xlsx
san_felipe_raw <- readxl::read_xlsx(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/antiquated_resaves","DataturReportePublico_San_Felipe.xlsx"), sheet = 1) # must still have .htm weirdness... Error in read_fun(path = path, sheet_i = sheet, limits = limits, shim = shim,  : invalid Unicode point 56564
```

Try re-saving the raw file as a .CSV again...
```{r}
# for San Felipe still
san_felipe_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/resaved_as_csv","DataturReportePublico_San_Felipe_resaved.csv")) # it works!!!

san_felipe_clean <- san_felipe_raw %>% 
  set_names(.[9, ]) %>%  # make the column names what they should be
  slice(11:136) %>% 
  janitor::clean_names()

# clean further by trimming/replacing white spaces in the concept headers, as well as making it lower case (for my aesthetics haha)
san_felipe_clean_concept <- san_felipe_clean %>%
  mutate(
    concept = str_to_lower(concepto),
    concept = str_trim(concept),
    concept = str_replace_all(concept, " ", "_")
  ) %>% 
  relocate(concept, .after = concepto) %>% 
  select(-concepto)

# identify which rows are the topic headers, and then give each row a specific name for easier pivoting.
san_felipe_clean_by_topic <- san_felipe_clean_concept %>% 
  mutate(
    is_topic = !str_detect(concept, "^[0-9]_estrellas|[1-5]_estrellas|total_categorias") & !is.na(concept) # identify, true or false, if it is not the subheader (aka it IS a topic header)
  ) %>%
  # fill down header concepts
  mutate(header_concept = if_else(is_topic, concept, NA)) %>% # or NA_character if there are problems.
  fill(header_concept) %>% 
  # make the combined concept by topic: if it’s a subcategory, combine, if not, leave as original
  mutate(
    concept_full = if_else(concept != header_concept, paste0(header_concept, ":", concept), concept
    )
  )

# now, pivot longer, where we have the columns as the header_concept!
san_felipe_long <- san_felipe_clean_by_topic %>% 
  pivot_longer(
    cols = 2:100,
    names_to = "month_year",
    values_to = "value"
  ) # check with original data to ensure it makes sense!

# great.  now let's make a column to indicate the year, making the final San Felipe dataset
san_felipe_long_final <- san_felipe_long %>% 
  mutate(
    is_mo_yr = !str_detect(month_year, "acum|total"),
    temp = if_else(is_mo_yr, month_year, NA)
  ) %>% 
  separate(temp, into = c("month", "year"), sep = "_", remove = FALSE) %>% 
  filter(is_topic %in% FALSE) %>% # we do not want the rows of just titles
  select(header_concept, concept, concept_full, month_year, month, year, value) %>% 
  mutate(municipio = "san_felipe") # before rbinding with other cities!
```

### Function to clean the files based on this file

```{r}
clean_hotel_occupations_data <- function(municipio) {
  
  # where to find the file
  file_path <- here::here(
    "/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/resaved_as_csv",
    paste0("DataturReportePublico_", municipio, "_resaved.csv")
  )
  
  # read raw data
  raw_data <- read_csv(file_path)
  
  # clean the column names, remove top rows, and tidy
  cleaned <- raw_data %>% 
    set_names(.[9, ]) %>%  
    slice(11:136) %>% # should all be the same because of the way it is structured
    janitor::clean_names()
  
  # clean and standardize the "concepto" column
  cleaned_concept <- cleaned %>%
    mutate(
      concept = str_to_lower(concepto),
      concept = str_trim(concept),
      concept = str_replace_all(concept, " ", "_")
    ) %>% 
    relocate(concept, .after = concepto) %>% # these two lines of code are not really necessary because the func will do it all at once
    select(-concepto)
  
  # identify topic headers and combine them with subcategories
  by_topic <- cleaned_concept %>% 
    mutate(
      is_topic = !str_detect(concept, "^[0-9]_estrellas|[1-5]_estrellas|total_categorias") & !is.na(concept)
    ) %>%
    mutate(header_concept = if_else(is_topic, concept, NA_character_)) %>% # use NA_character_ for this stuff just in case
    fill(header_concept) %>% 
    mutate(
      concept_full = if_else(concept != header_concept, paste0(header_concept, ":", concept), concept)
    )
  
  # pivot longer
  long <- by_topic %>% 
    pivot_longer(
      cols = 2:100, # adjust if more columns
      names_to = "month_year",
      values_to = "value"
    )
  
  # separate year and month, filter, and tag city
  final <- long %>% 
    mutate(
      is_mo_yr = !str_detect(month_year, "acum|total"),
      temp = if_else(is_mo_yr, month_year, NA_character_)
    ) %>% 
    separate(temp, into = c("month", "year"), sep = "_", remove = FALSE) %>% 
    filter(is_topic == FALSE) %>% 
    select(header_concept, concept, concept_full, month_year, month, year, value) %>% 
    mutate(municipio = str_to_lower(municipio))
  
  return(final)
}
```

Okay! Let's try it with Cabo San Lucas.

```{r}
cabo_san_lucas_final <- clean_hotel_occupations_data("Cabo_San_Lucas") # hmmm... there are weird crosses when resaved as UTF... "†".. but it only works as a UTF-8!

la_paz_final <- clean_hotel_occupations_data("La_Paz") # okay it is happening more than once.  Need to figure that out.  Come back tomorrow!
```

## COME BACK!

For all at once:

```{r}
all_cities <- c("San_Felipe", "Guaymas", "La_Paz"......)

all_data <- map_dfr(all_cities, clean_city_hotel_data)
```



Let's look only at "total_categorias" within categories:

```{r}
san_felipe_total_categorias <- san_felipe_long_final %>% 
  filter(concept == "total_categorias")
```




