---
title: 'OHI GoC 2025: Tourism & Recreation: Quantity Gapfilling'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: inline
editor_options: 
  chunk_output_type: inline
---

# Seeing if the number of employees INEGIs DENUE "Servicios de alojamiento temporal y de preparación de alimentos y bebidas" has a relationship with tourist arrivals (SECTUR Hotel Monitoring Program)

See ~TR/denue_prep.qmd for the data downloading and cleaning up to this point.

## Bring in Accommodations

```{r}
library(tidyverse)
library(magrittr)
library(here)
library(janitor)
library(tmaptools)
library(sf)
library(mapview)

alojamientos_2019_2024 <- st_read(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas/int","alojamientos_2019_2024_goc_10km.shp"))
```

## Bring in Total Visitors

```{r}

```










# Old exploring script

```{r}
library(tidyverse)
library(magrittr)
library(here)
library(janitor)
library(tmaptools)
library(sf)
library(mapview)

sp_1 <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue_establishments/denue_00_72_1_shp/conjunto_de_datos","denue_inegi_72_1.shp"))

sp_1_diccionario_de_datos <- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue_establishments/denue_00_72_1_shp/diccionario_de_datos","denue_diccionario_de_datos.csv"))

sp_1_goc <- sp_1 %>% 
  filter(entidad %in% c("Baja California","Baja California Sur","Jalisco","Nayarit","Sinaloa","Sonora"))

sp_1_goc[] <- lapply(sp_1_goc, function(x) {
  if (is.character(x)) iconv(x, from = "", to = "UTF-8", sub = "byte") else x
})

# sp_2_valid <- sp_2 %>% 
#   select(id, geometry)

mapview(sp_1_goc)
```

NOTE: `fecha_alta` is "Fecha en la que la unidad económica se integró al Directorio Estadístico Nacional de Unidades Económicas y al Registro Estadístico de Negocios de México (RENEM)."

Aka, in English, it is the date on which the economic unit was added to the National Statistical Directory of Economic Units and the Statistical Registry of Businesses in Mexico (RENEM).


```{r}
sp_2 <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue_establishments/denue_00_72_2_shp/conjunto_de_datos","denue_inegi_72_2.shp"), options = "ENCODING=UTF-8")

# sp_2_dropgeo <- sp_2 %>%
#   st_drop_geometry()
# 
# sp_2_sf <- sp_2_dropgeo %>%
#   st_as_sf(coords = c("longitud", "latitud"),
#            crs = 4326, # WGS84
#            remove = FALSE)
# 
# sapply(sp_2, function(col) {
#   if (is.character(col)) {
#     any(is.na(iconv(col, from = "UTF-8", to = "UTF-8")))
#   } else {
#     FALSE
#   }
# })

unique(sp_2$entidad)

sp_2_goc <- sp_2 %>% 
  filter(entidad %in% c("Baja California","Baja California Sur","Jalisco","Nayarit","Sinaloa","Sonora"))

sp_2_goc[] <- lapply(sp_2_goc, function(x) {
  if (is.character(x)) iconv(x, from = "", to = "UTF-8", sub = "byte") else x
})

# sp_2_valid <- sp_2 %>% 
#   select(id, geometry)

mapview(sp_2_goc)
```


```{r}
sp_1_2 <- st_union(sp_1_goc, sp_2_goc)

mapview(sp_1_goc) + mapview(sp_2_goc)
```


```{r}
denue_71 <- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue_establishments/denue_00_71_csv/conjunto_de_datos","denue_inegi_71_.csv")) %>% 
  mutate()

unique(denue_71$localidad) # \x needs to be replaced
unique(denue_71$entidad)
# [1] "Aguascalientes"                  "Baja California"                 "Baja California Sur"             "Campeche"                       
#  [5] "Coahuila de Zaragoza"            "Colima"                          "Chiapas"                         "Chihuahua"                      
#  [9] "Ciudad de M\xe9xico"             "Durango"                         "Guanajuato"                      "Guerrero"                       
# [13] "Hidalgo"                         "Jalisco"                         "M\xe9xico"                       "Michoac\xe1n de Ocampo"         
# [17] "Morelos"                         "Nayarit"                         "Nuevo Le\xf3n"                   "Oaxaca"                         
# [21] "Puebla"                          "Quer\xe9taro"                    "Quintana Roo"                    "San Luis Potos\xed"             
# [25] "Sinaloa"                         "Sonora"                          "Tabasco"                         "Tamaulipas"                     
# [29] "Tlaxcala"                        "Veracruz de Ignacio de la Llave" "Yucat\xe1n"                      "Zacatecas" 

denue_71_goc <- denue_71 %>% 
  filter(entidad %in% c("Baja California","Baja California Sur","Jalisco","Nayarit","Sinaloa","Sonora"))

denue_71_sf <- denue_71_goc %>%
  st_as_sf(coords = c("longitud", "latitud"),
           crs = 4326, # WGS84
           remove = FALSE)
mapview(denue_71_sf)


denue_71_sf[] <- lapply(denue_71_sf, function(x) {
  if (is.character(x)) iconv(x, from = "", to = "UTF-8", sub = "byte") else x
})

mapview(denue_71_sf) ## not as complete as using the shp files from INEGI
```

### 2024

```{r}
sp_1_2024 <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue_establishments/denue_00_72_1_1124_shp/conjunto_de_datos","denue_inegi_72_1.shp"))

sp_1_2024_goc <- sp_1_2024 %>% 
  filter(entidad %in% c("Baja California","Baja California Sur","Jalisco","Nayarit","Sinaloa","Sonora"))

sp_1_2024_goc[] <- lapply(sp_1_2024_goc, function(x) {
  if (is.character(x)) iconv(x, from = "", to = "UTF-8", sub = "byte") else x
})

mapview(sp_1_2024_goc)
```


### 2023

```{r}
sp_1_2023 <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue_establishments/denue_00_72_1_1123_shp/conjunto_de_datos","denue_inegi_72_1.shp"))

sp_1_2023_goc <- sp_1_2023 %>% 
  filter(entidad %in% c("Baja California","Baja California Sur","Jalisco","Nayarit","Sinaloa","Sonora"))

sp_1_2023_goc[] <- lapply(sp_1_2023_goc, function(x) {
  if (is.character(x)) iconv(x, from = "", to = "UTF-8", sub = "byte") else x
})

mapview(sp_1_2023_goc)
```

### 2022

```{r}
sp_1_2022 <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue_establishments/denue_00_72_1_1122_shp/conjunto_de_datos","denue_inegi_72_1.shp"))

unique(sp_1_2022$entidad)
#  [1] "AGUASCALIENTES"         "BAJA CALIFORNIA"        "CAMPECHE"               "BAJA CALIFORNIA SUR"   
#  [5] "COAHUILA DE ZARAGOZA"   "COLIMA"                 "CHIAPAS"                "CHIHUAHUA"             
#  [9] "CIUDAD DE M\xc9XICO"    "DURANGO"                "GUANAJUATO"             "GUERRERO"              
# [13] "HIDALGO"                "JALISCO"                "M\xc9XICO"              "MICHOAC\xc1N DE OCAMPO"
# [17] "MORELOS"                "NAYARIT"                "NUEVO LE\xd3N"          "OAXACA"                
# [21] "PUEBLA" 

sp_1_2022_goc <- sp_1_2022 %>% 
  filter(entidad %in% c("BAJA CALIFORNIA","BAJA CALIFORNIA SUR","JALISCO","NAYARIT")) ## sinaloa and sonora are likely in sp_2!!

sp_1_2022_goc[] <- lapply(sp_1_2022_goc, function(x) {
  if (is.character(x)) iconv(x, from = "", to = "UTF-8", sub = "byte") else x
})

mapview(sp_1_2022_goc)
```

### Checking the differences between the years downloaded to see if we need to download each year's data, or if we can just go based off of the most recent shapefiles.

```{r}
## check the ids of the establishments
length(unique(sp_1_goc$id)) # this is 2025: 90669
length(unique(sp_1_2024_goc$id)) # 89759
length(unique(sp_1_2023_goc$id)) # 89759
length(unique(sp_1_2022_goc$id)) # 80378
```

I think that they may have accidentally saved the 2023 shapefile as the 2023 shapefile, as there are establishments that were added to the National Statistical Directory of Economic Units and the Statistical Registry of Businesses in Mexico (RENEM) in 2024 in the 2023 shapefile.


```{r}
## check which establishments are new between 2022 and 2024, and make sure that the fecha_alta reflects their correct date of addition

length(setdiff(unique(sp_1_2024_goc$id), unique(sp_1_2022_goc$id))) # 39269

# 89759 - 80378 = 9381
## This does not really make sense. 

length(setdiff(unique(sp_1_2022_goc$id), unique(sp_1_2024_goc$id))) # 29888

## okay.  So this means that establishments are removed when they no longer exist, I believe.  
```

So, let's see which of the establishments are present in 2024 only, and which are present in 2022 only.

```{r}
## 2024 only

only_2024 <- setdiff(unique(sp_1_2024_goc$id), unique(sp_1_2022_goc$id))

sp_1_2024_goc_ONLY24 <-sp_1_2024_goc %>% 
  filter(id %in% only_2024)

range(sp_1_2024_goc_ONLY24$fecha_alta) # "2010-07" "2024-11"
unique(sp_1_2024_goc_ONLY24$fecha_alta)
#  [1] "2024-11" "2023-11" "2019-11" "2010-07" "2014-12" "2020-11" "2024-05" "2021-05" "2016-10" "2016-01" "2011-03"
# [12] "2021-11" "2013-07" "2020-04" "2022-11" "2012-06"

## why might there be establishments that were added to the National Statistical Directory of Economic Units and the Statistical Registry of Businesses in Mexico (RENEM) in 2010 in the 2024 shapefile, that are not in the 2022 shapefile?
```

After some thought, this is what I have determined:

1.  DENUE versions are cumulative but filtered by active status.
  Every DENUE release only includes active establishments at the time of publication.
  If an establishment was closed (temporarily or permanently) in 2022, it might be missing from the 2022 file, even if it has reopened or been reactivated by 2024.

2. Reactivated establishments keep their original `fecha_alta`.
  That is why there are fecha_altas as far back as 2010 in the 2024 file that were not present in the 2022 file: these businesses first entered RENEM in 2010, later disappeared from the active roster, and were likely revalidated by INEGI in 2024.

3. Differences between years do NOT equal purely new/closed businesses.
  Some IDs are newly added (new businesses).
  Some are reactivations.
  Some disappear because they closed, changed name or location, or failed verification.
  
SO, there can be establishments split/merged (1 --> 2 or 2 --> 1), occasional ID reassignments, and geographic filtering differences.  **I will need to merge the sp_1 and sp_2 for each year and compare again.** 

## Merge sp_1_goc and sp_2_goc

```{r}
names(sp_1_goc) == names(sp_2_goc) ## TRUE
st_crs(sp_1_goc) == st_crs(sp_2_goc) ## TRUE

common_cols <- intersect(names(sp_1_goc), names(sp_2_goc))
common_cols # just double checking before binding

## bind sp_1_2_goc, uses less memory than st_union() or rbind()
sp_1_2_goc <- bind_rows(sp_1_goc[, common_cols], sp_2_goc[, common_cols])
```

#### Bring in coastal population (we likely won't need this here, replace with TOURIST ARRIVALS)

```{r}
# ohi_rgn_pop_10km <- read_csv(here("general/OHI_rgn_population_10km.csv")) # now we have the number of individuals living within 10km of the coast for each state.
```
#### Get coastline buffered 10 km inland (thanks Mel!!)

```{r}
regions_marine <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp")) %>%
  filter(location == "marine")
regions_land <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp")) %>%
  filter(location == "land")

coast_buffer <- st_buffer(regions_marine, dist = 10000) %>%
  st_union() %>%
st_sf()

inland_10 <- st_intersection(coast_buffer, regions_land)
inland_10_state_region <- st_intersection(inland_10, states)
mapview(inland_10_state_region[16,])

# removed the weird ones by hand, probably should have done this in a more reproducible way!
inland_10_state_region <- inland_10_state_region[-c(3, 4, 10, 11, 13, 14),] %>%
st_make_valid()
  
i <- which(st_geometry_type(inland_10_state_region) == "GEOMETRYCOLLECTION")
g <- inland_10_state_region[i,]
poly_only <- st_collection_extract(g, c("POLYGON"))
poly_only <- st_cast(st_union(poly_only), "MULTIPOLYGON")
inland_10_state_region$geometry[i] <- poly_only

mapview(inland_10_state_region)
```

#### Isolate coastal service establishments

```{r}
## match the crs of both objects
st_crs(sp_1_2_goc) == st_crs(inland_10_state_region) ## FALSE

st_crs(sp_1_2_goc) # WGS84
st_crs(inland_10_state_region) # MEX ITRF2008 / LLC

sp_1_2_goc <- st_transform(sp_1_2_goc, st_crs(inland_10_state_region)) # transform to match crs
```

```{r}
## spatial join to only keep service establishments within 10km of coastline
sp_1_2_goc_coastal <- st_join(
  sp_1_2_goc,
  inland_10_state_region,
  join = st_intersects,
  left = FALSE
) ## note: this only keep the point geometries
```

```{r}
## check to make sure it looks good
mapview(inland_10_state_region) + mapview(sp_1_2_goc_coastal, col.regions = "red") ## looks great!!!
```

#### Summarize the number of service establishments by region and by fecha_alta

Ways to do this: 

- Track the rate at which service establishments are added to the coast (this does not seem good, because it is not necessarily good to have more and more establishments -- that could be potentially a negative!)
- Make a ratio of number of establishments per total tourist arrivals (total # of service establishments for a region for that year / # of tourist arrivals in the coastal region for that year) and then say that based on a historical reference point (maybe 2019, before covid), the ratio should either be maintained or increase to get a score of 100.

Let's start with the second option, and we can adjust if we determine there is a better way to do this. 

To clarify, the ratio for every year should include the number of service establishments with a fecha_alta EQUAL to that year and all previous years.  So for example:

> If we are evaluating the ratio for the year 2022, I would want to count all observations of establishments with a `fecha_alta` less than or equal to 2022.  For 2024, it would be all observations of establishments with a `fecha_alta` less than or equal to 2024.

My only concern about this is that it may not change and will only increase if we are using the same shapefile.  So we may need to download all year's shapefile's separately and then evaluate the ratio separately. 

So we would essentially download all year's shapefiles available on INEGI DENUE (ex: denue_00_72_2_1124_shp, denue_00_72_2_1123_shp, denue_00_72_2_1122_shp, denue_00_72_2_1121_shp) for both sp_1 and sp_2.  

Then, we would bind all of them together after using mutate() to give each merged sp_1_2_goc an identifier for the year it was from.  So the cleaned denue_00_72_1_1122_shp & denue_00_72_2_1122_shp would have a column that just has the year 2022 for every observation to indicate the `data_year`.

In the completely bound dataset, of all data_years, we can then summarize and determine the number of service establishments by region and data_year.  Then those are all divided by the number of tourist arrivals for that year per region.

We can then compare the ratios across different years. Does this sound right??








