---
title: 'OHI GoC 2025: Tourism & Recreation: Quantity Gapfilling'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: inline
editor_options: 
  chunk_output_type: inline
---

# Seeing if the number of employees INEGIs DENUE "Servicios de alojamiento temporal y de preparación de alimentos y bebidas" has a relationship with tourist arrivals (SECTUR Hotel Monitoring Program)

See ~TR/denue_prep.qmd for the data downloading and cleaning up to this point.

## Bring in Accommodations

We will need to clean it up, because the number of employees for each business is binned.

*This is what the per_ocu variable description says in the metadata:*
"This includes personnel directly employed by the company and external personnel supplied by another company, who worked for the economic unit, under its direction and control, and who covered at least one-third of the workday. This may include permanent staff, paid temporary staff, or unpaid temporary staff.
Economic units are classified by employee size ranges, which allows for the identification of the size of Economic Units based on the number of personnel they employ; that is, according to their total workforce. For the purposes of the Directory, Economic Units are grouped into employee size ranges or strata, based on the total number of employees reported by each unit, as shown in the following table.
Number of Employees
1 = 0 to 5
2 = 6 to 10
3 = 11 to 30
4 = 31 to 50
5 = 51 to 100
6 = 101 to 250
7 = 251 and above" 

So, we will take the midpoint of each range.

```{r}
library(tidyverse)
library(magrittr)
library(here)
library(janitor)
library(tmaptools)
library(sf)
library(mapview)

alojamientos_2019_2024_50km <- st_read(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas/int","alojamientos_2019_2024_goc_50km.shp"))

## see what the variable per_ocu looks like
unique(alojamientos_2019_2024_50km$per_ocu)

# [1] "0 a 5 personas"        "6 a 10 personas"       "11 a 30 personas"      "31 a 50 personas"      "101 a 250 personas"    "51 a 100 personas"     "251 y m<e1>s personas" "251 y más personas" 

## use a case_when to reclassify the number of people employed based on the midpoint of the range.  NOTE that I am assuming that you can never have 0 employees, so the first midpoint is between 1 and 5.
alojamientos_est_perocu <- alojamientos_2019_2024_50km %>% 
  mutate(
    employees_estimated = case_when(
      per_ocu == "0 a 5 personas" ~ 3,
      per_ocu == "6 a 10 personas" ~ 8,
      per_ocu == "11 a 30 personas" ~ 20.5,
      per_ocu == "31 a 50 personas" ~ 40.5,
      per_ocu == "51 a 100 personas" ~ 75.5,
      per_ocu == "101 a 250 personas" ~ 175.5,
      per_ocu %in% c("251 y m<e1>s personas", "251 y más personas") ~ 300,
      TRUE ~ NA_real_
    )
  )

mapview(alojamientos_est_perocu, zcol = "employees_estimated")

unique(alojamientos_est_perocu$municipio)

# when 10km:
#  [1] "Mexicali"                 "Ensenada"                 "Muleg<e9>"                "Los Cabos"                "La Paz"                   "Loreto"                   "Puerto Vallarta"          "Cabo Corrientes"         
#  [9] "Bah<ed>a de Banderas"     "San Blas"                 "Compostela"               "Tecuala"                  "Puerto Pe<f1>asco"        "San Luis R<ed>o Colorado" "Hermosillo"               "Pitiquito"               
# [17] "Guaymas"                  "Empalme"                  "Benito Ju<e1>rez"         "Huatabampo"               "Elota"                    "Mazatl<e1>n"              "Escuinapa"                "San Ignacio"             
# [25] "Ahome"                    "Navolato"                 "Culiac<e1>n"              "Guasave"                  "Angostura"                "Rosario"                  "San Quint<ed>n"           "San Felipe"              
# [33] "San Quintín"              "Mulegé"                   "Bahía de Banderas"        "Puerto Peñasco"           "San Luis Río Colorado"    "Benito Juárez"            "Mazatlán"                 "Culiacán"                
# [41] "Caborca" 

# when 50km:
#  [1] "Mexicali"                   "Ensenada"                   "Muleg<e9>"                  "Comond<fa>"                
#  [5] "Los Cabos"                  "La Paz"                     "Loreto"                     "Puerto Vallarta"           
#  [9] "Talpa de Allende"           "Tomatl<e1>n"                "San Sebasti<e1>n del Oeste" "Cabo Corrientes"           
# [13] "Mascota"                    "Bah<ed>a de Banderas"       "Tepic"                      "San Blas"                  
# [17] "Compostela"                 "Acaponeta"                  "Ru<ed>z"                    "Xalisco"                   
# [21] "Tuxpan"                     "Tecuala"                    "Santiago Ixcuintla"         "Rosamorada"                
# [25] "Huajicori"                  "San Pedro Lagunillas"       "Puerto Pe<f1>asco"          "San Luis R<ed>o Colorado"  
# [29] "Hermosillo"                 "Caborca"                    "Pitiquito"                  "Navojoa"                   
# [33] "Cajeme"                     "Guaymas"                    "Huatabampo"                 "Etchojoa"                  
# [37] "Empalme"                    "Benito Ju<e1>rez"           "Elota"                      "Mazatl<e1>n"               
# [41] "Navolato"                   "Ahome"                      "Culiac<e1>n"                "El Fuerte"                 
# [45] "Escuinapa"                  "San Ignacio"                "Guasave"                    "Rosario"                   
# [49] "Mocorito"                   "Salvador Alvarado"          "Concordia"                  "Angostura"                 
# [53] "San Quint<ed>n"             "San Felipe"                 "San Quintín"                "Mulegé"                    
# [57] "Comondú"                    "Tomatlán"                   "San Sebastián del Oeste"    "Bahía de Banderas"         
# [61] "Ruíz"                       "Puerto Peñasco"             "San Luis Río Colorado"      "Benito Juárez"             
# [65] "Culiacán"                   "Mazatlán"    

unique(alojamientos_est_perocu$localidad)

# when 10km:
#  [1] "San Felipe"                            "Villas de Cortez"                      "Campo Beluga"                          "Campo Daggets"                         "Campo Sacrificio"                     
#   [6] "Punta Final (Villa Mar y Sol)"         "Campo Raquel y Larry"                  "Punta Bufeo [Campo Tur<ed>stico]"      "Bah<ed>a de los <c1>ngeles"            "La Ventana [Campo]"                   
#  [11] "Puerto San Francisquito"               "Heroica Muleg<e9>"                     "Santa Rosal<ed>a"                      "Cabo San Lucas"                        "Buenavista"                           
#  [16] "San Jos<e9> del Cabo"                  "La Cuadra Regina"                      "La Paz"                                "Chametla"                              "La Ribera"                            
#  [21] "El Sargento"                           "Puerto de Pichilingue"                 "Loreto"                                "La Ventana"                            "El Cardonal"                          
#  [26] "Nopol<f3>"                             "Buenos Aires"                          "Los Barriles"                          "Punta Pescadero"                       "Tripu<ed> [Trailer Park]"             
#  [31] "Boca de la Vinorama"                   "Danzante Bay"                          "Punta Prieta"                          "Ninguno"                               "El Leonero"                           
#  [36] "Puerto Vallarta"                       "Sun Resorts"                           "Mismaloya"                             "Yelapa"                                "Las Juntas y los Veranos"             
#  [41] "Le<f3>n de la Sierra"                  "La Quebrada"                           "Ixtapa"                                "Las Juntas"                            "Playa Gemelas"                        
#  [46] "Sayulita"                              "Nuevo Vallarta"                        "Guadalupe Victoria"                    "Rinc<f3>n de Guayabitos"               "Bucer<ed>as"                          
#  [51] "Aticama"                               "Cruz de Huanacaxtle"                   "El Porvenir"                           "La Pe<f1>ita de Jaltemba"              "Los Ayala"                            
#  [56] "Platanitos"                            "Lo de Marcos"                          "Chacala"                               "La Palma"                              "San Blas"                             
#  [61] "Matanch<e9>n (La Aguada)"              "San Francisco"                         "Corral del Risco (Punta de Mita)"      "Lidia [Bungalows]"                     "Playa de los Cocos"                   
#  [66] "La Joya"                               "P<e1>tzcuaro"                          "Monte<f3>n"                            "Punta Raza"                            "Las Varas"                            
#  [71] "La Manzanilla"                         "Flamingos"                             "Santa Cruz de Miramar"                 "Tondoroque"                            "Desarrollo Punta Mita"                
#  [76] "Bah<ed>a de Matanch<e9>n"              "El Crucero de Chacala"                 "Punta Negra (Careyeros)"               "San Jos<e9> del Valle"                 "Para<ed>so Sereno"                    
#  [81] "Mezcales"                              "Playa Escondida"                       "Universo [Colonia]"                    "Playa Novillero"                       "Grand Palladium Vallarta [Hotel]"     
#  [86] "San Vicente"                           "Las Jarretaderas"                      "San Clemente de Lima"                  "Arboledas [Bungalows]"                 "La Tranquila [Condominios]"           
#  [91] "Imanta"                                "Los Arroyos Verdes"                    "Dos Palmeras (Casa Pel<ed>canos)"      "El Tonino [Bloquera]"                  "Punta Monterrey"                      
#  [96] "Rancho Banderas"                       "Papelillos [Hotel]"                    "Puerto Pe<f1>asco"                     "Golfo de Santa Clara"                  "Bah<ed>a la Choya"                    
# [101] "Playa Arenosa"                         "Bah<ed>a de Kino"                      "Puerto Libertad"                       "Mayan Palace Hotel"                    "Playa Miramar"                        
# [106] "Las Conchas"                           "Heroica Guaymas"                       "San Carlos (San Carlos Nuevo Guaymas)" "Condominios Playa Blanca"              "Empalme"                              
# [111] "Villa Ju<e1>rez"                       "Agua Blanca"                           "Las Bocas"                             "Costa del Mar"                         "La Cruz"                              
# [116] "Mazatl<e1>n"                           "Color Marino"                          "Dimas (Estaci<f3>n Dimas)"             "Teacap<e1>n"                           "Topolobampo"                          
# [121] "Altata"                                "Isla del Bosque"                       "Eldorado"                              "El Saucito"                            "Celestino Gazca Villase<f1>or"        
# [126] "Las Glorias"                           "Fraccionamiento Isla Para<ed>so"       "La Reforma"                            "Gustavo D<ed>az Ordaz (El Carrizo)"    "La Escondida"                         
# [131] "Agua Verde"                            "Localidad Econ<f3>mica"                "Rancho Girolando"                      "El Patio"                              "Ejido Plan Nacional Agrario"          
# [136] "Alfonsinas"                            "Tailwind"                              "El Encuentro"                          "Los Mochis"                            "Zacualpan"                            
# [141] "Desarrollo Puerto Bah<ed>a"            "Litib<fa>"                             "Bahía de los Ángeles"                  "Punta Bufeo [Campo Turístico]"         "Heroica Mulegé"                       
# [146] "Santa Rosalía"                         "San José del Cabo"                     "Tripuí [Trailer Park]"                 "San Román"                             "Nopoló"                               
# [151] "Costa Vida Vallarta"                   "Rincón de Guayabitos"                  "Bucerías"                              "Bahía de Matanchén"                    "La Peñita de Jaltemba"                
# [156] "San José del Valle"                    "Dos Palmeras (Casa Pelícanos)"         "Pátzcuaro"                             "Desarrollo Puerto Bahía"               "Litibú"                               
# [161] "Puerto Peñasco"                        "Bahía de Kino"                         "Villa Juárez"                          "Mazatlán"                              "Campo Santa Fe"                       
# [166] "Teacapán"                              "Dimas (Estación Dimas)"                "Motel Popeye"                          "Gustavo Díaz Ordaz (El Carrizo)"       "Playa del Sol"                        
# [171] "Pap<e1> Fern<e1>ndez [Campo]"          "Los Vientos"                           "Campo Lobo"                            "Cabo Golf Cars"                        "Cuesta Blanca"                        
# [176] "Las <c1>nimas de Quimixto"             "Lienzo Charro"                         "Mezcalitos"                            "Para<ed>so Escondido [Colonia]"        "Jalcocot<e1>n"                        
# [181] "Las Tijerillas"                        "Luis Pe<f1>a"                          "Hip<f3>lito Gonz<e1>lez L<f3>pez"      "Limoncito Hills"                       "La Playa Estates"                     
# [186] "Haixa (Litib<fa>)"                     "Playa Encanto"                         "Puerto Lobos"                          "Mayan Lakes"                           "Playa Huatabampito"                   
# [191] "Coch<f3>rit" 

# when 50km:
#   [1] "San Felipe"                                      "Villas de Cortez"                               
#   [3] "Campo Beluga"                                    "Campo Daggets"                                  
#   [5] "Campo Sacrificio"                                "Punta Final (Villa Mar y Sol)"                  
#   [7] "Campo Raquel y Larry"                            "Punta Bufeo [Campo Tur<ed>stico]"               
#   [9] "Bah<ed>a de los <c1>ngeles"                      "La Ventana [Campo]"                             
#  [11] "Catavi<f1>a"                                     "Puerto San Francisquito"                        
#  [13] "Rancho Escondido"                                "Piedra Blanca"                                  
#  [15] "Heroica Muleg<e9>"                               "Santa Rosal<ed>a"                               
#  [17] "San Isidro"                                      "San Lino"                                       
#  [19] "Cabo San Lucas"                                  "Buenavista"                                     
#  [21] "San Jos<e9> del Cabo"                            "La Cuadra Regina"                               
#  [23] "La Paz"                                          "Chametla"                                       
#  [25] "El Triunfo"                                      "La Acacia"                                      
#  [27] "La Ribera"                                       "El Sargento"                                    
#  [29] "Puerto de Pichilingue"                           "Loreto"                                         
#  [31] "San Miguel de Comond<fa>"                        "La Ventana"                                     
#  [33] "El Cardonal"                                     "Nopol<f3>"                                      
#  [35] "Buenos Aires"                                    "Los Barriles"                                   
#  [37] "Punta Pescadero"                                 "Tripu<ed> [Trailer Park]"                       
#  [39] "Boca de la Vinorama"                             "Danzante Bay"                                   
#  [41] "Punta Prieta"                                    "Ninguno"                                        
#  [43] "El Leonero"                                      "Puerto Vallarta"                                
#  [45] "Talpa de Allende"                                "La Cruz de Loreto"                              
#  [47] "Sun Resorts"                                     "San Sebasti<e1>n del Oeste"                     
#  [49] "Mismaloya"                                       "Yelapa"                                         
#  [51] "Las Juntas y los Veranos"                        "Mascota"                                        
#  [53] "El Tuito"                                        "Puerta de Enmedio"                              
#  [55] "Le<f3>n de la Sierra"                            "La Quebrada"                                    
#  [57] "Ixtapa"                                          "San Nicol<e1>s (La Troje)"                      
#  [59] "La Puntilla de Maito"                            "Las Juntas"                                     
#  [61] "Playa Gemelas"                                   "Sayulita"                                       
#  [63] "Nuevo Vallarta"                                  "Tepic"                                          
#  [65] "Guadalupe Victoria"                              "Rinc<f3>n de Guayabitos"                        
#  [67] "Bucer<ed>as"                                     "Aticama"                                        
#  [69] "Cruz de Huanacaxtle"                             "El Porvenir"                                    
#  [71] "La Pe<f1>ita de Jaltemba"                        "Los Ayala"                                      
#  [73] "Platanitos"                                      "Lo de Marcos"                                   
#  [75] "Chacala"                                         "La Palma"                                       
#  [77] "San Blas"                                        "Matanch<e9>n (La Aguada)"                       
#  [79] "San Francisco"                                   "Corral del Risco (Punta de Mita)"               
#  [81] "Lidia [Bungalows]"                               "Playa de los Cocos"                             
#  [83] "La Joya"                                         "P<e1>tzcuaro"                                   
#  [85] "Monte<f3>n"                                      "Punta Raza"                                     
#  [87] "Las Varas"                                       "Acaponeta"                                      
#  [89] "La Manzanilla"                                   "Flamingos"                                      
#  [91] "Santa Cruz de Miramar"                           "Tondoroque"                                     
#  [93] "Desarrollo Punta Mita"                           "Compostela"                                     
#  [95] "Ru<ed>z"                                         "San Juan de Abajo"                              
#  [97] "Xalisco"                                         "Bah<ed>a de Matanch<e9>n"                       
#  [99] "Tuxpan"                                          "El Crucero de Chacala"                          
# [101] "Tecuala"                                         "Punta Negra (Careyeros)"                        
# [103] "Santiago Ixcuintla"                              "Rosamorada"                                     
# [105] "San Jos<e9> del Valle"                           "Para<ed>so Sereno"                              
# [107] "Huajicori"                                       "Mezcales"                                       
# [109] "Playa Escondida"                                 "Villa Hidalgo"                                  
# [111] "Universo [Colonia]"                              "Playa Novillero"                                
# [113] "Grand Palladium Vallarta [Hotel]"                "San Vicente"                                    
# [115] "Mexcaltit<e1>n de Uribe"                         "Las Jarretaderas"                               
# [117] "L<e1>zaro C<e1>rdenas [Colonia]"                 "San Clemente de Lima"                           
# [119] "Arboledas [Bungalows]"                           "La Tranquila [Condominios]"                     
# [121] "Imanta"                                          "Los Arroyos Verdes"                             
# [123] "Dos Palmeras (Casa Pel<ed>canos)"                "San Cayetano"                                   
# [125] "El Tonino [Bloquera]"                            "San Pedro Lagunillas"                           
# [127] "Punta Monterrey"                                 "Rancho Banderas"                                
# [129] "Papelillos [Hotel]"                              "Puerto Pe<f1>asco"                              
# [131] "Golfo de Santa Clara"                            "Bah<ed>a la Choya"                              
# [133] "Playa Arenosa"                                   "Bah<ed>a de Kino"                               
# [135] "Miguel Alem<e1>n (La Doce)"                      "Plutarco El<ed>as Calles (La Y Griega)"         
# [137] "Puerto Libertad"                                 "Mayan Palace Hotel"                             
# [139] "Playa Miramar"                                   "Las Conchas"                                    
# [141] "Navojoa"                                         "Ciudad Obreg<f3>n"                              
# [143] "Heroica Guaymas"                                 "San Carlos (San Carlos Nuevo Guaymas)"          
# [145] "Condominios Playa Blanca"                        "Huatabampo"                                     
# [147] "Bacobampo"                                       "Empalme"                                        
# [149] "Esperanza"                                       "Villa Ju<e1>rez"                                
# [151] "Agua Blanca"                                     "Pueblo Yaqui"                                   
# [153] "Los Bahuises"                                    "V<ed>cam (Switch)"                              
# [155] "Las Bocas"                                       "Costa del Mar"                                  
# [157] "La Cruz"                                         "Campo el Salto"                                 
# [159] "Mazatl<e1>n"                                     "San Marcos"                                     
# [161] "Navolato"                                        "Los Mochis"                                     
# [163] "Culiac<e1>n Rosales"                             "Color Marino"                                   
# [165] "Dimas (Estaci<f3>n Dimas)"                       "Adolfo Ruiz Cortines"                           
# [167] "Leopoldo S<e1>nchez Celis"                       "Escuinapa de Hidalgo"                           
# [169] "San Ignacio"                                     "La Papalota"                                    
# [171] "Teacap<e1>n"                                     "El Rosario"                                     
# [173] "Mocorito"                                        "El Fuerte"                                      
# [175] "Topolobampo"                                     "Villa Uni<f3>n"                                 
# [177] "Altata"                                          "Guam<fa>chil"                                   
# [179] "Isla del Bosque"                                 "Guasave"                                        
# [181] "Juan Jos<e9> R<ed>os"                            "Quil<e1>"                                       
# [183] "Eldorado"                                        "El Saucito"                                     
# [185] "El Batall<f3>n"                                  "Pericos"                                        
# [187] "Celestino Gazca Villase<f1>or"                   "Las Glorias"                                    
# [189] "Costa Rica"                                      "Concordia"                                      
# [191] "Fraccionamiento Isla Para<ed>so"                 "Sistema de Producci<f3>n Intensiva del Noroeste"
# [193] "La Reforma"                                      "Gustavo D<ed>az Ordaz (El Carrizo)"             
# [195] "Licenciado Benito Ju<e1>rez (Campo Gobierno)"    "La Escondida"                                   
# [197] "Santa Eduwiges"                                  "Caimanero"                                      
# [199] "Callejones de Guasavito"                         "Agua Verde"                                     
# [201] "Le<f3>n Fonseca (Estaci<f3>n Verdura)"           "Nestle Culiac<e1>n"                             
# [203] "Localidad Econ<f3>mica"                          "Rancho Girolando"                               
# [205] "Ninguno [Motel]"                                 "El Patio"                                       
# [207] "Gabriel Leyva Solano (Benito Ju<e1>rez)"         "Ejido Plan Nacional Agrario"                    
# [209] "Alfonsinas"                                      "Tailwind"                                       
# [211] "Providencia"                                     "El Encuentro"                                   
# [213] "Zacualpan"                                       "Ruiz"                                           
# [215] "Desarrollo Puerto Bah<ed>a"                      "Litib<fa>"                                      
# [217] "Heroica Ciudad Concordia"                        "Productos Enlatados Madre Tierra"               
# [219] "Ferretera y Torniller<ed>a Zapata [FETOZA]"      "Walamito"                                       
# [221] "Bahía de los Ángeles"                            "Cataviña"                                       
# [223] "Punta Bufeo [Campo Turístico]"                   "Heroica Mulegé"                                 
# [225] "Santa Rosalía"                                   "San José del Cabo"                              
# [227] "Tripuí [Trailer Park]"                           "San Román"                                      
# [229] "Nopoló"                                          "San Sebastián del Oeste"                        
# [231] "San Nicolás (La Troje)"                          "Costa Vida Vallarta"                            
# [233] "Rincón de Guayabitos"                            "Bucerías"                                       
# [235] "Bahía de Matanchén"                              "La Peñita de Jaltemba"                          
# [237] "San José del Valle"                              "Dos Palmeras (Casa Pelícanos)"                  
# [239] "Pátzcuaro"                                       "Potrero las Lomas"                              
# [241] "Desarrollo Puerto Bahía"                         "Litibú"                                         
# [243] "Mexcaltitán de Uribe"                            "Lázaro Cárdenas [Colonia]"                      
# [245] "Los Tabachines"                                  "Puerto Peñasco"                                 
# [247] "Bahía de Kino"                                   "Miguel Alemán (La Doce)"                        
# [249] "Plutarco Elías Calles (La Y Griega)"             "Villa Juárez"                                   
# [251] "Ciudad Obregón"                                  "Vícam (Switch)"                                 
# [253] "Culiacán Rosales"                                "Mazatlán"                                       
# [255] "Guamúchil"                                       "Sistema de Producción Intensiva del Noroeste"   
# [257] "Campo Santa Fe"                                  "El Batallón"                                    
# [259] "Teacapán"                                        "La Caída"                                       
# [261] "Villa Unión"                                     "Dimas (Estación Dimas)"                         
# [263] "Quilá"                                           "Juan José Ríos"                                 
# [265] "Motel Popeye"                                    "Licenciado Benito Juárez (Campo Gobierno)"      
# [267] "Gustavo Díaz Ordaz (El Carrizo)"                 "Nestle Culiacán"                                
# [269] "León Fonseca (Estación Verdura)"                 "Motel Paradise"                                 
# [271] "Playa del Sol"                                   "Pap<e1> Fern<e1>ndez [Campo]"                   
# [273] "Los Vientos"                                     "Campo Lobo"                                     
# [275] "Cabo Golf Cars"                                  "Club Campestre el Madrigal"                     
# [277] "Cuesta Blanca"                                   "San Dionisio"                                   
# [279] "Las <c1>nimas de Quimixto"                       "Lienzo Charro"                                  
# [281] "Mezcalitos"                                      "Para<ed>so Escondido [Colonia]"                 
# [283] "Jalcocot<e1>n"                                   "Las Tijerillas"                                 
# [285] "Luis Pe<f1>a"                                    "Hip<f3>lito Gonz<e1>lez L<f3>pez"               
# [287] "Limoncito Hills"                                 "La Playa Estates"                               
# [289] "La Cantera"                                      "Haixa (Litib<fa>)"                              
# [291] "Playa Encanto"                                   "Puerto Lobos"                                   
# [293] "Mayan Lakes"                                     "Playa Huatabampito"                             
# [295] "Coch<f3>rit"                                     "Anglers Inn [Bill Chapman Jr.]"                 
# [297] "La Ca<ed>da"                                     "San Pedro"                                      
# [299] "El Bordo"                                        "TUMSA [Transportes]"                            
# [301] "El Diez"                                         "Cortijo la Piedra"  
```

```{r}
# aggregate the accommodation employee data by municipality and year
employees_by_muni_yr_50 <- alojamientos_est_perocu %>% 
  mutate(municipio = case_when(
    municipio == "Mexicali" ~ "San Felipe",
    TRUE ~ municipio
  )) %>% 
  group_by(municipio, estado, data_year, rgn_id) %>% 
  summarize(
    total_employees = sum(employees_estimated, na.rm = TRUE),
    num_businesses = n(),
    avg_employees_per_business = mean(employees_estimated, na.rm = TRUE),
    .groups = "drop"
  )
```


## Bring in Total Visitors

These shapefiles were written out from `~TR/tr_quantity.qmd`.

```{r}
## from 2018 - 2025
dom_goc <- st_read(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/hotel_monitoring_int","domestic_arrivals_sectur_goc.shp"))

int_goc <- st_read(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/hotel_monitoring_int","international_arrivals_sectur_goc.shp"))

## join back together to have both spatial attributes and the number of visitors
all_arrivals <- bind_rows(dom_goc, int_goc) 

total_arrivals_yr_muni <- all_arrivals %>% 
  group_by(year, municip, estado, rgn_id) %>% 
  summarize(arrivals = sum(value))

mapview(total_arrivals_yr_muni, zcol = "arrivals")
```


## Merge the datasets

First, clean them both up:

```{r}
## look for differences

unique(employees_by_muni_yr$municipio)
# [1] "Ahome"                    "Angostura"                "Bah<ed>a de Banderas"     "Bahía de Banderas"        "Benito Ju<e1>rez"         "Benito Juárez"            "Cabo Corrientes"          "Caborca"                 
#  [9] "Compostela"               "Culiac<e1>n"              "Culiacán"                 "Elota"                    "Empalme"                  "Ensenada"                 "Escuinapa"                "Guasave"                 
# [17] "Guaymas"                  "Hermosillo"               "Huatabampo"               "La Paz"                   "Loreto"                   "Los Cabos"                "Mazatl<e1>n"              "Mazatlán"                
# [25] "Muleg<e9>"                "Mulegé"                   "Navolato"                 "Pitiquito"                "Puerto Pe<f1>asco"        "Puerto Peñasco"           "Puerto Vallarta"          "Rosario"                 
# [33] "San Blas"                 "San Felipe"               "San Ignacio"              "San Luis R<ed>o Colorado" "San Luis Río Colorado"    "San Quint<ed>n"           "San Quintín"              "Tecuala"               



unique(total_arrivals_yr_muni$municip)
#  [1] "cabo_san_lucas"          "culiacan"                "el_fuerte"               "el_rosario"              "escuinapa"               "guaymas"                 "hermosillo"              "la_paz"                  "loreto"                 
# [10] "los_mochis"              "mazatlan"                "puerto_vallarta"         "san_felipe"              "san_jose_del_cabo"       "zona_corredor_los_cabos"
```
```{r}
## for 50km
employees_by_muni_yr_join_50 <- employees_by_muni_yr_50 %>%
  st_drop_geometry() %>% 
  mutate(
    municipio_clean = municipio %>%
      str_replace_all("Bah<ed>a", "Bahía") %>%
      str_replace_all("Ju<e1>rez", "Juárez") %>%
      str_replace_all("Culiac<e1>n", "Culiacán") %>%
      str_replace_all("Mazatl<e1>n", "Mazatlán") %>%
      str_replace_all("Muleg<e9>", "Mulegé") %>%
      str_replace_all("Pe<f1>asco", "Peñasco") %>%
      str_replace_all("R<ed>o", "Río") %>%
      str_replace_all("Ru<ed>z", "Ruíz") %>%
      str_replace_all("Quint<ed>n", "Quintín") %>%
      str_replace_all("Comond<fa>", "Comondú") %>%
      str_replace_all("San Sebasti<e1>n del Oeste", "San Sebastián del Oeste") %>%
      str_replace_all("Tomatl<e1>n", "Tomatlán") %>%
      str_replace_all("San Luis R<ed>o Colorado", "San Luis Río Colorado") %>%
      str_replace_all("<fa>", "ú") %>%
      str_replace_all("<e1>", "á") %>%
      str_replace_all("<ed>", "í") %>%
      str_replace_all("<f1>", "ñ") %>%
      str_replace_all("<e9>", "é") %>%
      str_to_lower() %>%
      str_replace_all(" ", "_"),
    
    estado = estado %>% 
      str_to_lower() %>%
      str_replace_all(" ", "_") %>% 
      ifelse(municipio_clean == "puerto_vallarta", "jalisco", .),
    
    data_year = as.numeric(data_year)
  )

```

```{r}
### for 10km
## denue accommodation employees
employees_by_muni_yr_join <- employees_by_muni_yr %>%
  st_drop_geometry() %>% 
  mutate(
    municipio_clean = municipio %>%
      str_replace_all("Bah<ed>a", "Bahía") %>%
      str_replace_all("Ju<e1>rez", "Juárez") %>%
      str_replace_all("Culiac<e1>n", "Culiacán") %>%
      str_replace_all("Mazatl<e1>n", "Mazatlán") %>%
      str_replace_all("Muleg<e9>", "Mulegé") %>%
      str_replace_all("Pe<f1>asco", "Peñasco") %>%
      str_replace_all("R<ed>o", "Río") %>%
      str_replace_all("Quint<ed>n", "Quintín") %>%
      
      # make lower case and remove spaces to match the arrivals data
      str_to_lower() %>%
      str_replace_all(" ", "_"),
    estado = estado %>% 
      str_to_lower() %>%
      str_replace_all(" ", "_"),
    data_year = as.numeric(data_year)
  )

unique(employees_by_muni_yr_join$municipio_clean)

## arrivals data tweaks for join
total_arrivals_aggregated_join <- total_arrivals_yr_muni %>%
  st_drop_geometry() %>% 
  mutate(
    municipio_clean = case_when(
      municip %in% c("cabo_san_lucas", "san_jose_del_cabo", "zona_corredor_los_cabos") ~ "los_cabos",
      municip == "los_mochis" ~ "ahome",
      municip == "el_rosario" ~ "rosario",
      municip == "culiacan" ~ "culiacán",
      municip == "mazatlan" ~ "mazatlán",
      TRUE ~ municip
    ),
    estado = estado %>% 
      str_to_lower() %>%
      str_replace_all(" ", "_"),
    year = as.numeric(year)
  ) %>%
  # reaggregate with changes
  group_by(year, municipio_clean, estado, rgn_id) %>%
  summarize(arrivals = sum(arrivals), .groups = "drop")

unique(total_arrivals_aggregated_join$municipio_clean)
```

### See the differences now in municipios

```{r}
# what the arrivals data has but the employment data does not
setdiff(unique(total_arrivals_aggregated_join$municipio_clean), unique(employees_by_muni_yr_join_50$municipio_clean))
# none!!!

# what DENUE data has but hotel arrivals data does not
setdiff(unique(employees_by_muni_yr_join$municipio_clean), unique(total_arrivals_aggregated_join$municipio_clean))
#  [1] "angostura"             "bahía_de_banderas"     "benito_juárez"         "cabo_corrientes"       "caborca"              
#  [6] "compostela"            "elota"                 "empalme"               "ensenada"              "guasave"              
# [11] "huatabampo"            "mulegé"                "navolato"              "pitiquito"             "puerto_peñasco"       
# [16] "san_blas"              "san_ignacio"           "san_luis_río_colorado" "san_quintín"           "tecuala"          
```

### Join the arrivals and accommodation employees data

```{r}
## join the datasets
combined_arrivals_employees_data <- total_arrivals_aggregated_join %>%
  left_join(employees_by_muni_yr_join_50, by = c("municipio_clean" = "municipio_clean", "year" = "data_year", "rgn_id" = "rgn_id", "estado" = "estado")) %>% 
  filter(year %in% 2019:2024)

# combined_arrivals_employees_data %>%
#   group_by(municipio_clean) %>%
#   summarize(
#     has_arrivals = sum(!is.na(arrivals)) > 0,
#     has_employees = sum(!is.na(total_employees)) > 0,
#     n_years = n()
#   ) %>%
#   print(n = Inf)
```


## Visitor and Accommodation Employees Relationship

```{r}
model1 <- lm(arrivals ~ total_employees, data = combined_arrivals_employees_data)
summary(model1)

library(ggplot2)

ggplot(combined_arrivals_employees_data, aes(x = total_employees, y = arrivals)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_minimal() +
  labs(
    title = "Total arrivals (domestic + international) vs Total accommodation employees",
    x = "Total employees",
    y = "Total arrivals"
  )


```

```{r}
## just to check
model2 <- lm(log(arrivals) ~ log(total_employees), data = combined_arrivals_employees_data)
summary(model2)
```

With municipio effects:
```{r}
model3 <- lm(arrivals ~ total_employees + municipio_clean, data = combined_arrivals_employees_data)
summary(model3)
```

```{r}
model4 <- lm(arrivals ~ total_employees + estado, data = combined_arrivals_employees_data)
summary(model4)
```
```{r}
AIC(model1) # 1983.689
AIC(model4) # 1981.345

BIC(model1) # 1990.348
BIC(model4) # 1996.881
```



# Gapfilling tourist arrivals 






We will be using model3 for any tourist centers that have gaps in certain years, but at least have some data.  For municipalities that have DENUE hotel employee data but not any SECTUR arrivals data/tourist centers, we will use model1.

```{r}
## use predict()
```


```{r}
## we will be prepping the df `combined_arrivals_employees_data` 
combined_arrivals_employees_data

combined_arrivals_employees_data_full <- full_join(employees_by_muni_yr_join_50, total_arrivals_aggregated_join, by = c("municipio_clean" = "municipio_clean", "data_year" = "year", "rgn_id" = "rgn_id", "estado" = "estado")) 
```

```{r}
## use predict() to fill arrivals by employees

filled_combined_arrivals_employees_data <- combined_arrivals_employees_data_full %>%
  mutate(
    arrivals_filled = ifelse(
      is.na(arrivals) & !is.na(total_employees),
      predict(model1, newdata = combined_arrivals_employees_data),
      arrivals
    )
  )

plot(model1, which = 1)  # residuals vs fitted
plot(model1, which = 2)  # QQ plot
```


```{r}
## make a vector of all municipios that can be filled with model3
munis_model3 <- unique(combined_arrivals_employees_data$municipio_clean)

## fill it
df_model3 <- combined_arrivals_employees_data_full %>%
  filter(municipio_clean %in% munis_model3)

df_model1 <- combined_arrivals_employees_data_full %>%
  filter(!municipio_clean %in% munis_model3)
```


```{r}
## fill using model3
df_model3_filled <- df_model3 %>%
  filter(data_year %in% 2019:2024) %>% 
  mutate(
    arrivals_filled_model3 = ifelse(
      is.na(arrivals) & !is.na(total_employees),
      predict(model3, newdata = df_model3),
      arrivals
    )
  )
```


```{r}
## fill using model 1
df_model1_filled <- df_model1 %>%
  filter(data_year %in% 2019:2024) %>% 
  mutate(
    arrivals_filled_model1 = ifelse(
      is.na(arrivals) & !is.na(total_employees),
      predict(model1, newdata = df_model1),
      arrivals
    )
  )
```

```{r}
## combine the two datasets
filled_combined <- bind_rows(
  df_model3_filled,
  df_model1_filled
) %>% 
  mutate(is_filled = ifelse(is.na(arrivals), "1", "0"))
```


```{r}
arrivals_employees_filled <- filled_combined %>% 
  mutate(arrivals_merged = case_when(
    is.na(arrivals) & !is.na(arrivals_filled_model3) ~ arrivals_filled_model3,
    is.na(arrivals) & is.na(arrivals_filled_model3) & !is.na(arrivals_filled_model1) ~ arrivals_filled_model1,
    TRUE ~ arrivals
  ))

library(plotly)

plot_data <- arrivals_employees_filled %>% 
  arrange(data_year)

plot_ly(
  data = plot_data,
  x = ~data_year,
  y = ~arrivals_merged,
  color = ~municipio_clean,
  type = 'scatter',
  mode = 'lines+markers'
) %>%
  layout(
    title = "Arrivals (Merged) Over Time by Municipio",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Arrivals")
  )

```

Check it diagnostically:

```{r}
ggplot(df_model3_filled, aes(x = arrivals, y = arrivals_filled_model3)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  theme_minimal() +
  labs(
    title = "Model3 Diagnostic: Actual vs Predicted Arrivals",
    x = "Actual arrivals",
    y = "Predicted arrivals (model3)"
  )
```
```{r}
ggplot(df_model1_filled, aes(x = arrivals, y = arrivals_filled_model1)) + ## there are no arrivals in this case, all need to be predicted.
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  theme_minimal() +
  labs(
    title = "Model1 Diagnostic: Actual vs Predicted Arrivals",
    x = "Actual arrivals",
    y = "Predicted arrivals (model1)"
  )
```

```{r}
hist(arrivals_employees_filled$arrivals_merged, breaks = 30,
     main = "Distribution of Filled Arrivals",
     xlab = "Arrivals")
```
```{r}
df_compare <- arrivals_employees_filled %>%
  mutate(source = ifelse(is.na(arrivals), "predicted", "actual"))

ggplot(df_compare, aes(x = arrivals_merged, fill = source)) +
  geom_histogram(alpha = 0.5, position = "identity") +
  theme_minimal()
```
```{r}
ggplot(arrivals_employees_filled,
       aes(x = data_year, y = arrivals_merged, color = is_filled)) +
  geom_line() +
  facet_wrap(~ municipio_clean, scales = "free_y") +
  theme_minimal() +
  labs(title = "Per-Municipio Time Series (Actual vs Filled)")
```

```{r}
## check that there are no negative predictions
arrivals_employees_filled %>%
  filter(arrivals_merged < 0) ## none!!
```

```{r}
df_model3_filled %>%
  filter(!is.na(arrivals)) %>%
  summarize(
    RMSE = sqrt(mean((arrivals - predict(model3, newdata = .))^2)),
    MAE  = mean(abs(arrivals - predict(model3, newdata = .)))
  )

df_model1_filled %>%
  filter(!is.na(arrivals)) %>%
  summarize(
    RMSE = sqrt(mean((arrivals - predict(model1, newdata = .))^2)),
    MAE  = mean(abs(arrivals - predict(model1, newdata = .)))
  )
```

```{r}
## diagnostics
set.seed(123)

train_idx <- sample(nrow(df_model3_orig), 0.8 * nrow(df_model3_orig))
train_data <- df_model3_orig[train_idx, ]
test_data  <- df_model3_orig[-train_idx, ]

model3_cv <- lm(arrivals ~ total_employees, data = train_data)

pred_test <- predict(model3_cv, newdata = test_data)

sqrt(mean((test_data$arrivals - pred_test)^2))  # RMSE

```

```{r}
arrivals_employees_filled %>%
  filter(is_filled == "1") %>%
  mutate(pred = ifelse(!is.na(arrivals_filled_model3),
                       predict(model3, newdata = .),
                       predict(model1, newdata = .))) %>%
  summarize(
    diff_mean  = mean(arrivals_merged - pred, na.rm = TRUE),
    diff_sd    = sd(arrivals_merged - pred, na.rm = TRUE),
    diff_range = range(arrivals_merged - pred, na.rm = TRUE)
  )

```



```{r}
ggplot(arrivals_employees_filled,
       aes(x = total_employees, y = arrivals_merged, color = is_filled)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_minimal()
```
## Write out

```{r}
write_csv(arrivals_employees_filled, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/gapfilled_int","arrivals_filled_by_DENUE_accommodation_employees_2019_2024.csv"))
```




# Old exploring script

```{r}
library(tidyverse)
library(magrittr)
library(here)
library(janitor)
library(tmaptools)
library(sf)
library(mapview)

sp_1 <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue_establishments/denue_00_72_1_shp/conjunto_de_datos","denue_inegi_72_1.shp"))

sp_1_diccionario_de_datos <- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue_establishments/denue_00_72_1_shp/diccionario_de_datos","denue_diccionario_de_datos.csv"))

sp_1_goc <- sp_1 %>% 
  filter(entidad %in% c("Baja California","Baja California Sur","Jalisco","Nayarit","Sinaloa","Sonora"))

sp_1_goc[] <- lapply(sp_1_goc, function(x) {
  if (is.character(x)) iconv(x, from = "", to = "UTF-8", sub = "byte") else x
})

# sp_2_valid <- sp_2 %>% 
#   select(id, geometry)

mapview(sp_1_goc)
```

NOTE: `fecha_alta` is "Fecha en la que la unidad económica se integró al Directorio Estadístico Nacional de Unidades Económicas y al Registro Estadístico de Negocios de México (RENEM)."

Aka, in English, it is the date on which the economic unit was added to the National Statistical Directory of Economic Units and the Statistical Registry of Businesses in Mexico (RENEM).


```{r}
sp_2 <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue_establishments/denue_00_72_2_shp/conjunto_de_datos","denue_inegi_72_2.shp"), options = "ENCODING=UTF-8")

# sp_2_dropgeo <- sp_2 %>%
#   st_drop_geometry()
# 
# sp_2_sf <- sp_2_dropgeo %>%
#   st_as_sf(coords = c("longitud", "latitud"),
#            crs = 4326, # WGS84
#            remove = FALSE)
# 
# sapply(sp_2, function(col) {
#   if (is.character(col)) {
#     any(is.na(iconv(col, from = "UTF-8", to = "UTF-8")))
#   } else {
#     FALSE
#   }
# })

unique(sp_2$entidad)

sp_2_goc <- sp_2 %>% 
  filter(entidad %in% c("Baja California","Baja California Sur","Jalisco","Nayarit","Sinaloa","Sonora"))

sp_2_goc[] <- lapply(sp_2_goc, function(x) {
  if (is.character(x)) iconv(x, from = "", to = "UTF-8", sub = "byte") else x
})

# sp_2_valid <- sp_2 %>% 
#   select(id, geometry)

mapview(sp_2_goc)
```


```{r}
sp_1_2 <- st_union(sp_1_goc, sp_2_goc)

mapview(sp_1_goc) + mapview(sp_2_goc)
```


```{r}
denue_71 <- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue_establishments/denue_00_71_csv/conjunto_de_datos","denue_inegi_71_.csv")) %>% 
  mutate()

unique(denue_71$localidad) # \x needs to be replaced
unique(denue_71$entidad)
# [1] "Aguascalientes"                  "Baja California"                 "Baja California Sur"             "Campeche"                       
#  [5] "Coahuila de Zaragoza"            "Colima"                          "Chiapas"                         "Chihuahua"                      
#  [9] "Ciudad de M\xe9xico"             "Durango"                         "Guanajuato"                      "Guerrero"                       
# [13] "Hidalgo"                         "Jalisco"                         "M\xe9xico"                       "Michoac\xe1n de Ocampo"         
# [17] "Morelos"                         "Nayarit"                         "Nuevo Le\xf3n"                   "Oaxaca"                         
# [21] "Puebla"                          "Quer\xe9taro"                    "Quintana Roo"                    "San Luis Potos\xed"             
# [25] "Sinaloa"                         "Sonora"                          "Tabasco"                         "Tamaulipas"                     
# [29] "Tlaxcala"                        "Veracruz de Ignacio de la Llave" "Yucat\xe1n"                      "Zacatecas" 

denue_71_goc <- denue_71 %>% 
  filter(entidad %in% c("Baja California","Baja California Sur","Jalisco","Nayarit","Sinaloa","Sonora"))

denue_71_sf <- denue_71_goc %>%
  st_as_sf(coords = c("longitud", "latitud"),
           crs = 4326, # WGS84
           remove = FALSE)
mapview(denue_71_sf)


denue_71_sf[] <- lapply(denue_71_sf, function(x) {
  if (is.character(x)) iconv(x, from = "", to = "UTF-8", sub = "byte") else x
})

mapview(denue_71_sf) ## not as complete as using the shp files from INEGI
```

### 2024

```{r}
sp_1_2024 <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue_establishments/denue_00_72_1_1124_shp/conjunto_de_datos","denue_inegi_72_1.shp"))

sp_1_2024_goc <- sp_1_2024 %>% 
  filter(entidad %in% c("Baja California","Baja California Sur","Jalisco","Nayarit","Sinaloa","Sonora"))

sp_1_2024_goc[] <- lapply(sp_1_2024_goc, function(x) {
  if (is.character(x)) iconv(x, from = "", to = "UTF-8", sub = "byte") else x
})

mapview(sp_1_2024_goc)
```


### 2023

```{r}
sp_1_2023 <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue_establishments/denue_00_72_1_1123_shp/conjunto_de_datos","denue_inegi_72_1.shp"))

sp_1_2023_goc <- sp_1_2023 %>% 
  filter(entidad %in% c("Baja California","Baja California Sur","Jalisco","Nayarit","Sinaloa","Sonora"))

sp_1_2023_goc[] <- lapply(sp_1_2023_goc, function(x) {
  if (is.character(x)) iconv(x, from = "", to = "UTF-8", sub = "byte") else x
})

mapview(sp_1_2023_goc)
```

### 2022

```{r}
sp_1_2022 <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue_establishments/denue_00_72_1_1122_shp/conjunto_de_datos","denue_inegi_72_1.shp"))

unique(sp_1_2022$entidad)
#  [1] "AGUASCALIENTES"         "BAJA CALIFORNIA"        "CAMPECHE"               "BAJA CALIFORNIA SUR"   
#  [5] "COAHUILA DE ZARAGOZA"   "COLIMA"                 "CHIAPAS"                "CHIHUAHUA"             
#  [9] "CIUDAD DE M\xc9XICO"    "DURANGO"                "GUANAJUATO"             "GUERRERO"              
# [13] "HIDALGO"                "JALISCO"                "M\xc9XICO"              "MICHOAC\xc1N DE OCAMPO"
# [17] "MORELOS"                "NAYARIT"                "NUEVO LE\xd3N"          "OAXACA"                
# [21] "PUEBLA" 

sp_1_2022_goc <- sp_1_2022 %>% 
  filter(entidad %in% c("BAJA CALIFORNIA","BAJA CALIFORNIA SUR","JALISCO","NAYARIT")) ## sinaloa and sonora are likely in sp_2!!

sp_1_2022_goc[] <- lapply(sp_1_2022_goc, function(x) {
  if (is.character(x)) iconv(x, from = "", to = "UTF-8", sub = "byte") else x
})

mapview(sp_1_2022_goc)
```

### Checking the differences between the years downloaded to see if we need to download each year's data, or if we can just go based off of the most recent shapefiles.

```{r}
## check the ids of the establishments
length(unique(sp_1_goc$id)) # this is 2025: 90669
length(unique(sp_1_2024_goc$id)) # 89759
length(unique(sp_1_2023_goc$id)) # 89759
length(unique(sp_1_2022_goc$id)) # 80378
```

I think that they may have accidentally saved the 2023 shapefile as the 2023 shapefile, as there are establishments that were added to the National Statistical Directory of Economic Units and the Statistical Registry of Businesses in Mexico (RENEM) in 2024 in the 2023 shapefile.


```{r}
## check which establishments are new between 2022 and 2024, and make sure that the fecha_alta reflects their correct date of addition

length(setdiff(unique(sp_1_2024_goc$id), unique(sp_1_2022_goc$id))) # 39269

# 89759 - 80378 = 9381
## This does not really make sense. 

length(setdiff(unique(sp_1_2022_goc$id), unique(sp_1_2024_goc$id))) # 29888

## okay.  So this means that establishments are removed when they no longer exist, I believe.  
```

So, let's see which of the establishments are present in 2024 only, and which are present in 2022 only.

```{r}
## 2024 only

only_2024 <- setdiff(unique(sp_1_2024_goc$id), unique(sp_1_2022_goc$id))

sp_1_2024_goc_ONLY24 <-sp_1_2024_goc %>% 
  filter(id %in% only_2024)

range(sp_1_2024_goc_ONLY24$fecha_alta) # "2010-07" "2024-11"
unique(sp_1_2024_goc_ONLY24$fecha_alta)
#  [1] "2024-11" "2023-11" "2019-11" "2010-07" "2014-12" "2020-11" "2024-05" "2021-05" "2016-10" "2016-01" "2011-03"
# [12] "2021-11" "2013-07" "2020-04" "2022-11" "2012-06"

## why might there be establishments that were added to the National Statistical Directory of Economic Units and the Statistical Registry of Businesses in Mexico (RENEM) in 2010 in the 2024 shapefile, that are not in the 2022 shapefile?
```

After some thought, this is what I have determined:

1.  DENUE versions are cumulative but filtered by active status.
  Every DENUE release only includes active establishments at the time of publication.
  If an establishment was closed (temporarily or permanently) in 2022, it might be missing from the 2022 file, even if it has reopened or been reactivated by 2024.

2. Reactivated establishments keep their original `fecha_alta`.
  That is why there are fecha_altas as far back as 2010 in the 2024 file that were not present in the 2022 file: these businesses first entered RENEM in 2010, later disappeared from the active roster, and were likely revalidated by INEGI in 2024.

3. Differences between years do NOT equal purely new/closed businesses.
  Some IDs are newly added (new businesses).
  Some are reactivations.
  Some disappear because they closed, changed name or location, or failed verification.
  
SO, there can be establishments split/merged (1 --> 2 or 2 --> 1), occasional ID reassignments, and geographic filtering differences.  **I will need to merge the sp_1 and sp_2 for each year and compare again.** 

## Merge sp_1_goc and sp_2_goc

```{r}
names(sp_1_goc) == names(sp_2_goc) ## TRUE
st_crs(sp_1_goc) == st_crs(sp_2_goc) ## TRUE

common_cols <- intersect(names(sp_1_goc), names(sp_2_goc))
common_cols # just double checking before binding

## bind sp_1_2_goc, uses less memory than st_union() or rbind()
sp_1_2_goc <- bind_rows(sp_1_goc[, common_cols], sp_2_goc[, common_cols])
```

#### Bring in coastal population (we likely won't need this here, replace with TOURIST ARRIVALS)

```{r}
# ohi_rgn_pop_10km <- read_csv(here("general/OHI_rgn_population_10km.csv")) # now we have the number of individuals living within 10km of the coast for each state.
```
#### Get coastline buffered 10 km inland (thanks Mel!!)

```{r}
regions_marine <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp")) %>%
  filter(location == "marine")
regions_land <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp")) %>%
  filter(location == "land")

coast_buffer <- st_buffer(regions_marine, dist = 10000) %>%
  st_union() %>%
st_sf()

inland_10 <- st_intersection(coast_buffer, regions_land)
inland_10_state_region <- st_intersection(inland_10, states)
mapview(inland_10_state_region[16,])

# removed the weird ones by hand, probably should have done this in a more reproducible way!
inland_10_state_region <- inland_10_state_region[-c(3, 4, 10, 11, 13, 14),] %>%
st_make_valid()
  
i <- which(st_geometry_type(inland_10_state_region) == "GEOMETRYCOLLECTION")
g <- inland_10_state_region[i,]
poly_only <- st_collection_extract(g, c("POLYGON"))
poly_only <- st_cast(st_union(poly_only), "MULTIPOLYGON")
inland_10_state_region$geometry[i] <- poly_only

mapview(inland_10_state_region)
```

#### Isolate coastal service establishments

```{r}
## match the crs of both objects
st_crs(sp_1_2_goc) == st_crs(inland_10_state_region) ## FALSE

st_crs(sp_1_2_goc) # WGS84
st_crs(inland_10_state_region) # MEX ITRF2008 / LLC

sp_1_2_goc <- st_transform(sp_1_2_goc, st_crs(inland_10_state_region)) # transform to match crs
```

```{r}
## spatial join to only keep service establishments within 10km of coastline
sp_1_2_goc_coastal <- st_join(
  sp_1_2_goc,
  inland_10_state_region,
  join = st_intersects,
  left = FALSE
) ## note: this only keep the point geometries
```

```{r}
## check to make sure it looks good
mapview(inland_10_state_region) + mapview(sp_1_2_goc_coastal, col.regions = "red") ## looks great!!!
```

#### Summarize the number of service establishments by region and by fecha_alta

Ways to do this: 

- Track the rate at which service establishments are added to the coast (this does not seem good, because it is not necessarily good to have more and more establishments -- that could be potentially a negative!)
- Make a ratio of number of establishments per total tourist arrivals (total # of service establishments for a region for that year / # of tourist arrivals in the coastal region for that year) and then say that based on a historical reference point (maybe 2019, before covid), the ratio should either be maintained or increase to get a score of 100.

Let's start with the second option, and we can adjust if we determine there is a better way to do this. 

To clarify, the ratio for every year should include the number of service establishments with a fecha_alta EQUAL to that year and all previous years.  So for example:

> If we are evaluating the ratio for the year 2022, I would want to count all observations of establishments with a `fecha_alta` less than or equal to 2022.  For 2024, it would be all observations of establishments with a `fecha_alta` less than or equal to 2024.

My only concern about this is that it may not change and will only increase if we are using the same shapefile.  So we may need to download all year's shapefile's separately and then evaluate the ratio separately. 

So we would essentially download all year's shapefiles available on INEGI DENUE (ex: denue_00_72_2_1124_shp, denue_00_72_2_1123_shp, denue_00_72_2_1122_shp, denue_00_72_2_1121_shp) for both sp_1 and sp_2.  

Then, we would bind all of them together after using mutate() to give each merged sp_1_2_goc an identifier for the year it was from.  So the cleaned denue_00_72_1_1122_shp & denue_00_72_2_1122_shp would have a column that just has the year 2022 for every observation to indicate the `data_year`.

In the completely bound dataset, of all data_years, we can then summarize and determine the number of service establishments by region and data_year.  Then those are all divided by the number of tourist arrivals for that year per region.

We can then compare the ratios across different years. Does this sound right??








