---
title: 'OHI GoC 2025: Tourism & Recreation'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# Merging all indicators into the TR current status score

This script brings in all current status scores for the components of TR, and puts them together. 

TR = avg(Quantity + Quality + Sustainability)

Quantity = tourist arrivals (avg int + dom)

Quality = avg(crime/coastal pop + activities)
activities = avg(2/6 * food&drink emp / tourist arrivals + 3/6 * cultural&gulf-related / tourist arrivals + 1/6 * other_recreational / tourist arrivals)

Sustainability = avg(distintivo S + CONANP returned fees + job quality)

```{r}
library(tidyverse)
library(sf)
library(mapview)
library(here)
```


```{r}
## for plotting:
goc_region_colors <- c(
  "Region_1" = "#1b9e77",
  "Region_2" = "#d95f02",
  "Region_3" = "#7570b3",
  "Region_4" = "#e7298a",
  "Region_5" = "#66a61e",
  "Region_6" = "#e6ab02",
  "Region_7" = "#a6761d",
  "Region_8" = "#666666",
  "Region_9" = "#1f78b4"
)

region_levels <- paste0("Region_", 1:9)

cs_quantity <- cs_quantity %>%
  mutate(rgn_id = factor(rgn_id, levels = region_levels))

cs_quality <- cs_quality %>%
  mutate(rgn_id = factor(rgn_id, levels = region_levels))

cs_sust <- cs_sust %>%
  mutate(rgn_id = factor(rgn_id, levels = region_levels))

all_cs_indicators <- all_cs_indicators %>%
  mutate(rgn_id = factor(rgn_id, levels = region_levels))
```


## Quantity

```{r}
cs_quantity <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quantity","tr_quantity_current_status_121025.csv")) %>% 
  rename(final_score = score) %>% 
  select(rgn_id, year, final_score)
```

```{r}
library(plotly)

quantity_plotly <- plotly::plot_ly(
  cs_quantity, 
  x = ~year, 
  y = ~final_score, 
  color = ~rgn_id, 
  colors = goc_region_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: Tourism & Recreation Quantity",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cs_quantity$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
quantity_plotly
```


## Quality

### Crime

```{r}
region_2_6_tibble <- tibble::tribble(
  ~year, ~score, ~rgn_id,
  2024, 100, "Region_2",
  2023, 100, "Region_2",
  2022, 100, "Region_2",
  2021, 100, "Region_2",
  2020, 100, "Region_2",
  2019, 100, "Region_2",
  2024, 100, "Region_6",
  2023, 100, "Region_6",
  2022, 100, "Region_6",
  2021, 100, "Region_6",
  2020, 100, "Region_6",
  2019, 100, "Region_6"
)

cs_quality_crime <- read_csv(here::here("TR/tr_quality_int/tr_qual_crime_score_120925.csv")) %>% 
  select(rgn_id, year = ano, score = current_status) %>% 
  filter(year %in% 2019:2024) %>% 
  bind_rows(region_2_6_tibble) %>%  # had no registered crimes.  might change this to just NA.
  complete(rgn_id, year = 2019:2024) %>% 
  mutate(score = ifelse(is.na(score), 100, score)) %>% 
  mutate(quality_type = "crime") %>% 
  mutate(rgn_id = factor(rgn_id, levels = region_levels))

crime_plotly <- plotly::plot_ly(
  cs_quality_crime, 
  x = ~year, 
  y = ~score, 
  color = ~rgn_id, 
  colors = goc_region_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "TR Quality: crime",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cs_quality_crime$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
crime_plotly
```

### Activities

```{r}
cs_quality_activities <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quality","final_activity_avg_score_121625.csv")) %>%   select(rgn_id, year, score = final_score) %>% 
  mutate(quality_type = "activities") %>% 
  mutate(rgn_id = factor(rgn_id, levels = region_levels))

activities_plotly <- plotly::plot_ly(
  cs_quality_activities, 
  x = ~year, 
  y = ~score, 
  color = ~rgn_id, 
  colors = goc_region_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "TR Quality: activities",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cs_quality_activities$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
activities_plotly
```


### Merge

```{r}
merge_quality <- bind_rows(cs_quality_activities, cs_quality_crime) %>% 
  select(rgn_id, year, score, quality_type) %>% 
  filter(year %in% 2019:2024) %>% 
  pivot_wider(
    names_from = quality_type,
    values_from = score
  ) %>% 
  mutate(crime = ifelse(is.na(crime), 100, crime)) ## EDIT THIS!!!!!

cs_quality <- merge_quality %>%
  rowwise() %>%
  mutate(
    final_score = mean(c(activities, crime))
  ) %>%
  ungroup() %>% 
  select(rgn_id, year, final_score)

write_csv(cs_quality, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quality","cs_quality_all_indicators_121625.csv"))
```

```{r}
library(plotly)

quality_plotly <- plotly::plot_ly(
  cs_quality, 
  x = ~year, 
  y = ~final_score, 
  color = ~rgn_id, 
  colors = goc_region_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: Tourism & Recreation Quality",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cs_quality$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
quality_plotly
```



## Sustainability

### Dist. S

```{r}
cs_sust_dist_s <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","dist_S_current_status_final_interpolated_121625.csv")) %>% 
  mutate(sust_type = "dist_s") %>% 
  select(rgn_id, year = data_year, score = score_interpolated, sust_type) %>% 
  filter(year %in% 2019:2024) %>% 
  mutate(rgn_id = factor(rgn_id, levels = region_levels))

dist_s_plotly <- plotly::plot_ly(
  cs_sust_dist_s, 
  x = ~year, 
  y = ~score, 
  color = ~rgn_id, 
  colors = goc_region_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "TR Sustainability: total dist. S hotels / total large hotels (interpolated)",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cs_sust_dist_s$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
dist_s_plotly
```

### CONANP fee retrieval from SHCP

```{r}
cs_sust_conanp <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","conanp_fee_return_current_status_final_121625.csv")) %>% 
  mutate(sust_type = "conanp",
         score = score * 100) %>% 
  filter(year %in% 2019:2024) %>% 
  mutate(rgn_id = factor(rgn_id, levels = region_levels))

conanp_plotly <- plotly::plot_ly(
  cs_sust_conanp, 
  x = ~year, 
  y = ~score, 
  color = ~rgn_id, 
  colors = goc_region_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "TR Sustainability: conanp fee retrieval from SHCP",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cs_sust_conanp$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
conanp_plotly
```

### Merge

```{r}
merge_sust <- bind_rows(cs_sust_dist_s, cs_sust_conanp) %>% 
  select(rgn_id, year, score, sust_type) %>% 
  pivot_wider(
    names_from = sust_type,
    values_from = score
  ) 

cs_sust <- merge_sust %>%
  rowwise() %>%
  mutate(
    final_score = mean(c(dist_s, conanp))
  ) %>%
  ungroup() %>% 
  select(rgn_id, year, final_score)

write_csv(cs_sust, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","cs_sust_all_indicators_121625.csv"))
```

```{r}
library(plotly)

sust_plotly <- plotly::plot_ly(
  cs_sust, 
  x = ~year, 
  y = ~final_score, 
  color = ~rgn_id,
  colors = goc_region_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: Tourism & Recreation Sustainability",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cs_sust$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
sust_plotly
```



# TR Score

```{r}
all_cs_indicators <- bind_rows(cs_sust, cs_quality, cs_quantity) %>% 
  group_by(rgn_id, year) %>% 
  summarize(tr_current_status = mean(final_score)) %>% 
  ungroup()

write_csv(all_cs_indicators, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output","tr_current_status_final_121825.csv"))

tr_goc_score <- all_cs_indicators %>% 
  filter(year == 2024) %>% 
  summarize(tr_goc_2024 = mean(tr_current_status)) # overall for this year: 81.51899!!!
```

```{r}
library(plotly)

tr_score_plotly <- plotly::plot_ly(
  all_cs_indicators, 
  x = ~year, 
  y = ~tr_current_status, 
  color = ~rgn_id, 
  colors = goc_region_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: Tourism & Recreation",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(all_cs_indicators$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
tr_score_plotly
```

Map viz:

```{r}
all_cs_indicators_2024 <- all_cs_indicators %>% 
  filter(year == 2024)
write_csv(all_cs_indicators_2024, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output","tr_current_status_final_2024_121825.csv"))

#### join with ohi rgn polygons
ohi <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp")) %>%
  filter(location == "marine") %>% 
  select(rgn_id)

all_cs_indicators_2024_ohigns <- full_join(all_cs_indicators_2024, ohi, by = "rgn_id") %>% 
  st_as_sf()
```


```{r}
mapview(all_cs_indicators_2024_ohigns, 
        zcol = "tr_current_status",
        col.regions = colorRampPalette(c('#FEE090', '#ABD9E9', '#4575B4', '#313695')),
        layer.name = "Tourism & Recreation Current Status Scores (2024)",
        legend = TRUE)

tr_at <- seq(0, 100, by = 10)

tr_cols <- colorRampPalette(c(
  '#8a0032', '#A50026', '#A50026', '#D73027', '#F46D43',
  '#FDAE61', '#FEE090', '#FEE090', '#ABD9E9', '#4575B4', '#313695'
))(length(tr_at) - 1)

mapview(all_cs_indicators_2024_ohigns, 
        zcol = "tr_current_status",
        col.regions = tr_cols,
        at = tr_at,
        layer.name = "Tourism & Recreation Current Status Scores (2024)",
        legend = TRUE)

mapview(
  all_cs_indicators_2024_ohigns,
  col.regions = "tr_cols",
  legend = FALSE
) %>%
  addLegend(
    position = "bottomright",
    pal = tr_pal,
    values = seq(0, 100, by = 1),
    title = "TR Current Status (0–100)"
  )
```

ggplot:
```{r}
tr_palette <- c(
  "#8a0032", "#A50026", "#A50026", "#D73027", "#F46D43",
  "#FDAE61", "#FEE090", "#FEE090", "#ABD9E9", "#4575B4", "#313695"
)

gg_tr <- ggplot() +
  geom_sf(
    data = all_cs_indicators_2024_ohigns,
    aes(fill = tr_current_status),
    color = "black",
    linewidth = 0.1
  ) +
  geom_sf(
    data = states,
    fill = "#d6cac0",
    color = "black",
    linewidth = 0.1
  ) +
  scale_fill_gradientn(
    colours = tr_palette,
    limits  = c(0, 100),        # min and max of the scale
    name    = "Tourism & Recreation\ncurrent status"
  ) +
  coord_sf(
    xlim = c(bbox_limits["xmin"], bbox_limits["xmax"]),
    ylim = c(bbox_limits["ymin"], bbox_limits["ymax"]),
    expand = FALSE
  ) +
  theme_void() +
  # North arrow & scale bar
  annotation_scale(
    location = "bl",
    width_hint = 0.25,
    bar_cols = c("#5f5a55", "white"),
    text_cex = 1
  ) +
  annotation_north_arrow(
    location = "bl",
    which_north = "true",
    pad_x = unit(0.0, "in"),
    pad_y = unit(0.3, "in"),
    style = north_arrow_fancy_orienteering
  ) +
  
  # Theme
  theme_void() +
  theme(
    text = element_text(family = "Calibri"),
    panel.background = element_rect(fill = "white", color = NA),
    plot.background  = element_rect(fill = "white", color = NA),
    legend.position  = "right",
    legend.key.height = unit(0.8, "cm"),
    legend.key.width = unit(0.5, "cm"),
    legend.title = element_text(size = 12),
    legend.text  = element_text(size = 10),
    plot.margin = margin(10, 10, 10, 10)
  ) 
gg_tr
```


Radar plot:
```{r}

all_cs_indicators_2024 <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output","tr_current_status_final_2024_121825.csv"))

radar_df <- all_cs_indicators_2024 %>%
  filter(!is.na(tr_current_status)) %>%
  arrange(rgn_id) %>%
  mutate(
    rgn_id = factor(rgn_id, levels = rgn_id)
  )

ggplot(
  radar_df,
  aes(
    x = rgn_id,
    y = tr_current_status,
    fill = rgn_id
  )
) +
  # radial bars
  geom_col(
    width = 0.99,
    color = "white",
    linewidth = 0.5,
    alpha = 0.9
  ) +
  # 100 reference ring
  geom_hline(
    yintercept = 100,
    color = "grey40",
    linewidth = 0.9
  ) +
  coord_polar(start = -pi / 2) +
  scale_y_continuous(
    limits = c(0, 100),
    breaks = c(25, 50, 75, 100),
    expand = c(0, 0)
  ) +
  scale_fill_manual(values = rgn_colors) +
  labs(
    title = "Tourism & Recreation current status by region (2024)",
    y = NULL,
    x = NULL
  ) +
  theme_minimal(base_family = "Calibri") +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.text.x = element_text(
      face = "bold",
      size = 12,
      color = "black"
    ),
    axis.text.y = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(color = "grey85"),
    legend.position = "none"
  ) + 
  geom_text(
    aes(label = round(tr_current_status, 1)),
    color = "black",
    size = 3,
    vjust = -0.3
  )




```

```{r}
library(dplyr)
library(ggplot2)
library(readr)
library(here)

all_cs_indicators_2024 <- read_csv(here::here(
  "/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output",
  "tr_current_status_final_2024_121825.csv"
))

rgn_labels <- c(
  "Region_1" = "Western Upper Gulf",
  "Region_2" = "Western Midriff Islands",
  "Region_3" = "Western Central Gulf",
  "Region_4" = "Western Southern Gulf",
  "Region_5" = "Eastern Upper Gulf",
  "Region_6" = "Eastern Midriff Islands",
  "Region_7" = "Southern Sonora Coast",
  "Region_8" = "Sinaloa Coast",
  "Region_9" = "Southeastern Gulf"
)

radar_df <- all_cs_indicators_2024 |>
  dplyr::filter(!is.na(tr_current_status)) |>
  dplyr::arrange(rgn_id) |>
  dplyr::mutate(
    rgn_id = factor(rgn_id, levels = unique(rgn_id))
  )

ggplot(
  radar_df,
  aes(
    x = rgn_id,
    y = tr_current_status,
    fill = rgn_id
  )
) +
  geom_col(
    width = 0.98,
    color = "white",
    linewidth = 0.4,
    alpha = 0.9
  ) +
  geom_hline(
    yintercept = 100,
    color = "grey60",
    linewidth = 0.6
  ) +
  coord_polar(start = -pi / 2, clip = "off") +
  scale_y_continuous(
    limits = c(0, 100),
    breaks = c(25, 50, 75, 100),
    expand = c(0, 0)
  ) +
  scale_x_discrete(
    labels = rgn_labels        # names must match factor levels
  ) +
  scale_fill_manual(values = rgn_colors) +
  labs(
    title = "Tourism & Recreation current status by region (2024)",
    y = NULL,
    x = NULL
  ) +
  theme_minimal(base_family = "Calibri") +
  theme(
    plot.margin = margin(t = 15, r = 15, b = 15, l = 15),  # <- more space
    plot.title  = element_text(face = "bold", size = 14),
    axis.text.x = element_text(
    face   = "bold",
    size   = 8,
    color  = "black",
    angle  = 10,      # try 0, 20, 30, 40, etc.
    vjust  = 5,
    hjust  = 10,
    margin = margin(t = -5)
  ),
    axis.text.y = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(color = "grey90"),
    legend.position = "none"
  ) +
  geom_text(
    aes(label = round(tr_current_status, 1)),
    color = "black",
    size = 3,
    vjust = -0.2,
    family = "Calibri"
  )




```

