---
title: 'OHI GoC 2025: Tourism & Recreation'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# Merging all indicators into the TR current status score

This script brings in all current status scores for the components of TR, and puts them together. 

TR = avg(Quantity + Quality + Sustainability)

Quantity = tourist arrivals (avg int + dom)

Quality = avg(crime/coastal pop + activities)
activities = avg(2/6 * food&drink emp / tourist arrivals + 3/6 * cultural&gulf-related / tourist arrivals + 1/6 * other_recreational / tourist arrivals)

Sustainability = avg(distintivo S + CONANP returned fees + job quality)

```{r}
library(tidyverse)
library(sf)
library(mapview)
library(here)
```


```{r}
## for plotting:
goc_region_colors <- c(
  "Region_1" = "#1b9e77",
  "Region_2" = "#d95f02",
  "Region_3" = "#7570b3",
  "Region_4" = "#e7298a",
  "Region_5" = "#66a61e",
  "Region_6" = "#e6ab02",
  "Region_7" = "#a6761d",
  "Region_8" = "#666666",
  "Region_9" = "#1f78b4"
)

region_levels <- paste0("Region_", 1:9)

cs_quantity <- cs_quantity %>%
  mutate(rgn_id = factor(rgn_id, levels = region_levels))

cs_quality <- cs_quality %>%
  mutate(rgn_id = factor(rgn_id, levels = region_levels))

cs_sust <- cs_sust %>%
  mutate(rgn_id = factor(rgn_id, levels = region_levels))

all_cs_indicators <- all_cs_indicators %>%
  mutate(rgn_id = factor(rgn_id, levels = region_levels))
```


## Quantity

```{r}
cs_quantity <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quantity","tr_quantity_current_status_121025.csv")) %>% 
  rename(final_score = score) %>% 
  select(rgn_id, year, final_score)
```

```{r}
library(plotly)

quantity_plotly <- plotly::plot_ly(
  cs_quantity, 
  x = ~year, 
  y = ~final_score, 
  color = ~rgn_id, 
  colors = goc_region_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: Tourism & Recreation Quantity",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cs_quantity$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
quantity_plotly
```


## Quality

### Crime

```{r}
region_2_tibble <- tibble::tribble(
  ~year, ~score, ~rgn_id,
  2024, 100, "Region_2",
  2023, 100, "Region_2",
  2022, 100, "Region_2",
  2021, 100, "Region_2",
  2020, 100, "Region_2",
  2019, 100, "Region_2"
)

cs_quality_crime <- read_csv(here::here("TR/tr_quality_int/tr_qual_crime_score_120925.csv")) %>% 
  select(rgn_id, year = ano, score = current_status) %>% 
  filter(year %in% 2019:2024) %>% 
  bind_rows(region_2_tibble) %>%  # had no registered crimes.  might change this to just NA.
  complete(rgn_id, year = 2019:2024) %>% 
  mutate(score = ifelse(is.na(score), 100, score)) %>% 
  mutate(quality_type = "crime") %>% 
  mutate(rgn_id = factor(rgn_id, levels = region_levels))

crime_plotly <- plotly::plot_ly(
  cs_quality_crime, 
  x = ~year, 
  y = ~score, 
  color = ~rgn_id, 
  colors = goc_region_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "TR Quality: crime",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cs_quality_crime$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
crime_plotly
```

### Activities

```{r}
cs_quality_activities <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quality","final_activity_avg_score_121625.csv")) %>%   select(rgn_id, year, score = final_score) %>% 
  mutate(quality_type = "activities") %>% 
  mutate(rgn_id = factor(rgn_id, levels = region_levels))

activities_plotly <- plotly::plot_ly(
  cs_quality_activities, 
  x = ~year, 
  y = ~score, 
  color = ~rgn_id, 
  colors = goc_region_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "TR Quality: activities",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cs_quality_activities$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
activities_plotly
```


### Merge

```{r}
merge_quality <- bind_rows(cs_quality_activities, cs_quality_crime) %>% 
  select(rgn_id, year, score, quality_type) %>% 
  filter(year %in% 2019:2024) %>% 
  pivot_wider(
    names_from = quality_type,
    values_from = score
  ) %>% 
  mutate(crime = ifelse(is.na(crime), 100, crime)) ## EDIT THIS!!!!!

cs_quality <- merge_quality %>%
  rowwise() %>%
  mutate(
    final_score = mean(c(activities, crime))
  ) %>%
  ungroup() %>% 
  select(rgn_id, year, final_score)

write_csv(cs_quality, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quality","cs_quality_all_indicators_121625.csv"))
```

```{r}
library(plotly)

quality_plotly <- plotly::plot_ly(
  cs_quality, 
  x = ~year, 
  y = ~final_score, 
  color = ~rgn_id, 
  colors = goc_region_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: Tourism & Recreation Quality",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cs_quality$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
quality_plotly
```



## Sustainability

### Dist. S

```{r}
cs_sust_dist_s <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","dist_S_current_status_final_interpolated_121625.csv")) %>% 
  mutate(sust_type = "dist_s") %>% 
  select(rgn_id, year = data_year, score = score_interpolated, sust_type) %>% 
  filter(year %in% 2019:2024) %>% 
  mutate(rgn_id = factor(rgn_id, levels = region_levels))

dist_s_plotly <- plotly::plot_ly(
  cs_sust_dist_s, 
  x = ~year, 
  y = ~score, 
  color = ~rgn_id, 
  colors = goc_region_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "TR Sustainability: total dist. S hotels / total large hotels (interpolated)",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cs_sust_dist_s$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
dist_s_plotly
```

### CONANP fee retrieval from SHCP

```{r}
cs_sust_conanp <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","conanp_fee_return_current_status_final_121625.csv")) %>% 
  mutate(sust_type = "conanp",
         score = score * 100) %>% 
  filter(year %in% 2019:2024) %>% 
  mutate(rgn_id = factor(rgn_id, levels = region_levels))

conanp_plotly <- plotly::plot_ly(
  cs_sust_conanp, 
  x = ~year, 
  y = ~score, 
  color = ~rgn_id, 
  colors = goc_region_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "TR Sustainability: conanp fee retrieval from SHCP",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cs_sust_conanp$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
conanp_plotly
```

### Merge

```{r}
merge_sust <- bind_rows(cs_sust_dist_s, cs_sust_conanp) %>% 
  select(rgn_id, year, score, sust_type) %>% 
  pivot_wider(
    names_from = sust_type,
    values_from = score
  ) 

cs_sust <- merge_sust %>%
  rowwise() %>%
  mutate(
    final_score = mean(c(dist_s, conanp))
  ) %>%
  ungroup() %>% 
  select(rgn_id, year, final_score)

write_csv(cs_sust, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","cs_sust_all_indicators_121625.csv"))
```

```{r}
library(plotly)

sust_plotly <- plotly::plot_ly(
  cs_sust, 
  x = ~year, 
  y = ~final_score, 
  color = ~rgn_id,
  colors = goc_region_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: Tourism & Recreation Sustainability",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cs_sust$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
sust_plotly
```



# TR Score

```{r}
all_cs_indicators <- bind_rows(cs_sust, cs_quality, cs_quantity) %>% 
  group_by(rgn_id, year) %>% 
  summarize(tr_current_status = mean(final_score)) %>% 
  ungroup()

tr_goc_score <- all_cs_indicators %>% 
  filter(year == 2024) %>% 
  summarize(tr_goc_2024 = mean(tr_current_status)) # overall for this year: 81.51899!!!
```

```{r}
library(plotly)

tr_score_plotly <- plotly::plot_ly(
  all_cs_indicators, 
  x = ~year, 
  y = ~tr_current_status, 
  color = ~rgn_id, 
  colors = goc_region_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: Tourism & Recreation",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(all_cs_indicators$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
tr_score_plotly
```

