---
title: 'OHI GoC 2025: Tourism & Recreation'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# Merging all indicators into the TR current status score

This script brings in all current status scores for the components of TR, and puts them together. 

TR = avg(Quantity + Quality + Sustainability)

Quantity = tourist arrivals (avg int + dom)

Quality = avg(crime/coastal pop + activities)
activities = avg(2/6 * food&drink emp / tourist arrivals + 3/6 * cultural&gulf-related / tourist arrivals + 1/6 * other_recreational / tourist arrivals)

Sustainability = avg(distintivo S + CONANP returned fees + job quality)

```{r}
library(tidyverse)
library(sf)
library(mapview)
library(here)
```


```{r}
## for plotting:
rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

region_levels <- paste0("Region_", 1:9)

cs_quality <- cs_quality %>%
  mutate(rgn_id = factor(rgn_id, levels = region_levels))

cs_sust <- cs_sust %>%
  mutate(rgn_id = factor(rgn_id, levels = region_levels))

all_cs_indicators <- all_cs_indicators %>%
  mutate(rgn_id = factor(rgn_id, levels = region_levels))
```


## Quantity

```{r}
# read in quantity
cs_quantity <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quantity","tr_quantity_current_status_121025.csv")) %>% 
  rename(final_score = score) %>% 
  dplyr::select(rgn_id, year, final_score)

# ensure regions are factors
cs_quantity_clean <- cs_quantity %>%
  mutate(rgn_id = factor(rgn_id, levels = region_levels))
```

```{r}
library(plotly)

quantity_plotly <- plotly::plot_ly(
  cs_quantity, 
  x = ~year, 
  y = ~final_score, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: Tourism & Recreation Quantity",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cs_quantity$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
quantity_plotly
```


## Quality

### Crime

```{r}
# region_2_6_tibble <- tibble::tribble(
#   ~year, ~score, ~rgn_id,
#   2024, 100, "Region_2",
#   2023, 100, "Region_2",
#   2022, 100, "Region_2",
#   2021, 100, "Region_2",
#   2020, 100, "Region_2",
#   2019, 100, "Region_2",
#   2024, 100, "Region_6",
#   2023, 100, "Region_6",
#   2022, 100, "Region_6",
#   2021, 100, "Region_6",
#   2020, 100, "Region_6",
#   2019, 100, "Region_6"
# )

cs_quality_crime <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quality","tr_qual_crime_score_030626.csv")) %>% 
  dplyr::select(rgn_id, year = ano, score = current_status) %>% 
  filter(year %in% 2019:2024) %>% 
  mutate(quality_type = "crime")

# %>% 
#   bind_rows(region_2_6_tibble) %>%  # had no registered crimes.  might change this to just NA.
#   complete(rgn_id, year = 2019:2024) %>% 
#   mutate(score = ifelse(is.na(score), NA, score)) %>% 
#   mutate(quality_type = "crime") %>% 
#   mutate(rgn_id = factor(rgn_id, levels = region_levels))
# write_csv(cs_quality_crime, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quality","tr_qual_crime_score_120925.csv"))

crime_plotly <- plotly::plot_ly(
  cs_quality_crime, 
  x = ~year, 
  y = ~score, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "TR Quality: crime",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cs_quality_crime$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
crime_plotly
```

### Activities

```{r}
cs_quality_activities <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quality","final_activity_avg_score_030526.csv")) %>%   
  dplyr::select(rgn_id, year, score = final_score) %>% 
  mutate(quality_type = "activities") %>% 
  mutate(rgn_id = factor(rgn_id, levels = region_levels))

activities_plotly <- plotly::plot_ly(
  cs_quality_activities, 
  x = ~year, 
  y = ~score, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "TR Quality: Availability of activities and food",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cs_quality_activities$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
activities_plotly
```


### Merge

```{r}
merge_quality <- bind_rows(cs_quality_activities, cs_quality_crime) %>% 
  dplyr::select(rgn_id, year, score, quality_type) %>% 
  filter(year %in% 2019:2024) %>% 
  pivot_wider(
    names_from = quality_type,
    values_from = score
  ) 

cs_quality <- merge_quality %>%
  rowwise() %>%
  mutate(
    final_score = mean(c(activities, crime), na.rm = TRUE),
    final_score = ifelse(is.na(final_score), NA, final_score) # for rgn_2 2019:2020
  ) %>%
  ungroup() 

cs_quality_clean <- cs_quality %>% 
  dplyr::select(rgn_id, year, final_score)

write_csv(cs_quality_clean, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quality","cs_quality_all_indicators_032326.csv"))
```

```{r}
library(plotly)

quality_plotly <- plotly::plot_ly(
  cs_quality, 
  x = ~year, 
  y = ~final_score, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: Tourism & Recreation Quality",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cs_quality$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
quality_plotly
```



## Sustainability

### Dist. S

100+ employees
```{r}
cs_sust_dist_s <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","dist_S_current_status_final_2019_2024_100over_031326.csv")) %>% 
  mutate(sust_type = "dist_s") %>% 
  dplyr::select(rgn_id, year = cert_year, score = current_status, sust_type) %>% 
  filter(year %in% 2019:2024) %>% # to match other indicators
  mutate(rgn_id = factor(rgn_id, levels = region_levels))

dist_s_plotly <- plotly::plot_ly(
  cs_sust_dist_s, 
  x = ~year, 
  y = ~score, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "TR Sustainability: total dist. S hotels / total large hotels (interpolated)",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cs_sust_dist_s$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
dist_s_plotly
```


### CONANP fee retrieval from SHCP

```{r}
cs_sust_conanp_fee <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","conanp_fee_return_current_status_final_121625.csv")) %>% 
  mutate(sust_type = "conanp_fee",
         score = score * 100) %>% 
  filter(year %in% 2019:2025) %>% 
  mutate(rgn_id = factor(rgn_id, levels = region_levels))

conanp_plotly <- plotly::plot_ly(
  cs_sust_conanp_fee, 
  x = ~year, 
  y = ~score, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "TR Sustainability: conanp fee retrieval from SHCP",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cs_sust_conanp$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
conanp_plotly
```

### CONANP budget over time

Todas las cifras mencionadas están expresadas en pesos constantes del 2026 (términos reales).
<https://nossamexico.com/cuidar-lo-que-importa-el-presupuesto-para-el-cuidado-del-ambiente-y-las-areas-naturales-protegidas-en-el-ppef-2026/>

```{r}
cs_sust_conanp_budget <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","conanp_federal_budget_current_status_final_032026.csv")) %>% 
  dplyr::select(year, score = current_status_all) %>% 
  distinct() %>%  # avg of all programs for each year.  Will impact all regions the same.
  mutate(sust_type = "conanp_budget") %>% 
  filter(year %in% 2019:2025)  %>% # because everything else goes/will go through these years
  crossing(rgn_id = region_levels) %>% # instead of using complete(), because rgn_id does not exist here yet.
  mutate(rgn_id = factor(rgn_id, levels = region_levels))

```

### Blue Flag beaches

```{r}
cs_sust_blue_flag <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","cropped_blue_flag_current_status_final_by_yr_031926.csv")) %>% 
  dplyr::select(rgn_id, year, score = current_status) %>% 
  mutate(sust_type = "blue_flag") %>% 
  filter(year %in% 2019:2025) # because everything else goes/will go through these years
```


### Job Quality

Raw CS scores:
```{r}
cs_sust_job_quality <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","tr_job_quality_sustainability_current_status_merged_032326.csv")) %>% 
  dplyr::select(rgn_id, year = ano_censal, score = tr_job_quality_cs_raw) %>% 
  mutate(sust_type = "job_quality") %>% 
  filter(year %in% 2019:2025) # because everything else goes/will go through these years.... only one year of scores within these years (census every 5 years).
```

Using interpolated scores:
```{r}
cs_sust_job_quality <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","tr_job_quality_sustainability_current_status_merged_interpolated_032326.csv")) %>% 
  dplyr::select(rgn_id, year = ano_censal, score = score_interpolated) %>% 
  mutate(sust_type = "job_quality") %>% 
  filter(year %in% 2019:2025) # because everything else goes/will go through these years

```



### Merge

```{r}
merge_sust <- bind_rows(cs_sust_dist_s, cs_sust_conanp_fee, cs_sust_conanp_budget, cs_sust_blue_flag, cs_sust_job_quality) %>% 
  dplyr::select(rgn_id, year, score, sust_type) %>% 
  pivot_wider(
    names_from = sust_type,
    values_from = score
  ) 

cs_sust <- merge_sust %>%
  rowwise() %>%
  mutate(
    final_score = mean(c(dist_s, conanp_fee, conanp_budget, blue_flag, job_quality), na.rm = TRUE)
  ) %>%
  ungroup() 

cs_sust_clean <- cs_sust %>% 
  dplyr::select(rgn_id, year, final_score)

write_csv(cs_sust_clean, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","cs_sust_all_indicators_032326.csv"))
```

```{r}
library(plotly)

sust_plotly <- plotly::plot_ly(
  cs_sust, 
  x = ~year, 
  y = ~final_score, 
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: Tourism & Recreation Sustainability",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(cs_sust$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
sust_plotly
```



# TR Score

```{r}
all_cs_indicators <- bind_rows(cs_sust_clean, cs_quality_clean, cs_quantity_clean) %>% 
  group_by(rgn_id, year) %>% 
  summarize(tr_current_status = mean(final_score, na.rm = TRUE)) %>% 
  ungroup()

write_csv(all_cs_indicators, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output","tr_current_status_final_032326.csv"))

tr_goc_score <- all_cs_indicators %>% 
  filter(year == 2024) %>% 
  summarize(tr_goc_2024 = mean(tr_current_status)) # overall for this year (03/23/26): 70.48207!!!
```

```{r}
library(plotly)

tr_score_plotly <- plotly::plot_ly(
  all_cs_indicators, 
  x = ~year, 
  y = ~tr_current_status, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: Tourism & Recreation",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(all_cs_indicators$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
tr_score_plotly
```

Map viz:

```{r}
all_cs_indicators_2024 <- all_cs_indicators %>% 
  filter(year == 2024)
write_csv(all_cs_indicators_2024, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output","tr_current_status_final_2024_032326.csv"))

#### join with ohi rgn polygons
ohi <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp")) %>%
  filter(location == "marine") %>% 
  dplyr::select(rgn_id)

all_cs_indicators_2024_ohigns <- full_join(all_cs_indicators_2024, ohi, by = "rgn_id") %>% 
  st_as_sf()
```


```{r}
mapview(all_cs_indicators_2024_ohigns, 
        zcol = "tr_current_status",
        col.regions = colorRampPalette(c('#FEE090', '#ABD9E9', '#4575B4', '#313695')),
        layer.name = "Tourism & Recreation Current Status Scores (2024)",
        legend = TRUE)

tr_at <- seq(0, 100, by = 10)

tr_cols <- colorRampPalette(c(
  '#8a0032', '#A50026', '#A50026', '#D73027', '#F46D43',
  '#FDAE61', '#FEE090', '#FEE090', '#ABD9E9', '#4575B4', '#313695'
))(length(tr_at) - 1)

mapview(all_cs_indicators_2024_ohigns, 
        zcol = "tr_current_status",
        col.regions = tr_cols,
        at = tr_at,
        layer.name = "Tourism & Recreation Current Status Scores (2024)",
        legend = TRUE)

# mapview(
#   all_cs_indicators_2024_ohigns,
#   col.regions = "tr_cols",
#   legend = FALSE
# ) %>%
#   addLegend(
#     position = "bottomright",
#     pal = tr_pal,
#     values = seq(0, 100, by = 1),
#     title = "TR Current Status (0–100)"
#   )
```

ggplot:
```{r}
tr_palette <- c(
  "#8a0032", "#A50026", "#A50026", "#D73027", "#F46D43",
  "#FDAE61", "#FEE090", "#FEE090", "#ABD9E9", "#4575B4", "#313695"
)

gg_tr <- ggplot() +
  geom_sf(
    data = all_cs_indicators_2024_ohigns,
    aes(fill = tr_current_status),
    color = "black",
    linewidth = 0.1
  ) +
  geom_sf(
    data = states,
    fill = "#d6cac0",
    color = "black",
    linewidth = 0.1
  ) +
  scale_fill_gradientn(
    colours = tr_palette,
    limits  = c(0, 100),        # min and max of the scale
    name    = "Tourism & Recreation\nCurrent Status 2024"
  ) +
  coord_sf(
    # xlim = c(bbox_limits["xmin"], bbox_limits["xmax"]),
    # ylim = c(bbox_limits["ymin"], bbox_limits["ymax"]),
    expand = FALSE
  ) +
  theme_void() +
  # North arrow & scale bar
  # annotation_scale(
  #   location = "bl",
  #   width_hint = 0.25,
  #   bar_cols = c("#5f5a55", "white"),
  #   text_cex = 1
  # ) +
  # annotation_north_arrow(
  #   location = "bl",
  #   which_north = "true",
  #   pad_x = unit(0.0, "in"),
  #   pad_y = unit(0.3, "in"),
  #   style = north_arrow_fancy_orienteering
  # ) +
  
  # Theme
  theme(
    text = element_text(family = "Calibri"),
    panel.background = element_rect(fill = "white", color = NA),
    plot.background  = element_rect(fill = "white", color = NA),
    legend.position  = "right",
    legend.key.height = unit(0.8, "cm"),
    legend.key.width = unit(0.5, "cm"),
    legend.title = element_text(size = 12),
    legend.text  = element_text(size = 10),
    plot.margin = margin(10, 10, 10, 10)
  ) 
gg_tr
```


Radar plot:
```{r}

all_cs_indicators_2024 <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output","tr_current_status_final_2024_121825.csv"))

## for plotting:
rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

region_levels <- paste0("Region_", 1:9)

radar_df <- all_cs_indicators_2024 |>
  dplyr::filter(!is.na(tr_current_status)) |>
  dplyr::mutate(
    rgn_id = paste0("Region_", rgn_id),
    rgn_id = factor(rgn_id, levels = paste0("Region_", 1:9))
  )

ggplot(
  radar_df,
  aes(
    x = rgn_id,
    y = tr_current_status,
    fill = rgn_id
  )
) +
  geom_col(
    width = 0.98,
    color = "white",
    linewidth = 0.4
  ) +
  geom_hline(
    yintercept = 100,
    color = "grey60",
    linewidth = 0.6
  ) +
  coord_polar(start = -pi / 2) +
  scale_y_continuous(
    limits = c(0, 100),
    expand = c(0, 0)
  ) +
  scale_fill_manual(values = rgn_colors) +
  theme_void() +
  theme(
    legend.position = "none"
  )




```

```{r}
library(dplyr)
library(ggplot2)
library(readr)
library(here)

all_cs_indicators_2024 <- read_csv(here::here(
  "/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output",
  "tr_current_status_final_2024_121825.csv"
))

rgn_labels <- c(
  "Region_1" = "Western Upper Gulf",
  "Region_2" = "Western Midriff Islands",
  "Region_3" = "Western Central Gulf",
  "Region_4" = "Western Southern Gulf",
  "Region_5" = "Eastern Upper Gulf",
  "Region_6" = "Eastern Midriff Islands",
  "Region_7" = "Southern Sonora Coast",
  "Region_8" = "Sinaloa Coast",
  "Region_9" = "Southeastern Gulf"
)

rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)


radar_df <- all_cs_indicators_2024 |>
  dplyr::filter(!is.na(tr_current_status)) |>
  dplyr::arrange(rgn_id) |>
  dplyr::mutate(
    rgn_id = factor(rgn_id, levels = unique(rgn_id))
  )

ggplot(
  radar_df,
  aes(
    x = rgn_id,
    y = tr_current_status,
    fill = rgn_id
  )
) +
  geom_col(
    width = 0.98,
    color = "white",
    linewidth = 0.4,
    alpha = 0.9
  ) +
  geom_hline(
    yintercept = 100,
    color = "grey60",
    linewidth = 0.6
  ) +
  coord_polar(start = -pi / 2, clip = "off") +
  scale_y_continuous(
    limits = c(0, 100),
    breaks = c(25, 50, 75, 100),
    expand = c(0, 0)
  ) +
  # scale_x_discrete(
  #   labels = rgn_labels        # names must match factor levels
  # ) +
  scale_fill_manual(values = rgn_colors) +
  labs(
    title = "Tourism & Recreation current status by region (2024)",
    y = NULL,
    x = NULL
  ) +
  theme_minimal(base_family = "Calibri") +
  theme(
    plot.margin = margin(t = 15, r = 15, b = 15, l = 15),  # <- more space
    plot.title  = element_text(face = "bold", size = 14),
  #   axis.text.x = element_text(
  #   face   = "bold",
  #   size   = 8,
  #   color  = "black",
  #   angle  = 10,      # try 0, 20, 30, 40, etc.
  #   vjust  = 5,
  #   hjust  = 10,
  #   margin = margin(t = -5)
  # ),
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank(), 
    axis.text.y = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(color = "grey90"),
    legend.position = "none"
  ) 
  # geom_text(
  #   aes(label = round(tr_current_status, 1)),
  #   color = "black",
  #   size = 3,
  #   vjust = -0.2,
  #   family = "Calibri"
  # )




```

