---
title: 'OHI GoC 2025: QUALITY Tourism & Recreation'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: inline
editor_options: 
  chunk_output_type: inline
---

# TR Quality

## Rate of Return

Not possible unfortunately. Lack of data availability.

## Service Establishments

See `TR/denue_prep.qmd` and `TR/servicios_de_alojamiento.qmd`.


Read in all service establishments.

```{r}
library(tidyverse)
library(sf)
library(mapview)
library(here)
```

### Culture service establishments 

```{r}
output_dir <- here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_esparcimiento_culturales_y_deportivos_servicios_recreativos/int")

shp_files <- list.files(output_dir, pattern = "\\.shp$", full.names = TRUE)

# loop through and assign each to an object named after the file
for (s in shp_files) {

  # extract the clean object name
  obj_name <- s %>%
    basename() %>%
    str_remove("\\.shp$")

  message("Reading in: ", obj_name)
  obj_data <- st_read(s, options = "ENCODING=LATIN1", quiet = TRUE)
  assign(obj_name, obj_data, envir = .GlobalEnv)
}

```

```{r}
# # all combined datasets in a named list
# culture_combined_list <- list(
#   museums = museums_2019_2024_goc_50km,
#   historical_sites = historical_sites_2019_2024_goc_50km,
#   nature_heritage = nature_heritage_2019_2024_goc_50km,
#   nautical_services = marine_tourism_2019_2024_goc_50km
#   # recreation_other = recreation_other_2019_2024_goc_50km
# )
# 
# culture_combined_df <- bind_rows(culture_combined_list)
# 
# culture_counts_df <- culture_counts %>% 
#   st_drop_geometry()
# 
# culture_counts_plotly <- plot_ly(culture_counts_df,
#         x = ~data_year,
#         y = ~n,
#         color = ~category,
#         type = 'scatter',
#         mode = 'lines+markers') %>%
#   layout(
#     title = "DENUE observations by year (50km buffer)",
#     xaxis = list(title = "Data year"),
#     yaxis = list(title = "Total observations"),
#     legend = list(title=list(text="Recreation type"))
#   )
# culture_counts_plotly
```


```{r}
culture_gulf_50km <- bind_rows(historical_sites_2019_2024_goc_50km, marine_tourism_2019_2024_goc_50km, museums_2019_2024_goc_50km, nature_heritage_2019_2024_goc_50km, recreation_other_2019_2024_goc_50km)


culture_gulf_50km_clean <- culture_gulf_50km %>% 
  st_drop_geometry() %>% 
  mutate(nombre_act = nombre_act %>% 
           str_replace_all("<fa>", "ú") %>%
           str_replace_all("<e1>", "á") %>%
           str_replace_all("<ed>", "í") %>%
           str_replace_all("<f1>", "ñ") %>%
           str_replace_all("<e9>", "é"),
         
         municipio = municipio %>%
           str_replace_all("<fa>", "ú") %>%
           str_replace_all("<e1>", "á") %>%
           str_replace_all("<ed>", "í") %>%
           str_replace_all("<f1>", "ñ") %>%
           str_replace_all("<e9>", "é")
  ) %>% 
  mutate(municipio_clean = municipio %>% 
           str_to_lower() %>%
           str_replace_all(" ", "_"),
         municipio_clean = case_when(
           municipio == "San Ignacio Río Muerto" ~ "cajeme",
           municipio == "Bácum" ~ "cajeme",
           municipio == "Mexicali" ~ "san_luis_río_colorado",
           municipio == "Sinaloa" ~ "guasave",
           municipio == "Santa María del Oro" ~ "san_pedro_lagunillas",
           TRUE ~ municipio_clean
           )) %>% 
  mutate(
    employees_estimated = case_when(
      per_ocu == "0 a 5 personas" ~ 3,
      per_ocu == "6 a 10 personas" ~ 8,
      per_ocu == "11 a 30 personas" ~ 20.5,
      per_ocu == "31 a 50 personas" ~ 40.5,
      per_ocu == "51 a 100 personas" ~ 75.5,
      per_ocu == "101 a 250 personas" ~ 175.5,
      per_ocu %in% c("251 y m<e1>s personas", "251 y más personas") ~ 300,
      TRUE ~ NA_real_
    )
  ) %>% 
  group_by(municipio_clean, data_year, rgn_id) %>% 
  summarize(agg_employees = sum(employees_estimated)) %>% 
  ungroup()
```

Now, join with tourist arrivals. 

```{r}
arrivals_filled_by_DENUE_accommodation_employees <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/gapfilled_int","arrivals_filled_by_DENUE_accommodation_employees_2019_2024.csv"))

gapfilled_total_arrivals <- arrivals_filled_by_DENUE_accommodation_employees %>%
  mutate(country = "Mexico") %>% 
  select(data_year, municipio_clean, arrivals_merged, is_filled) 

joined_arrivals_culture_estab_employees <- left_join(culture_gulf_50km_clean, gapfilled_total_arrivals, by = c("data_year","municipio_clean")) %>% 
  filter(municipio_clean != "jala")
```

```{r}
culture_agg_by_rgn_emp_arr <- joined_arrivals_culture_estab_employees %>% 
  select(rgn_id, data_year, agg_employees, arrivals_merged) %>% 
  group_by(rgn_id, data_year) %>% 
  summarize(rgn_agg_emp = sum(agg_employees),
            rgn_agg_arr = sum(arrivals_merged)) 

culture_activity_emp_percap_tourist <- culture_agg_by_rgn_emp_arr %>% 
  group_by(rgn_id, data_year) %>% 
  mutate(emp_per_tourist = rgn_agg_emp / rgn_agg_arr) %>% 
  ungroup()

## plot the emp/tourist over time
culture_gulf_emp_per_tourist_plotly <- plotly::plot_ly(
  culture_activity_emp_percap_tourist, 
  x = ~data_year, 
  y = ~emp_per_tourist, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Culture & Gulf-related establishment employee per capita tourist arrival",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "emp_per_tourist")
  ) 
culture_gulf_emp_per_tourist_plotly

```

Now let's actually calculate current status for each region and year. The reference point is the 90th quantile?

current status = emp_per_tourist (current) / emp_per_tourist (90th quantile of all regions and years)

Where every region and year is compared to the highest.

```{r}
## find the 90th quantile
culture_activity_emp_percap_tourist_cs <- culture_activity_emp_percap_tourist %>% 
  mutate(percap_90th = quantile(emp_per_tourist, probs = 0.9),
         current_status = (emp_per_tourist / percap_90th) * 100) %>% 
  mutate(score = ifelse(current_status > 100, 100, current_status))

# write_csv(culture_activity_emp_percap_tourist_cs, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quality","tr_quality_culture_current_status_quantile_121625.csv"))
```


```{r}
library(plotly)

## plot the score over time
culture_quantile_score_plotly <- plotly::plot_ly(
  culture_activity_emp_percap_tourist_cs, 
  x = ~data_year, 
  y = ~score, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: Culture & Gulf-related activity availability -- employees in service establishments per tourist arrival, scaled to the 90th quantile",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(culture_activity_emp_percap_tourist_cs$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
culture_quantile_score_plotly

```
Using regional-specific 2019 reference:

```{r}
hist_culture_activity_emp_percap_tourist_cs <- culture_activity_emp_percap_tourist %>% 
  group_by(rgn_id) %>% 
  mutate(hist_ref = emp_per_tourist[data_year == 2019],
         current_status = (emp_per_tourist / hist_ref) * 100) %>% 
  mutate(score = ifelse(current_status > 100, 100, current_status))

# write_csv(hist_culture_activity_emp_percap_tourist_cs, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quality","tr_quality_culture_current_status_temporal2019_121625.csv"))
```

```{r}

## plot the score over time
culture_hist_score_plotly <- plotly::plot_ly(
  hist_culture_activity_emp_percap_tourist_cs, 
  x = ~data_year, 
  y = ~score, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: Culture & Gulf-related activity availability -- employees in service establishments per tourist arrival, reference of 2019",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(hist_culture_activity_emp_percap_tourist_cs$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
culture_hist_score_plotly

```

### Food and drink establishments

comida_bebidas_categories <- c(
  "Cafeterías, fuentes de sodas, neverías, refresquerías y similares",
  "Restaurantes con servicio de preparación de alimentos a la carta o de comida corrida",
  "Servicios de preparación de otros alimentos para consumo inmediato",
  "Restaurantes con servicio de preparación de antojitos",
  "Restaurantes con servicio de preparación de tacos y tortas",
  "Restaurantes con servicio de preparación de pizzas, hamburguesas, hot dogs y pollos rostizados para llevar",
  "Restaurantes con servicio de preparación de pescados y mariscos",
  "Bares, cantinas y similares",
  "Restaurantes de autoservicio",
  "Restaurantes que preparan otro tipo de alimentos para llevar",
  "Centros nocturnos, discotecas y similares",
  "Servicios de preparación de alimentos en unidades móviles",
  "Servicios de preparación de alimentos para ocasiones especiales",
  "Servicios de comedor para empresas e instituciones"
)

English translation:
  “Cafeterias, soda fountains, ice cream parlors, refreshment stands, and similar establishments,”
  “Restaurants offering à la carte or set menu food service,”
  “Other food preparation services for immediate consumption,”
  “Restaurants offering appetizer preparation services,”
  “Restaurants offering taco and torta preparation services,”
  “Restaurants offering pizza, hamburger, hot dog, and roast chicken preparation services for takeout,”
  “Restaurants offering fish and seafood preparation services,”
  “Bars, cantinas, and similar establishments,”
  “Self-service restaurants,”
  “Restaurants preparing other types of food for takeaway,”
  “Nightclubs, discos, and similar establishments,”
  “Mobile food preparation services,”
  “Food preparation services for special occasions,”
  “Catering services for companies and institutions.”
  
For now, let's include all of the types of food and drink service establishments.

```{r}
comida_bebidas_50km <- st_read(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas/int","update_comida_bebidas_2019_2024_goc_50km.shp"))


comida_bebidas_50km_clean <- comida_bebidas_50km %>% 
  st_drop_geometry() %>% 
  mutate(nombre_act = nombre_act %>% 
           str_replace_all("<fa>", "ú") %>%
           str_replace_all("<e1>", "á") %>%
           str_replace_all("<ed>", "í") %>%
           str_replace_all("<f1>", "ñ") %>%
           str_replace_all("<e9>", "é"),
         
         municipio = municipio %>%
           str_replace_all("<fa>", "ú") %>%
           str_replace_all("<e1>", "á") %>%
           str_replace_all("<ed>", "í") %>%
           str_replace_all("<f1>", "ñ") %>%
           str_replace_all("<e9>", "é")
  ) %>% 
  mutate(municipio_clean = municipio %>% 
           str_to_lower() %>%
           str_replace_all(" ", "_"),
         municipio_clean = case_when(
           municipio == "San Ignacio Río Muerto" ~ "cajeme",
           municipio == "Bácum" ~ "cajeme",
           municipio == "Mexicali" ~ "san_luis_río_colorado",
           municipio == "Sinaloa" ~ "guasave",
           municipio == "Santa María del Oro" ~ "san_pedro_lagunillas",
           TRUE ~ municipio_clean
           )) %>% 
  mutate(
    employees_estimated = case_when(
      per_ocu == "0 a 5 personas" ~ 3,
      per_ocu == "6 a 10 personas" ~ 8,
      per_ocu == "11 a 30 personas" ~ 20.5,
      per_ocu == "31 a 50 personas" ~ 40.5,
      per_ocu == "51 a 100 personas" ~ 75.5,
      per_ocu == "101 a 250 personas" ~ 175.5,
      per_ocu %in% c("251 y m<e1>s personas", "251 y más personas") ~ 300,
      TRUE ~ NA_real_
    )
  ) %>% 
  group_by(municipio_clean, data_year, rgn_id) %>% 
  summarize(agg_employees = sum(employees_estimated)) %>% 
  ungroup()
```



```{r}
# comida_bebidas_50km_df <- all_comida_bebidas_50km %>% 
#   st_drop_geometry() %>% 
#   group_by(data_year, municipio) %>% 
#   summarize(total_obs = n()) %>% 
#   ungroup()
# 
# food_drink_plotly <- plot_ly(comida_bebidas_50km_df,
#         x = ~data_year,
#         y = ~total_obs,
#         color = ~municipio,
#         type = 'scatter',
#         mode = 'lines+markers') %>%
#   layout(
#     title = "DENUE comida & bebida observations by year (50km buffer)",
#     xaxis = list(title = "Data year"),
#     yaxis = list(title = "Total observations")
#   )
# food_drink_plotly
```

Now, join with tourist arrivals. 

```{r}
arrivals_filled_by_DENUE_accommodation_employees <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/gapfilled_int","arrivals_filled_by_DENUE_accommodation_employees_2019_2024.csv"))

gapfilled_total_arrivals <- arrivals_filled_by_DENUE_accommodation_employees %>%
  mutate(country = "Mexico") %>% 
  select(data_year, municipio_clean, arrivals_merged, is_filled) 

joined_arrivals_food_bev_estab_employees <- left_join(comida_bebidas_50km_clean, gapfilled_total_arrivals, by = c("data_year","municipio_clean"))
```

```{r}
food_bev_agg_by_rgn_emp_arr <- joined_arrivals_food_bev_estab_employees %>% 
  select(rgn_id, data_year, agg_employees, arrivals_merged) %>% 
  group_by(rgn_id, data_year) %>% 
  summarize(rgn_agg_emp = sum(agg_employees),
            rgn_agg_arr = sum(arrivals_merged)) 

food_bev_activity_emp_percap_tourist <- food_bev_agg_by_rgn_emp_arr %>% 
  group_by(rgn_id, data_year) %>% 
  mutate(emp_per_tourist = rgn_agg_emp / rgn_agg_arr) %>% 
  ungroup()

## plot the emp/tourist over time
food_bev_emp_per_tourist_plotly <- plotly::plot_ly(
  food_bev_activity_emp_percap_tourist, 
  x = ~data_year, 
  y = ~emp_per_tourist, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Food & beverage establishment employee per capita tourist arrival",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "emp_per_tourist")
  ) 
food_bev_emp_per_tourist_plotly

```

ow let's actually caluclate current status for each region and year. The reference point is the 90th quantile?

current status = emp_per_tourist (current) / emp_per_tourist (90th quantile of all regions and years)

Where every region and year is compared to the highest 

```{r}
## find the 90th quantile
food_bev_activity_emp_percap_tourist_cs <- food_bev_activity_emp_percap_tourist %>% 
  mutate(percap_90th = quantile(emp_per_tourist, probs = 0.9),
         current_status = (emp_per_tourist / percap_90th) * 100) %>% 
  mutate(score = ifelse(current_status > 100, 100, current_status))

# write_csv(food_bev_activity_emp_percap_tourist_cs, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quality","tr_quality_food_bev_current_status_quantile_121125.csv"))
```


```{r}
library(plotly)

## plot the score over time
food_bev_quantile_score_plotly <- plotly::plot_ly(
  food_bev_activity_emp_percap_tourist_cs, 
  x = ~data_year, 
  y = ~score, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: Food & beverage activity availability -- employees in service establishments per tourist arrival, scaled to the 90th quantile",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(food_bev_activity_emp_percap_tourist_cs$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
food_bev_quantile_score_plotly

```
Using regional-specific 2019 reference:

```{r}
hist_food_bev_activity_emp_percap_tourist_cs <- food_bev_activity_emp_percap_tourist %>% 
  group_by(rgn_id) %>% 
  mutate(hist_ref = emp_per_tourist[data_year == 2019],
         current_status = (emp_per_tourist / hist_ref) * 100) %>% 
  mutate(score = ifelse(current_status > 100, 100, current_status))

# write_csv(hist_food_bev_activity_emp_percap_tourist_cs, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quality","tr_quality_food_bev_current_status_temporal2019_121625.csv"))
```

```{r}

## plot the score over time
food_bev_hist_score_plotly <- plotly::plot_ly(
  hist_food_bev_activity_emp_percap_tourist_cs, 
  x = ~data_year, 
  y = ~score, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: Food & beverage activity availability -- employees in service establishments per tourist arrival, reference of 2019",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(hist_food_bev_activity_emp_percap_tourist_cs$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
food_bev_hist_score_plotly

```




### Other recreation service establishments!

recreacion_categorias <- c(
  "Centros de acondicionamiento físico del sector privado",
  "Centros de acondicionamiento físico del sector público",
  "Clubes deportivos del sector privado",
  "Clubes deportivos del sector público",
  "Equipos deportivos profesionales",
  "Deportistas profesionales",
  "Clubes o ligas de aficionados",
  "Campos de golf",
  "Artistas, escritores y técnicos independientes",
  "Compañías de danza del sector privado",
  "Compañías de teatro del sector público",
  "Compañías de teatro del sector privado",
  "Grupos musicales del sector público",
  "Cantantes y grupos musicales del sector privado",
  "Otras compañías y grupos de espectáculos artísticos del sector privado",
  "Parques de diversiones y temáticos del sector privado",
  "Parques de diversiones y temáticos del sector público",
  "Parques acuáticos y balnearios del sector privado",
  "Parques acuáticos y balnearios del sector público",
  "Billares",
  "Casas de juegos electrónicos",
  "Boliches",
  "Otros servicios recreativos prestados por el sector privado",
  "Otros servicios recreativos prestados por el sector público",
  "Venta de billetes de lotería, pronósticos deportivos y otros boletos de sorteo",
  "Otros juegos de azar",
  "Casinos"
)


```{r}
other_rec_50km <- st_read(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_esparcimiento_culturales_y_deportivos_servicios_recreativos/int/recreation_other_2019_2024_goc_50km.shp"))

other_rec_50km_clean <- other_rec_50km %>% 
  st_drop_geometry() %>% 
  mutate(municipio_clean = municipio %>% 
           str_to_lower() %>%
           str_replace_all(" ", "_"),
         municipio_clean = case_when(
           municipio == "San Ignacio Río Muerto" ~ "cajeme",
           municipio == "Bácum" ~ "cajeme",
           municipio == "Mexicali" ~ "san_luis_río_colorado",
           municipio == "Sinaloa" ~ "guasave",
           municipio == "Santa María del Oro" ~ "san_pedro_lagunillas",
           TRUE ~ municipio_clean
           )) %>% 
  mutate(
    employees_estimated = case_when(
      per_ocu == "0 a 5 personas" ~ 3,
      per_ocu == "6 a 10 personas" ~ 8,
      per_ocu == "11 a 30 personas" ~ 20.5,
      per_ocu == "31 a 50 personas" ~ 40.5,
      per_ocu == "51 a 100 personas" ~ 75.5,
      per_ocu == "101 a 250 personas" ~ 175.5,
      per_ocu %in% c("251 y m<e1>s personas", "251 y más personas") ~ 300,
      TRUE ~ NA_real_
    )
  ) %>% 
  group_by(municipio_clean, data_year, rgn_id) %>% 
  summarize(agg_employees = sum(employees_estimated)) %>% 
  ungroup()
```

Now, join with tourist arrivals. 

```{r}
arrivals_filled_by_DENUE_accommodation_employees <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/gapfilled_int","arrivals_filled_by_DENUE_accommodation_employees_2019_2024.csv"))

gapfilled_total_arrivals <- arrivals_filled_by_DENUE_accommodation_employees %>%
  mutate(country = "Mexico") %>% 
  select(data_year, municipio_clean, arrivals_merged, is_filled) 

joined_arrivals_other_rec_estab_employees <- left_join(other_rec_50km_clean, gapfilled_total_arrivals, by = c("data_year","municipio_clean")) %>% 
  filter(municipio_clean != "jala")
```

```{r}
other_rec_agg_by_rgn_emp_arr <- joined_arrivals_other_rec_estab_employees %>% 
  select(rgn_id, data_year, agg_employees, arrivals_merged) %>% 
  group_by(rgn_id, data_year) %>% 
  summarize(rgn_agg_emp = sum(agg_employees),
            rgn_agg_arr = sum(arrivals_merged)) 

other_rec_activity_emp_percap_tourist <- other_rec_agg_by_rgn_emp_arr %>% 
  group_by(rgn_id, data_year) %>% 
  mutate(emp_per_tourist = rgn_agg_emp / rgn_agg_arr) %>% 
  ungroup()

## plot the emp/tourist over time
other_rec_gulf_emp_per_tourist_plotly <- plotly::plot_ly(
  other_rec_activity_emp_percap_tourist, 
  x = ~data_year, 
  y = ~emp_per_tourist, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Other recreational establishment employee per capita tourist arrival",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "emp_per_tourist")
  ) 
other_rec_gulf_emp_per_tourist_plotly

```

Now let's actually calculate current status for each region and year. The reference point is the 90th quantile?

current status = emp_per_tourist (current) / emp_per_tourist (90th quantile of all regions and years)

Where every region and year is compared to the highest.

```{r}
## find the 90th quantile
other_rec_activity_emp_percap_tourist_cs <- other_rec_activity_emp_percap_tourist %>% 
  mutate(percap_90th = quantile(emp_per_tourist, probs = 0.9),
         current_status = (emp_per_tourist / percap_90th) * 100) %>% 
  mutate(score = ifelse(current_status > 100, 100, current_status))

# write_csv(other_rec_activity_emp_percap_tourist_cs, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quality","tr_quality_other_rec_current_status_quantile_121225.csv"))
```


```{r}
library(plotly)

## plot the score over time
other_rec_quantile_score_plotly <- plotly::plot_ly(
  other_rec_activity_emp_percap_tourist_cs, 
  x = ~data_year, 
  y = ~score, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: Other recreation activity availability -- employees in service establishments per tourist arrival, scaled to the 90th quantile",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(other_rec_activity_emp_percap_tourist_cs$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
other_rec_quantile_score_plotly

```

Using regional-specific 2019 reference:

```{r}
hist_other_rec_activity_emp_percap_tourist_cs <- other_rec_activity_emp_percap_tourist %>% 
  group_by(rgn_id) %>% 
  mutate(hist_ref = emp_per_tourist[data_year == 2019],
         current_status = (emp_per_tourist / hist_ref) * 100) %>% 
  mutate(score = ifelse(current_status > 100, 100, current_status))

# write_csv(hist_other_rec_activity_emp_percap_tourist_cs, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quality","tr_quality_other_rec_current_status_temporal2019_121625.csv"))
```

```{r}

## plot the score over time
other_rec_hist_score_plotly <- plotly::plot_ly(
  hist_other_rec_activity_emp_percap_tourist_cs, 
  x = ~data_year, 
  y = ~score, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: Other recreation activity availability -- employees in service establishments per tourist arrival, reference of 2019",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(hist_other_rec_activity_emp_percap_tourist_cs$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
other_rec_hist_score_plotly

```

### Merging all activities

Quality = avg(crime/coastal pop + activities)
activities = avg(2/6 * food&drink emp / tourist arrivals + 3/6 * cultural&gulf-related / tourist arrivals + 1/6 * other_recreational / tourist arrivals)

```{r}
food_bev_current_status <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quality","tr_quality_food_bev_current_status_temporal2019_121625.csv")) %>% 
  mutate(activity_type = "food_bev")

## plot the score over time
food_bev_current_status_plotly <- plotly::plot_ly(
  food_bev_current_status, 
  x = ~data_year, 
  y = ~score, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "food&bev, reference of 2019",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(food_bev_current_status$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
food_bev_current_status_plotly
```


```{r}
culture_current_status <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quality","tr_quality_culture_current_status_temporal2019_121625.csv")) %>% 
  mutate(activity_type = "culture")

## plot the score over time
culture_current_status_plotly <- plotly::plot_ly(
  culture_current_status, 
  x = ~data_year, 
  y = ~score, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "culture&gulf, reference of 2019",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(culture_current_status$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
culture_current_status_plotly
```


```{r}
other_rec_current_status <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quality","tr_quality_other_rec_current_status_temporal2019_121625.csv")) %>% 
  mutate(activity_type = "other_rec")

## plot the score over time
other_rec_current_status_plotly <- plotly::plot_ly(
  other_rec_current_status, 
  x = ~data_year, 
  y = ~score, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "food&bev, reference of 2019",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(other_rec_current_status$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
other_rec_current_status_plotly
```

```{r}
merge_activites <- bind_rows(food_bev_current_status, culture_current_status, other_rec_current_status) %>% 
  select(rgn_id, year = data_year, score, activity_type) %>% 
  pivot_wider(
    names_from = activity_type,
    values_from = score
  )

score_merge_activites <- merge_activites %>%
  mutate(
    final_score =
      (2/5) * food_bev +
      (2/5) * culture +
      (1/5) * other_rec
  )

write_csv(score_merge_activites, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quality","final_activity_avg_score_121625.csv"))
```


##### activities barplot viz
```{r}
viz_activities_2024_plot <- score_merge_activites %>%
  separate(col = rgn_id, into = c("rgn", "rgn_id_num"), sep = "_", remove = FALSE) %>% 
  filter(year == 2024) %>% 
  select(rgn_id, rgn_id_num, final_score) %>% 
  arrange(final_score) %>% 
  mutate(
    rgn_id_num = factor(rgn_id_num, levels = rgn_id_num)
    ) 


rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

a <- ggplot(
  viz_activities_2024_plot,
  aes(x = rgn_id_num, y = final_score, fill = rgn_id)
) +
  geom_col(color = "black", linewidth = 0.4) +
  scale_fill_manual(values = rgn_colors) +
  labs(
    title = "Current activity availability compared to pre-COVID levels",
    x = " ",
    y = "Activity availability relative to 2019 (%)"
  ) +
  theme_minimal(base_family = "Calibri") +
  theme(
    plot.title = element_text(face = "bold", size = 18),
    axis.title.y = element_text(face = "bold", size = 16),
    axis.text.x = element_text(face = "bold", size = 16),
    axis.text.y = element_text(face = "bold", size = 16),
    axis.title.x = element_blank(),
    legend.position = "none"
  ) +
  geom_hline(
  yintercept = 100,
  linetype = "dashed",
  linewidth = 0.8,
  color = "black"
)

a

ggsave(
  filename = here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_viz",
    "activity_availability_by_region_2024.png"
  ),
  plot = a,
  width = 7,
  height = 4,
  units = "in",
  dpi = 300
)

```

Instead, a map for viz.  

```{r}
#### join with ohi rgn polygons
ohi <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp")) %>%
  filter(location == "marine") %>% 
  select(rgn_id)

viz_activities_2024_ohigns <- full_join(viz_activities_2024_plot, ohi, by = "rgn_id") %>% 
  st_as_sf()

mapview(
  viz_activities_2024_ohigns, 
  zcol = "final_score",
  col.regions = colorRampPalette(c( '#FEE090', '#ABD9E9', '#4575B4', '#313695'
)),
  layer.name = "Current activity availability (2024) compared to pre-COVID levels (2019)",
  legend = TRUE
)

### or, using rgn colors, and adding the numbers on the ppt?
ohi <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp")) %>%
  filter(location == "marine")

rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

ohi <- ohi %>%
  mutate(rgn_id = factor(rgn_id, levels = names(rgn_colors)))

mapview::mapview(
  ohi,
  zcol = "rgn_id",
  col.regions = rgn_colors,
  legend = TRUE
)
```



```{r}
## plot the score over time
score_merge_activites_plotly <- plotly::plot_ly(
  score_merge_activites, 
  x = ~year, 
  y = ~final_score, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Merged activities, reference of 2019",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(score_merge_activites$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
score_merge_activites_plotly

```

## Public bathroom availability

Read in the data:

```{r}
library(tidyverse)
library(sf)
library(mapview)

banos_publicos_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_public_bathrooms","banos_publicos_812130_stablecimientoseconomicos.csv"))

banos_publicos_sf <- banos_publicos_raw %>% 
  st_as_sf(coords = c("Longitud", "Latitud"), # making sf object
           crs = 4326, # WGS84
           remove = FALSE) %>% 
  janitor::clean_names()


mapview(banos_publicos_sf)
```

```{r}
inland_50 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer","inland_50km_goc.shp"))

banos_publicos_sf_trans <- st_transform(banos_publicos_sf, st_crs(inland_50))
banos_publicos_goc <- st_intersection(banos_publicos_sf_trans, inland_50)
mapview(banos_publicos_goc, zcol = "descripcion_del_sector_economico")
```



## Crime

### Read in SESNSP GoC data

```{r}
library(tidyverse)
library(here)
library(sf)
library(mapview)
```


```{r}
sesnsp_goc_10km <- st_read(here::here("/home/shares/ohi/OHI_GOC/goal_prep/ss/int","sesnsp_goc_municipios_10km.shp")) ## need to fix the names / make a document describing the abbreviations!!

unique(sesnsp_goc_10km$rgn_id) # for 10km: [1] "Region_1" "Region_3" "Region_4" "Region_5" "Region_7" "Region_8" [7] "Region_9"
# for 50km: [1] "Region_1" "Region_3" "Region_4" "Region_5" "Region_7" "Region_8" "Region_9"
```

Out of the crimes listed in `SS/v2025/int_sesnsp/types_of_crimes_goc_sesnsp.csv`, these are the ones I think affect tourism quality the most:

bien_juridico_afectado	tipo_de_delito	subtipo_de_delito
El patrimonio	Daño a la propiedad	Daño a la propiedad
El patrimonio	Extorsión	Extorsión
El patrimonio	Fraude	Fraude
El patrimonio	Otros delitos contra el patrimonio	Otros delitos contra el patrimonio
El patrimonio	Robo	Otros robos
El patrimonio	Robo	Robo a casa habitación
El patrimonio	Robo	Robo a institución bancaria
El patrimonio	Robo	Robo a negocio
El patrimonio	Robo	Robo a transeúnte en espacio abierto al público
El patrimonio	Robo	Robo a transeúnte en vía pública
El patrimonio	Robo	Robo a transportista
El patrimonio	Robo	Robo de autopartes
El patrimonio	Robo	Robo de maquinaria
El patrimonio	Robo	Robo de vehículo automotor
El patrimonio	Robo	Robo en transporte individual
El patrimonio	Robo	Robo en transporte público colectivo
El patrimonio	Robo	Robo en transporte público individual
La familia	Violencia de género en todas sus modalidades distinta a la violencia familiar	Violencia de género en todas sus modalidades distinta a la violencia familiar
La libertad y la seguridad sexual	Abuso sexual	Abuso sexual
La libertad y la seguridad sexual	Acoso sexual	Acoso sexual
La libertad y la seguridad sexual	Hostigamiento sexual	Hostigamiento sexual
La libertad y la seguridad sexual	Otros delitos que atentan contra la libertad y la seguridad sexual	Otros delitos que atentan contra la libertad y la seguridad sexual
La libertad y la seguridad sexual	Violación equiparada	Violación equiparada
La libertad y la seguridad sexual	Violación simple	Violación simple
La sociedad	Corrupción de menores	Corrupción de menores
La sociedad	Otros delitos contra la sociedad	Otros delitos contra la sociedad
La sociedad	Trata de personas	Trata de personas
La vida y la Integridad corporal	Feminicidio	Feminicidio
La vida y la Integridad corporal	Homicidio	Homicidio doloso
La vida y la Integridad corporal	Lesiones	Lesiones dolosas
La vida y la Integridad corporal	Otros delitos que atentan contra la vida y la integridad corporal	Otros delitos que atentan contra la vida y la integridad corporal
Libertad personal	Otros delitos que atentan contra la libertad personal	Otros delitos que atentan contra la libertad personal
Libertad personal	Rapto	Rapto
Libertad personal	Secuestro	Secuestro
Libertad personal	Tráfico de menores	Tráfico de menores
Otros bienes jurídicos afectados (del fuero común)	Amenazas	Amenazas
Otros bienes jurídicos afectados (del fuero común)	Contra el medio ambiente	Contra el medio ambiente
Otros bienes jurídicos afectados (del fuero común)	Delitos cometidos por servidores públicos	Delitos cometidos por servidores públicos
Otros bienes jurídicos afectados (del fuero común)	Evasión de presos	Evasión de presos
Otros bienes jurídicos afectados (del fuero común)	Narcomenudeo	Narcomenudeo
Otros bienes jurídicos afectados (del fuero común)	Otros delitos del Fuero Común	Otros delitos del Fuero Común


```{r}
# vector of subtypes I want to keep for TR quality
tr_qual_crime_subtypes <- c(
  "Daño a la propiedad",
  "Extorsión",
  "Fraude",
  "Otros delitos contra el patrimonio",
  "Otros robos",
  "Robo a casa habitación",
  "Robo a institución bancaria",
  "Robo a negocio",
  "Robo a transeúnte en espacio abierto al público",
  "Robo a transeúnte en vía pública",
  "Robo a transportista",
  "Robo de autopartes",
  "Robo de maquinaria",
  "Robo de vehículo automotor",
  "Robo en transporte individual",
  "Robo en transporte público colectivo",
  "Robo en transporte público individual",
  "Violencia de género en todas sus modalidades distinta a la violencia familiar",
  "Abuso sexual",
  "Acoso sexual",
  "Hostigamiento sexual",
  "Otros delitos que atentan contra la libertad y la seguridad sexual",
  "Violación equiparada",
  "Violación simple",
  "Corrupción de menores",
  "Otros delitos contra la sociedad",
  "Trata de personas",
  "Feminicidio",
  "Homicidio doloso",
  "Lesiones dolosas",
  "Otros delitos que atentan contra la vida y la integridad corporal",
  "Otros delitos que atentan contra la libertad personal",
  "Rapto",
  "Secuestro",
  "Tráfico de menores",
  "Amenazas",
  "Contra el medio ambiente",
  "Delitos cometidos por servidores públicos",
  "Evasión de presos",
  "Narcomenudeo",
  "Otros delitos del Fuero Común"
)

# filter sesnsp_goc by those subtypes
sesnsp_goc_tr_qual <- sesnsp_goc_10km %>%
  filter(sbtp_d_ %in% tr_qual_crime_subtypes)

unique(sesnsp_goc_tr_qual$rgn_id) # [1] "Region_1" "Region_3" "Region_4" "Region_5" "Region_7" "Region_8" "Region_9"

```

### Now, summarize the number of crimes by year and region (assuming that all crimes are weighted equally)

```{r}
sesnsp_goc_tr_qual_sum <- sesnsp_goc_tr_qual %>% 
  st_drop_geometry() %>% 
  group_by(ano, rgn_id) %>% 
  summarise(number_crimes = sum(annl_tt))

unique(sesnsp_goc_tr_qual_sum$rgn_id) # [1] "Region_1" "Region_3" "Region_4" "Region_5" "Region_7" "Region_8" "Region_9"
```

#### Divide by coastal population in each region

##### NOTE: NEED TO UPDATE...

1. We will have to update this because we eventually want to use 2015, 2020, and 2025 census data for population, not just static 2020.
2. I think maybe we should use the population per municipality (instead of population per OHI Region).  Because for some regions there are multiple municipios in the region that have data collected for them, versus some only have one or two.  

```{r}
coastal_10km_pop <- read_csv(here::here("general","OHI_rgn_population_10km.csv")) 

coastal_10km_pop_sumrgn9 <- coastal_10km_pop %>% 
  group_by(rgn_id) %>% 
  summarize(rgn_population_10km = sum(population_10km_2020))

sesnsp_goc_tr_sum_percapita <- left_join(sesnsp_goc_tr_qual_sum, coastal_10km_pop_sumrgn9, by = "rgn_id")

sesnsp_goc_tr_crimes_percapita <- sesnsp_goc_tr_sum_percapita %>% 
  group_by(ano, rgn_id) %>% 
  mutate(crimes_percapita = 100 * (number_crimes / rgn_population_10km))
```

#### Prelim Scores

```{r}
max_crime_percapita_allrgns <- max(sesnsp_goc_tr_crimes_percapita$crimes_percapita, na.rm = TRUE)

tr_qual_crime_score_alt <- sesnsp_goc_tr_crimes_percapita %>%
  mutate(
    current_status = (1 - (crimes_percapita / max_crime_percapita_allrgns)) * 100
  ) 

write_csv(tr_qual_crime_score, here::here("TR/tr_quality_int","tr_qual_crime_score_111025.csv"))
```

#### Plotly

```{r}
library(plotly)

## TR quality crime index score over time
tr_qual_crime_plotly <- plotly::plot_ly(
  tr_qual_crime_score,
  x = ~ano, 
  y = ~current_status, 
  color = ~rgn_id,
  type = "scatter", 
  mode = "lines"
) %>%
  layout(
    title = "CURRENT STATUS: Índice de delitos per cápita que afectan la calidad turística por región",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Score (out of 100)"),
    legend = list(title = list(text = "Región"))
  ) %>%
  add_lines(
    x = range(tr_qual_crime_score$ano), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (0 crimes = 100/100)",
    inherit = FALSE
  )

tr_qual_crime_plotly
```

### Assuming that when there is no data, that the score is 100...only for regions that have data at some point.

If there are no municipios in that OHI region then they should not be gapfilled like this.

```{r}
## fill in regions missing any values with 100 since the summarize may have dropped any regions with 0. fix this later??
tr_qual_crime_score_filled <- tr_qual_crime_score %>%
  select(ano, rgn_id, current_status) %>%
  complete(ano, rgn_id, fill = list(current_status = 100))

tr_qual_crime_plotly_filled <- plot_ly(
  tr_qual_crime_score_filled,
  x = ~ano,
  y = ~current_status,
  color = ~rgn_id,
  type = "scatter",
  mode = "lines"
) %>%
  layout(
    title = "CURRENT STATUS: Índice de delitos que afectan la calidad turística por región",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Score (out of 100)"),
    legend = list(title = list(text = "Región"))
  ) %>%
  add_lines(
    x = range(tr_qual_crime_score_filled$ano),
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (0 crimes = 100/100)",
    inherit = FALSE
  )

tr_qual_crime_plotly_filled
```

```{r}
# bring back geometry
inland_50 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer","inland_50km_goc.shp"))

tr_qual_crime_score_filled_sf <- inland_50 %>%
  left_join(tr_qual_crime_score_filled, by = "rgn_id")

# st_write(tr_qual_crime_score_filled_sf, here::here("TR/tr_quality_int","tr_qual_crime_score_filled_sf.shp"))

tr_qual_crime_score_filled_sf_2025 <- tr_qual_crime_score_filled_sf %>% 
  filter(ano %in% "2025")

mapview(tr_qual_crime_score_filled_sf_2025, zcol = "current_status")
```

### UPDATE: only including crime that directly impacts tourists, now at 50km not 10km!!

#### Read in SESNSP GoC data

```{r}
library(tidyverse)
library(here)
library(sf)
library(mapview)
```


```{r}
sesnsp_goc_50km <- st_read(here::here("/home/shares/ohi/OHI_GOC/goal_prep/ss/int","sesnsp_goc_municipios_50km.shp")) ## need to fix the names / make a document describing the abbreviations!!

unique(sesnsp_goc_50km$rgn_id) # for 50km: [1] "Region_1" "Region_3" "Region_4" "Region_5" "Region_7" "Region_8" "Region_9"

mapview(sesnsp_goc_50km)
```

#### Updated TR crimes

Conceptual rule: “included if it plausibly alters safety, cleanliness, aesthetics, or atmosphere in spaces tourists visit, which in turn affects visit quality."

```{r}
# vector of subtypes I want to keep for TR quality
updated_tr_qual_crime_subtypes <- c(
  "Daño a la propiedad",
  "Extorsión",
  "Fraude",
  "Otros delitos contra el patrimonio",
  "Otros robos",
  "Robo a casa habitación",
  "Robo a institución bancaria",
  "Robo a negocio",
  "Robo a transeúnte en espacio abierto al público",
  "Robo a transeúnte en vía pública",
  "Robo a transportista",
  "Robo de autopartes",
  "Robo de vehículo automotor",
  "Robo en transporte individual",
  "Robo en transporte público colectivo",
  "Robo en transporte público individual",
  "Violencia de género en todas sus modalidades distinta a la violencia familiar",
  "Abuso sexual",
  "Acoso sexual",
  "Hostigamiento sexual",
  "Otros delitos que atentan contra la libertad y la seguridad sexual",
  "Violación equiparada",
  "Violación simple",
  "Corrupción de menores",
  "Otros delitos contra la sociedad",
  "Trata de personas",
  "Feminicidio",
  "Homicidio doloso",
  "Lesiones dolosas",
  "Otros delitos que atentan contra la vida y la integridad corporal",
  "Otros delitos que atentan contra la libertad personal",
  "Rapto",
  "Secuestro",
  "Tráfico de menores",
  "Amenazas",
  "Contra el medio ambiente",
  "Narcomenudeo"
)

#### those that were dropped:
# c(
#   "Delitos cometidos por servidores públicos",
#   "Evasión de presos",
#   "Otros delitos del Fuero Común",
#   "Robo de maquinaria"
# )



# filter sesnsp_goc by those subtypes
sesnsp_goc_tr_qual <- sesnsp_goc_50km %>%
  filter(sbtp_d_ %in% updated_tr_qual_crime_subtypes)

unique(sesnsp_goc_tr_qual$rgn_id) # [1] "Region_1" "Region_3" "Region_4" "Region_5" "Region_7" "Region_8" "Region_9"
```

#### Summarize the number of crimes by year and region (assuming that all crimes are weighted equally)

```{r}
sesnsp_goc_tr_qual_sum <- sesnsp_goc_tr_qual %>% 
  st_drop_geometry() %>% 
  group_by(ano, rgn_id) %>% 
  summarise(number_crimes = sum(annl_tt))

unique(sesnsp_goc_tr_qual_sum$rgn_id) # [1] "Region_3" "Region_4" "Region_5" "Region_7" "Region_8" "Region_9" "Region_1"
```

#### Divide by coastal population in each region

##### NOTE: NEED TO UPDATE...

1. We will have to update this because we eventually want to use 2015, 2020, and 2025 census data for population, not just static 2020.
2. I think maybe we should use the population per municipality (instead of population per OHI Region).  Because for some regions there are multiple municipios in the region that have data collected for them, versus some only have one or two.  

```{r}
coastal_50km_pop <- read_csv(here::here("general/OHI_rgn_population_50km.csv")) 

coastal_50km_pop_sumrgn9 <- coastal_50km_pop %>% 
  group_by(rgn_id) %>% 
  summarize(rgn_population_50km = sum(population_50km_2020))

sesnsp_goc_tr_sum_percapita <- left_join(sesnsp_goc_tr_qual_sum, coastal_50km_pop_sumrgn9, by = "rgn_id")

sesnsp_goc_tr_crimes_percapita <- sesnsp_goc_tr_sum_percapita %>% 
  group_by(ano, rgn_id) %>% 
  mutate(tr_crimes_percapita = (number_crimes / rgn_population_50km)) %>% 
  ungroup()
```

Make a visualization for 2024:
```{r}
region_2_6_tibble <- tibble::tribble(
  ~ano, ~tr_crimes_percapita, ~rgn_id,
  2024, 0.0002, "Region_2",
  2024, 0.0002, "Region_6"
)

viz_crime_rates_2024 <- sesnsp_goc_tr_crimes_percapita %>% 
  filter(ano == "2024") %>% 
  bind_rows(region_2_6_tibble)

library(ggplot2)
library(dplyr)

rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

viz_crime_rates_2024_plot <- viz_crime_rates_2024 %>%
  separate(col = rgn_id, into = c("rgn", "rgn_id_num"), sep = "_", remove = FALSE) %>% 
  arrange(tr_crimes_percapita) %>% 
  mutate(
    crimes_per_1000 = tr_crimes_percapita * 1000,
    rgn_id_num = factor(rgn_id_num, levels = rgn_id_num)
    ) 

#### join with ohi rgn polygons
ohi <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp")) %>%
  filter(location == "marine") %>% 
  select(rgn_id)

viz_crime_rates_2024_ohigns <- full_join(viz_crime_rates_2024_plot, ohi, by = "rgn_id") %>% 
  st_as_sf()

mapview(viz_crime_rates_2024_ohigns, 
        zcol = "crimes_per_1000",
        col.regions = colorRampPalette(c("#FFD599", "#FFAC75", "#F82736", "darkred")),
        layer.name = "Crimes affecting tourists per 1,000 local inhabitants by OHI region (2024)",
        legend = TRUE)

####################
## bar plot

p <- ggplot(
  viz_crime_rates_2024_plot,
  aes(x = rgn_id_num, y = crimes_per_1000, fill = rgn_id)
) +
  geom_col(color = "black", linewidth = 0.4) +
  scale_fill_manual(values = rgn_colors) +
  labs(
    title = "Crimes affecting tourists per 1,000 local inhabitants by OHI region (2024)",
    x = "Region",
    y = "Crimes per 1,000 local inhabitants"
  ) +
  theme_minimal(base_family = "Calibri") +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold", size = 16),
    axis.text.x = element_text(face = "bold", size = 16),
    axis.text.y = element_text(face = "bold", size = 16),
    axis.title.x = element_blank(),
    legend.position = "none"
  )

p

ggsave(
  filename = here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_viz",
    "crime_rate_per_1000_by_region_2024.png"
  ),
  plot = p,
  width = 6.5,
  height = 5,
  units = "in",
  dpi = 300
)

```

Trend in recent years:

```{r}
last_3yrs <- c(2022, 2023, 2024)

trend_last3_slope <- sesnsp_goc_tr_crimes_percapita %>%
  st_drop_geometry() %>% 
  filter(ano %in% last_3yrs) %>%
  group_by(rgn_id) %>%
  summarise(
    slope_per_year = coef(lm(tr_crimes_percapita ~ ano))[2],
    .groups = "drop"
  )

trend_last3_pct <- sesnsp_goc_tr_crimes_percapita %>%
  st_drop_geometry() %>% 
  filter(ano %in% last_3yrs) %>%
  group_by(rgn_id) %>%
  summarise(
    crime_percap_2022 = tr_crimes_percapita[ano == 2022],
    crime_percap_2024 = tr_crimes_percapita[ano == 2024],
    pct_change_3yr = (crime_percap_2024 - crime_percap_2022) / crime_percap_2022 * 100,
    .groups = "drop"
  ) 

dom_trend_last3yrs <- full_join(trend_last3_pct, trend_last3_slope) %>% 
  filter(arrival_type == "dom") %>% 
  mutate(target_arrivals = arrivals_2024 * 1.098,
         proj_arrivals_2030 = arrivals_2024 + (6 * slope_per_year),
         on_track = ifelse(proj_arrivals_2030 >= target_arrivals, 1, 0))
```



#### Prelim Scores

```{r}
max_tr_crime_percapita_allrgns <- max(sesnsp_goc_tr_crimes_percapita$tr_crimes_percapita, na.rm = TRUE)

tr_qual_crime_score_alt <- sesnsp_goc_tr_crimes_percapita %>%
  mutate(
    current_status = (1 - (tr_crimes_percapita / max_tr_crime_percapita_allrgns)) * 100
  ) 

write_csv(tr_qual_crime_score_alt, here::here("TR/tr_quality_int","tr_qual_crime_score_120925.csv"))
```

#### Plotly

```{r}
library(plotly)

## TR quality crime index score over time
tr_qual_crime_plotly <- plotly::plot_ly(
  tr_qual_crime_score_alt,
  x = ~ano, 
  y = ~current_status, 
  color = ~rgn_id,
  type = "scatter", 
  mode = "lines+markers"
) %>%
  layout(
    title = "CURRENT STATUS: Índice de delitos per cápita que afectan la calidad turística por región",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Score (out of 100)"),
    legend = list(title = list(text = "Región"))
  ) %>%
  add_lines(
    x = range(tr_qual_crime_score_alt$ano), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (0 crimes = 100/100)",
    inherit = FALSE
  )

tr_qual_crime_plotly
```

### Assuming that when there is no data, that the score is 100...only for regions that have data at some point.

If there are no municipios in that OHI region then they should not be gapfilled like this.

```{r}
## fill in regions missing any values with 100 since the summarize may have dropped any regions with 0.

## for rgn 2 (which has no DOCUMENTED crimes from sesnsp, but definitely crime to some degree...... <https://bahiadelosangeles.org/safety.html>)
region_2_na <- tibble(
  ano = 2015:2025,
  rgn_id = "Region_2"
) %>%
  mutate(current_status = NA_real_)

tr_qual_crime_score_filled <- tr_qual_crime_score_alt %>% 
  select(ano, rgn_id, current_status) %>%
  complete(ano, rgn_id, fill = list(current_status = 100)) %>%   # only rgn 1 and rgn 2 needed to be filled.
  bind_rows(region_2_na)

tr_qual_crime_plotly_filled <- plot_ly(
  tr_qual_crime_score_filled,
  x = ~ano,
  y = ~current_status,
  color = ~rgn_id,
  type = "scatter",
  mode = "lines+markers"
) %>%
  layout(
    title = "CURRENT STATUS: Índice de delitos que afectan la calidad turística por región",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Score (out of 100)"),
    legend = list(title = list(text = "Región"))
  ) %>%
  add_lines(
    x = range(tr_qual_crime_score_filled$ano),
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (0 crimes = 100/100)",
    inherit = FALSE
  )

tr_qual_crime_plotly_filled
```

```{r}
# # bring back geometry
# inland_50 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer","inland_50km_goc.shp"))
# 
# tr_qual_crime_score_filled_sf <- inland_50 %>%
#   left_join(tr_qual_crime_score_filled, by = "rgn_id")
# 
# # st_write(tr_qual_crime_score_filled_sf, here::here("TR/tr_quality_int","tr_qual_crime_score_filled_sf.shp"))
# 
# tr_qual_crime_score_filled_sf_2025 <- tr_qual_crime_score_filled_sf %>% 
#   filter(ano %in% "2025")
# 
# mapview(tr_qual_crime_score_filled_sf_2025, zcol = "current_status")
```
