---
title: "Tourist data"
output: html_document
date: "2025-3-6"
editor_options: 
  chunk_output_type: console
---
Data: "Ocupación en hoteles y moteles por centro turístico, corte mensual DataTur"
Go to: <https://monitoreohotelero.sectur.gob.mx/Reportes/Menu.aspx>

To download:

> 1. Go to the “Generate Report” section.
> 2. Select the “Mensual”, or monthly, time frame.
> 3. Choose the option “Hotel and motel occupancy by tourist center, monthly DataTur snapshot.” / "Ocupación en hoteles y moteles por centro turístico, corte mensual. DataTur"
> 4. Click “Continue” and then “Advanced criteria” (Criterios Avanzados)
> 5. Select the municipality of interest, then all the variables of interest.  Specify "Serie" for "Cobertura Temporal corte mensual", starting from 2018 January (Enero) to 2025 Junio (June).
> 6. Finally, click on “Generate Report.”

The report can be downloaded in various formats and contains the information available as of the date of consultation.  It is important to note that this information is presented at the tourist center level, not the municipal level. However, it can be useful for territorial approximations in the Gulf of California region, particularly in key tourist destinations.

Tourist centers in GoC:  Costalegre, Puerto Vallarta, San Felipe, SAN QUINTIN, Cabo San Lucas, La Paz, Loreto, San José del Cabo, Zona Corredor los Cabos, Nuevo Nayarit, Tepic, 	Culiacán, El Fuerte, El Rosario, Escuinapa, Los Mochis, Mazatlán, San Ignacio (but no data, all NAs 2013-2026), Guaymas, Hermosillo

```{r}
library(mapview)
library(sf)
library(tidyverse)
library(here)
library(terra)
```


Okay!! So it looks like you CANNOT save the file as a .CSV first, instead you **must** to do it in this order:

1.  On the SECTUR report site, after generating the report, click "Exportar a Excel".  

2.  Open the ".xls" (which is really a .htm) in Excel, and then "Save As..." a UTF-8 .CSV. Rename it `DataturReportePublico_municipio_resaved.csv` where you replace municipio with the municipality you just saved. 

3.  Bring into mazu in `/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/resaved_as_csv` to be read in here.  This is the only way that you do not get any "†" in whitespaces.

**These instructions will be brought to the top of the script when it is cleaned up.**

# Clean data 

```{r}
clean_hotel_occupations_data <- function(municipio) {
  
  # where to find the file
  file_path <- here::here(
    "/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/SECTUR_resaved_032426",
    paste0("DataturReportePublico_", municipio, "_resaved.csv")
  )

  # read raw data
  # raw_data <- read_csv(file_path)
  raw_data <- read_csv(file_path, locale = readr::locale(encoding = "UTF-8")) # changed because having error: "Error in `map()`: ℹ In index: 1. Caused by error in `nchar()`: ! invalid multibyte string, element 1"
  
  # clean the column names, remove top rows, and tidy
  cleaned <- raw_data %>% 
    set_names(.[9, ]) %>%  
    slice(-c(1:10)) %>% # NOTE: not all are the same length, so some of them will have the bottom text (ex: ND No disponible) as part of the output, and there will have to be extra cleaning done at the end.
    janitor::clean_names()
  
  # clean and standardize the "concepto" column
  cleaned_concept <- cleaned %>%
    mutate(
      concept = str_to_lower(concepto),
      concept = str_trim(concept),
      concept = str_replace_all(concept, " ", "_")
    ) %>% 
    relocate(concept, .after = concepto) %>% # these two lines of code are not really necessary because the func will do it all at once
    dplyr::select(-c(concepto))
  
  # identify topic headers and combine them with subcategories
  by_topic <- cleaned_concept %>% 
    mutate(
      is_topic = !str_detect(concept, "^[1-5]_estrella[s]?|total_categorias") & !is.na(concept)
    ) %>%
    mutate(header_concept = if_else(is_topic, concept, NA_character_)) %>% # use NA_character_ for this stuff just in case
    fill(header_concept) %>% 
    mutate(
      concept_full = if_else(concept != header_concept, paste0(header_concept, ":", concept), concept)
    )
  
  # pivot longer
  long <- by_topic %>% 
    pivot_longer(
      cols = 2:173, # adjust if more columns (if you change the number of years by downloading them later on. make sure to stop before `is_topic` column)
      names_to = "month_year",
      values_to = "value"
    )
  
  # separate year and month, filter, and tag city
  final <- long %>% 
    mutate(
      is_mo_yr = !str_detect(month_year, "acum|total"),
      temp = if_else(is_mo_yr, month_year, NA_character_)
    ) %>% 
    separate(temp, into = c("month", "year"), sep = "_", remove = FALSE) %>% 
    filter(is_topic == FALSE) %>% 
    dplyr::select(header_concept, concept, concept_full, month_year, month, year, value) %>% 
    mutate(municipio = str_to_lower(municipio)) %>% 
    mutate(value = as.numeric(str_replace_all(value, ",", "")))
  
  return(final)
}
```


## Apply function to all municipios

```{r}
all_tourist_centers <- c(
  "Cabo_San_Lucas", "Culiacan", "El_Fuerte", "El_Rosario","Ensenada", "Escuinapa", "Guaymas", "Hermosillo", "La_Paz", "Loreto", "Los_Mochis", "Mazatlan","Mexicali","Nuevo_Nayarit", "Puerto_Vallarta", "San_Felipe", "San_Jose_del_Cabo", "San_Quintin","Tepic", "Zona_Corredor_Los_Cabos"
)

library(purrr)

all_data <- all_tourist_centers %>%
  imap_dfr(function(mun, nm) {
    message("Reading: ", nm)
    clean_hotel_occupations_data(mun)
  })

unique(all_data$municipio) # all 20 there!


all_data_clean <- all_data %>% 
  rename( 
    metric = header_concept,
    category = concept
  ) %>%  # for better clarity, since we are done cleaning and these names make more sense now
  mutate(
    concept_full = str_replace_all(concept_full, "\U{0eef2f}th>|\U{0ee832}|\U{0ee82e}", "_"),
    # metric_label = str_replace_all(metric, "_", " "),
    # metric_label = str_to_title(metric_label), # for later viz 
    metric = str_replace_all(metric, "\U{0eef2f}th>|\U{0ee832}|\U{0ee82e}", "_"),
    category = str_replace_all(category, "\U{0eef2f}th>|\U{0ee832}|\U{0ee82e}", "_"),
    value = str_replace_all(value, ",",""),
    year = str_replace_all(year, "p",""), # only in 2025, assume it means "preliminary"
    month = case_when(
      tolower(month) == "enero" ~ "01",
      tolower(month) == "febrero" ~ "02",
      tolower(month) == "marzo" ~ "03",
      tolower(month) == "abril" ~ "04",
      tolower(month) == "mayo" ~ "05",
      tolower(month) == "junio" ~ "06",
      tolower(month) == "julio" ~ "07",
      tolower(month) == "agosto" ~ "08",
      tolower(month) == "septiembre" ~ "09",
      tolower(month) == "octubre" ~ "10",
      tolower(month) == "noviembre" ~ "11",
      tolower(month) == "diciembre" ~ "12",
      TRUE ~ NA_character_  # otherwise
    ) # for lubridate later
  ) %>% 
  filter(!category %in% NA) 

unique(all_data_clean$concept_full) # check that there are no weirdnesses left

# data_check <- all_data_clean %>% filter(year %in% 2024) %>% filter(metric %in% "cuartos_ocupados_no_residentes") # double check with raw.  looks good!
```


```{r}
# so that we can have the yearly total, not just by month, because we will have the scores per year.
all_data_yearly <- all_data_clean %>% 
  filter(str_detect(month_year, "total_anual_")) %>% 
  dplyr::select(-c(month, year)) %>% 
  separate_wider_delim(., month_year, delim = "_anual_", names = c("type","year"))

## save into int
# write_csv(all_data_yearly, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/hotel_monitoring_int","hotel_data_cleaned_yearly.csv"))
write_csv(all_data_yearly, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/hotel_monitoring_int","hotel_data_cleaned_yearly_2013_2026.csv"))
```


# NEW Domestic vs. International

We are updating the reference point to be the exact number of visitors per year that are 30% more than the international visitors in 2024 and 9.8% more than domestic visitors in 2024.  That way, the score is a reflection of how close the number of arrivals are to the target (it likely would not be a linear increase until 2030).

Therefore, the target/baseline would be: 

**International**
The # int arrivals for that year / 
The # int arrivals in 2024 * 1.30

**Domestic**
The # domestic arrivals for that year / 
The # domestic arrivals in 2024 * 1.098

### Domestic

```{r}
all_data_yearly <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/hotel_monitoring_int","hotel_data_cleaned_yearly_2013_2026.csv"))

## "llegadas_de_turistas_residentes" = domestic visitors
dom_arrivals <- all_data_yearly %>% 
  filter(metric %in% "llegadas_de_turistas_residentes") %>% # domestic arrivals
  filter(category %in% "total_categorias") # since we do not care what star designation it is
```

```{r}
# let's make them an sf object
dom_geocode <- dom_arrivals %>% 
  mutate(country = "Mexico") %>% 
  mutate(estado = case_when(
    municipio %in% c("cabo_san_lucas", "la_paz", "loreto", "san_jose_del_cabo", "zona_corredor_los_cabos") ~ "Baja California Sur",
    municipio %in% c("san_felipe", "ensenada","mexicali","san_quintin") ~ "Baja California",
    municipio %in% c("guaymas", "hermosillo") ~ "Sonora",
    municipio %in% c("culiacan", "el_fuerte", "los_mochis", "mazatlan", "escuinapa", "el_rosario") ~ "Sinaloa",
    municipio %in% c("nuevo_nayarit","tepic") ~ "Nayarit",
    municipio %in% c("puerto_vallarta") ~ "Jalisco",
    TRUE ~ ""  # default case for any unmatched cities
  )) %>% 
  tidygeocoder::geocode(city = municipio, country = country, state = estado) # to give each city a lat, long
  
dom_geocode_sf <- dom_geocode %>% 
  mutate(lat = ifelse(municipio %in% "zona_corredor_los_cabos", 23.1539, lat)) %>% # adding in the coordinates manually from <https://latitude.to/articles-by-country/mx/mexico/56998/los-cabos-corridor>
  mutate(long = ifelse(municipio %in% "zona_corredor_los_cabos", -109.720233, long)) %>% 
  st_as_sf(coords = c("long", "lat"),
           crs = 4326, # WGS84
           remove = FALSE)

mapview(dom_geocode_sf)

############################################################
# Now, let us assign each tourist center to an OHI region. 
############################################################

goc_poly <- st_read(here("spatial/ohi_regions","goc_ohi_rgns.shp"))
st_crs(goc_poly) # not WGS84: Mexico ITRF2008 / LCC
mapview(goc_poly)

dom_geocode_sf_transform <- st_transform(dom_geocode_sf, crs = st_crs(goc_poly)) %>%  # transform the tourist centers to Mexico ITRF2008 / LCC
   mutate(value = as.numeric(value))
crs(dom_geocode_sf_transform)

dom_goc <- st_intersection(dom_geocode_sf_transform, goc_poly) %>% 
  mutate(arrivals_type = "domestic")
mapview(dom_goc) # now each tourist center has associated region data

##############################
#A cool plot to see them each:
##############################

# map of all goc tourist centers
mapview(dom_goc, zcol = "rgn_id") + mapview(inland_50, zcol = "rgn_id")
```


Like stated earlier, the reference point has been updated to use specific target values rather than annual percentage increases (<https://tourismanalytics.com/blog-posts/how-mexicos-forty-percent-tourism-growth-plan-aims-for-fifth-global-spot-by-2030-with-twenty-billion-dollar-investment-in-infrastructure-and-market-expansion>).

For **domestic tourism**, the 2030 target is defined as **9.8% higher than 2024 domestic visitor arrivals**.

Therefore, the current status for each year is calculated as follows:

**Domestic:**
$$
\text{Score} = \frac{\text{Domestic arrivals (year)}}{\text{Domestic arrivals in 2024} \times 1.098} \times 100
$$

A **score of 100** indicates that the number of arrivals has **reached the 2030 target level** (i.e., 9.8% or 30% higher than 2024 levels).

* A score < 100 means arrivals are **below the target trajectory**.
* A score > 100 means arrivals have **exceeded the 2030 target**.

Those two current status scores are then averaged to produce an overall score of how close each region is to making the 2030 target overall. 

```{r}
target_baseline_increase <- 1.098  # 9.8% more visitors than the amount in 2024
baseline_year <- 2024

dom_quantity_score <- dom_goc %>%
  group_by(municipio) %>% #### I made a mistake in the old version: we want to calculate current status scores per municipio first, then average those for the entire region. FOR NOW, let's just have the scores by municipio.
  mutate(
    year = as.numeric(year),
    
    # determine the baseline value for each municipio (2024)
    dom_baseline_arrivals = value[year == baseline_year][1],

    # to know how far from 2024 we are (note: negative numbers are because of years before 2024)
    years_since_baseline = year - baseline_year,

    # target arrivals for domestic visitors
    dom_target_arrivals = dom_baseline_arrivals * target_baseline_increase,   # for domestic tourism, the 2030 target is defined as 9.8% higher than 2024 domestic visitor arrivals.

    # make it a score
    dom_index_score = (value / dom_target_arrivals) * 100,
    dom_gap_to_target = dom_target_arrivals - value,
    dom_gap_percentage = ((dom_target_arrivals - value) / value) * 100 ## target arrivals - current arrivals / current arrivals... ideal is 0 or negative to reflect how much needs to be done (if you reach the target, it is negative, because nothing needs to be done!)
  ) %>%
  ungroup() %>%
  dplyr::select(municipio, rgn_id, year, dom_current_arrivals = value,
         dom_baseline_arrivals, dom_target_arrivals, dom_index_score,
         dom_gap_to_target, dom_gap_percentage, everything()) %>% 
  filter(!is.na(dom_baseline_arrivals)) %>%  # remove any municipalities that do not have values for 2024 arrivals.
  filter(year %in% 2013:2025)


##### double-checking that it all looks good #####
# for san felipe:
# > 139861 * 1.098
# [1] 153567.4
# > 153567.4-139861
# [1] 13706.4
# > 13706.4/139861
# [1] 0.09800016
# > 66410/153567.4
# [1] 0.4324486

# and cabo san lucas: 
# cabo_san_lucas. Region_4 2019 274384 243965 267873.6
# it makes sense that the dom_gap_percentage is negative, because in 2019 the number of domestic arrivals is actually more than the 2024 * 1.098.

```


```{r}
library(plotly)

## plot the score over time
updated_dom_score_plotly <- plotly::plot_ly(
  dom_quantity_score, 
  x = ~year, 
  y = ~dom_index_score, 
  color = ~municipio, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "CURRENT STATUS: Llegadas de turistas nacionales por municipio",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(dom_quantity_score$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
updated_dom_score_plotly

htmlwidgets::saveWidget(updated_dom_score_plotly, here::here("TR/sectur_quantity_exploration_int","dom_arrivals_score_101325.html"), selfcontained = TRUE)
```

YAY, this makes sense!! So for 2024 rows:

dom_baseline_arrivals = value *(for 2024)*

Therefore, using the equation to get the current status:

$$
\text{dom_index_score}_{2024} = \frac{\text{value}}{\text{value} \times 1.098} \times 100 = \frac{100}{1.098} = 91.07468
$$


```{r}
## compare to the patterns of domestic arrivals over time
dom_arrivals_plot <- plotly::plot_ly(
  dom_goc, 
  x = ~year, 
  y = ~value, 
  color = ~municipio, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "Llegadas de turistas nacionales por municipio (domestic arrivals)",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Llegadas // Arrivals")
  ) %>% 
  add_lines(
    x = range(dom_goc$year),
    y = c(100, 100),
    line = list(color = "white", width = 2),
    name = "",
    inherit = FALSE
  )
dom_arrivals_plot

htmlwidgets::saveWidget(dom_arrivals_plot, here::here("TR/sectur_quantity_exploration_int","dom_arrivals_plot_100825.html"), selfcontained = TRUE)
```

### International

```{r}
## "llegadas_de_turistas_no_residentes" = international visitors
int_arrivals <- all_data_yearly %>% 
  filter(metric %in% "llegadas_de_turistas_no_residentes")%>% # int arrivals
  filter(category %in% "total_categorias") # since we do not care what star designation it is
```

```{r}
# let's make them an sf object
int_geocode <- int_arrivals %>% 
  mutate(country = "Mexico") %>% 
  mutate(estado = case_when(
    municipio %in% c("cabo_san_lucas", "la_paz", "loreto", "san_jose_del_cabo", "zona_corredor_los_cabos") ~ "Baja California Sur",
    municipio %in% c("san_felipe") ~ "Baja California",
    municipio %in% c("guaymas", "hermosillo") ~ "Sonora",
    municipio %in% c("culiacan", "el_fuerte", "los_mochis", "mazatlan", "escuinapa", "el_rosario") ~ "Sinaloa",
    municipio %in% c("puerto_vallarta") ~ "Jalisco",
    TRUE ~ ""  # default case for any unmatched cities
  )) %>% 
  tidygeocoder::geocode(city = municipio, country = country, state = estado) # to give each city a lat, long
  
int_geocode_sf <- int_geocode %>% 
  mutate(lat = ifelse(municipio %in% "zona_corredor_los_cabos", 23.1539, lat)) %>% # adding in the coordinates manually from <https://latitude.to/articles-by-country/mx/mexico/56998/los-cabos-corridor>
  mutate(long = ifelse(municipio %in% "zona_corredor_los_cabos", -109.720233, long)) %>% 
  st_as_sf(coords = c("long", "lat"),
           crs = 4326, # WGS84
           remove = FALSE)

mapview(int_geocode_sf)

############################################################
# Now, let us assign each tourist center to an OHI region. 
############################################################

goc_poly <- st_read(here("spatial/ohi_regions","goc_ohi_rgns.shp"))
st_crs(goc_poly) # not WGS84: Mexico ITRF2008 / LCC
mapview(goc_poly)

int_geocode_sf_transform <- st_transform(int_geocode_sf, crs = st_crs(goc_poly)) %>%  # transform the tourist centers to Mexico ITRF2008 / LCC
   mutate(value = as.numeric(value))
st_crs(int_geocode_sf_transform)

int_goc <- st_intersection(int_geocode_sf_transform, goc_poly) %>% 
  mutate(arrivals_type = "international")
mapview(int_goc) # now each tourist center has associated region data

##############################
#A cool plot to see them each:
##############################

# map of all goc tourist centers
mapview(int_goc, zcol = "rgn_id") + mapview(goc_poly, zcol = "rgn_id")
```

Like stated earlier, the reference point has been updated to use specific target values rather than annual percentage increases (<https://tourismanalytics.com/blog-posts/how-mexicos-forty-percent-tourism-growth-plan-aims-for-fifth-global-spot-by-2030-with-twenty-billion-dollar-investment-in-infrastructure-and-market-expansion>).

For **international tourism**, the 2030 target is defined as **30% higher than 2024 international visitor arrivals**.

Therefore, the current status for each year is calculated as follows:

**International:**
$$
\text{Score} = \frac{\text{International arrivals (year)}}{\text{International arrivals in 2024} \times 1.30} \times 100
$$

A **score of 100** indicates that the number of arrivals has **reached the 2030 target level** (i.e., 9.8% or 30% higher than 2024 levels).

* A score < 100 means arrivals are **below the target trajectory**.
* A score > 100 means arrivals have **exceeded the 2030 target**.

Those two current status scores are then averaged to produce an overall score of how close each region is to making the 2030 target overall. 


```{r}
int_target_baseline_increase <- 1.30  # 30% more visitors than the amount in 2024
baseline_year <- 2024

int_quantity_score <- int_goc %>%
  group_by(municipio) %>% #### I made a mistake in the old version: we want to calculate current status scores per municipio first, then average those for the entire region. FOR NOW, let's just have the scores by municipio.
  mutate(
    year = as.numeric(year),
    
    # determine the baseline value for each municipio (2024)
    int_baseline_arrivals = value[year == baseline_year][1],

    # to know how far from 2024 we are (note: negative numbers are because of years before 2024)
    years_since_baseline = year - baseline_year,

    # target arrivals for international visitors
    int_target_arrivals = int_baseline_arrivals * int_target_baseline_increase,   # for international tourism, the 2030 target is defined as 30% higher than 2024 international visitor arrivals.

    # make it a score
    int_index_score = (value / int_target_arrivals) * 100,
    int_gap_to_target = int_target_arrivals - value,
    int_gap_percentage = ((int_target_arrivals - value) / value) * 100 ## target arrivals - current arrivals / current arrivals... ideal is 0 or negative to reflect how much needs to be done (if you reach the target, it is negative, because nothing needs to be done!)
  ) %>%
  ungroup() %>%
  dplyr::select(municipio, rgn_id, year, int_current_arrivals = value,
         int_baseline_arrivals, int_target_arrivals, int_index_score,
         int_gap_to_target, int_gap_percentage, everything()) %>% 
  filter(!is.na(int_baseline_arrivals)) %>%  # remove any municipalities that do not have values for 2024 arrivals.
  filter(year %in% 2013:2025)
```


```{r}
library(plotly)

## plot the score over time
updated_int_score_plotly <- plotly::plot_ly(
  int_quantity_score, 
  x = ~year, 
  y = ~int_index_score, 
  color = ~municipio, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "CURRENT STATUS: Llegadas de turistas internacionales por municipio",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(int_quantity_score$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
updated_int_score_plotly

htmlwidgets::saveWidget(updated_int_score_plotly, here::here("TR/sectur_quantity_exploration_int","int_arrivals_score_101325.html"), selfcontained = TRUE)
```


```{r}
## compare to the patterns of international arrivals over time
int_arrivals_plot <- plotly::plot_ly(
  int_goc, 
  x = ~year, 
  y = ~value, 
  color = ~municipio, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "Llegadas de turistas internacionales por municipio (international arrivals)",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Llegadas // Arrivals")
  ) %>% 
  add_lines(
    x = range(int_goc$year),
    y = c(100, 100),
    line = list(color = "white", width = 2),
    name = "",
    inherit = FALSE
  )
int_arrivals_plot

htmlwidgets::saveWidget(int_arrivals_plot, here::here("TR/sectur_quantity_exploration_int","int_arrivals_plot_100825.html"), selfcontained = TRUE)
```

# COME BACK: UPDATED UNTIL 2025 UNTIL HERE

# Gapfilled: Using DENUE employee data

In the script `~TR/gapfilling_arrivals_by_employees.qmd`, I predicted missing tourist arrival data using a model that used the number of employees in hotels/accommodations to predict the total number of arrivals.  To apply it here, we will need to determine the proportion of arrivals that are international vs domestic out of the total in each state.  That way, we can apply it to the regions that have missing data.

Essentially, we will determine for each state the proportion of domestic/total and international/total and just assume that same proportion for the regions we are estimating using the regression models.

For now, we have only been using 2019 - 2024.  This can change as we update the DENUE cleaning function, but for now we will filter to those years.

```{r}
## first, we will read in the cleaned sectur data, and only select arrivals.  We want to find the total number of arrivals, int + dom, for every year and municipio.
all_sectur_arrivals <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/hotel_monitoring_int","hotel_data_cleaned_yearly.csv")) %>% 
  filter(metric %in% c("llegadas_de_turistas_residentes", "llegadas_de_turistas_no_residentes")) %>% # domestic and international arrivals
  filter(category %in% "total_categorias")  %>% # since we do not care what star designation it is
  filter(year %in% 2019:2024) %>% 
  mutate(arrival_type = case_when(
    metric == "llegadas_de_turistas_residentes" ~ "dom",
    metric == "llegadas_de_turistas_no_residentes" ~ "int",
    TRUE ~ NA
  ))

# summarize to find the total number of arrivals for every year and municipio
all_sectur_arrivals_sum <- all_sectur_arrivals %>%
  group_by(municipio, year) %>% 
  summarize(total_arrivals = sum(value))

# obtain the proportion of the total for each value.  Double check that for the same municipio, the proportion of the dom + the proportion of the int = 1. Then, assign them each to states.
all_sectur_arrivals_w_prop <- left_join(all_sectur_arrivals, all_sectur_arrivals_sum, by = c("municipio","year")) %>%
  mutate(prop_of_total = value/total_arrivals) %>% 
  mutate(country = "Mexico") %>%  
  mutate(estado = case_when(
    municipio %in% c("cabo_san_lucas", "la_paz", "loreto", "san_jose_del_cabo", "zona_corredor_los_cabos") ~ "baja_california_sur",
    municipio %in% c("san_felipe") ~ "baja_california",
    municipio %in% c("guaymas", "hermosillo") ~ "sonora",
    municipio %in% c("culiacan", "el_fuerte", "los_mochis", "mazatlan", "escuinapa", "el_rosario") ~ "sinaloa",
    municipio %in% c("puerto_vallarta") ~ "jalisco",
    TRUE ~ ""  # default case for any unmatched cities
  ))  
  
```

For now, we will average the proportion across the years 2019:2024 for every state. This is because in some states, notably Sonora, we have very little information on the numbers of arrivals throughout all years.  So we cannot obtain a number for the proportion of dom vs. int for every year for each state. 

Note that Nayarit has no data at all from SECTUR, so I will assume it has the same proportions as Jalisco.  

I will need to check in with Mel on all of this!! -- checked, this is the best we can do (120925).

```{r}
# summarize the average proportion of int and dom for every state.
all_sectur_prop_by_state <- all_sectur_arrivals_w_prop %>% 
  group_by(arrival_type, estado) %>% 
  summarize(avg_prop = mean(prop_of_total, na.rm = TRUE))

# pivot wider for easier use when using these proportions to estimate the number of arrivals international vs domestic.
# all_sectur_prop_by_state_wide <- all_sectur_prop_by_state %>% 
#   pivot_wider(
#     names_from = arrival_type,
#     values_from = avg_prop
#   ) %>% 
#   rename(avg_prop_dom = dom,
#          avg_prop_int = int) 

## save it!!
write_csv(all_sectur_prop_by_state, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/gapfilled_int","sectur_avg_prop_by_state_2019_2024.csv"))
```

## Bring back arrivals that have been gapfilled

```{r}
arrivals_filled_by_DENUE_accommodation_employees <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/gapfilled_int","arrivals_filled_by_DENUE_accommodation_employees_2019_2024.csv"))

gapfilled_total_arrivals <- arrivals_filled_by_DENUE_accommodation_employees %>%
  mutate(country = "Mexico") %>% 
  select(year = data_year, country, estado, municipio_clean, rgn_id, arrivals_merged, is_filled) 

```

```{r}
## prep all sectur arrivals for a join with the gapfilled totals
all_sectur_arrivals_prep_join <- all_sectur_arrivals %>% 
  mutate(
    municipio_clean = case_when(
      municipio %in% c("cabo_san_lucas", "san_jose_del_cabo", "zona_corredor_los_cabos") ~ "los_cabos",
      municipio == "los_mochis" ~ "ahome", # los mochis is a city within the municipality of ahome!!
      municipio == "el_rosario" ~ "rosario",
      municipio == "culiacan" ~ "culiacán",
      municipio == "mazatlan" ~ "mazatlán",
      TRUE ~ municipio
    )
  ) %>% 
  # reaggregate with changes
  group_by(year, municipio_clean, arrival_type) %>%
  summarize(value = sum(value), .groups = "drop") %>% 
  select(year, municipio_clean, arrival_type, raw_arrivals = value) # leave in "municipio" to double check at first

## check!!
setdiff(unique(all_sectur_arrivals_prep_join$municipio_clean), unique(gapfilled_total_arrivals$municipio_clean)) # character(0)
setdiff(unique(gapfilled_total_arrivals$municipio_clean), unique(all_sectur_arrivals_prep_join$municipio_clean)) 

#  [1] "acaponeta"               "angostura"               "bahía_de_banderas"      
#  [4] "benito_juárez"           "cabo_corrientes"         "caborca"                
#  [7] "cajeme"                  "comondú"                 "compostela"             
# [10] "concordia"               "elota"                   "empalme"                
# [13] "ensenada"                "etchojoa"                "guasave"                
# [16] "huajicori"               "huatabampo"              "mascota"                
# [19] "mocorito"                "mulegé"                  "navojoa"                
# [22] "navolato"                "pitiquito"               "puerto_peñasco"         
# [25] "rosamorada"              "ruíz"                    "salvador_alvarado"      
# [28] "san_blas"                "san_ignacio"             "san_luis_río_colorado"  
# [31] "san_pedro_lagunillas"    "san_quintín"             "san_sebastián_del_oeste"
# [34] "santiago_ixcuintla"      "talpa_de_allende"        "tecuala"                
# [37] "tepic"                   "tomatlán"                "tuxpan"                 
# [40] "xalisco"  

## makes sense, those were all the municipalities that were gapfilled using DENUE accommodation employees!!
```


```{r}
## municipalities missing in SECTUR arrivals
gapfilled_municipios <- setdiff(unique(gapfilled_total_arrivals$municipio_clean), unique(all_sectur_arrivals_prep_join$municipio_clean))

## expand into dataframe with years 2019–2024 and arrival types int/dom
gapfilled_municipios_df <- expand.grid(
  municipio_clean = gapfilled_municipios,
  year = 2019:2024,
  arrival_type = c("int", "dom"),
  KEEP.OUT.ATTRS = FALSE, # a logical indicating the "out.attrs" attribute should be computed and returned.
  stringsAsFactors = FALSE # logical specifying if character vectors are converted to factors.
)

## join with the prepped sectur arrivals df
gapfilled_municipios_and_sectur_arrivals_prep_join <- full_join(all_sectur_arrivals_prep_join, gapfilled_municipios_df, by = c("municipio_clean","year","arrival_type"))

```

Join the gapfilled totals with the raw int and dom values for arrivals:

```{r}
joined_gapfilled_and_raw_arrivals <- full_join(gapfilled_total_arrivals, gapfilled_municipios_and_sectur_arrivals_prep_join, by = c("year","municipio_clean"), relationship = "many-to-many")

## check that the raw arrivals for int + dom add up to the "arrivals_merged" when "is_filled" = 0 (it was not predicted using DENUE employees for accommodations)
## looks good!
```

Now, join this data frame with the avg proportions by state.

```{r}
all_sectur_prop_by_state <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/gapfilled_int","sectur_avg_prop_by_state_2019_2024.csv"))

joined_gapfilled_arrivals_and_prop <- left_join(joined_gapfilled_and_raw_arrivals, all_sectur_prop_by_state, by = c("estado", "arrival_type")) %>% 
  mutate(avg_prop = case_when(
    estado == "nayarit" & arrival_type == "dom" ~ 0.64928457,
    estado == "nayarit" & arrival_type == "int" ~ 0.35071543,
    TRUE ~ avg_prop
  )) # assume for now that nayarit has the save avg proportion as jalisco, though this is definitely not true.
```

Now we can actually estimate the number of international arrivals and domestic arrivals for all predicted total arrivals from DENUE data.

```{r}
estimated_int_dom_arrivals <- joined_gapfilled_arrivals_and_prop %>% 
  mutate(estimated_arrivals = ifelse(is_filled == "1", avg_prop*arrivals_merged, raw_arrivals)) # if it was gapfilled, then estimate the number of int or dom arrivals by multiplying the average proportion of int or dom arrivals by state and the predicted total arrivals from the DENUE data. If they were not gapfilled, use the origianl int and dom arrival values.
```


### Plotly of arrivals

```{r}
library(plotly)

### INT Arrivals over time ###
estimated_int_arrivals <- estimated_int_dom_arrivals %>% 
  filter(arrival_type %in% "int") %>% 
  arrange(year)


int_arrivals_plotly <- plotly::plot_ly(
  estimated_int_arrivals, 
  x = ~year, 
  y = ~estimated_arrivals, 
  color = ~municipio_clean, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "International arrivals by year and municipio",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Number of arrivals")
  )
int_arrivals_plotly
```


```{r}
### DOM Arrivals ###
estimated_dom_arrivals <- estimated_int_dom_arrivals %>% 
  filter(arrival_type %in% "dom") %>% 
  arrange(year)

dom_arrivals_plotly <- plotly::plot_ly(
  estimated_dom_arrivals, 
  x = ~year, 
  y = ~estimated_arrivals, 
  color = ~municipio_clean, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "Domestic arrivals by year and municipio",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Number of arrivals")
  )
dom_arrivals_plotly

```


## CS: by municipio

```{r}
# baseline_year <- 2024
# 
# quantity_current_status <- estimated_int_dom_arrivals %>% 
#   mutate(year = as.numeric(year)) %>%
#   
#   # group by municipio + arrival type
#   group_by(municipio_clean, arrival_type) %>%
#   
#   mutate(
#     # baseline per municipio + type
#     baseline_arrivals = estimated_arrivals[year == baseline_year][1],
# 
#     # target multiplier varies by type
#     target_multiplier = case_when(
#       arrival_type == "dom" ~ 1.098, # domestic +9.8%
#       arrival_type == "int" ~ 1.30,  # international +30%
#       TRUE ~ NA_real_
#     ),
# 
#     # calculate targets
#     target_arrivals = baseline_arrivals * target_multiplier,
# 
#     # CURRENT STATUS (0 to 100)
#     current_status = (estimated_arrivals / target_arrivals) * 100,
# 
#     # absolute and % gap
#     gap_to_target = target_arrivals - estimated_arrivals,
#     gap_percentage = ((target_arrivals - estimated_arrivals) / estimated_arrivals) * 100
#   ) 
# 
# ## save this to outputs
# write_csv(quantity_current_status, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quantity","quantity_current_status_120125.csv"))
```

### Municip. plotlys

International:

```{r}
# library(plotly)
# 
# int_quantity_current_status <- quantity_current_status %>% 
#   filter(arrival_type %in% "int") %>% 
#   arrange(year)
# 
# ## plot the score over time
# final_int_score_plotly <- plotly::plot_ly(
#   int_quantity_current_status, 
#   x = ~year, 
#   y = ~current_status, 
#   color = ~municipio_clean, 
#   type = "scatter", 
#   mode = "lines"
# ) %>% 
#   layout(
#     title = "CURRENT STATUS: Llegadas de turistas internacionales por municipio",
#     xaxis = list(title = "Año // Year"),
#     yaxis = list(title = "Puntaje // Percent (out of 100)")
#   ) %>%
#   add_lines(
#     x = range(int_quantity_current_status$year), 
#     y = c(100, 100),
#     line = list(color = "black", width = 2, dash = "dash"),
#     name = "Meta // Goal 2030 (100/100)",
#     inherit = FALSE
#   )
# final_int_score_plotly
# 
# htmlwidgets::saveWidget(final_int_score_plotly, here::here("TR/sectur_quantity_exploration_int","int_current_status_gapfilled_by_municipio_120125.html"), selfcontained = TRUE)
```

Domestic:

```{r}
# library(plotly)
# 
# dom_quantity_current_status <- quantity_current_status %>% 
#   filter(arrival_type %in% "dom") %>% 
#   arrange(year)
# 
# ## plot the score over time
# final_dom_score_plotly <- plotly::plot_ly(
#   dom_quantity_current_status, 
#   x = ~year, 
#   y = ~current_status, 
#   color = ~municipio_clean, 
#   type = "scatter", 
#   mode = "lines"
# ) %>% 
#   layout(
#     title = "CURRENT STATUS: Llegadas de turistas nacionales por municipio",
#     xaxis = list(title = "Año // Year"),
#     yaxis = list(title = "Puntaje // Percent (out of 100)")
#   ) %>%
#   add_lines(
#     x = range(dom_quantity_current_status$year), 
#     y = c(100, 100),
#     line = list(color = "black", width = 2, dash = "dash"),
#     name = "Meta // Goal 2030 (100/100)",
#     inherit = FALSE
#   )
# final_dom_score_plotly
# 
# htmlwidgets::saveWidget(final_dom_score_plotly, here::here("TR/sectur_quantity_exploration_int","dom_current_status_gapfilled_by_municipio_120125.html"), selfcontained = TRUE)
```

## CS: by region

Somewhere along the way we lost the lat/longs for each municipio.  Lets add them back in and join with ohi regions, so that we can attribute each municipality to an OHI Region.

```{r}
estimated_int_dom_arrivals_latlong <- estimated_int_dom_arrivals %>% 
  tidygeocoder::geocode(city = municipio_clean, country = country, state = estado) %>%  # to give each city a lat, long
  mutate(coordinate_na = is.na(lat))

na_municipios_latlong <- estimated_int_dom_arrivals_latlong %>% 
  filter(coordinate_na == "TRUE") %>%
  distinct(municipio_clean)
  
  # manually found:
  ## tomatlán = 19.938547, -105.249363
  ## talpa_de_allende = 20.382398, -104.820536
  ## san_sebastián_del_oeste = 20.762123, -104.854319
  ## san_quintín = 23.709045, -110.396823
  ## san_pedro_lagunillas = 21.218043, -104.753181
  ## mascota = 20.527390, -104.792100
  ## ensenada = 28.443258, -112.880537 (if you look at employees_by_muni_yr_join_50, it is interchanged with san quintín for some of the hotels, because ensenada as a municipality stretches far along baja california on the interior gulf.)
  ## cabo_corrientes = 20.392035, -105.453138
  ## bahía_de_banderas = 20.759572, -105.292800

estimated_int_dom_arrivals_latlong_clean <- estimated_int_dom_arrivals_latlong %>%   
  mutate(
    lat = case_when(
      municipio_clean == "talpa_de_allende" ~ 20.382398,
      municipio_clean == "tomatlán" ~ 19.938547,
      municipio_clean == "san_sebastián_del_oeste" ~ 20.762123,
      municipio_clean == "san_quintín" ~ 23.709045,
      municipio_clean == "san_pedro_lagunillas" ~ 21.218043,
      municipio_clean == "mascota" ~ 20.527390,
      municipio_clean == "ensenada" ~ 28.443258,
      municipio_clean == "cabo_corrientes" ~ 20.392035,
      municipio_clean == "bahía_de_banderas" ~ 20.759572,
      TRUE ~ lat 
    ),
    long = case_when(
      municipio_clean == "talpa_de_allende" ~ -104.820536,
      municipio_clean == "tomatlán" ~ -105.249363,
      municipio_clean == "san_sebastián_del_oeste" ~ -104.854319,
      municipio_clean == "san_quintín" ~ -110.396823,
      municipio_clean == "san_pedro_lagunillas" ~ -104.753181,
      municipio_clean == "mascota" ~ -104.792100,
      municipio_clean == "ensenada" ~ -112.880537,
      municipio_clean == "cabo_corrientes" ~ -105.453138,
      municipio_clean == "bahía_de_banderas" ~ -105.292800,
      TRUE ~ long
    )
  )


estimated_int_dom_arrivals_sf <- estimated_int_dom_arrivals_latlong_clean %>% 
  st_as_sf(coords = c("long", "lat"),
           crs = 4326, # WGS84
           remove = FALSE) %>% 
  select(-rgn_id) # because many were missing and it is easier to just intersect with ohi rgns.
```

Now, let's assign regions to each one.

```{r}
ohi_region_assignments <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/hotel_monitoring_int","ohi_rgn_assign_for_municipio_clean.csv"))

estimated_int_dom_arrivals_rgn <- left_join(estimated_int_dom_arrivals_sf, ohi_region_assignments, by = c("municipio_clean")) 
```

Let's aggregate all tourism arrivals by region and type.

```{r}
estimated_int_dom_arrivals_ohirgn <- estimated_int_dom_arrivals_rgn %>% 
  group_by(rgn_id, arrival_type, year) %>% 
  summarize(agg_tourist_arrivals = sum(estimated_arrivals, na.rm = TRUE))
```

##### Dom viz

```{r}
last_3yrs <- c(2022, 2023, 2024)

trend_last3_slope <- estimated_int_dom_arrivals_ohirgn %>%
  st_drop_geometry() %>% 
  filter(year %in% last_3yrs) %>%
  group_by(rgn_id, arrival_type) %>%
  summarise(
    slope_per_year = coef(lm(agg_tourist_arrivals ~ year))[2],
    .groups = "drop"
  )

trend_last3_pct <- estimated_int_dom_arrivals_ohirgn %>%
  st_drop_geometry() %>% 
  filter(year %in% last_3yrs) %>%
  group_by(rgn_id, arrival_type) %>%
  summarise(
    arrivals_2022 = agg_tourist_arrivals[year == 2022],
    arrivals_2024 = agg_tourist_arrivals[year == 2024],
    pct_change_3yr = (arrivals_2024 - arrivals_2022) / arrivals_2022 * 100,
    .groups = "drop"
  ) 

dom_trend_last3yrs <- full_join(trend_last3_pct, trend_last3_slope) %>% 
  filter(arrival_type == "dom") %>% 
  mutate(target_arrivals = arrivals_2024 * 1.098,
         proj_arrivals_2030 = arrivals_2024 + (6 * slope_per_year),
         on_track = ifelse(proj_arrivals_2030 >= target_arrivals, 1, 0))
write_csv(dom_trend_last3yrs, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_viz","dom_trend_last3yrs_arrivals_121725.csv"))

#### join with ohi rgn polygons
ohi <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp")) %>%
  filter(location == "marine") %>% 
  select(rgn_id)

dom_trend_last3yrs_ohigns <- full_join(dom_trend_last3yrs, ohi, by = "rgn_id") %>% 
  st_as_sf()

mapview(dom_trend_last3yrs_ohigns, 
        zcol = "pct_change_3yr",
        col.regions = colorRampPalette(c('#8a0032', '#A50026', '#A50026', '#D73027', '#F46D43', '#FDAE61', '#FEE090', '#FEE090', '#ABD9E9', '#4575B4', '#313695')),
        layer.name = "Percent change of domestic arrivals (2022-2024)",
        legend = TRUE)

########################################

### across the gulf:
trend_overall <- estimated_int_dom_arrivals_ohirgn %>%
  st_drop_geometry() %>% 
  filter(year %in% last_3yrs) %>%
  group_by(arrival_type, year) %>%
  summarise(
    total_arrivals = sum(agg_tourist_arrivals, na.rm = TRUE),
    .groups = "drop"
  )

trend_overall_slope <- trend_overall %>%
  group_by(arrival_type) %>%
  summarise(
    slope_per_year = coef(lm(total_arrivals ~ year))[2]
  )

trend_overall_pct_change <- trend_overall %>%
  group_by(arrival_type) %>%
  summarise(
    pct_change_3yr =
      (total_arrivals[year == 2024] - total_arrivals[year == 2022]) /
      total_arrivals[year == 2022] * 100
  )

dom_trend_goc <- full_join(trend_overall_slope, trend_overall_pct_change) %>% 
  filter(arrival_type == "dom")
write_csv(dom_trend_goc, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_viz","dom_trend_goc_arrivals_121725.csv"))
```


```{r}
#### for visualizations: see the proportion of int vs dom for each rgn
prop_dom_by_rgn <- estimated_int_dom_arrivals_ohirgn %>% 
  group_by(rgn_id, year) %>% 
  mutate(rgn_agg_arr = sum(agg_tourist_arrivals)) %>% 
  ungroup() %>% 
  mutate(prop_rgn_total = agg_tourist_arrivals / rgn_agg_arr) %>% 
  filter(year == 2024) %>% 
  filter(arrival_type == "dom") %>% 
  st_drop_geometry()
write_csv(prop_dom_by_rgn, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_viz","prop_dom_by_rgn_arrivals_121725.csv"))

### join with ohi rgn polygons
ohi <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp")) %>%
  filter(location == "marine") %>% 
  select(rgn_id)

prop_dom_arrivals_ohigns <- full_join(prop_dom_by_rgn, ohi, by = "rgn_id") %>% 
  st_as_sf()

mapview(prop_dom_arrivals_ohigns, 
        zcol = "prop_state_total",
        col.regions = colorRampPalette(c("#096b91", "#a057ce")),
        layer.name = "Proportion of Regionwide Domestic Arrivals (2024)",
        legend = TRUE)

#### prop of total goc
prop_dom_all <- estimated_int_dom_arrivals_ohirgn %>% 
  filter(year == 2024) %>% 
  st_drop_geometry() %>% 
  group_by(year) %>% 
  mutate(all_agg_arr = sum(agg_tourist_arrivals)) %>% 
  ungroup() %>% 
  group_by(arrival_type, year) %>% 
  mutate(total_type_by_yr = sum(agg_tourist_arrivals)) %>% 
  ungroup() %>% 
  select(-c(rgn_id, agg_tourist_arrivals)) %>% 
  distinct() %>% 
  mutate(prop_dom_total = total_type_by_yr / all_agg_arr) %>% 
  filter(arrival_type == "dom")
```


```{r}
baseline_year <- 2024

quantity_current_status_rgn <- estimated_int_dom_arrivals_ohirgn %>% 
  mutate(year = as.numeric(year)) %>%
  
  # group by ohi_rgn + arrival type
  group_by(rgn_id, arrival_type) %>%
  
  mutate(
    # baseline per municipio + type
    baseline_arrivals = agg_tourist_arrivals[year == baseline_year][1],

    # target multiplier varies by type
    target_multiplier = case_when(
      arrival_type == "dom" ~ 1.098, # domestic +9.8%
      arrival_type == "int" ~ 1.30,  # international +30%
      TRUE ~ NA_real_
    ),

    # calculate targets
    target_arrivals = baseline_arrivals * target_multiplier,

    # CURRENT STATUS (0 to 100)
    current_status = (agg_tourist_arrivals / target_arrivals) * 100,
  ) %>% 
  st_drop_geometry()

## save this to outputs
write_csv(quantity_current_status_rgn, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quantity","int_dom_quantity_current_status_120925.csv"))

quantity_current_status_rgn <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quantity","int_dom_quantity_current_status_120925.csv"))
```

### Regional plotlys

International:

```{r}
library(plotly)

int_quantity_current_status <- quantity_current_status_rgn %>% 
  filter(arrival_type %in% "int") %>% 
  arrange(year)

## plot the score over time
final_int_score_plotly <- plotly::plot_ly(
  int_quantity_current_status, 
  x = ~year, 
  y = ~current_status, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: Llegadas de turistas internacionales por municipio",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(int_quantity_current_status$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
final_int_score_plotly

htmlwidgets::saveWidget(final_int_score_plotly, here::here("TR/sectur_quantity_exploration_int","int_current_status_gapfilled_by_municipio_120925.html"), selfcontained = TRUE)

## plot the number of arrivals over time
final_int_arrivals_plotly <- plotly::plot_ly(
  int_quantity_current_status, 
  x = ~year, 
  y = ~agg_tourist_arrivals, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Llegadas de turistas internacionales por region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Aggregated tourist arrivals // Llegadas turísticas agregadas")
  ) 
final_int_arrivals_plotly
```

Domestic:

```{r}
library(plotly)

dom_quantity_current_status <- quantity_current_status_rgn %>% 
  filter(arrival_type %in% "dom") %>% 
  arrange(year)

## plot the score over time
final_dom_score_plotly <- plotly::plot_ly(
  dom_quantity_current_status, 
  x = ~year, 
  y = ~current_status, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: Llegadas de turistas nacionales por municipio",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(dom_quantity_current_status$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
final_dom_score_plotly

htmlwidgets::saveWidget(final_dom_score_plotly, here::here("TR/sectur_quantity_exploration_int","dom_current_status_gapfilled_by_municipio_120925.html"), selfcontained = TRUE)

## plot the number of arrivals over time
final_dom_arrivals_plotly <- plotly::plot_ly(
  dom_quantity_current_status, 
  x = ~year, 
  y = ~agg_tourist_arrivals, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Llegadas de turistas nacionales por region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Aggregated tourist arrivals // Llegadas turísticas agregadas")
  ) 
final_dom_arrivals_plotly
```


### Average by region and year:

```{r}

tr_quantity_overall_cs <- quantity_current_status_rgn %>% 
  group_by(rgn_id, year) %>% 
  summarize(current_status_new = mean(current_status)) %>% 
  ungroup() %>% 
  rename(score = current_status_new) %>% 
  select(rgn_id, year, score)

write_csv(tr_quantity_overall_cs, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quantity","tr_quantity_current_status_121025.csv"))


```

```{r}
library(plotly)

final_quantity_current_status <- tr_quantity_overall_cs %>% 
  arrange(year)

## plot the score over time
final_score_plotly <- plotly::plot_ly(
  final_quantity_current_status, 
  x = ~year, 
  y = ~score, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: TR Quantity",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(final_quantity_current_status$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
final_score_plotly

htmlwidgets::saveWidget(final_score_plotly,here::here("home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quantity","plotly_quantity_current_status_121025.html"), selfcontained = TRUE)
```

