---
title: 'OHI GoC 2025: QUALITY Tourism & Recreation'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: inline
editor_options: 
  chunk_output_type: inline
---

# TR Quality

## Rate of Return

Working on this.

## Service Establishments

See `TR/servicios_de_alojamiento.qmd`.

## Public bathroom availability

Read in the data:

```{r}
library(tidyverse)
library(sf)
library(mapview)

banos_publicos_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_public_bathrooms","banos_publicos_812130_stablecimientoseconomicos.csv"))

banos_publicos_sf <- banos_publicos_raw %>% 
  st_as_sf(coords = c("Longitud", "Latitud"), # making sf object
           crs = 4326, # WGS84
           remove = FALSE) %>% 
  janitor::clean_names()


mapview(banos_publicos_sf)
```

```{r}
inland_50 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer","inland_50km_goc.shp"))

banos_publicos_sf_trans <- st_transform(banos_publicos_sf, st_crs(inland_50))
banos_publicos_goc <- st_intersection(banos_publicos_sf_trans, inland_50)
mapview(banos_publicos_goc, zcol = "descripcion_del_sector_economico")
```



## Crime

### Read in SESNSP GoC data

```{r}
sesnsp_goc_10km <- st_read(here::here("/home/shares/ohi/OHI_GOC/goal_prep/ss/int","sesnsp_goc_municipios_10km.shp")) ## need to fix the names / make a document describing the abbreviations!!

unique(sesnsp_goc_10km$rgn_id) # for 10km: [1] "Region_1" "Region_3" "Region_4" "Region_5" "Region_7" "Region_8" [7] "Region_9"
# for 50km: [1] "Region_1" "Region_3" "Region_4" "Region_5" "Region_7" "Region_8" "Region_9"
```

Out of the crimes listed in `SS/v2025/int_sesnsp/types_of_crimes_goc_sesnsp.csv`, these are the ones I think affect tourism quality the most:

bien_juridico_afectado	tipo_de_delito	subtipo_de_delito
El patrimonio	Daño a la propiedad	Daño a la propiedad
El patrimonio	Extorsión	Extorsión
El patrimonio	Fraude	Fraude
El patrimonio	Otros delitos contra el patrimonio	Otros delitos contra el patrimonio
El patrimonio	Robo	Otros robos
El patrimonio	Robo	Robo a casa habitación
El patrimonio	Robo	Robo a institución bancaria
El patrimonio	Robo	Robo a negocio
El patrimonio	Robo	Robo a transeúnte en espacio abierto al público
El patrimonio	Robo	Robo a transeúnte en vía pública
El patrimonio	Robo	Robo a transportista
El patrimonio	Robo	Robo de autopartes
El patrimonio	Robo	Robo de maquinaria
El patrimonio	Robo	Robo de vehículo automotor
El patrimonio	Robo	Robo en transporte individual
El patrimonio	Robo	Robo en transporte público colectivo
El patrimonio	Robo	Robo en transporte público individual
La familia	Violencia de género en todas sus modalidades distinta a la violencia familiar	Violencia de género en todas sus modalidades distinta a la violencia familiar
La libertad y la seguridad sexual	Abuso sexual	Abuso sexual
La libertad y la seguridad sexual	Acoso sexual	Acoso sexual
La libertad y la seguridad sexual	Hostigamiento sexual	Hostigamiento sexual
La libertad y la seguridad sexual	Otros delitos que atentan contra la libertad y la seguridad sexual	Otros delitos que atentan contra la libertad y la seguridad sexual
La libertad y la seguridad sexual	Violación equiparada	Violación equiparada
La libertad y la seguridad sexual	Violación simple	Violación simple
La sociedad	Corrupción de menores	Corrupción de menores
La sociedad	Otros delitos contra la sociedad	Otros delitos contra la sociedad
La sociedad	Trata de personas	Trata de personas
La vida y la Integridad corporal	Feminicidio	Feminicidio
La vida y la Integridad corporal	Homicidio	Homicidio doloso
La vida y la Integridad corporal	Lesiones	Lesiones dolosas
La vida y la Integridad corporal	Otros delitos que atentan contra la vida y la integridad corporal	Otros delitos que atentan contra la vida y la integridad corporal
Libertad personal	Otros delitos que atentan contra la libertad personal	Otros delitos que atentan contra la libertad personal
Libertad personal	Rapto	Rapto
Libertad personal	Secuestro	Secuestro
Libertad personal	Tráfico de menores	Tráfico de menores
Otros bienes jurídicos afectados (del fuero común)	Amenazas	Amenazas
Otros bienes jurídicos afectados (del fuero común)	Contra el medio ambiente	Contra el medio ambiente
Otros bienes jurídicos afectados (del fuero común)	Delitos cometidos por servidores públicos	Delitos cometidos por servidores públicos
Otros bienes jurídicos afectados (del fuero común)	Evasión de presos	Evasión de presos
Otros bienes jurídicos afectados (del fuero común)	Narcomenudeo	Narcomenudeo
Otros bienes jurídicos afectados (del fuero común)	Otros delitos del Fuero Común	Otros delitos del Fuero Común


```{r}
# vector of subtypes I want to keep for TR quality
tr_qual_crime_subtypes <- c(
  "Daño a la propiedad",
  "Extorsión",
  "Fraude",
  "Otros delitos contra el patrimonio",
  "Otros robos",
  "Robo a casa habitación",
  "Robo a institución bancaria",
  "Robo a negocio",
  "Robo a transeúnte en espacio abierto al público",
  "Robo a transeúnte en vía pública",
  "Robo a transportista",
  "Robo de autopartes",
  "Robo de maquinaria",
  "Robo de vehículo automotor",
  "Robo en transporte individual",
  "Robo en transporte público colectivo",
  "Robo en transporte público individual",
  "Violencia de género en todas sus modalidades distinta a la violencia familiar",
  "Abuso sexual",
  "Acoso sexual",
  "Hostigamiento sexual",
  "Otros delitos que atentan contra la libertad y la seguridad sexual",
  "Violación equiparada",
  "Violación simple",
  "Corrupción de menores",
  "Otros delitos contra la sociedad",
  "Trata de personas",
  "Feminicidio",
  "Homicidio doloso",
  "Lesiones dolosas",
  "Otros delitos que atentan contra la vida y la integridad corporal",
  "Otros delitos que atentan contra la libertad personal",
  "Rapto",
  "Secuestro",
  "Tráfico de menores",
  "Amenazas",
  "Contra el medio ambiente",
  "Delitos cometidos por servidores públicos",
  "Evasión de presos",
  "Narcomenudeo",
  "Otros delitos del Fuero Común"
)

# filter sesnsp_goc by those subtypes
sesnsp_goc_tr_qual <- sesnsp_goc_10km %>%
  filter(sbtp_d_ %in% tr_qual_crime_subtypes)

unique(sesnsp_goc_tr_qual$rgn_id) # [1] "Region_1" "Region_3" "Region_4" "Region_5" "Region_7" "Region_8" "Region_9"

```

### Now, summarize the number of crimes by year and region (assuming that all crimes are weighted equally)

```{r}
sesnsp_goc_tr_qual_sum <- sesnsp_goc_tr_qual %>% 
  st_drop_geometry() %>% 
  group_by(ano, rgn_id) %>% 
  summarise(number_crimes = sum(annl_tt))

unique(sesnsp_goc_tr_qual_sum$rgn_id) # [1] "Region_1" "Region_3" "Region_4" "Region_5" "Region_7" "Region_8" "Region_9"
```

#### Divide by coastal population in each region

##### NOTE: NEED TO UPDATE...

1. We will have to update this because we eventually want to use 2015, 2020, and 2025 census data for population, not just static 2020.
2. I think maybe we should use the population per municipality (instead of population per OHI Region).  Because for some regions there are multiple municipios in the region that have data collected for them, versus some only have one or two.  

```{r}
coastal_10km_pop <- read_csv(here::here("general","OHI_rgn_population_10km.csv")) 

coastal_10km_pop_sumrgn9 <- coastal_10km_pop %>% 
  group_by(rgn_id) %>% 
  summarize(rgn_population_10km = sum(population_10km_2020))

sesnsp_goc_tr_sum_percapita <- left_join(sesnsp_goc_tr_qual_sum, coastal_10km_pop_sumrgn9, by = "rgn_id")

sesnsp_goc_tr_crimes_percapita <- sesnsp_goc_tr_sum_percapita %>% 
  group_by(ano, rgn_id) %>% 
  mutate(crimes_percapita = 100 * (number_crimes / rgn_population_10km))
```

#### Prelim Scores

```{r}
max_crime_percapita_allrgns <- max(sesnsp_goc_tr_crimes_percapita$crimes_percapita, na.rm = TRUE)

tr_qual_crime_score_alt <- sesnsp_goc_tr_crimes_percapita %>%
  mutate(
    current_status = (1 - (crimes_percapita / max_crime_percapita_allrgns)) * 100
  ) 

write_csv(tr_qual_crime_score, here::here("TR/tr_quality_int","tr_qual_crime_score_111025.csv"))
```

#### Plotly

```{r}
library(plotly)

## TR quality crime index score over time
tr_qual_crime_plotly <- plotly::plot_ly(
  tr_qual_crime_score,
  x = ~ano, 
  y = ~current_status, 
  color = ~rgn_id,
  type = "scatter", 
  mode = "lines"
) %>%
  layout(
    title = "CURRENT STATUS: Índice de delitos per cápita que afectan la calidad turística por región",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Score (out of 100)"),
    legend = list(title = list(text = "Región"))
  ) %>%
  add_lines(
    x = range(tr_qual_crime_score$ano), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (0 crimes = 100/100)",
    inherit = FALSE
  )

tr_qual_crime_plotly
```

### Assuming that when there is no data, that the score is 100...only for regions that have data at some point.

If there are no municipios in that OHI region then they should not be gapfilled like this.

```{r}
## fill in regions missing any values with 100 since the summarize may have dropped any regions with 0. fix this later??
tr_qual_crime_score_filled <- tr_qual_crime_score %>%
  select(ano, rgn_id, current_status) %>%
  complete(ano, rgn_id, fill = list(current_status = 100))

tr_qual_crime_plotly_filled <- plot_ly(
  tr_qual_crime_score_filled,
  x = ~ano,
  y = ~current_status,
  color = ~rgn_id,
  type = "scatter",
  mode = "lines"
) %>%
  layout(
    title = "CURRENT STATUS: Índice de delitos que afectan la calidad turística por región",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Score (out of 100)"),
    legend = list(title = list(text = "Región"))
  ) %>%
  add_lines(
    x = range(tr_qual_crime_score_filled$ano),
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (0 crimes = 100/100)",
    inherit = FALSE
  )

tr_qual_crime_plotly_filled
```

```{r}
# bring back geometry
inland_50 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer","inland_50km_goc.shp"))

tr_qual_crime_score_filled_sf <- inland_50 %>%
  left_join(tr_qual_crime_score_filled, by = "rgn_id")

# st_write(tr_qual_crime_score_filled_sf, here::here("TR/tr_quality_int","tr_qual_crime_score_filled_sf.shp"))

tr_qual_crime_score_filled_sf_2025 <- tr_qual_crime_score_filled_sf %>% 
  filter(ano %in% "2025")

mapview(tr_qual_crime_score_filled_sf_2025, zcol = "current_status")
```


