---
title: "Tourist data"
output: html_document
date: "2025-3-6"
editor_options: 
  chunk_output_type: console
---
Data: "Ocupación en hoteles y moteles por centro turístico, corte mensual DataTur"
Go to: <https://monitoreohotelero.sectur.gob.mx/Reportes/Menu.aspx>

To download:

> 1. Go to the “Generate Report” section.
> 2. Select the “Mensual”, or monthly, time frame.
> 3. Choose the option “Hotel and motel occupancy by tourist center, monthly DataTur snapshot.”
> 4. Click “Continue” and then “Advanced criteria” (Criterios Avanzados)
> 5. Select the municipality of interest, then all the variables of interest.  Specify "Serie" for "Cobertura Temporal corte mensual", starting from 2018 January (Enero) to 2025 Junio (June).
> 6. Finally, click on “Generate Report.”

The report can be downloaded in various formats and contains the information available as of the date of consultation.  It is important to note that this information is presented at the tourist center level, not the municipal level. However, it can be useful for territorial approximations in the Gulf of California region, particularly in key tourist destinations.

```{r}
library(mapview)
library(sf)
library(tidyverse)
library(here)
library(terra)
```


Okay!! So it looks like you CANNOT save the file as a .CSV first, instead you **must** to do it in this order:

1.  On the SECTUR report site, after generating the report, click "Exportar a Excel".  

2.  Open the ".xls" (which is really a .htm) in Excel, and then "Save As..." a UTF-8 .CSV. Rename it `DataturReportePublico_municipio_resaved.csv` where you replace municipio with the municipality you just saved. 

3.  Bring into mazu in `/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/resaved_as_csv` to be read in here.  This is the only way that you do not get any "†" in whitespaces.

**These instructions will be brought to the top of the script when it is cleaned up.**

# Clean data 

```{r}
clean_hotel_occupations_data <- function(municipio) {
  
  # where to find the file
  file_path <- here::here(
    "/home/shares/ohi/OHI_GOC/_raw_data/SECTUR_Datatur_Hoteles/resaved_as_csv",
    paste0("DataturReportePublico_", municipio, "_resaved.csv")
  )
  
  # read raw data
  # raw_data <- read_csv(file_path)
  raw_data <- read_csv(file_path, locale = readr::locale(encoding = "UTF-8")) # changed because having error: "Error in `map()`: ℹ In index: 1. Caused by error in `nchar()`: ! invalid multibyte string, element 1"
  
  # clean the column names, remove top rows, and tidy
  cleaned <- raw_data %>% 
    set_names(.[9, ]) %>%  
    slice(-c(1:10)) %>% # NOTE: not all are the same length, so some of them will have the bottom text (ex: ND No disponible) as part of the output, and there will have to be extra cleaning done at the end.
    janitor::clean_names()
  
  # clean and standardize the "concepto" column
  cleaned_concept <- cleaned %>%
    mutate(
      concept = str_to_lower(concepto),
      concept = str_trim(concept),
      concept = str_replace_all(concept, " ", "_")
    ) %>% 
    relocate(concept, .after = concepto) %>% # these two lines of code are not really necessary because the func will do it all at once
    select(-concepto)
  
  # identify topic headers and combine them with subcategories
  by_topic <- cleaned_concept %>% 
    mutate(
      is_topic = !str_detect(concept, "^[1-5]_estrella[s]?|total_categorias") & !is.na(concept)
    ) %>%
    mutate(header_concept = if_else(is_topic, concept, NA_character_)) %>% # use NA_character_ for this stuff just in case
    fill(header_concept) %>% 
    mutate(
      concept_full = if_else(concept != header_concept, paste0(header_concept, ":", concept), concept)
    )
  
  # pivot longer
  long <- by_topic %>% 
    pivot_longer(
      cols = 2:100, # adjust if more columns
      names_to = "month_year",
      values_to = "value"
    )
  
  # separate year and month, filter, and tag city
  final <- long %>% 
    mutate(
      is_mo_yr = !str_detect(month_year, "acum|total"),
      temp = if_else(is_mo_yr, month_year, NA_character_)
    ) %>% 
    separate(temp, into = c("month", "year"), sep = "_", remove = FALSE) %>% 
    filter(is_topic == FALSE) %>% 
    select(header_concept, concept, concept_full, month_year, month, year, value) %>% 
    mutate(municipio = str_to_lower(municipio)) %>% 
    mutate(value = as.numeric(str_replace_all(value, ",", "")))
  
  return(final)
}
```


## Apply function to all municipios

```{r}
all_tourist_centers <- c(
  "Cabo_San_Lucas", "Culiacan", "El_Fuerte", "El_Rosario", "Escuinapa",
  "Guaymas", "Hermosillo", "La_Paz", "Loreto", "Los_Mochis",
  "Mazatlan", "Puerto_Vallarta", "San_Felipe", "San_Jose_del_Cabo", 
  "Zona_Corredor_Los_Cabos"
)

library(purrr)

all_data <- all_tourist_centers %>% 
  map(clean_hotel_occupations_data) %>% 
  bind_rows() # looks good! keep checking after lunch.

all_data_clean <- all_data %>% 
  rename( 
    metric = header_concept,
    category = concept
  ) %>%  # for better clarity, since we are done cleaning and these names make more sense now
  mutate(
    concept_full = str_replace_all(concept_full, "\U{0eef2f}th>|\U{0ee832}|\U{0ee82e}", "_"),
    # metric_label = str_replace_all(metric, "_", " "),
    # metric_label = str_to_title(metric_label), # for later viz 
    metric = str_replace_all(metric, "\U{0eef2f}th>|\U{0ee832}|\U{0ee82e}", "_"),
    category = str_replace_all(category, "\U{0eef2f}th>|\U{0ee832}|\U{0ee82e}", "_"),
    value = str_replace_all(value, ",",""),
    year = str_replace_all(year, "p",""), # only in 2025, assume it means "preliminary"
    month = case_when(
      tolower(month) == "enero" ~ "01",
      tolower(month) == "febrero" ~ "02",
      tolower(month) == "marzo" ~ "03",
      tolower(month) == "abril" ~ "04",
      tolower(month) == "mayo" ~ "05",
      tolower(month) == "junio" ~ "06",
      tolower(month) == "julio" ~ "07",
      tolower(month) == "agosto" ~ "08",
      tolower(month) == "septiembre" ~ "09",
      tolower(month) == "octubre" ~ "10",
      tolower(month) == "noviembre" ~ "11",
      tolower(month) == "diciembre" ~ "12",
      TRUE ~ NA_character_  # otherwise
    ) # for lubridate later
  ) %>% 
  filter(!category %in% NA) 

unique(all_data_clean$concept_full) # check that there are no weirdnesses left

# data_check <- all_data_clean %>% filter(year %in% 2024) %>% filter(metric %in% "cuartos_ocupados_no_residentes") # double check with raw.  looks good!
```


```{r}
# so that we can have the yearly total, not just by month, because we will have the scores per year.
all_data_yearly <- all_data_clean %>% 
  filter(str_detect(month_year, "total_anual_")) %>% 
  select(-c(month, year)) %>% 
  separate_wider_delim(., month_year, delim = "_anual_", names = c("type","year"))

## save into int
write_csv(all_data_yearly, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/int","hotel_data_cleaned_yearly.csv"))
```

# OLD Domestic vs. International

Since there are different targets within Plan México for Domestic versus International visitors:

> "Para lograr la meta de ser el quinto país visitado del mundo en el 2030 se estima que es necesario aumentar tres millones más de turistas cada año, así como incrementar:

> Los cuartos de hotel en 12 por ciento.
> Del turismo nacional en 9.8 por ciento.
> La llegada de visitantes internacionales en 30 por ciento."

What we will do is determine the current status for domestic and international visitors separately, and then average those two to produce the tourism:quantity current status.

```{r}
unique(all_data_yearly$metric)

#  [1] "cuartos_registrados_fin_per_o"      "cuartos_disponibles"               
#  [3] "cuartos_disponibles_promedio"       "cuartos_ocupados"                  
#  [5] "cuartos_ocupados_residentes"        "cuartos_ocupados_no_residentes"    
#  [7] "llegadas_de_turistas"               "llegadas_de_turistas_residentes"   
#  [9] "llegadas_de_turistas_no_residentes" "turistas_noche"                    
# [11] "turistas_noche_residentes"          "turistas_noche_no_residentes"      
# [13] "porcentaje_de_ocupaci_"             "porcentaje_de_ocupaci_esidentes"   
# [15] "porcentaje_de_ocupaci_o_residentes" "estad_promedio"                    
# [17] "estad_promedio_residentes"          "estad_promedio_no_residentes"      
# [19] "densidad_de_ocupaci_"               "densidad_de_ocupaci_esidentes"     
# [21] "densidad_de_ocupaci_o_residentes" 
```


### Domestic

```{r}
## "llegadas_de_turistas_residentes" = domestic visitors
dom_arrivals <- all_data_yearly %>% 
  filter(metric %in% "llegadas_de_turistas_residentes") %>% # domestic arrivals
  filter(category %in% "total_categorias") # since we do not care what star designation it is
```

```{r}
# let's make them an sf object
dom_geocode <- dom_arrivals %>% 
  mutate(country = "Mexico") %>% 
  mutate(estado = case_when(
    municipio %in% c("cabo_san_lucas", "la_paz", "loreto", "san_jose_del_cabo", "zona_corredor_los_cabos") ~ "Baja California Sur",
    municipio %in% c("san_felipe") ~ "Baja California",
    municipio %in% c("guaymas", "hermosillo") ~ "Sonora",
    municipio %in% c("culiacan", "el_fuerte", "los_mochis", "mazatlan", "escuinapa", "el_rosario") ~ "Sinaloa",
    municipio %in% c("puerto_vallarta") ~ "Jalisco",
    TRUE ~ ""  # default case for any unmatched cities
  )) %>% 
  tidygeocoder::geocode(city = municipio, country = country, state = estado) # to give each city a lat, long
  
dom_geocode_sf <- dom_geocode %>% 
  mutate(lat = ifelse(municipio %in% "zona_corredor_los_cabos", 23.1539, lat)) %>% # adding in the coordinates manually from <https://latitude.to/articles-by-country/mx/mexico/56998/los-cabos-corridor>
  mutate(long = ifelse(municipio %in% "zona_corredor_los_cabos", -109.720233, long)) %>% 
  st_as_sf(coords = c("long", "lat"),
           crs = 4326, # WGS84
           remove = FALSE)

mapview(dom_geocode_sf)

############################################################
# Now, let us assign each tourist center to an OHI region. 
############################################################

goc_poly <- st_read(here("spatial/ohi_regions","goc_ohi_rgns.shp"))
st_crs(goc_poly) # not WGS84: Mexico ITRF2008 / LCC
mapview(goc_poly)

dom_geocode_sf_transform <- st_transform(dom_geocode_sf, crs = st_crs(goc_poly)) %>%  # transform the tourist centers to Mexico ITRF2008 / LCC
   mutate(value = as.numeric(value))
crs(dom_geocode_sf_transform)

dom_goc <- st_intersection(dom_geocode_sf_transform, goc_poly) %>% 
  mutate(arrivals_type = "domestic")
mapview(dom_goc) # now each tourist center has associated region data

##############################
#A cool plot to see them each:
##############################

# map of all goc tourist centers
mapview(dom_goc, zcol = "rgn_id") + mapview(goc_poly, zcol = "rgn_id")
```


The reference point could be a 9.8% increase in domestic tourist arrivals by 2030, which starting from 2024 is 5 years (since you would want to reach that goal at the end of 2029 to reach the target *by* 2030). <https://tourismanalytics.com/blog-posts/how-mexicos-forty-percent-tourism-growth-plan-aims-for-fifth-global-spot-by-2030-with-twenty-billion-dollar-investment-in-infrastructure-and-market-expansion>

Therefore, each year, domestic tourism arrivals should increase by 1.96%.

The most current full year of data is 2024.  Plan México was announced on January 13, 2025, which means the 9.8% tourism growth target by 2030 would logically use 2024 as the baseline year.

So, the current status for domestic quantity could be:

(Current arrivals / Target arrivals) × 100

Score of 100 = on track to meet the 2030 target
Score < 100 = below target expected level
Score > 100 = exceeding target, yay!

*We will use linear growth, even if it is not as realistic, for now.*

```{r}
growth_rate_linear_dom <- 0.0196  # 1.96% per year (9.8/5, linear growth, for now)
baseline_year <- 2024

dom_quantity_score <- dom_goc %>%
  group_by(municipio, rgn_id) %>%
  mutate(
    year = as.numeric(year),
    
    # determine the baseline value for each municipio (2024)
    dom_baseline_arrivals = value[year == baseline_year][1],

    # to know how far from 2024 we are (note: negative numbers are because of years before 2024)
    years_since_baseline = year - baseline_year,

    # target arrivals for domestic visitors
    dom_target_arrivals = dom_baseline_arrivals * (1 + (growth_rate_linear_dom * years_since_baseline)),  # the target number of arrivals should change based on how many years it is from the baseline. For example, two years from 2024 should have a growth of 1.96% * 2.  

    # make it a score
    dom_index_score = (value / dom_target_arrivals) * 100,
    dom_gap_to_target = dom_target_arrivals - value,
    dom_gap_percentage = ((dom_target_arrivals - value) / value) * 100
  ) %>%
  ungroup() %>%
  select(municipio, rgn_id, year, dom_current_arrivals = value,
         dom_baseline_arrivals, dom_target_arrivals, dom_index_score,
         dom_gap_to_target, dom_gap_percentage, everything())
```


```{r}
library(plotly)

## plot the score over time
dom_score_plotly <- plotly::plot_ly(
  dom_quantity_score, 
  x = ~year, 
  y = ~dom_index_score, 
  color = ~municipio, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "CURRENT STATUS: Llegadas de turistas nacionales por municipio",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(dom_quantity_score$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
dom_score_plotly

htmlwidgets::saveWidget(dom_score_plotly, here::here("TR/sectur_quantity_exploration_int","dom_arrivals_score_100825.html"), selfcontained = TRUE)


## compare to the patterns of domestic arrivals over time
dom_arrivals_plot <- plotly::plot_ly(
  dom_goc, 
  x = ~year, 
  y = ~value, 
  color = ~municipio, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "Llegadas de turistas nacionales por municipio (domestic arrivals)",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Llegadas // Arrivals")
  ) %>% 
  add_lines(
    x = range(dom_goc$year),
    y = c(100, 100),
    line = list(color = "white", width = 2),
    name = "",
    inherit = FALSE
  )
dom_arrivals_plot

htmlwidgets::saveWidget(dom_arrivals_plot, here::here("TR/sectur_quantity_exploration_int","dom_arrivals_plot_100825.html"), selfcontained = TRUE)
```


### International

```{r}
## "llegadas_de_turistas_no_residentes" = international visitors
int_arrivals <- all_data_yearly %>% 
  filter(metric %in% "llegadas_de_turistas_no_residentes")%>% # int arrivals
  filter(category %in% "total_categorias") # since we do not care what star designation it is
```

```{r}
# let's make them an sf object
int_geocode <- int_arrivals %>% 
  mutate(country = "Mexico") %>% 
  mutate(estado = case_when(
    municipio %in% c("cabo_san_lucas", "la_paz", "loreto", "san_jose_del_cabo", "zona_corredor_los_cabos") ~ "Baja California Sur",
    municipio %in% c("san_felipe") ~ "Baja California",
    municipio %in% c("guaymas", "hermosillo") ~ "Sonora",
    municipio %in% c("culiacan", "el_fuerte", "los_mochis", "mazatlan", "escuinapa", "el_rosario") ~ "Sinaloa",
    municipio %in% c("puerto_vallarta") ~ "Jalisco",
    TRUE ~ ""  # default case for any unmatched cities
  )) %>% 
  tidygeocoder::geocode(city = municipio, country = country, state = estado) # to give each city a lat, long
  
int_geocode_sf <- int_geocode %>% 
  mutate(lat = ifelse(municipio %in% "zona_corredor_los_cabos", 23.1539, lat)) %>% # adding in the coordinates manually from <https://latitude.to/articles-by-country/mx/mexico/56998/los-cabos-corridor>
  mutate(long = ifelse(municipio %in% "zona_corredor_los_cabos", -109.720233, long)) %>% 
  st_as_sf(coords = c("long", "lat"),
           crs = 4326, # WGS84
           remove = FALSE)

mapview(int_geocode_sf)

############################################################
# Now, let us assign each tourist center to an OHI region. 
############################################################

goc_poly <- st_read(here("spatial/ohi_regions","goc_ohi_rgns.shp"))
st_crs(goc_poly) # not WGS84: Mexico ITRF2008 / LCC
mapview(goc_poly)

int_geocode_sf_transform <- st_transform(int_geocode_sf, crs = st_crs(goc_poly)) %>%  # transform the tourist centers to Mexico ITRF2008 / LCC
   mutate(value = as.numeric(value))
crs(int_geocode_sf_transform)

int_goc <- st_intersection(int_geocode_sf_transform, goc_poly) %>% 
  mutate(arrivals_type = "international")
mapview(int_goc) # now each tourist center has associated region data

##############################
#A cool plot to see them each:
##############################

# map of all goc tourist centers
mapview(int_goc, zcol = "rgn_id") + mapview(goc_poly, zcol = "rgn_id")
```

The reference point could be a 30% increase in international tourist arrivals by 2030, which starting from 2024 is 5 years (since you would want to reach that goal at the end of 2029 to reach the target *by* 2030). <https://tourismanalytics.com/blog-posts/how-mexicos-forty-percent-tourism-growth-plan-aims-for-fifth-global-spot-by-2030-with-twenty-billion-dollar-investment-in-infrastructure-and-market-expansion>

Therefore, each year, international tourism arrivals should increase by 6%.

The most current full year of data is 2024.  Plan México was announced on January 13, 2025, which means the 30% tourism growth target by 2030 would logically use 2024 as the baseline year.

So, the current status for domestic quantity could be:

(Current arrivals / Target arrivals) × 100

Score of 100 = on track to meet the 2030 target
Score < 100 = below target expected level
Score > 100 = exceeding target, yay!

*We will use linear growth, even if it is not as realistic, for now.*

```{r}
growth_rate_linear_int <- 0.06 # 1.96% per year (30/5, linear growth, for now)
baseline_year <- 2024

int_quantity_score <- int_goc %>%
  group_by(municipio, rgn_id) %>%
  mutate(
    year = as.numeric(year),
    
    # determine the baseline value for each municipio (2024)
    int_baseline_arrivals = value[year == baseline_year][1],

    # to know how far from 2024 we are (note: negative numbers are because of years before 2024)
    years_since_baseline = year - baseline_year,

    # target arrivals for international visitors
    int_target_arrivals = int_baseline_arrivals * (1 + (growth_rate_linear_int * years_since_baseline)),  # the target number of arrivals should change based on how many years it is from the baseline. For example, two years from 2024 should have a growth of 1.96% * 2.  

    # make it a score
    int_index_score = (value / int_target_arrivals) * 100,
    int_gap_to_target = int_target_arrivals - value,
    int_gap_percentage = ((int_target_arrivals - value) / value) * 100
  ) %>%
  ungroup() %>%
  select(municipio, rgn_id, year, int_current_arrivals = value,
         int_baseline_arrivals, int_target_arrivals, int_index_score,
         int_gap_to_target, int_gap_percentage, everything())
```


```{r}
library(plotly)

## plot the score over time
int_score_plotly <- plotly::plot_ly(
  int_quantity_score, 
  x = ~year, 
  y = ~int_index_score, 
  color = ~municipio, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "CURRENT STATUS: Llegadas de turistas internacionales por municipio",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(int_quantity_score$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
int_score_plotly

htmlwidgets::saveWidget(int_score_plotly, here::here("TR/sectur_quantity_exploration_int","int_arrivals_score_100825.html"), selfcontained = TRUE)


## compare to the patterns of international arrivals over time
int_arrivals_plot <- plotly::plot_ly(
  int_goc, 
  x = ~year, 
  y = ~value, 
  color = ~municipio, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "Llegadas de turistas internacionales por municipio (international arrivals)",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Llegadas // Arrivals")
  ) %>% 
  add_lines(
    x = range(int_goc$year),
    y = c(100, 100),
    line = list(color = "white", width = 2),
    name = "",
    inherit = FALSE
  )
int_arrivals_plot

htmlwidgets::saveWidget(int_arrivals_plot, here::here("TR/sectur_quantity_exploration_int","int_arrivals_plot_100825.html"), selfcontained = TRUE)
```




### Merge

```{r}
# join int_quantity_score and dom_quantity_score

## check for duplicates
dom_dups <- dom_quantity_score %>%
  st_drop_geometry() %>%
  count(municipio, rgn_id, year) %>%
  filter(n > 1)

int_dups <- int_quantity_score %>%
  st_drop_geometry() %>%
  count(municipio, rgn_id, year) %>%
  filter(n > 1)

dom_dups    # correct: they should not have any duplicates!!
int_dups

dom_df <- dom_quantity_score %>% st_drop_geometry()
int_df <- int_quantity_score %>% st_drop_geometry()

quantity_scores <- full_join(
  dom_df,
  int_df,
  by = c("municipio", "rgn_id", "year"),
  suffix = c(".dom", ".int")
)

quantity_scores %>%
  mutate(has_dom = !is.na(dom_index_score),
         has_int = !is.na(int_index_score)) %>%
  count(has_dom, has_int)

# Missing int
quantity_scores %>% filter(is.na(int_index_score)) %>% select(municipio, rgn_id, year)

# Missing dom
quantity_scores %>% filter(is.na(dom_index_score)) %>% select(municipio, rgn_id, year)

dom_geom <- dom_quantity_score %>%
  select(municipio, rgn_id, geometry) %>%
  distinct(municipio, rgn_id, .keep_all = TRUE)
```


```{r}
quantity_scores_sf <- quantity_scores %>%
  left_join(dom_geom, by = c("municipio", "rgn_id")) %>%
  st_as_sf() %>% 
  select(municipio, rgn_id, year, dom_index_score, int_index_score, geometry) %>% 
  mutate(
    current_status_quantity = (dom_index_score + int_index_score) / 2
  )
```

```{r}
## plot the score over time
quantity_score_plotly <- plotly::plot_ly(
  quantity_scores_sf, 
  x = ~year, 
  y = ~current_status_quantity, 
  color = ~municipio, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "CURRENT STATUS: Llegadas de turistas (todos) por municipio",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(quantity_scores_sf$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
quantity_score_plotly

htmlwidgets::saveWidget(quantity_score_plotly, here::here("TR/sectur_quantity_exploration_int","all_arrivals_score_100825.html"), selfcontained = TRUE)

```

```{r}
# ## averaged score per region
# quantity_scores_rgn <- quantity_scores_sf %>% 
#   group_by(rgn_id) %>% 
#   summarize(quantity_score_rgn = mean(current_status_quantity))
# 
# mapview(quantity_scores_rgn)
```

# NEW Domestic vs. International

We are updating the reference point to be the exact number of visitors per year that are 30% more than the international visitors in 2024 and 9.8% more than domestic visitors in 2024.  That way, the score is a reflection of how close the number of arrivals are to the target (it likely would not be a linear increase until 2030).

Therefore, the target/baseline would be: 

**International**
The # int arrivals for that year / 
The # int arrivals in 2024 * 1.30

**Domestic**
The # domestic arrivals for that year / 
The # domestic arrivals in 2024 * 1.098

### Domestic

```{r}
## "llegadas_de_turistas_residentes" = domestic visitors
dom_arrivals <- all_data_yearly %>% 
  filter(metric %in% "llegadas_de_turistas_residentes") %>% # domestic arrivals
  filter(category %in% "total_categorias") # since we do not care what star designation it is
```

```{r}
# let's make them an sf object
dom_geocode <- dom_arrivals %>% 
  mutate(country = "Mexico") %>% 
  mutate(estado = case_when(
    municipio %in% c("cabo_san_lucas", "la_paz", "loreto", "san_jose_del_cabo", "zona_corredor_los_cabos") ~ "Baja California Sur",
    municipio %in% c("san_felipe") ~ "Baja California",
    municipio %in% c("guaymas", "hermosillo") ~ "Sonora",
    municipio %in% c("culiacan", "el_fuerte", "los_mochis", "mazatlan", "escuinapa", "el_rosario") ~ "Sinaloa",
    municipio %in% c("puerto_vallarta") ~ "Jalisco",
    TRUE ~ ""  # default case for any unmatched cities
  )) %>% 
  tidygeocoder::geocode(city = municipio, country = country, state = estado) # to give each city a lat, long
  
dom_geocode_sf <- dom_geocode %>% 
  mutate(lat = ifelse(municipio %in% "zona_corredor_los_cabos", 23.1539, lat)) %>% # adding in the coordinates manually from <https://latitude.to/articles-by-country/mx/mexico/56998/los-cabos-corridor>
  mutate(long = ifelse(municipio %in% "zona_corredor_los_cabos", -109.720233, long)) %>% 
  st_as_sf(coords = c("long", "lat"),
           crs = 4326, # WGS84
           remove = FALSE)

mapview(dom_geocode_sf)

############################################################
# Now, let us assign each tourist center to an OHI region. 
############################################################

goc_poly <- st_read(here("spatial/ohi_regions","goc_ohi_rgns.shp"))
st_crs(goc_poly) # not WGS84: Mexico ITRF2008 / LCC
mapview(goc_poly)

dom_geocode_sf_transform <- st_transform(dom_geocode_sf, crs = st_crs(goc_poly)) %>%  # transform the tourist centers to Mexico ITRF2008 / LCC
   mutate(value = as.numeric(value))
crs(dom_geocode_sf_transform)

dom_goc <- st_intersection(dom_geocode_sf_transform, goc_poly) %>% 
  mutate(arrivals_type = "domestic")
mapview(dom_goc) # now each tourist center has associated region data

##############################
#A cool plot to see them each:
##############################

# map of all goc tourist centers
mapview(dom_goc, zcol = "rgn_id") + mapview(goc_poly, zcol = "rgn_id")
```


Like stated earlier, the reference point has been updated to use specific target values rather than annual percentage increases (<https://tourismanalytics.com/blog-posts/how-mexicos-forty-percent-tourism-growth-plan-aims-for-fifth-global-spot-by-2030-with-twenty-billion-dollar-investment-in-infrastructure-and-market-expansion>).

For **domestic tourism**, the 2030 target is defined as **9.8% higher than 2024 domestic visitor arrivals**.

Therefore, the current status for each year is calculated as follows:

**Domestic:**
$$
\text{Score} = \frac{\text{Domestic arrivals (year)}}{\text{Domestic arrivals in 2024} \times 1.098} \times 100
$$

A **score of 100** indicates that the number of arrivals has **reached the 2030 target level** (i.e., 9.8% or 30% higher than 2024 levels).

* A score < 100 means arrivals are **below the target trajectory**.
* A score > 100 means arrivals have **exceeded the 2030 target**.

Those two current status scores are then averaged to produce an overall score of how close each region is to making the 2030 target overall. 

```{r}
target_baseline_increase <- 1.098  # 9.8% more visitors than the amount in 2024
baseline_year <- 2024

dom_quantity_score <- dom_goc %>%
  group_by(municipio) %>% #### I made a mistake in the old version: we want to calculate current status scores per municipio first, then average those for the entire region. FOR NOW, let's just have the scores by municipio.
  mutate(
    year = as.numeric(year),
    
    # determine the baseline value for each municipio (2024)
    dom_baseline_arrivals = value[year == baseline_year][1],

    # to know how far from 2024 we are (note: negative numbers are because of years before 2024)
    years_since_baseline = year - baseline_year,

    # target arrivals for domestic visitors
    dom_target_arrivals = dom_baseline_arrivals * target_baseline_increase,   # for domestic tourism, the 2030 target is defined as 9.8% higher than 2024 domestic visitor arrivals.

    # make it a score
    dom_index_score = (value / dom_target_arrivals) * 100,
    dom_gap_to_target = dom_target_arrivals - value,
    dom_gap_percentage = ((dom_target_arrivals - value) / value) * 100 ## target arrivals - current arrivals / current arrivals... ideal is 0 or negative to reflect how much needs to be done (if you reach the target, it is negative, because nothing needs to be done!)
  ) %>%
  ungroup() %>%
  select(municipio, rgn_id, year, dom_current_arrivals = value,
         dom_baseline_arrivals, dom_target_arrivals, dom_index_score,
         dom_gap_to_target, dom_gap_percentage, everything())


##### double-checking that it all looks good #####
# for san felipe:
# > 139861 * 1.098
# [1] 153567.4
# > 153567.4-139861
# [1] 13706.4
# > 13706.4/139861
# [1] 0.09800016
# > 66410/153567.4
# [1] 0.4324486

# and cabo san lucas: 
# cabo_san_lucas. Region_4 2019 274384 243965 267873.6
# it makes sense that the dom_gap_percentage is negative, because in 2019 the number of domestic arrivals is actually more than the 2024 * 1.098.

```


```{r}
library(plotly)

## plot the score over time
updated_dom_score_plotly <- plotly::plot_ly(
  dom_quantity_score, 
  x = ~year, 
  y = ~dom_index_score, 
  color = ~municipio, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "CURRENT STATUS: Llegadas de turistas nacionales por municipio",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(dom_quantity_score$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
updated_dom_score_plotly

htmlwidgets::saveWidget(updated_dom_score_plotly, here::here("TR/sectur_quantity_exploration_int","dom_arrivals_score_101325.html"), selfcontained = TRUE)
```

YAY, this makes sense!! So for 2024 rows:

dom_baseline_arrivals = value *(for 2024)*

Therefore, using the equation to get the current status:

$$
\text{dom_index_score}_{2024} = \frac{\text{value}}{\text{value} \times 1.098} \times 100 = \frac{100}{1.098} = 91.07468
$$


```{r}
## compare to the patterns of domestic arrivals over time
dom_arrivals_plot <- plotly::plot_ly(
  dom_goc, 
  x = ~year, 
  y = ~value, 
  color = ~municipio, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "Llegadas de turistas nacionales por municipio (domestic arrivals)",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Llegadas // Arrivals")
  ) %>% 
  add_lines(
    x = range(dom_goc$year),
    y = c(100, 100),
    line = list(color = "white", width = 2),
    name = "",
    inherit = FALSE
  )
dom_arrivals_plot

htmlwidgets::saveWidget(dom_arrivals_plot, here::here("TR/sectur_quantity_exploration_int","dom_arrivals_plot_100825.html"), selfcontained = TRUE)
```

### International

```{r}
## "llegadas_de_turistas_no_residentes" = international visitors
int_arrivals <- all_data_yearly %>% 
  filter(metric %in% "llegadas_de_turistas_no_residentes")%>% # int arrivals
  filter(category %in% "total_categorias") # since we do not care what star designation it is
```

```{r}
# let's make them an sf object
int_geocode <- int_arrivals %>% 
  mutate(country = "Mexico") %>% 
  mutate(estado = case_when(
    municipio %in% c("cabo_san_lucas", "la_paz", "loreto", "san_jose_del_cabo", "zona_corredor_los_cabos") ~ "Baja California Sur",
    municipio %in% c("san_felipe") ~ "Baja California",
    municipio %in% c("guaymas", "hermosillo") ~ "Sonora",
    municipio %in% c("culiacan", "el_fuerte", "los_mochis", "mazatlan", "escuinapa", "el_rosario") ~ "Sinaloa",
    municipio %in% c("puerto_vallarta") ~ "Jalisco",
    TRUE ~ ""  # default case for any unmatched cities
  )) %>% 
  tidygeocoder::geocode(city = municipio, country = country, state = estado) # to give each city a lat, long
  
int_geocode_sf <- int_geocode %>% 
  mutate(lat = ifelse(municipio %in% "zona_corredor_los_cabos", 23.1539, lat)) %>% # adding in the coordinates manually from <https://latitude.to/articles-by-country/mx/mexico/56998/los-cabos-corridor>
  mutate(long = ifelse(municipio %in% "zona_corredor_los_cabos", -109.720233, long)) %>% 
  st_as_sf(coords = c("long", "lat"),
           crs = 4326, # WGS84
           remove = FALSE)

mapview(int_geocode_sf)

############################################################
# Now, let us assign each tourist center to an OHI region. 
############################################################

goc_poly <- st_read(here("spatial/ohi_regions","goc_ohi_rgns.shp"))
st_crs(goc_poly) # not WGS84: Mexico ITRF2008 / LCC
mapview(goc_poly)

int_geocode_sf_transform <- st_transform(int_geocode_sf, crs = st_crs(goc_poly)) %>%  # transform the tourist centers to Mexico ITRF2008 / LCC
   mutate(value = as.numeric(value))
crs(int_geocode_sf_transform)

int_goc <- st_intersection(int_geocode_sf_transform, goc_poly) %>% 
  mutate(arrivals_type = "international")
mapview(int_goc) # now each tourist center has associated region data

##############################
#A cool plot to see them each:
##############################

# map of all goc tourist centers
mapview(int_goc, zcol = "rgn_id") + mapview(goc_poly, zcol = "rgn_id")
```

Like stated earlier, the reference point has been updated to use specific target values rather than annual percentage increases (<https://tourismanalytics.com/blog-posts/how-mexicos-forty-percent-tourism-growth-plan-aims-for-fifth-global-spot-by-2030-with-twenty-billion-dollar-investment-in-infrastructure-and-market-expansion>).

For **international tourism**, the 2030 target is defined as **30% higher than 2024 international visitor arrivals**.

Therefore, the current status for each year is calculated as follows:

**International:**
$$
\text{Score} = \frac{\text{International arrivals (year)}}{\text{International arrivals in 2024} \times 1.30} \times 100
$$

A **score of 100** indicates that the number of arrivals has **reached the 2030 target level** (i.e., 9.8% or 30% higher than 2024 levels).

* A score < 100 means arrivals are **below the target trajectory**.
* A score > 100 means arrivals have **exceeded the 2030 target**.

Those two current status scores are then averaged to produce an overall score of how close each region is to making the 2030 target overall. 


```{r}
int_quantity_score <- int_goc %>%
  group_by(municipio) %>% #### I made a mistake in the old version: we want to calculate current status scores per municipio first, then average those for the entire region. FOR NOW, let's just have the scores by municipio.
  mutate(
    year = as.numeric(year),
    
    # determine the baseline value for each municipio (2024)
    int_baseline_arrivals = value[year == baseline_year][1],

    # to know how far from 2024 we are (note: negative numbers are because of years before 2024)
    years_since_baseline = year - baseline_year,

    # target arrivals for international visitors
    int_target_arrivals = int_baseline_arrivals * target_baseline_increase,   # for international tourism, the 2030 target is defined as 9.8% higher than 2024 international visitor arrivals.

    # make it a score
    int_index_score = (value / int_target_arrivals) * 100,
    int_gap_to_target = int_target_arrivals - value,
    int_gap_percentage = ((int_target_arrivals - value) / value) * 100 ## target arrivals - current arrivals / current arrivals... ideal is 0 or negative to reflect how much needs to be done (if you reach the target, it is negative, because nothing needs to be done!)
  ) %>%
  ungroup() %>%
  select(municipio, rgn_id, year, int_current_arrivals = value,
         int_baseline_arrivals, int_target_arrivals, int_index_score,
         int_gap_to_target, int_gap_percentage, everything())
```


```{r}
library(plotly)

## plot the score over time
updated_int_score_plotly <- plotly::plot_ly(
  int_quantity_score, 
  x = ~year, 
  y = ~int_index_score, 
  color = ~municipio, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "CURRENT STATUS: Llegadas de turistas internacionales por municipio",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(int_quantity_score$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
updated_int_score_plotly

htmlwidgets::saveWidget(updated_int_score_plotly, here::here("TR/sectur_quantity_exploration_int","int_arrivals_score_101325.html"), selfcontained = TRUE)
```


```{r}
## compare to the patterns of international arrivals over time
int_arrivals_plot <- plotly::plot_ly(
  int_goc, 
  x = ~year, 
  y = ~value, 
  color = ~municipio, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "Llegadas de turistas internacionales por municipio (international arrivals)",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Llegadas // Arrivals")
  ) %>% 
  add_lines(
    x = range(int_goc$year),
    y = c(100, 100),
    line = list(color = "white", width = 2),
    name = "",
    inherit = FALSE
  )
int_arrivals_plot

htmlwidgets::saveWidget(int_arrivals_plot, here::here("TR/sectur_quantity_exploration_int","int_arrivals_plot_100825.html"), selfcontained = TRUE)
```



# NEXT STEPS: I need to join with goc_poly and assign the values within the regions to the entire region. 
