---
title: 'OHI GoC 2025: QUALITY Tourism & Recreation'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# TR Quality

## Rate of Return

Not possible unfortunately. Lack of data availability.

## Service Establishments

See `TR/denue_prep.qmd` and `TR/servicios_de_alojamiento.qmd`.


Read in all service establishments.

```{r}
library(tidyverse)
library(sf)
library(mapview)
library(here)
```

### Culture service establishments 

```{r}
output_dir <- here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_esparcimiento_culturales_y_deportivos_servicios_recreativos/int")

shp_files <- list.files(output_dir, pattern = "\\.shp$", full.names = TRUE)

# loop through and assign each to an object named after the file
for (s in shp_files) {

  # extract the clean object name
  obj_name <- s %>%
    basename() %>%
    str_remove("\\.shp$")

  message("Reading in: ", obj_name)
  obj_data <- st_read(s, options = "ENCODING=LATIN1", quiet = TRUE)
  assign(obj_name, obj_data, envir = .GlobalEnv)
}

```

```{r}
# # all combined datasets in a named list
# culture_combined_list <- list(
#   museums = museums_2019_2024_goc_50km,
#   historical_sites = historical_sites_2019_2024_goc_50km,
#   nature_heritage = nature_heritage_2019_2024_goc_50km,
#   nautical_services = marine_tourism_2019_2024_goc_50km
#   # recreation_other = recreation_other_2019_2024_goc_50km
# )
# 
# culture_combined_df <- bind_rows(culture_combined_list)
# 
# culture_counts_df <- culture_counts %>% 
#   st_drop_geometry()
# 
# culture_counts_plotly <- plot_ly(culture_counts_df,
#         x = ~data_year,
#         y = ~n,
#         color = ~category,
#         type = 'scatter',
#         mode = 'lines+markers') %>%
#   layout(
#     title = "DENUE observations by year (50km buffer)",
#     xaxis = list(title = "Data year"),
#     yaxis = list(title = "Total observations"),
#     legend = list(title=list(text="Recreation type"))
#   )
# culture_counts_plotly
```


```{r}
culture_gulf_50km <- bind_rows(historical_sites_2019_2024_goc_50km, marine_tourism_2019_2024_goc_50km, museums_2019_2024_goc_50km, nature_heritage_2019_2024_goc_50km, recreation_other_2019_2024_goc_50km)


culture_gulf_50km_clean <- culture_gulf_50km %>% 
  st_drop_geometry() %>% 
  mutate(nombre_act = nombre_act %>% 
           str_replace_all("<fa>", "ú") %>%
           str_replace_all("<e1>", "á") %>%
           str_replace_all("<ed>", "í") %>%
           str_replace_all("<f1>", "ñ") %>%
           str_replace_all("<e9>", "é"),
         
         municipio = municipio %>%
           str_replace_all("<fa>", "ú") %>%
           str_replace_all("<e1>", "á") %>%
           str_replace_all("<ed>", "í") %>%
           str_replace_all("<f1>", "ñ") %>%
           str_replace_all("<e9>", "é")
  ) %>% 
  mutate(municipio_clean = municipio %>% 
           str_to_lower() %>%
           str_replace_all(" ", "_"),
         municipio_clean = case_when(
           municipio == "San Ignacio Río Muerto" ~ "cajeme",
           municipio == "Bácum" ~ "cajeme",
           municipio == "Mexicali" ~ "san_luis_río_colorado",
           municipio == "Sinaloa" ~ "guasave",
           municipio == "Santa María del Oro" ~ "san_pedro_lagunillas",
           TRUE ~ municipio_clean
           )) %>% 
  mutate(
    employees_estimated = case_when(
      per_ocu == "0 a 5 personas" ~ 3,
      per_ocu == "6 a 10 personas" ~ 8,
      per_ocu == "11 a 30 personas" ~ 20.5,
      per_ocu == "31 a 50 personas" ~ 40.5,
      per_ocu == "51 a 100 personas" ~ 75.5,
      per_ocu == "101 a 250 personas" ~ 175.5,
      per_ocu %in% c("251 y m<e1>s personas", "251 y más personas") ~ 300,
      TRUE ~ NA_real_
    )
  ) %>% 
  group_by(municipio_clean, data_year, rgn_id) %>% 
  summarize(agg_employees = sum(employees_estimated)) %>% 
  ungroup()
```

Now, join with tourist arrivals. 

```{r}
arrivals_filled_by_DENUE_accommodation_employees <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/gapfilled_int","arrivals_filled_by_DENUE_accommodation_employees_2019_2024.csv"))

gapfilled_total_arrivals <- arrivals_filled_by_DENUE_accommodation_employees %>%
  mutate(country = "Mexico") %>% 
  dplyr::select(data_year, municipio_clean, arrivals_merged, is_filled) 

joined_arrivals_culture_estab_employees <- left_join(culture_gulf_50km_clean, gapfilled_total_arrivals, by = c("data_year","municipio_clean")) %>% 
  filter(municipio_clean != "jala")
```

```{r}
# aggregate the number of recreation employees and tourist arrivals by region and year
culture_agg_by_rgn_emp_arr <- joined_arrivals_culture_estab_employees %>% 
  dplyr::select(rgn_id, data_year, agg_employees, arrivals_merged) %>% 
  group_by(rgn_id, data_year) %>% 
  summarize(rgn_agg_emp = sum(agg_employees),
            rgn_agg_arr = sum(arrivals_merged)) 

# calculate the number of employees per tourist arrivals by dividing the two
culture_activity_emp_percap_tourist <- culture_agg_by_rgn_emp_arr %>% 
  group_by(rgn_id, data_year) %>% 
  mutate(emp_per_tourist = rgn_agg_emp / rgn_agg_arr) %>% 
  ungroup()

## plot the emp/tourist over time
culture_gulf_emp_per_tourist_plotly <- plotly::plot_ly(
  culture_activity_emp_percap_tourist, 
  x = ~data_year, 
  y = ~emp_per_tourist, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Culture & Gulf-related establishment employee per capita tourist arrival",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "emp_per_tourist")
  ) 
culture_gulf_emp_per_tourist_plotly

```

Now let's actually calculate current status for each region and year. The reference point is the 90th quantile?

current status = emp_per_tourist (current) / emp_per_tourist (90th quantile of all regions and years)

Where every region and year is compared to the highest.

```{r}
## find the 90th quantile
culture_activity_emp_percap_tourist_cs <- culture_activity_emp_percap_tourist %>% 
  mutate(percap_90th = quantile(emp_per_tourist, probs = 0.9),
         current_status = (emp_per_tourist / percap_90th) * 100) %>% 
  mutate(score = ifelse(current_status > 100, 100, current_status))

# write_csv(culture_activity_emp_percap_tourist_cs, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quality","tr_quality_culture_current_status_quantile_121625.csv"))
```


```{r}
library(plotly)

## plot the score over time
culture_quantile_score_plotly <- plotly::plot_ly(
  culture_activity_emp_percap_tourist_cs, 
  x = ~data_year, 
  y = ~score, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: Culture & Gulf-related activity availability -- employees in service establishments per tourist arrival, scaled to the 90th quantile",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(culture_activity_emp_percap_tourist_cs$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
culture_quantile_score_plotly

```
Using regional-specific 2019 reference:

```{r}
hist_culture_activity_emp_percap_tourist_cs <- culture_activity_emp_percap_tourist %>% 
  group_by(rgn_id) %>% 
  mutate(hist_ref = emp_per_tourist[data_year == 2019],
         current_status = (emp_per_tourist / hist_ref) * 100) %>% 
  mutate(score = ifelse(current_status > 100, 100, current_status))

# write_csv(hist_culture_activity_emp_percap_tourist_cs, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quality","tr_quality_culture_current_status_temporal2019_121625.csv"))
```

```{r}

## plot the score over time
culture_hist_score_plotly <- plotly::plot_ly(
  hist_culture_activity_emp_percap_tourist_cs, 
  x = ~data_year, 
  y = ~score, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: Culture & Gulf-related activity availability -- employees in service establishments per tourist arrival, reference of 2019",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(hist_culture_activity_emp_percap_tourist_cs$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
culture_hist_score_plotly

```

### Food and drink establishments

comida_bebidas_categories <- c(
  "Cafeterías, fuentes de sodas, neverías, refresquerías y similares",
  "Restaurantes con servicio de preparación de alimentos a la carta o de comida corrida",
  "Servicios de preparación de otros alimentos para consumo inmediato",
  "Restaurantes con servicio de preparación de antojitos",
  "Restaurantes con servicio de preparación de tacos y tortas",
  "Restaurantes con servicio de preparación de pizzas, hamburguesas, hot dogs y pollos rostizados para llevar",
  "Restaurantes con servicio de preparación de pescados y mariscos",
  "Bares, cantinas y similares",
  "Restaurantes de autoservicio",
  "Restaurantes que preparan otro tipo de alimentos para llevar",
  "Centros nocturnos, discotecas y similares",
  "Servicios de preparación de alimentos en unidades móviles",
  "Servicios de preparación de alimentos para ocasiones especiales",
  "Servicios de comedor para empresas e instituciones"
)

English translation:
  “Cafeterias, soda fountains, ice cream parlors, refreshment stands, and similar establishments,”
  “Restaurants offering à la carte or set menu food service,”
  “Other food preparation services for immediate consumption,”
  “Restaurants offering appetizer preparation services,”
  “Restaurants offering taco and torta preparation services,”
  “Restaurants offering pizza, hamburger, hot dog, and roast chicken preparation services for takeout,”
  “Restaurants offering fish and seafood preparation services,”
  “Bars, cantinas, and similar establishments,”
  “Self-service restaurants,”
  “Restaurants preparing other types of food for takeaway,”
  “Nightclubs, discos, and similar establishments,”
  “Mobile food preparation services,”
  “Food preparation services for special occasions,”
  “Catering services for companies and institutions.”
  
For now, let's include all of the types of food and drink service establishments.

```{r}
comida_bebidas_50km <- st_read(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas/int","update_comida_bebidas_2019_2024_goc_50km.shp"))


comida_bebidas_50km_clean <- comida_bebidas_50km %>% 
  st_drop_geometry() %>% 
  mutate(nombre_act = nombre_act %>% 
           str_replace_all("<fa>", "ú") %>%
           str_replace_all("<e1>", "á") %>%
           str_replace_all("<ed>", "í") %>%
           str_replace_all("<f1>", "ñ") %>%
           str_replace_all("<e9>", "é"),
         
         municipio = municipio %>%
           str_replace_all("<fa>", "ú") %>%
           str_replace_all("<e1>", "á") %>%
           str_replace_all("<ed>", "í") %>%
           str_replace_all("<f1>", "ñ") %>%
           str_replace_all("<e9>", "é")
  ) %>% 
  mutate(municipio_clean = municipio %>% 
           str_to_lower() %>%
           str_replace_all(" ", "_"),
         municipio_clean = case_when(
           municipio == "San Ignacio Río Muerto" ~ "cajeme",
           municipio == "Bácum" ~ "cajeme",
           municipio == "Mexicali" ~ "san_luis_río_colorado",
           municipio == "Sinaloa" ~ "guasave",
           municipio == "Santa María del Oro" ~ "san_pedro_lagunillas",
           TRUE ~ municipio_clean
           )) %>% 
  mutate(
    employees_estimated = case_when(
      per_ocu == "0 a 5 personas" ~ 3,
      per_ocu == "6 a 10 personas" ~ 8,
      per_ocu == "11 a 30 personas" ~ 20.5,
      per_ocu == "31 a 50 personas" ~ 40.5,
      per_ocu == "51 a 100 personas" ~ 75.5,
      per_ocu == "101 a 250 personas" ~ 175.5,
      per_ocu %in% c("251 y m<e1>s personas", "251 y más personas") ~ 300,
      TRUE ~ NA_real_
    )
  ) %>% 
  group_by(municipio_clean, data_year, rgn_id) %>% 
  summarize(agg_employees = sum(employees_estimated)) %>% 
  ungroup()
```



```{r}
# comida_bebidas_50km_df <- all_comida_bebidas_50km %>% 
#   st_drop_geometry() %>% 
#   group_by(data_year, municipio) %>% 
#   summarize(total_obs = n()) %>% 
#   ungroup()
# 
# food_drink_plotly <- plot_ly(comida_bebidas_50km_df,
#         x = ~data_year,
#         y = ~total_obs,
#         color = ~municipio,
#         type = 'scatter',
#         mode = 'lines+markers') %>%
#   layout(
#     title = "DENUE comida & bebida observations by year (50km buffer)",
#     xaxis = list(title = "Data year"),
#     yaxis = list(title = "Total observations")
#   )
# food_drink_plotly
```

Now, join with tourist arrivals. 

```{r}
arrivals_filled_by_DENUE_accommodation_employees <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/gapfilled_int","arrivals_filled_by_DENUE_accommodation_employees_2019_2024.csv"))

gapfilled_total_arrivals <- arrivals_filled_by_DENUE_accommodation_employees %>%
  mutate(country = "Mexico") %>% 
  select(data_year, municipio_clean, arrivals_merged, is_filled) 

joined_arrivals_food_bev_estab_employees <- left_join(comida_bebidas_50km_clean, gapfilled_total_arrivals, by = c("data_year","municipio_clean"))
```

```{r}
# aggregate food and drink employees and tourist arrivals by region and year
food_bev_agg_by_rgn_emp_arr <- joined_arrivals_food_bev_estab_employees %>% 
  dplyr::select(rgn_id, data_year, agg_employees, arrivals_merged) %>% 
  group_by(rgn_id, data_year) %>% 
  summarize(rgn_agg_emp = sum(agg_employees),
            rgn_agg_arr = sum(arrivals_merged)) 

# divide the number of employees by the number of tourist arrivals
food_bev_activity_emp_percap_tourist <- food_bev_agg_by_rgn_emp_arr %>% 
  group_by(rgn_id, data_year) %>% 
  mutate(emp_per_tourist = rgn_agg_emp / rgn_agg_arr) %>% 
  ungroup()

## plot the emp/tourist over time
food_bev_emp_per_tourist_plotly <- plotly::plot_ly(
  food_bev_activity_emp_percap_tourist, 
  x = ~data_year, 
  y = ~emp_per_tourist, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Food & beverage establishment employee per capita tourist arrival",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "emp_per_tourist")
  ) 
food_bev_emp_per_tourist_plotly

```

ow let's actually caluclate current status for each region and year. The reference point is the 90th quantile?

current status = emp_per_tourist (current) / emp_per_tourist (90th quantile of all regions and years)

Where every region and year is compared to the highest 

```{r}
## find the 90th quantile
food_bev_activity_emp_percap_tourist_cs <- food_bev_activity_emp_percap_tourist %>% 
  mutate(percap_90th = quantile(emp_per_tourist, probs = 0.9),
         current_status = (emp_per_tourist / percap_90th) * 100) %>% 
  mutate(score = ifelse(current_status > 100, 100, current_status))

# write_csv(food_bev_activity_emp_percap_tourist_cs, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quality","tr_quality_food_bev_current_status_quantile_121125.csv"))
```


```{r}
library(plotly)

## plot the score over time
food_bev_quantile_score_plotly <- plotly::plot_ly(
  food_bev_activity_emp_percap_tourist_cs, 
  x = ~data_year, 
  y = ~score, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: Food & beverage activity availability -- employees in service establishments per tourist arrival, reference to the 90th quantile",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(food_bev_activity_emp_percap_tourist_cs$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
food_bev_quantile_score_plotly

```
Using regional-specific 2019 reference:

```{r}
hist_food_bev_activity_emp_percap_tourist_cs <- food_bev_activity_emp_percap_tourist %>% 
  group_by(rgn_id) %>% 
  mutate(hist_ref = emp_per_tourist[data_year == 2019],
         current_status = (emp_per_tourist / hist_ref) * 100) %>% 
  mutate(score = ifelse(current_status > 100, 100, current_status))

# write_csv(hist_food_bev_activity_emp_percap_tourist_cs, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quality","tr_quality_food_bev_current_status_temporal2019_121625.csv"))
```

```{r}

## plot the score over time
food_bev_hist_score_plotly <- plotly::plot_ly(
  hist_food_bev_activity_emp_percap_tourist_cs, 
  x = ~data_year, 
  y = ~score, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: Food & beverage activity availability -- employees in service establishments per tourist arrival, reference of 2019",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(hist_food_bev_activity_emp_percap_tourist_cs$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
food_bev_hist_score_plotly

```




### Other recreation service establishments!

recreacion_categorias <- c(
  "Centros de acondicionamiento físico del sector privado",
  "Centros de acondicionamiento físico del sector público",
  "Clubes deportivos del sector privado",
  "Clubes deportivos del sector público",
  "Equipos deportivos profesionales",
  "Deportistas profesionales",
  "Clubes o ligas de aficionados",
  "Campos de golf",
  "Artistas, escritores y técnicos independientes",
  "Compañías de danza del sector privado",
  "Compañías de teatro del sector público",
  "Compañías de teatro del sector privado",
  "Grupos musicales del sector público",
  "Cantantes y grupos musicales del sector privado",
  "Otras compañías y grupos de espectáculos artísticos del sector privado",
  "Parques de diversiones y temáticos del sector privado",
  "Parques de diversiones y temáticos del sector público",
  "Parques acuáticos y balnearios del sector privado",
  "Parques acuáticos y balnearios del sector público",
  "Billares",
  "Casas de juegos electrónicos",
  "Boliches",
  "Otros servicios recreativos prestados por el sector privado",
  "Otros servicios recreativos prestados por el sector público",
  "Venta de billetes de lotería, pronósticos deportivos y otros boletos de sorteo",
  "Otros juegos de azar",
  "Casinos"
)


```{r}
other_rec_50km <- st_read(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_esparcimiento_culturales_y_deportivos_servicios_recreativos/int/recreation_other_2019_2024_goc_50km.shp"))

other_rec_50km_clean <- other_rec_50km %>% 
  st_drop_geometry() %>% 
  mutate(municipio_clean = municipio %>% 
           str_to_lower() %>%
           str_replace_all(" ", "_"),
         municipio_clean = case_when(
           municipio == "San Ignacio Río Muerto" ~ "cajeme",
           municipio == "Bácum" ~ "cajeme",
           municipio == "Mexicali" ~ "san_luis_río_colorado",
           municipio == "Sinaloa" ~ "guasave",
           municipio == "Santa María del Oro" ~ "san_pedro_lagunillas",
           TRUE ~ municipio_clean
           )) %>% 
  mutate(
    employees_estimated = case_when(
      per_ocu == "0 a 5 personas" ~ 3,
      per_ocu == "6 a 10 personas" ~ 8,
      per_ocu == "11 a 30 personas" ~ 20.5,
      per_ocu == "31 a 50 personas" ~ 40.5,
      per_ocu == "51 a 100 personas" ~ 75.5,
      per_ocu == "101 a 250 personas" ~ 175.5,
      per_ocu %in% c("251 y m<e1>s personas", "251 y más personas") ~ 300,
      TRUE ~ NA_real_
    )
  ) %>% 
  group_by(municipio_clean, data_year, rgn_id) %>% 
  summarize(agg_employees = sum(employees_estimated)) %>% 
  ungroup()
```

Now, join with tourist arrivals. 

```{r}
arrivals_filled_by_DENUE_accommodation_employees <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/gapfilled_int","arrivals_filled_by_DENUE_accommodation_employees_2019_2024.csv"))

gapfilled_total_arrivals <- arrivals_filled_by_DENUE_accommodation_employees %>%
  mutate(country = "Mexico") %>% 
  dplyr::select(data_year, municipio_clean, arrivals_merged, is_filled) 

joined_arrivals_other_rec_estab_employees <- left_join(other_rec_50km_clean, gapfilled_total_arrivals, by = c("data_year","municipio_clean")) %>% 
  filter(municipio_clean != "jala")
```

```{r}
# aggregate the number of other rec employees and tourist arrivals to each region and year
other_rec_agg_by_rgn_emp_arr <- joined_arrivals_other_rec_estab_employees %>% 
  dplyr::select(rgn_id, data_year, agg_employees, arrivals_merged) %>% 
  group_by(rgn_id, data_year) %>% 
  summarize(rgn_agg_emp = sum(agg_employees),
            rgn_agg_arr = sum(arrivals_merged)) 

# calculate employees per tourist arrival 
other_rec_activity_emp_percap_tourist <- other_rec_agg_by_rgn_emp_arr %>% 
  group_by(rgn_id, data_year) %>% 
  mutate(emp_per_tourist = rgn_agg_emp / rgn_agg_arr) %>% 
  ungroup()

## plot the emp/tourist over time
other_rec_gulf_emp_per_tourist_plotly <- plotly::plot_ly(
  other_rec_activity_emp_percap_tourist, 
  x = ~data_year, 
  y = ~emp_per_tourist, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Other recreational establishment employee per capita tourist arrival",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "emp_per_tourist")
  ) 
other_rec_gulf_emp_per_tourist_plotly

```

Now let's actually calculate current status for each region and year. The reference point is the 90th quantile?

current status = emp_per_tourist (current) / emp_per_tourist (90th quantile of all regions and years)

Where every region and year is compared to the highest.

```{r}
## find the 90th quantile
other_rec_activity_emp_percap_tourist_cs <- other_rec_activity_emp_percap_tourist %>% 
  mutate(percap_90th = quantile(emp_per_tourist, probs = 0.9),
         current_status = (emp_per_tourist / percap_90th) * 100) %>% 
  mutate(score = ifelse(current_status > 100, 100, current_status))

# write_csv(other_rec_activity_emp_percap_tourist_cs, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quality","tr_quality_other_rec_current_status_quantile_121225.csv"))
```


```{r}
library(plotly)

## plot the score over time
other_rec_quantile_score_plotly <- plotly::plot_ly(
  other_rec_activity_emp_percap_tourist_cs, 
  x = ~data_year, 
  y = ~score, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: Other recreation activity availability -- employees in service establishments per tourist arrival, reference as the 90th quantile",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(other_rec_activity_emp_percap_tourist_cs$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
other_rec_quantile_score_plotly

```

Using regional-specific 2019 reference:

```{r}
hist_other_rec_activity_emp_percap_tourist_cs <- other_rec_activity_emp_percap_tourist %>% 
  group_by(rgn_id) %>% 
  mutate(hist_ref = emp_per_tourist[data_year == 2019],
         current_status = (emp_per_tourist / hist_ref) * 100) %>% 
  mutate(score = ifelse(current_status > 100, 100, current_status))

# write_csv(hist_other_rec_activity_emp_percap_tourist_cs, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quality","tr_quality_other_rec_current_status_temporal2019_121625.csv"))
```

```{r}

## plot the score over time
other_rec_hist_score_plotly <- plotly::plot_ly(
  hist_other_rec_activity_emp_percap_tourist_cs, 
  x = ~data_year, 
  y = ~score, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: Other recreation activity availability -- employees in service establishments per tourist arrival, reference of 2019",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(hist_other_rec_activity_emp_percap_tourist_cs$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
other_rec_hist_score_plotly

```

### Merging all activities

Quality = avg(crime/coastal pop + activities)
activities = avg(2/5 * food&drink emp / tourist arrivals + 2/5 * cultural&gulf-related / tourist arrivals + 1/5 * other_recreational / tourist arrivals)

```{r}
food_bev_current_status <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quality","tr_quality_food_bev_current_status_quantile_121125.csv")) %>% 
  mutate(activity_type = "food_bev") %>% 
  mutate(score = ifelse(rgn_id %in% c("Region_2","Region_3","Region_6"), NA, score))

# food_bev_current_status <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quality","tr_quality_food_bev_current_status_quantile_121125.csv")) %>% 
#   mutate(activity_type = "food_bev")

## plot the score over time
food_bev_current_status_plotly <- plotly::plot_ly(
  food_bev_current_status, 
  x = ~data_year, 
  y = ~score, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "food&bev, reference of 2019",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(food_bev_current_status$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
food_bev_current_status_plotly
```


```{r}
culture_current_status <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quality","tr_quality_culture_current_status_quantile_121625.csv")) %>% 
  mutate(activity_type = "culture")%>% 
  mutate(score = ifelse(rgn_id %in% c("Region_2","Region_3","Region_6"), NA, score))

# culture_current_status <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quality","tr_quality_culture_current_status_quantile_121625.csv")) %>% 
#   mutate(activity_type = "culture")

## plot the score over time
culture_current_status_plotly <- plotly::plot_ly(
  culture_current_status, 
  x = ~data_year, 
  y = ~score, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "culture&gulf, reference of 2019",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(culture_current_status$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
culture_current_status_plotly
```


```{r}
other_rec_current_status <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quality","tr_quality_other_rec_current_status_quantile_121225.csv")) %>% 
  mutate(activity_type = "other_rec") %>% 
  mutate(score = ifelse(rgn_id %in% c("Region_2","Region_3","Region_6"), NA, score))

# other_rec_current_status <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quality","tr_quality_other_rec_current_status_quantile_121225.csv")) %>% 
#   mutate(activity_type = "other_rec")

## plot the score over time
other_rec_current_status_plotly <- plotly::plot_ly(
  other_rec_current_status, 
  x = ~data_year, 
  y = ~score, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "food&bev, reference of 2019",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(other_rec_current_status$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
other_rec_current_status_plotly
```

```{r}
merge_activites <- bind_rows(food_bev_current_status, culture_current_status, other_rec_current_status) %>% 
  dplyr::select(rgn_id, year = data_year, score, activity_type) %>% 
  pivot_wider(
    names_from = activity_type,
    values_from = score
  )

score_merge_activites <- merge_activites %>%
  mutate(
    final_score =
      (2/5) * food_bev +
      (2/5) * culture +
      (1/5) * other_rec
  )

write_csv(score_merge_activites, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_quality","final_activity_avg_score_030526.csv"))
```


##### activities barplot viz
```{r}
# viz_activities_2024_plot <- score_merge_activites %>%
#   separate(col = rgn_id, into = c("rgn", "rgn_id_num"), sep = "_", remove = FALSE) %>% 
#   filter(year == 2024) %>% 
#   dplyr::select(rgn_id, rgn_id_num, final_score) %>% 
#   arrange(final_score) %>% 
#   mutate(
#     rgn_id_num = factor(rgn_id_num, levels = rgn_id_num)
#     ) 
# 
# 
# rgn_colors <- c(
#   "Region_1" = "#C03729",
#   "Region_2" = "#E68C7C",
#   "Region_3" = "#FC8F24",
#   "Region_4" = "#dcaa38",
#   "Region_5" = "#cbd2a0",
#   "Region_6" = "#98a25a",
#   "Region_7" = "#747428",
#   "Region_8" = "#627391",
#   "Region_9" = "#8d61a9"
# )
# 
# a <- ggplot(
#   viz_activities_2024_plot,
#   aes(x = rgn_id_num, y = final_score, fill = rgn_id)
# ) +
#   geom_col(color = "black", linewidth = 0.4) +
#   scale_fill_manual(values = rgn_colors) +
#   labs(
#     title = "Current activity availability compared to pre-COVID levels",
#     x = " ",
#     y = "Activity availability relative to 2019 (%)"
#   ) +
#   theme_minimal(base_family = "Calibri") +
#   theme(
#     plot.title = element_text(face = "bold", size = 18),
#     axis.title.y = element_text(face = "bold", size = 16),
#     axis.text.x = element_text(face = "bold", size = 16),
#     axis.text.y = element_text(face = "bold", size = 16),
#     axis.title.x = element_blank(),
#     legend.position = "none"
#   ) +
#   geom_hline(
#   yintercept = 100,
#   linetype = "dashed",
#   linewidth = 0.8,
#   color = "black"
# )
# 
# a
# 
# ggsave(
#   filename = here::here(
#     "/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_viz",
#     "activity_availability_by_region_2024.png"
#   ),
#   plot = a,
#   width = 7,
#   height = 4,
#   units = "in",
#   dpi = 300
# )

```

Instead, a map for viz.  

```{r}
# #### join with ohi rgn polygons
# ohi <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp")) %>%
#   filter(location == "marine") %>% 
#   select(rgn_id)
# 
# viz_activities_2024_ohigns <- full_join(viz_activities_2024_plot, ohi, by = "rgn_id") %>% 
#   st_as_sf()
# 
# mapview(
#   viz_activities_2024_ohigns, 
#   zcol = "final_score",
#   col.regions = colorRampPalette(c( '#FEE090', '#ABD9E9', '#4575B4', '#313695'
# )),
#   layer.name = "Current activity availability (2024) compared to pre-COVID levels (2019)",
#   legend = TRUE
# )
# 
# ### or, using rgn colors, and adding the numbers on the ppt?
# ohi <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp")) %>%
#   filter(location == "marine")
# 
# rgn_colors <- c(
#   "Region_1" = "#C03729",
#   "Region_2" = "#E68C7C",
#   "Region_3" = "#FC8F24",
#   "Region_4" = "#dcaa38",
#   "Region_5" = "#cbd2a0",
#   "Region_6" = "#98a25a",
#   "Region_7" = "#747428",
#   "Region_8" = "#627391",
#   "Region_9" = "#8d61a9"
# )
# 
# ohi <- ohi %>%
#   mutate(rgn_id = factor(rgn_id, levels = names(rgn_colors)))
# 
# mapview::mapview(
#   ohi,
#   zcol = "rgn_id",
#   col.regions = rgn_colors,
#   legend = TRUE
# )
```



```{r}
## plot the score over time
score_merge_activites_plotly <- plotly::plot_ly(
  score_merge_activites, 
  x = ~year, 
  y = ~final_score, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Merged activities, reference of 90th quantile",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(score_merge_activites$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
score_merge_activites_plotly

```

## Public bathroom availability

Read in the data:

```{r}
library(tidyverse)
library(sf)
library(mapview)

banos_publicos_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_public_bathrooms","banos_publicos_812130_stablecimientoseconomicos.csv"))

banos_publicos_sf <- banos_publicos_raw %>% 
  st_as_sf(coords = c("Longitud", "Latitud"), # making sf object
           crs = 4326, # WGS84
           remove = FALSE) %>% 
  janitor::clean_names()


mapview(banos_publicos_sf)
```

```{r}
inland_50 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer","inland_50km_goc.shp"))

banos_publicos_sf_trans <- st_transform(banos_publicos_sf, st_crs(inland_50))
banos_publicos_goc <- st_intersection(banos_publicos_sf_trans, inland_50)
mapview(banos_publicos_goc, zcol = "descripcion_del_sector_economico")
```



## Crime

#### Read in SESNSP GoC data

```{r}
library(tidyverse)
library(here)
library(sf)
library(mapview)
```


```{r}
# bring in SESNSP data on crime incidence, cleaned in `~/ohi_goc_prep/SS/v2025/safety_sesnsp.qmd`
sesnsp_goc_states <- st_read(here::here("/home/shares/ohi/OHI_GOC/goal_prep/ss/int", "sesnsp_anual_goc_states_sf_w2025.shp")) ## need to fix the names / make a document describing the abbreviations!!

unique(sesnsp_goc_states$rgn_id) # for 50km: [1] "Region_1" "Region_3" "Region_4" "Region_5" "Region_7" "Region_8" "Region_9"

mapview(sesnsp_goc_states)
```

Limit to the municipalities that are coastal GoC:
```{r}
## bring in the municipality to rgn_id code CSV I made...isolate_goc_municipios.qmd led to the manually made .CSV: `~/home/shares/ohi/OHI_GOC/spatial/municipality_to_rgnid/municipality_to_rgnid.csv`

municipality_to_rgnid <- read_csv(here::here("/home/shares/ohi/OHI_GOC/spatial/municipality_to_rgnid","municipality_to_rgnid.csv")) %>% 
  dplyr::select(-c("notes","municipio_raw")) %>% 
  janitor::clean_names()

setdiff(unique(municipality_to_rgnid$municipio), unique(sesnsp_goc_states$municip)) 
#  [1] "Mulege"                        "San Luis Rio Colorado"        
#  [3] "General Plutarco Elias Calles" "San Ignacio Rio Muerto"       
#  [5] "Bacum"                         "Benito Juarez"                
#  [7] "Culiacan"                      "Mazatlan"                     
#  [9] "Cosala"                        "Bahia De Banderas"            
# [11] "Ruiz"                          "Santa Maria Del Oro"          
# [13] "Ahuacatlan"                    "Amatlan De Cañas"             
# [15] "Ixtlan Del Rio"   

municipality_to_rgnid <- read_csv(here::here("/home/shares/ohi/OHI_GOC/spatial/municipality_to_rgnid","municipality_to_rgnid.csv")) %>% 
  dplyr::select(-c("notes","municipio_raw")) %>% 
  janitor::clean_names() %>%
  
  # changes so that we can filter sesnsp_goc_states by the municipalities bordering the Gulf:
  mutate(municipio = str_trim(municipio, side = "both"),
         municipio = str_to_title(municipio),
         estado = str_to_title(estado)) %>% 
  mutate(municipio = case_when(
    municipio %in% "Mulege" ~ "Mulegé",
    municipio %in% "San Luis Rio Colorado" ~ "San Luis Río Colorado",
    municipio %in% "San Ignacio Rio Muerto" ~ "San Ignacio Río Muerto",
    municipio %in% "Benito Juarez" ~ "Benito Juárez",
    municipio %in% "Culiacan" ~ "Culiacán",
    municipio %in% "Mazatlan" ~ "Mazatlán",
    municipio %in% "Cosala" ~ "Cosalá",
    municipio %in% "Bahia De Banderas" ~ "Bahía de Banderas",
    municipio %in% "Ruiz" ~ "Ruíz",
    municipio %in% "Santa Maria Del Oro" ~ "Santa María del Oro",
    municipio %in% "Ahuacatlan" ~ "Ahuacatlán",
    municipio %in% "Amatlan De Cañas" ~ "Amatlán de Cañas",
    municipio %in% "Ixtlan Del Rio" ~ "Ixtlán del Río",
    TRUE ~ municipio
  ))
  
setdiff(unique(municipality_to_rgnid$municipio), unique(sesnsp_goc_states$municip)) # NA

## join the two
sesnsp_goc_municipios <- inner_join(sesnsp_goc_states, municipality_to_rgnid, by = c("entidad" = "estado","municip" = "municipio")) %>% 
  filter(!municip %in% "Ensenada") %>% 
  filter(!municip %in% "Mexicali")
```


#### Updated TR crimes

Conceptual rule: “included if it plausibly alters safety, cleanliness, aesthetics, or atmosphere in spaces tourists visit, which in turn affects visit quality."

```{r}
# vector of subtypes I want to keep for TR quality
updated_tr_qual_crime_subtypes <- c(
  "Daño a la propiedad",
  "Extorsión",
  "Fraude",
  "Otros delitos contra el patrimonio",
  "Otros robos",
  "Robo a casa habitación",
  "Robo a institución bancaria",
  "Robo a negocio",
  "Robo a transeúnte en espacio abierto al público",
  "Robo a transeúnte en vía pública",
  "Robo a transportista",
  "Robo de autopartes",
  "Robo de vehículo automotor",
  "Robo en transporte individual",
  "Robo en transporte público colectivo",
  "Robo en transporte público individual",
  "Violencia de género en todas sus modalidades distinta a la violencia familiar",
  "Abuso sexual",
  "Acoso sexual",
  "Hostigamiento sexual",
  "Otros delitos que atentan contra la libertad y la seguridad sexual",
  "Violación equiparada",
  "Violación simple",
  "Corrupción de menores",
  "Otros delitos contra la sociedad",
  "Trata de personas",
  "Feminicidio",
  "Homicidio doloso",
  "Lesiones dolosas",
  "Otros delitos que atentan contra la vida y la integridad corporal",
  "Otros delitos que atentan contra la libertad personal",
  "Rapto",
  "Secuestro",
  "Tráfico de menores",
  "Amenazas",
  "Contra el medio ambiente",
  "Narcomenudeo"
)

#### those that were dropped:
# c(
#   "Delitos cometidos por servidores públicos",
#   "Evasión de presos",
#   "Otros delitos del Fuero Común",
#   "Robo de maquinaria"
# )



# filter sesnsp_goc by those subtypes
sesnsp_goc_tr_qual <- sesnsp_goc_municipios %>%
  filter(sbtp_d_ %in% updated_tr_qual_crime_subtypes) %>% 
  filter(!municip %in% "Mexicali")

unique(sesnsp_goc_tr_qual$rgn_id) # [1] "Region_1" "Region_3" "Region_4" "Region_5" "Region_7" "Region_8" "Region_9"
```

Three categories of crimes:
```{r}
# high personal danger, there is a possibility that the tourist is the direct victim of violence

crimes_against_person <- c( 
  
  # sexual crimes
  "Abuso sexual",
  "Acoso sexual",
  "Hostigamiento sexual",
  "Otros delitos que atentan contra la libertad y la seguridad sexual",
  "Violación equiparada",
  "Violación simple",
  
  # violence and bodily harm
  "Feminicidio",
  "Homicidio doloso",
  "Lesiones dolosas",
  "Otros delitos que atentan contra la vida y la integridad corporal",
  
  # deprivation of liberty
  "Otros delitos que atentan contra la libertad personal",
  "Rapto",
  "Secuestro",
  "Tráfico de menores",
  
  # coercion and threats
  "Amenazas",
  "Extorsión",
  "Violencia de género en todas sus modalidades distinta a la violencia familiar",
  
  # robbery with personal exposure
  "Robo a transeúnte en espacio abierto al público",
  "Robo a transeúnte en vía pública",
  "Robo en transporte individual",
  "Robo en transporte público colectivo",
  "Robo en transporte público individual",
  
  # may not impact tourist directly, but really could, and could be directly linked to other crimes that directly harm the tourist
  "Corrupción de menores",
  "Otros delitos contra la sociedad",
  "Trata de personas",
  "Narcomenudeo"
)

# tourist might lose money or belongings, but not directly physically threatened
crimes_against_property <- c( 
  "Daño a la propiedad",
  "Fraude",
  "Otros delitos contra el patrimonio",
  "Otros robos",
  "Robo a casa habitación",
  "Robo a institución bancaria",
  "Robo a negocio",
  "Robo a transportista",
  "Robo de autopartes",
  "Robo de vehículo automotor",
  "Contra el medio ambiente"
)
```



#### Summarize the number of crimes by year and region (assuming that all crimes are weighted equally)

No weighting:
```{r}
# summarize the number of crimes by each region, not weighing any crime more than another, because they all affect tourist quality.  
sesnsp_goc_tr_qual_sum <- sesnsp_goc_tr_qual %>% 
  st_drop_geometry() %>% 
  filter(!is.na(rgn_id)) %>% 
  group_by(ano, rgn_id) %>% 
  summarise(number_crimes = sum(annl_tt)) %>% 
  ungroup()

unique(sesnsp_goc_tr_qual_sum$rgn_id) # [1] "Region_3" "Region_4" "Region_5" "Region_7" "Region_8" "Region_9" "Region_1"
```

With weighting:
```{r}
# assign weights by crime category
crime_weights <- tibble(
  sbtp_d_ = c(crimes_against_person, crimes_against_property),
  weight = c(rep(2, length(crimes_against_person)), 
             rep(1, length(crimes_against_property)))
)

# summarize weighted crimes by year and region
sesnsp_goc_tr_qual_sum_weighted <- sesnsp_goc_tr_qual %>% 
  st_drop_geometry() %>% 
  filter(!is.na(rgn_id)) %>% 
  left_join(crime_weights, by = "sbtp_d_") %>% 
  group_by(ano, rgn_id) %>% 
  summarise(number_crimes = sum(annl_tt * weight)) %>% 
  ungroup()

unique(sesnsp_goc_tr_qual_sum$rgn_id)
```


```{r}
## for plotting:
rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

region_levels <- paste0("Region_", 1:9)

#### plotly of crimes over time
sesnsp_goc_tr_qual_sum_factor <- sesnsp_goc_tr_qual_sum %>% 
  mutate(
    ano = as.numeric(ano),
    rgn_id = factor(rgn_id, levels = region_levels)
  ) %>%
  arrange(rgn_id, ano)

crime_raw_plotly <- plotly::plot_ly(
  sesnsp_goc_tr_qual_sum_weighted, 
  x = ~ano, 
  y = ~number_crimes, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Raw number of crimes that would affect a tourist's visit quality",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Total number of crimes")
  )
crime_raw_plotly
```

### Divide by coastal population in each region

##### Bring in population

```{r}
pop_goc <- read_csv(here::here("/home/shares/ohi/OHI_GOC/spatial/population_municipal","population_goc_municipality_inegi_2000_2020.csv"))

setdiff(unique(municipality_to_rgnid$municipio), unique(pop_goc$desc_municipio)) 
# [1] "General Plutarco Elias Calles" "Bacum" "San Felipe"

# add in San Felipe as `desc_municipio`, and the `ano_censal` and `pop_count` from this document: 17,143 https://www.inegi.org.mx/contenidos/productos/prod_serv/contenidos/espanol/bvinegi/productos/nueva_estruc/702825197735.pdf 
# San Felipe	Mexicali	13,123	14,845	16,721	17,143
# Source: Instituto Nacional de Estadística Geografía e Informática, México.
san_felipe_rows <- data.frame(
  cve_entidad = rep("02", 4),
  desc_entidad = rep("Baja California", 4),
  cve_municipio = rep("7", 4),
  desc_municipio = rep("San Felipe", 4),
  indicador = rep("From INEGI aggregation of localities", 4),
  unidad_medida = rep("People", 4),
  rgn_id = rep("Region_1", 4),
  directly_coastal = rep("Y", 4),
  ano_censal = c(2000, 2005, 2010, 2020),
  pop_count = c(13123, 14845, 16721, 17143)
)

# https://www.economia.gob.mx/datamexico/en/profile/geo/bacum?redirect=true
bacum_rows <- data.frame(
  cve_entidad = rep("26", 2),
  desc_entidad = rep("Sonora", 2),
  cve_municipio = rep("12", 2),
  desc_municipio = rep("Bacum", 2),
  indicador = rep("From INEGI aggregation of localities", 2),
  unidad_medida = rep("People", 2),
  rgn_id = rep("Region_7", 2),
  directly_coastal = rep("Y", 2),
  ano_censal = c(2010, 2020),
  pop_count = c(22820, 23151)
)

pop_goc <- rbind(pop_goc, san_felipe_rows, bacum_rows)

setdiff(unique(municipality_to_rgnid$municipio), unique(pop_goc$desc_municipio))  # leave "General Plutarco Elias Calles" because it is not coastal GoC.

pop_goc_municipalities <- inner_join(pop_goc, municipality_to_rgnid, by = c("desc_entidad" = "estado","desc_municipio" = "municipio", "rgn_id")) %>% 
  filter(!is.na(rgn_id)) %>% 
  filter(!desc_municipio %in% "Mexicali") # not representative of GoC, rather border.
```

```{r}
## aggregate by region of all GoC-bordering municipalities. NOT yet adjusted for only the coastal parts of Region_2 and Region_3
agg_pop_rgnid <- pop_goc_municipalities %>% 
  group_by(rgn_id, ano_censal, unidad_medida) %>% 
  summarize(total_rgn_pop = sum(pop_count, na.rm = TRUE)) %>% 
  ungroup()
```


### Interpolate municipal population for all years

```{r}
library(zoo)

## in order for it to join with the job data, we need to interpolate and determine the population in years between.
year_adj_pop_rgnid <- agg_pop_rgnid %>% 
  mutate(year_interpolated = as.numeric(ano_censal)) %>%
  group_by(rgn_id) %>% 
  complete(year_interpolated = 2000:2025) # expand to interpolate

year_adj_pop_rgnid_interpolated <- year_adj_pop_rgnid %>%
  arrange(rgn_id, year_interpolated) %>%
  group_by(rgn_id) %>% 
  mutate(
    population_interpolated = case_when(
      # Region_2: NA for all years before 2020 (San Quintin created in 2020)
      rgn_id == "Region_2" & year_interpolated < 2020 ~ NA_real_,
      # Region_2: hold 2020 value forward through 2025
      rgn_id == "Region_2" & year_interpolated >= 2020 ~ 117568.0,
      # All other regions: interpolate as before
      TRUE ~ na.approx(
        total_rgn_pop,
        x = year_interpolated,
        rule = 2,
        na.rm = FALSE
      )
    ),
    is_gapfilled = ifelse(is.na(ano_censal), 1, 0)
  ) %>%
  fill(ano_censal, .direction = "downup") %>% 
  rename(pop_ano_censal_original = ano_censal) %>% # rename so that when it is joined, we know which two censuses are being interpolated from.
  fill(total_rgn_pop, .direction = "downup") %>% # so we can see what the original population value was
  fill(unidad_medida, .direction = "downup") %>%
  dplyr::select(rgn_id, pop_ano_censal_original, population_original = total_rgn_pop, year_interpolated, population_interpolated, unidad_medida, is_gapfilled) %>% 
  ungroup()

## note that the interpolated population amount from 2020-2025 is the same for each region.  It does not project the growth before it happens.

## save this
# write_csv(year_adj_pop_rgnid_interpolated, here::here("/home/shares/ohi/OHI_GOC/spatial/population_municipal","interpolated_population_per_rgn_coastalmunicipios_2000_2025.csv"))
```

```{r}
sesnsp_goc_tr_sum_percapita <- left_join(sesnsp_goc_tr_qual_sum_weighted, year_adj_pop_rgnid_interpolated, by = c("rgn_id", "ano" = "year_interpolated")) %>% 
  group_by(ano, rgn_id) %>% 
  mutate(tr_crimes_percapita = (number_crimes / population_interpolated)) %>% 
  ungroup()
```


```{r}
crime_percap_plotly <- plotly::plot_ly(
  sesnsp_goc_tr_sum_percapita, 
  x = ~ano, 
  y = ~tr_crimes_percapita, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Per capita local inhabitant: crimes that would affect a tourist's visit quality",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Crimes per capita")
  )
  # add_lines(
  #   x = range(sesnsp_goc_tr_qual_sum_factor$ano),
  #   y = c(100, 100),
  #   line = list(color = "black", width = 2, dash = "dash"),
  #   name = "Meta // Goal = 0 crimes",
  #   inherit = FALSE
  # )
crime_percap_plotly
```

# Stopped here

Make a visualization for 2024:
```{r}
region_2_6_tibble <- tibble::tribble(
  ~ano, ~tr_crimes_percapita, ~rgn_id,
  2024, 0.0002, "Region_2",
  2024, 0.0002, "Region_6"
)

viz_crime_rates_2024 <- sesnsp_goc_tr_crimes_percapita %>% 
  filter(ano == "2024") %>% 
  bind_rows(region_2_6_tibble)

library(ggplot2)
library(dplyr)

rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

viz_crime_rates_2024_plot <- viz_crime_rates_2024 %>%
  separate(col = rgn_id, into = c("rgn", "rgn_id_num"), sep = "_", remove = FALSE) %>% 
  arrange(tr_crimes_percapita) %>% 
  mutate(
    crimes_per_1000 = tr_crimes_percapita * 1000,
    rgn_id_num = factor(rgn_id_num, levels = rgn_id_num)
    ) 

#### join with ohi rgn polygons
ohi <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp")) %>%
  filter(location == "marine") %>% 
  select(rgn_id)

viz_crime_rates_2024_ohigns <- full_join(viz_crime_rates_2024_plot, ohi, by = "rgn_id") %>% 
  st_as_sf()

mapview(viz_crime_rates_2024_ohigns, 
        zcol = "crimes_per_1000",
        col.regions = colorRampPalette(c("#FFD599", "#FFAC75", "#F82736", "darkred")),
        layer.name = "Crimes affecting tourists per 1,000 local inhabitants by OHI region (2024)",
        legend = TRUE)

####################
## bar plot

p <- ggplot(
  viz_crime_rates_2024_plot,
  aes(x = rgn_id_num, y = crimes_per_1000, fill = rgn_id)
) +
  geom_col(color = "black", linewidth = 0.4) +
  scale_fill_manual(values = rgn_colors) +
  labs(
    title = "Crimes affecting tourists per 1,000 local inhabitants by OHI region (2024)",
    x = "Region",
    y = "Crimes per 1,000 local inhabitants"
  ) +
  theme_minimal(base_family = "Calibri") +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold", size = 16),
    axis.text.x = element_text(face = "bold", size = 16),
    axis.text.y = element_text(face = "bold", size = 16),
    axis.title.x = element_blank(),
    legend.position = "none"
  )

p

ggsave(
  filename = here::here(
    "/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_viz",
    "crime_rate_per_1000_by_region_2024.png"
  ),
  plot = p,
  width = 6.5,
  height = 5,
  units = "in",
  dpi = 300
)

```

Trend in recent years:

```{r}
last_3yrs <- c(2022, 2023, 2024)

trend_last3_slope <- sesnsp_goc_tr_crimes_percapita %>%
  st_drop_geometry() %>% 
  filter(ano %in% last_3yrs) %>%
  group_by(rgn_id) %>%
  summarise(
    slope_per_year = coef(lm(tr_crimes_percapita ~ ano))[2],
    .groups = "drop"
  )

trend_last3_pct <- sesnsp_goc_tr_crimes_percapita %>%
  st_drop_geometry() %>% 
  filter(ano %in% last_3yrs) %>%
  group_by(rgn_id) %>%
  summarise(
    crime_percap_2022 = tr_crimes_percapita[ano == 2022],
    crime_percap_2024 = tr_crimes_percapita[ano == 2024],
    pct_change_3yr = (crime_percap_2024 - crime_percap_2022) / crime_percap_2022 * 100,
    .groups = "drop"
  ) 

dom_trend_last3yrs <- full_join(trend_last3_pct, trend_last3_slope) %>% 
  filter(arrival_type == "dom") %>% 
  mutate(target_arrivals = arrivals_2024 * 1.098,
         proj_arrivals_2030 = arrivals_2024 + (6 * slope_per_year),
         on_track = ifelse(proj_arrivals_2030 >= target_arrivals, 1, 0))
```



#### Prelim Scores

```{r}
max_tr_crime_percapita_allrgns <- max(sesnsp_goc_tr_crimes_percapita$tr_crimes_percapita, na.rm = TRUE)

tr_qual_crime_score_alt <- sesnsp_goc_tr_crimes_percapita %>%
  mutate(
    current_status = (1 - (tr_crimes_percapita / max_tr_crime_percapita_allrgns)) * 100
  ) 

write_csv(tr_qual_crime_score_alt, here::here("TR/tr_quality_int","tr_qual_crime_score_120925.csv"))
```

#### Plotly

```{r}
library(plotly)

## TR quality crime index score over time
tr_qual_crime_plotly <- plotly::plot_ly(
  tr_qual_crime_score_alt,
  x = ~ano, 
  y = ~current_status, 
  color = ~rgn_id,
  type = "scatter", 
  mode = "lines+markers"
) %>%
  layout(
    title = "CURRENT STATUS: Índice de delitos per cápita que afectan la calidad turística por región",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Score (out of 100)"),
    legend = list(title = list(text = "Región"))
  ) %>%
  add_lines(
    x = range(tr_qual_crime_score_alt$ano), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (0 crimes = 100/100)",
    inherit = FALSE
  )

tr_qual_crime_plotly
```

### Assuming that when there is no data, that the score is 100...only for regions that have data at some point.

If there are no municipios in that OHI region then they should not be gapfilled like this.

```{r}
## fill in regions missing any values with 100 since the summarize may have dropped any regions with 0.

## for rgn 2 (which has no DOCUMENTED crimes from sesnsp, but definitely crime to some degree...... <https://bahiadelosangeles.org/safety.html>)
region_2_na <- tibble(
  ano = 2015:2025,
  rgn_id = "Region_2"
) %>%
  mutate(current_status = NA_real_)

tr_qual_crime_score_filled <- tr_qual_crime_score_alt %>% 
  select(ano, rgn_id, current_status) %>%
  complete(ano, rgn_id, fill = list(current_status = 100)) %>%   # only rgn 1 and rgn 2 needed to be filled.
  bind_rows(region_2_na)

tr_qual_crime_plotly_filled <- plot_ly(
  tr_qual_crime_score_filled,
  x = ~ano,
  y = ~current_status,
  color = ~rgn_id,
  type = "scatter",
  mode = "lines+markers"
) %>%
  layout(
    title = "CURRENT STATUS: Índice de delitos que afectan la calidad turística por región",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Score (out of 100)"),
    legend = list(title = list(text = "Región"))
  ) %>%
  add_lines(
    x = range(tr_qual_crime_score_filled$ano),
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (0 crimes = 100/100)",
    inherit = FALSE
  )

tr_qual_crime_plotly_filled
```

```{r}
# # bring back geometry
# inland_50 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer","inland_50km_goc.shp"))
# 
# tr_qual_crime_score_filled_sf <- inland_50 %>%
#   left_join(tr_qual_crime_score_filled, by = "rgn_id")
# 
# # st_write(tr_qual_crime_score_filled_sf, here::here("TR/tr_quality_int","tr_qual_crime_score_filled_sf.shp"))
# 
# tr_qual_crime_score_filled_sf_2025 <- tr_qual_crime_score_filled_sf %>% 
#   filter(ano %in% "2025")
# 
# mapview(tr_qual_crime_score_filled_sf_2025, zcol = "current_status")
```
