---
title: 'OHI GoC 2025: SUSTAINABILITY Tourism & Recreation'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# TR Sustainability

## Distintivo S

"https://www.gob.mx/presidencia/prensa/el-objetivo-para-turismo-en-el-plan-mexico-es-ser-el-5to-pais-mas-visitado-del-mundo-para-el-2030-presidenta-claudia-sheinbaum?idiom=es"

To achieve the goal of being the fifth most visited country in the world by 2030, it is estimated that it is necessary to increase the number of tourists by three million each year, as well as increase:
 

The economic impact of international tourists by 46 percent.
Hotel rooms by 12 percent.
The generation of tourism jobs by 27 percent.
The tourism Gross Domestic Product (GDP) by 9 percent.
Domestic tourism by 9.8 percent.
*The number of service providers certified in sustainability by 31.5 percent.*
The average expenditure per international tourist by 3.3 percent.
The arrival of international visitors by 30 percent
The increase in DNR revenue by 40.5 percent.

However, instead we will impose a more stringent and easily comprehensible reference point --> any hotel that has 100+ employees in the coastal area of the Gulf (50km) should be certified in Distintivo S.

```{r}
library(tidyverse)
library(here)
library(janitor)
```

```{r}
## read the given spreadsheet from SECTUR via PNT
dist_S_2023_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/distintivo_S/tabulated_distintivo_S", "tabula-Directorio_de_empresas_vigentes_Distintivo_S.csv"))
```

```{r}
# # from PNT request
# dist_S_raw <- read_excel(here::here("/home/shares/ohi/OHI_GOC/_raw_data/distintivo_S/requested_PNT/4-260126 Respuesta UDT 340027000021125/Anexo SST 340027000021125","Anexo SST DISTINTIVO S  2025.xls")) %>% 
#   janitor::clean_names() 
# 
# colnames(dist_S_raw)
# 
# # first, define col_types vector based on column positions
# # fecha columns: 3, 15, 16, 18, 19, 21, 22, 24, 25, 27, 28, 30, 31, 33, 34, 36, 37, 39, 40, 42, 43, 45, 46, 48, 49
# col_types_vec <- c(
#   "text",   # 1  no_consecutivos
#   "text",   # 2  no_de_folio_distintivo
#   "date",   # 3  fecha_de_inicio_del_programa_en_la_empresa_inicio_historico
#   "text",   # 4  nombre_comercial_de_la_empresa
#   "text",   # 5  direccion
#   "text",   # 6  colonia
#   "text",   # 7  c_p
#   "text",   # 8  ciudad
#   "text",   # 9  estado
#   "text",   # 10 localidad
#   "text",   # 11 lada
#   "text",   # 12 giro_general
#   "text",   # 13 distintivo_nuevo_renovacion
#   "text",   # 14 renovacion_primera_segunda_tercera_cuarta_quinta
#   "date",   # 15 fecha_emision_2013
#   "date",   # 16 fecha_de_expiracion_2014
#   "text",   # 17 tipo_de_distintivo_17
#   "date",   # 18 fecha_de_emision_2014
#   "date",   # 19 fecha_de_expiracion_2015
#   "text",   # 20 tipo_de_distintivo_20
#   "date",   # 21 fecha_de_emision_2015
#   "date",   # 22 fecha_de_expiracion_2016
#   "text",   # 23 tipo_de_distintivo_23
#   "date",   # 24 fecha_de_emision_2016
#   "date",   # 25 fecha_de_expiracion_2017
#   "text",   # 26 tipo_de_distintivo_26
#   "date",   # 27 fecha_de_emision_2017
#   "date",   # 28 fecha_de_expiracion_2018
#   "text",   # 29 tipo_de_distintivo_29
#   "date",   # 30 fecha_de_emision_2018
#   "date",   # 31 fecha_de_expiracion_2019
#   "text",   # 32 tipo_de_distintivo_32
#   "date",   # 33 fecha_de_emision_2019
#   "date",   # 34 fecha_de_expiracion_2020
#   "text",   # 35 tipo_de_distintivo_35
#   "date",   # 36 fecha_de_emision_2020
#   "date",   # 37 fecha_de_expiracion_2021
#   "text",   # 38 tipo_de_distintivo_38
#   "date",   # 39 fecha_de_emision_2021
#   "date",   # 40 fecha_de_expiracion_2022
#   "text",   # 41 tipo_de_distintivo_41
#   "date",   # 42 fecha_de_emision_2022
#   "date",   # 43 fecha_de_expiracion_2023
#   "text",   # 44 tipo_de_distintivo_44
#   "date",   # 45 fecha_de_emision_2023
#   "date",   # 46 fecha_de_expiracion_2024
#   "text",   # 47 tipo_de_distintivo_47
#   "date",   # 48 fecha_de_emision_2024
#   "date",   # 49 fecha_de_expiracion_2025
#   "text",   # 50 tipo_de_distintivo_50
#   "text",   # 51 status_vigente_vencido
#   "text"    # 52 ultimo_organismo_evaluador
# )
# 
# ## re-read it in now that we know the col_types!!
# dist_S_raw <- read_excel(
#   here::here("/home/shares/ohi/OHI_GOC/_raw_data/distintivo_S/requested_PNT/4-260126 Respuesta UDT 340027000021125/Anexo SST 340027000021125",
#              "Anexo SST DISTINTIVO S  2025.xls"),
#   col_types = col_types_vec
# ) %>% 
#   janitor::clean_names()
```

```{r}
# after manually fixing the historical dates:
col_types_vec <- c(
  "text",   # 1  no_consecutivos
  "text",   # 2  no_de_folio_distintivo
  "date",   # 2.5  ADDING IN UPDATED DATE!!!!
  "date",   # 3  fecha_de_inicio_del_programa_en_la_empresa_inicio_historico
  "text",   # 4  nombre_comercial_de_la_empresa
  "text",   # 5  direccion
  "text",   # 6  colonia
  "text",   # 7  c_p
  "text",   # 8  ciudad
  "text",   # 9  estado
  "text",   # 10 localidad
  "text",   # 11 lada
  "text",   # 12 giro_general
  "text",   # 13 distintivo_nuevo_renovacion
  "text",   # 14 renovacion_primera_segunda_tercera_cuarta_quinta
  "date",   # 15 fecha_emision_2013
  "date",   # 16 fecha_de_expiracion_2014
  "text",   # 17 tipo_de_distintivo_17
  "date",   # 18 fecha_de_emision_2014
  "date",   # 19 fecha_de_expiracion_2015
  "text",   # 20 tipo_de_distintivo_20
  "date",   # 21 fecha_de_emision_2015
  "date",   # 22 fecha_de_expiracion_2016
  "text",   # 23 tipo_de_distintivo_23
  "date",   # 24 fecha_de_emision_2016
  "date",   # 25 fecha_de_expiracion_2017
  "text",   # 26 tipo_de_distintivo_26
  "date",   # 27 fecha_de_emision_2017
  "date",   # 28 fecha_de_expiracion_2018
  "text",   # 29 tipo_de_distintivo_29
  "date",   # 30 fecha_de_emision_2018
  "date",   # 31 fecha_de_expiracion_2019
  "text",   # 32 tipo_de_distintivo_32
  "date",   # 33 fecha_de_emision_2019
  "date",   # 34 fecha_de_expiracion_2020
  "text",   # 35 tipo_de_distintivo_35
  "date",   # 36 fecha_de_emision_2020
  "date",   # 37 fecha_de_expiracion_2021
  "text",   # 38 tipo_de_distintivo_38
  "date",   # 39 fecha_de_emision_2021
  "date",   # 40 fecha_de_expiracion_2022
  "text",   # 41 tipo_de_distintivo_41
  "date",   # 42 fecha_de_emision_2022
  "date",   # 43 fecha_de_expiracion_2023
  "text",   # 44 tipo_de_distintivo_44
  "date",   # 45 fecha_de_emision_2023
  "date",   # 46 fecha_de_expiracion_2024
  "text",   # 47 tipo_de_distintivo_47
  "date",   # 48 fecha_de_emision_2024
  "date",   # 49 fecha_de_expiracion_2025
  "text",   # 50 tipo_de_distintivo_50
  "text",   # 51 status_vigente_vencido
  "text"    # 52 ultimo_organismo_evaluador
)

dist_S_raw_2 <- read_excel(
  here::here("/home/shares/ohi/OHI_GOC/_raw_data/distintivo_S/requested_PNT/4-260126 Respuesta UDT 340027000021125/Anexo SST 340027000021125",
             "updated_dates_Anexo_SST_DISTINTIVO_S_2025.xlsx"),
  col_types = col_types_vec
) %>% 
  janitor::clean_names() %>% 
  mutate(localidad = str_trim(localidad), # remove whitespaces between, before, and after words.
         localidad = str_squish(localidad),
         nombre_comercial_de_la_empresa = str_trim(nombre_comercial_de_la_empresa),
         nombre_comercial_de_la_empresa = str_squish(nombre_comercial_de_la_empresa))
```


### Geocode the hotels

```{r}
# see how estados are formatted
unique(dist_S_raw_2$estado)

#  [1] "Oaxaca"              "Guerrero"            "Quintana Roo"       
#  [4] "Nayarit"             "Jalisco"             "Sonora"             
#  [7] "Baja California Sur" "Ciudad de México"    "Nuevo León"         
# [10] "Colima"              "Yucatán"             "Veracruz"           
# [13] "Hidalgo"             "Chiapas"             "Campeche"           
# [16] "Puebla"              "Morelos"             "Sinaloa"            
# [19] "San Luis Potosí"     "Guanajuato"          "Zacatecas"          
# [22] "Chihuahua"           "Querétaro"           "Estado de México"   
# [25] "Coahuila"            "Baja California"     "Aguascalientes"     
# [28] "TAMAULIPAS"          "Tamaulipas"          "OAXACA"             
# [31] "PUEBLA"              "NAYARIT"             "NUEVO LEÓN"         
# [34] "MORELOS"             "BAJA CALIFORNIA SUR" "QUINTANA ROO" 

dist_S_goc_estados <- dist_S_raw_2 %>% 
  filter(estado %in% c("Nayarit","Jalisco","Sonora","Baja California Sur","Baja  Calfornia Sur", "BAJA CALIFORNIA SUR","Baja California","Sinaloa")) %>%
  mutate(country = "Mexico")

dist_S_goc_estados_geocode <- dist_S_goc_estados %>% 
  tidygeocoder::geocode(city = localidad, country = country, state = estado) # to give each city a lat, long

```

Fix all lat/longs that tidygeocoder was not able to do.

```{r}
isna_lat <- dist_S_goc_estados_geocode %>% 
  filter(is.na(lat))
# [1] "Dreams VillaMagna Resort & Spa"  (found manually on website <https://www.rivieranayarit.com/hotels/nuevo-nayarit-dreams-villamagna/> that address is Paseo de los Cocoteros Lotes 32 y 33, Fraccionamiento Nuevo Vallarta, Nayarit, México, C.P 63732, and then on Google Maps) = 20.712368, -105.302158    
# [2] "Sunscape Puerto Vallarta Resort & Spa" (found manually on Google Maps) = 20.642284, -105.238695
# [3] "IBEROSTAR SELECTION PLAYA MITA" (found manually on Google Maps) = 20.790514, -105.488250

## fix the lats and longs that were NA after `geocode()`
dist_S_goc_estados_geocode_fix <- dist_S_goc_estados_geocode %>% 
  mutate(lat = case_when(
    nombre_comercial_de_la_empresa == "Dreams VillaMagna Resort & Spa" ~ 20.712368,
    nombre_comercial_de_la_empresa == "Sunscape Puerto Vallarta Resort & Spa" ~ 20.642284,
    nombre_comercial_de_la_empresa == "IBEROSTAR SELECTION PLAYA MITA" ~ 20.790514,
    TRUE ~ lat
  ),
  long = case_when(
    nombre_comercial_de_la_empresa == "Dreams VillaMagna Resort & Spa" ~ -105.302158,
    nombre_comercial_de_la_empresa == "Sunscape Puerto Vallarta Resort & Spa" ~ -105.238695,
    nombre_comercial_de_la_empresa == "IBEROSTAR SELECTION PLAYA MITA" ~ -105.488250,
    TRUE ~ long
  ))
```


```{r}
dist_S_geocode_sf <- dist_S_goc_estados_geocode_fix %>% 
  st_as_sf(coords = c("long", "lat"),
           crs = 4326, # WGS84
           remove = FALSE)

library(mapview)
mapview(dist_S_geocode_sf)
plot(dist_S_geocode_sf)

# one in tijuana and one in guadalajara that should be filtered out.
```


<!-- ### Number of certifications over time -->

<!-- ```{r} -->
<!-- dist_S_presence <- dist_S_geocode_sf %>%  -->
<!--   distinct(nombre_comercial, year_documented) %>%  -->
<!--   mutate(present = 1) %>%  -->
<!--   tidyr::pivot_wider( -->
<!--     names_from = year_documented, -->
<!--     values_from = present, -->
<!--     values_fill = 0 -->
<!--   ) -->

<!-- heat_df <- dist_S_presence %>%  -->
<!--   arrange(nombre_comercial) %>%  -->
<!--   mutate(id = row_number()) %>%  -->
<!--   pivot_longer(cols = c(`2015`,`2017`,`2023`), -->
<!--                names_to = "year", -->
<!--                values_to = "present") -->

<!-- plot_ly( -->
<!--   heat_df, -->
<!--   x = ~year, -->
<!--   y = ~nombre_comercial, -->
<!--   z = ~present, -->
<!--   type = "heatmap", -->
<!--   colorscale = "Blues" -->
<!-- ) %>%  -->
<!--   layout(title = "Presence of Distintivo S Certification Over Time", -->
<!--          xaxis = list(title = "Year"), -->
<!--          yaxis = list(title = "Business")) -->
<!-- ``` -->

#### Standardize hotel names

```{r}
library(stringi)

# trime all whitespaces, str to lower, and remove accents
dist_S_all_clean <- dist_S_geocode_sf %>% 
  mutate(
    nombre_comercial_clean = nombre_comercial_de_la_empresa %>% 
      str_to_lower() %>% 
      str_squish() %>% 
      stri_trans_general("Latin-ASCII") %>% # remove accents
      str_replace_all("[^a-z0-9 ]", "") %>% # remove punctuation
      str_replace_all("\\s+", " ") # normalize spaces
  )

# unique(dist_S_all_clean$nombre_comercial_clean)

# go one step futher and remove parts of the names that are not unique to the hotel/business
dist_S_all_clean_2 <- dist_S_all_clean %>% 
  mutate(
    nombre_comercial_clean = nombre_comercial_clean %>% 
      str_replace_all("\\b(hotel|resort|spa|suites|suite|all inclusive|the|grupo)\\b", "") %>% 
      str_squish()
  )

# unique(dist_S_all_clean_2$nombre_comercial_clean)
```


```{r}



%>% 
  arrange(desc(nombre_comercial_clean)) %>% 
  group_by(nombre_comercial_clean) %>% 
  fill(folio_del_distintivo, .direction = "up") %>% 
  fill(sitio_web, .direction = "down") ## note that pueblo bonito mazatlan and emerald bay are two separate resorts.
```


#### Classify OHI Rgn ID

```{r}
library(sf)

## first, join with OHI GoC Regions

inland_50 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer","inland_50km_goc.shp"))
dist_S_all_clean_2_trans <- st_transform(dist_S_all_clean_2, st_crs(inland_50))
dist_S_goc_rgn <- st_intersection(dist_S_all_clean_2_trans, inland_50)

mapview(dist_S_goc_rgn, zcol = "rgn_id") + mapview(inland_50, zcol = "rgn_id")  # looks good!
mapview(dist_S_goc_rgn, zcol = "rgn_id") + mapview(inland_50, zcol = "rgn_id") + mapview(dist_S_all_clean_2) # double checking with the original layered on top

## NOTE: we lose a hotel in culiacan if we do 10km, which fits into the 50km buffer.
```

#### Last date certified? How many times certified?

```{r}
library(dplyr)

expiration_cols <- c(
  "fecha_de_expiracion_2014",
  "fecha_de_expiracion_2015",
  "fecha_de_expiracion_2016",
  "fecha_de_expiracion_2017",
  "fecha_de_expiracion_2018",
  "fecha_de_expiracion_2019",
  "fecha_de_expiracion_2020",
  "fecha_de_expiracion_2021",
  "fecha_de_expiracion_2022",
  "fecha_de_expiracion_2023",
  "fecha_de_expiracion_2024",
  "fecha_de_expiracion_2025"
)

dist_S_goc_rgn_recent_yr <- dist_S_goc_rgn %>%
  rowwise() %>%  # go row by row, looking across columns for each hotel
  mutate(
    
    # --- most_recent_expiration_year ---
    # we want to know the last year a hotel had an expiration date recorded/when did they last have an active certification??
    most_recent_expiration_year = {
      
      # grab all the expiration column values for this hotel
      vals <- c_across(all_of(expiration_cols))
      
      # find which columns actually have a date (are not NA) because each non-NA means the hotel had a cert that expired that year
      non_empty <- which(!is.na(vals))
      
      # if every expiration column is NA, this hotel has no certification history at all... this would NOT happen, but it is a good check for us to make sure this is working as it should.
      if (length(non_empty) == 0) {
        NA_real_  # just return NA
        
      } else {
        # otherwise, grab the value from the last non-NA column (the most recent one)
        # max(non_empty) gives us the index of the rightmost non-NA column
        last_val <- vals[[max(non_empty)]]
        
        # the columns are POSIXt datetime objects, not plain strings, so we can use format() to pull out just the year
        as.numeric(format(last_val, "%Y"))
      }
    },
    
    # --- times_certified ---
    # how many certification cycles has this hotel completed between 2014-2025?
    # each expiration column = one potential cycle.
    # non-NA = they did it, NA = they didn't.
    times_certified = {
      
      # all the expiration columns again for this hotel
      vals <- c_across(all_of(expiration_cols))
      
      # count the non-NAs:
      # times_certified = 1 means they certified once and never came back.
      # times_certified = 12 means they renewed every single year since 2014!!
      sum(!is.na(vals))
    },
    
    # --- years_in_program ---
    # how many years since the hotel first joined? from start year to 2025.
    years_in_program = {
      start_year <- if (!is.na(updated_fecha_de_inicio_del_programa_en_la_empresa_inicio_historico)) {
        # use the official start date if we have it
        as.numeric(format(updated_fecha_de_inicio_del_programa_en_la_empresa_inicio_historico, "%Y"))
      } else {
        # otherwise, dig through the expiration columns and grab the earliest one...
        vals <- c_across(all_of(expiration_cols))
        earliest_val <- vals[[min(which(!is.na(vals)))]]
        as.numeric(format(earliest_val, "%Y"))
      }
      2025 - start_year
    },
    
    # --- certification_consistency_pct ---
    # times_certified / years_in_program
    # 100% = renewed every year since joining, 25% = only 1 in every 4 years, and so on.
    certification_consistency_pct = (times_certified / years_in_program) * 100
    
  ) %>%
  ungroup() # drop the rowwise grouping

# see how the distribution looks across hotels
table(dist_S_goc_rgn_recent_yr$times_certified, useNA = "always")
```


```{r}
dist_S_goc_rgn_recent_yr_select <- dist_S_goc_rgn_recent_yr %>% 
  dplyr::select(most_recent_expiration_year, updated_fecha_de_inicio_del_programa_en_la_empresa_inicio_historico, nombre_comercial_de_la_empresa, nombre_comercial_clean, times_certified, years_in_program, certification_consistency_pct, rgn_id, estado, localidad, direccion) %>% 
  st_as_sf()

st_write(dist_S_goc_rgn_recent_yr_select, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int","distintivo_S_certified_50km_030926.gpkg"), append = FALSE) 
```

#### Counts by rgn

```{r}
# count the 
dist_S_counts_by_rgn <- dist_S_goc_rgn_recent_yr %>% 
  st_drop_geometry() %>% 
  distinct(rgn_id, nombre_comercial_clean, most_recent_expiration_year) %>% 
  group_by(rgn_id, most_recent_expiration_year) %>% 
  summarise(n_hotels = n(), .groups = "drop")
```


#### plotly over time

```{r}
dist_S_counts_by_rgn <- dist_S_counts_by_rgn %>% 
  mutate(year = as.numeric(year_documented))

library(plotly)

plot_ly(
  dist_S_counts_by_rgn,
  x = ~year,
  y = ~n_hotels,
  color = ~factor(rgn_id),
  type = 'scatter',
  mode = 'lines+markers'
) %>% 
  layout(
    title = "Distintivo S Certified Hotels by GoC Region (2015–2023)",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Number of Certified Hotels"),
    legend = list(title = list(text = "Region ID"))
  )

```


# STOPPED HERE

## Compare with DENUE Accommodations

```{r}
alojamientos_2019_2024_50km <- st_read(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas/int","update_alojamientos_2019_2024_goc_50km.shp"))

library(stringi)

alojamientos_2019_2024_50km_clean <- alojamientos_2019_2024_50km %>% 
  mutate(
    nombre_comercial_clean = nom_estab %>% 
      str_to_lower() %>% 
      str_squish() %>% 
      stri_trans_general("Latin-ASCII") %>% # remove accents
      str_replace_all("[^a-z0-9 ]", "") %>% # remove punctuation
      str_replace_all("\\s+", " ") %>%  # normalize spaces
      str_replace_all("\\b(hotel|resort|spa|suites|suite|all inclusive|the|grupo)\\b", "") %>% 
      str_squish()
  ) %>% 
  mutate(
    employees_estimated = case_when(
      per_ocu == "0 a 5 personas" ~ 3,
      per_ocu == "6 a 10 personas" ~ 8,
      per_ocu == "11 a 30 personas" ~ 20.5,
      per_ocu == "31 a 50 personas" ~ 40.5,
      per_ocu == "51 a 100 personas" ~ 75.5,
      per_ocu == "101 a 250 personas" ~ 175.5,
      per_ocu %in% c("251 y m<e1>s personas", "251 y más personas") ~ 300,
      TRUE ~ NA_real_
    )
  )

all_hotels <- unique(alojamientos_2019_2024_50km_clean$nombre_comercial_clean)
```


```{r}
dist_S_goc_50km <- st_read(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int","distintivo_S_certified_50km_030926.gpkg"))

dist_S_goc_rgn_recent_yr_select <- dist_S_goc_rgn_recent_yr %>% 
  dplyr::select(most_recent_expiration_year, updated_fecha_de_inicio_del_programa_en_la_empresa_inicio_historico, nombre_comercial_de_la_empresa, nombre_comercial_clean, times_certified, years_in_program, certification_consistency_pct, rgn_id, estado, localidad, direccion) %>% 
  st_as_sf()

dist_S_businesses <- dist_S_goc_rgn_recent_yr %>% 
  dplyr::select(nombre_comercial_de_la_empresa, nombre_comercial_clean, estado, localidad, rgn_id) %>% 
  mutate(across(c(nombre_comercial_de_la_empresa, nombre_comercial_clean, estado, localidad), 
                ~tolower(str_squish(.)))) %>% 
  distinct()
  
unique(dist_S_goc_50km$nmbr_c_)
print(dist_S_businesses)
```

```{r}
dist_S_businesses_names <- unique(dist_S_businesses$nombre_comercial_clean)

setdiff(dist_S_businesses_names, all_hotels)
##those not present in y (all_hotels) but are present in x:


setdiff(all_hotels, dist_S_businesses_names)
```

### Join w DENUE, find avg size of Distintivo S hotels.

```{r}
dist_s_denue_id <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int","DistintivoS_x_DENUE_Accommodations.csv")) %>% 
  janitor::clean_names() %>% 
  filter(!is.na(id_denue)) %>%  # if it did not have a denue id, it is either not a hotel or it did not have enough information to know which hotel it was. This was done through manual searches *after* using setdiff first.
  select(hotel_resort_name, id_denue)

join_denue_alojamientos_dist_s <- left_join(dist_s_denue_id, alojamientos_2019_2024_50km_clean, by = c("id_denue" = "id"))

length(unique(join_denue_alojamientos_dist_s$id_denue)) # 31, less than 35 because some had slightly different names but the same DENUE id.
```

Find the avg size of the hotels that have already been able to obtain dist. S, as a standard for the size of hotel that could potentially obtain the certification:

```{r}
# across all regions:
size_join_denue_alojamientos_dist_s <- join_denue_alojamientos_dist_s %>% 
  summarize(
    avg_employees_dist_s = mean(employees_estimated, na.rm = TRUE),
    sd = sd(employees_estimated, na.rm = TRUE),
    n = sum(!is.na(employees_estimated)),
    se = sd / sqrt(n)
  )

```

For now, filter out to only all the DENUE hotels that are that size or larger. 

```{r}
missing_regions <- expand.grid(
  rgn_id = c("Region_1", "Region_2", "Region_3"),
  data_year = c(2015, 2017, 2023),
  total_large_hotels = 0
) 

large_hotels_denue <- alojamientos_2019_2024_50km_clean %>% 
  filter(data_year %in% 2023) %>% 
  filter(employees_estimated >= 226) %>% 
  group_by(rgn_id, data_year) %>% 
  summarize(total_large_hotels = n()) %>% 
  st_drop_geometry() %>% 
  bind_rows(large_hotels_2015_2017) %>% 
  bind_rows(missing_regions) %>% # makes sure that every region is represented here.
  group_by(rgn_id, data_year) %>% 
  summarize(total_large_hotels = sum(total_large_hotels))

```

Now, summarize the number of hotels per rgn with dist. S: 

```{r}
sum_dist_S_goc_50km <- dist_S_goc_50km %>% 
  group_by(rgn_id, yr_dcmn) %>% 
  summarize(total_dist_s = n()) %>% 
  mutate(data_year = as.numeric(yr_dcmn)) %>% 
  st_drop_geometry()

joined_distS_vs_large_hotels <- full_join(sum_dist_S_goc_50km, large_hotels_denue, by = c("rgn_id","data_year"))

## regions that have large hotels but no distintivo S:
joined_distS_vs_large_hotels_fix <- joined_distS_vs_large_hotels %>% 
  mutate(total_dist_s = ifelse(is.na(total_dist_s) & data_year %in% c(2015, 2017, 2023), 0, total_dist_s)) # to show that for those regions and years, there were no dist. S certified hotels.
```

## Current Status

```{r}
dist_S_current_status <- joined_distS_vs_large_hotels_fix %>% 
  mutate(current_status = (total_dist_s / total_large_hotels) * 100) %>% 
  mutate(current_status = ifelse(total_dist_s == 0 & total_large_hotels == 0, 100, current_status)) %>%  # because if there are no large hotels to get certified, then they should get a 100% (for now).
  select(rgn_id, data_year, total_dist_s, total_large_hotels, current_status)

### table to visualize
library(gt)
library(dplyr)

region_levels <- paste0("Region_", 1:9)

dist_S_current_status %>%
  mutate(
    rgn_id = factor(rgn_id, levels = region_levels),
    current_status = current_status / 100
  ) %>%
  arrange(rgn_id) %>%
  gt() %>%
  cols_label(
    rgn_id = "Region",
    data_year = "Year",
    total_dist_s = "Certified Hotels (Distintivo S)",
    total_large_hotels = "Total Large Hotels (+250 employees)",
    current_status = "Certification Coverage"
  ) %>%
  fmt_percent(
    columns = current_status,
    decimals = 1
  ) %>%
  cols_align(
    align = "center",
    everything()
  ) %>%
  tab_header(
    title = "Distintivo S Certification Coverage",
    subtitle = "Certified hotels as a share of total large hotels (+250 employees)"
  ) %>%
  opt_table_outline() %>%
  opt_all_caps()

```


```{r}
dist_S_current_status_final <- dist_S_current_status %>% 
  select(rgn_id, data_year, current_status) %>% 
  ungroup()

# write_csv(dist_S_current_status_final, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","dist_S_current_status_final_121225.csv"))
write_csv(dist_S_current_status_final, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","dist_S_current_status_final_75over_010726.csv"))

dist_S_cs_final <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","dist_S_current_status_final_75over_010726.csv"))
dist_S_cs_old <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","dist_S_current_status_final_121225.csv"))

# avg_2023_dist2 <- dist_S_cs_final %>% ## because of Region_1,2,3 it is not representative to do an average because there are no large hotels in those rgns.
#   filter(data_year == 2023) %>% 
#   summarize(pct_large_hotels_goc = mean(current_status))

avg_2023_dist2 <- dist_S_cs_final %>% 
  filter(!rgn_id %in% c("Region_1", "Region_2", "Region_3")) %>% 
  filter(data_year == 2023) %>%
  summarize(pct_large_hotels_goc = mean(current_status)) # now seems more accurate.  Only ~8.26%
```


### Gapfill

```{r}
library(zoo)

dist_S_current_status_complete <- dist_S_current_status_final %>% 
  complete(rgn_id, data_year = 2015:2024) %>% 
  mutate(current_status = as.numeric(current_status))  


dist_S_current_status_gapfilled_interpolated <- dist_S_current_status_complete %>%
  arrange(rgn_id, data_year) %>%
  group_by(rgn_id) %>% 
  mutate(
    score_interpolated = na.approx(
      current_status,
      x = data_year,
      rule = 2,
      na.rm = FALSE
    ),
    is_gapfilled = ifelse(is.na(current_status), 1, 0)
  ) %>%
  ungroup()

write_csv(dist_S_current_status_gapfilled_interpolated, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","dist_S_current_status_final_interpolated_75over_010726.csv"))

# dist_S_current_status_gapfilled_lm <- dist_S_current_status_complete %>% ## lm feels unstable because we only have three points for each rgn_id.  Let's stick to interpolation here.
#   group_by(rgn_id) %>% 
#   mutate(
#     score_lm = {
#       fit <- lm(current_status ~ data_year, na.action = na.exclude)
#       predict(fit, newdata = data.frame(data_year = data_year))
#     }) %>% 
#   mutate(
#     score_final = if_else(
#       !is.na(current_status),
#       current_status,
#       score_lm),
#     is_gapfilled = ifelse(is.na(current_status), 1, 0)
#   )

library(plotly)

## plot the score over time
dist_S_current_status_gapfilled_interpolated_plotly <- plotly::plot_ly(
  dist_S_current_status_gapfilled_interpolated, 
  x = ~data_year, 
  y = ~score_interpolated, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "TR Sustainability: total dist. S hotels / total large hotels (interpolated)",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Interpolated score")
  ) %>%
  add_lines(
    x = range(dist_S_current_status_final$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
dist_S_current_status_gapfilled_interpolated_plotly
```


```{r}
library(plotly)

dist_S_current_status_final <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","dist_S_current_status_final_121225.csv"))

## plot the score over time
final_score_plotly <- plotly::plot_ly(
  dist_S_current_status_final, 
  x = ~data_year, 
  y = ~current_status, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "TR Sustainability: total dist. S hotels / total large hotels",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(dist_S_current_status_final$data_year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal 2030 (100/100)",
    inherit = FALSE
  )
final_score_plotly

large_hotels_denue_plot <- large_hotels_denue %>% 
  arrange(data_year)

## plot the score over time
denue_large_hotels_plotly <- plotly::plot_ly(
  large_hotels_denue_plot, 
  x = ~data_year, 
  y = ~total_large_hotels, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "Total DENUE Large Hotels (>226 employees) across the GoC (50km)",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Number of hotels")
  ) 
denue_large_hotels_plotly

```


