---
title: 'OHI GoC 2025: SUSTAINABILITY Tourism & Recreation'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# TR Sustainability

## Job quality in tourism

```{r}
library(tidyverse)
library(mapview)
library(sf)
library(here)
library(janitor)
```

### Read in LE data

```{r}
## bring in data cleaned from LE of the CE
ce_totals_clean_years_adjusted <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/ce_int","ce_cleaned_totals_per_yr_municipio_activity.csv"))
```


Now, limit to the tourism sectors.

```{r}
tr_ce_totals_clean_years_adjusted <- ce_totals_clean_years_adjusted %>% 
  filter(general_sector %in% "marine_tourism")

unique(tr_ce_totals_clean_years_adjusted$specific_sector)
# [1] "accommodations_food_drink" "marine_tourist_transport"  "travel_planners"          
# [4] "coastal_recreation"    

unique(tr_ce_totals_clean_years_adjusted$eng_sector_desc)
#  [1] "Hotels and motels, and similar"                                                                       
#  [2] "Food and alcoholic and non-alcoholic beverage preparation services; includes restaurants and similar."
#  [3] "Tourist water transport, includes 48721 and 487210."                                                  
#  [4] "Boarding houses, guest houses, and furnished apartments and houses with hotel services."              
#  [5] "Travel agencies."                                                                                     
#  [6] "Campgrounds and recreational lodges."                                                                 
#  [7] "Tourist marinas."                                                                                     
#  [8] "Other reservation services."                                                                          
#  [9] "Museums, historical sites, zoos, and similar."                                                        
# [10] "Organization of excursions and tourist packages for travel agencies (tour operators)."    

```


#### Adjustment for municipalities that cover both coastal and pacific sides, and sectors that are not strictly coastal

Region_3 and Region_4 have municipalities that span both the gulf and pacific sides of the peninsula, so we need to adjust to ensure those represent the number of jobs in the Gulf.  Also, some sectors such as tourism, are not explicitly marine-related.  So we only consider the "coastal" jobs in our analysis, assuming that if the tourism jobs are coastal, then they have some tie to the Gulf.  

Many people travel to see the Gulf specifically, and accommodations, food, drink, etc. support those visits.  And those visits not only support the local economy, but also create connections between people and the Gulf.

```{r}
inland_50_population_2020 <- read_csv(here::here("/home/shares/ohi/OHI_GOC/spatial/population_inland_50km","pop2020_by_region_vector_inland_50.csv"))

coastal_population_multipliers <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/le","coastal_adjustment_multiplier_2020.csv"))
```

```{r}
## sum by rgn
tr_total_jobs_per_rgn_and_activity <- tr_ce_totals_clean_years_adjusted %>% 
  group_by(ano_censal, rgn_id, specific_sector) %>% 
  summarize(
    
    # job metrics
    h000a_rgn = sum(h000a, na.rm = TRUE),
    h000b_rgn = sum(h000b, na.rm = TRUE),
    h000c_rgn = sum(h000c, na.rm = TRUE),
    h010a_rgn = sum(h010a, na.rm = TRUE),
    h010b_rgn = sum(h010b, na.rm = TRUE),
    h010c_rgn = sum(h010c, na.rm = TRUE),
    
    # remunerations metrics
    j000a_rgn = sum(j000a, na.rm = TRUE),
    j300a_rgn = sum(j300a, na.rm = TRUE),
    j400a_rgn = sum(j400a, na.rm = TRUE),
    j600a_rgn = sum(j600a, na.rm = TRUE)
    
  ) %>% 
  ungroup()
```


```{r}
tr_coastal_total_jobs_per_rgn_and_activity <- tr_total_jobs_per_rgn_and_activity %>% 
  full_join(coastal_population_multipliers, by = "rgn_id") %>% 
  mutate(
    
    # fix San Felipe
    # coastal_pop_adjustment = ifelse(rgn_id %in% "Region_1" & ano_censal %in% "2023", 1, coastal_pop_adjustment),
    
    # h000a
    first_adj_coastal_h000a = ifelse(rgn_id %in% c("Region_3","Region_4"), h000a_rgn * coastal_pop_adjustment, h000a_rgn),
    coastal_h000a = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), h000a_rgn * coastal_pop_adjustment, first_adj_coastal_h000a),
    
    # h000b
    first_adj_coastal_h000b = ifelse(rgn_id %in% c("Region_3","Region_4"), h000b_rgn * coastal_pop_adjustment, h000b_rgn),
    coastal_h000b = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), h000b_rgn * coastal_pop_adjustment, first_adj_coastal_h000b),
    
    # h000c
    first_adj_coastal_h000c = ifelse(rgn_id %in% c("Region_3","Region_4"), h000c_rgn * coastal_pop_adjustment, h000c_rgn),
    coastal_h000c = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), h000c_rgn * coastal_pop_adjustment, first_adj_coastal_h000c),
    
    # h010a
    first_adj_coastal_h010a = ifelse(rgn_id %in% c("Region_3","Region_4"), h010a_rgn * coastal_pop_adjustment, h010a_rgn),
    coastal_h010a = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), h010a_rgn * coastal_pop_adjustment, first_adj_coastal_h010a),
    
    # h010b
    first_adj_coastal_h010b = ifelse(rgn_id %in% c("Region_3","Region_4"), h010b_rgn * coastal_pop_adjustment, h010b_rgn),
    coastal_h010b = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), h010b_rgn * coastal_pop_adjustment, first_adj_coastal_h010b),
    
    # h010c
    first_adj_coastal_h010c = ifelse(rgn_id %in% c("Region_3","Region_4"), h010c_rgn * coastal_pop_adjustment, h010c_rgn),
    coastal_h010c = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), h010c_rgn * coastal_pop_adjustment, first_adj_coastal_h010c),
    
    # j000a
    first_adj_coastal_j000a = ifelse(rgn_id %in% c("Region_3","Region_4"), j000a_rgn * coastal_pop_adjustment, j000a_rgn),
    coastal_j000a = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), j000a_rgn * coastal_pop_adjustment, first_adj_coastal_j000a),
    
    # j300a
    first_adj_coastal_j300a = ifelse(rgn_id %in% c("Region_3","Region_4"), j300a_rgn * coastal_pop_adjustment, j300a_rgn),
    coastal_j300a = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), j300a_rgn * coastal_pop_adjustment, first_adj_coastal_j300a),
    
    # j400a
    first_adj_coastal_j400a = ifelse(rgn_id %in% c("Region_3","Region_4"), j400a_rgn * coastal_pop_adjustment, j400a_rgn),
    coastal_j400a = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), j400a_rgn * coastal_pop_adjustment, first_adj_coastal_j400a),
    
    # j600a
    first_adj_coastal_j600a = ifelse(rgn_id %in% c("Region_3","Region_4"), j600a_rgn * coastal_pop_adjustment, j600a_rgn),
    coastal_j600a = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), j600a_rgn * coastal_pop_adjustment, first_adj_coastal_j600a)
    
  ) %>% 
  dplyr::select(-c(starts_with("first_adj_")))
```

Aggregating the metrics for each region:
```{r}
total_per_rgn <- tr_coastal_total_jobs_per_rgn_and_activity %>% 
  group_by(ano_censal, rgn_id) %>% 
  summarize(
    
    # job metrics
    total_h000a = sum(coastal_h000a, na.rm = TRUE),
    total_h000b = sum(coastal_h000b, na.rm = TRUE),
    total_h000c = sum(coastal_h000c, na.rm = TRUE),
    total_h010a = sum(coastal_h010a, na.rm = TRUE),
    total_h010b = sum(coastal_h010b, na.rm = TRUE),
    total_h010c = sum(coastal_h010c, na.rm = TRUE),
    
    # remuneration metrics
    total_j000a = sum(coastal_j000a, na.rm = TRUE),
    total_j300a = sum(coastal_j300a, na.rm = TRUE),
    total_j400a = sum(coastal_j400a, na.rm = TRUE),
    total_j600a = sum(coastal_j600a, na.rm = TRUE)
    
  ) %>% 
  mutate(across(starts_with("total_"), round))
```


### Indicators of job quality

Maintained Salary
J000A / H010A to represent maintained/stable/increasing salary.

Maintained Stability
H010A / H000A = % of paid workers
J600A / H000A = severance pay scaled to the total number of jobs

Gender Equality
H010B/H000B  vs H010C/H000C  = % of paid workers men vs. women
Ideal state: women ratio is at least the same or equal to get a score of 100.


#### Maintained Salary

J000A = Total compensation (in millions of pesos): This includes all regular and extraordinary payments and contributions, in cash and in kind, before any deductions, to compensate the work of the company’s employees, in the form of wages and salaries, social benefits, and profits distributed to staff, whether such payment is calculated based on a workday, the amount of work performed (piecework), or a base salary supplemented by commissions for sales or other activities. Includes: employer contributions to social security schemes; payments made to personnel on leave or temporary leave. Excludes: payments for severance or compensation; payments to third parties for the provision of employed personnel; payments consisting exclusively of commissions for personnel who did not receive a fixed salary or wage; payments of fees for professional services contracted on an infrequent basis.

```{r}
# J000A / H010A to represent maintained/stable/increasing salary.
tr_job_salary <- total_per_rgn %>% 
  mutate(maintained_salary = total_j000a / total_h010a)
```


#### Maintained Stability

```{r}
# H010A / H000A = % of paid workers
# J600A / H000A = severance pay scaled to the total number of jobs

tr_job_stability <- total_per_rgn %>% 
  mutate(stability_formalized = total_h010a / total_h000a,
         stability_severance = total_j600a / total_h000a)
```


#### Gender Equality

```{r}
# the % formalized workers in the tourism sector should be the same between men and women (so that of the women working, the same percentage or more of them are formalized when compared to the number of formalized men out of total men working)
tr_job_gender_eq <- total_per_rgn %>% 
  mutate(men_formalized = total_h010b / total_h000b,
         women_formalized = total_h010c / total_h000c,
         prop_eq = women_paid / men_paid) # goal: prop_eq >= 1
```


##### plotlys

```{r}

```


### Current Status

```{r}

```


### Merged Current Status

```{r}

```







