---
title: 'OHI GoC 2025: SUSTAINABILITY Tourism & Recreation'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# TR Sustainability: Investment in potential tourism management

Investment in potential tourism management = avg(CONANP s046 Budget, CONANP v003 Budget, CONANP v005 Budget, CONANP Return of Cobro de Derechos)

<https://docs.google.com/spreadsheets/d/1jBPN54SIoEPb8fLxgRwTCLOrEeSdx_HRHQ-HPczcyYc/edit?gid=896923982#gid=896923982>

## CONANP Budget

### descripcion

*s046_sustainable_conservation*
Programa de Conservación para el Desarrollo Sostenible (PROCODES)

*v003_anp_management*
(antes E016-16) Conservación y Manejo de Áreas Naturales Protegidas

*v005_protection_restoration*
(antes U040-16) Programa para la Protección y Restauración de Ecosistemas y Especies en Riesgo (U025-16, del 2012 hasta el 2018)

*presupuesto_total_mdp_2026*
total federal budget to CONANP in millions of 2026 pesos.


```{r}
fed_budget_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int/conanp_int/fed_budget_programas_importantes_2012_2026.csv"))

fed_budget_sust_tourism <- fed_budget_raw %>% 
  dplyr::select(year, v003_anp_management, s046_sustainable_conservation, v005_protection_restoration, presupuesto_total_mdp_2026)
```

```{r}
plot(fed_budget_sust_tourism$year, fed_budget_sust_tourism$s046_sustainable_conservation)
plot(fed_budget_sust_tourism$year, fed_budget_sust_tourism$v003_anp_management)
plot(fed_budget_sust_tourism$year, fed_budget_sust_tourism$v005_protection_restoration)
plot(fed_budget_sust_tourism$year, fed_budget_sust_tourism$presupuesto_total_mdp_2026)
```


I will use 2012 - 2018 as the historial reference period to determine if the budget is increasing, decreasing, or staying relatively constant over time.

### Find hist avg

```{r}
fed_budget_sust_tourism_long <- fed_budget_sust_tourism %>% 
  pivot_longer(
    cols = v003_anp_management:presupuesto_total_mdp_2026,
    names_to = "programas",
    values_to = "mdp_2026"
  )

w_hist_avg_fed_budget <- fed_budget_sust_tourism_long %>% 
  filter(year %in% 2012:2018) %>% 
  group_by(programas) %>% 
  summarize(
    hist_avg_mdp_2026 = mean(mdp_2026)
  ) %>% 
  ungroup()
```


### Current status

#### Each program

```{r}
current_status_fed_budget <- left_join(fed_budget_sust_tourism_long, w_hist_avg_fed_budget) %>% 
  mutate(current_status_raw = (mdp_2026 / hist_avg_mdp_2026) * 100,
         current_status = ifelse(current_status_raw > 100, 100, current_status_raw)) 
```


##### plotly

```{r}
library(plotly)

fed_budget_plotly <- plotly::plot_ly(
  current_status_fed_budget, 
  x = ~year, 
  y = ~current_status_raw, 
  color = ~programas, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: CONANP budget of programs (millones de pesos 2026)",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(current_status_fed_budget$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
fed_budget_plotly
```


#### Expand for each rgn and year

```{r}
# # without presupuesto_total_mdp_2026
# current_status_fed_budget_merged_wo_presupuesto <- current_status_fed_budget %>% 
#   filter(!programas %in% "presupuesto_total_mdp_2026") %>% 
#   expand_grid(
#     rgn_id = paste0("Region_", 1:9),
#     year = 2012:2026
#   ) %>%
#   left_join(pct_returned_df, by = "year") %>% 
#   rename(score = pct_returned)

# without presupuesto_total_mdp_2026
current_status_conanp_budget_expanded <- current_status_fed_budget %>% 
  filter(!programas %in% "presupuesto_total_mdp_2026") %>% 
  expand_grid(
    rgn_id = paste0("Region_", 1:9)
  )
```

<!-- ```{r} -->
<!-- all_fed_budget_plotly <- plotly::plot_ly( -->
<!--   current_status_fed_budget_merged_wo_presupuesto,  -->
<!--   x = ~year,  -->
<!--   y = ~current_status_all,  -->
<!--   type = "scatter",  -->
<!--   mode = "lines+markers" -->
<!-- ) %>%  -->
<!--   layout( -->
<!--     title = "CURRENT STATUS: CONANP budget of programs (millones de pesos 2026)", -->
<!--     xaxis = list(title = "Año // Year"), -->
<!--     yaxis = list(title = "Puntaje // Percent (out of 100)") -->
<!--   ) %>% -->
<!--   add_lines( -->
<!--     x = range(current_status_fed_budget_merged_wo_presupuesto$year),  -->
<!--     y = c(100, 100), -->
<!--     line = list(color = "black", width = 2, dash = "dash"), -->
<!--     name = "Meta // Goal (100/100)", -->
<!--     inherit = FALSE -->
<!--   ) -->
<!-- all_fed_budget_plotly -->
<!-- ``` -->

##### Save to Mazu

```{r}
write_csv(current_status_fed_budget_merged_wo_presupuesto, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","conanp_federal_budget_current_status_final_032026.csv"))
```


## CONANP Returned Fees

Original location: <https://docs.google.com/spreadsheets/d/1jBPN54SIoEPb8fLxgRwTCLOrEeSdx_HRHQ-HPczcyYc/edit?gid=0#gid=0>

```{r}

pct_returned_df <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/NOSSA_cuidar_lo_que_importa/d2026","CONANP_fees_returned_2007_2026.csv")) %>% 
  janitor::clean_names() %>% 
  dplyr::select(year, pct_returned)

expanded_pct_returned <- expand_grid(
  rgn_id = paste0("Region_", 1:9),
  year = 2007:2024
) %>%
  left_join(pct_returned_df, by = "year") %>% 
  mutate(programas = "return_cobros_from_shcp",
         current_status = pct_returned * 100)

write_csv(expanded_pct_returned, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","conanp_fee_return_current_status_final_121625.csv"))

```


## Avg the CS of each program and the return of cobro de derechos together for an overall "Investment in potential tourism management" score.


```{r}
cs_conanp_budget_w_return <- current_status_conanp_budget_expanded %>% 
  full_join(expanded_pct_returned) %>% 
  dplyr::select(year, rgn_id, programas, current_status)

unique(cs_investment_tourism_management$programas)

cs_investment_tourism_management <- cs_conanp_budget_w_return %>% 
  group_by(year, rgn_id) %>% # though every region has the same value for that year!!
  mutate(current_status_all = mean(current_status)) %>% 
  ungroup()
```

## Save to Mazu

```{r}
write_csv(cs_investment_tourism_management, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","investment_tourism_management_final_033026.csv"))
```


