---
title: 'OHI GoC 2025: SUSTAINABILITY Tourism & Recreation'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# TR Sustainability

## Job quality in tourism

```{r}
library(tidyverse)
library(mapview)
library(sf)
library(here)
library(janitor)
```

### Read in LE data

```{r}
## bring in data cleaned from LE of the CE
ce_totals <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/ce_int","ce_cleaned_totals_per_yr_municipio_activity.csv"))
```


Now, limit to the tourism sectors.

```{r}
tr_ce_totals <- ce_totals %>% 
  filter(general_sector %in% "marine_tourism")

unique(tr_ce_totals$specific_sector)
# [1] "accommodations_food_drink" "marine_tourist_transport"  "travel_planners"          
# [4] "coastal_recreation"    

unique(tr_ce_totals$eng_sector_desc)
#  [1] "Hotels and motels, and similar"                                                                       
#  [2] "Food and alcoholic and non-alcoholic beverage preparation services; includes restaurants and similar."
#  [3] "Tourist water transport, includes 48721 and 487210."                                                  
#  [4] "Boarding houses, guest houses, and furnished apartments and houses with hotel services."              
#  [5] "Travel agencies."                                                                                     
#  [6] "Campgrounds and recreational lodges."                                                                 
#  [7] "Tourist marinas."                                                                                     
#  [8] "Other reservation services."                                                                          
#  [9] "Museums, historical sites, zoos, and similar."                                                        
# [10] "Organization of excursions and tourist packages for travel agencies (tour operators)."    

```

#### Adjust for inflation

This will be done using the package `priceR`.  Documentation is located here: <https://cran.r-project.org/web/packages/priceR/readme/README.html> 

"adjust_for_inflation() automatically converts between nominal and real dollars, or in/deflates prices from one year’s prices to another’s.  It works for 304 countries / areas (see them with all by running show_countries())."

I adjusted all values in pesos to constant 2023 pesos to remove the effects of inflation and ensure that differences across 2003, 2008, 2013, 2018, and 2023 represent real changes in economic activity rather than changes in the general price level. Using constant pesos preserves internal consistency for regional comparisons within Mexico without introducing unnecessary distortions from exchange rates or purchasing power parity conversions, which are only required for cross-country analysis.

```{r}
library(priceR)

tr_ce_totals_clean_years_adjusted <- tr_ce_totals %>% 
  mutate(
    j000a_adj = adjust_for_inflation(
      price = j000a,
      from_date = ano_censal,
      country = "Mexico",
      to_date = 2023
      ), 
    j400a_adj = adjust_for_inflation(
      price = j400a,
      from_date = ano_censal,
      country = "Mexico",
      to_date = 2023
      ),
    j600a_adj = adjust_for_inflation(
      price = j600a,
      from_date = ano_censal,
      country = "Mexico",
      to_date = 2023
      ),
    j300a_adj = adjust_for_inflation(
      price = j300a,
      from_date = ano_censal,
      country = "Mexico",
      to_date = 2023
      )
  ) %>% 
  dplyr::select(-c(j000a, j300a, j400a, j600a))
```


#### Adjustment for municipalities that cover both coastal and pacific sides, and sectors that are not strictly coastal

Region_3 and Region_4 have municipalities that span both the gulf and pacific sides of the peninsula, so we need to adjust to ensure those represent the number of jobs in the Gulf.  Also, some sectors such as tourism, are not explicitly marine-related.  So we only consider the "coastal" jobs in our analysis, assuming that if the tourism jobs are coastal, then they have some tie to the Gulf.  

Many people travel to see the Gulf specifically, and accommodations, food, drink, etc. support those visits.  And those visits not only support the local economy, but also create connections between people and the Gulf.

```{r}
inland_50_population_2020 <- read_csv(here::here("/home/shares/ohi/OHI_GOC/spatial/population_inland_50km","pop2020_by_region_vector_inland_50.csv"))

coastal_population_multipliers <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/le","coastal_adjustment_multiplier_2020.csv"))
```

```{r}
## sum by rgn
tr_total_jobs_per_rgn_and_activity <- tr_ce_totals_clean_years_adjusted %>% 
  group_by(ano_censal, rgn_id, specific_sector) %>% 
  summarize(
    
    # job metrics
    h000a_rgn = sum(h000a, na.rm = TRUE),
    h000b_rgn = sum(h000b, na.rm = TRUE),
    h000c_rgn = sum(h000c, na.rm = TRUE),
    h010a_rgn = sum(h010a, na.rm = TRUE),
    h010b_rgn = sum(h010b, na.rm = TRUE),
    h010c_rgn = sum(h010c, na.rm = TRUE),
    
    # remunerations metrics
    j000a_rgn = sum(j000a_adj, na.rm = TRUE),
    j300a_rgn = sum(j300a_adj, na.rm = TRUE),
    j400a_rgn = sum(j400a_adj, na.rm = TRUE),
    j600a_rgn = sum(j600a_adj, na.rm = TRUE)
    
  ) %>% 
  ungroup()
```


```{r}
tr_coastal_total_jobs_per_rgn_and_activity <- tr_total_jobs_per_rgn_and_activity %>% 
  full_join(coastal_population_multipliers, by = "rgn_id") %>% 
  mutate(
    
    # fix San Felipe
    # coastal_pop_adjustment = ifelse(rgn_id %in% "Region_1" & ano_censal %in% "2023", 1, coastal_pop_adjustment),
    
    # h000a
    first_adj_coastal_h000a = ifelse(rgn_id %in% c("Region_3","Region_4"), h000a_rgn * coastal_pop_adjustment, h000a_rgn),
    coastal_h000a = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), h000a_rgn * coastal_pop_adjustment, first_adj_coastal_h000a),
    
    # h000b
    first_adj_coastal_h000b = ifelse(rgn_id %in% c("Region_3","Region_4"), h000b_rgn * coastal_pop_adjustment, h000b_rgn),
    coastal_h000b = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), h000b_rgn * coastal_pop_adjustment, first_adj_coastal_h000b),
    
    # h000c
    first_adj_coastal_h000c = ifelse(rgn_id %in% c("Region_3","Region_4"), h000c_rgn * coastal_pop_adjustment, h000c_rgn),
    coastal_h000c = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), h000c_rgn * coastal_pop_adjustment, first_adj_coastal_h000c),
    
    # h010a
    first_adj_coastal_h010a = ifelse(rgn_id %in% c("Region_3","Region_4"), h010a_rgn * coastal_pop_adjustment, h010a_rgn),
    coastal_h010a = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), h010a_rgn * coastal_pop_adjustment, first_adj_coastal_h010a),
    
    # h010b
    first_adj_coastal_h010b = ifelse(rgn_id %in% c("Region_3","Region_4"), h010b_rgn * coastal_pop_adjustment, h010b_rgn),
    coastal_h010b = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), h010b_rgn * coastal_pop_adjustment, first_adj_coastal_h010b),
    
    # h010c
    first_adj_coastal_h010c = ifelse(rgn_id %in% c("Region_3","Region_4"), h010c_rgn * coastal_pop_adjustment, h010c_rgn),
    coastal_h010c = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), h010c_rgn * coastal_pop_adjustment, first_adj_coastal_h010c),
    
    # j000a
    first_adj_coastal_j000a = ifelse(rgn_id %in% c("Region_3","Region_4"), j000a_rgn * coastal_pop_adjustment, j000a_rgn),
    coastal_j000a = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), j000a_rgn * coastal_pop_adjustment, first_adj_coastal_j000a),
    
    # j300a
    first_adj_coastal_j300a = ifelse(rgn_id %in% c("Region_3","Region_4"), j300a_rgn * coastal_pop_adjustment, j300a_rgn),
    coastal_j300a = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), j300a_rgn * coastal_pop_adjustment, first_adj_coastal_j300a),
    
    # j400a
    first_adj_coastal_j400a = ifelse(rgn_id %in% c("Region_3","Region_4"), j400a_rgn * coastal_pop_adjustment, j400a_rgn),
    coastal_j400a = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), j400a_rgn * coastal_pop_adjustment, first_adj_coastal_j400a),
    
    # j600a
    first_adj_coastal_j600a = ifelse(rgn_id %in% c("Region_3","Region_4"), j600a_rgn * coastal_pop_adjustment, j600a_rgn),
    coastal_j600a = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), j600a_rgn * coastal_pop_adjustment, first_adj_coastal_j600a)
    
  ) %>% 
  dplyr::select(-c(starts_with("first_adj_")))
```

Aggregating the metrics for each region:
```{r}
total_per_rgn <- tr_coastal_total_jobs_per_rgn_and_activity %>% 
  group_by(ano_censal, rgn_id) %>% 
  summarize(
    
    # job metrics
    total_h000a = sum(coastal_h000a, na.rm = TRUE),
    total_h000b = sum(coastal_h000b, na.rm = TRUE),
    total_h000c = sum(coastal_h000c, na.rm = TRUE),
    total_h010a = sum(coastal_h010a, na.rm = TRUE),
    total_h010b = sum(coastal_h010b, na.rm = TRUE),
    total_h010c = sum(coastal_h010c, na.rm = TRUE),
    
    # remuneration metrics
    total_j000a = sum(coastal_j000a, na.rm = TRUE),
    total_j300a = sum(coastal_j300a, na.rm = TRUE),
    total_j400a = sum(coastal_j400a, na.rm = TRUE),
    total_j600a = sum(coastal_j600a, na.rm = TRUE)
    
  ) %>% 
  mutate(across(starts_with("total_"), round))
```


### Indicators of job quality

Maintained Salary
J000A / H010A to represent maintained/stable/increasing salary.

Maintained Stability
H010A / H000A = % of paid workers
J600A / H000A = severance pay scaled to the total number of jobs

Gender Equality
H010B/H000B  vs H010C/H000C  = % of paid workers men vs. women
Ideal state: women ratio is at least the same or equal to get a score of 100.


#### Maintained Salary

J000A = Total compensation (in millions of pesos): This includes all regular and extraordinary payments and contributions, in cash and in kind, before any deductions, to compensate the work of the company’s employees, in the form of wages and salaries, social benefits, and profits distributed to staff, whether such payment is calculated based on a workday, the amount of work performed (piecework), or a base salary supplemented by commissions for sales or other activities. Includes: employer contributions to social security schemes; payments made to personnel on leave or temporary leave. Excludes: payments for severance or compensation; payments to third parties for the provision of employed personnel; payments consisting exclusively of commissions for personnel who did not receive a fixed salary or wage; payments of fees for professional services contracted on an infrequent basis.

```{r}
# J000A / H010A to represent maintained/stable/increasing salary.
tr_job_salary <- total_per_rgn %>% 
  mutate(maintained_salary = total_j000a / total_h010a) %>% 
  ungroup()
```


#### Maintained Stability

```{r}
# H010A / H000A = % of paid workers
# J600A / H000A = severance pay scaled to the total number of jobs

tr_job_stability <- total_per_rgn %>% 
  mutate(stability_formalized = total_h010a / total_h000a,
         stability_severance = total_j600a / total_h000a) %>% 
  ungroup()
```


#### Gender Equality

```{r}
# the % formalized workers in the tourism sector should be the same between men and women (so that of the women working, the same percentage or more of them are formalized when compared to the number of formalized men out of total men working)
tr_job_gender_eq <- total_per_rgn %>% 
  mutate(men_formalized = total_h010b / total_h000b,
         women_formalized = total_h010c / total_h000c,
         prop_eq = women_formalized / men_formalized) %>% # goal: prop_eq >= 1  
  ungroup()
```


##### plotlys

```{r}
# --- MAINTAINED SALARY ---

# total_j000a: Total compensation
total_j000a_plotly <- plotly::plot_ly(
  tr_job_salary,
  x = ~ano_censal,
  y = ~total_j000a,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines"
) %>%
  layout(
    title = "Total compensation (J000A, millions of pesos) per region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Total compensation (millions of pesos)")
  )
total_j000a_plotly

# total_h010a: Total paid workers
total_h010a_plotly <- plotly::plot_ly(
  tr_job_salary,
  x = ~ano_censal,
  y = ~total_h010a,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines"
) %>%
  layout(
    title = "Total paid/formalized workers (H010A) per region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Number of paid workers")
  )
total_h010a_plotly

# maintained_salary: J000A / H010A
maintained_salary_raw_plotly <- plotly::plot_ly(
  tr_job_salary,
  x = ~ano_censal,
  y = ~maintained_salary,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines"
) %>%
  layout(
    title = "Maintained salary (J000A / H010A): compensation per paid worker per region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Compensation per paid worker (millions of pesos)")
  )
maintained_salary_raw_plotly
```


```{r}
# --- MAINTAINED STABILITY ---

# total_h000a: Total workers
total_h000a_plotly <- plotly::plot_ly(
  tr_job_stability,
  x = ~ano_censal,
  y = ~total_h000a,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines"
) %>%
  layout(
    title = "Total workers (H000A) per region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Total number of workers")
  )
total_h000a_plotly

# total_j600a: Severance pay
total_j600a_plotly <- plotly::plot_ly(
  tr_job_stability,
  x = ~ano_censal,
  y = ~total_j600a,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines"
) %>%
  layout(
    title = "Total severance pay (J600A, millions of pesos) per region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Severance pay (millions of pesos)")
  )
total_j600a_plotly

# stability_formalized: H010A / H000A
stability_formalized_raw_plotly <- plotly::plot_ly(
  tr_job_stability,
  x = ~ano_censal,
  y = ~stability_formalized,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines"
) %>%
  layout(
    title = "Job stability: % formalized/paid workers (H010A / H000A) per region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Proportion of paid workers")
  )
stability_formalized_raw_plotly

# stability_severance: J600A / H000A
stability_severance_raw_plotly <- plotly::plot_ly(
  tr_job_stability,
  x = ~ano_censal,
  y = ~stability_severance,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines"
) %>%
  layout(
    title = "Job stability: severance pay per total worker (J600A / H000A) per region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Severance pay per worker (millions of pesos)")
  )
stability_severance_raw_plotly
```


```{r}
# --- GENDER EQUALITY ---

# total_h000b: Total male workers
total_h000b_plotly <- plotly::plot_ly(
  tr_job_gender_eq,
  x = ~ano_censal,
  y = ~total_h000b,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines"
) %>%
  layout(
    title = "Total male workers (H000B) per region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Total male workers")
  )
total_h000b_plotly

# total_h000c: Total female workers
total_h000c_plotly <- plotly::plot_ly(
  tr_job_gender_eq,
  x = ~ano_censal,
  y = ~total_h000c,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines"
) %>%
  layout(
    title = "Total female workers (H000C) per region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Total female workers")
  )
total_h000c_plotly

# total_h010b: Total paid male workers
total_h010b_plotly <- plotly::plot_ly(
  tr_job_gender_eq,
  x = ~ano_censal,
  y = ~total_h010b,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines"
) %>%
  layout(
    title = "Total paid male workers (H010B) per region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Total paid male workers")
  )
total_h010b_plotly

# total_h010c: Total paid female workers
total_h010c_plotly <- plotly::plot_ly(
  tr_job_gender_eq,
  x = ~ano_censal,
  y = ~total_h010c,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines"
) %>%
  layout(
    title = "Total paid female workers (H010C) per region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Total paid female workers")
  )
total_h010c_plotly

# men_formalized: H010B / H000B
men_formalized_plotly <- plotly::plot_ly(
  tr_job_gender_eq,
  x = ~ano_censal,
  y = ~men_formalized,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines"
) %>%
  layout(
    title = "% formalized male workers (H010B / H000B) per region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Proportion of paid male workers")
  )
men_formalized_plotly

# women_formalized: H010C / H000C
women_formalized_plotly <- plotly::plot_ly(
  tr_job_gender_eq,
  x = ~ano_censal,
  y = ~women_formalized,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines"
) %>%
  layout(
    title = "% formalized female workers (H010C / H000C) per region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Proportion of paid female workers")
  )
women_formalized_plotly

# prop_eq: women_formalized / men_formalized
prop_eq_raw_plotly <- plotly::plot_ly(
  tr_job_gender_eq,
  x = ~ano_censal,
  y = ~prop_eq,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines"
) %>%
  layout(
    title = "Gender equality: ratio of formalized women to men workers (H010C/H000C vs H010B/H000B) per region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "prop_eq (goal: >= 1)")
  ) %>%
  add_lines(
    x = range(tr_job_gender_eq$ano_censal),
    y = c(1, 1),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (1)",
    inherit = FALSE
  )
prop_eq_raw_plotly
```


### Current Status

For TR Job Quality, it is that stability_formalized, stability_severance, prop_eq, and maintained_salary will either be equal to or greater than the avg value from 2003:2018.

```{r}
tr_job_hist_avg <- left_join(tr_job_gender_eq, tr_job_stability) %>% 
  left_join(tr_job_salary) %>% 
  filter(ano_censal %in% 2003:2018) %>% 
  group_by(rgn_id) %>% 
  mutate(prop_eq_hist = mean(prop_eq, na.rm = TRUE),
         stability_severance_hist = mean(stability_severance, na.rm = TRUE),
         stability_formalized_hist = mean(stability_formalized, na.rm = TRUE),
         maintained_salary_hist = mean(maintained_salary, na.rm = TRUE)) %>% 
  dplyr::select(rgn_id, prop_eq_hist, stability_severance_hist, stability_formalized_hist, maintained_salary_hist) %>% 
  distinct()

```


```{r}
tr_job_current_status <- left_join(tr_job_gender_eq, tr_job_stability) %>% 
  left_join(tr_job_salary) %>% 
  dplyr::select(-c(total_h000a:total_j600a)) %>% 
  left_join(tr_job_hist_avg, by = c("rgn_id")) %>% 
  filter(!rgn_id %in% "Region_2") %>%  # no data previous to 2023, so no reference available.
  group_by(rgn_id, ano_censal) %>% 
  mutate(
    prop_eq_cs_raw = (prop_eq / prop_eq_hist) * 100,
    stability_severance_cs_raw = (stability_severance / stability_severance_hist) * 100,
    stability_formalized_cs_raw = (stability_formalized / stability_formalized_hist) * 100,
    maintained_salary_cs_raw = (maintained_salary / maintained_salary_hist) * 100
  ) %>% 
  mutate(
    prop_eq_cs = ifelse(prop_eq_cs_raw > 100, 100, prop_eq_cs_raw),
    prop_eq_capped = ifelse(prop_eq > 1, 1, prop_eq), # so that we can use just the ratio between formalized women and men, not comparing to hist avg
    prop_eq_capped = prop_eq_capped * 100,
    stability_severance_cs = ifelse(stability_severance_cs_raw > 100, 100, stability_severance_cs_raw),
    stability_formalized_cs = ifelse(stability_formalized_cs_raw > 100, 100, stability_formalized_cs_raw),
    maintained_salary_cs = ifelse(maintained_salary_cs_raw > 100, 100, maintained_salary_cs_raw)
  ) %>% 
  ungroup()
```

##### plotlys

Change for the `_raw` version of the current status to see the uncapped versions.

```{r}
# prop_eq_cs
prop_eq_cs_plotly <- plotly::plot_ly(
  tr_job_current_status, 
  x = ~ano_censal, 
  y = ~prop_eq_cs, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "RAW CURRENT STATUS of gender equality (% formalized women vs men), reference of average value 2003-2018",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "prop_eq / avg_prop_eq_2003_2018")
  ) %>%
  add_lines(
    x = range(tr_job_current_status$ano_censal), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
prop_eq_cs_plotly

# stability_severance_cs
stability_severance_cs_plotly <- plotly::plot_ly(
  tr_job_current_status, 
  x = ~ano_censal, 
  y = ~stability_severance_cs, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "RAW CURRENT STATUS of job stability (severance pay per total jobs), reference of average value 2003-2018",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "stability_severance / avg_stability_severance_2003_2018")
  ) %>%
  add_lines(
    x = range(tr_job_current_status$ano_censal), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
stability_severance_cs_plotly

# stability_formalized_cs
stability_formalized_cs_plotly <- plotly::plot_ly(
  tr_job_current_status, 
  x = ~ano_censal, 
  y = ~stability_formalized_cs, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "RAW CURRENT STATUS of job stability (% paid/formalized workers), reference of average value 2003-2018",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "stability_formalized / avg_stability_formalized_2003_2018")
  ) %>%
  add_lines(
    x = range(tr_job_current_status$ano_censal), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
stability_formalized_cs_plotly

# maintained_salary_cs
maintained_salary_cs_plotly <- plotly::plot_ly(
  tr_job_current_status, 
  x = ~ano_censal, 
  y = ~maintained_salary_cs, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "RAW CURRENT STATUS of maintained salary (total compensation per paid worker), reference of average value 2003-2018",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "maintained_salary / avg_maintained_salary_2003_2018")
  ) %>%
  add_lines(
    x = range(tr_job_current_status$ano_censal), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
maintained_salary_cs_plotly
```


### Merged Current Status

For now, since we do *not* have the number of individuals who were dismissed, retired, or terminated, the `stability_severance` metric does not make sense and is not accurate.  This is because this is the definition of J600A:

"Payments for compensation or severance pay. This refers to the monetary compensation granted to workers for dismissal, retirement, or termination of their employment contract with the economic unit."

If we scale by the number of jobs, it is possible that there are a high number of jobs for a certain year but a low number of people that were dismissed, and so the amount given for severance was actually fairly high per capita.  But, since we do not have the number of individuals who are dismissed, we are not able to track that information. 

Therefore, for now I will average `maintained_salary_cs`, `stability_formalized_cs`, and `prop_eq` (*NOTE* that I am just using the raw comparison of the ratio of formalized workers of women / men, because the goal is for the % of formalized female workers to be >= the % of formalized male workers.  If we want to have an ideal state of "growth towards that goal" then we can use the historical avg, `prop_eq_cs`).

```{r}
tr_job_current_status_merge <- tr_job_current_status %>% 
  group_by(rgn_id, ano_censal) %>% 
  mutate(tr_job_quality_cs_raw = mean(c(prop_eq_capped, stability_formalized_cs, maintained_salary_cs))) %>% 
  ungroup()
```

### Interpolate between years

```{r}
library(zoo)

tr_job_current_status_merge_interpolated <- tr_job_current_status_merge %>% 
  complete(rgn_id, ano_censal = 2003:2025) %>% 
  mutate(tr_job_quality_cs_raw = as.numeric(tr_job_quality_cs_raw)) %>% 
  arrange(rgn_id, ano_censal) %>%
  group_by(rgn_id) %>% 
  mutate(
    score_interpolated = na.approx(
      tr_job_quality_cs_raw,
      x = ano_censal,
      rule = 2,
      na.rm = FALSE
    ),
    is_gapfilled = ifelse(is.na(tr_job_quality_cs_raw), 1, 0)
  ) %>%
  ungroup()
```


##### Write to Mazu

```{r}
write_csv(tr_job_current_status_merge, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","tr_job_quality_sustainability_current_status_merged_032326.csv"))
```

```{r}
write_csv(tr_job_current_status_merge_interpolated, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","tr_job_quality_sustainability_current_status_merged_interpolated_032326.csv"))
```







