---
title: 'OHI GoC 2025: QUALITY Tourism & Recreation'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: inline
editor_options: 
  chunk_output_type: inline
---

# TR Sustainability

## Distintivo S

```{r}
library(tidyverse)
library(here)
library(janitor)
```

```{r}
## read the pdfs in
dist_S_2015_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/distintivo_S/tabulated_distintivo_S", "tabula-Directorio_Nacional_de_Empresas_DISTINTIVO_S_JUN_2015_VF.csv"), skip = 1)

dist_S_2017_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/distintivo_S/tabulated_distintivo_S", "tabula-EMPRESAS_DISTINTIVO_S_JULIO_2017.csv"), skip = 1)

dist_S_2023_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/distintivo_S/tabulated_distintivo_S", "tabula-Directorio_de_empresas_vigentes_Distintivo_S.csv"))
```

### Clean and join

```{r}
## 2015
dist_S_2015 <- dist_S_2015_raw %>% 
  janitor::clean_names()

colnames(dist_S_2015) 
# [1] "no"                             "nombre_comercial_de_la_empresa" "estado"                        
# [4] "municipio_localidad"            "sitio_web" 

## 2017
dist_S_2017 <- dist_S_2017_raw %>% 
  janitor::clean_names()

colnames(dist_S_2017) 
# [1] "no"                             "nombre_comercial_de_la_empresa" "estado"                        
# [4] "localidad"

## 2023
dist_S_2023 <- dist_S_2023_raw %>% 
  janitor::clean_names()

colnames(dist_S_2023) 
# [1] "no"                   "folio_del_distintivo" "nombre_comercial"     "ciudad"               "estado" 
```

Okay, it appears that the "no" column is not very helpful, because the hotels do not maintain their number over time. 

The distinctive folio number is only available in 2023. 

Let's prep them for join.

```{r}
## 2015
dist_S_2015_join <- dist_S_2015 %>% 
  rename(nombre_comercial = nombre_comercial_de_la_empresa,
         localidad = municipio_localidad) %>% 
  mutate(year_documented = "2015") %>% 
  select(-no)

## 2017
dist_S_2017_join <- dist_S_2017 %>% 
  rename(nombre_comercial = nombre_comercial_de_la_empresa) %>% 
  mutate(year_documented = "2017") %>% 
  select(-no)

## 2023
dist_S_2023_join <- dist_S_2023 %>% 
  rename(localidad = ciudad) %>% 
  mutate(year_documented = "2023") %>% 
  select(-no)

```

```{r}
## join them
dist_S_all <- bind_rows(dist_S_2015_join,
                        dist_S_2017_join,
                        dist_S_2023_join)
```

### Geocode the hotels

```{r}
dist_S_goc_estados <- dist_S_all %>% 
  filter(estado %in% c("Nayarit","Jalisco","Sonora","Baja California Sur","Baja  Calfornia Sur", "Baja California  Sur","Baja  California Sur","Sinaloa")) %>%  ## note: no baja california.
  mutate(country = "Mexico")

dist_S_goc_estados_geocode <- dist_S_goc_estados %>% 
  tidygeocoder::geocode(city = localidad, country = country, state = estado) # to give each city a lat, long
```

Fix all lat/longs that tidygeocoder was not able to do.

```{r}
na_dist_S_goc_estados_geocode <- dist_S_goc_estados_geocode %>% 
  filter(is.na(lat))

unique(na_dist_S_goc_estados_geocode$nombre_comercial)

# [1] "Palladium Vallarta" 20.7153, -105.2464    
# [2] "Campo de Golf Litibú" https://www.rivieranayarit.com/activity/golf/litibu/  20.797847, -105.483116       
# [3] "Hotel Sheraton Hacienda del Mar Resort & SPA" 22.914233, -109.837037 
# [4] "Grand Velas Riviera Nayarit" 20.707216, -105.301422                 
# [5] "IBEROSTAR SELECTION PLAYA MITA" 20.790676, -105.489015
# [6] Gran Paladium Costa Mujeres Resort Nayarit 20.753902, -105.455937

dist_S_goc_estados_geocode_fixed <- dist_S_goc_estados_geocode %>% 
  mutate(
    lat = case_when(
      is.na(lat) & nombre_comercial == "Palladium Vallarta" ~ 20.7153,
      is.na(lat) & nombre_comercial == "Campo de Golf Litibú" ~ 20.797847,
      is.na(lat) & nombre_comercial == "Hotel Sheraton Hacienda del Mar Resort & SPA" ~ 22.914233,
      is.na(lat) & nombre_comercial == "Grand Velas Riviera Nayarit" ~ 20.707216,
      is.na(lat) & nombre_comercial == "IBEROSTAR SELECTION PLAYA MITA" ~ 20.790676,
      is.na(lat) & nombre_comercial == "Gran Paladium Costa Mujeres Resort" ~ 20.790676,
      TRUE ~ lat
    ),
    long = case_when(
      is.na(long) & nombre_comercial == "Palladium Vallarta" ~ -105.2464,
      is.na(long) & nombre_comercial == "Campo de Golf Litibú" ~ -105.483116,
      is.na(long) & nombre_comercial == "Hotel Sheraton Hacienda del Mar Resort & SPA" ~ -109.837037,
      is.na(long) & nombre_comercial == "Grand Velas Riviera Nayarit" ~ -105.301422,
      is.na(long) & nombre_comercial == "IBEROSTAR SELECTION PLAYA MITA" ~ -105.489015,
      is.na(long) & nombre_comercial == "Gran Paladium Costa Mujeres Resort" ~ -105.455937,
      TRUE ~ long
    )
  )
```


```{r}
dist_S_geocode_sf <- dist_S_goc_estados_geocode_fixed %>% 
  st_as_sf(coords = c("long", "lat"),
           crs = 4326, # WGS84
           remove = FALSE)

mapview(dist_S_geocode_sf, zcol = "year_documented")

```


### Number of certifications over time

```{r}
dist_S_presence <- dist_S_geocode_sf %>% 
  distinct(nombre_comercial, year_documented) %>% 
  mutate(present = 1) %>% 
  tidyr::pivot_wider(
    names_from = year_documented,
    values_from = present,
    values_fill = 0
  )

heat_df <- dist_S_presence %>% 
  arrange(nombre_comercial) %>% 
  mutate(id = row_number()) %>% 
  pivot_longer(cols = c(`2015`,`2017`,`2023`),
               names_to = "year",
               values_to = "present")

plot_ly(
  heat_df,
  x = ~year,
  y = ~nombre_comercial,
  z = ~present,
  type = "heatmap",
  colorscale = "Blues"
) %>% 
  layout(title = "Presence of Distintivo S Certification Over Time",
         xaxis = list(title = "Year"),
         yaxis = list(title = "Business"))
```

#### Standardize hotel names

```{r}
library(stringi)

dist_S_all_clean <- dist_S_geocode_sf %>% 
  mutate(
    nombre_comercial_clean = nombre_comercial %>% 
      str_to_lower() %>% 
      str_squish() %>% 
      stri_trans_general("Latin-ASCII") %>% # remove accents
      str_replace_all("[^a-z0-9 ]", "") %>% # remove punctuation
      str_replace_all("\\s+", " ") # normalize spaces
  )

dist_S_all_clean_2 <- dist_S_all_clean %>% 
  mutate(
    nombre_comercial_clean = nombre_comercial_clean %>% 
      str_replace_all("\\b(hotel|resort|spa|suites|suite|all inclusive|the|grupo)\\b", "") %>% 
      str_squish()
  ) %>% 
  arrange(desc(nombre_comercial_clean)) %>% 
  group_by(nombre_comercial_clean) %>% 
  fill(folio_del_distintivo, .direction = "up") %>% 
  fill(sitio_web, .direction = "down") ## note that pueblo bonito mazatlan and emerald bay are two separate resorts.
```


### Current status score


#### Classify OHI Rgn ID

"https://www.gob.mx/presidencia/prensa/el-objetivo-para-turismo-en-el-plan-mexico-es-ser-el-5to-pais-mas-visitado-del-mundo-para-el-2030-presidenta-claudia-sheinbaum?idiom=es"

To achieve the goal of being the fifth most visited country in the world by 2030, it is estimated that it is necessary to increase the number of tourists by three million each year, as well as increase:
 

The economic impact of international tourists by 46 percent.
Hotel rooms by 12 percent.
The generation of tourism jobs by 27 percent.
The tourism Gross Domestic Product (GDP) by 9 percent.
Domestic tourism by 9.8 percent.
*The number of service providers certified in sustainability by 31.5 percent.*
The average expenditure per international tourist by 3.3 percent.
The arrival of international visitors by 30 percent
The increase in DNR revenue by 40.5 percent.

```{r}
library(sf)
library(mapview)

## first, join with OHI GoC Regions

inland_10 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_10km_buffer","inland_10km_goc.shp"))
dist_S_all_clean_2_trans <- st_transform(dist_S_all_clean_2, st_crs(inland_10))
dist_S_goc_rgn <- st_intersection(dist_S_all_clean_2_trans, inland_10)

mapview(dist_S_goc_rgn, zcol = "rgn_id") + mapview(inland_10, zcol = "rgn_id")

st_write(dist_S_goc_rgn, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int","distintivo_S_certified_10km.shp")) ## NOTE: we lose a hotel in culiacan which would fit into the 50km buffer if we so chose to do so. 
```

#### Counts by rgn

```{r}
dist_S_counts_by_rgn <- dist_S_goc_rgn %>% 
  st_drop_geometry() %>% 
  distinct(rgn_id, nombre_comercial_clean, year_documented) %>% 
  group_by(rgn_id, year_documented) %>% 
  summarise(n_hotels = n(), .groups = "drop")
```


#### plotly over time

```{r}
dist_S_counts_by_rgn <- dist_S_counts_by_rgn %>% 
  mutate(year = as.numeric(year_documented))

library(plotly)

plot_ly(
  dist_S_counts_by_rgn,
  x = ~year,
  y = ~n_hotels,
  color = ~factor(rgn_id),
  type = 'scatter',
  mode = 'lines+markers'
) %>% 
  layout(
    title = "Distintivo S Certified Hotels by GoC Region (2015–2023)",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Number of Certified Hotels"),
    legend = list(title = list(text = "Region ID"))
  )

```


## Compare with DENUE Accommodations

```{r}
alojamientos_2019_2024_10km <- st_read(here::here("/home/shares/ohi/OHI_GOC/_raw_data/inegi_denue/d2025/servicios_alojamiento_temporal_y_preparación_alimentos_y_bebidas/int","alojamientos_2019_2024_goc_10km.shp"))

library(stringi)

alojamientos_2019_2024_10km_clean <- alojamientos_2019_2024_10km %>% 
  mutate(
    nombre_comercial_clean = nom_estab %>% 
      str_to_lower() %>% 
      str_squish() %>% 
      stri_trans_general("Latin-ASCII") %>% # remove accents
      str_replace_all("[^a-z0-9 ]", "") %>% # remove punctuation
      str_replace_all("\\s+", " ") %>%  # normalize spaces
      str_replace_all("\\b(hotel|resort|spa|suites|suite|all inclusive|the|grupo)\\b", "") %>% 
      str_squish()
  )

all_hotels <- unique(alojamientos_2019_2024_10km_clean$nombre_comercial_clean)


```

```{r}
dist_S_goc_10km <- st_read(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int","distintivo_S_certified_10km.shp"))

dist_S_businesses <- unique(dist_S_goc_10km$nmbr_c_)
print(dist_S_businesses)
```

```{r}
setdiff(dist_S_businesses, all_hotels)
##those not present in y (all_hotels) but are present in x:


setdiff(all_hotels, dist_S_businesses)
```

# CONANP fees over time

```{r}
library(tidyverse)
library(here)
library(janitor)
```


Read in the fees over time for protected areas.  

This data is directly from CONANP: it was tabulated using tabula.  The original pdfs are located at `~/home/shares/ohi/OHI_GOC/_raw_data/CONANP/d2025`.  See this issue comment for more details: <https://github.com/OHI-GOC/OHI_GOC_issues/issues/13#issuecomment-3598113291>


```{r}
raw_ingresos_conanp <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/CONANP/d2025/tabula","tabula-RECAUDACION_INAI 340008300023725 2012-2024.csv"))

clean_ingresos_conanp <- raw_ingresos_conanp %>% 
  fill(region_area_natural_protegida, .direction = "down") %>% # come back to this and check that it makes sense.
  mutate(ent_fed_clean = case_when(
     ent_fed == "B C" ~ "baja_california",
     ent_fed == "B C S"~ "baja_california_sur", 
     ent_fed == "BC Y BCS" ~ "mix_baja", # all are pacific side <https://descubreanp.conanp.gob.mx/en/conanp/ANP?suri=77>
     ent_fed == "SON" ~ "sonora",
     ent_fed %in% c("SIN", "SIN.") ~ "sinaloa",  
     ent_fed == "NAY." ~ "nayarit",
     ent_fed == "MICH." ~ "michoacan",
     ent_fed == "JAL." ~ "jalisco",
     ent_fed == "JAL/COL"~"jalisco_colima", # too deep inland <https://descubreanp.conanp.gob.mx/es/conanp/ANP?suri=174>
     ent_fed == "ZAC."~ "zacatecas",
     ent_fed == "MEX/MICH" ~ "mex_michoacan",
     ent_fed == "COL." ~ "colima",
     TRUE ~ NA_character_
  )) %>% 
  mutate(region_area_natural_protegida = stringr::str_to_lower(region_area_natural_protegida),
         region_area_natural_protegida_clean = stringr::str_replace_all(region_area_natural_protegida, " ", "_"),
         region_area_natural_protegida_clean = stringr::str_replace_all(region_area_natural_protegida, "\\.", "")) %>% 
  select(region_area_natural_protegida_clean, ent_fed_clean, articulo, ingresos_2012:total)
```


```{r}
clean_ingresos_conanp_goc <- clean_ingresos_conanp %>% 
  filter(ent_fed_clean %in% c("baja_california_sur","baja_california","sonora","sinaloa","nayarit","jalisco")) %>% 
  filter(!region_area_natural_protegida_clean %in% c("islas de la bahia de chamela (islas la pajarera, cocinas, mamut, colorada, san pedro, san agustin, san andres y negrita, y los islotes los anegados, novillas, mosca y submarino)","islas de la bahia de chamela (islas la pajarera, cocinas, mamut,\ncolorada, san pedro, san agustin, san andres y negrita, y los islotes\nlos anegados, novillas, mosca y submarino)","la primavera","zicuiran infiernillo","sierra de manantlan","sierra de quila","chamela cuixmala","playa mismaloya","playa cuitzmala","playa el tecuan","playa teopa"))
## pacifico mexicano profundo: "no visitation due to the characteristics of the ANP" <https://descubreanp.conanp.gob.mx/swb/conanp/ANP?suri=119> ### playa mismaloya is just outside the ohi regions

all_possible_goc_protected_areas <- clean_ingresos_conanp_goc %>% 
  select(region_area_natural_protegida_clean, ent_fed_clean) %>% 
  distinct(region_area_natural_protegida_clean, ent_fed_clean) %>% 
  write_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int/conanp_int","all_possible_goc_protected_areas_OG.csv"))

###NOTE: there is an issue with Islas Marias!!! The numbers are incorrect and were not registered correctly in the tabulated data!!!!!
```

Read in the list of GoC protected areas (manually searched):

```{r}
library(tidygeocoder)
library(dplyr)

list_goc_protected_areas <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int/conanp_int","all_possible_goc_protected_areas_UPDATE_120825.csv")) %>% 
  filter(coastal_goc %in% c("yes", "almost")) %>% 
  mutate(country = "Mexico") 

# %>% 
#   tidygeocoder::geocode(city = region_area_natural_protegida_clean, country = country, state = ent_fed_clean)

list_goc_protected_areas <- read_csv(
  here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int/conanp_int",
             "all_possible_goc_protected_areas_UPDATE_120825.csv")
) %>%
  filter(coastal_goc %in% c("yes", "almost")) %>%
  mutate(country = "Mexico") 

vector_goc_protected_areas <- unique(list_goc_protected_areas$region_area_natural_protegida_clean)

# %>%
#   tidygeocoder::geocode(
#     address = region_area_natural_protegida_clean,
#     method = "osm",
#     lat = latitude,
#     long = longitude,
#     limit = 1
#   )
```

Read in polygons for anpmx:

```{r}
library(sf)
library(mapview)

anpmx <- st_read(here::here("/home/shares/ohi/OHI_GOC/_raw_data/CONANP/anpmx", "anpmx.shp"))
mapview(anpmx) # yup, these are the same as the pdf!!!
```

```{r}
normalize_name <- function(x) {
  x |>
    str_to_lower() |>                            # lowercase
    stringi::stri_trans_general("Latin-ASCII") |> # remove accents
    str_replace_all("[[:punct:]]", " ") |>        # remove punctuation
    str_replace_all("\\s+", " ") |>               # collapse whitespace
    str_trim()                                    # trim ends
}

anpmx_clean <- anpmx %>%
  janitor::clean_names() %>% 
  mutate(nombre_clean = normalize_name(nombre))

apmx_clean_goc <- anpmx_clean %>% 
  filter(nombre_clean %in% vector_goc_protected_areas)

inland_50 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer","inland_50km_goc.shp"))
marine_ohi <- st_read(here::here("spatial/marine_ohi","marine_ohi.shp")) 

st_crs(inland_50) == st_crs(marine_ohi) ## TRUE

land_and_water_ohi_rgns <- st_union(inland_50, marine_ohi)

land_and_water_ohi_rgns_union <- land_and_water_ohi_rgns %>% 
  select(geometry) %>% 
  st_union(.)



mapview(land_and_water_ohi_rgns, zcol = "rgn_id")

```

Limit to only GoC Protected areas:

```{r}
st_crs(anpmx_clean) == st_crs(inland_50)

ampmx_clean_transformed <- st_transform(anpmx_clean, st_crs(inland_50))
st_crs(ampmx_clean_transformed) == st_crs(inland_50)

anpmx_inland_goc <- st_intersection(ampmx_clean_transformed, inland_50)
mapview(anpmx_inland_goc)

anpmx_marine_goc <- st_intersection(ampmx_clean_transformed, marine_ohi)
mapview(anpmx_marine_goc)

anpmx_all_goc <- bind_rows(anpmx_inland_goc, anpmx_marine_goc) %>% 
  select(nombre_clean, geometry) %>% 
  group_by(nombre_clean) %>% 
  summarise(geometry = st_union(geometry), .groups = "drop")
  
mapview(anpmx_all_goc)

## now join with ohi regions again....
goc_ohi_rgns <- st_read(here::here("spatial/ohi_regions","goc_ohi_rgns.shp")) %>% 
  select(rgn_id, geometry)
mapview(goc_ohi_rgns)

anpmx_all_goc_rgns <- st_intersection(anpmx_all_goc, goc_ohi_rgns)
mapview(anpmx_all_goc_rgns, zcol = "rgn_id")
```


## Filter protected areas to those that are in the GoC


```{r}
list_goc_protected_areas <- read_csv(
  here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int/conanp_int",
             "all_possible_goc_protected_areas_UPDATE_120825.csv")
) %>%
  filter(coastal_goc %in% c("yes", "almost")) %>%
  mutate(country = "Mexico") 

vector_goc_protected_areas <- unique(list_goc_protected_areas$region_area_natural_protegida_clean)


pivot_longer_ingresos_conanp <- pivot_longer(
  data = clean_ingresos_conanp,
  cols = ingresos_2012:ingresos_2024,
  names_to = "name",
  values_to = "value"
) %>% 
  rename(total_2012_2024 = "total") %>% 
  mutate(year = str_replace(name, "ingresos_", "")) %>% 
  filter(region_area_natural_protegida_clean %in% vector_goc_protected_areas)
```

### Assign OHI Regions

Because some polygons for the protected areas span multiple regions, at first I thought I needed to figure out a proportion of fees to each area.  But each area is managed by a specific state, which fits into the OHI GoC regions.  

But, according to the legal provision governing fees (e.g. “Artículo 198” for marine/coastal protected areas, and “Artículo 198 A” for terrestrial ones): the income from those fees “se destinarán a la Comisión Nacional de Áreas Naturales Protegidas, para la conservación y aprovechamiento sustentable de las Áreas Naturales Protegidas.” <https://www.conanp.gob.mx/acciones/articulo_198.php>

And, some studies have shown that very little actually goes back into the management of the protected areas, but instead more to the admin of the Secretaria de Hacienda y Credito Publico (SHCP).  <https://nossamexico.com/cuidar-lo-que-importa-las-areas-naturales-protegidas-y-el-cobro-de-derechos-un-analisis-desde-el-noroeste-de-mexico/>

Regardless, we can still assume that the fees are collected at the state level and sent to the federal CONANP agency.  This still in some ways impacts the ability for the protected areas to be managed.  So, I will do a case_when() based on the state to assign regions.


```{r}
## assign ohi goc regions based on the state handling fee collections.
ohi_rgn_ingresos_conanp <- pivot_longer_ingresos_conanp %>% 
  mutate(rgn_id = case_when(
    ent_fed_clean == "baja_california" ~ "Region_2", # not Region_1 because if you look at the `anpmx_all_goc` we see that none of the protected areas in that area are actually managed, or have the federal entity, of Baja California.  Even the islas del golfo are managed by Guaymas.
    ent_fed_clean == "baja_california_sur" ~ "Region_4", # before el vizcaino because that is the only one in Region 3.
    region_area_natural_protegida_clean %in% c("nopolo") ~ "Region_4",
    region_area_natural_protegida_clean %in% c("el vizcaino") ~ "Region_3",
    region_area_natural_protegida_clean %in% c("alto golfo de california y delta del rio colorado", "el pinacate y gran desierto de altar") ~ "Region_5",
    region_area_natural_protegida_clean %in% c("isla de san pedro martir") ~ "Region_6",
    region_area_natural_protegida_clean %in% c("islas del golfo de california, guaymas") ~ "Region_7",
    ent_fed_clean == "sinaloa" ~ "Region_8",
    region_area_natural_protegida_clean %in% c("islas del golfo de california, zona sur, sinaloa","playa ceuta") ~ "Region_8",
    ent_fed_clean == "nayarit" ~ "Region_9",
    region_area_natural_protegida_clean %in% c("cuenca alimentadora del distrito nacional de riego 043 estado de", "isla isabel","islas marias","islas marietas") ~ "Region_9",
  )) %>% 
  filter(region_area_natural_protegida_clean != "sierra de alamos rio cuchujaqui")

## i personally think that sierra de alamos rio cuchujaqui is too far inland, but I know we have other things that far inland.... we can reassess later, but right now I am DROPPING it because it just looks much too removed from the Gulf (representative species etc) <https://descubreanp.conanp.gob.mx/es/conanp/ANP?suri=150>
```

## Sum by GoC Rgn

```{r}
sum_ingresos_conanp_goc <- ohi_rgn_ingresos_conanp %>% 
  group_by(year, rgn_id) %>% 
  summarize(annual_collection_pesos = sum(value, na.rm = TRUE))

```

Now, let's make the current status:

For each year, it will be the amount of fees collected divided by the average amount from years 2012 to 2022, as a way to determine if it has been stable.

(current year of fees collected) / (avg fees collected for that region from 2012 - 2022) * 100

Where it should be equal or more than that historical average to get a score of 100.

```{r}
## calc reference (historical) average per region, 2012-2022
ref_avg_ingresos <- sum_ingresos_conanp_goc %>% 
  filter(year >= 2012, year <= 2022) %>% 
  group_by(rgn_id) %>% 
  summarize(ref_avg_pesos = mean(annual_collection_pesos, na.rm = TRUE)) %>% 
  ungroup()

## join to full dataset
status_ingresos <- sum_ingresos_conanp_goc %>% 
  left_join(ref_avg_ingresos, by = "rgn_id") %>% 
  mutate(
    current_status_raw = (annual_collection_pesos / ref_avg_pesos) * 100,
    current_status = as.numeric(pmin(current_status_raw, 100)), # cap at 100
    year = as.numeric(year)
  ) %>% 
  ungroup() %>% 
  arrange(rgn_id, year)

write_csv(status_ingresos, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int/conanp_int","status_ingresos_120825.csv"))

```

Plotly of current status over time:

```{r}
library(plotly)

plot_ly(
  data = status_ingresos,
  x = ~year,
  y = ~current_status,
  split = ~rgn_id,
  color = ~rgn_id,
  type = "scatter", 
  mode = "lines"
) %>%
  layout(
    title = "CONANP fee collection status over time (relative to average fees from 2012-2022)",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Status (0 to 100)", range = c(0, 110)),
    legend = list(title = list(text = "OHI GoC Region"))
  ) %>% 
  add_lines(
    x = range(status_ingresos$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )

plot_ly(
  data = status_ingresos,
  x = ~year,
  y = ~current_status_raw,
  split = ~rgn_id,
  color = ~rgn_id,
  type = "scatter", 
  mode = "lines"
) %>%
  layout(
    title = "CONANP fee collection status over time (relative to average fees from 2012-2022)",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Status (0 to 100)", range = c(0, 1062.01)),
    legend = list(title = list(text = "OHI GoC Region"))
  ) %>% 
  add_lines(
    x = range(status_ingresos$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )

```



