---
title: 'OHI GoC 2025: QUALITY Tourism & Recreation'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: inline
editor_options: 
  chunk_output_type: inline
---

# TR Sustainability

## Distintivo S

```{r}
library(tidyverse)
library(here)
library(janitor)
```

```{r}
## read the pdfs in
dist_S_2015_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/distintivo_S/tabulated_distintivo_S", "tabula-Directorio_Nacional_de_Empresas_DISTINTIVO_S_JUN_2015_VF.csv"), skip = 1)

dist_S_2017_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/distintivo_S/tabulated_distintivo_S", "tabula-EMPRESAS_DISTINTIVO_S_JULIO_2017.csv"), skip = 1)

dist_S_2023_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/distintivo_S/tabulated_distintivo_S", "tabula-Directorio_de_empresas_vigentes_Distintivo_S.csv"))
```

### Clean and join

```{r}
## 2015
dist_S_2015 <- dist_S_2015_raw %>% 
  janitor::clean_names()

colnames(dist_S_2015) 
# [1] "no"                             "nombre_comercial_de_la_empresa" "estado"                        
# [4] "municipio_localidad"            "sitio_web" 

## 2017
dist_S_2017 <- dist_S_2017_raw %>% 
  janitor::clean_names()

colnames(dist_S_2017) 
# [1] "no"                             "nombre_comercial_de_la_empresa" "estado"                        
# [4] "localidad"

## 2023
dist_S_2023 <- dist_S_2023_raw %>% 
  janitor::clean_names()

colnames(dist_S_2023) 
# [1] "no"                   "folio_del_distintivo" "nombre_comercial"     "ciudad"               "estado" 
```

Okay, it appears that the "no" column is not very helpful, because the hotels do not maintain their number over time. 

The distinctive folio number is only available in 2023. 

Let's prep them for join.

```{r}
## 2015
dist_S_2015_join <- dist_S_2015 %>% 
  rename(nombre_comercial = nombre_comercial_de_la_empresa,
         localidad = municipio_localidad) %>% 
  mutate(year_documented = "2015") %>% 
  select(-no)

## 2017
dist_S_2017_join <- dist_S_2017 %>% 
  rename(nombre_comercial = nombre_comercial_de_la_empresa) %>% 
  mutate(year_documented = "2017") %>% 
  select(-no)

## 2023
dist_S_2023_join <- dist_S_2023 %>% 
  rename(localidad = ciudad) %>% 
  mutate(year_documented = "2023") %>% 
  select(-no)

```

```{r}
## join them
dist_S_all <- bind_rows(dist_S_2015_join,
                        dist_S_2017_join,
                        dist_S_2023_join)
```

### Geocode the hotels

```{r}
dist_S_goc_estados <- dist_S_all %>% 
  filter(estado %in% c("Nayarit","Jalisco","Sonora","Baja California Sur","Baja  Calfornia Sur", "Baja California  Sur","Baja  California Sur","Sinaloa")) %>%  ## note: no baja california.
  mutate(country = "Mexico")

dist_S_goc_estados_geocode <- dist_S_goc_estados %>% 
  tidygeocoder::geocode(city = localidad, country = country, state = estado) # to give each city a lat, long
```

Fix all lat/longs that tidygeocoder was not able to do.

```{r}
na_dist_S_goc_estados_geocode <- dist_S_goc_estados_geocode %>% 
  filter(is.na(lat))

unique(na_dist_S_goc_estados_geocode$nombre_comercial)

# [1] "Palladium Vallarta" 20.7153, -105.2464    
# [2] "Campo de Golf Litibú" https://www.rivieranayarit.com/activity/golf/litibu/  20.797847, -105.483116       
# [3] "Hotel Sheraton Hacienda del Mar Resort & SPA" 22.914233, -109.837037 
# [4] "Grand Velas Riviera Nayarit" 20.707216, -105.301422                 
# [5] "IBEROSTAR SELECTION PLAYA MITA" 20.790676, -105.489015
# [6] Gran Paladium Costa Mujeres Resort Nayarit 20.753902, -105.455937

dist_S_goc_estados_geocode_fixed <- dist_S_goc_estados_geocode %>% 
  mutate(
    lat = case_when(
      is.na(lat) & nombre_comercial == "Palladium Vallarta" ~ 20.7153,
      is.na(lat) & nombre_comercial == "Campo de Golf Litibú" ~ 20.797847,
      is.na(lat) & nombre_comercial == "Hotel Sheraton Hacienda del Mar Resort & SPA" ~ 22.914233,
      is.na(lat) & nombre_comercial == "Grand Velas Riviera Nayarit" ~ 20.707216,
      is.na(lat) & nombre_comercial == "IBEROSTAR SELECTION PLAYA MITA" ~ 20.790676,
      is.na(lat) & nombre_comercial == "Gran Paladium Costa Mujeres Resort" ~ 20.790676,
      TRUE ~ lat
    ),
    long = case_when(
      is.na(long) & nombre_comercial == "Palladium Vallarta" ~ -105.2464,
      is.na(long) & nombre_comercial == "Campo de Golf Litibú" ~ -105.483116,
      is.na(long) & nombre_comercial == "Hotel Sheraton Hacienda del Mar Resort & SPA" ~ -109.837037,
      is.na(long) & nombre_comercial == "Grand Velas Riviera Nayarit" ~ -105.301422,
      is.na(long) & nombre_comercial == "IBEROSTAR SELECTION PLAYA MITA" ~ -105.489015,
      is.na(long) & nombre_comercial == "Gran Paladium Costa Mujeres Resort" ~ -105.455937,
      TRUE ~ long
    )
  )
```


```{r}
dist_S_geocode_sf <- dist_S_goc_estados_geocode_fixed %>% 
  st_as_sf(coords = c("long", "lat"),
           crs = 4326, # WGS84
           remove = FALSE)

mapview(dist_S_geocode_sf, zcol = "year_documented")

```


### Number of certifications over time

```{r}
dist_S_presence <- dist_S_geocode_sf %>% 
  distinct(nombre_comercial, year_documented) %>% 
  mutate(present = 1) %>% 
  tidyr::pivot_wider(
    names_from = year_documented,
    values_from = present,
    values_fill = 0
  )

heat_df <- dist_S_presence %>% 
  arrange(nombre_comercial) %>% 
  mutate(id = row_number()) %>% 
  pivot_longer(cols = c(`2015`,`2017`,`2023`),
               names_to = "year",
               values_to = "present")

plot_ly(
  heat_df,
  x = ~year,
  y = ~nombre_comercial,
  z = ~present,
  type = "heatmap",
  colorscale = "Blues"
) %>% 
  layout(title = "Presence of Distintivo S Certification Over Time",
         xaxis = list(title = "Year"),
         yaxis = list(title = "Business"))
```

#### Standardize hotel names

```{r}
library(stringi)

dist_S_all_clean <- dist_S_geocode_sf %>% 
  mutate(
    nombre_comercial_clean = nombre_comercial %>% 
      str_to_lower() %>% 
      str_squish() %>% 
      stri_trans_general("Latin-ASCII") %>%   # remove accents
      str_replace_all("[^a-z0-9 ]", "") %>%  # remove punctuation
      str_replace_all("\\s+", " ")           # normalize spaces
  )

dist_S_all_clean_2 <- dist_S_all_clean %>% 
  mutate(
    nombre_comercial_clean = nombre_comercial_clean %>% 
      str_replace_all("\\b(hotel|resort|spa|suites|suite|all inclusive|the|grupo)\\b", "") %>% 
      str_squish()
  ) %>% 
  arrange(desc(nombre_comercial_clean)) %>% 
  group_by(nombre_comercial_clean) %>% 
  fill(folio_del_distintivo, .direction = "up") %>% 
  fill(sitio_web, .direction = "down") ## note that pueblo bonito mazatlan and emerald bay are two separate resorts.
```


### Current status score


#### Classify OHI Rgn ID

"https://www.gob.mx/presidencia/prensa/el-objetivo-para-turismo-en-el-plan-mexico-es-ser-el-5to-pais-mas-visitado-del-mundo-para-el-2030-presidenta-claudia-sheinbaum?idiom=es"

To achieve the goal of being the fifth most visited country in the world by 2030, it is estimated that it is necessary to increase the number of tourists by three million each year, as well as increase:
 

The economic impact of international tourists by 46 percent.
Hotel rooms by 12 percent.
The generation of tourism jobs by 27 percent.
The tourism Gross Domestic Product (GDP) by 9 percent.
Domestic tourism by 9.8 percent.
*The number of service providers certified in sustainability by 31.5 percent.*
The average expenditure per international tourist by 3.3 percent.
The arrival of international visitors by 30 percent
The increase in DNR revenue by 40.5 percent.

```{r}
library(sf)
library(mapview)

## first, join with OHI GoC Regions

inland_10 <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/inland_10km_buffer","inland_10km_goc.shp"))
dist_S_all_clean_2_trans <- st_transform(dist_S_all_clean_2, st_crs(inland_10))
dist_S_goc_rgn <- st_intersection(dist_S_all_clean_2_trans, inland_10)

mapview(dist_S_goc_rgn, zcol = "rgn_id") + mapview(inland_10, zcol = "rgn_id")

st_write(dist_S_goc_rgn, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int","distintivo_S_certified_10km.shp")) ## NOTE: we lose a hotel in culiacan which would fit into the 50km buffer if we so chose to do so. 
```

#### Counts by rgn

```{r}
dist_S_counts_by_rgn <- dist_S_goc_rgn %>% 
  st_drop_geometry() %>% 
  distinct(rgn_id, nombre_comercial_clean, year_documented) %>% 
  group_by(rgn_id, year_documented) %>% 
  summarise(n_hotels = n(), .groups = "drop")
```


#### plotly over time

```{r}
dist_S_counts_by_rgn <- dist_S_counts_by_rgn %>% 
  mutate(year = as.numeric(year_documented))

library(plotly)

plot_ly(
  dist_S_counts_by_rgn,
  x = ~year,
  y = ~n_hotels,
  color = ~factor(rgn_id),
  type = 'scatter',
  mode = 'lines+markers'
) %>% 
  layout(
    title = "Distintivo S Certified Hotels by GoC Region (2015–2023)",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Number of Certified Hotels"),
    legend = list(title = list(text = "Region ID"))
  )

```

