---
title: 'OHI GoC 2025: SUSTAINABILITY Tourism & Recreation'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# TR Sustainability

# CONANP Returned Fees

```{r}
pct_returned_tbl <- tibble::tribble(
  ~year, ~pct_returned,
  2024, 0.96,
  2023, 0.94,
  2022, 0.29,
  2021, 0.36,
  2020, 0.65,
  2019, 0.51,
  2018, 0.12,
  2017, 0.00,
  2016, 0.00,
  2015, 0.00,
  2014, 0.00,
  2013, 0.00,
  2012, 0.47,
  2011, 0.76,
  2010, 0.92,
  2009, 0.81,
  2008, 0.94,
  2007, 0.95
)

expanded_pct_returned <- expand_grid(
  rgn_id = paste0("Region_", 1:9),
  year = 2007:2024
) %>%
  left_join(pct_returned_tbl, by = "year") %>% 
  rename(score = pct_returned)

write_csv(expanded_pct_returned, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","conanp_fee_return_current_status_final_121625.csv"))

```


# Fed Budget

```{r}
fed_budget_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/sust_int/conanp_int/fed_budget_programas_importantes_2012_2026.csv"))

fed_budget_sust_tourism <- fed_budget_raw %>% 
  dplyr::select(year, v003_anp_management, s046_sustainable_conservation, v005_protection_restoration, presupuesto_total_mdp_2026)
```

```{r}
plot(fed_budget_sust_tourism$year, fed_budget_sust_tourism$s046_sustainable_conservation)
plot(fed_budget_sust_tourism$year, fed_budget_sust_tourism$v003_anp_management)
plot(fed_budget_sust_tourism$year, fed_budget_sust_tourism$v005_protection_restoration)
plot(fed_budget_sust_tourism$year, fed_budget_sust_tourism$presupuesto_total_mdp_2026)
```


I will use 2012 - 2018 as the historial reference period to determine if the budget is increasing, decreasing, or staying relatively constant over time.

## Find hist avg

```{r}
fed_budget_sust_tourism_long <- fed_budget_sust_tourism %>% 
  pivot_longer(
    cols = v003_anp_management:presupuesto_total_mdp_2026,
    names_to = "programas",
    values_to = "mdp_2026"
  )

w_hist_avg_fed_budget <- fed_budget_sust_tourism_long %>% 
  filter(year %in% 2012:2018) %>% 
  group_by(programas) %>% 
  summarize(
    hist_avg_mdp_2026 = mean(mdp_2026)
  ) %>% 
  ungroup()
```


## Current status

### Each program

```{r}
current_status_fed_budget <- left_join(fed_budget_sust_tourism_long, w_hist_avg_fed_budget) %>% 
  mutate(current_status_raw = (mdp_2026 / hist_avg_mdp_2026) * 100,
         current_status = ifelse(current_status_raw > 100, 100, current_status_raw)) 
```


#### plotly

```{r}
library(plotly)

fed_budget_plotly <- plotly::plot_ly(
  current_status_fed_budget, 
  x = ~year, 
  y = ~current_status_raw, 
  color = ~programas, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: CONANP budget of programs (millones de pesos 2026)",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(current_status_fed_budget$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
fed_budget_plotly
```


#### descripcion


*s046_sustainable_conservation*
Programa de Conservación para el Desarrollo Sostenible (PROCODES)


*v003_anp_management*
(antes E016-16) Conservación y Manejo de Áreas Naturales Protegidas

*v005_protection_restoration*
(antes U040-16) Programa para la Protección y Restauración de Ecosistemas y Especies en Riesgo (U025-16, del 2012 hasta el 2018)

*presupuesto_total_mdp_2026*
total federal budget to CONANP in millions of 2026 pesos.


### Merged

```{r}
# with presupuesto_total_mdp_2026
current_status_fed_budget_merged <- current_status_fed_budget %>% 
  group_by(year) %>% 
  mutate(current_status_all = mean(current_status)) %>% 
  ungroup()

# without presupuesto_total_mdp_2026
current_status_fed_budget_merged_wo_presupuesto <- current_status_fed_budget %>% 
  filter(!programas %in% "presupuesto_total_mdp_2026") %>% 
  group_by(year) %>% 
  mutate(current_status_all = mean(current_status)) %>% 
  ungroup()
```

```{r}
all_fed_budget_plotly <- plotly::plot_ly(
  current_status_fed_budget_merged_wo_presupuesto, 
  x = ~year, 
  y = ~current_status_all, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "CURRENT STATUS: CONANP budget of programs (millones de pesos 2026)",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(current_status_fed_budget_merged_wo_presupuesto$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
all_fed_budget_plotly
```

### Save to Mazu

```{r}
write_csv(current_status_fed_budget_merged_wo_presupuesto, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","conanp_federal_budget_current_status_final_032026.csv"))
```


### (incorrect) Alt methods

Instead of calculating CS for each program and averaging the CS across these programs, lets add them all together and find the CS of the overall added mdp per program per year.

This is incorrect because the programs are not equally weighted in importance, instead the program that has more money to it overall drives the current status score more. However, we want to reflect the fact that each of these programs will be maintained or increasing, not that the overall budget (regardless of where it goes) is improving.  Though that is important, it would apply more to the general resilience of protected areas, rather than "investment in maintained/enforced sustainable tourism and recreation". which is what this component of the score will reflect.

```{r}
# take the pivoted df
fed_budget_sust_tourism_long <- fed_budget_sust_tourism %>% 
  pivot_longer(
    cols = v003_anp_management:presupuesto_total_mdp_2026,
    names_to = "programas",
    values_to = "mdp_2026"
  )

# sum the budget to all three programs
fed_budget_sust_tourism_sum <- fed_budget_sust_tourism_long %>% 
  filter(!programas %in% "presupuesto_total_mdp_2026") %>% # remove so we can see the programs highlighted by Ana Figueroa
  group_by(year) %>% 
  summarize(
    total_mdp_de_2026 = sum(mdp_2026)
  ) %>% 
  ungroup()

## find the hist avg for the summed budget
hist_avg_fed_budget_sust_tourism_sum <- fed_budget_sust_tourism_sum %>% 
  mutate(hist_avg_mdp_de_2026 = mean(total_mdp_de_2026))

## current status
alt_current_status_fed_budget <- hist_avg_fed_budget_sust_tourism_sum %>% 
  mutate(current_status_raw = (total_mdp_de_2026 / hist_avg_mdp_de_2026) * 100,
         current_status = ifelse(current_status_raw > 100, 100, current_status_raw))

# plot and compare
alt_fed_budget_plotly <- plotly::plot_ly(
  alt_current_status_fed_budget, 
  x = ~year, 
  y = ~current_status, 
  type = "scatter", 
  mode = "lines+markers"
) %>% 
  layout(
    title = "alt CURRENT STATUS: CONANP budget of programs (millones de pesos 2026)",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Puntaje // Percent (out of 100)")
  ) %>%
  add_lines(
    x = range(alt_current_status_fed_budget$year), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
alt_fed_budget_plotly
```














