---
title: 'OHI GoC 2025: SUSTAINABILITY Tourism & Recreation'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# TR Sustainability

## Job quality in tourism

```{r}
library(tidyverse)
library(mapview)
library(sf)
library(here)
library(janitor)
```

### Read in LE data

```{r}
## bring in data cleaned from LE of the CE
ce_totals <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/ce_int","ce_cleaned_totals_per_yr_municipio_activity.csv"))
```


Now, limit to the tourism sectors.

```{r}
tr_ce_totals <- ce_totals %>% 
  filter(general_sector %in% "marine_tourism")

unique(tr_ce_totals$specific_sector)
# [1] "accommodations_food_drink" "marine_tourist_transport"  "travel_planners"          
# [4] "coastal_recreation"    

unique(tr_ce_totals$eng_sector_desc)
#  [1] "Hotels and motels, and similar"                                                                       
#  [2] "Food and alcoholic and non-alcoholic beverage preparation services; includes restaurants and similar."
#  [3] "Tourist water transport, includes 48721 and 487210."                                                  
#  [4] "Boarding houses, guest houses, and furnished apartments and houses with hotel services."              
#  [5] "Travel agencies."                                                                                     
#  [6] "Campgrounds and recreational lodges."                                                                 
#  [7] "Tourist marinas."                                                                                     
#  [8] "Other reservation services."                                                                          
#  [9] "Museums, historical sites, zoos, and similar."                                                        
# [10] "Organization of excursions and tourist packages for travel agencies (tour operators)."    

```

#### Adjust for inflation

This will be done using the package `priceR`.  Documentation is located here: <https://cran.r-project.org/web/packages/priceR/readme/README.html> 

"adjust_for_inflation() automatically converts between nominal and real dollars, or in/deflates prices from one year’s prices to another’s.  It works for 304 countries / areas (see them with all by running show_countries())."

I adjusted all values in pesos to constant 2023 pesos to remove the effects of inflation and ensure that differences across 2003, 2008, 2013, 2018, and 2023 represent real changes in economic activity rather than changes in the general price level. Using constant pesos preserves internal consistency for regional comparisons within Mexico without introducing unnecessary distortions from exchange rates or purchasing power parity conversions, which are only required for cross-country analysis.

```{r}
library(priceR)

tr_ce_totals_clean_years_adjusted <- tr_ce_totals %>% 
  mutate(
    j000a_adj = adjust_for_inflation(
      price = j000a,
      from_date = ano_censal,
      country = "Mexico",
      to_date = 2023
      ), 
    j400a_adj = adjust_for_inflation(
      price = j400a,
      from_date = ano_censal,
      country = "Mexico",
      to_date = 2023
      ),
    j600a_adj = adjust_for_inflation(
      price = j600a,
      from_date = ano_censal,
      country = "Mexico",
      to_date = 2023
      ),
    j300a_adj = adjust_for_inflation(
      price = j300a,
      from_date = ano_censal,
      country = "Mexico",
      to_date = 2023
      )
  ) %>% 
  dplyr::select(-c(j000a, j300a, j400a, j600a))
```


#### Adjustment for municipalities that cover both coastal and pacific sides, and sectors that are not strictly coastal

Region_4 and Region_3 have municipalities that span both the gulf and pacific sides of the peninsula, so we need to adjust to ensure those represent the number of jobs in the Gulf.  Also, some sectors such as tourism, are not explicitly marine-related.  So we only consider the "coastal" jobs in our analysis, assuming that if the tourism jobs are coastal, then they have some tie to the Gulf.  

Many people travel to see the Gulf specifically, and accommodations, food, drink, etc. support those visits.  And those visits not only support the local economy, but also create connections between people and the Gulf.

```{r}
inland_50_population_2020 <- read_csv(here::here("/home/shares/ohi/OHI_GOC/spatial/population_inland_50km","pop2020_by_region_vector_inland_50.csv"))

coastal_population_multipliers <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/le","coastal_adjustment_multiplier_2020.csv"))
```

```{r}
## sum by rgn
tr_total_jobs_per_rgn_and_activity <- tr_ce_totals_clean_years_adjusted %>% 
  group_by(ano_censal, rgn_id, specific_sector) %>% 
  summarize(
    
    # job metrics
    h000a_rgn = sum(h000a, na.rm = TRUE),
    h000b_rgn = sum(h000b, na.rm = TRUE),
    h000c_rgn = sum(h000c, na.rm = TRUE),
    h010a_rgn = sum(h010a, na.rm = TRUE),
    h010b_rgn = sum(h010b, na.rm = TRUE),
    h010c_rgn = sum(h010c, na.rm = TRUE),
    
    # remunerations metrics
    j000a_rgn = sum(j000a_adj, na.rm = TRUE),
    j300a_rgn = sum(j300a_adj, na.rm = TRUE),
    j400a_rgn = sum(j400a_adj, na.rm = TRUE),
    j600a_rgn = sum(j600a_adj, na.rm = TRUE)
    
  ) %>% 
  ungroup()
```


```{r}
tr_coastal_total_jobs_per_rgn_and_activity <- tr_total_jobs_per_rgn_and_activity %>% 
  full_join(coastal_population_multipliers, by = "rgn_id") %>% 
  mutate(
    
    # fix San Felipe
    # coastal_pop_adjustment = ifelse(rgn_id %in% "Region_1" & ano_censal %in% "2023", 1, coastal_pop_adjustment),
    
    # h000a
    first_adj_coastal_h000a = ifelse(rgn_id %in% c("Region_3","Region_4"), h000a_rgn * coastal_pop_adjustment, h000a_rgn),
    coastal_h000a = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), h000a_rgn * coastal_pop_adjustment, first_adj_coastal_h000a),
    
    # h000b
    first_adj_coastal_h000b = ifelse(rgn_id %in% c("Region_3","Region_4"), h000b_rgn * coastal_pop_adjustment, h000b_rgn),
    coastal_h000b = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), h000b_rgn * coastal_pop_adjustment, first_adj_coastal_h000b),
    
    # h000c
    first_adj_coastal_h000c = ifelse(rgn_id %in% c("Region_3","Region_4"), h000c_rgn * coastal_pop_adjustment, h000c_rgn),
    coastal_h000c = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), h000c_rgn * coastal_pop_adjustment, first_adj_coastal_h000c),
    
    # h010a
    first_adj_coastal_h010a = ifelse(rgn_id %in% c("Region_3","Region_4"), h010a_rgn * coastal_pop_adjustment, h010a_rgn),
    coastal_h010a = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), h010a_rgn * coastal_pop_adjustment, first_adj_coastal_h010a),
    
    # h010b
    first_adj_coastal_h010b = ifelse(rgn_id %in% c("Region_3","Region_4"), h010b_rgn * coastal_pop_adjustment, h010b_rgn),
    coastal_h010b = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), h010b_rgn * coastal_pop_adjustment, first_adj_coastal_h010b),
    
    # h010c
    first_adj_coastal_h010c = ifelse(rgn_id %in% c("Region_3","Region_4"), h010c_rgn * coastal_pop_adjustment, h010c_rgn),
    coastal_h010c = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), h010c_rgn * coastal_pop_adjustment, first_adj_coastal_h010c),
    
    # j000a
    first_adj_coastal_j000a = ifelse(rgn_id %in% c("Region_3","Region_4"), j000a_rgn * coastal_pop_adjustment, j000a_rgn),
    coastal_j000a = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), j000a_rgn * coastal_pop_adjustment, first_adj_coastal_j000a),
    
    # j300a
    first_adj_coastal_j300a = ifelse(rgn_id %in% c("Region_3","Region_4"), j300a_rgn * coastal_pop_adjustment, j300a_rgn),
    coastal_j300a = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), j300a_rgn * coastal_pop_adjustment, first_adj_coastal_j300a),
    
    # j400a
    first_adj_coastal_j400a = ifelse(rgn_id %in% c("Region_3","Region_4"), j400a_rgn * coastal_pop_adjustment, j400a_rgn),
    coastal_j400a = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), j400a_rgn * coastal_pop_adjustment, first_adj_coastal_j400a),
    
    # j600a
    first_adj_coastal_j600a = ifelse(rgn_id %in% c("Region_3","Region_4"), j600a_rgn * coastal_pop_adjustment, j600a_rgn),
    coastal_j600a = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), j600a_rgn * coastal_pop_adjustment, first_adj_coastal_j600a)
    
  ) %>% 
  dplyr::select(-c(starts_with("first_adj_")))
```

Aggregating the metrics for each region:
```{r}
total_per_rgn <- tr_coastal_total_jobs_per_rgn_and_activity %>% 
  group_by(ano_censal, rgn_id) %>% 
  summarize(
    
    # job metrics
    total_h000a = sum(coastal_h000a, na.rm = TRUE),
    total_h000b = sum(coastal_h000b, na.rm = TRUE),
    total_h000c = sum(coastal_h000c, na.rm = TRUE),
    total_h010a = sum(coastal_h010a, na.rm = TRUE),
    total_h010b = sum(coastal_h010b, na.rm = TRUE),
    total_h010c = sum(coastal_h010c, na.rm = TRUE),
    
    # remuneration metrics
    total_j000a = sum(coastal_j000a, na.rm = TRUE),
    total_j300a = sum(coastal_j300a, na.rm = TRUE),
    total_j400a = sum(coastal_j400a, na.rm = TRUE),
    total_j600a = sum(coastal_j600a, na.rm = TRUE)
    
  ) %>% 
  mutate(across(starts_with("total_"), round))
```


### Indicators of job quality

Maintained Salary
J000A / H010A to represent maintained/stable/increasing salary.

Maintained Stability
H010A / H000A = % of paid workers (essentially overall % formalized jobs in tourism sector)

Gender Equality
H010B/H000B  vs H010C/H000C  = % of paid workers men vs. women
Ideal state: women % formalization is at least the same or equal to the men's % formalization get a score of 100.

**NOTES:**

(1) Unfortunately, we likely cannot include `stability_severance` here, because we do not have a metric from the Economic Census that represents the number of people dissmissed, retired, or terminated during that year.  Scaling by the total number of jobs in tourism for that year would be inaccurate, because severance pay is only given to those who leave.  For example, if we have a severance pay of 100, and there are 4 people who were terminated, but 50 people who currently are employed, then the true severance-per-capita pay would be 25 pesos/person, but using the methods in this script, it would be 2 pesos/person. Even if we tried to normalize by the number of economic units, it would not be truly representative of the stability in each region, because one economic unit could contain 0 or 500 terminated employees. So, for now, since we do *not* have the number of individuals who were dismissed, retired, or terminated, the `stability_severance` metric does not make sense and is not accurate.  This is because this is the definition of J600A: "Payments for compensation or severance pay. This refers to the monetary compensation granted to workers for dismissal, retirement, or termination of their employment contract with the economic unit."  If we scale by the number of jobs, it is possible that there are a high number of jobs for a certain year but a low number of people that were dismissed, and so the amount given for severance was actually fairly high per capita.  But, since we do not have the number of individuals who are dismissed, we are not able to track that information.  

(2) We decided to include maintained salary, despite it not being included in the Livelihoods & Economies goal. I personally think that it would not be good to include it as a metric for job quality in the overall LE goal, since we do not want it interpreted as a "good" or "bad" individual salary, which it cannot represent since there is very likely a large skew in terms of what each individual employee in an economic unit is being paid.  For example, the CEO may be getting paid 60% of the overall J000A, and all the entry-level employees are getting paid a very small amount (1%) that is actually not of good quality, and they are each getting paid less than they would in another OHI GoC Region. And I think that is especially true in largely unformalized jobs. *However,* I think it could be a metric for TR Sustainability, as we have already narrowed the sector to marine tourism, which is mostly formalized sector. And perhaps we can emphasize that if there is a drop in the average total remunerations in a year without a drop in the number of employees, that it is very likely that the salaries of those jobs are dropping as well, which is certainly negative. 

Therefore, for now I will average `maintained_salary_cs`, `total_pct_paid_workers_cs`, and `prop_eq` (*NOTE* that I am just using the raw comparison of the ratio of formalized workers of women / men, because the goal is for the % of formalized female workers to be >= the % of formalized male workers).


#### Maintained Salary

J000A = Total compensation (in millions of pesos): This includes all regular and extraordinary payments and contributions, in cash and in kind, before any deductions, to compensate the work of the company’s employees, in the form of wages and salaries, social benefits, and profits distributed to staff, whether such payment is calculated based on a workday, the amount of work performed (piecework), or a base salary supplemented by commissions for sales or other activities. Includes: employer contributions to social security schemes; payments made to personnel on leave or temporary leave. Excludes: payments for severance or compensation; payments to third parties for the provision of employed personnel; payments consisting exclusively of commissions for personnel who did not receive a fixed salary or wage; payments of fees for professional services contracted on an infrequent basis.

```{r}
# J000A / H010A to represent maintained/stable/increasing salary.

tr_job_remunerations <- total_per_rgn %>% 
  mutate(maintained_salary = total_j000a / total_h010a) %>% 
  ungroup()
```


#### Maintained Stability

```{r}
# H010A / H000A = % of paid workers

tr_job_paid_workers <- total_per_rgn %>% 
  mutate(total_pct_paid_workers = total_h010a / total_h000a) %>% 
  ungroup()
```


#### Gender Equality

```{r}
# the % formalized workers in the tourism sector should be the same between men and women (so that of the women working, the same percentage or more of them are formalized when compared to the number of formalized men out of total men working)

tr_job_gender_eq <- total_per_rgn %>% 
  mutate(men_formalized = total_h010b / total_h000b,
         women_formalized = total_h010c / total_h000c,
         prop_eq = women_formalized / men_formalized) %>% # goal: prop_eq >= 1  
  ungroup()
```


##### metric plotlys

```{r}
# --- MAINTAINED SALARY ---

# total_j000a: Total compensation
total_j000a_plotly <- plotly::plot_ly(
  tr_job_remunerations,
  x = ~ano_censal,
  y = ~total_j000a,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines"
) %>%
  layout(
    title = "Total compensation (J000A, millions of pesos) per region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Total compensation (millions of pesos)")
  )
total_j000a_plotly

# total_h010a: Total paid workers
total_h010a_plotly <- plotly::plot_ly(
  tr_job_remunerations,
  x = ~ano_censal,
  y = ~total_h010a,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines"
) %>%
  layout(
    title = "Total paid/formalized workers (H010A) per region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Number of paid workers")
  )
total_h010a_plotly

# maintained_salary: J000A / H010A
maintained_salary_raw_plotly <- plotly::plot_ly(
  tr_job_remunerations,
  x = ~ano_censal,
  y = ~maintained_salary,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines"
) %>%
  layout(
    title = "Maintained salary (J000A / H010A): compensation per paid worker per region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Compensation per paid worker (millions of pesos)")
  )
maintained_salary_raw_plotly
```


```{r}
# --- MAINTAINED STABILITY ---

# total_h000a: Total workers
total_h000a_plotly <- plotly::plot_ly(
  tr_job_paid_workers,
  x = ~ano_censal,
  y = ~total_h000a,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines"
) %>%
  layout(
    title = "Total workers (H000A) per region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Total number of workers")
  )
total_h000a_plotly

# total_pct_paid_workers: H010A / H000A
total_pct_paid_workers_raw_plotly <- plotly::plot_ly(
  tr_job_paid_workers,
  x = ~ano_censal,
  y = ~total_pct_paid_workers,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines"
) %>%
  layout(
    title = "Job stability: % formalized/paid workers (H010A / H000A) per region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Proportion of paid workers")
  )
total_pct_paid_workers_raw_plotly
```


```{r}
# --- % FORMALIZED EQUALITY ACROSS GENDER---

# total_h000b: Total male workers
total_h000b_plotly <- plotly::plot_ly(
  tr_job_gender_eq,
  x = ~ano_censal,
  y = ~total_h000b,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines"
) %>%
  layout(
    title = "Total male workers (H000B) per region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Total male workers")
  )
total_h000b_plotly

# total_h000c: Total female workers
total_h000c_plotly <- plotly::plot_ly(
  tr_job_gender_eq,
  x = ~ano_censal,
  y = ~total_h000c,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines"
) %>%
  layout(
    title = "Total female workers (H000C) per region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Total female workers")
  )
total_h000c_plotly

# total_h010b: Total paid male workers
total_h010b_plotly <- plotly::plot_ly(
  tr_job_gender_eq,
  x = ~ano_censal,
  y = ~total_h010b,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines"
) %>%
  layout(
    title = "Total paid male workers (H010B) per region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Total paid male workers")
  )
total_h010b_plotly

# total_h010c: Total paid female workers
total_h010c_plotly <- plotly::plot_ly(
  tr_job_gender_eq,
  x = ~ano_censal,
  y = ~total_h010c,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines"
) %>%
  layout(
    title = "Total paid female workers (H010C) per region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Total paid female workers")
  )
total_h010c_plotly

# men_formalized: H010B / H000B
men_formalized_plotly <- plotly::plot_ly(
  tr_job_gender_eq,
  x = ~ano_censal,
  y = ~men_formalized,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines"
) %>%
  layout(
    title = "% formalized male workers (H010B / H000B) per region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Proportion of paid male workers")
  )
men_formalized_plotly

# women_formalized: H010C / H000C
women_formalized_plotly <- plotly::plot_ly(
  tr_job_gender_eq,
  x = ~ano_censal,
  y = ~women_formalized,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines"
) %>%
  layout(
    title = "% formalized female workers (H010C / H000C) per region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "Proportion of paid female workers")
  )
women_formalized_plotly

# prop_eq: women_formalized / men_formalized
prop_eq_raw_plotly <- plotly::plot_ly(
  tr_job_gender_eq,
  x = ~ano_censal,
  y = ~prop_eq,
  color = ~rgn_id,
  colors = rgn_colors,
  type = "scatter",
  mode = "lines"
) %>%
  layout(
    title = "Gender equality: ratio of formalized women to men workers (H010C/H000C vs H010B/H000B) per region",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "prop_eq (goal: >= 1)")
  ) %>%
  add_lines(
    x = range(tr_job_gender_eq$ano_censal),
    y = c(1, 1),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (1)",
    inherit = FALSE
  )
prop_eq_raw_plotly
```


### Metric CS

```{r}
tr_job_hist_avg <- left_join(tr_job_gender_eq, tr_job_paid_workers) %>% 
  left_join(tr_job_remunerations) %>% 
  filter(ano_censal %in% 2003:2018) %>% 
  group_by(rgn_id) %>% 
  mutate(
    
    total_pct_paid_workers_hist = mean(total_pct_paid_workers, na.rm = TRUE), # just to see, even though we will use a threshold to calculate current status
    maintained_salary_hist = mean(maintained_salary, na.rm = TRUE)
    
  ) %>% 
  dplyr::select(rgn_id, total_pct_paid_workers_hist, maintained_salary_hist) %>% 
  distinct()

```


```{r}
tr_job_current_status <- left_join(tr_job_gender_eq, tr_job_paid_workers) %>% 
  left_join(tr_job_remunerations) %>% 
  dplyr::select(-c(total_h000a:total_j600a)) %>% 
  left_join(tr_job_hist_avg, by = c("rgn_id")) %>% 
  mutate(formalization_ideal = 0.90) %>% 
  filter(!rgn_id %in% "Region_4") %>%  # no data previous to 2023, so no reference available.
  group_by(rgn_id, ano_censal) %>% 
  mutate(
    
    total_pct_paid_workers_cs_raw = (total_pct_paid_workers / formalization_ideal) * 100, # how close are they to 90% formalized
    total_pct_paid_workers_cs = ifelse(total_pct_paid_workers_cs_raw > 100, 100, total_pct_paid_workers_cs_raw),
    
    maintained_salary_cs_raw = (maintained_salary / maintained_salary_hist) * 100,
    maintained_salary_cs = ifelse(maintained_salary_cs_raw > 100, 100, maintained_salary_cs_raw),
    
    prop_eq_cs = ifelse(prop_eq > 1, 100, prop_eq * 100) # so that we can use just the ratio between formalized women and men, not comparing to hist avg
    
  ) %>% 
  ungroup()
```

##### cs plotlys

Change for the `_raw` version of the current status to see the uncapped versions.

```{r}
# prop_eq_cs
prop_eq_cs_plotly <- plotly::plot_ly(
  tr_job_current_status, 
  x = ~ano_censal, 
  y = ~prop_eq_cs, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "CURRENT STATUS of gender equality (% formalized women vs men), ideal is that they are equal (1:1)",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "% formalized women / % formalized men")
  ) %>%
  add_lines(
    x = range(tr_job_current_status$ano_censal), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
prop_eq_cs_plotly
```


```{r}
# total_pct_paid_workers_cs
total_pct_paid_workers_cs_plotly <- plotly::plot_ly(
  tr_job_current_status, 
  x = ~ano_censal, 
  y = ~total_pct_paid_workers_cs, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "CURRENT STATUS of job stability (% paid workers), in reference to an ideal of 90% jobs formalized",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "total_pct_paid_workers / ideal_pct_paid_workers")
  ) %>%
  add_lines(
    x = range(tr_job_current_status$ano_censal), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
total_pct_paid_workers_cs_plotly
```


```{r}
# maintained_salary_cs
maintained_salary_cs_plotly <- plotly::plot_ly(
  tr_job_current_status, 
  x = ~ano_censal, 
  y = ~maintained_salary_cs, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "CURRENT STATUS of maintained salary (total compensation per paid worker), reference of average value 2003-2018",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "maintained_salary / avg_maintained_salary_2003_2018")
  ) %>%
  add_lines(
    x = range(tr_job_current_status$ano_censal), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
maintained_salary_cs_plotly

# RAW maintained_salary_cs
raw_maintained_salary_cs_plotly <- plotly::plot_ly(
  tr_job_current_status, 
  x = ~ano_censal, 
  y = ~maintained_salary_cs_raw, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "RAW CURRENT STATUS of maintained salary (total compensation per paid worker), reference of average value 2003-2018",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "maintained_salary / avg_maintained_salary_2003_2018")
  ) %>%
  add_lines(
    x = range(tr_job_current_status$ano_censal), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
raw_maintained_salary_cs_plotly
```


### Merged CS

The simple average of all current status of each component of the TR Job Quality indicator, for one overall current status score each year and region. 

```{r}
tr_job_current_status_merge <- tr_job_current_status %>% 
  group_by(rgn_id, ano_censal) %>% 
  mutate(tr_job_quality_cs_raw = mean(c(prop_eq_cs, total_pct_paid_workers_cs, maintained_salary_cs))) %>% 
  ungroup()
```

### Interpolate between years

```{r}
library(zoo)

tr_job_current_status_merge_interpolated <- tr_job_current_status_merge %>% 
  complete(rgn_id, ano_censal = 2003:2025) %>% 
  mutate(tr_job_quality_cs_raw = as.numeric(tr_job_quality_cs_raw)) %>% 
  arrange(rgn_id, ano_censal) %>%
  group_by(rgn_id) %>% 
  mutate(
    score_interpolated = na.approx(
      tr_job_quality_cs_raw,
      x = ano_censal,
      rule = 2,
      na.rm = FALSE
    ),
    is_gapfilled = ifelse(is.na(tr_job_quality_cs_raw), 1, 0)
  ) %>%
  ungroup()
```


##### Write to Mazu

```{r}
write_csv(tr_job_current_status_merge, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","tr_job_quality_sustainability_current_status_merged_033026.csv"))
```

```{r}
write_csv(tr_job_current_status_merge_interpolated, here::here("/home/shares/ohi/OHI_GOC/goal_prep/tr/v2025/output_sustainability","tr_job_quality_sustainability_current_status_merged_interpolated_033026.csv"))
```








