---
title: "Sense of Place (SP): LSP choropleth maps"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    theme: cerulean
editor_options:
  chunk_output_type: console
---

# Purpose

Static choropleth maps of **Lasting Special Places (LSP)** scores by OHI Gulf of California region, plus **process maps** that illustrate how each LSP type (scripts 2–7) is scored.

- overall LSP current status (2025 snapshot)
- each LSP category current status (where a region has no score: solid white)
- method illustrations: special protected areas, dive sites, archaeology, fishing spots, special habitats, all other places

Choropleths follow the TR / Access style (states underlay, marine OHI regions, 0–100 scale, white for NA). Process maps use focused extents and rule-based symbology.

Outputs (PDF only) are written to:

`goal_prep/sp/lsp/v2025/viz` on Mazu, or `sp/lsp/v2025/viz` when using the local mirror.

```{r setup}
library(tidyverse)
library(sf)
library(here)
library(ggplot2)
library(grid)

# Prefer Mazu goal_prep paths; fall back to local workspace mirror
goc_root <- "/home/shares/ohi/OHI_GOC"
local_lsp_root <- here("sp/lsp/v2025")
mazu_lsp_root <- file.path(goc_root, "goal_prep/sp/lsp/v2025")

lsp_prep_root <- if (dir.exists(file.path(mazu_lsp_root, "output"))) {
  mazu_lsp_root
} else if (dir.exists(file.path(local_lsp_root, "output"))) {
  local_lsp_root
} else {
  mazu_lsp_root
}

lsp_output_dir <- file.path(lsp_prep_root, "output")
lsp_int_dir <- file.path(lsp_prep_root, "int")
viz_dir <- file.path(lsp_prep_root, "viz")
dir.create(viz_dir, recursive = TRUE, showWarnings = FALSE)

# first existing file among candidates
first_path <- function(...) {
  paths <- c(...)
  hit <- paths[file.exists(paths)]
  if (length(hit) == 0) NA_character_ else hit[[1]]
}

safe_read_csv <- function(path, ...) {
  if (is.na(path) || !file.exists(path)) return(NULL)
  tryCatch(read_csv(path, show_col_types = FALSE, ...), error = function(e) NULL)
}

safe_st_read <- function(path, ...) {
  if (is.na(path) || !file.exists(path)) return(NULL)
  tryCatch(st_read(path, quiet = TRUE, ...), error = function(e) NULL)
}

# TR / Access-style ramp (0 = dark red, 100 = dark blue)
lsp_palette <- c(
  "#8a0032", "#A50026", "#A50026", "#D73027", "#F46D43",
  "#FDAE61", "#FEE090", "#FEE090", "#ABD9E9", "#4575B4", "#313695"
)
lsp_na_fill <- "white"
lsp_legend_title <- "Current status score"

normalize_rgn_id <- function(x) {
  x %>%
    as.character() %>%
    str_squish() %>%
    str_replace_all(" ", "_")
}

rgn_labels <- c(
  "Region_1" = "Western Upper Gulf",
  "Region_2" = "Western Midriff Islands",
  "Region_3" = "Western Central Gulf",
  "Region_4" = "Western Southern Gulf",
  "Region_5" = "Eastern Upper Gulf",
  "Region_6" = "Eastern Midriff Islands",
  "Region_7" = "Southern Sonora Coast",
  "Region_8" = "Sinaloa Coast",
  "Region_9" = "Southeastern Gulf"
)
```

# Spatial layers

```{r spatial}
ohi_marine <- st_read(
  here("spatial/ohi_regions/goc_ohi_rgns.shp"),
  quiet = TRUE
) %>%
  filter(location == "marine") %>%
  mutate(rgn_id = normalize_rgn_id(rgn_id)) %>%
  dplyr::select(rgn_id, geometry)

states <- st_read(
  here("spatial/ohi_regions/states.shp"),
  quiet = TRUE
) %>%
  st_transform(st_crs(ohi_marine))
```

# Score tables

```{r scores}
read_rgn_condition <- function(path, value_col = "condition") {
  read_csv(path, show_col_types = FALSE) %>%
    mutate(
      rgn_id = normalize_rgn_id(rgn_id),
      score = suppressWarnings(as.numeric(.data[[value_col]]))
    ) %>%
    dplyr::select(rgn_id, score) %>%
    group_by(rgn_id) %>%
    summarize(score = mean(score, na.rm = TRUE), .groups = "drop")
}

# Prefer v3 condition filenames; fall back to older script output names
rgn_csv <- function(...) {
  path <- first_path(file.path(lsp_output_dir, c(...)))
  if (is.na(path)) {
    warning("Missing region CSV among: ", paste(c(...), collapse = ", "))
    return(tibble(rgn_id = character(), score = numeric()))
  }
  read_rgn_condition(path)
}

lsp_overall_path <- first_path(
  file.path(lsp_output_dir, "lsp_goal_current_status_by_region.csv"),
  file.path(lsp_output_dir, "lsp_current_status.csv")
)
lsp_overall <- read_csv(lsp_overall_path, show_col_types = FALSE) %>%
  mutate(
    rgn_id = normalize_rgn_id(rgn_id),
    score = suppressWarnings(as.numeric(current_status))
  ) %>%
  dplyr::select(rgn_id, score)

score_layers <- list(
  "lsp_overall_2025" = list(
    data = lsp_overall,
    title = "Lasting Special Places (LSP)\nCurrent status 2025"
  ),
  "lsp_special_protected_areas" = list(
    data = rgn_csv("condition_region_protected_areas.csv", "rgn_protected_area_condition.csv"),
    title = "LSP: Special Protected Areas"
  ),
  "lsp_priority_dive_sites" = list(
    data = rgn_csv("condition_region_dive_sites.csv", "rgn_dive_sites_condition.csv"),
    title = "LSP: Priority Dive Sites"
  ),
  "lsp_iconic_fishing_spots" = list(
    data = rgn_csv("condition_region_fishing_sites.csv", "rgn_fishing_sites_condition.csv"),
    title = "LSP: Iconic Fishing Spots"
  ),
  "lsp_archaeological_sites" = list(
    data = rgn_csv(
      "condition_region_archaeological_sites.csv",
      "rgn_ewg_archaeological_sites_condition.csv"
    ),
    title = "LSP: Archaeological Sites"
  ),
  "lsp_special_habitats" = list(
    data = rgn_csv("condition_region_special_habitats.csv", "rgn_special_habitats_condition.csv"),
    title = "LSP: Special Habitats"
  ),
  "lsp_all_other_special_places" = list(
    data = rgn_csv(
      "condition_region_other_marine_places.csv",
      "rgn_final_other_protections_condition.csv"
    ),
    title = "LSP: All other Special Places"
  )
)
```

# Choropleth helper

Join every score table to the full set of nine marine regions so missing categories stay NA (solid white), not dropped.

```{r map-helper}
join_scores_to_ohi <- function(score_df) {
  ohi_marine %>%
    left_join(score_df, by = "rgn_id") %>%
    mutate(
      rgn_label = unname(rgn_labels[rgn_id]),
      score = ifelse(is.nan(score), NA_real_, score)
    )
}

make_lsp_choropleth <- function(
    score_df,
    legend_title = lsp_legend_title,
    plot_title = NULL,
    label_scores = FALSE
) {
  dat <- join_scores_to_ohi(score_df)

  p <- ggplot() +
    geom_sf(
      data = states,
      fill = "#d6cac0",
      color = "grey10",
      linewidth = 0.1
    ) +
    geom_sf(
      data = dat,
      aes(fill = score),
      color = "black",
      linewidth = 0.2
    ) +
    scale_fill_gradientn(
      colours = lsp_palette,
      limits = c(0, 100),
      na.value = lsp_na_fill,
      name = "Score",
      breaks = seq(0, 100, by = 25)
    ) +
    labs(
      title = plot_title,
      subtitle = NULL,
      caption = NULL
    ) +
    coord_sf(expand = FALSE) +
    theme_void(base_size = 12) +
    theme(
      panel.background = element_rect(fill = "white", color = NA),
      plot.background = element_rect(fill = "white", color = NA),
      plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
      legend.position = "bottom",
      legend.direction = "horizontal",
      legend.key.height = unit(0.35, "cm"),
      legend.key.width = unit(1.2, "cm"),
      legend.title = element_text(size = 10),
      legend.text = element_text(size = 9),
      plot.subtitle = element_blank(),
      plot.caption = element_blank(),
      plot.margin = margin(6, 6, 6, 6)
    )

  if (label_scores) {
    p <- p +
      geom_sf_text(
        data = dat %>%
          mutate(label = ifelse(is.na(score), "NA", as.character(round(score)))),
        aes(label = label),
        size = 3,
        color = "grey10",
        fun.geometry = function(g) st_point_on_surface(st_make_valid(g))
      )
  }

  p
}

# PDF + PNG (PNG is what the website embeds from this folder on Mazu)
save_lsp_map <- function(gg, stem, width = 7.5, height = 7.5, dpi = 200) {
  if (is.null(gg) || !inherits(gg, "ggplot")) {
    message("Skip save (not a ggplot): ", stem)
    return(invisible(NULL))
  }
  pdf_path <- file.path(viz_dir, paste0(stem, ".pdf"))
  png_path <- file.path(viz_dir, paste0(stem, ".png"))
  tryCatch(
    ggsave(pdf_path, gg, width = width, height = height, device = cairo_pdf, bg = "white"),
    error = function(e) {
      ggsave(pdf_path, gg, width = width, height = height, device = "pdf", bg = "white")
    }
  )
  tryCatch(
    ggsave(png_path, gg, width = width, height = height, dpi = dpi, bg = "white"),
    error = function(e) message("PNG save failed for ", stem, ": ", conditionMessage(e))
  )
  message("Wrote PDF: ", pdf_path)
  if (file.exists(png_path)) message("Wrote PNG: ", png_path)
  invisible(list(pdf = pdf_path, png = if (file.exists(png_path)) png_path else NA_character_))
}
```

# Build, display, and save maps

```{r maps, fig.width=7.5, fig.height=7.5}
saved <- list()

for (stem in names(score_layers)) {
  layer <- score_layers[[stem]]
  gg <- make_lsp_choropleth(
    layer$data,
    legend_title = lsp_legend_title,
    plot_title = layer$title,
    label_scores = FALSE
  )
  print(gg)
  saved[[stem]] <- save_lsp_map(gg, stem)
}
```

# Multi-panel of category maps

```{r category-panel, fig.width=10, fig.height=10}
cat_long <- bind_rows(
  lapply(names(score_layers), function(stem) {
    if (stem == "lsp_overall_2025") return(NULL)
    score_layers[[stem]]$data %>%
      mutate(
        layer = stem,
        panel_title = score_layers[[stem]]$title
      )
  })
)

panel_sf <- ohi_marine %>%
  select(rgn_id) %>%
  st_drop_geometry() %>%
  tidyr::crossing(panel_title = unique(cat_long$panel_title)) %>%
  left_join(cat_long, by = c("rgn_id", "panel_title")) %>%
  left_join(ohi_marine, by = "rgn_id") %>%
  st_as_sf() %>%
  mutate(score = ifelse(is.nan(score), NA_real_, score))

gg_panel <- ggplot() +
  geom_sf(
    data = states,
    fill = "#d6cac0",
    color = "grey10",
    linewidth = 0.05
  ) +
  geom_sf(
    data = panel_sf,
    aes(fill = score),
    color = "black",
    linewidth = 0.15
  ) +
  scale_fill_gradientn(
    colours = lsp_palette,
    limits = c(0, 100),
    na.value = lsp_na_fill,
    name = "Score",
    breaks = seq(0, 100, by = 25)
  ) +
  facet_wrap(~ panel_title, ncol = 3) +
  labs(title = "LSP category scores", subtitle = NULL, caption = NULL) +
  coord_sf(expand = FALSE) +
  theme_void(base_size = 11) +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    strip.text = element_text(size = 9, face = "bold"),
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.title = element_text(size = 10),
    plot.title = element_text(face = "bold", size = 12, hjust = 0.5),
    plot.margin = margin(6, 6, 6, 6)
  )

gg_panel
save_lsp_map(gg_panel, "lsp_categories_panel", width = 10, height = 10)
```

# Process maps: scoring methods by LSP type

Static **method illustrations** for each LSP prep script (2–7). Review each object in the console with `print()`. Mid-script saves use `save_process_maps` (default FALSE). A final chunk exports **all** ggplot process objects as square PNG + PDF (title only) to the viz folder.

```{r process-helpers}
# --- toggle: leave FALSE while exploring; end-of-script PDF/PNG export still runs ---
save_process_maps <- FALSE

# Square exports for website / Word: title only, short legends, no subtitle or caption
viz_square <- 7.5

# Collect plot + intended file stem for the final export
process_plot_registry <- list()

process_theme <- function(base_size = 11) {
  theme_minimal(base_size = base_size) +
    theme(
      legend.position = "bottom",
      legend.direction = "horizontal",
      legend.box = "horizontal",
      legend.justification = "center",
      legend.title = element_blank(),
      legend.text = element_text(size = base_size - 1),
      legend.key.size = unit(0.45, "cm"),
      legend.margin = margin(t = 2, b = 0),
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = base_size - 2),
      axis.text.y = element_text(size = base_size - 2),
      plot.title = element_text(
        face = "bold", size = base_size + 1, hjust = 0.5,
        margin = margin(b = 6)
      ),
      plot.subtitle = element_blank(),
      plot.caption = element_blank(),
      plot.margin = margin(6, 6, 6, 6)
    )
}

# Drop leftover subtitle/caption; keep title. Use after building any process ggplot.
finish_process_plot <- function(gg, title = NULL) {
  if (is.null(gg) || !inherits(gg, "ggplot")) return(gg)
  if (!is.null(title)) {
    gg <- gg + labs(title = title, subtitle = NULL, caption = NULL)
  } else {
    gg <- gg + labs(subtitle = NULL, caption = NULL)
  }
  gg + process_theme()
}

register_process_plot <- function(gg, stem, width = viz_square, height = viz_square) {
  if (is.null(gg) || !inherits(gg, "ggplot")) return(invisible(NULL))
  process_plot_registry[[stem]] <<- list(gg = gg, width = width, height = height)
  invisible(gg)
}

show_process_map <- function(gg, stem, width = viz_square, height = viz_square) {
  if (is.null(gg)) {
    message("No plot object for ", stem)
    return(invisible(NULL))
  }
  # Enforce title-only + compact theme at display/save time
  gg <- finish_process_plot(gg)
  print(gg)
  register_process_plot(gg, stem, width = width, height = height)
  if (isTRUE(save_process_maps)) {
    save_lsp_map(gg, stem, width = width, height = height)
    message("Saved: ", stem)
  } else {
    message("Console only (mid-run). Final chunk will write square PNG/PDF for ", stem)
  }
  invisible(gg)
}

as_bbox <- function(xmin, xmax, ymin, ymax, crs = 4326) {
  st_bbox(c(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax), crs = st_crs(crs))
}

# Six Gulf of California states map window (avoid nation-wide Mexico extents from ANP/ZRP layers)
goc_xlim <- c(-117.2, -104.5)
goc_ylim <- c(18.8, 32.7)
goc_bb <- as_bbox(goc_xlim[1], goc_xlim[2], goc_ylim[1], goc_ylim[2])

coord_goc_process <- function() {
  coord_sf(
    xlim = goc_xlim,
    ylim = goc_ylim,
    expand = FALSE,
    crs = st_crs(4326),
    default_crs = st_crs(4326)
  )
}

# Crop safely even when layer CRS differs from the target bbox CRS
crop_sf <- function(x, bb) {
  if (is.null(x) || nrow(x) == 0) return(x)
  tryCatch(
    {
      x <- st_make_valid(x)
      if (is.na(st_crs(x))) st_crs(x) <- 4326
      bb_poly <- st_as_sfc(st_bbox(bb))
      if (st_crs(bb_poly) != st_crs(x)) {
        bb_poly <- st_transform(bb_poly, st_crs(x))
      }
      suppressWarnings(st_crop(x, bb_poly))
    },
    error = function(e) x
  )
}

crop_goc <- function(x) crop_sf(x, goc_bb)

# Fishing site legend (matches prep: ZRP, then TURF, then protected-area score)
fish_scoring_rule <- function(in_zrp, in_turf, pa_condition = NA_real_) {
  case_when(
    !is.na(in_zrp) & as.logical(in_zrp) ~ "ZRP",
    !is.na(in_turf) & as.logical(in_turf) ~ "TURF",
    !is.na(pa_condition) & is.finite(pa_condition) & pa_condition > 0 ~ "In PA",
    TRUE ~ "Other"
  )
}

fish_color_vals <- c(
  "ZRP" = "#1a5276",
  "TURF" = "#C03729",
  "In PA" = "#5a8f3a",
  "Other" = "#555555"
)
fish_shape_vals <- c(
  "ZRP" = 16,
  "TURF" = 15,
  "In PA" = 17,
  "Other" = 4
)

filter_points_in_bbox <- function(pts, bb) {
  if (is.null(pts) || nrow(pts) == 0) return(pts)
  pts <- st_transform(pts, 4326)
  coords <- st_coordinates(pts)
  pts %>%
    mutate(.x = coords[, 1], .y = coords[, 2]) %>%
    filter(
      .x >= bb["xmin"], .x <= bb["xmax"],
      .y >= bb["ymin"], .y <= bb["ymax"]
    ) %>%
    dplyr::select(-.x, -.y)
}

# Always WGS84 for process maps (choropleth `states` may be in OHI projected CRS)
process_states <- tryCatch(
  {
    src <- if (exists("states") && !is.null(states)) {
      states
    } else {
      safe_st_read(here("spatial/ohi_regions/states.shp"))
    }
    if (is.null(src)) NULL else crop_goc(st_transform(src, 4326))
  },
  error = function(e) NULL
)

process_anp_sf <- safe_st_read(first_path(
  file.path(goc_root, "spatial/from_hem_POEM_ANP/232_ANP-ITRF08_16062025/232_ANP-ITRF08_16062025.shp")
))
if (!is.null(process_anp_sf)) {
  process_anp_sf <- process_anp_sf %>%
    st_zm(drop = TRUE, what = "ZM") %>%
    st_make_valid() %>%
    st_transform(4326) %>%
    st_simplify(dTolerance = 0.002, preserveTopology = TRUE) %>%
    crop_goc()
}

all_protected_sf <- safe_st_read(first_path(
  file.path(goc_root, "spatial/protected_areas_wdpa_conanp_and_hem/all_goc_protected_areas_v2_2025.shp"),
  file.path(goc_root, "spatial/protected _areas_wdpa_and_hem/all_goc_protected_areas_2025.shp")
))
if (!is.null(all_protected_sf)) {
  all_protected_sf <- all_protected_sf %>%
    st_zm(drop = TRUE, what = "ZM") %>%
    st_make_valid() %>%
    st_transform(4326)
  nm <- names(all_protected_sf)
  name_col <- nm[tolower(nm) %in% c("name", "nombre", "orig_name", "wdpa_pid")][1]
  if (!is.na(name_col) && name_col != "name") {
    all_protected_sf <- all_protected_sf %>% rename(name = !!sym(name_col))
  }
  all_protected_sf <- crop_goc(all_protected_sf)
}

ewg_coords_v4 <- safe_read_csv(first_path(
  file.path(lsp_prep_root, "identifying_lasting_special_places_collapsed_lsp_protections_v4.csv"),
  file.path(mazu_lsp_root, "identifying_lasting_special_places_collapsed_lsp_protections_v4.csv"),
  file.path(local_lsp_root, "identifying_lasting_special_places_collapsed_lsp_protections_v4.csv")
))
if (!is.null(ewg_coords_v4)) {
  names(ewg_coords_v4) <- names(ewg_coords_v4) %>%
    tolower() %>%
    str_replace_all("[^a-z0-9]+", "_") %>%
    str_replace_all("^_|_$", "") %>%
    str_replace_all("_+", "_")
  if (!"lasting_special_place_lsp" %in% names(ewg_coords_v4)) {
    hit <- names(ewg_coords_v4)[str_detect(names(ewg_coords_v4), "lasting_special_place")]
    if (length(hit) > 0) {
      ewg_coords_v4 <- ewg_coords_v4 %>% rename(lasting_special_place_lsp = !!sym(hit[[1]]))
    }
  }
  ewg_coords_v4 <- ewg_coords_v4 %>%
    mutate(
      lat = suppressWarnings(as.numeric(lat)),
      long = suppressWarnings(as.numeric(long)),
      lasting_special_place_lsp = str_squish(as.character(lasting_special_place_lsp))
    ) %>%
    filter(!is.na(lat), !is.na(long), !is.na(lasting_special_place_lsp))
}

# SIMEC raw table (optional; used when we want pure i-efectividad fill)
simec_raw <- safe_read_csv(first_path(
  file.path(goc_root, "_raw_data/SIMEC/identifying_lasting_special_places_SIMEC.csv"),
  file.path(lsp_prep_root, "identifying_lasting_special_places_SIMEC.csv")
))
if (!is.null(simec_raw)) {
  names(simec_raw) <- tolower(names(simec_raw))
}
```

## Script 2 — Special protected areas (static mapview)

Same join as the protected-area prep note: condition table joined onto all protected area polygons by WDPA name, then filled by **overall condition**.

**SIMEC i-efectividad when available, else Ramsar management-plan rule, else NA fill.**

```{r process-protected-areas, fig.width=7.5, fig.height=7.5}
pa_cond <- safe_read_csv(first_path(
  file.path(lsp_output_dir, "condition_lsp_protected_areas.csv"),
  file.path(lsp_output_dir, "protected_area_condition.csv")
))

gg_pa_process <- NULL
if (!is.null(pa_cond) && !is.null(all_protected_sf)) {
  pa_cond <- pa_cond %>%
    mutate(
      overall_condition = suppressWarnings(as.numeric(overall_condition)),
      wdpa_orig_name = str_squish(str_replace_all(as.character(wdpa_orig_name), "\\n", " "))
    )

  # Script 2 mapview equivalent: table first, geometry by WDPA name
  protected_condition_w_geometry <- left_join(
    pa_cond,
    all_protected_sf %>%
      mutate(name = str_squish(str_replace_all(as.character(name), "\\n", " "))) %>%
      dplyr::select(name, geometry),
    by = c("wdpa_orig_name" = "name")
  ) %>%
    st_as_sf() %>%
    filter(!st_is_empty(geometry))

  n_matched <- sum(!is.na(protected_condition_w_geometry$overall_condition))
  n_requested <- nrow(pa_cond)
  message(
    "PA rows with geometry + condition: ", n_matched,
    " of ", n_requested, " in condition table"
  )

  gg_pa_process <- ggplot() +
    {
      if (!is.null(process_states) && nrow(process_states) > 0) {
        geom_sf(data = process_states, fill = "#d6cac0", color = "grey40", linewidth = 0.12)
      }
    } +
    {
      if (!is.null(all_protected_sf) && nrow(all_protected_sf) > 0) {
        geom_sf(
          data = all_protected_sf,
          fill = NA,
          color = "grey55",
          linewidth = 0.08,
          alpha = 0.4
        )
      }
    } +
    geom_sf(
      data = protected_condition_w_geometry %>% filter(!is.na(overall_condition)),
      aes(fill = overall_condition),
      color = "grey20",
      linewidth = 0.15,
      alpha = 0.92
    ) +
    scale_fill_gradientn(
      colours = lsp_palette,
      limits = c(0, 100),
      na.value = "white",
      name = "Score"
    ) +
    coord_goc_process() +
    labs(
      title = "Special protected areas",
      subtitle = NULL,
      caption = NULL
    ) +
    process_theme()
}

gg_pa_process
show_process_map(gg_pa_process, "process_special_protected_areas")
```

## Script 3 — Priority dive sites

### Gulf-wide overall condition (static mapview)

Static version of:

`mapview(dive_sites_condition_sf, zcol = "overall_condition") + mapview(all_protected)`

```{r process-dive-sites-condition-map, fig.width=7.5, fig.height=7.5}
dive_cond <- safe_read_csv(first_path(
  file.path(lsp_output_dir, "condition_lsp_dive_sites.csv"),
  file.path(lsp_output_dir, "dive_sites_condition.csv")
))

gg_dive_condition <- NULL
if (!is.null(dive_cond)) {
  dive_cond <- dive_cond %>%
    mutate(
      overall_condition = suppressWarnings(as.numeric(overall_condition)),
      latitud = suppressWarnings(as.numeric(latitud)),
      longitud = suppressWarnings(as.numeric(longitud)),
      nombre_sitio_de_buceo = str_squish(as.character(nombre_sitio_de_buceo))
    ) %>%
    filter(!is.na(latitud), !is.na(longitud), !is.na(overall_condition)) %>%
    distinct(nombre_sitio_de_buceo, rgn_id, latitud, longitud, .keep_all = TRUE)

  dive_sites_condition_sf <- dive_cond %>%
    st_as_sf(coords = c("longitud", "latitud"), crs = 4326, remove = FALSE)

  gg_dive_condition <- ggplot() +
    {
      if (!is.null(process_states) && nrow(process_states) > 0) {
        geom_sf(data = process_states, fill = "#d6cac0", color = "grey40", linewidth = 0.12)
      }
    } +
    {
      if (!is.null(all_protected_sf) && nrow(all_protected_sf) > 0) {
        geom_sf(
          data = all_protected_sf,
          fill = "#98a25a",
          color = "grey35",
          linewidth = 0.1,
          alpha = 0.28
        )
      }
    } +
    geom_sf(
      data = dive_sites_condition_sf,
      aes(color = overall_condition),
      size = 2.2,
      alpha = 0.95
    ) +
    scale_color_gradientn(
      colours = lsp_palette,
      limits = c(0, 100),
      na.value = "grey50",
      name = "Score"
    ) +
    coord_goc_process() +
    labs(
      title = "Priority dive sites",
      subtitle = NULL,
      caption = NULL
    ) +
    process_theme()
}

gg_dive_condition
show_process_map(gg_dive_condition, "process_priority_dive_sites_condition")
```

### Espíritu Santo zoom (scoring rule)

Same decision rule as the EWG page: inside scored ANP → SIMEC; outside → NOM floor.

```{r process-dive-sites, fig.width=7.5, fig.height=7.5}
dive_gpkg <- safe_st_read(first_path(
  file.path(lsp_int_dir, "prep_dive_sites_priority_with_region.gpkg"),
  file.path(lsp_int_dir, "dive_sites_w_priority_goc_rgnid.gpkg"),
  file.path(mazu_lsp_root, "int/prep_dive_sites_priority_with_region.gpkg"),
  file.path(mazu_lsp_root, "int/dive_sites_w_priority_goc_rgnid.gpkg"),
  file.path(local_lsp_root, "int/dive_sites_w_priority_goc_rgnid.gpkg")
))

gg_dive_process <- NULL
if (!is.null(dive_cond) && !is.null(dive_gpkg)) {
  dive_cond_rules <- dive_cond %>%
    mutate(
      overall_condition = suppressWarnings(as.numeric(overall_condition)),
      pa_condition = suppressWarnings(as.numeric(protected_area_condition)),
      nombre_sitio_de_buceo = str_squish(as.character(nombre_sitio_de_buceo)),
      scoring_rule = if_else(
        !is.na(pa_condition) & is.finite(pa_condition) & pa_condition > 0,
        "Inside PA",
        "Outside PA"
      )
    ) %>%
    distinct(nombre_sitio_de_buceo, .keep_all = TRUE)

  dives <- dive_gpkg %>%
    st_transform(4326) %>%
    mutate(nombre_sitio_de_buceo = str_squish(as.character(nombre_sitio_de_buceo))) %>%
    left_join(
      dive_cond_rules %>% dplyr::select(nombre_sitio_de_buceo, overall_condition, scoring_rule),
      by = "nombre_sitio_de_buceo"
    )

  # Leave headroom north of PN Espíritu Santo (ymax 24.55 was cropping the park box)
  bb <- as_bbox(-110.55, -110.12, 24.05, 24.72)
  pa_focus <- crop_sf(if (!is.null(process_anp_sf)) process_anp_sf else all_protected_sf, bb)
  dive_focus <- filter_points_in_bbox(dives, bb)

  if (!is.null(dive_focus) && nrow(dive_focus) > 0) {
    gg_dive_process <- ggplot() +
      {
        if (!is.null(pa_focus) && nrow(pa_focus) > 0) {
          geom_sf(data = pa_focus, fill = "#98a25a", color = "gray30", linewidth = 0.2, alpha = 0.45)
        }
      } +
      geom_sf(
        data = dive_focus,
        aes(color = scoring_rule, shape = scoring_rule),
        size = 2.4,
        alpha = 0.95
      ) +
      scale_color_manual(
        name = NULL,
        values = c(
          "Inside PA" = "#C03729",
          "Outside PA" = "#333333"
        ),
        na.value = "gray50"
      ) +
      scale_shape_manual(
        name = NULL,
        values = c(
          "Inside PA" = 16,
          "Outside PA" = 17
        ),
        na.value = 4
      ) +
      coord_sf(xlim = c(bb["xmin"], bb["xmax"]), ylim = c(bb["ymin"], bb["ymax"]), expand = FALSE) +
      labs(
        title = "Priority dive sites near Espiritu Santo",
        subtitle = NULL,
        caption = NULL
      ) +
      process_theme() +
      guides(color = guide_legend(nrow = 1, title = NULL), shape = guide_legend(nrow = 1, title = NULL))
  }
}

gg_dive_process
show_process_map(gg_dive_process, "process_priority_dive_sites")
```

## Script 5 — Iconic fishing spots (Gulf-wide points) + ZRP overview

Gulf-wide maps are cropped to the six Gulf of California states. Site colors mark how each place was scored:

- **ZRP** = inside a zona de refugio pesquero (fishing refuge; polygon wash is ZRP only)
- **TURF** = cooperative fishing territory (points only; national TURF polygons are not plotted)
- **In PA** = not ZRP/TURF, but overlaps a scored protected area
- **Other** = none of the above on the prep table

```{r process-fishing-gulf, fig.width=7.5, fig.height=7.5}
fish_cond <- safe_read_csv(first_path(
  file.path(lsp_output_dir, "condition_lsp_fishing_sites.csv"),
  file.path(lsp_output_dir, "fishing_sites_condition.csv")
))
zrp_sf <- safe_st_read(first_path(
  file.path(goc_root, "_raw_data/TNC_geoportal_fishing_refuge_zones/d2026/Zonas_Refugios_Pesqueros_-8805549763229037385.geojson")
))
if (!is.null(zrp_sf)) {
  # Nation-wide layer: keep only Gulf window
  zrp_sf <- zrp_sf %>% st_make_valid() %>% st_transform(4326) %>% crop_goc()
}

gg_fish_loreto <- NULL
if (!is.null(fish_cond) && !is.null(ewg_coords_v4)) {
  if (!"in_turf" %in% names(fish_cond)) fish_cond$in_turf <- FALSE
  if (!"protected_area_condition" %in% names(fish_cond)) {
    fish_cond$protected_area_condition <- NA_real_
  }

  fish_cond <- fish_cond %>%
    mutate(
      lasting_special_place_lsp = str_squish(as.character(lasting_special_place_lsp)),
      overall_condition = suppressWarnings(as.numeric(overall_condition)),
      in_zrp = as.logical(in_zrp),
      in_turf = as.logical(in_turf),
      pa_condition = suppressWarnings(as.numeric(protected_area_condition)),
      scoring_rule = fish_scoring_rule(in_zrp, in_turf, pa_condition),
      scoring_rule = factor(scoring_rule, levels = c("ZRP", "TURF", "In PA", "Other"))
    )

  fish_sf <- ewg_coords_v4 %>%
    mutate(lasting_special_place_lsp = str_squish(as.character(lasting_special_place_lsp))) %>%
    inner_join(
      fish_cond %>% dplyr::select(
        lasting_special_place_lsp, overall_condition, scoring_rule, in_zrp, in_turf
      ),
      by = "lasting_special_place_lsp"
    ) %>%
    st_as_sf(coords = c("long", "lat"), crs = 4326, remove = FALSE)

  message("Iconic fishing sites plotted (Gulf of California): ", nrow(fish_sf))
  message(
    "  ZRP: ", sum(fish_sf$scoring_rule == "ZRP", na.rm = TRUE),
    " | TURF: ", sum(fish_sf$scoring_rule == "TURF", na.rm = TRUE),
    " | In PA: ", sum(fish_sf$scoring_rule == "In PA", na.rm = TRUE),
    " | Other: ", sum(fish_sf$scoring_rule == "Other", na.rm = TRUE)
  )

  if (nrow(fish_sf) > 0) {
    gg_fish_loreto <- ggplot() +
      {
        if (!is.null(process_states) && nrow(process_states) > 0) {
          geom_sf(data = process_states, fill = "#d6cac0", color = "grey40", linewidth = 0.12)
        }
      } +
      {
        if (!is.null(zrp_sf) && nrow(zrp_sf) > 0) {
          geom_sf(
            data = zrp_sf,
            aes(fill = "ZRP polygon"),
            color = "#1a5276",
            linewidth = 0.15,
            alpha = 0.4
          )
        }
      } +
      geom_sf(
        data = fish_sf,
        aes(color = scoring_rule, shape = scoring_rule),
        size = 2.8,
        alpha = 0.95
      ) +
      scale_fill_manual(
        name = NULL,
        values = c("ZRP polygon" = "#5dade2")
      ) +
      scale_color_manual(name = NULL, values = fish_color_vals, drop = FALSE) +
      scale_shape_manual(name = NULL, values = fish_shape_vals, drop = FALSE) +
      coord_goc_process() +
      labs(
        title = "Iconic fishing spots (ZRP vs TURF)",
        subtitle = NULL,
        caption = NULL
      ) +
      process_theme() +
      guides(
        fill = guide_legend(order = 1, override.aes = list(alpha = 0.5, shape = NA)),
        color = guide_legend(order = 2, nrow = 1, title = NULL),
        shape = guide_legend(order = 2, nrow = 1, title = NULL)
      )
  }
}

gg_fish_loreto
show_process_map(gg_fish_loreto, "process_iconic_fishing_gulf")
```

### Fishing refuges (ZRPs) and TURF sites — overview

Blue wash = ZRP polygons in the Gulf crop. Circles = ZRP sites; squares = TURF sites (no national TURF polygons, so TURFs appear as points only).

```{r process-zrp-overview, fig.width=7.5, fig.height=7.5}
gg_zrp_overview <- NULL
has_fish_sf <- exists("fish_sf") && !is.null(fish_sf) && inherits(fish_sf, "sf") && nrow(fish_sf) > 0
if ((!is.null(zrp_sf) && nrow(zrp_sf) > 0) || has_fish_sf) {
  turf_pts <- NULL
  zrp_pts <- NULL
  if (has_fish_sf) {
    zrp_pts <- fish_sf %>% filter(scoring_rule == "ZRP")
    turf_pts <- fish_sf %>% filter(scoring_rule == "TURF")
  }

  gg_zrp_overview <- ggplot() +
    {
      if (!is.null(process_states) && nrow(process_states) > 0) {
        geom_sf(data = process_states, fill = "#d6cac0", color = "grey40", linewidth = 0.12)
      }
    } +
    {
      if (!is.null(all_protected_sf) && nrow(all_protected_sf) > 0) {
        geom_sf(data = all_protected_sf, fill = "#98a25a", color = NA, alpha = 0.12)
      }
    } +
    {
      if (!is.null(zrp_sf) && nrow(zrp_sf) > 0) {
        geom_sf(
          data = zrp_sf,
          aes(fill = "ZRP"),
          color = "#1a5276",
          linewidth = 0.2,
          alpha = 0.55
        )
      }
    } +
    {
      if (!is.null(zrp_pts) && nrow(zrp_pts) > 0) {
        geom_sf(data = zrp_pts, aes(color = "ZRP site", shape = "ZRP site"), size = 2.6)
      }
    } +
    {
      if (!is.null(turf_pts) && nrow(turf_pts) > 0) {
        geom_sf(data = turf_pts, aes(color = "TURF site", shape = "TURF site"), size = 3.0)
      }
    } +
    scale_fill_manual(name = NULL, values = c("ZRP" = "#5dade2")) +
    scale_color_manual(
      name = NULL,
      values = c("ZRP site" = "#1a5276", "TURF site" = "#C03729")
    ) +
    scale_shape_manual(
      name = NULL,
      values = c("ZRP site" = 16, "TURF site" = 15)
    ) +
    coord_goc_process() +
    labs(
      title = "ZRPs and TURF fishing sites",
      subtitle = NULL,
      caption = NULL
    ) +
    process_theme() +
    guides(
      fill = guide_legend(order = 1),
      color = guide_legend(order = 2, nrow = 1),
      shape = guide_legend(order = 2, nrow = 1)
    )
}

gg_zrp_overview
show_process_map(gg_zrp_overview, "process_zrp_overview")
```

## Script 4 — Archaeological sites (table + optional buffer diagram)

There are only about 20 scored sites. A table is often clearer than a full Gulf map of points. Columns show the condition score and how it was assigned.

For each site we first check for a nearby INAH inventory match (10 km). If none, we look for a nearby natural protected area (500 m) and use its management score when available. Otherwise we use a partial legal cover score of 30.

```{r process-arch-table}
arch_cond <- safe_read_csv(first_path(
  file.path(lsp_output_dir, "condition_lsp_archaeological_sites.csv"),
  file.path(lsp_output_dir, "ewg_archaeological_sites_condition.csv")
))

arch_table <- NULL
if (!is.null(arch_cond)) {
  arch_table <- arch_cond %>%
    mutate(
      overall_condition = suppressWarnings(as.numeric(overall_condition)),
      pa_condition = suppressWarnings(as.numeric(protected_area_condition)),
      in_inah = as.logical(in_INAH_directory),
      scoring_rule = case_when(
        !is.na(in_inah) & in_inah ~ "INAH",
        !is.na(pa_condition) & pa_condition > 0 ~ "In PA",
        !is.na(overall_condition) & overall_condition <= 35 ~ "Legal only",
        TRUE ~ "Other"
      ),
      site = str_trunc(str_squish(as.character(lasting_special_place_lsp)), 70)
    ) %>%
    dplyr::select(
      rgn_id,
      site,
      overall_condition,
      scoring_rule,
      in_INAH = in_inah,
      nombre_INAH,
      protected_area_condition = pa_condition
    ) %>%
    arrange(rgn_id, desc(overall_condition))

  # print for console; knitr will also render the table in the HTML doc
  print(arch_table, n = Inf)
  knitr::kable(
    arch_table %>% dplyr::select(rgn_id, site, overall_condition, scoring_rule),
    caption = "Archaeological and cultural places: condition score and how it was assigned"
  )
}
```

Optional: static map of archaeological points with **500 m** (protected area) and **10 km** (INAH) buffers. All three Region 9 sites have coordinates in the inventory; nearby shell-mound sites are offset slightly so they do not stack into one red dot. Extent includes the full Gulf and all of Jalisco.

```{r process-arch-buffer-map, fig.width=7.5, fig.height=7.5}
gg_arch_buffer <- NULL

# Shared plain-language scoring labels for archaeology products
arch_rule_label <- function(in_inah, pa_condition, overall_condition) {
  case_when(
    !is.na(in_inah) & in_inah ~ "INAH",
    !is.na(pa_condition) & pa_condition > 0 ~ "In PA",
    !is.na(overall_condition) & overall_condition <= 35 ~ "Legal only",
    TRUE ~ "Other"
  )
}

# Normalize names so long multi-line strings still join to the coordinate inventory
normalize_lsp_name <- function(x) {
  x %>%
    as.character() %>%
    str_replace_all("[\\r\\n]+", " ") %>%
    str_squish()
}

if (!is.null(arch_cond)) {
  arch_rule <- arch_cond %>%
    mutate(
      lasting_special_place_lsp = normalize_lsp_name(lasting_special_place_lsp),
      overall_condition = suppressWarnings(as.numeric(overall_condition)),
      pa_condition = suppressWarnings(as.numeric(protected_area_condition)),
      in_inah = as.logical(in_INAH_directory),
      scoring_rule = arch_rule_label(in_inah, pa_condition, overall_condition),
      rgn_id = normalize_rgn_id(rgn_id)
    ) %>%
    dplyr::select(rgn_id, lasting_special_place_lsp, overall_condition, scoring_rule)

  # Coordinates: v4 inventory, then prepared EWG sheet (same columns if both present)
  coord_candidates <- list(
    ewg_coords_v4,
    safe_read_csv(first_path(
      file.path(lsp_prep_root, "google_sheets/ewg_lsp_inventory_prepared.csv"),
      file.path(mazu_lsp_root, "google_sheets/ewg_lsp_inventory_prepared.csv"),
      file.path(local_lsp_root, "google_sheets/ewg_lsp_inventory_prepared.csv"),
      file.path(goc_root, "goal_prep/sp/lsp/v2025/google_sheets/ewg_lsp_inventory_prepared.csv")
    ))
  )

  arch_coords_list <- lapply(coord_candidates, function(df) {
    if (is.null(df)) return(NULL)
    nm <- names(df)
    # tolerate either clean or raw column names
    if (!"lasting_special_place_lsp" %in% nm) {
      hit <- nm[str_detect(tolower(nm), "lasting.?special.?place")]
      if (length(hit) == 0) return(NULL)
      df <- df %>% rename(lasting_special_place_lsp = !!sym(hit[[1]]))
    }
    lat_col <- names(df)[tolower(names(df)) %in% c("lat", "latitude")][1]
    lon_col <- names(df)[tolower(names(df)) %in% c("long", "lon", "longitude")][1]
    if (is.na(lat_col) || is.na(lon_col)) return(NULL)
    df %>%
      mutate(
        lasting_special_place_lsp = normalize_lsp_name(lasting_special_place_lsp),
        lat = suppressWarnings(as.numeric(.data[[lat_col]])),
        long = suppressWarnings(as.numeric(.data[[lon_col]]))
      ) %>%
      filter(!is.na(lat), !is.na(long), lasting_special_place_lsp != "") %>%
      dplyr::select(lasting_special_place_lsp, lat, long)
  })
  arch_coords_list <- arch_coords_list[!vapply(arch_coords_list, is.null, logical(1))]
  if (length(arch_coords_list) == 0) {
    message("No coordinate inventory found for archaeology map.")
    arch_coords <- tibble(lasting_special_place_lsp = character(), lat = numeric(), long = numeric())
  } else {
    arch_coords <- bind_rows(arch_coords_list) %>%
      distinct(lasting_special_place_lsp, .keep_all = TRUE)
  }

  arch_pts <- arch_rule %>%
    left_join(arch_coords, by = "lasting_special_place_lsp")

  missing_coords <- arch_pts %>% filter(is.na(lat) | is.na(long))
  if (nrow(missing_coords) > 0) {
    message(
      "Arch sites without coordinates (cannot plot): ",
      paste(str_trunc(missing_coords$lasting_special_place_lsp, 50), collapse = "; ")
    )
  }

  arch_pts <- arch_pts %>%
    filter(!is.na(lat), !is.na(long)) %>%
    st_as_sf(coords = c("long", "lat"), crs = 4326, remove = FALSE)

  message(
    "Arch map: ", nrow(arch_pts), " of ", nrow(arch_rule),
    " scored sites have coordinates and will be plotted."
  )

  if (nrow(arch_pts) > 0) {
    # True locations for buffers
    pts_proj <- st_transform(arch_pts, 32612)
    buf500 <- st_buffer(pts_proj, dist = 500) %>% st_transform(4326)
    buf10k <- st_buffer(pts_proj, dist = 10000) %>% st_transform(4326)

    # Fan out points that share nearly the same place so Gulf-scale maps show each site
    # (e.g. EL ROBLITO and CUMULO de conchas are ~3 km apart and stack otherwise).
    arch_pts_plot <- arch_pts %>%
      mutate(
        cell = paste0(round(long / 0.1), "_", round(lat / 0.1)),
        short_label = lasting_special_place_lsp %>%
          str_replace("\\s*Important for.*$", "") %>%
          str_replace("\\s*https?://\\S+", "") %>%
          str_trunc(32)
      ) %>%
      group_by(cell) %>%
      mutate(
        n_near = n(),
        ang = if_else(n_near > 1, 2 * pi * (row_number() - 1) / n_near, 0),
        # ~0.22 deg (~24 km) keeps stacked sites readable on the full-Gulf map
        r_off = if_else(n_near > 1, 0.22, 0),
        long_plot = long + r_off * cos(ang),
        lat_plot = lat + r_off * sin(ang)
      ) %>%
      ungroup() %>%
      st_drop_geometry() %>%
      st_as_sf(coords = c("long_plot", "lat_plot"), crs = 4326, remove = FALSE)

    bb_c <- st_bbox(buf10k)
    pad <- 0.2
    bb <- as_bbox(
      min(as.numeric(bb_c["xmin"]) - pad, -115.5),
      max(as.numeric(bb_c["xmax"]) + pad, -101.5),
      min(as.numeric(bb_c["ymin"]) - pad, 18.9),
      max(as.numeric(bb_c["ymax"]) + pad, 32.5)
    )
    if (!all(is.finite(c(bb["xmin"], bb["xmax"], bb["ymin"], bb["ymax"]))) ||
        (bb["xmax"] - bb["xmin"]) < 0.2 ||
        (bb["ymax"] - bb["ymin"]) < 0.2) {
      bb <- as_bbox(-115.5, -101.5, 18.9, 32.5)
    }

    states_f <- process_states
    if (!is.null(states_f) && nrow(states_f) > 0) {
      states_f <- st_transform(states_f, 4326)
    }
    anp_src <- if (!is.null(process_anp_sf)) process_anp_sf else all_protected_sf
    anp_f <- anp_src
    if (!is.null(anp_f) && nrow(anp_f) > 0) {
      anp_f <- st_transform(anp_f, 4326)
    }

    message(
      "Arch sites plotted: ", nrow(arch_pts_plot),
      "; map lon [", round(bb["xmin"], 2), ", ", round(bb["xmax"], 2),
      "], lat [", round(bb["ymin"], 2), ", ", round(bb["ymax"], 2), "]"
    )
    r9 <- arch_pts_plot %>% filter(rgn_id %in% c("Region_9", "Region 9"))
    if (nrow(r9) > 0) {
      message("Region 9 sites on map: ", paste(r9$short_label, collapse = "; "))
    }

    gg_arch_buffer <- ggplot() +
      {
        if (!is.null(states_f) && nrow(states_f) > 0) {
          geom_sf(data = states_f, fill = "#d6cac0", color = "grey40", linewidth = 0.12)
        }
      } +
      {
        if (!is.null(anp_f) && nrow(anp_f) > 0) {
          geom_sf(data = anp_f, fill = "#98a25a", color = NA, alpha = 0.18)
        }
      } +
      geom_sf(data = buf10k, fill = NA, color = "#5dade2", linewidth = 0.3, linetype = "dashed") +
      geom_sf(data = buf500, fill = NA, color = "#C03729", linewidth = 0.4) +
      geom_sf(
        data = arch_pts_plot,
        aes(color = scoring_rule, shape = scoring_rule),
        size = 2.8,
        alpha = 0.95
      ) +
      {
        if (requireNamespace("ggrepel", quietly = TRUE)) {
          ggrepel::geom_text_repel(
            data = st_drop_geometry(arch_pts_plot),
            aes(x = long_plot, y = lat_plot, label = short_label),
            size = 2.4,
            color = "grey20",
            max.overlaps = 40,
            min.segment.length = 0,
            box.padding = 0.25,
            segment.color = "grey50",
            segment.size = 0.2,
            seed = 1,
            inherit.aes = FALSE
          )
        } else {
          geom_text(
            data = st_drop_geometry(arch_pts_plot),
            aes(x = long_plot, y = lat_plot, label = short_label),
            size = 2.2,
            color = "grey20",
            check_overlap = TRUE,
            inherit.aes = FALSE
          )
        }
      } +
      scale_color_manual(
        name = NULL,
        values = c(
          "INAH" = "#1a5276",
          "In PA" = "#C03729",
          "Legal only" = "#333333",
          "Other" = "#7f8c8d"
        )
      ) +
      scale_shape_manual(
        name = NULL,
        values = c(
          "INAH" = 15,
          "In PA" = 16,
          "Legal only" = 17,
          "Other" = 18
        )
      ) +
      coord_sf(
        xlim = c(bb["xmin"], bb["xmax"]),
        ylim = c(bb["ymin"], bb["ymax"]),
        expand = FALSE,
        crs = st_crs(4326)
      ) +
      labs(
        title = "Archaeological and cultural sites",
        subtitle = NULL,
        caption = NULL
      ) +
      process_theme() +
      guides(color = guide_legend(nrow = 1, title = NULL), shape = guide_legend(nrow = 1, title = NULL))
  }
}

gg_arch_buffer
show_process_map(gg_arch_buffer, "process_archaeological_buffers")
```

### Upper Gulf zoom — light mapview for ANP QC

Heavy ggplot + full ANP geometry was crashing RStudio. This chunk only:

1. Filters Upper Gulf archive points  
2. Crops ANPs to that window (no `st_union`)  
3. Adds 500 m buffers  
4. Prints a small QC table  
5. Opens **mapview** (zoom/pan/click)

```{r process-arch-upper-gulf}
# Lightweight Upper Gulf explorer (avoids ggplot st_union crash)
if (!requireNamespace("mapview", quietly = TRUE)) {
  message("Install mapview first: install.packages(\"mapview\")")
} else {
  library(mapview)
  mapviewOptions(fgb = FALSE)  # more stable in some RStudio setups
}

gg_arch_upper_gulf <- NULL  # not used; mapview opens in Viewer

if (!is.null(arch_cond) && !is.null(ewg_coords_v4)) {
  arch_rule_ug <- arch_cond %>%
    mutate(
      lasting_special_place_lsp = str_squish(as.character(lasting_special_place_lsp)),
      overall_condition = suppressWarnings(as.numeric(overall_condition)),
      pa_condition = suppressWarnings(as.numeric(protected_area_condition)),
      in_inah = as.logical(in_INAH_directory),
      scoring_rule = case_when(
        !is.na(in_inah) & in_inah ~ "INAH",
        !is.na(pa_condition) & pa_condition > 0 ~ "In PA",
        !is.na(overall_condition) & overall_condition <= 35 ~ "Legal only",
        TRUE ~ "Other"
      ),
      short_label = lasting_special_place_lsp %>%
        str_replace("\\s*\\(.*$", "") %>%
        str_trunc(40)
    ) %>%
    dplyr::select(
      lasting_special_place_lsp, short_label, overall_condition,
      pa_condition, scoring_rule
    )

  arch_all <- ewg_coords_v4 %>%
    mutate(lasting_special_place_lsp = str_squish(as.character(lasting_special_place_lsp))) %>%
    inner_join(arch_rule_ug, by = "lasting_special_place_lsp") %>%
    st_as_sf(coords = c("long", "lat"), crs = 4326, remove = FALSE)

  bb_ug <- as_bbox(-115.15, -113.15, 30.95, 32.15)
  arch_ug <- filter_points_in_bbox(arch_all, bb_ug)

  if (is.null(arch_ug) || nrow(arch_ug) == 0) {
    message("No arch points in Upper Gulf window; check coords / bb_ug.")
  } else {
    # Crop ANPs first (small geometry) — do NOT union full GOC layer
    anp_src <- if (!is.null(all_protected_sf)) all_protected_sf else process_anp_sf
    anp_ug <- NULL
    if (!is.null(anp_src) && nrow(anp_src) > 0) {
      anp_ug <- crop_sf(st_transform(anp_src, 4326), bb_ug)
      if (!is.null(anp_ug) && nrow(anp_ug) > 0) {
        anp_ug <- anp_ug %>%
          st_make_valid() %>%
          mutate(anp_label = if ("name" %in% names(.)) as.character(name) else "ANP")
      }
    }

    # 500 m buffers only for these few points
    buf500_ug <- st_transform(arch_ug, 32612) %>%
      st_buffer(dist = 500) %>%
      st_transform(4326)

    # Light intersection QC (cropped ANPs only)
    if (!is.null(anp_ug) && nrow(anp_ug) > 0) {
      hit <- suppressWarnings(st_intersects(buf500_ug, anp_ug, sparse = TRUE))
      arch_ug$touches_anp_500m <- lengths(hit) > 0
      nm <- if ("name" %in% names(anp_ug)) anp_ug$name else rep("ANP", nrow(anp_ug))
      arch_ug$anp_name_hit <- vapply(hit, function(ix) {
        if (length(ix) == 0) NA_character_ else paste(unique(as.character(nm[ix])), collapse = "; ")
      }, character(1))
    } else {
      arch_ug$touches_anp_500m <- NA
      arch_ug$anp_name_hit <- NA_character_
    }
    arch_ug$qc_review <- arch_ug$scoring_rule == "Legal only" &
      !is.na(arch_ug$touches_anp_500m) & arch_ug$touches_anp_500m

    message("Upper Gulf sites: ", nrow(arch_ug))
    print(
      st_drop_geometry(arch_ug) %>%
        dplyr::select(
          short_label, scoring_rule, overall_condition, pa_condition,
          touches_anp_500m, anp_name_hit, qc_review
        ) %>%
        arrange(desc(qc_review), scoring_rule, short_label)
    )

    if (requireNamespace("mapview", quietly = TRUE)) {
      m <- mapview(
        arch_ug,
        zcol = "scoring_rule",
        layer.name = "Arch sites (prep score)",
        cex = 6,
        legend = TRUE
      )
      if (!is.null(anp_ug) && nrow(anp_ug) > 0) {
        m <- m + mapview(
          anp_ug,
          col.regions = "#98a25a",
          alpha.regions = 0.35,
          layer.name = "ANPs (cropped)",
          legend = FALSE,
          homebutton = FALSE
        )
      }
      m <- m + mapview(
        buf500_ug,
        col.regions = "#C03729",
        alpha.regions = 0.05,
        color = "#C03729",
        lwd = 2,
        layer.name = "500 m ANP-join buffer",
        legend = FALSE,
        homebutton = FALSE
      )

      # Surface explicitly for Viewer
      map_arch_upper_gulf <<- m
      print(m)
      message("mapview object: map_arch_upper_gulf (print again if Viewer is blank)")
    } else {
      message("mapview not available — table above is still valid for QC.")
    }
  }
}
```

## Script 6 — Special habitats (how scoring works + visual)

### Confirmed from `script6_special_habitats.qmd`

1. **Each habitat feature** gets `overall_condition = pmax(protected_area_condition, nom_condition)`:
   - Coquina NOM floor = **30**
   - Coral / seagrass / mangrove NOM floor = **70**
2. **Protected-area test:** habitat geometry is buffered by **100 m**, then `st_intersects` against **non-dissolved** ANPs. If it **touches** a PA (after that buffer), it is flagged `in_protected_area` and receives that PA’s full SIMEC (or Ramsar plan rule)—**not** an area-weighted fraction of overlap. Multiple overlapping PAs are averaged with `mean()`.
3. **Regional special-habitats score:** first average feature scores **within each habitat type** for the region, then average those **type-level** means. So habitat **types present in the region are treated as equally important** (coral one unit, seagrass one unit, …)—not equal weight per individual patch. (Comment in code says “features,” but the combine step is mean of type rows.)

```{r process-habitats-method-figure, fig.width=7.5, fig.height=7.5}
hab_feat <- safe_read_csv(file.path(lsp_output_dir, "special_habitats_features_condition.csv"))
gg_hab_features <- NULL

if (!is.null(hab_feat)) {
  # console summary of equal-type aggregation recipe
  type_rgn <- hab_feat %>%
    mutate(
      overall_condition = suppressWarnings(as.numeric(overall_condition)),
      rgn_id = normalize_rgn_id(rgn_id)
    ) %>%
    group_by(rgn_id, habitat_type) %>%
    summarize(
      type_mean = mean(overall_condition, na.rm = TRUE),
      n_features = n(),
      .groups = "drop"
    ) %>%
    mutate(
      rgn_label = factor(
        unname(rgn_labels[rgn_id]),
        levels = unname(rgn_labels)
      )
    )

  rgn_hab <- type_rgn %>%
    group_by(rgn_id, rgn_label) %>%
    summarize(
      condition = mean(type_mean, na.rm = TRUE),
      n_types = n(),
      n_features_total = sum(n_features),
      .groups = "drop"
    )

  message("Regional special-habitats condition = mean of per-type means (equal type weights):")
  print(rgn_hab)

  # Habitat colors (type-linked): seagrass green, mangrove brown, coral dark red, coquina purple-grey
  habitat_fills <- c(
    "Coquina Reef" = "#6E667A",
    "Coral Reef" = "#7A1212",
    "Mangrove" = "#7A4F2D",
    "Seagrass (Zostera marina)" = "#2F6B3A"
  )

  # Visual A: strip of type means contributing to each region
  gg_hab_types <- ggplot(type_rgn, aes(x = habitat_type, y = type_mean, fill = habitat_type)) +
    geom_col(width = 0.7, show.legend = FALSE) +
    geom_text(
      aes(label = sprintf("%.0f", type_mean)),
      vjust = -0.35,
      size = 2.6,
      color = "grey20"
    ) +
    scale_fill_manual(values = habitat_fills, na.value = "grey60") +
    facet_wrap(~ rgn_label, ncol = 3) +
    coord_cartesian(ylim = c(0, 108)) +
    labs(
      title = "Special habitats by type and region",
      subtitle = NULL,
      caption = NULL,
      x = NULL,
      y = "Mean condition"
    ) +
    theme_minimal(base_size = 10) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
      strip.text = element_text(face = "bold", size = 8),
      plot.title = element_text(face = "bold", size = 11),
      plot.subtitle = element_blank(),
      plot.caption = element_blank()
    )

  gg_hab_types
  show_process_map(gg_hab_types, "process_special_habitats_type_means")

  # Visual B: feature inventory — in PA vs NOM floor only (tabular console)
  rule_summary <- hab_feat %>%
    mutate(
      overall_condition = suppressWarnings(as.numeric(overall_condition)),
      pa_condition = suppressWarnings(as.numeric(protected_area_condition)),
      in_pa = as.logical(in_protected_area),
      scoring_rule = case_when(
        !is.na(in_pa) & in_pa & !is.na(pa_condition) ~ "Touches PA",
        TRUE ~ "Legal floor"
      )
    ) %>%
    count(habitat_type, scoring_rule)

  print(rule_summary)
  knitr::kable(rule_summary, caption = "Special habitat features by PA-touch vs NOM-only rule")
}

ggsave(
  "/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/viz/process_special_habitats_type_means.png",
  gg_hab_types, width = 7.5, height = 7.5
)

gg_hab_types

# Optional polygon map: coquina habitats showing PA-touch rule (Upper Gulf)
coquina_sf <- safe_st_read(first_path(
  file.path(lsp_int_dir, "coquina_reefs_condition.gpkg"),
  file.path(local_lsp_root, "int/coquina_reefs_condition.gpkg")
))
gg_hab_coquina <- NULL
if (!is.null(coquina_sf) && nrow(coquina_sf) > 0) {
  coquina_sf <- coquina_sf %>%
    st_zm(drop = TRUE, what = "ZM") %>%
    st_make_valid() %>%
    st_transform(4326)

  if ("in_protected_area" %in% names(coquina_sf)) {
    coquina_sf <- coquina_sf %>%
      mutate(in_pa = as.logical(in_protected_area))
  } else if ("protected_area_condition" %in% names(coquina_sf)) {
    coquina_sf <- coquina_sf %>%
      mutate(in_pa = !is.na(suppressWarnings(as.numeric(protected_area_condition))))
  } else {
    coquina_sf <- coquina_sf %>% mutate(in_pa = FALSE)
  }
  coquina_sf <- coquina_sf %>%
    mutate(
      overall_condition = suppressWarnings(as.numeric(overall_condition)),
      in_pa = ifelse(is.na(in_pa), FALSE, in_pa),
      pa_rule = factor(
        if_else(in_pa, "Touches PA", "Legal floor"),
        levels = c(
          "Touches PA",
          "Legal floor"
        )
      )
    )

  # Extent from coquina features (with pad), not a hard-coded window alone
  bb_c <- st_bbox(coquina_sf)
  pad <- 0.35
  bb <- as_bbox(
    bb_c["xmin"] - pad, bb_c["xmax"] + pad,
    bb_c["ymin"] - pad, bb_c["ymax"] + pad
  )
  # Guard against degenerate / empty extent
  if (!all(is.finite(c(bb["xmin"], bb["xmax"], bb["ymin"], bb["ymax"]))) ||
      (bb["xmax"] - bb["xmin"]) < 0.05 ||
      (bb["ymax"] - bb["ymin"]) < 0.05) {
    bb <- as_bbox(-114.9, -112.4, 29.3, 32.0)
  }

  states_c <- crop_sf(process_states, bb)
  anp_c <- crop_sf(
    if (!is.null(process_anp_sf)) process_anp_sf else all_protected_sf,
    bb
  )
  coq_c <- crop_sf(coquina_sf, bb)
  if (is.null(coq_c) || nrow(coq_c) == 0) coq_c <- coquina_sf

  message(
    "Coquina map: ", nrow(coq_c), " features; bbox lon [",
    round(bb["xmin"], 2), ", ", round(bb["xmax"], 2), "], lat [",
    round(bb["ymin"], 2), ", ", round(bb["ymax"], 2), "]"
  )

  gg_hab_coquina <- ggplot() +
    {
      if (!is.null(states_c) && nrow(states_c) > 0) {
        geom_sf(data = states_c, fill = "#d6cac0", color = "grey40", linewidth = 0.1)
      }
    } +
    {
      if (!is.null(anp_c) && nrow(anp_c) > 0) {
        geom_sf(
          data = anp_c,
          fill = "#98a25a",
          color = "gray30",
          alpha = 0.18,
          linewidth = 0.12
        )
      }
    } +
    geom_sf(
      data = coq_c,
      aes(fill = pa_rule),
      color = "grey15",
      linewidth = 0.2,
      alpha = 0.9
    ) +
    scale_fill_manual(
      name = NULL,
      values = c(
        "Touches PA" = "#C03729",
        "Legal floor" = "#6E667A"
      ),
      drop = FALSE
    ) +
    coord_sf(
      xlim = c(bb["xmin"], bb["xmax"]),
      ylim = c(bb["ymin"], bb["ymax"]),
      expand = FALSE,
      crs = st_crs(4326)
    ) +
    labs(
        title = "Coquina reefs and protected areas",
        subtitle = NULL,
        caption = NULL
      ) +
    process_theme()

  gg_hab_coquina
  show_process_map(gg_hab_coquina, "process_special_habitats_pa_touch")
}
```

## Script 7 — All other LSPs (dive-style map + codified protection table)

Outside scored ANPs, condition comes from the expert **codified protection spreadsheet** (`codified_protection_scores_v3.csv`). Sites that land in a PA use SIMEC/Ramsar like dives.

```{r process-other-map, fig.width=7.5, fig.height=7.5}
other_cond <- safe_read_csv(file.path(lsp_output_dir, "final_other_protections_condition.csv"))

gg_other_process <- NULL
if (!is.null(other_cond) && !is.null(ewg_coords_v4)) {
  other_cond <- other_cond %>%
    mutate(
      lsp_name = str_squish(as.character(lsp_name)),
      overall_condition = suppressWarnings(as.numeric(overall_condition)),
      pa_condition = suppressWarnings(as.numeric(protected_area_condition)),
      scoring_rule = if_else(
        !is.na(pa_condition) & pa_condition > 0,
        "Inside PA",
        "Outside PA"
      )
    )

  other_sf <- ewg_coords_v4 %>%
    mutate(lsp_name = str_squish(as.character(lasting_special_place_lsp))) %>%
    inner_join(
      other_cond %>% dplyr::select(lsp_name, overall_condition, scoring_rule),
      by = "lsp_name"
    ) %>%
    st_as_sf(coords = c("long", "lat"), crs = 4326, remove = FALSE)

  message("Other LSPs plotted (Gulf-wide): ", nrow(other_sf))
  message(
    "  Inside PA: ",
    sum(other_sf$scoring_rule == "Inside PA", na.rm = TRUE),
    " | Outside PA (codified): ",
    sum(other_sf$scoring_rule == "Outside PA", na.rm = TRUE)
  )

  # Gulf-wide: all points + full underlays in WGS84 (no tight bbox — avoids CRS zoom crashes)
  states_f <- process_states
  if (!is.null(states_f) && nrow(states_f) > 0) {
    states_f <- st_transform(states_f, 4326)
  }
  anp_src <- if (!is.null(process_anp_sf)) process_anp_sf else all_protected_sf
  anp_f <- anp_src
  if (!is.null(anp_f) && nrow(anp_f) > 0) {
    anp_f <- st_transform(anp_f, 4326)
  }

  if (nrow(other_sf) > 0) {
    gg_other_process <- ggplot() +
      {
        if (!is.null(states_f) && nrow(states_f) > 0) {
          geom_sf(data = states_f, fill = "#d6cac0", color = "grey40", linewidth = 0.12)
        }
      } +
      {
        if (!is.null(anp_f) && nrow(anp_f) > 0) {
          geom_sf(data = anp_f, fill = "#98a25a", color = NA, alpha = 0.2)
        }
      } +
      geom_sf(
        data = other_sf,
        aes(color = scoring_rule, shape = scoring_rule),
        size = 2.6,
        alpha = 0.95
      ) +
      scale_color_manual(
        name = NULL,
        values = c(
          "Inside PA" = "#C03729",
          "Outside PA" = "#333333"
        )
      ) +
      scale_shape_manual(
        name = NULL,
        values = c(
          "Inside PA" = 16,
          "Outside PA" = 17
        )
      ) +
      coord_goc_process() +
      labs(
        title = "Other marine places of importance",
        subtitle = NULL,
        caption = NULL
      ) +
      process_theme() +
      guides(color = guide_legend(nrow = 1, title = NULL), shape = guide_legend(nrow = 1, title = NULL))
  }
}

gg_other_process
show_process_map(gg_other_process, "process_all_other_special_places")
```

### Codified protection spreadsheet (for sites outside SIMEC)

```{r process-other-codified-table}
codified <- safe_read_csv(first_path(
  file.path(lsp_int_dir, "codified_protection_scores_v3.csv"),
  file.path(local_lsp_root, "int/codified_protection_scores_v3.csv"),
  file.path(lsp_prep_root, "int/codified_protection_scores_v3.csv")
))

if (!is.null(codified)) {
  cod_show <- codified %>%
    mutate(
      lsp_name = str_trunc(str_squish(as.character(lsp_name)), 55),
      estimated_condition_of_protection = suppressWarnings(as.numeric(estimated_condition_of_protection)),
      other_legal = str_trunc(str_squish(as.character(other_legal_protections_other_than_ANPs)), 50)
    ) %>%
    dplyr::select(
      lsp_name,
      estimated_condition = estimated_condition_of_protection,
      level,
      other_legal
    ) %>%
    arrange(desc(estimated_condition), lsp_name)

  print(cod_show, n = 30)
  knitr::kable(
    cod_show,
    caption = "Codified protection scores (int/codified_protection_scores_v3.csv) used when other LSPs fall outside scored PAs"
  )
} else {
  message("Codified spreadsheet not found under int/codified_protection_scores_v3.csv")
}
```

```{r process-maps-console-objects}
message(
  "Process plot objects for console review:\n",
  "  gg_pa_process, gg_dive_condition, gg_dive_process, gg_fish_loreto, gg_zrp_overview,\n",
  "  gg_arch_buffer, map_arch_upper_gulf (mapview), gg_hab_types (if habitats ran),\n",
  "  gg_hab_coquina, gg_other_process\n",
  "Tables: arch_table, cod_show (if loaded)\n",
  "Registered for PDF export: ", paste(names(process_plot_registry), collapse = ", ")
)
```

# Save all plots as PDF to viz

Writes every choropleth and every registered process map as **PDF only** into `viz_dir` (Mazu when available).

```{r save-all-plots-pdf}
message("Viz folder: ", viz_dir)

# Choropleths + multi-panel (rebuild objects so re-running this chunk alone still works)
pdf_export_log <- list()

for (stem in names(score_layers)) {
  layer <- score_layers[[stem]]
  gg <- make_lsp_choropleth(
    layer$data,
    legend_title = lsp_legend_title,
    plot_title = layer$title,
    label_scores = FALSE
  )
  pdf_export_log[[stem]] <- save_lsp_map(gg, stem, width = viz_square, height = viz_square)
}

if (exists("gg_panel") && inherits(gg_panel, "ggplot")) {
  pdf_export_log[["lsp_categories_panel"]] <- save_lsp_map(
    gg_panel, "lsp_categories_panel", width = 10, height = 10
  )
}

# Process maps collected via show_process_map()
if (length(process_plot_registry) == 0) {
  message("No process plots registered yet. Run process-map chunks first.")
} else {
  for (stem in names(process_plot_registry)) {
    item <- process_plot_registry[[stem]]
    gg_export <- finish_process_plot(item$gg)
    w <- if (!is.null(item$width)) item$width else viz_square
    h <- if (!is.null(item$height)) item$height else viz_square
    pdf_export_log[[stem]] <- save_lsp_map(gg_export, stem, width = w, height = h)
  }
}

message(
  "PDF + PNG export finished. ", length(pdf_export_log),
  " stems under:\n  ", viz_dir
)
invisible(pdf_export_log)
```

# Session info

```{r}
sessionInfo()
```




## Check conditon of each type

```{r}
special_protec <- read_csv(here::here(lsp_output_dir, "rgn_protected_area_condition.csv"))
dive_cond <- read_csv(here::here(lsp_output_dir, "rgn_dive_sites_condition.csv"))
fish_cond <- read_csv(here::here(lsp_output_dir, "rgn_fishing_sites_condition.csv"))
arch_cond <- read_csv(here::here(lsp_output_dir, "rgn_ewg_archaeological_sites_condition.csv"))
hab_cond <- read_csv(here::here(lsp_output_dir, "rgn_special_habitats_condition.csv"))
other_cond <- read_csv(here::here(lsp_output_dir, "rgn_final_other_protections_condition.csv"))
```
