---
title: "Sense of Place (SP): LSP choropleth maps"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    theme: cerulean
editor_options:
  chunk_output_type: console
---

# Purpose

Static choropleth maps of **Lasting Special Places (LSP)** scores by OHI Gulf of California region, plus **process maps** that illustrate how each LSP type (scripts 2–7) is scored.

- overall LSP current status (2025 snapshot)
- each LSP category current status (where a region has no score: solid white)
- method illustrations: special protected areas, dive sites, archaeology, fishing spots, special habitats, all other places

Choropleths follow the TR / Access style (states underlay, marine OHI regions, 0–100 scale, white for NA). Process maps use focused extents and rule-based symbology.

Outputs (PNG and PDF) are written to:

`goal_prep/sp/lsp/v2025/viz` on Mazu, or `sp/lsp/v2025/viz` when using the local mirror.

```{r setup}
library(tidyverse)
library(sf)
library(here)
library(ggplot2)
library(grid)

# Prefer Mazu goal_prep paths; fall back to local workspace mirror
goc_root <- "/home/shares/ohi/OHI_GOC"
local_lsp_root <- here("sp/lsp/v2025")
mazu_lsp_root <- file.path(goc_root, "goal_prep/sp/lsp/v2025")

lsp_prep_root <- if (dir.exists(file.path(mazu_lsp_root, "output"))) {
  mazu_lsp_root
} else if (dir.exists(file.path(local_lsp_root, "output"))) {
  local_lsp_root
} else {
  mazu_lsp_root
}

lsp_output_dir <- file.path(lsp_prep_root, "output")
lsp_int_dir <- file.path(lsp_prep_root, "int")
viz_dir <- file.path(lsp_prep_root, "viz")
dir.create(viz_dir, recursive = TRUE, showWarnings = FALSE)

# first existing file among candidates
first_path <- function(...) {
  paths <- c(...)
  hit <- paths[file.exists(paths)]
  if (length(hit) == 0) NA_character_ else hit[[1]]
}

safe_read_csv <- function(path, ...) {
  if (is.na(path) || !file.exists(path)) return(NULL)
  tryCatch(read_csv(path, show_col_types = FALSE, ...), error = function(e) NULL)
}

safe_st_read <- function(path, ...) {
  if (is.na(path) || !file.exists(path)) return(NULL)
  tryCatch(st_read(path, quiet = TRUE, ...), error = function(e) NULL)
}

# TR / Access-style ramp (0 = dark red, 100 = dark blue)
lsp_palette <- c(
  "#8a0032", "#A50026", "#A50026", "#D73027", "#F46D43",
  "#FDAE61", "#FEE090", "#FEE090", "#ABD9E9", "#4575B4", "#313695"
)
lsp_na_fill <- "white"
lsp_legend_title <- "Current status score"

normalize_rgn_id <- function(x) {
  x %>%
    as.character() %>%
    str_squish() %>%
    str_replace_all(" ", "_")
}

rgn_labels <- c(
  "Region_1" = "Western Upper Gulf",
  "Region_2" = "Western Midriff Islands",
  "Region_3" = "Western Central Gulf",
  "Region_4" = "Western Southern Gulf",
  "Region_5" = "Eastern Upper Gulf",
  "Region_6" = "Eastern Midriff Islands",
  "Region_7" = "Southern Sonora Coast",
  "Region_8" = "Sinaloa Coast",
  "Region_9" = "Southeastern Gulf"
)
```

# Spatial layers

```{r spatial}
ohi_marine <- st_read(
  here("spatial/ohi_regions/goc_ohi_rgns.shp"),
  quiet = TRUE
) %>%
  filter(location == "marine") %>%
  mutate(rgn_id = normalize_rgn_id(rgn_id)) %>%
  dplyr::select(rgn_id, geometry)

states <- st_read(
  here("spatial/ohi_regions/states.shp"),
  quiet = TRUE
) %>%
  st_transform(st_crs(ohi_marine))
```

# Score tables

```{r scores}
read_rgn_condition <- function(path, value_col = "condition") {
  read_csv(path, show_col_types = FALSE) %>%
    mutate(
      rgn_id = normalize_rgn_id(rgn_id),
      score = suppressWarnings(as.numeric(.data[[value_col]]))
    ) %>%
    dplyr::select(rgn_id, score) %>%
    group_by(rgn_id) %>%
    summarize(score = mean(score, na.rm = TRUE), .groups = "drop")
}

lsp_overall <- read_csv(
  file.path(lsp_output_dir, "lsp_current_status.csv"),
  show_col_types = FALSE
) %>%
  mutate(
    rgn_id = normalize_rgn_id(rgn_id),
    score = suppressWarnings(as.numeric(current_status))
  ) %>%
  dplyr::select(rgn_id, score)

score_layers <- list(
  "lsp_overall_2025" = list(
    data = lsp_overall,
    title = "Lasting Special Places (LSP)\nCurrent status 2025"
  ),
  "lsp_special_protected_areas" = list(
    data = read_rgn_condition(file.path(lsp_output_dir, "rgn_protected_area_condition.csv")),
    title = "LSP: Special Protected Areas"
  ),
  "lsp_priority_dive_sites" = list(
    data = read_rgn_condition(file.path(lsp_output_dir, "rgn_dive_sites_condition.csv")),
    title = "LSP: Priority Dive Sites"
  ),
  "lsp_iconic_fishing_spots" = list(
    data = read_rgn_condition(file.path(lsp_output_dir, "rgn_fishing_sites_condition.csv")),
    title = "LSP: Iconic Fishing Spots"
  ),
  "lsp_archaeological_sites" = list(
    data = read_rgn_condition(file.path(lsp_output_dir, "rgn_ewg_archaeological_sites_condition.csv")),
    title = "LSP: Archaeological Sites"
  ),
  "lsp_special_habitats" = list(
    data = read_rgn_condition(file.path(lsp_output_dir, "rgn_special_habitats_condition.csv")),
    title = "LSP: Special Habitats"
  ),
  "lsp_all_other_special_places" = list(
    data = read_rgn_condition(file.path(lsp_output_dir, "rgn_final_other_protections_condition.csv")),
    title = "LSP: All other Special Places"
  )
)
```

# Choropleth helper

Join every score table to the full set of nine marine regions so missing categories stay NA (solid white), not dropped.

```{r map-helper}
join_scores_to_ohi <- function(score_df) {
  ohi_marine %>%
    left_join(score_df, by = "rgn_id") %>%
    mutate(
      rgn_label = unname(rgn_labels[rgn_id]),
      score = ifelse(is.nan(score), NA_real_, score)
    )
}

make_lsp_choropleth <- function(
    score_df,
    legend_title = lsp_legend_title,
    plot_title = NULL,
    label_scores = FALSE
) {
  dat <- join_scores_to_ohi(score_df)

  p <- ggplot() +
    geom_sf(
      data = states,
      fill = "#d6cac0",
      color = "grey10",
      linewidth = 0.1
    ) +
    geom_sf(
      data = dat,
      aes(fill = score),
      color = "black",
      linewidth = 0.2
    ) +
    scale_fill_gradientn(
      colours = lsp_palette,
      limits = c(0, 100),
      na.value = lsp_na_fill,
      name = legend_title,
      breaks = seq(0, 100, by = 25)
    ) +
    labs(
      title = plot_title,
      caption = "White = no score (NA)"
    ) +
    coord_sf(expand = FALSE) +
    theme_void(base_size = 12) +
    theme(
      panel.background = element_rect(fill = "white", color = NA),
      plot.background = element_rect(fill = "white", color = NA),
      plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
      legend.position = "right",
      legend.key.height = unit(0.9, "cm"),
      legend.key.width = unit(0.45, "cm"),
      legend.title = element_text(size = 11),
      legend.text = element_text(size = 9),
      plot.caption = element_text(size = 9, hjust = 0, color = "grey30"),
      plot.margin = margin(10, 10, 10, 10)
    )

  if (label_scores) {
    p <- p +
      geom_sf_text(
        data = dat %>%
          mutate(label = ifelse(is.na(score), "NA", as.character(round(score)))),
        aes(label = label),
        size = 3,
        color = "grey10",
        fun.geometry = function(g) st_point_on_surface(st_make_valid(g))
      )
  }

  p
}

save_lsp_map <- function(gg, stem, width = 8, height = 7) {
  png_path <- file.path(viz_dir, paste0(stem, ".png"))
  pdf_path <- file.path(viz_dir, paste0(stem, ".pdf"))
  ggsave(png_path, gg, width = width, height = height, dpi = 300, bg = "white")
  tryCatch(
    ggsave(pdf_path, gg, width = width, height = height, device = cairo_pdf, bg = "white"),
    error = function(e) {
      ggsave(pdf_path, gg, width = width, height = height, device = "pdf", bg = "white")
    }
  )
  message("Wrote:\n  ", png_path, "\n  ", pdf_path)
  invisible(list(png = png_path, pdf = pdf_path))
}
```

# Build, display, and save maps

```{r maps, fig.width=8, fig.height=7}
saved <- list()

for (stem in names(score_layers)) {
  layer <- score_layers[[stem]]
  gg <- make_lsp_choropleth(
    layer$data,
    legend_title = lsp_legend_title,
    plot_title = layer$title,
    label_scores = FALSE
  )
  print(gg)
  saved[[stem]] <- save_lsp_map(gg, stem)
}
```

# Multi-panel of category maps

```{r category-panel, fig.width=12, fig.height=10}
cat_long <- bind_rows(
  lapply(names(score_layers), function(stem) {
    if (stem == "lsp_overall_2025") return(NULL)
    score_layers[[stem]]$data %>%
      mutate(
        layer = stem,
        panel_title = score_layers[[stem]]$title
      )
  })
)

panel_sf <- ohi_marine %>%
  select(rgn_id) %>%
  st_drop_geometry() %>%
  tidyr::crossing(panel_title = unique(cat_long$panel_title)) %>%
  left_join(cat_long, by = c("rgn_id", "panel_title")) %>%
  left_join(ohi_marine, by = "rgn_id") %>%
  st_as_sf() %>%
  mutate(score = ifelse(is.nan(score), NA_real_, score))

gg_panel <- ggplot() +
  geom_sf(
    data = states,
    fill = "#d6cac0",
    color = "grey10",
    linewidth = 0.05
  ) +
  geom_sf(
    data = panel_sf,
    aes(fill = score),
    color = "black",
    linewidth = 0.15
  ) +
  scale_fill_gradientn(
    colours = lsp_palette,
    limits = c(0, 100),
    na.value = lsp_na_fill,
    name = lsp_legend_title,
    breaks = seq(0, 100, by = 25)
  ) +
  facet_wrap(~ panel_title, ncol = 3) +
  labs(caption = "White = no score (NA)") +
  coord_sf(expand = FALSE) +
  theme_void(base_size = 11) +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    strip.text = element_text(size = 9, face = "bold"),
    legend.position = "right",
    plot.caption = element_text(size = 9, hjust = 0, color = "grey30"),
    plot.margin = margin(8, 8, 8, 8)
  )

gg_panel
save_lsp_map(gg_panel, "lsp_categories_panel", width = 12, height = 10)
```

# Process maps: scoring methods by LSP type

Static **method illustrations** for each LSP prep script (2–7). Review each object in the console with `print()`; **nothing is written to disk** until you set `save_process_maps <- TRUE` below.

```{r process-helpers}
# --- toggle: leave FALSE while exploring in the console ---
save_process_maps <- FALSE

show_process_map <- function(gg, stem, width = 9, height = 6.5) {
  if (is.null(gg)) {
    message("No plot object for ", stem)
    return(invisible(NULL))
  }
  print(gg)
  if (isTRUE(save_process_maps)) {
    save_lsp_map(gg, stem, width = width, height = height)
    message("Saved: ", stem)
  } else {
    message("Console only (not saved). Set save_process_maps <- TRUE to write PNG/PDF for ", stem)
  }
  invisible(gg)
}

process_theme <- function(base_size = 11) {
  theme_minimal(base_size = base_size) +
    theme(
      legend.position = "bottom",
      legend.box = "vertical",
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
      plot.title = element_text(face = "bold", size = base_size + 1),
      plot.subtitle = element_text(size = base_size - 1, color = "grey30"),
      plot.caption = element_text(size = base_size - 2, hjust = 0, color = "grey40")
    )
}

as_bbox <- function(xmin, xmax, ymin, ymax, crs = 4326) {
  st_bbox(c(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax), crs = st_crs(crs))
}

# Crop safely even when layer CRS differs from the target bbox CRS
crop_sf <- function(x, bb) {
  if (is.null(x) || nrow(x) == 0) return(x)
  tryCatch(
    {
      x <- st_make_valid(x)
      if (is.na(st_crs(x))) st_crs(x) <- 4326
      bb_poly <- st_as_sfc(st_bbox(bb))
      if (st_crs(bb_poly) != st_crs(x)) {
        bb_poly <- st_transform(bb_poly, st_crs(x))
      }
      suppressWarnings(st_crop(x, bb_poly))
    },
    error = function(e) x
  )
}

filter_points_in_bbox <- function(pts, bb) {
  if (is.null(pts) || nrow(pts) == 0) return(pts)
  pts <- st_transform(pts, 4326)
  coords <- st_coordinates(pts)
  pts %>%
    mutate(.x = coords[, 1], .y = coords[, 2]) %>%
    filter(
      .x >= bb["xmin"], .x <= bb["xmax"],
      .y >= bb["ymin"], .y <= bb["ymax"]
    ) %>%
    dplyr::select(-.x, -.y)
}

# Always WGS84 for process maps (choropleth `states` may be in OHI projected CRS)
process_states <- tryCatch(
  {
    src <- if (exists("states") && !is.null(states)) {
      states
    } else {
      safe_st_read(here("spatial/ohi_regions/states.shp"))
    }
    if (is.null(src)) NULL else st_transform(src, 4326)
  },
  error = function(e) NULL
)

process_anp_sf <- safe_st_read(first_path(
  file.path(goc_root, "spatial/from_hem_POEM_ANP/232_ANP-ITRF08_16062025/232_ANP-ITRF08_16062025.shp")
))
if (!is.null(process_anp_sf)) {
  process_anp_sf <- process_anp_sf %>%
    st_zm(drop = TRUE, what = "ZM") %>%
    st_make_valid() %>%
    st_transform(4326) %>%
    st_simplify(dTolerance = 0.002, preserveTopology = TRUE)
}

all_protected_sf <- safe_st_read(first_path(
  file.path(goc_root, "spatial/protected_areas_wdpa_conanp_and_hem/all_goc_protected_areas_v2_2025.shp"),
  file.path(goc_root, "spatial/protected _areas_wdpa_and_hem/all_goc_protected_areas_2025.shp")
))
if (!is.null(all_protected_sf)) {
  all_protected_sf <- all_protected_sf %>%
    st_zm(drop = TRUE, what = "ZM") %>%
    st_make_valid() %>%
    st_transform(4326)
  nm <- names(all_protected_sf)
  name_col <- nm[tolower(nm) %in% c("name", "nombre", "orig_name", "wdpa_pid")][1]
  if (!is.na(name_col) && name_col != "name") {
    all_protected_sf <- all_protected_sf %>% rename(name = !!sym(name_col))
  }
}

ewg_coords_v4 <- safe_read_csv(first_path(
  file.path(lsp_prep_root, "identifying_lasting_special_places_collapsed_lsp_protections_v4.csv"),
  file.path(mazu_lsp_root, "identifying_lasting_special_places_collapsed_lsp_protections_v4.csv"),
  file.path(local_lsp_root, "identifying_lasting_special_places_collapsed_lsp_protections_v4.csv")
))
if (!is.null(ewg_coords_v4)) {
  names(ewg_coords_v4) <- names(ewg_coords_v4) %>%
    tolower() %>%
    str_replace_all("[^a-z0-9]+", "_") %>%
    str_replace_all("^_|_$", "") %>%
    str_replace_all("_+", "_")
  if (!"lasting_special_place_lsp" %in% names(ewg_coords_v4)) {
    hit <- names(ewg_coords_v4)[str_detect(names(ewg_coords_v4), "lasting_special_place")]
    if (length(hit) > 0) {
      ewg_coords_v4 <- ewg_coords_v4 %>% rename(lasting_special_place_lsp = !!sym(hit[[1]]))
    }
  }
  ewg_coords_v4 <- ewg_coords_v4 %>%
    mutate(
      lat = suppressWarnings(as.numeric(lat)),
      long = suppressWarnings(as.numeric(long)),
      lasting_special_place_lsp = str_squish(as.character(lasting_special_place_lsp))
    ) %>%
    filter(!is.na(lat), !is.na(long), !is.na(lasting_special_place_lsp))
}

# SIMEC raw table (optional; used when we want pure i-efectividad fill)
simec_raw <- safe_read_csv(first_path(
  file.path(goc_root, "_raw_data/SIMEC/identifying_lasting_special_places_SIMEC.csv"),
  file.path(lsp_prep_root, "identifying_lasting_special_places_SIMEC.csv")
))
if (!is.null(simec_raw)) {
  names(simec_raw) <- tolower(names(simec_raw))
}
```

## Script 2 — Special protected areas (Gulf-wide condition choropleth)

EWG special protected areas joined to polygons and filled with prep **`overall_condition`**:

**SIMEC i-efectividad when available → else Ramsar management-plan rule → else leave unshown (NA / no fill for that record).**

```{r process-protected-areas, fig.width=9, fig.height=7}
pa_cond <- safe_read_csv(file.path(lsp_output_dir, "protected_area_condition.csv"))

gg_pa_process <- NULL
if (!is.null(pa_cond) && !is.null(all_protected_sf)) {
  pa_cond <- pa_cond %>%
    mutate(
      overall_condition = suppressWarnings(as.numeric(overall_condition)),
      join_name = str_squish(str_replace_all(as.character(wdpa_orig_name), "\\n", " "))
    )

  pa_sf <- all_protected_sf %>%
    mutate(join_name = str_squish(str_replace_all(as.character(name), "\\n", " "))) %>%
    inner_join(
      pa_cond %>%
        dplyr::select(join_name, overall_condition, lasting_special_place_description, i_efectividad_name),
      by = "join_name",
      relationship = "many-to-many"
    ) %>%
    filter(!is.na(overall_condition))

  # join failures are common if names differ slightly — report for console review
  n_matched <- length(unique(pa_sf$join_name))
  n_requested <- length(unique(pa_cond$join_name[!is.na(pa_cond$join_name) & pa_cond$join_name != ""]))
  message("PA polygons matched: ", n_matched, " of ", n_requested, " named protected areas in condition table")

  gg_pa_process <- ggplot() +
    {
      if (!is.null(process_states) && nrow(process_states) > 0) {
        geom_sf(data = process_states, fill = "#d6cac0", color = "grey40", linewidth = 0.12)
      }
    } +
    geom_sf(
      data = pa_sf,
      aes(fill = overall_condition),
      color = "grey20",
      linewidth = 0.15,
      alpha = 0.9
    ) +
    scale_fill_gradientn(
      colours = lsp_palette,
      limits = c(0, 100),
      na.value = "white",
      name = "Overall\ncondition\n(0–100)"
    ) +
    coord_sf(expand = FALSE) +
    labs(
      title = "Process: Special Protected Areas (script 2)",
      subtitle = "Gulf-wide: EWG special protected areas colored by overall condition",
      caption = paste0(
        "Fill = overall_condition (SIMEC i-efectividad when available; else Ramsar plan rule; ",
        "records with no score are omitted). Matched ", n_matched, "/", n_requested, " named PAs."
      )
    ) +
    process_theme()
}

gg_pa_process  # console review first
show_process_map(gg_pa_process, "process_special_protected_areas", width = 9, height = 7)
```

## Script 3 — Priority dive sites (keep Espíritu Santo example)

Same decision rule as the EWG page: inside scored ANP → SIMEC/RMPP; outside → default **58**.

```{r process-dive-sites, fig.width=9, fig.height=6.5}
dive_cond <- safe_read_csv(file.path(lsp_output_dir, "dive_sites_condition.csv"))
dive_gpkg <- safe_st_read(first_path(
  file.path(lsp_int_dir, "dive_sites_w_priority_goc_rgnid.gpkg"),
  file.path(mazu_lsp_root, "int/dive_sites_w_priority_goc_rgnid.gpkg"),
  file.path(local_lsp_root, "int/dive_sites_w_priority_goc_rgnid.gpkg")
))

gg_dive_process <- NULL
if (!is.null(dive_cond) && !is.null(dive_gpkg)) {
  dive_cond <- dive_cond %>%
    mutate(
      overall_condition = suppressWarnings(as.numeric(overall_condition)),
      pa_condition = suppressWarnings(as.numeric(protected_area_condition)),
      nombre_sitio_de_buceo = str_squish(as.character(nombre_sitio_de_buceo)),
      scoring_rule = if_else(
        !is.na(pa_condition),
        "Inside scored protected area (SIMEC or RMPP)",
        "Outside scored protected area (default 58)"
      )
    ) %>%
    distinct(nombre_sitio_de_buceo, .keep_all = TRUE)

  dives <- dive_gpkg %>%
    st_transform(4326) %>%
    mutate(nombre_sitio_de_buceo = str_squish(as.character(nombre_sitio_de_buceo))) %>%
    left_join(
      dive_cond %>% dplyr::select(nombre_sitio_de_buceo, overall_condition, scoring_rule),
      by = "nombre_sitio_de_buceo"
    )

  bb <- as_bbox(-110.50, -110.15, 24.10, 24.55)
  pa_focus <- crop_sf(if (!is.null(process_anp_sf)) process_anp_sf else all_protected_sf, bb)
  dive_focus <- filter_points_in_bbox(dives, bb)

  if (!is.null(dive_focus) && nrow(dive_focus) > 0) {
    gg_dive_process <- ggplot() +
      {
        if (!is.null(pa_focus) && nrow(pa_focus) > 0) {
          geom_sf(data = pa_focus, fill = "#98a25a", color = "gray30", linewidth = 0.2, alpha = 0.45)
        }
      } +
      geom_sf(
        data = dive_focus,
        aes(color = scoring_rule, shape = scoring_rule),
        size = 2.4,
        alpha = 0.95
      ) +
      scale_color_manual(
        name = "Dive site scoring rule",
        values = c(
          "Inside scored protected area (SIMEC or RMPP)" = "#C03729",
          "Outside scored protected area (default 58)" = "#333333"
        ),
        na.value = "gray50"
      ) +
      scale_shape_manual(
        name = "Dive site scoring rule",
        values = c(
          "Inside scored protected area (SIMEC or RMPP)" = 16,
          "Outside scored protected area (default 58)" = 17
        ),
        na.value = 4
      ) +
      coord_sf(xlim = c(bb["xmin"], bb["xmax"]), ylim = c(bb["ymin"], bb["ymax"]), expand = FALSE) +
      labs(
        title = "Process: Priority Dive Sites (script 3)",
        subtitle = "Espíritu Santo / Bahía de La Paz — inside scored ANPs vs unprotected default",
        caption = "Red circles: SIMEC/RMPP for containing PA. Black triangles: default 58."
      ) +
      process_theme() +
      guides(color = guide_legend(nrow = 2), shape = guide_legend(nrow = 2))
  }
}

gg_dive_process
show_process_map(gg_dive_process, "process_priority_dive_sites")
```

## Script 5 — Iconic fishing spots (Gulf-wide points) + ZRP overview

Gulf-wide map of all iconic fishing points, dive-style binary symbology:

- **Red circles** = site is inside a ZRP (`in_zrp`; condition 100)
- **Black triangles** = otherwise (ANP SIMEC / TURF / other path in prep)

**TURFs** are not drawn as polygons. For EWG discussion use **Aceves-Bueno et al. 2023**. Light blue wash under the points shows ZRP polygons for context.

```{r process-fishing-gulf, fig.width=9, fig.height=7}
fish_cond <- safe_read_csv(file.path(lsp_output_dir, "fishing_sites_condition.csv"))
zrp_sf <- safe_st_read(first_path(
  file.path(goc_root, "_raw_data/TNC_geoportal_fishing_refuge_zones/d2026/Zonas_Refugios_Pesqueros_-8805549763229037385.geojson")
))
if (!is.null(zrp_sf)) {
  zrp_sf <- zrp_sf %>% st_make_valid() %>% st_transform(4326)
}

gg_fish_loreto <- NULL
if (!is.null(fish_cond) && !is.null(ewg_coords_v4)) {
  fish_cond <- fish_cond %>%
    mutate(
      lasting_special_place_lsp = str_squish(as.character(lasting_special_place_lsp)),
      overall_condition = suppressWarnings(as.numeric(overall_condition)),
      in_zrp = as.logical(in_zrp),
      # Dive-style binary: ZRP vs everything else (TURF/ANP geography not mapped here)
      scoring_rule = if_else(
        !is.na(in_zrp) & in_zrp,
        "Inside ZRP (condition 100)",
        "Outside ZRP (ANP / TURF / other path)"
      )
    )

  fish_sf <- ewg_coords_v4 %>%
    mutate(lasting_special_place_lsp = str_squish(as.character(lasting_special_place_lsp))) %>%
    inner_join(
      fish_cond %>% dplyr::select(lasting_special_place_lsp, overall_condition, scoring_rule, in_zrp),
      by = "lasting_special_place_lsp"
    ) %>%
    st_as_sf(coords = c("long", "lat"), crs = 4326, remove = FALSE)

  message("Iconic fishing sites plotted (Gulf-wide): ", nrow(fish_sf))
  message(
    "  In ZRP: ", sum(fish_sf$scoring_rule == "Inside ZRP (condition 100)", na.rm = TRUE),
    " | Outside ZRP: ", sum(fish_sf$scoring_rule == "Outside ZRP (ANP / TURF / other path)", na.rm = TRUE)
  )

  if (nrow(fish_sf) > 0) {
    gg_fish_loreto <- ggplot() +
      {
        if (!is.null(process_states) && nrow(process_states) > 0) {
          geom_sf(data = process_states, fill = "#d6cac0", color = "grey40", linewidth = 0.12)
        }
      } +
      {
        if (!is.null(zrp_sf) && nrow(zrp_sf) > 0) {
          geom_sf(
            data = zrp_sf,
            fill = "#5dade2",
            color = "#1a5276",
            linewidth = 0.15,
            alpha = 0.35
          )
        }
      } +
      geom_sf(
        data = fish_sf,
        aes(color = scoring_rule, shape = scoring_rule),
        size = 2.8,
        alpha = 0.95
      ) +
      scale_color_manual(
        name = "Fishing site scoring rule",
        values = c(
          "Inside ZRP (condition 100)" = "#C03729",
          "Outside ZRP (ANP / TURF / other path)" = "#333333"
        )
      ) +
      scale_shape_manual(
        name = "Fishing site scoring rule",
        values = c(
          "Inside ZRP (condition 100)" = 16,
          "Outside ZRP (ANP / TURF / other path)" = 17
        )
      ) +
      coord_sf(expand = FALSE) +
      labs(
        title = "Process: Iconic Fishing Spots (script 5)",
        subtitle = "Gulf-wide: red = inside ZRP; black = outside ZRP. Blue wash = ZRP polygons.",
        caption = paste0(
          "Prep hierarchy: ZRP → TURF (Aceves-Bueno et al. 2023; not plotted) → ANP SIMEC. ",
          "Symbology highlights the top rule (ZRP) only."
        )
      ) +
      process_theme() +
      guides(color = guide_legend(nrow = 2), shape = guide_legend(nrow = 2))
  }
}

gg_fish_loreto
show_process_map(gg_fish_loreto, "process_iconic_fishing_gulf", width = 9, height = 7)
```

### Fishing refuges (ZRP) Gulf-wide — where the tools are

Separate overview so experts can see **where ZRPs are** without fishing points. TURF geography will be discussed with the **Aceves-Bueno et al. 2023** map (not plotted in this script).

```{r process-zrp-overview, fig.width=9, fig.height=7}
gg_zrp_overview <- NULL
if (!is.null(zrp_sf) && nrow(zrp_sf) > 0) {
  gg_zrp_overview <- ggplot() +
    {
      if (!is.null(process_states) && nrow(process_states) > 0) {
        geom_sf(data = process_states, fill = "#d6cac0", color = "grey40", linewidth = 0.12)
      }
    } +
    {
      if (!is.null(all_protected_sf) && nrow(all_protected_sf) > 0) {
        geom_sf(data = all_protected_sf, fill = "#98a25a", color = NA, alpha = 0.15)
      }
    } +
    geom_sf(
      data = zrp_sf,
      fill = "#5dade2",
      color = "#1a5276",
      linewidth = 0.2,
      alpha = 0.55
    ) +
    coord_sf(expand = FALSE) +
    labs(
      title = "Context: Zonas de Refugio Pesquero (ZRP) in the GOC",
      subtitle = "Fishing refuges used for ZRP = 100 scoring (script 5)",
      caption = "Green wash = ANP layer for context. TURFs: use Aceves-Bueno et al. 2023 figure for EWG discussion (not plotted here)."
    ) +
    process_theme()
}

gg_zrp_overview
show_process_map(gg_zrp_overview, "process_zrp_overview", width = 9, height = 7)
```

## Script 4 — Archaeological sites (table + optional buffer diagram)

There are only ~20 scored sites. A **kable table** is clearer than a cluttered Gulf choropleth. Columns show which hierarchical rule applied.

Prep buffers (from script 4): **10 km** around EWG points for the INAH directory join, **500 m** for the ANP join. Full site score = INAH 100 if matched, else PA SIMEC path, else codified floor **30**.

```{r process-arch-table}
arch_cond <- safe_read_csv(file.path(lsp_output_dir, "ewg_archaeological_sites_condition.csv"))

arch_table <- NULL
if (!is.null(arch_cond)) {
  arch_table <- arch_cond %>%
    mutate(
      overall_condition = suppressWarnings(as.numeric(overall_condition)),
      pa_condition = suppressWarnings(as.numeric(protected_area_condition)),
      in_inah = as.logical(in_INAH_directory),
      scoring_rule = case_when(
        !is.na(in_inah) & in_inah ~ "INAH directory (100)",
        !is.na(pa_condition) & pa_condition > 0 ~ "ANP / special protection (SIMEC path)",
        !is.na(overall_condition) & overall_condition <= 35 ~ "Codified legal floor (30)",
        TRUE ~ "Other / special case"
      ),
      site = str_trunc(str_squish(as.character(lasting_special_place_lsp)), 70)
    ) %>%
    dplyr::select(
      rgn_id,
      site,
      overall_condition,
      scoring_rule,
      in_INAH = in_inah,
      nombre_INAH,
      protected_area_condition = pa_condition
    ) %>%
    arrange(rgn_id, desc(overall_condition))

  # print for console; knitr will also render the table in the HTML doc
  print(arch_table, n = Inf)
  knitr::kable(
    arch_table %>% dplyr::select(rgn_id, site, overall_condition, scoring_rule),
    caption = "Archaeological / cultural LSPs: condition and hierarchical scoring rule (script 4)"
  )
}
```

Optional: static map of archaeological points with **500 m** (ANP) and **10 km** (INAH) buffers. Extent from the sites themselves in WGS84 (Gulf-wide), not a hard-coded crop that can break when underlays are still in a projected CRS.

```{r process-arch-buffer-map, fig.width=9, fig.height=7}
gg_arch_buffer <- NULL
if (!is.null(arch_cond) && !is.null(ewg_coords_v4)) {
  arch_rule <- arch_cond %>%
    mutate(
      lasting_special_place_lsp = str_squish(as.character(lasting_special_place_lsp)),
      overall_condition = suppressWarnings(as.numeric(overall_condition)),
      pa_condition = suppressWarnings(as.numeric(protected_area_condition)),
      in_inah = as.logical(in_INAH_directory),
      scoring_rule = case_when(
        !is.na(in_inah) & in_inah ~ "INAH directory (100)",
        !is.na(pa_condition) & pa_condition > 0 ~ "ANP / special protection (SIMEC path)",
        !is.na(overall_condition) & overall_condition <= 35 ~ "Codified legal floor (30)",
        TRUE ~ "Other / special case"
      )
    ) %>%
    dplyr::select(lasting_special_place_lsp, overall_condition, scoring_rule)

  arch_pts <- ewg_coords_v4 %>%
    mutate(lasting_special_place_lsp = str_squish(as.character(lasting_special_place_lsp))) %>%
    inner_join(arch_rule, by = "lasting_special_place_lsp") %>%
    st_as_sf(coords = c("long", "lat"), crs = 4326, remove = FALSE)

  if (nrow(arch_pts) > 0) {
    # Buffers in UTM (meters), then back to WGS84 for display
    pts_proj <- st_transform(arch_pts, 32612)
    buf500 <- st_buffer(pts_proj, dist = 500) %>% st_transform(4326)
    buf10k <- st_buffer(pts_proj, dist = 10000) %>% st_transform(4326)

    # Extent from all sites + 10 km buffer ring, with pad; south enough for northern Jalisco
    bb_c <- st_bbox(buf10k)
    pad <- 0.15
    bb <- as_bbox(
      bb_c["xmin"] - pad,
      bb_c["xmax"] + pad,
      min(as.numeric(bb_c["ymin"]) - pad, 20.5),
      bb_c["ymax"] + pad
    )
    # Fall back to Gulf envelope if something is degenerate
    if (!all(is.finite(c(bb["xmin"], bb["xmax"], bb["ymin"], bb["ymax"]))) ||
        (bb["xmax"] - bb["xmin"]) < 0.2 ||
        (bb["ymax"] - bb["ymin"]) < 0.2) {
      bb <- as_bbox(-115.5, -105.5, 20.5, 32.5)
    }

    states_f <- process_states
    if (!is.null(states_f) && nrow(states_f) > 0) {
      states_f <- st_transform(states_f, 4326)
    }
    anp_src <- if (!is.null(process_anp_sf)) process_anp_sf else all_protected_sf
    anp_f <- anp_src
    if (!is.null(anp_f) && nrow(anp_f) > 0) {
      anp_f <- st_transform(anp_f, 4326)
    }

    message(
      "Arch sites plotted: ", nrow(arch_pts),
      "; map lon [", round(bb["xmin"], 2), ", ", round(bb["xmax"], 2),
      "], lat [", round(bb["ymin"], 2), ", ", round(bb["ymax"], 2), "]"
    )

    gg_arch_buffer <- ggplot() +
      {
        if (!is.null(states_f) && nrow(states_f) > 0) {
          geom_sf(data = states_f, fill = "#d6cac0", color = "grey40", linewidth = 0.12)
        }
      } +
      {
        if (!is.null(anp_f) && nrow(anp_f) > 0) {
          geom_sf(data = anp_f, fill = "#98a25a", color = NA, alpha = 0.18)
        }
      } +
      geom_sf(data = buf10k, fill = NA, color = "#5dade2", linewidth = 0.3, linetype = "dashed") +
      geom_sf(data = buf500, fill = NA, color = "#C03729", linewidth = 0.4) +
      geom_sf(
        data = arch_pts,
        aes(color = scoring_rule, shape = scoring_rule),
        size = 2.4,
        alpha = 0.95
      ) +
      scale_color_manual(
        name = "Arch site scoring rule",
        values = c(
          "INAH directory (100)" = "#1a5276",
          "ANP / special protection (SIMEC path)" = "#C03729",
          "Codified legal floor (30)" = "#333333",
          "Other / special case" = "#7f8c8d"
        )
      ) +
      scale_shape_manual(
        name = "Arch site scoring rule",
        values = c(
          "INAH directory (100)" = 15,
          "ANP / special protection (SIMEC path)" = 16,
          "Codified legal floor (30)" = 17,
          "Other / special case" = 18
        )
      ) +
      coord_sf(
        xlim = c(bb["xmin"], bb["xmax"]),
        ylim = c(bb["ymin"], bb["ymax"]),
        expand = FALSE,
        crs = st_crs(4326)
      ) +
      labs(
        title = "Process: Archaeological sites (script 4)",
        subtitle = "Points by scoring rule; red ring = 500 m (ANP join); blue dashed = 10 km (INAH join)",
        caption = "Light green wash = ANPs. Map extent is from all scored sites in WGS84 (not a mid-Baja crop)."
      ) +
      process_theme() +
      guides(color = guide_legend(nrow = 2), shape = guide_legend(nrow = 2))
  }
}

gg_arch_buffer
show_process_map(gg_arch_buffer, "process_archaeological_buffers", width = 9, height = 7)
```

### Upper Gulf zoom — QC for codified-floor (black) points vs ANPs

Prep ANP rule uses a **500 m buffer** + intersection with `all_protected`. Sites scored as codified floor (**30**) should **not** touch an ANP after that buffer; if they do, they should take the ANP SIMEC path instead (unless PA condition collapses to 0—for example a special Ramsar rule—then floor 30 can still appear).

This zoom shows Upper Gulf points with ANP polygons, **500 m rings**, short labels, and an orange halo when a **floor-scored** site’s buffer still intersects an ANP.

```{r process-arch-upper-gulf, fig.width=9, fig.height=7}
gg_arch_upper_gulf <- NULL

if (!is.null(arch_cond) && !is.null(ewg_coords_v4)) {
  arch_rule_ug <- arch_cond %>%
    mutate(
      lasting_special_place_lsp = str_squish(as.character(lasting_special_place_lsp)),
      overall_condition = suppressWarnings(as.numeric(overall_condition)),
      pa_condition = suppressWarnings(as.numeric(protected_area_condition)),
      in_inah = as.logical(in_INAH_directory),
      scoring_rule = case_when(
        !is.na(in_inah) & in_inah ~ "INAH directory (100)",
        !is.na(pa_condition) & pa_condition > 0 ~ "ANP / special protection (SIMEC path)",
        !is.na(overall_condition) & overall_condition <= 35 ~ "Codified legal floor (30)",
        TRUE ~ "Other / special case"
      )
    ) %>%
    dplyr::select(
      lasting_special_place_lsp, overall_condition, pa_condition, scoring_rule
    )

  arch_all <- ewg_coords_v4 %>%
    mutate(lasting_special_place_lsp = str_squish(as.character(lasting_special_place_lsp))) %>%
    inner_join(arch_rule_ug, by = "lasting_special_place_lsp") %>%
    st_as_sf(coords = c("long", "lat"), crs = 4326, remove = FALSE)

  # Upper Gulf / Peñasco – El Golfo / western Upper Gulf shore
  bb_ug <- as_bbox(-115.15, -113.15, 30.95, 32.15)
  arch_ug <- filter_points_in_bbox(arch_all, bb_ug)

  if (!is.null(arch_ug) && nrow(arch_ug) > 0) {
    anp_src <- if (!is.null(all_protected_sf)) {
      all_protected_sf
    } else {
      process_anp_sf
    }
    if (!is.null(anp_src) && nrow(anp_src) > 0) {
      anp_src <- st_transform(st_make_valid(anp_src), 4326)
    }

    # Independent spatial QC: does the prep 500 m buffer touch any ANP?
    pts_proj <- st_transform(arch_ug, 32612)
    buf500_ug <- st_buffer(pts_proj, dist = 500) %>% st_transform(4326)
    touches_anp <- rep(FALSE, nrow(arch_ug))
    anp_names_hit <- rep(NA_character_, nrow(arch_ug))

    if (!is.null(anp_src) && nrow(anp_src) > 0) {
      anp_u <- anp_src %>% st_union() %>% st_make_valid()
      hits <- suppressWarnings(st_intersects(buf500_ug, anp_u, sparse = TRUE))
      touches_anp <- lengths(hits) > 0

      # best-effort name of first intersecting ANP polygon (for console table)
      hit_poly <- suppressWarnings(st_intersects(buf500_ug, anp_src, sparse = TRUE))
      nm_col <- if ("name" %in% names(anp_src)) "name" else names(anp_src)[1]
      anp_names_hit <- vapply(hit_poly, function(ix) {
        if (length(ix) == 0) return(NA_character_)
        paste(unique(as.character(anp_src[[nm_col]][ix])), collapse = "; ")
      }, character(1))
    }

    arch_ug <- arch_ug %>%
      mutate(
        touches_anp_500m = touches_anp,
        anp_name_hit = anp_names_hit,
        short_label = lasting_special_place_lsp %>%
          str_replace("\\s*\\(.*$", "") %>%
          str_trunc(28),
        # QC flag for floor sites that would be expected to take ANP path if SIMEC > 0
        qc_issue = scoring_rule == "Codified legal floor (30)" & touches_anp_500m,
        plot_class = case_when(
          qc_issue ~ "Floor score BUT buffer touches ANP (review)",
          TRUE ~ scoring_rule
        )
      )

    message("Upper Gulf arch sites: ", nrow(arch_ug))
    print(
      st_drop_geometry(arch_ug) %>%
        dplyr::select(
          short_label, scoring_rule, overall_condition, pa_condition,
          touches_anp_500m, anp_name_hit, qc_issue
        ) %>%
        arrange(desc(qc_issue), scoring_rule, short_label)
    )
    n_issue <- sum(arch_ug$qc_issue, na.rm = TRUE)
    message(
      "Codified-floor sites whose 500 m buffer still touches an ANP: ", n_issue,
      if (n_issue > 0) " (shown with orange stroke)" else " (none)"
    )

    states_ug <- process_states
    if (!is.null(states_ug) && nrow(states_ug) > 0) {
      states_ug <- st_transform(states_ug, 4326)
    }
    anp_ug <- if (!is.null(anp_src)) crop_sf(anp_src, bb_ug) else NULL

    gg_arch_upper_gulf <- ggplot() +
      {
        if (!is.null(states_ug) && nrow(states_ug) > 0) {
          geom_sf(data = states_ug, fill = "#d6cac0", color = "grey40", linewidth = 0.15)
        }
      } +
      {
        if (!is.null(anp_ug) && nrow(anp_ug) > 0) {
          geom_sf(
            data = anp_ug,
            fill = "#98a25a",
            color = "#4a5c2a",
            linewidth = 0.35,
            alpha = 0.4
          )
        }
      } +
      geom_sf(
        data = buf500_ug,
        fill = NA,
        color = "#C03729",
        linewidth = 0.45,
        alpha = 0.9
      ) +
      geom_sf(
        data = arch_ug,
        aes(color = plot_class, shape = plot_class),
        size = 3.2,
        stroke = 0.9,
        alpha = 0.98
      ) +
      geom_text(
        data = st_drop_geometry(arch_ug),
        aes(x = long, y = lat, label = short_label),
        size = 2.5,
        hjust = 0,
        nudge_x = 0.02,
        color = "grey15",
        check_overlap = TRUE
      ) +
      scale_color_manual(
        name = "Prep score / QC",
        values = c(
          "INAH directory (100)" = "#1a5276",
          "ANP / special protection (SIMEC path)" = "#C03729",
          "Codified legal floor (30)" = "#333333",
          "Other / special case" = "#7f8c8d",
          "Floor score BUT buffer touches ANP (review)" = "#e67e22"
        )
      ) +
      scale_shape_manual(
        name = "Prep score / QC",
        values = c(
          "INAH directory (100)" = 15,
          "ANP / special protection (SIMEC path)" = 16,
          "Codified legal floor (30)" = 17,
          "Other / special case" = 18,
          "Floor score BUT buffer touches ANP (review)" = 17
        )
      ) +
      coord_sf(
        xlim = c(bb_ug["xmin"], bb_ug["xmax"]),
        ylim = c(bb_ug["ymin"], bb_ug["ymax"]),
        expand = FALSE,
        crs = st_crs(4326)
      ) +
      labs(
        title = "Process: Archaeological sites — Upper Gulf zoom (script 4)",
        subtitle = "Red rings = 500 m ANP-join buffers. Green fill + outline = ANP polygons.",
        caption = paste0(
          "Black triangles = codified floor (30) in prep. Orange = floor score but 500 m buffer ",
          "intersects an ANP (possible mismatch or PA condition = 0). Check console table."
        )
      ) +
      process_theme() +
      guides(color = guide_legend(nrow = 3), shape = guide_legend(nrow = 3))
  } else {
    message("No archaeological points found in Upper Gulf window; adjust bb_ug if needed.")
  }
}

gg_arch_upper_gulf
show_process_map(gg_arch_upper_gulf, "process_archaeological_upper_gulf", width = 9, height = 7)
```

## Script 6 — Special habitats (how scoring works + visual)

### Confirmed from `script6_special_habitats.qmd`

1. **Each habitat feature** gets `overall_condition = pmax(protected_area_condition, nom_condition)`:
   - Coquina NOM floor = **30**
   - Coral / seagrass / mangrove NOM floor = **70**
2. **Protected-area test:** habitat geometry is buffered by **100 m**, then `st_intersects` against **non-dissolved** ANPs. If it **touches** a PA (after that buffer), it is flagged `in_protected_area` and receives that PA’s full SIMEC (or Ramsar plan rule)—**not** an area-weighted fraction of overlap. Multiple overlapping PAs are averaged with `mean()`.
3. **Regional special-habitats score:** first average feature scores **within each habitat type** for the region, then average those **type-level** means. So habitat **types present in the region are treated as equally important** (coral one unit, seagrass one unit, …)—not equal weight per individual patch. (Comment in code says “features,” but the combine step is mean of type rows.)

```{r process-habitats-method-figure, fig.width=10, fig.height=7}
hab_feat <- safe_read_csv(file.path(lsp_output_dir, "special_habitats_features_condition.csv"))
gg_hab_features <- NULL

if (!is.null(hab_feat)) {
  # console summary of equal-type aggregation recipe
  type_rgn <- hab_feat %>%
    mutate(
      overall_condition = suppressWarnings(as.numeric(overall_condition)),
      rgn_id = normalize_rgn_id(rgn_id)
    ) %>%
    group_by(rgn_id, habitat_type) %>%
    summarize(
      type_mean = mean(overall_condition, na.rm = TRUE),
      n_features = n(),
      .groups = "drop"
    )

  rgn_hab <- type_rgn %>%
    group_by(rgn_id) %>%
    summarize(
      condition = mean(type_mean, na.rm = TRUE),
      n_types = n(),
      n_features_total = sum(n_features),
      .groups = "drop"
    )

  message("Regional special-habitats condition = mean of per-type means (equal type weights):")
  print(rgn_hab)

  # Habitat colors (type-linked): seagrass green, mangrove brown, coral dark red, coquina purple-grey
  habitat_fills <- c(
    "Coquina Reef" = "#6E667A",
    "Coral Reef" = "#7A1212",
    "Mangrove" = "#7A4F2D",
    "Seagrass (Zostera marina)" = "#2F6B3A"
  )

  # Visual A: strip of type means contributing to each region
  gg_hab_types <- ggplot(type_rgn, aes(x = habitat_type, y = type_mean, fill = habitat_type)) +
    geom_col(width = 0.7, show.legend = FALSE) +
    geom_text(
      aes(label = sprintf("%.0f", type_mean)),
      vjust = -0.35,
      size = 2.6,
      color = "grey20"
    ) +
    scale_fill_manual(values = habitat_fills, na.value = "grey60") +
    facet_wrap(~ rgn_id, ncol = 3) +
    coord_cartesian(ylim = c(0, 108)) +
    labs(
      title = "Process: Special Habitats (type means by region)",
      subtitle = "Regional LSP habitat score = average of the type bars shown (equal type weight)",
      x = NULL,
      y = "Mean feature condition"
    ) +
    theme_minimal(base_size = 10) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
      strip.text = element_text(face = "bold", size = 8),
      plot.title = element_text(face = "bold", size = 11),
      plot.subtitle = element_text(size = 9, color = "grey30")
    )

  gg_hab_types
  show_process_map(gg_hab_types, "process_special_habitats_type_means", width = 10, height = 7)

  # Visual B: feature inventory — in PA vs NOM floor only (tabular console)
  rule_summary <- hab_feat %>%
    mutate(
      overall_condition = suppressWarnings(as.numeric(overall_condition)),
      pa_condition = suppressWarnings(as.numeric(protected_area_condition)),
      in_pa = as.logical(in_protected_area),
      scoring_rule = case_when(
        !is.na(in_pa) & in_pa & !is.na(pa_condition) ~ "Touches PA (full SIMEC / Ramsar, after 100 m buffer)",
        TRUE ~ "Outside PA (NOM habitat floor only)"
      )
    ) %>%
    count(habitat_type, scoring_rule)

  print(rule_summary)
  knitr::kable(rule_summary, caption = "Special habitat features by PA-touch vs NOM-only rule")
}

# Optional polygon map: coquina habitats showing PA-touch rule (Upper Gulf)
coquina_sf <- safe_st_read(first_path(
  file.path(lsp_int_dir, "coquina_reefs_condition.gpkg"),
  file.path(local_lsp_root, "int/coquina_reefs_condition.gpkg")
))
gg_hab_coquina <- NULL
if (!is.null(coquina_sf) && nrow(coquina_sf) > 0) {
  coquina_sf <- coquina_sf %>%
    st_zm(drop = TRUE, what = "ZM") %>%
    st_make_valid() %>%
    st_transform(4326)

  if ("in_protected_area" %in% names(coquina_sf)) {
    coquina_sf <- coquina_sf %>%
      mutate(in_pa = as.logical(in_protected_area))
  } else if ("protected_area_condition" %in% names(coquina_sf)) {
    coquina_sf <- coquina_sf %>%
      mutate(in_pa = !is.na(suppressWarnings(as.numeric(protected_area_condition))))
  } else {
    coquina_sf <- coquina_sf %>% mutate(in_pa = FALSE)
  }
  coquina_sf <- coquina_sf %>%
    mutate(
      overall_condition = suppressWarnings(as.numeric(overall_condition)),
      in_pa = ifelse(is.na(in_pa), FALSE, in_pa),
      pa_rule = factor(
        if_else(in_pa, "Yes → SIMEC / Ramsar then max with NOM", "No → NOM floor only (coquina 30)"),
        levels = c(
          "Yes → SIMEC / Ramsar then max with NOM",
          "No → NOM floor only (coquina 30)"
        )
      )
    )

  # Extent from coquina features (with pad), not a hard-coded window alone
  bb_c <- st_bbox(coquina_sf)
  pad <- 0.35
  bb <- as_bbox(
    bb_c["xmin"] - pad, bb_c["xmax"] + pad,
    bb_c["ymin"] - pad, bb_c["ymax"] + pad
  )
  # Guard against degenerate / empty extent
  if (!all(is.finite(c(bb["xmin"], bb["xmax"], bb["ymin"], bb["ymax"]))) ||
      (bb["xmax"] - bb["xmin"]) < 0.05 ||
      (bb["ymax"] - bb["ymin"]) < 0.05) {
    bb <- as_bbox(-114.9, -112.4, 29.3, 32.0)
  }

  states_c <- crop_sf(process_states, bb)
  anp_c <- crop_sf(
    if (!is.null(process_anp_sf)) process_anp_sf else all_protected_sf,
    bb
  )
  coq_c <- crop_sf(coquina_sf, bb)
  if (is.null(coq_c) || nrow(coq_c) == 0) coq_c <- coquina_sf

  message(
    "Coquina map: ", nrow(coq_c), " features; bbox lon [",
    round(bb["xmin"], 2), ", ", round(bb["xmax"], 2), "], lat [",
    round(bb["ymin"], 2), ", ", round(bb["ymax"], 2), "]"
  )

  gg_hab_coquina <- ggplot() +
    {
      if (!is.null(states_c) && nrow(states_c) > 0) {
        geom_sf(data = states_c, fill = "#d6cac0", color = "grey40", linewidth = 0.1)
      }
    } +
    {
      if (!is.null(anp_c) && nrow(anp_c) > 0) {
        geom_sf(
          data = anp_c,
          fill = "#98a25a",
          color = "gray30",
          alpha = 0.18,
          linewidth = 0.12
        )
      }
    } +
    geom_sf(
      data = coq_c,
      aes(fill = pa_rule),
      color = "grey15",
      linewidth = 0.2,
      alpha = 0.9
    ) +
    scale_fill_manual(
      name = "Touches PA\n(100 m buffer + intersects)?",
      values = c(
        "Yes → SIMEC / Ramsar then max with NOM" = "#C03729",
        "No → NOM floor only (coquina 30)" = "#6E667A"
      ),
      drop = FALSE
    ) +
    coord_sf(
      xlim = c(bb["xmin"], bb["xmax"]),
      ylim = c(bb["ymin"], bb["ymax"]),
      expand = FALSE,
      crs = st_crs(4326)
    ) +
    labs(
      title = "Process: Special Habitats (coquina × PA intersection)",
      subtitle = "Any intersection after 100 m buffer gets the full PA score for that match (not area-weighted)",
      caption = "Red = coquina feature touches a PA. Purple-grey = NOM floor only (30). Light green wash = ANPs."
    ) +
    process_theme()

  gg_hab_coquina
  show_process_map(gg_hab_coquina, "process_special_habitats_pa_touch", width = 9, height = 7)
}
```

## Script 7 — All other LSPs (dive-style map + codified protection table)

Outside scored ANPs, condition comes from the expert **codified protection spreadsheet** (`codified_protection_scores_v3.csv`). Sites that land in a PA use SIMEC/Ramsar like dives.

```{r process-other-map, fig.width=9, fig.height=7}
other_cond <- safe_read_csv(file.path(lsp_output_dir, "final_other_protections_condition.csv"))

gg_other_process <- NULL
if (!is.null(other_cond) && !is.null(ewg_coords_v4)) {
  other_cond <- other_cond %>%
    mutate(
      lsp_name = str_squish(as.character(lsp_name)),
      overall_condition = suppressWarnings(as.numeric(overall_condition)),
      pa_condition = suppressWarnings(as.numeric(protected_area_condition)),
      scoring_rule = if_else(
        !is.na(pa_condition) & pa_condition > 0,
        "Inside scored protected area (SIMEC / Ramsar)",
        "Outside PA → codified protection score (expert table)"
      )
    )

  other_sf <- ewg_coords_v4 %>%
    mutate(lsp_name = str_squish(as.character(lasting_special_place_lsp))) %>%
    inner_join(
      other_cond %>% dplyr::select(lsp_name, overall_condition, scoring_rule),
      by = "lsp_name"
    ) %>%
    st_as_sf(coords = c("long", "lat"), crs = 4326, remove = FALSE)

  message("Other LSPs plotted (Gulf-wide): ", nrow(other_sf))
  message(
    "  Inside PA: ",
    sum(other_sf$scoring_rule == "Inside scored protected area (SIMEC / Ramsar)", na.rm = TRUE),
    " | Outside PA (codified): ",
    sum(other_sf$scoring_rule == "Outside PA → codified protection score (expert table)", na.rm = TRUE)
  )

  # Gulf-wide: all points + full underlays in WGS84 (no tight bbox — avoids CRS zoom crashes)
  states_f <- process_states
  if (!is.null(states_f) && nrow(states_f) > 0) {
    states_f <- st_transform(states_f, 4326)
  }
  anp_src <- if (!is.null(process_anp_sf)) process_anp_sf else all_protected_sf
  anp_f <- anp_src
  if (!is.null(anp_f) && nrow(anp_f) > 0) {
    anp_f <- st_transform(anp_f, 4326)
  }

  if (nrow(other_sf) > 0) {
    gg_other_process <- ggplot() +
      {
        if (!is.null(states_f) && nrow(states_f) > 0) {
          geom_sf(data = states_f, fill = "#d6cac0", color = "grey40", linewidth = 0.12)
        }
      } +
      {
        if (!is.null(anp_f) && nrow(anp_f) > 0) {
          geom_sf(data = anp_f, fill = "#98a25a", color = NA, alpha = 0.2)
        }
      } +
      geom_sf(
        data = other_sf,
        aes(color = scoring_rule, shape = scoring_rule),
        size = 2.6,
        alpha = 0.95
      ) +
      scale_color_manual(
        name = "Other-LSP scoring rule",
        values = c(
          "Inside scored protected area (SIMEC / Ramsar)" = "#C03729",
          "Outside PA → codified protection score (expert table)" = "#333333"
        )
      ) +
      scale_shape_manual(
        name = "Other-LSP scoring rule",
        values = c(
          "Inside scored protected area (SIMEC / Ramsar)" = 16,
          "Outside PA → codified protection score (expert table)" = 17
        )
      ) +
      coord_sf(expand = FALSE, crs = st_crs(4326)) +
      labs(
        title = "Process: All other Special Places (script 7)",
        subtitle = "Gulf-wide: red = PA SIMEC path; black = codified protection score when outside ANPs",
        caption = "Codified scores are expert-assigned estimates and need EWG review (see table below)."
      ) +
      process_theme() +
      guides(color = guide_legend(nrow = 2), shape = guide_legend(nrow = 2))
  }
}

gg_other_process
show_process_map(gg_other_process, "process_all_other_special_places", width = 9, height = 7)
```

### Codified protection spreadsheet (for sites outside SIMEC)

```{r process-other-codified-table}
codified <- safe_read_csv(first_path(
  file.path(lsp_int_dir, "codified_protection_scores_v3.csv"),
  file.path(local_lsp_root, "int/codified_protection_scores_v3.csv"),
  file.path(lsp_prep_root, "int/codified_protection_scores_v3.csv")
))

if (!is.null(codified)) {
  cod_show <- codified %>%
    mutate(
      lsp_name = str_trunc(str_squish(as.character(lsp_name)), 55),
      estimated_condition_of_protection = suppressWarnings(as.numeric(estimated_condition_of_protection)),
      other_legal = str_trunc(str_squish(as.character(other_legal_protections_other_than_ANPs)), 50)
    ) %>%
    dplyr::select(
      lsp_name,
      estimated_condition = estimated_condition_of_protection,
      level,
      other_legal
    ) %>%
    arrange(desc(estimated_condition), lsp_name)

  print(cod_show, n = 30)
  knitr::kable(
    cod_show,
    caption = "Codified protection scores (int/codified_protection_scores_v3.csv) used when other LSPs fall outside scored PAs"
  )
} else {
  message("Codified spreadsheet not found under int/codified_protection_scores_v3.csv")
}
```

```{r process-maps-console-objects}
message(
  "Process plot objects for console review:\n",
  "  gg_pa_process, gg_dive_process, gg_fish_loreto, gg_zrp_overview,\n",
  "  gg_arch_buffer, gg_hab_types (if habitats ran), gg_hab_coquina, gg_other_process\n",
  "Tables: arch_table, cod_show (if loaded)\n",
  "save_process_maps is currently: ", save_process_maps
)
```

# Session info

```{r}
sessionInfo()
```


## Check conditon of each type

```{r}
special_protec <- read_csv(here::here(lsp_output_dir, "rgn_protected_area_condition.csv"))
dive_cond <- read_csv(here::here(lsp_output_dir, "rgn_dive_sites_condition.csv"))
fish_cond <- read_csv(here::here(lsp_output_dir, "rgn_fishing_sites_condition.csv"))
arch_cond <- read_csv(here::here(lsp_output_dir, "rgn_ewg_archaeological_sites_condition.csv"))
hab_cond <- read_csv(here::here(lsp_output_dir, "rgn_special_habitats_condition.csv"))
other_cond <- read_csv(here::here(lsp_output_dir, "rgn_final_other_protections_condition.csv"))
```
