---
title: 'OHI GoC 2026: SAF municipal vs regional crime-rate aggregation'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    theme: cerulean
editor:
  chunk_output_type: console
---

# Overview

The main safety script aggregates crimes and population to the **region** first, then divides:

`region_rate = sum(muni crimes) / sum(muni population)`

That is a population-weighted regional rate. A few large, low-crime municipalities (for example along the Sinaloa Coast) can pull the regional rate down even if smaller municipalities have higher rates.

This script compares that approach with the **unweighted mean of municipal rates** within each region:

`mean_muni_rate = mean(crime_i / pop_i)` across municipalities in the region

Region_2, Region_3, and Region_4 use the same LE `coastal_pop_adjustment` crime scaling as the main safety script (crimes × coastal share; full municipal population denominator).

Municipal population uses **CONAPO** annual mid-year totals (`pobproy_ggrupos.csv`, 1990–2040), not interpolated INEGI census anchors. That includes San Felipe and San Quintín as their own municipios.

# 1. Paths and packages

```{r}
#| label: setup

library(tidyverse)
library(here)
library(scales)
library(plotly)

safety_int <- "/home/shares/ohi/OHI_GOC/goal_prep/ss/safety_int"
safety_output <- "/home/shares/ohi/OHI_GOC/goal_prep/ss/safety_output"

annual_sesnsp_path <- here::here(safety_int, "sesnsp_anual_coastal_excl_mexicali_w2025.csv")
type_of_crimes_path <- here::here(safety_output, "types_of_crimes_goc_sesnsp_excl_mexicali_w2025.csv")
matrix_path <- here::here(safety_int, "saf_collapsed_crime_categories_matrix.csv")
coastal_adj_path <- "/home/shares/ohi/OHI_GOC/goal_prep/le/coastal_adjustment_multiplier_2020.csv"
# CONAPO municipal reconstruction/projections 1990-2040 (annual POB_TOTAL)
conapo_path <- "/home/shares/ohi/OHI_GOC/_raw_data/CONAPO_demographic_indicators/d2026/pobproy_ggrupos.csv"

rgn_labels <- c(
  "Region_1" = "Western Upper Gulf",
  "Region_2" = "Western Midriff Islands",
  "Region_3" = "Western Central Gulf",
  "Region_4" = "Western Southern Gulf",
  "Region_5" = "Eastern Upper Gulf",
  "Region_6" = "Eastern Midriff Islands",
  "Region_7" = "Southern Sonora Coast",
  "Region_8" = "Sinaloa Coast",
  "Region_9" = "Southeastern Gulf"
)

goal_code_lookup <- tibble::tribble(
  ~goal_code, ~ohi_goal,
  "FP-FIS", "food_provisioning_fisheries",
  "FP-MAR", "food_provisioning_mariculture",
  "AO-AFO", "coastal_opportunities_artisanal_fishing_opportunities",
  "AO-LCA", "coastal_opportunities_coastal_access",
  "NP", "natural_products",
  "BD", "biodiversity",
  "HS-PRO", "habitat_services_coastal_protection",
  "HS-SEQ", "habitat_services_carbon_storage",
  "HS-FIL", "habitat_services_filtration",
  "CW", "clean_waters",
  "TR", "tourism_recreation",
  "SP", "sentido_de_pertenencia",
  "LE-LIV", "livelihoods_economies_livelihoods",
  "LE-ECO", "livelihoods_economies_economies"
)

subtipo_name_fixes <- c(
  "Violencia de género distinta a la violencia familiar" =
    "Violencia de género en todas sus modalidades distinta a la violencia familiar",
  "Otros delitos contra la libertad y la seguridad sexual" =
    "Otros delitos que atentan contra la libertad y la seguridad sexual",
  "Otros delitos contra la vida y la integridad corporal" =
    "Otros delitos que atentan contra la vida y la integridad corporal",
  "Otros delitos contra la libertad personal" =
    "Otros delitos que atentan contra la libertad personal"
)
```

# 2. Build municipal crime counts by goal

```{r}
#| label: muni_crimes

types_of_crime <- read_csv(type_of_crimes_path, show_col_types = FALSE)
sesnsp_goc <- read_csv(annual_sesnsp_path, show_col_types = FALSE) %>%
  mutate(
    cve_municipio_int = as.integer(cve_municipio),
    municipio_norm = stringi::stri_trans_general(municipio, "Latin-ASCII") %>%
      toupper() %>% stringr::str_squish()
  ) %>%
  filter(!(municipio_norm == "MEXICALI" | cve_municipio_int == 2002))

collapsed_matrix <- read_csv(matrix_path, show_col_types = FALSE)

crimes_affecting_ohi <- collapsed_matrix %>%
  mutate(
    subtipo_de_delito = dplyr::recode(subtipo_de_delito_es, !!!subtipo_name_fixes)
  ) %>%
  left_join(types_of_crime, by = "subtipo_de_delito") %>%
  tidyr::pivot_longer(
    cols = all_of(goal_code_lookup$goal_code),
    names_to = "goal_code",
    values_to = "is_affecting"
  ) %>%
  left_join(goal_code_lookup, by = "goal_code") %>%
  mutate(is_affecting = as.integer(is_affecting)) %>%
  filter(is_affecting == 1)

stopifnot(!any(is.na(crimes_affecting_ohi$crime_id)))

sesnsp_goc_wid <- left_join(sesnsp_goc, types_of_crime) %>%
  dplyr::select(ano, rgn_id, cve_municipio, municipio, crime_id, annual_total)

# municipal totals by goal
muni_crimes <- left_join(crimes_affecting_ohi, sesnsp_goc_wid, by = "crime_id") %>%
  filter(!is.na(rgn_id), !is.na(cve_municipio)) %>%
  group_by(rgn_id, ano, cve_municipio, municipio, ohi_goal) %>%
  summarise(total_crimes = sum(annual_total, na.rm = TRUE), .groups = "drop")

# Option B: LE coastal-share scaling for Region_2, Region_3, Region_4
regions_with_coastal_adj <- c("Region_2", "Region_3", "Region_4")
coastal_adj <- read_csv(coastal_adj_path, show_col_types = FALSE) %>%
  mutate(
    rgn_id = as.character(rgn_id),
    coastal_pop_adjustment = dplyr::coalesce(coastal_pop_adjustment, 1)
  )

muni_crimes <- muni_crimes %>%
  mutate(rgn_id = as.character(rgn_id)) %>%
  left_join(coastal_adj, by = "rgn_id") %>%
  mutate(
    coastal_pop_adjustment = dplyr::coalesce(coastal_pop_adjustment, 1),
    total_crimes_raw = total_crimes,
    total_crimes = dplyr::if_else(
      rgn_id %in% regions_with_coastal_adj,
      total_crimes_raw * coastal_pop_adjustment,
      total_crimes_raw
    )
  )

muni_crimes %>%
  count(rgn_id, municipio) %>%
  arrange(rgn_id, municipio) %>%
  knitr::kable(caption = "Municipalities present in coastal SESNSP by region")
```

# 3. Municipal population from CONAPO (annual, no interpolation)

CONAPO provides mid-year municipal population for every year 1990–2040, including post-split municipios (San Felipe `02007`, San Quintín `02006`). We sum `HOMBRES` + `MUJERES` `POB_TOTAL` and join on SESNSP `cve_municipio` = CONAPO `CLAVE`.

```{r}
#| label: muni_pop

conapo_raw <- read_csv(conapo_path, show_col_types = FALSE)

muni_pop_conapo <- conapo_raw %>%
  mutate(
    cve_municipio = as.integer(CLAVE),
    ano = as.integer(ANO),
    pob_total = as.numeric(POB_TOTAL)
  ) %>%
  group_by(cve_municipio, ano, municipio_conapo = NOM_MUN, entidad_conapo = NOM_ENT) %>%
  summarise(population = sum(pob_total, na.rm = TRUE), .groups = "drop")

# Restrict to years present in the crime series
years_needed <- sort(unique(muni_crimes$ano))
muni_pop_interp <- muni_pop_conapo %>%
  filter(ano %in% years_needed) %>%
  rename(municipio_pop = municipio_conapo)

# Coverage check: should include San Felipe (2007) and San Quintín (2006)
muni_pop_interp %>%
  filter(cve_municipio %in% c(2006L, 2007L)) %>%
  group_by(cve_municipio, municipio_pop) %>%
  summarise(
    min_ano = min(ano),
    max_ano = max(ano),
    pop_2020 = population[ano == 2020][1],
    pop_2025 = population[ano == 2025][1],
    .groups = "drop"
  ) %>%
  knitr::kable(caption = "CONAPO population for San Quintín (2006) and San Felipe (2007)")

muni_pop_interp %>%
  summarise(
    min_ano = min(ano),
    max_ano = max(ano),
    n_years = dplyr::n_distinct(ano),
    n_muni = dplyr::n_distinct(cve_municipio)
  ) %>%
  knitr::kable(caption = "CONAPO municipal population coverage (crime years)")

summary(muni_pop_interp$population)
```

# 4. Municipal rates and both regional aggregations

```{r}
#| label: compare_methods

muni_rates <- muni_crimes %>%
  mutate(
    rgn_id = as.character(rgn_id),
    cve_municipio = as.integer(cve_municipio)
  ) %>%
  left_join(
    muni_pop_interp %>% dplyr::select(cve_municipio, ano, population, municipio_pop),
    by = c("cve_municipio", "ano")
  ) %>%
  filter(!is.na(population), population > 0) %>%
  mutate(
    muni_crime_rate = total_crimes / population,
    muni_crime_rate_unscaled = total_crimes_raw / population
  )

# Which coastal crime municipios failed to match CONAPO?
muni_crimes %>%
  mutate(cve_municipio = as.integer(cve_municipio)) %>%
  distinct(rgn_id, cve_municipio, municipio) %>%
  anti_join(muni_pop_interp %>% distinct(cve_municipio), by = "cve_municipio") %>%
  knitr::kable(caption = "Coastal SESNSP municipios with no CONAPO population match (should be empty)")

# Method A: region aggregate first (matches main script logic at muni resolution)
rate_region_aggregate <- muni_rates %>%
  group_by(rgn_id, ano, ohi_goal) %>%
  summarise(
    n_muni = dplyr::n(),
    total_crimes = sum(total_crimes, na.rm = TRUE),
    total_population = sum(population, na.rm = TRUE),
    rate_region_aggregate = total_crimes / total_population,
    .groups = "drop"
  )

# Method B: mean of municipal rates (unweighted)
rate_muni_mean <- muni_rates %>%
  group_by(rgn_id, ano, ohi_goal) %>%
  summarise(
    n_muni = dplyr::n(),
    rate_muni_mean = mean(muni_crime_rate, na.rm = TRUE),
    rate_muni_median = median(muni_crime_rate, na.rm = TRUE),
    .groups = "drop"
  )

rate_compare <- rate_region_aggregate %>%
  left_join(rate_muni_mean, by = c("rgn_id", "ano", "ohi_goal", "n_muni")) %>%
  mutate(
    ratio_muni_mean_over_aggregate = rate_muni_mean / rate_region_aggregate,
    rgn_label = dplyr::recode(rgn_id, !!!rgn_labels, .default = rgn_id)
  )

# mean across goals for region-year overview
compare_by_region <- rate_compare %>%
  group_by(rgn_id, rgn_label, ano) %>%
  summarise(
    mean_rate_region_aggregate = mean(rate_region_aggregate, na.rm = TRUE),
    mean_rate_muni_mean = mean(rate_muni_mean, na.rm = TRUE),
    mean_ratio = mean(ratio_muni_mean_over_aggregate, na.rm = TRUE),
    .groups = "drop"
  )

latest_ano <- max(compare_by_region$ano, na.rm = TRUE)

compare_by_region %>%
  filter(ano == latest_ano) %>%
  arrange(desc(mean_rate_region_aggregate)) %>%
  knitr::kable(
    digits = 4,
    caption = paste0(
      "Region rates in ", latest_ano,
      ": population-weighted aggregate vs mean of municipal rates"
    )
  )

# Sinaloa Coast detail: municipal rates in latest year (mean across goals)
muni_rates %>%
  filter(as.character(rgn_id) == "Region_8", ano == latest_ano) %>%
  group_by(municipio, cve_municipio, population) %>%
  summarise(
    mean_muni_rate = mean(muni_crime_rate, na.rm = TRUE),
    mean_crimes = mean(total_crimes, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_muni_rate)) %>%
  knitr::kable(
    digits = 4,
    caption = paste0("Sinaloa Coast (Region_8) municipal rates in ", latest_ano, " (mean across goals)")
  )

write_csv(rate_compare, here::here(safety_output, "rate_compare_muni_mean_vs_region_aggregate.csv"))
write_csv(compare_by_region, here::here(safety_output, "rate_compare_by_region_overview.csv"))
write_csv(muni_rates, here::here(safety_output, "muni_crime_rates_by_goal.csv"))
```

# 5. Plots

```{r}
#| label: plots
#| fig-width: 10
#| fig-height: 6

compare_long <- compare_by_region %>%
  tidyr::pivot_longer(
    cols = c(mean_rate_region_aggregate, mean_rate_muni_mean),
    names_to = "method",
    values_to = "mean_rate"
  ) %>%
  mutate(
    method = dplyr::recode(
      method,
      mean_rate_region_aggregate = "Region aggregate (sum crimes / sum pop)",
      mean_rate_muni_mean = "Mean of municipal rates"
    ),
    rgn_id = factor(rgn_id, levels = names(rgn_labels))
  )

ggplot(
  compare_long,
  aes(x = ano, y = mean_rate, color = method, group = method)
) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.3) +
  facet_wrap(~ rgn_id, ncol = 3, scales = "free_y", labeller = labeller(rgn_id = rgn_labels)) +
  scale_x_continuous(breaks = pretty_breaks(n = 4)) +
  labs(
    title = "Regional crime rates: aggregate vs mean of municipalities",
    subtitle = "Mean across OHI goals; R2/R3/R4 use LE coastal-share crime scaling",
    x = "Year",
    y = "Mean crime rate",
    color = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 9),
    axis.text.x = element_text(angle = 35, hjust = 1, size = 7)
  )

ggplot(
  compare_long,
  aes(x = ano, y = mean_rate, color = method, group = method)
) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.3) +
  facet_wrap(~ rgn_id, ncol = 3, scales = "fixed", labeller = labeller(rgn_id = rgn_labels)) +
  scale_x_continuous(breaks = pretty_breaks(n = 4)) +
  labs(
    title = "Regional crime rates: aggregate vs mean of municipalities",
    subtitle = "Shared y-axis",
    x = "Year",
    y = "Mean crime rate",
    color = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 9),
    axis.text.x = element_text(angle = 35, hjust = 1, size = 7)
  )
```

# 6. Interactive municipal rates over time

```{r}
#| label: plotly_muni_rates
#| fig-width: 11
#| fig-height: 7

muni_rates_overview <- muni_rates %>%
  mutate(
    rgn_label = dplyr::recode(rgn_id, !!!rgn_labels, .default = rgn_id),
    rgn_id = factor(rgn_id, levels = names(rgn_labels))
  ) %>%
  group_by(rgn_id, rgn_label, ano, cve_municipio, municipio, population) %>%
  summarise(
    mean_muni_rate = mean(muni_crime_rate, na.rm = TRUE),
    mean_crimes = mean(total_crimes, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    hover = paste0(
      municipio, "<br>",
      rgn_label, "<br>",
      "Year: ", ano, "<br>",
      "Mean rate: ", scales::number(mean_muni_rate, accuracy = 0.0001), "<br>",
      "Population: ", scales::comma(round(population))
    )
  )

p_muni_plotly <- plot_ly(
  muni_rates_overview,
  x = ~ano,
  y = ~mean_muni_rate,
  color = ~municipio,
  colors = "Set2",
  type = "scatter",
  mode = "lines+markers",
  text = ~hover,
  hoverinfo = "text",
  marker = list(size = 5),
  line = list(width = 1.5)
) %>%
  layout(
    title = list(
      text = "Municipal crime rates over time<br><sup>Mean across OHI goals; R2/R3/R4 coastal-share scaled</sup>",
      x = 0
    ),
    xaxis = list(title = "Year", dtick = 1),
    yaxis = list(title = "Mean crime rate"),
    legend = list(title = list(text = "Municipality"), font = list(size = 9)),
    hovermode = "closest"
  )

p_muni_plotly

# Faceted by region (easier to compare within region)
p_muni_by_rgn <- ggplot(
  muni_rates_overview,
  aes(
    x = ano,
    y = mean_muni_rate,
    color = municipio,
    group = municipio,
    text = hover
  )
) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.2) +
  facet_wrap(~ rgn_id, ncol = 3, scales = "free_y", labeller = labeller(rgn_id = rgn_labels)) +
  scale_x_continuous(breaks = pretty_breaks(n = 4)) +
  labs(
    title = "Municipal crime rates by region",
    subtitle = "Mean across OHI goals; hover for municipality detail",
    x = "Year",
    y = "Mean crime rate",
    color = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold", size = 9),
    axis.text.x = element_text(angle = 35, hjust = 1, size = 7)
  )

ggplotly(p_muni_by_rgn, tooltip = "text")
```

# 7. Interactive municipal rates over time (no coastal-share scaling)

Same plots as section 6, but using raw municipal crimes / municipal population (no LE `coastal_pop_adjustment`). Regions other than R2/R3/R4 are identical to section 6.

```{r}
#| label: plotly_muni_rates_unscaled
#| fig-width: 11
#| fig-height: 7

muni_rates_overview_unscaled <- muni_rates %>%
  mutate(
    rgn_label = dplyr::recode(rgn_id, !!!rgn_labels, .default = rgn_id),
    rgn_id = factor(rgn_id, levels = names(rgn_labels))
  ) %>%
  group_by(rgn_id, rgn_label, ano, cve_municipio, municipio, population) %>%
  summarise(
    mean_muni_rate = mean(muni_crime_rate_unscaled, na.rm = TRUE),
    mean_crimes = mean(total_crimes_raw, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    hover = paste0(
      municipio, "<br>",
      rgn_label, "<br>",
      "Year: ", ano, "<br>",
      "Mean rate (unscaled): ", scales::number(mean_muni_rate, accuracy = 0.0001), "<br>",
      "Population: ", scales::comma(round(population))
    )
  )

p_muni_plotly_unscaled <- plot_ly(
  muni_rates_overview_unscaled,
  x = ~ano,
  y = ~mean_muni_rate,
  color = ~municipio,
  colors = "Set2",
  type = "scatter",
  mode = "lines+markers",
  text = ~hover,
  hoverinfo = "text",
  marker = list(size = 5),
  line = list(width = 1.5)
) %>%
  layout(
    title = list(
      text = "Municipal crime rates over time<br><sup>Mean across OHI goals; no coastal-share scaling</sup>",
      x = 0
    ),
    xaxis = list(title = "Year", dtick = 1),
    yaxis = list(title = "Mean crime rate"),
    legend = list(title = list(text = "Municipality"), font = list(size = 9)),
    hovermode = "closest"
  )

p_muni_plotly_unscaled

p_muni_by_rgn_unscaled <- ggplot(
  muni_rates_overview_unscaled,
  aes(
    x = ano,
    y = mean_muni_rate,
    color = municipio,
    group = municipio,
    text = hover
  )
) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.2) +
  facet_wrap(~ rgn_id, ncol = 3, scales = "free_y", labeller = labeller(rgn_id = rgn_labels)) +
  scale_x_continuous(breaks = pretty_breaks(n = 4)) +
  labs(
    title = "Municipal crime rates by region (no coastal-share scaling)",
    subtitle = "Raw municipal crimes / municipal population; hover for detail",
    x = "Year",
    y = "Mean crime rate",
    color = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold", size = 9),
    axis.text.x = element_text(angle = 35, hjust = 1, size = 7)
  )

ggplotly(p_muni_by_rgn_unscaled, tooltip = "text")
```

# Interpretation notes

- If **mean of municipal rates >> region aggregate**, large low-crime municipalities are dominating the regional denominator (plausible for Sinaloa Coast).
- If the two methods are similar, municipalities within the region have similar rates.
- This script does not change the main SAF scores; it is diagnostic only.
- Region_2, Region_3, and Region_4 crimes are scaled by LE `coastal_pop_adjustment` to down-weight Pacific-side incidence (section 6). Section 7 shows the unscaled municipal rates for comparison.
- Municipal denominators are CONAPO annual `POB_TOTAL` (HOMBRES + MUJERES). The main SAF scores in `ss_safety_goal_specific.qmd` still use the regional INEGI interpolated population file unless that is updated separately.
