---
title: 'OHI GoC 2026: SS Safety Data Preparation (SESNSP)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor:
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options:
  chunk_output_type: console
---

# Overview

THINGS TO ADD: (1) Update the INEGI population data from the newest census when it is released in 2025; (2) Incorporate weights based on the type of crime?

In this script, we will take the annual crime incidents for each coastal municipality that we cleaned in `SS/v2025/safety/ss_safety_dataprep.qmd` and create individual subsets for each goal. This will be based on expert judgement, which identified which types of crimes impact specific OHI GoC goals. 

1. **Load packages.** We load the tidyverse and spatial libraries used throughout the script.
2. **Read cleaned SESNSP municipal data through 2025.** 

Intermediate files are saved to `/home/shares/ohi/OHI_GOC/goal_prep/ss/safety_int`. Final outputs are saved to `/home/shares/ohi/OHI_GOC/goal_prep/ss/safety_output`.

# Background on SESNSP

The **Secretariado Ejecutivo del Sistema Nacional de Seguridad Publica** (SESNSP) is the federal agency responsible for coordinating public security policy in México. Among its core functions, SESNSP compiles and publishes official statistics on crime incidence nationwide. These data are derived from **averiguaciones previas** and **carpetas de investigacion** (preliminary inquiries and case files) registered by state and federal prosecutors.

For the Safety and Security (SS) pressure, specifically Safety, we use the municipal level open data series, which reports the count of registered crimes by municipality, year, month, and crime classification. The official portal is here:

<https://www.gob.mx/sesnsp/acciones-y-programas/datos-abiertos-de-incidencia-delictiva>

For this update we use the 2015 through 2025 methodology series. The data dictionary and full methodology are documented here:

<https://sspcgob-my.sharepoint.com/:x:/g/personal/cni_sspc_gob_mx/Eft_H-La80NInrfnyRu9PCkBfyIyy-yr9hl8imxAyeLXMg?e=xvI38U>

**Incidencia Delictiva Fuero Común: Nueva Metodología**

| Campo | Descripción |
|:---|:---|
| Año | Año de registro de las averiguaciones previas y/o carpetas de investigación. |
| Clave_Ent | Clave de la entidad, según el Marco Geoestadístico Nacional (MGN) del Instituto Nacional de Geografía y Estadística (INEGI). |
| Entidad | Entidad federativa de registro de las averiguaciones previas y/o carpetas de investigación. |
| Bien jurídico afectado | Primera clasificación de los delitos en las averiguaciones previas y/o carpetas de investigación. |
| Tipo de delito | Segunda clasificación de los delitos. |
| Subtipo de delito | Tercera clasificación de los delitos. |
| Modalidad | Cuarta clasificación de los delitos. |
| Enero through diciembre | Mes de registro de las averiguaciones previas y/o carpetas de investigación. |

## 1. Load packages

```{r}
#| label: step_01_load_packages

library(tidyverse) # data manipulation and visualization
library(magrittr) # pipe operators
library(here) # project relative paths
library(janitor) # clean column names
library(readr) # read csv files
library(purrr) # functional programming tools
library(mapview) # interactive maps
library(sf) # spatial data handling
library(stringi) # normalize spanish place names

safety_int <- "/home/shares/ohi/OHI_GOC/goal_prep/ss/safety_int" # intermediate files
safety_output <- "/home/shares/ohi/OHI_GOC/goal_prep/ss/safety_output" # final outputs
annual_sesnsp_path <- "/home/shares/ohi/OHI_GOC/goal_prep/ss/safety_int/sesnsp_anual_coastal.csv" # cleaned SESNSP data
annual_sesnsp_sf_path <- "/home/shares/ohi/OHI_GOC/goal_prep/ss/safety_output/sesnsp_coastal_municipios_w2025.gpkg"
matrix_path <- "/home/shares/ohi/OHI_GOC/goal_prep/ss/safety_int/Safety_Security_Methodology_Hem_Crime_x_OHI_Goal.csv" # expert matrix
type_of_crimes_path <- "/home/shares/ohi/OHI_GOC/goal_prep/ss/safety_output/types_of_crimes_goc_sesnsp_w2025.csv" # has crimeID
inegi_population_path <- "/home/shares/ohi/OHI_GOC/spatial/population_municipal/interpolated_population_per_rgn_coastalmunicipios_2000_2025.csv" # update in September 2026!!
```

## 2. Read cleaned municipal data through 2025

```{r}
#| label: step_02_read_sesnsp

sesnsp_goc <- read_csv(here::here(annual_sesnsp_path))
sesnsp_goc_sf <- st_read(here::here(annual_sesnsp_sf_path))
```


```{r}
# read in the type of crimes and join with the data to give each crime an identifier
types_of_crime <- read_csv(here(type_of_crimes_path))

sesnsp_goc_wid <- left_join(sesnsp_goc, types_of_crime) %>% 
  dplyr::select(c(ano, rgn_id, clave_ent, entidad, cve_municipio, cve_municipio, municipio, crime_id, annual_total)) %>% 
  filter(!municipio %in% "Mexicali")
```


## 3. Clean Expert Matrix

```{r}
#| label: step_03_expert_matrix

hem_matrix <- read_csv(here::here(matrix_path)) 

hem_matrix_clean <- hem_matrix %>% 
  janitor::clean_names() %>% 
  dplyr::select(-x0_no_afecta_1_si_afecta) %>% 
  left_join(.,types_of_crime) %>% 
  tidyr::pivot_longer(
    cols = food_provisioning_fisheries:research_ocean_literacy, # one row per goal/subgoal
    names_to = "ohi_goal",
    values_to = "is_affecting"
  )

crimes_affecting_ohi <- hem_matrix_clean %>% 
  filter(is_affecting %in% "1")
unique(crimes_affecting_ohi$crime_id)

write_csv(hem_matrix_clean, here::here(safety_int, "hem_expert_matrix_clean_070626.csv"))
write_csv(crimes_affecting_ohi, here::here(safety_int, "crimes_affecting_ohi_070626.csv"))
```

## 4. Annual crimes per goal

```{r}
annual_incidences_per_crime_and_yr <- left_join(crimes_affecting_ohi, sesnsp_goc_wid)

unique(annual_incidences_per_crime_and_yr$crime_id) # same as crimes_affecting_ohi, good.
unique(annual_incidences_per_crime_and_yr$is_affecting) # good, only 1.  This means that all the crimes in this datasheet affect the goals.
```


```{r}
incidences_per_goal <- annual_incidences_per_crime_and_yr %>% 
  group_by(rgn_id, ano, ohi_goal) %>% 
  summarize(
    total_crimes_reported = sum(annual_total)
  ) %>% 
  ungroup()

unique(incidences_per_goal$ano)
unique(incidences_per_goal$rgn_id)
unique(incidences_per_goal$rgn_id)
```

## 5. Standardize by population

This will need to be updated when we have new INEGI population data from the census, released in September of 2026.

```{r}
# note that INEGI's population census will be updated through 2025 (done every 5 years) in September 2026.  We will need to update this then!!
interpolated_population_inegi_2000_2025 <- read_csv(here::here(inegi_population_path)) %>% 
  mutate(ano = year_interpolated) %>% 
  dplyr::select(rgn_id, ano, population_interpolated)
```


```{r}
# divide the number of crimes by the population for every region, year, and ohi goal
rate_of_incidence_per_goal <- incidences_per_goal %>% 
  left_join(interpolated_population_inegi_2000_2025) %>% 
  mutate(crime_rate = total_crimes_reported / population_interpolated)

write_csv(rate_of_incidence_per_goal, here::here(safety_output, "rate_of_incidence_per_goal_2015_2025.csv"))
```

### 5a. Viz

```{r}
library(tidyverse)
library(here)
library(scales)
library(systemfonts)   # for checking/registering fonts
library(showtext)      # renders text with the requested font

# ---- 0. font setup: calibri -----------------------------------------
font_family_to_use <- "Calibri"
showtext::showtext_auto()
showtext::showtext_opts(dpi = 300)  # match the dpi used in the commented ggsave() calls below

# ---- region labels & colors ------------------------------------------
rgn_labels <- c(
  "Region_1" = "Western Upper Gulf",
  "Region_2" = "Western Midriff Islands",
  "Region_3" = "Western Central Gulf",
  "Region_4" = "Western Southern Gulf",
  "Region_5" = "Eastern Upper Gulf",
  "Region_6" = "Eastern Midriff Islands",
  "Region_7" = "Southern Sonora Coast",
  "Region_8" = "Sinaloa Coast",
  "Region_9" = "Southeastern Gulf"
)

rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

# ---- 1. load data --------------------------------------------------
rate_of_incidence_per_goal <- read_csv(
  here::here(safety_output, "rate_of_incidence_per_goal_2015_2025.csv")
)

# make sure types are right (adjust column names if yours differ)
rate_of_incidence_per_goal <- rate_of_incidence_per_goal %>%
  mutate(
    ano      = as.integer(ano),
    rgn_id   = as.factor(rgn_id),
    ohi_goal = as.factor(ohi_goal)
  )

# a consistent theme for everything below
theme_set(
  theme_minimal(base_family = "Calibri") +
  theme(
    plot.title = element_text(face = "bold", size = 5),
    strip.text = element_text(face = "bold", size = 4),
    axis.text.x = element_text(angle = 35, hjust = 1, size = 3),
    axis.text.y = element_text(size = 3),
    axis.title.y = element_text(size = 3) 
  )
)
theme_set(theme_minimal())  # or theme_minimal(), whatever your normal default is

# ---- 2. overall crime rate trend (all regions/goals pooled) -------
overall_trend <- rate_of_incidence_per_goal %>%
  group_by(ano) %>%
  summarise(
    total_crimes = sum(total_crimes_reported, na.rm = TRUE),
    total_pop    = sum(population_interpolated, na.rm = TRUE),
    mean_rate    = mean(crime_rate, na.rm = TRUE),
    .groups = "drop"
  )

p_overall <- ggplot(overall_trend, aes(x = ano, y = mean_rate)) +
  geom_line(linewidth = 1, color = "#2C6E9E") +
  geom_point(size = 2, color = "#2C6E9E") +
  scale_x_continuous(breaks = pretty_breaks()) +
  labs(
    title = "Overall Crime Rate Over Time",
    subtitle = "Mean crime rate across all regions and OHI goals",
    x = "Year", y = "Mean crime rate"
  )
p_overall

# ggsave(here::here(safety_output, "plot_overall_trend.png"),
#        p_overall, width = 8, height = 5, dpi = 300)

# ---- 3. trend by ohi goal (small multiples) ------------------------
by_goal <- rate_of_incidence_per_goal %>%
  group_by(ano, ohi_goal) %>%
  summarise(mean_rate = mean(crime_rate, na.rm = TRUE), .groups = "drop")

p_by_goal <- ggplot(by_goal, aes(x = ano, y = mean_rate)) +
  geom_line(color = "#2C6E9E", linewidth = 1) +
  geom_point(color = "#2C6E9E", size = 1.5) +
  facet_wrap(~ ohi_goal, scales = "free_y", ncol = 4) +
  scale_x_continuous(breaks = pretty_breaks(n = 4)) +
  labs(
    title = "Crime Rate Over Time by OHI Goal",
    x = "Year", y = "Mean crime rate"
  )
p_by_goal

# ggsave(here::here(safety_output, "plot_by_goal.png"),
#        p_by_goal, width = 12, height = 8, dpi = 300)

# ---- 4. trend by region (small multiples) --------------------------
by_region <- rate_of_incidence_per_goal %>%
  group_by(ano, rgn_id) %>%
  summarise(mean_rate = mean(crime_rate, na.rm = TRUE), .groups = "drop")

p_by_region <- ggplot(by_region, aes(x = ano, y = mean_rate, color = rgn_id)) +
  geom_line(linewidth = 1, show.legend = FALSE) +
  geom_point(size = 1.5, show.legend = FALSE) +
  facet_wrap(~ rgn_id, scales = "free_y", labeller = labeller(rgn_id = rgn_labels)) +
  scale_color_manual(values = rgn_colors) +
  scale_x_continuous(breaks = pretty_breaks(n = 4)) +
  labs(
    title = "Crime Rate Over Time by Region",
    x = "Year", y = "Mean crime rate"
  )
p_by_region

# ggsave(here::here(safety_output, "plot_by_region.png"),
#        p_by_region, width = 12, height = 8, dpi = 300)

# ---- 5. all regions on one plot, colored, faceted by goal ----------
p_region_lines_by_goal <- ggplot(
  rate_of_incidence_per_goal,
  aes(x = ano, y = crime_rate, color = rgn_id, group = rgn_id)
) +
  geom_line(alpha = 0.8) +
  facet_wrap(~ ohi_goal, scales = "free_y", ncol = 4) +
  scale_color_manual(values = rgn_colors, labels = rgn_labels) +
  scale_x_continuous(breaks = pretty_breaks(n = 4)) +
  labs(
    title = "Crime Rate Trajectories by Region, Faceted by OHI Goal",
    x = "Year", y = "Crime rate", color = "Region"
  ) +
  theme(legend.position = "bottom")
p_region_lines_by_goal

# ggsave(here::here(safety_output, "plot_region_lines_by_goal.png"),
#        p_region_lines_by_goal, width = 14, height = 9, dpi = 300)

# ---- 6. heatmap: region x year, faceted by goal --------------------
p_heatmap <- ggplot(
  rate_of_incidence_per_goal,
  aes(x = ano, y = rgn_id, fill = crime_rate)
) +
  geom_tile(color = "white", linewidth = 0.2) +
  facet_wrap(~ ohi_goal, ncol = 4) +
  scale_fill_viridis_c(option = "magma", name = "Crime rate") +
  scale_y_discrete(labels = rgn_labels) +
  scale_x_continuous(breaks = pretty_breaks(n = 4)) +
  labs(
    title = "Crime Rate Heatmap: Region x Year, by OHI Goal",
    x = "Year", y = "Region"
  )
p_heatmap

# ggsave(here::here(safety_output, "plot_heatmap.png"),
#        p_heatmap, width = 14, height = 9, dpi = 300)

# ---- 7. single-goal / single-region deep dive (example) ------------
# swap "food_provisioning_fisheries" for whatever goal you want
p_single_goal <- rate_of_incidence_per_goal %>%
  filter(ohi_goal == "food_provisioning_fisheries") %>%
  ggplot(aes(x = ano, y = crime_rate, color = rgn_id, group = rgn_id)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  scale_color_manual(values = rgn_colors, labels = rgn_labels) +
  scale_x_continuous(breaks = pretty_breaks()) +
  labs(
    title = "Crime Rate Over Time: Food Provisioning - Fisheries",
    subtitle = "By region",
    x = "Year", y = "Crime rate", color = "Region"
  )
p_single_goal

# ggsave(here::here(safety_output, "plot_single_goal_example.png"),
#        p_single_goal, width = 9, height = 6, dpi = 300)
```


## 6. Current status of crime per goal

```{r}

```









