---
title: 'OHI GoC 2026: SS Safety Data Preparation (SESNSP)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor:
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options:
  chunk_output_type: console
---

# Overview

THINGS TO ADD: (1) Update the INEGI population data from the newest census when it is released in 2025; (2) Incorporate weights based on the type of crime?

In this script, we will take the annual crime incidents for each coastal municipality that we cleaned in `SS/v2025/safety/ss_safety_dataprep.qmd` and create individual subsets for each goal. This will be based on expert judgement, which identified which types of crimes impact specific OHI GoC goals. 

1. **Load packages.** We load the tidyverse and spatial libraries used throughout the script.
2. **Read cleaned SESNSP municipal data through 2025.** 

Intermediate files are saved to `/home/shares/ohi/OHI_GOC/goal_prep/ss/safety_int`. Final outputs are saved to `/home/shares/ohi/OHI_GOC/goal_prep/ss/safety_output`.

# Background on SESNSP

The **Secretariado Ejecutivo del Sistema Nacional de Seguridad Publica** (SESNSP) is the federal agency responsible for coordinating public security policy in México. Among its core functions, SESNSP compiles and publishes official statistics on crime incidence nationwide. These data are derived from **averiguaciones previas** and **carpetas de investigacion** (preliminary inquiries and case files) registered by state and federal prosecutors.

For the Safety and Security (SS) pressure, specifically Safety, we use the municipal level open data series, which reports the count of registered crimes by municipality, year, month, and crime classification. The official portal is here:

<https://www.gob.mx/sesnsp/acciones-y-programas/datos-abiertos-de-incidencia-delictiva>

For this update we use the 2015 through 2025 methodology series. The data dictionary and full methodology are documented here:

<https://sspcgob-my.sharepoint.com/:x:/g/personal/cni_sspc_gob_mx/Eft_H-La80NInrfnyRu9PCkBfyIyy-yr9hl8imxAyeLXMg?e=xvI38U>

**Incidencia Delictiva Fuero Común: Nueva Metodología**

| Campo | Descripción |
|:---|:---|
| Año | Año de registro de las averiguaciones previas y/o carpetas de investigación. |
| Clave_Ent | Clave de la entidad, según el Marco Geoestadístico Nacional (MGN) del Instituto Nacional de Geografía y Estadística (INEGI). |
| Entidad | Entidad federativa de registro de las averiguaciones previas y/o carpetas de investigación. |
| Bien jurídico afectado | Primera clasificación de los delitos en las averiguaciones previas y/o carpetas de investigación. |
| Tipo de delito | Segunda clasificación de los delitos. |
| Subtipo de delito | Tercera clasificación de los delitos. |
| Modalidad | Cuarta clasificación de los delitos. |
| Enero through diciembre | Mes de registro de las averiguaciones previas y/o carpetas de investigación. |

## 1. Load packages

```{r}
#| label: step_01_load_packages

library(tidyverse) # data manipulation and visualization
library(magrittr) # pipe operators
library(here) # project relative paths
library(janitor) # clean column names
library(readr) # read csv files
library(purrr) # functional programming tools
library(mapview) # interactive maps
library(sf) # spatial data handling
library(stringi) # normalize spanish place names

safety_int <- "/home/shares/ohi/OHI_GOC/goal_prep/ss/safety_int" # intermediate files
safety_output <- "/home/shares/ohi/OHI_GOC/goal_prep/ss/safety_output" # final outputs
annual_sesnsp_path <- "/home/shares/ohi/OHI_GOC/goal_prep/ss/safety_int/sesnsp_anual_coastal.csv" # cleaned SESNSP data
annual_sesnsp_sf_path <- "/home/shares/ohi/OHI_GOC/goal_prep/ss/safety_output/sesnsp_coastal_municipios_w2025.gpkg"
matrix_path <- "/home/shares/ohi/OHI_GOC/goal_prep/ss/safety_int/Safety_Security_Methodology_Hem_Crime_x_OHI_Goal.csv" # expert matrix
type_of_crimes_path <- "/home/shares/ohi/OHI_GOC/goal_prep/ss/safety_output/types_of_crimes_goc_sesnsp_w2025.csv" # has crimeID
inegi_population_path <- "/home/shares/ohi/OHI_GOC/spatial/population_municipal/interpolated_population_per_rgn_coastalmunicipios_2000_2025.csv" # update in September 2026!!
```

## 2. Read cleaned municipal data through 2025

```{r}
#| label: step_02_read_sesnsp

sesnsp_goc <- read_csv(here::here(annual_sesnsp_path))
sesnsp_goc_sf <- st_read(here::here(annual_sesnsp_sf_path))
```


```{r}
# read in the type of crimes and join with the data to give each crime an identifier
types_of_crime <- read_csv(here(type_of_crimes_path))

sesnsp_goc_wid <- left_join(sesnsp_goc, types_of_crime) %>% 
  dplyr::select(c(ano, rgn_id, clave_ent, entidad, cve_municipio, cve_municipio, municipio, crime_id, annual_total)) %>% 
  filter(!municipio %in% "Mexicali")
```


## 3. Clean Expert Matrix

```{r}
#| label: step_03_expert_matrix

hem_matrix <- read_csv(here::here(matrix_path)) 

hem_matrix_clean <- hem_matrix %>% 
  janitor::clean_names() %>% 
  dplyr::select(-x0_no_afecta_1_si_afecta) %>% 
  left_join(.,types_of_crime) %>% 
  tidyr::pivot_longer(
    cols = food_provisioning_fisheries:research_ocean_literacy, # one row per goal/subgoal
    names_to = "ohi_goal",
    values_to = "is_affecting"
  )

crimes_affecting_ohi <- hem_matrix_clean %>% 
  filter(is_affecting %in% "1")
unique(crimes_affecting_ohi$crime_id)

# write_csv(hem_matrix_clean, here::here(safety_int, "hem_expert_matrix_clean_070626.csv"))
# write_csv(crimes_affecting_ohi, here::here(safety_int, "crimes_affecting_ohi_070626.csv"))
```

## 4. Annual crimes per goal

```{r}
annual_incidences_per_crime_and_yr <- left_join(crimes_affecting_ohi, sesnsp_goc_wid)

unique(annual_incidences_per_crime_and_yr$crime_id) # same as crimes_affecting_ohi, good.
unique(annual_incidences_per_crime_and_yr$is_affecting) # good, only 1.  This means that all the crimes in this datasheet affect the goals.
```


```{r}
incidences_per_goal <- annual_incidences_per_crime_and_yr %>% 
  group_by(rgn_id, ano, ohi_goal) %>% 
  summarize(
    total_crimes_reported = sum(annual_total)
  ) %>% 
  ungroup()

unique(incidences_per_goal$ano)
unique(incidences_per_goal$rgn_id)
unique(incidences_per_goal$rgn_id)
```

## 5. Standardize by population

This will need to be updated when we have new INEGI population data from the census, released in September of 2026.

```{r}
# note that INEGI's population census will be updated through 2025 (done every 5 years) in September 2026.  We will need to update this then!!
interpolated_population_inegi_2000_2025 <- read_csv(here::here(inegi_population_path)) %>% 
  mutate(ano = year_interpolated) %>% 
  dplyr::select(rgn_id, ano, population_interpolated)
```


```{r}
# divide the number of crimes by the population for every region, year, and ohi goal
rate_of_incidence_per_goal <- incidences_per_goal %>% 
  left_join(interpolated_population_inegi_2000_2025) %>% 
  mutate(crime_rate = total_crimes_reported / population_interpolated)

# write_csv(rate_of_incidence_per_goal, here::here(safety_output, "rate_of_incidence_per_goal_2015_2025.csv"))
```

### 5a. Viz

```{r}
library(tidyverse)
library(here)
library(scales)
library(systemfonts)   # for checking/registering fonts
library(showtext)      # renders text with the requested font

# ---- 0. font setup: calibri -----------------------------------------
font_family_to_use <- "Calibri"
showtext::showtext_auto()
showtext::showtext_opts(dpi = 300)  # match the dpi used in the commented ggsave() calls below

# ---- region labels & colors ------------------------------------------
rgn_labels <- c(
  "Region_1" = "Western Upper Gulf",
  "Region_2" = "Western Midriff Islands",
  "Region_3" = "Western Central Gulf",
  "Region_4" = "Western Southern Gulf",
  "Region_5" = "Eastern Upper Gulf",
  "Region_6" = "Eastern Midriff Islands",
  "Region_7" = "Southern Sonora Coast",
  "Region_8" = "Sinaloa Coast",
  "Region_9" = "Southeastern Gulf"
)

rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

# ---- 1. load data --------------------------------------------------
rate_of_incidence_per_goal <- read_csv(
  here::here(safety_output, "rate_of_incidence_per_goal_2015_2025.csv")
)

# make sure types are right (adjust column names if yours differ)
rate_of_incidence_per_goal <- rate_of_incidence_per_goal %>%
  mutate(
    ano      = as.integer(ano),
    rgn_id   = as.factor(rgn_id),
    ohi_goal = as.factor(ohi_goal)
  )

# a consistent theme for everything below
theme_set(
  theme_minimal(base_family = "Calibri") +
  theme(
    plot.title = element_text(face = "bold", size = 5),
    strip.text = element_text(face = "bold", size = 4),
    axis.text.x = element_text(angle = 35, hjust = 1, size = 3),
    axis.text.y = element_text(size = 3),
    axis.title.y = element_text(size = 3) 
  )
)
theme_set(theme_minimal())  # or theme_minimal(), whatever your normal default is

# ---- 2. overall crime rate trend (all regions/goals pooled) -------
overall_trend <- rate_of_incidence_per_goal %>%
  group_by(ano) %>%
  summarise(
    total_crimes = sum(total_crimes_reported, na.rm = TRUE),
    total_pop    = sum(population_interpolated, na.rm = TRUE),
    mean_rate    = mean(crime_rate, na.rm = TRUE),
    .groups = "drop"
  )

p_overall <- ggplot(overall_trend, aes(x = ano, y = mean_rate)) +
  geom_line(linewidth = 1, color = "#2C6E9E") +
  geom_point(size = 2, color = "#2C6E9E") +
  scale_x_continuous(breaks = pretty_breaks()) +
  labs(
    title = "Overall Crime Rate Over Time",
    subtitle = "Mean crime rate across all regions and OHI goals",
    x = "Year", y = "Mean crime rate"
  )
p_overall

# ggsave(here::here(safety_output, "plot_overall_trend.png"),
#        p_overall, width = 8, height = 5, dpi = 300)

# ---- 3. trend by ohi goal (small multiples) ------------------------
by_goal <- rate_of_incidence_per_goal %>%
  group_by(ano, ohi_goal) %>%
  summarise(mean_rate = mean(crime_rate, na.rm = TRUE), .groups = "drop")

p_by_goal <- ggplot(by_goal, aes(x = ano, y = mean_rate)) +
  geom_line(color = "#2C6E9E", linewidth = 1) +
  geom_point(color = "#2C6E9E", size = 1.5) +
  facet_wrap(~ ohi_goal, scales = "free_y", ncol = 4) +
  scale_x_continuous(breaks = pretty_breaks(n = 4)) +
  labs(
    title = "Crime Rate Over Time by OHI Goal",
    x = "Year", y = "Mean crime rate"
  )
p_by_goal

# ggsave(here::here(safety_output, "plot_by_goal.png"),
#        p_by_goal, width = 12, height = 8, dpi = 300)

# ---- 4. trend by region (small multiples) --------------------------
by_region <- rate_of_incidence_per_goal %>%
  group_by(ano, rgn_id) %>%
  summarise(mean_rate = mean(crime_rate, na.rm = TRUE), .groups = "drop")

p_by_region <- ggplot(by_region, aes(x = ano, y = mean_rate, color = rgn_id)) +
  geom_line(linewidth = 1, show.legend = FALSE) +
  geom_point(size = 1.5, show.legend = FALSE) +
  facet_wrap(~ rgn_id, scales = "free_y", labeller = labeller(rgn_id = rgn_labels)) +
  scale_color_manual(values = rgn_colors) +
  scale_x_continuous(breaks = pretty_breaks(n = 4)) +
  labs(
    title = "Crime Rate Over Time by Region",
    x = "Year", y = "Mean crime rate"
  )
p_by_region

# ggsave(here::here(safety_output, "plot_by_region.png"),
#        p_by_region, width = 12, height = 8, dpi = 300)

# ---- 5. all regions on one plot, colored, faceted by goal ----------
p_region_lines_by_goal <- ggplot(
  rate_of_incidence_per_goal,
  aes(x = ano, y = crime_rate, color = rgn_id, group = rgn_id)
) +
  geom_line(alpha = 0.8) +
  facet_wrap(~ ohi_goal, scales = "free_y", ncol = 4) +
  scale_color_manual(values = rgn_colors, labels = rgn_labels) +
  scale_x_continuous(breaks = pretty_breaks(n = 4)) +
  labs(
    title = "Crime Rate Trajectories by Region, Faceted by OHI Goal",
    x = "Year", y = "Crime rate", color = "Region"
  ) +
  theme(legend.position = "bottom")
p_region_lines_by_goal

# ggsave(here::here(safety_output, "plot_region_lines_by_goal.png"),
#        p_region_lines_by_goal, width = 14, height = 9, dpi = 300)

# ---- 6. heatmap: region x year, faceted by goal --------------------
p_heatmap <- ggplot(
  rate_of_incidence_per_goal,
  aes(x = ano, y = rgn_id, fill = crime_rate)
) +
  geom_tile(color = "white", linewidth = 0.2) +
  facet_wrap(~ ohi_goal, ncol = 4) +
  scale_fill_viridis_c(option = "magma", name = "Crime rate") +
  scale_y_discrete(labels = rgn_labels) +
  scale_x_continuous(breaks = pretty_breaks(n = 4)) +
  labs(
    title = "Crime Rate Heatmap: Region x Year, by OHI Goal",
    x = "Year", y = "Region"
  )
p_heatmap

# ggsave(here::here(safety_output, "plot_heatmap.png"),
#        p_heatmap, width = 14, height = 9, dpi = 300)

# ---- 7. single-goal / single-region deep dive (example) ------------
# swap "food_provisioning_fisheries" for whatever goal you want
p_single_goal <- rate_of_incidence_per_goal %>%
  filter(ohi_goal == "food_provisioning_fisheries") %>%
  ggplot(aes(x = ano, y = crime_rate, color = rgn_id, group = rgn_id)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  scale_color_manual(values = rgn_colors, labels = rgn_labels) +
  scale_x_continuous(breaks = pretty_breaks()) +
  labs(
    title = "Crime Rate Over Time: Food Provisioning - Fisheries",
    subtitle = "By region",
    x = "Year", y = "Crime rate", color = "Region"
  )
p_single_goal

# ggsave(here::here(safety_output, "plot_single_goal_example.png"),
#        p_single_goal, width = 9, height = 6, dpi = 300)
```


## 6. Current status of crime per goal

We convert crime rates into a current status score for each region, year, and OHI goal. Higher crime rates produce lower scores. For each goal, we find the maximum crime rate across all regions and years, then score each observation as:

`current_status = (1 - (crime_rate / max_crime_rate_for_that_goal)) * 100`

A region and year with zero crime scores 100. The region and year with the highest crime rate for that goal scores 0.

```{r}
#| label: step_06_current_status

rate_of_incidence_per_goal <- read_csv( # reload if running this section alone
  here::here(safety_output, "rate_of_incidence_per_goal_2015_2025.csv")
)

summary(rate_of_incidence_per_goal$crime_rate) # overall crime rate distribution

# for each ohi goal, find the maximum crime rate across all regions and years
max_crime_rate_by_goal <- rate_of_incidence_per_goal %>%
  group_by(ohi_goal) %>%
  summarise(
    max_crime_rate = max(crime_rate, na.rm = TRUE),
    .groups = "drop"
  )

current_status_per_goal <- rate_of_incidence_per_goal %>%
  left_join(max_crime_rate_by_goal, by = "ohi_goal") %>%
  mutate(
    current_status = (1 - (crime_rate / max_crime_rate)) * 100 # 100 = best, 0 = worst
  )

summary(current_status_per_goal$current_status) # overall status distribution
unique(current_status_per_goal$ohi_goal)

# summary of current status for each goal separately
current_status_summary_by_goal <- split(
  current_status_per_goal$current_status,
  current_status_per_goal$ohi_goal
) %>%
  lapply(summary)

current_status_summary_by_goal

# also as a tidy table for easier comparison across goals
current_status_summary_table <- current_status_per_goal %>%
  group_by(ohi_goal) %>%
  summarise(
    n = dplyr::n(),
    min = min(current_status, na.rm = TRUE),
    q1 = quantile(current_status, 0.25, na.rm = TRUE),
    median = median(current_status, na.rm = TRUE),
    mean = mean(current_status, na.rm = TRUE),
    q3 = quantile(current_status, 0.75, na.rm = TRUE),
    max = max(current_status, na.rm = TRUE),
    .groups = "drop"
  )

current_status_summary_table

write_csv(
  current_status_per_goal,
  here::here(safety_output, "current_status_per_goal_2015_2025.csv")
)

write_csv(
  current_status_summary_table,
  here::here(safety_output, "current_status_summary_by_goal_2015_2025.csv")
)
```

### 6a. Current status plots by goal

We reuse the same region colors and labels defined in section 5a. For convenience we redefine them here so this section can also be run on its own.

```{r}
#| label: step_06a_status_plots
#| fig-width: 8
#| fig-height: 4.5
#| fig-dpi: 96

library(scales)

# turn off showtext if it was left on from section 5a; that dpi mismatch makes text huge in the plot pane
if (requireNamespace("showtext", quietly = TRUE)) {
  showtext::showtext_auto(FALSE)
}

rgn_labels <- c(
  "Region_1" = "Western Upper Gulf",
  "Region_2" = "Western Midriff Islands",
  "Region_3" = "Western Central Gulf",
  "Region_4" = "Western Southern Gulf",
  "Region_5" = "Eastern Upper Gulf",
  "Region_6" = "Eastern Midriff Islands",
  "Region_7" = "Southern Sonora Coast",
  "Region_8" = "Sinaloa Coast",
  "Region_9" = "Southeastern Gulf"
)

rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

theme_status <- theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11, color = "grey40"),
    axis.title = element_text(size = 11),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10),
    legend.position = "bottom",
    legend.key.size = unit(0.5, "cm"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(6, 8, 4, 6)
  )

current_status_per_goal <- read_csv(
  here::here(safety_output, "current_status_per_goal_2015_2025.csv")
) %>%
  mutate(
    ano = as.integer(ano),
    rgn_id = factor(rgn_id, levels = names(rgn_labels)),
    ohi_goal = as.character(ohi_goal)
  )

goal_list <- sort(unique(current_status_per_goal$ohi_goal))

p_status_list <- list() # named list so we can call any goal plot later

for (g in goal_list) {
  goal_title <- stringr::str_replace_all(g, "_", " ") %>%
    stringr::str_to_title()

  p_status_list[[g]] <- current_status_per_goal %>%
    filter(ohi_goal == g) %>%
    ggplot(aes(x = ano, y = current_status, color = rgn_id, group = rgn_id)) +
    geom_line(linewidth = 1.1, alpha = 0.9) +
    geom_point(size = 2) +
    scale_color_manual(values = rgn_colors, labels = rgn_labels) +
    scale_x_continuous(breaks = pretty_breaks(n = 6)) +
    scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 25)) +
    labs(
      title = goal_title,
      subtitle = "Current status (100 = best)",
      x = NULL,
      y = "Status",
      color = NULL
    ) +
    guides(color = guide_legend(nrow = 3, byrow = TRUE)) +
    theme_status

  # ggsave(
  #   here::here(safety_output, paste0("plot_current_status_", g, ".png")),
  #   p_status_list[[g]], width = 7, height = 4, dpi = 300
  # )
}

names(p_status_list) # list available goal plot names
```

#### Call individual current status plots by goal

```{r}
#| label: step_06a_print_status_by_goal
#| fig-width: 8
#| fig-height: 4.5
#| fig-dpi: 96

p_status_list[["biodiversity_habitats"]]
p_status_list[["biodiversity_species"]]
p_status_list[["clean_waters"]]
p_status_list[["coastal_opportunities_artisanal_fishing_opportunities"]]
p_status_list[["coastal_opportunities_coastal_access"]]
p_status_list[["food_provisioning_fisheries"]]
p_status_list[["food_provisioning_mariculture"]]
p_status_list[["habitat_services_carbon_storage"]]
p_status_list[["habitat_services_coastal_protection"]]
p_status_list[["habitat_services_filtration"]]
p_status_list[["livelihoods_economies_economies"]]
p_status_list[["livelihoods_economies_livelihoods"]]
p_status_list[["natural_products"]]
p_status_list[["research_ocean_literacy"]]
p_status_list[["sentido_de_pertenencia"]]
p_status_list[["tourism_recreation"]]
```

### 6b. Crimes included in each goal

For each OHI goal, we list the crimes that expert judgement marked as affecting that goal in `Safety_Security_Methodology_Hem_Crime_x_OHI_Goal.csv`.

```{r}
#| label: step_06b_crimes_by_goal

crimes_affecting_ohi <- read_csv(
  here::here(safety_int, "crimes_affecting_ohi_070626.csv")
)

crimes_by_goal <- crimes_affecting_ohi %>%
  dplyr::select(
    ohi_goal,
    crime_id,
    bien_juridico_afectado,
    tipo_de_delito,
    subtipo_de_delito,
    dplyr::any_of(c("eng_subtipo_de_delito", "eng_subtipo_de_delito_2"))
  ) %>%
  distinct() %>%
  arrange(ohi_goal, crime_id)

# save the full lookup for reference
write_csv(
  crimes_by_goal,
  here::here(safety_output, "crimes_included_per_goal_2015_2025.csv")
)

# print one table per goal
for (g in goal_list) {
  goal_title <- stringr::str_replace_all(g, "_", " ") %>%
    stringr::str_to_title()

  goal_crimes <- crimes_by_goal %>%
    filter(ohi_goal == g) %>%
    dplyr::select(-ohi_goal)

  cat("\n\n#### ", goal_title, "\n\n", sep = "")
  cat("Number of crimes included: ", nrow(goal_crimes), "\n\n", sep = "")

  print(
    knitr::kable(
      goal_crimes,
      caption = paste0("Crimes included in ", goal_title),
      row.names = FALSE
    )
  )
}
```

### 6c. Trying out different ways to calculate current status scores

Yes: using the **maximum crime rate across all goals, regions, and years** will usually push almost all scores even closer to 100. Only the single worst goal-region-year observation gets a status of 0. Every other observation is divided by an equal or larger denominator than in the per-goal max method, so scores can only stay the same or increase.

The **90th quantile** uses a less extreme reference than the absolute max, but is still closer to the upper tail than the 75th. Observations at or above that quantile are capped at a status of 0 (we use `pmax(0, ...)` so scores do not go negative). That should spread scores out more than the absolute max.

We also try a **per-region baseline**: for each region and goal, the reference is that region's mean crime rate from 2015 through 2020. Same formula, but now the score asks "how does this year compare to this region's own recent baseline for this goal?" If the current rate equals the 2015 through 2020 average, status is 0. If the rate is lower (safer), status rises toward 100. If the rate is higher than the baseline, we cap at 0.

Here we compare five methods side by side:

1. **Per-goal max** (what we already use): reference = max crime rate within each goal across regions and years
2. **Global max**: reference = max crime rate across all goals, regions, and years
3. **Per-goal 90th quantile**: reference = 90th percentile of crime rate within each goal; values above it score 0
4. **Global 90th quantile**: reference = 90th percentile across all goals, regions, and years; values above it score 0
5. **Per-region 2015 through 2020 mean**: reference = mean crime rate for that region and goal during 2015 through 2020; values above it score 0

```{r}
#| label: step_06c_status_alternatives
#| fig-width: 9
#| fig-height: 5
#| fig-dpi: 96

if (requireNamespace("showtext", quietly = TRUE)) {
  showtext::showtext_auto(FALSE)
}

rate_of_incidence_per_goal <- read_csv(
  here::here(safety_output, "rate_of_incidence_per_goal_2015_2025.csv")
)

# method 1: per-goal max (baseline)
max_by_goal <- rate_of_incidence_per_goal %>%
  group_by(ohi_goal) %>%
  summarise(ref_per_goal_max = max(crime_rate, na.rm = TRUE), .groups = "drop")

# method 2: one global max across all goals, regions, and years
ref_global_max <- max(rate_of_incidence_per_goal$crime_rate, na.rm = TRUE)
ref_global_max

# method 3: per-goal 90th quantile
q90_by_goal <- rate_of_incidence_per_goal %>%
  group_by(ohi_goal) %>%
  summarise(ref_per_goal_q90 = quantile(crime_rate, 0.90, na.rm = TRUE), .groups = "drop")

# method 4: global 90th quantile
ref_global_q90 <- quantile(rate_of_incidence_per_goal$crime_rate, 0.90, na.rm = TRUE)
ref_global_q90

# method 5: per-region mean crime rate for that goal during 2015 through 2020
ref_rgn_baseline <- rate_of_incidence_per_goal %>%
  filter(ano >= 2015, ano <= 2020) %>%
  group_by(rgn_id, ohi_goal) %>%
  summarise(
    ref_rgn_mean_2015_2020 = mean(crime_rate, na.rm = TRUE),
    .groups = "drop"
  )

ref_rgn_baseline

status_alternatives <- rate_of_incidence_per_goal %>%
  left_join(max_by_goal, by = "ohi_goal") %>%
  left_join(q90_by_goal, by = "ohi_goal") %>%
  left_join(ref_rgn_baseline, by = c("rgn_id", "ohi_goal")) %>%
  mutate(
    status_per_goal_max = (1 - (crime_rate / ref_per_goal_max)) * 100,
    status_global_max = (1 - (crime_rate / ref_global_max)) * 100,
    status_per_goal_q90 = pmax(0, (1 - (crime_rate / ref_per_goal_q90)) * 100), # cap at 0 above q90
    status_global_q90 = pmax(0, (1 - (crime_rate / ref_global_q90)) * 100), # cap at 0 above global q90
    # per-region baseline: handle zero baseline separately to avoid dividing by zero
    status_rgn_mean_2015_2020 = dplyr::case_when(
      is.na(ref_rgn_mean_2015_2020) ~ NA_real_,
      ref_rgn_mean_2015_2020 == 0 & crime_rate == 0 ~ 100,
      ref_rgn_mean_2015_2020 == 0 & crime_rate > 0 ~ 0,
      TRUE ~ pmax(0, (1 - (crime_rate / ref_rgn_mean_2015_2020)) * 100)
    )
  )

# overall summaries
summary(status_alternatives$status_per_goal_max)
summary(status_alternatives$status_global_max)
summary(status_alternatives$status_per_goal_q90)
summary(status_alternatives$status_global_q90)
summary(status_alternatives$status_rgn_mean_2015_2020)

# tidy long form for comparison tables and plots
status_long <- status_alternatives %>%
  dplyr::select(
    rgn_id, ano, ohi_goal,
    status_per_goal_max,
    status_global_max,
    status_per_goal_q90,
    status_global_q90,
    status_rgn_mean_2015_2020
  ) %>%
  tidyr::pivot_longer(
    cols = starts_with("status_"),
    names_to = "method",
    values_to = "current_status"
  ) %>%
  mutate(
    method = dplyr::recode(
      method,
      status_per_goal_max = "Per-goal max",
      status_global_max = "Global max",
      status_per_goal_q90 = "Per-goal 90th quantile",
      status_global_q90 = "Global 90th quantile",
      status_rgn_mean_2015_2020 = "Per-region mean 2015-2020"
    ),
    method = factor(
      method,
      levels = c(
        "Per-goal max",
        "Global max",
        "Per-goal 90th quantile",
        "Global 90th quantile",
        "Per-region mean 2015-2020"
      )
    )
  )

# summary by goal and method
status_alt_summary <- status_long %>%
  group_by(method, ohi_goal) %>%
  summarise(
    n = dplyr::n(),
    min = min(current_status, na.rm = TRUE),
    q1 = quantile(current_status, 0.25, na.rm = TRUE),
    median = median(current_status, na.rm = TRUE),
    mean = mean(current_status, na.rm = TRUE),
    q3 = quantile(current_status, 0.75, na.rm = TRUE),
    max = max(current_status, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(method, ohi_goal)

status_alt_summary

# overall summary by method only
status_alt_summary_overall <- status_long %>%
  group_by(method) %>%
  summarise(
    n = dplyr::n(),
    min = min(current_status, na.rm = TRUE),
    q1 = quantile(current_status, 0.25, na.rm = TRUE),
    median = median(current_status, na.rm = TRUE),
    mean = mean(current_status, na.rm = TRUE),
    q3 = quantile(current_status, 0.75, na.rm = TRUE),
    max = max(current_status, na.rm = TRUE),
    pct_above_95 = mean(current_status >= 95, na.rm = TRUE) * 100, # how piled up near 100
    pct_at_zero = mean(current_status == 0, na.rm = TRUE) * 100,
    .groups = "drop"
  )

status_alt_summary_overall

write_csv(
  status_alternatives,
  here::here(safety_output, "current_status_alternatives_2015_2025.csv")
)

write_csv(
  status_alt_summary,
  here::here(safety_output, "current_status_alternatives_summary_by_goal_2015_2025.csv")
)

write_csv(
  status_alt_summary_overall,
  here::here(safety_output, "current_status_alternatives_summary_overall_2015_2025.csv")
)
```

#### Compare score distributions across methods

```{r}
#| label: step_06c_compare_plots
#| fig-width: 10
#| fig-height: 5
#| fig-dpi: 96

# redefine in case this chunk is run without 6a
rgn_labels <- c(
  "Region_1" = "Western Upper Gulf",
  "Region_2" = "Western Midriff Islands",
  "Region_3" = "Western Central Gulf",
  "Region_4" = "Western Southern Gulf",
  "Region_5" = "Eastern Upper Gulf",
  "Region_6" = "Eastern Midriff Islands",
  "Region_7" = "Southern Sonora Coast",
  "Region_8" = "Sinaloa Coast",
  "Region_9" = "Southeastern Gulf"
)

rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

p_box_methods <- ggplot(status_long, aes(x = method, y = current_status, fill = method)) +
  geom_boxplot(outlier.size = 0.8, alpha = 0.8) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 25)) +
  labs(
    title = "Current status distributions by scoring method",
    subtitle = "All goals, regions, and years",
    x = NULL,
    y = "Current status"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 20, hjust = 1),
    plot.title = element_text(face = "bold")
  )

p_box_methods
```

#### Method comparison plots for each goal

One faceted plot per OHI goal, with the four scoring methods side by side. Plots are stored in `p_compare_list` so we can call any goal individually.

```{r}
#| label: step_06c_compare_by_goal
#| fig-width: 10
#| fig-height: 7
#| fig-dpi: 96

goal_list_compare <- sort(unique(status_long$ohi_goal))

p_compare_list <- list() # named list so we can call any goal plot later

for (g in goal_list_compare) {
  goal_title <- stringr::str_replace_all(g, "_", " ") %>%
    stringr::str_to_title()

  p_compare_list[[g]] <- status_long %>%
    filter(ohi_goal == g) %>%
    mutate(rgn_id = factor(rgn_id, levels = names(rgn_labels))) %>%
    ggplot(aes(x = ano, y = current_status, color = rgn_id, group = rgn_id)) +
    geom_line(linewidth = 0.9, alpha = 0.9) +
    geom_point(size = 1.5) +
    facet_wrap(~ method, ncol = 3) +
    scale_color_manual(values = rgn_colors, labels = rgn_labels) +
    scale_x_continuous(breaks = pretty_breaks(n = 5)) +
    scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 25)) +
    labs(
      title = paste0(goal_title, ": status under each scoring method"),
      x = NULL,
      y = "Current status",
      color = NULL
    ) +
    guides(color = guide_legend(nrow = 3, byrow = TRUE)) +
    theme_minimal(base_size = 11) +
    theme(
      legend.position = "bottom",
      plot.title = element_text(face = "bold"),
      strip.text = element_text(face = "bold", size = 9)
    )

  # ggsave(
  #   here::here(safety_output, paste0("plot_status_methods_", g, ".png")),
  #   p_compare_list[[g]], width = 10, height = 7, dpi = 300
  # )
}

names(p_compare_list) # list available goal plot names
```

#### Call individual method comparison plots by goal

```{r}
#| label: step_06c_print_compare_by_goal
#| fig-width: 10
#| fig-height: 7
#| fig-dpi: 96

p_compare_list[["biodiversity_habitats"]]
p_compare_list[["biodiversity_species"]]
p_compare_list[["clean_waters"]]
p_compare_list[["coastal_opportunities_artisanal_fishing_opportunities"]]
p_compare_list[["coastal_opportunities_coastal_access"]]
p_compare_list[["food_provisioning_fisheries"]]
p_compare_list[["food_provisioning_mariculture"]]
p_compare_list[["habitat_services_carbon_storage"]]
p_compare_list[["habitat_services_coastal_protection"]]
p_compare_list[["habitat_services_filtration"]]
p_compare_list[["livelihoods_economies_economies"]]
p_compare_list[["livelihoods_economies_livelihoods"]]
p_compare_list[["natural_products"]]
p_compare_list[["research_ocean_literacy"]]
p_compare_list[["sentido_de_pertenencia"]]
p_compare_list[["tourism_recreation"]]
```

