---
title: 'OHI GoC 2026: SS Safety Data Preparation (SESNSP)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor:
  markdown:
    wrap: sentence
  chunk_output_type: inline
editor_options:
  chunk_output_type: inline
---

# Overview

In this script we will prepare municipal crime data from SESNSP for the OHI GoC Safety and Security goal. The workflow below reads the newest available data through 2025, summarizes crimes by municipality, isolates Gulf of California coastal municipalities using our existing lookup table, joins those records to official INEGI municipal boundaries, and exports final tables and shapefiles.

1. **Load packages.** We load the tidyverse and spatial libraries used throughout the script.
2. **Read SESNSP municipal data through 2025.** We import the combined municipal CSV and standardize column names.
3. **Summarize annual crime counts by municipality.** Reshape monthly columns and sum to annual totals by crime type.
4. **Filter to Gulf of California states.** Keep obs from the six states that define our study region.
5. **Isolate coastal municipalities using the lookup table.**
   a. We read the municipality to region lookup table.
   b. We join SESNSP records to directly coastal municipalities.
6. **Join with INEGI municipal boundaries and map.**
   a. We read the INEGI municipal shapefile.
   b. We join coastal municipios to polygons and attach crime data.
   c. We preview the results in mapview.
7. **Export crime types for Gulf of California municipalities.** We save a lookup table of all crime categories in our study area.

Intermediate files are saved to `/home/shares/ohi/OHI_GOC/goal_prep/ss/safety_int`. Final outputs are saved to `/home/shares/ohi/OHI_GOC/goal_prep/ss/safety_output`.

# Background on SESNSP

The **Secretariado Ejecutivo del Sistema Nacional de Seguridad Publica** (SESNSP) is the federal agency responsible for coordinating public security policy in Mexico. Among its core functions, SESNSP compiles and publishes official statistics on crime incidence nationwide. These data are derived from **averiguaciones previas** and **carpetas de investigacion** (preliminary inquiries and case files) registered by state and federal prosecutors.

For the Safety and Security (SS) pressure, specifically Safety, we use the municipal level open data series, which reports the count of registered crimes by municipality, year, month, and crime classification. The official portal is here:

<https://www.gob.mx/sesnsp/acciones-y-programas/datos-abiertos-de-incidencia-delictiva>

For this update we use the 2015 through 2025 methodology series. The data dictionary and full methodology are documented here:

<https://sspcgob-my.sharepoint.com/:x:/g/personal/cni_sspc_gob_mx/Eft_H-La80NInrfnyRu9PCkBfyIyy-yr9hl8imxAyeLXMg?e=xvI38U>

**Incidencia Delictiva Fuero Común: Nueva Metodología**

| Campo | Descripción |
|:---|:---|
| Año | Año de registro de las averiguaciones previas y/o carpetas de investigación. |
| Clave_Ent | Clave de la entidad, según el Marco Geoestadístico Nacional (MGN) del Instituto Nacional de Geografía y Estadística (INEGI). |
| Entidad | Entidad federativa de registro de las averiguaciones previas y/o carpetas de investigación. |
| Bien jurídico afectado | Primera clasificación de los delitos en las averiguaciones previas y/o carpetas de investigación. |
| Tipo de delito | Segunda clasificación de los delitos. |
| Subtipo de delito | Tercera clasificación de los delitos. |
| Modalidad | Cuarta clasificación de los delitos. |
| Enero through diciembre | Mes de registro de las averiguaciones previas y/o carpetas de investigación. |

We downloaded the file `Municipal-Delitos-2015-2025_ene2026.csv` from this series and placed it in Mazu.

## 1. Load packages

```{r}
#| label: step_01_load_packages

library(tidyverse) # data manipulation and visualization
library(magrittr) # pipe operators
library(here) # project relative paths
library(janitor) # clean column names
library(readr) # read csv files
library(purrr) # functional programming tools
library(mapview) # interactive maps
library(sf) # spatial data handling
library(stringi) # normalize spanish place names

safety_int <- "/home/shares/ohi/OHI_GOC/goal_prep/ss/safety_int" # intermediate files
safety_output <- "/home/shares/ohi/OHI_GOC/goal_prep/ss/safety_output" # final outputs
muni_lookup_path <- "/home/shares/ohi/OHI_GOC/spatial/municipality_to_rgnid/municipality_to_rgnid_update.csv" # coastal municipio lookup
inegi_muni_path <- "/home/shares/ohi/OHI_GOC/spatial/delimitaciones_frontera_municipales_inegi/INEGI_Municipal_.shp" # inegi municipal polygons
```

## 2. Read SESNSP municipal data through 2025

We read the combined municipal file once rather than loading separate yearly spreadsheets. After cleaning column names we rename two fields so they match our earlier workflow.

```{r}
#| label: step_02_read_sesnsp

sesnsp_all <- readr::read_csv( # read the full municipal csv
  here::here(
    "/home/shares/ohi/OHI_GOC/_raw_data/SESNSP/d2026/Municipal-Delitos-2015-2025_ene2026",
    "Municipal-Delitos-2015-2025_ene2026.csv"
  ),
  locale = readr::locale(encoding = "latin1"), # spanish characters in source file
  show_col_types = FALSE
) %>%
  janitor::clean_names() # convert column names to snake case

unique(sesnsp_all$ano) # confirm all years are present.
# v2026: yes! 2015 through 2025
```

## 3. Summarize annual crime counts by municipality

Now we can pivot monthly columns to long format and sum across months to obtain an annual total for each crime type in each municipality.

```{r}
#| label: step_03_summarize_annual

sesnsp_all_long <- sesnsp_all %>% # reshape monthly columns to long format
  tidyr::pivot_longer(
    cols = enero:diciembre, # one row per month
    names_to = "mes",
    values_to = "n_delitos"
  )

sesnsp_anual <- sesnsp_all_long %>% # sum monthly counts to annual totals
  dplyr::group_by(
    ano,
    clave_ent,
    entidad,
    cve_municipio,
    municipio,
    bien_juridico_afectado,
    tipo_de_delito,
    subtipo_de_delito
  ) %>%
  dplyr::summarise(
    annual_total = sum(n_delitos, na.rm = TRUE),
    .groups = "drop"
  )

unique(sesnsp_anual$entidad) # list all states in the dataset
```

## 4. Filter to Gulf of California states

We keep municipalities in the six Mexican states that define the Gulf of California study region.

```{r}
#| label: step_04_filter_goc

goc_states <- c( # six states that border the goc
  "Baja California",
  "Baja California Sur",
  "Jalisco",
  "Nayarit",
  "Sinaloa",
  "Sonora"
)

sesnsp_anual_goc_states <- sesnsp_anual %>% # keep only goc states
  dplyr::filter(entidad %in% goc_states)

readr::write_csv(sesnsp_anual_goc_states, here::here(safety_int, "sesnsp_anual_goc_states.csv")) # save filtered table
```

## 5. Isolate coastal municipalities using the lookup table

Rather than geocoding municipality names, we can use the existing `municipality_to_rgnid` lookup table to identify Gulf of California coastal municipios and assign each one to an OHI region.

### 5a. Read the municipality to region lookup table

```{r}
#| label: step_05a_read_lookup

normalize_name <- function(x) { # helper to standardize place names for matching
  x %>%
    stringi::stri_trans_general("Latin-ASCII") %>%
    toupper() %>%
    stringr::str_squish()
}

muni_lookup <- readr::read_csv(muni_lookup_path) %>% # read curated goc municipio list
  dplyr::mutate(
    municipio = normalize_name(municipio), # clean lookup municipio names
    estado = normalize_name(estado), # clean lookup state names
    E03 = as.integer(E03), # inegi state code (first two digits of id)
    E04 = as.integer(E04), # inegi municipio code (last three digits of id)
    ID = sprintf("%02d%03d", E03, E04) # combined id to match inegi_muni$ID
  ) %>%
  dplyr::filter(directly_coastal == "Y") # keep only directly coastal municipios
```

### 5b. Join SESNSP records to coastal municipalities

The INEGI shapefile stores a single five digit `ID` for each municipio: the first two digits are the state code (`E03`) and the last three are the municipio code (`E04`). For example, Mexicali in Baja California (`E03` = 2, `E04` = 2) has `ID` = `"02002"`. We build the same `ID` from SESNSP's `cve_municipio` field and join on that key.

```{r}
#| label: step_05b_join_coastal

sesnsp_anual_goc_states_join <- sesnsp_anual_goc_states %>% # build inegi id from sesnsp municipio code
  dplyr::mutate(
    ID = sprintf("%05d", as.integer(cve_municipio)), # five digit id: state plus municipio
    municipio_norm = normalize_name(municipio), # standardized municipio name
    entidad_norm = normalize_name(entidad) # standardized state name
  )

not_in_coastal <- sesnsp_anual_goc_states_join %>% # check goc records that did not match coastal lookup
  dplyr::anti_join(muni_lookup, by = "ID") %>%
  dplyr::distinct(entidad, municipio, cve_municipio, ID) # non coastal municipios are expected here

sesnsp_anual_coastal <- sesnsp_anual_goc_states %>% # attach region ids to sesnsp crime records
  dplyr::inner_join(
    muni_lookup %>%
      dplyr::select(ID, rgn_id, E03, E04, both_pacific_gulf, directly_coastal),
    by = "ID"
  ) %>% 
  filter(!is.na(rgn_id)) # if rgn_id is NA, then it is a municipality on the pacific-side (facing the open ocean, not in the GoC)

unique(sesnsp_anual_coastal$municipio)
unique(sesnsp_anual_coastal$rgn_id)

readr::write_csv(sesnsp_anual_coastal, here::here(safety_int, "sesnsp_anual_coastal.csv")) # save joined crime table
```

## 6. Join with INEGI municipal boundaries and map

### 6a. Read the INEGI municipal shapefile

The municipal boundary file lives on mazu at `sftp://lecuona@mazu.nceas.ucsb.edu/home/shares/ohi/OHI_GOC/spatial/delimitaciones_frontera_municipales_inegi/INEGI_Municipal_.shp`. When working on mazu we read it from the shared path below.

```{r}
#| label: step_06a_read_inegi

inegi_muni <- sf::st_read(inegi_muni_path) # all official inegi municipal polygons sf

mapview(inegi_muni)
colnames(inegi_muni) # [1] "ID"       "NOMBRE"   "geometry"
```

### 6b. Join coastal municipios to polygons and attach crime data

To be able to see the municipalities spatially, we can join the lookup table to INEGI polygons on the shared five digit `ID`, then attach the summarized SESNSP records.

```{r}
#| label: step_06b_join_polygons

coastal_muni_sf <- inegi_muni %>% # isolate coastal municipio polygons by id
  dplyr::inner_join(
    muni_lookup %>%
      dplyr::select(ID, municipio, estado, rgn_id, E03, E04, both_pacific_gulf, directly_coastal),
    by = "ID"
  ) # NOTE THAT 25019, Sinaloa, Eldorado is a new municipality as of 2024...

sesnsp_coastal_sf <- coastal_muni_sf %>% # attach annual crime data to each coastal polygon
  dplyr::left_join(
    sesnsp_anual_coastal %>%
      dplyr::select(
        ID, ano, entidad, municipio, cve_municipio,
        bien_juridico_afectado, tipo_de_delito, subtipo_de_delito, annual_total
      ),
    by = "ID"
  )
```

### 6c. Preview coastal municipios in mapview

```{r}
#| label: step_06c_mapview

mapview::mapview(coastal_muni_sf, zcol = "rgn_id") # map coastal municipios colored by ohi region

mapview::mapview(sesnsp_coastal_sf, zcol = "rgn_id") # map municipios with attached crime attributes
```

## 7. Export crime types for Gulf of California municipalities

Now that we have isolated coastal municipalities in the Gulf of California, we make a lookup table of all crime types recorded by SESNSP in our study area. We also save the full spatial dataset for the next stage of goal preparation.

```{r}
#| label: step_07_export_crime_types

sesnsp_goc_crimes <- sesnsp_anual_coastal %>% # list unique crime classifications in study area
  dplyr::select(
    bien_juridico_afectado,
    tipo_de_delito,
    subtipo_de_delito
  ) %>%
  dplyr::distinct() %>% 
  mutate(crime_id = seq(1:55))

readr::write_csv( # save final crime type lookup table
  sesnsp_goc_crimes,
  here::here(safety_output, "types_of_crimes_goc_sesnsp_w2025.csv")
)

sf::st_write( # save final coastal municipio shapefile with crime data
  sesnsp_coastal_sf,
  here::here(safety_output, "sesnsp_coastal_municipios_w2025.gpkg")
)
```

## Next steps for Safety (SAF) within the Safety & Security (SS) pressure 

We will use the exported shapefile and crime type table in the next stage of goal preparation, where we select relevant crime categories and calculate regional scores.

