---
title: 'OHI GoC 2025: CCE'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# Sense of Pertenencia: Coastal Community Engagement (CCE)

## Adoption of beaches & maintained cultural heritage site employees (non-natural cultural sites)

CCE is the arithmetic mean of two subcomponents:

1. **Ama tu Playa and Adopta una Playa**: percent of SEMARNAT program inventory beaches already adopted in each OHI region (see `beach_adoption/ama_tu_playa.qmd`). Years 2019-2024 are scored as 0 because the campaign begins with the 2025-2030 strategy. 2025 uses the observed adoption rate.
2. **Non natural cultural places**: employees in cultural establishments per local coastal population, scaled to the **75th quantile** (capped score 0 to 100). For the SP merge we temporarily use `cultural_places_employees_per_pop_current_status_75th_quantile_gapfill2025.csv` (2025 copied from 2024 until DENUE 2025 exists). Alternative unit based and other quantile / 2019 reference versions are explored in `non_natural_cultural_lsp/non_natural_cultural_places.qmd`.

Ama tu Playa is part of Mexico's *Estrategia Nacional de Limpieza y Conservación de Playas y Costas de México 2025-2030*. It aims for beaches free of plastic waste by 2030 through community beach adoption and ongoing cleanups (<https://www.adoptaunaplaya.com/playas/>).

**Region 2 note:** no beaches from the program inventory fall in Region 2, so that region's Ama tu Playa row is explicit `ama_status = NA` with `reason = "no beaches in program inventory"` (not scored as 0 or 100). The full CCE and SP aggregation (arithmetic mean across subcomponents) lives in `cursor_generated/sp_current_status.qmd`.

```{r}
library(tidyverse)
library(here)
```

```{r}
ama_tu_playa_progress <- read_csv(here::here(
  "/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026/output",
  "current_status_ama_tu_playa_progress_040126.csv"
))

# support older single year csvs that lack a year column
if (!"year" %in% names(ama_tu_playa_progress)) {
  ama_tu_playa_progress <- ama_tu_playa_progress %>%
    mutate(year = 2025L)
}

# ensure region_2 is present with an explicit na + reason if an older csv is missing it
if (!"reason" %in% names(ama_tu_playa_progress)) {
  ama_tu_playa_progress <- ama_tu_playa_progress %>%
    mutate(reason = NA_character_)
}

if (!"Region_2" %in% ama_tu_playa_progress$rgn_id) {
  ama_tu_playa_progress <- bind_rows(
    ama_tu_playa_progress,
    tidyr::crossing(
      rgn_id = "Region_2",
      year = 2019:2025
    ) %>%
      mutate(
        adopted_beaches = NA_real_,
        total_beaches = 0,
        pct_adopted = NA_real_,
        current_status = NA_real_, # ama_status
        reason = "no beaches in program inventory"
      )
  )
}

ama_tu_playa_progress <- ama_tu_playa_progress %>% arrange(rgn_id, year)
colnames(ama_tu_playa_progress)
ama_tu_playa_progress %>% filter(rgn_id == "Region_2")
```

```{r}
cultural_places_scaled_to_population_cs <- read_csv(here::here(
  "/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026/output",
  "cultural_places_employees_per_pop_current_status_75th_quantile_gapfill2025.csv"
))
colnames(cultural_places_scaled_to_population_cs)
```
