################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
# Visualization
library(ggplot2)
library(plotly)
library(htmlwidgets)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```


# Explore: Mexico FAO data ------------------------------

```{r Mexico FAO Data}
## FAO data, some of the zeros were originally reported as NA (so exercise caution there)
harvest <- read_csv("https://raw.githubusercontent.com/OHI-Science/ohiprep_v2025/refs/heads/main/globalprep/mar/v2025/output/mar_harvest_tonnes.csv")
mex_harvest <- filter(harvest, rgn_id == 135)
mex_harvest <- filter(mex_harvest, year == 2017) %>% data.frame()

# data from Seafood Watch (Monterey Bay)
sustainability <- read_csv("https://raw.githubusercontent.com/OHI-Science/ohiprep_v2025/refs/heads/main/globalprep/mar/v2025/output/mar_sustainability.csv")
mex_sust <- filter(sustainability, taxa_code %in% mex_harvest$taxa_code) %>%
 filter(year==2025) %>% data.frame()

# determine whether sustainability was gapfilled (if so, fairly low confidence, depending on method used and taxa)
gf <- read_csv("https://raw.githubusercontent.com/OHI-Science/ohiprep_v2025/refs/heads/main/globalprep/mar/v2025/output/mar_sustainability_gf.csv")
mex_gf <- filter(gf, taxa_code %in% mex_harvest$taxa_code) %>% data.frame()
```


# Explore: CONAPESCA data shared by Juan Carlos (Sept 26 2025) ------------------------------

```{r CONAPESCA data shared by JC}
# These data were shared with us by Juan Carlos Villaseñor and originally accessed on September 26, 2025 from this GitHub repository: https://github.com/mex-fisheries

MexLandings_2000.present_latlong <- read_csv("/home/shares/ohi/OHI_GOC/_raw_data/CONAPESCA/d2025/SharedByJuanCarlos/mex_landings/data/clean/mex_landings_2000_present_with_lat_long.csv")
table(MexLandings_2000.present_latlong$state)
# n = 2,316,370 records from Baja, B.C.S., and Sonora from January 2000 through January 2025

MexLandings_2000.present<- readRDS(here("/home/shares/ohi/OHI_GOC/_raw_data/CONAPESCA/d2025/SharedByJuanCarlos/mex_landings/data/clean/mex_landings_2000_present.rds"))
# n = 12,153,879 records from all states in Mexico from January 2000 through January 2025
table(MexLandings_2000.present$state)

MexLandings_2018.present<- readRDS(here("/home/shares/ohi/OHI_GOC/_raw_data/CONAPESCA/d2025/SharedByJuanCarlos/mex_landings/data/clean/mex_conapesca_apertura_2018_present.rds"))

# Skimming through the dat. What does it look like?
head(MexLandings_2000.present)

table(MexLandings_2000.present$acuaculture_production) # Yes aquaculture production

max(MexLandings_2000.present$year)
max(MexLandings_2000.present$month[MexLandings_2000.present$year])
# Records up to January, 2025

```

```{r Basic filtering}
States_of_Interest<- c("BAJA CALIFORNIA", "BAJA CALIFORNIA SUR", "SONORA", "SINALOA", "NAYARIT", "JALISCO")

MexLandings_filtered<- MexLandings_2000.present %>% filter(state %in% States_of_Interest)
# n obs.= 4,713,144 (-7,440,735 obs)
# PRODUCTION NOTE: this operation filtered out 61% of all records and suggests that 39% of all fisheries records in Mexico (from 2000-present) come from these 6 states.
table(MexLandings_filtered$fleet)
# large_scale small_scale 
#      181823     1438462 
table(MexLandings_filtered$acuaculture_production) #there are also NAs in this column...

# WC.Fisheries_filtered<- MexLandings_filtered %>% filter(acuaculture_production== "NO") # n obs.= 3,701,563 (-1,011,581 obs)
# PRODUCTION NOTE: this operation filtered out 21% of all records from states of interest, suggesting that wild-catch fisheries account for 79% of all records.
WC.Fisheries_filtered<- MexLandings_filtered %>% filter(acuaculture_production== "NO" | is.na(acuaculture_production)) # 4684000 obs.
table(WC.Fisheries_filtered$year)

Mariculture_filtered<- MexLandings_filtered %>% filter(acuaculture_production== "SÍ")
# n obs.= 29,144 (-4,684,000 obs)
# PRODUCTION NOTE: this operation filtered out 99% of all records from states of interest and suggests that mariculture operations account for 1% of all records. This may or may not track with Mexico's total aquaculture production. Preliminary figure suggests that ~13.5% of Mexico's total fishery production is sources from aquaculture in 2024: https://mexicobusiness.news/agribusiness/news/mexicos-fisheries-and-aquaculture-sectors-see-growth-2024?utm_source=chatgpt.com

OTRAS_filtered<- MexLandings_filtered %>% filter(main_species_group== "OTRAS")
# n obs.= 715.547 (-1,011,581 obs)
# PRODUCTION NOTE: this operation filtered out 21% of all records from states of interest, suggesting that wild-catch fisheries account for 79% of all records.

# NOTES ON PRELIMINARY ANALYSIS: 
# SOME RECORDS ARE NOT APPENDED WITH INFORMATION ABOUT RECORD TYPE (i.e., no information on wild-catch versus aquaculture). These records are excluded from this preliminary analyses.
```
```{r Vessel summary table}
Vessel_Summary1 <- MexLandings_filtered %>%
  filter(year >= 2019 & year <=2024) %>%
  select(year, state, landing_site_key, eu_rnpa, eu_name, vessel_name) %>%
  group_by(year, state, landing_site_key) %>%
  summarise(
    Count_n_rnpa = n_distinct(eu_rnpa, na.rm = TRUE),
    Count_n_eu_name = n_distinct(eu_name, na.rm = TRUE),
    Count_n_unique_vessels = n_distinct(vessel_name, na.rm = TRUE),
    .groups = "drop"
  ) 

Vessel_Summary2 <- MexLandings_filtered %>%
  filter(year >= 2019 & year <=2024) %>%
  select(year, state, eu_rnpa, eu_name, vessel_name) %>%
  group_by(year, state) %>%
  summarise(
    Count_n_rnpa = n_distinct(eu_rnpa, na.rm = TRUE),
    Count_n_eu_name = n_distinct(eu_name, na.rm = TRUE),
    Count_n_unique_vessels = n_distinct(vessel_name, na.rm = TRUE),
    .groups = "drop"
  ) 
```

```{r W.C. cumulative percentage table}
# Summary table sorted by cumulative contribution of each fishery, per state, over the last 5 years:
WC.Fisheries_cumulative.contributions <- WC.Fisheries_filtered %>%
  filter(year >= 2019) %>%
  select(state, main_species_group, live_weight) %>%
  group_by(state, main_species_group) %>%
  summarise(
    Total_Live_Biomass = sum(live_weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(state) %>%
  mutate(
    State_Total = sum(Total_Live_Biomass, na.rm = TRUE),
    Percent_of_State = 100 * Total_Live_Biomass / State_Total
  ) %>%
  arrange(state, desc(Percent_of_State)) %>%
  mutate(
    Cumulative_Percent = cumsum(Percent_of_State)
  ) %>%
  ungroup() %>%
  mutate(
    Percent_of_State = sprintf("%.1f", Percent_of_State),
    Cumulative_Percent = sprintf("%.1f", Cumulative_Percent)
  )
```

```{r Plot cumulative percent}
df <- WC.Fisheries_cumulative.contributions %>%
  mutate(Percent_of_State = as.numeric(Percent_of_State)) %>%
  arrange(state, desc(Percent_of_State)) %>%
  mutate(
    # reorder within state for plotting
    main_species_group_ordered = reorder_within(main_species_group, Percent_of_State, state, fun = max, order = TRUE)
  )


W.C.Fisheries_cumulative_plot<- ggplot(df, aes(x = main_species_group_ordered, y = Percent_of_State)) +
  geom_segment(aes(xend = main_species_group_ordered, y = 0, yend = Percent_of_State), color = "pink") +
  geom_point(color = "tomato", size = 3) +
  facet_wrap(~state, scales = "free_y") +
  coord_flip() +
  scale_x_reordered() +   # <-- important for correct label display
  labs(
    title = "Species Group Contribution per State (Ordered by Descending Percent)",
    x = "Main Species Group",
    y = "Percentage of State Total (%)"
  ) +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))

W.C.Fisheries_cumulative_plot
W.C.Fisheries_cumulative_interactive <- ggplotly(W.C.Fisheries_cumulative_plot)
```

```{r Save plot}
ggsave(
  filename = "FIS/v2025/W.C. Fisheries Cumulative Plot_Sep 29 2025.png",  # file name
  plot = W.C.Fisheries_cumulative_plot,            # the plot object
  width = 10,                        # width in inches
  height = 10,                        # height in inches
  dpi = 300                         # resolution (dots per inch)
)

# Save your plotly object as HTML
htmlwidgets::saveWidget(W.C.Fisheries_cumulative_interactive, "FIS/v2025/W.C. Fisheries Cumulative Plotly_Sep 29 2025.html")
```

```{r MAR cumulative percentage table}
# Create a summary table of biomass landed from mariculture in the last 20 years... no data since 2019...
Mariculture_summary <- Mariculture_filtered %>%
  # filter(year >= 2019) %>%
  select(state, main_species_group, live_weight, landed_weight, value) %>%
  group_by(state, main_species_group) %>%
  summarise(
    Total_Live_Biomass = sum(live_weight, na.rm = TRUE),
    Total_Landed_Biomass = sum(landed_weight, na.rm = TRUE),
    Total_Value = sum(value, na.rm = TRUE),
    n_Records = n(),
    .groups = "drop"  # optional: ungroup after summarising
  )

# Summary table sorted by cumulative contribution of each mariculture stock, per state, over the last 20 years (no explicitly logged data since 2019)
Mariculture_cumulative.contributions <- Mariculture_filtered %>%
  #filter(year >= 2019) %>%
  select(state, main_species_group, live_weight) %>%
  group_by(state, main_species_group) %>%
  summarise(
    Total_Live_Biomass = sum(live_weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(state) %>%
  mutate(
    State_Total = sum(Total_Live_Biomass, na.rm = TRUE),
    Percent_of_State = 100 * Total_Live_Biomass / State_Total
  ) %>%
  arrange(state, desc(Percent_of_State)) %>%
  mutate(
    Cumulative_Percent = cumsum(Percent_of_State)
  ) %>%
  ungroup() %>%
  mutate(
    Percent_of_State = sprintf("%.1f", Percent_of_State),
    Cumulative_Percent = sprintf("%.1f", Cumulative_Percent)
  )
```

```{r Plot cumulative percent}
df <- Mariculture_cumulative.contributions %>%
  mutate(Percent_of_State = as.numeric(Percent_of_State)) %>%
  arrange(state, desc(Percent_of_State)) %>%
  mutate(
    # reorder within state for plotting
    main_species_group_ordered = reorder_within(main_species_group, Percent_of_State, state, fun = max, order = TRUE)
  )


Mariculture_cumulative_plot<- ggplot(df, aes(x = main_species_group_ordered, y = Percent_of_State)) +
  geom_segment(aes(xend = main_species_group_ordered, y = 0, yend = Percent_of_State), color = "lightblue") +
  geom_point(color = "blue", size = 3) +
  facet_wrap(~state, scales = "free_y") +
  coord_flip() +
  scale_x_reordered() +   # <-- important for correct label display
  labs(
    title = "Species Group Contribution per State (Ordered by Descending Percent)",
    x = "Main Species Group",
    y = "Percentage of State Total (%)"
  ) +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))

Mariculture_cumulative_plot
Mariculture_cumulative_interactive <- ggplotly(Mariculture_cumulative_plot)
```

```{r Save plot}
ggsave(
  filename = "MAR/v2025/Mariculture Stocks Cumulative Plot_Oct 1 2025.png",  # file name
  plot = Mariculture_cumulative_plot,            # the plot object
  width = 10,                        # width in inches
  height = 10,                        # height in inches
  dpi = 300                         # resolution (dots per inch)
)

# Save your plotly object as HTML
htmlwidgets::saveWidget(Mariculture_cumulative_interactive, "MAR/v2025/Mariculture Cumulative Plotly_Oct 1 2025.html")
```

```{r Summary table 2021}
# Create a summary table of biomass landed from w.c. fisheries in the last 5 years:
JCdata_Fisheries_summary <- WC.Fisheries_filtered %>%
  filter(year == 2021) %>%
  select(state, main_species_group, live_weight, landed_weight, value) %>%
  group_by(state, main_species_group) %>%
  summarise(
    Total_Live_Biomass_JCdata = sum(live_weight, na.rm = TRUE),
    Total_Landed_Biomass_JCdata = sum(landed_weight, na.rm = TRUE),
    Total_Value_JCdata = sum(value, na.rm = TRUE),
    n_Records_JCdata = n(),
    .groups = "drop"  # optional: ungroup after summarising
  )
```


# Explore: CONAPESCA data shared by dataMares (October 6 2025) ------------------------------

```{r CONAPESCA data 2001 thru 2023}
CONAPESCA_data<- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/CONAPESCA/d2025/SharedByDataMares/", "dataMares_CONAPESCA_Data_Oct6_2025.csv"))
# 4755692 obs.

# Make new date columns
CONAPESCA_data <- CONAPESCA_data %>%
  mutate(
    fecha = str_trim(fecha),  # Remove leading/trailing spaces
    fecha = na_if(fecha, "N/A"),
    fecha = na_if(fecha, "null"),
    fecha = na_if(fecha, ""),
    
    Date = parse_date_time(fecha, orders = c("dmy", "ymd")),  # Handle both formats
    year = year(Date),
    month = month(Date),
    day = day(Date)
  )

table(is.na(CONAPESCA_data$Date))          # How many are NA? NONE.

min(CONAPESCA_data$year)
max(CONAPESCA_data$year)
# weird... a couple of weird entries for date / year

CONAPESCA_data_filtered<- CONAPESCA_data %>% filter(year>= 2000)
CONAPESCA_data_filtered<- CONAPESCA_data_filtered %>% filter(year<= 2025)
# 4755602 obs.

table(CONAPESCA_data_filtered$`Tipo de pesca`)
table(CONAPESCA_data_filtered$year)
table(CONAPESCA_data_filtered$`Especie nombre general`)
table(CONAPESCA_data_filtered$zona.captura)
table(CONAPESCA_data_filtered$entidad.federativa)
head(CONAPESCA_data_filtered)

min(CONAPESCA_data_filtered$year) # 2000
max(CONAPESCA_data_filtered$year) # 2023

Otras<- CONAPESCA_data_filtered %>% filter(`Especie nombre general`=="OTRAS")
table(Otras$Especie_nombre_cientifico)

# Clean column names
CONAPESCA_data_filtered <- CONAPESCA_data_filtered %>%
  clean_names()
```
```{r Species per taxa table}
Species_per_Taxa <- CONAPESCA_data_filtered %>%
  select(especie_nombre_general, especie_nombre_cientifico) %>%
  group_by(especie_nombre_general) %>%
  summarise(
    n_species = n_distinct(especie_nombre_cientifico),
    species_list = paste(unique(especie_nombre_cientifico), collapse = ", ")
  )
```

```{r Summary table 2021}
# Create a summary table of biomass landed from w.c. fisheries in 2021
dataMares_Fisheries_summary <- CONAPESCA_data_filtered %>%
  filter(year == 2021) %>%
  select(entidad_federativa, especie_nombre_general, peso_kg, ganancia_total_mxn) %>%
  group_by(entidad_federativa, especie_nombre_general) %>%
  summarise(
    Total_Biomass_kg_dataMares = sum(peso_kg, na.rm = TRUE),
    Total_Value_mxn_dataMares = sum(ganancia_total_mxn, na.rm = TRUE),
    n_Records_dataMares = n(),
    .groups = "drop"  # optional: ungroup after summarising
  )

names(dataMares_Fisheries_summary)[1:2] <- c("state", "main_species_group")
```

```{r CPUE - dataMares Data}
CONAPESCA_data_filtered <- CONAPESCA_data_filtered %>%
  filter(dias_efectivos > 0) %>%
  mutate(CPUE_est = peso_kg / dias_efectivos)

max(CONAPESCA_data_filtered$CPUE_est) #NaN
min(CONAPESCA_data_filtered$CPUE_est) #NaN

CPUE_summary <- CONAPESCA_data_filtered %>%
  select(year, entidad_federativa, especie_nombre_general, dias_efectivos, peso_kg, ganancia_total_mxn, CPUE_est) %>%
  group_by(year, entidad_federativa, especie_nombre_general) %>%
  summarise(
    Total_Biomass_kg_dataMares = sum(peso_kg, na.rm = TRUE),
    Total_Days_Spent_Fishing = sum(dias_efectivos, na.rm= TRUE),
    Avg_CPUE_dataMares = round(mean(CPUE_est, na.rm= TRUE)),
    Total_Value_mxn_dataMares = sum(ganancia_total_mxn, na.rm = TRUE),
    n_Records_dataMares = n(),
    .groups = "drop"  # optional: ungroup after summarising
  )

glimpse(CPUE_summary)

Top10_species <- CONAPESCA_data_filtered %>%
  group_by(especie_nombre_general) %>%
  summarise(Total_Biomass = sum(peso_kg, na.rm = TRUE)) %>%
  arrange(desc(Total_Biomass)) %>%
  slice_head(n = 10) %>%
  pull(especie_nombre_general)

Top10_data <- CPUE_summary %>%
  filter(especie_nombre_general %in% Top10_species)

library(ggplot2)

CPUE_plot<- ggplot(Top10_data, aes(x = year, y = Avg_CPUE_dataMares, color = especie_nombre_general)) +
  facet_wrap(~ entidad_federativa, scales = "free_y") +
  geom_point(size = 2, alpha= 0.5) +
  geom_line(size= 1) +
  labs(
    title = "Preliminary Look at CPUE Over Time",
    subtitle = "Top 10 Species by Biomass (dataMares data)",
    x = "Year",
    y = "Average CPUE (kg/day)",
    color = "Species"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")

CPUE_plot

```

```{r Save plot}
ggsave(
  filename = "FIS/v2025/PreliminaryLook_CPUE_Oct6_2025.png",
  plot = CPUE_plot,
  width = 12,
  height = 10,
  dpi = 300
)
```


# Compare: J.C. versus dataMares data sets (year 2021) ------------------------------

```{r Compare these two datasets from 2021}
# Summary tables:
dataMares_Fisheries_summary <- CONAPESCA_data_filtered %>%
  filter(year == 2021) %>%
  select(entidad_federativa, especie_nombre_general, peso_kg, ganancia_total_mxn) %>%
  group_by(entidad_federativa, especie_nombre_general) %>%
  summarise(
    Total_Biomass_kg_dataMares = sum(peso_kg, na.rm = TRUE),
    Total_Value_mxn_dataMares = sum(ganancia_total_mxn, na.rm = TRUE),
    n_Records_dataMares = n(),
    .groups = "drop"  # optional: ungroup after summarising
  )

names(dataMares_Fisheries_summary)[1:2] <- c("state", "main_species_group")

JCdata_Fisheries_summary <- WC.Fisheries_filtered %>%
  filter(year == 2021) %>%
  select(state, main_species_group, live_weight, landed_weight, value) %>%
  group_by(state, main_species_group) %>%
  summarise(
    Total_Live_Biomass_JCdata = sum(live_weight, na.rm = TRUE),
    Total_Landed_Biomass_JCdata = sum(landed_weight, na.rm = TRUE),
    Total_Value_JCdata = sum(value, na.rm = TRUE),
    n_Records_JCdata = n(),
    .groups = "drop"  # optional: ungroup after summarising
  )


# Join tables:
Fisheries_summaries_merged <- full_join(
  dataMares_Fisheries_summary,
  JCdata_Fisheries_summary,
  by = intersect(names(dataMares_Fisheries_summary)[1:2],
                 names(JCdata_Fisheries_summary)[1:2])
)

Fisheries_summaries_merged <- Fisheries_summaries_merged %>%
  select(sort(names(.)))

# Count Discrepancy (assuming JC data is baseline)
Fisheries_summaries_merged$Count_discrepancy<- (Fisheries_summaries_merged$n_Records_JCdata - Fisheries_summaries_merged$n_Records_dataMares)
Fisheries_summaries_merged$Count_discrepancy.per <- round(
  100 * (Fisheries_summaries_merged$n_Records_JCdata - Fisheries_summaries_merged$n_Records_dataMares) / 
  Fisheries_summaries_merged$n_Records_JCdata)


# Biomass Discrepancy (assuming JC data is baseline)
Fisheries_summaries_merged$Biomass_discrepancy<- (Fisheries_summaries_merged$Total_Landed_Biomass_JCdata - Fisheries_summaries_merged$Total_Biomass_kg_dataMares)
 Fisheries_summaries_merged$Biomass_discrepancy.per <- round( 
  100 * (Fisheries_summaries_merged$Total_Landed_Biomass_JCdata - Fisheries_summaries_merged$Total_Biomass_kg_dataMares) / 
  Fisheries_summaries_merged$Total_Landed_Biomass_JCdata)
 
 
# Value Discrepancy (assuming JC data is baseline)
Fisheries_summaries_merged$Value_discrepancy<- (Fisheries_summaries_merged$Total_Value_JCdata - Fisheries_summaries_merged$Total_Value_mxn_dataMares)
 Fisheries_summaries_merged$Value_discrepancy.per <- round( 
  100 * (Fisheries_summaries_merged$Total_Value_JCdata - Fisheries_summaries_merged$Total_Value_mxn_dataMares) / 
  Fisheries_summaries_merged$Total_Value_JCdata)

```

```{r Save this data table}
write.csv(Fisheries_summaries_merged, "FIS/v2025/CONAPESCAdata_Summaries_Merged_EMF_Oct9_2025.csv", row.names = FALSE)
```


# Regenerate: J.C. CONAPESCA data with species names ------------------------------
```{r Regenerating J.C. CONAPESCA data with species names}
# How many species are fished in the GoC?
Species_data_dissertation <- read_csv("/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/CONAPESCA_SpMasterFile_EMF.csv")
length(unique(Species_data_dissertation$ScientificName))
# At least 496 unique species, according to a study I did in my dissertation (focused on a smaller sub-set of fisheries and area).

MexLandings_2000.present_wSpecies<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fis/v2025/int/juan_carlos/mex_landings_2000_present_EMF_edited.rds"))
# n = 12153879 obs.

length(unique(MexLandings_2000.present_wSpecies$species_name)) # 2052 unique species in all of Mexico.
# But remember... We don't need species from all of Mexico! Only the GoC region. Let's filter by state.

States_of_Interest<- c("BAJA CALIFORNIA", "BAJA CALIFORNIA SUR", "SONORA", "SINALOA", "NAYARIT", "JALISCO")

Landings_NWMexico <- MexLandings_2000.present_wSpecies %>% filter(state %in% States_of_Interest)
# n obs.= 4713144
length(unique(Landings_NWMexico$species_name)) 
# 1827 unique species in States of Interest.

# Filter OUT sites from Baja that are on the Pacific side:
# Lat-Long file (from Mel's repository):
Lat_Long_MasterFile<- read_csv("/home/frazier/ohi_goc_prep/FP/FIS/v2025/data/Baja_Sonora_states_key.csv")
colnames(Lat_Long_MasterFile) <- c("state", "landing_site", "landing_site_primary", "lat", "long", "GoC")

# Append lat-long info to landings data
temp_df <- Landings_NWMexico %>%
  left_join(
    Lat_Long_MasterFile,
    by = c("state", "landing_site")  # columns to match
  )

table(temp_df$GoC)
# 4713144 (total) - 996729 (no) = 3716415 obs. that are explicitly as "yes" in the GoC or NA. I have removed entries that are definitely "no" not in the Gulf.

# Generate filtered landings data for the GoC:
Landings_GoC <- temp_df %>%
  filter(GoC != "no" | is.na(GoC))
# 3716415 obs.

# Delete temp_df, not needed anymore.
rm(temp_df)

length(unique(Landings_GoC$species_name))
# 1693 unique species in the Gulf of California according to Juan Carlos' version of the CONAPESCA data base, keeping in mind that these are not really SPECIES (per se) but rather log-book entries for species.
length(unique(Landings_GoC$main_species_group))
# From 57 main species groups.
```
1693 unique species in the Gulf of California according to Juan Carlos' version of the CONAPESCA data base, keeping in mind that these are not really SPECIES (per se) but rather log-book entries for species.

```{r Save the Landings_GoC dataframe}
saveRDS(Landings_GoC, file="/home/shares/ohi/OHI_GOC/goal_prep/fis/v2025/int/juan_carlos/GoC_Landings_2000.present_EMF_created.rds")
```


# Explore: Species Names in CONAPESCA data ------------------------------

```{r Load in the Landings_GoC dataframe / skip the full file prep}
Landings_GoC<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fis/v2025/int/juan_carlos/GoC_Landings_2000.present_EMF_created.rds"))
# n = 3716415 obs.
```

```{r What if we do some filtering by groups?}
# How many species (log-book entries) are there per grouping?
Species_summary_table <- Landings_GoC %>%
  select(main_species_group, species_name) %>%
  group_by(main_species_group) %>%
  summarize(number_species_logbook = n_distinct(species_name))

# Filter out OTRAS, ESPECIE GENERICA, PECES DE ORNATO, & ORNATO
Landings_GoC_filtered <- Landings_GoC %>%
  filter(!main_species_group %in% c("OTRAS", "ESPECIE GENERICA", "PECES DE ORNATO", "ORNATO"))
# 3185035 obs.

length(unique(Landings_GoC_filtered$species_name))
# 1246 unique species, exluding otras and ornamentals...
# Still a lot...

# Start filtering out species names from notes about their preparation: 

# List of all key terms (these terms are not names, they are preparation notes):
keyword_list <- c(
  "DESC", "IND", "ENT", "FCO", "DESV", "S", "FCA", "DES", "DSC", 
  "ECO", "CONG", "CONGELADO", "DE CULTIVO", "DE CULTIVO VIVO", 
  "FILETE", "FILETE DE", "ECA", "CABEZA DE", "FILETE DE ECO", 
  "PIEL DE ECA", "PIEL DE SECA", "ALETA ECA DE", "DESCABEZADO", "FRESCO", "SECO DE", "ALETA", "SECA", "SECA DE", "SECO", "SECO DE", "DE ALTA MAR", "ENTERO",
  "POLVO DE", "CONCHA DE", "COCINERO"
)

# Sort by descending length to match multi-word keywords first
keyword_list <- keyword_list[order(nchar(keyword_list), decreasing = TRUE)]

GoC_Sp <- Landings_GoC_filtered %>%
  select(main_species_group, recorded_species_name = species_name) %>%
  distinct() %>%
  
  # Extract keywords as whole words with optional punctuation
  mutate(
    keywords = str_extract_all(
      recorded_species_name,
      paste0("\\b(", paste(keyword_list, collapse="|"), ")\\b[.,]?")
    )
  ) %>%
  
  # Pad for unnesting (if you want fixed number of columns)
  mutate(keywords = lapply(keywords, function(x) { length(x) <- 5; x })) %>%
  unnest_wider(keywords, names_sep = "_") %>%
  rename(
    keyword_1 = keywords_1,
    keyword_2 = keywords_2,
    keyword_3 = keywords_3,
    keyword_4 = keywords_4,
    keyword_5 = keywords_5
  ) %>%
  
  # Remove keywords from species name, except for "TIBURON ALETA BLANCA", "BERRUGATA ALETA AMARILLA", "CORVINA ALETA CORTA", "CORVINA ALETA AMARILLA"
  rowwise() %>%
  mutate(
    recorded_species_name_clean = if_else(
      recorded_species_name == c("TIBURON ALETA BLANCA", "BERRUGATA ALETA AMARILLA", "CORVINA ALETA CORTA", "CORVINA ALETA AMARILLA"),
      recorded_species_name,  # leave as-is
      str_remove_all(
        recorded_species_name,
        paste0("\\b(", paste(keyword_list, collapse="|"), ")\\b[.,]?")
      )
    ),
    recorded_species_name_clean = str_remove_all(recorded_species_name_clean, "\\bY\\b"),
    recorded_species_name_clean = str_squish(recorded_species_name_clean)
  ) %>%
  ungroup()

GoC_Sp <- GoC_Sp %>%
  arrange(main_species_group, recorded_species_name_clean)

unique(GoC_Sp$keyword_1)
unique(GoC_Sp$recorded_species_name_clean)

#write.csv(GoC_Sp, "/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/PrelimLook_CONAPESCA_SpeciesNames_Oct302025/AllGoC_SpKey_Oct312025.csv", row.names = FALSE)
```

```{r Test case: how many sp of Abulon and Anchoveta}
unique(Landings_GoC$main_species_group)

Abulon_Landings<- Landings_GoC %>% filter(main_species_group=="ABULON")
Abulon_Sp<- unique(Abulon_Landings$species_name)
# 19 "species", that is common names

# BOTETE DOES NOT EXIST IN THIS DATASET
# Botete_Landings<- Landings_GoC %>% filter(main_species_group=="BOTETE")
# Botete_Sp<- unique(Botete_Landings$species_name)

Anchoveta_Landings<- Landings_GoC %>% filter(main_species_group=="ANCHOVETA")
Anchoveta_Sp<- unique(Anchoveta_Landings$species_name)
# 9 "species", that is common names

# Write to CSV
# Combine vectors as columns:
Combined_Sp <- data.frame(
  main_species_group  = c(rep("ANCHOVETA", length(Anchoveta_Sp)),
              rep("ABULON", length(Abulon_Sp))),
  common_name = c(Anchoveta_Sp, Abulon_Sp)
)

#write.csv(Combined_Sp, "/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/Abulon&Anchoveta_Sp_Oct302025.csv", row.names = FALSE)
```

```{r ABALONE}
# Added information about each species manually and offline. 
# Re-import this new file:
Species_Info <- read_csv("/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/Abulon&Anchoveta_Sp_Oct302025.csv")

##### ABALONE
Abulon_Landings<- Landings_GoC %>% filter(main_species_group=="ABULON")

Abulon_Landings<- left_join(Abulon_Landings, Species_Info, 
          by = c("main_species_group", "species_name"="recorded_species_name"))

Abulon_Landings_summary<- Abulon_Landings %>%
  filter(year > 2014 & year < 2025) %>%
  select(year, state, family, estimated_common_name, live_weight) %>%
  group_by(year, state, family, estimated_common_name) %>%
  summarize(TotalBiomass_annual= sum(live_weight))

# Let's plot what this looks like:
Abulon_Biomass_plot<- ggplot(Abulon_Landings_summary,
       aes(x = year,
           y = TotalBiomass_annual,        # match your new column name
           color = estimated_common_name,
           group = estimated_common_name)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  facet_wrap(~ state, scales = "free_y") +
  labs(
    title = "Abalone Total Biomass (2015–2024)",
    x = "Year",
    y = "Total Biomass (kg)",
    color = "Common Name"
  ) +
  theme_bw(base_size = 13) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

# Save the plot as PNG
# ggsave(filename = "/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/PrelimLook_CONAPESCA_SpeciesNames_Oct302025/PrelimLook_SpBiomassPlots_Nov2025/Abalone.Biomass_plot_Oct302025.png",
#        plot = Abulon_Biomass_plot,
#        width = 10,        # width in inches
#        height = 6,        # height in inches
#        dpi = 300)         # resolution
```

```{r ANCHOVY}
##### ANCHOVY
Anchoveta_Landings<- Landings_GoC %>% filter(main_species_group=="ANCHOVETA")

Anchoveta_Landings<- left_join(Anchoveta_Landings, Species_Info, 
          by = c("main_species_group", "species_name"="recorded_species_name"))

Anchoveta_Landings_summary<- Anchoveta_Landings %>%
  filter(year > 2014 & year < 2025) %>%
  select(year, state, family, estimated_common_name, live_weight) %>%
  group_by(year, state, family, estimated_common_name) %>%
  summarize(TotalBiomass_annual= sum(live_weight))

# Let's plot what this looks like:
Anchoveta_Biomass_plot<- ggplot(Anchoveta_Landings_summary,
       aes(x = year,
           y = TotalBiomass_annual,        # match your new column name
           color = estimated_common_name,
           group = estimated_common_name)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  facet_wrap(~ state, scales = "free_y") +
  labs(
    title = "Anchovy Total Biomass (2015–2024)",
    x = "Year",
    y = "Total Biomass (kg)",
    color = "Common Name"
  ) +
  theme_bw(base_size = 13) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

# Save the plot as PNG
# ggsave(filename = "/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/PrelimLook_CONAPESCA_SpeciesNames_Oct302025/PrelimLook_SpBiomassPlots_Nov2025/Anchovy.Biomass_plot_Oct302025.png", 
#        plot = Anchoveta_Biomass_plot,
#        width = 10,        # width in inches
#        height = 6,        # height in inches
#        dpi = 300)         # resolution
```

```{r Top Landed Species}
# What if I filtered by the top catch groupings over the last 10 years?
Landings_GoC_cumulative.contributions <- Landings_GoC_filtered %>%
  filter(year > 2014 & year < 2025) %>%
  select(state, main_species_group, live_weight) %>%
  group_by(state, main_species_group) %>%
  summarise(
    Total_Live_Biomass = sum(live_weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(state) %>%
  mutate(
    State_Total = sum(Total_Live_Biomass, na.rm = TRUE),
    Percent_of_State = 100 * Total_Live_Biomass / State_Total
  ) %>%
  arrange(state, desc(Percent_of_State)) %>%
  mutate(
    Cumulative_Percent = cumsum(Percent_of_State)
  ) %>%
  ungroup() %>%
  mutate(
    Percent_of_State = sprintf("%.1f", Percent_of_State),
    Cumulative_Percent = sprintf("%.1f", Cumulative_Percent)
  )

# Create a list of main_species_group for which percent_state catch was >= than 10% (during this timeframe)

Landings_GoC_TopGroupings_summary <- Landings_GoC_filtered %>%
  filter(year > 2014 & year < 2025) %>%
  select(state, main_species_group, live_weight) %>%
  group_by(state, main_species_group) %>%
  summarise(Total_Live_Biomass = sum(live_weight, na.rm = TRUE), .groups = "drop") %>%
  group_by(state) %>%
  mutate(
    State_Total = sum(Total_Live_Biomass, na.rm = TRUE),
    Percent_of_State = 100 * Total_Live_Biomass / State_Total
  ) %>%
  filter(Percent_of_State >= 10) %>%
  ungroup()

TopLandedGroupings<- unique(Landings_GoC_TopGroupings_summary$main_species_group)

# Which groups would we be dropping?
setdiff(Landings_GoC_filtered$main_species_group,
        Landings_GoC_TopGroupings_summary$main_species_group)

unique(Landings_GoC_filtered$main_species_group)
unique(Landings_GoC_TopGroupings_summary$main_species_group)

# Okay, so now let's filter for only the Top Landed Groupings
Landings_GoC_TopGroupings <- Landings_GoC %>%
  filter(main_species_group %in% TopLandedGroupings)
# 1757236 obs.

length(unique(Landings_GoC_TopGroupings$species_name))
# Only 600 species according to this filtration!
unique(Landings_GoC_TopGroupings$species_name)

# But how much (biomass) information do we lose when we filter like this?
TotalBiomass_TopGroupings<- Landings_GoC_TopGroupings %>%
  filter(year > 2014 & year < 2025) %>%
  select(state, live_weight) %>%
  group_by(state) %>%
  summarize(TotalBiomass_TopGroupings_20142024= sum(live_weight))

TotalBiomass_AllGroupings<- Landings_GoC %>%
  filter(year > 2014 & year < 2025) %>%
  select(state, live_weight) %>%
  group_by(state) %>%
  summarize(TotalBiomass_AllGroupings_20142024= sum(live_weight))

# Merge these two summary tables together by common columns:
BiomassCompare_AllvsTopGroupings <- TotalBiomass_AllGroupings %>%
  left_join(
    TotalBiomass_TopGroupings,
    by = c("state")  # columns to match
  )

BiomassCompare_AllvsTopGroupings$Biomass_Discrepancy<- BiomassCompare_AllvsTopGroupings$TotalBiomass_AllGroupings_20142024 - BiomassCompare_AllvsTopGroupings$TotalBiomass_TopGroupings_20142024

BiomassCompare_AllvsTopGroupings$TopGroupings_PercentTotal<- 100 * (BiomassCompare_AllvsTopGroupings$TotalBiomass_TopGroupings_20142024 /
BiomassCompare_AllvsTopGroupings$TotalBiomass_AllGroupings_20142024)

# fewer number of decimal digits for display
BiomassCompare_AllvsTopGroupings <- BiomassCompare_AllvsTopGroupings %>%
  mutate(
    across(where(is.numeric), ~ round(., 2))  # rounds all numeric columns to 2 decimals
  )

# The short answer is that we lose anywhere from 15-41% of the biomass, depending on the state
```
Next steps: export these [top-landed] species as a dataframe and I'll append the species, genus, and family info offline.


#### Create: Species IDs for Top-Landed Taxa
```{r Filter out keywords from recorded_species_name to get recorded_species_name_clean}
# List of all key terms
keyword_list <- c(
  "DESC", "IND", "ENT", "FCO", "DESV", "S", "FCA", "DES", "DSC", 
  "ECO", "CONG", "CONGELADO", "DE CULTIVO", "DE CULTIVO VIVO", 
  "FILETE", "FILETE DE", "ECA", "CABEZA DE", "FILETE DE ECO", 
  "PIEL DE ECA", "PIEL DE SECA", "ALETA ECA DE", "DESCABEZADO", "FRESCO", "SECO DE", "ALETA", "SECA", "SECA DE", "SECO", "SECO DE", "DE ALTA MAR", "ENTERO",
  "POLVO DE", "CONCHA DE", "COCINERO"
)

# Sort by descending length to match multi-word keywords first
keyword_list <- keyword_list[order(nchar(keyword_list), decreasing = TRUE)]

TopLanded_Sp <- Landings_GoC_TopGroupings %>%
  select(main_species_group, recorded_species_name = species_name) %>%
  distinct() %>%
  
  # Extract keywords as whole words with optional punctuation
  mutate(
    keywords = str_extract_all(
      recorded_species_name,
      paste0("\\b(", paste(keyword_list, collapse="|"), ")\\b[.,]?")
    )
  ) %>%
  
  # Pad for unnesting (if you want fixed number of columns)
  mutate(keywords = lapply(keywords, function(x) { length(x) <- 5; x })) %>%
  unnest_wider(keywords, names_sep = "_") %>%
  rename(
    keyword_1 = keywords_1,
    keyword_2 = keywords_2,
    keyword_3 = keywords_3,
    keyword_4 = keywords_4,
    keyword_5 = keywords_5
  ) %>%
  
  # Remove keywords from species name, except for "TIBURON ALETA BLANCA"
  rowwise() %>%
  mutate(
    recorded_species_name_clean = if_else(
      recorded_species_name == "TIBURON ALETA BLANCA",
      recorded_species_name,  # leave as-is
      str_remove_all(
        recorded_species_name,
        paste0("\\b(", paste(keyword_list, collapse="|"), ")\\b[.,]?")
      )
    ),
    recorded_species_name_clean = str_remove_all(recorded_species_name_clean, "\\bY\\b"),
    recorded_species_name_clean = str_squish(recorded_species_name_clean)
  ) %>%
  ungroup()

TopLanded_Sp <- TopLanded_Sp %>%
  arrange(main_species_group, recorded_species_name_clean)

unique(TopLanded_Sp$keyword_1)
unique(TopLanded_Sp$recorded_species_name_clean)
# 281 names

#write.csv(TopLanded_Sp, "/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/PrelimLook_CONAPESCA_SpeciesNames_Oct302025/TopLanded_Sp_Oct312025.csv", row.names = FALSE)
```

```{r Use taxize to determine scientific names where possible}
library(taxize)
library(worrms)
library(stringr)
library(tidyr)
library(dplyr)
library(rgbif)

# Step 1: Extract scientific names from parentheses
TopLanded_Sp$scientific_name <- sub(".*\\(([^)]+)\\).*", "\\1", TopLanded_Sp$recorded_species_name_clean)
TopLanded_Sp$scientific_name[!grepl("\\(", TopLanded_Sp$recorded_species_name_clean)] <- NA
TopLanded_Sp$scientific_name <- trimws(TopLanded_Sp$scientific_name)

# Convert to proper binomial capitalization
capitalize_binomial <- function(x){
  x <- tolower(x)                    # all lowercase
  words <- strsplit(x, " ")[[1]]     # split words
  if(length(words) >= 2){
    words[1] <- paste0(toupper(substring(words[1],1,1)), substring(words[1],2))  # Genus capitalized
    words[2] <- words[2]  # species lowercase
    return(paste(words[1], words[2]))
  } else {
    return(x)
  }
}

TopLanded_Sp$scientific_name_clean <- sapply(TopLanded_Sp$scientific_name, function(x){
  if(!is.na(x)) capitalize_binomial(x) else NA
})

# Step 2: Unique names to query
taxa_list <- unique(na.omit(TopLanded_Sp$scientific_name_clean))

# Step 3: Get GBIF IDs safely
# function to get classification from GBIF
get_gbif_classification <- function(name){
  # search for the name
  res <- name_backbone(name = name, strict = FALSE)  # fuzzy search
  if(is.null(res$species)) res$species <- NA
  if(is.null(res$genus)) res$genus <- NA
  if(is.null(res$family)) res$family <- NA
  data.frame(
    scientific_name = name,
    species = res$species,
    genus = res$genus,
    family = res$family,
    stringsAsFactors = FALSE
  )
}

gbif_results <- do.call(rbind, lapply(taxa_list, get_gbif_classification))

# Step 4: Merge
# Merge GBIF taxonomy back to the original dataframe
TopLanded_Sp <- merge(
  TopLanded_Sp,
  gbif_results,
  by.x = "scientific_name_clean",   # column in your original df
  by.y = "scientific_name",         # column in gbif_results
  all.x = TRUE                      # keep all rows from TopLanded_Sp
)

#write.csv(TopLanded_Sp, "/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/PrelimLook_CONAPESCA_SpeciesNames_Oct302025/TopLanded_Sp_Nov032025.csv", row.names = FALSE)
```

```{r Mannually adding species info Nov 3-Nov 5 2025}
dataMares_CONAPESCA_data<- read_csv(here("/home/shares/ohi/OHI_GOC/_raw_data/CONAPESCA/d2025/SharedByDataMares/", "dataMares_CONAPESCA_Data_Oct6_2025.csv"))
# 4755692 obs.

#### Manually idetifying species from common names - Nov 3 through Nov 5
TEST<- dataMares_CONAPESCA_data %>% filter(`Especie nombre general`=="MOJARRA")
unique(TEST$`Especie nombre`)
unique(TEST$Especie_nombre_cientifico)
unique(TEST$Especie_nombre_cientifico[which(TEST$`Especie nombre`=="PALA")])

get_gbifid_("Vieja", method= "backbone")

#### SPECIES LIST - Nov 5 2025
TopLanded_Sp_List<- read_csv(here("/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/PrelimLook_CONAPESCA_SpeciesNames_Oct302025/TopLanded_Sp_Nov052025.csv"))
# 600 obs., 17 columns
unique(TopLanded_Sp_List$common_name_english_estimated)

# Helpful summary table for Mel - Nov 5
TopLanded_Sp_List_summarized <- TopLanded_Sp_List %>%
  select(common_name_english_estimated, species_estimated, genus_estimated, family_estimated) %>%
  group_by(
    species = species_estimated,
    genus = genus_estimated,
    family = family_estimated
  ) %>%
  summarize(
    associated_common_names = paste(unique(common_name_english_estimated), collapse = ", "),
    .groups = "drop"
  ) %>%
  arrange(species, genus, family)

unique(TopLanded_Sp_List_summarized$species)
```

```{r Save this summarized list species names for Mel - Nov 5 2025}
write.csv(TopLanded_Sp_List_summarized, "PrelimLook_CONAPESCA_SpeciesNames_Oct302025/TopLanded_Sp_Summarized_Nov052025.csv", row.names = FALSE)
```


#### Plot: Biomass of Top-Landed Taxa / Species
Added information about each top-landed species manually, offline.
```{r Append species info to the CONAPESCA landings data}
TopLanded_Sp_List <- read_csv("/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/PrelimLook_CONAPESCA_SpeciesNames_Oct302025/TopLanded_Sp_Nov052025.csv")
TopLanded_Sp_List<- TopLanded_Sp_List[-c(1)] #delete the index column
```

```{r CAMARON}
##### CAMARON
Camaron_Landings <- Landings_GoC_TopGroupings %>% filter(main_species_group=="CAMARON")
# 555356 obs.

Camaron_Landings<- left_join(Camaron_Landings, TopLanded_Sp_List, 
          by = c("main_species_group", "species_name"="recorded_species_name"))
unique(Camaron_Landings$species_estimated)

Camaron_Landings_summary<- Camaron_Landings %>%
  filter(year > 2014 & year < 2025) %>%
  select(year, state, family_estimated, species_estimated, live_weight) %>%
  group_by(year, state, family_estimated, species_estimated) %>%
  summarize(TotalBiomass_annual= sum(live_weight))

# Let's plot what this looks like:
Camaron_Biomass_plot<- ggplot(Camaron_Landings_summary,
       aes(x = year,
           y = TotalBiomass_annual,        # match your new column name
           color = species_estimated,
           group = species_estimated)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  facet_wrap(~ state, scales = "free_y") +
  labs(
    title = "Shrimp Total Biomass (2015–2024)",
    x = "Year",
    y = "Total Biomass (kg)",
    color = "Common Name"
  ) +
  theme_bw(base_size = 13) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

# Save the plot as PNG
# ggsave(filename = "/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/PrelimLook_CONAPESCA_SpeciesNames_Oct302025/Shrimp.Biomass_plot_Nov032025.png",
#        plot = Camaron_Biomass_plot,
#        width = 10,        # width in inches
#        height = 6,        # height in inches
#        dpi = 300)         # resolution
```

```{r CORVINA}
##### CORVINA
Covina_Landings<- Landings_GoC_TopGroupings %>% filter(main_species_group=="CORVINA")
# 172667 obs.

Covina_Landings<- left_join(Covina_Landings, TopLanded_Sp_List, 
          by = c("main_species_group", "species_name"="recorded_species_name"))

unique(Covina_Landings$species_estimated)

Corvina_Landings_summary<- Covina_Landings %>%
  filter(year > 2014 & year < 2025) %>%
  select(year, state, family_estimated, species_estimated, live_weight) %>%
  group_by(year, state, family_estimated, species_estimated) %>%
  summarize(TotalBiomass_annual= sum(live_weight))

# Let's plot what this looks like:
Corvina_Biomass_plot<- ggplot(Corvina_Landings_summary,
       aes(x = year,
           y = TotalBiomass_annual,        # match your new column name
           color = species_estimated,
           group = species_estimated)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  facet_wrap(~ state, scales = "free_y") +
  labs(
    title = "Corvina Total Biomass (2015–2024)",
    x = "Year",
    y = "Total Biomass (kg)",
    color = "Common Name"
  ) +
  theme_bw(base_size = 13) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

Corvina_Biomass_plot

# Save the plot as PNG
# ggsave(filename = "/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/PrelimLook_CONAPESCA_SpeciesNames_Oct302025/Corvina.Biomass_plot_Nov032025.png",
#        plot = Corvina_Biomass_plot,
#        width = 10,        # width in inches
#        height = 6,        # height in inches
#        dpi = 300)         # resolution
```

```{r ATUN}
##### ATUN
Atun_Landings<- Landings_GoC_TopGroupings %>% filter(main_species_group=="ATUN")
# 13729 obs.

Atun_Landings<- left_join(Atun_Landings, TopLanded_Sp_List, 
          by = c("main_species_group", "species_name"="recorded_species_name"))

Atun_Landings_summary<- Atun_Landings %>%
  filter(year > 2014 & year < 2025) %>%
  select(year, state, family_estimated, species_estimated, live_weight) %>%
  group_by(year, state, family_estimated, species_estimated) %>%
  summarize(TotalBiomass_annual= sum(live_weight))

# Let's plot what this looks like:
Atun_Biomass_plot<- ggplot(Atun_Landings_summary,
       aes(x = year,
           y = TotalBiomass_annual,        # match your new column name
           color = species_estimated,
           group = species_estimated)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  facet_wrap(~ state, scales = "free_y") +
  labs(
    title = "Tuna Total Biomass (2015–2024)",
    x = "Year",
    y = "Total Biomass (kg)",
    color = "Common Name"
  ) +
  theme_bw(base_size = 13) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

Atun_Biomass_plot

# Save the plot as PNG
# ggsave(filename = "/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/PrelimLook_CONAPESCA_SpeciesNames_Oct302025/Tuna.Biomass_plot_Nov032025.png",
#        plot = Atun_Biomass_plot,
#        width = 10,        # width in inches
#        height = 6,        # height in inches
#        dpi = 300)         # resolution
```

```{r GUACHINANGO}
##### GUACHINANGO
Guachinango_Landings<- Landings_GoC_TopGroupings %>% filter(main_species_group=="GUACHINANGO")
# 115796 obs.

Guachinango_Landings<- left_join(Guachinango_Landings, TopLanded_Sp_List, 
          by = c("main_species_group", "species_name"="recorded_species_name"))

unique(Guachinango_Landings$species_estimated)

Guachinango_Landings_summary<- Guachinango_Landings %>%
  filter(year > 2014 & year < 2025) %>%
  select(year, state, family_estimated, species_estimated, live_weight) %>%
  group_by(year, state, family_estimated, species_estimated) %>%
  summarize(TotalBiomass_annual= sum(live_weight))

# Let's plot what this looks like:
Guachinango_Biomass_plot<- ggplot(Guachinango_Landings_summary,
       aes(x = year,
           y = TotalBiomass_annual,        # match your new column name
           color = species_estimated,
           group = species_estimated)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  facet_wrap(~ state, scales = "free_y") +
  labs(
    title = "Snapper Total Biomass (2015–2024)",
    x = "Year",
    y = "Total Biomass (kg)",
    color = "Common Name"
  ) +
  theme_bw(base_size = 13) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

Guachinango_Biomass_plot

# Save the plot as PNG
# ggsave(filename = "/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/PrelimLook_CONAPESCA_SpeciesNames_Oct302025/PrelimLook_SpBiomassPlots_Nov2025/Snapper.Biomass_plot_Nov062025.png",
#        plot = Guachinango_Biomass_plot,
#        width = 10,        # width in inches
#        height = 6,        # height in inches
#        dpi = 300)         # resolution
```

```{r JUREL}
##### JUREL
Jurel_Landings<- Landings_GoC_TopGroupings %>% filter(main_species_group=="JUREL")
# 43338 obs.

Jurel_Landings<- left_join(Jurel_Landings, TopLanded_Sp_List, 
          by = c("main_species_group", "species_name"="recorded_species_name"))

Jurel_Landings_summary<- Jurel_Landings %>%
  filter(year > 2014 & year < 2025) %>%
  select(year, state, family_estimated, live_weight) %>%
  group_by(year, state, family_estimated) %>%
  summarize(TotalBiomass_annual= sum(live_weight))

# Let's plot what this looks like:
Jurel_Biomass_plot<- ggplot(Jurel_Landings_summary,
       aes(x = year,
           y = TotalBiomass_annual,        # match your new column name
           color = family_estimated,
           group = family_estimated)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  facet_wrap(~ state, scales = "free_y") +
  labs(
    title = "Jurel Total Biomass (2015–2024)",
    x = "Year",
    y = "Total Biomass (kg)",
    color = "Common Name"
  ) +
  theme_bw(base_size = 13) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

Jurel_Biomass_plot

# Save the plot as PNG
# ggsave(filename = "/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/PrelimLook_CONAPESCA_SpeciesNames_Oct302025/PrelimLook_SpBiomassPlots_Nov2025/Jack.Biomass_plot_Nov062025.png",
#        plot = Jurel_Biomass_plot,
#        width = 10,        # width in inches
#        height = 6,        # height in inches
#        dpi = 300)         # resolution
```

```{r MOJARRA}
##### MOJARRA
Mojarra_Landings<- Landings_GoC_TopGroupings %>% filter(main_species_group=="MOJARRA")
# 324205 obs.

Mojarra_Landings<- left_join(Mojarra_Landings, TopLanded_Sp_List, 
          by = c("main_species_group", "species_name"="recorded_species_name"))
unique(Mojarra_Landings$species_estimated)

Mojarra_Landings_summary<- Mojarra_Landings %>%
  filter(year > 2014 & year < 2025) %>%
  select(year, state, family_estimated, live_weight) %>%
  group_by(year, state, family_estimated) %>%
  summarize(TotalBiomass_annual= sum(live_weight))

# Let's plot what this looks like:
Mojarra_Biomass_plot<- ggplot(Mojarra_Landings_summary,
       aes(x = year,
           y = TotalBiomass_annual,        # match your new column name
           color = family_estimated,
           group = family_estimated)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  facet_wrap(~ state, scales = "free_y") +
  labs(
    title = "'Mojarra' Total Biomass (2015–2024)",
    x = "Year",
    y = "Total Biomass (kg)",
    color = "Common Name"
  ) +
  theme_bw(base_size = 13) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

Mojarra_Biomass_plot

# Save the plot as PNG
# ggsave(filename = "/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/PrelimLook_CONAPESCA_SpeciesNames_Oct302025/PrelimLook_SpBiomassPlots_Nov2025/Mojarra.Biomass_plot_Nov062025.png",
#        plot = Mojarra_Biomass_plot,
#        width = 10,        # width in inches
#        height = 6,        # height in inches
#        dpi = 300)         # resolution
```

```{r JUREL}
##### JUREL
Jurel_Landings<- Landings_GoC_TopGroupings %>% filter(main_species_group=="JUREL")
# 43338 obs.

Jurel_Landings<- left_join(Jurel_Landings, TopLanded_Sp_List, 
          by = c("main_species_group", "species_name"="recorded_species_name"))

Jurel_Landings_summary<- Jurel_Landings %>%
  filter(year > 2014 & year < 2025) %>%
  select(year, state, family_estimated, live_weight) %>%
  group_by(year, state, family_estimated) %>%
  summarize(TotalBiomass_annual= sum(live_weight))

# Let's plot what this looks like:
Jurel_Biomass_plot<- ggplot(Jurel_Landings_summary,
       aes(x = year,
           y = TotalBiomass_annual,        # match your new column name
           color = family_estimated,
           group = family_estimated)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  facet_wrap(~ state, scales = "free_y") +
  labs(
    title = "Jurel Total Biomass (2015–2024)",
    x = "Year",
    y = "Total Biomass (kg)",
    color = "Common Name"
  ) +
  theme_bw(base_size = 13) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

Jurel_Biomass_plot

# Save the plot as PNG
# ggsave(filename = "/home/ferrer/ohi_goc_prep/FP/EMF_CONAPESCA_Exploration/PrelimLook_CONAPESCA_SpeciesNames_Oct302025/PrelimLook_SpBiomassPlots_Nov2025/Jack.Biomass_plot_Nov052025.png",
#        plot = Jurel_Biomass_plot,
#        width = 10,        # width in inches
#        height = 6,        # height in inches
#        dpi = 300)         # resolution
```

# ------------------------------