---
title: "Gulf of California fishing effort by gear and taxa"
output: html_document
date: "`r format(Sys.Date())`"
---

Summarize **VMS fishing effort** (vessel-hours) for the six GoC states.

**Data**

- Vessel registry: `vessel.csv`
- Yearly VMS pings: `vms_{year}.csv`
- Catch by economic unit and taxonomic group: `CatchPerEconomicUnit_group_summary_Jun102026.csv`

**Effort unit:** sum of `hours` per ping, capped at **3 h** per ping.

```{r setup}
library(tidyverse)
library(here)
library(patchwork)

GOC_STATES <- c(
  "BAJA CALIFORNIA",
  "BAJA CALIFORNIA SUR",
  "JALISCO",
  "NAYARIT",
  "SINALOA",
  "SONORA"
)

VMS_ROOT <- "/home/shares/ohi/OHI_GOC/_raw_data/vms_juan_carlos"
CATCH_FILE <- here(
  "FP/FP_Scores/FIS_Scores/FIS_CPUE_SubScores/CPUE_In_Progress",
  "CatchPerEconomicUnit_group_summary_Jun102026.csv"
)
FIGURES_DIR <- here("_pressures/vessels/figures")
EFFORT_OUT <- here("_pressures/vessels/data/fishing_effort_by_major_taxonomic_group.csv")

CATCH_YEAR_MIN <- 2018L
HOURS_CAP_PER_PING <- 3
FIGURE_DPI <- 300

MAIN_TAXONOMIC_GROUPS <- c(
  "Anchovy & Sardine (small pelagics)",
  "Demersal marine fishes",
  "Pelagic marine fish (excl. anchovy & sardine)",
  "Sharks",
  "Shrimps, prawns"
)
```

## Helper functions

```{r helpers}
# Turn a group name into a safe file name.
figure_slug <- function(text) {
  str_replace_all(tolower(text), "[^a-z0-9]+", "_")
}

# Read one year of VMS pings and return effort by economic unit.
load_effort_by_economic_unit <- function(year, vessel_registry) {
  vms_file <- file.path(VMS_ROOT, "vms", sprintf("vms_%s.csv", year))

  read_csv(vms_file, show_col_types = FALSE) %>%
    select(-name, -port, -economic_unit) %>%
    left_join(vessel_registry, by = "vessel_rnpa") %>%
    filter(!is.na(hours), hours > 0, !is.na(eu_rnpa)) %>%
    mutate(hours_capped = pmin(hours, HOURS_CAP_PER_PING)) %>%
    group_by(year, eu_rnpa) %>%
    summarize(
      hours_raw = sum(hours),
      hours_capped = sum(hours_capped),
      .groups = "drop"
    )
}

# Keep the five main groups; put everything else in "Other".
collapse_to_main_taxonomic_groups <- function(group_name) {
  if_else(group_name %in% MAIN_TAXONOMIC_GROUPS, group_name, "Other")
}

# Shared styling for line plots with years on the x-axis.
theme_effort_plot <- function() {
  theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      plot.subtitle = element_text(color = "grey35", size = 10),
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
    )
}
```

## Load vessel registry

```{r load_vessel_registry}
vessel_registry <- read_csv(
  file.path(VMS_ROOT, "vessel.csv"),
  show_col_types = FALSE
) %>%
  filter(state %in% GOC_STATES)
```

## Load VMS effort by year and economic unit

One row per `year` × `eu_rnpa`. This table is used to decide whether an economic
unit has VMS effort in a given year.

```{r load_effort_by_year}
vms_years_on_disk <- list.files(
  file.path(VMS_ROOT, "vms"),
  pattern = "^vms_[0-9]{4}\\.csv$"
) %>%
  str_extract("[0-9]{4}") %>%
  as.integer() %>%
  sort()

years_to_load <- vms_years_on_disk[vms_years_on_disk >= CATCH_YEAR_MIN]

effort_by_eu_year <- map_dfr(years_to_load, function(year) {
  message("Loading VMS effort for ", year)
  load_effort_by_economic_unit(year, vessel_registry)
})

effort_2024_by_eu <- effort_by_eu_year %>%
  filter(year == 2024)

message(
  "Economic units with VMS effort: ",
  n_distinct(effort_by_eu_year$eu_rnpa), " across all years; ",
  nrow(effort_2024_by_eu), " in 2024"
)
```

## Load catch and label VMS vs non-VMS (per year)

An economic unit is **VMS-linked** only in years where it appears in
`effort_by_eu_year`.

```{r load_catch_and_label_vms}
catch_by_group <- read_csv(CATCH_FILE, show_col_types = FALSE)

economic_units_with_vms <- effort_by_eu_year %>%
  distinct(year, eu_rnpa) %>%
  mutate(has_vms_effort = TRUE)

catch_by_group <- catch_by_group %>%
  left_join(economic_units_with_vms, by = c("year", "eu_rnpa")) %>%
  mutate(
    has_vms_effort = !is.na(has_vms_effort),
    vms_record = if_else(has_vms_effort, "yes", "no")
  ) %>%
  select(-has_vms_effort) %>%
  group_by(eu_rnpa, year) %>%
  mutate(total_eu_catch_kg = sum(total_catch_kg)) %>%
  ungroup()

# Quick check: catch economic units missing from 2024 VMS effort
catch_eu_missing_from_vms_2024 <- setdiff(
  catch_by_group %>% filter(year == 2024) %>% pull(eu_rnpa),
  effort_2024_by_eu %>% pull(eu_rnpa)
)

message(
  "Catch economic units in 2024 without VMS match: ",
  length(catch_eu_missing_from_vms_2024)
)
```

## VMS vs non-VMS fleet structure

```{r vms_vs_nonvms_summary}
fleet_summary_by_year <- catch_by_group %>%
  group_by(year, vms_record) %>%
  summarize(
    n_economic_units = n_distinct(eu_rnpa),
    total_catch_kg = sum(total_catch_kg),
    .groups = "drop"
  ) %>%
  group_by(year) %>%
  mutate(
    share_of_units = n_economic_units / sum(n_economic_units),
    share_of_catch = total_catch_kg / sum(total_catch_kg)
  ) %>%
  ungroup()

summary(lm(data=filter(fleet_summary_by_year, vms_record=="no"), share_of_units~year))

summary(lm(data=filter(fleet_summary_by_year, vms_record=="yes"), n_economic_units~year))
summary(lm(data=filter(fleet_summary_by_year, vms_record=="no"), n_economic_units~year))
summary(lm(data=filter(fleet_summary_by_year, vms_record=="yes"), total_catch_kg~year))
summary(lm(data=filter(fleet_summary_by_year, vms_record=="no"), total_catch_kg~year))

ggplot(fleet_summary_by_year, aes(year, share_of_units, color = vms_record)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_color_manual(
    values = c("yes" = "#2c7fb8", "no" = "#d95f0e"),
    name = "VMS-linked"
  ) +
  labs(
    title = "Share of economic units with vs without VMS effort",
    x = "Year",
    y = "Share of economic units"
  ) +
  theme_effort_plot()

ggplot(fleet_summary_by_year, aes(year, share_of_catch, color = vms_record)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_color_manual(
    values = c("yes" = "#2c7fb8", "no" = "#d95f0e"),
    name = "VMS-linked"
  ) +
  labs(
    title = "Share of catch from VMS-linked vs non-VMS economic units",
    x = "Year",
    y = "Share of catch"
  ) +
  theme_effort_plot()

fleet_summary_by_year
```

## Non-VMS sector trends

```{r nonvms_trends}
non_vms_summary_by_year <- catch_by_group %>%
  filter(vms_record == "no") %>%
  group_by(year) %>%
  summarize(
    n_economic_units = n_distinct(eu_rnpa),
    total_catch_kg = sum(total_catch_kg),
    .groups = "drop"
  ) %>%
  mutate(catch_kg_per_eu = total_catch_kg / n_economic_units)

ggplot(non_vms_summary_by_year, aes(year, catch_kg_per_eu)) +
  geom_line(linewidth = 0.9, color = "#d95f0e") +
  geom_point(size = 2.5, color = "#d95f0e") +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(
    title = "Catch per non-VMS economic unit",
    x = "Year",
    y = "Catch (kg per economic unit)"
  ) +
  theme_effort_plot()

ggplot(non_vms_summary_by_year, aes(year, n_economic_units)) +
  geom_line(linewidth = 0.9, color = "#d95f0e") +
  geom_point(size = 2.5, color = "#d95f0e") +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(
    title = "Number of non-VMS economic units",
    x = "Year",
    y = "Economic units"
  ) +
  theme_effort_plot()

summary(lm(total_catch_kg ~ year, data = non_vms_summary_by_year))
summary(lm(n_economic_units ~ year, data = non_vms_summary_by_year))
summary(lm(catch_kg_per_eu ~ year, data = non_vms_summary_by_year))

non_vms_summary_by_year
```

Adding in taxonmic grouping
(note might need to weight the unit by proportion of catch)
```{r}

non_vms_summary_by_year <- catch_by_group %>%
  filter(vms_record == "no") %>%
  group_by(year, Major_Functional_Group) %>%
  summarize(
    n_economic_units = n_distinct(eu_rnpa),
    total_catch_kg = sum(total_catch_kg),
    .groups = "drop"
  ) %>%
  mutate(catch_kg_per_eu = total_catch_kg / n_economic_units)


```


## Fishing effort by major taxonomic group with VMS data

For each economic unit with both catch and VMS effort:

1. Split VMS hours across taxonomic groups in proportion to catch.
2. Sum by year and collapsed taxonomic group.
3. Compute catch per hour.

```{r effort_by_major_taxonomic_group}
summarize_one_year_by_taxonomic_group <- function(year) {
  vms_effort <- effort_by_eu_year %>%
    filter(year == {{ year }}) %>%
    transmute(year, eu_rnpa, hours = hours_capped)

  catch_this_year <- catch_by_group %>%
    filter(year == {{ year }})

  if (nrow(vms_effort) == 0 || nrow(catch_this_year) == 0) {
    return(NULL)
  }

  vms_effort %>%
    inner_join(catch_this_year, by = c("year", "eu_rnpa")) %>%
    mutate(
      taxonomic_group = collapse_to_main_taxonomic_groups(Major_Functional_Group),
      catch_share = total_catch_kg / total_eu_catch_kg,
      hours_for_group = hours * catch_share
    ) %>%
    group_by(year, taxonomic_group) %>%
    summarize(
      hours_effort = sum(hours_for_group, na.rm = TRUE),
      catch_kg = sum(total_catch_kg, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      catch_kg_per_hour = if_else(hours_effort > 0, catch_kg / hours_effort, NA_real_)
    )
}

years_with_catch_and_vms <- sort(
  intersect(unique(catch_by_group$year), unique(effort_by_eu_year$year))
)

message(
  "Summarizing taxonomic groups for years: ",
  paste(years_with_catch_and_vms, collapse = ", ")
)

effort_by_taxonomic_group <- map_dfr(
  years_with_catch_and_vms,
  summarize_one_year_by_taxonomic_group
) %>%
  rename(major_taxonomic_group = taxonomic_group) %>%
  arrange(year, major_taxonomic_group)

dir.create(dirname(EFFORT_OUT), recursive = TRUE, showWarnings = FALSE)
write_csv(effort_by_taxonomic_group, EFFORT_OUT)

message("Wrote ", nrow(effort_by_taxonomic_group), " rows → ", EFFORT_OUT)

effort_by_taxonomic_group
```

## Catch per hour by taxonomic group

```{r plot_catch_per_hour, fig.height=4, fig.width=7}
dir.create(FIGURES_DIR, showWarnings = FALSE, recursive = TRUE)

all_taxonomic_groups <- effort_by_taxonomic_group %>%
  distinct(major_taxonomic_group) %>%
  arrange(major_taxonomic_group) %>%
  pull(major_taxonomic_group)

for (group_name in all_taxonomic_groups) {
  group_data <- effort_by_taxonomic_group %>%
    filter(major_taxonomic_group == group_name)

  catch_per_hour_plot <- ggplot(group_data, aes(year, catch_kg_per_hour)) +
    geom_line(linewidth = 0.9, color = "steelblue") +
    geom_point(size = 2.5, color = "steelblue") +
    scale_x_continuous(breaks = scales::breaks_width(1)) +
    scale_y_continuous(labels = scales::label_comma()) +
    labs(
      title = group_name,
      subtitle = "Catch per VMS-allocated hour (kg/h)",
      x = "Year",
      y = "Catch (kg per hour)"
    ) +
    theme_effort_plot()

  figure_path <- file.path(
    FIGURES_DIR,
    paste0("catch_kg_per_hour_", figure_slug(group_name), ".png")
  )

  ggsave(figure_path, catch_per_hour_plot, width = 7, height = 4, dpi = FIGURE_DPI, bg = "white")
  message("Wrote figure → ", figure_path)
  print(catch_per_hour_plot)
}
```

## Catch and effort over time (five main groups)

```{r plot_catch_and_effort, fig.height=8, fig.width=10}
effort_main_groups <- effort_by_taxonomic_group %>%
  filter(major_taxonomic_group %in% MAIN_TAXONOMIC_GROUPS) %>%
  mutate(
    major_taxonomic_group = factor(major_taxonomic_group, levels = MAIN_TAXONOMIC_GROUPS),
    catch_tonnes = catch_kg / 1000
  )

for (group_name in MAIN_TAXONOMIC_GROUPS) {
  group_data <- effort_main_groups %>%
    filter(major_taxonomic_group == group_name)

  catch_plot <- ggplot(group_data, aes(year, catch_tonnes)) +
    geom_line(linewidth = 0.9, color = "#2c7fb8") +
    geom_point(size = 2.5, color = "#2c7fb8") +
    scale_x_continuous(breaks = scales::breaks_width(1)) +
    scale_y_continuous(labels = scales::label_comma()) +
    labs(title = group_name, x = NULL, y = "Catch (tonnes)") +
    theme_effort_plot()

  effort_plot <- ggplot(group_data, aes(year, hours_effort)) +
    geom_line(linewidth = 0.9, color = "#d95f0e") +
    geom_point(size = 2.5, color = "#d95f0e") +
    scale_x_continuous(breaks = scales::breaks_width(1)) +
    scale_y_continuous(labels = scales::label_comma()) +
    labs(x = "Year", y = "VMS effort (hours)") +
    theme_effort_plot()

  combined_plot <- catch_plot / effort_plot

  figure_path <- file.path(
    FIGURES_DIR,
    paste0("catch_and_effort_", figure_slug(group_name), ".png")
  )

  ggsave(figure_path, combined_plot, width = 7, height = 5, dpi = FIGURE_DPI, bg = "white")
  message("Wrote figure → ", figure_path)
  print(combined_plot)
}
```

```{r plot_catch_and_effort_overview, fig.height=6, fig.width=12}
effort_main_long <- effort_main_groups %>%
  select(year, major_taxonomic_group, catch_tonnes, hours_effort) %>%
  pivot_longer(
    cols = c(catch_tonnes, hours_effort),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(
    metric = recode(
      metric,
      catch_tonnes = "Catch (tonnes)",
      hours_effort = "VMS effort (hours)"
    )
  )

overview_plot <- ggplot(effort_main_long, aes(year, value, group = 1)) +
  geom_line(linewidth = 0.85) +
  geom_point(size = 2) +
  facet_grid(metric ~ major_taxonomic_group, scales = "free_y") +
  scale_x_continuous(breaks = scales::breaks_width(1)) +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(
    title = "Catch and VMS effort by major taxonomic group",
    subtitle = "Five main groups (2018–2024)",
    x = "Year",
    y = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8),
    strip.text = element_text(size = 8)
  )

overview_path <- file.path(FIGURES_DIR, "catch_and_effort_main_groups_overview.png")
ggsave(overview_path, overview_plot, width = 12, height = 6, dpi = FIGURE_DPI, bg = "white")
message("Wrote figure → ", overview_path)

overview_plot
```
