---
title: "Gulf of California fishing effort by gear and taxa (2024 VMS)"
output: html_document
date: "`r format(Sys.Date())`"
---

Summarize **2024 VMS fishing effort** (vessel-hours) for the six GoC states, grouped by
permit **gear type** and **target taxa** from the vessel registry.

**Data:** `vessel.csv` + `vms_2024.csv` (same paths as [`vessel_pressure.Rmd`](vessel_pressure.Rmd)).

**Effort unit:** sum of `hours` per ping, with each ping capped at **3 h** (limits long gaps
between VMS reports).

**Two views of taxa:**

1. **Permit combo** — full `target_species` string (e.g. `CAMARÓN | ESCAMA`).
2. **Taxa token** — split on `|` (e.g. `CAMARÓN`, `ESCAMA`); hours are **re-counted** for
   each token on multi-target permits (use for “which taxa appear on active permits,” not
   for mutually exclusive totals).

```{r setup}
library(tidyverse)
library(here)

GOC_STATES <- c(
  "BAJA CALIFORNIA",
  "BAJA CALIFORNIA SUR",
  "JALISCO",
  "NAYARIT",
  "SINALOA",
  "SONORA"
)
```

## Load registry and VMS

```{r load_data}
vessels <- read_csv(
  "/home/shares/ohi/OHI_GOC/_raw_data/vms_juan_carlos/vessel.csv",
  show_col_types = FALSE
) %>%
  filter(state %in% GOC_STATES)

vms <- read_csv(
  "/home/shares/ohi/OHI_GOC/_raw_data/vms_juan_carlos/vms/vms_2024.csv",
  show_col_types = FALSE
) %>%
  select(-name, -port, -economic_unit)

vessel_effort <- vms %>%
  left_join(vessels, by = "vessel_rnpa", suffix = c("_vms", "_registry"))

message("Registry vessels (GoC): ", nrow(vessels))
message("2024 VMS pings: ", format(nrow(vms), big.mark = ","))
message("Joined pings: ", format(nrow(vessel_effort), big.mark = ","))
```

## Prepare effort (capped hours)
## note: I thought about excluding fast moving vessels (because they aren't fishing), but 
## then decided that is "effort" (i.e., moving around looking for a good place to fish)
```{r prepare_effort}
effort <- vessel_effort %>%
  filter(!is.na(hours), hours > 0) %>%
  mutate(hours_capped = pmin(hours, 3)) 

tibble(
  metric = c("Pings with hours > 0", "Total vessel-hours (raw)", "Total vessel-hours (capped at 3 h)"),
  value = c(
    nrow(effort),
    sum(effort$hours),
    sum(effort$hours_capped)
  )
)
```

## Effort by gear type

Permit-level `gear_type` (may include `|` for multi-gear permits).

```{r effort_by_gear}
effort_by_gear <- effort %>%
  mutate(gear_type = coalesce(gear_type, "(missing)")) %>%
  group_by(gear_type) %>%
  summarise(
    n_pings = n(),
    n_vessels = n_distinct(vessel_rnpa),
    hours_raw = sum(hours),
    hours_capped = sum(hours_capped),
    .groups = "drop"
  ) %>%
  mutate(pct_hours = round(100 * hours_capped / sum(hours_capped), 1)) %>%
  arrange(desc(hours_capped))

effort_by_gear
```

```{r plot_gear, fig.height=6}
effort_by_gear %>%
  slice_head(n = 15) %>%
  mutate(gear_type = fct_reorder(gear_type, hours_capped)) %>%
  ggplot(aes(x = hours_capped, y = gear_type)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "2024 VMS effort by permit gear type",
    x = "Vessel-hours (capped at 3 h per ping)",
    y = NULL
  ) +
  theme_minimal()
```

## Effort by target species (permit taxa combo)

Full `target_species` string as registered (combinations separated by `|`).

```{r effort_by_target_combo}
effort_by_target <- effort %>%
  mutate(target_species = coalesce(target_species, "(missing)")) %>%
  group_by(target_species) %>%
  summarise(
    n_pings = n(),
    n_vessels = n_distinct(vessel_rnpa),
    hours_capped = sum(hours_capped),
    .groups = "drop"
  ) %>%
  mutate(pct_hours = round(100 * hours_capped / sum(hours_capped), 1)) %>%
  arrange(desc(hours_capped))

effort_by_target %>%
  print(n = 45)
```

## Effort by gear × target species

Primary summary table for gear and taxa together.

```{r effort_by_gear_target}
effort_by_gear_target <- effort %>%
  mutate(
    gear_type = coalesce(gear_type, "(missing)"),
    target_species = coalesce(target_species, "(missing)")
  ) %>%
  group_by(gear_type, target_species) %>%
  summarise(
    n_pings = n(),
    n_vessels = n_distinct(vessel_rnpa),
    hours_capped = sum(hours_capped),
    .groups = "drop"
  ) %>%
  group_by(gear_type) %>%
  mutate(pct_within_gear = round(100 * hours_capped / sum(hours_capped), 1)) %>%
  ungroup() %>%
  mutate(pct_total = round(100 * hours_capped / sum(effort$hours_capped), 1)) %>%
  arrange(gear_type, desc(hours_capped))

effort_by_gear_target %>%
  slice_head(n = 20)
```

```{r plot_gear_target, fig.height=7}
effort_by_gear_target %>%
  filter(hours_capped >= 1000) %>%
  mutate(
    label = str_trunc(paste(gear_type, target_species, sep = " · "), 50),
    label = fct_reorder(label, hours_capped)
  ) %>%
  ggplot(aes(x = hours_capped, y = label)) +
  geom_col(fill = "darkolivegreen4") +
  labs(
    title = "2024 VMS effort: gear × target species (≥ 1,000 vessel-hours)",
    x = "Vessel-hours (capped at 3 h per ping)",
    y = NULL
  ) +
  theme_minimal()
```

## Effort by taxa token (split permit targets)

Each `target_species` entry is split on `|`; multi-target permits contribute the **full**
ping hours to **each** listed taxon.

```{r taxa_token_helper}
## Split permit target_species on "|" and map to coarse taxon labels.
split_target_taxa <- function(target_species) {
  if (length(target_species) != 1L || is.na(target_species) || !nzchar(target_species)) {
    return("(missing)")
  }
  tokens <- str_split(target_species, "\\s*\\|\\s*")[[1]] %>%
    str_trim() %>%
    unique()
  case_when(
    str_detect(tokens, "SARDINA Y ANCHOVETA") ~ "SARDINA Y ANCHOVETA",
    str_detect(tokens, "CAMARÓN") ~ "CAMARÓN",
    str_detect(tokens, "ESCAMA") ~ "ESCAMA",
    str_detect(tokens, "ATÚN") ~ "ATÚN",
    str_detect(tokens, "TIBURÓN") ~ "TIBURÓN",
    str_detect(tokens, "OTRAS") ~ "OTRAS",
    TRUE ~ tokens
  )
}
```

```{r effort_by_taxa_token}
effort_by_taxa <- effort %>%
  mutate(taxon = map(target_species, split_target_taxa)) %>%
  unnest(taxon) %>%
  group_by(taxon) %>%
  summarise(
    n_pings = n(),
    n_vessels = n_distinct(vessel_rnpa),
    hours_capped = sum(hours_capped),
    .groups = "drop"
  ) %>%
  mutate(
    pct_of_capped_total = round(100 * hours_capped / sum(effort$hours_capped), 1),
    note = "Hours double-counted across taxa when permit lists multiple targets"
  ) %>%
  arrange(desc(hours_capped))

effort_by_taxa
```

## Effort by gear × taxa token

```{r effort_by_gear_taxa_token}
effort_by_gear_taxon <- effort %>%
  mutate(
    gear_type = coalesce(gear_type, "(missing)"),
    taxon = map(target_species, split_target_taxa)
  ) %>%
  unnest(taxon) %>%
  group_by(gear_type, taxon) %>%
  summarise(
    n_vessels = n_distinct(vessel_rnpa),
    hours_capped = sum(hours_capped),
    .groups = "drop"
  ) %>%
  arrange(gear_type, desc(hours_capped))

effort_by_gear_taxon %>%
  group_by(gear_type) %>%
  slice_head(n = 5) %>%
  ungroup()
```

## Effort by registry taxa flags

Binary permit fields (`target_shrimp`, `target_tuna`, etc.): hours attributed to each
flag = 1 (a vessel with multiple flags is counted in each).

```{r effort_by_target_flags}
flag_cols <- c(
  target_finfish = "ESCAMA (finfish flag)",
  target_sardine = "Sardine flag",
  target_shrimp = "Shrimp flag",
  target_tuna = "Tuna flag",
  target_shark = "Shark flag",
  target_other = "Other flag"
)

effort_by_flag <- map_dfr(names(flag_cols), function(col) {
  effort %>%
    filter(.data[[col]] == 1) %>%
    summarise(
      taxon_flag = flag_cols[[col]],
      n_pings = n(),
      n_vessels = n_distinct(vessel_rnpa),
      hours_capped = sum(hours_capped)
    )
}) %>%
  mutate(pct_of_capped_total = round(100 * hours_capped / sum(effort$hours_capped), 1)) %>%
  arrange(desc(hours_capped))

effort_by_flag
```

## Save summary tables

```{r write_outputs}
dir.create(here("_pressures/vessels/data"), recursive = TRUE, showWarnings = FALSE)

write_csv(effort_by_gear, here("_pressures/vessels/data/effort_by_gear_2024.csv"))
write_csv(effort_by_target, here("_pressures/vessels/data/effort_by_target_species_2024.csv"))
write_csv(effort_by_gear_target, here("_pressures/vessels/data/effort_by_gear_target_2024.csv"))
write_csv(effort_by_taxa, here("_pressures/vessels/data/effort_by_taxon_2024.csv"))
write_csv(effort_by_gear_taxon, here("_pressures/vessels/data/effort_by_gear_taxon_2024.csv"))
write_csv(effort_by_flag, here("_pressures/vessels/data/effort_by_taxon_flag_2024.csv"))
```
