---
title: "Gilman gear crosswalk — Mexican permits to ghost-gear risk scores"
output: html_document
date: "`r format(Sys.Date())`"
---

## Purpose

Map each Gulf of California vessel permit combination (`gear_type` × `target_species`) to
[Gilman et al. (2021)](https://www.nature.com/articles/s41598-021-86123-3) FAO gear categories
and attach ghost-gear risk scores used in **Pressure 3** (ghost gear) in
`step2_vessel_pressure.Rmd`.

**Prerequisite:** run the gear × target classification in step2 first, which writes
`data/gear_target_benthic_destructive.csv`.

**Outputs:**

| File | Description |
|------|-------------|
| `data/gear_gilman_crosswalk.csv` | Main lookup: permit combo → Gilman scores |
| `data/gilman_table2_reference_updated.csv` | Table 2 + composite categories |
| `data/gilman_table_s2_ghost_fishing_updated.csv` | Supplement S2 + composite ghost scores |

## Mapping logic (three paths)

Every permit combo is assigned a Gilman gear category using one of three methods:

1. **`table2_single`** — one unambiguous gear token (`ARRASTRE`, `CERCO`, `PALANGRE`, `VARA`).
   Look up Table 2 directly. High confidence.

2. **`dominant_guess`** — ambiguous or missing gear (`OTRO`, `PESCA MULTIPLE`, `NA`, or
   multi-token permits without two distinct gears). Infer the Gilman category from
   **target species**. Lower confidence; notes flag uncertain cases.

3. **`composite_mean`** — permit lists **2+ distinct gears** separated by `|`
   (e.g. `ARRASTRE | PALANGRE`). Take the **arithmetic mean** of Table 2 scores (R, C, A, O)
   across mapped component categories. `OTRO` and `PESCA MULTIPLE` tokens are **ignored**
   for compositing (e.g. `ARRASTRE | PALANGRE | PESCA MULTIPLE` → mean of trawl + longline).

**Ghost gear index used in pressure rasters:**

- `H_gear` = R × O (general gear hazard; not used for ghost pressure)
- **`H_gear_ghost`** = R × `ghost_fishing_mean` (Supplement Table S2) — this is what gets
  multiplied by capped VMS hours in the ghost-gear layer

```{r setup}
library(tidyverse)
library(here)

DISTINCT_GEARS <- c("ARRASTRE", "CERCO", "PALANGRE", "VARA")
AMBIGUOUS_GEARS <- c("OTRO", "PESCA MULTIPLE")
```

## Load inputs

```{r load_inputs}
combos <- read_csv(
  here("_pressures/vessels/data/gear_target_benthic_destructive.csv"),
  show_col_types = FALSE
)

gilman_ref <- read_csv(
  here("_pressures/vessels/data/gilman_table2_reference.csv"),
  show_col_types = FALSE
)

ghost_s2_ref <- read_csv(
  here("_pressures/vessels/data/gilman_table_s2_ghost_fishing.csv"),
  show_col_types = FALSE
)

tibble(
  dataset = c("Permit combos", "Gilman Table 2", "Gilman S2 ghost fishing"),
  n_rows = c(nrow(combos), nrow(gilman_ref), nrow(ghost_s2_ref))
)
```

Gilman Table 2 categories in our reference extract:

```{r preview_gilman_ref}
gilman_ref %>%
  select(gilman_gear_category, R, C, A, O) %>%
  arrange(gilman_gear_category)
```

## Helper: parse permit gear strings

Mexican permits can list multiple gears separated by `|`. We split tokens and identify which
are distinct gears vs ambiguous placeholders.

```{r parse_gear_tokens}
parse_gear_tokens <- function(gear_type) {
  gear_chr <- coalesce(gear_type, "")
  if (gear_chr == "") {
    return(tibble(
      gear_tokens = list(character(0)),
      distinct_gears = list(character(0)),
      has_ambiguous_token = TRUE,
      is_multi_gear = FALSE
    ))
  }
  tokens <- str_split(gear_chr, "\\s*\\|\\s*")[[1]] %>%
    str_trim() %>%
    unique()
  distinct <- tokens[tokens %in% DISTINCT_GEARS] %>% unique()
  ignored_ambiguous <- tokens[tokens %in% AMBIGUOUS_GEARS] %>% unique()
  has_ambiguous <- length(ignored_ambiguous) > 0L ||
    any(tokens == "") ||
    is.na(gear_type)
  tibble(
    gear_tokens = list(tokens),
    distinct_gears = list(distinct),
    ignored_ambiguous_gears = list(ignored_ambiguous),
    has_ambiguous_token = has_ambiguous,
    is_multi_gear = length(tokens) > 1L
  )
}
```

Example parses:

```{r parse_examples}
tibble(gear_type = c(
  "ARRASTRE",
  "ARRASTRE | PALANGRE",
  "ARRASTRE | PALANGRE | PESCA MULTIPLE",
  "OTRO",
  NA_character_
)) %>%
  rowwise() %>%
  mutate(parsed = list(parse_gear_tokens(gear_type))) %>%
  unnest_wider(parsed)
```

## Helper: map a single Mexican gear token

When a permit has one clear gear token, map it to a Gilman category using gear + target rules.

```{r map_single_mexican_gear}
map_single_mexican_gear <- function(mex_gear, target_species) {
  target_chr <- coalesce(target_species, "")
  has_shrimp <- str_detect(target_chr, "CAMARÓN")
  has_finfish <- str_detect(target_chr, "ESCAMA")
  has_sardine <- str_detect(target_chr, "SARDINA")
  has_tuna <- str_detect(target_chr, "ATÚN")
  has_shark <- str_detect(target_chr, "TIBURÓN")
  has_tuna_only <- has_tuna & !has_shrimp & !has_finfish & !has_shark & !has_sardine

  case_when(
    mex_gear == "ARRASTRE" & has_sardine ~ "Trawl, midwater",
    mex_gear == "ARRASTRE" & has_tuna_only ~ "Trawl, midwater",
    mex_gear == "ARRASTRE" ~ "Trawl, bottom",
    mex_gear == "CERCO" & has_tuna & !has_sardine ~
      "Purse seine, tuna, including drifting and anchored FADs",
    mex_gear == "CERCO" & has_tuna & has_sardine ~ "Purse seine, non-tuna",
    mex_gear == "CERCO" & has_tuna ~
      "Purse seine, tuna, including drifting and anchored FADs",
    mex_gear == "CERCO" ~ "Purse seine, non-tuna",
    mex_gear == "PALANGRE" & has_tuna ~ "Longline, pelagic",
    mex_gear == "PALANGRE" & has_shark ~ "Longline, demersal",
    mex_gear == "PALANGRE" ~ "Longline, demersal",
    mex_gear == "VARA" ~ "Pole-and-line, including anchored FADs",
    TRUE ~ NA_character_
  )
}
```

## Helper: composite and score averaging

For multi-gear permits, average Table 2 and S2 scores across component Gilman categories.

```{r composite_helpers}
use_composite_mean <- function(gear_type) {
  parsed <- parse_gear_tokens(gear_type)
  length(parsed$distinct_gears[[1]]) >= 2L
}

build_composite_category <- function(component_categories) {
  sorted <- sort(unique(component_categories))
  paste0("Composite: ", paste(sorted, collapse = " + "))
}

mean_gilman_scores <- function(categories, gilman_ref) {
  gilman_ref %>%
    filter(gilman_gear_category %in% categories) %>%
    summarise(
      across(c(R, C, A, O), ~ mean(.x, na.rm = TRUE)),
      .groups = "drop"
    )
}

mean_ghost_scores <- function(categories, ghost_ref) {
  ghost_ref %>%
    filter(gilman_gear_category %in% categories) %>%
    summarise(
      across(
        c(ghost_initial_mortality, ghost_duration_efficiency, ghost_vulnerability_species),
        ~ mean(.x, na.rm = TRUE)
      ),
      .groups = "drop"
    ) %>%
    mutate(
      ghost_fishing_mean = mean(
        c(ghost_initial_mortality, ghost_duration_efficiency, ghost_vulnerability_species),
        na.rm = TRUE
      )
    ) %>%
    mutate(across(everything(), ~ round(.x, 4)))
}
```

## Helper: dominant-category guess (ambiguous permits)

When compositing does not apply, infer the best Gilman category from target species and any
gear hints on the permit.

```{r map_gilman_dominant}
map_gilman_dominant <- function(gear_type, target_species) {
  gear_chr <- coalesce(gear_type, "")
  target_chr <- coalesce(target_species, "")

  has_trawl <- str_detect(gear_chr, "ARRASTRE")
  has_purse <- str_detect(gear_chr, "CERCO")
  has_longline <- str_detect(gear_chr, "PALANGRE")
  has_vara <- str_detect(gear_chr, "VARA")
  has_shrimp <- str_detect(target_chr, "CAMARÓN")
  has_finfish <- str_detect(target_chr, "ESCAMA")
  has_sardine <- str_detect(target_chr, "SARDINA")
  has_tuna <- str_detect(target_chr, "ATÚN")
  has_shark <- str_detect(target_chr, "TIBURÓN")
  has_otras <- str_detect(target_chr, "OTRAS")
  has_tuna_only <- has_tuna & !has_shrimp & !has_finfish & !has_shark & !has_sardine
  gear_missing <- is.na(gear_type) | gear_chr == ""
  target_missing <- is.na(target_species) | target_chr == ""
  ambiguous_gear <- gear_missing | gear_chr == "OTRO" | gear_chr == "PESCA MULTIPLE"

  category <- case_when(
    gear_missing & target_missing ~ NA_character_,
    has_sardine & has_purse & !has_trawl ~ "Purse seine, non-tuna",
    has_sardine & !has_trawl & !has_purse & !has_longline ~ "Purse seine, non-tuna",
    has_tuna & has_purse & !has_trawl ~
      "Purse seine, tuna, including drifting and anchored FADs",
    has_vara & !str_detect(gear_chr, "\\|") ~ "Pole-and-line, including anchored FADs",
    has_shrimp & (has_trawl | ambiguous_gear) ~ "Trawl, bottom",
    has_trawl & has_sardine ~ "Trawl, midwater",
    has_trawl & has_tuna_only ~ "Trawl, midwater",
    has_trawl ~ "Trawl, bottom",
    has_longline & has_tuna ~ "Longline, pelagic",
    has_longline & has_shark ~ "Longline, demersal",
    has_longline ~ "Longline, demersal",
    has_purse & has_tuna ~ "Purse seine, tuna, including drifting and anchored FADs",
    has_purse ~ "Purse seine, non-tuna",
    gear_chr == "OTRO" & has_shrimp ~ "Trawl, bottom",
    gear_chr == "OTRO" & has_sardine ~ "Purse seine, non-tuna",
    gear_chr == "OTRO" & has_tuna ~ "Longline, pelagic",
    gear_chr == "OTRO" & has_shark ~ "Longline, demersal",
    gear_chr == "OTRO" & has_finfish ~ "Longline, demersal",
    gear_chr == "OTRO" & has_otras ~
      "Hand dredge, harpoon, spear, lance, tongs, rakes, hand-collected, including diving—shore- and boat-based",
    gear_chr == "PESCA MULTIPLE" & has_shrimp ~ "Trawl, bottom",
    gear_chr == "PESCA MULTIPLE" & has_sardine ~ "Purse seine, non-tuna",
    gear_chr == "PESCA MULTIPLE" & has_tuna ~ "Longline, pelagic",
    gear_chr == "PESCA MULTIPLE" & has_shark ~ "Longline, demersal",
    gear_chr == "PESCA MULTIPLE" & has_finfish ~ "Longline, demersal",
    gear_missing & has_shrimp ~ "Trawl, bottom",
    gear_missing & has_sardine ~ "Purse seine, non-tuna",
    gear_missing & has_tuna ~ "Longline, pelagic",
    gear_missing & has_shark ~ "Longline, demersal",
    gear_missing & has_finfish ~ "Longline, demersal",
    str_detect(gear_chr, "\\|") & has_tuna & has_purse ~
      "Purse seine, tuna, including drifting and anchored FADs",
    str_detect(gear_chr, "\\|") & has_tuna & has_longline ~ "Longline, pelagic",
    str_detect(gear_chr, "\\|") & has_shark & has_longline ~ "Longline, demersal",
    str_detect(gear_chr, "\\|") & has_shrimp & has_longline ~ "Trawl, bottom",
    str_detect(gear_chr, "\\|") & has_vara ~ "Pole-and-line, including anchored FADs",
    str_detect(gear_chr, "\\|") & has_trawl ~ "Trawl, bottom",
    str_detect(gear_chr, "\\|") & has_longline ~ "Longline, demersal",
    str_detect(gear_chr, "\\|") & has_purse ~ "Purse seine, non-tuna",
    TRUE ~ "Hand dredge, harpoon, spear, lance, tongs, rakes, hand-collected, including diving—shore- and boat-based"
  )

  method <- case_when(
    is.na(category) ~ NA_character_,
    ambiguous_gear | gear_missing ~ "dominant_guess",
    gear_chr %in% c("OTRO", "PESCA MULTIPLE") ~ "dominant_guess",
    str_detect(gear_chr, "\\|") ~ "dominant_guess",
    TRUE ~ "table2_single"
  )

  confidence <- case_when(
    is.na(category) ~ NA_character_,
    method == "dominant_guess" & (ambiguous_gear | gear_chr == "OTRO") ~ "low",
    method == "dominant_guess" & str_detect(gear_chr, "\\|") ~ "medium",
    method == "dominant_guess" ~ "medium",
    TRUE ~ "high"
  )

  notes <- case_when(
    is.na(category) ~ "Unclassified permit (gear and target both missing).",
    method == "dominant_guess" &
      str_detect(gear_chr, "OTRO|PESCA MULTIPLE") ~
      "Ambiguous permit gear; dominant Gilman category inferred from target. Gillnets not in registry.",
    method == "dominant_guess" & str_detect(gear_chr, "\\|") ~
      "No pair of distinct gears (ARRASTRE/CERCO/PALANGRE/VARA) for composite; dominant category used.",
    gear_chr == "OTRO" ~
      "Permit gear ambiguous; mapped from target. Artisanal gillnets not in registry.",
    has_tuna & has_purse ~
      "Tuna purse seine mapped with FAD category; FAD use not recorded in registry.",
    has_shrimp & has_longline & !has_trawl ~
      "Shrimp on longline permit uncommon; verify against fishery.",
    has_shrimp & has_purse & !has_trawl ~
      "Shrimp + purse seine; may understate risk if trawl also used.",
    TRUE ~ NA_character_
  )

  tibble(
    gilman_gear_category = category,
    gilman_mapping_method = method,
    gilman_mapping_confidence = confidence,
    gilman_component_categories = NA_character_,
    mapping_notes = notes
  )
}
```

## Helper: one row of the crosswalk

Chooses composite mean vs dominant guess, then attaches Table 2 scores.

```{r map_gilman_row}
map_gilman_row <- function(gear_type, target_species, gilman_ref) {
  parsed <- parse_gear_tokens(gear_type)
  if (use_composite_mean(gear_type)) {
    distinct <- parsed$distinct_gears[[1]]
    ignored <- parsed$ignored_ambiguous_gears[[1]]
    components <- map_chr(distinct, ~ map_single_mexican_gear(.x, target_species))
    composite_name <- build_composite_category(components)
    scores <- mean_gilman_scores(components, gilman_ref) %>%
      mutate(across(c(R, C, A, O), ~ round(.x, 4)))

    ignored_note <- if (length(ignored) > 0) {
      paste0(
        " Ignored on permit: ",
        paste(ignored, collapse = ", "),
        "."
      )
    } else {
      ""
    }

    tibble(
      gilman_gear_category = composite_name,
      gilman_mapping_method = "composite_mean",
      gilman_mapping_confidence = if (length(ignored) > 0) "medium" else "medium",
      gilman_component_categories = paste(sort(components), collapse = "; "),
      mapping_notes = paste0(
        "Composite mean of Gilman Table 2 for permit gears (",
        paste(distinct, collapse = ", "),
        "). R/C/A/O are arithmetic means of component categories.",
        ignored_note
      ),
      R = scores$R,
      C = scores$C,
      A = scores$A,
      O = scores$O
    )
  } else {
    dominant <- map_gilman_dominant(gear_type, target_species)
    scores <- gilman_ref %>%
      filter(gilman_gear_category == dominant$gilman_gear_category) %>%
      select(R, C, A, O) %>%
      mutate(across(everything(), ~ round(.x, 4)))
    if (nrow(scores) == 0) {
      scores <- tibble(R = NA_real_, C = NA_real_, A = NA_real_, O = NA_real_)
    }
    bind_cols(dominant, scores)
  }
}
```

## Build the permit crosswalk

```{r build_crosswalk}
crosswalk <- combos %>%
  mutate(row_id = row_number()) %>%
  rowwise() %>%
  mutate(mapping = list(map_gilman_row(gear_type, target_species, gilman_ref))) %>%
  ungroup() %>%
  unnest_wider(mapping) %>%
  select(-row_id) %>%
  mutate(
    H_gear = round(R * O, 4),
    gilman_reference = "Gilman et al. 2021 Sci Rep s41598-021-86123-3 Table 2"
  ) %>%
  relocate(
    gear_type, target_species, n_vessels,
    gilman_gear_category, gilman_mapping_method, gilman_mapping_confidence,
    gilman_component_categories,
    R, C, A, O, H_gear,
    bottom_destructive, confidence,
    mapping_notes, gilman_reference
  )
```

### Mapping method summary

```{r mapping_summary}
count(crosswalk, gilman_mapping_method, gilman_mapping_confidence, sort = TRUE)
```

### Composite categories

```{r composite_summary}
crosswalk %>%
  filter(gilman_mapping_method == "composite_mean") %>%
  count(gilman_gear_category, gilman_component_categories, sort = TRUE)
```

### Low-confidence mappings (review list)

```{r low_confidence}
crosswalk %>%
  filter(gilman_mapping_confidence == "low" | !is.na(mapping_notes)) %>%
  select(
    gear_type, target_species, n_vessels,
    gilman_gear_category, gilman_mapping_method, gilman_mapping_confidence,
    mapping_notes
  ) %>%
  arrange(desc(n_vessels))
```

## Attach Supplement Table S2 ghost-fishing scores

Build updated reference tables that include composite categories, then join ghost scores
onto the crosswalk and compute **`H_gear_ghost`**.

```{r ghost_scores}
gilman_ref_updated <- gilman_ref %>%
  mutate(
    is_composite = FALSE,
    component_categories = NA_character_,
    source = "Gilman et al. 2021 Table 2"
  )

composite_ref <- crosswalk %>%
  filter(gilman_mapping_method == "composite_mean") %>%
  distinct(gilman_gear_category, R, C, A, O, gilman_component_categories) %>%
  mutate(
    is_composite = TRUE,
    component_categories = gilman_component_categories,
    source = "Composite mean of Table 2 components"
  ) %>%
  select(-gilman_component_categories)

gilman_ref_updated <- bind_rows(gilman_ref_updated, composite_ref) %>%
  arrange(is_composite, gilman_gear_category)

ghost_s2_updated <- ghost_s2_ref %>%
  mutate(
    is_composite = FALSE,
    component_categories = NA_character_
  )

composite_ghost <- crosswalk %>%
  filter(gilman_mapping_method == "composite_mean") %>%
  distinct(gilman_gear_category, gilman_component_categories) %>%
  rowwise() %>%
  mutate(
    components = list(str_split(gilman_component_categories, "; ")[[1]]),
    ghost_scores = list(mean_ghost_scores(components, ghost_s2_ref))
  ) %>%
  ungroup() %>%
  unnest_wider(ghost_scores) %>%
  mutate(
    is_composite = TRUE,
    component_categories = gilman_component_categories,
    source = "Composite mean of Supplement Table S2 (S3.1) components",
    notes = paste(
      "Arithmetic mean of ghost_initial_mortality, ghost_duration_efficiency,",
      "and ghost_vulnerability_species across component categories;",
      "ghost_fishing_mean is mean of those three."
    )
  ) %>%
  select(
    gilman_gear_category,
    ghost_initial_mortality,
    ghost_duration_efficiency,
    ghost_vulnerability_species,
    ghost_fishing_mean,
    source,
    notes,
    is_composite,
    component_categories
  )

ghost_s2_updated <- bind_rows(ghost_s2_updated, composite_ghost) %>%
  arrange(is_composite, gilman_gear_category)

ghost_lookup <- ghost_s2_updated %>%
  select(
    gilman_gear_category,
    ghost_initial_mortality,
    ghost_duration_efficiency,
    ghost_vulnerability_species,
    ghost_fishing_mean,
    gilman_ghost_reference = source
  )

crosswalk <- crosswalk %>%
  left_join(ghost_lookup, by = "gilman_gear_category") %>%
  mutate(
    H_gear_ghost = round(R * ghost_fishing_mean, 4),
    gilman_ghost_reference = coalesce(
      gilman_ghost_reference,
      "Gilman et al. 2021 Supplement Table S2 (S3.1)"
    )
  ) %>%
  relocate(
    gear_type, target_species, n_vessels,
    gilman_gear_category, gilman_mapping_method, gilman_mapping_confidence,
    gilman_component_categories,
    R, C, A, O, H_gear,
    ghost_initial_mortality, ghost_duration_efficiency,
    ghost_vulnerability_species, ghost_fishing_mean, H_gear_ghost,
    bottom_destructive, confidence,
    mapping_notes, gilman_reference, gilman_ghost_reference
  )
```

Distribution of ghost risk by mapping method:

```{r ghost_risk_summary}
crosswalk %>%
  filter(!is.na(H_gear_ghost)) %>%
  group_by(gilman_mapping_method) %>%
  summarise(
    n = n(),
    median_H_gear_ghost = median(H_gear_ghost),
    .groups = "drop"
  )
```

Preview of final crosswalk (highest vessel counts first):

```{r crosswalk_preview}
crosswalk %>%
  arrange(desc(n_vessels)) %>%
  select(
    gear_type, target_species, n_vessels,
    gilman_gear_category, H_gear_ghost,
    gilman_mapping_method, gilman_mapping_confidence
  ) %>%
  head(20)
```

## Write outputs

```{r write_outputs}
write_csv(
  gilman_ref_updated,
  here("_pressures/vessels/data/gilman_table2_reference_updated.csv")
)

write_csv(
  ghost_s2_updated,
  here("_pressures/vessels/data/gilman_table_s2_ghost_fishing_updated.csv")
)

write_csv(
  crosswalk,
  here("_pressures/vessels/data/gear_gilman_crosswalk.csv")
)

message("Wrote ", nrow(gilman_ref_updated), " rows to gilman_table2_reference_updated.csv")
message("  (", sum(gilman_ref_updated$is_composite), " composite categories)")
message("Wrote ", nrow(ghost_s2_updated), " rows to gilman_table_s2_ghost_fishing_updated.csv")
message("  (", sum(ghost_s2_updated$is_composite), " composite ghost-fishing categories)")
message("Wrote ", nrow(crosswalk), " rows to gear_gilman_crosswalk.csv")
```

**Next step:** use `gear_gilman_crosswalk.csv` in `step2_vessel_pressure.Rmd` (Pressure 3:
ghost gear filters and rasters).
