---
title: "Untitled"
output: html_document
date: "2026-06-01"
---

Getting pressure layers for vessel traffic.

```{r}
library(tidyverse)
library(here)
library(terra)
library(sf)
```

## Determine rescaling factor that will be used for all years.
We used the year 2024 for this.
```{r}
vessels <- read_csv("/home/shares/ohi/OHI_GOC/_raw_data/vms_juan_carlos/vessel.csv") %>%
  filter(state %in% c("BAJA CALIFORNIA", "BAJA CALIFORNIA SUR", "JALISCO",
                      "NAYARIT", "SINALOA", "SONORA"))
vms <- read_csv("/home/shares/ohi/OHI_GOC/_raw_data/vms_juan_carlos/vms/vms_2024.csv") %>%
  select(-name, -port, -economic_unit)

vessel_pressure <- left_join(vms, vessels, by=c("vessel_rnpa"))

### only 5% without gear/fishing information, pretty good!
60962 / 1248398

ggplot(vessel_pressure, aes(x = lon, y = lat)) +
  geom_point(alpha = 0.05, size = 0.15, stroke = 0) +
  coord_fixed() +
  labs(
    title = "VMS pings",
    subtitle = "All points in vessel_pressure",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal()

hours_sf <- vessel_pressure %>%
  filter(!is.na(hours), hours > 0) %>%
  mutate(hours = ifel(hours>3, 3, hours)) %>%
  filter(!is.na(lat), !is.na(lon)) %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
  st_transform(crs(ohi_marine))

hours_vect <- vect(hours_sf)

hours_count <- rasterize(
  hours_vect,
  y = ohi_marine,
  field = "hours",
  fun = "sum",
  background = 0
)

names(hours_count) <- "hours_count"

#global(hours_count, "sum", na.rm = TRUE)
#global(hours_count, "mean", na.rm = TRUE)

q99 <- quantile(values(hours_count), 0.99, na.rm = TRUE)
plot(ifel(hours_count > q99, 1, hours_count), main = "VMS hours per 1 km cell (2024)")
```


## Total hours on 1 km OHI marine raster

Sum of VMS hours per 1 km cell, aligned to `ohi_marine_raster_1km.tif` (EPSG:6372). Non-marine cells are NA.

```{r}
ohi_marine <- rast(here("spatial/ohi_regions/ohi_marine_raster_1km.tif"))

for(year in 2000:2025){
#year = 2024
vessels <- read_csv("/home/shares/ohi/OHI_GOC/_raw_data/vms_juan_carlos/vessel.csv") %>%
  filter(state %in% c("BAJA CALIFORNIA", "BAJA CALIFORNIA SUR", "JALISCO",
                      "NAYARIT", "SINALOA", "SONORA"))
  
vms <- read_csv(sprintf("/home/shares/ohi/OHI_GOC/_raw_data/vms_juan_carlos/vms/vms_%s.csv", year)) %>%
  select(-name, -port, -economic_unit)

vessel_pressure_tmp <- left_join(vms, vessels, by=c("vessel_rnpa"))

  hours_sf_tmp <- vessel_pressure_tmp %>%
  filter(!is.na(hours), hours > 0) %>%
  mutate(hours = ifel(hours>3, 3, hours)) %>%
  filter(!is.na(lat), !is.na(lon)) %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
  st_transform(crs(ohi_marine))

hours_vect_tmp <- vect(hours_sf_tmp)

hours_count_tmp <- rasterize(
  hours_vect,
  y = ohi_marine,
  field = "hours",
  fun = "sum",
  background = 0
)

names(hours_count_tmp) <- "hours_count"

hours_count_rescaled <- ifel(hours_count_tmp > q99, q99, hours_count_tmp)

writeRaster(
   hours_count,
   here(sprintf("/home/shares/ohi/OHI_GOC/goal_prep/pressures/vessels/benthic_destructive_fishing/vms_hours_%s_1km.tif", year)),
   overwrite = TRUE
 )

writeRaster(
   hours_count_rescaled,
   here(sprintf("/home/shares/ohi/OHI_GOC/goal_prep/pressures/vessels/benthic_destructive_fishing/vms_hours_%s_1km_rescaled.tif", year)),
   overwrite = TRUE
 )
}
```

# demersal destructive fishing pressure

## Gear type × target species combinations

All unique pairs in the GoC vessel registry (six states), with vessel counts.
`bottom_destructive` flags combinations likely to include **bottom trawling** with benthic
impact (permit-level inference). Rules: **sardine/anchovy** and **tuna-only** targets never
bottom trawl; **shrimp** with ambiguous gear (`OTRO`, missing gear, `PESCA MULTIPLE`) **is**
bottom trawl; `NA` + `NA` is unclassified.

```{r}
gear_target_combos <- vessels %>%
  count(gear_type, target_species, name = "n_vessels", .drop = FALSE) %>%
  mutate(
    gear_chr = coalesce(gear_type, ""),
    target_chr = coalesce(target_species, ""),
    has_trawl = str_detect(gear_chr, "ARRASTRE"),
    has_purse_seine = str_detect(gear_chr, "CERCO"),
    has_longline = str_detect(gear_chr, "PALANGRE"),
    has_pesca_multiple = gear_chr == "PESCA MULTIPLE",
    has_shrimp_target = str_detect(target_chr, "CAMARÓN"),
    has_finfish_target = str_detect(target_chr, "ESCAMA"),
    has_sardine_target = str_detect(target_chr, "SARDINA"),
    has_tuna_target = str_detect(target_chr, "ATÚN"),
    has_shark_target = str_detect(target_chr, "TIBURÓN"),
    has_tuna_only = has_tuna_target & !has_shrimp_target & !has_finfish_target &
      !has_shark_target & !has_sardine_target,
    has_otras_target = str_detect(target_chr, "OTRAS"),
    gear_missing = is.na(gear_type) | gear_chr == "",
    target_missing = is.na(target_species) | target_chr == "",
    ambiguous_gear = gear_missing | gear_chr == "OTRO" | has_pesca_multiple,
    bottom_destructive = case_when(
      gear_missing & target_missing ~ NA,
      ## Sardine/anchovy: never bottom trawl (even if ARRASTRE on permit)
      has_sardine_target ~ FALSE,
      ## Tuna-only target: never bottom trawl (pelagic fishery; even if ARRASTRE on permit)
      has_tuna_only ~ FALSE,
      ## Shrimp + ambiguous gear (OTRO, NA gear, PESCA MULTIPLE): treat as bottom trawl
      has_shrimp_target & ambiguous_gear ~ TRUE,
      ## Trawl on permit (non-sardine targets)
      has_trawl ~ TRUE,
      ## Shrimp with explicit non-trawl gear (e.g. PALANGRE + CAMARÓN)
      has_shrimp_target ~ FALSE,
      ## Pelagic / hook gear without trawl
      !has_trawl & (has_purse_seine | has_longline) ~ FALSE,
      ## Unspecified buckets without shrimp
      !has_trawl & (gear_chr == "OTRO" | has_otras_target) ~ FALSE,
      TRUE ~ FALSE
    ),
    confidence = case_when(
      gear_missing & target_missing ~ NA_character_,
      has_sardine_target ~ "high",
      has_tuna_only ~ "high",
      has_shrimp_target & ambiguous_gear ~ "low",
      has_trawl & has_shrimp_target & !has_longline & !has_pesca_multiple ~ "high",
      has_trawl & has_shrimp_target ~ "medium",
      has_trawl & has_finfish_target & !has_shrimp_target ~ "medium",
      has_trawl ~ "medium",
      has_shrimp_target ~ "high",
      !has_trawl & (has_purse_seine | has_longline) ~ "high",
      !has_trawl & gear_chr == "OTRO" & has_otras_target ~ "high",
      !has_trawl & (has_tuna_target) & !has_shrimp_target ~ "high",
      TRUE ~ "high"
    )
  ) %>%
  select(-gear_chr, -target_chr, -has_trawl, -has_purse_seine, -has_longline,
         -has_pesca_multiple, -has_shrimp_target, -has_finfish_target,
         -has_sardine_target, -has_tuna_target, -has_shark_target, -has_tuna_only,
         -has_otras_target,
         -gear_missing, -target_missing, -ambiguous_gear) %>%
  arrange(desc(n_vessels), gear_type, target_species)

dir.create(here("_pressures/vessels/data"), recursive = TRUE, showWarnings = FALSE)
write_csv(
  gear_target_combos,
  here("_pressures/vessels/data/gear_target_benthic_destructive.csv")
)

gear_target_combos
```


## Filter pings for potential bottom-destructive trawling

Keep VMS pings that plausibly reflect **bottom trawling on the continental shelf**,
using permit metadata (`gear_target_combos`), depth, and speed. Each step below
drops pings that fail that step; the table shows how many rows remain.

**Also applied:** each ping’s `hours` is capped at **3 h** before summing (limits inflation
from long gaps between VMS reports).

**Not applied here (consider later):** port/shore distance, `confidence` weights,
soft-bottom sediment mask, multi-year VMS files.

```{r filter_step_helper}
## Shared by bottom-trawl and ghost-pressure filter chunks.
filter_step <- function(label, n_in, n_out) {
  message(
    sprintf(
      "%s: %s → %s pings (removed %s)",
      label,
      format(n_in, big.mark = ","),
      format(n_out, big.mark = ","),
      format(n_in - n_out, big.mark = ",")
    )
  )
}
```

```{r filter_bottom_destructive_pings}
## Lookup: one row per gear + target combination (from registry, not from VMS pings).
destructive_lookup <- gear_target_combos %>%
  select(gear_type, target_species, bottom_destructive)

## Attach permit-based bottom-trawl flag to every ping.
pings_with_permit <- vessel_pressure %>%
  left_join(destructive_lookup, by = c("gear_type", "target_species"))

n_start <- nrow(pings_with_permit)

## 1) Drop pings with missing gear or target on the vessel permit (NA + NA combos, etc.).
pings_known_permit <- pings_with_permit %>%
  filter(!is.na(gear_type), !is.na(target_species))
filter_step("1. Known gear and target on permit", n_start, nrow(pings_known_permit))

## 2) Keep only permit combinations classified as bottom-destructive (see gear_target_combos chunk).
pings_destructive <- pings_known_permit %>%
  filter(bottom_destructive == TRUE)
filter_step("2. Bottom-destructive permit combo", nrow(pings_known_permit), nrow(pings_destructive))

## 3) Shelf only: depth_m is negative below sea level → water depth = -depth_m.
pings_shelf <- pings_destructive %>%
  mutate(water_depth_m = -depth_m) %>%
  filter(!is.na(water_depth_m), water_depth_m <= 200)
filter_step("3. Water depth ≤ 200 m", nrow(pings_destructive), nrow(pings_shelf))

## 4) Drop fast transit legs (speed between consecutive pings, in knots).
pings_trawling_speed <- pings_shelf %>%
  filter(!is.na(implied_speed_knots), implied_speed_knots <= 6)
filter_step("4. Implied speed ≤ 6 knots", nrow(pings_shelf), nrow(pings_trawling_speed))

## 5) Valid positions and positive time step (for later sum(hours) rasters).
pings_valid <- pings_trawling_speed %>%
  filter(!is.na(lat), !is.na(lon), !is.na(hours), hours > 0)

filter_step("5. Valid lat/lon and hours > 0", nrow(pings_trawling_speed), nrow(pings_valid))

## 6) Cap hours at 3 h so one long gap between pings cannot add >3 vessel-hours to a cell.
n_capped <- sum(pings_valid$hours > 3, na.rm = TRUE)
vessel_pressure_bottom <- pings_valid %>%
  mutate(hours = pmin(hours, 3))

message(
  sprintf(
    "6. Cap hours at 3 h: %s pings had hours > 3 before cap (all %s rows kept)",
    format(n_capped, big.mark = ","),
    format(nrow(vessel_pressure_bottom), big.mark = ",")
  )
)

## Summary table
tibble(
  step = c(
    "All pings (with permit fields joined)",
    "After: gear and target known",
    "After: bottom-destructive permit",
    "After: depth ≤ 200 m",
    "After: speed ≤ 6 kn",
    "After: valid position and hours > 0",
    "Final: hours capped at 3 h per ping"
  ),
  n_pings = c(
    n_start,
    nrow(pings_known_permit),
    nrow(pings_destructive),
    nrow(pings_shelf),
    nrow(pings_trawling_speed),
    nrow(pings_valid),
    nrow(vessel_pressure_bottom)
  ),
  pct_of_start = round(100 * c(
    n_start,
    nrow(pings_known_permit),
    nrow(pings_destructive),
    nrow(pings_shelf),
    nrow(pings_trawling_speed),
    nrow(pings_valid),
    nrow(vessel_pressure_bottom)
  ) / n_start, 1)
)

vessel_pressure_bottom
```

## Bottom-destructive trawl effort raster (sum of hours, 2024)

Sum of capped **`hours`** per 1 km cell for filtered pings (`vessel_pressure_bottom`).
Values are **vessel-hours** on the 2024 VMS track (permit + depth + speed filters).
Aligned to `ohi_marine_raster_1km.tif` (EPSG:6372); non-marine cells are NA.

```{r bottom_trawl_hours_raster}
ohi_marine <- rast(here("spatial/ohi_regions/ohi_marine_raster_1km.tif"))

bottom_sf <- vessel_pressure_bottom %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE) %>%
  st_transform(crs(ohi_marine))

bottom_vect <- vect(bottom_sf)

bottom_trawl_hours <- rasterize(
  bottom_vect,
  y = ohi_marine,
  field = "hours",
  fun = "sum",
  background = 0
)
bottom_trawl_hours <- ifel(is.na(ohi_marine), NA, bottom_trawl_hours)
names(bottom_trawl_hours) <- "bottom_trawl_hours"

## QC: raster total should match sum of ping hours (small diff if points fall outside extent).
hours_in_pings <- sum(vessel_pressure_bottom$hours, na.rm = TRUE)
hours_in_raster <- global(bottom_trawl_hours, "sum", na.rm = TRUE)[1, 1]
message(
  "Sum of hours in pings: ", format(round(hours_in_pings), big.mark = ","),
  " | Sum in raster: ", format(round(hours_in_raster), big.mark = ",")
)

plot(log(bottom_trawl_hours), main = "Bottom-destructive trawl hours per 1 km cell (2024)")

dir.create(here("_pressures/vessels/output"), recursive = TRUE, showWarnings = FALSE)
writeRaster(
  bottom_trawl_hours,
  here("_pressures/vessels/output/bottom_trawl_hours_2024_1km.tif"),
  overwrite = TRUE
)
```

# Ghost gear risk

Attach Gilman et al. (2021) scores to **every VMS ping** via the vessel permit
(`gear_type` + `target_species`). Crosswalk: `data/gear_gilman_crosswalk.csv`
(built from `data/gear_target_benthic_destructive.csv` via
`scripts/build_gilman_crosswalk.R`).

**Per-ping score:** `ghost_gear_risk` = `H_gear_ghost` = **R × ghost_fishing_mean**
(Table 2 ALDFG leakage rate **R** × Supplement Table S2 ghost-fishing mean). This is **not**
`H_gear × ghost_fishing_mean` (that would multiply in Table 2 outcome **O** as well).
`H_gear` = R × O is the general gear hazard index; we use the ghost-specific index only.

Pings whose permit has **both** gear and target missing stay **NA** (unclassified combo in
the crosswalk).

```{r ghost_gilman_lookup}
gilman_by_permit <- read_csv(
  here("_pressures/vessels/data/gear_gilman_crosswalk.csv"),
  show_col_types = FALSE
) %>%
  select(
    gear_type,
    target_species,
    gilman_gear = gilman_gear_category,
    gilman_R = R,
    ghost_fishing_mean,
    H_gear_ghost)

gilman_by_permit
```

```{r join_ghost_gear_risk}
## One row per permit combo → repeated on every ping for that vessel.
ghost_pressure <- vessel_pressure %>%
  left_join(gilman_by_permit, by = c("gear_type", "target_species")) %>%
  mutate(
    ghost_gear_risk = H_gear_ghost
  ) %>%
  select(-H_gear_ghost)

ghost_pressure %>%
  select(
    vessel_rnpa, gear_type, target_species,
    gilman_gear, ghost_gear_risk,
    hours, lat, lon
  ) %>%
  slice_head(n = 5)
```

## Filter pings for ghost gear pressure

Keep VMS pings that plausibly reflect **fishing activity** (not fast transit), with a
known permit and Gilman ghost score. Each step drops pings that fail that step.

**Also applied:** `hours` capped at **3 h**; per-ping contribution =
`ghost_gear_risk × hours_capped` (risk-weighted vessel-hours).

**Not applied here:** bottom-destructive permit filter, depth ≤ 200 m, port distance.

```{r filter_ghost_pressure_pings}
n_start <- nrow(ghost_pressure)


## 1) Drop pings without a Gilman ghost score (e.g. gear and target both missing on permit).
ghost_scored <- ghost_pressure %>%
  filter(!is.na(ghost_gear_risk))
filter_step("1. Gilman ghost score available", nrow(ghost_pressure), nrow(ghost_scored))

## 2) Drop fast transit legs (speed between consecutive pings, in knots).
ghost_fishing_speed <- ghost_scored %>%
  filter(!is.na(implied_speed_knots), implied_speed_knots <= 6)
filter_step("2. Implied speed ≤ 6 knots", nrow(ghost_scored), nrow(ghost_fishing_speed))

## 3) Valid positions and positive time step (for raster sum).
ghost_valid <- ghost_fishing_speed %>%
  filter(!is.na(lat), !is.na(lon), !is.na(hours), hours > 0)
filter_step("3. Valid lat/lon and hours > 0", nrow(ghost_fishing_speed), nrow(ghost_valid))

## 5) Cap hours at 3 h and multiply by permit-level ghost gear risk.
n_capped <- sum(ghost_valid$hours > 3, na.rm = TRUE)
ghost_pressure <- ghost_valid %>%
  mutate(
    hours_capped = pmin(hours, 3),
    ghost_risk_hours = ghost_gear_risk * hours_capped
  )

message(
  sprintf(
    "5. Cap hours at 3 h: %s pings had hours > 3 before cap (all %s rows kept)",
    format(n_capped, big.mark = ","),
    format(nrow(ghost_pressure), big.mark = ",")
  )
)

## Summary table
tibble(
  step = c(
    "All pings (with Gilman scores joined)",
    "After: gear and target known",
    "After: ghost_gear_risk available",
    "After: speed ≤ 6 kn",
    "After: valid position and hours > 0",
    "Final: hours capped; ghost_risk_hours computed"
  ),
  n_pings = c(
    n_start,
    nrow(ghost_pressure),
    nrow(ghost_scored),
    nrow(ghost_fishing_speed),
    nrow(ghost_valid),
    nrow(ghost_pressure)
  ),
  pct_of_start = round(100 * c(
    n_start,
    nrow(ghost_pressure),
    nrow(ghost_scored),
    nrow(ghost_fishing_speed),
    nrow(ghost_valid),
    nrow(ghost_pressure)
  ) / n_start, 1)
)

ghost_pressure
```

## Ghost gear pressure raster (sum of risk-weighted hours, 2024)

Sum of **`ghost_risk_hours`** per 1 km cell for filtered pings (`ghost_pressure`).
Values are **risk × vessel-hours** (Gilman `ghost_gear_risk` × capped VMS hours).
Aligned to `ohi_marine_raster_1km.tif` (EPSG:6372); non-marine cells are NA.

**QC note:** Total over all pings is usually **higher** than the raster sum because VMS
points on land or outside the marine mask are dropped when non-marine cells are set to
NA. Compare **on-marine ping sum** to the raster total (should be close).

```{r ghost_pressure_hours_raster}
ohi_marine <- rast(here("spatial/ohi_regions/ohi_marine_raster_1km.tif"))

ghost_sf <- ghost_pressure %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE) %>%
  st_transform(crs(ohi_marine))

ghost_vect <- vect(ghost_sf)

ghost_risk_hours <- rasterize(
  ghost_vect,
  y = ohi_marine,
  field = "ghost_risk_hours",
  fun = "sum",
  background = 0
)
ghost_risk_hours <- ifel(is.na(ohi_marine), NA, ghost_risk_hours)
names(ghost_risk_hours) <- "ghost_risk_hours"

## QC: ping sum exceeds raster sum when pings fall on land or outside the OHI marine mask.
risk_hours_in_pings <- sum(ghost_pressure$ghost_risk_hours, na.rm = TRUE)
risk_hours_in_raster <- global(ghost_risk_hours, "sum", na.rm = TRUE)[1, 1]

marine_at_ping <- extract(ohi_marine, ghost_vect, ID = FALSE)[[1]]
risk_hours_on_marine <- sum(
  ghost_pressure$ghost_risk_hours[!is.na(marine_at_ping)],
  na.rm = TRUE
)
risk_hours_off_marine <- sum(
  ghost_pressure$ghost_risk_hours[is.na(marine_at_ping)],
  na.rm = TRUE
)
pct_off_marine <- round(100 * risk_hours_off_marine / risk_hours_in_pings, 1)

message(
  "Sum of ghost_risk_hours in pings (all locations): ",
  format(round(risk_hours_in_pings), big.mark = ","),
  " | On OHI marine cells only: ",
  format(round(risk_hours_on_marine), big.mark = ","),
  " | Off marine / outside mask (excluded from raster): ",
  format(round(risk_hours_off_marine), big.mark = ","),
  " (", pct_off_marine, "%)"
)
message(
  "Sum in raster (marine cells only): ",
  format(round(risk_hours_in_raster), big.mark = ","),
  " | Diff vs on-marine ping sum: ",
  format(round(risk_hours_on_marine - risk_hours_in_raster), big.mark = ","),
  " (small values = edge of extent or rasterize rounding)"
)

plot(log(ghost_risk_hours), main = "Ghost gear risk-weighted hours per 1 km cell (2024)")

dir.create(here("_pressures/vessels/output"), recursive = TRUE, showWarnings = FALSE)
writeRaster(
  ghost_risk_hours,
  here("_pressures/vessels/output/ghost_risk_hours_2024_1km.tif"),
  overwrite = TRUE
)
```
