---
title: "Gulf of California: Mexican VMS (BigQuery) — obtain and explore"
date: "Interactive notebook (not knitted)"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

## Explore data and then save the datasets we want!
NOTE: Pings are typically one hour apart. Although there are anomolies.

## Loading packages and setting options
```{r}
## Load packages: bigrquery (Google BigQuery), DBI (database interface), tidyverse (dplyr, etc.).
pacman::p_load(
  bigrquery,
  DBI,
  tidyverse
)

cfg <- list(
  ## Your GCP *project id* (the one with billing linked). Queries bill here.
  billing_project   = "goc-vms",
  ## If NULL: use cached Google login from a prior bq_auth().
  ## If set to "you@gmail.com": bq_auth(email = ...) for that account.
  bq_email          = NULL,
  ## Hard cap (bytes) BigQuery may scan per query; raises an error if a query would exceed it.
  max_bytes_billed  = 10737418240
  ## Table or view name inside dataset mex_vms (see list_tables chunk for names).
  ##vms_table         = "mex_vms_processed_latest",
  ## Calendar year used in the “one-year” examples and the GoC sample pull.
#  explore_year      = 2024L
)


## Tell bigrquery: abort any single query that would bill more than this many bytes scanned.
options(bigrquery.max_bytes_billed = cfg$max_bytes_billed)

## Google identity for BigQuery API. Opens a browser the first time; then uses cache.
if (!is.null(cfg$bq_email) && is.character(cfg$bq_email) && nzchar(cfg$bq_email)) {
  bq_auth(email = cfg$bq_email)
} else {
  bq_auth()
}

## Open a DBI connection to BigQuery:
## - project + dataset: where the *tables* live (public mex-fisheries data).
## - billing: YOUR project id — that account pays for bytes scanned.
## - allowLargeResults: needed for some large result patterns (kept as in mex_fisheries README).
con <- dbConnect(
  bigquery(),
  project = "mex-fisheries",
  dataset = "mex_vms",
  billing = cfg$billing_project,
  use_legacy_sql = FALSE,
  allowLargeResults = TRUE
)

```

## Some functions for revealing costs
```{r bq_helpers, echo=TRUE}
## Helpers live in bq_vms_helpers.R (same folder). Run the setup chunk first (bigrquery loaded).
pacman::p_load(here)
source(here::here("_pressures/vessels/bq_vms_helpers.R"))
```



##  View all tables

```{r list_tables, echo=TRUE}
## INFORMATION_SCHEMA.TABLES lists every table/view in dataset mex_vms (names + type).
## mex_vms_processed_latest is the most current.
sql_tables <- "
SELECT table_name, table_type
FROM `mex-fisheries.mex_vms.INFORMATION_SCHEMA.TABLES`
ORDER BY table_name
"
## Run SQL synchronously; result is a data.frame in R (small).
DBI::dbGetQuery(con, sql_tables) %>% print(n=100)
```

## Explore the variables in each table
```{r}

## Lazy remote table: 
## No data is transferred until you dplyr::collect() or dbGetQuery() runs SQL.
mex_vms <- tbl(con, "mex_vms_processed_latest")
# peak at data:
mex_vms
# To get more information on each variable:
sql_columns <- paste0(
  "
  SELECT column_name, data_type, is_nullable
  FROM `mex-fisheries.mex_vms.INFORMATION_SCHEMA.COLUMNS`
  WHERE table_name = '", "mex_vms_processed_latest", "'
  ORDER BY ordinal_position
  "
)
DBI::dbGetQuery(con, sql_columns) %>% print(n=100)

# get range of years
## MIN/MAX year across the whole VMS object. Can still scan a lot of data — dry-run shows estimate.
sql_years <- paste0(
  "
  SELECT
    MIN(year) AS year_min,
    MAX(year) AS year_max
  FROM ", fully_qualified_vms_table("mex_vms_processed_latest"), "
  "
)
bq_print_dry_run(sql_years, cfg$billing_project, "Year MIN/MAX (full table scan warning)")
## Same SQL again, but this time for real — returns two numbers in R.
DBI::dbGetQuery(con, sql_years)

## How many rows in 2024
sql_count_year <- paste0(
  "
  SELECT year, COUNT(*) AS n_rows
  FROM ", fully_qualified_vms_table("mex_vms_processed_latest"), "
  WHERE year = ", as.integer(2024), "
  GROUP BY year
  "
)
bq_print_dry_run(sql_count_year, cfg$billing_project, paste0("COUNT(*) for year = ", 2024))
DBI::dbGetQuery(con, sql_count_year)

###
mex_vessel <- tbl(con, "vessel_info_latest")
mex_vessel

segment <- tbl(con, "segment_info_v_20240515")
segment

raw <- tbl(con, "mex_vms_v_20240515")
raw

spatial <- tbl(con, "spatial_features_v_20240312")
spatial

```


## Now get the data

### VMS data
Open the channel
```{r}
## Lazy remote table: 
## No data is transferred until you dplyr::collect() or dbGetQuery() runs SQL.
mex_vms <- tbl(con, "mex_vms_processed_latest")

# peak at data:
mex_vms
```

Subset the data, and download a small subset of the data to see what it looks like.

```{r explore_goc_dry_run_then_collect, echo=TRUE}
## Rough bounding box around the Gulf of California (WGS84 lat/lon). Edit if you want tighter/wider.
bbox <- list(lat_min = 20, lat_max = 32, lon_min = -115, lon_max = -105)

year = 2024
## dplyr verbs build a query; still lazy until collect().
## !! forces cfg$explore_year to be a concrete value in SQL (not a column name).
vms_goc <- mex_vms |>
  filter(
    year == !!as.integer(year),
    lat >= bbox$lat_min, lat <= bbox$lat_max,
    lon >= bbox$lon_min, lon <= bbox$lon_max
  )

## Some potential subsetting ideas
## head(1000): LIMIT 1000 in SQL — caps rows returned, NOT necessarily bytes scanned.
 vms_sample <- vms_goc |>
  #dplyr::select(any_of(c("vessel_rnpa", "lat", "lon", "year", "timestamp", "speed"))) |>
  head(1000L)

## Show the exact SQL dbplyr will send — useful to copy into BigQuery console for debugging.
sql_sample <- dbplyr::remote_query(vms_sample)
bq_print_dry_run(sql_sample, cfg$billing_project, "Filtered sample (dbplyr remote_query)")

## THIS LINE runs the query and pulls data into R (uses network + bytes scanned per dry-run).
vms_data <- vms_sample |>
  collect()

```

------------------------------------------------------------------------

## Visualize vessel tracks

Requires **`vms_data`** from the chunk above (local tibble of pings). Tracks connect pings **in time order** within each `vessel_rnpa`.

**Note:** `head(1000)` returns a **mixed sample** of pings (not necessarily full trips). For clearer tracks, pull more rows or filter to fewer vessels before `collect()`.

**Full-year maps:** use the **static** plot below (or density rasters later). **mapview** is only practical for a **small** number of vessels/lines — see `max_vessels_mapview`.

```{r plot_tracks_setup}
pacman::p_load(sf, ggplot2)

## mapview is optional (interactive); skip this chunk if you only want the static map.
if (!requireNamespace("mapview", quietly = TRUE)) {
  message("Install mapview for interactive maps: install.packages(\"mapview\")")
}

## Map extent (same as exploration bbox if you ran that chunk; otherwise from data).
if (!exists("bbox")) {
  bbox <- list(
    lat_min = min(vms_data$lat, na.rm = TRUE),
    lat_max = max(vms_data$lat, na.rm = TRUE),
    lon_min = min(vms_data$lon, na.rm = TRUE),
    lon_max = max(vms_data$lon, na.rm = TRUE)
  )
}

## Time column for ordering pings (processed table uses datetime).
time_col <- intersect(names(vms_data), c("datetime", "timestamp"))[1]
if (length(time_col) == 0) {
  stop("vms_data needs a datetime or timestamp column to order tracks.")
}
```

```{r plot_tracks_build_sf}
## Points as sf (WGS84).
pts_sf <- vms_data |>
  dplyr::filter(!is.na(lat), !is.na(lon)) |>
  dplyr::arrange(vessel_rnpa, .data[[time_col]]) |>
  sf::st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE)

## One line per vessel (needs at least 2 pings).
## dplyr::summarise() on sf + sfg geometry in one step fails; use list cols then st_sfc().
tracks_sf <- pts_sf |>
  sf::st_drop_geometry() |>
  dplyr::group_by(vessel_rnpa) |>
  dplyr::filter(dplyr::n() >= 2L) |>
  dplyr::summarise(
    n_pts = dplyr::n(),
    track_year = dplyr::first(year),
    lon = list(lon),
    lat = list(lat),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    geometry = purrr::map2(
      lon, lat,
      function(x, y) sf::st_linestring(cbind(x, y))
    )
  ) |>
  dplyr::select(-lon, -lat) |>
  sf::st_as_sf(crs = 4326)

message(
  "Tracks: ", nrow(tracks_sf), " vessels with >= 2 pings; ",
  nrow(pts_sf), " points total."
)
```


### Interactive map (mapview) — small subsets only

```{r plot_tracks_mapview, eval=FALSE}
## Set eval=FALSE so knitting does not open a browser; run this chunk interactively.
if (!requireNamespace("mapview", quietly = TRUE)) {
  stop("Install mapview to run this chunk.")
}

## Keep mapview light: cap how many track lines are drawn.
max_vessels_mapview <- 25L
tracks_mv <- tracks_sf |>
  dplyr::slice_head(n = max_vessels_mapview)

pts_mv <- pts_sf |>
  dplyr::filter(vessel_rnpa %in% tracks_mv$vessel_rnpa)

mapview::mapviewOptions(fgb = TRUE)

mapview::mapview(
  tracks_mv,
  color = "steelblue",
  lwd = 2,
  layer.name = paste0("Tracks (first ", max_vessels_mapview, " vessels)")
) +
  mapview::mapview(
    pts_mv,
    color = "darkorange",
    cex = 2,
    layer.name = "Pings"
  )
```

## Get the vms data (all the data):

```{r}
## Lazy remote table: 
## No data is transferred until you dplyr::collect() or dbGetQuery() runs SQL.
mex_vms <- tbl(con, "mex_vms_processed_latest")

# peak at data:
mex_vms

## Rough bounding box around the Gulf of California (WGS84 lat/lon). Edit if you want tighter/wider.
bbox <- list(lat_min = 20, lat_max = 32, lon_min = -115, lon_max = -105)

for (yr in 2000:2025) {
  out_path <- here(
    "/home/shares/ohi/OHI_GOC/_raw_data/vms_juan_carlos/vms",
    sprintf("vms_%s.csv", yr)
  )

  if (file.exists(out_path)) {
    message("Skipping ", yr, " — already saved: ", out_path)
    next
  }

  ## dplyr verbs build a query; still lazy until collect().
  ## !! forces yr to be a concrete value in SQL (not a column name).
  vms_goc <- mex_vms |>
    filter(
      year == !!as.integer(yr),
      lat >= bbox$lat_min, lat <= bbox$lat_max,
      lon >= bbox$lon_min, lon <= bbox$lon_max
    )

  ## Show the exact SQL dbplyr will send — useful to copy into BigQuery console for debugging.
  sql_sample <- dbplyr::remote_query(vms_goc)
  bq_print_dry_run(sql_sample, cfg$billing_project, paste0("GoC year ", yr))

  ## THIS LINE runs the query and pulls data into R (uses network + bytes scanned per dry-run).
  vms_data <- vms_goc |>
    collect()

  write_csv(vms_data, out_path)
  message("Saved ", yr, ": ", nrow(vms_data), " rows -> ", out_path)
}

```

## Get the corresponding vessel data
```{r}

  ## Lazy remote table: 
## No data is transferred until you dplyr::collect() or dbGetQuery() runs SQL.
mex_vessel <- tbl(con, "vessel_info_latest")

# peak at data:
mex_vessel

  out_path <- here(
    "/home/shares/ohi/OHI_GOC/_raw_data/vms_juan_carlos/vessel.csv")


  if (file.exists(out_path)) {
    message("Skipping ", yr, " — already saved: ", out_path)
    next
  }

  ## dplyr verbs build a query; still lazy until collect().
  vessel_goc <- mex_vessel

  ## Show the exact SQL dbplyr will send — useful to copy into BigQuery console for debugging.
  sql_sample <- dbplyr::remote_query(vessel_goc)
  bq_print_dry_run(sql_sample, cfg$billing_project)

  ## THIS LINE runs the query and pulls data into R (uses network + bytes scanned per dry-run).
  vms_data <- vessel_goc |>
    collect()

  write_csv(vms_data, out_path)
  message("Saved :", nrow(vms_data), " rows -> ", out_path)

