################################################################################ 

For questions contact: Erica Ferrer, Ph.D. [emferrer\@ucsd.edu](mailto:emferrer@ucsd.edu){.email} - OR - [ferrer\@nceas.ucsb.edu](mailto:ferrer@nceas.ucsb.edu){.email}

################################################################################ 

What's different about this key from April's key?

~108 lines effectively unchanged (copy-paste from April).
4 lines minor comment/header edits.
23 lines deleted (STEP 15).
~51 lines added (STEP 3b2 + one reference line).
~95 lines rewritten (merge, export, 3c–3d, next steps).
~147 lines in May that don’t match April line-for-line (adds + rewrites + line shift from deletion).


# PRE-WORK ------------------------------

```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
library(fuzzyjoin)
library(sf)
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```

# STEP 1: Load in Mel's lat-long keys from Baja & Sonora, and Sinaloa ------------------------------

```{r Append lat-long data to sites in Baja BCS Sonora and Sinaloa}
# Filter OUT sites from Baja that are on the Pacific side using lat-Long file from Mel's repository:

# Integrated Hem georefcatchlocs (STEP 3) and georeflocskm permit/INEGI georef (STEP 3) - May 19 2026
BajaSonora_LatLongKey<- read_csv("/home/frazier/ohi_goc_prep/FP/Exploratory_Research/FIS_TeamResearch/data/Baja_Sonora_states_key.csv")
colnames(BajaSonora_LatLongKey) <- c("state", "landing_site", "OHI_rgn", "lat", "long", "GoC")

Jalisco_LatLongKey<- read_csv(here("/home/frazier/ohi_goc_prep/FP/Exploratory_Research/FIS_TeamResearch/data/landing_sites_jalisco.csv"))
colnames(Jalisco_LatLongKey) <- c("landing_site", "CONAPESCA_landing_key", "est_habitat", "lat", "long", "site_notes", "OHI_rgn", "state")


# Clean the Baja-Sonora key...
BajaSonora_LatLongKey_clean <- BajaSonora_LatLongKey %>%
  
  # Normalize state names:
  mutate(state = toupper(state)) %>%
  
  # Remove any rows where landing_site is NA (NAs cause matching issues later on):
  filter(!is.na(landing_site)) %>%
  
  # Remove any direct duplicates of landing_site:
  group_by(state, landing_site) %>%
  slice(1) %>%  # keeps the first occurrence per key
  ungroup()


# Look into the rows that were removed in the de duplication of BajaSonora_LatLongKey...
removed_rows <- anti_join(BajaSonora_LatLongKey, BajaSonora_LatLongKey_clean, 
                          by = c("landing_site"))

# Convert lat-long entries to numeric entries, accounting for unique cases where the minus (-) is actually an n dash or m dash:
BajaSonora_LatLongKey_clean <- BajaSonora_LatLongKey_clean %>%
  mutate(
    long = as.numeric(str_replace_all(long, "[−–—]", "-")),
    lat  = as.numeric(str_replace_all(lat, "[−–—]", "-"))
  )

# Clean the Jalisco key...
Jalisco_LatLongKey_clean <- Jalisco_LatLongKey %>%
  # Normalize state names
  mutate(state = toupper(state)) %>%
  
  # Remove any rows where landing_site is NA (NAs cause matching issues later on):
  filter(!is.na(landing_site)) %>%
  
  # Keep only one row per CONAPESCA_landing_key; if duplicate keys exist, keep first
  group_by(CONAPESCA_landing_key) %>%
  slice(1) %>%  # keeps the first occurrence per key
  ungroup() %>%
  
  # Optional: remove any direct duplicates of landing_site if there are rows with NA in the CONAPESCA_landing_key
  group_by(state, landing_site) %>%
  slice(1) %>%  # keeps the first occurrence per key
  ungroup()

# Look into the rows that were removed in the de-duplication of Jalisco_LatLongKey...
removed_rows <- anti_join(Jalisco_LatLongKey, Jalisco_LatLongKey_clean, 
                          by = c("CONAPESCA_landing_key", "landing_site"))

# Convert lat-long entries to numeric entries, accounting for unique cases where the minus (-) is actually an n dash or m dash:
Jalisco_LatLongKey_clean <- Jalisco_LatLongKey_clean %>%
  mutate(
    long = as.numeric(str_replace_all(long, "[−–—]", "-")),
    lat  = as.numeric(str_replace_all(lat, "[−–—]", "-"))
  )

# Add the "GoC" helpful column here based on a latitude cut off for the Nayarit/Jalisco OHI region, region 9 starting at 20.40895:
Jalisco_LatLongKey_clean <- Jalisco_LatLongKey_clean %>%
  mutate(GoC = if_else(lat < 20.40895, "no", NA_character_))
```

```{r Merge Mel's Jalisco Baja and Sonora key into one master key}
Mels_LatLongKey <- bind_rows(BajaSonora_LatLongKey_clean, Jalisco_LatLongKey_clean) %>%
  filter(!is.na(landing_site)) %>%  # remove the NA from this data frame as this will create a many-to-many matching issue later on
  mutate(long = as.numeric(long))
# 1176 obs., 9 vars.

# Remove unused data frames:
rm(BajaSonora_LatLongKey, Jalisco_LatLongKey, removed_rows)

# Take a look
unique(Mels_LatLongKey$state)
unique(Mels_LatLongKey$OHI_rgn)
```

# STEP 2: Integrate Hem's catch-location lat-long key (georeflocskm) ------------------------------
```{r Load and prepare georeflocskm}
load("/home/shares/ohi/OHI_GOC/_raw_data/CONAPESCA/d2025/SharedByHemMorzaria/permit_data/georeflocskm.rda")
Hems_LatLongKey0 <- georeflocskm
#rm(georeflocskm)
```

```{r Flag potential duplicate entries in Hem's key}

colnames(Hems_LatLongKey0) <- c(
  "state", "landing_site", "geometry", "lon_deci", "lat_deci",
  "fishery_office", "fishery_office_ent", "distance", "distance_km"
)

Hems_LatLongKey <- Hems_LatLongKey0 %>%
  mutate(
    state = toupper(stringr::str_squish(state)),
    landing_site = toupper(
      stringi::stri_trans_general(landing_site, id = "Latin-ASCII")
    ),
    landing_site = stringr::str_squish(landing_site)
  ) %>%
  select(state, landing_site, fishery_office, lat_deci, lon_deci, distance_km) %>%
  distinct()


# Annotate Hem's key with alerts about disagreement; and compress the key
# Annotation thresholds:
threshold_alert  <- 2
threshold_low    <- 1
threshold_medium <- 0.5


Hems_LatLongKey_compressed <- Hems_LatLongKey %>%
  select(state, landing_site, fishery_office, lat_deci, lon_deci, distance_km) %>%
  
  group_by(state, landing_site) %>%
  
  summarise(
    
    Hems_lat_range = max(lat_deci, na.rm = TRUE) - min(lat_deci, na.rm = TRUE),
    Hems_lon_range = max(lon_deci, na.rm = TRUE) - min(lon_deci, na.rm = TRUE),
    
    Hems_est_lat = median(lat_deci, na.rm = TRUE),
    Hems_est_lon = median(lon_deci, na.rm = TRUE),
    
    notes_re_HemsKey = case_when(
      is.na(Hems_lat_range) | is.na(Hems_lon_range) ~
        "coordinate data not found in Hem's key",

      Hems_lat_range > threshold_alert | Hems_lon_range > threshold_alert ~
        "alert level 3: multiple state-landing_site matches in Hem's key with lat-long values >2 degrees apart",

      Hems_lat_range > threshold_low | Hems_lon_range > threshold_low ~
        "alert level 2: multiple state-landing_site matches in Hem's key with lat-long values ~1 degree apart",

      Hems_lat_range > threshold_medium | Hems_lon_range > threshold_medium ~
        "alert level 1: multiple state-landing_site matches in Hem's key with lat-long values ~0.5 degree apart",

      TRUE ~ "high coherence in Hem's key regarding state-landing_site location"
    ),

    .groups = "drop"
  )

# rm(Hems_LatLongKey_compressed)

table(Hems_LatLongKey_compressed$notes_re_HemsKey)

# alert level 1: multiple state-landing_site matches in Hem's key with lat-long values ~0.5 degree apart 
#                                                                                                    589 
#   alert level 2: multiple state-landing_site matches in Hem's key with lat-long values ~1 degree apart 
#                                                                                                   1072 
#  alert level 3: multiple state-landing_site matches in Hem's key with lat-long values >2 degrees apart 
#                                                                                                   1107 
#                                      high coherence in Hem's key regarding state-landing_site location 
#                                                                                                  24128  

# SANITY CHECK: which values really don't agree?
TEST<- Hems_LatLongKey_compressed %>%
  filter(notes_re_HemsKey== "alert level 3: multiple state-landing_site matches in Hem's key with lat-long values >2 degrees apart")
# n=1107 - Basically, values that have different offices. I feel good enough about this to proceed though.

# SANITY CHECK: Number of unique state + landing_site combinations
n_unique_sites <- Hems_LatLongKey_compressed %>%
  distinct(state, landing_site) %>%
  nrow()
# 26896 unique state-landing_site combinations.
```


# STEP 4: Create a backbone for a key... ------------------------------

```{r Prepare to merge keys}
dim(Hems_LatLongKey_compressed) # 26896 obs., 7 variables
dim(Mels_LatLongKey) # 1176 obs., 9 variables

# Shave down Mel's key:
Mels_LatLongKey_simple<- Mels_LatLongKey %>%
  select(state, landing_site, lat, long, OHI_rgn, GoC, est_habitat, site_notes)

colnames(Mels_LatLongKey_simple)<- c("state", "landing_site", "lat_Mel", "lon_Mel", "OHI_region_draft", "GoC", "est_habitat", "site_notes_Mel")
# 1176 obs. of 8 variables

table(Mels_LatLongKey_simple$GoC)
# 400 sites in the GoC, 588 sites likely not in the GoC
```

```{r Load the Mexican Pacific}
Mex.Pacific_Landings<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/Mex.Pacific_Landings_2000.present_May042026.rds"))

dim(Mex.Pacific_Landings)
# 6406303 obs., 21 vars.
```

```{r States_of_Interest}
unique(Mex.Pacific_Landings$state)

States_of_Interest<- c("BAJA CALIFORNIA", "BAJA CALIFORNIA SUR",
                       "SONORA", "SINALOA",
                       "NAYARIT", "JALISCO")
```

```{r Create a backbone of a key to merge to...}
# Create a backbone to merge to...
colnames(Mex.Pacific_Landings)

LatLong_merge.skeleton<- Mex.Pacific_Landings %>%
  select(state, landing_site) %>%
  distinct()
# 3060 distinct state-landingsite combos in the Mexican Pacific that I am going to try to fill using information from Hem's key or Mel's key. Privilege information from Hem's key.
```

# Cleaning
```{r  CLEAN NAMES FUNCTION}
clean_name <- function(x) {

  x <- iconv(x, from = "", to = "UTF-8", sub = "")
  x <- stringi::stri_trans_general(x, "Latin-ASCII")

  x %>%
    toupper() %>%
    trimws() %>%
    gsub("\\bDE\\b", "", .) %>%
    gsub("\\s+", " ", .) %>%
    trimws()
}
```

```{r Clean names for landing_site across LatLong_merge.skeleton}
# Clean the landing_site names:
LatLong_merge.skeleton<- LatLong_merge.skeleton %>%
  mutate(landing_site_clean= clean_name(landing_site)) %>%
  relocate(landing_site_clean, .after= landing_site)

Hems_LatLongKey<- Hems_LatLongKey %>%
  mutate(landing_site_clean= clean_name(landing_site),
         office_name_clean= clean_name(fishery_office)) %>%
  relocate(landing_site_clean, .after= landing_site) %>%
  relocate(office_name_clean, .after= fishery_office)

Hems_LatLongKey_compressed<- Hems_LatLongKey_compressed %>%
  mutate(landing_site_clean= clean_name(landing_site)) %>%
  relocate(landing_site_clean, .after= landing_site)

Mels_LatLongKey_simple<- Mels_LatLongKey_simple %>%
  mutate(landing_site_clean= clean_name(landing_site)) %>%
  relocate(landing_site_clean, .after= landing_site)
```

```{r Make sure to compress all keys after cleaning}
# Compress: LatLong_merge.skeleton
colnames(LatLong_merge.skeleton)

LatLong_merge.skeleton_compr<- LatLong_merge.skeleton %>%
  select(state, landing_site_clean) %>%
  group_by(state, landing_site_clean) %>%
  slice(1) %>%  # keeps the first occurrence per key
  ungroup()
# From 3060 obs down to 3038

# Compress: Hems_LatLongKey_compressed
colnames(Hems_LatLongKey_compressed)

Hems_LatLongKey_compr<- Hems_LatLongKey_compressed %>%
  select(state, landing_site_clean, Hems_lat_range, Hems_lon_range, Hems_est_lat, Hems_est_lon, notes_re_HemsKey) %>%
  group_by(state, landing_site_clean) %>%
  slice(1) %>%  # keeps the first occurrence per key
  ungroup()
# From 26896 obs down to 26866

# Compress: Mels_LatLongKey_simple
colnames(Mels_LatLongKey_simple)

Mels_LatLongKey_compr<- Mels_LatLongKey_simple %>%
  select(state, landing_site_clean, lat_Mel, lon_Mel, GoC, est_habitat, site_notes_Mel) %>%
  group_by(state, landing_site_clean) %>%
  slice(1) %>%  # keeps the first occurrence per key
  ungroup()
# From 1176 obs to 1168
```


# STEP 5: Create LatLong_SiteKey ------------------------------

```{r Merge to the backbone}
# LOGIC (May 19 2026):
# Start with the skeleton
# Privilege Hem’s key over Mel’s, unless Hem's key is empty or has high disagreement:

LatLong_SiteKey <- LatLong_merge.skeleton_compr %>%
  
  left_join(Mels_LatLongKey_compr, by = c("state", "landing_site_clean")) %>%
  
  left_join(
    Hems_LatLongKey_compr %>%
      select(state, landing_site_clean, Hems_lon_range, Hems_lat_range, Hems_est_lat, Hems_est_lon, notes_re_HemsKey),
    by = c("state", "landing_site_clean")
  ) %>%
  
  mutate(
    
    # Availability of Hem's key, yes or no?
    has_Hem = !is.na(Hems_est_lat) & !is.na(Hems_est_lon),
    
    # Disagreement in Hem's key?
    high_disagreement = (Hems_lon_range > 2 | Hems_lat_range > 2),

    # lat-lon difference between Hem and Mel's keys (flag high disagreement):
    keys_lat_diff = if_else(has_Hem & !is.na(lat_Mel),
                       abs(Hems_est_lat - lat_Mel), NA_real_),

    keys_lon_diff = if_else(has_Hem & !is.na(lon_Mel),
                       abs(Hems_est_lon - lon_Mel), NA_real_),
    
    # Assign lat-long values:
    # final coordinates: Hem's estimates unless high disagreement, fallback to Mel's estimates
    
    lat_est = case_when(
      has_Hem & is.na(lat_Mel) ~ Hems_est_lat,
      has_Hem & (!is.na(lat_Mel)) & high_disagreement ~ lat_Mel,
      has_Hem & (!is.na(lat_Mel)) ~ Hems_est_lat,
      !has_Hem & !is.na(lat_Mel) ~ lat_Mel,
      TRUE ~ NA_real_
      ),
    
    lon_est = case_when(
      has_Hem & is.na(lon_Mel) ~ Hems_est_lon,
      has_Hem & (!is.na(lon_Mel)) & high_disagreement ~ lon_Mel,
      has_Hem & (!is.na(lon_Mel)) ~ Hems_est_lon,
      !has_Hem & !is.na(lon_Mel) ~ lon_Mel,
      TRUE ~ NA_real_
      ),
    
    # Assign confidence values:
    lat_long_confidence = case_when(
      
      has_Hem & (!is.na(lat_Mel)) & high_disagreement ~
        "medium confidence; Mel's estimate; found in both keys but high disagreement in Hem's",

      has_Hem & (!is.na(lat_Mel)) ~
        "high confidence; Hem's estimate; Hem and Mel's keys agree",
      
      has_Hem & is.na(lat_Mel) ~
        "high confidence; Hem's estimate; found in Hem's key",

      has_Hem & (!is.na(lat_Mel)) &
        (keys_lat_diff > 2 | keys_lon_diff > 2) ~
        "check required; Hem's estimate; Hem and Mel's keys disagree by >2º",

      !has_Hem & !is.na(lat_Mel) ~
        "medium confidence; Mel's estimate; found only in Mel's key",

      TRUE ~ NA_character_
    )
  ) %>%
  relocate(lat_est, lon_est, .after = last_col())

# 3038 obs.

table(LatLong_SiteKey$lat_long_confidence)
#                                  high confidence; Hem's estimate; found in Hem's key 
#                                                                                  351 
#                            high confidence; Hem's estimate; Hem and Mel's keys agree 
#                                                                                  501 
# medium confidence; Mel's estimate; found in both keys but high disagreement in Hem's 
#                                                                                   78 
#                           medium confidence; Mel's estimate; found only in Mel's key 
#                                                                                  571 

table(LatLong_SiteKey$GoC)

# SANITY CHECK: how many lat-long values are missing from the states of interest?
TEST<- LatLong_SiteKey %>%
  filter(state %in% States_of_Interest) %>%
  filter(is.na(lat_est))

# 530 sites in the States_of_Interest are missing lat-long values (down from 538 thanks to name cleaning!)
```

```{r Check to see how many entries are covered by this key alone}
NW.Mex_Landings<- Mex.Pacific_Landings %>%
  filter(state %in% States_of_Interest)

dim(NW.Mex_Landings)
# 4713144 obs.

# Clean the landing_site names in NW.Mex_Landings:
NW.Mex_Landings<- NW.Mex_Landings %>%
  mutate(landing_site_clean= clean_name(landing_site),
         office_name_clean= clean_name(office_name)) %>%
  relocate(landing_site_clean, .after= landing_site) %>%
  relocate(office_name_clean, .after= office_name)

# Merge key based on landing_site_clean names:
NW.Mex_Landings <- NW.Mex_Landings %>%
  left_join(LatLong_SiteKey, by = c("state", "landing_site_clean"))
# 4713144 obs.

# SANITY CHECK: how many records are missing lat-long values?
Landings_missingLatLong<- NW.Mex_Landings %>%
  filter(is.na(lat_est))
# 1/4 of records (n= 1,213,865) are not assigned... 
# May 20: Down from 1213865 to 1205911 (-7954) thanks to name cleaning!
# This doesn't really work for me.
# Can we make this key better?

# table(Landings_missingLatLong$landing_site)

# Let's take a look at a site we KNOW should be assigned like "BAHIA DE KINO" but isn't:
TEST2<- Landings_missingLatLong %>%
  filter(landing_site=="BAHIA DE KINO")
```


# STEP 6: Create LatLong_OfficeKey ------------------------------

In situations where we can't merge by state-landing_site let's merge by state-office

```{r Key skeleton}
colnames(Hems_LatLongKey)

LatLong_OfficeKey0<- Hems_LatLongKey %>%
  select(state, office_name_clean, Hems_est_lat= lat_deci, Hems_est_lon= lon_deci) %>%
  group_by(state, office_name_clean) %>%
  mutate(
    lat_mean = mean(Hems_est_lat , na.rm= TRUE),
    lon_mean = mean(Hems_est_lon , na.rm= TRUE)
  )
# 36486 obs., 6 vars.

# Reorganize this key:
colnames(LatLong_OfficeKey0)

LatLong_OfficeKey<- LatLong_OfficeKey0 %>%
  select(state, office_name_clean, est_lat_mean= lat_mean, est_lon_mean= lon_mean) %>%
  distinct()
# Down to 73 obs., 4 vars.
```

```{r Create LatLong_OfficeKey}
# NOTE: CLEAN NAMES function found above

# Match based on state-office_name_clean:

Landing_missing.matched <- Landings_missingLatLong %>%
  left_join(
    LatLong_OfficeKey,
    by = c("state", "office_name_clean")
  )

# Then find any values that are unmatched to a lat-long. Is this a small enough number I can live with?
unmatched <- Landing_missing.matched %>%
  filter(is.na(est_lat_mean))
# Only 241371 out of 4713144 un-matched (= 5%). I can live with this!

table(unmatched$state)
```

```{r PURG? May 20 2026}
LatLong_OfficeKey$lat_long_confidence<- c("medium; Hem's estimate; estimate derived as a mean at the office level across sites")
```


# STEP 8: Add OHI region to the key ------------------------------

```{r Loading in OHI regions and buffer shape files}
# OHI regions:
OHI_regions_shape <- sf::st_read(
  "/home/shares/ohi/OHI_GOC/spatial/OHI_regions/marine/ohi_goc_final_071025.shp",
  options = "ENCODING=LATIN1"
)

OHI_regions_shape <- st_make_valid(OHI_regions_shape)

# Quick plot:
ggplot(OHI_regions_shape) +
  geom_sf(aes(fill = rgn_id)) +
  theme_minimal()

# 50km buffer:
OHI_buffer_shape <- sf::st_read(
  "/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer/",
  options = "ENCODING=LATIN1"
)

OHI_buffer_shape <- st_make_valid(OHI_buffer_shape)

# Quick plot:
ggplot(OHI_buffer_shape) +
  geom_sf(aes(fill = rgn_id)) +
  theme_minimal()
```

```{r Spatial join: assign OHI region to LatLong_SiteKey and LatLong_OfficeKey based on these shapes}
LatLong_SiteKey_final<- LatLong_SiteKey %>%
  filter(!is.na(lat_est))

LatLong_OfficeKey_sf <- st_as_sf(
  LatLong_OfficeKey,
  coords = c("est_lon_mean", "est_lat_mean"),
  crs = st_crs(OHI_regions_shape)
)
```



```{r Join the regions shape file and buffer}
OHI_combined_shape <- st_transform(OHI_regions_shape, st_crs(OHI_buffer_shape))

OHI_combined_shape <- st_make_valid(OHI_combined_shape)

# choose a target CRS and force it for region and buffer:
target_crs <- st_crs(OHI_buffer_shape)

regions <- st_transform(OHI_regions_shape, target_crs)
buffers <- OHI_buffer_shape

# Bind:
OHI_combined_shape <- bind_rows(
  regions %>% select(rgn_id, geometry),
  buffers %>% select(rgn_id, geometry)
)

# Quick plot:
ggplot(OHI_combined_shape) +
  geom_sf(aes(fill = rgn_id)) +
  theme_minimal()

# Make sure this key is clean by creating a union:
OHI_combined_shape <- OHI_combined_shape %>%
  group_by(rgn_id) %>%
  summarise(geometry = st_union(geometry), .groups = "drop")
```


# STEP 7: Export key(s) ------------------------------
```{r Clean up key before export}
# Final sites key:
LatLong_SiteKey_finalized<- LatLong_SiteKey

colnames(LatLong_SiteKey_finalized)

LatLong_SiteKey_finalized<- LatLong_SiteKey_finalized %>% 
  select(
    state, landing_site, GoC, est_habitat,
    lat_est, lon_est, lat_long_confidence,
    Hems_est_lat, Hems_est_lon, Hems_est_lat, Hems_est_lon,
    lat_Mel, lon_Mel,
    OHI_region_draft, site_notes_Mel
  )

colnames(LatLong_SiteKey_finalized)<- c(
  "state", "landing_site", "inside_GOC", "est_habitat",
  "lat_est", "lon_est", "latlon_est_flag",
  "HemKey_lat", "HemKey_lon", "HemPermitKey_lat", "HemPermitKey_lon",
  "MelKey_lat", "MelKey_lon",
  "OHI_region_draft", "site_notes_Mel"
)

LatLong_SiteKey_finalized <- LatLong_SiteKey_finalized %>%
  mutate(
    latlon_est_flag = if_else(
      is.na(latlon_est_flag),
      "alert 5: unable to assign lat-long based on state and landing site",
      NA_character_
    )
  )

sum(is.na(LatLong_SiteKey_finalized$lat_est))

# How many / which values are missing?
subset<- LatLong_SiteKey_finalized %>% 
  filter(is.na(lat_est))

LatLong_SiteKey_finalized<- LatLong_SiteKey_finalized %>%
  filter(is.na(lat_est))
# 1508 sites with lat-long details

# Final offices key:
LatLong_OfficeKey_finalized<- LatLong_OfficeKey
```


```{r PURG? remove unused dataframes}
rm(
  subset,
  BajaSonora_LatLongKey_clean, Jalisco_LatLongKey_clean,
  LatLong_merge.skeleton, LatLong_SiteKey,
  Mels_LatLongKey, Mels_LatLongKey_simple
)

rm(n_unique_sites, threshold_alert, threshold_low, threshold_medium)
```





# UNDER CONSTRUCTION / PURG - May 20 2026
# STEP 9: Add lat-long and OHI_region info to landings data ------------------------------
```{r OHI LatLong_SiteKey_finalized and RegionKey_pending}
LatLong_SiteKey_finalized<- read_csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/Sites_LatLongKey_May192026.csv")

RegionKey_pending<- read_csv(here("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/Spatial_Keys/RegionKey_UNDERCONSTRUCTION_Mar102026.csv"))

colnames(RegionKey_pending)
colnames(RegionKey_pending) <- c("state", "OHI_region_draft", "OHI_rgn_number", "OHI_rgn_name", "exclude_rgn", "OHI_rgn_notes")
```

```{r Append lat-long and OHI_region info to landings data}
temp_df <- Mex.Pacific_Landings %>%
  left_join(
    LatLong_SiteKey_finalized,
    by = c("state", "landing_site")
  )

table(temp_df$inside_GOC)

temp_df <- temp_df %>%
  left_join(
    RegionKey_pending,
    by = c("state", "OHI_region_draft")
  )

table(temp_df$inside_GOC)

colnames(temp_df)

test<- temp_df %>%
  filter(exclude_rgn=="exclude")

table(test$inside_GOC)
table(test$state)
table(test$OHI_rgn_number)

Mex.Pacific_Landings<- temp_df

rm(temp_df, test)
```

## STEP 9b: Save the NW Mexico landings appended with lat-long info in a separate .rds file:
```{r Save the Northwest Mexico dataframe: May 19 2026}
Mex.Pacific_Landings<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/Mex.Pacific_Landings_2000.present_May042026.rds"))
```

# NEXT STEPS: Use of the lat-long key can be found in "CONAPESCA_CleaningWorkflow_May2026.Rmd"
found here: /home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys

# STEP: Attach lat-long data to NW Mexico dataframe. (co-authors: Melanie Frazier, Hem Nalini Morzaria) ----------------------

# Flag these in the lat-long key as not in the Gulf [UNDER CONSTRUCTION May 19 2026]
office_name== c("EL ROSARIO", "ENSENADA", "TIJUANA")

# STEP: Attach lat-long and OHI_region info based on official landing site:
NOTE TO SELF: Come back to this! The lat long keys are imperfect. We'll want to fix this as much as possible.

# PURG
```{r Try cleaning the landing_site names}
# STEP 2: Trim down Landings_missingLatLong so we can do another merge
colnames(Landings_missingLatLong)

Landings_missingLatLong<- Landings_missingLatLong %>%
  select(-c(lat_Mel:lon_est))

# STEP 3: Clean the landing_site and office_name columns and place them next to their uncleaned counter-part:
Landings_missingLatLong <- Landings_missingLatLong %>%
  mutate(landing_site_clean = clean_name(landing_site),
         office_name_clean = clean_name(office_name)) %>%
  relocate(office_name, .before = office_name_clean) %>%
  relocate(landing_site, .before = landing_site_clean)

# SANITY CHECK: check entries where landing_site_clean is different than landing_site
TEST <- Landings_missingLatLong %>%
  filter(landing_site != landing_site_clean) %>%
  select(landing_site, landing_site_clean) %>%
  mutate(
    landing_site = iconv(landing_site,
                         from = "",
                         to = "UTF-8",
                         sub = ""),
    landing_site_clean = iconv(landing_site_clean,
                               from = "",
                               to = "UTF-8",
                               sub = ""),
    nchar_original = nchar(landing_site),
    nchar_clean = nchar(landing_site_clean)
  )

# These make sense. Many of them had "DE" removed.


# STEP 4: Create a modified key:
LatLong_SiteKey_modified<- LatLong_SiteKey %>%
  filter(!is.na(lat_est)) %>%
  mutate(landing_site_clean = clean_name(landing_site)) %>%
  relocate(landing_site, .before = landing_site_clean)

colnames(LatLong_SiteKey_modified)

# Start with the skeleton
LatLong_SiteKey_modified <- LatLong_SiteKey_modified %>%
  select(state, landing_site_clean, GoC, est_habitat, site_notes_Mel, lat_long_confidence, lat_est, lon_est) %>%
  distinct()

  mutate(
    # Determine if Hem should override Mel
    use_HemsKey = !is.na(Hems_est_lat) & !is.na(Hems_est_lon),
    
    # Compute difference between Hem and Mel
    keys_lat_diff = ifelse(!is.na(Hems_est_lat) & !is.na(lat_Mel), abs(Hems_est_lat - lat_Mel), 0),
    keys_lon_diff = ifelse(!is.na(Hems_est_lon) & !is.na(lon_Mel), abs(Hems_est_lon - lon_Mel), 0),
    
    # Decide final lat/lon
    lat_est = case_when(
      use_HemsKey & keys_lat_diff <= 2 & keys_lon_diff <= 2 ~ Hems_est_lat,  # Hem overrides when reasonable
      TRUE ~ lat_Mel  # Otherwise, keep Mel
    ),
    lon_est = case_when(
      use_HemsKey & keys_lat_diff <= 2 & keys_lon_diff <= 2 ~ Hems_est_lon,
      TRUE ~ lon_Mel
    ),
    
    # Decide final confidence
    lat_long_confidence = case_when(
      use_HemsKey & (keys_lat_diff > 2 | keys_lon_diff > 2) ~
        "alert 4: the lat-long values in Hem's key do not match the lat-long values in Mel's key by more than 2 degrees",
      use_HemsKey & (keys_lat_diff <= 2 & keys_lon_diff <= 2) ~ lat_long_confidence,  # keep Hem's original
      TRUE ~ NA_character_  # default when using Mel or no Hem info
    )
  )
```

```{r Match exact names}
# 2. EXACT MATCH FIRST
# I want to merge two dataframes in R together with slightly different column names. I want to merge Landings_missingLatLong to LatLong_SiteKey based on a match between Landings_missingLatLong$landing_site_clean and LatLong_SiteKey$landing_site

Landing_missing.matched <- Landings_missingLatLong %>%
  left_join(
    LatLong_SiteKey,
    by = c("state", "landing_site_clean" = "landing_site")
  )

# 1213865 (Good! As expected, no many-to-many matching)

table(Landing_missing.matched$state)

# 3. FIND UNMATCHED

unmatched <- Landing_missing.matched %>%
  filter(is.na(lat_mean))

table(unmatched$state)
```