################################################################################ 

For questions contact: Erica Ferrer, Ph.D. [emferrer\@ucsd.edu](mailto:emferrer@ucsd.edu){.email} - OR - [ferrer\@nceas.ucsb.edu](mailto:ferrer@nceas.ucsb.edu){.email}

################################################################################ 

# PRE-WORK ------------------------------

```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```

# STEP 1: Load in CONAPESCA RAW data, records from all of Mexico ------------------------------

This step is completed in the file shared by Juan Carlos: "03_combine_landings_edited.R" file path: /home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/JC_Scripts_EMF.edited/03_combine_landings_edited.R

# STEP 2: Work with CONAPESCA STITCHED data, records from all of Mexico ------------------------------

A stitched version of these data is available as one single file: "AllMex_landings_2000_present_EMF_edited.rds" file path: /home/shares/ohi/OHI_GOC/goal_prep/fp/juan_carlos/AllMex_landings_2000_present_EMF_edited.rds

```{r Load in CONAPESCA STITCHED data}
All.Mex_Landings_2000.present<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/juan_carlos/AllMex_landings_2000_present_EMF_edited.rds"))
# n = 12153879 obs., 20 vars.
```

# STEP 3: Create weight_kg_corrected column ------------------------------

```{r Landed weight correction}
# Kind of looks like sometimes 'landed weight' is 'live weight' and vice versa. Fix this:
All.Mex_Landings_2000.present$weight_kg_corrected <- ifelse(
  All.Mex_Landings_2000.present$live_weight >= All.Mex_Landings_2000.present$landed_weight,
  All.Mex_Landings_2000.present$live_weight,
  All.Mex_Landings_2000.present$landed_weight
)
# n = 12153879 obs., 21 vars.
```

# STEP 4a: Filter down CONAPESCA data to states in Northwest Mexico ------------------------------

```{r Filter out data from Northwest Mexico}
length(unique(All.Mex_Landings_2000.present$species_name)) 
# 2052 unique species in all of Mexico.
# But, we don't need species from ALL of Mexico, only the GoC region for now - so let's filter by state.

States_of_Interest<- c("BAJA CALIFORNIA", "BAJA CALIFORNIA SUR", "SONORA", "SINALOA", "NAYARIT", "JALISCO")

NW.Mex_Landings <- All.Mex_Landings_2000.present %>% filter(state %in% States_of_Interest)
# n = 4713144 obs., 21 vars.

length(unique(NW.Mex_Landings$species_name)) 
# 1827 unique species in States of Interest.
```

# STEP 4b: Attach lat-long and OHI_region info based on official landing site ------------------------------
NOTE TO SELF: Come back to this! The lat long keys are imperfect. We'll want to fix this as much as possible.
```{r Append lat-long data to sites in Baja BCS Sonora and Sinaloa}
# Filter OUT sites from Baja that are on the Pacific side using lat-Long file from Mel's repository:
# NOTE TO SELF: THIS PART IS SUBJECT TO CHANGE once we integrate lat-long data from Hem. Mar 10 2026
BajaSonora_LatLong_Key<- read_csv("/home/frazier/ohi_goc_prep/FP/FIS/v2025/data/Baja_Sonora_states_key.csv")
colnames(BajaSonora_LatLong_Key) <- c("state", "landing_site", "OHI_rgn", "lat", "long", "GoC")

Jalisco_LatLong_Key<- read_csv(here("/home/frazier/ohi_goc_prep/FP/FIS/v2025/data/landing_sites_jalisco.csv"))
colnames(Jalisco_LatLong_Key) <- c("landing_site", "CONAPESCA_landing_key", "habitat", "lat", "long", "site_notes", "OHI_rgn", "state")


#### CLEAN LAT-LONG KEYS: ####
# Clean the Baja-Sonora key...
BajaSonora_LatLong_Key_clean <- BajaSonora_LatLong_Key %>%
  
  # Normalize state names:
  mutate(state = toupper(state)) %>%
  
  # Remove any rows where landing_site is NA (NAs cause matching issues later on):
  filter(!is.na(landing_site)) %>%
  
  # Remove any direct duplicates of landing_site:
  group_by(state, landing_site) %>%
  slice(1) %>%  # keeps the first occurrence per key
  ungroup()

# Look into the rows that were removed in the de duplication of BajaSonora_LatLong_Key...
removed_rows <- anti_join(BajaSonora_LatLong_Key, BajaSonora_LatLong_Key_clean, 
                          by = c("landing_site"))

# Convert lat-long entries to numeric entries, accounting for unique cases where the minus (-) is actually an n dash or m dash:
BajaSonora_LatLong_Key_clean <- BajaSonora_LatLong_Key_clean %>%
  mutate(
    long = as.numeric(str_replace_all(long, "[−–—]", "-")),
    lat  = as.numeric(str_replace_all(lat, "[−–—]", "-"))
  )

# Clean the Jalisco key...
Jalisco_LatLong_Key_clean <- Jalisco_LatLong_Key %>%
  # Normalize state names
  mutate(state = toupper(state)) %>%
  
  # Remove any rows where landing_site is NA (NAs cause matching issues later on):
  filter(!is.na(landing_site)) %>%
  
  # Keep only one row per CONAPESCA_landing_key; if duplicate keys exist, keep first
  group_by(CONAPESCA_landing_key) %>%
  slice(1) %>%  # keeps the first occurrence per key
  ungroup() %>%
  
  # Optional: remove any direct duplicates of landing_site if there are rows with NA in the CONAPESCA_landing_key
  group_by(state, landing_site) %>%
  slice(1) %>%  # keeps the first occurrence per key
  ungroup()

# Look into the rows that were removed in the de duplication of Jalisco_LatLong_Key...
removed_rows <- anti_join(Jalisco_LatLong_Key, Jalisco_LatLong_Key_clean, 
                          by = c("CONAPESCA_landing_key", "landing_site"))

# Convert lat-long entries to numeric entries, accounting for unique cases where the minus (-) is actually an n dash or m dash:
Jalisco_LatLong_Key_clean <- Jalisco_LatLong_Key_clean %>%
  mutate(
    long = as.numeric(str_replace_all(long, "[−–—]", "-")),
    lat  = as.numeric(str_replace_all(lat, "[−–—]", "-"))
  )

# Add the "GoC" helpful column here based on a latitude cut off for the Nayarit/Jalisco OHI region, region 9 starting at 20.40895:
Jalisco_LatLong_Key_clean <- Jalisco_LatLong_Key_clean %>%
  mutate(GoC = if_else(lat < 20.40895, "no", NA_character_))

#### MERGE LAT-LONG KEYS: ####
LatLong_MasterKey_pending <- bind_rows(BajaSonora_LatLong_Key_clean, Jalisco_LatLong_Key_clean) %>%
  filter(!is.na(landing_site)) %>%  # remove the NA from this data frame as this will create a many-to-many matching issue later on
  mutate(long = as.numeric(long))
# 1176 obs., 9 vars.

# Remove unused data frames:
rm(BajaSonora_LatLong_Key, Jalisco_LatLong_Key, removed_rows)

# Take a look
unique(LatLong_MasterKey_pending$state)
unique(LatLong_MasterKey_pending$OHI_rgn)

# My regional key (temporary, subject to change, UNDER CONSTRUCTION):
RegionKey_pending<- read_csv(here("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/RegionKey_UNDERCONSTRUCTION_Mar102026.csv"))
# Previously: RegionKey_UNDERCONSTRUCTION_Dec012025.csv
colnames(RegionKey_pending) <- c("state", "OHI_rgn", "OHI_rgn_number", "OHI_rgn_name", "exclude_rgn", "exclude_notes")
```

```{r Append lat-long and OHI_region info to landings data}
# Append lat-long info to landings data:
temp_df <- NW.Mex_Landings %>%
  left_join(
    LatLong_MasterKey_pending,
    by = c("state", "landing_site")  # matching columns
  )

table(temp_df$GoC)
# 4713144 obs., 28 vars.
# 1090368 "no" not in the GOC.

# Append RegionKey_pending info to landings data:
temp_df <- temp_df %>%
  left_join(
    RegionKey_pending,
    by = c("state", "OHI_rgn")  # matching columns
  )
# 4713144 obs., 32 vars.
table(temp_df$GoC)
# 1090368 "no" not in the GOC as we'd expect!

NW.Mex_Landings<- temp_df
# 4713144 obs., 32 vars.

# Delete temp_df, not needed anymore.
rm(temp_df)
```

# STEP 4c: Save the NW Mexico landings appended with lat-long as a separate .rds file
```{r Save the Northwest Mexico dataframe: Mar 11 2026}
saveRDS(GOC_Landings, file="/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/NWMex_Landings_2000.present_EMF.Mar112026.rds")
# 4713144 obs., 32 vars.
```

# STEP 5a: Filter down data to sites in the GOC, exclude all sites that definitely AREN'T in the Gulf ------------------------------
Additional research will be required for some landing sites. Details pending additional lat-long key work. See notes here (as of Mar 11 2026): https://docs.google.com/document/d/1ANHEdRi8aWC0qLNaM1_c_nLtn5GihTFgHDcU96bry5Y/edit?tab=t.0

```{r filter down data to only the GOC}
# Generate filtered landings data for the GoC:
GOC_Landings <- NW.Mex_Landings %>%
  filter(GoC != "no" | is.na(GoC))
# 3622776 obs., 32 vars.

# Check that this all makes sense:
length(unique(GOC_Landings$species_name))
length(unique(GOC_Landings$main_species_group))
# 1703 unique species from 57 main species groups in the Gulf of California, keeping in mind that these are not really SPECIES (per se) but rather log-book entries for species.
```
NOW 1703 (previously 1693) unique species in the Gulf of California according to Juan Carlos' version of the CONAPESCA data base, keeping in mind that these are not really SPECIES, per se, but rather log-book entries for species.

# STEP 5b: Save the GOC landings dataframe as a separate .rds file
```{r Save the GOC_Landings dataframe: Mar 11 2026}
saveRDS(GOC_Landings, file="/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Landings_2000.present_EMF.Mar112026.rds")
# 3622776 obs., 32 vars.
# Previous iteration: GOC_FIS.landings_CLEAN_as.of.Dec092025.rds
```


# STEP 6a: Separate out key words from the official record for common name ------------------------------

Keeping in mind that some key terms reflect animal parts but are actually part of the species common name like "aleta de" versus "aleta blanca"

** QUICK SANITY CHECK: Is "aleta de aleta" an issue? Short answer is no, double "aleta" does not seem to be an issue.
GOC_Landings %>%
  rename(recorded_taxa_name = species_name) %>%
  mutate(
    aleta_count = stringi::stri_count_fixed(recorded_taxa_name, "ALETA")
  ) %>%
  count(aleta_count)

```{r Filter out keywords from recorded_taxa_name to get recorded_taxa_name_clean}
# Part A. Keyword list, excluding "ALETA" which is handled separately -----------------
Keyword_List <- c(
  "S.C.", "S/C", "S",
  "ENTERO", "ENTERA", "IND", "ENT",
  "FRESCO", "FRESCA", "FCA", "FCO", "FC",
  "ECO",
  "CONG", "CONGELADO",
  "VIVO", "VIVA", "VIVAS", "VIVITA",
  "DE CULTIVO", "DE CULTIVO VIVO", "ABASTO RANCHO", "RANCHO",
  "HUEVA DE",
  "FILETE", "FILETE DE", "ECA", "FILETE DE ECO", "DESHUESADA", "CARNE", "CARNE DE",
  "DESCABEZADO", "DES", "DSC", "DESC", "DESV", "DESVICERADO",
  "CABEZA DE", "BUCHE", "BUCHE DE", "TRIPA DE", "COLAS DE", "MANOS DE", "MANOS",
  "ANCAS DE", "ANCAS", "ANCA",
  "PIEL DE ECA", "PIEL DE SECA", "PIEL DE", "PIEL",
  "SECO DE", "SECA", "SECA DE", "SECO", "SECO DE",
  "COCINERO", "COCIDO", "COCIDA", "COCIDAS", "ASADO",
  "DE ALTA MAR",
  "CONCHA DE", "CONCHA",
  "POLVO DE",
  "CARNADA", "P/CARNADA",
  "DE AGUA DULCE",
  "MANTO", "MANTO DE", "PULPA DE", "PULPA FRESCA",
  "PULPO DE", "TENTACULOS DE", "TUBO DE", "TUBO",
  "NO APTO PARA EMPAQUE"
  # Remove ALETA from main keywords; we'll handle this separately
  #"ALETA", "ALETA ECA DE", "ALETA FCA DE"
)

# Order the key words longest first (note that this is necessary for the next steps):
Keyword_List <- Keyword_List[order(nchar(Keyword_List), decreasing = TRUE)]

# Escape regex meta-characters:
Keyword_List_escaped <- stringi::stri_replace_all_regex(
  Keyword_List, "([.|(){}+*?^$\\[\\]\\\\])", "\\\\$1"
)

# Final regex:
keyword_pattern <- paste0(
  "(?<![A-Za-z0-9])(",
  paste(Keyword_List_escaped, collapse = "|"),
  ")(?![A-Za-z0-9])"
)


# Part B. Take care of special ALETA phrases (handled separately from other key words) -----------------
special_aleta_adj <- c("BLANCA", "AMARILLA", "CORTA", "AZUL")

remove_aleta_except_special <- function(x) {
  stringi::stri_replace_all_regex(
    x,
    pattern = paste0("\\bALETA\\b(?!\\s+(?:", paste(special_aleta_adj, collapse = "|"), "))"),
    replacement = "",
    vectorize_all = FALSE
  )
}


# Part C. MAIN PARSING PIPELINE (RUN TIME = ~5 min, optimized but still computationally heavy) -----------------

GOC_Landings_parsed <- GOC_Landings %>%
  
  # Rename columns
  rename(
    recorded_taxa_name = species_name,
    recorded_taxa_group = main_species_group
  ) %>%
  
  # Extract keywords
  mutate(
    keywords_list = stringi::stri_extract_all_regex(recorded_taxa_name, keyword_pattern)
  ) %>%
  mutate(keywords_list = map(keywords_list, ~ { length(.x) <- 5; .x })) %>%
  unnest_wider(keywords_list, names_sep = "_") %>%
  rename_with(~ paste0("keywords_", 1:5), starts_with("keywords_list_")) %>%
  
  # Clean recorded_taxa_name
  mutate(
    recorded_taxa_name_clean = recorded_taxa_name %>%
      remove_aleta_except_special() %>%                      # Remove ALETA unless special
      stringi::stri_replace_all_regex(keyword_pattern, "") %>% # Remove other keywords
      stringi::stri_replace_all_regex("\\b(Y|DE|P/)\\b", "") %>% # Remove standalone Y/DE/P/
      stringi::stri_replace_all_regex("[.,/\\\\]+", " ") %>%      # Remove punctuation
      str_squish()
  ) %>%
  
  # Extract scientific name
  mutate(
    recorded_scientific_name = stringi::stri_extract_first_regex(recorded_taxa_name_clean, "\\(([^)]+)\\)") %>%
      stringi::stri_replace_all_regex("\\(|\\)", ""),
    scientific_name_recorded_T.F = !is.na(recorded_scientific_name),
    recorded_scientific_name = if_else(
      scientific_name_recorded_T.F,
      stringi::stri_trans_totitle(recorded_scientific_name),
      NA_character_
    )
  ) %>%
  
  # Extract common name
  mutate(
    common_name_clean = stringi::stri_trim_both(
      stringi::stri_replace_first_regex(recorded_taxa_name_clean, "\\s*\\([^)]*\\)", "")
    ),
    common_name_clean = tolower(common_name_clean)
  )
# 3622776 obs., 41 vars.


# Part D. Trim and reorder columns -----------------
GOC_Landings_clean <- GOC_Landings_parsed %>%

  select(
    # General info
    state, office_name, landing_site, landing_site_key, GoC, lat, long,
    year, month,
    receipt_date_fixed, period_start_fixed, period_end_fixed,
    fleet, acuaculture_production, eu_rnpa, eu_name, vessel_rnpa, vessel_name,
    
    # Species info
    recorded_taxa_group, recorded_taxa_name,
    
    # Cleaned / derived names
    recorded_taxa_name_clean, recorded_scientific_name,
    scientific_name_recorded_T.F, common_name_clean,
    
    # Keywords
    keywords_1, keywords_2, keywords_3, keywords_4, keywords_5,
    
    # Landing details
    value, landed_weight, live_weight, weight_kg_corrected
 
  )
# 3622776 obs., 32 vars. (but they are different variables than before)

# How many common names are there in this dataset?
length(unique(GOC_Landings_clean$common_name_clean))
length(unique(GOC_Landings_clean$recorded_taxa_group))
# NOW 737 common names (previously 734, and before that, 973 with the less comprehensive keywords_list), over 57 taxonomic groups
```

** QUICK SANITY CHECK: did this work for taxa like yellow-fin tuna? Yes, it seems to have worked!
test <- GOC_Landings_clean %>%
  filter(str_detect(recorded_taxa_name, "ATUN ALETA AMARILLA"))


# STEP 6b: OPTIONAL Save the GOC_Landings_parsed dataframe as a separate .rds file
```{r Save this dataframe - landings keywords & species for top groupings: Nov 10 2025}
# saveRDS(GOC_Landings_clean, file="/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Landings_parsed_2000.present_EMF.Mar112026.rds")
```


# STEP 7: Merge the cleaned master sp key with landings data ------------------------------
```{r Load species key}
MasterSpKey_pending <- read_csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/Species_Keys/CleanedMaster_SpKey_Dec012025.csv")
# 802 obs of 41 variables
```

```{r Shave down the size of Master_SpKey_simple}
Master_SpKey_simple<- MasterSpKey_pending %>% 
  select(recorded_taxa_group, recorded_taxa_name_clean, common_name_clean,
         estimated_species_name, species_est_certainty, estimated_genus_name, estimated_family_name,
         IHH_order, IHH_class, IHH_functional_major_groups_sm, IHH_fishery)
# 802 obs., 11 vars.
```

```{r Merge landings with filtered sp key}
GOC_Landings_clean <- GOC_Landings_clean %>%
  left_join(
    Master_SpKey_simple,
    by = c(
      "recorded_taxa_group",
      "recorded_taxa_name_clean",
      "common_name_clean"
    )
  )
# n = 3622776 obs. as expected, 41 vars.
```

## Save GOC_Landings with species as separate .rds file

```{r Save this dataframe - landings keywords & species for top groupings: Nov 10 2025}
saveRDS(GOC_Landings_clean, file="/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Landings_wSpecies_2000.present_EMF.Mar112026.rds")
```

STOPPED HERE, Mar 11 2026