################################################################################ 

For questions contact: Erica Ferrer, Ph.D. [emferrer\@ucsd.edu](mailto:emferrer@ucsd.edu){.email} - OR - [ferrer\@nceas.ucsb.edu](mailto:ferrer@nceas.ucsb.edu){.email}

################################################################################ 

# PRE-WORK ------------------------------

```{r Clear environment}
rm(list=ls(all=T))
```

# STEP 3: Attach lat-long data to NW Mexico dataframe. ------------------------------

# Flag these in the lat-long key as not in the Gulf [UNDER CONSTRUCTION April 3 2026]
office_name== c("EL ROSARIO", "ENSENADA", "TIJUANA")

## STEP 3a: Attach lat-long and OHI_region info based on official landing site:
NOTE TO SELF: Come back to this! The lat long keys are imperfect. We'll want to fix this as much as possible.
```{r Append lat-long data to sites in Baja BCS Sonora and Sinaloa}
# Filter OUT sites from Baja that are on the Pacific side using lat-Long file from Mel's repository:
# NOTE TO SELF: THIS PART IS SUBJECT TO CHANGE once we integrate lat-long data from Hem. Mar 10 2026
BajaSonora_LatLongKey<- read_csv("/home/frazier/ohi_goc_prep/FP/Exploratory_Research/FIS_TeamResearch/data/Baja_Sonora_states_key.csv")
colnames(BajaSonora_LatLongKey) <- c("state", "landing_site", "OHI_rgn", "lat", "long", "GoC")

Jalisco_LatLongKey<- read_csv(here("/home/frazier/ohi_goc_prep/FP/Exploratory_Research/FIS_TeamResearch/data/landing_sites_jalisco.csv"))
colnames(Jalisco_LatLongKey) <- c("landing_site", "CONAPESCA_landing_key", "est_habitat", "lat", "long", "site_notes", "OHI_rgn", "state")


#### CLEAN LAT-LONG KEYS: ####
# Clean the Baja-Sonora key...
BajaSonora_LatLongKey_clean <- BajaSonora_LatLongKey %>%
  
  # Normalize state names:
  mutate(state = toupper(state)) %>%
  
  # Remove any rows where landing_site is NA (NAs cause matching issues later on):
  filter(!is.na(landing_site)) %>%
  
  # Remove any direct duplicates of landing_site:
  group_by(state, landing_site) %>%
  slice(1) %>%  # keeps the first occurrence per key
  ungroup()

# Look into the rows that were removed in the de duplication of BajaSonora_LatLongKey...
removed_rows <- anti_join(BajaSonora_LatLongKey, BajaSonora_LatLongKey_clean, 
                          by = c("landing_site"))

# Convert lat-long entries to numeric entries, accounting for unique cases where the minus (-) is actually an n dash or m dash:
BajaSonora_LatLongKey_clean <- BajaSonora_LatLongKey_clean %>%
  mutate(
    long = as.numeric(str_replace_all(long, "[−–—]", "-")),
    lat  = as.numeric(str_replace_all(lat, "[−–—]", "-"))
  )

# Clean the Jalisco key...
Jalisco_LatLongKey_clean <- Jalisco_LatLongKey %>%
  # Normalize state names
  mutate(state = toupper(state)) %>%
  
  # Remove any rows where landing_site is NA (NAs cause matching issues later on):
  filter(!is.na(landing_site)) %>%
  
  # Keep only one row per CONAPESCA_landing_key; if duplicate keys exist, keep first
  group_by(CONAPESCA_landing_key) %>%
  slice(1) %>%  # keeps the first occurrence per key
  ungroup() %>%
  
  # Optional: remove any direct duplicates of landing_site if there are rows with NA in the CONAPESCA_landing_key
  group_by(state, landing_site) %>%
  slice(1) %>%  # keeps the first occurrence per key
  ungroup()

# Look into the rows that were removed in the de duplication of Jalisco_LatLongKey...
removed_rows <- anti_join(Jalisco_LatLongKey, Jalisco_LatLongKey_clean, 
                          by = c("CONAPESCA_landing_key", "landing_site"))

# Convert lat-long entries to numeric entries, accounting for unique cases where the minus (-) is actually an n dash or m dash:
Jalisco_LatLongKey_clean <- Jalisco_LatLongKey_clean %>%
  mutate(
    long = as.numeric(str_replace_all(long, "[−–—]", "-")),
    lat  = as.numeric(str_replace_all(lat, "[−–—]", "-"))
  )

# Add the "GoC" helpful column here based on a latitude cut off for the Nayarit/Jalisco OHI region, region 9 starting at 20.40895:
Jalisco_LatLongKey_clean <- Jalisco_LatLongKey_clean %>%
  mutate(GoC = if_else(lat < 20.40895, "no", NA_character_))

#### MERGE LAT-LONG KEYS TO CREATE MASTER LAT-LONG KEYS: ####
Mels_LatLongKey_pending <- bind_rows(BajaSonora_LatLongKey_clean, Jalisco_LatLongKey_clean) %>%
  filter(!is.na(landing_site)) %>%  # remove the NA from this data frame as this will create a many-to-many matching issue later on
  mutate(long = as.numeric(long))
# 1176 obs., 9 vars.

# Remove unused data frames:
rm(BajaSonora_LatLongKey, Jalisco_LatLongKey, removed_rows)

# Take a look
unique(Mels_LatLongKey_pending$state)
unique(Mels_LatLongKey_pending$OHI_rgn)
```

# STEP 3b: Integrate Hem's Lat-Long key ------------------------------
```{r Loading in Hem's lat-long}
load("/home/shares/ohi/OHI_GOC/goal_prep/fp/hem_morzaria/georefcatchlocs.rda")
Hems_LatLongKey0<- georefcatchlocs

rm(georefcatchlocs)

# Explore Hem's key
colnames(Hems_LatLongKey0)

# table(Hems_LatLongKey0$NOM_LOC) # This is the equivalent of landing_site.
# table(Hems_LatLongKey0$NOM_ENT) # Okay, so fishery_office_ent is essentially the state. We can match by this column.
# 
# table(Hems_LatLongKey0$fishery_office)
# table(Hems_LatLongKey0$fishery_office_ent) # Okay, so fishery_office_ent is also the state? I think this is the state of the fishery office. Ignore for now.

# Before we can merge we need to re-name some columns:
colnames(Hems_LatLongKey0)<- c("catch_code", "rnp_code", "landing_site", "state", "fishery_office", "fishery_office_ent", "lat_deci_Hem", "lon_deci_Hem" )
```

```{r Look through Hem's key}
Hems_LatLongKey_simple<- Hems_LatLongKey0 %>% 
  select(state, landing_site, fishery_office, lat_deci_Hem, lon_deci_Hem)
# Could include or exclude fishery_office. These offices are no longer used by CONAPESCA.

Hems_LatLongKey_simple<- Hems_LatLongKey_simple %>% 
  distinct()
# 829 obs.

length(unique(Hems_LatLongKey_simple$landing_site))
# 691 unique landing sites, at least by name, in Hem's key

length(unique(Mels_LatLongKey_pending$landing_site))
# 1068 "unique" landing sites, at least by name, in Mel's key

# Let's take a closer look at landing sites with the same name but different lat-longs in both keys. Are they at least in the same state? Are their lat-longs similar?
Site_duplicates_HemsKey <- Hems_LatLongKey_simple %>%
  select(state, landing_site, fishery_office) %>%
  distinct() %>%
  group_by(state, landing_site) %>%
  filter(n() > 1)
# From this it is clear that Hem's key is assigning a unique lat-long value to every combination of state, landing_site, AND fishery_office. This creates a situation where we might have more than one unique combination of lat-long values per state and landing_site (due to multiple fishery offices); however, the lat-long values are fairly similar in most cases....

Site_duplicates_OHIKey<- Mels_LatLongKey_pending[
  duplicated(Mels_LatLongKey_pending[, c("state", "landing_site")]) |
  duplicated(Mels_LatLongKey_pending[, c("state", "landing_site")], fromLast = TRUE),
]
# 0 duplicates. Yay!

# EXAMPLE:
subset<- Hems_LatLongKey_simple %>%
  filter(landing_site== "EL SIARIC")
# Lat-longs are all pretty similar here. I think it makes sense to just kind of pick one combo and move on. 

# TAKING WHAT WE CAN FROM HEM'S KEY:
# Let's take one representative value for lat-long from Hem's key and flag any duplicate site lat-longs that are more than 1 degree apart:

# Define thresholds
threshold_alert <- 2      # alert if differences > 2 degrees
threshold_low <- 1        # low confidence if differences > 1 degree
threshold_medium <- 0.5   # medium confidence if differences > 0.5 degree

# Step 1: Compute lat/lon ranges and assign confidence levels
Hems_LatLong_flags <- Hems_LatLongKey_simple %>%
  group_by(state, landing_site) %>%
  summarise(
    lat_range = max(lat_deci_Hem, na.rm = TRUE) - min(lat_deci_Hem, na.rm = TRUE),
    lon_range = max(lon_deci_Hem, na.rm = TRUE) - min(lon_deci_Hem, na.rm = TRUE),
    lat_long_confidence = case_when(
      lat_range > threshold_alert | lon_range > threshold_alert ~
        "alert 3: multiple lat-long values in Hem's key that are more than 2 degrees apart",
      lat_range > threshold_low | lon_range > threshold_low ~
        "alert 2: multiple lat-long values in Hem's key that are more than 1 degree apart",
      lat_range > threshold_medium | lon_range > threshold_medium ~
        "alert 1: multiple lat-long values in Hem's key that are more than 0.5 degree apart",
      TRUE ~ "high confidence"
    ),
    .groups = "drop"
  )

# Step 2: Keep a representative lat-long (first occurrence)
Hems_LatLongKey_compressed <- Hems_LatLongKey_simple %>%
  group_by(state, landing_site) %>%
  slice(1) %>%
  ungroup()

# Step 3: Merge the confidence/alert information
Hems_LatLongKey_compressed <- Hems_LatLongKey_compressed %>%
  left_join(Hems_LatLong_flags %>% select(state, landing_site, lat_long_confidence),
            by = c("state", "landing_site"))

# SANITY CHECK: Looks good!
print(Hems_LatLongKey_compressed)
# 724 obs. of 5 variables

# SANITY CHECK: Number of unique state + landing_site combinations
n_unique_sites <- Hems_LatLongKey_compressed %>%
  distinct(state, landing_site) %>%
  nrow()
# 724 unique state-landing_site combinations.

# remove dataframes to keep workspace clean:
rm(Hems_LatLong_flags, Hems_LatLongKey_simple, 
   Site_duplicates_HemsKey, Site_duplicates_OHIKey,
   subset)
```

```{r Create a backbone of a key to merge to}
Hems_LatLongKey_compressed # 724 obs. of 6 variables
Mels_LatLongKey_pending # 1176 obs. of 9 variables

# Shave down Mel's key:
Mels_LatLongKey_simple<- Mels_LatLongKey_pending %>%
  select(state, landing_site, lat, long, OHI_rgn, GoC, est_habitat, site_notes)
colnames(Mels_LatLongKey_simple)<- c("state", "landing_site", "lat_Mel", "lon_Mel", "OHI_region_draft", "GoC", "est_habitat", "site_notes_Mel")
# 1176 obs. of 8 variables
table(Mels_LatLongKey_simple$GoC)
# 400 sites in the GoC, 588 sites likely not in the GoC

# Create a backbone to merge to...
colnames(NW.Mex_Landings)

LatLong_merge.skeleton<- NW.Mex_Landings %>%
  select(state, landing_site) %>%
  distinct()
# 2046 distinct state-landingsite combos in (NW Mexico) that need to be filled using information from Hem's key or Mel's key. Privilege information from Hem's key.

# LOGIC:
# Take information from Hem's key and Mels key to populate the LatLong_merge.skeleton but in a preferential order. First, populate LatLong_merge.skeleton with information from Mel's key. Then, if landing_site information is available from Hem's key, supersede the state and landing_site info given by Mel's key. 
# If the lat-long values from Hem's key are non-NA and > 2 degrees different from Mel's non-NA lat-long values assign a lat.long_confidence of "low confidence, the lat-long values in Hem's key do not match the lat-long values in Mel's key by more than 2 degrees.

# Start with the skeleton
LatLong_merged <- LatLong_merge.skeleton %>%
  # Join Mel's key first (preferential population)
  left_join(Mels_LatLongKey_simple, by = c("state", "landing_site")) %>%
  # Join Hem's key
  left_join(Hems_LatLongKey_compressed %>%
              select(state, landing_site, lat_deci_Hem, lon_deci_Hem, lat_long_confidence),
            by = c("state", "landing_site")) %>%
  mutate(
    # Determine if Hem should override Mel
    use_HemsKey = !is.na(lat_deci_Hem) & !is.na(lon_deci_Hem),
    
    # Compute difference between Hem and Mel
    lat_diff = ifelse(!is.na(lat_deci_Hem) & !is.na(lat_Mel), abs(lat_deci_Hem - lat_Mel), 0),
    lon_diff = ifelse(!is.na(lon_deci_Hem) & !is.na(lon_Mel), abs(lon_deci_Hem - lon_Mel), 0),
    
    # Decide final lat/lon
    lat_final = case_when(
      use_HemsKey & lat_diff <= 2 & lon_diff <= 2 ~ lat_deci_Hem,  # Hem overrides when reasonable
      TRUE ~ lat_Mel  # Otherwise, keep Mel
    ),
    lon_final = case_when(
      use_HemsKey & lat_diff <= 2 & lon_diff <= 2 ~ lon_deci_Hem,
      TRUE ~ lon_Mel
    ),
    
    # Decide final confidence
    lat_long_confidence = case_when(
      use_HemsKey & (lat_diff > 2 | lon_diff > 2) ~
        "alert 4: the lat-long values in Hem's key do not match the lat-long values in Mel's key by more than 2 degrees",
      use_HemsKey & (lat_diff <= 2 & lon_diff <= 2) ~ lat_long_confidence,  # keep Hem's original
      TRUE ~ NA_character_  # default when using Mel or no Hem info
    )
  )

# Great! Seems like this worked. 
table(LatLong_merged$lat_long_confidence)
table(LatLong_merged$GoC)

#                             alert 1: multiple lat-long values in Hem's key that are more than 0.5 degree apart 
#                                                                                                             21 
#                               alert 2: multiple lat-long values in Hem's key that are more than 1 degree apart 
#                                                                                                             19 
#                              alert 3: multiple lat-long values in Hem's key that are more than 2 degrees apart 
#                                                                                                             11 
# alert 4: the lat-long values in Hem's key do not match the lat-long values in Mel's key by more than 2 degrees 
#                                                                                                             40 
#                                                                                                high confidence 
#                                                                                                            499
```

```{r Clean up key before export}
LatLongKey_finalized<- LatLong_merged
# 2046 obs., 13 vars.

colnames(LatLongKey_finalized)

LatLongKey_finalized<- LatLongKey_finalized %>% 
  select(state, landing_site, GoC, est_habitat, 
         lat_final, lon_final, lat_long_confidence, lat_deci_Hem, lon_deci_Hem, lat_Mel, lon_Mel, 
         OHI_region_draft, site_notes_Mel)

colnames(LatLongKey_finalized)<- c("state", "landing_site", "inside_GOC", "est_habitat", "lat_final", "lon_final", "FLAGS", "HemKey_lat", "HemKey_lon", "MelKey_lat", "MelKey_lon", "OHI_region_draft", "site_notes_Mel")

sum(is.na(LatLongKey_finalized$lat_final))
# 680 out of 2026 sites are missing lat-long info. Primarily in Sinaloa and Nayarit, which makes some sense for our purposes.

subset<- LatLongKey_finalized %>% 
  filter(is.na(lat_final))
table(subset$state)
    # BAJA CALIFORNIA BAJA CALIFORNIA SUR             JALISCO             NAYARIT             SINALOA              SONORA 
    #              54                  17                  25                 193                 324                  67

# remove dataframes to keep workspace clean:
rm(subset, 
   Hems_LatLongKey0, Hems_LatLongKey, Hems_LatLongKey_compressed, Hems_LatLong_flags, Hems_LatLongKey_simple,
   BajaSonora_LatLongKey, BajaSonora_LatLongKey_clean,
   Jalisco_LatLongKey, Jalisco_LatLongKey_clean,
   LatLong_merge.skeleton, LatLong_merged,
   Mels_LatLongKey_pending, Mels_LatLongKey_simple)

rm(n_unique_sites, threshold_alert, threshold_low, threshold_medium)
```

```{r Export this new lat-long key}
# Save this dataframe:
write.csv(LatLongKey_finalized, "/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/Sites_LatLongKey_Apr22026.csv", row.names = FALSE)
# 2046 obs., 13 vars.
```

# STEP 3c: Add lat-long and OHI_region info to landings data ------------------------------

Use of the lat-long key can be found in "CONAPESCA_CreatingLatLongKey_Mar2026.Rmd"
found here: /home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys