################################################################################ 

For questions contact: Erica Ferrer, Ph.D. [emferrer\@ucsd.edu](mailto:emferrer@ucsd.edu){.email} - OR - [ferrer\@nceas.ucsb.edu](mailto:ferrer@nceas.ucsb.edu){.email}

################################################################################ 

What's different about this key from April's key?

~108 lines effectively unchanged (copy-paste from April).
4 lines minor comment/header edits.
23 lines deleted (STEP 15).
~51 lines added (STEP 3b2 + one reference line).
~95 lines rewritten (merge, export, 3c–3d, next steps).
~147 lines in May that don’t match April line-for-line (adds + rewrites + line shift from deletion).


# PRE-WORK ------------------------------

```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
library(fuzzyjoin)
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```

# STEP 1: Attach lat-long data to NW Mexico dataframe. (co-authors: Melanie Frazier, Hem Nalini Morzaria) ----------------------

# Flag these in the lat-long key as not in the Gulf [UNDER CONSTRUCTION May 19 2026]
office_name== c("EL ROSARIO", "ENSENADA", "TIJUANA")

# STEP 2: Attach lat-long and OHI_region info based on official landing site:
NOTE TO SELF: Come back to this! The lat long keys are imperfect. We'll want to fix this as much as possible.

```{r Append lat-long data to sites in Baja BCS Sonora and Sinaloa}
# Filter OUT sites from Baja that are on the Pacific side using lat-Long file from Mel's repository:

# Integrated Hem georefcatchlocs (STEP 3b) and georeflocskm permit/INEGI georef (STEP 3b2) May 19 2026
BajaSonora_LatLongKey<- read_csv("/home/frazier/ohi_goc_prep/FP/Exploratory_Research/FIS_TeamResearch/data/Baja_Sonora_states_key.csv")
colnames(BajaSonora_LatLongKey) <- c("state", "landing_site", "OHI_rgn", "lat", "long", "GoC")

Jalisco_LatLongKey<- read_csv(here("/home/frazier/ohi_goc_prep/FP/Exploratory_Research/FIS_TeamResearch/data/landing_sites_jalisco.csv"))
colnames(Jalisco_LatLongKey) <- c("landing_site", "CONAPESCA_landing_key", "est_habitat", "lat", "long", "site_notes", "OHI_rgn", "state")


#### CLEAN LAT-LONG KEYS: ####
# Clean the Baja-Sonora key...
BajaSonora_LatLongKey_clean <- BajaSonora_LatLongKey %>%
  
  # Normalize state names:
  mutate(state = toupper(state)) %>%
  
  # Remove any rows where landing_site is NA (NAs cause matching issues later on):
  filter(!is.na(landing_site)) %>%
  
  # Remove any direct duplicates of landing_site:
  group_by(state, landing_site) %>%
  slice(1) %>%  # keeps the first occurrence per key
  ungroup()


# Look into the rows that were removed in the de duplication of BajaSonora_LatLongKey...
removed_rows <- anti_join(BajaSonora_LatLongKey, BajaSonora_LatLongKey_clean, 
                          by = c("landing_site"))

# Convert lat-long entries to numeric entries, accounting for unique cases where the minus (-) is actually an n dash or m dash:
BajaSonora_LatLongKey_clean <- BajaSonora_LatLongKey_clean %>%
  mutate(
    long = as.numeric(str_replace_all(long, "[−–—]", "-")),
    lat  = as.numeric(str_replace_all(lat, "[−–—]", "-"))
  )

# Clean the Jalisco key...
Jalisco_LatLongKey_clean <- Jalisco_LatLongKey %>%
  # Normalize state names
  mutate(state = toupper(state)) %>%
  
  # Remove any rows where landing_site is NA (NAs cause matching issues later on):
  filter(!is.na(landing_site)) %>%
  
  # Keep only one row per CONAPESCA_landing_key; if duplicate keys exist, keep first
  group_by(CONAPESCA_landing_key) %>%
  slice(1) %>%  # keeps the first occurrence per key
  ungroup() %>%
  
  # Optional: remove any direct duplicates of landing_site if there are rows with NA in the CONAPESCA_landing_key
  group_by(state, landing_site) %>%
  slice(1) %>%  # keeps the first occurrence per key
  ungroup()

# Look into the rows that were removed in the de-duplication of Jalisco_LatLongKey...
removed_rows <- anti_join(Jalisco_LatLongKey, Jalisco_LatLongKey_clean, 
                          by = c("CONAPESCA_landing_key", "landing_site"))

# Convert lat-long entries to numeric entries, accounting for unique cases where the minus (-) is actually an n dash or m dash:
Jalisco_LatLongKey_clean <- Jalisco_LatLongKey_clean %>%
  mutate(
    long = as.numeric(str_replace_all(long, "[−–—]", "-")),
    lat  = as.numeric(str_replace_all(lat, "[−–—]", "-"))
  )

# Add the "GoC" helpful column here based on a latitude cut off for the Nayarit/Jalisco OHI region, region 9 starting at 20.40895:
Jalisco_LatLongKey_clean <- Jalisco_LatLongKey_clean %>%
  mutate(GoC = if_else(lat < 20.40895, "no", NA_character_))


#### MERGE LAT-LONG KEYS TO CREATE MASTER LAT-LONG KEYS: ####
Mels_LatLongKey_pending <- bind_rows(BajaSonora_LatLongKey_clean, Jalisco_LatLongKey_clean) %>%
  filter(!is.na(landing_site)) %>%  # remove the NA from this data frame as this will create a many-to-many matching issue later on
  mutate(long = as.numeric(long))
# 1176 obs., 9 vars.

# Remove unused data frames:
rm(BajaSonora_LatLongKey, Jalisco_LatLongKey, removed_rows)

# Take a look
unique(Mels_LatLongKey_pending$state)
unique(Mels_LatLongKey_pending$OHI_rgn)
```

# STEP 3: Integrate Hem's catch-location lat-long key (georeflocskm) ------------------------------
```{r Load and prepare georeflocskm}
load("/home/shares/ohi/OHI_GOC/_raw_data/CONAPESCA/d2025/SharedByHemMorzaria/permit_data/georeflocskm.rda")
Hems_PermitLatLongKey0 <- georeflocskm
#rm(georeflocskm)
```

```{r Flag potential duplicate entries in Hem's key}
# Columns: NOM_ENT, NOM_LOC, deci_lat, deci_lon, fishery_office, fishery_office_ent, distance_km
colnames(Hems_PermitLatLongKey0) <- c(
  "state", "landing_site", "geometry", "lon_deci", "lat_deci",
  "fishery_office", "fishery_office_ent", "distance", "distance_km"
)

Hems_LatLongKey_simple <- Hems_PermitLatLongKey0 %>%
  mutate(
    state = toupper(stringr::str_squish(state)),
    landing_site = toupper(
      stringi::stri_trans_general(landing_site, id = "Latin-ASCII")
    ),
    landing_site = stringr::str_squish(landing_site)
  ) %>%
  select(state, landing_site, fishery_office, lat_deci, lon_deci, distance_km) %>%
  distinct()

# Thresholds:
threshold_alert  <- 2
threshold_low    <- 1
threshold_medium <- 0.5

Hems_LatLong_flags <- Hems_LatLongKey_simple %>%
  group_by(state, landing_site) %>%
  summarise(
    lat_range = max(lat_deci, na.rm = TRUE) - min(lat_deci, na.rm = TRUE),
    lon_range = max(lon_deci, na.rm = TRUE) - min(lon_deci, na.rm = TRUE),
    lat_long_confidence = case_when(
      lat_range > threshold_alert | lon_range > threshold_alert ~
        "alert 3: multiple lat-long values in Hem's permit/INEGI key that are more than 2 degrees apart",
      lat_range > threshold_low | lon_range > threshold_low ~
        "alert 2: multiple lat-long values in Hem's permit/INEGI key that are more than 1 degree apart",
      lat_range > threshold_medium | lon_range > threshold_medium ~
        "alert 1: multiple lat-long values in Hem's permit/INEGI key that are more than 0.5 degree apart",
      TRUE ~ "high confidence"
    ),
    .groups = "drop"
  )

Hems_LatLongKey_collated <- Hems_LatLongKey_simple %>%
  group_by(state, landing_site) %>%
  slice(1) %>%
  ungroup() %>%
  left_join(
    Hems_LatLong_flags %>% select(state, landing_site, lat_long_confidence),
    by = c("state", "landing_site")
  )

rm(Hems_LatLongKey_simple, Hems_LatLong_flags)

table(Hems_LatLongKey_collated$lat_long_confidence)

# alert 1: multiple lat-long values in Hem's permit/INEGI key that are more than 0.5 degree apart 
#                                                                                             589 
#   alert 2: multiple lat-long values in Hem's permit/INEGI key that are more than 1 degree apart 
#                                                                                            1072 
#  alert 3: multiple lat-long values in Hem's permit/INEGI key that are more than 2 degrees apart 
#                                                                                            1107 
#                                                                                 high confidence 
#                                                                                           24128 

# SANITY CHECK:
TEST<- Hems_LatLongKey_collated %>%
  filter(lat_long_confidence== "alert 3: multiple lat-long values in Hem's permit/INEGI key that are more than 2 degrees apart")

# SANITY CHECK: Number of unique state + landing_site combinations
n_unique_sites <- Hems_LatLongKey_collated %>%
  distinct(state, landing_site) %>%
  nrow()
# 26896 unique state-landing_site combinations.
```


# STEP 4: Create a backbone of a key to merge to... ------------------------------

```{r Prepare to merge keys}
dim(Hems_LatLongKey_collated) # 26896 obs., 7 variables
dim(Mels_LatLongKey_pending) # 1176 obs., 9 variables

# Shave down Mel's key:
Mels_LatLongKey_simple<- Mels_LatLongKey_pending %>%
  select(state, landing_site, lat, long, OHI_rgn, GoC, est_habitat, site_notes)

colnames(Mels_LatLongKey_simple)<- c("state", "landing_site", "lat_Mel", "lon_Mel", "OHI_region_draft", "GoC", "est_habitat", "site_notes_Mel")
# 1176 obs. of 8 variables

table(Mels_LatLongKey_simple$GoC)
# 400 sites in the GoC, 588 sites likely not in the GoC
```

```{r Load the Mexican Pacific}
Mex.Pacific_Landings<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/Mex.Pacific_Landings_2000.present_May042026.rds"))

dim(Mex.Pacific_Landings)
# 6406303 obs., 21 vars.
```

```{r States_of_Interest}
unique(Mex.Pacific_Landings$state)

States_of_Interest<- c("BAJA CALIFORNIA", "BAJA CALIFORNIA SUR",
                       "SONORA", "SINALOA",
                       "NAYARIT", "JALISCO")
```

```{r Create a backbone of a key to merge to...}
# Create a backbone to merge to...
colnames(Mex.Pacific_Landings)

LatLong_merge.skeleton<- Mex.Pacific_Landings %>%
  select(state, landing_site) %>%
  distinct()
# 3060 distinct state-landingsite combos in the Mexican Pacific that I am going to try to fill using information from Hem's key or Mel's key. Privilege information from Hem's key.
```

```{r Merge to the backbone}
# LOGIC (May 19 2026):
# Mel's key -> Hem catchlocs (georefcatchlocs) when within 2 degrees of Mel 
# -> Hem permit/INEGI georef (georeflocskm) to fill remaining NAs.

colnames(Hems_LatLongKey_collated)

# Start with the skeleton
LatLong_SiteKey <- LatLong_merge.skeleton %>%
  # Join Mel's key first (preferential population)
  left_join(Mels_LatLongKey_simple, by = c("state", "landing_site")) %>%
  # Join Hem's key
  left_join(Hems_LatLongKey_collated %>%
              select(state, landing_site, lat_deci, lon_deci, lat_long_confidence),
            by = c("state", "landing_site")) %>%
  mutate(
    # Determine if Hem should override Mel
    use_HemsKey = !is.na(lat_deci) & !is.na(lon_deci),
    
    # Compute difference between Hem and Mel
    lat_diff = ifelse(!is.na(lat_deci) & !is.na(lat_Mel), abs(lat_deci - lat_Mel), 0),
    lon_diff = ifelse(!is.na(lon_deci) & !is.na(lon_Mel), abs(lon_deci - lon_Mel), 0),
    
    # Decide final lat/lon
    lat_est = case_when(
      use_HemsKey & lat_diff <= 2 & lon_diff <= 2 ~ lat_deci,  # Hem overrides when reasonable
      TRUE ~ lat_Mel  # Otherwise, keep Mel
    ),
    lon_est = case_when(
      use_HemsKey & lat_diff <= 2 & lon_diff <= 2 ~ lon_deci,
      TRUE ~ lon_Mel
    ),
    
    # Decide final confidence
    lat_long_confidence = case_when(
      use_HemsKey & (lat_diff > 2 | lon_diff > 2) ~
        "alert 4: the lat-long values in Hem's key do not match the lat-long values in Mel's key by more than 2 degrees",
      use_HemsKey & (lat_diff <= 2 & lon_diff <= 2) ~ lat_long_confidence,  # keep Hem's original
      TRUE ~ NA_character_  # default when using Mel or no Hem info
    )
  )

table(LatLong_SiteKey$lat_long_confidence)
table(LatLong_SiteKey$GoC)

# SANITY CHECK: how many lat-long values are missing from the states of interest?
TEST<- LatLong_SiteKey %>%
  filter(state %in% States_of_Interest) %>%
  filter(is.na(lat_est))

# 538 sites in the States_of_Interest are missing lat-long values
```

```{r Check to see how many entries are covered by this key alone}
NW.Mex_Landings<- Mex.Pacific_Landings %>%
  filter(state %in% States_of_Interest)

dim(NW.Mex_Landings)
# 4713144 obs.

NW.Mex_Landings <- NW.Mex_Landings %>%
  left_join(LatLong_SiteKey, by = c("state", "landing_site"))
# 4713144 obs.

# SANITY CHECK: how many records are missing lat-long values?
Landings_missingLatLong<- NW.Mex_Landings %>%
  filter(is.na(lat_est))
# 1/4 of records (n= 1,213,865) are not assigned... 
# This doesn't really work for me.
# Can we make this key better?

# table(Landings_missingLatLong$landing_site)

# Let's take a look at a site we KNOW should be assigned like "BAHIA DE KINO" but isnt:
TEST2<- Landings_missingLatLong %>%
  filter(landing_site=="BAHIA DE KINO")
```

```{r In situations where we can't merge by state-landing_site let's merge by state-office}
LatLong_OfficeKey0<- Hems_LatLongKey_collated %>%
  select(state, fishery_office, lat_deci, lon_deci) %>%
  group_by(state, fishery_office) %>%
  mutate(
    lat_mean = mean(lat_deci , na.rm= TRUE),
    lon_mean = mean(lon_deci , na.rm= TRUE)
  )

# Reorganize this key:
colnames(LatLong_OfficeKey0)

LatLong_OfficeKey<- LatLong_OfficeKey0 %>%
  select(state, office_name= fishery_office, lat_mean, lon_mean) %>%
  distinct()

```

```{r}

#--------------------------
# 1. CLEAN NAMES
#--------------------------

clean_name <- function(x) {

  # fix encoding
  x <- iconv(x, from = "", to = "UTF-8", sub = "")

  x %>%
    toupper() %>%
    trimws() %>%
    gsub("\\bDE\\b", "", .) %>%
    gsub("\\s+", " ", .)
}

LatLong_OfficeKey <- LatLong_OfficeKey %>%
  mutate(office_clean = clean_name(office_name)) %>%
  select(-c(office_name))

Landings_missingLatLong <- Landings_missingLatLong %>%
  mutate(office_clean = clean_name(office_name)) %>%
  select(-c(office_name))

#--------------------------
# 2. EXACT MATCH FIRST
#--------------------------

matched_exact <- Landings_missingLatLong %>%
  left_join(
    LatLong_OfficeKey,
    by = c("state", "office_clean")
  )

table(matched_exact$state)

#--------------------------
# 3. FIND UNMATCHED
#--------------------------

unmatched <- matched_exact %>%
  filter(is.na(lat_mean))

table(unmatched$state)

#--------------------------
# 4. FUZZY MATCH ONLY UNMATCHED
#--------------------------

fuzzy_matches <- unmatched %>%
  rowwise() %>%
  mutate(
    best_match = {
      candidates <- LatLong_OfficeKey %>%
        filter(state == cur_data()$state)

      if(nrow(candidates) == 0) NA_character_
      else {
        d <- stringdist::stringdist(
          office_clean,
          candidates$office_clean,
          method = "jw"
        )

        candidates$office_name[which.min(d)]
      }
    }
  )
```

```{r Clean up key before export}
# Final sites key:
LatLong_SiteKey_finalized<- LatLong_SiteKey

colnames(LatLong_SiteKey_finalized)

LatLong_SiteKey_finalized<- LatLong_SiteKey_finalized %>% 
  select(
    state, landing_site, GoC, est_habitat,
    lat_est, lon_est, lat_long_confidence,
    lat_deci, lon_deci, lat_deci, lon_deci,
    lat_Mel, lon_Mel,
    OHI_region_draft, site_notes_Mel
  )

colnames(LatLong_SiteKey_finalized)<- c(
  "state", "landing_site", "inside_GOC", "est_habitat",
  "lat_est", "lon_est", "latlon_est_flag",
  "HemKey_lat", "HemKey_lon", "HemPermitKey_lat", "HemPermitKey_lon",
  "MelKey_lat", "MelKey_lon",
  "OHI_region_draft", "site_notes_Mel"
)

LatLong_SiteKey_finalized <- LatLong_SiteKey_finalized %>%
  mutate(
    latlon_est_flag = if_else(
      is.na(latlon_est_flag),
      "alert 5: unable to assign lat-long based on state and landing site",
      NA_character_
    )
  )

sum(is.na(LatLong_SiteKey_finalized$lat_est))

subset<- LatLong_SiteKey_finalized %>% 
  filter(is.na(lat_est))

# Final offices key:
LatLong_OfficeKey_finalized<- LatLong_OfficeKey
```

```{r Export these new lat-long keys}
write.csv(
  LatLong_SiteKey_finalized,
  "/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/Spatial_Keys/Sites_LatLongKey_May192026.csv",
  row.names = FALSE
)

write.csv(
  LatLong_OfficeKey_finalized,
  "/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/Spatial_Keys/Offices_LatLongKey_May192026.csv",
  row.names = FALSE
)
```

```{r remove unused dataframes}
rm(
  subset,
  Hems_LatLongKey0, Hems_LatLongKey_collated,
  Hems_LatLongKey_collated,
  BajaSonora_LatLongKey_clean, Jalisco_LatLongKey_clean,
  LatLong_merge.skeleton, LatLong_SiteKey,
  Mels_LatLongKey_pending, Mels_LatLongKey_simple
)

rm(n_unique_sites, threshold_alert, threshold_low, threshold_medium)
```

# STEP 3c: Add lat-long and OHI_region info to landings data ------------------------------
```{r OHI LatLong_SiteKey_finalized and RegionKey_pending}
LatLong_SiteKey_finalized<- read_csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/Sites_LatLongKey_May192026.csv")

RegionKey_pending<- read_csv(here("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/Spatial_Keys/RegionKey_UNDERCONSTRUCTION_Mar102026.csv"))

colnames(RegionKey_pending)
colnames(RegionKey_pending) <- c("state", "OHI_region_draft", "OHI_rgn_number", "OHI_rgn_name", "exclude_rgn", "OHI_rgn_notes")
```

```{r Append lat-long and OHI_region info to landings data}
temp_df <- Mex.Pacific_Landings %>%
  left_join(
    LatLong_SiteKey_finalized,
    by = c("state", "landing_site")
  )

table(temp_df$inside_GOC)

temp_df <- temp_df %>%
  left_join(
    RegionKey_pending,
    by = c("state", "OHI_region_draft")
  )

table(temp_df$inside_GOC)

colnames(temp_df)

test<- temp_df %>%
  filter(exclude_rgn=="exclude")

table(test$inside_GOC)
table(test$state)
table(test$OHI_rgn_number)

Mex.Pacific_Landings<- temp_df

rm(temp_df, test)
```

## STEP 3d: Save the NW Mexico landings appended with lat-long info in a separate .rds file:
```{r Save the Northwest Mexico dataframe: May 19 2026}
Mex.Pacific_Landings<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/Mex.Pacific_Landings_2000.present_May042026.rds"))
```

# NEXT STEPS: Use of the lat-long key can be found in "CONAPESCA_CleaningWorkflow_May2026.Rmd"
found here: /home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys
