################################################################################ 

For questions contact: Erica Ferrer, Ph.D. [emferrer\@ucsd.edu](mailto:emferrer@ucsd.edu){.email} - OR - [ferrer\@nceas.ucsb.edu](mailto:ferrer@nceas.ucsb.edu){.email}

################################################################################ 

What's different about this key from April's key?

~108 lines effectively unchanged (copy-paste from April).
4 lines minor comment/header edits.
23 lines deleted (STEP 15).
~51 lines added (STEP 3b2 + one reference line).
~95 lines rewritten (merge, export, 3c–3d, next steps).
~147 lines in May that don’t match April line-for-line (adds + rewrites + line shift from deletion).


# PRE-WORK ------------------------------

```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
library(fuzzyjoin)
library(sf) # for work on shape fines
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```

## clean_names function
```{r Load clean_names function}
# Takes text and standardizes it, which we will need for merging later on:

clean_names <- function(x) {

  x <- iconv(x, from = "", to = "UTF-8", sub = "")
  x <- stringi::stri_trans_general(x, "Latin-ASCII")

  x %>%
    toupper() %>%
    trimws() %>%
    gsub("\\bDE\\b", "", .) %>%
    gsub("\\s*\\([^\\)]*\\)", "", .) %>%   # remove parentheses + contents
    gsub("\\s+", " ", .) %>%
    trimws()
}
```

# STEP 0: Load in relevant fisheries landings data ------------------------------

```{r Load the Mexican Pacific - Mex.Pacific_Landings0}
Mex.Pacific_Landings0<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/Mex.Pacific_Landings0_2000.present_May042026.rds"))

dim(Mex.Pacific_Landings0)
# 6406303 obs., 21 vars.
```

```{r States_of_Interest}
unique(Mex.Pacific_Landings0$state)

States_of_Interest<- c("BAJA CALIFORNIA", "BAJA CALIFORNIA SUR",
                       "SONORA", "SINALOA",
                       "NAYARIT", "JALISCO")
```

```{r Generate NW.Mex_Landings0 dataframe}
NW.Mex_Landings0<- Mex.Pacific_Landings0 %>%
  filter(state %in% States_of_Interest)

dim(NW.Mex_Landings0)
# 4713144 obs., 21 obs.
```

```{r Clean the landing_site and office_name_clean names & create NW.Mex_Landings dataframe}
# Clean the landing_site names in NW.Mex_Landings: (est. 1-2 minute run time)
NW.Mex_Landings<- NW.Mex_Landings0 %>%
  mutate(landing_site_clean= clean_names(landing_site),
         office_name_clean= clean_names(office_name)) %>%
  relocate(landing_site_clean, .after= landing_site) %>%
  relocate(office_name_clean, .after= office_name)
# 4713144 obs., 23 obs.
```


# FIRST TEST: How well does this revised key (Hems_LatLong_Site.Office.Key) work on its own, no alterations?

# STEP 1: Test how good is Hems_LatLong_Site.Office.Key for matching with no additional edits 
Load in Hem's catch-location lat-long key (georeflocskm)

```{r Load and prepare georeflocskm}
load("/home/shares/ohi/OHI_GOC/_raw_data/CONAPESCA/d2025/SharedByHemMorzaria/permit_data/georeflocskm.rda")
Hems_LatLongKey0 <- georeflocskm
#rm(georeflocskm)
```

```{r Clean Hem's key}
# Check column names first:
colnames(Hems_LatLongKey0) <- c(
  "state", "landing_site", "geometry", "lon_deci", "lat_deci",
  "fishery_office", "fishery_office_ent", "distance", "distance_km"
)

# Clean the landing_site names:
Hems_LatLongKey_clean <- Hems_LatLongKey0 %>%
  mutate(landing_site_clean= clean_names(landing_site),
         office_name_clean= clean_names(fishery_office))

# Re-arrange columns:
colnames(Hems_LatLongKey_clean)

Hems_LatLongKey_clean<- Hems_LatLongKey_clean %>%
  select(state, landing_site, landing_site_clean, fishery_office, office_name_clean, lon_deci, lat_deci, geometry, distance, distance_km)
```

```{r Create Hems_LatLong_Site.Office.Key}
# Compress Hems_LatLongKey_clean by state-office_name_clean-landing_site_clean matches:
Hems_LatLong_Site.Office.Key <- Hems_LatLongKey_clean %>%
  group_by(state, landing_site_clean, office_name_clean) %>%
  summarise(
    site.office_mean_lon = mean(lon_deci, na.rm = TRUE),
    site.office_mean_lat = mean(lat_deci, na.rm = TRUE),
    site.office_lon_range = max(lon_deci, na.rm = TRUE) - min(lon_deci, na.rm = TRUE),
    site.office_lat_range = max(lat_deci, na.rm = TRUE) - min(lat_deci, na.rm = TRUE),
    n_site.offices = n(),
    .groups = "drop"
  ) %>%
  
  mutate(
    confidence_in_office.site_latlon = case_when(

      site.office_lon_range > 2 |
      site.office_lat_range > 2 ~
        "low; lat or lon estimates meaned at the state-office-site level differ by >2º in the georeflocskm key",

      site.office_lon_range > 1 |
      site.office_lat_range > 1 ~
        "medium; lat or lon estimates meaned at the state-office-site level differ by 1-2º in the georeflocskm key",

      site.office_lon_range > 0.5 |
      site.office_lat_range > 0.5 ~
        "medium-high; lat or lon estimates meaned at the state-office-site level differ by 0.5-1º in the georeflocskm key",

      TRUE ~
        "high; high coherence among lat-lon estimates meaned at the state-office-site level"
    )
  )

# 30276 obs., 9 vars.


table(Hems_LatLong_Site.Office.Key$confidence_in_office.site_latlon)
# high; high coherence
# 28886 
# medium-high; lat or lon estimates differ by 0.5-1º in the georeflocskm key 
# 900 
# medium; lat or lon estimates differ by 1-2º in the georeflocskm key 
# 443
# low; lat or lon estimates differ by >2º in the georeflocskm key
# 47

# Clean up / simplify the confidence column:
Hems_LatLong_Site.Office.Key <- Hems_LatLong_Site.Office.Key %>%
  mutate(
    lat.lon_confidence_simple =
      sub(";.*", "", confidence_in_office.site_latlon)
  )

table(Hems_LatLong_Site.Office.Key$lat.lon_confidence_simple)
      #  high         low      medium medium-high 
      # 28886          47         443         900 
```

```{r Test: how good is Hems_LatLong_Site.Office.Key for matching with no additional edits?}
# Does the extra work actually improve the matching? 

NW.Mex_Landings_matched1<- NW.Mex_Landings %>%
  left_join(
    Hems_LatLong_Site.Office.Key,
    by = c("state", "office_name_clean", "landing_site_clean")
  )

# 4713144 obs., 29 obs.
# Excellent! This worked, but...

# SANITY CHECK: How many entries are unmatched after the merge?
unmatched<- NW.Mex_Landings_matched1 %>%
  filter(is.na(site.office_mean_lat))
# 2607660 obs. are unmatched (55%), so this strategy ALONE does not work well.

length(unmatched$site.office_mean_lat) / length(NW.Mex_Landings_matched1$site.office_mean_lat)

# Let's try to understand sites didnt match:
unmatched_slice<- unmatched %>%
  slice_sample(n= 200)
# I see that there could be several entries captured with a simple state-landing_site_clean match.
```

# FIRST OUTCOME: Use of the Hems_LatLong_Site.Office.Key does not work that well. Let's try a few simple slice, dices, and creative merges that may improve matching.


# STEP 2: Split Hem's key down to the level of landing_site_clean to improve matching  ------------------------------

```{r Compress Hems_LatLong_Site.Office.Key even further to match by state and landing_site_clean}
# Compress by state and landing_site_clean:
Hems_LatLong_Site.Key <- Hems_LatLong_Site.Office.Key %>%
  group_by(state, landing_site_clean) %>%
  summarise(
    site_mean_lon = mean(site.office_mean_lon, na.rm = TRUE),
    site_mean_lat = mean(site.office_mean_lat, na.rm = TRUE),

    site_lon_range = max(site.office_mean_lon, na.rm = TRUE) - min(site.office_mean_lon, na.rm = TRUE),
    site_lat_range = max(site.office_mean_lat, na.rm = TRUE) - min(site.office_mean_lat, na.rm = TRUE),
    n_sites = n(),
    
    .groups = "drop"
  ) %>%
  
  mutate(
    confidence_in_site_latlon = case_when(

      site_lon_range > 2 |
      site_lat_range > 2 ~
        "low; lat or lon estimates meaned at the state-site level differ by >2º in the georeflocskm key",

      site_lon_range > 1 |
      site_lat_range > 1 ~
        "medium; lat or lon estimates meaned at the state-site level differ by 1-2º in the georeflocskm key",

      site_lon_range > 0.5 |
      site_lat_range > 0.5 ~
        "medium-high; lat or lon estimates meaned at the state-site level differ by 0.5-1º in the georeflocskm key",

      TRUE ~
        "high; high coherence among lat-lon estimates meaned at the state-site level"
    )
  )

#  25130 obs., 13 vars. (fewer rows than before, good!)

# Clean up / simplify the confidence column:
Hems_LatLong_Site.Key <- Hems_LatLong_Site.Key %>%
  mutate(
    lat.lon_confidence_simple =
      sub(";.*", "", confidence_in_site_latlon)
  )
```

```{r Sequential second merge}
# Join using the more specific key first: Hems_LatLong_Site.Office.Key
# Join again using the fallback key: Hems_LatLong_Site.Key
# Only use fallback lat/long values where the first join did not populate them.

colnames(Hems_LatLong_Site.Office.Key)
colnames(Hems_LatLong_Site.Key)

NW.Mex_Landings_matched2 <- NW.Mex_Landings_matched1 %>%
  
  # Second merge with first fallback key (Hems_LatLong_Site.Key)
  left_join(
    Hems_LatLong_Site.Key %>%
      distinct(),
    
    by = c("state", "landing_site_clean")
  )

# 4713144 obs., 35 obs. (nice!!)

# SANITY CHECK: How many entries are unmatched after the merge?
unmatched<- NW.Mex_Landings_matched2 %>%
  filter(is.na(site_mean_lon))

dim(unmatched)
# 1444423 (31%) of entries are still unmatched; can we do better?

length(unmatched$site.office_mean_lat) / length(NW.Mex_Landings_matched2$site.office_mean_lat)

# Let's try to understand sites didn't match:
unmatched_slice<- unmatched %>%
  slice_sample(n= 200)
# I see that there could be several entries captured with a state-office matches.

# SANITY CHECK: the sites that did match this time, what do those look like?
matched_slice<- NW.Mex_Landings_matched2 %>%
  filter(is.na(site.office_mean_lon) & !is.na(site_mean_lon)) %>%
  slice_sample(n= 200)
# Neat! These look good an make sense. For example, in my data I find the landing_site_clean == CORTINA LA PRESA ADOLFO LOPEZ MATEOS (office_name_clean == CULIACAN, state == SINALOA), with no exact match in Hem's key. But when I search only by state and landing_site_clean, I find a match!
```
# SECOND OUTCOME: This key works okay, but can probably do even better.


# STEP 3: Split Hem's key down to the level of office_name_clean to improve matching ------------------------------

```{r Compress Hems_LatLong_Site.Office.Key even further to match by state and office_name_clean}
# Compress by office_name_clean
Hems_LatLong_Office.Key <- Hems_LatLong_Site.Office.Key %>%
  group_by(state, office_name_clean) %>%
  summarise(
    office_mean_lon = mean(site.office_mean_lon, na.rm = TRUE),
    office_mean_lat = mean(site.office_mean_lat, na.rm = TRUE),
    
    office_lon_range = max(site.office_mean_lon, na.rm = TRUE) - min(site.office_mean_lon, na.rm = TRUE),
    office_lat_range = max(site.office_mean_lat, na.rm = TRUE) - min(site.office_mean_lat, na.rm = TRUE),
    n_offices = n(),
    .groups = "drop"
  ) %>%
  
  mutate(
    confidence_in_office_latlon = case_when(

      office_lon_range > 2 |
      office_lat_range > 2 ~
        "low; lat or lon estimates meaned at the state-office level differ by >2º in the georeflocskm key",

      office_lon_range > 1 |
      office_lat_range > 1 ~
        "medium; lat or lon estimates meaned at the state-office level differ by 1-2º in the georeflocskm key",

      office_lon_range > 0.5 |
      office_lat_range > 0.5 ~
        "medium-high; lat or lon estimates meaned at the state-office level differ by 0.5-1º in the georeflocskm key",

      TRUE ~
        "high; high coherence among lat-lon estimates meaned at the state-office level"
    )
  )

#  73 obs., 8 vars. (smaller than before, good!)

# Clean up / simplify the confidence column:
Hems_LatLong_Office.Key <- Hems_LatLong_Office.Key %>%
  mutate(
    lat.lon_confidence_simple =
      sub(";.*", "", confidence_in_office_latlon)
  )
```

```{r Sequential third merge}
# The idea is:
# Join using Hem's specific key first.
# Join using the fallback key ()
# Only use fallback lat/long values where the first join did not populate them.

colnames(Hems_LatLong_Site.Office.Key)
colnames(Hems_LatLong_Site.Key)
colnames(Hems_LatLong_Office.Key)

NW.Mex_Landings_matched3 <- NW.Mex_Landings_matched2 %>%
  
  # Third merge with second fallback key (Hems_LatLong_Office.Key):
  left_join(
    Hems_LatLong_Office.Key %>%
      distinct(),
    by = c("state", "office_name_clean")
  )
  
# 4713144 obs., 41 obs. (nice!!)

# SANITY CHECK: the sites that did match this time, what do those look like?
colnames(NW.Mex_Landings_matched3)
matched_slice<- NW.Mex_Landings_matched3 %>%
  filter(is.na(site.office_mean_lat) & is.na(site_mean_lat) & !is.na(office_mean_lat)) %>%
  slice_sample(n= 200)
# Great! This seems to have worked for state-landing_sites not found in Hem's key. Take this example:
# landing_site_clean== MAICULIS, office_name_clean == EL FUERTE, state == SINALOA... I can't find this match in Hem's key but I can find EL FUERTE in SINALOA.
```

```{r Fill missing values from fallback merges & clean up}
# Note that the coalesce function works from left to right, so R will use site.office_mean_lon if not NA, otherwise site_mean_lon, otherwise office_mean_lon
colnames(NW.Mex_Landings_matched3)

NW.Mex_Landings_matched3 <- NW.Mex_Landings_matched3 %>%
  mutate(
    est_lon = coalesce(
      site.office_mean_lon,
      site_mean_lon,
      office_mean_lon
    ),

    est_lat = coalesce(
      site.office_mean_lat,
      site_mean_lat,
      office_mean_lat
    ),

    lat.lon_source = case_when(
      !is.na(site.office_mean_lon) ~ "estimate at the state-office-site level",
      !is.na(site_mean_lon) ~ "estimate at the state-site level",
      !is.na(office_mean_lon) ~ "estimate at the state-office level",
      TRUE ~ NA_character_
    ),
    
      lat.lon_confidence = coalesce(
      confidence_in_office.site_latlon,
      confidence_in_site_latlon,
      confidence_in_office_latlon
    )
  )

dim(NW.Mex_Landings_matched3)
# 4713144 obs., 45 vars.

# Check lat-long estimate sources:
table(NW.Mex_Landings_matched3$lat.lon_source)
     # estimate at the state-office level estimate at the state-office-site level 
     #                            1108040 (24.5%)                    2105484 (44.6%)
     #   estimate at the state-site level 
     #                            1163237 (24.6%)

sum(is.na(NW.Mex_Landings_matched3$lat.lon_source))
# 336383 unmatched (7.1%)

# Check confidence values:
table(NW.Mex_Landings_matched3$lat.lon_confidence)
     # estimate at the state-office level estimate at the state-office-site level 
     #                            1108040 (24.5%)                    2105484 (44.6%)
     #   estimate at the state-site level 
     #                            1163237 (24.6%)

# Clean up / simplify the confidence column:
NW.Mex_Landings_matched3 <- NW.Mex_Landings_matched3 %>%
  mutate(
    lat.lon_confidence_simple =
      sub(";.*", "", lat.lon_confidence)
  )

table(NW.Mex_Landings_matched3$lat.lon_confidence_simple)
    #    high             low              medium           medium-high 
    # 3105807 (66%)       248691 (5%)      808243 (17%)     214020 (5%)
# Looks pretty good, yay!
```

```{r How many entries are unmatched now?}
# SANITY CHECK: How many entries are unmatched after THIS merge?
unmatched<- NW.Mex_Landings_matched3 %>%
  filter(is.na(est_lon))
# Only 336383 (7%) of entries are still unmatched.

# Let's try to understand sites didn't match:
unmatched_slice<- unmatched %>%
  slice_sample(n= 200)
```

```{r plot unmatched entries over time}
table(unmatched$year)

unmatched_summary <- unmatched %>%
  group_by(year) %>%
  summarise(n_count = n())

# Plot these missing values over time:
unmatched_plot1<- ggplot(unmatched_summary, aes(x = year, y = n_count)) +
  geom_line() +
  geom_point() +
  labs(
    x = "Year",
    y = "Count of Unmatched",
    title = "Unmatched Records Over Time"
  ) +
  theme_bw()

unmatched_plot1
```

# THIRD OUTCOME: This key works pretty well! Only missing lat-longs for 7% of entries, and the majority of entries are high confidence.


# STEP 4: Add information from Mel's keys to supplement missing and low-confidence values ------------------------------
Additional refinements to the key with Mel's key (only if needed)

```{r Load and clean Mel's Baja + BCS + Sonora key}
# Load in Mel's Baja-Sonora key:
BajaSonora_LatLongKey<- read_csv("/home/frazier/ohi_goc_prep/FP/Exploratory_Research/FIS_TeamResearch/data/Baja_Sonora_states_key.csv")
# 906 obs., 6 vars.

# Cleaning:
colnames(BajaSonora_LatLongKey) <- c("state", "landing_site", "OHI_rgn", "lat", "long", "GoC")

# Normalize state and landing_site names:
BajaSonora_LatLongKey<- BajaSonora_LatLongKey %>%
  mutate(state = clean_names(state),
         landing_site_clean= clean_names(landing_site))
  
# Convert lat-long entries to numeric entries, accounting for unique cases where the minus (-) is actually an n dash or m dash:
BajaSonora_LatLongKey_clean <- BajaSonora_LatLongKey %>%
  mutate(
    long = as.numeric(str_replace_all(long, "[−–—]", "-")),
    lat  = as.numeric(str_replace_all(lat, "[−–—]", "-"))
  )

# Now assign mean values for compressed key:
BajaSonora_LatLongKey_clean <- BajaSonora_LatLongKey_clean %>%
  
  filter(!is.na(landing_site_clean)) %>%
  
  group_by(state, landing_site_clean) %>%
  summarise(
    mean_lon = mean(long, na.rm = TRUE),
    mean_lat = mean(lat, na.rm = TRUE),
    
    lon_range = max(long, na.rm = TRUE) - min(long, na.rm = TRUE),
    lat_range = max(lat, na.rm = TRUE) - min(lat, na.rm = TRUE),
    n_sites = n(),
    
    GoC = na.omit(GoC)[1], #keep one value for GoC
    
    .groups = "drop"
  ) %>%
  
  mutate(
    confidence_in_latlon = case_when(

      lon_range > 2 |
      lat_range > 2 ~
        "low; lat or lon estimates meaned at the state-site level differ by >2º in Mel's key",

      lon_range > 1 |
      lat_range > 1 ~
        "medium; lat or lon estimates meaned at the state-site level differ by 1-2º in Mel's key",

      lon_range > 0.5 |
      lat_range > 0.5 ~
        "medium-high; lat or lon estimates meaned at the state-site level differ by 0.5-1º in Mel's key",

      TRUE ~
        "high; high coherence among lat-lon estimates meaned at the state-site level"
    )
  )

# 858 obs., 9 vars.
```

```{r Load and clean Mel's Jalisco key}
# Load in Mel's Jalisco key:
Jalisco_LatLongKey<- read_csv(here("/home/frazier/ohi_goc_prep/FP/Exploratory_Research/FIS_TeamResearch/data/landing_sites_jalisco.csv"))
# 293 obs., 8 vars.

# Cleaning:
colnames(Jalisco_LatLongKey) <- c("landing_site", "landing_site_key", "est_habitat", "lat", "long", "site_notes", "OHI_rgn", "state")

# Normalize state and landing_site names:
Jalisco_LatLongKey<- Jalisco_LatLongKey %>%
  mutate(state = clean_names(state),
         landing_site_clean= clean_names(landing_site)) %>%
  relocate(state, .before= landing_site) %>%
  relocate(landing_site_clean, .after= landing_site) 
  
# Convert lat-long entries to numeric entries, accounting for unique cases where the minus (-) is actually an n dash or m dash:
Jalisco_LatLongKey_clean <- Jalisco_LatLongKey %>%
  mutate(
    long = as.numeric(str_replace_all(long, "[−–—]", "-")),
    lat  = as.numeric(str_replace_all(lat, "[−–—]", "-"))
  )

# Now assign mean values for compressed key:
Jalisco_LatLongKey_clean <- Jalisco_LatLongKey %>%
  
  mutate(
    long = as.numeric(str_replace_all(as.character(long), "[−–—]", "-")),
    lat  = as.numeric(str_replace_all(as.character(lat), "[−–—]", "-"))
  ) %>%
  
  filter(
    !is.na(landing_site_clean),
    !is.na(long),
    !is.na(lat)
  ) %>%
  
  group_by(state, landing_site_clean) %>%

  summarise(
    mean_lon = mean(long, na.rm = TRUE),
    mean_lat = mean(lat, na.rm = TRUE),
    lon_range = max(long, na.rm = TRUE) - min(long, na.rm = TRUE),
    lat_range = max(lat, na.rm = TRUE) - min(lat, na.rm = TRUE),
    n_sites = n(),
    
    est_habitat = na.omit(est_habitat)[1], #keep one value for est_habitat
    site_notes  = na.omit(site_notes)[1], #keep one value for site_notes
  
    .groups = "drop"
  ) %>%
  
  mutate(
    confidence_in_latlon = case_when(

      lon_range > 2 |
      lat_range > 2 ~
        "low; lat or lon estimates meaned at the state-site level differ by >2º in Mel's key",

      lon_range > 1 |
      lat_range > 1 ~
        "medium; lat or lon estimates meaned at the state-site level differ by 1-2º in Mel's key",

      lon_range > 0.5 |
      lat_range > 0.5 ~
        "medium-high; lat or lon estimates meaned at the state-site level differ by 0.5-1º in Mel's key",

      TRUE ~
        "high; high coherence among lat-lon estimates meaned at the state-site level"
    )
  )

# 258 obs., 8 vars.

# Add the "GoC" helpful column here based on a latitude cut off for the Nayarit/Jalisco OHI region
# Region 9 starts at latitude= 20.40895:
Jalisco_LatLongKey_clean <- Jalisco_LatLongKey_clean %>%
  mutate(
    GoC = case_when(
      mean_lat < 20.40895 ~ "no",
      mean_lat > 20.40895 ~ "yes",
      TRUE ~ NA_character_
    )
  )
```

```{r Merge Mel's Baja + Sonora key with Jalisco key}
Mels_LatLong_Site.Key0 <- bind_rows(BajaSonora_LatLongKey_clean, Jalisco_LatLongKey_clean) %>%
  # remove the NA from this data frame as this will create a many-to-many matching issue:
  filter(!is.na(landing_site_clean)) %>%  
  mutate(mean_lon = as.numeric(mean_lon),
         mean_lat = as.numeric(mean_lat))
# 1116 obs., 11 vars.

# Take a look
unique(Mels_LatLong_Site.Key0$state)

# Remove unused data frames:
# rm(BajaSonora_LatLongKey, Jalisco_LatLongKey, removed_rows)

# Clean up / simplify the confidence column:
Mels_LatLong_Site.Key0 <- Mels_LatLong_Site.Key0 %>%
  mutate(
    lat.lon_confidence_simple =
      sub(";.*", "", confidence_in_latlon)
  )

table(Mels_LatLong_Site.Key0$lat.lon_confidence_simple)
       # high         low      medium medium-high 
       # 1093          12           5           6 
```

# STEP 5: Merge Hem's Site.Key and Mel's Site.Key ------------------------------
```{r Clean and prepare both keys for merging}
colnames(Hems_LatLong_Site.Key)
colnames(Mels_LatLong_Site.Key0)

# Shave down Mel's key:
Mels_LatLong_Site.Key<- Mels_LatLong_Site.Key0 %>%
  select(state, landing_site_clean,
         site_mean_lon= mean_lon, site_mean_lat= mean_lat, 
         site_lon_range= lon_range, site_lat_range= lat_range, n_sites= n_site.offices, 
         confidence_in_site_latlon= confidence_in_latlon, lat.lon_confidence_simple)
# 1116 obs., 9 variables

colnames(Hems_LatLong_Site.Key)
colnames(Mels_LatLong_Site.Key)

# Step 1: rename confidence so we can compare
Hems_LatLong_Site.Key <- Hems_LatLong_Site.Key %>%
  mutate(source = "hem") %>%
  rename(conf_hem = confidence_in_site_latlon)

Mels_LatLong_Site.Key <- Mels_LatLong_Site.Key %>%
  mutate(source = "mel") %>%
  rename(conf_mel = confidence_in_site_latlon)

# Step 2: join Mel onto Hems
integrated_LatLong_Site.Key0 <- bind_rows(
  Hems_LatLong_Site.Key,
  Mels_LatLong_Site.Key
)
# 26246 obs., 11 vars.

# Step 3: create a confidence ranking
confidence_rank <- c(
  "low" = 1,
  "medium" = 2,
  "medium-high" = 3,
  "high" = 4
)

integrated_LatLong_Site.Key0 <- integrated_LatLong_Site.Key0 %>%
  mutate(conf_rank = confidence_rank[lat.lon_confidence_simple])

colnames(integrated_LatLong_Site.Key0)

integrated_LatLong_Site.Key <- integrated_LatLong_Site.Key0 %>%
  select(state, landing_site_clean, source, conf_rank, 
         lat.lon_confidence_simple, site_mean_lon, site_mean_lat) %>%
  tidyr::pivot_wider(
    names_from = source,
    values_from = c(conf_rank, lat.lon_confidence_simple, site_mean_lon, site_mean_lat),
    names_sep = "_"
  )

# Step 4: decide which source wins:
# overwrite Hem's estimate ONLY when Mel's estimate is better
integrated_LatLong_Site.Key <- integrated_LatLong_Site.Key %>%
  mutate(
    hem_rank = coalesce(conf_rank_hem, 0),
    mel_rank = coalesce(conf_rank_mel, 0),

    mel_wins = mel_rank > hem_rank,

    chosen_source = if_else(mel_wins, "mel", "hem"),

    final_lat = case_when(
      mel_wins ~ site_mean_lat_mel,
      TRUE     ~ site_mean_lat_hem
    ),

    final_lon = case_when(
      mel_wins ~ site_mean_lon_mel,
      TRUE     ~ site_mean_lon_hem
    ),

    final_conf_rank = if_else(mel_wins, conf_rank_mel, conf_rank_hem),

    final_conf = if_else(
      mel_wins,
      lat.lon_confidence_simple_mel,
      lat.lon_confidence_simple_hem
    )
  )
# 25644 obs., 18 vars.

# SANITY CHECKS: does the number of times mel_wins == chosen_source
table(integrated_LatLong_Site.Key$mel_wins)
table(integrated_LatLong_Site.Key$chosen_source)
# Yes! Mel's key trumps or supplements hem's 641 times out of 25644 instances (2%)

# Add back the GOC details:
colnames(integrated_LatLong_Site.Key)
colnames(Mels_LatLong_Site.Key0)

# Shave down Mel's key A LOT:
added_details<- Mels_LatLong_Site.Key0 %>%
  select(state, landing_site_clean,
         GoC, est_habitat, site_notes
         )
l# 1116 obs., 5 variables

 integrated_LatLong_Site.Key<- integrated_LatLong_Site.Key %>%
   left_join(
    added_details,
    by = c("state", "landing_site_clean")
  )
# 25644 obs., 21 vars.

table(integrated_LatLong_Site.Key$GoC)
#  no yes 
# 567 548 
# Excellent!
```


# STEP 6: Create a backbone for a new "master" key... ------------------------------

```{r Create a backbone for my master key}
# Create a backbone to merge to...
colnames(Mex.Pacific_Landings0)

Mex.Pacific_Landings<- Mex.Pacific_Landings0 %>%
  mutate(landing_site_clean= clean_names(landing_site),
         office_name_clean= clean_names(office_name)) %>%
  relocate(landing_site_clean, .after= landing_site) %>%
  relocate(office_name_clean, .after= office_name)
# 6406303 obs., 23 vars.
  
# Create skeleton:
LatLong_merge.skeleton<- Mex.Pacific_Landings %>%
  select(state, office_name_clean, landing_site_clean) %>%
  distinct()
# 5404 distinct state-office-site combos in the Mexican Pacific that I am going to try to fill using information from Hem's keys and / or Mel's key.
```

```{r Join in information from keys sequentially to create LatLong_Master.Key}
# if LatLong_merge.skeleton has state, office, and landing_site merge in lat and long from Hem_LatLong_
# Site.Office.Key
# if LatLong_merge.skeleton has only state and landing_site merge in lat and long from  integrated_LatLong_
# Site.Key
# if LatLong_merge.skeleton has only state and office merge in lat and long from  Hem_LatLong_Office.Key

LatLong_Master.Key <- LatLong_merge.skeleton %>%
  
  # OPTION 1) most specific match: state + office + landing_site
  left_join(
    Hems_LatLong_Site.Office.Key %>%
      select(state, office_name_clean, landing_site_clean,
             lat_site.office = site.office_mean_lat,
             lon_site.office = site.office_mean_lon),
    by = c("state", "office_name_clean", "landing_site_clean")
  ) %>%
  
  # OPTION 2) fallback: state + landing_site
  left_join(
    integrated_LatLong_Site.Key %>%
      select(state, landing_site_clean,
             lat_site = final_lat,
             lon_site = final_lon),
    by = c("state", "landing_site_clean")
  ) %>%
  
  # OPTION 3) fallback: state + office
  left_join(
    Hems_LatLong_Office.Key %>%
      select(state, office_name_clean,
             lat_office = office_mean_lat,
             lon_office = office_mean_lon),
    by = c("state", "office_name_clean")
  )

LatLong_Master.Key <- LatLong_Master.Key %>%
  mutate(
    final_lat = coalesce(lat_site.office, lat_site, lat_office),
    final_lon = coalesce(lon_site.office, lon_site, lon_office)
  )
# 5404 obs., 11 vars.

# SANITY CHECK: In which cases is the the final_lat pulled from office?
TEST<- LatLong_Master.Key %>%
  filter(final_lat== lat_office) %>%
  slice_sample(n= 200)
```

# WE ARE HERE - May 22 2026!!!!!!!

## STEP 6B: Cleaning

```{r Make sure to compress all keys before merging}
dim(Hems_LatLong_Site.Office.Key) # 30276 obs., 9 variables
dim(Hems_LatLong_Site.Key) # 25130 obs., 9 variables
dim(Hems_LatLong_Office.Key) # 73 obs., 9 variables
dim(Mels_LatLong_Site.Key0) # 1116 obs., 12 variables

# Compress: LatLong_merge.skeleton
# colnames(LatLong_merge.skeleton)
 
# LatLong_merge.skeleton_compr<- LatLong_merge.skeleton %>%
#   select(state, office_name_clean, landing_site_clean) %>%
#   group_by(state, office_name_clean, landing_site_clean) %>%
#   slice(1) %>%  # keeps the first occurrence per key
#   ungroup()
# # From 5404 obs down to 5404

# # Compress: Hems keys
# colnames(Hems_LatLong_Site.Office.Key)

# Hems_LatLong_Site.Office.Key_compr<- Hems_LatLong_Site.Office.Key %>%
#   group_by(state, landing_site_clean, office_name_clean) %>%
#   slice(1) %>%  # keeps the first occurrence per key
#   ungroup()
# From 30276 obs down to 30276

# Hems_LatLong_Site.Key_compr<- Hems_LatLong_Site.Key %>%
#   group_by(state, landing_site_clean) %>%
#   slice(1) %>%  # keeps the first occurrence per key
#   ungroup()
# # Down from 25130 obs to 25130

# Hems_LatLong_Office.Key_compr<- Hems_LatLong_Office.Key %>%
#   group_by(state, office_name_clean) %>%
#   slice(1) %>%  # keeps the first occurrence per key
#   ungroup()
# # Down from 73 obs to 73

# Compress: Mels_LatLong_Site.Key
# colnames(Mels_LatLong_Site.Key)
# 
# Mels_LatLong_Site.Key_compr<- Mels_LatLong_Site.Key %>%
#   group_by(state, landing_site_clean) %>%
#   slice(1) %>%  # keeps the first occurrence per key
#   ungroup()
# Down from 1116 obs to 1116

# Compression has been completed by now.
rm(LatLong_merge.skeleton_compr,
   Hems_LatLong_Site.Office.Key_compr, 
   Hems_LatLong_Site.Key_compr, 
   Hems_LatLong_Office.Key_compr, 
   Mels_LatLong_Site.Key_compr)
```


# STEP 7: Generate "master" key ------------------------------

```{r Merge to the backbone}
# LOGIC (May 19 2026):
# Start with the skeleton
# Privilege Hem’s key over Mel’s, unless Hem's key is empty or has high disagreement:

LatLong_SiteKey <- LatLong_merge.skeleton %>%
  
  left_join(Mels_LatLong_Site.Key, 
            by = c("state", "landing_site_clean")) %>%
  
  left_join(
    Hems_LatLongKey %>%
      select(state, landing_site_clean, Hems_lon_range, Hems_lat_range, Hems_est_lat, Hems_est_lon, notes_re_HemsKey),
    by = c("state", "landing_site_clean")
  ) %>%
  
  mutate(
    
    # Availability of Hem's key, yes or no?
    has_Hem = !is.na(Hems_est_lat) & !is.na(Hems_est_lon),
    
    # Disagreement in Hem's key?
    high_disagreement = (Hems_lon_range > 2 | Hems_lat_range > 2),

    # lat-lon difference between Hem and Mel's keys (flag high disagreement):
    keys_lat_diff = if_else(has_Hem & !is.na(lat_Mel),
                       abs(Hems_est_lat - lat_Mel), NA_real_),

    keys_lon_diff = if_else(has_Hem & !is.na(lon_Mel),
                       abs(Hems_est_lon - lon_Mel), NA_real_),
    
    # Assign lat-long values:
    # final coordinates: Hem's estimates unless high disagreement, fallback to Mel's estimates
    
    lat_est = case_when(
      has_Hem & is.na(lat_Mel) ~ Hems_est_lat,
      has_Hem & (!is.na(lat_Mel)) & high_disagreement ~ lat_Mel,
      has_Hem & (!is.na(lat_Mel)) ~ Hems_est_lat,
      !has_Hem & !is.na(lat_Mel) ~ lat_Mel,
      TRUE ~ NA_real_
      ),
    
    lon_est = case_when(
      has_Hem & is.na(lon_Mel) ~ Hems_est_lon,
      has_Hem & (!is.na(lon_Mel)) & high_disagreement ~ lon_Mel,
      has_Hem & (!is.na(lon_Mel)) ~ Hems_est_lon,
      !has_Hem & !is.na(lon_Mel) ~ lon_Mel,
      TRUE ~ NA_real_
      ),
    
    # Assign confidence values:
    lat_long_confidence = case_when(
      
      has_Hem & (!is.na(lat_Mel)) & high_disagreement ~
        "medium confidence; Mel's estimate; found in both keys but high disagreement in Hem's",

      has_Hem & (!is.na(lat_Mel)) ~
        "high confidence; Hem's estimate; Hem and Mel's keys agree",
      
      has_Hem & is.na(lat_Mel) ~
        "high confidence; Hem's estimate; found in Hem's key",

      has_Hem & (!is.na(lat_Mel)) &
        (keys_lat_diff > 2 | keys_lon_diff > 2) ~
        "check required; Hem's estimate; Hem and Mel's keys disagree by >2º",

      !has_Hem & !is.na(lat_Mel) ~
        "medium confidence; Mel's estimate; found only in Mel's key",

      TRUE ~ NA_character_
    )
  ) %>%
  relocate(lat_est, lon_est, .after = last_col())

# 3038 obs.

table(LatLong_SiteKey$lat_long_confidence)
#                                  high confidence; Hem's estimate; found in Hem's key 
#                                                                                  351 
#                            high confidence; Hem's estimate; Hem and Mel's keys agree 
#                                                                                  501 
# medium confidence; Mel's estimate; found in both keys but high disagreement in Hem's 
#                                                                                   78 
#                           medium confidence; Mel's estimate; found only in Mel's key 
#                                                                                  571 

table(LatLong_SiteKey$GoC)

# SANITY CHECK: how many lat-long values are missing from the states of interest?
TEST<- LatLong_SiteKey %>%
  filter(is.na(lat_est)) %>%
  slice_sample(n= 200)

# 530 sites in the States_of_Interest are missing lat-long values (down from 538 thanks to name cleaning!)
```







```{r Merge LatLong_SiteKey to NW.Mex_Landings}
# Merge key based on landing_site_clean names:
NW.Mex_Landings <- NW.Mex_Landings %>%
  left_join(LatLong_SiteKey, by = c("state", "landing_site_clean"))
# 4713144 obs.

# SANITY CHECK: how many records are missing lat-long values?
Landings_missingLatLong<- NW.Mex_Landings %>%
  filter(is.na(lat_est))
# 1/4 of records (n= 1,213,865) are not assigned... 
# May 20: Down from 1213865 to 1205911 (-7954) thanks to name cleaning!
# This doesn't really work for me.
# Can we make this key better?

# table(Landings_missingLatLong$landing_site)

# Let's take a look at a site we KNOW should be assigned like "BAHIA DE KINO" but isn't:
TEST2<- Landings_missingLatLong %>%
  filter(landing_site=="BAHIA DE KINO")
```


# STEP 8: Generate "integrated" LatLong_OfficeKey ------------------------------

In situations where we can't merge by state-landing_site let's merge by state-office

```{r Key skeleton}
colnames(Hems_LatLongKey_clean)

LatLong_OfficeKey0<- Hems_LatLongKey_clean %>%
  select(state, office_name_clean, Hems_est_lat= lat_deci, Hems_est_lon= lon_deci) %>%
  group_by(state, office_name_clean) %>%
  mutate(
    lat_mean = mean(Hems_est_lat , na.rm= TRUE),
    lon_mean = mean(Hems_est_lon , na.rm= TRUE)
  )
# 36486 obs., 6 vars.

# Reorganize this key:
colnames(LatLong_OfficeKey0)

LatLong_OfficeKey<- LatLong_OfficeKey0 %>%
  select(state, office_name_clean, est_lat_mean= lat_mean, est_lon_mean= lon_mean) %>%
  distinct()
# Down to 73 obs., 4 vars.
```

```{r Create LatLong_OfficeKey}
# NOTE: CLEAN NAMES function found above

# Match based on state-office_name_clean:

Landing_missing.matched <- Landings_missingLatLong %>%
  left_join(
    LatLong_OfficeKey,
    by = c("state", "office_name_clean")
  )

# Then find any values that are unmatched to a lat-long. Is this a small enough number I can live with?
unmatched <- Landing_missing.matched %>%
  filter(is.na(est_lat_mean))
# Only 241371 out of 4713144 un-matched (= 5%). I can live with this!

table(unmatched$state)
```

```{r Note from May 20 2026}
LatLong_OfficeKey$lat_long_confidence<- c("medium; Hem's estimate; estimate derived as a mean at the office level across sites")
```


# STEP 9: Add OHI region to the integrated key ------------------------------

```{r Loading in OHI regions and buffer shape files}
# OHI regions:
OHI_regions_shape <- sf::st_read(
  "/home/shares/ohi/OHI_GOC/spatial/OHI_regions/marine/ohi_goc_final_071025.shp",
  options = "ENCODING=LATIN1"
)

OHI_regions_shape <- st_make_valid(OHI_regions_shape)

# Quick plot:
region_plot<- ggplot(OHI_regions_shape) +
  geom_sf(aes(fill = rgn_id)) +
  theme_bw()

region_plot

# 50km buffer:
OHI_buffer_shape <- sf::st_read(
  "/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer/inland_50km_goc.shp",
  options = "ENCODING=LATIN1"
)

OHI_buffer_shape <- st_make_valid(OHI_buffer_shape)

# Quick plot:
buffer_plot<- ggplot(OHI_buffer_shape) +
  geom_sf(aes(fill = rgn_id)) +
  theme_bw()

buffer_plot
```

```{r Spatial join: assign OHI region to LatLong_OfficeKey based on these shapes}
# A few important shape checks:
st_crs(LatLong_OfficeKey_sf)
st_crs(OHI_buffer_shape)

st_is_valid(OHI_regions_shape)
OHI_regions_shape <- st_make_valid(OHI_regions_shape)

st_is_valid(OHI_buffer_shape)
OHI_buffer_shape  <- st_make_valid(OHI_buffer_shape)

sf::sf_use_s2(FALSE)

# Convert office lat-longs to shapefile points
LatLong_OfficeKey_sf <- st_as_sf(
  LatLong_OfficeKey,
  coords = c("est_lon_mean", "est_lat_mean"),
  crs = 4326
)

# Optional ID for stable joins
LatLong_OfficeKey_sf$office_id <- seq_len(nrow(LatLong_OfficeKey_sf))

# Joins
regions_join <- st_join(
  LatLong_OfficeKey_sf,
  OHI_regions_shape,
  join = st_within
)

LatLong_OfficeKey_sf <- st_transform(
  LatLong_OfficeKey_sf,
  st_crs(OHI_buffer_shape)
)

buffer_join <- st_join(
  LatLong_OfficeKey_sf,
  OHI_buffer_shape,
  join = st_within
)

# Combine outputs:
final_join <- regions_join %>%
  mutate(
    buffer_rgn_id = buffer_join$rgn_id,
    location      = buffer_join$location,
    estado        = buffer_join$estado,
    ecorgn        = buffer_join$ecorgn
  )
```

```{r Do the same thing for LatLong_SiteKey_final}
LatLong_SiteKey_final<- LatLong_SiteKey %>%
  filter(!is.na(lat_est))

# Convert office lat-longs to shapefile points
LatLong_SiteKey_sf <- st_as_sf(
  LatLong_SiteKey_final,
  coords = c("lon_est", "lat_est"),
  crs = 4326
)

# Optional ID for stable joins
LatLong_SiteKey_sf$office_id <- seq_len(nrow(LatLong_SiteKey_sf))

# Joins
regions_join2 <- st_join(
  LatLong_SiteKey_sf,
  OHI_regions_shape,
  join = st_within
)

LatLong_SiteKey_sf <- st_transform(
  LatLong_SiteKey_sf,
  st_crs(OHI_buffer_shape)
)

buffer_join2 <- st_join(
  LatLong_SiteKey_sf,
  OHI_buffer_shape,
  join = st_within
)

# Combine outputs:
final_join2 <- regions_join2 %>%
  mutate(
    buffer_rgn_id = buffer_join2$rgn_id,
    location      = buffer_join2$location,
    estado        = buffer_join2$estado,
    ecorgn        = buffer_join2$ecorgn
  )

final_join2<- final_join2 %>%
  relocate(GoC, .before= rgn_id)

final_join2 <- final_join2 %>%
  mutate(
    region_est = case_when(
      !is.na(rgn_id) ~ rgn_id,
      !is.na(buffer_rgn_id) ~ buffer_rgn_id,
      TRUE ~ NA_character_
    )
  )

TEST<- final_join2 %>%
  filter(is.na(region_est))

table(TEST$state)
# Very few entries in states of Nayarit, Sinaloa, and Sonora. This is a good thing. It suggests that our lat longs are getting us pretty close. The values listed in Baja California, BCS, and Jalisco could be real.
    # BAJA CALIFORNIA BAJA CALIFORNIA SUR             JALISCO             NAYARIT             SINALOA              SONORA 
    #             214                 261                 218                  23                  33                  15 

table(TEST$GoC)
#  no yes 
# 526  23 
# More evidence these entries are real.
```

NOTE TO SELF: Tomorrow morning, come up with another flag system that flags things inside or outside of the GOC based on the state. Then, if inside the GOC, interpolate most likely region. If not inside the GOC (unknown) then exclude.
UNDER CONSTRUCTION, May 21 2026


# STEP 10: Export key(s) ------------------------------
```{r Clean up key before export}
# Final sites key:
LatLong_SiteKey_finalized<- LatLong_SiteKey

colnames(LatLong_SiteKey_finalized)

LatLong_SiteKey_finalized<- LatLong_SiteKey_finalized %>% 
  select(
    state, landing_site, GoC, est_habitat,
    lat_est, lon_est, lat_long_confidence,
    Hems_est_lat, Hems_est_lon, Hems_est_lat, Hems_est_lon,
    lat_Mel, lon_Mel,
    OHI_region_draft, site_notes_Mel
  )

colnames(LatLong_SiteKey_finalized)<- c(
  "state", "landing_site", "inside_GOC", "est_habitat",
  "lat_est", "lon_est", "latlon_est_flag",
  "HemKey_lat", "HemKey_lon", "HemPermitKey_lat", "HemPermitKey_lon",
  "MelKey_lat", "MelKey_lon",
  "OHI_region_draft", "site_notes_Mel"
)

LatLong_SiteKey_finalized <- LatLong_SiteKey_finalized %>%
  mutate(
    latlon_est_flag = if_else(
      is.na(latlon_est_flag),
      "alert 5: unable to assign lat-long based on state and landing site",
      NA_character_
    )
  )

sum(is.na(LatLong_SiteKey_finalized$lat_est))

# How many / which values are missing?
subset<- LatLong_SiteKey_finalized %>% 
  filter(is.na(lat_est))

LatLong_SiteKey_finalized<- LatLong_SiteKey_finalized %>%
  filter(is.na(lat_est))
# 1508 sites with lat-long details

# Final offices key:
LatLong_OfficeKey_finalized<- LatLong_OfficeKey
```


```{r PURG? remove unused dataframes}
rm(
  subset,
  BajaSonora_LatLongKey_clean, Jalisco_LatLongKey_clean,
  LatLong_merge.skeleton, LatLong_SiteKey,
  Mels_LatLong_Site.Key0, Mels_LatLong_Site.Key
)

rm(n_unique_sites, threshold_alert, threshold_low, threshold_medium)
```





# PURG ??? - May 21 2026

# STEP X: Add lat-long and OHI_region info to landings data ------------------------------
```{r OHI LatLong_SiteKey_finalized and RegionKey_pending}
LatLong_SiteKey_finalized<- read_csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/Sites_LatLongKey_May192026.csv")

RegionKey_pending<- read_csv(here("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/Spatial_Keys/RegionKey_UNDERCONSTRUCTION_Mar102026.csv"))

colnames(RegionKey_pending)
colnames(RegionKey_pending) <- c("state", "OHI_region_draft", "OHI_rgn_number", "OHI_rgn_name", "exclude_rgn", "OHI_rgn_notes")
```

```{r Append lat-long and OHI_region info to landings data}
temp_df <- Mex.Pacific_Landings0 %>%
  left_join(
    LatLong_SiteKey_finalized,
    by = c("state", "landing_site")
  )

table(temp_df$inside_GOC)

temp_df <- temp_df %>%
  left_join(
    RegionKey_pending,
    by = c("state", "OHI_region_draft")
  )

table(temp_df$inside_GOC)

colnames(temp_df)

test<- temp_df %>%
  filter(exclude_rgn=="exclude")

table(test$inside_GOC)
table(test$state)
table(test$OHI_rgn_number)

Mex.Pacific_Landings0<- temp_df

rm(temp_df, test)
```

## SAVE: the NW Mexico landings appended with lat-long info in a separate .rds file:
```{r Save the Northwest Mexico dataframe: May 19 2026}
Mex.Pacific_Landings0<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/Mex.Pacific_Landings0_2000.present_May042026.rds"))
```


# STEP Y: Attach lat-long data to NW Mexico dataframe. (co-authors: Melanie Frazier, Hem Nalini Morzaria) ----------------------

## Flag these in the lat-long key as not in the Gulf [UNDER CONSTRUCTION May 19 2026]
office_name== c("EL ROSARIO", "ENSENADA", "TIJUANA")

# STEP Z: Attach lat-long and OHI_region info based on official landing site:
NOTE TO SELF: Come back to this! The lat long keys are imperfect. We'll want to fix this as much as possible.

## NEXT STEPS: Use of the lat-long key can be found in "CONAPESCA_CleaningWorkflow_May2026.Rmd"
found here: /home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys
