################################################################################ 

For questions contact: Erica Ferrer, Ph.D. [emferrer\@ucsd.edu](mailto:emferrer@ucsd.edu){.email} - OR - [ferrer\@nceas.ucsb.edu](mailto:ferrer@nceas.ucsb.edu){.email}

################################################################################ 

# PRE-WORK ------------------------------

```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```

# STEP 1: Load in CONAPESCA RAW data ------------------------------

This step is completed in the file shared by Juan Carlos: "03_combine_landings_edited.R" file path: /home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/JC_Scripts_EMF.edited/03_combine_landings_edited.R

# STEP 2: Work with CONAPESCA STITCHED data ------------------------------

A stitched version of these data is available as one single file: "AllMex_landings_2000_present_EMF_edited.rds" file path: /home/shares/ohi/OHI_GOC/goal_prep/fp/juan_carlos/AllMex_landings_2000_present_EMF_edited.rds

```{r Load in CONAPESCA STITCHED data}
MexLandings_2000.present<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/juan_carlos/AllMex_landings_2000_present_EMF_edited.rds"))
# n = 12153879 obs., 20 vars.
```

# STEP 3: Create weight_kg_corrected column ------------------------------

```{r Landed weight correction}
# Kind of looks like sometimes 'landed weight' is 'live weight' and vice versa. Fix this:
MexLandings_2000.present$weight_kg_corrected <- ifelse(
  MexLandings_2000.present$live_weight >= MexLandings_2000.present$landed_weight,
  MexLandings_2000.present$live_weight,
  MexLandings_2000.present$landed_weight
)
# n = 12153879 obs., 21 vars.
```

# STEP 4a: Filter down to Northwest Mexico data ------------------------------

```{r Filter out data from Northwest Mexico}
length(unique(MexLandings_2000.present$species_name)) 
# 2052 unique species in all of Mexico.
# But, we don't need species from ALL of Mexico, only the GoC region for now - so let's filter by state.

States_of_Interest<- c("BAJA CALIFORNIA", "BAJA CALIFORNIA SUR", "SONORA", "SINALOA", "NAYARIT", "JALISCO")

NW.Mex_Landings <- MexLandings_2000.present %>% filter(state %in% States_of_Interest)
# n = 4713144 obs., 21 vars.

length(unique(NW.Mex_Landings$species_name)) 
# 1827 unique species in States of Interest.
```

# STEP 4b: Attach lat-long and OHI_region info based on landing site ------------------------------
NOTE TO SELF: Come back to this! The lat long keys are imperfect. We'll want to fix this as much as possible.
```{r Append lat-long data to sites in Baja BCS Sonora and Sinaloa}
# Filter OUT sites from Baja that are on the Pacific side using lat-Long file from Mel's repository:
# NOTE TO SELF: THIS PART IS SUBJECT TO CHANGE once we integrate lat-long data from Hem. Mar 10 2026
BajaSonora_LatLong_Key<- read_csv("/home/frazier/ohi_goc_prep/FP/FIS/v2025/data/Baja_Sonora_states_key.csv")
colnames(BajaSonora_LatLong_Key) <- c("state", "landing_site", "OHI_rgn", "lat", "long", "GoC")

Jalisco_LatLong_Key<- read_csv(here("/home/frazier/ohi_goc_prep/FP/FIS/v2025/data/landing_sites_jalisco.csv"))
colnames(Jalisco_LatLong_Key) <- c("landing_site", "CONAPESCA_landing_key", "habitat", "lat", "long", "site_notes", "OHI_rgn", "state")


#### CLEANING LAT-LONG KEYS: ####
# Clean the Baja-Sonora key...
BajaSonora_LatLong_Key_clean <- BajaSonora_LatLong_Key %>%
  
  # Normalize state names:
  mutate(state = toupper(state)) %>%
  
  # Remove any rows where landing_site is NA (NAs cause matching issues later on):
  filter(!is.na(landing_site)) %>%
  
  # Remove any direct duplicates of landing_site:
  group_by(state, landing_site) %>%
  slice(1) %>%  # keeps the first occurrence per key
  ungroup()

# Look into the rows that were removed in the de duplication of BajaSonora_LatLong_Key...
removed_rows <- anti_join(BajaSonora_LatLong_Key, BajaSonora_LatLong_Key_clean, 
                          by = c("landing_site"))

# Convert lat-long entries to numeric entries, accounting for unique cases where the minus (-) is actually an n dash or m dash:
BajaSonora_LatLong_Key_clean <- BajaSonora_LatLong_Key_clean %>%
  mutate(
    long = as.numeric(str_replace_all(long, "[−–—]", "-")),
    lat  = as.numeric(str_replace_all(lat, "[−–—]", "-"))
  )

# Clean the Jalisco key...
Jalisco_LatLong_Key_clean <- Jalisco_LatLong_Key %>%
  # Normalize state names
  mutate(state = toupper(state)) %>%
  
  # Remove any rows where landing_site is NA (NAs cause matching issues later on):
  filter(!is.na(landing_site)) %>%
  
  # Keep only one row per CONAPESCA_landing_key; if duplicate keys exist, keep first
  group_by(CONAPESCA_landing_key) %>%
  slice(1) %>%  # keeps the first occurrence per key
  ungroup() %>%
  
  # Optional: remove any direct duplicates of landing_site if there are rows with NA in the CONAPESCA_landing_key
  group_by(state, landing_site) %>%
  slice(1) %>%  # keeps the first occurrence per key
  ungroup()

# Look into the rows that were removed in the de duplication of Jalisco_LatLong_Key...
removed_rows <- anti_join(Jalisco_LatLong_Key, Jalisco_LatLong_Key_clean, 
                          by = c("CONAPESCA_landing_key", "landing_site"))

# Convert lat-long entries to numeric entries, accounting for unique cases where the minus (-) is actually an n dash or m dash:
Jalisco_LatLong_Key_clean <- Jalisco_LatLong_Key_clean %>%
  mutate(
    long = as.numeric(str_replace_all(long, "[−–—]", "-")),
    lat  = as.numeric(str_replace_all(lat, "[−–—]", "-"))
  )

# Add the "GoC" helpful column here based on a latitude cut off for the Nayarit/Jalisco OHI region, region 9 starting at 20.40895:
Jalisco_LatLong_Key_clean <- Jalisco_LatLong_Key_clean %>%
  mutate(GoC = if_else(lat < 20.40895, "no", NA_character_))

#### MERGE LAT-LONG KEYS: ####
LatLong_MasterKey_pending <- bind_rows(BajaSonora_LatLong_Key_clean, Jalisco_LatLong_Key_clean) %>%
  filter(!is.na(landing_site)) %>%  # remove the NA from this data frame as this will create a many-to-many matching issue later on
  mutate(long = as.numeric(long))
# 1176 obs., 9 vars.

# Remove unused data frames:
rm(BajaSonora_LatLong_Key, Jalisco_LatLong_Key, removed_rows)

# Take a look
unique(LatLong_MasterKey_pending$state)
unique(LatLong_MasterKey_pending$OHI_rgn)

# My regional key (temporary, subject to change, UNDER CONSTRUCTION):
RegionKey_pending<- read_csv(here("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/RegionKey_UNDERCONSTRUCTION_Mar102026.csv"))
# Previously: RegionKey_UNDERCONSTRUCTION_Dec012025.csv
colnames(RegionKey_pending) <- c("state", "OHI_rgn", "OHI_rgn_number", "OHI_rgn_name", "exclude_rgn", "exclude_notes")
```

```{r Append lat-long and OHI_region info to landings data}
# Append lat-long info to landings data:
temp_df <- NW.Mex_Landings %>%
  left_join(
    LatLong_MasterKey_pending,
    by = c("state", "landing_site")  # matching columns
  )

table(temp_df$GoC)
# 4713144 obs., 28 vars.
# 1090368 "no" not in the GOC.

# Append RegionKey_pending info to landings data:
temp_df <- temp_df %>%
  left_join(
    RegionKey_pending,
    by = c("state", "OHI_rgn")  # matching columns
  )
# 4713144 obs., 32 vars.
table(temp_df$GoC)
# 1090368 "no" not in the GOC as we'd expect!

NW.Mex_Landings<- temp_df
# 4713144 obs., 32 vars.

# Delete temp_df, not needed anymore.
rm(temp_df)
```

# STEP 4c: Save the NW Mexico landings with lat-long info as an rds file
```{r Save the Northwest Mexico dataframe: Mar 11 2026}
saveRDS(GOC_Landings, file="/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/NWMex_Landings_2000.present_EMF.Mar112026.rds")
# 4713144 obs., 32 vars.
```

# STEP 5: Filter down to GOC data ------------------------------
```{r filter out GOC data}
# Generate filtered landings data for the GoC:
GOC_Landings <- NW.Mex_Landings %>%
  filter(GoC != "no" | is.na(GoC))
# 3622776 obs., 32 vars.

# Check that this all makes sense:
length(unique(GOC_Landings$species_name))
length(unique(GOC_Landings$main_species_group))
# 1703 unique species from 57 main species groups in the Gulf of California, keeping in mind that these are not really SPECIES (per se) but rather log-book entries for species.
```
NOW 1703 (previously 1693) unique species in the Gulf of California according to Juan Carlos' version of the CONAPESCA data base, keeping in mind that these are not really SPECIES, per se, but rather log-book entries for species.

## Save the GOC landings rds file
```{r Save the GOC_Landings dataframe: Mar 11 2026}
saveRDS(GOC_Landings, file="/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Landings_2000.present_EMF.Mar112026.rds")
# 3622776 obs., 32 vars.
# Previous iteration: GOC_FIS.landings_CLEAN_as.of.Dec092025.rds
```

STOPPED HERE

# PART 4: Merge the cleaned master sp key with landings data ------------------------------

```{r Load data frames if needed}
Master_SpKey_Cleaned <- read_csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Data&Keys/CONAPESCA_Keys/Species_Keys/CleanedMaster_SpKey_Dec012025.csv")
# 802 obs of 41 variables
# Was previously found here: "/home/ferrer/ohi_goc_prep/general/Species_Keys/CleanedMaster_SpKey_Dec012025.csv"

GOC_Landings_Cleaned<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fis/v2025/int/juan_carlos/GOC_Landings_wKeyWords_2000.present_Dec012025.rds"))
# n = 3716415 obs.
```

```{r Shave down the size of Master_SpKey_filtered}
Master_SpKey_filtered<- Master_SpKey_Cleaned %>% 
  select(recorded_taxa_group, recorded_taxa_name_clean, common_name_clean,
         estimated_species_name, species_est_certainty, estimated_genus_name, estimated_family_name,
         IHH_order, IHH_class, IHH_functional_major_groups_sm, IHH_fishery)
# 802 obs., 11 vars.
```

```{r Merge landings with filtered sp key}
GOC_Landings_Cleaned_wSp <- GOC_Landings_Cleaned %>%
  left_join(
    Master_SpKey_filtered,
    by = c(
      "recorded_taxa_group",
      "recorded_taxa_name_clean",
      "common_name_clean"
    )
  )
# 3716415 obs. as expected, 41 vars.
```

## Save GOC_Landings_Cleaned_wSp rds file

```{r Save this dataframe - landings keywords & species for top groupings: Nov 10 2025}
saveRDS(GOC_Landings_Cleaned_wSp, file="/home/shares/ohi/OHI_GOC/goal_prep/fis/v2025/int/juan_carlos/GOC_Landings_wSp_2000.present_Dec012025.rds")
```
