################################################################################ 

For questions contact: Erica Ferrer, Ph.D. [emferrer\@ucsd.edu](mailto:emferrer@ucsd.edu){.email} - OR - [ferrer\@nceas.ucsb.edu](mailto:ferrer@nceas.ucsb.edu){.email}

################################################################################ 

What's different about this key from April's key?

~108 lines effectively unchanged (copy-paste from April).
4 lines minor comment/header edits.
23 lines deleted (STEP 15).
~51 lines added (STEP 3b2 + one reference line).
~95 lines rewritten (merge, export, 3c–3d, next steps).
~147 lines in May that don’t match April line-for-line (adds + rewrites + line shift from deletion).


# PRE-WORK ------------------------------

```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
library(fuzzyjoin)
library(sf) # for work on shape fines
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```

## clean_names function
```{r Load clean_names function}
# Takes text and standardizes it, which we will need for merging later on:

clean_names <- function(x) {

  x <- iconv(x, from = "", to = "UTF-8", sub = "")
  x <- stringi::stri_trans_general(x, "Latin-ASCII")

  x %>%
    toupper() %>%
    trimws() %>%
    gsub("\\bDE\\b", "", .) %>%
    gsub("\\s*\\([^\\)]*\\)", "", .) %>%   # remove parentheses + contents
    gsub("\\s+", " ", .) %>%
    trimws()
}
```


# FIRST TEST: How well does this revised key (Hems_LatLongKey_clean.compressed) work on its own, no alterations?

# STEP 1: Load in Hem's catch-location lat-long key (georeflocskm) ------------------------------

```{r Load and prepare georeflocskm}
load("/home/shares/ohi/OHI_GOC/_raw_data/CONAPESCA/d2025/SharedByHemMorzaria/permit_data/georeflocskm.rda")
Hems_LatLongKey0 <- georeflocskm
#rm(georeflocskm)
```

```{r Clean Hem's key}
# Check column names first:
colnames(Hems_LatLongKey0) <- c(
  "state", "landing_site", "geometry", "lon_deci", "lat_deci",
  "fishery_office", "fishery_office_ent", "distance", "distance_km"
)

# Clean the landing_site names:
Hems_LatLongKey_clean <- Hems_LatLongKey0 %>%
  mutate(landing_site_clean= clean_names(landing_site),
         office_name_clean= clean_names(fishery_office))

# Re-arrange columns:
colnames(Hems_LatLongKey_clean)

Hems_LatLongKey_clean<- Hems_LatLongKey_clean %>%
  select(state, landing_site, landing_site_clean, fishery_office, office_name_clean, lon_deci, lat_deci, geometry, distance, distance_km)
```

```{r Compress Hems_LatLongKey_clean}

Hems_LatLongKey_clean.compressed <- Hems_LatLongKey_clean %>%
  group_by(state, landing_site_clean, office_name_clean) %>%
  summarise(
    mean_lon = mean(lon_deci, na.rm = TRUE),
    mean_lat = mean(lat_deci, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  st_as_sf(
    coords = c("mean_lon", "mean_lat"),
    crs = 4326,
    remove= FALSE
  )

# 30276 obs., 6 vars.

# Re-name:
Hems_LatLongKey_clean.compressed <- Hems_LatLongKey_clean.compressed %>%
  rename(mean_geometry = geometry)

st_geometry(Hems_LatLongKey_clean.compressed) <- "mean_geometry"
# 30276 obs., 6 vars.
```


# STEP 2: Load in relevant fisheries landings data ------------------------------

```{r Load the Mexican Pacific - Mex.Pacific_Landings}
Mex.Pacific_Landings<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/Mex.Pacific_Landings_2000.present_May042026.rds"))

dim(Mex.Pacific_Landings)
# 6406303 obs., 21 vars.
```

```{r States_of_Interest}
unique(Mex.Pacific_Landings$state)

States_of_Interest<- c("BAJA CALIFORNIA", "BAJA CALIFORNIA SUR",
                       "SONORA", "SINALOA",
                       "NAYARIT", "JALISCO")
```

```{r Generate NW.Mex_Landings0 dataframe}
NW.Mex_Landings0<- Mex.Pacific_Landings %>%
  filter(state %in% States_of_Interest)

dim(NW.Mex_Landings0)
# 4713144 obs., 21 obs.
```

```{r Clean the landing_site names to create NW.Mex_Landings}
# Clean the landing_site names in NW.Mex_Landings:
NW.Mex_Landings<- NW.Mex_Landings0 %>%
  mutate(landing_site_clean= clean_names(landing_site),
         office_name_clean= clean_names(office_name)) %>%
  relocate(landing_site_clean, .after= landing_site) %>%
  relocate(office_name_clean, .after= office_name)
# 4713144 obs., 23 obs.
```


# STEP 3A: Test how good is Hems_LatLongKey_clean.compressed for matching with no additional edits ------------------------------

```{r Test: how good is Hems_LatLongKey_clean.compressed for matching with no additional edits?}
# Does the extra work actually improve the matching? 

NW.Mex_Landings_matched1<- NW.Mex_Landings %>%
  left_join(
    Hems_LatLongKey_clean.compressed,
    by = c("state", "office_name_clean", "landing_site_clean")
  )

# 4713144 obs., 26 obs.
# Excellent! This worked.

# SANITY CHECK: How many entries are unmatched after the merge?
unmatched<- NW.Mex_Landings_matched1 %>%
  filter(is.na(mean_lat))
# 2607660 (55%) are unmatched, so this strategy alone does not work well.

# Let's try to understand sites didnt match:
unmatched_slice<- unmatched %>%
  slice_sample(n= 200)
# I see that there could be several entries captured with a simple state-landing_site_clean match.
```

# FIRST OUTCOME: Direct use of key does not work that well. Let's try a few simple slice, dices, and merges that may improve matching.


# STEP 3B: Split Hem's key down to the level of landing_site_clean to improve matching  ------------------------------

```{r Compress Hems_LatLongKey_clean.compressed even further to match only by state and landing_site_clean}
# Compress by state and landing_site_clean:
Hems_LatLongKey_site.compressed <- Hems_LatLongKey_clean.compressed %>%
  group_by(state, landing_site_clean) %>%
  summarise(
    mean_lon = mean(mean_lon, na.rm = TRUE),
    mean_lat = mean(mean_lat, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  st_as_sf(
    coords = c("mean_lon", "mean_lat"),
    crs = 4326,
    remove= FALSE
  )

#  25130 obs., 5 vars. (smaller than before, good!)
```

```{r Sequential second merge}
# Join using the more specific key first: Hems_LatLongKey_clean.compressed
# Join again using the fallback key: Hems_LatLongKey_site.compressed
# Only use fallback lat/long values where the first join did not populate them.

colnames(Hems_LatLongKey_clean.compressed)
colnames(Hems_LatLongKey_site.compressed)

NW.Mex_Landings_matched2 <- NW.Mex_Landings_matched1 %>%
  
  # Second merge: fallback key
  left_join(
    Hems_LatLongKey_site.compressed %>%
      select(state, landing_site_clean,
             mean_lon, mean_lat, mean_geometry) %>%
      distinct(),
    
    by = c("state", "landing_site_clean"),
    suffix = c("", "_fallback")
  ) %>%
  
  # Fill missing values from fallback merge
  mutate(
    est_lon = coalesce(mean_lon, mean_lon_fallback),
    est_lat = coalesce(mean_lat, mean_lat_fallback)
  )

# 4713144 obs., 31 obs. (nice!!)

# SANITY CHECK: How many entries are unmatched after the merge?
unmatched<- NW.Mex_Landings_matched2 %>%
  filter(is.na(est_lon))

dim(unmatched)
# 1444423 (31%) of entries are still unmatched.

# Let's try to understand sites didn't match:
unmatched<- unmatched %>%
  slice_sample(n= 200)
# I see that there could be several entries captured with a state-office match.
```


# STEP 3C: Split Hem's key down to the level of office_name)clean to improve matching ------------------------------

```{r Compress Hems_LatLongKey_clean.compressed even further to match only by state and office_name_clean}
# Compress by office_name_clean
Hems_LatLongKey_office.compressed <- Hems_LatLongKey_clean.compressed %>%
  group_by(state, office_name_clean) %>%
  summarise(
    mean_lon = mean(mean_lon, na.rm = TRUE),
    mean_lat = mean(mean_lat, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  st_as_sf(
    coords = c("mean_lon", "mean_lat"),
    crs = 4326,
    remove= FALSE
  )

#  73 obs., 5 vars. (smaller than before, good!)
```

```{r Sequential third merge}
# The idea is:
# Join using Hem's specific key first.
# Join using the fallback key ()
# Only use fallback lat/long values where the first join did not populate them.

colnames(Hems_LatLongKey_clean.compressed)
colnames(Hems_LatLongKey_site.compressed)
colnames(Hems_LatLongKey_office.compressed)

NW.Mex_Landings_matched3 <- NW.Mex_Landings_matched2 %>%
  
  # Third merge: second fallback key
  left_join(
    Hems_LatLongKey_office.compressed %>%
      select(state, office_name_clean,
             mean_lon, mean_lat, mean_geometry) %>%
      distinct(),
    
    by = c("state", "office_name_clean"),
    suffix = c("", "_fallback2")
  ) %>%
  
  # Fill missing values from fallback merge
  mutate(
    est_lon = coalesce(mean_lon, mean_lon_fallback, mean_lon_fallback2),
    est_lat = coalesce(mean_lat, mean_lat_fallback, mean_lat_fallback2)
  )

# 4713144 obs., 34 obs. (nice!!)

# SANITY CHECK: How many entries are unmatched after THIS merge?
unmatched<- NW.Mex_Landings_matched3 %>%
  filter(is.na(est_lon))
# Only 336383 (7%) of entries are still unmatched.

# Let's try to understand sites didn't match:
unmatched_slice<- unmatched %>%
  slice_sample(n= 200)
# 
```

# SECOND OUTCOME: This key works pretty well!


# STEP 4: Annotate information about hem's key ------------------------------

```{r Flag duplicate landing_site entries in Hem's key}
# Annotate Hem's key with alerts about disagreement and then compress the key

# Annotation thresholds:
threshold_alert  <- 2
threshold_low    <- 1
threshold_medium <- 0.5


Hems_LatLongKey_clean.compressed <- Hems_LatLongKey_clean %>%
  select(state, landing_site, fishery_office, lat_deci, lon_deci, distance_km) %>%
  
  group_by(state, landing_site) %>%
  
  summarise(
    
    Hems_lat_range = max(lat_deci, na.rm = TRUE) - min(lat_deci, na.rm = TRUE),
    Hems_lon_range = max(lon_deci, na.rm = TRUE) - min(lon_deci, na.rm = TRUE),
    
    Hems_est_lat = median(lat_deci, na.rm = TRUE),
    Hems_est_lon = median(lon_deci, na.rm = TRUE),
    
    notes_re_HemsKey = case_when(
      is.na(Hems_lat_range) | is.na(Hems_lon_range) ~
        "coordinate data not found in Hem's key",

      Hems_lat_range > threshold_alert | Hems_lon_range > threshold_alert ~
        "alert level 3: multiple state-landing_site matches in Hem's key with lat-long values >2 degrees apart",

      Hems_lat_range > threshold_low | Hems_lon_range > threshold_low ~
        "alert level 2: multiple state-landing_site matches in Hem's key with lat-long values ~1 degree apart",

      Hems_lat_range > threshold_medium | Hems_lon_range > threshold_medium ~
        "alert level 1: multiple state-landing_site matches in Hem's key with lat-long values ~0.5 degree apart",

      TRUE ~ "high coherence in Hem's key regarding state-landing_site location"
    ),

    .groups = "drop"
  )

# rm(Hems_LatLongKey_clean.compressed)

table(Hems_LatLongKey_clean.compressed$notes_re_HemsKey)

# alert level 1: multiple state-landing_site matches in Hem's key with lat-long values ~0.5 degree apart 
#                                                                                                    589 
#   alert level 2: multiple state-landing_site matches in Hem's key with lat-long values ~1 degree apart 
#                                                                                                   1072 
#  alert level 3: multiple state-landing_site matches in Hem's key with lat-long values >2 degrees apart 
#                                                                                                   1107 
#                                      high coherence in Hem's key regarding state-landing_site location 
#                                                                                                  24128  

# SANITY CHECK: which values really don't agree?
TEST<- Hems_LatLongKey_clean.compressed %>%
  filter(notes_re_HemsKey== "alert level 3: multiple state-landing_site matches in Hem's key with lat-long values >2 degrees apart")
# n=1107 - Basically, values that have different offices. I feel good enough about this to proceed though.

# SANITY CHECK: Number of unique state + landing_site combinations
n_unique_sites <- Hems_LatLongKey_clean.compressed %>%
  distinct(state, landing_site) %>%
  nrow()
# 26896 unique state-landing_site combinations.
```


# STEP 5: Add infomration from Mel's lat-long keys ------------------------------
Additional refinements to the key with Mel's key (only if needed)

```{r Append Mel's lat-long data to sites in Baja + BCS + Sonora & Jalisco}
# Filter OUT sites from Baja that are on the Pacific side using lat-Long file from Mel's repository:

# Integrated Hem georefcatchlocs (STEP 3) and georeflocskm permit/INEGI georef (STEP 3) - May 19 2026
BajaSonora_LatLongKey<- read_csv("/home/frazier/ohi_goc_prep/FP/Exploratory_Research/FIS_TeamResearch/data/Baja_Sonora_states_key.csv")
colnames(BajaSonora_LatLongKey) <- c("state", "landing_site", "OHI_rgn", "lat", "long", "GoC")

Jalisco_LatLongKey<- read_csv(here("/home/frazier/ohi_goc_prep/FP/Exploratory_Research/FIS_TeamResearch/data/landing_sites_jalisco.csv"))
colnames(Jalisco_LatLongKey) <- c("landing_site", "CONAPESCA_landing_key", "est_habitat", "lat", "long", "site_notes", "OHI_rgn", "state")


# Clean the Baja-Sonora key...
BajaSonora_LatLongKey_clean <- BajaSonora_LatLongKey %>%
  
  # Normalize state names:
  mutate(state = toupper(state)) %>%
  
  # Remove any rows where landing_site is NA (NAs cause matching issues later on):
  filter(!is.na(landing_site)) %>%
  
  # Remove any direct duplicates of landing_site:
  group_by(state, landing_site) %>%
  slice(1) %>%  # keeps the first occurrence per key
  ungroup()


# Look into the rows that were removed in the de duplication of BajaSonora_LatLongKey...
removed_rows <- anti_join(BajaSonora_LatLongKey, BajaSonora_LatLongKey_clean, 
                          by = c("landing_site"))

# Convert lat-long entries to numeric entries, accounting for unique cases where the minus (-) is actually an n dash or m dash:
BajaSonora_LatLongKey_clean <- BajaSonora_LatLongKey_clean %>%
  mutate(
    long = as.numeric(str_replace_all(long, "[−–—]", "-")),
    lat  = as.numeric(str_replace_all(lat, "[−–—]", "-"))
  )

# Clean the Jalisco key...
Jalisco_LatLongKey_clean <- Jalisco_LatLongKey %>%
  # Normalize state names
  mutate(state = toupper(state)) %>%
  
  # Remove any rows where landing_site is NA (NAs cause matching issues later on):
  filter(!is.na(landing_site)) %>%
  
  # Keep only one row per CONAPESCA_landing_key; if duplicate keys exist, keep first
  group_by(CONAPESCA_landing_key) %>%
  slice(1) %>%  # keeps the first occurrence per key
  ungroup() %>%
  
  # Optional: remove any direct duplicates of landing_site if there are rows with NA in the CONAPESCA_landing_key
  group_by(state, landing_site) %>%
  slice(1) %>%  # keeps the first occurrence per key
  ungroup()

# Look into the rows that were removed in the de-duplication of Jalisco_LatLongKey...
removed_rows <- anti_join(Jalisco_LatLongKey, Jalisco_LatLongKey_clean, 
                          by = c("CONAPESCA_landing_key", "landing_site"))

# Convert lat-long entries to numeric entries, accounting for unique cases where the minus (-) is actually an n dash or m dash:
Jalisco_LatLongKey_clean <- Jalisco_LatLongKey_clean %>%
  mutate(
    long = as.numeric(str_replace_all(long, "[−–—]", "-")),
    lat  = as.numeric(str_replace_all(lat, "[−–—]", "-"))
  )

# Add the "GoC" helpful column here based on a latitude cut off for the Nayarit/Jalisco OHI region, region 9 starting at 20.40895:
Jalisco_LatLongKey_clean <- Jalisco_LatLongKey_clean %>%
  mutate(GoC = if_else(lat < 20.40895, "no", NA_character_))
```

```{r Append Mel's lat-long data to sites in Jalisco Baja and Sonora key into one master key}
Mels_LatLongKey <- bind_rows(BajaSonora_LatLongKey_clean, Jalisco_LatLongKey_clean) %>%
  filter(!is.na(landing_site)) %>%  # remove the NA from this data frame as this will create a many-to-many matching issue later on
  mutate(long = as.numeric(long))
# 1176 obs., 9 vars.

# Remove unused data frames:
rm(BajaSonora_LatLongKey, Jalisco_LatLongKey, removed_rows)

# Take a look
unique(Mels_LatLongKey$state)
unique(Mels_LatLongKey$OHI_rgn)
```


# STEP 6: Create a backbone for a new "integrated" key... ------------------------------

```{r Prepare to merge keys}
dim(Hems_LatLongKey_clean.compressed) # 26896 obs., 7 variables
dim(Mels_LatLongKey) # 1176 obs., 9 variables

# Shave down Mel's key:
Mels_LatLongKey_simple<- Mels_LatLongKey %>%
  select(state, landing_site, lat, long, OHI_rgn, GoC, est_habitat, site_notes)

colnames(Mels_LatLongKey_simple)<- c("state", "landing_site", "lat_Mel", "lon_Mel", "OHI_region_draft", "GoC", "est_habitat", "site_notes_Mel")
# 1176 obs. of 8 variables

table(Mels_LatLongKey_simple$GoC)
# 400 sites in the GoC, 588 sites likely not in the GoC
```

```{r Create a backbone of a key to merge to...}
# Create a backbone to merge to...
colnames(Mex.Pacific_Landings)

LatLong_merge.skeleton<- Mex.Pacific_Landings %>%
  select(state, landing_site) %>%
  distinct()
# 3060 distinct state-landingsite combos in the Mexican Pacific that I am going to try to fill using information from Hem's key or Mel's key. Privilege information from Hem's key.
```


## STEP 6B: Cleaning

```{r Clean names for landing_site across LatLong_merge.skeleton}
# Clean the landing_site names:
LatLong_merge.skeleton<- LatLong_merge.skeleton %>%
  mutate(landing_site_clean= clean_names(landing_site)) %>%
  relocate(landing_site_clean, .after= landing_site)

Hems_LatLongKey_clean<- Hems_LatLongKey_clean %>%
  mutate(landing_site_clean= clean_names(landing_site),
         office_name_clean= clean_names(fishery_office)) %>%
  relocate(landing_site_clean, .after= landing_site) %>%
  relocate(office_name_clean, .after= fishery_office)

Hems_LatLongKey_clean.compressed<- Hems_LatLongKey_clean.compressed %>%
  mutate(landing_site_clean= clean_names(landing_site)) %>%
  relocate(landing_site_clean, .after= landing_site)

Mels_LatLongKey_simple<- Mels_LatLongKey_simple %>%
  mutate(landing_site_clean= clean_names(landing_site)) %>%
  relocate(landing_site_clean, .after= landing_site)
```

```{r Make sure to compress all keys after cleaning}
# Compress: LatLong_merge.skeleton
colnames(LatLong_merge.skeleton)

LatLong_merge.skeleton_compr<- LatLong_merge.skeleton %>%
  select(state, landing_site_clean) %>%
  group_by(state, landing_site_clean) %>%
  slice(1) %>%  # keeps the first occurrence per key
  ungroup()
# From 3060 obs down to 3038

# Compress: Hems_LatLongKey_clean.compressed
colnames(Hems_LatLongKey_clean.compressed)

Hems_LatLongKey_compr<- Hems_LatLongKey_clean.compressed %>%
  select(state, landing_site_clean, Hems_lat_range, Hems_lon_range, Hems_est_lat, Hems_est_lon, notes_re_HemsKey) %>%
  group_by(state, landing_site_clean) %>%
  slice(1) %>%  # keeps the first occurrence per key
  ungroup()
# From 26896 obs down to 26866

# Compress: Mels_LatLongKey_simple
colnames(Mels_LatLongKey_simple)

Mels_LatLongKey_compr<- Mels_LatLongKey_simple %>%
  select(state, landing_site_clean, lat_Mel, lon_Mel, GoC, est_habitat, site_notes_Mel) %>%
  group_by(state, landing_site_clean) %>%
  slice(1) %>%  # keeps the first occurrence per key
  ungroup()
# From 1176 obs to 1168
```


# STEP 7: Generate "integrated" LatLong_SiteKey ------------------------------

```{r Merge to the backbone}
# LOGIC (May 19 2026):
# Start with the skeleton
# Privilege Hem’s key over Mel’s, unless Hem's key is empty or has high disagreement:

LatLong_SiteKey <- LatLong_merge.skeleton_compr %>%
  
  left_join(Mels_LatLongKey_compr, by = c("state", "landing_site_clean")) %>%
  
  left_join(
    Hems_LatLongKey_compr %>%
      select(state, landing_site_clean, Hems_lon_range, Hems_lat_range, Hems_est_lat, Hems_est_lon, notes_re_HemsKey),
    by = c("state", "landing_site_clean")
  ) %>%
  
  mutate(
    
    # Availability of Hem's key, yes or no?
    has_Hem = !is.na(Hems_est_lat) & !is.na(Hems_est_lon),
    
    # Disagreement in Hem's key?
    high_disagreement = (Hems_lon_range > 2 | Hems_lat_range > 2),

    # lat-lon difference between Hem and Mel's keys (flag high disagreement):
    keys_lat_diff = if_else(has_Hem & !is.na(lat_Mel),
                       abs(Hems_est_lat - lat_Mel), NA_real_),

    keys_lon_diff = if_else(has_Hem & !is.na(lon_Mel),
                       abs(Hems_est_lon - lon_Mel), NA_real_),
    
    # Assign lat-long values:
    # final coordinates: Hem's estimates unless high disagreement, fallback to Mel's estimates
    
    lat_est = case_when(
      has_Hem & is.na(lat_Mel) ~ Hems_est_lat,
      has_Hem & (!is.na(lat_Mel)) & high_disagreement ~ lat_Mel,
      has_Hem & (!is.na(lat_Mel)) ~ Hems_est_lat,
      !has_Hem & !is.na(lat_Mel) ~ lat_Mel,
      TRUE ~ NA_real_
      ),
    
    lon_est = case_when(
      has_Hem & is.na(lon_Mel) ~ Hems_est_lon,
      has_Hem & (!is.na(lon_Mel)) & high_disagreement ~ lon_Mel,
      has_Hem & (!is.na(lon_Mel)) ~ Hems_est_lon,
      !has_Hem & !is.na(lon_Mel) ~ lon_Mel,
      TRUE ~ NA_real_
      ),
    
    # Assign confidence values:
    lat_long_confidence = case_when(
      
      has_Hem & (!is.na(lat_Mel)) & high_disagreement ~
        "medium confidence; Mel's estimate; found in both keys but high disagreement in Hem's",

      has_Hem & (!is.na(lat_Mel)) ~
        "high confidence; Hem's estimate; Hem and Mel's keys agree",
      
      has_Hem & is.na(lat_Mel) ~
        "high confidence; Hem's estimate; found in Hem's key",

      has_Hem & (!is.na(lat_Mel)) &
        (keys_lat_diff > 2 | keys_lon_diff > 2) ~
        "check required; Hem's estimate; Hem and Mel's keys disagree by >2º",

      !has_Hem & !is.na(lat_Mel) ~
        "medium confidence; Mel's estimate; found only in Mel's key",

      TRUE ~ NA_character_
    )
  ) %>%
  relocate(lat_est, lon_est, .after = last_col())

# 3038 obs.

table(LatLong_SiteKey$lat_long_confidence)
#                                  high confidence; Hem's estimate; found in Hem's key 
#                                                                                  351 
#                            high confidence; Hem's estimate; Hem and Mel's keys agree 
#                                                                                  501 
# medium confidence; Mel's estimate; found in both keys but high disagreement in Hem's 
#                                                                                   78 
#                           medium confidence; Mel's estimate; found only in Mel's key 
#                                                                                  571 

table(LatLong_SiteKey$GoC)

# SANITY CHECK: how many lat-long values are missing from the states of interest?
TEST<- LatLong_SiteKey %>%
  filter(state %in% States_of_Interest) %>%
  filter(is.na(lat_est))

# 530 sites in the States_of_Interest are missing lat-long values (down from 538 thanks to name cleaning!)
```


```{r Merge LatLong_SiteKey to NW.Mex_Landings}

# Merge key based on landing_site_clean names:
NW.Mex_Landings <- NW.Mex_Landings %>%
  left_join(LatLong_SiteKey, by = c("state", "landing_site_clean"))
# 4713144 obs.

# SANITY CHECK: how many records are missing lat-long values?
Landings_missingLatLong<- NW.Mex_Landings %>%
  filter(is.na(lat_est))
# 1/4 of records (n= 1,213,865) are not assigned... 
# May 20: Down from 1213865 to 1205911 (-7954) thanks to name cleaning!
# This doesn't really work for me.
# Can we make this key better?

# table(Landings_missingLatLong$landing_site)

# Let's take a look at a site we KNOW should be assigned like "BAHIA DE KINO" but isn't:
TEST2<- Landings_missingLatLong %>%
  filter(landing_site=="BAHIA DE KINO")
```


# STEP 8: Generate "integrated" LatLong_OfficeKey ------------------------------

In situations where we can't merge by state-landing_site let's merge by state-office

```{r Key skeleton}
colnames(Hems_LatLongKey_clean)

LatLong_OfficeKey0<- Hems_LatLongKey_clean %>%
  select(state, office_name_clean, Hems_est_lat= lat_deci, Hems_est_lon= lon_deci) %>%
  group_by(state, office_name_clean) %>%
  mutate(
    lat_mean = mean(Hems_est_lat , na.rm= TRUE),
    lon_mean = mean(Hems_est_lon , na.rm= TRUE)
  )
# 36486 obs., 6 vars.

# Reorganize this key:
colnames(LatLong_OfficeKey0)

LatLong_OfficeKey<- LatLong_OfficeKey0 %>%
  select(state, office_name_clean, est_lat_mean= lat_mean, est_lon_mean= lon_mean) %>%
  distinct()
# Down to 73 obs., 4 vars.
```

```{r Create LatLong_OfficeKey}
# NOTE: CLEAN NAMES function found above

# Match based on state-office_name_clean:

Landing_missing.matched <- Landings_missingLatLong %>%
  left_join(
    LatLong_OfficeKey,
    by = c("state", "office_name_clean")
  )

# Then find any values that are unmatched to a lat-long. Is this a small enough number I can live with?
unmatched <- Landing_missing.matched %>%
  filter(is.na(est_lat_mean))
# Only 241371 out of 4713144 un-matched (= 5%). I can live with this!

table(unmatched$state)
```

```{r Note from May 20 2026}
LatLong_OfficeKey$lat_long_confidence<- c("medium; Hem's estimate; estimate derived as a mean at the office level across sites")
```


# STEP 9: Add OHI region to the key ------------------------------

```{r Loading in OHI regions and buffer shape files}
# OHI regions:
OHI_regions_shape <- sf::st_read(
  "/home/shares/ohi/OHI_GOC/spatial/OHI_regions/marine/ohi_goc_final_071025.shp",
  options = "ENCODING=LATIN1"
)

OHI_regions_shape <- st_make_valid(OHI_regions_shape)

# Quick plot:
ggplot(OHI_regions_shape) +
  geom_sf(aes(fill = rgn_id)) +
  theme_minimal()

# 50km buffer:
OHI_buffer_shape <- sf::st_read(
  "/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer/inland_50km_goc.shp",
  options = "ENCODING=LATIN1"
)

OHI_buffer_shape <- st_make_valid(OHI_buffer_shape)

# Quick plot:
ggplot(OHI_buffer_shape) +
  geom_sf(aes(fill = rgn_id)) +
  theme_minimal()
```

```{r Spatial join: assign OHI region to LatLong_OfficeKey based on these shapes}
# A few important shape checks:
st_crs(LatLong_OfficeKey_sf)
st_crs(OHI_buffer_shape)

st_is_valid(OHI_regions_shape)
OHI_regions_shape <- st_make_valid(OHI_regions_shape)

st_is_valid(OHI_buffer_shape)
OHI_buffer_shape  <- st_make_valid(OHI_buffer_shape)

sf::sf_use_s2(FALSE)

# Convert office lat-longs to shapefile points
LatLong_OfficeKey_sf <- st_as_sf(
  LatLong_OfficeKey,
  coords = c("est_lon_mean", "est_lat_mean"),
  crs = 4326
)

# Optional ID for stable joins
LatLong_OfficeKey_sf$office_id <- seq_len(nrow(LatLong_OfficeKey_sf))

# Joins
regions_join <- st_join(
  LatLong_OfficeKey_sf,
  OHI_regions_shape,
  join = st_within
)

LatLong_OfficeKey_sf <- st_transform(
  LatLong_OfficeKey_sf,
  st_crs(OHI_buffer_shape)
)

buffer_join <- st_join(
  LatLong_OfficeKey_sf,
  OHI_buffer_shape,
  join = st_within
)

# Combine outputs:
final_join <- regions_join %>%
  mutate(
    buffer_rgn_id = buffer_join$rgn_id,
    location      = buffer_join$location,
    estado        = buffer_join$estado,
    ecorgn        = buffer_join$ecorgn
  )
```

```{r Do the same thing for LatLong_SiteKey_final}
LatLong_SiteKey_final<- LatLong_SiteKey %>%
  filter(!is.na(lat_est))

# Convert office lat-longs to shapefile points
LatLong_SiteKey_sf <- st_as_sf(
  LatLong_SiteKey_final,
  coords = c("lon_est", "lat_est"),
  crs = 4326
)

# Optional ID for stable joins
LatLong_SiteKey_sf$office_id <- seq_len(nrow(LatLong_SiteKey_sf))

# Joins
regions_join2 <- st_join(
  LatLong_SiteKey_sf,
  OHI_regions_shape,
  join = st_within
)

LatLong_SiteKey_sf <- st_transform(
  LatLong_SiteKey_sf,
  st_crs(OHI_buffer_shape)
)

buffer_join2 <- st_join(
  LatLong_SiteKey_sf,
  OHI_buffer_shape,
  join = st_within
)

# Combine outputs:
final_join2 <- regions_join2 %>%
  mutate(
    buffer_rgn_id = buffer_join2$rgn_id,
    location      = buffer_join2$location,
    estado        = buffer_join2$estado,
    ecorgn        = buffer_join2$ecorgn
  )

final_join2<- final_join2 %>%
  relocate(GoC, .before= rgn_id)

final_join2 <- final_join2 %>%
  mutate(
    region_est = case_when(
      !is.na(rgn_id) ~ rgn_id,
      !is.na(buffer_rgn_id) ~ buffer_rgn_id,
      TRUE ~ NA_character_
    )
  )

TEST<- final_join2 %>%
  filter(is.na(region_est))

table(TEST$state)
# Very few entries in states of Nayarit, Sinaloa, and Sonora. This is a good thing. It suggests that our lat longs are getting us pretty close. The values listed in Baja California, BCS, and Jalisco could be real.
    # BAJA CALIFORNIA BAJA CALIFORNIA SUR             JALISCO             NAYARIT             SINALOA              SONORA 
    #             214                 261                 218                  23                  33                  15 

table(TEST$GoC)
#  no yes 
# 526  23 
# More evidence these entries are real.
```

NOTE TO SELF: Tomorrow morning, come up with another flag system that flags things inside or outside of the GOC based on the state. Then, if inside the GOC, interpolate most likely region. If not inside the GOC (unknown) then exclude.
UNDER CONSTRUCTION, May 21 2026


# STEP 10: Export key(s) ------------------------------
```{r Clean up key before export}
# Final sites key:
LatLong_SiteKey_finalized<- LatLong_SiteKey

colnames(LatLong_SiteKey_finalized)

LatLong_SiteKey_finalized<- LatLong_SiteKey_finalized %>% 
  select(
    state, landing_site, GoC, est_habitat,
    lat_est, lon_est, lat_long_confidence,
    Hems_est_lat, Hems_est_lon, Hems_est_lat, Hems_est_lon,
    lat_Mel, lon_Mel,
    OHI_region_draft, site_notes_Mel
  )

colnames(LatLong_SiteKey_finalized)<- c(
  "state", "landing_site", "inside_GOC", "est_habitat",
  "lat_est", "lon_est", "latlon_est_flag",
  "HemKey_lat", "HemKey_lon", "HemPermitKey_lat", "HemPermitKey_lon",
  "MelKey_lat", "MelKey_lon",
  "OHI_region_draft", "site_notes_Mel"
)

LatLong_SiteKey_finalized <- LatLong_SiteKey_finalized %>%
  mutate(
    latlon_est_flag = if_else(
      is.na(latlon_est_flag),
      "alert 5: unable to assign lat-long based on state and landing site",
      NA_character_
    )
  )

sum(is.na(LatLong_SiteKey_finalized$lat_est))

# How many / which values are missing?
subset<- LatLong_SiteKey_finalized %>% 
  filter(is.na(lat_est))

LatLong_SiteKey_finalized<- LatLong_SiteKey_finalized %>%
  filter(is.na(lat_est))
# 1508 sites with lat-long details

# Final offices key:
LatLong_OfficeKey_finalized<- LatLong_OfficeKey
```


```{r PURG? remove unused dataframes}
rm(
  subset,
  BajaSonora_LatLongKey_clean, Jalisco_LatLongKey_clean,
  LatLong_merge.skeleton, LatLong_SiteKey,
  Mels_LatLongKey, Mels_LatLongKey_simple
)

rm(n_unique_sites, threshold_alert, threshold_low, threshold_medium)
```





# PURG ??? - May 21 2026

# STEP X: Add lat-long and OHI_region info to landings data ------------------------------
```{r OHI LatLong_SiteKey_finalized and RegionKey_pending}
LatLong_SiteKey_finalized<- read_csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/Sites_LatLongKey_May192026.csv")

RegionKey_pending<- read_csv(here("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/Spatial_Keys/RegionKey_UNDERCONSTRUCTION_Mar102026.csv"))

colnames(RegionKey_pending)
colnames(RegionKey_pending) <- c("state", "OHI_region_draft", "OHI_rgn_number", "OHI_rgn_name", "exclude_rgn", "OHI_rgn_notes")
```

```{r Append lat-long and OHI_region info to landings data}
temp_df <- Mex.Pacific_Landings %>%
  left_join(
    LatLong_SiteKey_finalized,
    by = c("state", "landing_site")
  )

table(temp_df$inside_GOC)

temp_df <- temp_df %>%
  left_join(
    RegionKey_pending,
    by = c("state", "OHI_region_draft")
  )

table(temp_df$inside_GOC)

colnames(temp_df)

test<- temp_df %>%
  filter(exclude_rgn=="exclude")

table(test$inside_GOC)
table(test$state)
table(test$OHI_rgn_number)

Mex.Pacific_Landings<- temp_df

rm(temp_df, test)
```

## SAVE: the NW Mexico landings appended with lat-long info in a separate .rds file:
```{r Save the Northwest Mexico dataframe: May 19 2026}
Mex.Pacific_Landings<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/Mex.Pacific_Landings_2000.present_May042026.rds"))
```


# STEP Y: Attach lat-long data to NW Mexico dataframe. (co-authors: Melanie Frazier, Hem Nalini Morzaria) ----------------------

## Flag these in the lat-long key as not in the Gulf [UNDER CONSTRUCTION May 19 2026]
office_name== c("EL ROSARIO", "ENSENADA", "TIJUANA")

# STEP Z: Attach lat-long and OHI_region info based on official landing site:
NOTE TO SELF: Come back to this! The lat long keys are imperfect. We'll want to fix this as much as possible.

## NEXT STEPS: Use of the lat-long key can be found in "CONAPESCA_CleaningWorkflow_May2026.Rmd"
found here: /home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys
