---
title: "Untitled"
output: html_document
date: "2025-04-29"
---

Exploring the INEGI-ENOE employment data. 
https://www.inegi.org.mx/app/descarga/
Microdata/National Survey of Occupation and Employment (ENOE), population aged 15 years and over/Database
p3 is code for industry
https://www.inegi.org.mx/contenidos/productos/prod_serv/contenidos/espanol/bvinegi/productos/nueva_estruc/702825198411.pdf
I got the 2024, 4th quarter data.


```{r setup, include=FALSE}

library(tidyverse)

```

Exploring the data.

## getting states around Gulf of California
```{r}

tmp1 <- read_csv("/home/shares/ohi/OHI_GOC/inegi_enoe_employment/ENOE_COE1T424.csv")
sum(tmp1$fac_tri)
# 108,728,324
dim(tmp1)
# 345,680

check <- filter(tmp1, p3 %in% c("6224", "6226", "6227"))
 sum(check$fac_tri)

state_names <- tibble(
  ent = 1:32,
  state = c("Aguascalientes", "Baja California", "Baja California Sur", "Campeche", "Coahuila",
            "Colima", "Chiapas", "Chihuahua", "CDMX", "Durango",
            "Guanajuato", "Guerrero", "Hidalgo", "Jalisco", "Mexico State",
            "Michoacán", "Morelos", "Nayarit", "Nuevo León", "Oaxaca",
            "Puebla", "Querétaro", "Quintana Roo", "San Luis Potosí", "Sinaloa",
            "Sonora", "Tabasco", "Tamaulipas", "Tlaxcala", "Veracruz",
            "Yucatán", "Zacatecas")
)

tmp1a <- tmp1 %>%
  left_join(state_names, by = "ent") %>%
  filter(state %in% c("Baja California", "Baja California Sur", "Sonora", "Sinaloa", "Nayarit", "Jalisco"))
sum(tmp1a$fac_tri)
## 17,407,179
dim(tmp1a)
## 58,936

```

fac_tri is the expansion factor for each respondent in the quarterly sample. It tells you how many people in the total population that one person represents in that quarter.

Think of it like:

“This person stands in for ~X people in the Mexican population in this quarter.”
It should approximate the working-age population (12+) of Mexico for the quarter, in households covered by ENOE.

As of 2023–2024, this might be around 100 to 110 million depending on the month, but not the full national population (~130 million), because:

ENOE excludes kids under 12,

It samples households, not institutional housing (like prisons),

It may be limited to urban/rural areas depending on the wave.

> sum(tmp1$fac_tri)
[1] 108728324



## Get employed (worked at least one hour past week)

```{r}

tmp1b <- tmp1a %>%
   filter(p1 == 1)  # 1 = worked at least 1 hour last week
sum(tmp1b$fac_tri)
## 9,412,046
dim(tmp1b)
## 32,816

```


## determine job codes of relevance to marine jobs
```{r}
unique((tmp1b$p3))

```