---
title: "Untitled"
output: html_document
date: "2025-04-29"
---

Exploring the INEGI-ENOE employment data. 
https://www.inegi.org.mx/app/descarga/
Microdata/National Survey of Occupation and Employment (ENOE), population aged 15 years and over/Database
p3 is code for industry
https://www.inegi.org.mx/contenidos/productos/prod_serv/contenidos/espanol/bvinegi/productos/nueva_estruc/702825198411.pdf

p4A North American codes (fishing is more complete):
http://internet.contenidos.inegi.org.mx/contenidos/productos/prod_serv/contenidos/espanol/bvinegi/productos/nueva_estruc/702825197247.pdf

p3 and p4A have a lot of overlap but p4A is more complete, getting people who are involved in fishing industry beyond just the fisherman. 


I got the 2024, 4th quarter data. The COE data are the basic version (vs. the extended version).

Breakdown of tables: 
Occupation data: https://www.inegi.org.mx/contenidos/productos/prod_serv/contenidos/espanol/bvinegi/productos/nueva_estruc/702825198411.pdf

https://www.economia.gob.mx/datamexico/en/profile/industry/fishing


```{r setup, include=FALSE}

library(tidyverse)
library(here)

```

## Exploring the data.
There are 5 dataframes associated with these data. Get a feel for which ones are useful
survey is stored:
•	In the housing table (vivt), the identification data of the selected households is stored, as well as information about the number of residents in the dwelling and how many households it includes. This information is collected through questions 1, 2, and 3 of the sociodemographic questionnaire.
•	In the households table (hogt), the identification of each household found within a dwelling is stored, along with the survey dates and the outcome of both the preliminary and final interviews.
•	The sociodemographic table (sdemt) contains characteristics of household residents, such as residency status, age, sex, etc. These data are gathered through questions 5 to 23 of the sociodemographic questionnaire.
•	The occupation and employment questionnaire (COE), due to its size, is stored in two data tables: COE1 and COE2. In the basic version, the first table includes questions 1 through 5h, and the second includes questions 6 through 9. In the extended version, the first table stores the same questions, while the second includes questions 6 through 11.


```{r}
# This table provides data at the household level. There is municipality data, which may be useful
# for higher resolution data (if we need better than state or city). AGEB is a geostatistical area similar to a neighborhood,
# but no data. Maybe this is available in other formats? Otherwise, I don't think we will use this table.
# The data that seems useful is present in other datasets.
tmp_viv <- read_csv("/home/shares/ohi/OHI_GOC/_raw_data/inegi_enoe_employment/ENOE_VIVT424.csv")
summary(tmp_viv)
table(tmp_viv$est)
table(tmp_viv$est_d_tri)
table(tmp_viv$ageb)

#This table is also at the household level and doesn't really seem to provide additional information that 
# will be useful for our analysis, maybe r_def = 1 indicates a complete interview (although that is in the next table as well)
tmp_hog <- read_csv("/home/shares/ohi/OHI_GOC/_raw_data/inegi_enoe_employment/ENOE_HOGT424.csv")
summary(tmp_hog)

## This has information that could be helpful in determining job quality and sex breakdown
#r_def = complete interview (value = 1)
#ent = state
#sex 1= male 2 = female
#eda = age
#clase2, 1 = employed
#ing7c, classification of income level
#dur9c, working hours category level
#tue2, business classification
#tue3, gov/non-gov
#sub_o, underemployed, by hours
#medica5c, health benefits
#anios_esc = years of schooling
#hrsocuup - hours worked in a week
#ingocup = monthly income

tmp_dem <- read_csv("/home/shares/ohi/OHI_GOC/_raw_data/inegi_enoe_employment/ENOE_SDEMT424.csv")
summary(tmp_dem)
table(tmp_dem$est_d_tri)
tmp_dem <- tmp_dem %>%
  mutate(unique_id = paste(tmp_dem$con, tmp_dem$upm, sep="_")) %>%
 # mutate(unique_id = paste(unique_id, tmp_dem$v_sel, sep="_")) %>%
  mutate(unique_id = paste(unique_id, tmp_dem$n_ren, sep="_")) %>%
   mutate(unique_id = paste(unique_id, tmp_dem$n_hog, sep="_")) %>%
  mutate(unique_id = paste(unique_id, tmp_dem$n_pro_viv, sep="_")) %>%
  mutate(con_upm = paste(tmp_dem$con, tmp_dem$upm, sep="_")) %>%
  mutate(con_upm_nren =paste(con_upm, tmp_dem$n_ren))
sum(duplicated(tmp_dem$unique_id))
tmp_dem[(duplicated(tmp_dem$unique_id)),]

```

These are census data. But the fac_tri value indicates how many people each row (i.e., person represents).
Checking that this matches up!

fac_tri is the expansion factor for each respondent in the quarterly sample. It tells you how many people in the total population that one person represents in that quarter.

Think of it like:

“This person stands in for ~X people in the Mexican population in this quarter.”
It should approximate the working-age population (12+) of Mexico for the quarter, in households covered by ENOE.

As of 2023–2024, this might be around 100 to 110 million depending on the month, but not the full national population (~130 million), because:

ENOE excludes kids under 12,

It samples households, not institutional housing (like prisons),

It may be limited to urban/rural areas depending on the wave.

```{r}
tmp1 <- read_csv("/home/shares/ohi/OHI_GOC/_raw_data/inegi_enoe_employment/ENOE_COE1T424.csv")
sum(tmp1$fac_tri)
# 108,728,324
# population is 131.9 million people
# population 12+ is about 107 million - so this seems about right

dim(tmp1)
# 345,680
summary(tmp1)

## check number of people in fishing jobs
check1 <- filter(tmp1, p3 %in% c("6211")) #Industry data
 sum(check1$fac_tri)
# 91,8994 
 # From here: https://www.economia.gob.mx/datamexico/en/profile/industry/fishing?occupationMetrics=workforceOption&quarterFinal=option20243&quarterInitial1=option20243
 # 2024 third quarter was 91.8k. So this checks out.
 
 dim(check1) # you can see that a relatively small number of respondents is representing the larger population

check2 <- filter(tmp1, p4a %in% c("1141")) # This is the North America Industry
table(check2$p3) # can see that most match the 6211, but there are several other occupations included in there - which I think makes sense
 sum(check2$fac_tri)
 dim(check2)
 
```


## Obtaining a unique ID and join data

```{r}

jobs1 <- read_csv("/home/shares/ohi/OHI_GOC/_raw_data/inegi_enoe_employment/ENOE_COE1T424.csv")

jobs1 <- jobs1
jobs1 <- jobs1 %>%
  mutate(unique_id = paste(jobs1$con, jobs1$upm, sep="_")) %>%
  mutate(unique_id = paste(unique_id, jobs1$v_sel, sep="_")) %>%
  mutate(unique_id = paste(unique_id, jobs1$n_ren, sep="_")) %>%
  mutate(unique_id = paste(unique_id, jobs1$n_hog, sep="_")) %>%
  mutate(unique_id = paste(unique_id, jobs1$n_pro_viv, sep="_")) %>%
  select(unique_id, r_def, upm, v_sel, n_ren, n_hog, n_pro_viv, cd_a, ent, p1, p1e, p3, p4a, p4d1,
         fac_tri)

sum(duplicated(jobs1$unique_id)) # should be no duplicates.

jobs2 <- read_csv("/home/shares/ohi/OHI_GOC/_raw_data/inegi_enoe_employment/ENOE_COE2T424.csv")

jobs2 <- jobs2 %>%
  mutate(unique_id = paste(jobs2$con, jobs2$upm, sep="_")) %>%
  mutate(unique_id = paste(unique_id, jobs2$v_sel, sep="_")) %>%
  mutate(unique_id = paste(unique_id, jobs2$n_ren, sep="_")) %>%
  mutate(unique_id = paste(unique_id, jobs2$n_hog, sep="_")) %>%
  mutate(unique_id = paste(unique_id, jobs2$n_pro_viv, sep="_")) %>%
  select(unique_id, p6c, p6d, p6e, p6e_c, p6f, p7a, p7c, p6e_c, p6f_c)

sum(duplicated(jobs2$unique_id))

length(setdiff(jobs2$unique_id, jobs1$unique_id)) # should be no differences

jobs <- left_join(jobs1, jobs2, by = "unique_id")

```


# Identify the relevant states
```{r}
state_names <- tibble(
  ent = 1:32,
  state = c("Aguascalientes", "Baja California", "Baja California Sur", "Campeche", "Coahuila",
            "Colima", "Chiapas", "Chihuahua", "CDMX", "Durango",
            "Guanajuato", "Guerrero", "Hidalgo", "Jalisco", "Mexico State",
            "Michoacán", "Morelos", "Nayarit", "Nuevo León", "Oaxaca",
            "Puebla", "Querétaro", "Quintana Roo", "San Luis Potosí", "Sinaloa",
            "Sonora", "Tabasco", "Tamaulipas", "Tlaxcala", "Veracruz",
            "Yucatán", "Zacatecas")
)

jobs <- jobs %>%
  left_join(state_names, by = "ent") %>%
  filter(state %in% c("Baja California", "Baja California Sur", "Sonora", "Sinaloa", "Nayarit", "Jalisco"))
sum(jobs$fac_tri)
## 17,407,179
dim(jobs)
## 58,936

## This seems reasonable. Total population in 2020 of these states is 
(3769020 + 798447 + 2944840 + 3026943 + 1235456 + 8348151) * 0.81
# about 81% are 12+ in age, this value seems reasonable.

```


## Identify relevant codes and combine the two occupation datasets
### Primary jobs
```{r}

occupations <- read_csv(here("LIV/data/occupations.csv"))

occ <- occupations %>%
  filter(data == "occupation") %>%
  #filter(!is.na(category1)) %>%
  mutate(p3 = code) %>%
  select(-data, -code, p3, cat1_occ = category1, cat2_occ= category2, occ_description = description)

occ_marine <- occ %>%
  filter(!is.na(cat1_occ))

industry <- occupations %>%
  filter(data == "north_america_industry") %>%
  #filter(!is.na(category1)) %>%
  mutate(p4a = code)%>%
  select(-data, -code, p4a, cat1_ind = category1, cat2_ind= category2, ind_description = description)
  
ind_marine <- industry %>%
  filter(!is.na(cat1_ind))

occs_unique_id <- jobs %>%
  dplyr::filter(p3 %in% occ_marine$p3) %>%
  select(unique_id) %>%
  pull() # N = 4188

ind_unique_id <- jobs %>%
  dplyr::filter(p4a %in% ind_marine$p4a) %>%
  select(unique_id) %>%
  pull() # N = 4811

setdiff(occs_unique_id, ind_unique_id) 
setdiff(ind_unique_id, occs_unique_id)

marine_ids <- unique(c(occs_unique_id, ind_unique_id)) #N = 5564

marine_jobs_primary <- jobs %>%
  dplyr::filter(unique_id %in% marine_ids) %>%
  left_join(occ, by="p3") %>%
  left_join(industry, by= "p4a") %>%
  mutate(occ_ind = paste(occ_description, ind_description, sep = " --- ")) %>%
  mutate(occ_ind_id = paste(p3, p4a, sep = "---")) %>%
  group_by(occ_ind_id) %>%
  mutate(count = n()) %>%
  arrange(-count) %>%
  ungroup() 

marine_jobs_count <- marine_jobs_primary %>%
  select(occ_ind, occ_ind_id, count) %>%
  unique()

write_csv(marine_jobs_count, here("LIV/data/occupations_industry_blank.csv"))

```

## Identify employment in marine categories
Started to do this: Worked at least one hour past week (p1==1), and if not will return (p1e == 1,2,3).
But the data aren't actually in the provided data. So I will have to assume that if they have an occupation then
they are employed. 
```{r}
occu_cats <- read_csv(here("LIV/data/occupations_industry.csv"))

jobs2 <- jobs %>%
  mutate(occ_ind_id = paste(p3, p4a, sep="---")) %>%
  left_join(occu_cats, by="occ_ind_id") 

jobs2 <- jobs2 %>%
  filter(tourism == "x") %>%
  mutate(fac_tri_correct = 1)

table(jobs2$category1)
table(jobs2$category2)
table(jobs2$category3)

## divide out the combined aquaculture and fisheries based on overall employment
aqua <- filter(jobs2, jobs2$category1=="aquaculture")
fish <- filter(jobs2, jobs2$category1=="fisheries")

aqua_correct <- sum(aqua$fac_tri)/(sum(aqua$fac_tri) + sum(fish$fac_tri))

fish_aqua <- filter(jobs2, category1=="aquaculture-fisheries")
  
fish_aqua_correct <- rbind(fish_aqua %>% mutate(category1="aquaculture"),
                           fish_aqua %>% mutate(category1="fisheries")) %>%
                    mutate(fac_tri_correct = ifelse(category1=="aquaculture", 0.2, 0.8) ) 

jobs3 <- jobs2 %>%
  filter(category1 != "aquaculture-fisheries") 
jobs3 <- rbind.data.frame(jobs3, fish_aqua_correct)


```

### Plots of data
```{r}

# get fully marine related
table(jobs3$category1)

marine_jobs <- jobs3 %>%
  filter(category1 %in% c("aquaculture", "fisheries", "transportation-marine", "culture-entertainment-marine")) %>%
  group_by(category1, state) %>%
  summarize(employed = sum(fac_tri, na.rm=TRUE)) %>% 
  ungroup() %>%
  group_by(category1) %>%
  mutate(total_employed = sum(employed)) %>%
  ungroup() %>%
  mutate(category1 = reorder(category1, -total_employed))

ggplot(marine_jobs, aes(x = category1, y = employed, fill = state)) +
  geom_bar(stat = "identity") +
  labs(x = "Occupation Category", y = "Number Employed", fill = "State") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


# get tourism employment, at least partially marine related
table(jobs3$category3)

marine_jobs <- jobs3 %>%
  filter(category3 %in% "tourism") %>%
  mutate(category1 = gsub("-marine", "", category1)) %>%
  filter(category1 != "aquaculture") %>%
  group_by(category1, state) %>%
  summarize(employed = sum(fac_tri, na.rm=TRUE)) %>% 
  ungroup() %>%
  group_by(category1) %>%
  mutate(total_employed = sum(employed)) %>%
  ungroup() %>%
  mutate(category1 = reorder(category1, -total_employed))

ggplot(marine_jobs, aes(x = category1, y = employed, fill = state)) +
  geom_bar(stat = "identity") +
  labs(x = "Occupation Category", y = "Number Employed", fill = "State") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# get research
table(jobs3$category3)

marine_jobs <- jobs3 %>%
  filter(category3 %in% "research") %>%
  group_by(category2, state) %>%
  summarize(employed = sum(fac_tri, na.rm=TRUE)) %>% 
  ungroup() %>%
  group_by(category2) %>%
  mutate(total_employed = sum(employed)) %>%
  ungroup() %>%
  mutate(category2 = reorder(category2, -total_employed))

ggplot(marine_jobs, aes(x = category2, y = employed, fill = state)) +
  geom_bar(stat = "identity") +
  labs(x = "Occupation Category", y = "Number Employed", fill = "State") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
