---
title: 'OHI GoC 2025: Livelihoods & Economies'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: inline
editor_options: 
  chunk_output_type: inline
---

# LE: Sistema Automatizado de Información Censal (SAIC) // Automated Census Information System (SAIC)

1. Select the census year(s).
2. Indicates the geographical area.
3. Choose the economic activities.
4. Indicates the census variables.
5. Click on Consult.

Additionally, you can view information by employment strata and learn about the census results.
If you prefer, you can download all the information by selecting and following the steps indicated in the Bulk Download icon.

Methodological document: /home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025/saic_historico_metodologico_ce2024.pdf
* Note 1: Corrected figures. See note

* Note 2: As of Wednesday, July 30, 2025, the query of information in the Automated Census Information System was reestablished for the years 2003, 2008, 2013, as a result of the review of the query of the composition of the strata of employed personnel. The information published today replaces the previous one, which was affected in the cuts of "Strata of employed personnel", this was due to a technical adjustment in the information integration procedure used.

* Note 3: The monetary information presented here corresponds to the value in current prices of the year of each census event, therefore, to make comparisons between the information of the different years available in this system, it is suggested to use the Implicit Price Index (IPI) available in the Goods and Services Accounts (detailed). Base year 2018 . Specifically, during the presentation of the final results, the Revised 2023 IPI value was used, for example, for the Gross Census Value Added the following inflators per year were used: for 2008, 1.92081442376959; for 2013, 1.59502439833556 and for 2018, 1.26412683998505; and for Remunerations the following were used: for 2008, 1.93153962083199; for 2013, 1.5759; and for 2018, 1.2895. To calculate other variables and to see the complete calculation method, they can be found in the "Methodological Document" of this system.

* Note 4: The employed personnel reported in this Statistical Program is conceptually different (see metadata) from the definition of jobs affiliated with the Mexican Social Security Institute.

---

For each of the municipalities mentioned in my [comment above](https://github.com/OHI-GOC/OHI_GOC_issues/issues/20#issuecomment-3383649433), I downloaded the data for every relevant sector through the [Automated Census Information System (SAIC)](https://www.inegi.org.mx/app/saic/).  Each sector's dataset has variables that apply to both livelihoods and economies.  

They are now located here: ~/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025

I downloaded the sectors separately so that we can keep track of the types of sectors we have, but I will likely bind all of these datasets into one.  Since everything had to be checked (physically pressing the button haha) in order to download, it felt easiest to keep the job/economy, years, and municipio variables the same, and just vary the sectors selected.

These are all of the datasets/sectors downloaded:

* SAIC_Exporta_20251021_163827445_fishing.csv
* SAIC_Exporta_20251021_171653871_organizations.csv
* SAIC_Exporta_20251021_171548286_public_restrooms_parking.csv
* SAIC_Exporta_20251021_171034329_food_services.csv
* SAIC_Exporta_20251021_17859901_accommodations.csv
* SAIC_Exporta_20251021_17745611_tourist_marinas.csv
* SAIC_Exporta_20251021_17550467_museums_historical_naturalparks.csv
* SAIC_Exporta_20251021_17421711_job_training_services.csv
* SAIC_Exporta_20251021_1712765_hazardous_waste.csv
* SAIC_Exporta_20251021_165918141_professional_scientific_technical_services.csv
* SAIC_Exporta_20251021_165616924_possible_natural_products.csv
* SAIC_Exporta_20251021_16511703_prep_package_fish_shellfish.csv
* SAIC_Exporta_20251021_16482728_supply_naturalgas.csv
* SAIC_Exporta_20251021_16471476_energy_generation.csv
* SAIC_Exporta_20251021_164547329_water_collection_treatment_supply.csv
* SAIC_Exporta_20251021_164332559_all_mining.csv
* SAIC_Exporta_20251021_164117726_oil_gas.csv
* SAIC_Exporta_20251021_164011725_aquaculture.csv

**For every municipio, there is a "Municipal Total" row, which based on the SAIC Metodológico 2024 document from INEGI, represents the complete total for all registered economic units and activity types in that municipality.**

Here’s how it works according to the methodology:
- Each “Municipal total” aggregates the values of all economic units within that municipality -- across all sectors, subsectors, and classes classified under the SCIAN México 2023 system.
- This includes all the summed up values for variables such as (for example!!):
  - EU (Economic Units): all establishments registered in that municipality.
  - J000A (Total remuneration): total payroll and compensation for all workers in that municipality.
  - K000A (Total expenditures on goods and services): total operational expenditures across all sectors.

The document explicitly explains that these aggregates are constructed so users can obtain a “panorama del comportamiento económico” (economic overview) at the municipal, state, or national level.

That means we can compare sectoral data (like “Clase 713930 Marinas turísticas”) to the “Municipal total” to determine the proportion or contribution of that specific sector to the total economic activity in any municipio (if they participate in that sector).

> For example:
> If the municipal total remuneration is 46,766.636 million pesos, and the “Temporary accommodation and food services” sector has 1,989.833 million pesos, then that sector accounts for roughly 4.3% of total remuneration in Mexicali.

# Exploring SAIC

```{r}
library(tidyverse)
library(sf)
library(mapview)
```

## Accommodations

```{r}
## read in data
saic_accommodations <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251021_17859901_accommodations.csv"), skip = 4)

saic_accommodations_liv <- saic_accommodations %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)"))

saic_accommodations_liv_loscabos <- saic_accommodations_liv %>% 
  filter(municipio %in% "008 Los Cabos")
```

## Fishing

```{r}
## read in data
saic_fishing <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251021_163827445_fishing.csv"), skip = 4)

saic_fishing_liv <- saic_fishing %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)"))

saic_fishing_liv_loscabos <- saic_fishing_liv %>% 
  filter(municipio %in% "008 Los Cabos") # none

saic_fishing_eco <- saic_fishing %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))
```

## Aquaculture

```{r}
## read in data
saic_aquaculture <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251021_164011725_aquaculture.csv"), skip = 4)

saic_aquaculture_liv <- saic_aquaculture %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)"))

saic_aquaculture_liv_loscabos <- saic_aquaculture_liv %>% 
  filter(municipio %in% "008 Los Cabos") # none

saic_aquaculture_eco <- saic_aquaculture %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))
```

## Prep fish & shellfish

```{r}
## read in data
saic_fish_processing <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251021_16511703_prep_package_fish_shellfish.csv"), skip = 4)

saic_fish_processing_liv <- saic_fish_processing %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)"))

saic_fish_processing_liv_loscabos <- saic_fish_processing_liv %>% 
  filter(municipio %in% "008 Los Cabos") # none

saic_fish_processing_eco <- saic_fish_processing %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))
```

## Tourist marinas

```{r}
## read in data
saic_tourist_marina <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251021_17745611_tourist_marinas.csv"), skip = 4)

saic_tourist_marina_liv <- saic_tourist_marina %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)"))

saic_tourist_marina_liv_loscabos <- saic_tourist_marina_liv %>% 
  filter(municipio %in% "008 Los Cabos") 

saic_tourist_marina_eco <- saic_tourist_marina %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))
```

## Museums, historical, natural parks

```{r}
## read in data
saic_natural_parks <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251021_17550467_museums_historical_naturalparks.csv"), skip = 4)

saic_natural_parks_liv <- saic_natural_parks %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)")) # not many...

saic_natural_parks_liv_loscabos <- saic_natural_parks_liv %>% 
  filter(municipio %in% "008 Los Cabos") 

saic_natural_parks_eco <- saic_natural_parks %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))
```

## Public restrooms, parking

```{r}
## read in data
saic_restrooms_parking <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251021_171548286_public_restrooms_parking.csv"), skip = 4)

saic_restrooms_parking_liv <- saic_restrooms_parking %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)"))

saic_restrooms_parking_liv_loscabos <- saic_restrooms_parking_liv %>% 
  filter(municipio %in% "008 Los Cabos") 


saic_restrooms_parking_eco <- saic_restrooms_parking %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))
```

## Food services

```{r}
## read in data
saic_food_services <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251021_171034329_food_services.csv"), skip = 4)

saic_food_services_liv <- saic_food_services %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)"))

saic_food_services_liv_loscabos <- saic_food_services_liv %>% 
  filter(municipio %in% "008 Los Cabos") # 722512: Restaurantes con servicio de preparación de pescados y mariscos

saic_food_services_eco <- saic_food_services %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))
```

## Energy generation

```{r}
## read in data
saic_energy_generation <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251021_16471476_energy_generation.csv"), skip = 4)

# 221111 Generation of electricity from fossil fuels
# 221112 Generation of electricity from hydropower
# 221113 Generation of electricity from solar energy
# 221114 Generation of electricity from wind energy
# 221119 Generation of electricity from other types of energy

saic_energy_generation_liv <- saic_energy_generation %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)"))

saic_energy_generation_liv_loscabos <- saic_energy_generation_liv %>% 
  filter(municipio %in% "008 Los Cabos") # 722512: Restaurantes con servicio de preparación de pescados y mariscos

saic_energy_generation_eco <- saic_energy_generation %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))
```

## Supply natural gas

```{r}
## read in data
saic_natural_gas_supply <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251021_16482728_supply_naturalgas.csv"), skip = 4)

saic_natural_gas_supply_liv <- saic_natural_gas_supply %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)")) # none anywhere really

saic_natural_gas_supply_liv_loscabos <- saic_natural_gas_supply_liv %>% 
  filter(municipio %in% "008 Los Cabos")

saic_natural_gas_supply_eco <- saic_natural_gas_supply %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))
```


## Oil and gas: NONE?

```{r}
## read in data
saic_oil <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251021_164117726_oil_gas.csv"), skip = 4)

saic_oil_liv <- saic_oil %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)")) # none anywhere really

saic_oil_liv_loscabos <- saic_oil_liv %>% 
  filter(municipio %in% "008 Los Cabos")

saic_oil_eco <- saic_oil %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))
```

## Water collection, treatment, supply

```{r}
## read in data
saic_water <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251022_13388238_water_treatment_supply.csv"), skip = 4)

saic_water_liv <- saic_water %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)")) # only in some places like in Nayarit...

saic_water_eco <- saic_water %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))
```

## Professional and scientific services

```{r}
## read in data
saic_sci_services <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251021_165918141_professional_scientific_technical_services.csv"), skip = 4)

saic_sci_services_liv <- saic_sci_services %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)")) # none anywhere really

saic_sci_services_eco <- saic_sci_services %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))
```

## Mining

```{r}
## read in data
saic_mining <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251021_164332559_all_mining.csv"), skip = 4)

# Clase 212391 Minería de sal
# none except Navolato 2023...

saic_mining_liv <- saic_mining %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)")) # none anywhere really

saic_mining_eco <- saic_mining %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))
```


## Hazardous waste

```{r}
## read in data
saic_hazardous_waste <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251021_1712765_hazardous_waste.csv"), skip = 4)

saic_hazardous_waste_liv <- saic_hazardous_waste %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)")) # none anywhere really

saic_hazardous_waste_eco <- saic_hazardous_waste %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))
```

## Possible NP

```{r}
## read in data
saic_possible_np <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251021_165616924_possible_natural_products.csv"), skip = 4)

# only culiacan... and one or two others...

saic_possible_np_liv <- saic_possible_np %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)")) # none anywhere really

saic_possible_np_eco <- saic_possible_np %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))
```


## Job training services

For those unemployed.

```{r}
## read in data
saic_job_training <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251021_17421711_job_training_services.csv"), skip = 4) # only one year for Hermosillo available.

saic_job_training_liv <- saic_job_training %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)")) # none anywhere really

saic_job_training_eco <- saic_job_training %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))
```


## Associations and organizations

813 Associations and organizations

```{r}
## read in data
saic_organizations <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251022_1414982_organizations.csv"), skip = 4) # a lot of data, but might not even be helpful.

saic_organizations_liv <- saic_organizations %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)")) # a lot more data!!

saic_organizations_eco <- saic_organizations %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))
```


## Combine

```{r}
## all _liv objects but exclude _liv_loscabos
liv_objs <- ls(pattern = "_liv$")
liv_objs <- liv_objs[!grepl("_liv_loscabos$", liv_objs)]

all_liv <- mget(liv_objs) %>%
  bind_rows(.id = "source")

## all _eco
all_eco <- mget(ls(pattern = "_eco$")) %>%
  bind_rows(.id = "source")

unique(all_liv$source)
unique(all_eco$source)
```


# Sector shares

```{r}
library(tidyverse)

# List of all sector datasets
sector_datasets <- list(
  accommodations = saic_accommodations_liv,
  fishing = saic_fishing_liv,
  aquaculture = saic_aquaculture_liv,
  fish_processing = saic_fish_processing_liv,
  tourist_marina = saic_tourist_marina_liv,
  natural_parks = saic_natural_parks_liv,
  restrooms_parking = saic_restrooms_parking_liv,
  food_services = saic_food_services_liv,
  energy_generation = saic_energy_generation_liv,
  natural_gas = saic_natural_gas_supply_liv,
  oil_gas = saic_oil_liv,
  water_services = saic_water_liv,
  sci_services = saic_sci_services_liv,
  mining = saic_mining_liv,
  hazardous_waste = saic_hazardous_waste_liv,
  possible_np = saic_possible_np_liv,
  job_training = saic_job_training_liv,
  organizations = saic_organizations_liv
)

# Function to calculate sector share of municipal total
calculate_sector_share <- function(df) {
  
  df %>%
    group_by(ano_censal, entidad, municipio) %>%
    # extract municipal total for that year & municipio
    mutate(
      municipal_total = h001a_personal_ocupado_total[actividad_economica == "Total municipal"],
      proportion = h001a_personal_ocupado_total / municipal_total
    ) %>%
    ungroup()
}

# Apply to all sectors and combine
sector_shares <- map_df(sector_datasets, calculate_sector_share, .id = "sector")

# Optional: select key columns
sector_shares_summary <- sector_shares %>%
  select(sector, ano_censal, entidad, municipio, actividad_economica, h001a_personal_ocupado_total, municipal_total, proportion)

# View example
sector_shares_summary %>% filter(municipio == "008 Los Cabos")

```



