################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```


# STEP 1: Generate timeseries by OHI region and genus ------------------------------

```{r Load in GOC_MarineLandings0 as necessary}
GOC_MarineLandings0<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Marine.Landings_Sector.Sorted.Georef_PATCHED_Jun032026.rds"))
# 3246583 obs., 58 var.

# NOTE TO SELF: This source file may change in later Summer / Fall run. - EMF, June 4 2026
```

# **DECISION POINT: Selecting which sectors to include**
```{r Select Sectors_of_Interest}
table(GOC_MarineLandings0$sector_estimate_draft)

# SANITY CHECK: Which species are included in the freshwater or marine sector?
TEST<- GOC_MarineLandings0 %>%
  filter(sector_estimate_draft== "freshwater or marine: fishery sector yet to be classified")

table(TEST$recorded_taxa_group)
# Species like Mojarra (the overwhelming majority of catch), trout, and catfish, which can move from freshwater to marine systems and are a bit ambiguous in terms of where they can be caught.


MarineFishery_Sectors<- c("marine: fishery sector yet to be classified",
                          "marine: small-scale capture",
                          "marine: large-scale capture")
  
PossibleMarineFishery_Sectors<- c("freshwater or marine: large-scale capture",
                          "freshwater or marine: small-scale capture",
                          "freshwater or marine: fishery sector yet to be classified")

GOC_MarineFisheryLandings0<- GOC_MarineLandings0 %>%
  filter(sector_estimate_draft %in% c(MarineFishery_Sectors, PossibleMarineFishery_Sectors))

# 3137044 obs., 58 var.
```

```{r Generate GOC_MarineFisheryLandings_summary}
colnames(GOC_MarineFisheryLandings0)

GOC_MarineFisheryLandings_summary<- GOC_MarineFisheryLandings0 %>%
  select(year, 
         state, office_name_clean, landing_site_clean,
         GOC_est, GOC_Include_Exclude, OHI_region_est, OHI_region_name,
         eu_rnpa, eu_name, sector_estimate_draft, 
         Major_Functional_Group, recorded_taxa_group, estimated_taxa_group,
         recorded_taxa_name, recorded_taxa_name_clean,
         estimated_species_name, estimated_genus_name,
         landed_weight, live_weight, weight_kg_corrected, value
         )

GOC_MarineFisheryLandings_summary<- GOC_MarineFisheryLandings_summary %>%
  group_by(year, OHI_region_est, estimated_genus_name) %>%
  summarize(
    total_weight_mt= sum(weight_kg_corrected, na.rm = TRUE) / 1000,
    total_value_peso= sum(value, na.rm = TRUE)
  )
```

# STEP 2: Load and merge the various FIS scores ------------------------------

```{r Load the indicator dataframes as necessary}
# Read in datafiles as necessary:
CPUE_Scores0<- read.csv("/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_CPUE_Scores/FIS_CPUE_Scores_Jun032026.csv")
# 101 obs., 11 vars.

RAM_Scores0<- read.csv("/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_RAM_Scores/FIS_RAM_Scores_Jun032026.csv")
# 216 obs., 14 vars.

Carta_Scores0<- read.csv("/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_Carta_Scores/FIS_CartaScores_Jun032026.csv")
# 1299 obs., 17 vars.

CMSY_Scores0<- read.csv("/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_CMSY_Scores/FIS_CMSY_Scores_Jun012026.csv")
# 1569 obs., 11 vars.
```

## RAM Data
```{r Augment the RAM dataframe by collapsing all stock info into one score per genus per year}
colnames(RAM_Scores0)

RAM_Scores <- RAM_Scores0 %>%
  group_by(year, Genus) %>%
  summarize(
    RAM_BBmsy = mean(RAM_BBmsy_final, na.rm = TRUE),
    n_Formal_Assessments = sum(!is.na(RAM_BBmsy_numeric)),
    .groups = "drop"
  ) %>%
  mutate(
    RAM_provenance = case_when(
      n_Formal_Assessments == 0 ~ "value was interpolated based on formal assessment from other years (RAM or other*)",
      n_Formal_Assessments == 1 ~ "value was sourced from 1 formal assessment (RAM or other*)",
      n_Formal_Assessments >= 2 ~ "value was sourced from 2 or more formal assessments (RAM or other*)"
    )
  )
# 168 obs., 5 var. 

# lowercase to make sure matching works
RAM_Scores <- RAM_Scores %>%
  mutate(estimated_genus_name = tolower(Genus),
         Stock_Status_RAM= RAM_BBmsy) %>%
  select(-c(Genus, RAM_BBmsy))
# 168 obs., 6 var. 
```

```{r Rename the Stock_Status column in Carta_Scores}
Carta_Scores<- Carta_Scores0 %>%
  mutate(
    Stock_Status_Carta= Stock_Status
  ) %>%
  select(-c(Stock_Status))
```

```{r Rename the Stock_Status column in Stock_Status_CMSY}
CMSY_Scores<- CMSY_Scores0 %>%
  mutate(
    Stock_Status_CMSY= B.Bmsy
  ) %>%
  rename(year= yr)
```

```{r Create Stock Status Dataframe}
# Merge CMSY_Scores and Carta_Scores by genus, year, and region:
StockStatus_Scores0 <- full_join(
  CMSY_Scores,
  Carta_Scores,
  by = c("year",
         "estimated_genus_name")
)
# 2689 obs., 26 vars.

StockStatus_Scores0 <- full_join(
  StockStatus_Scores0,
  RAM_Scores,
  by = c("year",
         "estimated_genus_name")
)
# 2753 obs., 29 vars.
```

```{r Clean up this dataframe}
colnames(StockStatus_Scores0)

StockStatus_Scores<- StockStatus_Scores0 %>%
  select(year, Major_Functional_Group, estimated_genus_name, estimated_family_name,
         OHI_region_carta= OHI_region_est,
         Stock_Status_RAM, Stock_Status_Carta, Stock_Status_CMSY,
         Carta_provenance= carta_provenance, RAM_provenance, n_Formal_Assessments
         ) %>%
  mutate(Stock_Status_RAM_capped = pmin(Stock_Status_RAM, 1),
         Stock_Status_Carta_capped = pmin(Stock_Status_Carta, 1), 
         Stock_Status_CMSY_capped = pmin(Stock_Status_CMSY, 1))
```

# **DECISION POINT: Pick 1 Stock Status**
```{r Assign Stock Status 'Pick' column}
# A hierarchy of which score to pick!
StockStatus_Scores <- StockStatus_Scores %>%
  mutate(
    StockStatus_Pick = case_when(
      !is.na(Stock_Status_RAM_capped) ~ Stock_Status_RAM_capped,
      !is.na(Stock_Status_Carta_capped) ~ Stock_Status_Carta_capped,
      !is.na(Stock_Status_CMSY_capped) ~ Stock_Status_CMSY_capped,
      TRUE ~ NA_real_
    ),
    StockStatus_Pick = pmin(StockStatus_Pick, 1)  # optional: makes sure the values are capped at 1
  )

ScoringPeriod_low= 2018
ScoringPeriod_high= 2024

StockStatus_Scores<- StockStatus_Scores %>% 
  filter(year >= ScoringPeriod_low & year <= ScoringPeriod_high)
# 1593 obs. (down from 2753 obs.), 15 vars.
```


Sometimes Stock Status is reported at the year-genus-region level, but very often it's only reported at the year-genus level. I need to make a guide where I have each year-genus-region combination and when region-level details are missing, let's assume that the year-genus details apply to all regions.

A clean way to handle this is:
Create a complete grid of all year × genus × region combinations.
Join in the region-specific stock status values.
Join in the year-genus stock status values as a fallback.
Use coalesce() so that region-specific values take precedence, and genus-level values are used when region-level values are missing.

```{r}
guide <- expand_grid(
  year = unique(StockStatus_Scores$year),
  estimated_genus_name = unique(StockStatus_Scores$estimated_genus_name),
  OHI_region_carta = unique(StockStatus_Scores$OHI_region_carta)
) %>%
  filter(!is.na(OHI_region_carta))

# region-level records
region_status <- StockStatus_Scores %>%
  filter(!is.na(OHI_region_carta))

# genus-level records
genus_status <- StockStatus_Scores %>%
  filter(is.na(OHI_region_carta)) %>%
  select(
    year,
    estimated_genus_name,
    StockStatus_Pick
  ) %>%
  rename(StockStatus_genus = StockStatus_Pick)

guide_filled <- guide %>%
  left_join(
    region_status %>%
      select(
        year,
        estimated_genus_name,
        OHI_region_carta,
        StockStatus_Pick
      ),
    by = c("year", "estimated_genus_name", "OHI_region_carta")
  ) %>%
  left_join(
    genus_status,
    by = c("year", "estimated_genus_name")
  ) %>%
  mutate(
    StockStatus_final = coalesce(
      StockStatus_Pick,      # use region-specific if available
      StockStatus_genus      # otherwise use genus-level value
    )
  )
```



# STEP 3: Calculate GEOMETRIC MEAN of stock landings * Stock_Status ------------------------------

```{r Weights and geometric mean}
# Calculate the proportion of catch each "Stock" (i.e. genus) contributed to its regional total during that 5-year reference period.
colnames(StockStatus_Scores)

ReferencePeriod_low= 2018
ReferencePeriod_high= 2022

weights_data <- status_data %>%
dplyr::filter(year > 1990 & year < 2010) %>%
dplyr::group_by(stock_id, region_id) %>%
dplyr::mutate(StockCatch = sum(catch)) %>% # total tonnes across all years for each stock and region
dplyr::ungroup() %>%
dplyr::group_by(region_id) %>%
dplyr::mutate(TotalCatch = sum(catch)) %>% # total tonnes catch across all years and stocks for each region
dplyr::ungroup() %>%
dplyr::mutate(wprop = StockCatch / TotalCatch) %>% # proportion of catch each stock contributed to the total catch
dplyr::distinct(stock_id, region_id, wprop)

colnames(GOC_MarineFisheryLandings_summary)

Weights_data <- GOC_MarineFisheryLandings_summary %>%
  filter(year >= ReferencePeriod_low & year <= ReferencePeriod_high) %>%
  group_by(estimated_genus_name, OHI_region_est) %>%
  mutate(StockCatch_2018.22 = sum(total_weight_mt)) %>% # total tonnes across all years for each stock and region during this reference period, 2018-2022
  ungroup() %>%
  group_by(OHI_region_est) %>%
  mutate(TotalCatch_2018.22 = sum(total_weight_mt, na.rm = TRUE)) %>% # total tonnes catch across all years and stocks for each region
  ungroup() %>%
  mutate(wprop_2018.22 = StockCatch_2018.22 / TotalCatch_2018.22) %>% # proportion of catch each stock contributed to the total catch
  distinct(estimated_genus_name, OHI_region_est, wprop_2018.22)


# SANITY CHEC: Do the weights add up to 1?
TEST<- Weights_data %>%
  group_by(OHI_region_est) %>%
  summarize(sum_w = sum(wprop_2018.22, na.rm= TRUE))
# Yup, looks good!
```

```{r Add weights data to Stock Status df:}
StockStatus_Scores_finalized <- left_join(StockStatus_Scores, 
                                      Weights_data, 
                                      by = c("OHI_region_carta", "estimated_genus_name"))


# This is the geometric mean part:
StockStatus_GeoMean <- StockStatus_Scores_finalized %>%
  group_by(OHI_region_est, year) %>%
  summarize(
    GeoMean_StockStatus = prod(StockStatus_capped ^ wprop_2018.22, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r Save this dataframe}
write.csv(StockStatus_GeoMean, "/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_StockStatusScores_GeoMean_Jun012026.csv", row.names = FALSE)
```

```{r Plot Stock Status Geo Mean over time}
#### PLOT: Stock_Status (Geometric Mean) OVER TIME ####

#### Region Colors ####
Region_colors <- c(
  "royalblue",
  "skyblue",
  "orange",
  "sienna1",
  "plum",
  "purple4",
  "aquamarine3",
  "gold2",
  "tomato"
)

Regions_of_Interest <- c(
  "Region_1", "Region_2", "Region_3",
  "Region_4", "Region_5", "Region_6",
  "Region_7", "Region_8", "Region_9"
)

Regions_of_Interest<- tolower(Regions_of_Interest)

StockStatus_GeoMean_overTime <- StockStatus_GeoMean %>%
  filter(OHI_region_est %in% Regions_of_Interest) %>%
  ggplot(aes(
    x = yr,
    y = GeoMean_StockStatus,
    color = OHI_region_est
  )) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ OHI_region_est, scales = "fixed") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = Region_colors) +
  labs(
    title = expression(bold("FIS: Regional " * bolditalic("Stock_Status") * " Scores (Geometric Mean)")),
    y = "Stock_Status Score (from 0 to 1)",
    x = "Year"
  ) +
  ylim(0, 1.1) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 14, face = "bold")
    # plot.title = element_text(size = 18, face = "bold")
  )

StockStatus_GeoMean_overTime
  
  # Save as PNG
  # ggsave(
  #   filename = sprintf("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_StockStatus_Scores/StockStatus_GeoMeanScores_PlotOverTime_Dec162025.png"),
  #   plot = StockStatus_GeoMean_overTime,
  #   width = 12,
  #   height = 12,
  #   dpi = 300
  # )

```

```{r Plot CPUE Scores over time}
#### PLOT: CPUE (weighted average) OVER TIME ####

Region_colors <- c(
  "royalblue",
  "skyblue",
  "orange",
  "sienna1",
  "plum",
  "purple4",
  "aquamarine3",
  "gold2",
  "tomato"
)

Regions_of_Interest <- c(
  "Region_1", "Region_2", "Region_3",
  "Region_4", "Region_5", "Region_6",
  "Region_7", "Region_8", "Region_9"
)

CPUE_WeightedMean_overTime <- CPUE_Scores %>%
  filter(OHI_region_est %in% Regions_of_Interest) %>%
  ggplot(aes(
    x = yr,
    y = weighted_CPUE_acrossAllGroups,
    color = OHI_region_est
  )) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ OHI_region_est, scales = "fixed") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
  
  scale_color_manual(values = Region_colors) +
  
  labs(
    title = expression(bold("FIS: Regional " * bolditalic("CPUE") * " Scores relative to Baseline (2003–2013) (Weighted Average)")),
    y = "CPUE Scores (CPUE_present : CPUE_baseline capped at 1)",
    x = "Year"
  ) +
  ylim(0, 1.1) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 14, face = "bold")
    # plot.title = element_text(size = 18, face = "bold")
  )

CPUE_WeightedMean_overTime
```

```{r Save as PNG}
  # ggsave(
  #   filename = sprintf("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_CPUE_Scores/FIS_CPUEScores_WeightedMean_PlotOverTime_Dec162025.png"),
  #   plot = CPUE_WeightedMean_overTime,
  #   width = 12,
  #   height = 12,
  #   dpi = 300
  # )

```


# Merge Scores 
```{r Merge ALL dataframes for final OHI Score}
# Merge:
# lowercase to make sure matching works
CPUE_Scores <- CPUE_Scores %>%
  mutate(OHI_Region = tolower(OHI_Region))

OHI_Scores_df <- CPUE_Scores %>%
  full_join(
    StockStatus_GeoMean,
    by = c("Year", "OHI_Region")
  )

OHI_Scores_df <- OHI_Scores_df %>%
  mutate(FIS_Score = rowMeans(cbind(weighted_CPUE_acrossAllGroups, GeoMean_StockStatus), na.rm = TRUE))

# Multiply by 100 and round to second decimal place:
OHI_Scores_df$FIS_Score_Scale100 <- round(
  OHI_Scores_df$FIS_Score * 100,
  digits = 1
)
```

```{r Filter out the OHI_Scores_df and save it}
# SAVE THIS DATAFRAME
# write.csv(OHI_Scores_df, "/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_Scores_Jun012026.csv", row.names = FALSE)
```

```{r Graph FIS scores per region over time}
#### PLOT: FIS SCORES OVER TIME ####
Region_colors <- c(
  "royalblue",
  "skyblue",
  "orange",
  "sienna1",
  "plum",
  "purple4",
  "aquamarine3",
  "gold2",
  "tomato"
)

FIS_Scores_overTime<- OHI_Scores_df %>%
  filter(OHI_Region %in% Regions_of_Interest) %>%
  
  ggplot(aes(x = Year, y = FIS_Score_Scale100, color = OHI_Region)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ OHI_Region, scales = "fixed") +
  
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey50") +
  
  labs(
    title = "FIS Scores per Region Over Time",
    x = "Year",
    y = "FIS Score (0–100)"
  ) +
  
  ylim(50, 110) +
  
  scale_color_manual(values = Region_colors) +
  
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 16, face = "bold")
  )

FIS_Scores_overTime
```

```{r Save as PNG}
  # ggsave(
  #   filename = sprintf("/home/ferrer/ohi_goc_prep/FP/FIS/v2025/OHI_FIS_Scores/FIS_Scores_PlotOverTime_Dec172025.png"),
  #   plot = FIS_Scores_overTime,
  #   width = 12,
  #   height = 12,
  #   dpi = 300
  # )
```