################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```

## clean_names_lowercase function
```{r Load clean_names_lowercase function}
# Takes text and standardizes it, which we will need for merging later on:

clean_names_lowercase <- function(x) {

  x <- iconv(x, from = "", to = "UTF-8", sub = "")
  x <- stringi::stri_trans_general(x, "Latin-ASCII")

  x %>%
    tolower() %>%
    trimws()
}
```

## lowercase_function function
```{r lowercase_function}
# Turns all entries lower-case
lowercase_function<- function(x){
  if(is.character(x)){
    tolower(x)
  } else if (is.factor(x)){
    tolower(as.character(x))
  } else {
    x
  }
}
```


# STEP 1: Load in and summarize GOC_MarineLandings0
```{r Load in GOC_MarineLandings0 as necessary}
GOC_MarineLandings0<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Marine.Landings_Sector.Sorted.Georef_PATCHED_Jun032026.rds"))
# 3246583 obs., 58 var.

# NOTE TO SELF: This source file may change in later Summer / Fall run. - EMF, June 4 2026
```

# **DECISION POINT: Selecting which sectors to include**
```{r Select Sectors_of_Interest}
table(GOC_MarineLandings0$sector_estimate_draft)

# SANITY CHECK: Which species are included in the freshwater or marine sector?
TEST<- GOC_MarineLandings0 %>%
  filter(sector_estimate_draft== "freshwater or marine: fishery sector yet to be classified")

table(TEST$recorded_taxa_group)
# Species like Mojarra (the overwhelming majority of catch), trout, and catfish, which can move from freshwater to marine systems and are a bit ambiguous in terms of where they can be caught.


MarineFishery_Sectors<- c("marine: fishery sector yet to be classified",
                          "marine: small-scale capture",
                          "marine: large-scale capture")
  
PossibleMarineFishery_Sectors<- c("freshwater or marine: large-scale capture",
                          "freshwater or marine: small-scale capture",
                          "freshwater or marine: fishery sector yet to be classified")

GOC_MarineFisheryLandings0<- GOC_MarineLandings0 %>%
  filter(sector_estimate_draft %in% c(MarineFishery_Sectors, PossibleMarineFishery_Sectors))

# 3137044 obs., 58 var.
```

```{r Clean up names of GOC_MarineFisheryLandings}
GOC_MarineFisheryLandings <- GOC_MarineFisheryLandings0 %>%
  mutate(
    OHI_region_est = clean_names_lowercase(OHI_region_est),
    estimated_genus_name = clean_names_lowercase(estimated_genus_name),
    Major_Functional_Group = clean_names_lowercase(Major_Functional_Group)
  )

# 3137044 obs., 58 var.
```


# STEP 1: Generate GOC_Stocks_summary and GOC_Groups_summary ------------------------------

```{r Generate GOC_Stocks_summary}
colnames(GOC_MarineFisheryLandings)

GOC_Stocks_summary<- GOC_MarineFisheryLandings %>%
  select(year, 
         state, office_name_clean, landing_site_clean,
         GOC_est, GOC_Include_Exclude, OHI_region_est, OHI_region_name,
         eu_rnpa, eu_name, sector_estimate_draft, 
         Major_Functional_Group, recorded_taxa_group, estimated_taxa_group,
         recorded_taxa_name, recorded_taxa_name_clean,
         estimated_species_name, estimated_genus_name,
         landed_weight, live_weight, weight_kg_corrected, value
         )

GOC_Stocks_summary<- GOC_Stocks_summary %>%
  group_by(year, OHI_region_est, estimated_genus_name) %>%
  summarize(
    total_weight_mt= sum(weight_kg_corrected, na.rm = TRUE) / 1000,
    total_value_peso= sum(value, na.rm = TRUE)
  )
# 12696 obs.
```

```{r Generate GOC_Groups_summary}
colnames(GOC_MarineFisheryLandings)

GOC_Groups_summary<- GOC_MarineFisheryLandings %>%
  group_by(year, OHI_region_est, Major_Functional_Group) %>%
  summarize(
    total_weight_mt= sum(weight_kg_corrected, na.rm = TRUE) / 1000,
    total_value_peso= sum(value, na.rm = TRUE)
  )
# 2779 obs.
```

```{r How much catch is ungrouped unknown or not assigned to a region between 2018 and 2024 for CPUE?}
table(GOC_Groups_summary$Major_Functional_Group)

GOC_Groups_summary %>%
  filter(year >= 2018 & year <= 2024) %>%
  mutate(
    in_filter = is.na(Major_Functional_Group) |
                Major_Functional_Group == "unknown" |
                is.na(OHI_region_est)
  ) %>%
  group_by(year) %>%
  summarise(
    total_mass_mt = sum(total_weight_mt, na.rm = TRUE),
    mass_in_filter = sum(total_weight_mt[in_filter], na.rm = TRUE),
    mass_not_in_filter = sum(total_weight_mt[!in_filter], na.rm = TRUE),
    percent_unaccounted = (mass_in_filter/total_mass_mt)*100
  )

# Less than 1% unaccounted for by CPUE in any given year.
```

# STEP 2: Load and merge the various FIS scores ------------------------------

```{r Load the indicator dataframes as necessary}
RAM_Scores0<- read.csv("/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_RAM_Scores/FIS_RAM_Scores_Jun032026.csv")
# 216 obs., 14 vars.

Carta_Scores0<- read.csv("/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_Carta_Scores/FIS_CartaScores_Jun052026.csv")
# 1480 obs., 17 vars.

CMSY_Scores0<- read.csv("/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_CMSY_Scores/FIS_CMSY_Scores_Jun012026.csv")
# 1569 obs., 11 vars.
```

## RAM Data
```{r Augment the RAM dataframe by collapsing all stock info into one score per genus per year}
colnames(RAM_Scores0)

RAM_Scores <- RAM_Scores0 %>%
  group_by(year, Genus) %>%
  summarize(
    RAM_BBmsy = mean(RAM_BBmsy_final, na.rm = TRUE),
    n_Formal_Assessments = sum(!is.na(RAM_BBmsy_numeric)),
    .groups = "drop"
  ) %>%
  mutate(
    RAM_provenance = case_when(
      n_Formal_Assessments == 0 ~ "value was interpolated based on formal assessment from other years (RAM or other*)",
      n_Formal_Assessments == 1 ~ "value was sourced from 1 formal assessment (RAM or other*)",
      n_Formal_Assessments >= 2 ~ "value was sourced from 2 or more formal assessments (RAM or other*)"
    )
  )
# 168 obs., 5 var.
# The total number of observations is reduced because we are meaning across species to the genus level.

# lowercase to make sure matching works
RAM_Scores <- RAM_Scores %>%
  mutate(estimated_genus_name = tolower(Genus),
         Stock_Status_RAM= RAM_BBmsy) %>%
  select(-c(Genus, RAM_BBmsy))
# 168 obs., 6 var. 

# Add 0 helpful columns for merging:
# RAM_Scores$StockStatus_source<- c("formal assessment (RAM or other*)")
# RAM_Scores$StockStatus_est.level<- c("genus level, across the GOC")
# RAM_Scores$OHI_region_est<- c("GOC")
```

```{r Rename the Stock_Status column in Carta_Scores}
Carta_Scores<- Carta_Scores0 %>%
  mutate(
    Stock_Status_Carta= Stock_Status_final
  ) %>%
  select(-c(Stock_Status_final))

# Add 0 helpful column for merging:
# Carta_Scores$StockStatus_source<- c("Carta")
# Carta_Scores$StockStatus_est.level<- c("genus-region level")
```

```{r Rename the Stock_Status column in Stock_Status_CMSY}
CMSY_Scores<- CMSY_Scores0 %>%
  mutate(
    Stock_Status_CMSY= B.Bmsy
  ) %>%
  rename(year= yr)

# Add 1 helpful columns for merging:
# CMSY_Scores$StockStatus_source<- c("CMSY")
# CMSY_Scores$StockStatus_est.level<- c("genus level, across the GOC")
# CMSY_Scores$OHI_region_est<- c("GOC")
CMSY_Scores$CMSY_provenance<- c("CMSY model run from May 28")
```

# Join dataframes together the long way:

```{r Create a backbone to merge to and then add}
# Join in RAM data, merge by year x genus:
StockStatus_Scores0 <- GOC_Stocks_summary %>%
  left_join(RAM_Scores, by = c("year", "estimated_genus_name"))
# 12696 obs.

# Join in Carta data, merge is more specific by year x genus x region:
colnames(Carta_Scores)

StockStatus_Scores00 <- StockStatus_Scores0 %>%
  left_join(Carta_Scores, by = c("year", "estimated_genus_name", "OHI_region_est"))
# 12696 obs.

# Join in CMSY data, merge is more specific by year x genus x region:
colnames(CMSY_Scores)

StockStatus_Scores000 <- StockStatus_Scores00 %>%
  left_join(CMSY_Scores, by = c("year", "estimated_genus_name"))
# 12696 obs., 34 varsity.
```

```{r Clean up this dataframe}
colnames(StockStatus_Scores000)

StockStatus_Scores<- StockStatus_Scores000 %>%
  select(year, Major_Functional_Group, estimated_genus_name, estimated_family_name,
         OHI_region_est,
         total_weight_mt, total_value_peso,
         Stock_Status_RAM, Stock_Status_Carta, Stock_Status_CMSY,
         Carta_provenance= carta_provenance, RAM_provenance
         ) %>%
  mutate(Stock_Status_RAM_capped = pmin(Stock_Status_RAM, 1),
         Stock_Status_Carta_capped = pmin(Stock_Status_Carta, 1), 
         Stock_Status_CMSY_capped = pmin(Stock_Status_CMSY, 1))
```

# **DECISION POINT: Pick 1 Stock Status**
```{r Pick one StockStatus_final}
# A hierarchy of which score to pick!
StockStatus_Scores <- StockStatus_Scores %>%
  mutate(
    StockStatus_final = coalesce(
      Stock_Status_RAM_capped,
      Stock_Status_Carta_capped,
      Stock_Status_CMSY_capped
    ),
    StockStatus_source = case_when(
      !is.na(Stock_Status_RAM_capped)   ~ "RAM",
      !is.na(Stock_Status_Carta_capped) ~ "Carta",
      !is.na(Stock_Status_CMSY_capped)  ~ "CMSY",
      TRUE ~ NA_character_
    ),
    StockStatus_est.level = case_when(
      !is.na(Stock_Status_RAM_capped)   ~ "genus level",
      !is.na(Stock_Status_Carta_capped) ~ "genus x region level",
      !is.na(Stock_Status_CMSY_capped)  ~ "genus level",
      TRUE ~ NA_character_
    )
  )

# Still some NAs in here. Let's proceed.
```

```{r How much is left unassigned at the stock level?}
unique(StockStatus_Scores$estimated_genus_name)
# 196 unique genera in landings

StockStatus_Scores %>%
  filter(year >= 2018 & year <= 2024) %>%
  mutate(
    in_filter = is.na(StockStatus_final) |
                is.na(OHI_region_est)
  ) %>%
  group_by(year) %>%
  summarise(
    total_mass_mt = sum(total_weight_mt, na.rm = TRUE),
    mass_in_filter = sum(total_weight_mt[in_filter], na.rm = TRUE),
    mass_not_in_filter = sum(total_weight_mt[!in_filter], na.rm = TRUE),
    percent_unaccounted = (mass_in_filter/total_mass_mt)*100
  )

# Less than ~32% of total biomass unaccounted for by Stock Status in any given year.
# Estimates seem poorer in 2023 and 2024 (this is because genus estimates are poorer these years)...
```

```{r Save this dataframe as needed}
write.csv(StockStatus_Scores, 
          "/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/StockStatus_Estimates_Jun072026.csv", 
          row.names = FALSE)
```


# STEP 3: Calculate GEOMETRIC MEAN of stock landings * Stock_Status ------------------------------

```{r Load in this dataframe as needed}
StockStatus_Scores<- read.csv("/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/StockStatus_Estimates_Jun072026.csv")
```

```{r Weights and geometric mean}
# Calculate the proportion of catch each "Stock" (i.e. genus) contributed to its regional total during the study period:
colnames(StockStatus_Scores)

StudyPeriod_low= 2018
StudyPeriod_high= 2024

Weights_data <- GOC_Stocks_summary %>%
  filter(year >= StudyPeriod_low & year <= StudyPeriod_high) %>%
  group_by(estimated_genus_name, OHI_region_est) %>%
  mutate(StockCatch_mt = sum(total_weight_mt)) %>% # total tonnes across all years for each stock and region during the study period
  ungroup() %>%
  group_by(OHI_region_est) %>%
  mutate(TotalCatch_mt = sum(total_weight_mt, na.rm = TRUE)) %>% # total tonnes catch across all years and stocks for each region
  ungroup() %>%
  mutate(wprop_2018.24 = StockCatch_mt / TotalCatch_mt) %>% # proportion of catch each stock contributed to the total catch
  distinct(estimated_genus_name, OHI_region_est, wprop_2018.24)


# SANITY CHECK: Do the weights add up to 1?
TEST<- Weights_data %>%
  group_by(OHI_region_est) %>%
  summarize(sum_w = sum(wprop_2018.24, na.rm= TRUE))
# Yup, looks good!
```

```{r Add weights data to Stock Status df:}
colnames(Weights_data)

StockStatus_Scores_finalized <- left_join(StockStatus_Scores, 
                                      Weights_data, 
                                      by = c("OHI_region_est", "estimated_genus_name"))


# This is the geometric mean part:
StockStatus_GeoMean <- StockStatus_Scores_finalized %>%
  group_by(OHI_region_est, year) %>%
  summarize(
    GeoMean_StockStatus = prod(StockStatus_final ^ wprop_2018.24, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(StockStatus_SubScore= GeoMean_StockStatus * 100)
```

```{r Save this dataframe}
write.csv(StockStatus_GeoMean, 
          "/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_StockStatus_SubScores_Jun052026.csv", 
          row.names = FALSE)
```

# >> COMPLETE: StockStatus_SubScores!

```{r}
StockStatus_GeoMean<- read.csv("/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_StockStatus_SubScores_Jun052026.csv")
```


# PLOT: Stock_Status Sub-Score (Geometric Mean) over time
```{r Plot Stock Status Geo Mean over time}
# Region Colors:
Region_colors <- c(
  "royalblue",
  "skyblue",
  "orange",
  "sienna1",
  "plum",
  "purple4",
  "aquamarine3",
  "gold2",
  "tomato"
)

Regions_of_Interest <- c(
  "Region_1", "Region_2", "Region_3",
  "Region_4", "Region_5", "Region_6",
  "Region_7", "Region_8", "Region_9"
)

Regions_of_Interest<- tolower(Regions_of_Interest)

StockStatus_GeoMean_overTime <- StockStatus_GeoMean %>%
  filter(OHI_region_est %in% Regions_of_Interest,
         year>=2018 & year <= 2024) %>%
  ggplot(aes(
    x = year,
    y = StockStatus_SubScore,
    color = OHI_region_est
  )) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ OHI_region_est, scales = "fixed") +
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = Region_colors) +
  labs(
    title = expression(bold("FIS: Regional " * bolditalic("Stock_Status") * " Sub-Scores (Geometric Mean)")),
    y = "Stock_Status Sub-Score (from 0 to 1)",
    x = "Year"
  ) +
  ylim(0, 101) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 14, face = "bold")
  )

StockStatus_GeoMean_overTime
```

```{r Save this plot}
# Save as PNG
ggsave(
    filename = sprintf("/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_Visualizations_Jun2026/StockStatus_SubScores_plot_Jun042026.png"),
    plot = StockStatus_GeoMean_overTime,
    width = 12,
    height = 12,
    dpi = 300
  )
```

# Closer look into the scores of Region 1:
```{r Whats happening in Region 1 lately?}
colnames(StockStatus_Scores)

Region1_Look <- StockStatus_Scores %>%
  filter(year>= StudyPeriod_low & year<= StudyPeriod_high, 
         OHI_region_est == "region_1") %>%
  group_by(year, Major_Functional_Group) %>%
  summarize(
    mean_StockStatus = mean(StockStatus_final, na.rm = TRUE),
    mode_StockStatus_est.level =
      names(which.max(table(na.omit(StockStatus_source)))),
    .groups = "drop"
  )
# Looks like the change in small pelagics might be driving the changes, and those changes aren't only reflected in the CMSY.

ggplot(Region1_Look,
       aes(x = year,
           y = mean_StockStatus,
           color = mode_StockStatus_est.level,
           group = Major_Functional_Group)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  
  facet_wrap(~ Major_Functional_Group, scales = "free_y") +
  
  scale_color_brewer(palette = "Dark2") +
  
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.2)
  ) +
  
  labs(
    x = "Year",
    y = "Mean Stock Status",
    color = "Average assessment level",
    title = "Region 1: Mean Stock Status through Time by Functional Group",
    subtitle = "Diagnosing decline in FIS"
  ) +
  theme_bw() +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

```{r Let's continue this close look}
Region1_CloserLook <- StockStatus_Scores %>%
  filter(year>= StudyPeriod_low & year<= StudyPeriod_high, 
         OHI_region_est == "region_1",
         Major_Functional_Group == "Pelagic marine fish (excl. anchovy & sardine)") %>%
  group_by(year, estimated_genus_name) %>%
  summarize(
    mean_StockStatus = mean(StockStatus_final, na.rm = TRUE),
    mode_StockStatus_est.level =
      names(which.max(table(na.omit(StockStatus_source)))),
    .groups = "drop"
  )

ggplot(Region1_CloserLook,
       aes(x = year,
           y = mean_StockStatus,
           color = mode_StockStatus_est.level,
           group = estimated_genus_name)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  
  facet_wrap(~ estimated_genus_name, scales = "free_y") +
  
  scale_color_brewer(palette = "Dark2") +
  
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.2)
  ) +
  
  labs(
    x = "Year",
    y = "Mean Stock Status",
    color = "Average assessment level",
    title = "Region 1: Mean Stock Status through Time by Functional Group",
    subtitle = "Diagnosing decline in FIS"
  ) +
  theme_bw() +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "bottom"
  )

# Here we go. Catch for scomberomorus has been incredibly high in recent years.

Region1_CloserLook <- StockStatus_Scores %>%
    filter(year>= StudyPeriod_low & year<= StudyPeriod_high, 
         OHI_region_est == "region_1",
         estimated_genus_name == "scomberomorus")

# This decline could be real, frankly. Look at the declining Carta and CMSY scores.
```


# STEP 4: Calculate GEOMETRIC MEAN of stock landings * CPUE ------------------------------
```{r Load in CPUE data as necessary}
CPUE_Scores0<- read.csv("/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_CPUE_Scores/FIS_CPUE_Scores_Jun032026.csv")
# 101 obs., 11 vars.

# Winnow down and clean this dataframe:
colnames(CPUE_Scores0)

CPUE_Scores<- CPUE_Scores0 %>%
  select(year, Major_Functional_Group, CPUE_score, CPUE_score_capped) %>%
  filter(year >= StudyPeriod_low & year <= StudyPeriod_high) %>%
  mutate(
    Major_Functional_Group = clean_names_lowercase(Major_Functional_Group)
    )
# 101 obs., 6 vars.
```

```{r Weights and geometric mean}
# Calculate the proportion of catch each "Stock" (i.e. genus) contributed to its regional total during that 5-year reference period.
colnames(CPUE_Scores)

StudyPeriod_low= 2018
StudyPeriod_high= 2024

GroupWeights_data <- GOC_Groups_summary %>%
  filter(year >= StudyPeriod_low & year <= StudyPeriod_high) %>%
  
  group_by(Major_Functional_Group, OHI_region_est) %>%
  # total tonnes across all years for each stock and region during the study period 2018-2024
  mutate(GroupCatch_2018.24 = sum(total_weight_mt)) %>% 

  ungroup() %>%
  
  group_by(OHI_region_est) %>%
  mutate(TotalCatch_2018.24 = sum(total_weight_mt, na.rm = TRUE)) %>% # total tonnes catch across all years and stocks for each region
  ungroup() %>%
  
  mutate(wprop_2018.24 = GroupCatch_2018.24 / TotalCatch_2018.24) %>% # proportion of catch each stock contributed to the total catch
  distinct(Major_Functional_Group, OHI_region_est, wprop_2018.24)


# SANITY CHECK: Do the weights add up to 1?
TEST<- GroupWeights_data %>%
  group_by(OHI_region_est) %>%
  summarize(sum_w = sum(wprop_2018.24, na.rm= TRUE))
# Yup, looks good!
```

```{r Add weights data to Stock Status df:}
colnames(GOC_Groups_summary)
colnames(GroupWeights_data)
colnames(CPUE_Scores)

CPUE_Scores_GeoMean0 <- left_join(GOC_Groups_summary %>% filter(year>= StudyPeriod_low & year<= StudyPeriod_high),
                                   GroupWeights_data,
                                   by = c("Major_Functional_Group", "OHI_region_est"))
# 818 obs., 6 vars.

CPUE_Scores_GeoMean00 <- left_join(CPUE_Scores_GeoMean0,
                                   CPUE_Scores,
                                   by = c("year", "Major_Functional_Group"))
# 818 obs., 8 vars.

# This is the geometric mean part:
CPUE_Scores_GeoMean <- CPUE_Scores_GeoMean00 %>%
  group_by(OHI_region_est, year) %>%
  summarize(
    GeoMean_CPUE = prod(CPUE_score_capped ^ wprop_2018.24, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    CPUE_SubScore= GeoMean_CPUE * 100
  )
```

# >> COMPLETE: CPUE_SubScores!

# PLOT: CPUE Sub-Score (Geometric Mean) over time
```{r Plot CPUE Scores over time}
Region_colors <- c(
  "royalblue",
  "skyblue",
  "orange",
  "sienna1",
  "plum",
  "purple4",
  "aquamarine3",
  "gold2",
  "tomato"
)

Regions_of_Interest <- c(
  "region_1", "region_2", "region_3",
  "region_4", "region_5", "region_6",
  "region_7", "region_8", "region_9"
)

CPUE_GeoMean_overTime <- CPUE_Scores_GeoMean %>%
  filter(OHI_region_est %in% Regions_of_Interest) %>%
  ggplot(aes(
    x = year,
    y = CPUE_SubScore,
    color = OHI_region_est
  )) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  
  facet_wrap(~ OHI_region_est, scales = "fixed") +
  
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
  
  scale_color_manual(values = Region_colors) +
  
  labs(
    title = expression(bold("FIS: Regional " * bolditalic("CPUE") * " Sub-Scores (Geometric Mean)")),
    y = "CPUE Scores (CPUE_present : CPUE_baseline capped at 1)",
    x = "Year"
  ) +
  ylim(0, 101) +
  
  theme_bw(base_size = 14) +
  
  theme(
    legend.position = "none",
    strip.text = element_text(size = 14, face = "bold")
    # plot.title = element_text(size = 18, face = "bold")
  )

CPUE_GeoMean_overTime
```

```{r Save this plot}
# Save as PNG
ggsave(
    filename = sprintf("/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_Visualizations_Jun2026/CPUE_SubScores_plot_Jun042026.png"),
    plot = CPUE_GeoMean_overTime,
    width = 12,
    height = 12,
    dpi = 300
  )
```


# Merge Scores:
# **DECISION POINT: CPUE versus StockStatus weights!**
```{r Merge ALL dataframes for final OHI Score}
# Check dimensions:
dim(CPUE_Scores_GeoMean) # 70
colnames(CPUE_Scores_GeoMean)

dim(StockStatus_GeoMean) # 255
colnames(StockStatus_GeoMean)

# Merge:
FIS_Scores <- CPUE_Scores_GeoMean %>%
  full_join(
    StockStatus_GeoMean %>% filter(year >= StudyPeriod_low & year <= StudyPeriod_high),
    by = c("year", "OHI_region_est")
  ) %>%
  relocate(StockStatus_SubScore, .before= CPUE_SubScore) %>%
  relocate(GeoMean_StockStatus, .before= GeoMean_CPUE)
# 70 obs. Good! 

# FIS score is 1/3 CPUE versus 2/3 StockStatus:

FIS_Scores<- FIS_Scores %>%
  mutate(
    FIS_Score = (2/3)*StockStatus_SubScore + (1/3)*CPUE_SubScore 
  )
```

```{r Save FIS_Scores}
write.csv(FIS_Scores, 
          "/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_Scores_Jun052026.csv", 
          row.names = FALSE)
# 70 obs., 7 vars.
```

# >> COMPLETE: FIS_Scores!

# PLOT: FIS SCORES OVER TIME
```{r Graph FIS scores per region over time}
Region_colors <- c(
  "royalblue",
  "skyblue",
  "orange",
  "sienna1",
  "plum",
  "purple4",
  "aquamarine3",
  "gold2",
  "tomato"
)

FIS_Scores_overTime<- FIS_Scores %>%
  filter(OHI_region_est %in% Regions_of_Interest) %>%
  
  ggplot(aes(x = year, y = FIS_Score, color = OHI_region_est)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  
  facet_wrap(~ OHI_region_est, scales = "fixed") +
  
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey50") +
  
  labs(
    title = "FIS Scores per Region Over Time",
    x = "Year",
    y = "FIS Score (0–100)"
  ) +
  
  ylim(0, 101) +
  
  scale_color_manual(values = Region_colors) +
  
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 16, face = "bold")
  )

FIS_Scores_overTime
```

```{r Save as PNG}
  ggsave(
    filename = sprintf("/home/ferrer/ohi_goc_prep/FP/FP_Scores/FIS_Scores/FIS_Visualizations_Jun2026/FIS_Scores_plot_Jun052026.png"),
    plot = FIS_Scores_overTime,
    width = 12,
    height = 12,
    dpi = 300
  )
```