################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
library(sf) # for work on shape fines
# Visualization
library(ggplot2)
library(colorspace)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/_habitats/")
```

```{r Load clean_names_lowercase function}
clean_names_lowercase <- function(x) {

  x <- iconv(x, from = "", to = "UTF-8", sub = "")
  x <- stringi::stri_trans_general(x, "Latin-ASCII")

  x %>%
    tolower() %>%
    trimws()
}
```


# STEP 0: Load in Region_Key
```{r Region key}
Region_Key0<- read.csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Keys/Spatial_Keys/RegionKey_FINAL_May262026.csv")

colnames(Region_Key0)

Region_Key<- Region_Key0 %>%
  select(OHI_region= OHI_region_est, OHI_region_name) %>%
  distinct()
```

# STEP 1: Load in NRSI_data0

```{r Load in NRSI_data0}
NRSI_data0<- read.csv("/home/shares/ohi/OHI_GOC/_raw_data/dataMares/nrsi_per_reef_year_ALL_not_filtered.csv")

NRSI_data<- NRSI_data0
```

```{r Looking for sites in Nayarit}
unique(NRSI_data$reef)
table(NRSI_data$reef)

Fully_Protected_Reefs<- NRSI_data %>%
  filter(protection== "Fully Protected")

Not_Protected_Reefs<- NRSI_data %>%
  filter(protection== "Not Protected")

# Can't seem to find the sites near Nayarit... - Aug 5 2026
# Reached out to Eduardo, now I found it - Aug 10 2026
```

# STEP 2: Add lat-long details to NRSI_data

```{r Loading in OHI regions and buffer shape files}
# OHI regions:
OHI_regions_metadata <- readLines("/home/shares/ohi/OHI_GOC/spatial/OHI_regions/marine/ohi_goc_final_071025.prj")
# CRS: EPSG:4326 in practice, poorly labeled in meta-data.

OHI_regions_shape0 <- sf::st_read(
  "/home/shares/ohi/OHI_GOC/spatial/OHI_regions/marine/ohi_goc_final_071025.shp",
  options = "ENCODING=LATIN1"
)

st_crs(OHI_regions_shape0)

OHI_regions_shape0 <- st_make_valid(OHI_regions_shape0)

# Quick plot:
OHI_Region_plot<- ggplot(OHI_regions_shape0) +
  geom_sf(aes(fill = rgn_id)) +
  theme_bw()

OHI_Region_plot

# 50km buffer:
OHI_buffer_metadata <- readLines("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer/inland_50km_goc.prj")
# Projected CRS name: Mexico_ITRF2008_LCC

OHI_buffer_shape0 <- sf::st_read(
  "/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer/inland_50km_goc.shp",
  options = "ENCODING=LATIN1"
)

st_crs(OHI_buffer_shape0)

OHI_buffer_shape0 <- st_make_valid(OHI_buffer_shape0)

# Quick plot:
OHI_Buffer_plot<- ggplot(OHI_buffer_shape0) +
  geom_sf(aes(fill = rgn_id)) +
  theme_bw()

OHI_Buffer_plot
```

```{r Some quick cleaning}
# Let's use CRS 4326 for all projections now.
crs_use = 4326

# Change CRS on OHI_buffer_shape to match CRS on OHI_regions_shape
OHI_regions_shape <- st_transform(OHI_regions_shape0, crs_use)

# Change CRS on OHI_buffer_shape to match CRS on OHI_regions_shape
OHI_buffer_shape <- st_transform(OHI_buffer_shape0, crs_use)

# Remove columns and extract useful info:
OHI_buffer_shape<- OHI_buffer_shape0 %>%
  select(-c(estado))

additional_info<- OHI_buffer_shape0 %>%
  select(rgn_id, ecorgn) %>%
  st_drop_geometry()

# Add new columns to OHI_regions_shape
OHI_regions_shape <- OHI_regions_shape %>%
  left_join(
    additional_info,
    by = "rgn_id"
  ) %>%
  mutate(
    location= c("sea")
  ) %>%
  relocate(location, .after= rgn_id)

rm(additional_info)
```

```{r Join OHI geometric key to NRSI_data}
# Step 0 — define a Coordinate Reference System (CRS) once and stick to it:
crs_use <- 4326

# Alternative could be: CRS 5070 is the USA Contiguous Albers Equal Area (good proxy for northern Mexico too). EPSG:5070 is centered on the U.S., so Baja / Sonora are near the edge of its optimal zone.

# Step 1 — Convert master table to spatial points
# Turn NRSI_data into an sf object
colnames(NRSI_data)

master_sf <- NRSI_data %>%
  filter(!is.na(longitude), !is.na(latitude)) %>%
  st_as_sf(coords = c("longitude", "latitude"), 
           crs = crs_use,
           remove = FALSE) %>%
  st_transform(crs_use)

# Step 2 — Transform polygons ONCE
buffer_sf <- OHI_buffer_shape %>%
  st_transform(crs_use)

sea_sf <- OHI_regions_shape %>%
  st_transform(crs_use)

# Step 3 — Spatial join to land buffer polygons
# Join points → OHI_buffer_shape
master_land <- st_join(
  master_sf,
  buffer_sf,
  join = st_within,
  left = TRUE
)

# Re-name columns to be safe:
master_land <- master_land %>%
  rename(
    rgn_id_land = rgn_id,
    ecorgn_land = ecorgn,
    location_land = location
  )

# Step 5 — Spatial join to sea regions
# Join same (or filtered) points → OHI_regions_shape
master_all <- st_join(
  master_land,
  sea_sf,
  join = st_within,
  left = TRUE
)

# Step 6 — Fix duplicate column names cleanly
master_all <- master_all %>%
  rename(
    rgn_id_sea = rgn_id,
    ecorgn_sea = ecorgn,
    location_sea = location
  )
```

# NOTE TO SELF: This ^ is a pretty sloppy join. Perfect this later. - Aug 5 2026


# STEP 3: Compute NRSI_Scores per region ------------------------------
Scores = average NSRI score by OHI region

```{r Summarize NRSI and join Region_Key}
colnames(master_all)
table(master_all$rgn_id_sea)

# Region_2 Region_3 Region_4 Region_6 Region_7 Region_9 
#       95       76     1391       18        2       53 

NRSI_Scores_summarized<- master_all %>%
  group_by(year, rgn_id_sea) %>%
  summarize(
    mean_NRSI = mean(nrsi_01, na.rm = TRUE),
    n_reefs = n_distinct(reef, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    OHI_region = clean_names_lowercase(rgn_id_sea)
  )
# 55 obs., 6 vars.

colnames(NRSI_Scores_summarized)

# Generate NRSI_Scores with region details merged in:
NRSI_Scores0<- NRSI_Scores_summarized %>%
  select(OHI_region, year, mean_NRSI, n_reefs)
# 55 obs., 5 vars.

NRSI_Scores0 <- sf::st_drop_geometry(NRSI_Scores0)
# 55 obs., 4 vars.
```

```{r Gapfill NRSI_Scores0}

# First, expand the NRSI dataframe to include one row per year per region, even if that region has no data: 
NRSI_Scores <- NRSI_Scores0 %>%
  complete(
    OHI_region = c(paste0("region_", 1:9)),
    year = full_seq(year, 1)
  )  %>%
  left_join(
    Region_Key,
    by= c("OHI_region" = "OHI_region_est")
  )

# 252 obs., 5 vars.

# Gapfill missing years and NRSI scores using the zoo package:
NRSI_Scores <- NRSI_Scores %>%

  # Fill variables that don't change within a region
  group_by(OHI_region) %>%
  fill(OHI_region_name, .direction = "downup") %>%
  mutate(
    provenance = if_else(
      is.na(mean_NRSI),
      "gapfilled",
      "observed"
    ),
    
    # Interpolate the missing NRSI values
    mean_NRSI = na.approx(
      mean_NRSI,
      x = year,
      na.rm = FALSE
    )
  ) %>%
  ungroup()
# 252 obs., 6 vars.
```

## DECISION POINT: Take the global average across regions for which we have actual data:
In this case, we have data from regions 2, 3, 4, 6, 7, and 9, but the reefs from 2-6 are best sampled.
```{r Take the global average}
unique(NRSI_Scores$OHI_region)

# Regions_of_Interest<- c("region_2", "region_3", "region_4", "region_6")

Regions_of_Interest <- NRSI_Scores %>%
  filter(provenance == "observed") %>%
  distinct(OHI_region) %>%
  pull(OHI_region)

NRSI_global_mean <- NRSI_Scores %>%
  filter(OHI_region %in% Regions_of_Interest) %>%
  group_by(year) %>%
  summarise(
    mean_NRSI = if (all(is.na(mean_NRSI))) {
      NA_real_
    } else {
      mean(mean_NRSI, na.rm = TRUE)
    },
    n_reefs = if (all(is.na(n_reefs))) {
      NA_integer_
    } else {
      sum(n_reefs, na.rm = TRUE)
    },
    
    # n_regions = n_distinct(OHI_region, na.rm = TRUE),
    
    # n_regions_observed field counts the distinct OHI regions whose provenance is "observed" in each year:
    n_regions_observed = n_distinct(
      OHI_region[provenance == "observed"]
    ),
    
    #Derive the global provenance from the provenance of the regional observations contributing to each year:
    provenance = case_when(
      all(provenance == "observed", na.rm = TRUE) ~ "observed",
      all(provenance == "gapfilled", na.rm = TRUE) ~ "gapfilled",
      TRUE ~ "partially_observed"
    ),

    .groups = "drop"
  )


NRSI_global_mean$OHI_region<- c("global average")
NRSI_global_mean$OHI_region_name<- c("global average")


NRSI_Scores_all <- bind_rows(NRSI_Scores, NRSI_global_mean)
# 280 obs., 7 vars. 

colnames(NRSI_Scores_all)
unique(NRSI_Scores_all$provenance)

# NA out provenance values for which mean_NRSI is = NA.
NRSI_Scores_all <- NRSI_Scores_all %>%
  mutate(
    provenance = if_else(
      is.na(mean_NRSI),
      NA,
      provenance
    )
  )
# 280 obs., 7 vars. 
```

# SAVE NRSI_Scores_all ------------------------------

```{r Save NRSI_Scores}
write.csv(NRSI_Scores_all, 
          "/home/ferrer/ohi_goc_prep/BD/HAB/HAB_rocky_reefs/NRSI_Scores_df_Aug312026.csv", 
          row.names = FALSE)
# 280 obs., 7 vars. 
```

# PLOT mean_NRSI over time ------------------------------

```{r Load in NRSI_Scores_all as needed}
NRSI_Scores_all0<- read.csv("/home/ferrer/ohi_goc_prep/_habitats/HAB_rocky_reefs/Exploratory_Research/NRSI_Scores_df_Aug312026.csv")

NRSI_Scores_all<- NRSI_Scores_all0
```

```{r Plot mean_NRSI over time}
colnames(NRSI_Scores_all)

# Re-level a few things for plotting purposes:
NRSI_Scores_all <- NRSI_Scores_all %>%
  mutate(
    OHI_region = factor(
      OHI_region,
      levels = c(
        setdiff(unique(OHI_region), "global average"),
        "global average"
      )
    ),
    OHI_region_name = factor(
      OHI_region_name,
      levels = c(
        setdiff(unique(OHI_region_name), "global average"),
        "global average"
      )
    ),
    provenance = factor(
      provenance,
      levels = c(
        "observed",
        "partially_observed",
        "gapfilled"
      )
    )
  )

# Plot 1, panel per region:

Region_colors<- c(
  "Western Upper Gulf" = "#C03729",
  "Western Midriff Islands" = "#E68C7C",
  "Western Central Gulf" = "#FC8F24",
  "Western Southern Gulf" = "#dcaa38",
  "Eastern Upper Gulf" = "#cbd2a0",
  "Eastern Midriff Islands" = "#98a25a",
  "Southern Sonora Coast" = "#747428",
  "Sinaloa Coast" = "#627391",
  "Southeastern Gulf" = "#8d61a9",
  "global average" = "darkgrey")


NRSI_Scores_OverTime <- NRSI_Scores_all %>%
  filter(!is.na(OHI_region)) %>%
  
  ggplot(
    aes(
      x = year,
      y = mean_NRSI,
      color = OHI_region_name
    )
  ) +
  
  geom_line() +
  
  geom_point(
    aes(
      shape = provenance,
      fill = if_else(
        provenance == "observed",
        OHI_region_name,
        NA_character_
      )
    ),
    size = 2,
    stroke = 1.2
  ) +
  
  facet_wrap(~ OHI_region_name) +
  
  labs(
    title = "NRSI scores per Region",
    subtitle = "updated: Aug 31 2026",
    caption = "global average equals mean NRSI scores across all observed regions",
    y = "Normalized Reef Status Indicator (0 to 1)",
    x = "Year",
    color = "Region",
    shape = "Data provenance"
  ) +
  
  scale_color_manual(
    values = Region_colors
  ) +
  
  scale_fill_manual(
    values = Region_colors,
    na.value = "white",
    guide = "none"
  ) +
  
  scale_shape_manual(
    values = c(
      observed = 22,
      partially_observed = 22,
      gapfilled = 22
    ),
    labels = c(
      observed = "Observed",
      partially_observed = "Partially observed",
      gapfilled = "Gap-filled"
    )
  ) +
  
  guides(
    color = guide_legend(
      override.aes = list(
        shape = 0,
        fill = NA
      )
    ),
    shape = guide_legend(
      override.aes = list(
        color = "black",
        fill = c("black", "white", "white", "white")
      )
    )
  ) +
  
  theme_bw(base_size = 14) +
  theme(
    strip.text = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 16, face = "bold")
  ) +
  
  ylim(c(0,1))

NRSI_Scores_OverTime

```


```{r Save plot as PNGs}
ggsave(
  plot = NRSI_Scores_OverTime,
  filename = sprintf("/home/ferrer/ohi_goc_prep/BD/HAB/HAB_rocky_reefs/NRSI_Scores_plot_Aug312026.png"),
  width = 12,
  height = 12,
  dpi = 300
)
```

# Generate ROK_Scores ------------------------------

DECISION POINT: Reference point is equal to the highest NRSI value ever observed

```{r Calculate the Rocky Reef reference point}
ROK_ref_point<- max(NRSI_Scores_all$mean_NRSI[NRSI_Scores_all$provenance=="observed"], na.rm= TRUE)
# 0.6034904
```

```{r Calculate NRSI 3-year running mean}
# Calculate ROK_Scores as the 3-year running mean of mean_NRSI per region.

NRSI_Scores_all <- NRSI_Scores_all %>%
  arrange(OHI_region, year) %>%
  group_by(OHI_region) %>%
  mutate(
    mean_NRSI_3yr = rollapply(
      mean_NRSI,
      width = 3,
      FUN = mean,
      na.rm = TRUE,
      fill = NA,
      align = "right"
    )
  ) %>%
  ungroup()
# 280 obs., 8 vars. 
```

```{r Calculate original ROK_Scores}
# Calculate the Rocky Reef (ROK) score as the 3-year running mean divided by the reference point x 100:
ROK_Scores<- NRSI_Scores_all %>%
  mutate(
    ROK_score_raw = (mean_NRSI_3yr / ROK_ref_point) * 100
  )
# 280 obs., 9 vars.
```

```{r Calculate final ROK_Scores swapping out global mean as needed}
ROK_Scores <- ROK_Scores %>%
  group_by(year) %>%
  mutate(
    ROK_score_final = case_when(
      OHI_region %in% c("region_2", "region_3", "region_4", "region_6") ~ ROK_score_raw,
      OHI_region %in% c("region_1", "region_5", "region_7", "region_8", "region_9", "global average") ~
        first(ROK_score_raw[OHI_region == "global average"]),
      TRUE ~ NA_real_
    )
  ) %>%
  ungroup()
# 280 obs., 10 vars.
```



# SAVE ROK_Scores ------------------------------

```{r Save ROK_Scores}
write.csv(ROK_Scores, 
          "/home/ferrer/ohi_goc_prep/BD/HAB/HAB_rocky_reefs/ROK_Scores_1.0_df_Aug312026.csv", 
          row.names = FALSE)
# 280 obs., 10 vars. 
```


# PLOT ROK_Scores over time ------------------------------

```{r Load in ROK_Scores as needed}
ROK_Scores0<- read.csv("/home/ferrer/ohi_goc_prep/_habitats/HAB_rocky_reefs/ROK_outputs/ROK_Scores_1.0_df_Aug312026.csv")

ROK_Scores<- ROK_Scores0
```


```{r Plot mean_NRSI over time}
colnames(ROK_Scores)

# Re-level a few things for plotting purposes:
ROK_Scores <- ROK_Scores %>%
  mutate(
    OHI_region = factor(
      OHI_region,
      levels = c(
        setdiff(unique(OHI_region), "global average"),
        "global average"
      )
    ),
    OHI_region_name = factor(
      OHI_region_name,
      levels = c(
        setdiff(unique(OHI_region_name), "global average"),
        "global average"
      )
    ),
    provenance = factor(
      provenance,
      levels = c(
        "observed",
        "partially_observed",
        "gapfilled"
      )
    )
  )

# Plot 1, panel per region:

unique(ROK_Scores$OHI_region_name)

Region_colors<- c(
  "Western Upper Gulf" = "#C03729",
  "Western Midriff Islands" = "#E68C7C",
  "Western Central Gulf" = "#FC8F24",
  "Western Southern Gulf" = "#dcaa38",
  "Eastern Upper Gulf" = "#cbd2a0",
  "Eastern Midriff Islands" = "#98a25a",
  "Southern Sonora Coast" = "#747428",
  "Sinaloa Coast" = "#627391",
  "Southeastern Gulf" = "#8d61a9",
  "global average" = "darkgrey")


ROK_Scores_OverTime <- ROK_Scores %>%
  filter(!is.na(OHI_region) & !OHI_region=="global average") %>%
  
  ggplot(
    aes(
      x = year,
      y = ROK_score_final,
      color = OHI_region_name
    )
  ) +
  
  geom_line() +
  
  geom_point(
    shape = 16,       # filled circle
    size = 2,
    stroke = 1.2
  ) +
  
  facet_wrap(~ OHI_region_name) +
  
  labs(
    title = "Rocky Reefs Scores per Region",
    subtitle = "updated: Aug 31 2026",
    caption = "ROK score in regions 1, 5, 7, 8, and 9 are equal to the global average",
    y = "ROK Score (0 to 100)",
    x = "Year",
    color = "Region"
  ) +
  
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey50") +
  
  ylim(0, 101) +
  
  scale_color_manual(
    values = Region_colors
  ) +
  
  theme_bw(base_size = 14) +
  theme(
    strip.text = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 16, face = "bold")
  )

ROK_Scores_OverTime

```

```{r Save plot as PNGs}
ggsave(
  plot = ROK_Scores_OverTime,
  filename = sprintf("/home/ferrer/ohi_goc_prep/BD/HAB/HAB_rocky_reefs/ROK_Scores_1.0_plot_Aug312026.png"),
  width = 12,
  height = 12,
  dpi = 300
)
```


# Current Spatial Extent of Rocky Reef (not needed for BD scoring but here if it's helpful):

```{r Plot Rocky Reef Sargassum}
RockyReef_shape <- st_read("/home/shares/ohi/OHI_GOC/_raw_data/Hem_data/goc_habitat_data/Arrecife rocoso/Arrecife rocoso.shp") |> st_zm()

plot(st_geometry(RockyReef_shape))

RockyReef_shape <- RockyReef_shape |>
  st_transform(st_crs(OHI_regions))
```

```{r Explore rocky sargassum}
# Read in Arrecife rocoso (rocky reef) data
rocky_reef <- st_read(here("/home/shares/ohi/OHI_GOC/_raw_data/Hem_data/goc_habitat_data/Arrecife rocoso/Arrecife rocoso.shp")) |> 
  st_make_valid()
```


```{r Read in OHI region boundaries}
# Read in OHI region boundaries
OHI_regions <- st_read(here("spatial", "ohi_regions", "goc_ohi_rgns.shp")) |> st_make_valid()

plot(st_geometry(OHI_regions))
st_crs(OHI_regions)
st_crs(RockyReef_shape)

st_bbox(OHI_regions)
st_bbox(RockyReef_shape)

```

```{r Plot}
library(tmap)

tmap_mode("view")

tm_shape(OHI_regions) +
  tm_polygons(col = "white", border.col = "gray", outline = 0.2) +
tm_shape(RockyReef_shape) +
  tm_polygons(col = "blue", border.col = "blue")
```
