---
title: "OHI GoC: Gulf news engagement per capita, 2019-2025 (GDELT GKG)"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
editor:
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options:
  chunk_output_type: console
---

# What this script does

This script builds a news **engagement-per-capita** indicator for the Gulf of California. The default path now **downloads free GKG zip files** and filters them in R. BigQuery is optional and currently off (the free 1 TB quota on `ohi-goc-cce-gdelt` is used up).

GDELT 2.0 is an open index of worldwide news. About every 15 minutes it reads articles in many languages, translates most non-English coverage into English, finds people, places, organizations, events, and themes, and publishes the result as open metadata.

**How to run.** Restart R, then **Run All**. The script downloads **January 2024** GKG 15-minute zips from `data.gdeltproject.org` (~3,000 files, often 1 to 3 hours), keeps only **Spanish originals** that use **Spanish Gulf names** (`golfo de california`, `mar de cortés` / `cortes` / `cortez`) **and** **Spanish marine/coast terms** (pesca, manglar, vaquita, playa, and related words), then clips points to the 50 km study area. Re-run is safe: finished timestamps are skipped. Change `zip_start` / `zip_end` to try another month. Monthly GKG files go to `/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026/int/gkg_by_month/`. Only region-year current-status scores go to `.../v2026/output/`.

Every search is limited in **two** ways, because a map clip alone is not enough:

1. **The article must name the Gulf.** GDELT cannot nest two OR lists, so each query pairs one ocean-engagement topic list with one required Gulf / Sea of Cortez phrase (`"gulf of california"`, `"golfo de california"`, `"sea of cortez"`, `"mar de cortes"`). A protest in La Paz about an inland road, with no Gulf or ocean language, should not come through.
2. **The mapped location must sit in the study area.** After the search, we keep only points inside the OHI Gulf marine regions, **including islands**, plus the **50 km inland buffer**:

`/home/shares/ohi/OHI_GOC/spatial/OHI_regions/marine_and_inland50.gpkg`

The topic lists are meant to reflect **the marine environment and people’s engagement with the ocean**: protest and *manifestaciones* tied to the Gulf, megaprojects on the coast, and coverage of fishing, mangroves, beaches, whales, vaquita, and related coastal care. It is still **not** a finished Coastal Community Engagement score. GDELT measures news language, not belonging. More articles can mean more conflict or more tourism press, not more sense of place.

**Spanish coverage and Spanish ANEW.** Gulf news will be in Spanish. The GEO and DOC searches can still use English keywords because GDELT translates Spanish articles into English for search, but we restrict to **Spanish-language originals** (`sourcelang:spanish`). Emotion is **not** taken from that English translation. For Spanish articles, GDELT automatically scores the original Spanish text with the **Spanish adaptation of ANEW** inside GCAM (Global Content Analysis Measures). Those scores live in the GKG `GCAM` field as dictionary 22 (`v22.1` valence, `v22.2` arousal, `v22.3` dominance, plus familiarity, concreteness, and imageability). The GEO and DOC APIs do not return GCAM. Section 8 reads Spanish ANEW from BigQuery GKG. An optional 24-hour GKG zip loop remains in section 6 if `download_gkg <- TRUE`.

Cite Spanish ANEW as: Redondo, Jaime, Isabel Fraga, Isabel Padrón, and Montserrat Comesaña. 2007. "The Spanish adaptation of ANEW (affective norms for English words)." *Behavior Research Methods* 39 (3): 600-605.

Codebook: <http://data.gdeltproject.org/documentation/GCAM-MASTER-CODEBOOK.TXT>
Spanish ANEW in GCAM: <https://blog.gdeltproject.org/spanish-edition-of-anew-now-available-in-gcam/>

**Time windows you should know about**

- **This run (default):** free GKG zip download for **1-31 January 2024**, Spanish Gulf + Spanish marine filters, then clip in R.
- **GEO 2.0 API** (optional snapshot): last **7 days**. Set `run_gdelt <- TRUE` if you want it. Not a time series.
- **DOC 2.0 API** (optional snapshot): about the last **3 months**. Same flag. Not a time series.
- **BigQuery** (optional): partitioned GKG 2019-2025. Needs billing. `use_bigquery` is FALSE.

GDELT searches the **English translations** of articles, so English keywords also retrieve Spanish coverage. We still require `sourcelang:spanish` (GEO/DOC) or Spanish ANEW `v22.1` in GCAM (GKG) so the articles we keep are Spanish originals.

Docs:

- GDELT 2.0 announcement: <https://blog.gdeltproject.org/gdelt-2-0-our-global-world-in-realtime/>
- GEO 2.0 API: <https://blog.gdeltproject.org/gdelt-geo-2-0-api-debuts/>
- DOC 2.0 API: <https://blog.gdeltproject.org/gdelt-doc-2-0-api-debuts/>
- Data overview: <https://www.gdeltproject.org/data.html>

```{r}
library(tidyverse)
library(here)
library(sf)
library(httr)
library(jsonlite)
library(janitor)

# This 50 km buffer has self-intersections. s2 (spherical) geometry
# rejects them ("Edge 4 crosses edge 6"); GEOS (planar) does not.
sf::sf_use_s2(FALSE)

# Optional 7-day GEO / 3-month DOC snapshot. Leave FALSE for the
# 2019-2025 indicator.
run_gdelt <- FALSE
try_mapview <- FALSE

geo_timespan <- "7d"     # GEO API maximum is 1 week
doc_timespan <- "3months" # DOC API looks back about 3 months
geo_maxpoints <- 1000

# Optional 24-hour GKG zip test. Not the historical series.
download_gkg <- FALSE
gkg_lookback_hours <- 24

# Free path: download GKG 15-minute zips and filter in R.
# One month is the test. Expand zip_start/zip_end after it works.
download_gkg_zips <- TRUE
zip_start <- as.Date("2024-01-01")
zip_end <- as.Date("2024-01-31")

# BigQuery path. Leave FALSE while the free 1 TB quota is exhausted.
use_bigquery <- FALSE
bq_project_default <- "ohi-goc-cce-gdelt"
# The free 1 TB/month quota is already exhausted on this project.
# Enable billing (trial credits are fine) before querying again:
# https://console.cloud.google.com/billing?project=ohi-goc-cce-gdelt
series_start <- as.Date("2019-01-01")
series_end <- as.Date("2025-12-31")
reference_years <- 2019:2022
series_years <- as.integer(format(series_start, "%Y")):as.integer(format(series_end, "%Y"))
query_events <- FALSE
# GCAM/Spanish ANEW is a huge column. FALSE = engagement counts only (cheap).
# TRUE = also pull valence/arousal (much more data scanned).
gkg_include_anew <- FALSE
bq_max_gb_per_year <- 200
# GDELT 2.0 GKG / GCAM coverage begins 18 February 2015 (not used while series starts 2019).
gkg_v2_start <- as.Date("2015-02-18")

# Intermediate GKG zips, article tables, and working CSVs.
# Current-status scores only go to output_dir.
cce_root_candidates <- c(
  "/home/shares/ohi/OHI_GOC/goal_prep/sp/cce/v2026",
  here::here("sp", "cce", "v2026"),
  here::here("goal_prep", "sp", "cce", "v2026")
)
cce_root_hits <- cce_root_candidates[dir.exists(cce_root_candidates)]
cce_root <- if (length(cce_root_hits) > 0) {
  cce_root_hits[[1]]
} else {
  here::here("CCE", "gdelt_goc_news")
}

int_dir <- file.path(cce_root, "int")
gkg_month_dir <- file.path(int_dir, "gkg_by_month")
output_dir <- file.path(cce_root, "output")
dir.create(gkg_month_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(int_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
message("GKG monthly files: ", gkg_month_dir)
message("Other intermediates: ", int_dir)
message("Current-status scores: ", output_dir)

gdelt_geo <- NULL
gdelt_in_goc <- NULL
gdelt_doc <- tibble()
gkg_gulf_es <- tibble()
gkg_bq <- tibble()

if (!nzchar(Sys.getenv("BIGQUERY_PROJECT", unset = ""))) {
  Sys.setenv(BIGQUERY_PROJECT = bq_project_default)
}
message("BIGQUERY_PROJECT = ", Sys.getenv("BIGQUERY_PROJECT"))
```

```{r}
rgn_labels <- c(
  "Region_1" = "Western Upper Gulf",
  "Region_2" = "Western Midriff Islands",
  "Region_3" = "Western Central Gulf",
  "Region_4" = "Western Southern Gulf",
  "Region_5" = "Eastern Upper Gulf",
  "Region_6" = "Eastern Midriff Islands",
  "Region_7" = "Southern Sonora Coast",
  "Region_8" = "Sinaloa Coast",
  "Region_9" = "Southeastern Gulf"
)

add_rgn_labels <- function(df) {
  df %>%
    mutate(
      rgn_name = recode(as.character(rgn_id), !!!rgn_labels, .default = as.character(rgn_id))
    )
}

first_existing <- function(paths) {
  hits <- paths[file.exists(paths)]
  if (length(hits) == 0) {
    stop(
      "Could not find marine_and_inland50.gpkg. Tried:\n",
      paste0("  - ", paths, collapse = "\n")
    )
  }
  hits[[1]]
}

# GDELT GEO returns HTTP 404 when a query has no matches. Timeouts are
# common. Never let one failed query abort the rest of the pull.
gdelt_request <- function(url, timeout_sec = 180, times = 3) {
  tryCatch(
    httr::RETRY(
      "GET",
      url,
      httr::user_agent("OHI-GOC-GDELT/1.0 (open research; Gulf of California)"),
      httr::timeout(timeout_sec),
      times = times,
      pause_base = 2,
      pause_cap = 16,
      terminate_on = c(400, 404)
    ),
    error = function(e) {
      warning("Request failed: ", conditionMessage(e), "\n  ", url)
      NULL
    }
  )
}

extract_hrefs <- function(html) {
  html <- ifelse(is.na(html), "", as.character(html))
  matches <- stringr::str_match_all(html, 'href=\"([^\"]+)\"')
  purrr::map_chr(matches, function(m) {
    if (nrow(m) == 0) return(NA_character_)
    paste(unique(m[, 2]), collapse = " | ")
  })
}

extract_gcam_value <- function(gcam, key) {
  gcam <- ifelse(is.na(gcam), "", as.character(gcam))
  key_re <- gsub("\\.", "\\\\.", key)
  m <- stringr::str_match(gcam, paste0("(?:^|,)", key_re, ":([^,]+)"))
  as.numeric(m[, 2])
}

spanish_anew_ids <- c(
  valence = "v22.1",
  arousal = "v22.2",
  dominance = "v22.3",
  familiarity = "v22.10",
  concreteness = "v22.11",
  imageability = "v22.12"
)

parse_spanish_anew <- function(df) {
  gcam_col <- if ("gcam" %in% names(df)) df$gcam else df$GCAM
  df %>%
    mutate(
      anew_valence = extract_gcam_value(gcam_col, spanish_anew_ids[["valence"]]),
      anew_arousal = extract_gcam_value(gcam_col, spanish_anew_ids[["arousal"]]),
      anew_dominance = extract_gcam_value(gcam_col, spanish_anew_ids[["dominance"]]),
      anew_familiarity = extract_gcam_value(gcam_col, spanish_anew_ids[["familiarity"]]),
      anew_concreteness = extract_gcam_value(gcam_col, spanish_anew_ids[["concreteness"]]),
      anew_imageability = extract_gcam_value(gcam_col, spanish_anew_ids[["imageability"]])
    )
}

# GKG V2Locations: Type#Name#CountryCode#ADM1[#ADM2]#Lat#Long#FeatureID;
# Find the first numeric pair that looks like latitude, longitude.
latlon_from_parts <- function(parts) {
  nums <- suppressWarnings(as.numeric(parts))
  ok <- which(!is.na(nums) & nzchar(parts))
  if (length(ok) < 2) return(NULL)
  for (i in seq_len(length(ok) - 1)) {
    lat <- nums[ok[i]]
    lon <- nums[ok[i + 1]]
    if (!is.na(lat) && !is.na(lon) && abs(lat) <= 90 && abs(lon) <= 180) {
      return(c(lon = lon, lat = lat))
    }
  }
  NULL
}

parse_gkg_location_points <- function(v2locations, locations) {
  v2locations <- ifelse(is.na(v2locations), "", as.character(v2locations))
  locations <- ifelse(is.na(locations), "", as.character(locations))

  parse_one <- function(v2, loc) {
    pts <- tibble(longitude = numeric(), latitude = numeric())
    add_chunks <- function(text, sep) {
      if (!nzchar(text)) return(pts)
      chunks <- unlist(strsplit(text, ";", fixed = TRUE))
      for (ch in chunks) {
        parts <- unlist(strsplit(ch, sep, fixed = TRUE))
        ll <- latlon_from_parts(parts)
        if (!is.null(ll)) {
          pts <- bind_rows(pts, tibble(longitude = unname(ll["lon"]), latitude = unname(ll["lat"])))
        }
      }
      pts
    }
    pts <- add_chunks(v2, "#")
    if (nrow(pts) == 0) pts <- add_chunks(loc, "#")
    if (nrow(pts) == 0) pts <- add_chunks(loc, ";")
    distinct(pts)
  }

  purrr::map2(v2locations, locations, parse_one)
}

gulf_text_regex <- regex(
  "golfo de california|gulf of california|mar de cort[eé]s|mar de cortez|sea of cortez|sea of cortes",
  ignore_case = TRUE
)

# Spanish-only keep filters for the free zip pull.
gulf_es_regex <- regex(
  "golfo de california|mar de cort[eé]s|mar de cortez",
  ignore_case = TRUE
)

marine_es_regex <- regex(
  paste(
    "pesca|pesquer|manglar|\\bmangle\\b|vaquita|totoaba|ballena|delf[ií]n",
    "lobo marino|tibur[oó]n|tortuga|playa|\\bcostas?\\b|costera|costero",
    "litoral|marino|mar[ií]timo|oc[eé]ano|estero|arrecife|coral",
    "sargazo|humedal|marea|camar[oó]n|sardina|jaiba|almeja",
    "\\banp\\b|conanp|semarnat|captura incidental|pesca incidental",
    "area natural protegida|área natural protegida|vedas?",
    "malec[oó]n|muelle|humedales|laguna costera|zona federal mar[ií]tima",
    sep = "|"
  ),
  ignore_case = TRUE
)

ocean_engagement_pattern <- paste(
  "pesca|pesquer|manglar|vaquita|ballena|playa|costera|marino|",
  "limpieza|totoaba|manifest|protest|megaproyecto|malecon|malec[oó]n|",
  "fishing|fishery|mangrove|whale|beach|coastal|marine|ocean|cleanup|",
  "protected area|bycatch|theme_protest|env_|water_|fish_",
  sep = ""
)
```

# 1. Gulf study area: marine regions, islands, and 50 km inland

```{r}
goc_path <- first_existing(c(
  "/home/shares/ohi/OHI_GOC/spatial/OHI_regions/marine_and_inland50.gpkg",
  "/home/shares/ohi/OHI_GOC/spatial/ohi_regions/marine_and_inland50.gpkg",
  here::here("spatial", "OHI_regions", "marine_and_inland50.gpkg"),
  here::here("spatial", "ohi_regions", "marine_and_inland50.gpkg"),
  here::here("..", "spatial", "OHI_regions", "marine_and_inland50.gpkg")
))

message("Using study area: ", goc_path)

goc <- st_read(goc_path, quiet = TRUE) %>%
  st_transform(4326) %>%
  st_make_valid()

# Keep a region id if the file already has one.
rgn_col <- intersect(
  names(goc),
  c("rgn_id", "rgn", "region_id", "OHI_rgn", "ohi_rgn", "id")
)
if (length(rgn_col) == 1 && rgn_col != "rgn_id") {
  goc <- goc %>% rename(rgn_id = all_of(rgn_col))
}

goc_union <- suppressWarnings(
  goc %>%
    st_union() %>%
    st_make_valid() %>%
    st_as_sf()
)
if (is.na(st_crs(goc_union))) st_crs(goc_union) <- st_crs(goc)

bbox <- st_bbox(goc_union)
# Planar centroid is fine here: s2 is off, and this point is only for the
# optional GEO near: box, not for the 2019-2025 BigQuery clip.
centroid <- suppressWarnings(st_centroid(st_geometry(goc_union)))
centroid_lon <- as.numeric(st_coordinates(centroid)[1, 1])
centroid_lat <- as.numeric(st_coordinates(centroid)[1, 2])

# GEO's near: operator is a bounding box around a radius, not a true circle.
# Use a radius that covers the Gulf bbox, then clip tightly with the geopackage.
if (requireNamespace("geosphere", quietly = TRUE)) {
  half_diag_km <- as.numeric(
    geosphere::distHaversine(
      c(bbox["xmin"], bbox["ymin"]),
      c(bbox["xmax"], bbox["ymax"])
    ) / 1000 / 2
  )
  near_radius_km <- ceiling(half_diag_km + 25)
} else {
  near_radius_km <- 650
  message("geosphere is not installed. Using near_radius_km = 650.")
}

tibble(
  west = unname(bbox["xmin"]),
  south = unname(bbox["ymin"]),
  east = unname(bbox["xmax"]),
  north = unname(bbox["ymax"]),
  centroid_lon = centroid_lon,
  centroid_lat = centroid_lat,
  near_radius_km = near_radius_km
)
```

If `mapview` is available, this is the exact polygon GDELT points must land in.

```{r}
if (isTRUE(try_mapview) && requireNamespace("mapview", quietly = TRUE)) {
  tryCatch(
    mapview::mapview(goc, layer.name = "OHI marine + 50 km inland + islands"),
    error = function(e) {
      warning("mapview failed (", conditionMessage(e), "). Restart R if this is a C++ NULL symbol error.")
      NULL
    }
  )
}
```

# 2. Keyword sets: Gulf name AND ocean engagement

GDELT does not allow two OR blocks in one query. Space between terms is AND. So we do **not** search "protest anywhere in Mexico, then keep Gulf dots." We search:

`[ocean / community topic OR list] "[required Gulf phrase]"`

and repeat that for each Gulf name. That is why the next tibble has 24 rows (4 topics × 6 Gulf phrases). `query_name` is only a label we store in the output CSVs. **GDELT uses `query_text`.** The doubled names you saw (`protest.protest___gulf_of_california_`) were an R list-flattening artifact, not a GDELT error, and not the cause of the GEO 404s.

```{r}
# Required. An article has to use one of these names for the water body.
gulf_phrases <- c(
  '"gulf of california"',
  '"golfo de california"',
  '"sea of cortez"',
  '"sea of cortes"',
  '"mar de cortes"',
  '"mar de cortez"'
)

# One OR block per theme. These are about the ocean, the coast, or people
# organizing around them. Not generic city politics.
topic_sets <- list(
  protest = paste(
    "(manifestacion OR manifestaciones OR protesta OR protestas",
    "OR protester OR protest OR manifestation OR \"movimiento social\"",
    "OR \"social movement\")"
  ),
  megaproject = paste(
    "(megaproyecto OR megaproyectos OR malecon OR malecón",
    "OR \"desarrollo turistico\" OR marina OR megaproject",
    "OR \"mega project\" OR \"tourist development\")"
  ),
  ocean_environment = paste(
    "(pesca OR pesqueria OR pesquería OR manglar OR vaquita OR ballena",
    "OR playa OR costera OR marino OR limpieza OR totoaba",
    "OR \"area natural protegida\" OR fishing OR fishery OR mangrove",
    "OR whale OR beach OR coastal OR marine OR ocean OR cleanup",
    "OR \"protected area\" OR bycatch)"
  )
)

# GKG theme plus a required Gulf phrase. Theme names are GDELT's.
theme_sets <- list(
  gkg_protest = "theme:PROTEST"
)

# Labels for our output tables only. GDELT never sees these names.
# GDELT receives query_text (topic OR-list + one quoted Gulf phrase).
gulf_phrase_slug <- function(phrase) {
  phrase %>%
    str_replace_all('"', "") %>%
    str_to_lower() %>%
    str_replace_all("[^a-z]+", "_") %>%
    str_replace_all("^_|_$", "")
}

cross_gulf_queries <- function(topic_block, topic_name) {
  setNames(
    paste(topic_block, gulf_phrases),
    paste(topic_name, gulf_phrase_slug(gulf_phrases), sep = "__")
  )
}

# Flatten without R's outer.inner name prefix (protest.protest__...).
query_parts <- unname(c(
  purrr::imap(topic_sets, cross_gulf_queries),
  purrr::imap(theme_sets, cross_gulf_queries)
))
query_sets <- setNames(
  unlist(query_parts, use.names = FALSE),
  unlist(lapply(query_parts, names), use.names = FALSE)
)

add_geo_operators <- function(core_query) {
  paste(
    core_query,
    "sourcelang:spanish",
    "locationcc:MX",
    sprintf("near:%s,%s,%skm", centroid_lat, centroid_lon, near_radius_km)
  )
}

# Inventory of searches. query_name is a local label; query_text is what GDELT gets.
tibble(
  query_name = names(query_sets),
  query_text = unname(query_sets)
)
```

# 3. GEO 2.0 API: locations mentioned near the keywords

The GEO API returns points for places named close to the query in **Spanish-language** news (searched via GDELT's English translation, restricted with `sourcelang:spanish`). We ask for city and landmark resolution (`geores=2`) so a vague mention of "Mexico" does not become a point.

Then we **intersect with `marine_and_inland50.gpkg`**, which is the real geographic limit: Gulf waters, islands, and 50 km inland. The `near:` radius is only a coarse net.

```{r}
gdelt_geo_url <- function(query, timespan = geo_timespan, maxpoints = geo_maxpoints) {
  httr::modify_url(
    "https://api.gdeltproject.org/api/v2/geo/geo",
    query = list(
      query = query,
      mode = "PointData",
      format = "GeoJSON",
      timespan = timespan,
      maxpoints = maxpoints,
      geores = 2
    )
  )
}

fetch_gdelt_geo <- function(query, query_name) {
  url <- gdelt_geo_url(query)
  message("GEO query [", query_name, "]: ", url)

  resp <- gdelt_request(url)
  if (is.null(resp)) return(NULL)

  code <- httr::status_code(resp)
  # GEO returns 404 when the query has no mapped points this week.
  if (code == 404) {
    message("GEO [", query_name, "]: no matches (HTTP 404).")
    return(NULL)
  }
  if (code >= 400) {
    warning("GEO [", query_name, "]: HTTP ", code)
    return(NULL)
  }

  raw <- httr::content(resp, as = "text", encoding = "UTF-8")
  if (!grepl("FeatureCollection|Feature", raw)) {
    warning("GEO query [", query_name, "] did not return GeoJSON. First 300 characters:\n", substr(raw, 1, 300))
    return(NULL)
  }

  tmp <- tempfile(fileext = ".geojson")
  writeLines(raw, tmp, useBytes = TRUE)
  out <- tryCatch(
    st_read(tmp, quiet = TRUE),
    error = function(e) {
      warning("Could not parse GeoJSON for [", query_name, "]: ", e$message)
      NULL
    }
  )
  if (is.null(out) || nrow(out) == 0) return(NULL)

  out %>%
    mutate(
      query_name = query_name,
      query_text = query,
      article_urls = extract_hrefs(if ("html" %in% names(.)) html else NA_character_)
    )
}

gdelt_geo <- NULL
if (isTRUE(run_gdelt)) {
  geo_queries <- purrr::imap(as.list(query_sets), ~ add_geo_operators(.x))

  geo_raw <- purrr::imap(geo_queries, fetch_gdelt_geo) %>%
    purrr::compact()

  gdelt_geo <- if (length(geo_raw) == 0) {
    warning("No GEO features returned. The API may be down, or the week may have no matches.")
    NULL
  } else {
    bind_rows(geo_raw) %>%
      st_transform(4326)
  }
}

if (!is.null(gdelt_geo)) gdelt_geo
```

# 4. Clip to the 50 km inland Gulf layer and assign OHI regions

```{r}
if (!is.null(gdelt_geo) && nrow(gdelt_geo) > 0) {
  gdelt_pts <- gdelt_geo %>%
    st_cast("POINT", warn = FALSE) %>%
    st_make_valid()

  # Drop empty or invalid points before the join.
  gdelt_pts <- gdelt_pts[!st_is_empty(gdelt_pts), ]

  goc_join <- goc
  if (!"rgn_id" %in% names(goc_join)) {
    goc_join$rgn_id <- NA_character_
  }

  gdelt_in_goc <- st_join(
    gdelt_pts,
    goc_join %>% dplyr::select(any_of(c("rgn_id", "rgn_name"))),
    join = st_intersects,
    left = FALSE
  ) %>%
    add_rgn_labels() %>%
    mutate(
      longitude = st_coordinates(geometry)[, 1],
      latitude = st_coordinates(geometry)[, 2]
    )

  message(
    nrow(gdelt_pts), " GEO points in the coarse near-box; ",
    nrow(gdelt_in_goc), " remain inside marine + 50 km inland + islands."
  )

  gdelt_in_goc_tbl <- gdelt_in_goc %>%
    st_drop_geometry() %>%
    distinct(query_name, name, latitude, longitude, rgn_id, rgn_name, count, article_urls, .keep_all = TRUE)

  write_csv(gdelt_in_goc_tbl, file.path(int_dir, "gdelt_geo_points_in_goc.csv"))
  st_write(
    gdelt_in_goc,
    file.path(int_dir, "gdelt_geo_points_in_goc.gpkg"),
    delete_dsn = TRUE,
    quiet = TRUE
  )

  gdelt_in_goc_tbl %>%
    count(rgn_name, query_name, name = "n_locations") %>%
    arrange(rgn_name, desc(n_locations))
} else {
  gdelt_in_goc <- NULL
  gdelt_in_goc_tbl <- tibble()
  message("No GEO points to clip.")
}
```

```{r}
if (!is.null(gdelt_in_goc) && isTRUE(try_mapview) && requireNamespace("mapview", quietly = TRUE)) {
  tryCatch(
    mapview::mapview(
      goc_union,
      col.regions = "gray90",
      alpha.regions = 0.2,
      layer.name = "Study area"
    ) +
      mapview::mapview(
        gdelt_in_goc,
        zcol = "query_name",
        layer.name = "GDELT locations in GoC"
      ),
    error = function(e) {
      warning("mapview failed (", conditionMessage(e), "). Restart R if this is a C++ NULL symbol error.")
      NULL
    }
  )
}
```

# 5. DOC 2.0 API: article list for the same Gulf + ocean topics

DOC cannot clip to the geopackage. It still uses the same **Gulf phrase plus ocean-engagement** queries, Mexican outlets (`sourcecountry:mexico`), and Spanish originals (`sourcelang:spanish`).

```{r}
gdelt_doc_url <- function(query, mode = "ArtList", timespan = doc_timespan, maxrecords = 75) {
  httr::modify_url(
    "https://api.gdeltproject.org/api/v2/doc/doc",
    query = list(
      query = query,
      mode = mode,
      format = "json",
      timespan = timespan,
      maxrecords = maxrecords,
      sort = "datedesc"
    )
  )
}

fetch_gdelt_doc <- function(query, query_name) {
  # query already includes a required Gulf / Sea of Cortez phrase.
  full_query <- paste(query, "sourcecountry:mexico", "sourcelang:spanish")
  url <- gdelt_doc_url(full_query)
  message("DOC query [", query_name, "]: ", url)

  resp <- gdelt_request(url, timeout_sec = 180, times = 4)
  if (is.null(resp)) return(NULL)

  code <- httr::status_code(resp)
  if (code == 404) {
    message("DOC [", query_name, "]: no matches (HTTP 404).")
    return(NULL)
  }
  if (code >= 400) {
    warning("DOC [", query_name, "]: HTTP ", code)
    return(NULL)
  }

  parsed <- tryCatch(
    jsonlite::fromJSON(httr::content(resp, as = "text", encoding = "UTF-8")),
    error = function(e) NULL
  )
  arts <- parsed$articles
  if (is.null(arts) || length(arts) == 0) return(NULL)

  as_tibble(arts) %>%
    mutate(query_name = query_name, query_text = full_query)
}

gdelt_doc <- tibble()
if (isTRUE(run_gdelt)) {
  doc_raw <- purrr::imap(query_sets, fetch_gdelt_doc) %>%
    purrr::compact()

  gdelt_doc <- if (length(doc_raw) == 0) {
    tibble()
  } else {
    bind_rows(doc_raw) %>%
      distinct(url, query_name, .keep_all = TRUE)
  }

  if (nrow(gdelt_doc) > 0) {
    write_csv(gdelt_doc, file.path(int_dir, "gdelt_doc_articles_gulf_ocean.csv"))
  }

  gdelt_doc
}
```

# 6. Spanish ANEW emotion scores from GCAM

Spanish ANEW is **usable**, and it is the right emotion lexicon here, because it is applied to Spanish text rather than to an English translation. GDELT stores it in the GKG `GCAM` string as dictionary **22**:

| GCAM id | What it measures (Spanish ANEW) | Typical scale |
|---------|----------------------------------|---------------|
| `v22.1` | Valence / pleasure (all) | about 1 (unpleasant) to 9 (pleasant), 5 is middle |
| `v22.2` | Arousal (all) | about 1 (calm) to 9 (excited) |
| `v22.3` | Dominance (all) | about 1 (out of control) to 9 (in control) |
| `v22.10` | Familiarity | higher = more familiar wording |
| `v22.11` | Concreteness | higher = more concrete |
| `v22.12` | Imageability | higher = easier to picture |

`c22.*` are word counts for the same dimensions. We use the scored `v22.*` values. These fields appear only when GDELT ingested the article as Spanish. English ANEW (`v19.*`) is a different dictionary and should not be the primary score for this Gulf corpus.

This chunk downloads recent GKG 2.1 files (open zip files, no login), keeps rows that look like Spanish Gulf coverage, and parses Spanish ANEW. Set `gkg_lookback_hours` to 168 for a week. For months of history, use the BigQuery block in section 8.

```{r}
gulf_text_pattern <- gulf_text_regex

gkg_col_names <- c(
  "gkgrecordid", "date", "sourcecollectionidentifier", "sourcecommonname",
  "documentidentifier", "counts", "v2counts", "themes", "v2themes",
  "locations", "v2locations", "persons", "v2persons", "organizations",
  "v2organizations", "v2tone", "dates", "gcam", "sharingimage",
  "relatedimages", "socialimageembeds", "socialvideoembeds", "quotations",
  "allnames", "amounts", "translationinfo", "extras"
)

gkg_stamp <- function(t) {
  t <- as.POSIXlt(t, tz = "UTC")
  t$min <- (t$min %/% 15) * 15
  t$sec <- 0
  format(t, "%Y%m%d%H%M%S")
}

fetch_one_gkg <- function(stamp) {
  url <- paste0("http://data.gdeltproject.org/gdeltv2/", stamp, ".gkg.csv.zip")
  tmp <- tempfile(fileext = ".zip")
  resp <- tryCatch(
    httr::RETRY(
      "GET",
      url,
      httr::write_disk(tmp, overwrite = TRUE),
      httr::user_agent("OHI-GOC-GDELT/1.0"),
      httr::timeout(120),
      times = 3,
      terminate_on = c(404)
    ),
    error = function(e) NULL
  )

  if (is.null(resp) || httr::status_code(resp) != 200 ||
      !file.exists(tmp) || file.info(tmp)$size < 1000) {
    unlink(tmp)
    return(NULL)
  }

  dat <- tryCatch(
    {
      unzip_dir <- tempfile()
      dir.create(unzip_dir)
      on.exit(unlink(unzip_dir, recursive = TRUE), add = TRUE)
      utils::unzip(tmp, exdir = unzip_dir)
      inner <- list.files(unzip_dir, full.names = TRUE)[1]
      if (is.na(inner) || !file.exists(inner)) return(NULL)
      read_tsv(
        inner,
        col_names = gkg_col_names,
        col_types = cols(.default = col_character()),
        quote = "",
        progress = FALSE
      )
    },
    error = function(e) NULL
  )
  unlink(tmp)
  dat
}

gkg_gulf_es <- tibble()
if (isTRUE(run_gdelt) && isTRUE(download_gkg)) {
  stamps <- seq(
    Sys.time() - gkg_lookback_hours * 3600,
    Sys.time(),
    by = 15 * 60
  ) %>%
    unique() %>%
    purrr::map_chr(gkg_stamp) %>%
    unique()

  message("Downloading ", length(stamps), " GKG files for Spanish ANEW (GCAM v22).")

  gkg_raw <- purrr::map(stamps, fetch_one_gkg) %>%
    purrr::compact()

  message("Parsed ", length(gkg_raw), " of ", length(stamps), " GKG files.")

  gkg_gulf_es <- if (length(gkg_raw) == 0) {
    warning("No GKG files could be parsed. Check network access to data.gdeltproject.org.")
    tibble()
  } else {
    gkg_all <- bind_rows(gkg_raw)
    message("GKG rows before Gulf/Spanish filter: ", nrow(gkg_all))
    gkg_all %>%
      filter(
        !is.na(gcam),
        str_detect(gcam, "v22\\.1:"),
        str_detect(
          paste(locations, v2locations, documentidentifier, extras, sep = " "),
          gulf_text_pattern
        )
      ) %>%
      parse_spanish_anew() %>%
      filter(!is.na(anew_valence)) %>%
      distinct(documentidentifier, .keep_all = TRUE)
  }

  if (nrow(gkg_gulf_es) > 0) {
    write_csv(
      gkg_gulf_es %>%
        dplyr::select(
          date, sourcecommonname, documentidentifier, translationinfo,
          starts_with("anew_"), v2tone, themes
        ),
      file.path(int_dir, "gdelt_gkg_spanish_anew_gulf.csv")
    )
    gkg_gulf_es %>%
      summarise(
        n_articles = n(),
        mean_valence = mean(anew_valence, na.rm = TRUE),
        mean_arousal = mean(anew_arousal, na.rm = TRUE),
        mean_dominance = mean(anew_dominance, na.rm = TRUE)
      )
  } else if (length(gkg_raw) > 0) {
    message(
      "GKG files parsed, but none matched Spanish ANEW + Gulf text in this ",
      gkg_lookback_hours, "-hour window. That can be a true zero for a short lookback."
    )
    gkg_gulf_es
  }
}
```

If DOC returned URLs, you can join them to the GKG table on `documentidentifier` to attach Spanish ANEW scores to the article list.

# 6b. Free test: one month of GKG zips (Spanish + marine)

This path does not use BigQuery. It downloads the public 15-minute GKG zips and keeps a row only if **all** of these are true:

1. **Spanish original:** GCAM dictionary 22 (`v22.1`) or `TranslationInfo` contains `srclc:spa`.
2. **Spanish Gulf name:** `golfo de california` or `mar de cortés` / `cortes` / `cortez` (not English "gulf of california" / "sea of cortez").
3. **Spanish marine or coastal language:** fishing, mangroves, vaquita, totoaba, whales, beaches, coast, protected areas, and related terms. English marine words are not enough to keep a row.

January 2024 is about 2,976 files. Missing 15-minute stamps are skipped. Finished stamps are stored so a second run continues where it left off.

```{r}
gkg_month_dir <- file.path(int_dir, "gkg_by_month")
dir.create(gkg_month_dir, recursive = TRUE, showWarnings = FALSE)
zip_month_tag <- format(zip_start, "%Y-%m")
zip_csv <- file.path(gkg_month_dir, paste0("gkg_gulf_es_marine_", zip_month_tag, ".csv"))
zip_progress <- file.path(gkg_month_dir, paste0("stamps_done_", zip_month_tag, ".txt"))

gulf_line_grep <- "golfo de california|mar de cort"

is_spanish_original <- function(gcam, translationinfo) {
  gcam <- ifelse(is.na(gcam), "", as.character(gcam))
  translationinfo <- ifelse(is.na(translationinfo), "", as.character(translationinfo))
  str_detect(gcam, "v22\\.1:") | str_detect(translationinfo, regex("srclc:spa", ignore_case = TRUE))
}

filter_spanish_gulf_marine <- function(dat) {
  if (is.null(dat) || nrow(dat) == 0) return(tibble())
  blob <- paste(
    dat$locations, dat$v2locations, dat$documentidentifier,
    dat$extras, dat$themes, dat$v2themes,
    sep = " "
  )
  dat %>%
    mutate(
      is_spanish = is_spanish_original(gcam, translationinfo),
      has_gulf_es = str_detect(blob, gulf_es_regex),
      has_marine_es = str_detect(blob, marine_es_regex)
    ) %>%
    filter(is_spanish, has_gulf_es, has_marine_es)
}

fetch_one_gkg_filtered <- function(stamp) {
  url <- paste0("http://data.gdeltproject.org/gdeltv2/", stamp, ".gkg.csv.zip")
  tmp <- tempfile(fileext = ".zip")
  resp <- tryCatch(
    httr::RETRY(
      "GET",
      url,
      httr::write_disk(tmp, overwrite = TRUE),
      httr::user_agent("OHI-GOC-GDELT/1.0"),
      httr::timeout(120),
      times = 2,
      terminate_on = c(404)
    ),
    error = function(e) NULL
  )
  if (is.null(resp) || httr::status_code(resp) != 200 ||
      !file.exists(tmp) || file.info(tmp)$size < 1000) {
    unlink(tmp)
    return(NULL)
  }

  dat <- tryCatch(
    {
      unzip_dir <- tempfile()
      dir.create(unzip_dir)
      on.exit(unlink(unzip_dir, recursive = TRUE), add = TRUE)
      utils::unzip(tmp, exdir = unzip_dir)
      inner <- list.files(unzip_dir, full.names = TRUE)[1]
      if (is.na(inner) || !file.exists(inner)) return(NULL)
      raw_lines <- readLines(inner, warn = FALSE, encoding = "UTF-8")
      if (length(raw_lines) == 0) return(NULL)
      hit <- grepl(gulf_line_grep, raw_lines, ignore.case = TRUE, useBytes = TRUE)
      if (!any(hit)) return(tibble())
      parsed <- read_tsv(
        I(paste(raw_lines[hit], collapse = "\n")),
        col_names = gkg_col_names,
        col_types = cols(.default = col_character()),
        quote = "",
        progress = FALSE
      )
      filter_spanish_gulf_marine(parsed)
    },
    error = function(e) NULL
  )
  unlink(tmp)
  dat
}

done_stamps <- if (file.exists(zip_progress)) {
  unique(readLines(zip_progress, warn = FALSE))
} else {
  character()
}

if (isTRUE(download_gkg_zips)) {
  zip_stamps <- seq(
    as.POSIXct(paste(zip_start, "00:00:00"), tz = "UTC"),
    as.POSIXct(paste(zip_end, "23:45:00"), tz = "UTC"),
    by = 15 * 60
  ) %>%
    purrr::map_chr(gkg_stamp) %>%
    unique()

  zip_stamps <- setdiff(zip_stamps, done_stamps)
  message(
    "GKG zip pull ", zip_month_tag, ": ",
    length(zip_stamps), " stamps left (",
    length(done_stamps), " already done)."
  )

  n_ok <- 0L
  n_hit <- 0L
  for (i in seq_along(zip_stamps)) {
    stamp <- zip_stamps[[i]]
    one <- fetch_one_gkg_filtered(stamp)
    write(stamp, zip_progress, append = TRUE)
    if (!is.null(one) && nrow(one) > 0) {
      n_hit <- n_hit + nrow(one)
      write_csv(one, zip_csv, append = file.exists(zip_csv))
    }
    n_ok <- n_ok + 1L
    if (i %% 50 == 0 || i == length(zip_stamps)) {
      message(
        sprintf(
          "  %d / %d files, %d Spanish Gulf-marine rows so far",
          i, length(zip_stamps), n_hit
        )
      )
    }
  }

  if (file.exists(zip_csv)) {
    gkg_bq <- read_csv(zip_csv, show_col_types = FALSE) %>%
      janitor::clean_names() %>%
      parse_spanish_anew() %>%
      mutate(
        document_date = as.Date(substr(as.character(date), 1, 8), format = "%Y%m%d"),
        ocean_engagement = TRUE
      ) %>%
      distinct(documentidentifier, .keep_all = TRUE)
    write_csv(gkg_bq, file.path(gkg_month_dir, "gdelt_zip_spanish_gulf_marine.csv"))
    message("Kept ", nrow(gkg_bq), " distinct Spanish Gulf-marine articles for ", zip_month_tag)
    gkg_bq %>%
      summarise(
        n_articles = n(),
        mean_valence = mean(anew_valence, na.rm = TRUE),
        mean_arousal = mean(anew_arousal, na.rm = TRUE)
      )
  } else {
    message("No Spanish Gulf-marine rows in this month.")
  }
}
```

# 7. What to save and how to repeat this

On the NCEAS shares the script writes here (or to the matching `sp/cce/v2026` folder if you are working locally):

**GKG month files** (`.../int/gkg_by_month/`):

- `gkg_gulf_es_marine_YYYY-MM.csv`: Spanish Gulf + Spanish marine rows from the free zip pull
- `stamps_done_YYYY-MM.txt`: finished 15-minute timestamps so a rerun can skip them
- `gdelt_zip_spanish_gulf_marine.csv`: the combined month extract after Spanish + Gulf + marine filters

**Other intermediates** (`.../int/`): GEO/DOC snapshots, clipped articles, coastal population, annual rates, the full status table (with counts, per-capita rates, and sentiment), and BigQuery yearly caches in `gkg_by_year/`.

**Current-status scores only** (`.../output/`):

- `gdelt_news_engagement_current_status.csv`: one row per OHI region and year, with `rgn_id`, `year`, and `current_status` (uncapped engagement vs the 2019-2022 mean). Gulf-wide rows stay in the intermediate status table.

To repeat the 2019-2025 indicator, keep this script, the geopackage path, `BIGQUERY_PROJECT`, and the monthly CSVs in `int/gkg_by_month/`. GEO/DOC are a rolling window and are off by default.

**Limits that matter for CCE**

1. GEO and DOC are optional snapshots (7 days / 3 months). The 2019-2025 series is partitioned BigQuery GKG in section 8.
2. A Gulf place-name in an article is not the same as ocean engagement. Spot-check URLs. A tourism listicle that says "Sea of Cortez" can still pass the name filter.
3. GDELT geocoding is automated. Same-name places get mixed up. Always spot-check URLs.
4. This still should not be averaged into CCE until there is an agreed keyword list, an agreed archive of each pull, and a reference point.

# 8. Historical series, 2019-2025: engagement per capita and sentiment

The GEO and DOC APIs cannot build this series. Use **`gdelt-bq.gdeltv2.gkg_partitioned`** with `_PARTITIONTIME`, one year at a time. Keep **Spanish originals** (GCAM dictionary 22 / Spanish ANEW), require a **Gulf of California / Sea of Cortez / Mar de Cortés** name, then clip mapped coordinates to `marine_and_inland50.gpkg`. The unpartitioned `gkg` table is several terabytes and will exceed the free query quota.

**What "engagement" and "sentiment" mean here**

- **Engagement** is media volume: distinct Spanish Gulf articles per year, after the spatial clip. The scored layer further requires ocean / coast / protest / megaproject language so a generic tourism mention is less likely to count. We divide by **coastal locality population** in the same 50 km study area (INEGI 2020 localities, the same population family CCE already uses). The rate is articles per 100,000 coastal residents. Population is a **static 2020** denominator; we do not yet have yearly coastal-locality counts.
- **Sentiment** is mean Spanish ANEW valence (`v22.1`, about 1 = unpleasant to 9 = pleasant). Valence is already a rate, so we do **not** divide it by population. Arousal (`v22.2`) is kept as a companion (emotional intensity).

**Reference point (2019-2022)**

1. Compute one annual engagement-per-capita value for each year 2019 through 2022.
2. Average those four yearly values. That mean is the historical reference.
3. For every year 2019-2025:

$$\mathrm{status}_{r,t} = 100 \times \frac{\text{engagement per 100,000}_{r,t}}{\text{mean of 2019-2022 annual values}}$$

We do **not** cap at 100, so a rise above the historical mean is visible. Status = 100 means matching the 2019-2022 average. Each region uses **its own** 2019-2022 mean, and we also compute a Gulf-wide series (distinct articles Gulf-wide / sum of regional population) with one Gulf-wide 2019-2022 mean.

`use_bigquery` is already `TRUE`. The default project is `ohi-goc-cce-gdelt`. Enable billing if the free 1 TB quota is already used. After `gcloud auth application-default login`, Run All. Yearly CSVs in `output/gkg_by_year/` are reused so a later year can resume after a quota stop.

Do not mix English ANEW (`v19`) into these results.

```{r}
bq_project <- Sys.getenv("BIGQUERY_PROJECT", unset = "")
gkg_cache <- file.path(int_dir, "gdelt_bigquery_spanish_anew_gulf.csv")
gkg_year_dir <- file.path(int_dir, "gkg_by_year")
dir.create(gkg_year_dir, recursive = TRUE, showWarnings = FALSE)
gulf_re_sql <- paste(
  "golfo de california",
  "gulf of california",
  "mar de cort[eé]s",
  "mar de cortez",
  "sea of cortez",
  "sea of cortes",
  sep = "|"
)

gkg_year_sql <- function(year) {
  d0 <- max(series_start, as.Date(sprintf("%d-01-01", year)))
  d1 <- min(series_end, as.Date(sprintf("%d-12-31", year)))
  d2 <- d1 + 1
  select_cols <- paste(
    "DATE, SourceCommonName, DocumentIdentifier, TranslationInfo,",
    "V2Locations, Locations"
  )
  if (isTRUE(gkg_include_anew)) {
    select_cols <- paste(
      select_cols,
      ", V2Tone, V2Themes, Extras,",
      "REGEXP_EXTRACT(GCAM, r'(?:^|,)v22\\.1:([^,]+)') AS anew_valence,",
      "REGEXP_EXTRACT(GCAM, r'(?:^|,)v22\\.2:([^,]+)') AS anew_arousal,",
      "REGEXP_EXTRACT(GCAM, r'(?:^|,)v22\\.3:([^,]+)') AS anew_dominance"
    )
  }
  paste0(
    "SELECT ", select_cols, " ",
    "FROM `gdelt-bq.gdeltv2.gkg_partitioned` ",
    "WHERE _PARTITIONTIME >= TIMESTAMP('", d0, "') ",
    "AND _PARTITIONTIME < TIMESTAMP('", d2, "') ",
    "AND DATE >= ", format(d0, "%Y%m%d"), "000000 ",
    "AND DATE <= ", format(d1, "%Y%m%d"), "235959 ",
    "AND LOWER(IFNULL(TranslationInfo, '')) LIKE '%srclc:spa%' ",
    "AND (",
    "REGEXP_CONTAINS(LOWER(IFNULL(V2Locations, '')), r'", gulf_re_sql, "') ",
    "OR REGEXP_CONTAINS(LOWER(IFNULL(DocumentIdentifier, '')), r'", gulf_re_sql, "')",
    ")"
  )
}

gkg_bq <- tibble()

if (isTRUE(use_bigquery) && nrow(gkg_bq) == 0) {
  if (!requireNamespace("bigrquery", quietly = TRUE)) {
    stop("Install bigrquery to use this section: install.packages('bigrquery')")
  }
  if (!nzchar(bq_project)) {
    stop("Set BIGQUERY_PROJECT to your Google Cloud project id.")
  }

  writeLines(
    gkg_year_sql(series_years[[1]]),
    file.path(int_dir, "gdelt_bigquery_gkg_query.sql")
  )

  year_files <- character()
  for (yr in series_years) {
    year_csv <- file.path(gkg_year_dir, sprintf("gkg_gulf_%d.csv", yr))
    if (file.exists(year_csv)) {
      message("Reusing cached ", year_csv)
      year_files <- c(year_files, year_csv)
      next
    }

    sql <- gkg_year_sql(yr)
    gb <- tryCatch(
      {
        job <- bigrquery::bq_perform_query(sql, billing = bq_project, dry_run = TRUE)
        bytes <- job$statistics$query$totalBytesProcessed
        if (is.null(bytes)) bytes <- job$statistics$totalBytesProcessed
        bytes <- as.numeric(bytes)
        if (length(bytes) != 1 || is.na(bytes)) NA_real_ else bytes / 1e9
      },
      error = function(e) NA_real_
    )
    if (isTRUE(length(gb) == 1 && !is.na(gb))) {
      message(sprintf("Year %d dry run: about %.2f GB to scan", yr, gb))
    }

    message("Querying partitioned GKG for ", yr, " ...")
    one <- tryCatch(
      {
        tbl <- bigrquery::bq_project_query(
          bq_project,
          sql,
          maximum_bytes_billed = as.numeric(bq_max_gb_per_year) * 1e9
        )
        bigrquery::bq_table_download(tbl)
      },
      error = function(e) {
        warning(conditionMessage(e), call. = FALSE)
        if (grepl("quotaExceeded|free query bytes|maximumBytesBilled", conditionMessage(e), ignore.case = TRUE)) {
          message(
            "Stopped at ", yr, ". The free 1 TB quota on this project is used up.\n",
            "Path forward: enable billing (trial credits are OK):\n",
            "  https://console.cloud.google.com/billing?project=", bq_project, "\n",
            "Then re-run this chunk. Years already saved in gkg_by_year/ will be reused."
          )
        }
        NULL
      }
    )
    if (is.null(one) || !is.data.frame(one)) break

    one <- one %>%
      janitor::clean_names() %>%
      mutate(
        document_date = as.Date(substr(as.character(date), 1, 8), format = "%Y%m%d")
      )
    if ("anew_valence" %in% names(one)) {
      one <- one %>%
        mutate(across(starts_with("anew_"), as.numeric)) %>%
        filter(!is.na(anew_valence))
    }
    one <- one %>%
      filter(document_date >= series_start, document_date <= series_end) %>%
      distinct(documentidentifier, .keep_all = TRUE)

    write_csv(one, year_csv)
    year_files <- c(year_files, year_csv)
    message("Saved ", nrow(one), " Spanish Gulf rows for ", yr)
  }

  if (length(year_files) > 0) {
    gkg_bq <- purrr::map(year_files, ~ read_csv(.x, show_col_types = FALSE)) %>%
      bind_rows() %>%
      janitor::clean_names() %>%
      distinct(documentidentifier, .keep_all = TRUE)
    write_csv(gkg_bq, gkg_cache)
  }
} else if (file.exists(gkg_cache)) {
  message("use_bigquery is FALSE. Reusing ", gkg_cache)
  gkg_bq <- read_csv(gkg_cache, show_col_types = FALSE) %>%
    janitor::clean_names()
  if (!"anew_valence" %in% names(gkg_bq) && "gcam" %in% names(gkg_bq)) {
    gkg_bq <- parse_spanish_anew(gkg_bq)
  }
  if (!"document_date" %in% names(gkg_bq)) {
    gkg_bq <- gkg_bq %>%
      mutate(document_date = as.Date(substr(as.character(date), 1, 8), format = "%Y%m%d"))
  }
} else {
  message(
    "No GKG table yet. Set use_bigquery <- TRUE and BIGQUERY_PROJECT, ",
    "or place a previous pull at:\n  ", gkg_cache
  )
}
```

## Coastal population (2020 localities in the 50 km study area)

Same INEGI locality population CCE uses, intersected here with the **50 km** Gulf layer so the denominator matches the article clip. If the shapefile is not on this machine, per-capita columns stay empty and article counts are still written.

```{r}
pop_path_candidates <- c(
  "/home/shares/ohi/OHI_GOC/goal_prep/pressures/population_localities/inegi_boundaries_pop.shp",
  here::here("goal_prep", "pressures", "population_localities", "inegi_boundaries_pop.shp"),
  here::here("spatial", "population_localities", "inegi_boundaries_pop.shp")
)
pop_hits <- pop_path_candidates[file.exists(pop_path_candidates)]
pop_path <- if (length(pop_hits) > 0) pop_hits[[1]] else NULL

pop_by_rgn <- tibble(rgn_id = character(), rgn_name = character(), pobtot = numeric())

if (!is.null(pop_path)) {
  message("Using coastal population: ", pop_path)
  cities <- st_read(pop_path, quiet = TRUE) %>%
    janitor::clean_names() %>%
    st_make_valid() %>%
    st_transform(4326)

  if (!"pobtot" %in% names(cities)) {
    stop("Population shapefile needs a pobtot column.")
  }

  goc_join <- goc
  if (!"rgn_id" %in% names(goc_join)) {
    goc_join$rgn_id <- NA_character_
  }

  city_pts <- suppressWarnings(st_centroid(cities, of_largest_polygon = TRUE))
  pop_by_rgn <- st_join(
    city_pts,
    goc_join %>% dplyr::select(any_of("rgn_id")),
    join = st_intersects,
    left = FALSE
  ) %>%
    st_drop_geometry() %>%
    group_by(rgn_id) %>%
    summarise(pobtot = sum(as.numeric(pobtot), na.rm = TRUE), .groups = "drop") %>%
    add_rgn_labels()

  write_csv(pop_by_rgn, file.path(int_dir, "gdelt_coastal_population_by_region.csv"))
  pop_by_rgn
} else {
  message("Population shapefile not found. Engagement per capita will be NA.")
  pop_by_rgn
}
```

## Clip GKG locations, annual rates, and the 2019-2022 reference

```{r}
gkg_in_goc <- tibble()
engagement_annual <- tibble()
engagement_status <- tibble()
sentiment_annual <- tibble()

if (nrow(gkg_bq) > 0) {
  loc_col <- if ("v2locations" %in% names(gkg_bq)) gkg_bq$v2locations else NA_character_
  loc_old <- if ("locations" %in% names(gkg_bq)) gkg_bq$locations else NA_character_
  theme_col <- if ("v2themes" %in% names(gkg_bq)) gkg_bq$v2themes else ""
  extra_col <- if ("extras" %in% names(gkg_bq)) gkg_bq$extras else ""
  url_col <- if ("documentidentifier" %in% names(gkg_bq)) gkg_bq$documentidentifier else ""

  gkg_pts <- gkg_bq %>%
    mutate(
      pts = parse_gkg_location_points(loc_col, loc_old),
      ocean_engagement = str_detect(
        paste(theme_col, loc_col, extra_col, url_col, sep = " "),
        marine_es_regex
      )
    ) %>%
    tidyr::unnest(pts, keep_empty = FALSE)

  if (nrow(gkg_pts) > 0) {
    goc_join <- goc
    if (!"rgn_id" %in% names(goc_join)) goc_join$rgn_id <- NA_character_

    gkg_in_goc <- gkg_pts %>%
      st_as_sf(coords = c("longitude", "latitude"), crs = 4326, remove = FALSE) %>%
      st_join(
        goc_join %>% dplyr::select(any_of("rgn_id")),
        join = st_intersects,
        left = FALSE
      ) %>%
      add_rgn_labels()

    write_csv(
      gkg_in_goc %>% st_drop_geometry(),
      file.path(int_dir, "gdelt_gkg_articles_in_goc.csv")
    )
  }
}

days_2015 <- as.numeric(as.Date("2015-12-31") - gkg_v2_start) + 1
annualize_n <- function(n, year) {
  ifelse(year == 2015, n * 365 / days_2015, n)
}

if (nrow(gkg_in_goc) > 0) {
  articles <- gkg_in_goc %>%
    st_drop_geometry() %>%
    mutate(year = as.integer(format(document_date, "%Y"))) %>%
    filter(year >= min(series_years), year <= max(series_years))
  if (!"anew_valence" %in% names(articles)) articles$anew_valence <- NA_real_
  if (!"anew_arousal" %in% names(articles)) articles$anew_arousal <- NA_real_

  rgn_year <- articles %>%
    distinct(documentidentifier, rgn_id, rgn_name, year, ocean_engagement, .keep_all = TRUE) %>%
    group_by(rgn_id, rgn_name, year) %>%
    summarise(
      n_gulf = n_distinct(documentidentifier),
      n_gulf_ocean = n_distinct(documentidentifier[ocean_engagement]),
      mean_valence = mean(anew_valence, na.rm = TRUE),
      mean_arousal = mean(anew_arousal, na.rm = TRUE),
      .groups = "drop"
    )

  gulf_year <- articles %>%
    distinct(documentidentifier, year, .keep_all = TRUE) %>%
    group_by(year) %>%
    summarise(
      n_gulf = n_distinct(documentidentifier),
      n_gulf_ocean = n_distinct(documentidentifier[ocean_engagement]),
      mean_valence = mean(anew_valence, na.rm = TRUE),
      mean_arousal = mean(anew_arousal, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(rgn_id = "Gulf", rgn_name = "Gulf of California")

  pop_gulf <- pop_by_rgn %>%
    summarise(
      rgn_id = "Gulf",
      rgn_name = "Gulf of California",
      pobtot = sum(pobtot, na.rm = TRUE)
    )

  pop_join <- bind_rows(
    pop_by_rgn %>% mutate(rgn_id = as.character(rgn_id)),
    pop_gulf
  )

  rgn_ids <- unique(c(
    as.character(pop_by_rgn$rgn_id),
    as.character(rgn_year$rgn_id),
    "Gulf"
  ))

  rgn_grid <- tidyr::expand_grid(rgn_id = rgn_ids, year = series_years) %>%
    left_join(
      bind_rows(rgn_year, gulf_year) %>% mutate(rgn_id = as.character(rgn_id)),
      by = c("rgn_id", "year")
    ) %>%
    left_join(
      pop_join %>% dplyr::select(rgn_id, pobtot, rgn_name_pop = rgn_name),
      by = "rgn_id"
    ) %>%
    mutate(
      rgn_name = dplyr::coalesce(
        rgn_name,
        rgn_name_pop,
        recode(rgn_id, !!!rgn_labels, .default = rgn_id)
      ),
      n_gulf = replace_na(n_gulf, 0),
      n_gulf_ocean = replace_na(n_gulf_ocean, 0),
      n_gulf_annualized = annualize_n(n_gulf, year),
      n_gulf_ocean_annualized = annualize_n(n_gulf_ocean, year),
      n_engagement = n_gulf_ocean_annualized,
      engagement_per_100k = if_else(
        !is.na(pobtot) & pobtot > 0,
        1e5 * n_engagement / pobtot,
        NA_real_
      )
    ) %>%
    dplyr::select(-rgn_name_pop)

  ref_by_rgn <- rgn_grid %>%
    filter(year %in% reference_years, !is.na(engagement_per_100k)) %>%
    group_by(rgn_id) %>%
    summarise(
      n_reference_years = n(),
      engagement_ref = mean(engagement_per_100k, na.rm = TRUE),
      valence_ref = mean(mean_valence, na.rm = TRUE),
      .groups = "drop"
    )

  engagement_annual <- rgn_grid

  engagement_status <- rgn_grid %>%
    left_join(ref_by_rgn, by = "rgn_id") %>%
    mutate(
      engagement_status = if_else(
        !is.na(engagement_ref) & engagement_ref > 0,
        100 * engagement_per_100k / engagement_ref,
        NA_real_
      ),
      valence_status = if_else(
        !is.na(valence_ref) & valence_ref > 0,
        100 * mean_valence / valence_ref,
        NA_real_
      ),
      reference_note = paste(
        "Mean of annual engagement per 100,000 coastal residents, 2019-2022.",
        "Status is uncapped."
      )
    )

  write_csv(engagement_annual, file.path(int_dir, "gdelt_engagement_per_capita_annual.csv"))
  write_csv(engagement_status, file.path(int_dir, "gdelt_engagement_status_2019_2022_ref.csv"))

  sentiment_annual <- engagement_status %>%
    dplyr::select(
      rgn_id, rgn_name, year, n_gulf, n_gulf_ocean,
      mean_valence, mean_arousal, valence_ref, valence_status
    )
  write_csv(sentiment_annual, file.path(int_dir, "gdelt_sentiment_annual.csv"))

  current_status_scores <- engagement_status %>%
    filter(rgn_id != "Gulf") %>%
    transmute(
      rgn_id,
      year,
      current_status = engagement_status
    ) %>%
    arrange(rgn_id, year)
  write_csv(
    current_status_scores,
    file.path(output_dir, "gdelt_news_engagement_current_status.csv")
  )

  engagement_status %>%
    filter(rgn_id == "Gulf") %>%
    dplyr::select(
      year, n_gulf_ocean, n_engagement, pobtot, engagement_per_100k,
      engagement_ref, engagement_status, mean_valence, valence_status
    )
} else if (nrow(gkg_bq) > 0) {
  message("GKG rows downloaded, but none had coordinates inside the 50 km Gulf layer.")
}
```

```{r}
#| fig-cap: "Gulf-wide news engagement per 100,000 coastal residents, 2019-2025. Engagement is the count of distinct Spanish GKG articles that name the Gulf of California / Sea of Cortez and also use ocean, coast, protest, or megaproject language, clipped to marine waters, islands, and 50 km inland. The dashed line is the mean of the 2019-2022 annual values (the historical reference)."
if (nrow(engagement_status) > 0) {
  gulf_line <- engagement_status %>% filter(rgn_id == "Gulf")
  ref_val <- unique(stats::na.omit(gulf_line$engagement_ref))[1]
  ggplot(gulf_line, aes(year, engagement_per_100k)) +
    geom_hline(yintercept = ref_val, linetype = "dashed", color = "gray40") +
    geom_line() +
    geom_point() +
    scale_x_continuous(breaks = series_years) +
    labs(
      x = "Year",
      y = "Articles per 100,000 coastal residents",
      title = "Gulf-wide engagement per capita vs 2019-2022 mean"
    ) +
    theme_minimal()
}
```

```{r}
#| fig-cap: "Current-status style score for Gulf-wide engagement: 100 times that year's engagement per capita divided by the 2019-2022 mean of annual values. 100 means matching the historical average. Values above 100 mean engagement is higher than that average. Not capped."
if (nrow(engagement_status) > 0) {
  gulf_line <- engagement_status %>% filter(rgn_id == "Gulf")
  ggplot(gulf_line, aes(year, engagement_status)) +
    geom_hline(yintercept = 100, linetype = "dashed", color = "gray40") +
    geom_line() +
    geom_point() +
    scale_x_continuous(breaks = series_years) +
    labs(
      x = "Year",
      y = "Engagement status (100 = 2019-2022 mean)",
      title = "Is Gulf news engagement rising relative to 2019-2022?"
    ) +
    theme_minimal()
}
```

```{r}
#| fig-cap: "Regional engagement status, 2019-2025. Each region is compared to its own 2019-2022 mean of annual engagement per 100,000 coastal residents. The dashed line at 100 is that region's historical average, not a Gulf-wide goal."
if (nrow(engagement_status) > 0) {
  rgn_line <- engagement_status %>% filter(rgn_id != "Gulf")
  ggplot(rgn_line, aes(year, engagement_status, color = rgn_name)) +
    geom_hline(yintercept = 100, linetype = "dashed", color = "gray40") +
    geom_line() +
    geom_point() +
    scale_x_continuous(breaks = series_years) +
    labs(
      x = "Year",
      y = "Engagement status (100 = that region's 2019-2022 mean)",
      color = "Region"
    ) +
    theme_minimal()
}
```

```{r}
#| fig-cap: "Mean Spanish ANEW valence (sentiment) in the same Spanish Gulf articles, 2019-2025. The dashed line is the mean of the 2019-2022 annual mean-valence values. This is not per capita."
if (nrow(engagement_status) > 0) {
  gulf_line <- engagement_status %>% filter(rgn_id == "Gulf")
  ref_val <- unique(stats::na.omit(gulf_line$valence_ref))[1]
  ggplot(gulf_line, aes(year, mean_valence)) +
    geom_hline(yintercept = ref_val, linetype = "dashed", color = "gray40") +
    geom_line() +
    geom_point() +
    scale_x_continuous(breaks = series_years) +
    coord_cartesian(ylim = c(1, 9)) +
    labs(
      x = "Year",
      y = "Mean Spanish ANEW valence (1-9)",
      title = "Gulf-wide news sentiment vs 2019-2022 mean"
    ) +
    theme_minimal()
}
```

Optional protest-event count from the Events table (CAMEO root 14). This is a separate check, not the per-capita engagement indicator.

```{r}
if (isTRUE(use_bigquery) && isTRUE(query_events) && nzchar(bq_project)) {
  events_sql <- paste0(
    "SELECT GLOBALEVENTID, SQLDATE, EventRootCode, EventCode, ",
    "ActionGeo_FullName, ActionGeo_ADM1Code, ActionGeo_Lat, ActionGeo_Long, ",
    "SOURCEURL, NumMentions, NumSources, GoldsteinScale, AvgTone ",
    "FROM `gdelt-bq.gdeltv2.events_partitioned` ",
    "WHERE _PARTITIONTIME >= TIMESTAMP('", series_start, "') ",
    "AND _PARTITIONTIME < TIMESTAMP('", series_end + 1, "') ",
    "AND ActionGeo_CountryCode = 'MX' ",
    "AND EventRootCode = '14' ",
    "AND ActionGeo_Lat IS NOT NULL AND ActionGeo_Long IS NOT NULL ",
    "AND ActionGeo_Lat BETWEEN ", bbox["ymin"], " AND ", bbox["ymax"], " ",
    "AND ActionGeo_Long BETWEEN ", bbox["xmin"], " AND ", bbox["xmax"], " ",
    "AND SQLDATE >= ", format(series_start, "%Y%m%d"), " ",
    "AND SQLDATE <= ", format(series_end, "%Y%m%d"), " ",
    "AND (REGEXP_CONTAINS(LOWER(IFNULL(ActionGeo_FullName, '')), r'", gulf_re_sql, "') ",
    "OR REGEXP_CONTAINS(LOWER(IFNULL(SOURCEURL, '')), r'", gulf_re_sql, "'))"
  )

  events <- bigrquery::bq_project_query(bq_project, events_sql) %>%
    bigrquery::bq_table_download()

  if (nrow(events) > 0) {
    events_sf <- events %>%
      st_as_sf(coords = c("ActionGeo_Long", "ActionGeo_Lat"), crs = 4326, remove = FALSE)

    events_in_goc <- st_join(
      events_sf,
      goc %>% dplyr::select(any_of("rgn_id")),
      join = st_intersects,
      left = FALSE
    ) %>%
      add_rgn_labels()

    write_csv(
      events_in_goc %>% st_drop_geometry(),
      file.path(int_dir, "gdelt_bigquery_protest_events_in_goc.csv")
    )

    events_in_goc %>%
      st_drop_geometry() %>%
      mutate(year = as.integer(substr(as.character(SQLDATE), 1, 4))) %>%
      count(year, rgn_name, name = "n_events")
  }
}
```

