---
title: 'OHI GoC 2025: Species Present in the Gulf of California'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: inline
editor_options: 
  chunk_output_type: console
---

# Summary

The species that inhabit the ocean have a vital role in shaping the existing marine environment. They are valued for their beauty and inherent right to exist, as well as their role in supporting productive habitats that provide many benefits for people.

From the largest predators to microscopic plankton, these species depend upon each other for survival. The interaction of species that have naturally evolved in a given location is essential for ecosystem structure and stability. In addition, the maintenance of large numbers of species provides diverse genetic resource that makes it possible to avoid functional collapse, should conditions change in the future.

This sub-goal assesses the health of all marine species present in the Gulf of California, including endangered species and species in relatively good conditions. The presence of species that are not at risk leads to a higher score.

# Exploring data sources

## The Gulf of California Invertebrate Database:The Invertebrate Portion of the Macrofauna Golfo Database

Link to download: <https://www.desertmuseum.org/center/seaofcortez/database.php>

Downloaded: April 29, 2025

Directions to download: Select Shallowest depth *equals 0 meters*; Greatest depth *is less than or equal to 15,000 meters* [artifical and arbitrary max, knowing that the deepest parts of the GoC are 10,000 meters]. Then download CSV file.  For v2025, there were 2340 results found.

Acronyms used in the database: <https://www.desertmuseum.org/center/seaofcortez/acronyms.php>

Data stored: home/shares/ohi/OHI_GOC/_raw_data/GoC_invert_database_desertmuseum/d2025/seaofcortez_export_0_15000meters.csv

## IUCN Red List v4

### Setup

```{r}
library(rredlist)

# use the IUCN API website (https://api.iucnredlist.org/), click the “Generate a token” link at the top of the web page, and fill out the form to apply for a token. You should then receive a token after completing the form. After receiving a token, open the .Renviron file on your computer (e.g., using usethis::edit_r_environ())
usethis::edit_r_environ() # and add the token as "IUCN_REDLIST_KEY"

# verify access to IUCN Red List API
rredlist::rl_use_iucn() # make sure you did the above ^!

# --------------------------------------

library(ohicore)
library(tidyverse)
library(stringr)
library(here)
library(janitor)
library(plotly)
library(tictoc)
library(ggplot2)
library(leaflet)
library(sf)
library(viridis)
library(dplyr)


# ---- sources! ----
# source(here("workflow", "R", "common.R")) # file creates objects to process data
## set the mazu and neptune data_edit share based on operating system
dir_M             <- c('Windows' = '//mazu.nceas.ucsb.edu/ohi',
                       'Darwin'  = '/Volumes/ohi',    ### connect (cmd-K) to smb://mazu/ohi
                       'Linux'   = '/home/shares/ohi')[[ Sys.info()[['sysname']] ]]

# ---- set year and file path info ----
current_year <- 2025 # Update this in the future!!
version_year <- paste0("v",current_year)
data_dir_version_year <- paste0("d", current_year)

# ---- data directories ----

# raw data directory (on Mazu)
raw_data_dir <- here::here(dir_M, "OHI_GOC", "_raw_data")

# for IUCN raw data dir
iucn_raw_dir <- here::here(dir_M, "OHI_GOC", "_raw_data", "IUCN_spatial", data_dir_version_year)

# for coastline characteristics intermediates
int_dir <- here::here("SPP_ICO", "spp", version_year, "int")

# final output dir
output_dir <- here("SPP_ICO", "spp", version_year, "output")

# spatial data for GoC
goc_spatial <- here("spatial")

# this CRS might be better for visualization, explore.
gulf_crs <- st_crs("+proj=aea +lat_1=23 +lat_2=30 +lat_0=25 +lon_0=-110 +datum=WGS84 +units=m +no_defs")
```

GoC Shapefiles:
```{r}
# read in ecoregion shapefiles in 
eco_hr_latlon <- st_read(here(goc_spatial, "GoC_polygon_ecoregion_latlon.shp"))
# mapview(eco_hr_latlon)
head(eco_hr_latlon)

# eco_hr_eqarea <- st_read(here(goc_spatial, "GoC_polygon_ecoregion_eqarea.shp"))
# mapview(eco_hr_eqarea)

# check CRS of ecoregion boundary
st_crs(eco_hr_latlon) # WGS84, can be used to crop global data

# transform to gulf_CRS
# eco_hr_albers <- st_transform(eco_hr_eqarea, crs = "+proj=aea +lat_1=23 +lat_2=30 +lat_0=25 +lon_0=-110 +datum=WGS84 +units=m +no_defs")
# st_crs(eco_hr_albers) # double check it worked well
# mapview(eco_hr_albers) # looks good!
```

### Download Mexico IUCN Data

For spatial data, manual polygon downloads are available at <https://www.iucnredlist.org/resources/spatial-data-download>.  The API will give the list of the status and name of all species in Mexico, but it does not have any spatial data associated with it (such as locations/coordinates seen).  After downloading all marine groups manually and cropping it to the GoC, we will be able to match it with listed species and their status from `mex_species`.  It is possible there are more taxonomic group distribution data than are listed, which is okay and still good to explore.  

Note: there are 19 groups (separate downloads) for spatial data.  Mazu location is /home/shares/ohi/OHI_GOC/_raw_data/IUCN_spatial/d2025.  After requesting each group, go to your account page under "saved downloads".  There, all requested downloads will be present.  Download each group and unzip, then bring into Mazu for the appropriate year. 

```{r}
mex_species <- rredlist::rl_countries("MX")  # GoC in MX

# isolate assessment results
mex_species_assessments <- mex_species[["assessments"]] %>% # can drop scopes for now
  select(-scopes)
View(mex_species_assessments) 

# write to mazu for easier access
write_csv(mex_species_assessments, here("/home/shares/ohi/OHI_GOC/goal_prep/spp/iucn_redlist/v2025", "mex_species_assessments.csv"))
```


#### 1. Abalone

```{r}
# read in shapefile for polygons from Mazu
abalone_global_raw <- st_read(here::here(iucn_raw_dir, "ABALONES","ABALONES.shp"))
# plot(abalone_global_raw) # looks good! already in WGS 84. 28 attributes

# clip to GoC by cropping the global data to the gulf ecoregion boundary
abalone_global_wgs84 <- st_transform(abalone_global_raw, crs = 4326) # ensure it is wgs84
unique(st_is_valid(abalone_global_wgs84))  # check validity: some are FALSE.  need to fix using `st_make_valid()` because some polygons cross each other
unique(st_geometry_type(abalone_global_wgs84))  # confirm geometry type: all multipolygons

## make valid, some issues with polygon validity
abalone_global_wgs84_valid <- st_make_valid(abalone_global_wgs84)
unique(st_is_valid(abalone_global_wgs84_valid)) # all true ~
head(abalone_global_wgs84_valid) # looking good

eco_hr_wgs84 <- st_transform(eco_hr_latlon, crs = 4326) # ensure it is wgs84

abalone_dist_goc_wgs84 <- st_intersection(abalone_global_wgs84_valid, eco_hr_wgs84)
head(abalone_dist_goc_wgs84) # only species in GoC: Haliotis rufescens

mapview(abalone_dist_goc_wgs84)

write_sf(abalone_dist_goc_wgs84, here::here(int_dir, "abalone_dist_goc_wgs84.shp"))
```


#### 2. Cone Snails

```{r}
# read in shapefile for polygons from Mazu
cone_snails_global_raw <- st_read(here::here(iucn_raw_dir, "CONE_SNAILS","CONE_SNAILS.shp"))
plot(cone_snails_global_raw[1]) # looks good! already in WGS 84. 28 attributes

# clip to GoC by cropping the global data to the gulf ecoregion boundary
cone_snails_global_wgs84 <- st_transform(cone_snails_global_raw, crs = 4326) # ensure it is wgs84
unique(st_is_valid(cone_snails_global_wgs84))  # check validity: all valid!!
unique(st_geometry_type(cone_snails_global_wgs84))  # confirm geometry type: all multipolygons

eco_hr_wgs84 <- st_transform(eco_hr_latlon, crs = 4326) # ensure it is wgs84

## make valid, some issues with polygon validity
cone_snails_global_wgs84_valid <- st_make_valid(cone_snails_global_wgs84)
unique(st_is_valid(cone_snails_global_wgs84_valid)) # all true ~
head(cone_snails_global_wgs84_valid) # looking good

cone_snail_dist_goc_wgs84 <- st_intersection(cone_snails_global_wgs84, eco_hr_wgs84)
length(unique(cone_snail_dist_goc_wgs84$sci_name)) # 30 cone snail species!!!

mapview(cone_snail_dist_goc_wgs84) # wide range of distribution. Could narrow down depending on if they are on the list 

write_sf(cone_snail_dist_goc_wgs84, here::here(int_dir, "cone_snail_dist_goc_wgs84.shp"))
```


#### 3. Croakers & Drums

```{r}
# read in shapefile for polygons from Mazu
croakers_drums_global_raw <- st_read(here::here(iucn_raw_dir, "CROAKERS_DRUMS","CROAKERS_DRUMS.shp"))
plot(croakers_drums_global_raw[1]) # looks good! already in WGS 84. 28 attributes

# clip to GoC by cropping the global data to the gulf ecoregion boundary
croakers_drums_global_wgs84 <- st_transform(croakers_drums_global_raw, crs = 4326) # ensure it is wgs84
unique(st_is_valid(croakers_drums_global_wgs84))  # check validity: NOT all valid, st_make_valid()
unique(st_geometry_type(croakers_drums_global_wgs84))  # confirm geometry type: all multipolygons

eco_hr_wgs84 <- st_transform(eco_hr_latlon, crs = 4326) # ensure it is wgs84

## make valid, some issues with polygon validity
croakers_drums_global_wgs84_valid <- st_make_valid(croakers_drums_global_wgs84)
unique(st_is_valid(croakers_drums_global_wgs84_valid)) # stil NOT all true...

## explore those that are not valid
invalid_indices <- which(!st_is_valid(croakers_drums_global_wgs84_valid))
problem_polygons <- croakers_drums_global_wgs84_valid[invalid_indices,]
mapview(problem_polygons) # not in Mexico, can drop

# drop the invlaid polygons
valid_logical <- st_is_valid(croakers_drums_global_wgs84_valid)
croakers_drums_global_wgs84_drop <- croakers_drums_global_wgs84_valid[valid_logical,] # now, 296 obs, correct! Dropped two invalide polygons

croakers_drums_dist_goc_wgs84 <- st_intersection(croakers_drums_global_wgs84_drop, eco_hr_wgs84)
length(unique(croakers_drums_dist_goc_wgs84$sci_name)) # 49 croaker and drum species!!!

mapview(croakers_drums_dist_goc_wgs84) # wide range of distribution. Could narrow down depending on if they are on the list.. on first glance I see totoaba :)

write_sf(croakers_drums_dist_goc_wgs84, here::here(int_dir, "croakers_drums_dist_goc_wgs84.shp"))
```

#### 4. Eels

```{r}
# read in shapefile for polygons from Mazu
eels_global_raw <- st_read(here::here(iucn_raw_dir, "EELS","EELS.shp"))
plot(eels_global_raw[1]) # looks good! already in WGS 84. 28 attributes

# clip to GoC by cropping the global data to the gulf ecoregion boundary
eels_global_wgs84 <- st_transform(eels_global_raw, crs = 4326) # ensure it is wgs84
unique(st_is_valid(eels_global_wgs84))  # check validity: NOT all valid, st_make_valid()
unique(st_geometry_type(eels_global_wgs84))  # confirm geometry type: all multipolygons

eco_hr_wgs84 <- st_transform(eco_hr_latlon, crs = 4326) # ensure it is wgs84

## make valid, some issues with polygon validity
eels_global_wgs84_valid <- st_make_valid(eels_global_wgs84)
unique(st_is_valid(eels_global_wgs84_valid)) # stil NOT all true...

## explore those that are not valid
invalid_indices <- which(!st_is_valid(eels_global_wgs84_valid))
problem_polygons <- eels_global_wgs84_valid[invalid_indices,]
mapview(problem_polygons) # none are within GoC, can drop

# drop the invlaid polygons
valid_logical <- st_is_valid(eels_global_wgs84_valid)
eels_global_wgs84_drop <- eels_global_wgs84_valid[valid_logical,] # Dropped invalid polygons

eels_dist_goc_wgs84 <- st_intersection(eels_global_wgs84_drop, eco_hr_wgs84)
length(unique(eels_dist_goc_wgs84$sci_name)) # 79 eel species in GoC!!!

mapview(eels_dist_goc_wgs84) # cool!! wide range of dist

write_sf(eels_dist_goc_wgs84, here::here(int_dir, "eels_dist_goc_wgs84.shp"))
```

#### 5. Groupers

```{r}
# read in shapefile for polygons from Mazu
groupers_global_raw <- st_read(here::here(iucn_raw_dir, "GROUPERS","GROUPERS.shp"))
plot(groupers_global_raw[1]) # looks good! already in WGS 84. 28 attributes

# clip to GoC by cropping the global data to the gulf ecoregion boundary
groupers_global_wgs84 <- st_transform(groupers_global_raw, crs = 4326) # ensure it is wgs84
unique(st_is_valid(groupers_global_wgs84))  # check validity: all valid!!!!
unique(st_geometry_type(groupers_global_wgs84))  # confirm geometry type: all multipolygons

eco_hr_wgs84 <- st_transform(eco_hr_latlon, crs = 4326) # ensure it is wgs84

groupers_dist_goc_wgs84 <- st_intersection(groupers_global_wgs84, eco_hr_wgs84)
length(unique(groupers_dist_goc_wgs84$sci_name)) # 17 grouper species in GoC!!!

mapview(groupers_dist_goc_wgs84) # all ova!

write_sf(groupers_dist_goc_wgs84, here::here(int_dir, "groupers_dist_goc_wgs84.shp"))
```

#### 6. Hagfish

```{r}
# read in shapefile for polygons from Mazu
hagfish_global_raw <- st_read(here::here(iucn_raw_dir, "HAGFISH","HAGFISH.shp"))
plot(hagfish_global_raw[1]) # looks good! already in WGS 84. 28 attributes

# clip to GoC by cropping the global data to the gulf ecoregion boundary
hagfish_global_wgs84 <- st_transform(hagfish_global_raw, crs = 4326) # ensure it is wgs84
unique(st_is_valid(hagfish_global_wgs84))  # check validity: not all valid
unique(st_geometry_type(hagfish_global_wgs84))  # confirm geometry type: all multipolygons

## make valid, some issues with polygon validity
hagfish_global_wgs84_valid <- st_make_valid(hagfish_global_wgs84)
unique(st_is_valid(hagfish_global_wgs84_valid)) # all true now!

eco_hr_wgs84 <- st_transform(eco_hr_latlon, crs = 4326) # ensure it is wgs84

hagfish_dist_goc_wgs84 <- st_intersection(hagfish_global_wgs84_valid, eco_hr_wgs84)
length(unique(hagfish_dist_goc_wgs84$sci_name)) # 4 hagfish species in GoC!!!

mapview(hagfish_dist_goc_wgs84) # all southern gulf

write_sf(hagfish_dist_goc_wgs84, here::here(int_dir, "hagfish_dist_goc_wgs84.shp"))
```

#### 7. Lobsters

```{r}
# read in shapefile for polygons from Mazu
lobsters_global_raw <- st_read(here::here(iucn_raw_dir, "LOBSTERS","LOBSTERS.shp"))
plot(lobsters_global_raw[1]) # looks good! already in WGS 84. 28 attributes

# clip to GoC by cropping the global data to the gulf ecoregion boundary
lobster_global_wgs84 <- st_transform(lobsters_global_raw, crs = 4326) # ensure it is wgs84
unique(st_is_valid(lobster_global_wgs84))  # check validity: all valid!!
unique(st_geometry_type(lobster_global_wgs84))  # confirm geometry type: all multipolygons

eco_hr_wgs84 <- st_transform(eco_hr_latlon, crs = 4326) # ensure it is wgs84

lobsters_dist_goc_wgs84 <- st_intersection(lobster_global_wgs84, eco_hr_wgs84)
length(unique(lobsters_dist_goc_wgs84$sci_name)) # 9 lobster species in GoC!!! but the year compiled is 2011...

mapview(lobsters_dist_goc_wgs84) # whole gulf!

write_sf(lobsters_dist_goc_wgs84, here::here(int_dir, "lobsters_dist_goc_wgs84.shp"))
```

#### * 8. Mammals (Marine Only)

```{r}
# read in shapefile for polygons from Mazu
marine_mammals_global_raw <- st_read(here::here(iucn_raw_dir, "MAMMALS_MARINE_ONLY","MAMMALS_MARINE_ONLY.shp"))
plot(marine_mammals_global_raw[1]) # looks good! already in WGS 84. 28 attributes

# clip to GoC by cropping the global data to the gulf ecoregion boundary
marine_mammals_global_wgs84 <- st_transform(marine_mammals_global_raw, crs = 4326) # ensure it is wgs84
unique(st_is_valid(marine_mammals_global_wgs84))  # check validity: not all valid...
unique(st_geometry_type(marine_mammals_global_wgs84))  # confirm geometry type: all multipolygons

eco_hr_wgs84 <- st_transform(eco_hr_latlon, crs = 4326) # ensure it is wgs84

## make valid, some issues with polygon validity
marine_mammals_global_wgs84_valid <- st_make_valid(marine_mammals_global_wgs84)
unique(st_is_valid(marine_mammals_global_wgs84_valid)) # not all true still

## explore those that are not valid
invalid_indices <- which(!st_is_valid(marine_mammals_global_wgs84_valid))
problem_polygons <- marine_mammals_global_wgs84_valid[invalid_indices,]
mapview(problem_polygons) # some are within the GoC...

# dissolve the polygons without losing attributes:
library(igraph)

# find overlapping clusters
overlap_matrix <- st_intersects(marine_mammals_global_wgs84_valid)
# graph <- graph_from_adjacency_matrix(overlap_matrix)
# clusters <- components(graph)$membership
# 
# # merge polygons by cluster
# merged_data <- your_data %>%
#   mutate(cluster = clusters) %>%
#   group_by(cluster) %>%
#   summarize(do_union = TRUE)

marine_mammals_dist_goc_wgs84 <- st_intersection(marine_mammals_global_wgs84_valid, eco_hr_wgs84)
length(unique(marine_mammals_dist_goc_wgs84$sci_name)) 

mapview(marine_mammals_dist_goc_wgs84)

write_sf(marine_mammals_dist_goc_wgs84, here::here(int_dir, "marine_mammals_dist_goc_wgs84.shp"))
```

#### 9. Mangroves (HAB)

```{r}
# read in shapefile for polygons from Mazu
mangroves_global_raw <- st_read(here::here(iucn_raw_dir, "MANGROVES","MANGROVES.shp"))

plot(mangroves_global_raw[1]) # looks good! already in WGS 84. 28 attributes

# clip to GoC by cropping the global data to the gulf ecoregion boundary
mangroves_global_wgs84 <- st_transform(mangroves_global_raw, crs = 4326) # ensure it is wgs84
unique(st_is_valid(mangroves_global_wgs84))  # check validity: all valid!!
unique(st_geometry_type(mangroves_global_wgs84))  # confirm geometry type: all multipolygons

eco_hr_wgs84 <- st_transform(eco_hr_latlon, crs = 4326) # ensure it is wgs84

mangroves_dist_goc_wgs84 <- st_intersection(mangroves_global_wgs84, eco_hr_wgs84)
length(unique(mangroves_dist_goc_wgs84$sci_name)) # 4 mangrove species in GoC!!! but the year compiled is 2010...

mapview(mangroves_dist_goc_wgs84) # whole gulf!

write_sf(mangroves_dist_goc_wgs84, here::here(int_dir, "mangroves_dist_goc_wgs84.shp"))
```

#### * 10. Marinefish

```{r}
# # read in shapefile for polygons from Mazu
# marinefish_global_raw <- st_read(here::here(iucn_raw_dir, "MARINEFISH","MARINEFISH.shp"))
# 
# plot(mangroves_global_raw[1]) # looks good! already in WGS 84. 28 attributes
# 
# # clip to GoC by cropping the global data to the gulf ecoregion boundary
# mangroves_global_wgs84 <- st_transform(mangroves_global_raw, crs = 4326) # ensure it is wgs84
# unique(st_is_valid(mangroves_global_wgs84))  # check validity: all valid!!
# unique(st_geometry_type(mangroves_global_wgs84))  # confirm geometry type: all multipolygons
# 
# eco_hr_wgs84 <- st_transform(eco_hr_latlon, crs = 4326) # ensure it is wgs84
# 
# mangroves_dist_goc_wgs84 <- st_intersection(mangroves_global_wgs84, eco_hr_wgs84)
# length(unique(mangroves_dist_goc_wgs84$sci_name)) # 4 mangrove species in GoC!!! but the year compiled is 2010...
# 
# mapview(mangroves_dist_goc_wgs84) # whole gulf!
# 
# write_sf(mangroves_dist_goc_wgs84, here::here("/home/shares/ohi/OHI_GOC/_raw_data/IUCN_spatial/d2025/int/mangroves_dist_goc_wgs84.shp"))
```


#### * 11. Reef Forming Corals

```{r}
# # read in shapefile for polygons from Mazu
# reef_forming_corals_global_raw <- st_read(here::here(iucn_raw_dir, "REEF_FORMING_CORALS","REEF_FORMING_CORALS.shp"))
# 
# plot(reef_forming_corals_global_raw[1]) # looks good! already in WGS 84. 28 attributes
# 
# # clip to GoC by cropping the global data to the gulf ecoregion boundary
# reef_forming_corals_global_wgs84 <- st_transform(reef_forming_corals_global_raw, crs = 4326) # ensure it is wgs84
# unique(st_is_valid(reef_forming_corals_global_wgs84))  # check validity: all valid!!
# unique(st_geometry_type(reef_forming_corals_global_wgs84))  # confirm geometry type: all multipolygons
# 
# eco_hr_wgs84 <- st_transform(eco_hr_latlon, crs = 4326) # ensure it is wgs84
# 
# reef_forming_corals_dist_goc_wgs84 <- st_intersection(reef_forming_corals_global_wgs84, eco_hr_wgs84)
# length(unique(reef_forming_corals_dist_goc_wgs84$sci_name)) # 4 mangrove species in GoC!!! but the year compiled is 2010...
# 
# mapview(reef_forming_corals_dist_goc_wgs84) # whole gulf!
# 
# write_sf(reef_forming_corals_dist_goc_wgs84, here::here("/home/shares/ohi/OHI_GOC/_raw_data/IUCN_spatial/d2025/int/reef_forming_corals_dist_goc_wgs84.shp"))
```


#### 12. Salmonids

```{r}
# read in shapefile for polygons from Mazu
salmonids_global_raw <- st_read(here::here(iucn_raw_dir, "SALMONIDS","SALMONIDS.shp"))

plot(salmonids_global_raw[1]) # looks good! already in WGS 84. 28 attributes

# clip to GoC by cropping the global data to the gulf ecoregion boundary
salmonids_global_wgs84 <- st_transform(salmonids_global_raw, crs = 4326) # ensure it is wgs84
unique(st_is_valid(salmonids_global_wgs84))  # check validity: not all valid!!
unique(st_geometry_type(salmonids_global_wgs84))  # confirm geometry type: all multipolygons

eco_hr_wgs84 <- st_transform(eco_hr_latlon, crs = 4326) # ensure it is wgs84

## make valid, some issues with polygon validity
salmonids_global_wgs84_valid <- st_make_valid(salmonids_global_wgs84)
unique(st_is_valid(salmonids_global_wgs84_valid)) # not all true still

## explore those that are not valid
invalid_indices <- which(!st_is_valid(salmonids_global_wgs84_valid))
problem_polygons <- salmonids_global_wgs84_valid[invalid_indices,]
mapview(problem_polygons) # none are in the goc!! lets drop them.

# drop the invlaid polygons
valid_logical <- st_is_valid(salmonids_global_wgs84_valid)
salmonids_global_wgs84_drop <- salmonids_global_wgs84_valid[valid_logical,] 

salmonids_dist_goc_wgs84 <- st_intersection(salmonids_global_wgs84_drop, eco_hr_wgs84)
length(unique(salmonids_dist_goc_wgs84$sci_name)) # coho and rainbow trout in gulf of cali!

mapview(salmonids_dist_goc_wgs84) # only near cabo, compiled info in 2020

write_sf(salmonids_dist_goc_wgs84, here::here("/home/shares/ohi/OHI_GOC/_raw_data/IUCN_spatial/d2025/int/salmonids_dist_goc_wgs84.shp"))
```


#### 13. Seabreams, Snappers, Grunts

```{r}
# read in shapefile for polygons from Mazu
seabreams_snappers_grunts_global_raw <- st_read(here::here(iucn_raw_dir, "SEABREAMS_SNAPPERS_GRUNTS","SEABREAMS_SNAPPERS_GRUNTS.shp"))

plot(seabreams_snappers_grunts_global_raw[1]) # looks good! already in WGS 84. 28 attributes

# clip to GoC by cropping the global data to the gulf ecoregion boundary
seabreams_snappers_grunts_global_wgs84 <- st_transform(seabreams_snappers_grunts_global_raw, crs = 4326) # ensure it is wgs84
unique(st_is_valid(seabreams_snappers_grunts_global_wgs84))  # check validity: ALL TRUE!
unique(st_geometry_type(seabreams_snappers_grunts_global_wgs84))  # confirm geometry type: all multipolygons

eco_hr_wgs84 <- st_transform(eco_hr_latlon, crs = 4326) # ensure it is wgs84

seabreams_snappers_grunts_dist_goc_wgs84 <- st_intersection(seabreams_snappers_grunts_global_wgs84, eco_hr_wgs84)
length(unique(seabreams_snappers_grunts_dist_goc_wgs84$sci_name)) # 36 species!

mapview(seabreams_snappers_grunts_dist_goc_wgs84) # whole gulf!

write_sf(seabreams_snappers_grunts_dist_goc_wgs84, here::here("/home/shares/ohi/OHI_GOC/_raw_data/IUCN_spatial/d2025/int/seabreams_snappers_grunts_dist_goc_wgs84.shp"))
```


#### 14. Seagrasses (HAB)

```{r}
# read in shapefile for polygons from Mazu
seagrasses_global_raw <- st_read(here::here(iucn_raw_dir, "SEAGRASSES","SEAGRASSES.shp"))

plot(seagrasses_global_raw[1]) # looks good! already in WGS 84. 28 attributes

# clip to GoC by cropping the global data to the gulf ecoregion boundary
seagrasses_global_wgs84 <- st_transform(seagrasses_global_raw, crs = 4326) # ensure it is wgs84
unique(st_is_valid(seagrasses_global_wgs84))  # check validity: ALL TRUE!
unique(st_geometry_type(seagrasses_global_wgs84))  # confirm geometry type: all multipolygons

eco_hr_wgs84 <- st_transform(eco_hr_latlon, crs = 4326) # ensure it is wgs84

seagrasses_dist_goc_wgs84 <- st_intersection(seagrasses_global_wgs84, eco_hr_wgs84)
length(unique(seagrasses_dist_goc_wgs84$sci_name)) # 5 seagrass species!!

mapview(seagrasses_dist_goc_wgs84) # whole gulf!

write_sf(seagrasses_dist_goc_wgs84, here::here("/home/shares/ohi/OHI_GOC/_raw_data/IUCN_spatial/d2025/int/seagrasses_dist_goc_wgs84.shp"))
```

#### * 15. Sharks, Rays, Chimaeras

```{r}
# read in shapefile for polygons from Mazu
sharks_rays_chimaeras_global_raw <- st_read(here::here(iucn_raw_dir, "SHARKS_RAYS_CHIMAERAS","SHARKS_RAYS_CHIMAERAS.shp"))

plot(sharks_rays_chimaeras_global_raw[1]) # looks good! already in WGS 84. 28 attributes

# clip to GoC by cropping the global data to the gulf ecoregion boundary
sharks_rays_chimaeras_global_wgs84 <- st_transform(sharks_rays_chimaeras_global_raw, crs = 4326) # ensure it is wgs84
unique(st_is_valid(sharks_rays_chimaeras_global_wgs84))  # check validity: not all true.
unique(st_geometry_type(sharks_rays_chimaeras_global_wgs84))  # confirm geometry type: all multipolygons

eco_hr_wgs84 <- st_transform(eco_hr_latlon, crs = 4326) # ensure it is wgs84

## make valid, some issues with polygon validity
sharks_rays_chimaeras_global_wgs84_valid <- st_make_valid(sharks_rays_chimaeras_global_wgs84)
unique(st_is_valid(sharks_rays_chimaeras_global_wgs84_valid)) # not all true still

## explore those that are not valid
invalid_indices <- which(!st_is_valid(sharks_rays_chimaeras_global_wgs84_valid))
problem_polygons <- sharks_rays_chimaeras_global_wgs84_valid[invalid_indices,]
mapview(problem_polygons) # some are in GoC...

# # drop the invlaid polygons
# valid_logical <- st_is_valid(sharks_rays_chimaeras_global_wgs84_valid)
# sharks_rays_chimaeras_global_wgs84_drop <- sharks_rays_chimaeras_global_wgs84_valid[valid_logical,] 

sharks_rays_chimaeras_dist_goc_wgs84 <- st_intersection(sharks_rays_chimaeras_global_wgs84_valid, eco_hr_wgs84)
length(unique(sharks_rays_chimaeras_dist_goc_wgs84$sci_name)) # 

mapview(sharks_rays_chimaeras_dist_goc_wgs84) # 

write_sf(sharks_rays_chimaeras_dist_goc_wgs84, here::here("/home/shares/ohi/OHI_GOC/_raw_data/IUCN_spatial/d2025/int/sharks_rays_chimaeras_dist_goc_wgs84.shp"))
```

#### 16. Sturgeons & Paddlefishes

```{r}
# read in shapefile for polygons from Mazu
sturgeons_paddlefish_global_raw <- st_read(here::here(iucn_raw_dir, "STURGEONS_PADDLEFISHES","STURGEONS_PADDLEFISHES.shp"))

plot(sturgeons_paddlefish_global_raw[1]) # looks good! already in WGS 84. 28 attributes

# clip to GoC by cropping the global data to the gulf ecoregion boundary
sturgeons_paddlefish_global_wgs84 <- st_transform(sturgeons_paddlefish_global_raw, crs = 4326) # ensure it is wgs84
unique(st_is_valid(sturgeons_paddlefish_global_wgs84))  # check validity: not all true.
unique(st_geometry_type(sturgeons_paddlefish_global_wgs84))  # confirm geometry type: all multipolygons

eco_hr_wgs84 <- st_transform(eco_hr_latlon, crs = 4326) # ensure it is wgs84

## make valid, some issues with polygon validity
sturgeons_paddlefish_global_wgs84_valid <- st_make_valid(sturgeons_paddlefish_global_wgs84)
unique(st_is_valid(sturgeons_paddlefish_global_wgs84_valid)) # not all true still...

## explore those that are not valid
invalid_indices <- which(!st_is_valid(sturgeons_paddlefish_global_wgs84_valid))
problem_polygons <- sturgeons_paddlefish_global_wgs84_valid[invalid_indices,]
mapview(problem_polygons) # none are in the goc!! lets drop them.

# drop the invlaid polygons
valid_logical <- st_is_valid(sturgeons_paddlefish_global_wgs84_valid)
sturgeons_paddlefish_global_wgs84_drop <- sturgeons_paddlefish_global_wgs84_valid[valid_logical,] 

sturgeons_paddlefish_dist_goc_wgs84 <- st_intersection(sturgeons_paddlefish_global_wgs84_drop, eco_hr_wgs84)
length(unique(sturgeons_paddlefish_dist_goc_wgs84$sci_name)) # only 1 species, Acipenser medirostris

mapview(sturgeons_paddlefish_dist_goc_wgs84) # near la paz, compiled 2021

write_sf(sturgeons_paddlefish_dist_goc_wgs84, here::here("/home/shares/ohi/OHI_GOC/_raw_data/IUCN_spatial/d2025/int/sturgeons_paddlefish_dist_goc_wgs84.shp"))
```

#### 17. Syngnathiform Fishes

```{r}
# read in shapefile for polygons from Mazu
syngnathiform_global_raw <- st_read(here::here(iucn_raw_dir, "SYNGNATHIFORM_FISHES","SYNGNATHIFORM_FISHES.shp"))

plot(syngnathiform_global_raw[1]) # looks good! already in WGS 84. 28 attributes

# clip to GoC by cropping the global data to the gulf ecoregion boundary
syngnathiform_global_wgs84 <- st_transform(syngnathiform_global_raw, crs = 4326) # ensure it is wgs84
unique(st_is_valid(syngnathiform_global_wgs84))  # check validity: not all true.
unique(st_geometry_type(syngnathiform_global_wgs84))  # confirm geometry type: all multipolygons

eco_hr_wgs84 <- st_transform(eco_hr_latlon, crs = 4326) # ensure it is wgs84

## make valid, some issues with polygon validity
syngnathiform_global_wgs84_valid <- st_make_valid(syngnathiform_global_wgs84)
unique(st_is_valid(syngnathiform_global_wgs84_valid)) # not all true still

## explore those that are not valid
invalid_indices <- which(!st_is_valid(syngnathiform_global_wgs84_valid))
problem_polygons <- syngnathiform_global_wgs84_valid[invalid_indices,]
mapview(problem_polygons) # none are in the goc!! lets drop them.

# drop the invlaid polygons
valid_logical <- st_is_valid(syngnathiform_global_wgs84_valid)
syngnathiform_global_wgs84_drop <- syngnathiform_global_wgs84_valid[valid_logical,] 

syngnathiform_dist_goc_wgs84 <- st_intersection(syngnathiform_global_wgs84_drop, eco_hr_wgs84)
length(unique(syngnathiform_dist_goc_wgs84$sci_name)) # 9 synathiform fishes! compiled 2010-2016

mapview(syngnathiform_dist_goc_wgs84) # wide range throughout!

write_sf(syngnathiform_dist_goc_wgs84, here::here("/home/shares/ohi/OHI_GOC/_raw_data/IUCN_spatial/d2025/int/syngnathiform_dist_goc_wgs84.shp"))
```

#### 18. Tunas, Billfishes, Swordfish

```{r}
# read in shapefile for polygons from Mazu
tunas_billfishes_swordfish_global_raw <- st_read(here::here(iucn_raw_dir, "TUNAS_BILLFISHES_SWORDFISH","TUNAS_BILLFISHES_SWORDFISH.shp"))

plot(tunas_billfishes_swordfish_global_raw[1]) # looks good! already in WGS 84. 28 attributes

# clip to GoC by cropping the global data to the gulf ecoregion boundary
tunas_billfishes_swordfish_global_wgs84 <- st_transform(tunas_billfishes_swordfish_global_raw, crs = 4326) # ensure it is wgs84
unique(st_is_valid(tunas_billfishes_swordfish_global_wgs84))  # check validity: not all true.
unique(st_geometry_type(tunas_billfishes_swordfish_global_wgs84))  # confirm geometry type: all multipolygons

eco_hr_wgs84 <- st_transform(eco_hr_latlon, crs = 4326) # ensure it is wgs84

## make valid, some issues with polygon validity
tunas_billfishes_swordfish_global_wgs84_valid <- st_make_valid(tunas_billfishes_swordfish_global_wgs84)
unique(st_is_valid(tunas_billfishes_swordfish_global_wgs84_valid)) # not all true still

## explore those that are not valid
invalid_indices <- which(!st_is_valid(tunas_billfishes_swordfish_global_wgs84_valid))
problem_polygons <- tunas_billfishes_swordfish_global_wgs84_valid[invalid_indices,]
mapview(problem_polygons) # none are in the goc!! lets drop them.

# drop the invlaid polygons
valid_logical <- st_is_valid(tunas_billfishes_swordfish_global_wgs84_valid)
tunas_billfishes_swordfish_global_wgs84_drop <- tunas_billfishes_swordfish_global_wgs84_valid[valid_logical,] 

tunas_billfishes_swordfish_dist_goc_wgs84 <- st_intersection(tunas_billfishes_swordfish_global_wgs84_drop, eco_hr_wgs84)
length(unique(tunas_billfishes_swordfish_dist_goc_wgs84$sci_name)) # 19 species! very recently compiled

mapview(tunas_billfishes_swordfish_dist_goc_wgs84) # full distribution through gulf

write_sf(tunas_billfishes_swordfish_dist_goc_wgs84, here::here("/home/shares/ohi/OHI_GOC/_raw_data/IUCN_spatial/d2025/int/tunas_billfishes_swordfish_dist_goc_wgs84.shp"))
```

#### 19. Wrassess & Parrotfish

```{r}
# read in shapefile for polygons from Mazu
wrasses_parrotfishes_global_raw <- st_read(here::here(iucn_raw_dir, "WRASSES_PARROTFISHES","WRASSES_PARROTFISHES.shp"))

plot(wrasses_parrotfishes_global_raw[1]) # looks good! already in WGS 84. 28 attributes

# clip to GoC by cropping the global data to the gulf ecoregion boundary
wrasses_parrotfishes_global_wgs84 <- st_transform(wrasses_parrotfishes_global_raw, crs = 4326) # ensure it is wgs84
unique(st_is_valid(wrasses_parrotfishes_global_wgs84))  # check validity: not all true.
unique(st_geometry_type(wrasses_parrotfishes_global_wgs84))  # confirm geometry type: all multipolygons

eco_hr_wgs84 <- st_transform(eco_hr_latlon, crs = 4326) # ensure it is wgs84

## make valid, some issues with polygon validity
wrasses_parrotfishes_global_wgs84_valid <- st_make_valid(wrasses_parrotfishes_global_wgs84)
unique(st_is_valid(wrasses_parrotfishes_global_wgs84_valid)) # all true now!!!

wrasses_parrotfishes_dist_goc_wgs84 <- st_intersection(wrasses_parrotfishes_global_wgs84_valid, eco_hr_wgs84)
length(unique(wrasses_parrotfishes_dist_goc_wgs84$sci_name)) # 24 species in GoC!

mapview(wrasses_parrotfishes_dist_goc_wgs84) # in shallower waters

write_sf(wrasses_parrotfishes_dist_goc_wgs84, here::here("/home/shares/ohi/OHI_GOC/_raw_data/IUCN_spatial/d2025/int/wrasses_parrotfishes_dist_goc_wgs84.shp"))
```

#### Join all groups


```{r}
# define the paths for both directories
int_dir <- here("SPP_ICO", "spp", version_year, "int")  # local int files
iucn_dir <- file.path(dir_M, "OHI_GOC", "_raw_data", "IUCN_spatial", data_dir_version_year)  # Mazu files

# all distribution files from BOTH directories
shp_paths <- c(
  list.files(int_dir, pattern = "_dist_goc_wgs84\\.shp$", full.names = TRUE),
  list.files(iucn_dir, pattern = "_dist_goc_wgs84\\.shp$", full.names = TRUE)
)

# make sure files exist
stopifnot(all(file.exists(shp_paths)))

# different capitalizations and spellings in some
col_names <- lapply(shp_paths, function(p) names(st_read(p, quiet = TRUE)))
print(unique(col_names))

# read all shapefiles, clean column names, and collect all unique column names
sf_list <- map(shp_paths, ~{
  sf <- st_read(.x, quiet = TRUE)
  sf <- clean_names(sf)
  sf
})

# find all unique column names across all shapefiles
all_cols <- sf_list %>%
  map(~ names(.)) %>%
  unlist() %>%
  unique()

# add missing columns as NA to each sf object, so all have the same columns
sf_list_harmonized <- map(sf_list, ~{
  missing_cols <- setdiff(all_cols, names(.))
  if (length(missing_cols) > 0) {
    for (col in missing_cols) {
      .[[col]] <- NA
    }
  }
  # reorder columns to match all_cols
  . <- .[, all_cols]
  .
})

# combine them all into one sf object
all_species_goc <- bind_rows(sf_list_harmonized)

# write to mazu for easier access
write_sf(all_species_goc, here("/home/shares/ohi/OHI_GOC/goal_prep/spp/iucn_redlist/v2025", "all_species_goc.shp"))
```

#### Interactive DT

```{r}
library(DT)

# remove geometry column, keeping only attributes
all_species_goc_attributes <- st_drop_geometry(all_species_goc)

# write to mazu for easier access
write_csv(all_species_goc_attributes, here("/home/shares/ohi/OHI_GOC/goal_prep/spp/iucn_redlist/v2025", "all_species_goc_attributes.csv"))

# make interactive DT table!
datatable(all_species_goc_attributes, options = list(scrollX = TRUE))
```


### Cross Reference the IUCN Status Codes

Use `mex_species_assessments` and `all_species_goc` to determine if the status code has been correctly updated.

```{r}
# isolate columns of interest and rename for better clarification
all_species_goc_attributes_xref <- all_species_goc_attributes %>% 
  select(id_no, sci_name, category) %>% 
  rename(id_no = id_no,
         sci_name = sci_name,
         category_goc = category)

mex_species_assessments_xref <- mex_species_assessments %>% 
  select(id_no = sis_taxon_id,
         sci_name = taxon_scientific_name,
         category_mex = red_list_category_code)

# join the spatial and redlist data
mex_species_assessments_join <- full_join(mex_species_assessments_xref, all_species_goc_attributes_xref, by = c("sci_name", "id_no")) %>% 
  mutate(category_maintained = case_when(
    category_mex == category_goc ~ "1",
    TRUE ~ "0"
  ))

# the number of maintained species
goc_species_assessments_xref <- mex_species_assessments_join %>% 
  filter(category_maintained == "1") %>% 
  distinct()

length(unique(goc_species_assessments_xref$sci_name))

# write to mazu for easier access
write_csv(goc_species_assessments_xref, here("/home/shares/ohi/OHI_GOC/goal_prep/spp/iucn_redlist/v2025", "goc_species_assessments_xref.csv"))
```


### Interactive ways to display the IUCN data


```{r}
# rename for joining
mex_species_assessments_rename <- mex_species_assessments %>% 
  rename(id_no = sis_taxon_id,
         sci_name = taxon_scientific_name,
         category_mex = red_list_category_code)

# join the species with the iucn URLs from the mex redlist
goc_species_assessments_xref_viz <- left_join(goc_species_assessments_xref, mex_species_assessments_rename, by = c("id_no","sci_name", "category_mex")) %>% 
  distinct() %>% 
  filter(latest == "TRUE")

datatable(goc_species_assessments_xref_viz)

# write to mazu for easier access
write_csv(goc_species_assessments_xref_viz, here("/home/shares/ohi/OHI_GOC/goal_prep/spp/iucn_redlist/v2025", "goc_species_assessments_xref_viz.csv"))
```

```{r}
library(plotly)

# sum the counts for each category_mex
pie_data <- goc_species_assessments_xref_viz %>%
  count(category_mex)

# make the interactive pie chart for the categories!!
pie_cat_fig <- plot_ly(
  data = pie_data,
  labels = ~category_mex,
  values = ~n,
  type = "pie",
  textinfo = "label+percent",
  insidetextorientation = "radial"
) %>%
  layout(title = "Species by Mexican Red List Category")

pie_cat_fig
```

```{r}
library(reactable)

reactable(
  goc_species_assessments_xref_viz,
  groupBy = "category_mex",
  searchable = TRUE,
  filterable = TRUE,
  striped = TRUE,
  highlight = TRUE,
  columns = list(
    category_mex = colDef(name = "Red List Category"),
    sci_name = colDef(name = "Scientific Name"),
    year_published = colDef(name = "Year Published"),
    
  )
)

```




