---
title: "Gulf of California beach access — main workflow"
output: html_document
date: "`r Sys.Date()`"
---

This notebook builds the analyst locality list into beach worksheets, clipped support layers, and per-locality support maps.

**External inputs** (not stored in `_pressures/access`):

- INEGI locality polygons: `/home/shares/ohi/OHI_GOC/goal_prep/pressures/population_localities/inegi_boundaries_pop.shp`
- INEGI roads and sites: `/home/shares/ohi/OHI_GOC/_raw_data/inegi_roads/.../red_vial.shp`, `sitio_de_interes.shp`
- CONANP MPAs: `/home/shares/ohi/OHI_GOC/_raw_data/CONANP/anpmx/anpmx.shp`

**Local inputs** in `data/`:

- `GofC_access_localities.csv` — analyst locality names with INEGI keys
- `beach.shp` — Gulf beach linework
- `paths_osm_Jan7_2025.shp`, `parking_osm_Jan7_2025.shp` — Gulf-wide OSM cache (clipped per locality)

Run chunks in order from the repo root.

```{r setup, message=FALSE}
library(here)
library(dplyr)
library(sf)
library(purrr)

access_dir <- here("_pressures/access")
data_dir <- file.path(access_dir, "data")
source(file.path(access_dir, "access_utils.R"), local = TRUE)

paths <- default_external_paths(data_dir)
paths$osm_paths <- st_read(paths$osm_paths, quiet = TRUE) |> st_make_valid()
paths$osm_parking <- st_read(paths$osm_parking, quiet = TRUE) |> st_make_valid()
paths$mpa <- st_read(paths$mpa, quiet = TRUE) |> st_make_valid()

localities <- load_access_localities(data_dir)
locality_names <- sort(unique(localities$city))
length(locality_names)
```

## 1. Overview map of all localities

```{r overview-map, eval=FALSE}
build_localities_overview_map(data_dir, access_dir, paths = default_external_paths(data_dir))
```

Output: `output/all_localities_map.html`

## 2. Identify beaches within each 20 km buffer

Writes one CSV per analyst locality to `data/locality_beaches/{slug}_beaches.csv`. Existing expert scores in those CSVs are preserved.

```{r identify-beaches, eval=FALSE}
identify_all_locality_beaches(data_dir, paths = default_external_paths(data_dir))
```

## 3. Extract support layers per locality

Full INEGI/OSM roads, paths, parking, sites, and MPAs clipped to each locality's 20 km study region (no thinning). Outputs GeoPackages under `data/locality_support/{slug}/`.

Set `REFETCH_OSM <- TRUE` only if the Gulf-wide OSM cache is missing (slow; needs network).

```{r extract-support, eval=FALSE}
REFETCH_OSM <- FALSE

for (nm in locality_names) {
  message("=== Extract: ", nm, " ===")
  extract_locality_support(
    locality_nm = nm,
    data_dir = data_dir,
    access_dir = access_dir,
    paths = if (REFETCH_OSM) default_external_paths(data_dir) else paths,
    refetch_osm = REFETCH_OSM
  )
}

bind_rows(map(locality_names, \(nm) {
  p <- file.path(locality_support_dir(nm, data_dir), "extraction_summary.csv")
  if (file.exists(p)) readr::read_csv(p, show_col_types = FALSE) else NULL
})) |>
  readr::write_csv(file.path(data_dir, "locality_support", "extraction_summary_all_localities.csv"))
```

## 4. Build per-locality support maps

Interactive Leaflet maps with 20 km buffer, roads, paths, parking, beaches, and INEGI sites.

```{r support-maps, eval=FALSE}
for (nm in locality_names) {
  message("=== Map: ", nm, " ===")
  build_locality_support_map(nm, data_dir, access_dir)
}
```

Output: `localities_support_layers/{slug}_support_layers.html` (+ auto-generated `_files/` assets)
