---
title: "Beach access — La Paz & Mazatlán pilot scoring"
output: html_document
date: "`r Sys.Date()`"
---

Pilot workflow for expert beach access scores in **La Paz** and **Mazatlán**:

1. Merge legacy expert CSVs into locality beach worksheets
2. Build scored-beach Leaflet maps
3. Aggregate length-weighted scores by locality and OHI marine region

Legacy sources live in `data/pilot/` (`LaPaz.csv`, `beach_access.csv`).

```{r setup, message=FALSE}
library(here)
library(dplyr)
library(sf)
library(purrr)
library(readr)

access_dir <- here("_pressures/access")
data_dir <- file.path(access_dir, "data")
pilot_dir <- file.path(data_dir, "pilot")
out_dir <- file.path(access_dir, "output", "pilot")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

source(file.path(access_dir, "access_utils.R"), local = TRUE)
paths <- default_external_paths(data_dir)

pilot_localities <- c("La Paz", "Mazatlán")
```

## 1. Merge legacy expert scores

Re-run after regenerating beach CSVs in `access_workflow.Rmd` so pilot scores are not lost.

```{r merge-legacy}
merge_city_expert <- function(locality_nm, expert_df) {
  out_path <- locality_beach_csv_path(locality_nm, data_dir)
  if (!file.exists(out_path)) {
    warning("Template not found: ", out_path)
    return(invisible(NULL))
  }
  template <- read_csv(out_path, show_col_types = FALSE, locale = locale(encoding = "UTF-8"))
  merged <- merge_expert_scores(template, expert_df)
  write_csv(merged, out_path)
  n_scored <- sum(!is.na(merged$equitable_access_score))
  message(locality_nm, ": ", n_scored, "/", nrow(merged), " beaches scored")
  invisible(merged)
}

lapaz_expert <- read_csv(
  file.path(pilot_dir, "LaPaz.csv"),
  show_col_types = FALSE,
  locale = locale(encoding = "UTF-8")
)
merge_city_expert("La Paz", lapaz_expert)

mazatlan_legacy <- read_csv(
  file.path(pilot_dir, "beach_access.csv"),
  show_col_types = FALSE,
  locale = locale(encoding = "UTF-8")
)
merge_city_expert("Mazatlán", legacy_mazatlan_to_expert(mazatlan_legacy))
```

## 2. Scored-beach maps

```{r score-maps}
for (nm in pilot_localities) {
  build_beach_access_map(nm, data_dir, access_dir, paths = paths)
}
```

Output: `output/pilot/{slug}_beach_access_map.html`

## 3. Aggregate scores

Length-weighted `equitable_access_score` by locality and by OHI marine region.

```{r aggregate}
localities <- load_access_localities(data_dir)
active <- sort(unique(localities$city))

all_beaches <- map_dfr(active, \(nm) {
  p <- locality_beach_csv_path(nm, data_dir)
  if (!file.exists(p)) return(NULL)
  read_csv(p, show_col_types = FALSE, locale = locale(encoding = "UTF-8"))
}) |>
  mutate(
    ID_Beach = as.integer(ID_Beach),
    equitable_access_score = suppressWarnings(as.numeric(equitable_access_score)),
    length_km = as.numeric(length_km)
  )

write_csv(all_beaches, file.path(out_dir, "beach_access_all_cities.csv"))

scored <- all_beaches |> filter(!is.na(equitable_access_score))

by_city <- scored |>
  group_by(city) |>
  summarise(
    n_beaches = n(),
    total_length_km = sum(length_km, na.rm = TRUE),
    mean_access_score = weighted.mean(equitable_access_score, length_km, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(city)

write_csv(by_city, file.path(out_dir, "beach_access_by_city.csv"))
by_city

ohi_path <- here("spatial/ohi_regions/goc_ohi_rgns.shp")
if (!file.exists(ohi_path)) {
  warning("OHI regions not found; skipping regional summary.")
} else if (nrow(scored) == 0) {
  warning("No scored beaches yet.")
} else {
  beach <- st_read(paths$beach, quiet = TRUE) |> filter(ID_Beach %in% scored$ID_Beach)
  beach_scored <- beach |>
    left_join(scored, by = "ID_Beach", suffix = c("", ".city"))
  for (col in intersect(expert_columns(), names(beach_scored))) {
    city_col <- paste0(col, ".city")
    if (city_col %in% names(beach_scored)) {
      beach_scored[[col]] <- coalesce(beach_scored[[city_col]], beach_scored[[col]])
      beach_scored[[city_col]] <- NULL
    }
  }
  ohi_rgns <- st_read(ohi_path, quiet = TRUE) |> filter(location == "marine")
  beach_rgn <- beach_scored |>
    st_transform(st_crs(ohi_rgns)) |>
    st_join(ohi_rgns |> select(rgn_id, ecorgn), join = st_intersects, left = TRUE) |>
    st_drop_geometry()

  by_region <- beach_rgn |>
    filter(!is.na(rgn_id)) |>
    group_by(rgn_id, ecorgn) |>
    summarise(
      n_beaches = n(),
      total_length_km = sum(length_km, na.rm = TRUE),
      mean_access_score = weighted.mean(equitable_access_score, length_km, na.rm = TRUE),
      .groups = "drop"
    ) |>
    arrange(rgn_id)

  write_csv(by_region, file.path(out_dir, "beach_access_by_ohi_region.csv"))
  by_region
}
```
