---
title: 'OHI GoC 2025: Livelihoods Number of Jobs'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# LIV: Number of Jobs from the Censos Economicos (CE)

For 2024, Ensenada needs to be dropped.  See difference between before the municipality San Quintín was made in 2020 and after.

```{r}
library(sf)

municipio_codigo_sf <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/land_admin_boundaries", "mex_admin2_joined_w_municipio_codigos.shp"))
mapview(municipio_codigo_sf) # before 2020

## AFTER 2020: <https://commons.wikimedia.org/wiki/File:Municipios_Baja_California.png>
```


## Save subset of job *number* metrics

Reminder:
* The number of jobs, *H000A*;
* The employer contributions to social security schemes *J300A*,
* Other social benefits *J400A*;
* Total revenue from the supply of goods and services (millions of pesos) *M000A*; and
* Total revenue (millions of pesos) *A800A*

The longer description of H000A: "Personnel employed by the company. This includes personnel hired directly by the company, whether permanent, temporary, or unpaid, unionized or non-unionized, who worked during the reference year for the economic unit, subject to its management and control, covering at least one-third of the working hours of the unit. Includes: personnel who worked outside the economic unit under its labor and legal control; pieceworkers; striking workers; persons on sick leave, vacation, or temporary leave. Excludes: pensioners and retirees; personnel on unlimited leave and personnel who worked exclusively for fees or commissions without receiving a base salary; as well as personnel from companies hired to provide services such as cleaning, gardening, or security, among others."

```{r}
ce_cleaned_totals_per_yr_municipio_activity <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/ce_int","ce_cleaned_totals_per_yr_municipio_activity.csv")) %>% 
  filter(!(municipio == "ENSENADA" & ano_censal == "2024"))

ce_job_totals <- ce_cleaned_totals_per_yr_municipio_activity %>% 
  select(-c("a800a","m000a", "j400a","j300a"))
```

Aggregating all the activities per municipio:
```{r}
total_jobs_per_municipio <- ce_job_totals %>% 
  group_by(ano_censal, municipio, rgn_id) %>% 
  summarize(jobs_per_mun = sum(h000a, na.rm = TRUE), # double check this.
            ue_per_mun = sum(ue, na.rm = TRUE)) # I do not think that UE (economic units) are a good way of tracking over time, because you may have increasing or decreasing jobs despite a stagnant number of UEs.  And there is very unlikely a relationship between number of jobs and number of UEs.  But, we could look into that in order to gapfill 2019 and 2024 which has gaps due to confidentiality...
```

Aggregating the number of jobs for each region:
```{r}
total_jobs_per_rgn <- total_jobs_per_municipio %>% 
  group_by(ano_censal, rgn_id) %>% 
  summarize(total_jobs = sum(jobs_per_mun, na.rm = TRUE),
            total_ue = sum(ue_per_mun, na.rm = TRUE))
```

Aggregating instead by region, keeping activities separate:
```{r}
total_jobs_per_rgn_and_activity <- ce_job_totals %>% 
  group_by(ano_censal, rgn_id, sector_marino_propuesto) %>% # not using codigo, notas_breve, because those are too specific
  summarize(jobs_per_act = sum(h000a, na.rm = TRUE), 
            ue_per_act = sum(ue, na.rm = TRUE))
```



## Bring in population

```{r}
pop_goc <- read_csv(here::here("/home/shares/ohi/OHI_GOC/spatial/population_municipal","population_goc_municipality_inegi_2000_2020.csv"))

## aggregate by region
agg_pop_rgnid <- pop_goc %>% 
  pivot_longer(cols = x2000:x2020,
               names_to = c("year"),
               values_to = "pop_count"
  ) %>% 
  group_by(rgn_id, year, unidad_medida) %>% 
  summarize(total_rgn_pop = sum(pop_count, na.rm = TRUE))
```


```{r}
library(zoo)

## in order for it to join with the job data, we need to interpolate and determine the population in years between.
year_adj_pop_rgnid <- agg_pop_rgnid %>% 
  separate_wider_delim(year, delim = "x", names = c("delete","ano_censal")) %>% # clean up year
  select(-delete) %>% 
  mutate(year_interpolated = as.numeric(ano_censal)) %>% 
  complete(year_interpolated = 2000:2024) # expand to interpolate

year_adj_pop_rgnid_interpolated <- year_adj_pop_rgnid %>%
  arrange(rgn_id, year_interpolated) %>%
  group_by(rgn_id) %>% 
  mutate(
    population_interpolated = na.approx(
      total_rgn_pop,
      x = year_interpolated,
      rule = 2,
      na.rm = FALSE
    ),
    is_gapfilled = ifelse(is.na(ano_censal), 1, 0)
  ) %>%
  fill(ano_censal, .direction = "downup") %>% 
  rename(pop_ano_censal_original = ano_censal) %>% # rename so that when it is joined, we know which two censuses are being interpolated from.
  fill(total_rgn_pop, .direction = "downup") %>% # so we can see what the original population value was
  fill(unidad_medida, .direction = "downup") %>%
  select(rgn_id, pop_ano_censal_original, population_original = total_rgn_pop, year_interpolated, population_interpolated, unidad_medida, is_gapfilled) %>% 
  ungroup()

## note that the interpolated population amount from 2020-2024 is the same for each region.  It does not project the growth before it happens.

## save this
# write_csv(year_adj_pop_rgnid_interpolated, here::here("/home/shares/ohi/OHI_GOC/spatial/population_municipal","interpolated_population_per_rgn_coastalmunicipios_2000_2020.csv"))
```

Now for each year, make a column of the population.

```{r}
join_jobs_pop <- inner_join(total_jobs_per_rgn, year_adj_pop_rgnid_interpolated, by = c("ano_censal" = "year_interpolated", "rgn_id")) %>% 
  select(rgn_id, ce_ano_censal = ano_censal, total_jobs, total_population_interpolated_for_ce_year = population_interpolated, population_original, pop_ano_censal_original, unidad_medida, total_ue) # reordering and dropping total_rgn_pop because none of the years of the job census align with the population census.  So that column is currently NA. And the gapfill column is all 1.
```


## Jobs scaled to population

```{r}
jobs_scaled_to_population <- join_jobs_pop %>% 
  mutate(marine_jobs_per_local_pop = total_jobs/total_population_interpolated_for_ce_year,
         marine_ue_per_local_pop = total_ue/total_population_interpolated_for_ce_year) %>% 
  ungroup()
```

### Plotlys

```{r}
library(plotly)

## plot marine jobs per local pop
num_jobs_plotly <- plotly::plot_ly(
  jobs_scaled_to_population, 
  x = ~ce_ano_censal, 
  y = ~marine_jobs_per_local_pop, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "Number of marine-related jobs per region, SCALED to regional population",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "marine_jobs_per_local_pop")
  ) 
num_jobs_plotly

## plot marine ue's per local pop
num_ue_plotly <- plotly::plot_ly(
  jobs_scaled_to_population, 
  x = ~ce_ano_censal, 
  y = ~marine_ue_per_local_pop, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "Number of marine-related Economic Units (UEs) per region, SCALED to regional population",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "marine_ue_per_local_pop")
  ) 
num_ue_plotly
```

## Reference point (average value 2004-2014)

```{r}
historical_jobs_scaled_to_population <- jobs_scaled_to_population %>% 
  filter(ce_ano_censal %in% c("2004","2009","2014")) %>% 
  group_by(rgn_id) %>%  
  summarize(avg_value_2004_2014 = mean(marine_jobs_per_local_pop))
```

## Current status

First, join back the reference point, then divide the current value of marine jobs/pop by the historical average from 2004 - 2014 as a reference.

If it is equal to the historical average, or above, the region will receive the current status of 100.  Otherwise, it will reflect the reduction.

```{r}
ce_jobs_current_status <- jobs_scaled_to_population %>% 
  select(rgn_id, ce_ano_censal, marine_jobs_per_local_pop) %>% 
  left_join(historical_jobs_scaled_to_population) %>% 
  mutate(current_status_raw = marine_jobs_per_local_pop/avg_value_2004_2014 * 100,
         current_status = ifelse(current_status_raw >= 100, 100, current_status_raw))
```

### Plotlys 

```{r}
## plot num jobs current status
num_jobs_cs_plotly <- plotly::plot_ly(
  ce_jobs_current_status, 
  x = ~ce_ano_censal, 
  y = ~current_status, 
  color = ~rgn_id, 
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "Current status of marine-related jobs per region, SCALED to regional population, reference of average value 2004-2014",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "marine_jobs_per_local_pop / avg_value_2004_2014")
  ) 
num_jobs_cs_plotly
```


