---
title: 'OHI GoC 2025: Livelihoods Quality of Jobs'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# LIV: Quality of Jobs from the Censos Economicos (CE)



## Save subset of job *quality* metrics

Reminder:
* The number of jobs, *H000A*;
* The employer contributions to social security schemes *J300A*,
* Other social benefits *J400A*;
* Total revenue from the supply of goods and services (millions of pesos) *M000A*; and
* Total revenue (millions of pesos) *A800A*

Specifically: 

*J300A* is "Contribuciones patronales a regímenes de seguridad social. Son todas las aportaciones monetarias que la unidad económica cubrió con sus recursos a instituciones de seguridad social en beneficio de los trabajadores remunerados."

*J300A* is “Employer contributions to social security schemes. These are all monetary contributions that the economic unit covered with its resources to social security institutions for the benefit of paid workers.”

---

*J400A* is "Otras prestaciones sociales. Son los pagos que la unidad económica realizó a instituciones privadas en beneficio de sus trabajadores o que otorgó directamente en especie al personal remunerado, en complemento o adición a los sueldos y salarios, tales como servicios médicos privados, despensas, primas de seguros, servicios educativos, ayudas para estudio y guarderías. Excluye: las contribuciones patronales a regímenes de seguridad social, compra de equipo, uniformes y ropa de trabajo; costos de capacitación; primas vacacionales; erogaciones para actividades deportivas y recreativas; gastos por concepto de pasajes, viáticos y alimentación; además de todos aquellos gastos reembolsables al trabajador."

*J400A* is "Other social benefits. These are payments that the economic unit made to private institutions for the benefit of its workers or that it granted directly in kind to paid personnel, in addition to or as a supplement to wages and salaries, such as private medical services, food allowances, insurance premiums, educational services, study grants, and childcare. Excludes: employer contributions to social security schemes, purchase of equipment, uniforms, and work clothes; training costs; vacation bonuses; expenditures for sports and recreational activities; expenses for travel, per diem, and meals; in addition to all expenses reimbursable to the worker."

Because of this, I will be summing J300A and J400A for job quality.

```{r}
ce_cleaned_totals_per_yr_municipio_activity <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/ce_int","ce_cleaned_totals_per_yr_municipio_activity.csv")) %>% 
  filter(!(municipio == "ENSENADA" & ano_censal == "2024"))

ce_quality_totals <- ce_cleaned_totals_per_yr_municipio_activity %>% 
  select(-c("a800a","m000a"))  # leave in number of jobs (h000a) so that we can scale to the number of jobs.
```


## Adjust for inflation

This will be done using the package `priceR`.  Documentation is located here: <https://cran.r-project.org/web/packages/priceR/readme/README.html> 

"adjust_for_inflation() automatically converts between nominal and real dollars, or in/deflates prices from one year’s prices to another’s.  It works for 304 countries / areas (see them with all by running show_countries())."

```{r}
library(priceR)

ce_quality_totals_adj <- ce_quality_totals %>%
  mutate(sum_j300a_j400a_milpesos = j300a + j400a) %>% 
  mutate(sum_j300a_j400a_pesos = sum_j300a_j400a_milpesos * 1e6) %>% 
  mutate(
    quality_real_pesos_2024 = adjust_for_inflation(
      price = sum_j300a_j400a_pesos,
      from_date = ano_censal,
      country = "Mexico",
      to_date = 2024
      )
  ) %>% 
  select(ano_censal, rgn_id, estado, e03, municipio, e04, codigo, sector_marino_propuesto, notas_breve, sum_j300a_j400a_pesos, quality_real_pesos_2024, h000a, ue)
```


<!-- ## Aggregate by region and activity -->

<!-- ### By region -->

<!-- ```{r} -->
<!-- total_quality_byrgn <- ce_revenue_totals_adj %>%  -->
<!--   group_by(ano_censal, rgn_id) %>%  -->
<!--   summarize(total_rev_adj = sum(revenue_real_pesos_2024, na.rm = TRUE), -->
<!--             total_jobs = sum(h000a, na.rm = TRUE), -->
<!--             total_ue = sum(ue, na.rm = TRUE)) -->
<!-- ``` -->


<!-- ### By sector -->

<!-- ```{r} -->
<!-- total_revenue_byrgn_and_activity <- ce_revenue_totals_adj %>%  -->
<!--   group_by(ano_censal, rgn_id, sector_marino_propuesto) %>% # not using codigo, notas_breve, because those are too specific -->
<!--   summarize(total_rev_adj = sum(revenue_real_pesos_2024, na.rm = TRUE), -->
<!--             total_jobs = sum(h000a, na.rm = TRUE), -->
<!--             total_ue = sum(ue, na.rm = TRUE)) -->
<!-- ``` -->


<!-- ## Scale to the number of jobs -->

<!-- By region: -->
<!-- ```{r} -->
<!-- ce_revenue_adj_perjob <- total_revenue_byrgn %>%  -->
<!--   mutate(rev_per_job = total_rev_adj/total_jobs) -->
<!-- ``` -->

<!-- By region and sector: -->
<!-- ```{r} -->
<!-- ce_revenue_sector_adj_perjob <- total_revenue_byrgn_and_activity %>%  -->
<!--   mutate(rev_per_job = total_rev_adj/total_jobs) -->
<!-- ``` -->


<!-- ## Reference point (average value 2004-2014) -->

<!-- By region: -->
<!-- ```{r} -->
<!-- historical_rev_scaled_to_jobs <- ce_revenue_adj_perjob %>%  -->
<!--   filter(ano_censal %in% c("2004","2009","2014")) %>%  -->
<!--   group_by(rgn_id) %>%   -->
<!--   summarize(avg_value_2004_2014 = mean(rev_per_job)) -->
<!-- ``` -->

<!-- By region and sector: -->
<!-- ```{r} -->
<!-- historical_rev_sector_scaled_to_jobs <- ce_revenue_sector_adj_perjob %>%  -->
<!--   filter(ano_censal %in% c("2004","2009","2014")) %>%  -->
<!--   group_by(rgn_id, sector_marino_propuesto) %>%   -->
<!--   summarize(avg_value_2004_2014 = mean(rev_per_job)) -->
<!-- ``` -->

<!-- ## Current status -->

<!-- First, join back the reference point, then divide the current value of marine jobs/pop by the historical average from 2004 - 2014 as a reference. -->

<!-- If it is equal to the historical average, or above, the region will receive the current status of 100.  Otherwise, it will reflect the reduction. -->

<!-- By region: -->
<!-- ```{r} -->
<!-- ce_rev_current_status <- ce_revenue_adj_perjob %>%  -->
<!--   select(rgn_id, ano_censal, rev_per_job) %>%  -->
<!--   left_join(historical_rev_scaled_to_jobs) %>%  -->
<!--   mutate(current_status_raw = rev_per_job/avg_value_2004_2014 * 100, -->
<!--          current_status = ifelse(current_status_raw >= 100, 100, current_status_raw)) %>%  -->
<!--   ungroup() -->
<!-- ``` -->

<!-- By region and sector: -->
<!-- ```{r} -->
<!-- ce_rev_sector_current_status <- ce_revenue_sector_adj_perjob %>%  -->
<!--   select(rgn_id, ano_censal, sector_marino_propuesto, rev_per_job) %>%  -->
<!--   left_join(historical_rev_sector_scaled_to_jobs) %>%  -->
<!--   mutate(current_status_raw = rev_per_job/avg_value_2004_2014 * 100, -->
<!--          current_status = ifelse(current_status_raw >= 100, 100, current_status_raw)) %>%  -->
<!--   ungroup() -->

<!-- ## now, average into one score per region: -->
<!-- ce_rev_sector_current_status_final <- ce_rev_sector_current_status %>%  -->
<!--   group_by(ano_censal, rgn_id) %>%  -->
<!--   summarize(avg_cs = mean(current_status, na.rm = TRUE)) %>%  -->
<!--   ungroup -->
<!-- ``` -->

<!-- ### Plotlys  -->

<!-- #### By region -->

<!-- ```{r} -->
<!-- library(plotly) -->

<!-- ## Plot revenue current status by region -->
<!-- rev_cs_plotly <- plotly::plot_ly( -->
<!--   ce_rev_current_status, -->
<!--   x = ~ano_censal, -->
<!--   y = ~current_status, -->
<!--   color = ~factor(rgn_id), -->
<!--   type = "scatter", -->
<!--   mode = "lines+markers" -->
<!-- ) %>% -->
<!--   layout( -->
<!--     title = "Current Status of Marine Revenue per Job by Region (Constant 2024 Pesos)", -->
<!--     xaxis = list(title = "Census Year"), -->
<!--     yaxis = list( -->
<!--       title = "Revenue per Job / 2004-2014 Average (%)", -->
<!--       range = c(0, 110) -->
<!--     ), -->
<!--     legend = list(title = list(text = "Region")) -->
<!--   ) -->

<!-- rev_cs_plotly -->

<!-- ``` -->


<!-- #### By region and sector -->

<!-- ```{r} -->
<!-- library(plotly) -->

<!-- ## Plot revenue current status by region -->
<!-- rev_cs_sector_plotly <- plotly::plot_ly( -->
<!--   ce_rev_sector_current_status_final, -->
<!--   x = ~ano_censal, -->
<!--   y = ~avg_cs, -->
<!--   color = ~factor(rgn_id), -->
<!--   type = "scatter", -->
<!--   mode = "lines+markers" -->
<!-- ) %>% -->
<!--   layout( -->
<!--     title = "Current Status of Marine Revenue per Job by Region and Averaged Across Sector (Constant 2024 Pesos)", -->
<!--     xaxis = list(title = "Census Year"), -->
<!--     yaxis = list( -->
<!--       title = "Revenue per Job / 2004-2014 Average (%)", -->
<!--       range = c(0, 110) -->
<!--     ), -->
<!--     legend = list(title = list(text = "Region")) -->
<!--   ) -->

<!-- rev_cs_sector_plotly -->
<!-- ``` -->












