---
title: 'OHI GoC 2025: Livelihoods Number of Jobs'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# LIV: Number of Jobs from the Censos Economicos (CE)

This script will be used to obtain current status scores for the Livelihoods (LIV) subgoal, for the "number of marine-related jobs" component.  Note that the reference point determined by our expert working groups was that there are stable and more jobs over time. 

*Outline of script:*

1. Bring in cleaned data from script 1.
2. Apply coastal GoC adjustment, which was made using `~spatial/population/population_by_municipality_1990_2020.qmd`.
3. Population census data from INEGI is taken every 5 years.  Since we have data for 2020, 2015, 2010, 2005, and 2000, use `zoo` and interpolate between the years to have population estimates for each year that CE census data was collected (2003-2023).  The interpolated population estimates for each municipality are here: `~/home/shares/ohi/OHI_GOC/spatial/population_municipal/interpolated_population_per_rgn_coastalmunicipios_2000_2020.csv`
4. Scale the marine-related jobs to the coastal population. 
5. Determine the average jobs-per-person value from the years 2003-2013 as a reference point. If the current value is equal to the historical average, or above, the region will receive the current status of 100.  Otherwise, it will reflect the reduction.
6. Finally, use current/baseline to calculate the LIV_number_jobs current status scores from 2003-2023 for all 9 Goc OHI Regions. 
7. Save to Mazu.
8. Explore using `plotly`.


```{r}
## for plotting:
rgn_colors <- c(
  "Region_1" = "#C03729",
  "Region_2" = "#E68C7C",
  "Region_3" = "#FC8F24",
  "Region_4" = "#dcaa38",
  "Region_5" = "#cbd2a0",
  "Region_6" = "#98a25a",
  "Region_7" = "#747428",
  "Region_8" = "#627391",
  "Region_9" = "#8d61a9"
)

region_levels <- paste0("Region_", 1:9)
```

To see the changes made in Baja California in 2020 and 2021: 
```{r}
library(sf)

municipio_codigo_sf <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/land_admin_boundaries", "mex_admin2_joined_w_municipio_codigos.shp"))
mapview(municipio_codigo_sf) # before 2020

## AFTER 2020: <https://commons.wikimedia.org/wiki/File:Municipios_Baja_California.png>
```

This was already adjusted for in the municipality_to_rgnid.csv, but in the data from the year 2023, we need to make sure that Region_1 is only represented by the municipality of San Felipe, NOT Mexicali. 


## Save subset of job *number* metrics

Reminder of the metrics we currently have in `ce_cleaned_totals_per_yr_municipio_activity.csv`:

* The number of jobs, *H000A*;
* The employer contributions to social security schemes *J300A*,
* Other social benefits *J400A*;
* Total revenue from the supply of goods and services (millions of pesos) *M000A*; and
* Total revenue (millions of pesos) *A800A*

Of which we only need  *H000A* for this script.

The longer description of H000A: "Personnel employed by the company. This includes personnel hired directly by the company, whether permanent, temporary, or unpaid, unionized or non-unionized, who worked during the reference year for the economic unit, subject to its management and control, covering at least one-third of the working hours of the unit. Includes: personnel who worked outside the economic unit under its labor and legal control; pieceworkers; striking workers; persons on sick leave, vacation, or temporary leave. Excludes: pensioners and retirees; personnel on unlimited leave and personnel who worked exclusively for fees or commissions without receiving a base salary; as well as personnel from companies hired to provide services such as cleaning, gardening, or security, among others."

Filter for the job number metric:
```{r}
ce_cleaned_totals_per_yr_municipio_activity <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/ce_int","ce_cleaned_totals_per_yr_municipio_activity.csv")) %>% 
  filter(!(municipio == "MEXICALI" & ano_censal == "2023"))

ce_job_totals <- ce_cleaned_totals_per_yr_municipio_activity %>% 
  dplyr::select(-c("a800a","m000a", "j400a","j300a"))
```

Aggregating all the activities per municipio:
```{r}
total_jobs_per_municipio <- ce_job_totals %>% 
  group_by(ano_censal, municipio, rgn_id) %>% 
  summarize(jobs_per_mun = sum(h000a, na.rm = TRUE), # double check this.
            ue_per_mun = sum(ue, na.rm = TRUE)) # I do not think that UE (economic units) are a good way of tracking over time, because you may have increasing or decreasing jobs despite a stagnant number of UEs.  And there is very unlikely a relationship between number of jobs and number of UEs.  But, we could look into that in order to gapfill 2019 and 2024 which has gaps due to confidentiality...
```

Aggregating instead by region and general sector:
```{r}
total_jobs_per_rgn_and_activity <- ce_job_totals %>% 
  group_by(ano_censal, rgn_id, general_sector, specific_sector) %>% # not using codigo, notas_breve, because those are too specific
  summarize(jobs_per_act = sum(h000a, na.rm = TRUE), 
            ue_per_act = sum(ue, na.rm = TRUE)) %>% 
  ungroup()
```


### Add in adjustment for municipalities that cover both coastal and pacific sides, and sectors that are not strictly coastal

Region_3 and Region_4 have municipalities that span both the gulf and pacific sides of the peninsula, so we need to adjust to ensure those represent the number of jobs in the Gulf.  Also, some sectors such as tourism, are not explicitly marine-related.  So we only consider the "coastal" jobs in our analysis, assuming that if the tourism jobs are coastal, then they have some tie to the Gulf.  

Many people travel to see the Gulf specifically, and accommodations, food, drink, etc. support those visits.  And those visits not only support the local economy, but also create connections between people and the Gulf.

```{r}
inland_50_population_2020 <- read_csv(here::here("/home/shares/ohi/OHI_GOC/spatial/population_inland_50km","pop2020_by_region_vector_inland_50.csv"))

coastal_population_multipliers <- agg_pop_rgnid %>% 
  filter(ano_censal %in% 2020) %>% 
  full_join(inland_50_population_2020) %>% 
  mutate(prop_coastal = pop_sum_2020_inland_50 / total_rgn_pop,
         coastal_pop_adjustment = ifelse(prop_coastal > 1, 1, prop_coastal)) %>% 
  ungroup() %>% 
  dplyr::select(rgn_id, coastal_pop_adjustment) %>% 
  filter(!is.na(rgn_id))

### save for future coastal adjustments
# write_csv(coastal_population_multipliers, here::here("/home/shares/ohi/OHI_GOC/goal_prep/le","coastal_adjustment_multiplier_2020.csv"))
```

```{r}
coastal_total_jobs_per_rgn_and_activity <- total_jobs_per_rgn_and_activity %>% 
  full_join(coastal_population_multipliers, by = "rgn_id") %>% 
  mutate(
    
    # fix San Felipe
    coastal_pop_adjustment = ifelse(rgn_id %in% "Region_1" & ano_censal %in% "2023", 1, coastal_pop_adjustment),
    
    # jobs
    first_adj_coastal_jobs_per_act = ifelse(rgn_id %in% c("Region_3","Region_4"), jobs_per_act * coastal_pop_adjustment, jobs_per_act),
    coastal_jobs_per_act = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), jobs_per_act * coastal_pop_adjustment, first_adj_coastal_jobs_per_act),
    
    # ue
    first_adj_coastal_ue_per_act = ifelse(rgn_id %in% c("Region_3","Region_4"), ue_per_act * coastal_pop_adjustment, ue_per_act),
    coastal_ue_per_act = ifelse(specific_sector %in% c("coastal_recreation","travel_planners","accommodations_food_drink","environmental_research_consulting"), ue_per_act * coastal_pop_adjustment, first_adj_coastal_ue_per_act)) %>%  # if in any of these sectors, adjust for 50km inland
  
  dplyr::select(-c(jobs_per_act, ue_per_act, first_adj_coastal_jobs_per_act, first_adj_coastal_ue_per_act))
```

Aggregating the number of jobs for each region:
```{r}
total_jobs_per_rgn <- coastal_total_jobs_per_rgn_and_activity %>% 
  group_by(ano_censal, rgn_id) %>% 
  summarize(total_jobs = sum(coastal_jobs_per_act, na.rm = TRUE),
            total_ue = sum(coastal_ue_per_act, na.rm = TRUE)) %>% 
  mutate(total_jobs = round(total_jobs),
         total_ue = round(total_ue))
```


## Bring in population

```{r}
pop_goc <- read_csv(here::here("/home/shares/ohi/OHI_GOC/spatial/population_municipal","population_goc_municipality_inegi_2000_2020.csv"))

## aggregate by region
agg_pop_rgnid <- pop_goc %>% 
  group_by(rgn_id, ano_censal, unidad_medida) %>% 
  summarize(total_rgn_pop = sum(pop_count, na.rm = TRUE)) %>% 
  ungroup()
```


### Interpolate municipal population for all years

```{r}
library(zoo)

## in order for it to join with the job data, we need to interpolate and determine the population in years between.
year_adj_pop_rgnid <- agg_pop_rgnid %>% 
  mutate(year_interpolated = as.numeric(ano_censal)) %>%
  group_by(rgn_id) %>% 
  complete(year_interpolated = 2000:2023) # expand to interpolate

year_adj_pop_rgnid_interpolated <- year_adj_pop_rgnid %>%
  arrange(rgn_id, year_interpolated) %>%
  group_by(rgn_id) %>% 
  mutate(
    population_interpolated = na.approx(
      total_rgn_pop,
      x = year_interpolated,
      rule = 2,
      na.rm = FALSE
    ),
    is_gapfilled = ifelse(is.na(ano_censal), 1, 0)
  ) %>%
  fill(ano_censal, .direction = "downup") %>% 
  rename(pop_ano_censal_original = ano_censal) %>% # rename so that when it is joined, we know which two censuses are being interpolated from.
  fill(total_rgn_pop, .direction = "downup") %>% # so we can see what the original population value was
  fill(unidad_medida, .direction = "downup") %>%
  dplyr::select(rgn_id, pop_ano_censal_original, population_original = total_rgn_pop, year_interpolated, population_interpolated, unidad_medida, is_gapfilled) %>% 
  ungroup()

## note that the interpolated population amount from 2020-2023 is the same for each region.  It does not project the growth before it happens.

## save this
# write_csv(year_adj_pop_rgnid_interpolated, here::here("/home/shares/ohi/OHI_GOC/spatial/population_municipal","interpolated_population_per_rgn_coastalmunicipios_2000_2020.csv"))
```


Now for each year, make a column of the population.

```{r}
join_jobs_pop <- inner_join(total_jobs_per_rgn, year_adj_pop_rgnid_interpolated, by = c("ano_censal" = "year_interpolated", "rgn_id")) %>% 
  dplyr::select(rgn_id, ce_ano_censal = ano_censal, total_jobs, total_population_interpolated_for_ce_year = population_interpolated, population_original, pop_ano_censal_original, unidad_medida, total_ue) # reordering and dropping total_rgn_pop because none of the years of the job census align with the population census.  So that column is currently NA. And the gapfill column is all 1.

join_jobs_pop_coastal_adj <- join_jobs_pop %>% 
  left_join(coastal_population_multipliers) %>% 
  mutate(coastal_population_interpolated_for_ce_year = total_population_interpolated_for_ce_year * coastal_pop_adjustment)
```


## Jobs scaled to population

```{r}
jobs_scaled_to_population <- join_jobs_pop_coastal_adj %>% 
  mutate(marine_jobs_per_local_pop = total_jobs/coastal_population_interpolated_for_ce_year,
         marine_ue_per_local_pop = total_ue/coastal_population_interpolated_for_ce_year) %>% 
  ungroup()
```



### Exploration so far: plotlys

```{r}
sectors_of_interest <- c(
  "fishing",
  "mariculture",
  "seafood_processing_trade",
  "port_dock_construction_management",
  "boat_building",
  "tidal_energy",
  "oil_gas_maritime_transport",
  "cargo_goods_maritime_transport",
  "marine_tourist_transport",
  "coastal_recreation",
  "travel_planners",
  "accommodations_food_drink",
  "environmental_research_consulting"
)

# filter the summarized data for only these sectors
ce_by_sector_year_filtered <- coastal_total_jobs_per_rgn_and_activity %>%
  left_join(inland_50_population_2020) %>% 
  mutate(jobs_per_pop = coastal_jobs_per_act / pop_sum_2020_inland_50) %>% 
  group_by(specific_sector, ano_censal, rgn_id) %>%
  summarise(
    total_jobs = sum(jobs_per_pop, na.rm = TRUE),
    n_ue = sum(coastal_ue_per_act, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(specific_sector %in% sectors_of_interest)

library(plotly)
library(dplyr)
library(scales)

# Lists to store plots
plot_jobs_list  <- list()

for (sector in sectors_of_interest) {
  df_sector <- ce_by_sector_year_filtered %>%
    filter(specific_sector == sector)

  if (nrow(df_sector) == 0) next  # skip if sector is missing

  # plot 1: jobs for this sector
  p_jobs <- plot_ly(
    df_sector,
    x = ~ano_censal,
    y = ~total_jobs,
    color = ~rgn_id,
    colors = rgn_colors,
    type = "scatter",
    mode = "lines+markers",
    text = ~paste("Jobs:", comma(total_jobs), "<br>UE:", n_ue),
    hovertemplate = "%{text}<extra></extra>"
  ) %>%
    layout(
      title = paste("JOBS per coastal pop:", sector),
      xaxis = list(title = "Census Year"),
      yaxis = list(title = "Total Jobs (H000A) / coastal population")
    )

  # store plots
  plot_jobs_list[[sector]]   <- p_jobs
}

plot_jobs_list[["fishing"]]
plot_jobs_list[["mariculture"]]
plot_jobs_list[["seafood_processing_trade"]]
plot_jobs_list[["port_dock_construction_management"]]
plot_jobs_list[["boat_building"]]
plot_jobs_list[["oil_gas_maritime_transport"]]
plot_jobs_list[["cargo_goods_maritime_transport"]]
plot_jobs_list[["marine_tourist_transport"]]
plot_jobs_list[["coastal_recreation"]]
plot_jobs_list[["travel_planners"]]
plot_jobs_list[["accommodations_food_drink"]]
plot_jobs_list[["environmental_research_consulting"]]

```


```{r}
library(plotly)

## plot marine jobs per local pop
num_jobs_plotly <- plotly::plot_ly(
  jobs_scaled_to_population, 
  x = ~ce_ano_censal, 
  y = ~marine_jobs_per_local_pop, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "Number of marine-related jobs per region, SCALED to regional coastal population",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "marine_jobs_per_local_pop")
  ) 
num_jobs_plotly

## plot marine ue's per local pop
num_ue_plotly <- plotly::plot_ly(
  jobs_scaled_to_population, 
  x = ~ce_ano_censal, 
  y = ~marine_ue_per_local_pop, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "Number of marine-related Economic Units (UEs) per region, SCALED to coastal regional population",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "marine_ue_per_local_pop")
  ) 
num_ue_plotly
```


## Reference point (average value 2004-2014)

```{r}
historical_jobs_scaled_to_population <- jobs_scaled_to_population %>% 
  filter(ce_ano_censal %in% c("2003","2008","2013")) %>% 
  group_by(rgn_id) %>%  
  summarize(avg_value_2003_2013 = mean(marine_jobs_per_local_pop))
```


## Current status

First, join back the reference point, then divide the current value of marine jobs/pop by the historical average from 2004-2014 as a reference.

If it is equal to the historical average, or above, the region will receive the current status of 100.  Otherwise, it will reflect the reduction.

```{r}
ce_jobs_current_status <- jobs_scaled_to_population %>% 
  dplyr::select(rgn_id, ce_ano_censal, marine_jobs_per_local_pop) %>% 
  left_join(historical_jobs_scaled_to_population) %>% 
  mutate(current_status_raw = marine_jobs_per_local_pop/avg_value_2003_2013 * 100,
         current_status = ifelse(current_status_raw >= 100, 100, current_status_raw))
```

### Save to Mazu

```{r}
write_csv(ce_jobs_current_status, here::here("/home/shares/ohi/OHI_GOC/goal_prep/le/liv","ce_jobs_current_status.csv"))
```

### Plotlys 

```{r}
## plot num jobs current status -- NOT CAPPED
num_jobs_cs_plotly_noCAP <- plotly::plot_ly(
  ce_jobs_current_status, 
  x = ~ce_ano_censal, 
  y = ~current_status_raw, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "RAW CURRENT STATUS of marine-related jobs per region, SCALED to regional coastal population, reference of average value 2003-2013",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "marine_jobs_per_local_pop / avg_value_2003_2013")
  ) %>%
  add_lines(
    x = range(ce_jobs_current_status$ce_ano_censal), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
num_jobs_cs_plotly_noCAP
```


```{r}
## plot num jobs current status
num_jobs_cs_plotly <- plotly::plot_ly(
  ce_jobs_current_status, 
  x = ~ce_ano_censal, 
  y = ~current_status, 
  color = ~rgn_id, 
  colors = rgn_colors,
  type = "scatter", 
  mode = "lines"
) %>% 
  layout(
    title = "CURRENT STATUS of marine-related jobs per region, SCALED to regional coastal population, reference of average value 2003-2013",
    xaxis = list(title = "Año // Year"),
    yaxis = list(title = "marine_jobs_per_local_pop / avg_value_2003_2013")
  )  %>%
  add_lines(
    x = range(ce_jobs_current_status$ce_ano_censal), 
    y = c(100, 100),
    line = list(color = "black", width = 2, dash = "dash"),
    name = "Meta // Goal (100/100)",
    inherit = FALSE
  )
num_jobs_cs_plotly
```












