################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```


# OLDER RESEARCH (December 2025 - February 2026)  ------------------------------

# Load in MAR dataframe (data as of Dec 09 2025)
```{r Load MAR dataframe}
# Load file, only if needed:
Mariculture_Landings<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/mar/v2025/int/GOC_MAR.landings_CLEAN_as.of.Feb242026.rds"))
# 130432 obs., 50 vars.
```

```{r Investigate the mariculture flags}
table(Mariculture_Landings$common_name_clean)
table(Mariculture_Landings$aquaculture_flag)
table(Mariculture_Landings$acuaculture_production)
table(Mariculture_Landings$fleet) # lots of NAs

subset<- Mariculture_Landings %>% filter(acuaculture_production== c("NO", "no"))
# 7822 obs.

unique(subset$common_name_clean)
# NOTE TO SELF: Many types of ostion flagged in this subset as "NOT" aquaculture production, which is at odds with what we've been told...

subset<- Mariculture_Landings %>% filter(common_name_clean== c("tiburon", "tiburon azul", "tiburon tunero", "tiburon sedoso")) 
# NOTE TO SELF: Also weird, several shark entries flagged as aquaculture, even though we know this all but impossible.
```

```{r Top Mariculture Sp.}
table(Mariculture_Landings$common_name_clean)
table(Mariculture_Landings$aquaculture_flag)

Mariculture_Landings_Summary<- Mariculture_Landings %>% 
  filter(year >= 2014, year <= 2024) %>%
  select(state, recorded_taxa_group, recorded_taxa_name_clean, weight_kg_corrected, value) %>%
  group_by(state, recorded_taxa_group, recorded_taxa_name_clean) %>%
  summarize(total_catch_2014.2024= sum(weight_kg_corrected, na.rm = TRUE),
            total_value_2014.2024= sum(value, na.rm = TRUE))

unique(Mariculture_Landings_Summary$recorded_taxa_name_clean)

Top_MaricultureSp <- Mariculture_Landings_Summary %>%
  filter(total_catch_2014.2024 >= 10000) %>%
  pull(recorded_taxa_name_clean) %>%
  unique()

Top_MaricultureSp # NOTE TO SELF: 33 top species. Some of them are fairly general or could be freshwater.

Top_MaricultureSp_Summary<- Mariculture_Landings %>% 
  filter(year >= 2014, year <= 2024) %>%
  filter(recorded_taxa_name_clean %in% Top_MaricultureSp) %>%
  select(recorded_taxa_group, recorded_taxa_name_clean, weight_kg_corrected, value) %>%
  group_by(recorded_taxa_group, recorded_taxa_name_clean) %>%
  summarize(total_catch_2014.2024= sum(weight_kg_corrected, na.rm = TRUE),
            total_value_2014.2024= sum(value, na.rm = TRUE))
```

```{r Save Landings_GoC_TopMariculture_Summary to find MAR sustainability scores}
write.csv(Top_MaricultureSp_Summary, "/home/ferrer/ohi_goc_prep/FP/MAR_research/Mariculture_TopCatches_2014.2024_Feb242026.csv", row.names = FALSE)
```

```{r Mariculture Key}
# UNDER CONSTRUCTION (I did this manually, Feb 24 2026)
Mariculture_GroupKey<- read.csv("~/ohi_goc_prep/FP/MAR_research/Mariculture_GroupKey_UNDERCONSTRUCTION_Feb242026.csv")
```

```{r Merge and exclude}
Mariculture_Landings <- Mariculture_Landings %>%
  left_join(Mariculture_GroupKey,
            by = c("recorded_taxa_group", "recorded_taxa_name_clean"))

Mariculture_Landings<- Mariculture_Landings %>% filter(!MAR_Key_Manual=="Exclude")

Mariculture_Landings_Summary<- Mariculture_Landings %>% 
  filter(year >= 2014, year <= 2024) %>%
  select(year, MAR_Key_Manual, weight_kg_corrected, value) %>%
  group_by(MAR_Key_Manual, year) %>%
  summarize(annual_catch= sum(weight_kg_corrected, na.rm = TRUE),
            annual_value= sum(value, na.rm = TRUE))
```

```{r PLOT}
# Plot annual catch over time
ggplot(Mariculture_Landings_Summary, 
       aes(x = year, y = log10(annual_catch), color = MAR_Key_Manual, group = MAR_Key_Manual)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::comma) +  # Adds commas to y-axis
  labs(
    title = "Annual Catch Over Time by MAR_Key_Manual",
    x = "Year",
    y = "Annual Catch (tons)",
    color = "Mariculture Category"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    text = element_text(size = 12)
  )
```

# RECENT RESEARCH (late March, early April 2026)  ------------------------------

# Generate MARICULTURE List for the GOC
Curation is under construction Apr 29 2026

```{r Load GOC_Landings if needed}
GOC_Landings<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Landings_SectorSorted_2000.present_EMF.Apr032026.rds"))
# 3312602 obs., 62 vars.

GOC_Landings_sector<- GOC_Landings_sector
rm(GOC_Landings_marine)
```

```{r Filter out the MARICULTURE sector}
# Let's assume that the records are most accurate since 2018.

table(GOC_Landings$sector_estimate_draft)

Mariculture_Sectors<- c("marine: aquaculture")

GOC_MaricultureProduction<- GOC_Landings %>% 
  filter(sector_estimate_draft %in% Mariculture_Sectors)
# 105932 obs., 62 vars.

range(GOC_MaricultureProduction$year)
```

```{r Look into this first-draft list before proceeding}
# Taking the information I've gleaned here to refine my sector estimates above.
table(GOC_MaricultureProduction$estimated_genus_name)
table(GOC_MaricultureProduction$Major_Functional_Group)
table(GOC_MaricultureProduction$estimated_genus_name)
```

```{r Generate a table / list of common n > 10 records aquaculture species since 2018}
Mex.Pacific_MajorMariculture_summary <- GOC_MaricultureProduction %>%
  filter(year > 2017) %>%
  filter(!is.na(estimated_genus_name) & estimated_genus_name != "NEI" & estimated_genus_name != "unknown") %>%
  select(estimated_genus_name, Major_Functional_Group) %>%
  group_by(Major_Functional_Group, estimated_genus_name) %>%
  summarise(
    n_records_since2018 = n(),
    .groups = "drop"
  )

# Filter out the most common genera:
Mex.Pacific_MajorMariculture_summary<- Mex.Pacific_MajorMariculture_summary %>%
  filter(n_records_since2018 > 10)
## 11 obs. (down from 15)

MaricultureGenera_of_Interest<- unique(Mex.Pacific_MajorMariculture_summary$estimated_genus_name)
```

```{r Investigate records for unexpected taxa}
table(Mex.Pacific_MajorMariculture_summary$estimated_genus_name)
table(GOC_MaricultureProduction$reported_aquaculture_production) # is no = 1053
sum(is.na(GOC_MaricultureProduction$reported_aquaculture_production)) # is na = 104204

# Investigate weird entries
subset1<- GOC_MaricultureProduction %>%
  filter(year > 2017) %>%
  filter(estimated_genus_name == "Lutjanus") 
table(subset1$eu_name)
# Makes sense! All records from Ocean Farms in La Paz.

subset2<- GOC_MaricultureProduction %>%
  filter(year > 2017) %>%
  filter(estimated_genus_name == "Argopecten") 
table(subset2$keywords_1)
# Makes sense! All records say "de cultivo" explicitly.

subset4<- GOC_MaricultureProduction %>%
  filter(year > 2017) %>%
  filter(estimated_genus_name == "Dosinia") 
table(subset4$keywords_1)
# Makes sense! All records say "de cultivo" explicitly.

subset5<- GOC_MaricultureProduction %>%
  filter(year > 2017) %>%
  filter(estimated_genus_name == "Megapitaria") 
table(subset5$keywords_1)
# Makes sense! All records say "de cultivo" explicitly.
```

```{r Summarize MaricultureGenera_of_Interest over time}
colnames(GOC_MaricultureProduction)

GOC_AquacultureProduction_summary <- GOC_MaricultureProduction %>%
  filter(estimated_genus_name %in% MaricultureGenera_of_Interest) %>%
  select(year, estimated_genus_name, weight_kg_corrected, value) %>%
  group_by(estimated_genus_name, year) %>%
  summarise(
    annual_production_mt = sum(weight_kg_corrected, na.rm = TRUE) / 1000,
    annual_value_pesos   = sum(value, na.rm = TRUE),
    .groups = "drop"
  )
# 97 obs., 4 vars.
```

```{r Plot CONAPESCA aquaculture}
GOC_AquacultureProduction_summary %>%
  filter(year < 2025) %>%
  filter(estimated_genus_name=="Penaeus") %>%
ggplot(aes(x = year, y = annual_production_mt)) +
  geom_line(color = "tomato") +
  labs(x = "Year", y = "Production (mt)",
       title = "CONAPESCA Aquaculture Production of Shrimp (Penaeus)") +
  theme_minimal()

GOC_AquacultureProduction_summary %>%
ggplot(aes(x = year, y = annual_production_mt)) +
  geom_line(color = "tomato") +
  facet_wrap(~ estimated_genus_name, scales = "free_y") +
  labs(x = "Year", y = "Production (mt)",
       title = "CONAPESCA Aquaculture Production by genus") +
  theme_minimal()
```

The CONAPESCA aquaculture data look super weird.
We might want to go with FAO data in this case. 
But what if we summarize CONAPESCA data at a higher level?

# Summarize at the family level?
```{r We'll want to add family info to the CONAPESCA data based on genus}
AquacultureFamily_key0<- read.csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/AquacultureFamilyKey_INCOMPLETE_Apr042026.csv")
# 113 obs., 7 vars.

# Trim key and simplify (as needed):
colnames(AquacultureFamily_key0)
AquacultureFamily_key_filtered<- AquacultureFamily_key0 %>%
  select(estimated_genus_name, estimated_taxa_group, Major_Functional_Group,
         new_estimated_family_name) %>%
  distinct()

# if estimated_genus_name is.na, "NEI", or "unknown" & estimated_taxa_group = "OTRAS" then Major_Functional_Group should be NA, and new_estimated_family_name should be NA

AquacultureFamily_key_filtered <- AquacultureFamily_key0 %>%
  # Select only relevant columns
  select(
    estimated_genus_name, 
    estimated_taxa_group, 
    Major_Functional_Group,
    new_estimated_family_name
  ) %>%
  # Keep only unique rows
  distinct() %>%
  # Apply rules for placeholder genera
  mutate(
    # Rule 1: if estimated_genus_name is NA, "NEI", or "unknown", new_estimated_family_name = NA
    new_estimated_family_name = if_else(
      is.na(estimated_genus_name) | estimated_genus_name %in% c("NEI", "unknown"),
      NA_character_,
      new_estimated_family_name
    ),
    # Rule 2: if estimated_genus_name is NA, "NEI", or "unknown" AND estimated_taxa_group = "OTRAS", Major_Functional_Group = NA
    Major_Functional_Group = if_else(
      (is.na(estimated_genus_name) | estimated_genus_name %in% c("NEI", "unknown")) & 
        estimated_taxa_group == "OTRAS",
      NA_character_,
      Major_Functional_Group
    )
  ) %>%
  distinct()
# 92 obs., 3 vars.

AquacultureFamily_key_filtered<- AquacultureFamily_key_filtered %>%
  select(-Major_Functional_Group) %>%
  distinct()

# table(AquacultureFamily_key_filtered$estimated_genus_name)
# No genera duplicates. Looks good!
```

```{r Merge AquacultureFamily_key_filtered to GOC_AquacultureProduction_summary}
# Make estimated_genus_name lowercase so the merge goes smoothly:
AquacultureFamily_key_filtered$estimated_genus_name <- tolower(AquacultureFamily_key_filtered$estimated_genus_name)

GOC_MaricultureProduction$estimated_genus_name <- tolower(GOC_MaricultureProduction$estimated_genus_name)

GOC_MaricultureProduction$estimated_family_name <- tolower(GOC_MaricultureProduction$estimated_family_name)

# Merge in key:
GOC_MaricultureProduction <- left_join(GOC_MaricultureProduction, AquacultureFamily_key_filtered, 
                           by = c("estimated_genus_name", "estimated_taxa_group"))
# 105932 obs., 63 vars.
```

```{r Take a quick look}
TEST <- GOC_MaricultureProduction %>%
  filter(year > 2017) %>%
  select(new_estimated_family_name, Major_Functional_Group) %>%
  group_by(Major_Functional_Group, new_estimated_family_name) %>%
  summarise(
    n_records_since2018 = n(),
    .groups = "drop"
  )
```

```{r Let's just look at Penaeidae for now}
GOC_MaricultureProduction$new_estimated_family_name <- tolower(GOC_MaricultureProduction$new_estimated_family_name)

GOC_Penaeidae<- GOC_MaricultureProduction %>%
  filter(new_estimated_family_name== "penaeidae")
# 83046 obs.
range(GOC_Penaeidae$year)
# From 2006 to 2025

colnames(GOC_Penaeidae)

GOC_Penaeidae <- GOC_Penaeidae %>%
  # Select and rename columns
  select(
    estimated_family_name = new_estimated_family_name, 
    year,
    weight_kg_corrected, 
    Major_Functional_Group
  ) %>%
  # Create new column for production in metric tons
  mutate(
    production_mt = weight_kg_corrected / 1000,
    # Add source column with constant value
    source = "CONAPESCA"
  )

GOC_Penaeidae<- GOC_Penaeidae %>%
  select(-c(weight_kg_corrected, Major_Functional_Group))

FAO_Penaeidae<- FAO_MaricultureHarvest %>%
    filter(estimated_family_name== "penaeidae")
range(FAO_Penaeidae$year)
# From 2006 to 2025

FAO_Penaeidae<- FAO_Penaeidae %>%
  mutate(
    source = "FAO"
  ) %>%
  select(-c(Major_Functional_Group))
```

```{r Join and plot}
# Combine the dataframes
combined_df <- bind_rows(GOC_Penaeidae, FAO_Penaeidae)

# Quick check
head(combined_df)

# Plot production over years, color by source
ggplot(combined_df, aes(x = year, y = production_mt, color = source)) +
  geom_line() +      # line plot
  geom_point() +     # add points
  labs(
    title = "Penaeidae Production Over Years",
    x = "Year",
    y = "Production (mt)"
  ) +
  theme_minimal()
```


# NEWEST RESEARCH (as of APRIL 14 2026)  ------------------------------

# STEP 1: Generate MARICULTURE list for all of the Mexican Pacific based on CONAPESCA COSECHA ------------------------------

```{r Load Mex.Pacific_Landings_sector if needed}
Mex.Pacific_Landings_sector<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/Mex.Pacific_Landings_SectorSorted_2000.present_EMF.Apr152026.rds"))
# 6406303 obs., 62 vars.
```

```{r TESTING TESTING 1 2 3: Make all oyster production mariculture}
colnames(Mex.Pacific_Landings_sector)

Mex.Pacific_Landings_sector <- Mex.Pacific_Landings_sector %>%
  mutate(
    sector_estimate_draft = if_else(
      estimated_taxa_group == "OSTION",
      "marine: aquaculture",
      sector_estimate_draft
    )
  )
```

```{r Filter out the MARICULTURE sector}
# Let's assume that the records are most accurate since 2018.

table(Mex.Pacific_Landings_sector$sector_estimate_draft)

Mariculture_Sectors<- c("marine: aquaculture")

Mex.Pacific_MaricultureHarvest<- Mex.Pacific_Landings_sector %>% 
  filter(sector_estimate_draft %in% Mariculture_Sectors)
# 150037 obs., 62 vars.
# With all oysters (Apr 28 2026): 192933 (+42k) obs., 62 vars.

range(Mex.Pacific_MaricultureHarvest$year)
```

```{r Look into this first-draft mariculture list before proceeding}
# Taking the information I've gleaned here to refine my sector estimates above.
table(Mex.Pacific_MaricultureHarvest$estimated_genus_name)
table(Mex.Pacific_MaricultureHarvest$Major_Functional_Group)
```

```{r Generate a table of most common mariculture species since 2018}
Mex.Pacific_MajorMariculture_summary <- Mex.Pacific_MaricultureHarvest %>%
  filter(year > 2018) %>%
  filter(!is.na(estimated_genus_name) & estimated_genus_name != "NEI" & estimated_genus_name != "unknown") %>%
  select(estimated_genus_name, Major_Functional_Group) %>%
  group_by(Major_Functional_Group, estimated_genus_name) %>%
  summarise(
    n_records_since2018 = n(),
    .groups = "drop"
  )

# Filter out the most common genera:
Mex.Pacific_MajorMariculture_summary<- Mex.Pacific_MajorMariculture_summary %>%
  filter(n_records_since2018 > 10)
## 11 obs.

MaricultureGenera_of_Interest<- unique(Mex.Pacific_MajorMariculture_summary$estimated_genus_name)
```

NOTE TO SELF: This isn't quite picking up the species I would have expected it to. For example, I would expect to see some Abalone mariculture, but I don't...

FOLLOW UP: Haliotis (abalone) is in there, but not after 2018. This would also explain the big wobble on aquaculture landings through time...

```{r Sanity check: where is Abalone?}
# SANITY CHECK: Where is Abalone?
subset<- Mex.Pacific_MaricultureHarvest %>%
  filter(estimated_genus_name == "Haliotis")
# Only flagged in 2006. This suggests that my method of estimating sector is misisng catches. Makes sense if these catches weren't obviously flagged as aquaculture: not marked as aquaculture, no key words that would indicate aquaculture, etc.
```

```{r Investigate records for unexpected taxa}
table(Mex.Pacific_MajorMariculture_summary$estimated_genus_name)
table(Mex.Pacific_MaricultureHarvest$reported_aquaculture_production) # is no = 1053
sum(is.na(Mex.Pacific_MaricultureHarvest$reported_aquaculture_production)) # is na = 104204

# Investigate specific entries
subset1<- Mex.Pacific_MaricultureHarvest %>%
  filter(year > 2017) %>%
  filter(estimated_genus_name == "Lutjanus") 
table(subset1$eu_name)
# Makes sense! All records from Ocean Farms in La Paz.

subset2<- Mex.Pacific_MaricultureHarvest %>%
  filter(year > 2017) %>%
  filter(estimated_genus_name == "Argopecten") 
table(subset2$keywords_1)
# Makes sense! All records say "de cultivo" explicitly.

subset4<- Mex.Pacific_MaricultureHarvest %>%
  filter(year > 2017) %>%
  filter(estimated_genus_name == "Dosinia") 
table(subset4$keywords_1)
# Makes sense! All records say "de cultivo" explicitly.

subset5<- Mex.Pacific_MaricultureHarvest %>%
  filter(year > 2017) %>%
  filter(estimated_genus_name == "Megapitaria") 
table(subset5$keywords_1)
# Makes sense! All records say "de cultivo" explicitly.
```

```{r Summarize MaricultureGenera_of_Interest over time}
colnames(Mex.Pacific_MaricultureHarvest)

Mex.Pacific_MaricultureHarvest_summary <- Mex.Pacific_MaricultureHarvest %>%
  filter(estimated_genus_name %in% MaricultureGenera_of_Interest) %>%
  select(year, estimated_genus_name, weight_kg_corrected, value) %>%
  group_by(estimated_genus_name, year) %>%
  summarise(
    annual_production_mt = sum(weight_kg_corrected, na.rm = TRUE) / 1000,
    annual_value_pesos   = sum(value, na.rm = TRUE),
    .groups = "drop"
  )
# 117 obs., 4 vars.
```

```{r MaricultureFamilies_of_Interest based on most common CONAPESCA landings}
colnames(Mex.Pacific_MaricultureHarvest)

AquacultureFamilies_of_Interest <- Mex.Pacific_MaricultureHarvest %>%
  filter(estimated_genus_name %in% MaricultureGenera_of_Interest) %>%
  distinct(estimated_family_name) %>%
  pull(estimated_family_name)

print(AquacultureFamilies_of_Interest)
```
Looks good!


# STEP 2: Let's look into the Aquaculture PRODUCCION AVISOS (co-author: Leonardo Vazquez) ------------------------------
```{r AVISOS MAYORES}
Mex_AVISOS_2018 <- read.csv(
  "/home/shares/ohi/OHI_GOC/_raw_data/CONAPESCA/d2025/SharedByLeonardo/AVISOS_ MAYORES_MENORES_COSECHA_2018.csv",
  skip = 2,
  header = TRUE,
  fileEncoding = "latin1"
)

Mex_AVISOS_2018 <- tolower(Mex_AVISOS_2018)

# 35 obs., 62115 vars.
```

Let's see if THESE avisos are = Juan Carlos' avisos for the year 2018, state of Sonora
```{r Comparing AVISOS shared by Leo versus JC}
colnames(Mex_AVISOS_2018)
colnames(All.Mex_Landings_2000.present)

Sonora_Leo<- Mex_AVISOS_2018 %>%
  filter(aa_o_corte == 2018) %>%
  filter(nombre_estado == "SONORA")
# 29294 entries for 35 variables

Sonora_JC<- All.Mex_Landings_2000.present %>%
  filter(year == 2018) %>%
  filter(state == "SONORA")
# 29294 entries for 20 variables.

table(Sonora_Leo$nombre_principal)
     #      ALMEJA          ANCHOVETA              BAGRE            BANDERA            BAQUETA          BARRILETE          BERRUGATA 
     #          1126                614                 86                  1                749                  1               1012 
     #      CABRILLA            CALAMAR            CAMARON            CARACOL              CARPA              CAZON            CORVINA 
     #           331                 98               6803                874                 39                498               2133 
     #   GUACHINANGO              JAIBA              JUREL           LANGOSTA           LENGUADO               LISA             LOBINA 
     #           299               2569                243                 63               1275                133                  4 
     #      MACARELA               MERO            MOJARRA             OSTION              OTRAS            PAMPANO              PARGO 
     #           653                  2                247                 65               3573                 25                481 
     # PEPINO DE MAR             PIERNA              PULPO   RAYA Y SIMILARES             ROBALO              RONCO RUBIA Y VILLAJAIBA 
     #            16                115                324               1097                 21                  7                 29 
     #         RUBIO            SARDINA             SIERRA            TIBURON 
     #             1               2248                565                874 
                 
table(Sonora_JC$main_species_group)
     #        ALMEJA          ANCHOVETA              BAGRE            BANDERA            BAQUETA          BARRILETE          BERRUGATA 
     #          1126                614                 86                  1                749                  1               1012 
     #      CABRILLA            CALAMAR            CAMARON            CARACOL              CARPA              CAZON            CORVINA 
     #           331                 98               6803                874                 39                498               2133 
     #   GUACHINANGO              JAIBA              JUREL           LANGOSTA           LENGUADO               LISA             LOBINA 
     #           299               2569                243                 63               1275                133                  4 
     #      MACARELA               MERO            MOJARRA             OSTION              OTRAS            PAMPANO              PARGO 
     #           653                  2                247                 65               3573                 25                481 
     # PEPINO DE MAR             PIERNA              PULPO   RAYA Y SIMILARES             ROBALO              RONCO RUBIA Y VILLAJAIBA 
     #            16                115                324               1097                 21                  7                 29 
     #         RUBIO            SARDINA             SIERRA            TIBURON 
     #             1               2248                565                874 
```
Excellent! AVISOS shared by Leo or JC are effectively equivalent. The seem to come from the same raw data, same ideas, same number.
Now let's look in to the aquaculture avisos specifically.

```{r AVISOS PRODUCCION - i.e. Aquaculture}
Aquaculture_AVISOS_2018 <- read.csv(
  "/home/shares/ohi/OHI_GOC/_raw_data/CONAPESCA/d2025/SharedByLeonardo/Produccion acuicola/AVISOS_PRODUCCION_2018.csv",
  skip = 1,
  header = TRUE,
  fileEncoding = "latin1"
)

Aquaculture_AVISOS_2018 <- tolower(Aquaculture_AVISOS_2018)
# 5014 obs., 24 vars.
```

```{r Clean up the Aquaculture_AVISOS_2018 df}
options(scipen = 999)

unique(Aquaculture_AVISOS_2018$unidad_medida)
table(Aquaculture_AVISOS_2018$unidad_medida)
# [1] "PIEZAS"   "MILES"    "MILLONES" ""

conversion_factor <- c(
  "PIEZAS" = 1,
  "MILES" = 1e3,
  "MILLONES" = 1e6
)

Aquaculture_AVISOS_2018 <- Aquaculture_AVISOS_2018 %>%
  mutate(
    cantidad_clean = as.numeric(
      gsub(",", "", trimws(cantidad))
    ),
    unidad_medida = toupper(trimws(unidad_medida)),
    cantidad_std = case_when(
      unidad_medida == "PIEZAS" ~ cantidad_clean,
      unidad_medida == "MILES" ~ cantidad_clean * 1e3,
      unidad_medida == "MILLONES" ~ cantidad_clean * 1e6,
      TRUE ~ NA_real_
    )
  )

Aquaculture_AVISOS_2018$precio_pesos_clean <- as.numeric(
  gsub(",", "",
       trimws(Aquaculture_AVISOS_2018$precio_pesos)
  )
)

Aquaculture_AVISOS_2018$valor_pesos_clean <- as.numeric(
  gsub(",", "",
       trimws(Aquaculture_AVISOS_2018$valor_pesos)
  )
)
```

```{r Explore Aquaculture_AVISOS_2018}
colnames(Aquaculture_AVISOS_2018)
table(Aquaculture_AVISOS_2018$nombre_especie)

Sinaloan_Shrimp_Aquaculutre_2018<- Aquaculture_AVISOS_2018 %>%
  filter(ano_corte == 2018) %>%
  filter(nombre_especie == "CAMARON BLANCO") %>%
  filter(nombre_estado == "SINALOA")
# 519 entries for 24 variables.

# "AQUAPACIFIC, SA DE CV" is an economic unit listed here. Let's see if we can find it in the other dataset...
```

```{r Compare economic unit names from Mex_AVISOS_2018 to Mex.Pacific_Landings_sector}
colnames(Mex_AVISOS_2018)

shared_vals_All <- intersect(All.Mex_Landings_2000.present$eu_name,
                         Aquaculture_AVISOS_2018$unidad_economica)
print(shared_vals_All)
# 147 shared values for all of Mexico

shared_vals_Pacific <- intersect(Mex.Pacific_Landings_sector$eu_name,
                         Aquaculture_AVISOS_2018$unidad_economica)
print(shared_vals_Pacific)
# 77 shared values for the Mexican Pacific

shared_vals_Pacific.Flagged <- intersect(Mex.Pacific_MaricultureHarvest$eu_name,
                         Aquaculture_AVISOS_2018$unidad_economica)
print(shared_vals_Pacific.Flagged)
# 31 shared values for the Mexican Pacific that I was able to estimate as mariculture using my flag system alone. This means that my flags are missing some economic units that probably are producing aquaculture.

test<- Mex_AVISOS_2018 %>%
  filter(aa_o_corte == 2018) %>%
  filter(unidad_economica == "EARTH OCEAN FARMS S DE RL CV")
# 31 entries for 20 variables.

test2<- Mex.Pacific_Landings_sector %>%
  filter(year == 2018) %>%
  filter(eu_name == "EARTH OCEAN FARMS S DE RL CV")
# 31 entries for 62 variables. Excellent!

test3<- Mex.Pacific_Landings_sector %>%
  filter(eu_name == "JORGE ALBERTO GALVAN MORA")
# 5 entries for 20 variables.
# Interesting. These entries are not marked as aquaculture, but there are key words that my filters picked up. Good!
```

```{r Which aquaculture producers is my flag system overlooking?}
shared_vals_Pacific <- intersect(Mex.Pacific_Landings_sector$eu_name,
                         Aquaculture_AVISOS_2018$unidad_economica)
print(shared_vals_Pacific)
# 77 shared values for the Mexican Pacific

shared_vals_Pacific.Flagged <- intersect(Mex.Pacific_MaricultureHarvest$eu_name,
                         Aquaculture_AVISOS_2018$unidad_economica)
print(shared_vals_Pacific.Flagged)
# 31 shared values for the Mexican Pacific that I was able to estimate as mariculture using my flag system alone. This means that my flags are missing some economic units that probably are producing aquaculture.

EUs_overlooked <- setdiff(shared_vals_Pacific, shared_vals_Pacific.Flagged)
print(EUs_overlooked)

Landings_overlooked<- Mex.Pacific_Landings_sector %>%
  filter(eu_name %in% EUs_overlooked)
min(Landings_overlooked$year)
max(Landings_overlooked$year)
# 11494 obs. overlooked across all years.

table(Landings_overlooked$sector_estimate_draft)

              # marine: fishery sector yet to be classified                               marine: large-scale capture 
              #                                        5476                                                      2190 
              #                 marine: small-scale capture                                       marine: aquaculture 
              #                                        2439                                                         0 
              #                     freshwater: aquaculture                           freshwater: small-scale capture 
              #                                        1109                                                        16 
              #             freshwater: large-scale capture           freshwater: fishery sector yet to be classified 
              #                                           0                                                        17 
              #           freshwater or marine: aquaculture                 freshwater or marine: small-scale capture 
              #                                         172                                                        34 
              #   freshwater or marine: large-scale capture freshwater or marine: fishery sector yet to be classified 
              #                                           0                                                        41 

# Looks like: a lot of these were CORRECTLY flagged as aquaculture just freshwater or yet to be classified.
# About 40% (2439 + 2190=4629) of these "overlooked" observations might be INCORRECTLY marked as small-scale or large-scale industrial fisheries. Let's look into those.
# Turns out these entries are legit, probably not aquaculture. At least there were marked as NOT aquaculture by the fishers, even if the EU name might suggest aquaculture. Cool!

test4<- Landings_overlooked %>%
  filter(sector_estimate_draft=="marine: large-scale capture")
# 2190 obs.
table(test4$reported_fleet)
table(test4$Major_Functional_Group)
# These are all landings that were explicitly reported as large-scale. I think my filters might be working...

test5<- Landings_overlooked %>%
  filter(sector_estimate_draft=="marine: small-scale capture")
# 2439 obs.
table(test5$reported_fleet)
table(test5$Major_Functional_Group)
# These are all landings that were explicitly reported as small-scale and NOT aquaculture. I think my filters are working as expected here.
```

- Okay, so what I'm getting from this exercise is that my flags are actually working pretty well. 
- My sector estimates probably aren't too far off.
- BUT ALSO, there could be important aquaculture entries excluded from the CONAPESCA AVISOS_MAYORES_MENORES_COSECHA file because the production wasn't actually officially "harvested".

- To this point, Leonardo writes: "To clarify what these "Produccion "data includes, I am attaching the template which is mandatory for aquaculturists, aquaculture laboratories, and fishers to report the quantity and type of organisms they have cultivated or collected from nature. You'll see number of organisms and life stage of this "production" but how much of this was harvested? that you'll find it in the integrated databases as "cosecha" (harvest). As mentioned, one thing is to produce in the lab or sometimes collect from the field, and another thing is to grow, harvest and sell the products. There are several species that are "produced" in laboratories for research goals and looking to develop the cultivation but just a few are harvested and reported."

- Based on all this, I think it makes the most sense to compare the FAO aquaculture data to the data listed in the COSECHA AVISOS as oppossed to the PRODUCCION AVISOS (Aquauculture Avisos)  which is confounded with other types of aquaculture data...

```{r Summarize Aquaculture_AVISOS_2018 versus FAO}
colnames(Aquaculture_AVISOS_2018)

# Summarized Production for Aquaculture_AVISOS_2018
table(Aquaculture_AVISOS_2018$fase)
# I think we want to filter out Post-Larva and Juveniles

Estimated_MAR_2018 <- Aquaculture_AVISOS_2018 %>%
  filter(fase %in% c("POSTLARVA", "JUVENILES")) %>%
  group_by(nombre_especie) %>%
  summarise(total_cantidad = sum(cantidad_clean, na.rm = TRUE),
            total_cantidad_std = sum(cantidad_std, na.rm = TRUE), 
            .groups = "drop")

# Summarized Production for FAO Aquaculture (2018)
```


# STEP 3: Investigate FAO Aquaculture Data for Eastern Pacific ------------------------------

```{r Load in FAO_FamilyGenus_key_filtered}
FAO_FamilyGenus_key0<- read.csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/FAO_FamilyGenusKey_COMPLETE_Apr042026.csv")
# 193 obs., 7 vars.

# Trim key and simplify (as needed):
colnames(FAO_FamilyGenus_key0)

FAO_FamilyGenus_key_filtered<- FAO_FamilyGenus_key0 %>%
  select(estimated_genus_name, Major_Functional_Group, 
         estimated_family_name= new_estimated_family_name) %>%
  distinct(estimated_family_name, .keep_all = TRUE)
# 105 obs., 3 vars.
```

```{r Look into FAO aquaculture data}
# Load data
FAO_Landings_Data0<- read.csv('/home/ferrer/ohi_goc_prep/FP/CMSY_Analysis/CMSY_Revised_April2026/Generating_Model_Priors/Priors_Data/FAOMexicoData_19762023_downloadedApr042026.csv')
# FAO timeseries data organized with Stock (taxa) as # columns and years as # rows.
    
FAO_Landings_Data<- FAO_Landings_Data0

# We've got to do a little work on these data before they're ready to use.
colnames(FAO_Landings_Data)

# Clean column names
tolower <- colnames(FAO_Landings_Data) %>%
  # Remove trailing periods
  str_replace_all("\\.$", "") %>%
  # Replace multiple periods with a single underscore
  str_replace_all("\\.+", "_") %>%
  # Remove leading X_ for years
  str_replace("^X_(\\d+)$", "\\1") %>%
  # Convert to lowercase
  tolower()

# Apply cleaned names to the dataframe
colnames(FAO_Landings_Data) <- tolower

# Delete the columns that start with s or s_, we don't need these.
FAO_Landings_Data <- FAO_Landings_Data %>%
  select(-starts_with("s"))

# Filter for aquaculture production:
table(FAO_Landings_Data$detailed_production_source_name)

FAO_Landings_filtered<- FAO_Landings_Data %>%
  filter(detailed_production_source_name %in% c("Aquaculture production (marine)", 
                                                "Aquaculture production (brackishwater)"))

# Delete other columns we don't need:
colnames(FAO_Landings_filtered)

FAO_Landings_filtered<- FAO_Landings_filtered %>%
  select(-c(country_name, fao_major_fishing_area_name,
            unit_name, unit))

# Transpose these data such that years are rows per taxa (and rename species column):
FAO_MaricultureHarvest <- FAO_Landings_filtered %>%
  # Rename species column
  rename(estimated_family_name = asfis_species_scientific_name,
         FAO_source = detailed_production_source_name) %>%
  
  # Step 1: go long
  pivot_longer(
    cols = -estimated_family_name,
    names_to = "year",
    values_to = "production_mt"
  ) %>%
  
  # Step 2: clean year
  mutate(year = as.integer(year),
         production_mt = as.numeric(production_mt))

FAO_MaricultureHarvest <- FAO_Landings_filtered %>%
  # Step 1: rename columns
  rename(
    estimated_family_name = asfis_species_scientific_name,
    FAO_source = detailed_production_source_name
  ) %>%
  
  # Step 2: go long
  pivot_longer(
    cols = -c(estimated_family_name, FAO_source),
    names_to = "year",
    values_to = "production_mt"
  ) %>%
  
  # Step 3: clean year
  mutate(
    year = as.integer(year),
    production_mt = as.numeric(production_mt)
  )
# 2325 obs., 4 vars.

FAO_MaricultureHarvest <- FAO_MaricultureHarvest %>%
  group_by(estimated_family_name, year) %>%
  summarize(marine.brackish_production_mt = sum(production_mt)) %>%
  arrange(year) %>%
  filter(year >= min(year[marine.brackish_production_mt != 0], na.rm = TRUE))
# 351 obs., 3 vars.

# How many families are there in this database?
unique(FAO_MaricultureHarvest$estimated_family_name)
# 25 families (really 24 families, excluding "others")
```

```{r Keep workspace clean}
rm(FAO_Landings_Data)
rm(FAO_Landings_filtered)
```

```{r Load in FAO_FamilyGenus_key_filtered}
FAO_FamilyGenusKey_pending<- read.csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/FAO_FamilyGenusKey_COMPLETE_Apr042026.csv")
# 193 obs., 7 vars.

# Trim key and simplify (as needed):
colnames(FAO_FamilyGenusKey_pending)

FAO_FamilyGenusKey_filtered<- FAO_FamilyGenusKey_pending %>%
  select(Major_Functional_Group, 
         estimated_family_name= new_estimated_family_name) %>%
  distinct(estimated_family_name, .keep_all = TRUE)
# 105 obs., 2 vars.
```

```{r Merge FAO_FamilyGenusKey_filtered to FAO data}
# Make family lowercase so the merge goes smoothly:
FAO_FamilyGenusKey_filtered$estimated_family_name <- tolower(FAO_FamilyGenusKey_filtered$estimated_family_name)

FAO_MaricultureHarvest$estimated_family_name <- tolower(FAO_MaricultureHarvest$estimated_family_name)

# Merge in key:
FAO_MaricultureHarvest <- left_join(FAO_MaricultureHarvest, FAO_FamilyGenusKey_filtered,
                           by = "estimated_family_name")

# Rename columns:
colnames(FAO_MaricultureHarvest)

FAO_MaricultureHarvest<- FAO_MaricultureHarvest %>%
  rename(FAO_est_production = marine.brackish_production_mt)
```

```{r Plot FAO aquaculture over time}
# FAO Production for all families
FAO_MaricultureHarvest %>%
  filter(year > 2018 & year < 2025) %>%
  
  ggplot(aes(x = year, y = FAO_est_production)) +
  geom_line(color = "steelblue") +
  
  facet_wrap(~ estimated_family_name, scales = "free_y") +
  
  labs(x = "Year", y = "Production (mt)",
       title = "FAO Aquaculture Production (by family)") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# FAO Production for AquacultureFamilies_of_Interest
AquacultureFamilies_of_Interest <- tolower(AquacultureFamilies_of_Interest)

FAO_MaricultureHarvest %>%
  filter(estimated_family_name %in% AquacultureFamilies_of_Interest) %>%
  filter(year > 2018 & year < 2025) %>%
  
  ggplot(aes(x = year, y = FAO_est_production)) +
  geom_line(color = "steelblue") +
  
  facet_wrap(~ estimated_family_name, scales = "free_y") +
  
  labs(x = "Year", y = "Production (mt)",
       title = "FAO Aquaculture Production (by family)") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```


# STEP 4: Compare CONAPESCA COSECHA to FAO data
```{r Summarize mariculture harvest over all years from CONAPESCA COSECHA estimates}
colnames(Mex.Pacific_MaricultureHarvest)

# Summarized Production for CONAPESCA cosecha production in all years.
COSECHA_MaricultureHarvest <- Mex.Pacific_MaricultureHarvest %>%
# filter(year == 2018) %>%
  group_by(year, estimated_family_name) %>%
  summarise(total_est_production_mt = (sum(weight_kg_corrected, na.rm = TRUE)/1000),
            .groups = "drop")

COSECHA_MaricultureHarvest$estimated_family_name<- tolower(COSECHA_MaricultureHarvest$estimated_family_name)

COSECHA_MaricultureHarvest <- COSECHA_MaricultureHarvest %>%
  mutate(
    estimated_family_name = tolower(estimated_family_name),
    year = as.integer(year)
  ) %>%
  rename(CONAPESCA_est_production = total_est_production_mt)
```

```{r Merge these two dataframes}

# SANITY CHECK: Check overlap before merge
intersect(
  FAO_MaricultureHarvest$estimated_family_name,
  COSECHA_MaricultureHarvest$estimated_family_name
)

# Step 3: Merge (recommended approach)
FAO.COSECHA_MaricultureHarvest <- FAO_MaricultureHarvest %>%
  full_join(
    COSECHA_MaricultureHarvest,
    by = c("year", "estimated_family_name")
  )
  
# 610 obs., 5 vars.

# Merge in key:
FAO.COSECHA_MaricultureHarvest<- FAO.COSECHA_MaricultureHarvest %>%
  select(-c(Major_Functional_Group))

FAO.COSECHA_MaricultureHarvest <- left_join(FAO.COSECHA_MaricultureHarvest, FAO_FamilyGenusKey_filtered,
                           by = "estimated_family_name")
```

```{r Filter FAO.COSECHA_MaricultureHarvest}
# I want to filter an manipulate FAO.COSECHA_MaricultureHarvest in a few ways. First, I want to filter only years since 2000. then, I want to drop any rows where FAO_est_production and CONAPESCA_est_production both equal zero or NA. Finally, I want to estimate the biomass discrepancy between FAO_est_production and CONAPESCA_est_production in a new column.

FAO.COSECHA_MaricultureHarvest <- FAO.COSECHA_MaricultureHarvest %>%
  filter(year >= 2018) %>%
  mutate(
    valid_pair =
      !is.na(FAO_est_production) &
      !is.na(CONAPESCA_est_production) &
      FAO_est_production != 0 &
      CONAPESCA_est_production != 0,

    biomass_discrepancy = ifelse(
      valid_pair,
      FAO_est_production - CONAPESCA_est_production,
      NA_real_
    ),

    ratio_CONAPESCA.FAO = ifelse(
      valid_pair,
      round(
        ((CONAPESCA_est_production / FAO_est_production) * 100),
        2
      ),
      NA_real_
    )
  )

FAO.COSECHA_MaricultureHarvest <- FAO.COSECHA_MaricultureHarvest %>%
  filter(
    !(coalesce(FAO_est_production, 0) == 0 &
      coalesce(CONAPESCA_est_production, 0) == 0)
  )

colnames(FAO.COSECHA_MaricultureHarvest)

write.csv(FAO.COSECHA_MaricultureHarvest,
          "/home/ferrer/ohi_goc_prep/FP/Exploratory_Research/MAR_ExploratoryResearch/FAOvsCOSECHA_Apr292026.csv",
          row.names = FALSE)

# Save table at html:
library(kableExtra)

FAO.COSECHA_MaricultureHarvest %>%
  knitr::kable(digits = 2, format = "html") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE
  ) %>%
  row_spec(0, bold = TRUE)

ratio_bg <- ifelse(
  is.na(FAO.COSECHA_MaricultureHarvest$ratio_CONAPESCA.FAO),
  "white",  # or "#ffffff"
  ifelse(
    FAO.COSECHA_MaricultureHarvest$ratio_CONAPESCA.FAO < 70 |
      FAO.COSECHA_MaricultureHarvest$ratio_CONAPESCA.FAO > 130,
    "#ffe6e6",
    "lightgreen"
  )
)

FAO.COSECHA_Harvest_table<- FAO.COSECHA_MaricultureHarvest %>%
  knitr::kable(digits = 2, format = "html") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  column_spec(
    which(names(FAO.COSECHA_MaricultureHarvest) == "ratio_CONAPESCA.FAO"),
    background = ratio_bg
  )
  

FAO.COSECHA_Harvest_table

save_kable(FAO.COSECHA_Harvest_table, file = "/home/ferrer/ohi_goc_prep/FP/Exploratory_Research/MAR_ExploratoryResearch/FAOvsCOSECHA_Apr292026.html")
```

```{r Plot COSECHA vs FAO over time}
FAO.COSECHA_MarHarvest_long <- FAO.COSECHA_MaricultureHarvest %>%
  pivot_longer(
    cols = c(FAO_est_production, CONAPESCA_est_production),
    names_to = "source",
    values_to = "production_mt"
  )

# Fill NA values with 0

# Just oyster production: 
FAO.COSECHA_MarHarvest_long %>%
  filter(estimated_family_name=="ostreidae") %>%
  
  ggplot(aes(x = year, y = production_mt, color = source)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ estimated_family_name, scales = "free_y") +
  labs(
    x = "Year",
    y = "Production (mt)",
    color = "Source"
  ) +
  theme_bw()

# All production:
FAO.COSECHA_MarHarvest_long %>%
  filter(estimated_family_name %in% AquacultureFamilies_of_Interest) %>%
  filter(year >= 2018 & year < 2025) %>%
  
  ggplot(aes(x = year, y = production_mt, color= source)) +
  geom_point() +
  geom_line() +
  
  facet_wrap(~ estimated_family_name, scales = "free_y") +
  
  labs(x = "Year", y = "Production (mt)",
       title = "Production in the Mex Pacific (CONAPESCA vs. FAO)") +
  
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
OUTCOME (Apr 28 2026): We tested that these are effectively equivalent and because there's no consistent bias, we went with CONAPESCA data.


# STEP 4: Generate MAR scores, DRAFT as of Apr 28 ------------------------------

## General plan ------------------------------
1. Load in MAR seafood watch scores
1B. Prepare seafood watch scores as needed

2. Load in MAR tonnage from CONAPESCA
2B. Prepare CONAPESCA tonnage as needed

3. Load in MAR tonnage from FAO
3B. Prepare FAO tonnage as needed

4. Weight the CONAPESCA tonnage by the "correction factor" from the FAO, IF NEEDED.
If within 10% [25%], comfortable going with CONAPESCA data.
--> Bring a comparison table to meeting.

5. Multiply MAR seafood watch scores by tonnage per region to get a "weighted tonnage"

If CONAPESCA data is really off, still have FAO data and proportions
Which way should we go based on these numbers?
```{r Summarize my COSECHA data in all years all states}
colnames(Mex.Pacific_MaricultureHarvest)
table(Mex.Pacific_MaricultureHarvest$OHI_region_draft)
table(Mex.Pacific_MaricultureHarvest$OHI_rgn_number)

test<- Mex.Pacific_MaricultureHarvest %>%
  filter(is.na(estimated_genus_name))
table(test$common_name_clean)

# Summarized Production for CONAPESCA cosecha production in all years.
COSECHA_MaricultureHarvest_region <- Mex.Pacific_MaricultureHarvest %>%
 filter(year >= 2018) %>%
  group_by(estimated_genus_name, year, OHI_rgn_number) %>%
  summarise(total_est_production_mt = (sum(weight_kg_corrected, na.rm = TRUE)/1000),
            .groups = "drop")

# Summarized Production for FAO Aquaculture (2018)
```

## There are taxa that cannot be identified to the species, geneus, or even family level. These will need to be dealt with for final aquaculture estimates.
```{r Review PENDING Aquaculture key}
AquacultureFamily_key0<- read.csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/AquacultureFamilyKey_INCOMPLETE_Apr042026.csv", 
                           header= TRUE)
```

```{r Load in TEMPORARY MAR weights}
MAR_Weights_TEMPORARY0<- read.csv("/home/ferrer/ohi_goc_prep/FP/MAR_Scores/MAR_WeightingScores_TEMPORARY_Apr282026.csv", 
                           header= TRUE)

colnames(MAR_Weights_TEMPORARY0)

MAR_Weights_TEMPORARY<- MAR_Weights_TEMPORARY0 %>%
  select(Family, estimated_genus_name= Genus, Estimated_MAR_Score, MAR_Est_Reasoning)
```

```{r Merge COSECHA by year region genus to MAR Weights by genus}
TEMPORARY_MAR_df <- COSECHA_MaricultureHarvest_region %>%
  left_join(
    MAR_Weights_TEMPORARY,
    by = c("estimated_genus_name")
  )

TEMPORARY_MAR_df<- TEMPORARY_MAR_df %>%
  filter(estimated_genus_name %in% MaricultureGenera_of_Interest) %>%
  filter(year < 2025)

colnames(TEMPORARY_MAR_df)

TEMPORARY_MAR_df<- TEMPORARY_MAR_df %>%
  select(OHI_rgn_number, year, Family, 
          total_est_production_mt, Estimated_MAR_Score, MAR_Est_Reasoning)

# Scores dataframe:
TEMPORARY_MAR_scores <- TEMPORARY_MAR_df %>%
  group_by(OHI_rgn_number, year) %>%
  summarise(
    total_production = sum(total_est_production_mt, na.rm = TRUE),
    weighted_MAR = sum(total_est_production_mt * Estimated_MAR_Score, na.rm = TRUE) / total_production,
    .groups = "drop"
  )

# Exploratory plots, proof of concept:

# OHI Region = BCS_Region_3_4_TBD
TEMPORARY_MAR_df %>%
  filter(OHI_rgn_number == "BCS_Region_3_4_TBD") %>%
ggplot(aes(x = year, y = total_est_production_mt, color = Family)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Mariculture Production over time in BCS_Region_3_4_TBD",
    x = "Year",
    y = "Total Production (mt)",
    color = "Family"
  ) +
  theme_minimal()

TEMPORARY_MAR_scores %>%
  filter(OHI_rgn_number == "BCS_Region_3_4_TBD") %>%
ggplot(aes(x = year, y = weighted_MAR)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Mariculture Scores over time in BCS_Region_3_4_TBD",
    x = "Year",
    y = "MAR score"
  ) +
  theme_minimal()

# OHI Region = Region_9 (Lower Baja and La Paz area)
TEMPORARY_MAR_df %>%
  filter(OHI_rgn_number == "Region_9") %>%
ggplot(aes(x = year, y = total_est_production_mt, color = Family)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Mariculture Production over time in Region_9",
    x = "Year",
    y = "Total Production (mt)",
    color = "Family"
  ) +
  theme_minimal()

TEMPORARY_MAR_scores %>%
  filter(OHI_rgn_number == "Region_9") %>%
ggplot(aes(x = year, y = weighted_MAR)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Mariculture Scores over time in Region_9",
    x = "Year",
    y = "MAR score"
  ) +
  theme_minimal()

# Facet wrap:
TEMPORARY_MAR_df %>%
  ggplot(aes(x = year, y = total_est_production_mt, color = Family)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ OHI_rgn_number, scales = "free_y") +
  labs(
    title = "Mariculture Production over time by region",
    x = "Year",
    y = "Total Production (mt)",
    color = "Family"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )


# Facet wrap:
TEMPORARY_MAR_df %>%
  ggplot(aes(x = year, y = total_est_production_mt, color = Family)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ OHI_rgn_number, scales = "free_y") +
  labs(
    title = "Mariculture Production over time by region",
    x = "Year",
    y = "Total Production (mt)",
    color = "Family"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
