################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```

# Load in MAR dataframe (data as of Dec 09 2025)
```{r Load MAR dataframe}
# Load file, only if needed:
Mariculture_Landings<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/mar/v2025/int/GOC_MAR.landings_CLEAN_as.of.Feb242026.rds"))
# 130432 obs., 50 vars.
```

```{r Investigate the mariculture flags}
table(Mariculture_Landings$common_name_clean)
table(Mariculture_Landings$aquaculture_flag)
table(Mariculture_Landings$acuaculture_production)
table(Mariculture_Landings$fleet) # lots of NAs

subset<- Mariculture_Landings %>% filter(acuaculture_production== c("NO", "no"))
# 7822 obs.

unique(subset$common_name_clean)
# NOTE TO SELF: Many types of ostion flagged in this subset as "NOT" aquaculture production, which is at odds with what we've been told...

subset<- Mariculture_Landings %>% filter(common_name_clean== c("tiburon", "tiburon azul", "tiburon tunero", "tiburon sedoso")) 
# NOTE TO SELF: Also weird, several shark entries flagged as aquaculture, even though we know this all but impossible.
```

```{r Top Mariculture Sp.}
table(Mariculture_Landings$common_name_clean)
table(Mariculture_Landings$aquaculture_flag)

Mariculture_Landings_Summary<- Mariculture_Landings %>% 
  filter(year >= 2014, year <= 2024) %>%
  select(state, recorded_taxa_group, recorded_taxa_name_clean, weight_kg_corrected, value) %>%
  group_by(state, recorded_taxa_group, recorded_taxa_name_clean) %>%
  summarize(total_catch_2014.2024= sum(weight_kg_corrected, na.rm = TRUE),
            total_value_2014.2024= sum(value, na.rm = TRUE))

unique(Mariculture_Landings_Summary$recorded_taxa_name_clean)

Top_MaricultureSp <- Mariculture_Landings_Summary %>%
  filter(total_catch_2014.2024 >= 10000) %>%
  pull(recorded_taxa_name_clean) %>%
  unique()

Top_MaricultureSp # NOTE TO SELF: 33 top species. Some of them are fairly general or could be freshwater.

Top_MaricultureSp_Summary<- Mariculture_Landings %>% 
  filter(year >= 2014, year <= 2024) %>%
  filter(recorded_taxa_name_clean %in% Top_MaricultureSp) %>%
  select(recorded_taxa_group, recorded_taxa_name_clean, weight_kg_corrected, value) %>%
  group_by(recorded_taxa_group, recorded_taxa_name_clean) %>%
  summarize(total_catch_2014.2024= sum(weight_kg_corrected, na.rm = TRUE),
            total_value_2014.2024= sum(value, na.rm = TRUE))
```

```{r Save Landings_GoC_TopMariculture_Summary to find MAR sustainability scores}
write.csv(Top_MaricultureSp_Summary, "/home/ferrer/ohi_goc_prep/FP/MAR_research/Mariculture_TopCatches_2014.2024_Feb242026.csv", row.names = FALSE)
```

```{r Mariculture Key}
# UNDER CONSTRUCTION (I did this manually, Feb 24 2026)
Mariculture_GroupKey<- read.csv("~/ohi_goc_prep/FP/MAR_research/Mariculture_GroupKey_UNDERCONSTRUCTION_Feb242026.csv")
```

```{r Merge and exclude}
Mariculture_Landings <- Mariculture_Landings %>%
  left_join(Mariculture_GroupKey,
            by = c("recorded_taxa_group", "recorded_taxa_name_clean"))

Mariculture_Landings<- Mariculture_Landings %>% filter(!MAR_Key_Manual=="Exclude")

Mariculture_Landings_Summary<- Mariculture_Landings %>% 
  filter(year >= 2014, year <= 2024) %>%
  select(year, MAR_Key_Manual, weight_kg_corrected, value) %>%
  group_by(MAR_Key_Manual, year) %>%
  summarize(annual_catch= sum(weight_kg_corrected, na.rm = TRUE),
            annual_value= sum(value, na.rm = TRUE))
```

```{r PLOT}
# Plot annual catch over time
ggplot(Mariculture_Landings_Summary, 
       aes(x = year, y = log10(annual_catch), color = MAR_Key_Manual, group = MAR_Key_Manual)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::comma) +  # Adds commas to y-axis
  labs(
    title = "Annual Catch Over Time by MAR_Key_Manual",
    x = "Year",
    y = "Annual Catch (tons)",
    color = "Mariculture Category"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    text = element_text(size = 12)
  )
```


# SUMMARY STATS ABOUT OUR DATA: return to this later.
```{r How much of biomass includes information about species, genus, family?}
Landings_GoC %>%
  pull(estimated_species_name) %>%
  {c(n_na = sum(is.na(.)),
      n_unknown = sum(. == "unknown", na.rm = TRUE),
      n_NEI = sum(. == "NEI", na.rm = TRUE)
    ) }
# 1008085 (1839 + 205110 + 765582) out of 3559463 records (27%) for species are NA, "unknown", or "NEI"
# Meaning we have species info for 73% of records.

Landings_GoC %>%
  pull(estimated_genus_name) %>%
  {c(n_na = sum(is.na(.)),
     n_unknown = sum(. == "unknown", na.rm = TRUE),
     n_NEI = sum(. == "NEI", na.rm = TRUE)
    ) }
# 1008085 (206949 + 0 + 765582) out of 3559463 records (27%) for genus are NA, "unknown", or "NEI"
# Meaning we have genus info for 73% of records.

Landings_GoC %>%
  pull(estimated_family_name) %>%
  {c(n_na = sum(is.na(.)),
     n_unknown = sum(. == "unknown", na.rm = TRUE),
     n_NEI = sum(. == "NEI", na.rm = TRUE),
     n_family_info_pending = sum(. == "family missing but genus available", na.rm = TRUE)
    ) }
# 336348 (185825 + 0 + 150523) out of 3559463 records (9%) for family are NA, "unknown", or "NEI". 29438 (~1%) records have information about genus but family info pending.
# Meaning we already have or can get family info for 91% of records.
```


# Mariculture work as of APRIL 14 2026:

# STEP 13: Generate MARICULTURE List  ------------------------------

```{r Load GOC_Landings_marine if needed}
GOC_Landings_marine<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Landings_MARINE_2000.present_EMF.Apr032026.rds"))
# 3312602 obs., 62 vars.
```

```{r FILTER OUT MARICULTURE SECTORS OF INTEREST:}
table(GOC_Landings_marine$sector_estimate_draft)

# FILTER OUT MARICULTURE SECTORS OF INTEREST: ----
# Let's assume that the records are most accurate since 2018.

Mariculture_Sectors<- c("marine: aquaculture")

GOC_MaricultureProduction<- GOC_Landings_marine %>% 
  filter(sector_estimate_draft %in% Mariculture_Sectors)
# 105932 obs., 62 vars.

range(GOC_MaricultureProduction$year)
```

```{r Look into this first-draft list before proceeding}
# Taking the information I've gleaned here to refine my sector estimates above.
table(GOC_MaricultureProduction$estimated_genus_name)
table(GOC_MaricultureProduction$Major_Functional_Group)
table(GOC_MaricultureProduction$estimated_genus_name)
```

```{r Generate a table / list of common n > 10 records aquaculture species since 2018}
GOC_MajorAquacultureGenera_summary <- GOC_MaricultureProduction %>%
  filter(year > 2017) %>%
  filter(!is.na(estimated_genus_name) & estimated_genus_name != "NEI" & estimated_genus_name != "unknown") %>%
  select(estimated_genus_name, Major_Functional_Group) %>%
  group_by(Major_Functional_Group, estimated_genus_name) %>%
  summarise(
    n_records_since2018 = n(),
    .groups = "drop"
  )

# Filter out the most common genera:
GOC_MajorAquacultureGenera_summary<- GOC_MajorAquacultureGenera_summary %>%
  filter(n_records_since2018 > 10)
## 11 obs. (down from 15)

AquacultureGenera_of_Interest<- unique(GOC_MajorAquacultureGenera_summary$estimated_genus_name)
```

```{r Investigate records for unexpected taxa}
table(GOC_MajorAquacultureGenera_summary$estimated_genus_name)
table(GOC_MaricultureProduction$reported_aquaculture_production) # is no = 1053
sum(is.na(GOC_MaricultureProduction$reported_aquaculture_production)) # is na = 104204

# Investigate weird entries
subset1<- GOC_MaricultureProduction %>%
  filter(year > 2017) %>%
  filter(estimated_genus_name == "Lutjanus") 
table(subset1$eu_name)
# Makes sense! All records from Ocean Farms in La Paz.

subset2<- GOC_MaricultureProduction %>%
  filter(year > 2017) %>%
  filter(estimated_genus_name == "Argopecten") 
table(subset2$keywords_1)
# Makes sense! All records say "de cultivo" explicitly.

subset4<- GOC_MaricultureProduction %>%
  filter(year > 2017) %>%
  filter(estimated_genus_name == "Dosinia") 
table(subset4$keywords_1)
# Makes sense! All records say "de cultivo" explicitly.

subset5<- GOC_MaricultureProduction %>%
  filter(year > 2017) %>%
  filter(estimated_genus_name == "Megapitaria") 
table(subset5$keywords_1)
# Makes sense! All records say "de cultivo" explicitly.
```

```{r Summarize AquacultureGenera_of_Interest over time}
colnames(GOC_MaricultureProduction)

GOC_AquacultureProduction_summary <- GOC_MaricultureProduction %>%
  filter(estimated_genus_name %in% AquacultureGenera_of_Interest) %>%
  select(year, estimated_genus_name, weight_kg_corrected, value) %>%
  group_by(estimated_genus_name, year) %>%
  summarise(
    annual_production_mt = sum(weight_kg_corrected, na.rm = TRUE) / 1000,
    annual_value_pesos   = sum(value, na.rm = TRUE),
    .groups = "drop"
  )
# 97 obs., 4 vars.
```

# Compare to FAO aquaculture data [UNDER CONSTRUCTION Apr 4 2026]

```{r Load in FAO_FamilyGenus_key_filtered}
FAO_FamilyGenus_key0<- read.csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/FAO_FamilyGenusKey_COMPLETE_Apr042026.csv")
# 193 obs., 7 vars.

# Trim key and simplify (as needed):
colnames(FAO_FamilyGenus_key0)
FAO_FamilyGenus_key_filtered<- FAO_FamilyGenus_key0 %>%
  select(estimated_genus_name, Major_Functional_Group, 
         estimated_family_name= new_estimated_family_name) %>%
  distinct(estimated_family_name, .keep_all = TRUE)
# 105 obs., 3 vars.
```

```{r Merge FAO_FamilyGenus_key_filtered to GOC_AquacultureProduction_summary}
# Make estimated_genus_name lowercase so the merge goes smoothly:
FAO_FamilyGenus_key_filtered$estimated_genus_name <- tolower(FAO_FamilyGenus_key_filtered$estimated_genus_name)

GOC_AquacultureProduction_summary$estimated_genus_name <- tolower(GOC_AquacultureProduction_summary$estimated_genus_name)

# Merge in key:
GOC_AquacultureProduction_summary <- left_join(GOC_AquacultureProduction_summary,
                                               FAO_FamilyGenus_key_filtered, 
                           by = "estimated_genus_name")
# 97 obs., 6 vars.
```

```{r Look into FAO aquaculture data}
# Load data
FAO_Landings_Data0<- read.csv('/home/ferrer/ohi_goc_prep/FP/CMSY_Analysis/CMSY_Revised_April2026/Generating_Model_Priors/Priors_Data/FAOMexicoData_19762023_downloadedApr042026.csv')
# FAO timeseries data organized with Stock (taxa) as # columns and years as # rows.
    
FAO_Landings_Data<- FAO_Landings_Data0

# We've got to do a little work on these data before they're ready to use.
colnames(FAO_Landings_Data)

# Clean column names
clean_names <- colnames(FAO_Landings_Data) %>%
  # Remove trailing periods
  str_replace_all("\\.$", "") %>%
  # Replace multiple periods with a single underscore
  str_replace_all("\\.+", "_") %>%
  # Remove leading X_ for years
  str_replace("^X_(\\d+)$", "\\1") %>%
  # Convert to lowercase
  tolower()

# Apply cleaned names to the dataframe
colnames(FAO_Landings_Data) <- clean_names

# Delete the columns that start with s or s_, we don't need these.
FAO_Landings_Data <- FAO_Landings_Data %>%
  select(-starts_with("s"))

# Filter out aquaculture production:
table(FAO_Landings_Data$detailed_production_source_name)

FAO_Landings_filtered<- FAO_Landings_Data %>%
  filter(detailed_production_source_name %in% c("Aquaculture production (marine)", 
                                                "Aquaculture production (brackishwater)"))

# Delete other columns we don't need:
colnames(FAO_Landings_filtered)
FAO_Landings_filtered<- FAO_Landings_filtered %>%
  select(-c(country_name, fao_major_fishing_area_name, detailed_production_source_name,
            unit_name, unit))

# Transpose these data such that years are rows per taxa (and rename species column):
FAO_AquacultureProduction <- FAO_Landings_filtered %>%
  # Rename species column
  rename(estimated_family_name = asfis_species_scientific_name) %>%
  
  # Step 1: go long
  pivot_longer(
    cols = -estimated_family_name,
    names_to = "year",
    values_to = "production_mt"
  ) %>%
  
  # Step 2: clean year
  mutate(year = as.integer(year),
         production_mt = as.numeric(production_mt))

FAO_AquacultureProduction <- FAO_AquacultureProduction %>%
  arrange(year) %>%
  filter(year >= min(year[production_mt != 0], na.rm = TRUE))

# How many families are there in this database?
unique(FAO_AquacultureProduction$estimated_family_name)
# 24 family, excluding "others"
```

```{r Keep workspace clean}
rm(FAO_Landings_Data)
rm(FAO_Landings_filtered)
```

```{r Load in FAO_FamilyGenus_key_filtered}
FAO_FamilyGenus_key0<- read.csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/FAO_FamilyGenusKey_COMPLETE_Apr042026.csv")
# 193 obs., 7 vars.

# Trim key and simplify (as needed):
colnames(FAO_FamilyGenus_key0)
FAO_FamilyGenus_key_filtered<- FAO_FamilyGenus_key0 %>%
  select(Major_Functional_Group, 
         estimated_family_name= new_estimated_family_name) %>%
  distinct(estimated_family_name, .keep_all = TRUE)
# 105 obs., 2 vars.
```

```{r Merge FAO_FamilyGenus_key_filtered to FAO data}
# Make family lowercase so the merge goes smoothly:
FAO_FamilyGenus_key_filtered$estimated_family_name <- tolower(FAO_FamilyGenus_key_filtered$estimated_family_name)

FAO_AquacultureProduction$estimated_family_name <- tolower(FAO_AquacultureProduction$estimated_family_name)

# Merge in key:
FAO_AquacultureProduction <- left_join(FAO_AquacultureProduction, FAO_FamilyGenus_key_filtered, 
                           by = "estimated_family_name")

```

```{r Plot FAO aquaculture}
ggplot(FAO_AquacultureProduction, aes(x = year, y = production_mt)) +
  geom_line(color = "steelblue") +
  facet_wrap(~ estimated_family_name, scales = "free_y") +
  labs(x = "Year", y = "Production (mt)",
       title = "FAO Aquaculture Production (by family)") +
  theme_minimal()
```

```{r Plot CONAPESCA aquaculture}
GOC_AquacultureProduction_summary %>%
  filter(year < 2025) %>%
  filter(estimated_genus_name=="Penaeus") %>%
ggplot(aes(x = year, y = annual_production_mt)) +
  geom_line(color = "tomato") +
  labs(x = "Year", y = "Production (mt)",
       title = "CONAPESCA Aquaculture Production of Shrimp (Penaeus)") +
  theme_minimal()

GOC_AquacultureProduction_summary %>%
ggplot(aes(x = year, y = annual_production_mt)) +
  geom_line(color = "tomato") +
  facet_wrap(~ estimated_genus_name, scales = "free_y") +
  labs(x = "Year", y = "Production (mt)",
       title = "CONAPESCA Aquaculture Production by genus") +
  theme_minimal()
```

The CONAPESCA aquaculture data look super weird.
We might want to go with FAO data in this case. 
But what is we summarize CONAPESCA data at a higher level?

# Different approach [UNDER CONSTRUCTION Apr 4 2026]
```{r We'll want to add family info to the CONAPESCA data based on genus}
AquacultureFamily_key0<- read.csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/AquacultureFamilyKey_INCOMPLETE_Apr042026.csv")
# 113 obs., 7 vars.

# Trim key and simplify (as needed):
colnames(AquacultureFamily_key0)
AquacultureFamily_key_filtered<- AquacultureFamily_key0 %>%
  select(estimated_genus_name, estimated_taxa_group, Major_Functional_Group,
         new_estimated_family_name) %>%
  distinct()

# if estimated_genus_name is.na, "NEI", or "unknown" & estimated_taxa_group = "OTRAS" then Major_Functional_Group should be NA, and new_estimated_family_name should be NA

AquacultureFamily_key_filtered <- AquacultureFamily_key0 %>%
  # Select only relevant columns
  select(
    estimated_genus_name, 
    estimated_taxa_group, 
    Major_Functional_Group,
    new_estimated_family_name
  ) %>%
  # Keep only unique rows
  distinct() %>%
  # Apply rules for placeholder genera
  mutate(
    # Rule 1: if estimated_genus_name is NA, "NEI", or "unknown", new_estimated_family_name = NA
    new_estimated_family_name = if_else(
      is.na(estimated_genus_name) | estimated_genus_name %in% c("NEI", "unknown"),
      NA_character_,
      new_estimated_family_name
    ),
    # Rule 2: if estimated_genus_name is NA, "NEI", or "unknown" AND estimated_taxa_group = "OTRAS", Major_Functional_Group = NA
    Major_Functional_Group = if_else(
      (is.na(estimated_genus_name) | estimated_genus_name %in% c("NEI", "unknown")) & 
        estimated_taxa_group == "OTRAS",
      NA_character_,
      Major_Functional_Group
    )
  ) %>%
  distinct()
# 92 obs., 3 vars.

AquacultureFamily_key_filtered<- AquacultureFamily_key_filtered %>%
  select(-Major_Functional_Group) %>%
  distinct()

# table(AquacultureFamily_key_filtered$estimated_genus_name)
# No genera duplicates. Looks good!
```

```{r Merge AquacultureFamily_key_filtered to GOC_AquacultureProduction_summary}
# Make estimated_genus_name lowercase so the merge goes smoothly:
AquacultureFamily_key_filtered$estimated_genus_name <- tolower(AquacultureFamily_key_filtered$estimated_genus_name)

GOC_MaricultureProduction$estimated_genus_name <- tolower(GOC_MaricultureProduction$estimated_genus_name)

GOC_MaricultureProduction$estimated_family_name <- tolower(GOC_MaricultureProduction$estimated_family_name)

# Merge in key:
GOC_MaricultureProduction <- left_join(GOC_MaricultureProduction, AquacultureFamily_key_filtered, 
                           by = c("estimated_genus_name", "estimated_taxa_group"))
# 105932 obs., 63 vars.
```

```{r Take a quick look}
TEST <- GOC_MaricultureProduction %>%
  filter(year > 2017) %>%
  select(new_estimated_family_name, Major_Functional_Group) %>%
  group_by(Major_Functional_Group, new_estimated_family_name) %>%
  summarise(
    n_records_since2018 = n(),
    .groups = "drop"
  )
```

```{r Let's just look at Penaeidae for now}
GOC_MaricultureProduction$new_estimated_family_name <- tolower(GOC_MaricultureProduction$new_estimated_family_name)

GOC_Penaeidae<- GOC_MaricultureProduction %>%
  filter(new_estimated_family_name== "penaeidae")
# 83046 obs.
range(GOC_Penaeidae$year)
# From 2006 to 2025

colnames(GOC_Penaeidae)

GOC_Penaeidae <- GOC_Penaeidae %>%
  # Select and rename columns
  select(
    estimated_family_name = new_estimated_family_name, 
    year,
    weight_kg_corrected, 
    Major_Functional_Group
  ) %>%
  # Create new column for production in metric tons
  mutate(
    production_mt = weight_kg_corrected / 1000,
    # Add source column with constant value
    source = "CONAPESCA"
  )

GOC_Penaeidae<- GOC_Penaeidae %>%
  select(-c(weight_kg_corrected, Major_Functional_Group))

FAO_Penaeidae<- FAO_AquacultureProduction %>%
    filter(estimated_family_name== "penaeidae")
range(FAO_Penaeidae$year)
# From 2006 to 2025

FAO_Penaeidae<- FAO_Penaeidae %>%
  mutate(
    source = "FAO"
  ) %>%
  select(-c(Major_Functional_Group))
```

```{r Join and plot}
# Combine the dataframes
combined_df <- bind_rows(GOC_Penaeidae, FAO_Penaeidae)

# Quick check
head(combined_df)

# Plot production over years, color by source
ggplot(combined_df, aes(x = year, y = production_mt, color = source)) +
  geom_line() +      # line plot
  geom_point() +     # add points
  labs(
    title = "Penaeidae Production Over Years",
    x = "Year",
    y = "Production (mt)"
  ) +
  theme_minimal()
```

# PRE-WORK ------------------------------

```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```

## STEP 1: Load the NW Mexico landings appended with lat-long info:
```{r Load CONAPESCA data all Mexican Pacific, NW.Mex_Landings}
NW.Mex_Landings<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/NWMex_Landings_2000.present_EMF.Apr022026.rds"))
# 4713144 obs., 36 vars.
```

NOTE TO SELF: Need all states from Northwest Mexico! - Apr 14 2026
