################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```

# Load in MAR dataframe (data as of Dec 09 2025)
```{r Load MAR dataframe}
# Load file, only if needed:
Mariculture_Landings<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/mar/v2025/int/GOC_MAR.landings_CLEAN_as.of.Feb242026.rds"))
# 130432 obs., 50 vars.
```

```{r Investigate the mariculture flags}
table(Mariculture_Landings$common_name_clean)
table(Mariculture_Landings$aquaculture_flag)
table(Mariculture_Landings$acuaculture_production)
table(Mariculture_Landings$fleet) # lots of NAs

subset<- Mariculture_Landings %>% filter(acuaculture_production== c("NO", "no"))
# 7822 obs.

unique(subset$common_name_clean)
# NOTE TO SELF: Many types of ostion flagged in this subset as "NOT" aquaculture production, which is at odds with what we've been told...

subset<- Mariculture_Landings %>% filter(common_name_clean== c("tiburon", "tiburon azul", "tiburon tunero", "tiburon sedoso")) 
# NOTE TO SELF: Also weird, several shark entries flagged as aquaculture, even though we know this all but impossible.
```

```{r Top Mariculture Sp.}
table(Mariculture_Landings$common_name_clean)
table(Mariculture_Landings$aquaculture_flag)

Mariculture_Landings_Summary<- Mariculture_Landings %>% 
  filter(year >= 2014, year <= 2024) %>%
  select(state, recorded_taxa_group, recorded_taxa_name_clean, weight_kg_corrected, value) %>%
  group_by(state, recorded_taxa_group, recorded_taxa_name_clean) %>%
  summarize(total_catch_2014.2024= sum(weight_kg_corrected, na.rm = TRUE),
            total_value_2014.2024= sum(value, na.rm = TRUE))

unique(Mariculture_Landings_Summary$recorded_taxa_name_clean)

Top_MaricultureSp <- Mariculture_Landings_Summary %>%
  filter(total_catch_2014.2024 >= 10000) %>%
  pull(recorded_taxa_name_clean) %>%
  unique()

Top_MaricultureSp # NOTE TO SELF: 33 top species. Some of them are fairly general or could be freshwater.

Top_MaricultureSp_Summary<- Mariculture_Landings %>% 
  filter(year >= 2014, year <= 2024) %>%
  filter(recorded_taxa_name_clean %in% Top_MaricultureSp) %>%
  select(recorded_taxa_group, recorded_taxa_name_clean, weight_kg_corrected, value) %>%
  group_by(recorded_taxa_group, recorded_taxa_name_clean) %>%
  summarize(total_catch_2014.2024= sum(weight_kg_corrected, na.rm = TRUE),
            total_value_2014.2024= sum(value, na.rm = TRUE))
```

```{r Save Landings_GoC_TopMariculture_Summary to find MAR sustainability scores}
write.csv(Top_MaricultureSp_Summary, "/home/ferrer/ohi_goc_prep/FP/MAR_research/Mariculture_TopCatches_2014.2024_Feb242026.csv", row.names = FALSE)
```

```{r Mariculture Key}
# UNDER CONSTRUCTION (I did this manually, Feb 24 2026)
Mariculture_GroupKey<- read.csv("~/ohi_goc_prep/FP/MAR_research/Mariculture_GroupKey_UNDERCONSTRUCTION_Feb242026.csv")
```

```{r Merge and exclude}
Mariculture_Landings <- Mariculture_Landings %>%
  left_join(Mariculture_GroupKey,
            by = c("recorded_taxa_group", "recorded_taxa_name_clean"))

Mariculture_Landings<- Mariculture_Landings %>% filter(!MAR_Key_Manual=="Exclude")

Mariculture_Landings_Summary<- Mariculture_Landings %>% 
  filter(year >= 2014, year <= 2024) %>%
  select(year, MAR_Key_Manual, weight_kg_corrected, value) %>%
  group_by(MAR_Key_Manual, year) %>%
  summarize(annual_catch= sum(weight_kg_corrected, na.rm = TRUE),
            annual_value= sum(value, na.rm = TRUE))
```

```{r PLOT}
# Plot annual catch over time
ggplot(Mariculture_Landings_Summary, 
       aes(x = year, y = log10(annual_catch), color = MAR_Key_Manual, group = MAR_Key_Manual)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::comma) +  # Adds commas to y-axis
  labs(
    title = "Annual Catch Over Time by MAR_Key_Manual",
    x = "Year",
    y = "Annual Catch (tons)",
    color = "Mariculture Category"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    text = element_text(size = 12)
  )
```


# SUMMARY STATS ABOUT OUR DATA: return to this later.
```{r How much of biomass includes information about species, genus, family?}
Landings_GoC %>%
  pull(estimated_species_name) %>%
  {c(n_na = sum(is.na(.)),
      n_unknown = sum(. == "unknown", na.rm = TRUE),
      n_NEI = sum(. == "NEI", na.rm = TRUE)
    ) }
# 1008085 (1839 + 205110 + 765582) out of 3559463 records (27%) for species are NA, "unknown", or "NEI"
# Meaning we have species info for 73% of records.

Landings_GoC %>%
  pull(estimated_genus_name) %>%
  {c(n_na = sum(is.na(.)),
     n_unknown = sum(. == "unknown", na.rm = TRUE),
     n_NEI = sum(. == "NEI", na.rm = TRUE)
    ) }
# 1008085 (206949 + 0 + 765582) out of 3559463 records (27%) for genus are NA, "unknown", or "NEI"
# Meaning we have genus info for 73% of records.

Landings_GoC %>%
  pull(estimated_family_name) %>%
  {c(n_na = sum(is.na(.)),
     n_unknown = sum(. == "unknown", na.rm = TRUE),
     n_NEI = sum(. == "NEI", na.rm = TRUE),
     n_family_info_pending = sum(. == "family missing but genus available", na.rm = TRUE)
    ) }
# 336348 (185825 + 0 + 150523) out of 3559463 records (9%) for family are NA, "unknown", or "NEI". 29438 (~1%) records have information about genus but family info pending.
# Meaning we already have or can get family info for 91% of records.
```


# Mariculture work as of APRIL 14 2026:

# STEP 13: Generate MARICULTURE List  ------------------------------

```{r Load Mex.Pacific_Landings if needed}
Mex.Pacific_Landings<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/Mex.Pacific_Landings_SectorSorted_2000.present_EMF.Apr152026.rds"))
# 3312602 obs., 62 vars.

Mex.Pacific_Landings_sector<- Mex.Pacific_Landings_sector
rm(Mex.Pacific_Landings_marine)
```

```{r FILTER OUT MARICULTURE SECTORS OF INTEREST:}
table(Mex.Pacific_Landings$sector_estimate_draft)

# FILTER OUT MARICULTURE SECTORS OF INTEREST: ----
# Let's assume that the records are most accurate since 2018.

Mariculture_Sectors<- c("marine: aquaculture")

Mex.Pacific_MaricultureProduction<- Mex.Pacific_Landings %>% 
  filter(sector_estimate_draft %in% Mariculture_Sectors)
# 105932 obs., 62 vars.

range(Mex.Pacific_MaricultureProduction$year)
```

```{r Look into this first-draft list before proceeding}
# Taking the information I've gleaned here to refine my sector estimates above.
table(Mex.Pacific_MaricultureProduction$estimated_genus_name)
table(Mex.Pacific_MaricultureProduction$Major_Functional_Group)
table(Mex.Pacific_MaricultureProduction$estimated_genus_name)
```

```{r Generate a table / list of common n > 10 records aquaculture species since 2018}
Mex.Pacific_MajorAquacultureGenera_summary <- Mex.Pacific_MaricultureProduction %>%
  filter(year > 2017) %>%
  filter(!is.na(estimated_genus_name) & estimated_genus_name != "NEI" & estimated_genus_name != "unknown") %>%
  select(estimated_genus_name, Major_Functional_Group) %>%
  group_by(Major_Functional_Group, estimated_genus_name) %>%
  summarise(
    n_records_since2018 = n(),
    .groups = "drop"
  )

# Filter out the most common genera:
Mex.Pacific_MajorAquacultureGenera_summary<- Mex.Pacific_MajorAquacultureGenera_summary %>%
  filter(n_records_since2018 > 10)
## 11 obs. (down from 15)

AquacultureGenera_of_Interest<- unique(Mex.Pacific_MajorAquacultureGenera_summary$estimated_genus_name)
```

```{r Investigate records for unexpected taxa}
table(Mex.Pacific_MajorAquacultureGenera_summary$estimated_genus_name)
table(Mex.Pacific_MaricultureProduction$reported_aquaculture_production) # is no = 1053
sum(is.na(Mex.Pacific_MaricultureProduction$reported_aquaculture_production)) # is na = 104204

# Investigate weird entries
subset1<- Mex.Pacific_MaricultureProduction %>%
  filter(year > 2017) %>%
  filter(estimated_genus_name == "Lutjanus") 
table(subset1$eu_name)
# Makes sense! All records from Ocean Farms in La Paz.

subset2<- Mex.Pacific_MaricultureProduction %>%
  filter(year > 2017) %>%
  filter(estimated_genus_name == "Argopecten") 
table(subset2$keywords_1)
# Makes sense! All records say "de cultivo" explicitly.

subset4<- Mex.Pacific_MaricultureProduction %>%
  filter(year > 2017) %>%
  filter(estimated_genus_name == "Dosinia") 
table(subset4$keywords_1)
# Makes sense! All records say "de cultivo" explicitly.

subset5<- Mex.Pacific_MaricultureProduction %>%
  filter(year > 2017) %>%
  filter(estimated_genus_name == "Megapitaria") 
table(subset5$keywords_1)
# Makes sense! All records say "de cultivo" explicitly.
```

```{r Summarize AquacultureGenera_of_Interest over time}
colnames(Mex.Pacific_MaricultureProduction)

GOC_AquacultureProduction_summary <- Mex.Pacific_MaricultureProduction %>%
  filter(estimated_genus_name %in% AquacultureGenera_of_Interest) %>%
  select(year, estimated_genus_name, weight_kg_corrected, value) %>%
  group_by(estimated_genus_name, year) %>%
  summarise(
    annual_production_mt = sum(weight_kg_corrected, na.rm = TRUE) / 1000,
    annual_value_pesos   = sum(value, na.rm = TRUE),
    .groups = "drop"
  )
# 97 obs., 4 vars.
```



```{r Plot CONAPESCA aquaculture}
GOC_AquacultureProduction_summary %>%
  filter(year < 2025) %>%
  filter(estimated_genus_name=="Penaeus") %>%
ggplot(aes(x = year, y = annual_production_mt)) +
  geom_line(color = "tomato") +
  labs(x = "Year", y = "Production (mt)",
       title = "CONAPESCA Aquaculture Production of Shrimp (Penaeus)") +
  theme_minimal()

GOC_AquacultureProduction_summary %>%
ggplot(aes(x = year, y = annual_production_mt)) +
  geom_line(color = "tomato") +
  facet_wrap(~ estimated_genus_name, scales = "free_y") +
  labs(x = "Year", y = "Production (mt)",
       title = "CONAPESCA Aquaculture Production by genus") +
  theme_minimal()
```

The CONAPESCA aquaculture data look super weird.
We might want to go with FAO data in this case. 
But what is we summarize CONAPESCA data at a higher level?

# Different approach [UNDER CONSTRUCTION Apr 4 2026]
```{r We'll want to add family info to the CONAPESCA data based on genus}
AquacultureFamily_key0<- read.csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/AquacultureFamilyKey_INCOMPLETE_Apr042026.csv")
# 113 obs., 7 vars.

# Trim key and simplify (as needed):
colnames(AquacultureFamily_key0)
AquacultureFamily_key_filtered<- AquacultureFamily_key0 %>%
  select(estimated_genus_name, estimated_taxa_group, Major_Functional_Group,
         new_estimated_family_name) %>%
  distinct()

# if estimated_genus_name is.na, "NEI", or "unknown" & estimated_taxa_group = "OTRAS" then Major_Functional_Group should be NA, and new_estimated_family_name should be NA

AquacultureFamily_key_filtered <- AquacultureFamily_key0 %>%
  # Select only relevant columns
  select(
    estimated_genus_name, 
    estimated_taxa_group, 
    Major_Functional_Group,
    new_estimated_family_name
  ) %>%
  # Keep only unique rows
  distinct() %>%
  # Apply rules for placeholder genera
  mutate(
    # Rule 1: if estimated_genus_name is NA, "NEI", or "unknown", new_estimated_family_name = NA
    new_estimated_family_name = if_else(
      is.na(estimated_genus_name) | estimated_genus_name %in% c("NEI", "unknown"),
      NA_character_,
      new_estimated_family_name
    ),
    # Rule 2: if estimated_genus_name is NA, "NEI", or "unknown" AND estimated_taxa_group = "OTRAS", Major_Functional_Group = NA
    Major_Functional_Group = if_else(
      (is.na(estimated_genus_name) | estimated_genus_name %in% c("NEI", "unknown")) & 
        estimated_taxa_group == "OTRAS",
      NA_character_,
      Major_Functional_Group
    )
  ) %>%
  distinct()
# 92 obs., 3 vars.

AquacultureFamily_key_filtered<- AquacultureFamily_key_filtered %>%
  select(-Major_Functional_Group) %>%
  distinct()

# table(AquacultureFamily_key_filtered$estimated_genus_name)
# No genera duplicates. Looks good!
```

```{r Merge AquacultureFamily_key_filtered to GOC_AquacultureProduction_summary}
# Make estimated_genus_name lowercase so the merge goes smoothly:
AquacultureFamily_key_filtered$estimated_genus_name <- tolower(AquacultureFamily_key_filtered$estimated_genus_name)

Mex.Pacific_MaricultureProduction$estimated_genus_name <- tolower(Mex.Pacific_MaricultureProduction$estimated_genus_name)

Mex.Pacific_MaricultureProduction$estimated_family_name <- tolower(Mex.Pacific_MaricultureProduction$estimated_family_name)

# Merge in key:
Mex.Pacific_MaricultureProduction <- left_join(Mex.Pacific_MaricultureProduction, AquacultureFamily_key_filtered, 
                           by = c("estimated_genus_name", "estimated_taxa_group"))
# 105932 obs., 63 vars.
```

```{r Take a quick look}
TEST <- Mex.Pacific_MaricultureProduction %>%
  filter(year > 2017) %>%
  select(new_estimated_family_name, Major_Functional_Group) %>%
  group_by(Major_Functional_Group, new_estimated_family_name) %>%
  summarise(
    n_records_since2018 = n(),
    .groups = "drop"
  )
```

```{r Let's just look at Penaeidae for now}
Mex.Pacific_MaricultureProduction$new_estimated_family_name <- tolower(Mex.Pacific_MaricultureProduction$new_estimated_family_name)

GOC_Penaeidae<- Mex.Pacific_MaricultureProduction %>%
  filter(new_estimated_family_name== "penaeidae")
# 83046 obs.
range(GOC_Penaeidae$year)
# From 2006 to 2025

colnames(GOC_Penaeidae)

GOC_Penaeidae <- GOC_Penaeidae %>%
  # Select and rename columns
  select(
    estimated_family_name = new_estimated_family_name, 
    year,
    weight_kg_corrected, 
    Major_Functional_Group
  ) %>%
  # Create new column for production in metric tons
  mutate(
    production_mt = weight_kg_corrected / 1000,
    # Add source column with constant value
    source = "CONAPESCA"
  )

GOC_Penaeidae<- GOC_Penaeidae %>%
  select(-c(weight_kg_corrected, Major_Functional_Group))

FAO_Penaeidae<- FAO_AquacultureProduction %>%
    filter(estimated_family_name== "penaeidae")
range(FAO_Penaeidae$year)
# From 2006 to 2025

FAO_Penaeidae<- FAO_Penaeidae %>%
  mutate(
    source = "FAO"
  ) %>%
  select(-c(Major_Functional_Group))
```

```{r Join and plot}
# Combine the dataframes
combined_df <- bind_rows(GOC_Penaeidae, FAO_Penaeidae)

# Quick check
head(combined_df)

# Plot production over years, color by source
ggplot(combined_df, aes(x = year, y = production_mt, color = source)) +
  geom_line() +      # line plot
  geom_point() +     # add points
  labs(
    title = "Penaeidae Production Over Years",
    x = "Year",
    y = "Production (mt)"
  ) +
  theme_minimal()
```

# PRE-WORK ------------------------------

```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```

# STEP 1: Load in CONAPESCA RAW data, attend to records from all of Mexico. ------------------------------

## STEP 1a: Work with CONAPESCA **RAW** data (records from all of Mexico):
This very first step is completed in the file shared by Juan Carlos: "03_combine_landings_edited.R" file path: /home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/JC_Scripts_EMF.edited/03_combine_landings_edited.R

## STEP 1b: Work with CONAPESCA **STITCHED** data (records from all of Mexico):

A stitched version of these data is available as one single file: "AllMex_landings_2000_present_EMF_edited.rds" file path: /home/shares/ohi/OHI_GOC/goal_prep/fp/juan_carlos/AllMex_landings_2000_present_EMF_edited.rds

```{r Load in CONAPESCA STITCHED data}
All.Mex_Landings_2000.present<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/fis/v2025/int/juan_carlos/AllMex_landings_2000_present_EMF_edited.rds"))

# n = 12153879 obs., 20 vars.
```


## STEP 1c: Create weight_kg_corrected column (records from all of Mexico):

```{r Landed weight correction}
# Kind of looks like sometimes 'landed weight' is 'live weight' and vice versa. Fix this with weight_kg_corrected:
All.Mex_Landings_2000.present$weight_kg_corrected <- ifelse(
  All.Mex_Landings_2000.present$live_weight >= All.Mex_Landings_2000.present$landed_weight,
  All.Mex_Landings_2000.present$live_weight,
  All.Mex_Landings_2000.present$landed_weight
)
# n = 12153879 obs., 21 vars.
```

```{r Convert state names to UTF-8 text}
table(All.Mex_Landings_2000.present$state)
        # AGUASCALIENTES        BAJA CALIFORNIA    BAJA CALIFORNIA SUR               CAMPECHE                CHIAPAS 
        #           2582                 520154                1171129                1481876                 246989 
        #      CHIHUAHUA CIUDAD DE MÃ\u0089XICO       CIUDAD DE MÉXICO               COAHUILA                 COLIMA 
        #           1444                     10                      8                   3302                 205852 
        #        DURANGO             GUANAJUATO               GUERRERO                HIDALGO                JALISCO 
        #           4230                   3187                 765569                  13275                 230966 
        #         MEXICO              MICHOACAN                MORELOS                NAYARIT             NUEVO LEON 
        #           3121                 125690                   3988                 786570                    155 
        #         OAXACA                 PUEBLA              QUERETARO           QUINTANA ROO        SAN LUIS POTOSI 
        #         349059                   1513                   5549                 118935                  18896 
        #        SINALOA                 SONORA                TABASCO             TAMAULIPAS               TLAXCALA 
        #        1370011                 634314                 868938                1040082                   2652 
        #       VERACRUZ                YUCATAN              ZACATECAS 
        #        1231282                 935106                   7445 

All.Mex_Landings_2000.present$state_fixed <-
  All.Mex_Landings_2000.present$state

All.Mex_Landings_2000.present$state_fixed[
  grepl("CIUDAD DE M", All.Mex_Landings_2000.present$state_fixed)
] <- "CIUDAD DE MÉXICO"

table(All.Mex_Landings_2000.present$state_fixed)
   #   AGUASCALIENTES     BAJA CALIFORNIA BAJA CALIFORNIA SUR            CAMPECHE             CHIAPAS           CHIHUAHUA 
   #             2582              520154             1171129             1481876              246989                1444 
   # CIUDAD DE MÉXICO            COAHUILA              COLIMA             DURANGO          GUANAJUATO            GUERRERO 
   #               18                3302              205852                4230                3187              765569 
   #          HIDALGO             JALISCO              MEXICO           MICHOACAN             MORELOS             NAYARIT 
   #            13275              230966                3121              125690                3988              786570 
   #       NUEVO LEON              OAXACA              PUEBLA           QUERETARO        QUINTANA ROO     SAN LUIS POTOSI 
   #              155              349059                1513                5549              118935               18896 
   #          SINALOA              SONORA             TABASCO          TAMAULIPAS            TLAXCALA            VERACRUZ 
   #          1370011              634314              868938             1040082                2652             1231282 
   #          YUCATAN           ZACATECAS 
   #           935106                7445 

# Looks good!

All.Mex_Landings_2000.present$state <-
  All.Mex_Landings_2000.present$state_fixed
```

NOTE TO SELF: Will want to integrate this patch into the final standardized data paper. - Apr 15 2026

## STEP 1d: optional save...
Pending, if needed.


# STEP 2: Filter down CONAPESCA data to states in the Mexican Pacific.  ------------------------------

```{r Filter out data from Northwest Mexico}
length(unique(All.Mex_Landings_2000.present$species_name)) 
# 2052 unique species in all of Mexico.

# Filter states in the Mexican Pacific, n= 11
States_of_Interest<- c("BAJA CALIFORNIA", "BAJA CALIFORNIA SUR", "SONORA", "SINALOA", "NAYARIT", "JALISCO",
                       "COLIMA", "MICHOACAN", "GUERRERO", "OAXACA", "CHIAPAS")

Mex.Pacific_Landings <- All.Mex_Landings_2000.present %>% filter(state %in% States_of_Interest)
# n = 6406303 obs., 21 vars.

length(unique(Mex.Pacific_Landings$species_name)) 
# 1929 unique species in States of Interest.
```

# STEP 3c: Add lat-long and OHI_region info to landings data ------------------------------
```{r OHI LatLongKey_finalized and RegionKey_pending}
# Final lat-long key (subject to change as Apr 2 2026):
LatLongKey_finalized<- read_csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/Spatial_Keys/Sites_LatLongKey_Apr22026.csv")
# 2046 obs., 13 vars.

# My regional key (temporary, subject to change, UNDER CONSTRUCTION):
RegionKey_pending<- read_csv(here("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/Spatial_Keys/RegionKey_UNDERCONSTRUCTION_Mar102026.csv"))
# 14 obs., 6 vars.
# Previously: RegionKey_UNDERCONSTRUCTION_Dec012025.csv

colnames(RegionKey_pending)
colnames(RegionKey_pending) <- c("state", "OHI_region_draft", "OHI_rgn_number", "OHI_rgn_name", "exclude_rgn", "OHI_rgn_notes")
```

```{r Append lat-long and OHI_region info to landings data}
# Append lat-long info to landings data:
temp_df <- Mex.Pacific_Landings %>%
  left_join(
    LatLongKey_finalized,
    by = c("state", "landing_site")  # matching columns
  )
# 6406303 obs., 33 vars.

table(temp_df$inside_GOC)
# 1090368 "no" not in the GOC.

# Append RegionKey_pending info to landings data:
temp_df <- temp_df %>%
  left_join(
    RegionKey_pending,
    by = c("state", "OHI_region_draft")  # matching columns
  )
# 6406303 obs., 37 vars.

table(temp_df$inside_GOC)
# 1090368 "no" not in the GOC - same as before - as we'd expect!

colnames(temp_df)

# SANITY CHECK: Looking into the exclude_rgn column:
test<- temp_df %>%
  filter(exclude_rgn=="exclude")

table(test$inside_GOC)
table(test$state)
table(test$OHI_rgn_number)
  # BCS_Region_3_4_TBD  Region_1_2_Baja_TBD Region_9_Jalisco_TBD 
  #             184929                39899               156952 


Mex.Pacific_Landings<- temp_df
# 6406303 obs., 37 vars.

# Delete temp_df, not needed anymore.
rm(temp_df)
Mex.Pacific_Landings<- Mex.Pacific_Landings %>%
  select(-c(state_fixed))
# 6406303 obs., 36 vars.
```

## STEP 3d: Save the Mexican Pacific landings appended with lat-long info in a separate .rds file:
```{r Save the Northwest Mexico dataframe: Mar 11 2026}
saveRDS(Mex.Pacific_Landings, file="/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/MexPacific_Landings_2000.present_EMF.Apr152026.rds")
# 6406303 obs., 36 vars.
```

# Step 5a: Load in Mex.Pacific_Landings dataframe
```{r Load Mex.Pacific_Landings if needed}
Mex.Pacific_Landings<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/MexPacific_Landings_2000.present_EMF.Apr152026.rds"))
# 6406303 obs., 36 vars.
```

```{r Filter out keywords from recorded_taxa_name to get recorded_taxa_name_clean}

# PART A. Keyword list [UNDER CONSTRUCTION] -----------------

# NOTE: Keyword list excludes "ALETA" which is handled separately.

Keyword_list <- c(
# Processing state or product form (e.g., frozen, full fish, filleted, beheaded, dried, etc.):
  "S.C.", "S/C", "S",
  "FRESCO", "FRESCA", "FCA", "FCO", "FC",
  "ECO",
  "CONG", "CONGELADO",
  "SECO DE", "SECA", "SECA DE", "SECO", "SECO DE",
  "COCINERO", "COCIDO", "COCIDA", "COCIDAS", "ASADO",
  "ENTERO", "ENTERA", "IND", "ENT",
  "FILETE", "FILETE DE", "ECA", "FILETE DE ECO", "DESHUESADA", "CARNE", "CARNE DE",
  "DESCABEZADO", "DES", "DSC", "DESC", "DESV", "DESVICERADO",

# Animal parts:
  "MANTO", "MANTO DE",
  "PULPA DE", "PULPA FRESCA", "PULPO DE", 
  "POLVO DE",
  "TENTACULOS DE", "TUBO DE", "TUBO",
  "CABEZA DE", "BUCHE", "BUCHE DE", "TRIPA DE", "COLAS DE", "MANOS DE", "MANOS",
  "ANCAS DE", "ANCAS", "ANCA",
  "PIEL DE ECA", "PIEL DE SECA", "PIEL DE", "PIEL",

# Remove ALETA from main keywords; we'll handle these keywords separately:
  #"ALETA", "ALETA ECA DE", "ALETA FCA DE"

# Shells and pearls:
  "CONCHA DE", "CONCHA", "CONCHA NACAR",
  "PERLAS", "MEDIAS PERLAS",

# From where / habitat type:
  "DE ALTA MAR",
  "DE AGUA DULCE",

# Live culture or cultivation techniques:
  "VIVO", "VIVA", "VIVAS", "VIVITA",
  "DE CULTIVO", "DE CULTIVO VIVO", "CULTIVADAS", "CULTIVADOS", # NOTE TO SELF (Ap 27 2026) cultivados is a new add! Regenerate sector estimate.
  "ABASTO RANCHO", "RANCHO",

# Eggs:
  "HUEVA DE",

# Used as bait:
  "CARNADA", "P/CARNADA",

# Other:
  "NO APTO PARA EMPAQUE"
)

# 81 keywords as of Mar 23 2026.

# Order the key words longest first (note that this is necessary for the next steps):
Keyword_list <- Keyword_list[order(nchar(Keyword_list), decreasing = TRUE)]

# Escape regex meta-characters:
Keyword_list_escaped <- stringi::stri_replace_all_regex(
  Keyword_list, "([.|(){}+*?^$\\[\\]\\\\])", "\\\\$1"
)

# Final regex:
Keyword_pattern <- paste0(
  "(?<![A-Za-z0-9])(",
  paste(Keyword_list_escaped, collapse = "|"),
  ")(?![A-Za-z0-9])"
)


# PART B. Flag special "ALETA" phrases (handled differently from other key words) -----------------
special_aleta_cases <- c("BLANCA", "AMARILLA", "CORTA", "AZUL")

remove_aleta_except_special <- function(x) {
  stringi::stri_replace_all_regex(
    x,
    pattern = paste0("\\bALETA\\b(?!\\s+(?:", paste(special_aleta_cases, collapse = "|"), "))"),
    replacement = "",
    vectorize_all = FALSE
  )
}

# NOTE TO SELF: How did I get this list of key words originally? I think I started out with some preliminary version of the separate, parse functionality, quickly spotted the beginnings of a keywords list and then grew the list over time. Come back to this at some point to recreate this early part of the workflow to document it (reproducibility).
```

# STEP 5b: Separate out keywords from the original landing record using the parse regex in PART C:
```{r Separate keywords from landing record based on parse function}

# PART C. MAIN PARSING PIPELINE -----------------
# RUN TIME= ~10 min, optimized but still computationally heavy

Mex.Pacific_Landings_parsed <- Mex.Pacific_Landings %>%
  
  # Rename columns:
  rename(
    recorded_taxa_name = species_name,
    recorded_taxa_group = main_species_group
  ) %>%
  
  # Extract keywords:
  mutate(
    keywords_list = stringi::stri_extract_all_regex(recorded_taxa_name, Keyword_pattern)
  ) %>%
  mutate(keywords_list = map(keywords_list, ~ { length(.x) <- 5; .x })) %>%
  unnest_wider(keywords_list, names_sep = "_") %>%
  rename_with(~ paste0("keywords_", 1:5), starts_with("keywords_list_")) %>%
  
  # Clean recorded_taxa_name:
  mutate(
    recorded_taxa_name_clean = recorded_taxa_name %>%
      remove_aleta_except_special() %>%                      # Remove ALETA unless special
      stringi::stri_replace_all_regex(Keyword_pattern, "") %>% # Remove other keywords
      stringi::stri_replace_all_regex("\\b(Y|DE)\\b", "") %>%   # Remove standalone modifiers like Y / DE
      stringi::stri_replace_all_regex("P\\s*/\\s*", "") %>%       # remove P/ or P /
      stringi::stri_replace_all_regex("[.,/\\\\]+", " ") %>%      # Remove punctuation
      str_squish()
  ) %>%
  
  # Extract scientific name:
  mutate(
    recorded_scientific_name = stringi::stri_extract_first_regex(recorded_taxa_name_clean, "\\(([^)]+)\\)") %>%
      stringi::stri_replace_all_regex("\\(|\\)", ""),
    scientific_name_recorded_T.F = !is.na(recorded_scientific_name),
    recorded_scientific_name = if_else(
      scientific_name_recorded_T.F,
      stringi::stri_trans_totitle(recorded_scientific_name),
      NA_character_
    )
  ) %>%
  
  # Extract common name:
  mutate(
    common_name_clean = stringi::stri_trim_both(
      stringi::stri_replace_first_regex(recorded_taxa_name_clean, "\\s*\\([^)]*\\)", "")
    ),
    common_name_clean = tolower(common_name_clean)
  )
# 6406303 obs., 45 vars.

test<- Mex.Pacific_Landings_parsed %>%
  filter(state=="COLIMA")

# PART D. Trim and reorder columns -----------------
colnames(Mex.Pacific_Landings_parsed)

# Re-name columns, then trim:
Mex.Pacific_Landings_parsed<- Mex.Pacific_Landings_parsed %>%
  rename(est_latlon_FLAG= FLAGS,
         est_lat= lat_final,
         est_lon= lon_final)

Mex.Pacific_Landings_parsed <- Mex.Pacific_Landings_parsed %>%

  select(
    # General info from record:
    state, office_name, landing_site, landing_site_key, 
    est_lat, est_lon, est_latlon_FLAG, inside_GOC, est_habitat,
    year, month,
    receipt_date_fixed, period_start_fixed, period_end_fixed,
    fleet, acuaculture_production, eu_rnpa, eu_name, vessel_rnpa, vessel_name,
    
    # Species info from record:
    recorded_taxa_group, recorded_taxa_name,
    
    # Cleaned and derived names:
    recorded_taxa_name_clean, recorded_scientific_name,
    scientific_name_recorded_T.F, common_name_clean,
    
    # Keywords that have been parsed:
    keywords_1, keywords_2, keywords_3, keywords_4, keywords_5,
    
    # Landing details from record:
    value, landed_weight, live_weight, weight_kg_corrected,
    
    # OHI region and notes:
    OHI_rgn_number, OHI_region_draft ,OHI_rgn_name, OHI_rgn_notes, exclude_rgn
 
  )
# 6406303 obs., 40 vars.

# How many common names are there in this data frame?
length(unique(Mex.Pacific_Landings_parsed$common_name_clean))
length(unique(Mex.Pacific_Landings_parsed$recorded_taxa_group))
# NOW 814 common names (previously 732 in the GOC alone), over 57 taxonomic groups
```

## VIEW ALL KEY WORDS... SANITY CHECKS: Did this work as expected? Does the Keyword_list need more work?
```{r SANITY CHECK 1: Investigating Keywords}
# Extract all unique keywords from keyword columns 1 through 5. Delete exact duplicates. 
Keyword_vector <- Mex.Pacific_Landings_parsed %>%
  select(keywords_1:keywords_5) %>%
  unlist() %>%
  unique() %>%
  na.omit()

# Why is Keyword_vector shorter than main Keyword_list?
setdiff(Keyword_list, Keyword_vector)
# 8 keywords that aren't featured in the parsed data set:
# "FILETE DE ECO", "PIEL DE ECA", "MANTO DE", "TUBO DE", "RANCHO", "BUCHE", "ANCAS", "ECO", "ECA"    

# Example for "TUBO DE"
Mex.Pacific_Landings_parsed %>%
  select(recorded_taxa_name) %>%
  filter(str_detect(recorded_taxa_name, regex("TUBO", ignore_case = TRUE)))

# Example for "RANCHO"
Mex.Pacific_Landings_parsed %>%
  select(recorded_taxa_name) %>%
  filter(str_detect(recorded_taxa_name, regex("RANCHO", ignore_case = TRUE)))

# Longest-first ordering in the regex ensures: multi-word phrases match fully BEFORE matching component words.
# Component words will not appear separately in the extraction. So, for example, "VIVO P/ABASTO RANCHO" gets abstracted as "VIVO" and "ABASTO RANCHO" before "RANCHO" can be extracted.

# >> This is why my unique keyword vector has fewer elements than your list of 81 — some keywords are never extracted as standalone because they are “absorbed” into longer matches.
```

```{r SANITY CHECK 2: Did we properly keywords and modifiers from recorded_taxa_name_clean?}
# Look for "DE" or "P/" as a full word
DE_remaining <- Mex.Pacific_Landings_parsed %>%
  filter(str_detect(recorded_taxa_name_clean, "\\bDE\\b"))
P_remaining <- Mex.Pacific_Landings_parsed %>%
  filter(str_detect(recorded_taxa_name_clean, "\\bP\\b"))

# Count how many rows
nrow(DE_remaining)
nrow(P_remaining)
# Count = 0, good.

# Rows where "NACAR" appears anywhere (case-insensitive)
NACAR_remaining <- Mex.Pacific_Landings_parsed %>%
  filter(str_detect(recorded_taxa_name_clean, regex("NACAR", ignore_case = TRUE)))

# Count how many rows
nrow(NACAR_remaining)
# Count = 0, good.

# Any keyword column containing NACAR (case-insensitive)
NACAR_as_keyword <- Mex.Pacific_Landings_parsed %>%
  filter(
    str_detect(keywords_1, regex("NACAR", ignore_case = TRUE)) |
    str_detect(keywords_2, regex("NACAR", ignore_case = TRUE)) |
    str_detect(keywords_3, regex("NACAR", ignore_case = TRUE)) |
    str_detect(keywords_4, regex("NACAR", ignore_case = TRUE)) |
    str_detect(keywords_5, regex("NACAR", ignore_case = TRUE))
  )

# Count rows
nrow(NACAR_as_keyword)

# Optional: preview
head(NACAR_as_keyword %>% select(recorded_taxa_name, keywords_1:keywords_5))
# Looks great! 

# Remove these dataframes to keep workspace clear:
rm(DE_remaining, P_remaining, NACAR_remaining, NACAR_as_keyword)
```

```{r SANITY CHECK 3: Checking on the unique ALETA cases}
# QUICK SANITY CHECK: did this work for taxa like yellow-fin and blue-fin tuna?
test <- Mex.Pacific_Landings_parsed %>%
  filter(str_detect(recorded_taxa_group, "ATUN"))
# 32444 obs.

table(test$common_name_clean)
#  Yes, parsing seems to have worked!

# Remove this dataframe:
rm(test)
```

# STEP 5c: Save the Mex.Pacific_Landings_parsed dataframe as a separate .rds file - OPTIONAL:
```{r Save this dataframe - landings with keywords: Mar 23 2026}
colnames(Mex.Pacific_Landings_parsed)
# Looks good! Go ahead and save.

saveRDS(Mex.Pacific_Landings_parsed, file="/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/Mex.Pacific_Landings_KeywordsParsed_2000.present_EMF.Apr152026.rds")
# 6406303 obs., 40 vars.
```

# > IF SKIPPING STEP 5 - parsing keywords, then move to: ------------------------------

# STEP 6: Merge the curated Mex.Pacific_Landings_parsed (now ) with the most-up-to-date species key (MasterSpKey_pending). ------------------------------

## STEP 6a: Load in Mex.Pacific_Landings_parsed dataframe (and rename to Mex.Pacific_Landings_cleaned):
```{r Load Mex.Pacific_Landings_parsed if needed}
Mex.Pacific_Landings_parsed<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/Mex.Pacific_Landings_KeywordsParsed_2000.present_EMF.Apr152026.rds"))
# 6406303 obs., 40 vars.
```

```{r rename to Mex.Pacific_Landings_cleaned}
Mex.Pacific_Landings_species<- Mex.Pacific_Landings_parsed

# Remove dataframes (as appropriate) to keep workspace clean:
rm(Mex.Pacific_Landings)
rm(Mex.Pacific_Landings_parsed)
rm(All.Mex_Landings_2000.present)
rm(RegionKey_pending)
rm(LatLongKey_finalized)
```

```{r Load MasterSpKey_pending}
MasterSpKey_pending <- read_csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/Species_Keys/CleanedMaster_SpKey_Mar252026.csv")
# 807 obs of 41 variables
```

```{r Shave down the size of Master_SpKey_pending}
colnames(MasterSpKey_pending)

Master_SpKey_simple<- MasterSpKey_pending %>% 
  select(recorded_taxa_group, recorded_taxa_name_clean, common_name_clean,
         estimated_species_name, species_est_certainty, estimated_genus_name, estimated_family_name,
         IHH_order, IHH_class, IHH_functional_major_groups_sm, IHH_fishery)
# 807 obs., 11 vars.
```

## STEP 6b: Merge Mex.Pacific_Landings_cleaned with Master_SpKey_simple:
```{r Merge GOC landings with Master_SpKey_simple}
Mex.Pacific_Landings_species <- Mex.Pacific_Landings_species %>%
  left_join(
    Master_SpKey_simple,
    by = c(
      "recorded_taxa_group",
      "recorded_taxa_name_clean",
      "common_name_clean"
    )
  )
# n = 6406303 obs. as before, now 48 vars.
```

```{r Trim Mex.Pacific_Landings_species as needed}
colnames(Mex.Pacific_Landings_species)

Mex.Pacific_Landings_species <- Mex.Pacific_Landings_species %>%
  select(
    # Location:
    state, office_name, landing_site, landing_site_key,
    est_lat, est_lon, est_latlon_FLAG, inside_GOC, est_habitat,
    
    # Time:
    year, month, receipt_date_fixed, period_start_fixed, period_end_fixed,
    
    # Fleet and vessel info:
    fleet, acuaculture_production, eu_rnpa, eu_name, vessel_rnpa, vessel_name,
    
    # Taxa, recorded and cleaned names:
    recorded_taxa_group, recorded_taxa_name, recorded_taxa_name_clean,
    recorded_scientific_name, scientific_name_recorded_T.F, common_name_clean,
    
    # Keywords:
    keywords_1, keywords_2, keywords_3, keywords_4, keywords_5,
    
    # Taxa details, estimated:
    estimated_species_name, species_est_certainty, estimated_genus_name, estimated_family_name,
    
    # Catch and value:
    landed_weight, live_weight, weight_kg_corrected, value,
    
    # OHI region details:
    OHI_rgn_number, OHI_region_draft, OHI_rgn_name, OHI_rgn_notes, exclude_rgn,
    
    # Classification (IHH):
    IHH_order, IHH_class, IHH_functional_major_groups_sm, IHH_fishery
  )
```

```{r SANITY CHECK species missing}
test<- Mex.Pacific_Landings_species %>%
  filter(is.na(estimated_species_name))

table(test$recorded_taxa_group)

  #           ABULON              ALGAS             ALMEJA             BESUGO            CAMARON            CARACOL 
  #                1                214               2294                  1                  2                 12 
  #            CARPA             CHARAL              JUREL           LANGOSTA         LANGOSTINO             LOBINA 
  #                1                 13              78540                  5                  2                  1 
  #             MERO            MOJARRA             ORNATO              OTRAS    PECES DE ORNATO      PEPINO DE MAR 
  #               14                  5                  8               1150                  1                  2 
  # RAYA Y SIMILARES              RONCO RUBIA Y VILLAJAIBA             TRUCHA 
  #                6                 57                  3                  2
```


## STEP 6c: Save Mex.Pacific_Landings_species (keywords parsed with species) as separate .rds file - OPTIONAL:
```{r Save this dataframe - GOC landings with keywords with species names as of Mar 11 2026}
saveRDS(Mex.Pacific_Landings_species, file="/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/Mex.Pacific_Landings_Parsed.wSpecies_2000.present_EMF.Apr152026.rds")
# 6406303 obs., now 48 vars.
```


# STEP 7: Merge the Mex.Pacific_Landings_species with the most up-to-date group key (MajorGroupKey_pending) to create Mex.Pacific_Landings_classed. ------------------------------

# NOTE TO SELF: Species key is up-to-date and so is MajorGroupKey_pending as of - Mar 25 2026
To learn more about how this key was created, see: CONAPESCA_CreatingTaxonomicKeys_Mar132026.rmd

```{r Load in GOC data as needed}
Mex.Pacific_Landings_species<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/Mex.Pacific_Landings_Parsed.wSpecies_2000.present_EMF.Apr152026.rds"))
# 6406303 obs., 48 vars.
```

```{r Load in revised MajorGroupKey_pending}
MajorGroupKey_pending<- read_csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/MajorGroupKey_Mar252026.csv")
# 526 obs. 13 vars.
```

```{r Complicated merge landings data with MajorGroupKey_pending}
Special_groups <- c("OTRAS", "ESPECIE GENERICA", "FAUNA", "ORNATO", "PECES DE ORNATO", "NEI")

#STEP 1: Join in non-OTRAS, simple join
# For non-otra taxa, we only want to look up based on the recorded_taxa_group and estimated_genus_name. From that look-up we want to assign Major_Functional_Group, Assignment_for_Major_Functional_Group
NONOTRAS_MajorGroupKey_lookup <- MajorGroupKey_pending %>%
  filter(!recorded_taxa_group %in% Special_groups) %>%
  select(-c("common_name_clean", "IHH_functional_major_groups_sm", "estimated_species_name", "Manual_Classification_Notes", "species_est_certainty", "species_est_reasoning", "genus_est_reasoning", "ornamentalFLAG_1")) %>%
  distinct(
    recorded_taxa_group,
    estimated_genus_name,
    .keep_all = TRUE
  )

# 219 obs. 5 vars.
  
join_NormalGroups <- Mex.Pacific_Landings_species %>%
  filter(!recorded_taxa_group %in% Special_groups) %>%
  left_join(
    NONOTRAS_MajorGroupKey_lookup,
    by = c("recorded_taxa_group","estimated_genus_name")
  )
# 5463338 obs. (the majority of the data, as expected) and 51 vars (+3 new vars).

# SANITY CHECK
subset<- join_NormalGroups %>% filter(is.na(Major_Functional_Group))
table(subset$recorded_taxa_group)
# 62 obs. So a few obs from the new states where we are missing a Major_Functional_Group assignment. Ignore these for now. 62 obs. is pretty neglible at this point, but at some point fix this. - NOTE TO SELF

# STEP 2: JOIN IN OTRAS
# In contrast: for otras taxa, we want to look up based on the recorded_taxa_group and common_name_clean. From that look-up we want to assign estimated_taxa_group, estimated_species_name, species_est_certainty, species_est_reasoning, estimated_genus_name, genus_est_reasoning, Major_Functional_Group, Assignment_for_Major_Functional_Group, Manual_Classification_Notes, and ornamentalFLAG_1
OTRAS_MajorGroupKey_lookup <- MajorGroupKey_pending %>%
  filter(recorded_taxa_group %in% Special_groups) %>%
  distinct(
    recorded_taxa_group,
    common_name_clean,
    .keep_all = TRUE
  )
# 278 obs. 13 vars.

join_SpecialGroups <- Mex.Pacific_Landings_species %>%
  filter(recorded_taxa_group %in% Special_groups) %>%
  left_join(
    OTRAS_MajorGroupKey_lookup,
    by = c("recorded_taxa_group", "common_name_clean"),
    suffix = c("", "_new")
  ) %>%
  mutate(
    across(
      ends_with("_new"),
      ~ .,
      .names = "{sub('_new$', '', .col)}"
    )
  )

join_SpecialGroups <- join_SpecialGroups %>%
  select(-ends_with("_new"))

# 942965 obs. (as expected) and 55 vars. (+7 new vars).

# STEP 3:
Mex.Pacific_Landings_classed <- bind_rows(join_NormalGroups, join_SpecialGroups)
# 6406303 obs. (as expected) and 55 vars.
```

## STEP 7a: Assign new class for ANCHOVETA and SARDINA = "Small pelagic marine fishes"

```{r What if we assign and pull out small pelagic fishes?}
unique(Mex.Pacific_Landings_classed$recorded_taxa_group)

# Small pelagics
Small_Pelagics<- c("ANCHOVETA", "SARDINA") 

Mex.Pacific_Landings_classed <- Mex.Pacific_Landings_classed %>%
  mutate(
    Major_Functional_Group = case_when(
      recorded_taxa_group %in% Small_Pelagics ~ "Anchovy & Sardine (small pelagics)",
      Major_Functional_Group == "Pelagic marine fishes" ~ "Pelagic marine fish (excl. anchovy & sardine)",
      TRUE ~ Major_Functional_Group
    )
  )
# 6406303 obs. (as expected) and 55 vars.

table(Mex.Pacific_Landings_classed$Major_Functional_Group)

           # Anchovy & Sardine (small pelagics)                                   Cephalopods 
           #                             154620                                        135299 
           #                        Crustaceans                        Demersal marine fishes 
           #                             293016                                       2726374 
           #                  Diadromous fishes                                   Echinoderms 
           #                              11451                                         55036 
           #                  Freshwater fishes                                       Jellies 
           #                             569408                                         15980 
           #                       Marine algae                             Marine fishes NEI 
           #                              10831                                         52364 
           #    Marine or freshwater fishes NEI                            Marine species NEI 
           #                             237591                                         66257 
           #         Molluscs excl. Cephalopods                         Other freshwater taxa 
           #                             339546                                          3991 
           #         Other marine invertebrates Pelagic marine fish (excl. anchovy & sardine) 
           #                                 31                                        690395 
           #                             Sharks                               Shrimps, prawns 
           #                             319935                                        715814 
           #                            unknown 
           #                               7144 

# SANITY CHECK: Which records are captured by the "Marine or freshwater fishes NEI" category?
subset<- Mex.Pacific_Landings_classed %>% 
  filter(Major_Functional_Group=="Marine or freshwater fishes NEI")
# 134361 obs. from 2000 to 2025 (makes sense).
min(subset$year)
max(subset$year)

table(subset$est_habitat)
table(subset$common_name_clean)
table(subset$estimated_taxa_group)
# Nice! These look like taxa that, as expected, could be freshwater or marine (BAGRE, CARPA, MOJARRA, TRUCHA). More information required to make an educated guess. 

  # BAGRE   CARPA MOJARRA  TRUCHA 
  # 69928    1474  161130    5059 
# Mostly freshwater taxa.
```

# SANITY CHECK: How many records have we accounted for with MajorGroupKey_pending? ----
```{r SANITY CHECK: How many records have we accounted for with our MajorGroupKey_pending?}
table(Mex.Pacific_Landings_classed$Major_Functional_Group)
subset<- Mex.Pacific_Landings_classed %>% filter(is.na(Major_Functional_Group))
table(subset$recorded_taxa_group)
#             ABULON             BESUGO             ORNATO              OTRAS    PECES DE ORNATO              RONCO 
#                  1                  1                  8               1149                  1                 57 
# RUBIA Y VILLAJAIBA 
#                  3 
# This shows that 1220 out of all 6406303 (0.01%) entries are missing a Major_Functional_Group assignment. Good enough for now!
```

```{r Summarize landing records and total biomass per Major_Functional_Group}
# Summarize landing records and total biomass per Major_Functional_Group, including NAs as "N/A"
# Ensure NAs are included in the factor
Mex.Pacific_Landings_classed$Major_Functional_Group[is.na(Mex.Pacific_Landings_classed$Major_Functional_Group)] <- "N/A"
Mex.Pacific_Landings_classed$IHH_functional_major_groups_sm[is.na(Mex.Pacific_Landings_classed$IHH_functional_major_groups_sm)] <- "N/A"

# Summarize REVISED landing records and biomass:
NEW.LandingsPercent_summary <- Mex.Pacific_Landings_classed %>%
  # Convert NA to a level so it's counted
  mutate(Major_Functional_Group = ifelse(
    is.na(Major_Functional_Group),
    "NA",
    Major_Functional_Group
  )) %>%
  group_by(Major_Functional_Group) %>%
  summarise(
    NEW.Percent_TotLandingRecords = round(n() / nrow(Mex.Pacific_Landings_classed) * 100, 1),
    NEW.Percent_TotalBiomass = round(
      sum(weight_kg_corrected, na.rm = TRUE) / sum(Mex.Pacific_Landings_classed$weight_kg_corrected, na.rm = TRUE) * 100,
      1
    ),
    .groups = "drop"
  ) %>%
  # Sort descending by biomass
  arrange(desc(NEW.Percent_TotalBiomass))

NEW.LandingsPercent_summary
# This suggests that, from 2000 to 2025, < 1% of all landing records and ~0% of all biomass come from taxa that are NOT classified to the Major_Functional_Group level. The rest come from major functional groups like demersal fish and shrimp. I can live with this! 

# What did these numbers look like before based on IHH groupings?
# Build summary table with counts, percent of landing records, and percent of total biomass

OLD.LandingsPercent_summary <- Mex.Pacific_Landings_classed %>%
  mutate(IHH_functional_major_groups_sm = ifelse(
    is.na(IHH_functional_major_groups_sm),
    "NA",
    IHH_functional_major_groups_sm
  )) %>%
  group_by(IHH_functional_major_groups_sm) %>%
  summarise(
    OLD.Percent_TotLandingRecords = round(n() / nrow(Mex.Pacific_Landings_classed) * 100, 1),
    OLD.Percent_TotBiomass = round(
      sum(weight_kg_corrected, na.rm = TRUE) / sum(Mex.Pacific_Landings_classed$weight_kg_corrected, na.rm = TRUE) * 100, 
      1
    ),
    .groups = "drop"
  )

# Join these two tables to compare.
library(fuzzyjoin)

# Perform fuzzy join using string distance
MERGED.LandingsPercent_summary <- stringdist_left_join(
  NEW.LandingsPercent_summary,
  OLD.LandingsPercent_summary, 
  by = c("Major_Functional_Group" = "IHH_functional_major_groups_sm"),
  method = "jw",           # Jaro-Winkler distance (good for names)
  max_dist = 0.2,          # Adjust threshold as needed (0 = exact, ~0.2 = close match)
  distance_col = "dist"    # Optional: keep the match distance
) %>%
  arrange(dist)             # Optional: sort by closest match


# Reorder columns
MERGED.LandingsPercent_summary <- MERGED.LandingsPercent_summary %>%
  select(
    Major_Functional_Group,
    OLD.Percent_TotBiomass,
    NEW.Percent_TotalBiomass,
    OLD.Percent_TotLandingRecords,
    NEW.Percent_TotLandingRecords
  ) %>%
  # Order rows:
  arrange(
    # Then by OLD percent landing records descending
    desc(OLD.Percent_TotBiomass)
  )

# Save this dataframe:
# write.csv(MERGED.LandingsPercent_summary, "/home/ferrer/ohi_goc_prep/FP/CONAPESCA_CleanUp/Catch_byTaxaGroup_summarytable_Apr22026.csv", row.names = FALSE)
```

## STEP 7b: Save Mex.Pacific_Landings_classed (keywords parsed with species names and major taxanomic group) as separate .rds file - OPTIONAL:
```{r Save this dataframe - GOC landings with keywords with species names and major taxonomic group as of Mar 25 2026}
saveRDS(Mex.Pacific_Landings_classed, file="/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/Mex.Pacific_Landings_Parsed.wSpecies&MajorGroups_2000.present_EMF.Apr152026.rds")
# 6406303 obs., and 55 vars.
```

```{r Remove dataframes to keep workspace clean}
rm(join_NormalGroups, join_SpecialGroups,
   MERGED.LandingsPercent_summary, NEW.LandingsPercent_summary, OLD.LandingsPercent_summary,
   NONOTRAS_MajorGroupKey_lookup, OTRAS_MajorGroupKey_lookup)
```

# STEP 8: Take a look into key taxonomic groups. ------------------------------
... Can find this work in the CONAPESCA_CleaningWorkflow rmd. It's not necessary here.

# > SHIFTING GEARS: Work on estimating sector  ------------------------------

# STEP 9: FLAG and estimate landings from AQUACULTURE based on a series of logical rules. ------------------------------
I think aquaculture and mariculture in the region is being vastly under-counted and under-reported.
Let's implement a series of rules to estimate if records are associated with aquaculture or not.

```{r Load in dataframe if needed}
Mex.Pacific_Landings_classed0<- Mex.Pacific_Landings_classed

Mex.Pacific_Landings_classed0<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/Mex.Pacific_Landings_Parsed.wSpecies&MajorGroups_2000.present_EMF.Apr152026.rds"))
# 6406303 obs., 55 vars.

colnames(Mex.Pacific_Landings_classed0)
```

## STEP 9a: FLAG AQUACULTURE:

```{r Create a Mex.Pacific_Landings_aquaculture dataframe}
Mex.Pacific_Landings_aquaculture<- Mex.Pacific_Landings_classed0

Mex.Pacific_Landings_aquaculture$est_aquaculture_production<- c(NA)
# n = 6406303 obs., 56 vars.
```

## VIEW: ALL KEY WORDS... Does this list need more work?
```{r Investigating Keywords}
# Extract all unique keywords from keyword columns 1 through 5. Delete exact duplicates. 
Keyword_vector <- Mex.Pacific_Landings_classed %>%
  select(keywords_1:keywords_5) %>%
  unlist() %>%
  unique() %>%
  na.omit()

Keyword_vector
```

First, let's flag landing_site names and eu_name that are associated with aquaculture.
```{r Estimate likelihood of aquaculture based on name of landing_site and eu_name}
#### FLAG 1 - Flag aquaculture based on landing_site and eu_name (3 levels: YES, NO, POSSIBLY): ####
vector<- Mex.Pacific_Landings_aquaculture %>%
  filter(year > 2024) %>%
  select(eu_name) %>%
  distinct() %>%
  slice_sample(n = 200)

#### Clues from landing_site and eu_name [UNDER CONSTRUCTION, Apr 15 2206] ####
LandingSite_strong.clues<- c("CULTIVADORES", "ESTANQUE", "VIVEROS", "ACUACULTURA")
# NOTE TO SELF: keep working on this list. - Apr 15 2206
LandingSite_clues<- c("CULT", "DREN", "BORDO", "PREDIO")

EUname_clues<- c("ACUICOLA", "FARMS", "MARICULTURA",
                        "MARICULTIVOS",
                        "AGROACUICOLA", "AGRICOLA",
                        "ACUACULTURA", "AQUACULTURE",
                        "GRANJA", "GRANJA CAMARONERA",
                        "RANCHO")

# Build regex patterns dynamically
LandingSite_strong_pattern <- paste0("\\b(", paste(LandingSite_strong.clues, collapse = "|"), ")\\b")
LandingSite_weak_pattern <- paste0("\\b(", paste(LandingSite_clues, collapse = "|"), ")\\b")
EUname_pattern <- paste0("\\b(", paste(EUname_clues, collapse = "|"), ")\\b")

Mex.Pacific_Landings_aquaculture <- Mex.Pacific_Landings_aquaculture %>%
  
  mutate(
    acuaculture_production_norm = str_to_lower(str_trim(acuaculture_production))
  ) %>%
  
  mutate(
    aquacultureFLAG_site.details = case_when(
      str_detect(landing_site, regex(LandingSite_strong_pattern, ignore_case = TRUE)) ~ "YES",
      str_detect(landing_site, regex(LandingSite_weak_pattern, ignore_case = TRUE)) ~ "POSSIBLY",
      str_detect(eu_name, regex(EUname_pattern, ignore_case = TRUE)) ~ "POSSIBLY",
      TRUE ~ "NO"
    )
  ) %>%
  
  select(-acuaculture_production_norm)

# SANITY CHECK:
# Check that we pulled out landing sites that make sense:
table(Mex.Pacific_Landings_aquaculture$aquacultureFLAG_site.details)

 #      NO POSSIBLY      YES 
 # 6228465   177300      538 
 # Right off the bat this seems good. Fewer NO, more possibles, less dirrect YES. Don't want to be OVERLY CONFIDENT in this estimate.

subset<- Mex.Pacific_Landings_aquaculture %>%
  filter(aquacultureFLAG_site.details %in% c(
    "YES",
    "POSSIBLY"
  )) %>%
  distinct(landing_site, eu_name, aquacultureFLAG_site.details) %>%
  arrange(landing_site, eu_name)
# 1835 sites that mostly all make sense to me! If an organism came from one of these sites I would assume it was a farmed species, even if the record was NOT marked as aquaculture.

# Let's review these sites manually:
# write.csv(
#   Mex.Pacific_Landings_aquaculture %>%
#     filter(est_aquaculture_production != "uncertain") %>%
#     distinct(landing_site, acuaculture_production, est_aquaculture_production, est_aquaculture_certainty, aquacultureFLAG_site.details),
#   "/home/ferrer/ohi_goc_prep/FP/Exploratory_Research/MAR_ExploratoryResearch/AquacultureSite_ClassificationCheck_Mar122026.csv",
#   row.names = FALSE
# )
# After a manual review, this logic seems okay for now... let's see how this affects of estimates of aquaculture.
```

Next, let's implement rules based on key words and taxa. What taxa are likely aquaculture?
```{r Flagging aquaculture landings based on key words}
colnames(Mex.Pacific_Landings_aquaculture)

table(Mex.Pacific_Landings_aquaculture$recorded_taxa_group)
table(Mex.Pacific_Landings_aquaculture$estimated_taxa_group)

#### FLAG 2 - Flag aquaculture based on key words, Aquaculture_keywords [LIST IS PENDING REVIEW / UNDER CONSTRUCTION] ####
Aquaculture_keywords<- c(
  "DE CULTIVO", "DE CULTIVO VIVO", "CULTIVADAS", "ABASTO RANCHO", "RANCHO", "PERLAS", "MEDIAS PERLAS"
)

# Flag rows with aquaculture keywords:
Mex.Pacific_Landings_aquaculture <- Mex.Pacific_Landings_aquaculture %>%
mutate(
  aquacultureFLAG_keywords = case_when(
    rowSums(across(keywords_1:keywords_5, ~ str_to_upper(str_trim(.x)) %in% Aquaculture_keywords), na.rm = TRUE) > 0 ~ "YES",
    TRUE ~ "NO"
  )
)

table(Mex.Pacific_Landings_aquaculture$aquacultureFLAG_keywords)
# 135680 observations explicitly flagged as aquaculture due to key words.
```

```{r Flag certain aquaculture and NOT-aquaculture taxa}
#### FLAG 3 - Flag known, likely, and NOT aquaculture taxa [LIST IS PENDING REVIEW / UNDER CONSTRUCTION] ####
# In theory, names could come from a number of columns: estimated_taxa_group, recorded_taxa_group, recorded_taxa_name, common_name_clean

table(Mex.Pacific_Landings_aquaculture$estimated_taxa_group)
table(Mex.Pacific_Landings_aquaculture$common_name_clean)

Possible_AquacultureTaxa <- c("OSTION", "JUREL", "CAMARON", "CARPA", "MOJARRA", "TRUCHA", "ALMEJA", "ABULON", "ALGAS") # estimated_taxa_group
Definite_AquacultureTaxa <- c("TILAPIA") # common_name_clean
ProbablyNot_AquacultureGroups <- c("Anchovy & Sardine (small pelagics)", "Cephalopods", "Echinoderms") # Major_Functional_Group

Mex.Pacific_Landings_aquaculture <- Mex.Pacific_Landings_aquaculture %>%
  mutate(
    aquacultureFLAG_specific.taxa = case_when(
      str_to_upper(str_trim(common_name_clean)) %in% Definite_AquacultureTaxa ~ "YES",
      str_to_upper(str_trim(estimated_taxa_group)) %in% Possible_AquacultureTaxa ~ "POSSIBLY",
      str_trim(Major_Functional_Group) %in% ProbablyNot_AquacultureGroups ~ "PROBABLY NOT",
      TRUE ~ "NO"
    )
  )

table(Mex.Pacific_Landings_aquaculture$aquacultureFLAG_specific.taxa)

     #      NO     POSSIBLY PROBABLY NOT          YES 
     # 4335101      1315696       344955       410551 
```

## Flag what CANNOT be aquaculture based on what we know about the taxa
```{r Flag Major_Functional_Groups that are definitely NOT aquaculture like sharks}
#### FLAG 4 - Flag Major_Functional_Groups that are DEFINITELY not aquaculture [LIST IS PENDING REVIEW / UNDER CONSTRUCTION] ####

table(Mex.Pacific_Landings_aquaculture$estimated_taxa_group)
table(Mex.Pacific_Landings_aquaculture$Major_Functional_Group)

# DEFINITELY NOT:
CannotBe_AquacultureGroups <- c("Jellies", "Sharks") # Major_Functional_Groups
CannotBe_AquacultureGenera <- c("Istiophorus", "Makaira", "Tetrapturus", "Xiphias", "Acanthocybium", "Dosidicus") # estimated_genus_name

Mex.Pacific_Landings_aquaculture <- Mex.Pacific_Landings_aquaculture %>%
  mutate(
    NOTaquacultureFLAG_specific.groups = case_when(
      str_trim(Major_Functional_Group) %in% CannotBe_AquacultureGroups ~ "CANNOT BE AQUACULTURE",
      str_trim(estimated_genus_name) %in% CannotBe_AquacultureGenera ~ "CANNOT BE AQUACULTURE",
      TRUE ~ NA
    )
  )

table(Mex.Pacific_Landings_aquaculture$NOTaquacultureFLAG_specific.groups)

# CANNOT BE AQUACULTURE 
#                409569 
```

Create a combined aquaculture flag based on the previous three flags (aquacultureFLAG_site.details, aquacultureFLAG_keywords, aquacultureFLAG_specific.taxa) and the new flag NOTaquacultureFLAG_specific.groups

Here's the logic for the combined flag:

• If at least one flag is "YES", then flags indicate that aquaculture is highly likely and aquacultureFLAG_combined = "YES" (my YES flags are pretty robust at this point)
• If one more flags is "POSSIBLY", then aquacultureFLAG_combined is "possibly aquaculture" but not certain
• If no flags indicate aquaculture, then aquacultureFLAG_combined is "NO"

• BUT, if the NOTaquacultureFLAG_specific.groups indicates CANNOT BE AQUACULTURE, then combined flag is "NO"


```{r Create a combined aquaculture flag}
#### FLAG 5 - Create a combined aquaculture flag based on the previous four flags ####
Mex.Pacific_Landings_aquaculture <- Mex.Pacific_Landings_aquaculture %>%
  
  # Remember, put the strongest condition first:
  mutate(
    aquacultureFLAG_combined = case_when(
      
      if_any(c(NOTaquacultureFLAG_specific.groups), ~ .x == "CANNOT BE AQUACULTURE") ~ "CANNOT BE AQUACULTURE",
      
      if_any(c(aquacultureFLAG_specific.taxa), ~ .x == "PROBABLY NOT") ~ "PROBABLY NOT",
      
      if_any(c(aquacultureFLAG_site.details,
               aquacultureFLAG_keywords,
               aquacultureFLAG_specific.taxa), ~ .x == "YES") ~ "YES",
      
      if_any(c(aquacultureFLAG_site.details,
               aquacultureFLAG_keywords,
               aquacultureFLAG_specific.taxa), ~ .x == "POSSIBLY") ~ "POSSIBLY",
      
      TRUE ~ "NO"
    )
  )

# SANITY CHECK:
subset<- Mex.Pacific_Landings_aquaculture %>% 
  filter(aquacultureFLAG_site.details %in% c("YES", "POSSIBLY") & aquacultureFLAG_combined=="CANNOT BE AQUACULTURE")

table(subset$estimated_taxa_group)
# > 3647 where site names / details are incongruous with the taxa being caught. For example, giant squid and sharks are not used for aquaculture.
# Maybe these are taxa ASSOCIATED with aquaculture, or found new aquaculture sites but not actually aquaculture. :)

# This is good! It's a sign that our aquacultureFLAG_combined is working and trumping other clues.
```

## STEP 9b: ESTIMATE AQUACULTURE:
Here is the logic and set of decisions designed to estimate aquaculture:

Decision Table:
Report      aquacultureFLAG_combined  aquaculture_estimate    certainty
  yes       YES                       aquaculture             high; aquaculture flags agree with original report
  yes       POSSIBLY                  aquaculture             high; aquaculture flags agree with original report
  yes       NO                        aquaculture	            medium; no flags for aquaculture but original report says aquaculture
  yes       PROBABLY NOT              possibly aquaculture    low; aquaculture flags do not agree with original report
  yes       CANNOT BE AQUACULTURE     not aquaculture         medium; aquaculture flags suggest cannot be aquaculture, original report likely wrong
  
  no        YES	                      likely aquaculture	    medium; flags suggest aquaculture but original report says no, report could be wrong
  no        POSSIBLY                  possibly aquaculture    low; flags suggest possibly aquaculture but original report says no
  no        NO                        not aquaculture         high; aquaculture flags agree with original report
  no        PROBABLY NOT              not aquaculture         high; aquaculture flags agree with original report
  no        CANNOT BE AQUACULTURE     not aquaculture         high; aquaculture flags agree with original report
  
  NA        YES                       likely aquaculture      medium; no data from original report but flags suggest most likely aquaculture
  NA        POSSIBLY                  possibly aquaculture    low; no data from original report but flags suggest possibly aquaculture
  NA        NO                        unlikely aquaculture    medium; no data from original report and flags do not suggest aquaculture
  NA        PROBABLY NOT              unlikely aquaculture    high; no data from original report and flags suggest probably not aquaculture
  NA        CANNOT BE AQUACULTURE     not aquaculture         high; no data from original report and flags suggest definitely not aquaculture

```{r Generate estimates and confidence values}
#### Estimate aquaculture and confidence values based on the comparison of aquacultureFLAG_combined versus the official report: ####
Mex.Pacific_Landings_aquaculture <- Mex.Pacific_Landings_aquaculture %>%
  mutate(
    # Normalize original report:
    acuaculture_original.report = str_to_lower(str_trim(acuaculture_production)),

    # ESTIMATE OF AQUACULTURE:
    est_aquaculture_production = case_when(
      
      # Report = YES
      acuaculture_original.report == "yes" & aquacultureFLAG_combined %in% c("YES", "POSSIBLY", "NO") ~ "aquaculture",
      acuaculture_original.report == "yes" & aquacultureFLAG_combined == "PROBABLY NOT" ~ "possibly aquaculture",
      acuaculture_original.report == "yes" & aquacultureFLAG_combined == "CANNOT BE AQUACULTURE" ~ "not aquaculture",
      
      # Report = NO
      acuaculture_original.report == "no" & aquacultureFLAG_combined == "YES" ~ "likely aquaculture",
      acuaculture_original.report == "no" & aquacultureFLAG_combined == "POSSIBLY" ~ "possibly aquaculture",
      acuaculture_original.report == "no" & aquacultureFLAG_combined %in% c("NO", "PROBABLY NOT", "CANNOT BE AQUACULTURE") ~ "not aquaculture",
      
      # Report = NA
      is.na(acuaculture_original.report) & aquacultureFLAG_combined == "YES" ~ "likely aquaculture",
      is.na(acuaculture_original.report) & aquacultureFLAG_combined == "POSSIBLY" ~ "possibly aquaculture",
      is.na(acuaculture_original.report) & aquacultureFLAG_combined == "NO" ~ "unlikely aquaculture",
      is.na(acuaculture_original.report) & aquacultureFLAG_combined == "PROBABLY NOT" ~ "unlikely aquaculture",
      is.na(acuaculture_original.report) & aquacultureFLAG_combined == "CANNOT BE AQUACULTURE" ~ "not aquaculture",
      
      TRUE ~ "uncertain"
    ),

    # CERTAINTY ESTIMATES:
    est_aquaculture_certainty = case_when(
      
      # Report = YES
      acuaculture_original.report == "yes" & aquacultureFLAG_combined %in% c("YES", "POSSIBLY") ~ "high; aquaculture flags agree with original report",
      acuaculture_original.report == "yes" & aquacultureFLAG_combined == "NO" ~ "medium; no flags for aquaculture but original report says aquaculture",
      acuaculture_original.report == "yes" & aquacultureFLAG_combined == "PROBABLY NOT" ~ "low; aquaculture flags do not agree with original report",
      acuaculture_original.report == "yes" & aquacultureFLAG_combined == "CANNOT BE AQUACULTURE" ~ "medium; aquaculture flags suggest cannot be aquaculture, original report likely wrong",
      
      # Report = NO
      acuaculture_original.report == "no" & aquacultureFLAG_combined == "YES" ~ "medium; flags suggest aquaculture but original report says no, report could be wrong",
      acuaculture_original.report == "no" & aquacultureFLAG_combined == "POSSIBLY" ~ "low; flags suggest possibly aquaculture but original report says no",
      acuaculture_original.report == "no" & aquacultureFLAG_combined %in% c("NO", "PROBABLY NOT", "CANNOT BE AQUACULTURE") ~ "high; aquaculture flags agree with original report",
      
      # Report = NA
      is.na(acuaculture_original.report) & aquacultureFLAG_combined == "YES" ~ "medium; no data from original report but flags suggest most likely aquaculture",
      is.na(acuaculture_original.report) & aquacultureFLAG_combined == "POSSIBLY" ~ "low; no data from original report but flags suggest possibly aquaculture",
      is.na(acuaculture_original.report) & aquacultureFLAG_combined == "NO" ~ "medium; no data from original report and flags do not suggest aquaculture",
      is.na(acuaculture_original.report) & aquacultureFLAG_combined == "PROBABLY NOT" ~ "high; no data from original report and flags suggest probably not aquaculture",
      is.na(acuaculture_original.report) & aquacultureFLAG_combined == "CANNOT BE AQUACULTURE" ~ "high; no data from original report and flags suggest definitely not aquaculture",
      
      TRUE ~ "not enough information to assign aquaculture estimate"
    )
  ) %>%
  select(-acuaculture_original.report)

# 6406303 obs., 62 vars.

# SANITY CHECK:
table(Mex.Pacific_Landings_aquaculture$est_aquaculture_production)

         # aquaculture   likely aquaculture      not aquaculture possibly aquaculture unlikely aquaculture 
         #       38680               531058              3999687              1266052               570826 
               
subset<- Mex.Pacific_Landings_aquaculture %>% filter(acuaculture_production=="yes" & est_aquaculture_production =="not aquaculture")
table(subset$estimated_taxa_group) 
# Excellent! Most of the aquaculture explicitly reported as aquaculture is estimated to be aquaculture, but the ones that were reported as aquaculture but estimated to not be aquaculture make sense! Sharks, squid, and swordfish:
# CALAMAR   CAZON   OTRAS (swordfish) TIBURON 
#     276     506     188             1620 

min(subset$year)
max(subset$year)
# Very weird, the official aquaculture reports only go from 2006 to 2017

subset<- Mex.Pacific_Landings_aquaculture %>% filter(est_aquaculture_production==c("aquaculture", "likely aquaculture"))
# 285061 obs. The data set grows a lot!
```

```{r Re-level est_aquaculture_production column}
# Re-level est_aquaculture_production to be a little more logical:
Mex.Pacific_Landings_aquaculture <- Mex.Pacific_Landings_aquaculture %>%
  mutate(
    est_aquaculture_production = factor(
      est_aquaculture_production,
      levels = c(
        "aquaculture",
        "likely aquaculture",
        "possibly aquaculture",
        "unlikely aquaculture",
        "not aquaculture"
      )
    )
  )

# 6406303 obs., 62 vars.
```

```{r Summarize reported aquaculture versus estimated aquaculture}
table(Mex.Pacific_Landings_aquaculture$acuaculture_production)
table(Mex.Pacific_Landings_aquaculture$est_aquaculture_production)

         # aquaculture   likely aquaculture possibly aquaculture unlikely aquaculture      not aquaculture 
         #       38680               531058              1266052               570826              3999687

Reported.Aquaculture_summary<- Mex.Pacific_Landings_aquaculture %>% 
  filter(acuaculture_production== c("yes")) %>%
  filter(year >= 2014, year <= 2024) %>%
  select(year, weight_kg_corrected, value) %>%
  group_by(year) %>%
  summarize(reported_annual_catch= sum(weight_kg_corrected, na.rm = TRUE),
            reported_annual_value= sum(value, na.rm = TRUE))

Estimated.Aquaculture_summary<- Mex.Pacific_Landings_aquaculture %>% 
  filter(est_aquaculture_production== c("aquaculture", "likely aquaculture")) %>%
  filter(year >= 2014, year <= 2024) %>%
  select(year, weight_kg_corrected, value) %>%
  group_by(year) %>%
  summarize(est_annual_catch= sum(weight_kg_corrected, na.rm = TRUE),
            est_annual_value= sum(value, na.rm = TRUE))

Aquaculture_summary<- Estimated.Aquaculture_summary %>%
  left_join(Reported.Aquaculture_summary, by = "year") %>%
  select(year, reported_annual_catch, est_annual_catch, reported_annual_value, est_annual_value)
```

```{r Plot reported aquaculture versus estimated aquaculture}
EstvReported_Aquaculture_plot<- ggplot(Aquaculture_summary, aes(x = year)) +
  geom_line(aes(y = est_annual_catch, color = "Estimated catch"), linewidth = 1.2) +
  geom_point(aes(y = est_annual_catch, color = "Estimated catch"), size = 2) +
  
  geom_line(aes(y = reported_annual_catch, color = "Reported catch"), linewidth = 1.2) +
  geom_point(aes(y = reported_annual_catch, color = "Reported catch"), size = 2) +
  
  scale_color_manual(
    values = c(
      "Estimated catch" = "#1b9e77",
      "Reported catch" = "#d95f02"
    )
  ) +
  labs(
    x = "Year",
    y = "Catch (kg)",
    color = "",
    title = "Estimated vs Reported Aquaculture"
  ) +
  theme_minimal(base_size = 14) +
  scale_x_continuous(breaks = Aquaculture_summary$year) # ensures every year is a tick

EstvReported_Aquaculture_plot
```

```{r Save EstvReported_Aquaculture_plot}
# Save the plot as a high-resolution PNG
ggsave(
  filename = "/home/ferrer/ohi_goc_prep/FP/CONAPESCA_CleanUp/EstvReported_Aquaculture_plot_Apr152026.png",
  plot = EstvReported_Aquaculture_plot,
  width = 10,       # width in inches
  height = 6,       # height in inches
  dpi = 300,        # resolution
  bg = "white"      # ensures white background
)
```
Notice the spike in landings from 2017-2018? Of course, how could you miss it.
This jump coincides (temporally) with a change in reporting associated with the raw data itself. 
Data pre-2018 is reported here: mex_conapesca_avisos_2000_2019.rds
Data 2018 and onward is reported here: mex_conapesca_apertura_2018_present.rds
> NOTE TO SELF: Consult with Juan Carlos about these two datasets. What is fundamentally different about them?


# STEP 10: FLAG and estimate landings from FRESHWATER based on a series of logical rules. ------------------------------
```{r Flag freshwater landings}
table(Mex.Pacific_Landings_aquaculture$Major_Functional_Group)

# Possible freshwater fishes:
subset<- Mex.Pacific_Landings_aquaculture %>%
  filter(Major_Functional_Group== "Diadromous fishes")
table(subset$estimated_genus_name)
# Makes sense: Anguilla, Chanos

#### FLAG 6 - Flag freshwater based on keywords, Freshwater_keywords [LIST IS PENDING REVIEW / UNDER CONSTRUCTION] ####

Freshwater_keywords<- c(
  "DE AGUA DULCE"
)

#### FLAG 7A - Flag freshwater based on key taxa, Known_FreshwaterTaxa [LIST IS PENDING REVIEW / UNDER CONSTRUCTION] ####
Known_FreshwaterTaxa<- c(
  "tilapia",
  "rana", "rana toro",
  "trucha",
  "bagre canal"
) #common_name_clean

#### FLAG 7B - Flag known freshwater genera [LIST IS PENDING REVIEW / UNDER CONSTRUCTION] ####
Known_FreshwaterGenera<- c(
  "Cyprinus", # common carp; strictly freshwater.
  "Macrobrachium", # freshwater prawns/shrimps; mostly freshwater.
  "Hypophthalmichthys", # Asian carp; freshwater.
  "Chirostoma", # “charales,” endemic to freshwater lakes in Mexico.
  "Carassius", # goldfish/crucian carp; freshwater.
  "Micropterus", # black bass; freshwater.
  "Lepomis", # sunfish; freshwater.
  "Cyprinodon", # pupfish; freshwater.
  "Rana" # frogs
  ) # estimated_genus_name


#### FLAG 8 & 9 - Flag freshwater and NOT based on Major_Functional_Group and est_habitat type [UNDER CONSTRUCTION as of April 3 2026] ####

# Flag freshwater rows based on keywords (FLAGS 6 and 7), info from the IHH key, and estimated habitat type (based on landing site):
table(Mex.Pacific_Landings_aquaculture$Major_Functional_Group)

# Definitely freshwater:
Definite_FreshwaterGroups<- c("Freshwater fishes", "Other freshwater taxa")

# Could be freshwater or marine:
Unknown_MarineFreshwaterGroups<- c("Marine or freshwater fishes NEI", "unknown")
Likely_MarineGroups<- c("Diadromous fishes", "Molluscs excl. Cephalopods", "Shrimps, prawns")

# Definitely marine:
Definite_MarineGroups<- c("Anchovy & Sardine (small pelagics)", "Cephalopods", "Crustaceans", "Demersal marine fishes", "Echinoderms", "Jellies", "Marine algae", "Marine fishes NEI", "Marine species NEI", "Other marine invertebrates", "Pelagic marine fish (excl. anchovy & sardine)", "Sharks")
# NOTES:
# - Crustaceans could technically be split but are overwhelmingly marine.
# - Molluscs could technically be split but are overwhelmingly marine. There could be some freshwater clams hiding in there.
# - Shrimp technically split but overwhelmingly marine. When in doubt, lean marine.

# Logic:
# - First, any row whose functional group is explicitly in Definite_MarineGroups will immediately get "CANNOT BE FRESHWATER; MARINE"
# - Second, if a row didn’t hit the first condition, it is evaluated for strong freshwater evidence. "YES FRESHWATER"
# - Third, rows that didn’t match the first two conditions are considered. This is broader marine signal. "LIKELY MARINE"
# - Fourth, rows that didn’t match the first two conditions are considered. This is an uncertain signal. "POSSIBLY FRESHWATER OR MARINE"
# - Finally, Any row that didn’t meet any of the above conditions gets "NOT FLAGGED EITHER WAY".

Mex.Pacific_Landings_aquaculture <- Mex.Pacific_Landings_aquaculture %>%
  mutate(
    freshwater_flag = case_when(
      
      # Definitely NOT freshwater:
      (Major_Functional_Group %in% Definite_MarineGroups) ~ "CANNOT BE FRESHWATER; MARINE",
      
      # Definite freshwater: any definite criteria
      (Major_Functional_Group %in% Definite_FreshwaterGroups) |
      (common_name_clean %in% Known_FreshwaterTaxa) |
      (estimated_genus_name %in% Known_FreshwaterGenera) |
      (rowSums(across(keywords_1:keywords_5, ~ .x %in% Freshwater_keywords), na.rm = TRUE) > 0) |
      (est_habitat == "freshwater") ~ "YES FRESHWATER",
      
      # Likely marine: broader/uncertain criteria
      (Major_Functional_Group %in% Likely_MarineGroups) ~ "LIKELY MARINE",
      
      # Possible freshwater: broader/uncertain criteria
      (Major_Functional_Group %in% Unknown_MarineFreshwaterGroups) ~ "POSSIBLY FRESHWATER OR MARINE",
      
      # Otherwise, not freshwater
      TRUE ~ "NOT FLAGGED EITHER WAY"
    )
  )

table(Mex.Pacific_Landings_aquaculture$freshwater_flag)
# 661699 observations flagged as YES definitely "freshwater" based on freshwater keywords, the taxa common name, landing site, or Major_Functional_Group. Meanwhile, other entries, like "MOJARRA" could go either way and are flagged as possibly freshwater.

 # CANNOT BE FRESHWATER; MARINE                 LIKELY MARINE        NOT FLAGGED EITHER WAY POSSIBLY FRESHWATER OR MARINE 
 #                      4520138                       1041817                          1217                        181432 
 #               YES FRESHWATER 
 #                       661699 
```

```{r SANITY CHECK: Do the things flagged as freshwater make sense?}
# Lot's of Mojarra moved into the possible freshwater category...
subset<- Mex.Pacific_Landings_aquaculture %>%
  filter(freshwater_flag=="YES FRESHWATER")
# 661699 obs.

table(subset$estimated_taxa_group)

    # ALMEJA      BAGRE    CAMARON    CARACOL      CARPA LANGOSTINO     LOBINA    MOJARRA      OTRAS     TRUCHA 
    #   3829      44024       2778          2      69610      17586      32338     431929      59368        232 

# Let's take a closer look...
subset2<- subset %>%
  filter(recorded_taxa_group %in% c("ALMEJA", "CAMARON", "LANGOSTA", "LANGOSTINO"))

table(subset2$estimated_taxa_group)
table(subset2$estimated_species_name)
table(subset2$keywords_1)
# *Hm. But they are almost all (97%) flagged as "DE AGUA DUCLE" so yeah, that makes sense.
# I'm going to leave this as the "cost of doing business" for now but could be worth coming back to later.

subset3<- subset %>%
  filter(estimated_taxa_group %in% c("LOBINA"))
table(subset3$estimated_species_name)
# Florida bass. Makes sense!
```


```{r SANITY CHECK: Do the records classed as aquaculture and freshwater make sense?}
# SANITY CHECK:
table(Mex.Pacific_Landings_aquaculture$est_aquaculture_production)
subset<- Mex.Pacific_Landings_aquaculture %>% filter(est_aquaculture_production == "likely aquaculture")


table(subset$freshwater_flag)
# Note that out of the 279574 records that are "likely aquaculture", 173381 (~62%) are also flagged as freshwater species (things like tilapia, trout, and frogs). This makes sense!

# Meanwhile, the marine aquaculture species include shrimp, snapper, jacks, lots of bivalves and snails.
table(subset$recorded_taxa_name_clean)

subset2<- subset %>% filter(recorded_taxa_name_clean=="JUREL")
table(subset2$eu_name)
# NOTE TO SELF: Yes, these data for Jurel look really nice. These are definitely aquaculture jacks. Flags are doing a decent job so far!
```

```{r Reorder and rename columns in Mex.Pacific_Landings_aquaculture in new df Mex.Pacific_Landings_sector}
Mex.Pacific_Landings_sector <- Mex.Pacific_Landings_aquaculture %>%
  rename(reported_aquaculture_production = acuaculture_production,
         reported_fleet = fleet) %>%
  relocate(
    reported_aquaculture_production,
    NOTaquacultureFLAG_specific.groups,
    aquacultureFLAG_site.details,
    aquacultureFLAG_keywords,
    aquacultureFLAG_specific.taxa,
    aquacultureFLAG_combined,
    est_aquaculture_production,
    est_aquaculture_certainty,
    est_habitat,
    freshwater_flag,
    reported_fleet,
    .after = last_col()
  )
# 6406303 obs., 63 vars.
```

# STEP 11: Estimate SECTOR based on record details, estimated genera, estimated Major Functional Group, aquaculture flags, and freshwater flags. ------------------------------

What sector was actually reported?
```{r Create sector_estimate_reported column}
colnames(Mex.Pacific_Landings_sector)

Mex.Pacific_Landings_sector <- Mex.Pacific_Landings_sector %>%
  mutate(
    sector_estimate_reported = case_when(
      
      # More specific first
      reported_aquaculture_production == "yes" & reported_fleet == "small_scale" ~ "small-scale aquaculture",
      
      # General aquaculture
      reported_aquaculture_production == "yes" ~ "aquaculture",
      
      # Fisheries by fleet
      reported_fleet == "small_scale" ~ "small-scale fisheries",
      reported_fleet == "large_scale" ~ "large-scale fisheries",
      
      # Fallback for missing/unknown
      TRUE ~ "yet to be classified"
    )
  )

# SANITY CHECK:
table(Mex.Pacific_Landings_sector$sector_estimate_reported)
          # aquaculture large-scale fisheries small-scale fisheries  yet to be classified 
          #       42302                195953               1861039               4307009 

# Re-level
Mex.Pacific_Landings_sector <- Mex.Pacific_Landings_sector %>%
  mutate(
    sector_estimate_reported = factor(
      sector_estimate_reported,
      levels = c(
        "aquaculture",
        "yet to be classified",
        "small-scale fisheries",
        "large-scale fisheries"
      )
    )
  )

# REPORTED:
SectorLandings_summary_REPORTED<- Mex.Pacific_Landings_sector %>% 
  filter(year >= 2014, year <= 2024) %>%
  select(year, sector_estimate_reported, weight_kg_corrected, value) %>%
  group_by(year, sector_estimate_reported) %>%
  summarize(est_annual_catch_kg= sum(weight_kg_corrected, na.rm = TRUE),
            est_annual_value= sum(value, na.rm = TRUE))

SectorLandings_summary_REPORTED$est_annual_catch_mton <- SectorLandings_summary_REPORTED$est_annual_catch_kg/1000

table(SectorLandings_summary_REPORTED$sector_estimate_reported)
```

```{r Estimate sector}
table(Mex.Pacific_Landings_sector$est_aquaculture_production)   # adds up to: 6406303, as expected
table(Mex.Pacific_Landings_sector$freshwater_flag)              # adds up to: 6406303, as expected
table(Mex.Pacific_Landings_sector$reported_fleet)               # adds up to: 2056992 (some NAs)

#### ESTIMATE MOST LIKELY SECTOR based on report info and flags: ####

Mex.Pacific_Landings_sector <- Mex.Pacific_Landings_sector %>%
  mutate(
    # Helper columns
    is_aquaculture = est_aquaculture_production %in% c("aquaculture", "likely aquaculture"),
    is_likely.wildcatch = est_aquaculture_production %in% c("unlikely aquaculture", "not aquaculture", "possibly aquaculture"),
    is_freshwater = freshwater_flag %in% c("YES FRESHWATER"),
    is_marine.likely = freshwater_flag %in% c("CANNOT BE FRESHWATER; MARINE", "LIKELY MARINE"),
    is_fresh.or.marine = freshwater_flag %in% c("POSSIBLY FRESHWATER OR MARINE", "NOT FLAGGED EITHER WAY"),
    
    # Implement logic to estimate sector
    sector_estimate_draft = case_when(
      
      #### DEFINITE FRESHWATER SECTORS: ####
      is_freshwater & is_aquaculture ~ "freshwater: aquaculture",
      is_freshwater & is_likely.wildcatch & reported_fleet == "small_scale" ~ "freshwater: small-scale capture",
      is_freshwater & is_likely.wildcatch & reported_fleet == "large_scale" ~ "freshwater: large-scale capture",
      is_freshwater ~ "freshwater: fishery sector yet to be classified",
      
      #### DEFINITE & LIKELY MARINE SECTORS: ####
      is_marine.likely & is_aquaculture ~ "marine: aquaculture",
      is_marine.likely & is_likely.wildcatch & reported_fleet == "small_scale" ~ "marine: small-scale capture",
      is_marine.likely & is_likely.wildcatch & reported_fleet == "large_scale" ~ "marine: large-scale capture",
      is_marine.likely ~ "marine: fishery sector yet to be classified",
      
      #### SECTOR UNCERTAIN, COULD GO EITHER WAY: ####
      is_fresh.or.marine & is_aquaculture ~ 
        "freshwater or marine: aquaculture",
      is_fresh.or.marine & is_likely.wildcatch & reported_fleet == "small_scale" ~ 
        "freshwater or marine: small-scale capture",
      is_fresh.or.marine & is_likely.wildcatch & reported_fleet == "large_scale" ~ 
        "freshwater or marine: large-scale capture",
      is_fresh.or.marine ~ 
        "freshwater or marine: fishery sector yet to be classified",
      
      # Catch-all safeguard
      TRUE ~ NA_character_
    )
  ) %>%
  select(-is_aquaculture, -is_likely.wildcatch, -is_freshwater, -is_marine.likely, -is_fresh.or.marine)


# SANITY CHECK:
table(Mex.Pacific_Landings_sector$sector_estimate_draft)

                #         freshwater or marine: aquaculture freshwater or marine: fishery sector yet to be classified 
                #                                      5550                                                    126248 
                # freshwater or marine: large-scale capture                 freshwater or marine: small-scale capture 
                #                                      1132                                                     49719 
                #                   freshwater: aquaculture           freshwater: fishery sector yet to be classified 
                #                                    414151                                                    187124 
                #           freshwater: large-scale capture                           freshwater: small-scale capture 
                #                                        20                                                     60404 
                #                       marine: aquaculture               marine: fishery sector yet to be classified 
                #                                    150037                                                   3569462 
                #               marine: large-scale capture                               marine: small-scale capture 
                #                                    194755                                                   1647701 

table(Mex.Pacific_Landings_sector$est_aquaculture_production)
table(Mex.Pacific_Landings_sector$freshwater_flag)
table(Mex.Pacific_Landings_sector$reported_fleet)
```

```{r Re-level sector_estimate_draft}
# Re-level
Mex.Pacific_Landings_sector <- Mex.Pacific_Landings_sector %>%
  mutate(
    sector_estimate_draft = factor(
      sector_estimate_draft,
      levels = c(
        "marine: fishery sector yet to be classified",
        "marine: large-scale capture",
        "marine: small-scale capture",
        "marine: aquaculture",
        "freshwater: aquaculture",
        "freshwater: small-scale capture",
        "freshwater: large-scale capture",
        "freshwater: fishery sector yet to be classified",
        "freshwater or marine: aquaculture",
        "freshwater or marine: small-scale capture",
        "freshwater or marine: large-scale capture",
        "freshwater or marine: fishery sector yet to be classified"
      )
    )
  )
```

```{r Summarize by sector}
# ESTIMATED:
SectorLandings_summary <- Mex.Pacific_Landings_sector %>% 
  filter(year >= 2014, year <= 2024) %>%
  select(year, weight_kg_corrected, value, sector_estimate_draft) %>%
  group_by(year, sector_estimate_draft) %>%
  summarize(
    est_annual_catch_kg = sum(weight_kg_corrected, na.rm = TRUE),
    est_annual_value    = sum(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # Convert to metric tons
  mutate(est_annual_catch_mton = est_annual_catch_kg / 1000) %>%
  # Compute percent of annual total catch
  group_by(year) %>%
  mutate(percent_AnnualTotCatch = round(100 * est_annual_catch_kg / sum(est_annual_catch_kg, na.rm = TRUE))) %>%
  ungroup()

# Okay great. So from this we can see that the following sectors make up only 0-1% of total catch. I feel comfortable ignoring them then for the purposing of assessing food provisioning (maybe a different story for natural products) - Mar 24 2026.
# Small contributions from:
# freshwater: small-scale capture
# freshwater: large-scale capture
# freshwater: fishery sector yet to be classified
# freshwater or marine: aquaculture
# freshwater or marine: small-scale capture
# freshwater or marine: large-scale capture
# freshwater or marine: fishery sector yet to be classified

table(SectorLandings_summary$sector_estimate_draft)
```

```{r Trim and prepare Mex.Pacific_Landings_sector df for export}
colnames(Mex.Pacific_Landings_sector)

Mex.Pacific_Landings_sector<- Mex.Pacific_Landings_sector %>%
  select(-c(Manual_Classification_Notes, species_est_reasoning, genus_est_reasoning))

# 6406303 obs., 62 obs.
```

## Plotting sector estimates:
```{r Plot each sector of interest over time}
# SECTORS OF INTEREST:
Sectors_of_Interest<- c("marine: fishery sector yet to be classified" , "marine: small-scale capture", "marine: large-scale capture" , "marine: aquaculture", "freshwater: aquaculture")
  
# Define colors for each sector (fleet distinction = lighter for small-scale)
sector_colors <- c(
  # Marine (blues)
  "marine: fishery sector yet to be classified"       = "#cce0ff",  # very light blue for unknown
  "marine: small-scale capture"                       = "#6699cc",  # medium blue
  "marine: large-scale capture"                       = "#336699",  # darker blue
  "marine: aquaculture"                               = "#33cccc",  # teal/aquaculture cue
  
  # Freshwater (warm shades)
  "freshwater: aquaculture"                          = "#ff9933"  # bright orange
  # "freshwater: fishery sector yet to be classified" = "#ffe0cc",  # light orange for unknown - IGNORE THIS SECTOR FOR NOW.
  # "freshwater: small-scale capture"                  = "#ff6666",  # strong red/orange - IGNORE THIS SECTOR FOR NOW.
  # "freshwater: large-scale capture"                  = "#ffb366",  # lighter orange/yellow - IGNORE THIS SECTOR FOR NOW.
  
  # Freshwater or Marine (grey and purple shades) - IGNORE FOR NOW.
  # "freshwater or marine: fishery sector yet to be classified"  = "#e0e0e0",  # light grey
  # "freshwater or marine: small-scale capture"                    = "#b399cc",  # light purple
  # "freshwater or marine: large-scale capture"                    = "#8040a0",  # medium purple
  # "freshwater or marine: aquaculture"                             = "#999999"   # neutral grey
)

# Plot
Landings_bySector_plot<- SectorLandings_summary %>%
  filter(year >= 2018) %>%
  filter(sector_estimate_draft %in% Sectors_of_Interest) %>%
  ggplot(aes(x = year, y = est_annual_catch_mton, fill = sector_estimate_draft)) +
  
  # Area plot with slight border for separation
  geom_area(alpha = 0.9, color = "white", linewidth = 0.2) +
  
  # Manual fill palette
  scale_fill_manual(
    values = sector_colors,
    name = "Sector"
  ) +
  
  # Labels
  labs(
    x = "Year",
    y = "Estimated Annual Catch (metric tons)",
    title = "Estimated Catch by Sector (All of the Mexican Pacific)",
    subtitle = "Filtered for 2018 onwards; dropped sectors that contributed < 1% total catch"
  ) +
  
  # Minimal theme and sleek styling
  theme_bw(base_size = 14) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey85"),
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 13, color = "grey30")
  ) +
  
  # Optional: better legend layout
  guides(
    fill = guide_legend(nrow = 3, byrow = TRUE, title.position = "top")
  )

Landings_bySector_plot
```

Looks great! Save this plot.
```{r Save Landings_bySector_plot}
# Save the plot as a high-resolution PNG
ggsave(
  filename = "/home/ferrer/ohi_goc_prep/FP/CONAPESCA_CleanUp/Catch_bySector_plot_Apr152026.png",
  plot = Landings_bySector_plot,
  width = 10,       # width in inches
  height = 6,       # height in inches
  dpi = 300         # resolution
)
```

```{r Plot sector_estimate_reported}
# Plot sector_estimate_reported
REPORTEDLandings_bySector_plot<- SectorLandings_summary_REPORTED %>%
  filter(year >= 2018) %>%
  ggplot(aes(x = year, y = est_annual_catch_mton, fill = sector_estimate_reported)) +
  
  # Area plot with slight border for separation
  geom_area(alpha = 0.9, color = "white", linewidth = 0.2) +
  
  # Labels
  labs(
    x = "Year",
    y = "Reported Annual Catch (metric tons)",
    title = "Catch by Reported Sector",
    subtitle = "Filtered for 2018 onwards"
  ) +
  
  # Minimal theme and sleek styling
  theme_bw(base_size = 14) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey85"),
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 13, color = "grey30")
  ) +
  
  # Optional: better legend layout
  guides(
    fill = guide_legend(nrow = 3, byrow = TRUE, title.position = "top")
  )

REPORTEDLandings_bySector_plot
```

```{r Clean workspace as needed}
# remove dataframes to keep workspace clean:
rm(subset, subset2, subset3,
   Landings_bySector_plot, REPORTEDLandings_bySector_plot, EstvReported_Aquaculture_plot,
   Estimated.Aquaculture_summary, Reported.Aquaculture_summary, Aquaculture_summary,
   SectorLandings_summary, SectorLandings_summary_REPORTED, 
   ManualLatLong_patch, Office_Names, test, vector) # Patch dataframes. Remove later.
```

# Save the Mex.Pacific_Landings_sector dataframe
```{r Save the Mex.Pacific_Landings_sector dataframe}
saveRDS(Mex.Pacific_Landings_sector, file="/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/Mex.Pacific_Landings_SectorSorted_2000.present_EMF.Apr152026.rds")
# 6406303 obs., 62 vars.
```


# Mariculture work as of APRIL 14-27 2026:

# Generate MARICULTURE List for all of the Mexican Pacific  ------------------------------

```{r Load Mex.Pacific_Landings_sector if needed}
Mex.Pacific_Landings_sector<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/Mex.Pacific_Landings_SectorSorted_2000.present_EMF.Apr152026.rds"))
# 6406303 obs., 62 vars.
```

```{r FILTER OUT MARICULTURE SECTORS OF INTEREST:}
table(Mex.Pacific_Landings_sector$sector_estimate_draft)

# FILTER OUT MARICULTURE SECTORS OF INTEREST: ----
# Let's assume that the records are most accurate since 2018.

Mariculture_Sectors<- c("marine: aquaculture")

Mex.Pacific_MaricultureProduction<- Mex.Pacific_Landings_sector %>% 
  filter(sector_estimate_draft %in% Mariculture_Sectors)
# 150037 obs., 62 vars.

range(Mex.Pacific_MaricultureProduction$year)
```

```{r Look into this first-draft list before proceeding}
# Taking the information I've gleaned here to refine my sector estimates above.
table(Mex.Pacific_MaricultureProduction$estimated_genus_name)
table(Mex.Pacific_MaricultureProduction$Major_Functional_Group)
```

```{r Generate a table of common aquaculture species since 2018}
Mex.Pacific_MajorAquacultureGenera_summary <- Mex.Pacific_MaricultureProduction %>%
  filter(year > 2018) %>%
  filter(!is.na(estimated_genus_name) & estimated_genus_name != "NEI" & estimated_genus_name != "unknown") %>%
  select(estimated_genus_name, Major_Functional_Group) %>%
  group_by(Major_Functional_Group, estimated_genus_name) %>%
  summarise(
    n_records_since2018 = n(),
    .groups = "drop"
  )

# Filter out the most common genera:
Mex.Pacific_MajorAquacultureGenera_summary<- Mex.Pacific_MajorAquacultureGenera_summary %>%
  filter(n_records_since2018 > 10)
## 11 obs.

AquacultureGenera_of_Interest<- unique(Mex.Pacific_MajorAquacultureGenera_summary$estimated_genus_name)
```

NOTE TO SELF: This isn't quite picking up the species I would have expected it to. For example, I would expect to see some Abalone mariculture, but I don't...
FOLLOW UP: Haliotis (abalone) is in there, but not after 2018
This would also explain the big wobble on aquaculture landings through time...
```{r Sanity check}
# SANITY CHECK: Where is Abalone?
subset<- Mex.Pacific_MaricultureProduction %>%
  filter(estimated_genus_name == "Haliotis")
# Only flagged in 2006. This suggests that my method of estimating sector is misisng catches. Makes sense if these catches weren't obviously flagged as aquaculture: not marked as aquaculture, no key words that would indicate aquaculture, etc.
```

```{r Investigate records for unexpected taxa}
table(Mex.Pacific_MajorAquacultureGenera_summary$estimated_genus_name)
table(Mex.Pacific_MaricultureProduction$reported_aquaculture_production) # is no = 1053
sum(is.na(Mex.Pacific_MaricultureProduction$reported_aquaculture_production)) # is na = 104204

# Investigate specific entries
subset1<- Mex.Pacific_MaricultureProduction %>%
  filter(year > 2017) %>%
  filter(estimated_genus_name == "Lutjanus") 
table(subset1$eu_name)
# Makes sense! All records from Ocean Farms in La Paz.

subset2<- Mex.Pacific_MaricultureProduction %>%
  filter(year > 2017) %>%
  filter(estimated_genus_name == "Argopecten") 
table(subset2$keywords_1)
# Makes sense! All records say "de cultivo" explicitly.

subset4<- Mex.Pacific_MaricultureProduction %>%
  filter(year > 2017) %>%
  filter(estimated_genus_name == "Dosinia") 
table(subset4$keywords_1)
# Makes sense! All records say "de cultivo" explicitly.

subset5<- Mex.Pacific_MaricultureProduction %>%
  filter(year > 2017) %>%
  filter(estimated_genus_name == "Megapitaria") 
table(subset5$keywords_1)
# Makes sense! All records say "de cultivo" explicitly.
```

```{r Summarize AquacultureGenera_of_Interest over time}
colnames(Mex.Pacific_MaricultureProduction)

Mex.Pacific_MaricultureProduction_summary <- Mex.Pacific_MaricultureProduction %>%
  filter(estimated_genus_name %in% AquacultureGenera_of_Interest) %>%
  select(year, estimated_genus_name, weight_kg_corrected, value) %>%
  group_by(estimated_genus_name, year) %>%
  summarise(
    annual_production_mt = sum(weight_kg_corrected, na.rm = TRUE) / 1000,
    annual_value_pesos   = sum(value, na.rm = TRUE),
    .groups = "drop"
  )
# 117 obs., 4 vars.
```

# Let's look into the Aquaculture AVISOS data Leonardo sent me
```{r AVISOS MAYORES}
Mex_AVISOS_2018 <- read.csv(
  "/home/shares/ohi/OHI_GOC/_raw_data/CONAPESCA/d2025/SharedByLeonardo/AVISOS_ MAYORES_MENORES_COSECHA_2018.csv",
  skip = 2,
  header = TRUE,
  fileEncoding = "latin1"
)

Mex_AVISOS_2018 <- clean_names(Mex_AVISOS_2018)

# 35 obs., 62115 vars.
```

Let's see if THESE avisos are = Juan Carlos' avisos for the year 2018, state of Sonora
```{r Comparing AVISOS from Leo and JC}
colnames(Mex_AVISOS_2018)
colnames(All.Mex_Landings_2000.present)

Sonora_Leo<- Mex_AVISOS_2018 %>%
  filter(aa_o_corte == 2018) %>%
  filter(nombre_estado == "SONORA")
# 29294 entries for 35 variables

Sonora_JC<- All.Mex_Landings_2000.present %>%
  filter(year == 2018) %>%
  filter(state == "SONORA")
# 29294 entries for 20 variables.


table(Sonora_Leo$nombre_principal)
     #      ALMEJA          ANCHOVETA              BAGRE            BANDERA            BAQUETA          BARRILETE          BERRUGATA 
     #          1126                614                 86                  1                749                  1               1012 
     #      CABRILLA            CALAMAR            CAMARON            CARACOL              CARPA              CAZON            CORVINA 
     #           331                 98               6803                874                 39                498               2133 
     #   GUACHINANGO              JAIBA              JUREL           LANGOSTA           LENGUADO               LISA             LOBINA 
     #           299               2569                243                 63               1275                133                  4 
     #      MACARELA               MERO            MOJARRA             OSTION              OTRAS            PAMPANO              PARGO 
     #           653                  2                247                 65               3573                 25                481 
     # PEPINO DE MAR             PIERNA              PULPO   RAYA Y SIMILARES             ROBALO              RONCO RUBIA Y VILLAJAIBA 
     #            16                115                324               1097                 21                  7                 29 
     #         RUBIO            SARDINA             SIERRA            TIBURON 
     #             1               2248                565                874 
                 
table(Sonora_JC$main_species_group)
     #        ALMEJA          ANCHOVETA              BAGRE            BANDERA            BAQUETA          BARRILETE          BERRUGATA 
     #          1126                614                 86                  1                749                  1               1012 
     #      CABRILLA            CALAMAR            CAMARON            CARACOL              CARPA              CAZON            CORVINA 
     #           331                 98               6803                874                 39                498               2133 
     #   GUACHINANGO              JAIBA              JUREL           LANGOSTA           LENGUADO               LISA             LOBINA 
     #           299               2569                243                 63               1275                133                  4 
     #      MACARELA               MERO            MOJARRA             OSTION              OTRAS            PAMPANO              PARGO 
     #           653                  2                247                 65               3573                 25                481 
     # PEPINO DE MAR             PIERNA              PULPO   RAYA Y SIMILARES             ROBALO              RONCO RUBIA Y VILLAJAIBA 
     #            16                115                324               1097                 21                  7                 29 
     #         RUBIO            SARDINA             SIERRA            TIBURON 
     #             1               2248                565                874 

# Excellent! AVISOS are the same from Leo or JC. Same data source, same ideas.
# Now let's look in to the aquaculture avisos specifically.
```

```{r AVISOS PRODUCCION - i.e. Aquaculture}
Aquaculture_AVISOS_2018 <- read.csv(
  "/home/shares/ohi/OHI_GOC/_raw_data/CONAPESCA/d2025/SharedByLeonardo/Produccion acuicola/AVISOS_PRODUCCION_2018.csv",
  skip = 1,
  header = TRUE,
  fileEncoding = "latin1"
)

Aquaculture_AVISOS_2018 <- clean_names(Aquaculture_AVISOS_2018)
# 5014 obs., 24 vars.
```

```{r Clean up the Aquaculture_AVISOS_2018 df}
options(scipen = 999)

unique(Aquaculture_AVISOS_2018$unidad_medida)
table(Aquaculture_AVISOS_2018$unidad_medida)
# [1] "PIEZAS"   "MILES"    "MILLONES" ""

conversion_factor <- c(
  "PIEZAS" = 1,
  "MILES" = 1e3,
  "MILLONES" = 1e6
)

Aquaculture_AVISOS_2018 <- Aquaculture_AVISOS_2018 %>%
  mutate(
    cantidad_clean = as.numeric(
      gsub(",", "", trimws(cantidad))
    ),
    unidad_medida = toupper(trimws(unidad_medida)),
    cantidad_std = case_when(
      unidad_medida == "PIEZAS" ~ cantidad_clean,
      unidad_medida == "MILES" ~ cantidad_clean * 1e3,
      unidad_medida == "MILLONES" ~ cantidad_clean * 1e6,
      TRUE ~ NA_real_
    )
  )

Aquaculture_AVISOS_2018$precio_pesos_clean <- as.numeric(
  gsub(",", "",
       trimws(Aquaculture_AVISOS_2018$precio_pesos)
  )
)

Aquaculture_AVISOS_2018$valor_pesos_clean <- as.numeric(
  gsub(",", "",
       trimws(Aquaculture_AVISOS_2018$valor_pesos)
  )
)
```

```{r Explore Aquaculture_AVISOS_2018}
colnames(Aquaculture_AVISOS_2018)
table(Aquaculture_AVISOS_2018$nombre_especie)

Sinaloan_Shrimp_Aquaculutre_2018<- Aquaculture_AVISOS_2018 %>%
  filter(ano_corte == 2018) %>%
  filter(nombre_especie == "CAMARON BLANCO") %>%
  filter(nombre_estado == "SINALOA")
# 519 entries for 24 variables.

# "AQUAPACIFIC, SA DE CV" is an economic unit listed here. Let's see if we can find it in the other dataset...
```

```{r Compare economic unit names from Mex_AVISOS_2018 to Mex.Pacific_Landings_sector}
colnames(Mex_AVISOS_2018)

shared_vals_All <- intersect(All.Mex_Landings_2000.present$eu_name,
                         Aquaculture_AVISOS_2018$unidad_economica)
print(shared_vals_All)
# 147 shared values for all of Mexico

shared_vals_Pacific <- intersect(Mex.Pacific_Landings_sector$eu_name,
                         Aquaculture_AVISOS_2018$unidad_economica)
print(shared_vals_Pacific)
# 77 shared values for the Mexican Pacific

shared_vals_Pacific.Flagged <- intersect(Mex.Pacific_MaricultureProduction$eu_name,
                         Aquaculture_AVISOS_2018$unidad_economica)
print(shared_vals_Pacific.Flagged)
# 31 shared values for the Mexican Pacific that I was able to estimate as mariculture using my flag system alone. This means that my flags are missing some economic units that probably are producing aquaculture.

test<- Mex_AVISOS_2018 %>%
  filter(aa_o_corte == 2018) %>%
  filter(unidad_economica == "EARTH OCEAN FARMS S DE RL CV")
# 31 entries for 20 variables.

test2<- Mex.Pacific_Landings_sector %>%
  filter(year == 2018) %>%
  filter(eu_name == "EARTH OCEAN FARMS S DE RL CV")
# 31 entries for 62 variables. Excellent!

test3<- Mex.Pacific_Landings_sector %>%
  filter(eu_name == "JORGE ALBERTO GALVAN MORA")
# 5 entries for 20 variables.
# Interesting. These entries are not marked as aquaculture, but there are key words that my filters picked up. Good!
```

```{r Which aquaculture producers is my flag system overlooking?}
shared_vals_Pacific <- intersect(Mex.Pacific_Landings_sector$eu_name,
                         Aquaculture_AVISOS_2018$unidad_economica)
print(shared_vals_Pacific)
# 77 shared values for the Mexican Pacific

shared_vals_Pacific.Flagged <- intersect(Mex.Pacific_MaricultureProduction$eu_name,
                         Aquaculture_AVISOS_2018$unidad_economica)
print(shared_vals_Pacific.Flagged)
# 31 shared values for the Mexican Pacific that I was able to estimate as mariculture using my flag system alone. This means that my flags are missing some economic units that probably are producing aquaculture.

EUs_overlooked <- setdiff(shared_vals_Pacific, shared_vals_Pacific.Flagged)
print(EUs_overlooked)

Landings_overlooked<- Mex.Pacific_Landings_sector %>%
  filter(eu_name %in% EUs_overlooked)
min(Landings_overlooked$year)
max(Landings_overlooked$year)
# 11494 obs. overlooked across all years.

table(Landings_overlooked$sector_estimate_draft)

              # marine: fishery sector yet to be classified                               marine: large-scale capture 
              #                                        5476                                                      2190 
              #                 marine: small-scale capture                                       marine: aquaculture 
              #                                        2439                                                         0 
              #                     freshwater: aquaculture                           freshwater: small-scale capture 
              #                                        1109                                                        16 
              #             freshwater: large-scale capture           freshwater: fishery sector yet to be classified 
              #                                           0                                                        17 
              #           freshwater or marine: aquaculture                 freshwater or marine: small-scale capture 
              #                                         172                                                        34 
              #   freshwater or marine: large-scale capture freshwater or marine: fishery sector yet to be classified 
              #                                           0                                                        41 

# Looks like: a lot of these were CORRECTLY flagged as aquaculture just freshwater or yet to be classified.
# About 40% (2439 + 2190=4629) of these "overlooked" observations might be INCORRECTLY marked as small-scale or large-scale industrial fisheries. Let's look into those.
# Turns out these entries are legit, probably not aquaculture. At least there were marked as NOT aquaculture by the fishers, even if the EU name might suggest aquaculture. Cool!

test4<- Landings_overlooked %>%
  filter(sector_estimate_draft=="marine: large-scale capture")
# 2190 obs.
table(test4$reported_fleet)
table(test4$Major_Functional_Group)
# These are all landings that were explicitly reported as large-scale. I think my filters might be working...

test5<- Landings_overlooked %>%
  filter(sector_estimate_draft=="marine: small-scale capture")
# 2439 obs.
table(test5$reported_fleet)
table(test5$Major_Functional_Group)
# These are all landings that were explicitly reported as small-scale and NOT aquaculture. I think my filters are working as expected here.
```

- Okay, so what I'm getting from this exercise is that my flags are actually working pretty well. 
- My sector estimates probably aren't too far off.
- BUT ALSO, there could be important aquaculture entries excluded from the CONAPESCA AVISOS_MAYORES_MENORES_COSECHA file because the production wasn't actually officially "harvested".
- To this point, Leonardo writes: "To clarify what these "Produccion "data includes, I am attaching the template which is mandatory for aquaculturists, aquaculture laboratories, and fishers to report the quantity and type of organisms they have cultivated or collected from nature. You'll see number of organisms and life stage of this "production" but how much of this was harvested? that you'll find it in the integrated databases as "cosecha" (harvest). As mentioned, one thing is to produce in the lab or sometimes collect from the field, and another thing is to grow, harvest and sell the products. There are several species that are "produced" in laboratories for research goals and looking to develop the cultivation but just a few are harvested and reported."
- Based on all this, I think it makes the most sense to compare the FAO aquaculture data to the data listed in the PRODUCCION AVISOS (Aquauculture Avisos) files, not the COSECHA AVISOS which is confounded with other types of fisheries data.

```{r Summarize Aquaculture_AVISOS_2018 versus FAO}
colnames(Aquaculture_AVISOS_2018)

# Summarized Production for Aquaculture_AVISOS_2018
table(Aquaculture_AVISOS_2018$fase)
# I think we want to filter out Post-Larva and Juveniles

Estimated_MAR_2018 <- Aquaculture_AVISOS_2018 %>%
  filter(fase %in% c("POSTLARVA", "JUVENILES")) %>%
  group_by(nombre_especie) %>%
  summarise(total_cantidad = sum(cantidad_clean, na.rm = TRUE),
            total_cantidad_std = sum(cantidad_std, na.rm = TRUE), 
            .groups = "drop")

# Summarized Production for FAO Aquaculture (2018)
```

# Compare to FAO aquaculture data [UNDER CONSTRUCTION Apr 24 2026]

```{r Load in FAO_FamilyGenus_key_filtered}
FAO_FamilyGenus_key0<- read.csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/FAO_FamilyGenusKey_COMPLETE_Apr042026.csv")
# 193 obs., 7 vars.

# Trim key and simplify (as needed):
colnames(FAO_FamilyGenus_key0)
FAO_FamilyGenus_key_filtered<- FAO_FamilyGenus_key0 %>%
  select(estimated_genus_name, Major_Functional_Group, 
         estimated_family_name= new_estimated_family_name) %>%
  distinct(estimated_family_name, .keep_all = TRUE)
# 105 obs., 3 vars.
```

```{r Look into FAO aquaculture data}
# Load data
FAO_Landings_Data0<- read.csv('/home/ferrer/ohi_goc_prep/FP/CMSY_Analysis/CMSY_Revised_April2026/Generating_Model_Priors/Priors_Data/FAOMexicoData_19762023_downloadedApr042026.csv')
# FAO timeseries data organized with Stock (taxa) as # columns and years as # rows.
    
FAO_Landings_Data<- FAO_Landings_Data0

# We've got to do a little work on these data before they're ready to use.
colnames(FAO_Landings_Data)

# Clean column names
clean_names <- colnames(FAO_Landings_Data) %>%
  # Remove trailing periods
  str_replace_all("\\.$", "") %>%
  # Replace multiple periods with a single underscore
  str_replace_all("\\.+", "_") %>%
  # Remove leading X_ for years
  str_replace("^X_(\\d+)$", "\\1") %>%
  # Convert to lowercase
  tolower()

# Apply cleaned names to the dataframe
colnames(FAO_Landings_Data) <- clean_names

# Delete the columns that start with s or s_, we don't need these.
FAO_Landings_Data <- FAO_Landings_Data %>%
  select(-starts_with("s"))

# Filter out aquaculture production:
table(FAO_Landings_Data$detailed_production_source_name)

FAO_Landings_filtered<- FAO_Landings_Data %>%
  filter(detailed_production_source_name %in% c("Aquaculture production (marine)", 
                                                "Aquaculture production (brackishwater)"))

# Delete other columns we don't need:
colnames(FAO_Landings_filtered)
FAO_Landings_filtered<- FAO_Landings_filtered %>%
  select(-c(country_name, fao_major_fishing_area_name, detailed_production_source_name,
            unit_name, unit))

# Transpose these data such that years are rows per taxa (and rename species column):
FAO_AquacultureProduction <- FAO_Landings_filtered %>%
  # Rename species column
  rename(estimated_family_name = asfis_species_scientific_name) %>%
  
  # Step 1: go long
  pivot_longer(
    cols = -estimated_family_name,
    names_to = "year",
    values_to = "production_mt"
  ) %>%
  
  # Step 2: clean year
  mutate(year = as.integer(year),
         production_mt = as.numeric(production_mt))

FAO_AquacultureProduction <- FAO_AquacultureProduction %>%
  arrange(year) %>%
  filter(year >= min(year[production_mt != 0], na.rm = TRUE))

# How many families are there in this database?
unique(FAO_AquacultureProduction$estimated_family_name)
# 24 family, excluding "others"
```

```{r Keep workspace clean}
rm(FAO_Landings_Data)
rm(FAO_Landings_filtered)
```

```{r Load in FAO_FamilyGenus_key_filtered}
FAO_FamilyGenus_key0<- read.csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/FAO_FamilyGenusKey_COMPLETE_Apr042026.csv")
# 193 obs., 7 vars.

# Trim key and simplify (as needed):
colnames(FAO_FamilyGenus_key0)
FAO_FamilyGenus_key_filtered<- FAO_FamilyGenus_key0 %>%
  select(Major_Functional_Group, 
         estimated_family_name= new_estimated_family_name) %>%
  distinct(estimated_family_name, .keep_all = TRUE)
# 105 obs., 2 vars.
```

```{r Merge FAO_FamilyGenus_key_filtered to FAO data}
# Make family lowercase so the merge goes smoothly:
FAO_FamilyGenus_key_filtered$estimated_family_name <- tolower(FAO_FamilyGenus_key_filtered$estimated_family_name)

FAO_AquacultureProduction$estimated_family_name <- tolower(FAO_AquacultureProduction$estimated_family_name)

# Merge in key:
FAO_AquacultureProduction <- left_join(FAO_AquacultureProduction, FAO_FamilyGenus_key_filtered, 
                           by = "estimated_family_name")

```

```{r Plot FAO aquaculture}
ggplot(FAO_AquacultureProduction, aes(x = year, y = production_mt)) +
  geom_line(color = "steelblue") +
  facet_wrap(~ estimated_family_name, scales = "free_y") +
  labs(x = "Year", y = "Production (mt)",
       title = "FAO Aquaculture Production (by family)") +
  theme_minimal()
```

```{r Summarize FAO data in all years}
colnames(FAO_AquacultureProduction)

# Summarized Production for FAO_AquacultureProduction in all years
FAO_MAR_production <- FAO_AquacultureProduction %>%
 # filter(year == 2018) %>%
  group_by(year, estimated_family_name) %>%
  summarise(total_production_mt = sum(production_mt, na.rm = TRUE),
            .groups = "drop")

# Summarized Production for FAO Aquaculture (2018)
```

```{r Summarize my COSECHA data in all years}
colnames(Mex.Pacific_MaricultureProduction)

# Summarized Production for CONAPESCA cosecha production in all years.
COSECHA_MAR_production <- Mex.Pacific_MaricultureProduction %>%
# filter(year == 2018) %>%
  group_by(year, estimated_family_name) %>%
  summarise(total_est_production_mt = (sum(weight_kg_corrected, na.rm = TRUE)/1000),
            .groups = "drop")

# Summarized Production for FAO Aquaculture (2018)
```

```{r Aquaculture family of interest based on most common CONAPESCA landings}
colnames(Mex.Pacific_MaricultureProduction)

AquacultureFamilies_of_Interest <- Mex.Pacific_MaricultureProduction %>%
  filter(estimated_genus_name %in% AquacultureGenera_of_Interest) %>%
  distinct(estimated_family_name) %>%
  pull(estimated_family_name)

print(AquacultureFamilies_of_Interest)
```

```{r Merge these two dataframes}
# Step 1: Clean both datasets
clean_names <- function(x) {
  x %>%
    str_trim() %>%
    str_to_lower()
}

AquacultureFamilies_of_Interest<- clean_names(AquacultureFamilies_of_Interest)

FAO_MAR_production <- FAO_MAR_production %>%
  mutate(
    estimated_family_name = clean_names(estimated_family_name),
    year = as.integer(year)
  ) %>%
  rename(FAO_est_production = total_production_mt)

COSECHA_MAR_production <- COSECHA_MAR_production %>%
  mutate(
    estimated_family_name = clean_names(estimated_family_name),
    year = as.integer(year)
  ) %>%
  rename(CONAPESCA_est_production = total_est_production_mt)

# Step 2: Check overlap (important sanity check)
intersect(
  FAO_MAR_production$estimated_family_name,
  COSECHA_MAR_production$estimated_family_name
)

# Step 3: Merge (recommended approach)
merged_df <- FAO_MAR_production %>%
  left_join(
    COSECHA_MAR_production,
    by = c("year", "estimated_family_name")
  )

# I want to filter an manipulate merged_df in a few ways. First, I want to filter only years since 2000. then, I want to drop any rows where FAO_est_production and CONAPESCA_est_production both equal zero or NA. Finally, I want to estimate the biomass discrepancy between FAO_est_production and CONAPESCA_est_production in a new column.

merged_df <- merged_df %>%
  filter(year >= 2018) %>%
  filter(estimated_family_name %in% AquacultureFamilies_of_Interest) %>%
  filter(
    !(is.na(FAO_est_production) & is.na(CONAPESCA_est_production)) &
    !(FAO_est_production == 0 & CONAPESCA_est_production == 0)
  ) %>%
  mutate(
    biomass_discrepancy = FAO_est_production - CONAPESCA_est_production,
    percent.biomass_discrepancy =
      round(((FAO_est_production - CONAPESCA_est_production) /
               CONAPESCA_est_production) * 100, 2)
  )
```



General plan: 
1. Load in MAR seafood watch scores
1B. Prepare seafood watch scores as needed

2. Load in MAR tonnage from CONAPESCA
2B. Prepare CONAPESCA tonnage as needed

3. Load in MAR tonnage from FAO
3B. Prepare FAO tonnage as needed

4. Weight the CONAPESCA tonnage by the "correction factor" from the FAO, IF NEEDED.
If within 10% [25%], comfortable going with CONAPESCA data.
--> Bring a comparison table to meeting.

5. Multiply MAR seafood watch scores by tonnage per region to get a "weighted tonnage"

If CONAPESCA data is really off, still have FAO data and proportions
Which way should we go based on these numbers?
