---
title: 'OHI GoC 2024: Lasting Special Places (LSP)'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor:
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options:
  chunk_output_type: console
---

# Sentido de Pertenencia: Lasting Special Places

## Script 6: special habitats (coquina, coral, seagrass, mangrove)

In this script I score four habitat types the EWG cares about as polygons. I clip each map to OHI regions, then ask how well each habitat patch is protected. When a patch sits inside an ANP I use SIMEC (or a Ramsar plan rule if SIMEC is missing). When it does not, I still apply a legal floor from NOMs and related statutes, so habitat never scores below what the law already implies.

**What I do**

1. Build region clipped polygon layers for coquina, coral (Region 4 only), seagrass, and mangrove.
2. Load shared protected areas, SIMEC, and Ramsar tables.
3. For each habitat, buffer, join parks, score, keep the better of park score and NOM floor.
4. Average across habitats by region and write CSV plus intermediate geopackages.

```{r}
library(tidyverse)
library(here)
library(sf)
library(mapview)
library(stringdist)
library(stringi)
```

## Build habitat layers for scoring

I read TNC and HEM maps, intersect them with OHI regions, and tag readable `lsp_name` strings so I can trace rows later.

### Coquina reefs

```{r}
tnc_dir <- "/home/shares/ohi/OHI_GOC/_raw_data/TNC_geoportal_fishing_refuge_zones/d2026"
int_dir <- "/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/int"
dir.create(int_dir, recursive = TRUE, showWarnings = FALSE)

ohi_rgns <- st_read(here("spatial/ohi_regions/goc_ohi_rgns.shp"), quiet = TRUE)

coquina_reefs_tnc <- st_read(
  here(tnc_dir, "Arrecifes_de_coquina_-7828932449339036258.geojson"),
  quiet = TRUE
)

coquina_reef_lsps <- st_intersection(
  st_transform(coquina_reefs_tnc, st_crs(ohi_rgns)),
  ohi_rgns
) %>%
  mutate(
    lsp_name = paste0(
      FID, id, ": ",
      "Coquina Reefs in Libertad, Bahia Lobos, and Peñasco - AHW"
    )
  ) %>%
  dplyr::select(rgn_id, lsp_name)

mapview(coquina_reef_lsps, zcol = "rgn_id")
write_csv(coquina_reef_lsps, file.path(int_dir, "prep_special_habitat_coquina.csv"))
```

### Coral reefs

I only keep Region 4 (Cabo Pulmo, Los Frailes, Bahía San Gabriel), following expert notes.

```{r}
coral_reefs_tnc <- st_read(
  here(tnc_dir, "Arrecifes_de_coral_-7914658441806850050.geojson"),
  quiet = TRUE
)

# https://www.arcgis.com/home/item.html?id=ef3ac8f679674e60b0da5f57e3997c27
# "La capa se creó a través de la compilación, digitalización y combinación de mapas de arrecifes de coral para el Golfo de California provenientes de las siguientes fuentes:Bezaury-Creel &amp; Castro Campos. 2009. Base, Ulloa et al, ReefBase: A Global Information System for Coral Reefs. December, 2016. http://www.reefbase.org World Resources Institute, Reefs at Risk Revisited, 2016 Munguia-Vega et al. 2018. Ecological guidelines for designing a network of marine reserves in the unique biophysical environment of the Gulf of California. Reviews in Fish Biology and Fisheries 28: 749–776Reyes-Bonilla. 2003. Coral reefs of the pacific coast of Mexico. In: Latin American coral reefs: 331-349. J. Cortes (ed). Amsterdam: Elsevier. Reyes-Bonilla &amp; Lopez-Perez A. 2009. Corals and coral-reef communities in the Gulf of California. In: Atlas of coastal ecosystems in the western Gulf of California: 45-57. M. E. Johnson, J. Ledesma-Vazquez (eds). Tucson: The University of Arizona Press."

coral_reefs_lsps <- st_intersection(
  st_transform(coral_reefs_tnc, st_crs(ohi_rgns)),
  ohi_rgns
) %>%
  filter(rgn_id %in% "Region_4") %>%
  mutate(
    lsp_name = paste0(
      FID, ": ",
      "Only coral reefs in Gulf (Cabo Pulmo/Los Frailes & Bahía San Gabriel on Is. Espíritu Santo). RCB, BW"
    )
  ) %>%
  dplyr::select(rgn_id, lsp_name)

write_csv(coral_reefs_lsps, file.path(int_dir, "prep_special_habitat_coral.csv"))
```

### Seagrass

```{r}
seagrass_tnc <- st_read(
  here(tnc_dir, "Pastos_marinos_4840481632930554856.geojson"),
  quiet = TRUE
)

seagrass_lsps <- st_intersection(
  st_transform(seagrass_tnc, st_crs(ohi_rgns)),
  ohi_rgns
) %>%
  mutate(
    lsp_name = paste0(
      Habitat, ": ", Referencia, " - ",
      "Massive seagrass beds (Zostera marina), RCB, BW. The largest and most stable annual seagrass beds (Zostera marina) are in the eastern Pacific. The eelgrass in the Gulf of California is 100% annual, no perennial, and is considered a subspecies, JTC, BW"
    )
  ) %>%
  dplyr::select(rgn_id, lsp_name)

mapview(seagrass_lsps)
write_csv(seagrass_lsps, file.path(int_dir, "prep_special_habitat_seagrass.csv"))
```

### Mangroves

```{r}
mangroves_hem <- st_read(
  here("/home/shares/ohi/OHI_GOC/_raw_data/mangrove_Hem/USV_AGRUPADO/USV_AGRUPADO.shp"),
  quiet = TRUE
) %>%
  janitor::clean_names() %>%
  filter(descrip_ag %in% "MANGLAR")

mangrove_lsps <- st_intersection(
  st_transform(mangroves_hem, st_crs(ohi_rgns)),
  ohi_rgns
) %>%
  mutate(
    lsp_name = paste0(id, ": ", "Mangrove coastal lagoons, JTC."),
    feature_id = row_number()
  ) %>%
  dplyr::select(rgn_id, lsp_name, feature_id)

write_csv(mangrove_lsps, file.path(int_dir, "prep_special_habitat_mangrove.csv"))
```

## Shared protected area and score tables

```{r}
all_protected <- st_read(
  here(
    "/home/shares/ohi/OHI_GOC/spatial/protected_areas_wdpa_conanp_and_hem",
    "all_goc_protected_areas_v2_2025.shp"
  ),
  quiet = TRUE
)

lsp_google_sheets_dir <- "/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/google_sheets"

i_efectividad_raw <- read_csv(
  file.path(lsp_google_sheets_dir, "simec_i_efectividad_anp_2025.csv"),
  show_col_types = FALSE
) %>%
  janitor::clean_names()

ramsar_management_plans <- read_csv(
  file.path(lsp_google_sheets_dir, "ramsar_site_management_plan_presence.csv"),
  show_col_types = FALSE
)
```

## NOM and legal floors by habitat

These numbers are the floor when a patch is outside a scored park. Mangrove, seagrass, and coral have stronger statute support. Coquina is weaker, so its floor is lower.

```{r}
# NOM baseline condition for each special habitat
NOM_MANGROVE_CONDITION <- 70

# Ley General de Vida Silvestre Art. 60ter (prohibition of cutting/filling mangroves); Código Penal Art. 420bis (potential for prison); NOM-022-SEMARNAT-2003; NOM-059; LGEEPA	80	strong	"""Article 60 TER.- The removal, filling, transplanting, pruning, or any work or activity that affects the integrity of the hydrological flow of the mangrove is prohibited."" https://www.diputados.gob.mx/LeyesBiblio/pdf/LGVS.pdf 
# ""A prison sentence of two to ten years and a fine equivalent to three hundred to three thousand days' wages will be imposed on anyone who unlawfully....Damage, dry up or fill wetlands, mangroves, lagoons, estuaries or swamps; Damage reefs; Introduce or release into the natural environment any specimen of exotic flora or fauna that harms an ecosystem, or that hinders, alters or affects native or migratory species in the natural cycles of their reproduction or migration, or; Cause a fire in a forest, jungle, natural vegetation or forest lands, which damages natural elements, flora, fauna, ecosystems or the environment. An additional penalty of up to two years in prison and up to one thousand additional days of fines will be applied when the conduct described in this article is carried out in or affects a protected natural area, or the perpetrator or accomplice of the crime provided for in section IV, carries out the conduct to obtain a profit or economic benefit. https://mexico.justia.com/federales/codigos/codigo-penal-federal/libro-segundo/titulo-vigesimoquinto/capitulo-segundo/#articulo-420-bis 
# https://www.dof.gob.mx/nota_detalle.php?codigo=697013&fecha=10/04/2003#gsc.tab=0
# https://bmasf.mx/en/environmental-impact-assessment-in-mexico/"

NOM_SEAGRASS_CONDITION <- 70
# NOM-022 coastal wetlands standard; NOM-059 (Zostera marina: Special Protection (Pr)); LGEEPA; General Wildlife Law; Ramsar Convention	80	strong

NOM_CORAL_CONDITION <- 70
# Strict vessel liability on reefs under LGEEPA and the Federal Law for Environmental Liability (LFRA); General Wildlife Law; CITES Appendix II coral spp.; may be under NOM-059 but none of the coral species that have occurred in the GoC are listed currently.	80	strong	"Environmental Impact Assessment in Mexico - Bustamante + Freyre
# A prison sentence of two to ten years and a fine equivalent to three hundred to three thousand days' wages will be imposed on anyone who unlawfully....Damage, dry up or fill wetlands, mangroves, lagoons, estuaries or swamps; Damage reefs; Introduce or release into the natural environment any specimen of exotic flora or fauna that harms an ecosystem, or that hinders, alters or affects native or migratory species in the natural cycles of their reproduction or migration, or; Cause a fire in a forest, jungle, natural vegetation or forest lands, which damages natural elements, flora, fauna, ecosystems or the environment. An additional penalty of up to two years in prison and up to one thousand additional days of fines will be applied when the conduct described in this article is carried out in or affects a protected natural area, or the perpetrator or accomplice of the crime provided for in section IV, carries out the conduct to obtain a profit or economic benefit. 
# https://mexico.justia.com/federales/codigos/codigo-penal-federal/libro-segundo/titulo-vigesimoquinto/capitulo-segundo/#articulo-420-bis
# https://axaxl.com/fast-fast-forward/articles/mexicos-new-environmental-law 
# ""Strict liability holds companies responsible for accidents that occur, despite the best oversight and maintenance. If a perfectly maintained ship is operating on a coral reef, and the ship accidentally leaks hazardous waste and damages the reef, it isn’t the result of error or negligence. The ship still caused damage to the reef, and the ship company will be held responsible. Under strict liability, the plaintiff does not have to prove that damages were the result of negligence. He only has to prove that the damages occurred, and that the defendant was somehow responsible.  """

NOM_COQUINA_CONDITION <- 30
# No dedicated habitat statute; possible indirect LGEEPA coverage if NOM-059 species present; general environmental impact rules apply	30	weak
```

## Score function: parks, SIMEC, Ramsar, then NOM floor

Let's write one helper so all four habitats follow the same steps.

```{r}
normalize_name <- function(x) {
  x %>%
    stringi::stri_trans_general("Latin-ASCII") %>%
    str_to_upper() %>%
    str_remove_all(
      "\\bPARQUE NACIONAL\\b|\\bPARQUE\\b|\\bNACIONAL\\b|\\bZONA MARINA\\b|\\bARCHIPIELAGO\\b"
    ) %>%
    str_squish()
}

i_efectividad_norm <- i_efectividad_raw %>%
  mutate(nombre_norm = normalize_name(nombre))

assign_habitat_condition <- function(habitat_sf, name_col, habitat_label, nom_condition) {
  message("Processing: ", habitat_label, " (", nrow(habitat_sf), " features)")

  habitat_buffered <- st_buffer(
    st_transform(habitat_sf, st_crs(all_protected)),
    dist = 100
  )

  habitat_protected <- st_join(
    habitat_buffered,
    all_protected %>% dplyr::select(name, geometry),
    join = st_intersects,
    left = TRUE
  ) %>%
    mutate(in_protected_area = !is.na(name)) %>%
    rename(protected_area_name = name)

  habitat_norm <- habitat_protected %>%
    mutate(protected_area_norm = normalize_name(protected_area_name))

  matches <- stringdist::amatch(
    habitat_norm$protected_area_norm,
    i_efectividad_norm$nombre_norm,
    method = "jw",
    maxDist = 0.2
  )

  habitat_matched <- habitat_norm %>%
    mutate(
      nombre_matched = i_efectividad_norm$nombre[matches],
      nombre_norm_matched = i_efectividad_norm$nombre_norm[matches]
    )

  # QA table of fuzzy matches (as_tibble + plain print avoids print(n = Inf)
  # errors on sf-derived data frames / odd na.print options)
  match_check <- habitat_matched %>%
    st_drop_geometry() %>%
    distinct(protected_area_name, protected_area_norm, nombre_matched) %>%
    as_tibble()
  print(match_check)

  habitat_w_simec <- habitat_matched %>%
    left_join(i_efectividad_raw, by = c("nombre_matched" = "nombre")) %>%
    left_join(
      ramsar_management_plans,
      by = c("protected_area_name" = "wdpa_orig_name")
    ) %>%
    mutate(
      simec_score = ifelse(!is.na(i_efectividad), as.numeric(i_efectividad), NA_real_),
      ramsar_score = case_when(
        if_ramsar_has_management_plan == "y" ~
          as.numeric(max(i_efectividad_raw$i_efectividad, na.rm = TRUE)),
        if_ramsar_has_management_plan == "in_progress" ~ 60,
        if_ramsar_has_management_plan == "n" ~
          as.numeric(min(i_efectividad_raw$i_efectividad, na.rm = TRUE)),
        TRUE ~ NA_real_
      ),
      protected_area_condition = case_when(
        !is.na(simec_score) | !is.na(ramsar_score) ~ pmax(
          tidyr::replace_na(simec_score, -Inf),
          tidyr::replace_na(ramsar_score, -Inf)
        ),
        TRUE ~ NA_real_
      ),
      overall_condition = pmax(protected_area_condition, nom_condition, na.rm = TRUE),
      habitat_type = habitat_label
    )

  habitat_w_simec %>%
    group_by(rgn_id, {{ name_col }}) %>%
    summarize(
      overall_condition = max(overall_condition, na.rm = TRUE),
      protected_area_condition = max(protected_area_condition, na.rm = TRUE),
      in_protected_area = any(in_protected_area),
      habitat_type = first(habitat_type),
      geometry = st_union(geometry),
      .groups = "drop"
    )
}
```

## Run each habitat

```{r}
coquina_reefs_condition <- assign_habitat_condition(
  coquina_reef_lsps, lsp_name, "Coquina Reef", NOM_COQUINA_CONDITION
)
rgn_coquina_reefs_condition <- coquina_reefs_condition %>%
  st_drop_geometry() %>%
  group_by(rgn_id) %>%
  summarize(condition = mean(overall_condition, na.rm = TRUE), n_features = n(), .groups = "drop")

coral_reefs_condition <- assign_habitat_condition(
  coral_reefs_lsps, lsp_name, "Coral Reef", NOM_CORAL_CONDITION
)
rgn_coral_reefs_condition <- coral_reefs_condition %>%
  st_drop_geometry() %>%
  group_by(rgn_id) %>%
  summarize(condition = mean(overall_condition, na.rm = TRUE), n_features = n(), .groups = "drop")

seagrass_condition <- assign_habitat_condition(
  seagrass_lsps, lsp_name, "Seagrass (Zostera marina)", NOM_SEAGRASS_CONDITION
)
rgn_seagrass_condition <- seagrass_condition %>%
  st_drop_geometry() %>%
  group_by(rgn_id) %>%
  summarize(condition = mean(overall_condition, na.rm = TRUE), n_features = n(), .groups = "drop")

mangroves_condition <- assign_habitat_condition(
  mangrove_lsps, lsp_name, "Mangrove", NOM_MANGROVE_CONDITION
)
rgn_mangroves_condition <- mangroves_condition %>%
  st_drop_geometry() %>%
  group_by(rgn_id) %>%
  summarize(condition = mean(overall_condition, na.rm = TRUE), n_features = n(), .groups = "drop")
```

## Combine habitats by region and write

```{r}
rgn_special_habitats_condition <- bind_rows(
  rgn_coquina_reefs_condition,
  rgn_coral_reefs_condition,
  rgn_seagrass_condition,
  rgn_mangroves_condition
) %>%
  group_by(rgn_id) %>%
  summarize(
    condition = mean(condition, na.rm = TRUE),
    n_features = n(),
    .groups = "drop"
  )

print(rgn_special_habitats_condition)

out_dir <- "/home/shares/ohi/OHI_GOC/goal_prep/sp/lsp/v2025/output"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

write_csv(
  rgn_special_habitats_condition,
  file.path(out_dir, "condition_region_special_habitats.csv")
)

st_write(coquina_reefs_condition, file.path(int_dir, "condition_special_habitat_coquina.gpkg"), delete_dsn = TRUE)
st_write(coral_reefs_condition, file.path(int_dir, "condition_special_habitat_coral.gpkg"), delete_dsn = TRUE)
st_write(seagrass_condition, file.path(int_dir, "condition_special_habitat_seagrass.gpkg"), delete_dsn = TRUE)
st_write(mangroves_condition, file.path(int_dir, "condition_special_habitat_mangrove.gpkg"), delete_dsn = TRUE)
```








