################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```

## clean_names_lowercase function
```{r Load clean_names_lowercase function}
# Takes text and standardizes it, which we will need for merging later on:

clean_names_lowercase <- function(x) {

  x <- iconv(x, from = "", to = "UTF-8", sub = "")
  x <- stringi::stri_trans_general(x, "Latin-ASCII")

  x %>%
    tolower() %>%
    trimws()
}
```

## lowercase_function function
```{r lowercase_function}
# Turns all entries lower-case
lowercase_function<- function(x){
  if(is.character(x)){
    tolower(x)
  } else if (is.factor(x)){
    tolower(as.character(x))
  } else {
    x
  }
}
```


# PICK 3: CMSY Stock Assesments - genus level ------------------------------
- This task represents a considerable amount of work that was not completed in this file.
- To learn how the CMSY scores were generated, review this file:
CMSY_FinalAnalysis_May2026.Rmd.
- I used this file to generate the model priors, however, the actual CMSY script (RunCMSY_May282026.R) was run on my personal machine.

Important Notes: 
- I ran CMSY at genus level, since I actually think it’s a safer more reliable bet than species-level info and because – in many cases – fishermen seem to target taxa of the same genus interchangeably.

```{r Load in CMSY asessment results}
# Load in B/Bmsy output from CMSY:
CMSY_Timeseries0<- read.csv("/home/ferrer/ohi_goc_prep/FP/CMSY_Analysis/CMSY_FinalLook_May2026/Model_Output_Files/CMSY_Assessment_Timeseries_May282026.csv")
# 1569 obs., 14 vars.

colnames(CMSY_Timeseries0)

CMSY_Timeseries0<- CMSY_Timeseries0 %>%
  select(-c(X, cinfo.stock.i.))

table(CMSY_Timeseries0$stock)
```

```{r Load other useful dataframes: CMSY timeseries and key data}
# Useful genus key:
GenusFamilyGroup.Key0<- read.csv("~/ohi_goc_prep/FP/CONAPESCA_Keys/CONAPESCA_Keys/GenusFamilyGroupKey_COMPLETE_Apr042026.csv")
# 193 obs., 7 vars.

GenusFamilyGroup.Key<- GenusFamilyGroup.Key0

# Clean up the GenusFamilyGroup.Key key:
colnames(GenusFamilyGroup.Key)

GenusFamilyGroup.Key<- GenusFamilyGroup.Key %>%
  select(Major_Functional_Group, estimated_genus_name, 
         estimated_family_name= new_estimated_family_name,
         family_est_provenance= new_family_est_reasoning)
```

```{r Append info about Major_Functional_Group}
CMSY_Timeseries<- CMSY_Timeseries0 %>% 
  select(stock, yr,
         B, B.Bmsy, lcl.B.Bmsy, ucl.B.Bmsy,
         F.Fmsy, lcl.F.Fmsy, ucl.F.Fmsy)

# Convert the relevant columns to lower case
CMSY_Timeseries <- CMSY_Timeseries %>%
  mutate(estimated_genus_name = tolower(stock))

GenusFamilyGroup.Key <- GenusFamilyGroup.Key %>%
  mutate(estimated_genus_name = tolower(estimated_genus_name))

# Merge in Major_Functional_Group data:
CMSY_Timeseries <- CMSY_Timeseries %>%
  left_join(
    GenusFamilyGroup.Key,
    by = c("estimated_genus_name")
  )
# 1569 obs., 13 vars.
```

```{r Create CMSY_Timeseries dataframe}
colnames(CMSY_Timeseries)

CMSY_FIS_scores<- CMSY_Timeseries %>%
  select(estimated_genus_name, Major_Functional_Group, estimated_family_name,
         yr, B, B.Bmsy, lcl.B.Bmsy, ucl.B.Bmsy,
         F.Fmsy, lcl.F.Fmsy, ucl.F.Fmsy)
# 1569 obs., 11 vars.
```


# FINAL STEP: SAVE the CMSY_FIS_scores dataframe:
```{r Save this dataframe}
write.csv(CMSY_FIS_scores,
          "/home/ferrer/ohi_goc_prep/FP/FIS_Scores/FIS_RevisedScores_May2026/FIS_StockStatus_Scores_May2026/FIS_CMSYScores_June012026.csv", row.names = FALSE)
```

# For another time, not useful?, likely purge (June 1 2026):
```{r Summarize landings by genus / stock}
Genera_in_CMSY<- unique(CMSY_Timeseries$stock)
# 31 assessed genera

CMSY_LandingsData <- CMSY_LandingsData_original %>%
  mutate(estimated_genus_name = tolower(estimated_genus_name))

CMSY_LandingsData <- CMSY_LandingsData %>% 
  filter(estimated_genus_name %in% Genera_in_CMSY)
# 1126147 obs., 53 vars
unique(CMSY_LandingsData$estimated_genus_name)
# Yup! Looks good!

# Summarize by year and OHI region:
CMSYstocks_RegionalLandings_Summary <- CMSY_LandingsData %>% 
  filter(year >= 2014, year <= 2024) %>%
  select(year, OHI_Region_No, estimated_genus_name, weight_kg_corrected, value) %>%
  group_by(stock = estimated_genus_name, OHI_Region_No, yr = year) %>%
  summarize(
    total_catch_kg = sum(weight_kg_corrected, na.rm = TRUE),
    total_value_pesos = sum(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(stock, OHI_Region_No, yr)
# 2171 obs., 5 vars.

CMSYstocks_RegionalLandings_Summary <- CMSYstocks_RegionalLandings_Summary %>%
  left_join(
    CMSY_Timeseries,
    by = c("yr", "stock")
  )
# 2171 obs., 14 obs.
```

```{r Now filter to create the final dataframe}
colnames(CMSYstocks_RegionalLandings_Summary)

CMSY_FIS_scores<- CMSYstocks_RegionalLandings_Summary %>% select(yr, OHI_Region_No, stock,
                                                             total_catch_kg, B, B.Bmsy, lcl.B.Bmsy,
                                                             ucl.B.Bmsy,
                                                             F.Fmsy, lcl.F.Fmsy, ucl.F.Fmsy)
```