---
title: "Untitled"
output: html_document
date: "2025-04-29"
---

Exploring the INEGI-ENOE employment data. 
https://www.inegi.org.mx/app/descarga/
Microdata/National Survey of Occupation and Employment (ENOE), population aged 15 years and over/Database
p3 is code for industry
https://www.inegi.org.mx/contenidos/productos/prod_serv/contenidos/espanol/bvinegi/productos/nueva_estruc/702825198411.pdf

p4A North American codes (fishing is more complete):
http://internet.contenidos.inegi.org.mx/contenidos/productos/prod_serv/contenidos/espanol/bvinegi/productos/nueva_estruc/702825197247.pdf

p3 and p4A have a lot of overlap but p4A is more complete, getting people who are involved in fishing industry beyond just the fisherman. 


I got the 2024, 4th quarter data. The COE data are the basic version (vs. the extended version).

Breakdown of tables: 
Occupation data: https://www.inegi.org.mx/contenidos/productos/prod_serv/contenidos/espanol/bvinegi/productos/nueva_estruc/702825198411.pdf

https://www.economia.gob.mx/datamexico/en/profile/industry/fishing


```{r setup, include=FALSE}

library(tidyverse)
library(here)

```

## Exploring the data.
There are 5 dataframes associated with these data. Get a feel for which ones are useful
survey is stored:
•	In the housing table (vivt), the identification data of the selected households is stored, as well as information about the number of residents in the dwelling and how many households it includes. This information is collected through questions 1, 2, and 3 of the sociodemographic questionnaire.
•	In the households table (hogt), the identification of each household found within a dwelling is stored, along with the survey dates and the outcome of both the preliminary and final interviews.
•	The sociodemographic table (sdemt) contains characteristics of household residents, such as residency status, age, sex, etc. These data are gathered through questions 5 to 23 of the sociodemographic questionnaire.
•	The occupation and employment questionnaire (COE), due to its size, is stored in two data tables: COE1 and COE2. In the basic version, the first table includes questions 1 through 5h, and the second includes questions 6 through 9. In the extended version, the first table stores the same questions, while the second includes questions 6 through 11.


```{r}
# This table provides data at the household level. There is municipality data, which may be useful
# for higher resolution data (if we need better than state or city). AGEB is a geostatistical area similar to a neighborhood,
# but no data. Maybe this is available in other formats? Otherwise, I don't think we will use this table.
# The data that seems useful is present in other datasets.
tmp_viv <- read_csv("/home/shares/ohi/OHI_GOC/_raw_data/inegi_enoe_employment/ENOE_VIVT424.csv")
summary(tmp_viv)
table(tmp_viv$est)
table(tmp_viv$est_d_tri)
table(tmp_viv$ageb)

#This table is also at the household level and doesn't really seem to provide additional information that 
# will be useful for our analysis, maybe r_def = 1 indicates a complete interview (although that is in the next table as well)
tmp_hog <- read_csv("/home/shares/ohi/OHI_GOC/_raw_data/inegi_enoe_employment/ENOE_HOGT424.csv")
summary(tmp_hog)

tmp1 <- read_csv("/home/shares/ohi/OHI_GOC/_raw_data/inegi_enoe_employment/ENOE_COE1T424.csv")
sum(tmp1$fac_tri)
# 108,728,324
dim(tmp1)
# 345,680


## This has information that could be helpful in determining job quality and sex breakdown
# <https://www.inegi.org.mx/programas/enoe/15ymas/#microdatos> for metadata
# <https://www.inegi.org.mx/contenidos/programas/enoe/15ymas/doc/con_basedatos_proy2010.pdf>

#r_def = complete interview (value = 1)
#ent = state
#sex 1= male 2 = female
#eda = age
#clase2, 1 = employed
#ing7c, classification of income level
#dur9c, working hours category level
#tue2, business classification
#tue3, gov/non-gov
#sub_o, underemployed, by hours
#seg_soc, health benefits
#anios_esc = years of schooling
# dur_est - hours worked in a week
# ingocup = Monthly income of employed

tmp_dem <- read_csv("/home/shares/ohi/OHI_GOC/_raw_data/inegi_enoe_employment/ENOE_SDEMT424.csv")
summary(tmp_dem)
table(tmp_dem$est_d_tri)
tmp_dem <- tmp_dem %>%
  mutate(unique_id = paste(tmp_dem$con, tmp_dem$upm, sep="_")) %>%
 # mutate(unique_id = paste(unique_id, tmp_dem$v_sel, sep="_")) %>%
  mutate(unique_id = paste(unique_id, tmp_dem$n_ren, sep="_")) %>%
   mutate(unique_id = paste(unique_id, tmp_dem$n_hog, sep="_")) %>%
  mutate(unique_id = paste(unique_id, tmp_dem$n_pro_viv, sep="_")) %>%
  mutate(con_upm = paste(tmp_dem$con, tmp_dem$upm, sep="_")) %>%
  mutate(con_upm_nren =paste(con_upm, tmp_dem$n_ren))
sum(duplicated(tmp_dem$unique_id))
tmp_dem[(duplicated(tmp_dem$unique_id)),]
```

From <https://www.inegi.org.mx/contenidos/programas/enoe/15ymas/doc/clasificaciones_enoe_123.pdf>:

```{r}
# 02 Baja California
# 001 Ensenada
# 002 Mexicali
# 003 Tecate
# 004 Tijuana
# 005 Playas de Rosarito
# 006 San Quintín
# 007 San Felipe
# 
# 03 Baja California Sur
# 001 Comondú
# 002 Mulegé
# 003 La Paz
# 008 Los Cabos
# 009 Loreto
# 
# 14 Jalisco
# 001 Acatic
# 002 Acatlán de Juárez
# 003 Ahualulco de Mercado
# 004 Amacueca
# 005 Amatitán
# 006 Ameca
# 007 San Juanito de Escobedo
# 008 Arandas
# 009 El Arenal
# 010 Atemajac de Brizuela
# 011 Atengo
# 012 Atenguillo
# 013 Atotonilco el Alto
# 014 Atoyac
# 015 Autlán de Navarro
# 016 Ayotlán
# 017 Ayutla
# 018 La Barca
# 019 Bolaños
# 020 Cabo Corrientes
# 021 Casimiro Castillo
# 022 Cihuatlán
# 023 Zapotlán el Grande
# 024 Cocula
# 025 Colotlán
# 026 Concepción de Buenos Aires
# 027 Cuautitlán de García Barragán
# 028 Cuautla
# 029 Cuquío
# 030 Chapala
# 031 Chimaltitán
# 032 Chiquilistlán
# 033 Degollado
# 034 Ejutla
# 035 Encarnación de Díaz
# 036 Etzatlán
# 037 El Grullo
# 038 Guachinango
# 039 Guadalajara
# 040 Hostotipaquillo
# 041 Huejúcar
# 042 Huejuquilla el Alto
# 043 La Huerta
# 044 Ixtlahuacán de los Membrillos
# 045 Ixtlahuacán del Río
# 046 Jalostotitlán
# 047 Jamay
# 048 Jesús María
# 049 Jilotlán de los Dolores
# 050 Jocotepec
# 051 Juanacatlán
# 052 Juchitlán
# 053 Lagos de Moreno
# 054 El Limón
# 055 Magdalena
# 056 Santa María del Oro
# 057 La Manzanilla de la Paz
# 058 Mascota
# 059 Mazamitla
# 060 Mexticacán
# 061 Mezquitic
# 062 Mixtlán
# 063 Ocotlán
# 064 Ojuelos de Jalisco
# 065 Pihuamo
# 066 Poncitlán
# 067 Puerto Vallarta
# 068 Villa Purificación
# 069 Quitupan
# 070 El Salto
# 071 San Cristóbal de la Barranca
# 072 San Diego de Alejandría
# 073 San Juan de los Lagos
# 074 San Julián
# 075 San Marcos
# 076 San Martín de Bolaños
# 077 San Martín Hidalgo
# 078 San Miguel el Alto
# 079 Gómez Farías
# 080 San Sebastián del Oeste
# 081 Santa María de los Ángeles
# 082 Sayula
# 083 Tala
# 084 Talpa de Allende
# 085 Tamazula de Gordiano
# 086 Tapalpa
# 087 Tecalitlán
# 088 Tecolotlán
# 089 Techaluta de Montenegro
# 090 Tenamaxtlán
# 091 Teocaltiche
# 092 Teocuitatlán de Corona
# 093 Tepatitlán de Morelos
# 094 Tequila
# 095 Teuchitlán
# 096 Tizapán el Alto
# 097 Tlajomulco de Zúñiga
# 098 San Pedro Tlaquepaque
# 099 Tolimán
# 100 Tomatlán
# 101 Tonalá
# 102 Tonaya
# 103 Tonila
# 104 Totatiche
# 105 Tototlán
# 106 Tuxcacuesco
# 107 Tuxcueca
# 108 Tuxpan
# 109 Unión de San Antonio
# 110 Unión de Tula
# 111 Valle de Guadalupe
# 112 Valle de Juárez
# 113 San Gabriel
# 114 Villa Corona
# 115 Villa Guerrero
# 116 Villa Hidalgo
# 117 Cañadas de Obregón
# 118 Yahualica de González Gallo
# 119 Zacoalco de Torres
# 120 Zapopan
# 121 Zapotiltic
# 122 Zapotitlán de Vadillo
# 123 Zapotlán del Rey
# 124 Zapotlanejo
# 125 San Ignacio Cerro Gordo
# 
# 18 Nayarit
# 001 Acaponeta
# 002 Ahuacatlán
# 003 Amatlán de Cañas
# 004 Compostela
# 005 Huajicori
# 006 Ixtlán del Río
# 007 Jala
# 008 Xalisco
# 009 Del Nayar
# 010 Rosamorada
# 011 Ruíz
# 012 San Blas
# 013 San Pedro Lagunillas
# 014 Santa María del Oro
# 015 Santiago Ixcuintla
# 016 Tecuala
# 017 Tepic
# 018 Tuxpan
# 019 La Yesca
# 020 Bahía de Banderas
# 
# 25 Sinaloa
# 001 Ahome
# 002 Angostura
# 003 Badiraguato
# 004 Concordia
# 005 Cosalá
# 006 Culiacán
# 007 Choix
# 008 Elota
# 009 Escuinapa
# 010 El Fuerte
# 011 Guasave
# 012 Mazatlán
# 013 Mocorito
# 014 Rosario
# 015 Salvador Alvarado
# 016 San Ignacio
# 017 Sinaloa
# 018 Navolato
# 
# 26 Sonora
# 001 Aconchi
# 002 Agua Prieta
# 003 Alamos
# 004 Altar
# 005 Arivechi
# 006 Arizpe
# 007 Atil
# 008 Bacadéhuachi
# 009 Bacanora
# 010 Bacerac
# 011 Bacoachi
# 012 Bácum
# 013 Banámichi
# 014 Baviácora
# 015 Bavispe
# 016 Benjamín Hill
# 017 Caborca
# 018 Cajeme
# 019 Cananea
# 020 Carbó
# 021 La Colorada
# 022 Cucurpe
# 023 Cumpas
# 024 Divisaderos
# 025 Empalme
# 026 Etchojoa
# 027 Fronteras
# 028 Granados
# 029 Guaymas
# 030 Hermosillo
# 031 Huachinera
# 032 Huásabas
# 033 Huatabampo
# 034 Huépac
# 035 Imuris
# 036 Magdalena
# 037 Mazatán
# 038 Moctezuma
# 039 Naco
# 040 Nácori Chico
# 041 Nacozari de García
# 042 Navojoa
# 043 Nogales
# 044 Onavas
# 045 Opodepe
# 046 Oquitoa
# 047 Pitiquito
# 048 Puerto Peñasco
# 049 Quiriego
# 050 Rayón
# 051 Rosario
# 052 Sahuaripa
# 053 San Felipe de Jesús
# 054 San Javier
# 055 San Luis Río Colorado
# 056 San Miguel de Horcasitas
# 057 San Pedro de la Cueva
# 058 Santa Ana
# 059 Santa Cruz
# 060 Sáric
# 061 Soyopa
# 062 Suaqui Grande
# 063 Tepache
# 064 Trincheras
# 065 Tubutama
# 066 Ures
# 067 Villa Hidalgo
# 068 Villa Pesqueira
# 069 Yécora
# 070 General Plutarco Elías Calles
# 071 Benito Juárez
# 072 San Ignacio Río Muerto
```


```{r}
## limit the SDEM to the Gulf of California

## first, by state
state_names <- tibble(
  ent = 1:32,
  state = c("Aguascalientes", "Baja California", "Baja California Sur", "Campeche", "Coahuila",
            "Colima", "Chiapas", "Chihuahua", "CDMX", "Durango",
            "Guanajuato", "Guerrero", "Hidalgo", "Jalisco", "Mexico State",
            "Michoacán", "Morelos", "Nayarit", "Nuevo León", "Oaxaca",
            "Puebla", "Querétaro", "Quintana Roo", "San Luis Potosí", "Sinaloa",
            "Sonora", "Tabasco", "Tamaulipas", "Tlaxcala", "Veracruz",
            "Yucatán", "Zacatecas")
)

tmp_dem_goc_ent <- tmp_dem %>%
  left_join(state_names, by = "ent") %>%
  filter(state %in% c("Baja California", "Baja California Sur", "Sonora", "Sinaloa", "Nayarit", "Jalisco"))



## now, by municipio: 
# municipio is named starting at 001 for each state, so to give it a unique identifier, we need to combine "ent" and "mun" for each one. <https://www.inegi.org.mx/contenidos/programas/enoe/15ymas/doc/clasificaciones_enoe_123.pdf> 

tmp_dem_ent_mun <- tmp_dem_goc_ent %>% 
  mutate(ent_mun = paste(tmp_dem_goc_ent$ent, tmp_dem_goc_ent$mun, sep="_"))


library(tibble)
municipios <- tribble(
  ~ent_mun,        ~municipio,
  "2_1",           "Ensenada",
  "2_2",           "Mexicali",
  "2_3",           "Tecate",
  "2_4",           "Tijuana",
  "2_5",           "Playas de Rosarito",
  "2_6",           "San Quintín",
  "2_7",           "San Felipe",
  "3_1",           "Comondú",
  "3_2",           "Mulegé",
  "3_3",           "La Paz",
  "3_8",           "Los Cabos",
  "3_9",           "Loreto",
  "14_1",          "Acatic",
  "14_2",          "Acatlán de Juárez",
  "14_3",          "Ahualulco de Mercado",
  "14_4",          "Amacueca",
  "14_5",          "Amatitán",
  "14_6",          "Ameca",
  "14_7",          "San Juanito de Escobedo",
  "14_8",          "Arandas",
  "14_9",          "El Arenal",
  "14_10",         "Atemajac de Brizuela",
  "14_11",         "Atengo",
  "14_12",         "Atenguillo",
  "14_13",         "Atotonilco el Alto",
  "14_14",         "Atoyac",
  "14_15",         "Autlán de Navarro",
  "14_16",         "Ayotlán",
  "14_17",         "Ayutla",
  "14_18",         "La Barca",
  "14_19",         "Bolaños",
  "14_20",         "Cabo Corrientes",
  "14_21",         "Casimiro Castillo",
  "14_22",         "Cihuatlán",
  "14_23",         "Zapotlán el Grande",
  "14_24",         "Cocula",
  "14_25",         "Colotlán",
  "14_26",         "Concepción de Buenos Aires",
  "14_27",         "Cuautitlán de García Barragán",
  "14_28",         "Cuautla",
  "14_29",         "Cuquío",
  "14_30",         "Chapala",
  "14_31",         "Chimaltitán",
  "14_32",         "Chiquilistlán",
  "14_33",         "Degollado",
  "14_34",         "Ejutla",
  "14_35",         "Encarnación de Díaz",
  "14_36",         "Etzatlán",
  "14_37",         "El Grullo",
  "14_38",         "Guachinango",
  "14_39",         "Guadalajara",
  "14_40",         "Hostotipaquillo",
  "14_41",         "Huejúcar",
  "14_42",         "Huejuquilla el Alto",
  "14_43",         "La Huerta",
  "14_44",         "Ixtlahuacán de los Membrillos",
  "14_45",         "Ixtlahuacán del Río",
  "14_46",         "Jalostotitlán",
  "14_47",         "Jamay",
  "14_48",         "Jesús María",
  "14_49",         "Jilotlán de los Dolores",
  "14_50",         "Jocotepec",
  "14_51",         "Juanacatlán",
  "14_52",         "Juchitlán",
  "14_53",         "Lagos de Moreno",
  "14_54",         "El Limón",
  "14_55",         "Magdalena",
  "14_56",         "Santa María del Oro",
  "14_57",         "La Manzanilla de la Paz",
  "14_58",         "Mascota",
  "14_59",         "Mazamitla",
  "14_60",         "Mexticacán",
  "14_61",         "Mezquitic",
  "14_62",         "Mixtlán",
  "14_63",         "Ocotlán",
  "14_64",         "Ojuelos de Jalisco",
  "14_65",         "Pihuamo",
  "14_66",         "Poncitlán",
  "14_67",         "Puerto Vallarta",
  "14_68",         "Villa Purificación",
  "14_69",         "Quitupan",
  "14_70",         "El Salto",
  "14_71",         "San Cristóbal de la Barranca",
  "14_72",         "San Diego de Alejandría",
  "14_73",         "San Juan de los Lagos",
  "14_74",         "San Julián",
  "14_75",         "San Marcos",
  "14_76",         "San Martín de Bolaños",
  "14_77",         "San Martín Hidalgo",
  "14_78",         "San Miguel el Alto",
  "14_79",         "Gómez Farías",
  "14_80",         "San Sebastián del Oeste",
  "14_81",         "Santa María de los Ángeles",
  "14_82",         "Sayula",
  "14_83",         "Tala",
  "14_84",         "Talpa de Allende",
  "14_85",         "Tamazula de Gordiano",
  "14_86",         "Tapalpa",
  "14_87",         "Tecalitlán",
  "14_88",         "Tecolotlán",
  "14_89",         "Techaluta de Montenegro",
  "14_90",         "Tenamaxtlán",
  "14_91",         "Teocaltiche",
  "14_92",         "Teocuitatlán de Corona",
  "14_93",         "Tepatitlán de Morelos",
  "14_94",         "Tequila",
  "14_95",         "Teuchitlán",
  "14_96",         "Tizapán el Alto",
  "14_97",         "Tlajomulco de Zúñiga",
  "14_98",         "San Pedro Tlaquepaque",
  "14_99",         "Tolimán",
  "14_100",        "Tomatlán",
  "14_101",        "Tonalá",
  "14_102",        "Tonaya",
  "14_103",        "Tonila",
  "14_104",        "Totatiche",
  "14_105",        "Tototlán",
  "14_106",        "Tuxcacuesco",
  "14_107",        "Tuxcueca",
  "14_108",        "Tuxpan",
  "14_109",        "Unión de San Antonio",
  "14_110",        "Unión de Tula",
  "14_111",        "Valle de Guadalupe",
  "14_112",        "Valle de Juárez",
  "14_113",        "San Gabriel",
  "14_114",        "Villa Corona",
  "14_115",        "Villa Guerrero",
  "14_116",        "Villa Hidalgo",
  "14_117",        "Cañadas de Obregón",
  "14_118",        "Yahualica de González Gallo",
  "14_119",        "Zacoalco de Torres",
  "14_120",        "Zapopan",
  "14_121",        "Zapotiltic",
  "14_122",        "Zapotitlán de Vadillo",
  "14_123",        "Zapotlán del Rey",
  "14_124",        "Zapotlanejo",
  "14_125",        "San Ignacio Cerro Gordo",
  "18_1",          "Acaponeta",
  "18_2",          "Ahuacatlán",
  "18_3",          "Amatlán de Cañas",
  "18_4",          "Compostela",
  "18_5",          "Huajicori",
  "18_6",          "Ixtlán del Río",
  "18_7",          "Jala",
  "18_8",          "Xalisco",
  "18_9",          "Del Nayar",
  "18_10",         "Rosamorada",
  "18_11",         "Ruíz",
  "18_12",         "San Blas",
  "18_13",         "San Pedro Lagunillas",
  "18_14",         "Santa María del Oro",
  "18_15",         "Santiago Ixcuintla",
  "18_16",         "Tecuala",
  "18_17",         "Tepic",
  "18_18",         "Tuxpan",
  "18_19",         "La Yesca",
  "18_20",         "Bahía de Banderas",
  "25_1",          "Ahome",
  "25_2",          "Angostura",
  "25_3",          "Badiraguato",
  "25_4",          "Concordia",
  "25_5",          "Cosalá",
  "25_6",          "Culiacán",
  "25_7",          "Choix",
  "25_8",          "Elota",
  "25_9",          "Escuinapa",
  "25_10",         "El Fuerte",
  "25_11",         "Guasave",
  "25_12",         "Mazatlán",
  "25_13",         "Mocorito",
  "25_14",         "Rosario",
  "25_15",         "Salvador Alvarado",
  "25_16",         "San Ignacio",
  "25_17",         "Sinaloa",
  "25_18",         "Navolato",
  "26_1",          "Aconchi",
  "26_2",          "Agua Prieta",
  "26_3",          "Alamos",
  "26_4",          "Altar",
  "26_5",          "Arivechi",
  "26_6",          "Arizpe",
  "26_7",          "Atil",
  "26_8",          "Bacadéhuachi",
  "26_9",          "Bacanora",
  "26_10",         "Bacerac",
  "26_11",         "Bacoachi",
  "26_12",         "Bácum",
  "26_13",         "Banámichi",
  "26_14",         "Baviácora",
  "26_15",         "Bavispe",
  "26_16",         "Benjamín Hill",
  "26_17",         "Caborca",
  "26_18",         "Cajeme",
  "26_19",         "Cananea",
  "26_20",         "Carbó",
  "26_21",         "La Colorada",
  "26_22",         "Cucurpe",
  "26_23",         "Cumpas",
  "26_24",         "Divisaderos",
  "26_25",         "Empalme",
  "26_26",         "Etchojoa",
  "26_27",         "Fronteras",
  "26_28",         "Granados",
  "26_29",         "Guaymas",
  "26_30",         "Hermosillo",
  "26_31",         "Huachinera",
  "26_32",         "Huásabas",
  "26_33",         "Huatabampo",
  "26_34",         "Huépac",
  "26_35",         "Imuris",
  "26_36",         "Magdalena",
  "26_37",         "Mazatán",
  "26_38",         "Moctezuma",
  "26_39",         "Naco",
  "26_40",         "Nácori Chico",
  "26_41",         "Nacozari de García",
  "26_42",         "Navojoa",
  "26_43",         "Nogales",
  "26_44",         "Onavas",
  "26_45",         "Opodepe",
  "26_46",         "Oquitoa",
  "26_47",         "Pitiquito",
  "26_48",         "Puerto Peñasco",
  "26_49",         "Quiriego",
  "26_50",         "Rayón",
  "26_51",         "Rosario",
  "26_52",         "Sahuaripa",
  "26_53",         "San Felipe de Jesús",
  "26_54",         "San Javier",
  "26_55",         "San Luis Río Colorado",
  "26_56",         "San Miguel de Horcasitas",
  "26_57",         "San Pedro de la Cueva",
  "26_58",         "Santa Ana",
  "26_59",         "Santa Cruz",
  "26_60",         "Sáric",
  "26_61",         "Soyopa",
  "26_62",         "Suaqui Grande",
  "26_63",         "Tepache",
  "26_64",         "Trincheras",
  "26_65",         "Tubutama",
  "26_66",         "Ures",
  "26_67",         "Villa Hidalgo",
  "26_68",         "Villa Pesqueira",
  "26_69",         "Yécora",
  "26_70",         "General Plutarco Elías Calles",
  "26_71",         "Benito Juárez",
  "26_72",         "San Ignacio Río Muerto"
)

tmp_dem_ent_mun_join <- tmp_dem_ent_mun %>%
  left_join(municipios, by = "ent_mun") %>% 
  mutate(country = "Mexico")
```

Now, let's obtain their location via `tidygeocoder`:

```{r}
library(tidygeocoder)

tmp_dem_goc_lat_lon <- tmp_dem_ent_mun_join %>% 
  geocode(., city = "municipio", state = "state", country = "country") # done in 116.9 seconds!

# see what did not work
tmp_dem_goc_lat_lon_na <- tmp_dem_goc_lat_lon %>% 
  mutate(geocode_fail = is.na(.$lat)) %>% 
  filter(geocode_fail == "TRUE")

unique(tmp_dem_goc_lat_lon_na$municipio) # [1] "Bahía de Banderas"  "Zapotlán el Grande" "Casimiro Castillo"  NA  

## we know that Bahia de Banderas is within the Gulf. Casimiro Castillo and Zapotlán el Grande are NOT.
## I am going to manually put in the lat longs for "Bahia de Banderas"

tmp_dem_goc_lat_lon_fix <- tmp_dem_goc_lat_lon %>% 
  mutate(lat = ifelse(municipio == "Bahía de Banderas", 20.8074834, lat)) %>% # looks good!
  mutate(long = ifelse(municipio == "Bahía de Banderas", -105.2480864, long))

# now, isolate "Zapotlán el Grande" "Casimiro Castillo"  NA and drop them. 
tmp_dem_goc_lat_lon_na_fix <- tmp_dem_goc_lat_lon_fix %>% 
  mutate(geocode_fail = is.na(.$lat)) %>% 
  filter(geocode_fail == "TRUE")

tmp_dem_goc_sf <- tmp_dem_goc_lat_lon_fix %>% 
  filter(!municipio %in% unique(tmp_dem_goc_lat_lon_na_fix$municipio)) %>%  # unique(is.na(tmp_dem_goc_sf$lat)) [1] FALSE
  st_as_sf(coords = c("long", "lat"),
           crs = 4326, # WGS84
           remove = FALSE) # convert to sf object

library(mapview)
mapview(tmp_dem_goc_sf)

### there are no observations for cabo corrientes. 
# > test_1 <- tmp_dem_goc_lat_lon %>%  filter(municipio %in% "Cabo Corrientes")
# > test_2 <- tmp_dem_ent_mun_join %>%  filter(municipio %in% "Cabo Corrientes")
# > test_3 <- tmp_dem_ent_mun %>%  filter(ent_mun %in% "14_20") # all 0 obs 
```


```{r}
### now, let's look at completeness for tmp_dem of survey responses for all quality variables in the GoC states (not filtered for coastal municipios yet!!).

## for `zona`
unique(tmp_dem_goc_sf$zona) # 1, 2
tmp_dem_goc_sf$zona

## for `salario`
unique(tmp_dem_goc_sf$salario) # [1]  7468 11247
tmp_dem_goc_sf$salario # so weird... all are either number

## for `salario`
unique(tmp_dem_goc_sf$seg_soc) # [1] 0 1 2 3
tmp_dem_goc_sf$seg_soc 
table(tmp_dem_goc_sf$seg_soc)
#     0     1     2     3 
# 37975 16831 16281   585 
# a little over 50% of those who specified whether they have healthcare actually have access to healthcare.

## for `c_ocu11c`
unique(tmp_dem_goc_sf$c_ocu11c) # [1]  0  4  6  1  8  5  3 10  7  2  9 11
tmp_dem_goc_sf$c_ocu11c 
table(tmp_dem_goc_sf$c_ocu11c)
#     0     1     2     3     4     5     6     7     8     9    10    11 
# 37975  4376  1204   685  3353  8673  5478  1731  5545   327  2300    25 
# 0	No aplica
# 1	Profesionales, técnicos y trabajadores del arte
# 2	Trabajadores de la educación
# 3	Funcionarios y directivos
# 4	Oficinistas
# 5	Trabajadores industriales artesanos y ayudantes
# 6	Comerciantes
# 7	Operadores de transporte
# 8	Trabajadores en servicios personales
# 9	Trabajadores en protección y vigilancia
# 10	Trabajadores agropecuarios
# 11	No especificado

## for `medica5c`
unique(tmp_dem_goc_sf$medica5c) # [1] 0 3 1 4 2 5
tmp_dem_goc_sf$medica5c 
table(tmp_dem_goc_sf$medica5c)
#     0     1     2     3     4     5 
# 37975 12881   682 16149  3419   566 
# 0	No aplica	
# 1	Sin prestaciones
# 2	Solo acceso a instituciones de salud
# 3	Acceso de instituciones de salud y otras prestaciones	
# 4	No tiene acceso a instituciones de salud pero si a otras prestaciones
# 5	No especificado

## for `buscar5c`
unique(tmp_dem_goc_sf$buscar5c) # [1] 0 4 3 2 1 5
tmp_dem_goc_sf$buscar5c 
table(tmp_dem_goc_sf$buscar5c)
#     0     1     2     3     4     5 
# 37975   136   572   171 32792    26 
# 0	No aplica
# 1	Para tener otro empleo
# 2	Para cambiarse de trabajo
# 3	Otro	
# 4	No buscó
# 5	No especificado

## for `DUR_EST`
unique(tmp_dem_goc_sf$dur_est) # [1] 0 4 3 5 2 6 1
tmp_dem_goc_sf$dur_est 
table(tmp_dem_goc_sf$dur_est)
#     0     1     2     3     4     5     6 
# 37975   630  2033  5414 18381  6302   937 
# 1 = “Temporary absences with employment relationship”
# 2 = less than 15 horus,
# 3 = 15 -  34 hours
# 4 = 35 - 48 hours
# 5 = more than 48 hours
# 6 = not specified

## for `tue3`
unique(tmp_dem_goc_sf$tue3) # [1] 0 1 2
tmp_dem_goc_sf$tue3 
table(tmp_dem_goc_sf$tue3)
#     0     1     2 
# 67780  3575   317
# 0	No aplica
# 1	Administradas por los gobiernos
# 2	No administradas por los gobiernos


## for `busqueda`
unique(tmp_dem_goc_sf$busqueda) # [1] 0 2 1
table(tmp_dem_goc_sf$busqueda)
#     0     1     2 
# 37975   136 33561 
# 0	No aplica	
# 1	Si buscó un trabajo adicional
# 2	No buscó un trabajo adicional

## for `d_ant_lab`
table(tmp_dem_goc_sf$d_ant_lab)
#     0     1     2 
# 70872   720    80 

## for `d_cexp_est`
table(tmp_dem_goc_sf$d_cexp_est)
#   0     1     2     3     4 
# 70952   281   373    42    24  
# 0	No aplica	
# 1	Perdió o termino su empleo anterior	
# 2	Insatisfacción con el empleo anterior
# 3	Dejo o cerro un negocio propio
# 4	Otro

## for `dur_des`
table(tmp_dem_goc_sf$dur_des)
#     0     1     2     3     4     5     6 
# 70872   437   250    51    19     4    39 
# 0	No aplica	
# 1	Hasta 1 mes	
# 2	Más de 1 mes hasta 3 meses	
# 3	Más de 3 mes hasta 6 meses
# 4	Más de 6 meses hasta 1 año
# 5	Más de 1 año
# 6	No especificado
```

These are census data. But the fac_tri value indicates how many people each row (i.e., person represents).
Checking that this matches up!

fac_tri is the expansion factor for each respondent in the quarterly sample. It tells you how many people in the total population that one person represents in that quarter.

Think of it like:

“This person stands in for ~X people in the Mexican population in this quarter.”
It should approximate the working-age population (12+) of Mexico for the quarter, in households covered by ENOE.

As of 2023–2024, this might be around 100 to 110 million depending on the month, but not the full national population (~130 million), because:

ENOE excludes kids under 12,

It samples households, not institutional housing (like prisons),

It may be limited to urban/rural areas depending on the wave.

```{r}
tmp1 <- read_csv("/home/shares/ohi/OHI_GOC/_raw_data/inegi_enoe_employment/ENOE_COE1T424.csv")
sum(tmp1$fac_tri)
# 108,728,324
# population is 131.9 million people
# population 12+ is about 107 million - so this seems about right

dim(tmp1)
# 345,680
summary(tmp1)

## check number of people in fishing jobs
check1 <- filter(tmp1, p3 %in% c("6211")) #Industry data
 sum(check1$fac_tri)
# 91,8994  ## I got 91,994 ~ Sophia
 # From here: https://www.economia.gob.mx/datamexico/en/profile/industry/fishing?occupationMetrics=workforceOption&quarterFinal=option20243&quarterInitial1=option20243
 # 2024 third quarter was 91.8k. So this checks out.
 
 dim(check1) # you can see that a relatively small number of respondents is representing the larger population

check2 <- filter(tmp1, p4a %in% c("1141")) # This is the North America Industry
table(check2$p3) # can see that most match the 6211, but there are several other occupations included in there - which I think makes sense
 sum(check2$fac_tri)
 dim(check2)
 
```


## Obtaining a unique ID and join data

```{r}

jobs1 <- read_csv("/home/shares/ohi/OHI_GOC/_raw_data/inegi_enoe_employment/ENOE_COE1T424.csv")

jobs1 <- jobs1
jobs1 <- jobs1 %>%
  mutate(unique_id = paste(jobs1$con, jobs1$upm, sep="_")) %>%
  mutate(unique_id = paste(unique_id, jobs1$v_sel, sep="_")) %>%
  mutate(unique_id = paste(unique_id, jobs1$n_ren, sep="_")) %>%
  mutate(unique_id = paste(unique_id, jobs1$n_hog, sep="_")) %>%
  mutate(unique_id = paste(unique_id, jobs1$n_pro_viv, sep="_")) %>%
  select(unique_id, r_def, upm, v_sel, n_ren, n_hog, n_pro_viv, cd_a, ent, p1, p1e, p3, p4a, p4d1,
         fac_tri)

sum(duplicated(jobs1$unique_id)) # should be no duplicates.

jobs2 <- read_csv("/home/shares/ohi/OHI_GOC/_raw_data/inegi_enoe_employment/ENOE_COE2T424.csv")

jobs2 <- jobs2 %>%
  mutate(unique_id = paste(jobs2$con, jobs2$upm, sep="_")) %>%
  mutate(unique_id = paste(unique_id, jobs2$v_sel, sep="_")) %>%
  mutate(unique_id = paste(unique_id, jobs2$n_ren, sep="_")) %>%
  mutate(unique_id = paste(unique_id, jobs2$n_hog, sep="_")) %>%
  mutate(unique_id = paste(unique_id, jobs2$n_pro_viv, sep="_")) %>%
  select(unique_id, p6c, p6d, p6e, p6e_c, p6f, p7a, p7c, p6e_c, p6f_c)

sum(duplicated(jobs2$unique_id))

length(setdiff(jobs2$unique_id, jobs1$unique_id)) # should be no differences

jobs <- left_join(jobs1, jobs2, by = "unique_id")

```


# Identify the relevant states
```{r}
state_names <- tibble(
  ent = 1:32,
  state = c("Aguascalientes", "Baja California", "Baja California Sur", "Campeche", "Coahuila",
            "Colima", "Chiapas", "Chihuahua", "CDMX", "Durango",
            "Guanajuato", "Guerrero", "Hidalgo", "Jalisco", "Mexico State",
            "Michoacán", "Morelos", "Nayarit", "Nuevo León", "Oaxaca",
            "Puebla", "Querétaro", "Quintana Roo", "San Luis Potosí", "Sinaloa",
            "Sonora", "Tabasco", "Tamaulipas", "Tlaxcala", "Veracruz",
            "Yucatán", "Zacatecas")
)

jobs <- jobs %>%
  left_join(state_names, by = "ent") %>%
  filter(state %in% c("Baja California", "Baja California Sur", "Sonora", "Sinaloa", "Nayarit", "Jalisco"))
sum(jobs$fac_tri)
## 17,407,179
dim(jobs)
## 58,936

## This seems reasonable. Total population in 2020 of these states is 
(3769020 + 798447 + 2944840 + 3026943 + 1235456 + 8348151) * 0.81
# about 81% are 12+ in age, this value seems reasonable.

```


## Identify relevant codes and combine the two occupation datasets
### Primary jobs
```{r}

occupations <- read_csv(here("LIV/data/occupations.csv"))

occ <- occupations %>%
  filter(data == "occupation") %>%
  #filter(!is.na(category1)) %>%
  mutate(p3 = code) %>%
  select(-data, -code, p3, cat1_occ = category1, cat2_occ= category2, occ_description = description)

occ_marine <- occ %>%
  filter(!is.na(cat1_occ))

industry <- occupations %>%
  filter(data == "north_america_industry") %>%
  #filter(!is.na(category1)) %>%
  mutate(p4a = code)%>%
  select(-data, -code, p4a, cat1_ind = category1, cat2_ind= category2, ind_description = description)
  
ind_marine <- industry %>%
  filter(!is.na(cat1_ind))

occs_unique_id <- jobs %>%
  dplyr::filter(p3 %in% occ_marine$p3) %>%
  select(unique_id) %>%
  pull() # N = 4188

ind_unique_id <- jobs %>%
  dplyr::filter(p4a %in% ind_marine$p4a) %>%
  select(unique_id) %>%
  pull() # N = 4811

setdiff(occs_unique_id, ind_unique_id) 
setdiff(ind_unique_id, occs_unique_id)

marine_ids <- unique(c(occs_unique_id, ind_unique_id)) #N = 5564

marine_jobs_primary <- jobs %>%
  dplyr::filter(unique_id %in% marine_ids) %>%
  left_join(occ, by="p3") %>%
  left_join(industry, by= "p4a") %>%
  mutate(occ_ind = paste(occ_description, ind_description, sep = " --- ")) %>%
  mutate(occ_ind_id = paste(p3, p4a, sep = "---")) %>%
  group_by(occ_ind_id) %>%
  mutate(count = n()) %>%
  arrange(-count) %>%
  ungroup() 

marine_jobs_count <- marine_jobs_primary %>%
  select(occ_ind, occ_ind_id, count) %>%
  unique()

write_csv(marine_jobs_count, here("LIV/data/occupations_industry_blank.csv"))

```

## Identify employment in marine categories
Started to do this: Worked at least one hour past week (p1==1), and if not will return (p1e == 1,2,3).
But the data aren't actually in the provided data. So I will have to assume that if they have an occupation then
they are employed. 
```{r}
occu_cats <- read_csv(here("LIV/data/occupations_industry.csv"))

jobs2 <- jobs %>%
  mutate(occ_ind_id = paste(p3, p4a, sep="---")) %>%
  left_join(occu_cats, by="occ_ind_id") 

jobs2 <- jobs2 %>%
  filter(tourism == "x") %>%
  mutate(fac_tri_correct = 1)

table(jobs2$category1)
table(jobs2$category2)
table(jobs2$category3)

## divide out the combined aquaculture and fisheries based on overall employment
aqua <- filter(jobs2, jobs2$category1=="aquaculture")
fish <- filter(jobs2, jobs2$category1=="fisheries")

aqua_correct <- sum(aqua$fac_tri)/(sum(aqua$fac_tri) + sum(fish$fac_tri))

fish_aqua <- filter(jobs2, category1=="aquaculture-fisheries")
  
fish_aqua_correct <- rbind(fish_aqua %>% mutate(category1="aquaculture"),
                           fish_aqua %>% mutate(category1="fisheries")) %>%
                    mutate(fac_tri_correct = ifelse(category1=="aquaculture", 0.2, 0.8) ) 

jobs3 <- jobs2 %>%
  filter(category1 != "aquaculture-fisheries") 
jobs3 <- rbind.data.frame(jobs3, fish_aqua_correct)


```

### Plots of data
```{r}

# get fully marine related
table(jobs3$category1)

marine_jobs <- jobs3 %>%
  filter(category1 %in% c("aquaculture", "fisheries", "transportation-marine", "culture-entertainment-marine")) %>%
  group_by(category1, state) %>%
  summarize(employed = sum(fac_tri, na.rm=TRUE)) %>% 
  ungroup() %>%
  group_by(category1) %>%
  mutate(total_employed = sum(employed)) %>%
  ungroup() %>%
  mutate(category1 = reorder(category1, -total_employed))

ggplot(marine_jobs, aes(x = category1, y = employed, fill = state)) +
  geom_bar(stat = "identity") +
  labs(x = "Occupation Category", y = "Number Employed", fill = "State") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


# get tourism employment, at least partially marine related
table(jobs3$category3)

marine_jobs <- jobs3 %>%
  filter(category3 %in% "tourism") %>%
  mutate(category1 = gsub("-marine", "", category1)) %>%
  filter(category1 != "aquaculture") %>%
  group_by(category1, state) %>%
  summarize(employed = sum(fac_tri, na.rm=TRUE)) %>% 
  ungroup() %>%
  group_by(category1) %>%
  mutate(total_employed = sum(employed)) %>%
  ungroup() %>%
  mutate(category1 = reorder(category1, -total_employed))

ggplot(marine_jobs, aes(x = category1, y = employed, fill = state)) +
  geom_bar(stat = "identity") +
  labs(x = "Occupation Category", y = "Number Employed", fill = "State") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# get research
table(jobs3$category3)

marine_jobs <- jobs3 %>%
  filter(category3 %in% "research") %>%
  group_by(category2, state) %>%
  summarize(employed = sum(fac_tri, na.rm=TRUE)) %>% 
  ungroup() %>%
  group_by(category2) %>%
  mutate(total_employed = sum(employed)) %>%
  ungroup() %>%
  mutate(category2 = reorder(category2, -total_employed))

ggplot(marine_jobs, aes(x = category2, y = employed, fill = state)) +
  geom_bar(stat = "identity") +
  labs(x = "Occupation Category", y = "Number Employed", fill = "State") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

