---
title: 'OHI GoC 2025: Livelihoods & Economies'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: inline
editor_options: 
  chunk_output_type: inline
---

# LE: Sistema Automatizado de Información Censal (SAIC) // Automated Census Information System (SAIC)

1. Select the census year(s).
2. Indicates the geographical area.
3. Choose the economic activities.
4. Indicates the census variables.
5. Click on Consult.

Additionally, you can view information by employment strata and learn about the census results.
If you prefer, you can download all the information by selecting and following the steps indicated in the Bulk Download icon.

Methodological document: /home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025/saic_historico_metodologico_ce2024.pdf
* Note 1: Corrected figures. See note

* Note 2: As of Wednesday, July 30, 2025, the query of information in the Automated Census Information System was reestablished for the years 2003, 2008, 2013, as a result of the review of the query of the composition of the strata of employed personnel. The information published today replaces the previous one, which was affected in the cuts of "Strata of employed personnel", this was due to a technical adjustment in the information integration procedure used.

* Note 3: The monetary information presented here corresponds to the value in current prices of the year of each census event, therefore, to make comparisons between the information of the different years available in this system, it is suggested to use the Implicit Price Index (IPI) available in the Goods and Services Accounts (detailed). Base year 2018 . Specifically, during the presentation of the final results, the Revised 2023 IPI value was used, for example, for the Gross Census Value Added the following inflators per year were used: for 2008, 1.92081442376959; for 2013, 1.59502439833556 and for 2018, 1.26412683998505; and for Remunerations the following were used: for 2008, 1.93153962083199; for 2013, 1.5759; and for 2018, 1.2895. To calculate other variables and to see the complete calculation method, they can be found in the "Methodological Document" of this system.

* Note 4: The employed personnel reported in this Statistical Program is conceptually different (see metadata) from the definition of jobs affiliated with the Mexican Social Security Institute.

---

For each of the municipalities mentioned in my [comment above](https://github.com/OHI-GOC/OHI_GOC_issues/issues/20#issuecomment-3383649433), I downloaded the data for every relevant sector through the [Automated Census Information System (SAIC)](https://www.inegi.org.mx/app/saic/).  Each sector's dataset has variables that apply to both livelihoods and economies.  

They are now located here: ~/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025

I downloaded the sectors separately so that we can keep track of the types of sectors we have, but I will likely bind all of these datasets into one.  Since everything had to be checked (physically pressing the button haha) in order to download, it felt easiest to keep the job/economy, years, and municipio variables the same, and just vary the sectors selected.

These are all of the datasets/sectors downloaded:

* SAIC_Exporta_20251021_163827445_fishing.csv
* SAIC_Exporta_20251021_171653871_organizations.csv
* SAIC_Exporta_20251021_171548286_public_restrooms_parking.csv
* SAIC_Exporta_20251021_171034329_food_services.csv
* SAIC_Exporta_20251021_17859901_accommodations.csv
* SAIC_Exporta_20251021_17745611_tourist_marinas.csv
* SAIC_Exporta_20251021_17550467_museums_historical_naturalparks.csv
* SAIC_Exporta_20251021_17421711_job_training_services.csv
* SAIC_Exporta_20251021_1712765_hazardous_waste.csv
* SAIC_Exporta_20251021_165918141_professional_scientific_technical_services.csv
* SAIC_Exporta_20251021_165616924_possible_natural_products.csv
* SAIC_Exporta_20251021_16511703_prep_package_fish_shellfish.csv
* SAIC_Exporta_20251021_16482728_supply_naturalgas.csv
* SAIC_Exporta_20251021_16471476_energy_generation.csv
* SAIC_Exporta_20251021_164547329_water_collection_treatment_supply.csv
* SAIC_Exporta_20251021_164332559_all_mining.csv
* SAIC_Exporta_20251021_164117726_oil_gas.csv
* SAIC_Exporta_20251021_164011725_aquaculture.csv

**For every municipio, there is a "Municipal Total" row, which based on the SAIC Metodológico 2024 document from INEGI, represents the complete total for all registered economic units and activity types in that municipality.**

Here’s how it works according to the methodology:
- Each “Municipal total” aggregates the values of all economic units within that municipality -- across all sectors, subsectors, and classes classified under the SCIAN México 2023 system.
- This includes all the summed up values for variables such as (for example!!):
  - EU (Economic Units): all establishments registered in that municipality.
  - J000A (Total remuneration): total payroll and compensation for all workers in that municipality.
  - K000A (Total expenditures on goods and services): total operational expenditures across all sectors.

The document explicitly explains that these aggregates are constructed so users can obtain a “panorama del comportamiento económico” (economic overview) at the municipal, state, or national level.

That means we can compare sectoral data (like “Clase 713930 Marinas turísticas”) to the “Municipal total” to determine the proportion or contribution of that specific sector to the total economic activity in any municipio (if they participate in that sector).

> For example:
> If the municipal total remuneration is 46,766.636 million pesos, and the “Temporary accommodation and food services” sector has 1,989.833 million pesos, then that sector accounts for roughly 4.3% of total remuneration in Mexicali.

# Exploring SAIC

```{r}
library(tidyverse)
library(sf)
library(mapview)
```

## Accommodations

```{r}
## read in data
saic_accommodations <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251021_17859901_accommodations.csv"), skip = 4)

saic_accommodations_liv <- saic_accommodations %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)"))
summary(saic_accommodations_liv)

# unique(saic_accommodations_liv$ano_censal)
# [1] "2023"                          
# [2] "2018"                                     
# [3] "2013"                                 
# [4] "2008"                              
# [5] "2003"                              
# [6] NA                                  
# [7] "Nota: para las celdas que aparecen vacías, se omitieron los datos absolutos de las variables económicas, con el fin de garantizar el principio de confidencialidad y reserva."

## see which municipios are missing data for at least one year, and for how many years it is NA.
saic_accommodations_liv_check_complete <- saic_accommodations_liv %>% 
  filter(ano_censal %in% c("2023","2018","2013","2008","2003")) %>% 
  filter(actividad_economica %in% "Subsector 721 Servicios de alojamiento temporal") %>% 
  group_by(municipio) %>% 
  summarise(
    na_values = sum(is.na(h001a_personal_ocupado_total))
  )  ## if it is NA for every year, it has an na_values of 5. 
  
saic_accommodations_liv_noncomplete <- saic_accommodations_liv_check_complete %>% 
  filter(!na_values == 0) # any that have NAs for one or more years.

############## percent of municipios that have at least one year of data missing: #########
length(unique(saic_accommodations_liv_noncomplete$municipio))/length(unique(saic_accommodations_liv_check_complete$municipio)) * 100 # v2025: 23.52941% of municipios have at least one year missing data.

## check that 013 San Pedro Lagunillas has all NAs: YUP! good.

setdiff(unique(saic_accommodations_liv$municipio), unique(saic_accommodations_liv_complete$municipio)) # "012 Bácum" missing any registered economic units.
```

### Relationship between economic units registered and the number of individuals employed:

First, I am going to take a look at PV. 

```{r}
# separate out puerto vallarta
saic_accommodations_liv_puerto_vallarta <- saic_accommodations_liv %>% 
  filter(municipio %in% "067 Puerto Vallarta") %>% 
  filter(!actividad_economica %in% "Clase 713930 Marinas turísticas")

# fit a linear model
lm_accommodations_eu_employed_pv <- lm(h001a_personal_ocupado_total ~ ue_unidades_economicas, data = saic_accommodations_liv_puerto_vallarta)
summary(lm_accommodations_eu_employed_pv)

# Residuals:
#    Min     1Q Median     3Q    Max 
#  -5994  -5225  -1692   4497   9194 
# 
# Coefficients:
#                        Estimate Std. Error t value Pr(>|t|)    
# (Intercept)            5302.745   1015.487   5.222  8.2e-06 ***
# ue_unidades_economicas    5.666      0.225  25.183  < 2e-16 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Residual standard error: 5730 on 35 degrees of freedom
#   (13 observations deleted due to missingness)
# Multiple R-squared:  0.9477,	Adjusted R-squared:  0.9462 
# F-statistic: 634.2 on 1 and 35 DF,  p-value: < 2.2e-16

r2 <- summary(lm_accommodations_eu_employed_pv)$r.squared

# add best fit
plot(saic_accommodations_liv_puerto_vallarta$ue_unidades_economicas, saic_accommodations_liv_puerto_vallarta$h001a_personal_ocupado_total) + abline(lm_accommodations_eu_employed_pv, col = "darkgreen", lwd = 2, lty = 2)
```

Okay, I am going to see all the other municipios...

```{r}
# lm_all_municipios_accommodations <- saic_accommodations_liv %>%
#   filter(
#     !actividad_economica %in% "Clase 713930 Marinas turísticas",
#     !is.na(municipio)
#   ) %>%
#   group_by(municipio) %>%
#   summarise(
#     model = list(lm(h001a_personal_ocupado_total ~ ue_unidades_economicas, data = cur_data())),
#     r_squared = summary(model[[1]])$r.squared,
#     slope = coef(model[[1]])[2],
#     intercept = coef(model[[1]])[1]
#     # p_value = summary(model[[1]])$coefficients[2, 4] # ran into some municipios not having enough data so it was NA!!
#   )
# 
# print(lm_all_municipios_accommodations)

lm_all_municipios_accommodations_alt <- saic_accommodations_liv %>%
  filter(
    !actividad_economica %in% "Clase 713930 Marinas turísticas",
    !is.na(municipio)
  ) %>%
  group_by(municipio) %>%
  summarise(
    model = list(lm(h001a_personal_ocupado_total ~ ue_unidades_economicas, data = cur_data())),
    r_squared = {
      sm <- summary(model[[1]])
      if(length(sm$coefficients[,1]) >= 2) sm$r.squared else NA_real_
    },
    slope = {
      sm <- summary(model[[1]])
      if(length(sm$coefficients[,1]) >= 2) coef(model[[1]])[2] else NA_real_
    },
    intercept = {
      sm <- summary(model[[1]])
      if(length(sm$coefficients[,1]) >= 1) coef(model[[1]])[1] else NA_real_
    },
    p_value = {
      sm <- summary(model[[1]])
      if(length(sm$coefficients[,1]) >= 2) sm$coefficients[2,4] else NA_real_
    }
  )

print(lm_all_municipios_accommodations_alt) # wow, all significant and positive! Would be interesting to interpret the intercepts when we have really low numbers of economic units. 
```

## Fishing

```{r}
## read in data
saic_fishing <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251021_163827445_fishing.csv"), skip = 4)

saic_fishing_liv <- saic_fishing %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)"))
```

### Expanding

I am realizing that I need to expand the dataset so that every combination of ano_censal × municipio × actividad_economica exists, even if the values are missing (so we can clearly see which municipios or years lack data).

```{r}
# first, define the unique combinations we want for each dimension
years <- c("2023","2018","2013","2008","2003")
municipios <- unique(saic_fishing_liv$municipio)
actividades_fishing <- unique(saic_fishing_liv$actividad_economica)

# then we can make a full grid of all possible combinations
complete_grid_fishing <- expand_grid(
  ano_censal = years,
  municipio = municipios,
  actividad_economica = actividades_fishing
)

# join back with the OG dataset
saic_fishing_liv_expanded <- complete_grid_fishing %>%
  left_join(saic_fishing_liv, 
            by = c("ano_censal", "municipio", "actividad_economica"))

# final step: fill entidad where it is missing
saic_fishing_liv_expanded <- saic_fishing_liv_expanded %>%
  group_by(municipio) %>%
  fill(entidad, .direction = "downup") %>%
  ungroup()
```

### Now, how much data do we have?

```{r}
all_complete_fishing <- saic_fishing_liv_expanded %>%
  filter(!actividad_economica %in% "Total municipal") %>% 
  summarise(
    total_cells = n(),
    with_data = sum(!is.na(h001a_personal_ocupado_total)),
    completeness_pct = with_data / total_cells * 100
  )

by_muni_complete_fishing <- saic_fishing_liv_expanded %>%
  filter(!actividad_economica %in% "Total municipal") %>% 
  group_by(municipio) %>%
  summarise(
    n_records = n(),
    n_missing = sum(is.na(h001a_personal_ocupado_total)),
    completeness_pct = 100 * (1 - n_missing / n_records)
  ) %>%
  arrange(completeness_pct)


by_year_complete_fishing <- saic_fishing_liv_expanded %>%
  filter(!actividad_economica %in% "Total municipal") %>% 
  group_by(ano_censal) %>%
  summarise(
    total = n(),
    non_missing = sum(!is.na(h001a_personal_ocupado_total)),
    completeness_pct = 100 * non_missing / total
  )

by_actividad_complete_fishing <- saic_fishing_liv_expanded %>%
  filter(!actividad_economica %in% "Total municipal") %>% 
  group_by(actividad_economica) %>%
  summarise(
    total = n(),
    non_missing = sum(!is.na(h001a_personal_ocupado_total)),
    completeness_pct = 100 * non_missing / total
  ) %>%
  arrange(desc(completeness_pct)) # oof

library(ggplot2)

## uncensored
saic_fishing_liv_expanded %>%
  filter(!actividad_economica %in% "Total municipal") %>% 
  group_by(municipio, ano_censal) %>%
  summarise(has_data = any(!is.na(h001a_personal_ocupado_total))) %>%
  ggplot(aes(x = factor(ano_censal), y = municipio, fill = has_data)) +
  geom_tile(color = "white") +
  scale_fill_manual(values = c("TRUE" = "darkblue", "FALSE" = "grey80")) +
  theme_minimal() +
  labs(
    x = "Census Year",
    y = "Municipio",
    fill = "Has data for at least one of the sectors",
    title = "Fishing: Data availability by municipio and census year"
  )

## economic units
saic_fishing_liv_expanded %>%
  filter(!actividad_economica %in% "Total municipal") %>% 
  group_by(municipio, ano_censal) %>%
  summarise(has_data = any(!is.na(ue_unidades_economicas))) %>%
  ggplot(aes(x = factor(ano_censal), y = municipio, fill = has_data)) +
  geom_tile(color = "white") +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "grey80")) +
  theme_minimal() +
  labs(
    x = "Census Year",
    y = "Municipio",
    fill = "Has data for at least one of the sectors",
    title = "Fishing: Economic units present per municipio and census year"
  )
```
### Clean LIV

So, what I really want to know is out of the observations that have registered economic units, how many observations have data versus how many are suppressed?

First, let's make a clean dataset with columns that indicate whether the data has been suppressed or not!!

```{r}
# --- func to classify SAIC data completeness and confidentiality ---
classify_saic_status <- function(df,
                                 ue_col = "ue_unidades_economicas", # economic units
                                 var_check = "h001a_personal_ocupado_total") { # number of employees: if this column contains data, then all other variables do as well.
  df %>%
    mutate(
      data_status = case_when(
        # case 1: if there is a numeric UE, non-NA, and there is also data on number of employees... it should be valid!
        !is.na(.data[[ue_col]]) & .data[[ue_col]] > 0 & !is.na(.data[[var_check]]) ~ "valid",
        
        # case 2: if there is a non-NA numerica UE, but the employment data is NA... then it was likely suppressed. From the methodological document: "For cells that appear empty, the absolute values of the economic variables were omitted to ensure confidentiality."
        !is.na(.data[[ue_col]]) & .data[[ue_col]] > 0 & is.na(.data[[var_check]]) ~ "suppressed",
        
        # case 3: if there were no units registered no activity, and therefore in the raw dataset there was no observation for that economic activity, then we can say it has no activity.
        is.na(.data[[ue_col]]) ~ "no_activity",
        
        # case 4: if the UEs exist but the numbers are reported as zeros, then it is either a low number or rounded to zero: "Furthermore, when a zero appears in a row, it is because the respondents declared zero for the variable or because the reported value was less than $500,000, and since the figures are presented in millions of pesos, it was rounded to zero. Negative figures are indicated with a minus sign (-) to the left of the data."
        !is.na(.data[[ue_col]]) & .data[[ue_col]] > 0 & .data[[var_check]] == 0 ~ "low_or_rounded",
        
        TRUE ~ NA_character_ # otherwise, keep NA.
      )
    )
}

saic_fishing_liv_clean <- classify_saic_status(saic_fishing_liv_expanded) %>% 
  filter(!actividad_economica %in% "Total municipal",
         !is.na(actividad_economica))

saic_fishing_liv_clean_check <- saic_fishing_liv_clean %>% 
  select(ano_censal, entidad, municipio, actividad_economica, ue_unidades_economicas, h001a_personal_ocupado_total, data_status) ## looks great!! 001 Acaponeta looks about right: <https://www.emis.com/php/company-profile/MX/La_Princesa_De_Acaponeta_SPR_de_RL_es_11105923.html>, <https://dsiappsdev.semarnat.gob.mx/inai/F69/2021/138/4T/18NA2020PD043.pdf> 
```

### LIV plotly

```{r}
### only Subrama 11411 Pesca
employment_mun <- saic_fishing_liv_clean %>%
  filter(actividad_economica %in% "Subrama 11411 Pesca")

library(plotly)

fig <- plot_ly(
  data = employment_mun,
  x = ~ano_censal,
  y = ~h001a_personal_ocupado_total,
  color = ~municipio,
  text = ~municipio, 
  type = "scatter",
  mode = "lines+markers",
  hovertemplate = paste(
    "<b>Municipio:</b> %{text}<br>",
    "<b>Año:</b> %{x}<br>",
    "<b>Empleo total:</b> %{y:,.0f}<extra></extra>"
  )
) %>%
  layout(
    title = "Empleo pesca total por municipio: Subrama 11411 Pesca (SAIC)",
    xaxis = list(title = "Año censal"),
    yaxis = list(title = "Personas empleadas"),
    legend = list(title = list(text = "Municipio"))
  )

fig

htmlwidgets::saveWidget(fig, here::here("LIV/liv_int_plotlys","fishing_number_jobs_saic_110625.html"), selfcontained = TRUE)

### only UEs from Subrama 11411 Pesca
employment_mun <- saic_fishing_liv_clean %>%
  filter(actividad_economica %in% "Subrama 11411 Pesca")

library(plotly)

fig <- plot_ly(
  data = employment_mun,
  x = ~ano_censal,
  y = ~ue_unidades_economicas,
  color = ~municipio,
  text = ~municipio, 
  type = "scatter",
  mode = "lines+markers",
  hovertemplate = paste(
    "<b>Municipio:</b> %{text}<br>",
    "<b>Año:</b> %{x}<br>",
    "<b>Unidades Economicas total:</b> %{y:,.0f}<extra></extra>"
  )
) %>%
  layout(
    title = "Unidades economicas por municipio: Subrama 11411 Pesca (SAIC)",
    xaxis = list(title = "Año censal"),
    yaxis = list(title = "Personas empleadas"),
    legend = list(title = list(text = "Municipio"))
  )

fig

htmlwidgets::saveWidget(fig, here::here("LIV/liv_int_plotlys","fishing_ue_saic_110625.html"), selfcontained = TRUE)


## only Clase 114119 Pesca y captura de otros peces, crustáceos, moluscos y otras especies
employment_mun_moluscos <- saic_fishing_liv_clean %>%
  filter(actividad_economica %in% "Clase 114119 Pesca y captura de otros peces, crustáceos, moluscos y otras especies")

fig_moluscos <- plot_ly(
  data = employment_mun_moluscos,
  x = ~ano_censal,
  y = ~h001a_personal_ocupado_total,
  color = ~municipio,
  text = ~municipio, 
  type = "scatter",
  mode = "lines+markers",
  hovertemplate = paste(
    "<b>Municipio:</b> %{text}<br>",
    "<b>Año:</b> %{x}<br>",
    "<b>Empleo total:</b> %{y:,.0f}<extra></extra>"
  )
) %>%
  layout(
    title = "Empleo pesca total por municipio: Clase 114119 Pesca y captura de otros peces, crustáceos, moluscos y otras especies (SAIC)",
    xaxis = list(title = "Año censal"),
    yaxis = list(title = "Personas empleadas"),
    legend = list(title = list(text = "Municipio"))
  )

fig_moluscos

# htmlwidgets::saveWidget(fig_moluscos, here::here("LIV/liv_int_plotlys","fishing_moluscos_otros_number_jobs_saic_110625.html"), selfcontained = TRUE)

## NOW ITS UEs: only Clase 114119 Pesca y captura de otros peces, crustáceos, moluscos y otras especies
employment_mun_moluscos <- saic_fishing_liv_clean %>%
  filter(actividad_economica %in% "Clase 114119 Pesca y captura de otros peces, crustáceos, moluscos y otras especies")

fig_moluscos <- plot_ly(
  data = employment_mun_moluscos,
  x = ~ano_censal,
  y = ~ue_unidades_economicas,
  color = ~municipio,
  text = ~municipio, 
  type = "scatter",
  mode = "lines+markers",
  hovertemplate = paste(
    "<b>Municipio:</b> %{text}<br>",
    "<b>Año:</b> %{x}<br>",
    "<b>Unidades Economicas total:</b> %{y:,.0f}<extra></extra>"
  )
) %>%
  layout(
    title = "Unidades Economicas por municipio: Clase 114119 Pesca y captura de otros peces, crustáceos, moluscos y otras especies (SAIC)",
    xaxis = list(title = "Año censal"),
    yaxis = list(title = "Personas empleadas"),
    legend = list(title = list(text = "Municipio"))
  )

fig_moluscos

htmlwidgets::saveWidget(fig_moluscos, here::here("LIV/liv_int_plotlys","fishing_ue_moluscos_otros_number_jobs_saic_110625.html"), selfcontained = TRUE)
```


```{r}
## only Clase 114111 Pesca de camarón
employment_mun <- saic_fishing_liv_clean %>%
  filter(actividad_economica %in% "Clase 114111 Pesca de camarón")

library(plotly)

fig <- plot_ly(
  data = employment_mun,
  x = ~ano_censal,
  y = ~h001a_personal_ocupado_total,
  color = ~municipio,
  text = ~municipio, 
  type = "scatter",
  mode = "lines+markers",
  hovertemplate = paste(
    "<b>Municipio:</b> %{text}<br>",
    "<b>Año:</b> %{x}<br>",
    "<b>Empleo total:</b> %{y:,.0f}<extra></extra>"
  )
) %>%
  layout(
    title = "Empleo pesca total por municipio: Clase 114111 Pesca de camarón (SAIC)",
    xaxis = list(title = "Año censal"),
    yaxis = list(title = "Personas empleadas"),
    legend = list(title = list(text = "Municipio"))
  )

fig

htmlwidgets::saveWidget(fig, here::here("LIV/liv_int_plotlys","fishing_camaron_number_jobs_saic_110625.html"), selfcontained = TRUE)



## only Clase 114111 Pesca de camarón
employment_mun <- saic_fishing_liv_clean %>%
  filter(actividad_economica %in% "Clase 114111 Pesca de camarón")

library(plotly)

fig <- plot_ly(
  data = employment_mun,
  x = ~ano_censal,
  y = ~h001a_personal_ocupado_total,
  color = ~municipio,
  text = ~municipio, 
  type = "scatter",
  mode = "lines+markers",
  hovertemplate = paste(
    "<b>Municipio:</b> %{text}<br>",
    "<b>Año:</b> %{x}<br>",
    "<b>Empleo total:</b> %{y:,.0f}<extra></extra>"
  )
) %>%
  layout(
    title = "Empleo pesca total por municipio: Clase 114111 Pesca de camarón (SAIC)",
    xaxis = list(title = "Año censal"),
    yaxis = list(title = "Personas empleadas"),
    legend = list(title = list(text = "Municipio"))
  )

fig

htmlwidgets::saveWidget(fig, here::here("LIV/liv_int_plotlys","fishing_camaron_number_jobs_saic_110625.html"), selfcontained = TRUE)

## only Clase 114113 Pesca de sardina y anchoveta
employment_mun <- saic_fishing_liv_clean %>%
  filter(actividad_economica %in% "Clase 114113 Pesca de sardina y anchoveta")

library(plotly)

fig <- plot_ly(
  data = employment_mun,
  x = ~ano_censal,
  y = ~h001a_personal_ocupado_total,
  color = ~municipio,
  text = ~municipio, 
  type = "scatter",
  mode = "lines+markers",
  hovertemplate = paste(
    "<b>Municipio:</b> %{text}<br>",
    "<b>Año:</b> %{x}<br>",
    "<b>Empleo total:</b> %{y:,.0f}<extra></extra>"
  )
) %>%
  layout(
    title = "Empleo pesca total por municipio: Clase 114113 Pesca de sardina y anchoveta (SAIC)",
    xaxis = list(title = "Año censal"),
    yaxis = list(title = "Personas empleadas"),
    legend = list(title = list(text = "Municipio"))
  )

fig

htmlwidgets::saveWidget(fig, here::here("LIV/liv_int_plotlys","fishing_sardina_anchoveta_number_jobs_saic_110625.html"), selfcontained = TRUE)
```

### ECO

```{r}
saic_fishing_liv_loscabos <- saic_fishing_liv %>% 
  filter(municipio %in% "008 Los Cabos") # none

saic_fishing_eco <- saic_fishing %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))

## expanding ECO
# first, define the unique combinations we want for each dimension
years <- c("2023","2018","2013","2008","2003")
municipios <- unique(saic_fishing_liv$municipio)
actividades_fishing <- unique(saic_fishing_liv$actividad_economica)

# then we can make a full grid of all possible combinations
complete_grid_fishing <- expand_grid(
  ano_censal = years,
  municipio = municipios,
  actividad_economica = actividades_fishing
)

# join back with the OG dataset
saic_fishing_liv_expanded <- complete_grid_fishing %>%
  left_join(saic_fishing_liv, 
            by = c("ano_censal", "municipio", "actividad_economica"))

# final step: fill entidad where it is missing
saic_fishing_liv_expanded <- saic_fishing_liv_expanded %>%
  group_by(municipio) %>%
  fill(entidad, .direction = "downup") %>%
  ungroup()
```

## Aquaculture

```{r}
## read in data
saic_aquaculture <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251021_164011725_aquaculture.csv"), skip = 4)

saic_aquaculture_liv <- saic_aquaculture %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)"))
```

```{r}
# ----- for clase ---------
## see which municipios are missing data for at least one year, and for how many years it is NA.
saic_aquaculture_liv_check_complete <- saic_aquaculture_liv %>% 
  filter(ano_censal %in% c("2023","2018","2013","2008","2003")) %>% 
  filter(actividad_economica %in% "Clase 112513 Camaronicultura y acuicultura de otros crustáceos en agua salada") %>% 
  group_by(municipio) %>% 
  summarise(
    na_values = sum(is.na(h001a_personal_ocupado_total))
  )  ## if it is NA for every year, it has an na_values of 5. 

#### THIS IS NOT A GOOD WAY!!! BECAUSE SOME OF THE MUNICIPIOS DO NOT HAVE ENTRIES FOR EVERY YEAR!!!!
  
saic_aquaculture_liv_noncomplete <- saic_aquaculture_liv_check_complete %>% 
  filter(!na_values == 0) # any that have NAs for one or more years.

############## percent of municipios that have at least one year of data missing: #########
length(unique(saic_aquaculture_liv_noncomplete$municipio))/length(unique(saic_aquaculture_liv_check_complete$municipio)) * 100 # v2025: 96.66667% of municipios have at least one year missing data.

setdiff(unique(saic_aquaculture_liv$municipio), unique(saic_aquaculture_liv_check_complete$municipio)) #  
# [1] "006 San Quintín"           "007 San Felipe"            "002 Mulegé"                "008 Los Cabos"            
#  [5] "009 Loreto"                "020 Cabo Corrientes"       "067 Puerto Vallarta"       "008 Xalisco"              
#  [9] "011 Ruíz"                  "013 San Pedro Lagunillas"  "014 Santa María del Oro"   "020 Bahía de Banderas"    
# [13] "004 Concordia"             "005 Cosalá"                "007 Choix"                 "010 El Fuerte"            
# [17] "013 Mocorito"              "017 Sinaloa"               "020 Juan José Ríos"        "017 Caborca"              
# [21] "047 Pitiquito"             "055 San Luis Río Colorado" NA  
# are missing any registered economic units.
```


```{r}
# ----- for subrama --------
## see which municipios are missing data for at least one year, and for how many years it is NA.
saic_aquaculture_liv_check_complete <- saic_aquaculture_liv %>% 
  filter(ano_censal %in% c("2023","2018","2013","2008","2003")) %>% 
  filter(actividad_economica %in% "Subrama 11251 Acuicultura") %>% 
  group_by(municipio) %>% 
  summarise(
    na_values = sum(is.na(h001a_personal_ocupado_total))
  )  ## if it is NA for every year, it has an na_values of 5. 
  
saic_aquaculture_liv_noncomplete <- saic_aquaculture_liv_check_complete %>% 
  filter(!na_values == 0) # any that have NAs for one or more years.

############## percent of municipios that have at least one year of data missing: #########
length(unique(saic_aquaculture_liv_noncomplete$municipio))/length(unique(saic_aquaculture_liv_check_complete$municipio)) * 100 # v2025: 92.30769% of municipios have at least one year missing data for the subrama acricultura

setdiff(unique(saic_aquaculture_liv$municipio), unique(saic_aquaculture_liv_check_complete$municipio)) #  
#  [1] "007 San Felipe"            "008 Los Cabos"             "009 Loreto"                "020 Cabo Corrientes"      
#  [5] "011 Ruíz"                  "020 Bahía de Banderas"     "004 Concordia"             "007 Choix"                
#  [9] "013 Mocorito"              "017 Sinaloa"               "020 Juan José Ríos"        "047 Pitiquito"            
# [13] "055 San Luis Río Colorado" NA  
# are missing any registered economic units.
```
### Expanding

I am realizing that I need to expand the dataset so that every combination of ano_censal × municipio × actividad_economica exists, even if the values are missing (so we can clearly see which municipios or years lack data).

```{r}
# first, define the unique combinations we want for each dimension
years <- c("2023","2018","2013","2008","2003")
municipios <- unique(saic_aquaculture_liv$municipio)
actividades_aqua <- unique(saic_aquaculture_liv$actividad_economica)

# > years
# [1] "2023" "2018" "2013" "2008" "2003"
# > municipios <- unique(saic_aquaculture_liv$municipio)
# > municipios
#  [1] "002 Mexicali"              "006 San Quintín"           "007 San Felipe"           
#  [4] "001 Comondú"               "002 Mulegé"                "003 La Paz"               
#  [7] "008 Los Cabos"             "009 Loreto"                "020 Cabo Corrientes"      
# [10] "067 Puerto Vallarta"       "001 Acaponeta"             "004 Compostela"           
# [13] "008 Xalisco"               "010 Rosamorada"            "011 Ruíz"                 
# [16] "012 San Blas"              "013 San Pedro Lagunillas"  "014 Santa María del Oro"  
# [19] "015 Santiago Ixcuintla"    "016 Tecuala"               "017 Tepic"                
# [22] "018 Tuxpan"                "020 Bahía de Banderas"     "001 Ahome"                
# [25] "002 Angostura"             "004 Concordia"             "005 Cosalá"               
# [28] "006 Culiacán"              "007 Choix"                 "008 Elota"                
# [31] "009 Escuinapa"             "010 El Fuerte"             "011 Guasave"              
# [34] "012 Mazatlán"              "013 Mocorito"              "014 Rosario"              
# [37] "015 Salvador Alvarado"     "016 San Ignacio"           "017 Sinaloa"              
# [40] "018 Navolato"              "019 Eldorado"              "020 Juan José Ríos"       
# [43] "012 Bácum"                 "017 Caborca"               "018 Cajeme"               
# [46] "025 Empalme"               "029 Guaymas"               "030 Hermosillo"           
# [49] "033 Huatabampo"            "047 Pitiquito"             "048 Puerto Peñasco"       
# [52] "055 San Luis Río Colorado" NA                         
# > actividades <- unique(saic_aquaculture_liv$actividad_economica)
# > actividades
# [1] "Total municipal"                                                              
# [2] "Subrama 11251 Acuicultura"                                                    
# [3] "Clase 112515 Piscicultura en agua dulce"                                      
# [4] "Clase 112516 Piscicultura en agua salada"                                     
# [5] "Clase 112519 Otra acuicultura"                                                
# [6] "Clase 112513 Camaronicultura y acuicultura de otros crustáceos en agua salada"
# [7] "Clase 112514 Camaronicultura y acuicultura de otros crustáceos en agua dulce" 
# [8] NA      

# then we can make a full grid of all possible combinations
complete_grid_aquaculture <- expand_grid(
  ano_censal = years,
  municipio = municipios,
  actividad_economica = actividades_aqua
)

# join back with the OG dataset
saic_aquaculture_liv_expanded <- complete_grid_aquaculture %>%
  left_join(saic_aquaculture_liv, 
            by = c("ano_censal", "municipio", "actividad_economica"))

# final step: fill entidad where it is missing
saic_aquaculture_liv_expanded <- saic_aquaculture_liv_expanded %>%
  group_by(municipio) %>%
  fill(entidad, .direction = "downup") %>%
  ungroup()
```

### Now, how much data do we have?

```{r}
all_complete_aqua <- saic_aquaculture_liv_expanded %>%
  summarise(
    total_cells = n(),
    with_data = sum(!is.na(h001a_personal_ocupado_total)),
    completeness_pct = with_data / total_cells * 100
  )

by_muni_complete_aqua <- saic_aquaculture_liv_expanded %>%
  group_by(municipio) %>%
  summarise(
    n_records = n(),
    n_missing = sum(is.na(h001a_personal_ocupado_total)),
    completeness_pct = 100 * (1 - n_missing / n_records)
  ) %>%
  arrange(completeness_pct)


by_year_complete_aqua <- saic_aquaculture_liv_expanded %>%
  group_by(ano_censal) %>%
  summarise(
    total = n(),
    non_missing = sum(!is.na(h001a_personal_ocupado_total)),
    completeness_pct = 100 * non_missing / total
  ) # fairly consistent

by_actividad_complete_aqua <- saic_aquaculture_liv_expanded %>%
  group_by(actividad_economica) %>%
  summarise(
    total = n(),
    non_missing = sum(!is.na(h001a_personal_ocupado_total)),
    completeness_pct = 100 * non_missing / total
  ) %>%
  arrange(desc(completeness_pct)) # oof

library(ggplot2)

## uncensored
saic_aquaculture_liv_expanded %>%
  filter(!actividad_economica %in% "Total municipal") %>% 
  group_by(municipio, ano_censal) %>%
  summarise(has_data = any(!is.na(h001a_personal_ocupado_total))) %>%
  ggplot(aes(x = factor(ano_censal), y = municipio, fill = has_data)) +
  geom_tile(color = "white") +
  scale_fill_manual(values = c("TRUE" = "darkblue", "FALSE" = "grey80")) +
  theme_minimal() +
  labs(
    x = "Census Year",
    y = "Municipio",
    fill = "Has Data",
    title = "Aquaculture: Data availability by municipio and census year"
  )

## economic units
saic_aquaculture_liv_expanded %>%
  filter(!actividad_economica %in% "Total municipal") %>% 
  group_by(municipio, ano_censal) %>%
  summarise(has_data = any(!is.na(ue_unidades_economicas))) %>%
  ggplot(aes(x = factor(ano_censal), y = municipio, fill = has_data)) +
  geom_tile(color = "white") +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "grey80")) +
  theme_minimal() +
  labs(
    x = "Census Year",
    y = "Municipio",
    fill = "Has Data",
    title = "Aquaculture: Economic units present per municipio and census year"
  )
```
### Clean LIV

So, what I really want to know is out of the observations that have registered economic units, how many observations have data versus how many are suppressed?

First, let's make a clean dataset with columns that indicate whether the data has been suppressed or not!!

```{r}
# --- func to classify SAIC data completeness and confidentiality ---
classify_saic_status <- function(df,
                                 ue_col = "ue_unidades_economicas", # economic units
                                 var_check = "h001a_personal_ocupado_total") { # number of employees: if this column contains data, then all other variables do as well.
  df %>%
    mutate(
      data_status = case_when(
        # case 1: if there is a numeric UE, non-NA, and there is also data on number of employees... it should be valid!
        !is.na(.data[[ue_col]]) & .data[[ue_col]] > 0 & !is.na(.data[[var_check]]) ~ "valid",
        
        # case 2: if there is a non-NA numerica UE, but the employment data is NA... then it was likely suppressed. From the methodological document: "For cells that appear empty, the absolute values of the economic variables were omitted to ensure confidentiality."
        !is.na(.data[[ue_col]]) & .data[[ue_col]] > 0 & is.na(.data[[var_check]]) ~ "suppressed",
        
        # case 3: if there were no units registered no activity, and therefore in the raw dataset there was no observation for that economic activity, then we can say it has no activity.
        is.na(.data[[ue_col]]) ~ "no_activity",
        
        # case 4: if the UEs exist but the numbers are reported as zeros, then it is either a low number or rounded to zero: "Furthermore, when a zero appears in a row, it is because the respondents declared zero for the variable or because the reported value was less than $500,000, and since the figures are presented in millions of pesos, it was rounded to zero. Negative figures are indicated with a minus sign (-) to the left of the data."
        !is.na(.data[[ue_col]]) & .data[[ue_col]] > 0 & .data[[var_check]] == 0 ~ "low_or_rounded",
        
        TRUE ~ NA_character_ # otherwise, keep NA.
      )
    )
}

saic_aquaculture_liv_clean <- classify_saic_status(saic_aquaculture_liv_expanded) %>% 
  filter(!actividad_economica %in% "Total municipal",
         !is.na(actividad_economica))

saic_aquaculture_liv_clean_check <- saic_aquaculture_liv_clean %>% 
  select(ano_censal, entidad, municipio, actividad_economica, ue_unidades_economicas, h001a_personal_ocupado_total, data_status) ## looks great!! 001 Acaponeta looks about right: <https://www.emis.com/php/company-profile/MX/La_Princesa_De_Acaponeta_SPR_de_RL_es_11105923.html>, <https://dsiappsdev.semarnat.gob.mx/inai/F69/2021/138/4T/18NA2020PD043.pdf> 
```


```{r}
saic_aquaculture_liv_clean %>%
  count(data_status) %>%
  mutate(percentage = n / sum(n) * 100)

saic_aquaculture_liv_clean %>%
  group_by(ano_censal, data_status) %>%
  summarise(n = n()) %>%
  mutate(pct = n / sum(n) * 100)

saic_aquaculture_liv_clean %>%
  group_by(ano_censal, municipio, data_status) %>%
  summarise(n = n()) %>%
  mutate(pct = n / sum(n) * 100) %>% 
  arrange(desc(data_status))


# saic_aquaculture_liv_clean %>%
#   group_by(municipio, ano_censal) %>%
#   summarise(
#     suppressed = any(data_status == "suppressed"),
#     no_activity = all(data_status == "no_activity")
#   ) %>%
#   ggplot(aes(x = factor(ano_censal), y = municipio, fill = suppressed)) +
#   geom_tile(color = "white") +
#   scale_fill_manual(values = c("TRUE" = "darkred", "FALSE" = "grey80")) +
#   theme_minimal() +
#   labs(
#     title = "Municipios con datos suprimidos por confidencialidad (SAIC)",
#     x = "Año censal",
#     y = "Municipio",
#     fill = "Suprimido"
#   )
```

### Relationship between UE and employees

Okay, I am going to see all the municipios...

```{r}
lm_all_municipios_aquaculture <-  saic_aquaculture_liv_clean %>%
  filter(
    !is.na(h001a_personal_ocupado_total)
  ) %>%
  group_by(municipio) %>%
  summarise(
    model = list(lm(h001a_personal_ocupado_total ~ ue_unidades_economicas, data = cur_data())),
    r_squared = {
      sm <- summary(model[[1]])
      if(length(sm$coefficients[,1]) >= 2) sm$r.squared else NA_real_
    },
    slope = {
      sm <- summary(model[[1]])
      if(length(sm$coefficients[,1]) >= 2) coef(model[[1]])[2] else NA_real_
    },
    intercept = {
      sm <- summary(model[[1]])
      if(length(sm$coefficients[,1]) >= 1) coef(model[[1]])[1] else NA_real_
    },
    p_value = {
      sm <- summary(model[[1]])
      if(length(sm$coefficients[,1]) >= 2) sm$coefficients[2,4] else NA_real_
    }
  )

print(lm_all_municipios_aquaculture) 
```

### LIV plotly

```{r}
### only Subrama 11251 Acuicultura
employment_mun <- saic_aquaculture_liv_clean %>%
  filter(actividad_economica %in% "Subrama 11251 Acuicultura")

library(plotly)

fig <- plot_ly(
  data = employment_mun,
  x = ~ano_censal,
  y = ~h001a_personal_ocupado_total,
  color = ~municipio,
  text = ~municipio, 
  type = "scatter",
  mode = "lines+markers",
  hovertemplate = paste(
    "<b>Municipio:</b> %{text}<br>",
    "<b>Año:</b> %{x}<br>",
    "<b>Empleo total:</b> %{y:,.0f}<extra></extra>"
  )
) %>%
  layout(
    title = "Empleo acuícola total por municipio: Subrama 11251 Acuicultura (SAIC)",
    xaxis = list(title = "Año censal"),
    yaxis = list(title = "Personas empleadas"),
    legend = list(title = list(text = "Municipio"))
  )

fig

htmlwidgets::saveWidget(fig, here::here("LIV/liv_int_plotlys","aquaculture_number_jobs_saic_110625.html"), selfcontained = TRUE)

## only unidades economicas from Subrama 11251 Acuicultura

employment_mun <- saic_aquaculture_liv_clean %>%
  filter(actividad_economica %in% "Subrama 11251 Acuicultura")

library(plotly)

fig <- plot_ly(
  data = employment_mun,
  x = ~ano_censal,
  y = ~ue_unidades_economicas,
  color = ~municipio,
  text = ~municipio, 
  type = "scatter",
  mode = "lines+markers",
  hovertemplate = paste(
    "<b>Municipio:</b> %{text}<br>",
    "<b>Año:</b> %{x}<br>",
    "<b>Unidades Economicas total:</b> %{y:,.0f}<extra></extra>"
  )
) %>%
  layout(
    title = "Unidades Economicas por municipio: Subrama 11251 Acuicultura (SAIC)",
    xaxis = list(title = "Año censal"),
    yaxis = list(title = "Unidades Economicas"),
    legend = list(title = list(text = "Municipio"))
  )

fig

htmlwidgets::saveWidget(fig, here::here("LIV/liv_int_plotlys","aquaculture_ue_saic_110625.html"), selfcontained = TRUE)


## only Clase 112513 Camaronicultura y acuicultura de otros crustáceos en agua salada

employment_mun <- saic_aquaculture_liv_clean %>%
  filter(actividad_economica %in% "Clase 112513 Camaronicultura y acuicultura de otros crustáceos en agua salada")

fig <- plot_ly(
  data = employment_mun,
  x = ~ano_censal,
  y = ~h001a_personal_ocupado_total,
  color = ~municipio,
  text = ~municipio, 
  type = "scatter",
  mode = "lines+markers",
  hovertemplate = paste(
    "<b>Municipio:</b> %{text}<br>",
    "<b>Año:</b> %{x}<br>",
    "<b>Empleo total:</b> %{y:,.0f}<extra></extra>"
  )
) %>%
  layout(
    title = "Empleo acuícola total por municipio: Clase 112513 Camaronicultura y acuicultura de otros crustáceos en agua salada (SAIC)",
    xaxis = list(title = "Año censal"),
    yaxis = list(title = "Personas empleadas"),
    legend = list(title = list(text = "Municipio"))
  )

fig

# htmlwidgets::saveWidget(fig, here::here("LIV/liv_int_plotlys","aquaculture_camaronicultura_salada_number_jobs_saic_110625.html"), selfcontained = TRUE)

## only Clase 112516 Piscicultura en agua salada
employment_mun <- saic_aquaculture_liv_clean %>%
  filter(actividad_economica %in% "Clase 112516 Piscicultura en agua salada")

fig <- plot_ly(
  data = employment_mun,
  x = ~ano_censal,
  y = ~h001a_personal_ocupado_total,
  color = ~municipio,
  text = ~municipio, 
  type = "scatter",
  mode = "lines+markers",
  hovertemplate = paste(
    "<b>Municipio:</b> %{text}<br>",
    "<b>Año:</b> %{x}<br>",
    "<b>Empleo total:</b> %{y:,.0f}<extra></extra>"
  )
) %>%
  layout(
    title = "Empleo acuícola total por municipio: Clase 112516 Piscicultura en agua salada (SAIC)",
    xaxis = list(title = "Año censal"),
    yaxis = list(title = "Personas empleadas"),
    legend = list(title = list(text = "Municipio"))
  )

fig # none. all nas!!


## only Clase 112519 Otra acuicultura
employment_mun <- saic_aquaculture_liv_clean %>%
  filter(actividad_economica %in% "Clase 112519 Otra acuicultura")

fig <- plot_ly(
  data = employment_mun,
  x = ~ano_censal,
  y = ~h001a_personal_ocupado_total,
  color = ~municipio,
  text = ~municipio, 
  type = "scatter",
  mode = "lines+markers",
  hovertemplate = paste(
    "<b>Municipio:</b> %{text}<br>",
    "<b>Año:</b> %{x}<br>",
    "<b>Empleo total:</b> %{y:,.0f}<extra></extra>"
  )
) %>%
  layout(
    title = "Empleo acuícola total por municipio: Clase 112519 Otra acuicultura (SAIC)",
    xaxis = list(title = "Año censal"),
    yaxis = list(title = "Personas empleadas"),
    legend = list(title = list(text = "Municipio"))
  )

fig

```

```{r}

# --- Build plot with dropdown selector ---
municipios_list <- unique(employment_mun$municipio)

# Create one trace per municipio, hidden initially except the first
fig <- plot_ly()

for (i in seq_along(municipios_list)) {
  muni <- municipios_list[i]
  df_sub <- filter(employment_mun, municipio == muni)
  
  fig <- fig %>%
    add_trace(
      data = df_sub,
      x = ~ano_censal,
      y = ~empleo_total,
      type = "scatter",
      mode = "lines+markers",
      name = muni,
      text = ~municipio,  # <-- this makes the hover label correct!
      hovertemplate = paste(
        "<b>Municipio:</b> %{text}<br>",
        "<b>Año:</b> %{x}<br>",
        "<b>Empleo total:</b> %{y:,.0f}<extra></extra>"
      ),
      visible = if (i == 1) TRUE else FALSE
    )
}

# --- Build dropdown buttons ---
buttons <- lapply(seq_along(municipios_list), function(i) {
  vis <- rep(FALSE, length(municipios_list))
  vis[i] <- TRUE
  list(
    method = "update",
    args = list(list(visible = vis),
                list(title = paste("Empleo acuícola –", municipios_list[i]))),
    label = municipios_list[i]
  )
})

# --- Final layout with dropdown menu ---
fig <- fig %>%
  layout(
    title = paste("Empleo acuícola total por municipio (SAIC) –", municipios_list[1]),
    xaxis = list(title = "Año censal"),
    yaxis = list(title = "Personas empleadas"),
    updatemenus = list(
      list(
        y = 1.05,
        buttons = buttons,
        direction = "down",
        showactive = TRUE,
        xanchor = "left",
        yanchor = "top"
      )
    ),
    showlegend = FALSE
  )

fig
```





### ECO

```{r}
saic_aquaculture_liv_loscabos <- saic_aquaculture_liv %>% 
  filter(municipio %in% "008 Los Cabos") # none

saic_aquaculture_eco <- saic_aquaculture %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))
```

## Prep fish & shellfish

```{r}
## read in data
saic_fish_processing <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251021_16511703_prep_package_fish_shellfish.csv"), skip = 4)

saic_fish_processing_liv <- saic_fish_processing %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)"))

saic_fish_processing_liv_loscabos <- saic_fish_processing_liv %>% 
  filter(municipio %in% "008 Los Cabos") # none

saic_fish_processing_eco <- saic_fish_processing %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))
```

## Tourist marinas

```{r}
## read in data
saic_tourist_marina <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251021_17745611_tourist_marinas.csv"), skip = 4)

saic_tourist_marina_liv <- saic_tourist_marina %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)"))

saic_tourist_marina_liv_loscabos <- saic_tourist_marina_liv %>% 
  filter(municipio %in% "008 Los Cabos") 

saic_tourist_marina_eco <- saic_tourist_marina %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))
```

## Museums, historical, natural parks

```{r}
## read in data
saic_natural_parks <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251021_17550467_museums_historical_naturalparks.csv"), skip = 4)

saic_natural_parks_liv <- saic_natural_parks %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)")) # not many...

saic_natural_parks_liv_loscabos <- saic_natural_parks_liv %>% 
  filter(municipio %in% "008 Los Cabos") 

saic_natural_parks_eco <- saic_natural_parks %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))
```

## Public restrooms, parking

```{r}
## read in data
saic_restrooms_parking <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251021_171548286_public_restrooms_parking.csv"), skip = 4)

saic_restrooms_parking_liv <- saic_restrooms_parking %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)"))

saic_restrooms_parking_liv_loscabos <- saic_restrooms_parking_liv %>% 
  filter(municipio %in% "008 Los Cabos") 


saic_restrooms_parking_eco <- saic_restrooms_parking %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))
```

## Food services

```{r}
## read in data
saic_food_services <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251021_171034329_food_services.csv"), skip = 4)

saic_food_services_liv <- saic_food_services %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)"))

saic_food_services_liv_loscabos <- saic_food_services_liv %>% 
  filter(municipio %in% "008 Los Cabos") # 722512: Restaurantes con servicio de preparación de pescados y mariscos

saic_food_services_eco <- saic_food_services %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))
```

## Energy generation

```{r}
## read in data
saic_energy_generation <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251021_16471476_energy_generation.csv"), skip = 4)

# 221111 Generation of electricity from fossil fuels
# 221112 Generation of electricity from hydropower
# 221113 Generation of electricity from solar energy
# 221114 Generation of electricity from wind energy
# 221119 Generation of electricity from other types of energy

saic_energy_generation_liv <- saic_energy_generation %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)"))

saic_energy_generation_liv_loscabos <- saic_energy_generation_liv %>% 
  filter(municipio %in% "008 Los Cabos") # 722512: Restaurantes con servicio de preparación de pescados y mariscos

saic_energy_generation_eco <- saic_energy_generation %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))
```

## Supply natural gas

```{r}
## read in data
saic_natural_gas_supply <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251021_16482728_supply_naturalgas.csv"), skip = 4)

saic_natural_gas_supply_liv <- saic_natural_gas_supply %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)")) # none anywhere really

saic_natural_gas_supply_liv_loscabos <- saic_natural_gas_supply_liv %>% 
  filter(municipio %in% "008 Los Cabos")

saic_natural_gas_supply_eco <- saic_natural_gas_supply %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))
```


## Oil and gas: NONE?

```{r}
## read in data
saic_oil <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251021_164117726_oil_gas.csv"), skip = 4)

saic_oil_liv <- saic_oil %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)")) # none anywhere really

saic_oil_liv_loscabos <- saic_oil_liv %>% 
  filter(municipio %in% "008 Los Cabos")

saic_oil_eco <- saic_oil %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))
```

## Water collection, treatment, supply

```{r}
## read in data
saic_water <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251022_13388238_water_treatment_supply.csv"), skip = 4)

saic_water_liv <- saic_water %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)")) # only in some places like in Nayarit...

saic_water_eco <- saic_water %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))
```

## Professional and scientific services

```{r}
## read in data
saic_sci_services <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251021_165918141_professional_scientific_technical_services.csv"), skip = 4)

saic_sci_services_liv <- saic_sci_services %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)")) # none anywhere really

saic_sci_services_eco <- saic_sci_services %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))
```

## Mining

```{r}
## read in data
saic_mining <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251021_164332559_all_mining.csv"), skip = 4)

# Clase 212391 Minería de sal
# none except Navolato 2023...

saic_mining_liv <- saic_mining %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)")) # none anywhere really

saic_mining_eco <- saic_mining %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))
```


## Hazardous waste

```{r}
## read in data
saic_hazardous_waste <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251021_1712765_hazardous_waste.csv"), skip = 4)

saic_hazardous_waste_liv <- saic_hazardous_waste %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)")) # none anywhere really

saic_hazardous_waste_eco <- saic_hazardous_waste %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))
```

## Possible NP

```{r}
## read in data
saic_possible_np <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251021_165616924_possible_natural_products.csv"), skip = 4)

# only culiacan... and one or two others...

saic_possible_np_liv <- saic_possible_np %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)")) # none anywhere really

saic_possible_np_eco <- saic_possible_np %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))
```


## Job training services

For those unemployed.

```{r}
## read in data
saic_job_training <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251021_17421711_job_training_services.csv"), skip = 4) # only one year for Hermosillo available.

saic_job_training_liv <- saic_job_training %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)")) # none anywhere really

saic_job_training_eco <- saic_job_training %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))
```


## Associations and organizations

813 Associations and organizations

```{r}
## read in data
saic_organizations <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/SAIC/d2025","SAIC_Exporta_20251022_1414982_organizations.csv"), skip = 4) # a lot of data, but might not even be helpful.

saic_organizations_liv <- saic_organizations %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(h|i)")) # a lot more data!!

saic_organizations_eco <- saic_organizations %>%
  janitor::clean_names() %>% 
  select(ano_censal, entidad, municipio,actividad_economica, ue_unidades_economicas, matches("^(a|m)"))
```


## Combine

```{r}
## all _liv objects but exclude _liv_loscabos
liv_objs <- ls(pattern = "_liv$")
liv_objs <- liv_objs[!grepl("_liv_loscabos$", liv_objs)]

all_liv <- mget(liv_objs) %>%
  bind_rows(.id = "source")

## all _eco
all_eco <- mget(ls(pattern = "_eco$")) %>%
  bind_rows(.id = "source")

unique(all_liv$source)
unique(all_eco$source)
```


# Sector shares

```{r}
library(tidyverse)

# List of all sector datasets
sector_datasets <- list(
  accommodations = saic_accommodations_liv,
  fishing = saic_fishing_liv,
  aquaculture = saic_aquaculture_liv,
  fish_processing = saic_fish_processing_liv,
  tourist_marina = saic_tourist_marina_liv,
  natural_parks = saic_natural_parks_liv,
  restrooms_parking = saic_restrooms_parking_liv,
  food_services = saic_food_services_liv,
  energy_generation = saic_energy_generation_liv,
  natural_gas = saic_natural_gas_supply_liv,
  oil_gas = saic_oil_liv,
  water_services = saic_water_liv,
  sci_services = saic_sci_services_liv,
  mining = saic_mining_liv,
  hazardous_waste = saic_hazardous_waste_liv,
  possible_np = saic_possible_np_liv,
  job_training = saic_job_training_liv,
  organizations = saic_organizations_liv
)

# Function to calculate sector share of municipal total
calculate_sector_share <- function(df) {
  
  df %>%
    group_by(ano_censal, entidad, municipio) %>%
    # extract municipal total for that year & municipio
    mutate(
      municipal_total = h001a_personal_ocupado_total[actividad_economica == "Total municipal"],
      proportion = h001a_personal_ocupado_total / municipal_total
    ) %>%
    ungroup()
}

# Apply to all sectors and combine
sector_shares <- map_df(sector_datasets, calculate_sector_share, .id = "sector")

# Optional: select key columns
sector_shares_summary <- sector_shares %>%
  select(sector, ano_censal, entidad, municipio, actividad_economica, h001a_personal_ocupado_total, municipal_total, proportion)

# View example
sector_shares_summary %>% filter(municipio == "008 Los Cabos")

```



