################################################################################

Erica Ferrer
emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################

# PRE-WORK ------------------------------
```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```

# Region Key (December 2025) ------------------------------
```{r Region Keys}
# KEYS FROM MEL (including lat, long):
BajaSonoraKey<- read_csv(here("/home/frazier/ohi_goc_prep/FP/FIS/v2025/data/Baja_Sonora_states_key.csv"))

BajaSonoraKey <- BajaSonoraKey %>%
  mutate(
    long = as.numeric(long),
    lat  = as.numeric(lat)
  )

JaliscoKey<- read_csv(here("/home/frazier/ohi_goc_prep/FP/FIS/v2025/data/landing_sites_jalisco.csv"))

JaliscoKey <- JaliscoKey %>%
  rename(long= lon) %>% 
  mutate(
    long = as.numeric(long),
    lat  = as.numeric(lat)
  )

JaliscoKey_clean <- JaliscoKey %>%
  # Normalize state names
  mutate(state = toupper(state)) %>%
  
  # Remove any rows where landing_site is NA (can cause matching issues later)
  filter(!is.na(landing_site)) %>%
  
  # Keep only one row per landing_site_key; if duplicate keys exist, keep first
  group_by(landing_site_key) %>%
  slice(1) %>%  # keeps the first occurrence per key
  ungroup() %>%
  
  # Optional: also deduplicate by landing_site if there are rows with NA landing_site_key
  group_by(landing_site) %>%
  slice(1) %>%
  ungroup()

# NOTE TO SELF: Come back to this! The Jalisco key is imperfect, but we'll have to fix this in the new year.
# Identify rows that were removed in the deduplication of JaliscoKey
removed_rows <- anti_join(JaliscoKey, JaliscoKey_clean, 
                          by = c("landing_site_key", "landing_site"))


#### MERGE THESE WORKING KEYS FOR NOW: ####
RegionKey_working <- bind_rows(BajaSonoraKey, JaliscoKey_clean) %>%
  filter(!is.na(landing_site)) %>%  # remove the NA from this dataframe as this will create a many-to-many matching issue later on
  mutate(long = as.numeric(long))
# 1176 obs., 9 vars.

RegionKey_working <- RegionKey_working %>% mutate(long = as.numeric(long))
rm(JaliscoKey, JaliscoKey_clean, BajaSonoraKey, removed_rows)

# My regional key:
RegionKey_temporary<- read_csv(here("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Data&Keys/CONAPESCA_Keys/RegionKey_UNDERCONSTRUCTION_Dec02024.csv"))
```

# Load: Landings data with species info (December 2025) ------------------------------
```{r Load in landings data with species info}
Landings_GoC_Original_wSp<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/juan_carlos/GoC_Landings_wSp_2000.present_Dec012025.rds"))
# 3716415 obs, 41 vars.

Landings_GoC_Original_wSp <- Landings_GoC_Original_wSp %>% 
  select(-lat, -long) # remove lat, long for now to do the merge

Landings_GoC <- Landings_GoC_Original_wSp %>%
  left_join(
    RegionKey_working %>%
      select(-landing_site_key),
    by = c("state", "landing_site")
  )
# 3716415 obs, 45 vars.

# A few quick checks:
table(Landings_GoC$GoC)
# All in the GoC or not otherwise noted.
table(Landings_GoC$state)
table(Landings_GoC_Original_wSp$state)
table(Landings_GoC$OHI_rgn)
```

```{r Add OHI region to landings w species}
Landings_GoC <- Landings_GoC %>%
  mutate(Region_pick = if_else(!is.na(OHI_rgn), OHI_rgn, state))
# 3716415 obs, 46 vars. as expected!

Landings_GoC <- Landings_GoC %>%
  left_join(
    RegionKey_temporary %>%
      select(-Region_Name_Proposed),  # remove this column for now
    by = c("Region_pick" = "Region_Mel_or_CONAPESCA")
  )
# 3716415 obs, 47 vars. as expected!

subset<- Landings_GoC %>% filter(is.na(OHI_Region_No))
table(subset$state)
# 156952 obs., 47 vars.
# Only Jalisco entries are not keyed. This makes sense. These are parts of Jalisco that either aren't in Region 9 or we can't be totally sure they are in Region 9.

Landings_GoC<- Landings_GoC %>% filter(!is.na(OHI_Region_No)) 
# 3559463 obs., 47 vars.
# Removed some landings (-156952) from Jalisco.
```

```{r Landed weight correction}
# Kind of looks like sometimes 'landed weight' is 'live weight' and vice versa. Fix this:
Landings_GoC$weight_kg_corrected <- ifelse(
  Landings_GoC$live_weight >= Landings_GoC$landed_weight,
  Landings_GoC$live_weight,
  Landings_GoC$landed_weight
)
```

```{r Filtering out freshwater and MAR}
colnames(Landings_GoC)
table(Landings_GoC$recorded_taxa_group)
table(Landings_GoC$acuaculture_production)

#### Filter out aquaculture landings ####

aquaculture_keywords<- c(
  "DE CULTIVO", "DE CULTIVO VIVO", "ABASTO RANCHO", "RANCHO"
)

##### AQUACULTURE TAXA: UNDER CONSTRUCTION - NOTE TO SELF
aquaculture_taxa<- c(
  "OSTION"
)

# Flag aquaculture rows:
Landings_GoC <- Landings_GoC %>%
  mutate(
    aquaculture_flag = (
      coalesce(tolower(acuaculture_production) == "yes", FALSE) |
      (recorded_taxa_group %in% aquaculture_taxa) |
      (rowSums(across(keywords_1:keywords_5, ~ .x %in% aquaculture_keywords), na.rm = TRUE) > 0)
    )
  )

table(Landings_GoC$aquaculture_flag)
# 136090 observations explicitly flagged as aquaculture (when we include all "OSTION")

#### Filter out freshwater landings ####
freshwater_keywords<- c(
  "DE AGUA DULCE"
)

freshwater_taxa<- c(
  "tilapia",
  "rana", "rana toro",
  "trucha",
  "bagre canal"
)

# Flag freshwater rows:
Landings_GoC <- Landings_GoC %>%
  mutate(
    freshwater_flag = coalesce(
      (common_name_clean %in% freshwater_taxa) |
      (rowSums(across(keywords_1:keywords_5, ~ .x %in% freshwater_keywords), na.rm = TRUE) > 0) |
      (habitat == "freshwater") |
      (IHH_functional_major_groups_sm =="Freshwater fishes"),
      FALSE
    )
  )

table(Landings_GoC$freshwater_flag)
# 286013 observations explicitly flagged as freshwater landings (either based on the taxa, keyword, or landing site)
```

```{r Filter out freshwater and MAR species}
#### Now filter out freshwater & aquaculture landings from main landings ####
FreshFisheries_Landings <- Landings_GoC %>%
  filter(freshwater_flag & !aquaculture_flag)
# 280355 obs., 50 vars

FreshAquaculture_Landings <- Landings_GoC %>%
  filter(freshwater_flag & aquaculture_flag)
# 5658 obs., 50 vars

Mariculture_Landings <- Landings_GoC %>%
  filter(!freshwater_flag & aquaculture_flag)
# 130432 obs., 50 vars

MarineFisheries_Landings <- Landings_GoC %>%
  filter(!(aquaculture_flag | freshwater_flag))
# 3143018 obs., 50 vars

sum(is.na(MarineFisheries_Landings$OHI_Region_No))
table(MarineFisheries_Landings$OHI_Region_No)
# Seems like all entries have an OHI region number...
```
NOTE TO SELF: There could still be some freshwater species hiding in here. For example, the "MOJARRA" general category is mixed freshwater and marine demersal fish.

# Save the FIS and MAR dataframes as of Feb 24 2026
```{r Save the FIS and MAR dataframes as of Feb 24 2026}
saveRDS(MarineFisheries_Landings, "/home/shares/ohi/OHI_GOC/goal_prep/fp/fis/v2025/int/GOC_FIS.landings_CLEAN_as.of.Feb242026.rds")

saveRDS(Mariculture_Landings, "/home/shares/ohi/OHI_GOC/goal_prep/fp/mar/v2025/int/GOC_MAR.landings_CLEAN_as.of.Feb242026.rds")
```



# Fix Major_Functional_Group as of Feb 24 2026
```{r FIXING MAJOR FUNCTIONAL GROUPS}
# NOTE TO SELF: Return to this when needed (Dec 09 2025). Returned the next day, Dec 10 2025.
Genus_MajorGroup_Key_ALL <- Landings_GoC %>%
  select(recorded_taxa_group, estimated_genus_name, IHH_functional_major_groups_sm) %>%
  distinct(recorded_taxa_group, estimated_genus_name, IHH_functional_major_groups_sm, .keep_all = TRUE)
# 367 obs., 3 vars.
# NOW: 381 obs., 3 vars.

# See the issue here? One genus could be associated with multiple recorded_taxa_groups. We need the unique combination of estimated_genus_name & recorded_taxa_group to key out the larger dataset. But, in the end, we'll want to be able to group by ESTIMATED_taxa_group to avoid issues of splitting the catch from a single genus across two larger functional groups.

# ---- Create Major_Functional_Group values based IHH major functional groups assigned to recorded_taxa_group ----

# Step 1: Assign one Major_Functional_Group per recorded_taxa_group, respecting that a few of the categories (like "OTRAS") are inherently different.
skip_groups <- c("OTRAS", "ESPECIE GENERICA", "FAUNA", "ORNATO", "NEI")

Genus_MajorGroup_Key_ALL$Assignment_for_Major_Functional_Group = NA_character_  # initialize flag column

Genus_MajorGroup_Key_ALL <- Genus_MajorGroup_Key_ALL %>%
  group_by(recorded_taxa_group) %>%
  mutate(
    
    # Most frequent non-NA value for IHH_functional_major_groups within the recorded_taxa_group
    group_major = {
      x <- IHH_functional_major_groups_sm
      x <- x[!is.na(x) & x != ""]
      if (length(x) == 0) NA_character_ else {
        tbl <- table(x)
        names(tbl)[which.max(tbl)]
      }
    },
    
    # Assign only to non-skipped groups
    Major_Functional_Group = if_else(
      !(recorded_taxa_group %in% skip_groups),
      group_major,
      NA_character_
    ),
      Assignment_for_Major_Functional_Group = if_else(
      !(recorded_taxa_group %in% skip_groups) & is.na(Assignment_for_Major_Functional_Group),
      "Assigned_Step_1: based on recorded_taxa_group and IHH_functional_major_groups",
      Assignment_for_Major_Functional_Group
    )
  ) %>%
  ungroup() %>%
  select(-group_major)

# Put Assignment_for_Major_Functional_Group column at the end:
Genus_MajorGroup_Key_ALL <- Genus_MajorGroup_Key_ALL %>%
  relocate(Assignment_for_Major_Functional_Group, .after = last_col())


# Step 2: Genus-level assignment (ignores cases where genus = NA or NEI)
# Calculates the most common functional group within each genus; uses that majority value to fill missing (or “skip”) records; leaves everything else unchanged
Genus_MajorGroup_Key_ALL <- Genus_MajorGroup_Key_ALL %>%
  group_by(estimated_genus_name) %>%
  mutate(

    # ---- Determine genus_major ONLY IF genus is real (not NA, not NEI) ----
    genus_major = {
      genus_value <- cur_group()$estimated_genus_name   # single value

      if (!is.na(genus_value) && genus_value != "NEI") {

        x <- c(Major_Functional_Group, IHH_functional_major_groups_sm)
        x <- x[!is.na(x) & x != ""]
        if (length(x) == 0) NA_character_ else {
          tbl <- table(x)
          names(tbl)[which.max(tbl)]
        }

      } else {
        # If genus is NA or NEI, then cannot derive Functional_Group
        NA_character_
      }
    },

    Major_Functional_Group = case_when(

      # ---- CASE 1: if genus is NA or NEI → leave Functional_Group unchanged ----
      (is.na(estimated_genus_name) | estimated_genus_name == "NEI") ~ Major_Functional_Group,

      # ---- CASE 2: non-skip taxa with missing Functional_Group → fill Functional_Group with genus_major ----
      !(recorded_taxa_group %in% skip_groups) &
        is.na(Major_Functional_Group) ~ genus_major,

      # ---- CASE 3: if skip taxa → fill Functional_Group *only* from genus_major ----
      recorded_taxa_group %in% skip_groups ~ genus_major,

      # ---- CASE 4: keep original Functional_Group ----
      TRUE ~ Major_Functional_Group
    ),
      Assignment_for_Major_Functional_Group = case_when(
      ((!(recorded_taxa_group %in% skip_groups) & is.na(Assignment_for_Major_Functional_Group) & !is.na(Major_Functional_Group)) |
       (recorded_taxa_group %in% skip_groups & is.na(Assignment_for_Major_Functional_Group) & !is.na(Major_Functional_Group))) ~ "Assigned_Step_2: based on genus (which has been keyed out elsewhere)",
      TRUE ~ Assignment_for_Major_Functional_Group
    )
  ) %>%
  ungroup() %>%
  select(-genus_major)
  

# Step 3: Enforce genus-level consistency for Major_Functional_Group (ignores cases where genus = NA or NEI; assign "OTRAS" that have a verified genus)
Genus_MajorGroup_Key_ALL <- Genus_MajorGroup_Key_ALL %>%
  group_by(estimated_genus_name) %>%
  mutate(
    new_value = if (
      !is.na(cur_group()$estimated_genus_name) &&
      cur_group()$estimated_genus_name != "NEI" &&
      n() > 1
    ) {
      x <- Major_Functional_Group[!is.na(Major_Functional_Group) & Major_Functional_Group != ""]
      if(length(x) == 0) Major_Functional_Group else names(which.max(table(x))) 
    } else {
      Major_Functional_Group
    },
    Assignment_for_Major_Functional_Group = if_else(
      Major_Functional_Group != new_value & !is.na(new_value),
      "Assigned_Step_3: forced consistency across the genus",
      Assignment_for_Major_Functional_Group
    ),
    Major_Functional_Group = new_value
  ) %>%
  ungroup() %>%
  select(-new_value)
  

# Step 4: Manual patches for specific taxonomic groups (and two typos)
Genus_MajorGroup_Key_ALL <- Genus_MajorGroup_Key_ALL %>%
  mutate(
    Major_Functional_Group = case_when(
      estimated_genus_name == "Sphoeriodes" ~ "Marine fishes NEI", # Pressumably, is a typo of "Sphoeroides"
      estimated_genus_name == "Lujtanus" ~ "Demersal marine fishes", # Pressumably, is a typo of "Lutjanus"
      recorded_taxa_group == "MOJARRA" ~ "Marine fishes NEI",
      recorded_taxa_group == "ANCHOVETA" ~ "Pelagic marine fishes",
      recorded_taxa_group == "PECES ORNATO" ~ "Marine fishes NEI",
      recorded_taxa_group %in% c("ALGAS", "SARGAZO") ~ "Marine algae",
      recorded_taxa_group %in% c("PEPINO DE MAR", "ERIZO") ~ "Other marine invertebrates",
      recorded_taxa_group %in% c("PETO", "TIBURON", "CAZON") ~ "Sharks",
      TRUE ~ Major_Functional_Group
    ),
     Assignment_for_Major_Functional_Group = case_when(
      (estimated_genus_name == "Sphoeriodes" |
       estimated_genus_name == "Lujtanus" |
       recorded_taxa_group %in% c("MOJARRA", "ANCHOVETA", "PECES ORNATO", "ALGAS", "SARGAZO",
                                  "PEPINO DE MAR", "ERIZO", "PETO", "TIBURON", "CAZON")) ~ "Assigned_Step_4: manual patch",
      TRUE ~ Assignment_for_Major_Functional_Group
    )
  )


# Step 5: If Major_Functional_Group is NA then Assignment_for_Major_Functional_Group= "could not be assigned"
Genus_MajorGroup_Key_ALL <- Genus_MajorGroup_Key_ALL %>%
  mutate(
    Assignment_for_Major_Functional_Group = if_else(
      is.na(Major_Functional_Group),
      "Alert: Major_Functional_Group could not be assigned (at present)",
      Assignment_for_Major_Functional_Group
    )
  )


# Step 6: Assign estimated_taxa_group by looking at the most common recorded_taxa_group for each genus
Genus_MajorGroup_Key_ALL <- Genus_MajorGroup_Key_ALL %>%
  group_by(estimated_genus_name) %>%
  mutate(
    # Step 1: calculate mode of recorded_taxa_group within the genus
    genus_mode_group = if (!is.na(cur_group()$estimated_genus_name[1]) &&
                           cur_group()$estimated_genus_name[1] != "NEI") {
      tbl <- table(recorded_taxa_group)
      names(tbl)[which.max(tbl)]
    } else {
      NA_character_
    },
    
    # Step 2: assign estimated_taxa_group
    estimated_taxa_group = if_else(
      !is.na(genus_mode_group) & recorded_taxa_group != genus_mode_group,
      genus_mode_group,          # use mode for genus when recorded_taxa_group differs
      recorded_taxa_group        # otherwise keep recorded_taxa_group
    )
  ) %>%
  ungroup() %>%
  select(-genus_mode_group)


# Step 7: A couple of manual patches for specific taxa where we actually lost info due to "OTRAS" (e.g., Lutjanus and "Lujtanus, & Oligoplites)
Genus_MajorGroup_Key_ALL <- Genus_MajorGroup_Key_ALL %>%
  mutate(
    old_estimated_taxa_group = estimated_taxa_group,

    estimated_taxa_group = case_when(
      estimated_genus_name %in% c("Lutjanus","Lujtanus") ~ "GUACHINANGO",
      estimated_genus_name %in% c("Oligoplites") ~ "SARDINA",
      estimated_genus_name %in% c("Pseudobatos") ~ "RAYA Y SIMILARES",
      estimated_genus_name %in% c("Scomberomorus") ~ "SIERRA",
      estimated_genus_name %in% c("Selene") ~ "RONCO",
      estimated_genus_name %in% c("Trachinotus") ~ "PAMPANO",
      TRUE ~ estimated_taxa_group
    ),

    Assignment_for_Major_Functional_Group = case_when(
      !is.na(old_estimated_taxa_group) &
      old_estimated_taxa_group != estimated_taxa_group ~
        "Assigned_Step_7: manual patch for estimated_taxa_group based on specific genus",

      TRUE ~ Assignment_for_Major_Functional_Group
    )
  ) %>%
  select(-old_estimated_taxa_group)

# subset<- Genus_MajorGroup_Key_ALL %>% filter(is.na(Major_Functional_Group))  %>% select(estimated_genus_name)
# print(subset)

# Step 8: Manual patches — force tag for specific genera
Genus_MajorGroup_Key_ALL <- Genus_MajorGroup_Key_ALL %>%
  mutate(
    Major_Functional_Group = case_when(
      # Pelagic marine fishes
      estimated_genus_name %in% c("Makaira","Tetrapturus","Trachurus",
                                  "Brevoortia","Ethmidium","Belone",
                                  "Hemiramphus", "Hemirhamphus") ~ "Pelagic marine fishes",
      
      # Demersal marine fishes
      estimated_genus_name %in% c("Gillichthys","Helicolenus","Bathystoma",
                                  "Bellator","Bairdiella","Cheilotrema",
                                  "Chaetodipterus","Holocentrus","Alphestes",
                                  "Diplodus","Spondyliosoma","Albula",
                                  "Cathorops") ~ "Demersal marine fishes",
      
      # Sharks
      estimated_genus_name %in% c("Pristis","Pseudocarcharias") ~ "Sharks",
      
      # Crustaceans
      estimated_genus_name %in% c("Pagurus","Chaceon","Calappa") ~ "Crustaceans",
      
      # Molluscs excl. Cephalopods
      estimated_genus_name %in% c("Atrina","Pinna") ~ "Molluscs excl. Cephalopods",
      
      # Other marine invertebrates
      estimated_genus_name %in% c("Actinia") ~ "Other marine invertebrates",
      
      # Freshwater fishes
      estimated_genus_name %in% c("Lepidoseus","Cyprinodon") ~ "Freshwater fishes",
      
      # Diadromous fishes
      estimated_genus_name %in% c("Dorosoma","Clupanodon") ~ "Diadromous fishes",
      
      # Other freshwater taxa
      estimated_genus_name %in% c("Rana") ~ "Other freshwater taxa",
      
      # Keep existing value if not in manual list
      TRUE ~ Major_Functional_Group
    ),
    
    # Force tag for all genera you manually worked with
    Assignment_for_Major_Functional_Group = case_when(
      estimated_genus_name %in% c("Makaira","Tetrapturus","Trachurus",
                                  "Brevoortia","Ethmidium","Belone",
                                  "Hemiramphus",
                                  "Gillichthys","Helicolenus","Bathystoma",
                                  "Bellator","Bairdiella","Cheilotrema",
                                  "Chaetodipterus","Holocentrus","Alphestes",
                                  "Diplodus","Spondyliosoma","Albula",
                                  "Cathorops",
                                  "Pristis","Pseudocarcharias",
                                  "Pagurus","Chaceon","Calappa",
                                  "Atrina","Pinna",
                                  "Actinia",
                                  "Lepidoseus","Cyprinodon",
                                  "Dorosoma","Clupanodon",
                                  "Rana") ~
        "Assigned_Step_8: manual patch for Major_Functional_Group based on specific genus",
      
      TRUE ~ Assignment_for_Major_Functional_Group
    )
  )

# Step 9: Rearrange for ease of reading:
Genus_MajorGroup_Key_ALL <- Genus_MajorGroup_Key_ALL %>%
  select(
    recorded_taxa_group,
    estimated_taxa_group,
    estimated_genus_name,
    IHH_functional_major_groups_sm,
    Major_Functional_Group,
    Assignment_for_Major_Functional_Group
  )
    
# 367 obs., 6 vars.
# NOW: 381 obs., 3 vars.
```

#### SAVE GENUS TO MAJOR_FUNCTIONAL_GROUP KEY
```{r Save this key!}
Genus_MajorGroup_Key<- Genus_MajorGroup_Key_ALL %>% 
  select(recorded_taxa_group, estimated_genus_name, estimated_taxa_group, Major_Functional_Group) %>%
  distinct()
# 336 obs., 4 varsity.

write.csv(Genus_MajorGroup_Key, "~/ohi_goc_prep/FP/CONAPESCA_Data&Keys/CONAPESCA_Keys/Genus_to_MajorGroup_Key_Feb242026.csv", row.names = FALSE)
```