---
title: 'OHI GoC 2025: Clean Waters - Eutrophication'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

# Clean Waters: Eutrophication

The Clean Waters goal aims to capture the degree to which local waters are unpolluted by human-made causes based on the various contamination impacts it has on those that enjoy the Gulf.

This was because experts agreed that pollution has negative impacts on the ecological health of the system, human health for those who enjoy swimming in the GoC, and also food quality, such as the levels of heavy metals in fish.

In this subgoal, we will specifically be addressing the amount of nitrogen in coastal sites over time. The ideal state is that the waters are completely unpolluted and safe for both humans and organisms living in the ecosystem.

See <https://github.com/OHI-GOC/OHI_GOC_issues/issues/94> for more information.


```{r}
library(tidyverse)
library(here)
library(sf)
library(mapview)
library(stringdist)
library(stringi)
```
```{r}
supplementary_material_table6_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/romero_gil_2026_estimated_nitrogen_discharge_goc/d2026/romero_gil_2026_supplementary_material_table6.csv"))

supplementary_material_table6_clean <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/romero_gil_2026_estimated_nitrogen_discharge_goc/d2026/romero_gil_2026_supplementary_material_table6.csv")) %>% 
  janitor::clean_names() %>% 
  mutate(municipality = str_to_lower(municipality))
```

```{r}
municipality_to_rgnid <- read_csv(here::here("/home/shares/ohi/OHI_GOC/spatial/municipality_to_rgnid/municipality_to_rgnid.csv")) %>% 
  mutate(municipality = str_to_lower(municipio),
         municipality = str_trim(municipality)) %>% 
  select(municipality, rgn_id)
```

```{r}
table6_rgnid <- left_join(supplementary_material_table6_clean, municipality_to_rgnid) %>% 
  filter(!municipality %in% "ensenada") # pacific side, not gulf
```

Mean wastewater, 10%L c for agriculture, 40% livestock (that is even too high), and I feel like 100% for shrimp

```{r}
nitrogen_inputs <- table6_rgnid %>% 
  select(rgn_id, municipality, wastewater_mean, agriculture_10_percent, livestock_40_percent, shrimp_100_percent)
```

## Reference point

<https://www.dof.gob.mx/nota_detalle.php?codigo=5645374&fecha=11/03/2022#gsc.tab=0>

Official Mexican Standard NOM-001-SEMARNAT-2021, which establishes the permissible limits of pollutants in wastewater discharges into receiving bodies owned by the nation, states that in Mexican marine zones, the Total Nitrogen permissible limits are 25, 30, and 35 mg/L for the PM: Monthly Average, PS: Daily Average, and VI: Instantaneous Value, respectively.

In this Standard, the total nitrogen has been defined as the sum of the concentrations of total *Kjeldahl* nitrogen, nitrites and nitrates, expressed as mg/L of nitrogen. 

The issue is that the modeled amounts are annual values, versus estimated monthly averages.  Additionally, the modeled Total Nitrogen (TN) does not have a specified total mass of ocean water processed per year in metric tons, as it is modeling the estimated TN *entering* the Gulf of California.

I will assume that the estimated TN in tons/year are heterogeneously distributed across all months, meaning that every month, the discharge contributes about the same TN to the GoC.  

Therefore, the ideal is a total 0 TN discharge entering waterways, and a maximum of 25 mg/L entering Mexican marine waters.


```{r}
monthly_tn_tons_estimated <- nitrogen_inputs %>% 
  # monthly estimated values
  mutate(monthly_wastewater_mean = wastewater_mean/12,
         monthly_agriculture_10_percent = agriculture_10_percent/12,
         monthly_livestock_40_percent = livestock_40_percent/12,
         monthly_shrimp_100_percent = shrimp_100_percent/12) %>% 
  mutate(total_monthly_metric_tons_tn = monthly_wastewater_mean + monthly_agriculture_10_percent + monthly_livestock_40_percent + monthly_shrimp_100_percent) %>% 
  
  # daily estimated values
  mutate(daily_wastewater_mean = wastewater_mean/365,
         daily_agriculture_10_percent = agriculture_10_percent/365,
         daily_livestock_40_percent = livestock_40_percent/365,
         daily_shrimp_100_percent = shrimp_100_percent/365) %>% 
  mutate(total_daily_metric_tons_tn = daily_wastewater_mean + daily_agriculture_10_percent + daily_livestock_40_percent + daily_shrimp_100_percent)
```

Note that 1 metric ton = 1,000,000,000 mg.

```{r}
monthly_tn_mg_estimated <- monthly_tn_tons_estimated %>% 
  mutate(total_monthly_mg_tn = total_monthly_metric_tons_tn * 1e9)
```

However, this amount will surely be diluted by the seawater in the GoC.

To calculate the theoretical volume of seawater required to dilute a given amount of nitrogen, we can use mass balance equations (?).

```{r}
estimated_water_volume

monthly_tn_mg_per_liter_estimated <- monthly_tn_mg_estimated %>% 
  mutate(total_monthly_mg_per_liter_tn = total_monthly_mg_tn / estimated_water_volume)
```


^ **This method was very difficult because it is not possible to accurately estimate the volume of water impacted.**

## Alternate reference point and method

I will be reviewing literature to determine if there is an accepted maximum of Total Nitrogen (TN) metric tons per year that enters into marine waters, which could then be used as our reference point.

To do this in a reproductive way, I will be using google scholar, and search using these key phrases:

- "total maximum daily load" nitrogen marine "metric tons"
- "nitrogen loading threshold" eutrophication coastal
- "critical nitrogen load" marine embayment
- TRIX index nitrogen threshold Mediterranean
- "Gulf of California" nitrogen eutrophication loading


I will also research these other frameworks:

- HELCOM (Baltic Sea)
- OSPAR (North Atlantic)
- EPA Chesapeake Bay TMDL
- other Gulf of California environmental assessments




























