---
title: 'OHI GoC 2025: Clean Waters Eutrophication'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: console
editor_options: 
  chunk_output_type: console
---

## Clean Waters: Eutrophication

The Clean Waters goal aims to capture the degree to which local waters are unpolluted by human-made causes based on the various contamination impacts it has on those that enjoy the Gulf.

This was because experts agreed that pollution has negative impacts on the ecological health of the system, human health for those who enjoy swimming in the GoC, and also food quality, such as the levels of heavy metals in fish.

In this subgoal, we will specifically be addressing the amount of nitrogen in coastal sites over time. The ideal state is that the waters are completely unpolluted and safe for both humans and organisms living in the ecosystem.

See <https://github.com/OHI-GOC/OHI_GOC_issues/issues/94> for more information.

### 1. Read in JARG values

```{r}
library(tidyverse)
library(here)
library(sf)
library(mapview)
library(stringdist)
library(stringi)
```

```{r}
supplementary_material_table6_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/romero_gil_2026_estimated_nitrogen_discharge_goc/d2026/romero_gil_2026_supplementary_material_table6.csv"))

supplementary_material_table6_clean <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/romero_gil_2026_estimated_nitrogen_discharge_goc/d2026/romero_gil_2026_supplementary_material_table6.csv")) %>% 
  janitor::clean_names() %>% 
  mutate(municipality = str_to_lower(municipality))
```

```{r}
municipality_to_rgnid <- read_csv(here::here("/home/shares/ohi/OHI_GOC/spatial/municipality_to_rgnid/municipality_to_rgnid.csv")) %>% 
  mutate(municipality = str_to_lower(municipio),
         municipality = str_trim(municipality)) %>% 
  dplyr::select(municipality, rgn_id)
```

```{r}
table6_rgnid <- left_join(supplementary_material_table6_clean, municipality_to_rgnid) %>% 
  filter(!municipality %in% "ensenada") # pacific side, not gulf
```

Mean wastewater, 10%L c for agriculture, 40% livestock (that is even too high), and 100% for shrimp.

```{r}
nitrogen_inputs <- table6_rgnid %>% 
  dplyr::select(rgn_id, municipality, wastewater_mean, agriculture_10_percent, livestock_40_percent, shrimp_100_percent) %>% 
  mutate(municipality = tolower(stri_trans_general(municipality, "Latin-ASCII"))) # clean up municipal names, specifically puerto penasco to not have the ñ.
```

#### OLD reference point....

<https://www.dof.gob.mx/nota_detalle.php?codigo=5645374&fecha=11/03/2022#gsc.tab=0>

Official Mexican Standard NOM-001-SEMARNAT-2021, which establishes the permissible limits of pollutants in wastewater discharges into receiving bodies owned by the nation, states that in Mexican marine zones, the Total Nitrogen permissible limits are 25, 30, and 35 mg/L for the PM: Monthly Average, PS: Daily Average, and VI: Instantaneous Value, respectively.

In this Standard, the total nitrogen has been defined as the sum of the concentrations of total *Kjeldahl* nitrogen, nitrites and nitrates, expressed as mg/L of nitrogen. 

The issue is that the modeled amounts are annual values, versus estimated monthly averages.  Additionally, the modeled Total Nitrogen (TN) does not have a specified total mass of ocean water processed per year in metric tons, as it is modeling the estimated TN *entering* the Gulf of California.

I will assume that the estimated TN in tons/year are heterogeneously distributed across all months, meaning that every month, the discharge contributes about the same TN to the GoC.  

Therefore, the ideal is a total 0 TN discharge entering waterways, and a maximum of 25 mg/L entering Mexican marine waters.


<!-- ```{r} -->
<!-- monthly_tn_tons_estimated <- nitrogen_inputs %>%  -->
<!--   # monthly estimated values -->
<!--   mutate(monthly_wastewater_mean = wastewater_mean/12, -->
<!--          monthly_agriculture_10_percent = agriculture_10_percent/12, -->
<!--          monthly_livestock_40_percent = livestock_40_percent/12, -->
<!--          monthly_shrimp_100_percent = shrimp_100_percent/12) %>%  -->
<!--   mutate(total_monthly_metric_tons_tn = monthly_wastewater_mean + monthly_agriculture_10_percent + monthly_livestock_40_percent + monthly_shrimp_100_percent) %>%  -->

<!--   # daily estimated values -->
<!--   mutate(daily_wastewater_mean = wastewater_mean/365, -->
<!--          daily_agriculture_10_percent = agriculture_10_percent/365, -->
<!--          daily_livestock_40_percent = livestock_40_percent/365, -->
<!--          daily_shrimp_100_percent = shrimp_100_percent/365) %>%  -->
<!--   mutate(total_daily_metric_tons_tn = daily_wastewater_mean + daily_agriculture_10_percent + daily_livestock_40_percent + daily_shrimp_100_percent) -->
<!-- ``` -->

<!-- Note that 1 metric ton = 1,000,000,000 mg. -->

<!-- ```{r} -->
<!-- monthly_tn_mg_estimated <- monthly_tn_tons_estimated %>%  -->
<!--   mutate(total_monthly_mg_tn = total_monthly_metric_tons_tn * 1e9) -->
<!-- ``` -->

However, this amount will surely be diluted by the seawater in the GoC.

To calculate the theoretical volume of seawater required to dilute a given amount of nitrogen, we can use mass balance equations (?).

<!-- ```{r} -->
<!-- estimated_water_volume -->

<!-- monthly_tn_mg_per_liter_estimated <- monthly_tn_mg_estimated %>%  -->
<!--   mutate(total_monthly_mg_per_liter_tn = total_monthly_mg_tn / estimated_water_volume) -->
<!-- ``` -->


^ **This method was very difficult because it is not possible to accurately estimate the volume of water impacted.**

I will be reviewing literature to determine if there is an accepted maximum of Total Nitrogen (TN) metric tons per year that enters into marine waters, which could then be used as our reference point.

To do this in a reproductive way, I will be using google scholar, and search using these key phrases:

- "total maximum daily load" nitrogen marine "metric tons"
- "nitrogen loading threshold" eutrophication coastal
- "critical nitrogen load" marine embayment
- TRIX index nitrogen threshold Mediterranean
- "Gulf of California" nitrogen eutrophication loading


I will also research these other frameworks:

- HELCOM (Baltic Sea)
- OSPAR (North Atlantic)
- EPA Chesapeake Bay TMDL
- other Gulf of California environmental assessments

There were no values that we could refer to that would give us an idea of an ideal or maximum amount of nitrogen discharge from land systems into the ocean. This is because diffusion/dilution in the water column depends on currents, bathymetry, volume of the receiving water body, and the harm caused by the discharge depends on where it enters, not just how much.  And unfortunately, these values are currently at the municipal level, not summarized within each watershed and taken to a pourpoint. 

**In the future, that should be addressed, which will hopefully allow us to rerun the analysis and obtain status scores throughout time as well.  If that ends up happening, and the analysis will be redone by future analysts, then I recommend you look at these possible source-specific reference points that are potentially actionable: <https://github.com/OHI-GOC/OHI_GOC_issues/issues/94#issuecomment-4501656035>.**


### 2. Standardize by coastal length of each municipality


#### 2a. Isloate coastlines of each municipality

To allow for comparison of potential eutrophication pressure across regions of different sizes, total estimated tons of nitrogen are standardized by the length of each region's coastline in kilometers. 

Coastline data were obtained from Natural Earth (version 4.1.0, 1:10m scale; naturalearthdata.com), which derives primarily from World Data Bank 2. The global coastline shapefile was clipped to the Gulf of California study area and reprojected to match the Mexico ITRF2008 Lambert Conformal Conic coordinate reference system. Island coastlines were removed by erasing a 1,500 m buffer around the Islas del Golfo de California protected area polygon prior to calculating lengths, so that only mainland coastline contributed to each region's total. 

Coastline segments were then intersected with each OHI region boundary and summed using st_length in the sf package in R, which returned lengths in meters that were subsequently converted to kilometers. They are then intersected with buffered municipal polygons to obtain coastline lengths for each of the municipalities within each OHI region.

Total estimated mg of pharmaceuticals per region and year were then divided by the corresponding coastline length to produce a standardized pressure metric of mg of pharmaceuticals per kilometer of coastline.

```{r}
world_coastline <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/ne_10m_coastline/ne_10m_coastline.shp"))
marine_and_inland_10 <- st_read(here("/home/shares/ohi/OHI_GOC/spatial/OHI_regions/marine_and_inland10.gpkg"))

st_crs(marine_and_inland10) == st_crs(world_coastline) # FALSE

goc_coastline <- st_intersection(st_transform(world_coastline, st_crs(marine_and_inland10)), marine_and_inland10)
mapview(goc_coastline)
```


```{r}
## I want to get rid of the island coastlines!!
islands_polygons <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/from_hem_POEM_ANP/232_ANP-ITRF08_16062025/232_ANP-ITRF08_16062025.shp")) %>% 
  janitor::clean_names() %>% 
  filter(nombre %in% "Islas del Golfo de California") %>% 
  st_transform(st_crs(goc_coastline))
mapview(islands_polygons)

# erase the island polygons from the municipality polygons, using st_difference, which removes the island area from any municipality that overlaps it
goc_coastline_no_islands <- goc_coastline %>%
  st_difference(st_union(st_buffer(islands_polygons, dist = 1500))) %>% # crop out island areas
  st_make_valid()
mapview(goc_coastline_no_islands) + mapview(goc_coastline)

st_write(goc_coastline, here::here("/home/shares/ohi/OHI_GOC/spatial/OHI_regions","goc_coastline.gpkg"))
st_write(goc_coastline_no_islands, here::here("/home/shares/ohi/OHI_GOC/spatial/OHI_regions","goc_coastline_no_islands.gpkg"))
```

```{r}
municipal_polygons <- st_read(here::here("/home/shares/ohi/OHI_GOC/spatial/delimitaciones_frontera_municipales_inegi/INEGI_Municipal_.shp"))

target_municipalities <- unique(nitrogen_inputs$municipality)
length(target_municipalities) # 28

target_states <- c(
  "Baja California Sur", "Baja California", "Sonora", "Sinaloa", "Nayarit", "Jalisco"
)

municipal_polygons_clean <- municipal_polygons %>%
  mutate(
    NOMBRE = iconv(NOMBRE, from = "latin1", to = "UTF-8"),
    state = str_trim(str_extract(NOMBRE, "^[^,]+")),
    municipality = str_trim(str_extract(NOMBRE, "(?<=,).+$")),
    municipality_lower = tolower(stri_trans_general(municipality, "Latin-ASCII"))
  ) %>%
  filter(state %in% target_states) %>%
  filter(municipality_lower %in% target_municipalities) %>% 
  filter(!NOMBRE %in% "Sonora, Rosario") # JARG does not have this municipality, would get confused with Sinaloa, Rosario

length(target_municipalities)
#[1] 28
length(unique(municipal_polygons_clean$municipality_lower))
#[1] 28

# check which target municipalities are missing
matched <- municipal_polygons_clean$municipality_lower

setdiff(matched, target_municipalities) # character(0)
setdiff(target_municipalities, matched) # character(0)

missing <- target_municipalities[!target_municipalities %in% matched]
if (length(missing) > 0) {
  cat("missing municipalities:\n")
  print(missing)
} # nothing printed!!

## NOW, lets make a buffer around the polygons, and use that to intersect with the coastline vector.
municipal_polygons_clean_meters <- st_transform(municipal_polygons_clean, st_crs(goc_coastline_no_islands)) # make sure it is in meters
municipal_polygons_clean_buffer <- municipal_polygons_clean_meters %>% 
  st_buffer(., dist = 2000)
mapview(municipal_polygons_clean_meters) + mapview(municipal_polygons_clean_buffer)

municipal_coastline1 <- st_intersection(goc_coastline, municipal_polygons_clean_meters)
municipal_coastline2 <- st_intersection(goc_coastline, municipal_polygons_clean_buffer)

mapview(municipal_coastline1, zcol = "municipality_lower") + mapview(municipal_coastline2, zcol = "municipality_lower") + mapview(goc_coastline) + mapview(municipal_coastline2, zcol = "scalerank")

## now remove the islands
municipal_coastline2_noislands <- municipal_coastline2 %>% 
  filter(!scalerank %in% c("5","6"))
mapview(municipal_coastline2_noislands, zcol = "municipality_lower")
```

```{r}
goc_ohi_rgns <- st_read(here::here("spatial/ohi_regions","goc_ohi_rgns.shp"))

union_rgns <- goc_ohi_rgns %>% 
  st_union() %>% 
  st_make_valid()

watershed_all <- st_read(file.path("/home/shares/ohi/stressors_2021/_dataprep/nutrients/watersheds_pourpoints/watersheds_all_4326.shp"))
st_crs(watershed_all)
# mapview(watershed_all)

watershed_goc <- st_intersection(st_transform(watershed_all, st_crs(union_rgns)), union_rgns)
mapview(watershed_goc, zcol = "basin_id") + mapview(poly_concordia)
```


It is better to use the buffer to avoid losing any of the coastline in Sinaloa.  Also, filter by `scalerank` to remove the islands.

```{r}
## find length of the coastline for each region using `st_length`
st_crs(municipal_coastline2_noislands) # meters, Mexico ITRF2008 / LCC 

## calculate coastline length per municipio
coastline_length_by_mun <- municipal_coastline2_noislands %>%
  mutate(segment_length_m = as.numeric(st_length(.))) %>%  # length of each segment in meters
  st_drop_geometry() %>%
  group_by(municipality_lower) %>%
  summarize(coastline_length_m = sum(segment_length_m, na.rm = TRUE), .groups = "drop") %>%
  mutate(coastline_length_km = coastline_length_m / 1000)

# save to mazu
write_csv(coastline_length_by_mun, here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/eutrophication/int","coastline_length_by_municipality_JARG.csv"))
```


#### 2b. Divide each municipality's tons of nitrogen by the km coastline

```{r}
# concordia does not have a coastline, most of the municipality falls on watersheds that lead to mazatlan.
concordia_nitrogen <- nitrogen_inputs %>%
  mutate(all_sources_nitrogen = wastewater_mean + agriculture_10_percent + livestock_40_percent + shrimp_100_percent) %>%
  filter(municipality == "concordia") %>%
  pull(all_sources_nitrogen)

## join coastline length to total mg estimates and divide
total_mun_nitrogen_standardized <- nitrogen_inputs %>%
  mutate(all_sources_nitrogen = wastewater_mean + agriculture_10_percent + livestock_40_percent + shrimp_100_percent) %>% 
  mutate(all_sources_nitrogen = case_when(
    municipality == "mazatlan" ~ all_sources_nitrogen + concordia_nitrogen,
    TRUE ~ all_sources_nitrogen
  )) %>%
  filter(municipality != "concordia") %>% 
  left_join(coastline_length_by_mun %>% dplyr::select(municipality_lower, coastline_length_km), by = c("municipality" = "municipality_lower")) %>%
  mutate(ton_per_km_coastline = all_sources_nitrogen / coastline_length_km) %>% 
  arrange(ton_per_km_coastline)

summary(total_mun_nitrogen_standardized)
kable(total_mun_nitrogen_standardized)
```

```{r}
## save to mazu
write_csv(total_mun_nitrogen_standardized, here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/eutrophication/int","ton_per_km_coastline_municipality.csv"))
```


### 3. Rescale using the maximum value


```{r}
total_mun_nitrogen_standardized <- read_csv(here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/eutrophication/int","ton_per_km_coastline_municipality.csv"))

total_mun_nitrogen_rescaled <- total_mun_nitrogen_standardized %>%
  dplyr::select(rgn_id, municipality, all_sources_nitrogen, coastline_length_km, ton_per_km_coastline) %>% 
  mutate(rescaled = ton_per_km_coastline / max(ton_per_km_coastline, na.rm = TRUE))

summary(total_mun_nitrogen_rescaled)
kable(total_mun_nitrogen_rescaled)
```


### 4. Regional weighted average

Take the average of all values within each `rgn_id`, weighting by the length of the municipal coastline in km.

```{r}
current_status <- total_mun_nitrogen_rescaled %>%
  group_by(rgn_id) %>%
  summarize(
    pressure = weighted.mean(rescaled, w = coastline_length_km, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  complete(rgn_id = c("Region_2", "Region_9")) %>% 
  mutate(pressure = ifelse(rgn_id %in% c("Region_2"), 0, pressure)) %>%  # no modeled nitrogen inputs...
  mutate(status = 1 - pressure,
         current_status = status * 100) 

summary(current_status)
kable(current_status)
```

#### 4a. Gapfilling

We do not have information on Region_9, but it is not the same as Region_2 (which was evaluted by JARG's paper yet did not have values within the supplemental table).  

Therefore, we will use Region_8s values to very roughly gapfill, so that the status score is not NA, which would be inaccurate.

```{r}
current_status_gapfill <- current_status %>% 
  mutate(current_status = ifelse(rgn_id %in% c("Region_9"), 90.02575, current_status), # use Region_8's value to roughly gapfill Region_9.
         gapfilled = ifelse(rgn_id %in% c("Region_9"), "y", "n"))
```

```{r}
library(ggplot2)

current_status_gapfill <- data.frame(
  rgn_id = c("Region_1","Region_2","Region_3","Region_4","Region_5",
             "Region_6","Region_7","Region_8","Region_9"),
  current_status = c(28.1, 100, 99.9, 99.1, 97.8, 93.6, 91.4, 90.0, 90.0),
  gapfilled = c("n","n","n","n","n","n","n","n","y")
)

rgn_colors <- c(
  Region_1 = "#C03729", Region_2 = "#E68C7C", Region_3 = "#FC8F24",
  Region_4 = "#dcaa38", Region_5 = "#cbd2a0", Region_6 = "#98a25a",
  Region_7 = "#747428", Region_8 = "#627391", Region_9 = "#8d61a9"
)

ggplot(current_status_gapfill, aes(x = rgn_id, y = current_status, fill = rgn_id)) +
  geom_col() +
  scale_fill_manual(values = rgn_colors) +
  scale_y_continuous(limits = c(0, 100), expand = c(0, 0)) +
  labs(x = NULL, y = "Current status", title = "Eutrophication current status by region") +
  guides(fill = "none") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```



## Save to Mazu

```{r}
write_csv(current_status_gapfill, here::here("/home/shares/ohi/OHI_GOC/goal_prep/cw/v2025/eutrophication/output","eutrophication_cs.csv"))
```






















