---
title: 'OHI GoC 2025: Clean Waters: HAEDAT (HABs) Mexico Specific'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: false
    theme: cerulean
    highlight-style: haddock
  pdf:
    toc: true
editor: 
  markdown:
    wrap: sentence
  chunk_output_type: inline
editor_options: 
  chunk_output_type: inline
---


# Mexico Specific Harmful Algal Event Database (HAEDAT) Exploration

**Using this site instead: <https://haedat.iode.org/>**

What is the Harmful Algal Event Database?
The Harmful Algal Event Database (HAEDAT) is a component of the Harmful Algal Information system (HAIS) within the "International Oceanographic Data and Information Exchange" (IODE) of the "Intergovernmental Oceanographic Commission" (IOC) of UNESCO, and in cooperation with ICES, and PICES.

The HAEDAT is a meta database containing records of harmful algal events. HAEDAT contains records from the ICES area (North Atlantic) since 1985, and from the PICES area (North Pacific) since 2000. IOC Regional networks in South America, South Pacific and Asia, and North Africa are preparing to contribute. Guidance on submission of data and questions re HAEDAT can be found here.

HAEDAT Disclaimer: The HAEDAT database contains information based on yearly national reports by ICES and PICES member states. The available information on individual events varies greatly from event to event or country to country. Monitoring intensity, number of monitoring stations, number of samplings, stations, etc. also varies greatly and therefore there is not a direct proportionality between recorded events and actual occurrences of e.g. toxicity in a given region. Furthermore, areas with numerous recorded occurrences of HAE's, but with an efficient monitoring and management programmes, may have very few problems and a low risk of intoxications, whereas rare HAE's in other areas may cause severe problems and represent significant health risks.

To download:

1. "Browse Events" tab at the top menu
2. Click on "MEXICO"
3. Click on "[Download these events as a CSV file]" at the top right.
4. Open in Excel and resave as an .xlsx to fix delimiter issue.

## Exploration

```{r}
library(tidyverse)
library(here)
library(janitor)
library(sf)
library(mapview)
library(readxl)

# ISSUE WITH DELIMITERS: resave as xlsx!!!
# mex_hab_occurrence_raw <- read_csv(here::here("/home/shares/ohi/OHI_GOC/_raw_data/HAEDAT_mex/d2025","haedat_search.csv"))

# resaved as .xlsx
mex_hab_occurrence_raw <- read_excel(here::here("/home/shares/ohi/OHI_GOC/_raw_data/HAEDAT_mex/d2025","haedat_search.xlsx"), sheet = 1, col_names = TRUE)

```



```{r}
# clean by removing NA rows!
mex_hab_occurrence <- mex_hab_occurrence_raw %>% 
  janitor::clean_names() %>% 
  filter(!is.na(event_name)) %>% 
  mutate(event_year = as.character(event_year)) 

unique(mex_hab_occurrence$region)

# [1] "Sinaloa"                                    "Quintana Roo"                               "Baja California Sur"                       
#  [4] "Golfo de California"                        "MulegÈ, Baja California Sur"                "Navolato, Sinaloa"                         
#  [7] "Colima"                                     "El Caribe"                                  "Chiapas"                                   
# [10] "SInaloa"                                    "Baja California"                            "Oaxaca"                                    
# [13] "Jalisco"                                    "Sonora"                                     "Tamaulipas, north of Gulf of Mexico"       
# [16] "Guerrero"                                   "Yucatan"                                    "Yucat·n"                                   
# [19] "Nayarit"                                    "Campeche"                                   "Peninsula de Yucat·n and Mexican Caribbean"
# [22] "Sinalo"                                     "Todos Santos Bay"                           NA                                          
# [25] "Tabasco"                                    "Tamaulipas"                                 "Tehuantepec"                               
# [28] "Veracruz"                                   "Los Mochis, Sinaloa"                        "Michoac·n"                                 
# [31] "Bja California Sur"                         "La Paz, B. C. S."                           "Todos Santos Bay, B. C. N." 

mex_hab_goc <- mex_hab_occurrence %>% 
  filter(!region %in% c("Colima","Guerrero","Tabasco","Veracruz","Quintana Roo","El Caribe","Yucatan","Campeche","Todos Santos Bay","Tamaulipas","Chiapas","Oaxaca","Tamaulipas, north of Gulf of Mexico","Yucat·n","Peninsula de Yucat·n and Mexican Caribbean","Tehuantepec","Michoac·n","Todos Santos Bay, B. C. N."))

unique(mex_hab_goc$region)
#  [1] "Sinaloa"                     "Baja California Sur"         "Golfo de California"         "MulegÈ, Baja California Sur"
#  [5] "Navolato, Sinaloa"           "SInaloa"                     "Baja California"             "Jalisco"                    
#  [9] "Sonora"                      "Nayarit"                     "Sinalo"                      NA                           
# [13] "Los Mochis, Sinaloa"         "Bja California Sur"          "La Paz, B. C. S."
```

Let's take a look at the lats/longs that are already provided in the dataset:
```{r}
# mex_hab_goc_sf <- mex_hab_goc %>% 
#   filter(!is.na(latitude)) %>% 
#   st_as_sf(coords = c("longitude", "latitude"),
#            crs = 4326, # WGS84
#            remove = FALSE)
# mapview(mex_hab_goc_sf)
```

Unfortunately, some of these look accurate, and the others look very inaccurate. For example, MX-12-022 should be in Puerto Peñasco, but it is not. 


```{r}
unique(mex_hab_goc$location_text) ### soooo many issues. Using chatgpt to help me write the ase_when because it is so much...

location_text_fix <- mex_hab_goc %>%
  mutate(location_clean = case_when(
    
    # --- Mazatlán area ---
    location_text %in% c("Bahia Mazatl·n", "BahÌa de Mazatl·n", "Mazatlan", "Mazatlan Bay and Urias Estuary") ~ "Bahía de Mazatlán",
    
    # --- Altata area ---
    location_text %in% c("BahÌa Altata Ensenada de Pabellones", "Bay Altata Norte", 
                         "BahÌa Altata Norte", "Altata Norte", "BahÌa de Altata") ~ "Bahía de Altata", # in sinaloa.  The "BahÌa Altata Ensenada de Pabellones" has additional notes saying it is in "Sistema Lagunar Altata-Ensenada de Pabellones, Navolato, Sinaloa", great!
    
    # --- La Paz area ---
    location_text %in% c("Bahia de La Paz", "Bahia de la Paz", "BahÌa de la Paz", 
                         "La Ensenada de la Paz Shore", "Caleta Balandra", 
                         "Pichilingue", "Playa Coromuel") ~ "Bahía de La Paz", # it says that the Pichilingue observation is in BCS, not sinaloa.  All in la paz!! Did a double check.
    
    # --- Bahía Concepción ---
    location_text %in% c("Bahia ConcepciÛn", "BahÌa ConcepciÛn") ~ "Bahía Concepción",
    
    # --- Bahía de Los Ángeles ---
    location_text %in% c("BahÌa de Los ¡ngeles") ~ "Bahía de Los Ángeles",
    
    ###### STOPPED HERE, REVIEW THE REST!!!! ##########
    
    # --- Alto Golfo de California / San Felipe cluster ---
    location_text %in% c("Alto Golfo de California", 
                         "San Felipe", 
                         "San Felipe Puertecitos, Puerto de San Felipe y Alto Golfo de California",
                         "BahÌa del Colorado") ~ "Alto Golfo de California",
    
    # sinaloa, by topolobampo
    location_text %in% c("BahÌa del Colorado") ~ "Lechuguilla", # 25.743302, -109.345434
    
    # --- Bahía de Ballenas / Punta Abreojos cluster ---
    location_text %in% c("BahÌa de Ballenas", "RincÛn de las Ballenas", "Punta Abreojos") ~ "Bahía de Ballenas",
    
    # --- Bahía de Tortugas ---
    location_text %in% c("BahÌa de Tortugas") ~ "Bahía de Tortugas",
    
    # --- Bahía de Todos Santos / Ensenada area ---
    location_text %in% c("BahÌa de Todos Santos", "BAHIA DE TODOS SANTOS") ~ "Bahía de Todos Santos",
    
    # --- Bahía Kino ---
    location_text %in% c("Bahia Kino") ~ "Bahía Kino",
    
    # --- Topolobampo ---
    location_text %in% c("Topolobampo lagoon system") ~ "Bahía de Topolobampo",
    
    # --- Puerto Vallarta / Bahía de Banderas cluster ---
    location_text %in% c("Puerto Vallarta", "Nuevo Vallarta", "BahÌa de Banderas", 
                         "BahÌa la Manzanilla", "La Cruz de Huanacaxtle", "Barra de Navidad") ~ "Bahía de Banderas",
    
    # --- Cabo San Lucas area ---
    location_text %in% c("Cabo San Lucas") ~ "Cabo San Lucas",
    
    # --- Guerrero Negro ---
    location_text %in% c("Guerrero Negro") ~ "Guerrero Negro",
    
    # --- Mulegé ---
    location_text %in% c("MulegÈ") ~ "Mulegé",
    
    # --- Navolato / La Reforma (Sinaloa coast) ---
    location_text %in% c("Navolato", "La Reforma") ~ "Costa de Navolato",
    
    # --- Miscellaneous valid Gulf islands ---
    location_text %in% c("Isla Coronado", "Isla San JerÛnimo", "Isla EspÌritu Santo", 
                         "San Jose Island", "Isabel Island") ~ "Islas del Golfo de California",
    
    # --- Non-Gulf (to filter out later) ---
    location_text %in% c("Mexican Caribbean Coast", "Lake Catemaco", "Big Lagoonˇand Eel River",
                         "Alameda Oriente Lake", "Chapultepec Lake",
                         "Olympic Rowing and Canoeing Track ëëVirgilio Uribeíí",
                         "Bahia de Acapulco", "BahÌas San Luis, Gonzaga, Guadalupe, Los ƒngeles, de las ¡nimas, San Rafael") ~ NA_character_,
    
    # --- Default: keep as-is if not matched ---
    TRUE ~ location_text
  ))

```



```{r}
mex_hab_goc_clean <- mex_hab_goc %>% 
  mutate(region = case_when( # fix all that need fixing
    
     # BCS
    region %in% c("Baja California Sur", 
                  "Bja California Sur", 
                  "MulegÈ, Baja California Sur", 
                  "La Paz, B. C. S.") ~ "Baja California Sur",

    # Sinaloa
    region %in% c("Sinaloa", 
                  "SInaloa", 
                  "Sinalo", 
                  "Navolato, Sinaloa", 
                  "Los Mochis, Sinaloa") ~ "Sinaloa",
    TRUE ~ region
  )) %>% 
  tidygeocoder::geocode(city = location_text, country = country_name, state = region)  # to give each city a lat, long, if possible 
  
mex_hab_goc_cleaner <- mex_hab_goc_clean %>% 
  mutate(longitude = ifelse(is.na(longitude), long, longitude)) %>% 
  mutate(latitude = ifelse(is.na(latitude), lat, latitude)) %>% 
  tidygeocoder::geocode(country = country_name, state = region)

mex_hab_goc_cleanest <- mex_hab_goc_cleaner %>% 
  janitor::clean_names() %>% 
  mutate(longitude = ifelse(is.na(longitude), long_88, longitude)) %>% 
  mutate(latitude = ifelse(is.na(latitude), lat_87, latitude)) %>% 
  mutate(latitude = ifelse(location_text == "Bahía de la Paz", 24.312716332893793, latitude)) %>%  # 24.312716332893793, -110.32933847561796
  mutate(latitude = ifelse(location_text == "Bahia de La Paz", 24.312716332893793, latitude)) %>% 
  mutate(longitude = ifelse(location_text == "Bahia de La Paz", -110.32933847561796, longitude)) %>% 
  filter(!location_text %in% c("Bahia de Acapulco", "BAHIA DE TODOS SANTOS")) 
  
# mex_hab_goc_cleanest_na <- mex_hab_goc_cleanest %>% 
  # filter(is.na(latitude))

mex_hab_goc_cleanest_sf <- mex_hab_goc_cleanest %>% 
  st_as_sf(coords = c("longitude", "latitude"),
           crs = 4326, # WGS84
           remove = FALSE)

mapview(mex_hab_goc_cleanest_sf, zcol = "event_year")
```


```{r}
summary(mex_hab_goc_clean)

library(gtsummary)

mex_hab_summary_tbl <- mex_hab_goc_clean %>% 
  tbl_summary(include = c(event_year, region, occurred_before, humans_affected, fish_affected, causative_species_name0))

print(mex_hab_summary_tbl)
```








