################################################################################ 

For questions contact: Erica Ferrer, Ph.D. [emferrer\@ucsd.edu](mailto:emferrer@ucsd.edu){.email} - OR - [ferrer\@nceas.ucsb.edu](mailto:ferrer@nceas.ucsb.edu){.email}

################################################################################ 

# PRE-WORK ------------------------------

```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```

# STEP 1: Load in CONAPESCA RAW data, records from all of Mexico ------------------------------

This step is completed in the file shared by Juan Carlos: "03_combine_landings_edited.R" file path: /home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/JC_Scripts_EMF.edited/03_combine_landings_edited.R

# STEP 2: Work with CONAPESCA STITCHED data, records from all of Mexico ------------------------------

A stitched version of these data is available as one single file: "AllMex_landings_2000_present_EMF_edited.rds" file path: /home/shares/ohi/OHI_GOC/goal_prep/fp/juan_carlos/AllMex_landings_2000_present_EMF_edited.rds

```{r Load in CONAPESCA STITCHED data}
All.Mex_Landings_2000.present<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/juan_carlos/AllMex_landings_2000_present_EMF_edited.rds"))
# n = 12153879 obs., 20 vars.
```

# STEP 3: Create weight_kg_corrected column ------------------------------

```{r Landed weight correction}
# Kind of looks like sometimes 'landed weight' is 'live weight' and vice versa. Fix this:
All.Mex_Landings_2000.present$weight_kg_corrected <- ifelse(
  All.Mex_Landings_2000.present$live_weight >= All.Mex_Landings_2000.present$landed_weight,
  All.Mex_Landings_2000.present$live_weight,
  All.Mex_Landings_2000.present$landed_weight
)
# n = 12153879 obs., 21 vars.
```

# STEP 4a: Filter down CONAPESCA data to states in Northwest Mexico ------------------------------

```{r Filter out data from Northwest Mexico}
length(unique(All.Mex_Landings_2000.present$species_name)) 
# 2052 unique species in all of Mexico.
# But, we don't need species from ALL of Mexico, only the GoC region for now - so let's filter by state.

States_of_Interest<- c("BAJA CALIFORNIA", "BAJA CALIFORNIA SUR", "SONORA", "SINALOA", "NAYARIT", "JALISCO")

NW.Mex_Landings <- All.Mex_Landings_2000.present %>% filter(state %in% States_of_Interest)
# n = 4713144 obs., 21 vars.

length(unique(NW.Mex_Landings$species_name)) 
# 1827 unique species in States of Interest.
```

# STEP 4b: Attach lat-long and OHI_region info based on official landing site ------------------------------
NOTE TO SELF: Come back to this! The lat long keys are imperfect. We'll want to fix this as much as possible.
```{r Append lat-long data to sites in Baja BCS Sonora and Sinaloa}
# Filter OUT sites from Baja that are on the Pacific side using lat-Long file from Mel's repository:
# NOTE TO SELF: THIS PART IS SUBJECT TO CHANGE once we integrate lat-long data from Hem. Mar 10 2026
BajaSonora_LatLong_Key<- read_csv("/home/frazier/ohi_goc_prep/FP/FIS/v2025/data/Baja_Sonora_states_key.csv")
colnames(BajaSonora_LatLong_Key) <- c("state", "landing_site", "OHI_rgn", "lat", "long", "GoC")

Jalisco_LatLong_Key<- read_csv(here("/home/frazier/ohi_goc_prep/FP/FIS/v2025/data/landing_sites_jalisco.csv"))
colnames(Jalisco_LatLong_Key) <- c("landing_site", "CONAPESCA_landing_key", "est_habitat", "lat", "long", "site_notes", "OHI_rgn", "state")


#### CLEAN LAT-LONG KEYS: ####
# Clean the Baja-Sonora key...
BajaSonora_LatLong_Key_clean <- BajaSonora_LatLong_Key %>%
  
  # Normalize state names:
  mutate(state = toupper(state)) %>%
  
  # Remove any rows where landing_site is NA (NAs cause matching issues later on):
  filter(!is.na(landing_site)) %>%
  
  # Remove any direct duplicates of landing_site:
  group_by(state, landing_site) %>%
  slice(1) %>%  # keeps the first occurrence per key
  ungroup()

# Look into the rows that were removed in the de duplication of BajaSonora_LatLong_Key...
removed_rows <- anti_join(BajaSonora_LatLong_Key, BajaSonora_LatLong_Key_clean, 
                          by = c("landing_site"))

# Convert lat-long entries to numeric entries, accounting for unique cases where the minus (-) is actually an n dash or m dash:
BajaSonora_LatLong_Key_clean <- BajaSonora_LatLong_Key_clean %>%
  mutate(
    long = as.numeric(str_replace_all(long, "[−–—]", "-")),
    lat  = as.numeric(str_replace_all(lat, "[−–—]", "-"))
  )

# Clean the Jalisco key...
Jalisco_LatLong_Key_clean <- Jalisco_LatLong_Key %>%
  # Normalize state names
  mutate(state = toupper(state)) %>%
  
  # Remove any rows where landing_site is NA (NAs cause matching issues later on):
  filter(!is.na(landing_site)) %>%
  
  # Keep only one row per CONAPESCA_landing_key; if duplicate keys exist, keep first
  group_by(CONAPESCA_landing_key) %>%
  slice(1) %>%  # keeps the first occurrence per key
  ungroup() %>%
  
  # Optional: remove any direct duplicates of landing_site if there are rows with NA in the CONAPESCA_landing_key
  group_by(state, landing_site) %>%
  slice(1) %>%  # keeps the first occurrence per key
  ungroup()

# Look into the rows that were removed in the de duplication of Jalisco_LatLong_Key...
removed_rows <- anti_join(Jalisco_LatLong_Key, Jalisco_LatLong_Key_clean, 
                          by = c("CONAPESCA_landing_key", "landing_site"))

# Convert lat-long entries to numeric entries, accounting for unique cases where the minus (-) is actually an n dash or m dash:
Jalisco_LatLong_Key_clean <- Jalisco_LatLong_Key_clean %>%
  mutate(
    long = as.numeric(str_replace_all(long, "[−–—]", "-")),
    lat  = as.numeric(str_replace_all(lat, "[−–—]", "-"))
  )

# Add the "GoC" helpful column here based on a latitude cut off for the Nayarit/Jalisco OHI region, region 9 starting at 20.40895:
Jalisco_LatLong_Key_clean <- Jalisco_LatLong_Key_clean %>%
  mutate(GoC = if_else(lat < 20.40895, "no", NA_character_))

#### MERGE LAT-LONG KEYS: ####
LatLong_MasterKey_pending <- bind_rows(BajaSonora_LatLong_Key_clean, Jalisco_LatLong_Key_clean) %>%
  filter(!is.na(landing_site)) %>%  # remove the NA from this data frame as this will create a many-to-many matching issue later on
  mutate(long = as.numeric(long))
# 1176 obs., 9 vars.

# Remove unused data frames:
rm(BajaSonora_LatLong_Key, Jalisco_LatLong_Key, removed_rows)

# Take a look
unique(LatLong_MasterKey_pending$state)
unique(LatLong_MasterKey_pending$OHI_rgn)

# My regional key (temporary, subject to change, UNDER CONSTRUCTION):
RegionKey_pending<- read_csv(here("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/RegionKey_UNDERCONSTRUCTION_Mar102026.csv"))
# Previously: RegionKey_UNDERCONSTRUCTION_Dec012025.csv
colnames(RegionKey_pending) <- c("state", "OHI_rgn", "OHI_rgn_number", "OHI_rgn_name", "exclude_rgn", "exclude_notes")
```

```{r Append lat-long and OHI_region info to landings data}
# Append lat-long info to landings data:
temp_df <- NW.Mex_Landings %>%
  left_join(
    LatLong_MasterKey_pending,
    by = c("state", "landing_site")  # matching columns
  )

table(temp_df$GoC)
# 4713144 obs., 28 vars.
# 1090368 "no" not in the GOC.

# Append RegionKey_pending info to landings data:
temp_df <- temp_df %>%
  left_join(
    RegionKey_pending,
    by = c("state", "OHI_rgn")  # matching columns
  )
# 4713144 obs., 32 vars.
table(temp_df$GoC)
# 1090368 "no" not in the GOC as we'd expect!

NW.Mex_Landings<- temp_df
# 4713144 obs., 32 vars.

# Delete temp_df, not needed anymore.
rm(temp_df)
```

# STEP 4c: Save the NW Mexico landings appended with lat-long as a separate .rds file
```{r Save the Northwest Mexico dataframe: Mar 11 2026}
saveRDS(GOC_Landings, file="/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/NWMex_Landings_2000.present_EMF.Mar112026.rds")
# 4713144 obs., 32 vars.
```

# STEP 5a: Filter down data to sites in the GOC, exclude all sites that definitely AREN'T in the Gulf ------------------------------
Additional research will be required for some landing sites. Details pending additional lat-long key work. See notes here (as of Mar 11 2026): https://docs.google.com/document/d/1ANHEdRi8aWC0qLNaM1_c_nLtn5GihTFgHDcU96bry5Y/edit?tab=t.0

```{r filter down data to only the GOC}
# Generate filtered landings data for the GoC:
GOC_Landings <- NW.Mex_Landings %>%
  filter(GoC != "no" | is.na(GoC))
# 3622776 obs., 32 vars.

# Check that this all makes sense:
length(unique(GOC_Landings$species_name))
length(unique(GOC_Landings$main_species_group))
# 1703 unique species from 57 main species groups in the Gulf of California, keeping in mind that these are not really SPECIES (per se) but rather log-book entries for species.
```
NOW 1703 (previously 1693) unique species in the Gulf of California according to Juan Carlos' version of the CONAPESCA data base, keeping in mind that these are not really SPECIES, per se, but rather log-book entries for species.

# STEP 5b: Save the GOC landings dataframe as a separate .rds file
```{r Save the GOC_Landings dataframe: Mar 11 2026}
saveRDS(GOC_Landings, file="/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Landings_2000.present_EMF.Mar112026.rds")
# 3622776 obs., 32 vars.
# Previous iteration: GOC_FIS.landings_CLEAN_as.of.Dec092025.rds
```


# STEP 6a: Separate out key words from the official record for common name ------------------------------

# IF SKIPPING STEPS 1-5: load in GOC_Landings dataframe:
```{r Load GOC_Landings if needed}
GOC_Landings<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Landings_2000.present_EMF.Mar112026.rds"))
# 3622776 obs., 32 vars.
```

Keeping in mind that some key terms reflect animal parts but are actually part of the species common name like "aleta de" versus "aleta blanca"

** QUICK SANITY CHECK: Is "aleta de aleta" an issue? Short answer is no, double "aleta" does not seem to be an issue.
GOC_Landings %>%
  rename(recorded_taxa_name = species_name) %>%
  mutate(
    aleta_count = stringi::stri_count_fixed(recorded_taxa_name, "ALETA")
  ) %>%
  count(aleta_count)

```{r Filter out keywords from recorded_taxa_name to get recorded_taxa_name_clean}
# Part A. Keyword list, excluding "ALETA" which is handled separately -----------------

# NOTE TO SELF: How did I get this list of key words originally? I think I started out with some preliminary version of the separate, parse functionality, quickly spotted the beginnings of a keywords list and then grew the list over time. Come back to this at some point to recreate this early part of the workflow to document it (reproducibility).

Keyword_list <- c(
  "S.C.", "S/C", "S",
  "ENTERO", "ENTERA", "IND", "ENT",
  "FRESCO", "FRESCA", "FCA", "FCO", "FC",
  "ECO",
  "CONG", "CONGELADO",
  "VIVO", "VIVA", "VIVAS", "VIVITA",
  "DE CULTIVO", "DE CULTIVO VIVO", "ABASTO RANCHO", "RANCHO",
  "HUEVA DE",
  "FILETE", "FILETE DE", "ECA", "FILETE DE ECO", "DESHUESADA", "CARNE", "CARNE DE",
  "DESCABEZADO", "DES", "DSC", "DESC", "DESV", "DESVICERADO",
  "CABEZA DE", "BUCHE", "BUCHE DE", "TRIPA DE", "COLAS DE", "MANOS DE", "MANOS",
  "ANCAS DE", "ANCAS", "ANCA",
  "PIEL DE ECA", "PIEL DE SECA", "PIEL DE", "PIEL",
  "SECO DE", "SECA", "SECA DE", "SECO", "SECO DE",
  "COCINERO", "COCIDO", "COCIDA", "COCIDAS", "ASADO",
  "DE ALTA MAR",
  "CONCHA DE", "CONCHA",
  "POLVO DE",
  "CARNADA", "P/CARNADA",
  "DE AGUA DULCE",
  "MANTO", "MANTO DE", "PULPA DE", "PULPA FRESCA",
  "PULPO DE", "TENTACULOS DE", "TUBO DE", "TUBO",
  "NO APTO PARA EMPAQUE"
  # Remove ALETA from main keywords; we'll handle this separately
  #"ALETA", "ALETA ECA DE", "ALETA FCA DE"
)

# Order the key words longest first (note that this is necessary for the next steps):
Keyword_list <- Keyword_list[order(nchar(Keyword_list), decreasing = TRUE)]

# Escape regex meta-characters:
Keyword_list_escaped <- stringi::stri_replace_all_regex(
  Keyword_list, "([.|(){}+*?^$\\[\\]\\\\])", "\\\\$1"
)

# Final regex:
Keyword_pattern <- paste0(
  "(?<![A-Za-z0-9])(",
  paste(Keyword_list_escaped, collapse = "|"),
  ")(?![A-Za-z0-9])"
)


# Part B. Take care of special ALETA phrases (handled separately from other key words) -----------------
special_aleta_cases <- c("BLANCA", "AMARILLA", "CORTA", "AZUL")

remove_aleta_except_special <- function(x) {
  stringi::stri_replace_all_regex(
    x,
    pattern = paste0("\\bALETA\\b(?!\\s+(?:", paste(special_aleta_cases, collapse = "|"), "))"),
    replacement = "",
    vectorize_all = FALSE
  )
}


# Part C. MAIN PARSING PIPELINE (RUN TIME = ~5 min, optimized but still computationally heavy) -----------------

GOC_Landings_parsed <- GOC_Landings %>%
  
  # Rename columns
  rename(
    recorded_taxa_name = species_name,
    recorded_taxa_group = main_species_group
  ) %>%
  
  # Extract keywords
  mutate(
    keywords_list = stringi::stri_extract_all_regex(recorded_taxa_name, Keyword_pattern)
  ) %>%
  mutate(keywords_list = map(keywords_list, ~ { length(.x) <- 5; .x })) %>%
  unnest_wider(keywords_list, names_sep = "_") %>%
  rename_with(~ paste0("keywords_", 1:5), starts_with("keywords_list_")) %>%
  
  # Clean recorded_taxa_name
  mutate(
    recorded_taxa_name_clean = recorded_taxa_name %>%
      remove_aleta_except_special() %>%                      # Remove ALETA unless special
      stringi::stri_replace_all_regex(Keyword_pattern, "") %>% # Remove other keywords
      stringi::stri_replace_all_regex("\\b(Y|DE|P/)\\b", "") %>% # Remove standalone Y/DE/P/
      stringi::stri_replace_all_regex("[.,/\\\\]+", " ") %>%      # Remove punctuation
      str_squish()
  ) %>%
  
  # Extract scientific name
  mutate(
    recorded_scientific_name = stringi::stri_extract_first_regex(recorded_taxa_name_clean, "\\(([^)]+)\\)") %>%
      stringi::stri_replace_all_regex("\\(|\\)", ""),
    scientific_name_recorded_T.F = !is.na(recorded_scientific_name),
    recorded_scientific_name = if_else(
      scientific_name_recorded_T.F,
      stringi::stri_trans_totitle(recorded_scientific_name),
      NA_character_
    )
  ) %>%
  
  # Extract common name
  mutate(
    common_name_clean = stringi::stri_trim_both(
      stringi::stri_replace_first_regex(recorded_taxa_name_clean, "\\s*\\([^)]*\\)", "")
    ),
    common_name_clean = tolower(common_name_clean)
  )
# 3622776 obs., 41 vars.


# Part D. Trim and reorder columns -----------------
GOC_Landings_clean <- GOC_Landings_parsed %>%

  select(
    # General info
    state, office_name, landing_site, landing_site_key, GoC, lat, long, est_habitat,
    year, month,
    receipt_date_fixed, period_start_fixed, period_end_fixed,
    fleet, acuaculture_production, eu_rnpa, eu_name, vessel_rnpa, vessel_name,
    
    # Species info
    recorded_taxa_group, recorded_taxa_name,
    
    # Cleaned / derived names
    recorded_taxa_name_clean, recorded_scientific_name,
    scientific_name_recorded_T.F, common_name_clean,
    
    # Keywords
    keywords_1, keywords_2, keywords_3, keywords_4, keywords_5,
    
    # Landing details
    value, landed_weight, live_weight, weight_kg_corrected
 
  )
# 3622776 obs., 34 vars.

# How many common names are there in this dataset?
length(unique(GOC_Landings_clean$common_name_clean))
length(unique(GOC_Landings_clean$recorded_taxa_group))
# NOW 737 common names (previously 734, and before that, 973 with the less comprehensive keywords_list), over 57 taxonomic groups
```

** QUICK SANITY CHECK: did this work for taxa like yellow-fin tuna? Yes, it seems to have worked!
test <- GOC_Landings_clean %>%
  filter(str_detect(recorded_taxa_name, "ATUN ALETA AMARILLA"))


# STEP 6b: OPTIONAL Save the GOC_Landings_parsed dataframe as a separate .rds file
```{r Save this dataframe - landings keywords & species for top groupings: Mar 11 2026}
# saveRDS(GOC_Landings_clean, file="/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Landings_KeywordsParsed_2000.present_EMF.Mar122026.rds")
```

```{r Investigating Keywords}
#### VIEW: ALL KEY WORDS ####
# Extract all unique keywords from keyword columns 1 through 5. Delete exact duplicates. 
Keyword_vector <- GOC_Landings_clean %>%
  select(keywords_1:keywords_5) %>%
  unlist() %>%
  unique() %>%
  na.omit()
```
NOTE TO SELF: How did I get this list of key words originally? I think I started out with some preliminary version of the separate, parse functionality, quickly spotted the beginnings of a keywords list and then grew the list over time. Come back to this at some point to recreate this early part of the workflow to document it (reproducibility).

# STEP 7: Merge the cleaned master sp key with landings data ------------------------------

# IF SKIPPING STEPS 1-6: load in GOC_Landings_cleaned (keyword parsed) dataframe:
```{r Load GOC_Landings_clean if needed}
# GOC_Landings_clean<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Landings_KeywordsParsed_2000.present_EMF.Mar112026.rds"))
# 3622776 obs., 32 vars.
```

```{r Load MasterSpKey_pending}
MasterSpKey_pending <- read_csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/Species_Keys/CleanedMaster_SpKey_Dec012025.csv")
# 802 obs of 41 variables

GenusGroupKey_pending <- read_csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/Genus_to_MajorGroup_Key_Feb242026.csv")
unique(GenusGroupKey_pending$Major_Functional_Group)
#  [1] "Shrimps, prawns"            "Pelagic marine fishes"      "Demersal marine fishes"     "Sharks"                    
#  [5] "Crustaceans"                "Cephalopods"                NA                           "Molluscs excl. Cephalopods"
#  [9] "Freshwater fishes"          "Marine fishes NEI"          "Other marine invertebrates" "Diadromous fishes"         
# [13] "Marine algae"               "Other freshwater taxa"    
```

```{r Shave down the size of Master_SpKey_pending}
Master_SpKey_simple<- MasterSpKey_pending %>% 
  select(recorded_taxa_group, recorded_taxa_name_clean, common_name_clean,
         estimated_species_name, species_est_certainty, estimated_genus_name, estimated_family_name,
         IHH_order, IHH_class, IHH_functional_major_groups_sm, IHH_fishery)
# 802 obs., 11 vars.
```

```{r Merge GOC landings with Master_SpKey_simple}
GOC_Landings_clean <- GOC_Landings_clean %>%
  left_join(
    Master_SpKey_simple,
    by = c(
      "recorded_taxa_group",
      "recorded_taxa_name_clean",
      "common_name_clean"
    )
  )
# n = 3622776 obs. as before, now 42 vars.
```

## Save GOC_Landings with species as separate .rds file

```{r Save this dataframe - landings keywords & species for top groupings: Mar 11 2026}
# saveRDS(GOC_Landings_clean, file="/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Landings_wSpecies_2000.present_EMF.Mar122026.rds")
```


# STEP 7a: FLAG quaculture entries based on a series of logical rules
I think aquaculture and mariculture in the region is being vastly under-counted and under-reported.
Let's implement a series of rules to estimate if records are associated with aquaculture or not.

```{r Create a GOC_Landings_aquaculture dataframe}
GOC_Landings_aquaculture<- GOC_Landings_clean
GOC_Landings_aquaculture$est_aquaculture_production<- c(NA)
# n = 3622776 obs., 43 vars.
```

First, let's flag landing site names that are associated with aquaculture.
```{r Estimate likelihood of aquaculture based on name of landing site}
#### FLAG 1 - Flag aquaculture based on landing site name (3 flag levels: YES, NO, POSSIBLY): ####
GOC_Landings_aquaculture <- GOC_Landings_aquaculture %>%
  
  # Normalize original report
  mutate(
    acuaculture_production_norm = str_to_lower(str_trim(acuaculture_production))
  ) %>%
  
  # Step 1: landing site flag with YES / POSSIBLY / NO
  mutate(
    aquacultureFLAG_landing.site = case_when(
      str_detect(landing_site, regex("\\(CULT\\.?\\)|CULTIVADORES|ESTANQUE|VIVEROS", ignore_case = TRUE)) ~ "YES",
      str_detect(landing_site, regex("DREN|DRENES|BORDO|PREDIO", ignore_case = TRUE)) ~ "POSSIBLY",
      TRUE ~ "NO"
    )
  ) %>%
  
  # Drop helper column:
  select(-acuaculture_production_norm)

table(GOC_Landings_aquaculture$aquacultureFLAG_landing.site)

# PURG! I think this logic will be superceded by the logic below but this was helpful for initial checks. - Mar 12 2026
# Generate estimates and confidence values based on the comparison of landing_site flag values versus the official report:
# GOC_Landings_aquaculture <- GOC_Landings_aquaculture %>%
#   
#   # Step 2: estimate aquaculture production
#   mutate(
#     est_aquaculture_production = case_when(
#       aquacultureFLAG_landing.site == "YES" & acuaculture_production_norm %in% c(NA, "no") ~ "yes, with high confidence",
#       aquacultureFLAG_landing.site == "POSSIBLY" & is.na(acuaculture_production_norm) ~ "probably, medium confidence",
#       acuaculture_production_norm == "yes" ~ "yes, as reported",
#       acuaculture_production_norm == "no" & aquacultureFLAG_landing.site == "NO" ~ "no, as reported",
#       TRUE ~ "uncertain"
#     ),
#     
#     est_aquaculture_certainty = case_when(
#       est_aquaculture_production == "uncertain" ~
#         "not enough information to assign aquaculture estimate",
#       
#       est_aquaculture_production == "yes, with high confidence" & is.na(acuaculture_production_norm) ~
#         "no official report but landing site name suggests aquaculture with high confidence",
#       
#       est_aquaculture_production == "probably, medium confidence" & is.na(acuaculture_production_norm) ~
#         "no official report but landing site name suggests aquaculture with medium confidence",
#       
#       est_aquaculture_production == "yes, as reported" & acuaculture_production_norm == "yes" ~
#         "estimate is consistent with original report",
#       
#       est_aquaculture_production == "no, as reported" & acuaculture_production_norm == "no" ~
#         "estimate is consistent with original report",
#       
#       est_aquaculture_production == "yes, with high confidence" & acuaculture_production_norm == "no" ~
#         "original report likely misreported; estimate differs from official report because landing site name suggests aquaculture with high confidence",
#       
#       est_aquaculture_production == "probably, medium confidence" & acuaculture_production_norm == "no" ~
#         "original report says 'no', but landing site name suggests aquaculture with medium confidence",
#       
#       TRUE ~ "not enough information to assign aquaculture estimate"
#     )
#   ) %>%
#   
#   # Drop helper normalization column
#   select(-acuaculture_production_norm)
# 
# # SANITY CHECK:
# table(GOC_Landings_aquaculture$acuaculture_production)
# # 17496 yes
# table(GOC_Landings_aquaculture$est_aquaculture_production)
# # 17496 yes, as reported; 2190 yes, with high confidence; 914 probably, medium confidence
# # For a total of 3104 additional records!
# 
# # Check that we pulled out landing sites that make sense:
# GOC_Landings_aquaculture %>%
#   filter(est_aquaculture_production %in% c(
#     "yes, with high confidence",
#     "probably, with medium confidence"
#   )) %>%
#   distinct(landing_site) %>%
#   arrange(landing_site)
# # 46 sites that all make sense to me! If an organism came from one of these sites I would assume it was a farmed species, even if the record was NOT marked as aquaculture.
# 
# # Let's review these sites manually:
# write.csv(
#   GOC_Landings_aquaculture %>%
#     filter(est_aquaculture_production != "uncertain") %>%
#     distinct(landing_site, acuaculture_production, est_aquaculture_production, est_aquaculture_certainty, aquacultureFLAG_landing.site),
#   "/home/ferrer/ohi_goc_prep/FP/Exploratory_Research/MAR_ExploratoryResearch/AquacultureSite_ClassificationCheck_Mar122026.csv",
#   row.names = FALSE
# )
# After a manual review, this logic seems okay for now... let's see how this affects of estimates of aquaculture.
```

Next, let's implement rules based on key words and taxa:
```{r Flagging aquaculture landings based on key words}
colnames(GOC_Landings_aquaculture)
table(GOC_Landings_aquaculture$recorded_taxa_group)

#### FLAG 2 - Flag aquaculture key words ####
Aquaculture_keywords<- c(
  "DE CULTIVO", "DE CULTIVO VIVO", "ABASTO RANCHO", "RANCHO"
)

# ## WORK IN PROGRESS add cultivadas from common_name_clean to keywords..
# Aquaculture_Summary "cultivadas"

# Flag rows with aquaculture keywords:
GOC_Landings_aquaculture <- GOC_Landings_aquaculture %>%
mutate(
  aquacultureFLAG_keywords = case_when(
    rowSums(across(keywords_1:keywords_5, ~ str_to_upper(str_trim(.x)) %in% Aquaculture_keywords), na.rm = TRUE) > 0 ~ "YES",
    TRUE ~ "NO"
  )
)

table(GOC_Landings_aquaculture$aquacultureFLAG_keywords)
# 111947 observations explicitly flagged as aquaculture due to key words.
```

```{r Flag common aquaculture taxa like oysters and tilapia}
#### FLAG 3 - Flag known aquaculture taxa (LIST IS UNDER CONSTRUCTION) ####
# In theory, names could come from a number of columns: recorded_taxa_group, recorded_taxa_name, common_name_clean

Possible_AquacultureTaxa <- c("OSTION")
Definite_AquacultureTaxa <- c("TILAPIA")

GOC_Landings_aquaculture <- GOC_Landings_aquaculture %>%
  mutate(
    aquacultureFLAG_specific.taxa = case_when(
      str_to_upper(str_trim(common_name_clean)) %in% Definite_AquacultureTaxa ~ "YES",
      str_to_upper(str_trim(recorded_taxa_group)) %in% Possible_AquacultureTaxa ~ "POSSIBLY",
      TRUE ~ "NO"
    )
  )

table(GOC_Landings_aquaculture$aquacultureFLAG_specific.taxa)
#   NO          POSSIBLY    YES 
#   3408840     41066       172870 
```

Create a combined aquaculture flag based on the previous three flags: aquacultureFLAG_landing.site, aquacultureFLAG_keywords, aquacultureFLAG_specific.taxa

Here's the logic for the combined flag:
• If at least one flag is "YES", then flags indicate aquaculture is highly likely (my YES flags are pretty robust at this point)
• If one more flags is "POSSIBLY", then flags indicate the record is possibly aquaculture but not certain
• If combined flag is "NO", then flags do not indicate aquaculture

```{r Create a combined aquaculture flag}
#### FLAG 4 - Create a combined aquaculture flag based on the previous three flags ####
GOC_Landings_aquaculture <- GOC_Landings_aquaculture %>%
  mutate(
    aquacultureFLAG_combined = case_when(
      if_any(c(aquacultureFLAG_landing.site,
               aquacultureFLAG_keywords,
               aquacultureFLAG_specific.taxa), ~ .x == "YES") ~ "YES",
      
      if_any(c(aquacultureFLAG_landing.site,
               aquacultureFLAG_keywords,
               aquacultureFLAG_specific.taxa), ~ .x == "POSSIBLY") ~ "POSSIBLY",
      
      TRUE ~ "NO"
    )
  )

# SANITY CHECK:
subset<- GOC_Landings_aquaculture %>% filter(aquacultureFLAG_landing.site=="YES")
table(subset$aquacultureFLAG_combined)
# 2226 YES. Excellent. This worked, it matches the total length of the subset df as expected.
```

# STEP 7b: ESTIMATE quaculture entries based on a series of logical rules
Here is the logic and set of decisions designed to estimate aquaculture:
• If original report for acuaculture_production = "yes" & aquaculture_flags = "yes", then aquaculture_estimate = "aquaculture" & confidence = "high; flags agree with original report" 
• If original report for acuaculture_production = "no" & aquaculture_flags = "no", then aquaculture_estimate = "not aquaculture" & certainty = "high; flags agree with original report"

• If original report = "no" & aquaculture_flags = "yes", then aquaculture_estimate = "likely aquaculture" & certainty = "medium; flags suggest most likely aquaculture but original report says no" 
• If original report = "no" & aquaculture_flags = "possibly", then aquaculture_estimate = "possibly aquaculture" & certainty = "low; flags suggest possibly aquaculture but original report says no"

• If original report = NA & aquaculture_flags = "yes", then aquaculture_estimate = "likely aquaculture" & certainty = "high; no data from the original report but flags suggest most likely aquaculture"
• If original report = NA & aquaculture_flags = "possibly", then aquaculture_estimate = "possibly aquaculture" & certainty = "low; no data from the original report but flags suggest possibly aquaculture"
• If original report = NA & aquaculture_flags = "no", then aquaculture_estimate = "unlikely aquaculture" & certainty = "medium-high; no data from the original report but flags suggest probably not aquaculture"

Decision Table:
Report   aquacultureFLAG_combined  aquaculture_estimate    certainty
  yes       NO                        aquaculture	            high; no flags but original report says aquaculture
  yes       YES                       aquaculture             high; flags agree with original report
  no        NO                        not aquaculture         high; flags agree with original report
  no        YES	                      likely aquaculture	    medium; flags suggest most likely aquaculture but original report says no
  no        POSSIBLY                  possibly aquaculture    low; flags suggest possibly aquaculture but original report says no
  NA        YES                       likely aquaculture      medium-high; no data from original report but flags suggest most likely aquaculture
  NA        POSSIBLY                  possibly aquaculture    low; no data from original report but flags suggest possibly aquaculture
  NA        NO                        unlikely aquaculture    medium-high; no data from original report but flags suggest probably not aquaculture

```{r Generate estimates and confidence values}
#### Estimate aquaculture and confidence values based on the comparison of aquacultureFLAG_combined versus the official report: ####
GOC_Landings_aquaculture <- GOC_Landings_aquaculture %>%
  mutate(
    # Normalize original report for consistency (low case font):
    acuaculture_original.report = str_to_lower(str_trim(acuaculture_production)),
    
    # Final estimate based on original report and combined flags
    est_aquaculture_production = case_when(
      
      # Original report = yes (regardless of flags)
      acuaculture_original.report == "yes" ~ "aquaculture",
      
      # Original report = no & flags = NO
      acuaculture_original.report == "no" & aquacultureFLAG_combined == "NO" ~ "not aquaculture",
      
      # Original report = no & flags = YES
      acuaculture_original.report == "no" & aquacultureFLAG_combined == "YES" ~ "likely aquaculture",
      
      # Original report = no & flags = POSSIBLY
      acuaculture_original.report == "no" & aquacultureFLAG_combined == "POSSIBLY" ~ "possibly aquaculture",
      
      # Original report = NA & flags = YES
      is.na(acuaculture_original.report) & aquacultureFLAG_combined == "YES" ~ "likely aquaculture",
      
      # Original report = NA & flags = POSSIBLY
      is.na(acuaculture_original.report) & aquacultureFLAG_combined == "POSSIBLY" ~ "possibly aquaculture",
      
      # Original report = NA & flags = NO
      is.na(acuaculture_original.report) & aquacultureFLAG_combined == "NO" ~ "unlikely aquaculture",
      
      # Fallback for any other case
      TRUE ~ "uncertain"
    ),
    
    # Certainty messages corresponding to the estimate
    est_aquaculture_certainty = case_when(
      
      acuaculture_original.report == "yes" & aquacultureFLAG_combined == "YES" ~ "virtually certain; flags agree with original report",
      acuaculture_original.report == "yes" & aquacultureFLAG_combined == "NO" ~ "high; no flags for aquaculture but original report says it is",
      acuaculture_original.report == "yes" & aquacultureFLAG_combined == "POSSIBLY" ~ "virtually certain; flags agree with original report",
      
      acuaculture_original.report == "no" & aquacultureFLAG_combined == "NO" ~ "virtually certain; flags agree with original report",
      acuaculture_original.report == "no" & aquacultureFLAG_combined == "YES" ~ "medium; flags suggest likely aquaculture but original report says no; suspect record is misreported",
      acuaculture_original.report == "no" & aquacultureFLAG_combined == "POSSIBLY" ~ "low; flags suggest possibly aquaculture but original report says no; record could be misreported",
      
      is.na(acuaculture_original.report) & aquacultureFLAG_combined == "YES" ~ "high; no data from original report but flags suggest likely aquaculture",
      is.na(acuaculture_original.report) & aquacultureFLAG_combined == "POSSIBLY" ~ "low; no data from original report but flags suggest possibly aquaculture",
      is.na(acuaculture_original.report) & aquacultureFLAG_combined == "NO" ~ "medium-high; no data from original report but flags suggest unlikely aquaculture",
      TRUE ~ "not enough information to assign aquaculture estimate"
    )
  ) %>%
  
  # Drop the helper normalization column
  select(-acuaculture_original.report)

# SANITY CHECK:
table(GOC_Landings_aquaculture$est_aquaculture_production)
         # aquaculture   likely aquaculture      not aquaculture possibly aquaculture unlikely aquaculture 
         #       17496               281612              2622445                37649               663574 
               
subset<- GOC_Landings_aquaculture %>% filter(acuaculture_production==c("yes"))
table(subset$est_aquaculture_production) 
# Excellent, 100% of the aquaculture explicitly reported as aquaculture is estimated to be aquaculture (n= 17496).

min(subset$year)
max(subset$year)
# Very weird the aquaculture reports only go from 2006 to 2017

subset<- GOC_Landings_aquaculture %>% filter(est_aquaculture_production==c("aquaculture", "likely aquaculture"))
# 149382 obs. The data set grows a lot!!!

# Re-level est_aquaculture_production to be a little more logical:
GOC_Landings_aquaculture <- GOC_Landings_aquaculture %>%
  mutate(
    est_aquaculture_production = factor(
      est_aquaculture_production,
      levels = c(
        "aquaculture",
        "likely aquaculture",
        "possibly aquaculture",
        "unlikely aquaculture",
        "not aquaculture"
      )
    )
  )

# 3622776 obs., 48 vars.
```

```{r Summarize reported aquaculture versus estimated aquaculture}
table(GOC_Landings_aquaculture$acuaculture_production)
table(GOC_Landings_aquaculture$est_aquaculture_production)

SummaryOf_ReportedAquaculture<- GOC_Landings_aquaculture %>% 
  filter(acuaculture_production== c("yes")) %>%
  filter(year >= 2014, year <= 2024) %>%
  select(year, weight_kg_corrected, value) %>%
  group_by(year) %>%
  summarize(reported_annual_catch= sum(weight_kg_corrected, na.rm = TRUE),
            reported_annual_value= sum(value, na.rm = TRUE))

SummaryOf_EstimatedAquaculture<- GOC_Landings_aquaculture %>% 
  filter(est_aquaculture_production== c("aquaculture", "likely aquaculture")) %>%
  filter(year >= 2014, year <= 2024) %>%
  select(year, weight_kg_corrected, value) %>%
  group_by(year) %>%
  summarize(est_annual_catch= sum(weight_kg_corrected, na.rm = TRUE),
            est_annual_value= sum(value, na.rm = TRUE))

Aquaculture_Summary<- SummaryOf_EstimatedAquaculture %>%
  left_join(SummaryOf_ReportedAquaculture, by = "year") %>%
  select(year, reported_annual_catch, est_annual_catch, reported_annual_value, est_annual_value)
```

```{r Plot reported aquaculture versus estimated aquaculture}
ggplot(Aquaculture_Summary, aes(x = year)) +
  geom_line(aes(y = est_annual_catch, color = "Estimated catch"), linewidth = 1.2) +
  geom_point(aes(y = est_annual_catch, color = "Estimated catch"), size = 2) +
  
  geom_line(aes(y = reported_annual_catch, color = "Reported catch"), linewidth = 1.2) +
  geom_point(aes(y = reported_annual_catch, color = "Reported catch"), size = 2) +
  
  scale_color_manual(
    values = c(
      "Estimated catch" = "#1b9e77",
      "Reported catch" = "#d95f02"
    )
  ) +
  labs(
    x = "Year",
    y = "Catch (kg)",
    color = "",
    title = "Estimated vs Reported Aquaculture"
  ) +
  theme_minimal(base_size = 14)
```
Notice the spike in landings from 2017-2018? Of course, how could you miss it.
This jump coincides (temporally) with a change in reporting associated with the raw data itself. 
Data pre-2018 is reported here: mex_conapesca_avisos_2000_2019.rds
Data 2018 and onward is reported here: mex_conapesca_apertura_2018_present.rds
> NOTE TO SELF: Consult with Juan Carlos about these two datasets. What do different about them?


# STEP 8: Time to filter out freshwater taxa...
```{r Flag freshwater taxa}
#### FLAG 5 - Freshwater landings ####

Freshwater_keywords<- c(
  "DE AGUA DULCE"
)

Definite_FreshwaterTaxa<- c(
  "tilapia",
  "rana", "rana toro",
  "trucha",
  "bagre canal"
)

# Flag freshwater rows based on keywords, info from the IHH key, and estimated habitat type (based on landing site):
GOC_Landings_aquaculture <- GOC_Landings_aquaculture %>%
  mutate(
    freshwater_flag = coalesce(
      (common_name_clean %in% Definite_FreshwaterTaxa) |
      (rowSums(across(keywords_1:keywords_5, ~ .x %in% Freshwater_keywords), na.rm = TRUE) > 0) |
      (est_habitat == "freshwater") |
      (IHH_functional_major_groups_sm =="Freshwater fishes"),
      FALSE
    )
  )

table(GOC_Landings_aquaculture$freshwater_flag)
# 303293 observations explicitly flagged as freshwater landings (either based on the taxa, keyword, or landing site)
```

```{r SANITY CHECK: Do the records classed as aquaculture make sense?}
# SANITY CHECK:
table(GOC_Landings_aquaculture$est_aquaculture_production)
subset<- GOC_Landings_aquaculture %>% filter(est_aquaculture_production=="likely aquaculture")
table(subset$freshwater_flag)
table(subset$recorded_taxa_name_clean)
# Note that 172961 out of 281612 (~61%) records that are "likely aquaculture" are also flagged as freshwater species - things like Tilapia, trout. The marine species include snapper, jacks, lots of bivalves and snails.

subset2<- subset %>% filter(recorded_taxa_name_clean=="JUREL")
# NOTE TO SELF: Yes, these data for Jurel look really nice. These are definitely aquaculture jacks. Flags are doing a decent job so far!
```

```{r Reorder and rename columns in GOC_Landings_aquaculture in new df GOC_Landings_sorted}
GOC_Landings_sorted <- GOC_Landings_aquaculture %>%
  # rename(reported_aquaculture_production = acuaculture_production,
  #        reported_fleet = fleet) %>%
  relocate(
    reported_aquaculture_production,
    aquacultureFLAG_landing.site,
    aquacultureFLAG_keywords,
    aquacultureFLAG_specific.taxa,
    aquacultureFLAG_combined,
    est_aquaculture_production,
    est_aquaculture_certainty,
    est_habitat,
    freshwater_flag,
    reported_fleet,
    .after = last_col()
  )
# 3622776 obs., 49 vars.
```

# STEP 9: ESTIMATE SECTOR based on est_aquaculture_production and freshwater_flag
```{r Estimate sector}

#### ESTIMATE SECTOR based on report info and flags ####

GOC_Landings_sorted <- GOC_Landings_sorted %>%
  mutate(
    # Helper columns:
    is_aquaculture = est_aquaculture_production %in% c("aquaculture","likely aquaculture"),
    is_wildcatch = est_aquaculture_production %in% c("unlikely aquaculture","not aquaculture"),
    
    # Implement logic to estimate sector:
    sector_estimate_draft = case_when(
      
      #### FRESHWATER SECTORS ####
      freshwater_flag & is_aquaculture ~ "freshwater: aquaculture",
      freshwater_flag & is_wildcatch ~ "freshwater: wild capture",
      # Estimate: freshwater ornamentals > PENDING
      #freshwater_flag & ornamental_flag ~ "freshwater: ornamentals",
      freshwater_flag ~ "freshwater: yet to be classified",
 
      #### MARINE SECTORS ####     
      !freshwater_flag & is_aquaculture ~ "marine: mariculture",
      !freshwater_flag & is_wildcatch & reported_fleet == "small_scale" ~ "marine: small-scale fisheries",
      !freshwater_flag & is_wildcatch & reported_fleet == "large_scale" ~ "marine: large-scale fisheries",
      # Estimate: marine ornamentals > PENDING
      #!freshwater_flag & ornamental_flag ~ "marine: ornamentals",
      !freshwater_flag ~ "marine: yet to be classified",
      
      # Catch-all safeguard
      TRUE ~ NA_character_

    )
  ) %>%
  select(-is_aquaculture, -is_wildcatch)

# SANITY CHECK:
table(GOC_Landings_sorted$sector_estimate_draft) # Excellent! This all adds up to 3622776 obs.
         # freshwater: aquaculture         freshwater: wild capture freshwater: yet to be classified    marine: large-scale fisheries 
         #                  175109                           127130                             1054                           142404 
         #     marine: mariculture    marine: small-scale fisheries     marine: yet to be classified 
         #                  123999                           964673                          2088407 

table(GOC_Landings_sorted$est_aquaculture_production)
table(GOC_Landings_sorted$freshwater_flag)
table(GOC_Landings_sorted$fleet)

# Re-level
GOC_Landings_sorted <- GOC_Landings_sorted %>%
  mutate(
    sector_estimate_draft = factor(
      sector_estimate_draft,
      levels = c(
        "marine: yet to be classified",
        "marine: large-scale fisheries",
        "marine: small-scale fisheries",
        "marine: mariculture",
        "freshwater: aquaculture",
        "freshwater: wild capture",
        "freshwater: yet to be classified"
      )
    )
  )
```

```{r Summarize by sector}
# ESTIMATED:
SummaryOf_SectorLandings<- GOC_Landings_sorted %>% 
  filter(year >= 2014, year <= 2024) %>%
  select(year, weight_kg_corrected, value, sector_estimate_draft) %>%
  group_by(year, sector_estimate_draft) %>%
  summarize(est_annual_catch_kg= sum(weight_kg_corrected, na.rm = TRUE),
            est_annual_value= sum(value, na.rm = TRUE))

SummaryOf_SectorLandings$est_annual_catch_mton <- SummaryOf_SectorLandings$est_annual_catch_kg/1000

table(SummaryOf_SectorLandings$sector_estimate_draft)
```

```{r Plot each sector over time}
# Define colors for each sector (fleet distinction = lighter for small-scale)
sector_colors <- c(
  # Marine (blues)
  "marine: yet to be classified"       = "#cce0ff",  # very light blue for unknown
  "marine: small-scale fisheries"      = "#6699cc",  # medium blue
  "marine: large-scale fisheries"      = "#336699",  # darker blue
  "marine: mariculture"                = "#33cccc",  # teal/aquaculture cue
  
  # Freshwater (warm shades)
  "freshwater: yet to be classified"  = "#ffe0cc",  # light orange for unknown
  "freshwater: wild capture"           = "#ff6666",  # strong red/orange
  "freshwater: aquaculture"            = "#ffb366"   # lighter orange/yellow
)

# Plot
Landings_bySector_plot<- SummaryOf_SectorLandings %>%
  filter(year >= 2018) %>%
  ggplot(aes(x = year, y = est_annual_catch_mton, fill = sector_estimate_draft)) +
  
  # Area plot with slight border for separation
  geom_area(alpha = 0.9, color = "white", linewidth = 0.2) +
  
  # Manual fill palette
  scale_fill_manual(
    values = sector_colors,
    name = "Sector"
  ) +
  
  # Labels
  labs(
    x = "Year",
    y = "Estimated Annual Catch (metric tons)",
    title = "Estimated Catch by Sector",
    subtitle = "Filtered for 2018 onwards"
  ) +
  
  # Minimal theme and sleek styling
  theme_bw(base_size = 14) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey85"),
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 13, color = "grey30")
  ) +
  
  # Optional: better legend layout
  guides(
    fill = guide_legend(nrow = 3, byrow = TRUE, title.position = "top")
  )
```

Looks great! Save this plot.
```{r Save Landings_bySector_plot}
# Save the plot as a high-resolution PNG
ggsave(
  filename = "/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/Catch_bySector_plot_Mar132026.png",
  plot = Landings_bySector_plot,
  width = 10,       # width in inches
  height = 6,       # height in inches
  dpi = 300         # resolution
)
```

So that's estimated. What about what was actually reported?
```{r Plot what was actually reported}
colnames(GOC_Landings_sorted)

GOC_Landings_sorted <- GOC_Landings_sorted %>%
  mutate(
    sector_estimate_reported = case_when(
      
      # More specific first
      reported_aquaculture_production == "yes" & reported_fleet == "small_scale" ~ "small-scale aquaculture",
      
      # General aquaculture
      reported_aquaculture_production == "yes" ~ "aquaculture",
      
      # Fisheries by fleet
      reported_fleet == "small_scale" ~ "small-scale fisheries",
      reported_fleet == "large_scale" ~ "large-scale fisheries",
      
      # Fallback for missing/unknown
      TRUE ~ "yet to be classified"
    )
  )

# SANITY CHECK:
table(GOC_Landings_sorted$sector_estimate_reported)
         # aquaculture large-scale fisheries small-scale fisheries  yet to be classified 
         #        17496                142431               1052851               2409998 

# Re-level
GOC_Landings_sorted <- GOC_Landings_sorted %>%
  mutate(
    sector_estimate_reported = factor(
      sector_estimate_reported,
      levels = c(
        "aquaculture",
        "yet to be classified",
        "small-scale fisheries",
        "large-scale fisheries"
      )
    )
  )

# REPORTED:
SummaryOf_SectorLandings_REPORTED<- GOC_Landings_sorted %>% 
  filter(year >= 2014, year <= 2024) %>%
  select(year, sector_estimate_reported, weight_kg_corrected, value) %>%
  group_by(year, sector_estimate_reported) %>%
  summarize(est_annual_catch_kg= sum(weight_kg_corrected, na.rm = TRUE),
            est_annual_value= sum(value, na.rm = TRUE))

SummaryOf_SectorLandings_REPORTED$est_annual_catch_mton <- SummaryOf_SectorLandings_REPORTED$est_annual_catch_kg/1000

table(SummaryOf_SectorLandings_REPORTED$sector_estimate_reported)

# Plot
REPORTEDLandings_bySector_plot<- SummaryOf_SectorLandings_REPORTED %>%
  filter(year >= 2018) %>%
  ggplot(aes(x = year, y = est_annual_catch_mton, fill = sector_estimate_reported)) +
  
  # Area plot with slight border for separation
  geom_area(alpha = 0.9, color = "white", linewidth = 0.2) +
  
  # Labels
  labs(
    x = "Year",
    y = "Reported Annual Catch (metric tons)",
    title = "Catch by Reported Sector",
    subtitle = "Filtered for 2018 onwards"
  ) +
  
  # Minimal theme and sleek styling
  theme_bw(base_size = 14) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey85"),
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 13, color = "grey30")
  ) +
  
  # Optional: better legend layout
  guides(
    fill = guide_legend(nrow = 3, byrow = TRUE, title.position = "top")
  )

REPORTEDLandings_bySector_plot
```


# STEP 10: Let's look into jellyfish.
```{r}
# Create a vector of keywords
Jellyfish_keywords <- c("bola", "cañon", "canon", "aguamala", "medusa", "stomolophus")

# Filter using regex with ignore case
GOC_Landings_jellies <- GOC_Landings_sorted %>%
  filter(grepl(paste(Jellyfish_keywords, collapse = "|"), common_name_clean, ignore.case = TRUE))
# Found them!

table(GOC_Landings_jellies$recorded_taxa_group)
table(GOC_Landings_jellies$recorded_taxa_name_clean)

subset<- GOC_Landings_sorted %>% filter(is.na(estimated_genus_name))
table(subset$recorded_taxa_group)
table(subset$recorded_taxa_name_clean)
```


#### Now filter out freshwater & aquaculture landings from main landings ####

```{r Filter out freshwater and MAR species}

#### FRESHWATER
Freshwater_Landings <- GOC_Landings_aquaculture %>%
  filter(freshwater_flag)
# 286013 obs., 50 vars

FreshFisheries_Landings <- GOC_Landings_aquaculture %>%
  filter(freshwater_flag & !aquacultureFLAG_keywords)
# 280355 obs., 50 vars

FreshAquaculture_Landings <- GOC_Landings_aquaculture %>%
  filter(freshwater_flag & aquacultureFLAG_keywords)
# 5658 obs., 50 vars

#### MARINE - WILD CATCH AND MARICULTURE
Mariculture_Landings <- GOC_Landings_aquaculture %>%
  filter(!freshwater_flag & aquacultureFLAG_keywords)
# 130432 obs., 50 vars

MarineFisheries_Landings <- GOC_Landings_aquaculture %>%
  filter(!(aquacultureFLAG_keywords | freshwater_flag))
# 3143018 obs., 50 vars

sum(is.na(MarineFisheries_Landings$OHI_Region_No))
table(MarineFisheries_Landings$OHI_Region_No)
# Seems like all entries have an OHI region number...
```
NOTE TO SELF: There could still be some freshwater species hiding in here. For example, the "MOJARRA" general category is mixed freshwater and marine demersal fish.


# STEP 11: Interogate the EU and permit holder columns
```{r First look}
length(unique(GOC_Landings_clean$eu_name))
# 7414 unique economic units in the dataframe??

length(unique(GOC_Landings_clean$eu_rnpa))
# 6907 unique permits given to economic units in the dataframe??

length(unique(GOC_Landings_clean$vessel_name))
# 4179 unique vessel names in the dataframe

length(unique(GOC_Landings_clean$vessel_rnpa))
# 4169 unique vessel permit numbers in the dataframe

# So, for the most part, each vessel is registered to a unique vessel permit.'

# How many of the 3622776 observations have no economic unit name associated with them? Only ~2% of all records.
GOC_Landings_clean %>%
  count(eu_name == "NO DISPONIBLE")
# 72381 / 3622776

# How many of all observations have no vessel name associated with them? About 53% of all records.
GOC_Landings_clean %>%
  count(vessel_rnpa == "00000000")
# 1908721 / 3622776

# This tells us that about half of all records have no information about the fishers' vessel, but nearly all have permit info.
```

```{r Testing some ideas}
table(GOC_Landings_clean$acuaculture_production)

subset_aquaculture<- GOC_Landings_clean %>%
  filter(acuaculture_production == "yes")

vector_aquaculture<- unique(subset_aquaculture$landing_site)

subset2<- GOC_Landings_clean %>%
  filter(eu_name == "SCPP PUNTA TIBURON, S DE RL DE CV")
# Notice that some of these landings say nothing about aquaculture or not, whereas others do. It looks like the economic unit (possibly a cooperative?) does not dabble in aquaculture. 

subset3<- GOC_Landings_clean %>%
  filter(eu_name == "SCPP DE SERV LA CAIMANERA SC DE RL")

rm(subset, subset1, subset2, subset3)
```
Let's leave this section under construction. I'm not totally sure what to do with this yet. But it will impact how and if we can calculate CPUE.
Waiting to here back from Juan Carlos is if there is a key or roster for economic units...


STOPPED HERE, Mar 12 2026