################################################################################ 

For questions contact: Erica Ferrer, Ph.D. at emferrer@ucsd.edu - OR - ferrer@nceas.ucsb.edu

################################################################################ 

# PRE-WORK ------------------------------

```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```

# STEP 1: Load in CONAPESCA RAW data, attend to records from all of Mexico. ------------------------------

## STEP 1a: Work with CONAPESCA **RAW** data (records from all of Mexico):
This very first step is completed in the file shared by Juan Carlos: "03_combine_landings_edited.R" file path: /home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/JC_Scripts_EMF.edited/03_combine_landings_edited.R

## STEP 1b: Work with CONAPESCA **STITCHED** data (records from all of Mexico):

A stitched version of these data is available as one single file: "AllMex_landings_2000_present_EMF_edited.rds" file path: /home/shares/ohi/OHI_GOC/goal_prep/fp/juan_carlos/AllMex_landings_2000_present_EMF_edited.rds

```{r Load in CONAPESCA STITCHED data}
All.Mex_Landings_2000.present<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/juan_carlos/AllMex_landings_2000_present_EMF_edited.rds"))
# n = 12153879 obs., 20 vars.
```


## STEP 1c: Create weight_kg_corrected column (records from all of Mexico):

```{r Landed weight correction}
# Kind of looks like sometimes 'landed weight' is 'live weight' and vice versa. Fix this with weight_kg_corrected:
All.Mex_Landings_2000.present$weight_kg_corrected <- ifelse(
  All.Mex_Landings_2000.present$live_weight >= All.Mex_Landings_2000.present$landed_weight,
  All.Mex_Landings_2000.present$live_weight,
  All.Mex_Landings_2000.present$landed_weight
)
# n = 12153879 obs., 21 vars.
```

## STEP 1d: optional save...
Pending, if needed.


# STEP 2: Filter down CONAPESCA data to states in Northwest Mexico.  ------------------------------

```{r Filter out data from Northwest Mexico}
length(unique(All.Mex_Landings_2000.present$species_name)) 
# 2052 unique species in all of Mexico.
# But, we don't need species from ALL of Mexico, only the GoC region for now - so let's filter by state.

States_of_Interest<- c("BAJA CALIFORNIA", "BAJA CALIFORNIA SUR", "SONORA", "SINALOA", "NAYARIT", "JALISCO")

NW.Mex_Landings <- All.Mex_Landings_2000.present %>% filter(state %in% States_of_Interest)
# n = 4713144 obs., 21 vars.

length(unique(NW.Mex_Landings$species_name)) 
# 1827 unique species in States of Interest.
```


# STEP 3: Attach lat-long data to NW Mexico dataframe. [UNDER CONSTRUCTION, Mar 25 2026] ------------------------------

# Flag these in the lat-long key as not in the Gulf [UNDER CONSTRUCTION April 3 2026]
office_name== c("EL ROSARIO", "ENSENADA", "TIJUANA")

## STEP 3a: Attach lat-long and OHI_region info based on official landing site:
NOTE TO SELF: Come back to this! The lat long keys are imperfect. We'll want to fix this as much as possible.
```{r Append lat-long data to sites in Baja BCS Sonora and Sinaloa}
# Filter OUT sites from Baja that are on the Pacific side using lat-Long file from Mel's repository:
# NOTE TO SELF: THIS PART IS SUBJECT TO CHANGE once we integrate lat-long data from Hem. Mar 10 2026
BajaSonora_LatLongKey<- read_csv("/home/frazier/ohi_goc_prep/FP/Exploratory_Research/FIS_TeamResearch/data/Baja_Sonora_states_key.csv")
colnames(BajaSonora_LatLongKey) <- c("state", "landing_site", "OHI_rgn", "lat", "long", "GoC")

Jalisco_LatLongKey<- read_csv(here("/home/frazier/ohi_goc_prep/FP/Exploratory_Research/FIS_TeamResearch/data/landing_sites_jalisco.csv"))
colnames(Jalisco_LatLongKey) <- c("landing_site", "CONAPESCA_landing_key", "est_habitat", "lat", "long", "site_notes", "OHI_rgn", "state")


#### CLEAN LAT-LONG KEYS: ####
# Clean the Baja-Sonora key...
BajaSonora_LatLongKey_clean <- BajaSonora_LatLongKey %>%
  
  # Normalize state names:
  mutate(state = toupper(state)) %>%
  
  # Remove any rows where landing_site is NA (NAs cause matching issues later on):
  filter(!is.na(landing_site)) %>%
  
  # Remove any direct duplicates of landing_site:
  group_by(state, landing_site) %>%
  slice(1) %>%  # keeps the first occurrence per key
  ungroup()

# Look into the rows that were removed in the de duplication of BajaSonora_LatLongKey...
removed_rows <- anti_join(BajaSonora_LatLongKey, BajaSonora_LatLongKey_clean, 
                          by = c("landing_site"))

# Convert lat-long entries to numeric entries, accounting for unique cases where the minus (-) is actually an n dash or m dash:
BajaSonora_LatLongKey_clean <- BajaSonora_LatLongKey_clean %>%
  mutate(
    long = as.numeric(str_replace_all(long, "[−–—]", "-")),
    lat  = as.numeric(str_replace_all(lat, "[−–—]", "-"))
  )

# Clean the Jalisco key...
Jalisco_LatLongKey_clean <- Jalisco_LatLongKey %>%
  # Normalize state names
  mutate(state = toupper(state)) %>%
  
  # Remove any rows where landing_site is NA (NAs cause matching issues later on):
  filter(!is.na(landing_site)) %>%
  
  # Keep only one row per CONAPESCA_landing_key; if duplicate keys exist, keep first
  group_by(CONAPESCA_landing_key) %>%
  slice(1) %>%  # keeps the first occurrence per key
  ungroup() %>%
  
  # Optional: remove any direct duplicates of landing_site if there are rows with NA in the CONAPESCA_landing_key
  group_by(state, landing_site) %>%
  slice(1) %>%  # keeps the first occurrence per key
  ungroup()

# Look into the rows that were removed in the de duplication of Jalisco_LatLongKey...
removed_rows <- anti_join(Jalisco_LatLongKey, Jalisco_LatLongKey_clean, 
                          by = c("CONAPESCA_landing_key", "landing_site"))

# Convert lat-long entries to numeric entries, accounting for unique cases where the minus (-) is actually an n dash or m dash:
Jalisco_LatLongKey_clean <- Jalisco_LatLongKey_clean %>%
  mutate(
    long = as.numeric(str_replace_all(long, "[−–—]", "-")),
    lat  = as.numeric(str_replace_all(lat, "[−–—]", "-"))
  )

# Add the "GoC" helpful column here based on a latitude cut off for the Nayarit/Jalisco OHI region, region 9 starting at 20.40895:
Jalisco_LatLongKey_clean <- Jalisco_LatLongKey_clean %>%
  mutate(GoC = if_else(lat < 20.40895, "no", NA_character_))

#### MERGE LAT-LONG KEYS TO CREATE MASTER LAT-LONG KEYS: ####
Mels_LatLongKey_pending <- bind_rows(BajaSonora_LatLongKey_clean, Jalisco_LatLongKey_clean) %>%
  filter(!is.na(landing_site)) %>%  # remove the NA from this data frame as this will create a many-to-many matching issue later on
  mutate(long = as.numeric(long))
# 1176 obs., 9 vars.

# Remove unused data frames:
rm(BajaSonora_LatLongKey, Jalisco_LatLongKey, removed_rows)

# Take a look
unique(Mels_LatLongKey_pending$state)
unique(Mels_LatLongKey_pending$OHI_rgn)
```

# STEP 3b: Integrate Hem's Lat-Long key ------------------------------
```{r Loading in Hem's lat-long}
load("/home/shares/ohi/OHI_GOC/goal_prep/fp/hem_morzaria/georefcatchlocs.rda")
Hems_LatLongKey0<- georefcatchlocs

rm(georefcatchlocs)

# Explore Hem's key
colnames(Hems_LatLongKey0)

# table(Hems_LatLongKey0$NOM_LOC) # This is the equivalent of landing_site.
# table(Hems_LatLongKey0$NOM_ENT) # Okay, so fishery_office_ent is essentially the state. We can match by this column.
# 
# table(Hems_LatLongKey0$fishery_office)
# table(Hems_LatLongKey0$fishery_office_ent) # Okay, so fishery_office_ent is also the state? I think this is the state of the fishery office. Ignore for now.

# Before we can merge we need to re-name some columns:
colnames(Hems_LatLongKey0)<- c("catch_code", "rnp_code", "landing_site", "state", "fishery_office", "fishery_office_ent", "lat_deci_Hem", "lon_deci_Hem" )
```

```{r Look through Hem's key}
Hems_LatLongKey_simple<- Hems_LatLongKey0 %>% 
  select(state, landing_site, fishery_office, lat_deci_Hem, lon_deci_Hem)
# Could include or exclude fishery_office. These offices are no longer used by CONAPESCA.

Hems_LatLongKey_simple<- Hems_LatLongKey_simple %>% 
  distinct()
# 829 obs.

length(unique(Hems_LatLongKey_simple$landing_site))
# 691 unique landing sites, at least by name, in Hem's key

length(unique(Mels_LatLongKey_pending$landing_site))
# 1068 "unique" landing sites, at least by name, in Mel's key

# Let's take a closer look at landing sites with the same name but different lat-longs in both keys. Are they at least in the same state? Are their lat-longs similar?
Site_duplicates_HemsKey <- Hems_LatLongKey_simple %>%
  select(state, landing_site, fishery_office) %>%
  distinct() %>%
  group_by(state, landing_site) %>%
  filter(n() > 1)
# From this it is clear that Hem's key is assigning a unique lat-long value to every combination of state, landing_site, AND fishery_office. This creates a situation where we might have more than one unique combination of lat-long values per state and landing_site (due to multiple fishery offices); however, the lat-long values are fairly similar in most cases....

Site_duplicates_OHIKey<- Mels_LatLongKey_pending[
  duplicated(Mels_LatLongKey_pending[, c("state", "landing_site")]) |
  duplicated(Mels_LatLongKey_pending[, c("state", "landing_site")], fromLast = TRUE),
]
# 0 duplicates. Yay!

# EXAMPLE:
subset<- Hems_LatLongKey_simple %>%
  filter(landing_site== "EL SIARIC")
# Lat-longs are all pretty similar here. I think it makes sense to just kind of pick one combo and move on. 

# TAKING WHAT WE CAN FROM HEM'S KEY:
# Let's take one representative value for lat-long from Hem's key and flag any duplicate site lat-longs that are more than 1 degree apart:

# Define thresholds
threshold_alert <- 2      # alert if differences > 2 degrees
threshold_low <- 1        # low confidence if differences > 1 degree
threshold_medium <- 0.5   # medium confidence if differences > 0.5 degree

# Step 1: Compute lat/lon ranges and assign confidence levels
Hems_LatLong_flags <- Hems_LatLongKey_simple %>%
  group_by(state, landing_site) %>%
  summarise(
    lat_range = max(lat_deci_Hem, na.rm = TRUE) - min(lat_deci_Hem, na.rm = TRUE),
    lon_range = max(lon_deci_Hem, na.rm = TRUE) - min(lon_deci_Hem, na.rm = TRUE),
    lat_long_confidence = case_when(
      lat_range > threshold_alert | lon_range > threshold_alert ~
        "alert 3: multiple lat-long values in Hem's key that are more than 2 degrees apart",
      lat_range > threshold_low | lon_range > threshold_low ~
        "alert 2: multiple lat-long values in Hem's key that are more than 1 degree apart",
      lat_range > threshold_medium | lon_range > threshold_medium ~
        "alert 1: multiple lat-long values in Hem's key that are more than 0.5 degree apart",
      TRUE ~ "high confidence"
    ),
    .groups = "drop"
  )

# Step 2: Keep a representative lat-long (first occurrence)
Hems_LatLongKey_compressed <- Hems_LatLongKey_simple %>%
  group_by(state, landing_site) %>%
  slice(1) %>%
  ungroup()

# Step 3: Merge the confidence/alert information
Hems_LatLongKey_compressed <- Hems_LatLongKey_compressed %>%
  left_join(Hems_LatLong_flags %>% select(state, landing_site, lat_long_confidence),
            by = c("state", "landing_site"))

# SANITY CHECK: Looks good!
print(Hems_LatLongKey_compressed)
# 724 obs. of 5 variables

# SANITY CHECK: Number of unique state + landing_site combinations
n_unique_sites <- Hems_LatLongKey_compressed %>%
  distinct(state, landing_site) %>%
  nrow()
# 724 unique state-landing_site combinations.

# remove dataframes to keep workspace clean:
rm(Hems_LatLong_flags, Hems_LatLongKey_simple, 
   Site_duplicates_HemsKey, Site_duplicates_OHIKey,
   subset)
```

```{r Create a backbone of a key to merge to}
Hems_LatLongKey_compressed # 724 obs. of 6 variables
Mels_LatLongKey_pending # 1176 obs. of 9 variables

# Shave down Mel's key:
Mels_LatLongKey_simple<- Mels_LatLongKey_pending %>%
  select(state, landing_site, lat, long, OHI_rgn, GoC, est_habitat, site_notes)
colnames(Mels_LatLongKey_simple)<- c("state", "landing_site", "lat_Mel", "lon_Mel", "OHI_region_draft", "GoC", "est_habitat", "site_notes_Mel")
# 1176 obs. of 8 variables
table(Mels_LatLongKey_simple$GoC)
# 400 sites in the GoC, 588 sites likely not in the GoC

# Create a backbone to merge to...
colnames(NW.Mex_Landings)

LatLong_merge.skeleton<- NW.Mex_Landings %>%
  select(state, landing_site) %>%
  distinct()
# 2046 distinct state-landingsite combos in (NW Mexico) that need to be filled using information from Hem's key or Mel's key. Privilege information from Hem's key.

# LOGIC:
# Take information from Hem's key and Mels key to populate the LatLong_merge.skeleton but in a preferential order. First, populate LatLong_merge.skeleton with information from Mel's key. Then, if landing_site information is available from Hem's key, supersede the state and landing_site info given by Mel's key. 
# If the lat-long values from Hem's key are non-NA and > 2 degrees different from Mel's non-NA lat-long values assign a lat.long_confidence of "low confidence, the lat-long values in Hem's key do not match the lat-long values in Mel's key by more than 2 degrees.

# Start with the skeleton
LatLong_merged <- LatLong_merge.skeleton %>%
  # Join Mel's key first (preferential population)
  left_join(Mels_LatLongKey_simple, by = c("state", "landing_site")) %>%
  # Join Hem's key
  left_join(Hems_LatLongKey_compressed %>%
              select(state, landing_site, lat_deci_Hem, lon_deci_Hem, lat_long_confidence),
            by = c("state", "landing_site")) %>%
  mutate(
    # Determine if Hem should override Mel
    use_HemsKey = !is.na(lat_deci_Hem) & !is.na(lon_deci_Hem),
    
    # Compute difference between Hem and Mel
    lat_diff = ifelse(!is.na(lat_deci_Hem) & !is.na(lat_Mel), abs(lat_deci_Hem - lat_Mel), 0),
    lon_diff = ifelse(!is.na(lon_deci_Hem) & !is.na(lon_Mel), abs(lon_deci_Hem - lon_Mel), 0),
    
    # Decide final lat/lon
    lat_final = case_when(
      use_HemsKey & lat_diff <= 2 & lon_diff <= 2 ~ lat_deci_Hem,  # Hem overrides when reasonable
      TRUE ~ lat_Mel  # Otherwise, keep Mel
    ),
    lon_final = case_when(
      use_HemsKey & lat_diff <= 2 & lon_diff <= 2 ~ lon_deci_Hem,
      TRUE ~ lon_Mel
    ),
    
    # Decide final confidence
    lat_long_confidence = case_when(
      use_HemsKey & (lat_diff > 2 | lon_diff > 2) ~
        "alert 4: the lat-long values in Hem's key do not match the lat-long values in Mel's key by more than 2 degrees",
      use_HemsKey & (lat_diff <= 2 & lon_diff <= 2) ~ lat_long_confidence,  # keep Hem's original
      TRUE ~ NA_character_  # default when using Mel or no Hem info
    )
  )

# Great! Seems like this worked. 
table(LatLong_merged$lat_long_confidence)
table(LatLong_merged$GoC)

#                             alert 1: multiple lat-long values in Hem's key that are more than 0.5 degree apart 
#                                                                                                             21 
#                               alert 2: multiple lat-long values in Hem's key that are more than 1 degree apart 
#                                                                                                             19 
#                              alert 3: multiple lat-long values in Hem's key that are more than 2 degrees apart 
#                                                                                                             11 
# alert 4: the lat-long values in Hem's key do not match the lat-long values in Mel's key by more than 2 degrees 
#                                                                                                             40 
#                                                                                                high confidence 
#                                                                                                            499
```

```{r Clean up key before export}
LatLongKey_finalized<- LatLong_merged
# 2046 obs., 13 vars.

colnames(LatLongKey_finalized)

LatLongKey_finalized<- LatLongKey_finalized %>% 
  select(state, landing_site, GoC, est_habitat, 
         lat_final, lon_final, lat_long_confidence, lat_deci_Hem, lon_deci_Hem, lat_Mel, lon_Mel, 
         OHI_region_draft, site_notes_Mel)

colnames(LatLongKey_finalized)<- c("state", "landing_site", "inside_GOC", "est_habitat", "lat_final", "lon_final", "FLAGS", "HemKey_lat", "HemKey_lon", "MelKey_lat", "MelKey_lon", "OHI_region_draft", "site_notes_Mel")

sum(is.na(LatLongKey_finalized$lat_final))
# 680 out of 2026 sites are missing lat-long info. Primarily in Sinaloa and Nayarit, which makes some sense for our purposes.

subset<- LatLongKey_finalized %>% 
  filter(is.na(lat_final))
table(subset$state)
    # BAJA CALIFORNIA BAJA CALIFORNIA SUR             JALISCO             NAYARIT             SINALOA              SONORA 
    #              54                  17                  25                 193                 324                  67

# remove dataframes to keep workspace clean:
rm(subset, 
   Hems_LatLongKey0, Hems_LatLongKey, Hems_LatLongKey_compressed, Hems_LatLong_flags, Hems_LatLongKey_simple,
   BajaSonora_LatLongKey, BajaSonora_LatLongKey_clean,
   Jalisco_LatLongKey, Jalisco_LatLongKey_clean,
   LatLong_merge.skeleton, LatLong_merged,
   Mels_LatLongKey_pending, Mels_LatLongKey_simple)

rm(n_unique_sites, threshold_alert, threshold_low, threshold_medium)
```

```{r Export this new lat-long key}
# Save this dataframe:
write.csv(LatLongKey_finalized, "/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/Sites_LatLongKey_Apr22026.csv", row.names = FALSE)
# 2046 obs., 13 vars.
```

# NOT TO SELF: This integration should move to the .rmd file "CONAPESCA_CreatingLatLongKey_Mar2026.Rmd" [PENDING, Apr 2 2026]
found /home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys

# NOTE TO SELF: Will want to regenerate fishery landings files with new lat-long data. - Mar 25 2026. [COMPLETE Apr 2 2026.]

# STEP 3c: Add lat-long and OHI_region info to landings data ------------------------------
```{r OHI LatLongKey_finalized and RegionKey_pending}
# Final lat-long key (subject to change as Apr 2 2026):
LatLongKey_finalized<- read_csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/Sites_LatLongKey_Apr22026.csv")
# 2046 obs., 13 vars.

# My regional key (temporary, subject to change, UNDER CONSTRUCTION):
RegionKey_pending<- read_csv(here("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/RegionKey_UNDERCONSTRUCTION_Mar102026.csv"))
# 14 obs., 6 vars.
# Previously: RegionKey_UNDERCONSTRUCTION_Dec012025.csv

colnames(RegionKey_pending)
colnames(RegionKey_pending) <- c("state", "OHI_region_draft", "OHI_rgn_number", "OHI_rgn_name", "exclude_rgn", "OHI_rgn_notes")
```

```{r Append lat-long and OHI_region info to landings data}
# Append lat-long info to landings data:
temp_df <- NW.Mex_Landings %>%
  left_join(
    LatLongKey_finalized,
    by = c("state", "landing_site")  # matching columns
  )
# 4713144 obs., 32 vars.

table(temp_df$inside_GOC)
# 1090368 "no" not in the GOC.

# Append RegionKey_pending info to landings data:
temp_df <- temp_df %>%
  left_join(
    RegionKey_pending,
    by = c("state", "OHI_region_draft")  # matching columns
  )
# 4713144 obs., 36 vars.
table(temp_df$inside_GOC)
# 1090368 "no" not in the GOC - same as before - as we'd expect!

colnames(temp_df)

# Looking into the exclude_rgn column:
test<- temp_df %>%
  filter(exclude_rgn=="exclude")

table(test$inside_GOC)
table(test$state)
table(test$OHI_rgn_number)
  # BCS_Region_3_4_TBD  Region_1_2_Baja_TBD Region_9_Jalisco_TBD 
  #             184929                39899               156952 


NW.Mex_Landings<- temp_df
# 4713144 obs., 36 vars.

# Delete temp_df, not needed anymore.
rm(temp_df)
```

## STEP 3d: Save the NW Mexico landings appended with lat-long info in a separate .rds file:
```{r Save the Northwest Mexico dataframe: Mar 11 2026}
saveRDS(NW.Mex_Landings, file="/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/NWMex_Landings_2000.present_EMF.Apr022026.rds")
# 4713144 obs., 36 vars.
```

# STEP 4: Filter down to sites in the GOC based on lat-long data. ------------------------------
# Updated to include lat-longs from Hem's key as of April 2 2026.

```{r Load NW.Mex_Landings if needed}
NW.Mex_Landings<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/NWMex_Landings_2000.present_EMF.Apr022026.rds"))
# 4713144 obs., 36 vars.
```

## STEP 4a: Exclude all sites that definitely AREN'T in the Gulf, leave the others for now...
NOTE TO SELF: Additional research will be required for some landing sites. Details pending additional lat-long key work. See notes here (as of Mar 11 2026): https://docs.google.com/document/d/1ANHEdRi8aWC0qLNaM1_c_nLtn5GihTFgHDcU96bry5Y/edit?tab=t.0
```{r filter down data to only the GOC}
# Generate filtered landings data for the GoC:
GOC_Landings <- NW.Mex_Landings %>%
  filter(inside_GOC != "no" | is.na(inside_GOC))
# 3622776 obs., 35 vars.

# Check that this all makes sense:
length(unique(GOC_Landings$species_name))
length(unique(GOC_Landings$main_species_group))
# 1703 unique species from 57 main species groups in the Gulf of California, keeping in mind that these are not really SPECIES (per se) but rather log-book entries for species.

# trim dataframe if desired:
colnames(GOC_Landings)

GOC_Landings<- GOC_Landings %>%
  select(-c(HemKey_lat, HemKey_lon,
            MelKey_lat, MelKey_lon,
            site_notes_Mel)) %>%
  rename(est_latlon_FLAG= FLAGS,
         est_lat= lat_final,
         est_lon= lon_final)
# 3622776 obs., 31 vars. (-5 columns)
```
NOW 1703 unique species in the Gulf of California according to Juan Carlos' version of the CONAPESCA data base, keeping in mind that these are not really SPECIES, per se, but rather log-book entries for species.

## STEP 4b: Save the GOC_Landings dataframe as a separate .rds file:
```{r Save the GOC_Landings dataframe: Mar 11 2026}
saveRDS(GOC_Landings, file="/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Landings_2000.present_EMF.Apr022026.rds")
# 3622776 obs., 31 vars.
# Previous iteration: GOC_FIS.landings_CLEAN_as.of.Dec092025.rds
```

```{r Clean workspace as needed}
# remove dataframes (as needed) to keep workspace clean:
rm(All.Mex_Landings_2000.present, LatLongKey_finalized, RegionKey_pending)
```


# > IF SKIPPING STEPS 1-4 - curation of raw data files and the landings dataframe, then move to: : ------------------------------

# STEP 5: Separate out key words from the official record for common name. ------------------------------

# Step 5a: Load in GOC_Landings dataframe
```{r Load GOC_Landings if needed}
GOC_Landings<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Landings_2000.present_EMF.Apr022026.rds"))
# 3622776 obs., 31 vars.
```

Keeping in mind that some key terms reflect animal parts but are actually part of the species common name like "aleta de" versus "aleta blanca"

** QUICK SANITY CHECK: Is "aleta de aleta" an issue? Short answer is no; "aleta de aleta" does not seem to be an issue.
GOC_Landings %>%
  rename(recorded_taxa_name = species_name) %>%
  mutate(
    aleta_count = stringi::stri_count_fixed(recorded_taxa_name, "ALETA")
  ) %>%
  count(aleta_count)

```{r Filter out keywords from recorded_taxa_name to get recorded_taxa_name_clean}

# PART A. Keyword list [UNDER CONSTRUCTION] -----------------

# NOTE: Keyword list excludes "ALETA" which is handled separately.

Keyword_list <- c(
# Processing state or product form (e.g., frozen, full fish, filleted, beheaded, dried, etc.):
  "S.C.", "S/C", "S",
  "FRESCO", "FRESCA", "FCA", "FCO", "FC",
  "ECO",
  "CONG", "CONGELADO",
  "SECO DE", "SECA", "SECA DE", "SECO", "SECO DE",
  "COCINERO", "COCIDO", "COCIDA", "COCIDAS", "ASADO",
  "ENTERO", "ENTERA", "IND", "ENT",
  "FILETE", "FILETE DE", "ECA", "FILETE DE ECO", "DESHUESADA", "CARNE", "CARNE DE",
  "DESCABEZADO", "DES", "DSC", "DESC", "DESV", "DESVICERADO",

# Animal parts:
  "MANTO", "MANTO DE",
  "PULPA DE", "PULPA FRESCA", "PULPO DE", 
  "POLVO DE",
  "TENTACULOS DE", "TUBO DE", "TUBO",
  "CABEZA DE", "BUCHE", "BUCHE DE", "TRIPA DE", "COLAS DE", "MANOS DE", "MANOS",
  "ANCAS DE", "ANCAS", "ANCA",
  "PIEL DE ECA", "PIEL DE SECA", "PIEL DE", "PIEL",

# Remove ALETA from main keywords; we'll handle these keywords separately:
  #"ALETA", "ALETA ECA DE", "ALETA FCA DE"

# Shells and pearls:
  "CONCHA DE", "CONCHA", "CONCHA NACAR",
  "PERLAS", "MEDIAS PERLAS",

# From where / habitat type:
  "DE ALTA MAR",
  "DE AGUA DULCE",

# Live culture or cultivation techniques:
  "VIVO", "VIVA", "VIVAS", "VIVITA",
  "DE CULTIVO", "DE CULTIVO VIVO", "CULTIVADAS",
  "ABASTO RANCHO", "RANCHO",

# Eggs:
  "HUEVA DE",

# Used as bait:
  "CARNADA", "P/CARNADA",

# Other:
  "NO APTO PARA EMPAQUE"
)

# 81 keywords as of Mar 23 2026.

# Order the key words longest first (note that this is necessary for the next steps):
Keyword_list <- Keyword_list[order(nchar(Keyword_list), decreasing = TRUE)]

# Escape regex meta-characters:
Keyword_list_escaped <- stringi::stri_replace_all_regex(
  Keyword_list, "([.|(){}+*?^$\\[\\]\\\\])", "\\\\$1"
)

# Final regex:
Keyword_pattern <- paste0(
  "(?<![A-Za-z0-9])(",
  paste(Keyword_list_escaped, collapse = "|"),
  ")(?![A-Za-z0-9])"
)


# PART B. Flag special "ALETA" phrases (handled differently from other key words) -----------------
special_aleta_cases <- c("BLANCA", "AMARILLA", "CORTA", "AZUL")

remove_aleta_except_special <- function(x) {
  stringi::stri_replace_all_regex(
    x,
    pattern = paste0("\\bALETA\\b(?!\\s+(?:", paste(special_aleta_cases, collapse = "|"), "))"),
    replacement = "",
    vectorize_all = FALSE
  )
}

# NOTE TO SELF: How did I get this list of key words originally? I think I started out with some preliminary version of the separate, parse functionality, quickly spotted the beginnings of a keywords list and then grew the list over time. Come back to this at some point to recreate this early part of the workflow to document it (reproducibility).
```

# STEP 5b: Separate out keywords from the original landing record using the parse regex in PART C:
```{r Separate keywords from landing record based on parse function}

# PART C. MAIN PARSING PIPELINE -----------------
# RUN TIME= 5-6 min, optimized but still computationally heavy

GOC_Landings_parsed <- GOC_Landings %>%
  
  # Rename columns:
  rename(
    recorded_taxa_name = species_name,
    recorded_taxa_group = main_species_group
  ) %>%
  
  # Extract keywords:
  mutate(
    keywords_list = stringi::stri_extract_all_regex(recorded_taxa_name, Keyword_pattern)
  ) %>%
  mutate(keywords_list = map(keywords_list, ~ { length(.x) <- 5; .x })) %>%
  unnest_wider(keywords_list, names_sep = "_") %>%
  rename_with(~ paste0("keywords_", 1:5), starts_with("keywords_list_")) %>%
  
  # Clean recorded_taxa_name:
  mutate(
    recorded_taxa_name_clean = recorded_taxa_name %>%
      remove_aleta_except_special() %>%                      # Remove ALETA unless special
      stringi::stri_replace_all_regex(Keyword_pattern, "") %>% # Remove other keywords
      stringi::stri_replace_all_regex("\\b(Y|DE)\\b", "") %>%   # Remove standalone modifiers like Y / DE
      stringi::stri_replace_all_regex("P\\s*/\\s*", "") %>%       # remove P/ or P /
      stringi::stri_replace_all_regex("[.,/\\\\]+", " ") %>%      # Remove punctuation
      str_squish()
  ) %>%
  
  # Extract scientific name:
  mutate(
    recorded_scientific_name = stringi::stri_extract_first_regex(recorded_taxa_name_clean, "\\(([^)]+)\\)") %>%
      stringi::stri_replace_all_regex("\\(|\\)", ""),
    scientific_name_recorded_T.F = !is.na(recorded_scientific_name),
    recorded_scientific_name = if_else(
      scientific_name_recorded_T.F,
      stringi::stri_trans_totitle(recorded_scientific_name),
      NA_character_
    )
  ) %>%
  
  # Extract common name:
  mutate(
    common_name_clean = stringi::stri_trim_both(
      stringi::stri_replace_first_regex(recorded_taxa_name_clean, "\\s*\\([^)]*\\)", "")
    ),
    common_name_clean = tolower(common_name_clean)
  )
# 3622776 obs., 40 vars.


# PART D. Trim and reorder columns -----------------
colnames(GOC_Landings_parsed)

GOC_Landings_parsed <- GOC_Landings_parsed %>%

  select(
    # General info from record:
    state, office_name, landing_site, landing_site_key, 
    est_lat, est_lon, est_latlon_FLAG, inside_GOC, est_habitat,
    year, month,
    receipt_date_fixed, period_start_fixed, period_end_fixed,
    fleet, acuaculture_production, eu_rnpa, eu_name, vessel_rnpa, vessel_name,
    
    # Species info from record:
    recorded_taxa_group, recorded_taxa_name,
    
    # Cleaned and derived names:
    recorded_taxa_name_clean, recorded_scientific_name,
    scientific_name_recorded_T.F, common_name_clean,
    
    # Keywords that have been parsed:
    keywords_1, keywords_2, keywords_3, keywords_4, keywords_5,
    
    # Landing details from record:
    value, landed_weight, live_weight, weight_kg_corrected,
    
    # OHI region and notes:
    OHI_rgn_number, OHI_region_draft ,OHI_rgn_name, OHI_rgn_notes, exclude_rgn
 
  )
# 3622776 obs., 40 vars.

# How many common names are there in this data frame?
length(unique(GOC_Landings_parsed$common_name_clean))
length(unique(GOC_Landings_parsed$recorded_taxa_group))
# NOW 732 common names (previously 973 with the less comprehensive keywords_list), over 57 taxonomic groups
```

## VIEW ALL KEY WORDS... SANITY CHECKS: Did this work as expected? Does the Keyword_list need more work?
```{r SANITY CHECK 1: Investigating Keywords}
# Extract all unique keywords from keyword columns 1 through 5. Delete exact duplicates. 
Keyword_vector <- GOC_Landings_parsed %>%
  select(keywords_1:keywords_5) %>%
  unlist() %>%
  unique() %>%
  na.omit()

# Why is Keyword_vector shorter than main Keyword_list?
setdiff(Keyword_list, Keyword_vector)
# 8 keywords that aren't featured in the parsed data set:
# "FILETE DE ECO", "PIEL DE ECA", "MANTO DE", "TUBO DE", "RANCHO", "BUCHE", "ANCAS", "ECO", "ECA"    

# Example for "TUBO DE"
GOC_Landings_parsed %>%
  select(recorded_taxa_name) %>%
  filter(str_detect(recorded_taxa_name, regex("TUBO", ignore_case = TRUE)))

# Example for "RANCHO"
GOC_Landings_parsed %>%
  select(recorded_taxa_name) %>%
  filter(str_detect(recorded_taxa_name, regex("RANCHO", ignore_case = TRUE)))

# Longest-first ordering in the regex ensures: multi-word phrases match fully BEFORE matching component words.
# Component words will not appear separately in the extraction. So, for example, "VIVO P/ABASTO RANCHO" gets abstracted as "VIVO" and "ABASTO RANCHO" before "RANCHO" can be extracted.

# >> This is why my unique keyword vector has fewer elements than your list of 81 — some keywords are never extracted as standalone because they are “absorbed” into longer matches.
```

```{r SANITY CHECK 2: Did we properly keywords and modifiers from recorded_taxa_name_clean?}
# Look for "DE" or "P/" as a full word
DE_remaining <- GOC_Landings_parsed %>%
  filter(str_detect(recorded_taxa_name_clean, "\\bDE\\b"))
P_remaining <- GOC_Landings_parsed %>%
  filter(str_detect(recorded_taxa_name_clean, "\\bP\\b"))

# Count how many rows
nrow(DE_remaining)
nrow(P_remaining)
# Count = 0, good.

# Rows where "NACAR" appears anywhere (case-insensitive)
NACAR_remaining <- GOC_Landings_parsed %>%
  filter(str_detect(recorded_taxa_name_clean, regex("NACAR", ignore_case = TRUE)))

# Count how many rows
nrow(NACAR_remaining)
# Count = 0, good.

# Any keyword column containing NACAR (case-insensitive)
NACAR_as_keyword <- GOC_Landings_parsed %>%
  filter(
    str_detect(keywords_1, regex("NACAR", ignore_case = TRUE)) |
    str_detect(keywords_2, regex("NACAR", ignore_case = TRUE)) |
    str_detect(keywords_3, regex("NACAR", ignore_case = TRUE)) |
    str_detect(keywords_4, regex("NACAR", ignore_case = TRUE)) |
    str_detect(keywords_5, regex("NACAR", ignore_case = TRUE))
  )

# Count rows
nrow(NACAR_as_keyword)

# Optional: preview
head(NACAR_as_keyword %>% select(recorded_taxa_name, keywords_1:keywords_5))
# Looks great! 

# Remove these dataframes to keep workspace clear:
rm(DE_remaining, P_remaining, NACAR_remaining, NACAR_as_keyword)
```

```{r SANITY CHECK 3: Checking on the unique ALETA cases}
# QUICK SANITY CHECK: did this work for taxa like yellow-fin and blue-fin tuna?
test <- GOC_Landings_parsed %>%
  filter(str_detect(recorded_taxa_group, "ATUN"))
# 15374 obs.

table(test$common_name_clean)
#  Yes, parsing seems to have worked!

# Remove this dataframe:
rm(test)
```

# STEP 5c: Save the GOC_Landings_parsed dataframe as a separate .rds file - OPTIONAL:
```{r Save this dataframe - landings with keywords: Mar 23 2026}
colnames(GOC_Landings_parsed)
# Looks good! Go ahead and save.

# saveRDS(GOC_Landings_parsed, file="/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Landings_KeywordsParsed_2000.present_EMF.Apr022026.rds")
# 3622776 obs., 40 vars.
```

NOTE TO SELF: How did I get this list of key words originally? I think I started out with some preliminary version of the parse functionality and logic. From there, I think I quickly started to separate out a keywords list, the list grew over time, and I moved to this more sophisticated parse function. Come back to this at some point to recreate and / or document how one could accomplish this early part of the workflow. Document the original workflow or present a solution for the sake of reproducibility.


# > IF SKIPPING STEP 5 - parsing keywords, then move to: ------------------------------

# STEP 6: Merge the curated GOC_Landings_parsed (now ) with the most-up-to-date species key (MasterSpKey_pending). ------------------------------

## STEP 6a: Load in GOC_Landings_parsed dataframe (and rename to GOC_Landings_cleaned):
```{r Load GOC_Landings_parsed if needed}
GOC_Landings_parsed<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Landings_KeywordsParsed_2000.present_EMF.Apr022026.rds"))
# 3622776 obs., 40 vars.
```

```{r rename to GOC_Landings_cleaned}
GOC_Landings_species<- GOC_Landings_parsed

# Remove dataframes (as appropriate) to keep workspace clean:
rm(GOC_Landings_parsed)
rm(GOC_Landings)
```


```{r Load MasterSpKey_pending}
MasterSpKey_pending <- read_csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/Species_Keys/CleanedMaster_SpKey_Mar252026.csv")
# 807 obs of 41 variables
```

```{r Shave down the size of Master_SpKey_pending}
colnames(MasterSpKey_pending)

Master_SpKey_simple<- MasterSpKey_pending %>% 
  select(recorded_taxa_group, recorded_taxa_name_clean, common_name_clean,
         estimated_species_name, species_est_certainty, estimated_genus_name, estimated_family_name,
         IHH_order, IHH_class, IHH_functional_major_groups_sm, IHH_fishery)
# 807 obs., 11 vars.
```

## STEP 6b: Merge GOC_Landings_cleaned with Master_SpKey_simple:
```{r Merge GOC landings with Master_SpKey_simple}
GOC_Landings_species <- GOC_Landings_species %>%
  left_join(
    Master_SpKey_simple,
    by = c(
      "recorded_taxa_group",
      "recorded_taxa_name_clean",
      "common_name_clean"
    )
  )
# n = 3622776 obs. as before, now 48 vars.
```

```{r Trim GOC_Landings_species as needed}
colnames(GOC_Landings_species)

GOC_Landings_species <- GOC_Landings_species %>%
  select(
    # Location:
    state, office_name, landing_site, landing_site_key,
    est_lat, est_lon, est_latlon_FLAG, inside_GOC, est_habitat,
    
    # Time:
    year, month, receipt_date_fixed, period_start_fixed, period_end_fixed,
    
    # Fleet and vessel info:
    fleet, acuaculture_production, eu_rnpa, eu_name, vessel_rnpa, vessel_name,
    
    # Taxa, recorded and cleaned names:
    recorded_taxa_group, recorded_taxa_name, recorded_taxa_name_clean,
    recorded_scientific_name, scientific_name_recorded_T.F, common_name_clean,
    
    # Keywords:
    keywords_1, keywords_2, keywords_3, keywords_4, keywords_5,
    
    # Taxa details, estimated:
    estimated_species_name, species_est_certainty, estimated_genus_name, estimated_family_name,
    
    # Catch and value:
    landed_weight, live_weight, weight_kg_corrected, value,
    
    # OHI region details:
    OHI_rgn_number, OHI_region_draft, OHI_rgn_name, OHI_rgn_notes, exclude_rgn,
    
    # Classification (IHH):
    IHH_order, IHH_class, IHH_functional_major_groups_sm, IHH_fishery
  )
```


## STEP 6c: Save GOC_Landings_species (keywords parsed with species) as separate .rds file - OPTIONAL:
```{r Save this dataframe - GOC landings with keywords with species names as of Mar 11 2026}
# saveRDS(GOC_Landings_species, file="/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Landings_Parsed.wSpecies_2000.present_EMF.Apr022026.rds")
# n = 3622776 obs., 48 vars.
```


# STEP 7: Merge the GOC_Landings_species with the most up-to-date group key (MajorGroupKey_pending) to create GOC_Landings_classed. ------------------------------

# NOTE TO SELF: Species key is up-to-date and so is MajorGroupKey_pending as of - Mar 25 2026
To learn more about how this key was created, see: CONAPESCA_CreatingTaxonomicKeys_Mar132026.rmd

```{r Load in GOC data as needed}
GOC_Landings_species<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Landings_Parsed.wSpecies_2000.present_EMF.Apr022026.rds"))
# 3622776 obs. 48 vars.
```

```{r Load in revised MajorGroupKey_pending}
MajorGroupKey_pending<- read_csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/MajorGroupKey_Mar252026.csv")
# 526 obs. 13 vars.
```

```{r Complicated merge landings data with MajorGroupKey_pending}
Special_groups <- c("OTRAS", "ESPECIE GENERICA", "FAUNA", "ORNATO", "PECES DE ORNATO", "NEI")

#STEP 1: Join in non-OTRAS, simple join
# For non-otra taxa, we only want to look up based on the recorded_taxa_group and estimated_genus_name. From that look-up we want to assign Major_Functional_Group, Assignment_for_Major_Functional_Group
NONOTRAS_MajorGroupKey_lookup <- MajorGroupKey_pending %>%
  filter(!recorded_taxa_group %in% Special_groups) %>%
  select(-c("common_name_clean", "IHH_functional_major_groups_sm", "estimated_species_name", "Manual_Classification_Notes", "species_est_certainty", "species_est_reasoning", "genus_est_reasoning", "ornamentalFLAG_1")) %>%
  distinct(
    recorded_taxa_group,
    estimated_genus_name,
    .keep_all = TRUE
  )

# 219 obs. 5 vars.
  
join_NormalGroups <- GOC_Landings_species %>%
  filter(!recorded_taxa_group %in% Special_groups) %>%
  left_join(
    NONOTRAS_MajorGroupKey_lookup,
    by = c("recorded_taxa_group","estimated_genus_name")
  )
# 3101927 obs. (the majority of the data, as expected) and 51 vars (+3 new vars).

# SANITY CHECK
subset<- join_NormalGroups %>% filter(is.na(Major_Functional_Group))
table(subset$recorded_taxa_group)
# 0 obs. Great!

# STEP 2: JOIN IN OTRAS
# In contrast: for otras taxa, we want to look up based on the recorded_taxa_group and common_name_clean. From that look-up we want to assign estimated_taxa_group, estimated_species_name, species_est_certainty, species_est_reasoning, estimated_genus_name, genus_est_reasoning, Major_Functional_Group, Assignment_for_Major_Functional_Group, Manual_Classification_Notes, and ornamentalFLAG_1
OTRAS_MajorGroupKey_lookup <- MajorGroupKey_pending %>%
  filter(recorded_taxa_group %in% Special_groups) %>%
  distinct(
    recorded_taxa_group,
    common_name_clean,
    .keep_all = TRUE
  )
# 278 obs. 13 vars.

join_SpecialGroups <- GOC_Landings_species %>%
  filter(recorded_taxa_group %in% Special_groups) %>%
  left_join(
    OTRAS_MajorGroupKey_lookup,
    by = c("recorded_taxa_group", "common_name_clean"),
    suffix = c("", "_new")
  ) %>%
  mutate(
    across(
      ends_with("_new"),
      ~ .,
      .names = "{sub('_new$', '', .col)}"
    )
  )

join_SpecialGroups <- join_SpecialGroups %>%
  select(-ends_with("_new"))

# 520849 obs. (as expected) and 55 vars. (+7 new vars).

# STEP 3:
GOC_Landings_classed <- bind_rows(join_NormalGroups, join_SpecialGroups)
# 3622776 obs. (as expected) and 55 vars.
```

## STEP 7a: Assign new class for ANCHOVETA and SARDINA = "Small pelagic marine fishes"

```{r What if we assign and pull out small pelagic fishes?}
unique(GOC_Landings_classed$recorded_taxa_group)

# Small pelagics
Small_Pelagics<- c("ANCHOVETA", "SARDINA") 

GOC_Landings_classed <- GOC_Landings_classed %>%
  mutate(
    Major_Functional_Group = case_when(
      recorded_taxa_group %in% Small_Pelagics ~ "Anchovy & Sardine (small pelagics)",
      Major_Functional_Group == "Pelagic marine fishes" ~ "Pelagic marine fish (excl. anchovy & sardine)",
      TRUE ~ Major_Functional_Group
    )
  )
# 3622776 obs. (as expected) and 55 vars.

table(GOC_Landings_classed$Major_Functional_Group)

           # Anchovy & Sardine (small pelagics)                                   Cephalopods 
           #                              89647                                         88211 
           #                        Crustaceans                        Demersal marine fishes 
           #                             195881                                       1561744 
           #                  Diadromous fishes                                   Echinoderms 
           #                               7393                                         12075 
           #                  Freshwater fishes                                       Jellies 
           #                             242354                                         15978 
           #                       Marine algae                             Marine fishes NEI 
           #                               1723                                         13372 
           #    Marine or freshwater fishes NEI                            Marine species NEI 
           #                             134361                                         26719 
           #         Molluscs excl. Cephalopods                         Other freshwater taxa 
           #                             174768                                          2563 
           #         Other marine invertebrates Pelagic marine fish (excl. anchovy & sardine) 
           #                                 15                                        320186 
           #                             Sharks                               Shrimps, prawns 
           #                             171047                                        561207 
           #                            unknown 
           #                               3519 

# SANITY CHECK: Which records are captured by the "Marine or freshwater fishes NEI" category?
subset<- GOC_Landings_classed %>% 
  filter(Major_Functional_Group=="Marine or freshwater fishes NEI")
# 134361 obs. from 2000 to 2025 (makes sense).
min(subset$year)
max(subset$year)

table(subset$est_habitat)
table(subset$common_name_clean)
table(subset$estimated_taxa_group)
# Nice! These look like taxa that, as expected, could be freshwater or marine (BAGRE, CARPA, MOJARRA, TRUCHA). More information required to make an educated guess. 

  # BAGRE   CARPA MOJARRA  TRUCHA 
  # 36378     407   95130    2446 
# Mostly freshwater taxa.
```

# SANITY CHECK: How many records have we accounted for with MajorGroupKey_pending? ----
```{r SANITY CHECK: How many records have we accounted for with our MajorGroupKey_pending?}
table(GOC_Landings_classed$Major_Functional_Group)
subset<- GOC_Landings_classed %>% filter(is.na(Major_Functional_Group))
table(subset$recorded_taxa_group)
# OTRAS 
#    13 
# This shows that 13 out of all (n=3622776) entries are missing a Major_Functional_Group assignment;
# And, we know that out of all obs., 3519 are classified as "unknown" and 26719 "Marine species NEI" as their Major_Functional_Group.
```

```{r Summarize landing records and total biomass per Major_Functional_Group}
# Summarize landing records and total biomass per Major_Functional_Group, including NAs as "N/A"
# Ensure NAs are included in the factor
GOC_Landings_classed$Major_Functional_Group[is.na(GOC_Landings_classed$Major_Functional_Group)] <- "N/A"
GOC_Landings_classed$IHH_functional_major_groups_sm[is.na(GOC_Landings_classed$IHH_functional_major_groups_sm)] <- "N/A"

# Summarize REVISED landing records and biomass:
NEW.LandingsPercent_summary <- GOC_Landings_classed %>%
  # Convert NA to a level so it's counted
  mutate(Major_Functional_Group = ifelse(
    is.na(Major_Functional_Group),
    "NA",
    Major_Functional_Group
  )) %>%
  group_by(Major_Functional_Group) %>%
  summarise(
    NEW.Percent_TotLandingRecords = round(n() / nrow(GOC_Landings_classed) * 100, 1),
    NEW.Percent_TotalBiomass = round(
      sum(weight_kg_corrected, na.rm = TRUE) / sum(GOC_Landings_classed$weight_kg_corrected, na.rm = TRUE) * 100,
      1
    ),
    .groups = "drop"
  ) %>%
  # Sort descending by biomass
  arrange(desc(NEW.Percent_TotalBiomass))

NEW.LandingsPercent_summary
# This suggests that, from 2000 to 2025, < 1% of all landing records and < 0.05% of all biomass come from taxa that are NOT classified to the Major_Functional_Group level. The rest come from major functional groups like demersal fish and shrimp. I can live with this! 

# What did these numbers look like before based on IHH groupings?
# Build summary table with counts, percent of landing records, and percent of total biomass

OLD.LandingsPercent_summary <- GOC_Landings_classed %>%
  mutate(IHH_functional_major_groups_sm = ifelse(
    is.na(IHH_functional_major_groups_sm),
    "NA",
    IHH_functional_major_groups_sm
  )) %>%
  group_by(IHH_functional_major_groups_sm) %>%
  summarise(
    OLD.Percent_TotLandingRecords = round(n() / nrow(GOC_Landings_classed) * 100, 1),
    OLD.Percent_TotBiomass = round(
      sum(weight_kg_corrected, na.rm = TRUE) / sum(GOC_Landings_classed$weight_kg_corrected, na.rm = TRUE) * 100, 
      1
    ),
    .groups = "drop"
  )

# Join these two tables to compare.
library(fuzzyjoin)

# Perform fuzzy join using string distance
MERGED.LandingsPercent_summary <- stringdist_left_join(
  NEW.LandingsPercent_summary,
  OLD.LandingsPercent_summary, 
  by = c("Major_Functional_Group" = "IHH_functional_major_groups_sm"),
  method = "jw",           # Jaro-Winkler distance (good for names)
  max_dist = 0.2,          # Adjust threshold as needed (0 = exact, ~0.2 = close match)
  distance_col = "dist"    # Optional: keep the match distance
) %>%
  arrange(dist)             # Optional: sort by closest match


# Reorder columns
MERGED.LandingsPercent_summary <- MERGED.LandingsPercent_summary %>%
  select(
    Major_Functional_Group,
    OLD.Percent_TotBiomass,
    NEW.Percent_TotalBiomass,
    OLD.Percent_TotLandingRecords,
    NEW.Percent_TotLandingRecords
  ) %>%
  # Order rows:
  arrange(
    # Then by OLD percent landing records descending
    desc(OLD.Percent_TotBiomass)
  )

# Save this dataframe:
# write.csv(MERGED.LandingsPercent_summary, "/home/ferrer/ohi_goc_prep/FP/CONAPESCA_CleanUp/Catch_byTaxaGroup_summarytable_Apr22026.csv", row.names = FALSE)
```

## STEP 7b: Save GOC_Landings_classed (keywords parsed with species names and major taxanomic group) as separate .rds file - OPTIONAL:
```{r Save this dataframe - GOC landings with keywords with species names and major taxonomic group as of Mar 25 2026}
# saveRDS(GOC_Landings_classed, file="/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Landings_Parsed.wSpecies&MajorGroups_2000.present_EMF.Apr022026.rds")
# 3622776 obs., 55 vars.
```

```{r Remove dataframes to keep workspace clean}
rm(join_NormalGroups, join_SpecialGroups,
   MERGED.LandingsPercent_summary, NEW.LandingsPercent_summary, OLD.LandingsPercent_summary,
   NONOTRAS_MajorGroupKey_lookup, OTRAS_MajorGroupKey_lookup)
```

# STEP 8: Take a look into key taxonomic groups. ------------------------------
```{r Load dataframe if needed}
GOC_Landings_classed<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Landings_Parsed.wSpecies&MajorGroups_2000.present_EMF.Apr022026.rds"))
# 3622776 obs. 55 vars.

subset<- GOC_Landings_classed %>% filter(recorded_taxa_group == "PETO")
# Note, PETO is definitely a fish but not clear if its pelagic or demersal fish; it also identified in Hem's key as a shark, which I think is a mis-ID. Currently catalogued as demersal. Come back to this at some point.
```

## STEP 8a: Let's look into ALL groups from GOC_Landings_classed:
```{r Plot catch by major_functional_group over time}
# SUMARRY TABLE:
summary_MajorFunctionalGroup <- GOC_Landings_classed %>%
  group_by(year, Major_Functional_Group) %>%
  summarise(
    total_catch_kg = sum(weight_kg_corrected, na.rm = TRUE),
    .groups = "drop"
  )

unique(summary_MajorFunctionalGroup$Major_Functional_Group)

Groups_to_Exclude<- c("Shrimps, prawns", "Anchovy & Sardine (small pelagics)", "unknown", "N/A")

Catch_byGroup_plot<- summary_MajorFunctionalGroup %>%
  filter(year < 2025) %>%
  filter(!Major_Functional_Group %in% Groups_to_Exclude) %>%
ggplot(aes(x = year, y = total_catch_kg, fill = Major_Functional_Group)) +
  geom_area(alpha = 0.9) +
  scale_fill_viridis_d(option = "turbo") +
  labs(
    x = "Year",
    y = "Catch (kg)",
    fill = "Functional Group",
    title = "Catch by Major Functional Group Over Time"
  ) +
  theme_minimal(base_size = 14)

Catch_byGroup_plot
```

```{r Save Catch_byGroup_plot}
# Save the plot as a high-resolution PNG
ggsave(
  filename = "/home/ferrer/ohi_goc_prep/FP/CONAPESCA_CleanUp/Catch_byGroup_plot_Apr022026.png",
  plot = Catch_byGroup_plot,
  width = 10,       # width in inches
  height = 6,       # height in inches
  dpi = 300,        # resolution
  bg = "white"      # ensures white background
)
```

## STEP 8b: Pull out small pelagics (anchovy and sardine) and plot:
```{r Pull out small pelagic fishes anchovy and sardine}
table(GOC_Landings_classed$Major_Functional_Group)

# Plot major groups excluding catch from small pelagic fishes, freshwater taxa, and unknown taxa:
Groups_to_Exclude<- c("Anchovy & Sardine (small pelagics)", "Freshwater fishes", "Diadromous fishes", "Other freshwater taxa", "Marine or freshwater fishes NEI",  NA, "unknown")

# level the Major_Functional_Groups so they plot correctly:
summary_MajorFunctionalGroup$Major_Functional_Group <- factor(
  summary_MajorFunctionalGroup$Major_Functional_Group,
  levels = c(

  # Mostly freshwater:
    "Marine or freshwater fishes NEI",
    "Diadromous fishes",
    "Freshwater fishes",
    "Other freshwater taxa",
  
  # (Marine) misc:
    "Marine algae",
    "Marine species NEI",
    "unknown",
    
  # Marine inverts:
    "Shrimps, prawns",
    "Crustaceans",
    "Molluscs excl. Cephalopods",
    "Cephalopods",
    "Jellies",
    "Echinoderms",
    "Other marine invertebrates",
    
  # Marine fish:
    "Marine fishes NEI",
    "Anchovy & Sardine (small pelagics)",
    "Pelagic marine fish (excl. anchovy & sardine)",
    "Sharks",
    "Demersal marine fishes"
  )
)

Catch_woSmallPelagics_plot<- summary_MajorFunctionalGroup %>%
  filter(year>= 2018 & year < 2025) %>%
  filter(!Major_Functional_Group %in% Groups_to_Exclude) %>%
ggplot(aes(x = year, y = total_catch_kg, fill = Major_Functional_Group)) +
  geom_area(alpha = 0.9) +
  scale_fill_viridis_d(option = "turbo") +
  labs(
    x = "Year",
    y = "Catch (kg)",
    fill = "Functional Group",
    title = "Catch by Major Functional Group Over Time",
    subtitle = "Excluding small pelagic fish, suspected freshwater, and unknown taxa"
  ) +
  theme_minimal(base_size = 14)

Catch_woSmallPelagics_plot
```

```{r Save Catch_byGroup_plot}
# Save the plot as a high-resolution PNG
ggsave(
  filename = "/home/ferrer/ohi_goc_prep/FP/CONAPESCA_CleanUp/Catch_woSmallPelagics_plot_Apr022026.png",
  plot = Catch_woSmallPelagics_plot,
  width = 10,       # width in inches
  height = 6,       # height in inches
  dpi = 300,        # resolution
  bg = "white"      # ensures white background
)
```

## STEP 8c: Let's look into jellies:
```{r Isolate jellies}
# BEFORE THE FIX (on Mar 13):
# Something weird is happening with the merge. The capitalization is off, etc. Common name clean is not quite right, it's not lowercase so the merge doesn't work and we can't locate jellyfish.
     #           Cephalopods                Crustaceans     Demersal marine fishes          Diadromous fishes 
     #                 88211                     195861                    1583704                       7393 
     #           Echinoderms          Freshwater fishes               Marine algae          Marine fishes NEI 
     #                 12072                      70644                       1723                     326723 
     #    Marine species NEI Molluscs excl. Cephalopods      Other freshwater taxa Other marine invertebrates 
     #                 26719                     153540                       2562                         15 
     # Pelagic marine fishes                     Sharks            Shrimps, prawns                    unknown 
     #                379493                     171073                     561207                       3519 

# Create a vector of keywords
Jellyfish_keywords <- c("bola", "cañon", "canon", "aguamala", "medusa", "stomolophus")

# Filter using regex with ignore case
GOC_Landings_jellies <- GOC_Landings_classed %>%
  filter(grepl(paste(Jellyfish_keywords, collapse = "|"), common_name_clean, ignore.case = TRUE))
# 15978 obs. Found them!

GOC_Landings_jellies <- GOC_Landings_classed %>%
  filter(Major_Functional_Group== "Jellies")
# 15978 obs. Yes, this works as expected!

table(GOC_Landings_jellies$recorded_taxa_group)
table(GOC_Landings_jellies$recorded_taxa_name_clean)

table(GOC_Landings_classed$Major_Functional_Group)

# AFTER THE FIX (Mar 24 2026): Looks good!

#            Anchovy & Sardine (small pelagics)                                   Cephalopods                                   Crustaceans 
#                                         89647                                         88211                                        195868 
#                        Demersal marine fishes                             Diadromous fishes                                   Echinoderms 
#                                       1561183                                          7393                                         12075 
#                             Freshwater fishes                                       Jellies                                  Marine algae 
#                                        240988                                         15978                                          1723 
#                             Marine fishes NEI               Marine or freshwater fishes NEI                            Marine species NEI 
#                                         13372                                        133614                                         26719 
#                    Molluscs excl. Cephalopods                         Other freshwater taxa                    Other marine invertebrates 
#                                        172866                                          2563                                            15 
# Pelagic marine fish (excl. anchovy & sardine)                                        Sharks                               Shrimps, prawns 
#                                        319873                                        166561                                        560968 
#                                       unknown 
#                                          3519 
```

```{r Plot jellies over time}
summary_Jellies <- GOC_Landings_jellies %>%
  group_by(year) %>%
  summarise(
    total_landings_kg = sum(weight_kg_corrected, na.rm = TRUE),
    n_records = n(),
    .groups = "drop"
  )

ggplot(summary_Jellies, aes(x = year, y = total_landings_kg)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Jellyfish Landings Over Time",
    x = "Year",
    y = "Total Landed Weight (kg)"
  ) +
  theme_bw()
```

```{r SANITY CHECK into Molluscs excl. Cephalopods}
subset<- GOC_Landings_classed %>% 
  filter(Major_Functional_Group=="Molluscs excl. Cephalopods")
# 172866 obs.

table(subset$estimated_species_name)
table(subset$estimated_genus_name)
table(subset$recorded_taxa_group)
# These all look good, except for otras. What are otras in this case?

subset<- subset %>% 
  filter(recorded_taxa_group=="OTRAS")
# 21274 obs.

table(subset$recorded_taxa_name_clean)
# Okay, looks good.
```


## STEP 8b: Let's look into ornamentals:
```{r Ornamentals}
table(GOC_Landings_classed$ornamentalFLAG_1)
   #    NO POSSIBLY      YES 
   # 22119    20521     1634 
  # And a lot of NAs

summary_Ornamentals <- GOC_Landings_classed %>%
  filter(ornamentalFLAG_1=="YES") %>%
  group_by(year) %>%
  summarise(
    total_landings_kg = sum(weight_kg_corrected, na.rm = TRUE),
    n_records = n(),
    .groups = "drop"
  )

ggplot(summary_Ornamentals, aes(x = year, y = total_landings_kg)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Ornamentals - Landings Over Time (DRAFT)",
    x = "Year",
    y = "Total Landed Weight (kg)"
  ) +
  theme_bw()
```

```{r SANITY CHECK: Check in on NAs}
table(GOC_Landings_classed$Major_Functional_Group)
# 2965 "NA"s (and 3519 "unknown")

# Let's take a look at those NAs
subset<- GOC_Landings_classed %>%
  filter(Major_Functional_Group=="N/A")
# 2965 obs.

table(subset$recorded_taxa_group)
         # LOBINA           OTRAS PECES DE ORNATO 
         #   1063            1895               7 
```
# > SHIFTING GEARS: Work on estimating sector  ------------------------------

# STEP 9: FLAG and estimate landings from AQUACULTURE based on a series of logical rules. ------------------------------
I think aquaculture and mariculture in the region is being vastly under-counted and under-reported.
Let's implement a series of rules to estimate if records are associated with aquaculture or not.

```{r Load in dataframe if needed}
GOC_Landings_classed0<- GOC_Landings_classed

GOC_Landings_classed0<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Landings_Parsed.wSpecies&MajorGroups_2000.present_EMF.Apr022026.rds"))
# 3622776 obs. 55 vars.

colnames(GOC_Landings_classed0)
```

## STEP 9a: FLAG AQUACULTURE:

```{r Create a GOC_Landings_aquaculture dataframe}
GOC_Landings_aquaculture<- GOC_Landings_classed0

GOC_Landings_aquaculture$est_aquaculture_production<- c(NA)
# n = 3622776 obs., 56 vars.
```

## VIEW: ALL KEY WORDS... Does this list need more work?
```{r Investigating Keywords}
# Extract all unique keywords from keyword columns 1 through 5. Delete exact duplicates. 
Keyword_vector <- GOC_Landings_classed %>%
  select(keywords_1:keywords_5) %>%
  unlist() %>%
  unique() %>%
  na.omit()

Keyword_vector
```

First, let's flag landing_site names and eu_name that are associated with aquaculture.
```{r Estimate likelihood of aquaculture based on name of landing_site and eu_name}
#### FLAG 1 - Flag aquaculture based on landing_site and eu_name (3 levels: YES, NO, POSSIBLY): ####
vector<- GOC_Landings_aquaculture %>%
  filter(year > 2024) %>%
  select(eu_name) %>%
  distinct() %>%
  slice_sample(n = 200)

# Clues from landing_site and eu_name
LandingSite_strong.clues<- c("CULTIVADORES", "ESTANQUE", "VIVEROS")
LandingSite_clues<- c("CULT", "DREN", "BORDO", "PREDIO")

EUname_clues<- c("ACUICOLA", "FARMS", "MARICULTURA",
                        "MARICULTIVOS",
                        "AGROACUICOLA", "AGRICOLA",
                        "ACUACULTURA", "AQUACULTURE",
                        "GRANJA", "GRANJA CAMARONERA",
                        "RANCHO")

# Build regex patterns dynamically
LandingSite_strong_pattern <- paste0("\\b(", paste(LandingSite_strong.clues, collapse = "|"), ")\\b")
LandingSite_weak_pattern <- paste0("\\b(", paste(LandingSite_clues, collapse = "|"), ")\\b")
EUname_pattern <- paste0("\\b(", paste(EUname_clues, collapse = "|"), ")\\b")

GOC_Landings_aquaculture <- GOC_Landings_aquaculture %>%
  
  mutate(
    acuaculture_production_norm = str_to_lower(str_trim(acuaculture_production))
  ) %>%
  
  mutate(
    aquacultureFLAG_site.details = case_when(
      str_detect(landing_site, regex(LandingSite_strong_pattern, ignore_case = TRUE)) ~ "YES",
      str_detect(landing_site, regex(LandingSite_weak_pattern, ignore_case = TRUE)) ~ "POSSIBLY",
      str_detect(eu_name, regex(EUname_pattern, ignore_case = TRUE)) ~ "POSSIBLY",
      TRUE ~ "NO"
    )
  ) %>%
  
  select(-acuaculture_production_norm)

# SANITY CHECK:
# Check that we pulled out landing sites that make sense:
table(GOC_Landings_aquaculture$aquacultureFLAG_site.details)

 #      NO POSSIBLY      YES 
 # 3606674    13876     2226 

# AFTER MY FIX ON APR 3 2026: 
 #      NO POSSIBLY      YES 
 # 3497614   125023      139
 # Right off the bat this seems good. Fewer NO, more possibly, less dirrect YES. Don't want to be OVERLY CONFIDENT in this estimate.

subset<- GOC_Landings_aquaculture %>%
  filter(aquacultureFLAG_site.details %in% c(
    "YES",
    "POSSIBLY"
  )) %>%
  distinct(landing_site, eu_name, aquacultureFLAG_site.details) %>%
  arrange(landing_site, eu_name)
# 1406 sites that mostly all make sense to me! If an organism came from one of these sites I would assume it was a farmed species, even if the record was NOT marked as aquaculture.

# Let's review these sites manually:
# write.csv(
#   GOC_Landings_aquaculture %>%
#     filter(est_aquaculture_production != "uncertain") %>%
#     distinct(landing_site, acuaculture_production, est_aquaculture_production, est_aquaculture_certainty, aquacultureFLAG_site.details),
#   "/home/ferrer/ohi_goc_prep/FP/Exploratory_Research/MAR_ExploratoryResearch/AquacultureSite_ClassificationCheck_Mar122026.csv",
#   row.names = FALSE
# )
# After a manual review, this logic seems okay for now... let's see how this affects of estimates of aquaculture.
```

Next, let's implement rules based on key words and taxa. What taxa are likely aquaculture?
```{r Flagging aquaculture landings based on key words}
colnames(GOC_Landings_aquaculture)

table(GOC_Landings_aquaculture$recorded_taxa_group)
table(GOC_Landings_aquaculture$estimated_taxa_group)

#### FLAG 2 - Flag aquaculture based on key words, Aquaculture_keywords [LIST IS PENDING REVIEW / UNDER CONSTRUCTION] ####
Aquaculture_keywords<- c(
  "DE CULTIVO", "DE CULTIVO VIVO", "CULTIVADAS", "ABASTO RANCHO", "RANCHO", "PERLAS", "MEDIAS PERLAS"
)

# Flag rows with aquaculture keywords:
GOC_Landings_aquaculture <- GOC_Landings_aquaculture %>%
mutate(
  aquacultureFLAG_keywords = case_when(
    rowSums(across(keywords_1:keywords_5, ~ str_to_upper(str_trim(.x)) %in% Aquaculture_keywords), na.rm = TRUE) > 0 ~ "YES",
    TRUE ~ "NO"
  )
)

table(GOC_Landings_aquaculture$aquacultureFLAG_keywords)
# 111955 observations explicitly flagged as aquaculture due to key words.
```

```{r Flag certain aquaculture and NOT-aquaculture taxa}
#### FLAG 3 - Flag known, likely, and NOT aquaculture taxa [LIST IS PENDING REVIEW / UNDER CONSTRUCTION] ####
# In theory, names could come from a number of columns: estimated_taxa_group, recorded_taxa_group, recorded_taxa_name, common_name_clean

table(GOC_Landings_aquaculture$estimated_taxa_group)
table(GOC_Landings_aquaculture$common_name_clean)

Possible_AquacultureTaxa <- c("OSTION", "JUREL", "CAMARON", "CARPA", "MOJARRA", "TRUCHA", "ALMEJA", "ABULON", "ALGAS") # estimated_taxa_group
Definite_AquacultureTaxa <- c("TILAPIA") # common_name_clean
ProbablyNot_AquacultureGroups <- c("Anchovy & Sardine (small pelagics)", "Cephalopods", "Echinoderms") # Major_Functional_Group

GOC_Landings_aquaculture <- GOC_Landings_aquaculture %>%
  mutate(
    aquacultureFLAG_specific.taxa = case_when(
      str_to_upper(str_trim(common_name_clean)) %in% Definite_AquacultureTaxa ~ "YES",
      str_to_upper(str_trim(estimated_taxa_group)) %in% Possible_AquacultureTaxa ~ "POSSIBLY",
      str_trim(Major_Functional_Group) %in% ProbablyNot_AquacultureGroups ~ "PROBABLY NOT",
      TRUE ~ "NO"
    )
  )

table(GOC_Landings_aquaculture$aquacultureFLAG_specific.taxa)

# Numbers BEFORE revision on Mar 24 2026:
#   NO          POSSIBLY    YES 
#   3408840     41066       172870 

# Numbers AFTER revision on Mar 25 2026:
#      NO POSSIBLY      YES 
# 2561375   888531   172870 

# Numbers AFTER revision on April 3 2026:
     #      NO     POSSIBLY PROBABLY NOT          YES 
     # 2371442       888531       189933       172870 
     
# Just as many yeses (makes sense) and now more possibles. Now some things marked as probably not.
```

## Flag what CANNOT be aquaculture based on what we know about the taxa
```{r Flag Major_Functional_Groups that are definitely NOT aquaculture like sharks}
#### FLAG 4 - Flag Major_Functional_Groups that are DEFINITELY not aquaculture [LIST IS PENDING REVIEW / UNDER CONSTRUCTION] ####

table(GOC_Landings_aquaculture$estimated_taxa_group)
table(GOC_Landings_aquaculture$Major_Functional_Group)

# DEFINITELY NOT:
CannotBe_AquacultureGroups <- c("Jellies", "Sharks") # Major_Functional_Groups
CannotBe_AquacultureGenera <- c("Istiophorus", "Makaira", "Tetrapturus", "Xiphias", "Acanthocybium", "Dosidicus") # estimated_genus_name

GOC_Landings_aquaculture <- GOC_Landings_aquaculture %>%
  mutate(
    NOTaquacultureFLAG_specific.groups = case_when(
      str_trim(Major_Functional_Group) %in% CannotBe_AquacultureGroups ~ "CANNOT BE AQUACULTURE",
      str_trim(estimated_genus_name) %in% CannotBe_AquacultureGenera ~ "CANNOT BE AQUACULTURE",
      TRUE ~ NA
    )
  )

table(GOC_Landings_aquaculture$NOTaquacultureFLAG_specific.groups)

# Numbers AFTER revision on April 3 2026:
# CANNOT BE AQUACULTURE 
#                240202
```

Create a combined aquaculture flag based on the previous three flags (aquacultureFLAG_site.details, aquacultureFLAG_keywords, aquacultureFLAG_specific.taxa) and the new flag NOTaquacultureFLAG_specific.groups

Here's the logic for the combined flag:

• If at least one flag is "YES", then flags indicate that aquaculture is highly likely and aquacultureFLAG_combined = "YES" (my YES flags are pretty robust at this point)
• If one more flags is "POSSIBLY", then aquacultureFLAG_combined is "possibly aquaculture" but not certain
• If no flags indicate aquaculture, then aquacultureFLAG_combined is "NO"

• BUT, if the NOTaquacultureFLAG_specific.groups indicates CANNOT BE AQUACULTURE, then combined flag is "NO"


```{r Create a combined aquaculture flag}
#### FLAG 5 - Create a combined aquaculture flag based on the previous four flags ####
GOC_Landings_aquaculture <- GOC_Landings_aquaculture %>%
  
  # Remember, put the strongest condition first:
  mutate(
    aquacultureFLAG_combined = case_when(
      
      if_any(c(NOTaquacultureFLAG_specific.groups), ~ .x == "CANNOT BE AQUACULTURE") ~ "CANNOT BE AQUACULTURE",
      
      if_any(c(aquacultureFLAG_specific.taxa), ~ .x == "PROBABLY NOT") ~ "PROBABLY NOT",
      
      if_any(c(aquacultureFLAG_site.details,
               aquacultureFLAG_keywords,
               aquacultureFLAG_specific.taxa), ~ .x == "YES") ~ "YES",
      
      if_any(c(aquacultureFLAG_site.details,
               aquacultureFLAG_keywords,
               aquacultureFLAG_specific.taxa), ~ .x == "POSSIBLY") ~ "POSSIBLY",
      
      TRUE ~ "NO"
    )
  )

# SANITY CHECK:
subset<- GOC_Landings_aquaculture %>% 
  filter(aquacultureFLAG_site.details %in% c("YES", "POSSIBLY") & aquacultureFLAG_combined=="CANNOT BE AQUACULTURE")
table(subset$estimated_taxa_group)
# > 3647 where site names / details are incongruous with the taxa being caught. For example, giant squid and sharks are not used for aquaculture.
# Good! Sign that our aquacultureFLAG_combined is working and trumping other clues.
# Maybe these are taxa ASSOCIATED with aquaculture, or found new aquaculture sites but not actually aquaculture. :)
# Apr 3 2026.
```

## STEP 9b: ESTIMATE AQUACULTURE:
Here is the logic and set of decisions designed to estimate aquaculture:

Decision Table:
Report      aquacultureFLAG_combined  aquaculture_estimate    certainty
  yes       YES                       aquaculture             high; aquaculture flags agree with original report
  yes       POSSIBLY                  aquaculture             high; aquaculture flags agree with original report
  yes       NO                        aquaculture	            medium; no flags for aquaculture but original report says aquaculture
  yes       PROBABLY NOT              possibly aquaculture    low; aquaculture flags do not agree with original report
  yes       CANNOT BE AQUACULTURE     not aquaculture         medium; aquaculture flags suggest cannot be aquaculture, original report likely wrong
  
  no        YES	                      likely aquaculture	    medium; flags suggest aquaculture but original report says no, report could be wrong
  no        POSSIBLY                  possibly aquaculture    low; flags suggest possibly aquaculture but original report says no
  no        NO                        not aquaculture         high; aquaculture flags agree with original report
  no        PROBABLY NOT              not aquaculture         high; aquaculture flags agree with original report
  no        CANNOT BE AQUACULTURE     not aquaculture         high; aquaculture flags agree with original report
  
  NA        YES                       likely aquaculture      medium; no data from original report but flags suggest most likely aquaculture
  NA        POSSIBLY                  possibly aquaculture    low; no data from original report but flags suggest possibly aquaculture
  NA        NO                        unlikely aquaculture    medium; no data from original report and flags do not suggest aquaculture
  NA        PROBABLY NOT              unlikely aquaculture    high; no data from original report and flags suggest probably not aquaculture
  NA        CANNOT BE AQUACULTURE     not aquaculture         high; no data from original report and flags suggest definitely not aquaculture

```{r Generate estimates and confidence values}
#### Estimate aquaculture and confidence values based on the comparison of aquacultureFLAG_combined versus the official report: ####
GOC_Landings_aquaculture <- GOC_Landings_aquaculture %>%
  mutate(
    # Normalize original report:
    acuaculture_original.report = str_to_lower(str_trim(acuaculture_production)),

    # ESTIMATE OF AQUACULTURE:
    est_aquaculture_production = case_when(
      
      # Report = YES
      acuaculture_original.report == "yes" & aquacultureFLAG_combined %in% c("YES", "POSSIBLY", "NO") ~ "aquaculture",
      acuaculture_original.report == "yes" & aquacultureFLAG_combined == "PROBABLY NOT" ~ "possibly aquaculture",
      acuaculture_original.report == "yes" & aquacultureFLAG_combined == "CANNOT BE AQUACULTURE" ~ "not aquaculture",
      
      # Report = NO
      acuaculture_original.report == "no" & aquacultureFLAG_combined == "YES" ~ "likely aquaculture",
      acuaculture_original.report == "no" & aquacultureFLAG_combined == "POSSIBLY" ~ "possibly aquaculture",
      acuaculture_original.report == "no" & aquacultureFLAG_combined %in% c("NO", "PROBABLY NOT", "CANNOT BE AQUACULTURE") ~ "not aquaculture",
      
      # Report = NA
      is.na(acuaculture_original.report) & aquacultureFLAG_combined == "YES" ~ "likely aquaculture",
      is.na(acuaculture_original.report) & aquacultureFLAG_combined == "POSSIBLY" ~ "possibly aquaculture",
      is.na(acuaculture_original.report) & aquacultureFLAG_combined == "NO" ~ "unlikely aquaculture",
      is.na(acuaculture_original.report) & aquacultureFLAG_combined == "PROBABLY NOT" ~ "unlikely aquaculture",
      is.na(acuaculture_original.report) & aquacultureFLAG_combined == "CANNOT BE AQUACULTURE" ~ "not aquaculture",
      
      TRUE ~ "uncertain"
    ),

    # CERTAINTY ESTIMATES:
    est_aquaculture_certainty = case_when(
      
      # Report = YES
      acuaculture_original.report == "yes" & aquacultureFLAG_combined %in% c("YES", "POSSIBLY") ~ "high; aquaculture flags agree with original report",
      acuaculture_original.report == "yes" & aquacultureFLAG_combined == "NO" ~ "medium; no flags for aquaculture but original report says aquaculture",
      acuaculture_original.report == "yes" & aquacultureFLAG_combined == "PROBABLY NOT" ~ "low; aquaculture flags do not agree with original report",
      acuaculture_original.report == "yes" & aquacultureFLAG_combined == "CANNOT BE AQUACULTURE" ~ "medium; aquaculture flags suggest cannot be aquaculture, original report likely wrong",
      
      # Report = NO
      acuaculture_original.report == "no" & aquacultureFLAG_combined == "YES" ~ "medium; flags suggest aquaculture but original report says no, report could be wrong",
      acuaculture_original.report == "no" & aquacultureFLAG_combined == "POSSIBLY" ~ "low; flags suggest possibly aquaculture but original report says no",
      acuaculture_original.report == "no" & aquacultureFLAG_combined %in% c("NO", "PROBABLY NOT", "CANNOT BE AQUACULTURE") ~ "high; aquaculture flags agree with original report",
      
      # Report = NA
      is.na(acuaculture_original.report) & aquacultureFLAG_combined == "YES" ~ "medium; no data from original report but flags suggest most likely aquaculture",
      is.na(acuaculture_original.report) & aquacultureFLAG_combined == "POSSIBLY" ~ "low; no data from original report but flags suggest possibly aquaculture",
      is.na(acuaculture_original.report) & aquacultureFLAG_combined == "NO" ~ "medium; no data from original report and flags do not suggest aquaculture",
      is.na(acuaculture_original.report) & aquacultureFLAG_combined == "PROBABLY NOT" ~ "high; no data from original report and flags suggest probably not aquaculture",
      is.na(acuaculture_original.report) & aquacultureFLAG_combined == "CANNOT BE AQUACULTURE" ~ "high; no data from original report and flags suggest definitely not aquaculture",
      
      TRUE ~ "not enough information to assign aquaculture estimate"
    )
  ) %>%
  select(-acuaculture_original.report)

# 3622776 obs., 62 vars.

# SANITY CHECK:
table(GOC_Landings_aquaculture$est_aquaculture_production)

# Numbers BEFORE revision on Mar 24 2026:
         # aquaculture   likely aquaculture      not aquaculture possibly aquaculture unlikely aquaculture 
         #       17496               281612              2622445                37649               663574 

# Numbers AFTER revision on Mar 24 2026:
         # aquaculture   likely aquaculture      not aquaculture possibly aquaculture unlikely aquaculture 
         #       17496               281620              2141156               791707               390797 

# Numbers AFTER revisions on Apr 3 2026:
         # aquaculture   likely aquaculture      not aquaculture possibly aquaculture unlikely aquaculture 
         #       16469               279574              2186699               802339               337695 
               
subset<- GOC_Landings_aquaculture %>% filter(acuaculture_production=="yes" & est_aquaculture_production =="not aquaculture")
table(subset$estimated_taxa_group) 
# Excellent! Most of the aquaculture explicitly reported as aquaculture is estimated to be aquaculture.
# But the ones that were reported as aquaculture but estimated to not be aquaculture make sense! Shark and squid:
# CALAMAR   CAZON TIBURON 
#     220     141     242 

min(subset$year)
max(subset$year)
# Very weird, the official aquaculture reports only go from 2006 to 2017

subset<- GOC_Landings_aquaculture %>% filter(est_aquaculture_production==c("aquaculture", "likely aquaculture"))
# 147967 obs. The data set grows a lot!
```

```{r Re-level est_aquaculture_production column}
# Re-level est_aquaculture_production to be a little more logical:
GOC_Landings_aquaculture <- GOC_Landings_aquaculture %>%
  mutate(
    est_aquaculture_production = factor(
      est_aquaculture_production,
      levels = c(
        "aquaculture",
        "likely aquaculture",
        "possibly aquaculture",
        "unlikely aquaculture",
        "not aquaculture"
      )
    )
  )

# 3622776 obs., 62 vars.
```

```{r Summarize reported aquaculture versus estimated aquaculture}
table(GOC_Landings_aquaculture$acuaculture_production)
table(GOC_Landings_aquaculture$est_aquaculture_production)

         # aquaculture   likely aquaculture possibly aquaculture unlikely aquaculture      not aquaculture 
         #       16469               279574               802339               337695              2186699 

Reported.Aquaculture_summary<- GOC_Landings_aquaculture %>% 
  filter(acuaculture_production== c("yes")) %>%
  filter(year >= 2014, year <= 2024) %>%
  select(year, weight_kg_corrected, value) %>%
  group_by(year) %>%
  summarize(reported_annual_catch= sum(weight_kg_corrected, na.rm = TRUE),
            reported_annual_value= sum(value, na.rm = TRUE))

Estimated.Aquaculture_summary<- GOC_Landings_aquaculture %>% 
  filter(est_aquaculture_production== c("aquaculture", "likely aquaculture")) %>%
  filter(year >= 2014, year <= 2024) %>%
  select(year, weight_kg_corrected, value) %>%
  group_by(year) %>%
  summarize(est_annual_catch= sum(weight_kg_corrected, na.rm = TRUE),
            est_annual_value= sum(value, na.rm = TRUE))

Aquaculture_summary<- Estimated.Aquaculture_summary %>%
  left_join(Reported.Aquaculture_summary, by = "year") %>%
  select(year, reported_annual_catch, est_annual_catch, reported_annual_value, est_annual_value)
```

```{r Plot reported aquaculture versus estimated aquaculture}
EstvReported_Aquaculture_plot<- ggplot(Aquaculture_summary, aes(x = year)) +
  geom_line(aes(y = est_annual_catch, color = "Estimated catch"), linewidth = 1.2) +
  geom_point(aes(y = est_annual_catch, color = "Estimated catch"), size = 2) +
  
  geom_line(aes(y = reported_annual_catch, color = "Reported catch"), linewidth = 1.2) +
  geom_point(aes(y = reported_annual_catch, color = "Reported catch"), size = 2) +
  
  scale_color_manual(
    values = c(
      "Estimated catch" = "#1b9e77",
      "Reported catch" = "#d95f02"
    )
  ) +
  labs(
    x = "Year",
    y = "Catch (kg)",
    color = "",
    title = "Estimated vs Reported Aquaculture"
  ) +
  theme_minimal(base_size = 14) +
  scale_x_continuous(breaks = Aquaculture_summary$year) # ensures every year is a tick

EstvReported_Aquaculture_plot
```

```{r Save EstvReported_Aquaculture_plot}
# Save the plot as a high-resolution PNG
ggsave(
  filename = "/home/ferrer/ohi_goc_prep/FP/CONAPESCA_CleanUp/EstvReported_Aquaculture_plot_Apr032026.png",
  plot = EstvReported_Aquaculture_plot,
  width = 10,       # width in inches
  height = 6,       # height in inches
  dpi = 300,        # resolution
  bg = "white"      # ensures white background
)
```
Notice the spike in landings from 2017-2018? Of course, how could you miss it.
This jump coincides (temporally) with a change in reporting associated with the raw data itself. 
Data pre-2018 is reported here: mex_conapesca_avisos_2000_2019.rds
Data 2018 and onward is reported here: mex_conapesca_apertura_2018_present.rds
> NOTE TO SELF: Consult with Juan Carlos about these two datasets. What is fundamentally different about them?


# STEP 10: FLAG and estimate landings from FRESHWATER based on a series of logical rules. ------------------------------
```{r Flag freshwater landings}
table(GOC_Landings_aquaculture$Major_Functional_Group)

# Possible freshwater fishes:
subset<- GOC_Landings_aquaculture %>%
  filter(Major_Functional_Group== "Diadromous fishes")
table(subset$estimated_genus_name)
# Makes sense: Anguilla, Chanos

#### FLAG 6 - Flag freshwater based on keywords, Freshwater_keywords [LIST IS PENDING REVIEW / UNDER CONSTRUCTION] ####

Freshwater_keywords<- c(
  "DE AGUA DULCE"
)

#### FLAG 7A - Flag freshwater based on key taxa, Known_FreshwaterTaxa [LIST IS PENDING REVIEW / UNDER CONSTRUCTION] ####
Known_FreshwaterTaxa<- c(
  "tilapia",
  "rana", "rana toro",
  "trucha",
  "bagre canal"
) #common_name_clean

#### FLAG 7B - Flag known freshwater genera [LIST IS PENDING REVIEW / UNDER CONSTRUCTION] ####
Known_FreshwaterGenera<- c(
  "Cyprinus", # common carp; strictly freshwater.
  "Macrobrachium", # freshwater prawns/shrimps; mostly freshwater.
  "Hypophthalmichthys", # Asian carp; freshwater.
  "Chirostoma", # “charales,” endemic to freshwater lakes in Mexico.
  "Carassius", # goldfish/crucian carp; freshwater.
  "Micropterus", # black bass; freshwater.
  "Lepomis", # sunfish; freshwater.
  "Cyprinodon", # pupfish; freshwater.
  "Rana" # frogs
  ) # estimated_genus_name


#### FLAG 8 & 9 - Flag freshwater and NOT based on Major_Functional_Group and est_habitat type [UNDER CONSTRUCTION as of April 3 2026] ####

# Flag freshwater rows based on keywords (FLAGS 6 and 7), info from the IHH key, and estimated habitat type (based on landing site):
table(GOC_Landings_aquaculture$Major_Functional_Group)

# Definitely freshwater:
Definite_FreshwaterGroups<- c("Freshwater fishes", "Other freshwater taxa")

# Could be freshwater or marine:
Unknown_MarineFreshwaterGroups<- c("Marine or freshwater fishes NEI", "unknown")
Likely_MarineGroups<- c("Diadromous fishes", "Molluscs excl. Cephalopods", "Shrimps, prawns")

# Definitely marine:
Definite_MarineGroups<- c("Anchovy & Sardine (small pelagics)", "Cephalopods", "Crustaceans", "Demersal marine fishes", "Echinoderms", "Jellies", "Marine algae", "Marine fishes NEI", "Marine species NEI", "Other marine invertebrates", "Pelagic marine fish (excl. anchovy & sardine)", "Sharks")
# NOTES:
# - Crustaceans could technically be split but are overwhelmingly marine.
# - Molluscs could technically be split but are overwhelmingly marine. There could be some freshwater clams hiding in there.
# - Shrimp technically split but overwhelmingly marine. When in doubt, lean marine.

# Logic:
# - First, any row whose functional group is explicitly in Definite_MarineGroups will immediately get "CANNOT BE FRESHWATER; MARINE"
# - Second, if a row didn’t hit the first condition, it is evaluated for strong freshwater evidence. "YES FRESHWATER"
# - Third, rows that didn’t match the first two conditions are considered. This is broader marine signal. "LIKELY MARINE"
# - Fourth, rows that didn’t match the first two conditions are considered. This is an uncertain signal. "POSSIBLY FRESHWATER OR MARINE"
# - Finally, Any row that didn’t meet any of the above conditions gets "NOT FLAGGED EITHER WAY".

GOC_Landings_aquaculture <- GOC_Landings_aquaculture %>%
  mutate(
    freshwater_flag = case_when(
      
      # Definitely NOT freshwater:
      (Major_Functional_Group %in% Definite_MarineGroups) ~ "CANNOT BE FRESHWATER; MARINE",
      
      # Definite freshwater: any definite criteria
      (Major_Functional_Group %in% Definite_FreshwaterGroups) |
      (common_name_clean %in% Known_FreshwaterTaxa) |
      (estimated_genus_name %in% Known_FreshwaterGenera) |
      (rowSums(across(keywords_1:keywords_5, ~ .x %in% Freshwater_keywords), na.rm = TRUE) > 0) |
      (est_habitat == "freshwater") ~ "YES FRESHWATER",
      
      # Likely marine: broader/uncertain criteria
      (Major_Functional_Group %in% Likely_MarineGroups) ~ "LIKELY MARINE",
      
      # Possible freshwater: broader/uncertain criteria
      (Major_Functional_Group %in% Unknown_MarineFreshwaterGroups) ~ "POSSIBLY FRESHWATER OR MARINE",
      
      # Otherwise, not freshwater
      TRUE ~ "NOT FLAGGED EITHER WAY"
    )
  )

table(GOC_Landings_aquaculture$freshwater_flag)
# 287359 observations flagged as YES definitely "freshwater" based on freshwater keywords, the taxa common name, landing site, or Major_Functional_Group. 
# Meanwhile, other entries, like "MOJARRA" could go either way and are flagged as possibly freshwater.

# BEFORE April 3 updates:
#      NO     POSSIBLY      YES 
# 3226829     96104         299843 

# AFTER April 3 updates:
# CANNOT BE FRESHWATER; MARINE                 LIKELY MARINE        NOT FLAGGED EITHER WAY POSSIBLY FRESHWATER OR MARINE 
#                       2496598                        739242                            13                         99564 
#                YES FRESHWATER 
#                        287359                        13                        496351                        285775  
```

```{r SANITY CHECK: Do the things flagged as freshwater make sense?}
# Lot's of Mojarra moved into the possible freshwater category...
subset<- GOC_Landings_aquaculture %>%
  filter(freshwater_flag=="YES FRESHWATER")
# 287359 obs.

table(subset$estimated_taxa_group)

# AFTER April 3 updates:
         # ALMEJA           BAGRE         CAMARON         CARACOL           CARPA      LANGOSTINO          LOBINA         MOJARRA 
         #    509           30143            1749               1           36433            1834           31543          180509 
         # ORNATO           OTRAS PECES DE ORNATO          TRUCHA 
         #      2            4524              22              90 

# Let's take a closer look...
subset2<- subset %>%
  filter(recorded_taxa_group %in% c("ALMEJA", "CAMARON", "LANGOSTA", "LANGOSTINO"))

table(subset2$estimated_taxa_group)
table(subset2$estimated_species_name)
table(subset2$keywords_1)
# *Hm. But they are almost all (81%) flagged as "DE AGUA DUCLE" so yeah, that makes sense.
# I'm going to leave this as the "cost of doing business" for now but could be worth coming back to later.

subset3<- subset %>%
  filter(estimated_taxa_group %in% c("LOBINA"))
table(subset3$estimated_species_name)
# Florida bass. Makes sense!
```


```{r SANITY CHECK: Do the records classed as aquaculture and freshwater make sense?}
# SANITY CHECK:
table(GOC_Landings_aquaculture$est_aquaculture_production)
subset<- GOC_Landings_aquaculture %>% filter(est_aquaculture_production == "likely aquaculture")


table(subset$freshwater_flag)
# Note that out of the 279574 records that are "likely aquaculture", 173381 (~62%) are also flagged as freshwater species (things like tilapia, trout, and frogs). This makes sense!

# Meanwhile, the marine aquaculture species include shrimp, snapper, jacks, lots of bivalves and snails.
table(subset$recorded_taxa_name_clean)

subset2<- subset %>% filter(recorded_taxa_name_clean=="JUREL")
table(subset2$eu_name)
# NOTE TO SELF: Yes, these data for Jurel look really nice. These are definitely aquaculture jacks. Flags are doing a decent job so far!
```

```{r Reorder and rename columns in GOC_Landings_aquaculture in new df GOC_Landings_sector}
GOC_Landings_sector <- GOC_Landings_aquaculture %>%
  rename(reported_aquaculture_production = acuaculture_production,
         reported_fleet = fleet) %>%
  relocate(
    reported_aquaculture_production,
    NOTaquacultureFLAG_specific.groups,
    aquacultureFLAG_site.details,
    aquacultureFLAG_keywords,
    aquacultureFLAG_specific.taxa,
    aquacultureFLAG_combined,
    est_aquaculture_production,
    est_aquaculture_certainty,
    est_habitat,
    freshwater_flag,
    reported_fleet,
    .after = last_col()
  )
# 3622776 obs., 63 vars.
```

# MANUAL TEMPORARY PATCH: COME BACK TO THIS ERICA - Apr 3 2026
# Records from office_names on the Pacific side, where landing_site is missing are assumed not the Gulf of California.

```{r Patch for Pacific side landing sites TEMPORARY}
Office_Names <- GOC_Landings_sector %>%
  select(state, office_name, landing_site, inside_GOC) %>%
  filter(is.na(landing_site) & !is.na(office_name)) %>%
  distinct()
# 49 offices where is.na(landing_site)
```
# Filled this in manually. This is a temporary patch and should be fixed at some point.

```{r Load in manual patch}
ManualLatLong_patch<- read_csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/LatLong_OfficeName_ManualPatch_Apr032026.csv")
```

```{r Apply manual patch}
# The logic is as follows: In cases where landing_site is NA in GOC_Landings_sector, use the ManualLatLong_patch to update the inside_GOC column. Apply the update based on a match to state and office_name.

# Clean and standardize both tables:
GOC_Landings_sector <- GOC_Landings_sector %>%
  mutate(
    state_clean = str_to_upper(str_trim(stri_trans_general(state, "Latin-ASCII"))),
    office_clean = str_to_upper(str_trim(stri_trans_general(office_name, "Latin-ASCII"))),
    landing_site_na = is.na(landing_site) | landing_site == ""  # treat empty strings as missing
  )

ManualLatLong_patch <- ManualLatLong_patch %>%
  mutate(
    state_clean = str_to_upper(str_trim(stri_trans_general(state, "Latin-ASCII"))),
    office_clean = str_to_upper(str_trim(stri_trans_general(office_name, "Latin-ASCII")))
  )

# Apply patch only where landing_site is NA
GOC_patched <- GOC_Landings_sector %>%
  left_join(
    ManualLatLong_patch %>% select(state_clean, office_clean, inside_GOC_patch = inside_GOC),
    by = c("state_clean", "office_clean")
  ) %>%
  mutate(
    inside_GOC = if_else(landing_site_na, inside_GOC_patch, inside_GOC)
  ) %>%
  select(-inside_GOC_patch, -state_clean, -office_clean, -landing_site_na)

# 3622776 obs., 63 vars. (both as expected)


# SANITY CHECK: Did the patch work as expected?
subset<- GOC_patched %>%
  filter(office_name=="ENSENADA" & is.na(landing_site))
# Seems to!

Office_Names <- GOC_patched %>%
  select(state, office_name, landing_site, inside_GOC) %>%
  filter(is.na(landing_site) & !is.na(office_name)) %>%
  distinct()
# Yes, it worked!

#### APPLY MANUAL PATCH TO GOC_Landings_sector, RETURN TO THIS LATER ####
table(GOC_patched$inside_GOC)

GOC_Landings_sector <- GOC_patched %>%
  filter(inside_GOC== "yes" | is.na(inside_GOC))
# 3598085 obs., 63 vars. (as expected)
```


# STEP 11: Estimate SECTOR based on record details, estimated genera, estimated Major Functional Group, aquaculture flags, and freshwater flags. ------------------------------

What sector was actually reported?
```{r Create sector_estimate_reported column}
colnames(GOC_Landings_sector)

GOC_Landings_sector <- GOC_Landings_sector %>%
  mutate(
    sector_estimate_reported = case_when(
      
      # More specific first
      reported_aquaculture_production == "yes" & reported_fleet == "small_scale" ~ "small-scale aquaculture",
      
      # General aquaculture
      reported_aquaculture_production == "yes" ~ "aquaculture",
      
      # Fisheries by fleet
      reported_fleet == "small_scale" ~ "small-scale fisheries",
      reported_fleet == "large_scale" ~ "large-scale fisheries",
      
      # Fallback for missing/unknown
      TRUE ~ "yet to be classified"
    )
  )

# SANITY CHECK:
table(GOC_Landings_sector$sector_estimate_reported)
          # aquaculture large-scale fisheries small-scale fisheries  yet to be classified 
          #       17496                142431               1052851               2385307 

# Re-level
GOC_Landings_sector <- GOC_Landings_sector %>%
  mutate(
    sector_estimate_reported = factor(
      sector_estimate_reported,
      levels = c(
        "aquaculture",
        "yet to be classified",
        "small-scale fisheries",
        "large-scale fisheries"
      )
    )
  )

# REPORTED:
SectorLandings_summary_REPORTED<- GOC_Landings_sector %>% 
  filter(year >= 2014, year <= 2024) %>%
  select(year, sector_estimate_reported, weight_kg_corrected, value) %>%
  group_by(year, sector_estimate_reported) %>%
  summarize(est_annual_catch_kg= sum(weight_kg_corrected, na.rm = TRUE),
            est_annual_value= sum(value, na.rm = TRUE))

SectorLandings_summary_REPORTED$est_annual_catch_mton <- SectorLandings_summary_REPORTED$est_annual_catch_kg/1000

table(SectorLandings_summary_REPORTED$sector_estimate_reported)
```

```{r Estimate sector}
table(GOC_Landings_sector$est_aquaculture_production)   # adds up to: 3598085, as expected
table(GOC_Landings_sector$freshwater_flag)              # adds up to: 3598085, as expected
table(GOC_Landings_sector$reported_fleet)               # adds up to: 1195282 (some NAs)

#### ESTIMATE MOST LIKELY SECTOR based on report info and flags: ####

GOC_Landings_sector <- GOC_Landings_sector %>%
  mutate(
    # Helper columns
    is_aquaculture = est_aquaculture_production %in% c("aquaculture", "likely aquaculture"),
    is_likely.wildcatch = est_aquaculture_production %in% c("unlikely aquaculture", "not aquaculture", "possibly aquaculture"),
    is_freshwater = freshwater_flag %in% c("YES FRESHWATER"),
    is_marine.likely = freshwater_flag %in% c("CANNOT BE FRESHWATER; MARINE", "LIKELY MARINE"),
    is_fresh.or.marine = freshwater_flag %in% c("POSSIBLY FRESHWATER OR MARINE", "NOT FLAGGED EITHER WAY"),
    
    # Implement logic to estimate sector
    sector_estimate_draft = case_when(
      
      #### DEFINITE FRESHWATER SECTORS: ####
      is_freshwater & is_aquaculture ~ "freshwater: aquaculture",
      is_freshwater & is_likely.wildcatch & reported_fleet == "small_scale" ~ "freshwater: small-scale capture",
      is_freshwater & is_likely.wildcatch & reported_fleet == "large_scale" ~ "freshwater: large-scale capture",
      is_freshwater ~ "freshwater: fishery sector yet to be classified",
      
      #### DEFINITE & LIKELY MARINE SECTORS: ####
      is_marine.likely & is_aquaculture ~ "marine: aquaculture",
      is_marine.likely & is_likely.wildcatch & reported_fleet == "small_scale" ~ "marine: small-scale capture",
      is_marine.likely & is_likely.wildcatch & reported_fleet == "large_scale" ~ "marine: large-scale capture",
      is_marine.likely ~ "marine: fishery sector yet to be classified",
      
      #### SECTOR UNCERTAIN, COULD GO EITHER WAY: ####
      is_fresh.or.marine & is_aquaculture ~ 
        "freshwater or marine: aquaculture",
      is_fresh.or.marine & is_likely.wildcatch & reported_fleet == "small_scale" ~ 
        "freshwater or marine: small-scale capture",
      is_fresh.or.marine & is_likely.wildcatch & reported_fleet == "large_scale" ~ 
        "freshwater or marine: large-scale capture",
      is_fresh.or.marine ~ 
        "freshwater or marine: fishery sector yet to be classified",
      
      # Catch-all safeguard
      TRUE ~ NA_character_
    )
  ) %>%
  select(-is_aquaculture, -is_likely.wildcatch, -is_freshwater, -is_marine.likely, -is_fresh.or.marine)


# SANITY CHECK:
table(GOC_Landings_sector$sector_estimate_draft)

# Estimates BEFORE reclassification work to species key like MOJARRA (Mar 24 2026):
         # freshwater: aquaculture         freshwater: wild capture freshwater: yet to be classified    marine: large-scale fisheries 
         #                  175109                           127130                             1054                           142404 
         #     marine: mariculture    marine: small-scale fisheries     marine: yet to be classified 
         #                  123999                           964673                          2088407 

# Estimates AFTER reclassification work (Mar 24 2026):
                 #         freshwater or marine: aquaculture                  freshwater or marine: large-scale capture 
                 #                                      1843                                                        478 
                 # freshwater or marine: small-scale capture freshwater or marine: specific sector yet to be classified 
                 #                                     31528                                                      62305 
                 #                   freshwater: aquaculture                            freshwater: large-scale capture 
                 #                                    174541                                                         19 
                 #           freshwater: small-scale capture           freshwater: specific sector yet to be classified 
                 #                                     26374                                                      96625 
                 #                       marine: aquaculture                                marine: large-scale capture 
                 #                                    122732                                                     141931 
                 #               marine: small-scale capture               marine: specific sector yet to be classified 
                 #                                    947425                                                    2016975 

# Estimates AFTER new lat longs (Apr 2 2026):
                 #         freshwater or marine: aquaculture                  freshwater or marine: large-scale capture 
                 #                                      1843                                                        478 
                 # freshwater or marine: small-scale capture freshwater or marine: specific sector yet to be classified 
                 #                                     31489                                                      62294 
                 #                   freshwater: aquaculture                            freshwater: large-scale capture 
                 #                                    175608                                                         19 
                 #           freshwater: small-scale capture           freshwater: specific sector yet to be classified 
                 #                                     27059                                                      97157 
                 #                       marine: aquaculture                                marine: large-scale capture 
                 #                                    121665                                                     141931 
                 #               marine: small-scale capture               marine: specific sector yet to be classified 
                 #                                    946779                                                    2016454 

# After sector-specific improvements & removal of non-GOC sites... - adds up to 3598085 as expected. These look great! (Apr 3 2026)
                #         freshwater or marine: aquaculture freshwater or marine: fishery sector yet to be classified 
                #                                      1492                                                     64531 
                # freshwater or marine: large-scale capture                 freshwater or marine: small-scale capture 
                #                                      1063                                                     31981 
                #                   freshwater: aquaculture           freshwater: fishery sector yet to be classified 
                #                                    173580                                                     87650 
                #           freshwater: large-scale capture                           freshwater: small-scale capture 
                #                                        19                                                     24234 
                #                       marine: aquaculture               marine: fishery sector yet to be classified 
                #                                    105932                                                   2016090 
                #               marine: large-scale capture                               marine: small-scale capture 
                #                                    141346                                                    950167 



table(GOC_Landings_sector$est_aquaculture_production)
table(GOC_Landings_sector$freshwater_flag)
table(GOC_Landings_sector$reported_fleet)
```

```{r Re-level sector_estimate_draft}
# Re-level
GOC_Landings_sector <- GOC_Landings_sector %>%
  mutate(
    sector_estimate_draft = factor(
      sector_estimate_draft,
      levels = c(
        "marine: fishery sector yet to be classified",
        "marine: large-scale capture",
        "marine: small-scale capture",
        "marine: aquaculture",
        "freshwater: aquaculture",
        "freshwater: small-scale capture",
        "freshwater: large-scale capture",
        "freshwater: fishery sector yet to be classified",
        "freshwater or marine: aquaculture",
        "freshwater or marine: small-scale capture",
        "freshwater or marine: large-scale capture",
        "freshwater or marine: fishery sector yet to be classified"
      )
    )
  )
```

```{r Summarize by sector}
# ESTIMATED:
SectorLandings_summary <- GOC_Landings_sector %>% 
  filter(year >= 2014, year <= 2024) %>%
  select(year, weight_kg_corrected, value, sector_estimate_draft) %>%
  group_by(year, sector_estimate_draft) %>%
  summarize(
    est_annual_catch_kg = sum(weight_kg_corrected, na.rm = TRUE),
    est_annual_value    = sum(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # Convert to metric tons
  mutate(est_annual_catch_mton = est_annual_catch_kg / 1000) %>%
  # Compute percent of annual total catch
  group_by(year) %>%
  mutate(percent_AnnualTotCatch = round(100 * est_annual_catch_kg / sum(est_annual_catch_kg, na.rm = TRUE))) %>%
  ungroup()

# Okay great. So from this we can see that the following sectors make up only 0-1% of total catch. I feel comfortable ignoring them then for the purposing of assessing food provising (maybe a different story for natural products) - Mar 24 2026.
# Small contributions from:
# freshwater: small-scale capture
# freshwater: large-scale capture
# freshwater: fishery sector yet to be classified
# freshwater or marine: aquaculture
# freshwater or marine: small-scale capture
# freshwater or marine: large-scale capture
# freshwater or marine: fishery sector yet to be classified

table(SectorLandings_summary$sector_estimate_draft)
```

```{r Trim and prepare GOC_Landings_sector df for export}
colnames(GOC_Landings_sector)

GOC_Landings_sector<- GOC_Landings_sector %>%
  select(-c(Manual_Classification_Notes, species_est_reasoning, genus_est_reasoning))

# 3598085 obs., 62 obs.
```

## Plotting sector estimates:
```{r Plot each sector of interest over time}
# SECTORS OF INTEREST:
Sectors_of_Interest<- c("marine: fishery sector yet to be classified" , "marine: small-scale capture", "marine: large-scale capture" , "marine: aquaculture", "freshwater: aquaculture")
  
# Define colors for each sector (fleet distinction = lighter for small-scale)
sector_colors <- c(
  # Marine (blues)
  "marine: fishery sector yet to be classified"       = "#cce0ff",  # very light blue for unknown
  "marine: small-scale capture"                       = "#6699cc",  # medium blue
  "marine: large-scale capture"                       = "#336699",  # darker blue
  "marine: aquaculture"                               = "#33cccc",  # teal/aquaculture cue
  
  # Freshwater (warm shades)
  "freshwater: aquaculture"                          = "#ff9933"  # bright orange
  # "freshwater: fishery sector yet to be classified" = "#ffe0cc",  # light orange for unknown - IGNORE THIS SECTOR FOR NOW.
  # "freshwater: small-scale capture"                  = "#ff6666",  # strong red/orange - IGNORE THIS SECTOR FOR NOW.
  # "freshwater: large-scale capture"                  = "#ffb366",  # lighter orange/yellow - IGNORE THIS SECTOR FOR NOW.
  
  # Freshwater or Marine (grey and purple shades) - IGNORE FOR NOW.
  # "freshwater or marine: fishery sector yet to be classified"  = "#e0e0e0",  # light grey
  # "freshwater or marine: small-scale capture"                    = "#b399cc",  # light purple
  # "freshwater or marine: large-scale capture"                    = "#8040a0",  # medium purple
  # "freshwater or marine: aquaculture"                             = "#999999"   # neutral grey
)

# Plot
Landings_bySector_plot<- SectorLandings_summary %>%
  filter(year >= 2018) %>%
  filter(sector_estimate_draft %in% Sectors_of_Interest) %>%
  ggplot(aes(x = year, y = est_annual_catch_mton, fill = sector_estimate_draft)) +
  
  # Area plot with slight border for separation
  geom_area(alpha = 0.9, color = "white", linewidth = 0.2) +
  
  # Manual fill palette
  scale_fill_manual(
    values = sector_colors,
    name = "Sector"
  ) +
  
  # Labels
  labs(
    x = "Year",
    y = "Estimated Annual Catch (metric tons)",
    title = "Estimated Catch by Sector",
    subtitle = "Filtered for 2018 onwards; dropped sectors that contributed < 1% total catch"
  ) +
  
  # Minimal theme and sleek styling
  theme_bw(base_size = 14) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey85"),
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 13, color = "grey30")
  ) +
  
  # Optional: better legend layout
  guides(
    fill = guide_legend(nrow = 3, byrow = TRUE, title.position = "top")
  )

Landings_bySector_plot
```

Looks great! Save this plot.
```{r Save Landings_bySector_plot}
# Save the plot as a high-resolution PNG
ggsave(
  filename = "/home/ferrer/ohi_goc_prep/FP/CONAPESCA_CleanUp/Catch_bySector_plot_Apr032026.png",
  plot = Landings_bySector_plot,
  width = 10,       # width in inches
  height = 6,       # height in inches
  dpi = 300         # resolution
)
```

```{r Plot sector_estimate_reported}
# Plot sector_estimate_reported
REPORTEDLandings_bySector_plot<- SectorLandings_summary_REPORTED %>%
  filter(year >= 2018) %>%
  ggplot(aes(x = year, y = est_annual_catch_mton, fill = sector_estimate_reported)) +
  
  # Area plot with slight border for separation
  geom_area(alpha = 0.9, color = "white", linewidth = 0.2) +
  
  # Labels
  labs(
    x = "Year",
    y = "Reported Annual Catch (metric tons)",
    title = "Catch by Reported Sector",
    subtitle = "Filtered for 2018 onwards"
  ) +
  
  # Minimal theme and sleek styling
  theme_bw(base_size = 14) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey85"),
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 13, color = "grey30")
  ) +
  
  # Optional: better legend layout
  guides(
    fill = guide_legend(nrow = 3, byrow = TRUE, title.position = "top")
  )

REPORTEDLandings_bySector_plot
```

```{r Clean workspace as needed}
# remove dataframes to keep workspace clean:
rm(subset, subset2, subset3,
   Landings_bySector_plot, REPORTEDLandings_bySector_plot, EstvReported_Aquaculture_plot,
   Estimated.Aquaculture_summary, Reported.Aquaculture_summary, Aquaculture_summary,
   SectorLandings_summary, SectorLandings_summary_REPORTED,
   GOC_patched, ManualLatLong_patch, Office_Names, test, vector) # Patch dataframes. Remove later.
```

# Save the GOC_Landings_sector dataframe
```{r Save the GOC_Landings_sector dataframe}
saveRDS(GOC_Landings_sector, file="/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Landings_SectorSorted_2000.present_EMF.Apr032026.rds")
# 3598085 obs., 62 vars.
```


# STEP 12: Filter out freshwater & aquaculture landings from marine wild-capture and save these data frames as separate files. ------------------------------

```{r Load GOC_Landings_sector if needed}
GOC_Landings_sector<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Landings_SectorSorted_2000.present_EMF.Apr032026.rds"))
# 3598085 obs., 62 vars.
```

```{r Filter out freshwater and marine sectors}
subset<- GOC_Landings_sector %>%
  filter(is.na(sector_estimate_draft))
table(GOC_Landings_sector$sector_estimate_draft)
# 0. All records have a first-draft estimate for sector. Good!

Freshwater_Sectors<- c("freshwater: fishery sector yet to be classified", "freshwater: aquaculture", 
                       "freshwater: small-scale capture", "freshwater: large-scale capture")

Freshwater.Marine_Sectors<- c("freshwater or marine: fishery sector yet to be classified", "freshwater or marine: aquaculture",
                              "freshwater or marine: small-scale capture", "freshwater or marine: large-scale capture")

Marine_Sectors<- c("marine: fishery sector yet to be classified", "marine: aquaculture",
                   "marine: large-scale capture", "marine: small-scale capture")


#### FRESHWATER DATAFRAME (includes uncertain records): ####
GOC_Landings_freshwater <- GOC_Landings_sector %>%
  filter(sector_estimate_draft %in% c(Freshwater_Sectors, Freshwater.Marine_Sectors))
# 384550 obs., 62 vars (AFTER Apr 3 sector revision; note the change is marginal)

# Check that this worked:
table(GOC_Landings_freshwater$sector_estimate_draft)
# Looks good! Only freshwater or undetermined sectors as we'd expect.

#### MARINE DATAFRAME (includes uncertain records): ####
GOC_Landings_marine <- GOC_Landings_sector %>%
  filter(sector_estimate_draft %in% c(Marine_Sectors, Freshwater.Marine_Sectors))
# 3312602 obs., 62 vars (AFTER Apr 3 sector revision; note the change is marginal)

# Check that this worked:
table(GOC_Landings_marine$sector_estimate_draft)
# Looks good! Only marine or undetermined sectors as we'd expect.
```
NOTE TO SELF: There could still be some freshwater species hiding in here. For example, the "MOJARRA" general category is mixed freshwater and marine demersal fish.

# Save GOC_Landings_freshwater and GOC_Landings_marine dataframes
```{r Save GOC_Landings_freshwater and GOC_Landings_marine dataframes}
# GOC_Landings_freshwater: 
saveRDS(GOC_Landings_freshwater, file="/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Landings_FRESHWATER_2000.present_EMF.Apr032026.rds")
# 384550 obs., 62 vars.

# GOC_Landings_marine: 
saveRDS(GOC_Landings_marine, file="/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Landings_MARINE_2000.present_EMF.Apr032026.rds")
# 3312602 obs., 62 vars.
```


# STEP 13A: Generate MARICULTURE List

```{r Load GOC_Landings_marine if needed}
GOC_Landings_marine<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Landings_MARINE_2000.present_EMF.Apr032026.rds"))
# 3312602 obs., 62 vars.
```

```{r FILTER OUT MARICULTURE SECTORS OF INTEREST:}
table(GOC_Landings_marine$sector_estimate_draft)

# FILTER OUT MARICULTURE SECTORS OF INTEREST: ----
# Let's assume that the records are most accurate since 2018.

Mariculture_Sectors<- c("marine: aquaculture")

GOC_Landings_mariculture<- GOC_Landings_marine %>% 
  filter(year>2017) %>%
  filter(sector_estimate_draft %in% Mariculture_Sectors)
# 91680 obs., 62 vars.

range(GOC_Landings_mariculture$year)
```

```{r Look into this first-draft list before proceeding}
# Taking the information I've gleaned here to refine my sector estimates above.
table(GOC_Landings_mariculture$estimated_genus_name)
table(GOC_Landings_mariculture$Major_Functional_Group)
table(GOC_Landings_mariculture$estimated_genus_name)
```
```{r Generate a table / list of common n > 10 records aquaculture species since 2018}
AquacultureGenera_summary <- GOC_Landings_mariculture %>%
  filter(year > 2017) %>%
  filter(!is.na(estimated_genus_name) & estimated_genus_name != "NEI" & estimated_genus_name != "unknown") %>%
  select(estimated_genus_name, Major_Functional_Group) %>%
  group_by(Major_Functional_Group, estimated_genus_name) %>%
  summarise(
    n_records_since2018 = n(),
    .groups = "drop"
  )

AquacultureGenera_summary<- AquacultureGenera_summary %>%
  filter(n_records_since2018 > 10)
```


```{r Investigate records for unexpected taxa}
table(AquacultureGenera_summary$estimated_genus_name)
table(GOC_Landings_mariculture$reported_aquaculture_production) # is no = 1053
sum(is.na(GOC_Landings_mariculture$reported_aquaculture_production)) # is na = 104204

# Investigate weird entries
subset1<- GOC_Landings_mariculture %>%
  filter(year > 2017) %>%
  filter(estimated_genus_name == "Lutjanus") 
table(subset1$eu_name)
# Makes sense! All records from Ocean Farms in La Paz.

subset2<- GOC_Landings_mariculture %>%
  filter(year > 2017) %>%
  filter(estimated_genus_name == "Argopecten") 
table(subset2$keywords_1)
# Makes sense! All records say "de cultivo" explicitly.

subset4<- GOC_Landings_mariculture %>%
  filter(year > 2017) %>%
  filter(estimated_genus_name == "Dosinia") 
table(subset4$keywords_1)
# Makes sense! All records say "de cultivo" explicitly.

subset5<- GOC_Landings_mariculture %>%
  filter(year > 2017) %>%
  filter(estimated_genus_name == "Megapitaria") 
table(subset5$keywords_1)
# Makes sense! All records say "de cultivo" explicitly.
```

# STEP 13A: Generate ARTISNAL List
```{r FILTER OUT SSF SECTORS OF INTEREST:}
table(GOC_Landings_marine$sector_estimate_draft)

# FILTER OUT SSF SECTORS OF INTEREST: ----
# Let's assume that the records are most accurate since 2018.

SSF_Sectors<- c("marine: small-scale capture")

GOC_Landings_SSF<- GOC_Landings_marine %>% 
  filter(year>2017) %>%
  filter(sector_estimate_draft %in% SSF_Sectors)
# 950167 obs., 62 vars.

range(GOC_Landings_SSF$year)
```

```{r Generate a table / list of common n > 10 records SSF since 2018}
SSFGenera_summary <- GOC_Landings_SSF %>%
  filter(year > 2017) %>%
  filter(!is.na(estimated_genus_name) & estimated_genus_name != "NEI" & estimated_genus_name != "unknown") %>%
  select(estimated_genus_name, Major_Functional_Group) %>%
  group_by(Major_Functional_Group, estimated_genus_name) %>%
  summarise(
    n_records_since2018 = n(),
    .groups = "drop"
  )

SSFGenera_summary<- SSFGenera_summary %>%
  filter(n_records_since2018 > 10)
```


# STEP 14: Interrogate the economic unit (eu_rnpa) and vessel (vessel_name) columns. ------------------------------

```{r Load in dataframe if needed}
GOC_Landings_marine<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Landings_MARINE_2000.present_EMF.Apr032026.rds"))
# 3322933 obs., 61 vars.
```

```{r FILTER OUT MARINE SECTORS OF INTEREST:}
# FILTER OUT MARINE SECTORS OF INTEREST: ----
MarineFishery_Sectors<- c("marine: fishery sector yet to be classified" , "marine: small-scale capture", "marine: large-scale capture")

GOC_Landings_mar.fisheries<- GOC_Landings_marine %>% 
  filter(sector_estimate_draft %in% MarineFishery_Sectors)
# 3105164 obs., 61 vars.

range(GOC_Landings_mar.fisheries$year)
```

```{r First look}
length(unique(GOC_Landings_mar.fisheries$eu_name))
# 5002 unique economic units in the dataframe.

length(unique(GOC_Landings_mar.fisheries$eu_rnpa))
# 4647 unique permits given to economic units in the dataframe.

length(unique(GOC_Landings_mar.fisheries$vessel_name))
# 2141 unique vessel names in the dataframe.

length(unique(GOC_Landings_mar.fisheries$vessel_rnpa))
# 2134 unique vessel permit numbers in the dataframe.

# So, for the most part, each vessel is registered to a unique vessel permit.'

# How many of the 3229063 total observations have no economic unit name associated with them? Only ~2% of all records.
GOC_Landings_mar.fisheries %>%
  count(eu_name == "NO DISPONIBLE")
# 68006 / 3105164

# How many of all observations have no vessel name associated with them? About 51% of all records.
GOC_Landings_mar.fisheries %>%
  count(vessel_rnpa == "00000000")
# 1611882 / 3105164

# This tells us that about half of all records have no information about the fishers' vessel, but nearly all have permit info.
```

```{r Testing some ideas}
table(GOC_Landings_mar.fisheries$reported_aquaculture_production)
# No aquaculture hiding in here...

# Look into a specific unit:
subset2<- GOC_Landings_mar.fisheries %>%
  filter(eu_name == "SCPP PUNTA TIBURON, S DE RL DE CV")
table(subset2$reported_aquaculture_production)
# It looks like this particular economic unit (possibly a cooperative?) does not work in aquaculture. 

# Look into another specific unit:
subset3<- GOC_Landings_mar.fisheries %>%
  filter(eu_name == "SCPP DE SERV LA CAIMANERA SC DE RL")
table(subset3$reported_aquaculture_production)
# Notice that some of these economic units say nothing about aquaculture, whereas others do. 

# Keep workspace clean:
rm(subset2, subset3)
```
Let's leave this section under construction. I'm not totally sure what to do with this yet. But it will impact how and if we can calculate CPUE.

Waiting to here back from Juan Carlos is if there is a key or roster for economic units...

```{r Let's plot missing data over time}
# Summarize missing vessel IDs over time
vessel_missing_over_time <- GOC_Landings_mar.fisheries %>%
  mutate(missing_vessel = is.na(vessel_rnpa) | vessel_rnpa == "00000000" | vessel_rnpa == "0000000_") %>%
  group_by(year) %>%   # Replace 'year' with your time column
  summarise(
    total = n(),
    missing_count = sum(missing_vessel),
    missing_percent = 100 * missing_count / total
  )

# Plot missing vessel prevalence over time
ggplot(vessel_missing_over_time, aes(x = year, y = missing_percent)) +
  geom_line(color = "orange", size = 1.2) +
  geom_point(color = "tomato", size = 2) +
  labs(
    title = "Prevalence of Missing Vessel IDs Over Time",
    x = "Year",
    y = "Percentage of Records Missing Vessel ID"
  ) +
  scale_y_continuous(limits = c(0, 100)) +
  theme_bw()

# Vessel data is missing throughout the time series.

# Plot the opposite: what's not missing?
vessel_present_over_time <- GOC_Landings_mar.fisheries %>%
  mutate(missing_vessel = is.na(vessel_rnpa) | vessel_rnpa == "00000000" | vessel_rnpa == "0000000_") %>%
  group_by(year) %>%
  summarise(
    total = n(),
    present_count = sum(!missing_vessel),
    present_percent = 100 * present_count / total
  )

ggplot(vessel_present_over_time, aes(x = year, y = present_percent)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_point(color = "navy", size = 2) +
  
  labs(
    title = "Proportion of Records WITH Vessel IDs Over Time",
    x = "Year",
    y = "Percentage of Records with Vessel ID"
  ) +
  scale_y_continuous(limits = c(0, 100)) +
  theme_bw()

# Plot the opposite: what's not missing?
eu_present_over_time <- GOC_Landings_mar.fisheries %>%
  mutate(missing_eu = eu_name == "NO DISPONIBLE") %>%
  group_by(year) %>%
  summarise(
    total = n(),
    present_count = sum(!missing_eu),
    present_percent = 100 * present_count / total
  )

ggplot(eu_present_over_time, aes(x = year, y = present_percent)) +
  geom_line(color = "lavender", size = 1.2) +
  geom_point(color = "orchid", size = 2) +
  
  labs(
    title = "Proportion of Records WITH EU Details Over Time",
    x = "Year",
    y = "Percentage of Records with EU details"
  ) +
  scale_y_continuous(limits = c(0, 100)) +
  theme_bw()
```

```{r Investigate vessel names since 2018}
GOC_Landings_vessels<- GOC_Landings_mar.fisheries %>%
  filter(year >= 2018)
# 1104230 obs., 61 vars.

length(unique(GOC_Landings_vessels$vessel_name))
# 1297 unique vessel names in the dataframe.
length(unique(GOC_Landings_vessels$vessel_rnpa))
# 1279 unique vessel permit numbers in the dataframe.

# So, for the most part, each vessel is registered to a unique vessel permit.'

# How many of the 1104876 total observations have no economic unit name associated with them? Essentially 0% of records.
GOC_Landings_vessels %>%
  count(eu_name == "NO DISPONIBLE")
# 4 / 1104230

# How many of all observations have no vessel permit associated with them? 86% of all records.
GOC_Landings_vessels %>%
  count(vessel_rnpa == "00000000" | vessel_rnpa == "0000000_" | is.na(vessel_rnpa))
# 946953 / 1104230

# How many of all observations have no vessel NAME associated with them? About 86% of all records.
# Okay, so we know vessel name alone won't be a good estimator for CPUE...
GOC_Landings_vessels %>%
  count(is.na(vessel_name))
# 946953 / 1104230

# Take a look at a random assortment of rnpas versus names to understnad what's missing and what's not:
subset <- GOC_Landings_vessels %>%
  filter(is.na(vessel_name)) %>%
  select(vessel_rnpa, vessel_name, eu_name, eu_rnpa) %>%
  distinct() %>%
  slice_sample(n = 200)
# This shows they have real eu names, just no vessel details.
```

```{r Testing some ideas}
colnames(GOC_Landings_vessels)

# Case studies:
subset <- GOC_Landings_vessels %>%
  filter(eu_rnpa == "2508005234")

length(unique(subset$vessel_name))
# 9 vessels registered to this economic unit permit

subset2<- GOC_Landings_vessels %>%
  filter(eu_rnpa =="0203010103")

length(unique(subset2$vessel_name))
# 1 vessel name registered to this economic unit permit
length(unique(subset2$eu_name))
# 1 name for this economic unit
unique(subset2$vessel_rnpa)
# 1 vessel permit for this economic unit, 0000000_

range(subset2$year)
# From 2018 to 2025
unique(subset2$recorded_taxa_name_clean)
# Across several taxa
unique(subset2$sector_estimate_reported)
# And all small-scale (or unclassified)
```

## Esitmate CPUE per economic unit 2018-2024 
```{r Estimate CPUE per economic unit 2018-2024}
# Let's look through unit Economic Units. All records have a non-NA economic unit (at least since 2018)
EconomicUnits_df <- GOC_Landings_vessels %>%
  distinct(eu_rnpa)    # Keeps unique values as a one-column tibble
# 2849 obs.

GOC_Landings_vessels %>%
  count(is.na(eu_rnpa))


# Estimate Catch per economic unit 2018-2024: summarize weight_kg_corrected by eu_rnpa per year

CatchPerEU_summary <- GOC_Landings_vessels %>%
  filter(year >= 2018 & year <= 2024) %>%   # filter the years
  group_by(year, eu_rnpa) %>%              # group by year and economic unit
  summarize(
    total_catch_kg = sum(weight_kg_corrected, na.rm = TRUE)  # sum catch, ignore NAs
  ) %>%
  ungroup()  # remove grouping after summarizing

# Next, pull out the eu_rnpas that are present in all years 2018-2025. Using these eu_rnpas only, estimate the total catch of Major_Taxa_Group divided by the total eu_rnpa per year

# Step 1: Identify eu_rnpa present in all years
EU_all_years <- GOC_Landings_vessels %>%
  filter(year >= 2018 & year <= 2025) %>%
  group_by(eu_rnpa) %>%
  summarize(years_present = n_distinct(year), .groups = "drop") %>%
  filter(years_present == 8) %>%    # 2018–2025 = 8 years
  pull(eu_rnpa)                     # get vector of eu_rnpa
# 1003 eu.rnpas are present in all years 2018-2025

# Step 2: Filter data to only include these eu_rnpa
filtered_data <- GOC_Landings_vessels %>%
  filter(eu_rnpa %in% EU_all_years, year >= 2018 & year <= 2025)

# Step 3: Summarize catch by Major_Functional_Group per year
CPUE_summary <- filtered_data %>%
  group_by(year, Major_Functional_Group) %>%
  summarize(
    total_catch_mt = sum(weight_kg_corrected, na.rm = TRUE)/1000,
    n_eu = n_distinct(eu_rnpa),                    # number of economic units that year
    catch.mt_per_eu = total_catch_mt / n_eu          # divide total catch by number of units
  ) %>%
  ungroup()

# Looking promising! Let's plot it...
```

```{r Plot CPUE for Small Pelagics}
# PLOT FOR ALL GROUPS, excluding Small Pelagics:
# Remove NAs or 'unknown' if you don’t want them cluttering the plot
# First compute overall yearly mean (across all groups)

# First we need mean CPUE over years per group (2018-2024)
CPUE_mean_df <- CPUE_summary %>%
  filter(year > 2017 & year < 2025) %>%
  group_by(Major_Functional_Group) %>%
  summarise(mean_cpue = mean(catch.mt_per_eu), .groups = "drop")

CPUE_perEU_plot <- 
  
  CPUE_summary %>%
  filter(year < 2025) %>%
  
  ggplot(aes(x = year, y = catch.mt_per_eu)) +
  
  # Group-specific lines
  geom_line(aes(color = Major_Functional_Group), size = 1.1) +
  geom_point(aes(color = Major_Functional_Group), size = 2) +
  
  facet_wrap(~ Major_Functional_Group, scales = "free_y") +
  
  geom_hline(
  data = CPUE_mean_df,
  aes(yintercept = mean_cpue, group = Major_Functional_Group),
  color = "grey40",
  linetype = "dashed",
  linewidth = 0.8) +
  
  labs(
    title = "CPUE per Economic Unit by Functional Group (2018-2024)",
    x = "Year",
    y = "Catch per Economic Unit (mt)"
  ) +
  
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

CPUE_perEU_plot
```

```{r Save CPUE_perEU_plot}
# Save the plot as a high-resolution PNG
ggsave(
  filename = "/home/ferrer/ohi_goc_prep/FP/CONAPESCA_CleanUp/CPUE_perEU_plot_Apr022026.png",
  plot = CPUE_perEU_plot,
  width = 10,       # width in inches
  height = 6,       # height in inches
  dpi = 300         # resolution
)
```




# STEP X: Work on NATURAL PRODUCTS (ORNAMENTAL) List
```{r}

```



# > SHIFTING GEARS TO CMSY  ------------------------------

# STEP 15: Re-generate time series (from 2004-2024) per genus to use in CMSY [FIRST DRAFT COMPLETE; UNDER CONSTRUCTION]

```{r Load GOC_Landings_marine dataframe as needed}
GOC_Landings_marine<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/GOC_Landings_MARINE_2000.present_EMF.Apr032026.rds"))
# 3324153 obs., 61 vars.
```

```{r FILTER OUT MARINE SECTORS OF INTEREST:}
# FILTER OUT MARINE FISHERY SECTORS OF INTEREST: ----
MarineFishery_Sectors<- c("marine: fishery sector yet to be classified" , "marine: small-scale capture", "marine: large-scale capture")

GOC_Landings_mar.fisheries<- GOC_Landings_marine %>% 
  filter(sector_estimate_draft %in% MarineFishery_Sectors)
# 3105164 obs., 61 vars.

range(GOC_Landings_mar.fisheries$year)
```

```{r Sort by genera}
table(GOC_Landings_mar.fisheries$estimated_genus_name)
# Some of these genera don't appear very often. I want to filter genera for which we have more than 10 years worth of data and > 10 records per year.

# Summarize per genus and year
Genera_summary <- GOC_Landings_mar.fisheries %>%
  filter(!is.na(estimated_genus_name)) %>%
  select(estimated_genus_name, Major_Functional_Group, year) %>%
  group_by(estimated_genus_name, Major_Functional_Group, year) %>%
  summarise(
    n_records = n(),
    .groups = "drop"
  )

table(Genera_summary$Major_Functional_Group)
table(Genera_summary$estimated_genus_name)
# Great! No freshwater taxa here.

# Count how many years have more than 10 records; remove those genera with fewer than 10 records; remove unknown genera; remove genera with fewer than 10 years of data:

Genera_filtered <- Genera_summary %>%
  # remove those genera with fewer than 10 years of 10 records:
  filter(n_records > 10) %>%
  
  # # remove freshwater species:
  # filter(!Major_Functional_Group %in% FreshwaterTaxa) %>%
  
  # remove unknown genera:
  filter(!estimated_genus_name %in% c("unknown", "NEI")) %>%
  # remove genera with fewer than 10 years of data:
  group_by(estimated_genus_name, Major_Functional_Group) %>%
  summarise(n_years = n(), .groups = "drop") %>%
  filter(n_years > 10)
# 99 obs. / genera and 3 vars.

Genera_of_Interest<- Genera_filtered$estimated_genus_name
```

```{r Filter dataset to generate time series for Genera_of_Interest}
# Keep only genera that meet the criteria in the main dataset
GOC_Landings_cmsy <- GOC_Landings_marine %>%
  filter(estimated_genus_name %in% Genera_of_Interest)
# 2359797 obs., 61 vars.
```

```{r Generate time series}
colnames(GOC_Landings_cmsy)

CMSY.Landings_summary <- GOC_Landings_cmsy %>%
  # Ensure genus is not NA
  filter(!is.na(estimated_genus_name)) %>%
  group_by(year, estimated_genus_name, Major_Functional_Group) %>%
  summarise(
    total_biomass_kg = sum(weight_kg_corrected, na.rm = TRUE),  # total biomass
    total_revenue    = sum(value, na.rm = TRUE),               # total revenue
    .groups = "drop"
  ) %>%
  arrange(year, estimated_genus_name)

# View the summary
print(summary_table)
```

```{r Save these time series CMSY.Landings_summary}
write.csv(CMSY.Landings_summary, row.names= FALSE, file= "/home/ferrer/ohi_goc_prep/FP/CMSY_Analysis/CMSY_LandingsPerGenera_Apr022026.csv")
```

# COME BACK TO THIS TOMORROW: It's not wrong but it's messy. :)
```{r Create a summary table with genera per Major_Taxonomic_Group}
colnames(CMSY.Landings_summary)

unique(CMSY.Landings_summary$estimated_genus_name)

CMSY.Landings_summary$estimated_genus_name <- trimws(CMSY.Landings_summary$estimated_genus_name)
CMSY.Landings_summary$estimated_genus_name <- toupper(CMSY.Landings_summary$estimated_genus_name)

GeneraPerGroup_table <- CMSY.Landings_summary %>%
  select(Major_Functional_Group, estimated_genus_name) %>%
  distinct() %>%  # keep only unique combinations
  arrange(Major_Functional_Group, estimated_genus_name)  # sort by functional group and genus

# Create publication-ready table
Genera_Publication <- CMSY.Landings_summary %>%
  select(Major_Functional_Group, estimated_genus_name) %>%
  distinct() %>%                                 # remove exact duplicate rows
  group_by(estimated_genus_name) %>%             # group by genus
  summarise(
    Functional_Groups = paste(Major_Functional_Group, collapse = " | ")
  ) %>%
  ungroup() %>%
  arrange(Functional_Groups)                  # sort alphabetically

# View the table
Genera_Publication

library(knitr)
library(kableExtra)

# Simple table with borders
Genera_Publication %>%
  kable("html", caption = "Genera by Functional Group") %>%
  kable_styling(full_width = FALSE, position = "center")

write.csv(Genera_Publication, row.names= FALSE, file= "/home/ferrer/ohi_goc_prep/FP/CMSY_Analysis/CMSY_GeneraTable_Apr022026.csv")
```

# Okay! Timeseries are ready for CMSY.



Other notes:
Might be worth regenerating MasterSpKey with normalized unicode (stri_trans_nfkc). Although, to be perfectly honest, I don't think it will affect out fishery estimates at all if we code by genus.
Okay, I did this.

STOPPED HERE, April 2 2026