################################################################################ 

For questions contact: Erica Ferrer, Ph.D. [emferrer\@ucsd.edu](mailto:emferrer@ucsd.edu){.email} - OR - [ferrer\@nceas.ucsb.edu](mailto:ferrer@nceas.ucsb.edu){.email}

################################################################################ 

What's different about this key from April's key?

~108 lines effectively unchanged (copy-paste from April).
4 lines minor comment/header edits.
23 lines deleted (STEP 15).
~51 lines added (STEP 3b2 + one reference line).
~95 lines rewritten (merge, export, 3c–3d, next steps).
~147 lines in May that don’t match April line-for-line (adds + rewrites + line shift from deletion).


# PRE-WORK ------------------------------

```{r Clear environment}
rm(list=ls(all=T))
```

```{r Load Libraries & Set WD}
# Read In
library(here)
library(readxl)
library(readr)
# Data Processing
library(tidyverse)
library(dplyr)
library(tidytext)  # for reorder_within and scale_x_reordered
library(lubridate)
library(janitor)
library(stringi)
library(stringr)
library(data.table)
library(zoo) # used for gapfilling RAM data.
library(fuzzyjoin)
library(sf) # for work on shape fines
# Visualization
library(ggplot2)
library(ggforce)
library(plotly)
library(htmlwidgets)
library(htmltools)


setwd("/home/ferrer/ohi_goc_prep/FP/")
```

## clean_names function
```{r Load clean_names function}
# Takes text and standardizes it, which we will need for merging later on:

clean_names <- function(x) {

  x <- iconv(x, from = "", to = "UTF-8", sub = "")
  x <- stringi::stri_trans_general(x, "Latin-ASCII")

  x %>%
    toupper() %>%
    trimws() %>%
    gsub("\\bDE\\b", "", .) %>%
    gsub("\\s*\\([^\\)]*\\)", "", .) %>%   # remove parentheses + contents
    gsub("\\s+", " ", .) %>%
    trimws()
}
```

# STEP 0: Load in relevant fisheries landings data ------------------------------

```{r Load the Mexican Pacific - Mex.Pacific_Landings0}
Mex.Pacific_Landings0<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/Mex.Pacific_Landings_2000.present_May042026.rds"))

dim(Mex.Pacific_Landings0)
# 6406303 obs., 21 vars.
```

```{r States_of_Interest}
unique(Mex.Pacific_Landings0$state)

States_of_Interest<- c("BAJA CALIFORNIA", "BAJA CALIFORNIA SUR",
                       "SONORA", "SINALOA",
                       "NAYARIT", "JALISCO")
```

```{r Generate NW.Mex_Landings0 dataframe}
NW.Mex_Landings0<- Mex.Pacific_Landings0 %>%
  filter(state %in% States_of_Interest)

dim(NW.Mex_Landings0)
# 4713144 obs., 21 obs.
```

```{r Clean the landing_site and office_name_clean names & create NW.Mex_Landings dataframe}
# Clean the landing_site names in NW.Mex_Landings: (est. 1 minute run time)
NW.Mex_Landings<- NW.Mex_Landings0 %>%
  mutate(landing_site_clean= clean_names(landing_site),
         office_name_clean= clean_names(office_name)) %>%
  relocate(landing_site_clean, .after= landing_site) %>%
  relocate(office_name_clean, .after= office_name)
# 4713144 obs., 23 obs.
```


# FIRST TEST: How well does this revised key (Hems_LatLong_Site.Office.Key) work on its own, no alterations?

# STEP 1: Test how good is Hems_LatLong_Site.Office.Key for matching with no additional edits 
Load in Hem's catch-location lat-long key (georeflocskm)

```{r Load and prepare georeflocskm}
load("/home/shares/ohi/OHI_GOC/_raw_data/CONAPESCA/d2025/SharedByHemMorzaria/permit_data/georeflocskm.rda")
Hems_LatLongKey0 <- georeflocskm
#rm(georeflocskm)
```

```{r Check the CRS of georeflocskm}
# sf::st_geometry_type(georeflocskm$geometry)
# Geometry not an issue here.
```


```{r Clean Hem's key}
# Check column names first:
colnames(Hems_LatLongKey0) <- c(
  "state", "landing_site", "geometry", "lon_deci", "lat_deci",
  "fishery_office", "fishery_office_ent", "distance", "distance_km"
)

# Clean the landing_site names:
Hems_LatLongKey_clean <- Hems_LatLongKey0 %>%
  mutate(landing_site_clean= clean_names(landing_site),
         office_name_clean= clean_names(fishery_office))

# Re-arrange columns:
colnames(Hems_LatLongKey_clean)

Hems_LatLongKey_clean<- Hems_LatLongKey_clean %>%
  select(state, landing_site, landing_site_clean, fishery_office, office_name_clean, lon_deci, lat_deci, geometry, distance, distance_km)
```

```{r Create Hems_LatLong_Site.Office.Key}
# Compress Hems_LatLongKey_clean by state-office_name_clean-landing_site_clean matches:
Hems_LatLong_Site.Office.Key <- Hems_LatLongKey_clean %>%
  group_by(state, landing_site_clean, office_name_clean) %>%
  summarise(
    site.office_mean_lon = mean(lon_deci, na.rm = TRUE),
    site.office_mean_lat = mean(lat_deci, na.rm = TRUE),
    site.office_lon_range = max(lon_deci, na.rm = TRUE) - min(lon_deci, na.rm = TRUE),
    site.office_lat_range = max(lat_deci, na.rm = TRUE) - min(lat_deci, na.rm = TRUE),
    n_site.offices = n(),
    .groups = "drop"
  ) %>%
  
  mutate(
    confidence_in_office.site_latlon = case_when(

      site.office_lon_range > 2 |
      site.office_lat_range > 2 ~
        "low; lat or lon estimates meaned at the state-office-site level differ by >2º in the georeflocskm key",

      site.office_lon_range > 1 |
      site.office_lat_range > 1 ~
        "medium; lat or lon estimates meaned at the state-office-site level differ by 1-2º in the georeflocskm key",

      site.office_lon_range > 0.5 |
      site.office_lat_range > 0.5 ~
        "medium-high; lat or lon estimates meaned at the state-office-site level differ by 0.5-1º in the georeflocskm key",

      TRUE ~
        "high; high coherence among lat-lon estimates meaned at the state-office-site level"
    )
  )

# 30276 obs., 9 vars.


table(Hems_LatLong_Site.Office.Key$confidence_in_office.site_latlon)
# high; high coherence
# 28886 
# medium-high; lat or lon estimates differ by 0.5-1º in the georeflocskm key 
# 900 
# medium; lat or lon estimates differ by 1-2º in the georeflocskm key 
# 443
# low; lat or lon estimates differ by >2º in the georeflocskm key
# 47

# Clean up / simplify the confidence column:
Hems_LatLong_Site.Office.Key <- Hems_LatLong_Site.Office.Key %>%
  mutate(
    lat.lon_confidence_simple =
      sub(";.*", "", confidence_in_office.site_latlon)
  )

# Add source column:
Hems_LatLong_Site.Office.Key$lat.lon_source<- "Hem's state.site.office key"

table(Hems_LatLong_Site.Office.Key$lat.lon_confidence_simple)
      #  high         low      medium medium-high 
      # 28886          47         443         900 
```

```{r Test: how good is Hems_LatLong_Site.Office.Key for matching with no additional edits?}
# Does the extra work actually improve the matching? 

NW.Mex_Landings_matched1<- NW.Mex_Landings %>%
  left_join(
    Hems_LatLong_Site.Office.Key,
    by = c("state", "office_name_clean", "landing_site_clean")
  )

# 4713144 obs., 29 obs.
# Excellent! This worked, but...

# SANITY CHECK: How many entries are unmatched after the merge?
unmatched<- NW.Mex_Landings_matched1 %>%
  filter(is.na(site.office_mean_lat))
# 2607660 obs. are unmatched (55%), so this strategy ALONE does not work well.

length(unmatched$site.office_mean_lat) / length(NW.Mex_Landings_matched1$site.office_mean_lat)

# Let's try to understand sites didnt match:
unmatched_slice<- unmatched %>%
  slice_sample(n= 200)
# I see that there could be several entries captured with a simple state-landing_site_clean match.
```

# >> FIRST OUTCOME: Use of the Hems_LatLong_Site.Office.Key does not work that well. Let's try a few simple slice, dices, and creative merges that may improve matching.


# STEP 2: Split Hem's key down to the level of landing_site_clean to improve matching  ------------------------------

```{r Compress Hems_LatLong_Site.Office.Key even further to match by state and landing_site_clean}
# Compress by state and landing_site_clean:
Hems_LatLong_Site.Key0 <- Hems_LatLong_Site.Office.Key %>%
  group_by(state, landing_site_clean) %>%
  summarise(
    site_mean_lon = mean(site.office_mean_lon, na.rm = TRUE),
    site_mean_lat = mean(site.office_mean_lat, na.rm = TRUE),

    site_lon_range = max(site.office_mean_lon, na.rm = TRUE) - min(site.office_mean_lon, na.rm = TRUE),
    site_lat_range = max(site.office_mean_lat, na.rm = TRUE) - min(site.office_mean_lat, na.rm = TRUE),
    n_sites = n(),
    
    .groups = "drop"
  ) %>%
  
  mutate(
    confidence_in_site_latlon = case_when(

      site_lon_range > 2 |
      site_lat_range > 2 ~
        "low; lat or lon estimates meaned at the state-site level differ by >2º in the georeflocskm key",

      site_lon_range > 1 |
      site_lat_range > 1 ~
        "medium; lat or lon estimates meaned at the state-site level differ by 1-2º in the georeflocskm key",

      site_lon_range > 0.5 |
      site_lat_range > 0.5 ~
        "medium-high; lat or lon estimates meaned at the state-site level differ by 0.5-1º in the georeflocskm key",

      TRUE ~
        "high; high coherence among lat-lon estimates meaned at the state-site level"
    )
  )

#  25130 obs., 13 vars. (fewer rows than before, good!)

# Clean up / simplify the confidence column:
Hems_LatLong_Site.Key0 <- Hems_LatLong_Site.Key0 %>%
  mutate(
    lat.lon_confidence_simple =
      sub(";.*", "", confidence_in_site_latlon)
  )
```

```{r Sequential second merge}
# Join using the more specific key first: Hems_LatLong_Site.Office.Key
# Join again using the fallback key: Hems_LatLong_Site.Key
# Only use fallback lat/long values where the first join did not populate them.

colnames(Hems_LatLong_Site.Office.Key)
colnames(Hems_LatLong_Site.Key0)

NW.Mex_Landings_matched2 <- NW.Mex_Landings_matched1 %>%
  
  # Second merge with first fallback key (Hems_LatLong_Site.Key)
  left_join(
    Hems_LatLong_Site.Key0 %>%
      distinct(),
    
    by = c("state", "landing_site_clean")
  )

# 4713144 obs., 35 obs. (nice!!)

# SANITY CHECK: How many entries are unmatched after the merge?
unmatched<- NW.Mex_Landings_matched2 %>%
  filter(is.na(site_mean_lon))

dim(unmatched)
# 1444423 (31%) of entries are still unmatched; can we do better?

length(unmatched$site.office_mean_lat) / length(NW.Mex_Landings_matched2$site.office_mean_lat)

# Let's try to understand sites didn't match:
unmatched_slice<- unmatched %>%
  slice_sample(n= 200)
# I see that there could be several entries captured with a state-office matches.

# SANITY CHECK: the sites that did match this time, what do those look like?
matched_slice<- NW.Mex_Landings_matched2 %>%
  filter(is.na(site.office_mean_lon) & !is.na(site_mean_lon)) %>%
  slice_sample(n= 200)
# Neat! These look good an make sense. For example, in my data I find the landing_site_clean == CORTINA LA PRESA ADOLFO LOPEZ MATEOS (office_name_clean == CULIACAN, state == SINALOA), with no exact match in Hem's key. But when I search only by state and landing_site_clean, I find a match!
```
# >> SECOND OUTCOME: This key works okay, but can probably do even better.


# STEP 3: Split Hem's key down to the level of office_name_clean to improve matching ------------------------------

```{r Compress Hems_LatLong_Site.Office.Key even further to match by state and office_name_clean}
# Compress by office_name_clean
Hems_LatLong_Office.Key <- Hems_LatLong_Site.Office.Key %>%
  group_by(state, office_name_clean) %>%
  summarise(
    office_mean_lon = mean(site.office_mean_lon, na.rm = TRUE),
    office_mean_lat = mean(site.office_mean_lat, na.rm = TRUE),
    
    office_lon_range = max(site.office_mean_lon, na.rm = TRUE) - min(site.office_mean_lon, na.rm = TRUE),
    office_lat_range = max(site.office_mean_lat, na.rm = TRUE) - min(site.office_mean_lat, na.rm = TRUE),
    n_offices = n(),
    .groups = "drop"
  ) %>%
  
  mutate(
    confidence_in_office_latlon = case_when(

      office_lon_range > 2 |
      office_lat_range > 2 ~
        "low; lat or lon estimates meaned at the state-office level differ by >2º in the georeflocskm key",

      office_lon_range > 1 |
      office_lat_range > 1 ~
        "medium; lat or lon estimates meaned at the state-office level differ by 1-2º in the georeflocskm key",

      office_lon_range > 0.5 |
      office_lat_range > 0.5 ~
        "medium-high; lat or lon estimates meaned at the state-office level differ by 0.5-1º in the georeflocskm key",

      TRUE ~
        "high; high coherence among lat-lon estimates meaned at the state-office level"
    )
  )

#  73 obs., 8 vars. (smaller than before, good!)

# Clean up / simplify the confidence column:
Hems_LatLong_Office.Key <- Hems_LatLong_Office.Key %>%
  mutate(
    lat.lon_confidence_simple =
      sub(";.*", "", confidence_in_office_latlon)
  )

# Add source column:
Hems_LatLong_Office.Key$lat.lon_source<- "Hem's state.office key"
```

```{r Sequential third merge}
# The idea is:
# Join using Hem's specific key first.
# Join using the fallback key ()
# Only use fallback lat/long values where the first join did not populate them.

colnames(Hems_LatLong_Site.Office.Key)
colnames(Hems_LatLong_Site.Key0)
colnames(Hems_LatLong_Office.Key)

NW.Mex_Landings_matched3 <- NW.Mex_Landings_matched2 %>%
  
  # Third merge with second fallback key (Hems_LatLong_Office.Key):
  left_join(
    Hems_LatLong_Office.Key %>%
      distinct(),
    by = c("state", "office_name_clean")
  )
  
# 4713144 obs., 41 obs. (nice!!)

# SANITY CHECK: the sites that did match this time, what do those look like?
colnames(NW.Mex_Landings_matched3)
matched_slice<- NW.Mex_Landings_matched3 %>%
  filter(is.na(site.office_mean_lat) & is.na(site_mean_lat) & !is.na(office_mean_lat)) %>%
  slice_sample(n= 200)
# Great! This seems to have worked for state-landing_sites not found in Hem's key. Take this example:
# landing_site_clean== MAICULIS, office_name_clean == EL FUERTE, state == SINALOA... I can't find this match in Hem's key but I can find EL FUERTE in SINALOA.
```

```{r Fill missing values from fallback merges & clean up}
# Note that the coalesce function works from left to right, so R will use site.office_mean_lon if not NA, otherwise site_mean_lon, otherwise office_mean_lon
colnames(NW.Mex_Landings_matched3)

NW.Mex_Landings_matched3 <- NW.Mex_Landings_matched3 %>%
  mutate(
    est_lon = coalesce(
      site.office_mean_lon,
      site_mean_lon,
      office_mean_lon
    ),

    est_lat = coalesce(
      site.office_mean_lat,
      site_mean_lat,
      office_mean_lat
    ),

    lat.lon_source = case_when(
      !is.na(site.office_mean_lon) ~ "estimate at the state-office-site level",
      !is.na(site_mean_lon) ~ "estimate at the state-site level",
      !is.na(office_mean_lon) ~ "estimate at the state-office level",
      TRUE ~ NA_character_
    ),
    
      lat.lon_confidence = coalesce(
      confidence_in_office.site_latlon,
      confidence_in_site_latlon,
      confidence_in_office_latlon
    )
  )

dim(NW.Mex_Landings_matched3)
# 4713144 obs., 45 vars.

# Check lat-long estimate sources:
table(NW.Mex_Landings_matched3$lat.lon_source)
     # estimate at the state-office level estimate at the state-office-site level 
     #                            1108040 (24.5%)                    2105484 (44.6%)
     #   estimate at the state-site level 
     #                            1163237 (24.6%)

sum(is.na(NW.Mex_Landings_matched3$lat.lon_source))
# 336383 unmatched (7.1%)

# Check confidence values:
table(NW.Mex_Landings_matched3$lat.lon_confidence)
     # estimate at the state-office level estimate at the state-office-site level 
     #                            1108040 (24.5%)                    2105484 (44.6%)
     #   estimate at the state-site level 
     #                            1163237 (24.6%)

# Clean up / simplify the confidence column:
NW.Mex_Landings_matched3 <- NW.Mex_Landings_matched3 %>%
  mutate(
    lat.lon_confidence_simple =
      sub(";.*", "", lat.lon_confidence)
  )

table(NW.Mex_Landings_matched3$lat.lon_confidence_simple)
    #    high             low              medium           medium-high 
    # 3105807 (66%)       248691 (5%)      808243 (17%)     214020 (5%)
# Looks pretty good, yay!
```

```{r How many entries are unmatched now?}
# SANITY CHECK: How many entries are unmatched after THIS merge?
unmatched<- NW.Mex_Landings_matched3 %>%
  filter(is.na(est_lon))
# Only 336383 (7%) of entries are still unmatched.

# Let's try to understand sites didn't match:
unmatched_slice<- unmatched %>%
  slice_sample(n= 200)
```

```{r plot unmatched entries over time}
table(unmatched$year)

unmatched_summary <- unmatched %>%
  group_by(year) %>%
  summarise(n_count = n())

# Plot these missing values over time:
unmatched_plot1<- ggplot(unmatched_summary, aes(x = year, y = n_count)) +
  geom_line() +
  geom_point() +
  labs(
    x = "Year",
    y = "Count of Unmatched",
    title = "Unmatched Records Over Time"
  ) +
  theme_bw()

unmatched_plot1
```

# >> THIRD OUTCOME: This key works pretty well! Only missing lat-longs for 7% of entries, and the majority of entries are high confidence.


# STEP 4: Add information from Mel's keys to supplement missing and low-confidence values from Hem's keys: integrated_LatLong_Site.Key ------------------------------
Merge Hem's Site.Key and Mel's Site.Key to create integrated_LatLong_Site.Key

```{r Load and clean Mel's Baja + BCS + Sonora key}
# Load in Mel's Baja-Sonora key:
BajaSonora_LatLongKey<- read_csv("/home/frazier/ohi_goc_prep/FP/Exploratory_Research/FIS_TeamResearch/data/Baja_Sonora_states_key.csv")
# 906 obs., 6 vars.

# Cleaning:
colnames(BajaSonora_LatLongKey) <- c("state", "landing_site", "OHI_rgn", "lat", "long", "GoC")

# Normalize state and landing_site names:
BajaSonora_LatLongKey<- BajaSonora_LatLongKey %>%
  mutate(state = clean_names(state),
         landing_site_clean= clean_names(landing_site))
  
# Convert lat-long entries to numeric entries, accounting for unique cases where the minus (-) is actually an n dash or m dash:
BajaSonora_LatLongKey_clean <- BajaSonora_LatLongKey %>%
  mutate(
    long = as.numeric(str_replace_all(long, "[−–—]", "-")),
    lat  = as.numeric(str_replace_all(lat, "[−–—]", "-"))
  )

# Now assign mean values for compressed key:
BajaSonora_LatLongKey_clean <- BajaSonora_LatLongKey_clean %>%
  
  filter(!is.na(landing_site_clean)) %>%
  
  group_by(state, landing_site_clean) %>%
  summarise(
    mean_lon = mean(long, na.rm = TRUE),
    mean_lat = mean(lat, na.rm = TRUE),
    
    lon_range = max(long, na.rm = TRUE) - min(long, na.rm = TRUE),
    lat_range = max(lat, na.rm = TRUE) - min(lat, na.rm = TRUE),
    n_sites = n(),
    
    GoC = na.omit(GoC)[1], #keep one value for GoC
    
    .groups = "drop"
  ) %>%
  
  mutate(
    confidence_in_latlon = case_when(

      lon_range > 2 |
      lat_range > 2 ~
        "low; lat or lon estimates meaned at the state-site level differ by >2º in Mel's key",

      lon_range > 1 |
      lat_range > 1 ~
        "medium; lat or lon estimates meaned at the state-site level differ by 1-2º in Mel's key",

      lon_range > 0.5 |
      lat_range > 0.5 ~
        "medium-high; lat or lon estimates meaned at the state-site level differ by 0.5-1º in Mel's key",

      TRUE ~
        "high; high coherence among lat-lon estimates meaned at the state-site level"
    )
  )

# 858 obs., 9 vars.
```

```{r Load and clean Mel's Jalisco key}
# Load in Mel's Jalisco key:
Jalisco_LatLong.Key<- read_csv(here("/home/frazier/ohi_goc_prep/FP/Exploratory_Research/FIS_TeamResearch/data/landing_sites_jalisco.csv"))
# 293 obs., 8 vars.

# Cleaning:
colnames(Jalisco_LatLongKey) <- c("landing_site", "landing_site_key", "est_habitat", "lat", "long", "site_notes", "OHI_rgn", "state")

# Normalize state and landing_site names:
Jalisco_LatLong.Key<- Jalisco_LatLong.Key %>%
  mutate(state = clean_names(state),
         landing_site_clean= clean_names(landing_site)) %>%
  relocate(state, .before= landing_site) %>%
  relocate(landing_site_clean, .after= landing_site) 
  
# Convert lat-long entries to numeric entries, accounting for unique cases where the minus (-) is actually an n dash or m dash:
Jalisco_LatLong.Key_clean <- Jalisco_LatLong.Key %>%
  mutate(
    long = as.numeric(str_replace_all(long, "[−–—]", "-")),
    lat  = as.numeric(str_replace_all(lat, "[−–—]", "-"))
  )

# Now assign mean values for compressed key:
Jalisco_LatLong.Key_clean <- Jalisco_LatLong.Key %>%
  
  mutate(
    long = as.numeric(str_replace_all(as.character(long), "[−–—]", "-")),
    lat  = as.numeric(str_replace_all(as.character(lat), "[−–—]", "-"))
  ) %>%
  
  filter(
    !is.na(landing_site_clean),
    !is.na(long),
    !is.na(lat)
  ) %>%
  
  group_by(state, landing_site_clean) %>%

  summarise(
    mean_lon = mean(long, na.rm = TRUE),
    mean_lat = mean(lat, na.rm = TRUE),
    lon_range = max(long, na.rm = TRUE) - min(long, na.rm = TRUE),
    lat_range = max(lat, na.rm = TRUE) - min(lat, na.rm = TRUE),
    n_sites = n(),
    
    est_habitat = na.omit(est_habitat)[1], #keep one value for est_habitat
    site_notes  = na.omit(site_notes)[1], #keep one value for site_notes
  
    .groups = "drop"
  ) %>%
  
  mutate(
    confidence_in_latlon = case_when(

      lon_range > 2 |
      lat_range > 2 ~
        "low; lat or lon estimates meaned at the state-site level differ by >2º in Mel's key",

      lon_range > 1 |
      lat_range > 1 ~
        "medium; lat or lon estimates meaned at the state-site level differ by 1-2º in Mel's key",

      lon_range > 0.5 |
      lat_range > 0.5 ~
        "medium-high; lat or lon estimates meaned at the state-site level differ by 0.5-1º in Mel's key",

      TRUE ~
        "high; high coherence among lat-lon estimates meaned at the state-site level"
    )
  )

# 258 obs., 8 vars.

# Add the "GoC" helpful column here based on a latitude cut off for the Nayarit/Jalisco OHI region
# Region 9 starts at latitude= 20.40895:
Jalisco_LatLong.Key_clean <- Jalisco_LatLong.Key_clean %>%
  mutate(
    GoC = case_when(
      mean_lat < 20.40895 ~ "no",
      mean_lat > 20.40895 ~ "yes",
      TRUE ~ NA_character_
    )
  )
```

```{r Merge Mel's Baja + Sonora key with her Jalisco key}
Mels_LatLong_Site.Key0 <- bind_rows(BajaSonora_LatLongKey_clean, Jalisco_LatLong.Key_clean) %>%
  # remove the NA from this data frame as this will create a many-to-many matching issue:
  filter(!is.na(landing_site_clean)) %>%  
  mutate(mean_lon = as.numeric(mean_lon),
         mean_lat = as.numeric(mean_lat))
# 1116 obs., 11 vars.

# Take a look
unique(Mels_LatLong_Site.Key0$state)

# Remove unused data frames:
# rm(BajaSonora_LatLong.Key, Jalisco_LatLong.Key, removed_rows)

# Clean up / simplify the confidence column:
Mels_LatLong_Site.Key0 <- Mels_LatLong_Site.Key0 %>%
  mutate(
    lat.lon_confidence_simple =
      sub(";.*", "", confidence_in_latlon)
  )

table(Mels_LatLong_Site.Key0$lat.lon_confidence_simple)
       # high         low      medium medium-high 
       # 1093          12           5           6 
```

```{r Clean and prepare both Hems_LatLong_Site.Key0 and Mels_LatLong_Site.Key0 for merging}
colnames(Hems_LatLong_Site.Key0)
colnames(Mels_LatLong_Site.Key0)

# Shave down Mel's key:
Mels_LatLong_Site.Key<- Mels_LatLong_Site.Key0 %>%
  select(state, landing_site_clean,
         site_mean_lon= mean_lon, site_mean_lat= mean_lat, 
         site_lon_range= lon_range, site_lat_range= lat_range, n_sites, 
         confidence_in_site_latlon= confidence_in_latlon, lat.lon_confidence_simple)
# 1116 obs., 9 variables

colnames(Hems_LatLong_Site.Key0)
colnames(Mels_LatLong_Site.Key)

# Step 1: rename confidence so we can compare
Hems_LatLong_Site.Key <- Hems_LatLong_Site.Key0 %>%
  mutate(source = "hem") %>%
  rename(conf_hem = confidence_in_site_latlon)

Mels_LatLong_Site.Key <- Mels_LatLong_Site.Key %>%
  mutate(source = "mel") %>%
  rename(conf_mel = confidence_in_site_latlon)

# Step 2: join Mel onto Hems
integrated_LatLong_Site.Key0 <- bind_rows(
  Hems_LatLong_Site.Key,
  Mels_LatLong_Site.Key
)
# 26246 obs., 11 vars.

# Step 3: create a confidence ranking
confidence_rank <- c(
  "low" = 1,
  "medium" = 2,
  "medium-high" = 3,
  "high" = 4
)

integrated_LatLong_Site.Key0 <- integrated_LatLong_Site.Key0 %>%
  mutate(conf_rank = confidence_rank[lat.lon_confidence_simple])

colnames(integrated_LatLong_Site.Key0)
```

```{r After the merge - decide which source wins}
integrated_LatLong_Site.Key <- integrated_LatLong_Site.Key0 %>%
  select(state, landing_site_clean, source, conf_rank, 
         lat.lon_confidence_simple, site_mean_lon, site_mean_lat) %>%
  tidyr::pivot_wider(
    names_from = source,
    values_from = c(conf_rank, lat.lon_confidence_simple, site_mean_lon, site_mean_lat),
    names_sep = "_"
  )

# Step 4: decide which source wins:
# overwrite Hem's estimate ONLY when Mel's estimate is better
integrated_LatLong_Site.Key <- integrated_LatLong_Site.Key %>%
  mutate(
    hem_rank = coalesce(conf_rank_hem, 0),
    mel_rank = coalesce(conf_rank_mel, 0),

    mel_wins = mel_rank > hem_rank,
    
    # Add source column:
    lat.lon_source = if_else(mel_wins, "Mel's state.site key", "Hem's state.site key"),
    
    final_lat = case_when(
      mel_wins ~ site_mean_lat_mel,
      TRUE     ~ site_mean_lat_hem
    ),

    final_lon = case_when(
      mel_wins ~ site_mean_lon_mel,
      TRUE     ~ site_mean_lon_hem
    ),

    final_conf_rank = if_else(mel_wins, conf_rank_mel, conf_rank_hem),

    final_conf = if_else(
      mel_wins,
      lat.lon_confidence_simple_mel,
      lat.lon_confidence_simple_hem
    )
  )
# 25644 obs., 18 vars.

# SANITY CHECKS: does the number of times mel_wins == lat.lon_source
table(integrated_LatLong_Site.Key$mel_wins)
table(integrated_LatLong_Site.Key$lat.lon_source)
# Yes! Mel's key trumps or supplements hem's 641 times out of 25644 instances (2%)

# Add back the GOC details:
colnames(integrated_LatLong_Site.Key)
colnames(Mels_LatLong_Site.Key0)

# Shave down Mel's key A LOT:
added_details<- Mels_LatLong_Site.Key0 %>%
  select(state, landing_site_clean,
         GoC, est_habitat, site_notes
         )
# 1116 obs., 5 variables

 integrated_LatLong_Site.Key<- integrated_LatLong_Site.Key %>%
   left_join(
    added_details,
    by = c("state", "landing_site_clean")
  )
# 25644 obs., 21 vars.

table(integrated_LatLong_Site.Key$GoC)
#  no yes 
# 567 548 
# Excellent!
```


# STEP 5: Now create a "master" key ------------------------------

```{r Create a backbone for my master key}
# Create a backbone to merge to...
colnames(Mex.Pacific_Landings0)

Mex.Pacific_Landings<- Mex.Pacific_Landings0 %>%
  mutate(landing_site_clean= clean_names(landing_site),
         office_name_clean= clean_names(office_name)) %>%
  relocate(landing_site_clean, .after= landing_site) %>%
  relocate(office_name_clean, .after= office_name)
# 6406303 obs., 23 vars.
  
# Create skeleton:
LatLong_merge.skeleton<- Mex.Pacific_Landings %>%
  select(state, office_name_clean, landing_site_clean) %>%
  distinct()
# 5404 distinct state-office-site combos in the Mexican Pacific that I am going to try to fill using information from Hem's keys and / or Mel's key.
```

```{r Join in information from keys sequentially to create LatLong_Master.Key0}
# if LatLong_merge.skeleton has state, office, and landing_site merge in lat and long from Hem_LatLong_
# Site.Office.Key
# if LatLong_merge.skeleton has only state and landing_site merge in lat and long from  integrated_LatLong_
# Site.Key
# if LatLong_merge.skeleton has only state and office merge in lat and long from  Hem_LatLong_Office.Key

LatLong_Master.Key0 <- LatLong_merge.skeleton %>%
  
  # OPTION 1) most specific match: state + office + landing_site
  left_join(
    Hems_LatLong_Site.Office.Key %>%
      select(state, office_name_clean, landing_site_clean,
             lat_site.office = site.office_mean_lat,
             lon_site.office = site.office_mean_lon,
             lat.lon_source_site.office = lat.lon_source),
    by = c("state", "office_name_clean", "landing_site_clean")
  ) %>%
  
  # OPTION 2) fallback: state + landing_site
  left_join(
    integrated_LatLong_Site.Key %>%
      select(state, landing_site_clean, GoC,
             lat_site = final_lat,
             lon_site = final_lon,
             lat.lon_source_site = lat.lon_source),
    by = c("state", "landing_site_clean")
  ) %>%
  
  # OPTION 3) fallback: state + office
  left_join(
    Hems_LatLong_Office.Key %>%
      select(state, office_name_clean,
             lat_office = office_mean_lat,
             lon_office = office_mean_lon,
             lat.lon_source_office = lat.lon_source),
    by = c("state", "office_name_clean")
  ) %>%
  
  mutate(
    final_lat = coalesce(lat_site.office, lat_site, lat_office),
    final_lon = coalesce(lon_site.office, lon_site, lon_office),

    final_lat.lon_source = coalesce(
      lat.lon_source_site.office,
      lat.lon_source_site,
      lat.lon_source_office
    ),
    
    final_lat.lon_level = case_when(
      !is.na(lat_site.office) ~ "state_office_site",
      !is.na(lat_site)        ~ "state_site",
      !is.na(lat_office)      ~ "state_office",
      TRUE                    ~ NA_character_
    )
  )
# 5404 obs., 17 vars.

table(LatLong_Master.Key0$final_lat.lon_level)
     # state_office state_office_site        state_site 
     #          677               591              2374 

# Re-level final_lat.lon_source column:
LatLong_Master.Key0$final_lat.lon_source <- factor(
  LatLong_Master.Key0$final_lat.lon_source,
  levels = c(
    "Hem's state.site.office key",
    "Hem's state.site key",
    "Mel's state.site key",
    "Hem's state.office key"
  )
)

table(LatLong_Master.Key0$final_lat.lon_source)
# Hem's state.site.office key        Hem's state.site key        Mel's state.site key 
#                         591                        1358                        1016 
#      Hem's state.office key 
#                         677 
                        
# SANITY CHECK: In which cases is the the final_lat get pulled at the office level?
TEST<- LatLong_Master.Key0 %>%
  filter(final_lat.lon_level== "state_office") %>%
  slice_sample(n= 200)

table(TEST$final_lat.lon_source)
# Great! These are the cases we would expect.
```

```{r Trim LatLong_Master.Key0}
colnames(LatLong_Master.Key0)

LatLong_Master.Key<- LatLong_Master.Key0 %>%
  select(state, office_name_clean, landing_site_clean, GoC,
           final_lat, final_lon,
           final_lat.lon_level, final_lat.lon_source) %>%
  distinct()
# 5404 obs., 8 vars.

LatLong_Master.Key<- LatLong_Master.Key %>%
  mutate(
    final_lat.lon_source = as.character(final_lat.lon_source)
  )
```

```{r How many entries are unmatched now?}
NW.Mex_Landings_matched4<- NW.Mex_Landings %>%
  left_join(
    LatLong_Master.Key,
    by = c("state", "office_name_clean", "landing_site_clean")
  )

dim(NW.Mex_Landings_matched4)
# 4713144 obs., 28 vars.

sum(is.na(NW.Mex_Landings_matched4$final_lat))
# 237771 unmatched (5%)

# SANITY CHECK: Let's try to understand sites didn't match:
unmatched_slice<- NW.Mex_Landings_matched4 %>%
  filter(is.na(final_lat)) %>%
  slice_sample(n= 200)
# These entries don't make a ton of sense to me. We should be able to find a match at the office level...
```

```{r Final refinement - interpolate office-level entries based on landing site}
# Some sites have not been matched at the office level because Hem's office key missed them, yet sites assigned to the same office are assigned based on the integrated_LatLong_Site.Key. This means that I can use the information from the other sites in that office to say something about the office.

# I meaned across landing_sites to estimate lat-long at the state-office level:
LatLong_Master.Key <- LatLong_Master.Key %>%
  group_by(state, office_name_clean) %>%
  mutate(
    mean_lat = if (all(is.na(final_lat))) NA_real_ else mean(final_lat, na.rm = TRUE),
    mean_lon = if (all(is.na(final_lon))) NA_real_ else mean(final_lon, na.rm = TRUE),

    # identify rows being filled
    filled = is.na(final_lat) & !is.na(mean_lat),

    final_lat = coalesce(final_lat, mean_lat),
    final_lon = coalesce(final_lon, mean_lon),

    final_lat.lon_level = ifelse(
      filled,
      "state_office",
      final_lat.lon_level
    ),
    
    final_lat.lon_source = ifelse(
      filled,
      "Interpolated: I meaned across landing_sites to estimate lat-long at the state-office level",
      final_lat.lon_source
    )
  ) %>%
  select(-mean_lat, -mean_lon, -filled) %>%
  ungroup()
# 5404 obs., 7 vars.

table(LatLong_Master.Key$final_lat.lon_level)
     # state_office state_office_site        state_site 
     #          859               591              2374 
table(LatLong_Master.Key$final_lat.lon_source)

unmatched<- LatLong_Master.Key %>%
  filter(is.na(final_lat))
#   Hem's state.office key 
#   677 
#   Hem's state.site key 
#   1358 
#   Hem's state.site.office key 
#   591 
# Interpolated: I meaned across landing_sites to estimate lat-long at the state-office level 
#   182 
#   Mel's state.site key 
#   1016 

table(unmatched$state)
  # CHIAPAS    COLIMA  GUERRERO MICHOACAN   NAYARIT    OAXACA 
  #     309        95       526       311         1       338 
# Great. With the exception of 1 site in Nayarit, all the sites that have not been georeferenced lie in states definitively outside of the GOC. So I don't need to concern myself with this for now.
```

```{r How about now??}
NW.Mex_Landings_matched4<- NW.Mex_Landings %>%
  left_join(
    LatLong_Master.Key,
    by = c("state", "office_name_clean", "landing_site_clean")
  )

dim(NW.Mex_Landings_matched4)
# 4713144 obs., 45 vars.

sum(is.na(NW.Mex_Landings_matched4$final_lat))
# 1452 unmatched (>1%) Woohoo!! We've been able to georeference essentially every landing!

# SANITY CHECK: Let's try to understand sites didn't match:
unmatched<- NW.Mex_Landings_matched4 %>%
  filter(is.na(final_lat))

table(unmatched$state)
# All in Nayarit, so not a big issue.
```

# >> FOURTH OUTCOME: This key matches almost everything in Northwest Mexico! Unmatched = < 1%. Woohoo!!!

## Save LatLong_Master.Key (INCOMPLETE): 
```{r Save LatLong_Master.Key as csv}
write.csv(LatLong_Master.Key, 
          file= "/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/Spatial_Keys/LatLong_MasterKey_INCOMPLETE_May262026.csv", 
          row.names= FALSE)
```


# STEP 6: Add OHI region to the integrated LatLong_Master.Key ------------------------------
```{r Load in LatLong_Master.Key as needed}
LatLong_Master.Key<- read.csv(here("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/Spatial_Keys/LatLong_MasterKey_INCOMPLETE_May262026.csv"))
```

```{r Loading in OHI regions and buffer shape files}
# OHI regions:
OHI_regions_metadata <- readLines("/home/shares/ohi/OHI_GOC/spatial/OHI_regions/marine/ohi_goc_final_071025.prj")
# CRS: EPSG:4326 in practice, poorly labeled in meta-data.

OHI_regions_shape0 <- sf::st_read(
  "/home/shares/ohi/OHI_GOC/spatial/OHI_regions/marine/ohi_goc_final_071025.shp",
  options = "ENCODING=LATIN1"
)

st_crs(OHI_regions_shape0)

OHI_regions_shape0 <- st_make_valid(OHI_regions_shape0)

# Quick plot:
region_plot<- ggplot(OHI_regions_shape0) +
  geom_sf(aes(fill = rgn_id)) +
  theme_bw()

region_plot

# 50km buffer:
OHI_buffer_metadata <- readLines("/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer/inland_50km_goc.prj")
# Projected CRS name: Mexico_ITRF2008_LCC

OHI_buffer_shape0 <- sf::st_read(
  "/home/shares/ohi/OHI_GOC/spatial/inland_50km_buffer/inland_50km_goc.shp",
  options = "ENCODING=LATIN1"
)

st_crs(OHI_buffer_shape0)

OHI_buffer_shape0 <- st_make_valid(OHI_buffer_shape0)

# Quick plot:
buffer_plot<- ggplot(OHI_buffer_shape0) +
  geom_sf(aes(fill = rgn_id)) +
  theme_bw()

buffer_plot
```

```{r Some quick cleaning}
# Let's use CRS 4326 for all projections now.
crs_use = 4326

# Change CRS on OHI_buffer_shape to match CRS on OHI_regions_shape
OHI_regions_shape <- st_transform(OHI_regions_shape0, crs_use)

# Change CRS on OHI_buffer_shape to match CRS on OHI_regions_shape
OHI_buffer_shape <- st_transform(OHI_buffer_shape0, crs_use)

# Remove columns and extract useful info:
OHI_buffer_shape<- OHI_buffer_shape0 %>%
  select(-c(estado))

additional_info<- OHI_buffer_shape0 %>%
  select(rgn_id, ecorgn) %>%
  st_drop_geometry()

# Add new columns to OHI_regions_shape
OHI_regions_shape <- OHI_regions_shape %>%
  left_join(
    additional_info,
    by = "rgn_id"
  ) %>%
  mutate(
    location= c("sea")
  ) %>%
  relocate(location, .after= rgn_id)

rm(additional_info)
```

```{r Quick check: how long is LatLong_Master.Key with NAs removed?}
TEST<- LatLong_Master.Key %>%
  filter(!is.na(final_lat))
# 3824 obs. with lat-long values. Thus, I expect my shapefile key to also have this many observations.
```

```{r Join geomtric keys}
# Step 0 — define a Coordinate Reference System (CRS) once and stick to it:
crs_use <- 4326

# Alternative could be: CRS 5070 is the USA Contiguous Albers Equal Area (good proxy for northern Mexico too). EPSG:5070 is centered on the U.S., so Baja / Sonora are near the edge of its optimal zone.

# Step 1 — Convert master table to spatial points
# Turn LatLong_Master.Key into an sf object
master_sf <- LatLong_Master.Key %>%
  filter(!is.na(final_lon), !is.na(final_lat)) %>%
  st_as_sf(coords = c("final_lon", "final_lat"), 
           crs = crs_use,
           remove = FALSE) %>%
  st_transform(crs_use)

# Step 2 — Transform polygons ONCE
buffer_sf <- OHI_buffer_shape %>%
  st_transform(crs_use)

sea_sf <- OHI_regions_shape %>%
  st_transform(crs_use)

# Step 3 — Spatial join to land buffer polygons
# Join points → OHI_buffer_shape
master_land <- st_join(
  master_sf,
  buffer_sf,
  join = st_within,
  left = TRUE
)

# Re-name columns to be safe:
master_land <- master_land %>%
  rename(
    rgn_id_land = rgn_id,
    ecorgn_land = ecorgn,
    location_land = location
  )

# Step 5 — Spatial join to sea regions (run time: 2min)
# Join same (or filtered) points → OHI_regions_shape
master_all <- st_join(
  master_land,
  sea_sf,
  join = st_within,
  left = TRUE
)

# Step 6 — Fix duplicate column names cleanly
master_all <- master_all %>%
  rename(
    rgn_id_sea = rgn_id,
    ecorgn_sea = ecorgn,
    location_sea = location
  )
```

```{r SANITY CHECK: did rows get duplicated?}
# Row duplication risk
# If either: buffer polygons overlap, or sea polygons overlap, then st_join() can create:
# Multiple rows per single point
# This is the most common silent failure in spatial pipelines.

# Quick check:
master_all %>%
  st_drop_geometry() %>%
  count(state, office_name_clean, landing_site_clean) %>%
  filter(n > 1)

# Nope!
```

```{r Finishing touches on LatLong_Master.Key}
LatLong_Master.Key_regions <- st_drop_geometry(master_all)
# 3824 obs., 11 vars.

# Coalesce details in to fewer columns:
LatLong_Master.Key_regions <- LatLong_Master.Key_regions %>%
  mutate(
    location = coalesce(location_land, location_sea),
    ecorgn   = coalesce(ecorgn_land, ecorgn_sea),
    rgn_id   = coalesce(rgn_id_land, rgn_id_sea)
  )

colnames(LatLong_Master.Key_regions)

LatLong_Master.Key_regions<- LatLong_Master.Key_regions %>%
  filter(!is.na(final_lat)) %>%
  select(state, office_name_clean, landing_site_clean, GoC,
         final_lat, final_lon, final_lat.lon_level, final_lat.lon_source,
         rgn_id, ecorgn, location)
# 3824 obs., 11 vars.
```

# STEP 7: Integrate GOC and OHI_region flag system & create LatLong_Master.Key_finalized ------------------------------
```{r Assign OHI region and create LatLong_Master.Key_intermediate}
# If the record had a georeferenced location inside a specific OHI region or region buffer (50km), we assigned it to the Gulf of California and the corresponding OHI region (GOC_state.assign= TRUE, OHI_region= Region 1–9).
# If the record did not have a latitude-longitude value inside a specific OHI region or region buffer (50km) but was reported in Sinaloa, we assigned it to Region 8 (GOC_state.assign= TRUE, OHI_region= Region 8).
# If the record did not have a latitude-longitude value inside a specific OHI region or region buffer (50km) but was reported in Nayarit, we assigned it to Region 9 (GOC_state.assign= TRUE, OHI_region= Region 9).
# If the record did not have a latitude-longitude value inside a specific OHI region or region buffer (50km) but was reported in Sonora, we assigned it to the Gulf of California but did not assign it to a specific OHI region (GOC_state.assign= TRUE, OHI_region= NA).
# If the record did not have a latitude-longitude value inside a specific OHI region or region buffer (50km) and was reported in Baja California, Baja California Sur, or Jalisco, we did not assign it to the Gulf of California nor a specific OHI region (GOC_state.assign= NA, OHI_region= NA).

colnames(LatLong_Master.Key_regions)

LatLong_Master.Key_intermediate <- LatLong_Master.Key_regions %>%
  rename(OHI.region_geo.est = rgn_id,
         GOC_mel.est = GoC)

# FIRST: Assign OHI regions
 LatLong_Master.Key_intermediate<- LatLong_Master.Key_intermediate %>%
   mutate(
    
    # Assign based on state
    OHI_state.assign = case_when(
      
      # CASE 1: State is SINALOA or NAYARIT, fallback assignments:
      state == "SINALOA" ~ "Region_8",
      state == "NAYARIT" ~ "Region_9",
      
      # CASE 2: We cannot reliably assign the state to a single region:
      state %in% c(
        "SONORA",
        "BAJA CALIFORNIA",
        "BAJA CALIFORNIA SUR",
        "JALISCO"
      ) ~ NA_character_,
      
      TRUE ~ NA_character_
    )
  ) %>%
  relocate(OHI.region_geo.est, .before= OHI_state.assign)
# 3824 obs., 12 vars.
 
# THEN: Coalesce
# Priviledge the state autoassign, if ever there is a disagreement.
LatLong_Master.Key_intermediate <- LatLong_Master.Key_intermediate %>%
  mutate(
    OHI_region_est = coalesce(OHI_state.assign, OHI.region_geo.est)
  )

table(LatLong_Master.Key_intermediate$OHI_region_est)
sum(is.na(LatLong_Master.Key_intermediate$OHI_region_est))
# 1515 region NAs... Do they make sense?

no_region<- LatLong_Master.Key_intermediate %>%
  filter(is.na(OHI_region_est))

table(no_region$state)
    # BAJA CALIFORNIA BAJA CALIFORNIA SUR             JALISCO              SONORA 
    #             568                 551                 357                  39 
# Yes, these make sense.
```

```{r Assign whether or not it's inside the GOC}
# THEN: Assign whether or not it's inside the GOC

LatLong_Master.Key_intermediate <- LatLong_Master.Key_intermediate %>%
  mutate(
    # Assign based on state:
    GOC_state.assign = case_when(
      
      # CASE 1: States automatically assigned to the GOC:
      state %in% c("SINALOA", 
                   "SONORA",
                   "NAYARIT") ~ "yes",
      
      # CASE 2: States not automatically assigned to the GOC:
      state %in% c(
        "BAJA CALIFORNIA",
        "BAJA CALIFORNIA SUR",
        "JALISCO"
      ) ~ NA_character_,
      
      # everything else
      TRUE ~ NA_character_
    )
  ) %>%
  mutate(
    # Assign based on region:
    GOC_region.assign = case_when(
      
      # # CASE 1: States automatically assigned to the GOC:
      # OHI_region_est %in% c("Region_5",
      #                   "Region_6",
      #                   "Region_7",
      #                   "Region_8",
      #                   "Region_9") ~ "yes",
      # 
      # # CASE 2: Regions that cannot automatically be assigned to the GOC:
      # OHI_region_est %in% c("Region_1",
      #                   "Region_2",
      #                   "Region_3",
      #                   "Region_4") ~ NA_character_,
      
      # If Region # is assign at all, assign it to the GOC (new update as of May 26 2026):
      !is.na(OHI_region_est) ~ "yes",
      
      # everything else
      TRUE ~ NA_character_
    )
    ) %>%
  relocate(GOC_region.assign, .before= GOC_state.assign)


# THEN: Coalesce
# Priviledge the region auto-assign, then state auto-assign, then  mel's assignment if ever there is a disagreement.
LatLong_Master.Key_intermediate <- LatLong_Master.Key_intermediate %>%
  mutate(
    GOC_est = coalesce(GOC_region.assign, GOC_state.assign)
  )

table(LatLong_Master.Key_intermediate$GOC_est)
#  yes 
# 2348 

sum(is.na(LatLong_Master.Key_intermediate$GOC_est))
# 1476 GOC NAs. Great! This supports our OHI region results. 

non_GOC<- LatLong_Master.Key_intermediate %>%
  filter(is.na(GOC_est))

table(non_GOC$state)
    # BAJA CALIFORNIA BAJA CALIFORNIA SUR             JALISCO 
    #             568                 551                 357 
# Yes, these make sense.
```

```{r Investigating weird unexpected results}
TEST<- LatLong_Master.Key_intermediate %>%
  filter(state== "SONORA")

table(TEST$OHI_region_est)
# Region_1 Region_2 Region_4 Region_5 Region_6 Region_7 Region_8 Region_9 
#        2        1        4       62      112      239        5        2 
# The regions that I wouldn't expect actually do make sense when you look at the lat-long values and names for the specific sites. My guess would be they were reported in SONORA but caught elsewhere.
```

## Save LatLong_Master.Key_intermediate (INT): 
```{r Save key as csv}
write.csv(LatLong_Master.Key_intermediate, 
          file= "/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/Spatial_Keys/LatLong_MasterKey_INT_May262026.csv", 
          row.names= FALSE)
# 3824 obs., 16 obs.
```


# STEP 8: Add OHI region names and assign final OHI region and GOC values  ------------------------------
```{r}
LatLong_Master.Key_intermediate<- read.csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/Spatial_Keys/LatLong_MasterKey_INT_May262026.csv")
# 3824 obs., 16 obs.
```

```{r RegionKey_temporary}
colnames(LatLong_Master.Key_intermediate)

RegionKey_temporary<- LatLong_Master.Key_intermediate %>%
  select(state, OHI_region_est, GOC_est) %>%
  distinct(na.rm= TRUE)
```

```{r Save the RegionKey_temporary and work on it offline}
write.csv(RegionKey_temporary,
          file= "CONAPESCA_Keys/Spatial_Keys/RegionKey_TEMP_May262026.csv",
          row.names = FALSE)
```

```{r Re-load in RegionKey_finalized}
RegionKey_finalized<- read.csv("CONAPESCA_Keys/Spatial_Keys/RegionKey_FINAL_May262026.csv")
```

```{r Join in RegionKey_finalized}
colnames(RegionKey_finalized)

LatLong_Master.Key_finalized0 <- LatLong_Master.Key_intermediate %>%
  left_join(
    RegionKey_finalized,
    by = c("state", "OHI_region_est", "GOC_est")
  )

# 3824 obs., 18 obs.

table(LatLong_Master.Key_finalized0$GOC_Include_Exclude)
# exclude include 
#    1476    2348 
```

# STEP 9: Export full annotated LatLong_Master.Key_finalized ------------------------------
```{r Clean up key before export}
# Final sites key:
colnames(LatLong_Master.Key_finalized)

LatLong_Master.Key_finalized<- LatLong_Master.Key_finalized0 %>% 
  select(
    state, office_name_clean, landing_site_clean, 
    final_lat, final_lon, final_lat.lon_level, final_lat.lon_source,
    location, 
    OHI.region_geo.est, OHI_state.assign, OHI_region_est, OHI_region_name,
    GOC_mel.est, GOC_state.assign, GOC_region.assign, GOC_est, GOC_Include_Exclude
  )
# 3824 obs., 17 obs.
```

## Save LatLong_Master.Key_finalized: 
```{r Save key as csv}
write.csv(LatLong_Master.Key_finalized, 
          file= "/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/Spatial_Keys/LatLong_MasterKey_FINAL_May262026.csv", 
          row.names= FALSE)
# 3824 obs., 17 obs.
```


# STEP 10: Attach lat-long & OHI_region details to NW Mexico dataframe ----------------------
co-authors: Erica Ferrer, Melanie Frazier, Hem Nalini Morzaria
 
```{r OHI LatLong_Master.Key_finalized}
LatLong_Master.Key_finalized<- read_csv("/home/ferrer/ohi_goc_prep/FP/CONAPESCA_Scripts_and_Keys/CONAPESCA_Keys/Spatial_Keys/LatLong_MasterKey_FINAL_May262026.csv")
# 3824 obs., 17 obs.
```

```{r Load the sector-sorted Mexican Pacific - Mex.Pacific_Landings0}
Mex.Pacific_Landings0<- readRDS(here("/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/Mex.Pacific_Landings_SectorSorted_2000.present_May042026.rds"))

dim(Mex.Pacific_Landings0)
# 6406303 obs., 56 vars.
```

```{r States_of_Interest}
unique(Mex.Pacific_Landings0$state)

States_of_Interest<- c("BAJA CALIFORNIA", "BAJA CALIFORNIA SUR",
                       "SONORA", "SINALOA",
                       "NAYARIT", "JALISCO")
```

```{r Generate NW.Mex_Landings dataframe}
NW.Mex_Landings0<- Mex.Pacific_Landings0 %>%
  filter(state %in% States_of_Interest)

dim(NW.Mex_Landings0)
# 4713144 obs., 56 obs. As expected, excellent!

NW.Mex_Landings<- NW.Mex_Landings0 %>%
  mutate(landing_site_clean= clean_names(landing_site),
         office_name_clean= clean_names(office_name)) %>%
  relocate(landing_site_clean, .after= landing_site) %>%
  relocate(office_name_clean, .after= office_name)
# 4713144 obs., 58 obs. As expected, excellent!
```

```{r Append lat-long and OHI_region info to landings data}
NW.Mex_Landings_georeferenced0 <- NW.Mex_Landings %>%
  left_join(
    LatLong_Master.Key_finalized,
    by = c("state", "office_name_clean", "landing_site_clean")
  )
# 4713144 obs., 72 obs. As expected, excellent!

table(NW.Mex_Landings_georeferenced0$OHI_region_est)
sum(is.na(NW.Mex_Landings_georeferenced0$OHI_region_est)) # 1230390 NA
# Region_1 Region_2 Region_3 Region_4 Region_5 Region_6 Region_7 Region_8 Region_9 
#    55465    84081   171418   341561   169390   105857   353074  1370119   831789 

table(NW.Mex_Landings_georeferenced0$GOC_est)
#      no     yes (BEFORE MAY 26 UPDATE)
# 1082436 3379164 
#     yes (AFTER MAY 26 UPDATE)
# 3488592 
```

```{r Have a look at entries that are somewhat inconclusive...}
colnames(NW.Mex_Landings_georeferenced0)

# Region 2
TEST<- NW.Mex_Landings_georeferenced0 %>%
  filter(OHI_region_est=="Region_2" & GOC_mel.est== "no")

table(TEST$landing_site_clean)
# ZONA COLORADITO and other colorado sites I would count as within the Gulf...

# Region 3
TEST2<- NW.Mex_Landings_georeferenced0 %>%
  filter(OHI_region_est=="Region_3" & GOC_mel.est== "no")

table(TEST2$landing_site_clean)
# EL CARDON could go either way, SANTA MARIA could go either way...

# Region 4
TEST3<- NW.Mex_Landings_georeferenced0 %>%
  filter(OHI_region_est=="Region_4" & GOC_mel.est== "no")

table(TEST3$landing_site_clean)
# PUNTA BENTONITA not in the GOC but LAS CASITAS very well could be.

# Conclusion: Make a decision, document it, move on. See my final decision in RegionKey_finalized.
```

```{r How many records in the GOC versus not?}
GOC_df<- NW.Mex_Landings_georeferenced0 %>%
  filter(GOC_est == "yes")

sum(is.na(GOC_df$OHI_region_est))
# 5838 obs. have no OHI region assigned to them.

table(GOC_df$state)
# BEFORE MAY 26 UPDATE:
    # BAJA CALIFORNIA BAJA CALIFORNIA SUR             JALISCO             NAYARIT 
    #          175840              318364               95517              785118 
    #         SINALOA              SONORA 
    #         1370011              634314

# AFTER MAY 26 UPDATE:
    # BAJA CALIFORNIA BAJA CALIFORNIA SUR             JALISCO             NAYARIT             SINALOA 
    #          181539              471300               46310              785118             1370011 
    #          SONORA 
    #          634314

outside_GOC_df<- NW.Mex_Landings_georeferenced0 %>%
  filter(is.na(GOC_est))

table(outside_GOC_df$state)
# BAJA CALIFORNIA BAJA CALIFORNIA SUR             JALISCO             NAYARIT 
    #          338615              699829              184656                1452 

# These entries make sense except for Nayarit!
# The Nayarit is not a key issue (I checked). Instead, those sites in Nayarit cannot be estimated to a lat-long so I will need to assign OHI region and GOC manually.
```

```{r Final force of consistency in Nayarit Sonora and Sinaloa}
colnames(NW.Mex_Landings_georeferenced0)

NW.Mex_Landings_georeferenced <- NW.Mex_Landings_georeferenced0 %>%
  mutate(
    
    # Assign based on state:
    GOC_est = case_when(
      
      # CASE 1: States automatically assigned to the GOC:
      state %in% c("SINALOA", 
                   "SONORA",
                   "NAYARIT") ~ "yes",
      
      # Otherwise keep existing value
      TRUE ~ GOC_est
    ),
    
    GOC_Include_Exclude = case_when(
      
      # CASE 1: States automatically assigned to the GOC:
      state %in% c("SINALOA", 
                   "SONORA",
                   "NAYARIT") ~ "include",
      
      # Otherwise keep existing value
      TRUE ~ GOC_Include_Exclude
    ),
    
    OHI_region_est = case_when(
      
      state == "SINALOA" ~ "Region_8",
      state == "NAYARIT" ~ "Region_9",
      
      # Otherwise keep existing value
      TRUE ~ OHI_region_est
    ),
    
    OHI_region_name = case_when(
      
      state == "SINALOA" ~ "Sinaloan_Coast",
      state == "NAYARIT" ~ "Nayarit_Jalisco",
      
      # Otherwise keep existing value
      TRUE ~ OHI_region_name
    )
    
  )

# SANITY CHECK: Do the numbers look better?
GOC_df<- NW.Mex_Landings_georeferenced %>%
  filter(GOC_est == "yes")

sum(is.na(GOC_df$OHI_region_est))
# 5838 obs. have no OHI region assigned to them.

table(GOC_df$state)
# AFTER MAY 26 UPDATE:
    # BAJA CALIFORNIA BAJA CALIFORNIA SUR             JALISCO             NAYARIT             SINALOA 
    #          181539              471300               46310 now 786570 (was 785118) Good!   1370011 
    #          SONORA 
    #          634314
```

```{r Trim down dataframe}
colnames(NW.Mex_Landings_georeferenced)

NW.Mex_Landings_georeferenced<- NW.Mex_Landings_georeferenced %>%
  select(state, office_name, office_name_clean, landing_site, landing_site_clean, landing_site_key,
         year, month, receipt_date_fixed, period_start_fixed, period_end_fixed,
         eu_rnpa, eu_name, vessel_rnpa, vessel_name,
         recorded_taxa_group, recorded_taxa_name, recorded_taxa_name_clean, 
         recorded_scientific_name, scientific_name_recorded_T.F, common_name_clean,
         keywords_1, keywords_2, keywords_3, keywords_4, keywords_5,
         estimated_species_name, species_est_certainty, estimated_genus_name, estimated_family_name,
         landed_weight, live_weight, weight_kg_corrected, value,
         estimated_taxa_group, Major_Functional_Group, Assignment_for_Major_Functional_Group,
         est_habitat,
         reported_aquaculture_production, NOTaquacultureFLAG_specific.groups, aquacultureFLAG_site.details, aquacultureFLAG_keywords, aquacultureFLAG_specific.taxa, aquacultureFLAG_combined, est_aquaculture_production, est_aquaculture_certainty,
         reported_fleet, sector_estimate_reported, sector_estimate_draft,
         final_lat, final_lon, final_lat.lon_level, final_lat.lon_source,
         GOC_est, GOC_Include_Exclude, OHI_region_est, OHI_region_name
         )

# 4713144 obs., 57 var.
```


## SAVE: the NW Mexico landings appended with lat-long info in a separate .rds file:
```{r Save the Northwest Mexico dataframe: May 26 2026}
saveRDS(NW.Mex_Landings_georeferenced, file="/home/shares/ohi/OHI_GOC/goal_prep/fp/cleaned_CONAPESCA/NW.Mex_Landings_SectorSorted.Georef_2000.present_May262026.rds")
# 4713144 obs., 57 var.
```

```{r In conclusion: how many entries aren't in the GOC?}
table(NW.Mex_Landings_georeferenced$GOC_est)
# 3490044 of 4713144 total entries (74%) are assigned to the GOC.

# And of these entries, X% are definitively assigned to a region.
TEST<- NW.Mex_Landings_georeferenced %>%
  filter(!is.na(OHI_region_est))
# 3484206

table(TEST$OHI_region_est)
# Region_1 Region_2 Region_3 Region_4 Region_5 Region_6 Region_7 Region_8 Region_9 
#    55465    84081   171418   341561   169390   105857   353074  1370119   833241 
```

# STEP 10: CRS SENSITIVITY CHECK (optional,don't run this unless necessary)  ------------------------------
```{r SENSITIVITY CHECK: How much does CRS projection matter?}
# Step 0 — Define CRS
crs_wgs84  <- 4326
crs_albers <- 5070

# IMPORTANT: make behavior consistent across pipelines
sf_use_s2(FALSE)   # ensures GEOS is used in BOTH CRS workflows

# Step 1 — Build base point dataset (WGS84)
master_sf <- LatLong_Master.Key %>%
  filter(!is.na(final_lon), !is.na(final_lat)) %>%
  mutate(uid = row_number()) %>%   # stable ID for comparison
  st_as_sf(
    coords = c("final_lon", "final_lat"),
    crs = crs_wgs84,
    remove = FALSE
  )

# Step 2 — Create CRS variants of points
master_4326 <- master_sf
master_5070 <- st_transform(master_sf, crs_albers)

# Step 3 — Transform polygon layers consistently
buffer_4326 <- st_transform(OHI_buffer_shape, crs_wgs84)
buffer_5070 <- st_transform(OHI_buffer_shape, crs_albers)

sea_4326 <- st_transform(OHI_regions_shape, crs_wgs84)
sea_5070 <- st_transform(OHI_regions_shape, crs_albers)


# Step 4 — Spatial joins (4326 pipeline)
land_4326 <- st_join(
  master_4326,
  buffer_4326,
  join = st_within,
  left = TRUE
)

sea_join_4326 <- st_join(
  land_4326,
  sea_4326,
  join = st_within,
  left = TRUE
)


# Step 5 — Spatial joins (5070 pipeline)
land_5070 <- st_join(
  master_5070,
  buffer_5070,
  join = st_within,
  left = TRUE
)

sea_join_5070 <- st_join(
  land_5070,
  sea_5070,
  join = st_within,
  left = TRUE
)


# Step 6 — Convert to comparable tables
res_4326 <- sea_join_4326 %>%
  st_drop_geometry() %>%
  rename(
    rgn_land_4326 = rgn_id.x,
    rgn_sea_4326  = ecorgn.x
  )

res_5070 <- sea_join_5070 %>%
  st_drop_geometry() %>%
  rename(
    rgn_land_5070 = rgn_id.x,
    rgn_sea_5070  = ecorgn.x
  )


# Step 7 — Safe merge by UID (NOT row order)
comparison <- res_4326 %>%
  inner_join(res_5070, by = "uid")


# Step 8 — Measure differences
land_diff <- sum(
  comparison$rgn_land_4326 != comparison$rgn_land_5070,
  na.rm = TRUE
)

sea_diff <- sum(
  comparison$rgn_sea_4326 != comparison$rgn_sea_5070,
  na.rm = TRUE
)

# Great news! These are both zero. Seems like the CRS doesn't matter too much for assining region. 

# Step 8 — Flag mismatches explicitly
comparison <- comparison %>%
  mutate(
    land_match = rgn_land_4326 == rgn_land_5070,
    sea_match  = rgn_sea_4326  == rgn_sea_5070
  )

table(comparison$land_match) # TRUE: 1593
table(comparison$sea_match) # TRUE: 1593
sum(is.na(comparison$sea_match)) # NA: 2231

# No flags either. All looks good!
```

```{r Remove any dataframes we don't want to keep workspace clean}
rm(crs_wgs84, crs_albers, 
   master_4326,
   buffer_4326, buffer_5070, sea_4326, sea_5070,
   land_4326, sea_join_4326, land_5070, sea_join_5070,
   res_4326, res_5070,
   land_diff, sea_diff,
   land_diff_pct, sea_diff_pct,
   comparison)
```


# STEP 11: Other questions (UNDER CONSTRUCTION May 26 2026)
```{r Is office name actually obsolete since 2019?}
TEST<- NW.Mex_Landings_georeferenced %>%
  filter(year== 2022)

table(TEST$office_name_clean)

# Short answer: no
```


## NOTE: Flag these in the lat-long key as not in the Gulf [UNDER CONSTRUCTION May 19 2026]
office_name== c("EL ROSARIO", "ENSENADA", "TIJUANA")